% Tenth Philippic --- headnote
% philippic-10, February 43 BC, Rome

\textit{Delivered in the Senate in February 43 BC, the Tenth
Philippic is a speech of confirmation and of duel. A despatch had
arrived from Marcus Brutus --- styled in the formal decree by his
adoptive name, Quintus Caepio Brutus --- reporting that without
authorisation he had taken Macedonia, Illyricum, and all Greece
into the power of the Senate and the Roman people: he had drawn off
the legions that Gaius Antonius (the proconsul's brother) had meant
to seize, won over the cavalry bound for Syria, and received the
surrender of Dyrrachium from Publius Vatinius. The consul Pansa
convened the Senate at once, read the letter, and praised Brutus.
Cicero rises to second him --- and to answer the one ex-consul who
had spoken against confirming Brutus's command, Quintus Fufius
Calenus, the most persistent of Cicero's senatorial adversaries
through the whole Philippic season and, as Cicero needles him,
father-in-law of the very consul presiding.}

\textit{The argument turns on a single principle, stated flatly:
all the legions, all the forces that exist anywhere, belong to the
commonwealth, and a man forfeits every right to army and command
the moment he turns them against the state. By that rule Brutus's
self-raised army is no more suspect than the consular armies of
Hirtius, Pansa, Octavian, and Decimus Brutus, all praised for the
same cause; and Antony, who uses Roman troops against Rome, has no
claim on any of them. Cicero meets the objection that the veterans
distrust the name of Brutus the tyrannicide by widening the ground
beneath their feet: it is not the veterans alone who will fight for
liberty, and an order that lets itself be steered ``at the nod of
the veterans'' has already surrendered. From that comes the famous
central passage (\textsection 19--20), the hardest statement in the
series of the creed that death is preferable to slavery --- ``life
does not consist in breath; rather, there is none at all for one
who is a slave'' --- and that other nations may endure servitude
but Rome, bred to refer everything to dignity and virtue, cannot.}

\textit{The speech is laced with the running quarrel with Calenus:
the mockery of his praise of Brutus's letter as ``well and properly
written'' (praise, Cicero says, of Brutus's clerk, not Brutus), the
appeal to his pride in a promising son, the advice to talk less
with himself and more with his wise son-in-law. After a rapid
muster of Brutus's lieutenants --- Quintus Hortensius, who held
Macedonia for him; the younger Cicero, to whom Piso's legion gave
itself up; Gnaeus Domitius; Marcus Apuleius --- and a contemptuous
glance at Antony's hangers-on (Lucius Antonius, Trebellius restored
to favour by a cancellation of debts, Plancus Bursa, the squatters
Saxa and Cafo with their mime-actor neighbours on the Campanian
land), the speech closes, as the deliberative Philippics do, in a
formal motion (\textsection 25--26): a vote of thanks to Brutus, a
grant confirming him in command of Macedonia, Illyricum, and Greece
with power to raise money, contract loans, requisition grain, and
keep his forces close to Italy, and a parallel commendation
continuing Hortensius in the province until a successor is named.}
