% Fourteenth Philippic --- headnote
% philippic-14, 21 April 43 BC, Rome

\textit{Delivered in the Senate on 21 April 43 BC, the Fourteenth
Philippic is the last of the series and the last surviving speech of
Cicero. Dispatches had arrived from the consuls Hirtius and Pansa and
from the young Caesar reporting the victory at Forum Gallorum (14
April), where Antony's assault on Pansa's relief column had been thrown
back and Hirtius's fresh legions had turned the day into a rout. Rome,
which had put on the war-cloak for the peril of Decimus Brutus, was
minded to celebrate; a motion stood to resume the toga at once. Cicero
opens by resisting it. The war is not finished, he argues, while
Decimus Brutus still lies besieged in Mutina; the completion of the
war is the safety of Decimus Brutus, and the return to civilian dress
must be reserved for the victory complete, not spent on a battle,
however glorious, that has not yet raised the siege.}

\textit{The speech then turns on a single insistence: that the Senate
stop short of half-measures and name the thing it is fighting. It is
decreeing a thanksgiving while refusing to call Antony a public enemy;
but a thanksgiving for slaughter in a civil war is unheard-of --- not
Sulla, not Cinna, not Caesar after Pharsalus ever asked one --- so to
decree it at all is already to judge the defeated enemies. Cicero
presses the logic home: he moves to enlarge the thanksgiving to fifty
days, an unprecedented number, and to name Hirtius, Pansa, and Caesar
\textit{imperatores}, which by that very title brands Antony's men, the
fallen and the surviving alike, as enemies of the state. Around this
run the supporting movements: the savagery of Lucius Antonius at Parma,
set beside Hannibal and the Carthaginians; and a personal digression
(sections 14--16) in which Cicero rebuts a rumour, spread by the
Antonian faction in the city, that on this very day, the Parilia, he
meant to come down to the Forum with the fasces and make himself
tyrant --- a plot, he shows, to raise a hired mob against him and open
the way to a massacre.}

\textit{The peroration (sections 31--38) is the height of the speech
and one of the great funeral orations in Latin: the honours for the
soldiers of the Martian and Fourth legions who fell in the battle.
Cicero proposes the most ample monument that can be built, an altar of
valour to stand unto the memory of eternity, and equal rewards for the
parents, children, wives, and brothers of the dead; and he consoles the
bereaved with the immortality their kin have won --- \textit{O fortunate
death, which, owed to nature, was paid above all for the fatherland};
\textit{O happiest Sun, who before he set saw Antony fleeing with his
few}. The formal motion closes the speech and the Philippics together.
The hope is poignant in hindsight: Cicero hails the three commanders
and prays the gods to avert death from the wounded Pansa, but within
days both consuls would be dead --- Hirtius fallen, Pansa dead of his
wounds --- and the consular armies would pass to the young Caesar,
whose road thereafter led away from the Senate. Before the year was
out Cicero himself was proscribed and killed.}
