% Thirteenth Philippic --- headnote
% philippic-13, 20 March 43 BC, Rome

\textit{Delivered in the Senate on 20 March 43 BC, the Thirteenth
Philippic answers a fresh overture of peace. The war at Mutina was
still undecided: Decimus Brutus lay besieged, the consuls Hirtius and
Pansa and the young Caesar were in the field against Antony, and the
proposed second embassy had been abandoned (the subject of the Twelfth
Philippic). Now the question of an accommodation was revived from
another quarter. Marcus Aemilius Lepidus --- governor of Nearer Spain
and Narbonese Gaul, pontifex maximus, lately the architect of a
bloodless settlement that had brought the younger Pompeius back into
peace --- had sent the Senate a dispatch urging concord. The motion
before the house was Servilius's, on Lepidus's letter; Cicero rises to
assent to it, and to turn Lepidus's own authority against the peace he
counsels.}

\textit{The speech falls into three movements. In the first (sections
1--21) Cicero argues that no peace is possible with the Antonian
faction: a peace would be not peace but a charter of slavery, for the
Senate's own decrees against Antony would have to be unmade, and the
men one would have to take back are past redemption --- the long roll
of Antony's wrecked and bankrupt partisans is paraded with contempt.
The praise of Lepidus is woven through it, and with it the case for
the young Magnus Pompeius, son of Pompey the Great: restored by
Lepidus, he is to recover his father's house, his patrimony, and his
father's place in the augural college, which Cicero himself promises
to confer.}

\textit{The central and most celebrated movement (sections 22--48) is
a piece of forensic theatre. Cicero produces a letter Antony had
written to Hirtius and the young Caesar --- forwarded to him by
Hirtius --- and reads it out clause by clause, answering each with
ridicule and indignation. The letter convicts its author from his own
mouth: it confesses the murderous pact with Dolabella to torture and
kill Trebonius, Brutus, and Cassius; it gloats over Trebonius's death;
it abuses Hirtius, the ``boy'' Caesar, and Cicero himself (the
``trainer of gladiators''); and it coins, out of Antony's ``divine
piety,'' the superlative \textit{piissimus}, a word, Cicero notes,
that does not exist in the Latin tongue. By the close Cicero has used
Antony's own confessions to lay every parricide open to view. The
peroration (sections 49--50) refuses all reconciliation --- sooner
shall fire mingle with the waves than the commonwealth and the Antonii
be friends --- ranks the three possible ends (to conquer is best, to
die for liberty next, to live in disgrace worst), and closes with the
formal motion: assent to Servilius on Lepidus's dispatch, and a
further decree commending the younger Pompeius for offering his
service and his men to the Senate and people.}
