% Ninth Philippic --- headnote
% philippic-9, 4 February 43 BC, Rome

\textit{Delivered in the Senate on 4 February 43 BC, the day after
the Eighth Philippic, the Ninth is a speech in an altogether
different key: not invective but eulogy. The embassy Cicero had
opposed in the Sixth and Seventh Philippics --- three legates sent
in early January to carry the Senate's terms to Antony before
Mutina --- had come back with nothing but Antony's defiance; and
its senior member did not come back at all. Servius Sulpicius
Rufus, consul of 51 BC and the most celebrated jurist of the age,
had undertaken the winter journey already gravely ill, and died
near Antony's camp. On 4 February the consul Pansa proposed public
honours for him and spoke at length in his praise. Cicero rises to
second the motion --- and to answer the one man who had spoken
against its fullest form.}

\textit{That man was Publius Servilius Vatia Isauricus, who held
that the honour of a statue belonged only to an envoy killed by
the sword while on embassy, and who proposed a public tomb but no
statue. Cicero's answer is the legal heart of the speech, and it
is argued like the jurist it commemorates: our ancestors, he says,
looked to the cause of death, not its manner. The man for whom the
embassy itself proved fatal earned his monument as surely as one
cut down by violence, for the point of the honour was to make men
undertake dangerous embassies more boldly. He marshals the
precedents --- the four envoys whom Lars Tolumnius, king of Veii,
slew at Fidenae, and Gnaeus Octavius, murdered at Laodicea on an
embassy in which no danger had been foreseen, both honoured with
statues on the Rostra --- and turns the argument against Antony
himself: the man who was the cause of death was the killer, so that
Antony slew Sulpicius no less than Leptines slew Octavius. The
statue, Cicero insists, will stand as a permanent witness that the
war was grave enough to cost a legate his life, and an everlasting
testimony to the embassy Antony spurned.}

\textit{The speech is also an elegy for a friend. Cicero recalls
Sulpicius's reluctance, his pleading of ill health, the moment
when --- urged on by Pansa, and unwilling to set his own life
above the Senate's authority --- he drew his son and Cicero aside
and consented to go; his parting words, Cicero says, seemed an omen
of his fate. There is even a connoisseur's note in the honour
proposed: recalling Sulpicius's love of old Roman restraint and his
scorn for the extravagance of the age, Cicero decrees a statue of
bronze and on foot --- \textit{pedestris} --- rather than the
gilded equestrian kind first set up for Sulla, ``as though by his
own authority and will.'' He keeps Servilius's public tomb and adds
the statue to it, against the rule that statues go to many and tombs
to few. The speech closes, as the deliberative ones do, in a formal
motion (\textsection 15--17): the bronze pedestrian statue on the
Rostra, a reserved space of five feet for the family at the games,
an inscription recording that he died for the commonwealth, a
magnificent public funeral, and a public tomb thirty feet square in
the Esquiline field for Sulpicius and his line. It is the gentlest
speech of the series, and the only one given wholly to the dead.}
