% Fourth Philippic --- headnote
% philippic-4, 20 December 44 BC, Rome

\textit{The popular counterpart to the Third Philippic, delivered
to the people in the Forum on the very same day --- 20 December
44 BC --- only hours after the senatorial speech. The order is
fixed: in the morning the tribunes of the plebs convened the
Senate to provide a guard for the consuls-elect, Pansa and
Hirtius, who would take office on the Kalends of January; Cicero
addressed that meeting (the Third Philippic) and carried his
motion that Decimus Brutus and the Gallic municipalities, the
Martian and Fourth Legions and their officers Lucius Egnatuleius
and the rest, and Octavian himself, be commended for what they
had done against Antony. The Senate dispersed; Cicero went down
into the Forum and addressed the crowd that had gathered there.
This second speech is what they heard.}

\textit{It is short --- only sixteen sections --- and a single,
sustained announcement: that the Senate has just decreed Antony
an enemy in fact, though not yet in name (\textit{est hostis a
senatu nondum verbo appellatus, sed re iam iudicatus Antonius},
\textsection 1). Cicero builds the speech around that one
proposition. \textsection 1--6 set out the inferential chain:
if armies have rightly been raised against Antony, he is an
enemy; the Senate has just praised those who raised them
(Octavian, Egnatuleius, the legions); therefore by inescapable
inference the Senate has judged him an enemy --- and the
Martian and Fourth Legions, by deserting him for Octavian and
taking their stand at Alba, judged him an enemy before the
Senate did. \textsection 7--9 add the converging judgments of
Decimus Brutus, the province of Gaul, all Italy, and the people
themselves: every body that matters has reached the same
verdict. \textsection 10 closes the demonstration with the
gods --- prodigies and a universal consent which cannot have
been without divine prompting --- and \textsection 11--16 are
the hortatory close: Antony is no longer a wicked man but a
beast in a pit (\textit{immani taetraque belua quae, quoniam in
foveam incidit, obruatur}, \textsection 12); the Roman people,
conqueror of all nations, has its whole contest now with a
brigand and a Spartacus. The speech ends, characteristically,
on Cicero's own role: with Marcus Servilius bringing the
motion and Cicero, after a long withdrawal, acting as proposer
and leader (\textit{me auctore et principe}), the city has
caught fire again with the hope of liberty.}

\textit{The register is louder and more public than the Third
Philippic. The Senate had been addressed in measured legal
language --- the technical word \textit{hostis} held back, the
honours framed as a motion. The Forum is addressed at a higher
pitch of indignation: Antony is a plague (\textit{pestis}), a
beast (\textit{belua}), an assassin (\textit{percussor}), a
brigand (\textit{latro}), a Spartacus. The Tarquin material of
the Third Philippic is dropped; the comparison with Catiline
takes its place (\textsection 15), and is sharpened into a
boast: Catiline at least raised an army from nothing, whereas
Antony has lost the army he was given. The thirty-nine
sections of senatorial argument are compressed into sixteen
sections of public address, and the senatorial conclusion ---
that what the Senate has done amounts to the judgment of
\textit{hostis} --- is announced to the people as already
decided.}

\textit{Both speeches were carried, and the practical effect
was immediate: Octavian and the legions were now operating with
formal senatorial sanction; Decimus Brutus's defence of Cisalpine
Gaul was the Senate's defence; the siege of Mutina, which Antony
was even now beginning, was a siege against Rome. The formal
\textit{hostis} declaration would come only at the end of April
43, after Forum Gallorum and Mutina; for the present it was
held back, by the calculation Cicero makes explicit in
\textsection 1 --- the foundation has been laid; the rest will
follow.}
