% First Philippic --- headnote
% philippic-1, 2 September 44 BC, Rome

\textit{Delivered in the Senate on 2 September 44 BC, the first
of the fourteen orations against Mark Antony to which Cicero
himself, half-jocularly in a letter to Atticus, would lend
Demosthenes's title --- \textit{Philippics}, after the
Demosthenic speeches against Philip of Macedon. Antony had
summoned the Senate to vote a public \textit{supplicatio} of
thanksgiving in honour of the dead Caesar's memory, a measure
Cicero would not attend and could not in conscience support.
The speech recounts in its opening sections (\textsection 1--10)
Cicero's reluctance to come up to the City at all: his abortive
trip toward Greece in July (he had reached Leucopetra, on the
toe of Italy, before contrary winds drove him back), and his
return at the news, brought to him at Regium, that Brutus and
Cassius had issued a conciliatory edict and that the September
Kalends Senate would be calm. Hearing en route that Antony had
in fact convened a hostile Senate on 1 September and demanded
the consular Cicero's attendance with threats to demolish his
house, Cicero arrived a day late and on the morrow rose to
speak. The speech is the opening of the campaign that would
consume the last year and a half of his life and end in his
proscription on the Triumviral lists in November 43 and his
death at Formiae on 7 December.}

\textit{Its argument has a curiously double character. Cicero
addresses Antony as a friend who has gone astray, and is
generous with credit where it can be given: in \textsection
3--4 he praises Antony's abolition of the dictatorship as the
genuinely free act it was (``a kind of light seemed to have
been let in upon us, now that not only the kingship that we
had endured but even the fear of kingship had been lifted
away''); \textsection 31--32 dwell on the brilliance of
Antony's speech in the temple of Tellus on 17 March, on the
embassy of his small son sent up into the Capitol as a pledge
of peace. But \textsection 16--26 are the indictment proper:
Antony's measures of the summer, carried under the cloak of
Caesar's \textit{acta}, are not Caesar's acts at all but
forged authority --- the \textit{tertia decuria} that opens
juries to centurions and to ordinary legionaries of the
Larks, the appeals-law that empties the standing courts for
\textit{vis} and \textit{maiestas}, the colonial commissions
and exemptions. \textsection 27 fixes the speech's terms: if
Antony will let Cicero speak freely, well and good; if not,
let him at least be angry as a citizen rather than as an
armed man. \textsection 33--38 are the moral case --- that to
be feared rather than loved is \textit{invidiosum,
detestabile, imbecillum, caducum}, that no man is happy who
lives on terms by which he can be killed with glory to his
killer (the \textit{exitus C.~Caesaris}); and the apostrophic
\textit{O beatos illos} of \textsection 36 turns, with grief,
on Brutus's absence at his own praetorian games. The savage
personal invective for which the \textit{Philippics} would
become famous is reserved for the never-delivered Second
Philippic, written as a pamphlet in October and circulated
later in the year. The First Philippic, alone of the
fourteen, still keeps the form of an appeal.}
