% Second Philippic --- headnote
% philippic-2, October 44 BC, Rome

\textit{Cicero's most famous invective, and one of the
masterpieces of Latin oratory. The Second Philippic was never
spoken aloud. After the First Philippic of 2 September 44 BC,
Antony withdrew to his villa at Tibur and on 19 September
returned to the Senate with a prepared speech of seventeen
days' composition, savagely attacking Cicero (who, warned in
time, did not attend). Cicero answered with this pamphlet,
written through October at Puteoli, sent in draft to Atticus
for criticism on 25 October (\textit{Ad Atticum} 15.13a, 16.11),
and circulated through trusted hands during November --- but
withheld from publication until after Antony's break with the
Senate later that winter. The fiction of the speech is that it
is being delivered in the Senate on 19 September in immediate
reply to Antony; the form (the consular replying to a violent
attack, with all rhetorical resources unleashed) is modelled on
Demosthenes's \textit{On the Crown}, to which Cicero refers
ruefully in the letters of these weeks. The title
\textit{Philippic}, which Cicero himself had begun to use in
half-jest, fixes the analogy.}

\textit{The speech falls into two great movements. \textsection
1--41 are the defence, refuting Antony's individual charges in
order: the augural canvass and Brundisium ``mercy'' (1--10), the
consulship and the suppression of Catiline (11--20), the Ides of
March charge that Cicero counselled the killing (21--36) ---
which Cicero half-disclaims and half-embraces, in the great
\textit{O praeclarum illud convivium} of \textsection 32 (the
``Trojan-horse'' acceptance among the conspirators) and the
\textit{stilus}-as-stylus-and-dagger joke of \textsection 34
(``had that pen been mine that men say was, I should have
finished not only the one act but the whole play''). \textsection
37--41 dispatch the camp-of-Pompey and legacies charges, with
the brilliant Turselius joke. \textsection 42--58 are the
counter-attack: the savage Curio episode and the \textit{toga
muliebris}/\textit{stola} metaphor (44), the
Alexandria-with-Gabinius episode (48), the seed-of-civil-war
argument with the famous Helen-of-Troy comparison
(\textit{Ut Helena Troianis, sic iste huic rei publicae belli
causa}, 55), and the lictors-and-mime tour of the Italian
municipalities (57--58).}

\textit{\textsection 59--118 are the indictment proper: Antony
in the Caesarian war and after. \textsection 59--63 the
Brundisium return and second Italian tour, ending in the famous
Hippias-wedding vomiting scene (the master of horse filling his
own lap and the whole tribunal \textit{frustis esculentis vinum
redolentibus}). \textsection 64--69 the purchase of Pompey's
property at the post-Pharsalia auction --- the indignant
\textit{O audaciam immanem!} apostrophe, the \textit{Charybdis
tam vorax} cadence, the slaves' beds spread with Pompey's
purple coverlets. \textsection 70--78 the master-of-horse year,
the sham auction list, the comic Narbo-and-back midnight ride
in Gallic slippers and a cloak with the lover's letter for
Fulvia. \textsection 79--84 the false-augury Dolabella set-piece
--- the augur announcing \textit{Alio die} after the comitia
had run their course. \textsection 85--87 the great Lupercalia
scene: \textit{sedebat in rostris collega tuus amictus toga
purpurea, in sella aurea, coronatus}, and Antony offering the
diadem to the lamentation of the people. \textsection 88--92
the aftermath of the Ides --- Antony's flight on 15 March, the
funeral oration that loosed the mob, the magnificent abolition
of the dictatorship --- shadowed at once by the huckstering of
immunities and citizenships from the records at the temple of
Ops. \textsection 93--99 the forged Caesarian decrees (the
Deiotarus syngrapha, the freeing of Crete from tribute, the
exile-recalls), with the savage \textsection 99 attack on
Antony's treatment of his uncle Lucius Caesar and the sister
who was Antony's first wife. \textsection 100--108 the journey
through Italy with mimes and gamblers on the Campanian and
Leontine land, the desecration of Varro's villa at Casinum, and
the return to Rome with armed guards in the Forum. \textsection
109--112 demand to know on what religious grounds Antony, as
Caesar's flamen, neglects the praetexta on Caesar's festival
days, why armed Ituraeans are in the Forum, why the doors of
Concord are not open. \textsection 113--118 are the great
peroration: the example of the Liberators left for imitation,
the Roman people's young defenders, Antony's wife and her
\textit{tertia pensio}, and the closing offer of Cicero's own
body --- \textit{defendi rem publicam adulescens, non deseram
senex; contempsi Catilinae gladios, non pertimescam tuos} ---
sealed by the prayer \textit{ut moriens populum Romanum liberum
relinquam}.}

\textit{The speech was prophetic. Within fifteen months, on 7
December 43 BC, Cicero was overtaken at his villa at Formiae by
Antonian assassins and killed; his head and the hands that had
written the \textit{Philippics} were nailed to the Rostra in
the Forum. Of the surviving works of Latin prose, none has
more directly cost its author his life.}
