% Pro Rege Deiotaro --- headnote
% pro-rege-deiotaro, November 45 BC, Rome

\textit{Delivered before Julius Caesar in November 45 BC, the
third and last of Cicero's three so-called ``Caesarian''
speeches --- after \textit{Pro Marcello} in the Senate
(September 46) and \textit{Pro Ligario} in the Forum (late
November 46) --- and the strangest in its setting of the three.
The case was not heard in the Forum. Caesar, now Dictator for
the fourth time and recently returned from the Spanish campaign
against the younger Pompey, sat as sole judge in his own house
on the Palatine, with no public audience and no \textit{conventus
oratorum}; Cicero felt and made plain his unease at speaking
``within the walls of a private house'' (\textsection 5). The
defendant himself was not present. King Deiotarus of Galatia, an
old Roman client of long standing whom Pompey had decorated and
whom Caesar had visited at the royal estate after Zela in 47 BC,
had been charged with having tried to murder Caesar on that very
visit --- with armed men hidden in a room at Blucium where the
king's gifts had been laid out, and a second alleged ambush at
the bath. The accusers were two: Castor, the king's own
grandson, son of Deiotarus's daughter and the Tarcondimotid
Saocondarius, who hated his grandfather over a tetrarchal
inheritance dispute; and Phidippus, a runaway slave from the
king's household, the king's own physician, sent to Rome with
the royal embassy and there suborned by Castor.}

\textit{Cicero attacks the case on two fronts at once. First,
the prosecution itself is a monstrosity (\textsection 3): by
ancestral Roman law a slave may not even be examined under
torture against his master --- and here is a slave, untortured,
accusing his master in open court. Second, the alleged plot is
psychologically impossible: Deiotarus is a man of singular
\textit{frugalitas} (\textsection 26), too cautious to attempt
murder by force when poison would have been easier (\textsection
18), too sane to send the supposed accomplice to Rome
afterwards (\textsection 21), and bound to Caesar by hospitality
sealed at his own household altars. The argument from
\textit{ius hospitii} runs throughout: in \textsection 8 Cicero
swears Caesar to mercy ``by that right hand which as guest you
stretched out to the guest King Deiotarus,'' and in
\textsection 15 he piles up the impossibilities of killing a
\textit{hospes} in the sight of the household gods. \textsection
13 contains the most quoted line: Deiotarus came to Pompey
\textit{ut ad fugientem, non ut ad insequentem} --- ``as to one
fleeing, not as to one pursuing'' --- so that he joined a
partnership of peril, not of victory. The dancing-naked charge
(\textsection 26) Cicero ridicules into a passing meditation on
the kingly virtues, and the Blesamius letters (\textsection
33--34) he reduces to ``the little spiteful gossip of the
town,'' adding the famous demurral that no one has fallen in
Caesar's victory except in arms.}

\textit{The peroration (\textsection 40--43) drops forensic
attack altogether and turns frankly to clemency, setting before
Caesar the figure of two kings, father and son, whose royal
title hangs on this single private hearing. The four embassy
chiefs --- Hieras, Blesamius, Antigonus, Dorylaus --- offer
their own bodies for their kings, and Hieras goes so far as to
swear that he never left Caesar's side during the visit and to
volunteer himself as defendant in Deiotarus's place. The verdict
is not recorded; Caesar seems to have postponed judgement, and
the matter was overtaken five months later by the Ides of March.
Of the three Caesarian speeches this is the most subdued and the
most plainly forensic --- short on \textit{clementia}-encomium,
long on cross-examination --- and it carries through it the
discomfort of a speech delivered in conditions Cicero had never
known before and would not see again.}
