% Pro M. Marcello --- headnote
% pro-marcello, September 46 BC, Rome

\textit{Delivered in the Senate in September 46 BC, the first
of Cicero's three so-called ``Caesarian'' speeches (with
\textit{Pro Ligario} and \textit{Pro Rege Deiotaro}) and his
first public utterance in the Senate after Pharsalus. The
occasion was Caesar's pardon, in answer to a Senate motion
led by L. Calpurnius Piso (Caesar's father-in-law) and
Marcellus's cousin C. Marcellus, of M. Claudius Marcellus,
consul of 51 BC and one of the most stubborn Pompeian
senators --- the man who in his consulship had moved to
recall Caesar from Gaul and had famously had a Transpadane
of Comum publicly flogged to spite Caesar's grant of Latin
rights. After Pharsalus Marcellus had withdrawn to Mytilene
and refused to return; the pardon, given in his absence and
unexpectedly, broke the long silence (\textit{diuturnum
silentium}) Cicero himself had observed in the Senate since
the war's end, and gave him the occasion to speak again in
his old manner.}

\textit{The speech is an extended encomium of Caesar's
\textit{clementia}, but its architecture is more
philosophically dense than that label suggests. The opening
period (\textsection 1) piles up four \textit{tantum} and
\textit{tam} clauses on Caesar's gentleness, clemency,
moderation, and wisdom --- the four virtues the whole
oration will work through. \textsection 5--9 set Caesar's
military victories against the higher praise of moral
victory: there is no force ``so great that it cannot be
weakened and broken by iron'' (\textsection 8), but to
conquer one's own spirit, to lift up a noble adversary, is
the deed Cicero judges \textit{simillimum deo} --- ``most
like a god.'' \textsection 11--12 build to the speech's
famous antithesis: \textit{ipsam victoriam vicisse videris}
--- ``you seem to have conquered victory itself'' --- by
giving back to the conquered what victory's own right would
have taken from them. \textsection 23 charges Caesar with
the unfinished work of restoration: the courts to be set
up, public faith restored, lusts checked, laws renewed.}

\textit{\textsection 25--32 turn to the still-unsettled
commonwealth and the appeal that Caesar guard his own life
since the state's life depends on it. Cicero presses against
Caesar's reported saying that he had ``lived long enough for
nature or for glory'': enough perhaps for nature, perhaps
even for glory, but for the country ``certainly too little''
(\textsection 25). The \textit{vita} that matters is not the
one ``contained in body and breath'' but the one that lives
in posterity's memory (\textsection 28); and \textsection 27
locates this remaining work in Caesar's own
\textit{mens et ratio et consilium} --- ``mind and reason
and counsel.'' The closing \textsection 32 has the senators
pledge not only watches and guards but ``the interposing of
our own flanks and our own bodies'' to keep him alive;
\textsection 34 closes the ring opened by the first words
with a final personal thanks.}

\textit{Modern critics have read the speech with some
discomfort. Taken as flattery it can sound servile, the
republican consul of 63 BC bending the knee to the
dictator. But the encomium is also an unblinking statement
of the political condition --- that the commonwealth's life
now hangs from one mortal breath --- and an unmistakable
charge: that until Caesar settles the constitution
(\textsection 23, 27, 29) the work for which his
\textit{clementia} buys time is not done. The dignified
architecture and the insistence on the unfinished business
of the \textit{res publica} complicate the verdict of
servility, and Cicero himself in \textit{Ad Familiares}
treats the speech as a calculated public undertaking, not a
capitulation.}

\textit{Marcellus did not enjoy his pardon long. On the
journey home, in May 45 BC, he was murdered at the Piraeus
by a client, Magius Cilo --- the bitter sequel reported in
Servius Sulpicius Rufus's letter to Cicero (Fam.~4.12) and
already covered in this project.}
