% Pro Milone --- headnote
% pro-milone, 52 BC, Rome

\textit{For T. Annius Milo, written for the trial held
at Rome in April 52 BC. On 18 January of that year Milo
and P. Clodius Pulcher, the two great street-captains of
the late Republic, had met by chance on the Appian Way
near Bovillae; a quarrel between their armed retinues
ended with Clodius wounded, carried into a roadside inn,
and then finished off at Milo's order. Clodius's
followers brought the body back to Rome, and the mob
cremated it inside the Senate house, burning the Curia
Hostilia to the ground. In the disorder that followed
the Senate turned to Pompey, who was made sole consul
(\textit{consul sine collega}) and given charge of
restoring order. Milo was tried before a special court
constituted under Pompey's new law on public violence,
the lex Pompeia de vi; the prosecution was led by the
young Appius Claudius, Clodius's nephew, with Mark
Antony among the accusers.}

\textit{Cicero does not deny the killing. His defence
turns the case on its hinge: not whether Milo killed
Clodius, but whether the killing was lawful --- and a
killing in self-defence against an ambush, he argues, is
sanctioned by a law older than any statute, the unwritten
and inborn law of nature (\textsection\textsection
10--11), from which comes the speech's most famous
sentence, that amid arms the laws fall silent. The
narrative (\textsection\textsection 24--31) labours to
show that it was Clodius who lay in wait for Milo, not
the reverse; the argument from motive presses the
Cassian \textit{cui bono} --- to whose benefit was the
death? --- and answers that Clodius's whole career of
violence made his removal a deliverance for the
commonwealth. Only after the legal and political ground
is won does Cicero turn to the emotions: Milo's refusal
to beg, his own grief and the memory of his exile, and
the closing apostrophe to the land that would be blessed
to receive so brave a man.}

\textit{The speech as we have it was never delivered in
this form. Pompey had ringed the Forum with soldiers,
and the hostile, armed setting --- the strange new look
of a court that Cicero describes in the opening
sections --- unnerved him; the version he managed to
speak fell short, and Milo was convicted, by thirty-eight
votes to thirteen, and went into exile at Massilia. What
survives is the oration Cicero afterward wrote up as he
wished he had given it, and it became at once a model of
the art. The ancient anecdote has Milo, reading it in
exile, remark that he was glad Cicero had not spoken so
well in court --- for otherwise he would not now be
enjoying the excellent red mullet of Massilia.}
