% In Pisonem --- headnote
% in-pisonem, August/September 55 BC, Rome, in the Senate

\textit{In Pisonem, delivered in the Senate in August or
September 55 BC: the most violent invective in the surviving
Ciceronian corpus, after Lucius Calpurnius Piso Caesoninus
returned from his Macedonian governorship and dared (in a
senatorial speech against Cicero) to defend his record.
Piso had been one of the two consuls of 58 BC --- the year of
Cicero's exile --- and Cicero had paid for that exile, or so
he was now persuaded, with the bribery of the consular
provinces. The pent-up hatred of nine years pours out here.}

\textit{The opening (the original first column or so) is lost
and survives only in the fragments preserved by Asconius and
the late grammarians: the great cry \textit{pro di immortales!
qui hic inluxit dies}, the savage caricature of Piso's mother
as Insubrian Gallic stock and his maternal grandfather as a
huckster of Placentia, the line on the family carriage and the
brutal \textit{te beluam ex utero, non hominem fudit}. From
\textsection 1 the continuous text begins with the famous
\textit{iamne vides, belua}: Piso's gloomy dishonest face, the
deception of the consular electorate. \textsection\textsection
2--7 are the great \textit{me-vs-tibi} parallel: Cicero's
consulship of 63 BC --- the agrarians, Catiline, the Allobrogan
arrests, the closing of the year with Catulus's salutation
\textit{parens patriae} --- contrasted clause-for-clause with
Piso's of 58 BC, the year that began with Clodius's repeal of
the lex Aelia Fufia, the restoration of the corrupt collegia,
the abolition of the censorial \textit{notio}. \textsection
8--72 work through the year of exile: Piso's complicity with
Clodius, the abandonment of the Senate, the reception of the
exiled consul into the Macedonian command, and Cicero's own
reception by the cities on his return.}

\textit{From \textsection 73 the second great theme opens: the
moral and physical disgrace of Piso's eastern command and his
philosophical pretensions --- the Greek Epicurean Philodemus
in his entourage, the dancing-girls, the Macedonian frontier
disasters, the ruinous occupation of the Dyrrachian and
Apollonian provinces. \textsection\textsection 89--99 close on
the contrast with Pompey's eastern triumphs and the future
judgment of history. The speech is the political twin of
\textit{Post Reditum} (57 BC) and \textit{De Provinciis
Consularibus} (mid-56 BC); read with those, it is the closing
panel of Cicero's three-part settlement of the exile-year
account.}
