% Pro L. Cornelio Balbo --- headnote
% pro-balbo, summer 56 BC, Rome

\textit{For L. Cornelius Balbus, delivered at Rome in the
summer of 56 BC. Balbus, born at Gades (Cadiz) in
Hispania Ulterior, had served Pompey in the Sertorian
war and been granted Roman citizenship by Pompey under
the lex Gellia-Cornelia of 72 BC, which authorized the
imperator to grant the franchise on the advice of his
consilium. He had become a close intimate of Caesar (a
quaestor on Caesar's staff in Spain in 61 BC, and prefect
of engineers in his praetorship and consulship), and at
the time of this trial had been one of the lubricators
of the Luca conference of April 56 BC. The prosecution
was brought under the lex Papia of 65 BC (which had
established a quaestio on usurped citizenship); the
prosecutor's case --- the only charge --- was that Balbus
was born at Gades, a federate state, and that the
Gaditan treaty either forbade or did not allow the
transfer of citizenship.}

\textit{The defence was the post-Luca front in court
form: M. Crassus opening, Pompey speaking second
(``yesterday's speech,'' \textsection 2--4, the long
praise of which is the speech's opening figure), Cicero
closing. The legal architecture (Cicero's brief at
\textsection 17) is to leave to Crassus and Pompey the
detailed legal exposition and to give the closing in
the form of a defence not so much of Balbus as of
Pompey's act --- and through Pompey of the whole
generals'-citizenship-grant tradition that runs from
Marius and Sulla through to Caesar and the Caesarian
governors of his own command.}

\textit{The argument falls into three movements.
\textsection 5--17 sets out the core dilemma: the only
charge is the act of Pompey himself, and either Pompey
acted in ignorance (impossible: he is the man of
incredible authority and breadth of experience whose
footprints can be heard, as Cicero says at \textsection
13, in every shore and harbour of the empire) or he
acted knowingly (in which case who would dare to say
he had violated treaty?). \textsection 18--37 is the
properly legal section: the prosecutor has misunderstood
the technical word \textit{fundus} (a federate or
Latin people becomes \textit{fundus} when it confirms a
Roman law within its own jurisdiction --- a function of
the people's right, not of the federate-vs-free
distinction); the right of changing citizenship is a
two-way street that runs from us to them and from them
to us; the lex Gellia-Cornelia's ``\textit{si quid
sacrosanctum est}'' exception cannot bind because no
sacrosanct condition was ever Roman-vote-confirmed in
the Gaditan treaty; the closing fine reading at
\textsection 35--37 of the Gaditan treaty's
\textit{maiestatem populi Romani comiter conservanto}
clause (\textit{comiter} means kindly, not communiter
--- and the very phrase shows that the Gaditans are the
inferior party).}

\textit{\textsection 38--55 catalogues the historical
practice. C. Marius is the foundational example
(\textsection 46--49: M. Annius Appius the Iguvian, two
whole cohorts of Camertes, the famous defence of his
own grants by Marius in person under the Licinian-Mucian
quaestio). The roll-call at \textsection 50--51 covers
Pompey's father at Ravenna, P. Crassus the Heraclean,
Sulla the Massiliot, Q. Metellus Pius the Saguntine, M.
Crassus the federate from Aveia, and finally Pompey's
own grants (the Gaditan Hasdrubal, the Mamertines, the
Uticans, the Saguntine Fabii). Section 51 is the famous
quotation from Ennius's \textit{Annales}: ``\textit{the
man who shall strike the enemy will be Carthaginian to
me, whoever he is.}'' \textsection 52--55 give the
jurors' (the Mamertine-Cassius case under the Papian
law), the people's (the Latin-treaty precedent), and
the Senate's (the Greek priestess Calliphana of Velia)
parallels.}

\textit{The peroration \textsection 56--65 is the
political speech behind the legal one. \textsection
56--58: Balbus has no enemies; he is being attacked
because his friends are powerful. \textsection 59--62:
Pompey calls those enemies to fight him directly, not
through the company and the following. \textsection 61
is the most explicit surviving statement of Cicero's
post-Luca accommodation: ``some things we wished,
contended for, made trial of; they were not won''; the
fifteen-day supplicatio for Caesar in 57 BC, the army-
pay vote, the ten legates, the de-Provinciis-vote against
succession --- these are signal of the change of mind
the times required. \textsection 63--65 closes on
Caesar himself: do not let Balbus's intimacy with the
absent commander be his ruin; the empire's frontiers
are now at the bounds of his deeds. The verdict (we know
from no surviving source) was certainly an acquittal:
Balbus rose afterwards to the consulship of 40 BC, the
first man not of Roman or Latin birth to hold it.}
