% De Provinciis Consularibus --- headnote
% de-provinciis-consularibus, June 56 BC, Rome

\textit{On the Consular Provinces, delivered in the Senate at
Rome in June 56 BC. Under the lex Sempronia of C. Gracchus
the Senate fixed, before the consular elections, the two
provinces that the consuls of the next year would draw by
lot. Two were already in front of the house: the Gauls,
which Caesar held under the lex Vatinia of 59 BC and the
senatorial decree adding Transalpine Gaul; and Syria and
Macedonia, held by Gabinius and Piso, the two consuls of
58 BC who had stood by while Clodius drove Cicero into
exile. The chance of the year was that any combination of
these four provinces could be voted: take Gaul from Caesar,
or take Syria and Macedonia from Gabinius and Piso, or
some splitting motion. Cicero's vote, here delivered after
P. Servilius Vatia Isauricus had spoken before him, is for
Syria and Macedonia.}

\textit{The argument falls in three movements. \textsection
1--17 is the indictment of Gabinius and Piso: the harassed
Thessalonians, the lost army of Piso in Macedonia, the
auctioned Achaean tribute, the Byzantine maidens leaping
into wells to escape his lust, the plundered statues of
Greece; in Syria Gabinius the ``Semiramis,'' bought by
Ariobarzanes for murder, harrying Roman publicans into
slavery to Jews and Syrians, the consular thanksgiving
which the Senate refused him for letters of victory it
declined to credit. Both are fit to be torn out --- the
praetorian-province alternative will not do, since it can
be vetoed; only the lex Sempronia route is left.}

\textit{The pivot at \textsection 18 is the famous
about-face. The same speaker who had stood, two years
earlier, against Caesar's whole programme, now argues that
Caesar must be left in Gaul. The case is the strategic
state of the war (\textsection 31--35: Pompey has
pacified the eastern world; Caesar's exploits in Gaul go
beyond Marius's defence at Aquae Sextiae and Pomptinus's
containment of the Allobroges; the Alps no longer the
fortress they once were, the work nearly but not quite
finished, withdrawal would invite renewal); a long
sequence of historical exempla (\textsection 18--22) on
laying aside personal hostility for the commonwealth's
good --- Tiberius Gracchus the elder for Scipio Africanus's
brother, the Metelli for Marius in Gaul, M. Lepidus the
censor and M. Fulvius reconciled in the Campus, Q. Metellus
Nepos with Cicero himself --- and a personal narrative
(\textsection 40--44) of the long acquaintance between
Cicero and Caesar from their youth, Caesar's repeated
offers of partnership refused (the board of five, the
three consulars, a legateship of any honour), and the
critical fact established (\textsection 43) that Pompey
stands witness for Caesar that he wished Cicero recovered
from exile. The closing argument at \textsection 45--46
is the legal point that the same men who treat the Julian
laws as un-passed (no observation of the heavens) also
treat the lex Clodia against Cicero as lawfully passed: a
self-contradiction, since it was the same curiate-law
proceeding that made Clodius plebeian as had been used.}

\textit{The speech is the first public sign that the
optimate front of \textit{Pro Sestio} (March 56) had not
held. The conference of Luca, in mid-April 56, had reset
the terms of the triumvirate: Caesar's command in Gaul to
be extended for five more years, Pompey and Crassus the
consuls of 55 BC. Cicero's vote here for Caesar's
continuation, the second pillar of the post-Luca
realignment together with the abandoned attack on the
Campanian land settlement, is delivered in the language of
the senatorial-Caesarian common cause; but the rage of
the opening pages on Gabinius and Piso (whom Cicero would
in fact see prosecuted: Gabinius unsuccessfully two years
later, Piso never) is unmistakably authentic, and the
speech sustains both registers without breaking voice ---
the great trick of Cicero's middle period, when private
grievance and public reasoning are bound together by the
same rhetorical architecture. The thanksgiving for
Caesar's Gallic victories, fifteen days, twice the length
that had ever been voted before, was Cicero's own motion
\textsection 26--27.}
