% In Vatinium --- headnote
% in-vatinium, March 56 BC, Rome

\textit{Cicero against P. Vatinius, delivered at Rome in early
March 56 BC. The piece is not a free-standing oration but the
record of a cross-examination (\textit{interrogatio}) of a
hostile witness, conducted in the course of the \textit{Pro
Sestio} trial: Vatinius had appeared as one of the chief
witnesses for the prosecution against Sestius, and Cicero, after
delivering the \textit{Pro Sestio}, took him in hand. The form
shows in every paragraph: nearly every section opens with
\textit{quaero} or \textit{volo audire de te} ``I ask of you,''
``I want to hear from you,'' and the body is a rolling
catalogue of incriminating questions. The text we have has a
manuscript lacuna at the close of \textsection 4 (the
serpent-from-the-hole simile breaks off mid-clause); \textsection
5 resumes mid-sentence on the C. Cornelius defence of 65 BC.}

\textit{Vatinius is the great political bogey of Cicero's middle
years: tribune of the plebs in 59 BC, the chief instrument of
Caesar's consular legislation (the lex Vatinia gave Caesar his
five-year command in Cisalpine Gaul and Illyricum), and one of
the principal architects of the assault on the consul Bibulus.
Cicero takes him point by point through that tribunate ---
\textsection\textsection 14--29 are essentially a single sustained
indictment --- and through a few private scandals besides. The
opening (\textsection\textsection 1--10) sets the contemptuous
frame: Vatinius is not worth refuting as a witness, only worth
harrying; the parallel between the two men --- Cicero recalled
by the consensus of the city, Vatinius hated by it --- closes
with the demand that Vatinius answer which of the two it was
better for the state should have been born a citizen.}

\textit{The body of the cross-examination is structured around
the three pillars of legal-religious order that Vatinius's
tribunate trampled: the \textbf{auspices}
(\textsection\textsection 14--17, 20), the
\textbf{lex Aelia et Fufia} restraining tribunician
legislation (\textsection\textsection 18, 23), and the
\textbf{authority of the Senate} (\textsection\textsection 35--36,
on the legation Vatinius arranged for himself by his own bill).
Around these are arranged the charges of personal violence
(\textsection\textsection 21--23, the leading of Bibulus to
prison and the siege of his house), the Vettius affair
(\textsection\textsection 24--26, the suborned informer placed
on the rostra to mark out for proscription a Senate-bound
catalogue --- Bibulus, Lucullus, Curio, Domitius, Lentulus the
Flamen Martialis, L. Paulus --- and then strangled in the
Tullianum to silence the bribery), the alternate-rejection law
sprung against C. Antonius (\textsection\textsection 27--28,
with the parallel praise of Q. Maximus), the swift wealth from
treaties with cities, kings, and tetrarchs in a single year
(\textsection 29), the funereal toga at Q. Arrius's
\textit{epulum} in the temple of Castor
(\textsection\textsection 30--32 --- a set-piece on the public
feast turned into a deathly thing), and the Memmius
prosecution under the lex Licinia Junia
(\textsection\textsection 33--34, the unheard-of appeal to a
tribune --- Clodius by name --- and the storming of the
praetor's tribunal in the Forum).}

\textit{The closing run (\textsection\textsection 37--41) turns
on Vatinius's candidacy for the praetorship of 55 BC. Cicero's
own lex Tullia de ambitu (the canvassing law of 63 BC) explicitly
forbids gladiatorial games in the two years before standing,
and Vatinius is giving them; Caesar at Aquileia is reported as
saying that ``Vatinius in his tribunate did nothing for free; a
man who has placed everything in money ought to bear the lack of
office with equanimity'' (\textsection 38). The peroration
(\textsection 39) catalogues the public loathing --- relations,
fellow tribesmen, neighbours, kinsmen, even the scrofulas
``migrated from your foul face'' --- and asks why he should
prefer the praetorship to death. \textsection\textsection 40--41
return to the Sestius trial: the inconsistency of praising T.
Annius (Milo) here while having given false testimony against him
elsewhere; and the closing trap --- Vatinius had argued in his
own testimony that Sestius should not have been arraigned for
\textit{vis} at all, and yet has just testified against him on
that very charge.}

\textit{The interrogation was a hostile triumph: Sestius was
acquitted unanimously on 11 March 56 BC (Q. fr. 2.4
\textsection 1). It was also, in the longer arc, a tactical
mistake. Caesar took the offence personally: when Cicero in
56--54 BC moved (under the post-Luca compact) to defend the
men he had attacked, the most conspicuous of those rapprochements
was with Vatinius himself. Cicero would defend Vatinius for
\textit{de repetundis} in 54 BC --- the speech does not survive,
but the letters report it (Att. 4.16, Q. fr. 2.16) ---
and would correspond with him cordially in the late forties
(Fam. 5.9--11a, from Vatinius governor of Illyricum, are
amiable and even warm). The \textit{In Vatinium} stands, then,
as the high-water mark of the brief optimate-Pompeian-Sestian
front of late 57 -- early 56 BC, just before Luca: the moment
when Cicero, riding the unanimous Sestius acquittal, could
afford to flay one of Caesar's own creatures in open court. It
is also a small classroom in the rhetorical interrogation: the
sustained \textit{quaero, volo audire, fecerisne, dixerine} ---
the verb forms of cross-examination accumulated into a sustained
prosecution-by-question.}
