% Pro P. Sestio --- headnote
% pro-sestio, March 56 BC, Rome

\textit{Cicero for P. Sestius, delivered at Rome in March 56
BC. Sestius --- the tribune of the plebs of 57 BC, the
``actor and defender of my cause'' as Cicero calls him here
(\textsection 75) --- had been indicted under the lex Plautia
de vi for the political violence of his year. The actual
violence had been directed at him: in the first half of 57
BC, while pushing the bill for Cicero's recall, he was
attacked by Clodius's gangs in the temple of Castor and
left for dead, with more than twenty wounds. The prosecution,
mounted by Clodius's circle (with the new aedile, P. Albinovanus,
as accuser, and the witness Vatinius taken on by Cicero in
cross-examination as the surviving \textit{In Vatinium}),
was less an attack on Sestius than the second half of the
campaign that had failed against Cicero. The defence team
was the optimate front --- Hortensius opening (\textsection 3,
14), Crassus, M. Licinius Calvus also speaking; Cicero
closing. The acquittal was unanimous on the fifth day before
the Ides of March (11 March), as Q.fr. 2.4 reports.}

\textit{The speech sets the case in the form of a great
\textit{narratio} of the year of exile (\textsection 15--71).
The first set-piece is the joint portrait of the two consuls
of 58 BC, Gabinius and Piso, ``two whirlwinds of the
commonwealth'' (\textsection 15--35) --- the perfumed Gabinius
of the Puteal moneylenders' arcade and the curled hair, the
shaggy Piso of the schooled brow, sold the provinces in
advance for letting Clodius's bill against Cicero pass.
\textsection 36--52 is the apology for not fighting:
Cicero's account, set against Marius's exile and Metellus
Numidicus's, of why a private armed defence would have
brought slaughter and not preservation. \textsection 53--63
returns to the year, with the publication of the Cyprian king
(Ptolemy of Cyprus's auctioned kingdom; the Cyprian commission
that fell on Cato as the most-conspicuous moderator of the year
that needed moderating); \textsection 64--71 the recall
campaign, Pompey's late stirring, the Sestius journey to
Caesar (\textsection 71). \textsection 72--85 is the
narrative of Sestius's tribunate proper, with the violence
on 8 K. Feb. and the near-death in the temple of Castor,
the famous narrow escape of the Clodian hireling Numerius
who was almost killed for the unfortunate near-coincidence
of name with Q. Numerius Rufus (Gracchus). \textsection 86--95:
the parallel with Milo, who is here the great companion
figure --- Cicero's defence of guarding-by-private-force is
explicitly grounded in Milo's precedent.}

\textit{The peroration, \textsection 96--143, is the great
political digression of Cicero's whole career: the famous
analysis of the \textit{optimates} as not a party but a
disposition --- ``all are \textit{optimates} who are neither
guilty, nor wicked by nature, nor furious, nor hindered by
domestic evils'' (\textsection 97), the ``cum dignitate
otium'' formula at \textsection 98 (peace with standing as
the goal of statecraft), and the catalogue of the foundations
of that peace: religion, the auspices, the powers of the
magistracies, the authority of the Senate, the laws, the
custom of the ancestors, the courts, jurisdiction, good
faith, the provinces, the allies, the praise of empire,
military affairs, the treasury (\textsection 98). The
optimate digression then catalogues the proofs that the
true Roman people --- the \textit{populus} as opposed to
the \textit{contio} packed with hired claqueurs --- has been
plainly with Cicero: the consular contio under Lentulus
(\textsection 107), the centuriate-assembly recall vote
(\textsection 109), the games at the Apollonian \textit{ludi}
where Aesopus the actor pleaded Cicero's case from the lines
of Accius's \textit{Eurysaces} and \textit{Brutus}
(\textsection 120--123), the gladiatorial show of Scipio
Nasica where Sestius's appearance from the column of Maenius
brought roars of approval ``from the railings of the Forum''
(\textsection 124). \textsection 132--135 is the answer to
the prosecutor's gibe about ``the nation of the
\textit{optimates}'' and the close-in of C. Cato, the
gladiator-flouter and friend of the act-of-Caesar-rescinders.
\textsection 136--143 closes on the seven-times praise of the
\textit{optimas vita} for the young men of Rome --- the
example of Themistocles, Aristides, Miltiades, Hannibal cast
out by their own cities and yet immortal in the memory of
their adversaries --- and the final list of the Roman
forefathers (Bruti, Camilli, Ahalae, Decii, Curii, Fabricii,
Maximi, Scipiones, Lentuli, Aemilii) whose place in the
\textit{coetus deorum immortalium} the speech sets as the goal.
The sight-stopping close (\textsection 144--147) returns to
the persons present in the court --- Sestius, Milo, the boy
in the praetexta, the young P. Lentulus pleading for his
father --- and the imploration of the jury that, having wanted
Cicero safe, they preserve those through whom he was recovered.}

\textit{The speech was the political founding-document of
the post-exile Cicero. It is also the speech in which the
optimate consensus of late 57 BC at last finds its theory:
the formula \textit{cum dignitate otium} would be quoted
back at Cicero by his successors and by his prosecutors for
the rest of his life. The unanimous acquittal of 11 March
(Q.fr. 2.4 \textsection 1) was the high-water mark of the
optimate-Pompeian-Sestian front --- which would break apart
within five weeks at the conference of Luca, when Caesar,
Pompey, and Crassus reset the terms.}
