% Pro L. Flacco --- headnote
% pro-flacco, mid (probably August) 59 BC

\textit{Cicero's defence of Lucius Valerius Flaccus, late
governor of Asia (62 BC), prosecuted in mid-59 BC for
\textit{repetundae} (provincial extortion). Flaccus was a man
to whom Cicero owed a particular debt: as urban praetor of 63
BC he had intercepted the Allobrogian envoys at the Mulvian
Bridge and brought back to the consul the sealed letters
that exposed the conspiracy --- the documentary proof on
which the senatorial debate of 5 December and the executions
of that night turned. The prosecutor was Decimus Laelius, son
of a Pompeian friend, with subscribers of his own; and the
common talk in Asia, which Cicero will not deny but cannot
quite refute, was that Pompey himself had set Laelius on. The
defence was divided: Quintus Hortensius spoke first and took
the Mithridatic charge that turns on the witness Mithridates
of Pergamum, while Cicero took everything else --- the
critique of Greek and Asian witnesses as a class, the
city-by-city analysis (Acmonia, Dorylaeum, Temnus, Tralles,
Pergamum, Apollonis), the complaint about the Jewish
temple-gold which Flaccus had embargoed at Pompey's
precedent, the four Roman accusers (Decianus, Andro
Sextilius, Lucceius, Falcidius), and the closing.}

\textit{The speech survives partly through medieval
manuscripts (\textsection 1--5 and 6--106 with two closing
fragments) and partly through palimpsest readings of a
Bobbio codex (the Bobiensian fragments \textit{frBob1--12} and
the longer \textit{frMed}), which fill the lacuna between
\textsection 5 and \textsection 6 and preserve the opening of
the \textit{narratio} on Flaccus's career: military tribune
under P. Servilius in Cilicia, quaestor under M. Piso in
Spain, legate in the Cretan war under Metellus, urban praetor
in 63, governor of Asia in 62. The body of the speech then
unfolds at length: \textsection 9--26 the indictment of the
whole class of Greek witnesses (the proverb \textit{da mihi
testimonium mutuum}, the contrast of Roman \textit{religio}
with Greek \textit{contio}-shouting); \textsection 27--66
the city-by-city refutation, with the ethnographic set piece
on the three Greek races against the four nations of Asia
(``a Phrygian is generally bettered by being beaten,'' the
Lydian slave of comedy); \textsection 66--69 the famous
defence of Flaccus's gold-export ban, with its single
sentence on Jewish religion and the Roman state that has
echoed into modern reading lists; \textsection 70--85 the
Roman complainants, with the Decianus--Apollonis story of the
abducted wife and battered father; \textsection 86--93
Lurco, Sextilius, Falcidius, and the absent witness whose
mother and sister produce his letters; and the great
peroration (\textsection 94--106) which is the political heart
of the speech.}

\textit{The peroration is the reason for our reading. The
trial of 59 BC, four and a half years after the executions
in the Tullianum, is for Cicero a stalking-horse trial of
the Catilinarian executors: the prosecution of C. Antonius
earlier in the year had decked Catiline's tomb with flowers,
and to convict Flaccus would be (Cicero says) a victim
offered to the manes of Lentulus. \textsection 102, ``O
night, that almost brought eternal darkness upon this city
\ldots O those Nones of December which fell in my
consulship!'' is the consular's full-throated last
self-celebration before exile; \textsection 103, ``If
anything graver shall happen --- I, I say, Flaccus, shall
have betrayed you,'' is the consular's first acknowledgment,
in the surviving corpus, that the political weather has
turned and that the men who saved the state are now the
hunted. Caesar is consul, the triumvirate is in its first
high summer, P. Clodius has been transferred to the plebs in
March under Caesar's auspices and stands for the tribunate of
58. Flaccus was acquitted (\textit{Att.} 2.25 reports the
dignity preserved); within twelve months Cicero himself was
in exile.}
