% Pro P. Sulla --- headnote
% pro-sulla, mid-62 BC

\textit{Cicero's defence of Publius Cornelius Sulla --- nephew of
the dictator and consul-designate of 65 BC, condemned for
\textit{ambitus} (electoral bribery) along with his colleague
P. Autronius Paetus and stripped of the consulship before he
held it --- now charged under the Lex Plautia (a law on public
violence) with complicity in both the so-called ``first
Catilinarian conspiracy'' of 66/65 BC and the conspiracy of
63 BC. The prosecutor is Lucius Manlius Torquatus the
younger, son of the consul of 65 BC who, with Lucius Aurelius
Cotta, took the consulship that Sulla and Autronius had won
and lost; the prosecution is therefore the second strike of
the same family quarrel, the first having stripped Sulla of
office at the canvassing trial, this one seeking now to
strip him of citizenship and the right to the fire and
water of Italy. Cicero shares the defence with Quintus
Hortensius: Hortensius, who knew the elder Torquatus's
counsels, takes the charges relating to the ``first
conspiracy''; Cicero, who as consul tracked down and crushed
the second, takes those relating to his own consulship. The
trial probably falls in the spring or early summer of 62 BC.}

\textit{The strategic problem the speech has to solve is the
problem of consistency. Cicero in December 63 had executed
Lentulus, Cethegus, and three others without trial under the
\textit{senatus consultum ultimum}; Autronius, who had fled,
he had since refused to defend; against Autronius indeed he
had given evidence. To stand up now for a man whose
consulship had been forfeited along with Autronius's, who
had been Autronius's colleague in disgrace as in office, is
on the surface a sharp inconsistency, and Torquatus has
opened by accusing Cicero of regnum --- of arrogating to
himself a private kingship in the choice of whom to ruin and
whom to save. Cicero answers that the cases are wholly
unlike: Autronius was named by the Allobroges' deposition,
fought to break up his canvassing trial with hired
gladiators, threw himself into the conspiracy from beginning
to end; Sulla on the contrary lived through the whole crisis
quietly at Naples, was named by no informer, accused by no
witness, mentioned only once and then equivocally by Cassius
to the Allobroges, and bore his earlier disaster with such
modesty as to constitute his best testimony. The argument
from the absence of evidence is unfolded with great skill
(\textsection 11--67), each charge --- the gladiators bought
in Faustus's name, the dispatch of Sittius to Spain, the
men of Pompeii and the colonists, the bill of L. Caecilius
to relieve the canvassing penalty --- met in turn.}

\textit{The peroration shifts from charge to character and
finally to pathos. The conspirators were not men but
\textit{beluae quaedam ex portentis}, a kind of monstrous
beasts in human form, the eruption of long and now
despaired-of diseases of the commonwealth. Sulla, against
that, is the man whose voice never wounded a citizen, who
under the dictator pleaded for many lives, who in his
disaster has avoided the city out of shame; the case is no
longer about his life but about his burial. The speech
closes on his small son, who knows the paths of the
courts better than the paths of the Campus and the schools,
and on Cicero's plea to the judges to mitigate by mildness
the severity of these months. Sulla was acquitted; he lived
quietly through the 50s and (according to Caesar's
\textit{Civil War}) commanded the right wing at Pharsalus
in 48 BC.}
