% Pro A. Licinio Archia Poeta --- headnote
% pro-archia, summer/autumn 62 BC

\textit{Cicero's defence of the Greek poet Aulus Licinius
Archias of Antioch on a charge that he was not in fact a
Roman citizen --- brought, on a hostile reading of the
\textit{lex Papia} of 65 BC (which had ordered the
expulsion of those falsely claiming citizenship), by an
otherwise unknown prosecutor named Grattius. The political
context is the long campaign against the followers of
Lucullus: Archias had come to Rome in the consulship of
Marius and Catulus (102 BC) and had been at once taken into
the household of the Luculli, and had attended Marcus
Lucullus to Sicily and Lucius Lucullus to the Mithridatic
war, on which he was now writing the great epic. Two years
earlier the Manilian command had stripped Lucullus of the
eastern war and given it to Pompey; in 63 Lucullus's enemies
had blocked his triumph until 62; the prosecution of Archias
is a small late stroke in that campaign, aimed not at the
poet for himself but at his patron. The trial falls in the
summer or autumn of 62 BC, before a single praetor (Quintus
Cicero, the orator's brother) and a panel of judges.}

\textit{The technical case is short and dispatched in the
first eleven sections. Archias was enrolled as a citizen of
Heraclea, a federated town; the law of Silvanus and Carbo
(89 BC, the \textit{lex Plautia Papiria}) gave Roman
citizenship to citizens of federated towns who had a domicile
in Italy and registered before the praetor within sixty
days, all of which Archias did. The Heraclean rolls
themselves were burned in the Italian war, but Marcus
Lucullus is present to testify and the Heracleans have sent
legates with a public deposition; Archias's registration
stands in the records of the praetor Q. Metellus, the most
scrupulous of men. Cicero closes the technical part of the
speech with a list: he made wills under Roman law, took
inheritances from Roman citizens, was put up to the treasury
for benefits by Lucullus the proconsul. The case for the
citizenship is settled.}

\textit{The remaining three-quarters of the speech is the
\textsection 12--32 digression --- the most famous defence
of letters in Latin literature, and the one for which the
trial has been remembered for two thousand years while every
other detail of it has gone. Cicero asks pardon at the start
for ``a kind of speech that is almost new and unaccustomed''
and for treating, in court, of \textit{studia humanitatis}.
The argument moves from his own debt to Archias and to
letters (\textsection 12--14), to the cooperation of nature
and learning in producing greatness (Africanus, Laelius, the
elder Cato, all enriched by letters), to the famous
characterization at \textsection 16: ``these studies sharpen
youth, delight age; they adorn prosperity, in adversity
afford a refuge and consolation; they delight us at home, do
not hinder us abroad; they pass the night with us, they
travel, they go to the country.'' Then to the holiness of the
poet's name and the role of Archias's verses in extending the
glory of Roman arms beyond the narrow bounds of Latin into
the universal Greek (\textsection 19--24, with Alexander at
the tomb of Achilles, ``O fortunate young man, who hast found
in Homer the herald of thy virtue!''), to the Roman precedent
(Ennius and the Scipios; Accius and Brutus; Fulvius
consecrating the spoils of Mars to the Muses), and finally to
Cicero's own confession that virtue desires no wages but the
praise and glory of being remembered, and that he too has
hoped, in his consulship, that his deeds were being scattered
into the everlasting memory of the world. Archias, he hopes,
will give them the eternal testimony.}

\textit{The poem on the consulship was apparently never
finished; Cicero in the end wrote his own. Archias was
acquitted. The speech became, in late antiquity and again in
the Renaissance, the central manifesto of the studia
humanitatis: Petrarch's discovery of a manuscript at Li\`ege
in 1333 helped open the recovery of Cicero's prose, and
Petrarch's own treatise \textit{De Otio Religioso} draws
directly on \textsection 16. The reader who comes to it for
the first time should be warned that some sentences will be
familiar.}
