% De Lege Agraria I --- headnote
% de-lege-agraria-1, 1 January 63 BC

\textit{Cicero's first speech as consul, delivered in the Senate
on the Kalends of January 63 BC --- the very first day of his
year of office --- against the agrarian bill of the tribune
Publius Servilius Rullus. The bill proposed the establishment
of a board of ten commissioners (\textit{decemviri agris
dandis adsignandis iudicandis}) with extraordinary imperium
for five years: power to sell off vast tracts of public land
in Italy and the provinces (including the lands of Mithridates
in the territory still being conquered by Pompey, and the
remains of the Carthaginian and Corinthian lands consecrated
by Scipio Africanus and Mummius), to use the proceeds to buy
private land in Italy --- compulsorily, since the holders
could not refuse to sell at the decemvirs' price --- and to
plant colonies of Roman citizens upon it, particularly at
Capua and on the rich Campanian lands long held in trust by
the state. The decemvirs would have power to declare any tract
public, to assess the rest, and to spend the receipts; their
nominee was understood to be Crassus, with Caesar standing
behind the legislation. The opening of the speech is lost; what
survives are four short fragments preserved by other authors
and the substantial closing portion (here \textsection 1
through \textsection 27).}

\textit{Cicero's argument is the consul's argument. The bill
puts up the entire patrimony of the Roman people for sale,
\textit{sub praecone}, under the auctioneer's voice; it
delivers to ten men the imperium of magistrates without their
accountability; it threatens Pompey's command and his army's
ground in the East; it proposes to colonise Capua again as a
second Rome, with all the cities round it garrisoned by
decemviral planters. The closing pivot --- which sets the key
for the consulship as a whole --- is the consul's announcement
that he will not, when his year ends, take a province; that
the magistracy will be conducted free of the appetites that
make a magistrate vulnerable to a tribune; that the danger to
Rome is no longer from any king or nation outside, but
\textit{inclusum malum, intestinum ac domesticum} --- ``the
evil is enclosed, internal, domestic.'' The line will turn
out, by November, to have been more nearly literal than
metaphorical.}

\textit{The bill failed, after Cicero followed this senatorial
speech with a far more sustained address to the people in the
Forum (\textit{De Lege Agraria II}) and a third speech, also
to the people, on a particular provision (\textit{De Lege
Agraria III}). Rullus dropped the proposal.}
