% In Catilinam IV --- headnote
% in-catilinam-4, 5 December 63 BC

\textit{The fourth Catilinarian, delivered in the Senate on
5 December 63 BC --- the famous Nones of December debate on
the punishment of the conspirators in custody (Lentulus,
Cethegus, Statilius, Gabinius, Ceparius). Cicero presides as
consul. The Senate is divided. Decimus Silanus, the
consul-designate who is asked first, has spoken for death;
Tiberius Nero has spoken for postponing the question; Gaius
Caesar, when his turn came, has spoken for the alternative of
confiscation of goods and life imprisonment apportioned among
the Italian municipalities, with no possibility of subsequent
remission either by the Senate or by a popular assembly.
Cicero rises to expound the two opinions and let the Senate
choose, but it is plain from the speech which he prefers and
why. The Sallustian record makes the next speakers Lutatius
Catulus and Cato; Cato will speak next for death, the Senate
will follow him, and that night the prisoners will be taken
by Cicero in person to the Tullianum and strangled in turn.}

\textit{Cicero's argument is the consul's. The opening
disclaimer of personal danger and concern for his family
(\textsection 1--3) sets up the moral position from which the
rest is spoken; the analysis of the two opinions in
\textsection 7--11 is one of the most exact pieces of
parliamentary politics in the corpus. Caesar's opinion is
generously described --- a hostage of his perpetual goodwill
toward the commonwealth, a true mind \textit{popularis}
distinguished from the levity of mob-orators --- but Cicero
points out, by way of asking how Caesar can hold the
conspirators not citizens for the purposes of the Lex
Sempronia, that the consequence of Caesar's own reasoning is
that the death penalty is licit, since the conspirators are
no longer citizens at all. The roll-call of the orders
(\textsection 14--17) is a portrait of the city standing
together: the Roman knights, the tribunes of the treasury,
the scribes, the freeborn of every grade, the freedmen, the
shopkeepers (whose ``stock-in-trade is sustained by a throng
of citizens, is nourished by quiet''), even the slaves of any
tolerable condition. The consular pivot is the closing
section's catalogue of what Cicero has refused on behalf of
the city: the army, the province, the triumph, the badges of
praise; in return he asks only the memory of his consulship,
and commends to the Senate his little son.}

\textit{The conspirators were executed that night ---
\textit{vixerunt}, ``they have lived,'' was Cicero's word to
the crowd as he came down from the Tullianum. The execution
without trial of Roman citizens by senatorial decree, on the
authority of the senatus consultum ultimum, would five years
later furnish Clodius the pretext for Cicero's exile. The
voice that closes this speech --- ``you have a consul who will
not hesitate either to obey your decrees and to defend, as
long as he lives, what you shall have decided'' --- was
hearing in the same room, on this same night, the steps of
the same future.}
