% Pro L. Murena --- headnote
% pro-murena, late November 63 BC

\textit{Cicero's defence as consul of the consul-elect Lucius
Licinius Murena on a charge of \textit{ambitus} (electoral
bribery), brought under Cicero's own bribery law (the Lex
Tullia) by Servius Sulpicius Rufus, the defeated rival
candidate, and by Marcus Cato, the tribune-elect, with the
younger Servius Sulpicius and Gaius Postumus joining the
prosecution. The trial falls in the last days of November 63
BC, between the Second Catilinarian (9 November) and the
Third (3 December): Catiline is in the field at Manlius's
camp, Lentulus and the rest of the urban conspirators are at
large in the city, and the consul-designate of January is on
trial for his life. To convict Murena would leave one consul
on the Kalends of January --- and that consul (Decimus
Silanus, the other consul-elect) without a colleague in the
year of the Catilinarian crisis. Cicero, who as consul has
just driven the bribery law through, defends a man accused
under it; that anomaly is the speech's first business, and the
proem (\textsection 1--10) explains it as the consul's duty
to the commonwealth.}

\textit{The body of the speech is a triptych. The first part
(\textsection 11--14) deals with the alleged scandals of
Murena's life --- so thin that the prosecution has had to
fall back on calling him a dancer, which Cicero turns into a
small set piece on what kind of man would dance. The second
(\textsection 15--53) is the famous comparison of Murena's
soldierly career under Lucullus in the Mithridatic war with
Sulpicius's career as a jurist at Rome --- with the digression
on the absurdity of legal study (the Sabine farm dialogue at
\textsection 26, the joke that Cicero will declare himself
a lawyer in three days, the rule that those who could not
turn out as orators slip into the law as those who could not
turn out as harpers slip into the pipe-playing). The third
(\textsection 54--85) takes up the actual bribery charges,
mostly from Cato; what gives the section its colour is its
parallel digression on Stoic morality, where Cicero (with
elaborate respect for Cato as a man) parodies Cato's
adherence to the Zenonian rigorism --- the wise man pardons
no fault, all faults are equal, the wise man never changes
his opinion --- and proposes that some study with Plato and
Aristotle would have made Cato no better but a little
gentler. The contrast with Scipio at home with Panaetius (and
with Cato's own great-grandfather, the most affable of
Stoics) lands the joke without insult.}

\textit{The closing peroration (\textsection 78--90) drops
the wit and turns to the Catilinarian danger. The Trojan
horse is within the walls; the consul has weeks to live in
office; if Murena is condemned, on the Kalends of January
there will be only one consul, and that consul without a
colleague in the year that is to be. The set piece is the
projection of where the convicted man would go: home to see
his father's laurelled portrait disfigured, to his mother who
has just kissed her son the consul, into exile in the East
where he had been legate or in Transalpine Gaul where his
brother now governs. Murena was acquitted; he served the
consulship; the conspirators in the city were executed by
order of the Senate within a fortnight.}
