% In Catilinam I --- headnote
% in-catilinam-1, 8 November 63 BC

\textit{The first of the four Catilinarian speeches and the
most famous opening in Latin oratory: \textit{Quo usque
tandem abutere, Catilina, patientia nostra?} ``How long, at
last, Catiline, will you go on abusing our patience?''
Delivered on the morning of 8 November 63 BC in the temple of
Jupiter Stator on the slope of the Palatine, where the Senate
had been hastily reconvened (its usual chamber being thought
unsafe). The day before, a meeting of the conspirators at the
house of Marcus Laeca had distributed parts of Italy among the
plotters and resolved on the murder of the consul before
dawn; two Roman knights had volunteered for the killing and
been turned away from the door of Cicero's house, which had
been forewarned and reinforced. The senatus consultum
ultimum had been passed on 21 October, more than two weeks
earlier, but no consular hand had yet fallen on the
conspirators in the city.}

\textit{The speech is a single sustained act of frontal
address. Catiline is sitting in the Senate; the consular
benches around him have emptied at his approach; and Cicero
speaks not over him, not past him, but at him --- with the
rest of the senators as witnesses. The argument is that
Catiline must leave Rome of his own motion, taking the dregs
of the conspiracy with him, so that the war may at last
declare itself a war and not a brigandage hidden within the
walls. The two great prosopopoeiae are the speech's spine: in
\textsection 18 Cicero gives voice to the \textit{patria},
the country herself, addressing Catiline (``go away from me
and take this fear from me''); in \textsection 27 the same
\textit{patria} addresses Cicero, demanding to know why he
suffers an enemy to leave rather than dragging him to death.
The speech opens and closes with Jupiter Stator: at the start
the god is thanked for keeping the city safe through the
plot's many escapes; at the end the god is invoked, by his
true name as Stayer, to keep Catiline and his confederates
from the temples and walls of the city and to visit them, in
life and death, with eternal punishments.}

\textit{Catiline rose, made a brief and bitter answer (which
does not survive), and that night left Rome for Manlius's
camp at Faesulae in Etruria. The Second Catilinarian, the
\textit{contio} delivered in the Forum the next day,
announces his departure to the people.}
