% De Lege Agraria III --- headnote
% de-lege-agraria-3, January 63 BC

\textit{Cicero's third speech against the Rullan agrarian
bill, delivered to the people in the Forum a few days after
\textit{De Lege Agraria II}. The latter half of the speech is
lost; the surviving portion (sixteen sections) is a self-contained
counter-attack against the charge that Cicero in opposing the
bill is acting as patron of the Sullan possessors --- of the
Septimii, Turranii, and the rest who held the lands proscribed
in 82 BC.}

\textit{Cicero's response is to drag a single suppressed
clause of the bill into the daylight: clause XL, which
provides that whatever any man has held since the consulship of
Marius and Carbo (i.e. since 82 BC --- Sulla's name decorously
left unspoken) shall stand in the best private right.
Rullus's ``Marian'' formula is the consular dating of the year
Sulla destroyed; the clause itself is a quiet amnesty, going
beyond Sulla's own laws and the Lex Valeria interregni, that
gives Sullan plunder a better title than ancestral inheritance
--- because better than mortgaged is unmortgaged, better than
taxed is untaxed, better than liable to a praetor's interdict
is held by the best right under the civil law. The further
giveaway: the clause embraces not just \textit{data}
(``given'') but \textit{possessa} (``possessed'') --- and so
sweeps in those who have come into possession by force, by
stealth, by precarious tenure. Rullus, defending Sullan
holdings, is far more Sullan than Sulla. The personal sting
is that Rullus's own father-in-law (Valgius, the same heavily
endowed Hirpinian landowner of the Casinate set-piece in
\textsection 14) is the prime beneficiary.}

\textit{The closing reversal is the rhetorical pivot: Rullus
charges Cicero with defending Sullan possessors. Cicero answers
that what he is in fact defending is the people's possession of
the Scantian Forest, of the Campanian land, of the lands of
Italy, Sicily, and the provinces. Then comes the warning that
will turn out to be prophetic for the year ahead: the
Campanian land is being held in reserve not for you, citizens,
but for the kind of men who are fitted for violence, for
crime, for slaughter --- an army is being set up against you,
against your liberty, against Pompey; against this city,
Capua. The bill failed; Rullus dropped the proposal.}
