% De Lege Agraria II --- headnote
% de-lege-agraria-2, January 63 BC

\textit{The principal speech against the Rullan agrarian
bill, delivered to the Roman people in the Forum a few days
after the speech in the Senate (\textit{De Lege Agraria I}).
This is the first time Cicero ever addresses the people from
the Rostra as consul, and he opens by setting himself out as
the test case for what \textit{popularis} means: a new man
without ancestral portrait-busts to plead for him; the only
new man within memory to win the consulship at the first
attempt and in his year, by viva voce acclamation rather than
ballot; a man who comes before the people not from the cradle
but from the campaign field. The keynote --- repeated in three
forms --- is that there is nothing so popular as peace, as
liberty, as quiet, and that what looks popular (\textit{lex
agraria}) is in fact a snare against all three.}

\textit{The bill itself, as Cicero unfolds it clause by
clause, sets up a board of ten commissioners with five-year
extraordinary imperium, with a hand in selling the entire
patrimony of the Roman people: the public lands of Italy, the
royal lands of Macedonia and Bithynia, the lands of Mithridates
in Pontus and Cappadocia (still in the field of Pompey's war,
sold here \textit{sub praecone} while the commander himself
holds command), even Alexandria and Egypt under colour of
King Alexa's contested will. From the proceeds, lands in
Italy are to be bought and colonies planted upon them ---
including, climactically, at Capua. Cicero builds the bill
into a kingship: the decemvirs are kings in everything but
the name; their judgments admit no appeal; their auctions are
held in the dark of Paphlagonia and the solitude of
Cappadocia; their satellites are the surveyors of equestrian
rank set as bodyguards. Pompey is excluded by name from the
decemvirate; Pompey's spoils alone are exempted from the
decemviral inquiry, which Cicero reads as not honour but the
prosecutors' fear that the man would not endure the same
contempt as the rest.}

\textit{The set piece is the long discussion of Capua
(\textsection 76--97). Our ancestors, Cicero argues, took the
land from the Campanians, took the magistrates and the
commonwealth, and left the city only as a granary and a place
to lodge the ploughmen --- because they reckoned that only
three cities of all on earth, Carthage, Corinth, and Capua,
could sustain the weight and the name of empire. The other two
they destroyed; Capua they gutted from within. To plant five
thousand colonists at Capua, with a hundred decurions and ten
augurs and six pontiffs, and to call their magistrates not
duumviri but \textit{praetors} (as Marcus Brutus's earlier,
ill-omened colony had done), is to set up a second Rome
against this one. The whole digression is, in effect, an essay
in the political geography of the western Mediterranean and a
moral fable about how character is formed by place.}

\textit{The speech closes on the consul's note: peace,
tranquillity, quiet are what the people want; the consul's
office is the consul's defence of them; and the consul has
established concord with his colleague (Antonius Hybrida ---
unnamed but evidently the man) against those who said the two
would be enemies. The bill failed; Rullus dropped the
proposal.}
