% Pro A. Cluentio --- headnote
% pro-cluentio, 66 BC

\textit{The longest of Cicero's forensic speeches, at 202
sections. Defence in 66 BC of Aulus Cluentius Habitus of Larinum,
charged with the poisoning of his stepfather Statius Albius
Oppianicus and with two further alleged poisonings (of Vibius
Capax and of Cluentius's own brother-in-law, the younger
Oppianicus). The speech is built in two halves: the first against
the long-running public hatred (\textit{invidia}) attaching to
Cluentius from his prosecution of his stepfather eight years
earlier --- the so-called \textit{iudicium Iunianum} of 74 BC,
which was condemned by tribunician agitation as having been
corrupted by Cluentius's bribery, and which has hung over him
ever since. Cicero argues at length that the trial was indeed
corrupted, but in the opposite direction: by Oppianicus, through
the judge Staienus, who took 640{,}000 sesterces and pocketed it
when he saw the case was lost. The set-piece narrative of the
older Oppianicus's career --- ten or more murders by poison and
forgery, including his own children, his uncle, his
mother-in-law, and the wife of Aulus Aurius (whose widow Sassia,
Cluentius's own mother, he then married) --- forms the bedrock of
the case. The argument that Roman knights are not held by the
Cornelian law against the corrupting of trials (\textit{ne quis
iudicio circumveniretur}) is offered but explicitly waived at
Cluentius's own request, on the ground that he wishes to be
defended by the matter, not by the law. The closing peroration
addresses the figure of Sassia: the mother who long since
married her son-in-law, married her daughter's husband, and now
seeks her own son's death; whose journey to Rome through the
neighbouring towns is shown causing horror in every place she
passes. The whole town of Larinum has come up to Rome to support
Cluentius. The speech ends with the famous concluding line: ``in
public meetings is the place of hatred, in trials of truth.'' The
case is celebrated also for Quintilian's later remark that Cicero
\textit{tenebras offudisse iudicibus} --- ``cast darkness over
the judges.'' The acquittal followed.}
