% In C. Verrem Actio Secunda V --- headnote
% in-verrem-2-5, autumn 70 BC

\textit{Book V of the five-book Actio Secunda --- ``de suppliciis,''
the closing book of the prosecution of Verres, on his cruelties.
Cicero opens by demolishing Verres's main line of defence ---
that whatever he did at home, he was at least a good general who
kept Sicily safe from a slave war and from pirates. He shows
that there was no slave war in Sicily; that the Sicilian fleet
under Cleomenes (the husband of Verres's mistress Nice, set in
command of the fleet so that Verres could keep Nice on the
shore) was sent to sea half-empty because the rowers and soldiers
had been discharged for cash; that the pirate Heracleo with four
small light vessels burned the entire Roman fleet on the beach
at Helorus while Verres lay drunk; and that Heracleo afterwards
sailed unopposed into the Syracusan harbour, his rowers
splashing the eyes of the praetor and his men eating Sicilian
grain while the Sicilian sailors fed on roots of palms. To cover
the loss, Verres charged the captains of the lost ships with
betrayal and had them executed at the Achradina column, exacting
fees from their parents to lighten the strokes of the axe. The
real chief pirate captured by Caesetius and Tadius vanishes,
ransomed back into the trade; a substitute is kept alive in the
inland town of Centuripa to be brought forth at trial as a
witness. Roman citizens caught in Sicilian harbours, accused of
being Sertorian fugitives, are thrown into the Latomiae
stone-quarries at Syracuse and killed. The book builds to its
single most famous passage: the Roman citizen Publius Gavius of
Consa, escaped from the stone-quarries, caught at Messana, and
crucified by Verres on the Strait at the spot where the
condemned man could see Italy from the cross --- crying, beneath
the rods, only the words \textit{civis Romanus sum}: ``I am a
Roman citizen.'' The closing peroration calls upon the gods
whose images Verres has plundered (Jupiter, Juno, Minerva,
Latona and Apollo and Diana, Mercury, Hercules, the Mother of
Ida, Castor and Pollux, Ceres and Libera) and asks of them, of
the senate, and of the Roman people one thing: a worthy outcome.
The book is the architectural climax of the corpus, and its
last sentence is the prosecutor's wish that he may hereafter
defend the good rather than be compelled to accuse the dishonest.}
