% Pro Quinctio — headnote

\noindent
Cicero's earliest surviving speech, delivered in 81 BC when he was twenty-five
or twenty-six and a \textit{novus homo} just beginning his career at the bar.
The case is a civil one. Publius Quinctius's deceased brother Gaius had been
in business partnership in Gaul with Sextus Naevius --- formerly an auctioneer
in Rome --- and on Gaius's sudden death the partnership had passed to
Publius. Naevius, after years of silence about any debt, contrived (Cicero
charges) that a bond for a court appearance be technically broken; obtained
from the praetor Burrienus an order that Quinctius's goods be advertised
under the edict against fraudulently absent debtors; sent agents to throw
Quinctius forcibly from the Gallic estate; and now, before the judge Gaius
Aquilius, was pressing Quinctius to argue first, in the manner of a defendant
in a capital trial, on the standing of his own reputation and fortunes.

The opposing counsel was Quintus Hortensius --- the leading orator of Rome
before Cicero himself --- and a phalanx of senatorial advocates, including
the consular Lucius Philippus. Cicero's brief was to maintain three negative
propositions: that Naevius had no cause to apply for possession of the
goods; that the requirements of the praetor's edict had not been met; and
that Naevius had not, in fact, taken legal possession at all. The speech is
notable for the exuberance of its rhetorical structure --- the tricolon at
\textsection 95 is among the most elaborate in the corpus --- and for the
glimpse it gives of a young Cicero negotiating, with self-deprecating
courtesy, his stage debut against the man who was already the established
master.

The speech survives without its opening (the partial state of the text
is acknowledged in \textsection 1, where Cicero plunges into the matter
\textit{in medias res}), and one passage near \textsection 85 is supplied
from a later paraphrase. The verdict is unknown; tradition holds that
Cicero won.
