% Philippica IX (philippic-9) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} I could wish, senators, that the immortal gods had brought it about that we were giving thanks to Servius Sulpicius alive, rather than seeking honours for him dead. Nor indeed do I doubt that, if that man had been able to bring back his report on the embassy, his return would have been both welcome to you and a source of safety to the commonwealth --- not that Lucius Philippus and Lucius Piso lacked either zeal or care in so great a duty and so great a charge, but that, since Servius Sulpicius surpassed them in age and all men in wisdom, when he was suddenly torn from the cause he left the whole embassy bereft and crippled.

\ciceroSection{2} But if a just honour has ever been paid to any envoy who died, none more just will be found than in the case of Servius Sulpicius. The others who met death on an embassy set out toward an uncertain peril to their lives, with no fear of death: Servius Sulpicius set out with some hope of reaching Marcus Antonius, none of returning. And though he was in such a state that, if exertion were added to his grave ill health, he had no confidence in himself, he did not refuse to make the attempt --- even with his last breath, if he could bring any aid to the commonwealth. And so neither the violence of winter, nor the snows, nor the length of the journey, nor the roughness of the roads, nor his worsening sickness held him back; and when he had at last come to the meeting and conference with the man to whom he had been sent, in the very midst of his care and study of how to discharge his duty, he departed from life.

\ciceroSection{3} As in other things, so in this, Gaius Pansa, you have acted admirably, in that you both urged us to honour Servius Sulpicius and yourself spoke much and copiously in his praise. To what you have said I would add nothing beyond my motion, did I not think that Publius Servilius, a most distinguished man, must be answered --- he who held that this honour of a statue should be granted to no one except a man killed by the sword while on an embassy. But I, senators, interpret the mind of our ancestors thus: that they held the cause of death, not its manner, was the thing to be looked to. For they willed that a monument should stand to the man for whom the embassy itself had proved fatal, so that in dangerous wars men might undertake the office of envoy more boldly. We must not, then, search out the precedents of our ancestors, but unfold the purpose from which the precedents themselves were born.

\ciceroSection{4} Lars Tolumnius, king of the Veientes, slew four envoys of the Roman people at Fidenae, whose statues stood on the Rostra down to within my own memory: a just honour --- for to those who had met death for the commonwealth our ancestors gave, in exchange for a brief life, an enduring remembrance. We see on the Rostra the statue of Gnaeus Octavius, a famous and great man, the first to bring the consulship into that family which afterwards flourished with the bravest of men. No one then begrudged him his newness; no one failed to honour his merit. And yet the embassy of Octavius was one beneath which lay no suspicion of danger. For when he had been sent by the Senate to look into the dispositions of kings and free peoples, and above all to forbid the grandson of King Antiochus --- the Antiochus who had waged war against our ancestors --- to keep a fleet and to maintain elephants, he was killed at Laodicea, in the gymnasium, by a certain Leptines.

\ciceroSection{5} There was rendered to him then by our ancestors a statue in exchange for his life, which for many years honoured his line, and now alone remains as the memorial of so great a family. And yet to him, and to Tullus Cluvius and Lucius Roscius and Spurius Antius and Gaius Fulcinius, who were cut down by the king of the Veientes, it was not the blood poured out in death but death itself, met for the commonwealth, that brought the honour. And so, senators, if it had been chance that brought Servius Sulpicius his death, I should grieve indeed at so great a wound to the commonwealth, but I should think his death ought to be honoured not with a monument but with public mourning. As it is, who doubts that it was the embassy itself that took his life from him? For he carried his death away with him --- a death which, had he stayed among us, he might have escaped by his own care, by the diligence of an excellent son and a most devoted wife.

\ciceroSection{6} But he, seeing that if he did not obey your authority he would be untrue to himself, and that if he did obey, the charge taken up for the commonwealth's sake would be the end of his life, chose rather to die at the commonwealth's greatest crisis than to seem to have served the commonwealth less than he might have. In many cities along the road he travelled he had the chance to restore and tend himself. There was at hand a generous offer of hospitality, befitting the dignity of so great a man, and the urging of those who had been sent with him that he rest and look to his own life. But he, hurrying, hastening, eager to carry out your commands, held to this resolve while the disease fought against him.

\ciceroSection{7} And when Antony was thrown into the greatest disturbance by his coming --- because the demands served upon him by your order had been framed by the authority and judgment of Servius Sulpicius --- he made plain how he hated the Senate, when he bore the death of the Senate's spokesman with gladness and insolence. Antony, then, killed Servius Sulpicius no less than Leptines killed Octavius, or the king of the Veientes killed the men I named just now: for surely he brought death who was the cause of death. And therefore I think it bears even upon the memory of posterity that there should stand a record of what the Senate's judgment was concerning this war. For the statue itself will be a witness that the war was so grave that the death of an envoy won the remembrance of an honour.

\ciceroSection{8} But if you will recall, senators, the excuse Servius Sulpicius made for undertaking the embassy, no doubt will be left that by honouring him dead we make amends for the wrong we did him living. For you, senators --- it is a hard thing to say, but it must be said all the same --- you, I say, deprived Servius Sulpicius of his life. For when you saw him pleading his sickness more by his very state than by his words, you were not, indeed, cruel --- for what is less in keeping with this order? --- but, since you hoped there was nothing that could not be accomplished by his authority and wisdom, you stood the more vehemently against his excuse, and drove from his resolve a man who had always judged your common consent the weightiest thing.

\ciceroSection{9} But when the urging of the consul Pansa was added --- heavier than the ears of Servius Sulpicius had learned to bear --- then at last he drew his son and me aside, and spoke in such a way that he said he set your authority above his own life. Marvelling at his virtue, we did not dare to oppose his will. His son was moved by an extraordinary devotion; my own grief yielded little to his distress: but each of us was forced to give way before the greatness of his spirit and the weight of his words, when indeed, to the great praise and gratitude of you all, he promised that he would do what you wished, and would not shun the danger of a course of which he himself had been the author. As he hastened to carry out your commands, we escorted him on his way the next morning. And as he parted, he spoke to me in such a way that his words seemed an omen of his fate.

\ciceroSection{10} Give back, then, senators, the life to him from whom you took it. For the life of the dead rests in the memory of the living. Bring it about that the man whom you, unknowing, sent to his death may have immortality from you. If by your decree you set up a statue to him on the Rostra, no forgetfulness of posterity will darken his embassy. For the rest of the life of Servius Sulpicius will be commended to all remembrance by many illustrious monuments. The fame of all mortals will forever celebrate his gravity, his constancy, his good faith, his outstanding care and prudence in guarding the commonwealth. Nor indeed will there be passed over in silence that admirable and incredible and almost divine knowledge of his in interpreting the laws and in unfolding what is fair. If all men of every age who have had in this state an understanding of the law were brought together into one place, they could not be compared with Servius Sulpicius. For he was no more master of the law than of justice.

\ciceroSection{11} Thus the conclusions that proceeded from the laws and the civil code he always referred back to ease and to equity, nor did he prefer to set actions at law on foot rather than to do away with disputes. He has no need, then, of this monument of a statue; he has others, greater. For this statue will be the witness of an honourable death, the other the memorial of a glorious life --- so that it will prove a monument of a grateful Senate more than of a famous man.

\ciceroSection{12} The devotion of the son, too, will be seen to have counted for much toward the father's honour; and though, stricken with grief, he is not present, you ought nonetheless to be of such a mind as if he were present. He is in such a state that no one ever grieved more for the death of an only son than he mourns for his father. And indeed I think it bears even upon the good name of the son of Servius Sulpicius that he should be seen to have rendered his father the honour due. And yet Servius Sulpicius could leave behind no clearer monument than the very image of his own character --- a son of his virtue, his constancy, his devotion, his genius --- whose grief can be lightened either by this honour of yours or by no consolation at all.

\ciceroSection{13} But to me, recalling the many conversations I had with Servius Sulpicius in our friendship, it seems that --- if the dead have any feeling --- a bronze statue, and that one on foot, would be more pleasing to him than a gilded equestrian one, of the kind first set up for Lucius Sulla. For Servius wonderfully loved the restraint of our ancestors, and reproached the extravagance of this age. And so, as if I were to consult the man himself as to what he would wish, I decree a statue of bronze, on foot, as though by his own authority and will: a statue that will lessen and soothe the great grief and longing of the citizens by the honour of a monument. And this proposal of mine, senators, must necessarily be confirmed by the proposal of Publius

\ciceroSection{14} Servilius: for he who held that a tomb ought to be decreed at public expense for Servius Sulpicius did not vote a statue. For if the death of an envoy, without slaughter and the sword, calls for no honour, why does he decree the honour of burial, which can be held the greatest a dead man can have? But if he grants to Servius Sulpicius that which was not given to Gnaeus Octavius, why does he hold that what was given to the one ought not to be given to the other? Our ancestors, to be sure, decreed statues to many, tombs to few. But statues perish through storm and age; the sanctity of tombs, however, lies in the very soil, which can be moved or destroyed by no force; and, as all else is blotted out, so tombs grow more sacred with age.

\ciceroSection{15} Let this man, then, be exalted even by that honour --- a man to whom no honour can be paid that is not his due; let us be grateful in adorning the death of one to whom we can now render no other thanks. Let there be branded, too, the criminal audacity of Marcus Antonius, who wages an abominable war. For with these honours paid to Servius Sulpicius there will remain an everlasting testimony to the embassy that Antony spurned and cast aside. For these reasons I propose as follows: ``Whereas Servius Sulpicius, son of Quintus, of the Lemonian tribe, Rufus, in a most difficult time for the commonwealth, afflicted by a grave and dangerous illness, set the authority of the Senate and the safety of the commonwealth above his own life, and strove against the force and severity of his disease so as to reach the camp of Marcus Antonius, to which the Senate had sent him; and, when he had already come near the camp, overwhelmed by the force of the disease, lost his life at a most critical time for the commonwealth; and whereas his death was in keeping with a life passed most blamelessly and most honourably, in which Servius

\ciceroSection{16} Sulpicius was often of great service to the commonwealth, both as a private citizen and in his magistracies: since such a man met death for the commonwealth while on an embassy, it is the pleasure of the Senate that a bronze statue of Servius Sulpicius, on foot, be set up on the Rostra by the decision of this order; and that around that statue his children and descendants have a space of five feet in every direction at the games and the gladiatorial shows; and that the reason be inscribed upon the base, namely that he met death for the commonwealth; and that Gaius Pansa and Aulus Hirtius, the consuls, one or both, if it seem good to them, instruct the urban quaestors to let out the contract for that base and statue to be made and set up on the Rostra, and that, at whatever price they let it, they see to it that so great a sum be assigned and paid to the contractor. And since the Senate has formerly shown its authority in the funerals and the honours of brave men, it is its pleasure that he be borne forth on his last day as magnificently as possible.

\ciceroSection{17} And since Servius Sulpicius, son of Quintus, of the Lemonian tribe, Rufus, has so deserved of the commonwealth that he ought to be adorned with these distinctions, the Senate resolves, and judges it to be for the good of the commonwealth, that the curule aediles remit, for the funeral of Servius Sulpicius, son of Quintus, of the Lemonian tribe, Rufus, the edict which they have concerning funerals; and that Gaius Pansa the consul assign a site for a tomb in the Esquiline field, or in whatever place shall seem good, thirty feet in every direction, where Servius Sulpicius may be buried; which tomb shall be his own, his children's, and his descendants', under the same right as a tomb granted at public expense under the best title.''
