% Philippica VI (philippic-6) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} I think you have heard, citizens, what was done in the Senate, what each man's opinion was. For the matter, debated since the Kalends of January, was concluded a little while ago --- less severely, indeed, than was fitting, but still not altogether weakly. A delay has been brought to the war, not its cause removed. So what Publius Apuleius asked of me --- a man joined to me by many services and by the closest familiarity, and a great friend to you --- I will answer in such a way that you may learn the things at which you were not present. The most resolute and excellent consuls had cause, on the Kalends of January, to bring the matter of the commonwealth before the Senate first, because of what the Senate, on my motion, decreed on the thirteenth day before the Kalends of January.

\ciceroSection{2} On that day, citizens, for the first time the foundations of the commonwealth were laid: for after a long interval the Senate was so free that you too at last could be free. At that time, even had that day been about to bring me the end of my life, I had received reward great enough, when you all together with one mind and voice cried out that the commonwealth had again been preserved by me. Roused by that so great and so signal judgment of yours, I came into the Senate on the Kalends of January so that I might remember what part you had imposed upon me to sustain. And so, when I saw a nefarious war waged against the commonwealth, I thought no delay should be interposed in pursuing Marcus Antonius, and I judged that this most reckless man --- who, after committing many nefarious acts before, was now in this hour attacking a commander of the Roman people and besieging your most faithful and most valiant colony --- should be pursued by war: I decreed that a state of tumult should be declared; I said that the courts should be suspended, the military cloaks taken up, so that all might more keenly and more weightily press on to the avenging of the wrongs done to the commonwealth, if they saw every mark of the gravest war taken up by the Senate.

\ciceroSection{3} This opinion, citizens, so prevailed for three days that, although no division had been taken, nevertheless except for a few all seemed about to assent to me. But today, by some hope or other thrown in their way, the Senate was more relaxed. For more men followed the opinion that we should test, through legates, how much the authority of the Senate and your unanimity would weigh with Antony. I understand, citizens, that this opinion is rejected by you, and not unjustly. For to whom shall we send legates? To the man who, after squandering and pouring out the public money, after imposing on the commonwealth laws by force and against the auspices, after scattering an assembly, after besieging the Senate, summoned legions from Brundisium to crush the commonwealth --- and, abandoned by them, broke into Gaul with a band of brigands, attacks Brutus, blockades Mutina? What common ground of terms, of fair dealing, of negotiation can you have with this gladiator?

\ciceroSection{4} Although, citizens, that is no legation, but a denunciation of war if he does not obey: for it was so decreed, as though legates were being sent to Hannibal. For men are being sent to give him notice not to attack the consul-elect, not to blockade Mutina, not to lay waste the province, not to hold a levy, to be in the power of the Senate and the Roman people. Easily, of course, will he obey this notice --- so as to be in the power of the senators and in yours, the man who has never been in his own! For what has he ever done by his own judgment? He has always been dragged where lust dragged him, where fickleness, where frenzy, where drunkenness; two unlike kinds of men have always held him, pimps and brigands; he so delights in domestic adulteries and in public murders, that he sooner obeyed a most rapacious woman than the Senate and the Roman people.

\ciceroSection{5} So, what I did a little while ago in the Senate, I will do before you. I testify, I give warning, I foretell that Marcus Antonius will do none of the things which are entrusted to the legates; he will lay waste the fields, he will besiege Mutina, he will hold levies wherever he can. For he is the man who has always despised the judgment and the authority of the Senate, always despised your will and your power. Or will he do what was just now decreed --- that he should lead his army across the river Rubicon, which is the boundary of Gaul, provided that he should bring it no nearer the city of Rome than two hundred miles? Shall he obey this notice, shall he allow himself to be circumscribed by the Rubicon and by two hundred miles?

\ciceroSection{6} Not such a man is Antony; for if he were, the Senate would not have come to give him notice, as it did to Hannibal at the outset of the Punic War, not to attack Saguntum. But that he is being called off from Mutina in such a way as to be kept from the city like a deadly fire --- what ignominy that carries, what a judgment of the Senate! What of this, that the Senate gives the legates instructions to approach Decimus Brutus and his soldiers, and to demonstrate to them that their highest services and benefits to the commonwealth are pleasing to the Senate and the Roman people, and that this thing will be to them for great praise and great honour? Do you suppose Antony will allow the legates to enter Mutina, to depart from there in safety? He will never allow it, believe me. I know his violence, I know his shamelessness, I know his audacity.

\ciceroSection{7} Nor indeed ought we to think of him as we do of any human being, but as of some most savage beast. Since these things are so, what the Senate has decreed is not altogether weak: the legation has some severity about it: would that it had no delay! For while in most matters to be done dilatoriness and procrastination are hateful, in this war above all dispatch is needed. Help must be brought to Decimus Brutus, every force from every quarter must be collected; we cannot, without crime, exhibit any delay in the liberation of so great a citizen.

\ciceroSection{8} Or could he not, had he judged Antony consul, and Gaul Antony's province, have handed over his legions and the province to Antony, gone home, celebrated a triumph, given his opinion first in this order until he entered upon office?

\ciceroSection{9} What was there in it? But because he remembered that he was a Brutus, and was born for your liberty and not for his own leisure --- what else did he do but, almost with his own body, keep Antony from Gaul? Should legates have gone to such a man, or legions? But let us pass over what is past: let the legates make haste, as I see they are going to do; you, prepare your cloaks. For it has been so decreed that, if Antony has not obeyed the authority of the Senate, we go to the cloaks. To them we shall go; he will not obey: we shall lament that so many days have been lost for action. I do not fear, citizens, that, when Antony has heard that I have affirmed this both in the Senate and in the assembly --- that he will never be in the power of the Senate --- he, to confute me, so that I may seem to have foreseen nothing, will turn about and obey the Senate. He will never do it; he will not begrudge me this glory; he would rather have me thought wise by you than himself thought temperate.

\ciceroSection{10} What? Even if he himself wished it, do we suppose his brother Lucius would suffer it? Lately, indeed, at Tibur, I think, when Marcus Antonius seemed to him to waver, he is said to have threatened his brother with death. Will even from this Asiatic murmillo the mandates of the Senate, the words of legates, be heard? For he will not be able to be sundered from his brother, especially with so great a weight of authority. For this man, among those of Antony's circle, is an Africanus: he is reckoned more than Lucius Trebellius, more than Titus Plancus, than Extitius, a noble youth. Plancus, indeed --- who, condemned by every vote amid your loudest applause, somehow flung himself into the crowd and so returned in mourning that he seemed dragged back rather than restored --- Lucius despises as if Plancus had been interdicted from fire and water: he says at last that there ought to be no place in the curia for the man who set fire to the curia.

\ciceroSection{11} For Trebellius he now greatly cherishes; he used to hate him, when Trebellius opposed the abolition of debts; now he carries him in his eyes, after he saw that Trebellius himself could not be safe without the abolition of debts. For I think you have heard, citizens, what you have even been able to see --- that every day the sureties and creditors of Lucius Trebellius gather together. O Faith! --- for this, I think, is the cognomen Trebellius assumed --- what greater faith can there be than to defraud your creditors, to flee from your house, to go to arms because of debt? Where is that applause at the triumph, often at the games, where is the aedileship conferred with the highest enthusiasm of the good? Who is there who does not think that this man acted rightly by accident, and criminally by his own depravity?

\ciceroSection{12} But I return to your darling and delight, Lucius Antonius, who has taken you all into his trust. You deny it? Surely there is no one of you who does not have a tribe? Certainly none. And yet thirty-five tribes have adopted that man as their patron. Do you cry out again? Look at that gilded equestrian statue on your left, on which what is inscribed? \textit{To their patron from the thirty-five tribes.} Then Lucius Antonius is the patron of the Roman people. An evil plague upon him! --- for I assent to your cry. Not only is this brigand a man whom none would wish to have as a client, but who was ever of such great means, of such great deeds, that he would dare to call himself the patron of the Roman people, victor and lord of all nations?

\ciceroSection{13} In the Forum we see a statue of Lucius Antonius, as we see that of Quintus Tremulus, who conquered the Hernici, in front of the temple of Castor. O incredible shamelessness! Did he take so much upon himself, because at Mylasa, as a murmillo, he cut the throat of a Thracian, his own friend? In what way could we bear this man, if in this Forum, with you watching, he had fought to the death? But this is only one statue. The other is from the Roman equestrians equipped at public expense, who likewise add the inscription, \textit{to their patron.} When did that order ever adopt a patron? If anyone, it should have been me. But I leave myself aside; what censor, what commander did they ever so adopt? He divided land among them. O sordid men, who took it; wicked man, who gave!

\ciceroSection{14} The military tribunes who had served twice in Caesar's army set up a statue too. What sort of order is this? There were many, in many legions, over so many years. To these too he divided Semurium. The Campus Martius was the one thing left --- unless he had first fled with his brother. But this allotment of lands, a little while ago, citizens, was undone by the vote of Lucius Caesar, that most illustrious man and most outstanding senator: for in agreement with him we annulled the acts of the septemvirs. The benefits of Nucula lie void; their patron Antony grows cold. For the possessors will go away with the easier mind: they had made no outlay; they had not yet stocked it, partly because they were not confident, partly because they had no means.

\ciceroSection{15} But that triumphant statue, of which, in better times, I could not have spoken without laughing: \textit{To Lucius Antonius, from the Middle Janus, their patron.} Is that so? Is the Middle Janus now in Lucius Antonius's clientship? Who was ever found in that Janus to enter on his books a thousand sesterces lent to Lucius Antonius? But too much about trifles: let us return to the cause and the war; though it was not inappropriate that certain personages should be reviewed by you, so that you might be able silently to consider with whom this war is to be waged. For my part I urge you, citizens, even if something else were better, nevertheless to await the return of the legates with an even mind. Speed has been taken from the cause; nevertheless some good has been added to the cause.

\ciceroSection{16} For when the legates have reported back what they will certainly report --- that Antony is not in your power, not in the Senate's --- what citizen will be so wicked as to think he ought to be held a citizen? For now there are a few, but still more than is worthy of the commonwealth, who speak in this fashion: \textit{Shall we not even wait for the legates?} The very state of the commonwealth itself will surely wring that voice and pretence of clemency out of them. For which reason --- to confess it to you, citizens --- on today's day I contended less, I laboured less to have the Senate, in agreement with me, decree a state of tumult, order the cloaks to be taken up. I preferred to be praised by all twenty days from now for my opinion than to be reviled by a few today.

\ciceroSection{17} Therefore, citizens, await the return of the legates and swallow the annoyance of a few days. When they have come back, if they bring peace, judge me eager for it; if war, judge me foreseeing. Or shall I not see ahead for my fellow citizens, shall I not think day and night about your liberty, about the safety of the commonwealth? For what do I not owe you, citizens, who, sprung as I am from myself, preferred me to the most noble men in every office? Am I ungrateful? Who less so? I, who, once offices had been gained, performed in the Forum the same labours which I had performed in seeking them. Unpractised in public life? Who more exercised? I, who for now twenty years have waged war with impious citizens. Wherefore,

\ciceroSection{18} citizens, with counsel as far as I am able, and with labour all but more than I am able, I shall watch and keep watch for you. For who is the citizen --- especially in this rank in which you have wished me to stand --- so forgetful of your kindness, so unmindful of his country, so hostile to his own dignity, that this so great unanimity of yours does not rouse him, does not inflame him? Many and great assemblies I held as consul; in many I was present: never have I seen one so great as yours is now. You all feel one thing, you are all eager for one thing --- to turn aside Marcus Antonius's enterprises from the commonwealth, to extinguish his frenzy, to crush his audacity. All orders want the same thing; the municipalities, the colonies, all Italy lean the same way. And so a Senate firm of its own accord you have made firmer by your authority.

\ciceroSection{19} The time has come, citizens --- later, to be sure, than was fitting for the Roman people, but nevertheless so ripe that it can no longer be put off by even an hour. There was, so to speak, some fatal mischance which we endured, since it had in some fashion to be endured: now if there is any, it will be voluntary. It is not divinely lawful for the Roman people to be slaves --- whom the immortal gods willed to rule over all nations. The matter has been brought to the extremest crisis; the decision is upon liberty. Either you must conquer, citizens --- which surely both by your piety and by this so great concord you will accomplish --- or you must endure anything rather than be slaves. Other nations can bear slavery; liberty belongs to the Roman people by right.
