% Philippica VIII (philippic-8) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Yesterday's business was conducted more confusedly, Gaius Pansa, than the practice of your consulship demanded. You seemed to me to hold out too weakly against those to whom you are not accustomed to yield. For though the Senate's courage was what it usually is, and though all saw that in fact there was a war, and though some thought that this word ought to be removed, your own inclination, in the division, leaned toward leniency. And so, on account of the harshness of a word, with you as its author, our motion was defeated; that of Lucius Caesar, a most distinguished man, prevailed --- who, with the word's ferocity taken away, was milder in his speech than in his vote. Although he, indeed, before he gave his opinion, excused himself on grounds of kinship. He had done the same thing, when I was consul, in the case of his sister's husband, as he did at this time in the case of his sister's son: so that he might both be moved by his sister's grief and provide for the safety of the Roman people.

\ciceroSection{2} And yet Caesar himself in a way instructed you, senators, not to assent to him, when he said this: that he would have delivered a different opinion, one worthy of himself and of the commonwealth, were he not hindered by kinship. So he is the uncle: are you too, then, uncles, who assented to him? But what was the dispute about? Some thought the name of war ought not to be set down in the motion; they preferred to call it a tumult --- men ignorant not only of the facts but even of the words. For there can be a war such that there is no tumult; but a tumult cannot exist without a war.

\ciceroSection{3} For what else is a tumult but a disturbance so great that a greater fear arises? --- from which the very name ``tumult'' is derived. And so our ancestors named only the Italian tumult, because it was at home, and the Gallic tumult, because it was on Italy's border, and no other besides. That a tumult is graver than a war can be understood from this: that in a war exemptions hold good, but in a tumult they do not. So it comes about, as I have said, that there can be a war without a tumult, but no tumult without a war.

\ciceroSection{4} For indeed, since there is nothing intermediate between war and peace, it must follow that a tumult, if it is not a part of war, is a part of peace --- than which what could be said or supposed more absurd? But too much about a word: let us rather look at the matter, senators --- which, I am aware, is sometimes made worse by the word. We do not want this to seem a war. What authority, then, do we give the municipalities and the colonies for shutting Antony out? What, that soldiers be raised without compulsion, without penalty, by zeal and goodwill? What, that they pledge monies to the commonwealth? For if the name of war is taken away, the zeal of the municipalities will be taken away; and the consensus of the Roman people, which has already committed itself to the cause, must necessarily be weakened, if we grow faint.

\ciceroSection{5} But why say more? Decimus Brutus is under assault: is it not war? Mutina is besieged: is not even this war? Gaul is being laid waste: what peace could be more secure! And that one --- who can call it war? We have sent a consul, a most valiant man, with an army --- a man who, although he was weak from a grave and lingering illness, thought no excuse ought to be allowed him when he was being summoned to the defence of the commonwealth. Gaius Caesar, indeed, did not wait for your decrees, especially since that step belonged to his time of life: he took up the war against Antony of his own accord. For the time for decreeing was not yet come; but if he had let slip the time for waging the war, he saw that, with the commonwealth crushed, nothing could be decreed.

\ciceroSection{6} So these men now, and their armies, are living in peace! He is no enemy whose garrison Hirtius dislodged from Claterna; he is no enemy who stands armed against a consul, who assaults a consul-designate; and those words are neither hostile nor warlike which Pansa read out a moment ago from his colleague's despatch: ``I have dislodged the garrison; I have taken Claterna; the cavalry are routed; battle is joined; some are killed.'' What peace could be greater? Levies have been decreed throughout all Italy, with exemptions abolished; tomorrow the war-cloaks will be taken up; the consul has said that he will go down to the Forum with an armed guard.

\ciceroSection{7} Is this not a war --- or is it so great a war as never yet was? For in the other wars, and above all the civil ones, the struggle was waged over the commonwealth itself. Sulla against Sulpicius, over the validity of the laws which Sulla said had been passed by force; Cinna against Octavius, over the votes of the new citizens; again Sulla against Marius and Carbo, that the unworthy might not lord it, and that the most cruel murder of the most illustrious men might be punished. The causes of all these wars were born from a contest over the commonwealth. Of the most recent civil war I do not care to speak: I do not know its cause, I detest its outcome.

\ciceroSection{8} This is the fifth civil war being waged --- and all of them have fallen within our own lifetime --- but the first to be waged not amid dissension and discord among citizens but amid the greatest agreement and an incredible concord. All want the same thing, defend the same thing, feel the same thing. When I say all, I except those whom no one thinks worthy of citizenship. What, then, is the cause set at the heart of this war? We defend the temples of the immortal gods, we the walls, we the homes and dwellings of the Roman people, the altars, the hearths, the tombs of our ancestors; we defend the laws, the courts, liberty, our wives, our children, our country: against this Marcus Antonius strives, for this he fights --- to throw all these things into confusion, to overturn them, to count the plunder of the commonwealth the very object of the war, to scatter our fortunes in part and in part to parcel them out among parricides.

\ciceroSection{9} In this so unequal kind of war, the most wretched thing is this: that he promises his bandits, first, houses --- for he affirms that he will divide up the city among them; next, that through whatever gate they please he will lead them out to wherever they wish. All the Cafos, all the Saxas, and the rest of the plagues that follow Antony are marking out for themselves the finest mansions, the gardens, the estates at Tusculum and at Alba. And even the country folk --- if those creatures are men and not rather cattle --- are carried along by empty hope as far as the watering-places and Puteoli. So Antony has something to promise his men. What of us? Have we anything of the kind? May the gods grant us better! For this is the very thing we are working at: that no one hereafter be able to promise anything of the sort. Unwillingly I say it, but it must be said: Caesar's auction-spear, senators, brings to many wicked men both hope and audacity. For they have seen men become rich in an instant out of beggars: and so they are forever longing to see the spear --- those men who threaten our goods, to whom Antony promises everything.

\ciceroSection{10} What? What do we promise our armies? Things far better and greater. For the promise of crimes is ruinous both to those who await it and to those who make it: we promise our soldiers liberty, laws, rights, courts, the empire of the whole earth, dignity, peace, tranquillity. Antony's promises, then, are bloody, foul, criminal, hateful to gods and men, neither lasting nor wholesome: ours, by contrast, are honourable, unblemished, glorious, full of gladness, full of devotion.

\ciceroSection{11} Here, too, Quintus Fufius --- a brave and energetic man, my friend --- recounts to me the advantages of peace. As though, in fact, if peace were the thing to be praised, I could not do it just as aptly. For have I defended peace only once? Have I not always pursued tranquillity? --- a thing that, useful to all good men, is so especially to me. For what course could my own industry have held without cases in the Forum, without laws, without courts? --- which cannot exist once civil peace is taken away.

\ciceroSection{12} But I ask you, Calenus, what of you? Do you call slavery peace? Our ancestors, indeed, took up arms not only that they might be free but that they might rule: do you think arms should be thrown down so that we may be slaves? What juster cause is there for waging war than the throwing-off of slavery? In which condition, even if the master should not be troublesome, it is still most wretched that he could be, if he chose. Nay, other causes may be just; this one is necessary. Unless perhaps you think this does not concern you, because you hope to be a partner in Antony's domination. In which you slip twice: first, because you thrust your own interests in among the common ones; second, because you think anything stable or pleasant exists in a despotism. Even if it has served you before, it will not serve you forever.

\ciceroSection{13} Why, you used even to complain about that man --- what do you think you will do about a beast? And you say you are one who has always wished for peace, always wished all citizens safe. An honourable speech --- but only so if you mean good citizens, useful ones, citizens loyal to the commonwealth; but if you would have safe those who are citizens by nature, enemies by will, what then is the difference between you and them? Your father, indeed --- whom in my youth I had as an aged authority, a stern and prudent man --- used to give the first place among all citizens to Publius Nasica, who killed Tiberius Gracchus: by that man's courage, counsel, and greatness of spirit he judged the commonwealth had been set free.

\ciceroSection{14} What? Have we received anything different from our fathers? So this citizen, if you had lived in those times, would not have won your approval, because he had not wished all men safe. ``Whereas Lucius Opimius the consul has spoken concerning the commonwealth, concerning that matter they have decreed thus: that Lucius Opimius the consul should defend the commonwealth.'' This the Senate did in words; Opimius, in arms. Would you, then, if you had been there, think him a reckless citizen, or a cruel one --- or Quintus Metellus, whose four sons were consulars; or Publius Lentulus, princeps senatus; or the several other men of the highest rank who, armed, alongside Opimius the consul, pursued Gracchus onto the Aventine? In which battle Lentulus took a grave wound; Gracchus was killed, and Marcus Fulvius, a consular, and his two young sons. Those men, then, are to be blamed --- for, you see, they did not wish all citizens safe.

\ciceroSection{15} Let us come to nearer instances. In the consulship of Gaius Marius and Lucius Valerius, the Senate gave the commonwealth into their charge to be defended: Lucius Saturninus, tribune of the plebs, and Gaius Glaucia, praetor, were killed. On that day all the Scauri, the Metelli, the Claudii, the Catuli, the Scaevolae, the Crassi took up arms. Do you think those consuls, or those most illustrious men, are to be blamed? I wished Catiline to perish. Did you, who would have all men safe, wish Catiline safe? This is the difference, Calenus, between my view and yours. I do not wish any citizen to commit a thing for which he must be punished with death; you think that, even if he has committed it, he must be preserved. In the body, if there is anything of such a kind that it harms the rest of the body, we allow it to be burned and cut away, so that some single limb may perish rather than the whole body. So in the body of the commonwealth: that the whole may be safe, let whatever is pestilent be cut off.

\ciceroSection{16} A hard saying! But that one is much harder: ``Let the wicked be safe, the criminal, the impious; let the innocent be destroyed, the honourable, the good --- the whole commonwealth!'' In one man, Quintus Fufius, I confess you saw more than I did. I judged Publius Clodius a ruinous citizen, criminal, lustful, impious, bold, villainous; you, on the contrary, judged him holy, temperate, innocent, modest --- a citizen to be kept and longed for. In this one thing I grant that you saw most clearly, and that I erred greatly. But as for your saying that I am accustomed to deal with you angrily --- it is not so. That I deal vehemently, I confess; angrily, I deny. I do not, on the whole, lightly grow angry with friends, not even when they deserve it.

\ciceroSection{17} And so I can dissent from you without abuse of words; without the deepest pain of mind I cannot. For is my disagreement with you a small one, or over a small matter? Do I favour this man, you that one? No --- rather, I favour Decimus Brutus, you Marcus Antonius: I desire that a colony of the Roman people be preserved, you are eager that it be stormed. Can you deny this --- you who interpose every delay by which Brutus may be weakened and Antony made stronger? For how long will you say that you want peace? The thing is being carried on; the siege-sheds have been brought up; the fighting is most fierce. To intervene, we sent three of the foremost men of the state. These Antony scorned, rejected, repudiated: yet you persist as Antony's most steadfast defender.

\ciceroSection{18} And indeed, that he may seem the better senator, he denies that he is bound to be Antony's friend: that, although Antony was under a great obligation to him, Antony came out against him. See how great is his love of country: though he is angry with the man, he yet defends Antony for the commonwealth's sake. As for you, Quintus Fufius, when you are so bitter against the people of Massilia, I do not hear you with an even mind. For how long will you assail Massilia? Does not even the triumph make an end of the war --- the triumph in whose procession that city was borne, the city without which our ancestors never triumphed over the Transalpine nations? At which time, indeed, the Roman people groaned: for although all had griefs proper to their own affairs, yet there was no citizen who reckoned the miseries of that most loyal community foreign to himself.

\ciceroSection{19} Caesar himself, who had been most angry with them, nevertheless, on account of that community's singular dignity and loyalty, remitted each day some part of his anger: can a community so loyal not content you by any disaster of its own? Again, perhaps, you will say I am growing angry. But I say everything without anger, and yet not without pain of mind: I judge that no one is an enemy to that community who is a friend to this one. What your reasoning is, Calenus, I cannot work out. Before, we could not deter you from being a man of the people; now we cannot entreat you to be one. Enough, and more, with Fufius --- all of it without hatred, none of it without pain. But I suppose that a man who bears a son-in-law's complaint with moderation will bear a friend's with an even mind.

\ciceroSection{20} I come to the remaining consulars, of whom there is none --- and I say this by my own right --- who does not have with me some bond of goodwill: with some the greatest, with some a moderate one, with none none at all. How shameful a dawn yesterday broke for us --- for us consulars, I mean! Legates again? As though he were making a truce? Before the very faces and eyes of the legates he battered Mutina with his artillery; he was displaying his siege-works and his fortification to the legates; not for a single point of time, while the legates were present, did the assault draw breath. To this man, legates? Why? So that by their return you may be more violently terrified?

\ciceroSection{21} For my part, although I had voted against the sending of legates, I yet consoled myself with this: that when they came back, scorned and rejected by Antony, and reported to the Senate that he not only had not departed from Gaul, as we had decreed, but had not even withdrawn from Mutina, and that they had had no opportunity of meeting Decimus Brutus --- then it would come to pass that all of us, inflamed with hatred, roused by indignation, would come to Decimus Brutus's aid with arms, with horses, with men. But we have grown even more languid since, after we have looked closely at not only the audacity and crime of Marcus Antonius but his insolence and arrogance as well.

\ciceroSection{22} Would that Lucius Caesar were in health, that Servius Sulpicius were alive: this cause would be pleaded far better by three than it is now pleaded by one. I will say this in sorrow rather than in reproach. We have been deserted --- deserted, I say, senators --- by our leading men. But --- I have said it often already --- in a peril so great, all who think rightly and bravely will be consulars. The legates ought to have brought us spirit: they brought us fear --- though to me, indeed, none --- however well they think of the man to whom they were sent, from whom, moreover, they accepted instructions.

\ciceroSection{23} By the immortal gods! Where is that custom and courage of our ancestors? Gaius Popilius, among our forefathers, when he had been sent as legate to King Antiochus and had announced, in the words of the Senate, that he should withdraw from Alexandria, which he was besieging --- when the king kept putting off the time --- with a little rod drew a circle around him as he stood, and said that he would report to the Senate unless the king had answered him, before he stepped out of that circle, what he meant to do. Splendid! For he had brought with him the very face of the Senate and the authority of the commonwealth. The man who does not obey that authority --- instructions are not to be accepted from him; rather, he himself is to be repudiated.

\ciceroSection{24} Should I accept instructions from a man who scorns the Senate's instructions? Or judge that he has anything in common with the Senate --- a man who besieges a commander of the Roman people while the Senate forbids it? But what instructions! With what arrogance, what stupidity, what haughtiness! And why was he giving them to our legates, when he was sending Cotyla to us --- the ornament and citadel of his friends, a man of aedilician rank? If indeed he was an aedile at the time when, by Antony's order, the public slaves flogged him with their straps at a banquet.

\ciceroSection{25} But how modest the instructions! We are men of iron, senators, if we deny this man anything. ``Both provinces,'' he says, ``I give back: I lay down my army: I do not refuse to be a private citizen.'' For these are his very words. He seems to be coming round. ``I forget everything; I return to goodwill.'' But what does he add? ``If you will give to my six legions, to my cavalry, to my praetorian cohort, estates and land.'' He demands rewards even for those for whom, if he were to ask mere pardon, he would be judged the most shameless of men. He adds besides that the lands which he himself, together with Dolabella, has granted, those to whom they were given should keep.

\ciceroSection{26} Here is the Campanian land and the Leontine, which two our ancestors counted the refuges of the grain-supply. He makes provision for actors, for gamblers, for pimps; he makes provision, too, for Cafo and Saxa, those pugnacious and brawny centurions whom he has stationed amid the troops of male and female actors. He demands besides that the sum total of his autographs and notebooks, and the decrees of his colleague, should stand. Why does he trouble himself that each man should keep what he has bought, if the man who sold it keeps what he received? And that the accounts at the temple of Ops be not touched: that is, that the seven hundred million be not recovered; that it be no crime in the Septemviri that they did what they did. Nucula, I suppose, prompted this: he was afraid, perhaps, of losing so great a body of clients. He wants provision made, too, for those who are with him, for whatever they have done against the laws. For Mustela and Tiro he looks ahead: for himself he takes no trouble.

\ciceroSection{27} For what has he ever committed? Has he laid a hand on public money, or killed a man, or kept armed men about him? But why should he trouble himself about these men? For he demands that his judiciary law be not repealed. With that obtained, what has he to fear? That one of his men should be condemned by a Cydas, a Lysiades, a Curius? And yet he does not press us with many demands; he remits something, and relaxes. ``The toga-clad Gaul,'' he says, ``I give back; the long-haired I demand'' --- at his ease, you see, he prefers to be --- ``with six legions,'' he says, ``and those filled up out of Decimus Brutus's army,'' not merely from his own levy; and he is to hold it for so long as Marcus Brutus and Gaius Cassius shall hold their provinces as consuls or proconsuls. At the elections for this province his brother Gaius --- for it is his year --- has already met with a rebuff. ``But I myself,'' he says, ``am to hold it for five years.''

\ciceroSection{28} But that the law of Caesar forbids --- and you defend the acts of Caesar. These demands, Lucius Piso, and you, Lucius Philippus, leading men of the state --- could you, I will not say bear in spirit, but receive with your ears? But, as I suspect, there was a certain terror: you were with him neither as legates nor as consulars; you could maintain neither your own dignity nor that of the commonwealth. And yet, somehow --- by a certain wisdom, I suppose, which I could not have shown --- you came back not too greatly angered. To you Marcus Antonius granted nothing --- to you, men of the highest distinction, legates of the Roman people: yet what did we not concede to Cotyla, Marcus Antonius's legate? To one for whom there was no right that the gates of this city should stand open, to him this temple stood open; to him there was admittance to the Senate; he, yesterday, was entering your opinions and every word in his note-tablets; and to him even men who have held the highest honours were peddling themselves, against their own dignity.

\ciceroSection{29} O immortal gods! How great a thing it is to maintain the role of a leading man in the commonwealth! --- a role that ought to serve the citizens not only in their minds but in their very eyes. To receive an enemy's legate into one's house, to admit him to one's chamber, even to draw him aside, is the act of a man who thinks nothing of his dignity and too much of his danger. But what is the danger? For if it comes to the utmost crisis, either liberty is prepared for the victor or death set before the vanquished: of which the one is to be wished for, the other no one can escape. But a shameful flight from death is worse than any death. For this, indeed, I am not brought to believe ---

\ciceroSection{30} --- that there are certain men who envy another's constancy, who resent his toil, who take it ill that an unbroken will to aid the commonwealth wins approval from both the Senate and the Roman people. All of us, indeed, ought to do this, and it was --- not only among our ancestors but even lately --- the highest praise of the consulars: to keep watch, to be present in spirit, always to think or to do or to say something for the commonwealth.

\ciceroSection{31} I remember, senators, Quintus Scaevola the augur, in the Marsic War, when he was of the greatest age and of broken health, every day, as soon as it grew light, giving to all the opportunity of meeting him: nor did anyone in that war see him in his bed, and, old and feeble as he was, he was the first to come into the Senate-house. His industry I should most wish to be imitated by those who ought to imitate it; but, failing that, that they should not envy another's toil.

\ciceroSection{32} For indeed, senators, since after six years we have entered upon the hope of liberty, and have endured slavery longer than thrifty and diligent captives are wont to do, what watches, what anxieties, what toils ought we to refuse for the sake of freeing the Roman people? For my own part, senators, although those who have held this honour are accustomed to wear the toga even when the state is in war-cloaks, I have nevertheless resolved not to differ in dress from you and the rest of the citizens, in so great an atrocity of the time and so great a disturbance of the commonwealth. For we consulars are not conducting ourselves in this war in such a way that the Roman people will look with an even mind upon the insignia of our honour --- when some of us are so timid that they have cast away all memory of the Roman people's benefactions; some so turned away from the commonwealth that they openly show that they favour the enemy, that they easily endure our legates being scorned and mocked by Antony, and would have Antony's legate relieved. For they were saying that this man ought not to be barred from returning to Antony, and in the matter of receiving the same man they were correcting my motion. To them I will give way. Let Varius return to his commander --- but on this condition, that he never come back to Rome. To the rest, however, if they lay aside their error and return to goodwill with the commonwealth, I think pardon and impunity should be granted.

\ciceroSection{33} For these reasons I move as follows: that of those who are with Marcus Antonius, whoever shall have departed from arms and shall have come over, before the next Ides of March, either to the consuls Gaius Pansa or Aulus Hirtius, or to Decimus Brutus, commander and consul-designate, or to Gaius Caesar, propraetor --- to them let it be held no fault that they were with Marcus Antonius. If any of those who are with Marcus Antonius shall have done anything that seems worthy of honour or reward, that the consuls Gaius Pansa and Aulus Hirtius, one or both, if it seem good to them, refer the matter of his honour or reward to the Senate on the earliest possible day. If anyone, after this decree of the Senate, shall have set out to join Antony --- except Lucius Varius --- the Senate will judge that he has acted against the commonwealth.
