% Philippica V (philippic-5) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Never, senators, have these Kalends of January seemed longer in coming to me; and the same thing, I understood, seemed true to every one of you these last days. For those who wage war against the commonwealth were not waiting for this day; whereas we, just when most of all our counsel ought to come to the aid of the common safety, were not being called into the Senate. But the address of the consuls has lifted away that complaint of the days now past --- for they have spoken in such a way that the Kalends seem rather long-wished-for than late in coming. And as the consuls' address has lifted up my spirit and brought hope not only of preserving our safety but even of recovering our former dignity, so I should have been thrown into confusion by the proposal of the man who was asked his opinion first, had I not had confidence in your valour and your firmness.

\ciceroSection{2} For this day, senators, has dawned upon you, this power has been granted to you, that you may declare to the Roman people how much valour, how much firmness, how much weight there is in the counsel of this order. Recall what day it was thirteen days ago, what consent there was among you, what valour, what firmness; how great the praise, how great the glory, how great the gratitude you have won from the Roman people. And on that day, senators, you so resolved the matter that nothing now remains open to you except an honourable peace or a necessary war.

\ciceroSection{3} Does Marcus Antonius want peace? Let him lay down his arms, let him ask, let him sue for it. He will find no one more even-handed than me, with whom, while he was commending himself to impious citizens, he preferred to be an enemy than a friend. There is nothing, surely, that can be granted to a man waging war; there will perhaps be something that can be conceded to a man asking for it. But to send legates to one against whom you passed the gravest and severest judgment thirteen days ago is no longer mere fickleness, but, to speak my mind, derangement. First you praised the commanders who had undertaken war against him on their own initiative; then the veteran soldiers who, having been settled in colonies by Antony, set the liberty of the Roman people before his bounty.

\ciceroSection{4} What of the Martian Legion? What of the Fourth? Why are they praised? For if they abandoned their own consul, they are to be reproached; if they abandoned an enemy of the commonwealth, they are rightly praised. And yet, when you did not yet have consuls, you decreed that the question of rewards for the soldiers and honours for the commanders should be brought up at the earliest opportunity. Does it please you, at one and the same time, to grant rewards to those who have taken up arms against Antony and to send legates to Antony? So that it will now be a thing to blush at, that the legions decreed more honourably than the Senate: for the legions decreed to defend the Senate against Antony, and the Senate decrees legates to Antony. Is this to confirm the spirit of our soldiers, or to weaken their valour?

\ciceroSection{5} This is what twelve days have achieved: that the man whom none was found to defend, save Cotyla, now has even consular patrons. I wish they were all asked their opinion before me! --- though I suspect what some of them, who will be asked after me, are about to say --- I could speak against them more easily if anything were apparent. For there is a notion that someone will decree to Marcus Antonius that further Gaul which Plancus now holds. What else is this than to lavish upon the enemy every weapon for civil war: first the sinews of war, money without limit, of which he now stands in need; then cavalry, as much as he wishes? Cavalry, do I say? He will hesitate, no doubt, to bring with him the barbarian nations! He who does not see this is brainless; he who, seeing, decrees it, is impious.

\ciceroSection{6} Will you equip a wicked and abandoned citizen with the money, the infantry, the cavalry, the resources of the Gauls and the Germans? These excuses count for nothing: ``he is my friend.'' Let him be the country's friend first. ``He is my kinsman.'' Can any kinship be closer than that of one's country, in which parents themselves are also contained? ``He gave me money.'' I should like to see the man who would dare to say this. But when I have laid open what is at stake, it will be easy to settle what opinion you should pronounce, or which you should follow. What is at stake is whether the power shall be given to Marcus Antonius to crush the commonwealth, to butcher honest men, to make a present of the City and the lands to his brigands, to weigh down the Roman people with slavery --- or whether he shall be permitted to do none of these things. Hesitate about what you should do! ``But these charges do not apply to Antony.''

\ciceroSection{7} Not even Cotyla would dare to say that. For what charge does not apply to the man who, while saying that he defends the acts of Caesar, has himself overturned those laws of his which we could most have praised? Caesar wished to drain the marshes; this man has handed over the whole of Italy to be divided up by a moderate fellow, Lucius Antonius. What? Did the Roman people accept this law? What? Could it be carried under the auspices? But our modest augur acts on the auspices without his colleagues. And yet those auspices do not need an augur's interpretation: who does not know that, when Jupiter thunders, it is not lawful to transact business with the people? The tribunes of the plebs carried a law on the provinces against the acts of Gaius Caesar: Caesar's term was two years, theirs was six. Did the Roman people accept this law too? What? Was it promulgated? What? Was it not passed before it was written down? What? Did we not see the deed done before anyone could suspect that it was about to be done? Where is the Lex Caecilia et Didia,

\ciceroSection{8} where is the promulgation of three market-days, where is the penalty under the recent Lex Iunia et Licinia? Can these laws stand without the death of the rest of the laws? Was there a way for anyone to make his way into the Forum? And what thunderings further, what a tempest! So that, even if the auspices did not move Marcus Antonius, it would still seem strange that he could bear up under and sustain so great a force of storm, of rain, of whirlwinds. The law, then, which an augur says he carried not only with Jupiter thundering but with what was almost a heavenly outcry forbidding it --- can he hesitate to admit that this was carried in defiance of the auspices?

\ciceroSection{9} What? When the colleague with whom he carried it was the very colleague whom he himself had made flawed by his own report --- did our good augur think this had no bearing on the auspices? But of the auspices we, perhaps, shall be the interpreters, who are colleagues of his own: are we then to seek out interpreters of arms as well? In the first place, all the entrances to the Forum were so blocked that, even if no armed man stood in your way, you still could not have entered the Forum unless the barriers were torn down. So in fact were the guards posted: as the approaches of enemies are barred from a city by fortlets and earthworks, just so you could see the people and the tribunes of the plebs driven back from entering the Forum.

\ciceroSection{10} For these reasons I vote that all the laws which Marcus Antonius is said to have carried were carried by force and against the auspices, and that the people is not bound by these laws. If Marcus Antonius is said to have carried any law confirming the acts of Caesar, or abolishing the dictatorship in perpetuity, or settling colonies on lands, I vote that the same laws be carried afresh under the auspices, that they may bind the people. For however good the matter, he carried it improperly and by violence; nevertheless those laws are not to be held valid, and all the audacity of an insane gladiator must be repudiated by our authority.

\ciceroSection{11} But that squandering of public money is on no account to be borne, by which he made off with seven hundred million sesterces in false entries and donations, so that it seems like a portent that so great a sum of the Roman people could vanish in so short a time. What of those countless monstrous gains which the house of Marcus Antonius drained dry? He sold forged decrees; kingdoms, citizenships, immunities --- once the money was taken --- he ordered to be incised in bronze. He used to say that he was conducting all this from the memoranda of Gaius Caesar, of which he himself was the author. Inside the house, the markets of the whole commonwealth were busy; a woman more fortunate for herself than for her husbands held an auction of provinces and kingdoms; exiles were restored as it were by law without a law. Unless these things are rescinded by the authority of the Senate, now that we have entered upon the hope of recovering the commonwealth, no image of a free state will be left.

\ciceroSection{12} Nor was it only by fictitious memoranda and saleable autographs that uncountable wealth was heaped up in that house --- with Antony saying, of what he sold, that he was acting from the acts of Caesar; he was also entering as false decrees of the Senate matters for which money had been taken; bonds were sealed; decrees of the Senate which had never been made were filed in the treasury. To the disgrace of this even foreign nations were witness. Treaties meanwhile were being made, kingdoms granted, peoples and provinces declared free, and the false records of these very matters, with the Roman people groaning, were posted up all over the Capitol. By these means so great a sum of money was piled up in one house that, if money of this kind were called in by law, the commonwealth would never lack money. He also carried a law on the courts --- the man of chaste and unimpaired life, the authority on courts and on law. In this he deceived us. He used to say that he had set up as judges centurions, legionaries, and ``larks'': but in fact he chose gamblers, he chose exiles, he chose Greeks --- O glorious bench of judges! O admirable dignity of council!

\ciceroSection{13} My mind longs to plead the defence of an accused man before that council! I am fond of Cydas of Crete, the prodigy of his island, a man of the most reckless and abandoned sort. But suppose he were not so: does he know Latin? Is he of the type and form of a judge? Does he know --- which is the greatest thing --- our laws and customs? Does he, finally, know our men? For Crete, gentlemen, is better known to you than Rome is to Cydas. The selection and scrutiny of judges, even among our own citizens, used to be done. But who knows --- or could know --- one Cortynius as a judge? For Lysiades the Athenian most of us do know: he is the son of Phaedrus, the celebrated philosopher; and besides he is a witty fellow, so that he can fit in easily with Manius Curius, his bench-mate and gaming-mate.

\ciceroSection{14} I ask, then: if Lysiades, when summoned as a judge, does not answer, and pleads that he is a member of the Areopagus and ought not to be judging cases at Rome and Athens at the same time --- will the man who presides over the court of a little Greek judge, sometimes in pallium and sometimes in toga, accept that excuse? Or will he disregard the most ancient laws of the Athenians? What a bench, again, good gods! A judge from Crete --- and the most worthless of them. How is the accused to approach this man, how to come up to him? They are a hard people. ``But the Athenians are compassionate.'' I suppose Curius is not cruel either, who takes his chance with Fortune every day. There are likewise judges chosen who will perhaps be excused: they have a legal excuse, that they have changed their soil by reason of exile and have not afterwards been restored.

\ciceroSection{15} Would that madman have chosen these men as judges, filed their names with the treasury, entrusted to them a great part of the commonwealth, had he conceived of any appearance of a commonwealth? And I have spoken only of the better-known judges: those whom you know less I have not wished to name --- dancers, lyre-players, the whole chorus, in short, of Antony's revels, you must know, have been thrown into the third panel of judges. There is the reason why a law so excellent and so glorious was carried in the greatest downpour of rain, in storm, with winds, with squalls, with whirlwinds, amid lightning and thunder: that we might have for judges men whom no one would wish even as guests. The vastness of his crimes, the consciousness of his misdeeds, the plundering of that money whose account was made up in the temple of Ops --- these devised the third panel; and shameful judges were not sought out until honourable judges had given up hope of saving the guilty.

\ciceroSection{16} But what brazenness, what filthiness of mire it took, to dare to choose these men as judges! By their selection a double disgrace was branded upon the commonwealth: one, that the judges were so shameful; the other, that it was laid open and made known how many shameful men we had in the citizen body. These laws, then, and the rest of the same kind, even if they had been carried without violence and under valid auspices, I should still vote to abrogate: but now why should I vote to abrogate laws which I judge were never carried at all?

\ciceroSection{17} Are not those things to be marked with the gravest disgraces and memorials of this order, for the memory of posterity: that one man, Marcus Antonius, in this city since the founding of the city, openly kept armed men with him? Which neither our kings did, nor those who, after the kings had been driven out, wished to seize the kingship. I remember Cinna; I saw Sulla; only the other day, Caesar: for these three, since the civil community was freed by Lucius Brutus, were more powerful than the whole commonwealth. I cannot affirm that they were surrounded by no weapons; but this I do say: not by many, and not in the open.

\ciceroSection{18} But this plague was followed by a column of armed men. Crassicius, Mustela, Tiro --- flaunting their swords --- led troops like themselves through the Forum; barbarian archers held an assigned place in the column. And whenever the procession reached the temple of Concord, the steps would be filled, the litters set down --- not that he wished his shields to be hidden, but lest his household, if they carried the shields themselves, should grow weary. But this was foulest of all, not in sight only but even in hearing: that armed men, brigands, cut-throats were stationed in the very cella of Concord; that a prison was made of the temple; that, with the doors of Concord shut and brigands moving about among the benches of the Senate, the senators should pronounce their opinions.

\ciceroSection{19} If I did not come to that place on the Kalends of September, he said he would send carpenters and pull down my house. A great matter, I suppose, was at stake: he was bringing up the question of a public thanksgiving. I came the next day: he himself did not come. I spoke on the commonwealth, less freely indeed than is my custom, yet more freely than the threats of danger demanded. But that vehement and violent man, who would shut out this practice of speaking freely --- for Lucius Piso had done this same thing, with the greatest praise, thirty days before --- proclaimed his enmity against me; he ordered me to be present in the Senate on the thirteenth day before the Kalends of October. Meanwhile he himself, in Scipio's villa at Tibur, declaimed against me for seventeen days, working up a thirst; for that is usually his motive for declaiming.

\ciceroSection{20} When the day on which he had ordered me to be present came, then indeed in a hollow square he came to the temple of Concord and, against me in my absence, vomited up a speech out of his most impure mouth. On that day, had it been permitted to me by my friends to go into the Senate as I desired, he would have made the beginning of slaughter from me; for so he had resolved. And when once he had stained his sword with the crime, no consideration could have brought an end to the slaughter except weariness and satiety. For Lucius his brother was present, an Asiatic gladiator, who had fought as a murmillo at Mylasa: he was thirsting for our blood, having spilled much of his own in that gladiatorial fight. This man set a price upon your money; he marked down the holdings, both town and country; his beggary joined to greed was bearing down upon our fortunes; he divided up estates among whom and to whom he pleased; there was no recourse for the private citizen, no appeal to equity. Each possessor had only what Antony the partitioner had left him.

\ciceroSection{21} Although these acts, if you render the laws void, cannot stand, I still vote that they be marked separately under their own heading, and that it be judged that there were no septemvirs, that nothing should stand which is said to have been done by them. As for Marcus Antonius, who can judge him a citizen rather than a most foul and most cruel enemy --- who, sitting before the temple of Castor, said in the hearing of the Roman people that none should live save the victor? Do you suppose, senators, that he spoke more threateningly than he would have acted? And as for what he dared to say in public assembly --- that, when he had left office, he would be at the gates of the City with an army, and would enter the City whenever he wished --- what was that, but a proclamation of slavery to the Roman people?

\ciceroSection{22} And his journey to Brundisium --- what was the haste, what the hope, except that he might bring back a vast army to the City, or rather, into the City? And his selection of centurions: what unbridled rage of an ungoverned mind! When his bravest legions had shouted down his promises, he ordered to come to him at his lodging those centurions whom he had learned to be of right mind toward the commonwealth, and forced them to be cut down before his feet and his wife's --- the wife whom that grave commander had taken with him to the army. With what spirit, do you think, would this man have come against us, whom he hated, when he had been so cruel to those he had never seen? And how greedy for the moneys of the rich, when he had thirsted for the blood of the poor? Their own goods, such as they were, he at once distributed among his comrades and drinking-mates.

\ciceroSection{23} And while he was raging and was already advancing standards hostile to his country from Brundisium, Gaius Caesar, by the kindness of the immortal gods, by the divine greatness of his spirit, his genius, his counsel --- though by his own initiative and his exceptional valour, yet with the approval of my authority --- went to the colonies of his father, called together the veteran soldiers, in a few days made an army, and held back the headlong assault of the brigands. But afterwards, when the Martian Legion saw so excellent a leader, it did nothing else but make us at last be free; which the Fourth Legion imitated. On hearing this news, when he had called the Senate together and had brought in a consular to give an opinion judging Gaius Caesar an enemy, he suddenly collapsed.

\ciceroSection{24} Afterwards, with neither solemn sacrifices made nor vows pronounced, he did not set out --- he fled, in his military cloak. But where? Into the province of citizens most steadfast and most brave, who could not have borne him even had he come in such a way as to bring no war: ungoverned, irascible, full of insults, arrogant, always demanding, always plundering, always drunk. But that man, whose worthlessness no one could bear even when at peace, has carried war into the province of Gaul; he hems in Mutina, the most steadfast and most splendid colony of the Roman people; he assaults Decimus Brutus, commander, consul-elect, a citizen born not for himself but for us and for the commonwealth.

\ciceroSection{25} Is Hannibal then an enemy, and Antony a citizen? What did Hannibal do as an enemy that this man has not either done or is doing or is plotting and devising? What did the whole march of the Antonii consist in but devastations, lootings, slaughters, plunderings? Which Hannibal did not commit, because he was reserving much for his own use; whereas these men, who live for the hour, have thought neither of the fortunes and goods of the citizens nor even of their own advantage. To this man --- good gods! --- it pleases someone to send legates? These men know the form of the commonwealth, the laws of war, the precedents of our ancestors; they think what the majesty of the Roman people, what the severity of the Senate requires? You decree legates? If you decree them in order to plead, he will despise you; if in order to command, he will not listen. In short, however stern the instructions we give the legates, the very name of legates will quench the present fire of the Roman people, will break the spirit of the municipalities and of Italy. To omit these things, great as they are: certainly this legation will bring delay and slowness to the war.

\ciceroSection{26} Although some say --- as I hear some are about to say --- ``Let the legates set out: the war shall none the less be prepared,'' nevertheless the very name of legates will soften men's spirits and delay the speed of the war. The greatest turnings of the times, senators, come about by the slightest causes, both in every contingency of the commonwealth and especially in war, and most of all in civil war, which is generally governed by opinion and rumour. No one will ask with what instructions we have sent the legates: the very name of a legation sent unprompted will look like a sign of fear. Let Antony withdraw from Mutina, let him cease to assault Brutus, let him depart from Gaul. He is not to be entreated with words: he must be compelled by arms.

\ciceroSection{27} For we are not sending to a Hannibal that he withdraw from Saguntum, to whom the Senate once sent Publius Valerius Flaccus and Quintus Baebius Tampilus --- who, if Hannibal did not obey, were ordered to go to Carthage. Where do we order our men to go, if Antony does not obey? We are sending to one of our own citizens, that he should not assault a commander, not assault a colony of the Roman people. Is it so? Is this to be asked of him through legates? What is the difference, by the immortal gods, whether he assaults this City, or the bulwark of this City, a colony of the Roman people stationed there as a defence? The cause of the Second Punic War, which Hannibal waged against our ancestors, was the siege of Saguntum. Rightly were legates sent to him: they were sent to a Carthaginian, they were sent on behalf of the enemies of Hannibal, our allies. What likeness, then, is here? We are sending to a citizen, that he may not invest a commander of the Roman people, not an army, not a colony, not assault them, not lay waste the fields, not be an enemy.

\ciceroSection{28} Come now, suppose he obeys: can we, or do we wish to, treat him as a citizen? On the thirteenth day before the Kalends of January you crushed him with your decrees; you appointed that on the Kalends of January these matters should be brought before you which you see are now brought, concerning the honours and rewards of those who have well deserved and now deserve well of the commonwealth: among whom you judged that Gaius Caesar was first --- he who turned aside Marcus Antonius's nefarious assaults from the City into Gaul. Then the veteran soldiers, who were the first to follow Caesar; then those heavenly and divine legions, the Martian and the Fourth, you commended; to them, when they had not only abandoned but were even waging war upon their consul, you have pledged honours and rewards. And on the same day, when the edict of Decimus Brutus, that most outstanding citizen, was brought in and posted, you praised his act, and the war which he had undertaken on his own initiative you confirmed by public authority.

\ciceroSection{29} What else, then, did you do on that day, but judge Antony an enemy? After your decrees, either he will be able to look upon you with composure, or you upon him without the greatest pain? His own crime shut him out from the commonwealth, drew him apart, severed him --- and not his crime alone but, as it seems to me, a certain fortune of the commonwealth. If he should obey the legates and return to Rome, do you suppose that a standard for abandoned citizens to rally to will ever be wanting? But this I fear less: there are other things which I fear more, and on which I dwell. He will never obey the legates. I know the man's madness, his arrogance; I know the desperate counsels of his friends, to whom he is given over.

\ciceroSection{30} Lucius indeed, his brother, in his quality as a foreign gladiator, takes the lead of the household. Granted that the man himself were of sound mind --- which he never will be --- it will still not be allowed him to be so on their account. The time, meanwhile, will be worn away; the preparations for war will grow cold. Whence to this point has the war been drawn out, but from delay and procrastination? As soon as the Senate could be held freely, after the brigand's departure --- or rather, his desperate flight --- I always pressed that we be called together. On the day on which we were first called together, I was present myself, though the consuls-elect were absent; with your fullest agreement I laid down by my motion the foundations of the commonwealth --- later, to be sure, than was fitting (for I could not do it earlier), but yet, if from that time no day had been let pass, we should certainly have had no war at all.

\ciceroSection{31} Every evil, while it is being born, is easily crushed; once inveterate, it generally grows stronger. But then the Kalends of January were being waited for --- not rightly, perhaps. Yet let us leave past things behind: even this delay, until the legates set out, until they return? The waiting for them brings doubt about the war. And in a doubtful war, what zeal can there be for the levy? Therefore, senators, I move that no mention of legates be made; I think the matter must be carried through without any delay, and at once; I say that a state of tumult must be decreed, the courts suspended, the military cloaks taken up, a levy held, exemptions revoked, throughout the City and throughout all Italy except Gaul.

\ciceroSection{32} If these things are done, the mere opinion and rumour of our sternness will overwhelm the madness of the wicked gladiator. He will perceive that he has undertaken war against the commonwealth; he will experience the sinews and the strength of a unanimous Senate. For now he keeps saying that it is a contest of parties. What parties? One party has been beaten; the other is from the very heart of Gaius Caesar's party --- unless perhaps we suppose that Caesar's party is being assaulted by the consuls Pansa and Hirtius and by the son of Gaius Caesar. But this war is not stirred up out of the dissension of parties, but out of the criminal hope of the most abandoned citizens, by whom our goods and fortunes have been marked down and already distributed according to each one's choice.

\ciceroSection{33} I have read a letter of Antony's which he had sent to a certain septemvir, a capital fellow, his colleague. ``What you have set your heart on, you shall see to; what you have desired, you shall certainly have.'' Behold the man to whom we are to send legates, to whom we are to put off making war: a man who has not even committed our fortunes to the lottery, but has so handed us over to each man's caprice that he has left nothing untouched even for himself which has not already been promised to someone. With this man, senators --- with this man, I say --- it must be fought out, and at once; the slowness of legates is to be repudiated.

\ciceroSection{34} Therefore, that we may not have to make many decrees daily, I move that the whole commonwealth be entrusted to the consuls, and that they be permitted to defend the commonwealth and to see to it that the commonwealth take no harm. I move also that it be no fault to those who are in the army of Marcus Antonius, if they leave him before the Kalends of February. If you decree these things, senators, in a short time you will recover the liberty of the Roman people and your own authority. But if you act more leniently, you will yet decree the same things --- only, perhaps, more slowly. Concerning the commonwealth, on what you have referred to us, I seem to have decreed enough.

\ciceroSection{35} The other matter is concerning honours: about which I understand I must speak next. But the order which is wont to be kept in asking for opinions, the same I shall keep in honouring brave men. From Brutus, then, the consul-elect, let us take our opening, after the custom of our forefathers. To omit his earlier deeds --- which are very great, but have hitherto been praised more by men's private judgments than publicly --- with what words shall we be able to attain to the praises of this very moment? Such valour seeks no reward other than this glory of praise; though it lack even that, it will still be content with itself --- although, set as it were in the memory of grateful citizens as in the light, it rejoices. Praise, then, the testimony of our judgment and witness, must be paid to Brutus.

\ciceroSection{36} Therefore, senators, I move that a decree of the Senate be drawn up in these words: ``Whereas Decimus Brutus, commander, consul-elect, holds the province of Gaul in the power of the Senate and the Roman people, and whereas, in so short a time, with the highest zeal of the municipalities and colonies of the province of Gaul, which has well deserved and now deserves well of the commonwealth, he has enrolled and got together so great an army --- that he has done this rightly and in due order and in accordance with the interest of the commonwealth; and that this most outstanding service of Decimus Brutus toward the commonwealth is and shall be gratifying to the Senate and the Roman people: accordingly the Senate and the Roman people judge that, by the labour, counsel, valour, and incredible zeal of Decimus Brutus, commander, consul-elect, and by the consensus of the province of Gaul, the commonwealth has been succoured in its most difficult time.''

\ciceroSection{37} For so great a service of Brutus to the commonwealth, senators, and so great a benefit, what honour is great enough that is not owed to him? For if Gaul had lain open to Marcus Antonius, if, with the municipalities and colonies overwhelmed unprepared, he had been able to break into that further Gaul, what terror would have hung over the commonwealth? That most insane of men, headlong and devious in all his counsels, would no doubt have hesitated to make war on us not only with his own army but with all the savagery of the barbarian world, so that we could not have checked his fury even with the rampart of the Alps. This, then, is the gratitude that must be paid to Decimus Brutus, who --- before your authority had intervened --- by his own counsel and judgment, did not receive him as a consul but kept him off from Gaul as an enemy, and chose rather to be himself besieged than that this City should be. Let him therefore have, for this so great and so glorious deed, the everlasting witness of our decree; and let Gaul, which always has presided and presides over this empire and our common liberty, be deservedly and truly praised, that she did not surrender her own forces to Antony but set them against him.

\ciceroSection{38} And I also move that the most ample honours be decreed to Marcus Lepidus for his outstanding services to the commonwealth. He has always wished the Roman people free, and he gave the greatest sign of his will and his judgment on that day when, while Antony was placing the diadem on Caesar, he turned aside and by his groans and his sadness made clear how great a hatred he had of slavery, how greatly he desired the Roman people free, how he had borne what he had borne more from the necessity of the times than by his own judgment. With what restraint he then used himself in that time of the civic body which followed the death of Caesar --- who of us can forget? These are great things, but our speech hastens on to greater.

\ciceroSection{39} For what --- O immortal gods! --- could happen more admirable to all nations, what more to be wished for by the Roman people, than that, when the greatest civil war was on, the issue of which we all dreaded, it should be extinguished rather by wisdom and mercy than that the matter should be brought to a crisis by arms and the sword? And if Caesar had had the same way of acting in that foul and pitiable war, to leave aside the father, we should now have safe the two sons of Gnaeus Pompey, that supreme and singular man: to whom certainly filial duty ought not to have been a snare. Would that Marcus Lepidus could have saved them all! That he would have done so, he declared in what he was able to do, when he restored Sextus Pompey to the citizenship --- the greatest ornament of the commonwealth, the most illustrious monument of his clemency. Heavy was that fortune of the Roman people, heavy was that fate. For when Pompey the father, who was the light of the Roman people's empire, was extinguished, there was killed a son most like his father.

\ciceroSection{40} But all things seem to me to have been expiated, by the judgment of the immortal gods, with Sextus Pompey preserved for the commonwealth. For this just and great cause, and because Marcus Lepidus has turned the most dangerous civil war, and the greatest, to peace and concord by his humanity and wisdom, I move that a decree of the Senate be drawn up in these words: ``Whereas the commonwealth has been administered by Marcus Lepidus, commander, pontifex maximus, time and again both well and prosperously, and the Roman people has perceived that kingly domination is most displeasing to him; and whereas by his labour, valour, counsel, and singular clemency and gentleness a most bitter civil war has been extinguished,

\ciceroSection{41} and Sextus Pompey, son of Gnaeus, Magnus, has by the authority of this order laid down his arms, and by Marcus Lepidus, commander, pontifex maximus, with the highest goodwill of the Senate and the Roman people has been restored to the citizenship: the Senate and the Roman people, in return for the very many and very great services of Marcus Lepidus to the commonwealth, place great hope in his valour, his authority, his good fortune, of peace, quiet, concord, and liberty; and the Senate and the Roman people will be mindful of his services to the commonwealth; and it pleases them, by the resolution of this order, that a gilded equestrian statue be set up to him on the Rostra, or in whatever other place in the Forum he himself shall choose.'' This honour, senators, seems to me the greatest --- first, because it is just; for it is not bestowed only on the prospect of times to come, but is rendered for the fullest services already given. Nor can we point to anyone else as having had this honour bestowed by the Senate on the Senate's own free and unconstrained judgment.

\ciceroSection{42} I come to Gaius Caesar, senators: who, if he had not been, who of us could have remained? Antony was flying to the City from Brundisium, a man most ungovernable, ablaze with hatred, with hostile spirit against all honest men, with an army. What could be set against his audacity and his crime? We had as yet no commanders, no troops; there was no public counsel, no liberty; necks had to be given over to his nefarious cruelty; we were all seeking flight, which itself had no way out.

\ciceroSection{43} Who at that moment, what god, brought to us, brought to the Roman people, this divine young man? Who, when all things lay open to that pestilential citizen for our destruction, suddenly, beyond all hope, arose and got together an army to be set against the fury of Marcus Antonius, before anyone suspected that he was even thinking of it. Great honours were paid to Gnaeus Pompey, when he was a young man, and rightly; for he came to the aid of the commonwealth --- but he was much more robust in age, and more prepared by the eagerness of soldiers seeking a leader, and in a different kind of war. For Sulla's cause was not pleasing to all. The multitude of the proscribed declares it --- the very great calamities of so many municipalities.

\ciceroSection{44} But Caesar, many years younger, armed veterans now eager for rest; he embraced that cause which was most welcome to the Senate, most welcome to the people, most welcome to all Italy, most welcome to gods and men. And Pompey came to the great command and the victorious army of Lucius Sulla: Caesar attached himself to no one; he was himself the prince in making an army and getting together a guard. Pompey had the territory of Picenum hostile to the party of his adversaries; this man, out of Antony's own friends but better friends to liberty, made up an army against Antony. By Pompey's resources Sulla ruled; by this man's protection Antony's domination was crushed.

\ciceroSection{45} Let us therefore give command to Caesar, without which the business of the army cannot be conducted, the army held together, the war waged: let him be propraetor, with the fullest right that any man could enjoy. This honour, though it is great for his age, yet bears upon the necessity of carrying through what is to be done, not only upon his dignity. Therefore let us seek that which we shall scarcely attain to this day. But often, I hope, the opportunity of distinguishing this young man will fall to us and to the Roman people;

\ciceroSection{46} for the present time, however, I vote that the decree be passed thus: ``Whereas Gaius Caesar, son of Gaius, pontifex, propraetor, at a most critical moment for the commonwealth, has urged the veteran soldiers to defend the liberty of the Roman people and has enrolled them; and whereas the Martian and the Fourth Legions, with the greatest zeal and with the most excellent consent for the commonwealth, with Gaius Caesar as their leader and author, are defending and have defended the commonwealth and the liberty of the Roman people; and whereas Gaius Caesar, propraetor, has set out to the help of the province of Gaul with his army, has brought cavalry, archers, and elephants under his own power and that of the Roman people, and at the most difficult time for the commonwealth has come to the aid of the safety and dignity of the Roman people --- for these causes it pleases the Senate that Gaius Caesar, son of Gaius, pontifex, propraetor, be a senator, and give his opinion in the place of an ex-praetor; and that his standing, in whatever magistracy he shall seek, be so reckoned as it would be allowed to be reckoned under the laws, if in the preceding year he had been quaestor.'' For what reason is there,

\ciceroSection{47} senators, why we should not wish him to attain the most ample honours as soon as possible? For when by the leges annales they used to set a riper age for the consulship, they were guarding against the rashness of youth: Gaius Caesar at the outset of his life has taught us that we ought not to wait, when valour is excellent and exceptional, for the advance of age. And so our ancestors, of that truly ancient stock, had no leges annales; these were brought in many years later by ambition, that there might be set stages of canvassing between equals. Thus often a great natural endowment of valour, before it could profit the commonwealth, has been extinguished.

\ciceroSection{48} Yet truly among the ancients the Rulli, the Decii, the Corvini, and many others, and in more recent memory Africanus the elder, Titus Flamininus, made consuls while still very young, performed such deeds that they increased the empire of the Roman people and adorned its name. What of Alexander the Macedonian, who, when from his early youth he had begun to do the greatest deeds, did he not die in his thirty-third year? An age which is, by our laws, ten years younger than the consular. From which it can be judged that the race of valour is swifter than the race of age. For as for what those who envy Caesar feign to fear --- that he may not be able to hold himself in, may not moderate himself, may not, lifted up by our honours, use his resources too intemperately ---

\ciceroSection{49} the nature of things is such, senators, that whoever has caught the sense of true glory, and whoever has felt himself held a dear citizen and a saving one of the commonwealth by the Senate, by the Roman knights, and by the whole Roman people, judges that nothing is to be set beside this glory. Would that Gaius Caesar --- the father, I mean --- as a young man had had the fortune to be most dear to the Senate and to all the best men! When he had neglected to attain to this, he wasted all the force of his genius (which was the highest) in popular shallowness. And so, since he had no regard either for the Senate or for honest men, he opened up for himself that path to enlarging his power which the virtue of a free people could not bear. But his son's reasoning is the furthest possible thing from his own: he is dear to all, and most of all dearest to the best. In him the hope of liberty is set; from him safety has already been received; for him the highest honours are both sought out and ready.

\ciceroSection{50} Whose singular prudence, then, do we admire, while of his folly we are afraid? For what is more foolish than to prefer power useless, resources envied, headlong and slippery desire of domination, to true, weighty, solid glory? Has he seen this as a boy --- and shall he not see it when he has advanced in age? ``But he is hostile to certain most illustrious and excellent citizens.'' No fear of this need exist. Caesar has remitted all his enmities to the commonwealth; he has set up the commonwealth as the judge over himself, the moderator of all his counsels and acts. For he has come to the commonwealth in such a spirit that he should strengthen, not overthrow it. I have all the feelings of the young man known to me. Nothing is dearer to him than the commonwealth, nothing weightier than your authority, nothing more to be wished than the judgment of honest men, nothing sweeter than true glory.

\ciceroSection{51} Therefore from him you should not only fear nothing but expect greater and better things; nor in him, who has set out to free Decimus Brutus from siege, should you fear that the memory of a private grief may linger which could be more potent with him than the safety of the civic body. I dare even to bind my faith, senators, to you, to the Roman people, to the commonwealth --- which, indeed, when no force was constraining me, I should not dare to do; for I should be afraid of the dangerous opinion of rashness in so great a matter. I promise, I undertake, I pledge, senators, that Gaius Caesar will always be such a citizen as he is today, and as we ought most to wish and to pray that he be.

\ciceroSection{52} Since this is so, of Caesar at this time enough will have been said. Nor do I think that I should be silent about Lucius Egnatuleius, a most steadfast and most resolute citizen and most friendly to the commonwealth; but a testimony to his exceptional valour must be paid, in that he brought the Fourth Legion to Caesar, to be a guard for the consuls, the Senate, the Roman people, and the commonwealth. For this cause it pleases me that Lucius Egnatuleius be permitted to seek, to take, and to hold magistracies three years before the legal time. In which, senators, not so much an advantage is granted to Lucius Egnatuleius as an honour: for in such a matter it is enough to be named.

\ciceroSection{53} Concerning the army of Gaius Caesar, I move that the decree be passed thus: ``It pleases the Senate that the veteran soldiers who, having followed the authority of Caesar, pontifex, propraetor, have defended and are defending the liberty of the Roman people and the authority of this order, shall themselves and their children have exemption from military service; and that Gaius Pansa and Aulus Hirtius, the consuls, either or both, if it should seem good to them, take cognisance of what land in those colonies into which the veteran soldiers were settled is being held contrary to the Lex Iulia, so that it may be distributed to the veteran soldiers; and that they take cognisance separately of the Campanian land, and consider the matter of enlarging the rewards of the veteran soldiers; and that to the Martian Legion and the Fourth Legion, and to those soldiers who from the Second and the Thirty-fifth Legions have come over to Gaius Pansa and Aulus Hirtius the consuls and have given in their names, because to them the authority of the Senate and the liberty of the Roman people is and has been most dear, exemption from military service, for themselves and their children, be granted, save in the Gallic or Italic tumult; and that those legions, the war being concluded, be discharged; and that whatever sum of money Gaius Caesar, pontifex, propraetor, has promised to the soldiers of those legions individually, the same be paid; and that Gaius Pansa and Aulus Hirtius, the consuls, either or both, if it should seem good to them, take account of land which can be divided without injury to private parties, and so give and assign it to those soldiers, to the Martian Legion and to the Fourth Legion, as land has been most amply given and assigned to any soldiers.'' I have spoken, consuls, on all those matters concerning which you have referred to us; which, if they shall be decreed without delay and in good time, you will more easily make ready those things which the moment and necessity demand. There is need of speed: of which, had we made use, we should --- as I have often said --- have no war at all.
