% Philippica III (philippic-3) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Later, senators, than the situation of the commonwealth demanded, but at last we have been summoned together; which I, for my part, was calling for daily, since I saw that a wicked war against our altars and hearths, against our lives and fortunes, was being not prepared but already waged by an abandoned and lost man. The Kalends of January are awaited, but Antony does not await them: he is attempting to make an armed assault upon the province of Decimus Brutus, that high and singular man, intending --- equipped and ready from that province --- to march upon the City; from which he threatens that he will come.

\ciceroSection{2} What waiting is there, then, or what delay even of the smallest space of time? For although the Kalends of January are close at hand, yet a short time is long for the unprepared. The day --- or rather the hour --- often brings great disasters if no provision has been made; and a fixed day is not, like sacrifices, the kind of thing for which deliberations are usually deferred. If either the Kalends of January had fallen on the day Antony first fled from the City, or those Kalends had not been awaited, by now we should have no war at all. By the authority of the Senate and the consent of the Roman people we should easily have broken the audacity of an insane man. This I am confident that the consuls-elect will do, as soon as they enter office; for they are of the highest spirit, the highest counsel, and a singular concord. My own haste, however, is greedy not only for victory but also for speed.

\ciceroSection{3} For how long shall a war so great, so cruel, so wicked be warded off by private undertakings? Why does public authority not come to its aid as soon as possible? Gaius Caesar, a young man --- almost still a boy --- with an incredible and almost divine mind and courage, at the very moment when Antony's frenzy was burning hottest, and when his cruel and ruinous return from Brundisium was being feared, when none of us asked it, none of us thought of it, none of us even wished it --- because it seemed it could not be done --- raised the firmest of armies out of the unconquered stock of the veteran soldiers, and poured out his patrimony: though I have not used the word I should have used --- he did not pour it out: he invested it in the safety of the commonwealth.

\ciceroSection{4} Although such thanks as are owed to him cannot be returned, yet such must be returned as the greatest our hearts can hold. For who is so ignorant of affairs, so heedless of the commonwealth, that he does not see this: that if Marcus Antonius had been able to come from Brundisium to Rome, as he threatened, with the forces he expected to have, no kind of cruelty would have been passed over by him? --- he who in his host's house at Brundisium ordered the throats of the bravest and best of the citizens to be cut; before whose feet, as they died, his wife's mouth, it was well known, was sprinkled with their blood. Steeped in this cruelty, when he was coming on far more enraged against all good men than he had been against those whom he had slaughtered, on whom of us, or on what good man at all, would he have spared his hand?

\ciceroSection{5} From this plague Caesar, by a private undertaking --- for it could not have been done otherwise --- delivered the commonwealth. Had he not been born into this commonwealth, by Antony's crime we should have no commonwealth at all. For so I see, so I judge: if one young man had not checked the frantic onsets and the most cruel attempts of that man, the commonwealth would have perished utterly. To him, on this day of all days, senators --- for we now meet for the first time in such a way that, by his kindness, we are free to say what we feel --- to him authority must be granted, that he may defend the commonwealth which he has not only taken in hand of his own accord, but which we have also entrusted to him. Nor indeed about the Martian Legion ---

\ciceroSection{6} since after this long interval it is now allowed us to speak about the commonwealth, can silence be held. Who, one man, was ever braver, who more devoted to the commonwealth, than the whole Martian Legion? Having judged Antony to be an enemy of the Roman people, it refused to be the companion of his madness: it left the consul --- which it surely would not have done, had it judged him to be consul whom it saw to be intending nothing else, attempting nothing else, but the slaughter of citizens and the destruction of the State. And that legion has taken its stand at Alba. What city could it have chosen either more suitable for action, or more faithful, or of braver men, or of citizens more devoted to the commonwealth?

\ciceroSection{7} Imitating the valour of this legion, the Fourth Legion, under the lead of Lucius Egnatuleius the quaestor, an excellent and most valiant citizen, has followed the authority and the army of Gaius Caesar. We must therefore see to it, senators, that the things which that most illustrious young man and the most outstanding of all men has done of his own accord and is doing, be confirmed by our authority; and that the wonderful unanimity of the veterans, those most valiant men, and then of the Martian and the Fourth Legions, in the recovery of the commonwealth, be ratified by our praise and testimony; and that on this very day we pledge that their advantages, honours, and rewards shall be our concern as soon as the consuls-elect enter office.

\ciceroSection{8} And these things which I have said about Caesar and his army have long been known to us. For by the admirable valour of Caesar and the steadfastness of the veteran soldiers and of those legions which by their best judgment followed our authority, the freedom of the Roman people, and the valour of Caesar, Antony has been thrown back from our throats. But these, as I have said, are earlier matters: this new edict, however, of Decimus Brutus --- which has just been posted --- certainly cannot be passed over in silence. He promises that he will hold the province of Gaul in the power of the Senate and Roman people. O citizen born for the commonwealth, mindful of his name, an imitator of his ancestors! For the freedom that, with Tarquin expelled, was so longed for by our forefathers, is not more so than the freedom we now must preserve, with Antony driven out.

\ciceroSection{9} They, from the founding of the City, had learned to obey kings: we, with the kings expelled, had been overtaken by forgetfulness of slavery. And that Tarquin whom our ancestors could not endure was not held and called cruel or impious, but proud: a vice which we have often put up with in private men, our ancestors could not put up with even in a king. Lucius Brutus would not endure a proud king: shall Decimus Brutus suffer the criminal and impious Antony to be king? What did Tarquin do like the countless things which Antony does and has done? Even the kings had a senate: yet, as when Antony holds a senate, armed barbarians did not take part in the king's council. The kings kept the auspices; which this consul-and-augur has disregarded, not only by passing laws against the auspices, but with a colleague proposing them with him whom he himself had rendered defective by lying auspices.

\ciceroSection{10} What king, again, was ever so signally shameless as to have all the advantages, the favours, the rights of his kingdom for sale? What immunity, what citizenship, what reward has this man not sold --- to single men, to states, to whole provinces? Of Tarquin we have heard nothing low, nothing sordid: yet in this man's house gold was weighed out among the wool-baskets, money was counted out; in one house all the men whose interest it was struck their bargains for the whole empire of the Roman people. Of cruel punishments inflicted on Roman citizens we have no record from Tarquin: but this man both at Suessa cut the throats of those he had taken into custody, and at Brundisium slaughtered some three hundred bravest men and most excellent citizens.

\ciceroSection{11} Finally, Tarquin was waging war on behalf of the Roman people at the time when he was expelled: Antony was bringing an army against the Roman people at the time when, deserted by the legions, he took fright at the name of Caesar and at his army, and, neglecting the customary sacrifices, before dawn pronounced those vows which he will never pay, and at this moment is attempting to invade a province of the Roman people. A greater benefit, therefore, does the Roman people have, and look for, from Decimus Brutus than our ancestors received from Lucius Brutus, the first of this race and name which is, above all others, to be preserved.

\ciceroSection{12} Since all servitude is wretched, but it is intolerable to serve a foul, lewd, effeminate creature, never even in his fear sober --- the man, then, who shuts him out from Gaul, by a private undertaking moreover, judges --- and judges most truly --- that he is no consul. We must therefore see to it, senators, that we confirm by our public authority the private undertaking of Decimus Brutus. Nor indeed ought you to have thought Marcus Antonius a consul after the Lupercalia: for on that day, when in the sight of the Roman people, naked, anointed, drunk, he addressed the assembly, and set about putting a diadem on his colleague --- on that day he abdicated not only the consulship but liberty itself. For he would surely have had to become a slave on the spot, if Caesar had been willing to receive the badge of kingship from his hand. Shall I, then, count this man a consul, this man a Roman citizen, this man a free man, this man, finally, a man, who on that foul and scandalous day showed both what could be suffered while Gaius Caesar lived, and what he himself longed to gain on Caesar's death?

\ciceroSection{13} Nor indeed can I be silent about the valour, the steadfastness, the dignity of the province of Gaul. It is the flower of Italy, the firm support of the empire of the Roman people, the ornament of our dignity. So great is the unanimity of the towns and colonies of the province of Gaul, that all seem to have conspired to defend the authority of this order and the majesty of the Roman people. Therefore, tribunes of the plebs, although you have brought before us nothing but the question of a guard, so that the consuls may safely hold the Senate on the Kalends of January, yet you seem to me, with great wisdom and the best of intent, to have given us the power of speaking on the whole commonwealth. For when you judged that the Senate could not safely be held without a guard, you decreed at the same time that Antony's crime and audacity were inside the walls.

\ciceroSection{14} I shall therefore embrace all matters in my motion, with you, as I see, not unwilling: that authority be given by us to the most outstanding leaders, that hope of rewards be held out to the bravest soldiers, and that Antony be judged --- not by word, but in fact --- not only no consul but also a public enemy. For if he is consul, the legions that left the consul have deserved a flogging to death, Caesar is a criminal, Brutus is a traitor, who by private undertaking raised armies against the consul. If, on the other hand, new honours must be sought out for the soldiers in view of their divine and immortal merit, and yet to the leaders thanks cannot even be returned, who is there that does not reckon him an enemy whom those who pursue him with arms are judged to be the preservers of the commonwealth?

\ciceroSection{15} But how insulting in his edicts, how barbarous, how rude! First, against Caesar, what abuse he heaped up, drawn from the recollection of his own unchastities and debaucheries! For who is more chaste than this young man, who more modest, what brighter example of the old purity have we among our youth? Who, on the other hand, more foul than the man who reviles him? He throws low birth in the teeth of Gaius Caesar's son, whose natural father, had life lasted him, would have been made consul. ``A mother from Aricia.'' You would think he meant from Tralles, or Ephesus. See how we are looked down on --- all of us who come from the towns --- that is, all of us without exception: for how many of us do not? But what country town does he not scorn, who so despises Aricia --- a town most ancient in its antiquity, allied by treaty, almost a neighbour in distance, most honourable in the lustre of its townsmen?

\ciceroSection{16} From Aricia came the Voconian and the Atinian laws; from Aricia many curule chairs both in our fathers' memory and in our own; from Aricia the most polished and most numerous of the Roman knights. But if you do not approve an Arician wife, why do you approve a Tusculan one? Although the father of this most sanctified and best of women, Marcus Atius Balbus, a man honourable in the first rank, was a praetorian: of your own wife, a good woman, certainly at any rate a wealthy one, the father was a certain Bambalio, a man of no account whatever. Nothing was more contemptible than that man, who got his cognomen out of insult, on account of the hesitancy of his tongue and the stupor of his heart. ``But his grandfather was noble.'' Tuditanus, I take it, the one who used to stand on the Rostra in a robe and tragic buskins and scatter coins to the people. I could wish he had left his descendants this contempt for money! There you have a glorious nobility of blood.

\ciceroSection{17} And how does it come about that the son born of Julia seems ill-born to you, when you are accustomed to glory in the same descent on your mother's side? What madness, again, is it to say a word against the low birth of wives, when your own father took to wife Numitoria of Fregellae, a traitor's daughter, and you yourself have begotten children of a freedman's daughter? But this is for those most illustrious men, Lucius Philippus, who has a wife from Aricia, and Gaius Marcellus, who has a daughter of an Arician woman, to look into: I know for certain that they do not regret the standing of these most excellent women. The same man also addresses Quintus Cicero, my brother's son, in his edict, and the madman does not see that his addressing him is a recommendation. For what could have happened to that young man more to be desired than to be known to all the world as a sharer in Caesar's counsels, and an enemy to Antony's frenzy?

\ciceroSection{18} But even this gladiator dared write that he had plotted parricide against his father and his uncle. O the wonderful impudence, the audacity, the temerity! To dare write this against that young man whom both I and my brother, on account of his sweetest and best of characters and his outstanding ability, vie with each other in loving, and every hour hold in our eyes, our ears, our embrace? For with the same edicts he does not know whether to hurt me or to praise me. When he threatens those same best of citizens with the punishment which I exacted from the basest and most criminal, he seems to praise me --- as though he wished to imitate the deed; but when he scratches at the memory of that most splendid act, he supposes that he is rousing among his own sort some unpopularity against me.

\ciceroSection{19} But what did he himself do? When he had posted so many edicts, he gave notice that the Senate should attend in full numbers on the twenty-fourth of November: on that very day, he himself did not attend. And how did he give notice? These, I think, are the words at the end: ``If anyone does not attend, all will be able to think that he was the author both of my destruction and of utterly ruinous designs.'' What are these ruinous designs? Are they those which look to the recovery of the freedom of the Roman people? Of which designs I confess that I both am, and have been, an adviser to Caesar and an encourager. Although he needed no man's counsel, yet, as the saying goes, I spurred the runner on. As for your destruction, what good man would not be its author, when in it consist the safety and life of every best citizen, the freedom of the Roman people, and the public dignity?

\ciceroSection{20} But when he had roused us with so fierce an edict, why did he himself not attend? Do you suppose it was anything sad or grave? Held back by wine and feasting --- if such things are to be called feasts and not cook-shops --- he disregarded the day of his edict: he put it off to the twenty-eighth of November. He ordered us to attend on the Capitol; into which temple he himself climbed I know not by what passage of the Gauls. Men met when summoned, and even some considerable men, but forgetful of their own dignity. For such was the day, such the report, such the man who had called the Senate, that it was shameful to a senator to fear nothing. Yet even to those very men who had come together, he did not dare say a word about Caesar, although he had decided to bring his case before the Senate: a certain consular had brought a written motion.

\ciceroSection{21} What else is not daring to lay a charge against the man who was leading an army against him as consul, than to declare himself the public enemy? For necessarily one or the other had to be the enemy; nor could the contending leaders be judged otherwise. If, then, Caesar is the enemy, why does the consul lay no charge before the Senate? But if Caesar was not to be condemned by the Senate, what can he say but that, by his silence about Caesar, he has confessed himself to be the enemy? Whom in his edicts he calls a Spartacus, in the Senate he does not dare even call dishonest. But amid the most grievous matters, what fits of laughter he excites! I have committed to memory the little sentences of a certain edict, which he seems to think most witty: but a man who could understand what he meant, I have yet to find.

\ciceroSection{22} ``No insult is one which is offered by a worthy man.'' First, what is meant by ``worthy''? For many are worthy even of evil --- as he himself is. Or does he mean ``offered by one who has dignity''? But what dignity could be greater than mine? Then again, what is meant by ``to offer an insult''? Who talks like this? Then: ``Nor is the fear which an enemy proclaims.'' What, then? Does fear get proclaimed by a friend? Other things like these in succession. Is it not better to be dumb than to say what no one can understand? See why his teacher has been turned from an orator into a ploughman, possessing on the public land of the Leontine plain two thousand acres tax-free --- so that he may make an even greater fool of the stupid man on a public salary.

\ciceroSection{23} But these things, perhaps, are lighter matters: what I ask is, why was he so mild in the Senate, when he had been so savage in his edicts? For what was the use of threatening with death Lucius Cassius, a tribune of the plebs, a most valiant and steadfast citizen, if he came into the Senate? of driving Decimus Carfulenus, a man of right feeling for the commonwealth, out of the Senate by violence and threats of death? of forbidding Tiberius Cannutius --- by whom he had been castigated, with full right, in many most honourable public addresses --- not only the temple but even the approach of the Capitol? What decree of the Senate was he afraid they would veto? About the public thanksgiving, I imagine, for Marcus Lepidus, that most illustrious man. And that was indeed the danger --- that, for one of whose extraordinary honour we were every day taking thought, his ordinary honour should be obstructed.

\ciceroSection{24} And, lest he should seem to have given notice without cause that the Senate should be in attendance --- since he had been about to bring a matter of public concern before them --- when news was brought him about the Fourth Legion, he collapsed in mind, and, hurrying to escape, passed by mere division the decree of the Senate about the public thanksgiving --- a thing which had never been done before. But what a setting out afterwards, what a journey in the war-cloak, what avoiding of men's eyes, of the light, of the City, of the Forum, what a wretched flight, how foul, how disgraceful! Splendid, however, were the decrees of the Senate that evening of the same day: the religious assignment of provinces, the divine timing by which what was suited to each man came to each.

\ciceroSection{25} You do excellently then, tribunes of the plebs, in bringing before us the question of a guard for the consuls and the Senate, and deservedly we are all bound both to feel and to give you the greatest thanks. For how can we be free of danger amid so great a greed and audacity of men? But that broken and lost man --- what verdicts more grievous than those of his own friends does he await about himself? His closest intimate, a man bound to me, Lucius Lentulus, and Publius Naso, devoid of every greed, gave their judgment that they had no province at all, and that there had been no proper allotment by Antony. The same was done by Lucius Philippus, a man most worthy of his father, his grandfather, and his ancestors; in the same opinion was Gaius Turranius, a man of the highest integrity and innocence; the same was done by Spurius Oppius; even those who, out of regard for friendship with Marcus Antonius, conceded more to him perhaps than they wished --- Marcus Piso, my own relative, an excellent man and citizen, and Marcus Vehilius, of equal innocence --- said that they would defer to the authority of the Senate.

\ciceroSection{26} What shall I say of Lucius Cinna? whose tested integrity in many great matters makes the glory of this most honourable act less to be wondered at --- a man who utterly disregarded his province, just as with equal greatness and steadiness of spirit Gaius Cestius rejected his. Who, then, are left whom the divine lot delights? Titus Annius, Marcus Gallius. O fortunate each of them! for they wanted nothing more. Gaius Antonius, Macedonia. This man, too, is fortunate! for this was the province he had always on his lips. Gaius Calvisius, Africa. Nothing more fortunate! for he had only just left Africa, and as if foreseeing, he had left two of his legates at Utica behind him for his return. Then Marcus Cusinius, Sicily; Quintus Cassius, Spain. I have nothing to suspect: it is, I suppose, the lots for those two provinces that were the less divine.

\ciceroSection{27} O Gaius Caesar --- I mean the younger --- what a deliverance you have brought to the commonwealth, how unforeseen, how sudden! For who in fleeing did what he did, what would he have done as he came on? In an address he had said that he would be the guardian of the City, and that he would have an army at the walls of the City till the Kalends of May. O illustrious guardian of the sheep, as men say, a wolf! Was Antony to be the guardian of the City, or its plunderer and harasser? And he indeed said that he would come into the City and go out again when he pleased. What of this? Did he not, with the people hearing, sitting before the temple of Castor, say that, except him who had been victorious, none should live?

\ciceroSection{28} On this day, senators, for the first time after a long interval, we plant our foot upon the possession of liberty; of which, so far as I could, I have been not only the defender but also the preserver. But when I could no longer do that, I was quiet, and bore that disaster of the times, and that grief, neither abjectly nor without some dignity. But who can endure this most foul beast --- and how? What is there in Antony but lust, cruelty, insolence, audacity? Out of these vices, the whole of him is glued together. There appears in him nothing free-born, nothing moderate, nothing decent, nothing chaste.

\ciceroSection{29} Wherefore, since the matter has been brought to this point of crisis --- whether he shall pay the penalty to the commonwealth, or we shall be slaves --- let us at last, by the immortal gods, senators, take up the spirit and valour of our fathers, that we may either recover the freedom proper to the Roman race and name, or set death before slavery. Many things which in a free state ought not to have been borne we have borne and endured, some perhaps in hope of recovering liberty, others out of an excessive desire of living: but if we have borne those things which necessity compelled us to bear, which a certain almost fated force imposed --- which yet, in itself, we did not bear --- shall we also bear the most foul and most cruel domination of this impure brigand?

\ciceroSection{30} What will this man do, if he can, when enraged, who, when he could be angry with no one, was the enemy of all good men? What will he, victorious, not dare, who, having obtained no victory, did such great crimes after Caesar's death --- emptied his crowded house, plundered his gardens, transferred to himself every ornament from them, sought a pretext for slaughter and arson from his funeral, by two or three decrees of the Senate passed well and for the State's good restored the rest of public business to his own profit and plunder; sold immunities, freed states, lifted whole provinces out of the right of the empire of the Roman people, recalled exiles, had false laws and false decrees inscribed under Gaius Caesar's name and set up on the Capitol in bronze, and instituted a domestic market for all these things; imposed laws on the Roman people, with arms and garrisons shut out the people and the magistrates from the Forum, packed the Senate with armed men, shut armed men into the cell of Concord while he held the Senate, ran down to the legions at Brundisium, cut the throats of the best-feeling centurions among them, and tried to come with an army to Rome for our destruction and for the partitioning of the City?

\ciceroSection{31} And he, dragged off from this onslaught by the counsel and forces of Caesar, by the consent of the veterans, by the valour of the legions, not even broken by fortune, has not diminished his audacity, has not ceased either to rush on madly or to rave. Into Gaul he is leading a mutilated army; with one legion --- and that one wavering --- he is waiting for his brother Lucius, than whom no one more like himself can he find. But that man, from murmillo a leader, from gladiator a commander, what havocs has he wrought wherever he has set foot! He empties the wine-cellars, slaughters the herds of cattle and whatever else of livestock he comes upon; the soldiers feast; he himself, that he may imitate his brother, drowns himself in wine; the fields are laid waste, the villas plundered, mothers of families, virgins, freeborn boys snatched away and handed over to the soldiers. These same things, wherever he has led an army, Marcus Antonius has done.

\ciceroSection{32} To these most foul brothers, then, will you open the gates? Will you ever receive them into the City? Will you not, with the moment offered, the leaders ready, the spirits of the soldiers stirred, the Roman people in concord, all Italy roused to recover liberty, profit by the kindness of the immortal gods? With this moment lost, there will be no other. He will be hemmed in at his back, his front, his flanks, if he comes into Gaul. Nor must he be pressed only with arms but with our decrees. Great is the force, great the divine power, of a Senate of one and the same mind. Do you not see the Forum packed, and the Roman people roused to the hope of recovering its liberty? --- which, after this long interval, when it sees us here in such numbers, also hopes that we have met as free men.

\ciceroSection{33} Awaiting this day, I have avoided Marcus Antonius's criminal arms, when, attacking me in my absence, he did not see for what time I was reserving myself and my strength. For if then, when he was demanding from me the start of the slaughter, I had chosen to answer, now I should not be able to take counsel for the commonwealth. But having gained this opportunity, no time at all shall I let pass, senators, neither of day nor of night, in which I shall not think out whatever ought to be thought concerning the freedom of the Roman people and your dignity, and as for what ought to be done and carried out, I shall not only not refuse it but seek it out and demand it. This I did so long as I was allowed; I ceased so long as I was not allowed. Now it is not only allowed but necessary, unless we prefer to be slaves than, sooner than be slaves, to decide the matter in spirit and in arms.

\ciceroSection{34} The immortal gods have given us these defences: to the City, Caesar; to Gaul, Brutus. For if the one had been able to crush the City, immediately --- if the other had been able to seize Gaul, a little later --- every best man must have perished, the rest must have been slaves. This opportunity, therefore, that has been offered, hold fast, by the immortal gods, senators, and remember that you are the chief counsellors of the greatest council in the wide world. Give the Roman people the signal that your counsel will not fail the commonwealth, since he declares that his valour will not fail. There is no need for me to admonish you. There is no one so foolish as not to see that, if we sleep through this moment, not only a cruel and proud, but even an infamous and disgraceful, dominion must be borne by us.

\ciceroSection{35} You know Antony's insolence, you know his friends, you know his whole household. To serve men lustful, petulant, foul, lewd, gamblers, drunkards, is the height of misery joined with the height of dishonour. If now --- which may the gods avert! --- the final fate of the commonwealth has come, let us do what noble gladiators do, that they may fall with honour: let us, the chiefs of the world and of all nations, fall with dignity rather than serve in disgrace.

\ciceroSection{36} Nothing is more to be loathed than dishonour, nothing fouler than servitude. For glory and for liberty we were born: let us either keep these, or die with dignity. Too long have we hidden what we felt; now it is open at last; on both sides all men make plain what they feel, what they want. There are impious citizens --- for the love we bear the commonwealth, too many; but compared with the multitude of right-feeling citizens, only very few --- whom, for our crushing of them, the immortal gods have given the commonwealth an unbelievable strength and good fortune. For to the defences we already have will soon be added the consuls, of the highest prudence, valour, and concord, men who for many months have been considering and meditating about the freedom of the Roman people. With these as our authors and leaders, with the gods aiding, with us watchful and looking far ahead, with the Roman people in concord, we shall surely be free in a short space. The recollection of slavery will make liberty sweeter.

\ciceroSection{37} For these reasons --- since the tribunes of the plebs have brought before us the question that the Senate may be held safely on the Kalends of January, and that opinions on the highest matters of the commonwealth may be expressed freely --- on that matter I propose as follows: that Gaius Pansa and Aulus Hirtius, the consuls-elect, see to it that the Senate may safely be held on the Kalends of January. That, whereas an edict of Decimus Brutus, imperator and consul-elect, has been posted, the Senate judges Decimus Brutus, imperator and consul-elect, to deserve the best of the commonwealth, since he is defending the authority of the Senate and the liberty and dominion of the Roman people;

\ciceroSection{38} and that, whereas he is holding the province of Hither Gaul, of the best, bravest, and most loyal citizens of the commonwealth, and his army in the power of the Senate, he and his army, the towns and colonies of the province of Gaul, have acted and are acting rightly and in order and in accordance with the commonwealth. That the Senate judges it of the utmost concern to the commonwealth that the provinces be held by Decimus Brutus and Lucius Plancus, imperators and consuls-elect, and likewise by the others who hold provinces, under the Julian Law, until by decree of the Senate a successor has been appointed to each, and that these men take pains that those provinces and those armies be in the power of the Senate and Roman people and a defence to the commonwealth. And that, since by the work, valour, and counsel of Gaius Caesar, and by the highest agreement of the veteran soldiers, who, following his authority, have been and are a defence to the commonwealth, the Roman people has been and at this time is being defended against the most grievous perils;

\ciceroSection{39} and since the Martian Legion has taken its stand at Alba, in a most loyal and most valiant town, and has betaken itself to the authority of the Senate and the freedom of the Roman people; and since the Fourth Legion, using equal counsel and the same valour under the lead of Lucius Egnatuleius, an excellent quaestor and an outstanding citizen, is defending and has defended the authority of the Senate and the freedom of the Roman people --- it shall be a great care to the Senate, and shall be in the future, that for such great deserts of theirs toward the commonwealth, honours be granted to them and thanks be returned. That it is the pleasure of the Senate that Gaius Pansa and Aulus Hirtius, consuls-elect, when they have entered office, if it seem good to them, refer these matters at the earliest opportunity to this order, in such fashion as may seem to them in accordance with the commonwealth and their own good faith.

