% Philippica II (philippic-2) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} By what fate of mine, senators, am I to say it has come about that, these twenty years past, no man has been an enemy of the commonwealth without at the same time declaring war on me as well? Nor indeed is there any need for me to name them: you yourselves remember alongside me. They paid me penalties heavier than I should have wished for: but you, Antony, I marvel that you do not shudder at the ends of those whose deeds you imitate. And in others I marvelled at this less. For not one of them was an enemy to me of his own choice: every one of them had been provoked by me for the commonwealth's sake. You, not so much as injured by a word, in order to seem more reckless than Catiline and more frenzied than Clodius, have unprovoked assailed me with insults, and have thought that your estrangement from me would be your recommendation to impious citizens.

\ciceroSection{2} What am I to suppose? That he holds me in contempt? I do not see in my life, nor in my standing, nor in my conduct, nor in this modest measure of talent of mine, what Antony could find to look down upon. Or did he believe that I could most easily be torn down in the Senate? An order which has given to many of its most distinguished citizens the testimony of having well managed the commonwealth, but to me alone the testimony of having saved it? Or did he wish to contend with me in the trial of speaking? This indeed is a kindness. For what fuller, what richer subject than to speak both for me and against Antony? It must be this, surely: he did not think he could be approved by his own sort as an enemy of his country, unless he were also an enemy of mine.

\ciceroSection{3} Before I reply to his other points, on the friendship which he has charged me with breaking --- a charge I judge to be of the gravest kind --- I shall say a few words. He has complained that I once came into court against his interest. Was I not to come against a stranger on behalf of a familiar friend and connection? Was I not to come against influence gathered not on hope of virtue but by the bloom of youth? Was I not to come against an injury that this man secured by the favour of a most unjust officer, not by the right of a praetor? But this, I take it, you brought up to commend yourself to the lowest order, since all men remembered that you were the son-in-law of a freedman, and your children the grandsons of Quintus Fadius, a freedman. But you had handed yourself over, you said, to my school --- for so you put it --- you had been a frequent visitor at my house. By Hercules, if you had done so, you would have looked better to your reputation, better to your decency. But you did not do it; nor, even had you wished to, could you have done it on Gaius Curio's terms.

\ciceroSection{4} You said you had ceded the canvass for the augurate to me. O incredible audacity! O effrontery beyond words! At the time when, with the whole college asking for me, Gnaeus Pompeius and Quintus Hortensius nominated me as augur --- for one was not allowed to be nominated by more than two --- you were neither solvent, nor did you suppose you could be safe in any way except by the overthrow of the commonwealth. Could you in fact have stood for the augurate at that time, when Curio was not in Italy? Or could you, at the time when you were made augur, have carried a single tribe without Curio? Whose intimates were even condemned for violence, because they had been too zealous in your behalf.

\ciceroSection{5} ``But I have enjoyed a kindness from you.'' Of what kind? Although that very thing you cite, I have always carried before me openly: I preferred to confess myself in your debt rather than to seem ungrateful in the eyes of any less prudent man. But what kindness? That you did not kill me at Brundisium? Could you have killed a man whom the very victor himself --- who, as you yourself used to boast, had conferred on you the chieftainship from among his band of brigands --- had wished to be safe, had ordered to come back into Italy? Grant that you could have. What is any other kindness of brigands, senators, except that they can tell over the men to whom they granted life by not taking it? And if that were a kindness, the men who killed the very one by whom they had been spared --- men whom you yourself are accustomed to call most distinguished --- would never have won such glory. What sort of kindness, again, is it that you have refrained from a wicked crime? In which it ought to have been less pleasing to me that I was not killed by you than wretched that you could do that thing with impunity.

\ciceroSection{6} But suppose it a kindness, since no greater could be had from a brigand: in what respect can you call me ungrateful? Or ought I not to have complained of the ruin of the commonwealth, for fear of seeming ungrateful to you? In that complaint --- wretched indeed and grievous, but for me, by reason of the rank in which the Senate and the Roman people have placed me, a necessary one --- what was said by me in insult, what not in moderation, what not in friendship? What temperance, after all, was required, complaining of Marcus Antonius and yet refraining from abuse, when you had squandered what was left of the commonwealth, when in your house everything was for sale in the foulest market, when you confessed that laws which had never been promulgated were carried both about you and by you, when you as augur had suppressed the auspices and as consul the tribunician veto, when you were guarded in the most disgraceful fashion by armed men, when day by day in your disreputable household you committed every kind of impurity, being worn out with wine and with debauchery?

\ciceroSection{7} Yet I, as if my contest were with Marcus Crassus, with whom I have had many a great quarrel, and not with a single worthless gladiator, while complaining gravely about the commonwealth, said nothing about the man. And so today I shall make him understand how great a kindness he received from me on that occasion. Furthermore, he even read out a letter which he said I had sent to him --- a man devoid of any breeding, and ignorant of common civility. For who that has the slightest knowledge of decent men's usages ever brought into public and read aloud, on the occasion of some offence, a letter sent to him by a friend? What is this but to take from life the partnership of life, to take away the conversation of friends in absence? How many jokes commonly appear in letters which, brought out, would seem inept; how many serious things which still must in no way be made public!

\ciceroSection{8} Let this stand to his rudeness: now look at his incredible stupidity. What have you, eloquent man, to set against me --- you who nevertheless appear like a weasel to Sextus Seius and to Tiro Numisius? Since these very men stand at this very moment with their swords in sight of the Senate, I too will think you eloquent, if you show how you would defend them as common cutthroats. But what at last can you allege, if I deny that I ever sent you that letter? By what witness can you convict me? By the handwriting? In which you have a profitable proficiency. How can you, when the letter is in a secretary's hand? I am beginning to envy your tutor, who is paid the enormous fee I shall presently disclose to teach you to know nothing.

\ciceroSection{9} For what is less the part, I do not say of an orator, but of a human being, than to throw something at an adversary which, if he deny it in a word, the man who threw it cannot push any further? But I do not deny it, and on that very point I convict you not only of rudeness but of insanity. For what word is there in that letter which is not full of courtesy, of obligation, of goodwill? The whole of your charge is that I do not think ill of you in that letter, that I write to you as to a citizen, as to a good man, not as to a criminal and a brigand. Yet I, for my part, even if I had the right after the provocation, shall not produce your letters --- letters in which you ask that you may be allowed, through my agency, to recall a certain man from exile, and swear that you will not do it against my will: and you obtain that from me. For why should I have set myself in the way of your audacity, which neither the authority of this order nor the estimation of the Roman people nor any laws could restrain?

\ciceroSection{10} But still, what was there for you to be asking of me, if the man on whose behalf you asked had already been brought back by a law of Caesar's? But of course he wanted it to be my favour --- in a matter in which there could be no favour even of his own, once a law had been passed. But since I have, senators, much to say both on my own behalf and against Marcus Antonius, of you I ask the one thing: that you hear me kindly while I speak for myself; the other I shall myself bring about --- that when I speak against him, you shall listen attentively. At the same time I make this one entreaty: that if you know my moderation and self-restraint both in life in general and in speech, you should not think, when I have today answered him as he has provoked, that I have forgotten myself. I shall not handle him as a consul --- not that he treats me as a consular. And yet he is in no way a consul, whether because he lives as he does, or because he conducts the commonwealth as he does, or because he was made consul as he was; I, beyond any dispute, am a consular.

\ciceroSection{11} So that you might understand, then, what sort of consul he was professing himself to be, he flung my consulship in my teeth. A consulship, senators, which in name was mine, in fact yours. For what did I establish, what did I do, what did I undertake, except by the counsel, the authority, the judgment of this order? And these things you, a wise man --- not only an eloquent one --- dared to vituperate in the presence of those by whose counsel and wisdom they had been done? Yet who was found to vituperate my consulship besides yourself and Publius Clodius? Whose fate, indeed, awaits you, as it awaited Gaius Curio, since you have in your house what was fatal to both of them.

\ciceroSection{12} My consulship does not please Marcus Antonius. But it pleased Publius Servilius --- to name first among the consulars of that age the one most recently dead; it pleased Quintus Catulus, whose authority shall live forever in this commonwealth; it pleased the two Luculli, Marcus Crassus, Quintus Hortensius, Gaius Curio, Gaius Piso, Manius Glabrio, Manius Lepidus, Lucius Volcatius, Gaius Figulus, Decimus Silanus, Lucius Murena, who were then consuls-elect; the same that pleased the consulars also pleased Marcus Cato, who in many things looked ahead as he was leaving life, but especially in this --- that he did not see you consul. The most decisive judgment of all upon my consulship was Gnaeus Pompeius's, who, as soon as he set eyes on me when leaving Syria, embraced me and gave me joy, saying that it was by my kindness he was to see his country again. But why do I list them one by one? In a most full session of the Senate the verdict was such that there was no man who did not give me thanks as to a parent, no one who did not refer his life, his fortunes, his children, and the commonwealth itself, as received back from me.

\ciceroSection{13} But since the commonwealth has been bereft of those many such men whom I have named, let us come to the living --- the two who, of the number of consulars, are still left. Lucius Cotta, a man of the highest talent and the highest prudence, decreed a public thanksgiving in the amplest terms for those very acts of mine which you reproach; and to him those very consulars I just named, and the whole Senate, gave their assent. That is an honour which, since the founding of this city, no man in the civil toga before me has received.

\ciceroSection{14} Lucius Caesar, your uncle, with what a speech, with what firmness, with what gravity did he deliver his judgment against his own sister's husband --- your stepfather! Him you ought to have had as the author and teacher of all your counsels and of your whole life; you preferred to be like your stepfather rather than your uncle. I, a stranger, made use of his counsel as consul: you, his sister's son, have you ever referred a single matter of state to him? To whom does he refer? Immortal gods! To those, of course, whose very birthdays we have to hear announced. Antony has not come down to the Senate today.

\ciceroSection{15} Why not? He is throwing a birthday-feast in his pleasure-gardens. For whom? I shall name nobody: imagine some Phormio, some Gnatho, or even some Ballio. O the foul and disgraceful character of the man, his unbearable shamelessness, worthlessness, lust! When you have so close at hand a leading senator, a citizen above all his fellows --- you refer no business of the commonwealth to him, but you refer it to men who have nothing of their own and are draining yours. Your consulship, no doubt, is for the public health; mine was a public ruin. Have you lost shame so utterly along with chastity that you dared to say this in the very temple in which I consulted that Senate which once flourishing presided over the lands of the earth, but in which you have set down the most abandoned of men with their swords?

\ciceroSection{16} Yet you have even dared --- and what is there you would not dare? --- to say that the Capitoline slope, while I was consul, was full of armed slaves. To force those infamous senatorial decrees through, I was bringing violence to bear, I suppose, upon the Senate. O wretch, whether you do not know these things --- for you know nothing good --- or whether you do know them, that you can speak with such effrontery in the presence of such men! For what Roman knight, what young man of noble blood besides yourself, what man of any order who remembered that he was a citizen, was not on the Capitoline slope, when the Senate was in this temple? Who did not give in his name? Although neither the secretaries were enough nor the tablets sufficient to hold their names.

\ciceroSection{17} For when wicked men, accused of parricide against their country, were confessing, driven by the testimony of accomplices, by their own handwriting, by the voice almost of the letters themselves, that they had agreed to set fire to the city, to butcher the citizens, to lay Italy waste, to destroy the commonwealth --- who was there who would not be roused to the defence of the common safety, especially when the Senate and the Roman people had a leader of such a kind that, if there were any such now, the same thing would have befallen you as befell them? He says that the body of his stepfather was not given by me for burial. Yet not even Publius Clodius ever said this --- and since I was rightly his enemy, I grieve that you have now surpassed him in every vice.

\ciceroSection{18} What again came into your mind, to recall to our memory that you had been brought up at the house of Publius Lentulus? Or were you afraid we should not think that nature alone could have brought you out so utterly worthless, had there not been the addition of training as well? But you were so far out of your wits that throughout the whole of your speech you fought against yourself --- saying not only things that did not hang together, but things sharply disjoined and contrary, so that there was less of a contest between me and you than between you and yourself. You confessed that your stepfather had been involved in so great a crime, and yet complained that he was punished for it. So you praised what is properly mine, and reproached what belongs wholly to the Senate. For the arrest of the guilty was my act, the punishment was the Senate's. The eloquent man does not understand that the very person he is speaking against is praised by himself, while those before whom he speaks are reviled.

\ciceroSection{19} And now what audacity --- I do not call it that, for he covets to be called bold --- but what stupidity, in which he conquers all men, though it is the very thing he least wishes to be called, that he should make mention of the Capitoline slope, when armed men move about among our benches, when in this very temple of Concord --- immortal gods! --- in which, when I was consul, those saving opinions were delivered by which we have lived down to this day, men are posted with swords drawn. Accuse the Senate; accuse the equestrian order, which was then joined with the Senate; accuse all the orders, all the citizens --- provided only that you confess that this very order is at this moment besieged by Ituraeans. You say these things so impudently not because of audacity, but because you do not see how violently your statements collide with one another. Surely you have no sense at all. For what is more foolish than, when you yourself have taken up arms ruinous to the commonwealth, to reproach another with arms that brought it safety?

\ciceroSection{20} And in one passage you even wished to be witty. Good gods, how that did not become you! In which you are not without some blame yourself. For you might have got some salt from your mime-actress wife. ``Let arms yield to the toga.'' What? Did they not then yield? But later the toga yielded to your arms. Let us therefore enquire which was the better thing --- that the arms of criminals should yield to the liberty of the Roman people, or that our liberty should yield to your arms. Nor indeed shall I answer you any further about my verses: I shall say only this briefly: that you know neither those nor any letters at all; and that I never failed either the commonwealth or my friends, and yet, through all kinds of monuments of mine, I have achieved as much by my spare-time labours as to bring some usefulness to the youth and some praise to the Roman name through my night-watches and my writings. But these are not topics for the present hour: let us look at greater things.

\ciceroSection{21} You said that Publius Clodius was killed by my counsel. What would men have thought, if he had been killed at the time when you yourself pursued him with a sword in the forum, in sight of the Roman people, and would have finished the business, had he not flung himself onto the staircase of a bookshop and, with the entrance blocked, checked your onset? I confess that I favoured you in that --- though even you do not say I urged it on. As for Milo, I could not even favour him; for he had finished the business before anyone suspected him about to do it. ``But I urged it on.'' Was Milo's spirit then such that he could not benefit the commonwealth without a prompter? ``But I rejoiced.'' What then? In the universal joy of the city, was I alone to be sad?

\ciceroSection{22} Although on the death of Clodius an inquiry was set up, hardly established with sufficient prudence --- for what need was there of an inquiry by a new law into a man who had killed another, when there was already an inquiry established by law? --- still an inquiry was made. So what no man said against me when the case was being tried, you have now been found, all these years afterwards, to say.

\ciceroSection{23} As to what you have dared to say --- and that at length --- that by my agency Pompey was separated from Caesar's friendship, and that on that account, through my fault, the civil war was born: in that, you have not erred in the substance, but, what is the chief thing, in the dates. While that excellent citizen Marcus Bibulus was consul, I omitted nothing, so far as I could do and strive, to call Pompey away from his union with Caesar. In that, Caesar was the more fortunate. For he himself separated Pompey from intimacy with me. But after Pompey had given himself wholly over to Caesar, why should I have tried to drag him away from him? It would have been folly to hope, effrontery to urge it.

\ciceroSection{24} Yet there were two occasions on which I urged Pompey to something against Caesar. Reproach me for these, if you can: the one, that he should not extend Caesar's command for five years more; the other, that he should not suffer a law to be carried that an account should be taken of him though absent. Had I persuaded him on either of these, we should never have fallen into these miseries. And I, again, when Pompey had now made over all his resources, both his own and those of the Roman people, to Caesar, and was beginning too late to see what I had long before foreseen, and when I saw that an unholy war was being brought upon the country, did not cease to be the advocate of peace, concord, and a settlement. And that saying of mine is known to many: ``Would, Gnaeus Pompeius, that you had either never entered a partnership with Gaius Caesar, or never broken it off. The one would have been a mark of your weight, the other of your prudence.'' These, Marcus Antonius, were always my counsels both about Pompey and about the commonwealth. Had they prevailed, the commonwealth would stand; you, by your villainies, your destitution, your infamy, would have fallen.

\ciceroSection{25} But these are old matters; the recent one is that Caesar was killed by my counsel. I am afraid now, senators --- and that would be the basest of things --- that I should seem to have suborned a sham prosecutor to be brought against me, who not only would adorn me with my own praises but would even load me with the praises of others. For who ever heard my name in that fellowship of a most glorious deed? Whose name, on the other hand, of any who had been in that number, was kept secret? Kept secret, did I say? Whose was not at once made public? I should sooner say that some men have boasted of having been in that fellowship although they were not partners, than that any one of them who actually was, was willing to be hidden.

\ciceroSection{26} How probable, again, is it that, among so many men, some of them obscure and some of them young men, no one keeping his secret, my name could have lain hidden? Indeed, if instigators had been needed for the freeing of their country, with the actors ready to hand, would I have prompted Brutuses, each of whom saw daily before him the image of Lucius Brutus, and one of whom saw also that of Ahala? Would these men, then, with such ancestors, have sought counsel rather from strangers than from their own, rather from outside than at home? What? Did Gaius Cassius, born in that family which has never endured the lordship --- not even the power --- of anyone, want me as instigator, I suppose: a man who would, even without these most distinguished men, have finished the business in Cilicia at the mouth of the river Cydnus, had Caesar put in to the bank he had chosen rather than to the opposite one?

\ciceroSection{27} Was it I who roused Gnaeus Domitius to the recovery of liberty --- not the death of his most distinguished father, nor his uncle's, nor his own stripped honour, but my authority? Or did I persuade Gaius Trebonius, whom I should not have dared even to advise? Whereto the commonwealth owes him still greater gratitude, who preferred the liberty of the Roman people to the friendship of one man, and chose rather to be the overthrower of despotism than to share in it. Or did Lucius Tillius Cimber follow me as his author? Whom I marvelled at for doing that deed more than I had thought him likely to do it; and marvelled at on this account, that he had been unmindful of the favours he had received, mindful of his country. What of the two Servilii --- shall I call them the Cascas, or the Ahalas? --- do you think these too were stirred up by my authority rather than by their love of the commonwealth? It would be long to go through the rest: and it was a glorious thing for the commonwealth that they were so many, a glorious thing for the men themselves.

\ciceroSection{28} But now remember in what fashion this acute man convicted me. ``When Caesar had been killed,'' he says, ``Brutus at once held up his bloody dagger, and called Cicero by name, congratulating him on the recovery of liberty.'' Why me, in particular? Because I knew of it? See whether this was not rather the cause of his calling me --- that, having performed an act very like those I had performed, he called me as the most fitting witness that he had stood forth as a rival of my fame.

\ciceroSection{29} But you --- most stupid of all men --- do you not understand that if (the thing of which you accuse me) it is a crime to have wished Caesar killed, it is also a crime to have rejoiced at Caesar's death? For what difference is there between one who urges a deed and one who approves it? Or what does it matter whether I willed that it should be done, or am glad that it was done? Is there anyone, then --- excepting those who rejoiced that he reigned --- who either did not wish that to be done or did not approve it when done? All, then, are guilty. For all good men, so far as in them lay, killed Caesar: to some counsel was wanting, to some courage, to some opportunity; the will was wanting to none.

\ciceroSection{30} But pay heed to the man's stupor --- or rather, his bestial dullness. For thus he spoke: ``Brutus, whom I name in honour, holding the bloody dagger, called out Cicero's name: from which it ought to be understood that he was an accomplice.'' So I am called criminal by you because you suspect I suspected something; while he who carried the dripping dagger before him is named by you in honour. Let it be so. Let there be this stupor in your words: how much greater in your matters and opinions? Settle this at last, consul: do you wish the cause of the Brutuses, of Gaius Cassius, of Gnaeus Domitius, of Gaius Trebonius, and the rest to be what? Sleep off the drunkenness, I say, and breathe it out. Or must firebrands be set to you to rouse you, who sleep through so great a cause? Will you never understand that you must decide whether those who carried out that business were murderers or vindicators of liberty?

\ciceroSection{31} Attend then a little while, and take up for one moment of time the thought of a sober man. I, who am, as I myself confess, an intimate of those men, and, as you charge, their associate, deny that there can be any middle term: I confess that they, unless they are the liberators of the Roman people and the saviours of the commonwealth, are worse than cut-throats, worse than murderers, worse even than parricides --- if indeed it is a more dreadful thing to kill the father of one's country than one's own. And you, a wise and considered man, what do you say? If they are parricides, why have they always been named in honour by you both in this order and before the Roman people? Why was Marcus Brutus, on your motion, exempted from the laws should he have been absent from the city more than ten days? Why were the Apollinarian games celebrated with such unbelievable honour to Marcus Brutus? Why were provinces given to Brutus and Cassius? Why were quaestors added to them, why was the number of their legates increased? Yet these things were done through you. They are not, then, murderers. It follows that, by your own judgment, they are liberators, since there can be no third thing.

\ciceroSection{32} How is it? Am I confusing you? For perhaps you do not sufficiently understand what is said with too great distinctness. But this is the sum of my conclusion: that since by your own judgment they have been acquitted of crime, they have been judged by the same judgment most worthy of the highest rewards. And so I now unsay my speech. I shall write to them that, if any chance ask, as you charged me with, whether the report be true, they should not deny it to anyone. For I fear that to have been concealed by them from this thing would be no honour to me, and to have shrunk back when invited would be the basest shame to me. For what thing was ever, holy Jupiter, done greater not only in this city but in all lands? What more glorious, what more commended to the eternal memory of men? In the partnership of this design, do you, as into the Trojan horse, enclose me with the chief men?

\ciceroSection{33} I do not refuse it; I even give you thanks, in whatever spirit you do it. For so great a thing is it, that I do not compare with its praise the odium you wish to stir up against me. For what is more blessed than those whom you say have been driven out and banished by you? What place is there either so deserted or so uncivilized that, when they have come to it, does not seem to address them and to seek them out? What men so rude that, when they have looked upon them, do not think they have gained the greatest fruit of life? What posterity will be so unmindful, what books will be found so ungrateful, as not to follow their glory with the remembrance of immortality? Then enroll me, for my part, among such a number.

\ciceroSection{34} But there is one thing I fear you will not approve: for if I had been in their counsel, I should have removed from the commonwealth not only the king but the kingship; and, if that pen had been mine that men say was, then, believe me, I should have finished not only the one act but the whole play. Although if to have wished Caesar killed is a crime, look, I beg you, Antony, what is to become of you, who are known beyond all doubt to have taken this counsel with Gaius Trebonius at Narbo, and on whose account, because of your partnership in that counsel, we saw you, when Caesar was being killed, drawn aside by Trebonius. But I, however --- see how I deal with you, not as an enemy --- I praise that you once thought rightly; I give thanks that you did not betray it; I forgive that you did not do it.

\ciceroSection{35} The deed required a man. Yet if any one brought you into court and made use of the Cassian formula, ``to whose advantage was it?'' --- have a care, I beg you, that you do not stick fast. Although indeed it was, as you yourself used to say, to the advantage of all who were unwilling to be slaves, and to you above all, who not only are not a slave but reign --- you who freed yourself from an enormous debt at the temple of Ops; you who, by the same set of ledgers, dissipated a countless sum of money; to whose house so many things were brought from Caesar's; in whose house there is the most lucrative manufactory of forged memoranda and signatures, the most shameful auction-block of lands, of towns, of exemptions, of revenues.

\ciceroSection{36} For indeed what other thing could have come to the rescue of your destitution and your debt except the death of Caesar? You seem somehow disturbed: do you secretly fear that this charge may be seen to touch you? I free you from your fear: no one will ever believe it; it is not your part to deserve well of the commonwealth; the commonwealth has its illustrious authors for that most beautiful deed; I say only that you rejoice at it, I do not allege that you did it. I have replied to the gravest charges: now those that remain must also be answered.

\ciceroSection{37} You threw at me Pompey's camp and that whole period of mine. At which time, indeed, if, as I said, my counsel and my authority had prevailed, you would today be in want, we should be free, the commonwealth would not have lost so many generals and armies. For I confess that when I foresaw what afterwards came to pass, I was in as great a grief as the rest of the best citizens, had they foreseen the same, would have been. I grieved, I grieved, senators, that the commonwealth, once preserved by your counsels and by mine, was within a short time about to perish. Nor indeed was I so untaught or so ignorant of affairs as to be broken in spirit out of greed for life --- a life which, if I retained it, would consume me with anguish, and, if I gave it up, would release me from all troubles. I wished those most distinguished men, those lights of the commonwealth, to live: so many consulars, so many of praetorian rank, so many honourable senators, and besides that the whole flower of the nobility and the youth, and the armies of the best citizens: for if these had lived, however unfair the terms of peace --- for to me every peace with our fellow citizens seemed better than civil war --- we should be holding the commonwealth today.

\ciceroSection{38} Had this opinion of mine prevailed, and had not those very men, on whose lives I was advising, stood most strongly against me out of an inflated hope of victory --- to leave the rest aside --- you, certainly, would never have remained in this order, or rather never have remained in this city. But, you say, my speech estranged the goodwill of Gnaeus Pompeius from me. Did Pompey love any man more, did he hold conversation or counsel with anyone more often? Which indeed was a great thing --- that men who differed on the highest matters of the commonwealth should remain in the same habit of friendship. He saw what I felt and looked to, and I, on the other hand, saw the same of him. I looked first to the safety of the citizens, in order that we might afterwards look to their dignity; he, to their present dignity. Because each of us had something he was following, our disagreement was the more tolerable.

\ciceroSection{39} But what that exceptional and almost godlike man felt about me, those know who followed him in his flight from Pharsalia to Paphos. Never was there any mention by him of me except an honourable one, except such as was full of the warmest friendly regret, when he confessed that I had seen more, but that he had hoped for better. And in the name of such a man you dare to assail me, when you confess yourself the friend of him whose property you have purchased at auction? But let that war be passed over, in which you were too fortunate. Nor shall I answer about the jokes you said I made in camp: those were camps full of cares; yet still men, however turbulent affairs may be, if only they are men, are sometimes relaxed in spirit.

\ciceroSection{40} But that the same man reproaches my sadness and the same my jocularity is great proof that in both I held a measure. You have said that no legacies come to me. Would that this charge of yours were true! More of my friends and connections would be alive. But how did that come into your mind? For I have entered on the books legacies of more than twenty millions of sesterces. Although in this respect I confess you are more fortunate. No one but a friend has made me his heir, that so, with whatever benefit there was, some grief at heart should be joined; while you, Lucius Rubrius of Casinum --- a man you had never seen --- made you his heir.

\ciceroSection{41} And see indeed how much he loved you, this man whose very colour, black or white, you do not know. He passed over his brother's son --- of Quintus Fufius, a most honourable Roman knight and his closest friend --- whom he had always openly named as his heir, and does not even mention him: he made you his heir, whom he had never seen, or certainly never greeted. Tell me, please, if it is not too much trouble: Lucius Turselius --- what was his face, what his stature, of what municipality, of what tribe? ``I know nothing,'' you will say, ``except what estates he held.'' So, disinheriting his brother, he was making you his heir. And besides this, you broke into many large estates of utter strangers, having driven out their true heirs by force, as though you were heir.

\ciceroSection{42} Although what I have most marvelled at was this, that you dared to make mention of legacies, when you yourself had not taken your father's. So, that you might collect these things, you most insane of men, you have been declaiming so many days in another man's villa? Although you indeed, as your most intimate friends say, declaim for the sake of breathing out your wine, not for the sake of sharpening your wit. But yet, for the sake of a joke, you keep at hand a master --- a rhetorician elected by your vote and that of your fellow-drinkers, to whom you have allowed to say what he pleases against you --- a man witty enough, but with easy matter for sayings against you and yours. But see what a difference there is between you and your grandfather. He spoke gradually what was helpful to his case; you rush along with what is irrelevant.

\ciceroSection{43} But what a fee was paid to that rhetorician! Hear, hear, senators, and learn the wounds of the commonwealth. Two thousand acres of the Leontine plain you assigned to the rhetorician Sextus Clodius, and free of tax, that for so great a fee of the Roman people you should learn to know nothing. Did this too, you most audacious of men, come from Caesar's memoranda? But I shall speak in another place of the Leontine land and of the Campanian, which lands he has fouled by snatching them from the commonwealth and giving them to the basest possessors. Now, since I have replied sufficiently to his charges, certain things must be said about the corrector and reformer of us himself. Nor shall I pour out all at once: so that, if there should be a further contest (and there will be), I may always come fresh; a faculty which the multitude of his vices and sins amply grants me.

\ciceroSection{44} Shall we then inspect you from boyhood? So I think; let us begin from the beginning. Do you hold it in memory that, still wearing the boy's bordered toga, you went bankrupt? ``It is the fault of my father,'' you will say. I grant it. Indeed, this is a defence full of filial piety. Yet this still belongs to your audacity, that you sat in the fourteen rows, when by the Roscian law a fixed place was set down for bankrupts, even for one who had gone bankrupt by the fault of fortune, not his own. You took on the man's toga, which you straightway turned into a woman's. At first a common prostitute; the wage of your shame was fixed, and not a small one; but soon Curio intervened, who drew you off from your trade as a courtesan, and, as if he had given you a matron's stole, settled you in a fixed and steady marriage.

\ciceroSection{45} No boy bought for lust was ever so much in the power of his owner as you were in Curio's. How often did his father throw you out of his house, how often did he set guards to keep you from crossing the threshold? Yet you, with night for accomplice, with lust urging, with the wage driving, were let down through the tiles. Such infamies that house could endure no longer. Do you know that I am speaking of things known to me most fully? Recall that time when the elder Curio lay grieving on his bed; his son, throwing himself at my feet, weeping, commended you to me; begged that I would defend him against his own father, if he should demand back the six million sesterces --- for he said he had gone bail for you to that amount. He himself, burning with love, kept declaring that, because he could not bear the longing of being parted from you, he would go into exile.

\ciceroSection{46} What great evils of a most flourishing family did I then quiet, or rather lift away! I persuaded the father to discharge his son's debt; to ransom the young man, endowed with the highest hope both of spirit and of talent, with the means of the family estate; and to keep him not only from your intimacy but even from meeting you, by the law and power of a father. Since you knew that all this had been done through me, would you have dared to challenge me with insults, had you not trusted those swords we see?

\ciceroSection{47} But let us now leave aside the rapes and the infamies: there are some things I cannot honourably name; while you, on the other hand, are the freer from constraint because you have admitted to deeds about you which you could not even hear from a modest enemy. But look at the rest of the course of his life, which I shall run through swiftly. For my spirit hurries on to the things he did in the civil war, in the gravest miseries of the commonwealth, and to those he does daily. Hear them attentively, as you do, I beg --- though they are far better known to you than to me. For in such matters the spirit should be roused not only by the knowledge of things but by their recollection; although, I think, we should cut through the middle, lest we come too late to the end.

\ciceroSection{48} In the tribunate he was the intimate of Clodius --- yes, of him whose kindnesses to me he keeps reminding me of; he was the firebrand of all his fires, in whose house, even then, he was contriving something. He himself best knows what I mean. Thence the journey to Alexandria against the authority of the Senate, against the commonwealth, against religion; but he had Gabinius as his commander, with whom anything could be done in the rightest fashion. What then was the manner of his return, or of what kind? From Egypt, first into furthest Gaul rather than home. What home? For every man at that time had his own house, but you had none anywhere. Home, do I say? Where on the earth was there for you to set foot in your own, except the one Misenum, which you held with partners as men hold the Sisapo mines?

\ciceroSection{49} You come from Gaul to stand for the quaestorship. Dare to say that you came to your mother before you came to me. I had received earlier a letter from Caesar that I should suffer satisfaction to be made me by you: and so I did not even let you speak about a reconciliation. Afterwards I was courted by you, you cultivated by me at your canvass for the quaestorship; at which time, with the approval of the Roman people, you attempted to kill Publius Clodius in the forum, and although you tried it of your own accord, not at my prompting, yet you said publicly that you did not think you would ever, except by killing him, make satisfaction to me for the wrongs you had done me. In which I much wonder why you say that Milo did that deed at my prompting, when you, of your own offer, were proposing to me the same thing, and I never urged you on. Although, if you persisted in it, I preferred that the matter should be referred to your glory than to my favour.

\ciceroSection{50} You were made quaestor: then at once, without a decree of the Senate, without lots, without a law, you ran off to Caesar. For that was the one refuge, in your destitution, debt, worthlessness, and ruined manner of life, that you thought there was upon the earth. There, when you had filled yourself both with his largesses and with your own robberies --- if to fill is to gulp down what one at once pours out again --- you flew, in want, to the tribunate, that in that office, if you could, you might be like your man. Hear now, I beg, not what he has done shamefully and intemperately against himself and his domestic honour, but what he has done impiously and wickedly against us and our fortunes --- that is, against the universal commonwealth. For from his crime, you will find, the beginning of all our evils was born.

\ciceroSection{51} For when in the consulship of Lucius Lentulus and Gaius Marcellus, on the Kalends of January, you were trying to prop up the commonwealth as it slipped and was almost falling, and were willing, even for Gaius Caesar himself, if he was of sound mind, to consult --- then this man set his tribunate, sold and made over, against your counsels, and offered his neck to that axe which has cut off many men for lesser offences. Against you, Marcus Antonius, this Senate decreed --- this Senate still entire, with so many of its lights not yet extinguished --- what is usually decreed in the manner of our ancestors against an enemy in the toga. And you have dared to speak against me before the senators, when I had been judged the saviour by this order, and you the enemy of the commonwealth? The mention of that crime of yours has been broken off, but the memory has not been wiped out. So long as the race of men, so long as the name of the Roman people shall stand --- which shall stand forever, if you will allow it --- that pestilent veto of yours shall be named.

\ciceroSection{52} What was being done in the Senate eagerly, what rashly, when you, one young man, prevented the whole order from deciding about the safety of the commonwealth --- not once but more than once --- nor did you suffer anything to be done about the authority of the Senate with you in the way? But what was being done, except that you should not wish the commonwealth utterly to be destroyed and overturned, when neither the leaders of the state, by entreating, nor your elders, by warning, nor the assembled Senate, by acting, could move you from a vote that was sold and given up in advance? Then, after many things had been tried before, that wound was of necessity inflicted on you which has been inflicted upon few before you, none of whom escaped uninjured:

\ciceroSection{53} then this order gave arms against you to the consuls, and to the rest of the commands and powers: which you would not have escaped, had you not betaken yourself to the arms of Caesar. You, you, I say, Marcus Antonius, gave to Gaius Caesar, eager to throw all into confusion, the first cause for bringing war against his country. For what else was he alleging, what cause was he putting forward for his most insane plan and act, except that the veto had been disregarded, the tribunician right abolished, Antonius hemmed in by the Senate? I pass over how false these things are, how light, especially when there can be no just cause at all for any man to take up arms against his country. But of Caesar nothing: you, at any rate, must confess that the cause of a most ruinous war was lodged in your person.

\ciceroSection{54} O wretched man if you understand this, more wretched still if you do not understand that this is committed to writing, this is handed down to memory, that not even the posterity of all the ages will ever be unmindful of it: that consuls were driven out of Italy, and with them Gnaeus Pompeius, who was the ornament and the light of the Roman people's command, all the consulars who through ill-health could not follow that disaster and that flight, the praetors, the men of praetorian rank, the tribunes of the plebs, a great part of the Senate, the whole growing crop of the youth --- in one word, the commonwealth driven and banished from its proper seats!

\ciceroSection{55} As, then, in seeds is the cause of trees and shrubs, so of this most mournful war you were the seed. You grieve that three armies of the Roman people have been killed: Antonius killed them. You long for our most distinguished citizens: them too Antonius snatched from you. The authority of this order has been struck down: Antonius struck it down. In short, all the evils we have seen since --- and what evil have we not seen? --- if we reckon rightly, we shall set down to the account of Antonius alone. As Helen was to the Trojans, so was this man to this commonwealth, the cause of war, the cause of pestilence and ruin. The rest of his tribunate was like its beginning. He carried through all the things which the Senate, with the commonwealth still safe, had brought about could not be done. Yet learn the crime upon crime that he has piled.

\ciceroSection{56} He was restoring many men who had been ruined: among them no mention of his own uncle. If he was strict, why not against all? If he was merciful, why not toward his own people? But I leave the rest aside: Licinius Lenticula, condemned of gaming by the dicing-law, his own fellow-gambler, he restored --- as though it were not lawful to play with one already condemned, but in order that what he had lost at dice he might pay off by the favour of the law. What reason did you give the Roman people why this man ought to be restored? That he was, no doubt, prosecuted in absence; that the case had been judged unheard; that there had been no real legal inquiry into gaming; that he had been overborne by force and arms; finally --- what was said of your own uncle --- that the court had been corrupted with money? None of these. But he was a good man and worthy of the commonwealth. That has no bearing on the matter; yet for my part, since being condemned is in itself nothing, if it were so I should forgive him. But the most worthless of all men, who did not hesitate even to play at dice in the forum, this man, condemned by the law on gaming, he has restored to full standing: does he not most openly profess his own taste?

\ciceroSection{57} In that same tribunate, when Caesar, on his way to Spain, had committed Italy to him to trample on, what a journey he made, what a tour of the municipal towns! I know that I am moving in things made widely public by the talk of all men, and that what I am saying and shall say is better known to all who were then in Italy than to me who was not. Yet I shall mark down each detail, although in no way will my speech come up to your knowledge. For what ever in the world was heard of, of so great an outrage, so great a baseness, so great a disgrace?

\ciceroSection{58} A tribune of the plebs was being carried in a war-chariot; laurelled lictors went before him; among them an open litter was borne in which a mime-actress was carried, whom the honest townsmen of the country towns, coming out of necessity to meet him, greeted not by her well-known and theatrical name, but as Volumnia. There followed a four-wheeled coach with pimps, the basest of his companions; his mother, cast aside, followed behind the mistress of her impure son as if she had been her daughter-in-law. O the calamitous fertility of that miserable woman! With the footprints of these infamies that fellow stamped all the municipal towns, the prefectures, the colonies, in short the whole of Italy.

\ciceroSection{59} Of the rest of his deeds, senators, the censure is indeed difficult and slippery. He took part in the war; he glutted himself with the blood of fellow-citizens utterly unlike himself: he was fortunate, if there can be any fortune in crime. But since we wish provision to be made for the veterans --- although the soldiers' case is unlike yours, for they followed, you sought out a leader --- nevertheless, that you may not call me into hatred with them, I shall say nothing of the kind of war it was. As victor you returned from Thessaly to Brundisium with your legions. There you did not kill me. A great kindness! For I admit you could have. And yet there was no man of those who were then with you who did not think me worth sparing.

\ciceroSection{60} For such is the love of one's country that I was held sacred even by your legions, because they remembered that it had been saved by me. But suppose that you gave me what you did not take away, and that I have my life from you because it was not snatched from me: was I to allow this kindness of yours, through all your insults, to be guarded by me as I was guarding it, especially when you saw you were going to hear what follows?

\ciceroSection{61} You came to Brundisium, into the very lap and embrace of your little mime. What of it? Am I lying? How wretched it is to be unable to deny what it is most shameful to confess! If you felt no shame before the country towns, did you feel none even before the veteran army? For what soldier was there who did not see her at Brundisium? Who did not know she had come so many days' journey to congratulate you? Who did not grieve to learn so late what a worthless creature he had followed?

\ciceroSection{62} Again, a tour through Italy with that same mime as companion; into the country towns a cruel and miserable settling of soldiers; in the city a foul plundering of gold, of silver, and most of all of wine. To which it was added that, with Caesar unaware, since he was at Alexandria, by the favour of his friends Antony was appointed master of horse. Then he supposed himself entitled by right to live with Hippias and to hand over the public post-horses to Sergius the mime-player; then he had picked out for his residence not this house which he now ill maintains, but the house of Marcus Piso. Why should I produce his decrees, his plunderings, the inheritances he conferred and the inheritances he tore away? Want compelled him; he had nowhere to turn: not yet had so vast a legacy come to him from Lucius Rubrius, nor from Lucius Turselius; not yet had he succeeded as sudden heir into the place of Gnaeus Pompeius and of many others who were away. He had to live in the manner of brigands, having only what he could plunder.

\ciceroSection{63} But let us pass over those deeds that belong to a more sturdy wickedness: let us speak rather of the most worthless kind of frivolity. You, with those jaws of yours, those flanks, that gladiator's solidity of the whole body, had drained off such a quantity of wine at Hippias's wedding that you were forced, the next day, to vomit in the sight of the Roman people. O the thing foul not only to see but to hear! If at dinner, in those huge cups of yours, this had happened to you, who would not think it disgraceful? But in an assembly of the Roman people, transacting public business, as master of horse, for whom even a belch would be a disgrace, this man, vomiting, filled his own lap and the whole tribunal with bits of food reeking of wine. But these things he himself confesses to be among his squalors: let us come to his splendours.

\ciceroSection{64} Caesar withdrew himself from Alexandria, fortunate, as he himself thought, but in my judgment --- since he who is unfortunate for the commonwealth can be fortunate in no respect --- not fortunate at all. The spear was set up before the temple of Jupiter Stator, and the property of Gnaeus Pompeius --- wretched me! for though my tears have been spent, my grief stays fixed within me --- the property, I say, of Gnaeus Pompeius Magnus was made subject to the bitterest voice of the auctioneer. In that one matter alone the commonwealth, forgetful of its servitude, groaned aloud; and though men's spirits were enslaved and all things were held in fear, yet the groan of the Roman people was free. When all were waiting to see what man would be found so impious, so deranged, so hostile to gods and men, that he would dare to approach that wickedness of buying a confiscated estate, no one was found but Antony --- and that, though there were so many around that spear who would have dared all things else: one man was found who would dare what the audacity of all the rest had fled and recoiled from.

\ciceroSection{65} So great then a stupor possessed you, or, to speak more truly, so great a frenzy, that, first, when you of all men are the buyer of confiscated property, and second when you are the buyer of Pompey's confiscated property, you do not know yourself accursed to the Roman people, do not know yourself loathsome, do not know that all the gods, all men are and shall be your enemies? And how insolently the glutton at once leapt upon the fortunes of that man by whose valour the Roman people was more dread to foreign nations, by whose justice more dear! Into that man's wealth then, when he had suddenly gorged himself, he revelled with joy, an actor out of a mime, just now in want, suddenly rich. But, as it is in some poet, ``Ill gained, ill spent.''

\ciceroSection{66} It is incredible, and like a portent, by what means he poured away in so few --- I do not say months but days --- a treasure so vast. There was a huge quantity of wine, a very great weight of finest silver, costly clothing, much and elegant furniture, and many splendid pieces in many places --- not, indeed, those of a luxurious man, but the furnishings of an abundant one. In a few days nothing of these was left. What Charybdis was ever so voracious?

\ciceroSection{67} Do I call her Charybdis? She, if she ever was, was a single creature: Ocean itself, by Hercules, scarcely seems to have been able so quickly to engulf so many things, so scattered, set in places so remote. Nothing was locked, nothing sealed, nothing inventoried. Whole storehouses were given over to the basest of men; some things the male mime-players plundered, others the female; the house was packed with gamblers, full of drunkards; whole days through they drank, and in many places at once; gambling losses were piled on too --- for not always was this man fortunate; on slaves' beds in their cells you could see Gnaeus Pompeius's purple coverlets spread. Cease, then, to wonder that these things were consumed so quickly. Not only the patrimony of one man, however ample, as that was, but cities and kingdoms, so great a worthlessness could swiftly have devoured. But he took the house too, and the gardens.

\ciceroSection{68} O monstrous audacity! Did you even dare to enter that house, to cross that most sacred threshold, to show your most impure face to the household gods of those rooms? A house which for a long time no man could look upon, no man pass by without tears --- in this house are you not ashamed to lodge so long? In which, however empty of wits you are, still nothing can give you pleasure. When you look upon those beaks of ships in the vestibule, do you suppose that you are entering your own house? It cannot be. However devoid of mind and feeling you may be --- as you are --- yet you know yourself and your possessions and your people. Nor, indeed, do I believe you can ever, waking or in dreams, have a quiet mind. It must be that, violent and frenzied as you are, when the figure of that single man is set before you, you start up in terror from sleep, and rave often even when awake.

\ciceroSection{69} For my part I pity the very walls and roofs. For what had that house ever seen except what was chaste, what except from the best custom and the holiest training? For that man was, senators, as you know, as illustrious abroad as he was admirable at home; not more to be praised for his foreign achievements than for his domestic discipline. In his halls now there are stables instead of bedrooms, taverns instead of dining rooms. Although now he denies it: do not ask; he has reformed. He bade his wife --- the famous one --- take her things and go; he took her keys from her by the Twelve Tables; he turned her out. What a man esteemed as a citizen, how approved! Of his whole life there is nothing more honourable than that he divorced his mime.

\ciceroSection{70} But how often he says: ``Antony, both consul and Antony!'' This is to say, both consul and most unchaste, both consul and most worthless of men. For what else is Antony? For if rank had been signified by the name, your grandfather, I believe, would once have said that he was both consul and Antony. He never said it. My colleague, your uncle, would have said it too, unless you are the only Antony. But I leave aside those faults which are not proper to the very part by which you have harried the commonwealth: I return to your own proper part, that is, to the civil war, which was born, kindled, and undertaken by your doing.

\ciceroSection{71} From which war, both through your timidity and through your lusts, you absented yourself. You had tasted citizen blood, or rather drained it down; you had been a standard-bearer in the line at Pharsalus; you had killed Lucius Domitius, a most illustrious and most noble man, and many besides who had fled from the battle, men whom Caesar, as he did some, might perhaps have saved, you most cruelly pursued and butchered. Such things being done and so many, what was the reason that you did not follow Caesar into Africa, especially as so great a part of the war was left? And so what place did you obtain with Caesar himself after his return from Africa? In what rank were you? You, whose quaestor he had been the commander, you, master of horse to him the dictator, prime mover of the war, the author of cruelty, the partner of his plunder, his ``son'' by his will, as you yourself used to say --- you were dunned for the money you owed for the house, for the gardens, for the auction-lot.

\ciceroSection{72} At first you answered with plain ferocity, and --- not that I may seem to be all against you --- what you said was nearly equitable and just: ``Money? From me? Gaius Caesar? Why rather than I from him? Could he have conquered without me? Indeed he could not. I brought to him the cause of the civil war; I proposed pernicious laws; I bore arms against the consuls and commanders of the Roman people, against the Senate and the Roman people, against our ancestral gods, our altars, our hearths, against our country. Has he conquered for himself alone? Whose deed is in common, why should not their plunder be in common?'' You demanded your right; but what was that to the matter?

\ciceroSection{73} He had more power. And so, after he had shaken off your protests, he sent his soldiers both upon you and upon your sureties, when suddenly that famous list was brought out by you. What laughter from men, that there should be such a list, of such variety, of so many possessions, of which, apart from a part of Misenum, there was nothing the auctioneer could call his own. But the look of the auction itself was wretched: Pompey's clothing, not abundant and that stained; some battered silver vessels of his; squalid slaves --- so that we mourned that anything from those relics remained for us to see.

\ciceroSection{74} Yet this auction the heirs of Lucius Rubrius forbade by Caesar's decree. The scoundrel was at a loss: he had nowhere to turn. Indeed, at those very times an assassin sent by him was said to have been seized in Caesar's own house with a dagger; about whom Caesar openly inveighed against you in the Senate with complaint. Caesar set out for Spain, with a few days extended to you for paying because of your indigence. Even then you do not follow. So good a gladiator, so soon dismissed? Should anyone then fear this man, who in his own party --- that is, in his own fortunes --- was so timid?

\ciceroSection{75} At length he set off for Spain; but, as he says, he could not arrive safely. How then did Dolabella arrive? Either, Antony, that cause should not have been undertaken, or, when you had undertaken it, defended to the end. Three times Caesar fought to the finish with citizens, in Thessaly, in Africa, in Spain. Dolabella was present in all these battles; in the Spanish one he even received a wound. If you ask my judgment, I should have wished he were not: but, while the plan from the beginning was to be censured, his constancy was to be praised. But what are you? At that time the sons of Gnaeus Pompeius were first seeking back their country. Granted; let this have been the common cause of the party. They were seeking back besides their ancestral gods, altars, hearths, their household Lar, into which you had broken. While they were seeking with arms what was theirs by law --- though in things so most unjust what can there be of equity? --- still, whom was it most equitable to fight against the sons of Gnaeus Pompeius? Whom? You, the auction-buyer of his goods.

\ciceroSection{76} Was it that you, at Narbo, might vomit on the tables of your hosts, while Dolabella in your stead fought in Spain? And what a return from Narbo! He even kept asking why I had so suddenly turned back from my journey. I lately explained, senators, the cause of my return. I wished, if I could, to be of use to the commonwealth even before the Kalends of January. For, as to your asking how I came back, first by light, not by darkness; then in shoes and toga, in no Gallic slippers and no cloak. But you even look at me, and that, as it seems, in anger. You would surely be reconciled with me now, if you knew how ashamed I am of your worthlessness, of which you yourself are not ashamed. Of all the disgraces of all men I have never seen, never heard of, anything more shameful. You who fancied yourself master of horse, you who for the next year were soliciting --- or rather asking for --- the consulship, ran through the country towns and colonies of Gaul, from which we, when the consulship was sought, not asked for, used to seek the consulship, in Gallic slippers and a cloak.

\ciceroSection{77} But mark the man's frivolity. When at about the tenth hour of the day he had reached Red Rocks, he hid in a certain little tavern, and there, concealing himself, drank himself drunk until evening; from there he was hurried in a light cart to the city, came to his house with his head muffled. The doorkeeper says: ``Who are you?'' ``A messenger from Marcus.'' He is brought straight to the woman for whose sake he had come, and handed her a letter. As she read it weeping --- for it was written in the manner of a lover; the head of the letter was that henceforth he would have nothing with that mime-actress; that he had cast off all his love from there and poured it into her --- as the woman wept the more abundantly, the tender-hearted man could not bear it, uncovered his head, fell upon her neck. O worthless man! For what else am I to say? Nothing can I say more aptly. So, to make a woman set eyes upon you, my catamite, unhoped-for, when you showed yourself unexpectedly --- for that you struck the city with nightly terror and Italy with many days of fear?

\ciceroSection{78} And at home indeed you had a cause in love, abroad one yet more disgraceful, lest Lucius Plancus should sell up your sureties. Brought forward into a public meeting by a tribune of the plebs, when you had answered that you had come on your own business, you turned even the people into wits at your expense. But enough of trifles: let us come to greater things. You went very far out to meet Gaius Caesar on his return from Spain. You went quickly, came back quickly, that he might learn you to be, if not brave, yet at least energetic. You became, somehow, his familiar again. Caesar had this trait wholly: any man whom he had known to be plainly ruined in debt and in want, if he had known the same man also to be a wastrel and bold, this man he took most willingly into intimacy.

\ciceroSection{79} So by these splendid recommendations you were ordered to be returned consul, and that, with the man himself. I do not complain about Dolabella, who was then pushed, induced, deluded. In that matter, how great was the bad faith of each of you toward Dolabella, who is ignorant? Caesar induced him to stand, and then turned aside what had been promised and accepted, transferring it to himself; you put your own consent to his bad faith. The Kalends of January come; we are gathered into the Senate: Dolabella inveighed against him much more copiously and more elaborately than I do now.

\ciceroSection{80} But what, in his anger, he said, good gods! First, when Caesar had shown that, before he set out, he would order Dolabella to be consul --- and they deny him to be a king, who both did always such a thing and said it ---  but when Caesar had spoken thus, then this good augur said that he had been endowed with such a priesthood that he could either obstruct the elections by the auspices or vitiate them, and asserted that he would do it. In this thing first learn the man's incredible stupidity.

\ciceroSection{81} What then? That which you said you could do by the right of your priesthood, could you have done any less, if you had not been augur but had been consul? Look out --- perhaps you could have done it even more easily. For we have only the right of declaration; consuls and the other magistrates have also the right of observation. Granted: this is incompetence, for prudence is not to be demanded of a man who is never sober; but mark his impudence. Many months before, in the Senate, he said he would either obstruct Dolabella's elections by the auspices, or do what he did. Can anyone foretell what flaw there is going to be in the auspices, unless he has resolved to watch the heavens? Which neither is permissible during the holding of the comitia by the laws, and if anyone has watched, he ought to announce it not when the comitia have been held but before they are held. But his impudence is entangled with his ignorance: he neither knows what becomes an augur, nor does he do what becomes a man of shame.

\ciceroSection{82} And recall his consulship from that day down to the Ides of March. What attendant was ever so abject, so low? He himself was good for nothing; he asked for everything; thrusting his head into the back of his colleague's litter, he begged from his colleague the favours he was peddling. Behold the day of Dolabella's comitia. The sorting of the prerogative century: he keeps quiet. It is announced: he says nothing. The first class is called, then, as is the custom, votes; then the second class --- all of which was done quicker than I have said it.

\ciceroSection{83} The business concluded, the good augur --- you would have said it was Gaius Laelius --- says: ``On another day.'' O singular impudence! What had you seen, what had you perceived, what had you heard? For neither did you say that you had been watching the heavens, nor do you say so today. That flaw, then, occurred which you had foreseen even on the Kalends of January, and had foretold so far in advance. So, by Hercules, with great damage --- I hope --- rather to yourself than to the commonwealth, you have invented auspices; you have entangled the Roman people in religious obligation; you, augur to augur, you, consul to consul, made your evil announcement. I will say no more, lest I seem to overturn the acts of Dolabella, which must one day be referred to our college. But mark the man's arrogance and insolence.

\ciceroSection{84} As long as you wish, Dolabella is a flawed consul; again, when you wish, he was created with the auspices intact. If there is no force in those words by which an augur announces his decision, as you announced, confess that, when you said ``On another day,'' you were not sober; but if there is some force in those words, I ask the augur from his colleague what that force is. But that, from the many achievements of Marcus Antonius, my speech may not happen to leap past one of singular beauty, let us come to the Lupercalia. He does not dissemble, senators: it is plain that he is shaken; he sweats, he turns pale. Let him do what he will, only let him not turn sick --- as he did in the Porch of Minucia. What defence can there be of such great disgracefulness? I am eager to hear, that I may see where the great fee of the rhetorician shows itself, where the field of Leontini appears.

\ciceroSection{85} Your colleague was sitting on the Rostra, wrapped in a purple toga, on a golden chair, crowned. You go up, you approach the chair --- so you were a Lupercus that you ought to have remembered you were a consul --- you display a diadem. A groan throughout the Forum. Where did the diadem come from? You had not picked one up that had been thrown down; you had brought it from home, a crime premeditated and considered. You were placing the diadem upon him to the lamentation of the people; he was rejecting it to their applause. You, therefore, you alone, criminal, were found, who, while you were the author of monarchy and would have him whom you had as colleague for your master, were trying at the same moment what the Roman people could bear and endure.

\ciceroSection{86} Nay, even you were trying to win pity: a suppliant you threw yourself at his feet. Asking what? To be a slave? You should have asked it for yourself alone, who from boyhood had so lived as to endure everything, that you would readily be a slave; from us and from the Roman people you certainly had no such commission. O that splendid eloquence of yours, when you spoke in a public meeting naked! What more shameful than this, what more foul, what more worthy of every punishment? Are you waiting for us to goad you with the spur? This, if you have any part of feeling, lacerates you; this speech draws blood. I am afraid I may diminish the glory of the highest men; yet I shall speak, moved by indignation. What is more unworthy than that he should live who placed the diadem, when all confess that he was justly killed who rejected it?

\ciceroSection{87} He even had it set down in the calendar at the Lupercalia: ``To Gaius Caesar, dictator perpetual, Marcus Antonius the consul by command of the people offered the kingship; Caesar refused.'' I now scarcely wonder you disturb the peace; that you hate not only the city but even the daylight; that with the most desperate brigands you drink not only from day to day but from one day into the next. For where can you find rest in peace? What place can there be for you in the laws, in the courts, which, so far as in you lay, you abolished by your kingly domination? Was Lucius Tarquinius therefore driven out, Spurius Cassius, Spurius Maelius, Marcus Manlius slain, that many ages after, a king at Rome should be set up by Marcus Antonius, which is impious?

\ciceroSection{88} But let us go back to the auspices, of which Caesar was going to treat in the Senate on the Ides of March. I ask: what would you then have done? I had heard, indeed, that you had come prepared, because you thought I was going to speak about your fabricated auspices, which yet had to be obeyed. The fortune of the commonwealth took away that day. Did the death of Caesar take away also your judgment about the auspices? But I have come upon the time which must take precedence of the matters on which my speech had entered. What flight of yours, what panic on that glorious day, what despair of life from consciousness of your crimes, when from that flight, by the kindness of those who wished you safe, if you should be sane, you stole home in secret!

\ciceroSection{89} O my auguries of things to come, ever in vain most true! I told those liberators of ours on the Capitol, when they wished me to go to you to urge you to defend the commonwealth, that, as long as you feared, you would promise all things; as soon as you ceased to fear, you would be like yourself. So while the rest of the consulars went and came, I held to my judgment: I neither saw you that day nor the next, nor did I believe that any alliance of best citizens with a most savage enemy could be confirmed by any covenant. On the third day I came to the temple of Tellus, and that against my will, since all the approaches were beset by armed men.

\ciceroSection{90} What a day was that for you, Antony? Although you have suddenly stood forth as my enemy, yet I pity you, because you have envied yourself. What a man you would have been, immortal gods, and how great, if you had been able to keep the mind of that day! We should have peace, which had been made through a noble boy as hostage, the grandson of Marcus Bambalio. Although fear was making you good --- not a lasting teacher of duty --- audacity, which never leaves you when fear is absent, made you wicked. And yet, then, when men thought you most excellent --- I dissenting --- you most criminally took charge of the funeral of the tyrant, if that was a funeral.

\ciceroSection{91} That fine eulogy was yours, that lamentation, that exhortation; you, you, I say, lit those torches, both those with which he was half-burned and those with which the house of Lucius Bellienus was burned to the ground. You let loose upon our houses those onsets of desperate men, and largely of slaves, which we ourselves drove back with force and hand. Yet the same man, as if the soot were wiped off, on the following days passed in the Capitol splendid decrees of the Senate, that no tablet of immunity or of any privilege should be posted after the Ides of March. You yourself remember what you said about exiles, you know what you said about immunity. The best thing of all was that you abolished the name of the dictatorship from the commonwealth in perpetuity: by which act you seemed to have conceived so great a hatred of kingship that you were taking away all fear of it on account of the late dictator. To others the commonwealth seemed constituted,

\ciceroSection{92} but to me by no means, who, with you at the helm, feared every kind of shipwreck. Have I, then, been deceived? Or could he longer remain unlike himself? Before your very eyes throughout the whole Capitol tablets were being posted, and not only were immunities sold to individuals, but to whole peoples: citizenship was granted no longer one by one, but to whole provinces. Therefore, if these things stand which cannot stand while the commonwealth stands, you have lost, senators, whole provinces; and not the revenues only but the empire too of the Roman people has been diminished by this domestic huckstering.

\ciceroSection{93} Where is the seven hundred million which is on the tablets at the temple of Ops? Money indeed of ill omen, but yet money which, if it were not paid back to those whose it was, might rescue us from tributes. How is it that, of the forty million sesterces which you owed on the Ides of March, you had ceased to owe them before the Kalends of April? There are indeed countless things bought of your people not without your knowledge; but one outstanding decree, posted in the Capitol, concerning King Deiotarus, most loyal friend of the Roman people: which when it was set forth, there was no one who in his very pain could keep from laughing.

\ciceroSection{94} For who was ever a greater enemy to anyone than Caesar to Deiotarus? As much so as to this order, to the equestrian, to the Massiliotes, to all whom he perceived to hold the commonwealth of the Roman people dear. From a man, then, from whom living, neither present nor absent, King Deiotarus had ever obtained any just or generous thing, he became, when dead, well-favoured. Caesar had reproached his guest to his face, had reckoned up his money, had placed in his tetrarchy one of his Greek companions, had taken away Armenia which the Senate had given him. These things he stripped away while living, he returns when dead.

\ciceroSection{95} But in what terms! Now it seemed to him just, now not unfair. Wonderful complication of words! But Deiotarus never --- for I was always present for him in his absence --- said that anything we asked for him seemed just to himself. A bond for ten million sesterces was made by his envoys, good men but timid and inexperienced, without our advice, without the advice of the king's other guest-friends, in the women's quarters --- the place in which many things have been and are sold. With regard to this bond, you had better, I think, consider what you are to do; for the king himself of his own accord, with no record of Caesar's, as soon as he heard of Caesar's death, recovered his property by his own war.

\ciceroSection{96} The wise man knew it had always been law that what tyrants had snatched away, when the tyrants were killed, those from whom it had been snatched should recover. No jurist, then, not even that one who is the sole jurist for you, through whom you do these things, will say that anything is owed under that bond for things that had been recovered before the bond. For he did not buy them from you, but possessed them himself before you sold to him what was his. He was a man. We, indeed, are to be despised, who hate the author and defend the acts.

\ciceroSection{97} What shall I say of the countless memoranda, what of the innumerable handwritten orders? Of these even retailers are at work, who hawk them in public like gladiators' bills. And so such heaps of money are heaped up in his house that the money is now being weighed, not counted. But how blind is greed! Lately a tablet was posted by which the wealthiest cities of the Cretans are freed from tribute, and it is decreed that after Marcus Brutus's proconsulship Crete shall not be a province. Are you in possession of your mind? Are you not to be put in chains? Could Crete be freed by Caesar's decree after the departure of Marcus Brutus, when Crete had no connection with Brutus while Caesar was alive? But by the sale of this decree --- lest you think nothing was done --- you have lost the province of Crete. Altogether no man was the buyer of anything who lacked this seller.

\ciceroSection{98} And as for the law about exiles which you posted, did Caesar pass it? I attack no man's calamity: only I complain, first, that the returns of those whose case Caesar judged dissimilar have been polluted; next, I do not see why you do not grant the same to the rest, for not more than three or four are left. Those who are in a similar calamity, why do they not enjoy a similar mercy of yours, why do you hold them in the position of your uncle? Concerning him you were unwilling to bring forward a bill when you brought it for the rest: yet you even drove him to seek the censorship, and engineered a candidacy that aroused both the laughter and the complaints of men. But why did you not hold those elections?

\ciceroSection{99} Was it because a tribune of the plebs reported a sinister thunderbolt? When your interest is at stake, there are no auspices; when your relatives', then you become scrupulous. What? Did you not also let him down in the office of one of the Seven? For there came in your way one whom you feared, I suppose, lest you should be unable, with your head intact, to refuse him. You loaded with every insult the man whom, in your father's place, if there were any natural piety in you, you ought to have honoured. You cast off his daughter, your own kinswoman, after seeking out and previously inspecting another match. That was not enough: you charged a most chaste woman with shameful conduct. What can be added to that? Not content with that, in a most crowded Senate on the Kalends of January, with the uncle himself sitting there, you dared to say that your cause of hatred with Dolabella was that you had discovered he had committed adultery with your sister and wife. Who can interpret whether you were more impudent in saying it in the Senate, or more depraved in saying it against Dolabella, or fouler in saying it before her father's hearing, or more cruel against that wretched woman in speaking so filthily, so impiously.

\ciceroSection{100} But let us go back to the handwritten orders. What hearing was given by you? For Caesar's acts were confirmed by the Senate for the sake of peace: namely, what Caesar had done, not what Antony said Caesar had done. Whence do these things break out? On whose authority are they produced? If they are false, why are they approved? If they are true, why are they sold? But the resolution had been that from the Kalends of June, you, with a council, would examine Caesar's acts. What council was there? Whom did you ever convoke? What Kalends of June did you wait for? Or those to which you returned, after traversing the colonies of the veterans, hedged about with arms? O that famous tour of yours in April and May, when you even tried to settle Capua as a colony! How you left there, or rather almost did not leave, we know.

\ciceroSection{101} The city you threaten. Make the attempt, that at last that ``almost'' may be removed! But how distinguished is that journey of yours! What shall I say of the preparation of luncheons, of your frenzied wine-bibbing? These are your losses; those losses are ours: the Campanian land, which, when it was being taken from the revenues to be given to soldiers, we still thought a great wound was being inflicted on the commonwealth --- this you were dividing up among your fellow-revellers and fellow-gamblers. Male mimes I mean and female mimes, senators, settled on the Campanian land. Why do I now complain about the Leontine field? Since once both these ploughlands, the Campanian and the Leontine, were reckoned among the patrimony of the Roman people as fruitful and productive. To his doctor three thousand acres: what, if he had cured you? To a rhetorician two thousand: what, if he could have made you eloquent?

\ciceroSection{102} But let us go back to the journey and to Italy. You led a colony to Casilinum, where Caesar had earlier led one. You consulted me by letter about Capua, you yourself indeed, though I should have given the same reply about Casilinum: whether you could rightly lead a new colony where a colony already existed. I replied that into a colony which had been led with auspices, while it was still intact, no new colony could rightly be led; but that new colonists could be enrolled, I wrote. But you, lifted up by your insolence, overturning all the auspicial law, led a colony to Casilinum, where one had been led a few years before, that you might raise the standard, that you might drive round the plough; with whose share you almost grazed the gate of Capua, that the territory of a flourishing colony might be diminished.

\ciceroSection{103} From this disturbance of religious things you fly to the estate of Marcus Varro, that most chaste and most upright man, at Casinum. By what right, with what face? ``By the same,'' you will say, ``as that by which I came to the estates of the heirs of Lucius Rubrius, of the heirs of Lucius Turselius, to my other countless possessions.'' And if it is from the auction-spear, let the spear stand, let the records stand, only Caesar's, not yours --- by which you owed, not by which you discharged yourself. Of Varro's estate at Casinum, who says it was sold, who saw the spear of that sale, who heard the auctioneer's voice? You say you sent a man to Alexandria to buy it from Caesar; for to wait for the man himself was too much. But who ever heard --- and no man's safety was a care to more --- that anything was withdrawn from the fortunes of Varro? What? If Caesar even wrote to you to return it, what can adequately be said of so great a shamelessness? Take away those swords for a little while which we see: you will at once understand that one thing is the cause of Caesar's spear, another the cause of your own boldness and rashness. For not only the owner, but any friend, neighbour, host, agent will keep you out of those quarters. But how many days in that country house have you been most foully reveling! From the third hour the drinking, the playing, the vomiting went on. O the wretched walls themselves --- ``by how unlike a master ---'' although how is this fellow a master? --- yet by how unlike were they held! Marcus Varro wished it as the retreat of his studies, not of his lusts.

\ciceroSection{104} What things in that villa used formerly to be discussed, what considered, what committed to letters! The rights of the Roman people, the records of our ancestors, every method of wisdom and every method of learning. But with you as occupant --- not as master --- all the place echoed with the voices of drunkards, the pavements swam with wine, the walls were soaked, freeborn boys mingled with the boys for hire, harlots among matrons. Men came to pay their respects from Casinum, from Aquinum, from Interamna: not one was admitted. Quite rightly: for in a man so utterly base the marks of distinction were being made cheap.

\ciceroSection{105} When, setting out from there for Rome, he was approaching Aquinum, there came out to meet him, as is usual in a populous country town, a sizeable crowd. But this fellow was carried through the town in a closed litter as if he were dead. Foolish people of Aquinum: but at any rate they lived on the road. What of the people of Anagnia? They, though off the road, came down to greet him as if he were consul. It is incredible to say, but among all it was settled that no one was greeted in return, especially when he had two men of Anagnia with him, Mustela and Laco, of whom the first is chief of his swords, the other of his cups.

\ciceroSection{106} Why should I recall those threats and insults with which he inveighed against the Sidicines, harassed the people of Puteoli, because they had adopted Gaius Cassius and the Bruti as patrons? With great enthusiasm, judgment, goodwill, affection, not, like you and Basilus, by force and arms, nor like others of your sort, whose clients no one would wish to be, much less to be their client. Meanwhile, while you are absent, what a day was that of your colleague's, when he overthrew that tomb in the Forum which you were accustomed to revere? When the news was brought to you, as those who were there with you agreed, you collapsed. What happened next I do not know --- fear, I believe, prevailed, and arms --- you indeed dragged your colleague down from heaven and brought it about not, indeed, that he should be like you, but at least that he should be unlike himself.

\ciceroSection{107} But what a return after this to Rome, what a disturbance throughout the whole city! We remembered Cinna at the height of his power, Sulla afterwards in his domination, and lately Caesar in his reign. Perhaps there had been swords then, but concealed, and not so many. But your barbarism --- what is it, how vast? They follow in a hollow square with their swords; we see litters borne by shield-bearers. And to these things now, senators, become familiar, we have grown calloused by custom. On the Kalends of June, when, as had been determined, we wished to come into the Senate, struck with terror we suddenly scattered.

\ciceroSection{108} But this fellow, who did not need a Senate, neither missed anyone and was rather glad at our withdrawal, and at once accomplished those marvellous deeds of his. He who had defended Caesar's handwritten orders for his own profit, overturned Caesar's laws --- and those most distinguished --- that he might shake the commonwealth. He prolonged the years of provincial governments; and the same man, who ought to have been the defender of Caesar's acts, both in public and private matters rescinded Caesar's acts. In public matters nothing is graver than law; in private matters most binding is a testament. Some laws he abolished without publication, others he published in order to abolish them. He made void a will, a thing which has always been maintained even for the meanest citizens. The statues, the paintings which Caesar bequeathed to the people together with his gardens, he took, some to Pompey's gardens, some to Scipio's villa.

\ciceroSection{109} And you, scrupulous in the memory of Caesar, you love him dead? What greater honour had he attained than to have a sacred couch, an image, a pediment, a flamen? Is Marcus Antonius, then, flamen, as to Jupiter, as to Mars, as to Quirinus, so to the divine Julius? Why then do you delay? Why are you not consecrated? Choose the day, see who is to consecrate you: we are colleagues; no one will refuse. O detestable man, whether because you are priest of a tyrant or because of a dead man! I ask next: do you not know what day it is today? Do you not know that yesterday was the fourth day in the circus of the Roman Games? And that you yourself proposed to the people that a fifth day besides should be assigned to Caesar? Why are we not in the toga praetexta? Why do we suffer the honour granted to Caesar by your own law to be neglected? Or did you let the thanksgivings be polluted by the addition of a day, and refuse the same to the sacred couches? Either remove religion from everything, or preserve it everywhere.

\ciceroSection{110} You ask whether the sacred couch pleases me, the pediment, the flamen. None of these things pleases me. But you who defend Caesar's acts, what can you say to show why you defend some and care nothing for others? Or do you perhaps wish to confess that you measure everything by your own profit, not by his dignity? What then to this? For I look for your eloquence. I have known your grandfather a most eloquent man, but you a still plainer one in speaking. He never spoke in public unclothed: we have seen the breast of you, a simple man. Will you answer to these things, or will you dare to open your mouth at all? Will you find anything in all my long speech which you confidently think you can answer? But let us pass over the past.

\ciceroSection{111} This one day, this one, I say, today, this point of time at which I am speaking, defend, if you can. Why is the Senate hedged in by a ring of armed men, why do your bodyguards hear me with their swords, why are the doors of the temple of Concord not open, why do you bring down into the Forum men of all nations the most barbarous, Ituraeans with arrows? He says he does it for his own protection. Is it not better to perish a thousand times than to be unable to live in one's own city without an armed guard? But this is no guard, believe me: you must be hedged about by the affection and goodwill of the citizens, not by arms.

\ciceroSection{112} The Roman people will tear these arms from you and wrest them away --- I would they may, while we are safe! But however you deal with us, while you use these counsels, you cannot, believe me, last long. For that wife of yours, who is by no means greedy --- whom I describe without insult --- has owed too long to the Roman people her third instalment. The Roman people has men to whom it may entrust the helm of the commonwealth: wherever in the lands they are, there is all the bulwark of the commonwealth --- or rather the commonwealth itself, which thus far has only avenged itself, has not yet recovered itself. The commonwealth indeed surely has young men of the noblest birth ready as its defenders. As much as they wish, let them yield to leisure, considering peace; still they shall be recalled by the commonwealth. And the name of peace is sweet and the thing itself wholesome; but between peace and slavery there is the greatest difference. Peace is tranquil liberty, slavery the last of all evils, to be repelled not only by war but even by death.

\ciceroSection{113} But if those liberators of ours have taken themselves from our sight, yet they have left the example of the deed. They did what no one had done. Brutus pursued Tarquin in war, who was then king when one was allowed to be at Rome; Spurius Cassius, Spurius Maelius, Marcus Manlius were slain on suspicion of seeking the kingship: these first attacked with swords, not one seeking the kingship but one reigning. Which deed, since in itself it is glorious and divine, lies also exposed for imitation, especially since they have won that glory which scarcely seems to be capable of being contained by heaven. For though there was sufficient reward in the very consciousness of a most beautiful deed, yet to a mortal man immortality, I think, is not to be despised.

\ciceroSection{114} Recall, then, Marcus Antonius, that day on which you abolished the dictatorship; set before your eyes the joy of the Senate and the Roman people; compare it with this huckstering of yours and of your people: then you will understand how great is the difference between praise and profit. But just as some men through some disease and dullness of the senses do not taste the sweetness of food, so the lustful, the greedy, the criminal have no taste of true praise. But if praise cannot allure you to right action, can not even fear call you back from the foulest deeds? You do not fear the courts: if from innocence, I commend you; but if from violence, do you not understand what is to be feared by him who fears not the courts in that way?

\ciceroSection{115} But if you do not fear brave men and distinguished citizens, because they are kept from your body by arms, your own people, believe me, will not bear with you long. And what life is it, day and night to fear those of your household? Unless, indeed, you have men bound by greater benefits than that man had certain of those by whom he was killed, or unless you are at all comparable with him. There was in him genius, reason, memory, letters, care, thought, diligence; he had performed deeds in war --- disastrous, I admit, to the commonwealth, but yet great; for many years he had meditated reigning, and with great toil and great dangers had brought about what he had thought; he had soothed the unskilled multitude with gifts, monuments, donatives, banquets; his own he had bound by rewards, his adversaries by the show of clemency. In short --- he had already brought to a free state, partly by fear, partly by patience, the habit of slavery.

\ciceroSection{116} With him for the lust of mastery I can compare you, but in other respects you are in no way to be likened to him. But of the many evils which were burnt into the commonwealth by him, this much of good there is, that the Roman people has now learnt how much trust to put in each, to whom it should commit itself, of whom it should beware. These things you do not consider, nor do you understand that it is enough for brave men to have learnt how beautiful in the thing, how grateful in the benefit, how glorious in fame, it is to kill a tyrant? Or, when men did not bear with him, will they bear with you?

\ciceroSection{117} Hereafter, believe me, men will hasten to this work in rivalry, nor will the slowness of opportunity be waited for. Look back, I beseech, at last on the commonwealth, Marcus Antonius; consider those from whom you sprang, not those with whom you live: with me, as you will: be reconciled with the commonwealth. But concerning yourself, you yourself will judge; concerning myself, I will myself declare. I defended the commonwealth as a young man, I shall not desert it as an old man: I despised the swords of Catiline, I shall not be afraid of yours. Nay, I would willingly even offer my body, if the liberty of the citizens can be redeemed at once by my death, so that the indignation of the Roman people may at last bring forth what it has long been in labour with!

\ciceroSection{118} For if nearly twenty years ago in this very temple I declared that death could not come untimely to a consular, how much more truly shall I now declare it cannot come so to an old man! For me indeed, senators, death is now even to be wished for, my work in those things which I obtained and which I performed being done. Only these two things I pray: one, that dying I may leave the Roman people free --- nothing greater than this can be granted me by the immortal gods --- the other, that to each man it may turn out as he deserves of the commonwealth.
