% Pro Rege Deiotaro (pro-rege-deiotaro) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Though in every more serious case, Gaius Caesar, I am wont at the outset of speaking to be more deeply stirred than either my practice or my years would seem to demand, in this case so many things unsettle me that, while my good faith brings to me as much zeal in the defending of the safety of King Deiotarus, fear takes from me as much in the way of ability. To begin with, I am pleading for the life and fortunes of a king --- a thing which, though not unfair when it is your own life that has been at risk, is yet so unprecedented, a king to stand on trial for his life, that before this hour it has never been heard of.

\ciceroSection{2} Next, the king whom in former days I was wont to honour, with the whole Senate, for his unbroken services to our commonwealth --- this same king I am now compelled to defend against a most atrocious charge. There is added the fact that I am thrown into confusion by the cruelty of the one accuser and the unfitness of the other. Cruel Castor --- not to call him criminal and impious --- a grandson who has brought his grandfather into peril of his life and has fastened the terror of his own youth upon the man whose old age he ought to have guarded and shielded, and has drawn the credential of his coming-of-age from impiety and from crime; who has bought off his grandfather's slave with bribes to accuse his master, and has spirited him away from the feet of the king's envoys.

\ciceroSection{3} As for the runaway --- when I saw the face of a slave accusing his master, and an absent master, and a master most devoted to our commonwealth, when I heard his words, I was not so much grieved at the stricken condition of the throne as I was alarmed for the common fortunes of us all. For though by the custom of our ancestors a slave may not be examined against his master even under torture --- the examination in which pain can wring out the true word even from one unwilling --- there has now sprung up a slave who, the very master whom he could not have called upon while on the rack, that master he accuses without his bonds.

\ciceroSection{4} There is one other thing, Gaius Caesar, that troubles me at times --- though, when I look closely at you, I cease to fear it. For it is in itself unfair, yet by your wisdom it is made altogether fair. I mean that to speak before the very man against whose life one is charged with having taken counsel of crime is, when you consider it in itself, a hard thing; for hardly any man, sitting as judge in his own peril, will show himself fairer to the defendant than to himself. But your surpassing and singular nature, Caesar, diminishes that fear for me. For I am not so afraid of what you will judge concerning King Deiotarus as I am aware of what you wish the rest of the world to judge concerning you.

\ciceroSection{5} I am moved also by the strangeness of the place itself, that I should be pleading a case as great as none ever brought into the lists of debate within the walls of a private house, that I should plead outside any public gathering and outside that thronged audience on which the zeal of orators is wont to rest its strength. In your eyes, in your face and looks, I find my repose; you alone I look upon, to you alone my whole speech is turned --- circumstances which are most weighty for my hope of obtaining the truth, but slighter for the rousing of the spirit and for the whole onrush and strain of speaking.

\ciceroSection{6} For if I were pleading this case, Gaius Caesar, in the Forum, with you as both hearer and judge, how great a vigour the throng of the Roman people would bring me! For what citizen would not be favourable to the king whose every year, he would remember, has been spent in the wars of the Roman people? I should look upon the Senate-house, I should look out upon the Forum, I should at last call heaven itself to witness; thus, as I recalled the benefits both of the immortal gods and of the Roman people and of the Senate towards King Deiotarus, words could in no way fail me.

\ciceroSection{7} But since these walls make the scope narrower and the conduct of so great a case is enfeebled by the setting, it is for you, Caesar --- who have so often pleaded for others --- to refer my present state of mind to your own self, that both your fairness and your attentiveness in hearing may the more easily lessen this disquiet of mine. But before I speak about the accusation itself, let me say a few words about the accusers' hope; who, though they seem to be strong neither in talent nor in practical experience, have nevertheless come to this case not without some hope and design.

\ciceroSection{8} They were not unaware that you had been angry with King Deiotarus; they remembered that he had been visited with certain inconveniences and losses on account of the offence given to your spirit, and they had come to know that you were both angry with him and friendly to themselves; and they thought that, since they were speaking before your very self about your own peril, a trumped-up charge would easily settle in a mind that had been chafed. From this fear, then, Caesar, set us free first of all, by your good faith and your steadiness and your clemency, so that we may not suspect any portion of your anger to be still seated in you. By that right hand of yours, I entreat you --- which as guest you stretched out to the guest King Deiotarus --- by that right hand, I say, more firm not so much in wars or in battles as in pledges and in good faith. You wished to enter his house, you wished to renew an old guest-friendship; you the household gods received, you as a friend and at peace the altars and hearths of King Deiotarus beheld.

\ciceroSection{9} You are easily entreated, Caesar, and once entreated, you are won over. No enemy ever appeased you who felt that any leavings of bitterness had stayed behind in you. Yet who has not heard of your complaints against Deiotarus? You never accused him as an enemy, but as a friend who had discharged his duty too little, in that he had leaned towards Gnaeus Pompey's friendship more than towards yours. For even this fault, you used to say, you would have pardoned him, if at that time he had sent auxiliaries to Pompey, or even if he had sent his son and pleaded his own age in excuse.

\ciceroSection{10} Thus, when you were freeing him from the gravest charges, you were leaving him a very small fault. And so you not only inflicted no penalty on him: you freed him of every fear, you acknowledged him as your guest, you left him a king. For he was not carried forward by hatred of you, but slipped through the common error. That king whom the Senate had often addressed by that very title in decrees most full of honour, and who from his earliest youth had counted that order most weighty and most sacred --- this man, a distant foreigner of another race, was thrown into the same confusion as ourselves, born and ever turning in the heart of the commonwealth.

\ciceroSection{11} When he heard that, by the authority of a Senate of one mind, arms had been taken up, that consuls, praetors, tribunes of the plebs, and we who were imperators had been given the commonwealth to defend, he was stirred in spirit, and being a man most friendly to this empire, he was in dread for the safety of the Roman people, in which he saw his own to be also bound up. Yet in this utmost fear he judged that he ought to keep quiet. He was most disturbed of all, however, when he heard that the consuls had fled out of Italy, that all the men of consular rank had fled --- so it was reported to him --- that the whole Senate, all of Italy, had streamed out. For to such messages and rumours the road eastward lay open, and no true report came after them. He heard nothing about your terms of settlement, nothing about your zeal for concord and peace, nothing about the conspiracy of certain men against your dignity. Such being the case, nevertheless he held himself in check up to the moment when envoys and letters from Gnaeus Pompey came to him.

\ciceroSection{12} Forgive him, forgive him, Caesar, if King Deiotarus yielded to the authority of that man whom we all followed; on whom, when gods and men had heaped every distinction, you yourself most of all had heaped distinctions both numerous and great. For if your own deeds have brought obscurity upon the praises of all the rest, we have not on that account forgotten the memory of Gnaeus Pompey. How great his name was, how great his resources, how great his glory in every kind of war, how great the honours paid him by the Roman people, how great by the Senate, how great by yourself --- who does not know? He had so far surpassed those before him in glory as you have surpassed all men. And so the wars of Gnaeus Pompey, his victories, triumphs, consulships, we used to number with wonder: yours we cannot reckon up.

\ciceroSection{13} To him, then, King Deiotarus came, in this wretched and fated war --- a man he had previously aided in just and lawful wars, a man with whom he was joined not only by guest-friendship but even by intimate familiarity; and he came either invited as a friend, or summoned as an ally, or called out as one who had learnt to obey the Senate: in short, he came as to one fleeing, not as to one pursuing, that is, into the partnership of peril, not of victory. And so, once the battle of Pharsalus had been fought, he withdrew from Pompey; he was unwilling to follow an unbounded hope; he held that satisfaction had been made --- for duty, if he owed anything, for error, if anything had escaped his knowledge --- and he betook himself home, and, while you were waging the Alexandrian war, served your interests.

\ciceroSection{14} It was he who sustained the army of Gnaeus Domitius, that most distinguished man, with his own roofs and supplies; it was he who sent money to Ephesus, to the man whom you had chosen out of your own people as the most loyal and the most approved by all; it was he who, holding auctions a second and a third time, gave money for you to use in the war; it was he who exposed his own body to peril, and stood with you in the battle-line against Pharnaces, and held your enemy to be his own. These things, indeed, were so received by you, Caesar, that you adorned him with the most ample title and name of king.

\ciceroSection{15} This man, then, not only freed of peril by you but also adorned with the most ample honour, is now charged with having wished to murder you in his own house --- which, unless you judge him utterly mad, you surely cannot even suspect. For to pass over how great a crime it would be to slay a guest in the very sight of the household gods, how great a savagery to extinguish the brightest light of all nations and all memory, how great a ferocity not to dread the conqueror of the world, how great an inhuman and ungrateful spirit to be found a tyrant by the man who had given him the name of king --- to pass all this over, how great a madness it would have been to rouse against himself alone every king (and many were his neighbours), every free people, every ally, every province, in short, all the arms of all the world? In what way could he have been torn apart from his kingdom, his house, his wife, his most beloved son --- a crime of such magnitude not only carried out but even conceived?

\ciceroSection{16} But, I suppose, this man, being rash and headlong, did not see these things. Who more thoughtful than he, who more guarded, who more prudent? Yet in this place I think that Deiotarus should be defended not so much by his talent and good sense as by the loyalty and the religion of his whole life. The man's uprightness is known to you, Gaius Caesar; known his character, known his constancy. To anyone, again, who has merely heard the name of the Roman people, has not Deiotarus's integrity, gravity, virtue, good faith been heard of? A deed, then, that would not fall to a man without sense by reason of fear of present ruin, nor to a criminal man unless he were also wholly out of his mind --- this you fabricate as the design of the best of men and of a man by no means a fool?

\ciceroSection{17} And how, not only not credibly, but not even plausibly! ``When,'' he says, ``you came to Blucium and turned aside into the house of your royal host, there was a certain room in which had been laid out the things with which the king had decided to present you. To this place, after the bath, before you reclined at table, he wished to lead you. For armed men had been placed in that very room to murder you.'' Behold the charge, behold the cause why a runaway accuses a king, a slave a master! For my part, by Hercules, Caesar, at the outset, when the case was first laid before me in this form, that Phidippus the physician, the king's slave, who had been sent along with the envoys, had been bribed by that young man, I was struck by this suspicion: ``He has suborned the physician as informer; clearly he will fabricate some charge of poison.'' Though far from the truth, the thing was not far removed from the usual style of accusations.

\ciceroSection{18} What does the physician say? Nothing about poison. Yet that could have been done, first, more secretly --- in a drink, in food; and secondly, more safely, because when done it can be denied. Had he killed you openly, he would have turned against himself not only the hatreds but the very arms of all nations: had he done it by poison, he could never have hidden it from the divine will of Jove the protector of hospitality, though men perhaps he could have hidden it from. What therefore he could have attempted with more secrecy and accomplished with more caution, this he refused to entrust to you (sic!) --- to a cunning physician and to a slave he reckoned faithful: but about arms, about steel, about ambushes, he was unwilling to keep you in the dark?

\ciceroSection{19} And with what wit the charge is patched together! ``Your usual luck,'' he says, ``which has often saved you, saved you then too: you said you did not wish to inspect them.'' What followed? Did Deiotarus, since the deed was not at the moment accomplished, at once disband his army? Was there no other place for an ambush? ``But you had said you were coming back to that same place after dinner, and so you did.'' Was it then so great a matter to keep armed men for an hour or two in the same place where they had been stationed? When you had been at the dinner in friendly and pleasant fashion, you went over there, as you had said: in which place you found Deiotarus to be towards you such as King Attalus was towards Publius Africanus, to whom, as we read, he sent the most magnificent gifts from Asia, all the way to Numantia --- gifts which Africanus accepted in the sight of the army. When Deiotarus, then present, had done the like in a kingly spirit and after the kingly fashion, you withdrew into the bedchamber.

\ciceroSection{20} I beseech you, Caesar, call back the memory of that hour, set before your eyes that day, recollect the looks of the men who gazed on you in admiration. Was there any trembling, any disturbance, anything that did not bear the stamp of a most weighty and most blameless man's discipline? Modest, quiet, in keeping with the bearing of such a man? What reason, then, can be devised why he should have wished to kill you washed, but not when you had dined?

\ciceroSection{21} ``He put it off,'' he says, ``to the following day, so that, when you had come to the fortress, he might there carry out what he had planned.'' I do not see any reason for changing the place, but still the matter is darkly handled. ``When,'' he says, ``you had said that you wished to vomit after dinner, they began to lead you to the bath: for there the ambush was. Yet your same luck saved you: you said you preferred the bedchamber.'' May the gods destroy you, runaway! So you are not only worthless and dishonest, but also a witless idiot. What? Had he set up bronze statues in the bath, which could not be moved from the bath into the bedchamber? You have the charges of the ambush: for he said nothing more. ``Of these things,'' he says, ``I was an accomplice.'' What then? Was Deiotarus so out of his wits that the very man he held as accomplice of so great a crime he would send away from himself, would even send to Rome, where he knew that both his bitterest enemy his grandson, and Gaius Caesar against whom he had laid an ambush, were to be found --- especially when this was the one man who could give evidence about him in his absence?

\ciceroSection{22} ``And my brothers,'' he says, ``because they were accomplices, he cast in chains.'' When, then, he was binding those whom he had with him, he was sending you, untied, to Rome --- you who knew the same things which you say they knew? The rest of the accusation was twofold: one part, that the king had always been on the watch, alienated from you in spirit; the other, that he had got together a great army against you. About the army I shall speak briefly, as about the rest. Never did King Deiotarus have such forces as he could wage war upon the Roman people with --- only such as he could guard his own borders from raids and brigands and could send as auxiliaries to our commanders. And in former days he could indeed maintain greater forces; now he can scarcely keep up the slender ones he has.

\ciceroSection{23} ``But he sent to a certain Caecilius'': only those whom he sent, because they refused to go, he cast in chains. I do not ask how likely it is either that the king had no others to send, or that those who had been sent disobeyed, or that those who in so great a matter were not obedient should rather have been chained than killed. But all the same, when he was sending to Caecilius, did he not know that that cause had been beaten --- or did he think this Caecilius of yours a great man? A man whom assuredly the one who knows our people best would either despise because he did not know him, or, if he did know him, despise him for that very reason.

\ciceroSection{24} He adds also this: that the cavalry he sent were not of the best. I believe, Caesar, nothing to compare with your own cavalry: but he sent from those he had, picked men. He says that someone or other from that number was judged a slave. I do not so believe; I have not heard so. But even if it had been the case, I should reckon that no fault of the king's. But ``estranged in spirit from you'' --- in what way? He hoped, I suppose, that your Alexandrian campaign would have a difficult outcome on account of the nature of the country and of the river. Yet at that very time he gave money, he supported the army, he failed in nothing towards the man you had set in charge of Asia; he was at hand for you the victor not only for hospitality but for danger and for the battle-line.

\ciceroSection{25} Then followed the African war. Grave rumours about you ran abroad, which even stirred up that crazy Caecilius. With what spirit was the king then --- he who held auctions and chose rather to strip himself than not to supply you with money? ``But at that very time,'' he says, ``he was sending men to Nicaea and Ephesus to catch the rumours from Africa and report them to him quickly.'' And so when it had been reported to him that Domitius had perished in a shipwreck, that you were besieged in a fort, he said about Domitius a Greek verse to the same effect as the Latin one we ourselves have: ``Let friends perish, provided enemies fall along with them.'' But this verse he, if he had been your bitterest enemy, would still never have uttered: for he himself is mild, the verse savage. And how could he have been a friend to Domitius, while an enemy to you? An enemy, again, to you --- and why? When he remembered that he, who by the law of war could even have been put to death, had by you been established as king together with his son.

\ciceroSection{26} What next? Where does the rascal go on to? He says that Deiotarus, carried away by this joy, drowned himself in wine and danced naked at a banquet. What cross can bring punishment enough on this runaway? Did anyone ever see Deiotarus dancing, or drunken? All the virtues are in that king --- which I do not think escapes you, Caesar --- but most of all a singular and admirable frugality: though I know that kings are not usually praised by this word. To be called a thrifty man does not have much of praise in it for a king: brave, just, severe, weighty, of great spirit, generous, beneficent, liberal --- these are the kingly praises; the other is for private life. Let each take it as he will: I, for my part, judge frugality, that is to say modesty and temperance, to be the greatest virtue. This he has had from his earliest years, marked and known by all Asia, by our magistrates and legates, and especially by the Roman knights who have done business in Asia.

\ciceroSection{27} By many steps of services towards our commonwealth he climbed to this royal title; yet whatever time was free from the wars of the Roman people, he gave to acquaintances, friendships, business dealings and accounts with our countrymen, so that he was held not only a noble tetrarch but also a most excellent father of a family, and a most attentive farmer and cattle-breeder. He, then, who as a young man not yet endowed with so great glory did nothing except in the strictest and gravest fashion --- did he, with such a reputation and at such an age, take to dancing?

\ciceroSection{28} You ought, Castor, rather to imitate the character and the discipline of your grandfather, than to slander a man most excellent and most distinguished through a runaway's mouth. Even if you had had a dancer for grandfather, and not a man from whom examples of modesty and chastity might be sought, still this abuse would suit his age not at all. The studies which he had imbibed from his earliest years --- not how to dance, but how to handle weapons well and horses superbly --- had now, with the whole of his age, failed him. And so, when many had lifted Deiotarus into the saddle, we used to wonder that as an old man he could keep his seat; but this young man --- who was my soldier in Cilicia, my fellow-soldier in Greece --- when in our army he was riding with the picked cavalry which his father had sent along with him to Pompey, what charges he used to make! how he displayed himself, how he showed himself off, how he yielded to no one in that cause in zeal and eagerness!

\ciceroSection{29} Then, when the army was lost, I, who had always been an advocate of peace, and after the battle of Pharsalus a counsellor not for laying down arms but for casting them away, could not bring this man over to my advice, both because he himself was burning with zeal for that war, and because he thought he owed it to his father to do so. Lucky is your house, which has obtained not only impunity but even licence to accuse: wretched is Deiotarus, who is accused both by a man who served in the same camp, and not only before you but by his own household. Are you, Castor, not able with your prosperous fortune to be content without the ruin of your kin?

\ciceroSection{30} Let there be enmities, by all means --- though there ought not to have been: for it was King Deiotarus who called your family out of the darkness, cast down and obscure, into the light. Who heard of your father before he heard whose son-in-law he was? But however ungratefully and impiously you spurn the name of kinship, still you might have carried on the quarrel in human fashion: not pursue him with a fabricated charge, not strive after his life, not arraign him on a capital count. Be it so: let even this magnitude of bitterness and hatred be granted: should every right of life and of common safety, indeed of common humanity itself, be trampled upon? To solicit a slave by words, to corrupt him with hope and promises, to lead him away from his master's home, to arm him against his master --- this is to declare not against one kinsman but against every household a wicked war. For if this corrupting of a slave goes not only unpunished but is even ratified by so great an authority, no walls will guard our safety, no laws, no rights. For when what is within the house and is our own can fly out with impunity and fight against us, then in the master's place is servitude, and in the slave's place mastery.

\ciceroSection{31} O the times, O the manners! That Gnaeus Domitius whom we saw as boys --- consul, censor, pontifex maximus --- when as tribune of the plebs he had haled Marcus Scaurus, the chief man of the state, before the people's court, and Scaurus's slave had come secretly to his house and said that he was going to inform against his master, ordered the man to be seized and brought to Scaurus. See what a difference --- though unfairly do I compare Castor with Domitius: yet that man sent back his enemy's slave; you spirited him away from his grandfather. That man refused to hear him unbribed; you have bribed him. That man rejected the slave as helper against his master; you have even taken him on as accuser. But, you will say, this slave was bribed by you only once.

\ciceroSection{32} When he had been brought to you and had been with you, did he not run back to the envoys? Did he not come to this Gnaeus Domitius? Did he not, in the hearing of Servius Sulpicius, that most distinguished man, who happened then to be dining at Domitius's, and of this Titus Torquatus, that excellent young man, confess that he had been bribed by you, that he had been driven into deceit by your promises? What is this so ungovernable, so cruel, so unbounded inhumanity? Did you come to this city for the purpose that you might corrupt this city's rights and precedents, and stain our state's humanity with your domestic savagery?

\ciceroSection{33} And how shrewdly the charges are gathered together! ``Blesamius'' --- for it was under his name, an excellent man and not unknown to you, that he was abusing you --- ``Blesamius,'' he says, ``used to write to the king that you were in disfavour, that you were thought a tyrant, that, by your statue's being set among those of the kings, men's minds were violently offended, that there was no applause for you.'' Do you not see, Caesar, that these things have been gathered up by them out of the little spiteful gossip of the town? Was Blesamius likely to write that Caesar was a tyrant? For he would have seen the heads of many citizens, many men, by Caesar's order, harried, scourged, put to death; many houses ruined and overturned; the Forum filled with armed soldiers! What we have always seen in civil victory, in your victory we have not seen.

\ciceroSection{34} You alone, I say, you, Gaius Caesar, are he in whose victory no one has fallen except in arms. And the man whom we, born free in the highest freedom of the Roman people, hold to be not only no tyrant but the most clement of victors --- can he seem to Blesamius, who lives under a monarchy, a tyrant? As for the statue, who complains --- and one statue at that, when he sees so many? For surely it is much to be envied that he should have statues, whose trophies we do not envy. For if the place gives the envy, there is no place at least, for a statue, more famous than the Rostra. About the applause, what shall I answer? An applause which has never been wanted by you, and which has sometimes been silenced by the very wonder of men struck dumb in admiration, and perhaps passed over because nothing common can be thought worthy of you.

\ciceroSection{35} I think nothing has been omitted by me, but something has been reserved to the last part of the case. And what is this something? That my speech may quite reconcile you to Deiotarus. For I no longer fear that you may be angry with him; this I fear, that you may suspect him to be at all angry with you: which is very far off, believe me, Caesar. For what he remembers is what he keeps through you, not what he has lost: nor does he reckon that he has been punished by you, but, since you judged that much should be granted to many, he did not refuse that you should take it from him, who had been on the other side, that you might give it.

\ciceroSection{36} For indeed if Antiochus the Great, that king of Asia, when --- after he had been overcome by Lucius Scipio and ordered to reign only as far as Taurus --- he had lost all this Asia which is now our province, was wont to say that the Roman people had dealt kindly with him, in that, freed of an overlarge charge, he might use the modest bounds of his kingdom: how much more easily can Deiotarus comfort himself in the same way. For Antiochus had paid the penalty of his madness; this man has paid the penalty of his error. You have granted everything to Deiotarus, Caesar, in that you have left the royal title both to him and to his son. With this title kept and preserved, he holds that no benefit from the Roman people, no judgement of the Senate concerning him, has been diminished. He is of great and uplifted spirit; he will never sink under enemies, not even under fortune.

\ciceroSection{37} He thinks that he has brought forth much by his deeds in the past, and that he holds much in spirit and virtue, which by no means can he lose. For what fortune, what chance, what so great a wrong, can blot out the decrees passed about Deiotarus by all our generals? He has been honoured by all who, after he was of age to be in the camp, waged wars in Asia, Cappadocia, Pontus, Cilicia, Syria: as for the Senate's judgements about him, so many, so full of honour, recorded in the public letters and monuments of the Roman people, what age shall ever bury them, what oblivion so great wipe them out? What shall I say of his virtue, of his greatness of spirit, his weight, his constancy? Qualities which all the learned and wise have called the chief goods, and some indeed the only good, and have said that virtue is sufficient not only for living well, but for living happily.

\ciceroSection{38} Reflecting on these things and pondering them day and night, he not only is not angry with you --- for he would be not only ungrateful but mad --- but he refers all the tranquillity and quiet of his old age to your clemency as a gift received. In which spirit, while he stood before, so I do not doubt that by your letters, of which I have read the copy, which you gave to this Blesamius for him at Tarraco, he has further raised himself up and shaken himself loose from every anxiety. For you bid him hope well and be of good cheer --- which I know that you do not write to no purpose. For I remember that in almost the same words you wrote to me, and that I was bidden by your letters to hope well, and not to no purpose.

\ciceroSection{39} I am indeed at pains for the cause of King Deiotarus, with whom the commonwealth has knit me in friendship, whom the will of each of us has joined to me in guest-friendship, whom intercourse has bound to me in intimacy, but with whom his great services to me and to my army have created the closest of ties: but while I am at pains for him, I am also at pains for many most distinguished men whom you ought once for all to have pardoned, that your benefit may not be called into doubt, that an everlasting anxiety may not settle in men's minds, that it may not come about that any of those who have once been freed by you from fear should begin to fear you.

\ciceroSection{40} I ought not, Caesar, to do what is wont to be done in such great perils: to try by what means I may stir your pity by speaking. There is no need. It is wont of itself to run to meet the suppliant and the unfortunate, called forth by no man's speech. Set before yourself two kings, and gaze upon in your mind what you cannot in the eye: you will surely give to pity what you have denied to anger. There are many monuments of your clemency, but most of all the unimpaired safety of those to whom you have given safety. Which, if in private persons they are matters of glory, much more shall they be recounted in the case of kings. Always in this state has the royal name been sacred, but most sacred of all that of kings who are our allies and friends.

\ciceroSection{41} That name these kings feared, with you as victor, that they should not lose; and now that it has been kept and by you confirmed, they trust that they will hand it on even to their descendants. Their bodies, for the safety of their kings, these royal envoys offer up --- Hieras, and Blesamius, and Antigonus, long since known to you and to all of us, and Dorylaus, gifted with the same good faith and virtue, who lately was sent as envoy to you along with Hieras: men most beloved of their kings, and to you also, as I hope, approved.

\ciceroSection{42} Inquire of Blesamius whether he has written anything to the king against your dignity. Hieras for his part takes the whole case upon himself, and against those charges sets himself up as defendant in the king's stead. He calls upon your memory, in which you are most powerful; he says that he never withdrew a foot from you in Deiotarus's tetrarchy; that he was at hand for you on the first borders, escorted you to the last; that when you had come out of the bath, he was with you; when you had inspected those gifts after dinner; when you had lain down in the bedchamber; and that he showed you the same constant attendance on the next day.

\ciceroSection{43} Wherefore, if any of these things charged against him was conceived, he does not refuse that you should judge the deed his. For which reason, Gaius Caesar, I would have you reflect that your sentence on this day will bring either, with utmost dishonour, the most wretched ruin upon kings, or, with safety, unimpaired renown: of which the one is the wish of cruelty, the other --- to preserve them --- is for your clemency.
