% Pro M. Marcello (pro-marcello) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} To the long silence, senators, which I have observed in these last times --- not from any fear, but partly from grief, partly from a sense of shame --- this day has brought an end; and the same day has been the beginning of my speaking, in the old manner, what I would and what I felt. For so great a gentleness, so unaccustomed and unheard-of a clemency, so great a moderation at the height of supreme power over everything, and finally so incredible and almost divine a wisdom --- these I can in no way pass over in silence.

\ciceroSection{2} For now that M. Marcellus has been given back to you, senators, and to the commonwealth, I judge that not only his voice and authority but my own has been preserved and restored to you and to the commonwealth. For it grieved me, senators, and I was deeply tormented, when I saw such a man --- one who had been on the same side as I --- not in the same fortune as I; nor could I persuade myself, nor did I think it lawful, to move along our old course while he, the rival and imitator of my studies and labours, was as it were torn away from me as a comrade and companion. Therefore, Gaius Caesar, you have both opened up to me the practice of my former life, which had been closed off, and have raised for all these men, as it were, a kind of signal that they may hope well for the commonwealth.

\ciceroSection{3} For it has been understood, certainly by me in many things and above all in my own case, but a little while ago by everyone, when you granted M. Marcellus to the Senate and to the commonwealth --- and that with his offences against you expressly recalled --- that you set the authority of this order and the dignity of the commonwealth above your own grievances and suspicions. He indeed has reaped on this day the greatest reward of all his life before, both in the highest agreement of the Senate and in your weightiest and greatest judgement. From which surely you understand how much praise there is in a kindness given, when in one received there is such glory.

\ciceroSection{4} Truly fortunate is the man from whose preservation a joy almost as great comes to everyone as to himself --- a thing that has happened to him deservedly and by the best right. For who is more pre-eminent than he in nobility, or in uprightness, or in zeal for the best arts, or in blamelessness, or in any kind of praise? No man has so great a stream of talent, no force or such abundance of speaking or writing exists, which could --- I shall not say adorn, but even narrate --- your deeds, Gaius Caesar. And yet I affirm this, and shall say it with your good leave: there is no praise in them ampler than the praise you have won this day.

\ciceroSection{5} I am often used to setting before my eyes, and to bringing it up gladly in frequent conversation, that all the deeds of our own commanders, of all foreign nations and of the most powerful peoples, of all the most famous kings, cannot be compared with yours --- not in the size of their contests, not in the number of their battles, not in the variety of their regions, not in the speed of completion, not in the unlikeness of their wars; nor indeed could the most far-flung lands have been traversed more quickly by anyone's steps than they have been ranged over, I shall not say by your marches, but by your victories.

\ciceroSection{6} These things, unless I should confess them to be so great that the mind and thought of scarcely any man can take them in, I should be out of my senses. But there are nonetheless other things still greater. For warlike praises some men are used to belittling with words, and to drawing them away from the commanders and sharing them out with many, so that they may not be properly the generals' own. And certainly in arms the valour of soldiers, the lay of the ground, the help of allies, fleets, supplies are of great assistance; but Fortune by a kind of right of her own claims the largest part for herself, and whatever has been carried through prosperously she counts almost entirely her own.

\ciceroSection{7} But of this glory, Gaius Caesar, which you have just now won, you have no partner at all. The whole of it, however great it is --- and certainly it is the greatest --- the whole of it, I say, is yours. Nothing of that praise does the centurion pluck for himself, nothing the prefect, nothing the cohort, nothing the squadron; nor indeed does that very mistress of human affairs, Fortune, offer herself for partnership in this glory: she gives way to you, she confesses it to be wholly and properly your own. For temerity is never mixed with wisdom, nor is chance admitted into deliberation.

\ciceroSection{8} You have subdued nations barbarous in their savagery, innumerable in multitude, unbounded in their territories, abounding in every kind of resource; yet you have conquered things whose nature and condition was such that they could be conquered. For there is no force so great that it cannot be weakened and broken by iron and strength. But to conquer one's own spirit, to check one's anger, to deal moderately with the conquered, to lift up an adversary outstanding in nobility, in talent, in courage --- not only to lift him up from where he lies but even to enlarge his former dignity: the man who does this, I do not compare with the greatest of men, but I judge him most like a god.

\ciceroSection{9} And so, Gaius Caesar, your praises in war shall be celebrated, not only in our own writings and tongues but in those of almost every nation, nor shall any age ever fall silent of your praises; yet things of this sort, somehow, even when they are read, seem to be drowned out by the shouting of soldiers and the sound of trumpets. But when we hear or read of some deed done with clemency, with gentleness, with justice, with moderation, with wisdom --- especially in anger, which is the enemy of deliberation, and in victory, which is by nature insolent and proud --- with what zeal we are set on fire, so that we often love men we have never seen, not only in deeds done but even in fiction!

\ciceroSection{10} But you, whom we behold present before us, whose mind and feelings and face we see --- so that whatever the fortune of war has left to the commonwealth, that you wish to be kept safe --- with what praises shall we extol you, with what zeal shall we attend you, with what good will shall we embrace you? The very walls, by all that's holy, as it seems to me, of this Senate-house are eager to give you thanks, because in a short time that authority will return to these its own and its ancestors' seats. For my part, when just now together with you I saw the tears of Gaius Marcellus, a most excellent man endowed with a memorable devotion, the memory of all the Marcelli flooded my heart --- to whom, even though they are dead, by your preservation of M. Marcellus you have given back their dignity, and you have nearly rescued from extinction a most noble family now reduced to a few.

\ciceroSection{11} This day you rightly set above your greatest and innumerable triumphs. For this deed is the proper possession of Gaius Caesar alone; the rest were done under your leadership, great indeed, but with a great and numerous companionship. Of this deed, however, you are both the leader and the companion: a deed so great that to your trophies and monuments age shall bring an end --- for there is nothing made by labour and hand that ancient time does not finish and consume ---

\ciceroSection{12} but this justice and lenience of yours shall flower more each day. So that just as much as length of time shall draw from your works, so much shall it add to your praises. And all the other victors of civil wars you had already conquered in equity and pity: but on this very day you have conquered yourself. I fear that what I am about to say cannot be understood in the hearing as I myself feel it in the thinking: you seem to have conquered victory itself, when you have given back to the conquered the things which had been taken from them. For when by victory's own right all of us, the conquered, ought to have been killed, we have been preserved by the judgement of your clemency. Rightly, then, are you the one man unconquered, by whom even the condition and force of victory itself has been conquered.

\ciceroSection{13} And mark, senators, how widely this judgement of Gaius Caesar's reaches. For all of us who by some unhappy and deadly fate of the commonwealth were driven to those arms --- although we are held by some fault of human error --- have certainly been acquitted of any crime. For when at your entreaty he preserved M. Marcellus for the commonwealth, and gave me back to myself and likewise to the commonwealth though no man entreated for me, and gave the rest of those most distinguished men back both to themselves and to their country (whose presence and dignity you see in this very gathering): he was not bringing enemies into the Senate-house, but judging that by most men the war had been undertaken rather out of ignorance, and out of a false and groundless fear, than out of any greed or cruelty.

\ciceroSection{14} In which war I always thought we should listen to peace, and I always grieved that not only peace itself but even the speech of citizens demanding peace was being rejected. For I never followed those arms, nor any civil arms, and my counsels were always the allies of peace and the toga, not of war and weapons. The man I followed I followed out of a private duty, not a public one; and the loyal memory of a grateful spirit had such power with me that without any greed, indeed without even any hope, knowingly and consciously, I rushed as it were towards a voluntary destruction.

\ciceroSection{15} And this plan of mine was not at all hidden. For both in this order, while the matter was still untouched, I said many things in favour of peace, and in the war itself I held the same view, even at peril of my life. From which no man will be so unjust an estimator of facts as to doubt what Caesar's wish about the war was, when as soon as he could he voted that the makers of peace should be preserved, and was angrier with the rest. And this was perhaps less marvellous at a time when the outcome was uncertain and the fortune of war hung in the balance: but a victor who loves the makers of peace surely declares that he had preferred not to fight than to conquer.

\ciceroSection{16} And of this thing I am M. Marcellus's witness. For our feelings agreed, as in peace always, so then even in war. How often, and with what grief, did I see him dreading both the insolence of certain men and the very ferocity of victory itself! The more grateful, therefore, Gaius Caesar, ought your generosity to be to us who saw those things. For now it is not the causes that have to be compared with each other, but the victories.

\ciceroSection{17} We have seen your victory bounded by the close of the battles: a sword bare of its sheath we have not seen in the city. The citizens we lost were struck down by the violence of Mars, not by the wrath of victory; so that no man should doubt that, if he could, Gaius Caesar would call back many from the world below, since out of the same battle-line he preserves those whom he can. As for the other side, I say no more than what we all feared --- that their victory would have been all too prone to wrath.

\ciceroSection{18} For some of them threatened not only those still in arms but at times even those who were idle, and used to say that we must consider not what each man had thought but where he had been; so that to me at any rate the immortal gods seem, even if they exacted penalties from the Roman people for some offence in stirring up so great and so mournful a civil war, now either appeased or at last sated, to have set the whole hope of safety upon the clemency and wisdom of the victor.

\ciceroSection{19} Rejoice therefore in that so outstanding a good of yours, and enjoy your fortune and glory together with the very nature and character that are your own --- from which surely the greatest fruit and delight come to a wise man. Of your other deeds, when you remember them, although you will very often congratulate your virtue, yet you will most often congratulate your good luck: but of us, whom you have wished to be alive in the commonwealth alongside you, every time you think, just so often you will think of your greatest kindnesses, just so often of your incredible generosity, just so often of your singular wisdom: which I shall venture to call not only the highest goods, but indeed the only goods. For there is such a splendour in true praise, such a dignity in greatness of spirit and in counsel, that these things appear to be the gifts of virtue, while the rest seem only lent by fortune.

\ciceroSection{20} Do not, then, grow weary in preserving good men --- men who have fallen, not from any private greed or perversity, but from a sense of duty perhaps foolish, certainly not wicked, and from a kind of vision of the commonwealth. For there is no fault on your part, if certain men feared you; and on the contrary it is the highest praise, that they have felt that you were least to be feared.

\ciceroSection{21} Now I come to that weightiest complaint and most appalling suspicion of yours, which has to be guarded against not by you alone any more than by all the citizens, and most of all by us who have been preserved by you. Though I hope it is false, still I shall never play it down: for your watchfulness is our watchfulness. And if I must err on one side or the other, I had rather seem too timid than too little prudent. But who is so deranged as this? Of your own people? --- and yet, who more your own than those to whom you, against their hope, have given back their safety? Or from the number of those who were with you? Such madness in any man is not to be believed --- that he should not set above his own the life of the leader under whom he has won everything that is greatest. Or, if none of your people plan any crime, is there need to beware that none of your enemies do? Who? For all of those who were enemies have either lost their lives by their own stubbornness, or have kept them by your pity, so that either none of the enemy survive, or those who were enemies are now most friendly.

\ciceroSection{22} And yet, since in the minds of men there are such great hiding-places and such deep recesses, let us by all means heighten your suspicion: for at the same time we shall heighten our diligence. For who of all men is so ignorant of affairs, so untrained in public matters, so careless of his own safety and of the common safety, as not to understand that his own is contained in yours, and that all men's lives hang from your single life? For my part, thinking of you day and night as I ought, I dread merely human chances and the uncertain accidents of health and the common frailty of nature, and I grieve that, while the commonwealth ought to be immortal, it rests upon the breath of one mortal.

\ciceroSection{23} But if, in addition to human chances and the uncertain movements of health, there is also added crime and the conspiracy of ambush, what god, if he wished, could come to the aid of the commonwealth, do we suppose? Everything must be raised up by you, Gaius Caesar, by you alone, which you perceive lies fallen, struck and prostrated by the very assault of war, which had to be: the courts must be set up, public faith restored, the lusts checked, the next generation increased, all things which have now slipped away and flowed apart must be bound up by strict laws.

\ciceroSection{24} In so great a civil war, in such an ardour of spirits and of arms, it could not be refused but that the commonwealth, shaken, should --- whatever the outcome of war had been --- lose much of its dignity's ornaments and its stability's defences, and that both leaders, while in arms, should do many things which the same men in the toga would have forbidden to be done. All these wounds of war must now be healed by you, and none can be the physician for them except you.

\ciceroSection{25} And so I heard with unwilling ears that famous saying of yours, most splendid and most wise: ``I have lived long enough for nature or for glory.'' Long enough, if you say so, perhaps for nature; I shall add, if you please, for glory too: but for your country --- which is the greatest thing --- certainly too little. Therefore lay aside, I beg, that learned men's prudence about despising death: do not, at our peril, be a philosopher. For it has often come to my ears that you keep saying the same thing too often --- that you have lived enough for yourself. I believe it; but I should hear it gladly, if you lived for yourself alone, or if you had been born for yourself alone. The safety of all citizens and the whole commonwealth your deeds have embraced; you are so far from the completion of those greatest works that you have not yet laid the foundations of what you are planning. And here will you measure the term of your life, not by the safety of the commonwealth, but by the equanimity of your mind? What if this is not enough even for glory --- of which, however wise you may be, you will not deny that you are most eager?

\ciceroSection{26} ``Do we leave so little behind?'' you will say. On the contrary, enough for very many others, however many they be; for you alone, too little. For whatever it is, however ample, that is little when there is something ampler. And if, Gaius Caesar, this was to be the end of your immortal deeds --- that, the adversary having been overcome, you should leave the commonwealth in that condition in which it now is --- see, I beg you, that your divine virtue may not earn more wonder than glory; if indeed glory is the bright and far-spread fame of great services to one's fellow citizens or to one's country or to the whole race of men.

\ciceroSection{27} This therefore is the part that remains to you; this is the act that is left; in this it must be laboured, that you set the commonwealth in order, and that you yourself first of all may enjoy it in the highest tranquillity and ease: then, if you please, when you shall have both paid what you owe to your country and filled up nature herself with the fullness of living, then say that you have lived long enough. For what indeed is this very thing, ``long,'' in which there is anything final? When the end comes, all past pleasure counts as nothing, because no further is to be. Yet that spirit of yours has never been content with these narrow bounds which nature has given us for living: it has always blazed with the love of immortality.

\ciceroSection{28} Nor is this life of yours to be reckoned the life which is contained in body and breath: that, I say, that is your life, which shall flourish in the memory of all ages, which posterity shall nourish, which eternity herself shall ever watch over. To this you must devote yourself, to this show yourself: which has long had many things to wonder at; now it also awaits things to praise. Posterity shall surely be stupefied as it hears and reads of your commands, your provinces, the Rhine, the Ocean, the Nile, your innumerable battles, your incredible victories, your monuments, your gifts, your triumphs.

\ciceroSection{29} But unless this city is settled by your counsels and your institutions, your name shall only wander far and wide, and have no settled seat or sure dwelling. There shall be even among those yet unborn, as there has been among us, great disagreement, when some shall lift your deeds with praises to heaven, but others perhaps shall find something wanting, even the greatest thing, unless you have quenched the fire of civil war by the safety of your country --- so that the one shall seem to have been the work of fate, the other of your judgement. Serve, then, those judges who shall judge you many ages hence, and I rather think more incorruptibly than we do; for they shall judge without love and without greed, and again without hatred and without envy.

\ciceroSection{30} And this, even if at that time, as some falsely think, it shall not concern you, certainly does concern you now --- to be such a man that no forgetfulness shall ever darken your praises. The wills of the citizens were divided, and their opinions were pulled apart. For we disagreed not only in counsels and zeal but in arms and camps as well. There was a certain obscurity, there was a contest between the most illustrious leaders; many were in doubt what was best, many what was expedient for themselves, many what was fitting, some even what was lawful.

\ciceroSection{31} The commonwealth has now passed through this mournful and fated war: he has conquered, who did not let fortune kindle his hatred but softened it with his goodness; and who did not judge that all those with whom he was angry deserved also either exile or death. Arms have been laid down by some, struck from the hands of others. He is a thankless and unjust citizen who, freed from the peril of arms, still keeps his spirit under arms --- so that even the better man is he who fell in the line of battle, who poured out his life in the cause. For what to some appears stubbornness, to others may appear constancy.

\ciceroSection{32} But now all disagreement has been broken by arms, extinguished by the equity of the victor: it remains that all should will the same thing, all who have anything not only of wisdom but even of sanity. Unless you, Gaius Caesar, are safe, and remain in that disposition which you have shown both before and most of all today, we cannot be safe. Therefore all of us who wish these things to be kept safe both exhort and beseech you to take thought for your life, for your safety; and all of us promise you --- since you think there is something underneath that must be guarded against --- to speak for others what I feel about myself: not only watches and guards, but the interposing of our own flanks and our own bodies.

\ciceroSection{33} But that my speech may end where it began, we all give you, Gaius Caesar, the greatest thanks, and feel still greater thanks. For all feel the same thing, as you could perceive from the prayers and tears of all. But because it is not necessary that they should all speak standing, they wish at any rate that it be said by me, to whom in some way it is necessary; and what it is fitting should be done, now that M. Marcellus has been given back by you to this order, to the Roman people, and to the commonwealth, I understand to be done. For I see that all rejoice, not as for the safety of one man alone, but as for the safety of all.

\ciceroSection{34} As for the highest good will, which mine toward him has always been known to all to be such that I should yield to no one save to Gaius Marcellus, his most excellent and most loving brother --- since I have shown it by anxiety, by care, by labour, as long as there was any doubt about his safety: certainly at this time, freed from great cares, vexations, griefs, I owe it the more to show. And so, Gaius Caesar, I give you thanks in such a way that, although in all things I have not only been preserved by you but adorned, yet to your innumerable services towards me alone, this deed of yours --- which I did not think could now happen --- has added a great crown.
