% Pro Cn. Plancio (pro-plancio) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Because of the outstanding and singular loyalty of Cn. Plancius, gentlemen, in guarding my safety, when I saw so many good men supporting the honour due to him, I felt no ordinary pleasure: for I saw that the man whose service had been my salvation was now finding the memory of my own ordeal to be his ally. But when I heard that men who were partly my enemies, partly my enviers, were backing this prosecution, and that the very thing which had aided Cn. Plancius in his canvass was now turning against him in this trial, I grieved, gentlemen, and bore it bitterly --- that this man's safety should be the more endangered for the very reason that he had sheltered my own safety and life with his goodwill, as a guard and protection.

\ciceroSection{2} But now, gentlemen, the sight of you, this gathering of yours, restores and refreshes my mind, as I look upon and contemplate each one of you. For in this number I see no one to whom my safety was not dear, no one whose supreme service to me does not stand on record, no one to whom I am not bound by an everlasting memory of the kindness. And so I have no fear that Cn. Plancius's guardianship of my safety should tell against him before the very men who themselves most wished to see me safe; and it more often comes into my mind, gentlemen, that I should wonder at M. Laterensis --- a man most devoted to my dignity and my safety --- for having chosen this man above all others as his defendant, than that I should fear it might seem to you that he did so with good reason.

\ciceroSection{3} And yet I do not take so much to myself nor presume so far, gentlemen, as to think that Cn. Plancius has won impunity by his services toward me. Unless I show his life to be of the utmost integrity, his character most temperate, his loyalty, self-restraint, devotion, and innocence to be of the highest --- I will refuse nothing where his punishment is concerned. But if I make good everything that is to be expected of good men, then I will ask of you, gentlemen, that to the man by whose compassion my safety was guarded you grant your own compassion, at my pleading. For my part, to the rest of the labours which I undertake in this case --- greater than in others --- I take on this trouble besides: that I must speak not only for Cn. Plancius, whose safety I am bound to protect no less than my own, but also for myself, about whom the prosecutors have said almost more than about the matter and the defendant.

\ciceroSection{4} And yet, gentlemen, if anything in me myself has been censured in such a way that it is detached from him, that does not greatly trouble me; for I am not afraid that, because grateful men are very rarely found, therefore, when they say I am too grateful to him, this can be turned into a charge against me. But the things they have stirred up in such a way as either to say that Cn. Plancius's services toward me were less than I myself proclaimed, or, if they were of the highest, to deny nonetheless that they ought to weigh with you as heavily as I supposed --- these things I must handle, gentlemen, and with measure, lest I myself give offence, and only then, when I have answered the charges, lest the defendant seem to have been defended not so much by his own innocence as by the recollection of my own ordeal.

\ciceroSection{5} But in a case easy and straightforward, gentlemen, a most difficult and slippery method of defence is set before me. For if it were merely necessary for me to speak against Laterensis, even that itself would be painful, given so great an intimacy between us and so great a friendship. For this is the old law of just and true friendship, which has long held between him and me: that friends should always wish the same things; nor is there any surer bond of friendship than agreement and partnership in counsels and in wishes. Yet for me the most painful thing in this matter is not to speak against him, but far more this: that I must speak against him in a case in which a certain comparison of the men themselves seems to have to be made.

\ciceroSection{6} For Laterensis asks --- and on this one point above all he presses --- by what virtue, by what merit, by what dignity Plancius surpassed him. And so, if I yield to that man's distinctions, which are many and great, not only must I sacrifice this man's dignity, but I must also accept the suspicion of bribery; but if I set this man before him, my speech must be held insulting, and I must say the very thing he demands of me: that Laterensis was surpassed in dignity by Plancius. Thus either the good name of a most dear friend must be wounded, if I follow that line of the prosecution, or the safety of one who has deserved supremely well of me must be abandoned. But, Laterensis, let me confess that I am being carried along blind and headlong in this case, if I shall ever have said that you could have been surpassed in dignity by Plancius or by anyone. And so I will withdraw from the comparison to which you summon me, and will come to the one to which the case itself leads me.

\ciceroSection{7} What? Do you think the people is a judge of dignity in the matter of magistracies? Perhaps sometimes it is; would that it always were! But it is very rarely, and, if ever it is, it is in the conferring of those magistracies to which it thinks its own safety entrusted; at these lighter elections, honour is won by the diligence and influence of the candidates, not by those distinctions which we see in you. For where the people is concerned, he who either envies or favours is always an unfair judge of dignity. And yet you can establish nothing in yourself, Laterensis, that is the peculiar mark of your own praise, without its being shared by you with Plancius.

\ciceroSection{8} But this whole point will be dealt with in another place; now I argue only about the right of the people, which both can and is sometimes accustomed to pass over the deserving; nor, if one who ought not to have been passed over has been passed over by the people, is one who was not passed over to be condemned by the jury. For if it were so, then what the senators among our forefathers could not retain --- that they should be the censors of the elections --- the jury would now hold, which would be far less to be borne. For in those days the man who had been elected did not hold the magistracy, if the senators had not given their sanction; but now it is demanded of you that you reverse the people's verdict by the destruction of the man who was elected. And so now, since by the door I did not wish I have entered upon the case, I seem to hope that my speech will be so far from the slightest suspicion of giving you offence that I should rather reproach you, for unfairly bringing your own dignity into hazard, than that I should try to touch it with any insult.

\ciceroSection{9} Do you think your self-restraint, your industry, your devotion to the commonwealth, your virtue, your innocence, your loyalty, your labours, have been broken and cast down and rejected, because you were not made aedile? See at last, Laterensis, how greatly I differ from you. If, so help me god, there were only ten good men in the state --- wise, just, and grave --- who had judged you unworthy of the aedileship, I should think a graver judgement passed upon you than this which you fear may seem to have been passed by the people. For the people does not always judge at elections, but is for the most part moved by influence, yields to entreaties, favours those by whom it has been most courted; and finally, even if it does judge, it is led to judgement by no discrimination or wisdom, but sometimes by impulse and even by a kind of recklessness. For in the crowd there is no deliberation, no reason, no discernment, no diligence; and the wise have always said that what the people has done is to be borne, not always to be praised. And so, when you say that you ought to have been made aedile, you are accusing the people's fault, not your competitor's.

\ciceroSection{10} Grant that you were worthier than Plancius --- on which very point I will contend with you a little later in such a way as to preserve your dignity --- but grant that you were worthier: then it is not the competitor by whom you were beaten, but the people by whom you were passed over, that is at fault. And here you ought first to consider this: that at elections, especially those for the aedileship, there is the people's enthusiasm, not its judgement; that those votes are coaxed, not sifted; that those who cast their vote consider what each man owes to themselves more often than what is owed to each by the commonwealth. But if you would rather it were a judgement, then it is not for you to overturn it, but to bear it.

\ciceroSection{11} ``The people judged badly.'' But it judged. ``It ought not.'' But it could. ``I will not bear it.'' But many of the most distinguished and wisest citizens have borne it. For this is the condition of free peoples, and especially of this people which is foremost, lord and conqueror of all nations: that by its votes it can either give or take away from each man what it wishes. And it is for us --- for us who are tossed in this storm and on the waves of the people --- to bear the people's wishes with moderation, to win over those that are estranged, to keep those we have gained, to calm those that are stirred up; if we do not think honours great, then not to be the people's servant; but if we seek them, then not to weary of supplication.

\ciceroSection{12} I come now to the part of the people itself, that I may argue against you in its words rather than my own. If it should meet you and could speak with one voice, it would say this: ``I did not set Plancius before you, Laterensis; but, since you were equally good men, I conferred my favour rather upon the man who had striven for it from me than upon the one who had not entreated me too humbly.'' You will answer, I suppose, that, relying on the splendour and antiquity of your family, you did not think you had to canvass so hard. But that man will indeed call you back to its own practices and to the examples of its forefathers; it will say that it has always wished to be asked, always to be entreated; that it preferred M. Seius --- a man who could not keep even his equestrian splendour unimpaired through the calamity of a trial --- to a man most noble, most innocent, most eloquent, M. Piso; that it set ahead of Q. Catulus --- born in the highest family, a man most wise and most blameless --- I do not say C. Serranus, a most foolish man (for he was nevertheless noble), nor C. Fimbria, a new man (for he had both spirit enough and judgement), but Cn. Mallius, not only ignoble but without virtue, without talent, of a life even despised and sordid.

\ciceroSection{13} ``My eyes longed for you,'' it says, ``when you were at Cyrene; for I preferred that I, rather than our allies, should enjoy your virtue, and the more it mattered, the more it was withdrawn from me, when I did not see you. Then, when I thirsted for your virtue, you deserted and abandoned me. For you had begun to seek the tribunate of the plebs at a time which called for that eloquence and virtue of yours; and since you abandoned that canvass --- if you meant by this that in so great a storm you could not steer, I had doubts about your virtue; if that you were unwilling, about your will; but if, as I rather understand, you were reserving yourself for other occasions, then I too,'' the Roman people will say, ``have called you back to those occasions for which you reserved yourself. Seek, then, that magistracy in which you can be of great use to me; whoever the aediles shall be, the same games are made ready for me; but it matters very greatly to me who the tribunes of the plebs are. So either give me back what you had promised, or, if --- what matters less to me --- that perhaps pleases you more, I will give you back that aedileship even though you seek it carelessly; but that you may attain the highest honours in keeping with your dignity, I advise you to learn to entreat me a little more diligently.''

\ciceroSection{14} This is the people's speech; but mine, Laterensis, is this: that a juror ought not to ask why you were beaten, provided only you were not beaten by bribery. For if, every time a man is passed over who ought not to have been passed over, it shall be necessary for the man who was elected to be condemned, then there is no longer any reason to entreat the people, no reason to await the counting, no reason to await the announcement of the votes. As soon as I see who the candidates are, I shall say:

\ciceroSection{15} ``This man is of a consular family, that one of a praetorian; the rest I see are from the equestrian order; they are all without stain, all equally good men and upright; but the gradations must be kept: let the praetorian yield to the consular stock, let the equestrian order not contend with the praetorian name.'' Enthusiasm would be abolished, electoral support extinguished, no contests, no liberty of the people in conferring magistracies, no awaiting of the votes; nothing, as for the most part happens, would turn out contrary to expectation, there would henceforth be no variety in elections. But if it very often happens that we are surprised both at some who are elected and at some who are not; if the Campus and those waves of the elections, like a deep and boundless sea, so seethe with a kind of tide that they advance toward some and recede from others --- shall we nonetheless, amid the surge of enthusiasms and the movement of recklessness, look for some measure and deliberation and reason?

\ciceroSection{16} Therefore do not call me to the comparison of the two of you, Laterensis. For if the ballot is welcome to the people --- which lays open men's faces but conceals their minds, and gives them that liberty to do what they wish while promising what they are asked --- why do you press, in a trial, that what does not happen on the Campus should happen here? ``This man is worthier than that one'' is exceedingly hard to say. How then is it fairer? Thus, I suppose: what is at issue, what is enough for the juror, is this: ``This man was elected.'' ``Why he rather than I?'' Either I do not know, or I will not say, or, finally --- what would be most grievous to me if I said it, yet which I might say with impunity --- ``not rightly.'' For what would you gain, if I were to use that last line of defence: that the people did what it wished, not what it ought?

\ciceroSection{17} What? If I defend the people's act too, Laterensis, and show that Cn. Plancius did not creep up to office, but came to it by the course that has always lain open to men sprung from this equestrian order of ours --- can I not strip from your speech the comparison of you two, which cannot be handled without insult, and bring you at last to the case and the charge? If, because he is the son of a Roman knight, he ought to have been the inferior, then all the sons of Roman knights have canvassed against you. I say no more; yet I do wonder at this: why you are angry above all with the man who stood furthest from you. For my part, if at any time, as happens, I am jostled in a crowd, I do not blame the man who is at the top of the Sacred Way when I am shoved at the Fabian arch, but the one who runs into me and falls against me. You are not angry with Q. Pedius, a brave man, nor with this A. Plotius, a most distinguished man and my friend; and you think yourself thrust out by the man who pushed these aside, rather than by those who leaned upon you yourself.

\ciceroSection{18} But still, this is your first contest with Plancius, in the matter of your birth and family, in which you are beaten by him; for why should I not confess what must be confessed? But I was beaten by my competitors no more here than I was, both on other occasions and in my canvass for the consulship. Yet see whether these very things which you despise told in his favour. For let us make the comparison thus. Your name is consular in both families. Do you doubt, then, that all who favour the nobility, who think it the finest thing, who are led by your ancestral images and names, made you aedile? For my part I do not doubt it. But if there are too few who love the nobility, is that fault ours? Indeed, let us come to the head and the source of each man's stock.

\ciceroSection{19} You are from the most ancient town of Tusculum, from which come very many consular families --- among them the Iuventian --- so many as do not come from all the rest of the towns together; this man is from the prefecture of Atina, not so ancient, not so honoured, not so near the city. How much difference do you mean this to make to the reckoning of a canvass? First, which do you think favour their own people more, the Atinates or the Tusculans? The one --- and this I can easily know, on account of the neighbourhood --- when they saw the father of this most distinguished and excellent man, Cn. Saturninus, as aedile and as praetor, rejoiced in a wonderful way, because he had been the first to bring the curule chair not only into that family but into that prefecture; the others --- because, I suppose, the town is crammed with consular men, for that they are not malicious I know for certain --- I have never perceived to rejoice more vehemently at the honour of their own. We have this, our towns have it.

\ciceroSection{20} What shall I say of myself, of my brother? At our honours the very fields and mountains, I might almost say, rejoiced. Do you ever see any Tusculan boast of M. Cato, that man foremost in every virtue, or of Ti. Coruncanius, his fellow townsman, or of so many Fulvii? No one says a word. But whatever man of Arpinum you fall in with, even if you do not wish it, you will nonetheless have to hear something perhaps even about me, but certainly about C. Marius. So this man, first of all, had the burning enthusiasm of his own people; you, only as much as could exist in men already sated with honours.

\ciceroSection{21} Next, your fellow townsmen are indeed men of the greatest splendour, but still few, if at least they are compared with the Atinates; this man's prefecture is full of the bravest men, so that none in all Italy can be called more populous --- which multitude indeed you now see, gentlemen, in mourning garb and grief, suppliants before you. These so many Roman knights, so many tribunes of the treasury --- for we have dismissed the plebs from the trial, who all attended at the elections --- what strength, what dignity did they bring to this man's canvass? For not only the Teretine tribe, of which I shall speak in another place, but dignity, the gaze of all eyes, a solid and sturdy and unremitting crowd, did they furnish. For the towns are strongly moved also by the bond of neighbourhood.

\ciceroSection{22} Everything I say of Plancius I say from my own experience; for we are neighbours of the Atinates. A neighbourhood is to be praised, even loved, that keeps that old custom of duty --- not darkened by malice, not inured to lies, not painted over, not deceitful, not schooled in the artifice of pretence, whether of the suburbs or even of the city. No man of Arpinum failed to support Plancius, no man of Sora, none of Casinum, none of Aquinum. That most populous tract of Venafrum and Allifae, that whole region of ours, so rough and mountainous, so faithful and plain and supportive of its own, thought itself adorned by his honour, thought itself increased by his dignity; and from these same towns there are present now Roman knights, publicly, with an embassy and testimony, and they are now in no less anxiety than they were then in enthusiasm. For it is graver to be stripped of one's fortunes than not to be increased in dignity.

\ciceroSection{23} Therefore, just as there were other, more brilliant things in you, Laterensis, which your ancestors had left you, so Plancius surpassed you in this matter not only of his town but of his neighbourhood --- unless perhaps the neighbourhood of Labicum or Gabii or Bovillae was helping you, from which towns there are now scarcely to be found men to fetch the meat at the Latin festival. Let us add, if you like, the very thing which you think tells against him too: that his father is a tax-farmer. And who does not know how great a help that order is in seeking office? For the flower of the Roman knights, the ornament of the state, the bulwark of the commonwealth, is contained in the order of the publicani.

\ciceroSection{24} Who is there, then, who would deny that this order's enthusiasm in Plancius's election was singular? And not without reason, whether because his father is a man who has long been the chief of the tax-farmers, or because he himself was uniquely esteemed by the partners, or because he canvassed most diligently, or because he was entreating on behalf of his son, or because this man's own supreme services to that order, in his quaestorship and tribunate, stood on record, or because in honouring this man they thought they were honouring their order and looking after their own children. Something besides --- I will say it timidly, but it must nonetheless be said --- for not by resources, not by invidious influence, not by power scarcely to be borne, but by the recalling of a kindness, by compassion, by entreaties, did I too bring something. I appealed to the people, tribe by tribe; I humbled myself and made supplication; men who offered themselves to me of their own accord, by Hercules, who of their own accord made promises --- these too I asked. The cause of my asking prevailed, not influence.

\ciceroSection{25} Nor, if a most eminent man --- one to whom nothing could rightly be denied that he asks --- failed, as you say, to obtain something for someone, am I therefore arrogant because I say that I prevailed. For, to pass over the fact that I was labouring on behalf of a man who was strong by his own resources, the request itself is always most effective which is joined above all to the duty of close connection. For I was not asking in such a way that I seemed to be canvassing because he was my friend, because my neighbour, because I had always been most intimate with his father, but as if for a parent and the guardian of my safety. It was not my power but the cause of my asking that prevailed. No one rejoiced at my restoration, no one grieved at my wrong, to whom this man's compassion toward me was not welcome.

\ciceroSection{26} For if, before my return, good men commonly offered themselves to Cn. Plancius of their own accord, when he was seeking the tribunate --- to a man for whom my name in absence had counted toward his honour --- do you not think that my entreaties in person availed him? Or do the colonists of Minturnae stand in everlasting praise because they snatched C. Marius from civil steel and from impious hands, because they received him under their roof, because, worn out with hunger and the waves, they revived him, because they gathered him provisions for the journey, because they gave him a ship, because, as he left the land he had saved, they escorted him with vows, with omens, and with tears --- and yet do you wonder that to Plancius this loyalty, this compassion, this virtue counted toward his honour, because he received me, whether driven out by force or withdrawing by design, helped me, guarded me, and preserved me for these men, for the Senate and the Roman people, that they might have one to bring back?

\ciceroSection{27} The deeds I have spoken of could, by Hercules, have covered over Cn. Plancius's faults --- so you need not wonder that in the life of which I am now about to speak, so many and so great supports have been a help to this office. For this is the man who, as a youth, having set out with A. Torquatus into Africa, was so esteemed by that gravest and most upright man, most worthy of all praise and honour, as both the bond of shared quarters and the modesty of a most modest young man demanded; which, were he present, that man would declare no less than this man's cousin and father-in-law, T. Torquatus, declares it --- a man equal to the other in every virtue and praise, joined to him indeed by the greatest ties of kinship and connection, yet by such great ties of love that those bonds of relationship seem slight. Afterward in Crete he was the tent-companion of Saturninus, his kinsman, and a soldier of this Q. Metellus; and since he was most thoroughly approved by him, and is so to this day, he ought to hope that he has been approved by all. In that province C. Sacerdos was legate --- a man, what courage, what constancy! L. Flaccus --- what a man, what a citizen! What they think of this man, they declare by their constant attendance and their testimony.

\ciceroSection{28} In Macedonia he was military tribune, and afterward quaestor in the same province. First, Macedonia loves him so, as these chief men of their communities make plain; who, though sent on another errand, yet, moved by his sudden peril, sit by him, labour for him, and judge that, if they have stood ready at his side, they will do something more welcome to their own communities than if they had completed their embassy and their instructions. L. Apuleius, indeed, thinks so much of this man that he has surpassed, by his good offices and goodwill, that custom of our forefathers which prescribes that praetors stand in a parent's place to their own quaestors. He was tribune of the plebs --- not perhaps so vehement as those whom you justly praise, but certainly of such a kind that, had all men always been so, no vehement tribune would ever have been missed.

\ciceroSection{29} I pass over those things which, even if they are less upon the stage, are at any rate praised when they are brought into the open: that he lives well with his own --- first with his father (for in my judgement filial piety is the foundation of all the virtues), whom he reveres as a god (for a parent is to his children not much otherwise), but loves as a comrade, a brother, an equal. What shall I say of his bearing toward his uncle, toward his connections, toward his kinsmen, toward this Cn. Saturninus, a most distinguished man? How great a desire for this office of his you believe these men felt, when you see their partnership in grief! What shall I say of myself, who in this man's peril seem to myself to be the defendant? What of these so many men of such standing whom you see in changed dress? And these are the proofs, gentlemen, solid and clear-stamped; these are the marks of probity not painted over with a forensic gloss, but branded with the household tokens of truth. Easy is that running-up to greet and that crowd-pleasing flattery; it is looked at, not handled; it appears at a distance, it is not sifted, not taken into the hands.

\ciceroSection{30} A man, then, adorned with everything, as much from without as from within --- in some respects lower than you (I mean in birth and name), in others higher: in the zeal of his townsmen, of his neighbours, of the partnerships of revenue, in the memory of my own dark days; equal in virtue, integrity, modesty --- do you wonder that this man was made aedile? Do you splash that brightness of his life with those stains of yours? You hurl adulteries at him, which no one could recognize, not by a name, nor even by a suspicion. You call him a two-husbanded man, so as to coin even words, not charges alone. You say that someone was taken off by him into the province for the sake of his lust --- which is no charge, but an unpunished lie in slander; that a little actress was carried off, which is said to have been done at Atina by the young men under a certain old usage against stage-folk, very much a country custom.

\ciceroSection{31} O youth elegantly spent, against which, even when what was lawful is thrown up, that very thing too is found false! Someone was let out of prison --- and indeed let out by an oversight, let out, as you have learned, at the request of a connection and an excellent young man; and the same man was afterward sought out by the praetor's order. And these, and no others, are the slanders flung at the life of a man about whose modesty, scruple, and integrity you might harbour doubt. ``But the father,'' he says, ``ought even to count against the son.'' O word harsh and unworthy of your probity, Laterensis! That a father, in a trial on a man's standing, that a father, in a struggle for his fortunes, that a father, before such men as these, ought to count against his son? A father who, were he the basest, the most sordid of men, would yet by the very name of father carry weight with merciful and compassionate jurors --- would carry weight, I say, by the common feeling of all and by the sweetest commendation of nature.

\ciceroSection{32} But when Cn. Plancius is a Roman knight of such antiquity in the equestrian name that his father, his grandfather, all his forebears were Roman knights, and held the highest rank both of dignity and of favour in a most flourishing prefecture; then when he himself, in the legions of the commander P. Crassus, served among the most distinguished men, the Roman knights, in the highest splendour; when afterward he was the first among his own people, a most upright and just arbiter of very many matters, the founder of the greatest companies of revenue, the manager of very many --- if not only was nothing in him ever blamed, but everything was praised, will that man yet count against his most honourable son as a father --- a father who could shield even a less honourable man, and a stranger, by his authority or his favour?

\ciceroSection{33} ``He said something rather harshly once,'' he says. Nay, perhaps rather freely. ``But that very thing,'' he says, ``is not to be borne.'' Are they, then, to be borne who complain of this --- that they cannot bear the freedom of a Roman knight? Where is that ancient way, where that fairness of right, where that old liberty which, crushed by civil ills, ought now to be lifting up its head and at last, revived, raising itself again? Shall I recall the slanders of Roman knights against the noblest men, the things the publicans said against Q. Scaevola --- a man surpassing all in talent, justice, integrity --- harshly, fiercely, and freely? When the consul P. Nasica, on returning home after a suspension of business he had proclaimed by edict, asked the crier Granius in the middle of the Forum why he was downcast --- whether because the auctions had been put off --- ``No indeed,'' said Granius, ``but because the embassies have.'' The same man, to a most powerful man, the tribune of the plebs M. Drusus, who was contriving many things in the commonwealth, when he had greeted him and, as happens, said, ``How goes it, Granius?'' answered, ``No, rather, how goes it with you, Drusus?'' That man with rougher wit often grazed the will of L. Crassus, of M. Antonius, and went unpunished: now our state is so far crushed by our arrogance that the liberty which once belonged to a crier in raising a laugh is now not allowed to a Roman knight in raising a lament.

\ciceroSection{34} For what utterance of Plancius's was ever of insult rather than of grief? And of what did he ever complain, except when he was driving off a wrong from his allies and from himself? When the Senate was hindered from rendering to the Roman knights the answer that had always been granted even to enemies, that wrong was a grief to all the publicans, but he bore that very grief a little more openly. That common feeling in others perhaps lay hidden; he had what he felt in spirit along with the rest more ready than the rest both in his face and on his tongue.

\ciceroSection{35} And yet, gentlemen --- for I recognize it from myself --- a great many things are heaped upon Plancius which were never said by him. Because I myself say something now and then, drawn not by zeal but either by the heat of speaking or under provocation, and because, as happens with many, something now and then slips out which, if not exactly witty, is at any rate perhaps not boorish --- whatever anyone has said, they say I said it. But I, for my part, if there is anything that seems to me clever and worthy of a freeborn, learned man, do not disown it; I grow vexed when the things of others, unworthy of me, are heaped upon me. As for his having been the first to vote for the law on the publicans, at the time when a most distinguished man, the consul, gave to that order through the people what, had it been allowed, he would have given through the Senate --- if there is a charge in this because he cast his vote, what publican did not cast it? If because he was first to vote, do you wish that to be a matter of the lot, or of the man who was carrying that law? If of the lot, there is no charge in chance; if of the consul, then you establish that this man too was judged, by a man of the highest standing, the leader of his order.

\ciceroSection{36} But let us come at last to the case. In it you, under the name of the Lex Licinia (which is the law on sodalicia, the electoral clubs), have embraced all the laws on canvassing; for you have followed nothing else in this law except the named jurors. And if this kind of juror is fair in any matter except in this tribal business, I do not understand why the Senate willed that in this one kind the tribes should be named by the prosecutor, and did not carry over the same naming into the other cases, and finally willed that in canvassing itself the rejection of jurors should be made by alternate strikes, and, though it left out no kind of harshness, yet thought that this one was to be passed over.

\ciceroSection{37} Tell me --- is the reason for this at last obscure, or was it both debated, at the time when that matter was being handled in the Senate, and discussed most fully yesterday by Q. Hortensius, with whom the Senate then agreed? This, then, was our view: whoever was a briber of any tribe, and whoever, through that combination which is named a sodality more honourably than truly, corrupted any tribe by base bribery, that man was best known to those men who were of that tribe. Thus the Senate judged that, when there were named to a defendant the very tribes which he had bound to himself by bribery, the same men would be both witnesses and jurors. A harsh kind of trial, certainly --- yet, if either a man's own tribe, or the one most closely connected with him, were named, scarcely to be refused.

\ciceroSection{38} But you, Laterensis, what tribes did you name? The Teretine, I suppose. That, surely, was fair, and surely it was expected, and it was worthy of your consistency. The very tribe of which you keep crying out that Plancius was the seller and corrupter and broker --- that tribe, above all (men most severe and most grave), you ought to have named. But the Voltinian? --- for you take pleasure in trumping up some charge or other about that tribe too. Why, then, did you not name this very one? What has Plancius to do with the Lemonian, what with the Oufentine, what with the Clustumine? For the Maecian you wished not to be one that should judge, but one that should be struck out.

\ciceroSection{39} Do you doubt, then, gentlemen, that M. Laterensis by his own choice selected you not for the meaning of the law, but for some hope of his own about the verdict? Do you doubt that, since he did not name those tribes in which Plancius has great connections, he judged them attended on by this man's good offices, not corrupted by his bribery? For what can you say as to why that naming does not carry the highest harshness, once that principle is removed which we followed in passing the decree?

\ciceroSection{40} That you should pick out from the whole people either your own friends, or my enemies, or finally those whom you judge inexorable, inhuman, cruel; that you should call together --- I being ignorant, unsuspecting, unaware --- both your own connections and those of your friends, men hostile either to me or even to my defenders, and join to them besides those whom you think harsh by nature and hostile to all; then that you should pour them out suddenly, so that I see the seating of my jurors gathered before I could so much as suspect who they were going to be, and that before them, with not even five struck out (which in the most recent case was settled by the advice of the council), you should force me to plead a case touching all my fortunes?

\ciceroSection{41} For it does not follow, if Plancius so lived as to offend no one knowingly, or if you so erred as to name them in your ignorance --- so that we, against your will, came nonetheless before jurors and not before executioners --- that on that account this naming is in itself not harsh. Or did not, lately, the most illustrious citizens refuse to bear the name of a named juror, when out of one hundred and twenty-five jurors, leading men of the equestrian order, the defendant struck out seventy-five and kept fifty, and they threw everything into confusion rather than obey that law and condition? And shall we --- with jurors drawn not from selected men but from the whole people, jurors not named for striking out but set up by the prosecutor --- bear it so that we strike out no one?

\ciceroSection{42} Nor do I now complain of the unfairness of the law, but I show that your deed is at odds with the meaning of the law; and that harsh trial, if you had done as the Senate resolved and the people ordered --- so as to name to this man both his own tribe and those attended on by him --- not only would I not complain, but I would think this man acquitted, with those named as jurors who could likewise be witnesses; and I think now not much otherwise. For when you named these tribes, you made plain that you preferred to use jurors unknown rather than known; you fled the meaning of the law, you rejected all fairness, you chose that the case should be turned over in the dark rather than in the light.

\ciceroSection{43} ``The Voltinian tribe was corrupted by this man; he had the Teretine for sale.'' What would he say before the Voltinians, or before the jurors of his own tribe? Nay rather, what would you say? Which juror out of them would you have either as a silent witness or even rouse to speak? For indeed, if the defendant were naming tribes, Plancius would perhaps name the Voltinian on account of connection and neighbourhood, but his own he would certainly have named. Or if he had to name a presiding officer, whom in the end would he sooner have named than this C. Alfius whom he has --- to whom he must be best known, his neighbour, his fellow-tribesman, a most grave and most just man? Whose fairness indeed, and that goodwill toward Cn. Plancius's safety which the man displays without any suspicion of partiality, easily declares that the jurors of his own tribe were not to be shunned by a man for whom you may see a presiding officer of his own tribe was something to be longed for.

\ciceroSection{44} Nor do I now blame your plan, that you did not name those tribes by which this man was best known, but I show, from your own conduct, that the Senate's plan was not kept. For who of those men would have listened to you then, or what would you say? That Plancius was a broker? Their ears would spit it out, no one would acknowledge it, they would reject it. Or that he was a man of favour? They would gladly hear it, and we would confess it without fear. For do not think, Laterensis, that by those laws on canvassing which the Senate willed to be sanctioned this was done --- that the soliciting of votes, that attentiveness, that favour should be abolished. There have always been good men who wished to be men of favour among their own fellow-tribesmen;

\ciceroSection{45} nor indeed was our order so harsh toward the common people that it should refuse them to be cultivated by our moderate generosity; nor is this to be forbidden to our children --- that they should attend on their fellow-tribesmen, that they should esteem them, that they should be able to bring their tribe together for their connections, that they should look for a like service from them in their own canvassing. For these things are full of duty, full of attentiveness, full even of antiquity. In this very kind we ourselves both took part, when the seasons of our own ambition demanded it, and we have seen the most illustrious men do so, and we wish that today there be as many men of favour as possible. The drawing-up of fellow-tribesmen into squads, the parcelling-out of the people, votes bound fast by bribery --- these would rouse the severity of the Senate and the force and indignation of all good men. Teach this, bring this forward, lean upon this, Laterensis: that Plancius marshalled men into squads, enrolled them, was a broker, fixed the price, distributed the money; then I will wonder that you were unwilling to use those weapons which the law gave you. For with fellow-tribesmen as jurors, if those things are true, we could bear not only their severity but not even their faces.

\ciceroSection{46} Since you have shunned this method, and have been unwilling to have those jurors in whose hands, over this man's offence, there would have been both the most certain knowledge and the gravest indignation --- what will you say before these men, who silently ask of you why you have laid this burden upon them, why you have chosen them above all, why, finally, you have preferred them to divine the matter rather than have those who knew it judge? I say, Laterensis, that Plancius both is himself a man of favour and had in his canvassing many men of favour eager for him; and if you call these men sodales --- club-fellows --- you befoul a dutiful friendship with a charged name; but if, because they are men of favour, you think they are to be accused, do not wonder that you did not attain, by spurning the friendships of men of favour, what your dignity would have demanded.

\ciceroSection{47} For just as I show that Plancius is a man of favour in his own tribe --- because he has done kindly by many, has gone surety for many, has sent very many into the working bands by his father's authority and favour, because finally he has bound up the whole prefecture of Atina by every good office, through himself, through his father, through his forebears --- so do you show that he was a broker, that he bribed, that he enrolled, that he marshalled fellow-tribesmen into squads. And if you cannot, do not take generosity away from our order, do not think favour a crime, do not sanction attentiveness with a penalty. And so, faltering in this tribal charge of the sodalicia, you have betaken yourself to the common case of canvassing --- in which let us at last, if you please, cease to contend with vulgar and well-worn declamation.

\ciceroSection{48} For thus I deal with you. Pick out one tribe, whichever is convenient to you; you show --- which you ought to do --- through what broker, by what distributor it was corrupted; I, if you cannot do this (which, as my opinion holds, you will not even begin to do), will show through whom he carried it. Is this not a true contest? Does it please you that the matter be handled thus? Can I set foot to foot, as they say, any more, or come any nearer? Why are you silent, why do you dissemble, why do you shuffle? Again and again I press and bear down, I pursue, I demand, and indeed I clamour for the charge. Whatever tribe, I say, you choose that Plancius carried, do you point out the flaw, if you can; I will show by what means he carried it. Nor will this method be one for Plancius and another for you, Laterensis. For as, of the tribes you carried, if I were now to ask it of you, you could set out by whose zeal you carried them, so I make this contest: that I, to you yourself as my adversary, will render the account of whatever tribe you demand.

\ciceroSection{49} But why do I deal thus? As though Plancius were not already, at the earlier assembly, designated aedile --- an assembly which the consul first began to hold, a man of the highest authority in all matters, and the very author of these laws on canvassing; and then began to hold suddenly, beyond all expectation, so that, even had anyone planned to bribe, no space was given for getting things ready. The tribes were called, the vote was cast, the ballots were counted out: Plancius prevailed by far the most; there was not, and could not be, any suspicion of bribery. Do you say so, then? One prerogative century has so much authority that no one ever carried it first without being declared either consul at those very assemblies or at any rate for that year; and do you wonder that Plancius was made aedile, in whom not a small part of the people but the whole people declared its will, in whose honour not the part of one tribe but the whole assembly was a prerogative to the assembly?

\ciceroSection{50} And at that time, Laterensis, if you had been willing to do it, or if you had thought it consonant with your gravity to do what many nobles have often done --- that, when they had been less strong in the votes than they had thought, then, the assembly being adjourned, they would throw themselves down and beseech the Roman people with broken and humbled spirit --- I do not doubt that the whole multitude would have turned itself toward you. For hardly ever has the nobility, especially when whole and innocent, been rejected as a suppliant by the Roman people. But if your gravity and your greatness of soul were worth more to you --- as they ought to have been --- than the aedileship, do not, when you have what you would rather have, hanker after what you reckoned of less worth. For my own part, I always laboured most of all, first to be worthy of office, secondly to be thought worthy of it; the third thing for me was what for most men is the first, the office itself, which ought at last to be pleasant to those men alone to whose dignity the Roman people gives its testimony, not a favour to their ambition.

\ciceroSection{51} You ask, too, Laterensis, what you are to answer to your ancestral images, what to a most distinguished and excellent man, your father, now dead. Do not brood on those things, and beware rather lest that complaint of yours and your excessive grief be rebuked by those wisest of men. For your father saw Appius Claudius, a most noble man, not made aedile while his own brother, a most powerful and illustrious citizen, C. Claudius, was alive, and the same man made consul without a rebuff; he saw a man most closely joined to himself, an outstanding man, L. Volcatius; he saw M. Piso, who, after taking such a little knock in that aedileship, attained the highest honours from the Roman people. And your grandfather indeed would tell you of P. Nasica's rebuff for the aedileship --- a citizen than whom I judge no one in this commonwealth braver --- and of C. Marius's, who, after suffering two rebuffs for the aedileship, was made consul seven times, and of L. Caesar's, of Cn. Octavius's, of M. Tullius's, all of whom we know to have been made consuls after being passed over for the aedileship.

\ciceroSection{52} But why do I gather up rebuffs for the aedileship? --- which have often been of such a kind that to those who were passed over it seemed a kindness done by the people. The military tribune L. Philippus, of the highest nobility and eloquence; the quaestor C. Caelius, a most illustrious and brave young man; the tribunes of the plebs P. Rutilius Rufus, C. Fimbria, C. Cassius, Cn. Orestes --- these were not elected, whom yet we know all to have been made consuls. These things your father and your forebears will say to you of their own accord, not for the sake of consoling you, nor indeed to free you from any blame --- which you fear may seem to have been taken on by you --- but to exhort you to hold to that course which you took up from your earliest age. For nothing, believe me, Laterensis, has been taken from you. ``Taken from,'' I say; if, by Hercules, you are willing to interpret truly what has happened, something has even been signified about your courage. For do not suppose that there was not some great stir over that candidacy of yours, on account of which you held off from swearing a certain oath. You announced, a young man, what you felt about the highest matters of the commonwealth --- more bravely indeed than not a few who have run their course of honours, but more openly than the reckoning either of your ambition or of your age demanded.

\ciceroSection{53} For this reason, in a divided people, do not suppose that there were none whose minds that bold spirit of yours offended; men who, while you were off your guard, could perhaps now dislodge you from your place, but who will surely never move you when you are looking ahead and on your watch. Or were you led by those arguments of his? ``Can you doubt,'' he says, ``that a coalition was formed, when Plancius carried most of the tribes together with Plotius?'' Could they have been carried together, then, unless they had carried the tribes together? But some he carried by almost the same number of votes. Of course --- since by then they had already come in all but elected and proclaimed at the earlier election. And yet not even that need have raised a suspicion of coalition. For our forefathers would never have established a lot-drawing for the aedileship had they not seen that it could happen that competitors should be equal in votes.

\ciceroSection{54} And you say that at the earlier election the Aniensis tribe was conceded by Plotius to Pedius, the Teretina by Plancius to you; but that now both were torn away from each, lest matters should come to a tight pass. How does it square that, before the people's will was yet known, these men whom you say were even then in league should have made a sacrifice of their own tribes, by which you were to be helped --- and that the same men, once they had already tested what they could do, should have turned grasping and tight-fisted? For they were afraid, I suppose, of a narrow result --- as though the thing could come into any contest or any close decision! And yet, by calling A. Plotius, a man of the highest distinction, into the same charge, you give it away that you have laid hold of a man by whom you were not interrogated. For when you complained that you have more witnesses about the Voltinia tribe than the votes you carried in it, you give it away either that you are producing as witnesses men who passed you over because they had taken bribes, or that you did not carry even their votes for free.

\ciceroSection{55} But that charge about the money said to have been seized in the Circus Flaminius was hot while the matter was fresh; now in the case it has gone cold. For you point out neither whose that money was, nor what tribe, nor what distributor. And in fact the man who was then being called into the charge, when he had been brought before the consuls, complained bitterly that he had been unfairly bandied about by your people. If he was a distributor --- and especially the distributor of the man you had on trial --- why was he not made a defendant by you? Why did you not, by his condemnation, secure some prejudgement toward this trial? But you neither bring these things forward nor trust in them; another reckoning, another scheme has roused you to the hope of crushing this man. Great are your resources; your influence reaches wide; you have many friends, many men devoted to you, many supporters of your renown. Many envy this man; to many his father, an excellent man, seems to cling too tightly to the rights and liberty of the equestrian order; many too are the common enemies of all defendants, who always give testimony about bribery as though either they sway the minds of the jurors by their witness, or it is pleasing to the Roman people, or they will more easily on that account attain the standing they desire.

\ciceroSection{56} You will not see me, gentlemen, fighting these men in my old manner --- not that it would be right for me to shirk anything that the safety of Plancius requires, but because there is no need for me to pursue with my voice what you see with your minds; and because those very men whom I see ready to testify have deserved so well of me that you ought to take their reproach upon your own prudence and to release my modesty from it. This one thing I beg and entreat of you with all my strength, gentlemen, both for the sake of the man I defend and for the sake of the common danger: do not think that the fortunes of the innocent should be subjected to invented hearsay, to talk scattered and sown abroad.

\ciceroSection{57} Many friends of the prosecutor, some men hostile to us too, many common detractors and enviers of all men, have invented much. And nothing is so swift as slander; nothing is more easily sent out, nothing more quickly taken up, more widely spread. Nor will I ever demand, if you find the source of the slander, that you overlook it or pass it by. But if anything flows without a head, or if anything is of such a kind that no author for it stands forth --- and whoever heard it will seem to you either so careless that he has forgotten where he heard it, or to hold his author so worthless that he did not think him worth remembering --- of that common saying of his, ``I heard it,'' we beg that it do no harm to an innocent defendant.

\ciceroSection{58} But I come now to L. Cassius, my good friend, out of whose speech I have not even taken to task that Iuventius of yours, whom that young man, graced with every humanity and virtue, said to have been the first man from the plebs made curule aedile. On which point, Cassius, if I should answer you thus --- that the Roman people did not know it, and there was no one to tell us of it, especially now that Congus is dead --- you would not, I think, be surprised, since I myself, no stranger to the study of antiquity, confess that I heard it here for the first time from you. And since your speech was most elegant and most subtle, worthy both of the zeal and of the modesty of a Roman knight, and since you were so heard by these men that great honour was paid both to your talent and to your humanity, I will answer the things you said, most of which were about myself --- in which the stings themselves, if you had any in reproaching me, fell upon me not unwelcome.

\ciceroSection{59} You asked whether I thought the road to attaining office had been easier for me, the son of a Roman knight, or whether it would be easier for my son, because his family would be of consular rank. Now though I wish him everything rather than myself, yet I never wished his approaches to office easier than mine were. Nay more, lest perhaps he should think I have gained him offices already rather than shown him the way to attain them, I am wont to teach him these lines --- though his age is not yet weighed down for precepts --- which that king sprung from Jove gave to his own sons: ``One must keep watch always; many are the snares for good men.'' What many envy --- you know the rest. Did not that grave and gifted poet write what he wrote, not to rouse those royal boys, who were now nowhere, but to rouse us and our children to labour and to praise?

\ciceroSection{60} You ask what more Plancius could have attained had he been the son of Cn. Scipio. He could not have been made aedile any more than he was; but he would have this advantage --- that less envy would be felt against him. For the steps of office are equal for the highest men and the lowest; the steps of glory are unequal. Which of us calls himself the equal of M'. Curius, of C. Fabricius, of C. Duellius? Which of A. Atilius Calatinus, which of Cn. and P. Scipio, which of Africanus, of Marcellus, of Maximus? Yet we have attained the same steps of office that they attained. For in virtue there are many ascents, so that he most excels in glory who most surpasses in virtue; but the people's bound of office is the consulship, a magistracy which by now nearly eight hundred men have attained. Of these, if you search diligently, you will scarcely find a tenth part worthy of glory. But no one ever argued as you do: ``Why is this man made consul? What more could he have attained, were he L. Brutus, who freed the state from the domination of kings?'' In office, nothing more; in praise, much. So then Plancius was made quaestor, and tribune of the plebs, and aedile no less than if he had been born in the highest station --- but these honours countless others, born in his own station, have attained.

\ciceroSection{61} You bring forward the triumphs of T. Didius and C. Marius, and ask what is comparable in Plancius --- as though those men whom you mention took up their magistracies because they had triumphed, and did not triumph because magistracies were committed to them in which, the matter being well managed, they could triumph. You ask what camp he has seen --- the man who was a soldier in Crete under this same commander, and a tribune of the soldiers in Macedonia, and who as quaestor withdrew from his military service only as much time as he chose to transfer to guarding me. You ask whether he is eloquent.

\ciceroSection{62} No --- and, what is the next best thing, he does not even seem so to himself. Whether he is a jurist? As if there were anyone to say that this man gave him a false answer about the law! For all arts of that sort are reproached in those who, having professed them, cannot give satisfaction --- not in those who confess that they have stood apart from such studies. Virtue, probity, integrity in a candidate are what is looked for, not glibness of tongue, not art, not learning. Just as, in buying slaves, however thrifty a man we may buy as a smith or a plasterer, we are wont to take it ill if he happens to be ignorant of the arts we sought in buying him; but if we have bought one to set over the farm, to put in charge of the flock, we care for nothing in him but thrift, labour, watchfulness --- so the Roman people chooses its magistrates as it were the stewards of the commonwealth: in whom, if there is besides some art, it readily allows it, but if not, it is content with their virtue and innocence. For how few are eloquent, how few are skilled in law, even if you count those who wish to be? But if besides these there is no one worthy of office, what then is to become of so many excellent and most accomplished citizens?

\ciceroSection{63} You bid Plancius speak of the faults of Laterensis. He can say nothing, except that he thought you too quick to anger against yourself. The same man you exalt Laterensis with praises. I readily allow you to do at length what does not pertain to the trial, and, while prosecuting, to dwell so long on what I as defender could without danger confess. And indeed I not only confess that there are the highest distinctions in Laterensis, but I even reproach you because you do not enumerate them, and instead hunt up certain other things, empty and trifling. That he gave games at Praeneste --- what then? Did not other quaestors do so? That at Cyrene he was generous toward the tax-farmers and just toward the allies --- who denies it? But so many things are done at Rome that what happens in the provinces is scarcely heard.

\ciceroSection{64} I do not fear, gentlemen, that I shall seem to arrogate anything to myself if I speak of my own quaestorship. For however much it flourished, yet I judge that I was afterward in such great commands that I need not seek much of my glory back from the praise of a quaestorship. But even so I do not fear that anyone will dare to say that any quaestorship in Sicily was either more illustrious or more welcome than mine. By Hercules, I will say this truly: at that time I thought that men at Rome talked of nothing else but my quaestorship. In the highest dearness of grain I had sent the greatest quantity; courteous to the businessmen, just to the merchants, generous to the contractors, abstinent toward the allies, I had seemed to all most diligent in every duty; and certain honours, unheard of, had been devised for me by the Sicilians.

\ciceroSection{65} And so I was leaving with this hope --- that I supposed the Roman people would of its own accord heap everything upon me. But when, leaving the province for the sake of making my journey, I had chanced to come to Puteoli in those days, when very many men, and the most fashionable, are wont to be in those parts, I nearly collapsed, gentlemen, when a certain man asked me on what day I had left Rome and whether there was any news. When I had answered him that I was leaving my province: ``By Hercules, yes,'' he said, ``from Africa, I think.'' To him, now nettled and disdainful: ``No --- from Sicily,'' I said. Then a certain man, like one who knew everything: ``What? Don't you know,'' he said, ``that this man was quaestor at Syracuse?'' Why say more? I gave up being nettled and made myself one of those who had come to the waters.

\ciceroSection{66} But that affair, gentlemen, I rather think profited me more than if all men had then congratulated me. For after I perceived that the ears of the Roman people were rather dull, but its eyes keen and sharp, I gave up thinking about what men would hear of me; I saw to it that thereafter they should see me present every day, I dwelt before their eyes, I pressed the Forum; neither my doorkeeper nor sleep frightened anyone away from meeting me. Why should I say anything of my crowded hours --- I to whom not even leisure was ever leisurely? For those speeches which you say, Cassius, you are wont to read when you are at leisure, I wrote at the games and on holidays, that I might never be wholly at leisure. For that saying of M. Cato, which he wrote at the beginning of his Origins, I have always thought splendid and noble: that of famous and great men an account no less of their leisure than of their business ought to stand on record. And so, if I have any praise --- and how great it is I do not know --- it was got at Rome, sought in the Forum; and my private counsels public misfortunes too have approved, so that even the highest affairs of state had to be conducted by me at home, and a city had to be saved within the city.

\ciceroSection{67} The same road, then, Cassius, is built for Laterensis; the same course for virtue toward glory --- this much the easier, perhaps, that I climbed up here sprung from myself and leaning upon myself, while his outstanding virtue will be helped by the recommendation of his ancestors. But to return to Plancius: he was never absent from the city except by lot, by law, by necessity; he did not prevail by the same things by which perhaps some men do, but he prevailed by constant attendance, he prevailed by attentiveness to his friends, he prevailed by generosity; he was before men's eyes, he canvassed, he used that manner of life by which very many new men have attained the same honours with the least envy.

\ciceroSection{68} For as to what you say, Cassius --- that I owe no more to Plancius than to all good men, because my safety was equally dear to them --- I confess that I owe a debt to all good men. But even those to whom I owe it, good men and citizens, said at the aedilician election that they on my account owed something to Plancius. But suppose that I do owe a debt to many, and to Plancius among them: ought I then to throw all my accounts into confusion, or to discharge to the rest, when each man's day has come, this debt which now, while it is demanded, presses me? And yet the owing of money and the owing of gratitude are unlike. For he who pays money at once no longer has what he has returned; but he who owes it keeps what is another's. Gratitude, on the other hand, both he who returns has, and he who has it returns by the very fact that he has it. Nor will I now cease to owe a debt to Plancius if I discharge this one; nor would I render it him any less by my will itself, had this trouble not befallen him.

\ciceroSection{69} You ask of me, Cassius, what more I could do for my brother, who is dearest to me, what more for my children, than nothing to me can be more delightful, than what I do for Plancius --- and you do not see that by my very love for these of mine I am most of all spurred and roused to defend this man's safety. For to them nothing is more wished than the safety of the man by whom they know my own was defended; and I myself never look upon them but, remembering that I was preserved for them through this man, I recall his service to me. You bring up that Opimius was condemned, the very preserver of the commonwealth; you add Calidius, by whose law Q. Metellus was restored to citizenship; you reproach my entreaties for Plancius, because neither was Opimius freed in his own name nor Calidius in Metellus's. About Calidius I answer you only this, which I myself saw: that Q. Metellus Pius, when consul, made supplication to the Roman people at the praetorian election when Q. Calidius was a candidate --- and that, consul and most noble man as he was, he did not hesitate to call that man the patron of himself and of his most noble house.

\ciceroSection{70} On which point I ask of you whether you think that in Calidius's trial Metellus Pius, had he been able to be at Rome, or his father, had he lived, would not have done what I do in the trial of Plancius. For as to the disaster of Opimius --- would that it could be torn out of the memory of men! That was a wound to the commonwealth, a disgrace to this empire, a shame to the Roman people --- it is not to be thought a verdict. For what heavier axe could those jurors --- if jurors and not parricides of their fatherland they are to be named --- have driven into the commonwealth than when they cast out of the state the man who as praetor had freed the commonwealth from a war on its borders, as consul from a war at home?

\ciceroSection{71} But, you say, I make Plancius's service too great a thing, and, as you put it, I magnify it in words. As though I ought to be grateful by your judgement and not my own! ``What was his service so great?'' he says. ``Was it that he did not cut your throat?'' No, rather that he did not suffer my throat to be cut. And on this point, Cassius, you even cleared my enemies, and said that there were no plots laid against my life by them. Laterensis advanced the same thing. For which reason I will say more of that man a little later; of you I only ask this --- whether you think the hatred of my enemies against me was moderate (what hatred of any barbarians whatsoever was ever so monstrous and so cruel against an enemy?), or whether there was in them any fear of reputation or of punishment, in those whose steel you saw, throughout that whole year, in the Forum, whose fire in the shrines, whose violence ranging through the whole city. Unless perhaps you think that they spared my life because they feared nothing about my return. And do you think there was anyone so witless that, while these men lived and the city and the Senate-house stood, he did not think I would return, if I lived? For which reason you ought not, being the man and the citizen you are, to proclaim that my life was preserved by the loyalty of friends and not assailed by the restraint of my enemies.

\ciceroSection{72} I will now answer you, Laterensis, less vehemently perhaps than I was provoked by you, but surely no less considerately and no less in friendship. For there was, first, that rather harsh charge --- that I was lying and inventing for the occasion in what I said of Plancius. Of course I, a wise man, contrived a reason to seem bound by the greatest bonds of service, when I was free and unfettered! What then? Were the ties of intimacy, of neighbourhood, of my father's friendship too few, too unjust, for me to defend Plancius? Had these not existed, I should be afraid, I suppose, of acting basely if I defended a man of this splendour and this dignity. Doubtless I had to invent a most ingenious reason, to say that I owed everything to a man whom I ought rather to owe a debt to. But even common soldiers do this against their will --- give the civic crown and confess that they were saved by someone --- not because it is base to be snatched, fallen in the line, from the hands of the enemy (for that cannot befall any but a brave man fighting hand to hand), but because they shrink from the burden of the service, since it is a very great thing to owe to a stranger the same that you owe to a parent.

\ciceroSection{73} Do I, when others dissemble even lesser services that are real, lest they should seem to be under obligation --- do I lyingly claim to be bound by a service for which gratitude cannot even seem capable of being returned? Or are you, Laterensis, ignorant of this? You who, being most friendly to me, when you had wished even to share the danger of your life with me, when you had attended me in that grim and bitter grief and departure not with your tears alone but with your spirit, your body, your means, when you had defended my children and my wife in my absence with your resources and your aid --- so always dealt with me, that you remitted and conceded to me to spend all my zeal upon the honour of Cn. Plancius, because you said that his service to me was welcome even to you yourself.

\ciceroSection{74} But that I am saying nothing new, nothing for the occasion --- is there not also that speech as witness which was the first delivered by me in the Senate? In which, when I had named very few in giving thanks --- because all could in no way be enumerated, and yet it would be a crime that anyone be passed over, and I had resolved to name only those who had been the leaders and, as it were, the standard-bearers of our cause --- among these I gave thanks to Plancius. Let the speech be read out, which on account of the greatness of the matter was delivered from a written text: in which I, a cunning man, was surrendering myself to one to whom I owed nothing of any consequence, and was binding fast the servitude of so great a service by an everlasting testimony. I do not wish to have read out the rest of what was set down by me in writing; I pass it by, lest I should seem either to bring it forward for the occasion, or to use a kind of literature that may seem better suited to my own studies than to the custom of the courts.

\ciceroSection{75} And you even cry out, Laterensis: ``How long do you keep saying these things? You made no headway in the case of Cispius; your entreaties are worn out by now.'' Will you then throw Cispius in my teeth --- a man well deserving of me, whom, since I had come to know him with you for witness, I defended with you for that very prompter? And will you say ``how long?'' to a man whom you say could not, for all my contending for Cispius, gain his cause? For that word ``how long?'' could carry this odium: ``He was granted you, that man was pardoned at your request; you make no end; we cannot bear it.'' But to say ``how long?'' to a man who has laboured for one and has not even gained that very thing --- it is the part of a mocker rather than of one finding fault; unless perhaps I alone have so conducted myself in the courts, have so lived both with these men and among these men, am and always have been such a patron in causes, such a citizen in the commonwealth, that I alone am to be set up by you as one who may never gain anything from the jurors.

\ciceroSection{76} And you cast in my teeth that little tear at the trial of Cispius. For thus you said: ``I saw that little tear of yours.'' See how I repent of your word. Not a little tear only, but many tears and weeping with sobbing you could have seen. Was I, who, moved by the tears of my own people in my absence, should have laid aside the grudges he had against me, and should have been not only no assailant of my safety, as my enemies had thought, but even its defender --- was I, in this man's peril, not to show my grief?

\ciceroSection{77} But you, Laterensis, who then said that my tears were welcome, now wish those same tears to seem invidious. You deny that the tribunate of Plancius brought any help to my standing, and at this point --- which you can do most truly --- you recall the godlike services of L. Racilius, a most brave and most steadfast man, toward me. To him indeed, as to Cn. Plancius, I have never dissembled that I owe a very great debt, and always will avow it; for he thought that no contentions, no enmities, no risks of life were to be shunned by him, neither for the commonwealth nor for me. And would that, as grateful as I am toward him, the Roman people had likewise been allowed --- through the violence and wrong of men --- to return its gratitude to him! But if Plancius did not strive for the same ends in his tribunate, you ought to judge that it was not the will that was wanting in him, but that I, since I already owed so much to Plancius, was content with the services of Racilius.

\ciceroSection{78} Or do you really think that the jurors will act the less willingly for my sake because you charge me with being grateful? Or, when the senators by that decree of the Senate which was made in the monument of Marius --- by which my safety was commended to all nations --- gave thanks to Cn. Plancius alone (for he alone of the magistrates was the defender of my safety), shall I not think gratitude must be returned by me to the man to whom the Senate thought thanks must be given for me? And when you see this, with what disposition toward you do you suppose me to be, Laterensis? That there is any danger so great, any labour so great, any struggle so great, that I would shun it not only for your safety but even for your standing? On which account I am the more --- I will not say wretched, for that word is foreign to virtue, but surely hard-pressed: not because I owe a debt to many (for the burden of a service is light, the burden of gratitude), but because the accounts often fall due together, on account of the contentions among themselves of certain men well deserving of me, so that at one and the same time I fear I may scarcely be able to seem grateful to all.

\ciceroSection{79} But I shall weigh these things by my own measures: not only what I owe to each man, but also what is in each man's interest, and what each man's circumstance demands of me. At stake is your zeal, or, if you like, your reputation, the credit of your aedileship; but at stake for Cn. Plancius are his life, his country, his fortunes. You wished me safe; this man brought it about that I could even be alive. And yet I am pulled apart and torn by grief, and in a case so unequal I am pained that I have offended you. But, as the God of Faith is my witness, I will far sooner throw away my own safety for your sake than hand over the safety of Cn. Plancius to your contest.

\ciceroSection{80} For indeed, gentlemen, while I wish to be furnished with every virtue, there is nothing I would rather have than to be grateful and to be seen as grateful. For this is the one virtue which is not only the greatest but also the mother of all the rest. What is piety, but a will grateful toward our parents? Who are good citizens, who are those who deserve well of their country in war and at home, but those who remember their country's kindnesses? Who are holy, who are observers of religion, but those who pay back to the immortal gods the gratitude they have earned, with just honours and a mindful heart? What sweetness can life have, if friendships are taken away? And what friendship can there be among the ungrateful?

\ciceroSection{81} Who among us has been brought up as a free man, in whose mind there do not turn, with grateful recollection, the men who reared him, his teachers and instructors, and the very place, mute though it is, where he was nourished or taught? Whose resources can be so great, or ever have been, as to stand without the services of many friends? And those resources can surely not survive at all if memory and gratitude are taken away. For my part I think nothing so proper to a human being as to be bound not only by a kindness but by the very sign of goodwill; and, on the other hand, nothing so inhuman, so monstrous, so savage, as to bring it about that you seem not, I will not say unworthy of a kindness, but defeated by it.

\ciceroSection{82} Since this is so, I will now yield, Laterensis, to that charge of yours, and in the very thing in which there can be no excess, since you so wish it, I will grant that I am excessively grateful; and I will ask of you, gentlemen, to embrace with your kindness a man whom anyone who reproaches reproaches for being grateful beyond measure. Nor ought that to count toward making light of my gratitude which the same man said, that you are neither guilty men nor litigious, as if therefore my influence ought to weigh less with you. As though, in fact, I had not always preferred that these protections of mine, if there are any in me, should be ready for my friends rather than necessary to them! For of myself I dare say only this much: that my friendship has been a pleasure to more men than a defence; and I would be heartily ashamed of my life, if in my intimacy there were a place for no one save the litigious or the guilty.

\ciceroSection{83} But somehow you heaped these things upon me again and again, and you were truly insistent on this point: that you had not wished to throw the case into the games-season for this reason, lest I, after my custom, should say something about the sacred cars, for pity's sake, as I had done before in the case of other aediles. You accomplished something here; for you have stripped me of an ornament of my speech. I shall be laughed at if I make mention of the cars, now that you have foretold it; but without the cars, what shall I be able to say? Here too you added that I had, by my own law, sanctioned bribery with exile, so that I might be able to deliver more pitiable perorations. Does it not seem to you that the man is disputing with some declaimer, not with a disciple of toil and of the Forum?

\ciceroSection{84} ``For at Rhodes,'' he says, ``I was not'' --- he wishes I had been --- ``but I was,'' he says (I thought he was going to say among the Vaccaei) ``twice in Bithynia.'' If the place gives any handle for reproach, I do not see why you should think Nicaea sterner than Rhodes; if the cause is to be looked at, you too were in Bithynia with the highest dignity, and I at Rhodes with no less. For as to your reproaching me with defending too many men: would that you, who can, and the rest, who shirk it, were willing to relieve me of this labour! But it comes about, through your fastidiousness --- you who, in weighing cases, reject nearly all of them --- that most of them flow together to us, who can deny nothing to the wretched and the struggling.

\ciceroSection{85} You reminded me too that, since you had been in Crete, some witticism could have been spoken against your candidacy, and that I had let it slip. Which of us, then, is the more eager for a quip? I, who did not say what could have been said, or you, who have even said something against yourself? You said that you had sent no dispatches about your own exploits, because my own letters, which I had sent to someone, had done me harm. I do not grasp how they harmed me; I see they could have profited the commonwealth.

\ciceroSection{86} But these are lighter matters; those, however, are grave and great --- that my departure, which you had often wept over, you now wished, as it were, to reproach and lay something to my charge. For you said that not aid had failed me, but I had failed the aid. I confess indeed that, because I saw that aid did not fail me, for that reason I spared that aid. For who does not know what the state, what the crisis, what the storm in the commonwealth then was? Was it the terror of a tribune or the madness of a consul that moved me? Was it a great thing for me to fight it out with the sword against the remnants of those whom, while they flourished and were whole, I had conquered without the sword? The consuls, the most foul and base since the memory of man --- as both those beginnings and these recent ends of affairs have declared, of whom the one lost his army and the other sold it --- bought with their provinces, had deserted the Senate, the commonwealth, all good men; men who in army, in arms, in resources had the greatest power, while what they felt was not known. That frenzied voice, made effeminate by unspeakable debaucheries at sacred altars, kept ringing out most bitterly that both they and the consuls were of one mind with it; the destitute were being armed against the wealthy, the ruined against the good, slaves against their masters.

\ciceroSection{87} But the Senate was with me, and indeed in changed dress --- a thing undertaken by public counsel for me alone since the memory of man. Yet recall who were then the enemies under the name of consuls, who alone in this city did not allow the Senate to obey the Senate, and by their edict took from the senators not their mourning but the tokens of mourning. The whole order of knights was with me; yet that dancer, Catiline's consul, kept terrifying them in the public meetings with the threat of proscription. All Italy had gathered; yet upon it the fear of civil war and devastation was being thrust. These supports, eager and roused, I confess, Laterensis, that I could have used; but the struggle was not to be fought out by right, by laws, by debate --- for surely, especially in so good a cause, my own aid would never have failed me, that aid in which others have often abounded --- it had to be fought out by arms, by arms, I say; and for slaughter to be done by these against slaves and the leaders of slaves would have been ruinous to the Senate and the good men of the commonwealth.

\ciceroSection{88} That the wicked should be conquered by the good was, I confess, a glorious thing, if I had then seen any end to the conquering; but that I surely did not see. For where would there have been ready to my hand consuls as brave as L. Opimius, as C. Marius, as L. Flaccus, under whose leadership the commonwealth conquered wicked citizens by arms; or, if not as brave, then at least as just as P. Mucius, who defended that the arms which Scipio as a private man had taken up had been taken up, when Ti. Gracchus was slain, by the best of right? I would have had, then, to fight against the consuls. I say nothing more than this: I saw that grave adversaries stood ready for our victory, but no avengers for our destruction.

\ciceroSection{89} If I failed these supports of my safety because I was unwilling to fight, I will confess what you wish --- that not aid failed me, but I failed the aid; but if, the greater the zeal of good men was toward me, the more I thought I must take counsel for them and spare them, do you reproach in me what was reckoned to Q. Metellus's praise, and is so reckoned today, and always will be to his greatest glory? It is agreed --- as you can hear from many who were then present --- that he yielded against the will of good men, and that there was no doubt that he could have come off the superior in struggle and arms. So when he was defending his own act, not the Senate's, when he was holding to the persistence of his own opinion, not the safety of the commonwealth, yet for that very constancy, because he took that voluntary wound, he surpassed in glory and praise the most just and most illustrious triumphs of all the Metelli --- because both he did not wish those most wicked citizens themselves to be killed, and he saw to it that no good man should perish in the same slaughter. And I, with such great perils set before me --- when, if I were conquered, the destruction of the commonwealth, and if I conquered, an endless struggle was being prepared --- should I have brought it about that I be named the destroyer of the commonwealth, I who had been its saviour?

\ciceroSection{90} You say that I feared death. I, for my part, would not think even immortality should be accepted against the commonwealth, far less would I wish to die with the ruin of the commonwealth. For as to those who have given up their life for the commonwealth --- though you may call me out of my wits --- I have never, by Hercules, thought that they attained death rather than immortality. But if I had then fallen by the sword and hand of those impious men, the commonwealth would have lost forever a citizen defender of its own safety. Nay more, even if some force of disease or nature itself had consumed me, still the supports of posterity would have been diminished, because the example would have perished by my death of what kind the Senate and the Roman people would have been in restoring me. Or, if there had ever been a lust for life in me, would I, in the December of my consulship, have stirred up the weapons of all the parricides? Which, if I had been quiet for twenty days, would have fallen back into the watch of other consuls. Wherefore, if a lust for life against the commonwealth is base, surely my lust for death would have been far more base, accompanied by the ruin of the state.

\ciceroSection{91} As for your boast that you are a free man in public life, that I both confess and rejoice at, and I even congratulate you on it; but as for your denying it of me, in that I will not allow either you or anyone else to err any longer. For if anyone thinks that something has been taken from my freedom because I do not dissent from all the same men from whom I was wont to dissent before --- first, if I show myself grateful to those who have deserved well of me, do I not cease to fall under the charge of being a man too mindful and too grateful? But if at some time, without any harm to the commonwealth, I look also to my own safety and that of my own, surely I am not only not to be reproached, but, if I wished to rush to ruin, good men would beg me not to do it.

\ciceroSection{92} Indeed the commonwealth itself, if it could speak, would plead with me thus: that, since I had always served it, never myself, and yet had reaped from it fruits not glad and abundant, as ought to have been, but mingled with great bitterness, I should now serve myself and take counsel for my own; that it not only has enough from me, but even fears it may have repaid me too little for all that it had from me.

\ciceroSection{93} What then? If I think none of these things, and am in public life the same man I have always been, will you still demand my freedom? Which you place in this: that I should always fight to the death with those with whom I have at some time contended. But it is far otherwise. For we all ought to stand as it were within some sphere of the commonwealth, which, since it revolves, we should choose the part toward which its advantage and safety has turned us. I, for my part, do not call Cn. Pompey the author, the leader, the defender of my safety --- for these things perhaps look, in a private way, for the memory of services and for gratitude --- but I say this, which pertains to the safety of the commonwealth: should I not protect the man whom all concede to be the leading man in the state? Should I fail the praises of C. Caesar, which I see celebrated, first by the very many and most ample judgements of the Roman people, now of the Senate too, to which I have always bound myself? Then, by Hercules, I would confess that I had not held any judgement about the advantage of the commonwealth, but had been a friend or an enemy to men.

\ciceroSection{94} Or, when I see a ship holding her course under favourable winds, if she does not make for that harbour which I once approved, but for another no less safe and calm, should I fight with the storm at peril, rather than obey and yield to her, especially when safety is set before me? I, for my part, have learned these things, have seen these things, have read these things written; these things the records and the writings have handed down to us concerning the wisest and most illustrious men, both in this commonwealth and in other states: that the same opinions were not always defended by the same men, but whatever the condition of the commonwealth, the inclination of the times, the principle of concord demanded. This I both do, Laterensis, and always will do; and the freedom which you demand in me, which I have neither ever let go nor will let go, I will reckon to lie not in obstinacy but in a certain moderation.

\ciceroSection{95} Now I come to that last point, in which you said that, while I was extolling Plancius's service to me in words, I was making a citadel out of a sewer and worshipping a stone from a tomb as a god; for, you said, there had been for me no danger of treachery and no danger of death. I will explain the reckoning of that time briefly, and not unwillingly. For there is nothing in my own history less widely known, and less either made famous by my own recounting or heard of and known by men. For I, Laterensis, withdrawing from that conflagration of the laws, of right, of the Senate, of all good men --- when my own house, by its own blaze, was threatening a burning-up to the city and to all Italy unless I kept quiet --- made in my mind for Sicily, which both was itself joined to me as a single household, and was held by C. Vergilius, with whom alone, beyond almost anyone, both length of time and friendship, both my brother's colleagueship and the cause of the commonwealth, had allied me.

\ciceroSection{96} See now the darkness of those times. When the island itself was almost willing to come to meet me, that praetor --- often harassed in the meetings of the same tribune of the plebs on account of the same cause of the commonwealth --- I say nothing more than this: did not wish me to come into Sicily. What shall I say? That C. Vergilius, so good a citizen and man, lacked goodwill toward me, lacked memory of our common times, lacked piety, humanity, faith? None of those things is true, gentlemen; but he feared, that storm which we had not borne together with you, that he could not himself sustain it by his own resources. Then, my plan suddenly changed, I made haste to seek Brundisium by land; for the greatness of winter was barring the sea routes.

\ciceroSection{97} Since all those towns which lie from Vibo to Brundisium were under my protection, gentlemen, they secured me a safe journey, though many were threatening, at great fear to themselves. I came to Brundisium, or rather I approached its walls; the one city most friendly to me I turned aside from, which would more easily have suffered itself to be cut to pieces than that I should be snatched from its embrace. I betook myself to the gardens of M. Laenius Flaccus. Although every fear was set before him --- the confiscation of his goods, exile, death --- he preferred to endure these, if they should come to pass, than to give up the guarding of my life. By his hands, and those of his father, a most prudent and excellent old man, and of his brother, and of the sons of both, I was settled in a safe and faithful ship; and, hearing their prayers and vows for my return, I made haste to seek Dyrrachium, which was under my protection.

\ciceroSection{98} When I had come there, I learned --- what I had heard --- that Greece was crammed with the most criminal and unspeakable men, from whose hands that consulship of mine had wrung their impious sword and their pestilent fires; who, before they could hear of my arrival, while they were several days' journey from me, I pressed on into Macedonia and to Plancius. But this man, as soon as he learned that I had crossed the sea --- hear, hear and attend, Laterensis, that you may know what I owe to Plancius, and may at last confess that what I am doing I do both gratefully and dutifully; and that to this man, the things which he did for my safety, even if they are to do him little good, ought certainly not to do him harm! --- for as soon as he heard that I had reached Dyrrachium, at once, his lictors dismissed, his insignia thrown aside, his dress changed, he set out to me.

\ciceroSection{99} O bitter to me, gentlemen, the memory of that time and place, when this man fell upon me, when he embraced me and bedewed me with his tears and could not speak for grief! O thing cruel in the hearing and unspeakable in the sight! O all those remaining days, and those nights, in which this man, not leaving my side, led me to Thessalonica and into the quaestor's residence! Here I shall now say nothing more of the praetor of Macedonia than that he was always both an excellent citizen and a friend to me, but feared the same things as the rest; that Cn. Plancius was the one man, not who feared less, but who, if those things which were feared should come to pass, wished to undergo and endure them with me.

\ciceroSection{100} When L. Tubero, my kinsman, who had been a legate to my brother, came to me on his way back from Asia, and reported to me, with the most friendly spirit, that treachery which he had heard was being prepared for me by the conspiring exiles, I was making ready to go into Asia on account of the connection of that province with me and with my brother; he did not allow it. By force, by force, I say, Plancius held me back in his embrace, and for many months did not leave my side, the quaestor's part thrown off and the comrade's taken up.

\ciceroSection{101} O your wretched watches, Cn. Plancius, O your tearful vigils, O your bitter nights, O the unhappy guarding of my life! --- if indeed I, living, am of no use to you, who, perhaps, dead would have been of use. For I remember, I remember and shall never forget, that night when to you, watching, sitting beside me, grieving, I, wretched and led on by a false hope, kept promising certain vain and empty things: that I, if I were restored to my country, would render you thanks in person; but if either chance had taken my life from me, or some greater force had cut off my return, that these men --- these men, for what others did I then behold in my mind? --- would pay you in full, for me, all the rewards of those labours of yours. Why do you look at me, why do you ask again for my promises, why do you implore my faith? I promised you then nothing of my own resources, but I pledged the goodwill of these men toward me; these men I saw wished to mourn for me, to groan, to fight to the death for my life, even at peril of their own; of their longing, their grief, their complaints I heard something each day together with you; now I fear that I can render you nothing but tears, which you poured out in the greatest abundance amid my bitter sufferings.

\ciceroSection{102} For what else can I do but mourn, but weep, but embrace you together with my own safety? Safety can be given you by the same men who gave it back to me. You, however --- rise up, I beg! --- I will hold and embrace, and I will profess myself not only the pleader for your fortunes but their companion and partner; and, as I hope, no one will be of so cruel and inhuman a spirit, nor so unmindful, I will not say of my own services to good men, but of the services of good men to me, as to tear and wrench from me the saviour of my life. I do not beg of you, gentlemen, for a man adorned with kindnesses of mine, but for the guardian of my safety; I contend not by resources, not by authority, not by influence, but by prayers, by tears, by appeal to your pity; and together with me this most wretched and excellent father here implores you, and for one son two fathers plead.

\ciceroSection{103} Do not, gentlemen --- by yourselves, by your fortunes, by your children --- give to my enemies, and to those especially whom I have taken upon myself for your safety, the joy of boasting that you, now forgetful of me, have shown yourselves enemies of the safety of the man by whom my safety was preserved; do not weaken my spirit, both with grief and also with fear of a change in your goodwill toward me; allow me, what, relying on you, I have often promised to this man, to pay it to him from you.

\ciceroSection{104} And you, C. Flavus --- I beg and implore you, who were the partner of my counsels in my consulship, the sharer of my dangers, the helper in the things I did, and who wished me not only always safe but also honoured and flourishing --- preserve for me, through these men, the man through whom you see me preserved for you and for these. My saying more is hindered even by your tears, and yours, gentlemen, not only my own; by which, in the midst of my great fear, I am suddenly led into the hope that you will be the same in preserving this man as you were in preserving me, since by those tears of yours I am reminded of those which you have often and abundantly shed for me.
