% Pro C. Rabirio Postumo (pro-rabirio-postumo) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} If there is anyone, gentlemen of the jury, who thinks C. Rabirius deserves rebuke because he entrusted his fortunes --- his resources especially, well-founded and most excellently established --- to a king's power and caprice, let him add to his own verdict the judgement not only of mine but of the very man who made the venture; for to no one does that plan of his displease more sharply than to himself. And yet for the most part we do this: we weigh plans by their outcomes, and of the man for whom something has turned out well we say that he foresaw much, of the man for whom it went otherwise that he perceived nothing. Had good faith stood firm in the king, nothing would be called wiser than Postumus; because the king betrayed him, nothing is called more insane than this --- so that now it seems there is nothing in a wise man but the gift of divination.

\ciceroSection{2} But still, if there is anyone, gentlemen, who thinks that hope of Postumus's --- whether empty hope, or ill-considered calculation, or, to use the gravest word, temerity --- deserves blame, I do not fight against his opinion; this one thing, however, I beg: that, since he sees this man's plans punished most cruelly by Fortune herself, he should not think any further bitterness must be added to the ruins under which this man has been crushed. It is enough not to raise up men who have stumbled through imprudence; but to press the fallen, or to push along those already plunging headlong, is surely inhuman --- especially, gentlemen, since this is given almost by nature to mankind: that if in any family some praise has by chance flowered, those who are of that stock pursue it most eagerly, because virtues are celebrated in the talk of men and the memory of fathers; if indeed not only in the glory of military affairs did Scipio and Maximus, the sons, imitate Paulus, but even in the devoting of his life and in the very manner of his death the son imitated P. Decius. Let small things, then, be like great, gentlemen.

\ciceroSection{3} For when we were boys, this man's father, C. Curtius, was the leading man of the equestrian order, the bravest and greatest of tax-farmers; and men would not so have approved his greatness of spirit in the conduct of his business, had there not been in the same man an incredible kindness, so that in increasing his estate he seemed to seek not the spoils of avarice but the instrument of his goodness.

\ciceroSection{4} Born of this man --- although he had never seen his own father --- he was nonetheless drawn into the likeness of his father's discipline, both with Nature herself as guide, who has the greatest force, and by the constant talk of the household. He carried out much, he undertook much, he held large shares of the public contracts; he lent to peoples; in many provinces his business was conducted; he gave himself even to kings; to this Alexandrian himself he lent a great sum already long before; nor in the meantime did he ever cease to enrich his friends, to send them into business, to give them shares, to increase them with his means, to support them with his credit. Why say more? Both in greatness of spirit and in liberality he had reproduced his father's life and habit. Meanwhile Ptolemy, driven from his kingdom by treacherous counsels --- as the Sibyl said, Postumus perceived --- came to Rome. To him, needy and asking, this unfortunate man lent money, and not then for the first time; for he had lent to him as reigning king while absent; nor did he think he was lending rashly, since it was doubtful to no one that the man would be restored to his kingdom by the Senate and the Roman people.

\ciceroSection{5} But in his giving and lending he went too far, and lent not only his own money but his friends' too --- foolishly; who denies it, or who will now dare, because it fell out badly, to think it was well-advised? But it is hard, when you have entered upon something with great hope, not to pursue it through to the very end. The king was a suppliant, he asked for much, he promised everything, so that Postumus was now compelled to fear that he would lose what he had already lent, if he set a limit to his lending. And there was nothing more coaxing than that king, nothing more kindly than this man --- so that he might rather regret having begun than be free to leave off.

\ciceroSection{6} From this first arises that charge: they say the Senate was corrupted. O immortal gods! Is this that longed-for severity of the courts? Our corrupters plead their case; do we who were corrupted not plead ours? What then? Am I not to defend the Senate in this place, gentlemen? I ought to, in every place indeed; so well has that order deserved of me; but neither is that the matter at issue at this time, nor is that affair joined to Postumus's case. And yet, although money was supplied by Postumus for the expense of the journey, for that magnificence of equipment and the king's retinue, and the bonds were drawn up at Cn. Pompey's Alban estate when the king had set out from Rome --- still the man who was giving ought not to have inquired why the man who was receiving was then drawing it. For he lent not to a bandit but to a king; and not to a king hostile to the Roman people, but to one whose restoration he saw entrusted to the consul by the Senate; and not to a king alien to this empire, but to one with whom he had seen a treaty struck on the Capitol.

\ciceroSection{7} But if the creditor is at fault, and not the man who wickedly used the money he had borrowed, then let the man who has forged a sword and sold it be condemned, and not the man who with that sword has killed some citizen. Wherefore neither ought you, C. Memmius, to do this --- to wish the Senate, to whose authority you have given yourself from your youth, to be involved in such infamy --- nor ought I to defend what is not at issue. For Postumus's case, whatever it is, is severed from the Senate.

\ciceroSection{8} But if I show it likewise severed from Gabinius, surely you will have nothing to say. For this is a case of ``\textit{to whom that money came}'' --- a kind of little appendix, as it were, to a case already judged and condemned. The damages have been assessed against A. Gabinius; no sureties were given, nor were the damages paid in full to the people out of his goods. The Julian law orders that pursuit be made of those to whom came the money which the convicted man took. If this is new in the Julian law --- as many things are written there more severely and more strictly than in the old laws --- then let a new procedure of this kind of trial be introduced, by all means;

\ciceroSection{9} but if this clause has been transferred word for word, as many as it had not only in the Cornelian but even earlier in the Servilian law --- by the immortal gods! what are we doing, gentlemen, or what custom of new trials are we introducing into the commonwealth? For this practice was indeed known to all of you, but, if experience is the best teacher, it ought to be best known to me. I have prosecuted on extortion, I have sat as a juror, I have presided as praetor over the inquiry, I have defended very many; no part of it which could bring any opportunity of learning has been absent from me. So I maintain this: that no one has ever pleaded a ``\textit{to whom the money came}'' case who had not been named in the assessment of the damages. And in the damages no one was named except from the words of witnesses, or the account-books of private persons, or the records of states.

\ciceroSection{10} And so, at the bringing-in of the damages, those used to be present who had something to fear about themselves, and, when they had been named, if it seemed good, they used at once to speak against it; but if they had dreaded the fresh odium of that moment, they would answer afterward. And when they had done this, very many often won their case. But this is new, and altogether unheard-of before this time. In the damages Postumus's name is nowhere. In the damages, I say; only the other day you yourselves sat as jurors in the case against A. Gabinius --- did any witness name Postumus? A witness, indeed? Did the prosecutor? Did you, in short, hear Postumus's name in that whole trial?

\ciceroSection{11} Postumus is not, then, a defendant overflowing from that case which has been judged; rather one Roman knight has been seized as a defendant on extortion. By what account-books? Those that were not read out in the Gabinian trial. On what witness? By whom he was then nowhere named. By what assessment of damages? In which no mention of Postumus was made. By what law? By one that does not bind him. Here now, gentlemen, the matter is one for your deliberation, for your wisdom; you ought to look not to how much is permitted to you, but to what befits you. For if you ask what is permitted, you can remove from the state whomever you wish; the ballot is what gives the power; the same conceals caprice, of which there is no reason for anyone to fear the witness, unless he dreads his own.

\ciceroSection{12} Where then is a juror's wisdom? In this: that he weigh not only what he can do, but also what he ought to do, and that he remember not only how much has been allowed him, but also how far it has been entrusted to him. A ballot is handed to you as juror. By what law? The Julian law on extortion. About what defendant? About a Roman knight. But that order is not bound by that law. ``By that clause,'' he says, ``\textit{to whom that money came}.'' You heard nothing against Postumus when you were a juror against Gabinius, nothing after Gabinius was condemned when you assessed the damages against him. ``But now I hear it.'' Postumus, then, is a defendant under that law from which not only he himself but the whole order too is released and free.

\ciceroSection{13} Here I shall not implore you first, Roman knights, whose right is being put to the test in this trial, but you, senators, whose good faith toward this order is at stake; which indeed, as often before, so in this very case recently has been made known. For when, with the best and most outstanding consul, Cn. Pompey, putting the question on this very inquiry, there arose some --- but very few --- harsh opinions, opinions which would have decreed that tribunes, that prefects, that scribes, that all the attendants of magistrates be bound by this law: you, you yourselves, I say, and the Senate in full session resisted; and although at that time, on account of the offences of many, that moment had blazed up even to the peril of the innocent, yet, while you were quenching the hatred of us, you did not allow a new fire to be set beneath this order.

\ciceroSection{14} With this spirit, then, the Senate. What of you, Roman knights, what at last are you going to do? Glaucia used --- a foul man, but a shrewd one --- to advise the people that, when some law was being read out, they should attend to the first line. If it were ``\textit{the Dictator, the consul, the praetor, the master of horse},'' he need not trouble himself; let him know it had nothing to do with him; but if it were ``\textit{whoever, after this law},'' let him see that he be not bound up by some new inquiry.

\ciceroSection{15} Now you, Roman knights, look to it. You know that I, sprung from your number, have always felt everything on your behalf. None of these things do I say without great care and the highest love for your order. Other men have embraced other men and other orders; I have always embraced you. I warn and foretell, while the matter and the case are still untouched I give notice, I call all men and the gods to witness: while you can, while it is allowed, take care that you do not set a harsher condition for yourselves and for your order than you can bear. This evil will creep further, believe me, than you think.

\ciceroSection{16} When a most powerful and most noble tribune of the plebs, M. Drusus, was bringing in a new inquiry against the equestrian order --- ``\textit{if anyone has taken money for the sake of giving judgement}'' --- the Roman knights openly resisted. What? Did they wish this to be allowed? By no means; they judged that this kind of taking money was not only base but even abominable. And yet they argued thus: that those ought to be bound by those laws who by their own judgement had followed that condition of life. ``The most ample rank of the state delights you, the curule chair, the fasces, the commands, the provinces, the priesthoods, the triumphs, the very portrait handed down to the memory of posterity;

\ciceroSection{17} let there be, then, at the same time, some anxiety too, and a certain greater dread of the laws and the courts. We have never despised those things'' --- so they argued --- ``but we have followed this quiet and leisured life; and since it lacks honour, let it lack vexation too.'' ``You are as much a juror, being a knight, as I am a senator.'' So it is; but you sought that, I am compelled to this. Wherefore either let it be permitted me not to be a juror, or let me not be bound by a senator's law.

\ciceroSection{18} Will you, Roman knights, lose this right received from your fathers? I warn you not to do it. Men will be dragged into these trials out of every malice --- nay, out of the very talk of the malevolent --- unless you take care. If it were now reported to you that opinions were being spoken in the Senate to bind you by these laws, you would think you must run together to the Senate-house; if a law were being brought, you would fly to the Rostra. The Senate wished you to be free of this law, the people never bound you, you came together here unfettered; take care that you do not depart fettered.

\ciceroSection{19} For if it is to be Postumus's undoing --- a man who was neither tribune, nor prefect, nor a companion out of Italy, nor an intimate of Gabinius --- in what way hereafter will those defend themselves who, being of your order, have been entangled in these affairs with our magistrates? ``You,'' he says, ``drove Gabinius on to bring back the king.'' My good faith no longer permits me to deal more harshly with Gabinius. For the man whom, received back from such great enmities into friendship, I have defended with the highest zeal, this man, struck down, I ought not to violate. With whom, if Cn. Pompey's authority had not earlier reconciled me, his own fortune would now reconcile me.

\ciceroSection{20} But still, when you say this --- that Gabinius set out for Alexandria at Postumus's instigation --- if you give no credit to Gabinius's defence, do you forget even your own accusation? Gabinius said he had done it for the sake of the commonwealth, because he feared the fleet of Archelaus, because he thought the sea would be packed with pirates; he said too that it was permitted him by law. You, his enemy, deny it. I pardon you, and the more because it is against that verdict. I return then to your charge and accusation.

\ciceroSection{21} What were you bawling out? That ten thousand talents had been promised to Gabinius. A wonderfully smooth agent, of course, had to be found, to wheedle the man --- the most avaricious, as you would have it --- into not utterly scorning two hundred and forty million sesterces. Gabinius, whatever his purpose in doing it, did it surely on his own; whatever that intent was, it was Gabinius's. Whether, as he himself said, he sought glory, or, as you would have it, money, he sought it for himself, not for Rabirius; for Rabirius was not Gabinius's companion or follower, nor had he set out from Rome and hastened on the authority of Gabinius --- whose business this was not --- but on the authority of P. Lentulus, a most illustrious man, an authority proceeding from the Senate, with a fixed plan and a hope that was not doubtful.

\ciceroSection{22} ``But he was the king's finance minister.'' And indeed he was even in the king's custody, and violence was very nearly brought against his life; many things besides, which the king's caprice, which necessity forced on him, he endured. Of all which things there is one single reproach: that he entered the kingdom, that he entrusted himself to a king's power. But if we are asking the truth --- foolishly. For what is more foolish than for a Roman knight from this city, a citizen, I say, of this commonwealth, which is and always has been most especially free, to come into a place where he must obey another and serve him?

\ciceroSection{23} But shall I not, even so, in this matter pardon Postumus --- a man of moderate learning --- when I see that men of the highest wisdom have stumbled? We have heard that the one man easily the most learned of all Greece, Plato, was involved in the greatest dangers and plots through the injustice of Dionysius, the tyrant of Sicily, to whom he had entrusted himself; that Callisthenes, a learned man, the companion of Alexander the Great, was put to death by Alexander; that Demetrius, who was called the Phalerean, both noble and famous from the Athenian commonwealth, which he had managed most excellently, and from his learning, was deprived of life in that same Egyptian kingdom, an asp set against his body.

\ciceroSection{24} I plainly confess that nothing more deranged can be done than knowingly to come into a place where you are about to lose your liberty. But the folly of this very deed another, prior, folly now defends, which makes this most foolish act --- that he came into the kingdom, that he entrusted himself to the king --- seem to have been done wisely; if indeed it belongs not so much to a man always foolish as to a man wise too late, when he has been hampered by his own folly, to extricate himself in whatever way he can.

\ciceroSection{25} Wherefore let that stand and be fixed which can neither be moved nor changed: in respect of which the fair-minded say Postumus hoped, the unfair say he sinned, and he himself confesses that he was even mad --- that he lent his own money, his friends' money, to a king at so great a peril to his own fortunes --- yet, this once undertaken and contracted, those further things had to be endured, so that he might at last set free himself and his friends. So you may cast it up at me as often as you wish, that he wore the pallium, that he had certain insignia not of a Roman man: as often as you say any of these things, so often you say one and the same thing --- that he rashly lent money to a king, that he entrusted his fortunes and his good name to a king's caprice.

\ciceroSection{26} He had acted rashly, I admit; but the deed could now in no way be changed: either the pallium had to be put on at Alexandria, that he might be allowed to wear the toga at Rome, or all his fortunes had to be cast away, if he kept the toga. For the sake of pleasure and delight I have often seen at Naples not only well-known Roman citizens, but even noble youths and certain senators born of the highest rank --- not in their own gardens or suburban estates, but at Naples,

\ciceroSection{27} in that most crowded town, in a dark tunic; in that same place many have seen the famous L. Sulla, that commander, in a Greek cloak. And of L. Scipio, who waged the war in Asia and utterly defeated Antiochus, you see the statue on the Capitol not only with the cloak but even with Greek slippers; whose impunity was free not only from prosecution but even from gossip. Surely the excuse of necessity will more easily defend P. Rutilius Rufus; who, when he had been overwhelmed by Mithridates at Mytilene, avoided the king's cruelty by a change of dress. So that P. Rutilius, who was a model to our countrymen of virtue, of antiquity, of prudence, a consular, wore Greek shoes and the pallium; nor indeed did anyone think this was to be charged to the man, but to the occasion. Shall a charge of dress be brought against Postumus --- against the man in whom there was a hope that he might at last come through to his own fortunes?

\ciceroSection{28} For when he had come to Alexandria, gentlemen, this one method of saving his money was set before Postumus by the king: if he undertook the management and, as it were, the disbursement of the royal revenues. But this he could not do unless he were appointed \textit{dioecetes} --- for he who governs these matters goes by this name. The business seemed hateful to Postumus, but there was no possibility at all of refusal; the very name too was vexing, but the office bore this name among them --- it was not he who imposed it. He hated even that dress, but without it he could maintain neither that name nor that charge. So there was at work that force which, as our poet says, ``breaks and weakens the highest resources.''

\ciceroSection{29} ``Let him die,'' you will say; for that follows. He would surely have done it, had he been able to die without the greatest dishonour, his affairs so entangled. Do not, then, turn his misfortune into his fault, nor think the king's injury this man's crime, nor interpret as deliberate plan what came from necessity, nor as free will what came from force --- unless perhaps you think that even those who have fallen among enemies or among bandits are to be blamed, if under compulsion they do anything otherwise than as free men would. None of us is ignorant, even if we have not experienced it, of the way of kings. And these are the commands of kings: ``Take heed and obey my word, and beyond what is asked''; if more --- then those threats too: ``If at the second light I find you here, you shall die''; which things we ought to read and behold not only that we may take pleasure, but that we may learn also to beware and to escape.

\ciceroSection{30} ``But from this very thing a charge arises.'' For he says that when Postumus was collecting the money for Gabinius, he collected for himself a tenth of the sums levied. I do not understand what kind of thing this is --- whether he made a tenth an addition, as our collectors are wont to do with the one-percent, or a deduction from the total. If an addition, then eleven thousand talents came to Gabinius. But not only were ten thousand alleged by you, but even assessed by these jurors.

\ciceroSection{31} I add this too: how does it fit, either that to so heavy a burden of tribute, in collecting so great a sum, an addition of a thousand talents was made; or that, in so great a payment to a man --- the most avaricious, as you would have it --- a deduction of a thousand talents was conceded? For it was not in Gabinius's nature to remit so much of his own, nor in the king's to allow so much to be imposed on his subjects. ``But there will be witnesses, the Alexandrian envoys.'' They said nothing against Gabinius; nay rather, they praised Gabinius. Where then is that custom, where the practice of the courts, where the precedents? Is a man wont to testify against the one who collected the money, who did not testify against the one in whose name that money was being collected?

\ciceroSection{32} Come, if the man who did not testify is wont to do so, is the man who praised him also wont to do so? With the same witnesses --- and indeed not produced, but the witnesses' depositions read out --- the matter is wont to be brought to these cases as though prejudged. And my friend and intimate even says that the Alexandrians had the same reason for praising Gabinius as I had for defending the same man. My reason, C. Memmius, for defending Gabinius was the renewal of goodwill. Nor indeed do I regret to have enmities mortal, friendships everlasting.

\ciceroSection{33} For if you think I defended the case unwillingly, so as not to offend Cn. Pompey's feelings, you are vehemently ignorant both of him and of me. For neither would Pompey have wished me to do anything unwillingly for his sake, nor would I, to whom the liberty of all citizens had been most dear, have flung away my own. Nor, as long as I was most hostile to Gabinius, was Cn. Pompey not most friendly to me; nor, after I had granted, on his authority, the pardon I owed, did I do anything in pretence --- lest with my own treachery I should do an injury even to the very man on whom I had conferred the favour. For by not returning into goodwill with my enemy I was not wronging Pompey; but if, brought back through him, I had returned treacherously, I should of course have deceived myself most of all, but next to that I should have deceived him too.

\ciceroSection{34} But let us leave off about myself; let us return to those Alexandrians of yours. What a face they have, what audacity! Only the other day, before your eyes, in Gabinius's trial they were roused up at every third word; they denied that money had been given to Gabinius. Again and again Pompey's testimony was read out, that the king had written to him that no money had been given to Gabinius except for military purposes. ``At that time,'' he says, ``the Alexandrian witnesses were not believed.'' What then? They are believed now. Why? Because now they say what then they denied.

\ciceroSection{35} What then? Is this the standing of witnesses --- that the same men who were not believed when they denied should be believed when they assert? But if they spoke the truth then, with most severe brow, now they are lying; if they lied then, let them teach us with what countenance they are wont to speak the truth. We used to hear of Alexandria, now we know it. From there all the conjuring-tricks, from there, I say, all the deceptions, from there, in short, came all those plot-devices of their mimes. And nothing takes me longer, gentlemen, than to read men's faces.

\ciceroSection{36} They spoke just now here with us, at these very benches; with what raised eyebrows, refusing this charge of ten thousand! Now you know the tastelessness of the Greeks; they were acting it out with their shoulders --- for the sake of the moment then, I suppose; now of course there is no moment for it. Once a man has perjured himself, he ought not to be believed afterward, even if he swear by ever so many gods --- especially, gentlemen, since in these trials there is wont to be no place even for a new witness, and for that reason the same jurors are retained who served on the defendant's case, so that everything may be known to them and nothing new may be able to be invented.

\ciceroSection{37} ``\textit{To whom that money came}'' cases are not wont to result in a defendant's being condemned by trials properly their own. And so, if either Gabinius had given sureties, or the people had recovered from his goods as much as the sum of the damages, then however great a sum had come to Postumus from him, it would not be reclaimed; so that it can easily be understood that what, out of the money which had come to some defendant who was condemned, was shown in that first trial to have come to someone --- that is wont to be reclaimed by this kind of trial. But now what is being done? Where on earth are we? What so perverse, so back-to-front, can be said or devised?

\ciceroSection{38} The man is accused who did not take from the king, as Gabinius was judged to have done, but who lent the king a very great sum of money. So he gave to Gabinius who did not pay back to this man. Is that it? Come, tell me: when the man who owed money to Postumus has given it not to him but to Gabinius, then, Gabinius being condemned, is the case to be pleaded by that man to whom the money came, or by this man? ``But he has it and conceals it.'' For there are some who talk this way. What kind of ostentation and boasting is that, after all? If he had never had anything, still, if he had had it, there would be no reason why he should hide that he had it. But the man who had received two rich and ample patrimonies, and besides had increased his estate by good and honest dealings --- what reason at last would there be why he should wish to be thought to have nothing?

\ciceroSection{39} Or is it this: that, when, drawn on by the interest, he was lending, he was working to have as much as possible; but afterward, when he had exacted what he had lent, he was working to be thought in want? A new kind of glory he covets! ``For he was lord at Alexandria.'' Nay rather, he was under a most arrogant lordship; he himself endured custody, he saw his friends in chains, death often hovered before his eyes, naked and destitute he fled at the last from the kingdom.

\ciceroSection{40} ``But money was at some point remitted by exchange, Postumus's ships were brought to Puteoli, his wares were heard of and seen.'' Cheats indeed and counterfeit, of paper and linen and glass; and although many ships had been packed with them, the freight could not be paid. That arrival at Puteoli, the talk of that time, the running and the showing-off of the passengers, and Postumus's name then somewhat envied among the malevolent on account of the report of his money --- for one summer, I know not which, no more, it crammed their ears with those tales.

\ciceroSection{41} But the truth, gentlemen, if you wish to know it: had not C. Caesar's liberality --- the highest toward all, the same incredible toward this man --- come forth, we should not still have this Postumus in the Forum. He alone took up the burdens of many of this man's friends; and the obligations which many of his close associates, in Postumus's prosperity, had shared out and shouldered, now in his stricken fortunes one man bears them all. You see the shadow and the image of a Roman knight, gentlemen, preserved by the help and faith of one friend. Nothing can be torn from this man except this likeness of his former dignity, which Caesar alone guards and upholds; which indeed, in his most wretched circumstances, must nonetheless be reckoned the greatest gift to him --- unless of course this can be brought about by some moderate virtue: that so great a man should hold this man so dear, and a man stricken, above all, and absent, and amid so great a fortune of his own that to regard another's is a great thing, and amid so great an occupation with the greatest affairs which he is conducting and has conducted, that either it would be no wonder he should forget others, or, if he remembered, he could easily prove that he had forgotten.

\ciceroSection{42} For my part I have come to know many of C. Caesar's virtues, great and incredible; but the rest are set forth, as it were, on greater stages and are almost public. To choose ground for a camp, to draw up an army, to storm cities, to rout the enemy's line, to bear that violence of cold and winters which we scarcely endure beneath the roofs of this city, on those very days to pursue the enemy when even the beasts hide themselves in their lairs and all wars by the law of nations fall quiet --- these are indeed great things; who denies it? But they have been roused up by great rewards and by the everlasting memory of men. The less is it to be wondered at that the man does those things who has coveted immortality.

\ciceroSection{43} This is the true praise, which is celebrated not by the songs of poets, not by the monuments of annals, but is weighed by the judgement of the prudent. A Roman knight, his old friend, zealous, loving, attentive toward him, slipping not through caprice, not through the base outlays and losses of his desires, but through his enterprise in enlarging his patrimony --- this man he took up, he did not let him fall to ruin, he propped and upheld him with his means, his fortune, his faith, and today upholds him still, nor allows his friend, hanging in the balance, to crash down; nor does the splendour of his own name dazzle the keenness of that man's spirit, nor does the height of his fortune and glory stand in the way of the lights, as it were, of his mind.

\ciceroSection{44} Let those things be great indeed, which in very truth are great; about the judgement of my own mind, let each man feel as he will: for I set this liberality toward his own, this mindfulness of friendship amid such great resources, such great fortune, before all his other virtues. And this goodness of his, of a new kind, unfamiliar to illustrious and overmighty men, you, gentlemen, ought not only not to spurn and reject but even to embrace and to magnify --- and the more because you see these days have been taken up to undermine that man's dignity. From which nothing can be taken from him that he would not either bear bravely or easily restore; but if he hears that his closest friend has been despoiled of his honour, he will not bear it without great grief, and he will have lost that which he cannot hope to be able to recover.

\ciceroSection{45} These things ought to be enough for men who are not unfair, and even far too many for you, whom we are confident are the fairest of all. But that satisfaction may be given to everyone's suspicion, or malevolence, or cruelty: ``Postumus is hiding money, the king's riches lie concealed.'' Is there anyone out of so great a people who would wish the goods of C. Rabirius Postumus knocked down to himself for a single sesterce? But wretched me, with how great grief I said that! Ah, Postumus, are you the son of C. Curtius, by C. Rabirius's judgement and will a son, by birth his sister's son? Are you that man, liberal toward all your own, whose goodness has enriched many, who have poured out nothing, contributed nothing to any caprice? Are your goods, Postumus, knocked down by me for a single sesterce? O my wretched and bitter auctioneering!

\ciceroSection{46} But this even the poor man wishes: that he be even condemned by you, if his goods are so sold that to each man his full due is paid. He cares now for nothing else but his good faith, nor can you, if you should now be willing to forget your own clemency, tear anything else from him besides. Which, gentlemen, that you may not do, I beg and beseech you --- and the more, if money come from abroad is sought from a man to whom his own is not paid back. For against the man whom pity ought to have aided, envy has been sought out.

\ciceroSection{47} But now, since, as I hope, I have made good the faith I gave you, Postumus, I shall pay also the tears I owe; which indeed I saw most abundant from you in my own downfall. There hovers before my eyes that sorrowful night for all my household, when you brought your whole self with all your resources to me. With companions, with a guard, with even so great a weight of gold as that time demanded, you supported that departure of mine; you never failed my children in my absence, never my wife. I can summon up many witnesses, brought back to their country, of your liberality, which I have often heard was a great help to your father Curtius in a trial on a capital charge.

\ciceroSection{48} But now I fear everything; I dread the very odium of his goodness. For the weeping of so many men shows how dear you are to your own, and grief weakens me and chokes off my voice. I beseech you, gentlemen, do not tear from this best of men --- than whom no one was ever better --- the name of a Roman knight, and the enjoyment of this light, and the sight of you. He begs you for nothing else but that he be allowed to look upon this city with eyes unbowed, and to set his footstep in this Forum --- which very thing Fortune had torn from him, had not one friend's resources come to his aid.
