% Pro M. Aemilio Scauro (pro-scauro) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{a} It was above all to be wished for M. Scaurus, gentlemen, that, without having drawn upon himself the hatred of anyone, he might keep, without offence and without trouble, what he has always sought before all else: the dignity of his birth, his family, his name.

\ciceroSection{c} He underwent also the judgement of the people, on the inquiry of Cn. Domitius, tribune of the plebs.

\ciceroSection{d} He was made a defendant by Q. Servilius Caepio under the Servilian law, at a time when the courts lay in the hands of the equestrian order, and when, after P. Rutilius had been condemned, no one seemed so innocent that he did not fear those courts.

\ciceroSection{e} By the same man, under the Varian law, that guardian of the commonwealth was called into the charge of treason; he was harried by Q. Varius, tribune of the plebs.

\ciceroSection{f} For I did not merely admire that man, as all men did, but loved him with a particular love. He was the first who, when I was burning with the zeal for praise, drove me into the hope that by my own courage, without the support of fortune, I could arrive by labour and constancy wherever I had aimed.

\ciceroSection{g} And since the prosecution was heaped up rather by a kind of pile of charges than by any division of kinds and variety.

\ciceroSection{h} He said, then, that a certain Bostar of Nora, fleeing from Sardinia at Scaurus's coming, was buried before the dinner could be cleared away for this man.

\ciceroSection{i} If, in short, he could in no way have seized upon those goods unless Bostar were dead.

\ciceroSection{k} If, by Hercules, gentlemen, I were speaking for L. Tubulus --- the one man whom, out of all memory, we have received as the most criminal and most reckless --- even so I would not fear it, were it said that poison had been given by him to a guest or a fellow-diner at table, a man to whom he was neither heir nor enemy.

\ciceroSection{m} ...the goods, rather than that what he had should be sold. Come now: I have defended Scaurus, Triarius; do you defend your mother.

\ciceroSection{n} ...that you feared he was insolvent; that in the end the defendant wished to keep the goods which had been proscribed, unless...

\ciceroSection{o} When Aris would not give it, he tried to flee secretly from Sardinia.

\ciceroSection{p} They (the beavers) ransom themselves with that part of the body for the sake of which they are most sought after.

\ciceroSection{1q} So, I say, gentlemen, the matter stands; and this is no new thing argued by me, but a thing inquired into by others.

\ciceroSection{1r} Those things we have heard; this, indeed, we remember and have all but seen: that P. Crassus, of the same stock and name, killed himself rather than fall into the hands of his enemies.

\ciceroSection{2s} And neither could M'. Aquilius, who had held the same honours and had been most brave in war, match that deed of the elder Crassus...

\ciceroSection{2} ...he disfigured by the disgrace of his own old age and of his exploits. What of this? Could either the most illustrious of the Julii, or M. Antonius, endowed with the highest command, match the other Crassus in those same times?

\ciceroSection{3} What? In all the monuments of Greece --- which are more adorned in words than in substance --- who is found, when you have set aside Ajax and the fables (who himself, from grief at his disgrace, as the poet says, an insolent victor, could not bear to be vanquished), save the Athenian Themistocles, who punished himself with death?

\ciceroSection{4} But the Greeklings, to be sure, invent much; among whom they relate that even Cleombrotus of Ambracia threw himself down from the highest wall --- not because he had met with any bitterness, but, as I see it written among the Greeks, because, having read a book of the greatest philosopher Plato, gravely and elegantly written, on death, in which, as I think, Socrates, on that very day on which he was to die, argues at great length that that is death which we should reckon life --- when the soul is hemmed in by the body as in a prison --- and that that is life, when the same soul, freed from the chains of the body, has betaken itself back to the place from which it had sprung.

\ciceroSection{5} Did that Sardinian woman of yours, then, know or had she read Pythagoras or Plato? --- who, even so, praise death only in such a way that they forbid one to flee from life, and say that this is done against the covenant and the law of nature. Of voluntary death you will surely find no other just cause. And this he himself saw; for he let fall, at a certain point, that the woman had chosen rather to be despoiled of life than of her chastity.

\ciceroSection{6} But he recoiled at once and said no more about her chastity --- afraid, I suppose, lest he give us occasion for mockery and jest. For it is established that she was of the highest ugliness, and old besides. How, then, can there be --- however witty that Sardinian woman may have been --- any suspicion of lust or love?

\ciceroSection{7} And lest you suppose, Triarius, that what I am about to bring forward I am inventing myself in the speaking, and not learning the case from my client, I will set out for you what the opinions in Sardinia were about that woman's death --- for there were two --- so that all the more easily...

\ciceroSection{8} ...I have told you, he had long since loved that lustful and shameless mother in a notorious and well-known adultery. Since he feared this wife of his --- an old woman, and rich, and troublesome --- he was unwilling either to keep her in marriage because of her ugliness, or to put her away because of her dowry. And so, by an agreement made with the mother of Bostar, he formed the plan that both should come to Rome; there, he assured her, he would find some way to take her to wife.

\ciceroSection{9} Here the opinion was, as I said, twofold. One did not jar with the natural state of things: that Aris's wife, stirred up by the pain of being supplanted, when she had heard that Aris had betaken himself to Rome with that woman of his under a feigned show of fear and flight --- so that, since there had earlier been an intimacy between them, they might now be joined too in marriage --- burned with a woman's pain, and chose rather to die than to endure it.

\ciceroSection{10} The other no less probable, and, as I think, even more believed in Sardinia: that this Aris, your witness and your host, Triarius, when setting out for Rome, gave the task to his freedman --- not, indeed, to do violence to the old woman, for that would not have been right toward his patroness, but to squeeze her throat with two little fingers, and to gird it with a thin cord, so that she might be thought to have perished by hanging.

\ciceroSection{11} This suspicion indeed prevailed all the more for this reason: that, when the people of Nora were celebrating the rites of the dead and all had, after their custom, gone out of the town, then it was said that she had been hanged by the freedman. Now the departure and the solitude were a thing to be sought by the man who was throttling his patroness; for her who wished to die they were not needed.

\ciceroSection{12} And the suspicion was confirmed by this: that, the old woman dead, the freedman set out for Rome at once, as if the work were finished; while Aris, as soon as the freedman brought word of his wife's death, straightway at Rome took that mother of Bostar to wife.

\ciceroSection{13} See into what families --- how foul, how defiled, how base --- you are giving over this family, gentlemen. See by what witnesses you are moved; about what man, what stock, what name you cast your votes --- do you think you ought to forget? You discern the crimes of mothers against their children, of husbands against their wives; you see monstrous lusts mingled with cruelty. The authors of the two greatest charges --- the charges by which this whole case has been defamed among the ignorant or the envious --- you have, men disfigured by every outrage and disgrace.

\ciceroSection{14} Is there, then, in these charges, gentlemen, any suspicion left even now? Have they not been purged through, refuted, broken? How did this come about? Because you gave me, Triarius, something I could dissolve, something I could reason upon, something I could argue about; because the kind of charge was such that it did not hang wholly upon a witness, but such that the juror himself could weigh it by his own judgement.

\ciceroSection{15} Nor indeed, gentlemen, ought we to do anything else with an unknown witness than to seek out the force and nature of the matters themselves by argument, by conjecture, by suspicion. For a witness --- not only an African or a Sardinian, to be sure, if so they prefer to be called, but even any more refined and more scrupulous man --- can be driven on, deterred, fashioned, bent: he is master of his own will, and in him lies an unpunished licence of lying.

\ciceroSection{16} But the argument --- which alone is proper to the matter, for nothing else can truly be called an argument --- is the voice of the facts, the footprint of nature, the mark of truth; whatever it is, it must of necessity remain unchanged; for it is not invented by the orator, but taken up. And so, in that kind of accusation, if I were beaten, I would yield and give way; for I should be beaten by the fact, beaten by the case, beaten by the truth.

\ciceroSection{17} Would you lead in against me a column of Sardinians, and bands of them, and try to terrify me, not press me, with the din of Africans? I shall not, indeed, be able to argue, but I shall be able to take refuge in the good faith and mildness of these men, in the sworn oath of the jurors, in the fairness of the Roman people, who wished this family to be foremost in this city; I shall be able to implore the divine power of the immortal gods, who have always stood forth as the favourers of this stock and this name.

\ciceroSection{18} He demanded, he commanded, he wrested away, he forced. If you teach me by ledgers --- since the very making-up of the ledgers holds a kind of sequence and order of the business transacted --- I shall attend keenly and shall see what I must do in defence. If, in the end, you lean upon witnesses --- I will not say good and approved men, only let them be known --- I shall consider in what manner I must clash with each of them.

\ciceroSection{19} But if there is one colour, one voice, one nation of all the witnesses; if what they say they try to confirm by no arguments at all, nor even by any kind of writing, whether public or private --- which itself, even so, can be forged --- where am I to turn, gentlemen, or what am I to do? Am I to argue with them one by one? What? ``You had nothing to give.'' He will say he had it. Who shall know this, who shall judge? ``There was no ground for it.'' He will invent one. How shall we refute it? ``You could have refrained from giving, if you had wished.'' He will say it was wrested from him by force. What eloquence can, by arguing, confute the shamelessness of an unknown man?

\ciceroSection{20} I will not therefore deal with that conspiracy of Sardinians, nor with their wrung-out, forced, and solicited perjury, with subtlety; nor will I hunt out certain arguments fined down as with a needle; but against the onset of those men I will charge and clash with an onset of my own. Not each single man is to be dragged out by me from their battle-line, nor must I fight and contend with them one by one: that whole line is to be laid flat by a single charge.

\ciceroSection{21} For there is one greatest charge, the grain-charge, of all Sardinia, about which Triarius questioned all the Sardinians --- a kind of charge confirmed by one compact and consent of all their testimony. Before I touch that charge, I ask of you, gentlemen, to suffer me to lay, as it were, certain foundations of our whole defence. If these shall be set down and established, as my reasoning and my reflection bears, I shall dread no part of the accusation.

\ciceroSection{22} For I shall speak first about the kind of accusation itself; afterward about the Sardinians; then also a few words about Scaurus; and when these matters have been unfolded, then at last I shall come to this dreadful and fearsome grain-charge.

\ciceroSection{23} What sort of accusation is this, then, Triarius --- first, that you did not go to make inquiry? What was that confidence in crushing this man, so fierce, so sure? When I was a boy, I seem to have heard that L. Aelius, a man of freedman's birth, learned and witty, when he was avenging the wrongs done to his patron, laid an information against Q. Mutto, a man of the basest sort. When it was asked of him what province or what day for the witnesses he demanded, he asked for the eighth hour for himself, while he should make inquiry in the cattle-market.

\ciceroSection{24} Did you think you ought to do this same thing in the case of M. Aemilius Scaurus? ``For the case,'' he says, ``was brought to me at Rome.'' What? Did not the Sicilians bring the case of Sicily to me at Rome? And what men! Prudent by nature, shrewd by practice, learned in their training. Yet I thought that the case of the province must be learned and studied in the province itself.

\ciceroSection{25} Or was I not to look into the complaints and wrongs of the ploughmen amid the very crops and fields? I traversed, I say, Triarius --- in a most hard winter, indeed --- the valleys and hills of the Agrigentines. That most famous and most fruitful plain, the Leontine, all but taught me the case itself. I went to the huts of the ploughmen; men talked with me from the very plough-handle.

\ciceroSection{26} And so that case was so vividly drawn out that they seemed not to hear what I was saying, gentlemen, but to discern it and all but to touch it. For it seemed to me neither probable nor true that I, having taken up the patronage of a most loyal and most ancient province, should learn the case in my own bedchamber, as if it were the case of a single client.

\ciceroSection{27} I, lately, when the people of Reate, who were under my protection, wished me to plead their public case about the river Velinus and its channels before these consuls, did not think I should do enough either for the dignity of a most weighty prefecture or for my own good faith, unless not only the men but the very place and lake had taught me that case.

\ciceroSection{28} Nor would you have done otherwise, Triarius, if those Sardinians of yours had wished you to do it --- these men who were least of all willing for you to come into Sardinia, lest you should learn that everything was far otherwise than it had been brought before you: that there was no complaint of the multitude in Sardinia, no hatred of Scaurus among the people...

\ciceroSection{29} ...they say Etna burns with its breath; so would I have buried Verres with all Sicily as witness. But you set the case over for a second hearing with one witness produced. And with what a witness, immortal gods! Not enough that he was one, that he was unknown, that he was of no weight: did you finish off the first hearing with Valerius too as witness --- a man who, presented with citizenship by your father's benefaction, repaid his thanks to you not by distinguished services but by manifest perjury?

\ciceroSection{30} But if perchance the omen of your family's name has led you on, we, even so, after the manner of our forefathers --- because we hold it to be auspicious --- interpret it not toward ruin but toward safety. But all that haste and hurry of yours, in that you did away with the inquiry and with the whole prior hearing, laid open and made plain what was nonetheless not hidden: that this trial was not got up for the sake of judgement, but for the sake of the consular elections.

\ciceroSection{31} Here I shall in no place, gentlemen, find fault with Appius Claudius, a most brave and most distinguished man, joined to me, as I hope, by a faithful and firm return into goodwill. For let those have been the part either of one whom his own grief and suspicion compelled to act so, or of one who claimed this part for himself --- because either he did not notice whom he was injuring, or thought a return into goodwill would be easy for him;

\ciceroSection{32} I shall say only what is enough for the case and what can be least harsh or rough toward him. For what is there of disgrace in Appius Claudius being an enemy to M. Scaurus? What? Was not his grandfather an enemy to P. Africanus? What? That same man to me myself? What? I to him? Which enmities have perhaps sometimes brought grief to each of us, but disgrace, certainly, never.

\ciceroSection{33} The outgoing governor envied his successor; he wished him to be as much offended as possible, so that his own memory might stand out the more --- a thing not only not jarring with custom but even usual and very widespread. Nor indeed would this very daily matter, by itself, have moved Appius Claudius, endowed as he is with that humanity and wisdom, had he not thought that this man would be a rival to his brother C. Claudius.

\ciceroSection{34} Who, whether he should be patrician or plebeian --- for he had not yet settled it for certain --- Appius thought he would have contention with this man, and the greater for this: that he remembered him to have been a patrician in his candidacy for the pontificate, in the priesthood of the Salii, and in the rest. For which reason, in his own consulship he wished neither that his brother be turned away, nor, if this man were a patrician, did he see that he would be a match for Scaurus, unless he had struck him down by some fear or infamy.

\ciceroSection{35} Should I think this not to be conceded to a brother in his brother's most ample honour --- I especially, who feel, almost beyond all others, how much brotherly love avails? But the brother, you say, is no longer a candidate. What then? If that man --- held back at the supplication of all Asia, prevailed upon by the businessmen, by the tax-farmers, by all the allies and citizens --- set the advantage and safety of the province before his own honour, do you therefore think that a mind once galled could so easily be healed?

\ciceroSection{36} And yet, in all these matters, especially among barbarous men, opinion often avails more than the thing itself. The Sardinians are persuaded that they will do nothing more pleasing to Appius than if they detract from Scaurus's reputation; they are led on, too, by the hope of many advantages and rewards; they think a consul can do anything, especially when he promises unasked. About which I will now say no more.

\ciceroSection{37} And yet what I have said I have said no otherwise than if I were his brother --- not the one who both is so and has said much, but the one I have been wont to be in my own person. You ought, then, gentlemen, to set yourselves against the whole kind of this accusation, in which you see that nothing was undertaken by custom, nothing with measure, nothing with deliberation, nothing with integrity, but, on the contrary, everything was undertaken shamelessly, in confusion, hastily, headlong, by conspiracy, by command, by authority, by hope, by threats.

\ciceroSection{38} I come now to the witnesses, in whom I shall show that there is not only no good faith and authority, but not even the appearance or the image of witnesses. For, first, their very agreement destroys good faith --- an agreement laid bare by the Sardinians' compact and by the conspiracy that has been read out; next, that greed which has been taken up in the hope and the promise of rewards; lastly, the nation itself, whose vanity is so great that they think freedom is to be distinguished from slavery by nothing else than the licence of lying.

\ciceroSection{39} Nor do I say that we ought never to be moved by the complaints of the Sardinians. I am neither so inhuman nor so estranged from the Sardinians --- especially since my brother lately came away from them, after he had been put in charge of the grain-supply by the commission of Cn. Pompey; who both himself, in his own good faith and humanity, took thought for them, and was in turn most dear and welcome to them.

\ciceroSection{40} But let this refuge lie open to grief, let it lie open to just complaints; let the way be barred to conspiracy, let it be hedged off against treachery --- and this no more in the case of Sardinians than of Gauls, of Africans, of Spaniards. T. Albucius was condemned, and C. Megaboccus, on evidence from Sardinia, with some Sardinians even among those who spoke in their praise. Thus the very variety made for greater credit. For they were held fast by fair witnesses and by uncorrupted ledgers.

\ciceroSection{41} Now there is one voice, one mind --- not wrung out by grief, but feigned; and stirred up not by this man's wrongs, but by the promises and rewards of others. ``But the Sardinians were believed at some time.'' And perhaps they will be believed at some time, if they come uncorrupted, if untainted, if of their own accord, if not at someone's prompting, if free of obligation, if at liberty. If these conditions hold, even so let them rejoice and marvel that they are believed. But when all these are absent, will they still not look to themselves, will they not shudder at the reputation of their own race?

\ciceroSection{42} That the Phoenicians are the most deceitful of races, all the monuments of antiquity and all histories have handed down to us. Sprung from these, the Carthaginians, by their many rebellions, by their many violated and broken treaties, have shown that they have in no way degenerated. From the Carthaginians, with the race of Africans mixed in, sprang the Sardinians --- colonists not led out into Sardinia and there settled, but cast away and disowned.

\ciceroSection{43} Wherefore, since there was nothing untainted in this race full of pestilence, how strongly do we suppose it has been soured by so many transfusions? Here Cn. Domitius Sincaius will pardon me, a most distinguished man, my guest and friend; all those, in short, who were presented with citizenship by the same Cn. Pompey will pardon me --- whose praise, even so, we make use of; the other good men from Sardinia will pardon me, for I believe there are some.

\ciceroSection{44} Nor, when I speak of the vices of the race, do I except no one; but I must speak of the whole stock, in which perhaps some have, by their own character and humanity, conquered the vices of the stock and race itself. That a great part of it is without good faith, without fellowship and connection with our name, the matter itself declares. For what province is there besides Sardinia which has not a single state friendly and free to the Roman people?

\ciceroSection{45a} Africa herself, that parent of Sardinia, who waged so many and most bitter wars with our forefathers, defended herself from the alliance of the Punic wars not only by her most loyal kingdoms but even in the province itself, with Utica as witness. Hither Spain, by the death of the Scipios...

\ciceroSection{45b} Poor in resources, deceitful in race.

\ciceroSection{45c} There were found those who were even called brothers of the Roman people.

\ciceroSection{45d} When this name was heard, which has ranged through all nations.

\ciceroSection{45f} For when, out of many, one only remained to him --- Dolabella, his father's enemy, who together with his kinsman Q. Caepio had set his hand to the indictment against Scaurus's own father --- that enmity, he held, had not been taken up by him but left to him.

\ciceroSection{45g} What the devil sort of reasoning is that?

\ciceroSection{45h} The man whom the royal purple did not move --- did the Sardinians' sheepskin change him?

\ciceroSection{45k} Especially since the nearness and the throng of the place removes any suspicion of sloth or of greed.

\ciceroSection{45l} I, moreover, who have Alban columns --- I brought them on pack-saddles.

\ciceroSection{45m} Were you in want of a house? But you had one. Did money abound? But you were in need. You ran, out of your senses, against the columns; against another man's goods, a madman, you played the madman; you valued a house sunk down, blind, prostrate, at more than yourself and your own fortunes.

\ciceroSection{45n} Since you could not escape these things, will you nonetheless contend and demand that M. Aemilius --- with all his dignity, with his father's memory, with his grandfather's glory --- be given up to a most sordid, most empty, most worthless race, and to witnesses, I might almost say, in skins?

\ciceroSection{46o} On every side it supplies me with something to say for M. Scaurus, wherever not only my mind but even my eyes have fallen. That Senate-house testifies on your behalf concerning his father's most weighty leadership and most brave conduct; L. Metellus himself, this man's grandfather, seems to have established the most holy gods in that temple in your sight, gentlemen, that they might plead with you for the safety of their grandson --- because they themselves had often come with their aid to many who were in straits and imploring help.

\ciceroSection{47} That Capitol, made splendid with its three temples, the approaches to Jupiter Optimus Maximus, to Juno Regina, to Minerva adorned with his father's and even with this man's most ample gifts --- M. Scaurus before...

\ciceroSection{48} ...of that L. Metellus, pontifex maximus, who, when that temple was burning, threw himself into the midst of the fire and snatched from the flame that Palladium which, as a kind of pledge of our safety and empire, is kept under the guardianship of Vesta. Would that he could for a little while come forth! He would surely snatch his own stock from this flame, who had snatched from that conflagration the gods...

\ciceroSection{49} ...you indeed, M. Scaurus, I see --- I see, I say, not merely think of --- and not, indeed, without great mourning and grief of spirit, when I have looked upon your son's mourning garb and call you to mind. And would that, as in this whole case you have been before my eyes, so now you might present yourself to the minds of these men and cling fast in their hearts! Even one who chanced not to know you, your very aspect would tell that you were a leading man of the state.

\ciceroSection{50} In what manner am I now to address you? As a man? But you are not among us. As a dead man? But you live and flourish; you dwell in the minds and on the lips of all; and your divine spirit had nothing mortal, nor could anything of yours die save the body. In whatever manner, then, I...

\ciceroSection{51} ...altogether.
