% In Pisonem (in-pisonem) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{fr1} By the immortal gods! What a day, senators, has dawned upon me — most longed-for! — that I should see this portent of this place, this monster of the city, this prodigy of the state!

\ciceroSection{fr2} For my part I have wished nothing else; you perhaps might have preferred to hear that he was wasted by torment or sunk in the waves.

\ciceroSection{fr3} A disturbance of mind has driven him on, and a kind of darkness of crimes spreading over him, and the burning torches of the Furies.

\ciceroSection{fr4} On what rock has he not stumbled? Onto what weapon has he not run?

\ciceroSection{fr5} What has he dared to deny — or rather, what has he not confessed?

\ciceroSection{fr6} What is more sluggish than that man, what filthier, what more worthless, what more nerveless, what more stupid, what more obscure?

\ciceroSection{fr7} Turbulent, seditious, factious, ruinous.

\ciceroSection{fr8} What slightest sign of talent in you? Of talent? Indeed, of a freeborn and free man? You who by your very colour scorn your country, by your speech your race, by your character your name.

\ciceroSection{fr9} This is not meant to look down on Placentia, which he is wont to boast as the place of his origin; for neither does my own nature carry me to that, nor does the dignity of a municipality which has deserved most well of me bear it.

\ciceroSection{fr10} Here this man, having set out from his native place, by chance settled at Placentia; and a few years afterwards he was reckoned a citizen of that town — for it was then a town. For first he began to be reckoned a Gaul, then a Cisalpine, at last a Placentine.

\ciceroSection{fr11} There was a certain Insubrian, both merchant and crier; when he had come to Rome with his daughter, he dared to address a young noble, the son of a most thievish Caesoninus, and gave him his daughter in marriage. They say the man was called Calventius. To a light and rash man he gave his daughter.

\ciceroSection{fr12} That Insubrian grandfather adopted a man even greater for himself.

\ciceroSection{fr13} Your father had a more elegant father-in-law than Gaius Piso had in that grief of mine. To him I gave my daughter — to whom, if I had had then the power of all, I should especially of all have given her.

\ciceroSection{fr14} That mother of yours, brought from I-know-not-what lands, poured you out from her belly as a beast, not a man. Who poured you from her womb a brute, not a man.

\ciceroSection{fr15} Although your whole kindred is borne to you by waggon.

\ciceroSection{fr16} All those things — pretended, feigned, painted-over.

\ciceroSection{fr17} I had thought him austere, I had thought him gloomy, I had thought him grave; but I see an adulterer, I see a glutton, I see one taking shelter in his walls, in the squalor of his friends, hiding his lusts in shadows. I had thought him grave, I see an adulterer, I see a glutton.

\ciceroSection{fr18} Nor did anyone dare to sit by Gabinius or to speak to him in the curia.

\ciceroSection{fr19} What of this — that the wretch, when he could not speak, could not be silent?

\ciceroSection{1} Now do you see, beast, now do you feel what is the men's complaint about your face? No one complains that some Syrian, made out of the herd of the new-bought, has been made consul. For not that slavish colour of yours, not those hairy cheeks, not those rotten teeth deceived us. Your eyes, your eyebrows, your forehead, in short the whole face of you (which is a kind of silent speech of the mind) — this drove men into delusion, this deceived, betrayed, led astray those to whom you were unknown. Few of us knew those muddy vices of yours, few your slow wits, the dullness and weakness of your tongue. Never had your voice been heard in the Forum, never had a trial been made of your counsel, no deed of yours, not to say distinguished but not even noted, was known either in the field or at home. You crept up to office by men's mistake, by the recommendation of smoky portraits — to which you have nothing similar except the colour.

\ciceroSection{2} And does the same man boast to me that he attained every magistracy without rebuff? It is allowed to me to say of myself those things — and with true glory. For the Roman people gave all the offices to me — to me the man. For when you were made quaestor, even those who had never seen you yet committed that office to your name. Aedile you were made — Piso was made by the Roman people, not this Piso. The praetorship likewise was given to your forefathers. Those dead were known; you living, no one yet knew. Me the Roman people first as quaestor, first by ranking as aedile, first as praetor, by all the votes made — the people gave the honour to the man, not to family, to morals, not to ancestors, to virtue seen, not to nobility heard.

\ciceroSection{3} For what shall I say of the consulship — gained, or carried? Wretched me, that I now compare myself with this pestilence and ruin! But I shall say nothing for the sake of comparison; and yet I shall grasp those things which lie farthest apart. You were declared consul — I shall say nothing harsher than what all confess — in difficult times of the commonwealth, with the consuls divided, when you did not refuse this from those who said you were consul: that they did not even think you worthy of the light, unless you had stood out as worse than Gabinius. Me all Italy, me all the orders, me the entire state declared first consul not by tablet sooner than by voice. But I leave aside how each of us was made; let chance, granted, be mistress of the field. It is more magnificent to say how we have carried our consulship than how we got it.

\ciceroSection{4} I, on the Kalends of January, freed the senate and all good men from fear of the agrarian law and of the greatest largesses. I preserved the Campanian land — if it ought not to have been divided; if it ought to have been, I reserved it for better authors. I, in the case of Gaius Rabirius accused of high treason, upheld and defended, against hatred, the senate's authority interposed forty years before my consulship. I, by my own enmity (without any ill will of the senate towards the comitia), deprived young men, brave and good, but in such a state of fortune that, if they had got magistracies, they would have seemed about to overthrow the commonwealth's stability.

\ciceroSection{5} I tamed by my patience and indulgence Antonius my colleague, eager for a province, plotting many things in the commonwealth. The province of Gaul, equipped and adorned by the senate's authority with army and money — which I exchanged with Antonius, because I judged the times of the commonwealth so to bear — I laid down in public assembly, with the Roman people crying out against it. I bade Lucius Catiline, plotting not obscurely but openly the murder of the senate and the destruction of the city, depart from the city, that we might be safe by walls from one against whom we could not by laws. I wrenched out, in the last month of my consulship, the weapons aimed at the throats of the state from the wicked hands of the conspirators. I seized, brought forth, extinguished the torches already kindled for this city's burning.

\ciceroSection{6} Me Quintus Catulus, the chief of this order and author of public counsel, in a most crowded senate named ``Father of the Country.'' To me this most distinguished man, who sits next to you because of you, Lucius Gellius, with these listening, said the civic crown was owed by the commonwealth. To me, in my toga, the senate threw open the temples of the immortal gods — not, as for many, for the commonwealth well managed, but, as for none, for the commonwealth preserved, with a singular kind of supplication. When, leaving office, I was forbidden by a tribune of the plebs to say in the assembly what I had decided, and he allowed me only to swear, without any hesitation I swore that the commonwealth and this city was safe by my own work alone.

\ciceroSection{7} Me, the entire Roman people, in that assembly, gave not the congratulation of one day, but eternity and immortality, when they on their oath, with one voice and consensus, approved my so great and so weighty oath. At which time was that return of mine home from the Forum, that no one — except him who was with me — seemed to be in the number of citizens. So my consulship was carried through by me, that I did nothing without the senate's counsel, nothing without the Roman people's approval; that I always defended the curia from the rostra, the people in the senate; that I joined the multitude with the leading men, the equestrian order with the senate. I have set out my consulship briefly.

\ciceroSection{8} Dare now, O Fury, to speak of yours! Of which the start was the games at the Crossroads — then for the first time held since Lucius Iulius and Gaius Marcius's consulship, against this order's authority. These Quintus Metellus — I do harm to a most brave dead man, in setting against this hateful beast him whom this state has produced few equal to — but he, consul-elect, when a tribune of the plebs had ordered, by his own help, the masters to hold the games against senatorial decree, in his private capacity forbade them, and obtained by authority what he could not yet hold by power. You, when on the Kalends of January the day of the Compitalia fell, allowed Sextus Clodius — who never before had worn the bordered toga — to hold the games and to flit about in it: an impure man, most worthy not only of your face but even of your eye.

\ciceroSection{9} Therefore with these foundations laid for your consulship, three days afterwards, with you looking on and silent, by that fatal portent and prodigy of the commonwealth the Aelian and Fufian Law was overturned — the bulwarks and walls of tranquillity and peace; the colleges restored, not only those which the senate had abolished, but countless new ones from every dregs of the city and slavery were stirred up. By the same man, busy in unheard-of and wicked debaucheries, that ancient mistress of modesty and discipline — the censorship — was abolished, while you meanwhile, the urn of the commonwealth — you who say you were consul at Rome at the time — never signified your opinion by a word in such great wreckings of the state.

\ciceroSection{10} I do not yet say what you did, but what you allowed to be done. Nor does it greatly matter (especially in a consul) whether he himself by ruinous laws or wicked assemblies harasses the commonwealth, or suffers others to harass it. Or can there be any excuse — I shall not say for a consul thinking ill, but for one sitting, hesitating, sleeping in the greatest disturbance of the commonwealth? For nearly a hundred years we had held the Aelian and Fufian Law; for four hundred years the censorship's judgment and notice. These laws no wicked man dared not — though no one indeed could — overturn; and to lessen that power so that we might not be judged on our morals every fifth year, no one so prodigally insolent attempted. These things, O butcher! were buried in the prelude of your consulship.

\ciceroSection{11} Pursue the days that follow these funerals. Before the Aurelian tribunal — not even with you connivant (which itself would be a crime), but with you looking on with cheerier eyes than was your wont — a levy of slaves was held by him who never thought any deed or any suffering shameful for himself. Arms were placed in the temple of Castor — O betrayer of all temples! — with you watching, by that bandit to whom that temple, when you were consul, was a citadel of ruined citizens, a refuge of Catiline's old soldiers, a fortress of forensic banditry, an urn of all laws and religions. Not only my house but the whole Palatine was packed with the senate, with Roman knights, with the entire state, with all Italy — when you, not only to that man — for I leave aside private things, which can be denied; I am recalling what is open — not only, I say, to that man to whom you had given the first vote in your assembly, whom you used to ask third in giving opinion in the senate, did you never approach, but in all the plans which were being prepared for crushing me you not only took part but most cruelly presided.

\ciceroSection{12} What did you yourself dare to say, before my own son-in-law, your kinsman? — That Gabinius was in the most sordid want, that he could not stand without a province, that he had hope from a tribune of the plebs if he would join his counsels with him, but had despaired of the senate; that you were complying with this man's desire — as I had done with my colleague; that there was no reason why I should implore the protection of the consuls; that each man should look to himself. And these things I scarcely dare say. I fear lest there be anyone who does not yet see the marked baseness of the man wrapped under the coverings of his face. Yet I will say it. He himself shall recognise it, and shall, with some pain, recall his own crimes.

\ciceroSection{13} Do you remember, mud, when about the fifth hour I had come to you with Gaius Piso, you came forth, I-know-not-from-what-hovel, with your head wrapped, in slippers, and with that foul mouth of yours had breathed on us a most vile cookshop, you used as excuse your health: that you were said to be wont to be cured by certain wine-soaked drugs? When we had accepted the excuse — for what could we do? — we stood a little while in that smell and smoke of your dives; whence you cast us out, both by answering most insolently and by belching most disgracefully.

\ciceroSection{14} The same man, about that two-day stretch, was led forth into a public meeting by him to whom you offered your consulship as a kind of dagger; when you were asked what you thought about my consulship — a weighty authority, some Calatinus, no doubt, or Africanus or Maximus and not a Caesoninus Semiplacentinus Calventius — you answered, with one eyebrow raised to your forehead, the other lowered to your chin, that cruelty did not please you. Here that man most worthy of your praises praised you. You, gallowsbird, the consul, condemn the senate of cruelty in a public meeting? For not me, who obeyed the senate; for that wholesome and diligent motion was the consul's, but the verdict and judgment of the senate. When you reproach these, you show what kind of consul you would have been at that time, if it had so chanced to befall. By Hercules, you would have judged that Catiline ought to be helped with pay and grain. For what difference was there between Catiline and him to whom you sold the senate's authority, the safety of the state, the whole commonwealth, for the price of a province?

\ciceroSection{15} For what I as consul forbade Lucius Catiline to attempt, the consuls helped Publius Clodius to do. He wished to slay the senate; you abolished it. He wished to burn the laws; you abrogated them. He wished to ruin the country; you struck it down. What was done with you as consuls without arms? That hand of conspirators wished to burn the city; you burned the house of him on whose account the city was not burned. And not even those, if they had had a consul like you, would have thought of burning the city. For they did not wish to deprive themselves of homes — but with these standing they thought there would be no shelter for their crime. They sought the slaughter of citizens; you, slavery. Here you are even crueller. For freedom had been so implanted in this people, before you were consuls, that it should choose to die rather than serve.

\ciceroSection{16} But that twin of Catiline's and Lentulus's plans — that you drove me from my house, you drove Gnaeus Pompeius into his. For neither with me standing and remaining on guard for the city, nor with Pompey, conqueror of all peoples, resisting, did they ever judge they could destroy the commonwealth. Of me you even exacted penalties by which you might appease the dead Manes of the conspirators. All the hatred enclosed in the wicked feelings of the impious, you poured out upon me. To whose madness if I had not given way, on Catiline's pyre, with you as leaders, I should have been slain. What greater proof do you await that there was no difference between you and Catiline, than that you stirred up that same hand from among the half-dead remnants of Catiline; that you gathered all the desperate from every side; that you poured the prison upon me; that you armed the conspirators; that you wished to throw my body, and the life of all good men, against their steel and madness?

\ciceroSection{17} But now I come back to that splendid public speech of yours. Are you that man to whom cruelty is not pleasing? — who, when the senate had voted that its grief and pain should be declared by changing dress, when you saw the commonwealth in mourning by the grief of the most ample order — O our merciful man! — what do you do? What no tyrant in any barbary has done. For I leave aside that — that a consul should issue an edict that the senate's decree be not obeyed; than which nothing fouler can be done or thought. I come back to the mercy of him to whom the senate seemed too cruel in saving the country.

\ciceroSection{18} He dared to issue an edict, with that peer of his (whom yet he wished to surpass in every vice), that the senate, against what it itself had voted, should return to its garb. What tyrant has done this in any Scythia — that those whom he afflicted with grief, he should not allow to mourn? You leave the grief, you take away the marks of grief; you snatch away tears, not by consoling but by threatening. If the conscript fathers had changed their dress not by public counsel but by private duty or pity, even then it would be not to be borne — that this should not be allowed them through the prohibitions of your power. But when the senate in full session had so voted, and all the rest of the orders had already done so before, you, dragged out of the dim cookshop into the consulship with that shaved dancing-girl, forbade the senate of the Roman people to mourn the fall and ruin of the commonwealth. But he was even asking just now of me, what use I had had of his help; why I had not resisted my enemies with my own resources. As if not only I — who have often been a help to many — but anyone has ever been so destitute as to think himself not only safer with that man as champion, but better equipped with him as advocate or witness.

\ciceroSection{19} I, indeed, was wishing to lean on the counsel or the protection of that beast and putrid carcase! From this cast-out corpse I was seeking some help or ornament for myself. I was then seeking a consul — I say a consul — not, indeed, the kind of consul I could not find in this hog, who could defend the commonwealth's so great cause by his weight and counsel; but one who, like a stump and stake, if he had only stood, could yet uphold the title of the consulship. For since that whole cause of mine was consular and senatorial, I had needed the help both of the consul and of the senate. Of which the one had been turned, by you the consuls, even to my own ruin, the other had been wholly snatched from the commonwealth. And yet, if you ask my judgment, I should not have given way, and my country herself would have held me in her embrace, if I had had to fight to the death with that mortuary gladiator and with you and with your colleague.

\ciceroSection{20} For another was the case of that most outstanding man Q. Metellus, whom in my judgment I join with the praise of the immortal gods. Who, with that Gaius Marius, the most brave man and consul and consul for the sixth time, and his unconquered legions, judged he must yield, lest he fight with arms. What contest, then, would it have been for me of this kind? — with Gaius Marius, of course, or with someone equal to him, or with one bearded Epicurean, with the other Catiline's lantern-bearer consul? Nor, by Hercules, did I flee your eyebrow, or your colleague's cymbals, nor was I so timid that I, who had steered the ship of state in the greatest whirlwinds and waves and had set her safe in port, should fear the little cloud of your forehead or your colleague's polluted breath.

\ciceroSection{21} Other winds I have seen, other storms I have foreseen in my mind; with other tempests overhanging I did not yield, but twice offered myself alone for the safety of all. Therefore by my withdrawal then all those wicked swords fell from the most cruel hands, when indeed you, O madman and witless one!, while all good men hidden and shut in mourned, while the temples groaned, while the very roofs of the city wailed, you embraced that funeral creature conceived from wicked debaucheries, from civil bloodshed, from every harshness of crimes, and in the same temple, in the same trace of place and time, you took the verdicts not only of my funeral but of my country's.

\ciceroSection{22} Why should I tell of the banquets of those days, of your gaiety and congratulation, of the most uncontrolled drinking-bouts with your most squalid herds? Who saw you sober in those days, who doing anything worthy of a free man, who in public at all? When your colleague's house resounded with song and cymbals, and he himself naked danced at the banquet (in which when he twirled that dancer's circle, not even then did he fear fortune's wheel) — but this fellow, not so neat a glutton, not so musical, lay in the stench and filth of his Greeks; which (in those mournings of the commonwealth) was reported as a sort of Lapith and Centaur banquet, in which no one can say whether he drank more, or vomited, or poured forth.

\ciceroSection{23} Will you make any mention of your consulship, or dare to say you were consul at Rome? What? Do you think the consulship is in lictors and the bordered toga? Which ornaments you wished even Sextus Clodius to have when you were consul. Do you, you Clodian dog, think the consulship is declared by these badges? It is in mind one ought to be consul; in counsel, in faith, in weight, in vigilance, in care; in short, in the whole duty of the consulship, by guarding every office and (most of all, what the very meaning of the name prescribes) by consulting for the commonwealth. Or shall I think him a consul who did not think the senate was in the commonwealth? And shall I count him consul without that counsel without which not even kings could exist at Rome? For I now leave aside those things. When the levy of slaves was held in the Forum, when arms were brought to the temple of Castor by daylight and openly, and that temple, with the entrance taken away and the steps torn up, was held in arms by the remnants of the conspirators and by Catiline's collusive prosecutor, then his avenger; when Roman knights were exiled, when good men were driven from the Forum with stones, when it was not allowed to the senate even to help the commonwealth or even to mourn it; when the citizen whom this order, with Italy and all peoples assenting, had judged the saviour of the country, was driven by no judgment, no law, no custom, by slavery and arms — I shall not say with your help (which truly may be said), but at least with your silence: shall anyone then suppose there were consuls at Rome?

\ciceroSection{24} What robbers, then, if you are consuls — what raiders, what enemies, what traitors, what tyrants will be named? Great is the name, great the appearance, great the dignity, great the majesty of the consul. The narrowness of your breast does not contain it, that lightness does not receive it, that poverty of soul; the weakness of your talent does not bear, the unfamiliarity with prosperity does not sustain so great a person, so weighty, so severe. Seplasia, by Hercules, as I used to hear it said, as soon as it saw you, refused you as Campanian consul. It had heard of the Decii, the Magii, and had received something of that famous Vibellius Taurea; in whom, if that moderation which is wont to be in our consuls was lacking, yet there was the procession, there was the appearance, there was the gait at least worthy of Seplasia and Capua.

\ciceroSection{25} Indeed if your perfumers had seen Gabinius the duumvir, they would more quickly have recognised him. Those were combed locks, the moist fringes of curls, the flowing and rouged cheeks worthy of Capua — but that old Capua. For this present one overflows with a multitude of most splendid men, most brave men, most excellent citizens, and most friendly to me. Of whom no one looked at you in the bordered toga at Capua but groaned with longing for me — by whose counsel they remembered both the whole commonwealth and that very city had been preserved. They had given me a gilded statue; they had taken me alone as their patron; they reckoned they had life, fortunes, children from me; me both present they had defended by their decrees and embassies against your banditry, and absent — with Gnaeus Pompeius the chief mover, tearing the weapons of your crimes from the body of the commonwealth — they had recalled.

\ciceroSection{26} Or were you then consul, when on the Palatine my house was burning — not by any chance, but by fires thrown on, with you instigating? Was there ever in this city a greater fire to which the consul did not bring relief? But you, at that very time at your mother-in-law's, near my house (whose home you had thrown open for the emptying of mine), sat — not the extinguisher, but the author of the fire, and you yourself almost as consul ministered the flaming brands to the Clodian Furies. Or in the rest of the time did anyone reckon you consul, did anyone obey you, did anyone rise as you came into the curia, did anyone think you should be answered when you put a question? Is that year at last to be reckoned in the commonwealth, when the senate had been struck dumb, when the courts had fallen silent, when good men mourned, when the violence of your banditry flitted through the whole city, and not a single citizen but the very state itself yielded to your and Gabinius's crime and madness?

\ciceroSection{27} Not even then, muddy Caesoninus, did you emerge from the most wretched dregs of your nature, when at last the wakened virtue of the most distinguished man swiftly demanded both his true friend and the citizen who had deserved best, and his own former character. Nor did that man suffer the pestilence of your crimes to delay any longer in that commonwealth which he himself had adorned and increased. Yet that other, such as he is — overcome in baseness by you alone — Gabinius, just barely collected himself; but he did collect himself, and against his Clodius first feignedly, then unwillingly, at last yet for Gnaeus Pompey truly and vehemently fought. In which spectacle there was wonderful equity of the Roman people. Whichever of them perished, the people, like the lanista in such a pair, thought it gain — but immortal gain if both should fall.

\ciceroSection{28} But he yet was doing something; he was protecting the authority of the highest man. He himself was wicked, was a gladiator, but he was fighting with the wicked and with an equal gladiator. You, of course, religious and holy man, were unwilling to break the bond which by my blood you had struck in the bargain of the provinces. For that sister-tied adulterer had stipulated for himself that, if he gave you the province, the army, the money snatched from the bowels of the commonwealth, you should furnish yourself the comrade and helper of all his crimes. So in that tumult fasces were broken, he himself was struck, daily there were weapons, stones, flights; and at last the man was caught with a dagger heading for the senate, who, all agreed, had been posted to kill Gnaeus Pompey.

\ciceroSection{29} Did anyone hear, not only any motion or report, but any voice or complaint of yours at all? Do you think yourself to have been consul, when, while you held the imperium, he who had saved the commonwealth by the senate's authority, who had bound up all the parts of all peoples in three triumphs, declared he could not stand in safety in public, in fact, in Italy? Were you then consuls when, on whatever subject you began to say a word or to refer to the senate, the whole order cried out against you and showed you would do nothing unless you had first referred about me? When you, although bound by treaty, yet said you wished to do so, but were hindered by a law.

\ciceroSection{30} A law which to private men did not seem to be a law, branded by slaves, cut by violence, imposed by banditry, with the senate set aside, with all good men driven from the Forum, with the commonwealth captured, against all laws, written by no custom — that those who said they feared this should be consuls, the very calendars cannot bear, not to say the minds of men. For if you did not think that a law, which was, against all laws, the tribunician proscription of an unconvicted citizen and his unforfeited rights and goods — and yet you were bound by a bargain — who would think you not only consuls but free men, whose mind was crushed by reward, whose tongue was bound by hire? But if you alone thought it a law, would anyone reckon you then consuls or now consulars, who do not know the laws, the institutions, the morals, the rights of that state in which you wish to be in the number of the chief men?

\ciceroSection{31} Or when you set out for the bought-or-snatched provinces in your military cloaks, did anyone reckon you consuls? So, I suppose, if not by their crowded numbers in adorning and celebrating your departure, at least with good omens as for consuls, not with most gloomy ones as for enemies and traitors, did men accompany you. Did you also dare, you most monstrous and foul prodigy, to set my own departure — that witness of your wickedness and cruelty — in the place of curse and insult? At which time, conscript fathers, I gathered the immortal fruit of your love and judgment toward me. You, not by murmur but by voice and shout, broke that abandoned and half-living man's frenzy and impudence.

\ciceroSection{32} Will you place the senate's grief, the longing of the equestrian order, the squalor of Italy, the year-long silence of the curia, the lasting silence of the courts and Forum — and other such wounds which my departure inflicted on the commonwealth — in the place of insult? Which, if it had been the most calamitous, was yet rather worthy of pity than of contempt, and would be reckoned joined rather with glory than with disgrace; and at most that grief would have been mine, the wickedness and dishonour your own. But indeed — perhaps what I am about to say will seem wonderful to hear, but I shall surely say what I feel — when I have been affected with such great kindnesses by you, conscript fathers, and with such honours, I do not only not reckon that a calamity, but, if anything can be separated from the commonwealth in my regard (which can scarcely be), I judge that fortune was to be wished for and sought by me, privately, for the increase of my name.

\ciceroSection{33} And to compare your most joyful day with my most gloomy: which, do you think, is more to be desired by a good and wise man — to leave one's country in such a way that all his fellow citizens pray for safety, soundness, return (which happened to me) — or, what befell you on setting out, that all should curse, pray ill, wish that road of yours alone and unending? To me, by the gods, in such hatred of all mortals (and that just and owed), any flight would be more desirable than any province. But go on. For if that most turbulent time of mine surpasses your most tranquil, what shall I match the rest, which were full of dishonour to you, of dignity to me?

\ciceroSection{34} Me on the Kalends of January — the day which first dawned on the commonwealth after our destruction and ruin — the most crowded senate, with all Italy gathering, with the most distinguished and brave man, P. Lentulus, putting the motion, with one consenting voice, recalled. Me the same senate commended to foreign nations, me to our envoys and magistrates by its authority and by consular letters, not (as you, you Insubrian, dared to say) bereft of country, but, as the senate at that very time named me, ``citizen and saviour of the commonwealth.'' For my own single safety the senate thought the help of all citizens, of all from Italy who wished the commonwealth safe, must be implored by the consul's voice and letters. For the saving of my person, all Italy at one time, as if at a signal given, came to Rome. About my safety the speeches of P. Lentulus, that most outstanding man and best consul, of Cn. Pompey, the most distinguished and unconquered citizen, and of the rest of the leading men of the state were thronged and most thankful.

\ciceroSection{35} About me the senate so decreed, with Cn. Pompey as author and prince of the motion, that, if anyone had hindered my return, he should be reckoned in the number of enemies; and the senate's authority concerning me was declared in such words that to no one is a triumph more honourably written up than my safety and restoration. About me, when all the magistrates had promulgated the bill except one praetor (from whom it could not have been demanded — the brother of my enemy), and except two tribunes bought from the auction-block, the consul P. Lentulus carried a law in the comitia centuriata, on the motion of his colleague Q. Metellus, whom that same commonwealth which in his tribunate had divided from me, in his consulship, by the virtue and wisdom of that excellent and most just man, joined.

\ciceroSection{36} How that law was received, why need I say? From you I hear that no excuse to any citizen seemed sufficiently just for non-attendance; that in no comitia ever had there been so great a multitude of men, or one more splendid. This certainly I see — what the public records show — that you were the proposers, you the distributors, you the watchers of the tablets; and what in the honours of your kinsmen you do not do, by either the excuse of age or of office, that, in my safety, with no one asking, you did of your own accord.

\ciceroSection{37} Compare now, our Epicurus brought from the pigsty, not from the school — compare, if you dare, your absence with mine. You held the consular province, with the boundaries that the law of your desire — not those that the law of your son-in-law — had set. For by Caesar's most just and most excellent law, the free peoples were plainly and truly free; but by that law which no one but you and your colleague thought a law, all Achaia, Thessaly, Athens, the whole of Greece was made over to you. You had as great an army as not the senate or the Roman people had given you, but as your own lust had enrolled. You had emptied the treasury.

\ciceroSection{38} What deeds did you do with the imperium, the army, the consular province? What deeds he did, I ask! He, who as soon as he came — I do not yet record his rapines, I do not declare the moneys exacted, taken, levied, the slaughter of allies, the killing of guests, the perfidy, the savagery, the crimes — soon, if it shall please, I shall argue with him as with a thief, as with a temple-robber, as with an assassin. Now I shall set my plundered fortune against the flourishing fortune of an imperator. Who ever held a province with an army who sent no letters to the senate? But to hold so great a province with so great an army — Macedonia, especially, which so many barbarian peoples border that for the Macedonian generals the boundaries of the province have always been the same as those of swords and pikes; from which several have returned with praetorian command, but none with consular has returned safe without triumphing! This is new; that is much more so. This vulture of that province has been called — if it pleases the gods — \textit{imperator}.

\ciceroSection{39} Not even then, our Paulus, did you dare send dispatches with laurel to Rome? ``I sent them,'' he says. Who ever read them out, who demanded that they be read? For it now matters nothing to me whether you, oppressed by consciousness of your crimes, dared write nothing to the order which you had despised, struck down, destroyed; or whether your friends hid your dispatches and by their silence condemned your rashness and audacity. And I do not know whether I would rather you should seem to have had no shame in sending dispatches, while your friends had more shame and counsel — or that you should seem to have been more modest than you are wont, or that your deed should not have been condemned by your friends' judgment.

\ciceroSection{40} If you had not by your wicked insults to this order shut the curia against yourself for ever, what at last had been done or carried in your province about which you ought to have written to the senate with any congratulation? The vexation of Macedonia? Or the shameful loss of towns? Or the plundering of allies? Or the ravaging of fields? Or the fortifying of Thessalonica? Or the besieging of the military road? Or the destruction of our army by sword, hunger, cold, plague? You who wrote nothing to the senate, as in the city you were found worse than Gabinius, so in the province you have been a little more downcast than he.

\ciceroSection{41} For that gulf and glutton, born for his belly, not for praise and glory — when he had stripped the Roman knights in the province, when the publicans (joined to us by goodwill and by dignity) of all their fortunes, many of life and reputation, when he had done nothing else with that army save depopulate cities, lay waste fields, drain houses — dared (for what would he not dare?) to demand a thanksgiving from the senate by letters. O immortal gods! Do you also, you twin gulfs and rocks of the commonwealth, depress my fortune, raise yours? — when about me such senatorial decrees were made in my absence, such public meetings were held, there was such movement of all municipia and colonies, such decrees of the publicans, of the colleges, of all kinds and orders — which I should never dare wish for or even think of — and you have undergone everlasting marks of the foulest disgrace?

\ciceroSection{42} Or, if I should see you and Gabinius nailed on the cross, would I be affected with greater joy from the lacerating of your bodies than I am affected from your fame? No punishment is to be reckoned, by which good men too and brave can by some chance be afflicted. And these your voluptuous Greeks even say so. Whom would that you so listened to as they should have been listened to: never would you have plunged yourself into so many disgraces. But you listen at the trough, you listen in debauchery, you listen in food and wine. But these very men say, who define evil by pain, good by pleasure, that the wise man — even if shut up in Phalaris's bull and roasted with kindled fires — will yet say it is sweet, and is not in the least disturbed. Such force of virtue did they wish to lay down, that the good man could never not be blessed.

\ciceroSection{43} What is the punishment, then, what the chastisement? That, in my judgment, which can fall on no one save the guilty: the wickedness undertaken, the mind hindered and crushed, the hatred of good men, the marks branded by the senate, the loss of dignity. Nor does that M. Regulus seem to me afflicted with punishment, whom the Carthaginians, with eyelids cut off, bound on a machine, killed by sleeplessness; nor C. Marius, whom Italy, saved by him, saw plunged in the marshes of Minturnae; whom Africa, conquered by him, saw driven out and shipwrecked. For these are the weapons of fortune, not of fault. But punishment is the penalty of sin. Nor would I, if ever I should pray ill for you (which I have often done, and in which the gods have heard my prayers), pray for sickness or death or torment. That Thyestean curse is the poet's, that moves the hearts of the common people, not of the wise — that you, ``\textit{cast out by shipwreck, stuck somewhere on rough rocks, with bowels torn, should hang upon the side, scattering the rocks with gore, blood and black slime}.''

\ciceroSection{44} I should not bear it ill, if it had so happened; but that would still be human. M. Marcellus, who was three times consul, of the highest virtue, piety, military glory, perished at sea; who yet, on account of his virtue, lives in glory and praise. That death is to be reckoned in some kind of fortune, not in punishment. What is the punishment then, what the chastisement, what rocks, what crosses? — that two leaders should be in provinces of the Roman people, should have armies, should be called \textit{imperatores}, of whom the one was so unbridled by consciousness of his crimes and frauds that, from that province which has been of all the most triumphal, he dared send no letter to the senate; from which province a man most adorned in every dignity, L. Torquatus, with great deeds done, was called \textit{imperator} by the senate on my motion, whence in these few years we have seen the most just triumphs of Cn. Dolabella, C. Curio, M. Lucullus — from this province, with you as imperator, no message has been brought to the senate. From the other a letter has been brought, read out, brought before the senate.

\ciceroSection{45} Immortal gods! Should I have so wished — that my enemy be branded with that disgrace with which no one ever was; that the senate, which has now come into the habit of such kindness as to honour those who have managed the commonwealth well with new honours and the number of days and kind of words, should not believe the dispatches of this one man announcing things, should refuse to him demanding them? In these things I feed, I delight, I take full pleasure: that this order thinks of you no otherwise than of the keenest enemies; that the Roman knights, that the rest of the orders, that the whole state hates you; that no good man, no citizen finally — who only remembers that he is a citizen — there is, who does not flee you with his eyes, reject you with his ears, despise you with his soul, indeed shudder at the very recollection of your consulship.

\ciceroSection{46} These things I have always sought of you, these I have wished, these I have prayed for. More even has happened than I wished. For that you should lose the army I never, by Hercules, wished. That too has come about beyond my wish — but greatly to my pleasure. For it had never come into my mind to wish for you the madness and frenzy into which you have fallen. And yet, that should have been wished. Yet it had escaped me that the immortal gods have these surest punishments for the wicked and abandoned. Do not so suppose, conscript fathers, as you see on the stage, that wicked men are terrified by the impulse of the gods with the burning torches of the Furies. Each man's own fraud, his own crime, his own wickedness, his own audacity hurls him from sanity and mind. These are the Furies of the wicked, these the flames, these the torches.

\ciceroSection{47} Should I not think you mad, frenzied, mind-stripped, more mind-spent than that tragic Orestes or Athamas — you who first dared to do (for that is the head of the matter), and then a little while ago, with Torquatus, that most holy and weighty man, pressing you, to confess that you had left the province of Macedonia, into which you had transported so great an army, without a single soldier? I leave aside the loss of the greater part of the army; let that be of your bad luck. But what cause can you bring of disbanding the army? What power had you, what law, what senatorial decree, what right, what precedent? What else is it to rave? — not to know men, not to know laws, not the senate, not the state? To bloody one's body is a small thing; greater is this wounding of one's life, fame, safety.

\ciceroSection{48} If you had dismissed your household — which would have concerned no one but yourself — your friends would have thought you must be put under restraint. The protection of the commonwealth, the guardianship of the province, you dismissed without orders of the Roman people or senate, if you had been in your right mind? Lo, the other of you, having now poured out the greatest plunder he had drawn from the fortunes of the publicans, from the fields and cities of allies — when bottomless lusts had devoured part of that plunder, and a kind of new and unheard-of luxury part, and part also (in those very places where he plundered everything) the purchases for building this Tusculan mountain. When he was now in want, when that intolerable building of his had been left at a stand — he sold himself, his fasces, the army of the Roman people, the divine prohibition of the immortal gods, the answers of the priests, the senate's authority, the orders of the Roman people, the name and dignity of the empire — to the king of Egypt.

\ciceroSection{49} When the boundaries of the province he had had as great as he had wished, as he had desired, as by the price of my person and danger he had bought, he could not stay within them. He led the army out of Syria. How was that lawful, outside the province? He offered himself a hireling companion to the Alexandrian king. What is more shameful than this? He came into Egypt, he engaged with the Alexandrians. When had this order or this people undertaken this war? He took Alexandria. What else do we await from his madness, save that he send a letter to the senate about such great deeds done?

\ciceroSection{50} If this man were of his own mind, if he were not paying the heaviest penalties to his country and to the immortal gods through madness and frenzy, would he have dared (I leave aside going forth from his province, leading out the army, waging war on his own initiative, entering a kingdom without orders of the Roman people or senate — what very many old laws, and most plainly the lex Cornelia on majestas and the lex Iulia on extortion, forbid) — but I leave these aside. Would he, if he were not raging most keenly, dare to claim for himself a province which P. Lentulus, most friendly to this order, when he had it both by senatorial authority and by lot, had laid down with no hesitation, on a religious scruple, when, even if no scruple had hindered, the custom of the ancestors and examples and the heaviest punishments of the laws would yet have forbidden it?

\ciceroSection{51} Since we have begun a contention of fortunes, let us leave aside Gabinius's return — whose face, although he himself has cut himself off, I yet wait to see. Let us, if it pleases, compare your return with mine. Mine indeed was such that, from Brundisium right up to Rome, I saw a continuous procession of the whole of Italy. For there was no region, no municipality, no prefecture or colony from which men did not come publicly to congratulate me. Why should I tell of my arrivals, of the pourings-out of men from the towns, of the gatherings from the fields of heads of households with their wives and children, of those days which, as the festivals and solemnities of the immortal gods, were celebrated by all on the occasion of my coming and return?

\ciceroSection{52} That single day was to me the equal of immortality, on which I returned to my country, when I saw the senate come forth and the entire Roman people, when Rome herself, almost shaken from her seats, seemed to me to come forward to embrace her saviour. She so received me that not only all men and women of every kind, age, order, of every fortune and place, but even the very walls and roofs and temples of the city seemed to rejoice. Me on the days following, in that very house from which you had cast me, which you had plundered, which you had burnt, the pontifices, the consuls, the conscript fathers placed; and they decreed me — what before me to none — that the house should be rebuilt at public cost.

\ciceroSection{53} You have my return. Compare now in turn your own — since indeed, with the army lost, you brought home unhurt nothing save that old face of yours. Who knows by what way you came with your laurelled lictors? What \textit{Maeanders} did you twist along, while you sought every solitude? What turnings and bendings did you seek? What municipality saw you? What friend invited you? What guest looked on you? Was not night for you in the place of day, solitude in the place of crowds, a tavern in the place of a town, so that you seemed to be brought back not as a noble \textit{imperator} returning from Macedonia, but as a dead infamous one? But Rome itself, O disgrace of the family — I shall not say of the Calpurnii but of the Calventii — not of this city but of the Placentine municipality, not of your father's race but of trousered kindred — how did you enter? Who came to meet you, of these here, or of the rest of citizens — I shall not say — but of your own legates?

\ciceroSection{54} For with me was L. Flaccus, a man most unworthy of your legateship and worthier of those counsels by which he was joined with me in my consulship for saving the commonwealth — with me at the time when someone was telling that you had been seen wandering with your lictors not far from the gate. I know likewise that an outstanding brave man, expert in war and military affairs, my friend Q. Marcius — by the work of which legates you, as imperator, had been hailed in battle, when you had been far away — was at home, idle, on that arrival of yours.

\ciceroSection{55} But why should I count those who did not come to meet you? Indeed I say almost no one came, not even of the most dutiful candidate-class, although they had been generally admonished and asked, both that very day and many days before. Some little togas were ready for the lictors at the gate; receiving these, they threw off their cloaks, offered a new procession to their imperator. So this man, after such an army of so great a province, three years afterwards, brought himself into the city as Macedonian imperator in such a way that no most obscure trader's return was ever more deserted. In which yet that man, who is ready to defend himself, has rebuked me. When I had said he had entered by the Caelimontane gate, the bold man challenged me with a wager that he had not entered by the Esquiline. As if either I should have known that, or any of you had heard, or it mattered by what gate you entered — provided not by the triumphal one, which was always open to Macedonian proconsuls before you. You have been found who, endowed with consular imperium, did not triumph from Macedonia.

\ciceroSection{56} But you have heard, conscript fathers, the philosopher's voice. He has said he was never desirous of a triumph. O wickedness, O pestilence, O ruin! When you were extinguishing the senate, selling the authority of this order, handing over your consulship to the tribune of the plebs, overturning the commonwealth, betraying my person and safety for the one price of the province — if you were not desirous of a triumph, with what desire of what thing will you say you burned? For I have often seen those who seemed to me and to others more desirous of a province cover and conceal their desire by the name of a triumph. This D. Silanus the consul said in this order; this even my colleague used to say. For no one can desire and openly seek an army without setting before him the desire of a triumph.

\ciceroSection{57} If the senate and Roman people had compelled you, not seeking, even refusing, to undertake the war and lead the army, it would yet be the part of a narrow and downcast soul to despise the honour and dignity of a just triumph. For as it is the part of a frivolous mind to fish for empty rumour and to pursue all the shadows even of false glory, so it is the part of a soul fleeing the light and splendour to reject just glory, which is the most honourable fruit of true virtue. But when not only without the senate's demanding and compelling, but with the senate unwilling and crushed, not only with no zeal of the Roman people, but with no one giving a free vote, that province was for you the price of the state overthrown and ruined by you, and when this was the bargain of all your crimes — that, if you handed over the whole commonwealth to wicked bandits, Macedonia should be handed over to you with what boundaries you wished — when you were emptying the treasury, when you were robbing Italy of its youth, when you crossed the most vast sea in winter, if you despised a triumph, what was carrying you off, you most senseless robber, save the so blind desire of plunder and rapine?

\ciceroSection{58} Cn. Pompey can no longer use your counsel; for he has erred. He had not tasted that philosophy of yours. Three times now the foolish man has triumphed. Crassus, I am ashamed for you. What of the fact that, when by you the most fearful war had been brought to an end, you wished that laurel crown to be decreed to you so greatly by the senate? P. Servilius, Q. Metellus, C. Curio, L. Afranius, why did you not hear this so learned, so erudite man, before you were led into that error? For C. Pomptinus himself, my kinsman, it is no longer in his power either; for he is hindered by religious obligations undertaken. O foolish Camilli, Curii, Fabricii, Calatini, Scipios, Marcelli, Maximi! O senseless Paulus, rustic Marius, fathers of these our two consuls of no counsel, who triumphed!

\ciceroSection{59} But since we cannot change the past, why does this little manikin — this Epicurus made of clay and mud — delay to give these brilliant precepts of wisdom to that most distinguished and supreme imperator, his son-in-law? That man, believe me, is borne by glory; he burns, he is ablaze with desire of a just and great triumph. He has not learned those same things you have. Send him a little book — and, if you can already meet him in person, plan with what words you may compress and extinguish his kindled desire. You will avail with him, a man flying with desire of glory, you a moderate and constant man; with the unlearned, you the learned; with the son-in-law, the father-in-law. For you will say (as you are a man made for persuading, neat, perfect, polished from the school): ``What is it, Caesar, that the so-often decreed thanksgivings of so many days now please you so greatly? In which men are led by error, and which the gods neglect — who, as our divine Epicurus said, are not wont to be either propitious or angry to anyone.'' You will not, of course, gain credit when you argue this; for he will see that they are angry both at you and have been.

\ciceroSection{60} You will turn yourself to the other school. You will discourse on the triumph: ``What does that chariot have, what those leaders bound before the chariot, what the images of towns, what the gold, what the silver, what the legates on horseback and the tribunes, what the shouting of soldiers, what that whole pomp? These are empty, believe me, almost children's playthings — to catch applause, to be carried through the city, to wish to be looked at. From which there is nothing solid you can hold, nothing you can refer to bodily pleasure.''

\ciceroSection{61} ``Why do you not look at me — who, from the province from which T. Flamininus, L. Paulus, Q. Metellus, T. Didius, countless others, moved by lightness and desire, triumphed, returned in such a way that at the Esquiline gate I trampled the Macedonian laurel, that I myself with fifteen ill-clad men came thirsting to the Caelimontane gate? In which place a freedman had hired a house for me — that splendid imperator! — two days before that day. And if it had not been free, in the Campus Martius I should have set up my tent. Meanwhile, Caesar, with the triumphal litters neglected, my money is and shall be at home. I delivered the accounts to the treasury at once, as your law bade, nor did I obey your law in any other thing. If you should learn these accounts, you would understand that letters profited no one more than me. For they are so neatly and lettered written that the scribe at the treasury who returned them, when the accounts had been written out, rubbed his head with his left hand and muttered to himself: `The reckoning, by Hercules, is plain — the silver \textit{οἴχεται} — has gone.' '' By this speech I do not doubt you can call him back even as he is now ascending the chariot.

\ciceroSection{62} O darkness, O mud, O filth, O forgetful of your father's race, scarcely mindful of your mother's! That broken, lowly, downcast, sordid quality of yours is below what would seem worthy even of the Mediolanian crier, your grandfather. L. Crassus, the wisest man of our state, with magnifying glasses almost combed the Alps, that, where there was no enemy, he might find some cause for a triumph. With the same desire that great-talented man C. Cotta burned, with no certain enemy. Neither of them triumphed: one because the colleague snatched that honour from him, the other because death did so. The desire to triumph of M. Piso has just now been mocked by you, from which you said you greatly recoiled. He, although he had waged a less great war, as you said, yet did not think that honour to be despised. You, more learned than Piso, more prudent than Cotta, richer in counsel, talent, wisdom than Crassus, despise what those amateurs (as you call them) reckoned splendid.

\ciceroSection{63} If you reproach them for being desirous of the laurel crown when they had waged either small or no wars, you, with such peoples subdued, with such great deeds done, ought least to despise the fruit of your labours, the rewards of your dangers, the marks of your virtue. Nor have you indeed despised them, granted you be wiser than Themista, but you did not wish your iron face to be flogged by the senate's reproof. Now you see — since indeed I have been so my own enemy as to compare myself with you — that both my departure and my absence and my return have far excelled yours, so that those things have given me undying glory, while they have inflicted on you everlasting shame.

\ciceroSection{64} Are you, even in this daily, constant, urban life, going to set your splendour, your favour, your domestic celebrity, your forensic work, your counsel, help, authority, your senatorial vote, ahead of mine — or, to speak more truly, of any man's, even the lowest and most despised? Come now, the senate hates you (which you justly grant it does) as the destroyer and ruiner not only of its dignity and authority, but altogether of its order and name. The Roman knights cannot bear to look at you, from whose order a most outstanding and adorned man, L. Aelius, was in your consulship banished. The Roman plebs wishes you destroyed, on whose ill-name you have laid those things you had done about me through bandits and slaves. All Italy curses you, whose decrees and prayers you most haughtily rejected.

\ciceroSection{65} Make trial of this so great and so universal hatred, if you dare. The most splendidly equipped and most magnificent games are at hand, since men's memory — the like of which not only have never been, but I cannot even guess by what means could be hereafter. Give yourself to the people, commit yourself to the games. Do you fear the hiss? Where are your schools? Do you fear an outcry? — that is not even the philosopher's care. That hands be laid on you? — pain is an evil, as you argue; reputation, dishonour, infamy, baseness — words and trifles. But about this I do not doubt: he will not dare to come to the games. He will not enter a public banquet for the sake of dignity, unless perhaps to dine with P. Clodius — that is, with his loves — but plainly for his own pleasure. The games he will leave to us amateurs.

\ciceroSection{66} For he is wont in his disputations to set the pleasures of the belly before the delight of the eyes and ears. For although to you he seems only wicked, cruel, once a small thief, now even greedy, sordid, defiant, proud, deceitful, perfidious, impudent, audacious — know that there is nothing more luxurious, nothing more lustful, nothing more reckless, nothing more worthless. But do not think this luxury in him as you would think.

\ciceroSection{67} For there is one luxury which, although wholly faulty and base, is yet more worthy of a free-born and free man. With him there is nothing fine, nothing elegant, nothing exquisite — I shall praise an enemy — nay, scarcely anything sumptuous save lusts. No engraved silver; the largest cups, and those (lest he seem to despise his own people) Placentine; the table heaped not with shellfish or fish, but with much half-rotten meat. Squalid slaves serve, some even old men. The same man cook, the same hall-keeper. No baker at home, no wine-cellar; bread and wine from the petty dealer and from the cask; Greeks crammed five to a couch, often more; he himself alone. The drinking goes on as long as there is something to ladle from the jar. When he has heard the cock crow, he thinks his grandfather has come back to life: he orders the table cleared.

\ciceroSection{68} Someone will say: ``Whence are these things known to you?'' By Hercules, I shall not, for the sake of insult, describe anyone — especially an ingenious and learned man, with whose race I cannot, even if I should wish, be angry. There is a certain Greek who lives with Piso, a man (truly to say, for so I have known him) civilised, but only as long as he is either with others, or with himself. He, when he had seen this young man with that even then god-cursed face, did not despise his friendship, though he was indeed sought after; he gave himself to such intimacy that he plainly lived with him as one and scarcely ever departed from him. I am not speaking before the unlearned but, as I think, in a gathering of the most learned and humane men. You have surely heard the Epicurean philosophers said to measure all things to be sought by man by pleasure: rightly or otherwise, nothing to us — or, if to us, nothing to this time. But yet it is a slippery kind of speech for a young man not keenly understanding, and often headlong.

\ciceroSection{69} So this stallion, as soon as he heard pleasure being so greatly praised by the philosopher, did no fishing into it, but so roused all his pleasure-loving senses, so neighed in answer to that man's speech, that he reckoned not the master of virtue but the author of lust had been found by him. The Greek at first distinguishing and dividing how those things were said: this man limping (as they say) at the ball, holding tight to what he had received, calling on witnesses, wishing to seal it on tablets, judging that Epicurus spoke clearly. He says, I think, that he can understand no good with bodily pleasures removed.

\ciceroSection{70} What more? The pliable and most graceful Greek did not wish to be too pugnacious against the Roman people's general. But this man of whom I speak is polished not only in philosophy but in those other studies which most say the Epicureans neglect. He makes a poem so festive, so neat, so elegant, that nothing more witty can be made. In which one may reproach him, if anyone wishes, only lightly — not as wicked, not as audacious, not as impure, but as a Greekling, as a flatterer, as a poet. He fell, however — or rather slipped — into this man, deceived by the same eyebrow as so many wise men and so great a state. He could not call himself back, entangled by familiarity, and at the same time feared the report of inconstancy. Asked, invited, compelled, he wrote so many things to him, and even about him, that he expressed all his lusts, all his debaucheries, all the kinds of dinners and banquets, finally his adulteries, in his most dainty verses,

\ciceroSection{71} in which, if anyone wishes, he might gaze upon his life as in a mirror. Of which much, read and heard by many, I should recite, did I not fear that this very kind of speech in which I am now using would shrink from this place's custom; and at the same time I do not wish anything to be taken from the man who wrote them. Who, if he had been of better fortune in choosing a pupil, perhaps could have been more austere and weighty. But chance led him into this practice of writing very unworthy of a philosopher, since philosophy, as is reported, contains the discipline of virtue and duty and living well; and he who professes it seems to me to sustain the gravest character.

\ciceroSection{72} But the same chance, with him not knowing what he was professing, when he said he was a philosopher, polluted him with the mud and filth of that most foul and intemperate beast. Who, when just now he had praised the deeds of my consulship — which praise of a most base man was almost shameful to me myself — said: ``It was not that envy that hurt you, but your verses.'' Too great a punishment, in your view, was set up in your consulship for me, whether as a bad poet or as a free man. For I wrote: \textit{cedant arma togae} — ``Let arms yield to the toga.'' What then? This thing has stirred up those waves for you. But this, I think, was nowhere written in that elegy which in your consulship was carved on the tomb of the commonwealth: ``\textit{Do you wish, do you order} that, because M. Cicero made a verse'' — but ``because he set things right.'' Yet still,

\ciceroSection{73} since we have you not as Aristarchus but as Phalaris the grammarian — who put no mark to a bad verse, but pursued the poet with arms — I wish to know what at last you reproach in this verse: ``Let arms yield to the toga.'' ``You say,'' he says, ``that the highest imperator was to yield to your toga.'' Should I now, ass, teach you letters? There is no need of words but of cudgels. I did not say of \textit{this} toga which I am wearing, nor of arms, the shield or sword of one general; but, because the toga is the sign of peace and leisure, while on the other hand arms are of tumult and war, I, then speaking in the poets' way, wished this to be understood: that war and tumult would yield to peace and leisure.

\ciceroSection{74} Ask your familiar, that Greek poet of yours; he will approve the very kind and recognise it, nor wonder that you have no sense. ``But on the other line,'' he says, ``you stick: \textit{concedat laurea laudi} — `Let the laurel yield to praise.''' By Hercules, I owe you thanks; for I should be stuck if you had not freed me. For when you, fearful and trembling, with your own most thievish hands, snatched the laurel from the bloody fasces and threw it down at the Esquiline gate, you judged that the laurel had yielded — not only to the most ample, but even to the smallest praise. And by that speech you wish, criminal, this to be understood: that Pompey was made my enemy by that verse, so that, if the verse hurt me, the destruction would seem to have been sought for me by him whom that verse had offended.

\ciceroSection{75} I leave aside that the verse pertained to him in nothing; that it was not my part, when I had as much as I could decked him with many speeches and writings, to violate him by one verse. But suppose he was first offended; did he not compensate with one little verse for so many of my volumes of his praises? But if he had been moved, would he have been so cruel against the life — I shall not say of his closest friend, of one who had not so deserved of his praise, of the commonwealth, of a consular, of a senator, of a citizen, of a free man — for the sake of a verse? Do you understand what you are saying, before whom you say it, of whom you speak? You wish to embrace the most ample men in your and Gabinius's wickedness, and that not secretly. For a little while ago you said I was contending with those whom I despised, not touching those who could do more, against whom I ought to be angry. Of whom — for who does not understand whom you mean? — although the cause is not the same in all, yet that of all is approved by me.

\ciceroSection{76} Cn. Pompeius, with many opposing his zeal and love towards me, has always loved me, has always judged me most worthy of his association, has always wished me not only safe but most ample and adorned. Your frauds, your wickedness, your charges of my plots, of his perils — wickedly feigned — together with those of the men who, by the licence of intimacy, had set up by your impulse a home of their most wicked gossip in his ears; your desires of provinces — these brought it about that I was shut out, and that all who wished me, who wished his glory, who wished the commonwealth, to be safe, were kept from speech and access;

\ciceroSection{77} by which things it was brought about that he was not allowed plainly to stand by his own judgment, when certain men had not estranged his zeal from me, but had delayed his help. Did not L. Lentulus (who was then praetor) come to you, did not Q. Sanga, did not L. Torquatus the elder, did not M. Lucullus? Who all came to him, and many other mortals, and to entreat and beseech him at his Alban estate that he should not abandon my fortunes, joined as they were with the safety of the commonwealth. Whom he sent on to you and to your colleague, that you should undertake the public cause, that you should bring the matter before the senate; that he was unwilling without public counsel to fight against an armed tribune of the plebs; that, if the consuls defended the commonwealth by senatorial decree, he would take up arms. Do you, unhappy one, recall what you answered?

\ciceroSection{78} In which all of them indeed, but Torquatus beyond the rest, raged at the contumacy of your reply: that you were not so brave as Torquatus himself had been in his consulship, or as I had been; that there was no need of arms, no need of strife; that I could save the commonwealth a second time, if I had given way; that there would be infinite slaughter, if I had resisted. Then at the last, neither he nor his son-in-law nor his colleague would fail the tribune of the plebs. Here you, an enemy and traitor, say that I ought to be more hostile to others than to you? I know that

\ciceroSection{79} C. Caesar did not feel about the commonwealth as I did. But yet — what I have often said before of him in these men's hearing — he wished me to be the partner of his whole consulship and of those honours which he shared with his nearest. He offered, invited, asked me. I was not led, perhaps from too great desire of constancy, to that cause. I did not demand to be most dear to one to whose kindnesses I had not yielded my opinion. The matter was thought to be brought into contest in your consulship — whether the things he had done in the previous year would stand or be rescinded. Why should I say more? If he thought there was so much strength and virtue in me alone that what he himself had done would fall, if I had resisted, why should I not pardon him, if he set his own safety before mine?

\ciceroSection{80} But I leave the past. As Cn. Pompeius embraced me with all his zeal, his labours, the dangers of his life — when he travelled to the municipalities for me, implored the faith of Italy, sat constantly by P. Lentulus the consul, the author of my safety, made good the senate's vote, professed himself in public meetings not only the defender of my safety but a suppliant for me — to the will of this man (whom he understood could do much, whom he had recognised as not unfriendly to me) he joined as a partner and helper, C. Caesar. Now you see that to you I owe to be not unfriendly but a hostile enemy, while to those whom you describe I owe to be not only not angry but a friend; of whom one, what I shall remember, has always been as much my friend as his own; the other, what I shall forget, was sometimes more his own friend than mine.

\ciceroSection{81} Then this happens — that brave men, even if they have fought hand-to-hand with the sword, lay aside that hatred of contention together with the very fight and arms. Nor could that man ever hate me, not even at the time when we differed. Virtue (which you do not even know by face) has this: that brave men are delighted by its appearance and beauty even when set in an enemy. For my part I shall say what I feel from my soul, conscript fathers, and what I have often said before in your hearing. If C. Caesar had never been my friend, if always angry, if always he should reject my friendship and offer himself implacable and inappeasable to me — yet to him, having done such great deeds and doing them daily, I could not but be a friend. Whose imperium I set against and oppose to the ascent and crossing of the Gauls — not the wall of the Alps; against the most monstrous peoples of the Germans — not the trench of the Rhine, overflowing with those gulfs.

\ciceroSection{82} He has so brought it about that, if the mountains had subsided, the rivers had dried up, we should have Italy fortified, not by the protection of nature but by his victory and deeds. But when he seeks me, loves me, thinks me worthy of every praise — will you call me from your enmities to an old quarrel, will you so by your crimes rub afresh the past dooms of the commonwealth? Which indeed you, who well knew my conjunction with Caesar, mocked, when with quite trembling lips you asked of me why I did not lay an information against you. Although, as far as I am concerned, I shall never lessen that care of yours by a denial — yet I must consider how much trouble and burden I, the most friendly of men, would lay on him, hindered by such great affairs of the commonwealth and so great a war. Nor yet do I despair, although the youth grows slack and is not as it ought to be in desire of praise and glory, that there will be some who will not refuse to strip this cast-off carcase of the consular spoils — especially with the defendant so afflicted, so destitute, so weak, so nerveless, who have so behaved that you feared lest you seemed unworthy of the kindness, unless you had stood out as most like him by whom you had been sent.

\ciceroSection{83} Or do you think the stains of your imperium and the slaughters of your province have been too little tracked by us? Which we have indeed pursued, not by sniffing your tracks at the entry, but by your whole wallowings of body and lairs. Marked by us were both those first crimes at your arrival when, after taking money from the Dyrrachians for the killing of your guest Plator, you stayed at the house of the very man whose blood you had assigned. And him, having taken music-slaves and other gifts, fearing and much hesitating, you confirmed and ordered to come to Thessalonica on your faith. Whom you afflicted with no punishment, not even according to the custom of our ancestors, when the wretch was eager to bend his neck under the axes of his guest, but you ordered the doctor whom you had brought out with you to cut the man's veins.

\ciceroSection{84} On which there was even an addition to the killing of Plator: Pleuratus his comrade, whom you killed by lashes, broken with the highest old age. The same man you, Rabocentus the chief of the Bessian people, when you had sold yourself to King Cotys for three hundred talents, struck with the axe — when he had come to your camp as an envoy and was promising you great support and help of foot and horse from the Bessians; nor him only, but also the rest of the envoys who had come together. Whose heads, all of them, you sold to King Cotys. Against the Denseletae — a nation always obedient to this empire, which even in that defection of all the barbarians had protected Macedonia under praetor C. Sentius — you brought a wicked and cruel war; and although you might have used these as most faithful allies, you preferred to use them as most keen enemies. So you made the perpetual defenders of Macedonia its harassers and plunderers. They disturbed our revenues, captured cities, ravaged fields, led our allies into slavery, snatched up households, drove off cattle; the Thessalonians, when they had despaired of the town, they forced to fortify the citadel.

\ciceroSection{85} By you the most ancient and most holy temple of Iupiter Urius among the barbarians was plundered. Your crimes the immortal gods expiated upon our soldiers. When they were afflicted with a new kind of disease, and no one could recover who had once fallen ill, no one doubted that violated guests, killed envoys, peaceful and allied peoples provoked by wicked war, plundered temples, brought about so great a desolation. From a small particle you recognise the universal kind of your crimes and cruelty.

\ciceroSection{86} What of avarice, which is implicated in countless charges? Should I now unfold the sum? In each kind I shall say what is most known. Did you not — the eighteen million sesterces which you had set down as something like ``equipment-money'' in the sale of my person, allotted to you out of the treasury — leave them at usury at Rome? Did you not, when the Apollonians had given you 200 talents at Rome that they should not have to pay debts incurred, of your own accord assign Fufidius, a Roman knight and most adorned man, as creditor to your debtors? Did you not, when you had handed over winter quarters to your legate and prefect, wholly overturn the wretched cities, which were drained not only of goods but underwent the wicked insults and disgraces of lust? What measure had you for assessing grain, what for ``honorary''? — if indeed what is wrenched out by force and fear can be called ``honorary''. Which all alike, but most bitterly the Bottiaeans, the Byzantines, the Chersonese, Thessalonica felt. You alone were the master, you alone the assessor, you alone the seller, in the whole province for three years, of all grain.

\ciceroSection{87} Why should I bring forth the inquiries on capital matters, the bargainings of defendants, redemptions, the most bitter condemnations, the most lustful acquittals? When you sense that any place is known to me, you may by yourself recall how many and how great are the charges in that kind. What of that arms-factory? When the cattle of the whole province had been driven up, under the name of hides, did you renew that whole household and ancestral profit-business? For you had seen, when already a big boy, in the Italian War, your house filled with profit, when your father had presided over the making of arms. What? You remember that you made a province subject to the tax of the Roman people — with a fixed harbour-tax laid on each thing whatever was sold — a slave-girl to your publicans? What?

\ciceroSection{88} Centurionships openly sold; what? — companies assigned through your little slave; what? — pay to soldiers reckoned out through all the years from the cities by tables openly set up; what? — that journey into Pontus and that attempt of yours; what? — the weakening and casting-down of your soul on the news that Macedonia was made praetorian, when you fell senseless and dead not only at being succeeded but at Gabinius's not being succeeded; what? — a quaestor put in command, the men of aedilician rank rejected; the best of your legates, each one violated by you, military tribunes not received, M. Baebius, a brave man, killed by your order?

\ciceroSection{89} What of this — that you so often, distrusting and despairing of your affairs, lay in squalor, lamentations and grief? That to that demagogue priest you sent six hundred friends and allies to the wild beasts? That, when you could scarcely bear your grief and pain at being recalled, you betook yourself first to Samothrace, then thence to Thasus with your tender dancing-boys and with Autobulus, Athamas, Timocles, the handsome brothers? That returning thence in the villa of Euchadia, who had been Execestus's wife, you lay mourning some days, and thence threadbare, with all unaware and by night, came to Thessalonica? That, when you could not bear the throng of weepers and the storm of complaints, you fled into the out-of-the-way town of Beroea? In which town when rumour had filled you with hope (because you thought Q. Ancharius would not succeed you), how, criminal, did you renew yourself to your usual intemperance!

\ciceroSection{90} I leave aside the gold for the crown which long tortured you, when you now wished it, now did not. For your son-in-law's law forbade both that it be decreed and that you receive it, save with a triumph decreed. In which you could not vomit out the money already taken and devoured (as in the hundred talents of the Achaeans); you only changed the names and kinds of moneys. I leave aside the diplomas given indiscriminately throughout the province; I leave aside the number of ships and the sum of plunder; I leave aside the account of grain demanded and ordered; I leave aside the freedom snatched from peoples and from individuals adorned with privileges by name — of which there is nothing which is not by the lex Iulia diligently sanctioned not to be allowed.

\ciceroSection{91} Aetolia — which, set apart far from barbarian peoples, lying in the bosom of peace, is contained almost in the middle lap of Greece — O Punishment and Fury of allies! — by your departure you ruined wretchedly. Arsinoe, Stratus, Naupactus, as you yourself just now stated, noble cities and full, you confess to have been taken by enemies. By what enemies? Surely by those whom you, sitting at Ambracia at your first coming, forced to migrate from the towns of the Agrians and Dolopes and to leave their altars and hearths. This you at the end, splendid imperator — when there had been added to your old defeats the sudden destruction of Aetolia — disbanded your army; nor did you prefer to undergo any punishment that was due to so great a deed than that anyone should know the number and remnants of your soldiers.

\ciceroSection{92} And that you may see the likeness of two Epicureans in military matters and command: Albucius, when he had triumphed in Sardinia, was condemned at Rome; this man, when he expected a like end, set up trophies in Macedonia. And those things which all peoples have wished to be the marks and monuments of warlike praise and victory, our preposterous imperator set up — funereal indications of lost towns, of slaughtered legions, of the province bereft of garrison and remaining soldiers — for the lasting disgrace of his race and name. The same man, that there might be something to be inscribed and cut on the base of the trophies, when he had come to Dyrrachium on his return, was besieged by those very soldiers whom a little before he answered Torquatus had been disbanded by him as a kindness. To whom, having sworn that he would pay back the next day what was owed, he hid himself at home; thence in the dead of night, in slippers, in the dress of a slave, he embarked and avoided Brundisium and made for the farthest shores of the Adriatic Sea,

\ciceroSection{93} while meanwhile at Dyrrachium the soldiers began to besiege the house in which they thought he was, and, supposing the man hidden, set fire round it. Stirred by which fear, the Dyrrachians indicated that the slipper-shod imperator had fled by night. They themselves throw down, dash, smash, scatter a most lifelike statue of his, which he had wished to stand in the most frequented place lest the memory of so charming a man should perish. So the hatred which they had brought against the man himself, that they poured out on his image and likeness.

\ciceroSection{94} Since these things are so — for I do not doubt that, when you see I know these outstanding matters, you do not suppose that middle part and throng of your disgraces is unheard by me — there is no reason for you to exhort me, none for you to invite me. To be reminded is enough for me. But none other shall remind me except the time of the commonwealth, which seems to me to be approaching more than you ever have thought. Do you see, do you feel, with the Iudiciary law passed, what kind of judges we are about to have hereafter? Neither will whoever wishes be enrolled, nor will whoever does not wish not be enrolled. None will be cast into that order, none excluded. Neither shall ambition strive after favour, nor inequity after rivalry. Those judges shall judge whom the law itself, not the lust of men, has chosen. Since this is so, believe me, you will invite no one unwilling. The matter itself and the time of the commonwealth will either invite or dissuade myself (which I would not), or some other.

\ciceroSection{95} For my part, as I said a little before, I do not think the punishments are the same in men as perhaps most do — condemnations, expulsions, deaths. Finally I think nothing has any punishment that can fall on an innocent, on a brave, on a wise, on a good man and citizen. That condemnation which is demanded against you fell to P. Rutilius — who was the model in this state of innocence. The greater punishment seemed to me of the judges and of the commonwealth than that of Rutilius. L. Opimius was cast out of his country — he who as praetor and consul had freed the commonwealth from the greatest dangers. Not on him to whom the wrong was done, but on those who did it, the punishment of crime and conscience remained. But on the contrary, twice was Catiline acquitted; even that author of your province was let go, when he had brought debauchery to the cushions of the Bona Dea. Was there anyone in so great a state who thought him freed from incest, and not those who had so judged bound by the same crime?

\ciceroSection{96} Should I wait until 75 ballots are sorted out about you — about whom long ago all mortals of every kind, age, order have judged? For who thinks you worthy of access, of any honour, of any common greeting? All curse the memory of your consulship, your deeds, morals, finally face and name from the commonwealth. The legates who were with you alienated; the military tribunes hostile, the centurions, and any soldiers from such an army that remain — not disbanded by you but scattered — hate you, wish for your destruction, curse you. Achaia drained, Thessaly harassed, Athens torn, Dyrrachium and Apollonia emptied, Ambracia plundered, the Parthini and Bullienses mocked, Epirus cut down, the Locrians, Phocians, Boeotians burnt, Acarnania, Amphilochia, the Perrhaebian and Athamanian peoples sold, Macedonia given to the barbarians, Aetolia lost, the Dolopes and the neighbouring mountaineers driven out from towns and fields. Roman citizens who do business in those places have felt that you alone have come as despoiler, harasser, pirate, enemy of themselves and of the allies.

\ciceroSection{97} To all these countless judgments, the home judgment of the verdict of your condemnation has been added — a hidden return, a furtive journey through Italy, an entry into the city deserted by friends, no letters to the senate from the province, no congratulation for your three summer campaigns, no mention of triumph. You dare to say not only what you did, but not even in what places you were. From that fount and seed-bed of triumphs, when you had brought back the dry leaves of laurel, when you had left them thrown down at the gate, then you yourself pronounced about yourself that you seemed to have so done. If you had done nothing worthy of honour, where was the army, where the expense, where the imperium, where that province most fertile in thanksgivings and triumphs? But if you could have hoped for anything, if you had thought (which the name of imperator, which the laurelled fasces, which those trophies full of disgrace and ridicule declare you to have thought) — who is more wretched than you, who more condemned, you who dared neither to write to the senate that the commonwealth had been well managed by you, nor in person to say it?

\ciceroSection{98} Or do you suppose me — whom it has always so persuaded that each man's fortune is weighed not by outcomes but by deeds, and that not on the ballots of a few judges but on the verdicts of all citizens our reputation and fortune hangs — to think you uncondemned, when allies, when those bound by treaty, when free peoples, when stipendiaries, when traders, when publicans, when the entire state, when legates, when military tribunes, when the remaining soldiers who escaped the sword, hunger, disease, judge you most worthy of every torture; to whom no pardon for the greatest crimes can be given, neither before the senate, nor before the Roman knights, nor before any order, neither in the city nor in Italy, for forgiving; who hates yourself, who fears all, who dare entrust your cause to no one, who condemns yourself?

\ciceroSection{99} I have never sought your blood; I have never sought that final punishment of the law and the courts that could be common to good men and to bad. I have wished to see you cast down, contemned, despised by the rest, despaired of and abandoned by yourself, looking round at everything, frightened at whatever sound, distrusting your own affairs, without voice, without freedom, without authority, without any consular appearance, shuddering, trembling, fawning on all — and I have seen it. Therefore, if what you fear may happen befalls you, I shall not bear it ill. But if it shall fall out a little later, I shall yet enjoy your unworthiness and timidity. Nor shall I see you, fearing to become a defendant, less gladly than as defendant; nor shall I rejoice less when I see you always squalid, than when I should for a little while see you in the mourning-garb of the accused.

