% Pro P. Sestio (pro-sestio) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} If anyone before this, members of the jury, was wondering what the cause might be that, given the great resources of the commonwealth and the great standing of our empire, by no means enough citizens of brave and great spirit could be found who would dare to put themselves and their own preservation in jeopardy for the standing of the city and for the common liberty, from this time forward let him rather wonder if he sees any good and brave citizen, than if he sees one timid or one taking thought for himself rather than for the commonwealth. For, leaving aside thinking back on the case of any single one, you can see at a single glance these men: those who, with the Senate and with all good men, raised up the afflicted commonwealth and freed it from domestic banditry --- now sad, in mourning dress, on trial for life, for reputation, for citizenship, for fortunes, for their children; while those who have violated, harassed, thrown into confusion, and overturned everything divine and human are not only flitting about in eager joy, but are even contriving danger for the bravest and best citizens, fearing nothing themselves.

\ciceroSection{2} In which, while there are many things unworthy, nothing is less to be borne than this: that they now attempt to bring danger upon us, no longer through their own bandits, no longer through men ruined by want and crime, but through you --- through the best men against the best citizens; and those whom they could not destroy with stones, with iron, with torches, with violence, with hands and forces, these they now think to crush by your authority, your scruple, your verdicts. I, members of the jury --- the voice which I had thought I should be using to give thanks and to commemorate the kindness of those who have deserved the best of me --- am now compelled to use to ward off their dangers; that this voice may serve, above all, those by whose effort it has been restored both to me and to you and to the Roman people.

\ciceroSection{3} And although the case of P. Sestius has been pleaded out by Q. Hortensius, the most distinguished man and most eloquent, and nothing has been left undone by him that needed to be lamented for the commonwealth or argued for the defendant, still I shall set about speaking, lest my championing seem to have failed, of all men, the very man through whom it has been brought to pass that I have not failed the rest of the citizens. And I am of this mind, members of the jury: that in this case, and in this final place of speaking, the part assumed by me is rather of devotion than of defence, of complaint than of eloquence, of grief than of talent.

\ciceroSection{4} And so, if I act either more sharply or more freely than those who spoke before me, I beg of you that you grant my speech as much as you think should be granted both to a devoted grief and to a just anger. For neither can grief in any be more bound to duty than this grief of mine, taken up at the danger of a man who has deserved the best of me; nor is any anger more to be praised than that which kindles me at the wickedness of those who have judged that they should wage war against all the defenders of my preservation.

\ciceroSection{5} But since the rest have answered the individual charges one by one, I shall speak about the whole standing of P. Sestius --- about the kind of his life, about his nature, his character, his incredible love for the good men, his zeal for preserving the common safety and quiet --- and I shall try, if only I can attain it, that in this confused and comprehensive defence nothing should seem to have been omitted by me which pertains either to your inquiry, or to the defendant, or to the commonwealth. And since the tribunate of P. Sestius was placed by Fortune herself in the gravest times of the city, and amid the ruins of an overturned and afflicted commonwealth, I shall not approach those greatest and amplest matters before I have shown by what beginnings and foundations these so great praises in the highest things were raised up.

\ciceroSection{6} P. Sestius was born, members of the jury, of a father whom most of you remember --- a wise, holy, and severe man; who, when he had been made tribune of the plebs, the first among the noblest men, in the best times, did not so much wish to use the rest of the offices as to seem worthy of them. By his urging Sestius married the daughter of a most honourable and most distinguished man, C. Albinus, by whom he has this boy and his daughter, already given in marriage. By two men of the deepest weight and full of antiquity he was so approved that to each of them he was both especially dear and pleasing. The death of his daughter took from Albinus the name of father-in-law, but did not take away the affection or good will of that connection: today he loves Sestius so that you can most easily judge from this man's constancy at his side, or from his anxiety and trouble.

\ciceroSection{7} While his father still lived he married the daughter of a most excellent and most stricken man, L. Scipio. The famous devotion of P. Sestius shone clearly in this and was welcome to all: he set out at once for Massilia, that he might see and console his father-in-law, driven out by the surges of the commonwealth, lying in foreign lands --- a man who ought to have stood in the footsteps of his ancestors --- and brought his daughter to him, that the man might, by the unhoped-for sight and embrace, lay aside if not all then some part of his sorrow; and with the greatest zeal and offices, in keeping with that connection, he supported both that man's hardship as long as he lived and his daughter's loneliness. I could say much about his liberality, his domestic offices, his military tribunate, his abstinence in the provincial magistracy he held; but the standing of the commonwealth flits before my eyes and snatches me to itself, and urges me to leave these lesser matters.

\ciceroSection{8} He was quaestor to my colleague C. Antonius, members of the jury, by lot, but mine by partnership of counsels. I am hindered, by what I take to be a bond of duty, from setting forth how many things P. Sestius, when he was with my colleague, perceived, reported to me, foresaw long beforehand. About Antonius I say nothing except this one thing: that, in that highest fear and danger of the city, he never wished either, by denial, to remove the common dread of all, or, by dissembling, to settle the particular suspicion of certain men against himself. In holding up and steadying that colleague, if you used truthfully to praise my forbearance towards him conjoined with the highest care for the commonwealth, near-equal praise must be P. Sestius's, who so attended on his consul that he seemed to that man a good quaestor and to all of us the best citizen.

\ciceroSection{9} The same man, when that conspiracy had burst out from its hiding-places and shadows and was flitting openly in arms, came with an army to Capua, a city which we suspected was being tried, on account of its many advantages for war, by that impious and criminal band. C. Mevulanus, military tribune of Antonius, he flung headlong out of Capua --- a ruined man, and not obscurely engaged at Pisaurum and in other parts of Gallic territory in that conspiracy; and the same Sestius saw to it that C. Marcellus, when he had come not only to Capua but had even thrust himself, as if from zeal for arms, into a most numerous household, was driven out from that city. For this reason both that gathering at Capua --- which on account of the safety of that city preserved by my consulship adopted me alone as patron --- gave at my hands the greatest thanks to this P. Sestius, and at this present time the same men, the name changed, the colonists and the decurions, bravest and best men, by their testimony declare the kindness of P. Sestius, by their decree avert his danger.

\ciceroSection{10} Recite, please, L. Sestius, what the decurions of Capua have decreed, that now your boy's voice may signify something to your enemies --- what, when it has gathered strength, it may seem about to bring about. \textit{Decree of the Decurions.} I do not recite a decree wrung out by some duty of neighbourhood, or clientage, or public hospitality, or canvassing, or recommendation: I recite the memory of a danger lived through, the heralding of the fullest kindness, the voice of a duty present, the testimony of past time.

\ciceroSection{11} And in those same times, when Sestius had already freed Capua from fear, and when the Senate and all the good men, the domestic enemies seized and put down with me as leader, had drawn the city out from the greatest dangers, I summoned P. Sestius from Capua by letter together with that army he had with him at the time. Reading these letters, he flew at once to the city with incredible swiftness. And that you may recall the dreadfulness of that time, hear the letter, and stir up your memory to the thought of the past fear. \textit{Letter of Cicero, consul.} By this arrival of P. Sestius the assaults and attempts of the new tribunes of the plebs, who in the last days of my consulship were eager to harass the things I had done, and the rest of the conspiracy, were retarded.

\ciceroSection{12} And after it was understood that, with M. Cato as tribune of the plebs --- bravest and best citizen --- defending the commonwealth, the Senate and the Roman people could easily, by their own majesty, without an armed garrison, protect the standing of those who had defended the common safety at their own peril, Sestius with that army, with the highest swiftness, overtook Antonius. Here what should I parade --- by what means he urged the consul to action, what spurs he applied to a man perhaps eager for victory but yet too much in dread of the common Mars and chance of war? It is long to say those things, but this short word I shall say: if M. Petreius's outstanding spirit and love for the commonwealth, his standing virtue in the commonwealth, his highest authority among the soldiers, his wonderful experience in military matters had not been there, and if P. Sestius had not been a helper to him in stirring up Antonius, urging him on, accusing him, driving him --- a place would have been given in that war to winter; and never would Catiline, when he had emerged from the Apennine frost and those snows and, finding a whole summer free, had begun to seize beforehand the cattle-paths of Italy and the stables of the shepherds, have fallen in his most miserable doom without much blood and without the laying waste of all Italy.

\ciceroSection{13} This was the spirit, then, that P. Sestius brought to his tribunate --- to leave aside his quaestorship in Macedonia and come at last to these closer matters; although that singular provincial integrity is not to be passed over --- whose traces I lately saw in Macedonia not pressed lightly to the heralding of a brief time, but fixed for the everlasting memory of that province. But let us so pass these things by that, looking on them and looking back, we leave them: let us come to the tribunate, which by now itself calls to itself and in a way swallows up my speech, with strained zeal and at speed.

\ciceroSection{14} About which tribunate, indeed, it was so spoken of by Q. Hortensius that his speech seemed to contain not only the defence of the charges, but also to set forth, worthy of memory, an authority and discipline for the youth in the taking up of public affairs. But still, since the whole tribunate of P. Sestius sustained nothing other than my name and my cause, I think I must necessarily say of the same matters, if not more subtly, then certainly more grievously, lamented. In which speech if I wished to inveigh more sharply against certain men, who would not allow that I might brand with the freedom of my voice those by the rage of whose wickedness I have been violated? But I shall act with measure, and I shall serve this man's circumstances rather than my own grief: if any privately disagree with our preservation, let them lie hidden; if any did something at some time and the same now keep silent and quiet, let us also have forgotten; if any thrust themselves forward, harass us, we shall bear it as long as it can be borne; nor will my speech offend anyone except him who has so thrust himself forward that we may seem not to have charged at him but to have run into him.

\ciceroSection{15} But it is necessary, before I begin to speak about the tribunate of P. Sestius, that I set forth the whole shipwreck of last year's commonwealth, in which gathering of wreckage and refitting of the common safety all the deeds, words, and counsels of P. Sestius were turned. That had been a year, members of the jury, when, amid great commotion and the fear of many, the bow was bent against me alone --- as the ignorant of affairs commonly used to say --- but in real fact against the whole commonwealth, by the transfer to the plebs of a furious and ruined man, angry at me but far more sharply hostile to peace and the common safety. This man --- a most distinguished man and most friendly to me, with many opposing me, Cn. Pompeius --- had bound by every precaution, pact, and oath to do nothing in his tribunate against me; that wicked man, born from the cesspool of every crime, judged that he would too little break his pact, unless he had terrified, by his own peculiar dangers, that very person who had stood surety for another's danger.

\ciceroSection{16} This loathsome and monstrous beast, bound by the auspices, fettered by ancestral custom, constrained by the chains of the sacred laws, the consul suddenly let loose by a curiate law --- whether, as I judge, prevailed upon, or, as some thought, angry with me, but certainly ignorant and unforeseeing of the so great wickednesses and evils impending. He as tribune of the plebs was lucky in overturning the commonwealth by no sinews of his own --- for what sinews could there be in a life of that kind, the man drained by his brother's outrages, his sisters' debaucheries, every unheard-of lust?

\ciceroSection{17} But there was indeed surely a kind of fated fortune of the commonwealth, that that blind and witless tribune of the plebs got --- what shall I say? --- consuls? That I should call by this name the overturners of this empire, the betrayers of your standing, the enemies of all good men, who thought themselves adorned by those fasces and other insignia of the highest office and command for the destroying of the Senate, the afflicting of the equestrian order, the extinguishing of all rights and institutions of our ancestors? Of which, by the immortal gods, if you do not yet wish to recall the wickednesses and brands burnt into the commonwealth, look at their faces and their gait in your minds: their deeds will more easily occur to your minds, if you set their very faces before your eyes.

\ciceroSection{18} The one, drenched in unguents, with curled hair, looking down on the witnesses of his debaucheries and the old harassers of his early youth, puffed up by the gangs of the Puteal and the moneylenders --- by whom long ago, driven, lest in that Scyllaean strait of his debt as in a current he should cling to a column, he had taken refuge in the harbour of the tribunate --- despised the Roman knights, threatened the Senate, sold himself to the gangs and used to declare himself snatched away by them so that he should not have to plead in court on the charge of canvassing, and from the same men also said that, even with the Senate against him, he hoped for a province; and unless he obtained it, he judged that he would in no way be safe.

\ciceroSection{19} The other, good gods, how grimly he stalked, how truculent, how terrible to look upon! You would say you were looking upon some one of those bearded men, an example of ancient command, an image of antiquity, a column of the commonwealth. Dressed roughly in this plebeian and almost dingy purple of ours, his hair so unkempt that he seemed about to lift up the Seplasia --- that perfumers' street of the very Capua in which, for the sake of decking out his image, he was then holding the duumvirate. For what shall I say of his eyebrow? --- which then to men seemed not an eyebrow but a pledge for the commonwealth. So great was the gravity in his eye, so great the contraction of his brow, that the year seemed to lean on that eyebrow as on a surety.

\ciceroSection{20} This was the talk of all: ``Yet there is a great and firm support of the commonwealth; I have one whom I may set against that filth and mire; by his face, by Hercules, he will break his colleague's lust and lightness; the Senate will have for this year someone whom it can follow; an authority and leader for the good will not be wanting.'' To me, especially, men used to give thanks, that I was going to have, against a furious and reckless tribune, a consul both friend and connection, and at the same time brave and weighty. And of these two the one deceived no one. For who could have judged that a man could hold the helm of so great an empire and handle the rudders of the commonwealth in its greatest course and waves --- a man emerged suddenly out of the long shadows of brothels and debaucheries, finished off by wine, taverns, brothel-keepers, and adulteries? When this man, beyond hope, was placed in the highest position by another's resources --- one who, drunk, could not look on the impending storm, nor even bear to look at the unfamiliar daylight.

\ciceroSection{21} The other plainly deceived many on every side; for he was, by men's opinion, recommended by his very nobility, that flattering little go-between. We good men always favour nobility, both because it is useful to the commonwealth that nobles should be worthy of their ancestors, and because the memory of distinguished men, even of those who have deserved well of the commonwealth, has weight with us, even when they are dead. Because they saw him always sad, always silent, somewhat grim and uncultivated, and because he was of that name, so that frugality seemed engrafted in the family, they favoured him, rejoiced, and called the man, in their hope, to the integrity of his ancestors --- forgetful of his mother's family.

\ciceroSection{22} I, however --- to speak truly, members of the jury --- never thought that there was as much wickedness, recklessness, cruelty in the man as I myself, with the commonwealth, have felt. I knew the man was worthless and light, and from his youth, by men's mistake of false opinion, recommended; for his mind was hidden by his face, his outrages by his walls. But this barrier is neither lasting, nor drawn over so that it cannot be seen through by curious eyes. We saw the kind of life, the laziness, the inertia; those who had come a little closer looked into his shut-in lusts; finally, even speech gave us handles by which we could grasp his hidden feelings.

\ciceroSection{23} The learned man would praise certain philosophers, but he could not give their names; yet he praised most those who, beyond the rest, are said to be authors and praisers of pleasure --- of what kind, and at what time, and in what way, he did not ask: he had swallowed the very word with all the powers of mind and body. And he used to say that the same men taught splendidly that the wise do everything for their own sake, that to take up public affairs is not for the well-balanced, that nothing is more excellent than a leisured life full and packed with pleasures; while those who said that one must serve standing, take thought for the commonwealth, hold the account of duty rather than of advantage in all life, undergo dangers for the fatherland, take wounds, meet death --- those, he said, were prophesying and going mad.

\ciceroSection{24} From these constant and daily talks of his, and from the men with whom I saw him living in the inner part of his house, and from the very fact that the house gave off such a smoke that many tokens of his fellowship reeked, I judged thus: that nothing good was to be expected from those trifles, but, certainly, nothing bad to be feared. But it is so, members of the jury, that, if you should give a sword to a small boy or to a feeble old man or to a weak person, by his own assault he will hurt no one; but if it is brought near the unprotected body even of the bravest man, by its very edge and the strength of the iron it can wound: when to nerveless and bloodless men a consulship was given as a sword, men who could of themselves never strike anyone, they, armed in the name of the highest command, slaughtered the commonwealth. They made a pact openly with a tribune of the plebs, that they should receive from him the provinces they themselves wanted, an army and money to whatever extent they wanted, on this condition --- if they themselves should first hand over to the tribune of the plebs the commonwealth afflicted and bound. And that pact, they said, could be sealed and sanctified by my blood.

\ciceroSection{25} The thing being made known --- for so great a wickedness could neither be dissembled nor lie hidden --- bills are promulgated at one and the same time by the same tribune: about my destruction and about the consuls' provinces by name. Here at that point the Senate anxious, you, Roman knights, stirred up, all Italy moved, all citizens, finally, of every class and order judged that help for the supreme commonwealth was to be sought from the consuls and from the supreme command. But those alone, beside that furious tribune, were two whirlwinds of the commonwealth, who not only did not aid their headlong fatherland, but grieved that it was falling too slowly. They were called on by these men daily with the complaints of all good men and even with the prayers of the Senate to take up my cause, to do something, finally, to refer to the Senate. They harassed the most eminent of that order not only by refusing, but even by mocking.

\ciceroSection{26} Here suddenly, when an incredible multitude had gathered onto the Capitol from the whole city and from all Italy, all thought that they should change their dress, and that I should now be defended by every kind of private counsel --- since the commonwealth was bereft of public leaders. At the same time the Senate was in the temple of Concord, the very temple that brought back the memory of my consulship, when the whole order, weeping, begged the curled consul; for that other rough and severe consul shut himself up at home on purpose. With what arrogance then did that filth and stain reject the prayers of the most eminent order and the tears of the most distinguished citizens! How that glutton of his fatherland despised me myself! For why should I speak of his patrimony, which he lost at the time when he was making his profit? You came to the Senate, you, I say, Roman knights and all good men, in changed dress, and threw yourselves for my head at the feet of the most foul brothel-keeper; and when your prayers had been rejected by that bandit, a man of incredible loyalty, greatness of soul, and steadiness, L. Ninnius, referred to the Senate about the commonwealth, and the Senate in full session voted that they should change their dress for my preservation.

\ciceroSection{27} O that day, members of the jury --- mournful for the Senate and all good men, sorrowful for the commonwealth, for me grievous as private grief, glorious for the memory of posterity! For what could anyone take from all memory more illustrious than that, for one citizen, all the good privately of one mind and the whole Senate by public counsel had changed their dress? Which change indeed was not made then for the sake of pleading, but of mourning. For whom would you have entreated, when all were in mourning dress, and when this was sign enough that he was a wicked man who had not changed his dress? When this change of dress had been made amid such great grief in the city, I leave aside what that tribune --- pirate of all things divine and human --- did, who, when the noblest young men, the most honourable Roman knights, had attended as suppliants for my preservation, threw them before the swords and stones of his gangs. Of the consuls I speak, on whose loyalty the commonwealth ought to have leaned.

\ciceroSection{28} Half-dead he flies out from the Senate, with mind and face no less troubled than if a few years before he had fallen into a meeting of his creditors. He summons a contio; the consul makes such a speech as Catiline, victorious, would never have made: that men were mistaken if they still thought that the Senate could do anything in the commonwealth; that the Roman knights would pay the penalty for that day on which, when I was consul, they had been with swords on the slope of the Capitol; that the time had come for those who had been in fear --- he meant the conspirators, of course --- to take vengeance. If he had said only this, he would have been worthy of every punishment; for the speech of a consul, by itself, can shake a commonwealth. Now see what he did.

\ciceroSection{29} L. Lamia, who, both because of the great intimacy I had with his father, was uniquely fond of me, and on behalf of the commonwealth was even eager to meet death, he banished in a contio, and proclaimed by edict that he should be two hundred miles from the city, because he had dared to plead for a citizen, for a citizen of good service, for a friend, for the commonwealth. What is one to do with this man, or for what is one rather to keep so criminal an enemy? Who --- to leave aside the rest of the things he holds joined and shared with his monstrous and impure colleague --- has this one thing peculiarly his: that he drove out from the city, banished --- I do not say a Roman knight, not a most adorned and excellent man, not a most loyal citizen of the commonwealth, not a man at that very time mourning, with the Senate and all good men, the misfortune of his friend and of the commonwealth, but cast a Roman citizen out of his fatherland, by edict, without any trial.

\ciceroSection{30} Nothing have allies and Latins been used to bear more bitterly than this --- which very rarely happens --- to be ordered out of the city by the consuls; and yet for those there was a return to their own communities, to their own household Lares; and in that common discomfort no peculiar disgrace fell on anyone by name. But what is this? Will the consul, by edict, drive Roman citizens from their household gods? Will he expel them from the fatherland? Will he choose whom he wills, and condemn and cast out by name? If he had ever thought that you who now are would be in the commonwealth, if finally he had believed that any image or shadow of trials would be left in the city, would he ever have dared to remove the Senate from the commonwealth, to scorn the prayers of the Roman knights, to overturn, finally, the right and liberty of all citizens by new and unheard-of edicts?

\ciceroSection{31} Although you listen to me, members of the jury, with the most attentive minds and with the highest kindness, I yet fear that some of you may perhaps wonder why this speech of mine wishes to be so long or so far back-reaching, or what the offences of those who harassed the commonwealth before this man's tribunate have to do with the case of P. Sestius. But I have set myself this task: to show that all the counsels of P. Sestius and the mind of his whole tribunate were this --- that he should so far as possible heal the afflicted and ruined commonwealth. And if, in setting forth those wounds, I shall seem to say more about myself, forgive me; for both that ruin of mine you and all good men have judged to be the greatest wound of the commonwealth, and P. Sestius is on trial not in his own name but in mine: who, since he spent the whole power of his tribunate on my preservation, must necessarily have my cause of past time joined with the present defence of his.

\ciceroSection{32} The Senate, then, was in mourning; the city, by public counsel, was in dingy garb; there was no Italian municipium, no colony, no prefecture, no company at Rome of the tax-farmers, no college or council or any common assembly that had not then most honourably decreed concerning my preservation. When suddenly the two consuls publish an edict that the senators should return to their own dress. What consul ever forbade the Senate to obey its own decrees? What tyrant forbade the wretched to mourn? Is it too little, Piso --- to leave aside Gabinius --- that you so far deceived men that you neglected the authority of the Senate, despised the counsels of every best citizen, betrayed the commonwealth, brought down the consular name? Did you even dare to publish an edict that men should not mourn my, their own, the commonwealth's calamity, that they should not signify their grief by their dress? Whether that change of dress was for their own mourning or for the sake of pleading, who was ever so cruel as to forbid anyone to grieve for himself or to entreat for others?

\ciceroSection{33} What? Are men not used, of their own accord, to change their dress in the dangers of friends? On your own behalf, Piso, will no one change? Not even those whom, with no senatus consultum, indeed against the resisting Senate, you yourself appointed legates? Therefore the misfortune of a desperate man and a betrayer of the commonwealth those will perhaps mourn who wish: but the danger of a most flourishing citizen, joined to the danger of the city, by the good will of all the good and a man who has deserved best of the safety of the fatherland --- this the Senate may not mourn? And those same consuls, if they are to be called consuls, whom there is no one who does not think should be torn out not only from memory but from the calendar --- the pact of provinces already struck, brought forward into a contio in the Circus Flaminius by that fury and pestilence of the fatherland --- to your great groan --- with their voice and opinion approved everything that was then being done against me and against the commonwealth. With those same consuls sitting and looking on, a law was passed that the auspices should not be valid, that no one should announce against, that no one should interpose against the law, that on every comitial day a law should be allowed to be carried, that the Aelian and Fufian laws should not be valid: by which one bill alone who is there who does not understand that the whole commonwealth has been wiped out?

\ciceroSection{34} And with those same consuls looking on, an enrolment of slaves was being held before the Aurelian tribunal in the name of the political clubs (\textit{collegia}); when men were being recruited street by street, formed into ten-man squads, urged on to violence, to fighting, to slaughter, to plunder. With those same consuls arms were being openly carried into the temple of Castor; the steps of that same temple were being torn away; armed men held the Forum and the contiones; killings and stonings were taking place; there was no Senate, nothing left of magistracy: one man held all the power of all by arms and banditry --- not by any strength of his own, but, when he had drawn off the two consuls from the commonwealth by the pact of provinces, he ran wild, he lorded it, he made promises to others, by terror and fear he held many, by hope and promises even more.

\ciceroSection{35} When these things were of this kind, members of the jury --- when the Senate had no leaders, and instead of leaders had betrayers or rather open enemies; when the equestrian order was being summoned by the consuls as defendant; when the authority of all Italy was being repudiated; when some were being banished by name, others terrified by fear and danger; when there were arms in the temples, armed men in the Forum; when these things were not dissembled by the silence of the consuls but were approved by their voice and opinion; when we all saw the city not yet razed and overturned but already taken and overpowered --- still, members of the jury, by the so great zeal of the good, we should have resisted these so great evils, but other fears, other concerns and suspicions moved me.

\ciceroSection{36} For I shall set forth on this very day, members of the jury, the whole reason of my act and counsel, and I shall fail neither this so great zeal of yours in listening, nor indeed this so great multitude --- the like of which, in my memory, has never been at any trial. For if I, in so good a cause, with so great zeal of the Senate, with so incredible a consensus of all the good, with the people so prepared, all Italy, finally, equipped for every contention, gave way before the rage of a tribune of the plebs, that most despicable of men, and was alarmed at the lightness and recklessness of the most contemptible consuls, I confess that I was too timid, of no spirit, of no counsel.

\ciceroSection{37} For what was there like in the case of Q. Metellus? Whose cause, although all good men approved, yet neither the Senate publicly nor any single order distinctly nor by their decrees all Italy had taken up. He had acted with regard to a kind of his own glory rather than the manifest preservation of the commonwealth, when he alone had refused to swear to a law passed by violence: in the end he seemed to have been so brave on this condition, that he was exchanging the love of his fatherland for the glory of constancy. He had to do with the unconquered army of C. Marius; he had as enemy C. Marius, the preserver of his fatherland, then in his sixth consulship; he had to do with L. Saturninus, twice tribune of the plebs, a wakeful man, and one who, in his popular cause, had behaved if not with measure then certainly in a popular and abstinent way. He yielded, lest, defeated by brave men, he should fall with disgrace, or, victorious, deprive the commonwealth of many brave citizens.

\ciceroSection{38} My cause the Senate openly, the equestrian order most sharply, all Italy publicly, all the good men distinctly and zealously had taken up. I had done deeds of which I had been not the sole author but the leader of the wishes of all, and which pertained not only to my singular glory but to the common safety of all citizens and almost of nations; I had done them on this condition, that everyone should always stand for and protect my act. The contention I had was not with a victorious army, but with hired gangs stirred up to plunder the city; I had as enemy not C. Marius, the terror of enemies, the hope and bulwark of the fatherland, but two unwelcome prodigies, whom poverty, whom the magnitude of debt, whom lightness, whom wickedness had bound and made over to a tribune of the plebs.

\ciceroSection{39} Nor had I to do with Saturninus, who, because he knew that the corn-supply administration had been transferred from himself, the quaestor at Ostia, by way of disgrace, to the chief man both of the Senate and of the city, M. Scaurus, was pursuing his grief with great straining of mind --- but with the kept boy of wealthy buffoons, with the adulterer of his sister, with the priest of debaucheries, with a poisoner, with a forger of wills, with an assassin, with a brigand. Whom, if --- a thing easy to do and that ought to have been done, and which the best and bravest citizens demanded of me --- I had overpowered with violence and arms, I should not have feared lest anyone reprove violence repelled by violence, or mourn the death of ruined citizens, or rather of domestic enemies. But these things moved me: in every contio that fury was crying that the things he was doing against my preservation he was doing on the authority of Cn. Pompeius, a most distinguished man and most friendly to me both now and as long as it was permitted; M. Crassus, with whom I had every tie of friendship, a most brave man, was being declared by the same pestilence to be most hostile to my fortunes; C. Caesar, who, by no merit of mine, ought not to have been alienated from me, was being said by the same man, in daily contiones, to be most hostile to my preservation.

\ciceroSection{40} He was saying that, in the taking of his counsels, he would use these three as authors, as helpers in the doing of the thing; of whom, one had the largest army in Italy, two who were then private citizens could both lead the Roman people and prepare an army, if they wished, and that they would do so. And he was announcing to me not the people's verdict, nor any lawful contention, nor a hearing or pleading of the case, but violence, arms, armies, generals, camps. What then? Did the speech of an enemy --- empty especially --- so wickedly thrown against most distinguished men, move me? Yes, indeed; not his speech, but the silence of those against whom that so wicked a speech was being thrown. Who at the time, although for other reasons they kept silent, yet to men who feared everything, by keeping silent seemed to speak, by not denying seemed to confess. They, then, terrified at that time by some fear --- because they thought that those acts and the whole business of the previous year were being shaken by the praetors, weakened by the Senate and by the chief men of the city --- did not wish to alienate from themselves a popular tribune, and were saying that their own dangers were closer to them than mine.

\ciceroSection{41} But still, both Crassus said that my cause must be taken up by the consuls, and Pompey kept entreating their good faith and said that he, a private citizen, would not fail a publicly undertaken cause. That man, zealous for me, most eager to preserve the commonwealth --- certain men placed for that very business at my house warned him to be more cautious, and said that ambushes against his life had been laid by me at my own house; and they stirred up his suspicion, some by sending letters, others by messengers, others in person, so that the man --- though he certainly feared nothing from me --- thought that he must take care, lest those certain men should contrive anything in my name. Caesar himself, whom men ignorant of the truth especially thought angry with me, was at the gates, was holding command; his army was in Italy, and in that army the very tribune of the plebs, my enemy, had set his brother as commander.

\ciceroSection{42} When I saw these things, then --- for they were not hidden --- that the Senate, without which the city cannot stand, had been wholly removed from the city; that the consuls, who ought to have been the leaders of public counsel, had brought it about that, through them, public counsel itself was wholly being destroyed; that those who had most power were being set up in all contiones --- falsely, but still alarmingly --- as authors of my destruction; that contiones were being held against me daily; that no one was uttering a voice for me or for the commonwealth; that the standards of the legions were thought to be aimed at your throats and goods --- falsely, but yet so thought; that the old forces of the conspirators and that scattered, overpowered, unwelcome band of Catiline had been renewed by a new leader and an unhoped-for change of things: when I saw these things, what was I to do, members of the jury?

\ciceroSection{43} I know that then your zeal was not failing me, but my zeal was nearly failing yours. Should I, a private citizen, contend in arms against a tribune of the plebs? The good would have conquered the wicked, the brave the lazy; he would have been killed, the man who could only by this one medicine be driven off from the pestilence of the commonwealth. What then? Who would have stood for the rest? To whom, finally, was it doubtful that that tribunician blood, especially poured out by no public counsel, would have had the consuls as avengers and defenders? When a certain man had said in a contio that I must either perish once or conquer twice. What was it ``to conquer twice''? This indeed: that, if I had fought it out with a most witless tribune of the plebs, I should have to fight with the consuls and the rest of his avengers.

\ciceroSection{44} I, indeed --- even if I had to perish, and not to receive a stroke for me curable but for him who had inflicted it deadly --- still, members of the jury, I would have preferred to perish once than to conquer twice; for that other contention was of such a kind that neither defeated nor victorious could we have held the commonwealth. What if, in the first contention, defeated by tribunician violence, I had fallen in the Forum with many good men? The consuls, I suppose, would have called the Senate, which they had wholly destroyed from the city; would have called to arms men who had not allowed the commonwealth to be defended even by dress. After his death they would have parted from the tribune of the plebs --- they who had wished one and the same hour to be of my destruction and of their own rewards.

\ciceroSection{45} But one course remained to me, which perhaps some brave and sharp and great-souled man would say: ``You should have resisted, fought back, met death fighting.'' On which point I call you, you, fatherland, to witness, and you, household and ancestral gods --- that for the sake of your seats and temples, for the sake of the safety of my fellow-citizens (which has always been dearer to me than my own life) I have fled from a fight and slaughter. For if it had befallen me, sailing in some ship with my friends, members of the jury, that many pirates from many places threatened with their fleets to overpower that ship unless they had handed me over alone to them, and the passengers refused, and chose to perish with me rather than hand me over to enemies, I should rather have flung myself into the deep, that I might preserve the rest, than have led those so eager for me not only to certain death, but into the great risk of life.

\ciceroSection{46} But when, against this ship of state --- the helm torn from the Senate, drifting in the deep among the storms of seditions and discords --- so many armed fleets seemed about to charge unless I alone were given up; when proscription, slaughter, plunder were being announced; when some did not defend me from suspicion of their own danger, others were stirred up by old hatred of the good, others were envious, others judged me to stand in their way, others wished to avenge some grief of theirs, others hated the very commonwealth and this state of the good and our peace, and on all these so various counts demanded me alone --- should I rather have fought it out, with the highest --- I will not say destruction, but certainly peril of you and of your children, than that, what was hanging over all, I, one for all, should take upon myself and undergo?

\ciceroSection{47} The wicked would have been beaten. ``But they were citizens; and beaten by him, a private citizen, who without arms even, as consul, had preserved the commonwealth.'' But if the good men had been beaten, who would have been left? Do you not see that the matter would have come down to slaves? Or was death, as some think, to be met by me with even mind? What? Was I then fleeing death? Was there any thing I could think more to be desired? When I was carrying through those so great matters in so great a multitude of the wicked, did not death, did not exile flit before my eyes? Were not these things, in the very doing of the deed, sung by me as fates? Was it for me, in so great a grief of my dear ones, in so great a separation, in so great a bitterness, in so great a stripping of all the things which either nature or fortune had given me, to keep my life? Was I so untaught, so ignorant of things, so devoid of counsel or talent? Had I heard nothing, seen nothing, learned nothing in my own reading and inquiry? Did I not know that the course of life is short, of glory eternal? That, since death has been set as the limit for all, it is to be wished that the life owed to necessity should rather seem to be given to the fatherland than reserved to nature? Did I not know that among the wisest men this had been a contention --- some saying that the souls and feelings of men were extinguished at death; others, that the minds of wise and brave men were most awake and active when they had departed from the body; that of the two, the one was not to be fled (to lack feeling); the other even to be wished for (to feel a better feeling)?

\ciceroSection{48} Finally, since I had always referred everything to standing, and judged that without standing nothing in life was to be sought by a man, should I, a man of consular rank, after such great deeds done, fear death --- which even maidens at Athens, the daughters, I take it, of king Erechtheus, are said to have despised for the fatherland? Especially since I was of that city from which C. Mucius alone came into the camp of Porsenna and tried to kill him with death set before him; from which P. Decius, first the father, after a few years the son endowed with his father's virtue, devoted himself and his life, in the line of battle drawn up, for the safety and victory of the Roman people; from which countless others, partly for the gaining of praise, partly for the avoiding of disgrace, met death in various wars with the most even minds; in which city I myself remembered that the father of this M. Crassus, a most brave man, lest he should see his enemy victorious while he himself yet lived, drew off his own life with the same hand by which he had often offered death to enemies.

\ciceroSection{49} I, thinking these and many other things, saw this: that if my death had taken away the public cause, there would never be anyone who would dare to take up against ruined citizens the safety of the commonwealth; and so, not only if I had perished by violence, but even if I had been put out by sickness, I judged that the example of preserving the commonwealth would perish with me. For who would ever, if I had not been recalled by the Senate and the Roman people with the so great zeal of all the good --- which certainly, if I had been killed, could not have happened --- dare to lay hand on any part of the commonwealth, with the least odium to himself? I have therefore preserved the commonwealth by my withdrawal, members of the jury: I have driven slaughter from you and your children, devastation, fires, plunderings, by my own grief and mourning; and I alone have preserved the commonwealth twice --- once by glory, again by my hardship. And in this I shall never deny that I am a man, so that I might glory that I, without grief, have done without my best brother, my dearest children, my most loyal wife, your sight, my fatherland, this rank of office. Had I done this, what kindness would you have from me, when I had left for you those things which to me were of no value? This, in my own mind at least, ought to be the surest sign of my highest love for the fatherland --- that, although I could not be away from it without the highest grief, I preferred to bear this myself than that it should be shaken by the wicked.

\ciceroSection{50} I remembered, members of the jury, that divine man, born from the same roots as we for the safety of this empire, C. Marius, who, in his highest age, when he had fled from the violence of nearly lawful arms, first hid his old man's body in the marshes by sinking it down, then took refuge in the pity of the lowest and slenderest men of Minturnae, then in a very small boat, fleeing all ports and lands, came to the most deserted shores of Africa. And he kept his life, that he might not be unavenged, for the most uncertain hope and for the destruction of the commonwealth. I, who --- as many in the Senate said in my absence --- was living at the peril of the commonwealth, who for that reason was being commended to foreign nations by consular letters from the opinion of the Senate --- if I had abandoned my life, should I not have betrayed the commonwealth? In which now, when I am restored, the example of public good faith lives with me. And if it is kept immortal, who does not understand that this city will be immortal?

\ciceroSection{51} For the foreign wars of kings, of peoples, of nations have long since been so extinguished that we deal splendidly with those whom we suffer to be at peace; finally, from a war's victory hardly anyone has fallen into the envy of his fellow-citizens. Against domestic evils and the counsels of reckless citizens we must often resist; and the medicine of these dangers is in the keeping of the commonwealth --- which medicine, members of the jury, you would all have lost, if by my destruction the power had been wrenched from the Senate and the Roman people of declaring their grief about me. Wherefore I warn you, young men, and I lay it down as a precept by my own right --- you who look to standing, to the commonwealth, to glory: that, if some necessity at any time call you to defend the commonwealth against wicked citizens, you not be slacker, and shrink from brave counsels at the recollection of my misfortune.

\ciceroSection{52} First, there is no danger that anyone shall ever fall in with such consuls, especially if to those that has been paid in full what is owed them. Then, never now, as I hope, will any wicked man say that he is being attacked, with the counsel and aid of the good, in the commonwealth, while they are silent; nor will he set the terror of an armed army against the togaed; nor will there be just cause for a general sitting at the gates to allow his terror to be falsely paraded and set against. Nor, again, will the Senate ever be so oppressed that it has not even the power of supplication and mourning, nor the equestrian order so captive that Roman knights are banished by a consul. Although these things, and even many far greater that I deliberately pass by, befell, you see me, nevertheless, recalled to my pristine standing in a short time, with a moment of grief interposed, by the voice of the commonwealth.

\ciceroSection{53} But to return to that which throughout this whole speech I have set out for myself --- that the commonwealth was that year, by the wickedness of the consuls, brought low by every evil --- on that very day, mournful for me, sorrowful for all good men, when I had torn myself from the embrace of the fatherland and from your sight, and had yielded, from fear of your danger, not my own, to the rage, the wickedness, the perfidy, the weapons and threats of a man, and had left the fatherland --- which to me was dearest --- for the love of that very fatherland; when not only men, but the roofs of the city and the temples mourned my so horrible, so grave, so sudden misfortune; when none of you wished to look on the Forum, none on the Curia, none on the daylight: on that very day, I say --- ``day,'' did I say? rather, in the very same hour and at the same point of time as my and the commonwealth's destruction --- the province was put up to vote for Gabinius and Piso. By the immortal gods, guardians and preservers of this city and empire, what monstrosities, what wickednesses you saw in the commonwealth! A citizen had been driven out who had defended the commonwealth from the authority of the Senate with all the good men, and driven out for no other charge than that very one; he had been driven out without trial, by violence, with stones, with iron, finally with slaves stirred up; the law had been carried with the Forum laid waste and abandoned and given over to assassins and slaves --- and that the law that, that it should not be carried, the Senate had been in changed dress.

\ciceroSection{54} In so great a confusion of the city, the consuls did not allow even night to come between my destruction and their own booty: at once, with me struck down, to drink my blood and, with the commonwealth still breathing, to strip its spoils, they flew. I leave aside the congratulations, the feasts, the dividing-up of the treasury, the kindnesses, the hope, the promises, the booty, the joy of the few in the mourning of all. My wife was harassed; my children were sought for slaughter; my son-in-law --- and Piso my son-in-law --- was thrust away as suppliant from the feet of the consul Piso; my goods were plundered and were brought to the consuls; my house was burning on the Palatine: the consuls were feasting. And if they rejoiced at my discomfort, still they should have been moved by the danger of the city.

\ciceroSection{55} But, that I may now withdraw from my own case, recall the rest of the pestilences of that year --- so will you most easily perceive what great force of all remedies the commonwealth missed from its closest magistrates --- the multitude of laws, both of those carried and indeed of those that were promulgated. For they were carried by those consuls --- shall I say, while they were silent? rather, even with them approving --- so that the censorial power of marking and the gravest verdict of that most sacred magistracy on the commonwealth was abolished; so that the political clubs (\textit{collegia}) not only were those old ones restored against the senatus consultum, but innumerable others, new ones, were enrolled by one gladiator; so that, by the remission of the five-twelfths, almost a fifth part of the revenues was abolished; so that to Gabinius, in place of his Cilicia (which he had bargained for himself, if he should betray the commonwealth) Syria was given; and to that one glutton, by a new law, the power of deliberating twice on the same matter and of changing his province after the law was carried was given.

\ciceroSection{56} I leave aside that law which by one bill wiped out all the rights of religion, of the auspices, of the magistracies, all the laws which are about the right and the time of carrying laws; I leave aside every domestic stain. We were seeing even foreign nations shaken by the fury of that year. By a tribunician law that famous Pessinuntine priest of the Mother Goddess was driven out and stripped of his priesthood, and a shrine of the most sacred and most ancient religion was sold for great money to Brogitarus, an impure man and unworthy of that religion, especially since he had sought it not for the sake of worshipping it but of violating it. Men were named kings by the people who had never sought that even from the Senate. Exiles were brought back to Byzantium who had been condemned, at the very time when uncondemned citizens were being thrust out from their city.

\ciceroSection{57} King Ptolemy, who, although he had not yet himself been called an ally by the Senate, was yet the brother of that other king who, though in the same case, had already obtained that honour from the Senate; he was of the same line and same ancestors, of the same antiquity of alliance; finally, he was a king, if not yet an ally, at least not an enemy: at peace, quiet, relying on the empire of the Roman people, was enjoying his ancestral kingdom and royal leisure: about him, who was thinking nothing, suspecting nothing, with the same gangs casting the votes, a bill was passed that, sitting with his purple and sceptre and those royal insignia, he should be exposed to the public crier; and, with the Roman people commanding (which has been used to give back kingdoms even to kings conquered in war), a friendly king, with no injury recalled, with no things demanded back, was made public property along with all his goods.

\ciceroSection{58} Many bitter, many shameful, many turbulent things that year had; yet to that wickedness which their monstrous nature put forth against me, I do not know whether we may rightly say this is next. Antiochus the Great --- our ancestors, when he had been overcome by a great straining of war by land and sea, ordered to reign within Mount Taurus; Asia, with which they had punished him, they made over to Attalus, that he should reign in it. With Tigranes, king of the Armenians, we ourselves lately waged a serious and long war, when he, by inflicting injuries on our allies, had nearly provoked us to war. He was both himself, of his own, a vehement man, and he defended Mithridates, the sharpest enemy of this empire, driven from Pontus, with his own resources and kingdom; and when he was driven out by L. Lucullus, that highest man and general, he yet remained, with hostile mind, in his pristine intent with his remaining forces. This man Cn. Pompeius, when he had seen him cast down as suppliant in his own camp, raised up; and the royal insignia, which he had thrown off from his head, restored, and, with certain things commanded, ordered him to reign; and he thought a king set up by him no less glorious to himself and to this empire than he would seem if bound.

\ciceroSection{59} The man who was himself an enemy of the Roman people, and received the sharpest enemy into his kingdom, who fought, who joined standards, who almost contended for empire --- reigns today, and the name of friendship and alliance, which he had violated by arms, this he has obtained by prayers; that wretched Cyprian, who was always a friend, always an ally, about whom no harsher suspicion was ever brought to the Senate or to our generals --- alive, as they say, and seeing, with his food and clothing, was made public property. There you see why other kings should think their fortune stable, when by the example handed down from that mournful year they see that, through some tribune and six hundred gangs, they can be stripped of their fortunes and stripped of their whole kingdom.

\ciceroSection{60} But they wished even by that business to stain the splendour of M. Cato --- ignorant of what gravity, what integrity, what greatness of soul, what, finally, virtue is worth, which in a savage tempest is calm and shines in the shadows and, driven from its place, yet stays and clings in its fatherland and shines through itself always, nor is ever dimmed by another's filth. They thought that M. Cato was not to be honoured but banished, nor that the business should be entrusted to him but laid upon him --- they who in a contio openly said that they had torn out M. Cato's tongue, which had always been free against extraordinary powers. They will perceive, as I hope, in a short time, that that liberty remains, and that, if it can be, it is even greater because, with those consuls, M. Cato, even when he had despaired that anything could be accomplished by his authority, yet fought with his very voice and grief; and after my departure he so harassed Piso with weeping words --- weeping for my and the commonwealth's misfortune --- that that most ruined and shameless man came near already to repenting of his province.

\ciceroSection{61} Why, then, did he obey the bill? As if indeed he had not, before this, sworn to other laws also which he had thought unjustly carried! He does not offer himself to those rashnesses, that, when it is of no service to the commonwealth, he should deprive the commonwealth of himself the citizen. When I was consul, when he was tribune-elect, he offered his life into the risk; he gave that opinion whose unpopularity he saw must be made good with peril of his head; he spoke vehemently, he acted sharply; what he felt he carried before him; he was leader, author, doer of those things, not because he did not see his own danger, but in so great a tempest of the commonwealth he thought that he should think of nothing except the dangers of the fatherland.

\ciceroSection{62} His own tribunate followed. What shall I say of the singular greatness of his soul and his incredible virtue? You remember that day when, the temple seized by his colleague, with all of us fearing for the life of that man and citizen, he himself with the firmest spirit came into the temple, and by his authority calmed the shouting of men, by his virtue the rush of the wicked. He went into danger then, but he went into it for that cause whose magnitude it is now no longer needful for me to speak of. But if he had not obeyed that most criminal Cyprian bill, that disgrace would still cling to the commonwealth no less; for, the kingdom now being publicly auctioned, the proposal had been made about Cato himself by name; and if he had refused, do you doubt that violence would have been used against him, when all the acts of that year seemed to be shaken through that one man?

\ciceroSection{63} And he saw also this: that, since that stain of the publicly-auctioned kingdom remained on the commonwealth, which now no one could wash away, what good might come to the commonwealth out of evils, this would be more useful to be preserved by him than to be scattered by others. And he, even if by some other violence he were driven out from this city in those times, would have borne it easily. For he who had been without the Senate the previous year --- in which, if he had then come, he could yet have seen me as a partner in his commonwealth's counsels --- could he with even mind have been in this city when I was driven out, and in my name both the whole Senate had been condemned and his own opinion? He yielded to the same time as we, the same fury, the same consuls, the same threats, snares, dangers. We drew up a greater grief; he, no less grief of mind.

\ciceroSection{64} On these so many and so great injuries against allies, against kings, against free communities there ought to have been a complaint of the consuls: in the keeping of that magistracy kings and foreign nations have always been. Was any voice ever heard of those consuls? Although who would listen, if they wished most to complain? Should they complain about the Cyprian king, when they --- a citizen of mine, by no charge of mine, struggling under the name of his fatherland --- not only did not defend me standing but did not even cover me when I was lying down? I had yielded --- if you wish (which it was not) the plebs to have been alienated from me --- to unpopularity; if everything seemed to be moved, to the time; if violence was beneath, to arms; if the partnership of the magistrates, to a pact; if the danger of citizens, to the commonwealth.

\ciceroSection{65} Why, when about a citizen's life --- I do not argue what kind of citizen --- and goods a proscription was being carried, when both by the sacred laws and by the Twelve Tables it was sanctioned that no privilege might be enacted against anyone, and that nothing about a citizen's life might be put to a vote except in the centuriate assembly --- was no voice heard from the consuls, and was it laid down in that year, so far as it lay in those two pestilences of this empire, that, by right, by stirred-up gangs, any citizen at all might be cast out by name from the city by the assembly of a tribune of the plebs?

\ciceroSection{66} What of the things that were promulgated in that year, what promised to many, what drawn up, what hoped for, what plotted --- what shall I say? What place in the world was not by now destined for someone? What care of any public business could be thought of, wished for, fashioned, that had not been allotted and parcelled out? What kind of command, or what province, what plan either of casting or of melting down money was not being found? What region or coast of the lands was wider in which some kingdom was not being set up? Who was a king who, in that year, did not think he must either buy what he did not have, or buy back what he had? Who was seeking from the Senate a province, who money, who an embassy? For men condemned of violence, restitution; the canvassing for the consulship was being prepared for that very popular priest. These things the good men groaned at, the wicked hoped for; the tribune of the plebs was acting; the consuls were helping.

\ciceroSection{67} Here at last, later than he himself wished, Cn. Pompeius --- with the most unwilling of those who, by their own counsels and feigned terrors, had turned away the mind of that best and bravest man from the defence of my preservation --- stirred up that his own habit of conducting the commonwealth well, not put to sleep but retarded by some suspicion. He did not bear it, that man who had subdued the most criminal citizens, the sharpest enemies, the greatest nations, kings, wild and unheard-of races, an infinite band of pirates, even slaves, by virtue and by victory; who, with all wars by land and sea pressed down, had set the empire of the Roman people within the bounds of the world --- he did not bear that the commonwealth should be overturned by the wickedness of a few, which he himself had often saved not only by counsels but even by his blood. He came forward to the public cause, by his authority resisted the rest, complained about the past.

\ciceroSection{68} A certain inclination toward better hope seemed to come about. The Senate in full session decreed concerning my return on the Kalends of June, no one dissenting, on the motion of L. Ninnius, whose loyalty and virtue in my cause never trembled. That Ligus, whoever he was, an addition to my enemies, interposed. The matter, and our cause, was already in such a place that it seemed to lift its eyes and live. Whoever there was who had touched any part of the Clodian wickedness in my mourning, wherever he had come, whatever trial he had undergone, was being condemned: no one was found who would confess that he had cast a vote concerning me. My brother had departed from Asia in great sorrow, but in still much greater grief. To him, coming to the city, the whole community had come out with tears and groaning. The Senate spoke more freely; the Roman knights ran together. That Piso, my son-in-law, to whom it was not allowed to bear the fruit of his devotion either from me or from the Roman people, was demanding from his kinsman his father-in-law; the Senate was rejecting everything unless the consuls had first put the question about me.

\ciceroSection{69} When the matter was now being held in our hands, and when the consuls, by the pact of the provinces, had lost all freedom --- who, when private men in the Senate were demanding that they put the question on opinions about me, said that they feared the Clodian law: when they could no longer sustain this, a plot is entered into about the destruction of Cn. Pompeius. With which laid bare and the steel discovered, that man was shut up at home as long as my enemy was in his tribunate. About my return eight tribunes promulgated bills; from which it was understood that, in my absence, friends had not increased to me --- especially in that fortune in which there were not, even, some whom I had thought to be friends --- but that they had had always the same wish, not always the same liberty; for of the nine tribunes I had then had, one defected in my absence, who took for himself a surname from the masks of the Aelii, that he might seem to be more of that nation than of that race.

\ciceroSection{70} In this year, then --- the new magistrates being designated --- when all good men had turned every hope of a better state upon their loyalty, P. Lentulus, the chief, by his authority and his opinion, with Piso and Gabinius opposing, took up the cause; and on the proposition of eight tribunes of the plebs spoke the most outstanding opinion concerning me. He, when he saw it would pertain more to his own glory and to the favour of so great a kindness for that cause to be reserved whole for his consulship, yet preferred such a thing to be brought about more swiftly through others than more slowly through him.

\ciceroSection{71} At this same time, members of the jury, P. Sestius, while tribune-elect, undertook a journey to C. Caesar for my preservation. He thought it pertained both to the concord of the citizens and to the facility of carrying it through that Caesar's mind should not shrink from the cause. What he did, how much he made progress, has nothing to do with the case. For my part I judge that, if he were equal to us, as I think, nothing came forward from this man's effort; but if he were rather angry, not much. Yet you see the man's earnestness and integrity. I now enter on Sestius's tribunate, for this first journey, while tribune-elect, he undertook for the sake of the commonwealth. That year went out; men seemed to have drawn breath, not yet in fact, but in hope of the commonwealth's recovery. Two vultures in their general's cloaks went out with evil omens and curses. Would that those things which men then prayed for had befallen the men themselves! We should not have lost the province of Macedonia with its army, nor the cavalry in Syria and the best cohorts.

\ciceroSection{72} The tribunes of the plebs enter on their magistracy, all of whom had affirmed that they would promulgate concerning me. Of these the chief was bought up by my enemies --- the man whom men in mourning, in derision, used to call Gracchus, since this was also the fate of the city, that little dormouse drawn out from the briar-patches should try to gnaw the commonwealth. The other --- not that Serranus from the plough, but a man grafted from the deserted threshing-floor of Gavius Olelus, after the Gavii had been called out, into the Calatine Atilii --- suddenly, his name being entered into the records, took his name off the record. The Kalends of January come. You can know these things better; for my part, I say what I have heard. What was the crowding of the Senate; what the expectation of the people; what the gathering of envoys from all Italy; what the virtue, action, gravity of the consul P. Lentulus; what the moderation also of his colleague concerning me. Who, when he had said that enmities had been undertaken by him with me from a difference about the commonwealth, said he would commit them to the senators and to the times of the commonwealth.

\ciceroSection{73} Then the chief, being asked his opinion, L. Cotta said that which was most worthy of the commonwealth: that nothing had been done about me with right, nothing in the manner of our ancestors, nothing in the laws; that no one could be removed from the city without trial; that about a citizen's life nothing could not only be carried but not even decided except in the centuriate assembly; that that had been violence, the flame of a shaken commonwealth and of disturbed times; that, with the right and the courts removed, with a great revolution impending, I had bent aside a little and, in hope of the remaining tranquillity, had fled the present waves and tempest; therefore, since I in my absence had no less freed the commonwealth from great dangers than at one time present, that not only should I be restored but even adorned by the Senate. He argued also many things prudently, that the most senseless and most ruined enemy of decency and modesty had so written about me what he had written, with such words, things, opinions, that, even if it had been carried by right, still it could not have force; therefore that I, who was absent under no law, should not be restored by a law, but called back by the Senate's authority.

\ciceroSection{74} Of this man there was no one who would not say that he held the truest opinion. But after him Cn. Pompeius, being asked, having approved and praised Cotta's opinion, said that he, for the sake of my peace, that I might be free from every popular stirring, was of opinion that to the authority of the Senate the kindness also of the Roman people should be added towards me. When all in rivalry, one and another, more weightily and more honourably, had spoken concerning my preservation, and the division was being made without any variation, he stood up, as you know, this Atilius Gavianus; and although he had been bought, he did not dare to interpose; he asked for the night for deliberation. The Senate's outcry, complaints, prayers, my father-in-law cast at his feet. He affirmed that on the next day he would make no delay. He was believed; the meeting broke up. To that deliberator, meanwhile, with the long night interposed, the price was doubled. The days followed --- few altogether in the month of January --- in which it was lawful for the Senate to be held; but yet nothing was done except about me.

\ciceroSection{75} When by every delay, mockery, calumny the Senate's authority was being impeded, there came at last the day for an assembly to be held about me, the eighth before the Kalends of February. The author of the bill, a man most friendly to me, Q. Fabricius, occupied the temple a good while before dawn. Sestius was quiet that day --- this man who is the defendant on a charge of violence; this actor and defender of my cause goes nowhere forward, awaits the counsels of my enemies. What? Those by whose counsel P. Sestius is called into court --- in what manner do they bear themselves? When they had occupied the Forum, the Comitium, the Curia long before night with armed men and slaves for the most part, they make an attack on Fabricius, lay hands on him, kill some, wound many.

\ciceroSection{76} Coming into the Forum the best and most steady man, M. Cispius, tribune of the plebs, they drove out by violence; they made the greatest slaughter in the Forum; and all of them, with drawn and bloody swords in every part of the Forum, with their eyes sought my brother, the best man, the bravest, and most loving of me, with their voice demanded him. To whose weapons that man would gladly, in such grief and longing for me, have offered his own body, not for the sake of fighting back but of dying, had he not kept his life for the hope of my return. Yet he underwent that wicked violence of those criminal bandits; and, when he had come to entreat my preservation from the Roman people, driven from the Rostra he lay in the Comitium, and covered himself with the bodies of slaves and freedmen, and at that time defended his life with the protection of night and flight, not of right and of the courts.

\ciceroSection{77} You remember then, members of the jury, the Tiber filled with citizens' bodies, the sewers stuffed, the blood wiped up from the Forum with sponges, so that all judged so great a force and so magnificent a furnishing to be no private or plebeian but patrician and praetorian thing. You bring no charge against Sestius either before this time or on this very most turbulent day. Yet violence was busy in the Forum. Certainly: for when greater? Stonings we have very often seen, not so often, but yet too often, swords; but slaughter so great, so great heaps of bodies piled up --- except perhaps on that Cinnan and Octavian day --- who ever saw in the Forum? From what stirring of minds? For from the obstinacy or steadiness of an interceder there often arises sedition, from the fault and wickedness of a proposer with some advantage offered to the inexperienced or with bribery, from the wrangling of magistrates; it arises gradually first from shouting, then from some breaking-up of the contio, scarcely late and rarely it comes to blows. But who has heard of a sedition stirred up at night, with no word said, no contio summoned, no law carried?

\ciceroSection{78} Or is it likely that any Roman citizen or any free man came down into the Forum with a sword before daylight to keep the bill about me from being carried, except those who long since have been fattened on the blood of the commonwealth by that pestilential and ruined citizen? Here now I ask the very accuser, who complains that P. Sestius was in his tribunate with a multitude and with a great guard, whether he was there on that day. Certainly he was not. The cause of the commonwealth, then, was beaten, and beaten not by auspices, not by intercession, not by votes, but by violence, by the hand, by the sword. For if that praetor had announced against Fabricius who had said he had observed the heaven, the commonwealth would have taken a blow, but one which it could lament after taking; if his colleague had interposed against Fabricius, he would have hurt the commonwealth, but he would have hurt it by tribunician right. Will you let loose new-bought gladiators, foisted in for an expected aedileship, with assassins released from prison, before daylight? Will you cast down a magistrate from the temple, do the greatest slaughter, sweep the Forum? And, when you have done everything by violence and by arms, will you accuse him who has fortified himself by a guard, not to attack you, but to be able to defend his own life?

\ciceroSection{79} Yet from that very time Sestius did not aim at this --- that, fenced in by his own men, he should safely hold his magistracy in the Forum, govern the commonwealth. So, trusting in the sanctity of the tribunate, when he thought himself armed by the sacred laws not only against violence and the sword but even against words and interruption, he came into the temple of Castor, and announced against the consul. When suddenly that Clodian band, often by now victorious in the slaughter of citizens, shouts, is set on, attacks; some assail with swords the unarmed and unprepared tribune, others with broken pieces of the railings and clubs; from whom this man, having received many wounds, his body weakened and torn to pieces, threw himself down half-dead, and by no other means did he ward off death from himself except by the appearance of death. When they saw him lying on the ground and slashed with very many wounds, with the last breath, bloodless and finished, more from weariness and mistake than from pity or moderation they at last left off cutting at him.

\ciceroSection{80} And does Sestius plead his case on a charge of violence? Why so? Because he is alive. But that is no fault of his. The one final blow was wanting, which, had it been added, would have drawn out the remaining breath. Accuse Lentidius: he did not strike the place. Curse Titius, that Sabine from Reate, why he so rashly cried out that the man was killed. As for Sestius himself, what do you accuse? Did he fail the swords? Did he fight back? Did he, as gladiators are commonly bidden, refuse the iron? Or is this very thing the violence --- not to be able to die? Or is it that, as tribune of the plebs, he stained the temple with blood? Or that, when he had been carried away and first come to himself, he did not order himself to be carried back? Where is the charge? What do you reprove?

\ciceroSection{81} Here I ask, members of the jury: if on that day that Clodian people had brought about what they wished, if P. Sestius, who has been left for dead, had been killed --- would you have gone to arms? Would you have stirred your minds to that ancestral spirit and the virtue of your forefathers? Would you have, at last, demanded back the commonwealth from a deadly bandit? Or would you even then be quiet, hesitate, fear, when you saw the commonwealth crushed and trampled on by the most criminal assassins and by slaves? Of him whose death you would avenge, then, if indeed you were thinking of being free and of having a commonwealth --- about the virtue of him alive, what you ought to say, what to feel, what to think, what to judge --- do you think it should be doubted?

\ciceroSection{82} But indeed those very murderers, whose unbridled rage is fed by long impunity, so shuddered at the violence of their deed that, if the report of Sestius's death had been a little longer, they had thought of killing that Gracchus of theirs in order to transfer the charge upon us. The little countryman --- not careless --- perceived (for the worthless men could not keep silent) that his own blood was being sought to extinguish the unpopularity of the Clodian deed; he snatched up a muleteer's cloak with which he had first come to Rome for the elections, and covered himself with a reaper's basket. When some sought ``Numerius,'' others ``Quintius,'' he was saved by the mistake of the twin name. And you all know how far the man was in danger till it became known that Sestius was alive; which, had it not been disclosed a little sooner than I should have wished, they would not have been able by the death of their hireling to transfer the unpopularity to those whom they thought, but they would have lessened the infamy of a most bitter wickedness by a kind of welcome wickedness.

\ciceroSection{83} And if then P. Sestius, members of the jury, in the temple of Castor had given up that life which he scarcely retained, I do not doubt that, if only there were a Senate in the commonwealth, if the majesty of the Roman people had revived, at last a statue would have been set up in the Forum to him, killed for the commonwealth's sake. Nor indeed any of those whom you see, set in that place and on the Rostra by our ancestors after their death, ought to be preferred to P. Sestius either in bitterness of death or in spirit towards the commonwealth: who, when he had taken up the cause of an afflicted citizen, the cause of a friend, the cause of a man well-deserving of the commonwealth, the cause of the Senate, the cause of Italy, the cause of the commonwealth, and when, obeying the auspices and religion, he was announcing against what he had perceived, in broad daylight by criminal pestilences, in sight of gods and men, was killed in the most sacred temple, in the most sacred cause, in the most sacred magistracy. Will anyone, then, say his life should be stripped of its honours, whose death you would think should be honoured by an everlasting monument?

\ciceroSection{84} ``You bought, gathered, made ready men,'' he says. To do what? To besiege the Senate? To drive out uncondemned citizens? To plunder goods? To set fire to houses? To pull down roofs? To kindle the temples of the immortal gods? To drive out the tribunes of the plebs from the Rostra by the sword? To sell whatever provinces he wished to whomever he wished? To name kings? To bring back, through our envoys, into free communities men condemned of capital matters? To hold the chief man of the city beset by the sword? That he might be able to bring about these things, which could in no way come about except with the commonwealth crushed by arms --- for that purpose, I take it, P. Sestius prepared a band and forces. But it was not yet ripe; the matter itself did not yet drive good men to such guards. We were defeated, not at all by that band alone, but yet not without that: you mourned in silence.

\ciceroSection{85} The Forum had been seized the previous year, the temple of Castor as some citadel occupied by fugitives: there was silence. Everything was being done by the shouting, running together, violence, and force of men ruined by both want and recklessness: you bore it. Magistrates were being driven from the temples; others were altogether kept from approach and from the Forum: no one resisted. Gladiators from the praetor's escort, seized, brought into the Senate, confessing, thrown into chains by Milo, released by Serranus: no mention. The Forum strewn with the bodies of Roman citizens by night-slaughter: not only no new inquiry, but even the old courts were abolished. You saw a tribune of the plebs, having received more than twenty wounds, lying and dying. Of the other tribune of the plebs --- a divine man (for I shall say what I feel, and what all feel with me): a divine man, endowed with a singular, unheard-of, new greatness of soul, gravity, faith --- the house was attacked by sword, by torches, by the Clodian army.

\ciceroSection{86} And you in this place praise Milo, and you praise him rightly. For when did we ever see a man of such immortal virtue? Who, with no reward set before him except this --- which now is thought worn out and despised --- the verdict of the good, took up all dangers, the highest labours, the gravest contentions and enmities; who alone of all citizens, it seems to me, taught by deed, not by words, both what ought to be done by outstanding men in the commonwealth, and what is necessary --- that the wickedness of reckless men, overturners of the commonwealth, must be resisted by laws and by courts; if the laws should not have force, the courts not exist, if the commonwealth should be held down by violence and the consensus of reckless men with arms, that life and liberty must be defended with guard and forces. To feel this is of prudence; to do it, of bravery; both to feel and to do is of perfect and heaped-up virtue.

\ciceroSection{87} T. Annius Milo went forward to the commonwealth as tribune of the plebs --- about whose praise I shall speak more, not because he himself rather wishes these things to be said than thought, or I gladly bestow on him present the fruit of praise, especially since I cannot in words attain it, but because I judge that, if I have proved Milo's cause praised by the very voice of the prosecutor, you will judge Sestius's case in this charge equal: T. Annius, then, came to the cause of the commonwealth in such a way that he wished to recover the citizen torn away from his fatherland. The case simple, the plan steady, full of consensus of all, full of concord. He had his colleagues as helpers; the highest zeal of one consul, the spirit of the other almost reconciled; of the praetors one alien, the incredible wish of the Senate, the spirits of the Roman knights stirred to the cause, Italy upright. Two alone had been bought to obstruct; if those despised and contemptible men had not been able to sustain so great a thing, he saw he would carry through with no labour the cause he had taken up. He acted by authority, he acted by counsel, he acted through the highest order, he acted by the example of good and brave citizens: what was worthy of the commonwealth, what worthy of himself, who he himself was, what he should hope, what render to his ancestors, he most diligently considered.

\ciceroSection{88} To this gravity of the man that gladiator saw himself, if he should act by character, unable to be equal; he turned to the sword, the torches, daily slaughter, fires, plunderings with his army; he began to attack his house, to meet him on his ways, to provoke and terrify by violence. He did not move the man of the highest gravity and the highest steadiness; but although grief of mind, innate liberty, prompt and outstanding virtue urged the bravest man that he should break and beat back violence by violence, especially when it was so often offered, so great was the man's moderation, so great his counsel, that he kept down his grief and did not avenge himself by the same thing by which he had been provoked, but to bind that man, by the snares of the laws, if he could --- that man already exulting and dancing on so many funerals of the commonwealth.

\ciceroSection{89} He came down to accuse. Who ever did so, on so peculiar a public ground, with no enmities, no rewards, no demand of men or even opinion that he would ever do it? The man's spirits were broken; for, with this man as accuser, he was despairing of the disgrace of that pristine court of his. But behold, a consul, a praetor, a tribune of the plebs put forth new edicts of a new kind: that the defendant should not be present, should not be summoned, should not be inquired about, that no one should be allowed at all to make mention before the jurymen or of the courts! What was a man, born for virtue, standing, glory, to do, with the violence of criminal men strengthened, with laws and courts abolished? Should the tribune of the plebs, the most outstanding man, give his neck to a private man, the most ruined? Or should he afflict the cause he had taken up? Or should he stay at home? Both to be conquered he thought disgraceful, and to be deterred, and to be secretly snatched away. He aimed at this --- that, since he could not use the laws against him, he should not fear that man's violence either in his own peril or in the commonwealth's.

\ciceroSection{90} How then, in this kind of guard prepared, do you accuse Sestius, when you praise Milo? Or does he who defends his own roof, who from his hearth and altars wards off sword and flame, who wishes that he might be allowed to be safe in the Forum, in the temple, in the Curia --- does he by right prepare a guard? While he who, by the wounds which he sees daily on his whole body, is warned to protect his head and his neck and his throat and his sides by some guard --- this man do you think should be accused of violence?

\ciceroSection{91} For who of us, members of the jury, is ignorant that the nature of things has so brought it about that, at a certain time, men --- with neither natural nor civil right yet set down --- were spread out through the fields and dispersed and wandering, and had only so much as they had been able by hand and force, through slaughter and wounds, either to snatch away or to keep? Those, then, who first stood out as outstanding in virtue and counsel, having seen through the kind of human teachableness and talent, gathered the scattered into one place, and led them across from that wildness to justice and to gentleness. Then they fenced about with walls those things for the common usefulness which we call public, then the meeting-places of men, which afterwards were called communities, then the joined dwellings, which we call cities, and they invented both divine and human right.

\ciceroSection{92} And between this life polished by humanity and that monstrous one nothing differs so much as right and violence. Of these, whichever we are unwilling to use, we must use the other. We wish violence to be extinguished; right must necessarily prevail --- that is, the courts, in which all right is contained. The courts displease, or there are none; violence must necessarily lord it. All see this. Milo both saw it and acted, that he should try right, drive off violence. He wished to use the one, that virtue might overcome recklessness; he used the other necessarily, that virtue might not be overcome by recklessness. The same reckoning was Sestius's --- if not in accusing, for it was not necessary that the same be done by all, then certainly in the necessity of defending his own preservation, and of preparing a guard against violence and the hand.

\ciceroSection{93} O immortal gods! What outcome do you show us? What hope do you give the commonwealth? How rare a man will be found of such virtue that he should embrace each best cause of the commonwealth, devote himself to the good men, seek solid and true praise --- when he knows that those two near-fates of the commonwealth, Gabinius and Piso, the one daily draws an unreckonable weight of gold from the most peaceful and most opulent treasures of Syria, brings war on the quiet, that he may pour out their old and untouched riches into the deepest gulf of his own lusts, builds a villa in the sight of all so great that now that hovel seems to be a villa which the very tribune of the plebs once unfolded painted in contiones --- by which a chaste and not greedy man called the bravest and supreme citizen into unpopularity;

\ciceroSection{94} the other first sold peace to the Thracians and Dardanians at the highest price, then, that they might be able to make up the money, handed over Macedonia to be harassed and stripped by them, and that same man divided the goods of creditors, of Roman citizens, with their Greek debtors, exacts the largest sums from the people of Dyrrhachium, plunders the Thessalians, has imposed a fixed sum yearly on the Achaeans, and yet has left in no public or sacred place a statue or picture or ornament; and they so make sport, on whom every punishment and every penalty is owed by the best right, while the defendants are these two whom you see. I leave aside now Numerius, Serranus, Aelius --- the sweepings of the Clodian sedition; but yet these too even now flit about, as you see; nor, while you fear something for yourselves, will they ever fear for themselves.

\ciceroSection{95} For what shall I say of the aedile himself, who has even named a day and accused Milo on a charge of violence? Yet this man will never by any injury be brought to repent of having been of such virtue and so great firmness of mind in the commonwealth. But what young men who see these things --- where shall they bend their minds? He who attacked, broke down, set on fire the public monuments, the sacred buildings, the houses of his enemies; who was always packed with assassins, fenced with armed men, fortified with informers, with whom today there is overflowing abundance; who both stirred up a foreign band of malefactors and bought slaves fit for slaughter and in his tribunate poured out the whole prison into the Forum --- flits about as aedile, accuses one who in some part repressed his exulting fury. This man, who guarded himself in such a way that, in private things, he should defend his household gods, in public things, the rights and auspices of the tribunate, was not allowed by the Senate's authority to accuse with measure the man by whom he himself is wickedly accused.

\ciceroSection{96} No doubt this is what you specifically asked about me in your prosecution: what is our nation of \textit{optimates} --- so you put it. You ask a noble subject for the young to learn, and not a difficult one for me to teach through; about which I shall say a few things, members of the jury, and, as I judge, my speech will not be far from the usefulness of the hearers, nor from your duty, nor from the very case of P. Sestius. There have always been, in this city, two kinds of those who have been eager to bear themselves in public affairs and to bear themselves in them with distinction; of which kinds the one have wished both to be held and to be \textit{populares}, the others \textit{optimates}. Those who wished what they were doing and saying to be pleasing to the multitude were thought \textit{populares}; but those who so bore themselves that their counsels were approved by every best citizen were thought \textit{optimates}.

\ciceroSection{97} Who, then, is this best citizen? In number, if you ask, they are countless; for otherwise we could not stand. There are the leaders of public counsel; there are those who follow their school; there are men of the highest orders, to whom the Curia is open; there are the citizens of the towns, the country folk, the men of business; there are even freedmen \textit{optimates}. The number, as I have said, of this kind is widely and variously diffused. But the whole class, that the mistake may be removed, can be briefly circumscribed and defined. All are \textit{optimates} who are neither guilty, nor wicked by nature, nor furious, nor hindered by domestic evils. So be it, then: those whom you called a nation are those who are both upright and sound and well placed in their domestic affairs. Of these, those who in steering the commonwealth serve the wishes, advantages, and opinions of these are reckoned defenders of the \textit{optimates}, and themselves \textit{optimates} of the gravest and most distinguished citizens, and chief men of the city.

\ciceroSection{98} What, then, is set before these helmsmen of the commonwealth, on which they ought to look and to which they ought to direct their course? That which is most outstanding and most to be wished by all sound and good and blessed men --- \textit{cum dignitate otium}, peace with standing. Those who wish this, all are \textit{optimates}; those who bring it about are thought the highest men and the preservers of the city. For neither does it suit men to be so carried away by the standing of action that they should not look out for peace, nor to embrace any peace which shrinks from standing. And of this peace-with-standing these are the foundations, these the limbs, which must be guarded by the chief men and defended even at the peril of their head: religions, the auspices, the powers of magistrates, the authority of the Senate, the laws, the custom of our ancestors, the courts, the administration of justice, good faith, the provinces, the allies, the praise of empire, military affairs, the treasury.

\ciceroSection{99} To be the defender and patron of these so many and so great things is the part of a great spirit, of great talent and great steadiness. For in so great a number of citizens there is a great multitude of those who, either for fear of penalty, conscious of their own faults, seek new movements and reversals of the commonwealth, or who, on account of an inbred madness of mind, are nourished by the discords of citizens and by sedition, or who, on account of the entanglement of their estate, prefer to burn down in a common fire than in their own. Who, when they have got tutors and leaders for their zeals and vices, waves are stirred up in the commonwealth; so that they who have demanded for themselves the rudders of the fatherland must be wakeful, and must strain with all knowledge and diligence that, those things being preserved which I a little while ago said were the foundations and the limbs, they may keep the course and reach that harbour of peace and standing.

\ciceroSection{100} If I were to deny, members of the jury, that this way is rough and steep, or full of dangers and of snares, I should lie --- especially since I have not only always understood it, but felt it beyond others. With greater guards and resources the commonwealth is attacked than it is defended, because reckless and ruined men are urged on at a nod, and even of their own accord are stirred up against the commonwealth, while the good are somehow slower, and, with the beginnings of things neglected, are at last roused only by necessity itself; so that sometimes by hesitation and slowness, while they wish to keep peace even without standing, they themselves lose both.

\ciceroSection{101} But of those who wished to be defenders of the commonwealth, if they are too light, they fall away; if too timid, they fail. They alone remain and bear all things for the commonwealth's sake who are such as your father, M. Scaurus, was, who from C. Gracchus all the way to Q. Varius resisted all the seditious, whom no violence ever, no threats, no unpopularity shook; or such as Q. Metellus, your mother's uncle, who, when he had as censor marked a flourishing man in the popular kind, L. Saturninus, and had kept the imposter Gracchus from the census against the violence of the stirred-up multitude, and when, alone, he had refused to swear to that law which he had judged was not lawfully carried, preferred to be removed from the city than from his opinion; or --- to leave the old examples, of which there is a store worthy of the glory of this empire, and not to name any of those who live --- such as lately Q. Catulus was, whom neither the storm of danger nor the breeze of office could ever move from his course either by hope or by fear.

\ciceroSection{102} These men imitate, by the immortal gods, you who seek standing, who seek praise, who seek glory! These things are ample, these divine, these immortal; these are celebrated by report, committed to the monuments of the annals, propagated to posterity. There is labour, I do not deny; great dangers, I confess; ``many snares for the good men'' has been most truly said; ``but,'' he says, ``it is folly to demand that which many envy and many seek, unless you bring it to pass with the highest care and labour.'' I should have wished that he had not in another place said what wicked citizens may pick up: ``Let them hate, provided they fear''; for he had given outstanding precepts to the youth.

\ciceroSection{103} But yet this way and method of taking up the commonwealth was once more to be feared, when in many things the zeal of the multitude or the people's advantage disagreed with the usefulness of the commonwealth. The ballot-law was being carried by L. Cassius: the people thought their own liberty was the matter; the chief men disagreed, and in the safety of the \textit{optimates} feared the rashness of the multitude and the licence of the ballot. Ti. Gracchus was carrying his agrarian law: it was welcome to the people; the fortunes of the slenderer seemed to be set up; the \textit{optimates} strove against, both because they saw discord stirred up, and because, when the wealthy were being moved from their long possessions, they thought the commonwealth was being stripped of its defenders. The corn-law C. Gracchus was carrying: a thing pleasant to the plebs, for sustenance was supplied generously without labour; the good men resisted, because they thought the plebs were being called away from industry to laziness, and saw the treasury being drained.

\ciceroSection{104} Many things even within our memory, which I deliberately pass by, were in this contention --- popular desire dissenting from the counsel of the chief men. Now there is nothing in which the people dissents from the chosen and the chief men: it neither demands anything, nor is desirous of new things, and is delighted with its own peace and the standing of every best man and the glory of the whole commonwealth. So seditious and turbulent men, because by no bribery now can they stir up the Roman people, since the plebs, having lived through the gravest seditions and discords, embraces peace, hold hired contiones, and do not aim that they should say or carry what those wish to hear who are at the contio, but by price and wages they bring it about that, whatever they say, those should seem to wish to hear it.

\ciceroSection{105} Do you think that the Gracchi or Saturninus or any of those old men who were held \textit{populares} ever had a hired man in a contio? No one had; for the bribery itself and the hope of the advantage set before them stirred the multitude without any wages. So in those times those who were \textit{populares} did indeed give offence among grave and honourable men, but in the people's verdicts and every sign of approval flourished. They were applauded in the theatre; they obtained by votes what they had striven for; men loved their name, their speech, their face, their gait. But those who opposed that kind were thought grave and great men; they had much weight in the Senate, the most among good men, but they were not pleasing to the multitude; their wish was often offended in the votes; and indeed if any one of them ever received applause, he feared lest he had committed some fault. And yet, if any greater matter was at stake, that same people was most moved by these men's authority.

\ciceroSection{106} Now, unless I am mistaken, the city is in such a state that, if you remove the bands of hired men, all will seem to feel the same about the commonwealth. For in three places especially can the judgement and the will of the Roman people about the commonwealth be signified --- in a contio, in the elections, in the seating at games and at gladiatorial shows. What contio has there been in these years --- one not hired but a true one --- in which the consensus of the Roman people could not be perceived? Many were held about me by that most criminal gladiator, to which no one came uncorrupted, no one upright; no good man could look on that foul face, no good man could hear that fury-driven voice. Those contiones of ruined men were necessarily turbulent.

\ciceroSection{107} About me the consul P. Lentulus held a contio: a gathering of the Roman people came together; all orders, all Italy stood in that contio. He pleaded the case with the highest gravity and abundance of speaking, with such silence, with such approval of all, that nothing ever seemed to have come to the ears of the Roman people so popular. There was brought forward by him Cn. Pompeius, who showed himself to the Roman people not only the author of my preservation but even the suppliant. As his speech in contiones was always grave and welcome, so I contend that never was either the weight of his opinion or the pleasantness of his eloquence greater.

\ciceroSection{108} With what silence the rest of the chief men of the city were heard about me! Whom for that reason I do not name in this place, lest my speech, if I have said too little of any, seem ungrateful, if enough of all, endless. Now then, set against this the contio of that same enemy of mine about that same me, before the true people on the Campus Martius! Who was there who not only did not approve, but did not think it the most unworthy crime that the man not, I shall say, spoke, but lived and breathed? Who was there who did not judge the commonwealth stained by his voice, and himself, if he heard him, bound with crime?

\ciceroSection{109} I come to the elections --- whether you please of magistrates or of laws. We see laws often carried, many of them. I leave aside those that are carried in such a way that scarcely five, and those from another tribe, are found to cast a vote. About me, whom that ruin of the commonwealth said was a tyrant and snatcher of liberty, he says he carried a law. Who is there who confesses that he, when the bill was being carried against me, gave a vote? But when about that same me, by senatus consultum, the matter was being carried in the centuriate assembly, who is there who does not declare that he was present and gave a vote concerning my preservation? Which cause, then, ought to seem popular: that in which all the honours of the city, all ages, all orders consent as one, or that into which roused furies fly together as if to the funeral of the commonwealth?

\ciceroSection{110} Or if Gellius is anywhere there --- a man unworthy both of his brother, that most distinguished man and excellent consul, and of the equestrian order, whose name he keeps although he has done away with its ornaments --- will that be popular? For this man is given over to the Roman people. I have seen nothing more so. Who, when his youth, by reason of the highest offices of his stepfather L. Philippus, that highest man, could have flourished, was so far from \textit{popularis} that he ate up his goods alone; then, from an impure and impudent young man, after he had drawn down his father's estate from the riches of common men to the trifle of the philosophers, he wished to be thought a Greekling and idle, and suddenly gave himself to the study of letters. The Attic readers helped him not at all; his books were even often pawned for wine; the insatiable belly remained, the supplies failed. So he was always involved in the hope of new things; in the peace and tranquillity of the commonwealth he grew old. What sedition was there ever in which he was not the leader? What seditious man was there to whom he was not intimate? What turbulent contio whose stirrer-up he was not? To whom did he ever speak well as a good man? Speak well? Nay, what brave and good citizen did he not most impudently harass? Who, as I think, not for the sake of his lust, but to seem a friend of the people, married a freedwoman as wife.

\ciceroSection{111} He cast a vote about me, he was present, he took part in the feasts and congratulations of the parricides; in which, however, he avenged me, when with that mouth he kissed my enemies. Who, as if by my fault he had lost his goods, is, for that very cause, my enemy because he has nothing. Did I, Gellius, take away your patrimony, or did you eat it up? What? Did you, that gulf and abyss of a patrimony, glut yourself at my peril, so that, if I as consul had defended the commonwealth against you and your fellows, you would not wish me to be in the city? None of yours wishes to look at you; all flee your approach, your speech, your meeting; you your sister's son Postumius, a grave young man with senile judgement, marked, when in a great number he did not appoint you tutor for his children. But, carried away by hatred both of mine and of the commonwealth's name --- of which two he is more hostile to which I do not know --- I have said more than ought to have been said against the most furious and most needy carouser.

\ciceroSection{112} I return to that point: that, when business was being done against me, with the city taken and overpowered, Gellius, Firmidius, Titius, and furies of the same kind were leaders and authors of those hired bands, while the proposer himself shrank in nothing from the foulness, recklessness, filthiness of these. But when about my standing the matter was being carried, no one thought either the excuse of his health or of his old age sufficiently just; there was no one who did not judge that he was recalling the commonwealth into its seat together with me.

\ciceroSection{113} Let us see now the elections of magistrates. There was lately a college of tribunes, in which three were thought least, two vehemently, \textit{populares}. Of those who were not held \textit{populares}, to whom in that kind of hired contiones there was no power of standing up, I see two were made praetors by the Roman people; and so far as I could understand from the talk of the multitude and the votes, the Roman people openly bore --- in itself --- the steady and outstanding spirit of Cn. Domitius in his tribunate, and the loyalty and bravery of Q. Ancharius, that even if they could have done nothing, yet by the very wish itself they were welcome. Now what the estimation of C. Fannius is, we see; what the verdict of the Roman people in his honours will be, no one ought to doubt.

\ciceroSection{114} What of those two \textit{populares}? What did they do? The one, who yet had restrained himself, had carried nothing, had only felt about the commonwealth otherwise than men were expecting --- a man both good and harmless and always approved by good men --- because he understood too little, in his tribunate, what was approved by the true people, and because he thought that to be the Roman people which was at the contio, did not keep that place into which, if he had not wished to be \textit{popularis}, he would most easily have come. The other, who had so flung himself in popular reckoning that he held the auspices, the Aelian law, the Senate's authority, the consul, his colleagues, the verdict of the good men as nothing, sought the aedileship with good men and men of the first rank, but not with the most outstanding resources and influence: he did not carry his own tribe; the Palatine, finally, through which all those pestilences were said to harass the commonwealth, he lost; nor did he carry from those elections anything that good men wished except a defeat. You see, then, that the people itself, so to speak, is now no longer popular, who so vehemently rejects those who are held \textit{populares}, and judges those who are opposed to that kind most worthy of office.

\ciceroSection{115} Let us come to the games; for that aim of your minds and eyes upon me, members of the jury, makes me think that I may now use a more relaxed kind of speaking. The signs of the elections and contiones are sometimes true, sometimes corrupted and spoiled. The seatings of theatres and gladiators are altogether said to be wont, by the lightness of certain men, to raise thin and rare bought applause; and yet it is easy, when this is done, to see in what manner and by whom it is done, and what the unspoiled multitude does. What shall I say now of which men or to what kind of citizen the loudest applause is given? It is not lost on any of you. Granted this is light --- which it is not, since it is bestowed on every best man --- but if it is light, it is light to a grave man; to one indeed who hangs on the lightest things, who is held and led by the rumour and, as they themselves call it, the favour of the people, applause must seem immortality, the hiss death.

\ciceroSection{116} From you, then, Scaurus, especially I ask --- you who put on the most prepared and most magnificent games --- did any of those \textit{populares} look upon your games? Did any commit himself to the theatre and the Roman people? That very greatest player, not only spectator but actor and accompanist, who knows all his sister's interludes, who is brought into a gathering of women in place of a girl-musician, neither looked on your games in that burning tribunate of his, nor any others except those from which he scarcely escaped alive. Once, I say, in all that man committed himself to games --- when in the temple of Virtue honour was being given to virtue, and the monument of C. Marius, preserver of this empire, had supplied a seat of safety to the citizen of his town and the defender of the commonwealth.

\ciceroSection{117} At which time what the Roman people would show that it felt was declared in both kinds: first, when, on hearing the senatus consultum, applause was given by all to the matter itself and to the absent Senate; then to the senators one by one as they returned from the Senate to look on. But when the consul who himself was giving the games sat down, standing with hands stretched out, giving thanks and weeping with joy, they declared their good will and pity for me. But when that frantic man came in with his crazed mind set on, the Roman people scarcely contained itself; men scarcely held back their hatred from his impure and unspeakable body. Voices indeed and stretched-out palms and the shout of curses all poured forth.

\ciceroSection{118} But why do I speak of the spirit and virtue of the Roman people, descrying liberty out of long servitude --- in that man whom even the players, sitting in his presence as he was then canvassing for the aedileship, did not spare? For when the Comedy in the Toga \textit{Simulans} was being acted, as I think, the whole troupe in the most distinct chorus, looking the impure man in the face, declaimed: ``This, Titus, is the sequel after the beginnings and the end of your vicious life ---!'' He sat half-dead; and he who used to fill his contiones with the railing of singers was being thrust out by the very voices of the singers. And since the games have been mentioned, I shall not pass over even this: that in great variety of opinions there was never any place where anything said by the poet seemed to fall to our time, that either escaped the whole people, or which the actor himself did not press out.

\ciceroSection{119} And, please, in this place, members of the jury, do not think me drawn by some lightness into an unaccustomed kind of speaking, if I speak in the court of poets, of actors, of games. I am not so ignorant, members of the jury, of cases, not so unaccustomed in speaking, as to chase after speech of every kind and to pluck and taste every little flower from every side. I know what your gravity, this attendance, that gathering, the standing of P. Sestius, the magnitude of the danger, my age, my office demand. But I have taken upon myself in this place a certain teaching to the youth: who they are who are \textit{optimates}. In the unfolding of which it must be shown that not all are \textit{populares} who are thought so. This I shall most easily attain, if I have brought out the true and uncorrupted judgement of the whole people, and the inmost feelings of the city.

\ciceroSection{120} What was that, when the recent news of that senatus consultum which was passed in the temple of Virtue was carried to the games and the stage, before the largest gathering, the highest artist --- and, by Hercules, always of the best parties in the commonwealth as on the stage --- weeping with new joy and with grief and longing for me mingled, pleaded my cause before the Roman people in much weightier words than I myself could have pleaded for me? For he was pressing out the talent of the highest poet not only by his art, but also by his grief. With what force --- ``He who has aided the commonwealth with steadfast spirit, has stood, has stood with the Achivi'' --- ``With you,'' he was saying, ``has he stood,'' he was pointing to your ranks! He was called back by all --- ``In the doubtful matter has not doubted to offer his life nor spared his head.''

\ciceroSection{121} These things with what great shouts were spoken by him! When now, the gesture set aside, applause was being given to the words of the poet and the zeal of the actor and to our hopefulness: ``the highest friend in the highest war'' --- for the actor himself added that with friendly mind, and perhaps men out of some longing approved --- ``endowed with the highest talent.'' Now those things, with how great a groan of the Roman people, by the same man a little later in the same play, were acted! ``O father ---'' he thought that I, absent, was to be lamented as a father --- I, whom Q. Catulus, whom many others often in the Senate had named the father of the fatherland. With how great weeping over those fires and ruins of ours, when he was bewailing his father driven out, his fatherland afflicted, his house burned and overturned, he so acted that, having shown the former fortune, when he had turned himself, ``I have seen all these things on fire,'' stirred weeping even in enemies and the envious!

\ciceroSection{122} By the immortal gods! What? That same one --- in what manner he said that! These things to me indeed seem so to have been acted and written that they would have seemed splendid to be said, even by Q. Catulus, if he had come back to life; for he used freely to reprove and accuse the rashness sometimes of the people, the error sometimes of the Senate: ``O ungrateful Argives, thankless Greeks, mindless of kindness!'' That indeed was not true; for they were not ungrateful, but wretched, since to give back safety to him from whom they had received it was not allowed; nor was anyone ever more grateful to anyone alone, than were they all together to me. But still, that the most eloquent poet wrote on my behalf, the bravest actor performed --- not only the best, about me, when he was pointing to all orders, accusing the Senate, the Roman knights, the whole Roman people: ``You allow him to be in exile, you allow him to be driven out, having been driven out you bear it!'' What a sign there then was of all, what a declaration of will from the whole Roman people in the cause of a man not \textit{popularis}, I for my part used to hear: those who were present can more easily judge.

\ciceroSection{123} And since my speech has carried me to this point, the actor wept over my misfortune so often when he so grievously was acting my cause that that splendid voice of his was hindered by tears; nor did the poets, whose talents I have always loved, fail my time; and the Roman people approved these things not only by their applause but even by their groan. Was it then proper for Aesopus rather, on my behalf, or Accius, to say these things, if the Roman people were free, or the chief men of the city? I was named in the \textit{Brutus}: ``Tullius, who had set fast liberty for the citizens.'' It was called back a thousand times. Did the Roman people seem to judge too little that it was set up by me and by the Senate that which the ruined citizens charged had been overturned through us?

\ciceroSection{124} But the greatest verdict of the Roman people was declared in the seating of all at the gladiatorial show; for it was the show of Scipio, worthy both of him and of that Metellus to whom it was being given. And that was a kind of spectacle which is celebrated by every kind of crowd and every kind of men, in which the multitude is most delighted. Into this seating P. Sestius, tribune of the plebs, since he was doing nothing else in that magistracy except my cause, came and gave himself to the people --- not from desire of applause, but that our very enemies might see the wish of the whole people. He came, as you know, from the Maenian column: such applause was raised from all the spectacles all the way from the Capitoline, such from the railings of the Forum, that no greater consensus or more open expression of the whole Roman people was ever said to have been in any cause.

\ciceroSection{125} Where then were those moderators of contiones, masters of laws, drivers-out of citizens? Or is there some peculiar people belonging to the wicked citizens, to whom we shall have been offensive and hateful? For my part I judge there is no time of more crowded people than that of gladiators, neither of any contio nor of any elections. This countless multitude of men, this so great signification of the Roman people without any variation of all together, when in those very days the matter about me was thought going to be acted on --- what did it declare except that the safety and standing of the best citizens was dear to the whole Roman people?
