% In Vatinium (in-vatinium) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} \textsc{Cross-examination of P. Vatinius as a witness.} If, Vatinius, I had been willing to look only at what indignity demanded, I would have done what was strongly the pleasure of these gentlemen: dismissed you in silence --- you whose testimony, on account of the foulness of your life and the squalor of your household, is reckoned of no weight. For none of these men thought you ought either to be refuted as a serious adversary or questioned as a scrupulous witness. But a little while ago I was perhaps less restrained than I should have been; for, in the hatred of you --- in which, although on account of your crime against me I ought to surpass everyone, I am yet outdone by almost everyone --- I was so stirred up that, although I despised you no less than I hated you, I nevertheless wished to send you off harassed rather than disregarded.

\ciceroSection{2} And so, lest you wonder that this honour is being paid you by me --- that I question one whom no one judges worthy of meeting, of approach, of the vote, of citizenship, of the daylight --- no cause would have driven me to it but this: to crush that ferocity of yours, to break your audacity, and to slow that loquacity of yours, snared by a few questions of mine. Indeed, Vatinius, even if you had come under suspicion with P. Sestius falsely, you should still have pardoned me, if in the great peril of a man who has deserved best of me I had been willing to comply both with his moment and with his wish.

\ciceroSection{3} But that yesterday, while giving testimony, you lied --- when you affirmed that you had had absolutely no conversation with Albinovanus, not only about prosecuting Sestius, but on any subject whatsoever --- you yourself a moment ago unwittingly disclosed: when you said both that T. Claudius had taken counsel with you and had sought your advice for prosecuting P. Sestius, and that Albinovanus, whom you had previously said was scarcely known to you, had come to your house and spoken much with you, and lastly that the written texts of P. Sestius's contio speeches --- which Albinovanus neither knew nor could find --- you had given to him, and that they had been recited at this trial. In all of which, on the one count you confessed, that the prosecutors were drilled and suborned by you; on the other you refuted your own inconsistency, tangled with both fickleness and even perjury --- when you said that the man you had said was utterly estranged from you had been at your house, and that to a man whom from the beginning you had judged to be a sham-prosecutor you had given whatever books he had asked for in order to prosecute.

\ciceroSection{4} You are too vehement and ferocious by nature: you do not consider it lawful that any word should leave anyone's mouth that does not strike pleasantly and honourably on your ears. You came in furious with all; which I, the moment I caught sight of you, before you began to speak --- while Gellius, the wet-nurse of every seditious man, was previously giving his testimony --- perceived and foresaw. For suddenly, like a serpent from its hole, with eyes bulging, neck inflated, throat swollen, you thrust yourself in, so that there was renewed for me that old assault of yours upon \dots\

\ciceroSection{5} \dots\ I defended my old friend, who was nevertheless an intimate of yours, since in this state assault, of the sort you yourself sometimes employ, is wont to be censured, but defence never. But I ask of you: why was I not to defend C. Cornelius? Did Cornelius carry any law against the auspices? Did he disregard the Aelian, the Fufian law? Did he bring violence to bear on a consul? Did he hold a temple with armed men? Did he throw down a vetoing colleague by force? Did he pollute the rites, drain the treasury, plunder the commonwealth? These are yours, yours all of them: against Cornelius nothing of the sort was charged. He was said to have read the codex; he defended himself, with his own colleagues as witnesses, that he had read it not for the sake of reciting but of verifying. It was settled, however, that Cornelius on that day had dismissed the assembly and obeyed the veto. But you, to whom Cornelius's defence is displeasing --- what case, what face will you bring before your own patrons? To them you are already prescribing how great a reproach it will be to them if they defend you, when you have judged Cornelius's defence to be the sort of thing to be flung at me by way of slander.

\ciceroSection{6} And yet, Vatinius, remember this: a little after that defence of mine which you say was displeasing to good men, I was made consul --- with the highest goodwill of the entire Roman people and the singular zeal of every best citizen --- in the most magnificent fashion within human memory; and that I attained, by living modestly, all those things which you said you so often hoped to attain by predicting them shamelessly. As for the fact that you reproached me with my withdrawal, and that you wished to renew the grief and groaning of these men --- to whom that day was bitterest which to you was sweetest --- this much only I answer you: when you and the other plagues of the commonwealth were seeking a pretext for arms, and were eager, through the use of my name, to plunder the fortunes of the rich, to drink down the blood of the leading men of the state, to glut your cruelty and the long-standing hatred which you had inveterately conceived against good men, I preferred to break your wickedness and madness by yielding rather than by resistance.

\ciceroSection{7} Therefore I beg of you, Vatinius, that you pardon me, since I spared the country which I had saved; and, if I bear with you as the destroyer and harasser of the commonwealth, that you bear with me as its preserver and guardian. Then you reproach the withdrawal of a man whom you see was recalled by the longing of every citizen --- by the very mourning of the commonwealth itself. But you said that men toiled for my recall not on my account but on the account of the commonwealth: as if any man of outstanding spirit who has entered upon public life would judge anything more to be wished for than that he be loved by his fellow citizens for the commonwealth's sake!

\ciceroSection{8} Surely my nature is harsh, my approach difficult, my face severe, my answers haughty, my way of life arrogant; surely no one missed my company, no one my humanity, no one my counsel, no one my help --- by longing for which (to put it at the lowest) the Forum was sad, the Senate-house mute, every pursuit of the good arts at last fell silent. But let nothing have been done on my account: let us grant that all those decrees of the Senate, the bidding of the people, the resolutions of all Italy, of every guild and every association concerning me were made on the commonwealth's account. What then, you who are most untaught of solid praise and true standing, could have happened to me more remarkable? What more to be wished for, towards the immortality of glory and the everlasting memory of my name, than that all my fellow citizens should judge this --- that the safety of the state was bound up with the safety of one man, mine?

\ciceroSection{9} And this thrust of yours I return to you. For as you said that I was dear to the Senate and the Roman people not so much on my own account as on the commonwealth's, so I say of you --- though you are most foul in every fierceness and savagery --- that you are nevertheless hateful to the state not so much on your own account as on the commonwealth's. And, that I may at long last come to you, let this be my last word about myself. What each of us says about himself is, in fact, not a thing to be much asked after; what good men say --- that is what carries the greatest weight and moment.

\ciceroSection{10} There are two times at which the judgements of our fellow citizens about us may be looked at --- one of office, the other of safety. Office on such terms of goodwill from the Roman people has been conferred on few, and on me; safety with so great a zeal of the state has been restored to no one. But what men feel about you we have tested in the matter of office; in the matter of safety we await it. Even so, that I may compare you not with these leading men of the state who are present in support of P. Sestius, but with you yourself, a man not only of the most shameless sort but the foulest and the lowest --- I ask of you, Vatinius, you most arrogant and to me most hostile man, which after all you think to have been better and more excellent for this state, this commonwealth, this city, these temples, the treasury, the Senate-house, these men whom you see, their goods, their fortunes, their children, the rest of the citizens, in a word for the shrines of the immortal gods, the auspices, the rites: that I be born a citizen in this state or you? When you have answered me this --- either so shamelessly that men can scarcely keep their hands off you, or so painfully that those puffed-up parts of you at last burst --- then answer also, with a memory ready, what I shall ask you about yourself.

\ciceroSection{11} And that most shadowy time of the entry of your youth I shall let lie hidden. You may, by my leave, have dug through walls in your adolescence with impunity, plundered your neighbours, beaten your mother: let your worthlessness have this prize --- that the foulness of your youth be covered up by your own obscurity and squalor. You stood for the quaestorship along with P. Sestius; while this man said nothing except what he was actually doing, you said you were thinking of holding a second consulship. I ask of you: do you remember, when P. Sestius was made quaestor by every vote, that you barely, against everyone's wish, not by the people's favour but by the consul's, clung on at the bottom of the list?

\ciceroSection{12} In that office, when, with a great outcry, the water-province had fallen to you by lot, were you not sent by me as consul to Puteoli, to prevent the exporting from there of gold and silver? In which business, when you supposed that you had been sent not as a guard to detain the goods but as a customs-officer to share them out, and when you were rifling the houses, the warehouses, the ships of all men most thievishly, and were entangling business-men in the most inequitable trials, terrifying merchants stepping off ship and delaying those embarking --- do you remember that hands were laid on you at the assembly in Puteoli, that the complaints of the men of Puteoli were brought to me as consul? After your quaestorship, did you go out as legate to Further Spain under the proconsul C. Cosconius? When the journey into Spain is usually completed almost on foot, or, if a man wishes to sail, the route of sailing is fixed --- did you come into Sardinia, and from there into Africa? Were you (a thing it was not lawful for you to do without a decree of the Senate) in the kingdom of Hiempsal, were you in the kingdom of Mastanesosus, did you come to the strait through Mauretania? Whom do you know to have ever, since the founding of the city, reached that province as legate to Spain by such routes as these?

\ciceroSection{13} You were made tribune of the plebs --- for why should I question you about your scandals in Spain and your foulest thefts? I ask of you first, in general: what kind of wickedness and crime did you leave undone in that magistracy? And here at the outset I prescribe to you not to mingle your squalor with the splendour of the most distinguished men. Whatever I shall ask you, I shall ask of yourself alone; nor shall I drag you out from the standing of any most ample man, but from your own darkness; and all my weapons will be so cast at you that no one, as you are wont to say, will be wounded through your flank: they will stick fast in your lungs and your innards.

\ciceroSection{14} And since the beginnings of all great things are drawn from the immortal gods, I want you to answer me, you who are accustomed to call yourself a Pythagorean, and to hold up the name of a most learned man as a screen for your savage and barbarous ways: what so great perversion of mind has held you, what so great madness, that, when you had taken upon yourself unheard-of and impious rites --- when you are accustomed to call up the souls of the dead from below, when you are accustomed to slaughter the shades of the underworld with the entrails of boys --- you have held in contempt the auspices on which this city was founded, on which the whole commonwealth and the empire are sustained, and at the beginning of your tribunate gave notice to the Senate that the responses of the augurs and the arrogance of that college would be no obstacle to your acts?

\ciceroSection{15} Next I ask: did you keep faith in that matter? Did anything ever cause you delay in summoning the assembly and bringing forward a law, because you knew that on that day watch had been kept of the heavens? And since this is the one place where you say you have something in common with Caesar, I shall part you from him --- not only for the commonwealth's sake but also for Caesar's, lest some stain from your utter unworthiness seem to be spattered upon his standing. First I ask: are you, as Caesar does, committing your case to the Senate? Then, what authority belongs to a man who defends himself by another's deed, not his own? Then --- for at last the true voice will burst out of me, and I shall say without hesitation what I think --- if even C. Caesar had been somewhat more violent in some matter, if the greatness of contention, his zeal for glory, his outstanding spirit, his excelling nobility had driven him in some direction, what in that great man would even at the time have been bearable, and would have been blotted out by the very greatest deeds he afterwards performed --- that, gallows-bird, you will take to yourself, and the voice of Vatinius the brigand and sacrilegist will be heard demanding this: that the same be granted him as to Caesar?

\ciceroSection{16} For thus I question you. You were tribune of the plebs --- separate yourself from the consul: you had nine brave colleagues. Of them three you knew were daily keeping watch of the heavens, whom you ridiculed, whom you said were private men. Of these you see two now sitting in the praetexta --- you who sold the aedilician praetexta toga which you had had made up for nothing! --- the third you know, a young man as he was, to have attained from that beleaguered and battered tribunate the standing of a consular. The remaining six there were: of whom some plainly thought with you, some held some middle course; they all had laws promulgated, and among them many were carried by my friend C. Cosconius, our juror here --- whom you nearly burst at the sight of in aedilician office --- carried, and on my motion too.

\ciceroSection{17} I want you to answer me: did any one out of the whole college dare to bring forward a law except you alone? What audacity in you was so great, what so great a force, that what your nine colleagues judged ought to be feared by them, you alone --- emerged from the mire, easily of all men in all things the lowest --- judged ought to be despised, looked down upon, mocked? Do you know any tribune of the plebs, from the founding of the city, to have transacted business with the plebs when it was settled that watch had been kept of the heavens?

\ciceroSection{18} At the same time I want you to answer this also: when, with you as tribune of the plebs, the Aelian and Fufian laws still stood in the commonwealth --- laws that often weakened and held in check the madnesses of tribunes, against which laws no one before you ever attempted to act --- which laws indeed in the year following, with two men sitting in the temple (not consuls, but betrayers and plagues of this state), were burned up together with the auspices, with the vetoes, with all public right: did you ever hesitate to transact with the plebs against those laws, and to convoke the assembly? Of all the tribunes of the plebs, however seditious, have you heard of any so audacious that he ever called an assembly contrary to the Aelian or the Fufian law?

\ciceroSection{19} I ask this also of you: did you attempt, did you desire, finally did you even contemplate --- for the thing is of such a kind that, if it has only come into your mind, no one would judge you unworthy of any torment --- did you contemplate, in that intolerable not-kingship of yours (for that name you long to hear) but brigandage, getting yourself made augur in the place of Q. Metellus, so that whoever caught sight of you would take up a doubled grief and groaning, both from longing for the most distinguished citizen and from the office held by the foulest and the most wicked man? Did you so judge the commonwealth, with you as tribune, not merely shaken loose, not merely the state thrown out of joint, but this city captured and overturned, that we could bear Vatinius an augur?

\ciceroSection{20} At this point I ask: if you had been made augur --- the thing you had coveted, in the very contemplation of which we who hated you could scarcely endure our grief, while those to whom you were a darling could scarcely hold back their laughter --- if to the rest of the wounds by which you thought the commonwealth was being destroyed you had inflicted this mortal blow too of your augurate --- were you going to decree, the thing all augurs from Romulus down have decreed, that with Jupiter flashing lightning it is impious to transact business with the people? Or, because you had always so transacted, were you, augur as you were, about to dissolve the auspices?

\ciceroSection{21} And not to speak any longer about your augurate --- which I do unwillingly, because it sets me to remembering the ruins of the commonwealth; for never, while not only the majesty of these men but even the city itself was standing, did you imagine you would be augur --- but, to leave your dreams and come to your crimes, I want you to answer me: when M. Bibulus the consul --- I shall not call him a man thinking well of the commonwealth, lest you, a powerful fellow, take offence at me, who quarrelled with him, but a man certainly advancing nowhere, attempting nothing in the commonwealth, only in mind dissenting from your acts --- when you were leading him as consul off into chains, and from the Tabula Valeria your colleagues were ordering him to be set free, did you, with the tribunals run together into a continuous platform, make a bridge before the rostra, by which the most temperate and most steadfast consul of the Roman people, with rescue removed, with friends shut out, with the violence of ruined men whipped on, was being led, in a foulest and most miserable spectacle, not to prison, but to punishment, to death?

\ciceroSection{22} I ask: was anyone before you so wicked as to do this, so that we may know whether you are an imitator of old crimes or an inventor of new ones? And, the same you --- when, with these counsels and crimes of this sort in the name of C. Caesar (a most clement and most excellent man, but by your own wickedness and audacity), you had driven M. Bibulus out of the Forum, the Senate-house, the temples, all public places, and were keeping him shut up at home, and when, not by the majesty of his command, not by the right of the laws, but by the guard of his door and the watch of his walls the life of the consul was sheltered --- did you send a beadle to drag M. Bibulus from his house by force, so that, what has always been preserved among private men, that no one's house can be his exile, that should not, with you as tribune of the plebs, hold for a consul?

\ciceroSection{23} At the same time answer me this --- you who call us, who agree on the common safety, tyrants --- were you not, not a tribune of the plebs, but some intolerable tyrant out of the mire and out of the darkness, who first attempted to overturn the very commonwealth which had been founded by the discovery of the auspices, by the abolition of those same auspices; then who alone trampled underfoot and reckoned as nothing those most sacred laws --- I mean the Aelian and Fufian --- which lived through the ferocity of the Gracchi, the audacity of Saturninus, the swill of Drusus, the contention of Sulpicius, the gore of Cinna, even amid the arms of Sulla; who exposed a consul to death, besieged him shut up at home, tried to drag him out from under his own roof; who in that magistracy not only emerged from beggary but now even strikes us with terror at the thought of his wealth?

\ciceroSection{24} Were you of such cruelty as to attempt by your bill to lift up and destroy chosen men and the leading men of the state, when L. Vettius --- who had confessed in the Senate that he had been carrying a weapon, that he had wanted with his own hands to offer death to Cn. Pompeius, the highest and most illustrious of citizens --- you brought before the contio, set the informer on the rostra, in that augurated temple and place, I say, in which other tribunes of the plebs, for the sake of seeking authority, are accustomed to bring forward the leading men of the state? There you wished the informer Vettius to lend his tongue and his voice to your wickedness and your madness. Did L. Vettius, at your contio, asked by you, say that he had had as authors and instigators and partners of that crime those men whose removal from the state --- which you at that time were attempting --- would have left the state unable to stand? M. Bibulus, with whose confinement you were not content, you had wanted to kill, you had stripped of his consulship, you were longing to deprive of his country; L. Lucullus, whose achievements --- because you yourself, evidently from boyhood, had had your eye on the praises of imperators --- you envied the more vehemently; C. Curio, the perpetual enemy of all the wicked, the author of public counsel, the freest of men in defending the common liberty, together with his son, leading man of the youth, even more closely bound up with the commonwealth than ought to have been required of that age --- these you wished to destroy;

\ciceroSection{25} L. Domitius, whose dignity and splendour was, I suppose, dazzling Vatinius's eyes --- whom for the common hatred of the good men you hated, but for the future, on account of all the hope that there is and was concerning him, you had for some while been afraid of; L. Lentulus, this our juror, the Flamen Martialis, because at that time he was the rival of your Gabinius --- by the information of the same Vettius you wished to crush him: who, if he had then conquered that plague and ruin (which by your wickedness was not allowed him), the commonwealth would not have been conquered. His son too, by the same information and the same charge, you wished to add to his father's destruction; L. Paulus, who at that time as quaestor was holding Macedonia --- what a citizen, what a man! who by the laws had banished from the country two impious betrayers, enemies in the household, a man born for preserving the commonwealth --- you herded into the same information of Vettius and into the same number with these.

\ciceroSection{26} What then am I to complain of for myself? --- I who actually owe you thanks for not having thought I ought to be cut off from the number of the bravest citizens. But what was that great madness of yours: that, when Vettius had already finished his speech at your bidding and had marked out the lights of the state and had stepped down from the rostra, you suddenly recalled him, conferred with him in the sight of the Roman people, and then asked him whether he could name any others? Did you not press him to name C. Piso, my son-in-law, who --- in the greatest abundance of the best young men, with their equal continence, virtue, piety --- has left no peer; and likewise M. Laterensis, a man who day and night ponders the praise and the commonwealth? Did you not, you most foul enemy, promulgate an inquiry against so many most ample men of such standing, with the informer's reward for Vettius, with the most lavish prizes? When all these things had been rejected by all mortals, not by their will but by their outcry, did you not break the very Vettius's neck in prison, that there might survive no informer's record from a bribed informer, and an inquiry into that same crime might not be demanded against yourself?

\ciceroSection{27} And since you frequently invoke the law you carried about alternate rejections of jurors --- that all may understand you could not do even a right thing without crime --- I ask: when, an equitable law having been promulgated at the start of your magistracy, you had already by then carried many others, did you wait until C. Antonius should have been brought to trial before Cn. Lentulus Clodianus and, after he had been brought to trial, then straightway pass against him --- a man made defendant after your law --- so that a consular, shut out, miserable, in the puncture of a moment, was stripped of the benefit and the equity of your own law?

\ciceroSection{28} You will say you had a familiarity with Q. Maximus. A splendid defence of your crime! For Maximus's highest praise is this very thing: that, with the enmity taken up, the case undertaken, the questor and the bench chosen, he was unwilling to give to his enemy a more advantageous arrangement of rejections. Maximus did nothing foreign either to his own virtue or to those most distinguished men, the Pauli, the Maximi, the Africani, whose glory we not only hope to be renewed in his virtue, but already see renewed: your fraud, your evil-doing, your crime is this, that you put off to a moment of cruelty what you had promulgated under the name of mercy. And now C. Antonius consoles his misery only by this one thing, that he has chosen to hear, rather than to see, the images of his father and his brother and his brother's daughter placed not in the family hall but in the prison.

\ciceroSection{29} And since you despise the moneys of others, and boast most intolerantly of your own riches, I want you to answer me: did you, while tribune of the plebs, make treaties with cities, with kings, with tetrarchs? did you allocate sums from the treasury by your laws? did you snatch shares, then dearest of all, partly from Caesar and partly from the publicans? Since these things are so, I ask of you: have you not become rich, from a most poor man, in that very year in which the most stringent law about extortion was passed --- so that all might understand that, by you, not only our acts (we whom you call tyrants), but even your closest friend's own law has been treated with contempt? before whom you are even accustomed to bring charges against us, who are his closest friends, when you yourself, every time you say you are his connection, abuse him most contumeliously.

\ciceroSection{30} And this also I desire to learn from you: with what counsel or with what intent did you do it that at the public feast of my friend Q. Arrius you reclined in a black toga? Whom did you ever see do so, whom did you ever hear of? By what precedent, by what custom did you do it? You will say that you did not approve those thanksgivings. Splendid: let there have been no thanksgivings. Do you see that I am asking nothing of you about the case of that year, nothing about what may seem common to you with the highest men, but only about your own particular crimes? Let there have been no thanksgiving. Tell me, who ever dined in mourning? For that feast is funereal in this sense, that it is an offering belonging to the funeral, but the feast itself is of dignity.

\ciceroSection{31} But I leave aside the public feast of the Roman people, the festal day with its silver, its dress, all its display and ornament worth seeing. Who ever, in a private mourning, who ever at a family funeral, dined in a black toga? To whom, when he was coming out from the baths, was a black toga ever brought, except to you? When so many thousand men were reclining at table, when the master of the feast Q. Arrius himself was robed in white, you, with C. Fibulus in black and the rest of your furies, brought a funereal thing into the temple of Castor. Who at that moment did not groan, who did not grieve at the disaster of the commonwealth? What other talk was there at that feast but this --- that this state, so great and so weighty, was thrown beneath, not your madness only, but even your mockery?

\ciceroSection{32} Were you ignorant of this custom? Had you never seen a public feast? Had you never as a boy or a young man been among cooks? Had you not, a little before, glutted that old hunger of yours at the most magnificent feast of the most noble young man Faustus? Whom had you seen recline there in black? The master of the house and his friends in black togas before the dinner? What so great derangement held you, that, unless you had done what was not lawful, unless you had violated the temple of Castor, the name of the feast, the eyes of the citizens, the ancient custom, the authority of the man who had invited you, you thought it too feebly attested that you held those thanksgivings to be nothing?

\ciceroSection{33} I ask you this also, about something you committed as a private man --- in which indeed you will now no longer be allowed to say that your case is bound up with that of the most distinguished men: were you not arraigned under the Licinian and Junian law? did not C. Memmius the praetor decree, by that law, that you appear on the thirtieth day? When that day came, did you not do what in this commonwealth not only had never been done before, but is, in all memory, altogether unheard-of? did you not appeal to the tribunes of the plebs, in order not to plead your case --- I have spoken too lightly: that very thing would have been both novel and not to be borne --- did you not appeal by name to the plague of that year, the fury of the country, the storm of the commonwealth, Clodius? Who, nevertheless, when by right, by custom, by his power he could not block the trial, returned to that violence and madness of his and offered himself as leader to your soldiers. In which matter, lest you suppose anything to have been said by me against you rather than asked of you, I shall lay on myself no burden of testimony: what in a short time I see I shall have to say from this same place, I save up; and I shall not arraign you, but, as in the other matters, I shall ask.

\ciceroSection{34} I ask you, Vatinius: did anyone in this state, since the founding of the city, appeal to tribunes of the plebs that he might not plead his case? Did any defendant ever mount the tribunal of his own questor and throw him down by force, scatter the benches, hurl down the urns, and finally commit, in the disturbance of the trial, all those things on whose account the trials had been instituted? Do you know that at that time Memmius fled, that your prosecutors were torn out of your hands and your men's, that the presidents of the inquisitions were driven from the next-door tribunals, that in the Forum, in the daylight, with the Roman people looking on, the inquisition, the magistrates, the custom of the ancestors, the laws, the jurors, the defendant, the penalty were swept away? Do you know that all this, by the diligence of C. Memmius, has been entered and attested in the public records? This also I ask: when, after you had been arraigned, you had returned from your legation --- so that no one should suppose you were dodging the courts --- and when, with the choice between the two open to you, you kept saying that you preferred to plead, what consistency was there, when you had been unwilling to use the refuge of the legation, in your having fled, by a most shameless appeal, to wicked rescue?

\ciceroSection{35} And since mention has been made of your legation, I want to hear from you: by what decree of the Senate, after all, you became legate. From your gesture I understand what you answer: by your own law, you say. Are you then the very surest parricide of the country? Were you setting your eye on this --- that the senators should be utterly swept out of the commonwealth? Did you not even leave to the Senate this, which no one ever took away --- that legates be sent by the authority of that order? Did the public council seem to you so squalid, the Senate so beaten down, the commonwealth so wretched and prostrate, that it could no longer choose, after the custom of the ancestors, messengers of peace and war, spokesmen, interpreters, advisers in war-counsel, ministers of provincial service?
