% De Haruspicum Responsis (de-haruspicum-responsis) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Yesterday, senators, when both your dignity and the great press of Roman knights present (to whom the Senate was being granted access) had stirred me deeply, I judged that I had to put down P. Clodius's shameless impudence: when he was hindering the cause of the publicans by the most foolish questionings, doing service for P. Tullio the Syrian, and selling himself --- though he had been bought up entirely --- even before your eyes. So I held in check that frenzied and exulting man as soon as I threatened him with the danger of a trial: with two opening words I crushed all the gladiator's onset and ferocity.

\ciceroSection{2} And yet, ignorant who the consuls were, ashen and hot-flushed, he flung himself out of the Curia all at once, with certain broken and now empty threats, and with those terrors of the Pisonian and Gabinian period: whom, when he had gone out, I began to follow, and I had the greatest reward both from the rising-up of every one of you and from the company of the publicans. But suddenly out of his wits, without his own face, without colour, without voice, he stood still; then he looked back, and as soon as he caught sight of Cn. Lentulus the consul, he collapsed almost at the threshold of the Curia --- by recollection, I suppose, of his Gabinius, and by longing for Piso. Of his unbridled, headlong frenzy what shall I say? Or can he be more grievously wounded by my words than he was, in the very act, dispatched and butchered by that gravest of men, P. Servilius? Whose force and signal, almost divine gravity, even if I could now match, still I do not doubt that the weapons hurled by an enemy will seem lighter and duller than those his father's colleague let fly.

\ciceroSection{3} Yet I wish to set out the reasoning of my act to those who, yesterday, supposed me to have been carried by pain and to have gone almost further in anger than the considered reason of a wise man would have demanded. I did nothing in anger, nothing with an unmastered mind, nothing that was not long thought through and meditated long before. For I, senators, have always declared that I am a foe to two men: who, when they ought to have defended me and the commonwealth, when they could have saved us, when they were called to the consular duty by the very ensigns of that command, and to my safety not only by your authority but by your prayers, first deserted us, then betrayed us, lastly attacked us; and by the rewards of an unspeakable bargain wished me, together with the commonwealth, to be utterly crushed and snuffed out. Those punishments which they could not, under their own bloody and fatal command, either keep from the walls of allies or carry into the cities of enemies --- the cutting-down, the burning, the overturning, the depopulation, the wasteland --- these, with their plunder, they brought into all my houses and fields.

\ciceroSection{4} On these Furies and firebrands, on these, I say, ruinous prodigies and almost pestilences of this empire, I declare that I have undertaken an inexpiable war: not as great as my own pain and that of mine demanded, but as great as your pain and the pain of all good men demanded. Toward Clodius, indeed, my hatred is no greater today than it was on the day when I learned that he had been singed by the most religious fires, sent away in woman's dress out of an unholy seduction and out of the house of the pontifex maximus. Then, I say, then I saw, and long before I had foreseen, what storm was being raised, what tempest was hanging over the commonwealth. I saw that wickedness, so monstrous, the audacity so savage of a youth raving and noble and wounded, could not be confined within the bounds of peace: that evil would burst out at some time, if it had gone unpunished, to the destruction of the city. Not much, in fact, was added afterwards to my hatred.

\ciceroSection{5} For he did nothing against me out of hatred of me, but out of hatred of severity, hatred of standing, hatred of the commonwealth: he no more violated me than he violated the Senate, the Roman knights, all good men, all Italy. Nor was he in the end more wickedly disposed against me than against the immortal gods themselves. For he violated them by a wickedness no man before him had committed; toward me he was of the same temper which his bosom-friend Catiline, had he won, would have shown. So I never thought he should be prosecuted by me --- no more than that block of wood, of which we should not know what kind of man it was, did he not himself say he was a Ligurian. Why should I pursue this animal and beast, fattened on the fodder and acorns of my enemies? Who, if he sensed by what wickedness he has bound himself, I do not doubt is the most miserable of men; but if he does not see it, there is danger that he will defend himself by the excuse of stupor.

\ciceroSection{6} It is added too that, by the expectation of all, this victim seems to have been vowed and assigned to that bravest and most distinguished man, T. Annius. To snatch from him a glory already pledged and destined --- when I myself by his work have recovered both my standing and my safety --- would be very unjust. For just as that famous P. Scipio seems to me to have been born for the destruction and ruin of Carthage, who alone overturned by a kind of fated coming that city which had been by many commanders besieged, attacked, shaken, almost taken, so T. Annius seems to have been born for the suppressing, snuffing out, and razing to the ground of this pestilence, and granted as it were by a divine gift to the commonwealth. He alone has understood by what method an armed citizen --- one who with stones, with iron, drives some men to flight, holds others shut up at home; who terrifies the whole city, the Curia, the Forum, all the temples, with slaughter and burnings --- ought not only to be conquered but bound.

\ciceroSection{7} From a man so great, and one who has so deserved of me and of his country, I will never voluntarily snatch this defendant in particular --- whose enmities he has not only undertaken on my behalf but has even sought out. But if even now, ensnared as he is in the dangers of every law, entangled in the hatred of all good men, caught in the now not distant expectation of punishment, he is yet borne forward fumbling and tries, blocked though he is, to make some onset against me, I will resist; and either with Milo's consent or with his help I will repel his attempt: just as yesterday, when standing he silently threatened me, I touched only with my voice the beginning of the law and the trial. He sat down: I fell silent. Had he named the day, as he had hinted, I would have brought it about that the third day from the praetor was named for him at once. And let him so govern himself, and consider this: if he is content with the wickednesses he has committed, he is now consecrated to Milo; if he aims any weapon at me, I shall at once snatch up the arms of trials and laws.

\ciceroSection{8} A little while ago, senators, he held a contio which has been reported to me in full. The whole substance and drift of that contio, hear first; then, when you have laughed at the man's shamelessness, you shall hear from me about the whole. Of religion and the rites and ceremonies Clodius spoke in the contio, senators: Publius, I say, Publius Clodius complained that the rites and religious observances were being neglected, violated, polluted! No wonder it seems ridiculous to you: even his own contio laughed at the man, who, as he himself is wont to boast, has been transfixed by two hundred decrees of the Senate, all passed against him on behalf of religion --- and that man, who has brought defilement onto the cushions of the Good Goddess, who has violated those rites which it is forbidden for the eyes of a man to look on even unaware, not only with a man's gaze but with debauchery and seduction, complains in the contio that religion has been neglected.

\ciceroSection{9} And so now his next contio is awaited on chastity. For what difference does it make whether, having been driven from the most religious altars, he should complain of religion --- or whether, just emerged from his sister's bedroom, he should defend modesty and chastity? He read out in the contio this fresh response of the haruspices about the rumbling, in which, among many other things, this also is written, as you have heard, that ``sacred and religious places are being held profane.'' And he said in this matter that my house had been consecrated by that most religious priest, P. Clodius.

\ciceroSection{10} I am glad that out of this whole portent, which perhaps has been the most grave to be reported to this order in many years, a cause has been given to me of speaking, not only just but even necessary; for you will find that out of this whole prodigy and response we are forewarned of his wickedness and frenzy and of the impending greatest dangers almost by the very voice of Jupiter Best and Greatest.

\ciceroSection{11} But first I will expiate the religion of my house, if I shall be able to do so truly and without doubt on anyone's part; if even the slightest scruple shall seem to remain to anyone, not only with patience but with a willing mind I will obey the portents of the immortal gods and religion. But what house in this city is so empty of and so pure from any such suspicion of religion? Although your houses, senators, and those of the rest of the citizens are for the most part by far free from religion, yet my one house has been freed by all judgements: the only one in this city. For I appeal to you, Lentulus, and to you, Philippus. From this haruspical response the Senate decreed that you should refer to this order on sacred and religious places. Can you bring up my house, which, as I have said, alone in this whole city has been freed by every religion and by every judgement? Whose first enemy himself, in that storm and night of the commonwealth, when he had inscribed the rest of his crimes by that filthy stylus dipped in the mouth of Sex. Clodius, did not touch a single letter of religion; then the same house the Roman people, whose is the highest power in all matters, by the centuriate assembly, by the votes of every age and order, ordered to be of the same right as it had been; afterwards you, senators --- not because the matter was doubtful but so that to this Fury, if he should remain longer in the city he so longed to destroy, the very voice of speech might be denied --- decreed that the religion of my house be referred to the college of the pontiffs.

\ciceroSection{12} What religion is so great that we are not freed from it, in our doubts and in the gravest scruples, by the response and word of a single P. Servilius or M. Lucullus? Concerning public sacrifices, the greatest games, the rites of the household gods and Mother Vesta, that very sacrifice which is performed for the safety of the Roman people, which since the founding of Rome was first violated by the wickedness of this single chaste guardian of religion --- whatever three pontiffs had decided has always seemed to the Roman people, to the Senate, to the immortal gods themselves, holy enough, august enough, religious enough. But on my house, P. Lentulus, consul and pontiff; P. Servilius, M. Lucullus, Q. Metellus, M'. Glabrio, M. Messalla, L. Lentulus, flamen Martialis; P. Galba, Q. Metellus Scipio, C. Fannius, M. Lepidus, L. Claudius rex sacrorum, M. Scaurus, M. Crassus, C. Curio, Sex. Caesar flamen Quirinalis, Q. Cornelius, P. Albinovanus, Q. Terentius, the lesser pontiffs --- having heard the cause, with it pleaded twice, with the greatest assembly of the most distinguished and wisest citizens standing by --- with one mind all freed it from every religious obligation.

\ciceroSection{13} I deny that, since the establishment of the rites --- whose antiquity is the same as the city's --- in any matter whatever, not even concerning the chastity of the Vestal virgins, has so full a college given judgement. Although for the discussion of a crime it matters that as many as possible be present (for that is the rule of pontifical interpretation, since the same men hold the power of judges), the explanation of religion can be rightly done even by one experienced pontiff (a thing which in a capital trial is harsh and unjust); yet you will find this: that more pontiffs have judged on my house than ever judged on the rites of the virgins. The day after, the Senate, in fullest attendance, with you, Lentulus, the consul-elect, leader of the opinion, on the motion of P. Lentulus and Q. Metellus the consuls, when all the pontiffs of this order were present and when others who excelled in honours of the Roman people had spoken many words about the college's verdict, and when all of the same were present at the drawing-up, decreed that my house was seen to be freed from religion by the judgement of the pontiffs.

\ciceroSection{14} On this place, then, sacred above all, do the haruspices seem to be speaking, on the place which alone of all private places has this peculiar right, that by the very men who preside over the rites it has been judged not sacred? But bring up the matter, as you must do under the senatorial decree. Either the inquiry will be granted to you, who first delivered an opinion on this house and freed it from every religious obligation; or the Senate itself will judge, which, with that one priest of the rites alone dissenting, in fullest attendance previously judged; or, what surely will happen, it will be referred to the pontiffs, to whose authority, faith, and prudence our forefathers entrusted both private and public rites and religious matters. What then can they judge other than they have judged? Many are the houses in this city, senators, and I dare say almost all by the best of right, but yet by private right, by hereditary right, by right of authority, by right of mancipium, by right of bond: I deny that any other house exists fortified by the same private right as the one of best title, and at the same time by every public, human, and divine right outstanding;

\ciceroSection{15} since first it is built on the authority of the Senate at public expense, and then it has been fortified and hedged around by so many decrees of the Senate against the wicked violence of this gladiator. The first commission was given last year to those magistrates --- to whom in the gravest dangers the whole commonwealth is wont to be entrusted --- to see that I might be allowed to build without violence; then, when he had brought ruin to my dwelling with stones and fire and steel, the Senate decreed that those who had done it should be held under the law on violence, which is against those who attack the whole commonwealth. And on your motion --- O bravest and best consuls within memory! --- the same Senate, in fullest attendance, decreed that whoever should have violated my house would be acting against the commonwealth.

\ciceroSection{16} I deny that there exist on any public work, on any monument, on any temple, as many decrees of the Senate as on my house, which the Senate has held to be the only one since the city's founding to be built from the treasury, to be freed by the pontiffs, to be defended by the magistrates, to be punished by the courts. To P. Valerius for his great services to the commonwealth a public house was given on the Velia; to me, on the Palatine, restored. To him, the place; to me, even the walls and the roof. To him, what he himself should defend by private right; to me, what every magistrate should defend publicly. Which things, indeed, if I held either of myself or by others, I would not proclaim before you, lest I should seem to glory too much; but since they are given me by you, since they are being attempted by the tongue of him by whose hand they were before overthrown, you who restored me and my children with your own hands --- I am not speaking of my own deeds but of yours; nor do I fear that this proclamation of mine, of your kindnesses, may seem ungrateful rather than arrogant.

\ciceroSection{17} And yet if, after I have undergone such labours for the common safety, in refuting the slanders of wicked men a certain pang of spirit should at any time carry me away to glory, who would not pardon me? For yesterday I saw a certain man muttering, who, they said, was saying that I could not be borne, because, when I was asked by this same most foul parricide what city I was a citizen of, I answered, with you and the Roman knights approving, that I was the citizen of that city which had not been able to do without me. He, I think, groaned. What then was I to answer? --- I ask of him himself who cannot bear me. That I am a Roman citizen? It would have been a literal answer. Or to be silent? It would have been to abandon the matter. Can any man, engaged in great affairs amid envy, answer the insults of an enemy gravely enough without his own praise? But he himself not only answers whatever he can when challenged, but actually rejoices that he is reminded by his friends what he should answer.

\ciceroSection{18} But since my own case is dispatched, let us see now what the haruspices say. For I confess that I have been deeply stirred both by the magnitude of the portent and by the gravity of the response and by the one and consistent voice of the haruspices; nor am I one who, if perhaps I seem to be more than the rest who are equally busy with me engaged in the study of letters, take pleasure in or use at all those letters which would deter our minds and call them away from religion. I have, indeed, our forefathers as authors and masters of the cultivating of religion, whose wisdom seems to me to have been so great that those are wise enough and beyond who can not, I will not say match their wisdom, but at least see how great it was: who held that fixed and solemn rites are contained in the pontificate, the auspices for things well done in augury, the old prophecies of fate in the books of the Apolline seers, and the expiations of portents in the discipline of the Etruscans. Which is so great that within our memory it foretold to us, not obscurely but a little while before, first the dire beginnings of the Italic war; afterwards the almost-extreme crisis of the Sullan and Cinnan period; and then this recent conspiracy to set the city on fire and to destroy the empire.

\ciceroSection{19} Then, if I had any leisure, I have learned besides that many learned and wise men have both said and left in writing many things on the godhead of the immortal gods; which, although I see they are written down divinely, are yet of such a kind that our forefathers seem to have taught these men, not to have learned from them. For who is so out of his wits that, when he has looked up to the sky, does not feel that there are gods, and that those things are formed by such a great mind that scarcely anyone by any art could trace through the order and the necessary connection of things --- thinks all this happens by chance? Or, when he has understood that there are gods, does not understand that this great empire was born and increased and held by their godhead? However much we may, senators, love ourselves, yet not in number do we surpass the Spaniards, nor in strength the Gauls, nor in cleverness the Phoenicians, nor in arts the Greeks, nor finally in that very domestic and inborn sense of this nation and this land the Italians and Latins themselves --- but by piety and religion and by this one wisdom, that we have seen all things to be ruled and steered by the godhead of the gods, we have surpassed all peoples and nations.

\ciceroSection{20} So, that I may not say more on a matter least doubtful, lend your attention, and bring not only your ears but your minds to the haruspices' voice: ``\textit{Because in the Latinian field a noise was heard with a rumbling.}'' I leave aside the haruspices, I leave aside that ancient discipline, handed down to Etruria, as men say, by the immortal gods themselves: cannot we ourselves be haruspices? In the field nearby and suburban there was heard a hidden noise and horrible rumbling of arms. Who is there of those Giants, whom the poets tell us made war on the immortal gods, so impious that he would not confess that by such a new and great motion the gods are foretelling and proclaiming something great to the Roman people? Of the matter it is written: ``\textit{There are postilia owing to Jupiter, Saturn, Neptune, Tellus, the heavenly gods.}''

\ciceroSection{21} I hear by which gods being violated expiation is owed; but for what offences of men I ask. ``\textit{Games less diligently performed and polluted.}'' What games? I appeal to you, Lentulus --- yours is the priesthood, the chariot-cars, the choric prelude, the games, the libations and the banquets of the games --- and to you, pontiffs, to whom the epulones of Jupiter Best and Greatest, if anything has been omitted or committed wrongly, refer the matter; on whose word the same games, renewed and instituted afresh, are celebrated. What games have been performed less diligently? When and by what wickedness have they been polluted? You will answer, both for yourself and for your colleagues, and even for the college of pontiffs, that nothing has been despised by anyone's negligence or polluted by wickedness: that all the solemn and lawful elements of the games, with everything observed, have been kept with the greatest scruple.

\ciceroSection{22} What games, then, do the haruspices declare were performed less diligently and polluted? Those at which the immortal gods themselves, and that Idaean Mother, wished you --- you, Cn. Lentulus, by whose great-great-grandfather's hands she was received --- to be a spectator. And had you not on that day wished to look on the Megalesia, I do not know that we could now be alive and complaining of these matters. For an unnumbered force, stirred up, gathered out of every alley, of slaves, was on a sudden by this religious aedile loosed from arches and doorways and, on the signal given, hurled into the scena, and broke in. Yours then, yours, Cn. Lentulus, was the same courage that was once in your great-grandfather as a private man; you, your name, your imperium, your voice, your gaze, your assault --- the rising Senate and the Roman knights and all good men followed you, when he, that aedile, had handed over the Senate and the Roman people, bound in their seats and constrained at the show, hindered by crowd and narrowness, to the multitude of mocking slaves.

\ciceroSection{23} Or if a stage-dancer halted, or a flute-player suddenly fell silent, or that boy with both parents living failed to hold the chariot-rope, if he let go the strap, or if the aedile blundered in word or libation-cup --- the games are not duly performed, those errors are expiated, and the minds of the immortal gods are appeased by repeating the games. If the games have been turned from joy to fear, if they have not been interrupted but ended and removed, if those days, by the wickedness of him who wished to turn games to mourning, became for the whole state almost funeral days instead of festal --- shall we doubt what games that rumbling proclaims to have been polluted?

\ciceroSection{24} And if we wish to recall those things which have been handed down to us about each god: this Great Mother, whose games were violated, polluted, almost turned to slaughter and to the funeral of the city --- this one, I say, we have received as walking through fields and groves with a certain noise and rumbling. She therefore has shown to you, has shown to the Roman people, both the proofs of his crimes and the signs of dangers. For what shall I say of those games which our forefathers wished to be performed and celebrated on the Palatine, before the temple, in the very sight of the Great Mother, at the Megalensia? Which by use and institution are most chaste, solemn, religious; at which games for the first time the great elder P. Africanus, in his second consulship, gave a place to the Senate before the seats of the people --- that this filthy plague should pollute those games! At which, if any free man came either for the sake of viewing or even for religion, hands were laid on him; at which no matron came for the violence and the press of slaves. So those games --- whose religion is so great that, fetched from the farthest lands, it has settled in this city; the only games which are not even named by a Latin word, that by the very vocabulary the foreign religion sought out and undertaken in the name of the Great Mother is declared --- these games slaves performed, slaves looked on, the whole Megalesia, finally, under this aedile, was a slaves' festival.

\ciceroSection{25} O immortal gods! how could you have spoken to us more clearly, if you were present and dwelt with us? You have signified that the games have been polluted, you say it plainly. What more befouled, more disfigured, more perverse, more thrown into confusion can be said than that all slavery, freed by the magistrate's leave, was sent into one set of seats, set over against another, so that the one set of seats was thrown to the power of slaves, and the other was wholly of slaves? If a swarm of bees had at the games come into the scena or the seats, we should think haruspices ought to be sent for from Etruria; we see suddenly such great swarms of slaves let loose into the Roman people, hemmed in and shut up, and we are not stirred? And in the matter of the bee swarm perhaps the haruspices, from the writings of the Etruscans, would warn us to beware of slavery.

\ciceroSection{26} What we should beware of when signified by some separate and distinct portent, when the thing itself is a portent to itself, and when the danger is in that very thing from which the danger is foretold --- shall we not fear it? Did your father give such Megalesia? Did your uncle? He even makes mention of his lineage, when he has chosen to put on games after the example of Athenion or Spartacus rather than of a C. or Appius Claudius? Those, when they put on games, ordered slaves to leave the seats; you let slaves into one set, threw free men out of the other. And so those who before were separated from the free by the herald's voice, at your games separated free men from themselves not by voice but by hand. Did this not even occur to your mind, the priest of the Sibylline books, that our forefathers sought these rites from your books? --- if those are yours which you with impious mind hunt out, with violated eyes read, with polluted hands handle.

\ciceroSection{27} On the bidding of that prophetess, when Italy long since wearied by the Punic war and harried by Hannibal, our forefathers brought those rites adopted from Phrygia and set them up at Rome; which the man received who was judged the best of the Roman people, P. Scipio, and the woman who was thought the chastest of matrons, Q. Claudia (whose ancient severity in sacrifice your sister is wonderfully thought to imitate). Neither, then, did your forefathers connected with these religions, nor the priesthood itself by which this whole religion is constituted, nor the curule aedileship which is most accustomed to guard this religion, move you not to defile with every shame the most chaste games, to stain them with disgrace, to bind them in wickedness?

\ciceroSection{28} But why do I marvel at this? You who, after taking money, have laid waste Pessinus itself, the seat and dwelling of the Mother of the gods; and to Brogitarus the Galatian, an unclean and accursed man, whose envoys, when you were tribune, used to portion out money in the temple of Castor to your gangs --- to him you sold that whole place and shrine, you tore the priest from those very altars and cushions, you overturned all those things which antiquity, which the Persians, which the Syrians, which all the kings who have held Europe and Asia, have always cultivated with the highest religion; which our forefathers thought so holy that, although we had a city and an Italy crowded with shrines, yet our commanders in the greatest and most dangerous wars made vows to this goddess, and discharged them at Pessinus itself, before that very chief altar, and in that place and shrine.

\ciceroSection{29} And whereas Deiotarus most chastely guarded that shrine in his religious devotion --- he whom we have alone in the whole world most faithful to this empire and most loving of our name --- you handed him over, by his addiction to money for Brogitarus, as I said before. And yet this Deiotarus, often by the Senate judged worthy of the royal name, adorned by the testimonies of the most distinguished commanders --- you bid even to be called king together with Brogitarus. But the one is king by the Senate's verdict, through us; the other is named so by the money of Brogitarus, through you. I will judge the second a king, if he shall have where to pay you what you have lent him by bond. For while many things are royal in Deiotarus, that most of all: that he gave you no penny, that the part of your law which agreed with the Senate's verdict (that he himself be king) he did not refuse, that Pessinus, violated by your wickedness and stripped of priest and rites, he recovered in order to keep it in its ancient religion, that he does not suffer rites received from all antiquity to be polluted by Brogitarus, and that he prefers his own son-in-law to lack the gift you gave rather than that shrine to lack the antiquity of its religion. But to come back to the responses of the haruspices, of which the first is on the games --- who would not confess that the whole has been foretold and answered about that man's games?

\ciceroSection{30} Next, on sacred and religious places. O wonderful impudence! Do you dare speak about my house? Hand over yours either to the consuls or to the Senate or to the college of pontiffs. Mine indeed has been freed by all those three judgements, as I said before; but in those quarters which you, by manifestly killing Q. Seius, that excellent Roman knight, hold, I say there was a chapel and altars. By censorial registers, by the memory of many men, I will establish and demonstrate this: only let the matter be raised --- as it must be raised under the senatorial decree which has lately been passed --- I have what I should wish to say on religious places.

\ciceroSection{31} When I shall have spoken of your house --- in which the chapel has been built in such a way that another did the building, only you have to do the demolishing --- then I will see whether it is necessary for me to say anything about other places. For some think that to me belongs the opening up of the magmentarium of Tellus. They say it lay open recently, and I remember it. Now they say a most holy part and seat of a great religion is shut up in a private vestibule. Many things move me: that the temple of Tellus is in my charge; that the man who removed that magmentarium said in regard to my house, freed by the pontiffs' judgement, that he had been judged in his brother's favour. The religion of Tellus moves me too, in this dearness of grain, this sterility of fields, this scarcity of crops; and the more so because under the same portent a postilio is said to be owed to Tellus.

\ciceroSection{32} Perhaps we are speaking of old matters; though even if this is less written down in civil law, by the law of nature, by the common right of nations, it is sanctioned that mortals can acquire by use nothing from the immortal gods. Yet though we may neglect ancient things --- shall we neglect those that are happening at this very moment, that we see? Who does not know that L. Piso has at this very time pulled down the great and most holy chapel of Diana on the Caeliculus? Neighbours of the place are at hand; many even in this order are present who often performed family sacrifices in that very chapel at the appointed yearly day. And we ask which places the immortal gods miss, what they signify, of what they speak? That by Sex. Serranus the most holy chapels have been undermined, built over, crushed, defiled at last with the utmost shame --- do we not know it?

\ciceroSection{33} You could make my house religious? With what mind? --- the mind you had lost. With what hand? --- the hand with which you had ruined it. With what voice? --- the voice with which you had ordered it set on fire. By what law? --- the law you had not even framed under that impunity of yours. On what cushion? --- the cushion you had defiled. By what statue? --- the statue you had snatched from a courtesan's tomb and set up on a commander's monument. What does my house have of religion, except that it touches the wall of an unclean and sacrilegious man? And so, that no one of mine may unawares look into your house and see you performing those rites of yours, I shall raise the roof higher --- not so that I may look down on you, but so that you may not look on this city which you wished to destroy.

\ciceroSection{34} But let us see now the rest of the haruspices' responses. ``\textit{Envoys killed against right and divine law.}'' What is this? I see the talk is of the Alexandrians; which I do not refute. For I judge that the right of envoys, as it is fenced about with the protection of men, is also walled in by divine law. But I ask of him who as tribune poured out from the prison into the Forum all the witnesses, by whose judgement now all daggers and all poisons are handled, who has made bond with Hermarchus of Chios --- whether he knows that one most bitter adversary of Hermarchus, Theodosius, sent on embassy from a free city to the Senate, was struck by a dagger? Which I am sure has seemed to the immortal gods no less unworthy than the Alexandrians.

\ciceroSection{35} Nor do I now bring everything down on you alone. The hope of safety would be greater if besides you no one were unclean: there are more; both you have more confidence in yourself for this, and we almost in justice have less. Who does not know that Plator from Orestis, that part of Macedonia which is free, a man distinguished and noble in those parts, came as envoy to Thessalonica, to our (as he himself called himself) imperator? Whom he, on account of money which he could not extort from him, threw into chains, and sent in his own physician, who in the foulest and most cruel way cut the veins of an envoy, ally, friend, free man. He would not have his lictors' axes bloodied by the wickedness: but he stained the name of the Roman people with such a wickedness as can be expiated by no thing but his own punishment. What sort of executioners must we suppose this man to have, who uses even his own physicians not for safety but for slaughter?

\ciceroSection{36} But let us read out what comes next. ``\textit{Faith and oath neglected.}'' What this is in itself I do not easily interpret, but from what follows I suspect it is said about the manifest perjury of your jurors, from whom the money would once have been snatched away if they had not asked for protection from the Senate. The reason why I should suspect it is said about them is this: that I judge both that perjury to be most signal and notable in this state, and you yourself nevertheless are not called into the charge of perjury by those with whom you swore the conspiracy.

\ciceroSection{37} And I see in the haruspices' response this added: ``\textit{Ancient and secret sacrifices less diligently performed and polluted.}'' Are these the haruspices speaking, or our fathers' household gods? Many, I dare say, are the men on whom the suspicion of this misdeed could fall. Who besides this one man? Is it obscurely said which rites have been polluted? What can be said more plainly, what more religiously, what more gravely? ``\textit{Ancient and secret.}'' I deny that any words Lentulus, that grave and eloquent orator, used more often, when he was prosecuting you, than these now uttered out of the Etruscan books, turned and interpreted against you. For what sacrifice is so ancient as this which we have received from kings, of the same age as this city? What so secret as that which not only excludes curious eyes but even wandering ones, into which not only impropriety but not even ignorance can enter? Which sacrifice no one before P. Clodius in all memory has violated; no one ever attended; no one neglected; no man without horror beheld --- which is performed by the Vestal Virgins, performed for the Roman people, performed in that house which is in command, performed with incredible ceremony, performed for that goddess whose very name it is unlawful for men to know; whom he calls the Good Goddess on this account, that she has forgiven him such a crime. She has not forgiven you, believe me, no: unless perhaps you think it is forgiveness when the jurors let you out shaken and emptied, by their judgement absolved, by all condemned --- or that you have not, as the opinion of that religion has it, lost your eyes.

\ciceroSection{38} Who indeed before you saw those rites as a man knowingly, that anyone could know what penalty would follow that wickedness? Or would the blindness of your eyes hurt you more than that of your lust? Do you not feel even this --- that those eyelids of your great-great-grandfather, closed in his blindness, were rather to be wished than these of your sister blazing? But if you give close attention, you will understand that till now the punishments of men, not of the gods, are wanting. Men have defended you in a most foul affair; men have praised you, the most shameful and most guilty; men have freed you, almost confessing, by their judgement; for men the wrong of your defilement, brought on themselves, has not been pain; men have given you arms --- some against me, others later against that unconquered citizen; the kindnesses of men I now grant you should look for no greater:

\ciceroSection{39} but from the immortal gods what punishment can be greater for a man than frenzy and madness? Unless perhaps you think those whom you see in tragedies tortured and consumed by wound and pain of body undergo graver wraths of the immortal gods than those who are brought on stage out of their wits. Not so wretched are those wails and groans of Philoctetes, although they are bitter, as that exultation of Athamas, and the old age of the matricides. You, when you let out wild voices in the contiones, when you overturn the houses of citizens, when you drive the best of men from the Forum with stones, when you fling burning torches onto the roofs of neighbours, when you set sacred buildings on fire, when you stir up slaves, when you throw rites and games into confusion, when you do not distinguish wife and sister, when you do not feel whose bed you are entering --- then you rave like a Bacchant, then you go mad, then you pay those penalties which alone have been appointed by the immortal gods for the wickedness of men. For the weakness of our body undergoes many chances of itself; the body itself is often killed off by the slightest cause: the gods' weapons are fixed in the minds of the impious. So you are the more miserable when you are dragged into every fraud with eyes open than if you had no eyes at all.

\ciceroSection{40} But since enough has been said about all the things which the haruspices say have been committed, let us see what those same haruspices say is being warned by the immortal gods. They warn ``\textit{lest by the discord and dissension of the leading men, slaughter and dangers be created for the senators and the chief men, and they be deserted by the help of the divine power; whereby money and an army may pass to one rule, and a beating-back and diminution may follow.}'' These are all the haruspices' words: I add nothing of my own. Who, then, is bringing about discord among the optimates? That same man --- not by any force of wit or counsel of his own, but by a kind of error of ours; which indeed he, since it was not obscure, easily perceived. For the commonwealth is shaken the more shamefully on this account: that not even by him is it harassed in such a way that, like a brave man in the battle who has received wounds in front from a brave adversary, it can seem to fall honourably.

