% De Provinciis Consularibus (de-provinciis-consularibus) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} If any of you, senators, is waiting to hear which provinces I will move that we assign, let him consider with himself which men in particular ought to be pulled out of the provinces by me: he will not be in doubt as to what is fitting for me to feel, when he has thought what I am of necessity bound to feel. And if I were the first to deliver this opinion, you would surely praise it; if I were the only one, you would certainly forgive it; even if the opinion seemed to you a little less useful, you would yet grant some indulgence to my pain. As it is, senators, I am affected by no slight pleasure, both because what most of all serves the interest of the commonwealth --- that Syria and Macedonia be assigned --- means that my pain is not at variance with the common good, and because I have as my model P. Servilius, who delivered his opinion before me, a most distinguished man, of singular loyalty and goodwill both toward the commonwealth as a whole and toward my own restoration.

\ciceroSection{2} If he, both a little while ago and as often as occasion and opportunity for speaking has fallen to him, judged that Gabinius and Piso --- two prodigies of the commonwealth, all but two corpses of it --- ought to be branded, both for their other crimes and most of all for that signal wickedness and unprovoked cruelty against me, not only by his vote but by the gravity of his words, with what feeling toward those men ought I to be, whose safety they handed over as a pledge to fill out the measure of their own appetites? Yet in delivering this opinion I shall not obey my pain; I shall not be the slave of my anger. I will hold toward them the feeling each one of you ought to hold: that special and proper sense of pain which is mine --- though you have always counted it as shared with me --- I will set aside in giving my opinion, and will keep for the season of vengeance.

\ciceroSection{3} There are four provinces, senators, on which I understand that opinions have so far been delivered: the two Gauls, which we now see joined under one command; and Syria and Macedonia, which those two pestilent consuls, against your will and over your suppressed protest, seized as the rewards of an upended commonwealth. Two are to be assigned by us under the lex Sempronia. What ground for hesitation can there be about Syria and Macedonia? I pass over that those who now hold them gained them on these terms: they had not so much as touched them before they had condemned this order, banished your authority from the state, and most foully and most cruelly outraged the public faith, the perpetual safety of the Roman people, and me and all who belong to me.

\ciceroSection{4} I pass over all that happened at home and in the city --- so vast that Hannibal himself never wished this city as much harm as those men inflicted on it. I come to the provinces themselves. Of which Macedonia, formerly fortified by the works not so much of towers as of the trophies of many commanders, long since pacified by many victories and triumphs, is now so harried by barbarians --- to whom, because of his greed, peace has been snatched away --- that the Thessalonians, set in the very lap of our empire, are forced to abandon their town and fortify their citadel; that our great military road, which runs through Macedonia all the way to the Hellespont, is not only beset by raids of the barbarians but actually marked off, like a frontier, by Thracian camps. Thus those peoples who, to keep the peace, had paid in silver to our distinguished commander --- so that he could refill his emptied house --- have, for the peace they bought, brought a war upon us almost like a just war.

\ciceroSection{5} As for our army, that army gathered by an arrogant levy and the harshest conscription, it has perished to a man. I say this with great pain: the soldiers of the Roman people have been pitiably captured, killed, abandoned, scattered, consumed by neglect, hunger, disease, and devastation, so that --- and this is most outrageous --- the commander's wickedness seems to have been expiated upon his country and upon his army. And this Macedonia, with the neighbouring peoples now subdued and the barbarian frontier held down, was peaceful in itself and quiet, and we used to guard it with a slight garrison and a small force --- without even an imperium, by legates, by the mere name of the Roman people. Now, ravaged by a consular army and a consular command, it can scarcely recover itself by long peace. And meanwhile which of you has not heard, who is unaware, that the Achaeans pay a vast yearly tribute to L. Piso; that the entire customs and harbour-dues of the Dyrrachians have been turned to this one man's profit; that the city of the Byzantines, most loyal to you and to this empire, has been ravaged as if it were an enemy? After he could squeeze nothing out of beggars, could wring nothing by any violence from the wretched, he sent his cohorts into winter-quarters; over them he set those whom he judged would be the most diligent attendants of his crimes, the ministers of his appetites.

\ciceroSection{6} I leave aside the administration of justice, in a free state, against the laws and decrees of the Senate; I let go his butcheries; I pass over his lusts, of which the most bitter testimony stands as a signal monument of shame and almost as ground for a just hatred of our empire: it is established that maidens of the noblest birth flung themselves into wells, putting themselves to a voluntary death to escape an unavoidable disgrace. I pass over these things --- not because they are not most grave, but because at the moment I speak without witness. As for the city of the Byzantines itself, who is unaware that it had been most full and most adorned with statues? Statues which the Byzantines, drained as they were by the costs of the greatest wars --- when they were holding back, smashing, choking off at their very throats every onslaught of Mithridates, the whole armed force of Pontus boiling over and bursting toward Asia --- which they then, I say, and afterwards too, kept under the most scrupulous guard, those statues and the rest of the city's adornments;

\ciceroSection{7} but under your most ill-starred and most monstrous commander, Caesoninus Calventius, that free city, freed by the Senate and the Roman people for its outstanding services, has been so plundered and stripped that, had not C. Vergilius the legate, a brave and innocent man, intervened, the Byzantines would have had not one statue out of that vast number left them. What temple in Achaea, what place or grove in all Greece, was so holy that any image, any ornament has remained in it? You bought from a most foul tribune of the plebs --- in that shipwreck of the city which you, who were bound to be a pilot, had yourself overturned --- you bought, I say, at a great price, the right to give judgement on debts in free communities, against the decrees of the Senate and against the law of your son-in-law: and what you bought you sold on these terms: either you would not give judgement, or you would overturn the property of Roman citizens.

\ciceroSection{8} Of all this, senators, I say nothing now against the man himself: I am speaking about the province. So I pass over all those things you have often heard and hold fast in your minds, even if you should not be hearing them now. Nothing do I say about that effrontery of his within the city, which by his very presence he has fixed in your minds and before your eyes; nothing do I argue about his insolence, nothing about his contumacy, nothing about his cruelty; let those dark lusts of his lie hidden, which he covered up by brow and forehead, not by modesty and self-control: I argue only what is at issue --- the province. Will you not put him down? Will you suffer him to remain longer? Whose fortunes, the moment he set foot in the province, contended so with his depravity that no one could decide whether he was the more wanton or the more luckless.

\ciceroSection{9} Or is that Semiramis to be kept any longer in Syria? Whose journey into the province was such that king Ariobarzanes hired your consul to commit murder, as one might hire some Thracian; and on entering Syria his cavalry first underwent destruction, and then his best cohorts were cut to pieces. Accordingly, under that commander nothing else has ever been done in Syria but compacts of money with petty despots, settlements, plunderings, brigandage, slaughter --- the commander of the Roman people standing openly before his army drawn up, stretching out his right hand, not exhorting his soldiers to glory but shouting that everything was bought and must be bought by him.

\ciceroSection{10} As for the unhappy publicans --- and I am unhappy too, I say, in the unhappiness and pain of men who deserved so well of me --- he has handed them over into slavery to Jews and Syrians, peoples born to slavery. He decided from the start, and he has held to it, that he would deliver no judgement to a publican; he rescinded compacts made without any wrong; he abolished the watches; he freed many tributaries and stipendiaries; he forbade any publican or publican's slave to be in whatever town he was in or wherever he came. What more is there? He would be thought cruel, had he held toward the enemy the temper that he held toward Roman citizens --- and especially toward that order which has always been kept up, in keeping with its dignity, by the kindness of the magistrates.

\ciceroSection{11} And so, senators, you see that not by any rashness in their contracting nor by ignorance in carrying out their business, but by the greed, the arrogance, the cruelty of Gabinius, the publicans have been now almost ruined and overthrown. Whom indeed, in these straits of the treasury, you must yet rescue --- though many of them you can no longer help, men who, on account of that public enemy of the Senate, that bitter foe of the equestrian order and of all good men, have lost their possessions and their good name as well; whom no thrift, no self-restraint, no virtue, no labour, no distinction has been able to defend against the audacity of that glutton and brigand.

\ciceroSection{12} What of those who even now sustain themselves by the props of patrimony or the generosity of friends --- shall we suffer them to perish? If a man cannot enjoy his public contract because of an enemy, the censorial law itself protects him: shall the man whom this enemy will not suffer to enjoy it --- a man who, even if he is not so styled, is an enemy --- shall help not be brought to him? Keep then in the province longer the man who makes compacts with the enemy at the cost of the allies, and with the allies at the cost of the citizens; who reckons himself worth more than his colleague on this score: that the other deceived you by gloom and by face, while he himself never pretended to be less worthless than he was. Piso, on the other hand, glories on a different ground: that in a short time he has brought it to pass that C. Gabinius is not thought, alone of all men, the most worthless.

\ciceroSection{13} Were these men, then, even if they were not at some point to be brought home from their provinces --- would you not think they ought to be torn out? And would you keep these twin pestilences --- this disaster to allies, to soldiers, this ruin of publicans, this devastation of provinces, this stain on our empire? Yet you yourselves, in the previous year, recalled these very same men once they had reached their provinces. At that time, had your judgement been free, and had the matter not been put off so often and at last torn from your hands, you would have restored what you wanted --- your own authority --- by recalling those through whom it had been lost, and by stripping from them the very rewards which they had earned by their crime and the overthrow of their country.

\ciceroSection{14} If from that punishment they then escaped --- by the resources of others, not their own, against your most strenuous will --- yet they have undergone another, much greater and graver, punishment. For what punishment can fall more grievous upon a man who has, if not any shame for fame, at least some fear of legal sanction, than not to be believed, when he sends letters reporting on the commonwealth that the war has been well conducted? This the Senate decided when in full session it denied a thanksgiving to Gabinius: first, that nothing was to be believed by a man stained by every wickedness and outrage; next, that by a traitor --- and one whom they had known to be while present an open enemy of the commonwealth --- the commonwealth could not have been well managed; and last, that not even the immortal gods would wish their temples to be opened, and supplications offered to them, in the name of so foul and so wicked a man. So that other one is either himself a learned man and rather subtly schooled by his Greeks (in whose company he now gluttonises on the open stage, where formerly he used to do it behind the curtain), or he has friends with more sense than Gabinius has --- whose letters are nowhere produced.

\ciceroSection{15} Are these, then, the men we are to keep as our commanders? Of whom one does not dare to inform us why he is being saluted as imperator, and the other --- if his couriers do not slacken --- must in a few days repent his daring. His friends, if he has any --- if any can be friends to a beast so monstrous and so foul --- console themselves with this: that the Senate refused a thanksgiving to T. Albucius too. The first thing unlike here is that one was a campaign in Sardinia against bandits in sheepskins, conducted by a propraetor with a single auxiliary cohort; the other a war finished off against the greatest peoples and tyrants of Syria, with a consular army and a consular command. Then again, Albucius had decreed beforehand for himself, in Sardinia, the very thing he was asking from the Senate: it was common knowledge that the Greekling and frivolous fellow had as good as held a triumph in his own province, and the Senate, by refusing the thanksgiving, set a mark upon his rashness.

\ciceroSection{16} But let him take his comfort, and let him think this signal disgrace --- since it has been branded upon only one man besides himself --- a lighter thing, provided that he wait for the same end as the man whose example he uses for consolation: especially since in Albucius there was neither the lust of Piso nor the audacity of Gabinius, and yet by this single blow he fell --- the Senate's mark of shame.

\ciceroSection{17} And yet whoever assigns the two Gauls to the two consuls keeps both of these men; and whoever assigns one of the Gauls and either Syria or Macedonia still keeps one of them, and makes a different lot in cases of equal wickedness. ``I shall make those provinces praetorian,'' he says, ``so that Piso and Gabinius shall be succeeded at once.'' --- if he allows it! For then the tribune will be able to interpose; now he cannot. So I, the very same man who am now assigning Syria and Macedonia to those who shall have been designated consuls, I shall assign these same provinces as praetorian provinces too --- so that the praetors may have annual provinces, and we may see, as soon as may be, those whom we cannot bear to look upon. But, believe me, succession will never be made to those men, unless the question is brought up under the law that does not allow interposition on provinces. So this opportunity having been let slip, an entire year must be waited out by you; and meanwhile the calamity of the citizens, the misery of the allies, and the impunity of the most wicked of men is prolonged.

\ciceroSection{18} But suppose those men were the best of men --- still, in my opinion, I would not yet hold that C. Caesar must be succeeded. On this point I will say what I think, senators; and I will not fear that interruption from my closest friend, by which a little while ago my speech was broken off. He says, that excellent man, that I ought not to be more hostile to Gabinius than to Caesar; that the whole storm to which I yielded was raised by Caesar as instigator and abettor. If first I should answer him this --- that I take account of the common good, not of my own pain --- can I not make my case, when I claim to be doing what I can do by the example of the bravest and most distinguished citizens? Did not Tiberius Gracchus --- the father, I mean, whose sons would they had not fallen short of their father's gravity! --- attain such great praise on this score: that as tribune of the plebs, alone of all that college, he came to the aid of L. Scipio, the bitterest enemy both of him and of his brother Africanus; and swore in the contio that he had not made up the quarrel, but that it seemed to him beneath the dignity of the empire that, where the captive enemy generals had been led in Scipio's triumph, that Scipio should himself now be led to the same place who had led the triumph?

\ciceroSection{19} Who had more enemies than C. Marius? L. Crassus, M. Scaurus were strangers to him, all the Metelli his enemies: yet they not only did not call him --- their enemy --- back from Gaul by their votes, but on account of the war in Gaul they assigned to him an extraordinary province. The greatest war in Gaul has been fought; the greatest peoples have been subdued by Caesar, but not yet bound fast by laws, by sure jurisdiction, by a sufficiently firm peace. We see the war advanced and, to speak the truth, almost finished --- but on these terms: that, if the same man who began it pursues it to the end, we shall see all completed; if he is succeeded, there is danger that we may hear of the remnants of that great war stirred up afresh and renewed. Therefore I --- a senator, hostile, if you wish, to the man --- ought to be a friend, as I have always been, to the commonwealth.

\ciceroSection{20} What if I lay aside my own enmities for the sake of the commonwealth: who, in justice, will reproach me? Especially when I have always thought that the patterns of all my counsels and acts ought to be sought from the deeds of the greatest men. Was not that famous M. Lepidus, who was twice consul and pontifex maximus, praised --- not only by the testimony of memory but by the letters of the annals and the voice of the highest poet --- because, on the day he was made censor, with M. Fulvius his colleague, his bitterest enemy, he was reconciled to him on the spot in the Campus, that they might defend the common duty of the censorship by a common mind and will?

\ciceroSection{21} And, to leave aside antiquity --- which is beyond counting --- did not your own father, Philippus, at one time make peace with his bitterest enemies? With all of whom the same commonwealth reconciled him as had estranged him.

\ciceroSection{22} I pass over many examples, because I look on these lights and ornaments of our state, P. Servilius and M. Lucullus. Would that L. Lucullus too were sitting yonder! What feuds in this city were graver than the feuds between the Luculli and Servilius? Which feuds, in those bravest of men, the interest of the commonwealth and their own dignity not only quenched but actually carried over into friendship and the habit of intimacy. What of Q. Metellus Nepos --- did he not, when consul, in the temple of Jupiter Best and Greatest, moved both by your authority and by the incredible weight of P. Servilius's eloquence, in his very absence and by his own great kindness, make peace with me? Can I, then, be hostile to a man whose letters and fame and dispatches have my ears every day full of the new names of nations, peoples, places?

\ciceroSection{23} I burn, believe me, senators --- as you yourselves think and do for me --- with a kind of incredible love of country: which love of country once compelled me to come to its rescue against the gravest dangers, with a struggle that put my life at stake; and again, when I saw all weapons aimed from every side at our country, to face them and take the blow alone, on behalf of all. This my old and unbroken feeling toward the commonwealth brings me back, reconciles me, restores me to friendship with C. Caesar.

\ciceroSection{24} In the end let men think what they wish: I cannot be other than friend to anyone who has deserved well of the commonwealth. For if I declared not enmity but war on those who wished to destroy all this with flame and sword --- when some of them were my close friends, and some had even been freed from capital trials by my own defending --- why should the same commonwealth which could inflame me against friends not be able to placate me toward enemies? What hatred had I against P. Clodius, except that I judged the man would be a public danger to his country who, fired by the most shameful lust, had violated by one crime the two most sacred things, religion and chastity? Is there any doubt then, from the things he has done and does daily, that in attacking him I had more regard for the commonwealth than for my own peace --- while certain men, in defending him, had more regard for their own peace than for the common good?

\ciceroSection{25} I confess that I differed from C. Caesar in matters of state and felt with you; but now I assent to the same men with whom I formerly felt. For you yourselves, to whom L. Piso does not dare to send letters about his own doings, you who condemned the letters of Gabinius with a notable mark and a new disgrace, have decreed thanksgivings to C. Caesar in number such as no one ever received from one war, and with honour such as no one ever received at all. Why then should I look for some person to bring me back into friendship with him? The most august order has brought me back, and that order which is both the source and chief mover of public counsel and of all my own counsels. I follow you, senators; I obey you, I assent to you; you who, so long as you did not particularly approve C. Caesar's policies in the commonwealth, saw me also less closely linked with him; afterwards, when by his exploits you changed your minds and your wills, you saw me not only the companion of your opinion but its eulogist.

\ciceroSection{26} But what is there in this case that men should especially marvel at and reprove --- my own counsel, when I myself have previously decreed many things which had to do more with the standing of the man than with the necessity of the commonwealth? I voted by my own motion a thanksgiving of fifteen days. For the commonwealth as many days as for C. Marius were enough; for the immortal gods that same gratitude was not too small a measure as for the greatest wars; therefore the surplus of days was a tribute paid to the man's standing.

\ciceroSection{27} On which point I --- who as consul first proposed a thanksgiving of ten days when Mithridates was killed and the Mithridatic war finished by Cn. Pompey, and on whose motion a consular thanksgiving was first doubled (for you assented to me when, after the letters of the same Pompey were read out, you decreed a ten-day thanksgiving for the completion of all wars by sea and by land) --- I, I say, marvelled at the courage and greatness of soul of Cn. Pompey, in that, although he had himself been preferred above all others in every honour, he was assigning a fuller honour to another than he himself had obtained. Therefore in that thanksgiving which I proposed, the substance was paid to the immortal gods and to the institutions of our forefathers and to the commonwealth's interest, but the dignity of the words, the honour, the novelty, and the number of days were granted to Caesar's own praise and glory.

\ciceroSection{28} The pay of the army was lately referred to us: I not only voted for it but worked to bring you to vote for it; I answered many objectors; I was present at the drawing-up. There too I attributed more to the man than to any unspecified necessity. For I judged that even without this support of money he could keep the army by the booty already won, and finish the war; but I did not think that the lustre and adornment of his triumph should be diminished by our parsimony. The matter of the ten legates was raised --- whom some refused outright; some asked for precedents; some put off the day; some gave them without any ornament of words. There too I spoke so that all should understand I was doing, more amply, what I felt for the commonwealth's sake, on account of Caesar's own dignity.

\ciceroSection{29} But that same I, in now assigning the provinces --- I who handled all those earlier matters in silence --- am being interrupted, when in those previous causes the man's standing was a help to me, but in this case nothing else moves me but the calculation of the war, the supreme advantage of the commonwealth. For Caesar himself: what reason can he have for wishing to linger in his province, except so that what has been begun by him may be handed over completed to the commonwealth? The pleasantness of the regions, I suppose, the beauty of the cities, the courtesy and refinement of those peoples and nations, the desire of victory, the pushing-out of the empire's borders --- these things are keeping him there. What can be found rougher than those lands, more uncultivated than those towns, more savage than those peoples, more notable than so many victories, more remote than the Ocean? Or has his return to the country some occasion for offence? With the people, who sent him? Or with the Senate, by whom he was honoured? Does the day add to longing for him, or rather to forgetfulness; and does that laurel won by great dangers, after a long interval, lose its freshness? Therefore, if any do not love the man, there is no reason for them to call him back from the province: they call him back to glory, to a triumph, to congratulations, to the highest honour of the Senate, to the goodwill of the equestrian order, to the love of the people.

\ciceroSection{30} But if he, on account of the commonwealth's interest, does not hasten to enjoy this most signal good fortune --- in order to bring all those things to completion --- what ought I, a senator, to do, who, even if he should wish otherwise, would still have to take counsel for the commonwealth? I myself understand it thus, senators: that at this time, in assigning the provinces, we ought to take account of perpetual peace. For who does not feel that everything else is free for us from all peril and even from the suspicion of war?

\ciceroSection{31} Long since we see that vast sea, by whose seething not only sea-routes but cities and military roads were held in check, kept secure for the Roman people by the courage of Cn. Pompey from the Ocean to the farthest Pontus, as if it were one safe and closed harbour; those nations which by their numbers and very mass were able to overflow into our provinces have, by the same man, been partly cut down, partly held back, so that Asia, which used to mark the limit of our empire, is now itself ringed by three new provinces. I can speak of every region, of every kind of enemy: there is no people which has not either been so reduced as scarcely to exist, or so subdued as to keep quiet, or so pacified as to rejoice in our victory and our empire.

\ciceroSection{32} The Gallic war, senators, has been fought, with C. Caesar as commander; before, it had only been beaten back. Always our commanders thought those nations were rather to be repelled in war than provoked. Even the great C. Marius himself, whose godlike and outstanding courage came to the rescue when the Roman people had been struck by great mournings and slaughters, drove back the vast forces of the Gauls flowing into Italy: he did not himself penetrate to their cities and seats. Lately my own companion in labours, dangers, and counsels, C. Pomptinus, that bravest of men, broke in battle the war of the Allobroges, suddenly risen and stirred up by the wicked conspiracy, and tamed those who had provoked him; and content with that victory --- with the commonwealth freed from fear --- he kept quiet. C. Caesar's reckoning, I see, has been a very different one. He judged that he had to make war not only with those whom he saw already in arms against the Roman people, but that all of Gaul had to be brought under our sway.

\ciceroSection{33} And so he has fought most successfully, in the greatest battles, against the fiercest peoples, the Germans and the Helvetii; the rest he has terrified, beaten down, tamed, accustomed to obey the empire of the Roman people; and regions and peoples that no records, no voice, no fame had made known to us before, our commander, our army, the arms of the Roman people have traversed. Before, senators, we held but a path through Gaul; the rest of the parts were held by nations either hostile to our empire, or untrustworthy, or unknown, or at any rate savage and barbarous and warlike --- nations which there has never been a man who did not desire to see broken and tamed. No one ever thought wisely about our state, from the very beginning of this empire, who did not judge that Gaul was most of all to be feared by it; but on account of the violence and the multitude of those peoples, they had never before been fought with all together. We resisted always when provoked: at last, now, it has been brought about that the limit of our empire and the limit of those lands is one and the same.

\ciceroSection{34} The Alps had fortified Italy formerly, by nature, not without some divine help; for if that approach had lain open to the savagery and multitude of the Gauls, never would this city have offered a home and seat to the highest empire. Now they may sink as they please! For there is nothing beyond that height of mountains, all the way to the Ocean, that Italy need fear. Yet a single summer, or two at most --- by fear or hope, by punishment or rewards, by arms or laws --- can bind all Gaul fast in eternal chains: but if the country is left half-finished and bitter, although now cut down it will at some point rise up and break out again into renewed war.

\ciceroSection{35} Therefore let Gaul be in the keeping of the man to whose loyalty and courage and good fortune it has been entrusted. If he, adorned with the most splendid gifts of Fortune, were unwilling to put that goddess to the test more often than he must; if he were hastening to his fatherland, to the household gods, to the standing he sees set out for him in the state, to his most beloved children, to his most distinguished son-in-law; if he were eager to ride into the Capitol the victor with that signal laurel; if, in short, he feared some chance, which now can no longer add as much as it might take away --- yet we ourselves should still wish that all those things be brought to completion by the same man by whom they were begun. But seeing that he has long since done enough for his own glory, but not yet enough for the commonwealth, and prefers to come later to the fruits of his own labours rather than not to fulfil the task he has undertaken on behalf of the state --- we ought neither to call back a commander on fire to manage the commonwealth's business well, nor to throw into confusion and to hinder the whole plan of the Gallic war, now almost laid out.

\ciceroSection{36} For those opinions of distinguished men are by no means to be approved, of which the one decrees Further Gaul along with Syria, the other Hither Gaul. He who decrees the Further throws into confusion all the things I have just been discussing; he shows at the same time that he holds for law that which he denies to be law; and the part of the province on which interposition is not possible, this he tears away --- the part that has a defender, he leaves alone. And he so manages it that what was given by the people he does not violate; what the Senate gave, he as a senator hastens to take away. The other has regard for the plan of the Gallic war, and discharges the office of a good senator: a law which he does not regard as such, even that he keeps; for he sets a date in advance for the successor. And yet nothing seems to me more alien from the dignity and the discipline of our forefathers than that he who, on the Kalends of January, ought as consul already to have his province, should appear to have it as a thing betrothed but not yet decreed.

\ciceroSection{37} Suppose someone goes through his whole consulship without a province for whom no province had been decreed before he was elected. Will he draw lots, or not? For both not to draw lots is absurd, and not to have the thing one drew is absurd. Will he set out in the general's cloak? Whither? Where it will not be lawful for him to come before a fixed day. In January, in February he will not have a province: at last on the Kalends of March a province will suddenly be born to him.

\ciceroSection{38} And yet under these motions Piso will remain in his province. Which things, if they are grave in themselves, this is graver still: that to fine a commander by reduction of his province is an insult, and care must be taken that this happen not only to the highest of men but even to a moderate one. I understand, senators, that you have decreed many distinguished honours and almost unique ones to C. Caesar. If because he had so deserved, you were grateful; but if also that he might be most closely linked with this order, you were wise and far-seeing. This order has never embraced anyone with its honours and benefits who has thought any standing more outstanding than that which he had attained through you. No one has ever been able to be a leading man here who has preferred to be a popular man. But men who have either, on account of their unworthiness, distrusted themselves; or, on account of the disparagement of others, been driven from association with this order, have often been forced from this harbour into those waves almost of necessity. If, after that tossing and that popular course, they bring back a public matter well managed and turn their gaze again to the Senate-house and wish to be received by this most august body, they ought not only not to be turned away but actually to be sought after.

\ciceroSection{39} We are warned by the bravest man and best consul within memory to take care that the Hither Gaul not be assigned, against our will, to someone after the men who shall now have been designated consuls; and that thereafter forever it be held in popular and turbulent fashion by those who attack this order. This blow, although I do not despise it, senators --- especially when warned of it by the wisest of consuls and the most diligent guardian of peace and quiet --- yet I judge that it is more vehemently to be feared if I shall lessen the honour of most distinguished and most powerful men or shall reject their zeal toward this order. For that C. Iulius, having been adorned by the Senate with all signal or unprecedented marks of honour, should hand over this province from his own hand to one whom you would least wish; should leave not even freedom to that order through which he himself has won the most splendid glory --- to bring myself to suspect this is in no way possible. To begin with, what feeling each man will have, I do not know: what I have to hope, I see. This I, as senator, ought to do, so far as I can: that no distinguished or powerful man may seem to have just cause to be angry with this order.

\ciceroSection{40} And these things, even if I were Caesar's bitterest enemy, I should still feel for the sake of the commonwealth. But I think it not out of place --- so that I may be less often interrupted by some, or less often reproached by the silent thoughts of others --- to set out briefly what is the reason and basis of my dealings with Caesar. And first, the time of intimacy and habitual friendship which from the youth of all of us has held between him and me, and my brother, and our cousin C. Varro, I pass over. After I had entered fully into public life, I dissented from him on these terms: that, in the parting of opinion, we still remained joined in friendship.

