% Pro M. Caelio (pro-caelio) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} If anyone, members of the jury, were now perchance present, ignorant of our laws, our courts, our custom, he would surely wonder what so great atrocity there is in this case, that on festive days and during the public games, with every other forensic business broken off, this one trial alone is being held --- and he would not doubt that the defendant is charged with so great a crime that, if it were neglected, the state could not stand. The same man, when he heard that there is a law which orders daily inquiry into seditious and ruined citizens who in arms have besieged the senate, have brought violence on magistrates, have made war on the commonwealth, would not condemn the law, but would ask what crime was being moved in the trial. When he heard that no crime, no audacity, no violence is being called to trial, but that a young man of brilliant talent, industry, and popularity is being prosecuted by the son of a man whom he himself is bringing, and has brought, into court, while the resources of a courtesan are being used to assault him --- he would not censure the dutifulness of Atratinus himself, would judge that a woman's lust must be checked, and would think you, gentlemen, hard-driven, since not even in the common leisure of the city are you allowed to be at leisure.

\ciceroSection{2} For if you are willing to attend carefully and to judge truly of this entire case, you will conclude, members of the jury, that no one would have come down to this prosecution to whom it had been left to choose, nor, when he had come down, would have had any hope, unless he were leaning on someone's intolerable lust and excessively bitter hatred. But Atratinus, a most cultured and excellent young man, my close connection, I forgive, since he has the excuse either of dutifulness, or of necessity, or of his age. If he wished to prosecute, I attribute it to dutifulness; if he was ordered, to necessity; if he hoped for something, to youth. To the rest, not only is nothing to be forgiven, but they are to be sharply resisted.

\ciceroSection{3} And to me, members of the jury, this seems the entrance most suited to the defence of the youth of M. Caelius: that I should first reply to the things that the prosecutors said for the disfiguring of this case and the dragging down and stripping of his standing. His father has been thrown at him in various ways --- said either to be too little distinguished himself, or to have been treated too little dutifully by his son. As to standing: M. Caelius's father easily and silently answers for himself, even without my speech, before those who know him and his elders. To those, however, who, on account of his age (since he has now for some time spent less time in the Forum and with us), have not got an equal acquaintance with him, let them hold this: that whatever standing there can be in a Roman knight --- and that can certainly be the very greatest --- has always been held the highest in M. Caelius, and is held so today, not only by his own circle but by everyone to whom for any cause he could be known.

\ciceroSection{4} Moreover, that being the son of a Roman knight should be set down as a charge by prosecutors --- this neither these men ought to have allowed in their pleading, nor we in our defence. As to what you said about his dutifulness, this estimate is indeed ours, but the judgement is certainly the parent's. What we hold, you will hear from those under oath; what the parents feel, the tears of his mother and her unbelievable grief, the squalor of his father, and this present sorrow that you see, and his mourning, declare.

\ciceroSection{5} As to the charge that the young man is not approved by his fellow townsmen --- the people of \dots\ have never given greater honours to anyone present, members of the jury, than to M. Caelius in his absence: him in his absence they co-opted into the most ample order, and they conferred on him, when he was not even seeking them, things which they had refused to many seekers. The same men have now sent the most select men of our order and Roman knights with a delegation to this trial and with the most weighty and ornate testimonial. I seem to myself to have laid the foundations of my defence, the firmest of all if they rest on the judgement of his own people. Nor indeed could his youth be sufficiently commended to you if it were displeasing not only to a parent of such quality, but also to a township so distinguished and so weighty. As for myself,

\ciceroSection{6} to come back to me, it is from these same fountains that I have flowed forth into the public reputation of men, and this work of mine in the Forum and the order of my life has spread out into men's estimation a little more widely on the strength of the commendation and judgement of my own. As to what was thrown at him on the score of his chastity, and what was kept up by all his prosecutors not by way of charges but of voices and abuses --- this M. Caelius will never take so bitterly as to repent of not having been born ill-favoured. For these are the abuses widespread against all whose youth had a free-born form and look. But it is one thing to abuse, another to prosecute. Prosecution requires a charge that it may define the matter, mark out the person, prove by argument, confirm by witness; abuse has no aim except contumely, which, if it is thrown more saucily, is called railing, if more wittily, urbanity.

\ciceroSection{7} Which part of the prosecution I was indeed astonished at, and took it ill, that it had been chiefly entrusted to Atratinus. For neither was it fitting, nor did his age call for it, nor (as you could see) did the modesty of an excellent young man permit him to engage in such a speech. I could have wished some of you sturdier men had taken on this office of abuse: somewhat more freely and more strongly and more after our manner we should refute that licence of slander. With you, Atratinus, I shall deal more gently, both because your modesty restrains my speech, and because I owe it to my service to you and to your father to protect him. There is, however, this I want you to be admonished of:

\ciceroSection{8} first, that what you are, all may judge you to be --- that as far as you stand off from foulness in your conduct, so far you separate yourself from licence in your words; next, that you do not say of another things at which, if they were falsely returned upon you, you would blush. For who is there to whom that road does not lie open? who is there who could not, as petulantly as he wished, abuse this age and this standing --- even if without the least suspicion, yet not without an argument? But the fault of those parts of the case is on the men who wished you to plead them; the praise of your own modesty is that we saw you saying these things unwillingly, of your talent, that you said them ornately and elegantly.

\ciceroSection{9} But to that whole speech the defence is brief. For so far as the age of M. Caelius could give a foothold to that suspicion, it was first defended by his own modesty, and then also by his father's care and discipline. From the time the father gave him the toga of manhood --- I shall say nothing here of myself; let it be only as much as you yourselves estimate; this I shall say, that he was led from his father straight to me --- no one ever saw this M. Caelius in that bloom of his age except either with his father or with me, or being trained in the most chaste house of M. Crassus in the most honourable arts.

\ciceroSection{10} As to the familiarity with Catiline that has been thrown at Caelius, that ought to be far removed from such a suspicion. For you know that, when Caelius was already a young man, Catiline stood for the consulship with me. If he ever went over to him, or ever fell off from me --- although many fine young men were attached to that worthless and wicked man --- then let Caelius be reckoned to have been too closely Catiline's familiar. ``But afterwards we know and we saw that this man was even among that man's friends.'' Who denies it? But this period of life, which itself is by its own nature weak, and harassed besides by the lust of others, this it is that I am here defending. He was constantly with me when I was praetor; he did not know Catiline. Africa Catiline was at that time holding as praetor. Then followed the year in which Catiline pleaded his case on the charge of extortion. Caelius was with me; to that man he never came even as a supporter. Then came the year in which I stood for the consulship; Catiline was standing with me. Caelius never went near him; from me he never withdrew.

\ciceroSection{11} For so many years, then, he had passed his life in the Forum without suspicion, without ill repute, when he attached himself to Catiline standing a second time. To what limit, then, do you suppose his youth ought to have been kept under guard? In our time at least one year used to be set down for keeping the arm under the toga; for using exercise and the games of the Campus we were in the tunic; and the same was the rule of the camp and of the soldier's life, if we had at once begun to serve. At that age, unless a man defended himself by his own gravity, his chastity, both his domestic discipline and a certain natural goodness besides, however he had been kept watch over by his own people, he could not still escape genuine ill repute. But of him who had kept those first beginnings of his age sound and untouched, of his fame and his chastity, when now he had braced himself and was a man among men, no one spoke.

\ciceroSection{12} ``But Caelius attached himself to Catiline, when he had already been some years in the Forum.'' And many of every order and every age did the same. For Catiline, as I think you remember, possessed --- not painted out, but lightly sketched in --- very many likenesses of the greatest virtues. He used many wicked men, and yet he pretended to be devoted to the best of men. There were in him many enticements to lust; there were also certain spurs to industry and labour. The vices of lust were blazing in him; there flourished too the studies of war. Nor do I think there ever was such a monster on earth, blended together as he was out of contrary and various and warring natural pursuits and desires.

\ciceroSection{13} Who, at one time, was more agreeable to the more distinguished men, who more closely tied to the more disgraceful? who was sometimes a citizen of the better party, who a more dreadful enemy to this state? who in pleasures more polluted, who in labours more enduring? who in rapacity more grasping, who in lavishness more pouring out? Those qualities indeed, gentlemen, in that man were astonishing: to embrace many men in friendship, to retain them by attentiveness, to share with all what he had, to serve every man's needs with his money, his influence, his bodily labour, even with crime, if it were needed, and audacity; to turn his nature and govern it according to occasion, to twist and bend it this way and that --- with the strict to live strictly, with the relaxed pleasantly, with the elderly seriously, with the youthful affably, with criminals audaciously, with the libertines in luxury.

\ciceroSection{14} With this so varied and manifold a nature, when he had gathered every wicked and audacious man out of every land, so too he held many brave and good men by a certain semblance of a counterfeit virtue. Nor would that abandoned onset of destroying this empire ever have come from him, if so vast a savagery of so many vices had not rested on certain roots of complaisance and forbearance. Therefore let this charge, members of the jury, be spat out, and let no charge of familiarity with Catiline cling. For it is shared with many, even with some good men. He almost deceived even me --- me, I say --- when he seemed to me both a good citizen and eager for every best man and a steadfast and faithful friend; and his deeds I caught with my eyes before my opinion, with my hands before my suspicion. Whose great packs of friends, if Caelius too was among them, it is rather for him to be sorry to have made a mistake, just as I sometimes regret my own mistake about that same man, than for him to fear the charge of that friendship.

\ciceroSection{15} And so your speech has slid down from the abuses of unchastity to the odium of the conspiracy. For you laid down --- yet that hesitantly and curtly --- that this man was a partner of the conspiracy on account of his familiarity with Catiline; in which not only the charge would not stick, but the speech of an eloquent young man scarcely held together. For what so great madness was there in Caelius, what so great a wound either in his character and nature, or in his property and fortune? Where, indeed, was Caelius's name ever heard in that suspicion? I am speaking too much about a matter not in the least doubtful; this, however, I do say. Not only if he had been a partner in the conspiracy, but unless he had been a most bitter enemy of that crime, he would never have wished, by an accusation of conspiracy, to commend his own youth in particular.

\ciceroSection{16} Whence I rather think that, on the score of canvassing and these charges of clubs and bribery-agents, since I have fallen on the matter, I should answer in like fashion. For Caelius would never have been so out of his mind that, if he had stained himself with that boundless canvassing, he would have prosecuted another for canvassing; nor would he have looked for suspicion of a deed in another, of which he was wishing perpetual licence for himself; nor, if he thought he had once to undergo the peril of a canvassing prosecution, would he himself have brought another a second time on a charge of canvassing. Which, although he is doing it neither wisely and against my will, yet has the kind of eagerness that he seems rather to be hounding another's innocence than fearfully thinking about his own.

\ciceroSection{17} As to the debt that has been thrown at him, the criticised expenses, the demanded account-books --- see how few words I shall use in answer. Account-books a man who is in his father's power keeps none. He has never once contracted any new loan against the old. The expense of one kind has been thrown at him, of his housing: you said he lives at thirty thousand. Now at last I understand: P. Clodius's apartment block is for sale; this man lives in some little rooms there at, I think, ten thousand. You, while you wished to please him, accommodated your lie to his moment.

\ciceroSection{18} You blamed him for moving out of his father's house. That, at his age, is the very last thing to be blamed. For when, both from a public case, he had achieved a victory irksome to me indeed, but glorious to himself --- and on account of his age was now able to seek office --- he moved out from the father not only with the father's leave but with his urging; and, since the father's house was far from the Forum, that he might more easily go his rounds of our houses and himself be courted by his own circle, he rented a house on the Palatine, and not at a great rent. At which point I can say what M. Crassus, that most distinguished man, said a little while ago when he was complaining about the coming of king Ptolemy: ``Would that not in the Pelian grove'' --- and I might be allowed to weave on this song further: for never would the wandering mistress have given us this trouble, ``Medea sick at heart, by savage love wounded.'' For thus you will discover, gentlemen, what (when I have come to that place) I shall show: that this Palatine Medea and this moving were the cause to the young man either of all his ills or rather of all this talk.

\ciceroSection{19} Wherefore those things which I had perceived were already being prepared and patched up out of the prosecutors' speech, I do not greatly fear, members of the jury, relying on your prudence. For they were saying that there would be a witness, a senator, who would say he had been struck by Caelius at the pontifical elections. Of him I shall ask, if he comes forward, first why he did nothing immediately; next, if he preferred to complain rather than to act, why brought forward by you rather than of his own accord; and why so long after, rather than at once. If he replies sharply and acutely to these things, then at last I shall ask from what spring this senator flows. For if of himself he rises and is born, perhaps, as I do, I shall be moved; but if he is a little stream summoned and led from the very source of your accusation, I shall rejoice that, when your accusation rests on so great influence and so great resources, only one senator has been found who would care to oblige you. As to the witness Fufius.

\ciceroSection{20} Nor, however, do I dread that other class of nocturnal witnesses. For it has been said by them that there will be men who will say that their wives, returning from dinner, were assaulted by Caelius. Weighty men they will be, who will dare to say this on oath, when they must confess that they have never even tried, by meeting and arrangement, to seek redress for so great an injury. But the whole kind of this assault, members of the jury, you both look forward to in mind, and, when it shall be brought, you will have to repulse. For M. Caelius is not being prosecuted by the same men by whom he is being assaulted. Openly are weapons hurled at him; secretly they are passed up.

\ciceroSection{21} Nor do I say this so as to make it odious for those for whom it ought even to be glorious. They are doing their duty, defending their own, doing what the bravest men are wont to do; injured, they grieve; angered, they are carried away; provoked, they fight. But it is, members of the jury, for your wisdom to consider that, even if there is just cause for brave men to assault M. Caelius, it is not therefore a just cause for you to think rather of another's pain than of your own faith. You see, indeed, what a multitude is in the Forum, what kinds, what pursuits, what variety of men. Out of all this number, how many do you think there are who, when they think powerful, popular, eloquent men want something, are accustomed to offer themselves unbidden, to busy themselves, to promise testimony?

\ciceroSection{22} If any of this kind have perchance thrust themselves into this trial, shut out their lust, members of the jury, by your wisdom, that you may be seen at the same time to have looked out both for this man's safety and for your own scruple, and against perilous influences of men, for the condition of all citizens. I shall lead you away from witnesses, and shall not allow the truth of this trial, which can in no way be altered, to be lodged in the will of witnesses, which can most easily be feigned, and bent and twisted with no trouble. We shall act by arguments, we shall refute the charges by signs clearer than any daylight; thing will fight thing, case case, reason reason.

\ciceroSection{23} And so I am content that that part of the case has been weightily and elegantly pleaded out by M. Crassus --- about the Neapolitan seditions, about the assault on the Alexandrians at Puteoli, about Palla's property. I could have wished it had been said by the same man also about Dio. Yet about that very matter, what is there for you to wait for? The man who did it either does not fear, or even confesses; for he is a king. The man, however, who is said to have been his helper and accomplice, P. Asicius, has been freed by trial. What sort of charge, then, is it that he who committed it does not deny, that he who denied it has been acquitted, but at which this man should be afraid, who has been free not only from doing the deed but even from the suspicion of complicity? And if Asicius was helped more by his case than he was hurt by ill-will, will it injure this man, who has been spattered not only with no suspicion of that deed, but not even with any ill repute?

\ciceroSection{24} ``But Asicius was acquitted by collusion.'' It is most easy to answer that point, especially for me by whom that case was defended. But Caelius judges Asicius's case to be the best one; whatever sort it may be, he thinks it separate from his own. And not only Caelius, but also the most cultivated and learned young men, men with the rightest pursuits and the best arts, T. and C. Coponius, who, of all men, most grieved at the death of Dio, both for the love of learning and humanity, and also because they were bound to him by the bond of hospitality. Dio was staying with Titus, as you have heard; he had been known to Titus at Alexandria. What he or his most splendid brother think about M. Caelius you will hear, if they are produced, from the men themselves.

\ciceroSection{25} Therefore let these things be put aside, that we may at last come to those things on which the case rests. For I noticed, members of the jury, that you listened most attentively to my friend L. Herennius. In whose speech, although you were largely held by his talent and a certain genus of speaking, yet I sometimes feared that that subtly-built up speech to incriminate would creep little by little and softly into your minds. For he said much about luxury, much about lust, much about the vices of youth, much about morals; and he, who in his other life was mild and was wont to range most pleasantly in that kind of urbanity in which now almost everybody takes pleasure, was in this case some grim sort of paternal uncle, censor, schoolmaster: he scolded M. Caelius as no parent ever scolded; he discoursed at length on incontinence and intemperance. What more, members of the jury? I forgave you for listening attentively, because I myself shrank from that sad, that harsh kind of speech.

\ciceroSection{26} The first part of his speech moved me less: that Caelius had been a familiar of my friend Bestia, had dined at his house, had often gone to his house, had supported him for the praetorship. These things, which are plainly false, do not move me; for he said that men were dining together who are either away or who are obliged to say the same thing. Nor again does it move me that he said Caelius was a fellow-member with him among the Luperci. A wild brotherhood it is, indeed, of the German Luperci, and quite of the shepherds and the country: their woodland coming-together was instituted before humanity and the laws --- if indeed the brotherhood-fellows not only lay each other's names in court but even draw attention to the brotherhood when they are accusing, lest, perchance, anyone should not know it, they seem to fear!

\ciceroSection{27} But these things I leave; I answer those things which moved me more. The reproof of his pleasures was long, and even gentler, and had more disputation than savagery, on which account it was listened to the more attentively. For when P. Clodius, my friend, in his most weighty and vehement self-presentation, on fire, was driving everything along with the saddest words and the loudest voice, although I approved his eloquence, I was nevertheless not afraid; for I had seen him in not a few cases litigating in vain. To you, however, Balbus, I reply first by petition: if it is permissible, if it is right, that I defend a man who has refused no banquet, who has been in the gardens, who has put on perfumes, who has seen Baiae.

\ciceroSection{28} I myself have seen and heard of many in this state, not only those who had merely tasted with the tip of the lip this kind of life, and touched it (as it is said) with the very tip of their fingers, but those who had given the whole of their youth over to pleasures, who at some time emerged and (as is said) recovered themselves to the good fruit, and have been weighty and famous men. For there is granted, by the consent of all, some play to this age, and nature herself pours out the desires of youth. If they break out in such a way as to shake no one's life, to overthrow no one's house, they are wont to be held easy and bearable.

\ciceroSection{29} But you seemed to me to wish to gather up some odium against Caelius from the common ill-fame of youth. And so all that silence which was given to your speech was for this reason: because, with one defendant set forth, we were thinking of the vices of many. It is easy to denounce luxury. The day would now fail me, if I should try to bring out what can be said in that direction; on seducings, on adulteries, on impudence, on extravagances the speech is endless. Even if you set forth no defendant but those vices, still the matter itself can be denounced both copiously and weightily. But it is for your wisdom, members of the jury, not to be drawn off from the defendant, nor, when the prosecutor has set up the stings of your severity and weight in a thing, in vices, in habits, in times, to let them out upon the man, upon the defendant --- when he has been called into a kind of unjust hatred not on his own charge but on the vice of many.

\ciceroSection{30} And so I do not dare to answer your severity as is fitting. For it would have been mine to beg off the dispensations of youth and ask leave. I do not dare, I say; I use no refuges of age, I let go the rights granted to all. I ask only this: that, if there is any common ill-fame at this time of the debt, the petulance, the lusts of youth (which I see is great), let not another's misdeeds, the vices of an age and a time, harm this man. And I, who ask these things, do not refuse to answer most diligently the charges which are brought against him in particular. There are then two charges, of gold and of poison; in which one and the same person is involved. The gold was taken from Clodia; the poison was sought to be given to Clodia, as is said. All the rest are not charges but abuses, of a sauciness of railing rather than of public inquiry. ``Adulterer, debauchee, broker'' --- this is railing, not accusation. For there is no foundation, no seat to these charges; they are contumelious voices rashly let out by an angry prosecutor with no author. Of these two charges I see the source.

\ciceroSection{31} I see the spring, I see the certain name and head. Gold was needed; he took it from Clodia, took it without a witness, kept it as long as he wished. I see the greatest sign of some out-of-the-way familiarity. He wanted to kill the same woman; he sought poison, suborned slaves, prepared the draught, fixed the place, brought it secretly. Again, I see a great hatred to have arisen from a most cruel breach. The whole business in this case for us, members of the jury, is with Clodia, a woman not only noble but notorious; about whom I shall say nothing except for the sake of warding off a charge.

\ciceroSection{32} But you understand, in your eminent prudence, Cn. Domitius, that we have to do with this woman alone. If she does not say that she lent gold to Caelius, if she does not allege that poison was prepared for her by him, we are acting petulantly, if we name a matron otherwise than the sanctity of matrons demands. But if, with that woman set aside, no charge or resources are left them to assault M. Caelius, what else is there for us as advocates to do, but repulse those who hound him? Which I should do the more vehemently, were it not that enmities stand in my way with this woman's husband --- I meant to say her brother; I always make this slip. Now I shall act with restraint, and shall not go further than my faith and the case itself shall force me to: for I have never thought I had to wage women's enmities, especially with a woman whom all have always reckoned rather everyone's friend than anyone's enemy.

\ciceroSection{33} But still I shall ask of her first whether she would prefer that I deal with her severely and gravely and in the old style, or relaxedly and gently and politely. If in that austere manner and mode, I must call up some bearded one of those old men from the underworld --- not with the little beard which she takes pleasure in, but with that shaggy one we see in the antique statues and busts --- to scold the woman and speak for me, lest she perchance be angry with me. Let one, then, of this very family come up, and best of all the Blind one; for he will feel the least pain, who will not see her. He, surely, if he comes forth, will deal with her and speak as follows: ``Woman, what have you to do with Caelius, what with a young man, what with a stranger? Why have you been either so familiar as to lend gold, or so hostile as to fear poison? Have you not seen your father, have you not heard that your father's brother, your grandfather, your great-grandfather, your great-great-grandfather, your forefather of the fifth degree, were consuls?

\ciceroSection{34} did you not, finally, just now know yourself to have held the marriage of Q. Metellus, a most distinguished and bravest man, most loving of his country, who, the moment he had set his foot over the threshold, surpassed nearly every citizen in virtue, glory, and standing? When you had married out of a most ample family into a most distinguished house, why was Caelius so closely joined to you? Was he kinsman, connection, your husband's familiar? Nothing of the sort. What then was it, if not a kind of recklessness and lust? If our manly likenesses did not move you, did not even my own descendant, that Q. Claudia, urge you to be a rival of domestic praise in the glory of women --- not that virgin Vestal Claudia, who, embracing her father in his triumph, did not allow him to be dragged from the chariot by an enemy tribune of the plebs? Why have your brother's vices moved you rather than your father's good qualities, and your grandfather's, and those traced down through us in men and even in women? For this did I sunder the peace of Pyrrhus, that you should daily strike compacts of most foul loves; for this did I bring water in, that you should use it incestuously; for this did I lay down a road, that you should throng it in the company of strangers' husbands?''

\ciceroSection{35} But why, members of the jury, have I brought in so weighty a person, that I fear lest the same Appius should suddenly turn round and begin to prosecute Caelius with that censorial gravity of his? But I shall see to that later, and in such a way, members of the jury, that I trust I shall prove M. Caelius's life even before the most severe disputants. You, however, woman --- for now I speak with you in person, with no character introduced --- if you propose to make good what you do, what you say, what you charge, what you contrive, what you allege, you must give and set forth the reason of so great a familiarity, of so great a closeness, of so great a conjunction. The prosecutors, indeed, fling about lusts, loves, adulteries, Baiae, beaches, banquets, drinking-parties, songs, music, sailings, and they declare that they say nothing without your agreement. Which things, since you, by some unbridled and headlong mind, have wanted to be brought into the Forum and into court, you must either dissolve them and prove them false, or confess that we ought to give credence neither to your charge nor to your testimony.

\ciceroSection{36} But if you would rather have me act in a more polite fashion, this is how I shall deal with you. I shall remove that hard old man, almost a peasant; I shall therefore take one of these, and best of all your littlest brother, who is in this kind of thing the most polished --- who loves you most, who, on account of some I-suppose-timidity and certain idle nocturnal terrors, used always when a little boy to sleep with you his elder sister. Imagine him saying to you: ``Why are you in an uproar, sister? Why are you out of your senses? Why, having let your shouting begin, do you make a small thing great by your words? You have caught sight of the young man next door; his fairness, his height, his face and eyes have struck you; you wished to see him oftener; you have sometimes been in the same gardens; you, a noble lady, want to bind that young man, son of a thrifty and tight-fisted father, to you with your resources. You cannot; he kicks, spits out, repels, judges your gifts not worth that much. Take yourself elsewhere. You have gardens beside the Tiber, carefully laid out at the very place to which all the youth comes for swimming; from this place you can pick your bargains daily; why are you a nuisance to this man who scorns you?''''

\ciceroSection{37} I now turn back to you in turn, Caelius, and take upon myself paternal authority and severity. But I am at a loss which father best to take --- some Caecilian, vehement and hard: ``\textit{Now at last my soul burns up, now my heart is heaped up with anger}'' --- or that one: ``\textit{O wretched, O criminal!}'' Iron these fathers are: ``\textit{Am I to say what I would like, what I would have? --- which you so by your foul deeds make sure that I want in vain}'' --- such fathers are scarcely to be borne. Such a father would say: ``Why have you taken yourself into that courtesan's neighbourhood? Why, when you knew the snares, did you not flee? Why have you known any other man's woman? Squander it, scatter it; with my leave you may. If you come into want, you will hurt; not I. I have enough to enjoy what is left of my life.''

\ciceroSection{38} To this hard, blunt old man Caelius would answer that he had been led aside from the road by no lust. The proof? --- no expenses, no wastings, no new loans against old. ``But there was a rumour.'' How few can escape that, especially in so foul-mouthed a state? Do you wonder that the neighbour of that woman has been ill spoken of, when her own brother could not escape the talk of the unfair? But to a mild and merciful father, of the kind who is here --- ``\textit{He has broken in the doors; they will be put back. He has torn the dress; it will be patched up}'' --- Caelius's case is most easily disposed of. For what could there be in which he could not easily defend himself? I say nothing now against that woman; but if there were anyone unlike her, who set herself before all comers, who plainly always had a paramour openly named, into whose gardens, house, and Baiae the lusts of all men had a public-right of way, who even fed young men and supported the thrift of fathers with her expenses; if as a widow she lived freely, as a saucy one petulantly, as a rich one lavishly, as a lustful one in the courtesan's manner --- should I think a man an adulterer, if any greeted her a little less guardedly?

\ciceroSection{39} Someone will say: ``Is this then your discipline? Is it thus that you train young men? Is this why the parent commended this boy to you and put him in your keeping --- that he should set his youth in love and pleasures, and that you should defend this life and these pursuits?'' I, members of the jury, if there has been any of this hardness of soul and this temper of virtue and continence, that he should spit out every pleasure and finish the whole course of his life in the labour of his body and the strain of his mind --- whom no rest, no relaxation, no pursuits of his peers, no games, no banquet would delight, who should think nothing in life worth seeking except what is joined with praise and dignity --- this man, in my opinion, I think gifted and adorned with certain divine goods. From this kind of man I think were those Camilli, Fabricii, Curii, and all those who out of the smallest things made these so great.

\ciceroSection{40} But these kinds of virtues are now scarcely to be found, not only in our manners, but even in books. Even the leaves which contained that ancient severity have grown obsolete; nor only with us, who have followed this profession and reasoning of life in act rather than in words, but even with the Greeks, the most learned of men, to whom, when they could not act, it was nevertheless permitted to speak and write honourably and grandly: in the changed times of Greece, certain other precepts have arisen.

\ciceroSection{41} And so some have said that wise men do all things for the sake of pleasure, nor have learned men shrunk from this foulness of speech; others have thought pleasure must be joined with dignity, so that they might join, by their facility of speaking, things that most repugn each other; that one straight road to praise by labour --- those who have approved it are now nearly alone left in the schools. For nature herself has begotten in us many enticements by which a slumbering virtue might at times wink; she has shown to youth many slippery roads, on which it could scarcely set foot or step without some fall and slipping; she has given a variety of most pleasant things, by which not only this age but even maturity already strengthened might be caught.

\ciceroSection{42} Wherefore if you find anyone who casts away with his eyes the fairness of things, who is taken by no smell, no touch, no taste, who shuts out from his ears every sweetness, this man perhaps I, and a few, will think the gods are with; but most will think them angry. Therefore let this deserted and untrodden way, now blocked with leaves and boughs, be left aside. Let some play be granted to youth; let young manhood be a little freer; let not all things be denied to pleasures; let not always that true and straight reason prevail; let desire and pleasure sometimes win against reason, provided that in this kind there be kept that prescription and moderation. Let youth spare its own chastity, not despoil another's, not pour out its inheritance, not be slaughtered by usury, not break into another's house and family, not bring shame on the chaste, a stain on the upright, ill-fame on the good, not terrify any with violence, not be present at ambushes, keep clear of crime. At last, when it has obeyed pleasures, has given some time to the play of the age and to these idle desires of youth, let it call itself back at some point to the care of the household, of the Forum, and of the commonwealth, that it may seem to have cast away by satiety and to have despised by experience those things which earlier it had not seen by reason.

\ciceroSection{43} And many indeed within our memory and that of our fathers and forefathers, members of the jury, have been the highest of men and most distinguished citizens, in whom, when the lusts of youth had cooled, eminent virtues, with their age now firmed, came forth. Of whom I do not wish to name any one; you yourselves recall them with yourselves. For I do not wish to join even the smallest error of any brave and famous man with his greatest praise. If I were willing to do this, many of the highest and most ornamented men would be cited by me, of whom would be named on the one side too much licence in their youth, on the other lavish luxury, vast debt, expenses, lusts --- which afterwards, covered over by many virtues, anyone who wished could defend by the excuse of youth.

\ciceroSection{44} But indeed in M. Caelius --- for now I will speak the more confidently of his honourable pursuits, since I dare, relying on your wisdom, to confess certain things freely --- no luxury will be found, no expenses, no debt, no lust of banquets and brothels. That vice indeed of the belly and the gullet not only does not lessen with age in men but even increases. As for love-affairs and ``charms'' so called, which are not wont to be lasting troubles to those endowed with a firmer mind --- for they pass through their bloom early and quickly --- they have never held this man taken or hindered.

\ciceroSection{45} You have heard him when he was speaking on his own behalf, you heard him before when he was prosecuting --- I say this for the cause of defending, not of boasting --- and the kind of his eloquence, his fluency, the abundance of his thoughts and his words, you have, with your prudence, perceived. And in him you saw not only his talent shine forth (which often, even if it is not nourished by industry, yet has its own strength), but in him there was, unless my goodwill perchance deceived me, a method both built up by good arts and elaborated by care and watchings. And know, gentlemen, that those desires which are flung at Caelius and these pursuits of which I speak cannot easily be in the same man. For it cannot be that a mind given over to lust, hindered by love, longing, desire, often by too great abundance and even sometimes by want, should be able to bear up, not only in pleading but even in thinking, this whatever-it-is which we do in speaking, however we do it.

\ciceroSection{46} Or do you think there is any other reason why, with such great rewards of eloquence, with such great pleasure in speaking, such great praise, such great influence, such great honour, so few there are, and always have been, who toil at this labour? All pleasures must be trodden down, the pursuits of delight must be left aside, the play, the joke, the banquet, the conversation almost of intimates must be deserted. For this reason in this kind labour offends men and frightens them off the pursuit, not because either talent fails or boyhood's training.

\ciceroSection{47} Or would this man, if he had given himself up to that life, while still very young have called a consular man into court? Would this man, if he were fleeing labour, if he were held bound by pleasures, daily go his rounds in this battle-line, seek enmities, call into court, undergo capital peril, and, with the Roman people watching, fight now for so many months either for safety or for glory? Does that nearby neighbourhood smell of nothing, do men's reports declare nothing, does Baiae itself even speak nothing? Indeed Baiae does not only speak, but it cries out --- that the lust of one woman has slid down to such a point that she does not seek solitude and darkness and these coverings of disgraces, but in the foulest things rejoices in the most thronged frequency and the most brilliant daylight.

\ciceroSection{48} But if there be anyone who thinks that even courtesan loves are forbidden to youth, he is indeed very severe --- I cannot deny it --- but he stands far off not only from the licence of this age but also from the custom and the leave of the ancestors. For when has this not been done, when has it been censured, when has it not been allowed, when, lastly, has it been so that what was lawful was not lawful? Here I will define the very thing; I will name no woman; only so much I shall leave in the middle.

\ciceroSection{49} If any unmarried woman opens her house to all men's lust, and openly sets herself up in courtesan's life, has gotten the habit of the dinner-parties of the most foreign men --- if she does this in the city, if in her gardens, if in that frequented Baiae, if, lastly, she so bears herself, not only in her gait but in her dress and retinue, not only in the blaze of her eyes, the freedom of her speech, but even in her embracing, kissing, beach-going, sailings, banquets, that she seems not only a courtesan but even a forward and saucy courtesan --- if some young man chances to have been with this woman, does he, L. Herennius, seem to you an adulterer, or a lover; to have wished to storm her chastity, or to fill his lust?

\ciceroSection{50} I now forget your injuries, Clodia, I lay aside the memory of my pain; what was cruelly done by you against my own people in my absence, I overlook; let what I have said not be said against you. But of you yourself I ask, since the prosecutors say they have a charge from you and a witness for that charge, you yourself. If there be such a woman as I have described a moment ago, unlike you, of a courtesan's life and habits --- to have had with her any account, can it seem to you most foul or most disgraceful? If you are not such, as I prefer, what is it that they throw at Caelius? But if they wish you to be such, what reason is there that we should fear this charge, if you make light of it? Therefore give us a way and a method of defence. For either your modesty will defend us by saying that nothing has been done by M. Caelius too forwardly, or your shamelessness will give to him and to the rest a great means of defending themselves.

\ciceroSection{51} But since my speech now seems to have got out of the shoals and to have run past the rocks, what remains of the course shows itself most easy to me. There are, in one woman, two charges of the highest crimes: of the gold which is said to have been taken from Clodia, and of the poison which they accuse Caelius of having prepared for the killing of the same Clodia. The gold, you say, he took to give to L. Lucceius's slaves, by whom Dio of Alexandria, who was at that time staying with Lucceius, was to be killed. A great charge, whether of laying snares for a legate, or of soliciting slaves for the killing of their master's guest --- a counsel full of wickedness, full of audacity!

\ciceroSection{52} In which charge first I demand: did he say to Clodia why he took the gold, or did he not say? If he did not say, why did she give it? If he did say, she has bound herself by the same crime of complicity. Did you dare bring out the gold from your cabinet, did you dare strip your Venus of her ornaments --- you, the spoiler of the rest of them --- when you knew for what so great a crime this gold was being sought, for the killing of a legate, for the everlasting stain of crime upon L. Lucceius, a most holy and most upright man? Your generous mind ought not to have been an accomplice of so great a crime, your popular house its minister, that hospitable Venus of yours, finally, its helper.

\ciceroSection{53} Balbus saw this; he said Clodia was kept in the dark, and that Caelius had brought to her the story that he was looking for gold for the trimming of games. If he was as familiar with Clodia as you wish him to be, since you say so much about her lust, he certainly told her where he wanted the gold; if he was not so familiar, she did not give it. Thus, if Caelius told you the truth --- O immoderate woman --- you, knowing, gave gold for crime; if he did not dare to tell, you did not give it. Why should I now, with arguments which are countless, resist this charge? I can say that the morals of M. Caelius are most far from the atrocity of so great a crime; that least of all is it to be believed that to a man so talented and so prudent it should not have come into mind that a deed of such crime ought not to be entrusted to unknown and another's slaves. I can also, by the custom of the rest of the patrons and my own, ask the prosecutor where Caelius came together with Lucceius's slaves, what approach there was for him; if of his own accord, with what recklessness! if through another, through whom? I can travel through every lurking-place of suspicion in speaking; no cause, no place, no opportunity, no accomplice, no hope of carrying out or hiding the deed, no method, no trace of the great crime, will be found.

\ciceroSection{54} But these things, which are an orator's province, which to me, not on account of my talent but on account of this practice and use of speaking, could have brought some fruit, when seen to be put forward by my own elaboration, for brevity's sake I leave aside altogether. For I have, members of the jury, a man whom you can easily allow to be a partner of your scrupulousness and your oath: L. Lucceius, a most holy man and a most weighty witness, who would neither have not heard that so great a crime against his fame and his fortunes had been borne in by M. Caelius, nor have neglected it, nor have borne it. Or could that man, endowed with that humanity, those pursuits, those arts, that learning, have neglected the peril of the very man whom he loved on account of these very pursuits, and have omitted to take care, in his guest's case, of what he would severely receive if it were aimed at a stranger? What he would have grieved at if he had found out that it had been done through unknown hands, would he have neglected when attempted by his own slaves? What he would blame in fields and public places, would he lightly have borne if it were begun in the city and at his own house? What in the peril of some country man he would not have passed over, would he, an educated man, have thought ought to be dissembled in the snares of a most learned man?

\ciceroSection{55} But why should I keep you longer, gentlemen? Receive his sworn-conscience and authority, and learn diligently every word of his testimony. Read. \textit{Testimony of L. Lucceius.} What more do you wait for? Or do you think the case itself and truth can send out any voice on their own behalf? This is innocence's defence, this is the case's own speech, this is the one voice of truth. In the charge itself there is no suspicion, in the matter no argument, in the business which is said to have happened no trace of the speech, of the place, of the time; no witness, no accomplice is named; the whole charge is brought out from a hostile, ill-famed, cruel, criminal, and lustful house. But that house which is said to have been attempted by this nefarious crime is full of integrity, dignity, duty, religion; from which house there is read out to you authority bound by oath --- so that there be put in contention a thing not at all to be doubted: whether a rash, saucy, angry woman is judged to have made up the charge, or a weighty, wise, moderate man to have given testimony scrupulously.

\ciceroSection{56} The remaining charge, then, is of the poison; whose beginning I cannot find, nor unfold its end. For what cause was there for which Caelius should wish to give that woman poison? That he might not return the gold? She did not ask for it. That a charge might not stick? Did anyone bring one? Did anyone, finally, have made mention, if he had laid no name against any? Indeed you have heard L. Herennius say that he would not have been a trouble to Caelius even by a word, if for the second time he had not laid the name of Bestia (his connection) on the same charge after the first man had been acquitted. Is it credible, then, that so great a crime was committed without any cause? And do you not see that a charge of greatest crime is being feigned, in order that there may seem to have been a cause for the undertaking of another crime?

\ciceroSection{57} To whom, finally, did he entrust it, what helper did he use, what partner, what confidant, to whom did he commit so great a deed, to whom himself, to whom his own safety? To the woman's slaves? For so it has been objected. And was he so out of his senses --- he, to whom you certainly grant talent, even if you take other things away by your hostile speech --- as to entrust all his fortunes to another's slaves? But to what slaves? --- for this very point matters very greatly --- the very ones whom he understood not to use the common condition of slavery, but to live with their mistress more freely, more loosely, more familiarly? For who does not see, gentlemen --- or who is unaware --- that in such a household, in which the matron of the family lives in a courtesan's manner, in which nothing is done that ought to be brought outside, in which strange lusts, luxury, and finally all unheard-of vices and disgraces go round, that here the slaves are not slaves, with whom everything is shared, by whom things are done, who themselves take part in the same pleasures, to whom secrets are entrusted, on whom even some of the daily expense and luxury overflows? Was Caelius then not seeing this?

\ciceroSection{58} For if he was as familiar with the woman as you wish, he knew that those slaves too were familiars of their mistress. But if he had not so great an intimacy as is brought in by you, what so great familiarity could there be with her slaves? But of the poison itself, what reason is feigned? Where was it sought, in what way was it prepared, by what arrangement, to whom, in what place was it handed over? They say he had it at home, and tried its strength on a certain slave prepared for that very purpose, by whose very swift death the poison was, by him, approved.

\ciceroSection{59} Immortal gods! Why do you sometimes wink at the great crimes of men, or reserve the punishments of the present cheat for another day? For I saw, I saw and drank in that bitterest grief in life, when Q. Metellus was dragged from the bosom and embrace of his country, and when that man, who thought himself born for this empire, on the third day after he had flourished in the senate-house, on the rostra, in the commonwealth, in the most untouched age, in the best constitution, with the greatest strength, was, most unworthily, snatched from all good men and the whole state. At which time, dying, when his mind was now overwhelmed in the rest of its parts, he reserved his last sense for the memory of the commonwealth: looking at me as I wept, he signified with broken and dying voice how great a tempest hung over me, how great a storm over the state, and, often striking the wall which had been common to him and Q. Catulus, kept naming Catulus, often me, oftenest the commonwealth, that he grieved not so much that he was dying as that the country and I too were being despoiled of his protection.

\ciceroSection{60} If no force of sudden crime had taken him off, that man, how, as a consular, would he have stood up to his furious brother --- he, who as consul, when his brother was beginning to rage and thunder, said in the senate's hearing that he would kill him with his own hand! From this house, then, has that woman come forth and dared to speak about the swiftness of poison? Will she not herself fear the very house, that it may not utter some voice; will she not shudder at the conscious walls, at that funereal and mournful night? But I return to the charge; for the mention of that most distinguished and bravest man has both weakened my voice with weeping and hindered my mind with grief.

\ciceroSection{61} But still where the poison came from, in what way it was prepared, is not said. They say it was given to this P. Licinius, a modest and good young man, Caelius's familiar; that an arrangement was made with the slaves that they should come to the Senian baths; that Licinius would come to the same place and would deliver to them the box of poison. Here first I ask: what was the use of having it brought to that arranged place; why did those slaves not come to Caelius's home? If that great habit of Caelius, that great familiarity with Clodia, was still going on, what would there be of suspicion if a slave of the woman were seen at Caelius's? But if there now lay underneath a falling-out, if the habit was put out, if there had been a breach --- ``hence those tears,'' surely, and this is the cause of all these crimes and charges.

\ciceroSection{62} ``On the contrary,'' he says, ``when the slaves had brought the whole thing and Caelius's evil-doing to their mistress, the ingenious woman gave them instructions to promise everything to Caelius; but, that the poison, when it was being delivered by Licinius, might be openly seized, she ordered the place to be appointed at the Senian baths, that she might send friends thither who should hide; then suddenly, when Licinius had come and was delivering the poison, that they should leap out and seize the man.'' All which things, members of the jury, have a most easy way of being refuted. For why had she fixed on the public baths in particular? in which I do not find what hiding-place there could be for men in the toga. For if they were to be in the entrance of the baths, they would not be hidden; but if they wanted to throw themselves into the deepest part, they could not, in shoes and clothes, do so very conveniently, and perhaps they would not have been received --- unless perchance the powerful woman, by that quadrans-bargain of hers, had become familiar with the bath-keeper.

\ciceroSection{63} And indeed I was vehemently expecting to hear who those good and worthy witnesses of this poison openly seized would be said to be; for none have so far been named. But I do not doubt that they are most weighty men, who, first, are familiars of such a woman, and next have undertaken this commission, that they should be crammed into the baths --- a thing she would never have got from any but most honourable men, full of dignity, however powerful she may be. But why am I speaking of the dignity of those witnesses? Learn their virtue and diligence. ``They lay in wait in the baths.'' Distinguished witnesses! ``Then they leapt out at random.'' Restrained men! For so you fashion the story: when Licinius had come, was holding the box in his hand, was trying to deliver it, had not yet delivered it, then suddenly out flew those splendid witnesses without a name; and Licinius, when he had now stretched out his hand to deliver the box, drew it back, and from that sudden onset of the men cast himself into flight. O great force of truth, which against men's talents, cunning, cleverness, against the feigned snares of all, easily defends itself by itself!

\ciceroSection{64} As, for instance, this whole little fable of an old female poet of very many fables --- how it lacks plot, how it can find no end! For why? --- those many men (for there must have been not a few, that Licinius might both easily be seized and the matter be witnessed by many eyes) --- why did they let Licinius slip out of their hands? Why was Licinius any the less easy to seize when he had drawn back, lest he should hand over the box, than if he had handed it over? For they had been posted there to seize Licinius, that Licinius might be openly caught either when he was holding the poison or when he had handed it over. This was the woman's whole plan, this was the commission of these who had been asked: I cannot find why you should say they leapt out at random and before the time. They had been asked for this purpose, they had been placed for this purpose: that the poison, the snares, finally the very crime, might be openly seized.

\ciceroSection{65} Could they have leapt out at a more fitting time than when Licinius had come, when he was holding the box of poison in his hand? Which, when it had now been delivered to the slaves, if the woman's friends had suddenly come out of the baths and seized Licinius, he would have implored men's good faith and would have stoutly denied that the box had been delivered by him. How would they confute him? Would they say they had seen it? First they would call upon themselves the charge of the greatest crime; next they would say they saw what they could not have seen from the place where they were stationed. Therefore at the very time they showed themselves, when Licinius had come, was getting out the box, was stretching out his hand, was delivering the poison. The end then is of a mime, not a play; in which, when no \textit{clausula} is found, somebody slips out of the hands, then the clappers crash, the curtain rises.

\ciceroSection{66} For I ask why that woman's-pack of yours let go from its hands Licinius staggering, hesitating, giving way, trying to flee --- why they did not seize him, why they did not, by his own confession, by many eyes, finally by the voice of the deed itself, wring out the charge of so great a crime. Were they afraid that, being so many to one, the strong against the weak, the keen against the terrified, they could not get the better? No argument will be found in the matter, no suspicion in the case, no end of the charge. Therefore this case, from arguments, from conjecture, from those signs by which truth is wont to be brought to light, has been transferred entirely to witnesses. Whom indeed I, members of the jury, await not only without any fear, but even with some hope of pleasure.

\ciceroSection{67} My mind already greatly longs to see, first, the gaily-dressed young men, the familiars of a wealthy and noble woman; then the brave men set in ambush by their commandress and as a guard of the baths. Of whom I shall ask in what manner they were hidden, or where; whether it was the bath-tank or some Trojan horse that bore and covered so many invincible men waging a woman's war. Indeed I shall force them to answer this: why so many and such men did not seize this one and so weak a man whom you see, either standing or following him as he fled --- they who, never indeed, if they come up to that place, will get themselves out of it. As witty, as ready of tongue at banquets, as sometimes eloquent in their cups as they may be, the strength of the Forum is one thing, of the dining-room another; the rule of the bench is one thing, of the couch another; the public look of jurors and of revellers is not the same; the daylight, finally, of the sun is far one thing, of the lamps another. We shall therefore shake out all the dainty-things of these men, all their inanities, if they come forward. But let them hear me, let them work at some other office, take up some other favour, show themselves off in other matters; let them flourish in their charm with that woman, lord it in expense, cling to her, lie at her feet, serve her; but let them spare the head and fortunes of an innocent man.

\ciceroSection{68} ``But those slaves were freed by the judgement of her relations, the noblest and most distinguished of men.'' At last we have found something which that woman is said to have done by the judgement and authority of her own kinsmen, the bravest of men. But I am eager to know what argument that manumission has; in which either a charge against Caelius was sought, or an inquiry was prevented, or a reward was paid with cause to slaves privy to many things. ``But it pleased her relations.'' Why should it not have pleased them, when you said that you yourself had brought to them, not from another, but discovered by yourself, the matter?

\ciceroSection{69} Do we wonder, here too, if a most obscene tale has followed up that fictitious box? Nothing is which does not seem to fall in upon such a woman. The matter has been heard and made loud in talk. You perceive, gentlemen, in your minds long since what I want, or rather what I do not want to say. Which thing, even if it has happened, certainly was not done by Caelius --- for what would it have been to him? --- but was done perhaps by some young man not so much dull as forward. But if it is feigned, it is not indeed a modest, but yet not an inelegant lie; which assuredly the talk and opinion of men would never have approved, unless everything that was said with any foulness seemed neatly to fit upon her.

\ciceroSection{70} The case has been pleaded and pleaded out by me, members of the jury. You now understand how great a judgement you are upholding, how great a matter has been entrusted to you. You inquire concerning violence. The law that pertains to the empire, to the majesty, to the standing of the country, to the safety of all --- the law which Q. Catulus brought in armed dissension of citizens in nearly the last times of the commonwealth, and which, when the flame of my consulship had been quieted, extinguished the smouldering remains of the conspiracy --- by this law now Caelius's youth is sought, not for the punishments of the commonwealth, but for the lusts and dainties of a woman.

\ciceroSection{71} And in this connection too the condemnation of M. Camurtius and C. Caesernius is preached at us. O folly! Is it folly I should call it, or singular shamelessness? Do you dare, when you come from that woman, to make mention of those men? Do you dare to stir up the memory of so great a disgrace --- not extinguished indeed, but suppressed by time? On what charge and what fault did those men perish? On the charge that they followed up that same woman's grief and injury by the wicked debauchery of Vettius. Therefore, that the name of Vettius might be heard in the case, that that old penny-tale might be brought up again, was it for that that the case of Camurtius and Caesernius was renewed? Who, although they were not certainly held by the law on violence, were yet so caught up in that crime, that they did not seem fit to be loosed from the snares of any law.

\ciceroSection{72} But why is M. Caelius being called into this trial? On him no charge proper to this inquisition is brought, nor anything indeed of such a sort that, severed from the law, it is yet joined to your severity. His earliest age was given over to instruction and to those arts by which we are equipped for this forensic use, for taking part in the commonwealth, for office, glory, standing. Of his elders he had such friendships as he most wished to imitate for industry and self-control; and such pursuits among his peers that he seemed to be seeking the same course of praise that the best and noblest sought.

\ciceroSection{73} When then a little strength had been added to his age, he set out for Africa as a tent-mate of Q. Pompeius the proconsul, a most chaste man and most diligent in every duty; in which province there were both his father's affairs and possessions, and also a certain provincial experience not without good cause assigned by our forefathers to that age. He left thence most highly approved by Pompeius's judgement, as you will learn from the man's own testimony. He wished, by ancient custom and by the example of those young men who afterwards stood out in the state as great men and most distinguished citizens, that his industry should be known to the Roman people from some illustrious accusation.

\ciceroSection{74} I could wish his desire of glory had carried him elsewhere; but the time for this complaint has passed. He prosecuted C. Antonius, my colleague, to whom, poor man, the memory of his glorious benefit to the commonwealth was no help, while the opinion of an evil deed planned hurt him. Afterwards he never granted to any of his peers more share than to himself in the Forum, more in business and in his friends' cases, more weight among his own. All these things which only watchful, only sober, only industrious men can attain, he attained by labour and by diligence.

\ciceroSection{75} In this turning, so to speak, of his age --- for I shall hide nothing, relying on your humanity and wisdom --- the young man's reputation paused a moment at the goal-posts on account of his new acquaintance with that woman, and his unhappy neighbourhood, and the unaccustomedness of pleasures which, when they have been shut up too long and pressed and bound in the first age, suddenly sometimes pour out and erupt all together. Out of this life or rather out of this talk --- for it was by no means as great as men were saying --- but out of it, whatever there was, he has emerged, and has thrown himself out and lifted himself out of it; and so far is he from the ill-fame of that familiarity, that now he has even to ward off her enmity and hatred against him.
