% Post Reditum ad Quirites (post-reditum-ad-quirites) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} The thing I prayed for to Jupiter Best and Greatest and to the rest of the immortal gods, citizens, at the time when I devoted myself and my fortunes for your safety, your peace, and your concord --- that, if I had ever set my own reckonings before your safety, I should bear an everlasting punishment willingly taken on; but if both what I had done before I had done for the saving of the state, and undertaken that wretched journey for your safety, so that the hatred which villainous and audacious men had long held conceived against the commonwealth and against all good men, that hatred they should fix upon me alone rather than on every best citizen and on the whole state --- if I had been of this spirit toward you and your children, that at last the memory of me, with pity and longing, should hold you and the senators and all Italy: that I have been proved guilty of this self-devotion, by the judgement of the immortal gods, by the witness of the Senate, by the consent of Italy, by the confession of my enemies, and by your own divine and immortal kindness, I most rejoice.

\ciceroSection{2} For although, citizens, nothing is more to be wished for by a man than a prosperous, even, and continuous fortune in his life's course without a single offence --- still, if everything had been calm and peaceful for me, I should have lacked that incredible and almost divine pleasure of joy which I now enjoy by your kindness. What sweeter thing has been given by nature to the human race than children of one's own? My own indeed, both for my indulgence to them and for the outstanding gift of their nature, are dearer to me than my life: yet they were not received with so much pleasure as they are now restored.

\ciceroSection{3} Nothing was ever sweeter to anyone than my brother to me: I did not feel it so much when I was enjoying him as when I was without him --- and afterwards, when you gave me back to him and him to me. Each man's domestic estate is his delight: my recovered fortunes bring me more pleasure now than they brought when they were untouched. Friendships, daily intimacies, neighbourhoods, ties of clientage, the games at last and the festal days --- what pleasure they had I have understood by losing them more than by enjoying them.

\ciceroSection{4} As for honour, standing, place, order, your kindnesses --- though they always seemed most splendid to me, yet now, restored, they look more brilliant than if they had not been overshadowed. The country itself --- by the immortal gods! --- it can hardly be said what dearness, what pleasure she has: what a sight Italy is, what a multitude of towns, what a shape of regions, what fields, what crops, what beauty of city, what kindness of citizens, what dignity of state, what majesty in you! All of which I had once enjoyed in such a way as no man more; but, as good health is sweeter to those who have been raised up out of a heavy illness than to those who have never had a sick body, so all these things, when they have been longed for, please more than when they have been continually had.

\ciceroSection{5} To what end, then, do I argue this? To what end? That you may understand that no man has ever been of so great eloquence, of a kind of speaking so divine and incredible, that he could not only enlarge or adorn by speaking, but even count off or follow out, the greatness and the multitude of kindnesses you have bestowed on me and on my brother and on our children. From my parents, as was needed, I was begotten small; by you I have been born a man of consular rank. They gave me a brother whom I did not yet know what he would be; you have given him back proven and known to me by incredible devotion. The commonwealth I received in those days when she was almost lost; from you I have recovered her, the same one whom all once judged saved by one man's act. The immortal gods have given me children; you have given them back. Many things besides, prayed for from the immortal gods, we have obtained: had not your good will been there, we should be without all the gifts of the gods. Your offices, finally --- which we had won one by one, step by step --- we now have together from you, so that whatever before I owed to my parents, whatever to the immortal gods, whatever to you yourselves, that whole I owe at this time, all together, to the entire Roman people.

\ciceroSection{6} For while in your kindness itself there is so great a magnitude that I cannot encompass it in speech, in your zealous service the will of your spirits has been so declared that you seem to me not only to have lifted the calamity from me but also to have increased my standing. For not, as for P. Popilius, that most noble man, did young sons and many besides --- kinsmen by blood and by marriage --- entreat for my return; not as for Q. Metellus, that most distinguished man, did a son already of approved years, did the consular L. Diadematus, a man of the highest authority, did C. Metellus the censor, did their children, did Q. Metellus Nepos, who at that time was standing for the consulship, did the sons of his sisters, the Luculli, the Servilii, the Scipiones --- in great numbers Metelli or daughters of Metelli supplicated you and your fathers for the return of Q. Metellus. And if his own highest standing and greatest deeds done had not been enough, the devotion of his son, the prayers of his kindred, the squalor of the youths, the tears of the older men, were able to move the Roman people.

\ciceroSection{7} For C. Marius, who, after those old most distinguished consulars, in the memory of you and of your fathers, was the third consular before me to undergo a most undeserved fortune by his own outstanding glory --- the case was different. He did not return by entreaty: rather, in the secession of the citizens, he called himself back by army and arms. But me, naked of kindred, fortified by no relationship, with no fear of arms or tumult --- the divine and unheard-of authority and courage of C. Piso my son-in-law, and the daily tears and mourning squalor of my most wretched and best brother, entreated for from you.

\ciceroSection{8} There was one brother who could turn your eyes by his own squalor, who by his own weeping could renew the longing for me and the memory of me; who had resolved, citizens, that, if you would not give me back to him, he would undergo the same fortune. He was of so great a love for me that he denied it was right to be separated from me, not only in dwelling-place but even in tomb. For me, when I was present, the Senate and twenty thousand other men changed their dress; for me, when I was absent, you saw the squalor and dirt of one man. He alone, who could be in the Forum, was found to me a son in devotion, a parent in service, the same in love as he had always been --- a brother. For the squalor and grief of my wretched wife and the unceasing mourning of my most upright daughter and the longing for me of my small son and his childish tears were either kept within the necessary travels, or for the most part within the house and in shadow. Therefore your service to us is the greater, because not to a multitude of kindred but to ourselves did you give us back.

\ciceroSection{9} But as my kindred, whom I could not provide myself with, were not at hand to entreat for my calamity, so --- in that thing which my own courage ought to have provided --- helpers, sponsors, and exhorters of my restoration were so many that I far surpassed all who came before me in dignity and abundance. Never about P. Popilius, that most distinguished and bravest man, never about Q. Metellus, that most noble and most steadfast citizen, never about C. Marius, the guardian of the citizenry and of your empire, was mention made in the Senate.

\ciceroSection{10} The first two were restored by tribunician bills, with no authority of the Senate; Marius indeed was restored not only without the Senate but with the Senate suppressed. Nor in the return of C. Marius did the memory of his deeds count, but army and arms; in mine, that the Senate should count, the Senate ever demanded, and that this should at last bear fruit, when first it was permitted, brought it about by its numbers and authority. There was no movement of the towns and colonies in their returns; in mine, all Italy three times by their decrees called me back into our country. They came back with their enemies killed, after a great slaughter of citizens; I, with those by whom I had been thrown out still holding their provinces, my enemy --- on the motion of the consul, the best and gentlest of men, my other consul referring --- was brought back; while that enemy who had lent his voice to my ruin to the common foes of the state was alive only to the extent of his breath, in fact relegated below all the dead.

\ciceroSection{11} Never about P. Popilius did the bravest of consuls L. Opimius, never about Q. Metellus did C. Marius, who was an enemy, or even his successor, M. Antonius, that most eloquent man, with A. Albinus his colleague, exhort the Senate or the people. But on my behalf the previous consuls were always demanded that they should refer; but they were afraid that they might seem to act for favour's sake, because one was my kinsman by marriage, and the other I had defended on a capital charge. Bridled by their bargain over the provinces, they bore through that whole year the complaints of the Senate, the grief of good men, the groan of Italy. But on the Kalends of January, after the bereaved commonwealth had implored the good faith of a consul as her lawful guardian, P. Lentulus, consul, parent, god, salvation of our life, fortune, memory, and name --- as soon as he had referred about the solemn worship of the gods --- judged that there was no human business he should put before mine.

\ciceroSection{12} And on that very day the matter would have been settled, had not that tribune of the plebs, whom I, when consul, had honoured as quaestor with the greatest kindnesses --- when both the whole order and many of the highest men were begging him, and Cn. Oppius his father-in-law, the best of men, was lying weeping at his feet --- demanded a night for deliberation: a deliberation spent not in giving back the bribe (as some thought), but, as it became clear, in increasing it. Afterwards no other matter was acted on in the Senate. When the affair was being held back by various pretexts, with the will of the Senate yet plainly visible, the case was being brought to you in the month of January. Here is the difference between me and my enemies.

\ciceroSection{13} I, when I had seen men being openly enrolled and arranged in centuries on the tribunal of Aurelius; when I understood that the old forces of Catiline had been recalled to the hope of slaughter; when I saw that men of the very party of which I was reckoned among the leaders --- partly because they envied me, partly because they feared for themselves --- were either betrayers or deserters of my safety; when the two consuls, bought by the bargain of the provinces, had given themselves up as supporters to the enemies of the commonwealth; when they saw that their poverty, their greed, their lusts could not be filled, unless they had handed me over bound to the household enemies; when the Senate and the Roman knights were forbidden by edicts and orders to weep on my behalf and, with their dress changed, to entreat you; when all the bargains over the provinces, when all the treaties and reconciliations of favour with all parties were being sealed in my blood; when all good men were not unwilling to perish for me or with me --- I would not fight in arms for my safety, because I thought that both to win and to be conquered would be mournful for the commonwealth.

\ciceroSection{14} But my enemies, in the month of January when my case was being moved, with the bodies of citizens slaughtered, in a flood of blood, thought my return must be cut off. And so, while I was away, you had such a commonwealth that you thought she and I had to be restored on the same terms. As for me, in a state in which the Senate had no force, in which there was every impunity, no courts, where violence and the sword paraded in the Forum, where private men kept themselves up by the protection of their own walls and not of the laws, where the tribunes of the plebs were wounded with you looking on, where the houses of the magistrates were attacked with iron and torches, where the consul's fasces were broken, where the temples of the immortal gods were set on fire --- I judged there was no commonwealth. So neither, when she was driven out, did I think there was a place for me in this city, nor, if she should be restored, did I doubt that she would bring me back along with her.

\ciceroSection{15} Or could I, when I was most certain that P. Lentulus would be consul the next year --- who, in those very most dangerous times for the commonwealth, when he was curule aedile in my consulship, was a sharer in all my counsels and a partner in my dangers --- could I doubt that he, when I had been brought low by consular wounds, would, by consular medicine, bring me back to safety? With him as leader, his colleague --- a most clement and most upstanding man --- not opposing at first and afterwards even helping, almost all the rest of the magistrates were defenders of my safety: among whom T. Annius and P. Sestius, in outstanding spirit, courage, authority, protection, resources, stood forth in distinguished good will and divine zeal toward me; and, with the same P. Lentulus as proposer and the colleague jointly referring, the Senate in the largest numbers, with one dissenter and no one vetoing, adorned my standing in the most lavish words it could and commended my safety to you, to the towns and to all the colonies.

\ciceroSection{16} So me, naked of kindred, fortified by no relationship, the consuls, the praetors, the tribunes of the plebs, the Senate, all Italy ever entreated of you. Finally, all those adorned with your greatest kindnesses and offices, brought before you by the same man, not only exhorted you to save me, but were even authors, witnesses, and praisers of my deeds done. The leader in exhorting you and asking of you was Cn. Pompeius, the man who, of all who are, have been, or will be, is first in courage, wisdom, and glory: who alone, to me alone, his private friend, gave the same things he gave to the whole commonwealth --- safety, peace, dignity. His speech, as I have learned, was in three parts: first he taught you that the commonwealth had been saved by my counsels, and bound my cause together with the common safety, and exhorted you to defend the authority of the Senate, the standing of the state, and the fortunes of a citizen of good deserts; next, in his peroration, he set out that you were being asked by the Senate, asked by the Roman knights, asked by all Italy; finally he himself at the last, on behalf of my safety, did not only ask you but entreated you.

\ciceroSection{17} To this man, citizens, I owe as much as it is hardly right for one man to owe another. Following his counsels, P. Lentulus's opinion, the Senate's authority, you have placed me back in that position to which your kindnesses had raised me, by the same centuries by which you had placed me there. At the same time you heard, from the same place, the highest men, the most distinguished and most ample, the leaders of the state, all the consulars, all the praetors say the same thing: so that, by the testimony of all, it stood fixed that the commonwealth had been saved through me alone. And so, when P. Servilius, that man of the gravest weight and most distinguished citizen, had said that the commonwealth had been handed on safe to the magistrates after me by my own act, the rest gave the same opinion. But you heard at that time not only the authority but even the testimony of L. Gellius, that most distinguished man: who, because he had felt --- to his great peril --- that his own fleet had been almost meddled with, said in your public meeting that, if I had not been consul when I was consul, the commonwealth would have utterly perished.

\ciceroSection{18} Behold, citizens! With so many testimonies, with this authority of the Senate, with so great an agreement of Italy, with so great a zeal of all good men, with P. Lentulus pleading the case, with the rest of the magistrates of one mind, with Cn. Pompeius entreating, with all men favouring me, and finally with the immortal gods themselves approving my return by the abundance, plenty, and cheapness of crops, I, restored to myself, to my own people, and to the commonwealth, shall promise you so much as I am able to perform. First, that the same devotion which the most upright men show toward the immortal gods, the same I shall always show toward the Roman people; and that your divinity shall be in my whole life as weighty and as sacred to me as that of the immortal gods. Next, since the commonwealth itself has brought me back into the citizenry, I shall fail her in no place.

\ciceroSection{19} If anyone supposes that I have been changed in will, or weakened in courage, or broken in spirit, he is greatly in error. What violence and injury and the madness of villainous men could take from me, they have torn off, taken away, scattered: what cannot be removed from a brave man, that all remains and shall remain. I myself saw the bravest of men, my fellow-townsman, C. Marius --- since for us, as if by some fated necessity, war had to be waged, not only with those who would have wiped out these things, but even with fortune --- I saw him, when he was at the very pitch of old age, not only with his spirit not broken because of the greatness of his calamity, but strengthened and made new.

\ciceroSection{20} Whom I myself heard say that he had been miserable then, when he was without that country which he himself had freed from siege, when he heard that his goods were possessed and plundered by his enemies, when he saw his young son the partner of the same calamity, when, sunk in the marshes, by the rushing-up and the pity of the Minturnenses he had kept his body and life, when, taken across in a small ship to Africa, he had come destitute and a suppliant to those upon whom he himself had bestowed kingdoms; but that, having recovered his standing, he would not bring it about that, when those things he had lost had been restored to him, he should not have the courage of spirit which he had never lost. But this is the difference between me and him: that he, by the very thing in which he was strongest, took vengeance on his enemies --- by arms; I shall use what I have been used to, speech: since for that art there is place in war and in sedition, for this in peace and quiet.

\ciceroSection{21} Although he, with an angry spirit, did nothing but take vengeance on his enemies, I shall think about my friends only so much as the commonwealth allows me. Citizens, since four classes of men in all have wronged me --- one of those who, out of hatred for the commonwealth (because, against their wishes, I had saved her), were my bitterest enemies; another of those who, under the pretence of friendship, treacherously betrayed me; a third of those who, since they could not attain the same things by their own slothfulness, envied my praise and my standing; a fourth of those who, although they ought to have been guardians of the commonwealth, sold off my safety, the standing of the state, and the dignity of the command which was in their hands --- so I shall avenge their offences one by one, in the manner I was provoked by each: bad citizens, by managing the commonwealth well; treacherous friends, by trusting them in nothing and watching every move; the envious of my courage and glory, by serving them; the traffickers in provinces, by recalling them home and demanding from them an account of the provinces.

\ciceroSection{22} Although it is more my care, citizens, how I shall return thanks to you who have so well deserved of me, than how I shall pursue the injuries and cruelty of my enemies. For the reckoning of avenging an injury is easier than that of repaying a kindness, because to be superior to wicked men is less of a matter than to be made equal with good men. Then again it is not as needful to repay the badly deserving what one owes them, as it is to repay the best deserving.

\ciceroSection{23} Hatred can be softened by prayers, or set down for the sake of the times of the commonwealth and the common good, or held in check by the difficulty of avenging it, or laid to rest by mere length of time; not to honour the well-deserving, neither is it right to be entreated, nor is one bound in any way to remit it for the commonwealth's sake; nor is there an excuse of difficulty, nor is it fair to fix the memory of a kindness by time and day. Finally, the man who has been more relaxed in vengeance is openly praised for his judgement; but the man who is too slow in repaying so great kindnesses as you have bestowed on me is most heavily blamed, and not only is called ungrateful (which is itself grave), but must even be called impious. And in repaying duty the reckoning is unlike that of money owed: because the man who keeps money has not paid it, and the man who has paid it does not have it; but a kind return both the man who has made it has, and the man who has it has paid.

\ciceroSection{24} Therefore I shall cherish the memory of your kindness with an everlasting good will: not only so long as I shall draw my breath, but, even when life has failed me, those memorials of your service to me shall remain. In repaying gratitude I undertake again, and shall always make good, that, in counsels to be taken on behalf of the commonwealth, my care; in dangers to be warded off from the commonwealth, my spirit; in giving an opinion plainly, my good faith; in offending men's wishes for the sake of the commonwealth, my freedom; in bearing through labour, my industry; and, in increasing your conveniences, the good will of a grateful spirit, shall not fail.

\ciceroSection{25} And this care, citizens, shall be fixed in my spirit forever: that, both with you who hold for me the force and divinity of the immortal gods, and with your descendants and all the nations, I may seem most worthy of that state which has judged, by all its votes, that it could not keep its own dignity unless it had recovered me.
