% De Domo Sua (de-domo-sua) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Pontiffs, while many things have been by divine ordering devised and instituted by our ancestors, nothing is more splendid than that they wished the same men to preside over both the religious worship of the immortal gods and over the highest commonwealth: that the most ample and most distinguished citizens, by managing the commonwealth well, should preserve religion, and by interpreting religion wisely, should preserve the commonwealth. If at any time a great cause has lain in the judgement and power of the priests of the Roman people, this one is so great that the whole standing of the commonwealth, the safety, life, liberty, the altars, hearths, the household gods, the goods, fortunes, dwellings of all citizens seem to have been entrusted and committed to your wisdom, your good faith, and your authority.

\ciceroSection{2} You must today decide whether you prefer hereafter to strip mad and ruined magistrates of the protection of villainous and criminal citizens, or even to arm them with the religious authority of the immortal gods. For if that ruin and flame of the commonwealth shall have defended his own pestilent and deadly tribunate by divine religion --- a tribunate which he cannot defend by human equity --- we shall have to look for other rituals, other priests of the immortal gods, other interpreters of religion. But if, by your authority and wisdom, pontiffs, those things which were done in the commonwealth out of the madness of villainous men --- crushed by some, deserted by others, betrayed by still others --- if they shall be undone, there will be cause why we may rightly and deservedly praise the counsel of our ancestors in choosing the most ample men for the priesthoods.

\ciceroSection{3} But since that madman, if I had reproached anything I have felt during these days in the Senate about the commonwealth, thought that he would have some way of access to your ears, I shall set aside my own order of speaking. I shall reply not to the speech of a furious man --- which he cannot use --- but to his railing, the practice of which he has fortified both with intolerable insolence and with long-standing impunity. And first I demand this of you, vain and furious man: what so great a punishment for your villainies and disgraces is harassing you, that you suppose these men of such standing, who hold up the dignity of the commonwealth not only by their counsels but even by their very appearance, are angry with me because in giving my opinion I joined the safety of the citizens with the honour of Cn. Pompeius? --- and that they will think otherwise on the highest religious question at this time than they thought when I was absent?

\ciceroSection{4} ``You stood higher then,'' he says, ``with the pontiffs; but now, since you have brought yourself before the people, you must necessarily stand lower.'' Is it so indeed? That which is most flawed in an inexpert multitude --- variation, inconstancy, and a frequent change of opinions like the change of weather --- you would carry across to these men, whom gravity holds back from inconstancy, the fixed and defined right of the religious laws from a wilful opinion, the antiquity of precedents, the authority of writings and monuments? ``Are you,'' he says, ``the man whom the Senate could not do without, whom good men mourned, whom the commonwealth longed for, by whose restoration we thought the authority of the Senate restored --- and have you, on first arriving, betrayed it?'' I am not yet speaking of my own opinion: I shall first reply to your shamelessness.

\ciceroSection{5} This man --- you, deadly plague of the commonwealth --- this citizen, whom you forced to give way before the sword and arms and the terror of an army and the villainy of consuls and the threats of the most audacious of men, by the levy of slaves, by the besieging of temples, by the seizure of the Forum, by the suppression of the Curia --- you forced him to leave his home and his country, that good men might not fight with the wicked by the sword --- the citizen whom (you confess it yourself) the Senate longed for, all good men longed for, all Italy longed for, summoned, called back for the saving of the commonwealth: ``but you should not,'' you say, ``have come into the Senate, into the Capitol, on that turbulent day.''

\ciceroSection{6} I did not come, in fact: I kept myself at home as long as the time was turbulent, when it was clear that your slaves --- already long ago made ready by you for the slaughter of good men --- had come to the Capitol with you, with that band of villainous and ruined men of yours under arms. When this was reported to me, you should know that I stayed at home, and gave neither you nor your gladiators a chance for renewing the slaughter. After it was reported to me that the Roman people had come together on the Capitol because of the fear and shortage of grain, but that the agents of your villainies, terrified, had scattered --- some with their swords lost, some with them snatched away --- I came, not only without any forces and without armed men, but even with only a few friends.

\ciceroSection{7} For when P. Lentulus the consul, who had so well deserved of me and of the commonwealth; when Q. Metellus, who, although he had been my enemy and was your brother, had set my safety and standing above our quarrel and your prayers --- when these men were summoning me to the Senate, when so great a multitude of citizens, by their fresh service to me, was calling me by name to repay the favour, was I not to come, especially since you, with your band of fugitives, had clearly already gone away? Here you have even dared to call me, the guardian and defender of the Capitol and of all temples, ``the enemy of the Capitol'' --- because, when the two consuls held the Senate on the Capitol, I came there. Is there any time when it is shameful to have come to the Senate? Or was the matter under discussion such that I ought to have rejected the matter itself and condemned those who were moving it?

\ciceroSection{8} First I say that it is the part of a good senator always to come to the Senate; and I am not in agreement with those who decide, in less good times, not even to come to the Senate at all. They do not understand that this excessive perseverance of theirs has been very welcome and very pleasant to those whose spirits they wished to vex. ``But some,'' he says, ``out of fear, because they thought they were not safe in the Senate, withdrew.'' I do not blame them, nor ask whether anything had to be feared: I think every man should fear by his own judgement. ``Why,'' you ask, ``did I not fear?'' Because it was clear you had gone away. ``Why,'' you ask, ``when some good men thought they could not be safely in the Senate, did I not feel the same?'' ``Why,'' say I, ``when I had thought I could not be safely in the city at all, did they remain?'' Or are others permitted --- and rightly permitted --- in my fear to fear nothing for themselves, while I alone shall be obliged to be afraid both for myself and on behalf of others?

\ciceroSection{9} Or am I to be blamed because in giving my opinion I did not condemn the two consuls? Was it those very men I ought most of all to have damned, by whose law it was brought about that I --- uncondemned and the man best deserving of the commonwealth --- should not bear the punishment of the condemned? Whose very offences, on account of their distinguished good will toward me in saving me, not only I myself but all good men ought to bear --- was I, of all people, restored through them to my old dignity, to repudiate by my opinion their best counsel? But what opinion did I give? First the one which the talk of the people had already fixed in our minds; next the one which had been mooted in the Senate in the previous days; finally the one which the Senate, in great numbers, when it gave me its vote, then followed --- so that neither was the matter raised by me unexpected and fresh, nor, if there is any fault in the opinion, is it greater for him who gave it than for all who approved it.

\ciceroSection{10} ``But the judgement of the Senate was not free, on account of fear.'' If you make those who withdrew to have feared, grant that those who remained did not fear. But if without those who were then absent nothing could have been freely decreed --- when all were present, the matter began to be referred about removing the senatus consultum from the records: there was an outcry from the entire Senate. But I ask in the very opinion --- since I was its prime mover and proposer --- what is to be censured? Was there not cause for taking new counsel? Or was my part in that cause not the principal part? Or was there some other refuge to which we could turn? What greater cause could there be than famine? than sedition? than your counsels and those of yours, who, when an opening was given for stirring up the spirits of the inexperienced, thought you could renew, in the matter of the corn supply, your old deadly banditry?

\ciceroSection{11} Some grain provinces had no grain at all; some had sent it elsewhere, doubtless out of the greed of the sellers; some, that the relief might be the more welcome when they came to the rescue in the very hunger, kept it shut up under their own guards, that they might suddenly send it out anew. The thing was not in any uncertain opinion, but in present and visible danger before our eyes. Nor were we looking ahead by guesswork: we were already seeing it from experience. For when the price of grain was rising, so that now plain shortage and famine, not high price, was being feared, a coming-together took place at the temple of Concord, with the consul Metellus calling the Senate there. If this was due to the genuine grief and hunger of men, surely the consuls were able to take up the cause, and surely the Senate could take some counsel; but if the cause was the corn supply, you certainly were the inflamer and stirrer of sedition --- and was it not for all of us to do this very thing, to take away the matter from your madness?

\ciceroSection{12} What if both were true: that hunger goaded the people, and you, in this sore, like a swelling, stood out --- was not the medicine to be applied so much the greater, that should heal both this native ill and the inflicted one? There was both present scarcity and famine to come. Not enough: a stoning had taken place. If from the people's grief, with no one stirring it up, a great evil; if at P. Clodius's prompting, the usual villainy of a desperate man; if both --- so that the matter itself was such as of itself to stir up the multitude's spirits, and that there were leaders of sedition ready and armed --- does not the very commonwealth seem to have implored both the help of the consul and the good faith of the Senate? But indeed it is clear both were the case. That there was a great difficulty in the corn supply and the highest shortage of grain, so that men no longer feared a long-lasting high price but plain famine, no one denies. That this enemy of peace and quiet was likely to seize this cause for fires, slaughter, plunder --- pontiffs, I do not wish you to suspect, unless you see it.

\ciceroSection{13} Who are the men named openly in the Senate by your brother, the consul Q. Metellus, by whom he said he had been attacked with stones, and even struck? He named L. Sergius and M. Lollius. Who is this Lollius? Who is not even now without iron at your side; who, when you were tribune --- I shall say nothing of me, but who demanded that Cn. Pompeius be killed. Who is Sergius? Catiline's armour-bearer, the close attendant of your person, the standard-bearer of sedition, the inciter of the shop-keepers, condemned of injuries, an assassin, a stoner, a despoiler of the Forum, a besieger of the Curia. With these and such-like leaders, when in the high price of corn you were preparing sudden assaults against the consuls, against the Senate, against the goods and fortunes of the wealthy, under the pretext of the poor and the inexperienced --- when there could be no safety for you in peace --- when, with desperate captains, you had ranks of ruined men decuriated and listed --- was not the Senate to provide that this deadly torch should not catch hold in such great fuel for sedition?

\ciceroSection{14} So there was cause for new counsel: see now whether my part was not all but the principal part. Whom in that stoning was your Sergius naming, whom Lollius, whom the rest of the plagues? Whom did they say should provide the corn supply? Was it not me? What of this --- that nightly running-up-and-down of hired ruffians, set on by you yourself, demanding grain from me --- as though indeed I had presided over the corn supply, or held some grain pressed away, or had any sway at all in such a thing by office or by command? But the man, hanging over slaughter, had given out my name to his hirelings, had thrown it into the minds of the inexperienced. When in the temple of Jupiter Best and Greatest the Senate, in the largest numbers, with that one man dissenting, had decreed about my standing, suddenly on that very day a cheapness of corn unexpected followed the dearest price.

\ciceroSection{15} There were those who --- which I myself feel --- said the immortal gods had approved my return by their will. Some, however, brought the matter back to this reckoning and conjecture: that, since at my return the hope of peace and concord seemed to lie, while at my withdrawal there was the daily fear of sedition --- the price of corn changed, with the fear of war (now almost upon them) thrust away. Because at my return it became hard once again, the corn was demanded of me, since good men kept saying that there would be cheapness when I came. Finally, I was being named, not only by your hired ruffians at your prompting, but, with your forces driven off and scattered, by the entire Roman people, who at that time had come together on the Capitol --- when on that day I was less well, I was being summoned by name to the Senate.

\ciceroSection{16} I came, awaited; many opinions had already been given when I was asked for mine. I gave one most healthy for the commonwealth, and necessary for me. Plenty of grain, cheapness of corn, was being demanded of me; whether or not I might have any power in the matter was not being considered. I was being demanded by the entreaty of good men: I could not bear the railing of the wicked. I delegated to a more substantial friend --- not because I should have laid that burden on him, who had so deserved of me (for I myself should rather have given way), but because I saw that what we should all promise about Cn. Pompeius, that man would most easily bring through, by his good faith, by judgement, by courage, by authority, and last by his good fortune.

\ciceroSection{17} And so, whether the immortal gods grant this fruit of my return to the Roman people --- that, just as at my withdrawal there had been shortage of crops, hunger, devastation, slaughter, fires, plunder, the impunity of villainies, flight, terror, discord; so at my return the abundance of fields, plenty of crops, hope of peace, calm of spirits, courts, laws, concord of the people, the authority of the Senate seem to have been brought back together with me --- or whether I myself had to provide something by my arrival, by counsel, by authority, by care, in return for so great a kindness from the Roman people: I make it good, I promise it, I undertake --- I say nothing further, what is enough for the present I say --- the commonwealth shall not, on the score of corn supply, come into the danger to which it was being called.

\ciceroSection{18} Surely, then, in this duty which fell to me especially, my opinion is not to be censured? That the matter was a very great one, and of the highest peril, not only from famine but from slaughter, fires, devastation --- no one denies; when to the cause of the high price was added that scout of the common miseries, who has always lit the torches of his own villainy from the ills of the commonwealth. He says nothing should have been decreed for one man out of the regular order. I shall not now answer you as I would the rest --- that to Cn. Pompeius were committed, out of the regular order, very many wars, the most dangerous, the greatest, by sea and by land: of which, if anyone repents, he repents of the victory of the Roman people.

\ciceroSection{19} It is not so I argue with you. With those men this speech of mine could be conducted, who argue thus: that, if any matter must be referred to one man, they would refer it most particularly to Cn. Pompeius; but that they will give nothing extraordinary to anyone; while, since it has been given to Pompey, they are accustomed to adorn and uphold it as becoming his standing. From praising their opinion I am held back by Cn. Pompey's triumphs --- by which he, when he had been called out of the regular order to defend his country, increased the name of the Roman people and adorned the empire. Their consistency I approve --- a consistency which I too had to use, when on his motion he waged war out of the regular order with Mithridates and Tigranes.

\ciceroSection{20} But with them I can argue something. As for you, what is your shamelessness so great that you dare say that nothing should be given to anyone out of the regular order? --- you who, by a wicked law, when you had publicly confiscated, without his cause heard, Ptolemy, king of Cyprus, brother of the king of Alexandria who reigned by the same right; when you had bound the Roman people to villainy, when you had let loose the patronage of this empire onto the kingdom, the goods, the fortunes of a man with whose father and grandfather and ancestors there had been alliance and friendship with us --- you set M. Cato in charge of bringing his money home, and (if Ptolemy should defend his right) of waging the war.

\ciceroSection{21} You will say: ``What a man! the most upright, the most prudent, the bravest, the most loving of the commonwealth, of admirable virtue, judgement, and rule of life --- almost without an equal in earning praise!'' But what has that to do with you, who deny it true that any man should be set out of the regular order over any matter of the commonwealth? It is in this only I rebuke your inconsistency: the man whom in this affair you would not be putting forward for his own dignity, but smuggling away for your own villainy --- whom you had thrown to your Sergii, Lollii, Titii, and the rest of the leaders of slaughter and fire, whom you had said was the executioner of citizens, the principal of the killing of the uncondemned, the author of cruelty --- to him you have brought, by name, this honour and this command out of the regular order, in your own bill. And you were of such intemperance that you were unable to hide the reckoning of your own villainy:

\ciceroSection{22} you read out in a public meeting a letter which you said had been sent to you by C. Caesar --- ``Caesar to Pulcher'' --- when you even pleaded as a sign of love this --- that he used only the cognomina, and did not write \textit{proconsul} or \textit{tribune of the plebs} --- and then that he congratulated you for having removed M. Cato from your tribunate, and for having taken from him the freedom of speaking thereafter on extraordinary commands. Either he never sent you that letter; or, if he did send it, did not wish it to be read in a public meeting. But, whether he sent it or you forged it, certainly your design about Cato's office was made open by the recital of that letter.

\ciceroSection{23} But I leave aside Cato, whose outstanding courage, dignity, and (in that business which he conducted) good faith and self-restraint seem to cover the wickedness of your law and your action. What of this? --- Who gave to that man, the most disgraceful, the most criminal, the most contaminated since men were born, that opulent and fertile Syria, who war with the most pacified peoples, who that money set aside for buying lands, snatched out of the bowels of the treasury, who an unbounded command? To whom indeed, when you had given Cilicia, you changed the bargain and transferred Cilicia to the praetor, again out of the regular order: you gave Syria by name to Gabinius, with the price made bigger. What of this? --- to that most foul, most cruel, most deceitful man, marked with all the stains of every villainy and lust, L. Piso, did you not by name hand over free peoples (freed by many senatus consulta, even by a recent law of his own son-in-law) bound and tied up? When you had been paid the wages of your kindness and the price of the province with my blood, did you not still divide the treasury with him?

\ciceroSection{24} Indeed? You undid the consular provinces (which C. Gracchus, who alone was most popular, did not take away from the Senate but even sanctioned by law that they had to be assigned by the Senate every year), the same provinces --- decreed by the Senate by the lex Sempronia --- you undid, gave by name out of the regular order without the lot, not to consuls but to plagues of the commonwealth. And shall I be reproved by you because, in the greatest matter, almost despaired of, for the highest of men, often chosen for the gravest dangers of the commonwealth, we have often by name set him in charge? What then? If those measures which you, in those darknesses and blind clouds and storms of the commonwealth, when you had thrown the Senate down from the helm and shoved the people out of the ship, while you yourself, the arch-pirate with a flock of foulest brigands, were sailing under full canvas --- if those measures which you then promulgated, settled, promised, sold, you had been able to bring through, would there have been any place in the world free of extraordinary fasces and the Clodian command?

\ciceroSection{25} But at last roused --- I shall say what I felt and feel in his hearing, however he means to hear it --- at last roused, I say, the grief of Cn. Pompeius's spirit, hidden and buried too long deep within him, suddenly came to the aid of the commonwealth, and lifted up our state, broken with evils, lessened and weakened, cast down with fear, to some hope of liberty and of its old standing. Was this man, of all men, not to be set out of the regular order in charge of the corn supply? Yes, of course --- you handed over by your law, to that most foul gourmandiser, the foretaster of your lusts, the most beggared and most criminal man, Sex. Clodius, your kinsman by blood, who by his own tongue alienated even your own sister from you, all the grain, private and public, all the corn provinces, all the contractors, all the keys of the granaries: by which law first there was a high price, then shortage. Famine was hanging over us, fires, slaughter, plunder; your madness was hanging over the goods and fortunes of all.

\ciceroSection{26} The unseasonable plague even complains, out of the most filthy mouth of Sex. Clodius, that the corn supply has been snatched from him --- and that, in the highest of dangers, the commonwealth has implored the help of that man whose hand he should remember to have often saved and increased it! ``Nothing pleases Clodius to be carried out of the regular order!'' What of the law about me, which you say you carried, parricide, fratricide, sister-slayer? Did you not carry it out of the regular order? Or, about the destruction of a citizen --- who, as all gods and men have now judged, is the saver of the commonwealth, but who, as you yourself confess, was not only uncondemned but not even charged --- it was permitted you to carry not a law but a wicked privilegium, with the Senate mourning, with all good men in grief, with the prayers of all Italy spurned, with the commonwealth crushed and captive: but for me, with the Roman people imploring it, with the Senate demanding it, with the times of the commonwealth crying out for it --- it has not been permitted to give an opinion on the safety of the Roman people?

\ciceroSection{27} Indeed, in that opinion, if Cn. Pompey's standing has been increased and joined with the common good, surely I should be praised, if I should be seen to vote for the standing of the man who had brought help and aid to my safety. Let men cease, let them cease, to hope that I, restored, can be undermined by the same engines by which they earlier broke me when I was standing. For what pair of consular friendship has there ever been in this state more closely bound together than that of me and Cn. Pompeius? Who has more brilliantly spoken about his standing before the Roman people? who more often in the Senate? What was so great a labour, what a quarrel, what a contention, that I have not undertaken for his standing? What honour from him to me, what proclamation of my praise, what return of good will has he passed by? This our union ---

\ciceroSection{28} this conspiracy in managing the commonwealth well, this most pleasant partnership of life and of all duties, certain men by feigned conversations and false charges have torn apart, while the same men were warning him to fear me and to beware of me, and were saying to me that he was the most hostile of all to me alone --- so that I could not boldly enough ask of him what I should ask, and he, vexed with the suspicions and the villainy of certain men, did not promise to me as broadly as my time required.

\ciceroSection{29} A great penalty has been paid for my error, pontiffs, so that I am not only sorry of my folly but ashamed of it: I, when not some sudden moment of mine but old labours undertaken and foreseen long before had bound me with the bravest and most distinguished man, allowed myself to be torn from such a friendship, and did not understand which men I should withstand as open enemies, or whom not to trust as treacherous friends. So let men cease from time to time to swell me up with the same words: ``What does that man want? Does he not know how strong his authority is, what he has done, with what dignity he has been restored? Why is he honouring the man by whom he was deserted?''

\ciceroSection{30} I do not in fact think I was then deserted, but almost given up; nor do I think it should be made plain by me what was done in that flame of the commonwealth against me, or how, or by whom. If it was useful to the commonwealth that I, alone for all, should drink down that most undeserving calamity, this too is useful: that I should hide and be silent about whose villainy stoked the matter. But it is the part of an ungrateful man to be silent about this; therefore I shall most willingly proclaim that Cn. Pompeius, in zeal and authority, equally as each one of you, in resources, contention, prayers, and at last in dangers, laboured especially for my safety. He, P. Lentulus, while you thought of nothing day and night but my safety, took part in all your counsels; he was the gravest counsellor for setting the matter in motion, the most faithful partner for preparing it, the bravest helper for completing it; he went to the towns and colonies; he implored the help of all Italy, eager to help; he was the first in the Senate to give an opinion; and after he had given an opinion, he himself entreated the Roman people for my safety.

\ciceroSection{31} So you may set aside that speech of yours, that, after the opinion I gave on the corn supply, the spirits of the pontiffs were changed --- as if these men had a different feeling about Cn. Pompeius from mine, or did not know what I had to do for the expectation of the Roman people, for Cn. Pompeius's services to me, for the reckoning of my own time --- or even, if my opinion had perhaps offended the spirit of any pontiff (which I know to be otherwise), as if he would settle the matter differently, either as a pontiff on a question of religion or as a citizen on a question of the commonwealth, than the law of the rituals or the safety of the state should compel.

\ciceroSection{32} I understand, pontiffs, that I have said more outside the case than either common opinion or my own will required. But while I longed to clear myself before you, your good will in listening attentively to me has carried my speech further. I shall make this up by the brevity of the speech which belongs to the case itself and to your judgement: which, since it is divided into the law of religion and the law of the commonwealth, the part of religion (which is much more wordy) I shall pass over, and shall speak about the law of the commonwealth.

\ciceroSection{33} For what is either so arrogant as to try to teach the college of pontiffs about religion, about divine matters, ceremonies, sacred things; or so foolish as, if any man finds something in your books, to tell it to you; or so meddling as to wish to know those things which our ancestors wished you alone to be both consulted on and to know? I deny that, by public law, by the laws this state uses, any citizen could be afflicted with such a calamity without trial. This right, in this state, even when there were kings, I say there was; this has been handed to us by our ancestors; this, finally, is the proper mark of a free state: that nothing of a citizen's life, or of his goods, can be torn away without the judgement of the Senate or of the people, or of those who are appointed judges in each matter.

\ciceroSection{34} Do you see that I do not pull up by the roots all your acts, nor pursue what is plain --- that you did nothing at all by right, that you were not tribune of the plebs, that today you are a patrician? I am speaking before pontiffs, and the augurs are present. I am moving in the middle of public law. What is, pontiffs, the law of adoption? Surely that he should adopt who can no longer beget children of his own, and who, when he could, has tried. Then what each man's cause is for adopting, what the reckoning of birth and dignity, what of sacred rites, the college of pontiffs is accustomed to investigate. What of these things has been investigated in that adoption? A man twenty years old, even less, adopts a senator. For the sake of children? But he can beget; he has a wife; he will get children from her; the father, then, will disinherit his own son?

\ciceroSection{35} What of this --- why are the rites of the Clodian gens dying out, so far as it is in you? Of which all examination should have lain with the pontiffs when you were being adopted: unless perhaps you were asked thus --- whether you wished to disturb the commonwealth by seditions, and to be adopted for that reason --- not so that you should become his son, but so that you should be made tribune of the plebs and overturn the state from the foundations. You answered, I take it, that this is what you wanted. The pontiffs thought it a good cause: they approved it. The age of the adopter was not investigated, as in the case of Cn. Aufidius and M. Pupius --- of whom each in our memory, in extreme old age, the one adopted Orestes, the other Piso; which adoptions, like countless others, were followed by inheritances of name, of money, of sacred rites. You are neither a Fonteius (which you ought to have been), nor your father's heir; nor, with the paternal sacred rites lost, did you come into these adoptive ones. Thus, with rites overturned, with two gentes contaminated --- the one you deserted, the one you polluted --- with the lawful right of guardianship and inheritance under the law of the Quirites left behind, you have become, contrary to right, the son of the man who, in respect of his age, could have been your father.

\ciceroSection{36} I am speaking before pontiffs: I deny that this adoption of yours has been performed by pontifical right. First because your two ages are such that the man who adopted you might either, by his age, have been to you in the place of a son, or in the position he was in: next, because the cause of adopting is normally investigated --- that he should adopt who, what nature can no longer give him, may seek by lawful and pontifical right; and so adopt that there be no diminution either of the dignity of the lines or of the religion of the rites; and especially this: that no false claim, no fraud, no trickery, be brought into it: that this feigned adoption of a son may seem to imitate, as much as possible, the truth of begetting children.

\ciceroSection{37} What is a greater false claim than this --- that a beardless boy, in good health and married, should come and say that he wishes a senator of the Roman people to adopt him as his son; while everyone knows and sees that he is not being adopted that he may be settled as a son, but that he may pass out of the patricians and be able to become tribune of the plebs? Nor is this hidden: for the adopted is at once emancipated, that he may not be the son of the man who adopted him. Why then was he adopting? Approve this kind of adoption: now all the rites will perish whose guardians you ought to be, now no man will be left a patrician. For why should anyone wish that he should not be allowed to be made tribune of the plebs --- that the canvass of the consulship be narrower for him, that, when he could come into a priesthood, he should not come, because that office is not for a patrician? As soon as anything happens to anyone, by reason of which it is more convenient to be a plebeian, on a similar reckoning he will be adopted.

\ciceroSection{38} So in a short time the Roman people will have neither rex sacrorum, nor flamines, nor Salii, nor half the rest of its priesthoods, nor proposers of the centuriate and curiate assemblies; and the auspices of the Roman people, if patrician magistrates are not created, must perish, since there will be no interrex --- which thing requires both that the man should be a patrician himself and be put forward by the patricians. I have said before pontiffs that this adoption, approved by no decree of this college, performed against all pontifical right, is to be held for nothing; on which being removed, you understand that the whole of your tribunate has fallen.

\ciceroSection{39} I come to the augurs. Their books, if any are hidden away, I do not poke into. I am not too curious in seeking out the augurs' law. The things which I learned together with the people, which have often been answered in public meetings, I know. They deny it lawful that anything be done with the people when the heaven has been watched. On the day on which the curiate law about you is said to have been carried, do you dare deny that the heaven was being watched? A man here present, gifted with singular courage, steadfastness, and gravity --- M. Bibulus --- I maintain that on that very day the consul watched the heaven. Are you, then, weakening the acts of C. Caesar, that bravest of men? Not at all: nothing of mine matters now, except those weapons that out of his actions have been thrown into my body.

\ciceroSection{40} But these matters about the auspices, which I now touch on most briefly, were yours. You, with your tribunate already running off the rails and weakened, suddenly stood out as the patron of the auspices. You brought M. Bibulus into a public meeting, you brought out the augurs. To you the augurs answered, when you asked, that, when the heaven had been watched, business could not be done with the people; to you M. Bibulus, when you asked, replied that he had watched the heaven, and the same he said in a public meeting, when brought out by your brother Appius, that you, since you had been adopted against the auspices, had not been tribune at all. All your action in the later months was that whatever C. Caesar had done, since it had been done against the auspices, ought to be undone by the Senate; and that, if this should be done, you said you would carry me, the guardian of the city, back into the city on your own shoulders. See the man's madness: by his own tribunate he was held bound up with Caesar's acts.

\ciceroSection{41} If both pontiffs by the law of the rites and augurs by the religion of the auspices undo your whole tribunate, what more do you ask? Or some still plainer right of the people and of the laws? At about the sixth hour of the day I made some complaint in court, when I was defending C. Antonius my colleague, on the commonwealth, which seemed to me pertinent to that wretched man's cause. Wicked men carried these things to certain brave men far otherwise than they had been said by me. At the ninth hour of that very day you were adopted. If, in the case of other laws, what should be the three nundinae --- in the case of an adoption, three hours is enough --- I do not censure it. But if the same things should be observed, the Senate has judged that the people are not bound by the laws of M. Drusus, which were carried against the lex Caecilia and Didia.

\ciceroSection{42} You see now that, by every kind of law --- the law of sacred rites, the law of auspices, the law of statutes --- you were not tribune of the plebs. But I leave this whole matter aside not without cause. For I see certain most distinguished men, leaders of the state, in several cases, judging that you could lawfully transact business with the plebs --- who, even about me, when they said that the commonwealth had been carried out for burial by your bill, still said that the funeral, however wretched and bitter, had been duly proclaimed: that the bill which you had carried about a citizen so deserving of the commonwealth, you had proclaimed a funeral for the commonwealth, and that what you had carried with the auspices safe, you had done by right. So I shall be allowed, I think, not to weaken those proceedings by which they recognised your tribunate as duly constituted.

\ciceroSection{43} Suppose, then, that you were tribune of the plebs as legitimately as this very man P. Servilius --- a man most distinguished and ample in all things. By what right, by what custom, by what precedent, did you carry a law by name about the head of an uncondemned citizen? The sacred laws forbid it; the Twelve Tables forbid laws to be sworn against private men: this is what a privilegium is. No one ever has carried it. There is nothing more cruel, nothing more pernicious, nothing this state can less endure. The most wretched name of proscription, and all the bitterness of the Sullan time --- what does it have that is most marked for the memory of cruelty? I think, the punishment fixed against Roman citizens by name without trial.

\ciceroSection{44} Will you, then, pontiffs, by your judgement and authority give a tribune of the plebs this power, that he may proscribe whom he will? For I ask: what else is it, unless ``you would and order that M. Tullius shall not be in the citizenry, and that his goods shall be mine''? For so, in fact (though in different words), he carried it. Is this a plebiscite? this a law? this a bill? Can you bear this? can the state bear it --- that single citizens be removed from the citizenry by single little verses? I, indeed, have already gone through it. I fear no violence, no assault. I have filled up the spirits of the envious, I have appeased the hatreds of the wicked, I have even sated the perfidy and villainy of the betrayers; finally about my case, which seemed set up by the ruined citizens for hatred, all cities, all orders, all gods and men have now judged.

\ciceroSection{45} You yourselves, pontiffs, your children, and the rest of the citizens by your authority and wisdom must take counsel. For when so moderate are the judgements of the people set up by our ancestors --- first that capital punishment should not be combined with money penalty; next that no one should be charged on a day not appointed; that the magistrate should three times accuse, with a day intervening, before he names a fine or judges; that the fourth accusation should be on the day proclaimed three nundinae before, on which day the trial shall be; then that many things have been allowed for the appeasing of the judges and for the pity of the defendants; then that the people are to be entreated, the suffrage easily given for safety; finally, even, that, if any matter takes that day from him --- by auspices or by excuse --- the whole case and trial is removed: when these things are so in law, where there is a charge, an accuser, witnesses --- what is more shameful than that, when the man has not been ordered to be present, has not been summoned, has not been charged, hired men and assassins and beggared and ruined men should give a vote about his life, about his children, about all his fortunes --- and that this should be considered law?

\ciceroSection{46} And if this could be done about me, whom honour, dignity, the cause itself, the commonwealth defended; whose money was not even being demanded; against whom there stood nothing except the shifting of the times and the sloping of the common moment --- what then will be the case for those whose life is removed from popular office and from this brilliant favour, but whose fortunes are so great that too many beggared, too many spendthrift nobles set their hearts on them?

\ciceroSection{47} Give this licence to a tribune of the plebs, and look a little while into your spirits at the youth, especially those who already seem with desire to hover over the tribunician power. Whole colleges of tribunes of the plebs, by my faith, will be found, with this right confirmed, that come together about the goods of the wealthiest of men, especially with popular plunder and the hope of largess set out for them. But what did the skilled and crafty drafter of laws set down? ``Would you and order that to M. Tullius water and fire be denied?'' --- Cruel, wicked, not to be carried even against the most criminal citizen without trial! He did not carry that he be denied. What then? ``That he have been denied.'' O filth, O monster, O villainy! This law for you Clodius wrote, more foul than his own tongue: that a man be denied who has not been denied? Sextus, my friend, by your good leave, since you are now a logician and a nibbler at these subtleties --- can what has not been done, that it be done, be brought before the people, or be sanctioned by any words, or be confirmed by any votes?

\ciceroSection{48} With this drafter, this counsellor, this servant --- the most foul of men, indeed of all things both two-footed and four-footed, Sex. Clodius --- you destroyed the commonwealth. Nor were you so dull and so mad as not to know that there was a Clodius who acted against the laws and others who were used to draft laws. But neither of those, nor of the rest, in whom there was any moderation, was anyone in your power. Neither could you use the same drafters of laws as the rest, nor the same architects of works, nor bring in any pontiff you wished. Finally, not even in the partnership of plunder could you find a contractor or a surety outside the number of your gladiators --- not even, in your own proscription, anyone to give a vote, except a thief and an assassin.

\ciceroSection{49} And so, when you, flourishing and powerful, were flying through the middle of the Forum like a popular harlot, your friends --- happy and protected by your friendship alone --- who had committed themselves to the people, were so beaten back that they even lost your tribe, the Palatina; those who had come into court --- whether as accusers or as defendants --- with you praying for them, were condemned. Finally, even that newcomer Ligus, your venal under-signer and counter-signer, when by the testament and judgement of his brother M. Papirius he had been disowned, said he wished to pursue his death: he laid charge against Sex. Propertius --- but, partner of another man's tyranny and villainy, he did not dare to accuse, for fear of the calumnia.

\ciceroSection{50} Is this the law of which we are speaking, as if it seemed lawfully proposed --- whose every part each man touched (with finger, with voice, with plunder, with vote), wherever he came was driven off and convicted? What if the proscription is so written that it dissolves itself? For it is: ``Because M. Tullius referred a false senatus consultum.'' If, then, he referred a false senatus consultum, then there is a bill: if he did not refer one, there is none. Does it not seem to you sufficiently judged by the Senate that I had not only not falsified the authority of that order, but had been the one who, since the foundation of the city, had most carefully obeyed the Senate? In how many ways do I show that this thing of yours, which you call a law, is not a law? What of this --- that even on more matters than one, with one drawer of lots, you carried a bill: do you still suppose that what M. Drusus, that very good man, with M. Scaurus and L. Crassus as advisers, in most of his laws did not obtain --- can you, the man of every villainy and lust, with decimi and Clodii as your sponsors, obtain?

\ciceroSection{51} You carried a law about me, that I should not be received back --- not that I should leave; for you yourself could not say that it was not lawful for me to be at Rome. For what would you say? That I had been condemned? Surely not. Driven out? By what process? But not even this is written --- that I should leave. There is a penalty for whoever shall have received me, which all men ignored; expulsion is nowhere set down. But suppose it were so. What of this --- the demolition of public works? What of the inscribing of one's name? Does this seem to you anything other than a plunder of my goods? Apart from this, you could not even, under the lex Licinia, propose for yourself the charge of the work. What of this very thing which you are now arguing before pontiffs --- that you have consecrated my house, that you have built a monument on my premises, that you have dedicated a statue, and that you have done these things by one little bill: does this seem to be one and the same with what you carried by name about me?

\ciceroSection{52} It is one, by Hercules, just as it is one and the same that you carried by one law that the king of Cyprus, whose ancestors were always our allies and friends, should with all his goods be put up to the auctioneer, and that the Byzantine exiles should be brought back. ``I gave the same man the business in both matters,'' he says. What if he had given the same man the business of demanding back the cistophorus in Asia, of going from there to Spain, of, when he had returned to Rome, being allowed to canvass for the consulship, of, when he had been made consul, holding the province of Syria --- since he was writing about one man, would it be one matter?

\ciceroSection{53} If the Roman people had now been consulted on this matter, and you had not done everything through slaves and brigands, could it not have been that the people should have approved about the king of Cyprus and disapproved about the Byzantine exiles? What is, I ask, the other force, what the other meaning of the lex Caecilia and Didia, except this: that the people should not be compelled, in joined matters of many kinds, either to accept what they do not want or to reject what they want? What if you carried it by violence, is it then a law? Or can anything be considered done by right, when it is agreed it was done by violence? Or, if at the very passing of your law, with the city already taken, stones were thrown, but no hand-to-hand fight took place, were you on that account able to come at that ruin and inundation of the citizenry without the highest violence?

\ciceroSection{54} When on the tribunal of Aurelius you were openly enrolling not only freemen but even slaves, stirred up out of every neighbourhood --- you were not, I suppose, then preparing violence! When by your edicts you were ordering the shops to close, you were seeking not the violence of an inexpert multitude but the moderation and prudence of upright men. When you were carrying arms into the temple of Castor, you were planning nothing but that nothing might be done by violence. When indeed you tore up and removed the steps of the temple of Castor, then it was that, that you might be able to act with moderation, you drove audacious men back from the entrance and the ascent of that temple. When you ordered those men to be present, who in a meeting of good men had spoken about my safety, and broke up their advocacy by hands, by iron, by stones --- then indeed you showed that violence chiefly displeased you.

\ciceroSection{55} Yet this furious violence of an insane tribune of the plebs could easily have been overcome and broken by the courage or the multitude of good men. What of this --- when Syria was being given to Gabinius, Macedonia to Piso, to each an unbounded command, vast money, that they might allow you everything, help you, prepare for you a hand, troops, their picked centurions, money, slave households, and lift you up by their criminal public meetings, mock the authority of the Senate, threaten the Roman knights with death and proscription, terrify me with menaces, announce slaughter and combat against me, fill my house, full of good men, with the fear of proscription through their friends, strip me of the throng of good men, despoil me of the protection of the Senate, prevent the most ample order from not only fighting on my behalf but even from weeping and supplicating with their dress changed --- not even then was there violence?

\ciceroSection{56} Why, then, did I yield, or what fear was there? I will not say of myself: grant me to be timid by nature. What of those many thousands of bravest men? what of our Roman knights? what of the Senate? what of all good men, finally? If there was no violence, why did they accompany me weeping rather than either rebuke and hold me back, or in anger leave me? Or did I fear this --- that, if proceedings against me were begun by the custom and institution of our ancestors, I should not be able to stand them face to face?

\ciceroSection{57} Was it the trial I had to fear, if a day had been named --- or, without trial, the privilegium? The trial? My case was so disgraceful, of course; I, the man who, even were the case unknown, could not unfold it by speaking! Or because I could not prove the case? Whose merit is so great that, of itself, it commended not only itself, but even me when absent. Or would the Senate, would all the orders, would those who flew together to me from all Italy to bring me back, have been more sluggish, when I was present, in keeping me here and saving me, in a cause which the parricide himself now says was such that he complains of my having been awaited and called back, by all, to my old standing?

\ciceroSection{58} Or was there no danger in a trial: I feared the privilegium, that, if a fine were sworn against me when I was present, no one would step in? Was I so destitute of friends, or the commonwealth so naked of magistrates? What of this --- if the tribes had been called, would they have approved a proscription, I will not say against me, who had so deserved of their safety, but against any citizen at all? Or, if I had been present, would those old forces of the conspirators, your ruined and beggared soldiery, and the new band of the most criminal consuls, have spared my body? --- when, although I yielded to the cruelty and villainy of all of them, not even in my absence could I sate their minds with my grief.

\ciceroSection{59} For what wrong had my poor wife done you, whom you harassed, dragged about, tore at with every cruelty? what my daughter, whose unceasing weeping and mourning squalor was sweet to you, but turned the minds and eyes of all others? what my small son --- who, while I was away, was seen by no man except weeping and worn out: what had he done that you so often wished to kill him by ambush? what my brother? --- who, having come from his province some time after my departure and not thinking he should live unless I were restored, when his grief, his squalor incredible and unheard-of, seemed pitiable to all mortals --- how often did he slip out of your iron and your hands!

\ciceroSection{60} But why do I reproach the cruelty you used against me and mine, you who against my walls, my roof, my columns, my doorposts have made a kind of foreigner's war, an unspeakable war steeped in every hatred? For I do not think you, when after my departure you had swallowed up in hope and greed the fortunes of all the wealthy, the revenues of all the provinces, the goods of tetrarchs and kings, were blinded by greed for my silver and my furniture. I do not believe that the Campanian consul, with his dancer-colleague --- when to one of them you had bestowed all of Achaia, Thessaly, Boeotia, Greece, Macedonia, and all barbarian lands, the goods of Roman citizens; to the other Syria, Babylon, the Persians, the most undisturbed and most peaceful nations, to be plundered --- were that greedy for my thresholds and columns and doors.

\ciceroSection{61} Nor again did that hand and forces of Catiline think they would fill up their hunger with the rubble and shards of my roofs. But, as we are accustomed to raze enemy cities --- not of all enemies, but of those with whom we have undertaken a bitter and exterminating war --- not led by plunder, but by hatred --- because, against those whose cruelty had set our minds aflame, some war always seems to settle in their very houses and seats ---

\ciceroSection{62} Nothing had been carried about me; I had not been ordered to be present; I had not been summoned and was absent; I was, even by your judgement, a citizen safe, when my house on the Palatine, my villa at Tusculum, were being transferred --- one to one consul, the other to the other (consuls indeed they called them!) --- when marble columns out of my house, with the Roman people looking on, were being carried to the consul's mother-in-law; into the field of the next-door consul not the equipment or ornaments of the villa, but even the trees were being moved, with the villa itself, not from greed for plunder --- for what was there of plunder? --- but from hatred and cruelty, being utterly destroyed. The house was burning on the Palatine, not by chance but by deliberate firing; the consuls were feasting and were busy with the congratulations of the conspirators, when one said he had been Catiline's darling, the other Cethegus's cousin.

\ciceroSection{63} This violence, pontiffs, this villainy, this fury, by setting my body in the way I drove away from the necks of all good men, and the whole assault of discord, the whole long-gathered violence of the wicked, which now grown rotten in suppressed and silent hatred was bursting out, having found such audacious leaders, I caught with my body. Against me alone the consular firebrands were thrown by tribunician hands; against me clung all those wicked weapons of conspiracy which I once had beaten back. If, as many of the bravest men thought best, I had wished to fight with violence and arms against violence, either I should have conquered with a great slaughter of the wicked, but yet of citizens; or, with all good men killed (which would have been the dearest wish of those men), I should have fallen along with the commonwealth.

\ciceroSection{64} I was seeing that, with the Senate and Roman people alive, my return at the highest standing would be quick; nor was I understanding that for any longer time it could happen that I should not be allowed to be in that commonwealth which I had myself saved. If it were not allowed, I had heard and read that the most distinguished men of our state had thrown themselves among the enemies, to a plain death, for the safety of the army. Should I, for the safety of the whole commonwealth, hesitate, in this better situation than the Decii --- that they could not even hear of their own glory, while I could even be a spectator of my own praise? And so, broken, your madness made empty assaults; for the bitterness of my fall had taken in the entire violence of all the wicked: in such monstrous injury and in such ruins there was no room for new cruelty.

\ciceroSection{65} Cato had been next. What were you to do? It was not possible that the same measure as the wrong should be that of his standing. What could you do? Thrust him out to the Cyprian money? The plunder is lost. Plunder there will not be wanting from elsewhere; for the present he must be sent away. Thus M. Cato, hated, is, as if by way of kindness, relegated to Cyprus. Two are cast out, whom the wicked could not bear to see: one by a most disgraceful honour, the other by a most honourable calamity.

\ciceroSection{66} And that you may know that this man has always been the enemy not of men but of virtues: with me cast out, with Cato sent away, he turned upon the very man on whose authority and help in his public meetings he was saying, of all the things he was doing and had done, that he had done and was doing them all --- Cn. Pompeius, whom in everyone's judgement he saw to be far the leading man of the state. He did not think Pompey would any longer indulge his madness. When he had stolen, by the treachery of the king's slaves, the captive son of a friendly king out of his keeping, and by that injury had provoked the bravest of men, he hoped he could fight with him with the same forces by which I had been unwilling to do battle at the cost of good men --- and at first with the consuls helping him; afterwards Gabinius broke the bargain, but Piso remained in good faith.

\ciceroSection{67} What slaughters, what stonings, what flights this man then made; how easily, with iron and daily ambushes, when he had been left even by the firmest core of his own forces, he stripped Cn. Pompeius of the Forum and the Curia and shut him up in his house, you saw: from which you can judge how great that violence was at its rising and gathering, when this remnant, now scattered and put out, terrified Cn. Pompeius.

\ciceroSection{68} This a most prudent man, the most loving of the commonwealth, of me, and even of the truth itself, L. Cotta, saw in giving his opinion on the Kalends of January, when he held that no law should be carried about my return: who said that I had taken thought for the commonwealth, that I had yielded to the storm, that I had stood out as more friend to you and to the rest of the citizens than to myself, that I had been driven out by violence, by arms, by the dissensions of men and by slaughter set up and by a new tyranny; that nothing could have been carried about my life, that nothing was lawfully written or could have force, but that everything had been done against the laws and the custom of our ancestors, rashly, in turbulence, by violence, by madness. And if this was a law, neither was it lawful for the consuls to refer the matter to the Senate, nor for him to give an opinion: but since both were being done, he held that no law should be carried about me, lest that thing which was nothing should be judged a law. No opinion could have been truer, weightier, better, more useful to the commonwealth: for, with the man's villainy and madness branded, a similar disaster was being moved off the commonwealth for the future.

\ciceroSection{69} Nor did Cn. Pompeius, who gave a most distinguished opinion about me, nor you, pontiffs, who defended me by your opinions and authorities, fail to see that that thing was no law, but rather a flame of the time, a forbidding of villainy, a voice of madness; but you took thought lest at any time some popular hatred against us should overflow, if we were seen to be restored without the people's judgement. By the same counsel the Senate followed the opinion of the bravest of men, M. Bibulus: that you should decide about my house --- not because there was doubt that nothing had been done by him in accordance with the laws, the religious rules, or the law, but lest at any time, in such an abundance of the wicked, some man should arise who would say that some religious dedication still resided in my house. For that this was no law, as often as the Senate gave an opinion about me, it judged. Since indeed by that very writing of his, his giving an opinion was forbidden ---

\ciceroSection{70} this matter that pair, so similar, Piso and Gabinius, men afraid of laws and of trials, saw, when the Senate in fullest numbers daily demanded that they should refer about me, that they did not say they disapproved of the matter, but that they were prevented by his law. This was true: for they were prevented, but by that law which the same man had carried about Macedonia and Syria. This law you, P. Lentulus, neither as private man nor as consul ever thought a law. For when the tribunes of the plebs were referring, you, consul-elect, often gave an opinion about me; from the Kalends of January, until the matter was completed, you referred about me, you promulgated a law, you carried it: in which, if that were a law, you would have been allowed nothing. But indeed Q. Metellus, your colleague, that most distinguished man --- the law which men most strange to P. Clodius were judging to be a law, Piso and Gabinius, the same the brother of P. Clodius, when he referred about me to the Senate jointly with you, judged to be no law.

\ciceroSection{71} But indeed those who feared Clodius's laws --- how did they observe the rest? The Senate indeed, whose is the gravest judgement on the law of laws, as often as it was consulted about me, so often judged it no law. The same thing you, Lentulus, saw in the law you carried about me. For it was not so passed that I should be allowed to come to Rome, but that I should come; for you did not wish to carry, what was permitted, that it be permitted, but that I should be in the commonwealth in such a way that I should seem to have been called by the command of the Roman people rather than restored to manage the state.

\ciceroSection{72} This man you have even dared, monstrous plague, to call an exile --- when you yourself were marked by such villainies and disgraces that every place you came to, you made most like an exile. For what is an exile? In itself a name of calamity, not of disgrace. When then is it shameful? In truth, when it is the punishment of an offence; in men's opinion also when it is the punishment of a condemned man. Am I, then, undergoing the name for an offence of mine, or by a judgement passed? An offence? Now you yourself dare not say it, you whom your satellites here name a happy Catiline; nor any of those who used to. Not only is no man now so inexpert as to say that what I did in my consulship were offences, but no man is so hostile to his country as not to confess that the country was saved by my counsels.

\ciceroSection{73} For what is there in the lands so large or so small a counsel which has not, about my deeds, judged those things which were most to be wished for and most splendid for me? The highest counsel of the Roman people and of all peoples and nations and kings is the Senate: it decreed that all who wished the commonwealth safe should come together to defend me alone, and showed both that the commonwealth could not have stood, had I not been; and that none would be left, if I should not return.

\ciceroSection{74} Next to this dignity is the equestrian order: all the partnerships of all the public revenues passed the most ample and ornate decrees about my consulship and my deeds. The scribes, who deal with us in public accounts and records, wished that their judgement and decree about my services to the commonwealth should not be obscure. There is no college in this city, no \textit{pagani} or \textit{montani} (since our ancestors wished even of the urban plebs that there be gatherings, and as it were certain councils), which has not most lavishly decreed not only about my safety but about my standing.

\ciceroSection{75} For why should I recall those divine and immortal decrees of the towns and the colonies and of all Italy, by which I seem to have ascended into heaven by steps as it were, and not merely to have returned to my country? But that day, indeed, when, with you, P. Lentulus, carrying the law about me, the Roman people themselves saw and felt how great and of what dignity it was. For it is agreed that at no assemblies has the Campus Martius ever flourished with such concourse, with such splendour, of every kind of man, of every age, of every order. I leave aside the one judgement and the one consensus of cities, of nations, of provinces, of kings, finally of the world, on my services to all mortals: my arrival and entrance into the city, what was it? Did my country receive me as light and safety given back and restored to her, or as a cruel tyrant, as you Catilinarian rabble used to say of me?

\ciceroSection{76} And so that one day, on which the Roman people honoured me with its concourse and joy as it accompanied me from the gate to the Capitol and from there home, was of such pleasure to me that that criminal violence of yours seems to me not only not to have had to be warded off, but even to have been worth stirring up. So that calamity --- if so it should be called --- has burned out this whole kind of railing, that no one any longer dares to censure my consulship, approved by so many, so great, so splendid judgements, testimonies, and authorities. And if in your railing you not only object to me no shame, but even adorn my praise --- what madder thing can either happen to you or be feigned? For by one piece of railing you grant me to have saved my country twice: once, when I did what all men do not deny ought to be entrusted to immortality, if it can be done --- which you thought ought to be punished with execution; a second time, when I caught your assault, the assault inflamed against all good men, with my body --- that I might not, when I had saved that state by being unarmed, drag it under arms into danger.

\ciceroSection{77} So be it: there was no penalty in me for an offence; but there was for a trial. Whose? Who ever questioned me under any law? Who demanded a trial? Who named a day? Can a man, then, who is not condemned bear the punishment of one condemned? Is this tribunician, this popular? Although where can you call yourself popular, except when you have done it for the people? But when this right has been handed down by our ancestors, that no Roman citizen can lose either liberty or citizenship unless he himself shall have been the author of it (which you yourself were able to learn in your own case --- for I believe that, although in that adoption nothing was lawfully done, you were still asked whether you were the author, that P. Fonteius might have over you the power of life and death, as over a son): I ask, if you had refused or had been silent, if the thirty curiae nevertheless had ordered it, would that ordering have been valid? Surely not. Why so? Because the law was so set up by our ancestors, who were not feignedly and falsely popular but truly and wisely so, that no Roman citizen can lose his liberty against his will.

\ciceroSection{78} Nay rather, even if the decemvirs had unjustly judged a sacramentum about liberty, still, as often as anyone wished, they wished that in that one kind of case the matter judged could be brought back. As for citizenship, no man will ever lose it, against his will, by any order of the people. The Roman citizens who set out for Latin colonies could not become Latins, unless they had been the authors of it and had given in their names. Those who were condemned of capital crimes did not lose this citizenship before they had been received into that one to which they had come for the sake of changing soil --- that is, of changing it. And that this might be done, they did it not by the taking-away of citizenship, but by the denial of roof and water and fire.

\ciceroSection{79} The Roman people, with L. Sulla as dictator carrying the law, took citizenship from the towns by the centuriate assemblies; took their lands away. About the lands, this stood, for the people had power; about citizenship, it did not stand even so long as those arms of the Sullan time stood. Or could L. Sulla, conqueror, with the commonwealth recovered, by the centuriate assemblies snatch away the citizenship of the men of Volaterrae, when they were even then under arms; and today the men of Volaterrae enjoy this citizenship along with us not only as citizens but even as the best of citizens? Was P. Clodius, with the commonwealth overturned, in a packed assembly, with hired ruffians (not only of the beggared but even of the slaves), with Fidulius for chief --- who himself swears that he was not at Rome on that day --- able to take away citizenship from a man of consular rank?

\ciceroSection{80} If he was not there, what bolder thing than you, who had cut his name in? what more desperate, who could not, even by lying, sketch out a better author? But if he was the first to vote ``yes'' (which he could easily have done, who, for lack of a roof, had spent the night in the Forum), why does he not swear that he was at Cadiz, when you have proved that you were at Interamna? Do you, then, popular man, think it ought to be enough that our citizenship and freedom should be so guarded that, if at the demand of a tribune of the plebs ``Do you wish? Do you order?'' a hundred Fidulii had said that they wished and ordered, every one of us could lose his citizenship? At that time, then, our ancestors were not popular, who sanctioned about citizenship and freedom such laws as neither the violence of the times, nor the power of magistrates, nor the decrees of praetors, nor finally the power of the whole Roman people --- which in other matters is the greatest --- could undermine.

\ciceroSection{81} But you, even you, the snatcher of citizenship, carried a law on public injuries most welcome to one Menulla, an Anagnine fellow, who in return for that law set up a statue to you in my house --- so that the very place, in your great injury, refuted the law and the statue's inscription. This was much more grievous to the townsmen of Anagnia than the villainies which that same gladiator had committed at Anagnia.

\ciceroSection{82} What of this --- if it was not even written in that very bill (which Fidulius denies he voted on, but which you, that you may dignify the acts of your splendid tribunate with the standing of the man, embrace as your sponsor) --- if you carried nothing about me to the effect that I should not be in the number of citizens, even in the place to which the offices of the Roman people had set me, will you still violate, by your voice, the man whom, after the criminal villainy of the previous consuls, you see honoured by so many judgements of the Senate, of the Roman people, of all Italy --- whom not even when I was absent could you deny to be a senator under your law? For where had you carried that water and fire be denied me? --- a thing which C. Gracchus on P. Popilius, Saturninus on Metellus carried (the most seditious of men against the best and bravest of citizens): not (what could not be carried) that there be a denial, but that he be denied. Where did you provide that the censor should not enrol me in the Senate in my own place? --- a thing which is set down in the laws about all to whom, even when condemned, fire and water has been denied.

\ciceroSection{83} Demand these things from Clodius the drafter of your laws, order him to come forward. He is in hiding altogether; but, if you order him to be sought, they will find the man at your sister's house, hiding with his head bowed. But if your father --- by my faith, an extraordinary citizen, unlike all of you --- whom no one in his senses ever called an exile, who, when a tribune of the plebs had promulgated about him, refused to be present because of the inequity of that Cinnan time, and his command was abrogated --- if in him a lawful penalty had no shame because of the violence of the times: in me, to whom no day was named, who was not on trial, who was never summoned by a tribune of the plebs, can it be the penalty of a condemned man --- and one which is not even prescribed in the bill itself?

\ciceroSection{84} See what difference there is between that most unjust case of your father and our fortune and condition. Your father, the best of citizens, son of a most distinguished man --- who, if he were alive, with the severity he had, would surely keep you from being alive --- L. Philippus, censor, passed over his own uncle in reading the senatorial roll. For he could give no reason why those things were not valid which had been done in that commonwealth in which he had wished to be censor at those very times. Of me L. Cotta, an ex-censor, said in the Senate, on oath, that, if he had been censor when I was away, he would have read me out as senator in my own place.

\ciceroSection{85} Who put a substitute judge in my place? Who of my friends made his will at my withdrawal who did not assign me the same as if I were present? Who has hesitated to receive me, not only as citizen but as ally, against your law, and to help me? Finally the whole Senate, long before the law about me was carried, voted that thanks should be given to those states which had received M. Tullius --- only this? no, more --- a citizen who had so well deserved of the commonwealth. And do you, one pestilent citizen, deny that he is a citizen who has been restored, whom, when he was cast out, the whole Senate always thought not only a citizen but even a distinguished citizen?

\ciceroSection{86} Indeed, as the annals of the Roman people and the monuments of antiquity tell us, the famous Kaeso Quinctius, M. Furius Camillus, and C. Servilius Ahala --- though they had so well deserved of the commonwealth --- still bore the violence and anger of an inflamed people; condemned in the centuriate assemblies, when they had fled into exile, they were again, by the same people calmed, restored to their old standing. And if to those condemned, calamity not only did not lessen the glory of a most distinguished name, but even adorned it (for although it is more to be wished to finish the course of life without grief and without injury, yet to the immortality of glory it brings more to have been longed for by one's own citizens than never at all to have been violated): for me, restored without any judgement of the people at all, with the most ample judgements of all, will it hold the place of railing or of charge?

\ciceroSection{87} A brave and steadfast citizen in the best line was P. Popilius always; yet in his whole life nothing is more brilliant for praise than the calamity itself. Who would now remember that he had so well deserved of the commonwealth, had he not both been driven out by the wicked and restored by the good? Q. Metellus's command in soldiering was distinguished, his censorship outstanding, his whole life full of weight; yet to this man's praise the calamity of his time spread sempiternal memory. And if, even to those who were unjustly but yet by laws driven out, brought back, with their enemies killed, by tribunician bills --- not by the authority of the Senate, not by the centuriate assemblies, not by decrees of Italy, not by the longing of the citizenry --- the wrong done by their enemies was no shame, in me --- who set out unharmed, was away along with the commonwealth, returned with the highest standing while you were alive, with your brother estranged from you, with one consul bringing me back, with the other (the praetor) suffering it --- do you suppose your villainy ought to be my shame?

\ciceroSection{88} And if the Roman people, stirred by anger or hatred, had cast me from the citizenry, and the same people afterwards, recalling my services to the commonwealth, had collected itself, and had rebuked their own rashness and injury by my restoration --- still surely no man would be so mad as to think that such a judgement of the people ought to be a matter of standing for me rather than of disgrace. Now, however, when no man at all called me to a trial of the people, when I could not be condemned who had not been charged, when finally I had not even been driven out in such a way that, had I contested it, I could not have prevailed; on the contrary when I have always been defended by the Roman people, increased, ornamented --- what is it that any man can put himself before me on the grounds of being popular?

\ciceroSection{89} Or do you suppose that to be the Roman people which is made up of those who are hired for pay, who are pushed to bring violence against the magistrates, to besiege the Senate, to wish daily for slaughter, fires, plunder? --- whom you yourself could not gather thickly except with the shops shut, over which people you had set Lentidii, Lollii, Plaguleii, Sergii as leaders. O the appearance and dignity of the Roman people, which kings, foreign nations, and the furthest peoples should fear --- a multitude of men gathered together from slaves, from hired men, from the criminal, from the beggared!

\ciceroSection{90} That was the beauty of the Roman people, that the form which you saw in the Field, when even to you, against the authority and zeal of the Senate and of all Italy, the power of speaking was given. That is the people, the master of kings, the conqueror and ruler of all peoples, which on that most splendid day, you wretch, you saw --- when all the chiefs of the citizenry, all men of all orders and ages, thought they were giving their vote not on a citizen but on the safety of the state; when finally men had come into the Field, not with shops closed, but with their towns shut up.

\ciceroSection{91} I, with this people, if then there had either been consuls in the commonwealth or none at all, would have withstood your headlong fury and impious villainy without any labour. But I was unwilling to undertake a public cause against armed violence without public protection --- not because the violence of P. Scipio, that bravest man, against Ti. Gracchus, a private man, displeased me; for the act of Scipio, P. Mucius the consul (who had been thought somewhat sluggish in managing the commonwealth), by many senatus consulta not only defended once it was done, but even adorned. With me, either you had to be killed along with the consuls, or, with you alive, I had to fight in arms with you and with them.

\ciceroSection{92} There were many other things to fear at that time. The commonwealth, by my faith, would have come to slaves: so much hatred of good men was branded into the impious minds from that old conspiracy. Here you even forbid me to glory; you say that what I am used to proclaim about myself is not to be borne, and you, witty man, even bring in a piece of urbane and charming talk --- that I am used to say I am Jupiter, and the same to keep saying that Minerva is my sister. I am not so insolent in saying that I am Jupiter as I am unschooled in supposing Minerva to be Jupiter's sister; but still, I take to myself a virgin sister --- you do not let your sister be a virgin. But take care lest you yourself be used to call yourself Jupiter, since you can rightly call the same woman both sister and wife.

\ciceroSection{93} And since you reproach this, that I am used (you say) to proclaim too gloriously of myself: who has ever heard me speak of myself except when forced and out of necessity? For if, when thefts, largesses, lusts are objected to me, I am used to answer that the country was saved by my counsels, dangers, labours, I am to be considered not so much glorying in my deeds as confessing the charges. But if before these hardest times of the commonwealth nothing else was ever objected to me except the cruelty of that one time when I drove ruin from the country --- what then? was it more becoming for me not to answer this railing, or to answer it tamely?

\ciceroSection{94} Indeed, I always thought it was even of advantage to the commonwealth that I should keep, by words, the splendour and dignity of that most splendid deed which I had done by the authority of the Senate and the consent of all good men for the safety of the country, especially when to me alone, in this commonwealth, with the Roman people listening, it had been right to say on oath that this city and this commonwealth had been saved by my work. That railing of cruelty is now extinguished; for they see that I was longed for, sought back, summoned, by the zeal of all citizens, not as a cruel tyrant but as a most gentle parent.

\ciceroSection{95} Another has arisen: my withdrawal is objected to me. To this charge I cannot answer without my highest praise. For what, pontiffs, ought I to say? That I had fled because of the consciousness of an offence? But that which was given me as a charge was not only no offence, but the most splendid thing since men were born. That I had feared the people's judgement? But neither was that ever set in motion, and, if it had been, I should have come away with double glory. That the protection of good men was lacking to me? It is false. That I had feared death? It is shameful.

\ciceroSection{96} I must, therefore, say what I would not say unless forced --- for I have never said anything of myself loftily for the sake of taking up praise rather than driving off charge --- I say it, then, and as great as my voice can, I say: when the violence of all the lost and the conspirators, stirred up, with a tribune of the plebs as leader, with the consuls as authors, with the Senate stricken, with the Roman knights terrified, with the whole state in suspense and anxiety, was making an assault not so much against me as through me against all good men --- I saw that, if I should win, the slender remains of the commonwealth would be left, and if I should be conquered, none would be. When I had so judged, I wept for the parting from my unhappy wife, the loneliness of my dearest children, the case of my absent and most loving and best brother, the sudden ruins of a most secure household. But to all these things I set the life of my fellow citizens before; and I preferred that the commonwealth should suffer at the withdrawal of one man rather than that it should perish at the destruction of all. I hoped --- which has come to pass --- that I, while I lay flat, could be raised up by brave men still alive; that, if I had perished along with the good, the commonwealth could in no way be made anew.

\ciceroSection{97} I felt great and incredible grief, pontiffs: I do not deny it, nor do I take to myself that wisdom which some men were demanding in me, who were saying that I was of too broken a spirit and too downcast. Or could I, when I was being torn from such a variety of things --- which I pass over for this reason, that I cannot recall them, even now, without weeping --- have denied that I was a man and rejected the common feeling of nature? Then indeed, neither would I have called that act of mine praiseworthy, nor would I say any kindness had come from me to the commonwealth, if I had left for the commonwealth's sake those things which I would have lacked with calm spirit; and that hardness of spirit (just as that of the body, which when it is burned does not feel) I should have judged numbness rather than virtue.

\ciceroSection{98} To take on so great griefs of the spirit, and to bear, alone, with the city standing, the things that happen to the conquered when the city has been taken; even to see oneself torn from the embrace of one's own people, the roofs disturbed, the fortunes plundered, the country itself lost finally for the country's sake; to be despoiled of the most ample kindnesses of the Roman people, to be cast down from the highest grade of standing, to see one's enemies in praetextae, with death not yet bewailed, claiming the management of the funeral: to undergo all these things for the saving of the citizens, and that, while you bear it grievously --- not so wise as those who care for nothing, but as loving of yours and yourself as common humanity demands: that praise is splendid and divine. For he who deserts those things which he never thought dear and pleasant, with calm spirit, for the commonwealth's sake, declares no remarkable good will toward the commonwealth; but he who leaves, for the commonwealth's sake, those things from which he is torn with the greatest grief, to him the country is dear, whose safety he sets before love of his own people.

\ciceroSection{99} For which reason, although that fury and pestilence may burst, he shall hear this from me, since he has provoked me: I have twice saved the commonwealth --- I, who as toga-clad consul vanquished armed men, and as a private man yielded to armed consuls. From each time I bore the greatest fruit: from the earlier, that, on the authority of the Senate, I saw the Senate and all good men with their dress changed for the sake of my safety; from the later, that the Senate, the Roman people, and all mortals, both privately and publicly, judged that without my return the commonwealth could not be safe.

\ciceroSection{100} But this my return, pontiffs, is bound up in your judgement. For if you settle me in my own house, the thing which in my whole case you have always done, by your zeal, your counsels, your authority, your opinions, I see plainly and feel that I have been restored. But if my house is not only not given back to me, but even gives my enemy a monument of my grief, of his villainy, of the public calamity --- who will think this a return rather than a perpetual punishment? My house is in sight of nearly the whole city, pontiffs: in which, if there remains that thing --- not a monument of virtue but a tomb inscribed with my enemy's name --- I must rather emigrate somewhere else than dwell in that city in which I should see set up trophies of victory over me and over the commonwealth.

\ciceroSection{101} Or could I have such hardness of spirit, or such shamelessness of eyes, that, when the Senate has so often, with the consent of all, judged me the saver of this city, in it I could look upon my own house overturned --- not by my own private enemy but by the common enemy --- and by the same man a building set up before the eyes of the citizenry, that the weeping of good men might never be allowed to settle? The house of Sp. Maelius, who was reaching for kingship, was levelled; and, because the Roman people judged that this had befallen Maelius equitably, by the very name of the Aequimaelium the justice of the punishment was approved. The house of Sp. Cassius was, for the same cause, overturned, and on that place a temple of Tellus set up. In Vacci pratum was the house of M. Vaccus, which was confiscated and overturned, that his crime might be marked by the memory and the name of the place. M. Manlius, when he had driven back the Gauls' assault from the ascent of the Capitol, was not content with the glory of his deed; he was judged to have aimed at kingship: therefore his house, you see, was overturned and clothed in two groves. Shall I, then, undergo and bear that very greatest punishment which our ancestors thought could be set against criminal and unspeakable citizens, that to our descendants I may seem to have been not the extinguisher of conspiracy and crime but its author and leader?

\ciceroSection{102} Will the dignity of the Roman people, with the Senate alive, with you the chiefs of the public counsel, be able, pontiffs, to bear this stain of disgrace and inconsistency: that the house of M. Tullius Cicero shall be seen joined with the house of Fulvius Flaccus to the memory of a publicly fixed punishment? M. Flaccus was killed, by senatorial vote, because with C. Gracchus he had acted against the safety of the commonwealth; his house was overturned and confiscated; in which, some time later, Q. Catulus built a portico from the Cimbric spoils. But that firebrand and fury of the country, when she had with Piso and Gabinius as leaders captured the city, occupied it, held it --- at one and the same moment was wiping out the monuments of a most distinguished man dead, and joining my house with the house of Flaccus, that, with the punishment by which the Senate had marked the overthrower of the citizenry, the same fellow, with the Senate suppressed, might mark the man whom the senators had judged the guardian of the country.

\ciceroSection{103} Will you allow this portico to be on the Palatine and in the most beautiful place of the city, set up as the visible token of tribunician fury, of consular villainy, of the cruelty of the conspirators, of the calamity of the commonwealth, of my grief --- to the sempiternal memory of all peoples? --- which portico, by the love which you have for the commonwealth and have always had, you would long not only by your opinions but, if it were necessary, by your hands to break up, except perhaps the superstitious dedication of that most chaste priest deters someone.

\ciceroSection{104} O the matter that loose-living men cannot stop laughing at; while soberer men cannot hear without the greatest grief! P. Clodius, who tore religion from the house of the pontifex maximus, has he set it on mine? Do you, who are the priests of the rituals and rites, take him as your authority and master of public religion? O immortal gods --- for I wish you to hear this --- does P. Clodius care for your rites, fear your divinity, suppose all human affairs to be held in by your religion? Does he not mock the authority of all these highest men who are present? Does he not abuse, pontiffs, your gravity? Can the word ``religion'' fall or slip from that mouth of his? --- which mouth of his you, with the same mouth, in accusing the Senate for severely decreeing about religion, most foully and most foully violated.

\ciceroSection{105} Look, pontiffs, at this religious man, and, if you think well --- which is the part of good pontiffs --- warn him that there is a measure to religion: he ought not to be too superstitious. What need had you, fanatic, with old wives' superstition, to look in upon a sacrifice that was being made in another man's house? What such weakness of mind held you, that you did not think the gods could be sufficiently appeased unless you also entangled yourself in women's rites? Whom of your ancestors --- who both kept up the private rites and presided over the public priesthoods --- did you ever hear of as having been present when the sacrifice of Bona Dea was being performed? No one --- not even the man who has gone blind. From which it is understood that men have many false opinions in life: when that one, who knowingly had seen nothing that was unspeakable, lost the light of his eyes; while of this man, who polluted the rites not only by sight but by unchaste villainy and stuprum, the whole punishment of the eyes has been turned to blindness of mind. With this man as authority, so chaste, so religious, so holy, so devout, can you, pontiffs, not be moved, when he says that he with his own hands overturned the house of the best of citizens and with the same hands consecrated it?

\ciceroSection{106} What was your consecration? ``I had carried,'' he says, ``that it should be permitted to me.'' What of this --- you had not made the exception that, if anything could not by right be carried, it should not be considered carried? Will you, then, settle that it is right for the seats, the altars, the hearths, the household gods of every one of you to be subjected to the lust of a tribune? --- that the house of any man on whom each, through stirred-up men, has rushed, whom by an assault he has struck, may not only be afflicted (which is of present madness, like a sudden storm), but for the time to come bound by an everlasting religious dedication?

\ciceroSection{107} For myself, I have learned, pontiffs, in undertaking religious matters, that the chief thing is to interpret what the will of the immortal gods seems to be; nor is there any piety toward the gods unless there be an honourable opinion of their divinity and mind, so that one think nothing to be sought from them which is unjust and dishonourable, but to seek what is just and honourable. That ruin, when he held everything, could find no man to whom he might assign my house, to whom he might hand it over, to whom he might give it. He himself, when he was burning with greed for that place and that house, and had for that one cause wished, by his deadly bill, to be the master in my goods --- yet, even in that very madness of his, did not dare to possess my house, with desire of which he was inflamed: do you suppose that the immortal gods, by whose labour and counsel they themselves held their own temples, wished to migrate into a house broken and overthrown by the violence of the most criminal of men --- to its unspeakable banditry?

\ciceroSection{108} There is no citizen in so great a people, outside that contaminated and bloody hand of P. Clodius, who has touched any thing of my goods, who has not, by his own resources, defended me in that storm. But those who contaminated themselves by some contagion of plunder, of partnership, of buying --- could not escape the punishment of any private or public trial. From these goods, then, of which no man touched anything who was not held the most villainous in the judgement of all --- have the immortal gods coveted my house? That fair freedom of yours has driven out my household gods and family Lares, that it might settle itself, as it were in a captive seat?

\ciceroSection{109} What is more sacred, what more fortified by every religion, than the house of every one of the citizens? Here are altars, here are hearths, here are household gods; here rites, religion, ceremonies are contained; this is so sacred a place of refuge to all that no man can rightly be torn away from it. The more is this man's madness to be driven from your ears who, what our ancestors wished by religious means to be safe to us and sacred, has not only undermined against religion, but even, in religion's own name, overturned.

\ciceroSection{110} But what goddess is it? It must be a good one, since it has been dedicated by you. ``Liberty,'' he says, ``it is.'' You then settled in my house her whom you had banished from the whole city? You, when you denied that your colleagues, gifted with the highest power, were free; when entrance to the temple of Castor was open to no man; when this most distinguished man, born of the highest line, who had used the highest kindnesses of the people, a pontiff and a consular, gifted with singular goodness and modesty, whom I cannot enough wonder with what eyes you dare to look upon, with the Roman people listening, you ordered to be trampled by your footmen; when you drove me, uncondemned, out by tyrannical privilegia sworn against me; when you held shut up at home the leading man of the world; when you were occupying the Forum with armed gangs of ruined men --- you were placing in that house the image of liberty, a house which itself was the proof of your most cruel domination and of the most wretched servitude of the Roman people? Was she --- liberty --- to be driven, of all places, out of his house, who, had he not been there, would have brought the whole citizenry into the power of slaves?

\ciceroSection{111} But where was that Liberty of yours found? For I have asked diligently. A certain Tanagran courtesan is said to have been there. Her statue, in marble, was set up not far from Tanagra in a tomb. This a noble man, not unrelated to this religious priest of liberty, brought home for the adornment of his aedileship; for he had reckoned to surpass all the previous shows in splendour. So all the statues, paintings, ornaments which were left over in temples and public places throughout all Greece and all the islands, he brought home most frugally for the honour of the Roman people.

\ciceroSection{112} Afterwards, when he understood that, with his aedileship having been thwarted, he could be returned praetor by L. Piso the consul, if he should have a competitor with the same first letter of his name, he set down his aedileship in two places, partly in his strongbox and partly in his gardens. The statue, taken from a courtesan's tomb, he gave to that fellow, that it might be a sign more of those men than of public liberty. Should anyone dare violate this goddess --- the image of a courtesan, the ornament of a tomb, taken away by a thief, set up by a sacrilegious man? Shall this drive me from my own house? Shall this, the conqueress of an afflicted state, be adorned with the spoils of the commonwealth? Shall this be in that monument which has been set up that it should be the visible token of the suppressed Senate, for the memory of an everlasting disgrace?

\ciceroSection{113} O Q. Catulus! Shall I name the father first or the son? For the memory of the son is fresher and more bound up with my own affairs. Did so much deceive you, when you were thinking that the highest and daily greater rewards in the commonwealth were going to lie before me? You denied it right that two consuls in this state should be enemies of the commonwealth: men were found who handed the bound Senate over to a furious tribune, who by edicts and order forbade the senators to entreat for me and the people to be suppliants on my behalf, with whom looking on my house was disturbed, plundered, and who finally ordered the half-burned remnants of my fortunes to be carted to their own houses.

\ciceroSection{114} I come now to the father. You, Q. Catulus, when M. Fulvius's house had been the father-in-law of your brother, wished it to be a monument of your spoils, that of the man who had taken counsels harmful to the commonwealth all memory might be utterly removed from the sight and minds of men. If anyone, while you were building that portico, had said to you that the time would come when a tribune of the plebs --- one who had despised the authority of the Senate and the judgement of all good men --- would, with the consuls not only watching but even helping, disturb and overturn your monument, and would join it with the house of that citizen who, as consul, had defended the commonwealth on the authority of the Senate --- would you not answer that this could not happen unless the state had been overthrown?

\ciceroSection{115} But see the man's intolerable audacity, with a certain projecting and unbridled lust. Did this man ever think out a monument or any religious dedication? He wanted to live grandly and on a large scale, and to join two great and noble houses. At the same point of time at which my withdrawal took from him the cause of slaughter, he was demanding of Q. Seius that he should sell him his house. When that man refused, at first he threatened that he would block his light. Postumus affirmed that, while he lived, the house should never be that man's. The sharp youth understood from that very speech what should be done; he removed the man, most openly, with poison; he bought the house, with the bidders worn out, almost half as dear again as it was valued. To what, then, does this speech tend?

\ciceroSection{116} That house of mine is nearly all empty; scarcely a tenth part of my buildings has been added to Catulus's portico. The cause was the walking-place and the monument and that Tanagran liberty, with liberty itself oppressed. On the Palatine, with most beautiful prospect, he had coveted a portico with chambers, paved, of three hundred feet; a most ample peristyle, the rest in such a manner as easily to surpass all men's houses both in spaciousness and in dignity. And the religious man, when the same person was buying and selling my house, yet in those great shadows did not dare to write his own name to that purchase. He set up that Scato, a man wanting in his own virtue, that he --- who in the Marsian country, where he was born, had no longer any roof under which to take shelter from rain --- should say he had bought the most noble house on the Palatine. The lower part of the house he assigned, not to his own gens of the Fonteii, but to the Clodian; which gens he had abandoned --- a gens to which, of the many Clodii, no man has given his name except one ruined by either poverty or villainy. Will you, pontiffs, approve this so varied, so novel a will in every kind --- this shamelessness, this audacity, this greed?

\ciceroSection{117} ``A pontiff,'' he says, ``was present.'' Are you not ashamed, when the matter is being conducted before pontiffs, to say a pontiff and not the college of pontiffs was present, especially when as tribune of the plebs you could either announce them or even compel them? So be it: you did not bring in the college. What of this --- which one of the college, then, was present? For there was need of authority, which is in all of these men; but age and office augment dignity. There was need also of knowledge, which, although they all have got, yet certainly long experience makes more skilled.

\ciceroSection{118} Who, then, was present? ``My wife's brother,'' he says. If we seek authority, even though he is of an age that he has not yet acquired it, yet, however much authority there is in a young man, that must be considered the less because of so close a tie of marriage; but if knowledge was sought, who was less skilled than he who had come into the college a few days before? --- who was even more bound to you by recent kindness, when he saw himself preferred to your own brother, your wife's brother. Yet in this you provided that your brother could not accuse you. Do you, then, call this a dedication, to which you could call neither the college, nor a pontiff adorned with the offices of the Roman people, nor even a young man of any kind, when you had your closest friends in the college? He was present (if indeed he was present) whom you had pushed, whom your sister had asked, whom your mother had compelled.

\ciceroSection{119} See, then, pontiffs, what you settle in my case about the fortunes of all: are you to think that, by a pontiff's word --- if he shall touch the doorpost and say something --- the house of every one of us can be consecrated; or were the dedications of temples and shrines and the religious rites of them set up by our ancestors for the honour of the immortal gods, without any calamity to citizens? A tribune of the plebs has been found who, equipped with consular forces, in every assault of his madness has rushed against that citizen whom the commonwealth herself, when he was struck down, would lift up by her own hands.

\ciceroSection{120} What of this --- if any man like him (and there will not be lacking those who would imitate him) shall, by violence, strike down some man unlike me, to whom the commonwealth does not owe so much, and shall dedicate his house through a pontiff, will you, by that authority, settle that this ought to be valid? You say: ``Where will he find a pontiff?'' What of this --- can a pontiff and a tribune of the plebs not be the same man? M. Drusus, that most distinguished man, was a tribune of the plebs and a pontiff. So if he had touched the doorpost of the house of his enemy Q. Caepio and said a few words, would Caepio's house have been dedicated?

\ciceroSection{121} I say nothing about pontifical right, nothing about the words of the dedication itself, nothing about religion, ceremonies. I do not pretend that I do not know matters which, even if I knew, I would pretend not to know, lest I seem troublesome to others, even meddlesome to you; though many things flow out from your discipline that often come to our ears too. I seem to have heard that in the dedication of a temple it is right that a doorpost be held: for there is a doorpost where there is the entrance and the doors of a temple. No one ever held the doorpost of a walking-place in dedicating it. But if you have dedicated a statue or an altar, it can be moved from the place without religious offence. But this you will no longer be allowed to say, since you said that a pontiff held the doorpost.

\ciceroSection{122} But why do I speak of dedication, or argue about your law and religion contrary to what I had set out? I, in fact, even if I were to say that everything had been done with the solemn words and with the old and traditional usage, would still defend myself by the law of the commonwealth. Would the commonwealth, made new again, have been able to bear it that, when you, at the withdrawal of the citizen by whose work alone the Senate and all good men had so often judged the state to be safe --- with the commonwealth oppressed by the foulest banditry, with two most criminal consuls, you holding it in your hand --- should have dedicated the house of him who had refused that the country saved by him should perish in his name?

\ciceroSection{123} Give entrance to this religion, pontiffs: now you will find no way out for the common fortunes. Or if a pontiff has held a doorpost and has transferred words composed for the religion of the immortal gods to the destruction of citizens, the most holy name of religion shall be valid in injustice; if a tribune of the plebs has consecrated the goods of someone with words no less ancient and equally solemn, shall it not be valid? But indeed C. Atinius, in our fathers' memory, consecrated the goods of Q. Metellus --- who as censor had thrown him out of the Senate --- the grandfather of you, Q. Metellus, and of you, P. Servilius, and the great-grandfather of you, P. Scipio --- with a brazier set up in the Rostra and a flute-player called in. What then? Did that madness of a tribune of the plebs, drawn from a few precedents of very ancient times, do harm to Metellus, the highest and most distinguished man?

\ciceroSection{124} Surely not. We saw the same thing done by a tribune of the plebs to Cn. Lentulus, censor: did he therefore by religion bind any of Lentulus's goods? But why do I speak of others? You, you, I say, with head veiled, with public meeting summoned, with brazier set up, consecrated the goods of your friend Gabinius, on whom you had bestowed all the kingdoms of the Syrians, the Arabs, the Persians. If, then, nothing was done at that time, what could be done in my goods? But if it is valid, why does that gulf of a man, glutted along with you on the blood of the commonwealth, build up his villa at Tusculum to the sky out of the bowels of the treasury, while I have not been allowed to look upon my own ruins, the like of which throughout the city I did not allow?

\ciceroSection{125} I leave Gabinius aside. What of this --- by your own example, did not L. Ninnius, the bravest and best of all men, consecrate your goods? But if, because it pertains to you, you say that it ought not to be valid, those laws you have established in your splendid tribunate, with which you would refuse to be bound when they were turned against you, but with which you overturned others; but if that consecration is lawful, what is there which can be profane in your goods? Or has consecration no force, and dedication is religious? What did that obtestation of yours, the flute-player, the brazier, the prayers, the ancient words, then avail? Why did you wish to lie, to deceive, to abuse the divinity of the immortal gods to men's terror? For if that is valid --- I leave aside Gabinius --- your own house, certainly, and whatever else you have, has been consecrated to Ceres; but if that was a sport, what is more impure than you, who pollute all religious rites either by lying or by stuprum?

\ciceroSection{126} ``I now confess,'' he says, ``that in Gabinius's case I was unspeakable.'' You see how the punishment you set up for another has been turned upon yourself. But, you witness of every villainy and disgrace --- what you confess in the case of Gabinius, whose unchastity in boyhood, whose lusts in youth, whose disgrace and poverty in the rest of his life, whose banditry as consul we saw, to whom not even the calamity itself could come unjustly --- this you weaken in my case, and you say that what you did with one young man as witness is graver than what you did with a whole public meeting?

\ciceroSection{127} ``Dedication,'' he says, ``has great religious force.'' Does Numa Pompilius not seem to you to be speaking? Learn his speech, pontiffs --- and you, flamines: even you, rex sacrorum, learn from your kinsman, although he has abandoned that gens; but yet learn from a man given to the religious matters, the whole law of all rites. What of this --- in dedication, is it not asked who dedicates and what and how? Or do you so confound and disturb these things that whoever wishes can dedicate what he wishes, how he wishes? Who were you who were dedicating? By what right? By what law? By what precedent? By what authority? Where had the Roman people set you in charge of that thing? For I see that there is an old tribunician law which forbids that, without an order of the plebs, a building, a piece of land, an altar be consecrated. Nor at that time did Q. Papirius, who carried this law, perceive, nor did he suspect that the danger would be that the dwellings or possessions of uncondemned citizens should be consecrated. For neither was it right that this should be done, nor had any man done it, nor was there cause why he, in forbidding, should seem not so much to deter as to remind.

\ciceroSection{128} But because there were being consecrated buildings --- not the dwellings of private men, but those which are called sacred ---, fields were being consecrated, not as our estates, if anyone wished, but as a commander would consecrate fields captured from the enemy; altars were being set up which would bring religious force to the very place where they had been consecrated --- these things, unless the plebs had so ordered, he forbade to be done. If you interpret this as written about our houses and fields, I do not contend; but I ask what law has been carried that you should consecrate my house, where this power was given to you, by what right you did it. Nor am I now arguing about religion, but about all our goods; nor about pontifical right, but about public law. The Papirian law forbids that buildings be consecrated without an order of the plebs. Let this stand, indeed, about our houses and not about public temples: show me one word of consecration in your law itself, if it is a law and not the voice of your villainy and cruelty.

\ciceroSection{129} If at that time, in that shipwreck of the commonwealth, everything could have come into your mind, or if your secretary, in that fire of the citizenry, had not been making bonds with the Byzantine exiles and the legates of Brogitarus, but had been writing those things --- not decrees but monsters --- with a free mind for you, you would have got everything, if not in fact, at least in lawful words. But at one and the same time bonds were being made for sums of money, treaties for provinces were being struck, the namings of kings were for sale, the assignments of all slaves throughout the city by neighbourhoods were being made out, enemies were being reconciled into favour, new commands were being written for young men, poison was being prepared for the wretched Q. Seius, plans were being entered into for killing Cn. Pompeius, the defender and guardian of the empire; that the Senate be nothing, that good men always mourn, that the captured commonwealth, by the consuls' betrayal, be held by tribunician violence. When so many and such great things were going on, it is not strange (especially in his madness of spirit and blindness) that many things escaped him and you.

\ciceroSection{130} But see how great is the force of this Papirian law in such a matter --- not in such a one as you bring forward, full of villainy and madness. Q. Marcius the censor had made a statue of Concord and had set it up in public. When C. Cassius, censor, had transferred this statue into the Curia, he consulted your college whether anything seemed to be a cause why he should not dedicate that statue and the Curia to Concord. Compare, I beg you, pontiffs, man with man, time with time, matter with matter. He was a censor of the highest moderation and weight; this man a tribune of the plebs of singular villainy and audacity. That time was peaceful, set in the freedom of the people and the steering of the Senate; your time, with the freedom of the Roman people oppressed, with the authority of the Senate destroyed.

\ciceroSection{131} The matter was full of justice, of wisdom, of dignity (for the censor --- in whose hand our ancestors wished to be, what you have removed, the Senate's judgement on dignity --- wished a statue of Concord in the Curia and the Curia to be dedicated to that goddess). It was a splendid will, worthy of all praise; for he thought it was, as it were, prescribing that opinions should be given without the zeal of dissension, if he should bind the very seat and temple of public counsel by the religion of Concord. You, with iron, with fear, with edicts, with privilegia, with present forces of ruined men, with the terror and threats of an absent army, with the partnership and unspeakable compact of the consuls, when you held the citizenry oppressed in slavery, set up the image of Liberty more for the mockery of your own shamelessness than for the pretence of religion. He, in the Curia which could be dedicated without anyone's inconvenience; you, on the blood and almost on the bones of a citizen who had so well deserved of the commonwealth, set up the image not of public liberty but of licence.

\ciceroSection{132} And he yet referred to the college; you, to whom did you refer? If you were considering anything, if you were doing some private religious matter that had to be expiated or instituted, still by the old custom of the rest you would have referred it to a pontiff. When you were beginning a new shrine in the most distinguished place of the city, with an unspeakable and unheard-of practice, did you not think you had to refer to the public priests? But if it did not seem right to bring in the college of pontiffs, did none of these men seem to you fit, who excel in age, in office, in authority, with whom you might consult about the dedication? Whose dignity you did not despise but feared. Or would you dare to ask of P. Servilius or of M. Lucullus --- by whose counsel and authority I, as consul, snatched the commonwealth out of your hands and your jaws --- with what words or by what rite (this I say first) you should consecrate the house of a citizen; and next, of that citizen to whom the chief of the Senate, then all the orders, then all Italy, then all peoples had given testimony for the saving of this city and this empire?

\ciceroSection{133} What would you say, O unspeakable and pernicious stain of the citizenry? ``Be present, Lucullus, be present, Servilius, while I dedicate Cicero's house, that you may guide me in the words and hold the doorpost!'' You are indeed of singular audacity, but of singular shamelessness too. Yet your eyes, your face, your words would have fallen, when those men, who in their own dignity sustained the person of the Roman people and the authority of the empire, would have terrified you with the gravest words, and would have said it was not lawful for them to take part in your madness and to exult in the parricide of the country.

\ciceroSection{134} When you saw these things, then you turned to your kinsman by marriage --- not chosen by you but left over by the rest. Of him, however, I believe, if he is sprung from those who are reported to have learned the rites from Hercules himself, when his labours were done, that he was not so cruel in the hardships of a brave man as to set with his own hands a pyre on the head of one even still alive and breathing: who either said nothing or did nothing at all, and bore this punishment for his mother's rashness, that he should give a mute person and name in the offence; or, if he said something with stumbling words and touched the doorpost with a trembling hand, certainly he completed nothing duly, nothing chastely, nothing in the manner and rite. He had seen Murena, his stepfather, the consul-elect, bringing me as consul reports of the common destruction with the Allobroges; he had heard from him that he had twice received his safety from me --- once separately, again with everyone.

\ciceroSection{135} Therefore who is there who can suppose that to this new pontiff --- setting out his first religious matter after entering on the priesthood and putting forth his voice --- there did not happen both that his tongue was struck dumb, and that his hand grew numb, and that his mind, weakened by fear, fell, especially when he saw of so great a college neither the rex sacrorum, nor a flamen, nor a pontiff, and was being compelled, against his will, to be a partner of another man's villainy, and was bearing the gravest punishment of an unspeakable kinship by marriage?

