% Post Reditum in Senatu (post-reditum-in-senatu) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Senators, if for your immortal services to me and to my brother and to our children I shall have given you thanks too little fully, I beg and entreat you to think it is to be set down to my own nature rather than to the size of your kindnesses. For what such fertility of mind can there be, what such abundance of speaking, what such divine and incredible kind of utterance, by which any man could --- I shall not say embrace by speaking, but even survey by counting --- the whole of what you have together earned by giving to us? You have given me back my brother whom I longed for; me to my brother who loves me so much; their parents to our children, our children to us; you have given back rank, order, fortunes, this great commonwealth, this country than which nothing can be more sweet --- and at last, you have given us back to ourselves.

\ciceroSection{2} If we ought to hold our parents most dear, because life and patrimony and freedom and citizenship are handed to us from them; if the immortal gods, by whose kindness we have held these things and been increased in the rest; if the Roman people, by whose offices we have been placed in the greatest council and in the highest grade of standing and in this citadel of all the lands; if this very order, by whose magnificent decrees we have so often been honoured: then it is something measureless and unbounded that we owe to you --- you who, by your singular zeal and singular agreement, have at one and the same moment given us back the kindnesses of our parents, the gifts of the immortal gods, the offices of the Roman people, and your own many judgements about me. So that, while we owe much to you, much to the Roman people, things innumerable to our parents, all things to the immortal gods, those things separately we held before through them; now, all together, we have recovered through you.

\ciceroSection{3} And so, senators, I appear to have obtained through you what is not even to be wished for in a man --- a kind of immortality. For what time will there ever be when the memory and the fame of your kindnesses to me shall die? You who, at the very moment when, hemmed in by violence, by the sword, by fear, by threats, you were held under siege --- not long after my departure recalled me by common voice, on the motion of L. Ninnius, the bravest and best of men, whom that pestilential year had as the most loyal and the least fearful defender of my safety, had it been thought fit to fight. After the power of decreeing was no longer left to you, by the act of that tribune of the plebs who, when he could not in his own person tear at the commonwealth, had hidden himself under another man's villainy --- you were never silent about me, never failed to demand my safety from those consuls who had sold it.

\ciceroSection{4} And so by your zeal and authority it was brought about that that very year which I had wished to be fatal to me rather than to my country had eight tribunes of the plebs who promulgated for my safety and again and again referred the matter to you. For the consuls, modest men and law-abiding, were prevented by a law --- not the law against me, but the law passed against themselves, which my enemy promulgated, that, if those who had almost destroyed our state should rise from the dead, then I should return: by which act he confessed both that he longed for the lives of those men, and that the commonwealth would be in great danger if, with the enemies and assassins of the commonwealth alive again, I had not come back. And in that very year, when I had withdrawn, when the chief man of the citizenry was guarding his life not by the protection of laws but by his own walls, when the commonwealth was without consuls, robbed not only of its perpetual parents but even of its yearly guardians, when you were forbidden to give your opinions, when the heading of my proscription was being read out --- you never doubted to bind my safety together with the common safety.

\ciceroSection{5} But after, by the singular and outstanding courage of the consul P. Lentulus, you began on the Kalends of January to look out from the fog and darkness of the previous year into the light of the commonwealth; when the great standing of Q. Metellus, that most noble man and best of men, had come to the aid of the state, when the courage and good faith of the praetors and almost all the tribunes of the plebs had come to the aid of the state; when Cn. Pompeius --- in courage, in glory, in deeds easily the first of all peoples, of all ages, of all memory --- thought he could come safely into the Senate: such was your common consent about my safety that, while my body was absent, my standing had already returned to my country.

\ciceroSection{6} In that month indeed you were able to judge what difference there was between me and my enemies. I gave up my own safety, that the commonwealth should not be drenched in citizens' wounds on my account; they thought my return must be cut off, not by the votes of the Roman people, but by a flood of blood. And so afterwards you gave no answer to citizens, none to allies, none to kings; nothing the courts declared by their verdicts, nothing the people by their votes, nothing this order by its authority; you saw the Forum mute, the Curia tongueless, the city silent and broken.

\ciceroSection{7} At that very time, when the man had withdrawn who, with you for his backers, had stood up against slaughter and burning --- when you saw men flying about the whole city with iron and torches, the houses of the magistrates assaulted, the temples of the gods set on fire, the rods of fasces of the highest man and most distinguished consul broken, the most sacred body of a most brave and most upstanding tribune of the plebs not merely touched and violated by hand but wounded with iron and cut to pieces --- some magistrates, shaken by that carnage, drew back a little from my cause: some out of fear of death, some out of despair of the commonwealth. The rest there were whom no terror, no force, no hope, no fear, no promises, no threats, no weapons, no torches could drive away from your authority, from the dignity of the Roman people, and from my safety.

\ciceroSection{8} The chief man, P. Lentulus --- parent and god of our life, fortune, memory, and name --- thought this would be the proof of his courage, this the token of his spirit, this the very light of his consulship, if he should give me back to myself, to my own people, to you, and to the commonwealth. As soon as he was named consul-elect, he never doubted to deliver an opinion worthy of my safety and of the commonwealth. When he was forbidden by a tribune of the plebs, when that splendid clause of the law was being read out --- that no man should refer the matter to you, no man decree, no man debate, no man speak, no man cross to vote, no man be present at the writing --- he held that whole proscription to be no law: by which a citizen with the highest deserts of the commonwealth was being torn away by name, without trial, together with the Senate, from the commonwealth itself. And when he entered upon his magistracy, I shall not say what he did first, but what at all he did, except that, by saving me, he should set up your authority and standing for the time to come.

\ciceroSection{9} O immortal gods, what a kindness you seem to me to have given, that this year P. Lentulus was consul of the Roman people! And how much greater you would have given, had he been consul the year before. For I should not have needed the medicine of a consul, had I not fallen by a consul's wound. I had heard from the wisest of men and the best citizen and man, Q. Catulus, that there had often not been a single bad consul, but never two --- with the exception of that Cinnan time. So he used to say that my cause would always be the firmest, while there was even one consul left in the state. He spoke truly, if that condition --- the two-consul case, which had not before been seen in the commonwealth --- could have lasted as a continuing thing. For if Q. Metellus had been consul at that time as my enemy, do you doubt with what spirit he was likely to be in saving me, when you see he has been the prime mover and signatory in restoring me?

\ciceroSection{10} But there were those consuls whose minds, narrow, low, twisted, filled up with darkness and filth, could neither look up to nor sustain nor take in the very name of consul, the splendour of that office, the greatness of so great a command --- not consuls, but traffickers in provinces and sellers of your standing. Of whom one was demanding back from me Catiline --- his lover --- in many men's hearing; the other Cethegus, his cousin. These two, the most villainous since the memory of men, not consuls but bandits, did not merely desert me in a public and consular cause, but betrayed me, attacked me, and wished me to be stripped of every help, not only their own, but yours and that of all the orders. Yet one of them deceived neither me nor anyone.

\ciceroSection{11} For who could entertain any hope of any good in a man whose youth had been openly available to all kinds of lust? Who could not even keep men's filthy unrestraint away from the most sacred parts of his own body? Who, when he had got through his own estate no less briskly than later he got through the commonwealth, kept up his poverty and luxury by playing the pander in his own household? Who, had he not taken refuge on the altar of the tribunate, could not have escaped the praetor's force, his creditors' multitude, or the proscription of his goods? Who, in his magistracy, had he not carried the bill on the pirate war, would surely, driven by poverty and corruption, have turned pirate himself --- and would have done less harm to the commonwealth than that an unspeakable enemy and brigand walked within our walls? It was while he sat watching that a tribune of the plebs carried a law that the auspices should not be obeyed, that no man should announce against an assembly or against a comitia, that no man should veto a law --- that the Aelian and Fufian laws, which our ancestors had wished to be the surest protections of the commonwealth against tribunician madnesses, should have no force.

\ciceroSection{12} And the same man afterwards, when an innumerable multitude of good men in mourning had come down from the Capitol as suppliants to him, when the most noble youths and all the Roman knights had thrown themselves at the feet of the most shameless of pimps --- with what a face the curled-haired libertine spurned not only the tears of the citizens but the prayers of the country itself! And he was not content with that, but mounted to a public meeting and there said what, had Catiline her former husband risen from the dead, he would not have dared to say: that he would exact from the Roman knights the penalties for the Nones of December which had taken place in my consulship and for the Capitoline slope. Nor only did he say this, but he laid hands by name on whom he pleased, and as for L. Lamia, a Roman knight, a man of outstanding standing, most loving of my safety in his friendship, most loving of the commonwealth in his fortunes --- he, the masterful consul, ordered out of the city. And when you had voted that the senatorial dress should be changed, and you all had changed it, and all good men had done so before --- he, smeared with unguents, in the praetexta which all the praetors and aediles had then thrown off, mocked the squalor of you and the mourning of a most dear citizenry; and he did what no tyrant ever did --- said nothing to keep you from groaning over your evil in secret, but issued an edict that you should not openly mourn the country's ills.

\ciceroSection{13} But when in the Circus Flaminius he was led out into the public meeting --- not by a tribune of the plebs as a consul, but by an arch-pirate as a bandit --- he came forward first, with what consular weight! full of wine, of sleep, of debauchery, his hair dripping, his locks dressed, his eyes heavy, his cheeks flopping; in a thick, drunken voice he, as a man of grave authority, declared that the punishment of citizens not condemned by trial displeased him most violently. Where had this great authority of his lain hidden so long? Why had so outstanding a courage in this curled-haired dancing master been on holiday so long, in dens and gluttonies? For that other one, Caesoninus Calventius, has from his youth haunted the Forum, with nothing to recommend him except a feigned and crafty sourness --- no judgement, no abundance of speaking, no zeal for soldiering or for getting to know men, no liberality. Whom, going past, when you had seen him uncombed, bristly, mournful, even if you thought him rustic and uncivil, you would still not have judged him debauched and ruined.

\ciceroSection{14} Whether you stood in the Forum with this man or with a stump, you would have thought there was no difference --- without sense, without taste, tongueless, slow, an inhuman business; you would have called him only a Cappadocian just dragged out of a slave-pen. The same man at home, how lustful, how filthy, how unrestrained --- with pleasures admitted not by the front door but smuggled in by the back! Yet when even he begins to study letters, and the monstrous beast to philosophise with little Greeks, then he is an Epicurean: not deeply given to that doctrine, whatever it is, but caught by the one word ``pleasure.'' And he has masters, not from those impertinent ones who debate whole days about duty and about virtue, who exhort one to labour, to industry, to undergoing dangers for one's country --- but those who argue that no hour ought to be empty of pleasure, that in every part of the body some joy and delight should always be found.

\ciceroSection{15} These he uses as masters of the household for his lusts; these track and sniff out every pleasure; these are the founders and arrangers of the dinner-party; the same men weigh and assess the pleasures, give an opinion and judge how much should be assigned to each lust. Trained in their arts, he so despised this most prudent state that he reckoned all his lusts and all his shamefulness could be hidden, if only he carried his hard face into the Forum. He, in fact, did not deceive me at all --- for I had known, by the marriage tie of the Pisones, how far his maternal birth, of Transalpine blood, had drawn him from this race of men --- but he deceived you and the Roman people, not by judgement nor eloquence (which often happens with many), but by his wrinkles and his eyebrow.

\ciceroSection{16} L. Piso, did you dare with that eye --- I shall not say with that mind --- with that face, not life, with such an eyebrow, for I cannot say with such great deeds done, to combine your counsels with A. Gabinius for my destruction? Did not the smell of his unguents, the panting of his wine, the brow stamped with the tracks of the curling-iron lead you to this thought --- that, when you had been similar to him in fact, you should not be allowed any longer to use the cover of your own brow to hide such great shamefulness? With this man you dared to enter into a compact, that you should sell off, by the bargain of provinces, the consular dignity, the standing of the commonwealth, the authority of the Senate, and the fortunes of a citizen of the highest deserts? In your consulship, by your edicts and orders, the Senate of the Roman people was not allowed to come to the aid of the commonwealth --- not only by their opinions and their authority, but not even by their grief and their dress?

\ciceroSection{17} Did you suppose yourself consul at Capua --- in which city was once the home of pride --- as you were at that time, or at Rome, in whose state all the consuls before you obeyed the Senate? You dared, when led out into the Circus Flaminius with that companion of yours, to say that you had always been merciful? By which word you were demonstrating that the Senate, and all good men, when they were warding the plague off the country, had been cruel. Were you, the merciful, who had set me, your kinsman by marriage, as the foreman of the prerogative century in your election, who had asked me on the Kalends of January to give my opinion in the third place --- were you the merciful when you handed me over, bound, to the enemies of the commonwealth? You spurned my son-in-law, your cousin, with the proudest and cruellest words from your knees; you spurned your kinswoman, my daughter. And you, of singular clemency and pity --- when I had fallen, together with the commonwealth, by no tribunician but a consular blow --- were of such villainy and such intemperance that you would not allow even one hour to pass between my plague and your plunder, until at least the lamentation and the groaning of the city should fall silent!

\ciceroSection{18} It was not yet publicly known that the commonwealth had fallen, when the funeral arrangements were being settled to your account. At one and the same time my house was being plundered, was burning; my goods were being carried off from the Palatine to the consul next door, from my Tusculan villa to the other consul next door --- with the same hired ruffians acting as voters, with the same gladiator as carrier of the law, in a Forum empty not only of good men but even of free men, with the Roman people not knowing what was being done, the Senate suppressed and stricken: the treasury, provinces, legions, commands were being given as gifts to two impious and unspeakable consuls. The ruins these consuls left, you, the consuls, have shored up by your courage --- borne up by the great good faith and care of the tribunes of the plebs and of the praetors.

\ciceroSection{19} What shall I say of the most outstanding man, T. Annius? --- or who shall ever speak worthily enough of so great a citizen? Who, when he saw that a villainous citizen --- or rather a household enemy --- if it was permitted to use the laws, must be broken in court; but if violence itself impeded and removed the courts, then audacity must be overcome by courage, madness by fortitude, recklessness by judgement, brute force by resources, violence by violence --- first prosecuted him under the law on violence; then, when he saw that the courts had been removed by that same man, he took care that he might not be able to do everything by violence: who taught that neither houses nor temples nor the Forum nor the Curia could be defended without the highest courage and the greatest resources and forces from a domestic banditry; who first, after my withdrawal, drove fear from good men, hope from the audacious, terror from this order, and slavery from the citizenry.

\ciceroSection{20} Following this same plan with equal courage, spirit, and good faith, P. Sestius, for my safety, for your authority, for the standing of the state, never thought any enmities, any violence, any attacks, any peril of life had to be avoided. He so commended the cause of the Senate, harassed in the public meetings of villainous men, by his own care to the multitude, that nothing seemed so popular as your name, nothing once so dear to all as your authority. He defended me with everything that a tribune of the plebs could; the rest of his services he kept up just as if he had been my brother. By his clients, his freedmen, his household, his resources, his letters, I was so kept up that he seemed not only the helper of my calamity but its partner.

\ciceroSection{21} You have now seen the services and the zeal of the rest --- how desirous of me C. Cestilius was, how zealous of you, how unwavering in the cause. What of M. Cispius? to whom and to whose father and brother I feel how much I owe; for though their good will had been bruised by me in a private trial, the memory of my public benefit blotted out the private offence. Then T. Fadius, who was my quaestor; M. Curtius, whose father I served as quaestor: in zeal, in love, in spirit, neither failed our connection. C. Messius spoke much about me, both for friendship's sake and for the commonwealth's: he separately, at the start, promulgated a law about my safety.

\ciceroSection{22} Q. Fabricius --- if what he attempted on my behalf he had been able to bring through against violence and the sword, we should have recovered our state in the month of January. His good will pushed him toward my safety, violence held him back, your authority called him forward. As for the praetors, what spirit they had toward me you could judge: when L. Caecilius privately strove to keep me up by all his resources, and publicly promulgated a law about my safety with almost all of his colleagues, and gave the despoilers of my goods no power of bringing the matter into court. M. Calidius, named praetor-elect, by his very opinion declared at once how dear my safety was to him.

\ciceroSection{23} All the services of C. Septimius, Q. Valerius, P. Crassus, Sex. Quinctilius, C. Cornutus stood at the highest pitch both toward me and toward the commonwealth. While I gladly call these things to mind, so I do not unwillingly pass over the wicked acts of certain men against me. It is not part of my present case to remember injuries: which, even if I could avenge them, I should rather forget. My whole life must be turned in another direction --- to give thanks to those who have well deserved of me, to keep up friendships tested by fire, to wage war with my open enemies, to forgive my timid friends, to expose the betrayers, and to console the grief of my departure by the dignity of my return.

\ciceroSection{24} If in all my life there were no other duty left me except to be judged grateful enough toward the very leaders, chiefs, and authors of my safety, still I should suppose the brief span of my remaining life to be left me only for repaying, not for so much as recalling, the gratitude. For when shall I --- when shall any of my people --- repay this man and his children? What memory, what force of mind, what greatness of attention will be able to answer to so many and so great kindnesses? --- the man who first, when I lay afflicted, stretched out to me a consular hand of good faith; who called me from death to life, from despair to hope, from destruction to safety; who was of such love towards me, of such zeal for the commonwealth, that he thought out a way to ease my calamity and even to do it honour. For what could be more magnificent, what more splendid, could befall me than that, on his motion, you decreed that all from every part of Italy who wished the commonwealth safe should come together to me alone, a man broken and almost cast about the winds, that I might be restored and defended? --- so that, by the same voice with which only three times since the founding of Rome a consul had used for the whole commonwealth, before those alone who could hear his voice, by that same voice the entire Senate should rouse all the citizens out of all the fields and towns and the whole of Italy to defend the safety of one man.

\ciceroSection{25} What more splendid thing could I leave to my descendants than this --- that the Senate had judged that whoever did not defend me as a citizen had not wished the commonwealth safe? And so much weight had your authority, so much had the outstanding standing of the consul, that any man who did not come thought himself committed to a disgrace and a scandal. The same consul, when that incredible multitude had come to Rome --- almost Italy itself --- summoned you in the largest numbers up onto the Capitol. At which moment you could understand how much force the goodness of nature has, and true nobility. For Q. Metellus, both an enemy and the brother of an enemy, when he had perceived your wishes, set aside all his private hatreds. P. Servilius --- a most distinguished man and indeed the best, my warmest friend --- by a kind of divine weight of authority and of speaking, called him back to the deeds and virtues of his own line and of their common blood, so that he might have in his counsel both his brother (returned from the dead, the partner of my own things), and all the Metelli, the most outstanding citizens, almost called up out of Acheron --- among them that famous Numidicus Metellus, whose departure from his country once seemed honourable to all, but mournful nonetheless.

\ciceroSection{26} And so by divine providence Q. Metellus stood forth not only as the defender of my safety --- he who, until this one kindness, had been my enemy --- but even as a signatory of my standing. On that day, when you were four hundred and seventeen, and all the magistrates were present, only one dissented --- the man who, by his own law, thought even the Catilinarian conspirators were to be roused from the dead. And on that day, when by the gravest words and in the largest numbers you had judged that the commonwealth had been preserved by my counsels, the same consul saw to it that the same things were said by the leading men of the state in a public meeting on the next day; in which indeed he himself pleaded my case most splendidly, and brought it about, with all Italy standing by and listening, that no man could hear from any hireling or wretch a single bitter and hostile word against good men.

\ciceroSection{27} To these you yourselves added the rest --- not only the helps of safety but even the ornaments of standing. You decreed that no one should impede the matter on any pretext; if any did, you would consider it grievous and offensive --- he would be held to act against the commonwealth, the safety of good men, and the harmony of citizens, and the matter would at once be referred about him to you. And you ordered me to come back, even if longer they should pursue their pretexts. What more? --- that thanks be given to those who had come from the Italian towns? What more? --- that, on that day to which the matter had been postponed, they should be requested to come together with the same zeal? What more, finally, on that day which P. Lentulus appointed as a birthday for me and my brother and our children --- not only for the memory of our own time, but for the memory of all time --- on which day he summoned me to my country in the centuriate assembly (which our ancestors most particularly wished to be called and held as the proper assembly), so that the same centuries which had made me consul should approve my consulship ---

\ciceroSection{28} on that day what citizen was there who thought it right, of whatever age or health he might be, that he himself should not give his vote on my safety? When did you ever see so great a multitude on the field, so great a splendour of all Italy and of all the orders, when did you see returning officers and dispensers and guardians of the boxes of such standing? And so by the outstanding and divine kindness of P. Lentulus we were not led back into our country as some most distinguished citizens have been, but were brought back on horses with their distinguishing trappings and on a gilded chariot.

\ciceroSection{29} Can I ever be thought sufficiently grateful to Cn. Pompeius? --- who not only with you, who all felt the same, but with the whole Roman people, said that the safety of the Roman people was both kept up through me and bound together with mine; who commended my cause to the prudent, taught the inexperienced, and at the same time held in check the wicked by his authority, raised up the good; who exhorted the Roman people on my behalf as on behalf of a brother or a parent, and even entreated them; who, since he himself, fearing fighting and bloodshed, had kept to his house, then asked the previous tribunes to promulgate and refer about my safety; who, in a colony lately established, when he was himself holding the magistracy, in which there was no bought intercessor, set down the violence and cruelty of the privilegium against me by the authority of most honourable men and by public letters --- and was thought of as the leader of all Italy in calling on protection for my safety; who, though he had always been most loving to me, also worked to make his own intimate friends my friends.

\ciceroSection{30} With what services shall I repay the kindnesses of T. Annius? Whose whole reckoning, whole purpose, whole tribunate finally was nothing but a constant, perpetual, brave, undefeated defence of my safety. What shall I say of P. Sestius? --- who showed his good will and good faith toward me not only by the grief of his mind but by the wounds of his body. To you, senators, I have given thanks individually and shall give them: as a body I gave them at the start, so far as I could --- to give them ornately enough I cannot in any way. And though there are many men's special services to me, which can in no way be passed over in silence, still it is not for the present time and for my own fear to attempt to recall the kindnesses of individuals: for it is hard not to leave out somebody, and unspeakable to leave out anyone. I ought, senators, to honour you all together as if among the gods. But, as among the immortal gods themselves, we are not used always to worship and pray to the same ones, but now some, now others --- so it is among men who have done me divine services. All my life shall be enough to me for proclaiming and recalling their services to me.

\ciceroSection{31} Today, however, I have settled to give thanks by name to the magistrates, and out of private men to one only, who has gone to the Italian towns and colonies for my safety, has entreated the Roman people as a suppliant, has spoken the opinion which you, following him, have so given me back my standing. You always honoured me when I flourished; when I was hard pressed, by changing your dress and almost by your own grief, you defended me, so far as you were allowed. In our memory, senators were not used to change their dress even in their own perils: in my peril, the Senate did change its dress, so far as was permitted, even by the edicts of those who stripped my perils not only of their own protection but even of your intercession.

\ciceroSection{32} When these things had been thrown in my way, and I saw that I should have to fight, in private capacity, the same army I had as consul defeated not by arms but by your authority, I turned much over in my own mind. The consul had said in the public meeting that he would exact the penalties for the slope of the Capitol from the Roman knights. Some men were being attacked by name, some summoned, some banished. The approaches of the temples were being denied, not only by guards and force but even by demolition. The other consul, that he might not only desert me and the commonwealth, but even betray us to the enemies of the commonwealth, had bound himself by bargains for his own rewards. Another man was at the gates with command for many years and a great army, who I do not say was an enemy to me, but I know was silent when he was said to be an enemy.

\ciceroSection{33} Two parties were thought to exist in the commonwealth: the one was thought to be demanding me up out of enmity, the other timidly defending me out of suspicion of slaughter. But those who seemed to be demanding me up increased the fear of fighting in this respect: they never lessened men's suspicion and care by denying it. So when I saw the Senate bereft of its leaders, and myself partly attacked by magistrates, partly betrayed, partly abandoned --- saw slaves enrolled by name under the colour of guilds, saw all Catiline's forces called back to the hope of slaughter and burning under almost the same leaders, saw the Roman knights stricken by the fear of proscription, the towns of devastation, all by fear of slaughter --- I could have, senators, I could have, on the urging of many of the bravest men, defended myself by force and arms; nor did that same spirit of mine which is not unknown to you fail me. But I saw that, if I should conquer the present adversary, I had altogether too many others to conquer; if I should be conquered, many good men must perish both for me and with me, and even after me; that the avengers of the tribune's blood were ready, and that the punishment for my death was being held back for trial and posterity.

\ciceroSection{34} I would not, when as consul I had defended the common safety without the sword, defend my own as a private man with arms; I preferred that good men mourn my fortunes than despair of their own. And, if I alone had been killed, it seemed shameful to me; if with many, mournful for the commonwealth. If I had thought everlasting hardship had been set before me, I should rather have inflicted death upon myself than an everlasting grief. But when I saw that I should not be longer absent from this city than the commonwealth itself, neither did I think I had to remain when she had been driven out, and she, as soon as she was recalled, brought me back along with her. With me the laws were absent, with me the courts, with me the rights of magistrates, with me the authority of the Senate, with me liberty, with me even the abundance of crops, with me all the holinesses and religions of gods and men. If these things were always to be absent, I should rather mourn your fortunes than long for my own; but if at some time they were to be recalled, I understood that I had to come back along with them.

\ciceroSection{35} The most certain witness of this feeling of mine is this same man who was the guardian of my life, Cn. Plancius, who, having set aside all the trappings and conveniences of provincial office, set down his whole quaestorship in keeping me up and saving me. Who, had he been my quaestor when I was a commander, would have been in the place of a son: now will surely be in the place of a parent, since he has been the partner not of my command but of my grief.

\ciceroSection{36} For which reason, senators, since I have been restored to the commonwealth on the same terms as the commonwealth itself has been restored, I shall not in defending it lessen anything of my old freedom of speech --- I shall enlarge it. For if I defended that very thing when she owed me something, what ought I to do now, when I owe her the most? For what is there that can break or weaken my spirit, when you see the very calamity itself is a witness, not of any wrongdoing of mine, but of my divine services to the commonwealth? For it was brought upon me because I had defended the citizenry; and undertaken by my own will, that the commonwealth, defended by me, should not through me again be called into the last danger.

\ciceroSection{37} For me, not as for P. Popilius, that most noble man, did young sons, did a multitude of kinsmen entreat the Roman people; not as for Q. Metellus, the highest and most distinguished man, did a son already of proven youth, did the consulars L. and C. Metellus, did their children, did Q. Metellus Nepos who was then standing for the consulship, did the Luculli, the Servilii, the Scipiones, the daughters of the Metelli weeping and in mourning supplicate the Roman people; but one brother, who in his devotion was a son to me, in counsel a parent, in love (as he was) a brother --- by squalor and tears and daily prayers compelled the longing for my name to be renewed, and the memory of my deeds done to be brought back into use. He had resolved that, unless he had recovered me through you, he would undergo the same fortune and demand the same dwelling-place for life and death. Yet he never feared the greatness of the business, his own loneliness, the violence of the enemies, or their weapons.

\ciceroSection{38} The other was the championing defender of my fortunes, in the highest courage and devotion --- C. Piso, my son-in-law, who set at nothing the threats of my enemies, the enmities of his kinsman by marriage, his own kinsman, the consul, and Pontus and Bithynia, his quaestorian province, before my safety. The Senate never decreed anything for P. Popilius; never in this order was there mention of Q. Metellus. They were restored at last, by the killing of their tribunician enemies, by tribunician bills --- when one of them had obeyed the Senate, the other had fled violence and slaughter. As for C. Marius indeed, the third consular, in this men's memory before me, driven out by the storm of civil war, he was not only not restored by the Senate, but by his very return almost wiped out the whole Senate. There was no agreement of the magistrates about them, no summoning of the Roman people for the defence of the commonwealth, no movement of Italy, no decrees of the towns and colonies.

\ciceroSection{39} Therefore, since your authority has summoned me, the Roman people has called me, the commonwealth has implored me, all Italy has carried me back almost on its own shoulders --- I shall not let it happen, senators, that, since those things have been given me back which were not in my own power, I should not have those things which I myself can supply: especially since, having recovered what was lost, I never lost courage and good faith.
