% Pro L. Flacco (pro-flacco) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.
%
% Note: this speech survives partly through medieval manuscripts
% (sections 1-5, 6-106, with two fragments at the end) and partly
% through Bobiensian palimpsest fragments (frBob1-12, frMed) which
% fill the lacuna between sections 5 and 6. The fragments are
% rendered here as labelled.

\ciceroSection{1} When in the greatest dangers of this city and empire, in the gravest and bitterest disaster of the commonwealth, with Lucius Flaccus as the partner and helper of my counsels and my dangers, I was driving slaughter from you, your wives, your children, devastation from the temples, the shrines, the city, Italy, I hoped, judges, that I should rather be a helper of Lucius Flaccus to honour than a beseecher against his miseries. For what reward of dignity could the Roman people refuse to him, since to his ancestors it had always paid such rewards, when Lucius Flaccus, after almost five hundred years of the commonwealth, had recovered the old praise of the Valerian house in liberating the country?

\ciceroSection{2} But if perchance there should at some time arise a detractor of this kindness, or an enemy of his virtue, or one envious of his praise, I supposed that Lucius Flaccus would have to undergo the judgment rather of an unskilled multitude --- and that with no danger --- than of the wisest and most chosen men. For where the safety, then, not of citizens alone but of nations had been by the same authority and defence held and saved, I never thought that anyone, by their hands, would create danger for this man's fortunes and plots against him. And if it should at some time come to pass that someone should think of Lucius Flaccus's destruction, yet I never supposed, judges, that Decimus Laelius, the son of an excellent man, himself endowed with the best hope of the highest dignity, would take up an accusation that suited the hatred and madness of wicked citizens rather than his own virtue and the youth he had set himself. For when I had seen the most distinguished men often laying down most just enmities with citizens of good service, I did not suppose that any friend of the commonwealth, when Lucius Flaccus's love for his country had been seen, would proclaim a new enmity against him with no injury received.

\ciceroSection{3} But since, judges, many things have deceived us both in our private affairs and in the commonwealth, we bear what is to be borne. This much we ask of you: that you reckon all the resources of the commonwealth, the whole standing of the state, all the memory of times past, the safety of the present, the hope of what is to come, to be set down and fixed in your power, in your votes, in this one trial. If ever the commonwealth has implored the counsel, the gravity, the wisdom, the foresight of judges, at this time, this time I say, it implores them. You are not to give judgment about Lydians or Mysians or Phrygians, who have come driven and stirred up to this place, but about your own commonwealth, about the standing of the state, about the common safety, about the hope of all good men --- if any such hope still remains to support the minds and thoughts of brave citizens. All the other refuges of good men, defences of the innocent, supports of the commonwealth, counsels, aids, rights have fallen.

\ciceroSection{4} For whom shall I call upon, whom shall I beseech, whom shall I implore? The senate? But that very body itself seeks help from you, and feels that the confirmation of its authority has been entrusted to your power. The Roman knights? You will declare to fifty leading men of that order what your decision concerning all is. The Roman people? But it has handed over to you all its power over us. So that, unless in this place, unless before you, unless through you, judges, we hold not our authority --- which is lost --- but our safety, which hangs by a slender and last hope, there is nothing else to which we can flee. Unless, perhaps, you do not see what is being attempted in this trial, what is being done, the foundations of what cause are being laid down, judges.

\ciceroSection{5} Condemned has been the man who killed Catiline as he was bringing his standards against the country: what is there to keep the man who drove Catiline out of the city from being afraid? He is hurried to punishment who took down the proofs of the common ruin: why should the man feel safe who took care that they should be brought forward and exposed? The partners of his counsels, the agents and companions, are harassed: what should the authors, the leaders, the chiefs expect for themselves? And would that our enemies, and the enemies of all good men, would weigh together with me whether at that time all good men were our leaders or our companions in the saving of the common safety!

\ciceroSection{frBob1} \textit{[He preferred to say strangled.]}

\ciceroSection{frBob2} \textit{[What did this connection of mine intend by the small shield?]}

\ciceroSection{frBob3} \textit{[And what indeed of Decianus?]}

\ciceroSection{frBob4} \textit{[Would that it were properly mine! The senate then for the most part \ldots]}

\ciceroSection{frBob5} \textit{[The immortal gods, I say, [drove] Lentulus \ldots]}

\ciceroSection{frMed} \textit{[\ldots since this foreign matter agreed with his domestic life and nature. So I shall not allow you, Decimus Laelius, to take this on yourself, and to lay this down as a law and condition for the rest in time to come, for us in the present time: that, when you have marked his youth, when you have spattered the rest of his life with stains of disgrace, when you have brought forward the ruins of his private affairs, the blots in his household, the ill-fame at Rome, the vices and infamies of his time in Spain, Gaul, Cilicia, Crete --- provinces where he was conspicuously employed --- only then shall we hear what the men of Tmolus and the Doryleans think of Lucius Flaccus. The man whom so many and so weighty provinces wish safe, whom very many citizens from all Italy, bound to him by old ties of duty, defend, whom this country common to us all holds embraced for the recent memory of the highest service --- if even all Asia were demanding him for punishment, I should defend, I should resist. What if neither all of it nor the best part of it nor the uncorrupted nor of its own free will, neither rightly nor by custom nor truly nor scrupulously nor uprightly, but if driven, if solicited, if stirred up, if compelled, if impiously, if rashly, if greedily, if inconsistently it has sent its name into this trial through the most needy of witnesses, while it can truly complain of no injuries of its own --- shall, even so, judges, these things heard for a moment take away credit from things known long ago? I shall keep, then, this order as defender which the prosecutor has fled, and shall press the prosecutor and pursue him, and demand the charge unprompted from my opponent. What is it, Laelius? \ldots Did you mean anything by saying \ldots a man who has not lived in the shade, nor in the discipline and arts of that age? For as a boy he set out to the war when his father was consul. And by this very name, no doubt, something \ldots]}

\ciceroSection{frBob6} \textit{[But if neither the luxury of Asia, [overcame] the most weak time of his age \ldots]}

\ciceroSection{frBob7} \textit{[From this stage of his age he betook himself to the army of his uncle Gaius Flaccus.]}

\ciceroSection{frBob8} \textit{[Set out as military tribune with Publius Servilius, that gravest and most scrupulous of citizens.]}

\ciceroSection{frBob9} \textit{[Adorned by their fullest commendations, he was made quaestor.]}

\ciceroSection{frBob10} \textit{[\ldots Marcus Piso, who, if he had not received the cognomen of frugality, would himself have made one.]}

\ciceroSection{frBob11} \textit{[The same man undertook a new war and finished it.]}

\ciceroSection{frBob12} \textit{[Handed over not to the witnesses of Asia, but to its prosecutors as messmates.]}

\ciceroSection{6} This man, then, Laelius --- with what charges do you finally assault him? He was military tribune in Cilicia under the imperator Publius Servilius: that is passed over in silence. He was quaestor to Marcus Piso in Spain: no word about the quaestorship is let fall. He carried on the Cretan war for a great part of it, and bore it up alongside the supreme commander: on this period the prosecution is mute. The praetorian jurisdiction, a varied and many-sided business, open to suspicions and bad blood, is not touched. But in the gravest and most dangerous moment of the commonwealth's life, even by his enemies that very praetorship is praised. ``But he is hurt by the witnesses.'' Before I say by whom, by what hope, by what force, by what means stirred up, with what fickleness, what destitution, what perfidy, what audacity furnished, I shall speak of the whole class and of the condition of us all. By the immortal gods, judges! how the man who a year before had administered law at Rome was administering law in Asia a year later --- will you inquire from witnesses you do not know, and decide nothing yourselves by conjecture? When in so varied a jurisdiction, with so many decrees passed and the wills of so many influential men crossed, what voice was ever launched --- not of suspicion, which generally is false, but of anger or grief?

\ciceroSection{7} And is this the man on trial for avarice, who in the richest opportunity has escaped any base gain, and in a most slanderous community, in business most exposed to suspicion, has escaped every reproach, not only the formal charge? I pass over what must not be passed over: that no avaricious act of his in private affairs, no contention about money, no sordid spot in his household, can be brought forward. By what witnesses, then, can I refute these men if not by you?

\ciceroSection{8} Will that village fellow from Tmolus, a man known not only not to us but not even among his own people, teach you what sort of man Lucius Flaccus is? --- the man whom you knew as a most modest youth, as the most upright of men in the greatest of provinces, as the bravest soldier of your army, the most diligent commander, the most temperate legate and quaestor; the man whom you yourselves, in his presence, judged to be the most steadfast of senators, the most just of praetors, and a citizen most devoted to the commonwealth.

\ciceroSection{9} Of those for whom you ought to be witnesses to others, will you yourselves listen to other men as witnesses against them? And what witnesses? First, what is the common ground: Greeks. Not that I above all single men would withdraw faith from this nation; for if any of our people was ever not averse to that race in interest and goodwill, I think I am one, and was more so when there was more leisure. But in that number there are many good men, learned, modest, who have not been led to this trial; many shameless, unlettered, fickle, whom I see stirred up for various reasons. Yet this I say of the whole class of Greeks: I grant them letters, I allow the discipline of many arts, I do not deny them charm of speech, keenness of intellect, abundance of expression --- and finally, if they claim anything else for themselves, I do not contend. But the scruple and good faith of testimony that nation has never cultivated, and they do not know what the force, what the authority, what the weight of this whole matter may be.

\ciceroSection{10} Whence comes the saying ``lend me your testimony in return''? Is it thought to be the Gauls', or the Spaniards'? It is the Greeks' altogether --- so much so that even men who do not know Greek know in what words this is usually said by Greeks. Watch, then, with what face, with what assurance they speak; then you will understand with what scruple. They never answer us to the question, always more than is asked to the accuser; they never trouble how to prove what they say, only how to extricate themselves in saying it. Marcus Lurco gave evidence angry with Flaccus, because, as he himself said, his freedman had been condemned in a shameful trial. He said nothing that would wound him, though he wished to; for scruple stood in the way: yet what he did say --- with what shame, with what trembling, with what pallor he said it!

\ciceroSection{11} How ready a man was Publius Septimius, how angry about the trial and about the bailiff! Yet he hesitated; yet now and then scruple fought against his anger. Marcus Caelius was hostile because, when in a plain matter he had thought it sacrilege for a tax-farmer to judge against a tax-farmer, he had been struck off the panel of recuperatores; yet he held himself back and brought nothing into court to wound except his disposition. Had these men been Greeks, and had not our customs and discipline weighed more than pain and feud, every one of them would have said he was despoiled, harassed, ruined in fortune. A Greek witness has come forward with the will to wound; he rehearses not the words of his oath but the words of his injury; to be beaten, refuted, convicted he reckons the most disgraceful thing; to that he prepares himself, he cares for nothing else. And so the man chosen is not the best or the gravest, but the most shameless and the most talkative.

\ciceroSection{12} You, on the other hand, in private trials over the smallest matters, weigh a witness with care; even when you know the man's looks, his name, his tribe, you still think his character must be examined. But the man who gives evidence from among our people --- how he holds himself in, how he moderates every word, how he fears to say anything from greed, anything from anger, anything more or less than is necessary! Do you suppose them likewise --- men to whom an oath is a joke, testimony a game, your good opinion mere darkness, men to whom praise, payment, favour, congratulation lie all set out in shameless lying? But I shall not stretch my speech out: it could indeed be infinite, if I cared to unfold the levity of the whole race in giving testimony. I shall draw closer: I shall speak of these witnesses of yours.

\ciceroSection{13} We have run up against a vehement prosecutor, judges, and an enemy hateful and troublesome in every kind --- whom I hope, with these sinews of his, will be of great use both to his friends and to the commonwealth. But certainly he has taken up this case and accusation inflamed by an incredible eagerness. What an entourage he had in his investigation! Entourage, do I say? --- rather, what an army! What outlay, what expense, what largesse! Yet, however useful these things are to my case, I speak of them with hesitation, because I fear that Laelius may think that something of personal slight or invidious feeling has been sought against him in my speech, in matters which he undertook for his own glory's sake. I shall therefore leave this whole part alone; only this I shall ask of you, judges --- that, if you have heard anything yourselves by common report and conversation, of force, of armed bands, of arms, of troops, you remember it. It is for fear of the resentment these things stir up that, under this new and recent law, a fixed number of assistants in the investigation has been laid down.

\ciceroSection{14} But, to set aside this force --- how great are the things which, because they have been done under accusatorial right and custom, we cannot rebuke, but are forced to lament! First, that talk has been scattered all over Asia that Gnaeus Pompeius, because he was bitterly hostile to Lucius Flaccus, pressed Laelius --- his paternal friend, his close connection --- to bring this man to trial, and conferred on him his whole authority, favour, resources, and means for accomplishing this business. This seemed all the more probable to Greek minds because a little before they had seen Laelius on intimate terms with Flaccus in the same province. And Pompeius's authority, great as it is everywhere among all men as it ought to be, stands out in that province in particular, which he recently freed from the war both of pirates and of kings. He added this further: that he terrified those unwilling to leave home with a summons to give evidence, and stirred those who could not stand still at home with a generous and liberal travel-allowance.

\ciceroSection{15} So this young man, full of talent, moved the wealthy by fear, the poor by reward, the foolish by error; so were extorted those celebrated decrees that are read out --- declared by no votes nor by any authoritative bodies, bound by no oath, but by hand stretched out and the shout of a stirred-up crowd. O what a fine custom and discipline we have received from our ancestors, if only we kept hold of it! But somehow now it slips out of our hands. For none of our wisest and most scrupulous men wanted any force to belong to a public meeting; what the plebs decreed or what the people commanded, they wanted done with the public meeting dismissed, the parts distributed, the orders set out tribe by tribe, century by century, by classes, by ages, with the proposers heard, and the measure promulgated and known for many days.

\ciceroSection{16} But the entire commonwealths of the Greeks are governed by the recklessness of a seated assembly. And so, to leave aside this present-day Greece, which has long ago been struck and dashed by its own counsels, that ancient Greece which once flourished in resources, in empire, and in glory fell by this single evil --- the immoderate freedom and licence of public meetings. When the unskilled men, raw and ignorant of all things, sat down in the theatre, then they undertook useless wars, then they set seditious men over the commonwealth, then they cast out of the city the citizens who deserved best of her.

\ciceroSection{17} If these things were apt to happen at Athens, when she shone not only in Greece but among almost all nations, what moderation do you suppose there was in the assemblies of Phrygia and Mysia? Men of those nations more often than not throw our own assemblies into uproar; what do you finally suppose happens when they are by themselves? That Athenagoras of Cyme was beaten with rods, who had dared to export grain in a famine. An assembly was given to Laelius. He came forward; and he, a Greek among Greeks, did not speak of his own fault but complained of his punishment. They held out their hands; the decree was born. Is *this* testimony? The men of Pergamum, lately feasted, a little before sated with every largesse, since Mithridates, who held that crowd not by his own authority but by stuffing them with food, said he wished it, the cobblers and belt-makers shouted out for it. Is this the testimony of a community? I myself brought witnesses publicly out of Sicily; but those were the testimonies, not of an assembly stirred up, but of a sworn senate.

\ciceroSection{18} And so I have no contest now with the witness; it is for you, judges, to see whether these things are to be reckoned testimonies. A young man of good birth, of decent family, eloquent, comes into a Greek town with the largest and most splendid retinue, calls for an assembly, terrifies the wealthy and weighty men with the threat of a summons to give evidence so that they may not stand against him, and entices the poor and fickle with the hope of largesse, with public travel-money, with private favour as well. To stir up the artisans and shopkeepers and all that dregs of the cities --- what trouble is that, particularly against a man who recently held a power of the highest, but who, on the very name of that power, could not be the object of the highest love?

\ciceroSection{19} It is no wonder, surely, that men to whom our axes are an object of hatred, our magistrate's name a thing of bitterness, our pasture-tax, our tithes, our customs-dues a thing of death, lay greedy hands on whatever facility for hurting is held out to them! Remember, then, when you hear these decrees, that you are not hearing testimonies, you are hearing the recklessness of the mob, you are hearing the voice of every fickle man, you are hearing the noise of the unskilled, you are hearing a stirred-up assembly of the most fickle of nations. And so probe to the bottom the nature and reason of these charges; you will find now nothing but hope, nothing but terror and threats.

\ciceroSection{20} The cities have nothing in their treasury, nothing in their revenues. There are two ways of raising money: either a loan or a levy; yet no creditor's account-book is brought forward, nor is any record of a levy read out. How easy it is for them to enter false accounts and to record in the books whatever is convenient, learn, please, from a letter of Gnaeus Pompeius to Hypsaeus and from one of Hypsaeus to Pompeius. \textit{[Letters of Pompey and Hypsaeus.]} Do we seem to you, with these authorities, to have shown sufficiently the dissolute habits of the Greeks and their shameless licence? Unless, perhaps, we are to think that men who deceived Gnaeus Pompeius --- and that with him on the spot, with no one driving them to it --- were either timid or scrupulous when Laelius was pressing them, against an absent man and against Lucius Flaccus.

\ciceroSection{21} But suppose the books were uncorrupted at home; what authority or what credit can they have now? The law orders them to be brought to the praetor within three days, sealed with the judges' seals; on the thirtieth day they are scarcely brought. So that the books may not easily be tampered with, the law required them to be deposited under seal in a public place; but they are sealed already tampered with. What does it matter, then, whether they are brought before the judges so long after, or not brought at all? And what if the witnesses' enthusiasm is yoked with the prosecutor's --- shall they still be reckoned witnesses? Where, then, is that expectation that is wont to attend trials? For in the past, when the prosecutor had spoken sharply and vehemently, and the defender had answered humbly and submissively, the third place looked for was that of the witnesses, who spoke either without any partisanship at all, or with at least some dissimulation of their eagerness. But now what is this?

\ciceroSection{22} They sit together; they rise from the prosecutors' benches; they do not dissemble, they do not blush. Do I complain of the benches? Out of one house they come forward together; if they trip in a word, they will have nowhere to go back to. Can a man be a witness, then, whom the prosecutor questions without anxiety, and does not fear that he may answer something the prosecutor himself does not wish? Where, then, is that praise of the orator that used to be looked for either in the prosecutor or in the defender: ``He has questioned the witness well; he has approached him cleverly, taken him up; brought him where he wished; convicted him and reduced him to silence''?

\ciceroSection{23} What can you ask, Laelius, of a man who, before you have said ``I ask you,'' will pour out more than you yourself prescribed for him beforehand at home? But what can I, the defender, ask? for either the witness's speech is usually refuted, or his life touched. By what argument shall I refute the speech of one who says ``We gave,'' nothing more? One must speak against the man, then, when his speech offers no ground for argument. What shall I say against an unknown? It must be lamented, then, and bewailed, as I have long been doing, the whole inequity of this prosecution: first the common class of the witnesses, since the nation speaks that is the least scrupulous in the giving of testimony. I draw closer: I deny that those things are testimonies which you call decrees, and say they are the murmur of the destitute and a kind of reckless commotion of a Greek assembly. I shall enter further still. The man who handled the affair is not present; the man who is said to have counted out the money has not been brought; no private letters are produced; the public ones have been kept back in the prosecutors' power; the chief weight is in the witnesses; these men live with the enemies, sit with the adversaries, lodge with the prosecutors.

\ciceroSection{24} Do you suppose this will produce a real examination and uncovering of the truth, or some stain upon innocence and ruin of it? For there are many things of this kind, judges, that even if they may be neglected in the man whose case is at issue, must yet be feared in the precedent and the example they set. If I were defending some man born in the lowest place, of no splendour of life, with no commendation of fame, I should still beg of his fellow-citizens, by the common right of humanity and by mercy, that they not deliver one citizen to other citizens by witnesses unknown, witnesses stirred up, witnesses who are the prosecutor's bench-mates, dinner-companions, tent-mates --- to men Greek in fickleness and barbarian in cruelty --- nor hand on a dangerous imitation of the precedent to the rest hereafter.

\ciceroSection{25} But when it is the case of Lucius Flaccus that is being tried, of that house from which the man who was made first consul was the first consul in this state; by whose virtue, with the kings driven out, liberty was established in the commonwealth; which has lasted to this very day in an unbroken line of offices, commands, and the glory of deeds done; and when Lucius Flaccus, far from degenerating from this perennial and well-witnessed virtue of his ancestors, has, just as he saw his family's glory most flourishing, fallen in love with the praise of his country in setting his country's freedom free as praetor --- shall I in this defendant fear that some pernicious precedent will be handed on, when, even if he had erred in something, all good men would think that they ought to wink at it?

\ciceroSection{26} Yet for my part I do not even ask this; on the contrary, judges, I beg and beseech you that you contemplate the whole case as keenly as you can, with eyes most intent (as the saying is). Nothing will be found vouched for by scruple, nothing grounded on truth, nothing wrung out by pain; on the contrary, all will be found corrupted by lust, by anger, by partisanship, by money, by perjury.

\ciceroSection{27} Indeed, since now the universal eagerness of these men has been recognized, I shall come to the particular complaints and accusations of the Greeks. They complain that money was levied on the cities under the name of a fleet. We confess, judges, that this was done. But if this is a charge, it lies either in his not being permitted to make the levy, or in there being no need of ships, or in no fleet's having sailed under this praetor. That it was permitted, that you may understand, hear what the Senate decreed in my consulship --- when in fact it departed in nothing from the decrees of the continuous earlier years. \textit{[Decree of the Senate.]} It is next, then, for us to inquire whether or not there was need of a fleet. Shall, then, the Greeks decide this, or any foreign nations, or our praetors, our generals, our imperatores? For my own part I think that, in such a region and province as that, girt by sea, broken into harbours, encircled by islands, there had to be sailing not only for safety's sake but for the adornment of empire as well.

\ciceroSection{28} For this was the way and the greatness of mind in our ancestors --- that, while in private affairs and on their own outlay they lived content with the smallest, in the slenderest mode of life, in command and public dignity they referred everything to glory and splendour. For in domestic affairs the praise sought is for restraint, in public affairs for dignity. And if he had a fleet for safety's sake too, who will be so unfair as to find fault? ``There were no pirates.'' What of it? Could anyone guarantee there would be none? ``You are diminishing,'' he says, ``the glory of Pompeius.'' Nay rather, you are increasing the trouble.

\ciceroSection{29} For Pompeius cleared away the fleets of the pirates, their cities, their harbours, their hideouts; he completed the peace of the seas with the highest virtue and incredible speed. But that other thing he did not undertake, nor ought he to have undertaken --- that, if a small pirate-boat had appeared anywhere, he should be thought blameworthy. And so he himself in Asia, when he had completed every war by land and sea, still levied a fleet from these very same cities. And if at that time he judged a fleet necessary --- when by his presence and his name alone all could be safe and at peace --- what do you think had to be decided and done by Flaccus, when Pompeius had departed?

\ciceroSection{30} What? Did we not, on Pompeius's own motion, in the consulship of Silanus and Murena, decree that a fleet should sail in Italy? Did we not at that very time, when Lucius Flaccus was levying rowers in Asia, here pay out four million three hundred thousand sesterces for the upper and the lower seas? What? In the next year, was not money paid out to Marcus Curtius and Publius Sextilius the quaestors for a fleet? What? Were there not horsemen on the maritime coast all this time? For this is the divine glory of Pompeius: first, that those pirates who at the time when the maritime command was given him were straggling scattered over the whole sea, were all reduced into the power of the Roman people; then, that Syria is ours, that Cilicia is held, that Cyprus through king Ptolemy ventures nothing; and besides this, that Crete is ours by Metellus's virtue, that there is nowhere they may set out from, nowhere they may return to, that all the bays, headlands, shores, islands, and seaboard cities are kept under the bars of our empire.

\ciceroSection{31} If under the praetor Flaccus there had been no pirate at sea, his diligence would still not have to be blamed. For I should not consider that there was none for the reason that this man kept a fleet. What, then, if I prove by the testimony of Lucius Eppius, Lucius Agrius, Gaius Cestius, Roman knights, and of this most illustrious man Gnaeus Domitius, who was at that time legate in Asia, that at the very time when you say a fleet ought not to have been kept, several persons were captured by pirates --- shall Flaccus's plan in levying rowers still be censured? What if a noble Adramyttene, whose name is known to almost all of us, Atyanas the boxer, an Olympic victor, was even killed by the pirates? This is, among the Greeks --- since we are speaking of their seriousness --- almost greater and more glorious than to have triumphed at Rome. ``But you took no one.'' How many famous men have presided over the maritime coast, who, though they took no pirate, yet kept the sea safe! For there is chance in the taking --- place, wind, opportunity; the precaution of defence is easy, not only by the lurking-places of hidden spots but by the moderation and turning of the weather.

\ciceroSection{32} It remains to ask whether that fleet of yours sailed by oar and course, or only on paper and at expense. Can it be denied --- when all Asia is witness to it --- that the fleet was divided in two parts, one of which sailed above Ephesus, the other below? On this fleet that great man Marcus Crassus sailed from Aenus to Asia, on these ships Flaccus sailed from Asia to Macedonia. Where, then, is the diligence of the praetor to be looked for? In the number of the ships and in the equitable distribution of expense? Half of what Pompeius had levied he levied; could he have done it more sparingly? And he distributed the money on Pompeius's own scheme, which was itself patterned on Lucius Sulla's scheme. When Sulla had distributed the money on a proportionate scale through all the cities of Asia, both Pompeius and Flaccus followed that scheme in levying the expense. Nor is even that sum yet completed.

\ciceroSection{33} ``He does not enter it.'' True; what does he gain by it? For when he undertakes the burden of the levy, he is confessing the very thing you wish to be a charge. How, then, can it be shown that, by not entering this money in the books, he was making a charge for himself in a matter where there would have been no charge if he had entered it? But you say that my brother, who succeeded Lucius Flaccus, levied no money for rowers. I delight in every praise of my brother --- but in others heavier and greater. He decided one thing; he saw something else: he thought that whenever any rumour of pirates was heard, he could put a fleet together as quickly as he wished. In short, my brother first did this in Asia --- to relieve the cities of this expense for rowers; but a charge usually appears when someone establishes outlays which were not established before, not when a successor changes something in his predecessors' arrangements. Flaccus could not know what others would do afterwards; he saw what they had done.

\ciceroSection{34} But, since I have spoken of the common charge of all Asia, I shall come now to the individual cities; and let the first of them for us be the city of the Acmonenses. The crier calls in his loudest voice the legates of the Acmonenses; one Asclepiades comes forward. ``Let the rest come.'' Have you compelled the very crier to lie? For this man, I suppose, is so weighty as to sustain the name of his city by his own authority --- a man condemned at home in the most disgraceful trials, branded by the public records, the records of whose disgraces, adulteries, and debaucheries from the Acmonenses are still in existence; which I think must be passed over not only for their length but for the most disgraceful obscenity of their language. He says that 206,000 drachmae were given publicly. He has merely said it, shown nothing, brought forth nothing; but he added what, since it was a domestic matter, he certainly ought to have proved --- that he himself privately gave 206,000 drachmae. The shameless wretch never even dared to wish he had as much as he says was taken from him.

\ciceroSection{35} He says he gave it through Aulus Sextilius and through his own brothers. Sextilius could give; for the brothers, indeed, are partners in beggary. Let us hear Sextilius, then; let the brothers themselves come forward; let them lie as shamelessly as they will, and say that they gave what they never had; perhaps, brought up in person, they will say something in which they may be caught out. ``I have not brought Sextilius,'' he says. Hand over the books. ``I have not brought them.'' At least produce the brothers. ``I did not summon them.'' Are we, then, to fear as a charge or a piece of testimony what one Asclepiades --- needy in fortune, foul in life, condemned in repute --- has thrown out, relying on shamelessness and audacity, with no books and no authority?

\ciceroSection{36} The same man said that the testimonial which we were producing as given by the Acmonenses to Flaccus was false. The loss of that testimonial we should have prayed for. For when this distinguished representative of his own city had inspected the public seal, he said that his fellow-citizens and the rest of the Greeks were in the habit of sealing on the spur of the moment whatever was needed. Keep the testimonial then for yourself; for Flaccus's life and dignity does not lean on the testimony of the Acmonenses. You give me the very thing this case demands most: that there is no weight, no consistency, no firm judgment in Greek men, and finally no good faith in their testimony. Unless, indeed, your formula of testimony and speech can be drawn and distinguished as follows: that the cities are said to have granted Flaccus, in his absence, something; while when Laelius was present, in person pressing the case under the force of the law, by the right of accusation, terrifying and threatening with his own resources besides --- they wrote and sealed nothing for the moment.

\ciceroSection{37} For my part, judges, I have often seen great matters caught and held in the smallest things --- as in this Asclepiades. The testimonial we have produced was sealed with that Asiatic clay which is generally familiar to all of us, which all use not only in public letters but in private ones too which we daily see being sent by the tax-farmers and often to each one of us. Nor did the witness himself, after inspecting the seal, say we had produced a forgery; he simply put forward the levity of all Asia, on which we both freely and easily concede. Our testimonial, then, which he says was given to us for the moment, but which he confesses was given, is sealed with clay; in that other testimony said to have been given to the prosecutor we have seen wax.

\ciceroSection{38} Here, judges, if I thought you were to be moved by the decrees of the Acmonenses or by the documents of the rest of the Phrygians, I would shout out and contend with all my might, I would call the tax-farmers to witness, I would arouse the men of business, I would even implore your own knowledge; with the wax detected, I should be confident that the made-up audacity of the whole testimony had been openly seized and overthrown. As it is, I shall not press in any harder, nor flit about in this matter the more insolently, nor inveigh against this trifler as if he were some real witness, nor shall I dwell on the whole testimony of the Acmonenses, whether it was forged here, as it appears, or sent from home, as is said. For if I am content to remit the testimonial to men whom Asclepiades calls fickle, I shall not fear their testimony.

\ciceroSection{39} I come now to the testimony of the Doryleans; who, when produced, said they had lost their public records at some caves. O what shepherds these must be, and how greedy of writings, if it is true that they took nothing from these men but writings! But we suspect there is another reason: namely, that these men should not appear too unsubtle. The penalty, I think, at Dorylaeum is heavier than elsewhere for false and corrupted records. If they had brought forward genuine ones, there was no charge; if forged ones, there was a penalty. They thought it best to say they were lost.

\ciceroSection{40} Let them keep quiet, then, and let me put this in my profit-account and turn to other things. They will not. For some fellow comes in to fill the place and says he gave money privately. This now can in no way be borne. The man who reads from public records that have been in the prosecutor's power ought not to have authority; yet in some sort there seems to be a trial when those very records, of whatever kind, are produced. But when the man whom none of you have ever seen, whom no mortal has heard of, says only ``I gave,'' will you hesitate, judges, to vindicate a most noble citizen against this most obscure Phrygian? And the same man, lately, when in a suit about freedom three honourable and weighty Roman knights tried him --- he saying that the man whose freedom was being claimed was his kinsman --- they did not believe him. How does it come about that he who was no creditable witness for the pain and the blood of his own household is yet a weighty authority for a public injury?

\ciceroSection{41} And when this Dorylean was lately being carried out in great press and concourse of you all, Laelius was loading the odium of his death onto Lucius Flaccus. You do unjustly, Laelius, if you think your own bench-mates live at our peril; especially since we suppose it was through your own carelessness it happened. For you set out a basket of figs before a Phrygian who had never seen a tree. His death has lightened you of something --- you have lost a gluttonous guest; but how has it benefited Flaccus? The man who was strong enough to come this far, died after the sting was discharged and the testimony delivered. But that very pillar of your prosecution, Mithridates, when he had been kept by us as a witness for two days and had poured out everything he wanted, retired thwarted, refuted, and broken; he goes about with a corselet on. Our learned and wise gentleman fears that Lucius Flaccus may now bind himself in crime, when already he cannot escape that witness; and that the man who held himself in before delivering his testimony --- when he might still gain something by it --- now sets about adjoining a real charge of misdoing to a false charge of avarice. But since Quintus Hortensius has dealt subtly and at length with this witness and the whole Mithridatic charge, let us, as we set out, go on to the rest.

\ciceroSection{42} The chief of all the Greek-stirrers, who sits with the prosecutors, is that Heraclides of Temnus --- a foolish and talkative man, but, as he himself thinks, so learned that he calls himself the master even of the others. But this man, ambitious enough to greet you all and us daily, until that age of his could not get into the senate at Temnus, and he who professes to hand on the art of speaking even to the rest has himself been beaten in all the most disgraceful trials.

\ciceroSection{43} Of equal good fortune, the legate Nicomedes came along with him, who could on no terms come into the senate, and who was condemned for theft and for misconduct as partner. The leader of the legation, Lysanias, did indeed reach the senatorial order, but, when he embraced too much of public business, was condemned for embezzlement and lost both his property and his senatorial title. These three even wished to falsify the records of our treasury; for they professed they had nine slaves, when they had come without any companion at all. In the first place I see that Lysanias was present at the writing of the decree; whose brother's property, because he was not paying the people, was sold publicly under the praetor Flaccus. Besides there is Philippus, son-in-law of Lysanias, and Hermobius, whose brother Pollis was likewise condemned for public money. They say that they gave to Flaccus and those with him 2,200,000 drachmae.

\ciceroSection{44} I have to do here with a city most exact and most thorough in its accounts, in which not a single coin can be moved without five praetors, three quaestors, and four bankers, who are elected from the people among them. Out of this whole great number not a single one has been brought. They make out, indeed, that this money was given to Flaccus by name; while another and larger sum, when they were giving it to this same man, they say they entered as for the rebuilding of a sacred temple; which is most inconsistent. For either everything had to be entered in secret, or everything in the open. When they enter it openly to Flaccus, they fear nothing, they are afraid of nothing; when they enter it under a public work, the same men suddenly fear the same man whom they had despised. If the praetor gave it, as it is written, he paid through the quaestor; the quaestor through the public bank; the bank either out of revenue or out of a levy. There will never be anything resembling a charge in this, unless you unfold for me this whole account, with both the kinds of persons and of records.

\ciceroSection{45} Or take what is written in the same decree --- that the most distinguished men of the city, who had held the highest offices, were defrauded under this praetor. Why are they not present in court, why not named in the decree? For I do not believe that he who lifts himself up, Heraclides, is meant in that passage. For is *he* among the most distinguished citizens, the man whom Hermippus here led off as condemned, who did not receive even this present legation from his fellow-citizens but went seeking it as far as Tmolus, who has had no honour in his city ever, and on whom the matter usually entrusted to the most slender persons was once in his life uniquely entrusted? Under the praetor Titus Aufidius he was set over public grain. When he had taken money for it from the praetor Publius Varinius, he hid it from his fellow-citizens and laid the cost on them besides. After this had become known and exposed at Temnus by the letter of Publius Varinius, and after Gnaeus Lentulus, the former censor, the patron of the Temnitans, had written letters on the same matter --- no one ever saw Heraclides at Temnus afterwards.

\ciceroSection{46} And so that you may grasp his shamelessness, learn now, please, the very ground of grievance which set this most fickle man against Flaccus. He bought a Cymaean estate at Rome from the ward Meculonius. When he was making himself rich in words, while he had nothing but that very shamelessness which you see, he took money on loan from Sextus Stloga, this judge of ours, a man of the first rank, who recognizes the matter and does not fail to know the man; who nevertheless lent on the credit of the most select Publius Fulvius Neratus. When he was paying Stloga, he took money from Gaius and Marcus Fufius, Roman knights of the first rank. Here, by Hercules, he ``put out the crow's eye,'' as the saying goes. For he struck this Hermippus, an educated man, his fellow-citizen, to whom he ought to have been very well known. For on his credit he took money from the Fufii. Untroubled, Hermippus sets out for Temnus, while this fellow says he will pay the money he has taken on Hermippus's surety from the Fufii, out of his own students.

\ciceroSection{47} For our rhetorician here had certain wealthy pupils, whom he made by half more foolish than he had received them; yet none did he so far befool as to lend him a penny. And so when he had quietly slipped out of Rome and defrauded many men by petty borrowings, he came to Asia, and to Hermippus's question about the Fufian debt answered that he had paid all the money to the Fufii. Meanwhile, after no long interval, a freedman of the Fufii arrived with a letter for Hermippus; the money is demanded of Hermippus. Hermippus demands it of Heraclides; he himself, however, satisfies the absent Fufii and discharges his own credit; this man, hot and prevaricating, he pursues at law. The case is heard by recuperatores.

\ciceroSection{48} Do not suppose, judges, that the shamelessness of cheats and welshers is not one and the same in every place. He did all the same things our debtors are wont to do; he denied that he had ever taken any loan at all at Rome; he affirmed that he had never even heard the Fufii's name; while he tore Hermippus himself, that most modest and most excellent man, my old friend and host, the most splendid and adorned of his city, with every reproach and abuse. But while the man, voluble in a kind of headlong speed of speaking, was tossing himself about in that speech of his, suddenly --- as the Fufii's depositions and the names were read out --- he, the most audacious of men, took fright; the most loquacious fell silent. And so the recuperatores, in this matter least of all in doubt, gave judgment against him at the very first hearing. When he was not satisfying the judgment, he was assigned to Hermippus and led off by him.

\ciceroSection{49} You have here both the man's honour and the authority of his testimony, and the whole cause of the feud. He, dismissed by Hermippus when he had sold him a few slaves, betook himself to Rome, then returned to Asia, when my brother had now succeeded Flaccus. He went to him and laid out the case: that the recuperatores, forced by Flaccus's violence and fear, had given a false judgment against their will. My brother, with his usual fairness and prudence, decreed that, if he denied the judgment, he should go for the double; if he said they had been forced by fear, he should keep the same recuperatores. He refused; and as if nothing had been done, no judgment given, he began to demand from Hermippus on the spot the very slaves which he himself had sold him. The legate Marcus Gratidius, to whom appeal was made, refused to grant an action; he showed it was his ruling that what had been adjudicated should stand.

\ciceroSection{50} Once more this fellow, who had nowhere any place to take a stand, betook himself to Rome; Hermippus follows --- the man who never gave way to that shamelessness. Heraclides claims from Gaius Plotius, a senator and a man of the first rank who had been legate in Asia, certain slaves which he says were sold by force, after he had been adjudicated. Quintus Naso, that most distinguished man, who had been praetor, is taken as judge. When he showed that he was about to give sentence for Plotius, he abandoned that judge, and, since the matter could not legally come into court, he gave the whole case up. Do I seem to you, judges, sufficiently to be approaching the witnesses one by one, instead of, as I had at first proposed, only contending with the whole class?

\ciceroSection{51} I come to Lysanias of the same city, your private witness, Decianus; whom you, when you had known him as an ephebe at Temnus, because, naked, he had then delighted you, you have always wanted to keep naked. You took Apollonides off from Temnus; you laid out money to the very young man at heavy interest, with a pledge taken nonetheless. This pledge you say has been forfeited to you; today you hold and possess it. This man you have compelled to come and bear witness on the hope of recovering his paternal estate. Since he has not yet given his testimony, I wait to see what he is going to say. I know the kind of person, I know the habit, I know the lust. So, although I am sure of what he is ready to say, I shall yet argue nothing against him before he has spoken. For he will turn the whole thing about and invent something else. So let him keep what he has prepared, and I shall keep myself entire for what he brings forward.

\ciceroSection{52} I come now to that city on which I have bestowed many and great pains and offices, and which my brother before all others cherishes and loves. If this city, through good men and weighty men, had laid its complaints before you, I should be a little more moved. But now what am I to think? --- that the Trallians have committed their public case to Maeandrius, a needy, sordid man, without honour, without reputation, without property? Where were those Pythodori, Aetidemi, Lepisones, and the rest of the men known among us, noble among their own? Where was that magnificent and glorying ostentation of the city? Would they not have been ashamed, had they been pursuing this case in earnest, that not merely the legate, but any Trallian at all, should be called Maeandrius? To this legate, this public witness, would they have handed over Lucius Flaccus, the patron whose family from his father and his ancestors they had already inherited, to be sacrificed by the testimony of the city? It is not so, judges; surely it is not so.

\ciceroSection{53} I myself saw lately in a certain trial Philodorus the Trallian giving testimony, I saw Parrhasius, I saw Archidemus --- when this same Maeandrius was beside me as a kind of attendant, supplying me with what to say against his fellow-citizens and his city if I wished. For there is nothing more fickle than that man, nothing more needy, nothing more polluted. Therefore, if the Trallians make this man the spokesman of their complaints, the keeper of their records, the witness of their wrong, the agent of their grievances, let them put down their high spirits, repress their souls, set aside their arrogance, and confess that the image of their city has been expressed in the person of Maeandrius. But if those very men have always thought he should be trampled and trodden under foot at home, let them stop thinking there is any authority in a testimony for which no authority can be found. But I shall set out the matter as it stands, that you may know why this city has neither attacked Flaccus rigorously nor defended him generously.

\ciceroSection{54} It was angry with him over the Castricius case, on which Hortensius answered fully; against its will it had paid Castricius money long owed. From this came all the hatred, from this every offence. When Laelius came to angry men and rubbed open the Castricius wound by his speech, the leading men were silent, were not present at that meeting, were not willing to be authors of that decree or that testimony. So bare of the optimates was that meeting that the chief among chiefs was Maeandrius, by whose tongue, as by a fan of sedition, that meeting of beggars was at that time fanned about.

\ciceroSection{55} Hear, then, the just complaint and grievance of a city which is modest --- as I have always thought it --- and weighty, as they wish to be thought. Money which had been with them in the name of Flaccus's father from the cities, this they complain has been taken from them. In another place I shall ask what was permitted to Flaccus; for now I only ask of the Trallians: this money which they complain was taken from them --- do they say it is theirs, conferred upon them by the cities for their own use? I am eager to hear. ``We do not say so,'' he says. What then? ``Brought to us, entrusted to us in the name of Lucius Flaccus the father, for his festival days and his games.'' What then?

\ciceroSection{56} ``You,'' he says, ``were not allowed to take it.'' I shall see to that presently; but first I shall hold to this. A weighty, wealthy, and adorned city complains because it is not keeping what is another's; it says it has been despoiled, because it does not have what was not its own. Can anything more shameless be said or invented? A town was chosen, in which the whole sum of money from all Asia was to be set down for Lucius Flaccus's honours. This whole sum was diverted from his honours into traffic and usury; and it was recovered many years afterwards.

\ciceroSection{57} What injury was done to the city? ``But the city is grieved.'' I believe it: that profit which was already devoured by hope was torn out beyond hope. ``But it complains.'' It does so shamelessly: for we cannot rightly complain of every grief we feel. ``But it accuses with the gravest words.'' Not the city, but unskilled men stirred up by Maeandrius. And here time and again make sure you remember what the recklessness of the multitude is, what the levity proper to Greeks, what force a speech has in a seditious meeting. Here, in this gravest and most temperate community, when the Forum is full of trials, full of magistrates, full of the best men and citizens, when the Senate-house, the avenger of recklessness and moderator of duty, watches and besets the Rostra --- yet what waves you see stirred up in our public meetings! What do you think happens at Tralles? Or what was done at Pergamum? Unless, perhaps, these cities want it to be thought that they could be more easily moved and driven by a single letter of Mithridates to violate the friendship of the Roman people, their own loyalty, all the rights of duty and humanity, than to wound by their testimony the son of a man whose father, in arms, they had decreed should be driven from their walls.

\ciceroSection{58} Therefore, do not throw at me these names of noble cities; for those whom this family despised as enemies, the same it will never fear as witnesses. But you must confess: if your cities are governed by the counsel of their leaders, then this war was undertaken by these cities against the Roman people not by the recklessness of the multitude but by the policy of the optimates. But if that movement was stirred up by the recklessness of the unskilled, suffer me, then, to separate the offences of the mob from the public cause.

\ciceroSection{59} ``But it was not allowed to him,'' it is said, ``to take that money.'' Do you wish that it should have been allowed to Flaccus the father, or not? If it was allowed --- as certainly it was allowed --- to use money brought together for his honours, of which he himself was taking nothing, the son rightly took the father's money. If it was not allowed, yet on his death not the son alone but any heir whatsoever could most rightly have taken it. And in fact at the time, when the Trallians themselves had at heavy interest had this money in their hands for many years, they yet got from Flaccus everything they wanted; nor were they so shameless as to dare what Laelius said --- that Mithridates had carried this money off from them. For who was there who did not know that Mithridates was more zealous in adorning than in despoiling the Trallians?

\ciceroSection{60} If these things were said by me as they ought to be said, I should be acting more weightily, judges, than I have so far acted, on how much faith you ought to give to Asiatic witnesses; I should call your minds back to the memory of the Mithridatic war, to that wretched and cruel slaughter of all Roman citizens through so many cities at one moment of time, to our praetors handed over, our legates thrown into chains, the very memory of the Roman name with every footprint of empire wiped not only out of the Greeks' dwellings but even out of their writings. Mithridates they called master, him father, him saviour of Asia, him Euhius, Nysius, Bacchus, Liber.

\ciceroSection{61} It was one and the same time when all Asia was closing its gates to Lucius Flaccus the consul, and was not only receiving but inviting that Cappadocian into its cities. Suffer this: if we cannot forget, let us at least be silent. Let me complain rather of the levity of the Greeks than of their cruelty. Are these to have authority among us, with whom they would not have us be at all? For all they could, they slew our togaed citizens; they wiped out, so far as in them lay, the name of Roman citizens. In this city then they parade themselves which they hate, before those they look on unwillingly, in that commonwealth for whose overthrow not the will but the strength was lacking to them. Let them look upon this flower of legates and praisers of Flaccus from real and unspoiled Greece; then let them weigh themselves, then compare themselves with these, then, if they dare, prefer their own dignity to theirs.

\ciceroSection{62} The Athenians are present, from whom humanity, learning, religion, fruits of the earth, rights, laws are believed to have sprung up and to have been distributed into all lands; over the possession of whose city, on account of its beauty, even among the gods, tradition records, there was a contest; whose age is such that she is held to have given birth out of herself to her own citizens, and the very same earth is called their parent, their nurse, their fatherland; and whose authority is so great that the now broken and almost shattered name of Greece is held up by the praise of this single city.

\ciceroSection{63} The Lacedaemonians are present, the proven and renowned virtue of whose city is reckoned strengthened not only by nature but by discipline; who alone in the whole world for more than seven hundred years now have lived under one and the same way of life, with their laws never altered. Many legates from all Achaia are present, from Boeotia, from Thessalia, where Flaccus, lately, was set in command as legate to the imperator Metellus. Nor do I pass you over, Massilia, who knew Lucius Flaccus as a military tribune and quaestor: a city whose discipline and weight I shall rightly say is to be set above not only Greece but, I almost think, all nations; which, though removed so far from all the regions, the disciplines, and the language of the Greeks, though girt round at the ends of the earth by Gallic peoples and washed by the waves of barbarism, is so governed by the policy of the optimates that all men can more easily praise its institutions than rival them.

\ciceroSection{64} It is these whom Flaccus uses as praisers, these as witnesses to his innocence, that we may resist the desires of the Greeks by the help of Greeks. And yet who that has ever cared to know these things even moderately is ignorant that there are truly three races of Greeks? Of which one is the Athenians, that nation which was held to be of the Ionians; the second was called Aeolian; the third Dorian. And all this Greece, which has flourished in renown, in glory, in learning, in many arts, in empire and the fame of war, has held, as you know, and always held, a certain small space of Europe; the maritime coast of Asia she encircled with cities after she had won it by war --- not to bind it down with colonies as a thing conquered, but to hold it as a beleaguered place.

\ciceroSection{65} Therefore I ask of you, witnesses from Asia, that, when you wish truly to recall how much authority you bring into a trial, you describe Asia for yourselves, and remember not what foreigners are wont to say of you, but what you yourselves judge of your own race. For your Asia, as I think, is made up of Phrygia, Mysia, Caria, Lydia. Is this proverb ours, then, or yours, that ``a Phrygian is generally bettered by being beaten''? What? Has not this been put about by your own voice in all Caria: ``If you wish to try anything dangerous, it is best done in a Carian''? What again is more often used and celebrated in Greek conversation than, when a man is held in contempt, that he is called ``the lowest of the Mysians''? For what shall I say of Lydia? What Greek ever wrote a comedy in which the slave of the leading parts was not a Lydian? What injury, then, is done to you, if we determine that we shall stand on your own judgment about you?

\ciceroSection{66} For my own part it seems to me I have now said enough and more than enough about the Asian class of witnesses; yet it is for you, judges, to take in by your minds and reflection all the things which can be said of the levity, inconstancy, and greed of these men, even if I say less of them. There follows the matter of that Jewish gold. This, no doubt, is the reason this case is being heard not far from the Aurelian steps. For this charge has this place been sought, Laelius, by you and that crowd; you know how great a band it is, what concord, what weight in public meetings. So I shall speak in a low voice only that the judges may hear; for there are not lacking men to incite these against me and against every best citizen. I shall not help them to do it the more easily.

\ciceroSection{67} When gold under the name of the Jews was customarily exported every year out of Italy and out of all our provinces to Jerusalem, Flaccus by edict forbade its being exported from Asia. Who is there, judges, who could not truly praise this? That gold ought not to be exported, both often before and most weightily in my consulship the Senate decided. To resist this barbarian superstition was a mark of strictness; to despise the multitude of Jews when sometimes blazing in our public meetings, for the sake of the commonwealth, was a mark of the highest gravity. ``But when Gnaeus Pompeius captured Jerusalem in his victory, he touched nothing in that temple.''

\ciceroSection{68} In this, before all things, as in much else, he was wise: in so suspicious and slanderous a community he left no room for the talk of detractors. For I do not believe it was the religion of the Jews and of enemies that hindered that most outstanding general, but a sense of decency. Where, then, is the charge, since you nowhere reproach a theft, you approve the edict, you confess the judgment, you do not deny that it was sought and produced openly, that the affair was carried through by the first men of the state, the very thing declares? At Apamea openly, before the praetor's feet in the forum, a hundred pounds of gold a little less was weighed by Sextus Caesius, a Roman knight, a most chaste and most upright man; at Laodicea twenty pounds a little more by Lucius Peducaeus here, our judge; at Adramyttium a hundred by the legate Gnaeus Domitius; at Pergamum no great amount.

\ciceroSection{69} The reckoning of the gold balances; the gold is in the treasury. Theft is not charged; envy is sought; the speech is turned away from the judges, the voice is poured out upon the ring of bystanders and the crowd. Each city, Laelius, has its own religion; we have ours. While Jerusalem stood and the Jews were at peace, even so the religion of these sacred rites was repugnant to the splendour of this empire, the gravity of our name, the institutions of our ancestors; but more so now, since that nation has shown by arms what it thought of our empire, has shown by being defeated, by being farmed out, by being made a slave-people, how dear it was to the immortal gods.

\ciceroSection{70} And so, since you see what you wished to be a charge has wholly been turned to praise, let us come at last to the complaints of Roman citizens; among which let Decianus's be first. What injury, then, was done to you, Decianus? You do business in a free city. First let me be inquisitive. How long are you to do business, especially since you were born in such a place? You have spent thirty years now in the forum, but in that of Pergamum. After a long interval, when it pleases you to travel, you come to Rome, bringing a new face, an old name, Tyrian purple --- in which I envy you, that you have been so long elegant in a single suit of clothes.

\ciceroSection{71} But suppose you wish to do business; why not at Pergamum, Smyrna, Tralles, where there are many Roman citizens and the law is administered by our magistrate? Quiet pleases you, lawsuits, throngs, a praetor are hateful, you delight in the freedom of the Greeks. Why is it that you alone, then, have the men of Apollonis, the most loving of the Roman people, the most faithful of allies, in a worse case than ever Mithridates or even your own father had them? Why is it that they are not allowed to enjoy their freedom through you, why finally not allowed to be free? They are men out of all Asia the most frugal, most upright, most removed from the Greek luxury and levity, fathers of households content with their own, ploughmen, country folk; they have lands by nature very good and made better by diligence and cultivation. In these lands you wished to have estates. I should have preferred altogether, and it would have suited you better, if at length thick lands delighted you, that you had bought them somewhere here in Crustumerium or in Capena.

\ciceroSection{72} But, very well; it is Cato's saying that ``money is balanced by your feet.'' From the Tiber to the Caicus is a long way altogether --- in which place even Agamemnon would have wandered with his army, had he not found Telephus as a guide. But I concede that too: the town pleased him, the region delighted. You should have bought. Amyntas is in birth, in honour, in repute, in fortune the leading man of that city. His mother-in-law, a woman of weak counsel and rich enough, Decianus enticed to himself; and, while she did not know what was being done, set his household in possession of her estates; and the wife of Amyntas, then pregnant, he carried off from him; she bore a daughter at Decianus's house, and to this day at Decianus's house there are both Amyntas's wife and daughter.

\ciceroSection{73} Am I making any of this up, Decianus? All the gentlemen know it, the good men know it, in short the men of any acquaintance know it, the merchants of even the middle sort know it. Stand up, Amyntas, and demand back from Decianus --- not money, not estates, let him keep his mother-in-law to himself; let him restore your wife, give back to a wretched father his daughter. The limbs which he crippled with stones, with cudgels, with iron, the hands he bruised, the fingers he broke, the sinews he cut, he cannot restore; but the daughter, the daughter, I say, give back, Decianus, to her wretched father.

\ciceroSection{74} You wonder that these things did not satisfy Flaccus? Whom, pray, did you ever satisfy? You made false purchases, advertisements of estates with manifest fraud. By Greek laws a guardian had to be assigned for these women; you wrote down Polemocrates, your hireling and minister of your designs. Polemocrates was brought into court for fraud and trickery by Dion in the very name of this guardianship. What a flocking from the neighbouring towns from every side, what passion of mind, what complaint! Polemocrates was condemned by every vote; the sales were void, the advertisements were void. Do you make restitution? You go to the Pergamenes that they may take into their public records your splendid advertisements and purchases. They reject, they refuse them. And what men are these? The Pergamenes, your praisers. For so it seemed to me that you boasted of the Pergamenes' praise, as if you had attained the honour of your ancestors; and you thought yourself superior to Laelius in this --- that the city of Pergamum praised you. Is the city of Pergamum more honourable, then, than that of Smyrna?

\ciceroSection{75} They themselves do not even say so. I should wish I had so much leisure that I could read out the decree which the men of Smyrna passed about the dead Castricius: first, that he be carried into the town, which is not granted to others; next, that the ephebes carry him; finally, that a golden crown be set upon him in death. These things were not done for Publius Scipio, that most distinguished man, when he died at Pergamum. But Castricius --- with what words, immortal gods! ``the glory of the country, the ornament of the Roman people, the flower of youth'' they call him. So, Decianus, if you are eager for glory, I advise you to seek other ornaments; the Pergamenes have laughed at you.

\ciceroSection{76} What? Did you not understand that you were being made fun of, when they read out these words to you: ``most illustrious man, of outstanding wisdom, of singular genius''? Believe me, they were playing. When indeed they were placing a golden crown upon you in their letters, they were entrusting no more gold to you in fact than they would have to a jackdaw; even then could you not see through the men's wit and humour? Those very Pergamenes therefore rejected the advertisements which you brought. Publius Orbius, a wise and innocent man, decreed everything against you. With Publius Globulus, my close friend, you were more in favour. I wish that neither he nor I were sorry for it!

\ciceroSection{77} You say that Flaccus decreed unjustly in your case; you add as causes of your enmity, that your father, as tribune of the plebs, served notice on Lucius Flaccus the father, when he was curule aedile. But this should not have been very offensive even to Flaccus the father himself, especially since the man on whom notice was served was afterwards made praetor and consul, while the man who served the notice could not stand his ground in the city as a private man. But if you thought your enmity just, why, when Flaccus was a military tribune, were you a soldier in his legion, when by the laws of military service you might have escaped a tribune's unfairness? Why, again, did the praetor call you, his father's enemy, into his council? --- and how scrupulously such things are usually observed you all know.

\ciceroSection{78} We are now accused by men who were on our council. ``Flaccus made a decree.'' What other than he ought? ``Against your children.'' Did the Senate not vote the same? ``Against you in your absence.'' He decreed it, when you were on the spot, when you would not come forward; this is not a decree against an absent defendant, but against a hidden one. \textit{[Decree of the Senate and ruling of Flaccus.]} What if he had not made a decree but had given an edict --- who could truly find fault? Are you about to find fault even with my brother's letter, full of humanity and fairness, which on the matter of that woman he sent to me, when he himself \ldots required.

\ciceroSection{79} Read it. \textit{[Letter of Quintus Cicero.]} What? Did the people of Apollonis seize the chance and not refer this to Flaccus, was nothing done before Orbius, was nothing referred to Globulus? Did not legates of the Apollonidans, in my consulship, lay before our Senate every petition about the wrongs of one Decianus alone? But you have entered these estates in the census. I pass over that they are another's, I pass over that they are held by force, I pass over that they have been condemned by the Apollonidans, I pass over that they have been rejected by the Pergamenes, I pass over even that they have been restored to their integrity by our magistrates, I pass over that you have on them no right either by fact or by possession;

\ciceroSection{80} I ask only this --- whether those estates can be entered in the census, whether they have civil right, whether they are or are not mancipi, whether they can be pledged at the treasury or before the censor. In what tribe, finally, did you enter those estates? You have committed yourself --- if some heavier moment had come --- so that out of the same estates a tribute would have been levied both at Apollonis and at Rome. But, very well; you were boastful, you wished to have a great quantity of land entered, even land that cannot be divided among the Roman plebs. You were also entered for 130,000 sesterces in cash. This, I think, was not paid out by you. But I let this pass. You entered as yours the slaves of Amyntas, and you did him no injury in this. For Amyntas possesses those slaves. And at first he was indeed afraid, when he had heard that his own slaves had been entered as yours; he referred the matter to the lawyers. The opinion of all was, that if Decianus, by entering another's property in the census, could make it his own, he would be on the way to being the greatest man \ldots

\ciceroSection{81} You have here the cause of the enmity, in the heat of which Decianus brought this rich prosecution to Laelius. For thus Laelius complained, when he spoke of Decianus's perfidy: ``He who was my prompter, who brought the case to me, whom I followed --- he has been corrupted by Flaccus, he has deserted and betrayed me.'' Have you, then, called this most modest man, born of the most noble family, deserving best of the commonwealth, into the peril of all his fortunes --- a man for whom you yourself were prompter, on whose council you sat, with whom you maintained every step of your dignity? If it is permitted, I will defend Decianus, who comes under your suspicion through no fault of his own.

\ciceroSection{82} He is not corrupted, believe me. For what was there to be bought from him? That he should drag out the trial? When the law gave six hours in all, how much could he have taken from these hours, if he had been willing to humour you? It is, no doubt, what he himself suspects: you envied your subscriber's talent; because he readily set off any topic he had taken up, and questioned witnesses keenly, or perhaps could have brought it about that you should fall out of the people's talk, you therefore stuck Decianus away into the rear ranks. But, however unlikely it is that Decianus has been corrupted by Flaccus,

\ciceroSection{83} just so know it is in other matters: as in what Lucceius says, that Lucius Flaccus wished to give him two million sesterces to lead him from his loyalty. And do you accuse a man of avarice whom you say wished to throw away two million sesterces? For what was he buying when he was buying you? That you should come over to him? What part of the case were we to give you? Or that you should report Laelius's plans? --- what witnesses were going to come forward from him? What? Did we not see them? See them living together? Who does not know it? See that the records were in Laelius's power? Can it be doubted? Or was it that you should not be vehement, not copious, in prosecution? Now you do raise suspicion; for you so spoke that something seems to have been obtained from you.

\ciceroSection{84} ``But on Andro Sextilius a heavy and unbearable injury was done, in that, when his wife Valeria had died intestate, Flaccus dealt with the case as if the inheritance pertained to himself.'' In this what is to be reproved I am eager to know. That he laid claim to a falsehood? How do you make this out? ``She was free-born,'' he says. O learned in the law! What? Do not inheritances by the law come from free-born women? ``She had passed,'' he says, ``into manus.'' Now I hear; but I ask, by usus or by coemptio? By usus she could not; for from a legitimate guardianship nothing can be alienated except by the authority of all the guardians. By coemptio, then? By all the authorities, of whom you certainly will not say Flaccus was one.

\ciceroSection{85} It remains what he never ceased to shout: that, while he was praetor, he ought not to have transacted his own business or made mention of an inheritance. The greatest inheritances, I hear, have come to you, Lucius Lucullus, who are about to give your verdict on Lucius Flaccus, by your remarkable generosity and the greatest benefits to your people, when you held the province of Asia with consular imperium. Had any man said they were his own, would you have given them up? You, Titus Vettius, if any inheritance came to you in Africa, would you lose it by usus, or would you keep what was yours, with no greed and your dignity safe? But of that inheritance, even when Globulus was praetor, possession had been claimed in the name of Flaccus. It was not, then, oppression, not violence, not occasion, not season, not command, not the axes that drove Flaccus's mind to do an injury.

\ciceroSection{86} And there too Marcus Lurco, an excellent man, my close friend, turned the sting of his testimony; he said it was not right for a praetor to demand money from a private man in his province. Why, pray, Marcus Lurco, is it not right? To extort, to receive against the laws is not right; that to demand it is not right you will never show, unless you prove it is not allowed. Or, to take a free legation for the purpose of collecting a debt --- as you yourself lately and many good men have often done, which I do not censure but see that the allies complain of --- is right; and yet, if a praetor has not left his inheritance behind in his province, you think he should not only be censured but condemned? ``But,'' he says, ``Valeria had assigned the whole of her property as dowry.'' None of these matters can be made out, unless you show that she was not under Flaccus's guardianship. If she was, whatever dowry was assigned without his authority is null.

\ciceroSection{87} But, although Lurco bore himself in giving testimony with moderation up to his dignity, you saw that he was nevertheless angry with Flaccus. Nor did he hide the cause of his anger or think it should be passed over in silence; he complained that his freedman had been condemned under Flaccus as praetor. O wretched conditions of administering provinces, where diligence is full of feuds, neglect of vituperation; where strictness is dangerous, generosity ungrateful, conversation treacherous, flattery deadly, where the brow of all is friendly but the mind of many angry, where the angers are hidden and the flatteries open, where the people wait for praetors as they come, are subservient to them as they are present, desert them as they leave! But let me leave off complaining, lest I seem to praise our own decision in passing over provinces.

\ciceroSection{88} He sent letters about the bailiff of Publius Septimius, an honourable man, a bailiff who had committed murder; you could see Septimius blazing with anger. He gave a verdict by edict against the freedman of Lurco; Lurco is an enemy. What then? Was Asia to be handed over to the freedmen of well-favoured and splendid men? Or does Flaccus carry on some quarrel or other with your freedmen? Or, in the affairs of yourselves and your own, is strictness hateful to you, while you praise the same when you are sitting in judgment about us? But that Andro of yours, despoiled of his goods, as you say, has not come to give testimony. What if he should come?

\ciceroSection{89} The arbiter of the settlement was Gaius Caecilius --- what a splendid man, what a man of good faith, what a man of scruple! The witness to the seal was Gaius Sextilius, son of Lurco's sister, a man both modest and steadfast and weighty. If there was force, if fraud, if fear, if circumvention --- who forced the agreement to be made, who forced these men to be present? What? If the whole of that money was repaid to this young Lucius Flaccus --- if it was demanded, if it was got in by this Antiochus, the freedman of this young man's father, most approved by the elder Flaccus --- do we seem not only to escape the charge of avarice but to attain singular praise of generosity? For the inheritance which had come equally to both by law, he conceded to his young kinsman; he himself touched nothing of Valeria's goods. What he had decided to do, drawn on by the young man's modesty and not by an overlarge family fortune, that he not only did but did liberally and abundantly. From which it must be understood that he has not taken money against the laws, who has been so liberal in conceding an inheritance.

\ciceroSection{90} ``But the Falcidian charge is huge: he says he gave Flaccus fifty talents.'' Let us hear the man. He is not present. How then does he speak? His mother brings forward one letter, and his sister another; they say it is written to them by him that this great sum of money was given to Flaccus. So a man whom no one would believe were he holding the altar and swearing, will prove what he wishes through a letter, unsworn? And what a man! How disloyal to his own fellow-citizens! He who could have run through a tolerably elegant patrimony here with us preferred to scatter it on Greek dinner-parties.

\ciceroSection{91} What was the use of leaving this city, of foregoing such splendid liberty, of running the danger of a sea-voyage --- as if it were not allowed to eat one's goods at Rome? Now at last the playful son to a doting little mother, to an old woman least suspicious by nature, clears himself by letter, that he may seem not to have spent on the voyage out the money he had taken with him, but to have given it to Flaccus. But those Trallian rents had been sold under the praetor Globulus; Falcidius had bought them for nine hundred thousand sesterces. If he gives this great sum to Flaccus, he gives it of course that the purchase may stand. Then he is buying something certainly worth much more; he gives out of his profit, he takes nothing off the principal.

\ciceroSection{92} Why does he order Albanus to come, why does he flatter his mother besides, why does he go fishing through letters in the weakness of his sister and his mother --- in short, why do we not hear the man himself? He is detained, no doubt, in his province. The mother says he is not. ``He would have come,'' she says, ``had he been summoned.'' You certainly would have compelled him, had you placed any reliance on that witness; but you did not wish to draw him from his business. A great contest had been laid out for him, a great struggle with the Greeks --- who, however, lie defeated, as I think. For this single man has surpassed all Asia in the size of his cups and in drinking. But who, Laelius, told you about those letters? The women say they do not know who he is. He himself, then, has told you that he wrote to his mother and his sister?

\ciceroSection{93} Or did he even write at your asking? But you put no question to that most steadfast and most modest of men, Marcus Aebutius, the kinsman of Falcidius, you put no question to his son-in-law, of equal good faith, Gaius Manilius? --- who certainly could not have heard nothing of so great a sum of money, had it been given. Did you, Decianus, suppose that with these letters read out, with these little women produced, with that absent author praised, you would prove so great a charge --- especially since you yourself, by not bringing forward Falcidius, have made a judgment that a forged letter would have more weight than his own forged voice and feigned grief, were he present?

\ciceroSection{94} But why do I argue so long about Falcidius's letters, or about Andro Sextilius, or about Decianus's census, while I am silent about the safety of us all, about the fortunes of the state, about the highest concern of the commonwealth? You are sustaining the whole of it on your shoulders, judges --- yours, I say --- in this trial. You see in what shaking of the times, in what overturning and confusion of affairs we are placed. Among many other things, certain men are engaged above all in this --- that your minds, your judgments, your votes shall be found most hostile and unfriendly to every best citizen. You have given many heavy verdicts for the dignity of the commonwealth on the crime of the conspirators. They do not think the commonwealth sufficiently overturned, unless they have thrust down into the same punishment as the impious, the citizens who have deserved best of her.

\ciceroSection{95} Gaius Antonius has been brought down. So be it: he had a kind of infamy of his own; and yet, if I say what is mine to say, even he himself would not have been condemned with you for judges --- the man at whose condemnation Lucius Catiline's tomb was decked with flowers, and celebrated by the assembling and feasting of the most audacious of men, our own household enemies. Funeral rites have been performed for Catiline; now from Flaccus the penalties of Lentulus are demanded through you. To Publius Lentulus, who tried to bury you, slaughtered in the embrace of your children and wives, in the burning of your fatherland --- what victim more pleasing can you slaughter than if you have glutted with the blood of Lucius Flaccus that man's hatred, abominable against you all?

\ciceroSection{96} Let us, then, sacrifice to Lentulus, let us make offerings to Cethegus, let us call back the banished; let us in turn, if it so pleases, suffer the penalties of an excessive piety and a supreme love for our country. We are now denounced by informers, charges are forged against us, dangers are prepared. If they brought these things through others, if finally they had stirred up under the people's name a multitude of unskilled citizens, we could bear it with a calmer mind; but this cannot be borne --- that they think it possible, through senators and Roman knights who carried out all these things for the safety of all by common counsel, with one mind and one valour, that the very authors, leaders, chiefs of these matters should be despoiled of all their fortunes and driven from the state. Indeed, they perceive that the Roman people has the same mind and the same will; in every way it can, the Roman people signifies what it feels; there is no variety of opinion among men, no variety of will, no variety of speech.

\ciceroSection{97} Therefore, if anyone calls me thither, I come; the Roman people as judge I not only do not refuse but demand. Let force be absent, let iron and stones be removed, let hired gangs withdraw, let slaves be silent; there will be no one so unjust who has heard me, provided he be free and a citizen, who will not think rather of rewards for me than of penalty. O immortal gods! What is more wretched than this? We who wrenched the iron and the flame from Publius Lentulus's hands have confidence in the judgment of the unskilled multitude, and yet we fear the verdicts of the most select and most distinguished citizens!

\ciceroSection{98} Manius Aquilius, convicted by our fathers of many charges of avarice and of weighty testimonies, because he had bravely waged war with the runaway slaves, was acquitted by their verdict. I myself as consul lately defended Gaius Piso; who, because he had been a brave and steadfast consul, was preserved unharmed for the commonwealth. As consul I likewise defended Lucius Murena, the consul-elect. None of those judges, when most distinguished men were prosecuting, thought it right to listen to charges of bribery, since with Catiline now waging war they all knew, on my advising, that there had to be two consuls on the Kalends of January. An innocent man, a good man, adorned in every way, has been twice acquitted in this year with me defending him: Aulus Thermus. How great a joy of the Roman people followed for the sake of the commonwealth, how great a thanksgiving! Always grave and wise judges have considered, in deciding cases, what the utility of the state, what the common safety, what the times of the commonwealth required.

\ciceroSection{99} When the tablet shall be given to you, judges, it will be given not on Flaccus alone, it will be given on the leaders and authors of preserving the city, it will be given on all good citizens, it will be given on yourselves, on your children, on your life, on your country, on the common safety. You are not judging in this case about foreign nations, not about allies; you are judging about yourselves and about your own commonwealth.

\ciceroSection{100} If, however, the consideration of provinces moves you more than your own, I do not refuse, indeed; rather I demand that you be moved by the authority of the provinces. For we shall set against the province of Asia, first, a great part of the same province, which has sent legates and praisers for this man's perils; next the province of Gaul, the province of Cilicia, the province of Spain, the province of Crete; against the Greek Lydians, Phrygians, and Mysians shall stand the men of Marseilles, of Rhodes, the Lacedaemonians, the Athenians, all Achaia, Thessaly, Boeotia; against the witnesses Septimius and Caelius, Publius Servilius and Quintus Metellus, witnesses to this man's modesty and integrity, will fight back; the city's jurisdiction will answer the Asian jurisdiction; against the imputation of a single year, all Lucius Flaccus's age and his whole life will defend him.

\ciceroSection{101} And if it ought to profit Lucius Flaccus, judges, that he has shown himself, as military tribune, as quaestor, as legate, worthy of his ancestors with the most distinguished imperatores, the most splendid armies, the most weighty provinces, let it profit him that under your eyes, in the common dangers of all of us, he joined his own perils to mine; let the praisers from the most honourable municipia and colonies profit him; let even the splendid and true praise of the Senate and the Roman people profit him.

\ciceroSection{102} O night, that almost brought eternal darkness upon this city, when the Gauls were being summoned to war, Catiline to the city, the conspirators to iron and to fire, when I, Flaccus, calling on heaven and the night to witness, weeping I besought you weeping, when to your most outstanding and most tested loyalty I commended the safety of the city and of the citizens! Then, Flaccus, you, the praetor, took the messengers of the common ruin, you caught the pestilence of the commonwealth shut up in letters, you brought the proofs of the perils, the helps of safety, to me and to the Senate. What thanks were given you then by me, what by the Senate, what by all good men! Who would have thought that to you, that to that bravest of men Gaius Pomptinus, any good man would ever deny not safety only but any honour? O those Nones of December which fell in my consulship! What day I can truly call the birthday of this city, or at least its day of saving!

\ciceroSection{103} O night that this day followed, fortunate for this city --- wretched I, I fear funereal for us! What was Lucius Flaccus's spirit then --- I shall say nothing of mine --- what love of country, what valour, what gravity stood out! But why do I recall things which then, when they were being done, were exalted to heaven by the unanimous consent of all, the single voice of the Roman people, the single testimony of the whole world, but which now I fear may not only not profit but even do some harm? For I see that the memory of bad men is sometimes much sharper than that of good ones. I, if anything graver shall happen --- I, I say, Flaccus, shall have betrayed you. O that right hand of mine, my pledge, my promises, when I was promising you that, if we preserved the commonwealth, you would, all your life as long as you lived, be not only fortified but adorned by the protection of all good men. I thought, I hoped, that even if our honour were of less worth to you, our safety at least would be dear.

\ciceroSection{104} As for Lucius Flaccus, judges, if --- which omen may the immortal gods avert --- a heavy injury crush him, yet he will never repent that he looked out for your safety, took counsel for you, your children, your wives, your fortunes; he will always feel that he owed such a spirit to the dignity of his lineage, to his own piety, and to his country. As for you, do not repent that you did not spare such a citizen --- by the immortal gods, judges, take care! For how many are there who follow this party in the commonwealth, who wish to please you and the like of you, who hold the authority of every best and most distinguished man and order to be a great matter, when they see that other path easier to honours and to all that they have desired? But let the rest be theirs; let them have to themselves the power, the honours, the highest stores of every other advantage; let it be permitted to those who wished these things to be safe themselves to be safe.

\ciceroSection{105} Do not think, judges, that those whose case is whole, who have not yet come to honours, are not waiting on the outcome of this trial. If for Lucius Flaccus so great a love towards every good man, so great a zeal for the commonwealth, has been a calamity, who hereafter, do you think, will be so out of his mind as not to think that the way of life he had reckoned headlong and slippery is to be preferred to this level and steady one? But if you, judges, are weary of such citizens, declare it; they will change their minds who can; they will determine what to do whose case is whole; we who are already gone forward shall bear this ending of our temerity. But if you wish as many as possible to be of this mind, you will declare in this trial what you feel.

\ciceroSection{106} To this --- to this poor boy of yours and your children's suppliant, judges, you will give in this trial the precepts for living. If you preserve his father, you will prescribe what kind of citizen he himself ought to be; if you take him away, you will show that no fruit is held out by you for sound, steady, and grave reasoning. He, who --- since he is of an age to perceive feeling already from his father's grief, but cannot yet bear his father help --- begs you not to swell his own grief with a father's tears, his father's sorrow with his weeping; who looks even at me, calls me with his face, weeping in a manner implores my pledge and asks back that dignity which I once promised his father for his country's saving. Have pity on the family, judges, have pity on the bravest of fathers, have pity on the son; preserve a name most distinguished and bravest, for the sake of the lineage, or of antiquity, or of the man, to the commonwealth.

\ciceroSection{fr1} How well-disposed do you think this man to the Roman people, how loyal?

\ciceroSection{fr2} Inborn levity, and learned vanity.
