% Pro P. Sulla (pro-sulla) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} What I should most have wished, judges, was that Publius Sulla had been able to keep the splendour of his standing as it once was, and after the disaster he suffered to gather some fruit from his moderation. But since hostile chance has fallen out so that at the height of his honour he was overthrown by the common odium that attaches to canvassing and by the singular hatred of Autronius, and since amid these wretched and battered remains of his former fortune he still has men whose temper not even his ruin could glut --- though I take great heaviness of mind from his troubles, yet among my other griefs I bear with patience that an occasion has been given me on which good men may recognize a mildness and mercy in me once familiar to them all, now as it were laid aside, and on which the wicked and ruined citizens, mastered and beaten, will admit that with the commonwealth tottering I was vehement and brave, and with the commonwealth saved gentle and merciful.

\ciceroSection{2} And since Lucius Torquatus, my close friend and connection, judges, has thought that, if he should violate that intimacy and friendship of ours in his prosecution, he could detract something from the authority of my defence, I shall couple with the warding off of this man's danger the defence of my own duty. I should certainly not be using this kind of speech now, judges, were the matter mine alone --- for in many places the chance has been given me, and will often be given again, of speaking to my own praise. But just as he saw that as much as he stripped from my authority, by so much would he diminish this man's protection, so I, on my side, judge that if I prove to you the reasoning of my act and the consistency of this duty and defence, I shall also prove the case of Publius Sulla.

\ciceroSection{3} And first, Lucius Torquatus, this is what I ask of you: why do you set me apart from the rest of the most distinguished men, the leading citizens, in this duty and in the right of defence? What is the reason why what Quintus Hortensius has done, that man most distinguished and most adorned, is not censured by you, but mine is? For if a plot was formed by Publius Sulla to set this city on fire, to extinguish our empire, to wipe out our state, these things should bring me greater grief than they should bring Hortensius, greater hatred toward me; my judgment finally as to who is to be helped in such cases, who attacked, who defended, who deserted, ought to be the weightier. ``Just so,'' he says; ``for it was you who tracked down, you who exposed the conspiracy.''

\ciceroSection{4} When he says this, he does not consider that the man who exposed it took care that all should see what before had been hidden. So that conspiracy, if it has been exposed through me, is as exposed to Hortensius as it is to me. And when you see him --- a man endowed with such honour, authority, virtue, and judgment --- not hesitating to defend the innocence of Publius Sulla, I ask why the path to the case that lay open to Hortensius should have been shut off to me. I ask this too: if you think that I, who am defending him, should be censured, what then do you think of these greatest of men, these most distinguished citizens, by whose zeal and dignity you see this trial honoured, this case adorned, this man's innocence defended? For there is not just one mode of defence, the one set out in a speech: all who attend, who are anxious, who wish him saved, defend him each in his own measure of standing.

\ciceroSection{5} Or was I, when I saw on these benches the ornaments and lights of the commonwealth, to be unwilling to appear among them --- I, who climbed up to that very place, to this most exalted seat of dignity and honour, by my own great labours and dangers? And so that you may understand, Torquatus, whom you accuse, in case it offends you that I, who have defended no one in this kind of trial, do not fail Publius Sulla, recall who else are present in his support: you will see that my judgment in his case and in the others, and theirs, has been one and the same.

\ciceroSection{6} Which of us stood by Vargunteius? No one --- not even this Quintus Hortensius, who alone had previously defended him on a charge of canvassing. For he no longer thought himself bound by any duty to that man, since by committing so great a crime he had broken the bond of all duties. Which of us thought that Servius Sulla, that Publius, that Marcus Laeca, that Gaius Cornelius should be defended? Which of us stood by them? No one. Why? Because in other cases good men do not think that even guilty men, if they are connections, should be deserted; in this charge, however, the fault is not only one of fickleness but there is even a sort of contagion of crime, if you defend a man whom you suspect of being bound by parricide of the fatherland.

\ciceroSection{7} What of Autronius? Did not his comrades, his colleagues, his old friends, of whom he once had had abundance, did not all the leading men of the commonwealth fail him? Indeed many of them even injured him by their evidence. They had judged that the wickedness was such as not only ought not to be hidden by them, but rather to be opened up and brought into the light. So why is there anything to wonder at if you see me present in this case among the same men with whom you understand I was absent in the others? Unless, indeed, you wish me alone, beyond the rest, to be thought savage, harsh, inhuman, endowed with a singular barbarity and cruelty.

\ciceroSection{8} If you fix this character on me for the whole of my life, Torquatus, on account of my actions, you are very wrong indeed. By nature I am merciful, by my country's call severe; cruel, neither my country nor my nature would have me. The very vehement and harsh character which the time and the commonwealth then forced on me, my own will and very nature have already laid aside. For that called for severity for a brief time; this calls for mercy and mildness for the whole of life.

\ciceroSection{9} There is therefore no reason for you to drag me alone away out of so great a company of the most distinguished men. The duty is single, and the cause of all good men one and the same. There will be nothing for you to wonder at hereafter if in that quarter where you have noticed these you will see me also. For there is no cause of mine private to me in the commonwealth: a time for action was more peculiarly mine than the rest's, but the cause of grief and fear and danger was for all of us in common. Nor could I then have been the leader of our salvation, had others not been willing to be my companions. So what was peculiarly mine as consul, beyond the others, must now, as a private man, be common to me with the rest. Nor do I say this to share the odium, but to share the praise: I impart to no one a part of my burden, to all good men a share of the glory.

\ciceroSection{10} ``You gave evidence against Autronius,'' he says; ``you defend Sulla.'' This whole charge is of the kind, judges, that, if I am inconstant and frivolous, neither should credit have been given to my evidence nor authority to my defence; but if there is in me regard for the commonwealth, scruple in private duty, zeal for keeping the goodwill of good men, the prosecutor's last word should be that Sulla is being defended by me, that Autronius was injured by my testimony. For I see now that I bring not only zeal for defending cases, but some weight of opinion and authority too. This I shall use moderately, judges, and should not be using at all if he had not driven me to it.

\ciceroSection{11} Two conspiracies, Torquatus, are set up by you: one which is said to have been formed in the consulship of Lepidus and Volcatius, when your father was consul-designate; the other in mine. In each of these you say Sulla was involved. As to your father's, that bravest of men and best of consuls, you know I had no part in his counsels; you know that, although I was on the most familiar terms with you, yet I had no share in those proceedings or those discussions --- because, I take it, I was not yet deeply involved in public life, because I had not yet attained the goal of office I had set myself, because canvassing and forensic work drew me away from all such reflection.

\ciceroSection{12} Who, then, took part in your counsels? All these men whom you see attending here, and Quintus Hortensius first --- he, both because of his standing and dignity and uncommon spirit toward the commonwealth, and because of his close friendship and high love for your father, was moved as much by the dangers common to all as by the singular dangers of your father. So the charge of that conspiracy has been answered by the man who was present, who knew, who shared your counsel and your fear; and although his speech in beating off this charge was a most copious and most ornate one, yet there was no less authority in it than there was skill. As for that conspiracy, then, which is said to have been formed against you, reported to you, brought to light by you, I could not be a witness: not only have I learned nothing of it firsthand, but the rumour of that suspicion scarcely reached my ears.

\ciceroSection{13} The men who shared your counsel, who learned of these things together with you, who themselves were thought to be the targets at the time, who did not stand by Autronius, who gave grave evidence against him --- these defend this man, stand by him, declare in his danger that they were deterred not by a charge of conspiracy (so that they should be absent from the others) but by the wickedness of those men. The time and charge of my consulship, however, of the greatest conspiracy, will be defended by me. And this division of the defence between us has not been made fortuitously, judges, nor at random; but when we saw that we were being called as patrons in those charges of which we could be witnesses, each of us thought that he should take on himself the part on which he himself could know and judge something.

\ciceroSection{14} And since you have heard Hortensius diligently on the charges of the earlier conspiracy, attend first to this on the conspiracy that was formed in my consulship. Many things, when I was consul, did I hear about the supreme dangers of the commonwealth, many things did I investigate, many things did I learn; never any messenger about Sulla reached me, no informer, no letter, no suspicion. Much weight perhaps that voice ought to carry --- the voice of the man who as consul tracked down by his own counsel the plots against the commonwealth, exposed them by truth, avenged them by greatness of spirit --- when he says that he heard nothing about Publius Sulla, suspected nothing. But I do not yet use this voice for defending him: I shall use it rather to clear myself, that Torquatus may stop wondering that I, who did not stand by Autronius, defend Sulla.

\ciceroSection{15} For what was Autronius's case, what is Sulla's? He sought to break up the canvassing trial first by stirring up a riot of gladiators and runaway slaves, then --- as we all saw --- by stoning and a rush of mob. Sulla, when his own modesty and dignity were of no help to him, sought no resort. The other, once condemned, conducted himself --- not only in his counsels and discourse, but in his very face and look --- as a man manifestly hostile to the highest orders, dangerous to all good men, an enemy to his country. This man thought himself so broken and beaten by that disaster that he reckoned nothing remained to him of his former dignity except what he had kept by his own modesty.

\ciceroSection{16} In this present conspiracy, what was so close as that man's bond with Catiline, with Lentulus? What partnership in the best things has any men joined together so close as theirs in crime, in lust, in audacity? What infamy did Lentulus conceive without Autronius? What deed did Catiline commit without that same man? Meanwhile Sulla with these men sought neither night nor solitude, but did not even join them in ordinary speech and meeting.

\ciceroSection{17} Him the Allobroges, those most truthful informers in the gravest of matters, him the letters and messages of many men convicted; Sulla meanwhile no one accused, no one named. Lastly, when Catiline had been driven out --- or had been let out --- of the city, that man sent arms, horns, trumpets, fasces, standards, legions; left behind within the walls, awaited from outside, hemmed in only by the punishment of Lentulus, he turned at last to fear, never to sanity. This man on the contrary kept so quiet that for the whole of that time he was at Naples --- where men involved in this kind of suspicion are not thought to have been, and where the place itself is more suited to consoling broken spirits than to inflaming them. Therefore, on account of this great unlikeness of men and of cases, I have shown myself unlike in each.

\ciceroSection{18} For Autronius used to come to me, and used to come often, a suppliant with many tears begging that I would defend him, and recalled that he had been my schoolfellow in boyhood, my familiar in youth, my colleague in the quaestorship. He brought up many services of mine to him, some too of his own to me. By all this, judges, I was so bent and broken in spirit that already I was laying aside the memory of the plots he had made against me, was already forgetting that Gaius Cornelius had been sent by him to butcher me in my own house, in sight of my wife and children. Had he plotted these things against me alone, soft and gentle as I am of disposition, by Hercules I should never have held out against his tears and prayers; but when into my mind there came

\ciceroSection{19} the danger of my country, of you all, of this city, of those shrines and temples, of children at the breast, of matrons and maidens, and when those hostile and deadly torches and the universal fire of the whole city, when the weapons, when the slaughter, when the blood of citizens, when the ash of my country began to play before my eyes and to chafe my mind by memory, then at last I withstood him --- and not only that enemy and parricide but also those connections of his, the Marcelli, father and son, of whom the one held in my eyes the gravity of a parent, the other the sweetness of a son. Nor did I think it possible without supreme wickedness to defend in another's accomplice, when I knew it, the same crime which I had punished in others.

\ciceroSection{20} And I, the same man, could not endure Publius Sulla as a suppliant, nor look upon those same Marcelli weeping for this man's dangers, nor sustain the prayers of his connection Marcus Messala. For neither was the case opposed to nature, nor did either man or matter fight against my mercy. Nowhere had there been a name, nowhere a trace, no charge, no informer's word, no suspicion. I took the case up, Torquatus, I took it up, and gladly --- so that I, whom good men, as I hope, have always thought consistent, should not even by the wicked be called cruel.

\ciceroSection{21} Here our friend says, judges, that he cannot endure my kingship. What kingship, pray, Torquatus? My consulship, I take it: in which I commanded nothing and on the contrary obeyed the senators and all good men. In that magistracy a kingship was not, you know, set up by me, but put down. Or do you say that, having so great an empire, so great a power, I was no king then, but say I am king now as a private man? On what title, pray? ``Because the men against whom you gave evidence have been condemned; the man you defend hopes to be acquitted.'' Here on the matter of my testimony I make you this answer: if I lied, that you spoke against the same men too; but if I told the truth, then it is no kingship, when you say what is true on oath, to make it stick. About this man's hope I say only this: Publius Sulla expects from me no resources, no power, nothing in fact except the good faith of his defence.

\ciceroSection{22} ``If you had not taken the case,'' he says, ``he would never have stood up to me, but would have fled without trial.'' Even if I were to grant you this --- that Quintus Hortensius, a man of such weight, that men like these stand not by their own judgment but by mine; even if I were to give you what cannot be believed, that, had I not stood by him, these men would not have stood by him --- which of us is the king, the man whom innocent men do not resist, or the man who does not desert the disaster-stricken? But here also (a thing it was very far from necessary for you to do) you wished to be witty, when you said that Tarquin and Numa and I, third, were foreign kings. I let go now the question of king; I ask why you said that I was a foreigner. For if I am, it is less to be wondered at that I am king (since, as you say, two foreign kings have already been at Rome) than that I have been consul at Rome as a foreigner. ``This is what I say,'' he says: ``that you are from a country town.''

\ciceroSection{23} I admit it, and I add: from that country town from which now for the second time salvation has been sent to this city and empire. But I should much like to learn from you why those who come from country towns seem to you foreigners. No one ever flung that at the elder Marcus Cato, though he had the most enemies, no one at Tiberius Coruncanius, no one at Manius Curius, no one at this our own Gaius Marius, though many envied him. For my part I am vehemently glad to be a man at whom you, however much you may have wished, could throw no insult that does not fit a very great part of the citizens. But yet I think that, on account of the great cause of our connection, you must again and again be warned by me. All cannot be patricians; if you ask the truth, they do not even care; nor do those of your own age think themselves outdone by you on that account.

\ciceroSection{24} And if we seem to you foreigners, whose name and honour have already grown old in this city and in the talk and tongue of men, how must those competitors of yours seem to you foreigners --- men now picked from all Italy who will contend with you for office and for every dignity? Take care not to call any of them a foreigner, lest you be buried beneath the votes of foreigners. If they bring sinew and industry, believe me, they will shake out of you that vain show of words, and will rouse you often out of your sleep, and will not suffer themselves to be surpassed by you in office unless they are bested in virtue.

\ciceroSection{25} And if, judges, the rest of the patricians ought to think me and you foreigners, yet by Torquatus this fault would be passed over in silence; for he is himself, on his mother's side, a townsman --- of a most respectable and most noble stock, but a townsman of Asculum. Either, then, let him show that the men of Picenum alone are not foreigners, or let him be glad that I do not put my stock above his. Wherefore neither call me a foreigner hereafter, lest you be more grievously refuted, nor a king, lest you be laughed at. Unless perhaps it seems to you kingly so to live as to serve no man, no, not even any desire of one's own; to despise all lusts; to want neither gold nor silver nor anything else; to think freely in the senate; to consult the people's interest rather than its will; to yield to no one, to withstand many. If you think this kingly, I confess that I am king. But if it is my power, my dominion, or some arrogant or haughty word that moves you, why do you not bring it forth, rather than the odium of the term and the insult of bad-mouthing?

\ciceroSection{26} If, having laid such great services before the commonwealth, I were to demand of the senate and Roman people no other reward but honourable leisure, who would not grant it? Let the rest have for themselves their honours, their commands, their provinces, their triumphs, their other ornaments of distinguished praise; let it be permitted me, with quiet and tranquil mind, to enjoy the sight of the city which I had saved. What if I do not even demand this? What if my old labour, my anxiety, my services, my works, my watchings, are at the service of friends, are ready for all? What if neither friends miss my zeal in the Forum nor the commonwealth in the Senate-house; if not only no exemption on the score of my deeds, but no excuse of office or of age releases me from labour; if my goodwill, my industry, my house, my mind, my ears are open to all; if not even time is left me for the recollection and contemplation of those things I did for the safety of all --- yet shall this be called a kingship, of which a man willing to be the deputy can nowhere be found?

\ciceroSection{27} Far, far is the suspicion of kingship from me. If you ask who at Rome have tried to seize a kingship, that you may not have to unroll the memory of the annals, you will find them in the portraits of your own house. For my deeds, I take it, have lifted me up too high and inspired me with I-know-not-what proud spirits! Of those deeds, judges, so glorious, so immortal, this much I can say: that I, having rescued this city and the lives of all citizens out of the supreme dangers, shall have had reward enough if from this so great service to all mortals no danger comes back upon me.

\ciceroSection{28} For I remember in what state I did such great deeds, I understand in what city I am living. The Forum is full of those men whom I drove from your necks, judges, but did not put away from my own. Unless indeed you think that there were few who could attempt or hope to be able to wipe out so great an empire. From their hands I could snatch the torches and wrest the swords, as I did; their hardened and unholy intentions I could neither cure nor put away. Wherefore I am not unaware in how great peril I live amid so great a multitude of the wicked, since I see that I alone have undertaken eternal war with all the wicked.

\ciceroSection{29} But if perhaps you envy me my supports, and if these things seem to you kingly --- that all good men of every kind and order link their safety with mine --- console yourself with this: that the minds of all the wicked are most fiercely set against me alone. They hate me not only because I crushed their unholy attempts and their criminal madness, but the more on this account, that they think nothing similar can be attempted any more while I live.

\ciceroSection{30} But why, indeed, should I wonder if anything is wickedly said about me by the wicked, when Lucius Torquatus --- a man, first, who has himself laid these foundations of his youth, who has set this hope of the highest dignity before himself, and second, the son of Lucius Torquatus, that bravest of consuls, that most steadfast of senators, that ever-best of citizens --- is at moments carried away by the immoderation of his words? When he had spoken with lowered voice of the wickedness of Publius Lentulus, of the audacity of all the conspirators --- only just so much that you who approve those things could hear --- of the punishment, of the prison he was speaking with great and querulous voice.

\ciceroSection{31} In which the first absurdity was this: while he wished to make those things he had said softly approved by you, but did not wish those who were standing around the trial to hear them, he did not see that what he said clearly the men to whom he was making himself agreeable would so hear that you also would hear it, you who did not approve it. And the second is now an orator's failing --- not to see what each cause requires. For nothing is so foreign to one who would prosecute another for conspiracy as to seem to be lamenting the punishment and death of conspirators. When that tribune of the plebs does so who alone seems to have been left of them for the lamenting of conspirators, no one wonders; for it is hard to be silent when you grieve. But you, if you do anything of this kind --- not only as a young man of your sort, but in a case in which you wish to seem the avenger of conspiracy --- I greatly wonder at.

\ciceroSection{32} But what I rebuke most is this: that, with such talent and prudence as you have, you do not grasp the cause of the commonwealth --- you who think that those things which all good men in my consulship did for the common safety are not approved by the Roman plebs. Do you reckon any one of these men present here, to whom you were making yourself agreeable against their own will, either so wicked as to wish all this to perish, or so wretched that he would wish himself to perish and would have nothing he wished to be safe? Or does no one rebuke that most distinguished man of your stock and name, who deprived his own son of life that he might strengthen his command over others; do you rebuke the commonwealth, which killed her domestic enemies that she herself might not be killed by them?

\ciceroSection{33} Mark, then, Torquatus, how I shrink from the authority of my consulship! With the loudest voice, that all may hear, I say and shall always say. Be present in mind, all of you, citizens, by whose attendance I am greatly gladdened; lift up your minds and your ears, and attend to me as I speak (so he thinks) on matters that bring odium! I as consul, when an army of ruined citizens, gathered together by clandestine wickedness, had readied the cruellest and most lamentable destruction for the country, when for the going-down and ruin of the commonwealth Catiline in the camp and Lentulus in these temples and houses had been set up as leaders, by my counsels, by my labours, at the danger of my own head, without alarm, without levy, without arms, without an army --- five men seized and confessed --- I freed the city from burning, the citizens from slaughter, Italy from devastation, the commonwealth from death; I redeemed the lives of all citizens, the fabric of the world, this city itself, the dwelling-place of us all, the citadel of kings and foreign nations, the light of the peoples, the home of empire, by the punishment of five mad and ruined men.

\ciceroSection{34} Or did you suppose that I should not say without oath in court the things which I had said on oath at the greatest assembly? And this also I shall add, lest perhaps some wicked man should suddenly begin to love you, Torquatus, and to hope something from you, and that they all may hear it, I shall say it in the clearest voice. Of all those things which I undertook and carried out in my consulship for the safety of the commonwealth, that Lucius Torquatus, when he was my messmate in my consulship, and had been so before in my praetorship, came forward as actor, helper, sharer --- when he was a leader, an instigator, a standard-bearer of the youth. And his father, that man most loving of country, of greatest spirit, of highest counsel, of singular steadfastness, though he was ill, was yet present at all those events, never went apart from me, was the one who in zeal, in counsel, in authority gave me most help, conquering the weakness of his body by the virtue of his mind.

\ciceroSection{35} Do you see how I am snatching you out of the sudden favour of the wicked and reconciling you to all good men? Men who love you and hold you and will hold you always, nor, if perchance you should fall away from me, will on that account suffer you to fall away from them and from the commonwealth and from your own dignity. But now I return to the case, and I take you to witness, judges: a kind of necessity has been laid on me by him to speak so much of myself. For if Torquatus had accused Sulla alone, I too at this time should be doing nothing but defending the man who had been accused. But since he in that whole speech inveighed against me, and since (as I said at the start) he wished to strip the authority from my defence, even if my own indignation did not force me to answer, yet the case itself would have demanded this speech of me.

\ciceroSection{36} You say that Sulla was named by the Allobroges. Who denies it? But read the deposition and see in what manner he was named. They said that Lucius Cassius mentioned, with the others, that Autronius was on his side. I ask whether Cassius named Sulla. Nowhere. They say they themselves asked Cassius what Sulla's view was. See the diligence of the Gauls: who, knowing nothing of these men's lives or characters, and having heard only that they had suffered an equal disaster, asked whether they were of the same mind. What did Cassius say then? Even if he had answered that Sulla was of the same mind and on his side, even so it ought not, in my view, to count as a charge against this man. Why so? Because the man who would push barbarians into a war ought not to lessen their suspicion and to clear the men of whom they seemed to have some suspicion.

\ciceroSection{37} He did not, however, answer that Sulla was on his side. For it would have been absurd that, having named the rest unprompted, he should make no mention of Sulla unless reminded and questioned --- unless perhaps it is likely that the name of Publius Sulla had escaped Cassius's memory. If the man's nobility, if his blasted fortune, if the relics of his former dignity had not been so illustrious, yet the mention of Autronius would have brought back the memory of Sulla. Even, in my view, since Cassius was assembling the names of the leaders of the conspiracy to inflame the spirits of the Allobroges, and since he knew that foreign nations are most moved by nobility, he would not have named Autronius before Sulla.

\ciceroSection{38} And this in particular cannot be allowed: that the Gauls, when Autronius was named, thought, on account of likeness of disaster, that something was to be asked them about Sulla; while Cassius, if this man were in the same crime, could not have brought him to mind even after calling Autronius's name. But yet what did Cassius answer about Sulla? That he did not know for certain. ``He does not clear him,'' he says. I have already said: not even if he were positively to charge him --- and only when questioned --- would that seem to me a charge.

\ciceroSection{39} But for my part, in trials and in inquiries, I think the question is not whether someone is being cleared, but whether he is being charged. For when Cassius says that he does not know, is he covering for Sulla or showing well enough that he does not know? ``He covers for him before the Gauls.'' Why? ``So that they may not denounce him.'' What? If he had thought there was danger that they would ever denounce him, would he have admitted it about himself? ``He did not know, then.'' I see: Cassius was kept in the dark about Sulla alone --- for about the rest he certainly knew, since most of the plotting was put together in his house. He, unwilling to deny that Sulla was of their number lest he give the Gauls less hope, but not daring to lie, said he did not know. And this much is plain: when the man who knew about all denied that he knew about Sulla, his denial has the same force as if he had said that he knew this man was outside the conspiracy. For when a man's knowledge plainly extends to all, his ignorance about anyone ought to count as a clearing. But I no longer ask whether Cassius clears Sulla; this much is enough for me, that there is nothing in the deposition against Sulla.

\ciceroSection{40} Shut out from this charge, Torquatus rushes in upon me again, accuses me; says that I entered the deposition into the public records other than as it had been spoken. Immortal gods! --- for I attribute to you the things that are yours, and indeed I cannot grant my own talent so much as that I, on my own, should have foreseen so many things, so great, so various, so sudden, in that most turbulent storm of the commonwealth --- you indeed kindled my mind then with the desire of saving my country; you turned me from all other thoughts to the one safety of the commonwealth; you finally bore before my mind, in such great darkness of error and ignorance, the clearest light.

\ciceroSection{41} I saw this, judges: unless I had attested the senate's authority for this deposition, while memory was fresh, by public records, the time would come when not Torquatus nor any like Torquatus --- in this I have been much deceived --- but some shipwrecked man of his patrimony, an enemy of peace, a foe to good men, would say that the deposition had been given otherwise, the more easily that, when some wind had been roused against the best men, he might find in the troubles of the commonwealth some harbour for his own troubles. So when the informers had been brought into the senate, I appointed senators who should write down everything --- the informers' words, the questions asked, the answers given.

\ciceroSection{42} And what men! not only of the highest virtue and good faith --- for of that kind there was the largest supply in the senate --- but men whom I knew capable, by memory and knowledge and quickness of writing, of following with the greatest ease whatever was said: Gaius Cosconius, who was then praetor, Marcus Messala, who was then standing for the praetorship, Publius Nigidius, Appius Claudius. There is, I take it, no one who would think that good faith or talent was lacking to these men for a true record. What then? what did I do? Knowing that the deposition had been entered in the public records yet that those records, after our ancestors' fashion, were kept under private custody, I did not hide it, did not keep it at home, but at once ordered it to be copied by all the secretaries, distributed everywhere, made common, and published to the Roman people. I distributed it through all Italy, sent it out to all the provinces; and of that deposition, by which safety had been brought to all, I wished no one to be without his share.

\ciceroSection{43} And so I say there is no spot in the world where the name of the Roman people is, to which this deposition has not also been sent, written out word for word. In which I, in so sudden and brief and turbid a time, foresaw many things by divine help (as I said), not by my own counsel: first, that no one should be able to remember of the danger to the commonwealth or to anyone as much or as little as he wished; second, that none should ever be allowed to find fault with that deposition or charge that it was rashly believed; finally, that nothing should now be sought from me, nothing from my notes --- so that neither forgetfulness on my part nor too great a memory should be supposed, nor at last either base negligence or cruel diligence be imputed.

\ciceroSection{44} But yet from you, Torquatus, I ask: when your enemy had been informed against, when there was a full senate and a recent memory as witness, and when my secretaries (you my close friend and messmate) would have been ready to give you the deposition before they had even entered it in the book, if you had wished --- why, when you saw that things were being done otherwise, did you keep silent, allow it, not complain to me as one closest to you, or, since you are so easily carried into invective against your friends, why did you not protest more sharply and more vehemently? You, when your voice was never heard, when, with the deposition read, transcribed, made public, you were quiet, were silent --- do you suddenly forge so great a thing, and bring yourself to such a pass that, before convicting me of altering the deposition, you would have to confess on your own showing that you had been guilty of the highest negligence?

\ciceroSection{45} Should anyone's safety have been to me of such weight that I should have neglected my own? Should I, by whom the truth had been laid open, contaminate it with some lie? Should I, finally, help any man whom I should have thought to have made such cruel plots against the commonwealth, plots set up especially in my consulship? And if I had now forgotten my severity and steadfastness, was I so out of my mind that, when letters have been invented for posterity's sake, that they may serve as a defence against forgetfulness, I should think that the recent memory of the whole senate could be overthrown by my notebook?

\ciceroSection{46} I bear with you, Torquatus, I have long borne with you, and at times I myself call back and bend back my mind when it is roused to take revenge on your speech. I allow something to your anger, I give to your youth, I yield to friendship, I grant to your father. But unless you set a limit to yourself, you will compel me, forgetful of our friendship, to take account of my own dignity. No one ever brushed me with the slightest suspicion whom I have not overturned and broken. But this I want you to believe me: it is not those whom I think I can most easily overcome whom I am most apt to answer.

\ciceroSection{47} You, since you are not at all ignorant of my custom of speaking, do not abuse this new mildness of mine; do not think that the stings of my speech, which are stowed away, are shaken out; do not at all suppose that what is remitted and conceded to you is lost from me. Both those excuses for your injury weigh with me --- your angry mind, your youth, our friendship --- and yet I do not yet judge you to have strength enough that I ought to wrestle and grapple with you. But if you were stronger in experience and age, I should be the same I am wont to be when I am provoked; now I shall so deal with you that I may seem rather to have borne the injury than repaid the favour.

\ciceroSection{48} Nor indeed can I understand why you are angry with me. If because I defend the man you accuse, why am I not angry at you, because you accuse the man I defend? ``I,'' you say, ``am accusing my enemy.'' And I am defending my friend. ``You ought not to defend anyone in an inquiry on conspiracy.'' On the contrary, no one more --- when he, of all men, has himself learned much about others, never suspected anything of him. ``Why did you give evidence against others?'' Because I was forced. ``Why have they been condemned?'' Because they were believed. ``It is kingship to give evidence against whom you choose and defend whom you choose.'' On the contrary, it is slavery not to give evidence against whom you choose and not to defend whom you choose. And if you begin to consider whether it was more necessary for me to do this or that for you, you will see that more honourably might you have set a limit to your enmity than I to my humanity.

\ciceroSection{49} Yet, in fact, when the highest honour of your family was at stake --- that is, your father's consulship --- did that wisest of men, your father, not get angry with his closest connections when they both defended Sulla and praised him? He understood that we have been handed down this discipline by our ancestors: that no man's friendship hinder us from warding off dangers. But that contest was wholly unlike this trial. Then, with Publius Sulla shattered, the consulship was being secured for you, as it was secured. It was a contest of office. You kept clamouring that you were taking back what had been snatched from you: beaten on the Field, you wished to win in the Forum. Then those who fought against you for this man's safety, your closest friends with whom you were not angry, were stripping the consulship from you, were resisting your honour, and yet did this with your friendship unbroken, with duty whole, after the old example and practice of every best man.

\ciceroSection{50} But for myself --- what ornaments of yours do I oppose, what dignity of yours do I resist? What is there now that you would seek from this man? The honour has gone to your father, the insignia of honour have been brought to you. You come, dressed in this man's spoils, to mangle the man you have killed; I defend and protect a man fallen and stripped. And here you both rebuke me because I defend him and grow angry; I, on the other hand, not only am not angry with you but do not even rebuke what you have done. For I think that you set yourself what you thought you ought to do, and could be a sufficient judge of your own duty.

\ciceroSection{51} But the son of Gaius Cornelius prosecutes him, and that ought to count as much as if his father were giving evidence. O what a wise father Cornelius is, who, leaving aside what reward there is wont to be in giving evidence, what disgrace there is in confessing, has taken on the same end through his son's accusation! But what is it then that Cornelius is informing of through this boy? If old things, unknown to me, shared with Hortensius --- Hortensius has answered. But if, as you say, that attempt of Autronius and Catiline, when they wished to make slaughter on the Field at the consular elections held by me --- I myself saw Autronius then on the Field. But why did I say I saw? --- I saw; for you then, judges, were not anxious nor suspecting anything; I, screened by a strong guard of friends, then put down the forces and the attempt of Catiline and Autronius.

\ciceroSection{52} Now is there anyone who says that Sulla then breathed aspiration toward the Field? And yet, if at that time he had joined himself with Catiline in the partnership of crime, why was he keeping clear of him, why was he not with Autronius, why in a like case are like marks of guilt not found? But since Cornelius even now is doubting whether to inform, and (as you say) is shaping his son for this dim and shadowy informing, what then does he say of that night when, among the scythe-makers' shops, at Marcus Laeca's, on the night which followed the day after the Nones of November, in my consulship, by Catiline's summons there was a meeting? That night was the sharpest and bitterest of all the times of the conspiracy. Then was set Catiline's day for going out, then the terms for the rest's staying, then the dividing-up through the whole city of slaughter and of fires; then your father, Cornelius (a thing he at last admits at this date), demanded for himself that obliging task, that, when at first light he came to greet the consul, he should be admitted (after my custom and by the right of friendship) and butcher me on my couch.

\ciceroSection{53} At this time, when the conspiracy was burning at its hottest, when Catiline was going out to the army, Lentulus was being left behind in the city, Cassius set over the burnings, Cethegus over the slaughter, when it was being prescribed for Autronius to seize Etruria, when all things were being arranged, drawn up, prepared --- where was Sulla, Cornelius? At Rome? On the contrary, far away. In those regions toward which Catiline was making? Far further still. In the territory of Camerinum, of Picenum, of Gaul, the parts which the disease, as it were, of that madness had most pervaded? Nothing less. He was, as I said before, at Naples, in that part of Italy which was most free of any such suspicion.

\ciceroSection{54} What does Cornelius then disclose, what does he bring forward --- either he himself, or you who carry these instructions for him? That gladiators were bought in Faustus's name for slaughter and tumult? ``Just so: gladiators were brought in.'' Whom we see to be due by his father's will. ``A troop has been picked up.'' Which, if it had been passed over, another troop could have provided Faustus's games. Would, indeed, that this very thing could content not only the malice of the unfair but the expectation of the fair! ``It was done in great haste, when the time of the games was far away.'' As if, indeed, the time for giving the games were not now strongly approaching. ``Faustus not knowing it, when he neither knew nor wished, the troop was made ready.''

\ciceroSection{55} But there are letters of Faustus, by which he begs in entreaty from Publius Sulla that he buy gladiators, and these very gladiators --- not only sent to Sulla but to Lucius Caesar, Quintus Pompeius, Gaius Memmius, by whose advice the whole thing was done. ``But Cornelius, his freedman, was put in charge of the troop.'' If now in providing the troop there is no suspicion, who was put in charge has nothing to do with it; but yet he undertook the slavish duty of looking after the iron-gear, but was never in charge, and that business throughout was conducted by Bellum, Faustus's freedman.

\ciceroSection{56} ``But Sittius was sent by him into Further Spain to throw that province into confusion.'' First, judges, Sittius set out in the consulship of Lucius Julius and Gaius Figulus --- some good while before Catiline's madness and the suspicion of this conspiracy. Next, he set out not then for the first time, but when he had been in those same parts for several years some time before, on the same business; and he set out not just on his own account but on a necessary errand --- a great affair contracted with the king of Mauretania. Then, when he had gone, Sulla managing his estate and conducting it, by the sale of very many and very fine estates of Publius Sittius the same man's debt was paid off. So that the cause which drove the rest to crime --- the desire to keep their possessions --- was not in Sittius's case, his estates having been reduced.

\ciceroSection{57} Then how incredible, how absurd this is: that he who would make slaughter at Rome, would set this city ablaze, should send away from him his closest connection and dispatch him to the ends of the earth! Was it that he might more easily accomplish at Rome what he was attempting, if there were disorder in Spain? But these very things were going on by themselves, without any link. Or in such great matters, with counsels so new, so dangerous, so turbulent, did he think that the man most loving of him, most close, most joined to him by services, by daily intercourse, by use, was to be sent away? It is not likely that the man whom he would have always with him in prosperity, whom he would have with him in leisure, he should send away from him in adversity and in that very tumult he was preparing.

\ciceroSection{58} As for Sittius himself --- for I cannot desert the cause of an old friend and guest --- is the man, or is his family and breeding, of such a kind that this can be believed: that he should have wished to make war on the Roman people? That, when his father stood out, while the rest of his neighbours and the surrounding peoples deserted, in singular service and good faith to our commonwealth, the son should have thought that an unholy war must be undertaken by him against his country? Whose debt we see, judges, to have been contracted not from lust but from zeal in conducting business --- a man who owed at Rome on such terms that the largest sums were owed to him in provinces and in kingdoms; and when he sought these in, he did not allow his agents in his absence to undergo any of the burden; he chose rather to sell off all his possessions and to be stripped of a most well-furnished patrimony than that any delay should be made to any one of his creditors.

\ciceroSection{59} From this kind, indeed, judges, I have never feared --- when I was busy in that storm of the commonwealth. That was the kind of men terrible and to be greatly feared, who clung with so great a love to their possessions embracing them that you would say their limbs could sooner be torn and pulled away. Sittius never thought he had a kinship with his estates. So he saved himself not only from the suspicion of so great a wickedness, but even from all men's talk --- not by arms, but by his patrimony.





