% In Catilinam IV (in-catilinam-4) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} I see, senators, that the faces and eyes of you all are turned upon me; I see that you are anxious not only about your own danger and that of the commonwealth but, if that be put aside, also about mine. Your goodwill toward me is, in evil times, pleasing, and in pain, welcome; but, by the immortal gods, lay it aside, and forgetting my safety think of yourselves and of your children. If this condition of the consulship has been given me, that I should bear all bitterness, all pains and torments, I shall bear them not only bravely but also gladly --- provided only that by my labours dignity and safety be brought to you and to the Roman people.

\ciceroSection{2} I am that consul, senators, for whom not the Forum, in which all equity is contained, not the Field of Mars, consecrated by the consular auspices, not the Senate-house, the supreme refuge of all nations, not my house, the common refuge, not the bed given for rest, not, in short, this seat of office, has ever been free from the danger and snares of death. I have kept silent on many things, borne many, conceded many, healed many --- with some pain to myself --- in your fear. Now if the immortal gods have willed this end to my consulship, that I should snatch you and the Roman people from a most wretched slaughter, your wives and children and the Vestal virgins from a most bitter outrage, the temples and shrines and this most beautiful country of us all from a most foul flame, all Italy from war and devastation --- whatever fortune is set before me alone, let it be undergone. For if Publius Lentulus, drawn on by soothsayers, thought his name fated to the ruin of the commonwealth, why should I not rejoice that my consulship has come to be almost fated to the safety of the Roman people?

\ciceroSection{3} Therefore, senators, take counsel for yourselves, look out for the country, save yourselves, your wives, your children, your fortunes; defend the name and safety of the Roman people; cease to spare me and to think of me. For first I ought to hope that all the gods who preside over this city will return thanks to me as I deserve; then, if anything happens, I shall die with an even and prepared mind. For neither can a base death befall a brave man, nor an untimely one a consular, nor a wretched one a wise man. And yet I am not so iron as not to be moved by the grief of my dearest and most loving brother who is here, and by the tears of all those by whom you see me beset. Nor does my mind fail to call me back home --- by my wife, undone, my daughter cast down with fear, and my little son, whom the commonwealth seems to me to embrace as a hostage of my consulship; nor by him who, awaiting the end of this day, stands in my sight, my son-in-law. I am moved by all these things, but only that they may be safe with you, even if some violence shall crush me, rather than that both they and we together perish in the ruin of the commonwealth.

\ciceroSection{4} For which reason, senators, lean to the safety of the commonwealth, look around at all the storms which hang over you unless you provide. It is not Tiberius Gracchus, because he wished to be made tribune of the plebs a second time, not Gaius Gracchus, because he tried to stir up the agrarian party, not Lucius Saturninus because he killed Gaius Memmius, who is being brought into some peril and into the judgment of your severity: those are held who remained at Rome to burn the city, to slaughter you all, to receive Catiline; the letters are held, the seals, the hands, in short the confession of each. The Allobroges are solicited, the slaves stirred up, Catiline summoned --- that is, the design has been entered upon that, with all killed, no one should be left even to lament the name of the Roman people and to bewail the calamity of so great an empire.

\ciceroSection{5} All these things the informers have brought, the accused have confessed, you have already by many judgments judged: first, in that you returned thanks to me in singular words and decreed that by my courage and diligence the conspiracy of lost men had been brought to light; then in that you compelled Publius Lentulus to abdicate the praetorship; then in that you decreed that he and the rest concerning whom you had judged be given over into custody; and most of all in that you decreed a thanksgiving in my name --- which honour has been granted to no togate citizen before me; finally, yesterday, you gave the most ample rewards to the envoys of the Allobroges and to Titus Volturcius. All which things are of such a kind that those who have been delivered into custody by name seem without any doubt to have been condemned by you.

\ciceroSection{6} But I have determined to refer the matter to you, senators, as if untouched, both as to what you judge of the deed and what you decide on for the punishment. Those things I shall set forth which are the consul's. I have long seen that a great frenzy was abroad in the commonwealth, and that some new evils were being mixed and stirred up; but I never thought that a conspiracy this great, this ruinous, was being held by citizens. Now, whatever it is, in whatever direction your minds and opinions incline, you must decide before nightfall. How great a crime has been brought before you, you see. To it, if you think few are accessory, you are vehemently mistaken. This evil has been disseminated more widely than is supposed; it has flowed not only through Italy but has even crossed the Alps, and creeping in obscurity has now seized many provinces. To suppress it by holding off and putting it off cannot in any way be done; in whatever way it pleases you, it must be quickly punished by you.

\ciceroSection{7} I see that there are so far two opinions: one of Decimus Silanus, who decrees that those who tried to destroy these things should be visited with death; the other of Gaius Caesar, who removes the punishment of death and embraces all the bitterness of the other punishments. Each, both for his own dignity and for the magnitude of affairs, moves in the highest severity. The one judges that those who tried to deprive us all and the Roman people of life, who tried to destroy the empire, to extinguish the name of the Roman people, ought not to enjoy a single moment of life and this common breath; and he recalls that this kind of punishment has often been used against wicked citizens in this commonwealth. The other understands that death has been set up by the immortal gods not for the sake of punishment, but as either the necessity of nature or the rest from labours and miseries. So the wise have never met it unwillingly, the brave often even gladly. But chains --- and those everlasting --- have certainly been devised as a singular punishment of unspeakable crime. He orders them to be apportioned among the municipalities. The thing seems to have an unfairness in it, if you should wish to command; a difficulty, if you should ask.

\ciceroSection{8} Yet let it be decreed, if it pleases. For I shall undertake it, and, as I hope, shall find men who think it not against their dignity to refuse what for the safety of all you shall have decided. He adds a heavy penalty for the municipalities, if any of these prisoners shall break his chains; he sets dreadful guards around them, worthy of the crime of lost men; he provides that no one should be able to lighten the punishment of those whom he condemns either through the Senate or through the people; he takes away even hope, which alone is wont to console a man in his miseries. Besides, he orders the goods to be made public; he leaves to wicked men life alone --- which if he had taken away, he would have removed many miseries of mind and body in one pang, and all the punishments of crimes. So that some terror might be set before the wicked in life, the ancients of old wished such punishments to be set up among the dead for the impious, because, you see, they understood that, with these removed, death itself would not be greatly to be feared.

\ciceroSection{9} Now, senators, I see what is to my interest. If you shall have followed the opinion of Gaius Caesar, since he has followed in the commonwealth that path which is held to be popular, perhaps with this sponsor and witness of the opinion popular outbreaks against me will be less to be dreaded; but if the other, I do not know whether more trouble may not be brought upon me. Yet let the interest of the commonwealth conquer the reckoning of my dangers. For we have from Caesar, as both his own dignity and the greatness of his ancestors demanded, an opinion that is, as it were, a hostage of his perpetual goodwill toward the commonwealth. It has been understood what is the difference between the levity of mob-orators and a mind truly popular providing for the safety of the people.

\ciceroSection{10} I see that of those who wish to be held popular some are absent, lest, of course, they cast a vote on the life of Roman citizens. The same man both two days ago gave Roman citizens into custody, and decreed a thanksgiving for me, and yesterday rewarded the informers most amply. To no one is it now a doubt what he, who decreed custody for the accused, congratulation for the inquirer, reward for the informer, judged about the whole matter and case. But Gaius Caesar understands that the Lex Sempronia was set up about Roman citizens; that whoever is an enemy of the commonwealth can in no way be a citizen; finally, that the very mover of the Lex Sempronia paid the penalty to the commonwealth at the people's order. The same man does not think that Lentulus, that lavish prodigal, even can be called popular, since he has so bitterly, so cruelly thought about the ruin of the Roman people, the destruction of this city. So this most mild and most gentle man does not hesitate to commit Publius Lentulus to everlasting darkness and chains, and provides for the future that no one may afterwards be able to vaunt himself by lightening this man's punishment, and to be ``popular'' in the ruin of the Roman people. He adds the confiscation of the goods, that every torment of mind, and want and beggary of body besides, may follow.

\ciceroSection{11} For which reason, whether you decide on this, you will have given me a comrade for the assembly dear and pleasant to the people, or whether you prefer to follow Silanus's opinion, the Roman people will easily free me and you of any reproach of cruelty, and I shall maintain that it was much the gentler. And yet, senators, what cruelty can there be in punishing the savagery of so great a crime? For I judge from my own feeling. So may it be permitted me to enjoy with you the commonwealth saved, as I, in being more vehement in this case, am moved not by atrocity of mind --- for who is gentler than I? --- but by a singular humanity and mercy. For I seem to myself to see this city, the light of the world and the citadel of all nations, suddenly falling in one fire. I see in mind the wretched and unburied heaps of citizens in a buried country; before my eyes turns the look of Cethegus and his frenzy revelling in your slaughter.

\ciceroSection{12} When I have set before me Lentulus reigning, as he himself has confessed he had hoped from the fates, Gabinius beside him in purple, Catiline come with his army --- then the lamentation of mothers of households, then the flight of girls and boys, and the outrage of the Vestal virgins make me shudder. And because these things seem to me vehemently wretched and pitiable, on that account I shall show myself severe and vehement against those who wished to bring them to pass. For I ask: if any father of a household, with his children killed by a slave, his wife killed, his house burned, did not visit upon the slaves the most bitter punishment, would he seem merciful and pitying, or most inhuman and most cruel? To me, indeed, the man would seem dangerous and iron who did not soften his own pain and torment by the pain and torment of the guilty. So we, in the case of these men who wished to butcher us, our wives, our children, who tried to destroy the houses of each one of us and this universal dwelling-place of the commonwealth, who set on foot to settle the Allobrogan nation in the traces of this city and on the ash of an empire burned away --- if we shall be most vehement, we shall be held merciful; but if we shall wish to be more remiss, the report of supreme cruelty in the ruin of country and citizens must be undergone by us.

\ciceroSection{13} Unless to anyone Lucius Caesar, that bravest and most loyal man, seemed two days ago crueller, when he said that the husband of his own sister, that most select woman --- present and listening --- ought to be deprived of life; when he said that his own grandfather had been killed by a consul's order, and his grandfather's son, an underage boy, sent as legate by his father, had been killed in prison. What, of these, was a like deed; what design entered upon to destroy the commonwealth? The will to bounty was then abroad in the commonwealth, and a kind of contention of parties. And at that time the grandfather of this Lentulus, a most distinguished man, armed, pursued Gracchus. He even at that time received a heavy wound, that nothing of the sum of the commonwealth might be diminished; this man, to overthrow the foundations of the commonwealth, summons the Gauls, stirs up the slaves, calls Catiline, allots us to be butchered to Cethegus and the rest of the citizens to be killed to Gabinius, the city to be burnt to Cassius, all Italy to be wasted and despoiled to Catiline. I think you should fear less to seem to have decided some severer thing in this so monstrous and unspeakable crime: much more is it to be feared that by remission of penalty we shall seem cruel toward our country than that by severity of chastisement we shall seem too vehement against the bitterest enemies.

\ciceroSection{14} But what I overhear, senators, I cannot dissemble. For voices are thrown about which come to my ears, of those who seem to fear that I have not enough protection for what you decide today must be carried out. Everything, senators, has been provided and prepared and set in order, both with my supreme care and diligence and much more even with the will of the Roman people to keep the supreme empire and to preserve the common fortunes. All men of all orders are present, of every kind, in short of every age; the Forum is full, the temples around the Forum are full, all the approaches to this temple and place are full. For this is the only cause that has been found, since the founding of the city, in which all felt one and the same thing --- except those who, when they saw they must perish themselves, wished to perish with all rather than alone.

\ciceroSection{15} These men I except and willingly set apart; nor do I think they are to be counted in the number of wicked citizens but of the bitterest enemies. The rest, however, immortal gods! with what throng, what zeal, what courage do they agree for the common safety and dignity! Why should I here recall the Roman knights? who yield to you the supreme rank in counsel in such a way as to contend with you in love of the commonwealth; whom this day, called back from many years of dissension with this order to partnership and concord, joins with you and with this case. Which conjunction, if confirmed in my consulship, we shall keep perpetual in the commonwealth, I assure you that no civil and domestic evil shall hereafter come to any part of the commonwealth. With like zeal for defending the commonwealth I see the tribunes of the treasury, those bravest men, have come together; and the scribes likewise as a body, whom, though by chance this day had brought them together at the treasury, I see turned from the expectation of the lot to the common safety.

\ciceroSection{16} The whole multitude of the freeborn is present, even of the slenderest. For who is there to whom these temples, the look of the city, the holding of liberty, this very light, in short, and the common soil of country are not both dear and indeed sweet and pleasant? It is worth while, senators, to know the zeal of the freedmen, who, having attained by their own courage the fortune of this state, judge this truly to be their country, when certain men born here, and born of the highest place, have judged it to be not their country but the city of enemies. But why do I recall these orders and men, whom private fortunes, the common commonwealth, the liberty (which is sweetest of all things) have stirred up to defend the safety of the country? There is no slave, who is only of a tolerable condition of slavery, who does not shudder at the audacity of citizens, who does not desire these things to stand, who does not contribute as much will as he dares and can to the common safety.

\ciceroSection{17} For which reason if any of you is by chance moved by what has been heard --- that some pander of Lentulus's is running about among the shops, hoping that for a price the minds of the needy and unschooled can be solicited --- this thing has indeed been begun and tried, but no men were found so wretched in fortune or so lost in will as not to wish that the very place of their seat and work and daily gain, that the bed and their little couch, that this quiet course of their life, be safe. Indeed by far the greatest part of those who are in the shops, no, the whole class --- for that should rather be said --- is most loving toward quiet. For all their stock-in-trade, all their work and gain, are sustained by a throng of citizens, are nourished by quiet; whose gain, if it tends to be diminished by the closing of shops, what would happen if they were burned?

\ciceroSection{18} Since these things are so, senators, the protection of the Roman people is not lacking to you: see to it that you not seem to be lacking to the Roman people. You have a consul reserved out of many dangers and snares and out of the midst of death not for his own life but for your safety. All orders agree in mind, will, voice for the saving of the commonwealth. Beset by the firebrands and weapons of an impious conspiracy, the common country stretches out her hands as a suppliant to you; to you she commits herself, to you the lives of all her citizens, to you the citadel and the Capitoline, to you the altars of the Penates, to you that everlasting fire of Vesta, to you the temples and shrines of all the gods, to you the walls and houses of the city. Besides, on your own life, on the souls of your wives and children, on the fortunes of all, on your seats and hearths, you must judge today.

\ciceroSection{19} You have a leader who remembers you and forgets himself --- which is not always given; you have all the orders, all men, the entire Roman people, what we see today for the first time in a civil case, feeling one and the same thing. Consider what labours founded the empire, what courage made liberty firm, what kindness of the gods increased and heaped up these fortunes --- all that one night might have nearly destroyed. That this can never hereafter not only not be carried through but even thought of by citizens, must be provided this day. And these things I have spoken not to rouse you, who almost outrun me in zeal, but that my voice, which ought to be in the commonwealth foremost, might seem to have discharged its consular duty.

\ciceroSection{20} Now, before I return to your opinion, I shall say a few words about myself. I see that, as great as is the band of the conspirators, which you see is very great, I have brought upon myself a multitude of enemies of like size; but I judge that multitude base and weak and abject. But if at some time, stirred up by anyone's frenzy and crime, that band shall prevail more than your dignity and that of the commonwealth, yet, senators, I shall never repent of my deeds and counsels. For death, with which they perhaps threaten me, is prepared for all; the praise of life as great as you have honoured me with by your decrees, no man has attained: for to others you have always decreed congratulation for affairs well managed; to me alone for the commonwealth saved.

\ciceroSection{21} Let that famous Scipio be distinguished, by whose counsel and courage Hannibal was compelled to return to Africa and to depart from Italy; let the other Africanus be adorned with surpassing praise, who destroyed two cities most hostile to this empire, Carthage and Numantia; let that Paulus be reckoned a distinguished man, whose chariot the most powerful and most noble king Perseus once honoured; let Marius be in everlasting glory, who twice freed Italy from siege and from the fear of slavery; let Pompey be set before all, whose deeds and virtues are bounded by the same regions and limits as the sun's course: there will surely be among the praises of these some place for our glory --- unless perhaps it is more to throw open to us provinces whither we may go than to take care that even those who are absent should have a place to which to return as victors.

\ciceroSection{22} Although in one point the condition of foreign victory is better than of domestic --- because foreign-born enemies, when crushed, either serve, or, when received, count themselves bound by the kindness; but those who from the number of citizens, depraved by some madness, have once begun to be enemies of country --- these, when you have driven them off from the ruin of the commonwealth, you can neither coerce by force nor placate by kindness. For which reason I see that with lost citizens an everlasting war has been undertaken by me. That war I trust can easily be warded off from me and from mine, by your help and that of all loyal men, and by the memory of so great dangers, which not only in this people which has been saved but in the conversations and minds of all nations will always cling. Nor will any force be found so great that can break and shake your conjunction with the Roman knights, and so great a conspiring of all loyal men.

\ciceroSection{23} Since these things are so, in return for my command, for my army, for my province which I neglected, for my triumph and the rest of the badges of praise which by me, on account of the guarding of the city's and your safety, have been refused, in return for my provincial clientships and friendships --- which yet I keep up by my urban resources with no less labour than I procure --- in return, then, for all these things, in return for my singular zeal toward you, and in return for this diligence in saving the commonwealth, which you see, I ask of you nothing but the memory of this time and of my whole consulship. So long as that shall be fixed in your minds, I shall consider myself fenced in by the safest wall. But if the violence of the wicked shall disappoint and overcome my hope, I commend to you my little son, for whom there will assuredly be sufficient protection not only for safety but even for dignity, if you remember that he is the son of him who at his sole risk has saved all these things.

\ciceroSection{24} For which reason concerning the supreme safety of yourselves and of the Roman people, concerning your wives and children, concerning your altars and hearths, concerning the shrines and temples, concerning the houses and seats of the whole city, concerning the empire and liberty, concerning the safety of Italy, concerning the entire commonwealth, decide diligently, as you have begun, and bravely. You have a consul who will not hesitate either to obey your decrees and, what you shall have decided, to defend it as long as he lives, and to make it good by himself.
