% De Lege Agraria II (de-lege-agraria-2) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} It is established by custom, citizens, and by the institution of our ancestors, that those who through your kindness have attained the right of family portrait-busts should hold their first public address to link the gratitude they owe your kindness with the praise of their own forebears. In which speech a few are sometimes found worthy of their ancestors' place; but most achieve only this --- to make it look as though so much was owed to their ancestors that even something was left over to be paid out to their descendants. To me, citizens, it is not granted to speak before you about my ancestors --- not that they were not such men as you see those of us were who were begotten of their blood and trained by their teachings, but because they had no part in popular renown or in the light of public office.

\ciceroSection{2} About myself, however, I fear that to speak before you would be arrogant, to keep silent ungrateful. For both to recall by what efforts I have attained this dignity is exceedingly hard, and to be silent about your great kindnesses I cannot in any way. Therefore I shall observe a fixed measure and moderation in speaking: I shall recall what I have received from you; why I am worthy of your highest honour and your singular judgment, I shall myself say with restraint, if it must be said, supposing that you yourselves who passed the verdict will think the same.

\ciceroSection{3} You made me, after a very long interval reaching nearly to the limit of our memory, the first ``new man'' to be made consul; and that place which the nobility held fortified by garrisons and walled around by every device, you tore open under my leadership and willed it to lie open to virtue for the future. And you made me consul not only --- which is in itself most splendid --- but you made me so in a way in which few nobles in this city have been made consuls, and no new man before me. For if you will recall the new men, you will find that those who were made consul without a defeat were made so by long labour and some opportunity, when they had stood many years after their praetorship and rather later than their age and the laws would have allowed; while those who stood in their year did not become consul without a defeat. I am the only one of all the new men we can remember who stood for the consulship as soon as it was permitted, was made consul as soon as I stood --- so that your honour was sought to the day of my own time, not snatched up at the chance of another's candidacy, not extorted by long entreaties, but won by dignity.

\ciceroSection{4} That is the most splendid thing I mentioned just now, citizens --- that with this honour you adorned me, the first new man for many years, at my first standing, in my year. But still, more magnificent and more ornate than that, nothing can be: that at my election you cast not the silent ballot, that vindicator of secret liberty, but you carried before you a living voice as the index of your goodwill and zeal toward me. And so it was not the final counting of votes, but those first thronging assemblies of yours, not the single voices of the heralds, but the one voice of the entire Roman people, that proclaimed me consul.

\ciceroSection{5} This kindness of yours, citizens, so distinguished and so singular, I count great indeed for the joy and gladness of my mind, but greater still for my care and anxiety. For there are turning in my mind many heavy thoughts that grant me no part of either day's or night's rest: first, the keeping of my consulship --- a charge that is hard and great for everyone, but for me beyond the rest, since for me an error finds no pardon, a thing rightly done finds slim praise wrung out of unwilling givers; faithful counsel is not shown to me in doubt, certain support is not offered me in toil from the nobility.

\ciceroSection{6} If I alone were brought into some peril, I should bear it, citizens, with a more even mind. But it seems to me certain men, if in any matter they shall judge me to have slipped --- not in counsel only, but even by accident --- will censure all of you who set me before the nobility. For my part, citizens, I judge that I must endure all things rather than fail to conduct my consulship in such a way that in all my actions and counsels your action and counsel about me may be praised. There is added to me also that highest labour and most difficult method of holding the consulship, that I have decreed I must not use the same law and condition as previous consuls, who shunned approach to this place and the sight of you in part with great care, in part did not pursue it with vigour. But I, not only in this place, where it is most easily said, but in the Senate itself, where there seemed no place for such a voice, declared in my very first speech on the Kalends of January that I would be a consul friendly to the people.

\ciceroSection{7} For I cannot in any way bring it about that, when I see myself made consul not by the zeal of powerful men nor by the surpassing favour of a few, but by the judgment of the entire Roman people, made consul in a way that I should be set far before the most noble men, I should not seem to be a friend of the people both in this magistracy and in the whole of my life. But for the force and interpretation of this word I have great need of your wisdom. For a great error is in circulation because of the treacherous pretences of certain men who, when they fight against and obstruct the people's not only advantages but also safety, wish by their oratory to seem friends of the people.

\ciceroSection{8} What kind of commonwealth I received on the Kalends of January, citizens, I see well: full of anxiety, full of fear, in which there was no evil, no calamity, that the loyal did not dread and the wicked await; all turbulent designs against the present condition of the commonwealth and against your peace were said to be in part being undertaken, in part already undertaken when we were consuls-elect. Faith had been driven from the Forum, not by any single blow of fresh disaster, but by suspicion and the disturbance of the courts, the weakening of cases already judged. New dominations, extraordinary not commands but kingships, were thought to be the object of pursuit.

\ciceroSection{9} When I not only suspected this but plainly saw it --- for it was not being done in secret --- I said in the Senate that in this magistracy I would be a consul friendly to the people. For what is so dear to the people as peace? in which not only those whom nature has endowed with sense, but even roofs and fields, seem to me to rejoice. What so dear to the people as liberty? which you see sought not only by men but even by beasts, and set before all things. What so dear to the people as quiet? which is so pleasant a thing that you and your ancestors and every bravest man have thought the greatest labours should be undertaken so that one might at last be in quiet, especially in command and dignity. Indeed, on this account also we owe to our ancestors a particular praise and gratitude --- that by their labour it has been brought about that we may be in quiet without fear. How then can I not be a friend of the people, when I see all these things, citizens --- foreign peace, the liberty proper to your race and name, domestic quiet, in short all the things dear and ample to you --- entrusted to the keeping and as it were the patronage of my consulship?

\ciceroSection{10} For neither, citizens, ought it to seem to you pleasant or popular if some bounty has been promulgated, the kind of thing that can be paraded in words but in truth cannot be done unless the treasury is drained dry; nor are these to be reckoned popular: disturbances of the courts, the undoing of cases already judged, the restoration of the condemned --- which in stricken states, when affairs are already beyond hope, are wont to be the last terminations of disasters. Nor, if any men promise lands to the Roman people, if they are pushing one thing in secret while they parade another in hope and the colour of pretence, are they to be thought friends of the people. For I will speak truly, citizens: the very kind of agrarian law I cannot blame. For there comes to my mind that two most distinguished men, of the highest talent, most loving toward the Roman commons, Tiberius and Gaius Gracchus, settled the commons on public lands which had been previously held by private men. I am not, however, the kind of consul who, like very many, judges it impious to praise the Gracchi, by whose counsels, wisdom, and laws I see many parts of the commonwealth set in order.

\ciceroSection{11} Therefore, when at the start, just consul-designate, it was being announced to me that the tribunes-designate were drafting an agrarian bill, I was eager to learn what they had in mind. For I judged that, since we had to hold our offices in the same year, there ought to be some partnership between us in the well-administering of the commonwealth.

\ciceroSection{12} When I tried to insinuate myself in friendly fashion into their conversations and to lend my services, I was kept in the dark, shut out; and when I made it clear that, if a law were to seem to me useful to the Roman commons, I would be its sponsor and helper, they nevertheless rejected this generosity of mine and denied that I could be brought to approve any bounty. I made an end of offering myself, lest perhaps my eagerness should seem either treacherous or shameless. Meanwhile they did not cease to meet privately among themselves, to bring in certain private persons, to add the night and solitude to their secret gatherings. By which things how great a fear we were in, you can easily make out by conjecture from the anxiety in which you were yourselves at that time.

\ciceroSection{13} The tribunes of the plebs at last enter on their office. The public address of Publius Rullus is awaited, since he was both the chief of the agrarian bill and was bearing himself more truculently than the rest. When designate already he was practising another expression, another tone of voice, another walk, with shabbier clothing, his body unkempt and shaggy, his hair longer than before and his beard greater, so that by his eyes and looks he seemed to be announcing to all his tribunician violence and threatening the commonwealth. I was waiting for the man's law and his speech: at first no law is published, he orders an assembly to be called for the day before the Ides. There is a great rush of expectation. He unfolds quite a long speech, in very fine words. There was one thing that seemed to me defective: that out of so vast a throng no one could be found who was able to make out what he was saying. Whether he did this for the sake of stratagem, or whether he delights in this kind of eloquence, I do not know. Yet those who had stood somewhat sharper-witted in the assembly conjectured that he had wished to say I-know-not-what about an agrarian law. At length, when I was already designate, the law is published. At my order, several scribes hurry up at the same moment and bring me the law transcribed.

\ciceroSection{14} On every count I can assure you, citizens, that I came to read and study the law in this spirit: that, if I should perceive it to be fitted to your interests and useful, I should be its sponsor and helper. For neither does the consulship by nature, by separation, or by any inherent hatred have undertaken some kind of war with the tribuneship --- because very often loyal and brave consuls have stood up to seditious and wicked tribunes of the plebs, and because tribunician violence has sometimes resisted consular caprice. It is not the unlikeness of the powers that causes the dispute, but the disjunction of minds.

\ciceroSection{15} And so I took the law into my hands in this spirit: that I wished it to be suited to your interests, and of such a kind that a consul popular in deed, not in oratory, could honourably and gladly defend it. And from the first clause of the law to the last I find, citizens, that nothing else has been thought, nothing else undertaken, nothing else done, except this: that ten kings should be set up over the treasury, the revenues, all the provinces, the entire commonwealth, the kingdoms, the free peoples, in short the lords of the world --- under the pretence and the name of an agrarian law. So I assert, citizens, that by this fine and popular agrarian law nothing is given to you, all things are bestowed on certain men, lands are paraded before the Roman people, even your liberty is torn away, the fortunes of private men are increased, those of the state are drained, and finally --- which is most outrageous --- through a tribune of the plebs, whom our ancestors wished to be the protector of liberty and its guardian, kings are set up in the state.

\ciceroSection{16} When I shall have set these things out, citizens, if they shall seem to you false, I shall follow your authority, I shall change my own opinion. But if you perceive that snares are being set against your liberty under the pretence of a bounty, do not hesitate to defend, with no labour of your own, with the consul as your helper, the liberty won by your ancestors at vast cost of sweat and blood and handed down to you. The first clause of the agrarian law is one by which, as they think, you are tested gently in the matter of your liberty's diminution, to see how you can bear it. For he orders that the tribune of the plebs who has carried the law create the ten commissioners through seventeen tribes, so that whomever nine tribes have made, that man shall be a decemvir.

\ciceroSection{17} Here I ask why he took the beginning of his measures and his laws from this point --- that the Roman people be deprived of its vote. So often by agrarian laws have commissioners been set up: triumvirs, quinquevirs, decemvirs; I ask of this popular tribune of the plebs whether they were ever created except through the thirty-five tribes. For when it is fitting that all powers, commands, and commissions should proceed from the entire Roman people, certainly those most of all that are set up for some advantage and benefit of the people, in which both all together pick out the man whom they think will most consult the Roman people, and each one for himself, by his zeal and his vote, can pave a way to obtain the favour. To this tribune of the plebs above all it has come to deprive the entire Roman people of its votes, and to call a few tribes, not on a fixed condition of right but by the favour of a chance lot, to the exercise of liberty.

\ciceroSection{18} ``Likewise,'' he says, ``and in the same way,'' in another clause, ``as in the elections of the Pontifex Maximus.'' He did not even see this, that our ancestors were so popular that, where it was not religiously lawful for a man to be created by the people on account of the religious nature of the rites, in this case nevertheless on account of the magnitude of the priesthood they wished the people to be petitioned. And the same Gnaeus Domitius, tribune of the plebs, a most distinguished man, carried the same measure as to the other priesthoods, that, since the people could not from religious scruple appoint priests, a smaller part of the people should be summoned, and whoever was made by that part should be co-opted by the college.

\ciceroSection{19} See what the difference is between Gnaeus Domitius, tribune of the plebs, a most noble man, and Publius Rullus, who has tested, I suppose, your patience by saying that he is a noble. Domitius, in a thing that could not be done by the people on account of the religious rites, contrived by reason that the part of the people, as far as he could, as far as it was lawful, as far as it was permitted, should have its share. This man, in a thing that always was the people's own, that no one ever lessened, no one ever altered --- that those who were going to assign lands to the people should first receive the favour from the people before they gave it --- has tried to wrench from your hands and snatch away wholly. The other gave to the people in some manner what could in no way be given; this one tries by some method to wrench away what can in no way be taken.

\ciceroSection{20} One will ask, in such injury and such impudence, what end he had in view. He had no lack of design: faith toward the Roman commons, citizens, and fairness toward you and toward your liberty, he sorely lacked. For he orders him who carried the law to hold the elections for the creation of the ten commissioners. I will say it more plainly: Rullus, a man neither greedy nor grasping, orders Rullus to hold the elections. I do not yet object; I see others have done so. The thing no one ever did --- to deal with a smaller part of the people --- see whither it tends. He will hold the elections, he will choose to declare elected those for whom royal power is sought by this law. The whole people he himself does not trust, nor do those authors of these designs think they can rightly be entrusted with it.

\ciceroSection{21} The same Rullus will draw lots for the tribes. The lucky man will draw out the tribes he wants. And those whom nine tribes drawn out by the same Rullus shall have made decemvirs, these we shall have, as I shall presently show, as masters in all things. And these, that they may seem grateful and mindful of the favour, will confess that they owe something to certain men of those nine tribes; but for the remaining twenty-six tribes there will be nothing they will not think they can rightly refuse. What men, then, does he wish made decemvirs? Himself first. By what right? For there are old laws --- and not consular laws, if you think this counts for anything, but tribunician laws, very pleasing and dear to you and your ancestors: there is the Lex Licinia and another, the Lex Aebutia, which exempt not only the man who has carried a measure for some commission and power, but also his colleagues, kinsmen and connections, from having that power or commission entrusted to them.

\ciceroSection{22} For if you are looking out for the people, remove yourself from the suspicion of any advantage to yourself, give proof that you are looking only for the people's profit and gain, allow the power to come to others, the gratitude for the favour to yourself. For this thing is scarcely fitting for a free people, scarcely fitting for your spirit and magnificence. Who carried the law? Rullus. Who barred the larger part of the people from the votes? Rullus. Who presided at the elections, who summoned the tribes he wished, with no overseer of the lots, who created the decemvirs he wished? The same Rullus. Whom did he proclaim chief? Rullus. By Hercules, I scarcely think he could prove this acceptable to his slaves --- not to mention to you, masters of all nations. The best laws, then, will be done away by this law without exception; the same man will seek for himself by his own law a commission; the same man with the larger part of the people robbed of votes will hold elections; whomever he wishes, including himself, he will declare elected; and, of course, he will not refuse his colleagues, the co-signers of the agrarian law, to whom the first place in the index and in the heading of the law has been granted to him; the rest of the fruits of all things which lie set in the hope of this law are kept by joint guarantee and on equal share.

\ciceroSection{23} But see the man's diligence --- if you suppose either that Rullus thought of this, or that it could come into Rullus's mind. Those who contrived these things saw that, if you had been given the power of choosing from the entire people, in any matter where faith, integrity, virtue, authority were sought, you would without hesitation entrust it to Gnaeus Pompey as your chief. For the one man whom out of all you would pick to set in command of all the world's wars on land and sea, certainly in the appointing of decemvirs, whether faith were considered or honour, both could most safely be entrusted to him and most justly adorn him.

\ciceroSection{24} And so this law makes no exception of youth, of any legal impediment, of office, of any magistrate hindered by other duties or laws; even an accused man is not excepted from being eligible to be made decemvir; --- Gnaeus Pompey alone is excepted, lest with Publius Rullus --- I say nothing of the others --- he be made decemvir. For he orders him to declare himself in person, which has never been required in any other law, not even in those magistracies which have a fixed order of standing, lest, if the law were accepted, you should add him as a colleague to Rullus, a guardian and avenger of his greed. Here, since I see that you are stirred by the man's worth and by the law's insult, I shall renew what I said at the start --- a kingship is being prepared, your liberty by this law is being utterly taken away.

\ciceroSection{25} Or did you think otherwise? when a few men cast eyes of greed upon all that is yours, did you not think they would in particular act so as to drive Gnaeus Pompey from every guardianship of your liberty, from every power, commission, and patronage of your interests? They saw and they see that, if through your imprudence and my negligence you accept a law you do not know, it will come about that, when later you have learned the snares set for you, when you make the decemvirs, then you must think Pompey's protection should be opposed to all the law's flaws and crimes. And will it not be a small proof to you that domination and power over all things is being sought by certain men, when you see that the man whom they perceive to be the guardian of your liberty is being made to have no part in the dignity?

\ciceroSection{26} Now learn what power is given to the decemvirs, and how great. First, by a curiate law he equips them. This is unheard of and quite new --- that magistracy be given by a curiate law to those who have not first been given it by any election. He orders that law to be carried by that praetor of the Roman people who has been made first. But how? That those should hold the decemvirate whom the plebs has appointed. He has forgotten that none are appointed by the plebs. And does this man bind the world with new laws who does not remember in his third clause what he has written in his second? Here it is plain what right you have received from your ancestors, what is left to you by this tribune of the plebs. Our ancestors wished you to give your judgment twice on each magistracy. For when the centuriate law was carried for censors, when the curiate for the other patrician magistrates, then a second judgment was given on the same men, so that there might be the power of recalling the favour, if the people came to repent.

\ciceroSection{27} Now, citizens, you keep those first elections, the centuriate and the tribute; the curiate are kept only for the sake of the auspices. But this tribune of the plebs, because he saw that no one could hold a power without the order of the people or the plebs, has confirmed it by curiate elections which you do not enter, has done away with the tribute elections which were yours. So when our ancestors wished you to judge concerning each magistracy by two elections, this popular man has not left even one election in the people's power.

\ciceroSection{28} But see the man's scrupulousness and diligence. He saw and clearly perceived that without a curiate law the decemvirs could not have power, since they were appointed by nine tribes; he orders a curiate law to be carried about them; he commands the praetor. How absurdly he does it does not concern me. He orders the praetor who has been made first to carry the curiate law; if he cannot carry it, the one who has been made last --- so that he seems either to have toyed in such great matters or assuredly to have had I-know-not-what in view. But this thing, which is either so perverse as to be ridiculous, or so malicious as to be obscure, let us pass over; let us return to the man's scruples. He sees that without a curiate law nothing can be done by the decemvirs.

\ciceroSection{29} What then, if it shall not be carried? Mark the cunning. ``Then,'' he says, ``those decemvirs shall be of the same right as those who have been created under the best law.'' If this can be done in this state which far surpasses other states in the right of liberty --- that anyone obtain command or power without any election --- what is the use of ordering, in the third clause, a curiate law to be carried, when in the fourth you allow that without a curiate law they shall have the same right as if they had been created by the people under the best law? Kings are being set up, citizens, not decemvirs. And so they take their birth from these beginnings and foundations such that, not only when they shall have begun to discharge the magistracy, but even when they are being set up, every right, power, and liberty of yours is taken away.

\ciceroSection{30} See now how diligently he keeps the right of the tribunician power. When consuls have carried curiate laws, the tribunes of the plebs have often vetoed them --- nor do we complain that this is the tribunes' power; only, if anyone has used that power rashly, we count him a madman --- but here this tribune of the plebs takes away from the curiate law that the praetor will carry the power of veto. And while this is to be condemned in him because by means of a tribune of the plebs the tribunician power is diminished, it is also to be laughed at, because while a consul, if he has not got a curiate law, may not touch military business, yet this man, in the very case where he forbids veto, sets up a power that, even if a veto is interposed, will still be the same as if the law had been passed --- so that I cannot see why he either forbids veto or thinks anyone will veto, since the veto will only signify the vetoer's folly and not impede the matter.

\ciceroSection{31} Let the decemvirs, then, be set up neither by true elections --- that is, by the votes of the people --- nor by those shadowy elections, kept up for show and for the maintenance of antiquity, with the thirty lictors for the auspices' sake. See now what greater ornaments he confers upon those who have received from you no power, than upon all of us upon whom you have bestowed the most ample powers. He orders the decemvirs to have the auspices for planting colonies, and the keepers of the sacred chickens, ``with the same right,'' he says, ``as the triumvirs had under the Lex Sempronia.'' Do you dare even, Rullus, to make mention of the Lex Sempronia? does that very law not remind you that those triumvirs were created by the votes of the thirty-five tribes? And, since you are very far removed from Tiberius Gracchus's fairness and modesty, do you suppose that what was done on a most unlike footing ought to be of the same right?

\ciceroSection{32} Besides, he gives a power praetorian in name, in fact regal; he limits it to five years; he makes it perpetual, for he secures it with such resources and forces that against the unwilling it cannot in any way be taken away. Then he equips them with attendants, scribes, copyists, heralds, architects, and besides with mules, tents, saddle-blankets, baggage. He drains the expense from the treasury, replenishes it from the allies. He sets up two hundred surveyors of equestrian rank, twenty bodyguards apiece, the same as ministers and satellites of the power. So far you have only the form, citizens, only the appearance of tyrants; you see the trappings of power, not yet the power itself. For someone may perhaps say: ``what hurt do these things do me, the scribe, the lictor, the herald, the keeper of the sacred chickens?'' All these things are of such a kind, citizens, that the man who holds them without your votes seems either an unbearable king or a private madman.

\ciceroSection{33} Look at how great a power is granted; you will say it is not the insanity of private men but the unrestraint of kings. First, infinite power is given of raising innumerable money by the alienation, not the enjoyment, of your revenues. Then, over the world and over all nations, judicial cognizance is given without a council, punishment without appeal, censure without recourse to help.

\ciceroSection{34} For five years they will be able to pass judgment on consuls or on the very tribunes of the plebs; meanwhile no one will pass judgment on them. They will be allowed to seek magistracies, not allowed to plead a case in court. They will be able to buy lands from whomever they wish and at whatever price they wish, however dearly. They are allowed to plant new colonies, renew old ones, fill all Italy with their colonies. The supreme power is given of going through every province, of fining free peoples in their lands, of selling kingdoms. When they wish, they may be at Rome; when it suits them, anywhere they please they may roam with the supreme imperium and judgment of all things. Meanwhile let them dissolve public courts, withdraw from councils whomever they wish, decide each one alone of the greatest matters, hand it over to a quaestor, send a surveyor; let what the surveyor reports to the one man by whom he was sent be ratified.

\ciceroSection{35} A word fails me, citizens, when I call this power kingly --- but in fact a certain greater word is needed. For there was never any kingdom that did not contain itself, if not within some right, then at least within fixed regions. But this power is boundless: by the law's permission it embraces all the kingdoms, and your empire which extends most widely, and those things which are partly free from you, partly even unknown to you. They are given, then, first, the power to sell all those things which it was decreed by the Senate should be sold in the consulship of Marcus Tullius and Gnaeus Cornelius and afterwards.

\ciceroSection{36} Why is this so dim and unseeing? Could not all those things which the Senate decreed sold be set down by name in the law? There are two causes for this obscurity, citizens: one of shame --- if there can be any shame in such marked impudence --- the other of crime. For he does not dare to call by name those things the Senate decreed sold; for there are public places of the city, there are little chapels which since the restoration of the tribunician power no one has touched, which our ancestors wished to be partly the city's ornaments, partly refuges in danger. These the decemvirs will sell by tribunician law. To this will be added Mount Gaurus, the willow-groves at Minturnae, even that very saleable Herculanean way of many delights and great cash, and many other things which the Senate decreed sold from the narrowness of the treasury, but which the consuls did not sell because of the odium.

\ciceroSection{37} But these things perhaps from a sense of shame are passed over in silence in the law. The thing more to be believed and more to be feared is this: that to the decemviral audacity a great power is given of corrupting the public records and of forging decrees of the Senate which were never made --- since out of the number of those who were consuls in those years many are dead. Unless perchance it is unfair for us to suspect any boldness on the part of those whose greed seems to find the world too narrow.

\ciceroSection{38} You have one kind of selling, which I see seems great to you. But turn your minds to what follows; you will see this made into a kind of step and approach to the rest. ``Whatever lands, whatever places, buildings.'' What is there besides? Many things in slaves, in cattle, in gold, in silver, in ivory, in clothing, in furniture, in other things. Why should I say more? Did he think it would be invidious if he had named everything? He did not fear odium. What then? He thought it long, and he was afraid he might leave something out; he wrote in ``or anything else'' --- by which brevity you see that nothing is excepted. So whatever there is outside Italy that has been made public property of the Roman people in the consulship of Lucius Sulla and Quintus Pompeius or after, that he orders the decemvirs to sell.

\ciceroSection{39} By this clause, citizens, I say that all races, nations, provinces, and kingdoms have been handed over and bestowed on the decemvirs' authority, judgment, and power. First I ask this: is there any place anywhere that the decemvirs cannot say has been made public property of the Roman people? For when the same man who has declared so can also pass judgment, what is there that he cannot say, when it is allowed to him to pass the judgment as well? It will suit them to say that Pergamum, Smyrna, Tralles, Ephesus, Miletus, Cyzicum, in short the whole of Asia which has been recovered since the consulship of Lucius Sulla and Quintus Pompeius, has been made the Roman people's;

\ciceroSection{40} will speech be wanting for the discussion of that matter, or, when the same man both discusses and judges, can he not be brought to give a false judgment? or, if he refuses to condemn Asia, will he not appraise the terror of condemnation and his threats at as much as he wishes? What of this, that one cannot in any way speak against, what has now been settled and judged by us, since we have already entered upon the inheritance --- the kingdom of Bithynia, which certainly has been made public property of the Roman people: is there any reason why all the lands, cities, lakes, harbours, the whole of Bithynia in short, will not be sold by the decemvirs? What of Mytilene, which certainly has become yours, citizens, by the law of war and the right of victory, a city famous in the first rank for its nature and site and the layout of its buildings and its beauty, with pleasing and fertile lands? They are included, of course, under the same clause.

\ciceroSection{41} What of Alexandria and all Egypt --- how secretly it lies hid, how it is concealed, how stealthily the whole of it is handed over to the decemvirs! For who of you does not know that this kingdom is said to have been made the Roman people's by the will of King Alexa? Here I, the consul of the Roman people, not only pass no judgment but do not even bring out an opinion. For this great matter seems to me of weight, not only for deciding but even for speaking. I see those who affirm that the will was made; I am aware that the authority of the Senate exists for the entry on the inheritance --- when on Alexa's death we sent ambassadors to Tyre to recover the money he had deposited there.

\ciceroSection{42} I remember that Lucius Philippus often confirmed this in the Senate. Almost all agree that the man who at the present time holds that kingdom is neither of royal birth nor of royal mind. It is said on the other side that there is no will, that the Roman people ought not to seem grasping after every kingdom, that our men will migrate to those parts because of the goodness of the land and the abundance of all things.

\ciceroSection{43} On so great a matter Publius Rullus with the rest of his decemviral colleagues will pass judgment --- and which way will he judge? for either way is so great that in no way can it be conceded or borne. He will wish to be popular; he will adjudge it to the Roman people. So the same man will sell, by his own law, Alexandria, will sell Egypt, will be found judge, arbiter, master --- in short, king --- of a most wealthy and beautiful land. He will not, you say, take so much for himself; he will not grasp it; he will judge that Alexandria belongs to the king, and will judge it away from the Roman people.

\ciceroSection{44} First, why should the decemvirs judge concerning the inheritance of the Roman people, when you have wished the centumviri to judge concerning private inheritances? Then, who will plead the cause of the Roman people? Where will that case be tried? Who are these decemvirs whom we are to see adjudge the kingdom of Alexandria gratis to Ptolemy? But if Alexandria was being sought, why did they not run the same races at this time as they did in the consulship of Lucius Cotta and Lucius Torquatus? Why did they not, openly as before, why not in the same fashion as then, head straight and openly for that region? Or did they who could not hold the kingdom by the etesian winds and a straight course now think they would reach Alexandria by blind darkness and obscurity?

\ciceroSection{45} And consider this in your minds together, citizens. Foreign nations can scarcely bear our ambassadors, men of slight authority, who go on free embassies for the sake of private business. For the very name of empire is heavy, and is dreaded even in a slight person, because it is your name, not their own, that they make use of when they have set out from here. What do you suppose --- when these decemvirs with command, with fasces, with that picked youth of surveyors will roam through the whole world --- with what spirit, what fear, what danger the wretched nations will be?

\ciceroSection{46} There is terror in command; they will endure it. There is expense at his coming; they will bear it. Some service will be commanded; they will not refuse. But this --- how great it is, citizens! --- when that decemvir who has come to some city either expected as a guest or suddenly as a master shall declare that very place to which he has come, that very seat of hospitality into which he has been led, public property of the Roman people! What a calamity to the people if he shall say so --- what a profit to himself if he shall deny it! And the same men who grasp at these things sometimes complain that all lands and all seas have been entrusted to Gnaeus Pompey. As though it were the same thing to entrust many things and to bestow all things, to set a man over labour and trouble or over plunder and gain, to send him to liberate allies or to oppress them! Finally, if some honour is singular, does it make no difference whether the Roman people confers it on whom it will, or whether it is impudently filched from the Roman people by deceit of law?

\ciceroSection{47} You have understood how many things and how great the decemvirs will sell by the law's permission. It is not enough. When they have filled themselves with the blood of allies, of foreign nations, of kings, let them cut the sinews of the Roman people, lay hands on your revenues, break in upon the treasury. For there follows the clause in which he does not even allow --- if perchance money should be lacking, though so much can be received from the previous clauses that it ought not to be lacking --- but plainly, as if this thing would be your salvation, so he forces and orders the decemvirs to sell your revenues by name, citizens.

\ciceroSection{48} Read me out in order, from the law's text, the auction of the Roman people; which by Hercules I think will be a sorrowful and bitter announcement for this very herald.--- ``The auction.'' --- As in his own affairs, so in the commonwealth, he is a luxurious wastrel, who would sell the woods before the vineyards! You have surveyed Italy; pass on to Sicily. --- There is nothing in this province either in the towns or in the lands which our ancestors left to us as our own that he does not order to be sold.

\ciceroSection{49} What our ancestors left you, won by recent victory in the cities and territories of allies as both a bond of peace and a monument of war --- shall you, having received it from them, sell it on this man's motion? Here for a moment I seem to myself to stir your minds, citizens, while I lay open to you the snares which they have, in their own opinion, set deep against Gnaeus Pompey's dignity. And forgive me, I beg, if I appeal to such a man more than once. You, citizens, two years ago, when I was praetor, in this very place imposed on me this role: to defend, with you and so far as I could, the dignity of him in his absence. So far I have done what I could, neither led on by his friendship nor by hope of office and the most ample dignity, which I, though with his goodwill yet in his absence, attained through you.

\ciceroSection{50} For which reason, since I see that this whole law is fitted as a sort of engine for the overthrow of his power, both I shall resist the men's designs and I shall surely accomplish this --- that what I see, you all may not only see but hold fast. He orders to be sold what belonged to the Attalenses, the Phaselites, the Olympeni, and the Aperan, Oroandic, and Gedusan lands. These were made yours by the command and victory of Publius Servilius, that most illustrious man. He adds royal lands of Bithynia which the publicans now enjoy; then the Attalic lands in the Chersonese, in Macedonia, which were of King Philip or Perseus, which were also let out by the censors and yield a most certain revenue.

\ciceroSection{51} He enters into the same auction the rich and fertile Corinthian lands, and the Cyrenean lands which were of Apion, and lands in Spain near New Carthage, and in Africa he sells old Carthage itself --- which Publius Africanus, of course, did not consecrate by the Senate's decision out of religious regard for those seats and their antiquity, nor that the place itself might display the traces of the calamity of those who had contended with this city for empire; he was not, perhaps, as diligent as Rullus, or perhaps could not find a buyer for that place. But among these lands captured in old wars by the courage of the greatest commanders, he adds the royal lands of Mithridates, those that were in Paphlagonia, in Pontus, in Cappadocia, for the decemvirs to sell. Is it really so?

\ciceroSection{52} Without laws being given, without hearing the words of the commander, with the war not yet finished --- when King Mithridates, having lost his army, expelled from his kingdom, is nevertheless even now contriving something in the farthest lands and is defended from the unconquered hand of Gnaeus Pompey by the Maeotis and those marshes and the narrowness of passes and the height of mountains, when the commander is engaged in war, while in those places even now the very name of war remains --- shall the lands over which by ancestral custom all judgment and power should still be in Gnaeus Pompey's hands be sold by the decemvirs?

\ciceroSection{53} And, I suppose, Publius Rullus --- for he bears himself as if he were already designate decemvir --- will himself set out for the auction! He will, of course, before he comes into Pontus, send a letter to Gnaeus Pompey, of which I think the draft has already been composed by these men: ``Publius Servilius Rullus, tribune of the plebs, decemvir, sends greetings to Gnaeus Pompey, son of Gnaeus.'' I do not believe he will write ``Magnus,'' for he does not seem likely to grant in a word what he tries to take away by law. ``Please be so good as to be present for me at Sinope, and bring up your forces, while I sell by my law those lands which you took by your labour.'' Or will he not bring in Pompey? In his own province will he sell the spoils of the commander? Set before your eyes Rullus in Pontus between our camp and the enemy's, with the spear set up, holding an auction with his pretty surveyors.

\ciceroSection{54} Nor in this is the only insult, which is grave and signal and novel --- that anything won in war, before laws are given, while the commander is still administering the war, should not only be sold but auctioned off. Men aim certainly at more than the insult; they hope that, if to Pompey's enemies it be granted, with command, with judgment of all things, with infinite power, with countless money, not only to roam in those parts but even to come up to his own army, some snares may be set for him, something may be detracted from his army, his forces, his glory. They think that, if the army has any hope in Gnaeus Pompey, either of lands or of other rewards, it will not have it when it sees the power of all those things transferred to the decemvirs.

\ciceroSection{55} I do not greatly mind that they are foolish enough to hope these things, shameless enough to attempt them; my complaint is that they have despised me so far as to scheme such monstrosities in my consulship above all. And in all these lands and buildings to be sold, the decemvirs are permitted to sell ``in whatever places they wish.'' O confounded reasoning, O unbridled lust, O dissolute and lost designs! Revenues may not be let out anywhere except in this city, from this place, in this throng of yours. Shall our own things, made over and alienated from us forever, be allowed to be sold in the darkness of Paphlagonia and the solitude of Cappadocia?

\ciceroSection{56} When Lucius Sulla was selling at his ill-omened auction the goods of citizens not condemned, and was saying that he was selling his own booty, yet he sold from this place, nor did he dare flee the sight of those whose eyes he was offending. Shall the decemvirs sell your revenues not only without you as witnesses, citizens, but without even a public herald as witness? Then follows ``all lands outside Italy'' from no fixed time --- not as before, ``from the consulship of Sulla and Pompey.'' To the decemvirs is given the cognizance whether a thing is private or public; and on that land a vast assessment is laid.

\ciceroSection{57} How great a judgment this is, how unbearable, how kingly --- who can fail to see, that they should be able in whatever places they please, with no debate, no council, to make private property public and to free public property? In this clause is excepted the Recentoric land in Sicily, which I myself heartily rejoice to see excepted, citizens, both for the connections of those involved and for the fairness of the case. But what shamelessness this is! Those who hold the Recentoric land defend themselves by length of possession, not by right; by the Senate's mercy, not by the land's status. For they confess that the land is public; they say that they ought not to be moved from their possessions, their most ancient seats, and their household gods. And, if the Recentoric land is private, why is it excepted? but if public, what is this fairness, that the rest, even if private, should be allowed to be judged public, while this one should be excepted by name which confesses itself to be public? So the land of those who had some power with Rullus is excepted, and all the rest of the lands wherever they are, without any selection, without the Roman people's notice, without the Senate's judgment, will be assigned to the decemvirs?

\ciceroSection{58} And there is also another lucrative exception in the previous clause by which all things are sold, which will protect those lands about which provision is made by treaty. He has heard this matter often agitated in the Senate (not by me, but by others), and sometimes from this place: that King Hiempsal possesses lands on the seacoast which Publius Africanus adjudicated to the Roman people; yet that protection was given him afterwards by Gaius Cotta the consul by treaty. Because you have not ordered this treaty, Hiempsal fears that it may not be sufficiently firm and ratified. Whatever this treaty's status, your judgment is taken away, the whole treaty is accepted, approved. That he diminishes the decemviral auction I praise; that he makes provision for a friendly king I do not blame; that it is not done gratis I declare.

\ciceroSection{59} For there flutters before their eyes Juba, the king's son, a young man no less well moneyed than well haired. There hardly seems any place left to hold so great a heap of money; he increases, he adds, he piles up. ``Gold, silver from booty, from spoils, from the victor's crown, that has come to anyone and not been brought into the public treasury or spent on a monument'' --- this he orders to be declared before the decemvirs and reported to them. By this clause you see also the inquiry concerning most distinguished men who have waged the wars of the Roman people, and the judgment about money still remaining, transferred to the decemvirs. Of these things there will be no judgment of theirs as to how much each man's spoils were, what was paid in, what is left over; but for the future also this law is set up for your commanders --- that whoever has departed from a province must declare before these same decemvirs how much booty, how much in spoils, how much in coronary gold he has.

\ciceroSection{60} Yet here this excellent man excepts the man he loves, Gnaeus Pompey. Whence this love so unforeseen and so sudden? He who is excluded from the honour of the decemvirate almost by name, whose judgment and the giving of laws and the cognizance over the lands captured by his own courage are taken away --- to whose province, no, into whose very camp the decemvirs are sent with command, with infinite money, with the greatest power and judgment over all things --- from whom alone the right of a commander, which has always been preserved to all commanders, is torn away --- is he the only one excepted from being ordered to declare his spoils? Does this clause seem to give honour to the man, or to seek odium against him?

\ciceroSection{61} This Gnaeus Pompey remits to Rullus; he uses nothing of that benefit of the law, of the decemviral kindness. For if it is fair that commanders should not bring their booty and spoils to the monuments of the immortal gods or to the city's ornaments, but should bring them back to the decemvirs as to masters, Pompey claims for himself nothing special, nothing; he wishes to live in the common right, on the same footing as the rest. But if it is unfair, citizens, if disgraceful, if intolerable that these decemvirs should be set up as toll-collectors over the moneys of all, who are to shake down not only kings and men of foreign nations but even your own commanders --- then it does not seem to me they except Pompey for honour's sake, but because they fear that he might not be able to bear the same insult as the rest.

\ciceroSection{62} But Pompey is so disposed that whatever pleases you he will think he must bear; what you cannot bear, he will surely accomplish that you not be compelled to bear it longer against your will. Yet for all that, he provides that, if any money after our consulship be received from new revenues, the decemvirs may use it. New revenues, moreover, he sees will be those Pompey will have annexed. So with the spoils remitted, he thinks he ought to enjoy the revenues won by Pompey's courage. Let the money be raised, citizens, for the decemvirs as great as is in the world; let nothing be left out; let all cities, lands, kingdoms, in short even your revenues be sold; let the spoils of your commanders be added to the heap; you see how immense and monstrous wealth would be sought by the decemvirs in such auctions, in so many judgments, in so boundless a power over all things.

\ciceroSection{63} Now learn other immense and intolerable gains, that you may understand that this popular name of an agrarian law has been sought to gratify certain men's importunate greed. With this money he orders lands to be bought to which you may be transplanted. I am not accustomed, citizens, to address men with rougher language unless I have been provoked. I would it could be done that those who hope to be decemvirs could be named by me without insult; you would already see what kind of men those would be to whom you would entrust the power of selling and buying everything.

\ciceroSection{64} But what I do not yet decide I must say, you nevertheless can think over in your minds. This one thing certainly I seem to myself to be able to say with the greatest truth: at the time when this commonwealth had its Luscini, Calatini, Acidini --- men adorned not only with the people's honours and great deeds but with patience of poverty --- and at the time when there were the Catos, Phili, Laelii, whose wisdom and temperance in public and private, in forensic and domestic affairs you had observed --- yet a thing of this kind was committed to no one: that the same man should judge and sell, and do this for five years over the whole world, and the same man alienate the lands and revenues of the Roman people; and, when he had himself with no witness made the total sum of so much money out of his own will, then at last buy from whomever he wished what seemed good.

\ciceroSection{65} Entrust now, citizens, all these things to those men whom you suspect of sniffing for this decemvirate. You will find that part of them are men to whom for keeping nothing seems enough, part to whom for spending nothing seems enough. Here I do not even argue what is most evident, citizens --- that this custom was not left us by our ancestors, that lands should be bought from private owners for the public planting of the commons; in all laws, the commons have been planted on public lands. I confess that I expected something of this kind from this rough and savage tribune of the plebs; but this most lucrative and most disgraceful traffic of buying and selling I have always thought foreign to a tribunician proceeding, foreign to the dignity of the Roman people.

\ciceroSection{66} He orders lands to be bought. First I ask --- what lands and in what places? I do not want the Roman commons to hang in suspense and uncertainty by an obscure hope and a blind expectation. There is the Alban land, Setine, Privernate, Fundan, Vescine, Falernian, Liternine, Cuman, Nucerian. Good. By another gate the Capenate, Faliscan, Sabine, Reatine; by another the Venafran, Allifan, Trebulan. You have so much money that all these and others like them you might not only buy but heap up. Why do you not name them and define them, so that the Roman commons might at least be able to deliberate what is in their interest, what suits them, how much you ought to be entrusted with in buying and selling things? ``I define,'' he says, ``Italy.'' A region certain enough! For how very little does it matter whether you are planted at the foot of Mount Massicus or in the forest of Sila?

\ciceroSection{67} Come then, you do not define a place; what about the nature of the land? ``Indeed,'' he says, ``land that can be ploughed or tilled.'' ``Land that can be ploughed,'' he says, ``or tilled,'' not what has been ploughed or tilled. Is this a law, or the bill for a Veratian auction? on which is said to have been written: ``two hundred acres on which an olive grove can be made, three hundred acres where vineyards can be set out''? Will you buy this with that countless money --- what can be ploughed or tilled? What soil is so thin and lean that it cannot be scratched by a plough, or what is so rough a stony place that the farmers' tilling does not work upon it? ``The reason,'' he says, ``I cannot name lands is that I shall touch none against the owner's will.'' This, citizens, is much more lucrative than if he were taking from the unwilling: for a profit will be reckoned out of your money, and then at last land will be bought when it is to the buyer's and the seller's profit at once.

\ciceroSection{68} But see the force of the agrarian law. Not even those who hold public lands will give up their possession unless they are bought out on the best terms and with the largest money. The reckoning is reversed. Before, when an agrarian law was mentioned by a tribune of the plebs, those who held public lands or invidious possessions immediately took fright; this law enriches them in fortune, frees them from odium. For how many do you think there are, citizens, who cannot defend the breadth of their possessions, cannot bear the odium of Sullan lands, who wish to sell, can find no buyer, would even, by some method, want to lose those lands? They who but a moment ago day and night dreaded the very name of tribune, feared your power, were terrified by the mention of an agrarian law, will now even be sought out and entreated to deliver up to the decemvirs lands partly public, partly full of odium and full of danger, at whatever price they themselves wish. And this song this tribune of the plebs sings within himself, not to you.

\ciceroSection{69} He has a father-in-law, an excellent man, who in those darknesses of the commonwealth occupied as much land as he wanted. He wishes to come to his rescue, now sinking and overwhelmed, weighed down by the burden of Sullan estates, by his own law, that he may be allowed to lay down his odium and stow away his cash. And do you not hesitate to sell your revenues, won at the cost of so much of your ancestors' blood and sweat, in order to enrich Sullan possessors and free them from danger?

\ciceroSection{70} For this decemviral purchase has two kinds of land in view, citizens. The owners flee the one because of odium, the other because of waste. Sullan land joined together over the widest tracts by certain men has so much odium attached to it that it cannot bear even a single hiss of a true and bold tribune of the plebs. All that land, however cheaply it is bought up, will be charged to us at a vast price. The other kind of land is uncultivated for sterility, deserted and unoccupied for unhealthiness, and will be bought from those who see that, if they do not sell, they will have to abandon them. And, surely, this is what was said by this tribune of the plebs in the Senate: that the city commons have too much power in the commonwealth; that they ought to be drained off (so he used the word, as if speaking of some bilge and not of the order of the best citizens).

\ciceroSection{71} But you, citizens, if you will hear me, hold fast that possession of yours --- of favour, of liberty, of votes, of dignity, of city, of Forum, of games, of festal days, and of every other advantage --- unless you prefer to leave these things and the light of the commonwealth, and be settled in the dryness of Sipontum or in the territories of the Salpini, full of pestilence, with Rullus as your leader. Or let him say what lands he is going to buy; let him show both what he is going to give and to whom. Or that, when he has sold all the cities, lands, revenues, kingdoms, he then buy some sand or marshes, can you allow it, I ask? Yet this is excellent --- that by this law all things are sold first, the moneys raised and heaped up, before one clod is bought. Then he orders buying, forbids it from the unwilling.

\ciceroSection{72} I ask: if there are no willing sellers, what will become of the money? The law forbids it to be paid back, prohibits its being demanded. Therefore the decemvirs will hold all the money, no land will be bought for you; with the revenues alienated, the allies vexed, kings and all peoples emptied out, they will have the moneys and you will not have the lands. ``Easily,'' he says, ``they will be brought by the size of the money to want to sell.'' This, then, is a law by which we sell our own at what price we can, buy others' at what the owners wish.

\ciceroSection{73} And on these lands which shall have been bought by this law he orders colonies to be planted by these same decemvirs. What? is every place such that it makes no difference to the commonwealth whether a colony be planted there or not, or is there a place that requires a colony, or is there one that plainly refuses one? In which matter, as in the other parts of the commonwealth, it is worth while to recall the diligence of our ancestors, who placed colonies in suitable places against the suspicion of danger, so that they might seem not towns of Italy but bulwarks of the empire. These men will plant colonies on the lands they have bought; even if it does not profit the commonwealth?

\ciceroSection{74} ``And in whatever places besides shall seem good.'' What reason is there, then, why they may not plant a colony on the Janiculum and place their own garrison upon our head and neck? You, may you not define how many colonies, in what places, with what number of colonists, you wish to plant; may you seize a place that you have judged suitable for your own violence, fill it with the number, secure it with the garrison you wish, and with the revenues of the Roman people and all his resources may you constrain the very Roman people, oppress him, reduce him under that decemviral authority and power?

\ciceroSection{75} And as for his planning to garrison and occupy all Italy with his garrisons --- learn this, I beg, citizens. He permits the decemvirs to plant in all the municipalities, in all the colonies of all Italy, what colonists they wish, and orders lands to be given to those colonists. Are not, plainly, greater forces sought than your liberty can bear, and greater garrisons; is not, plainly, a kingship being set up; is not, plainly, your liberty being taken away? For when the same men shall hold all the money, the largest multitude, the same men shall garrison all Italy with their forces, the same men shall hold your liberty enclosed by their garrisons and colonies, what hope, what means of recovering your liberty will be left?

\ciceroSection{76} ``But the Campanian land,'' it is said, ``the most beautiful in the world, will be divided by this law, and a colony will be planted at Capua, a most ample and most ornate city.'' What can we say to this? Of your interest first I shall speak, citizens; then to the magnitude and dignity I shall return, so that, if anyone is delighted with the goodness of the land or the town, he may have nothing to expect, and if anyone is moved by the unworthiness of the matter, he may resist this pretended bounty. And first I shall speak of the town, in case anyone perchance is more pleased by Capua than by Rome. He orders five thousand colonists to be enrolled at Capua; in this number each man takes five hundred for himself.

\ciceroSection{77} I beg you, do not console yourselves; consider truly and diligently. Do you suppose there will be a place in this number for you or men like you, of unspoiled, quiet, peaceable habits? If there is for all of you or the greater part, although your honour bids me keep watch day and night and look upon every part of the commonwealth with intent eyes, yet for a little while, if it suits your interest, I shall close them. But if a place and a city is being sought for five thousand men picked for violence, crime, and slaughter --- a city which can make and equip war --- will you nevertheless allow forces to be strengthened against you in your name, garrisons armed, cities, lands, resources prepared?

\ciceroSection{78} For the Campanian land which they parade before you, they themselves have lusted after; they will plant their own men, in whose name they themselves may hold and enjoy it; they will buy in besides; they will join up those ten-acre plots. For if they shall say it is not allowed by law, neither is it allowed by the Cornelian law; yet we see, to set the distant aside, the Praeneste land held by a few. Nor do I see anything wanting to these men's moneys except estates of this kind, by whose support they could sustain the size of their households and the expenses of their Cumaean and Puteolan estates. But if he is looking out for your interest, let him come and dispute with me face to face about the division of the Campanian land.

\ciceroSection{79} I asked of him on the Kalends of January to what men and how he was going to distribute that land. He answered that he would begin with the Romilian tribe. First, what arrogance and insult is this, that one part of the people should be lopped off, the order of the tribes neglected, lands given to the country tribes who have lands first, before they are given to the city tribes, to whom this hope and pleasantness of land is held out? Or, if he denies this was said by him and intends to satisfy you all, let him bring forward; let him distribute it into ten-acre plots; let him propose your names from the Suburan tribe right down to the Arnian. If you understand that not only ten-acre plots cannot be given to you, but that so great a number of men cannot even be packed into the Campanian land, will you still allow the commonwealth to be vexed, the majesty of the Roman people to be despised, yourselves to be tricked any longer by a tribune of the plebs?

\ciceroSection{80} But if that land could come to you, would you not still prefer that it remain in your patrimony? Will you allow the one most beautiful estate of the Roman people, the head of your money, the ornament of peace, the support of war, the foundation of revenues, the granary of the legions, the consolation of the corn-supply, to perish? Or have you forgotten how, in the Italic War, when the other revenues were lost, with how great armies you fed yourselves on the produce of the Campanian land? Or do you not know that all those other magnificent revenues of the Roman people often hang on the slightest tilt of fortune, on the inclination of the times? What use will the harbours of Asia, the coast of Syria, all the overseas revenues, be to us at the slightest suspicion of pirates or of enemies?

\ciceroSection{81} But this revenue of the Campanian land, citizens, is of such a kind that, since it is at home and is sheltered by all the garrisons of the towns, it is wont to be neither threatened by wars nor variable in produce nor calamitous in climate or place. Our ancestors not only did not diminish what they had taken from the Campanians but even bought up what those held from whom it could not justly be taken. For which reason neither the two Gracchi, who thought most about the interests of the Roman commons, nor Lucius Sulla, who lavished everything on whom he wished without any scruple, dared touch the Campanian land. Rullus has appeared who would dispossess the commonwealth of that holding from which neither the Gracchi's generosity nor Sulla's domination had cast it down. The land which now you, passing by, call yours, and which travellers passing along the road, foreigners, hear is yours --- when it has been divided, will neither be yours nor be called yours. And what kind of men will hold it?

\ciceroSection{82} First, fierce, ready for violence, prepared for sedition, who, as soon as the decemvirs sound the signal, may be armed against citizens and ready for slaughter; then you will see the whole Campanian land carried over to a few men flowing with means and forces. To you meanwhile, who received from your ancestors those most beautiful seats of revenue won by arms, not a clod from your fathers' and grandfathers' possessions will be left. But how great will the difference be between your diligence and that of private men! What? When Publius Lentulus, who was princeps senatus, was sent by our ancestors into those parts to buy up with public money the private lands which encroached on the Campanian public ground, he is said to have reported that for no money could he buy a certain man's farm, and that the man who refused to sell said that he could not be brought to sell because, having many farms, he had never received bad news from that one alone.

\ciceroSection{83} Is that so, indeed? This case moved a private man; will it not move the Roman people not to deliver the Campanian land gratis to private men at Rullus's request? But the same Roman people can say of this revenue what that man is said to have said of his farm. Asia for many years gave you no produce in the Mithridatic War; the revenue of the Spanish provinces was none in Sertorian times; in the slave war Manius Aquilius even gave the Sicilian states a loan of grain. But from this revenue no bad news has ever been heard. The other revenues are stricken by the difficulties of war; with this revenue even the difficulties of war are sustained.

\ciceroSection{84} Then in this assignment of lands not even that can be said which is said in others, that lands ought not to be deserted by the commons and free men's tilling. For so I say: if the Campanian land is divided, the commons are turned out and driven from the lands, not settled and placed there. For all the Campanian land is tilled and held by the commons, and by a most upright and most modest commons --- a kind of men of the best habits, the best ploughmen and soldiers, who are by this people-loving tribune of the plebs being utterly cast out. And those wretched men, born and brought up on those lands, exercised in subduing the clods, will not have where suddenly to take themselves; while to these strong, sturdy, and bold satellites of the decemvirs the whole holding of the Campanian land will be handed over. And, as you now boast of your ancestors --- ``this land our ancestors left us'' --- so your descendants will boast of you: ``this land our fathers, having received from theirs, lost.''

\ciceroSection{85} I think this: if even now the Field of Mars were divided and to each of you a stand of two feet were assigned, you would still rather enjoy the whole at large than enjoy a small part as your own. For which reason, even if something out of this land were going to come to each of you which is now paraded before you but is being prepared for others, you would still hold it with more honour as a body than as individuals. Now indeed, since nothing of it concerns you, but it is being procured for others and snatched from you, will you not most fiercely resist this law for your lands as you would an armed enemy? He adds the Stellate plain to the Campanian land and assigns out of it twelve acres to each man. As though the Campanian land differed by a little from the Stellate.

\ciceroSection{86} But a multitude is sought, citizens, with which to fill all those towns. For I said before that it is permitted by the law that they may seize what municipalities they wish, what old colonies they wish, with their colonists. They will fill up the municipality of Cales, they will overpower Teanum, they will bind down Atella, Cumae, Naples, Pompeii, Nuceria with their garrisons; Puteoli indeed, which now is in its own power, uses its own right and liberty, they will wholly seize with a new people and foreign forces. Then that standard of a Campanian colony, dreadful to this empire, will be carried into Capua by the decemvirs; then against this Rome, the common country of us all, that other Rome will be sought.

\ciceroSection{87} Into that town men are wickedly attempting to transfer your commonwealth, in which town our ancestors wished there to be no commonwealth at all --- who reckoned that only three cities of all on earth, Carthage, Corinth, and Capua, could sustain the weight and the name of empire. Carthage was destroyed because, both with its supplies of men and by its very nature and place, girt with harbours, armed with walls, it seemed about to make sallies out of Africa and to threaten the two most fruitful islands of the Roman people. Of Corinth scarcely a trace was left. For it was placed in the narrows and at the throat of Greece, so that it held the keys of the country and almost joined two seas which differ most in navigation, separated by a slender division. The cities which were far from the eye of empire they not only crushed but, lest they should ever be able to revive and rise and pull themselves up again, took away utterly, as I said.

\ciceroSection{88} About Capua there was much and long deliberation; the public records exist, citizens; there are several decrees of the Senate. Wise men decided that, if they took the land from the Campanians, took away the magistrates, the senate, the public council out of that city, left no image of a commonwealth, there would be nothing for us to fear in Capua. And so you will find this written in the old records: that there be a city which could supply the things by which the Campanian land was tilled, that there be a place for the bringing in and storing of crops, that ploughmen worn out by the cultivation of fields might use the dwelling-places of a city; for that reason those buildings were not destroyed.

\ciceroSection{89} See what an interval lies between the counsels of our ancestors and the madness of these men. They wished Capua to be a refuge for ploughmen, a market for the country folk, a storehouse and granary of the Campanian land; these men, with the ploughmen driven out, your produce poured out and scattered, set up the same Capua as the seat of a new commonwealth, prepare a mass against the old commonwealth. But if our ancestors had thought that anyone in so brilliant an empire and so distinguished a discipline of the Roman people would be like Marcus Brutus or Publius Rullus --- for these are the two we have so far seen who would wish to transfer this commonwealth wholesale to Capua --- they would surely not have left the name of that city.

\ciceroSection{90} But they thought that at Corinth and Carthage, even if they took away the senate and magistrates and the land from the citizens, there would still not be lacking those who would restore those things and change everything before we could hear of it; but here, before the eyes of the Senate and the Roman people, nothing could arise that could not be quenched and suppressed before it was openly born and grown. Nor in this matter were men of divine mind and counsel mistaken. For after the consulship of Quintus Fulvius and Quintus Fabius --- in whose consulship Capua was overcome and taken --- nothing has been, I do not say done, but even thought of in that city against this commonwealth. Many wars were waged afterwards with kings: with Philip, Antiochus, Perseus, the Pseudo-Philip, Aristonicus, Mithridates, and the rest. Many heavy wars besides: the Third Punic, the Corinthian, the Numantine. Many domestic seditions in this commonwealth which I pass over; wars with allies, the Fregellan, the Marsic. In all these domestic and foreign wars Capua not only did us no harm but offered herself most opportunely to us both for equipping war and for fitting out armies and for receiving them within her roofs and walls.

\ciceroSection{91} There were no men in the city who could throw the commonwealth into confusion by evil meetings, by turbulent decrees of the senate, by unjust commands, and seek some pretext for revolution. For neither was there power for anyone to hold a meeting, nor to take public counsel; they were not carried away by lust for glory, because where there is no public honour, there cannot be lust for glory; they were not divided by contention or by canvassing. For there was nothing left over which they should contend, nothing which they should seek against another, nothing in which they should disagree. So that Campanian arrogance and intolerable ferocity our ancestors by reason and counsel led down into a most idle and slothful repose. Thus they both escaped the infamy of cruelty, in not destroying the most beautiful city of Italy, and provided much for the future, in that, with all the city's sinews cut, they left the city itself loose and weakened.

\ciceroSection{92} These counsels of our ancestors seemed to Marcus Brutus, as I said before, and to Publius Rullus matter for blame; nor do those omens and auspices of Marcus Brutus deter you, Publius Rullus, from a like madness. For both the man himself who planted the colony, and those who took magistracies at Capua under his appointment, and those who touched any part of that planting, of that honour, of that office, all suffered the most bitter punishments of the impious. And since I have made mention of Marcus Brutus and that time, I shall mention what I myself saw, when I had come to Capua after the colony had already been planted under Lucius Considius and Sextus Saltius, ``the praetors,'' as they themselves called them --- so that you may understand how much arrogance the place itself brings, which in the few days after the colony had been planted there could be perceived and understood.

\ciceroSection{93} For first, as I said, when in other colonies the magistrates are called duumviri, these wished to be called praetors. To whom, if their first year had brought this lust, do you not think within a few years they would have been reaching for the name of consuls? Then there walked before them lictors not with rods, but, as the city praetors here have, with two fasces apiece. Greater victims were set up in the Forum, which by these praetors from a tribunal --- as by us consuls --- with the council's assent, were approved and sacrificed to the herald and the flute-player. Then patres conscripti were summoned. The look on Considius's face was scarcely bearable to behold. The man whom we used to see at Rome ``shrivelled with squat leanness,'' contemned, abject --- when we saw him at Capua with Campanian disdain and a regal spirit, I seemed to myself to see again the Blossii and Vibellii.

\ciceroSection{94} And now what dread was there of those tunic-wearers! In the Albana and Seplasia what a rushing about of those asking what the praetor had decreed, where he was dining, where he had given notice! As for us who had come there from Rome, we were no longer called guests but foreigners and newcomers.

\ciceroSection{95} Those who looked ahead to these things --- I mean our ancestors, citizens --- do you not think they ought to be venerated and worshipped by us as among the immortal gods? For what did they see? This which now I beg you to perceive and learn. Manners are not so engendered in men by the stock of their birth and seed as by the things which from nature herself are supplied to us for the way of life by which we are nourished and live. The Carthaginians were fraudulent and lying, not by race but by nature of place, since on account of their harbours they were summoned to a zeal for deceit by their zeal for gain in many and various conversations of merchants and incomers. The Ligurians were hard and rustic; the land itself, yielding nothing except what was sought by much cultivation and great labour, taught them. The Campanians were always proud through the goodness of their lands and the greatness of their crops, the salubrity, the layout, the beauty of their city. From this abundance and overflow of all things first that arrogance was born by which from our ancestors they demanded a second consul from Capua, and then that luxury which conquered Hannibal --- still then unconquered in arms --- by pleasure.

\ciceroSection{96} Hither when these decemvirs shall have brought five thousand colonists by Rullus's law, set up a hundred decurions, ten augurs, six pontiffs, what spirits do you suppose, what onsets, what fierceness those men will have? Rome, set on hills and in valleys, raised aloft and suspended in upper rooms, with no excellent streets and very narrow lanes, they will laugh at and despise in comparison with their Capua, spread out in a most level place and most splendidly situated. As for fields, the Vatican and the Pupinia, with their rich and fertile plains, they will not, of course, think comparable. The abundance of neighbouring towns there they will compare with this through laughter and jest; Veii, Fidenae, Collatia, even by Hercules Lanuvium, Aricia, Tusculum --- these they will measure against Cales, Teanum, Naples, Puteoli, Cumae, Pompeii, Nuceria.

\ciceroSection{97} Lifted up and inflated by these things, perhaps not at once, but certainly, if they shall have taken on a little antiquity and strength, they will not be contained; they will go forward, they will carry everything along with them. A single private man, unless endowed with great wisdom, is hardly contained within the lattices and limits of duty in great fortunes and resources --- not to mention these colonists at Capua, sought out and chosen by Rullus and men like Rullus, set in a home of arrogance and luxurious seats, who will not at once be searching for some piece of crime or scandal, indeed even more than those old genuine Campanians: for those, born and brought up in long-standing fortune, were yet corrupted by an excessive abundance of all things; these, brought over from the highest indigence into the same abundance, will be moved not only by abundance but also by their unfamiliarity with it.

\ciceroSection{98} These traces of Marcus Brutus's crime, Publius Rullus, you preferred to follow rather than the monuments of our ancestors' wisdom; these things you with those advisers of yours have devised: that you might plunder our old revenues, search out new ones, set up against this city a new city to challenge in dignity; that you might bring under your right, your jurisdiction, your power, the cities, the nations, the provinces, the free peoples, the kings, in short the whole world; that, when you had drained all the money out of the treasury, gathered it from the revenues, exacted it from all the kings, peoples, and from our commanders, all moneys should still be paid to you at your nod; that the same men in part should from Sullan possessors charge to the Roman people, at whatever price you wished, lands hated for odium and in part deserted and pestilential, bought from your relatives and from yourselves; that you might fill all the municipalities and colonies of Italy with new colonists; that in whatever places it should seem good to you and as many as it should seem good to you, you might place colonies;

\ciceroSection{99} that you might encircle the whole commonwealth with your soldiers, your cities, your garrisons and hold it down oppressed; that you might be able to deprive the very Gnaeus Pompey, by whose protection the commonwealth has very often been guarded against the bitterest enemies and the most wicked citizens, of his victorious army and of the sight of these citizens; that nothing precious in gold and silver, nothing reckoned in number and slaves, nothing broken open by violence and the hand could be that you did not hold pressed down and snatched away; that you might meanwhile flit through nations, through all kingdoms with the highest command, with infinite judgment, with all your money; that you might come into Gnaeus Pompey's camp and the camp itself, if it suited you, you might sell; that meanwhile you might seek the remaining magistracies, freed from all laws, without fear of trials, without danger; that no one might be able to bring you before the Roman people, no one to lead you forth, no one to compel you into the Senate, no consul to coerce you, no tribune of the plebs to detain you.

\ciceroSection{100} That you have lusted after these things, in your folly and intemperance, I do not wonder; that you hoped to attain them in my consulship, I do greatly wonder. For when there ought to be in all consuls a heavy care and diligence in guarding the commonwealth, then there ought to be most of all in those who have been made consuls not in the cradle but on the campaign field. No ancestors of mine have given the Roman people surety on my behalf; credit was given me. From me you ought to demand what I owe; you ought to call upon me myself. As, when I was a candidate, no sponsors of my line commended me to you, so, if I have done anything wrong, there are no portrait-busts to plead for me before you. Therefore, only let life suffice me, which I shall try to defend from these men's crime and snares: I promise you this, citizens, in good faith --- you have entrusted the commonwealth to a vigilant man, not a timid one, to a diligent man, not a slothful one.

\ciceroSection{101} Am I the kind of consul to fear an assembly, to dread a tribune of the plebs, to make uproar often and without cause, to fear that I must dwell in prison if a tribune of the plebs has ordered me led there? When I am armed with your arms, equipped with the highest command, authority, and most ample insignia, I do not shrink from coming forward to this place, I am able, citizens, with you as my sponsors, to resist the wickedness of one man, nor do I fear that the commonwealth, fortified by such great defences, can be conquered or oppressed by these men. If I had been afraid before, yet at this assembly, before this people, I should certainly not be afraid. For who has ever recommended an agrarian law in so favourable an assembly as I have spoken against it? if to ``speak against'' it is rather than to disturb and overthrow it.

\ciceroSection{102} From which, citizens, it can be perceived that nothing is so popular as that which I, your popular consul, bring you for this year: peace, tranquillity, quiet. The things you feared when we were consuls-elect have been provided against by my counsel and reason that they should not be able to happen. Not only will you, who always wished to be in quiet, be in quiet, but I will render most quiet and most easy even those to whom quiet is hateful. For honours, powers, riches are wont to be procured for those out of tumult and from the dissensions of citizens; you, whose favour rests in your votes, your liberty in your laws, your right in trials and in the equity of magistrates, your private property in peace, ought by every method to keep your quiet.

\ciceroSection{103} For if those who through laziness live in idleness yet take pleasure in their base sluggishness from idleness itself, how fortunate you will be, citizens, if in this state which you have, sought not by your indolence but won by your virtue, you keep your quiet! I, from the concord which I have established with my colleague --- against the will of those men who said that we as consuls were and should be enemies --- have provided for all things, looked out for the corn-supply, recalled credit, given notice to the tribunes of the plebs not to stir up anything turbulent in my consulship. The highest and surest defence of our common fortunes, citizens, is this: that, just as in this day's most crowded assembly you have shown yourselves to me for your safety's sake, so in the rest of the times you show yourselves to the commonwealth. I promise, undertake, pledge this to you, and confirm it: that I shall accomplish that those who have envied my office shall yet acknowledge that you all in choosing your consul saw most.
