% In Catilinam I (in-catilinam-1) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} How long at last, Catiline, will you go on abusing our patience? How long will that fury of yours mock us? To what limit will your unbridled audacity hurl itself? Has the night-watch on the Palatine, has the guard about the city, has the fear of the people, has the gathering together of every loyal man, has this most fortified place for holding the Senate, has the look and countenance of these men here moved you not at all? Do you not feel that your designs lie open? Do you not see that your conspiracy is held already bound by the knowledge of all these men? What you did the night before last, what you did the night before that, where you were, whom you summoned, what plan you formed --- which one of us do you suppose does not know it?

\ciceroSection{2} O the times, O the manners! The Senate understands these things, the consul sees them; this man yet lives. Lives? No more --- he comes even into the Senate, takes part in the public council, marks down and designates with his eyes for slaughter every one of us. We meanwhile, brave men, think we are doing enough for the commonwealth if we avoid this man's fury and his weapons. To death you ought, Catiline, long since to have been led on the consul's order, to be turned upon yourself the doom which you have been so long contriving against us all.

\ciceroSection{3} A man of the highest rank, Publius Scipio, the pontifex maximus, as a private citizen put to death Tiberius Gracchus, who was only modestly shaking the foundation of the commonwealth: shall we as consuls bear with Catiline, who lusts to lay the world waste with slaughter and burnings? --- I pass over those antique examples, that Gaius Servilius Ahala killed with his own hand Spurius Maelius for striving after revolution. There was, there once was in this commonwealth a courage by which brave men coerced a citizen who threatened ruin with sharper punishments than they used against the bitterest enemy. We have a decree of the Senate against you, Catiline, vehement and weighty; the commonwealth does not lack counsel, nor does the authority of this order; we, we, I say openly, the consuls, are wanting.

\ciceroSection{4} The Senate decreed once that the consul Lucius Opimius should see to it that the commonwealth take no harm: no night intervened. Gaius Gracchus, of a most distinguished father, grandfather, and ancestors, was killed on certain suspicions of sedition. The consular Marcus Fulvius was killed with his children. By a like decree of the Senate the commonwealth was committed to the consuls Gaius Marius and Lucius Valerius: did death and the punishment of the commonwealth wait one day longer for Lucius Saturninus the tribune of the plebs and Gaius Servilius the praetor? But we have for twenty days now suffered the edge of the authority of these men to grow blunt. We have, you see, a decree of the Senate of that kind, but shut up in the tablets, as though hidden in its sheath --- by which decree of the Senate it would have been fitting that you, Catiline, were straightway killed. You live; and you live not to lay aside your audacity but to confirm it. I desire, senators, to be merciful; I desire in such great perils to the commonwealth not to seem irresolute; but already I now condemn myself for sluggishness and worthlessness.

\ciceroSection{5} A camp has been pitched in Italy in the gorges of Etruria against the Roman people; the number of the enemy grows day by day. Yet the commander of that camp and the leader of the enemy you see within the walls and even in the Senate, contriving daily some plague within the commonwealth itself. If now I should order you, Catiline, to be arrested, to be killed, I shall, I suppose, have to fear lest all loyal men say that this thing has been done by me too late, rather than that anyone should call it done too cruelly. But what ought long since to have been done, I am for a definite reason not yet brought to do. Then at last shall you be killed, when no man so wicked, so abandoned, so like yourself can be found, who would not confess that it was done by right.

\ciceroSection{6} As long as anyone shall be left who would dare defend you, you shall live --- and live as you live now, beset by my many and firm guards lest you be able to stir against the commonwealth. The eyes and ears of many men also will, without your perceiving it, as they have done up to now, watch and keep guard over you. For what is there, Catiline, that you yet wait for, if neither night with its darkness can hide your wicked meetings nor a private house contain within its walls the voices of your conspiracy, if all things break out into the light? Change now that mind of yours, believe me, forget slaughter and burning. You are caught on every side; brighter than daylight to us are all your designs --- which you may now go over with me.

\ciceroSection{7} Do you remember that on the twelfth day before the Kalends of November (21 October) I said in the Senate that on a certain day --- which day was to be the sixth day before the Kalends of November (27 October) --- Gaius Manlius, the satellite and minister of your audacity, would be in arms? Did the matter so great, so atrocious, so incredible escape me, Catiline --- or, what is far more wonderful, the day? I said the same in the Senate, that you had appointed the slaughter of the chief men for the fifth day before the Kalends of November (28 October), at the time when many leaders of the state fled from Rome, not so much to save themselves as to check your designs. Can you deny that on that very day, hemmed in by my guards and my diligence, you were unable to move against the commonwealth, when at the others' departure you said you were yet content with our slaughter --- of those of us who had stayed?

\ciceroSection{8} What of this --- when you were confident you would seize Praeneste on the very Kalends of November by a night-attack, did you not perceive that that colony was fortified by my order with my guards, garrisons, and watches? You do nothing, you set on foot nothing, you think nothing that I do not hear, that I do not even see and clearly perceive. Go over with me at last that night which has just gone by; you will already understand that I keep watch much more keenly for the safety of the commonwealth than you do for its ruin. I say that on the night before last you came among the scythe-makers --- I shall not act obscurely --- to the house of Marcus Laeca; that there came together to the same place several partners in the same madness and crime. Do you dare deny it? Why are you silent? I will convict you if you deny. For I see there are here in the Senate certain men who were with you.

\ciceroSection{9} O immortal gods! where in the world are we? what commonwealth do we have? in what city do we live? Here, here in our number, senators, in this most sacred and weighty council of the world, are men who plot the destruction of us all, the ruin of this city and even of the world. These I, the consul, see; I ask their opinion on the commonwealth; and those whom it would have been fitting to butcher with the sword, I do not yet wound with my voice! You were, then, at Laeca's that night, Catiline; you distributed parts of Italy; you settled where each man should set out for; you chose whom to leave at Rome, whom to take out with you; you marked off parts of the city for fires; you confirmed that you yourself would now be leaving; you said that there was still some little delay for you, because I yet lived. Two Roman knights were found to release you of that care, and undertook on that very night a little before dawn to kill me in my bed.

\ciceroSection{10} All these things I learned of when your gathering was scarcely yet broken up; my house I fortified and made firm with greater guards; I shut out those whom you had sent to me to greet me in the morning, when those very men had come whose coming to me at that hour I had foretold to many men of the highest rank. Since these things are so, Catiline, go on whither you have begun: come out at last from the city; the gates lie open; set off. That Manlian camp of yours has wanted you, the commander, too long. Take with you also all your men --- if not all, then as many as possible. Cleanse the city. You will free me from a great fear, provided that a wall stand between you and me. To dwell with us any longer you cannot; I shall not bear it, I shall not endure it, I shall not allow it.

\ciceroSection{11} Great thanks must be given to the immortal gods and to this very Jupiter Stator, the most ancient guardian of this city, that we have so often now escaped a plague so foul, so horrible, so hostile to the commonwealth. The supreme safety of the commonwealth must not be put at risk too often through one man. As long as you, Catiline, lay snares for me when I was consul-elect, I defended myself not by a public guard but by private diligence. When at the recent consular elections you wished to kill me, the consul, and your competitors on the Field of Mars, I crushed your wicked attempts by the protection and forces of friends, with no public alarm raised. In short, every time you aimed at me, I stood against you on my own --- though I saw that my ruin was bound up with great calamity to the commonwealth.

\ciceroSection{12} Now you openly aim at the entire commonwealth: the temples of the immortal gods, the houses of the city, the lives of every citizen, all Italy you call to ruin and devastation. Therefore, since what is the first thing, and what is proper to this empire and to the discipline of our ancestors, I do not yet dare to do, I shall do what is gentler in severity, more useful for the common safety. For if I shall order you killed, the rest of the band of conspirators will remain in the commonwealth; but if you, as I have for some time been urging you, shall depart, the great and pernicious bilge of your companions will be drained out of the city. What now, Catiline?

\ciceroSection{13} Do you hesitate to do at my command what already of your own accord you were doing? The consul orders an enemy to leave the city. Do you ask me whether into exile? I do not order it; but, if you ask my advice, I urge it. For what is there, Catiline, that can now please you in this city? in which there is no man, outside of that conspiracy of lost men, who does not fear you, no one who does not hate you. What brand of domestic disgrace is not seared upon your life? What scandal in private affairs does not cling to your reputation? What lust ever was absent from your eyes, what crime from your hands, what vileness from your whole body? To what young boy, whom you had snared in the lures of corruption, have you not held out either the dagger for audacity or the torch for lust?

\ciceroSection{14} What of this? When recently by the death of your former wife you had emptied a place for a new marriage, did you not heap upon this crime another, an incredible one besides? which I pass over and gladly suffer to be passed in silence, lest the savagery of so great a deed be seen either to have come into being in this city or to have gone unavenged. I pass over the ruins of your fortunes, all of which you will feel hanging over you on the next Ides; I come to those things which concern not the private disgrace of your vices, not your domestic difficulty and shame, but the supreme commonwealth and the life and safety of us all.

\ciceroSection{15} Can this light, Catiline, or the breath of this air be pleasant to you, when you know that there is no one of these men who does not know that on the day before the Kalends of January (31 December) in the consulship of Lepidus and Tullus you stood in the Comitium with a weapon, that you had prepared a band for the killing of the consuls and the leading men of the state, that not any change of mind on your part nor any fear of yours stood in the way of your crime and madness, but the Fortune of the Roman people? And let me leave those things aside --- for they are neither obscure nor are there few things committed since --- how often as consul-designate, how often when I was consul, have you tried to kill me! How many of your aims, so directed that they could not be parried, have I escaped by some slight twist and, as the saying goes, with the body alone! You do nothing, you accomplish nothing, and yet you do not cease to try and to wish.

\ciceroSection{16} How often has that dagger of yours been wrested from your hands, how often has it slipped by some accident and slid away! By what rites it was initiated by you and consecrated I do not know, that you think you must plant it in a consul's body. Now, indeed, what is that life of yours? For I shall now so speak with you that I may seem moved not by hatred --- by which I ought to be --- but by mercy, of which none is owed to you. You came a little while ago into the Senate. Who out of so great a throng, who out of so many of your friends and intimates, greeted you? If this has happened within the memory of mankind to no one, do you yet wait for the affront of speech, when by the most weighty judgment of silence you stand crushed? What of this --- that at your coming those benches were vacated, that all the consulars who very often had been marked down by you for slaughter, as soon as you sat down, left bare and empty that part of the benches: with what spirit, then, do you think this is to be borne?

\ciceroSection{17} If, by Hercules, my slaves feared me in the way that all your fellow-citizens fear you, I should think I had to leave my house. Do you not think you must leave the city? And if I saw myself unjustly so heavily suspected and resented by my fellow-citizens, I would rather forgo the sight of citizens than be looked upon by the hostile eyes of all. Do you, when by your conscience of your crimes you recognize that the just hatred of all is owed you and has long been owed, hesitate to avoid the sight and presence of those whose hearts and feelings you wound? If your parents feared and hated you, and you could in no way appease them, you would, I think, withdraw from their sight to some distance. Now your country, the common parent of us all, hates you and fears you, and judges that you have been thinking now this long while of nothing but her own murder: will you respect neither her authority nor follow her judgment nor dread her violence?

\ciceroSection{18} She so deals with you, Catiline, and after a manner silently speaks: ``No crime now for some years has come into being except through you, no scandal without you. To you alone the slayings of many citizens, to you alone the harrying and plundering of allies, has been free of penalty and licensed; you have had the strength not only to neglect laws and inquiries but to overthrow and break them through. Those earlier things, although they were not bearable, I yet bore as I could; but now that I am wholly in fear because of you alone, that whatever has rattled, Catiline is dreaded, that no plan seems to be able to be set on foot against me which is not akin to your crime --- this is not bearable. Therefore depart and take from me this fear; if it is true, that I be not crushed; if false, that at last sometime I may cease to fear.''

\ciceroSection{19} If your country, as I said, should so speak with you, ought she not to obtain her wish, even if she could not use force? What of this --- that you put yourself into custody, that for the sake of avoiding suspicion you said you wished to dwell with Manius Lepidus? When you were not received by him, you even dared to come to me and asked that I keep you under guard at my house. When you carried away from me too the same answer --- that I could in no way safely be under the same roof with you, since I was in great danger because we were contained within the same walls --- you went to Quintus Metellus the praetor. When you were rebuffed by him, you migrated to your comrade, that excellent man Marcus Metellus, whom you supposed, of course, would be most diligent in guarding you, most sagacious in suspecting, most stout in punishing! But how far must he seem he ought to be from prison and from chains, who has already judged himself worthy of a guard?

\ciceroSection{20} Since these things are so, Catiline, do you hesitate, if you cannot die with an even mind, to go away into some other lands, and to commit that life of yours, snatched from many just and owed punishments, to flight and solitude? ``Refer the matter,'' you say, ``to the Senate''; this you demand, and you say that, if this order shall decree that you should go into exile, you will obey. I shall not refer it, for that is foreign to my habits, and yet I shall make you understand what these men feel about you. Leave the city, Catiline, free the commonwealth from fear; into exile, if this is the word you wait for, set forth. What is it? Do you take any notice, do you mark the silence of these men? They suffer it, they keep silent. Why do you wait for the authority of those who speak, when you perceive the will of those who hold their peace?

\ciceroSection{21} But if I had said this same thing to that excellent young man Publius Sestius, or to that bravest man Marcus Marcellus, the Senate by its best right would already have laid violence and hands upon me, the consul, in this very temple. But about you, Catiline, when they keep quiet they approve, when they suffer it they decree, when they are silent they cry out --- and not these only, whose authority you find so dear (their lives so very cheap), but also those Roman knights, most honourable and most upright men, and the rest of those bravest citizens who stand around the Senate, whose throng you might have seen, whose zeal you might have perceived, whose voices a little while ago you might have heard. Whose hands and weapons I scarcely have for some time held back from you --- those same I shall easily bring to escort you, leaving these things you have long sought to ravage, all the way to the gates.

\ciceroSection{22} And yet why do I speak? That anything should break you, that you should ever amend yourself, that you should think of any flight, ponder any exile? May the immortal gods grant you that mind! Although I see, if struck with terror by my voice you have brought yourself to go into exile, what a storm of odium hangs over me, if not in the present time --- the memory of your crimes being so fresh --- yet for the future. But it is worth as much, provided that the calamity be your private one and be sundered from the dangers of the commonwealth. But that you should be moved by your own vices, dread the punishments of the laws, yield to the times of the commonwealth --- that is not to be looked for. For you are not such, Catiline, that either modesty has called you back from baseness, or fear from danger, or reason from madness.

\ciceroSection{23} Therefore, as I have often now said, set out; and if you wish to brew odium against me, an enemy, as you call yourself --- straight on into exile. I shall hardly bear men's talk if you do this, I shall hardly sustain the weight of that odium if you have gone into exile by the consul's order. But if you would rather serve my praise and glory, go out with that ungovernable band of criminals, betake yourself to Manlius, stir up lost citizens, sever yourself from the loyal, bring war upon your country, exult in your impious brigandage --- so that you may seem not to have been driven out by me to strangers but to have been invited to your own.

\ciceroSection{24} And yet why do I bid you, when I know already that men have been sent ahead by you to wait armed for you at the Forum Aurelium, when I know that a day has been agreed and fixed with Manlius, when I know that even that silver eagle which I trust will be the ruin and the death of you and all your men --- which had a wicked shrine set up for it in your house --- has been sent ahead? How will you be able to be without it longer, you that were wont to worship it as you set out for slaughter, from whose altars you so often turned that impious right hand to the killing of citizens?

\ciceroSection{25} You will go at long last where that unbridled and frenzied lust of yours has long been carrying you. Nor does this thing bring you sorrow, but a kind of incredible pleasure. To this madness nature bore you, your will trained you, your fortune kept you. Never have you longed not only for peace but even for war except a wicked one. You have got from the lost, those abandoned not only by every fortune but even by every hope, a band of wicked men sweated together.

\ciceroSection{26} Here what gladness will you enjoy, in what joys will you revel, in how great a pleasure will you riot, when in so great a number of your followers you will neither hear any good man nor see one! For the practice of this kind of life those toils of yours that are talked of were rehearsed --- to lie on the ground not only to lay siege to debauchery but also to undertake crime, to keep watch not only when plotting against husbands' sleep but against the goods of the leisured. Now you have where to display that famous patience of yours of hunger, of cold, of want of all things, by which in a short time you will feel yourself worn out.

\ciceroSection{27} So much have I accomplished, when I drove you back from the consulship: that you might rather as an exile try to harass the commonwealth than as consul vex it; and that what was wickedly undertaken by you might be called brigandage rather than war. Now, senators, that I may ward off and entreat from me a just complaint, almost, of my country, take in carefully, I beg you, what I shall say, and store it deep in your minds and hearts. For if my country, which is far dearer to me than my life, if all Italy, if the entire commonwealth should speak with me: ``Marcus Tullius, what are you doing? Will you suffer the man whom you have found to be an enemy, whom you see will be the leader of war, whom you perceive is awaited as commander in the camp of the enemy, the author of crime, the chief of conspiracy, the recruiter of slaves and lost citizens, to leave the city, so that he may seem not let out by you of the city but let in upon the city? Will you not order him to be cast in chains, hauled to death, slain by the highest punishment? What at last hinders you? The custom of our ancestors?

\ciceroSection{28} But very often even private men in this commonwealth have visited death upon ruinous citizens. Or laws which have been carried about the punishment of Roman citizens? But never in this city have those who deserted the commonwealth held the rights of citizens. Or do you fear odium with posterity? You return a fine thanks to the Roman people which has lifted you, a man known by yourself, with no commendation of ancestors, so promptly through every step of the offices to the highest command, if on account of odium or the fear of some danger you neglect the safety of your fellow-citizens.

\ciceroSection{29} But if there is any fear of odium, the odium of severity and courage is not more vehemently to be dreaded than that of sluggishness and worthlessness. Or, when by war Italy shall be wasted, when cities shall be vexed, when houses shall burn --- do you then not think you will burn up in the fire of odium?'' To these most sacred voices of the commonwealth and to the minds of those men who feel the same I shall answer in a few words. I, senators, if I had judged this best to be done --- that Catiline be visited with death --- would not have given that gladiator the use of one hour for living. For if the highest men and most distinguished citizens not only did not stain themselves with the blood of Saturninus and the Gracchi and Flaccus and several previous men but even honoured themselves by it, surely I had nothing to fear lest with this parricide of citizens killed any odium against me should overflow into posterity. But if it should hang most over me, yet I have always been of this mind --- to count odium gained by virtue glory, not odium.

\ciceroSection{30} And yet there are some in this order who either do not see what is hanging over us, or dissemble what they see; who have nourished Catiline's hope by their soft opinions and have strengthened the budding conspiracy by not believing in it; through whose authority many not only wicked but also unschooled men, if I had punished him, would say that the deed was done cruelly and like a king. As it is I see that, if he comes to the Manlian camp, where he is bound, no one will be so foolish as not to see that a conspiracy has been made, no one so wicked as not to confess it. But if this one man only is killed, I see that this plague of the commonwealth can be checked for a little, but not crushed in perpetuity. Whereas if he shall cast himself out and lead his men out with him and shall gather to the same place the rest of the wrecks collected from every quarter, there will be quenched and destroyed not only this so far-grown plague of the commonwealth but also the very stock and seed of all our evils.

\ciceroSection{31} For now this long while, senators, we have been moving in these dangers and snares of the conspiracy; but somehow or other the maturity of all crimes and of the long-standing fury and audacity has burst out into the time of our consulship. Now if from so great a band of brigands this one man shall be removed, we shall perhaps for some short time seem to have been relieved of care and fear; but the danger will remain and will be enclosed deep within the veins and the vitals of the commonwealth. As often men who are sick of a heavy disease, when they are tossed by heat and fever, if they have drunk cold water, seem at first to be relieved, but then are afflicted much more heavily and vehemently --- so this disease which is in the commonwealth, if relieved by this man's punishment, will, with the rest still alive, grow more grievous.

\ciceroSection{32} For which reason let the wicked withdraw, let them sever themselves from the loyal, let them gather into one place, let them at last, as I have often said, be sundered from us by a wall: let them cease to lay snares for the consul in his own house, to stand around the tribunal of the city praetor, to besiege the Senate-house with swords, to prepare fire-darts and torches for the burning of the city; in short, let it be written on every man's brow what he feels about the commonwealth. I promise you this, senators: that there will be in us, the consuls, so great a diligence, in you so great an authority, in the Roman knights so great a courage, in all loyal men so great a unanimity that, by Catiline's setting forth, you will see all things laid open, illumined, suppressed, and avenged.

\ciceroSection{33} With these omens, Catiline, with the supreme safety of the commonwealth, with your own ruin and destruction, and with the doom of those who have joined themselves to you in every crime and parricide, set out for an impious and wicked war. You, Jupiter, who were established by Romulus under the same auspices as this city, whom we truly name the Stayer of this city and this empire, you will keep this man and his confederates from your temples and from the rest, from the houses of the city and from the walls, from the lives and fortunes of all the citizens; and the men, enemies of loyal men, foes of their country, brigands of Italy, joined to one another by a compact of crime and an unspeakable partnership, you will visit, living and dead, with eternal punishments.
