% Pro L. Murena (pro-murena) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} What I prayed of the immortal gods, judges, after the custom and institution of our ancestors, on the day on which I, with the auspices taken, at the centuriate assembly proclaimed Lucius Murena consul --- that the matter might come well and happily to my own loyalty and my magistracy, to the people and the Roman commons --- I pray of the same immortal gods, that for the same man's consulship, together with his safety, the same may be obtained: that your minds and opinions may agree with the wills and votes of the Roman people; and that this matter may bring you and the Roman people peace, tranquillity, quiet, concord. If that solemn prayer of the assembly, consecrated by the consular auspices, has in itself as great a force and religion as the dignity of the commonwealth requires, I have likewise prayed that to those men also to whom this consulship has been given on my motion the matter may turn out auspiciously, happily, prosperously.

\ciceroSection{2} Since these things are so, judges, and since all the power of the immortal gods has either been transferred to you or certainly shared with you, the same man commends to your loyalty as consul whom previously he commended to the immortal gods, so that by the voice of the same man both declared consul and defended he may keep the kindness of the Roman people together with your safety and that of all citizens. And since in this duty the zeal of my defending him, and even the very taking up of the case, has been censured by the prosecutors, before I begin to speak for Lucius Murena I shall say a few words on my own behalf --- not because at this moment the defence of my own duty is more important to me than that of his safety, but so that, when my action has been approved by you, I may with greater authority repel the assaults of the enemies from his honour, his reputation, and all his fortunes.

\ciceroSection{3} And first to Marcus Cato, who directs his life to the certain rule of reason and most diligently weighs the moments of all duties --- I shall answer concerning my duty. Cato denies that it was right that I, both consul and the mover of the law on canvassing, and after a consulship so severely conducted, should touch the case of Lucius Murena. His censure moves me vehemently --- not only that I prove the reason of my deed to you, judges, to whom most of all I am bound, but to Cato himself, that gravest and most upright man. By whom, after all, Marcus Cato, is it juster that the consul be defended than by the consul? Who in the commonwealth can or ought to be more closely joined with me than the man to whom the commonwealth is now being handed over by me to be sustained, the commonwealth I have sustained with my great labours and dangers? But if in those matters which are reclaimed which are mancipia, that man should warrant the danger of the lawsuit who has bound himself by the bond, surely also more rightly in the trial of a consul-designate that consul above all who declared the consul ought to be the sponsor of the kindness of the Roman people and the defender of his danger.

\ciceroSection{4} And if, as in some states is wont to happen, a patron were appointed to this case publicly, that man should be given as defender to one adorned with the highest honour who, endowed with the same honour, would bring no less authority than skill to the speaking. But if those leaving harbour are wont with the highest zeal to give warning to those who are now sailing into harbour out of the deep both of the calculation of storms and of pirates and of places (because nature brings it about that we favour those who enter upon the same dangers we have completed), with what spirit, then, ought I to be --- almost now in sight of land out of a great tossing --- toward this man, on whom I see the greatest storms of the commonwealth must be undergone? Therefore if it is the part of a good consul not only to see what is being done but also to provide what is to be, I shall show in another place how much it concerns the common safety that there be two consuls in the commonwealth on the Kalends of January.

\ciceroSection{5} If it is so, not so much my duty toward the fortunes of a friend as the commonwealth ought to call me, the consul, to the defence of the common safety. For as for my having carried the law on canvassing, I certainly so carried it that I should not abrogate the law I long ago carried for myself --- on defending the dangers of citizens. For if I confessed that bribery had been committed and defended that as rightly done, I should be acting wickedly, even if another had carried the law; but since I am defending that nothing has been done against the law, what is there in the carrying of a law that should hinder my defence?

\ciceroSection{6} He says it is not a part of the same severity to have driven from the city, by words and almost by command, Catiline contriving the ruin of the commonwealth within the walls --- and now to speak for Lucius Murena. But I have always gladly played the parts of mildness and mercy, which nature herself taught me; the role of weight and severity I did not seek, but bore as imposed on me by the commonwealth, just as the dignity of this empire in the supreme danger of the citizens demanded. But if then, when the commonwealth was demanding force and severity, I conquered nature and was as vehement as I was compelled to be (not as I wished), now, when all the cases call me to mercy and to humanity, with how great zeal, then, ought I to serve my own nature and habit? About my duty in defending and the reasoning of your accusation perhaps in another part of the speech we shall have to speak.

\ciceroSection{7} But, judges, no less than Cato's accusation did the complaint of that wisest and most distinguished man, Servius Sulpicius, move me, who said that he bore it most grievously and bitterly that I, forgetful of intimacy and connection, was defending Lucius Murena's case against him. Him, judges, I am eager to satisfy and to call you in as arbiters. For if it is grave to be accused truly in friendship, even if you are accused falsely it is not to be neglected. I, Servius Sulpicius, in your candidacy I both confess that I owed and judge that I performed all zeal and duty toward you on account of our connection. There was nothing lacking to you in your seeking the consulship from me which ought to be sought from a friend or from a man of favour or from a consul. That time has passed; the reckoning has changed. So I judge, so I persuade myself: that I owed you, against the office of Murena, as much as you have dared to demand of me; against his safety, I owe nothing.

\ciceroSection{8} For if I was zealous for you when you were seeking the consulship, I ought not now, when you are seeking Murena himself, to be a helper on the same terms. And not only is this not to be praised, but it cannot even be conceded --- that, when our friends are accusing, we should not even defend the most distant strangers. With Murena, judges, both a great and an old friendship is mine, which in a struggle for life will not be put down by Servius Sulpicius for this reason --- that it was overcome by him in the contest for office. And if this cause did not exist, yet either the man's dignity or the magnitude of the office he attained would have seared upon me the highest infamy of arrogance and cruelty, if I had refused the case of so great a danger to a man so amply adorned both with his own and with the Roman people's distinctions. For neither is it permitted me, nor is it now an open question, that I not lend my labour to the relief of men's dangers. For when so great rewards have been given me for this industry as have been given to none before, I judge thus: that to lay down the labours which you took on in the candidacy, when you have attained the office, is the part of a man both crafty and ungrateful.

\ciceroSection{9} But if I am permitted to leave off, if with you as my authority I may, if no infamy of sloth, no baseness of arrogance, no fault of inhumanity is undertaken, then I gladly leave off. But if flight from labour convicts of laziness, the rejection of suppliants of arrogance, the neglect of friends of wickedness --- it must be that this case is of such a kind that no industrious or merciful or dutiful man can desert it. And of this matter you, Servius, may most easily make a conjecture from your own zeal. For if you think it necessary even to give answers to the adversaries of your friends consulting you on the law, and if you think it base that, with you as advocate, the very man against whom you have come should fall in his case --- do not be so unjust as, when your own springs are open even to your enemies, to think that ours ought to be closed even to our friends.

\ciceroSection{10} For if your intimacy had removed me from this case, and if this same thing had befallen Quintus Hortensius, Marcus Crassus, those most distinguished men, if likewise the rest from whom I see your favour is highly esteemed, the consul-designate would not have a defender in this state in which our ancestors wished a patron to be lacking to no one, however lowly. But I, judges, would have judged myself wicked if I had failed a friend, cruel if a wretched man, arrogant if a consul. Wherefore what must be given to friendship will be given by me amply --- so that I shall deal with you, Servius, just as if my brother, who is dearest to me, were in your place; what must be granted to duty, faith, religion, that I shall so moderate as to remember that I am speaking against a friend's zeal on behalf of a friend's danger.

\ciceroSection{11} I see, judges, that there were three parts to the whole accusation, and that one of them was engaged in censure of his life, the second in a comparison of dignity, the third in the charges of bribery. And of these three parts, that first one which ought to have been the gravest was so weak and slight that some sort of accusatorial law rather than a true means of speaking ill compelled them to say something about the life of Lucius Murena. For Asia was thrown up; which was sought by him not for pleasure and luxury but traversed in military labour. He, if as a young man he had not served under his father as commander, would seem either to have feared the enemy or his father's command, or to have been disowned by his parent. And when sons in the praetexta usually sit on the horses of those triumphing, was he to flee his father's triumph adorned with military gifts, so that for matters jointly carried out he might almost triumph at the same time as his father?

\ciceroSection{12} This man, indeed, judges, both was in Asia and to that bravest man, his parent, was a great help in dangers, a consolation in labours, and a partner in the congratulation of victory. And if Asia carries some suspicion of luxury, what should be praised is not having never seen Asia but having lived continently in Asia. For which reason it was not the name of Asia that ought to have been thrown up against Murena --- from which the praise of his family, the memory of his stock, the honour and glory of his name has been established --- but some scandal and disgrace either undertaken in Asia or carried out from Asia. To have served, indeed, in that war which the Roman people then was waging not only as the greatest but also as the only war was a mark of courage; to have served most willingly with his father as commander was a mark of piety; that the end of his service was his father's victory and triumph was a mark of good fortune. To slander, in these matters, no place is left, because praise has occupied all.

\ciceroSection{13} Cato calls Lucius Murena a dancer. The slander, if truly thrown up, is that of a vehement accuser; if falsely, that of a slanderous railer. Wherefore, since you are of such authority, Marcus Cato, you ought not to snatch up a slander out of the cross-streets or out of some scurrility of buffoons, nor rashly to call a consul of the Roman people a dancer; but to consider with what other vices it must be that the man is afflicted to whom that charge can truly be brought. For scarcely anyone dances sober, unless perhaps he is mad --- neither in solitude nor in a moderate and honourable banquet. Of a timely banquet, of a pleasant place, of many delights, dancing at the end is the companion. Do you snatch up against me what is necessarily the last of all vices, and leave aside those which removed make this vice not even possible? No base banquet, no love affair, no revel, no lust, no extravagance is shown; and, when the things are not found which bear the name of pleasure, even though they are vicious --- in the man in whom you cannot find the very luxury, in him do you think you will find the shadow of luxury?

\ciceroSection{14} So nothing can be said against the life of Lucius Murena, nothing, I say, at all, judges. So is the consul-designate defended by me, that no fraud, no greed, no perfidy, no cruelty, no insolent word in his life is brought forth. Well: the foundations of the defence are laid. For not yet by our praises (which I shall use later) but almost by his enemies' confession we are defending a good and upright man. With this established it is easier for me to come to the comparison of dignity, which was the second part of the accusation.

\ciceroSection{15} I see that the highest dignity is in you, Servius Sulpicius, of birth, of integrity, of industry, and of all the other adornments relying on which it is fitting to undertake the candidacy for the consulship. I see that these are equal in Lucius Murena --- and so equal that he could neither be conquered in dignity by you, nor surpass you in dignity. You despised the birth of Lucius Murena, you exalted your own. In which point, if you take it for granted that no one except a patrician is born of good stock, you make it seem that the plebs must again be called away to the Aventine. But if there are ample and honourable plebeian families, both the great-grandfather of Lucius Murena and his grandfather were praetor, and his father, when he had most amply and most honourably triumphed from his praetorship, left this step to the consulship the easier for him to attain because that office, owed already to the father, was sought by the son.

\ciceroSection{16} But your nobility, Servius Sulpicius, although it is the highest, is yet better known to literary men and historians, but rather more obscure to the people and the voters. For your father was of equestrian rank, your grandfather was distinguished by no celebrated praise. Therefore not from the recent talk of men but from the antiquity of the annals must the memory of your nobility be dug out. Wherefore I am wont always to count you in our number, because by your virtue and industry you have brought it about that, although you were the son of a Roman knight, you were thought worthy of the highest distinction. Nor has it ever seemed to me that there was less virtue in Quintus Pompeius, that new man and bravest hero, than in that most noble man Marcus Aemilius. For it is the part of the same spirit and talent both to hand on to one's posterity, as Pompey did, the magnitude of name which one has not received, and, like Scaurus, to renew by one's own virtue the almost dead memory of one's stock.

\ciceroSection{17} Although I now thought, judges, that by my labour it had been brought about that to many brave men no humbleness of birth should be thrown up --- I, who was concerned not only with recalling the Curii, the Catones, the Pompeii, those bravest men of antiquity, those new men, but these recent ones, the Marii and the Didii and the Caelii. But when after so great an interval I had broken open those barriers of the nobility, so that the approach to the consulship hereafter, as it was with our ancestors, would lie open no more to nobility than to virtue, I did not suppose that, when a consul-designate from an old and distinguished family was being defended by the son of a Roman knight as consul, the prosecutors would speak about the newness of his birth. Indeed it has happened to me to be a candidate with two patricians, one most wicked and most audacious, the other a most modest and excellent man; yet I overcame Catiline by dignity, Galba by favour. If this charge ought to be against a new man, I should certainly not have lacked enemies or detractors.

\ciceroSection{18} Let us therefore omit speaking of birth, in which there is great dignity in each; let us see the rest. ``He stood for the quaestorship together with me, and I was made first.'' One does not have to answer everything. For it does not escape any of you that, when many become equal in dignity, but only one can hold the first place, the order of dignity and of declaration is not the same --- because the declaration has its degrees, but dignity is very often the same in all. But the quaestorship of each was almost of equal weight in the lot. He had under the Lex Titia a province silent and quiet; you had that one which, when the quaestors draw lots, is even greeted with a shout, the Ostian, not so much favourable and bright as bustling and troublesome. The name of each settled in the quaestorship: for the lot gave you no field on which virtue might run forth and be recognized.

\ciceroSection{19} The space of the remaining time is called into the comparison. By each it has been treated in the most unlike fashion. Servius here pursued with us this urban service of giving answers, of writing, of warranting, full of anxiety and bile; he learned the civil law, he kept much watch, he laboured, he was at the service of many, he endured the folly of many, bore their arrogance, drank down their churlishness; he lived at others' arbitration, not his own. Great and pleasing to men is the praise --- that one man should toil in that knowledge which will profit many.

\ciceroSection{20} What was Murena doing meanwhile? He was legate to a most brave and most wise man, the supreme commander Lucius Lucullus; in which legation he led the army, joined standards, joined hand-to-hand, routed great forces of the enemy, took cities partly by force, partly by siege; that Asia full and at the same time delicate he so passed through that in it he left no trace of greed or of luxury. In the greatest war he so engaged that he carried out many great deeds without his commander, but his commander carried out none without him. And although I say these things in the presence of Lucius Lucullus himself, yet that we may not seem to have a licence of invention granted to us by him on account of our danger, all has been attested by public despatches, in which Lucius Lucullus has imparted as much praise as neither an ambitious commander nor a jealous one ought to grant another in the sharing of glory.

\ciceroSection{21} The highest honour and dignity is in each; which I, if Servius will let me, would set in equal and the same praise. But he will not let me; he assails the soldiering, harries this whole legation, thinks the consulship belongs to attendance and to these everyday labours. ``You were,'' he says, ``with the army; you have not touched the Forum for so many years; you have been away so long, and now, when you have come back after a long interval, you contend in dignity with those who have lived in the Forum?'' First, that attendance of ours, Servius, you do not know how much sometimes brings men of distaste, how much of satiety. To me indeed it was vehemently expedient that my favour was set in your eyes; but yet I have overcome the satiety of myself by my great labour, and you likewise perhaps; but yet to neither of us would absence have been a hindrance.

\ciceroSection{22} But to leave this aside and return to the comparison of pursuits and arts, who can doubt that for attaining the consulship the glory of military matters brings far more dignity than that of the civil law? You watch through the night to give answers to your consultants; he that he may come up at the right time with his army where he is heading. The crowing of cocks rouses you, the calls of trumpets him; you set up an action, he draws up a line of battle; you guard against your consultants being caught, he against cities or camps being taken; he holds and knows how the enemy's forces are warded off, you how rain-water is. He has been exercised in extending boundaries; you in regulating them. And surely --- for what I think must be said --- the virtue of military matters surpasses all the rest. This bore the name to the Roman people, this to this city begot eternal glory, this compelled the world to obey this empire; all urban affairs, all these distinguished pursuits of ours and this forensic praise and industry lie hidden under the protection and guard of warlike virtue. As soon as a suspicion of alarm has rattled, our arts straightaway fall silent.

\ciceroSection{23} And since you seem to me to kiss that knowledge of the law as your own little daughter, I shall not allow you to live in so great an error as to think that I-know-not-what which you have learned with such great care to be something distinguished. With other virtues --- of continence, of weight, of justice, of faith, with all the rest --- I have always judged you most worthy of the consulship and of every honour; that, indeed, you have learned the civil law --- I will not say you have wasted the labour, but I shall say this: there is no road in that discipline laid out toward the consulship. For all the arts which conciliate to us the zeal of the Roman people ought to have both an admirable dignity and a most welcome usefulness.

\ciceroSection{24} The highest dignity is in those who are distinguished by military glory: for everything which is contained in the empire and the constitution of the commonwealth is thought to be defended and sustained by them. Most welcome usefulness is also in them, since by their counsel, work, and danger we may use both the commonwealth and our own affairs. Weighty also and full of dignity is the faculty of speaking, which has often weighed in the consul's election; that, by counsel and by oratory, the minds both of the Senate and of the people and of the judges of cases can be governed. A consul is sought who can sometimes by his oratory restrain the frenzy of a tribune, who can sway the inflamed people, who can resist bounty. It is not strange if on this account often even unknown men attain the consulship, especially since this matter wins very great gratitude, the strongest friendships, the highest favour. Of which nothing, Sulpicius, is in your craft.

\ciceroSection{25} First, in so slim a knowledge there cannot be a great dignity. The matters are little; matters indeed for the most part busy in single syllables and apexes of letters. Then, even if anything was held in great account by our ancestors, since your craft has been laid open, despised it lies and hated. There was once a time when it was thought a few men could know whether anything could be done by right; for the calendars were not generally posted, those that gave answers had a great hold on the matter; the day was even sought after the manner of the Chaldeans. There was found a certain scribe, Gnaeus Flavius, who pecked at the eyes of crows and gave the citizens the calendars learned by heart day by day, and snatched the wisdom from these very prudent men. So they, angered (because they feared lest, with the calendars made common and known, actions could be taken without their advice) made up certain forms, that in everything they should be present.

\ciceroSection{26} Although it could most prettily be done thus: ``The Sabine farm is mine.'' ``No, mine''; then judgment --- they did not wish it. ``The farm,'' he says, ``which is in the field which is called Sabine.'' Verbose enough; what next? ``This I declare to be mine by Quirite right.'' What then? ``Therefore there I summon you to a struggle of hands by right.'' What he, from whom the claim was made, should answer to this so loquacious litigant he had not. The same lawyer crosses over in the manner of a Latin flute-player. ``Whence,'' he says, ``you summoned me to a struggle of hands by right, thence there I in turn summon you.'' Lest the praetor meanwhile should think himself fine and blessed and should say something of his own accord, for him too a formula was made up, absurd in the rest but indeed in this: ``With the bystanders of each present, I declare that road; go your road.'' That sage was at hand to teach them how to enter on the road. ``Return your road.'' Under the same guide they returned. These things even then to those bearded ancients seemed ridiculous, I suppose --- that men, when they had stood rightly and in place, were ordered to go away in order that, whence they had gone, they might at once return to the same place. With the same silliness all those things have been painted: ``When I see you in court'' and these: ``Will you say from what cause you have claimed?'' which while they were hidden, were necessarily sought from those who held them; but afterwards, when they were spread abroad and tossed in men's hands and shaken out, they were found most empty of prudence, but most full of fraud and folly.

\ciceroSection{27} For when very many things had been splendidly set up by laws, most of them by the wits of the lawyers were corrupted and depraved. They wished all women, on account of weakness of judgment, that the elders be in the power of guardians; these have invented kinds of guardians who would be contained in the power of the women. They wished sacred rites not to perish; by the wit of these, old men have been found for the sake of going through the form of \textit{coemptio} for the doing away of sacred rites. In every part of the civil law, in short, they have abandoned equity and held the words themselves --- so that, because in someone's books they had found that name set down by way of example, they thought all women who were going through \textit{coemptio} were called ``Gaiae.'' Now this seems wonderful to me --- that so many men, so witty, after so many years even now have not been able to settle whether ``the third day'' or ``the day after to-morrow,'' ``the judge'' or ``the arbiter,'' ``the matter'' or ``the suit,'' ought to be said.

\ciceroSection{28} And so, as I said, in that knowledge there was never consular dignity, since it was wholly composed of made-up and feigned matters; of favour, indeed, far less still. For what lies open to all and is equally ready for me and for my adversary, can in no way be welcome. So you have already lost not only the hope of conferring a kindness but even that --- which once was --- ``May I consult you?'' No one can be thought wise in that prudence which has no force either anywhere outside Rome or at Rome when business is suspended. No one can be held expert because in what all know they cannot in any way disagree among themselves. The matter, moreover, is not thought difficult because it is contained in very few and not at all obscure letters. So if you, vehemently busy man that I am, vex me, in three days I shall declare myself a lawyer. For the things that are done by writing are all written; nor is anything written so narrowly that I cannot add ``Concerning what matter the action is''; the things, indeed, on which one is consulted are answered with the smallest danger. If you have answered what you ought, you will seem to have answered the same as Servius; if otherwise, you will seem also to know and treat a doubtful point of law.

\ciceroSection{29} Wherefore not only must that famous military glory be set before your formulas and actions, but also the practice of speaking far and much surpasses that exercise of yours for honour. So most men, it seems to me, at the start much preferred this; afterwards, when they could not attain it, they slipped especially into that. As they say of Greek craftsmen, that those are pipe-players who could not become harpers, so we see that those who could not turn out to be orators come down to the study of law. Great is the labour of speaking, great the matter, great the dignity, supreme the favour. For from you a kind of healthful service is sought; from those who speak, salvation itself. Then your responses and decrees are often overthrown by speech, and without the defence of oratory cannot be firm. In which, if I had made enough progress, I should speak more sparingly of its praise; now I say nothing of myself, but of those who are or were great in speaking.

\ciceroSection{30} There are, then, two arts which can place men in the most ample step of dignity: one of the commander, the other of the good orator. By this the ornaments of peace are kept; by that the dangers of war are repelled. The other virtues, however, in themselves and by themselves are of much weight: justice, faith, modesty, temperance --- in which all see that you, Servius, excel. But now I dispute about pursuits adapted to honour, not about the inborn virtue of each. All those pursuits are shaken from our hands as soon as some new motion has begun to sound the war-call. For, as a witty poet and a very good author says, ``with the proclamation of battles is driven from the midst'' not only that wordy pretence of prudence of yours but even that mistress of things, ``wisdom; the matter is carried by force, the orator is despised'' --- not only the hateful and talkative one in speaking, but even ``the good; the rough soldier is loved'' --- and your study lies wholly idle. ``Not by right of struggling hand, but rather by iron,'' he says, ``do they reclaim the matter.'' If this is so, the Forum, I think, must yield to the camp, peace to soldiering, the pen to the sword, the shade to the sun; in short, let that be in the state the first thing on account of which the state itself is the chief of all.

\ciceroSection{31} But Cato shows that we are making these things too great with our words, and that we have forgotten that the whole Mithridatic war was waged against little women. I judge it far otherwise, judges, and I shall say a few things on the matter; for the case is not contained in this. For if all the wars which we waged with the Greeks are to be despised, let the triumph of Manius Curius over King Pyrrhus be derided, that of Titus Flamininus over Philip, that of Marcus Fulvius over the Aetolians, that of Lucius Paullus over King Perseus, that of Quintus Metellus over the Pseudo-Philip, that of Lucius Mummius over the Corinthians. But if these wars were the gravest and their victories the most welcome, why are the Asiatic nations and that enemy of yours despised? Yet from the monuments of old matters I see that the Roman people waged perhaps its greatest war with Antiochus; whose victor Lucius Scipio shared equal glory with his brother Publius. The praise which the latter, with Africa crushed, bore before himself by the surname itself, the same did this man take to himself from the name of Asia.

\ciceroSection{32} In which war, indeed, the surpassing virtue of Marcus Cato, your great-grandfather, shone out; in which, since he was such (as I judge for myself) as I see you to be, he would never have set out with Scipio if he had thought one had to make war with little women. Nor indeed, when the Senate had urged Publius Africanus to set out as legate to his brother --- since he himself a little before, with Hannibal driven out of Italy, cast out of Africa, with Carthage crushed, had freed the commonwealth from the greatest dangers --- would he have undertaken it, unless that grave and vehement war was being thought weighty. Yet if you consider diligently what Mithridates could do and what he carried out and what kind of man he was, surely you will set this king before all those with whom the Roman people waged war. Whom Lucius Sulla, with the greatest and bravest army, a fighting and keen and not unskilled commander --- to say nothing else --- when he had carried the war into him, sent away with all Asia at peace; whom Lucius Murena, the father of this man, very vehemently and very vigilantly harried, repressed in great part, left not crushed; which king, having taken several years to recruit his accounts and his forces of war, gathered such strength in hope and attempt that he thought he would join the Ocean with Pontus, the forces of Sertorius with his own.

\ciceroSection{33} To which war, when two consuls had been sent in such a way that one should pursue Mithridates, the other should guard Bithynia, the affairs of one, both by land and sea ruinous, much enlarged the king's resources and name. The deeds of Lucius Lucullus, however, came forth so great that no greater war can be recalled, nor any waged with greater counsel and courage. For when the entire onset of the war had stood about the walls of the Cyziceni, and Mithridates had thought that city would be the door of Asia for him --- through which once broken open and torn loose the whole province would lie open --- this was all carried out by Lucullus: that the city of most loyal allies should be defended and all the king's forces consumed in the long siege. What of that naval battle off Tenedos, when, on a strained course with most ardent leaders, the enemies' fleet, swelling in hope and spirit, was making for Italy --- do you think it was joined in a moderate contest and a small struggle? I leave aside the battles, I pass over the assaults of towns; cast out of his kingdom at length, however, he was so strong in counsel and authority that, with the king of the Armenians joined to him, he renewed himself with new resources and forces. And if I had now to speak of the achievements of our army and commander, I could recall many and great battles; but it is not this we are doing. This I say:

\ciceroSection{34} If this war, if this enemy, if that king were to be despised, neither would the Senate and the Roman people have judged with such great care that he must be undertaken, nor would they have waged it for so many years and with such great glory under Lucullus, nor would the Roman people with such great zeal have committed the closing out of that war to Gnaeus Pompey. Of whose battles, which are countless, the keenest seems to me that one which was joined with the king and fought with the highest contention. From which battle, when he had wrenched himself away and fled to the Bosphorus, where the army could not approach, even in the last fortune and in flight he kept the kingly name. So Pompey himself, with the kingdom occupied, with the enemy driven from all the coasts and known seats, yet placed so much weight in the soul of one man that, although he himself by victory possessed all that he had held and approached and hoped, yet he did not judge the war finished before life had cast him out. This enemy do you, Cato, despise --- with whom for so many years, in so many battles, so many commanders waged wars, whose life, when expelled and cast out, was so highly esteemed that, when his death was announced, they at last judged the war finished? In this war, then, we maintain that Lucius Murena was known as legate of bravest spirit, of the highest counsel, of the greatest labour; and that this work of his had no less dignity for the attaining of the consulship than this our forensic industry.

\ciceroSection{35} But, you will say, in the candidacy for the praetorship Servius was declared first. Do you proceed to deal with the people as if from a written contract, that the place which it has once given to anyone for an office, the same should be owed in the rest of his offices? For what strait, what Euripus do you suppose has so many motions, so great, so varied, agitations and exchanges of waves, as the workings of an assembly have perturbations and tides? A day intervened, a night put between, often disturbs everything; and the whole opinion sometimes a small breath of rumour changes. Often even without any open cause it falls out otherwise than you reckoned, so that sometimes even the people wonders that it has so come to pass --- as though it had not itself done it.

\ciceroSection{36} Nothing is more uncertain than the crowd, nothing more obscure than the wills of men, nothing more deceptive than the whole reckoning of an assembly. Who supposed that Lucius Philippus, of the highest talent, energy, favour, nobility, could be defeated by Marcus Herennius? Who that Quintus Catulus, surpassing in humanity, wisdom, integrity, by Gnaeus Mallius? Who that Marcus Scaurus, that gravest man, distinguished citizen, bravest senator, by Quintus Maximus? Not only were none of these things thought likely to be so, but even when they had been done, why they had been so done could not be understood. For, as storms are often stirred up by some certain sign of the sky, often unforeseen are roused by no certain reason from some obscure cause, so in this popular storm of the assembly often you may understand by what sign it was stirred, often the cause is so obscure that it seems to have been roused by chance.

\ciceroSection{37} Yet if a reason is to be given, two things were vehemently desired in the praetorship which both have profited Murena much in the consulship: one, the expectation of the games (which had grown both by some rumour and by the zeal and talk of his competitors); the other, that those whom in the province and the whole legation he had had as witnesses both of his liberality and of his virtue had not yet departed. Both of these fortune reserved for him for the candidacy of the consulship. For both Lucius Lucullus's army, which had come together for the triumph, was also at hand for Lucius Murena at the assembly; and the most ample show, which the candidacy of the praetorship had wanted, the praetorship restored.

\ciceroSection{38} Do these supports and aids of the consulship seem small to you --- the will of soldiers, which by itself has weight by numbers, by favour with their own men, and especially in declaring a consul has much authority with the entire Roman people --- the soldier-vote? For commanders are chosen at consular elections, not interpreters of words. Therefore that speech is weighty: ``He nursed me when wounded; he gave me a share of the booty; under this leader we took the camp, joined battle; he never imposed more labour on the soldier than he took on himself: himself both brave and even fortunate.'' How much do you suppose this is worth for men's repute and goodwill? For if such great religion attaches to those elections that hitherto the omen of the prerogative century has always counted, what wonder is it that in this man's case the report and talk of good fortune counted? But if you reckon these things lighter which are the gravest, and you set this urban canvassing before the military, do not so vehemently despise the elegance of his games and the magnificence of his stage --- which have profited him very greatly. For why should I say that the people and the throng of the unschooled are greatly delighted by games? It is less to be wondered at. Yet that is enough for this case, for the elections are of the people and the multitude. Wherefore, if to the people the magnificence of games is a pleasure, it is not to be wondered at that this profited Lucius Murena with the people.

\ciceroSection{39} But if we ourselves --- who are kept by business from the common pleasure, and in the very occupation can have many other delights --- yet are charmed and led by games, what wonder do you have at an unschooled multitude?

\ciceroSection{40} Lucius Otho, a brave man, my friend, restored to the equestrian order not only its dignity but also its pleasure. So this law which pertains to the games is the most welcome of all, because to the most honourable order, with its splendour, the fruit also of pleasantness has been restored. Wherefore games delight men, believe me, even those who dissemble it, not only those who confess it; which I myself perceived in my own candidacy. For we too had a stage as a competitor. If I, who as aedile had given three sets of games, was yet moved by Antonius's games, do you, who by chance had given none, suppose that none of this very silver stage of his which you mock was hostile to you?

\ciceroSection{41} But let all these things by all means be equal --- let forensic work be equal to military, military canvassing to urban; let it be the same to have given the most magnificent games and to have given none. What? In the praetorship itself, do you think there was nothing between your lot and his? His lot was that which all of us, your friends, were wishing for you: the dispensing of justice; in which the magnitude of the business wins glory, the bestowing of equity favour --- in which lot a wise praetor, such as he was, avoids offence by even-handedness in deciding, joins goodwill by gentleness in hearing. A distinguished province and one apt to the consulship, in which the praise of equity, integrity, accessibility is closed at the end with the pleasure of games.

\ciceroSection{42} What was your lot? Sad, harsh: the inquiry on embezzlement, on one side full of tears and squalor, on the other full of accusers and informers; jurors compelled against their will, kept against their will; a scribe condemned, a whole order alienated; a Sullan favouritism censured; many brave men and almost a part of the state offended; suits severely assessed --- whom it pleases forgets, whom it pains remembers. Finally, you did not wish to go to a province. I cannot blame this in you, since I myself approved it both as praetor and as consul. But yet a province brought Lucius Murena much good favour with the highest reputation. He held a levy in Umbria as he set out; the commonwealth gave him an opportunity for liberality, by which he joined to himself many tribes which are made up out of the towns of Umbria. He himself, moreover, in Gaul brought it about by fairness and diligence that our men exacted moneys already despaired of. You meanwhile were, of course, at Rome at the service of friends; I confess it; but yet consider this --- that the zeal of some friends is wont to be diminished toward those by whom they perceive that provinces are despised.

\ciceroSection{43} And since I have shown, judges, that the dignity for the candidacy of the consulship was equal but the fortune of provincial business unequal in Murena and in Sulpicius, I shall now speak more openly in what my friend, Servius, was inferior; and I shall say, with you listening, when the time is gone, what I often said to him alone when the matter was open. ``You do not know how to canvass for the consulship, Servius,'' I often said to you; and in those very things which I saw you both do and say with great and brave spirit I was wont to tell you that you seemed to me a brave accuser more than a wise candidate. First, the threats and warnings of accusing which you used to use daily are the part of a brave man, but they both turn the people's opinion away from the hope of attaining and weaken the zeal of friends. Somehow this always happens --- nor has it been observed in one or two but now in many --- that as soon as a candidate has seemed to be meditating an accusation, he seems to have despaired of the office.

\ciceroSection{44} What then? Are we not to pursue an injury received? Yes, indeed, vehemently; but there is one time for canvassing, another for pursuing. I want a candidate, especially for the consulship, to be brought into the Forum and the Field of Mars in great hope, with great spirit, with great resources. I do not like the inquiry of a candidate, the herald of his repulse; not the procuring of witnesses rather than supporters; not threats more than blandishments; not warnings rather than greetings --- particularly when by this new fashion almost all run together to the houses of all and from the candidates' faces make a guess how much spirit and ability each man seems to have.

\ciceroSection{45} ``Do you see that man, sad and downcast? He lies prostrate, he despairs, he has thrown away his spears.'' This rumour creeps. ``You know that he is meditating accusation, inquiring into competitors, seeking witnesses. Look for another now, since he himself despairs.'' By such rumours the most intimate friends of candidates are weakened, lay aside their zeal; either they cast aside the matter as certain or save their work and favour for the trial and accusation. To the same point is added that even the candidate himself cannot put his entire mind and all his care, work, and diligence into the candidacy. For the thought of accusation is added, no small matter but surely the greatest of all. For it is great to procure those means by which you can drive a man from the state --- particularly one not poor or weak --- who is defended both by himself and by his own and even by strangers. For we all run together to repel dangers; and even those of us who are not openly enemies render to the most distant strangers in the most extreme dangers the duties and services of the most intimate friends.

\ciceroSection{46} For which reason I, who have known the trouble both of canvassing and of defending and of accusing, have so understood that in canvassing the zeal is keenest, in defending the duty, in accusing the labour. So I judge that it can in no way be done that the same man both diligently set up and equip an accusation and a candidacy for the consulship. Few can sustain one; both, no one. You, when you had turned aside from the course of canvassing and brought your mind to accusing, thought you could satisfy each business. You greatly erred. For what day was there, after you entered upon that announcement of accusation, which you did not spend wholly in that pursuit? You demanded a law on canvassing, which was not lacking to you, for the Lex Calpurnia had been written most severely. Way was made for your wish and your dignity. But that whole law would perhaps have armed your accusation, if you had had a guilty defendant; but it stood against your candidacy.

\ciceroSection{47} A heavier penalty against the commons was demanded by your voice; the spirits of the slenderer were stirred. Exile against our order; the Senate yielded to your demand, but not gladly did it set up a harder condition than the common fortune on your motion. To the excuse of illness was added a penalty: the will of many was offended, who must either against the convenience of their health labour, or with the inconvenience of illness leave aside even the rest of the fruits of life. What then? Who carried these things? He who obeyed the authority of the Senate and your will; finally, he carried them whom they least profited. Those things, indeed, which by my supreme will the full Senate rejected --- do you think they have been moderately adverse to you? You demanded confusion of votes †of the prerogative of the Lex Manilia†, an equalizing of favour, of dignity, of votes. Honourable men, in their own neighbourhoods and towns full of favour, bore it with weight that by such a man it was being fought to take away every step of dignity and favour. The same man wished there to be ``elective'' jurors, so that the secret hatreds of citizens, which are now contained in silent disagreements, might break out against the fortunes of every best man.

\ciceroSection{48} All these things were paving for you the way of accusing, but blocking it for attaining. And out of all of them, that wound was thrown upon your candidacy --- not with me silent --- the greatest, about which by a most witty and most copious man, Quintus Hortensius, many things have been most weightily said. Hence too a harder place of speaking is given me: that, when before me both he had spoken and the man of the highest dignity, diligence, and ability of speaking, Marcus Crassus, I at the end should not act some part of the case but should speak about the whole as it should seem to me. So I move in nearly the same matters and, so far as I can, judges, I run to your satiety. But yet, Servius, what an axe do you suppose you threw against your own candidacy when you brought the Roman people into such a fear that they trembled lest Catiline be made consul, while you, with your candidacy laid down and cast away, were preparing your accusation?

\ciceroSection{49} For they saw you inquiring, sad yourself, your friends gloomy; they noticed observations, testifications, the leadings-aside of witnesses, the conferences of subscribers --- by which things even the very faces of candidates are wont to seem more obscure. Catiline meanwhile they saw alert and joyful, surrounded by a chorus of youth, walled in with informers and cut-throats, swollen both with the hope of soldiers and with the promises of my colleague (as he himself used to say), flowing about with an army of Arretine and Faesulan colonists; which throng of a most unlike kind was distinguished by men struck down by the calamity of the Sullan time. His look itself was full of fury, his eyes of crime, his speech of arrogance --- so that the consulship seemed to him already explored and stored at home. He despised Murena; Sulpicius he counted his accuser, not his competitor; he threatened him with violence, threatened the commonwealth.

\ciceroSection{50} What fear was thrown by these things upon all loyal men, and what despair of the commonwealth, if he had been made consul, do not wish to be reminded by me; you yourselves recall it with yourselves. For you remember when the voices of that wicked gladiator had become widespread which he was said to have used at a private meeting --- when he had said that a faithful defender of the wretched could not be found except him who himself was wretched; that the wounded and wretched ought not to trust the promises of the unhurt and the fortunate; therefore those who wished to fill again what was consumed, to recover what was snatched away, should look to what he himself owed, what he held, what he dared; that the man who was to be the leader and standard-bearer of the calamitous ought to be least timid and very calamitous.

\ciceroSection{51} Then, hearing these things, you remember that on my motion a decree of the Senate was passed that the elections should not be held the next day, that we might be able to act on these matters in the Senate. So the next day, with the Senate full, I called Catiline up and ordered him, if he wished anything, to speak about these matters which had been brought to me. And he, as he was always most open, did not clear himself but informed against himself and put on the stigma. For then he said there were two bodies of the commonwealth: one weak with a feeble head, the other firm without a head; to this, if it had so deserved of him, a head would not be lacking while he lived. The full Senate groaned, but did not yet decree severely enough for the indignity of the matter; for in part they were not brave in decreeing because they feared nothing, in part because they feared everything. He burst out of the Senate triumphing in joy --- a man who ought not to have left it alive at all, particularly since the same man, in the same order a few days before, had answered Cato, that bravest man, threatening and announcing trial, that, if any fire were kindled against his fortunes, he would put it out not with water but with ruin.

\ciceroSection{52} Stirred then by these matters, and because I knew that men already conspirators with swords were being led down by Catiline to the Field, I went down to the Field with the firmest guard of bravest men and with that broad and conspicuous breastplate --- not to cover me (for I knew that Catiline was wont to aim not at the side or the belly but at the head and neck), but so that all the loyal might notice and, when they saw the consul in fear and danger (which is what happened), might run together to help and protect him. So when they thought you, Servius, more remiss in canvassing, and saw Catiline inflamed both with hope and lust, all who wished to drive that plague from the commonwealth at once betook themselves to Murena.

\ciceroSection{53} Great is the sudden inclination of wills at consular elections, especially when it has leaned to a good man and one adorned with many other supports of canvassing. He, with the most honourable father and ancestors, with the most modest youth, with the most distinguished legation, with a praetorship approved in jurisdiction, welcome in the show, adorned in the province, having canvassed diligently, and having so canvassed as not to yield to one threatening or to threaten anyone --- is it any wonder that the sudden hope of Catiline's attaining the consulship was a great help to him?

\ciceroSection{54} Now I have the third place of the speech left, on the charges of canvassing --- thoroughly cleared by those who spoke before me, but, since Murena so wished, to be retreated by me; in which place to Gaius Postumus, my friend, that most ornate man, on the indications of distributors and on the moneys seized; to Servius Sulpicius the younger, a witty and good young man, on the centuries of knights; to Marcus Cato, a man surpassing in every virtue, on his own accusation, on the decree of the Senate, on the commonwealth, I shall answer.

\ciceroSection{55} But a few things which suddenly moved my mind I shall first complain of, concerning Lucius Murena's fortune. For although often before, judges, both from others' miseries and from my own daily cares and labours I judged those men fortunate who, removed from the pursuits of ambition, have followed the leisure and tranquillity of life, then indeed in these so great and so unforeseen dangers of Lucius Murena I have been so affected in mind that I cannot enough pity either the common condition of all of us or his outcome and fortune. Who first, while he tried to ascend one step of dignity from the continued offices of his family and ancestors, has come into danger lest he lose both what was left him and what has been won by himself; then on account of his pursuit of new praise is brought even into the peril of his old fortune.

\ciceroSection{56} Which things are grave, judges, but the bitterest is this --- that he has these accusers, men who have come down not from hatred of enmities to accusation, but from zeal for accusation to enmities. For, to leave aside Servius Sulpicius (whom I see moved against Lucius Murena not by injury but by the contention for office), it is a paternal friend, Gaius Postumus, who accuses --- an old, as he himself says, neighbour and intimate, who has brought forward many causes of intimacy but could recall no cause of quarrel. The son of his comrade accuses, Servius Sulpicius, by whose talents all his father's friends ought to be the more fortified. Marcus Cato accuses, who, as he was never alienated from Murena in any matter, was on this condition born to us in this state, that his resources, his talent, ought to be a protection to many even strangers, but a destruction scarcely to anyone, even an enemy.

\ciceroSection{57} I shall therefore answer Postumus first, who somehow seems to me, like a candidate for the praetorship into the consular field, to run as if a circus-vaulter into the four-horse course. If his competitors did no wrong, he conceded their dignity by giving up his pursuit; if any of them, indeed, distributed bribes, then a friend is to be sought who would pursue another's wrong rather than his own. Of Postumus's charges, of those of the young Servius . . .

\ciceroSection{58} I come now to Marcus Cato, who is the foundation and strength of the whole accusation; who however is so weighty an accuser and so vehement that I shall fear his authority much more than his charges. In which accuser, judges, I shall first beg this --- that nothing of Lucius Murena's dignity, nothing of his expectation of a tribuneship, nothing of the splendour and weight of his whole life harm him; in short, that those goods of Marcus Cato, which he has won so as to be able to profit many, may not be of harm to this man alone. Twice consul had been Publius Africanus and had destroyed the two terrors of this empire, Carthage and Numantia, when he accused Lucius Cotta. There was in him the highest eloquence, the highest faith, the highest integrity, an authority as great as in the imperium of the Roman people, which was held by his work. Often I have heard the elder men say this --- that this surpassing force and dignity of the accuser was of much profit to Lucius Cotta. The wisest men who were then judging that case did not wish anyone so to fall in the trial as to seem cast down by the excessive forces of his adversary.

\ciceroSection{59} What of this --- Servius Galba (for it has been handed down to memory) --- did not the Roman people snatch from your great-grandfather, that bravest and most flourishing man, Marcus Cato, leaning to his ruin? Always in this state both the people as a whole and wise judges far-foreseeing for posterity have stood against the too great resources of accusers. I do not want an accuser to bring power into court, no greater violence, no surpassing authority, no excessive favour. Let all these things avail for the safety of the innocent, for help to the powerless, for aid to the calamitous, but in the danger and ruin of citizens let them be rejected.

\ciceroSection{60} For if anyone shall by chance say that Cato would not have come down to accuse, unless he had previously judged the case, he will set up an unjust law, judges, and a wretched condition for men's dangers, if he shall think that the judgment of the accuser against the accused ought to count for some prejudgment. I cannot, Cato, on account of my singular mind's judgment of your virtue, blame your decision; some things perhaps I might shape and lightly amend. ``You commit not many faults,'' said the older master to that bravest man, ``but you do commit faults; I can rule you.'' But I cannot rule you. I should say most truly that you commit no fault, and that in no matter you are of such a kind that you should seem to need to be corrected rather than lightly bent. For nature herself fashioned you for honour, weight, temperance, magnitude of soul, justice, in short for all the virtues a great and lofty man. There has been added to that a doctrine not moderate or mild, but, as it seems to me, a little harsher and harder than either truth or nature suffers.

\ciceroSection{61} And since this speech is to be delivered by us neither in an unschooled multitude nor in any meeting of country folk, I shall a little more boldly dispute about the studies of humanity, which are both known and pleasant to me and to you. In Marcus Cato, judges, know that these goods which we see, divine and surpassing, are properly his own; that what we sometimes miss are all not from nature but from a teacher. For there was a certain man of the highest talent, Zeno, whose followers in his discoveries are called Stoics. His opinions and precepts are of this kind: that the wise man is never moved by favour, never forgives anyone's offence; that no man is merciful unless foolish and trifling; that it is not the part of a man to be entreated or appeased; that the wise alone are, even if most distorted, beautiful, if most beggarly, rich, if they serve in slavery, kings; while we who are not wise they call runaway slaves, exiles, enemies, in short madmen; that all faults are equal; that every offence is unspeakable crime, and that he who has strangled a barnyard cock without cause has offended no less than he who has throttled his own father; that the wise man supposes nothing, repents of nothing, is deceived in no matter, never changes his opinion.

\ciceroSection{62} This Marcus Cato, that man of the highest talent, led on by most learned authors, snatched up --- not for the sake of disputing, as a great part of men do, but in order to live so. The publicans demand something; ``beware lest favour have any weight.'' Some suppliants come, wretched and calamitous; ``you would be wicked and unspeakable if you did anything brought to it by mercy.'' Someone confesses he has erred and asks pardon for his offence; ``it is unspeakable wickedness to forgive.'' But the offence is light. ``All faults are equal.'' You have said something; ``it is fixed and settled.'' You were led not by the matter but by opinion; ``the wise man supposes nothing.'' You have erred in some matter; he thinks you are being slandered. From this discipline are these of ours: ``I said in the Senate that I would denounce the name of a consular candidate.'' You said it angry. ``The wise man,'' he says, ``never grows angry.'' But for the time. ``It is the part of a wicked man,'' he says, ``to deceive by lying; to change one's opinion is base, to be entreated is crime, to pity is scandal.''

\ciceroSection{63} But ours --- I shall confess, Cato, that I too in my youth, distrustful of my own talent, sought the supports of doctrine --- ours, I say, of Plato and Aristotle, moderate and temperate men, say that with the wise man favour sometimes counts; that to pity is the part of a good man; that there are distinct kinds of offences and unequal punishments; that with a constant man there is a place for forgiving; that the wise man himself often supposes something he does not know, sometimes grows angry, the same is entreated and appeased, what he has said sometimes (if so it is more right) changes, departs from his opinion now and then; that all virtues are moderated by a kind of mean.

\ciceroSection{64} If some fortune, Cato, with that nature of yours, had brought you to these masters, you would not indeed be a better man, nor braver, nor more temperate, nor juster --- for you cannot be --- but a little more inclined to gentleness. You would not accuse, drawn by no enmities, vexed by no injury, a most modest man endowed with the highest dignity and honour; you would think, when fortune had placed both you and Lucius Murena in the keeping of the same year, that you were joined to him by some bond of the commonwealth; that which you said atrociously in the Senate, either you would not have said, or, if you could, you would interpret in the milder part. And as to your very self,

\ciceroSection{65} so far as I divine by my opinion, you, now stirred up by some impetus of mind and lifted by the force of nature and of talent and burning with the recent zeal of precepts, will be bent by use, will be eased by time, will be softened by age. For those very preceptors and masters of virtue of yours seem to me to have set the limits of duties a little farther than nature would wish, so that, when in mind we had striven to the uttermost, we might yet stand where we ought. ``Forgive nothing.'' Rather, something, not everything. ``Do nothing for the sake of favour.'' Rather, resist favour, when duty and faith demand it. ``Do not be moved by mercy.'' Yes, in dissolving severity; but yet there is some praise of humanity. ``Stay in your opinion.'' Indeed, unless another better opinion shall have conquered the opinion.

\ciceroSection{66} Of this kind was that Scipio, who did not regret doing the same thing as you --- having a most learned man, Panaetius, at home; by whose discourse and precepts, although they were the same things which delight you, yet he was not made harsher but, as I have heard from old men, the gentlest. Who, indeed, was more affable than Gaius Laelius, who pleasanter from this same study, who weightier than he, who wiser? I can say the same things about Lucius Philus, about Gaius Gallus; but I shall now lead you to your own home. Do you suppose anyone was more agreeable, more sociable, more moderate to every reckoning of humanity than Cato, your great-grandfather? When you were speaking truly and weightily about his surpassing virtue, you said you had a domestic example to imitate. That example is set before you at home; but yet the likeness of his nature could come more to you who are sprung from him than to each one of us; while for imitation that pattern is set before me as much as before you. But if you sprinkle his affability and accessibility upon your weight and severity, those things will not indeed be better which now are the best, but certainly seasoned more pleasantly.

\ciceroSection{67} Wherefore, to return to what I was treating, take Cato's name out of the case for me, remove force, leave aside authority --- which in trials should either count nothing or count for safety --- and meet me on the charges themselves. What do you accuse, Cato, what do you bring into court, what do you allege? You accuse canvassing; I do not deny it. You blame me, that I should defend the very thing I have punished by law. I have punished canvassing, not innocence; canvassing itself, indeed, I shall accuse with you, if you wish. You said that on my motion a decree of the Senate was passed --- that, if for hire they had gone forth to meet candidates, if hired they had escorted them, if to gladiator-shows place by tribes had generally been given and likewise dinners had generally been given --- it would seem to be done against the Lex Calpurnia. So the Senate so judges that these things, if they have been done, seem to have been done against the law; it decrees what is not necessary, while it accommodates the candidates. For whether it has been done or no is vehemently in question; if it has been done, no one can doubt that it is against the law.

\ciceroSection{68} It is therefore ridiculous to leave uncertain what is doubtful, to judge what cannot be doubtful to anyone. And this is decreed at the demand of all the candidates --- that from the decree of the Senate it cannot be understood whose interest it is, nor against whom it is. Wherefore prove that those things were committed by Lucius Murena; then I myself will grant to you that they were committed against the law. ``Many came forth to meet him as he came down from the province.'' This is wont to happen to one canvassing for the consulship; for who is not met on his return? ``What was that crowd?'' First, even if I could not give you that account, what is there to wonder at, when such a man was arriving, a candidate for the consulship, that many should have come forth to meet him? If this had not been done, it would have seemed more to be wondered at.

\ciceroSection{69} What of this --- if I add that which does not differ from custom, that many were asked to come --- is it either criminal or to be wondered at, in a state in which we are wont, when asked, to come down at almost first night out of the farthest part of the city to escort the sons of the lowest men, that men should not have been reluctant to come forth at the third hour to the Field of Mars, especially when asked in the name of such a man? What if all the partnerships came, of whose number many sit as judges? what if many honourable men of our own order? what if that most dutiful nation of all candidates, which suffers no one not to enter the city honourably? what, finally, if our very accuser Postumus came to meet him with a quite considerable troop of his own? what is there to wonder at in that crowd? I leave aside clients, neighbours, fellow tribesmen, the entire army of Lucullus which had come for the triumph in those days; this I say --- that an unrewarded throng in this duty has never been wanting not only to anyone's dignity but not even to anyone's wish. ``But many escorted him.'' Prove for hire; I will grant it to be a charge.

\ciceroSection{70} With this set aside, what do you blame? ``What need is there,'' he says, ``of escorts?'' Do you ask of me what need there is of that which we have always used? Slender men have one place in our order for either earning a kindness or returning one --- this work and attendance in our candidacies. For neither can it be done nor is it to be demanded of us or of the Roman knights that they escort their friend candidates whole days; by whom if our house is thronged, if from time to time we are escorted to the Forum, if we are honoured by one walk along the basilica, we seem to be diligently attended and cultivated; that constant attendance is of the slenderer friends and of those not occupied, of whom the supply is not wont to be wanting to good and beneficent men.

\ciceroSection{71} Do not therefore, Cato, snatch from the lower kind of men this fruit of duty; allow those who hope all things from us themselves also to have something they can give us. If there shall be nothing besides their own vote --- the slender, even if they vote, count nothing in favour. They themselves, finally, as they are wont to say, cannot speak for us, cannot give surety, cannot call us to their house. And all these things they ask of us; nor do they think they can compensate by anything else what they get from us, except by their own service. So they have stood against both the Lex Fabia (which is on the number of escorts) and the decree of the Senate which was made under the consul Lucius Caesar. For there is no penalty which can shut out the attentiveness of the slenderer from this old institution of duties.

\ciceroSection{72} ``But shows have been given by tribes, and to dinner men were generally summoned.'' Although this thing was not done by Murena at all, judges, but by his friends in the customary fashion and measure, yet, prompted by the very thing, I recall how many of our points, Servius, these complaints raised in the Senate detracted. For what time was there, in our memory or our fathers', when this --- whether it is canvassing or liberality --- was not such that place both in the Circus and in the Forum was given to friends and fellow tribesmen? These rewards and conveniences slenderer men obtained from their fellow tribesmen by the old institution * * *

\ciceroSection{73} that a prefect of engineers once gave a place to his fellow tribesmen --- what shall they decide against the foremost men who in the Circus have prepared whole stands for their fellow tribesmen's sake? All these charges of escorts, of shows, of dinners alike have been thrown by the multitude upon your excessive diligence, Servius; in which Murena however is defended by the authority of the Senate. For what? Does the Senate think it a charge to come forth to meet him? No, but for hire. Prove it. To escort him in numbers? No, but hired. Prove it. To give a place to view, or to invite to dinner? Not at all, but generally, indiscriminately. What is generally? All. Therefore, if Lucius Natta, a young man of the highest place, of whom we already see what spirit and what kind of man he will be, wished to be in the centuries of knights and to be in favour both for this duty of connection and for the future, that will not be to his stepfather's reproach or charge; nor, if a Vestal virgin, his kinswoman and connection, gave him her own gladiator-place, has not she done piously and is not he removed from blame? All these are duties of connections, conveniences of the slender, services of candidates.

\ciceroSection{74} But Cato deals with me austerely and Stoically; he denies that it is right to lure goodwill by food, denies that the judgment of men in granting magistracies ought to be corrupted by pleasures. Therefore, if anyone summons to dinner for the sake of his candidacy, let him be condemned. ``So you,'' he says, ``would seek of me supreme command, supreme authority, the rudders of the commonwealth, by warming men's senses and softening their minds and applying pleasures? Were you seeking,'' he says, ``the pandering of a herd of dainty youth, or the empire of the world from the Roman people?'' A horrible speech; but use, life, manners, the state itself reject it. Nor indeed have the Lacedaemonians, the authors of that life and discourse, who at daily feasts recline on oak, nor the Cretans, none of whom ever tasted food reclining, kept their commonwealths better than the Romans, who divide the times of pleasure and labour --- of whom the one were destroyed by the single coming of our army, the others, by the protection of our empire, keep their discipline and their laws.

\ciceroSection{75} Wherefore, Cato, do not by too severe a speech blame the institutions of our ancestors which the matter itself, which the long duration of empire approves. There was a learned man among our fathers, of the same study, an honourable and noble man, Quintus Tubero. He, when Quintus Maximus was giving a public banquet to the Roman people in the name of Publius Africanus, his uncle, was asked by Maximus to spread the dining-couch, since Tubero was the son of the same Africanus's sister. And he, that most learned man and Stoic, spread little Punic couches with kid-skins and set out Samian ware, as if Diogenes the Cynic were dead and not the death of the divine Africanus being honoured; whom when Maximus on his last day was praising, he gave thanks to the immortal gods that that man had been born most of all in this commonwealth; for the empire of the world had to be where he was. At the celebration of his death the Roman people bore heavily this perverse wisdom of Tubero,

\ciceroSection{76} and so the most upright man, the best citizen, although he was the grandson of Lucius Paullus, the son of Africanus's sister (as I said), by these little kid-skins was thrown out of the praetorship. The Roman people hates private luxury, loves public magnificence; it loves not lavish feasts; squalor and inhumanity much less. It distinguishes the reckoning of duties and times, the alternation of labour and pleasure. For when you say that men's minds ought to be lured by no thing to the granting of magistracy except by dignity, you yourself, in whom is the highest dignity, do not keep this. For why do you ask anyone to be zealous for you, to help you? You ask me to be over you, to entrust myself to you. What then? Ought I to be asked this by you, or rather you by me to undertake labour and danger for my safety?

\ciceroSection{77} What of this --- that you have a name-prompter (\textit{nomenclator})? in this, indeed, you trick and deceive. For if it is honourable that your fellow citizens be addressed by you by name, it is base that they should be better known to your slave than to you. But if you do already know them, are they nevertheless to be addressed through a prompter when you are canvassing, as if you were uncertain? What of this --- that, when you have been prompted, you yet greet them as if you had known them yourself? What of this --- that, after you have been declared elected, you greet much more carelessly? All these things, if you trace them to the reckoning of the state, are right; but if you wish to weigh them against the precepts of the Stoic discipline, they are found most wrong. Wherefore neither must those fruits of games, gladiator-shows, banquets be torn from the Roman commons --- which all our ancestors set up --- nor must that goodwill be taken from candidates which signifies liberality more than bribery.

\ciceroSection{78} ``But the commonwealth has brought you to accuse.'' I believe, Cato, that you have come with that mind and that opinion; but you slip from imprudence. What I do, judges, I do for the sake both of friendship and of dignity for Lucius Murena, and I cry out and bear witness that I do it for the sake of peace, of quiet, of concord, of liberty, of safety, of the very life of all of us. Hear, hear the consul, judges --- I shall say nothing arrogant; only this --- thinking whole days and nights about the commonwealth! Lucius Catiline did not so far despise and contemn the commonwealth as to think that with the supply of men he led out with him he could crush this state. The contagion of his crime spreads more widely than anyone supposes; it pertains to more men. Within, within, I say, is the Trojan horse; under my consulship, you shall never be crushed by it while you sleep.

\ciceroSection{79} You ask of me whether I fear Catiline at all. Nothing; and I have taken care that no one should fear him. But his forces, which I see here, I say are to be feared; nor is the army of Lucius Catiline so much to be feared now as those men who are said to have deserted that army. For they have not deserted, but, left by him in look-outs and ambushes, have remained on our head and necks. These wish to cast down a sound consul and a good commander, joined both by nature and by fortune to the safety of the commonwealth, from the protection of the city, and to throw him from the keeping of the state by your votes. Whose iron and audacity I have driven back on the Field of Mars, weakened in the Forum, often crushed even at my own house, judges; if to these you shall hand over the other consul, they will gain much more by your votes than by their swords. It greatly matters, judges --- which I have, with many opposing, accomplished and made good --- that on the Kalends of January there be two consuls in the commonwealth.

\ciceroSection{80} Do not suppose that they are using ordinary counsels or accustomed paths. It is not a wicked law, not ruinous bounty, not some heard-of evil that is being sought against the commonwealth. The designs in this state, judges, are entered upon: of destroying the city, of slaughtering the citizens, of extinguishing the Roman name. And these things citizens (citizens, I say, if it is right to address them by this name) think and have thought against their own country. Their designs daily I meet, their audacity I weaken, their crime I resist. But I warn you, judges. My consulship is now at its end; do not snatch from me the substitute of my diligence, do not take away him to whom I wish to hand over the commonwealth, to be defended unharmed from these so great dangers.

\ciceroSection{81} And to these evils, judges, what other thing comes you do not see? You, you I appeal to, Cato; do you not foresee the storm of your year? For now in yesterday's assembly the ruinous voice of a tribune-designate, your colleague, has thundered; against whom much your mind, much all loyal men have provided, who called you to candidacy for the tribuneship. Everything which has been agitated through this triennium, from the time when you know that by Lucius Catiline and Gnaeus Piso the design of killing the Senate was entered upon, breaks out now into these days, into these months, into this time.

\ciceroSection{82} What place is there, judges, what time, what day, what night, when I am not snatched from the snares and points of these men --- not by my own only but much more by divine counsel --- and fly? They wish me killed not under my own name, but as the watchful consul to be moved from the protection of the commonwealth. Nor would they less wish, Cato, to remove you also by some method, if they could; which, believe me, they both pursue and contrive. They see how much spirit there is in you, how much talent, how much authority, how much defence of the commonwealth; but when they shall have seen the tribunician force stripped of consular authority and aid, then they think they will the more easily crush you, unarmed and weakened. For that no consul be substituted they do not fear. They see that this will be in the power of your colleagues; they hope that Decimus Silanus, that distinguished man, may be set up without a colleague, you without a consul, the commonwealth without protection.

\ciceroSection{83} In these so great matters, in these so great dangers, it is your part, Marcus Cato, who seem to me to have been born not for yourself but for your country, to see what is going on, to keep as a helper, defender, partner in the commonwealth a consul not greedy --- a consul (which this time most of all demands) settled by fortune for the embracing of quiet, by knowledge for waging war, by spirit and use for the sustaining of any business you wish. Although the whole power of this matter lies in you, judges; you in this case hold the entire commonwealth, you steer it. If Lucius Catiline with his council of wicked men whom he has led out with him could pass judgment in this matter, he would condemn Lucius Murena; if he could kill him, he would kill him. For the reckoning of his designs requires that the commonwealth be deprived of help, that against his frenzy the supply of commanders be diminished, that with the adversary driven off, a greater opportunity be given to the tribunes of the plebs of stirring up sedition and discord. Shall, then, men picked out from the most ample orders, the most honourable and the wisest, judge what that most dangerous gladiator, that enemy of the commonwealth, would judge?

\ciceroSection{84} Believe me, judges, in this case you will pass an opinion not on Lucius Murena's safety only but on your own. We have come into the extreme crisis; there is now nothing whence we may renew ourselves, nor where, slipped, we may stand. Not only must the help we have not be diminished, but new must, if it can be, be procured. For the enemy is not at the Anio (which in the Punic war seemed gravest), but in the city, in the Forum --- immortal gods! one cannot say it without a groan --- not no one even in that very shrine of the commonwealth, in the Senate-house itself, I say, is not no one an enemy. May the gods grant that my colleague, that bravest man, armed crush this wicked brigandage of Catiline! I, togate, with you and all loyal men as helpers, shall by counsel scatter and crush this danger which the commonwealth, having conceived, is bringing forth.

\ciceroSection{85} But what at last shall be done if these things slipped from our hands shall overflow into the year that follows? There will be one consul, and he occupied not in administering the war but in supplying a colleague. Those who shall hinder this * * * that monstrous, dangerous plague of Catiline will burst forth, by which * * * he threatens; into the suburban fields he will suddenly fly; in the city will be frenzy, in the Senate fear, in the Forum conspiracy, in the Field an army, in the fields desolation; in every seat and place we shall fear the sword and flame. Which now have long been preparing, those same things, if the commonwealth be furnished with its own protections, will easily be crushed both by the counsels of magistrates and by the diligence of private men.

\ciceroSection{86} Since these things are so, judges, first for the sake of the commonwealth (than which no thing ought to be more important to anyone) I, by my supreme and to you well-known diligence in the commonwealth, warn you; by my consular authority I urge; by the magnitude of the danger I beseech you, that you take counsel for quiet, for peace, for safety, for your own life and that of the rest of the citizens; then I, the same, drawn by the duty both of defender and friend, beg and beseech, judges, that you not by a new lamentation overwhelm the recent congratulation of a wretched man worn out by both bodily disease and grief of mind, Lucius Murena. Just now, adorned with the greatest kindness of the Roman people, he seemed fortunate, because he was the first to bring a consulship to an old family, the first to a most ancient town; now the same man, in squalor and dirt, worn out by disease, lost in tears and grief, is your suppliant, judges; he calls upon your faith, implores your mercy, looks upon your power and your resources.

\ciceroSection{87} Do not, by the immortal gods, judges, deprive him of this thing by which he hoped to be more honoured, together with the other honours he had won before and with all his dignity and fortune. And so Lucius Murena, judges, asks and beseeches you: if he has unjustly hurt no one, if he has violated no one's ear or will, if he has been hateful (to put it most lightly) to no one either at home or in the field, let there be a place with you for modesty, let there be a refuge for downcast men, let there be aid for shame. The stripping of the consulship ought to carry great mercy, judges; for at one stroke with the consulship are torn away all things; the consulship itself, indeed, in these times can have no odium, for it is set against the assemblies of the seditious, the snares of conspirators, the weapons of Catiline; in short, it is set alone against every danger and every injury.

\ciceroSection{88} Wherefore I do not see what there is to envy in Murena or in any of us in this distinguished consulship, judges; what however is to be pitied, both turns before my eyes and you can see and observe. If --- which omen Jupiter avert --- you should crush him by your votes, whither, wretched, will he turn? Home? to see that portrait of his most distinguished father, his parent, which a few days before he saw laurelled at his own congratulation, the same now disfigured by infamy and mourning? Or to his mother, who, wretched, just now kissed her son the consul, and now is in torment and anxious lest a little later she should see the same stripped of all dignity?

\ciceroSection{89} But why do I appeal to his mother or his house --- whom the new penalty of the law deprives of home and parent and the customary intercourse and sight of all his own? Will the wretched man go, then, into exile? Whither? to the parts of the East where he was for many years legate, where he led armies, where he carried out the greatest deeds? But it is great pain, whence you have departed with honour, to return there with infamy. Or shall he hide in the contrary part of the world, so that Transalpine Gaul, which lately saw him most gladly with the supreme command, may now see the same in mourning, in grief, in exile? In that province, moreover, with what spirit will he look upon Gaius Murena his brother? what will be his pain, what the other's mourning, what the lamentation of each --- and how great the perturbation of fortune and of speech, when in the very places where a few days before messengers and letters had spread the news that Murena had been made consul, and whence guests and friends had hastened to Rome to congratulate, suddenly should arise the messenger of his own calamity!

\ciceroSection{90} If these things are bitter, if wretched, if mournful, if most foreign to your mildness and mercy, judges --- preserve the kindness of the Roman people, give the commonwealth back its consul, give this to his very modesty, give it to his dead father, give it to his stock and family, give it also to Lanuvium, that most honourable town, which in this whole case you have seen present in numbers and in mourning. Do not from the ancestral rites of Juno Sospita (to whom every consul must do sacrifice) tear away its own and its native consul above all. Whom I to you, if my commendation has any weight or my confirmation any authority, I, consul, commend the consul, judges, in such a way that I promise and pledge that he will be most desirous of quiet, most zealous for loyal men, keenest against sedition, bravest in war, most hostile to this conspiracy which now shakes the commonwealth.
