% De Lege Agraria I (de-lege-agraria-1) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.
%
% The opening of the speech is lost; what survives are four short
% fragments preserved by other authors (1f-4f) and a substantial
% middle and closing portion (sections 1-27 in this Perseus text,
% which corresponds to roughly the second half of the speech as
% Cicero delivered it on 1 January 63 BC in the Senate).

\ciceroSection{1f} . . . a beardless youth . . .

\ciceroSection{2f} . . . they will seize Capua by planting colonists, fortify Atella with a garrison, hold Nuceria and Cumae with a multitude of their own, bind down all the other towns with garrisons.

\ciceroSection{3f} . . . The whole Propontis and Hellespont will therefore go up under the auctioneer's voice; the entire coast of the Lycians and the Cilicians will be assigned; Mysia and Phrygia will obey the same terms and the same statute.

\ciceroSection{4f} . . . The booty, the spoils, the lots of plunder, even the camp of Gnaeus Pompey while he sits there as commander, the ten commissioners (decemviri) will sell.

\ciceroSection{1} . . . [text breaks off]. What was sought openly, that is now attacked from below by mining tunnels. For the decemvirs will say --- as is being said by many and has often been said --- that after that same consulship, by the will of King Alexander, his kingdom became the property of the Roman people. Will you, then, give Alexandria, by these men's covert solicitation, to the very men whose open onslaught you resisted? Are these, by the immortal gods, the counsels of sober men, in your judgment, or the dreams of drunkards? --- the deliberations of the wise, or the wishes of the deranged?

\ciceroSection{2} See now in the next clause how this filthy gourmand throws the commonwealth into chaos, scatters and dissipates the possessions left to us by our ancestors, plays the spendthrift heir on the patrimony of the Roman people no less than on his own. He sets out in his law revenues that the decemvirs are to put up for sale --- that is, he posts an auction notice for public goods. He wants land bought up to be distributed; he is hunting for money. No doubt he will think up something and bring it forward. For in the previous clauses the dignity of the Roman people was being violated, the very name of our empire summoned to the common hatred of the world, peaceful cities, the lands of allies, the standings of kings handed over to the decemvirs as a gift; now ready cash, sure, counted out, is what is being hunted.

\ciceroSection{3} I wait to see what the watchful and acute tribune of the plebs will think up. ``Let it go for sale,'' he says, ``the Scantian Forest.'' Did you find this forest, then, among possessions left to the people, or among the censors' pasturelands? If you have rooted out something, found it, dug it from darkness --- though it is unfair, still by all means consume what is convenient, since you have brought it in. But are you to sell off the Scantian Forest while we are consuls and this Senate sits? Are you to lay a hand on any revenue at all, are you to wrench from the Roman people its provisions for war and its ornaments of peace? Then I shall judge myself a more useless consul than those bravest men who flourished among our ancestors, if the revenues that were procured for the Roman people in their consulships are judged not even capable of being kept in mine.

\ciceroSection{4} He puts up for sale the possessions of Italy in good order, every one. He is, to be sure, careful in this --- he leaves nothing out. He goes through all Sicily in the censors' tablets; he leaves no building, no fields. You have heard the auction of the Roman people advertised by a tribune of the plebs, fixed for January; and, I suppose, you have no doubt that the men who acquired these things by arms and courage refrained from selling them off for the treasury's sake precisely so that there would be something left for us to sell off as a bounty.

\ciceroSection{5} Look now at where they are aiming, more openly than before. For in the earlier part of the law it was made clear by me how they were attacking Pompey; now they will give themselves away. They order the lands of the Attalenses and the Olympeni to be sold, lands which the victory of Publius Servilius, that bravest of men, joined to the Roman people; then the royal lands in Macedonia, which were partly won by the courage of Titus Flamininus, partly by Lucius Paullus, who conquered Perseus; then the most excellent and fruitful Corinthian land, which by the command and good fortune of Lucius Mummius was annexed to the revenues of the Roman people; afterwards lands in Spain near New Carthage, possessed by the surpassing courage of the two Scipios; then they sell off old Carthage itself --- which Publius Africanus stripped of roof and walls and consecrated to the eternal memory of mankind, whether to mark the calamity of the Carthaginians, or to bear witness to our victory, or under some religious binding put upon him.

\ciceroSection{6} With these badges and fillets of empire put up for sale, with which our ancestors handed down to us the commonwealth adorned, they order the lands to be sold which King Mithridates possessed in Paphlagonia, in Pontus and Cappadocia. Are they not openly seen to be hunting Pompey's army with the auctioneer's spear close at hand, when they order the very lands to be sold in which he is even now waging war and operating?

\ciceroSection{7} And what about this --- that they fix no place for themselves for the auction they propose? For the decemvirs are permitted by the law to sell in whatever places seem good to them. Censors are not allowed to let out the revenues anywhere except in the sight of the Roman people; will these men be allowed to sell even in the farthest lands? Yet even the most worthless of men, when they have squandered their patrimonies, do this much: they hold their auctions in the auction halls rather than in cross-streets and back-lanes. This man by his law permits the decemvirs to sell the goods of the Roman people in whatever darkness suits them, in whatever solitude they wish.

\ciceroSection{8} Now, in all the provinces, kingdoms, and free peoples, how bitter, how terrifying, how profitable the criss-crossing of the decemvirs is going to be --- do you not see it? Even those to whom you have given commissions for the recovery of inheritances, who as private men, on private business, have gone abroad equipped with no great wealth and no supreme authority --- you hear well enough, I think, how heavy their arrivals tend to be on our allies.

\ciceroSection{9} For which reason, what amount of terror and evil do you suppose hangs by this law over all the nations, when ten commissioners are to be let loose upon the world with supreme imperium, supreme greed, and an infinite lust for everything? Heavy as their arrivals will be, terrifying as their fasces, their power of judgment is what will be unbearable: for it will be allowed to them to declare public whatever seems so to them, and to sell whatever they have so declared. Even what scrupulous men will not do --- to receive money for not selling --- even this very thing will be allowed them by law. From this, what plunderings, what compacts, what, in short, traffic in justice and in fortunes do you think there will be in every place?

\ciceroSection{10} For what was bounded in the earlier part of the law by the formula ``in the consulship of Sulla and Pompey'' they have now made open and unbounded again. For he orders the same decemvirs to lay a heavy assessment on every public tract, so that they may both free those tracts which it suits them to free, and declare public those which they please. In which trial it cannot be made out whether their severity is going to be the bitterer thing, or their generosity the more profitable. There are, however, two exceptions in the whole law, not so much unjust as suspicious. He excepts in the assessment the Recentoric land in Sicily, and in the lands to be sold those tracts about which protection is given by treaty. These last are in Africa, and are held by Hiempsal.

\ciceroSection{11} Here I ask: if Hiempsal is sufficiently protected by treaty, and the Recentoric land is private property, what need was there to except them? but if both that treaty has some doubt about it, and the Recentoric land is sometimes said to be public, who is likely to suppose that just two cases were found in the world that they would spare gratis? Is there indeed any coin so well hidden anywhere that the architects of this law have not sniffed it out? Provinces, free cities, allies, friends, and finally kings they drain dry; they lay their hands on the revenues of the Roman people.

\ciceroSection{12} It is not enough. Listen, listen, you who have held armies and waged wars by the most ample judgment of the people and the Senate: whatever has come to anyone from booty, from spoils, from gold for the victor's crown, which has not been spent on a monument or paid into the treasury --- this he orders to be reported to the decemvirs! By this clause they have many hopes; they set up an inquiry, by their own decision, against all generals and their heirs; but they reckon to take the largest sum of money from Faustus. The case sworn jurors refused to take up, these decemvirs of yours have taken up: perhaps they think it was passed over by the jurors precisely because it was being kept for them.

\ciceroSection{13} Then for the future also he provides most carefully that whatever money any general has, he must hand over directly to the decemvirs. Here, however, he excepts Pompey, in just the same fashion, it seems to me, as in that law by which foreigners are expelled from Rome Glaucippus is excepted: for by such an exception one man is not given a benefit, but one man is spared an injury. Yet the man whose share of the spoils he releases, on his revenues he encroaches. For he orders any money received after our consulship from new revenues to be at the decemvirs' disposal --- as though we did not understand that they are planning to sell off the very revenues which Gnaeus Pompey has annexed.

\ciceroSection{14} You see now, senators, how by every means and method the decemviral money has been built up and heaped together. The odium of this fund will be diminished: for it will be spent on land purchases. Excellent. Who, then, will buy this land? The same decemvirs. You, Rullus --- I leave the rest aside --- will buy what you want, sell what you want, do both at the price you want. For the worthy gentleman provides that he shall not buy from an unwilling seller. As though we did not see that to buy from an unwilling seller is a crime, and from a willing one a profit. How much land, to leave others aside, will your father-in-law sell you, and --- if I rightly know his fairness of mind --- sell you not unwillingly? Others will gladly do the same: to exchange the odium of holding for the cash, to take what they want, to give up what they can scarcely keep.

\ciceroSection{15} Now consider the boundless and intolerable licence of every kind of operation. Money has been gathered to buy lands; further, lands will not be bought from the unwilling. Suppose the holders agree among themselves not to sell --- what will happen? Will the money be paid back? It is not allowed. Will it be exacted? He forbids it. Yet very well: there is nothing that cannot be bought, if you give as much as the seller wishes. Let us strip the world bare, sell off the revenues, empty out the treasury --- so that, with the holders enriched whether to envy or to plague, the lands may yet be bought.

\ciceroSection{16} What then? What will be the planting of colonies on these lands, what the whole rationale and layout of the business? ``Colonies,'' he says, ``will be planted.'' How many? of what men? in what places? for who does not see that all these things have to be considered in colonies? Did you suppose, Rullus, that we and these confederates of yours, the contrivers of all this, were going to hand over to you the whole of Italy unarmed, that you might secure it with garrisons, occupy it with colonies, hold it bound and tied down by every kind of chain? For where is provision made that you shall not plant a colony on the Janiculum, that you shall not be able to crowd and crush this city with another city? ``We shall not do it,'' he says. First, I do not know that; second, I am afraid; finally, I shall not allow our salvation to be at your kindness rather than at our own foresight.

\ciceroSection{17} As for your wishing to fill all Italy with your colonies --- did you suppose none of us would understand what that meant? It is written: ``Into whatever municipalities and into whatever colonies the decemvirs wish, they may plant colonists whom they wish, and assign lands to them in whatever places they wish.'' So that, when they have occupied all Italy with their soldiers, no hope is left to us not only of keeping our dignity but even of recovering our liberty. And these things I am charging on suspicion and conjecture.

\ciceroSection{18} Now every man's every doubt will be cleared away: now they will openly show that the very name of this commonwealth, the seat of the city and of empire, finally this temple of Jupiter Best and Greatest, and this citadel of all nations, displeases them. They want colonists planted at Capua: that city against this city again, to set up; thither to carry their resources, to transfer the name of empire, is what they have in mind. The place which on account of the richness of its lands and the abundance of all things is said to have begotten arrogance and cruelty --- in that place these new colonists of ours, picked out by the decemvirs for every crime, will be settled. And, I suppose, in a city where men born in long-standing dignity and fortune could not bear an abundance of things with moderation, in that city these satellites of yours will modestly contain their insolence!

\ciceroSection{19} Our ancestors took away from Capua the magistracies, the senate, the common council, all, in short, the marks of a commonwealth, and left in the city nothing but the empty name ``Capua'' --- not from cruelty (for what could be milder than they, who often gave back their own to even foreign enemies once defeated?), but from policy: because they saw that, if any vestige of a commonwealth were kept within those walls, the city itself could provide a seat for empire. As for you, did you not see how ruinous these things were? --- unless, I suppose, you were eager to overthrow the commonwealth and to procure for yourselves a new dominion.

\ciceroSection{20} For what is to be guarded against in the planting of colonies? If luxury, Capua corrupted Hannibal himself; if arrogance, this seems to have been born there of the Campanian disdain; if a stronghold, that colony is not set in front of this city, but set against it. And how, immortal gods, it is armed! For in the Punic war whatever Capua could do, she did by her own self; now all the cities round about Capua will be occupied by colonists through the same decemvirs --- for to that end the law itself permits the decemvirs to plant colonists, whom they will, into all the towns they will. And they order the Campanian land and the Stellate plain to be divided for these colonists.

\ciceroSection{21} I do not complain of the diminution of the revenues, of the disgrace of this loss and damage; I pass over those things which there is no man who could not most gravely and most truly complain of --- that we have not been able to keep the head of the public patrimony, the most beautiful possession of the Roman people, the support of the corn supply, the granary of war, a revenue placed under the seal and the bars of the commonwealth; that we have, in the end, conceded that very land to Publius Rullus, the very land which by itself had stood up both to Sulla's despotism and to the bounties of the Gracchi; I do not say only this --- that this is a revenue in the commonwealth which, when others fail, remains; that, when interrupted, never falls quiet; that shines in peace, does not grow shabby in war, sustains the soldier, fears not the enemy. I pass over this whole speech and reserve it for the assembly: I am speaking now of the danger to our safety and our liberty.

\ciceroSection{22} For what part do you suppose will be left whole to you in the commonwealth, or in the keeping of your liberty and dignity, when Rullus and those whom you fear far more than Rullus shall have occupied Capua and the cities around Capua with every band of needy and wicked men, with every kind of resources, with all their silver and gold? These things, senators, I shall resist with vehemence and sharpness, and I will not allow these men in my consulship to bring out the schemes they have long meditated against the commonwealth.

\ciceroSection{23} You have erred, Rullus, gravely, you and not a few of your colleagues, who have hoped that you could be thought ``friends of the people'' in overthrowing the commonwealth, opposed by a consul who is a friend of the people in truth, not in show. I challenge you, I summon you to a public assembly, I want to use the Roman people as judge. For if we look around at all that is grateful and pleasant to the people, we shall find nothing so popular as peace, as concord, as quiet. You handed over to me a state troubled with suspicion, in suspense from fear, thrown into confusion by your laws and assemblies and colony-plantings. You have shown the wicked hope, struck the loyal with fear, taken faith from the Forum and dignity from the commonwealth.

\ciceroSection{24} When in this commotion and disturbance of minds and affairs the consul's voice and authority shall suddenly have shone forth on the Roman people in such darkness, when he shall have shown that there is nothing to fear, no army, no band, no colonies, no sale of revenues, no new command, no decemviral kingdom, no second Rome and no other seat of empire under our consulship, and the highest tranquillity of peace and quiet, --- I shall have to fear, I suppose, that this splendid agrarian law of yours may seem the more popular!

\ciceroSection{25} But when I have laid bare the wickedness of your designs, the deceit of the law, and the snares laid against the Roman people itself by tribunes of the plebs who pretend to be its friends --- shall I, I suppose, be afraid that I shall not be allowed to take my stand against you in the assembly? --- particularly when I have determined and resolved to conduct my consulship in the only way it can be conducted with weight and freedom: not to seek a province, nor honour, nor any embellishment or perquisite, nor anything at all that can be hindered by a tribune of the plebs.

\ciceroSection{26} The consul declares before a most crowded Senate on the Kalends of January that, if the present condition of the commonwealth holds and no business arises which he cannot honourably escape, he will not go to a province. So I shall conduct myself in this magistracy, senators, that I may be able to coerce a tribune of the plebs angry with the commonwealth, and to despise one angry with me. For which reason, by the immortal gods, gather yourselves together, tribunes of the plebs; desert those by whom, if you do not look out, you will in a short time be deserted; conspire with us, agree with the loyal, defend the common commonwealth with common zeal and love. The commonwealth has many hidden wounds, many ruinous designs of wicked citizens; there is no danger from outside, no king, no people, no nation to be feared; the evil is enclosed, internal, domestic. Each one of us in his place must work to heal it; we all must wish to make it whole.

\ciceroSection{27} You err, if you suppose that the Senate approves what is said by me, but the people are of another mind. All who would have themselves safe will follow the authority of a consul released from cupidities, free from offences, careful in dangers, not timid in struggles. But if any of you is led by the hope that he can fill his sails with honour by tumultuous methods, first let him cease to hope it in my consulship; second, let him take me myself for a lesson, whom he sees, born from the equestrian order, made consul --- which path of life leads good men most easily to honour and dignity. And if you, senators, declare to me your zeal for defending the common dignity, I shall surely accomplish what the commonwealth most wants: that the authority of this order, which it had with our ancestors, may now after a long interval be seen restored to the commonwealth.
