% In Catilinam III (in-catilinam-3) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} The commonwealth, citizens, and the lives of you all, your goods, fortunes, wives and children, and this dwelling-place of a most distinguished empire, this most fortunate and most beautiful city, you see this day --- through the immortal gods' supreme love towards you, through my labours, my counsels, my dangers --- snatched from flame and sword and almost from the jaws of fate, preserved for you and restored.

\ciceroSection{2} And if those days on which we are saved are no less pleasant and bright to us than those on which we are born --- because the joy of safety is sure, the condition of birth uncertain, and because we are born without sense, while we are saved with pleasure --- surely, since the man who founded this city we have raised to the immortal gods by goodwill and by repute, he too ought to be in honour with you and your descendants who has saved this same city, founded and enlarged. For we have put out the fires that were already laid almost beneath the whole city, its temples, shrines, houses, and walls, and that surrounded them; the same swords drawn against the commonwealth we have blunted, and we have struck their points away from your throats.

\ciceroSection{3} Since these things have been brought to light in the Senate, laid open, found out by me, I shall now expound them briefly to you --- so that you, who do not know and are waiting, may learn how great they were, how manifest, by what method investigated and seized. To begin with, when Catiline a few days ago burst out of the city, leaving behind at Rome the partners of his crime --- the keenest leaders of this wicked war --- I always kept watch and provided, citizens, how in such great and so hidden snares we might be safe. For at the time when I was casting Catiline out of the city --- I do not now fear the odium of this word, since the more dreaded odium is that he went out alive --- but at the time when I wished him driven into exile, I supposed either that the rest of the conspirators' band would go out with him, or that those who stayed behind would be feeble and weak without him.

\ciceroSection{4} And I, when I saw that those whom I knew to be inflamed with the highest madness and crime were with us and had remained at Rome, spent on this all my days and nights: that I might perceive and see what they were doing, what they were contriving --- so that, since on account of the incredible magnitude of the crime my speech would have less weight with your ears, I might so seize the matter that, only when you saw with your eyes the very wickedness, then at last you might in your hearts provide for your safety. So when I learned that envoys of the Allobroges, for the sake of stirring up a Transalpine war and a Gallic alarm, had been solicited by Publius Lentulus, and that they were being sent into Gaul to their fellow-citizens, with letters and instructions to Catiline by the same route, and that Titus Volturcius had been added to them as a companion, and that to him letters had been given for Catiline --- I thought the opportunity offered me, what was most difficult and what I had always begged of the immortal gods, that the whole matter should be manifestly caught not only by me but also by the Senate and by you.

\ciceroSection{5} So yesterday I summoned to me Lucius Flaccus and Gaius Pomptinus, the praetors, men most brave and most loving toward the commonwealth; I laid out the business; I showed what I wished done. They, who in everything concerning the commonwealth feel splendidly and well, took up the matter without refusal and without any delay; and as evening came on, they made their way secretly to the Mulvian Bridge, and there in the nearest country houses they posted themselves in two parts so that the Tiber and the bridge stood between them. To the same place they themselves had brought, without anyone's suspicion, many brave men; and I had sent from the prefecture of Reate several picked young men, whose service I use constantly in the defence of the commonwealth, with swords.

\ciceroSection{6} Meanwhile, about the third watch of the night ended, when the envoys of the Allobroges with a great escort were beginning to enter on the Mulvian Bridge --- and Volturcius along with them --- an attack is made upon them; swords are drawn both by their party and by ours. The matter was known to the praetors only; it was unknown to the rest. Then by the intervention of Pomptinus and Flaccus the fight that had begun is settled. Whatever letters were in that escort were handed over to the praetors with the seals unbroken; the men themselves, having been seized, are brought to me as day was beginning to dawn. And the most wicked plotter of all these crimes, Cimber Gabinius, I called to me at once, while he yet suspected nothing; then in the same way Lucius Statilius was sent for, and after him Cethegus; but most slowly Lentulus came --- because, I suppose, the previous night he had stayed up beyond his custom over the writing of the letters.

\ciceroSection{7} When the highest and most distinguished men of this state, who, on hearing of the matter, had come together to me in numbers in the morning, judged that the letters should be opened by me before being brought to the Senate, lest, if nothing should be found, so great a commotion should seem rashly thrown by me upon the state --- I refused to do so, lest in a public danger I should not bring an unbroken matter before the public council. For, citizens, even if the things which had been brought to me had not been found, yet I did not think that in such great dangers to the commonwealth there was any excessive diligence on my part to be feared. The full Senate I quickly assembled, as you saw.

\ciceroSection{8} And meanwhile, on the prompt of the Allobroges, I sent at once Gaius Sulpicius the praetor, a brave man, to bring out from Cethegus's house any weapons that were there. From which he brought out a very large number of daggers and swords. I brought in Volturcius without the Gauls; on the Senate's order I gave him the public assurance; I urged him to declare without fear what he knew. Then he, when he had with difficulty recovered from a great fear, said that he had received from Publius Lentulus instructions and a letter for Catiline --- that Catiline should make use of the protection of slaves and approach the city with his army as soon as possible; this with the design that, when they had set fire to the city on every side, in the manner that had been laid out and apportioned, and when they had carried out an unbounded slaughter of citizens, he might be at hand both to receive the fugitives and to join himself with these urban leaders.

\ciceroSection{9} The Gauls, brought in, said that an oath and a letter to their nation had been given them by Publius Lentulus, Cethegus, and Statilius, and that it was prescribed to them by these and by Lucius Cassius that they should send cavalry into Italy as soon as possible; foot-troops they would not lack. That Lentulus had assured them, on the strength of the Sibylline books and the responses of the haruspices, that he was the third of those Cornelii to whom rule of this city and of empire must come: Cinna and Sulla had been before him. The same man had said that this was the year fated for the destruction of this city and empire --- the tenth year after the acquittal of the Vestals, the twentieth after the burning of the Capitoline.

\ciceroSection{10} They said also that there had been a dispute between Cethegus and the rest, because Lentulus and the others wished the slaughter to be done and the city burnt at the Saturnalia, while to Cethegus that seemed too long. And, not to be long, citizens, we ordered the tablets to be brought forward which were said to have been given by each. First we showed them to Cethegus: he recognized the seal. We cut the thread; we read. It was written in his own hand to the Senate and people of the Allobroges that he would do what he had promised their envoys; he asked that they likewise do what their envoys had received from him. Then Cethegus, who had a little earlier made some answer about the swords and daggers that had been seized at his house and had said he had always been a fancier of fine ironwork, when the letter had been read, broken down and dejected by his conscience, suddenly fell silent. Statilius was brought in, and recognized both the seal and his own hand. The tablets were read out, in nearly the same sense; he confessed. Then I showed the tablets to Lentulus and asked whether he recognized the seal. He nodded. ``It is, indeed,'' said I, ``a known seal, the image of your grandfather, that most distinguished man, who singularly loved his country and his fellow-citizens; whose seal even mute should have called you back from such a crime.''

\ciceroSection{11} Letters were read out in the same fashion to the Senate and people of the Allobroges. If he wished to say anything about these matters, I gave him the chance. And he at first denied; but afterwards, when the whole indication had been laid out and made public, he rose, asked the Gauls what they had to do with him, why they had come to his house --- and likewise asked Volturcius. When they had answered him briefly and steadily through whom and how often they had come to him, and had asked him whether he had not spoken with them at all about the Sibylline books, then suddenly, mad with crime, he showed how great is the force of conscience. For when he could have denied it, suddenly --- contrary to everyone's opinion --- he confessed. So he was failed not only by that genius and practice in speaking by which he had always been strong, but also, through the force of his manifest and detected crime, by the impudence by which he used to surpass everyone, and by his wickedness.

\ciceroSection{12} Volturcius, however, suddenly demands that the letters be brought forward and opened which he said had been given him by Lentulus to Catiline. And there Lentulus, although vehemently disturbed, recognized the seal and his hand. The letter, however, was without a name, but ran thus: ``Who I am you will know from the man whom I have sent to you. Take care to be a man, and consider into what place you have advanced. See what is now necessary to you, and take care to add the supports of all to yourself, even of the lowest.'' Gabinius, then brought in --- when he had begun at first to answer impudently --- in the end denied none of the things which the Gauls were charging.

\ciceroSection{13} And to me, citizens, although those proofs and indications of the crime seemed quite certain --- the tablets, the seals, the hands, in short the confession of each --- yet much more certain were these: the colour, the eyes, the looks, the silence. For so were they stupefied, so they were gazing at the ground, so they were now and then exchanging furtive glances with one another, that they seemed not so much to be informed against by others as to be informing against themselves. The indications laid out and made public, citizens, I consulted the Senate concerning the supreme matter of state, what was to be done. Speeches were made by the leading men, most keen and most courageous, which the Senate followed without any variance. And since the decree of the Senate has not yet been written out, from memory I will set out for you, citizens, what the Senate decreed.

\ciceroSection{14} First, thanks were given to me in the most ample words, that by my courage, counsel, and foresight the commonwealth had been freed of the greatest dangers. Then Lucius Flaccus and Gaius Pomptinus, the praetors, since I had used their brave and faithful service, were deservedly and rightly praised. And to a brave man, my colleague, praise was rendered, that he had removed those who were partners in this conspiracy from his own and from the commonwealth's counsels. And so they decreed: that Publius Lentulus, when he had abdicated his praetorship, should be handed over to custody; likewise that Gaius Cethegus, Lucius Statilius, Publius Gabinius --- all of whom were present --- should be handed over to custody. And the same was decreed against Lucius Cassius, who had demanded for himself the management of burning the city; against Marcus Ceparius, to whom it had been disclosed that Apulia had been allotted for stirring up the shepherds; against Publius Furius, who is from those colonists whom Lucius Sulla settled at Faesulae; against Quintus Annius Chilo, who together with this Furius had always been engaged in this stirring up of the Allobroges; against Publius Umbrenus, a freedman, by whom it was clear that the Gauls had been first brought to Gabinius. And the Senate used such mildness, citizens, that out of so great a conspiracy and so great a multitude of domestic enemies, with the punishment of nine of the most lost men, the commonwealth saved, it judged that the minds of the rest could be healed.

\ciceroSection{15} And besides, a thanksgiving was decreed to the immortal gods for their singular merit, in my name --- which since this city was founded has befallen me first as togate citizen --- and decreed in these words: ``because I had freed the city from fires, the citizens from slaughter, Italy from war.'' Which thanksgiving, if compared with other thanksgivings, has this difference: the others have been set up for affairs well managed, this one alone for the commonwealth saved. And what had to be done first has been done and concluded. For Publius Lentulus, although by the indications laid open, by his own confessions, by the judgment of the Senate, he had lost not only the right of a praetor but also of a citizen, yet abdicated his magistracy --- so that the religious scruple which had not stopped Gaius Marius, that most distinguished man, from killing Gaius Glaucia the praetor (about whom nothing had been decreed by name), of that scruple we should be free in punishing Publius Lentulus as a private man.

\ciceroSection{16} Now since you, citizens, hold seized and arrested the wicked leaders of the most criminal and dangerous war, you must judge that all of Catiline's forces, all his hopes and resources, with these dangers driven from the city, have collapsed. Whom indeed I, when I was driving out of the city, foresaw in my mind, citizens, that with Catiline removed I had nothing to dread from Publius Lentulus's drowsiness, nor from Lucius Cassius's fat, nor from Gaius Cethegus's frenzied recklessness. Catiline alone of all those was to be feared --- but only so long as he was contained within the walls of the city. He knew everything, he held the approaches to all; he could call upon, try, solicit; he dared. He had a mind apt for crime, and to that mind neither tongue nor hand was wanting. Already for definite tasks he had definite men picked out and apportioned. Nor did he, when he had given some commission, think it done: there was nothing he himself did not undertake, run to, watch over, labour at; cold, thirst, hunger he could endure.

\ciceroSection{17} This man, so keen, so audacious, so prepared, so cunning, so vigilant in crime, so diligent in lost causes, --- if I had not driven him from his domestic plot into a brigandage in camp (I shall say what I think, citizens) --- I should not easily have driven this so great mass of evil from your necks. He would not have appointed for us the Saturnalia, nor would he have so long beforehand declared the day of the commonwealth's destruction and fate, nor would he have allowed his seal, his letters --- the manifest witnesses of his crime --- to be caught. Which now, with him absent, have been so handled that no theft in a private house was ever found out so openly as this conspiracy in the whole commonwealth has been manifestly seized. But if Catiline had remained in the city up to this day --- although, as long as he was here, I met and obstructed all his designs --- yet, to put it most lightly, we should have had to fight with him, nor should we ever, with him in the city as an enemy, have freed the commonwealth of such great dangers in such great peace, in such great quiet, in such great silence.

\ciceroSection{18} And yet all these things, citizens, have been so administered by me that they seem to have been done and provided by the nod and counsel of the immortal gods. And we may attain this by conjecture, in that the steering of so great matters scarcely seems to have been able to be of human counsel; but in fact they have so present in these times brought us help and aid that we can almost see them with our eyes. For, to leave aside those things --- the torches seen at night-time from the west, and the burning of the sky, the strokes of lightning, the earthquakes; to leave aside the rest, which were so many in our consulship that the immortal gods seemed to be foretelling these things which now happen --- this at least, citizens, which I am about to say, must neither be passed over nor left aside.

\ciceroSection{19} For surely you remember that, in the consulship of Cotta and Torquatus, several things on the Capitoline were struck from heaven --- when both the images of the gods were thrown down and the statues of ancient men were cast down, and the bronze tablets of the laws were melted, and even the founder of this city, Romulus, was touched, whom you remember to have stood gilded on the Capitoline, small and at the breast, gaping for the wolfish dugs. At that time, when the haruspices had come together from all Etruria, they said that slaughter and fires and the destruction of the laws and a civil and domestic war and the fall of the whole city and empire was approaching, unless the immortal gods, by every means appeased, by their divinity should bend the very fates.

\ciceroSection{20} And so by their responses both games for ten days were then held, and no thing that pertained to the placating of the gods was left undone. The same men ordered an image of Jupiter to be made larger, and to be set up on a high place, and turned --- as it had not been before --- toward the east. And they said that they hoped, if that statue which you see should look upon the rising of the sun and the Forum and the Senate-house, then those designs which had been entered upon in secret against the safety of the city and the empire would be brought to light so as to be discerned by the Senate and the Roman people. And those consuls let out the contract for the setting up of that statue. But so great was the slowness of the work that neither under the previous consuls nor under us was it set up before this very day.

\ciceroSection{21} Here who can be so averse to the truth, so headlong, so deranged in mind, as to deny that all these things which we see, and especially this city, are administered by the nod and power of the immortal gods? When it had been so answered that slaughter, fires, the destruction of the commonwealth were being prepared --- and that by citizens (which then, on account of the magnitude of the crimes, seemed incredible to some) --- you have perceived that those things were not only thought of by wicked citizens but also undertaken. And does not this, indeed, seem so present that it appears to have been done by the nod of Jupiter Best and Greatest? --- that, when this very day in the morning by my order both the conspirators and their informers were being led through the Forum to the temple of Concord, at that very moment the statue was being set up. With it set up and turned toward you and toward the Senate, you saw that everything which was being designed against the safety of all had been brought to light and laid open by the Senate and by you.

\ciceroSection{22} For which they are the more deserving of hatred and punishment, who not only at your dwellings and roofs but even at the temples and shrines of the gods have tried to bring deadly and wicked fires. Whom if I should say I have resisted, I should claim too much for myself and not be borne: he, he, Jupiter resisted; he willed the Capitoline, he these temples, he the entire city, he you all, to be safe. With the immortal gods as my leaders, citizens, I undertook this mind and will and came upon these so great indications. Why, that Allobroges' solicitation --- that by Lentulus and the rest of the domestic enemies, in such madness, so great matters were entrusted to unknown and barbarous men, and that letters were committed to them --- this surely would never have been, unless from the immortal gods counsel had been wrenched from such great audacity. And what of this --- that men of Gaul, from a state ill at peace, the one nation which seems to remain that could make war on the Roman people and not be unwilling to, neglected the hope of empire and the greatest things offered them by patrician men, and preferred your safety to their own resources --- do you not think this was done divinely? particularly since they could have overcome us not by fighting but by keeping silent.

\ciceroSection{23} For which reason, citizens, since a thanksgiving has been decreed at all the couches of the gods, celebrate those days with your wives and children. For many honours have often been justly given and rendered to the immortal gods, but never surely more justly. For you have been snatched from a most cruel and most wretched destruction --- snatched without slaughter, without blood, without an army, without battle; togate, with me alone as togate leader and commander, you have conquered.

\ciceroSection{24} Recall, citizens, all civil dissensions, not only those which you have heard of but those which you yourselves remember and have seen. Lucius Sulla crushed Publius Sulpicius; Gaius Marius, the guardian of this city, and many brave men he partly drove from the state, partly killed. Gnaeus Octavius the consul drove his colleague from the city by arms: this whole place ran with heaps of bodies and the blood of citizens. Then Cinna with Marius prevailed: then with most distinguished men killed the lights of the state were extinguished. The cruelty of this victory Sulla afterwards avenged: there is no need to say with what diminution of citizens and what calamity to the commonwealth. Marcus Lepidus dissented from the most distinguished and most brave Quintus Catulus: not so much his own destruction brought sorrow to the commonwealth as that of others.

\ciceroSection{25} And yet all those dissensions were of the kind that pertained not to the destruction but to the alteration of the commonwealth. They did not wish that there be no commonwealth, but that they themselves be the leading men in the one that was; nor did they wish this city to be burned, but themselves to flourish in this city. And yet all those dissensions, of which none sought the ruin of the commonwealth, were of such a kind that they were settled not by the reconciliation of concord but by the slaughter of citizens. But in this one war, the greatest and cruellest within the memory of mankind --- such a war as no barbarism ever waged with its own race --- in which war this law was set up by Lentulus, Catiline, Cethegus, and Cassius, that all who could be safe with the city safe should be counted in the number of enemies, I have so borne myself, citizens, that you all be saved; and, when your enemies had thought there would survive only as many citizens as had stood out against an unbounded slaughter, and only as much of the city as the flame had been unable to reach, I have preserved both the city and the citizens entire and unharmed.

\ciceroSection{26} For so great deeds, citizens, I shall ask of you no reward of virtue, no badge of honour, no monument of praise, except the everlasting memory of this day. In your minds I wish all my triumphs, all the ornaments of honour, the monuments of glory, the badges of praise to be stored and placed. Nothing mute can please me, nothing silent, nothing of a kind that even the less worthy can attain. By your memory, citizens, my deeds will be nourished, by talk they will grow, by the monuments of letters they will grow old and gain strength; and that same day, which I hope will be eternal, has been propagated, I see, both for the safety of the city and for the memory of my consulship; and at one and the same time in this commonwealth two citizens have arisen, of whom the one bounded the limits of your empire not by regions of the earth but by regions of heaven, the other has saved the seat and dwelling-place of this empire.

\ciceroSection{27} But since the fortune and condition of the things I have done is not the same as that of those who waged foreign wars --- because I must live with those whom I have conquered and subdued, while they have left their enemies either killed or crushed --- it is yours, citizens, if rightly the deeds of the rest profit them, to provide that mine never harm me. For that the wicked and impious minds of most audacious men should not be able to harm you, I provided; that they not harm me, it is yours to provide. And yet, citizens, to me myself nothing now can be done of harm by them. For there is great defence in the loyal, which has been laid up for me in perpetuity; great dignity in the commonwealth, which will always silently defend me; great force of conscience, which they who neglect, in wishing to violate me, will give themselves away.

\ciceroSection{28} For there is in us such a spirit, citizens, that we not only yield to no man's audacity but even of our own motion always provoke the wicked. But if every assault of the domestic enemies, driven from you, shall turn itself upon me alone, you must look to it, citizens, on what condition you afterwards wish those to be who have offered themselves on behalf of your safety to all odium and to all dangers. As for me myself, what is there that can be added to the fruit of life --- particularly when in your honour and in the glory of virtue I see nothing higher to which it would please me to ascend?

\ciceroSection{29} This I shall surely accomplish, citizens: that the things I have done in the consulship I shall as a private man defend and adorn; so that, if any odium has been undertaken in saving the commonwealth, it may hurt the envious, and may benefit me for glory. In short, I shall so conduct myself in the commonwealth that I shall always remember what I have done, and take care that those things may seem to have been done by virtue, not by chance. You, citizens, since now it is night, having venerated that Jupiter who is the guardian of this city and of yourselves, depart to your houses; and those, although now the danger has been driven off, defend nevertheless equally as on the previous night by guards and watches. That this may not have to be done by you longer, and that you may be able to be in everlasting peace, I shall provide, citizens.
