% Pro C. Rabirio Perduellionis Reo (pro-rabirio-perduellionis) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Although, citizens, it is not my custom at the opening of a speech to give the reason for my having undertaken this or that defence --- because in every fellow-citizen's hour of danger I have always thought the bond of common citizenship a sufficiently just claim on me --- nevertheless in this defence, where the life, the reputation, and the entire fortunes of Gaius Rabirius are at stake, the reason for my obligation must, I think, be set out: because the cause that has seemed most just to me for taking up his defence ought also to seem to you the most just for acquitting him.

\ciceroSection{2} For though long-standing friendship, the man's standing, the claims of common humanity, and the whole habit of my life have urged me to defend Gaius Rabirius, what compelled me to do so with the utmost zeal was something else: the safety of the commonwealth, the duty of a consul, the consulship itself, which you committed to me together with the safety of the commonwealth. For it is not, citizens, the guilt of any crime, not the odium of his way of life, not, finally, any old, just, and weighty private enmities of citizens that have summoned Gaius Rabirius into peril of his life: it is that the supreme protection of the majesty and authority of Rome which we have inherited from our ancestors should be torn out of the commonwealth; that hereafter the authority of the Senate, the imperium of the consuls, the consensus of loyal men should count for nothing against the plague and ruin of the state. To overthrow these things, the old age, the weakness, the friendlessness of one man have been put to the test.

\ciceroSection{3} If, then, it is the part of a good consul, when he sees all the supports of the commonwealth being shaken and torn loose, to bring help to his country, to come to the aid of the common safety and fortunes, to call upon the loyalty of his fellow citizens, to count his own safety after the safety of all --- it is also the part of loyal and brave citizens, such as you have shown yourselves at every crisis of the commonwealth, to block off every road to sedition, to fortify the commonwealth's defences, to think the highest authority resides in the consuls and the highest counsel in the Senate, and to judge that whoever has followed those things deserves praise and honour rather than punishment and execution.

\ciceroSection{4} Hence the labour in defending him is mine in particular, but the will to save the man must be one I share with you. For you ought to consider, citizens, that within the memory of mankind no greater matter, none more dangerous, none more requiring foresight from you all has ever been undertaken by a tribune of the plebs, defended by a consul, or laid before the Roman people. For nothing else is at issue in this case, citizens, except that there shall hereafter be no public council in the commonwealth, no concord of loyal men against the fury and audacity of the wicked, no refuge and stronghold of safety in the last extremities of the state.

\ciceroSection{5} Since this is so, first --- as is necessary in so vast a struggle for life, reputation, and all one's fortunes --- I beg of Jupiter Best and Greatest and the rest of the immortal gods and goddesses, by whose help and aid this commonwealth is ruled far more than by the reason and counsel of men, that they grant peace and indulgence; I pray them to permit this day to have dawned both for preserving this man's safety and for setting the commonwealth on its foundations. Then I beseech and implore you, citizens, whose power comes nearest to the divine power of the immortal gods, since at one and the same moment the life of Gaius Rabirius, a most wretched and most innocent man, and the safety of the commonwealth are committed to your hands and your votes, to bring to a man's fortunes that mercy, and to the commonwealth's safety that wisdom, which are habitually yours.

\ciceroSection{6} Now, since you, Titus Labienus, by your strict timing have stood in the way of my own diligence, and have driven me out of the orderly span allotted for the defence into the narrow joint of half an hour, I shall comply --- with what is most unjust, the prosecutor's terms, and with what is most pitiful, an enemy's authority. Yet by this prescription of half an hour you have left me the part of an advocate but stripped me of the part of a consul: for half an hour will be just about enough for the defence, but too little for the protest.

\ciceroSection{7} Unless perhaps you think I ought to answer at length about the religious sites and groves which you say my client has profaned. On that charge nothing has ever been said by you except that the same charge was made against Gaius Rabirius by Gaius Macer. In which I marvel that you remember what Macer, his enemy, charged him with, but have forgotten what fair-minded jurors on oath delivered as their verdict.

\ciceroSection{8} Or shall a long speech be poured out about the alleged embezzlement and the burning of the records office? On that charge a kinsman of Gaius Rabirius, Gaius Curtius, was honourably acquitted in a most distinguished trial in keeping with his integrity; while Rabirius himself was never so much as named as a suspect, never mind brought to trial. Or must I answer more carefully about the sister's son? whom you said was killed by my client, when an excuse of a death in the family was being looked for in order to delay a trial. What is more likely than that he held his sister's husband dearer than his sister's son --- so much dearer that the one was robbed of life with utter cruelty, while for the other a stay of two days was sought in postponement of the trial? Or must more be said about other men's slaves held in defiance of the Lex Fabia (a law against the kidnapping of free men), or about Roman citizens flogged or killed in defiance of the Lex Porcia (a law forbidding the scourging of citizens), when Gaius Rabirius is honoured by all Apulia with such enthusiasm and by Campania with such singular goodwill, when not only individuals but almost whole regions have come together to ward off the danger to him --- regions extending considerably wider than the bare name and limits of his neighbourhood would call for? And what need have I to prepare a long speech against what is set down in the same indictment of fines, that he has spared neither his own chastity nor anyone else's?

\ciceroSection{9} I rather suspect that this is precisely why a half hour has been allotted me by Labienus --- so that I should not say more about chastity. So you see that for these charges, which require an advocate's diligence, that half hour of yours has been more than long enough. The other part, on the killing of Saturninus, you wished to be too narrow and constricted: but that part calls not for an orator's talent but for a consul's protection, and demands it.

\ciceroSection{10} For as to the trial for high treason, the abolition of which you keep accusing me of, that charge is mine, not Rabirius's. And I would to heaven, citizens, that I had been either the first or the only man in this commonwealth to abolish it! Would that this thing he wishes to make my reproach were the proper testimony to my praise. For what could be wished for that I should rather have than to have driven the executioner from the Forum and the cross from the Field of Mars in my consulship? But that praise belongs first to our ancestors, citizens, who, when they had expelled the kings, kept no trace of the cruelty of kingship in the free people; then to the many brave men who wished your liberty to be guarded not by the sharpness of penalties but by the mildness of laws.

\ciceroSection{11} On which then of us, Labienus, is the title ``friend of the people'' to fall: on you, who think that against Roman citizens at the very public assembly the executioner and the chains ought to be brought, who order that on the Field of Mars at a comitia centuriata, in a place taken by the auspices, a cross be planted and set up for the punishment of citizens? or on me, who forbid the assembly to be polluted by contact with the executioner, who say that the Forum of the Roman people is to be cleansed of those traces of unspeakable crime, who maintain that the assembly should remain pure, the Field of Mars sacred, the body of every Roman citizen inviolate, the right of liberty whole?

\ciceroSection{12} A fine ``friend of the people'' you make for a tribune of the plebs --- guardian and defender of right and liberty! The Lex Porcia took the rods from the body of every Roman citizen; this man in his mercy has brought back the lash. The Lex Porcia tore the citizen's liberty from the lictor; Labienus, ``friend of the people,'' has handed it to the executioner. Gaius Gracchus passed a law that no judgment should be passed on the life of a Roman citizen without your order; this ``friend of the people'' has compelled, without your order, two officers (the duumviri) not to try a Roman citizen but, with no case stated, to condemn a Roman citizen on a capital charge.

\ciceroSection{13} You make mention to me of the Lex Porcia, of Gaius Gracchus, of these men's love of liberty, of any popular man you please --- you, who have tried to violate the liberty of this people, to test its mildness, to alter its discipline, with punishments unheard of and a savagery of language never heard before? For these formulas of yours, which delight you, the gentle and popular man --- ``Go, lictor, bind his hands'' --- belong not to this liberty and this mildness; they belong not even to Romulus or Numa Pompilius. They are the dirges of torture out of Tarquin, that most arrogant and most cruel of kings, which you, the mild and popular man, recall with such pleasure: ``Veil his head, hang him on the unlucky tree.'' Words, citizens, which long ago in this commonwealth were buried not only in the darkness of antiquity but in the light of liberty.

\ciceroSection{14} Could you really suppose, if that proceeding of yours were popular, if it had any element of fairness or of right in it, that Gaius Gracchus would have left it untouched? Did, I suppose, your uncle's death cause you a sharper grief than his brother's death caused Gaius Gracchus, and is the death of an uncle whom you never saw bitterer to you than was to him the death of the brother with whom he had lived in perfect harmony? With as much right do you avenge an uncle's death as he would have pursued a brother's, if he had wished to act on this principle? And did this Labienus of yours, your uncle, whoever he may have been, leave behind in the Roman people as much regret as Tiberius Gracchus had left? Is your filial duty greater than Gaius Gracchus's, or your spirit, or your judgment, or your resources, or your authority, or your eloquence? --- which, if they had been smallest in him, would still in comparison with your gifts have been counted greatest.

\ciceroSection{15} But when in all these things Gaius Gracchus surpassed all men, what gulf, then, do you suppose lies between him and you? And yet Gaius Gracchus would have died first of the most cruel death a thousand times over before he would have allowed the executioner to stand at his assembly --- the man whom the censors' laws sought to deprive not of the Forum only but even of this sky and air and dwelling-place in the city. This man dares to call himself ``friend of the people'' and to call me a stranger to your interests, when he has hunted out all his harshness of penalty and language not from the memory of yourselves and your fathers but from the records of annals and from the commentaries of kings, while I with all my powers, all my counsels, all my words and deeds have fought against and resisted his cruelty? Unless perhaps you wish for yourselves the condition which slaves, if they had not the hope of liberty set before them, could in no way bear.

\ciceroSection{16} A miserable thing is the disgrace of a public conviction, miserable the confiscation of one's goods, miserable exile; yet in every such calamity some trace of liberty is kept. And if at last death is set before us, let us die in liberty: let the executioner, the veiling of the head, the very name of the cross be far not only from the body of Roman citizens but even from their thought, from their eyes, from their ears. For the actual happening and endurance, but also the prospect, the expectation, the very mention of all these things, is unworthy of a Roman citizen and a free man. Or shall the kindness of their masters by the touch of one rod free our slaves from the fear of all such punishments, and shall not our deeds, our years lived, your honours, free us from the rods, from the hook, from the very terror of the cross?

\ciceroSection{17} Therefore, Labienus, I confess and even profess and proclaim it openly: by my counsel, my courage, my authority you have been driven back from that cruel, lawless proceeding of yours --- a proceeding that was not tribunician but kingly. In which proceeding, although you set at naught all the precedents of our ancestors, all our laws, all the authority of the Senate, all the religious rites and the public laws of the auspices --- nevertheless from me on these points in this so brief allotment of time you shall not hear: ample time will be given us for that argument.

\ciceroSection{18} Now I shall speak of the charge concerning Saturninus and the death of your most distinguished uncle. You charge that Lucius Saturninus was killed by Gaius Rabirius. But that Gaius Rabirius previously, on the testimony of many witnesses and with Quintus Hortensius defending him most copiously, showed to be false. For my own part, were the matter open to me, I should take up this charge, I should acknowledge it, I should confess it. Would that the case allowed me the latitude to be able to declare this --- that by Gaius Rabirius's hand Lucius Saturninus, an enemy of the Roman people, was put to death! --- That shouting of yours does not move me, but consoles me, since it shows that there are some unschooled citizens, but not many. Believe me, this Roman people which keeps silence here would never have made me consul if it thought that it could be flustered by your shouting. How much fainter the cry has grown already! Why don't you check the voice that betrays your folly, that bears witness to your few numbers!

\ciceroSection{19} I should gladly, I say, confess --- if I could do so truthfully, or even if it were open to me --- that Lucius Saturninus was killed by the hand of Gaius Rabirius, and I should think that deed a most splendid one. But since I cannot do that, I will confess what for praise will count for less, but for the charge no less. I confess that for the killing of Saturninus Gaius Rabirius took up arms. What now, Labienus? what graver confession do you await from me, or what greater charge against my client? unless you suppose there is some difference between the man who killed and the man who was armed for the purpose of killing. If it was a crime to kill Saturninus, arms cannot have been taken up against Saturninus without crime. If you grant that arms were taken up by right, you must grant that he was killed by right.

\ciceroSection{20} A decree of the Senate is passed, that the consuls Gaius Marius and Lucius Valerius should call upon the tribunes of the plebs and the praetors as they thought fit, and should see to it that the imperium and the majesty of the Roman people be preserved. They call upon all the tribunes of the plebs except Saturninus, and all the praetors except Glaucia: those who wished the commonwealth safe they bid take up arms and follow them. All obey: from the temple of Sancus and the public arsenals arms are issued to the Roman people, with the consul Gaius Marius distributing them. And here, to leave the rest aside, I ask you, Labienus, on your own account: when Saturninus was holding the Capitoline armed, when there were with him Gaius Glaucia, Gaius Saufeius, even that ex-shackled Gracchus from the workhouse --- I shall add, since you so wish, your uncle Quintus Labienus too --- and in the Forum on the other side were the consuls Gaius Marius and Lucius Valerius Flaccus, and behind them the entire Senate (and that very Senate which even you yourselves, who now bring these patres conscripti who are now in office into hatred for the easier disparagement of the present body, used to praise); when the equestrian order --- and what knights, immortal gods! the knights of our fathers' generation, who at that time held a great share of the commonwealth and the entire dignity of the courts --- when all men of all orders who thought their own safety was bound up in the safety of the commonwealth had taken up arms: what, finally, was Gaius Rabirius to do?

\ciceroSection{21} On your own account, I say, Labienus, I ask. When the consuls had called to arms by decree of the Senate, when armed Marcus Aemilius the princeps senatus had taken his stand in the Comitium, who, since he could scarcely move, did not think the slowness of his feet would be an obstacle for pursuit but an obstacle for flight; when, finally, Quintus Scaevola, broken with old age, ruined by disease, maimed and crippled in every limb, leaning on a spear, was showing both the strength of his spirit and the weakness of his body; when Lucius Metellus, Servius Galba, Gaius Serranus, Publius Rutilius, Gaius Fimbria, Quintus Catulus, and all who were then men of consular rank had taken up arms for the common safety; when all the praetors, the entire nobility and youth came running --- the two Domitii (Gnaeus and Lucius), Lucius Crassus, Quintus Mucius, Gaius Claudius, Marcus Drusus, when all the Octavii, the Metelli, the Iulii, the Cassii, the Catos, the Pompeii, when Lucius Philippus, Lucius Scipio, when Marcus Lepidus, when Decimus Brutus, when this very Publius Servilius, under whose command, Labienus, you served, when this Quintus Catulus here, then quite a young man, when this Gaius Curio, when, finally, all the most illustrious men were with the consuls: what then was it fitting for Gaius Rabirius to do? to lurk shut in and hidden in some out-of-the-way place, and to cover up his cowardice with walls and the protection of darkness? or to make for the Capitoline and there join your uncle and the rest who were taking refuge in death because of the foulness of their lives? or to come together with Marius, Scaurus, Catulus, Metellus, Scaevola, with all loyal men, in a partnership not of safety only but also of danger?

\ciceroSection{22} You yourself, Labienus, what would you do in such a case and at such a moment? When the calculation of cowardice was driving you to flight and into hiding, when the wickedness and madness of Lucius Saturninus was summoning you to the Capitoline, when the consuls were calling you to the safety and freedom of your country, what authority, then, what voice, whose lead would you most wish to follow, whose command to obey? ``My uncle was with Saturninus,'' he says. What about your father --- with whom was he? What about your relatives, the Roman knights? What about your whole prefecture, your district, your neighbourhood? What about the entire Picene country --- did it follow tribunician madness or consular authority?

\ciceroSection{23} I assert this for my part: what you now publicly declare about your uncle, no man as yet has confessed about himself; no one, I say, has been found so abandoned, so lost, so deserted not just by all honour but even by the pretence of honour, as to confess that he was on the Capitoline with Saturninus. But your uncle was. Granted; let him have been --- and have been so under no compulsion, under no despair of his own affairs, driven by no domestic wounds; let intimacy with Lucius Saturninus have led him to put friendship before country: was that any reason why Gaius Rabirius should have deserted the commonwealth, should have failed to appear in that armed assembly of loyal men, should have refused to obey the voice and command of the consuls?

\ciceroSection{24} But we see that in the nature of things three courses lay open: to be with Saturninus, or to be with the loyal, or to lie hidden. To lie hidden was as good as the most disgraceful death; to be with Saturninus was madness and crime; courage and honour and shame compelled a man to be with the consuls. Is this what you put on trial as a charge --- that Gaius Rabirius was with those whom he would have been most insane to attack and most disgraced to abandon? But Gaius Decianus, of whom you so often make mention, was condemned because, in prosecuting Publius Furius --- a man marked with every brand of dishonour --- with the utmost zeal of all loyal men, he dared to lament Saturninus's death in a public assembly; and Sextus Titius was condemned for keeping a portrait of Lucius Saturninus in his house. The Roman knights ruled in that trial that the man was a wicked citizen and not to be tolerated in the state who, by displaying the portrait of a man as hostile and seditious as that, would either honour his death, or stir up regret in the unschooled by an appeal to pity, or signal his own intent to imitate the wickedness.

\ciceroSection{25} And so it seems strange to me, Labienus, where you came by that portrait of yours; for after Sextus Titius was condemned, no one was found bold enough to keep one of them. Had you heard of this, or could you for your years have known it, you would never have brought to the Rostra and the public assembly the portrait that, posted up in his house, brought ruin and exile on Sextus Titius; nor would you ever have steered your craft onto those rocks where you might see the wreck of Sextus Titius's vessel and the foundering of Gaius Decianus's fortunes. But in all this you slip from inexperience. You have undertaken a cause older than your own memory, a cause which died before you were born; and a cause in which you yourself would assuredly have stood, had your years allowed it, you bring to trial.

\ciceroSection{26} Or do you not see, first, what men and what kind of men, now dead, you charge with the highest crime, and next, how many of those who are still alive you summon by the same charge into utmost danger of their lives? For if Gaius Rabirius committed a capital fraud because he bore arms against Lucius Saturninus, the years he was then will at least carry with them some plea against the danger to him; but Quintus Catulus, the father of this man here, in whom were the highest wisdom, surpassing virtue, singular humanity; Marcus Scaurus, of that gravity, that judgment, that prudence; the two Mucii, Lucius Crassus, Marcus Antonius (who was at that time outside the city with a guard), whose counsels and talents in this city were by far the greatest, and the rest, of equal dignity, the guardians and steersmen of the commonwealth --- how shall we defend these dead men?

\ciceroSection{27} What shall we say of those most honourable men and best of citizens, the Roman knights, who then together with the Senate defended the safety of the commonwealth? what of the tribunes of the treasury and of the men of every other order who at that time took up arms for the common liberty? But why do I speak of all those who obeyed the consuls' command? what shall the reputation of the consuls themselves come to? Lucius Flaccus --- a man always most diligent in the commonwealth, in the holding of magistracies, in the priesthood and in the rites at which he presided --- shall we condemn him, dead, of unspeakable crime and of parricide? Shall we add to this stain and infamy of death even the name of Gaius Marius? Gaius Marius, whom we may truly call father of the country, parent, I say, of your liberty and of this very commonwealth --- shall we condemn him, dead, of unspeakable crime and of parricide?

\ciceroSection{28} Indeed, if for Gaius Rabirius, because he went to arms, Titus Labienus thought a cross should be planted in the Field of Mars, what punishment, then, will be devised for the man who summoned him? And if a pledge was given to Saturninus, as you say very often, it was not Gaius Rabirius but Gaius Marius who gave it, and the same man who violated it, if he failed to keep it. What pledge, Labienus, could be given without a decree of the Senate? Are you so much a stranger in this city, so ignorant of our discipline and custom, that you do not know these things, that you seem to be travelling in a foreign state rather than holding a magistracy in your own?

\ciceroSection{29} ``What harm can these things now do to Gaius Marius,'' he says, ``since he is without sense and without life?'' Is that so, indeed? Would Gaius Marius have lived through such labours and such dangers if he had thought of himself and of his glory in hope and spirit no further than the limits of life demanded? Did he, I suppose, think, when he had routed innumerable hosts of enemies in Italy and freed the commonwealth from siege, that all his goods would die with him? It is not so, citizens; nor does any one of us conduct himself in the dangers of the commonwealth with praise and courage who is not led on by the hope and the fruit of posterity. And so, for many other reasons the minds of good men seem to me to be divine and eternal, but most of all because the spirit of every best and wisest man so foreknows of the future that it seems to look at nothing but what is everlasting.

\ciceroSection{30} Wherefore I, calling to witness the spirit of Gaius Marius and of the rest of the wisest and bravest citizens, which seem to me to have passed from the life of men into the worship and sanctity of the gods, declare that for their reputation, glory, and memory, no less than for the shrines and sanctuaries of my fathers, I think I must do battle. And were I obliged to take up arms in defence of their renown, I should take them up no less briskly than they took theirs in defence of the common safety. For nature has marked out a slender course of life for us, citizens, but a measureless one of glory. So if we shall do honour to those who have already departed this life, we shall leave for ourselves a juster condition for our own deaths. But if you neglect those, Labienus, whom we can no longer see, do you think no thought ought to be taken even for those whom you do see?

\ciceroSection{31} I assert that there is no man of all those who were at Rome on that day --- the day you bring to trial --- and was then of military age, who did not take up arms, who did not follow the consuls. All those whose actions on that day you can guess from their age are by you summoned by name on Gaius Rabirius's capital charge. ``But Rabirius killed Saturninus.'' Would that he had done so! I should not be pleading against the punishment, but demanding the reward. For if to Scaeva, the slave of Quintus Croto, who killed Lucius Saturninus, freedom was given, what reward would have been fitting to give a Roman knight? And if Gaius Marius, because he ordered the pipes that supplied water to the temple and seat of Jupiter Best and Greatest to be cut, because on the Capitoline slope of wicked citizens . . .  [text breaks off]

\ciceroSection{32} . . . [text breaks off]. And so the Senate in considering that case under my management was no more diligent or harsh than were all of you when, with your minds, your hands, your voices, you rejected the partition of the world and that very Campanian land itself.

\ciceroSection{33} I, the same man, cry out, declare, denounce, what the very author of this prosecution does. There is no king left, no people, no nation that you should fear; there is no foreign, no external evil that can creep into this commonwealth. If you wish this state to be immortal, if you wish this empire to be eternal, if you wish glory to remain everlasting, then we must guard against our own desires, against turbulent men greedy for revolution, against domestic evils, against plots from within.

\ciceroSection{34} Against these evils, however, your ancestors left you a great defence --- that voice of the consul: ``those who wish the commonwealth to be safe.'' Favour that voice, citizens; do not by your verdict take from me, do not snatch from the commonwealth, the hope of liberty, the hope of safety, the hope of dignity.

\ciceroSection{35} What should I do, if Titus Labienus had carried out a slaughter of citizens as Lucius Saturninus did, if he had broken open the prison, if he had seized the Capitoline with armed men? I should do the same as Gaius Marius did: refer the matter to the Senate; urge you to defend the commonwealth; armed myself, with you armed, stand against him. As it is, since there is no suspicion of arms, I see no weapons, no violence, no slaughter, no siege of the Capitoline and the citadel, but a pernicious accusation, a bitter trial, the whole business taken up by a tribune of the plebs against the commonwealth, I have thought I must call upon you not to arms but to the votes against an assault on your majesty. And so now I beg, beseech, urge you all. It is not the custom that a consul . . .  [text breaks off]

\ciceroSection{36} . . . [text breaks off]. He who has received face on, for the commonwealth, these scars and marks of valour, fears now lest he take a wound to his reputation; he whom the assaults of enemies could never move from his place, now stands in dread of the assault of his fellow-citizens, to which he must of necessity yield.

\ciceroSection{37} He no longer asks of you the ability to live well but to die honourably; he labours not so much to enjoy his own house as not to be deprived of his ancestors' tomb. He asks and beseeches you for nothing else but that you not deprive him of a lawful funeral and a death at home; that you suffer the man who never shunned any danger of death for his country to be allowed to die in his country.

\ciceroSection{38} I have spoken to the time set me by the tribune of the plebs; from you I ask and beg that you account this defence of mine faithful to a friend in danger and consular in loyalty to the commonwealth.

\ciceroSection{38b} . . . and was by far the dearest both to the entire Roman people and most of all to the equestrian order.
