% Pro Lege Manilia (De Imperio Cn. Pompei) (pro-lege-manilia)
% English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.
%
% Cicero's first political speech, delivered as praetor in 66 BC,
% supporting the Manilian law that gave Pompeius the command against
% Mithridates. The speech is structured as: kind of war, magnitude of
% war, choice of general -- with the third part the longest, an
% encomium of Pompey.

\ciceroSection{1} Although the sight of you in numbers has always been by far most pleasant to me, citizens, and this place has seemed the most ample for action and the most adorned for speaking, yet from this approach to praise, which has always been most open to the best men, not my own will up to now but the reckonings of my life undertaken from my entry into manhood have hindered me. For when before I dared not yet touch the authority of this place, and resolved that nothing should be brought hither save what was finished by talent and elaborated by industry, I thought that all my time should be spent on the times of my friends.

\ciceroSection{2} And so this place was never empty of those who defended your case, and my labour, busied chastely and uprightly in the perils of private men, has from your judgement attained the most ample fruit. For when, on account of the postponement of the elections, three times I was returned as first praetor by all the centuries, I easily understood, citizens, both what you judged of me and what you laid down for others. Now, when there is in me as much authority as you have wished there to be by giving me honours, and as much faculty for action as the almost daily practice of speaking from forensic use could bring to a watchful man, surely both what authority I have I shall use among those who have given it to me, and what I can attain in speaking I shall show especially to those who have judged that fruit also should be granted to that thing by their judgement.

\ciceroSection{3} And this above all I see I rightly should rejoice in: that in this --- to me unfamiliar --- manner of speaking from this place, such a case has been laid before me in which speech can be wanting to no man. For I must speak of the singular and outstanding virtue of Gnaeus Pompeius. But of this speech it is harder to find the end than the beginning. So that I must seek not so much abundance as measure in speaking.

\ciceroSection{4} And that my speech may set out from where the whole case is drawn: a grave and perilous war is being brought against your revenues and your allies by two most powerful kings, Mithridates and Tigranes, of whom the one being left, the other being provoked, each thinks that an opportunity has been offered him for seizing Asia. To Roman knights, most honourable men, letters are brought daily out of Asia, of whom great affairs are being carried on, occupied in working your revenues; who, on the connection which I have with that order, have brought to me the case of the commonwealth and the perils of their own affairs:

\ciceroSection{5} that several villages of Bithynia (which is now your province) have been burned out; that the kingdom of Ariobarzanes (which borders on your revenues) is wholly in the power of the enemy; that Lucius Lucullus, with great deeds done, is leaving that war; that he who has succeeded him is not prepared enough to administer such a war; that one general is being demanded and sought by all your allies and citizens for that war; that this same one alone is feared by the enemy --- besides, none.

\ciceroSection{6} You see what the cause is. Now consider yourselves what must be done. First, I think I must speak about the kind of war; then about the size of it; then about choosing a general. The kind is of that war which most ought to rouse and inflame your minds to the zeal of pursuing. In which the glory of the Roman people is at stake, which has been handed down to you by your ancestors --- great in all things, the highest in military matters. The safety of allies and friends is at stake, for which your ancestors waged many great and grave wars. The most certain and greatest revenues of the Roman people are at stake, which lost, you will lack both the ornaments of peace and the supports of war. The goods of many citizens are at stake, for which the commonwealth and they themselves must be consulted by you.

\ciceroSection{7} And since you have always been more eager for glory and greedy for praise than other peoples, that stain conceived in the previous Mithridatic war must be wiped out by you, which has now sunk deep and grown old in the name of the Roman people --- that he, who in one day in all Asia, in so many cities, by one message and one signal, ordered all Roman citizens to be killed and slaughtered, has not only up to now received no punishment worthy of his crime, but from that time has now reigned for the twenty-third year, and so reigns that he wishes not to hide himself in the lurking-places of Pontus or of Cappadocia, but to come forth from his ancestral kingdom and to be at large in your revenues, that is, in the light of Asia.

\ciceroSection{8} For up to now our generals have so contended with that king that they brought back from him the marks of victory, not victory itself. Lucius Sulla triumphed; Lucius Murena triumphed over Mithridates --- two bravest men and highest generals; but they so triumphed that the man, driven and overcome, kept reigning. But yet praise must be given those generals for what they did; pardon must be granted for what they left unfinished, because from that war the commonwealth recalled Sulla into Italy, Sulla recalled Murena.

\ciceroSection{9} But Mithridates devoted all the rest of the time not to forgetting the old war but to preparing a new one. Who afterwards, when he had built and equipped the largest fleets and gathered very great armies from whatever peoples he could, and was pretending to bring war on the Bosporans, his neighbours, sent envoys and letters all the way to Spain to those leaders with whom we were then waging war, that, when in two most distant and most diverse places by one counsel war was waged on land and sea by two enemy forces, you, distracted by the doubtful contention, should fight for the empire.

\ciceroSection{10} But yet the danger of one part --- the Sertorian and Spanish, which had much more strength and steadiness --- was driven off by Gnaeus Pompeius's divine counsel and singular virtue; in the other part the matter was so handled by Lucius Lucullus, the highest man, that the great and distinguished beginnings of his deeds seem to be ascribed not to his luck but to his virtue, while these last things which have lately happened seem to be ascribed not to fault but to fortune. But of Lucullus I shall speak in another place, and so I shall speak, citizens, that neither true praise should seem to be drawn from him by my speech, nor false praise feigned upon him.

\ciceroSection{11} Of the dignity and glory of your empire, since that is the beginning of my speech, see what spirit you ought to take up. Our ancestors often waged wars for our merchants or shippers too injuriously treated. With so many thousands of Roman citizens killed by one message and at one time --- with what spirit, after all, ought you to be? Because legates had been addressed too haughtily, your fathers willed Corinth, the light of all Greece, to be put out. Will you suffer that king to be unavenged who killed an envoy of the Roman people, a man of consular rank, after torturing him with chains and with floggings and with every punishment? They could not bear that the liberty of Roman citizens should be lessened. Will you neglect that life is taken away? They pursued the right of legation violated by a word. Will you leave a legate killed by every punishment?

\ciceroSection{12} See lest, as it was most beautiful for them to hand down to you so great a glory of empire, so it should be most foul for you not to be able to keep and preserve what you have received. What? That the safety of allies is called into the highest peril and crisis --- with what spirit, after all, ought you to bear that? King Ariobarzanes, an ally and friend of the Roman people, has been driven out of his kingdom. Two kings hang over all Asia, most hostile not only to you but also to your allies and friends. But all states throughout all Asia and Greece are compelled, on account of the size of the peril, to await your help: yet to demand a sure general from you (especially since you have sent another) they neither dare nor reckon they can do it without the highest peril.

\ciceroSection{13} For they see and feel this same as you: that there is one man in whom the highest things are all, and that he is near, by which they the more grievously are without him; whose very coming and name (although he came for the maritime war) yet they understand has held back and slowed the onsets of the enemy. These men, since it is not lawful to speak freely, ask you silently that you should also reckon them worthy whose safety you should commend to such a man, and this even the more because we send to other provinces such men with command that, even if they defend them from the enemy, yet their very coming into the cities of allies differs not much from a hostile storming. They were hearing of him before; now present they see him with such temperance, such gentleness, such humanity that those seem to be the most blessed with whom he most long stays.

\ciceroSection{14} Wherefore, if for the sake of allies, with no wrong of their own, our ancestors waged wars with Antiochus, with Philip, with the Aetolians, with the Carthaginians --- how much the more zealously does it become you, provoked by wrongs, to defend the safety of the allies together with the dignity of your empire, especially when it is a matter of your greatest revenues? For the revenues of other provinces, citizens, are so great that with them we can scarcely be content for keeping the provinces themselves. But Asia is so rich and fertile that by the abundance of its fields and the variety of its fruits and the size of its pastures and the multitude of those things which can be exported, it easily surpasses all lands. And so this province, citizens, if you wish to keep both the usefulness of war and the dignity of peace, must be defended not only from calamity but even from the fear of calamity.

\ciceroSection{15} For in other things, when calamity comes, then the loss is taken. But in the revenues, not only the coming of evil but even the very fear brings calamity. For when the forces of enemies are not far off, even if no breach has been made, yet the cattle are abandoned, the tilling of the field is deserted, the sailing of merchants comes to rest. So neither from the harbour, nor from the tithes, nor from the pasture-tax can revenue be kept; wherefore often the harvest of a whole year is lost by one rumour of peril and one terror of war.

\ciceroSection{16} What spirit then do you reckon those have either who pay the revenues to us, or those who work and exact them, when two kings with the largest forces are near at hand; when one running up of cavalry can in a very short time take away the revenue of a whole year; when the publicans reckon they hold at great peril the largest households which they have in the salt-pans, in the fields, in the harbours, and in the watch-stations? Do you suppose you can enjoy these things, unless you have kept those who are a fruit to you freed not only (as I said before) from calamity but even from the fear of calamity?

\ciceroSection{17} And not even this must be neglected by you which I had set as last for myself when I was about to speak about the kind of war: that this matter has to do with the goods of many Roman citizens, of whom by your own wisdom, citizens, account must be diligently taken. For both the publicans, most honourable and most adorned men, have brought their reckonings and resources into that province, of whom themselves the affairs and fortunes for their own sake ought to be a care to you. For if we have always reckoned that the revenues are the sinews of the commonwealth, surely that order which works them we shall rightly say is the support of the rest of the orders.

\ciceroSection{18} Then out of the rest of the orders busy and industrious men do business, some themselves in Asia, for whom in their absence you ought to provide; some have great sums of money placed in that province. So it is of your humanity to keep a great number of citizens from calamity; of your wisdom to see that the calamity of many citizens cannot be sundered from the commonwealth. For first, this is of small worth: that you, the publicans, should afterwards by victory recover the revenues lost. For neither will the same have means to take them on, on account of the calamity, nor will others have the will, on account of fear.

\ciceroSection{19} Then what the same Asia and that same Mithridates taught us at the start of the Asiatic war --- that, surely, taught by calamity, we ought to keep in memory. For then, when many in Asia lost great fortunes, we know that at Rome credit fell, with payments hindered. For it cannot be that many in one state should lose their property and fortunes without dragging more with them into the same calamity. From this peril keep the commonwealth. For (believe me what you yourselves see) this credit and this reasoning of moneys, which is at large at Rome, which is in the forum, is entangled with those Asiatic moneys and clings together. Those things cannot fall without these, shaken by the same motion, falling. Wherefore see whether you must hesitate to lean with all zeal toward that war in which the glory of your name, the safety of allies, the greatest revenues, the fortunes of very many citizens joined with the commonwealth, are being defended.

\ciceroSection{20} Since I have spoken about the kind of war, now of its size I shall say a few things. For this can be said: that the kind of war is necessary so that it must be waged, but is not so great that it must be greatly feared. In which one must most labour, lest perhaps those things which must be most diligently provided for shall seem to you must be despised. And that all may understand that I attribute to Lucius Lucullus as much praise as is owed to a brave man and a wise man and a great general, I say that at his coming the largest forces of Mithridates were equipped and drawn up in all things; and that the most distinguished and most friendly city of Asia to us, that of the Cyziceans, was pressed by the king himself with the greatest multitude and most vehemently besieged; which Lucius Lucullus by virtue, persistence, counsel freed from the highest perils of the siege.

\ciceroSection{21} By the same general a great and equipped fleet, which under the Sertorian leaders, inflamed with zeal and hatred, was being borne against Italy, was overcome and sunk; great forces of the enemy besides were destroyed in many battles; and the Pontus, which before had been closed to the Roman people from every approach, was opened to our legions; Sinope and Amisus, in which towns were homes of the king equipped and stuffed with all things, and very many other cities of Pontus and Cappadocia, were captured by one approach and coming. The king, despoiled of his ancestral and grandfather's kingdom, betook himself as suppliant to other kings and other peoples. And all these things were done with the allies of the Roman people kept safe and the revenues whole. Praise enough this is, I think; and so, citizens, that you may understand this: that by none of those who carp at this law and case has Lucius Lucullus been likewise praised from this place.

\ciceroSection{22} Perhaps now it will be asked in what manner, since these things are so, the rest of the war can be great. Learn, citizens; for not without cause this seems to be asked. First, Mithridates fled from his own kingdom in such a way as that famous Medea once is said to have fled from the same Pontus, of whom they tell that in her flight she scattered the limbs of her brother in those places by which her parent was pursuing her, that the gathering of them dispersed and the parent's grief should slow the speed of pursuing. So Mithridates fleeing left in Pontus the greatest store of gold and silver and of all most beautiful things, which he had received from his ancestors and had himself in the previous war heaped up in his kingdom from all Asia plundered. While our men were diligently gathering all these, the king himself escaped from their hands. So grief slowed him in the zeal of pursuing; them, joy.

\ciceroSection{23} Tigranes, king of Armenia, received him in that fear and flight, and confirmed him distrustful of his affairs and raised him afflicted and revived him ruined. Into whose kingdom afterwards, when Lucius Lucullus came with the army, more peoples too were stirred against our general. For fear had been thrown into those peoples whom the Roman people had never thought ought either to be provoked by war or to be tried; there was also another grave and vehement opinion which had pervaded the minds of the barbarian nations: that for the sake of plundering a most well-off and most religious shrine our army had been brought into those shores. So many and great peoples were being stirred up by a kind of new terror and fear. But our army, although it had taken a city out of Tigranes's kingdom and had used favourable battles, was yet moved by the excessive distance of places and by the longing for their own.

\ciceroSection{24} Here I shall say no more. For this was the end: that out of those places by our soldiers a more timely return rather than a longer advance was sought. But Mithridates had now confirmed himself and his band by the work of those who had withdrawn to him from his own kingdom, and was helped by the great accidental supports of many kings and peoples. Now this is generally wont to happen: that the afflicted fortunes of kings easily draw the resources of many to pity --- and most of those who are either kings or live in a kingdom, since to them the kingly name seems great and sacred.

\ciceroSection{25} So conquered he could effect as much as unhurt he had never dared to wish. For when he had received himself into his own kingdom, he was not content with what had happened to him beyond hope --- that, after he had been driven out, he should ever touch that land --- but he made an attack on our distinguished and victorious army. Allow me here, citizens, as the poets are wont who write Roman matters, to pass over our calamity, which was so great that it brought it to the general's ears not as a message from a battle but as a rumour from talk. Here in that very ill and most grave check of war, Lucius

\ciceroSection{26} Lucullus, who could perhaps in some part have remedied those hurts, by your order compelled --- you who thought that, by old example, a measure must be set to the length of imperium --- some of the soldiers (who had now finished their service) discharged, some he handed over to Manius Glabrio. Many things I pass over on purpose. But by guess see how great you should reckon that war to have become, which the most powerful kings join, the stirred peoples renew, the whole peoples take up, our new general receives with the old army driven away.

\ciceroSection{27} I think I have said enough as to why this war is by its very kind necessary, by its size perilous. It remains that I should speak about choosing a general for that war and putting him in charge of such great matters. Would, citizens, that you had so great a supply of brave and innocent men that this deliberation should be hard for you --- whom in particular you should reckon must be put in charge of such great matters and so great a war! But now, when there is one Gnaeus Pompeius who has surpassed in virtue not only the glory of those men who now are but even the memory of antiquity, what thing is there which can make any man's mind doubtful in this case?

\ciceroSection{28} For I so think: that in the highest general these four things ought to be: knowledge of the military art, virtue, authority, fortune. Who then was either ever or ought to have been more knowing than this man? Who from school and from the lessons of boyhood, in the greatest war and against the fiercest enemies, set out for his father's army and into the discipline of military service; who in his last boyhood was a soldier in the army of the highest general; in the entrance of manhood was himself the general of the largest army; who has more often clashed with an enemy than anyone has fought with a personal foe; has waged more wars than the rest have read; has finished off more provinces than others have desired; whose youth was trained for the knowledge of military art not by foreign precepts but by his own commands; not by the checks of war but by victories; not by stipends but by triumphs. What kind of war, finally, can there be in which the fortune of the commonwealth has not exercised him? The civil war, the African, the Transalpine, the Spanish (mixed of citizens and most warlike peoples), the slave war, the naval war --- the various and diverse kinds both of wars and of enemies, not only waged by him alone but also finished off, declare that there is nothing set in the use of the military which can escape the knowledge of this man.

\ciceroSection{29} Now of the virtue of Gnaeus Pompeius, what speech can be found equal? What is there which any man can bring forth either worthy of him or new to you or unheard by anyone? For not only are those generalship's virtues which are commonly reckoned: labour in business, fortitude in perils, industry in acting, swiftness in finishing, counsel in foreseeing --- which are so great in this one man as in all the rest of the generals whom we have either seen or heard of they have not been. Italy is witness, which he himself the conqueror Lucius

\ciceroSection{30} Sulla confessed had been freed by the virtue and aid of this man. Sicily is witness, which, surrounded by many perils on every side, he disentangled not by the terror of war but by the swiftness of counsel. Africa is witness, which, pressed by great forces of the enemy, overflowed with the blood of those very enemies. Gaul is witness, through which a way for our legions into Spain was opened by the slaughter of the Gauls. Spain is witness, which most often saw the most enemies overcome and laid low by him. Italy is witness again and the more often, which, when it was being pressed by the foul and perilous slave war, asked help of him in his absence; which war by the expectation of him was thinned and lessened, by his coming was taken away and buried.

\ciceroSection{31} Witnesses now indeed all coasts and all lands, peoples, nations; finally all seas, both wholly and on every coast all bays and harbours. For what place in the whole sea through these years either had so firm a guard as to be safe, or was so hidden as to lie in concealment? Who sailed who did not commit himself to the peril either of death or of slavery, when he sailed either in winter or on a sea filled with brigands? This great a war, so foul, so old, so widely divided and dispersed, who would ever have reckoned could be finished off either by all generals in one year or in all years by one general?

\ciceroSection{32} What province have you held free of brigands through these years? What revenue has been safe to you? What ally have you defended? To whom have you been a defence by your fleets? How many islands do you reckon have been deserted, how many cities of allies either left out of fear or captured by brigands? But why do I recall distant things? It once was --- it was the special way of the Roman people to wage war far from home and to defend with the bulwarks of empire the fortunes of allies, not their own roofs. Shall I say that the sea has been closed to your allies through these years, when your armies never crossed from Brundisium save in highest winter? Shall I complain that those who came to you from foreign nations were captured, when envoys of the Roman people have been ransomed? Shall I say that the sea was not safe for merchants, when twelve fasces have come into the power of brigands?

\ciceroSection{33} Shall I recall Cnidus or Colophon or Samos, most noble cities, and countless others captured, when you know that your harbours, and those harbours from which you draw life and breath, have been in the power of brigands? Or are you ignorant that the harbour of Caieta, most frequented and most full of ships, with the praetor looking on, was plundered by brigands; and that out of Misenum the children of him who had before waged war with the brigands were carried off by the brigands? For why should I complain of the Ostian misfortune and that disgrace and ignominy of the commonwealth, when, almost with you looking on, that fleet over which a consul of the Roman people had been set was captured and sunk by brigands? By the immortal gods! Could the incredible and divine virtue of one man in so short a time bring such great light to the commonwealth, that you, who lately were seeing the fleet of the enemy before the Tiber's mouth, now hear that there is no pirate ship within the mouth of Ocean?

