% Pro A. Cluentio (pro-cluentio) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.
%
% The longest of Cicero's forensic speeches (202 sections),
% delivered in 66 BC. The defence of Aulus Cluentius Habitus on
% charges of poisoning his stepfather Statius Albius Oppianicus,
% complicated by Cluentius's own earlier prosecution of Oppianicus
% for poisoning -- a notoriously corrupt trial that has weighed
% on Cluentius's reputation through eight years of public hostility.

\ciceroSection{1} I have noticed, gentlemen, that all the prosecutor's speech is divided into two parts, of which the one seemed to me to lean and to trust greatly on the now grown-old hatred of the Junian trial; the other only for the sake of custom, timidly and distrustfully, to touch the reasoning of the charges of poisoning, on which matter this inquiry has been set up by the law. So I am resolved to keep this same division of hatred and charges in my defence in such a way that all may understand that I have neither wished to escape by being silent nor to obscure by speaking.

\ciceroSection{2} But when I consider in what manner I must labour in each, the one part --- which is proper to your trial and to the lawful inquiry on poisoning --- seems to me to be very brief and not to require great contention in speaking. The other, however, which is far removed from the trial, which is more fitted to seditiously roused public meetings than to calm and moderate trials --- I see how much difficulty in pleading and how much labour it will have.

\ciceroSection{3} But in this difficulty this thing, gentlemen, comforts me: that you are wont to hear charges so as to seek of the speaker the dissolving of all of them; that you reckon that you ought not to grant the defendant more for safety than the defender has been able by purging the charges and proving by speaking. About hatred, however, you ought so to debate among yourselves that you should consider not what is said by us but what ought to be said. For in the charges Aulus Cluentius's own peril is at issue; in hatred the common cause. Wherefore the one part of the case we shall plead so that we should teach you; the other so that we should beg. In the one your diligence is to be joined to us; in the other your faith is to be implored. For there is no one who can withstand hatred without your aid and that of such men.

\ciceroSection{4} Indeed, as far as it concerns me, I do not know whither to turn. Shall I deny that there was that infamy of a corrupted trial? Shall I deny that this matter was agitated in public meetings, tossed about in trials, recalled in the senate? Shall I tear from men's minds so great an opinion, so deeply set, so old? It is not of our wit; it is, gentlemen, of your help, in this calamitous fame, as if in some most ruinous flame and in a common fire, to come to the help of this innocence.

\ciceroSection{5} For just as in other places truth has too little of stay and too little of strength, so in this place false hatred ought to be weak. Let it lord it in public meetings; let it lie low in trials. Let it have weight in the opinions and talk of the unskilled; let it be rejected by the wits of the prudent. Let it have vehement sudden onsets; let it grow old with the interval of space and the case learned. Finally let that definition of fair trials which has been handed to us by our ancestors be kept: that in trials both fault should be punished without hatred and hatred set down without fault.

\ciceroSection{6} Wherefore I ask of you, gentlemen, before I begin to speak about the case itself, these things: first, that which is most fair --- that you should bring no prejudgement here. For we shall lose not only the authority but even the name of judges, if we shall not judge here from the cases themselves --- if we shall bring our judgements ready-made from home to the cases. Next, if you have already taken any opinion in your minds, if reasoning shall pluck it out, if speech shall shake it, if at last truth shall wrest it free, do not resist; let it go from your minds either willingly or with calm. But when I shall speak and dissolve about each thing, do not yourselves silently set against your thought what is contrary, but await the end and suffer me to keep my own order of speaking. When I shall have closed, then if anything shall have been passed over, ask of it in your minds.

\ciceroSection{7} I, gentlemen, easily understand that I am coming to that case which has now for eight continuous years been heard from the contrary side, and is by men's own opinion silently almost convicted and condemned. But if some god shall have won me your goodwill for hearing me, I shall surely bring it to pass that you understand that nothing is so to be feared by a man as hatred; that nothing is so to be desired by an innocent man, with hatred undertaken, as a fair trial, in which alone at last some end and outcome of false infamy is found. Wherefore great hope holds me, that, if I shall be able to set forth what is in the case and to attain all things by speaking, this place and your seating, which they thought would be dreadful and fearful for Aulus Cluentius, shall at last be the harbour and refuge of his wretched and much-tossed fortune.

\ciceroSection{8} Although there are very many things which seem to me ought to be said, before I speak about the case, of the common perils of hatred, yet, lest your expectation be longer held in suspense by my speech, I shall come to the charge with that prayer, gentlemen, which I see I must use the more often: that you should hear me as if at this time this case were spoken first, as it is spoken --- not as if it had often been spoken and never proved. For on this day for the first time the means of dissolving that old charge has been given. Before this time error in this case and hatred has been at large. Wherefore, while I answer briefly and clearly to the prosecution of many years, I ask, gentlemen, that you should hear me, as you have set yourselves to do, kindly and attentively.

\ciceroSection{9} Aulus Cluentius is said to have corrupted by money the trial, that he might condemn his innocent enemy, Statius Albius. I shall show, gentlemen, first, since the head of that atrocity and hatred was that an innocent man was caught by money, that no one was ever called into court on greater charges, by graver witnesses; next, that such prejudgements have been made about him by the very judges by whom he was condemned that not only by the same but not even by any others could he in any way be acquitted. When I have shown these things, then I shall show what I see is most asked: that that trial was tried by money not by Cluentius, but against Cluentius. And I shall make you understand in that whole case what the matter itself bore, what error feigned, what hatred forged.

\ciceroSection{10} First then this is, from which it can be understood that Cluentius ought greatly to have trusted the case: that, trusting in the surest charges and witnesses, he came down to accuse. Here, gentlemen, I must briefly set forth those charges by which Albius was condemned. I ask of you, Oppianicus, to think that I am unwilling speaking of your father's case, drawn on by faith and the duty of defence. For if to you in the present I shall not be able, yet many means of giving satisfaction shall be given me at the time remaining; if I do not now satisfy Cluentius, the means of satisfying afterward shall not be given me. At the same time, who is there who ought to doubt to speak against the condemned and the dead for the unhurt and the living? When against him against whom it is said the condemnation has now taken away every danger of disgrace, and death even of grief; while he for whom we are speaking can take in nothing of offence without the bitterest sense of mind and trouble and the highest disgrace and foulness of life.

\ciceroSection{11} And that you may understand that Cluentius brought the name of Oppianicus into court not by a prosecutor's mind, not led on by any display or glory, but by unspeakable wrongs, by daily plots, with the peril of life set before his eyes --- I shall seek a beginning a little further off for showing the matter; which I ask, gentlemen, that you should not bear heavily. For with the beginnings learned, you will much more easily understand the ends. Aulus Cluentius Habitus, the father of this man, gentlemen, was not only of the town of Larinum (from which he was), but also of that region and neighbourhood, in virtue, standing, nobility, the chief. He, when he had died in the consulship of Sulla and Pompeius, left this man fifteen years of age, and a daughter, grown and marriageable, who in a short time after her father's death married Aulus Aurius Melinus, her cousin, a young man among the chief (as he was then reckoned) honourable and noble among his own. When this marriage was full of dignity,

\ciceroSection{12} full of concord, suddenly the unspeakable lust of a fierce woman arose, joined not only with disgrace but with crime. For Sassia, the mother of this Habitus --- for "mother" by me in all the case (although she shows hostile hatred and cruelty against him), "mother," I say, she shall be called; nor shall she ever so hear from me of her own crime and monstrousness as to lose the name of nature; for the more loving and more indulgent the very name of mother is, the more by greater hatred shall you reckon worthy the singular crime of that mother who for many years now, and now most of all, has wished her son killed --- this Habitus's mother, then, was caught with love (against what was right) for that young Melinus, her son-in-law. And not even that for long. In whatever way she could, she held herself in that lust. Then she began so to blaze with frenzy, was so kindled, was so carried by lust, that neither shame nor piety, nor the family's stain, nor men's standing, nor her son's grief, nor her daughter's mourning could call her back from that desire.

\ciceroSection{13} The young man's mind, not yet firm in counsel and reasoning, she lured by all those things by which that age can be caught and softened. The daughter, who was not only afflicted by that common feminine grief at such a husband's wrongs, but could not bear her mother's unspeakable taking-of-her-husband (about which she reckoned she could not even complain without crime), wished the rest to be ignorant of so great an ill of hers. In the hands and lap of this most loving brother of hers she was wasting away with mourning and tears.

\ciceroSection{14} But behold, a sudden divorce, which seemed to be the comfort of all ills! Cluentia leaves Melinus --- not unwilling, given such great wrongs; not gladly, given that he was her husband. Then indeed that distinguished and famous mother began openly to leap with joy, to triumph with gladness --- the conqueror of her daughter, not of her lust. She was unwilling that her standing should longer be hurt by hidden suspicions. The bridal bed which two years before she had spread for her own daughter as bride, in the same house, with her daughter cast out and driven away, she ordered to be adorned and spread for herself. The mother-in-law marries her son-in-law, with no auspices, no witnesses, with the dreadful omens of all.

\ciceroSection{15} O incredible crime of a woman, and (save this one) unheard in all life! O unbridled and untamed lust! O singular audacity! Did she not fear, if not the power of the gods and the fame of men, then that very night and those wedding torches, not the threshold of the chamber, not her daughter's bed, finally not the very walls, witnesses of the earlier marriage? She broke through and laid prostrate all things by greed and frenzy. Lust overcame shame, audacity fear, frenzy reason.

\ciceroSection{16} The son bore this common disgrace of family, kinship, and name heavily. But his trouble was being increased by the daily complaints and ceaseless weeping of his sister. He resolved nevertheless that nothing graver in such great wrongs and so great a crime of his mother had to be done by him than that he should not deal with her, lest those things which he could not see without the highest grief of mind, by his dealing with his mother, he should be reckoned not only to see but even to approve by his own judgement.

\ciceroSection{17} What was the beginning to him of the strife with his mother, you have heard. That this had to do with the case, when you have learned the rest, you will understand. For it does not escape me that, of whatever sort a mother be, yet in the trial of her son it is scarcely fitting that the foulness of a parent should be spoken. I should not be a fit man for any case, gentlemen, if this --- which is set and fixed in the common feelings of men and in nature itself --- I, who am drawn to defending men's perils, did not see. I easily understand that men ought not only to be silent about the wrongs of parents but even to bear them with calm mind. But I think those things which can be borne must be borne; those which can be silent, must be silent.

\ciceroSection{18} Aulus Cluentius has seen no calamity in his life, has come to no peril of death, has feared no ill, that has not all been forged and put forth from his mother. Who at this time would be silent about all things, and would suffer those things, if she could not by forgetfulness, yet by her own silence to be covered. But indeed it is so done that she can in no way at all be silent. For this very trial, this peril, that prosecution, all that supply of witnesses which is to be, was at the start adorned by his mother; by his mother at this time it is being arrayed and prepared with all her resources and supplies. She herself lately flew from Larinum to Rome for the sake of crushing him. The audacious, wealthy, cruel woman is at hand. She sets up the prosecutors, she arrays the witnesses; she rejoices in his squalor and filth; she desires his ruin; she longs to pour out all her own blood, provided only that she sees his poured out before. Unless you shall see all these things plainly in the case, reckon her named by us rashly. But if they shall be both open and unspeakable, you ought to forgive Cluentius for suffering these things to be said by me; you ought not forgive me, if I were silent.

\ciceroSection{19} Now I shall set out summarily on what charges Oppianicus was condemned, that you may see through both Aulus Cluentius's constancy and the reasoning of the prosecution. And first I shall show what was the cause of accusing, that you may see Aulus Cluentius did this very thing compelled by force and necessity.

\ciceroSection{20} When he had caught manifestly the poison which Oppianicus, his mother's husband, had prepared for him, and the matter was held not by guess but by eyes and hands, and there could be no doubt at all in the case, he accused Oppianicus. With what constancy and with what diligence I shall say afterwards. Now I have wished you to know this: that there was for him no other cause of accusing save that he should shun the peril of life set before him and the daily snares laid for his head, by this one method. And that you may understand Oppianicus was prosecuted on such charges that neither the prosecutor ought to have feared nor the defendant to have hoped, I shall set forth a few of the charges of that trial. Which learned, none of you will wonder that he, distrusting his affairs, fled to Staienus and to money.

\ciceroSection{21} There was a certain Dinaea, of Larinum, mother-in-law of Oppianicus, who had three sons --- Marcus and Numerius Aurius and Gnaeus Magius --- and a daughter Magia married to Oppianicus. Marcus Aurius, a young man, was captured at Asculum in the Italic war, fell into the hands of Quintus Sergius the senator (he who was condemned among the assassins), and was at his house in the slave-prison. But Numerius Aurius, his brother, died and left Gnaeus Magius his brother as heir. Afterwards Magia, the wife of Oppianicus, died. Last, the only son left of Dinaea, Gnaeus Magius, died. He made heir that young Oppianicus, his sister's son, and ordered him to share with Dinaea his mother. Meanwhile an informer came to Dinaea --- not obscure or uncertain --- to tell her that her son Marcus Aurius was alive and was in the Gallic field in slavery.

\ciceroSection{22} The woman, with her children lost, when the hope of recovering one son was shown to her, called together all her kinsmen and her son's connections, and weeping asked of them that they should take up the business, track down the young man, restore to her the son whom yet alone of many fortune had wished to leave her. When she had begun to do these things, she was overcome by disease. So she made a will of such kind, that to that son she bequeathed 400,000 sesterces, and instituted as heir that same Oppianicus, her grandson. And in those few days she died. Yet her kinsmen, as they had set themselves while Dinaea lived, so when she had died, set out to track down Marcus Aurius into the Gallic field with that same informer.

\ciceroSection{23} Meanwhile Oppianicus, as he was --- as you will find from many things --- of singular crime and audacity, through a certain Gallicanus, his familiar, first corrupted that informer with money. Then he saw to the taking off and killing of Marcus Aurius himself, with no great loss made. But those who had set out to track down and recover the kinsman send letters to Larinum to the Aurii, the kinsmen of that young man and his connections: that the method of tracking him down was hard for them, because they understood that the informer had been corrupted by Oppianicus. Which letters Aulus Aurius, a brave and resourceful man, noble at home and the very near kinsman of that Marcus Aurius, openly in the forum, with many hearing, with Oppianicus present, reads out; and with the clearest voice testifies that he will lodge the name of Oppianicus, if he should learn that Marcus Aurius had been killed.

\ciceroSection{24} Meanwhile in a short time those who had set out to the Gallic field return to Larinum. They announce that Marcus Aurius has been killed. The minds not only of the kinsmen but even of all the Larinates are moved, by hatred of Oppianicus and by pity of that young man. And so when Aulus Aurius, who before had given notice, had begun to pursue the man with cries and threats, he fled from Larinum and betook himself to the camp of the most distinguished man, Quintus Metellus.

\ciceroSection{25} But after that flight, the witness of crime and conscience, he never dared to commit himself to trials, never to laws, never unarmed to his enemies. But through that force and victory of Lucius Sulla, with armed men he flew up to Larinum in the highest fear of all. He took away the four-men whom the townsmen had made; he said that he had been made by Sulla, and three others besides, and that by the same it had been ordered him that Aulus Aurius (who had shown to him the lodging of his name and the peril of his life), and the second Aulus Aurius and his son Lucius, and Sextus Vibius (whom he was said to have used as go-between in corrupting that informer) should be proscribed and killed. So with these most cruelly killed, the rest were terrified by no middling fear of proscription and death from him. With these things made plain in the case and trial, who is there who could think that he could have been acquitted? And these are small things; learn the rest, that you may wonder that Oppianicus was not at any time condemned, but was unhurt for some time.

\ciceroSection{26} First see the audacity of the man. He longed to take Sassia in marriage, the mother of Habitus --- her, whose husband, Aulus Aurius, he had killed. Whether is he the more shameless who asked it, or she the more cruel, if she should marry, is hard to say. But yet learn the humanity and constancy of each.

\ciceroSection{27} Oppianicus asks that Sassia marry him, and contends greatly. But she does not wonder at his audacity, does not despise his shamelessness, finally does not dread that house of Oppianicus dripping with the blood of her husband. But because he had three sons, she answered that on this account she shrank from that marriage. Oppianicus, who had longed for Sassia's money, thought a remedy must be sought at home for that delay which was being brought to the marriage. For when he had by Novia an infant son, but his other son (born by Papia at Teanum, which is eighteen miles from Larinum) was being brought up at his mother's, he sent for the boy from Teanum suddenly without cause, which he had not been wont to do save on public games or festal days. The wretched mother, suspecting nothing of ill, sends him. He, when he had pretended that he was setting out for Tarentum, on that very day the boy --- when at the eleventh hour he had been seen in public well --- before night was dead, and the next day before it was light was burned.

\ciceroSection{28} And to the mother this great mourning was reported by the rumour of men sooner than by any one out of Oppianicus's household. She, when she had heard at one time that not only her son but also the office of the funeral was snatched from her, came at once to Larinum unconscious, and there held the funeral over from the start, with her son already buried. Ten days had not yet passed when that other infant son is killed. So Sassia at once marries Oppianicus, with mind glad now and hope most well-confirmed; nor wonder, who saw herself softened not by wedding gifts but by the funerals of his sons. So, what others on account of children are wont to be more eager for money, he reckoned it pleasant to lose his children on account of money.

\ciceroSection{29} I see, gentlemen, that, by your humanity, with these so great crimes briefly shown by me, you are vehemently moved. With what spirit then do you suppose those men were, who about these matters not only had to hear but also to judge? You hear about him over whom you are not judges, about him whom you do not see, about him whom you can no longer hate, about him who has satisfied both nature and the laws --- whom the laws have punished with exile, nature with death. You hear not from an enemy, you hear without witnesses, you hear when those things which can be most copiously said are spoken briefly and in passing. Those men were hearing about him over whom they were sworn to give their votes, about him whose unspeakable and crime-laden face, present, they were looking at, about him whom all hated on account of his audacity, about him whom they reckoned worthy of every punishment. They were hearing from prosecutors; they were hearing the words of many witnesses; they were hearing about each thing while it was being spoken gravely and at length by Publius Cannutius, a most eloquent man.

\ciceroSection{30} And is there anyone who, when he has learned these things, can suspect that Oppianicus was crushed in trial and circumvented as innocent? In a heap, gentlemen, I shall now set out the rest, that I may come to those things which are nearer to this case and more closely joined. Hold, please, in memory that this is not set forth by me --- that I should accuse the dead Oppianicus --- but, since I wish to persuade you that the trial was not corrupted by him, that I should use this as the start and foundation of the defence: that Oppianicus, a most criminal and most guilty man, was condemned. Who when he himself had given the cup to his wife Cluentia, the aunt of this Habitus, suddenly she in the middle of the drinking cried out that she was dying with the greatest pain, nor did she live longer than she spoke. For in that very speech and shout she died. And to this sudden death and the voice of the dying, all besides which are wont to be the signs and traces of poison were on the body of the dead woman. And by the same poison he killed his brother Gaius Oppianicus.

\ciceroSection{31} Nor is this enough. Although in the very fratricide no crime seems to be passed over, yet, that he might come to this unspeakable deed, he prepared his approach by other crimes before. For when his brother's wife Auria was pregnant, and the birth was now thought to be near, he killed the woman with poison, that at the same time what had been conceived from his brother might be killed. After he attacked his brother, who, late, with that cup of death now drained, when he was crying out about his own and his wife's destruction and longed to change his will, in the very signifying of this will died. So he killed the woman, lest by her birth he should be shut out from his brother's inheritance. But his brother's children he deprived of life before they could receive this light from nature --- that all might understand that nothing could be closed, nothing sacred, against him from whose audacity the children of his brother not even the guarding of the mother's body could keep safe.

\ciceroSection{32} I hold in memory a Milesian woman, when I was in Asia, who, because she had taken money from the second heirs and procured an abortion for herself by drugs, was condemned of a capital matter; nor unjustly, who had taken away the hope of a parent, the memory of a name, the help of a line, the heir of a household, a destined citizen of the commonwealth. By how much is Oppianicus, in the same wrong, worthy of greater punishment! If indeed she, when she had brought force on her own body, tortured herself; but he brought about the same through another's body's death and torture. Others do not seem able to undertake many parricides in single men; Oppianicus has been found to kill more in one body.

\ciceroSection{33} So when his uncle had learned this custom and audacity of his --- the uncle of that young man Oppianicus, Gnaeus Magius --- and he, when he was afflicted with a grave disease and was making that man, his sister's son, his heir, with friends called in and his mother Dinaea present, asked his wife if she was pregnant. When she answered that she was, he asked of her that, when he was dead, at Dinaea's house --- who was then her mother-in-law --- until she should give birth, she should live, and bring diligence that what she had conceived she might be able to keep and to bring forth safe. So he bequeathed her by will a great sum of money from his son if any should be born; from the second heir nothing he bequeathed. What he suspected of Oppianicus you see;

\ciceroSection{34} what he judged is not obscure. For whose son he made his heir, him he did not name as guardian for his children. What Oppianicus did, learn, that you may understand that Magius did not look far ahead in mind as he was dying. The money which had been bequeathed to the woman from her son if any should be born, that, ready, Oppianicus paid the woman not owed --- if this is to be called payment of a legacy and not the wage of an abortion. By which price taken and by many gifts besides (which then were being read out from Oppianicus's tablets), the hope which she had in her belly entrusted by her husband, conquered by greed, she sold to the crime of Oppianicus. Nothing now seems able to be added to this wickedness;

\ciceroSection{35} mark the outcome. The woman who, by her husband's adjuration, in those ten months ought not to know any house save her mother-in-law's, in the fifth month after her husband's death married Oppianicus himself. Which marriage was not long-lasting; for they had been joined not by the dignity of marriage but by the partnership of crime.

\ciceroSection{36} What? That slaughter of Asuvius of Larinum, a wealthy young man --- how famous it was then, the matter recent; how celebrated by the talk of all! There was a certain Avillius of Larinum, of ruined wickedness and the highest want, endowed with a kind of art fitted to rousing the lusts of young men. Who, when he had plunged himself by flatteries and assents deeply into Asuvius's company, Oppianicus at once began to hope that, by Avillius as if by some engine moved up, he could capture Asuvius's youth and storm his ancestral fortunes. The plan was thought up at Larinum; the matter was carried to Rome. For they reckoned it could be more easily entered upon in solitude, and more conveniently finished in a crowd. Asuvius set out with Avillius for Rome. Oppianicus followed them on their tracks. How they lived at Rome, in what banquets, in what disgraces, with how great and how lavish expenses --- with not only Oppianicus aware but a fellow-banqueter and helper --- it is long for me to say, especially as I hasten to other things. Learn the outcome of this feigned familiarity.

\ciceroSection{37} When the young man was at the house of a certain little woman, and where he had spent the night was tarrying for the next day --- Avillius, as had been arranged, pretends to be sick and to wish to make his will. Oppianicus brings to him sealers who knew neither Asuvius nor Avillius, and calls that man Asuvius himself; with the will sealed in Asuvius's name they depart. Avillius at once recovers; but Asuvius in that short time, as if he were going to a little garden, was led out to certain sand-pits beyond the Esquiline gate and killed.

\ciceroSection{38} When he was already missed for one and a second day and was not being found in those places where by custom he was being looked for, and Oppianicus in the forum of the Larinates was repeatedly saying that lately he and his friends had sealed his will, the freedmen of Asuvius and not a few friends, because it was agreed that Avillius had been with him on the day on which Asuvius had last been seen, and had been seen by many, attack the man and set him before the feet of Quintus Manlius, who was then triumvir. And there at once, with no witness, no informer, terrified by the consciousness of the recent evil deed, he sets out everything (as it was said by me a little before) and confesses that Asuvius had been killed by him at the counsel of Oppianicus. Oppianicus is dragged out from the house where he was hiding by Manlius;

\ciceroSection{39} the informer Avillius is held openly on the other side. What now do you ask of the rest? Most of you knew Manlius. He had never thought of honour from boyhood, of the pursuits of virtue, of any fruit of good standing; but, from a petulant and dishonest buffoon, in the discords of the state had now come by the votes of the people to that column at which by the abuse of many he had often been carried. So then he settles with Oppianicus, takes money from him, leaves the case --- both undertaken and so plain. And then in Oppianicus's case this Asuvian charge was confirmed both by many witnesses and indeed by his very will, in which it was agreed that Oppianicus's name was first --- the man whom you say was wretched and innocent, circumvented by a false trial.

\ciceroSection{40} What? Your grandmother, Oppianicus, Dinaea, of whom you are heir --- did your father not manifestly kill? To whom, when he had brought that famous doctor of his, now well known and often victorious, through whom he had killed very many, the woman cries out that she does not at all wish to be cured by him by whose curing she had lost all her own. Then he suddenly attacks a certain Anconitan man, Lucius Clodius, a circuit pharmacist, who by chance had then come to Larinum, and settles with him at 2,000 sesterces (which is shown from his very tablets). Lucius Clodius, since he was hurrying, who had many circuits remaining, as soon as he was brought in, finished the matter; with the first dose he took the woman off and afterwards stayed not a moment of time at Larinum. The same Dinaea making her will,

\ciceroSection{41} when Oppianicus, who had been her son-in-law, had taken the tablets, with his finger he wiped out the legacies; and, when he had done this in many places, after her death, lest he be convicted by the erasures, he had the will copied onto other tablets and sealed it with adulterated seals. Many things I pass over on purpose; for I fear that these very things may seem too many. Yet you ought to reckon that he was like himself in the rest of the parts of his life. The decurions all judged that he had corrupted the public records of Larinum, the censorial ones. With him no one now had business, no one transacted any matter; no one out of so many kinsmen and connections ever wrote him as guardian for his children; no one judged him worthy of approach, no one of meeting, no one of speech, no one of banquet. All shrank away, all turned aside, all fled him as if some monstrous and ruinous beast and pestilence.

\ciceroSection{42} Yet this man so audacious, so unspeakable, so guilty Habitus would never have prosecuted, gentlemen, if he could have passed it over with his life safe. Oppianicus was an enemy to him, was so. But yet he was his stepfather. Cruel and hostile to him was his mother, but his mother. Finally nothing was so far from accusation as Cluentius, both by nature and by will and by the ordered reasoning of his life. But when this condition was set before him --- that either he should justly and dutifully accuse, or bitterly and unworthily die --- he preferred to accuse in whatever way he could, than to die in that manner.

\ciceroSection{43} And that you may see this is so, I shall set out for you Oppianicus's deed manifestly learned and caught --- from which at the same time you will understand that both for him to accuse and for him to be condemned was necessary. There were certain Martiales at Larinum, public ministers of Mars, consecrated to that god by the old institutions and religious sanctities of the Larinates. Of whom, since there was a great enough number, and likewise (as in Sicily there are very many Venusians) so they at Larinum were reckoned in Mars's household, suddenly Oppianicus began to defend that they were all free and Roman citizens. Heavily the decurions of Larinum and all the townsmen took it. So they asked Habitus that he should take up that case and defend it publicly. Habitus, although he had removed himself from every business of that kind, yet for the sake of his place, for the antiquity of his line, for that --- because he reckoned himself born not for his own advantages but also for those of his townsmen and other connections --- did not wish to fail the will of all the Larinates so great.

\ciceroSection{44} With the case undertaken and brought to Rome, daily great contentions between Habitus and Oppianicus were stirred up out of the zeal of each for the defence. Oppianicus himself was of monstrous and bitter nature; his frenzy was kindled by the mother of Habitus, hostile and inimical to her son. The Magni, however, thought it of their interest that he should be moved away from the Martiales' case. There lay underneath also another greater cause which moved most the mind of Oppianicus, a most greedy and most audacious man.

\ciceroSection{45} For Habitus up to the time of that trial had never made any will. For he could not bring it into his mind to leave anything of that sort to his mother, nor by will to pass over altogether the name of his parent. When Oppianicus knew this --- for it was not obscure --- he understood that, with Habitus dead, all his goods would come to his mother; who, with money increased by him, with greater gain, deprived of her son, with less peril could be killed. So, set on fire by these things, by what reasoning he tried to take away Habitus by poison, learn.

\ciceroSection{46} Gaius and Lucius Fabricius were twin brothers, of the town of Aletrium, men like one another both in form and in morals, but most unlike their townsmen --- of whom what splendour, how almost equal, how almost in all constant and moderate the reasoning of life is, none of you (in my opinion) is ignorant. With these Fabricii Oppianicus had always most familiarly used. Now you almost all know how great force the likeness of pursuits and nature has for joining friendships. When they so lived as to think no gain shameful, when from them every fraud, all snares and circumventions of young men were born, and they were known to all by vices and dishonesty --- zealously, as I said, Oppianicus had attached himself to their familiarity many years before. So then he resolved through Gaius

\ciceroSection{47} Fabricius (for Lucius had died) to prepare snares for Habitus. At that time Habitus was of weak health. He was using a not ignoble but a respected man as doctor, Cleophantus, whose slave Diogenes Fabricius began to solicit by hope and price for giving poison to Habitus. The slave, not uncunning and (as the matter itself made plain) thrifty and upright, did not despise Fabricius's speech; he took the matter to his master. Cleophantus then conferred with Habitus. Habitus at once shared it with Marcus Baebius, a senator, his most familiar friend; of whom, of what faith, of what prudence, of what diligence he was, I think you remember. It pleased him that Habitus should buy Diogenes from Cleophantus, that the matter could the more easily either be caught by his information or shown to be false. To be brief: Diogenes is bought; in a few days the poison is procured. Many good men, when they had come up secretly, the money sealed which was being given for that thing was caught in the hands of Scamander, freedman of the Fabricii.

\ciceroSection{48} By the immortal gods! Will anyone, with these things known, say that Oppianicus was circumvented? Who was ever brought into court more audacious, who more guilty, who more open? What talent, what ability of speaking, what defence devised by anyone could withstand this one charge? At the same time, who is there who could doubt that, with this thing learned and manifestly caught, either death had to be met by Cluentius or accusation undertaken?

\ciceroSection{49} I think it is shown enough, gentlemen, that Oppianicus was accused on such charges that he could in no honourable manner be acquitted. Learn now that the defendant was so summoned that, with the matter once and again prejudged, he came into court already condemned. For Cluentius, gentlemen, first lodged the name of him in whose hands he had caught the poison. He was the freedman of the Fabricii, Scamander. The council was untouched, no suspicion of corrupted trial; the case in court was simple, the matter sure, one charge brought. Then Gaius Fabricius (he of whom I spoke before), who, when the freedman had been condemned, saw that the same peril was hanging over himself, who knew that I had a neighbourly bond with the Aletrines and great use with most of them, brought them in numbers to me at home. Who, although they thought of the man as it was necessary, yet, because he was of the same town, reckoned it was of their dignity to defend him in what ways they could; and they asked of me that I should do this and undertake Scamander's case --- in which case the whole peril of the patron was contained.

\ciceroSection{50} I, who could neither refuse anything to such men so loving me, nor reckoned that charge so great and so manifest --- as not even those very men did who were then commending the case to me --- promised them that I would do all that they wished. The matter began to be conducted; Scamander was summoned as defendant. The accuser was Publius Cannutius, a most ingenious man and practised in speaking. He accused Scamander indeed in three words: that the poison had been caught. All the weapons of the whole prosecution were thrown at Oppianicus. The cause of the snares was opened; the familiarity of the Fabricii was recalled; the man's life and audacity were brought forth; finally the whole prosecution, handled variously and gravely, was at the end concluded by the manifest catching of the poison. Here I then rose to answer --- with what care, immortal gods!

\ciceroSection{51} With what trouble of mind, with what fear! For my part, I always begin to speak with great fear; as often as I speak, so often do I seem to come to a trial not only of talent but also of virtue and duty, lest I should seem either to profess what I cannot (which is shamelessness), or not to bring about what I can (which is either treachery or carelessness). Then indeed I was so disturbed that I feared all things: if I should say nothing, lest I be reckoned the most childish; if I should say much in such a case, lest I be reckoned the most shameless. I gathered myself at length and so resolved: that I must act bravely. To that age which I then was it was wont to be reckoned to my praise, even if in less firm cases I had not failed men's perils. So I did. So I fought, so by every method I contended, so to all things I fled, as far as I was able, the remedies and refuges of cases, that I might attain --- this I shall say timidly --- that no one should think a patron had been wanting to that case.

\ciceroSection{52} But whatever I had grasped, the prosecutor at once wrenched out of my hands. If I had asked what enmities Scamander had with Habitus, he confessed that there were none; but said that Oppianicus, whose minister he had been, was and is most hostile to him. But if I had managed this --- that no profit was going to come to Scamander out of the death of Habitus --- he granted; but he said that to the wife of Oppianicus, a man practised in killing wives, all the goods of Habitus would come. When I had used that defence which in the cases of freedmen has always been reckoned most honourable --- that Scamander was approved by his patron --- he confessed; but asked by whom the patron himself was approved.

\ciceroSection{53} When I had dwelt in this in many words --- that snares had been laid for Scamander through Diogenes, and it had been arranged between them in another matter that Diogenes should bring a medicine, not poison; that this could happen to anyone --- he asked why he had come into such a hidden place, why alone, why with sealed money. Finally in this place the case was pressed by witnesses, by most honourable men. Marcus Baebius said that Diogenes had been bought by his counsel, and that with him present Scamander had been caught with the poison and the money. Publius Quintilius Varus, a man endowed with the highest religious obligation and the highest authority, said that, of the snares which were being laid for Habitus and of the soliciting of Diogenes, Cleophantus had spoken with him, the matter recent.

\ciceroSection{54} And in that trial, although we seemed to defend Scamander, in word he was the defendant, but in fact and in peril, by the whole prosecution, Oppianicus was. Nor was he bearing it obscurely, nor could he in any way pretend. He was present in numbers; he kept calling supporters; he fought with all zeal and favour. Finally --- which was the greatest harm to that case --- in this very place, as if he himself were the defendant, he sat. The eyes of all the judges were turned not on Scamander but on Oppianicus. His fear, his disturbance, his suspended and uncertain face, his frequent change of colour, which were before suspicious, made these things open and manifest.

\ciceroSection{55} When it was right to go to the council, Gaius Junius the inquisitor asked of the defendant, by that Cornelian law which then was, whether he wished the vote on himself to be cast in secret or openly. By Oppianicus's vote it was answered (since he was saying that Junius was a familiar of Habitus) that he wished it cast in secret. They went to council. By all the votes save one (which Staienus said was his) Scamander was condemned in the first hearing. Who then of all was there who, with Scamander condemned, did not reckon that judgement was made about Oppianicus? What is judged by that condemnation, save that it was poison which was being sought to be given to Habitus? What slighter suspicion was thrown on Scamander, or could be thrown, that he should be reckoned of his own will to have wished to kill Habitus?

\ciceroSection{56} And with this trial then made, and Oppianicus by fact and standing now (by law and pronouncement not yet) condemned, yet Habitus did not at once make Oppianicus defendant. He wished to learn whether the judges were strict only against those whom they had found to have had the poison themselves, or even judged the counsels and consciousnesses of such crimes worthy of punishment. So Gaius Fabricius, whom he reckoned to have been an accomplice in that deed on account of his familiarity with Oppianicus, he at once made defendant; and he obtained that the first place be set up for him on account of the connection of the case. Then Fabricius not only did not bring to me his neighbours and friends the Aletrines, but himself could afterwards use them neither as defenders nor as praisers.

\ciceroSection{57} For we thought it was the part of humanity to defend a matter (untouched) of a man not foreign, however suspicious; the part of shamelessness to try to shake what had been judged. So then he, by want and necessity compelled, in such a case fled to the brothers Caepasius, busy men and of such mind that whatever power of speaking was given they would set in honour and benefit. Now this is set up almost most unfairly: that, in diseases of the body, as each man is the most difficult, so the most noble and the best doctor is sought; in capital perils, as each cause is the most difficult, so the worst and most obscure patron is brought in. Unless perhaps this is the cause: that doctors must furnish nothing besides art, but speakers must furnish authority too.

\ciceroSection{58} The defendant is summoned, the case is conducted. In few words Cannutius accuses, as a matter already judged. The elder Caepasius begins to answer with a long and far-fetched preface. At first his speech is heard attentively. Oppianicus was raising his now downcast and pressed-down spirit; Fabricius himself was rejoicing. He did not understand that the spirits of the judges were being moved not by his eloquence but by the shamelessness of the defence. After he began to speak about the matter, to those things which were in the case he himself even added certain new wounds; so that, although he was doing it on purpose, he yet sometimes seemed not to be defending but to be playing the false ally. So when he thought he was speaking most cunningly, and had brought forth those most grave words from the inmost art: "Look back, gentlemen, on the fortunes of men, look back on doubtful and varying chances, look back on the old age of Gaius Fabricius" --- when he had said this "look back" often for the sake of adorning his speech, he himself looked back. But Gaius Fabricius, with head cast down, had departed from the benches.

\ciceroSection{59} Here the judges laugh; the patron grows angry and bears it bitterly that the case is snatched from him, and that he cannot say the rest from that "Look back, gentlemen" passage. And nothing was nearer happening than that he should pursue the man and with twisted neck bring him back to the benches, that he might close the rest. So then Fabricius was condemned, first by his own judgement (which is the gravest), then by the force of the law and the votes of the judges. What is there now that we should say more about Oppianicus's person and case? Before the same judges he was made defendant, when he was now condemned by these two prejudgements; but by the same judges who, by the condemnation of the Fabricii, had judged about Oppianicus, the first place was set up for him. He was accused on the gravest charges --- both those which have been briefly said by me, and many others besides which I now leave aside; he was accused before those who had condemned Scamander as Oppianicus's minister, and Gaius Fabricius as accomplice of the evil deed.

\ciceroSection{60} Whether, by the immortal gods! is it more wonderful that he was condemned, or that he at all dared to answer? For what could those judges do? Who, if they had condemned the innocent Fabricii, ought yet to be consistent with themselves and to agree with their previous judgements in Oppianicus's case. Or could they themselves have rescinded their own judgements, when others are wont in judging to provide that they should not differ from the judgements of others? And could those who had condemned Fabricius's freedman because he had been the minister in the evil deed, the patron because he was an accomplice --- acquit the very chief and architect of the crime? And could those who had condemned the rest, with no prejudgement made, yet on the case itself, free this man whom they had received as already twice condemned?

\ciceroSection{61} Then indeed those senatorial trials, marked not by false hatred but by true and notable foulness, and covered with disgrace and infamy, would have left no place for defence. For what then would those judges answer, if anyone should ask of them: "You condemned Scamander; on what charge?" "Surely because he had wished to kill Habitus through the doctor's slave by poison." "What did Scamander gain by Habitus's death?" "Nothing; but he was the minister of Oppianicus." "And you condemned Gaius Fabricius; why so?" "Because, when he himself most familiarly used Oppianicus, but his freedman had been caught in the evil deed, it was not made probable that he was without share of that counsel." If, then, they had acquitted Oppianicus himself, twice condemned by their own judgements, who could have borne so great a foulness of judgements, who so great an inconsistency of judged matters, who so great a lust of the judges?

\ciceroSection{62} But if you see this --- which now by all this speech has been made plain --- that in that trial it was necessary that the defendant be condemned, especially by the same judges who had made the two prejudgements, at the same time you must necessarily see this: that the prosecutor could have had no cause to wish to corrupt the trial. For I ask of you, Titus Attius, with all the rest of the proofs now left aside, whether you reckon the Fabricii also were innocent and condemned; whether you say those trials too were corrupted by money --- in which trials the one was acquitted by Staienus alone, the other condemned even himself. Come: if guilty, of what evil deed? Was anything besides the poison sought for, by which Habitus was to be killed, thrown at them? Was anything else handled in those trials save these snares laid for Habitus by Oppianicus through Fabricius? Nothing, nothing, I say, else, gentlemen, will you find. The memory stands; there are public records. Convict me if I lie. Read out the words of the witnesses; teach what besides this poison of Oppianicus was thrown at them in those trials, not only as a charge but even as an evil saying.

\ciceroSection{63} Many things can be said why it had necessarily been so judged; but I will meet your expectation, gentlemen. For although I am so heard by you that I do not think anyone has ever been more kindly or more attentively heard, yet your silent expectation has long called me elsewhere, which seems to speak against me: "What then? Do you deny that that trial was corrupted?" I do not deny; but I confirm that it was not corrupted by him. "By whom then was it corrupted?" I think first: if it had been uncertain what the outcome of that trial would be, it would yet be more like the truth that he had corrupted who had feared lest he himself be condemned, than that other who had feared lest the other be acquitted. Next, since it was not in doubt what had to be judged, surely rather he who in some way distrusted himself than he who in every reasoning was confident. Finally, surely rather he who twice had offended before those judges, than he who twice had had his case approved.

\ciceroSection{64} One thing surely no one will be so hostile to Cluentius as not to grant me: that, if it stand fixed that that trial was corrupted, it was corrupted either by Habitus or by Oppianicus. If I show that it was not by Habitus, I prove by Oppianicus; if I show by Oppianicus, I clear Habitus. Wherefore, although I have shown enough that there was no reason for him to corrupt this trial (from which it is understood that it was corrupted by Oppianicus), yet learn separately about him himself. And I shall not argue those things which are vehemently grave: that he had corrupted who had been in danger, who had feared, who had no hope of safety in another reckoning, who had always been of singular audacity. Many such things there are. But, since I have a thing not doubtful but open and manifest, the enumeration of single proofs is not necessary.

\ciceroSection{65} I say that to Gaius Aelius Staienus the judge Statius Albius gave a great sum of money for corrupting the trial. Does anyone deny it? You I call upon, Oppianicus; you, Titus Attius, of whom the one by eloquence, the other by silent piety laments that condemnation. Dare to deny that money was given by Oppianicus to Staienus the judge. Deny it; deny it, I say, in my place. Why are you silent? Or can you not deny what you have demanded back, what you have confessed, what you have carried off? With what face, then, do you make mention of a corrupted trial, when you confess that money was given by your own side to a judge before the trial, and was wrenched away after the trial?

\ciceroSection{66} In what manner, then, were these things done? I shall recall a little further off, gentlemen, and shall so open all the things which have lain hidden in long-lasting obscurity that you may seem to look upon them with your eyes. I ask of you that, as you have up to now attentively heard me, you should likewise hear what remains. Surely nothing shall be said by me which shall not seem worthy of this gathering and silence, worthy of your zeal and your ears. For as soon as Oppianicus, from the fact that Scamander had been made defendant, began to suspect what was hanging over him, he at once attached himself to the familiarity of a needy and audacious man, practised in corrupting trials, and at that time a judge: Staienus. And first, with Scamander as defendant, he had so far brought it about by gifts and presents that he used him as a more eager partisan than the faith of a judge demanded.

\ciceroSection{67} But afterwards, when Scamander had been acquitted by Staienus's vote alone, while the patron of Scamander had not been freed even by his own vote, he thought that sharper remedies must be brought to his own safety. Then from Staienus --- as from a man most acute in devising, most shameless in daring, most fierce in bringing about (for those things he both partly had and partly pretended he had) --- he began to seek help for his life and fortunes. Now you are not ignorant, gentlemen, that even beasts, warned by hunger, often return to the place where they have at any time been fed.

\ciceroSection{68} That Staienus had two years before, when he had taken on the case of the goods of Safinius Atella, said that for 600,000 sesterces he would corrupt the trial. Which, when he had taken from the orphan, he kept hidden, and, the trial held, gave back neither to Safinius nor to the buyers of the goods. Which money, when he had spent it and had left himself nothing not only for his desires but not even for his necessity, he resolved that he must return to the same plunders and trial-suppressions. So when he saw Oppianicus already ruined and slain by the two prejudgements, by his promises he raised him up cast down, and at the same time forbade him to despair of safety. Oppianicus then began to ask the man to show him the way of corrupting the trial.

\ciceroSection{69} And he, as was afterwards heard from Oppianicus himself, denied that there was anyone in the state save him who could bring this about. But at first he began to be heavy, because he was saying that he was seeking the aedileship with most noble men and feared envy and offence. Afterwards, won over, he first asked a very great sum, then came down to that which could be brought about: 640,000 sesterces he ordered to be brought home to him. Which money, as soon as it had been brought to him, the most impure man at once began to turn in such mind and thought: that nothing was more useful for his reckonings than that Oppianicus should be condemned; for, with him acquitted, that money would have either to be apportioned to the judges or given back to the man himself; with him condemned, no one would demand it back.

\ciceroSection{70} So he devises a singular thing. And these things, gentlemen, which are truly said by us, you will the more easily believe if you wish in your minds, after a long interval, to recall the life and nature of Gaius Staienus. For just as the opinion is of each man's morals, so it can be reckoned what was done or not done by him. Since he was needy, expensive, audacious, cunning, treacherous, and at his own house, most wretched in most empty places, he saw such great a sum of money set down, he began to turn his mind to every malice and fraud. "Should I give it to the judges? For myself then, save peril and infamy, what shall be sought? Shall I devise nothing wherefore Oppianicus must be condemned? What then? --- for there is nothing which cannot happen --- if some chance has perhaps snatched him from peril, must it not be given back? Let us push him as he is going down headlong," he says, "and lay him low ruined."

\ciceroSection{71} He takes this counsel: that he should promise money to certain very light judges; then keep it back afterwards, so that --- since he reckoned grave men of their own accord were going to judge strictly --- he should make those who were lighter, in their disappointment, angry with Oppianicus. So, as he was always preposterous and perverse, he begins from Bulbus and lightly nudges him --- now for some time having sought nothing, sad and yawning. "What of you?" he says. "Will you help me, Bulbus, that we should not serve the commonwealth gratis?" He, as soon as he had heard this "not gratis": "I shall follow whither you wish," he says; "but what do you bring?" Then he promises him 40,000, if Oppianicus should be acquitted; and he asks him to address the rest with whom he was wont to speak; and the framer of the whole business himself sprinkles his Gutta on this Bulbus.

\ciceroSection{72} So he seemed by no means bitter to those who had tasted some little hope from his speech. One and another day had passed when the matter seemed too little sure. The depository and confirmer of the money was being looked for. Then with cheerful face Bulbus addresses the man as flatteringly as he can: "What of you," he says, "Paetus?" --- for this surname Staienus had chosen for himself out of the images of the Aelii, lest, if he had made himself a Ligur, he should seem to use the surname rather of his nation than of his line --- "concerning the matter you spoke of with me, they are asking of me where the money is." Here that most dishonest deceiver, fed by trial-gain, who was now in spirit and mind brooding over that money which he had stored up, knits his brow --- recall his face and those feigned and pretended looks of his --- and he, who was wholly made out of fraud and lying, and who had even by zeal and a kind of artistry of malice seasoned the vices he had from nature, beautifully maintains that he had been let down by Oppianicus; and he adds this as testimony: that, by his vote, since all were openly going to give theirs, he would be condemned.

\ciceroSection{73} Talk had spread in the council that some mention of money had been at large among the judges. The matter was neither so hidden as it was being hidden, nor so open as it ought to be opened for the sake of the commonwealth. In that obscurity and doubt of all, it pleased Cannutius, an experienced man, who by a kind of scent of suspicion had felt that Staienus had been corrupted, and yet did not think the matter was finished, suddenly to pronounce: "They have spoken." Here then Oppianicus did not greatly fear; he reckoned the matter had been finished by Staienus.

\ciceroSection{74} They were going to council: thirty-two judges. By sixteen votes acquittal could be brought about. 40,000 sesterces distributed among single judges should bring about that number of votes, so that on top, in hope of greater rewards, Staienus's own seventeenth vote should be added. And by chance then --- because that had been done suddenly --- Staienus himself was not present. He was defending I know not what case before a judge. Habitus easily suffered this; Cannutius easily; but Oppianicus did not, nor his patron Lucius Quinctius. Who, since he was at that time tribune of the plebs, made the greatest disturbance against Gaius Junius the inquisitor of the trial, that they should not go to council without him; and when this seemed to him to be done too carelessly on purpose by the runners, he himself set out from the public trial to Staienus's private trial and ordered that to be dismissed by his power; he himself led Staienus to the benches.

\ciceroSection{75} They rise to council, when Oppianicus had said he wished the votes to be cast openly --- which was then the power --- so that Staienus could know what he owed to each. Various kinds of judges. The money-takers were few, but all angry. As those who in the field have been wont to take are wont to be most hostile to those candidates whose money they think has been kept back, so judges of that kind had then come hostile to the defendant. The rest reckoned him most guilty; but they were waiting for the votes of those whom they thought had been corrupted, that out of these they might decide by whom the trial seemed to be corrupted. And lo, such a drawing of lots that among the first Bulbus and Staienus and Gutta had to judge! There was the highest expectation of all what those light and money-taking judges would vote. And they all, without any doubt, condemn.

\ciceroSection{76} Here then a stumbling-block was thrown into men, and a certain doubt as to what had been done. Then wise men, of that old discipline of trials, who could neither acquit a most guilty man, nor (about him about whom suspicion had arisen that he had been assailed by money) wish, with that matter unknown, to be the first to condemn, said that it was not clear. But some strict men who had laid down this --- that what each man should do was to be regarded by the spirit with which he did it --- although others, with money taken, were judging the truth, yet none the less reckoned that they ought to be consistent with their previous judgements; so they condemned. Five there were in all who acquitted that innocent Oppianicus of yours --- led on either by imprudence or by pity or by some suspicion or by canvassing.

\ciceroSection{77} With Oppianicus condemned, at once Lucius Quinctius, a man most popular, who was wont to gather all the winds of rumours and meetings, reckoned that the means had been offered him to grow out of senatorial hatred --- because he reckoned that the trials of that order were now less approved by the people. One and a second public meeting, vehement and grave, is held. The tribune of the plebs kept crying that the judges had taken money to condemn an innocent defendant; he was saying that the fortunes of all were at stake; that there were no trials; that whoever had a wealthy enemy could not be safe. Men ignorant of the whole business, who had never seen Oppianicus, thinking that an excellent and most modest man had been pressed down by money, kindled by suspicion, began to call the matter into the open and to demand that whole case.

\ciceroSection{78} And at that very time, in the house of Titus Annius, a most honourable man, my connection and friend, by night Staienus came summoned by Oppianicus. The rest is now known to all: how Oppianicus dealt with him about the money, how he said he would give it back, how this whole speech was heard by good men who had then on purpose stood near in concealment, how the matter was made plain and brought into the forum, and all the money was wrenched and snatched from Staienus. The character of this Staienus was now known to the people and clearly seen, was free of no foul suspicion. That money had been kept back by him which he had vouched for the defendant, those who were in the meeting did not understand. For neither were they being taught. They felt that mention of money had been at large in the trial; they heard an innocent defendant had been condemned; they saw he had been condemned by Staienus's vote. That this had not been done by him gratis, knowing the man, they judged. A like suspicion stuck on Bulbus, on Gutta, on certain others.

\ciceroSection{79} So I confess --- for now I can without punishment confess this, especially in this place --- since not only the life but even the name of Oppianicus had before that time been unknown to the people, and it then most unworthily seemed that an innocent man had been circumvented by money --- this suspicion the wickedness of Staienus and the foulness of certain like judges of his afterwards increased --- and Lucius Quinctius pleaded the case, a man both endowed with the highest power and fitted for inflaming the spirits of the multitude --- the highest hatred and infamy was forged at that trial. And into this fresh flame I remember that Gaius Junius, who had been in charge of that inquiry, was thrown; and that man (of aedile rank, by all opinions now set up as praetor) was taken away not by debate of speaking but by the cry of men, from the forum and even from the state.

\ciceroSection{80} And I do not regret that I am defending the case of Aulus Cluentius at this time rather than then. For the case stays the same, which can in no way be changed. The unfairness of the time and hatred has gone back, so that what was of evil in the time hinders nothing, what was of good in the case helps. So now in what manner I am heard I feel, not only by those whose judgement and power it is, but even by those whose only standing it is. But then if I had spoken, I would not have been heard --- not because the matter was different (indeed it was the same), but the time was different. Learn it thus. Who would then dare to say that Oppianicus had been condemned guilty? Who now dares to deny it? Who then could prove that the trial had been tried by money by Oppianicus? Who at this time can deny it? To whom would it then be allowed to teach that Oppianicus had been made defendant only when he had been condemned by two recent prejudgements? Who is there who at this time would try to weaken it?

\ciceroSection{81} Wherefore, with hatred removed (which day has softened, my speech has pleaded off, your faith and equity has thrown back from the debate of truth), what is left further in the case? It stands fixed that money was at large in the trial. It is asked from where it set out: from the prosecutor or from the defendant. The prosecutor says these things: "First I was prosecuting on the gravest charges, so that there was no need of money. Then I was bringing a man condemned, so that even by money he could not be snatched. Last, even if he had been acquitted, the standing of all my fortunes would yet have remained whole." What against the defendant? "First, I was greatly fearing the very multitude and atrocity of the charges. Then, with the Fabricii condemned on account of consciousness of my own crime, I felt that I had been condemned. Last, I had come into such a chance that the whole standing of my fortunes was contained in the peril of one trial."

\ciceroSection{82} Come: since he had many and grave causes for corrupting the trial, this man none, let the very setting-out of the money be looked for. Cluentius made up the records most diligently. But this matter has surely this property: that nothing can either be added to or taken from one's estate without leaving traces. There are eight years now since this case has been turned in this preparation, when all things which have to do with this matter you have agitated, handled, sought out from his and from others' tablets. While meanwhile of Cluentian money you find no trace. What? Must Albian money be tracked by us by traces, or can we come even to the very lair, with you as guides? In one place are held 640,000 sesterces, are held with a most audacious man, are held with a judge. What more do you wish?

\ciceroSection{83} But Staienus was set up not by Oppianicus but by Cluentius for corrupting the trial. Why, when they were going to council, did Cluentius and Cannutius suffer him to be absent? Why, when they were sending to council, did they not look for Staienus the judge to whom they had given money? Oppianicus complained, Quinctius pressed; that the council should not be held without Staienus was brought about by tribunician power. But he condemned. For he had given this condemnation as a hostage to Bulbus and the rest, that he might seem let down by Oppianicus. Wherefore, if from that side the cause of corrupting the trial, if from that side the money, from that side Staienus, from that side finally all fraud and audacity is, but on this side modesty, an honourable life, no suspicion of money, no cause of corrupting the trial --- suffer, with truth made plain and every error taken away, the infamy of that foulness to pass over to where the rest of the evil deeds stand; let the hatred at last depart from him to whom you see fault has never come.

\ciceroSection{84} But, you will say, Oppianicus gave the money to Staienus not for corrupting the trial but for the winning of favour. Do you say this, Attius, endowed with such prudence, with so much use and practice! They say he is most wise to whom what is needed comes of itself into mind; that nearest comes that man who follows another's good devisings. In stupidity it is the contrary. He is less stupid into whose mind nothing comes than he who approves what stupidly comes into another's mind. So Staienus then, the matter recent, when he was being pressed in the throat --- whether he himself thought it up, or (as men were then saying) was warned by Publius Cethegus --- gave that fable of conciliation and favour.

\ciceroSection{85} For you can recall that this was then the talk of men: that Cethegus, because he hated the man and would not have his wickedness at large in the commonwealth, and because he saw that he who had confessed that he had taken money from a defendant secretly and out of order when he was a judge could not be safe --- gave him counsel less faithful. In this if Cethegus was dishonest, he seems to me to have wished to remove an adversary; but if the case was such that Staienus could not deny he had taken coins, while nothing was more perilous nor more foul than to confess for what thing he had taken --- the counsel of Cethegus is not to be reproached.

\ciceroSection{86} But the cause of Staienus then was different, your case now, Attius, is different. He, when he was pressed by the matter, whatever he said would say more honourably than if he should confess what had been done. But that you have now brought back the same thing which then was hooted off and cast out, I wonder. For how could Cluentius then return into favour with Oppianicus, who could not with his mother? In the public records the defendant and accuser were stuck together; the Fabricii had been condemned. Albius could not slip away by another accuser; nor could Cluentius leave the prosecution without the ignominy of a false charge.

\ciceroSection{87} Or rather should he act as a false ally? For that too has to do with corrupting the trial. But what need was there for that of a judge as depository? And altogether why should that whole matter be conducted through Staienus, a man most foreign to each side, most filthy, most foul, rather than through some good man, friend and connection of both? But why do I dispute these things in many words, as if of an obscure matter, when the very money which was given to Staienus by its number and sum shows not only how great it was but even for what it was? I say sixteen judges had to be corrupted that Oppianicus should be acquitted; that to Staienus 640,000 sesterces had been brought. If, as you say, for the sake of winning favour, what does that addition of forty thousand mean? If, as we say, that 40,000 sesterces should be given to sixteen judges, Archimedes could not better have laid it out.

\ciceroSection{88} But it is said that very many trials have been made by Cluentius that the trial was corrupted. Indeed, before this time that very thing in its own name has never been called into trial. So much agitated, so long tossed about is this matter, that on this day for the first time that case has been defended; on this day for the first time truth, trusting in these judges, has sent her voice against hatred. But yet what are these many trials? For I have prepared myself for all things and so set out that I should teach what trials made afterwards about that trial are spoken of: were partly more like a ruin or a storm than a trial and debate; partly weighed not at all against Habitus; partly even were for him; partly were of such a kind that they were neither ever called trials nor reckoned so.

\ciceroSection{89} Here, more that I should keep custom than because you do not of your own will do this, I shall ask of you, gentlemen, that you should attentively hear me while I dispute about each of these. Gaius Junius was condemned, who had been in charge of that trial. Add even this, if it pleases you: he was then condemned when he was the inquisitor of the trial. Not only to the case but not even to the law was any relaxation given through the tribune of the plebs. At the time when it was not lawful to take him from the trial to any other duty of the commonwealth, at that time he himself was snatched away to a trial. But to what trial? For your faces, gentlemen, invite me to say freely now what I had thought ought to be left silent.

\ciceroSection{90} What? That at last --- was it an inquiry or a debate or a trial? I shall reckon it was. Let anyone who wishes today, of that stirred-up people whose wishes were then served, say on what matter Junius pleaded his case. Whomever you ask, this he will answer: because he had taken money, because he had circumvented an innocent man. This is the opinion. But, if it were so, he ought to have been accused on this law on which Habitus is being accused. But he himself was holding the inquiry by that law. Quinctius would have waited a few days. But he wished neither as a private man to accuse, nor with hatred now calmed. You see therefore that all the prosecutor's hope was not in the case but in the time and the power.

\ciceroSection{91} He sought a fine. By what law? Because he had not sworn into the law (which has never been a hurt to anyone), and because Gaius Verres, the city praetor (a sacred and diligent man), did not have the substitution of his lot in that codex which was then brought forth erased. For these causes Gaius Junius was condemned, gentlemen --- most light and most weak ones, which ought not at all to have been brought into trial. So he was crushed not by the case but by the time.

\ciceroSection{92} Do you reckon that this trial ought to harm Cluentius? For what cause? If by the law Junius had not been substituted by lot, or if into some law he had at some time not sworn, was therefore by his condemnation anything being judged about Cluentius? "No," he says; "but he was condemned by those laws, because he had committed something against another law." Those who confess this --- can they at the same time defend that that was a trial? "Therefore," he says, "the Roman people was hostile to Gaius Junius for this: because that trial was thought corrupted through him." Has the case then changed at this time? Is there now another matter, another reasoning of that trial, another nature of the whole business than there was then? I think no thing of those which were done could have been changed.

\ciceroSection{93} What then is the cause that now our defence is heard with such silence, while then to Junius the means of defending himself was snatched? Because then in the case there was nothing save hatred, error, suspicion, daily public meetings stirred up seditiously and popularly. The same tribune of the plebs was prosecuting in the meetings, the same at the benches; he was coming to the trial not only from a public meeting but even with the public meeting itself. Those Aurelian steps, then new, seemed built as if for the theatre of that trial; which when the prosecutor had filled with stirred-up men, there was for the defendant the means not only of speaking but not even of rising.

\ciceroSection{94} Lately before Gaius Orchivius, my colleague, the place was not granted by the judges to Faustus Sulla on a matter of moneys remaining; not because they thought that Sulla was outside the law, or that the case of public money was contemned and cast aside, but because, with a tribune of the plebs prosecuting, they thought there could be no debate on equal terms. What shall I compare? Sulla with Junius? This tribune of the plebs with Quinctius? Or rather time with time? Sulla with the greatest resources, kinsmen, connections, friends, very many clients; while these for Junius were small and weak, sought and gathered by his own labour. Here a modest, modest tribune of the plebs, not only not seditious but even an adversary to the seditious; that one bitter, given to charges, a popular and turbulent man. This time calm and at peace; that one stirred up by all the storms of hatred. Although these things were so, in Faustus's case yet those judges decided that the defendant was pleading on unfair terms, when to his adversary the highest force of power was added to the right of accusation.

\ciceroSection{95} Which reckoning, gentlemen, you ought diligently, by your wisdom and humanity, to consider and deeply to see through: what evil, what peril of how great force the tribunician force can bring upon each one of us, with hatred forged especially and with public meetings stirred up seditiously. In the best of times, by Hercules --- when men kept themselves safe not by popular tossing but by dignity and innocence --- yet neither Publius Popilius nor Quintus Metellus, most distinguished and most ample men, could withstand the tribunician force; much less in these times, with these morals, with these magistrates, can we be safe without your wisdom and the remedies of trials.

\ciceroSection{96} That, then, was not a trial like a trial, gentlemen, was not, in which no measure was used, no manner and custom kept, no case defended. It was that force, and (as I have often said) a kind of ruin and storm, and anything rather than a trial or debate or inquiry. But if there is anyone who thinks that was a trial and that one must abide by these matters judged, he yet ought to sunder this case from that. For from him --- whether because he had not sworn into the law or because by the law he had not been substituted as judge --- a fine, it is said, was sought. But Cluentius's reasoning with those laws by which a fine was sought from Junius can in no part be joined.

\ciceroSection{97} But Bulbus too has been condemned. Add: "of treason," that you may understand that this trial is not joined with that. But this charge was thrown at him. I confess; but also that a legion was solicited by him in Illyricum has been made plain by the letters of Gaius Cosconius and by the testimonies of many --- which charge was the proper one of that inquiry, and a thing held by the law of treason. But this most of all hurt him. That now is divination; if I may use it, see lest my conjecture be much truer. For I so think: that Bulbus, because he, a worthless, foul, dishonest man, polluted with many disgraces, was led into trial, was therefore the more easily condemned. You pick out for me from Bulbus's whole case what is convenient for you, that you may say the judges followed it.

\ciceroSection{98} Wherefore this trial of Bulbus's ought no more to harm this case than those two trials recalled by the prosecutor --- of Publius Popilius and Tiberius Gutta, who pleaded a case of canvassing, who were accused by those who themselves had been condemned of canvassing. Whom I think were not on this account restored as before because they made plain that those men had taken money for judging, but because they proved to the judges that, since they had reproached others in the same kind in which themselves had offended, they ought to come to the prizes of the law. Wherefore I think no one doubts that that condemnation of canvassing can in no part be joined with Cluentius's case and your trial.

\ciceroSection{99} What of this --- that Staienus has been condemned? I do not say at this time, gentlemen --- which I do not know but ought to be said --- that he was condemned of treason; I do not read out the testimonies of most honourable men which were said against Staienus by those who were under the most distinguished man Mamercus Aemilius as legates, prefects, and military tribunes; by whose testimonies it was made plain that, by his work most of all, when he was quaestor, a sedition was forged in the army. Nor do I read those testimonies which were spoken about the 600,000 sesterces which he, when he had taken them under the name of the Safinian trial (just as afterwards in Oppianicus's), kept silent and held back.

\ciceroSection{100} I leave aside both these things and very many others which were said in that trial against Staienus. This I say: that there was then for Publius and Lucius Cominius, Roman knights, honourable and eloquent men, the same controversy with Staienus whom they were prosecuting which I now have with Attius. The Cominii said the same as I say: that Staienus had taken money from Oppianicus to corrupt the trial. Staienus said that he had taken it for the sake of winning favour.

\ciceroSection{101} This conciliation of his was laughed at, and the character of a good man taken on --- as in those gilded statues which he set up at the temple of Juturna, on which he wrote that the kings had been brought back into favour by him. All his frauds and falsehoods were stirred up; his whole life, occupied in such reasoning, was opened. His domestic want, his forensic gain, was brought into the open. The money-taking interpreter of peace and concord was not approved. So Staienus then, while he defended the same as Attius now, was condemned.

\ciceroSection{102} The Cominii, when they handled this which we have handled in the whole case, proved it. Wherefore, if by Staienus's condemnation it was judged that Oppianicus wished to corrupt the trial, that Oppianicus gave the money to a judge for buying votes; and since it has been laid down that in that fault either Cluentius or Oppianicus was; and no penny of Cluentius is found by any trace given to a judge, while Oppianicus's money was taken away from a judge after the trial was made --- can it be in doubt that that condemnation of Staienus is not only not against Cluentius but most of all confirms our case and defence?

\ciceroSection{103} So now I see that the trial of Junius is of such a kind that I think it must be called rather a sedition's incursion, the force of a multitude, a tribunician onset, than a trial. But if anyone should call it a trial, yet he must necessarily confess this: that the fine which was sought from Junius can in no way be joined with Cluentius's case. The Junian therefore was done by force; that of Bulbus and Popilius and Gutta is not against Cluentius; that of Staienus is even for Cluentius. Let us see if any other trial we can bring forth which is for Cluentius. Did Gaius Fidiculanius Falcula, who had condemned Oppianicus, plead a case --- especially when (which was most odious in that trial) he had sat for a few days from substitution? He pleaded, and twice indeed pleaded. For Lucius Quinctius had brought him into the highest hatred by daily seditious and turbulent meetings. In one trial a fine was sought from him, just as from Junius, because he had not sat by the duty of his own decury or by the law. He was accused at a somewhat calmer time than Junius, but on almost the same law and charge. Because no sedition or force or tumult was at large in the trial, in the first hearing he was most easily acquitted. I do not count this acquittal. For nevertheless it can be that, although he had not committed that fine, yet he had taken for judging --- as Staienus's case of money taken for judging on that law was nowhere pleaded. The proper charge of that inquiry was not that.

\ciceroSection{104} What was Fidiculanius said to have done? To have taken from Cluentius 400,000 sesterces. Of what order was he? Senatorial. By the law on which an account is wont to be sought from a senator in that kind --- of extortion --- by that law accused, he was most honourably acquitted. For the case was conducted in the manner of our ancestors, without force, without fear, without peril. All things were said and set out and shown. The judges were brought to think that not only could he honestly be condemned by the man who had not sat throughout, but, even if that judge had known nothing else save what was agreed had been the prejudgements about him, he ought to have heard nothing besides.

\ciceroSection{105} Then those five too, who, hunting the little rumours of unskilled men, had then acquitted him, now did not greatly wish their clemency to be praised. From whom if anyone should ask whether they sat as judges over Gaius Fabricius, they would say they had sat. If they were asked whether he had been accused on any charge save of that poison which had been sought to be given to Habitus, they would deny it. If they were then asked what they judged, they would say they had condemned. For no one acquitted. In the same way, if it had been asked of Scamander, surely they would have answered the same. Although he was acquitted by one vote, but that one no one of those would then have wished it to be called his.

\ciceroSection{106} Which then would more easily render the reckoning of his vote? He who says he was consistent both with himself and with the matter judged, or he who answered that he was lenient toward the chief of the evil deed, most vehement against his helpers and accomplices? About whose vote I ought not to dispute. For I do not doubt that those such men, struck by some sudden suspicion, swerved from their position. Wherefore the pity of those who acquitted I do not reproach. The constancy of those who in judging followed the previous judgements of their own accord (not by Staienus's fraud) I approve. The wisdom indeed of those who said that it was not clear to them I praise --- who could in no way acquit the man whom they had known most guilty and whom they themselves had twice already before condemned, while to condemn him, with so great an infamy of council and the suspicion of so atrocious a thing thrown in, they preferred a little later, with the matter made plain.

\ciceroSection{107} And lest you should judge them wise only from the deed, but rather should approve that what these men did was done most rightly and most wisely from the men themselves, who can be recalled more prudent in talent than Publius Octavius Balbus, more experienced in law, more diligent or more pure in faith, religious obligation, duty? He did not acquit. Who more constant than Quintus Considius, who more experienced in trials and that dignity which ought to be at large in public trials, who more outstanding in virtue, counsel, authority? Not even he acquitted. It is long thus to speak of each one's virtue, and the things which are now known to all do not seek the ornaments of words. What kind of man Marcus Iuventius Pedo was, of that old discipline of the trials; what kind Lucius Caulius Mergus, Marcus Basilus, Gaius Caudinus! All of whom in the public trials, with the commonwealth then flourishing, flourished. Of the same number Lucius Cassius, Gnaeus Heius of equal integrity and prudence; by the vote of none of whom was Oppianicus acquitted. And among all these the youngest in age, in talent and diligence and religion equal to those whom I have named before, Publius Saturius, was of the same opinion.

\ciceroSection{108} O singular innocence of Oppianicus! Toward which defendant who acquitted is reckoned canvassing, who delayed cautious, who condemned constant. These things, with Quinctius then acting, were shown neither in the public meeting nor in the trial. For neither he himself suffered them to be said, nor through the stirred-up multitude could anyone stand to speak. So he himself, after he overthrew Junius, abandoned the whole case. For in those few days he himself became a private man, and he understood that men's zeal had cooled. But if, in those days through which he prosecuted Junius, he had wished to prosecute Fidiculanius, the means of answering would not have been given to Fidiculanius. And at first indeed he was threatening all those judges who had condemned Oppianicus.

\ciceroSection{109} You knew the man's insolence; you knew his spirits and tribunician airs. What hatred was there, immortal gods! What pride, what ignorance of self, how grave and how unbearable arrogance! Who took even this very thing bitterly --- from which all those things were born --- that Oppianicus was not granted as a gift to him and to his defence; just as if it ought not have been sign enough that he was abandoned by all who had betaken himself to that patron. For at Rome there was the highest abundance of patrons, of most eloquent and most ample men, of whom surely some one would have defended a Roman knight, noble in his town, if he had thought such a case could honourably be defended.

\ciceroSection{110} For Quinctius indeed --- when had he ever before pleaded a case, at almost fifty years of age? Who had ever seen him not only in the place of patron but of a more polished supporter? Who, because he had taken the rostra (long empty, that place since the coming of Lucius Sulla deserted from the tribunician voice), and had recalled the multitude (long unaccustomed to public meetings) to the likeness of the old custom, was therefore for a little while more pleasant to a certain kind of men. And the same one in how great hatred he was afterwards with those very men through whom he had ascended into a higher place!

\ciceroSection{111} Nor unjustly. For see to it that you recall not only his morals and arrogance but also his face and dress and even that purple cast down to his ankles. He, as if it were not in any way to be borne that he had departed conquered from a trial, brought the matter from the benches to the rostra. And do we already complain often that for new men there are not great enough rewards in this state? I deny that they were ever anywhere greater. Where, if anyone born in an ignoble place so lives that he seems able to keep the dignity of nobility by virtue, he comes thither, as long as industry with innocence has gone with him.

\ciceroSection{112} But if anyone leans on this one thing --- that he is ignoble --- he often goes further than if the same man with the same vices were most noble. As Quinctius (for I shall say nothing about the rest), if he had been a noble man, who could have borne him with that pride and intolerance? Because he was in that place, they so bore him that, if he had something good from nature, they thought it ought to profit him; but his pride and arrogance they reckoned should rather be laughed at, on account of the lowness of the man, than feared. But that I may return thither: at the time when Fidiculanius was acquitted, you who recall trials made --- what then do you reckon was judged? Surely that he had judged gratis.

\ciceroSection{113} But he had condemned, but he had not heard the whole case, but in all the public meetings by Lucius Quinctius the tribune of the plebs he had been vehemently and often harassed. All those Quinctian things, then, the judges judged unjust, false, turbulent, popular, seditious. Be it so: Falcula could be innocent. Now therefore someone condemned Oppianicus gratis; now Junius did not substitute by lot those who would condemn with money taken; now someone can not have sat from the start and yet have condemned Oppianicus gratis. But if Falcula was innocent, I ask who is guilty. If he condemned gratis, who took? I deny that anything was thrown at any of those which was not thrown at Fidiculanius; that there was anything in Fidiculanius's case which was not the same in the rest's.

\ciceroSection{114} Either you must reproach this trial, whose prosecution leans on matters judged, or, if you grant this to be true, confess that Oppianicus was condemned gratis. Although it ought to be a great enough proof that out of so many judges, with Falcula acquitted, no other was made defendant. Why do you gather for me men condemned of canvassing on another law, on fixed charges, with very many witnesses? Since they themselves first ought to have been accused of extortion rather than of canvassing. For if in canvassing trials this hurt them, when they were pleading on another law, surely, if they had been brought on the proper law of this offence, it would have hurt much more.

\ciceroSection{115} Then if so great force was in this charge that, on whatever law each of those judges had been made defendant, he yet should perish by this blow, why, in such a multitude of accusers, with such great rewards, were the rest not made defendants? Here is brought forth what ought not to be called a trial: that against Publius Septimius Scaevola a fine was assessed under that name. What custom is in this matter, since I am speaking before most experienced men, I ought not to teach in many words. For never was that diligence which is wont to be brought to the rest of trials brought to the same defendant condemned.

\ciceroSection{116} In assessing fines the judges generally either, because they think the man whom they have once condemned is hostile to them, if any capital suit has been brought against him, do not admit it; or, because they reckon they have already finished, when they have judged about the defendant, they attend more carelessly to the rest. So very many have been acquitted of treason for whom, condemned of extortion, fines on treason had been assessed. And we see this happen daily: that, with a defendant condemned of extortion, those to whom it has been laid down (in assessing fines) that the moneys had come, the same judges acquit. When this happens, trials are not rescinded; but it is laid down that the assessment of fines is not a trial. Scaevola was condemned on other charges, with very many witnesses from Apulia. With every contention it was fought that this fine should be assessed for capital. Which thing, if it had had the weight of a matter judged, he would afterwards, by the same or by other enemies, have been made defendant on this very law.

\ciceroSection{117} What follows is what they call a trial, but our ancestors never named a trial nor regarded as much as a matter judged: the animadversion and authority of the censors. About which, before I begin to speak, I must say a very few words about my duty, that the reasoning both of this man's peril and of the rest of duties and friendships may seem to be kept by me. For I have friendship with the brave men who were last censors, both. With the one indeed (as most of you know) there is great use, and the highest connection has been set up between the duties of each.

\ciceroSection{118} Wherefore whatever I shall have to say of their entries, I shall say it with such mind that I would have all my speech reckoned not about their deed but about the censorial reasoning. From Lentulus, my familiar, who is named by me for honour's sake on account of his outstanding virtue and the highest honours which he has obtained from the Roman people, I shall easily get this, gentlemen: that the faith and diligence, the force of mind and freedom of speaking which he himself is wont to bring to the perils of friends, of these he should grant me to take so much for myself as I cannot pass over without his peril. By me yet, as is fair, all things shall be said cautiously and step by step, that neither the faith of this defence may seem abandoned nor anyone's dignity hurt or friendship violated.

\ciceroSection{119} I see therefore, gentlemen, that the censors animadverted on certain judges of that Junian council, when they entered this very cause. Here first I shall set forth this common point: that this state has never been so content with censorial animadversions as with matters judged. Nor shall I spend time on a known matter. As example I shall set down this one thing: that Gaius Geta, when he had been cast out of the senate by Lucius Metellus and Gnaeus Domitius the censors, was himself afterwards made censor; and that he, whose morals had been reproached by the censors, this same man was afterwards in charge of the morals both of the Roman people and of those who had animadverted on him. But if that were thought a trial, as the rest condemned by a foul trial are for ever deprived of all honour and dignity, so for men marked by ignominy there would be neither approach to honour nor return into the senate-house.

\ciceroSection{120} Now if any freedman of Gnaeus Lentulus or Lucius Gellius shall be condemned of theft, he, with all ornaments lost, shall never recover any part of his honour. But those whom Lucius Gellius and Gnaeus Lentulus themselves --- two censors, most distinguished men and wisest --- marked under the name of theft and of moneys taken, these not only returned to the senate but were even acquitted in the trials of those very things. Our ancestors wished no man to be a judge --- not only of any man's standing, but not even of the smallest matter of money --- save him on whom the parties had agreed. Wherefore in all laws in which it has been excepted on what causes either it shall not be lawful to take a magistracy, or to be chosen as judge, or to accuse another --- this thing of ignominy has been passed over. For they wished the cause of fear, not the punishment of life, to be in that power. So I shall not only show what now you see,

\ciceroSection{121} that very often the censorial entries have been taken away by the votes of the Roman people, but also by the trials of those who under oath were bound to determine with greater religious obligation and diligence. First the judges, the senators and Roman knights, in very many defendants now, against whom it has been entered that they took moneys against the laws, have obeyed their own religious obligation rather than the censors' opinion. Then the city praetors, who under oath ought to enrol all the best as picked judges, have never reckoned that for that thing the censorial ignominy ought to be a hindrance to them.

\ciceroSection{122} The censors themselves often have not stood by the judgements of earlier censors --- if you wish to call those judgements. And even the censors themselves among themselves so reckon their own judgements of such weight that the one not only reproaches the other's judgement but even rescinds it: that the one wishes to remove from the senate, the other keeps in and reckons worthy of the most ample order; that the one orders some entered as aerarii or moved from a tribe, the other forbids it. Wherefore why does it come into your mind to call these things "trials," which you see rescinded by the Roman people, repudiated by sworn judges, neglected by magistrates, changed by those who have got the same power, differ between colleagues?

\ciceroSection{123} Since these things are so, let us see what at last the censors are said to have judged about that corrupted trial. And first let us decide whether, because the censors have entered, it is so; or, because it was so, they have entered. If because they have entered, see what you do, lest into each one of us you grant in future a kingly power to censors; lest the censorial entry be able to bring no less of calamity to citizens than that most bitter proscription; lest we hereafter as much fear the censorial pen, whose point our ancestors have blunted by many remedies, as that dictatorial sword. But if what has been entered

\ciceroSection{124} ought to be grave because it is true, let us ask whether it is true or false. Let the censorial authorities be set aside; let what is not in the case be taken from the case. Show what money Cluentius gave, whence he gave, in what manner he gave; finally show one trace of money set out by Cluentius. Then convince that Oppianicus was a good man, an upright man; that nothing was ever thought otherwise of him, that nothing was prejudged. Then embrace the censorial authority; then defend that their judgement was joined to the matter.

\ciceroSection{125} But while it shall stand fixed that Oppianicus was the man who was judged to have corrupted the public records of the town with his own hand, who erased a will, who saw to the sealing of a false will under a substituted person, who killed the man in whose name it was sealed, who killed the uncle of his own son in slavery and in chains, who saw to it that his townsmen were proscribed and killed, who took in marriage the wife of the man he had killed, who gave money for an abortion, who killed his mother-in-law, who killed his wives, who at one time killed his brother's wife and the children hoped for and his brother himself, who finally killed his own children, who, when he wished to give poison to his stepson, was caught manifestly; with whose ministers and accomplices condemned, he himself, brought to trial, gave money to a judge for corrupting the votes of the judges --- while these things, I say, shall stand fixed about Oppianicus, and the charge of Cluentian money shall be held by no proof, what is there that this censorial will or opinion (whatever it was) can seem to help you or to oppress this innocent man?

\ciceroSection{126} What then did the censors follow? Not even themselves --- to put it most gravely --- will say anything else save talk and rumour. They will say that they learned nothing by witnesses, nothing by records, nothing by any grave proof; finally that they laid down nothing with the case learned. But if they had done so, yet it ought not to be so fixed that it could not be plucked out. I shall not use the supply of examples (which is the highest); I shall not bring forth an old matter, not some powerful or well-favoured man. Lately a slight man, a clerk to the aediles, Decimus Matrinius, when I had defended him before Marcus Junius and Quintus Publicius the praetors, and Marcus Plaetorius and Gaius Flaminius the curule aediles, I persuaded that they should under oath choose as clerk him whom these very same censors had left as aerarius. For when in the man no fault was found, they reckoned that what he had deserved (not what had been laid down about him) had to be asked.

\ciceroSection{127} For these things which they have entered about the corrupted trial --- who is there who reckons that they have been learned enough by them and judged diligently? Against Manius Aquilius and against Tiberius Gutta I see it has been entered. What is this? They judge that two alone were corrupted by money. The rest, of course, condemned gratis. Therefore he was not circumvented, not pressed down by money, not (as it was held in that Quinctian public meeting) all those who condemned Oppianicus must be set in fault and suspicion. Two alone I see by the authority of the censors are judged kin to that foulness. Or let them bring forth that they have learned about the rest something which they had learned about these two.

\ciceroSection{128} For this is least to be approved: that they have transferred to the censorial entries and authority an example from the military custom. For our ancestors so laid it down: that, if a disgrace of military matter had been committed by many, by lot it should be animadverted upon some, that fear, of course, should reach all, but punishment a few. Why does the same fit the censors in choice of dignity and in the judgement of citizens and in the animadversion of vices? For a soldier who has not held his place, who has feared the onset and force of the enemy, can the same afterwards be both a better soldier and a good man and a useful citizen. Wherefore lest in war, on account of fear of the enemy, anyone should fall into fault, an ampler fear of death and punishment was set up by our ancestors. But lest too many should undergo capital punishment, therefore that drawing of lots was prepared.

\ciceroSection{129} Will you do the same as censor in choosing the senate? If there shall be more who took money for condemning an innocent man, will you not animadvert on all, but pluck off as you wish, and pick a few of many for ignominy by lot? Will the senate-house then have, with you knowing and seeing, as senator, the Roman people as judge, the commonwealth as citizen without ignominy any man who for the ruin of an innocent man has changed his faith and religious obligation for money? And shall he who, drawn on by a price, has snatched fatherland, fortunes, children from an innocent citizen, not be branded by the censorial mark of strictness? You are the prefect of morals; you the master of the old discipline and strictness, if either you keep anyone in the senate knowingly stained by such great crime, or lay down that he who is in the same fault need not be afflicted with the same punishment. Or what condition of punishment our ancestors wished to be set forth in war for the timidity of a soldier --- shall you set up the same in peace for the wickedness of a senator? But if this example had to be transferred from the military matter to the censorial animadversion, that very thing ought to have been done by lot. But if to draw lots for punishment and to commit men's offence to the judgement of fortune is least censorial, surely in the offence of many a few ought not to be plucked off for ignominy and foulness.

\ciceroSection{130} But we all understand that in those entries a kind of popular wind was being sought. The matter had been tossed about in a public meeting by a seditious tribune. The case unknown, that thing had been approved by the multitude. To no one was it lawful to speak against; finally no one laboured to defend the contrary part. Furthermore those trials had come into great hatred. For a few months afterwards another vehement hatred was at large in the trials, from the marking of the tablets. The censors did not seem to be able to be passed over and neglected as a stain on the trials. Men whom they saw infamous in the rest of vices and in every disgrace, them they wished to mark by this entry too; and so much the more because at that very time, in those censors' time, the trials had been shared with the equestrian order, that they should seem by the ignominy of fit men to have reproached the matter by their authority.

\ciceroSection{131} But if it had been allowed me to plead this case before those very censors (or to anyone else), to men endowed with such prudence I should surely have proved it. For the matter shows that they themselves had nothing learned, nothing ascertained; that out of all that entry a kind of rumour and popular applause was being sought. For against Publius Popilius (who had condemned Oppianicus) Lucius Gellius entered: that he had taken money for the condemning of an innocent man. Now of how great divination is this very thing --- to know that the defendant whom he had perhaps never seen was innocent, when most wise men, gentlemen (to say nothing of those who condemned), with the case learned, said to themselves it was not clear!

\ciceroSection{132} But suppose so. Gellius condemns Popilius; he judges that he had taken money from Cluentius. Lentulus denies this. For Popilius, because he was the son of a freedman, he did not enrol in the senate, but he leaves him a senator's place at the games and the rest of the ornaments and frees him from all ignominy. Which when he does, he judges by his own vote that Oppianicus was condemned gratis. And the same Popilius afterwards Lentulus, in a canvassing trial, in evidence most diligently praises. Wherefore, if Lentulus did not stand by Lucius Gellius's judgement, nor was Gellius content with Lentulus's standing --- and if each censor thought that he must not abide by his colleague's opinion --- what is there why any of us should think all censorial entries fixed and ratified for ever?

\ciceroSection{133} But they animadverted on Habitus himself. For no foulness, indeed, for no --- I shall not say vice, but error --- of his whole life. For there can be no man more pure than this man, more honest, more diligent in keeping all duties; nor do they say otherwise, but they followed that same fame of the corrupted trial. Nor do they themselves think otherwise than we wish to be thought of his modesty, integrity, virtue. But they thought the prosecutor could not be passed over, when it had been animadverted upon the judges. About which whole matter, if I shall bring forth one deed from all antiquity, I shall say no more.

\ciceroSection{134} For the example of the highest and most distinguished man, Publius Africanus, does not seem to me must be passed over. Who, when he was censor and Gaius Licinius Sacerdos had come forward in the census of the knights, said with a clear voice (so that the whole assembly could hear) that he knew that man had perjured himself in set words; if anyone wished to speak against him, he would use his own testimony. Then, when no one spoke against, he ordered him to lead his horse across. So he, by whose discretion both the Roman people and foreign nations were wont to be content, himself by his own knowledge was not content for the ignominy of another. But if it had been allowed for him to do this in this man's case, easily before those very judges he would have withstood both false suspicion and the popularly stirred-up hatred.

\ciceroSection{135} One thing yet there is which most disturbs me, to which place I scarcely seem able to answer: that you read out a clause from the will of Gnaeus Egnatius the father --- a most honourable and most wise man, of course --- who had on this account disinherited his son: because he had taken money for the condemnation of Oppianicus. Of which man's lightness and inconsistency I shall say no more. This very will which you read out is of such a kind that he, when he was disinheriting that son whom he hated, joined as co-heirs to the son whom he loved most foreign men. But you, Attius, see, I think, diligently whether you wish the censorial judgement to be grave, or Egnatius's. If Egnatius's, what the censors entered about the rest is light; for Gnaeus Egnatius himself, whom you wish to be grave, they cast out of the senate. But if the censorial: this Egnatius whom his father disinherited by a censorial entry, the censors --- when they were casting out the father --- kept in the senate.

\ciceroSection{136} But the whole senate judged that that trial was corrupted. In what way? It took up the matter. Or could it have repudiated a thing of such kind brought before it? When a tribune of the plebs, the rouser of the people, had almost called the matter to hands; when an excellent and most innocent man was being said to have been circumvented by money; when the senatorial order was blazing with hatred --- could nothing be decreed; could that stirring-up of the multitude be repudiated without the highest peril of the commonwealth? But what was decreed? How justly, how wisely, how diligently! \textsc{If there are any by whose work it has come about that a public trial should be corrupted}. Does the senate seem to judge that it was so done, or, if it has been done, to bear it heavily and gravely? If Aulus Cluentius himself were asked his opinion about the trials, he would speak no other than they spoke whose opinions you say condemned Cluentius.

\ciceroSection{137} But I ask of you whether out of that senatorial decree Lucius Lucullus the consul, a most wise man, brought forward that law; whether a year afterwards Marcus Lucullus and Gaius Cassius, whom as consuls-designate the senate had then decreed the same. They did not bring it forward. And what you charge was done by Habitus's money (and confirm by no slightest suspicion) was done first through the equity and wisdom of those consuls --- so that what the senate had decreed for putting out that present fire of hatred, they thought ought not afterwards to be referred to the people. Then the Roman people itself, which (stirred before by Lucius Quinctius's feigned complaints) had demanded that thing and that proposal, the same, moved by the tears of Gaius Junius's son, a small boy, with the greatest cry and concourse repudiated the whole inquiry and law.

\ciceroSection{138} From which it could be understood (which has often been said): that, as the sea (which by its nature is calm) is stirred and troubled by the force of the winds, so the Roman people of itself is at peace; by the voices of seditious men it is stirred as by the most violent storms. There is also a remaining authority, very great, which I have shamefully almost passed over: it is said to be mine. Attius read from some speech, which he was saying was mine, a kind of urging of judges to honourable judging, and a recalling both of other trials which had not been approved and of that very Junian trial; just as if I had not from the start of this defence said that that trial was odious, or, when I was disputing about the infamy of trials, could have passed over what was so popular at that time.

\ciceroSection{139} For my part, if I said anything of that kind, I neither recalled it as known nor said it in evidence; and that speech was rather of my time than of judgement and authority. For when I was prosecuting and had set before me at the start that I should move the spirits both of the Roman people and of the judges, and was bringing forth all the offences of trials not from my own opinion but from men's rumour, that thing which had been so popularly agitated I could not pass over. But he errs vehemently if anyone reckons that he has our authorities sealed in our speeches which we have made in trials. For all those are of cases and of times, not of the men themselves or of the patrons. For if the cases themselves could speak for themselves, no one would call in a speaker. Now we are called in that we should say those things which are not laid down by our authority, but are drawn from the matter itself and the case.

\ciceroSection{140} They say that that ingenious man, Marcus Antonius, was wont to say "for this cause he had never written any speech, that, if at any time anything had been said by him not necessary, he could deny that he had said it." Just as if, if anything has been said or done by us, it should not be taken in by the memory of men, unless we have entrusted it to writings. For my part in that kind I more willingly follow the authority of many, and especially of the most eloquent and most wise man Lucius Crassus. Who, when he was defending Gnaeus Plancus, with Marcus Brutus accusing (a vehement and crafty man in speaking), when Brutus had had two readers set up to read out from Crassus's two speeches alternate chapters contrary to one another --- because in the dissuasion of that proposal which was being brought against the Narbonensian colony he detracted as much as possible from the senate's authority, while in the persuasion of the Servilian law he adorned the senate with the highest praises; and Brutus had read out many things said somewhat sharply against the Roman knights from that speech, that the spirits of those judges might be set on fire against Crassus --- Crassus is said to have been somewhat moved.

\ciceroSection{141} So in answering, first he set out the reasoning of each time, that the speech might seem to be made from the matter and from the case. Then, that Brutus might be able to understand what man and endowed not only with what eloquence but also with what charm and what wit he had provoked, he himself called up three readers with the little books which Marcus Brutus, the father of that prosecutor, left on the civil law. When their beginnings were being read out --- those which I think are known to you: "It chanced that we were in the country, in the Privernate territory, I and Brutus my son" --- he was demanding the Privernate estate. "We were at Albanum, I and Brutus my son" --- he was demanding the Alban. "It chanced when we had sat down at Tiburtum, I and Brutus my son" --- he was looking for the Tiburtine estate. He was saying that Brutus, a wise man, since he saw the worthlessness of his son, wished to attest by writing what estates he was leaving him. But that, if he had been able to write honourably that he had been with his son of that age in the baths, he would not have passed it over. Yet that he was looking for those baths from him not out of his father's books but out of the records and the census. Crassus then so avenged himself on Brutus that the latter regretted his reading. For he had borne heavily that he was reproached in those speeches which he had made about the commonwealth, in which perhaps consistency is more demanded. But I do not bear it heavily that those things were read out.

\ciceroSection{142} For neither were they foreign from that time which then was, nor from that case which was then being conducted. Nor have I taken on myself any burden, when I so spoke, which would forbid me to defend this case honourably and freely. But if I should wish to confess that I have now learned the case of Aulus Cluentius, that before I had been in that popular opinion, who could reproach this? Especially, gentlemen, when from yourselves it is most just to obtain what I from the start asked and now ask: that, if you have brought hither some grave opinion about that trial, this --- with the case seen through and the whole truth learned --- you should set down.

\ciceroSection{143} Now, since I have answered to all the things which were said by you, Titus Attius, about the condemnation of Oppianicus, you must necessarily confess that opinion has much deceived you, that you reckoned I would defend the case of Aulus Cluentius not by his own deed but by the law. For this you very often said: that it was so reported to you that I had in mind to defend this case by the protection of the law. Is that so? By friends, of course, we are imprudently betrayed; and there is some one of those whom we reckon as friends who carries our counsels to our adversaries. Who has reported this to you, who was so dishonest? To whom indeed have I told? No one (I think) is at fault. Surely the law itself reported it to you. But have I seemed to you so to defend that in the whole case I have made any mention of the law; have I seemed to defend this case otherwise than as if Habitus were held by the law? Surely (as a man ought to confirm) no place has been passed over by me of clearing him of that odious charge.

\ciceroSection{144} What then? Some will perhaps ask whether it displeases me to drive off a peril of life by the protection of the laws. To me indeed, gentlemen, it does not displease; but I use my own custom. In the trial of an honourable and prudent man, I am wont to use not only my own counsel but to obey much also the counsel and will of him whom I defend. For when this case was brought to me, who ought to know those laws to which we are called and in which we are at large, I said to Habitus at once: "Whoever has conspired that any one should be condemned --- he is free, but our order is held." And he began to ask and beseech me that I should not defend him by law. When I said what seemed best to me, he led me over to his opinion. For he was confirming, weeping, that he was not more eager for the keeping of his citizenship than for his standing.

\ciceroSection{145} I obeyed the man --- yet I therefore did so (for we ought not always to do this), because I saw that the case could most copiously be defended by itself without the law. I saw in this defence which I have now used more dignity; in that, which he did not wish me to use, less labour would there be. But if nothing else were being attempted save that we should hold this case, I should have closed with the law read out. Nor does that speech of Attius's move me, which says that it is an unworthy deed if a senator who has circumvented anyone in trial be held by the laws; if a Roman knight has done the same, that he be not held.

\ciceroSection{146} Although I should grant you that this is unworthy (whose kind I shall already see), you must necessarily grant me that it is by far the more unworthy, in that state which is held by laws, to depart from the laws. For this is the bond of the dignity which we enjoy in the commonwealth, this the foundation of liberty, this the spring of equity. The mind and spirit and counsel and opinion of the state is set in the laws. As our bodies cannot, without mind, so the state cannot, without law, use its parts as sinews and blood and limbs. The ministers of laws are the magistrates; the interpreters of the laws are the judges; finally we are all servants of the laws, that we may be free.

\ciceroSection{147} What is it, Quintus Naso, why you sit in that place? What force is there by which these judges, endowed with such dignity, are held in by you? You, gentlemen, why out of so great a multitude of citizens are so few you casting your sentence on men's fortunes? By what right has Attius said what he wished? Why has the means of speaking been given me so long? What do those clerks mean, what those lictors, what the rest whom I see attending this inquiry? I think all these things happen by law, and that this whole trial (as I said before) is ruled and administered as if by a kind of mind of the law. What then? Is this inquiry alone so governed? What of Marcus Plaetorius and Gaius Flaminius "for assassins"; what of Gaius Orchivius "of embezzlement"; what of mine "of extortion"; what of Gaius Aquilius's, before whom now a case of canvassing is being pleaded; what of the rest of the inquiries? Look round all the parts of the commonwealth: you will see all things happen under the command and prescription of laws.

\ciceroSection{148} If anyone before me, Titus Attius, should wish to make you defendant, you would cry out that you were not held by the law of extortion. Nor would this refusal of yours be a confession of money taken, but a turning aside of toil and unlawful peril. Now what is being attempted and what right is being set up by you, see. The law (by which law this inquiry has been set up) bids the inquisitor of the trial --- that is, Quintus Voconius --- with those judges who shall have fallen to him by lot (he calls upon you, judges) inquire about poison. Inquire against whom? It is unbounded. \textit{Whoever has made, sold, bought, kept, given}. What does the same law at once add? Read. \textit{And about his life let inquiry be made}. Whose? Of him who has conspired, agreed? It is not so. What then is it? Tell. \textit{Of him who has been military tribune of the first four legions, or who quaestor, or tribune of the plebs} --- thereafter he named all the magistrates --- \textit{or who has spoken his sentence in the senate, or shall have}. What then? \textit{Who of them has conspired, conspires, agreed, agrees that any man should be condemned in a public trial}. "Who of them"? Of whom? Of those, of course, who have been written above. What difference there is in the manner of writing --- although it is plain, the law itself yet teaches us. For where it binds all mortals, it speaks thus: \textit{Who has made, shall have made bad poison}. All men, women, children, slaves are called into court. If the same had been wished about conspiracy, it would have added: "Or who shall have conspired." Now it is so: \textit{And about his life let inquiry be made who has held a magistracy or who has spoken his sentence in the senate, who of them has conspired, conspires}. Is Cluentius such?

\ciceroSection{149} Surely he is not. Who then is Cluentius? One who yet did not wish his case to be defended by the law. So I cast aside the law; I obey Cluentius. Yet to you, Attius, I shall answer a few things which are sundered from this man's case. For there is something in this case which Cluentius reckons concerns himself; there is something which I reckon concerns me. He thinks it of his own interest that he should be defended by the matter itself and by the business done, not by the law. But I reckon it of my interest that I should not seem to have been overcome by Attius in any debate. For I have not this case alone to plead. To me this labour is set forth before all who can be content with this means of defence. I do not wish anyone of those who are present to think that I, by being silent, approve what has been said by Attius about the law. Wherefore, Cluentius, in your matter I obey you. I neither read out the law nor speak in this place for you; but those things which I think are looked for from me, I shall not abandon.

\ciceroSection{150} It seems unfair to you, Attius, that not all should be held by the same laws. First, that I should confess this is most unfair, it is of such a kind that those laws should be changed --- not that we should not obey those which are. Then what senator has ever refused this --- that, when he had attained a higher step of dignity by the kindness of the Roman people, on this account he should think he ought to use harder conditions of the laws? How many advantages there are which we lack, how many troublesome and hard things which we undergo! And all these things are yet made up by the advantage of honour and amplitude. Turn now to the equestrian order and into the rest of the orders the same conditions of life: they will not bear them. For they think that fewer snares of laws and conditions and trials ought to be set forth for them, who could either not ascend to the highest place of the state, or did not seek it.

\ciceroSection{151} And to leave aside all the other laws by which we are held but the rest of the orders are freed, this very law: "lest anyone be circumvented in trial," Gaius Gracchus brought in. He brought in that law on behalf of the plebs, not against the plebs. Afterwards Lucius Sulla, a man most far from the people's cause, when he was setting up the inquiry of that thing on this very law on which you now judge, did not yet dare to bind by a new kind of inquiry the Roman people, which he had received free from this kind. But if he had thought it could be done, on account of the hatred which he had against the equestrian order he would have done nothing more willingly than have brought all that bitterness of his proscription which he used against the old judges into this one inquiry.

\ciceroSection{152} Nor is anything now being attempted --- believe me, gentlemen, and look forward to what is to be foreseen --- save that Roman knights should be shut into the peril of this law. Nor is this attempted by all but by a few. For those senators who easily keep themselves safe by integrity and innocence (such as, to say truly, you are, and the rest who have lived without greed) wish the knights, the nearest to the senatorial order in dignity, to be most close-joined in concord. But those who wish themselves to be able to do all things, and that besides there should be nothing in any man or in any order, by this one fear think that they will reduce the Roman knights into their power: if it be set up that about those who have judged a matter trials of this kind can be made. For they see the authority of this order being confirmed; they see the trials being approved. With this fear set forth, they trust that they can pluck out the sting of your strictness.

\ciceroSection{153} For who would dare to judge truly and bravely about a man endowed with somewhat greater resources, when he saw that he himself would have to plead his case for what he had conspired or agreed? O brave men, the Roman knights, who withstood that most distinguished and most powerful man, Marcus Drusus, the tribune of the plebs, when he was attempting nothing else with all that nobility which then was, save that those who had judged a matter should be called into court by inquiries of this kind! Then Gaius Flavius Pusio, Gnaeus Titinius, Gaius Maecenas, those oaks of the Roman people, and the rest of the same order did not do the same as Cluentius now (so that they should reckon they took on themselves any fault by refusing), but most openly fought back; when both they refused and openly, most bravely and most honourably, said that they could have come to the most ample place by the judgement of the Roman people, if they had wished to bring their pursuits to seeking honours. That they had seen what splendour was in that life, what great ornaments, what dignity. Which they had not despised, but had been content with their own and their fathers' order, and had preferred to follow that calm and quiet life, removed from the storms of hatreds and trials of this kind.

\ciceroSection{154} Either, they said, the whole of life ought to be restored to them for seeking honours; or, since that could not be, the condition of life which, having sought office, they had abandoned, ought to remain. That it was unfair that those who, on account of the multitude of perils, had passed over the ornaments of honours and were deprived of the people's kindnesses, should yet not be without the perils of new trials. That a senator could not complain of this: because, with that condition set forth, he had begun to seek office; and because there were very many ornaments by which he could soften that trouble --- place, authority, splendour at home, name and favour among foreign nations, the bordered toga, the curule chair, the marks of office, the fasces, armies, commands, provinces. In which things our ancestors wished both the highest rewards to be set forth for things rightly done, and more perils for sins. They were not refusing this --- that they should be accused on the same law on which now Habitus is being accused, which then was the Sempronian, now is the Cornelian; for they understood that by that law the equestrian order was not held; but they were labouring lest by a new law they should be bound.

\ciceroSection{155} Habitus has not even at this time refused this --- that he should render the reckoning of his life on that law on which he is not held. Which condition if pleases you, let us all attempt that this inquiry should be carried as soon as possible into all orders. Meanwhile, by the immortal gods! since we hold all our advantages, our rights, our liberty, finally our safety from the laws, let us not depart from the laws. At the same time let us consider how unfair it is that the Roman people should now do something else --- has entrusted to you the commonwealth and its own fortunes, is without care, does not fear lest by a law which it has never itself ordered, and by an inquiry from which it thinks itself loosed and free, it be bound through a few judges.

\ciceroSection{156} For Titus Attius, a good and eloquent young man, so pleads his case: that all citizens are held by all laws. You hear and attend in silence, as you ought. Aulus Cluentius, a Roman knight, pleads his case on a law on which only senators and those who have held a magistracy are held. To me through him it is not lawful to refuse and to set up the protections of my defence in the citadel of the law. If Cluentius shall hold the case (as, leaning on your equity, we trust), all will think (which it shall be) that he held it on account of his innocence, since he was so defended; that in the law which he had not wished to touch there had been no protection.

\ciceroSection{157} Here now is something which has to do with me, of which I spoke before, which I owe to the Roman people; since this is the standing of my life: that all my care and toil is set in defending men's perils. I see how great and how perilous and how unbounded an inquiry is being attempted by the prosecutors, when they try to transfer to the Roman people that law which has been written into our order. In which law is: "Whoever has conspired" --- which how widely it spreads you see. "Has agreed" --- equally uncertain and unbounded. "Has consented" --- that indeed is both unbounded and obscure and hidden. "Or has spoken false testimony" --- who of the Roman commons has ever spoken testimony for whom this peril, with Titus Attius as author, is not seen prepared? For I confirm that surely no one will speak, if this trial is set forth before the Roman commons.

\ciceroSection{158} But this I promise to all: if to anyone perhaps trouble shall be stirred up by this law who is not held by the law, if he shall wish to use me as defender, I shall defend his case by the protection of the law and shall most easily prove it either before these judges or before such as these; and I shall use the whole defence of the law which now to use, by him to whose will I must obey, is not granted me. For I ought not to doubt, gentlemen, that, if any case of such kind be brought to you, of one who is not held by the law --- even if he should seem odious or offensive to many, even if you should hate him, even if you should be unwilling to acquit him --- yet you would acquit him and obey rather your religious obligation than your hatred.

\ciceroSection{159} For it is the part of a wise judge to consider that as much has been allowed him by the Roman people as has been entrusted and committed; and to remember that not only power has been given him, but faith has been placed in him. He can acquit whom he hates, condemn whom he does not hate; and always not what he himself wishes, but what the law and religious obligation compel, to consider; to mark on what law the defendant is summoned, about what defendant he learns, what matter is at issue in the inquiry. When these things are to be seen, then it is the part of a great man, gentlemen, and a wise one, when he has taken that tablet for judging, not to reckon himself alone or to think that whatever he has desired is lawful for him; but to have in council the law, religious obligation, equity, faith; lust however, hatred, envy, fear, and all desires to remove; and most to value the consciousness of his mind, which we have received from the immortal gods, which cannot be torn from us. Which, if it shall be the witness of the best counsels and deeds in our whole life, we shall live without any fear and with the highest honour.

\ciceroSection{160} If Titus Attius had either learned or thought of these things, surely he would not even have tried to say --- which he handled in many words --- that the judge ought to lay down what seems good to him and not be bound by the laws. About which things I seem to have said for Cluentius's will too much, for the dignity of the matter too little, for your prudence enough. The rest are very few; which, because they had to do with your inquiry, on this account they decided that they must be feigned by them and brought forth, lest they be found the foulest of all, if they had brought into court nothing save hatred. And that you may reckon I have necessarily handled in many words those things of which I have already spoken, attend the rest. Surely you will understand that those things which could be shown in few have most briefly been defended.

\ciceroSection{161} He said that to Gnaeus Decidius the Samnite (he who was proscribed) a wrong was done in his calamity by this man's household. By no one was he more liberally treated than by Cluentius. The resources of this man relieved him in his most inconvenient affairs; and this both he himself and all his friends and connections knew. That this man's bailiffs brought force and hand on the shepherds of Ancharius and of Pacenus. When some controversy of shepherds had arisen, as is wont, in the cattle-tracks, Habitus's bailiffs defended the matter of their master and the private possession. When complaint was made, the cause shown to them, they parted without trial and controversy.

\ciceroSection{162} When by the will of Publius Aelius a kinsman had been disinherited, this man --- more foreign --- was instituted heir. Publius Aelius did so on the deserving of Habitus; nor was he present at the making of the will, and that will was sealed by this man's enemy Oppianicus. That he denied a legacy to Florus from a will. It is not so. But when 30,000 sesterces had been written for 40,000, and security did not seem enough to him, he wished him to enter something received to his liberality. At first he denied it was owed; afterwards without controversy he paid. That after the war the wife of a certain Ceius the Samnite was demanded back from him. When he had bought the woman from the slave-dealers, at the time when he first heard she was free, he gave her back to Ceius without trial.

\ciceroSection{163} That there is a certain Ennius whose goods Habitus holds. This Ennius is some needy false accuser, the hireling of Oppianicus, who for very many years has been quiet. Then at some time he sued Habitus's slaves of theft. Lately he began to demand from Habitus himself. He in that private trial, believe me, perhaps with you yourselves as patrons, will not escape a false-charge. And even (as is reported to us) you suborn a man of many guests, a certain Aulus Bivius, an innkeeper from the Latin road, to say that hands were brought against him by Cluentius and his slaves in his own shop. About which man it is not necessary to say anything yet now. If he shall invite us (as he is wont), we shall so receive the man that he shall bear it heavily that he turned aside from the road.

\ciceroSection{164} You have, gentlemen, what in the whole case the prosecutors have, after meditating eight years, gathered about the morals of Aulus Cluentius (whom they wish to be an odious defendant) --- how light by their very kind, how false in matter, how short for answer! Learn now what has to do with your oath, what is of your trial, what burden has been laid on you by that law by which, compelled, you have come together hither: about the charges of poisoning. That all may understand in how few words this case could have been closed, and how many things have been said by me which had to do most with this man's wish, least with your judgement.

\ciceroSection{165} It has been thrown that Gaius Vibius Capax was taken away by poison by this Aulus Cluentius. Conveniently the man is at hand, endowed with the highest faith and every virtue, Lucius Plaetorius, a senator, who was the guest-friend and familiar of that Vibius. At his house Vibius lived at Rome; at his he was sick; at his he died. But Cluentius is heir. I say that he died intestate, and that the possession of his goods was given by the praetor's edict to this Numerius Cluentius (whom you see), the son of his sister, a most modest young man and among the chief honourable, a Roman knight.

\ciceroSection{166} The other charge of poisoning is against this young man Oppianicus: that, when in his marriage according to the custom of the Larinates a multitude of men was lunching, poison was prepared by the counsel of Habitus; that, when it was being given in mead, a certain Balbutius, his familiar, intercepted it, drank it, and at once died. If I were so handling this as if a charge had to be cleared by me, I should say in many words those things which now my speech briefly runs through.

\ciceroSection{167} What has Habitus ever admitted into himself, that this great deed should not seem to shrink from him? But why did he greatly fear Oppianicus, when that man could not say any word at all in this very case, while accusers cannot fail this man as long as his mother lives? --- which you will understand. Or that nothing might be lacking from the cause of his peril, that a new charge should be added to his case? But what was the time of giving the poison, on that day, in that throng? Through whom further was it given? Whence taken? What then is this intercepting of the cup? Why was it not given afresh? Many things there are which can be said. But I shall not allow myself to seem, by not speaking, to have wished to speak. The matter defends itself.

\ciceroSection{168} I deny that that young man, whom you have said died at once with the cup drained, died on that day at all. A great charge and a shameless lie! See through the rest. I say that he, since he had come to that lunch undigested and (as that age bears) had yet not spared himself, was sick for some days and so died. Who is the witness of this thing? The same who is the witness of his own mourning, his father; the father, I say, of that young man, whom on account of his grief of mind a very slight suspicion could have set up as witness from that place against Aulus Cluentius. He raises this man up by his testimony. Read it. But you, if it is not troublesome, rise for a little. Bear this grief of necessary recollection, in which I shall not longer dwell, since (which was the part of an excellent man) you have made it that your mourning should bring calamity and a false charge to no innocent man.

\ciceroSection{169} One charge of this kind yet remains to me, gentlemen, from which you can see what was said by me at the start of my speech: that whatever ill Aulus Cluentius has seen through these years, whatever trouble and business he has at this time, all has been forged from his mother. That Oppianicus was killed by poison given to him in bread by a certain Marcus Asellius, his familiar; and that this was done by the counsel of Habitus, you say. In which first I ask what cause there was for Habitus that he should wish to kill Oppianicus. For I confess there were enmities; but men wish their enemies to be afflicted with death either because they fear them or because they hate them.

\ciceroSection{170} For what fear was Habitus drawn on to try to undertake so great a deed against himself? What was there that anyone should now fear Oppianicus, afflicted with punishment for his evil deeds and cast out of the state? What did he fear? Lest he be assailed by a ruined man; or lest he be accused by a condemned man; or lest he be hurt by an exile's testimony? But if Habitus, because he hated his enemy, on this account did not wish him to enjoy life, was he so stupid that he reckoned what then he was living was life --- that of a condemned man, of an exile, of one abandoned by all, whom on account of the fierceness of his mind no one wished to receive under his roof, no one to approach, no one to address, no one to look upon?

\ciceroSection{171} Did Habitus then envy this man's life? If he hated him bitterly and deeply, ought he not to have wished him to live as long as possible? Did the enemy hasten his death, which alone was his refuge of calamity in his ills? Who, if he had had any spirit and virtue (as many brave men in such grief have often done), would himself have taken his life. For him why would the enemy wish to offer that which he himself ought to wish for himself? For now indeed what at last did death bring him of ill? Unless perhaps we are led by trifles and tales to think that he is bearing the punishments of the impious among the dead, and has met more enemies there than he has left here; that by the Avengers of his mother-in-law, of his wives, of his brother, of his children he has been driven headlong into the seat and region of the wicked. If these things are false (which all understand), what at last did death snatch from him save the sense of grief?

\ciceroSection{172} Come now: through whom was the poison given? Through Marcus Asellius. What had this man with Habitus? Nothing. And, since he had used Oppianicus most familiarly, rather even there was a strife. Did he then most of all entrust both his crime and that man's peril to him whom he knew was more offended with him, most familiar with Oppianicus? Why then have you, who have been roused by piety to accusing, suffered this Asellius to be unavenged for so long? Why have you not used Habitus's example: that through him who had brought the poison there should be a prejudgement against this man?

\ciceroSection{173} Now indeed how improbable, how unfamiliar, gentlemen, how new --- that the poison was given in bread! Could it more easily have been given in the cup? Could it have lain hidden more secretly in some part of bread than if the whole had been dissolved in the drink? Could it have spread more quickly when eaten than when drunk into the veins and into all parts of the body? Could it more easily escape notice in bread, if it had been observed, than in the cup, when it was so mixed that it could in no way be distinguished?

\ciceroSection{174} But he died by sudden death. If it had been so done, that thing yet, on account of the like chance of many, would have least firm a suspicion of poison. If it were suspicious, yet it would have to do with others rather than with Habitus. But on this very point men lie most shamelessly. That you may understand it, learn both his death and how, after his death, the charge was sought against Habitus by his mother.

\ciceroSection{175} When wandering and exiled and shut out from every quarter, Oppianicus had betaken himself to the Falernian estate to Lucius Quinctius, there first he fell into disease and was sick vehemently enough and long. When Sassia was with him, and was using a certain Sextus Albius, a settler, a strong man (who used to be with him), more familiarly than a most dissolute man could allow with his fortune unhurt, and was reckoning that the right of chaste and lawful marriage had been taken away by her husband's condemnation, a certain Nicostratus, a faithful little slave of Oppianicus, very inquisitive and least lying, is said to have been wont to report many things to his master. Meanwhile Oppianicus, when he was now recovering and could no longer bear the wickedness of the settler in the Falernian estate, and had set out hither toward the city --- for he was wont to keep something hired outside the gate --- is said to have fallen from his horse, and the man (of weak health) to have struck his side vehemently. And after he came to the city with a fever, in a few days he died. The reasoning of his death, gentlemen, is such that either it has no suspicion or, if it has any, that turns within the walls of his domestic crime.

\ciceroSection{176} After his death Sassia, the unspeakable woman, at once began to set snares against her son. She decided to hold an inquiry about her husband's death. She bought from Aulus Rupilius (whom Oppianicus had used as doctor) a certain Strato, just as that Habitus might do the same as Habitus had done in buying Diogenes. She said she would inquire about this Strato and about a certain Asclas, her own slave. Besides, that slave Nicostratus, whom she reckoned had been too talkative and too faithful to his master, she demanded from this young man Oppianicus for the inquiry. He, since he was at that time a boy, and that inquiry was said to be set up about his father's death --- although he reckoned that slave both was well-disposed to himself and had been to his father --- yet dared to refuse nothing. Friends and guest-friends of Oppianicus and of the woman herself, many honourable men adorned with all things, are called in. By all the most vehement tortures the inquiry is held. When the spirits of the slaves had been tried both with hope and with fear, that they should say something in the inquiry, yet (I think) led by the authority of the supporters, they remained in the truth and said they knew nothing.

\ciceroSection{177} The inquiry on that day, by the opinion of friends, was dismissed. After a long enough interval again they are called in. The inquiry is held afresh. No force of the most fierce tortures is passed over. The supporters turn aside and now scarcely can bear it. The cruel and harsh woman is in a fury that what she had thought was by no means coming forward as she had hoped. When now the torturer and the tortures themselves had been worn out, and yet she would not make an end, one of the supporters, a man both adorned with people's honours and endowed with the highest virtue, said that he understood that this was being done not so that the truth should be found, but that they should be compelled to say something false. After the rest had approved this, by the opinion of all it was laid down that they seemed to have been inquired into enough.

\ciceroSection{178} Nicostratus is given back to Oppianicus. She herself sets out for Larinum, mourning with her own people that now surely her son would be safe, to whom not only no true charge but not even a feigned suspicion would come; to whom not only the open assault of enemies, but not even the hidden snares of his mother could harm. After she came to Larinum, she who had pretended that she was persuaded that poison had been given before to her husband by that Strato, at once gave him a furnished and adorned shop at Larinum for practising medicine. One, a second, a third year Sassia was quiet, so that she seemed to wish and desire some calamity for her son rather than to plot and contrive it.

\ciceroSection{179} Then meanwhile, in the consulship of Quintus Hortensius and Quintus Metellus, that she might draw to this prosecution this Oppianicus (who was doing something else and thinking nothing of this kind) against his will, she promised her daughter to him --- her whom she had taken up from her son-in-law --- so that, with him bound by the marriage and at the same time tied by the hope of a will, she could have him in her power. At about this very time that doctor Strato did a theft in the house and a slaughter of this kind. When in the building there was a chest in which he knew there was some money and gold, by night he killed two fellow-slaves sleeping and threw them into the fish-pond. He himself cut out the bottom of the chest, and carried off ten thousand sesterces and five pounds of gold, with one of the slaves, a not large boy, knowing.

\ciceroSection{180} The next day, with the theft known, all suspicion was being moved on those slaves who were not appearing. When that cutting-out of the bottom in the chest was being noticed, men were asking how it could possibly have been done. One of Sassia's friends recalled that he had lately seen at some auction, among the small things, a hooked little saw with teeth on every side and twisted, by which that thing seemed possible to be so cut around. To be brief: it is sought out by the auctioneers; that little saw is found to have come to Strato. With this beginning of suspicion arisen and Strato openly charged, that knowing boy was struck with fear and reported the whole matter to his mistress. The men were found in the fish-pond. Strato was thrown into chains; and even in his shop the coins (by no means all) were found.

\ciceroSection{181} An inquiry on the theft is set up. For what else could anyone suspect? Or do you say this: that, with the chest plundered, the money taken, not all recovered, the men killed, an inquiry was set up about Oppianicus's death? Whom do you prove this to? What is there of less likeness to truth that you could bring forth? Then (to leave aside the rest), three years after Oppianicus's death was inquiry being held about his death? And kindled by ancient hatred, she demanded that same Nicostratus then without cause in the inquiry. Oppianicus at first refused. Afterwards, when she threatened that she would lead the daughter away, that she would change her will, the most faithful slave to a most cruel woman not into the inquiry he handed but plainly to the punishment he gave over.

\ciceroSection{182} Three years afterwards then an inquiry on the husband's death was being held afresh. But about what slaves? Some new thing, I trust, had been thrown forward; some new men had been called into suspicion. About Strato and about Nicostratus. What? Had inquiry not been made at Rome about these men? Indeed --- woman now mad not by disease but by crime --- when you had held an inquiry at Rome, when by the opinion of Titus Annius, Lucius Rutilius, Publius Saturius, and the rest of the most honourable men it had been laid down that they seemed to have been inquired into enough --- did you about the same matter, three years afterwards, about the same men, with no man called in (I do not say "no man," lest you say a settler perhaps was at hand, but with no good man), against your son's life try to hold an inquiry?

\ciceroSection{183} Or do you say this --- for it comes into my mind what could be said, although remember that this has not yet been said: that, while the inquiry was being held about the theft, Strato confessed something about poison? In this one way, gentlemen, often the truth, pressed down by the wickedness of many, comes up, and the defence of innocence, shut off, breathes again. Because either those who in fraud are cunning do not dare as much as they devise; or those whose audacity stands out and is thrown forth are abandoned by the counsels of malice. But if confident astuteness or cunning audacity were one, it could scarcely in any way be withstood. Was there no theft? But nothing was clearer at Larinum. Or did suspicion not have to do with Strato? But he was both charged from the saw and informed against by the knowing boy. Was not this being done in the inquiring? What other cause then was there of inquiring? Or what you must say, and what then Sassia kept saying: that, while inquiry was being made about the theft, then Strato in those same tortures spoke about poison?

