% Pro A. Caecina (pro-caecina) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.
%
% A civil case in 69 BC: Aulus Caecina, dispossessed by force from a
% Volaterran estate by Sex. Aebutius, sues under the interdictum de vi
% armata (the praetorian interdict on armed force). Cicero argues that
% Caecina was driven off by an armed band and the law accordingly
% requires Aebutius to give the estate back.

\ciceroSection{1} If, as much as audacity has weight in the field and in deserted places, so much shamelessness had force in the forum and in the courts, no less now in the case would Aulus Caecina yield to the shamelessness of Sextus Aebutius than then he yielded to his audacity in the use of force. But he reckoned both that this was the part of a considerate man --- not to fight with arms about a matter on which the law ought to decide --- and that this was the part of a constant man, that he should overcome by law and trial the man with whom he had been unwilling to contend by force and arms.

\ciceroSection{2} And to me indeed Aebutius seems to have been audacious above all in calling men together and in arming them; then shameless in the trial, not only because he dared to come into court (for that, although it is dishonestly done in an open matter, is yet by malice now customary), but because he did not hesitate to confess that very thing of which he is accused. Unless perhaps he had this reasoning: that, since, if force had been used by custom, he would not have been the higher in keeping possession (because it had been done against right and custom, that Aulus Caecina had fled with his friends, terrified by fear); now too in the trial, if the case were defended by the custom and institution of all, we should not be lower in pleading. But if it should depart from the custom, he, the more shamelessly he should plead, by so much should depart the higher. As if dishonesty in trial could do the same as boldness in force, or we had not the more willingly yielded to audacity then, by so much that we should the more easily withstand shamelessness now.

\ciceroSection{3} So it is by far another reasoning, recoverers, that I come now to plead the case in this hearing than I had come at the start. For then the hope of our case lay set in my defence; now in the adversary's confession; then in our witnesses; now indeed in theirs. About whom I was before in trouble lest, if they were dishonest, they should say something false; if they were thought honest, they should prove what they had said. Now my mind is at the calmest. For if they are good men, they help me, when they say under oath what I uncharge them with not under oath. But if they are less fit, they do not hurt me; for, when either credit is given them, credit is given to this very thing which we charge; or if faith is not given them, the faith of the adversary's witnesses is taken away.

\ciceroSection{4} But yet, when I consider their pleading of the case, I do not see what could be more shamelessly said. But when I consider your hesitation in judging, I fear lest what they seem to have done shamelessly they have done craftily and cunningly. For if they had denied that force had been used by armed men, they would easily be caught by most honourable witnesses in a plain matter. But if they had confessed and had defended that what at no time can rightly be done was at that time rightly done by them, they hoped --- which they have attained --- that they would lay upon you a cause of deliberating and a just delay and scruple in judging. At the same time they thought this would happen which is most unworthy: that in this case judgement should seem to be made not on the dishonesty of Sextus Aebutius, but on the civil law. In which thing, if I had to plead the case of one Aulus

\ciceroSection{5} Caecina alone, I should profess myself a fit enough defender, because I should answer for my faith and diligence. Which, when they are in the actor of a case, there is nothing in a matter especially open and simple for which an excelling talent should be needed. But since I must speak about that right which has to do with all, which has been established by our ancestors and kept up to this time --- which, when taken away, not only some part of right is lessened, but the very force which is most adverse to right seems to be confirmed by the trial --- I see this is a case for the highest talent, not that what is before our eyes should be shown, but that, if any error in such a matter should be set before you, all should not rather think me to have failed the case than you to have failed your scruple. Although I so persuade myself,

\ciceroSection{6} recoverers, that you have already twice doubted in this same case not so much on account of the obscure and doubtful reasoning of the law as because, since this trial seems to have to do with that man's whole standing, you have looked for a delay before condemning, and at the same time given him space to gather himself. Which, since it is now in the custom and good men like you in judging do this, must perhaps be the less reproached, but yet must seem the more lamented; because all trials have been devised either for ending controversies or for punishing evil deeds, of which the one is the lighter, because it both hurts less and very often is judged by a domestic arbiter; the other is the most vehement, because it has to do with the graver matters and asks not the honorary work of a friend but the strictness and force of a judge.

\ciceroSection{7} What is graver, and for whose sake trials have most of all been set up, that has now by bad custom been loosened. For as each thing is most foul, so it must be most chiefly and most quickly punished; but the same, because there is peril of standing, is most slowly judged. How then does it agree, that what was the cause of setting up the trial should be the cause of delay in judging? If anyone, what he has promised by sponsion (in which matter he has bound himself by one word) does not do, he is condemned by a quick trial without any scruple of the judge. He who through guardianship or partnership or business committed or by reasoning of trust has cheated anyone --- in this, the greater the offence, is the punishment to be the slower?

\ciceroSection{8} "It is a foul trial." Out of a foul deed, indeed. See then how unfairly it falls out, that, because the matter is unworthy, therefore foul standing follows; because foul standing follows, therefore an unworthy matter is not punished. And if any judge or recoverer should say to me: "You could have contended by a lighter action; you could have come to your right by an easier and more convenient trial. Wherefore either change your action, or do not press me to judge anyway" --- he would seem either too timid for it to be just that he should be a brave judge, or too eager for it to be just that he should be a wise one, if either he should lay down for me in what manner I should pursue my right, or himself dare not judge what has been laid before him. For if the praetor who gives the trials never lays down for the plaintiff what action he wishes him to use, see how unfair it is, with the matter now established, for the judge to ask what could have been done or what can be, not what has been.

\ciceroSection{9} But yet I should obey your too great kindness, if I could recover my right by some other reasoning. Now indeed, who is there who either thinks force used by armed men ought to be left unpunished, or could show me some lighter action for that thing? Out of which kind of fault, as they keep crying, either trials of wrong or capital trials have been established --- in this can you reproach our atrocity, when you see that nothing else has been done save that possession is sought back by the interdict? But, whether the peril of his standing or the doubt of the law has made you the slower up to now in judging, the cause of one of these matters you yourselves have already taken away by the trial often put off; the cause of the other I shall surely take away from you on this day, that you should not longer doubt about our controversy and the common right.

\ciceroSection{10} And if perhaps I shall seem to have sought the beginning of laying out the matter further off than the reasoning of the law about which the trial is and the nature of the case demand, I ask that you forgive me. For Aulus Caecina labours not less lest he seem to have pleaded by extreme right than lest he should not have got his fixed right. There was Marcus Fulcinius, recoverers, of the town of Tarquinii. Who both at home was reckoned among the chief honourable, and at Rome ran a not ignoble banking-house. He had in marriage Caesennia, of the same town, born in the highest place and a most well-respected woman, as he himself in life showed by many things and at his death declared in his will.

\ciceroSection{11} To this Caesennia he sold an estate in the Tarquinian field in those most difficult times of payment. Since he was using his wife's dowry counted out --- that the matter should be the safer for the woman --- he saw to it that the dowry was placed in that estate. Some time afterwards, the banking-house being now closed, Fulcinius bought certain estates lying near and joining to that estate of his wife. Fulcinius dies. (For many things which are in the matter, because they are remote from the case, I shall pass over.) By his will he makes heir the son whom he had by Caesennia. He bequeaths to Caesennia the use and fruit of all his goods, that she should enjoy them with her son.

\ciceroSection{12} A great honour from her husband would have been pleasant to the woman, if it had been allowed to be long-lasting. For she would have enjoyed the goods with him whom she wished to be heir to her own goods, and from whom she herself reaped the greatest fruit. But fortune took this fruit from her in good time. For in a short time the young Marcus Fulcinius died. He made Publius Caesennius his heir; he bequeathed to his wife a great weight of silver, and to his mother the greater part of his goods. So the women were called to share.

\ciceroSection{13} When this auction of the inheritance had been set up, that Aebutius, who already long had been fed on Caesennia's widowhood and loneliness and had insinuated himself into her familiarity by this method, that he should take up some business or other for her with some gain to himself, if any should arise, was at this time also occupied in those reckonings of the auction and partition. He even thrust himself in and pushed in, and brought Caesennia into this opinion, that the inexperienced woman thought nothing could be cunningly done where Aebutius was not present.

\ciceroSection{14} Which character you now know from daily life, recoverers --- the flatterer of women, the agent of widows, the over-litigious defender, worn out at court, witless and stupid among men, knowing of the law and crafty among women --- this character lay upon Aebutius. For he was such an Aebutius to Caesennia. Lest perhaps you ask, was he a kinsman? Nothing more foreign. A friend handed down by her father or husband? Nothing less. Who, then? He, that one whom I have just deformed --- a voluntary friend of the woman, joined by no connection but by feigned duty and pretended busyness, by labour at times more opportune than ever faithful.

\ciceroSection{15} When the auction had been set up at Rome, as I had begun to say, the friends and kinsmen of Caesennia urged --- which came also into the woman's mind --- that, since the power of buying that Fulcinian estate (which was joining to her old estate) was given, there was no reason to lose such an opportunity, especially when money was owed her from the partition; that she could nowhere place it better. So the woman decided to do this. She entrusts the buying of the estate for herself --- to whom? To whom do you suppose? Or does it not come into the minds of all of you that this was the office of him who was prepared for all the woman's affairs, without whom nothing could be carried on diligently enough or carefully enough?

\ciceroSection{16} You attend rightly. The business is given to Aebutius. He is at the table; Aebutius bids. Many buyers are deterred, partly by the favour of Caesennia, partly by the price. The estate is knocked down to Aebutius. Aebutius promises the money to the banker, by which testimony now the most excellent man uses to argue that it was bought for himself. As if either we deny that it was knocked down to him, or anyone then would doubt that it was being bought for Caesennia, when most knew it, almost all had heard it. Who had not heard could attain it by guess --- since money was owed Caesennia from that inheritance, since it was most expedient to place that money in estates, since there were estates which suited the woman the best, since they were for sale, since the bidder was one for whom no one would wonder that he was doing service for Caesennia, while no one could suspect he was buying for himself. With this purchase done, the money is paid by Caesennia.

\ciceroSection{17} Of which thing this man thinks no account can be rendered, because he himself has done away with the records; but that he has the banker's records in which the money was entered as outlay to him, and brought back as received. As if it ought to be done otherwise. When all things had been so done as we maintain, Caesennia held the estate and let it out. And not so long after she married Aulus Caecina. To put it in a few words: with a will made, the woman dies. She makes Caecina heir of eleven-twelfths and a half-twelfth; of two sextulae, Marcus Fulcinius, freedman of her former husband; on Aebutius she sprinkles a sextula. This sextula she wished to be his wage of attendance and trouble, if she had taken any. But he reckons by this sextula that he keeps a handle for all controversies.

\ciceroSection{18} Already at the start he dared to say that Caecina could not be heir of Caesennia, because he was of worse right than the rest of the citizens, on account of the misfortune of the Volaterrans and the civil calamity. So a timid and inexperienced man, who had neither spirit nor counsel enough, would not have thought the inheritance of so much worth that his citizenship should come into doubt. He yielded, I trust, to Aebutius as much as the latter wished he should have of Caesennia's goods. Indeed, as it was worthy of a brave and wise man, so he ground down and crushed his false charge and stupidity.

\ciceroSection{19} When he was in possession of the goods, and when this man was making too great a thing of his sextula, Caecina demanded an arbiter for the dividing of the estate as heir. And in those few days, after he sees that he can scrape nothing from Aulus Caecina by the terror of suit, he gives notice to him at Rome in the forum that that estate of which I spoke before --- whose buyer I have shown him to have been by Caesennia's order --- is his own and was bought by him for himself. What say you? Is that estate his which Caesennia, without any controversy, for four years (that is, from the time when the estate was sold), as long as she lived, possessed? "For," he says, "the use and fruit of that estate had been given to Caesennia by her husband's will."

\ciceroSection{20} When he was so maliciously bringing this new kind of suit, it pleased Caecina, on the opinion of his friends, to fix on a day on which they should come to the place itself, and Caecina should by custom be put off from the estate. They confer; the day is taken from the convenience of each. Caecina with friends comes on the day to the fortlet of Axia, from which place that estate of which we are speaking is not far. There he is made certain by many that Aebutius had gathered together very many men, free and slave, and armed them. While they partly wondered and partly did not believe it, behold Aebutius himself comes into the fortlet. He gives notice to Caecina that he has armed men; that he will not leave, if he should approach. It pleased Caecina and his friends still to try, as long as it should seem possible to be done with safety to life.

\ciceroSection{21} They go down from the fortlet; they set out to the estate. It seems rashly committed; but, I think, this was the cause: no one thought he would commit a thing as rashly as he was threatening it in words. And he sets armed men against all the entries by which approach could be made not only to that estate of which there was the controversy, but even to that nearest one about which there was no question. So at first, when Caecina wished to enter the old estate, because that was the nearest approach, men in numbers, armed, stood in the way.

\ciceroSection{22} Driven from this place, Caecina yet went where he could to that estate in which by the agreement force ought to be used. Olives in a straight row mark out the boundary of the farthest part of that estate. When approach was being made to these, this man with all his troops was at hand, and called his slave by name Antiochus to him, and with a clear voice ordered him to kill the man who entered that row of olives. The most prudent man (in my opinion) Caecina yet seems to me to have had in this matter more spirit than counsel. For although he saw the multitude of armed men and had heard that voice of Aebutius which I have recalled, yet he came nearer; and now, as he was going within the boundary of that place which the olives marked, he fled the onset of the armed Antiochus and the others, the weapons and the rush. At the same time his friends and supporters, terrified by fear, throw themselves into flight (in what manner you have heard their witness say).

\ciceroSection{23} With these things so done, the praetor Publius Dolabella issued the interdict, as is the custom: \textsc{on force used by armed men}, with no exception, only that he should restore from where he had cast him out. He said he had restored. The sponsion was made. About this sponsion you must judge. It was most to be desired by Caecina, recoverers, that he should have no controversy; secondly, that he should have it not with such a dishonest man; thirdly, that he should have it not with such a stupid one. For we are no less raised up by his stupidity than hurt by his dishonesty. He was dishonest, because he gathered men, armed them, used force with them gathered and armed. In this he hurt Caecina; in this same he raises him up. For for those very things which he most dishonestly did, he took witnesses, and uses those witnesses in the case.

\ciceroSection{24} So I am resolved, recoverers, before I come to my own defence and my witnesses, to use his confession and witnesses --- who confesses and so willingly confesses that he seems not only to admit but even to profess, recoverers: "I called men together; I gathered, I armed; with the terror of death and peril of life I stood in the way that you should not approach. With iron," he says, "with iron" (and he says this in court) "I cast you back and frightened you off." What? What do the witnesses say? Publius Vetilius, kinsman of Aebutius, that he had come as supporter to Aebutius with armed slaves. What besides? That there were several armed men. What else? That Aebutius had threatened Caecina. What shall I say of this witness, save this, recoverers: that you should not on that account believe him less because the man is reckoned less fit, but that you should believe him because from that side he says what is most foreign to that case? Aulus

\ciceroSection{25} Terentius, the second witness, charges not only Aebutius but himself also with a most evil deed. Of Aebutius he says this, that there were armed men. About himself he proclaims this: that he had ordered Antiochus, Aebutius's slave, to fall on Caecina with iron as he came up. What more shall I say of this man? On whom, although Caecina asked it of me, I never wished to say this, lest I should seem to charge him with a capital matter; about him I now doubt how to speak or to be silent, when he proclaims this of himself under oath.

\ciceroSection{26} Then Lucius Caelius said not only that Aebutius was with several armed men, but that Caecina had come there with very few supporters. Shall I detract from this witness? Whom I demand that you believe equally with my own witness. Publius Memmius followed, who recalled his own no small kindness to Caecina's friends, to whom he said that he had given a way through his brother's estate by which they could escape, when all were terrified by fear. To this witness I shall give thanks, that both in the matter he showed himself merciful and in his testimony scrupulous.

\ciceroSection{27} Aulus Atilius and his son Lucius Atilius said that there were armed men there and that they had brought their own slaves; even this besides: that, when Aebutius was threatening Caecina with trouble, then there Caecina demanded that the eviction be made by custom. Publius Rutilius said this same thing, and the more willingly that, after such a long time, in some trial credit might be reckoned to be given to his testimony. Two other witnesses said nothing about force, but about the matter itself and the buying of the estate: Publius Caesennius, the warrantor of the estate, not so much grave by authority as by body; and the banker Sextus Clodius, whose surname is Phormio, no less black nor less bold than that Phormio of Terence, said nothing of force, nothing besides which had to do with your trial.

\ciceroSection{28} But in the tenth place a witness awaited and reserved for the last spoke: a senator of the Roman people, the splendour of the order, the honour and ornament of the courts, the model of old religious sanctity, Fidiculanius Falcula. Who, when he had come so vehement and sharp that he not only wounded Caecina by his perjury but even seemed to me to be angry, I rendered so calm and soft that he did not dare (as you remember) to say a second time how many miles his estate was from the city. For when he had said less than 1,000 paces, the people with laughter cried back that they were the same. For all remembered how much he had received in the Albian trial.

\ciceroSection{29} What shall I say against him save what he cannot deny: that he came onto the council of a public inquiry when he was not a judge of that council; and on that council, when he had not heard the case and there was the power of putting it off, said that the matter was clear to him; when he had wished to judge about an unknown matter, preferred to condemn rather than to acquit; when, if they should condemn one less, the defendant could not be condemned, was at hand not to learn the case but to fill out the condemnation? Can anything graver be said against any man, than that he was led for a price to condemning a man whom he had never seen or heard of? Or can anything surer be charged than that which he against whom it is charged does not even by a nod try to weaken?

\ciceroSection{30} But yet that witness --- that you might easily understand he had not been present in mind, when the case was pleaded by them and the witnesses were speaking, but had only been thinking for so long about some other defendant --- when all before him had said as witnesses that there were several armed men with Aebutius, alone said there were not. He seemed to me at first the old hand to understand splendidly what stood in the way of the case, and only to err because he was weakening all the witnesses who had said before him. When suddenly --- behold the same as he is wont --- he said that two slaves alone had been armed. What can you do with this man? Will you not sometimes grant that by the excuse of the highest stupidity he should plead off the hatred of the highest dishonesty?

\ciceroSection{31} Whether, recoverers, did you not believe these witnesses, when you held that nothing was clear? But it was not in controversy that they were saying the truth. Or, in a multitude gathered, in arms, in weapons, in the present fear of death and the manifest peril of slaughter, was it doubtful to you whether some force seemed to be there or not? In what matters then can force be understood, if not in these? Or did that defence seem to you distinguished: "I did not cast him out, but I stood in the way; for I did not allow him to enter the estate, but I set armed men against him, that you should understand: had you set foot on the estate, you must at once perish?" What say you? He who has been frightened off, put to flight, driven by arms --- does he not seem to have been cast out?

\ciceroSection{32} Of the word we shall see later. Now let us set down the matter itself, which they do not deny, and ask the right and action of that matter. The matter laid down is what is not denied by the adversary: that Caecina, when he had come on the appointed day and at the appointed time that force and putting-off should be done by custom, was driven and shut out by force, by men gathered and armed. Since this is agreed, I, a man inexperienced in the law, ignorant of business and suits, think I have this action, that by the interdict I keep my right and pursue your wrong. Suppose I err in this and can in no way attain by this interdict what I wish. I wish to use you in this matter as my master.

\ciceroSection{33} I ask whether there is any action for this matter or none. Men ought not to be called together on account of a controversy of possession; a multitude ought not to be armed for the sake of keeping right; nor is anything so hostile to right as force, nor anything so opposed to equity as men gathered together and armed. Since it is so and the matter is of such a kind that it seems above all to be animadverted on by the magistrates, I again ask whether there is any action for this matter or none. Will you say there is none? I am eager to hear: he who in peace and leisure, when he has made a band, has prepared troops, has gathered a multitude of men, armed them, drawn them up; has driven, put to flight, turned aside men unarmed who had come to the appointed place for the sake of trying their right, by arms, by men, by terror and the peril of death --- shall this man say:

\ciceroSection{34} "I have indeed done all the things you say, and they are both turbulent and rash and perilous. What then? I have done them with impunity. For what to do with me out of civil and praetorian right you have not." Indeed? Recoverers, will you hear this and suffer it to be said before you more often? When our ancestors were of such great diligence and prudence that they fixed and pursued the rights of all things in all matters --- not only of so great matters but even of the most slight --- could they have passed over this one kind, perhaps the greatest, that, if anyone had compelled me to leave my home by arms, I should have an action; if anyone had forbidden me to enter, I should have none? I do not yet dispute about Caecina's case; I do not yet speak about the right of our possession. So much about your defence, Gaius

\ciceroSection{35} Piso, I ask. Since you so say and so lay down: that, if Caecina, when he was on the estate, had been cast out from there, then by this interdict he ought to have been restored; but now indeed he was in no way cast out from where he had not been; that we have got nothing by this interdict --- I ask: if today, as you returned home, gathered and armed men should keep you not only from the threshold and roof of your house but from the very approach and vestibule, what would you do? My friend Lucius Calpurnius warns you to say what he himself said before --- "of wrongs." What has that to do with the case of possession, what with restoring him who ought to be restored, what finally with civil right or with the praetor's notice and animadversion? You will plead the action of wrongs. I shall grant you more: not only have you pleaded but even have him condemned. Will you possess any the more? For the action of wrongs does not attain the right of possession, but tempers the grief of liberty lessened by trial and punishment.

\ciceroSection{36} Will the praetor meanwhile, Piso, be silent about so great a matter? Shall he have no way to restore you to your house? Who, whole days, either forbids force to be used or orders it to be restored when used; who interdicts about ditches, about drains, about the smallest controversies of waters and ways --- shall he suddenly become mute, and in the most atrocious matter have nothing to do? And shall the praetor have no way by custom and example to help Gaius Piso when forbidden his home and roof --- forbidden, I say, by gathered and armed men? For what will he say, or what shall you, having received so notable a wrong, ask? "Whence forbidden by force?" So no man ever interdicted. It is new --- I do not say uncustomary, but altogether unheard of. "Whence cast out?" What will you do, when they shall answer you what you now answer me, that they themselves stood in your way armed lest you should approach the house, that further you could in no way be cast out who had not approached?

\ciceroSection{37} "I am cast out," you say, "if any of mine is cast out at all." Now you do well. For you depart from the words and use equity. For if we wish to follow the very words, in what manner are you cast out, when your slave is cast out? But it is so as you say. I ought to understand that you have been cast out, even if you have not been touched. Is it not so? Come now: if not even any of yours shall have been moved from the place, and all shall have been kept and held in the house, you alone forbidden and frightened off your own house by force and arms --- shall you have this action which we have used, or some other, or none at all? To say that there is no action in such a notable and atrocious matter is neither of your prudence nor of your authority. If perhaps there is some other which has escaped me, say what it is; I am eager to learn.

\ciceroSection{38} If this is the one which we have used, with you as judge we must necessarily win. For I do not fear that you may say: in the same case under the same interdict you ought to be restored, but Caecina ought not. For to whom is it not plain that the goods, fortunes, possessions of all are called back into uncertainty, if in any part the meaning of this interdict has been lessened or weakened, if by the authority of such men the force of armed men shall seem to have been approved by trial --- in which trial it is said that the doubt has been not about arms but about words? Will he hold his case before you who has so defended himself: "I cast you back by armed men, I did not cast you out" --- so that so great a deed shall seem to have lain hidden not in the equity of defence but in one letter?

\ciceroSection{39} Will you lay down that there is no action for this matter, no right of trying it set up: for him who has stood in the way with armed men, who with a multitude gathered has forbidden some man not the entry but the very approach? What then? What weight has this thing, that anything seems to differ or to be set apart in any part: whether, when I have set my foot and made my track in possession, I am then driven out and cast out; or, by the same force and the same arms, I am met before, that I should not only enter but not even look on or breathe in? How does this differ from that, that he is forced to restore who has driven out one who has entered, but he is not forced who has driven back one entering?

\ciceroSection{40} See, by the immortal gods, what right for us, what condition for yourselves, what law finally for the state you wish to set up. Of this kind one action through this interdict which we have used has been set up. If that has no force, or if it does not have to do with this matter, what more carelessly or more stupidly can be said about our ancestors --- who have either passed over the action of so great a matter, or have set up one which by no means in words took in the case and reasoning of the right? This is perilous: that this interdict be loosened. It is full of snares for all that any matter of such a kind be set up that, when it has been done by arms, it cannot rightly be rescinded. But yet that is the most foul: that the most prudent men should be condemned of so great stupidity, that you should judge that the right and action of this matter did not come into our ancestors' minds.

\ciceroSection{41} "We may complain," he says; "yet by this interdict Aebutius is not held." Why so? "Because force was not done to Caecina." Can it be said in this case, where there were arms, where a gathered multitude of men, where men drawn up and placed at fixed places with iron, where threats, perils, and terrors of death --- there was no force? "No one," he says, "was killed or wounded." What say you? When we are speaking of a controversy of possession and of the contention of right between private men, will you deny that force was used, if slaughter and killing were not done? But the largest armies, I say, have often been driven and put to flight by the very terror and onset of the enemy without anyone's not only death but even wound.

\ciceroSection{42} For, recoverers, that is not the only force which reaches our body and life, but even much greater is that which, with the peril of death thrown in, by terror often moves the mind from its place and fixed station. So men wounded are often weakened in body, yet do not yield in spirit nor leave the place which they have determined to defend; while others are driven away whole. So that there is no doubt that greater force has been used against him whose mind has been terrified than against him whose body has been wounded.

\ciceroSection{43} But if we say that those armies have been driven by force which have fled by fear and often by a slight suspicion of peril; and if not only by the impact of shields nor by the clash of bodies nor by the close blow nor by the casting of weapons, but often by the very cry of soldiers or by the drawing-up and sight of standards great forces have been driven --- as both we have seen and we have heard --- shall what is called force in war not be called so in peace? And what is reckoned vehement in military matters, shall it be judged light in civil right? And what moves armed armies, shall it not seem to have moved the company of toga'd men? And shall a wound of body declare that force more than a terror of mind? And shall wounding be looked for, when flight is established?

\ciceroSection{44} For your witness said this: that to our supporters, terrified by fear, he showed a place by which they might flee. Those who not only fled but even sought a safe way for the flight itself --- shall force not seem to have been used to them? Why then did they flee? On account of fear. What did they fear? Force, of course. Can you then deny the beginnings, when you grant the ends? You confess that they fled terrified. You give the cause of the flight as the same that we all understand: arms, a multitude of men, the rush and onset of the armed. Where these are granted to have been done, there shall force be denied to have been done?

\ciceroSection{45} But this is now old and customary by the example of our ancestors in many matters: that, when men came to use force, if they had seen any armed however far off, they should at once depart, having taken witness, when they could most excellently make a sponsion: \textit{unless force has been used against the praetor's edict}. Indeed? Is it enough to know that there are armed men, that you should prove force used; is it not enough to fall into their hands? Shall the sight of armed men avail to prove force, but the rush and onset shall not? Shall he who has departed prove force used to himself more easily than he who has fled?

\ciceroSection{46} But this I say: if, as soon as Aebutius said in the fortlet to Caecina that he had gathered men and armed them, and that he, if he had come on, would not have departed, Caecina had at once departed, you ought not to have doubted that force had been used to Caecina. But if, as soon as he had seen armed men far off, he had withdrawn, you should the less doubt. For all is force which by peril compels us either to depart from anywhere or forbids us to approach. But if you should rule otherwise, see lest you rule this: that he who has departed alive has not had force used to him, lest you set this rule for all the controversies of possession --- that they should think they must contend and decide by arms, lest, just as in war a punishment is set by generals against the cowards, so in trials the case of those should be the worse who have fled than of those who have contended to the last.

\ciceroSection{47} When we are speaking of right and of lawful controversies of men, and in these things name "force," very slight force ought to be understood. I have seen armed men, however few: it is great force. I have departed, frightened off by one man's weapon: I have been cast out and driven away. If you so rule, not only will there be no reason why anyone hereafter should wish to fight to the last for possession, but not even why he should fight back. But if you understand no force without slaughter, without wounding, without blood, you will rule that men ought to be more eager for possession than for life.

\ciceroSection{48} Come now: about force I shall have you yourself, Aebutius, as judge. Answer, if it seems good to you. Did Caecina not wish, or was he not able, to come onto the estate? When you say that you stood in the way and drove him back, surely you grant that he wished to come. Can you then say that force was no hindrance to him to whom, although he longed and had come with that purpose, by gathered men it was not allowed to come? For if what he most wished he could in no way do, surely some force must have stood in the way. Or do you say why, when he wished to come, he did not come.

\ciceroSection{49} Now you cannot deny that force was used. The question is in what manner he is cast out who has not come. For he who is cast out must be moved and driven from a place. But how can this happen to him who has never been at all in that place from which he says he has been cast out? What if he had been there and had fled in fear, with armed men seen --- would you say he had been cast out? I think so. Do you say it? You who so diligently and so cunningly judge controversies by words and not by equity, and express rights not by common usefulness but by letters --- can you say that he has been cast out who shall not have been touched? What? Will you say "thrust out"? For by that word the praetors before were wont to use in this interdict. What say you? Can anyone be thrust out who is not touched? Indeed, if we wish to follow the word, must we not understand this: that he is thrust out who is laid hands upon? It is necessary, I say, if we wish to fit the matter to the word, that no man be ruled to have been thrust out who has not been understood, with force used, to have been moved by the hand and pushed headlong.

\ciceroSection{50} But cast out --- how can anyone be, save moved into a lower place from a higher? He can be driven, put to flight, finally thrown out. But that he can in no way be, who has not even been touched --- nay, even on a level and flat place. What then? Do we suppose this interdict was set up for the sake of those who say they have been cast headlong from higher places (for those we can truly call cast out), or, since the will and counsel and meaning of the interdict is understood, shall we reckon that the highest shamelessness or singular stupidity is to dwell in the error of words --- not only to leave aside the matter and the cause and the common usefulness, but even to betray it?

\ciceroSection{51} Or is this in doubt: that there is not such a great supply of words --- not only in our own tongue (which is said to be poor) but in any other --- that all things should be named by their fixed and proper words; nor truly is there any need of words, when the matter for whose sake the words have been sought is understood? What law, what senatorial decree, what magistrate's edict, what treaty or pact --- what (to come back to private matters) will, what trial or stipulation or formula of pact and agreement, cannot be weakened and torn apart, if we should wish to bend the matter to words and leave aside the counsel of those who have written and the reason and authority?

\ciceroSection{52} By Hercules, familiar and daily speech will not hold together if we shall hunt for words between us. Finally there will be no domestic command, if we grant our slaves this --- that they obey us according to words, and not according to that which can be understood from the words. I see that I must use examples of all these things; but to each one of you another example does not occur in every kind, which may be a witness that the law does not depend on words but words serve the counsels and authorities of men.

\ciceroSection{53} Splendidly and copiously did Lucius Crassus, by far the most eloquent man, a little before we came into the forum, defend this opinion in the Centumviral court, and easily, when against him the most prudent man, Quintus Mucius, was speaking, prove to all that Manius Curius, who had been instituted heir thus: \textit{when the posthumous son shall have died}, when the son not only had not died but had not even been born, ought to be heir. What? Was this provided for enough by the words? By no means. What thing then prevailed? The will. Which, if it could be understood while we are silent, we should not use words at all. Because it cannot, words have been devised --- not which should hinder, but which should show the will.

\ciceroSection{54} The law orders the use and authority of an estate to be two years; but we use the same right in houses, which are not named in the law. If a way be unrepaired, the law orders him to drive his beast wherever he wishes. Can it be understood from the very words that, if a way be unrepaired in Bruttium, it is lawful, if he wishes, to drive a beast through Marcus Scaurus's Tusculan estate? The action against the warrantor present is in these words: \textit{whenever I see you in court}. By this action that famous Appius the Blind could not have used, if so men in court would chase the words that they did not consider the matter for which the words are. If by will the orphan Cornelius were named heir, and he were now twenty years old --- with you as interpreters he would lose the inheritance.

\ciceroSection{55} Many examples come into my mind; into yours more, I am sure. But that we may not take in too many things and that the speech may not stray further from what is set forth, let us consider this very interdict about which the matter is. For you will understand in this very thing, if we set up the law in words, that we shall lose all the usefulness of this interdict, while we wish to be cunning and crafty. \textsc{Whence you or your household or your agent}. If your bailiff alone had cast me out, the household had not cast me out, I think, but someone of the household. Would you rightly say that you have restored? Why, what is easier than to prove to those who only know Latin that in one little slave the name of household is not valid? But if indeed you should not even have a slave save him who had cast me out, you would, of course, cry out: "If I have a household, I confess that you have been cast out by my household." For there is no doubt that, if we are led to judging the matter by the word and not by the matter, we should understand by household one which consists of several slaves; that one man is not a household. The word certainly not only demands

\ciceroSection{56} but compels this. But indeed the reason of the law and the force of the interdict and the will of the praetors and the counsel and authority of prudent men spits out this defence and reckons it as nothing. What then? Do those men not speak Latin? Indeed they speak just as much as is enough for understanding the will, since they have set themselves this purpose: that whether you yourself have cast me out, or any of yours, whether of the slaves or of the friends, they distinguish the slaves not by number but call them by the one name of "household";

\ciceroSection{57} but among the free, whoever it is, let him be called by the name of agent. Not because all are or are called agents who carry on any of our business, but because in this matter, the meaning of the interdict learned, they did not wish all the words to be subtly searched out. For there is not another reason of equity in one slave and in many; not another reason of right in this kind --- whether your agent has cast me out (he who is lawfully called an agent of all the affairs of one who is not in Italy or is away on the commonwealth's business, as it were a kind of master, that is, the substitute of another's right), or your tenant or your neighbour or your client or your freedman or whoever has used that force and casting-out at your request or in your name.

\ciceroSection{58} Wherefore, if to restoring him who has been cast out by force the same reasoning of equity has the same weight, that being understood, surely it is nothing to the matter what the force of the words and of the names is. You shall as much restore if your freedman has cast me out, who is set in charge of no business of yours, as if your agent has cast me out. Not because all are agents who carry on any of our business, but because to ask this in this matter is nothing to the purpose. You shall as much restore if one little slave, as if the whole household has done it. Not because one little slave is the same as a household, but because it is not asked in what words each thing is said, but what matter is being treated. Even (that we may withdraw further from the word, but not the slightest from equity) if you shall have no slave of yours, and all are foreign and hireling, yet they themselves shall be held in the kind and name of your "household."

\ciceroSection{59} Go on further to follow this same interdict. \textsc{Men gathered}. You shall have gathered no man; they shall have come together of their own accord. Surely he gathers who herds men together and calls them. Those have been gathered who have been herded together by some man into one place. If they have been not only not called together but not even come together, but have only been those who --- not, indeed, that force should be done, but for the sake of tilling or pasturing --- had been wont to be on the field, you will say that men were not gathered. By the word indeed you will overcome me with myself as judge; but in the matter you will not even stand with any judge. For the force of a multitude they wished to be restored, not of a multitude only called together. But because most often, where there is need of a multitude, men are wont to be gathered, therefore the interdict has been set up about those gathered. Which, even if it shall seem to differ in word, in matter shall yet be one and shall be valid in all cases in which one and the same case of equity is seen through. \textsc{Or armed}. What shall we say?

\ciceroSection{60} Whom can we truly call armed, if we wish to speak Latin? I think those who are equipped and adorned with shields and weapons. What then? If with clods or stones or cudgels you have driven anyone headlong from his estate, and you have been ordered to restore him whom you have cast out by armed men, will you say that you have restored? If words have weight, if cases are weighed not by reasoning but by sounds, say it on my warrant. You will surely overcome that those were not armed who threw stones which they themselves had taken from the ground; that turfs and clods are not arms; that those were not armed who, passing by, broke off a branch from a tree; that arms by their own names are some for covering, others for hurting; that those who had not these were unarmed.

\ciceroSection{61} But if there shall be any judgement of arms, then say so. When it shall be a judgement of right and of equity, beware lest in such a cold, such a starved cavil you should hide. For you will not find any judge or recoverer who, as if a soldier's arms had to be inspected, should so prove armed; but it shall be just as if they had been most armed, if they shall be found to have been so prepared that they could bring force on life or body.

\ciceroSection{62} And that you may the more understand how words have no weight: if you alone or any one man with shield and sword had made an attack on me and I had been so cast out, would you dare to say that the interdict was about armed men, but here there had been one armed man? You would not, I think, be so shameless. But see lest you be much more shameless now. For then indeed you could implore all mortals: that men in your business should forget Latin, that the unarmed should be judged armed, that, when the interdict was about several, but the matter had been committed by one, one man should be judged to be several men.

\ciceroSection{63} But in these cases not the words come into court, but that matter for whose sake these words have been thrown into the interdict. Force which had to do with life and body they wished to be restored without any exception. That generally is done through gathered and armed men. If by another counsel, with the same peril, it is done, they wished it to be by the same right. For there is not greater wrong if your household than if your bailiff, not if your slaves than if foreign and hireling, not if your agent than if your neighbour or your freedman, not if by gathered men than if by voluntary or even daily and domestic ones, not if by armed than if by unarmed who had the force of armed for hurting, not if by several than if by one armed. For by what things mostly such force is done, those things are named in the interdict. If by other things the same force is done, that, although not enclosed in the words of the interdict, is yet kept by the meaning of the law and its authority.

\ciceroSection{64} I now come to that thing of yours: "I did not cast him out; for I did not allow him to come on." I think you yourself, Piso, see how much narrower and more unjust this defence is than if you should use that other: "They were not armed, when they were with cudgels and stones." If, by Hercules, the choice were given to me --- not a copious man for speaking --- which I should prefer to defend: that he had not been cast out who had been met by force and arms as he entered, or that those had not been armed who were without shields and without iron --- although for proving I should see each thing weak and trifling, yet for speaking in the one I should seem to myself able to find something: that they were not armed who had neither anything of iron nor any shield. But here surely I should stick, if I had to defend that he who had been driven and put to flight had not been cast out.

\ciceroSection{65} And in your whole defence this seemed to me most wonderful: that you said the authority of those learned in the law need not be obeyed. Which I, although I now neither hear for the first time nor in this case alone, yet greatly wondered why you should say it. For others run to that line of speech when they think they have in the case the equitable and good which they should defend. If on the contrary it is contended in words and letters and (as it is wont to be said) by the highest right, they are wont to oppose to such injustice the name and dignity of the equitable and good. Then they laugh at that thing called "or-not." Then they call into hatred the snares of words and the traps of letters; then they cry out that the matter ought to be judged from the equitable and good, not from a cunning and crafty law; that to follow the writing belongs to the false accuser, and to the good judge to defend the will and authority of the writer.

\ciceroSection{66} But in this case, when you yourself are the one who defends himself by word and letter; when these are your parts: "Whence have you been cast out? Or whence you were forbidden to come? You were driven back, not cast out;" when this is your speech: "I confess that I gathered men, I confess that I armed them, I confess that I threatened you with death, I confess that this would be punished by the praetor's interdict, if will and equity had force; but I find in the interdict one word in which I may hide --- I did not cast you out from that place into which I forbade you to come" --- in this defence do you accuse those who are consulted, because they think the reason of equity, not of the word, ought to be considered?

\ciceroSection{67} And in this place you said Scaevola did not hold his case before the Centumviri --- whom I recalled before, when he did the same as you now (although he did it in some case, you in none), and yet that he proved to no one what he was defending, because he seemed to assail equity by words. While I wonder at this --- that you defended this in this matter at the wrong time and against what this case demanded --- I am wont to wonder also at this thing commonly defended in courts and sometimes by ingenious men: that those learned in the law ought not to be conceded, nor that civil law ought always to have force in cases.

\ciceroSection{68} For those who dispute this, if they say that those who are consulted establish something not rightly, ought not to say that those learned in the law need not be obeyed, but stupid men. But if they grant that they answer rightly and say that judgement should be given otherwise, they say that judgement ought to be given ill. For it cannot be that one thing should be judged about right, another thing answered; nor that anyone be reckoned learned in the law who establishes that to be law which ought not to be judged.

\ciceroSection{69} "But sometimes contrary judgement has been given." First, whether rightly or wrongly? If rightly, that was right which was judged. But if otherwise, there is no doubt whether the judges or those learned in the law are to be blamed. Next, if anything has been judged about a varying right, do they not establish rather against the men learned in the law, if it has been pronounced otherwise than pleased Mucius, than out of their authority, if (as Manilius established) so it has been judged? For Crassus himself did not so plead his case before the Centumviri that he should speak against the men learned in the law, but that he should teach this: that what Scaevola was defending was not law; and to that matter he brought not only reasons, but used Quintus Mucius, his father-in-law, and many most experienced men as authorities.

\ciceroSection{70} For he who thinks the civil law ought to be despised, that man tears apart the bonds not only of trials but of usefulness and of common life. But he who reproaches the interpreters of the law: if he says they are inexperienced in the law, he detracts from the men, not from the civil law; but if he thinks the experienced should not be obeyed, he does not hurt the men but shakes the laws and rights. Which surely must come into your minds: that there is nothing in the state to be kept so diligently as the civil law. For with this taken away, there is nothing by which it can be made certain to anyone what is one's own or what is another's; nothing which can be equal among all and one for all.

\ciceroSection{71} So in the rest of controversies and trials, when it is asked whether something has happened or not, whether what is brought forward is true or false, even a feigned witness is wont to be suborned and false records to be set in; sometimes by an honourable and probable name an error is thrown in the way of the good man as judge; to the dishonest the means is given, that, when he has knowingly judged wrongly, he may yet seem to have followed a witness or records. In the law there is nothing of this kind, recoverers --- no false records, no dishonest witness; finally, that excessive power which lords it in the state in this one kind alone is at rest. What it should do, in what manner approach the judge, finally where to stretch out a finger, it has not.

\ciceroSection{72} For this can be said to a judge by some not so modest as well-favoured man: "Judge that this happened or never happened; believe this witness; approve these records." This he cannot say: "Establish that for him to whom a son is born after the will the will is not broken; judge that what a woman has promised without the authority of a guardian is owed." There is no approach to such matters either of any man's power or favour. Finally, that this may seem the greater and more sacred, not even by a price can a judge in such a case be corrupted.

\ciceroSection{73} That witness of yours who dared to say that he was seen to have done it, of which he could not even know what he was being charged with --- he himself would never dare to judge that a husband was owed a dowry which a woman had named without anyone's authority. O distinguished thing, recoverers, and to be kept by you on this account! For what is the civil law? That which can neither be bent by favour nor broken by power nor adulterated by money. Which, if not only oppressed but even abandoned or kept too carelessly, there is nothing whereby anyone may reckon he has anything sure either to receive from his father or to leave to his children.

\ciceroSection{74} For what does it profit to have a house or estate left by one's father or by some method well got, if it is uncertain whether what is now yours by mancipium-right you can keep, if the civil law is too little defended and cannot be held by public law against any man's favour? What, I say, profits it to have an estate, if those rights of boundaries, of possessions, of waters and of paths --- which have been most diligently set out by our ancestors --- can in some way be confused and altered? Believe me: a greater inheritance has come to each of us in those very goods from law and from statutes than from those by whom those very things were left to us. For that an estate should come to me by someone's will can happen; that I should keep what has been made mine cannot happen without civil law. An estate can be left by a father; but \textit{usucapio} of an estate (that is, the end of trouble and peril of suits) is left not by the father but by the laws. The aqueducts, the right of drawing, the way, the path are left by the father; but the fixed authority of all these things is taken from the civil law.

\ciceroSection{75} Wherefore no less diligently those things which you have received from your ancestors --- the public patrimony of the law --- ought you to keep, than your own private property: not only because these are hedged in by the civil law, but also because the patrimony of one will be lost by the inconvenience of one, while the law cannot be lost without great inconvenience to the state. In this very case, recoverers, if we do not get this --- that he was cast out by force, by armed men, who, it is agreed, was driven and put to flight by force, by armed men --- Caecina shall not lose the matter (which he himself with brave mind would lose, if the time so bore); for the present he shall not be restored to possession; nothing more.

\ciceroSection{76} The cause of the Roman people, the right of the state, the goods, the fortunes, the possessions of all are called back into doubt and uncertainty. By your authority this shall be set up, this shall be laid down: with whom you contend hereafter about possession, if you have cast him out only as he was entering the field, you shall have to restore him; but if, as he was entering with an armed multitude, you have met him and so as he came have driven him back, put to flight, turned aside, you shall not restore. If this is the voice of the law --- that there is force not in slaughter alone but also in the mind --- of caprice, that, unless gore appears, force has not been used. If of the law, that he is cast out who is forbidden; of caprice, that no man can be cast out save from the place where he has set his footprint.

\ciceroSection{77} If of the law, that the matter and meaning and equity ought to have most weight; of caprice, that all law is twisted by word and letter --- you, recoverers, must establish which voices seem to you the more honourable and more useful. In this place it has fallen out most conveniently that he is not present who a little before was present and is wont to be often present with us in this case, the most adorned man, Gaius Aquilius. For with him present I should speak the more timidly of his virtue and prudence, because both he himself would be afflicted by some modesty over his own praise, and a like reasoning of modesty would slow me from the praise of one present. Whose authority, it has been said by that side, ought not to be too greatly conceded. I do not fear that I should say more about such a man than you should think or wish to be recalled before you.

\ciceroSection{78} Wherefore I shall say this: that his authority can never have too much weight, whose prudence the Roman people has seen through in providing safeguards, not in deceiving --- who has never sundered the reasoning of the civil law from equity, who for so many years has kept his talent, his labour, his faith ready and laid open before the Roman people; who is so just and good a man that he seems by nature, not by training, to be a counsellor; who is so experienced and prudent that out of the civil law not only some knowledge but even goodness seems to have been born; whose talent is so great, whose faith so approved, that whatever you draw from him, you feel you draw pure and clear.

\ciceroSection{79} Wherefore you make me very greatly grateful when you say he is the author of our defence. But this I wonder: he whom you say feels something against me, why do you call him as author for me, why do you name him "ours"? But yet what does that author of yours say? \textsc{In whatever words each thing has been done and pronounced}. I have met not no one of that kind of counsellors, I think, that very man by whose authority you say you handle this matter and set up the defence of the case. Who, when he had entered into this dispute with me --- that no one could be proved to have been cast out save from the place in which he had been --- confessed that the matter and meaning of the interdict was with me; he said I was shut out by the word; but he did not think that one could depart from the word.

\ciceroSection{80} When I used many examples out of all the memory of antiquity --- that the law and the reasoning of the equitable and good have often been sundered from the word and the writing in very many things, and that always that thing has had most weight which had in itself most authority and equity --- he comforted me and showed that in this very case there was nothing for me to be in trouble about. For the very words of the sponsion were on my side, if I should wish to attend diligently. "How then?" I said. "Because surely," he said, "Caecina was cast out by force, by armed men, from some place; if not from that place into which he wished to come, yet surely from that place whence he fled." "What then?" "The praetor," he said, "interdicted that he should be restored to where he had been cast out from --- that is, whatever place that was from where he had been cast out. But Aebutius, who confesses that Caecina has been cast out from some place, since he has said that he restored him,

\ciceroSection{81} must necessarily have made the sponsion ill." What is it, Piso? Does it please you that we fight by words? Does it please you that the case of law and equity --- and not of our possession, but altogether of all possessions --- be set up in a word? I have shown what seemed to me, what was wont to be done by our ancestors, what was worthy of the authority of those by whom judgement is to be given: that the true, the equitable, the useful for all is to be looked at; by what counsel and by what meaning, not by what words each thing was done. You call me to the word. I shall not come before I have refused. I deny it ought to be; I deny it can be obtained; I deny that there is any matter which can be enough taken in or guarded against or excepted, if, with some word omitted or set down ambiguously, with the matter and meaning known, not what is understood but what is said shall have force.

\ciceroSection{82} Since I have refused enough, I now come whither you call. I ask of you whether I was cast out, not from the Fulcinian estate. For the praetor did not order, "if I had been cast out from that estate," that I should be so restored, but "to where I had been cast out from." I have been cast out from the next neighbour's estate, by which I was approaching that estate; from the road; from somewhere certainly, whether private or public; to that I have been ordered to be restored. You have said you have restored. I deny that I have been restored according to the praetor's decree. What do we say to this? Either by your sword (as it is said), or by ours, your defence must be finished off.

\ciceroSection{83} If you take refuge in the meaning of the interdict and say that what must be inquired is about which estate the matter was when Aebutius was being ordered to restore, and you do not think the equity of the matter ought to be caught by the snare of the word, you are at large in my camp and garrisons. Mine, mine is that defence; this I cry out; this I call upon all men and gods to witness: that, since our ancestors guarded armed force by no defence of law, not the footprint of him who has been cast out, but the deed of him who has cast out, comes into court; that he is cast out who has been put to flight, that force has been used against him on whom the peril of death has been thrown. But if you flee and dread this place, and from this

\ciceroSection{84} (so to speak) field of equity call me back to those narrows of words and to all the corners of letters, you are shut up in those very snares which you try to set against me. "I did not cast out, but I cast back." This seems sharp to you; this is the point of your defence: into that very point your case must necessarily run. For I report this to you: if I have not been cast out from that place to which I was forbidden to come, yet I have been cast out from that place to which I came, whence I fled. If the praetor did not distinguish the place to which he ordered me to be restored, and ordered me to be restored, I have not been restored according to the decree.

\ciceroSection{85} I should like, recoverers, that you should so reckon this whole matter (if it shall seem to you craftier than my custom of defending bears): first, that another, not I, has thought of it; next, that I am not only not the deviser but not even the approver of this reasoning, and have brought it not for my defence, but have referred it to their defence. That I can in my own right say it was not necessary to ask in this matter which I have brought forward by what words the praetor interdicted, but about what place was at issue when he interdicted; nor in the force of armed men ought it to be looked into in what place the force was done, but whether it was done. But you can in no way defend this --- that in the matter where you wish, words ought to be looked at; in the matter where you do not, they ought not.

\ciceroSection{86} But yet, what is answered me to that which I have already said: that not only by the matter and meaning, but also by the words, this interdict has been so framed that nothing should seem must be changed? Mark, please, diligently, recoverers; for it is part of your wit to know not my, but our ancestors', prudence. For I am not about to say what I have devised, but what has not escaped them. When the interdict is about force, they understood there were two kinds of cases to which the interdict had to do: one, if anyone said he had been cast out from the place where he had been; the other, if anyone from the place where he was coming. And of these each, and nothing besides, can happen, recoverers.

\ciceroSection{87} Consider it thus. If anyone has cast my household out of my estate, he has cast me out \textit{from} that place; if anyone has met me with armed men outside my estate and forbidden me to enter, he has cast me out not \textit{from} that place but \textit{by} that place. For these two kinds of matters they found one word which would declare both things enough: that whether \textit{from} an estate or \textit{by} an estate I had been cast out, by one and the same interdict I should be restored: \textsc{whence you}. This word "whence" declares both --- both \textit{from} which place and \textit{by} which place. Whence was Cinna cast out? From the city. Whence Telesinus? By the city. Whence cast out the Gauls? From the Capitol. Whence those who were with Gracchus?

\ciceroSection{88} From the Capitol. You see therefore that by this one word "whence" two things are signified: \textit{from} which and \textit{by} which. But when he orders him to be restored thither, he so orders that, if the Gauls had asked of our ancestors that they be restored to the place whence they had been cast out, and could by some force attain this, they ought, I think, not to be restored to the tunnel by which they had attacked, but to the Capitol. For this is understood: \textsc{whence you have cast out, whether from which place or by which place, restore him there}. This now is simple: restore him to that place. Whether you have cast him out from this place, restore him to this place; whether by this place, restore him to that place, not from which but by which he has been cast out. Just as if anyone, when from the deep he had come up to his fatherland and, suddenly cast back by storm, should wish that, since he had been driven from his fatherland, he should be restored thither --- this, I think, he would wish: that fortune restore him to that place from which he had been driven, not into the open sea, but into the very city he was seeking. So, since we hunt for the force of words by the necessary likeness of things, he who asks that he be restored to that from which he was cast out (that is, whence he was cast out) asks this: that he be restored to that very place.

\ciceroSection{89} When the words lead us thither, the matter itself compels us to feel and to understand this. For, Piso (I now return to those beginnings of my defence): if anyone should cast you out of your house by force, by armed men, what will you do? You will pursue, I think, by this interdict which we have used. What? If anyone with armed men should forbid you, as you are returning home from the forum, to enter your house, what will you do? You will use the same interdict. Since therefore the praetor has interdicted that, where you have been cast out from, to that you should be restored, you yourself shall interpret this same as I say (and as is plain): when that word "whence" has weight in either matter, and to that you have been ordered to be restored, you ought as much to be restored to your house if you have been cast out from the vestibule, as if from the inner part of the house.

\ciceroSection{90} But that there should now be no doubt at all, recoverers, whether you wish to look at the matter or the words, but to judge for us, here, with all things now overwhelmed and lost, that defence arises: that he can be cast out who is then in possession; he who is not in possession in no way can. So, if I have been cast out from your house, it is not necessary that I be restored; if you yourself, it is necessary. Count how many things in that defence are false, Piso. And first mark this: that you have now been driven from that reasoning by which you said no man can be cast out save from where he was. Now you grant he can be. You say that he who is not in possession cannot be cast out.

\ciceroSection{91} Why then either is added in that daily interdict "whence he cast me out by force" the words \textsc{when I was in possession}, if no one can be cast out who is not in possession; or in this interdict on \textsc{armed men} is it not added, if it is necessary to ask whether he was in possession or not? You deny that one is cast out save who is in possession. I show: if anyone has been cast out without gathered or armed men, he who confesses he has cast him out wins the sponsion if he shows that he was not in possession. You deny that one is cast out save who is in possession. I show, out of this interdict on \textsc{armed men}, that he who can show that the man who has been cast out was not in possession must yet be condemned of the sponsion, if he confesses that he has been cast out.

\ciceroSection{92} Men are cast out in two ways: either without gathered or armed men, or by such reasoning and force. For two unlike things two distinct interdicts have been set up. In that daily force it is not enough that one can show himself cast out, unless he can show that, when he was in possession, then he was cast out. Not even that is enough, unless he shows that he so possessed that he possessed neither by force nor secretly nor by sufferance. So he who has said he has restored is wont often with a great voice to confess that he has cast out by force, but he adds this: "He was not in possession"; or even, when he has granted this very thing, he yet wins the sponsion if he makes it plain that the other had possessed from him either by force or secretly or by sufferance.

\ciceroSection{93} Do you see how many defences our ancestors wished him to be able to use who had used force without arms and a multitude? But this man, who from law, duty, good morals has fled to iron, to arms, to slaughter, you see naked and abandoned in the case --- so that he who armed had contended for possession, unarmed plainly contends about the sponsion. Is there any difference, Piso, between these interdicts? Is there any difference, whether in this is added \textsc{when A. Caecina was in possession} or not? Does the reasoning of the law move you, does the unlikeness of the interdicts, does the authority of our ancestors? If it had been added, it ought to be asked about it; it has not been added: shall it nevertheless be necessary? And I do not in this defend Caecina;

\ciceroSection{94} for Caecina was in possession, recoverers. And this, although it is outside the case, I shall yet run through briefly, that you may wish the man himself defended no less than the common right. That Caesennia possessed, on account of her use and fruit, you do not deny. The tenant who held the estate on lease from Caesennia, since he was on the estate from the same lease, is there any doubt that, if Caesennia was then in possession when the tenant was on the estate, after her death the heir possessed by the same right? Then Caecina himself, when he was going round his estates, came to that one and received accounts from the tenant.

\ciceroSection{95} There are testimonies on this matter. Afterwards, Aebutius, why did you give notice to Caecina about that estate rather than another (if you have any), if Caecina was not in possession? Caecina himself, again, why did he wish to be put off by custom, and so had answered you on the opinion of his friends and of Aquilius? But, you say, Sulla passed a law. Let me say nothing about that time, nothing complain about the calamity of the commonwealth: this I answer to you, that the same Sulla added in the same law: \textsc{if anything has been proposed which is not law, by that law nothing is proposed}. What is it that is not law, that the people cannot order or forbid? Not to go further off, that addition declares that there is something. For, were there not, this would not be added in all laws.

\ciceroSection{96} But I ask of you whether you would think, if the people should order that I be your slave, or you mine, that order would stand and be firm. You see this is nothing, and confess. In which matter first you grant that not whatever the people has ordered ought to stand. Next, you bring no reasoning why, if liberty cannot in any way be taken away, citizenship can be. For both in the same way concerning each matter has been handed down to us; and, if citizenship can once be taken away, liberty cannot be kept. For how can he be free in the right of the Quirites who is not in the number of the Quirites?

\ciceroSection{97} And I, while I was a young man pleading this case, proved it against the most eloquent man of our state, Gaius Cotta. When I was defending the freedom of an Aretine woman, and Cotta had thrown a religious scruple before the decemvirs that our oath could not be judged just, because citizenship had been taken from the Aretines; and I had vehemently contended that citizenship could not be taken away, the decemvirs in the first hearing did not give judgement. Afterwards, the matter searched and considered, they judged our oath just. And this was judged with Cotta speaking against and Sulla alive. But now in other matters, that all who are in the same case both plead by law and pursue their right, and use all civil law without the doubt of any magistrate or judge or experienced or inexperienced man --- why should I recall? I am sure that to none of you it is in doubt.

\ciceroSection{98} I do not fail to know that this is wont to be asked --- that you may hear from me what does not come into your mind: in what manner, if citizenship cannot be taken away, our citizens have often set out into Latin colonies. Either of their own will or by the fine of a law they set out. Which fine, if they had wished to suffer, they could have remained in citizenship. What? He whom the pater patratus has handed over, or his own father or the people has sold --- by what right does he lose his citizenship? That the citizenship may be loosed from religious obligation, a Roman citizen is handed over. Who, when he is received, is theirs to whom he has been handed; if they do not receive him (as the Numantines Mancinus), he keeps his case and his right of citizenship whole. If a father has sold him whom he had taken into his power, he sends him out from his power. Now, when the people sells him who has not become a soldier,

\ciceroSection{99} it does not take liberty from him, but judges that he is not free who, that he might be free, was unwilling to enter peril; but when it sells the uncensused, it judges this: that, since those who have been in lawful slavery are freed by the census, he who, when he was free, was unwilling to be censused, has himself adjudged liberty against himself. But if at the highest by these things liberty or citizenship can be taken away, those who recall these things do not understand: if our ancestors wished it could be taken away by these reasonings, did they not wish it in another way?

\ciceroSection{100} For as they bring these things forth out of the civil law, so I would they should bring those by which by law or proposal citizenship or liberty has been snatched away. For as for what concerns exile, of what kind it is can plainly be understood. For exile is not punishment, but a refuge and a haven from punishment. For because they wish to escape some penalty or calamity, they thither turn the soil --- that is, they change their seat and place. So in our law none will be found that, as among other states, any evil deed has been punished by exile. But when men shun chains, deaths, and ignominies which have been established by laws, they flee, as if to an altar, into exile. Who, if they should wish in their own state to undergo the force of the law, would not lose citizenship before life. Because they do not wish, citizenship is not taken from them, but is left and laid down by them. For, since by our law no one can be of two states, this citizenship is then lost when he who has fled is received into exile, that is, into another state.

\ciceroSection{101} It does not escape me, recoverers, although I am leaving aside very many things about this law, that I have yet slipped further than the reasoning of your trial demanded. But I did this not because I thought you longed for this defence in this case, but that all might understand that citizenship has not been taken from anyone nor can be taken. This I wished both those to know to whom Sulla wished to do wrong, and all the rest of the new and old citizens. For no reasoning can be brought why, if from any new citizen citizenship could be taken away, it could not be from all patricians, from all the most ancient citizens.

\ciceroSection{102} For that to this case nothing this had to do can be understood first from this: that you ought not to judge about that matter; next, that Sulla himself so passed about citizenship that he did not take away their bonds and inheritances. For he orders them to be of the same right as the people of Ariminum had been --- whom who is ignorant were of the twelve colonies, and could take inheritances from Roman citizens? But if citizenship could have been taken from Aulus Caecina by law, all good men would yet rather seek that reasoning: in what manner we might keep that most respected and most modest man --- endowed with the highest counsel, the highest virtue, the highest domestic authority --- as a citizen, freed of wrong, than that, as now (when he could lose nothing of the right of citizenship), there should be anyone save those like you, Sextus, in stupidity and shamelessness, who should say that citizenship has been taken from him.

\ciceroSection{103} Who, since, recoverers, he has not abandoned his right nor yielded anything to that man's audacity and petulance, of the rest now lays down the common cause and the right of the Roman people in your faith and scruple. He is the man, he has so wished himself approved by you and by men like you, that he has laboured no less in this case lest he should seem to contend something unjustly than lest he should seem to abandon it loosely; nor feared less to despise Aebutius than to be reckoned despised by him.

\ciceroSection{104} Wherefore, if there is anything outside the trial which is to be granted to the man, you have a man of singular modesty, of known virtue and approved faith, of the most ample name in all Etruria, by many signs of virtue and of humanity known in either fortune. If anything in the opposite party is to be offended at in the man, you have him (to say nothing more) who confesses he has gathered men. But if, with the men set aside, you ask about the case --- since the trial is on force, and he who is charged confesses he has used force with armed men, and tries to defend himself by word, not by equity, and you see that very word taken from him; that the authority of the wisest men makes with us; that it does not come into court whether Aulus Caecina was in possession or not --- yet that he can be shown to have been in possession; much less is it asked whether the estate is Caecina's or not, yet I have shown this very thing, that the estate is Caecina's --- since these things are so, decide what the times of the commonwealth warn you to judge about armed men, what his confession warns about force, what our settlement warns about equity, what the reasoning of the interdict warns about right.

