% Pro M. Fonteio (pro-fonteio) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.
%
% A defence in 69 BC of M. Fonteius, former governor of Transalpine
% Gaul (74-72 BC), prosecuted for extortion by his Gallic subjects.
% The manuscript is mutilated; the speech survives only in fragments.
% The opening sections are damaged.

\ciceroSection{1} \textit{[The opening is lost.]} ...or to have paid, or so to have settled as all others have settled? For thus I defend Marcus Fonteius, gentlemen, and so contend that, after the Valerian law was passed, when you were quaestor, up to the quaestorship of Titus Crispinus, no one paid otherwise; that he followed the authority of all his predecessors, and that all who came after followed his.

\ciceroSection{2} What do you accuse, what do you reproach? For when, in the dodrans and quadrans tablets which he says were instituted by Hirtuleius, he longs for an office of Fonteius, I cannot reckon whether he himself errs or wishes to lead you into error. For I ask of you, Marcus Plaetorius: can our cause be approved by you yourself, if, in the matter in which Marcus Fonteius is accused by you, he has as authority that very Hirtuleius whom you most praise; while in that matter in which you praise Hirtuleius, Fonteius is found to have done the same? You reproach the kind of payment; the public records show Hirtuleius paid in the same way. You praise him because he set up the dodrans tablets; the same Fonteius set up, and of the same kind of money. For lest perhaps you should be ignorant and reckon that those tablets had to do with a different reckoning of old debt --- they were set up for one cause and in one kind. For with the publicans who held Africa, who held the Aquileian harbour duty\ldots \textit{[The text breaks off.]}

\ciceroSection{3} \textit{[Resuming.]} ...no one, no one, I say, gentlemen, will be found who shall say that he gave one penny to Marcus Fonteius in his quaestorship, or that the latter took from that money which was paid for the treasury. In the records of no man will any sign of this theft be found, no trace of waste or diminution among those entries. And yet men, if any have been accused on this kind of inquiry, we see have been reproached first by witnesses; for it is hard that he who has given money to a magistrate should not either be led by hatred to speak or compelled by religious obligation. Then if by some favour the witnesses are deterred, the records at least remain uncorrupted and whole. Suppose all were most friendly to Fonteius, suppose so great a number of men --- most unknown and most foreign --- spared his life, looked out for his standing. The matter itself, however, and the reckoning of letters, and the drawing-up of the records have this force: that out of receipts and outlays, whatever is feigned, or stolen, or does not stand, appears. They have all entered moneys received by the Roman people; if they at once paid out or gave to others sums equally great, so that what is received by the Roman people is the outlay to someone, surely nothing can have been taken away. But if they have brought anything home, out of their chests, out of\ldots \textit{[Text broken.]}

\ciceroSection{4} ...the faith of gods and men! No witness is found in 23 million sesterces! Of how many men? Of more than six hundred. In what lands was the business done? In that, in that place, I say, which you see. Was money given out of order? Indeed not a single penny was moved without many letters. What then is this prosecution which can more easily climb the Alps than a few steps of the treasury, which more diligently defends the treasury of the Ruteni than of the Roman people, which more willingly uses unknown than known witnesses, foreign than domestic; which thinks the charge can be more plainly confirmed by the lust of barbarians than by the letters of our men?

\ciceroSection{5} The reckoning of two magistracies, in each of which he was occupied with handling and managing the greatest sums of money, the triumvirate and the quaestorship, is so rendered, gentlemen, that in those things which were done before our eyes, which had to do with many men, which were drawn up in public and private records, no sign of theft, no suspicion of any offence is found.

\ciceroSection{6} The Spanish legateship followed at a most troubled time of the commonwealth, when, with the coming of Lucius Sulla and the greatest armies into Italy, the citizens differed by force, not by trials and laws; and in this state of the commonwealth despaired of, what kind of\ldots \textit{[Text broken.]}

\ciceroSection{7} If no money has been counted out, of what money is the fiftieth?

\ciceroSection{8} The greatest quantity of grain from Gaul; the most ample troops of foot from Gaul; horsemen most in number from Gaul.

\ciceroSection{9} That afterwards the Gauls were going to drink it more diluted --- which they were going to reckon as poison.

\ciceroSection{10} That Plaetorius's mother, while she lived, kept a school; after she died, she had schoolmasters.

\ciceroSection{11} That under this praetor Gaul was overwhelmed by debt. From whom do they say loans of such great sums were made? From the Gauls? Nothing less. From whom then? From the Roman citizens who do business in Gaul. Why do we not hear their words? Why are no records of theirs brought forth? I press the prosecutor still and stand by him, gentlemen; I press him, I say, and demand the witnesses. I spend more labour and toil in this case in demanding witnesses than other defenders in refuting them. Boldly I say this, gentlemen, I do not rashly affirm it: Gaul is stuffed with businessmen, full of Roman citizens. No Gaul does any business without a Roman citizen. No coin in Gaul is moved without the records of Roman citizens.

\ciceroSection{12} See to what I come down, gentlemen, how far I seem to depart from my custom and from caution and diligence. Let one set of records be brought forth in which there is some trace which signifies that money was given to Marcus Fonteius. Let them produce one out of so great a number of businessmen, settlers, publicans, farmers, cattle-men, as witness; I shall grant the prosecution to be true. By the immortal gods! What is this case, what defence? Marcus Fonteius governed the province of Gaul, which is made up of those kinds of men and states which (to leave aside the old) partly within our own memory waged bitter and long-lasting wars with the Roman people; partly were lately subdued by our generals, lately overcome by war, lately marked by triumphs and monuments, lately fined of fields and cities by the senate; partly with Marcus Fonteius himself contended with iron and hand, and by his much sweat and toil fell under the imperium and rule of the Roman people.

\ciceroSection{13} In the same province is Narbo Martius, a colony of our citizens, the watchtower and bulwark of the Roman people set against and opposed to those very nations. There is also the city of Massilia, of which I have spoken before --- of bravest and most faithful allies, who have made up the perils of the Gallic wars by special rewards from the Roman people. There are besides the largest number of Roman citizens and knights, most honourable men. Over this province, which from this variety of kinds was made up, Marcus Fonteius (as I said) governed. Those who were enemies he subdued; those who had lately been so, he compelled to depart from those fields of which they had been fined. On the rest, who had often been overcome in great wars on this account that they should always obey the Roman people, he commanded great cavalries for those wars which were then being waged in the whole world by the Roman people; great sums of money for their pay; the greatest quantity of grain for sustaining the Spanish war.

\ciceroSection{14} He who carried this through is called into court; you who were not present at the matter learn the case together with the Roman people. Those speak against him to whom most unwillingly it was commanded; those speak who were compelled to depart from their fields by Gnaeus Pompeius's decree; those speak who out of the slaughter and flight of war now first dare to stand up against an unarmed Marcus Fonteius. What? The Narbonensian colonists --- what do they wish, what do they think? They wish him kept safe through you; they think themselves kept safe through him. What of the state of the Massilians? Him present they have furnished with those most ample honours which they had; you absent they ask and beseech that their religious obligation, their praise, their authority may seem to have had some weight before your minds.

\ciceroSection{15} What? What is the will of the Roman citizens? There is no one out of so great a number who does not reckon him to have deserved most excellently of the province, of the imperium, of the allies and of the citizens. Since therefore, gentlemen, you have learned who attacked Marcus Fonteius, who wish him defended, decide now what your equity, what the dignity of the Roman people demands: whether you would prefer to trust and to consult for your colonists, your businessmen, your most friendly and most ancient allies, than for those to whom you ought neither (on account of their anger) to give faith, nor (on account of their faithlessness) to give honour.

\ciceroSection{16} What? If I should bring even a greater abundance of most honourable men which can be a witness to his virtue and innocence, will the agreement of the Gauls yet have more weight than the highest authority of all? When Fonteius was governing Gaul, you know, gentlemen, that the largest armies of the Roman people were in the two Spains, and the most distinguished generals. How many Roman knights, how many tribunes of the soldiers, of what kind and how many and how often legates went out to them! Besides, the largest and most adorned army of Gnaeus Pompeius wintered in Gaul under Marcus Fonteius's command. Does fortune itself seem to you to have wished there to be witnesses and partners enough, fit enough, of those things which were done in Gaul under Marcus Fonteius as praetor? Whom can you produce out of so great a number of men in this case as witness? Who is there of that number whom you would approve as authority? That man we shall now use as praiser and witness.

\ciceroSection{17} Will you yet doubt longer, gentlemen, that what I set forth to you at the start is most true: that nothing else is being attempted by this trial save that, with Marcus Fonteius pressed down by the testimonies of those to whom many things have been commanded for the commonwealth's sake against their will, the rest may hereafter be the slower to command, when you see those attack him by whose victory the imperium of the Roman people cannot be safe? It has been thrown at him also that Marcus Fonteius made gain out of the repairing of roads, that either he should not compel them to repair, or he should not approve what was repaired. If both all were compelled to repair and the work of many was disapproved, surely each is false: both that price was given for exemption (when no one was free) and for approval (when many things were disapproved). What?

\ciceroSection{18} If we can lay this charge on the best names, and not so as to transfer the fault to others, but to teach that those were in charge of that repairing who can easily both render their duty and have it approved --- yet will you, trusting in angry witnesses, lay all things on Marcus Fonteius? When Marcus Fonteius was hindered by greater businesses of the commonwealth, and when it had to do with the commonwealth that the Domitian way be repaired, he gave the business to his legates, leading men, Gaius Annius Bellienus and Gaius Fonteius. So they were in charge; they commanded, in their own dignity, what seemed good and approved it. Which you, if from no other thing, at least from our letters which you have copied, sent and received, surely can know. Which if you have not read before, learn now from us what Fonteius wrote to his legates about these matters, and what they wrote back to him. \textsc{Letter sent to C. Annius the legate, to C. Fonteius the legate; letter received from C. Annius the legate, from C. Fonteius the legate}.

\ciceroSection{19} It is plain enough, I think, gentlemen, that this reasoning of repairing has nothing to do with Marcus Fonteius and was handled by those whom no one can reproach. Learn now of the wine-charge, which they wished to be the most odious and the greatest. The charge by Plaetorius, gentlemen, has been so set up: that to Marcus Fonteius it did not first come into mind in Gaul to set up a harbour duty on wine, but with this reasoning entered upon and set forth he set out from Rome. So that Titurius at Tolosa exacted four denarii on each amphora of wine under the name of harbour duty; at Crodunum Porcius and Munius three denarii and a victoriatus; at Vulchalo Servaeus two and a victoriatus; and in these places by them harbour duty was exacted if any went off the road to Cobiomagus (which is a village between Tolosa and Narbo) and were unwilling to go to Tolosa; that at Elesiodulum Gaius Annius exacted six denarii from those who carried wine to the enemy.

\ciceroSection{20} I see, gentlemen, that the charge is great both in its very kind --- for a tribute is said to have been laid on our crops, and I confess that a very great sum could have been gathered by that reasoning --- and in its odium very great; for above all his enemies have wished to spread this matter abroad in conversations. But I so think that the greater the charge be that shall be shown to be false, by so much greater is the wrong done by him who feigns it. For he wishes by the size of the matter so to take hold of the minds of those who hear that there shall be a difficult approach for truth. \textsc{On the wine-charge. On the war of the Vocontii. On the disposition of the winter quarters}.

\ciceroSection{21} "But the Gauls deny this." But the reasoning of things and the force of the proofs convicts. Can the judge then not believe witnesses? The greedy and angry and conspirators and removed from religious obligation he not only can but ought not to. For if, because the Gauls speak, therefore Marcus Fonteius must be reckoned guilty, what need have I of a wise judge, of a just inquisitor, of a not stupid speaker? For the Gauls speak; we cannot deny it. If you reckon that these are the parts of an ingenious and experienced and just judge: that, since witnesses speak, without any doubt one must believe --- safety itself cannot guard the innocence of brave men. But if in matters to be judged the wisdom of the judge holds no slight part, in weighing each thing and pondering it by its own weight --- see whether your parts in considering be not far greater and graver than mine in speaking.

\ciceroSection{22} For with me about each matter the witness must not only once be questioned, but even briefly; often even not questioned, lest either to one angry the means of speaking, or to one greedy the authority, should seem to be given. You can move the same matter in your minds the more often, and consider longer about each witness; and, if any we have not wished to question, you ought to reckon what cause was for our being silent. Wherefore, if you think this is laid down for the judge by law or by duty --- to believe witnesses --- there is nothing why one judge should be reckoned better or wiser than another. For the judgement of the ears is one and simple, and given by nature in common and equally to the foolish and the wise.

\ciceroSection{23} Where then can prudence shine forth, where can a foolish and credulous hearer be told from a religious and wise judge? Surely in this: that the things which are said by the witnesses are handed over to conjecture and to thought --- with what authority, with what calmness of mind, with what modesty, with what faith, with what religious obligation, with what zeal of good standing, with what care, with what fear they are said. Will you indeed doubt this in the testimonies of barbarian men, which often within our own and our fathers' memory the wisest judges thought ought not to be doubted in the case of the most distinguished men of our state? Who did not believe Gnaeus and Quintus Caepio, Lucius and Quintus Metellus as witnesses against Quintus Pompeius, a new man --- whose virtue, birth, and deeds the suspicion of greed and enmities took the faith and authority from in their testimony. Whom have we seen,

\ciceroSection{24} whom can we truly recall as equal in counsel, gravity, constancy, in the rest of virtues, in the ornaments of honour, talent, and deeds, to Marcus Aemilius Scaurus? Yet --- whose unsworn nod almost ruled the world --- by his sworn testimony belief was given neither against Gaius Fimbria nor against Gaius Memmius. Those who judged would not have this way open to enmities --- that whom each one hated, him by his testimony he could destroy. How great was the modesty in Lucius Crassus, what talent, what authority, who is ignorant? Yet he, whose very speech had the authority of testimony, could not by his very testimony --- those things which with hostile mind he said against Marcus Marcellus --- prove.

\ciceroSection{25} It was, it was, gentlemen, with those judges a divine and singular counsel, who reckoned that they were judging not only about the defendant but also about the prosecutor, about the witness: what was feigned, what was brought by fortune and the time, what was corrupted by price, what was perverted by hope or fear, what seemed to come from any greed or enmities. Which if the judge shall not embrace all by counsel, shall not look round in mind and thought; if, as each thing shall be said from that place, he shall reckon it said as from some oracle, surely it shall be enough --- as I said before --- that a not deaf judge be set in charge of this duty and office. There shall be nothing why some wise man, experienced in many things, should be sought for judging matters.

\ciceroSection{26} But did those Roman knights whom we have seen, who lately in the commonwealth and the courts most flourished, have such great spirit, such great strength that they did not believe Lucius Crassus, Marcus Scaurus as witness; and you fear not to believe the testimonies of the Volcae and the Allobroges? If credit ought not to have been given to a hostile witness --- was Crassus more hostile to Marcellus or Scaurus to Fimbria, out of civil pursuits and domestic envy, than the Gauls to him? Of whom those who are in the best case --- the knights --- were compelled most unwilling once and again and the more often to give grain and money; the rest, partly out of the old wars fined of land, partly out of this very war overcome and pressed down.

\ciceroSection{27} If those who, on account of some advantage of their own, seem to speak somewhat too eagerly, ought not to be believed, I think a greater advantage was set forth to the Caepiones and the Metelli out of the condemnation of Quintus Pompeius (when they had taken away the disparager of their pursuits) than to all Gaul out of the calamity of Marcus Fonteius, in which that province reckons its near-immunity and liberty has been placed. Or, if the men themselves should be looked at (which surely in a witness ought to have most weight), can any man, however ample, of Gaul be compared not only with the highest men of our state but with the lowest Roman citizen? Does Indutiomarus know what it is to give testimony? Is he moved by that fear by which each one of us, when we have been brought to that place?

\ciceroSection{28} Recall, gentlemen, with how great labour you are wont to take pains, not only what to say in testimony but also what words you use, lest any word seem set down too immoderately, lest any seem to have slipped out from any greed; finally with your face you labour, that no suspicion of greed could be signified, so that, when you come forward, there is some silent reckoning of you of modesty and religious obligation; and, when you depart, that may seem to have been diligently kept and held.

\ciceroSection{29} I trust Indutiomarus feared or considered these same things in his testimony --- who first that most considered word of our custom, "I think" (which we use even when under oath we say things we have ascertained, which we ourselves have seen), took out of his whole testimony, and said he "knew" all things. For he feared, of course, that he should lose any part of his standing with you and the Roman people --- lest any rumour of this kind should follow, that Indutiomarus, such a man, had spoken so eagerly, so rashly. He did not understand that he ought to render to his fellow citizens and to our prosecutors in his testimony nothing save voice and face and audacity.

\ciceroSection{30} Or do you think those nations are moved by the religious obligation of the oath and the fear of the immortal gods in giving testimony? Which differ so much from the custom and nature of the rest of the peoples that, while the rest take up wars for their religious sanctities, those take them up against the religious sanctities of all. Those, in waging war, ask peace and pardon from the immortal gods; these have waged war with the immortal gods themselves. These are the nations which once set out so far from their own seats, even to Delphi, to harass and despoil Pythian Apollo and the oracle of the world. By these same peoples (sacred and religious in testimony!) the Capitoline was besieged, and that Jupiter by whose name our ancestors wished the faith of testimonies to be bound.

\ciceroSection{31} Finally, can anything seem holy and religious to those who, even when at any time, drawn on by some fear, they think the gods must be appeased, defile their altars and temples with human sacrifices, so that they cannot worship religion itself unless they have first violated it by crime? For who is ignorant that they keep up to this day that monstrous and barbarous custom of sacrificing men? Wherefore of what faith, of what piety do you reckon those to be who think the immortal gods themselves can be most easily appeased by the crime and blood of men? Will you join your religious obligation with these as witnesses; from these will you reckon that anything has been said sacredly or moderately? This will your minds, so chaste,

\ciceroSection{32} so whole, take upon themselves: that, while all our legates who in those three years came into Gaul, all the Roman knights who were in that province, all the businessmen of that province, finally all in Gaul who are allies and friends of the Roman people, wish Marcus Fonteius safe, and praise him under oath privately and publicly --- that you should yet prefer to slay him with the Gauls? Whom would you seem to follow? The will of men? Will then the will of enemies have more weight with you than that of citizens? Or the dignity of the witnesses? Can you then prefer the unknown to the known, the unjust to the just, the foreign to the domestic, the greedy to the moderate, the hireling to the gratuitous, the impious to the religious, the most hostile to this empire and name to the good and faithful both allies and citizens?

\ciceroSection{33} Or do you doubt, gentlemen, that those nations all have and bear inborn enmities with the name of the Roman people? Do you so think that those men are at large here in their cloaks and trousers, with downcast and humble mind, as those are wont who, afflicted with wrongs, flee for help to the judges as suppliants and inferiors? Nothing less indeed. They on the contrary wander joyful and erect throughout all the forum, with certain threats and barbarous and monstrous terror of words. Which I would surely not believe, had I not several times heard from the prosecutors themselves at the same time as you, gentlemen, when they were warning that you should beware, lest with this man acquitted some new Gallic war should be stirred up.

\ciceroSection{34} If, gentlemen, all things failed Marcus Fonteius in his case --- if with foul youth, infamous life, ill-discharged offices which he held before your eyes, with the testimonies of good men against him, with legateships disgracefully spent, hated by all his own, he were called into court; if in that trial he were pressed by the testimonies and records of the colonists of the Roman people of Narbo, of the most faithful allies of Massilia, of all the Roman citizens --- yet you would have to take the greatest pains, lest those whom you had received from your fathers and forefathers so afflicted that they were to be despised, you should now seem to have feared and to have been moved by their threats and terror.

\ciceroSection{35} But now, when no good man hurts him, when all your citizens and allies praise him, when the same men attack who have most often attacked this city and this empire; when Marcus Fonteius's enemies threaten you and the Roman people, while his friends and kinsmen beg of you --- will you doubt to show not only to your own citizens (who are most led by glory and praise) but even to foreign nations and peoples that, in casting your votes, you preferred to spare a citizen than to yield to an enemy?

\ciceroSection{36} Great is this cause of acquittal, by Hercules, gentlemen, with the rest of causes: lest some notable stain and ignominy should be taken on by this empire, if it shall be carried into Gaul thus: that the senators and Roman knights, moved not by the testimonies of the Gauls but by their threats, judged the matter at their lust. Indeed, if they shall try to make war, then we shall have to summon up from the dead Gaius Marius, who could be a match in waging war for that threatening and arrogant Indutiomarus; we shall have to summon up Gnaeus Domitius and Quintus Maximus, who should a second time finish off and press down the nation of the Allobroges and the remains of the war by their own arms. Or, since this cannot be, we shall have to ask my friend Marcus Plaetorius to deter his new clients from making war, to plead off their angry minds and dreadful onsets; or, if he cannot, we shall ask Marcus Fabius, his subscriber, to soften the spirits of the Allobroges, since with them the name of the Fabii is most ample. Do those men wish either to be quiet (which the conquered and subdued are wont) or, when they threaten, to make us understand that they are setting before the Roman people not the fear of war but the hope of triumph?

\ciceroSection{37} But if even with a foul defendant you ought not to suffer that those men should think they had got anything by their threats, what do you think you ought to do in the case of Marcus Fonteius? About which man, gentlemen --- for I now seem to myself, with the case almost closed by two hearings, that I ought to say this --- about which man you have heard from his enemies even no feigned, not only no charge but no curse of reproach. Has any defendant ever, especially one engaged in this kind of life --- in seeking honours, in offices of power, in commanding --- been so accused that no reproach, no deed, no foulness which had been born from lust or petulance or audacity, was thrown at him by his prosecutor, if not true, yet feigned with some reasoning and suspicion?

\ciceroSection{38} Marcus Aemilius Scaurus, the highest man of our state, we know was accused by Marcus Brutus. The speeches stand, from which it can be understood that many things were said even against that very Scaurus, falsely; who denies it? Yet they were said and thrown by an enemy. How many things did Manius Aquilius hear in his trial; how many Lucius Cotta; finally Publius Rutilius! Who, although he was condemned, yet seems to me must be reckoned among the best and most innocent men. He, then, that most pure and most temperate man, heard many things in his case which had to do with the suspicion of debaucheries and lusts.

\ciceroSection{39} The speech is extant of the man (in my opinion) by far the most ingenious and most eloquent of our men, Gaius Gracchus; in which speech very many foul and disgraceful things are said against Lucius Piso. But against what man! Who was of such great virtue and integrity that even in those best times, when you could find no man worthless, alone he was named Frugi. When Gracchus ordered him to be summoned to a public meeting and the runner was asking which Piso (because there were several): "You compel me," he said, "to call my enemy Frugi." That man, then, whom not even an enemy could enough signify in addressing without first praising him, who by one surname was declared not only who he was but even what kind he was --- yet he was being called into a false and unjust charge of reproaches.

\ciceroSection{40} Marcus Fonteius has been so accused in two hearings that nothing has been thrown at him by which any trace of lust, of petulance, of cruelty, of audacity could be signified. Not only have they brought forward no deed of his, but they have not even reproached any saying. But if the same as they have either of will to crush him or of licence to evil-speaking they had either of mind for lying or of talent for feigning, Marcus Fonteius would have been of no better fortune for not hearing reproaches than those of whom I spoke before. Therefore, gentlemen, you see a frugal man --- frugal, I say, and in all the parts of his life moderate and temperate, full of modesty, full of duty, full of religious obligation --- placed in your faith and power, and so placed that he has been entrusted to your faith, given over to your power.

\ciceroSection{41} See therefore whether it is fairer that a most honourable man, a bravest man, a best citizen be handed over to most hostile and most monstrous nations or be given back to friends. Especially when there are so many things which beg in your minds for the safety of this innocent man: first the antiquity of his line, which we see set out from Tusculum, the most distinguished town, carved and noted in the monuments of his deeds; then the continuous praetorships, which flourished both in the rest of ornaments and especially in the standing of innocence; then the recent memory of his father, by whose blood not only the band of the Asculans (by whom he was killed) but the whole Italic war was stained with the spot of crime; finally himself --- both honourable and whole in all the parts of his life, in military matters of the highest counsel and the greatest spirit, and indeed by experience of waging wars among the chief of those men who now are most practised.

\ciceroSection{42} Wherefore, even if you are to be warned by me, gentlemen --- which you are not --- I seem to be able lightly, by my authority, to lay down this for you: that out of that kind of men whose virtue, industry, fortune in war is known, you should reckon that they must be diligently kept by you. For there was greater abundance of such men then in this commonwealth; which, although it was, yet was provided for not only their safety but their honour. What now must you do, with military pursuits worn out among the youth, with the bravest men and highest leaders consumed partly by age, partly by the discords of the state and the calamity of the commonwealth, when so many wars are either of necessity taken up by us or suddenly and unforeseen are born? Do you not think that the man himself must be reserved for the doubtful times of the commonwealth, and the rest must be set on fire by the zeal of praise and virtue?

\ciceroSection{43} Recall what legates lately in the Italic war Lucius Iulius, what Publius Rutilius, what Lucius Cato, what Gnaeus Pompeius had. You will know there was then Marcus Cornutus, Lucius Cinna, Lucius Sulla, men of praetorian rank, most experienced in waging war; besides Gaius Marius, Publius Didius, Quintus Catulus, Publius Crassus --- men learned in the knowledge of the military art not by letters but by deeds and victories. Come now, cast your eyes into the senate-house; look deeply into all the parts of the commonwealth. Do you see that nothing can happen so that such men shall be longed for, or, if it should happen, that the Roman people abounds in such men? If you shall diligently attend to these things, surely, gentlemen, you will rather wish to keep at home for yourselves and your children a man busy at the labours of war, brave at perils, experienced in use and discipline, prudent in counsels, lucky in chance and fortune, than to hand him and grant him as a gift to peoples most hostile to the Roman people and most cruel.

\ciceroSection{44} But the Gauls are brought up almost with hostile standards against Marcus Fonteius, and they press and urge with the highest zeal, with the highest audacity. I see, gentlemen. But with many and firm protections, with you as helpers, we shall stand against this monstrous and unbearable barbarism. First Macedonia is set against the onset of those men --- a faithful and friendly province to the Roman people. Which, since it says that itself and its cities have been kept safe by the counsel and the hand of Marcus Fonteius, as it itself was defended through him from the coming and depopulation of the Thracians, so now drives off the onsets and terrors of the Gauls from his head.

\ciceroSection{45} On the other side stands further Spain, which surely can not only by its religious obligation withstand those men's greed but also refute the perjuries of the wicked by its testimonies and praises. And from Gaul itself the most faithful and most grave help is taken. The whole state of the Massilians comes to the help of this wretched and innocent man, which labours not only on this account that it should seem to render an equal favour to him by whom itself was kept safe, but also because it thinks itself placed by that condition and that fate in those lands, that those nations should not be able to harm our men in any way.

\ciceroSection{46} Equally fights for Marcus Fonteius's safety the Narbonensian colony, which through him itself was lately freed from the siege of the enemy, and is now moved by the misery and perils of the same man. Finally, as is fitting in a Gallic war, as the rights and customs of our ancestors lay down, there is no Roman citizen who thinks he must use any excuse. All the publicans, farmers, cattle-men, and the rest of the businessmen of that province defend Marcus Fonteius with one mind and with one voice. But if Indutiomarus shall despise such great resources of our helps --- the leader of the Allobroges and the rest of the Gauls --- shall he then tear away and drag this man even from the embrace of his mother (a most choice and most wretched woman) before your eyes? Especially when on the other side a Vestal virgin holds her brother in her embrace and implores your faith, gentlemen, and that of the Roman people --- she who for you and your children for so many years has been occupied in appeasing the immortal gods, that she now, for her own and her brother's safety, may be able to appease your spirits.

\ciceroSection{47} What protection, what comfort is left for that wretched woman if he is lost? For other women can themselves bear protections for themselves and have at home a partner and sharer of all fortunes. But to this maiden what is there save a brother that can be either pleasant or dear? Do not allow, gentlemen, the altars of the immortal gods and of mother Vesta to be reminded by the daily lamentations of the maiden of your judgement. Look forward lest that eternal fire, kept by the nightly labours and watchings of Fonteia, should be said to have been put out by the tears of your priestess.

\ciceroSection{48} The Vestal virgin stretches out to you suppliant hands, the same which she is wont to stretch to the immortal gods on your behalf. Beware lest it be perilous and arrogant to reject the entreaty of her, whose prayers if the gods should despise, these things could not be safe. Do you see, gentlemen, suddenly --- the bravest man, Marcus Fonteius, by the recollection of his parent and sister, has poured forth tears? He who never was terrified in battle, who armed often threw himself into the hand and multitude of the enemy, since in such perils he reckoned that he was leaving to his own the same comforts which his father had left him --- this same man now with troubled spirit fears that not only he should not be an ornament and help to his own, but should leave to the wretches with the bitterest mourning eternal disgrace and ignominy.

\ciceroSection{49} O fortune very unequal, Marcus Fonteius, if you could have chosen that you should rather perish by the weapons of the Gauls than by their perjuries! For then virtue had been the partner of life, glory the companion of death. But now what grief is it that you should bear the punishments of victory and command at the discretion of those who have either been overcome by arms or have most unwilling obeyed! From this peril, gentlemen, defend a brave and innocent citizen. See to it that you may seem to have believed our witnesses more than the foreigners, to have consulted more for the safety of citizens than for the lust of enemies, to have reckoned weightier the entreaty of her who presides over your sacred rites than the audacity of those who have waged wars with all the sacred rites and shrines. Finally look forward, gentlemen --- which most has to do with the dignity of the Roman people --- that the prayers of a Vestal virgin should seem to have weighed more with you than the threats of the Gauls.

