% In C. Verrem Actio Secunda I (in-verrem-2-1) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.
%
% This is the first of the five published books of the Actio Secunda,
% the written prosecution Cicero published after Verres withdrew into
% exile rather than face the second hearing.

\ciceroSection{1} I think none of you, gentlemen, is unaware that during these days the talk of the multitude and the opinion of the Roman people has been that Gaius Verres would not answer in the second hearing and would not appear at the trial. Which rumour had spread not only because he had certainly resolved and made up his mind not to appear, but also because no one supposed any man so audacious, so senseless, so shameless, that, having been convicted of such unspeakable charges by so many witnesses, he would dare to look the jurors in the face or show his own face to the Roman people.

\ciceroSection{2} It is the same Verres he always was --- as forward to dare, so prepared to listen. He is here, he answers, he is defended; he does not even leave himself this much, that, when he is being held caught manifestly in the most disgraceful matters, if he were silent and absent, he might at least seem to have sought a modest end to his shamelessness. I bear it, gentlemen, and I do not take it ill that I shall reap the fruit of my labour, and you the fruit of your virtue. For if he had done what he had previously resolved --- not to be present --- it would have been somewhat less known than I needed it to be how I had laboured in preparing and constructing this prosecution; and your praise, gentlemen, would have been thin indeed and obscure.

\ciceroSection{3} Nor does the Roman people expect this from you, nor can it be content with this --- that the man who refused to be present should be condemned, and that you should be brave only against one whom no one dared to defend. On the contrary: let him be present, let him answer; let him be defended with the highest resources, with the highest zeal of the most powerful men. Let my diligence contend with the greed of all of them; your integrity with his money; the constancy of the witnesses with the threats and power of his patrons. Then at last all those things will appear to be conquered, when they have come into contention and rivalry. If he had been condemned in absence, he would seem to have looked not so much to his own advantage as to grudging your praise.

\ciceroSection{4} For no greater safety for the commonwealth can be found at this moment than that the Roman people understand that --- the prosecutor having diligently rejected the jurors --- the allies, the laws, the commonwealth can best be defended by senatorial council. Nor can any greater ruin come to all our fortunes than that, by the opinion of the Roman people, the standard of truth, integrity, faith, religion be judged absent from this order.

\ciceroSection{5} So it seems to me, gentlemen, that I have undertaken a great and most ailing and almost given-up part of the commonwealth, and that I have served not so much my own praise and reputation in this as yours. For I have come to the lifting of the resentment against the courts and the removal of their reproach: that, when this case has been judged according to the will of the Roman people, in some part the authority of the courts may seem to be re-established by my diligence; while if it should be wrongly judged, an end may at last be set to the controversy over the courts.

\ciceroSection{6} For without doubt, gentlemen, in this case that very thing is brought into crisis. The defendant is most guilty. If he is condemned, men will cease to say that money has greatest power in these courts. If he is acquitted, we shall cease to refuse the transfer of the courts. Although about his acquittal he himself no longer hopes, nor does the Roman people fear, there are some who wonder at his singular shamelessness --- that he is here, that he answers. To me, considering the rest of his audacity and madness, this does not seem wonderful. For he has committed many things impiously and unspeakably against gods and men, by the punishments of which crimes he is harried and led from his right mind and judgement.

\ciceroSection{7} The penalties of those Roman citizens drive him headlong --- whom he in part struck with the axe, in part killed in chains, in part hung upon the cross while they invoked the laws of liberty and citizenship. The household gods drag him to punishment, since he was found who would tear sons out of their parents' embrace and lead them to death, and who would demand from the parents the price of the burial of their children. The religions and ceremonies of all the rites and shrines violated, the images of the gods --- which not only have been carried off from their temples but lie thrust away in darkness and hidden by him --- do not allow his mind to stand still without madness and frenzy.

\ciceroSection{8} Nor does this man seem to me to be offering himself only to condemnation; nor is he content with this common penalty for greed --- a man who has bound himself by so many crimes. His monstrous and intolerable nature requires some singular punishment. It is not only sought, that with his condemnation his goods be restored to those from whom they were torn, but the religion of the immortal gods must be expiated, and the torment of Roman citizens and the blood of many innocents must be paid for by his punishment.

\ciceroSection{9} For we have brought into your court not a thief but a robber, not an adulterer but a stormer of chastity, not a sacrilegious man but an enemy of all sacred things and religion, not an assassin but the most cruel butcher of citizens and allies. So I think this is the one defendant within human memory whom it would be advantageous to have condemned. For who does not understand that, even if he is acquitted against the will of gods and men, he can in no way be torn out of the hands of the Roman people? Who does not see clearly that we shall be reckoned to have come off splendidly if the Roman people shall be content with this man's punishment alone, and shall not decide thus: that this man has not committed a greater crime against himself --- when he plundered shrines, when he killed so many innocent men, when he afflicted Roman citizens with death, torture, and the cross, when he released the leaders of the pirates after taking money --- than those, if any, who, having sworn, by their vote shall have freed this man, covered with so many great unspeakable crimes?

\ciceroSection{10} There is not, there is not in this man any room for any erring, gentlemen. He is not the kind of defendant, this is not the moment, this is not the council --- I fear lest I seem to speak too arrogantly before such men --- this is not even the prosecutor through whom a defendant so guilty, so ruined, so convicted, can either be secretly slipped away or torn from us with impunity. Shall I not prove to these jurors that Gaius Verres took money against the laws? Will such men endure that they did not credit so many senators, so many Roman knights, so many cities, so many honourable men from so distinguished a province, the letters of so many peoples and private men --- that they resisted so great a will of the Roman people?

\ciceroSection{11} Let them endure it: we shall find, if we can bring this man alive to another trial, those before whom we may prove that this man in his quaestorship turned aside the public money given to Gnaeus Carbo the consul; we shall find those whom we may persuade that this man, in another's name --- as you learned in the first hearing --- took money from the urban quaestors. There will be those who will censure his audacity in this too, that under several names he subtracted as much as suited him from the head of the tithe-grain. There will perhaps be those, gentlemen, who will think that embezzlement of his most keenly to be punished, in that he did not hesitate to carry off the monuments of Marcus Marcellus and Publius Africanus --- which in their name, but in fact were and were considered the Roman people's --- from the most religious shrines and from the cities of allies and friends.

\ciceroSection{12} Suppose he has come out from the embezzlement trial too: let him meditate about the pirate captains whom he released after taking money, let him see what to answer about those whom he kept hidden in his house in their place; let him consider not only how he can heal our charge but how to deal with his own confession. Let him remember that, in the previous hearing, stirred by the angry and hostile shout of the Roman people, he confessed that the pirate leaders had not been struck with the axe; that he was already then afraid lest it be brought against him as a charge that he had freed them with money. Let him admit --- which cannot be denied --- that, being a private man, he kept the pirate captains alive and unharmed in his house at Rome, until I made him stop. If he can show in that treason trial that this was permitted to him, I will admit it ought to have been allowed. Suppose he escapes from this too: I will go where the Roman people has long been calling me.

\ciceroSection{13} For of the right of liberty and of citizenship the Roman people thinks the judgement is its own --- and rightly so. Let him by all means break through the senatorial councils with his violence; let him burst the courts of all sorts; let him fly out from your strictness: believe me, he will be held by tighter snares before the Roman people. The Roman people will believe these Roman knights, who, produced before you as witnesses earlier, said in their own sight that a Roman citizen, who offered honourable men as bondsmen, was hung on the cross by him.

\ciceroSection{14} The thirty-five tribes will believe that most weighty and most distinguished man, Marcus Annius, who said that in his presence a Roman citizen was struck with the axe. The Roman people will hear that leading man, the Roman knight Lucius Flavius, who in his testimony said that his close friend Herennius, a businessman from Africa, when more than a hundred Roman citizens at Syracuse recognised him and defended him with tears, was struck with the axe. Lucius Suettius will give his good faith and authority and religion --- a man endowed with all distinctions, who said under oath before you that many Roman citizens were most cruelly put to death by violence, by his order, in the stone quarries. When I conduct this case from a higher position, by the kindness of the Roman people, I do not fear that any force can snatch this man from the votes of the Roman people, or that any service of my aedileship can be more ample or more pleasing to the Roman people.

\ciceroSection{15} Wherefore let all attempt all things in this trial. There is now nothing left in this case in which any man can err, gentlemen, except at your own peril. My method has been recognised in past matters; in what remains, it has been tested and provided for. I showed my zeal for the commonwealth at the time when, after a long interval, I brought back the old custom; and at the request of allies and friends of the Roman people --- who are also my close associates --- I laid the indictment of a most audacious man. Which deed of mine the most chosen and most distinguished men --- of whom several were among you --- so approved that to the man who had been Verres's quaestor, and was wronged by Verres and was pursuing just enmities, they refused not only the laying of the indictment, but even the subscription, when he asked for it.

\ciceroSection{16} I went to Sicily for the inquiry. In which business my industry was shown by the speed of my return, my diligence by the multitude of documents and witnesses, and my modesty and conscience by the fact that, when I had come as a senator to the allies of the Roman people --- I, who had been quaestor in that province --- I lodged with my own friends and connections rather than with those who had asked me for help, the defender of the common cause. To no one was my coming a labour or expense, either publicly or privately. I used in inquiry only as much force as the law gave me, not as much as the zeal of those whom this man had harassed could have given me.

\ciceroSection{17} When I returned to Rome from Sicily --- though Verres and his friends, refined and city-bred men, had spread rumours of this kind, that the witnesses might be slowed up: that I had been led off from a real prosecution by a great sum of money --- although it was credited by no one (because both from Sicily there were witnesses who had known me as quaestor in the province, and from here the most distinguished men, who as they themselves are well known, so know each one of us best) --- yet I feared lest anyone should doubt my fidelity and integrity, until we came to the rejection of the jurors. I knew that, in rejecting jurors in our own memory, some had not avoided the suspicion of a deal, while in the prosecution itself their industry and diligence was approved.

\ciceroSection{18} I rejected jurors in such a way that this is agreed: that since the present state of the commonwealth which we now use, no council of similar splendour and dignity has existed. Which praise this man says is common between him and me. He, when he had rejected Publius Galba as juror, kept Marcus Lucretius; and when his patron asked him why he had let his closest friends, Sextus Peducaeus, Quintus Considius, Quintus Junius, be rejected, he replied: "Because I knew them in giving judgement to be too independent in their voting and view."

\ciceroSection{19} So with the jurors rejected, I now hoped that my burden was shared with you. I thought that not only to those who knew me but to those who did not, my fidelity and diligence had been proved. In which I was not deceived. For at my elections, when this man used unlimited bribery against me, the Roman people judged that the man's money, which against my faith had had no power with me, ought to have no power with them against my honour. On which day, gentlemen, when you first sat as called against this defendant, who was so unjust to this order, who was so eager for new things, for new courts and new jurors, that he was not moved by your appearance and your assembling?

\ciceroSection{20} When in this matter your dignity rewarded me for my diligence, this is what I gained: that in the one hour in which I began to speak, I cut off from a defendant audacious, moneyed, lavish, and ruined, the hope of corrupting the trial; that on the first day, with so great a number of witnesses cited, the Roman people judged that, with this man acquitted, the commonwealth could not stand; that the second day took from this man's friends and defenders not only the hope of victory but the will to defend; that the third day so cast the man down that, by feigning illness, he debated not what he should answer but how not to answer. Then on the remaining days, by these charges, by these witnesses both city and provincial, he was so overwhelmed and crushed that, with the days of the games interposed, no one judged him postponed for further hearing, but condemned.

\ciceroSection{21} Wherefore, so far as concerns me, gentlemen, I have won. For I have not coveted the spoils of Gaius Verres, but the reputation of the Roman people. It was for me to come with a cause to the prosecution: what cause was more honourable than to be appointed and chosen as defender by so distinguished a province? To take counsel for the commonwealth: what so suits the commonwealth as in such great resentment against the courts to bring forward a man by whose condemnation the whole order could be in praise and favour with the Roman people? To show and persuade that a guilty man has been brought up: who in the Roman people is there who has not taken away from the previous hearing this conviction --- that all the crimes, thefts, and disgraces of those previously condemned, if they were brought together into one place, would scarcely be matched and compared with a small part of his?

\ciceroSection{22} Look you to and consult, gentlemen, what concerns your reputation, your estimation, and the common safety. Your splendour ensures that you cannot err without the highest detriment and danger to the commonwealth. For the Roman people cannot hope that there are others in the senate who can rightly judge, if you cannot. It is necessary, when it has despaired of the whole order, to seek another kind of men and another method of giving judgements. If this seems lighter to you because you think it is a heavy and inconvenient burden to give judgement, you ought to understand first that it makes a difference whether you yourselves throw off this burden, or whether, because you could not prove your fidelity and conscience to the Roman people, the power of giving judgement is for that reason taken from you. And then to consider also: in what danger we shall come before those jurors whom, on account of hatred of us, the Roman people shall have wished to give judgement on us.

\ciceroSection{23} I will tell you what I have understood, gentlemen. Know that there are some men whom such a hatred of our order possesses that they openly say that they wish this man --- whom they know to be most dishonest --- to be acquitted on this one count alone: that the courts may, with disgrace and shame, be taken from the senate. This has compelled me to deal with you in many words, gentlemen --- not my fear about your fidelity, but a new hope of theirs which, having suddenly drawn Verres back from the gate to the trial, made some suspect that not without cause his plan had been so suddenly changed.

\ciceroSection{24} Now, lest Hortensius use a new kind of complaint and say that a defendant is being crushed about whom the prosecutor says nothing; that nothing is so dangerous to the fortunes of the innocent as for adversaries to be silent; and that he should not praise my ability differently than I wish, when he says that, if I had said many things, I would have lifted up the man against whom I was speaking, but that, because I did not say them, I had ruined him: I shall give way to him. I will use a continuous oration --- not because this is now necessary, but that I may test whether he bears it more troublesomely that I was silent then or that I am speaking now.

\ciceroSection{25} Here you will perhaps be diligent to see that I do not remit any hour from my legal hours. Unless I shall have used up every moment which the law has granted me, you will complain, you will invoke the faith of gods and men: that Gaius Verres is being circumvented because the prosecutor refuses to speak as long as he is allowed. Shall I not be allowed not to use what the law has given me for my own sake? For the time of accusation has been given me for my own sake, that I may be able to set forth in my speech the charges and the case. If I do not use it, I do you no wrong, but I take something from my own right and convenience. "But the case must be heard." For this reason indeed: because otherwise the defendant, however guilty, cannot be condemned. Did you, then, take it ill that something was done by me to keep this man from being condemned? For with the case heard many can be acquitted; with it not heard, no one can be condemned.

\ciceroSection{26} "I take away the second hearing." That which the law has in itself most troublesome, that the case should be pleaded twice --- which is established more for my sake than yours, or no more for yours than mine. For if to plead twice is convenient, surely it is common to both. If it is needful that he who spoke later be refuted, it has been established for the prosecutor's sake that the matter should be conducted twice. But, I take it, Glaucia first proposed that the defendant be put off for a second hearing. Before that, judgement could be given at the first or "amplius" pronounced. Which then do you think the milder law? I think the older one, by which one could be either quickly acquitted or slowly condemned. I am restoring to you that lex Acilia, by which many were once accused, the case once pleaded, the witnesses once heard, and condemned --- by no means with charges so manifest or so great as those by which you are convicted. Suppose you were pleading your case not under this so atrocious law, but under that mildest one. I shall accuse; you will answer; with the witnesses produced, I shall send the matter to the council in such a way that, even if the law gave the power of "amplius," yet they would think it disgraceful for them not to give judgement at the first.

\ciceroSection{27} But if it is necessary for the case to be heard --- has it not been heard enough? We pretend, Hortensius, what we have often experienced in speaking. Who paid great attention to us in this kind of case, where something is said to have been snatched or carried off by anyone? Is not all the expectation of the jurors either in the documents or in the witnesses? I said in the first hearing that I would make plain that Gaius Verres had taken from Sicily forty million sesterces against the laws. What? Should I have made this plainer if I had narrated it thus? "There was a certain Dio of Halaesa, who, when his son was an heir of his kinsman to a very great inheritance under the praetor Gaius Sacerdos, had at that time no business or controversy. Verres, as soon as he touched the province, immediately gave letters at Messana, summoned Dio, set against him slanderers from his own bosom who said that the inheritance had fallen to Erucine Venus; about this matter he showed he himself would judge."

\ciceroSection{28} I can next unfold the whole matter, then say in conclusion what happened: that Dio counted out a million sesterces to win the most certain case; that besides, his herds of mares were carried off by him, that the silver, the embroidered tapestries he had he saw to be carried off. These things, neither when I said them nor when you denied them, would our speech be of great moment. At what time, then, would the juror prick up his ear and apply his mind? When Dio himself came forward; when others who were then in Sicily and present at Dio's affairs; when through those very days when Dio was pleading the case, it was found he had taken money on loan, called in his debts, sold properties; when documents of good men were produced; when those who gave money to Dio said that they had then heard those funds were being taken to be given to Verres; when Dio's friends, hosts, patrons, the most honourable men, said they had heard the same things.

\ciceroSection{29} I take it, when these things were happening, then you would listen --- as you have listened --- then the case would seem truly to be conducted. Thus I have done all things in the previous hearing in such a way that on no charge was there any in which any of you needed a continuous prosecution. I deny that anything has been said by the witnesses which was either obscure to any of you or required any orator's eloquence. For you remember that I myself acted in such a way that, in interrogating witnesses, I set forth and unfolded all charges; so that, when I had placed the whole matter in plain view, I then at last interrogated the witness. So not only do you, who must judge, hold our charges; but the Roman people too has learned the whole prosecution and the case. Although I speak of my deed as if I had done that of my own will rather than driven by your wrong.

\ciceroSection{30} You interposed a prosecutor who, when I had asked for myself only one hundred and ten days for Sicily, asked one hundred and eight for himself for Achaia. When you had snatched from me three months most fitted for action, you thought that I would yield up to you all the rest of this year's time --- so that, when I had used up our hours, you, with two sets of games interposed, would answer on the fortieth day after; then time would be drawn out so that we would come from Manius Glabrio the praetor and from a great part of these jurors to another praetor and other jurors.

\ciceroSection{31} If I had not seen this --- if all my acquaintances and strangers had not warned me that this was being done, this was being plotted, this was being laboured at --- that the matter should be put off to that moment --- I might, I suppose, fear that, if I wished to use my hours in prosecuting, my charges would not last out, my speech would fail, my voice and strength would give way; that I should not be able to prosecute twice the man whom no one had dared to defend in the first hearing. I have proved my plan both to the jurors and to the Roman people. There is no one who thinks the wrong and shamelessness of these men could have been opposed in any other way. For with what folly should I have been, if --- when those who had bought to rescue Verres set down the day in their bond ("if the matter goes to the council after the first of January") --- I should have fallen on that day myself, when I could have avoided it?

\ciceroSection{32} Now I must take careful account of the time given me for speaking, since it is in my mind to set out the whole case. So I shall pass over the first act of his most disgraceful and flagitious life. He shall hear nothing from me about the disgraces of his boyhood, nothing from that impure youth of his --- which kind it was, you either remember, or, from the man whom he has produced as most like himself, you can recognise. I will pass over everything that seems shameful to me to mention; nor will I consider only what it suits him to hear, but what it suits me to say. I beg you to grant this and to allow my modesty: that I may keep silent on some part of his shamelessness.

\ciceroSection{33} All that time which was before he came to magistracies and the commonwealth, let it be free and unbound for me. Let his nightly orgies and vigils be passed over in silence; let no mention be made of pimps, gamblers, panders. Let the losses, the disgraces which his father's estate, his own youth bore, be passed by. Let him profit by the silence about his earlier infamy; let his rest of his life suffer me to make this great loss of charges.

\ciceroSection{34} You were quaestor to Gnaeus Papirius the consul, fourteen years ago. From that day to this, what you have done, I summon to court: not an hour will be found empty of theft, crime, cruelty, disgrace. These are the years spent in the quaestorship and the legateship of Asia and the city praetorship and the Sicilian praetorship. So this same fourfold division will be of my whole prosecution. As quaestor, you drew the province by senatorial decree: the consular fell to you, that you should be with the consul Gnaeus Carbo and hold that province. There was at that time dissension among the citizens, about which I will say nothing of what you ought to have felt. This one thing I say: at such a time and in such a lot, you ought to have decided which side you preferred to feel for and to defend. Carbo bore it ill that a man of singular luxury and idleness had fallen to him as quaestor; nevertheless he honoured him with all kindnesses and offices. To make this short: the money was assigned and counted out. The quaestor set out for the province; he came to the consular army in Gaul as expected, with money. As soon as the first opportunity seemed to be given him --- learn how a man begins his magistracies and his administration of the commonwealth --- the public money turned aside, the quaestor deserted his consul, his army, his lot, and his province.

\ciceroSection{35} I see what I have done. He pricks up his ears; he hopes that some breath can be blown his way in this matter, by the will and defence of those to whom the name of the dead Gnaeus Carbo is hateful, with whom he hopes that abandonment and betrayal of his consul will be welcome. As if he had done this from a desire to defend the nobility or zeal for that party, and not most openly plundered the consul, the army, and the province, and run away because of the most shameless theft! For it is obscure and his deed of such a kind that anyone might suspect that Gaius Verres, because he could not endure new men, had crossed over to the nobility, that is, to his own kind --- and had done nothing for the sake of money!

\ciceroSection{36} Let us look at his accounts, how he rendered them: now he himself will show why he left Gnaeus Carbo, now he will betray himself. First learn the brevity. "Received," he says, "two million two hundred thirty-five thousand four hundred seventeen sesterces. Paid out for pay, grain, legates, the deputy quaestor, the praetorian cohort, one million six hundred thirty-five thousand four hundred seventeen sesterces. Remaining: at Ariminum, six hundred thousand sesterces." Is this how to render accounts? Have I or you, Hortensius, or anyone in the world, ever rendered them in this fashion? What is this? What shamelessness, what audacity? What example, out of so many men's accounts rendered, is of this kind? But that six hundred thousand sesterces, which he could not falsely set down to whom they had been given, which he writes were left at Ariminum, those very six hundred thousand which were left over --- Carbo did not touch them, nor did Sulla see them, nor were they brought back to the treasury. He chose Ariminum as his town --- a town which, when he was rendering his accounts, had been crushed and plundered. He did not suspect, what he will now perceive, that we have witnesses enough left from that calamity of the Ariminenses for this matter. Read on.

\ciceroSection{37} "To Publius Lentulus and Lucius Triarius, urban quaestors, the matter of the accounts was rendered." Read: "By decree of the senate." So that he might be permitted to render his account in this fashion, he became a Sullan suddenly --- not that honour and dignity might be restored to the nobility. But if you had fled empty from there, still that wicked desertion and unholy betrayal of your consul would be judged a crime. "Carbo was a bad citizen, a wicked consul, a seditious man." So he was to others; when did he begin to be so to you? After he had committed to you his money, his grain supply, all his accounts, and his army. For if he had earlier displeased you, you would have done what Marcus Piso did the next year. When the quaestorship had fallen to him to Lucius Scipio the consul, he did not touch the money, did not set out for the army; what he felt about the commonwealth, he felt without injuring his own faith, the custom of our ancestors, or the bond of the lot.

\ciceroSection{38} For if we wish to confound and confuse all these things, we shall make all life dangerous, hateful, and hostile --- if the lot will have no religion, the bond of mutual fortune (whether favourable or doubtful) no companionship, the customs and institutions of our ancestors no authority. He is the common enemy of all who has been the enemy of his own. No wise man has ever thought a betrayer should be trusted. Sulla himself, to whom this man's coming ought to have been most welcome, removed him from himself and from his army; he ordered him to be at Beneventum, with those whom he understood to be most friendly to his party, where he could in nothing harm the highest matter and cause. He afterwards bestowed rewards on him generously, granted him certain goods of the proscribed in the Beneventan country to be plundered. He gave honour as to a betrayer, not faith as to a friend.

\ciceroSection{39} Now however much there are men who hate the dead Gnaeus Carbo, they ought to think not what they wished to happen to him, but what is to be feared by themselves in such a case. This is a common evil, a common fear, a common danger. There are no more hidden snares than those which lie hidden in the pretence of duty or in some name of close association. For one who is openly an adversary you can easily avoid by being on guard; but this hidden, internal, household evil not only does not appear, but even crushes you before you can see and explore it. Is it really so?

\ciceroSection{40} You, when you have been sent as quaestor to the army --- the guardian not only of the money but also of the consul, partner in all matters and counsels, held in the place of a son, as the custom of our ancestors used --- shall you suddenly leave him, desert him, cross over to the adversaries? O crime! O monster to be exported to the farthest lands! For that nature which has committed so great a deed cannot be content with this one crime. It must always plot something of the kind, must always be engaged in similar audacity and treachery.

\ciceroSection{41} So this same man, whom Gnaeus Dolabella afterwards, when Gaius Malleolus had been killed, had as deputy quaestor --- I do not know whether this connection was not even greater than that with Carbo, and the judgement of the will ought not to count more than that of the lot --- the same man was to Gnaeus Dolabella as he had been to Gnaeus Carbo. For he turned the charges that had weight against himself onto Dolabella, and brought the whole case of his patron over to Dolabella's enemies and prosecutors; he himself spoke a most hostile and most dishonest testimony against the man to whom he had been legate, to whom he had been deputy quaestor. That wretched Gnaeus Dolabella was burnt down --- both by his unspeakable betrayal and by his dishonest and false testimony --- and most of all by the resentment against this man's own thefts and disgraces.

\ciceroSection{42} What shall you do with this man? For what hope shall you reserve so treacherous, so intolerable a beast --- who in Gnaeus Carbo neglected and violated the lot, in Gnaeus Dolabella the will, and both not only deserted but betrayed and assaulted? Do not, gentlemen, please, weigh these charges by the brevity of my speech rather than by the magnitude of the matters themselves; for I must necessarily hurry to set out for you everything that has been laid down for me.

\ciceroSection{43} Wherefore, his quaestorship being demonstrated and the theft and crime of his first magistracy seen through, attend to the rest. In which I shall pass over that period of the Sullan proscriptions and plunderings; nor will I let him take any defence for himself out of the common calamity. I shall accuse him of his own definite and proper crimes. Wherefore, with all this Sullan period excluded from the prosecution, learn his splendid legateship.

\ciceroSection{44} After Cilicia was decreed as Gnaeus Dolabella's province --- O immortal gods, with what greed, by what intrigues, did this man storm that legateship for himself! That was the beginning of Gnaeus Dolabella's greatest calamity. For as he set out, wherever he made his journey, he was such that he seemed to spread not as a legate of the Roman people but as a kind of calamity. In Achaia --- I will pass over all smaller things, of which something similar perhaps another man too has done at some time. I will say nothing but what is singular, what, if it were said against another defendant, would seem incredible --- he demanded money of the magistrate of Sicyon. Let this not be a charge against Verres: others have done the same. When the man would not give it, he punished him --- dishonestly, but not unheard of.

\ciceroSection{45} See what kind of punishment --- you will ask from what kind of man you judge him. He ordered a fire to be made of green and damp wood in a narrow place; there he left a free-born man, noble in his own city, an ally and friend of the Roman people, half-dead, tormented by the smoke. As for the statues, the painted panels he carried off from Achaia --- I will not speak in this place; I have another place set apart for showing his greed for these things. At Athens, you have heard that a great mass of gold was carried off from the temple of Minerva; this was said in Gnaeus Dolabella's trial. Said? Even assessed. You will find that Gaius Verres was not a sharer in this counsel, but the leader.

\ciceroSection{46} He came to Delos. There from the most religious shrine of Apollo by night he secretly took away the most beautiful and most ancient statues, and saw to it that they were thrown into his cargo ship. The next day, when those who lived at Delos saw the shrine plundered, they took it ill; for so great is the religion and antiquity of that shrine among them that they think Apollo himself was born in that very place. Yet they did not dare say a word, lest the matter perhaps concern Dolabella himself. Then suddenly the greatest tempests arose, gentlemen, so that not only could Dolabella, although he wished, not set out, but he could scarcely stay in the town: so great were the waves cast up. Here that ship of this brigand, loaded with sacred statues, was driven out and broken by the waves; on the shore those statues of Apollo are found. By Dolabella's order they are restored. The tempest is calmed; Dolabella sets out from Delos.

\ciceroSection{47} I do not doubt that, although there has never been any feeling of humanity in you, never any thought of religion, yet now in your fear and danger your own crimes come to your mind. Can any hope of safety be conveniently shown you, when you remember how impious, how wicked, how unspeakable you have been towards the immortal gods? Did you dare to plunder Apollo of Delos? Did you dare to lay impious and sacrilegious hands on that temple, so ancient, so sacred, so religious? If in your boyhood you were not so trained in those arts and disciplines that you should learn and know what has been entrusted to letters --- not even afterwards, when you came to those very places, were you able to learn what is handed down by memory and writing:

\ciceroSection{48} that Latona, after long wandering and flight, when pregnant and her time of delivery now upon her, fled to Delos and there bore Apollo and Diana? From which opinion of men that island is thought sacred to those gods, and the authority of its religion is so great, and always has been, that not even the Persians, when they had declared war on all of Greece, gods and men, and brought a fleet of a thousand ships to Delos, attempted to violate or touch anything. Did you, most dishonest and most senseless of men, dare to plunder this shrine? Was there ever a greed so great that it could overcome so great a religion? And if you did not think of these things then, do you not now even remember that no evil is so great but it has long been owed you for your crimes?

\ciceroSection{49} When he came to Asia --- why should I speak of his receptions, his lunches, dinners, horses, and gifts? I will not deal with Verres on his daily crimes. At Chios, I say, he carried off the most beautiful statues by force; likewise at Erythrae and Halicarnassus. At Tenedos --- I pass over the money he plundered --- Tenes himself, who is held by the Tenedians the most sacred god, who is said to have founded their city, from whose name Tenedos is named --- this very Tenes, I say, most beautifully wrought, whom you once saw in the Comitium, he carried off, with the great groan of the city.

\ciceroSection{50} As for that storming of the most ancient and most noble shrine of Samian Juno --- how grievous to the Samians, how bitter to all Asia, how known to all, how unheard of by no one of you! On which storming, when ambassadors had come from Samos to Gaius Nero in Asia, they took back this answer: that complaints of this kind, which concerned legates of the Roman people, ought to be referred not to the praetor but to Rome. What documents did this man take from there, what statues! Which I myself recently recognised in his house, when I had come for the sake of sealing them up.

\ciceroSection{51} Where now, Verres, are these statues? Those, I mean, which we recently saw at your house, set out by all the columns, in all the spaces between the columns, and finally in your park, under the open sky. Why, as long as you thought another praetor with the jurors you had drawn for the substitution would be entering on the council about you --- so long they were at home? After you saw that we preferred to use our witnesses than your hours, you left no statue at home except two which are in the middle of the house, which were carried off from Samos itself. Did you not think I would summon as witnesses on this matter your closest friends, who had always been at your house, of whom I might inquire whether they knew the statues had been there which were not now there?

\ciceroSection{52} What did you suppose those men would judge of you, when they saw that you were now fighting not against your prosecutor but against an inheritance-officer and a confiscation-broker? Concerning which matter, you heard Charidemus of Chios speak in evidence in the previous hearing: that, when he was a trierarch and was escorting Verres from Asia by Dolabella's order, he was at Samos with him, and that he then knew the shrine of Juno and the town of Samos were plundered. And afterwards he pleaded his case publicly before his fellow citizens at Chios, the Samians being his accusers. He was acquitted, because he made it plain that what the legates of the Samos said pertained to Verres, not to him.

\ciceroSection{53} Aspendus --- you know it is an old and noble town in Pamphylia --- is full of the best statues. I will not say this statue was taken from there and that one. I say this: that you left no statue at Aspendus, Verres. Everything from the shrines, from public places, openly, in everyone's sight, was carried out and exported on wagons. And even that famous Aspendian harpist, of whom you have often heard --- the one whom the Greeks proverbially say "plays everything within" --- he carried off and set up in his innermost chambers, so that he might seem to have surpassed even him in his own art.

\ciceroSection{54} At Perga we know there is the most ancient and most sacred shrine of Diana. That too has been stripped and plundered by you; from Diana herself, what gold she had, has been pulled off and carried away, I say. What, in evil hour, is this so great audacity and madness? For the cities of allies and friends to which you came under the right and name of legate --- if you had invaded them by force, with army and command, yet, I take it, the statues and ornaments which you carried off from those cities you would have carried off not to your own house, nor to the country-houses of your friends, but to Rome, to the public.

\ciceroSection{55} What shall I say of Marcus Marcellus, who took Syracuse, that most splendid of cities? What of Lucius Scipio, who waged war in Asia and conquered Antiochus, the most powerful king? What of Flamininus, who subdued King Philip and Macedon? What of Lucius Paulus, who overcame King Perses by force and valour? What of Lucius Mummius, who took the most beautiful and most ornate city of Corinth, full of every kind of treasure, and joined many cities of Achaia and Boeotia under the rule and dominion of the Roman people? Whose houses, when they flourished in honour and virtue, were empty of statues and panel paintings; but the whole city and the temples of the gods and all parts of Italy we see adorned with their gifts and monuments.

\ciceroSection{56} I am afraid lest these things may seem to someone perhaps too ancient and outworn now; for then all were equally of this kind, so that this praise of singular virtue and innocence seems to be of those men, indeed of those times, not of these. Publius Servilius, a most distinguished man, with the greatest deeds done, is here to give his vote on you. He took Olympus by force, by troops, by counsel, by valour --- an ancient city, increased and adorned with all things. I produce a recent example of a most brave man. For Servilius, commander of the Roman people, took Olympus, an enemy city, after you in those same parts as a quaestorian legate had pacified towns of allies and friends seen to be plundered and harassed.

\ciceroSection{57} What you carried off by crime and brigandage from the most religious shrines, we cannot see except in your own house and those of your friends. The statues and ornaments which Publius Servilius carried off from an enemy city taken by force and valour, by the law of war and the right of imperator, he brought to the Roman people, conveyed in his triumph, saw to be inscribed on the public tablets at the treasury. Learn from the public letters the diligence of so distinguished a man. Read: "Accounts rendered by Publius Servilius." You see that not only the number of the statues but the size, shape, and posture of each one is set down in the documents. Surely the joy of valour and victory is greater than this pleasure which is taken from lust and greed. I say Servilius keeps the booty of the Roman people much more carefully marked and recorded than you keep your thefts.

\ciceroSection{58} You will say that your statues and panel paintings were also an ornament to the city and the Forum of the Roman people. I remember; with the Roman people I saw the Forum and the Comitium adorned for show with magnificent ornament, but bitter and mournful to the senses and the mind. I saw everything ablaze with your thefts, the booty of the provinces, the spoils of allies and friends. At which time, gentlemen, this man got the greatest hope for his other crimes. For he saw that those who wished to be called the masters of the courts were the slaves of these greeds.

\ciceroSection{59} The allies and foreign nations then for the first time threw away all hope of their affairs and fortunes, because by chance there were at Rome at that time very many ambassadors from Asia and Achaia, who venerated in the Forum the statues of the gods carried off from their own shrines, and likewise, when they recognised other statues and ornaments, looked at them in tears, one in this place, another in that. The talk of all these we then heard was that there was no reason for anyone to doubt about the destruction of the allies and friends, when indeed they saw in the Forum of the Roman people, in that place where formerly those who had done injuries to allies were accused and condemned, there openly placed those things which had been taken and torn from the allies by crime.

\ciceroSection{60} Here I do not think he will deny that he has very many statues, innumerable panel paintings; but, I take it, he is accustomed sometimes to say that what he has seized and stolen he bought, since indeed into Achaia, Asia, Pamphylia, at public expense and in the name of an embassy, he was sent as a buyer of statues and panel paintings. I have received all the documents both of him and of his father, which I have read most diligently and digested: his father's, while he lived; yours, as long as you say you kept them. For in him, gentlemen, you will find this novel thing. We hear that some never kept books --- which is the false opinion men have of Antonius, for he kept them most diligently; but let it be a certain kind of conduct, by no means to be approved. We hear that another did not keep them from the start, but kept them from some time onward; this also has some method. But this is novel and ridiculous: that this man answered us when we asked him for his books, that he had kept them up to the consulship of Marcus Terentius and Gaius Cassius, but afterwards had stopped.

\ciceroSection{61} In another place we shall consider what kind of conduct this is. Now it does not concern me; for of these times in which I am now engaged I have both your books and your father's. You cannot deny that you have brought back very many of the most beautiful statues, very many of the best panel paintings. And would that you might deny it! Show me one --- in either your books or your father's --- bought: you have won. Even those two most beautiful statues which now stand by your impluvium, which for many years stood before the doors of Samian Juno, you cannot show how you bought; these two, I say, which alone are now in your house, which await the inheritance-officer, abandoned and left behind by the other statues.

\ciceroSection{62} But, I take it, only in these matters had he unbridled and unchecked greeds. The rest of his lusts were held in by some method or moderation. How many freeborn women, how many mothers of households, do you suppose he assaulted in that filthy and impure legateship? On what town did he set foot where he did not leave behind more traces of his rapes and disgraces than of his coming? But I shall pass over everything that can be denied. I shall leave aside even those things which are most certain and most plain. I shall pick out one of his unspeakable deeds, that I may the more easily come at last to Sicily, which has imposed this burden and labour upon me.

\ciceroSection{63} There is in the Hellespont a town, gentlemen, called Lampsacum --- among the most known and noble in the province of Asia. The Lampsacenes themselves are exceedingly courteous to all Roman citizens, and besides most settled and quiet, almost beyond the rest fitted to the highest Greek leisure rather than to any violence or tumult. It happened, when this man had begged of Gnaeus Dolabella that he be sent to King Nicomedes and King Sadala, and had asked for this journey as suiting his own profit rather than the public business, that he came on that journey to Lampsacum with the great calamity and almost ruin of the city. He was lodged with a certain doorkeeper as host, and his companions were placed likewise with other hosts. As was his way, and as his disgraceful lusts were prompting him, he immediately gave instructions to those companions of his, the most worthless and most disgraceful men, to look out and find any maiden or woman on whose account he might tarry longer at Lampsacum.

\ciceroSection{64} His companion was a certain Rubrius, a man made for his lusts, who, with marvellous art, wherever he came, used to investigate all such matters. He brought this report to him: that there was a certain Philodamus, easily the leading man of the Lampsacenes by birth, honour, wealth, and reputation; that he had a daughter who lived with her father (because she had no husband), a woman of singular beauty, but reckoned of the highest integrity and chastity. As soon as he heard this, the man so caught fire for what he had not only never seen himself but had not even heard from one who had seen, that he immediately said he wished to move to Philodamus's. The host doorkeeper, suspecting nothing, fearing lest he himself be offended in some way, began with great force to keep him. Verres, since he could not find a reason for leaving the host, began to make a way to his rape by another method: he said Rubrius, his darling, his helper and accomplice in all such matters, was poorly lodged. He ordered him to be moved to Philodamus.

\ciceroSection{65} When this was reported to Philodamus, although he was unaware how much evil was already being prepared for him and his children, he came to Verres. He pointed out that this duty was not his; that, since it was the part of hosts to receive guests, even so they were accustomed to receive praetors and consuls themselves, not the lackeys of legates. Verres, since he was carried away by one greed, ignored that whole demand and reasoning. He ordered Rubrius to be brought by force to a man who ought not to receive him. Here Philodamus, since he could not maintain his own right, laboured to keep his humanity and habit. The man, who had always been thought most hospitable and friendly to our citizens, did not wish to seem to have received that very Rubrius unwillingly. Magnificently and with adornment, since he was among the foremost of his people in resources, he prepared a banquet. He asked Rubrius to invite whomever he pleased, leaving the place to him alone if he saw fit. He even sent his son, a most select young man, abroad to a certain kinsman of his to dinner.

\ciceroSection{66} Rubrius invited Verres's companions; Verres made them all aware of what was needed. They came in good time; they reclined at table. Conversation arose between them, and an invitation that they should drink in the Greek manner; the host urged it; they called for larger cups; the banquet was kept up by everyone's talk and merriment. After things seemed to Rubrius to be heating up enough, he said: "Please, Philodamus, why do you not order your daughter to be called in to us?" The man, who was both of the highest gravity and of that age and a parent, was struck dumb at the dishonest man's saying. Rubrius pressed. Then he, that he might give some answer, said it was not the custom of the Greeks for women to recline in a banquet of men. Then another from another part: "But this is not to be borne; let the woman be called!" And at the same time Rubrius ordered his slaves to shut the door and themselves to stand at the doorway.

\ciceroSection{67} When Philodamus understood that this was being done, that this was being prepared --- that violence should be done to his daughter --- he called his slaves to him, ordered them to disregard himself, to defend his daughter; that someone should run out and bring this great household evil to the news of the son. Meanwhile a shouting arose throughout the whole house. Among the slaves of Rubrius and the host this leading and most honourable man is being tossed about in his own house. Each man for himself laid hands. Finally hot water is poured on Philodamus by Rubrius himself. When this was reported to the son, immediately, beside himself, he hurries to the house, that he might come to the help of his father's life and his sister's chastity. All the Lampsacenes, with the same feeling, as soon as they heard this --- moved both by Philodamus's standing and by the magnitude of the wrong --- gathered to the house at night. Here Verres's lictor Cornelius, who, with his slaves, had been stationed by Rubrius almost as a guard for carrying off the woman, is killed; some slaves are wounded; Rubrius himself is wounded in the crowd. He, when he saw such great tumults raised by his own greed, longed in some way to fly, if he could.

\ciceroSection{68} On the next day men assembled in a public meeting in the morning. They asked what was best to do. Each man, according to his authority, spoke to the people. There was no one whose feeling and speech was not this: that they should not fear, if the Lampsacenes had avenged this man's unspeakable crime by force and hand, lest the senate and Roman people should think that city must be punished. But if the legates of the Roman people used this right against allies and foreign nations --- that it was not allowed to keep the chastity of one's children safe from their lust --- it would be better to suffer anything than to live amid such great violence and bitterness.

\ciceroSection{69} When all felt this, and each spoke to that effect according to his own feeling and grief, all set out for that house in which Verres was lodged. They began to break in the door with stones, to press with steel, to gather wood and brushwood about it, and to set fire. Then the Roman citizens who were doing business at Lampsacum ran together. They begged the Lampsacenes that the name of legate weigh more with them than the wrong of the legate. They understood the man was impure and unspeakable; but since he had not accomplished what he had attempted, and would not be at Lampsacum afterwards, their offence would be lighter if they spared a wicked man than if they did not spare a legate.

\ciceroSection{70} So this man, much more wicked and worthless than that famous Hadrianus, was somewhat luckier. The latter, because the Roman citizens could not bear his greed, was burnt alive in his own house at Utica; and that is judged to have happened to him so deservedly that everyone rejoiced and no investigation was set up. This man, scorched by an embassy of allies, yet flew out from that flame and danger; and not yet has he been able to invent any cause why he committed it, or what happened, that he came into so great a danger. For he cannot say: "When I wished to quell a sedition, when I was levying grain, when I was collecting pay, when I was finally doing something for the commonwealth, that I gave a sharper order, that I punished, that I threatened." If he should say these things, even so he ought not to be forgiven, if by too harshly commanding he seemed to have brought the allies to so great a danger.

\ciceroSection{71} Now, since he himself dares neither to speak truly nor to invent falsely the cause of that tumult, but a most temperate man of his own order, who was then assistant to Gaius Nero --- Publius Tettius --- has said that he learned these same things at Lampsacum; and a man of every distinction, Gaius Varro, who was then military tribune in Asia, says he himself heard these same things from Philodamus --- can you doubt that fortune wished to snatch this man not so much from that danger as to keep him for your judgement? --- unless he says what he interrupted Hortensius with in Tettius's testimony in the previous hearing (at which time, indeed, he gave a sufficient sign that, if there were anything he could say, he could not be silent --- so that, as long as he was silent during the other witnesses, all could know he had nothing to say): he then said, that Philodamus and his son had been condemned by Gaius Nero.

\ciceroSection{72} About which, that I may not say much, I say only this: that Nero followed this --- and his council too --- that since it was agreed the lictor Cornelius had been killed, they thought no man ought to have the power of killing a man even in avenging a wrong. Wherein I see by Nero's judgement you were not acquitted of dishonesty, but they were condemned of murder. But what kind of condemnation was that? Listen, please, gentlemen, and pity the allies at last and show that there ought to be some defence for them in your trust. Because it seemed to all Asia that he had been killed lawfully --- that lictor of his (in name; in fact, the minister of his most dishonest greed) --- this man feared lest Philodamus be acquitted by Nero's judgement. He asks and begs Dolabella to leave his province, to go to Nero. He explains he cannot be safe, if Philodamus had been allowed to live and at some time come to Rome. Dolabella was moved.

\ciceroSection{73} He did what many reproached: leaving his army, his province, and his war, and setting out for Asia, into another's province, for the sake of the most worthless of men. After he came to Nero, he urged him to take cognisance of Philodamus's case. He himself had come to be on the council and speak first; he had even brought along his own prefects and military tribunes, all of whom Nero called to council. There was on the council also that most equitable judge, Verres himself. There were not a few toga-clad creditors of the Greeks, with whom the favour of every most dishonest legate is most useful for collecting their money.

\ciceroSection{74} That wretched man could find no defender. For who, among toga-clad men, would not be moved by Dolabella's favour? --- or among Greeks, by the same man's force and command? Then a Roman citizen from among the Lampsacenes' creditors is set up as prosecutor, who, if he said what Verres ordered, could collect money from the people through the same man's lictors. While all these things were being done with such great contention, with such great resources --- when many were prosecuting that wretched man, no one defending; when Dolabella was fighting with his prefects on the council; when Verres said his own fortunes were at stake, and gave testimony, and was on the council, and had procured a prosecutor --- when all these things were happening, and although it was agreed a man had been killed, yet so great was the force of his wrong, so great his dishonesty, was thought, that on Philodamus's case "amplius" was pronounced.

\ciceroSection{75} What now in the second hearing shall I bring forward of Gnaeus Dolabella's spirits, what of this man's tears and runnings about, what of Gaius Nero's --- a most excellent and most innocent man --- mind that in some matters was too timid and humble? Who in that matter did not have anything he ought to do --- unless perhaps what all then desired, that he should conduct the matter without Verres and without Dolabella. Whatever had been done without these, all would have approved. Then indeed what was pronounced was thought to have been judged not by Nero, but snatched from Nero by Dolabella. For Philodamus and his son are condemned by very few votes. Dolabella is present, presses, urges, that they be struck with the axe as soon as possible, so that as few as possible could hear of this man's unspeakable crime.

\ciceroSection{76} A bitter and miserable spectacle, grievous to the whole province of Asia, was set up in the Forum at Laodicea: an aged parent led to punishment, on the other side his son --- the one because he had defended the chastity of his children, the other because he had defended his father's life and his sister's reputation. Each wept, not for his own punishment, but the father for the son's death, the son for the father's. What floods of tears do you suppose Nero himself shed? What weeping was there of all Asia, what mourning and groaning of the Lampsacenes? --- that innocent men, nobles, allies and friends of the Roman people, were struck with the axe, on account of the singular worthlessness and most dishonest greed of a most flagitious man!

\ciceroSection{77} Now indeed, Dolabella, neither for you nor for your children, whom you have left wretched in want and solitude, can I have pity. Was Verres so much to you that you wished his lust to be paid for with the blood of innocent men? Was it for this you abandoned army and enemy --- so that by your own violence and cruelty you might lift the dangers of this most dishonest man? Did you, because you had set him in the place of a quaestor, suppose he would be your friend forever? Did you not know that by him Gnaeus Carbo the consul, whose actual quaestor he had been, was not only deserted but also stripped of his auxiliaries and money, unspeakably attacked and betrayed? You experienced, then, his treachery, when he himself went over to your enemies, when the guilty fellow gave the keenest testimony against you, when he refused to render his accounts at the treasury except after your condemnation.

\ciceroSection{78} Shall your lusts, Verres, be so great that the provinces of the Roman people, the foreign nations, cannot hold and bear them? Shall what you have seen, what you have heard, what you have desired, what you have meditated --- unless it be at hand at your nod, unless it obey your lust and greed --- shall men be unleashed, houses be stormed, cities not only at peace but indeed of allies and friends shall flee to violence and arms, that they may be able to ward off from themselves and their children the crime and lust of a legate of the Roman people? For I ask of you: were you besieged at Lampsacum? Did that multitude begin to set fire to the house in which you were lodged? Did the Lampsacenes wish to burn alive a legate of the Roman people? You cannot deny it. For I have your testimony which you gave before Nero; I have the letters which you sent to him.

\ciceroSection{79} Read this very passage from the testimony. "Testimony of Gaius Verres against Artemidorus. Not long after, into the house..." Was the Lampsacene state attempting to make war on the Roman people? Did it wish to revolt from our empire and name? For I see, both from what I have read and what I have heard, I understand: in any city in which not only is a legate of the Roman people besieged, not only attacked by fire, steel, hand, troops, but in some part violated --- unless satisfaction is publicly made --- it is customary for war to be declared and waged on that city.

\ciceroSection{80} What then was the cause why the whole Lampsacene state from the public meeting (as you yourself write) ran together to your house? For neither in the letters which you send to Nero, nor in your testimony, do you show any cause for so great a tumult. You say you were besieged, fire was brought, brushwood was put round, your lictor was killed, the power of going out in public was taken from you. The cause of so great a terror you hide. For if Rubrius had done some wrong in his own name and not by your impulse and your greed, they would have come to you to complain of your companion's wrong rather than to attack you. Since therefore the witnesses produced by us have said what cause there was for that tumult, while you have hidden it: does not this account which we have proposed get confirmation both from their testimony and from your own continued silence?

\ciceroSection{81} Will you then spare this man, gentlemen, whose offences are so great that those whom he wronged could not wait the lawful time for vengeance, nor could put off so great a force of grief to a later moment? You were besieged. By whom? By Lampsacenes. Barbarian men, I take it, or men who despised the name of the Roman people. On the contrary, by men most gentle by nature, custom, and discipline; men allies by the condition of the Roman people, slaves by fortune, suppliants by will. So that it is plain to all: unless the bitterness of the wrong, the force of the crime had been so great that the Lampsacenes thought they should rather die than endure it, they would never have come to that point of being moved more vehemently by hatred of your lust than by fear of the legateship.

\ciceroSection{82} Do not, by the immortal gods, force allies and foreign nations to use this refuge, which, unless you act, they will necessarily use. Nothing would ever have softened the Lampsacenes towards this man, had they not believed that he would pay penalties at Rome. Even though they had received such a wrong as no law could pursue worthily enough, yet they preferred to commit their inconveniences to our laws and courts than allow them to their own grief. You, when you had been besieged by so distinguished a state on account of your crime and disgrace, when you had compelled wretched and stricken men, as if despairing of our laws and courts, to flee to violence, to hands, to arms, when in towns and cities of friends you had presented yourself not as a legate of the Roman people but as a lustful and cruel tyrant, when before foreign nations you had violated the fame of our empire and name by your reproaches and disgraces, when you escaped from the steel of the Roman people's friends and flew out from the flame of allies --- do you hope that here will be your refuge? You err. They allowed you to come out alive, that you might fall here, not that you might rest here.

\ciceroSection{83} And you say that judgement has been given that you were besieged at Lampsacum unjustly, because Philodamus with his son was condemned. What if I show, if I make it plain by a worthless witness, but on this matter still suitable --- by you yourself, I say, as witness, I shall show that you yourself transferred the cause and fault of this besieging onto others, and that those whom you charged were not punished. Now Nero's judgement does not help you at all. Read the letter which he sent to Nero. "Letter of Gaius Verres to Nero. Themistagoras and Thessalus..." You write that Themistagoras and Thessalus stirred up the people. What people? Those who besieged you, who tried to burn you alive. Where do you pursue them, where do you accuse them, where do you defend the right and name of legate? Will you say it was done in Philodamus's case? Bring me Verres's own testimony.

\ciceroSection{84} Let us see what the same man said under oath. Read: "Asked by the prosecutor, he replied that he was not pursuing the matter in this trial: that he had it in mind to pursue it at another time." How then does Nero's judgement help you, how Philodamus's condemnation? You, when you were besieged as a legate, when, as you yourself wrote to Nero, a notable wrong had been done to the Roman people and the common cause of legates --- you did not pursue it. You say you have it in mind to pursue at another time. What time was that? When did you pursue? Why did you diminish the right of legateship, why did you desert and betray the cause of the Roman people, why did you leave your wrongs joined with the public ones? Ought you not to have referred the matter to the senate, to have complained of so atrocious wrongs, to have seen to it that those who stirred up the people were summoned by the consuls' letters?

\ciceroSection{85} Recently when Marcus Aurelius Scaurus demanded --- because he said he had been prevented by force at Ephesus, when he was quaestor, from leading away from the shrine of Diana his slave who had taken refuge in that asylum --- Pericles of Ephesus, a most noble man, was summoned to Rome, since he was alleged to have been the author of that wrong. You, if you had shown the senate that you had been treated as legate at Lampsacum in such a way that your companions were wounded, your lictor killed, you yourself besieged and almost burnt --- and that the leaders and chief authors of this matter (as you write) were Themistagoras and Thessalus --- who would not have been moved? Who would not have provided for himself out of the wrong done to you? Who would not have thought that in this matter your case and the common danger were involved? For the name of legate ought to be such that not only among the rights of allies but even among the weapons of enemies it goes safe.

\ciceroSection{86} Great is this Lampsacene charge of lust and most dishonest greed. Now hear another, of greed almost not lighter in its kind. He demanded a ship of the Milesians, which should escort him to Myndus for protection. They at once gave him an excellent skiff from their fleet, equipped and armed. With this protection he set out for Myndus. As to what he carried off publicly from the Milesians in wool, and the expense of his coming, the contumelies and wrongs to the Milesian magistrate --- although these can be said both truly and gravely and vehemently --- I shall pass over speaking, and reserve them all whole for the witnesses. Hear what can in no way be passed in silence and cannot be expressed worthily.

\ciceroSection{87} He orders the soldiers and rowers to return on foot from Myndus to Miletus. He himself sells the most beautiful skiff, chosen out of the ten Milesian ships, to Lucius Magius and Lucius Fannius, who lived at Myndus. These are the men whom the senate recently judged should be held in the number of enemies. In that vessel they sailed to all the enemies of the Roman people, all the way from Dianium, which is in Spain, to Sinope, which is in Pontus. O immortal gods, incredible greed and singular audacity! Did you dare to sell a ship from the fleet of the Roman people, which the Milesian state had given you to escort you? Did the magnitude of the deed not move you? --- if men's regard did not move you, did you not consider this either, that of this most dishonest theft (or rather of so unspeakable a piece of plunder) so distinguished and so noble a city would be a witness?

\ciceroSection{88} Or because Gnaeus Dolabella then attempted, at your request, to punish the man who had been in command of the skiff, when he had reported the affair to the Milesians, and had ordered the report (which had been entered into their public records under their laws) to be expunged --- did you for that reason think you had escaped from this charge? That opinion has greatly deceived you, and in many places. For you have always thought, and especially in Sicily, that you were sufficiently safeguarded for defence if you had either forbidden anything to be entered into the public records, or had compelled what had been entered to be deleted. How nothing this is --- although you have learned it from many cities of Sicily in the previous hearing --- yet learn it also in this very city. They are obedient indeed, while those who order are present. As soon as they have left, they not only enter what they were then forbidden, but they also subscribe the cause why it was not entered in the documents at the time. Those documents remain at Miletus; they remain, and as long as that city exists they will remain.

\ciceroSection{89} For ten ships, by the order of Lucius Murena, the Milesian people had built from tax-money for the Roman people, just as the other cities of Asia had each its share. Wherefore one of the ten --- not by the sudden coming of pirates, but by the brigandage of a legate; not by the violence of a tempest, but by this horrible tempest of allies --- they entered in their public records as having been lost.

\ciceroSection{90} There are at Rome envoys of the Milesians, the most noble men and chief of their state, who, although they are awaiting the month of February and the names of the consuls-elect, yet so great a deed they will not be able even when interrogated to deny, nor even when produced to keep silent. They will say --- I say --- both moved by religion and from fear of their own laws, what was done with that skiff. They will show that Gaius Verres, in that fleet which had been built against pirates, was himself a wicked pirate. When Gaius Malleolus, Gnaeus Dolabella's quaestor, was killed, Verres thought two inheritances had come to him: one of the deputy quaestor's office, for he was at once ordered by Dolabella to be deputy quaestor; the other of guardianship, for, since he was the guardian of the ward Malleolus, he made an assault on his goods.

\ciceroSection{91} For Malleolus had set out for the province so amply furnished that he left absolutely nothing at home. Besides, he had placed loans with various peoples and made bond-notes; he had brought with him a great weight of the best chased silver --- for he too was a comrade of Verres in this disease and greed. He left a great weight of silver, a large household, many craftsmen, many handsome men. Verres seized whatever silver pleased him, led off the slaves he wished, carried away the wines and other things easily got in Asia which Malleolus had left, sold the rest, and exacted the money.

\ciceroSection{92} When it was agreed that he had realized two and a half million sesterces, when he returned to Rome, he gave no letter to the ward, none to his mother, none to the guardians. The slave-craftsmen of the ward he kept at home; the handsome and lettered men he kept around his feet, saying they were his, that he had bought them. When the mother and grandmother of the boy more than once demanded that, if he would not give back the money or render the account, he should at least say how much money of Malleolus's he had brought back, demanded by many, he at last said one million sesterces. Then in the bottom margin of the wax he made the lowest entry in disgraceful erasure: paid to the slave Chrysogonus, six hundred thousand sesterces; received from the ward Malleolus. How out of one million sesterces six hundred thousand were made, how those six hundred thousand sesterces fitted exactly --- as those of Gnaeus Carbo's money were left over six hundred thousand --- how the entry to Chrysogonus was made, why this entry is the lowest and in erasure, you will judge.

\ciceroSection{93} Yet, although he had returned six hundred thousand sesterces as received, fifty thousand sesterces were not paid. The slaves --- after he was made defendant --- some have been returned, some are still kept; the savings of all and the under-slaves are kept. This is his splendid guardianship. Behold, to whom you should commit your children! Behold the memory of his dead comrade! Behold the fear for the reputation of the living! When all Asia had presented itself to be plundered and harried, when all Pamphylia had been laid open to your plunder, were you not content with these so rich things? Could you not keep your hands from the guardianship, your hands from the ward, your hands from the son of your comrade? Now it is not Sicilians, not the farmers (as you keep saying), not those who, by your decrees and edicts, have been roused and made hostile to you, who beset you. Malleolus was produced by me, and his mother and his grandmother, who, weeping, said that the boy had been overturned out of his ancestral goods by you.

\ciceroSection{94} What are you waiting for? That Malleolus himself rise from the dead and demand of you the duties of guardianship, comradeship, and friendship? Suppose he is here. Most greedy and filthy man, give back the goods to your comrade's son, if not what you carried off, at least what you have confessed! Why do you compel your comrade's son to send forth this his first voice in the Forum with grief and complaint? Why do you compel his comrade's wife, his comrade's mother-in-law, finally the whole household of his dead comrade, to give testimony against you? Why do you force the most modest and most select women, unaccustomed and unwilling, to come forward into so great an assembly of men? Read the testimonies of all. Testimony of mother and grandmother.

\ciceroSection{95} As to how he, when deputy quaestor, harassed the community of the Milyads, how he afflicted Lycia, Pamphylia, Pisidia, and all Phrygia by levying grain, by his arbitrary valuation, by that "Sicilian assessment" of his (which he then first invented) --- it is not necessary to demonstrate in words. Know this: under these heads --- the things which were done by him when he levied grain, hides, hair-cloth, sacks from the cities, and did not take them but exacted money for those things --- under these heads alone, against Gnaeus Dolabella the lawsuit was assessed at three million sesterces. All of which, even if it was done at Dolabella's will, was nevertheless conducted through this man.

\ciceroSection{96} I shall stand on a single heading; for there are many of the same kind. Read: "Of moneys assessed against Gnaeus Dolabella, ex-praetor: from the community of the Milyads..." I say that you compelled this, you assessed it, the money was counted out to you. And I show that, with the same violence and wrong, when you were exacting the greatest sums of money, you spread through all parts of the province like some ruinous storm and pestilence.

\ciceroSection{97} So Marcus Scaurus, who prosecuted Gnaeus Dolabella, kept this man in his power and command. The young man, when he had learned of many thefts and disgraces of his in the inquiry, acted shrewdly and cleverly. He showed Verres a very great volume of his deeds, took from the man what he wished against Dolabella, produced him as witness. Verres said what he thought the prosecutor wished. Of which kind, of witnesses who have stolen with him, if I had wished to use them, I would have had a great supply. Who, that they might free themselves from the danger of lawsuits and the connection of charges, promised to descend to whatever I wished.

\ciceroSection{98} I refused the will of all of them. There was no place in my camp for any betrayer, nor even for a deserter. Perhaps those are to be considered the better prosecutors who have done all this. So be it; but I wish my role to be praised most as defender, not prosecutor. He does not dare to render his accounts at the treasury before Dolabella was condemned. He gets from the senate that the day be deferred, because he says his books were sealed by Dolabella's prosecutors --- as if he did not have the power to copy them. He is the only one who never renders his accounts at the treasury. You have heard the quaestorian account, rendered in three little lines; the legateship's, only after the man had been condemned and exiled who could have refuted it. Now finally the praetorship's, which by senatorial decree he ought to have rendered at once, he has not rendered up to this time.

\ciceroSection{99} He said in the senate that he was waiting for his quaestor --- as if there were not, just as a quaestor can render his account without a praetor (as you, Hortensius, as all do), in the same manner a praetor without a quaestor. He said the same Dolabella had obtained this. The omen pleased the senators rather than the cause. They approved. But the quaestors too came long ago. Why did you not render? From the dregs of those accounts of your legateship and deputy-quaestorship come those entries which Dolabella was necessarily required to make good. From the lawsuits assessed against Dolabella, ex-praetor and ex-propraetor:

\ciceroSection{100} Where Dolabella entered as received from Verres less than Verres entered as paid to him --- 535,000 sesterces; and where Dolabella entered Verres as having received more than the same had in his books --- 232,000 sesterces; and where he entered him as having received more grain --- one million eight hundred thousand sesterces, which you, most upright man, had otherwise in your books. From this came those extraordinary moneys, which without any guide we can investigate at least in some part. From this is the account with Quintus and Gnaeus Postumus Curtius under many headings, none of which he has in his books. From this come the four hundred thousand sesterces counted out to Publius Tadius at Athens, which I shall make plain by witnesses. From this came the praetorship most openly bought --- unless that too is doubtful, how this man was made praetor.

\ciceroSection{101} A man, naturally, of approved industry or work, of distinguished reputation for self-restraint, or finally (which is the lightest of qualities) of constancy in attendance --- who before his quaestorship had lived with prostitutes and pimps; who had conducted his quaestorship as you have learned; who, after that wicked quaestorship at Rome, had scarcely stayed three days; who in his absence had not lain in oblivion but had been in everyone's continual mention of every disgrace --- he, suddenly, as he came to Rome, was made praetor on charm. More money, indeed, was given that he should not be prosecuted. To whom it was given, has nothing to do with me, nothing with the case. That it was given was easily agreed at the time among all in the recent affair.

\ciceroSection{102} Most foolish and senseless man, when you were making up your books and wishing to escape the charge of extraordinary money, did you suppose you would slip from all suspicion if you did not enter as paid to those whom you trusted with money, and brought no entry into your books, while the Curtii entered it under so many headings as received from you? Of what use was it to you not to have entered it as paid out to them? Did you think you would plead your case from your books alone?

\ciceroSection{103} But let us now come to that splendid praetorship, and to those charges which are better known to those present than to us who have come prepared and rehearsed for speaking. About which I do not doubt that I cannot avoid and escape the offence of negligence. For many will say so: "About that he said nothing, in which I was concerned. He did not touch that wrong which was done to me or to my friend, in which I was concerned." Towards all of you who know his wrongs --- that is, the entire Roman people --- I want it strongly understood that not by my negligence shall it come about that I leave many things out, but because I want others to be reserved whole for the witnesses, and many I judge to be passed over for reason of brevity and time. I will confess this also unwillingly: that I, when this man has not let any moment of time pass empty of offence, have not been able to know all that has been committed by him. Wherefore listen to me concerning the charges of his praetorship in such a way that, from each kind --- both of giving judgement and of demanding the patching of public buildings --- you require those things which are most worthy of a defendant in whom nothing small or moderate ought to be charged.

\ciceroSection{104} For when he was made praetor, who had risen with auspices from Chelidon's bed, he received by lot the city province more by his own and Chelidon's will than by the Roman people's. From the start, learn what kind he was in establishing the edict. Publius Annius Asellus died under the praetor Gaius Sacerdos. Since he had an only daughter and was not in the census, what nature urged him and no law forbade --- he made his daughter heir of his goods. The heir was the daughter. Everything was in the ward's favour: the equity of the law, the father's will, the edicts of praetors, the custom of the law as it was at the time when Asellus died.

\ciceroSection{105} This praetor-elect --- whether warned, or tested, or, with the keenness he himself has in such matters, came to this dishonesty without any guide or informer, I do not know. Just see the audacity and madness of the man --- he addresses the heir Lucius Annius, who had been instituted heir after the daughter (for I cannot be persuaded he was addressed by the man first). He says he can give him the inheritance by edict; he teaches the man what can be done. To the one the matter seemed good; to the other, saleable. He, although he is of singular audacity, sent feelers to the ward's mother. He preferred to take the money, lest he edict anything new, rather than interpose this so dishonest and inhuman edict.

\ciceroSection{106} The guardians did not see how to give the praetor money on the ward's account --- a great sum especially --- nor how to bring it into account, nor how to give it without their own danger. At the same time, they did not think this man would be so dishonest. Often appealed to, they utterly refused. He, at the discretion of the man to whom he was giving away the inheritance taken from the children, drew up an edict of how equitable a kind --- learn, please. "Since I understand the lex Voconia..." Who would have ever believed Verres would be the adversary of women? Or has he done anything against women for this reason: that the whole edict might not seem written at Chelidon's discretion? He says he wishes to oppose men's greed. Who, not in our times only, but even in our ancestors', has been more removed from greed? Say the rest, please --- for the man's gravity and knowledge of praetorian law delights me, and his authority. Read: "Whoever, from the censors Aulus Postumius and Quintus Fulvius, or after that, was rated... has done or shall have done."

\ciceroSection{107} "Has done or shall have done"? Who ever issued an edict in this manner? Who ever set out by edict the fraud or danger of a matter which could neither be reproved after the edict nor foreseen before the edict? Lawfully, by laws, by the authority of all who were consulted, Publius Annius had made his will --- not dishonest, not unfilial, not inhuman. But if he had so done, yet after his death no new law concerning his testament ought to have been established. The lex Voconia, of course, delighted you. You would have imitated that very Gaius Voconius, who by his law took the inheritance from no one --- neither maiden nor woman; he sanctioned for the future, that whoever should be in the census after those censors should not make a maiden or a woman his heir.

\ciceroSection{108} In the lex Voconia there is no "has done or shall have done"; nor is past time reproved in any law except for that thing which is so wicked and unspeakable in itself that, even if there were no law, it must greatly be avoided. And in these very things we see many things sanctioned by laws in such a way that what was done before is not called into court: the Cornelian on wills, on coin, several others, in which not some new right is established for the people, but it is sanctioned that, what has always been an evil deed, the inquiry on it shall belong to the people from a certain time.

\ciceroSection{109} If anyone establishes anything new in civil law, will he not allow that all that was done before be ratified? Bring me the leges Atinia, Furia, Fusia, the lex Voconia itself (as I said), and all the others on civil law: you will find this established in all of them --- the law for which the people may use it after that law. Those who attribute most to the edict say that the praetor's edict is a law for one year; you embrace more by edict than by law. If the Kalends of January bring the end of the praetor's edict, why does not the beginning of the edict also rise from the Kalends of January? Or shall no one be able by edict to advance into the year in which there will be another praetor; will he go back into that in which there was another praetor?

\ciceroSection{110} And if you had set this in your edict for a precedent, not for the case of one man, you would have composed it more cautiously. You write: "Whoever has made or shall have made an heir." What if he should bequeath more than comes to the heir or heirs? Which is permitted by the lex Voconia to him who is not in the census. Why do you not provide for this, when it is in the same kind? Because you embrace in your words not the matter but the case of the man, so that it readily appears you were moved by price, not by law. And if you had edicted this for the future --- although it would still be wicked, it would still be dishonest. But you could be reproved, you could not come into peril; for no one would commit it. As things are, the edict is of such a kind that any man understands it has not been written for the people, but for the second heirs of Publius Annius.

\ciceroSection{111} So when this article had been adorned by you with so many words and a hireling preface, was any praetor afterwards found who edicted the same thing? Not only did no one edict it, but no one feared lest someone would edict it. For after you as praetor, many were in the same situations. Among these recently Annaea, by the advice of many kinsmen, a moneyed woman, since she had not been rated, made her daughter heir by will. So great was men's judgement of his singular dishonesty, that what Gaius Verres of his own accord had instituted, no one feared anyone would be found who would wish to follow this institution. For you alone were found to whom it was not enough to correct the wills of the living, unless you also rescinded those of the dead.

\ciceroSection{112} You yourself took this from the Sicilian edict; you wished to decree from your urban edict if any matter should suddenly arise. Whatever defence you afterwards left yourself, in that you most offended, when by your provincial edict you yourself rejected your own authority. Nor do I doubt that, as this matter seems bitter and unworthy to me, who have my daughter most at heart, so it does to each of you who is moved with the same feeling and indulgence for daughters. For what has nature wished to be more pleasant, more dear to us? What is more worthy in which all our diligence and indulgence be spent?

\ciceroSection{113} Most dishonest of men, why have you done so great a wrong to dead Publius Annius? Why have you burned this grief into his ashes and bones, that you should snatch from his children their ancestral goods --- handed down by their father's will, by right, by laws --- and give them to whomever it suited you? Those with whom we share our goods alive, will the praetor, when we are dead, be able to take goods and fortunes from them? "Neither," he says, "shall I give the suit nor the possession." Will you, then, snatch the praetexta from the maiden, will you tear off the ornaments not only of fortune but even of free birth? We wonder that the Lampsacenes went to arms against this man; we wonder that he, leaving his province, fled secretly from Syracuse. If we had grieved another's plight as our own wrong, no trace of him would have been left in the Forum.

\ciceroSection{114} A father gives to his daughter; you forbid. The laws permit; you nevertheless interpose! Of his own goods he gives, not departing from law. What have you to censure? Nothing, I take it. But I grant it: forbid, if you can, if you have someone to listen to you, if anyone can be obedient to you. Will you take away will from the dead, goods from the living, right from all? Would the Roman people not have avenged this with its hand, had it not reserved you for this moment and this trial? After praetorian law was established, we have always used this rule: if the will was not produced, then according as anyone ought most to be heir, if he had died intestate, possession was given to him. Why this is most equitable is easy to say, but in a matter so customary it is enough to show that all before laid down the law thus, and that this is an old and traditional edict.

\ciceroSection{115} Learn another new edict of the man's in an old matter, and at the same time, while there is somewhere to learn civil law, hand over the youth to his discipline. Marvellous is the man's genius, marvellous his prudence. A certain Minucius died before this man's praetorship; his will was none; the inheritance came by law to the Minucian gens. If he had had an edict such as all before him and after him have had, possession would have been given to the Minucian gens. If anyone thought himself heir by a will not then in existence, he would either sue for the inheritance under the law, or, having received security for the suit, would make a sponsio and so contend for the inheritance. By this rule, I think, our ancestors and we have always proceeded. See how this man corrected it.

\ciceroSection{116} He composes the edict in such words that any man can understand it has been written for the cause of one man --- only that he does not name the man. He sets out the whole matter; the law, the custom, the equity, the edicts of all he disregards. "From the urban edict: if there is dispute about an inheritance --- if the possessor will not make the sponsio." Now what is for the praetor, which is the possessor? Ought not this to be asked, who ought to be the possessor? Therefore, because he is in possession, you do not move him from possession; if he were not in possession, would you not give? You write nothing else; you embrace nothing else by your edict but that case for which you had taken money. Now this is ridiculous:

\ciceroSection{117} "If there is dispute about an inheritance and tablets of the will sealed with no fewer signs than is required by law shall be brought before me, I shall give possession according to the tablets of the will." This is traditional. The next ought to follow: "If the tablets of the will shall not be produced." What does he say? That he will give to the man who says he is heir. What difference, then, whether they are produced or not? If he produces them, with one seal short of what the law requires, you will not give possession; if he produces no tablets at all, will you give? What now shall I say? That no one ever afterwards edicted thus? It would be very wonderful that no one was found who wished to be called like this man. He himself does not have this in his Sicilian edict; for he had already exacted the price. Likewise, in that edict of which I spoke before --- about giving possession of inheritances in Sicily --- he edicted the same thing as all at Rome but he. From the Sicilian edict: "If there is dispute about an inheritance..."

\ciceroSection{118} And, by the immortal gods! What can be said about this? For I again ask of you: just as a moment ago in that Annian article on women's inheritances, now in this one on the possessions of inheritances, why did you not transfer those articles into the provincial edict? Did you think those who live in the province more worthy than us, that they should use a fairer law? Or is one thing equitable at Rome, another in Sicily? For this cannot be said in this place: that many things must be edicted otherwise in the provinces. Not concerning the possessions of inheritances, not concerning women's inheritances. For in both kinds I see not only others, but you yourself edicted in just as many words as it is customary to edict at Rome. What at Rome with great infamy you had edicted, your price taken --- those you alone, lest you be ill-spoken-of in the province for nothing, took out of the Sicilian edict.

\ciceroSection{119} And although he composed the entire edict, while he was praetor-elect, at the discretion of those who would deal in law from him for their own advantage --- yet in his magistracy he decreed against this very edict of his own, without any conscience. So Lucius Piso filled many books of those matters in which he had interceded, because Verres had decreed otherwise than he had edicted. Which I do not think you have forgotten: what a multitude, what an order used to gather at Piso's tribunal under this praetor. If Verres had not had Piso as colleague, he would have been covered with stones in the Forum. But Verres's wrongs seemed lighter, because in Piso's equity and prudence there was a most ready refuge, which men used without labour, without trouble, without expense, even without an advocate.

\ciceroSection{120} For, please call to mind, gentlemen, what was the man's lust in giving judgement, what the variety of decrees, what the brokering, how empty were the houses of all who are accustomed to be consulted on civil law, how full and packed Chelidon's. From which woman, when one had come to him and a word had been whispered in his ear, sometimes he called back those between whom he had already decreed and changed the decree; sometimes between others he decreed the contrary, without any conscience, of what he had decreed for those just before.

\ciceroSection{121} Hence those men who, even from grief, were found ridiculous: of whom some, what you have often heard, said it was no wonder that "Verrine law" was so worthless. Others were even less witty, but, because they were sore, seemed witty: when they cursed Sacerdos for having left so worthless a "Verres" (boar). I would not mention these things --- for they are neither very pretty in saying, nor worthy of such severity --- did I not wish you to remember: this man's worthlessness and inequity were then in everybody's mouth and in common proverbs.

\ciceroSection{122} As for the Roman commons, shall I mention his arrogance first, or his cruelty? Without doubt, his cruelty is graver and more atrocious. Do you think these have forgotten how this man was accustomed to cut down the Roman commons with rods? Which matter even a tribune of the plebs argued in public meeting, when he produced before the sight of the Roman people the man whom this man had cut down with rods. Of which matter I shall give you opportunity, in due time, to be reminded.

\ciceroSection{123} As for his arrogance, who does not know? How he despised, scorned, never reckoned as free, the slenderest of men? Publius Trebonius made many good and honourable men his heirs; among these he made his own freedman one. He had had a brother Aulus Trebonius proscribed. To him, when he wished to provide, he wrote that the heirs should swear they would see to it that from each one's portion not less than a half should come to that proscribed Aulus Trebonius. The freedman swore. The other heirs went to Verres and showed they ought not to swear that they would do what was against the lex Cornelia, which forbade a proscribed man to be helped. They obtained that they should not swear. He gives them possession. This I do not censure; for it would have been unfair that anything from his brother's goods should be given to a proscribed and destitute man. The freedman, unless he had sworn by his patron's will, would have thought he was committing a crime.

\ciceroSection{124} So Verres said he would not give him the possession of the inheritance, lest he could help his proscribed patron, and at the same time as a punishment for having obeyed his other patron's will. You give possession to him who did not swear; I grant it. It is praetorian. You take from him who swore; by what example? He helps the proscribed; there is a law, there is a punishment. What does that have to do with the man who declares the law? Do you reproach him because he was helping a patron who was then in misfortunes, or because he was preserving the will of his other dead patron, from whom he had received the highest benefit? Which of these do you reproach? And this, indeed, the excellent man said from his bench: "Shall a freedman be heir to so wealthy a Roman knight!" O modest order, that he rose from there alive!

\ciceroSection{125} I could bring forward six hundred decrees in which, even though I were to say nothing, the very newness and inequity of the decrees declares that money intervened. But, that you may make a conjecture from one about the rest, hear what you learned in the previous hearing. There was Gaius Sulpicius Olympus. He died under the praetor Gaius Sacerdos, perhaps before Verres had begun to canvass for the praetorship. He made Marcus Octavius Ligus his heir. Ligus accepted the inheritance; he held it under Sacerdos as praetor without any controversy. After Verres entered office, by his edict --- which edict Sacerdos had not had --- the daughter of his patron Sulpicius claimed a sixth part. Lucius Gellius defended Ligus's case. He showed that his edict ought not to apply to those inheritances which had come before his praetorship; if this had then been the edict, perhaps Ligus would not have accepted the inheritance. The fair demand, the highest authority of the men, was overcome by price.

\ciceroSection{126} Ligus came to Rome. He did not doubt that, if he himself met Verres, he could move the man by the equity of his cause and his own authority. He came to his house. He laid out the matter; he showed how long ago his inheritance had come; what was easy for an able man in a most equitable case --- many things which could move anyone, he said. Finally he began to ask not to despise his authority and contemn his favour so far as to afflict him with so great a wrong. The man began to accuse Ligus, because in a windfall and inheritance matter he was so diligent, so attentive. He said he too should take account; he needed many things, his dogs whom he kept around him needed many. I cannot mention these things more plainly than you heard Ligus himself say in evidence.

\ciceroSection{127} What of it, Verres? Will not even these witnesses be believed, or do these things not pertain to the case? Shall we not believe Marcus Octavius, not Lucius Ligus? Who will believe us, to whom shall we believe? What is there that can be made plain by witnesses, if this is not? Or is what they say slight? Nothing is slighter than for the urban praetor in his magistracy to establish this rule of law: that for all those to whom an inheritance has come, the praetor must necessarily be co-heir. Do we doubt with what face this man was accustomed to address other men, lower in place, authority, order; with what face country-men from the municipalities; with what face, finally, freedmen --- whom he never thought free --- when he did not hesitate to demand money for giving judgement from Marcus Octavius Ligus, a man most adorned in place, order, name, virtue, ability, resources? As to how he conducted himself in the patching of public buildings, what shall I say? Those who felt it have said; there are others to say it; matters known and manifest have been brought forward and will be.

\ciceroSection{128} Gnaeus Fannius, a Roman knight, full brother of Quintus Titinius, your juror, said that he gave you money. Read: "Testimony of Gnaeus Fannius." Do not believe Gnaeus Fannius when he says it; do not, I say, you, Quintus Titinius, believe Gnaeus Fannius your brother; for he says an incredible thing. He charges Gaius Verres with greed and audacity --- vices which seem to belong to anyone rather than to this man. Quintus Tadius said, a man most close to this man's father, not foreign to his mother's family and name. He produced his books, by which he showed he had given money. Read: "Entries of Quintus Tadius." Read: "Testimony of Quintus Tadius." Shall we not believe even Tadius's books, even his testimony? What then in trials shall we follow? What else is to grant all offences and crimes to all but this --- not to believe the testimonies of honourable men and the books of good men?

\ciceroSection{129} For why should I speak of the daily talk and complaint of the Roman people, of this man's most shameless theft (or rather of his novel and singular brigandage)? That he dared, in the temple of Castor --- a most frequented and most distinguished monument, a temple set in the eyes and daily sight of the Roman people, where the senate is often called together, where most numerous gatherings are held daily on the greatest matters --- in that place, in everyone's talk, to leave an eternal monument of his audacity.

\ciceroSection{130} The temple of Castor, gentlemen, Publius Junius had to keep up under Lucius Sulla and Quintus Metellus the consuls. He died; he left a small ward as his son. When Lucius Octavius and Gaius Aurelius the consuls had let out the sacred buildings, and they had not been able to require everything to be patched up, nor had those praetors to whom this business had been given --- Gaius Sacerdos and Marcus Caesius --- a senatorial decree was made: that those buildings concerning whose patching no judgement and decision had been made, the praetors Gaius Verres and Publius Caelius should examine and decide. Of which power being granted to him, he so abused that, as you have learned from Gnaeus Fannius and Quintus Tadius --- but although he had most openly and shamelessly preyed on everything, he wished to leave this most clear sign of his brigandage, of which we could not only at some time hear but daily see.

\ciceroSection{131} He inquired who ought to deliver up the temple of Castor in good order. He knew that Junius himself had died; he wished to know to whom that matter pertained. He hears that the son is a ward. The man who had always so openly said that wards male and female were the surest plunder for praetors, said that a wished-for business had been brought to his lap. A monument of that size, of that workmanship --- although it was patched and whole --- yet he supposed he could find something in which he could plot and plunder.

\ciceroSection{132} The temple of Castor ought to have been delivered to Lucius Habonius. He happened to be the guardian of the ward Junius by the father's will. Already without any deduction it had been agreed with him in what manner it should be delivered. Verres summons Habonius to him. He asks if there is anything which has not been delivered by the ward, which ought to be required. When the man said --- as was the truth --- that the delivery was easy for the ward, that all the statues and gifts were in evidence, that the temple itself was whole in all its work --- it began to seem unworthy to him to depart from so great a temple and so great a work without rich plunder, especially from a ward.

\ciceroSection{133} He himself came to the temple of Castor. He inspected the temple. He sees the roof everywhere most beautifully panelled, and besides everything else new and whole. He turns himself, asks what to do. One of those dogs (whom he had told Ligus he kept many of around him) says: "Verres, you have nothing here to plot, unless perhaps you wish to require the columns to stand by the plumb-line." The man, ignorant of all things, asks what is "by the plumb-line." They tell him that scarcely any column can be by the plumb-line. "By Hercules," he says, "let us proceed thus: let the columns be required to stand by the plumb-line."

\ciceroSection{134} Habonius, who knew the law --- in which law only the number of the columns is given, of plumb-line no mention is made --- and who did not think it expedient to so accept, lest in the same way he must deliver, says it is not owed him, says it ought not to be required. Verres orders Habonius to be quiet, and at the same time shows him some hope of partnership. He easily restrains the modest and least obstinate man, confirms that he will so require the columns.

\ciceroSection{135} A new and unforeseen calamity of the ward is at once reported to Gaius Mustius, the ward's stepfather, who has lately died, to Marcus Junius the uncle, to Publius Titius the guardian, a most temperate man. They bring the matter to a leading man endowed with the highest duty and virtue, Marcus Marcellus, who was the ward's guardian. Marcus Marcellus comes to Verres. He asks of him, by his good faith and diligence, in many words, not to attempt by the highest wrong to overturn the ward Junius from his ancestral fortunes. Verres, who in hope and conception had already devoured that plunder, was moved neither by the equity of his speech nor by Marcus Marcellus's authority. So he replied that he would require it as he had shown.

\ciceroSection{136} When they saw all approaches to him difficult, all entries arduous and rather closed off --- with whom neither right, nor equity, nor pity, nor a kinsman's speech, nor a friend's wish, nor anyone's authority weighed against price, nor favour --- they decide it best to do what would have come into anyone's mind: to seek help from Chelidon, who under this praetor not only presided over the Roman people in civil law and the controversies of all private men, but even ruled in these matters of patching.

\ciceroSection{137} To Chelidon comes Gaius Mustius, a Roman knight, a publican, a most honourable man. There comes Marcus Junius, the uncle of the boy, a most temperate and chaste man. There comes a man of the highest modesty, the highest dutifulness, most distinguished in his order, Publius Titius the guardian. O, to many bitter, miserable, and unworthy praetorship of yours! To leave aside other things, with what shame, do you think, with what grief, did such men come to a courtesan's house? --- they who would never on any condition have undergone such disgrace, had not the reckoning of duty and bond compelled them. They came, as I say, to Chelidon. The house was full. New rights, new decrees, new judgements were being sought. "Let her give me possession; let her not take it from me; let her not give a trial against me; let her award the goods to me." Some were counting out money; from others bonds were being sealed. The house was packed not with a courtesan's gathering but with a praetor's crowd.

\ciceroSection{138} As soon as opportunity was first given, those whom I have named approach. Gaius Mustius speaks, lays out the matter, asks help, promises money. She replied as a courtesan, not inhumanly: she said she would gladly do it, and would speak with this man diligently. She told them to come back. They departed. The next day they returned. She says the man cannot be moved; that he says a very great sum can be made of that matter. I fear lest someone of the people, who was not present at the previous hearing, may suppose I am inventing these things, because they are incredible by reason of their notable disgrace. You, gentlemen, learned them earlier.

\ciceroSection{139} Publius Titius, the guardian of the ward Junius, said it under oath; Marcus Junius the guardian and uncle said it. Mustius would have said it, if he were alive; but the matter being recent, evidence about Mustius was heard. Lucius Domitius said it; who, although he knew that I had heard from the living Mustius (because I made very much use of him --- for in a trial in which he had at stake almost all his fortunes, with me as sole defender, Gaius Mustius won), although he knew, I say, that I knew Mustius was accustomed to bring all matters to him, yet he kept silent about Chelidon as long as he could, diverted his answer elsewhere. Such was the modesty in that most distinguished young man, the leader of the youth, that for a long time, while pressed by me, he would answer anything rather than name Chelidon. First he said that Verres's friends had been sent to him; then at last, compelled, he named Chelidon.

\ciceroSection{140} Are you not ashamed, Verres, to have conducted your praetorship at the discretion of the woman whom Lucius Domitius scarcely thought it honourable for himself to be named with? Rejected by Chelidon, they take a necessary plan: that they themselves take up the business. With Habonius the guardian --- the matter being scarcely of forty thousand sesterces --- they settle for two hundred thousand. Habonius brings the matter to Verres. To himself it seems a sufficiently great and shameless sum. Verres, who had reckoned on much more, takes Habonius up sharply with words, says he cannot be content with that compromise. To make this short: he confirms he will let the contract.

\ciceroSection{141} The guardians know nothing of this. What had been done with Habonius, they consider most certain. They fear no greater calamity to the ward. But Verres does not procrastinate. He begins to let the contract --- without proclamation or public notice of the day, at the most alien moment, the very Roman games being held, the Forum decked. So Habonius reports the agreement back to the guardians. Yet they hurry to the moment. Junius the uncle raises his finger to bid. Verres's colour was changed, his look, his speech, finally his mind failed. He began to think what to do; if the work were taken back by the ward, if it went away from the contractor he had himself set up, no plunder for himself. So he hits upon --- what? Nothing ingenious, nothing on which anyone could say "dishonest, but cleverly so." Expect from him nothing wily, nothing veteran. Everything will be found open, everything plain --- shamelessness, madness, audacity.

\ciceroSection{142} "If the ward takes back the work, the plunder is snatched from my hands. What then is the remedy? What? --- that the ward should not be allowed to take it." Where is that custom in the selling of pledges and lands of all consuls, censors, praetors, finally quaestors --- that he should be in the best position whose property is at stake, whose risk it is? He shuts out the only one to whom (I would almost say) the only power should have been given. For what? Does anyone aspire to my money against my will, does anyone approach? A work is being let to be repaired out of my money; I say I will repair. The approval is to be yours, who let. The people are secured by sureties and lands. And, if you do not think there is security, will you, the praetor, send into my goods whom you please, and not let me approach to defend my fortunes?

\ciceroSection{143} It is worth while to learn the law itself; you will say the same man wrote it as the edict on inheritance. Read: "Law for the doing of the work. The work which the ward Junius..." Speak --- speak, please, more loudly. "Gaius Verres, urban praetor, has added." The censorial laws are being corrected! For what? I see in many old laws "Gnaeus Domitius and Lucius Metellus the censors have added," "Lucius Cassius and Gnaeus Servilius the censors have added." Gaius Verres wishes something of the kind. Speak: what has he added? Read: "Whoever, from the censors Lucius Marcius and Marcus Perperna --- shall not admit a partner, nor give a share, nor take it back." Why so? Lest the work should be done badly? But your approval was on it. Lest there should not be sufficient resources? But the people were and would be more than secured by sureties and lands, if you wished.

\ciceroSection{144} Here, if the matter itself, if the unworthiness of your wrong did not move you, if the calamity of the ward, if the tears of his kinsmen, if the danger of Decimus Brutus (whose lands were under pledge), if the authority of the guardian Marcus Marcellus had no weight with you --- did you not even notice that this offence of yours would be of such a kind that you could neither deny it (for you put the law in your records) nor confess with any defence? The work is awarded for 560,000 sesterces, when the guardians cried out that they would do that work for 40,000, even at the discretion of that most unfair man. For what kind of work was it?

\ciceroSection{145} What you saw: all those columns which you see whitewashed --- a machine being applied --- without any expense, were thrown down and replaced with the same stones. This you let out for 560,000 sesterces. And among those columns, I say there are some which were not even moved by your contractor. I say there is one from which only the old stucco was knocked off and new put on. If I had thought that columns could be whitewashed for so great a sum, certainly I should never have canvassed for the aedileship.

\ciceroSection{146} But, that the matter might seem to be conducted and not snatched from the ward: "If anything for the sake of the work has been pulled down, let him repair." What was there for him to pull down, when each stone was put back in its own place? "Let the contractor give security for damage to him who received from the old contractor." It is mocking, when he orders Habonius to give security to himself. "Money in cash shall be paid." From whose goods? Of the man who, where you let the work for 560,000, cried out that he would do it for 40,000. From whose goods? Of the ward, whose age and solitude, even if there had been no guardians, the praetor ought to have defended. With the guardians defending, you took not only his ancestral fortunes but even the goods of the guardians. "This work let him do good in his own materials."

\ciceroSection{147} What is "in his own"? Some stone had to be cut and brought by his own machine; for to that place no stone, no timber was brought. Such was the work in that letting as a few day-labourers' workmen earned wages, and the manufacturing-cost of the machine. Do you suppose it less work to make one column entirely new with no recycled stone, or to set up four of those columns? No one doubts that to make a new one is much greater. I will show that in private houses, with long and difficult transport, columns of equal size were let at 200,000 sesterces.

\ciceroSection{148} But it is silly to argue in many words about so plain a shamelessness of his, especially since he in the whole law openly contemned everyone's talk and estimation, who even at the end added: "Let him have the recycled materials for himself" --- as if anything recycled was being taken from that work, and the whole work was not made up of recycled materials. But if the ward could not take it back, it was not necessary that the matter should come to him. Some other from the people could have approached the business. All were excluded no less openly than the ward. He fixed the day for doing the work as the Kalends of December; he lets out about the Ides of September. By the narrow time all were excluded.

\ciceroSection{149} What of it, then? How does Habonius come up to that day? No one is troublesome to Habonius, neither on the Kalends of December, nor on the Nones, nor on the Ides. Finally, somewhat before the work was done, Verres set out for the province. After he was made defendant, at first he denied that he could enter the work as accepted. When Habonius pressed him, he transferred the cause to me, because I had sealed his book. Habonius asks of me, sends friends. He easily obtains. Verres did not know what to do. If he had not entered it as accepted, he thought he would have some defence; but he understood that Habonius would lay open the whole matter --- although what could be more open than it now is? --- so that he might plead with one less witness, he entered the work as accepted from Habonius four years after he had set the day for the work.

\ciceroSection{150} On these terms, if any contractor from the people had approached, he would not have used the chance. When he had excluded the rest of the contractors by the day, then they did not wish to come into the discretion and power of one who would think the plunder snatched from him. Now, that we may not have to argue --- where that money went, the man himself makes plain. First, when Decimus Brutus pressed him more vehemently --- who counted out 560,000 of his own money --- because Verres could no longer bear it, with the work let, with sureties accepted, of the 560,000 he remitted 110,000 to Decimus Brutus. This, if it had been someone else's matter, he certainly could not have done. Then money was counted out to Cornificius, whom he cannot deny was his secretary. Finally Habonius's own books cry out that that plunder was Verres's. Read: "Entries of Habonius."

\ciceroSection{151} Here also in the previous hearing Quintus Hortensius complained that the ward Junius had come into your sight wearing the praetexta and stood there while his uncle was giving evidence; and he cried that I was acting popularly and stirring up odium because I produced the boy. What was there, Hortensius, that was popular in that boy? What odium-stirring? I had produced the son, I take it, of a Gracchus or a Saturninus or some such man, that by the very name and memory of the father I might stir the minds of the inexperienced multitude! He was the son of Publius Junius, a man of the Roman commons, whom his dying father commended both to the guardians and the kindred, both to the laws, to the equity of the magistrates, and to your trials.

\ciceroSection{152} This boy, despoiled of his ancestral goods and all his fortunes by Verres's wicked and unspeakable brigandage, came into court, if for nothing else, that at least he might see the man through whose action he had been many years in mourning slightly less worn than himself. So to you, Hortensius, not his age, but his case, not his clothing, but his fortune seemed "popular." Nor did it move you so much that he came in the toga praetexta, as that he came without the bulla. For the dress moved no one which custom and the right of free birth gave him. But that ornament of childhood which his father had given him, the sign and badge of fortune, that this had been torn from him by this brigand --- this men then took grievously and bitterly.

\ciceroSection{153} Nor were those tears popular more than ours, more than yours, Quintus Hortensius, more than these jurors who are about to give their vote --- because the case is common, the danger common. By common protection such dishonesty must be put out, like some fire. For we have small children. It is uncertain how long the life of each of us shall be. We must, while alive, take care and provide that their loneliness and childhood shall be guarded by the strongest protection. For who can defend our children's childhood against the dishonesty of magistrates? The mother, I take it. Of course, the mother was a great protection to the ward Annia, that leading lady; the less she invoked gods and men, this man took the ancestral fortunes from the infant girl. Will the guardians defend? It is most easy, of course, before a praetor of this kind, by whom in the case of the ward Junius the speech and will and authority of the guardian Marcus Marcellus was rejected!

\ciceroSection{154} Do we still ask what this man did in farthest Phrygia, what in the outermost parts of Pamphylia, what kind he was as a brigand himself in the war of the brigands --- a man who in the Forum of the Roman people is found a wicked pirate? Do we doubt what he plotted in the booty of enemies, who from Lucius Metellus's spoils made for himself such great spoils, who let out the whitewashing of four columns for more money than that man had let out the building of all of them? Shall we await what the witnesses from Sicily say? Who has ever looked upon that temple but is a witness of your greed, your wrong, your audacity? Who comes from the statue of Vortumnus to the Circus Maximus but is reminded at every step of your greed? --- which path of the holy chariots and that procession you so let out that you yourself do not dare to go that way. Will anyone think that you, when separated from Italy by the strait, refrained from the allies, who wished the temple of Castor to be a witness of your thefts? --- which the Roman people see daily, and the jurors will see even when they cast their vote on you.

\ciceroSection{155} And he also exercised a public trial in his praetorship; for not even this is to be passed over. A fine was sought before this praetor by Quintus Opimius, who was brought to trial nominally because, when he was tribune of the plebs, he had vetoed against the lex Cornelia; in fact because in his tribunate he had spoken against the will of some noble man. Of which trial if I should wish to say everything, many would have to be appealed to and wounded --- which is not necessary for me. I will say only this: a few men --- to call them lightly --- arrogant, with this man's help, in jest and play overturned Quintus Opimius from all his fortunes.

\ciceroSection{156} He even complains that we transacted the first hearing of his trial in only nine days, when before him in three hours Quintus Opimius, a senator of the Roman people, lost his goods, his fortunes, all his ornaments? On account of whose unworthy trial it was discussed in the senate again and again that this whole class of fines and trials of this kind should be abolished. Then in the selling of Quintus Opimius's goods, what plunderings, how openly, how dishonestly he did --- it is long to tell. This I say: unless I make this plain to you by the books of the most honourable men, suppose the whole thing made up by me for the moment.

\ciceroSection{157} A man who, from a senator of the Roman people's calamity --- when he had presided as praetor over his trial --- attempted to bring spoils home and tear off the spoils-of-war: shall he be able to deprecate any calamity of his own? For of that subsortition of jurors of Junius I say nothing. For what? Should I dare to speak against the books which you have produced? It is hard. For not only your authority and that of the jurors, but even the gold ring of your secretary deters me. I will not say what is hard to prove. I will say what I will show: that you have been heard by many leading men, when you said that you ought to be forgiven for having produced a false book. For the resentment with which Gaius Junius was burnt up --- by that, unless you had foreseen it, you yourself would then have had to perish.

\ciceroSection{158} In this manner he has learned to provide for himself and his safety: by entering in his books, both private and public, what had not been done; by removing what had been; and always taking out something, changing, altering. He goes so far that he cannot even find a defence of his own crimes without other crimes. A subsortition of his own jurors, of this kind, the most senseless man thought he could effect through his comrade Quintus Curtius, the juror in his own trial --- whom, unless I had resisted by force of the people and the shouting and clamour of men, would have substituted from your decury (a copy of which I should have had as fully as possible) those whom Verres had nodded to, into his council, without any cause.




