% In C. Verrem Actio Secunda IV (in-verrem-2-4) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.
%
% Book IV of the Actio Secunda --- de signis --- treats Verres's
% plundering of the art and sacred objects of Sicily: the Heius
% collection at Messana, the candelabrum of Antiochus, the temples
% of Ceres at Henna and Catina, the Diana of Segesta, the Syracusan
% triumphal sculptures.

\ciceroSection{1} I come now to that which this man calls (as he himself says) his pursuit, his friends call his disease and madness, the Sicilians call brigandage. By what name to call it I do not know. I shall lay the matter before you; you weigh it by its own weight, not by that of a name. Learn first the kind of thing itself, gentlemen; then perhaps you will not greatly inquire by what name you should call it. I say in all Sicily --- a province so wealthy, so old, of so many towns, of so many resourceful households --- that there was no silver vase, no Corinthian or Delian work, no gem or pearl, anything made of gold or ivory, no bronze or marble or ivory statue, no painting either on a panel or in cloth, that he did not seek out, inspect, and (where it pleased him) carry off.

\ciceroSection{2} I seem to say a great thing: mark also how I say it. For not for the sake of swelling a word or a charge do I take in all things: when I say that this man left nothing of that kind in the whole province, know that I speak Latin, not as a prosecutor. To put it more plainly: nothing in any man's house, not even a guest-friend's; nothing in public places, not even in shrines; nothing among Sicilians, nothing among Roman citizens; finally nothing whatsoever, that came before his eyes and his mind, neither private nor public nor profane nor sacred, did this man leave in all Sicily.

\ciceroSection{3} Whence then should I rather begin than from that city which alone was in your love and in your delight? Or out of what number rather than from your very praisers? For it will be more easily seen what kind of man you were among those who hate you, who accuse you, who pursue you, when you are found among your own Mamertines to have been a plunderer in the most dishonest fashion. Gaius Heius is a Mamertine --- all who have come to Messana easily grant me this --- the most adorned man of all in that city in every kind of thing. His house is, you might say, the best at Messana, certainly the best known and to our men the most open and the most hospitable. That house, before this man's coming, was so adorned that it was an ornament also to the city; for Messana itself, although adorned by its situation, walls, and harbour, is plainly empty and bare of those things in which this man delights.

\ciceroSection{4} There was at Heius's house a shrine, with great dignity, in the building, of great age, handed down from his ancestors, in which there were four most beautiful statues of the highest art and the highest fame, which could delight not only this man as a man of talent and understanding, but anyone of us, whom he calls amateurs: one a marble Cupid by Praxiteles --- for indeed I have learned the names of artists too while I have inquired into this man. The same artist, I think, made that other Cupid of the same kind, which is at Thespiae, on whose account Thespiae is visited; for there is no other reason for visiting it. And that famous Lucius Mummius, when he was carrying off the Thespiads (which now are by the temple of Felicitas) and the rest of the profane statues from that town, did not touch this marble Cupid, because it was consecrated.

\ciceroSection{5} But to return to that shrine: there was the statue I am speaking of, of Cupid, in marble; on the other side a Hercules, finely made, in bronze. He was said to be by Myron, as I think, and indeed it is so. Likewise before these gods were little altars, which could signify to anyone the religious sanctity of the shrine. There were besides two bronze statues, not of the largest size but of singular grace, in maiden's bearing and dress, which with hands lifted up bore on their heads certain sacred objects laid up in the manner of Athenian maidens; \textit{Canephori} they themselves were called. But their artist --- who? whose? you remind me well --- they said was Polyclitus. Whoever of us came to Messana was wont to look at these. To all these were open daily for visiting; the house was an ornament not so much to its master as to the state.

\ciceroSection{6} Gaius Claudius, whose aedileship we know was most magnificent, used this Cupid only so long as he had the forum adorned for the immortal gods and the Roman people; and, since he was a guest-friend of the Heii, but the patron of the Mamertine people, as he made use of their kindness in the lending, so was he himself diligent in the giving back. Lately, gentlemen, noble men of this kind --- but why do I say "lately"? Indeed but now and a very little while ago we have seen --- who adorned the forum and the basilicas not with the spoils of provinces but with the ornaments of friends, with the conveniences of guest-friends not with the thefts of the guilty; who yet returned their statues and ornaments to each man, and did not, having taken them out of the cities of allies and friends for four days under the pretence of an aedileship, then carry them off home and to their villas.

\ciceroSection{7} All these statues that I have spoken of, gentlemen, Verres carried off from Heius's shrine. Not one, I say, of these did he leave; nor any other besides one ancient wooden one, the good Fortune, I think; that this man would not keep in his house. By the faith of gods and men! What is this? What is this case, what is this shamelessness? The statues I am speaking of, before they were carried off by you, no man came with imperium to Messana who did not visit them. So many praetors, so many consuls were in Sicily, both in peace and in war, so many men of every kind --- I am not speaking of the upright, the innocent, the scrupulous --- so many greedy, so many dishonest, so many bold, of whom no one seemed to himself so vehement, so powerful, so noble that he dared to ask for, to take, or to touch anything from that shrine. Will Verres carry off whatever shall be most beautiful anywhere? Shall it be lawful to no one to have anything besides? Shall this man's one house take in so many most well-off houses? Did none of his predecessors touch them, that this man might carry them off? Did Gaius Claudius Pulcher restore them, that Gaius Verres might carry them off? But that Cupid did not require a pimp's house and the school of harlots; he was easily contained in his ancestral shrine. He knew he had been left to Heius by his ancestors in the inheritance of the sacred things; he was not seeking the heir of a harlot.

\ciceroSection{8} But why do I press so vehemently? I shall be repelled with one word. "I bought it," he says. Immortal gods, a famous defence! We sent into the province, with imperium and axes, a buyer to buy up all statues, painted panels, all the silver, gold, ivory, gems, to leave nothing for any man! For this defence, I see, opens to him for everything --- "I bought it." First, even if I should grant you what you wish --- that you bought (since in this whole kind you are about to use this one defence) --- I ask, what kind of trials did you reckon there were at Rome, if you reckoned that anyone was about to grant you this: that you in your praetorship and command bought up so many things so precious, indeed all things which were of any value, out of a whole province?

\ciceroSection{9} See the diligence of our ancestors, who suspected nothing yet of this kind, but yet provided for those things which could happen in small matters. They did not suppose that any man, who had set out into a province with power or with a legateship, would be so mad as to buy silver --- for it was given out of public funds; or clothes --- for they were furnished by the laws. They thought of slaves, which we all use and which are not furnished by the people: they ordained that no one should buy except in the place of a deceased one. If anyone had died at Rome? No, but in the place itself; for they did not wish you to furnish your own home in the province, but to fill up the use of the province itself.

\ciceroSection{10} What was the cause why they so diligently kept us from buyings in the provinces? This, gentlemen: that they thought it was a snatching, not a buying, when the seller was not allowed to sell at his own discretion. They understood that in the provinces, if he who was there with imperium and power wished to buy whatever was at any man's house, and that should be allowed him, the result would be that whatever each wished, whether it was for sale or not, he would carry off at the price he wished. Someone will say: "Do not deal with Verres in this fashion; do not search out his deeds by the reckoning of old religious scruple. Grant that he bought without punishment, only that he bought in good reckoning, nothing on the strength of his power, nothing from an unwilling man, nothing through wrong." So I shall deal with him: if Heius sold what he had for sale at the value he set on it, I cease to ask why you bought it.

\ciceroSection{11} What then must we do? Are we to use proofs in such a matter? We must inquire, I think, whether this Heius had any debt, whether he held an auction; if he held one, whether such great difficulty in money matters held him, whether such great want, such great force pressed him, that he should despoil his own shrine, that he should sell his ancestral gods. But I see the man held no auction; that he sold nothing ever save his own harvests; that he was not only in no debt but had much money of his own and always had had. If all these things were the contrary of what I say, yet he would not have been about to sell these things, which had for so many years been in his family and in his ancestors' shrine. "What if he was persuaded by the size of the money?" It is not credible that the man, so well-off, so honourable, would set money before his religious obligation and the monuments of his ancestors.

\ciceroSection{12} "These things are so; but yet men sometimes are drawn even from their own institutions by the size of the money." Let us see how great that money was which could draw Heius --- a most well-off, least greedy man --- from humanity, from piety, from religion. So you ordered him, I think, to enter into his account-books: "All these statues by Praxiteles, Myron, Polyclitus sold to Verres for 6,500 sesterces." So he entered them. Read. \textsc{From the account-books}. It pleases me that these famous names of artists, which these men set on high, have so fallen by Verres's assessment. A Cupid by Praxiteles for 1,600 sesterces! Surely from this was born "I prefer to buy than to ask."

\ciceroSection{13} Some will say, "What? Do you assess these things at a great price?" I, indeed, by my own reckoning and use, do not assess them. Yet I think they ought to be regarded by you in this manner: at how much they would be valued in the judgement of those who are zealous for these things, at how much they are wont to be sold, at how much these very statues, if they came up for open and free sale, might be sold; finally, at how much Verres himself values them. For never, had he reckoned that Cupid at 400 denarii, would he have committed himself, on its account, to talk among men and to such great reproach.

\ciceroSection{14} Who of you, then, does not know at how much these things are valued? Have we not seen at auction a bronze statue, not the largest, sold for 40,000 sesterces? What? If I should wish to name men who bought either at no less, or even at more, can I not? For what measure there is in these things of greed, the same is of valuation; it is hard to set an end of price unless you have set one to the lust. I see therefore that Heius was led neither by will nor by any difficulty of the time nor by the size of the money to sell these statues; that you, by that pretence of a buying, by force, by fear, by command, by the fasces, snatched and carried them off from the man whom, with the rest of the allies, the Roman people had committed not only to your power but to your faith.

\ciceroSection{15} What more, gentlemen, can be desired by me in this charge than that Heius himself should say these very things? Nothing surely. But let us not desire what is hard. Heius is a Mamertine; the Mamertine state alone publicly by common counsel praises this man. To all the rest of the Sicilians he is hateful; by these alone is he loved. But of this embassy which has been sent to praise him the chief is Heius; for he is the first man of the city. So that, perhaps, while he serves the public mandate, he may keep silent about his private wrongs.

\ciceroSection{16} Knowing and considering this, gentlemen, I yet entrusted the matter to Heius. I produced him in the first hearing, and yet I did this with no peril. For what could Heius answer, were he a dishonest man, were he unlike himself? That those statues were in his own house, and not with Verres? How could he say anything of the kind? Were he a man most foul and most shamelessly lying, he would say this: that he had had them for sale, and had sold them at his own price. The man, most noble in his own city, who would most have wished you to think truly of his religious obligation and dignity, first said that he was praising this man publicly, because so it had been laid upon him. Next, that he had not had them for sale, nor on any condition (had either been allowed him to choose) could he ever have been led to sell those things which had been left and handed down by his ancestors in the shrine.

\ciceroSection{17} Why do you sit, Verres? What do you await? Why do you say that you are surrounded and oppressed by the Centuripine state, by the Catinaean, by the Halaesine, by the Tyndaritan, by the Hennensian, by the Agyrinensian, and the rest of the cities of Sicily? Your second fatherland (as you used to say) Messana surrounds you --- your own Messana, I say, the helper of your crimes, the witness of your lusts, the receiver of your plunder and thefts. For the most ample man of that city is here, an envoy sent from home for the cause of this trial, the chief of your praise, who praises you publicly --- for so it has been laid upon him and ordered. Although when asked about the cargo-ship, you hold in memory what he answered: that it was built with public works publicly compelled, and that a Mamertine senator had publicly been in charge of its building. The same man flees privately to you, gentlemen; he uses this law by which the trial is held, the common citadel of allies. Although the law is on extortion, he says he is not asking back money, of which he does not so greatly miss what was snatched. He says he is asking back from you the sacred things of his ancestors; he demands again from you the household gods of his fathers.

\ciceroSection{18} Is there any modesty, Verres, any religion, any fear? You lived at Heius's house at Messana; you saw him performing divine rites before those gods in his own shrine almost daily. He is not moved by money; finally, things which were for ornament's sake he does not ask for: keep your Canephori, restore the images of the gods. Because he said this, because at the time given he modestly complained before you, an ally and friend of the Roman people, because he was loyal to his religious obligation not only in asking back his ancestral gods but even in his very testimony and oath, know that a man was sent by this man to Messana --- one of the envoys, that very man who was in public charge of building this man's ship --- who should ask the senate that Heius be afflicted with ignominy.

\ciceroSection{19} Most senseless of men, what did you suppose? That you would obtain it? Were you ignorant of how much he was esteemed by his fellow citizens, how much his authority was valued? But suppose you obtained it: suppose the Mamertines had set down something graver against Heius. How great do you think the authority of their praise will be, if they have established a punishment against the man whom it is agreed gave true testimony? But what kind of praise is it, when the praiser, when questioned, must of necessity wound? What? Are these praisers of yours not my witnesses? Heius is a praiser: he wounded most gravely. I shall produce the rest. They will keep silent willingly about what they can; they will say what is necessary unwillingly. Will they deny that the largest cargo-ship was built at Messana? Let them deny it, if they can. Will they deny that a Mamertine senator was in public charge of building that ship? Would that they would deny it! There are also other things, which I prefer to keep entire, to give them as little time as possible to meditate and confirm their perjury.

\ciceroSection{20} Shall this praise be reckoned in the number for you? Shall these men relieve you by their authority? Who neither owe you help if they could give it, nor can if they wished, on whom you have laid most private wrongs and insults; in whose town you have made many whole households for ever infamous by your debauches and disgraces. "But you have served them publicly." Not without great hurt to the commonwealth and the province of Sicily. They ought to give and were wont to give 60,000 modii of wheat bought to the Roman people: by you alone has it been remitted. The commonwealth has taken hurt because through you the right of empire has been lessened in one city; the Sicilians, because that very thing has not been deducted from the total of grain, but has been transferred onto the Centuripines and the Halaesines, free peoples, and on these has been laid more than they could bear.

\ciceroSection{21} You ought by treaty to have demanded a ship: you remitted it for three years; you have demanded no soldier in all those years. You have done as pirates are wont. Pirates, although they are common enemies of all, yet make some men friends to themselves, whom they not only spare but even enrich with plunder; and especially those who hold a town in an opportune place, where ships often must put in. That famous Phaselis which Publius Servilius took had not before been a city of Cilicians and pirates; the Lycians, Greek men, lived in it. But because it was so situated and so jutting out into the deep that the pirates leaving Cilicia often necessarily came to it, and, when they returned from these places, were carried thither, the pirates joined that town to themselves first by trade, then even by partnership.

\ciceroSection{22} The Mamertine state was not before dishonest; it was even an enemy of dishonest men, for it kept the baggage of Gaius Cato (the man who was consul). But of what man! A most distinguished and most powerful man --- who yet, when he had been consul, was condemned. So Gaius Cato, the grandson of two most distinguished men, Lucius Paulus and Marcus Cato, and the son of Publius Africanus's sister: when he was condemned at the time when strict trials were being held, his fine was assessed at 8,000 sesterces. With him these Mamertines were angry, who often spent more on a luncheon for Timarchides than the fine of Cato had been assessed.

\ciceroSection{23} But this state was a Phaselis to that brigand and Sicilian pirate. Hither all things were brought from every side; with these men they were left. What needed to be hidden, they kept set aside and laid up. Through these men what he wished he had loaded secretly and exported in concealment. Finally the largest ship, which loaded with thefts he should send to Italy, he had built and constructed at their hands. For these things, immunity was given by him from outlay, from labour, from military service, from all things at last. For three years they alone, not only in Sicily but (as I think) at this time in the whole world, were free, exempt, loose, and free from every outlay, trouble, and duty. Hence those Verria were born, hence to a banquet Sextus Cominius

\ciceroSection{24} was ordered to be dragged forth, at whom he tried to throw a cup with his own hand, whom with his throat twisted he ordered to be snatched from the banquet into chains and into darkness. Hence that cross on which this man hoisted up a Roman citizen with many looking on, which he dared to fix nowhere save among those with whom he shared all his crimes and brigandages. Do you also come to praise any of you? With what authority? Whether that which you ought to have before the senate, or that before the Roman people?

\ciceroSection{25} Is there any city, not only in our provinces but among the most distant nations, either so powerful or so free or even so monstrous and barbarous, or any king finally, who does not invite a senator of the Roman people under his roof and into his house? An honour which is paid not to the man only, but first to the Roman people, by whose kindness we have come into this order, then to the authority of the order, which, unless it shall be grave among allies and foreign nations, where shall be the name and dignity of empire? The Mamertines did not invite me publicly. When I say me, it is light; if they did not invite a senator of the Roman people, they took away the honour due not to the man but to the order. For to me Tullius lay open the most well-off and most ample house of Gnaeus Pompeius the basilisk, where, even had I been invited by you, I would yet have lodged. There was also the most honourable house of the Percennii (who now likewise are Pompeii), where my brother Lucius lodged with their highest goodwill. A senator of the Roman people, so far as it was in you, in your town lay and spent the night in public. No other city has ever committed this. "But you were calling our friend into court." Will you interpret what private business I am doing by lessening senatorial honour?

\ciceroSection{26} But these things we shall complain of then, if any matter about you shall be conducted by that order, the order which up to now has been despised by you alone. With what face have you committed yourselves to the sight of the Roman people? Did you not first tear down that cross, which still drips with the blood of a Roman citizen, which is fixed at your harbour and city, and cast it into the deep, and purify all that place, before you came to Rome and into this assize of these men? On the soil of the Mamertines, an ally and at peace, a monument of this man's cruelty has been set up. Was your city chosen to which, when men should come from Italy, they should see the cross of a Roman citizen before they had seen any friend of the Roman people? Which you are wont to show to the Reginians, whose state you envy, and likewise to your dependents, Roman citizens, that they may presume the less for themselves and the less despise you, when they see the right of citizenship dispatched by that punishment.

\ciceroSection{27} But you say you bought these. What? Have you forgotten to buy from the same Heius those Attalic tapestries, named throughout all Sicily? It was lawful in the same manner as the statues. For what was done? Or did you spare the records? But this escapes a senseless man: he thought it would be less famous what was carried off out of a wardrobe than what was carried off out of a shrine. But how did he carry them off? I cannot say more plainly than Heius himself said before you. When I had asked whether anything else of his goods had come to Verres, he answered that this man had sent to him to send him the tapestries to Agrigentum. I asked whether he had sent them; he answered (which was necessary) that he had been obedient to the praetor's word, that he had sent them. I asked whether they had reached Agrigentum; he said they had reached. I asked in what manner they had returned; he denied they had yet returned. There was laughter from the people and wonder from all of you.

\ciceroSection{28} Did it not come into your mind here to bid him enter that he had sold these too to you for 6,500 sesterces? Did you fear that your debt would grow, if those things should stand you at 6,500 sesterces which you could easily sell at 200,000? It would have been worth as much, believe me; you would have something to defend. No one would inquire how much that thing was worth. If only you could say that you had bought it, you would easily prove your case and your deed to whom you wished. As it is, with the tapestries you have nothing by which to extricate yourself.

\ciceroSection{29} What? From the phylarch of Centuripa, a wealthy and noble man, the most beautifully made trappings (which are said to have belonged to King Hiero) --- did you carry off, or buy? When I was in Sicily, I was so hearing both from the Centuripines and from the rest --- for the matter was not little known: they said you had carried off these trappings from the phylarch of Centuripa as well as the like noble ones from Aristus of Panhormus, as the third from Cratippus of Tyndaris. For if the phylarch had sold them, you would not, when you became defendant, have promised to give them back. Because you saw that more men knew, you considered: if you had given them back, you would have had less; the matter would not the less be witnessed. So you did not give them back. The phylarch said in evidence that, since he knew that disease of yours (as your friends call it), he had wished to hide the trappings from you; that, when he was asked by you, he denied that he had them, that he had laid them up with another, that they should not be found. That such was your sagacity that you inspected them through that very man with whom they were laid up. Then, caught, he could not deny it. So the trappings were taken from him, against his will, free.

\ciceroSection{30} Now in what fashion he was wont to find out and search out all these things, gentlemen, is worth knowing. There are at Cibyra two brothers, Tlepolemus and Hiero, of whom one (I think) used to model in wax, the other was a painter. These men, I think, when they had come into suspicion among their fellow citizens at Cibyra of having despoiled the temple of Apollo, fearing the penalty of trial and law, fled from home. Because they had come to know Verres as eager for their craft, then, when this man (as you have learned from the witnesses) had come to Cibyra with empty bonds, fleeing from home they betook themselves to him as exiles, when this man was in Asia. He kept them with him at that time and made much use of their work and counsel in the plunders and thefts of the legateship.

\ciceroSection{31} These are those to whom in the books Quintus Tadius enters that he gave at this man's order --- "to the Greek painters." These men, now well known and proved by the trial, he led with him into Sicily. After they had come thither, in a wonderful manner (you would say hunting dogs) they so smelled out and tracked everything that, where each thing was, they found it by some method. One thing by threatening, another by promising, another through slaves, another through the free, another through a friend, another through an enemy they would find. Whatever pleased them must be lost. Those whose silver was demanded had no other wish than that it should displease Hiero and Tlepolemus.

\ciceroSection{32} But by Hercules, gentlemen, I will tell you this: I remember Pamphilus of Lilybaeum, my friend and guest-friend, a noble man, telling me that, when this man had carried off from him by his power a hydria made by the hand of Boethus of distinguished workmanship and great weight, he had returned home truly sad and troubled, because such a vase --- which had been left him by his father and his ancestors, which he was wont to use for festal days, for the visits of guests --- had been carried off from him. "While I was sitting at home, sad," he said, "a Venusian came up. He bids me bring at once the embossed cups to the praetor. I was disturbed," he said. "I had two: I order both to be brought out, lest some greater evil arise, and to be carried with me to the praetor's house. When I come there, the praetor was resting; those Cibyratan brothers were walking about. When they saw me, 'Where are the cups, Pamphilus?' they say. I show them, sad. They praise them. I begin to complain that I shall have nothing of any value if even the cups are carried off. Then they, when they see me troubled, 'What will you give us not to have these carried off from you?' To be brief, they asked me," he said, "for a thousand sesterces. I said I would give them. Meanwhile the praetor calls, demands the cups." Then they had begun to say to the praetor that they had thought, from what they had heard, that Pamphilus's cups were of some value: that it was a poor business, not worthy that Verres should keep them in his silver. He says he too thinks so. So Pamphilus carries off his most excellent cups.

\ciceroSection{33} And by Hercules, although I knew that to understand these things was a kind of trifle, I was yet wont to wonder that this man had any sense at all in these very things, the man whom I knew to have nothing in any matter that resembled a man. Then for the first time I understood that those Cibyratan brothers had been there for this: that, in stealing, this man might use his own hands and their eyes. But he is so eager for this distinguished standing --- to be reckoned a knower in these things --- that lately (see the senselessness of the man!), after he was put off till the morrow, when already he was as if condemned and dead, on the morning of the circus games at Lucius Sisenna's house (a leading man), when the dining-couches were spread and the silver set out in the rooms, and as befits the dignity of Lucius Sisenna the house was full of most honourable men, he came up to the silver and began to consider each piece at leisure. Some wondered at his stupidity, that in the very trial he was increasing the suspicion of that very greed of which he was being accused; others at his senselessness --- who, when so many witnesses had spoken and he was put off, could remember any of those things. But the slaves of Sisenna, I trust --- who had heard what testimonies had been given against him --- never took their eyes from him nor their fingers from the silver.

\ciceroSection{34} It is the part of a good judge to make a guess of each man's greed and self-restraint from small things. He who, as defendant, and as defendant put off by law, by fact and by men's opinion almost condemned, could not in the largest gathering keep himself from handling and inspecting Lucius Sisenna's silver --- this man as praetor in a province will any one suppose could have kept his greed and his hands from the silver of the Sicilians?

\ciceroSection{35} But that we may return to Lilybaeum, whence the speech has gone aside: there is Diocles, the son-in-law of Pamphilus (from whom the hydria was taken), surnamed Popilius. From him, of all the vessels of the sideboard, just as they were set out, he carried them off. Suppose he says he bought them. For here, on account of the size of the theft, the entries (as I think) have been made. He bade Timarchides assess the silver --- in such a manner as he who has assessed most stingily for a gift to actors. Yet I have long erred in spending so many words about your buyings, and asking whether you bought or not, and how, and at what price --- when I can dispatch it in a word. Set out for me, in writing, what silver you got in the province of Sicily, whence each piece was, and at what price you bought it.

\ciceroSection{36} What follows? Although I ought not to ask these writings of you; for it would be fit that I should hold and produce your account-books. But you say you did not draw them up for some of these years. Set together for me what I demand --- about the silver; the rest I shall see to. "I have nothing in writing, nor can I produce it." What then will be? What do you suppose these judges can do? A house full of most beautiful statues already before your praetorship; many things set out at your villas; many laid up with friends; many given and granted to others. The records show no purchase. All the silver has been carried off out of Sicily; nothing has been left to anyone that he might wish to call his own. A dishonest defence is feigned: that the praetor bought up all that silver. Yet this very thing cannot be shown by the records. If, in the records you produce, it is not written in what manner you have what you have; and if of these years in which you say you have bought very many things you produce no records at all --- must you not be condemned both by the records produced and by those not produced?

\ciceroSection{37} You from Marcus Coelius, a Roman knight, a most choice young man, took at Lilybaeum what you wished; you from Gaius Cacurius, a busy and resourceful and most well-favoured man, did not hesitate to take all his furniture; you from Quintus Lutatius Diodorus, who by Quintus Catulus's kindness was made a Roman citizen by Lucius Sulla, with all knowing it, took at Lilybaeum the largest and most beautiful citrus-wood table. I do not throw it at you that you despoiled and plundered the man most worthy of your character, Apollonius son of Nicon of Drepanum (who is now called Aulus Clodius), of all his most beautifully made silver. I am silent. For he does not think he has been wronged, because, when he was already a ruined man and was thrusting his neck into the noose, you came to his help, and shared with him the ancestral goods snatched from the orphan-children of Drepanum. I rejoice even, if you took anything from him; and I say nothing was done by you more rightly. But from Lyso of Lilybaeum, the first man, with whom you lodged --- the statue of Apollo certainly ought not to have been carried off. You will say you bought it. I know it: for 1,000 sesterces. "So I think." I know it, I say. "I shall produce records." Yet that ought not to have been done. From the orphan-boy Heius, of whom Gaius Marcellus is guardian, from whom you had snatched a great sum of money, the saucers with insets at Lilybaeum --- do you say they were bought, or do you confess they were snatched?

\ciceroSection{38} But why do I gather his middling wrongs in such matters, which seem to consist only in this man's thefts and in the losses of those from whom he took? Hear, gentlemen, if you wish, a thing of such a kind that you can see his singular madness and frenzy --- not greed --- now. Diodorus is a Maltese, who before this gave his testimony before you. He has lived for many years now at Lilybaeum, a man both noble at home and, among those to whom he had betaken himself, splendid and well-favoured for his virtue. About him it was said to Verres that he had very fine embossed work, among them certain cups, called \textit{Thericlia}, made by the hand of Mentor with the highest art. When this man heard of it, he was so set on fire by greed not only of seeing but even of carrying off, that he summoned Diodorus to him and demanded them. The other, who would not unwillingly have had them, answered that he did not have them at Lilybaeum, that he had left them at Malta with a kinsman of his.

\ciceroSection{39} Then this man at once sends sure men to Malta. He writes to certain Maltese to seek out those vases. He asks Diodorus to give a letter to that kinsman of his. Nothing seemed longer to him than till he saw that silver. Diodorus, a thrifty and diligent man, who wished to keep his own, writes to his kinsman to answer those who came from Verres that he had sent that silver to Lilybaeum a few days before. He himself meanwhile withdraws; he preferred to be away from home for a while rather than, present, to lose that silver of the best workmanship. When this man heard, he was so disturbed that, without any doubt, he seemed to all to be mad and frantic. Because he could not himself snatch the silver from Diodorus, he was saying that vases of the best workmanship had been snatched from him. He threatened the absent Diodorus, cried out openly, sometimes could not hold his tears. We have heard in the fables of Eriphyle, of such a greed that, when she had seen a necklace (of gold and gems, I think), inflamed by its beauty, she betrayed her husband's life. Like to her was this man's greed --- nay, the keener and the more frantic, that she was desiring what she had seen, while this man's lusts were stirred not only by the eyes but by the ears.

\ciceroSection{40} He orders Diodorus to be sought through the whole province. He had now broken camp out of Sicily and packed up his vessels. The man, that he might in some way recall him to the province, devises this method (if it is to be called a method rather than madness). He sets up one of his own dogs to say that he wished to charge Diodorus the Maltese with a capital crime. At first it seemed wonderful to all that Diodorus should be a defendant --- a most quiet man, most far removed from any suspicion not only of any deed but even of the smallest fault. Then it became clear that all this was on account of the silver. This man does not hesitate to bid the name be entered; and then for the first time, I think, this man received the name of an absent man.

\ciceroSection{41} The thing is famous through all Sicily, that on account of greed for embossed silver men were being made defendants in capital cases, and not only being made defendants but absent. Diodorus at Rome in mourning runs round his patrons and guest-friends and tells everyone the story. Vehement letters are sent to this man by his father, likewise by his friends: that he should see what he was doing in Diodorus's case, how far he was going. The matter was famous and odious; the man was mad; he would perish on this charge alone unless he took care. This man even then thought his father, if not in the place of a parent, at least in the place of a man. He had not yet sufficiently equipped himself for trial; it was the first year of the province; he was not, as in the case of Sthenius, already gorged with money. So his frenzy was held back a little not by shame but by fear and dread. He does not dare condemn Diodorus in his absence; he removes him from the list of defendants. Meanwhile under this man as praetor Diodorus was nearly three years without his province and his home.

\ciceroSection{42} The rest --- not only Sicilians but even Roman citizens --- had laid this down: since this man was going so far in his greed, that there was nothing anyone could think to keep or hold at home of what should somewhat please him. After they understood that the brave man whom the province most expected, Quintus Arrius, was not to succeed --- they laid down that they could have nothing so closed and so hidden that it should not lie most open and most ready to this man's greed. Then he carries off from a splendid and well-favoured Roman knight, Gnaeus Calidius (whose son he knew was a senator and a judge of the Roman people), some noble silver horsemen which had been Quintus Maximus's.

\ciceroSection{43} Imprudent I have fallen here, gentlemen; for he bought, did not carry them off. I would I had not said it. He will toss himself about and ride upon these little horsemen. "I bought; I paid the money." I trust so. "Even the records will be produced." It is worth so much: give me the records. Cleanse, by all means, this Calidian charge, while I can look at the records. Yet what was the cause why Calidius at Rome was complaining that, when he had been doing business in Sicily for so many years, he was so despised, so looked down upon by you alone that he was even, with the rest of the Sicilians, despoiled --- if you had bought? What was the cause why he was confirming that he would demand back the silver from you, if he had sold it to you of his own will? Could you, moreover, see to it that you should not return it to Gnaeus Calidius? Especially when he was so familiar with Lucius Sisenna, your defender, and you had returned to the rest of Sisenna's familiars.

\ciceroSection{44} Finally, I do not think you will deny that to an honourable man (though not more well-favoured than Gnaeus Calidius), Lucius Curidius, you returned the silver through your friend Potamo. He indeed made the case of the rest harder before you. For when you had assured several that you would return it, after Curidius said in evidence that you had returned it to him, you made an end of returning, because you understood that, with the plunder let go from your hands, you could yet not escape testimony. To Gnaeus Calidius, a Roman knight, under all other praetors it had been allowed to have well-made silver; it had been allowed him to be able from his own household supplies, when he had invited a magistrate or any superior, to adorn and prepare a banquet. Many men were at Calidius's house with power and command: no one was found so senseless as to snatch that silver so distinguished and so noble; no one so audacious as to demand it; no one so shameless as to ask that he should sell it.

\ciceroSection{45} For it is arrogant, gentlemen, and not to be borne, that a praetor in his province should say to an honourable, well-off, splendid man, "Sell me your embossed vessels." For this is to say, "You are not worthy to have these things which are so well made; those are for my dignity." Are you, Verres, more worthy than Calidius? --- who, that I may not weigh your life or standing with his (for they are not to be compared); I shall weigh this very thing in which you feign yourself the higher: that you gave 300,000 sesterces to the distributors that you might be returned as praetor, 300,000 to the prosecutor that he should not be hateful to you. On this account do you despise and look down on the equestrian order? On this account did it seem unworthy to you that anything which pleased you should be possessed by Calidius rather than by you?

\ciceroSection{46} He has long been tossing himself about over Calidius; he tells everyone he bought it. From Lucius Papinius too, a leading man, a wealthy and honourable Roman knight --- did you buy a censer? Who said in evidence that you, when you had asked to inspect it, returned it with the inset torn out --- that you may understand there was understanding in the man, not greed; that he was eager for the artistry, not the silver. Nor was this restraint his only in Papinius's case; he kept up this practice in all the censers there were in Sicily. It is incredible how many and how distinguished there were. I trust, when Sicily was flourishing in resources and supplies, there were great works of art on that island. For there was no house, before this praetor, even a little well-off, in which these things were not, even if there was no other silver besides --- a great dish with figures and images of the gods, a bowl with which the women would use for sacred rites, a censer. And all these of old work and the highest artistry were so made that this might be supposed: that there had once been at the Sicilians' houses likewise the rest in proportion; but, although fortune had taken much, those things had nevertheless remained with them which religious sanctity had kept.

\ciceroSection{47} I have said, gentlemen, that there were many such things at almost all the Sicilians'; the same I confirm that there is now not even one. What is this? What monster, what portent have we sent into a province? Does he not seem to you to have set himself this task, that not one man's lust, not his own eyes, but the frenzies of all most greedy men he should fill, when he had returned to Rome? Who, as soon as he had come into any town, those Cibyratan dogs were let loose at once, who tracked and searched out everything. If any large vessel or greater work had been found, they brought it gladly. If they had been able to hunt down anything less of that kind, those at least were taken in place of little hares: dishes, bowls, censers.

\ciceroSection{48} Here what weeping of women, what lamentations are you to suppose were wont to be in these matters? Which perhaps may seem small to you, but they stir up great and bitter grief, especially in little women, when those things are torn from their hands which they were wont to use for sacred rites, which they had received from their own, which had always been in the family. Here do not wait for me to plead these charges from door to door --- that this man took a bowl from Aeschylus of Tyndaris, a dish likewise from Thraso of Tyndaris, a censer from Nymphodorus of Agrigentum. When I shall give witnesses from Sicily, let him pick whom he will whom I shall question about dishes, bowls, censers. Not only no town but not even any house a little well-off will be found that is free of this wrong. He, when he had come into a banquet, if he had seen anything embossed, could not, gentlemen, keep his hands off. Gnaeus Pompeius (who had been Philo) was a Tyndaritan. He gave him a dinner at his villa in Tyndaritan territory. He did what the Sicilians did not dare. He, because he was a Roman citizen, thought he could do it the more without punishment. He set out a dish on which were excellent figures. This man, as soon as he saw it, did not hesitate to take that emblem of his household and hospitable gods from the hospitable table; but yet, as I said before about this man's restraint, with the figures wrenched off, he returned the rest of the silver without any greed.

\ciceroSection{49} What? To Eupolemus of Calacte, a noble man, the guest-friend and most familiar of the Luculli, who is now with the army with Lucius Lucullus --- did he not do the same? He was dining at his house. He had set out the rest of the silver plain, lest he himself be left plain --- two cups, not large, but with insets. This man, as if a festive entertainment, lest he should depart from the banquet without his little crown, in that very place, with the guests looking on, saw to it that the insets were torn out. Nor do I now try to enumerate all this man's deeds; nor is there need, nor can it in any way be done. I produce before you only the tokens and examples of each kind of his various wickedness. For he did not so conduct himself in these matters as if he were sometime to render an account, but plainly as if he were either never going to be a defendant, or, the more he had carried off, the less danger he was going to come into court with. Who handled these things which I am speaking of now not in concealment, not through friends and go-betweens, but openly from his high place by his command and power.

\ciceroSection{50} When he had come to Catina, a wealthy, honourable, well-stocked town, he orders Dionysiarchus, the proagorus (that is, the highest magistrate), to be summoned to him. To him he openly commands that all the silver which was at any man's house at Catina should be sought out and brought to him. Phylarchus the Centuripine, a leading man by birth, by virtue, by money --- did you not hear him say the same thing on oath: that this man had given him the business and ordered him to seek out, in the city by far the largest and most well-off in all Sicily, all the silver from the Centuripines and to order it to be brought to him? At Agyrium likewise, by this man's command, the Corinthian vessels through Apollodorus (whom you heard as witness) were carried away to Syracuse.

\ciceroSection{51} But that is the best of all, that, when he had come to Haluntium, a busy and diligent praetor, he himself would not go up into the town, because the climb was hard and steep, but ordered Archagathus of Haluntium --- a man not only at home but throughout all Sicily among the chief noble --- to be summoned. To him he gave the business, that whatever there was of embossed silver at Haluntium, or even of Corinthian work, all should be carried at once down to the sea out of the town. Archagathus goes up into the town, a noble man who wished to be loved and respected by his own; he was bearing heavily the province given him by this man, nor knew what to do. He proclaims what has been ordered him; he bids all bring forth what they had. There was the highest fear; for the very tyrant did not depart far. Archagathus and the silver he was awaiting, lying in his litter, by the sea below the town.

\ciceroSection{52} What concourse do you suppose was made in the town, what cry, what weeping further of women? Who saw it, would say the Trojan horse had been brought in, that the city had been taken. Vessels carried out without their cases; others wrenched from the hands of women; doors of many broken in; bolts torn off. For what do you think? If shields are sometimes sought out from private houses in war and tumult, men yet give them unwilling, although they feel they are giving them for the common safety. Do not think anyone without the greatest grief brought forth embossed silver from his home for another to snatch. Everything is brought down. The Cibyratan brothers are summoned; a few they reject; from those they had approved, the crusts or the insets are torn off. So the Haluntians, with their delights shaken out of their houses, return home with plain silver.

\ciceroSection{53} What kind of broom of this sort, gentlemen, was there ever in any province? They were wont to divert something from the public most obscurely, through a magistrate. Even when sometimes from a private man, they took it secretly --- and these too were condemned. And if you ask, that I may say something against myself, I think those were prosecutors who pursued the thefts of such men by smell or by some lightly pressed footprint. For we, indeed, what are we doing in the case of Verres, whom we find rolled in mud with the marks of his whole body? It is a great thing to say something against him who, passing by, with his litter set down for a little, not by sleights of hand but openly by his power, by one command, plundered the whole town from door to door. And yet, that he might be able to say he had bought, he commands Archagathus to give to those whose silver had been taken some little coins for form's sake. Archagathus finds a few who would receive them; to these he gave. Yet this man did not return that money to Archagathus. Archagathus wished to demand it back at Rome; Gnaeus Lentulus Marcellinus dissuaded him, as you have heard him himself say. Read. \textsc{Testimony of Archagathus and Lentulus}.

\ciceroSection{54} And lest perhaps you think that he wished to gather without cause this great quantity of insets, see how much he reckoned of you, how much of the standing of the Roman people, how much of the laws and trials, how much of the Sicilian and businessman witnesses. After he had collected such a multitude of insets that he had not left even one to anyone, he set up the largest workshop at Syracuse in the palace. Openly he orders all the artists, the embossers and vase-makers, to be summoned, and he had several of his own. He shuts these in --- a great multitude of men. For eight months on end work did not fail them, with no vase being made save in gold. Then those things which he had wrenched off from dishes and censers he so cunningly fitted into golden cups, so neatly set into golden saucers, that you would say they had been born for that purpose. Yet the praetor himself, who says that by his vigilance there was peace in Sicily, was wont to sit in this workshop the greater part of the day, in a dark tunic and cloak.

\ciceroSection{55} I would not dare to bring forward these things, gentlemen, did I not fear that perhaps you should say you had heard more about him from others in conversation than from me at the trial. For who is there who has not heard about this workshop, about the gold vases, about that cloak of his? Name any good man you wish from the Syracusan assize; I will produce him. There will be no one who does not say he heard or saw this.

\ciceroSection{56} O the times, O the manners! I shall bring forward nothing too old. There are some not a few of you who knew Lucius Piso, the father of this Lucius Piso who was praetor. He, when he was praetor in Spain, in which province he was killed, somehow or other, while he was practising arms, broke and shattered a gold ring he was wearing. Wishing to make himself a ring, he ordered a goldsmith to be called into the forum at his chair at Corduba, and openly weighed out the gold; he orders the man to set up his chair in the forum and make the ring, with all present. Some will perhaps say this was too diligent. Thus far one will reproach him, if any wishes; nothing further. But it must be granted to him; for he was the son of Lucius Piso, of him who first brought the law on extortion.

\ciceroSection{57} It is laughable for me now to speak of Verres, after I have spoken of Piso Frugi. Yet see how much they differ. This man, when he was making vessels of gold for several sideboards, did not labour over what he should hear, not only in Sicily but even at Rome at trial; that man wished, in a half-ounce of gold, that all Spain should know whence the praetor's ring was made. No wonder that this man approved his own name; that man approved his surname. I cannot in any way pursue all this man's deeds either by memory or compass them by speech. I desire only to touch briefly on the kinds, as Piso's ring has just now reminded me of what had wholly slipped my mind. From how many honourable men do you suppose that this man took rings off their fingers? He never hesitated, whenever he was delighted by anyone's gem or ring. I shall say a thing incredible, but so famous that I do not think even he will deny it.

\ciceroSection{58} When a letter from Agrigentum had been brought to Valentius, his interpreter, this man by chance noted the seal in the wax. It pleased him; he asked whence the letter was; the answer was, from Agrigentum. He sent letters to those to whom he was wont, that the ring should be brought to him as soon as possible. So by this man's letters from Lucius Titius, a Roman citizen, a paterfamilias, a ring was drawn off his finger. But that greed of his is incredible. For when he was looking for thirty couches well-made, with the rest of the banquet ornaments, for each of the rooms, which he has not only at Rome but in all his villas, he would seem to be procuring too many things. Yet there was no well-off house in Sicily where he did not set up a weaving-shop.

\ciceroSection{59} A Segestan woman, very rich and noble, by name Lamia, made him for three years, with her household full of looms, coverlets, dyed in nothing save purple. Attalus, a wealthy man at Netum; Lyso at Lilybaeum; Critolaus at Aetna; at Syracuse Aeschrio, Cleomenes, Theomnastus; Archonidas at Helorus --- the day will fail me sooner than the names. "He himself was supplying the purple; only the labour was his friends'." I trust so; for now I am not pleased to bring all charges. As if even this were not enough for charge: that he had so much to give, that he wished to carry off so many things, this finally (which he grants) that the labour of his friends was used in matters of this kind.

\ciceroSection{60} Now, do you suppose that for any save him bronze couches and bronze candelabra were being made at Syracuse for three years on end? "He was buying." I trust so. But I am only making you certain, gentlemen, what this man did as praetor in the province, lest he seem perhaps to have been too negligent and not to have furnished and adorned himself enough when he had the power. I come now not to theft, not to greed, not to lust, but to a deed of such a kind that all unspeakable things seem to me to be contained and to lie within it: in which the immortal gods were violated, the standing and authority of the name of the Roman people was lessened, guest-friendship was despoiled and betrayed, all the most friendly kings, and the nations that are in their realm and rule, were alienated from us by this man's crime.

\ciceroSection{61} For you know that the kings of Syria, the boys, the sons of King Antiochus, were lately at Rome. Who had come not on account of the kingdom of Syria --- for that they held without controversy, as they had received it from their father and from their ancestors --- but they thought that the kingdom of Egypt belonged to them and to Selene, their mother. They, after they had been by the times of the commonwealth shut out and could not transact through the senate what they wished, set out into Syria into their father's kingdom. One of them, who is called Antiochus, wished to make his journey through Sicily, and so under this man as praetor came to Syracuse.

\ciceroSection{62} Here Verres reckoned an inheritance had come to him, since into his realm and hands he had come whom he had heard had many distinguished things with him, and whom he suspected. He sends gifts to the man liberally enough --- for his domestic use --- of oil, of wine as much as seemed good, even of wheat as much as was enough, out of his tithes. Then he summoned the king himself to dinner. He adorns the dining-room amply and magnificently. He sets out those things in which he abounded, very many and most beautiful silver vessels --- for he had not yet made the gold ones. He takes care that the banquet shall be furnished and prepared in all things. To be brief: the king departed in such a way that he reckoned both that this man was richly furnished and that he himself had been received with honour. Then the king himself summons the praetor to dinner. He sets out all his resources, much silver, not a few cups even of gold (which, as is the royal custom and especially in Syria, were marked with the most distinguished gems). There was also a wine-vase of one very large gem, hollowed out into a ladle, with a gold handle, of which (I trust) you have heard a witness fit enough and weighty enough, Quintus Minucius, speak.

\ciceroSection{63} This man took each vase in turn into his hands, praised, marvelled. The king rejoiced that the banquet was pleasing and welcome to the praetor of the Roman people. After they parted thence, this man thought of nothing else (as the matter itself made plain) save how he should send the king out of the province despoiled and stripped. He sends to ask for those vessels which he had seen most beautiful at his house: he says he wishes to show them to his own embossers. The king, who did not know him, gave them most willingly without any suspicion. He sends also to ask for the gem ladle: he wished, he said, to consider it more diligently. That too was sent him.

\ciceroSection{64} Now mark the rest, gentlemen, of which both you have heard, and the Roman people will not now first hear, and which has spread among foreign nations even to the most distant lands. A candelabrum of the most distinguished gems, finished by wonderful workmanship, those kings of whom I speak had brought to Rome, that they might set it up on the Capitoline. Because they had found the temple not yet finished, neither could they set it up nor wished commonly to show or bring it forth, that it should both seem the more magnificent when at its own time it should be set up in the cella of Jupiter Optimus Maximus, and the more distinguished when its beauty should reach men's eyes fresh and untouched. They resolved to take it back with them into Syria, that, when they had heard that the image of Jupiter Optimus Maximus had been dedicated, they should send envoys who, with the rest of things, should bring that singular and most beautiful gift to the Capitoline. The matter came to this man's ears in some manner.

\ciceroSection{65} For the king had wished it concealed, not because he feared or suspected anything, but lest many should anticipate it with their eyes before the Roman people should. This man asks the king and begs him in many words to send it to him. He says he longs to inspect it; that he will give the power of seeing it to no others. Antiochus, who was of a mind both boyish and royal, suspected nothing of this man's wickedness. He bids his men carry it wrapped to the praetorium as secretly as possible. When they had brought it there, and with the wrappings cast off had set it up, this man began to cry out that the thing was worthy of the kingdom of Syria, worthy of a royal gift, worthy of the Capitoline. For it was of that splendour which from the most distinguished and most beautiful gems it had to have, of such a variety of works that art seemed to vie with abundance, of such a size that it could be understood not for the apparatus of men, but for the adornment of the most ample temple, to have been made. When he seemed to have inspected it enough, they begin to take it up to carry it back. He says he wishes to consider it again and again; he is by no means yet sated. He bids them depart and leave the candelabrum. So they then return empty to Antiochus.

\ciceroSection{66} The king at first feared nothing, suspected nothing. One day, a second, more: it is not given back. Then he sends, if it should seem good, that he restore it. This man bids them return to him later. It seems wonderful to him; he sends again; it is not given back. He himself addresses the man, asks that he restore it. Learn the face of the man and his marked shamelessness. What he knew --- what he had heard from the king himself --- was to be set up on the Capitoline; what he saw was being kept for Jupiter Optimus Maximus, for the Roman people --- this he began to ask and most vehemently to demand that he should give him as a gift. When the king said that he was hindered both by the religious obligation to Jupiter Capitoline and by men's standing, because many nations were witnesses of that work and gift, this man began most fiercely to threaten the man. When he sees him no more moved by threats than by prayers, he suddenly bids the man depart from the province before night. He says that he has learned that pirates are coming from his kingdom to Sicily.

\ciceroSection{67} The king, in the largest assembly at Syracuse in the forum (lest anyone perhaps think that I am occupied in some obscure charge and feigning something by men's suspicion), in the forum at Syracuse, weeping and calling gods and men to witness, began to cry out that a candelabrum made of gems, which he was about to send to the Capitoline, which he had wished to be in that most distinguished temple a monument to the Roman people of his partnership and friendship, had been carried off from him by Gaius Verres. About the rest of the works of gold and gems, which were his own and yet were in the hands of that man, he was not concerned; this seemed to him a wretched and unworthy thing to be torn from him. Although it had been before consecrated by his and his brother's mind and thought, yet then in that gathering of Roman citizens he was giving, presenting, declaring, consecrating to Jupiter Optimus Maximus, and was calling Jupiter himself to witness his own will and his religious obligation. What voice, what lungs, what strength can sustain the complaint of this one charge? King Antiochus, who at Rome had been before the eyes of all of us for almost two years with his royal retinue and ornament, who was a friend and ally of the Roman people, with a most friendly father, grandfather, and ancestors, the most ancient and most distinguished kings, and the most opulent and greatest kingdom --- he was driven headlong out of the province of the Roman people.

\ciceroSection{68} How do you suppose foreign nations are about to receive this, how the fame of this deed of yours is about to come into the kingdoms of others and into the most distant lands, when they shall hear that by a praetor of the Roman people in the province a king has been violated, a guest-friend despoiled, an ally and friend of the Roman people cast out? Know that your name, gentlemen, and that of the Roman people will be a subject of hatred and bitterness to foreign nations, if this great wrong of his shall depart unpunished. So all will reckon, especially since this fame of the greed and lust of our men has spread, that this is the deed not of him alone, but also of those who have approved it. Many kings, many free states, many private men, opulent and powerful, have surely in mind to adorn the Capitoline as the dignity of the temple and the name of our empire requires. Who, if they shall understand that you have taken it heavily that this royal gift was made away with, will reckon their zeal and gifts welcome to you and to the Roman people. But if they shall hear that you have neglected this in so noble a king, in so singular a thing, in so bitter a wrong, they will not be so senseless as to spend toil, care, money on those things which they shall not reckon will be welcome to you.

\ciceroSection{69} In this place, Quintus Catulus, I call upon you; for I am speaking of your most distinguished and most beautiful monument. You ought to take up not only a judge's strictness in this charge, but almost the force of an enemy and a prosecutor. For your honour by that temple, by the kindness of the senate and Roman people, the eternal memory of your name is consecrated together with that temple. By you this care is to be taken up, by you this work is to be undertaken: that the Capitoline, as it has been the more magnificently restored, so should be the more copiously adorned than it was; that that flame may seem by the divine will to have come not to destroy the temple of Jupiter Optimus Maximus, but to demand a more distinguished and more magnificent one.

\ciceroSection{70} You have heard Quintus Minucius say that King Antiochus stayed at his house at Syracuse; that he himself knew it had been brought to this man, that he himself knew it was not given back. You have heard, and you will hear, all from the Syracusan assize, who shall say so: that in their hearing it had been dedicated and consecrated by King Antiochus to Jupiter Optimus Maximus. If you were not a judge and the matter had been brought to you, you above all would be the proper man to pursue it, you to demand, you to act. Wherefore I do not doubt with what mind you ought to be the judge of this charge, you who before another judge ought to be a much sharper actor and prosecutor than I am.

\ciceroSection{71} But to you, gentlemen, what can seem more unworthy or less to be borne than this? Will Verres have at home a candelabrum of Jupiter, finished in gems and gold, by whose splendour the temple of Jupiter Optimus Maximus ought to glow and be made bright --- shall it be set up in his house at the kind of banquets which shall blaze with domestic debaucheries and disgraces? In the house of this most foul pimp, together with the rest of Chelidon's hereditary ornaments, shall the ornaments of the Capitoline be set up? What sacred thing will there ever be for him, what religious obligation has there been --- who does not now feel himself bound by such great crime; who comes into court where he cannot even pray to Jupiter Optimus Maximus and seek his help in the manner of all? From whom even the immortal gods seek back their own in that trial which has been established for men to seek back their own things. Do we wonder that Minerva at Athens, Apollo at Delos, Juno at Samos, Diana at Perga, many gods besides throughout all Asia and Greece were violated by this man --- who could not keep his hands from the Capitoline? What private men adorn and shall adorn out of their own moneys, that Gaius Verres did not suffer to be adorned by kings.

\ciceroSection{72} So with this unspeakable crime conceived, in all Sicily afterwards he held nothing to be either sacred or religious. So he conducted himself in that province through three years that war was thought to have been declared by him not only against men but against the immortal gods. Segesta is a very ancient town in Sicily, gentlemen, which men show was founded by Aeneas as he fled from Troy and came into these places. So the Segestans reckon themselves joined to the Roman people not only by perpetual partnership and friendship, but also by kinship. This town once, when that state was waging war with the Carthaginians in its own name and of its own accord, was taken by force and destroyed by the Carthaginians, and all things which could be an ornament to the city were carried from that place to Carthage. There was at the Segestans' a bronze image of Diana, both endowed with the highest and most ancient religious sanctity and finished with singular work and artistry. Brought to Carthage, it had only changed place and men; the old religious obligation it kept; for on account of its singular beauty even to the enemy it seemed worthy that they should worship it most religiously.

\ciceroSection{73} Some generations later Publius Scipio took Carthage in the third Punic war. In which victory --- see the man's virtue and diligence, that you may both rejoice in the most distinguished examples of domestic virtue, and judge the more worthy of hatred this man's incredible audacity --- having called together all the Sicilians, since he had learned that Sicily had been longest and most often harassed by the Carthaginians, he orders all things to be sought out; he promises that it shall be of great care to him that all things should be restored to the cities to which they had belonged. Then those things which had been once at Himera, taken away (of which I spoke before), are restored to the Thermitans; then others to the Geloans, others to the Agrigentines; among which was even that famous bull which Phalaris (the most cruel of all tyrants) is said to have had, in which he was wont to send men down alive for punishment and to set fire beneath. Which bull, when Scipio was giving back to the Agrigentines, he is said to have said it was fair for them to consider whether it was more useful to the Agrigentines to serve their own or to obey the Roman people, when they had at the same time the same monument both of domestic cruelty and of our gentleness.

\ciceroSection{74} At that time to the Segestans with the greatest care this very Diana of which we are speaking is given back. She is brought back to Segesta. In her old seat with the highest congratulation and joy of the citizens she is set up. She had been set at Segesta on a base, very tall, on which in great letters the name of Publius Africanus was carved, and that he, after Carthage was taken, had restored her was inscribed. She was worshipped by the citizens, was visited by all comers. When I was quaestor, nothing was shown me by them sooner. The statue was very ample and tall, with a stole; yet in that great size there was the age and bearing of a maiden. Arrows hung from her shoulder; with the left hand she held a bow; in the right she carried a burning torch.

\ciceroSection{75} When this man, the enemy and the brigand of all sacred things and all religion, had seen it, as if he had been struck by that very torch, so he began to blaze with greed and madness. He commands the magistrates to take it down and to give it to him. He shows that nothing will be more welcome to him. They indeed said that this was unspeakable for them, and that they were held both by the highest religious obligation and by the highest fear of the laws and the courts. This man now begs them, now threatens, now shows hope, now fear. They sometimes set against him the name of Publius Africanus; they said that this was the Roman people's; that they had no power in this thing which a most distinguished general, with the city of the enemy taken, had wished to be a monument of the victory of the Roman people.

\ciceroSection{76} When this man pressed no more lightly --- nay, much more vehemently --- daily, the matter is treated in the senate. From all sides there is vehement protest. So at that time and on his first coming there is a flat refusal. Afterwards, whatever there was of burden in demanding sailors and rowers, in commanding grain, he was laying on the Segestans beyond the rest --- somewhat more than they could bear. Besides, he was summoning their magistrates, was calling each best and noblest man to himself, was dragging them round all the assize-towns of the province, was threatening each one in turn that he would be his ruin, was threatening all together that he would utterly overthrow that state. So at last the Segestans, conquered by many ills and great fear, decreed that they must obey the praetor's command. With great mourning and groaning of the whole state, with many tears and lamentations of all the men and women, the image of Diana is put out for taking down.

\ciceroSection{77} See how great a religious sanctity there was among the Segestans. Know, gentlemen, that no one was found, neither free nor slave, neither citizen nor stranger, who would dare to touch that statue. Know that some barbarian workmen were brought from Lilybaeum. They at last, ignorant of the whole business and the religious sanctity, with wages received, took it down. When it was being brought out of the town, what gathering of women do you think was made, what weeping of the elders? Some of whom even held in memory that day when this same Diana, brought back to Segesta from Carthage, had announced the victory of the Roman people by her return. How unlike was this day to that time! Then a general of the Roman people, a most distinguished man, was bringing back to the Segestans their fathers' gods, recovered out of the city of the enemy. Now from the city of the allies a praetor of the same people, a most foul and most impure man, was carrying off those very same gods by an unspeakable crime. What is more known throughout all Sicily than that all the matrons and maidens of Segesta met when Diana was being carried out of the town, anointed her with unguents, heaped her with wreaths and flowers, with incense, with burning perfumes accompanied her even to the borders of the territory?

\ciceroSection{78} If you, while in command, did not fear this great religious sanctity on account of greed and audacity, do you not now even, in such great peril of yourself and your children, dread it? What man do you think will help you with the immortal gods unwilling, or what god, with such religious sanctities of theirs violated? Did that Diana bring no religious obligation upon you in peace and quiet? She, who, although she had seen the two cities in which she had been placed taken and burned, was twice saved out of the flame and the sword of two wars; who, with her place changed by the victory of the Carthaginians, yet did not lose her religious obligation; who, by the virtue of Publius Africanus, recovered her religious obligation together with her place. With this crime undertaken, when the base was empty and on it the name of Publius Africanus was carved, the matter seemed unworthy and not to be borne to all: that not only had religious obligations been violated, but Gaius Verres had also taken away the glory of the deeds, the memory of the virtue, the monuments of victory of Publius Africanus, that bravest of men.

\ciceroSection{79} When this was reported to him about the base and the writing, he reckoned that men would come into forgetfulness of the whole business if he had taken away the base too as a sign of his crime. So they put it out for taking down by his command --- a contract which has been read to you out of the public records of the Segestans in the previous hearing. You now, Publius Scipio, you, I say, the most choice and the most adorned of young men, I call upon. From you I demand and require the duty owed to your line and your name. Why do you fight for this man, who has plundered the praise and honour of your family? Why do you wish him to be defended? Why do I take up your part, why do I sustain your duty? Why does Marcus Tullius look for the monuments of Publius Africanus, while Publius Scipio defends him who took them away? When the custom has been handed down by our ancestors that each man should so defend the monuments of his ancestors as not even to suffer them to be adorned with the name of others, will you stand by this man, who not in any part obstructed the monument of Publius Scipio, but utterly destroyed and took it away?

\ciceroSection{80} Who then, by the immortal gods, will keep watch over the memory of the dead Publius Scipio? Who over the monuments and tokens of his virtue, if you abandon or desert them, and not only suffer them despoiled, but even defend their despoiler and harasser? The Segestans are here, your clients, the allies and friends of the Roman people. They make you certain that Publius Africanus, after Carthage was destroyed, restored the image of Diana to their ancestors, and that it had been set up and dedicated among the Segestans in the name of that general; that Verres saw to it that this should be torn down and carried off, and that the very name of Publius Scipio should be erased and removed. They beg and beseech you, that you should restore to them their religious obligation, to your line its praise and glory; that what through Publius Africanus they had recovered out of the city of the enemy, this through you they may be able to keep safe out of the brigand's house. What can you either honourably answer them, or they do, save that they implore you and your faith? They are here and implore. You can keep up the amplitude of your domestic praise, Scipio; you can. All those things are in you which either fortune or nature gives men. I do not pluck off the fruit of your duty; I do not seek another's praise for myself. It is not for my modesty, with Publius Scipio, the most flourishing young man, alive and unhurt, to profess myself the champion of the monuments of Publius Scipio and their defender.

\ciceroSection{81} Wherefore, if you take up the patronage of your domestic praise, it will fit me not only to be silent about your monuments, but even to rejoice for Publius Africanus. Such is the fortune of the dead, that his honour is defended by those who are of the same family, and no foreign help is required. But if friendship with this man hinders you, if you reckon what I am asking of you to belong less to your duty, I shall succeed as substitute to your office; I shall take up the parts I had reckoned to be another's. Then let that distinguished nobility cease to complain that the Roman people willingly gives, and has always given, honours to industrious new men. There is nothing to complain of in this state, which on account of virtue rules over all nations, that virtue has the most power. Let there be among others the image of Publius Africanus; let other dead men be adorned by his virtue and his name. Such was that man, so deserving of the Roman people, that he ought to be commended not to one family but to the whole state. There is some share of mine, that I am of that state which he made ample, illustrious, and famous --- chiefly that in those things in my own way I am occupied of which he was chief: equity, industry, self-restraint, the defence of the wretched, the hatred of the dishonest. Which kinship of pursuits and arts is almost no less joined than that in which you delight, of birth and of name.

\ciceroSection{82} I demand back from you, Verres, the monument of Publius Africanus. The case of the Sicilians, which I undertook, I leave aside; the trial for extortion, let it not be at this time; the wrongs of the Segestans, let them be neglected. Let the base of Publius Scipio be restored; let the name of the unconquered general be carved on it; let the most beautiful statue, taken from Carthage, be set up. These things from you not the defender of the Sicilians, not your prosecutor, not the Segestans demand, but he who has undertaken to keep watch over and to preserve the praise and glory of Publius Africanus. I do not fear that this duty of mine will not be approved by Publius Servilius the judge --- who, when he has done the greatest deeds and is now at the height of setting up monuments of his own deeds and labouring on them, will surely wish to hand them on to be defended not only by his own descendants but also by all brave men and good citizens, not to be despoiled by the dishonest. I do not fear that you, Quintus Catulus, will be displeased --- whose monument, the most ample and most distinguished in the world, exists --- that there should be as many guardians of monuments as possible, and that all good men should reckon the defence of another's glory to belong to their own duty.

\ciceroSection{83} For my part I am moved by this man's other thefts and disgraces only so as to think them to be reproached. But here I am afflicted with such great grief that nothing seems to me more unworthy, nothing less to be borne. Will Verres adorn his house, full of debauchery, full of disgrace, full of dishonour, with the monuments of Africanus? Will Verres set up the monument of the most self-restrained and most pure man, the image of the maiden Diana, in that house in which the disgraces of harlots and pimps are always going on?

\ciceroSection{84} But this is the only monument of Africanus you have violated. What? From the Tyndaritans, did you not take away the most beautifully made image of Mercury, set up by the same Scipio's kindness? But how, immortal gods! How audaciously, how lustfully, how shamelessly! You heard the envoys of the Tyndaritans say lately --- most honourable men and chief of the city --- that Mercury, who was worshipped among them with yearly rites and the highest religious sanctity, whom Publius Africanus, after Carthage was taken, had given to the Tyndaritans not only as a monument of his own victory but also of their faith and partnership, was taken away by the force, the crime, the command of this man. Who, as soon as he came into that town, at once, as if it not only were proper to be done but even necessary, as if the senate had ordered this and the Roman people had bidden it, so on the spot ordered the statue to be torn down and carried away to Messana.

\ciceroSection{85} When this seemed unworthy to those who were present, incredible to those who heard, on his first visit it was not pressed by him. As he was leaving, he charges the proagorus Sopater (whose words you have heard) to take it down. When he refused, he threatens vehemently, and at once departs from the town. Sopater refers the matter to the senate; from all sides there is vehement protest. To be brief, he comes to them again somewhat later. He at once asks about the statue. The answer is that the senate does not allow it; that capital punishment had been established if anyone should touch it without the senate's leave. At the same time the religious sanctity is recalled. Then this man says, "What religious sanctity do you tell me of, what punishment, what senate? I shall not leave you alive; you shall die by the rods unless the statue is handed over to me." Sopater again, weeping, refers the matter to the senate; he sets out this man's greed and threats. The senate gives Sopater no answer, but disturbed and troubled departs. He, summoned by the praetor's messenger, sets out the matter; says it can in no way be done.

\ciceroSection{86} And these things --- for nothing of this man's shamelessness must be passed over --- were going on in the assembly openly from his chair and his higher place. It was the highest winter, the weather (as you have heard Sopater himself say) most cold, the rain heaviest, when this man bids the lictors throw Sopater headlong from the colonnade in which he himself was sitting into the forum and set him there naked. Hardly was this fully ordered when you would have seen him stripped and surrounded by the lictors. All thought it was going to come to this: that the wretched and innocent man would be flogged with rods. The opinion deceived these men. Did this man flog without cause an ally and friend of the Roman people? He is not so dishonest. Not all vices are in one man; he was never cruel. He received the man gently and kindly. There are equestrian statues of the Marcelli in the middle of the forum, as in almost the rest of the towns of Sicily. From these this man chose the statue of Gaius Marcellus, whose services to that city and to the whole province were most recent and greatest. On it he orders Sopater --- a man both noble at home and endowed with the highest magistracy --- to be straddled and bound.

\ciceroSection{87} With what torture he was afflicted must come into the mind of all, when he was bound, naked, in the open air, in the rain, in the cold. Yet there was no end to this wrong and cruelty until the people and the whole multitude, moved by the atrocity of the thing and by pity, by their cry compelled the senate to promise this man that statue of Mercury. They cried that the immortal gods themselves would avenge themselves; that meanwhile an innocent man should not perish. Then the senate in great number came to him and promised the statue. So Sopater, from the statue of Gaius Marcellus, when he was now nearly stiffened, scarcely alive, was carried off. I cannot accuse him in order, however much I wish: it is the work not only of talent but of a singular artistry.

\ciceroSection{88} This seems to be one charge, and it is set down by me as one --- about the Mercury of Tyndaris. There are several. But how I may divide and separate them I do not know. It is of moneys taken, that he carried off a statue of great value from allies. It is of embezzlement, that he did not hesitate to carry off a public statue of the Roman people, taken out of the spoils of the enemy, set up in the name of our general. It is of treason, that he dared to overthrow and carry off the monuments of our empire, our glory, our deeds. It is of crime, that he violated the highest religious obligations. It is of cruelty, that against an innocent man, against your ally and friend, he devised a new and singular kind of punishment.

\ciceroSection{89} But that further thing --- what it is I cannot now say; by what name to call it I do not know --- which was done on the statue of Gaius Marcellus. What is this? Was it because he was patron? What then? Whither does it look? Whether ought that thing to weigh for the help or for the calamity of clients and guest-friends? Or that you should show this --- that against your force in their patrons there is no protection? Who would not understand that in the command of a present dishonest man there is greater force than in the patronage of absent good men? Or indeed by this is signified your singular insolence, arrogance, contumacy. You thought, of course, that you were taking something from the amplitude of the Marcelli. So now the Marcelli are not the patrons of the Sicilians; Verres has been substituted in their place.

\ciceroSection{90} What such great virtue or dignity did you reckon there was in you that you should try to translate to yourself the patronage of so splendid, so illustrious a province, to take it away from its surest and most ancient patrons? Can you with that wickedness, stupidity, idleness keep up the patronage not only of all Sicily but of one most slight Sicilian? Was the statue of Marcellus to you as a gallows for the clients of the Marcelli? Did you out of his honour seek punishments against those very men who had paid him the honour? What next? What then did you reckon would be done with your statues? But that thing which happened? For the Tyndaritans threw down this man's statue --- which on account of the Marcelli they had ordered to be set up on a yet higher base --- as soon as they heard of his being put off. So now the fortune of the Sicilians has given you Gaius Marcellus as judge, that, by the religious sanctity of him to whose statue the Sicilians under you as praetor were tied up, we may give you over bound and fastened to these judges.

\ciceroSection{91} And first, gentlemen, this man was saying that the Tyndaritans had sold this Mercury to Marcus Marcellus this Aeserninus, and he was hoping that for his sake Marcus Marcellus himself was about to say so --- which never seemed credible to me, that a young man born in that station, the patron of Sicily, would lend his name to this man for transferring his charge. But the whole matter was so foreseen and provided against by me, that, even if a man had at the highest been found who longed to take Verres's fault and crime upon himself, he could yet do nothing. For I have brought such witnesses and carried away such writings, that of this man's deed there can be no doubt to anyone.

\ciceroSection{92} The public records show that the Mercury was carried to Messana at public expense; they say at how much; that to this business publicly Polea the legate had been in charge. What? Where is he? He is at hand, he is a witness. By the order of the proagorus Sopater. Who is this man? He who was bound to the statue. What? Where is he? You have seen the man and have heard his words. Demetrius the gymnasiarch saw to its taking down, since he was in charge of that place. What? Do we say this? Indeed he himself in person. Recently at Rome this very man promised to give back that statue to the envoys, if the testimony about that thing were taken away and security given that they would not give testimony --- this Zosippus and Ismenias said before you, most noble men and chief of the Tyndaritan state.

\ciceroSection{93} What? At Agrigentum, did you not take away another monument of the same Publius Scipio --- the most beautiful statue of Apollo, on whose thigh in tiny silver letters Myron's name was inscribed --- out of the most religious shrine of Aesculapius? Which thing, gentlemen, when this man had done in secret, when for his crime and that unspeakable theft he had brought in certain dishonest men as leaders and helpers, the city was vehemently moved. For at one time the Agrigentines were missing the kindness of Africanus, their domestic religious obligation, the ornament of the city, the token of victory, the witness of partnership. So by those who were chief in that city it was directed and the business given to the quaestors and aediles that they should keep watch by night at the sacred buildings. For at Agrigentum --- I trust, on account of the multitude of those men and their virtue, and because Roman citizens, brave and honourable men, very many, in that town live and do business with the Agrigentines themselves with the closest mind --- this man did not dare openly to demand or to take what pleased him.

\ciceroSection{94} The temple of Hercules is among the Agrigentines not far from the forum, exceedingly sacred and religious among them. There is a bronze image of Hercules himself, of which I could not easily say I have seen anything more beautiful --- although I do not so much understand in these matters as I have seen many things --- to the point, gentlemen, that his mouth and chin are a little more rubbed, because in prayers and thanksgivings men are wont not only to revere him but even to kiss him. To this temple, when this man was at Agrigentum, with Timarchides as leader, suddenly in the dead of night a gathering and onset of armed slaves is made. A cry is raised by the watchmen and the keepers of the shrine, who at first, when they tried to resist and defend, were beaten down by clubs and cudgels. Afterwards, with the bars torn off and the doors broken in, they try to take down the statue and shake it loose with crowbars. Meanwhile from the cry the rumour spread through the whole city that the household gods were being taken by storm --- not by the unexpected coming of the enemy nor by a sudden onset of brigands, but that out of the house and out of the praetor's staff a band of runaway slaves had come, drawn up and armed.

\ciceroSection{95} There was no man at Agrigentum either of so great age or of so weak strength but, roused by that news, rose up that night, and snatched up the weapon which chance offered each. So in a short time men gathered to the temple from the whole city. For more than an hour now very many men had been struggling to take down the statue. Yet meanwhile it was not shaken from any quarter, when some, with crowbars set under, were trying to move it; others, having tied it round all its limbs, to drag it to themselves with ropes. Then suddenly the Agrigentines run together; great is the stoning. The night-soldiers of that distinguished general put themselves to flight. Yet they take two very small images, that they might not return wholly empty to that brigand of religious sanctities. Never is it so ill with the Sicilians but they say something witty and apt, as in this matter they were saying: that this most monstrous boar (\textit{verres}) ought to be reckoned among the labours of Hercules no less than the Erymanthian boar.

\ciceroSection{96} This virtue of the Agrigentines the Assorines afterwards imitated, brave and faithful men, but by no means from so ample or noble a city. Chrysas is a river which flows through the fields of the Assorines. Among them he is held a god and is worshipped with the highest religious sanctity. His shrine is in the country, near the very road by which one goes from Assorus to Henna. In it is the most beautifully made image of Chrysas in marble. This man did not dare to demand it of the Assorines, on account of the singular religious sanctity of that shrine: he gives the business to Tlepolemus and Hiero. They by night, with a band gathered and armed, come up; they break in the doors of the shrine. The temple-keepers and watchmen feel it in good time. The signal which was familiar to the neighbourhood is given on the horn. Men run together from the fields. Tlepolemus is cast out and put to flight, nor was anything missed from the shrine of Chrysas save one very small bronze statue.

\ciceroSection{97} Of the Great Mother there is a temple at Engyium --- for now I must not only speak briefly about each, but I think very many things must be passed over, that we may come to the greater and more illustrious thefts and crimes of his in this kind. In this temple bronze breastplates and helmets, embossed in Corinthian work, and large hydriae of like kind and finished by the same artistry, that same Scipio --- a man surpassing in all things --- had set up and inscribed his name. Why should I now say more about this man, or complain? All those things, gentlemen, he carried off; nothing in the most religious shrine save the traces of religion violated and the name of Publius Scipio he left. The spoils of the enemy, the monuments of generals, the honours and ornaments of shrines, with these distinguished names lost, will hereafter be reckoned among the equipment and furniture of Verres.

\ciceroSection{98} You alone, of course, are delighted by Corinthian vessels; you discern the alloy of that bronze, you most cunningly perceive the lines of the works! These things Scipio --- that most learned and most humane man --- did not understand: you, with no good art, with no humanity, with no wit, with no learning, understand and judge! See lest he have surpassed not only by self-restraint but even by understanding both you and those of you who wish to be called elegant. For because he understood how beautiful they were, he therefore reckoned that they had been made not for men's luxury but for the adornment of shrines and towns, that to our descendants they should seem to be religious monuments.

\ciceroSection{99} Hear yet his singular greed, gentlemen, his audacity, his madness, especially in polluting those sacred things which it was not lawful to be touched not only by hands, but even violated in thought. There is a shrine of Ceres among the Catinaeans of the same religious sanctity as at Rome, as in the rest of places, as almost in the whole world. In that inmost shrine there was a most ancient statue of Ceres, of which the men did not know not only what kind it was, but not even that it was. For approach into that shrine is not for men; the rites are wont to be performed by women and maidens. This statue, by night, in secret, the slaves of this man took out of that most religious and most ancient place. The next day the priestesses of Ceres and the elders of that shrine, well-respected and noble women, refer the matter to their magistrates. To all it seemed bitter, unworthy, finally lamentable.

\ciceroSection{100} Then this man, moved by the atrocity of the business, that the suspicion of that crime might be removed from himself, gives a certain guest-friend of his the business that he should find someone whom he should accuse of having done it, and should see to it that he should be condemned on that charge, lest he himself be on the charge. The matter is not put off. For when this man had departed from Catina, the name of a certain slave is laid; he is accused; feigned witnesses are given against him. The whole senate of the Catinaeans was judging the matter by their laws. The priestesses are summoned. From them is asked secretly in the senate-house what they thought had been done, in what manner the statue had been carried off. They answer that the praetor's slaves had been seen in that place. The matter, which was not before obscure, by the priestesses' testimony began to be plain. They go to deliberation; that innocent slave is acquitted by all the votes --- that you may the more easily condemn this man by all the votes.

\ciceroSection{101} For what do you ask, Verres? What do you hope, what do you await? What god or man do you think will be of help to you? Did you dare to send slaves to despoil a shrine which the free might not approach even for the sake of adorning? Did you not hesitate to lay hands on those things from which the rights of religion compelled you to keep even your eyes? And yet, not even caught by the eyes did you fall into this fraud so wicked and so unspeakable. For you longed for what you had never seen; you fell in love, I say, with what you had not before looked upon. By your ears you conceived such great greed that no fear, no religion, no force of the gods, no men's standing could hold it in.

\ciceroSection{102} But, I trust, you had heard from a good man and on good authority. How can you, who could not have heard even from a man? You heard, then, from a woman, since neither could men have seen nor known it. What do you suppose, gentlemen, was that woman like, how chaste, who would speak with Verres? How religious, who would set forth the method of despoiling a shrine? And it is in no way wonderful that the rites which are performed in the highest chastity of men and women, those same were violated through this man's debauchery and disgrace. What then? Did he begin to seek by hearing this thing alone, since he himself had not seen it? Indeed several others; out of which I shall pick the despoiling of a most noble and most ancient shrine, about which you have heard the witnesses speak in the previous hearing. Now, please, hear those same things, and attend diligently, as you have done up to now.

\ciceroSection{103} Malta is an island, gentlemen, separated from Sicily by a sea broad enough and dangerous; in it is a town of the same name, to which this man never came, but which was nevertheless his weaving-shop for three years for the making of women's clothing. Not far from that town, on a promontory, there is an ancient shrine of Juno, which was always in such great religious sanctity that not only in those Punic wars which were waged and tossed about with naval forces almost in these places, but even in this great multitude of pirates, it was always inviolate and holy. Indeed even this is handed down to memory: that once the fleet of King Masinissa having put in to that place, his royal prefect carried off ivory tusks of incredible size out of the shrine and bore them to Africa and gave them to Masinissa. The king at first was delighted with the gift; afterwards, when he had heard whence they were, at once sent sure men in a quinquereme, that they should put back those tusks. So on them in Punic letters it was written that King Masinissa had received them in ignorance, and, when the matter became known, had taken care that they should be carried back and put back. There was besides a great quantity of ivory, many ornaments, among which were ivory victories of old work and finished with the highest art.

\ciceroSection{104} All these, that I may not delay you, by one onset and one message through Venus's slaves, whom he had sent for that purpose, he saw to be taken away and carried off. By the immortal gods! What man am I accusing? Whom am I pursuing by the laws or by the right of trial? About whom shall you give your sentence by tablet? The Maltese envoys publicly say that the temple of Juno was despoiled, that this man left nothing in the most religious shrine. To the place into which the fleets of the enemy often came, where pirates almost yearly are wont to winter, which neither did a brigand violate before nor ever an enemy touch --- by this man alone was so despoiled that nothing was left at all. Now will this man be called defendant, or I prosecutor, or this trial? For he is convicted on charges or summoned to court on suspicion! Gods are taken off, shrines harassed, cities stripped naked. Of all these things he has left himself no method either of denying or of defending. In all things he is convicted by me, is overcome by witnesses, is pressed by his own confession, is held in manifest crimes --- and yet stands and silently takes count with me of his deeds!

\ciceroSection{105} I seem to myself too long to dwell in one kind of charges. I feel, gentlemen, that the satiety of your ears and minds must be met. Wherefore I shall pass over many things. But for those things which I am about to say, refresh yourselves, please, gentlemen, by the immortal gods --- those very gods of whose religion I have long been speaking --- while I recall and bring forth that deed of his by which the whole province was moved. About which, if I shall seem to begin somewhat further off and to recall the memory of religion, forgive me. The size of the matter does not allow me to skim briefly over the atrocity of the charge.

\ciceroSection{106} This is an old opinion, gentlemen, which stands fixed by the most ancient writings and monuments of the Greeks: that the whole island of Sicily is consecrated to Ceres and Libera. As the rest of the nations think this, so the Sicilians themselves are so persuaded that it seems to be set and born in their minds. For both they think these goddesses were born in these places, and that fruits were first found in this land; and that Libera (whom they likewise call Proserpina) was carried off out of the grove of the Hennenses --- a place which, because it is set in the middle of the island, is called the navel of Sicily. When Ceres wished to track and seek her out, she is said to have kindled torches at those fires which burst forth from the top of Aetna; carrying which before her, she traversed the whole world.

\ciceroSection{107} Now Henna, where the things I am speaking of are recorded to have been done, is in a most lofty and high place, on the top of which is a level plain of fields and perennial waters. The whole place is cut and sheer from every approach. Round about it are very many lakes and groves and the most joyous flowers at every season of the year, so that the place itself seems to declare that rape of the maiden which we have heard of from boyhood. For nearby is a certain cave, turned to the north, of unbounded depth, by which they say Father Dis suddenly came out with his chariot, and, having seized the maiden from that place, carried her off with him; and suddenly not far from Syracuse plunged beneath the earth; and a lake suddenly arose in that place, where up to this time the Syracusans hold yearly festal days with the largest gathering of men and women. On account of the antiquity of this opinion, because the traces and almost the cradles of these gods are found in these places, throughout all Sicily there is privately and publicly a wonderful religious sanctity of Hennensian Ceres. For many marvels often declare her power and divinity. In many cases in the most difficult matters her present help has been brought, that this island seems by her not only to be loved but even to be inhabited and watched over.

\ciceroSection{108} Nor only the Sicilians, but the rest of the nations and peoples worship Hennensian Ceres most. For if the Athenian rites are sought with the highest greed --- to whom Ceres in that wandering is said to have come and to have brought fruits --- how great a religious obligation is it fitting should be that of those among whom it is agreed that she was born and found fruits? So under our fathers, in a bitter and difficult time of the commonwealth, when, with Tiberius Gracchus killed, the fear of great perils was being foretold by portents, in the consulship of Publius Mucius and Lucius Calpurnius, the Sibylline books were resorted to. Out of which it was found that the most ancient Ceres must be appeased. Then from the most ample college of decemvirs the priests of the Roman people, although there was in our city a most beautiful and most magnificent temple of Ceres, yet went all the way to Henna. For so great was the authority and antiquity of that religion that, when they went thither, they seemed to set out not for the temple of Ceres but for Ceres herself.

\ciceroSection{109} I shall not dwell longer on this. For long since I fear that my speech may seem foreign to the manner of trials and from the daily custom of speaking. This I say: that this very most ancient, most religious Ceres, the chief of all rites which are performed among all peoples and nations, was taken by Gaius Verres from her own temples and seats. Those of you who came to Henna saw the marble image of Ceres, and in another temple that of Libera. They are very ample and most beautiful, but not so ancient. There was a certain bronze one, of moderate size and singular work, with torches, very ancient, by far the most ancient of all those that are in that shrine. This he took away. And yet not content with that.

\ciceroSection{110} Before the temple of Ceres, in an open and uncovered place, are two statues, one of Ceres, the other of Triptolemus, most beautiful and most ample. Beauty was their peril; size their salvation, because their taking down and carrying off seemed most difficult. There stood on Ceres's right hand a great image of Victory, most beautifully made. This this man saw to it should be torn off the statue of Ceres and carried off. What sort of mind is his now in the recollection of his crimes, when I myself in their commemoration am stirred not only in mind but even shudder in body? For there comes into my mind that shrine, that place, that religious sanctity. There turn before my eyes all things --- that day, when, when I had come to Henna, the priestesses of Ceres with fillets and sacred boughs were before me, the public meeting and gathering of citizens, in which while I was speaking such great groanings and weepings were going on that the bitterest mourning seemed to be at large in the whole city.

\ciceroSection{111} They were not complaining of the commands of tithes, not of the plunderings of goods, not of the unjust trials, not of this man's harsh lusts, not of the force, not of the insults by which they had been harassed and oppressed; the divinity of Ceres, the antiquity of the rites, the religious sanctity of the shrine they wished to be expiated by the punishment of this most criminal and most audacious man; they said they were suffering and neglecting all the rest. The grief was so great that Verres seemed a second Orcus to have come to Henna, and not to have carried off Proserpina but Ceres herself. For that city does not seem a city, but a shrine of Ceres. The Hennenses think that Ceres lives among them, so that to me they seemed not citizens of that state, but all priests, all dwellers near and ministers of Ceres.

\ciceroSection{112} Did you, Verres, dare at Henna to take away the image of Ceres? Did you at Henna try to snatch the Victory from Ceres's hand and to take goddess from goddess? Of which things they did not dare to violate or to touch even who had in them all things that are nearer to crime than to religion. For under Publius Popilius and Publius Rupilius as consuls slaves, runaways, barbarians, enemies, held that place; but those slaves were not so much slaves of their masters as you of your lusts; nor those runaways so much from their masters as you from right and from the laws; nor those so much barbarians by tongue and birth as you by nature and morals; nor those so much enemies to men as you to the immortal gods. What entreaty therefore is left to him who has surpassed slaves in unworthiness, runaways in rashness, barbarians in crime, enemies in cruelty?

\ciceroSection{113} You have heard Theodorus and Numenius and Nicasion, the envoys of Henna, publicly say that they had this charge from their fellow citizens: that they should approach Verres and demand back from him the images of Ceres and Victory. If they had obtained this, then to keep up the old custom of the Hennenses, that publicly against him --- although he had harassed Sicily, yet, since they had received these things as established by their ancestors --- they should give no testimony. But if he had not given them back, then to be at the trial, then to instruct the judges of his wrongs, but most of all to complain of the religion. Do not, by the immortal gods, despise their complaints; do not despise and neglect them, gentlemen! The wrongs of the allies are at stake; the force of the laws is at stake; the standing and truth of trials is at stake. All of which are very great. But this is the greatest: with such great religious sanctity is the whole province bound, such great superstition out of this man's deed has taken hold of the minds of all the Sicilians, that whatever public and private hurts happen, on account of that, seem to come about by reason of his crime.

\ciceroSection{114} You have heard the Centuripines, the Agyrines, the Catinaeans, the Aetnenses, the Herbitenses, and very many others publicly say what desolation there was in the fields, what wasting, what flight of the farmers, how deserted, how untilled, how abandoned all things were. Although these things happened by many and various wrongs of his, yet this one cause has the most weight in the opinion of the Sicilians: that, with Ceres violated, all the cult and fruit of Ceres in those places they think have perished. Heal the religious obligation of the allies, gentlemen; keep your own. For neither is this a foreign religion to you nor an alien one. But even if it were, if you should not wish to take it on, yet against him who had violated it you ought to wish to set up sanction.

\ciceroSection{115} But now in the common religion of all peoples, and in those rites which our ancestors fetched and brought from foreign nations and worshipped --- which rites, as they were in fact, so they wished them to be called, Greek --- if we should desire to be careless and dissolute, how can we? I shall recall also the plundering of one most beautiful and most adorned of all cities, Syracuse, and bring it forth, gentlemen, that I may at last close and define the whole speech of this kind. There is hardly anyone of you who has not often heard in what manner Syracuse was taken by Marcus Marcellus, and sometimes even read it in the annals. Compare this peace with that war, the coming of this praetor with the victory of that general, his impure staff with the unconquered army of the other, this man's lusts with the other's self-restraint. By the man who took her, you will say Syracuse was founded; by him who received her established, you will say she was taken.

\ciceroSection{116} And now I leave aside those things which scattered by me in many places shall be said and have been said: that the forum of the Syracusans, which on Marcellus's coming in was kept clean of slaughter, on the coming of Verres ran with the blood of innocent Sicilians; that the harbour of the Syracusans, which then had been closed both to our fleets and to the Carthaginians', was open under this man as praetor to a Cilician light vessel and to brigands. I leave aside the force used against the freeborn, the matrons violated, which were not committed in the city even when taken either by the hatred of the enemy or by the licence of soldiers or by the custom of war or by the right of victory. I leave aside, I say, all these things which were carried out by him through three years. Hear those things which are joined to those things of which I spoke before.

\ciceroSection{117} You have often heard that the city of Syracuse is the largest of Greek cities, the most beautiful of all. It is, gentlemen, as it is said. For both by its situation it is at once strongly fortified and from every approach (whether by land or by sea) most distinguished to look upon, and it has harbours almost shut in by the buildings and embrace of the city: which, although they have separate approaches between themselves, are joined and flow together at the outlet. By their joining, the part of the town which is called the Island, separated by a narrow sea, is again joined and held together by a bridge.

\ciceroSection{118} The city is so great that it is said to consist of four cities; one of which is that I have spoken of, the Island, which, surrounded by two harbours, juts out to the mouth and entry of each. In it is the house which was King Hiero's, which the praetors are wont to use. In it are several sacred buildings, but two which far surpass the rest --- of Diana, and the other (which before this man's coming was most adorned), of Minerva. At the end of this Island is a fountain of fresh water, whose name is Arethusa, of incredible size, very full of fish, which would be wholly covered by the wave were it not separated by a rampart and breakwater of stones from the sea.

\ciceroSection{119} The second city of Syracuse is the one whose name is Achradina, in which is the largest forum, the most beautiful colonnades, the most adorned prytaneum, the most ample senate-house, the famous temple of Olympian Jupiter, and the rest of the parts of the city, which, divided by one continuous broad street and by many cross streets, are filled with private buildings. The third city is the one which (because in that part there was an old shrine of Fortune) is named Tycha, in which is a most ample gymnasium and several sacred buildings; and that part is worshipped and is most thickly inhabited. The fourth is the one which (because it was last built up) is named Neapolis. At the top of which is the largest theatre; besides, two splendid temples, one of Ceres, the other of Libera; and the most beautiful and largest image of Apollo, who is called Temenites: which, if this man had been able to carry, he would not have hesitated to take off.

\ciceroSection{120} Now I shall return to Marcellus, that these things may not seem recalled by me without cause. He, when he had taken so distinguished a city by force and resources, did not think it belonged to the praise of the Roman people to destroy and put out this beauty --- especially since from it no danger was being shown. So he so spared all buildings, public and private, sacred and profane, as if he had come with the army to defend them, not to assault them. In the adornment of the city he had a regard for victory, he had a regard for humanity. Of victory it seemed to him to be to carry many things to Rome which could be an ornament to the city; of humanity, not to despoil the city plainly --- especially one which he had wished to keep safe.

