% In C. Verrem Actio Secunda II (in-verrem-2-2) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.
%
% Book II of the Actio Secunda --- de iurisdictione --- treats Verres's
% administration of justice in Sicily.

\ciceroSection{1} I must necessarily pass over many things, gentlemen, that I may be able somehow at last to speak about those matters which have been entrusted to my faith. For I have undertaken the case of Sicily; that province has drawn me to this business. Yet, having taken up this burden and received the Sicilian case, I have embraced something somewhat larger in mind. For I have undertaken the case of the whole order; I have undertaken the case of the commonwealth, since I thought that judgement could only be given rightly when not only a dishonest defendant should be brought up, but also a diligent and firm prosecutor should come into court.

\ciceroSection{2} Wherefore I must come the more quickly to the case of Sicily, leaving the rest of his thefts and disgraces aside, that I may be able both to act with my strongest powers and to have time enough for speaking. And before I speak of the misfortunes of Sicily, a few things, it seems to me, must be said about the dignity, antiquity, and usefulness of the province. For while you ought to take diligent thought of all our allies and provinces, especially of Sicily, gentlemen, for many and most just causes. First, because of all foreign nations Sicily was the first to attach herself to the friendship and faith of the Roman people. She was the first of all to be called a "province," which is the ornament of empire. She first taught our ancestors how splendid it is to rule over foreign peoples. Of all alone she has been such in faith and goodwill towards the Roman people that the cities of that island, which had once entered into our friendship, never afterwards revolted; the most distinguished have remained in friendship continuously.

\ciceroSection{3} So our ancestors had the step of empire towards Africa from this province. For the great resources of Carthage would not so easily have fallen, had not that supply of grain and that refuge for our fleets lain open. Wherefore Publius Africanus, when Carthage was destroyed, adorned the cities of the Sicilians with the most beautiful statues and monuments, that he might place the most monuments of victory among those whom he reckoned were most rejoicing in the victory of the Roman people.

\ciceroSection{4} Finally that very Marcus Marcellus --- whose virtue the enemies, whose pity the conquered, whose faith the rest of the Sicilians saw in Sicily --- not only consulted for the allies in that war, but even, with the enemies overcome, restrained himself. The most beautiful city of Syracuse --- which, while strongly fortified by hand, was also closed by the nature of its position by land and sea --- when he had taken it by force and counsel, he not only allowed it to be safe, but left it so adorned that it would be at the same time a monument of victory, gentleness, and self-restraint, when men saw both what he had stormed, and whom he had spared, and what he had left. So great honour did he think should be paid to Sicily, that he thought even an enemy city ought not to be removed from an island of allies.

\ciceroSection{5} So we have always so used that province for every purpose, that whatever it could bring forth from itself we considered not to be born among us, but already laid up in our home. When did it not give us what it owed at the proper day? When did it not promise of itself what it judged was needed? When did it refuse what was commanded? So that wise Marcus Cato called Sicily "the storehouse of our commonwealth, the nurse of the Roman commons." We indeed have experienced, in the greatest and most difficult Italian war, that Sicily was for us not as a storehouse, but as that ancient and well-stocked treasury of our ancestors. For without any expense to ourselves, by supplying hides, tunics, and grain, she clothed, fed, and armed our largest armies.

\ciceroSection{6} What of those things which perhaps we do not even feel, gentlemen --- how great they are! That we have so many wealthier citizens, because they have a near, faithful, and profitable province where they easily run, where they gladly do business; whom Sicily in part dismisses with profit and gain by furnishing wares, in part keeps, that they may freely plough, pasture, do business, and finally place their seat and dwelling. Which is no small advantage to the commonwealth: that so great a number of citizens are kept so close to home in such good and fruitful matters.

\ciceroSection{7} And since our taxes and provinces are as it were certain estates of the Roman people, just as you take greatest delight in your nearby estates, so the Roman people delights in the suburb-like nearness of this province. Now indeed the patience, virtue, and frugality of the men themselves is such, gentlemen, that they seem closest to that ancient discipline of ours, not to this which has now grown common: nothing like the rest of the Greeks, no idleness, no luxury; on the contrary the highest labour in public and private affairs, the highest thrift, the highest diligence. So they love our men that they alone find neither publican nor businessman hateful.

\ciceroSection{8} The wrongs of our magistrates they have so endured for many that they never before this time fled by public counsel to the altar of laws and your protection --- although they had borne even that year which had so afflicted them that they could not be safe, unless Gaius Marcellus had come almost as by some fate, that twice from the same family the safety of Sicily should be established; and after that they had felt the unlimited command of Marcus Antonius. So they had received from their ancestors that the kindnesses of the Roman people towards the Sicilians were so great that they thought even the wrongs of our men ought to be borne.

\ciceroSection{9} The cities have given testimony in public against no man before this. They would have endured even this man, had he offended in a human manner, in customary fashion, finally in some one kind. But when they could not bear the luxury, the cruelty, the greed, the arrogance --- when they had lost all their advantages, rights, and benefits from the senate and Roman people through the crime and lust of one man --- they decided either to avenge his wrongs through you and pursue them, or, if they seemed unworthy to you for whom you should bring help and aid, to leave their cities and seats, since now, harassed by his wrongs, they had already left their fields.

\ciceroSection{10} With this plan they sought from Lucius Metellus by entire embassies that he succeed Verres as soon as possible. With this mind they so often before their patrons lamented their miseries. Moved by this grief, they laid before the consuls demands, which seemed to be not demands but charges against him. They even brought it about that I --- whose faith and self-restraint they had known --- almost from the state of my life, by their grief and tears, they led to prosecute the man, from whom my method and will was furthest removed; although in this case I seem to have undertaken many more parts of defence than of prosecution.

\ciceroSection{11} Finally from the whole province the most noble men and chief came publicly and privately. Each gravest and amplest city most vehemently pursued its wrongs. But how, gentlemen, did they come? For I seem to myself now that I ought to speak more freely before you for the Sicilians than perhaps they themselves wish: I shall consult their safety rather than their will. Do you think any defendant in any province has ever, in absence, been defended by such great resources, such great greed, against the inquiry of the prosecutor? The quaestors of both provinces who had served under Verres met me with their fasces.

\ciceroSection{12} Their successors, vehemently devoted to him, generously treated from his provisions, were no less keen against me. See what he could do, who had four most zealous defenders and champions in one province as quaestors --- but the praetor and his whole staff so zealous that it was easily plain that not so much was Sicily that province (which they had found empty), as Verres himself, who had departed full. They threatened the Sicilians if they decreed embassies against him, threatened those who had set out, generously promised others if they would praise him, kept by force and guards the gravest witnesses of private affairs, whom we had subpoenaed in their presence.

\ciceroSection{13} When all these things had been done, yet know that there was only one city, the Mamertine, which sent embassies publicly to praise him. The chief man of that embassy, the most noble citizen of the city, Gaius Heius, you heard speak under oath: that a very large cargo ship was built for Verres at Messana publicly with forced labour. And the same Mamertine envoy, his praiser, said that Verres had taken not only his goods but even the sacred things and household gods handed down from his ancestors out of his house. Splendid praise, when in two matters the work of envoys is consumed at the same time --- in praising and in demanding back! And by what reason that very city is friendly to him, will be told in its proper place. For you will find that the very causes of the Mamertines' goodwill towards him are sufficient causes for his condemnation. No other city, gentlemen, praises him by public counsel.

\ciceroSection{14} The force of the highest command had power among very few men --- not cities --- to such an extent that either the most worthless from the most wretched and most deserted towns were found who set out without the order of people or senate, or those who had been decreed as envoys against Verres, and had received public testimony and instructions, were kept back by force and fear. Which I have not unwillingly seen happen in a few cases, by which the rest of so many and such great and weighty cities --- finally all Sicily --- might have more authority with you, when you saw that no force could keep them back, no danger could prevent them from trying whether their complaints have any weight with you, of the most ancient and most faithful allies.

\ciceroSection{15} For what perhaps not a few of you have heard --- that he is being praised publicly by the Syracusans --- although in the previous hearing, from the testimony of Heraclius the Syracusan, you learned what kind that thing is, yet in another place I will show how the whole matter stands as far as that city is concerned. For you will understand that no men have anyone in such hatred as the Syracusans have and have had Verres. "But the Sicilians alone pursue him: the Roman citizens who do business in Sicily defend, love, wish him to be safe." First, if it were so, yet you in this inquiry on extortion, which has been established for the cause of allies by an allied law and trial, ought to hear the complaints of the allies.

\ciceroSection{16} But you could understand in the previous hearing that the most honourable Roman citizens from Sicily, on very many and the greatest matters --- both wrongs they themselves had received and those they knew had been done to others --- were giving testimony. I confirm what I understand, gentlemen, thus. I seem to myself to have done a pleasing thing to the Sicilians, that I have pursued their wrongs by my labour, my enmities, my danger. No less pleasing have I done, I understand, for our citizens, who think that the safety of their right, of their freedom, of their affairs and fortunes consists in his condemnation.

\ciceroSection{17} Wherefore concerning his Sicilian praetorship I do not refuse but you so hear me that, if to any kind of men --- whether of the Sicilians or of our citizens --- if to any order, whether of farmers or of cattle-men or of merchants, he has been approved; if he has not been the common enemy and brigand of all of these; finally if he has spared anyone in any matter, that you also spare him. Who, as soon as Sicily fell to him by lot as a province, immediately at Rome, before he set out from the city, began to inquire and discuss with his men by what means he could in that province make the most money in one year. He did not wish to learn in the doing --- although he was not raw and a recruit in the matter of provinces. He wished to come into Sicily prepared for plunder and meditated.

\ciceroSection{18} Splendid the omen of common talk and rumour cast forth by the multitude into that province, when from this man's name they most ridiculously divined what he would do in the province! For who could doubt --- when he recalled his flight and theft in the quaestorship, when he thought of the plunderings of towns and shrines in the legateship, when he saw the brigandages of the city praetorship in the Forum --- what kind of man he would be in the fourth act of his dishonesty? And, that you may understand he had sought at Rome not only kinds of stealing but also definite names, hear an argument by which you may more easily judge of his singular shamelessness.

\ciceroSection{19} On the day he touched Sicily --- see whether he came sufficiently prepared by that urban omen for sweeping out the province --- immediately he sends letters from Messana to Halaesa (which I think he wrote in Italy, for as soon as he stepped from the ship he gave them): that Dio of Halaesa should come to him at once; that he wished to take cognizance of an inheritance which had come to his son from a kinsman, Apollodorus Laphiro.

\ciceroSection{20} It was, gentlemen, a very great sum of money. This Dio, gentlemen, is the man now made Roman citizen by the kindness of Quintus Metellus, of whom by many leading witnesses and by the books of many you were satisfied in the previous hearing: that one million sesterces was counted out so that, with Verres taking cognizance, he might prevail in a case in which there could not be the slightest doubt. Besides, herds of the noblest mares were driven away; what silver and bedclothes were in his house were plundered. So Quintus Dion lost one million sesterces, on no other account than that an inheritance had come to him.

\ciceroSection{21} What of it? Under what praetor had this inheritance come to Dion's son? Under the same as Annia, daughter of the senator Publius Annius; under the same as the senator Marcus Ligus --- under the praetor Gaius Sacerdos. What of it? Did no one then trouble Dio? No more than Ligus under Sacerdos as praetor. What of it? Then who reported to Verres? No one --- unless you suppose that the four-time informers had been waiting at the strait. When he was at Rome, he heard that a very great inheritance had come to a certain Sicilian Dio. The heir was ordered to set up statues in the forum; if he did not set them up, his hereditas was forfeit to Erucine Venus. Although they had been set up by the will, he supposed, since the name of Venus was in it, that he would find a cause for malicious prosecution.

\ciceroSection{22} So he sets up someone to claim that inheritance for Erucine Venus. For not the quaestor claims it, as is the custom, who held Eryx mountain. A certain Naevius Turpio claims it --- this man's runner and emissary, the worst of all the four-time informers from that conventus, condemned for wrong-doing under Gaius Sacerdos as praetor. For the case was such that the praetor himself, although he was looking for a malicious prosecutor, could not find a more thoughtful one. He acquits this man for Venus, condemns him for himself. He preferred, of course, that men should sin rather than the gods: that he should rather take from Dio what was not allowed than that Venus should take what was not owed.

\ciceroSection{23} Why should I now read the testimony of Sextus Pompeius Chlorus, who pleaded Dio's case, who was present at all matters --- a most honourable man, although a Roman citizen for his virtue's sake long since, yet first and most noble of all the Sicilians? What of Quintus Caecilius Dion himself, a man most approved and most modest? What of Lucius Caecilius, of Lucius Ligus, of Titus Manilius, of Lucius Calenus --- by all of whose testimonies the matter of Dion's money has been confirmed? Marcus Lucullus said the same thing: that he had already known these wrongs of Dion, by reason of the hospitality which he had with him.

\ciceroSection{24} What? Has Lucullus, who was then in Macedonia, learned these things better than you, Hortensius, who were at Rome --- to whom Dio fled, who about Dion's wrongs most gravely complained by letters with Verres? Are these things new to you? Unforeseen? Now first do your ears receive this charge? Have you heard nothing from Dio, nothing from your mother-in-law Servilia, that leading lady, the old hostess of Dio? Do not many of my witnesses say what you yourself know? Did not the legal exception, not this man's innocence, snatch you from being a witness for me on this charge? Read: "Testimonies of Lucullus, Chlorus, Dion." Does the Venus-man not seem to you to have sought a great enough sum of money under Venus's name --- a man who had set out for the province from Chelidon's lap?

\ciceroSection{25} Hear another no less shameless malicious prosecution in a smaller sum. Sosippus and Philocrates are brothers of Agyrium. Their father died twenty-two years ago. In whose will, in a certain place, if anything were forfeit, there was a fine to Venus. In the very twentieth year, when so many praetors, so many quaestors, so many malicious prosecutors had in the meantime been in the province, the inheritance was demanded from these men in Venus's name. Verres takes cognizance of the case, receives money through Volcatius, near 400,000 sesterces, from the two brothers. You have heard the testimonies of many before. The Agyrium brothers won --- in such a way that they departed empty and destitute.

\ciceroSection{26} "But that money did not reach Verres." What kind of defence is this? Is it asserted in this matter, or attempted? For the matter is novel to me. Verres set up the malicious prosecutor; Verres ordered him to be present; Verres took cognizance; Verres gave judgement. The greatest sums of money were given. Those who gave prevailed in their cases. Will you defend yourself to me thus: "That money was not counted out to Verres"? I help you. My witnesses say the same: they say they gave to Volcatius. What such great force was in Volcatius that he could take 400,000 sesterces from two men? Would anyone, if Volcatius had come of his own accord, have given him a single half-as? Let him come now, let him try; no one will receive him under his roof. But I say more: I charge you with having taken forty million sesterces against the laws; I deny that any one penny was counted out to you yourself. But when on account of your decrees, on account of your edicts, on account of your commands, on account of your judgements, money was given, it was not to be asked by whose hand it was counted, but by whose wrong it was extorted.

\ciceroSection{27} Those chosen companions of yours were your hands. The prefects, the secretaries, the assistants, the doctors, the haruspices, the criers were your hands. The closer each was to you by kinship, by marriage, by some bond, the more he was thought your hand. That whole staff, which gave more evil to Sicily than if a hundred companies of fugitives had been there, was without controversy your hand. Whatever was taken by anyone of these, it must be judged not only given to you but counted out by your own hand. For if you approve this defence: "He did not himself receive it" --- you may abolish all trials on extortion. No defendant will ever be brought up so guilty that he cannot use this defence. For when Verres uses it, what defendant will ever afterwards be so ruined that he is not referred to the innocence of Quintus Mucius, if compared with him? Nor do they now seem so much to defend Verres, as in Verres to test the method of defence.

\ciceroSection{28} On which matter, gentlemen, you must take the highest care. This concerns the highest interest of the commonwealth, and the reputation of our order, and the safety of the allies. For if we wish to be reckoned innocent, we must answer not only for ourselves but for our companions. First of all care must be taken that we lead with us those who will look to our reputation and our life. Then, if in choosing men the hope of friendship has deceived us, we should both avenge it and dismiss them, always living in such a way that we judge an account must be rendered. This is from Africanus, that most generous man --- but the generosity to be approved is that which is without danger to one's reputation, as it was in him:

\ciceroSection{29} when a certain old hanger-on, of the number of his friends, could not obtain from him that he be led as prefect to Africa, and bore it ill: "Do not wonder," he said, "if you do not obtain this from me. I have long been asking him, whom I judge will hold my reputation dear, to set out as my prefect, and so far I cannot obtain it." For in truth it should be sought much more from men, if we wish to be safe and honourable, that they go with us into the province, than that it be conferred as a kindness. But when you both invited your friends into the province as if into plunder, and with them and through them plundered, and gave them gold rings in public meeting, did you not consider that an account must be rendered by you not only of your own deeds but also of theirs?

\ciceroSection{30} When he had set up these great and rich profits for himself out of those cases which he himself had instituted, with his council --- that is, with his staff --- to take cognizance of, then he had invented that infinite kind for seizing innumerable money. It is doubtful to no one but that all the money of all is placed in the power of those who give judgement, and of those who judge; that none of you can hold his houses, none his farm, none his ancestral goods, if, when these have been demanded by anyone from any one of you, the dishonest praetor (against whom no one can intervene) gives whomever he wishes as judge, and the worthless and frivolous judge judges what the praetor has bidden.

\ciceroSection{31} If, indeed, this also is added --- that the praetor gives judgement in such words that not even Lucius Octavius Balbus as judge, a man most skilled in law and duty, could judge otherwise --- if the trial were of such a kind: "Let Lucius Octavius be judge. If it appears that the Capenate farm, concerning which it is at issue, is by the right of the Quirites Publius Servilius's, and that farm is not restored to Quintus Catulus..." it would not be necessary for Lucius Octavius the judge either to compel Publius Servilius to restore the farm to Quintus Catulus, or to condemn the man whom he ought not. Of this kind was the entire praetorian law, the whole judicial business in Sicily, for three years under Verres as praetor. Decrees of this kind: "If he does not receive what you say is owed, prosecute. If he claims, lead him to court." He ordered Gaius Fuficius the claimant to be led, and Lucius Suettius, Lucius Racilius. Trials of this kind: those who were Roman citizens judged whether men were Sicilians, when Sicilians ought by their laws to be given to others; those who were Sicilians, whether men were Roman citizens.

\ciceroSection{32} But, that you may embrace the whole kind of trials, learn first the rights of the Sicilians, then this man's institutions. The Sicilians are by this rule: that what a citizen pursues against a citizen, he should contest at home by his own laws; what a Sicilian against a Sicilian not of the same state, that the praetor allots judges by the decree of the ten legates of Publius Rupilius (which he established by the opinion of the ten legates --- which they call the lex Rupilia). What a private man claims from a people or a people from a private man, the senate from some city is given to judge, with alternate cities being rejected. What a Roman citizen claims from a Sicilian, a Sicilian judge; what a Sicilian from a Roman citizen, a Roman citizen is given. For other matters chosen judges from the conventus of Roman citizens are usually proposed. Between farmers and tithe-collectors trials are held under the grain law which they call the lex Hieronica.

\ciceroSection{33} All these things, under this praetor, were not only thrown into confusion but plainly snatched from both Sicilians and Roman citizens. First, their own laws: when a citizen was pursuing a citizen, either he gave whatever judge it suited him --- a herald, a haruspex, his own physician --- or, if a trial had been established by laws and they had come to a fellow-citizen as judge, the citizen could not freely give judgement. For learn the man's edict, by which edict he had drawn all trials into his own power: "If anyone judged wrongly, he himself would take cognizance; when he had taken cognizance, he would punish." When he was doing this, no one doubted that --- when the judge thought another would judge his own judgement and he himself would come into peril of his life on it --- he would look to the will of the man who would judge him on his life on the spot.

\ciceroSection{34} No chosen judges from the conventus, none proposed from the businessmen. This abundance, which I am saying, was the staff of judges --- not of Quintus Scaevola (who nevertheless was not accustomed to give from his own staff), but of Gaius Verres. Of what kind do you suppose the staff was, of which this man was the chief? If anywhere you see the edict, "If any senate judges wrongly," I shall show that man too, if any one was given as judge, judged what he did not believe by the compulsion of this man. Out of the lex Rupilia no allotment, except when nothing concerned this man. By the lex Hieronica the trials of very many controversies were abolished under one heading. From the conventus and the businessmen no judges. You see what power he had: now learn the matters he conducted.

\ciceroSection{35} Heraclius, son of Hiero, is a Syracusan, a man among the foremost noble in his own home, and before this praetor probably the wealthiest of the Syracusans, now from no other calamity but this man's greed and wrong the poorest. To him an inheritance came easily of three million sesterces by the will of his kinsman Heraclius. The house was full of the best chased silver and much bedclothes and precious slaves. In which matters who is ignorant of this man's greeds and follies? It was talked of: that a great sum had been left to Heraclius; that not only Heraclius would be wealthy but also adorned with furniture, silver, clothing, slaves.

\ciceroSection{36} Verres also hears this; and at first by that gentler artifice of his he attempts to approach Heraclius, to ask him for things to inspect which he would not return. Then he is admonished by certain Syracusans --- these certain ones were Verres's connections, whose wives Verres never thought belonged to others, Cleomenes and Aeschrio, who you will understand from the rest of the charges how powerful with him they were and on what disgraceful cause. These, as I say, admonish the man that the matter is splendid, packed with everything; but Heraclius himself is a man older in years, not most prompt; that besides the Marcelli, he had no patron whom he could approach or appeal to in his own right; that in that will, by which he had been written as heir, he ought to set up statues in the palaestra. "We shall arrange that the palaestrites say that they have not been set up in accordance with the will, and demand the inheritance, since they say it has been forfeited to the palaestra."

\ciceroSection{37} The plan pleased Verres. For he so foresaw that, when so great an inheritance came into controversy and was sought in court, it could not be that he himself should depart without plunder. He approves the plan; he is the author that they should begin to act as soon as possible, and attack a man of that age (least litigious) as tumultuously as possible. A suit is laid against Heraclius. At first all wonder at the dishonesty of the malicious prosecution. Then those who knew this man partly suspected, partly plainly saw that an eye had been cast at the inheritance. Meanwhile the day arrived on which Verres had edicted that he would by custom and by the lex Rupilia allot suits at Syracuse. He had come prepared for the allotment of this suit. Then Heraclius shows him that he could not be allotted on that day, because the lex Rupilia forbade a suit to be allotted within thirty days of when it had been written. Thirty days had not yet passed. Heraclius hoped that, if he should escape that day, before the next allotment Quintus Arrius (whom the province was then most eagerly expecting) would have succeeded.

\ciceroSection{38} Verres put off the day of all the suits, and fixed that day so that he could allot this suit of Heraclius after thirty days, by the law. After that day came, Verres begins to pretend that he wishes to allot. Heraclius approaches with his advocates and asks that he be allowed to dispute on equal right with the palaestrites, that is, with the Syracusan people. The adversaries demand that judges be given for the matter, chosen from those cities which gathered at that conventus, those who seemed best to Verres. Heraclius on the contrary demands that judges be given by the lex Rupilia, that the institutions of his predecessors, the authority of the senate, the right of all the Sicilians be not departed from.

\ciceroSection{39} Why should I demonstrate this man's lust and crimes in giving judgement? Who of you has not learned of his city jurisdiction? Who under this praetor was ever able, with Chelidon unwilling, to bring a lawsuit? The province did not corrupt this man, as it has not no one. He was the same as he had been at Rome. When Heraclius said what all understood --- that there is a definite right for the Sicilians on what right they should contest among themselves; that there was a lex Rupilia which the consul Publius Rupilius had given by the opinion of the ten legates; that all consuls and praetors had always preserved this in Sicily --- Verres said he would not allot by the lex Rupilia: he gave five judges, those whom he found convenient.

\ciceroSection{40} What are you to do with this man? What punishment to find worthy of his lust? Although it had been written for you --- most worthless and most shameless of men --- in what fashion you should give judges among the Sicilians; although the authority of an imperator of the Roman people, the dignity of the ten legates, the most illustrious men, the senatorial decree intervened (by which senatorial decree Publius Rupilius by the opinion of the ten legates had established laws in Sicily); although all before you as praetor had preserved the Rupilian laws both in other matters and especially in trials --- did you dare for nothing in face of your plunder to count so many most sacred matters? Was no law to you, no religion, no shame for your reputation, no fear of trial? With you no man's authority weighty, no precedent which you wished to follow?

\ciceroSection{41} But, as I had begun to say --- with five judges given by no law, by no institution, by no rejection, by no allotment, out of his lust, not who would take cognizance of the cause but who would judge what had been ordered --- on that day nothing was done. They are ordered to be present the next day. Heraclius meanwhile, when he saw all snares being laid against his fortunes by the praetor, takes counsel by the opinion of his friends and kinsmen not to be present at the trial. So he flees from Syracuse that night. Verres next morning, having risen much earlier than ever before, orders the judges to be summoned. When he learned that Heraclius was not present, he begins to compel them to condemn the absent Heraclius. They remind him that, if it seems good to him, he should use his own institution and not compel them before the tenth hour to give judgement against an absent man for one present. They obtain this.

\ciceroSection{42} Meanwhile he himself and his friends and counsellors began to be much troubled, and to bear it ill that Heraclius had fled. They thought the condemnation of an absent man --- especially of so great a sum --- would be much more odious than if he had been condemned present. Added to this was that the judges had not been given by the lex Rupilia. They saw the matter would seem much more disgraceful and unjust. So while he wishes to correct this, his greed and dishonesty became more open. For he says he will not use those five judges. He orders --- what should from the beginning have been done by the lex Rupilia --- that Heraclius and those who had written the suit be summoned. He says he wishes to allot judges by the law. What he had not been able to obtain from him the day before --- when Heraclius had begged and beseeched with many tears --- it occurred to him the next day: that he ought by the lex Rupilia to allot the suits. He draws three from the urn. He orders them to condemn the absent Heraclius. So they condemn.

\ciceroSection{43} What madness, in evil hour, was this! When did you ever think you would have to render an account of your deeds? When did you reckon such men would hear of these matters? That an inheritance be claimed which is owed to no one for the praetor's plunder? That a city's name be interposed? That the foulest mask of malicious prosecution be put on a respectable city? Nor only this, but the matter so conducted that not even a pretence of equity be employed? For, by the immortal gods, what is the difference whether the praetor commands and compels by force someone to depart from all his goods, or gives a trial of this kind, by which trial without the case being heard he must be overturned from all his fortunes?

\ciceroSection{44} For surely you cannot deny that you ought to have allotted judges by the lex Rupilia, especially when Heraclius demanded it. But if you say this --- that with Heraclius's consent you departed from the law --- you yourself will entangle yourself, you yourself will be implicated by your defence. For why first did he not wish to be present, when he had judges from that number whom he had demanded? Then why did you allot other judges after his flight, if those who had previously been given you had given by the will of both? Then all the other suits at that conventus the quaestor Marcus Postumius allotted: this one alone you will be found to have allotted at that gathering.

\ciceroSection{45} "Therefore," someone will say, "he gave that inheritance to the Syracusan people." First, even if I were willing to confess this, you would still be necessarily condemned. For it is not permitted that with impunity we may snatch from someone and hand over to another. But you will find that he plundered that inheritance in such a way that he did very few things secretly: that the Syracusan people came into the greatest odium by his disgrace, by another's reward; that few Syracusans --- those who now say they have come for the public laudation --- both then were sharers of the plunder and now have come not to laud him but to a common assessment of suits. After he was condemned in absence, not only the inheritance which was disputed, which was three million sesterces, but all of Heraclius's own ancestral goods (which were not less in money) were handed over to the palaestra of the Syracusans, that is, to the Syracusans, in possession.

\ciceroSection{46} What praetorship is this? You snatch an inheritance which had come from a kinsman, had come by will, had come by laws --- those goods which the man who had made the will (somewhat before he died) had handed over to this Heraclius for use and possession; on which inheritance, since he had died somewhat before you became praetor, there had been no controversy, no one had made mention. But let it be so. Snatch the inheritance from kinsmen, give to palaestrites; plunder another's goods in the city's name; overthrow laws, wills, the wishes of the dead, the rights of the living. Was it also necessary to expel Heraclius from his ancestral goods? --- which, as soon as he had fled, how shamelessly, how openly, how bitterly, immortal gods, were those goods plundered! How calamitous that affair seemed to Heraclius, profitable to Verres, disgraceful to the Syracusans, miserable to all! For it is at once seen to: whatever chased silver was in those goods is brought to him, whatever Corinthian vessels, much bedclothes. No one doubted but that not only from that captured and crushed house, but from the whole province, all things must be brought to him. The slaves he wished he led off, others he distributed; an auction was held, in which his unconquered staff lorded it.

\ciceroSection{47} But this is splendid. The Syracusans who had presided over Heraclius's goods --- in word over collecting them, in fact over distributing them --- were rendering an account of these affairs in the senate. They said that several pairs of cups, valuable silver pitchers, much bedclothes, valuable slaves had been given to Verres. They said how much money had been given to each at his order. The Syracusans groaned, but yet endured. Suddenly it is read out that under one heading 300,000 sesterces had been given by the praetor's order. The greatest cry arose of all --- not only of the best men nor of those to whom it had always seemed unworthy that the goods of a private man had been snatched in the people's name by the highest wrong, but even those very authors of the wrong and partners in some little portion of the plunder began to cry out that he should keep the inheritance for himself. So great a clamour arose in the senate-house that the people came running.

\ciceroSection{48} The matter, learned by all the conventus, is quickly reported home to him. The man, hostile to those who had read it out, an enemy to all who had cried out, blazed with anger and bile. But yet he was then unlike himself. You know the man's face, you know his audacity. Yet then by the talk of the people and the cry and the manifest theft of a great sum he was disturbed. When he had collected himself, he calls the Syracusans to him. Since he could not deny that money had been given by them, he did not seek far away (for he could not have proved it), but the next-handy man, almost a second son, whom he said had carried off that money. He shows that he will compel him to return it. He, after he heard this, took account of his own dignity, age, and nobility. He spoke before the senate, he showed that nothing concerned him; about Verres, what all saw, even he did not speak obscurely. So those Syracusans afterwards set up a statue, and he, as soon as he could, left Verres and departed from the province.

\ciceroSection{49} And yet they say he is accustomed to complain sometimes that, wretched man, he is being weighed down not by his own offences and charges, but by those of his men. You held the province three years; your son-in-law, a chosen young man, was with you one year; your comrades, brave men, your legates, left you the first year. One legate, Publius Tadius, who remained, was not so long with you. Who, if he had always been with you, would yet most carefully have spared not only your reputation, but much more his own. What is it that you accuse others? What is it that you think you can not only divert your fault to someone, but share it with another?

\ciceroSection{50} Those 300,000 sesterces are counted out to the Syracusans. How they later returned to Verres through the back door, by documents and witnesses, gentlemen, I shall make plain to you. From this iniquity and dishonesty of his, gentlemen, --- because plunder from those goods had come to many Syracusans, against the will of the people and senate of Syracuse --- those crimes also were committed (through Theomnastus and Aeschrio and Dionysodorus and Cleomenes) most against the will of the city: first, that the whole city be plundered, on which I have set apart another place for speaking; that all the statues, through those men I have named, all the ivory from sacred buildings, all painted panels everywhere, finally the very statues of the gods which he wished, he carried off. Then, that in the senate-house at Syracuse --- that place which they call by the name "bouleuterion" --- a most honourable and most distinguished place among them, where there is a bronze statue of that very Marcus Marcellus (who, at the law of war and victory which would have allowed him to take it, preserved that place for the Syracusans and gave it back), in that very place they should set up a gilded statue of Verres and another of his son: so that, while the memory of this man remained, the Syracusan senate could not be in the senate-house without tears and groans.

\ciceroSection{51} Through these same partners of Verres's thefts, wrongs, and wives, the Marcelliae are abolished at Syracuse by his order, with the greatest groan and grief of the city. That festival they paid back as a debt to Gaius Marcellus's recent kindnesses, and bestowed on the line, name, and family of the Marcelli with the greatest goodwill. Mithridates in Asia, when he had occupied that whole province, did not abolish the Mucia. Although an enemy --- and an enemy in other matters too fierce and savage --- yet he did not wish to violate an honour to a man consecrated by the religion of the gods. Did you not wish to grant to the Syracusans one festal day to the Marcelli, through whom they obtained the power of celebrating the rest of their festal days?

\ciceroSection{52} But you splendidly set up Verria for them to celebrate, and that for that day what was needed for the rites and feasts should be let out for several years! Now, in such great shamelessness of his, something seems to be remitted, lest we strain everything, lest we seem to do everything with grief. For day, voice, and lungs would fail me, if I should now wish to cry aloud how miserable and unworthy it is for a festal day to exist by his name among those who consider themselves utterly extinguished by his work! O splendid Verria! Where, please, did you go, where did you not bring this day with you? For what house, what city did you visit, what shrine, finally, did you not leave overturned and swept clean? Wherefore let those things be called Verria, which are seen to be established not by name but by your hands and nature.

\ciceroSection{53} How easily wrong creeps in, and the habit of offending --- how not easily it is checked --- see, gentlemen. Bidis is a sufficiently slight town, not far from Syracuse. Here by far the first man of the city is a certain Epicrates. To him an inheritance of five hundred thousand sesterces had come from a certain woman, his kinswoman, and so close that --- even if she had died intestate --- Epicrates ought to be heir under the laws of the Bidini. That matter was recent which I described above, of Heraclius the Syracusan, who would not have lost his goods, had not an inheritance come to him. To this Epicrates also, as I said, an inheritance had come.

\ciceroSection{54} His enemies began to think that no less under the same praetor he could be overthrown from his goods, by which Heraclius had been overthrown. They establish the matter secretly. Through his interpreters they bring it to Verres. So the case is set up that the palaestrites of Bidis demand the inheritance from Epicrates, just as the Syracusan palaestrites had demanded it from Heraclius. You have never seen a praetor so palaestric. But he so defended the palaestrites that he himself went away the more anointed by them. Who at once, that something might be present, orders 80,000 sesterces to be counted out to one of his friends.

\ciceroSection{55} The matter could not be sufficiently hidden. Through one of those who had been present Epicrates is informed. At first he began to make light of it and disregard it, because the case had nothing that could be argued. Then, when he thought of Heraclius and knew this man's lust, he thought it most expedient to leave the province secretly. So he did. He set out for Rhegium. When this was heard, those who had given the money were boiling. They thought nothing could be done with Epicrates absent. For Heraclius had at least been present when judges were first given. About this man, who, before going to court, before any mention of controversy was even made, had departed --- they thought nothing could be done. The men set out for Rhegium. They meet Epicrates. They show, what he knew --- that they had given 80,000 sesterces. They ask him to see to it for them, by what had departed from themselves of money, to take security from himself in whatever manner he pleased: that no one would prosecute Epicrates about that inheritance.

\ciceroSection{56} Epicrates dismisses the men, badly received from him in many words. They return from Rhegium to Syracuse. They begin, as happens, to complain among many that they had given 80,000 sesterces in vain. The matter spread and began to be in everyone's mouth and talk. Verres returns to that Syracusan custom of his. He says he wishes to take cognizance about those 80,000 sesterces. He summons many. The Bidini say they gave to Volcatius. They do not add: "by his order." He calls Volcatius, orders the money to be returned. Volcatius, with most equal mind, brings the coins (who lost nothing), gives them back in the sight of many. The Bidini take the coins.

\ciceroSection{57} Someone will say: "Why then do you blame Verres in this --- a man who not only is no thief himself, but did not even allow another to be?" Pay attention; you will at once understand that this money, which only just appeared to leave him, returned by the same path. For what ought a praetor to have done, when, having taken cognizance with his council, he had learned that his companion had received money for the sake of corrupting the law, the decree, the judge --- in which matter the praetor's own life and reputation were concerned, but the Bidini had given money against the praetor's reputation and fortunes? Ought he not both to punish the man who had received and those who had given? You who had instituted to punish those who had judged wrongly --- which often happens through ignorance --- do you allow these to depart with impunity, who because of your decree, because of your judgement, thought money should be either given or received?

\ciceroSection{58} The same Volcatius was afterwards with you, a Roman knight, after such great disgrace received. For what is more disgraceful for a free-born man, what less worthy of a free man, than to be compelled in a great gathering by a magistrate to return a theft? If he had been of the mind that not only a Roman knight but any free man ought to be of, he would not have been able to look at you afterwards. He would have been your foe, your enemy, after such great contumely received --- unless he had then connived with you and served your reputation rather than his. How friendly to you he then was, not only as long as he was with you in the province, but even now when you have already been left by your other friends, both you understand and we can guess. Or is this the only argument that nothing was done with this man's knowledge --- that Volcatius did not get angry with him? --- that this man punished neither Volcatius nor the Bidini?

\ciceroSection{59} It is a great argument, but the greatest is this: that those very Bidini --- with whom this man ought to have been angry, from whom he learned that, because they could do nothing by right with Epicrates (even if he were present), for that reason his decree had been tested by money --- to those men, I say, Verres handed over not only that inheritance which had come to Epicrates, but, as in the Syracusan Heraclius, so in this case too --- somewhat more atrociously, because Epicrates had not been called at all --- his ancestral goods and fortunes to the Bidini. For he showed in a new way: if anyone should claim anything from an absent man, he would hear it. The Bidini approach, demand the inheritance. The agents demand that he refer them to their own laws or order a suit to be written under the lex Rupilia. The adversaries did not dare to speak against. No way out was found. They allege that the man had departed for the sake of fraud. They demand that he order the goods to be possessed.

\ciceroSection{60} Epicrates owed not a single penny to anyone. His friends said that, if anyone claimed anything, they would suffer a trial, would give security for satisfying the judgement. When all plans were chilled, by Verres's prompting they began to allege that Epicrates had corrupted public records --- from which suspicion he was furthest removed. They demand the action of that matter. Friends refuse, lest any trial or his own taking cognizance be established about his reputation in his absence. And at the same time they did not cease to demand that he refer them to their own laws.

\ciceroSection{61} Verres, having found a wide opening, when he sees there is something his friends would not defend in Epicrates's absence, declares that he will give the action of this matter on the first opportunity. When all clearly saw that not only had those coins which had departed from Verres in pretence returned to him, but he had afterwards taken much more, the friends ceased to defend Epicrates. Verres ordered the Bidini to take possession of and keep all of Epicrates's goods. To those 500,000 sesterces of inheritance, his own old fifteen hundred thousand sesterces of property was added. Was the matter so set up from the beginning, or so brought to its conclusion at the end? Or is the sum so small, or is the man Verres such, that what I have said could seem to have been done for nothing?

\ciceroSection{62} Now hear about the misery of the Sicilians, gentlemen. Both Heraclius the Syracusan and this Bidian Epicrates, expelled from all their goods, came to Rome. Wearing mourning, with the longest beards and hair, they were at Rome for almost two years. When Lucius Metellus set out for the province, then they --- well-recommended --- set out together with Metellus. Metellus, as soon as he came to Syracuse, undid both --- about Epicrates and about Heraclius. In the goods of each there was nothing that could be restored, except whatever had not been able to be moved from its place.

\ciceroSection{63} Metellus had splendidly done this on his first arrival --- to undo and make void all of this man's wrongs as far as he could. Because he had ordered Heraclius to be restored and it was not being done, whatever Syracusan senator was named by Heraclius, he ordered to be led to court. So very many were led. Epicrates indeed was at once restored. Other trials were undone at Lilybaeum, others at Agrigentum, others at Panhormus. The censuses which had been held under this praetor, Metellus had shown he would not preserve. The tithes which Verres had sold against the lex Hieronica, he had edicted he would sell again under the lex Hieronica. Everything Metellus did was such that he seemed not so much to be conducting his own praetorship as to be unweaving Verres's. As soon as I came into Sicily, he changed.

\ciceroSection{64} A certain Laetilius had come to him in those two days, a man not foreign to letters --- so this Verres always used him as a courier. He had brought several letters, among them one from home which had wholly changed the man. Suddenly Metellus began to say he wished everything for Verres's sake; that he had friendship and kinship with him. All wondered that this had only now come into his mind, after he had killed him with so many deeds and decrees. There were those who supposed that Laetilius had come from Verres as a legate to mention favour, friendship, and kinship. From that time he began to seek public laudations from the cities, not only deterring witnesses with words but even keeping them by force. Unless I had at my coming somewhat repressed his attempts, and had fought among the Sicilians not by Metellus's letters but by Glabrio's letter and the law, I could not have summoned so many witnesses here.

\ciceroSection{65} But, as I had begun to say, learn the miseries of the allies. That Heraclius and Epicrates went out far to meet me with all their household; arriving at Syracuse, they gave thanks weeping, wished to depart with me to Rome. Because there were several towns I still wished to visit, I arranged with the men what day they should be present for me at Messana. There they sent a messenger that they were being kept by the praetor. Whom I subpoenaed as witnesses, whose names I gave to Metellus, most eager to come, afflicted with the greatest wrongs --- they have not yet come. By such a right are the allies, that they may not even be allowed to lament their own misfortunes.

\ciceroSection{66} Now you have heard the testimony of Heraclius of Centuripa, an excellent and most noble youth, from whom 100,000 sesterces had been claimed by malicious prosecution and malice. Verres, by interposed penalties and compromises, saw to it that 400,000 sesterces was extorted; and because the judgement on the compromise had been made for Heraclius --- when a Centuripan citizen had given judgement between two citizens --- he ordered that to be void, and judged that judge to have judged falsely. He forbade him to be in the senate, to use any of the public privileges. If anyone struck him, he edicted that he would not give a trial of injury; whatever was sought from him, he would give a judge from his own staff; but to him no action of any matter would he give.

\ciceroSection{67} Such was the authority of this man that no one struck him --- although the praetor in his province by word permitted, by deed urged it --- nor did anyone claim anything from him, since Verres by his authority had shown the licence of malicious prosecution. But that grave disgrace remained on the man as long as Verres remained in the province. With this fear thrown into the judges --- by a new manner, by no precedent --- do you suppose that anything in Sicily has been judged but at this man's nod? Does it seem only this matter was at stake (which yet was at stake) --- that this money should be torn from Heraclius? Or also that, in which the greatest plunder lay --- that, under the name of judgements, the goods and fortunes of all should be in this one man's power?

\ciceroSection{68} Now in capital trials, why should I gather each particular matter and case? Out of many similar I shall take those which seem most to surpass in dishonesty. There was a certain Sopater of Halicyae, a man at home among the foremost in wealth and honour. When he had been accused of a capital matter by his enemies before the praetor Gaius Sacerdos, he was easily acquitted in that trial. Against this same Sopater the same enemies, before Gaius Verres (when he had succeeded Sacerdos), laid the indictment of the same matter. The case seemed easy to Sopater, both because he was innocent and because he did not think Verres would dare to disapprove of Sacerdos's judgement. The defendant is summoned. The case is conducted at Syracuse. The charges are handled by the prosecutor which had been previously dissolved not only by defence but also by judgement.

\ciceroSection{69} The case of Sopater was defended by Quintus Minucius, a Roman knight among the foremost splendid and honourable, and to you, gentlemen, not unknown. There was nothing in the case that seemed needing to be feared or even doubted at all. Meanwhile Verres's freedman and assistant Timarchides --- who is, as you have learned in the previous hearing from very many witnesses, the transactor and minister of all matters of this kind --- comes to Sopater. He warns the man not to trust too much in Sacerdos's judgement and his own case; that his prosecutors and enemies have it in mind to give the praetor money; but that the praetor for safety's sake prefers to receive, and at the same time prefers, if it can be done, that the matter once judged should not be undone. Sopater, since this had come on him unexpected and unforeseen, was much disturbed; nor in the moment had he what to answer Timarchides, except that he would consider what to do, and at the same time he showed himself in the greatest financial difficulty. Afterwards he referred the matter to friends, who, when they were authors to him of redeeming his safety, came to Timarchides. Having set out his difficulties, he brings the man down to 80,000 sesterces, and counts out that money to him.

\ciceroSection{70} After they came to plead the case, then truly all who defended Sopater were without fear or worry. There was no charge; the matter was judged; Verres had received the money. Who could doubt what was going to happen? The matter is not finished out that day; the trial is dismissed. A second time Timarchides comes to Sopater. He says his prosecutors are promising the praetor much greater money than this man had given. Therefore, if he had any sense, he should see what to do. The man --- although he was a Sicilian and a defendant, that is, with unfair right and adverse moment --- could not bear to hear Timarchides further. "Do," he said, "what you please. I shall not give more." His friends and defenders thought the same, and even more so because Verres, however he might present himself in that inquiry, yet had on his council honourable men from the Syracusan conventus, who had also been on Sacerdos's council when this same Sopater was acquitted. They reasoned thus: that those same men could in no way condemn Sopater on the same charge, with the same witnesses, who had previously acquitted him. So with this one hope they came to the trial.

\ciceroSection{71} After they had come there, when the same men as usual had assembled in numbers on the council, and the whole defence of Sopater rested on this one hope --- the numbers and dignity of the council, and the fact that they were (as I said) the same who had previously freed Sopater on that very charge --- learn the man's open dishonesty and audacity, hidden by no method, not even by dissimulation. Marcus Petilius, a Roman knight whom he had on his council, he orders to attend, because he was judge in a private case. Petilius refuses, because his friends, whom he wished to be on his council, Verres himself was keeping on his council. This generous man says he is not keeping any of those who wished to be present for Petilius. So all depart; for the others too obtain leave not to be kept, who said they wished to be present for one or the other of those who were then holding that trial. So this man alone is left with his most worthless staff.

\ciceroSection{72} Minucius, who was defending Sopater, did not doubt but that, since Verres had dismissed his council, he would not on that day inquire into the matter. When suddenly he is ordered to speak. He answers: "Before whom?" "Before me," he says, "if I seem suitable to you to judge of a Sicilian and Greek-ling." "You are suitable," he says, "but I would much wish there were present those who had been present before and had taken cognizance of the case." "Speak," he says; "those cannot be present." "By Hercules, indeed," said Minucius, "Petilius asked me also to be present for him on his council." And at the same time he began to leave the benches.

\ciceroSection{73} Verres, angry, follows the man with sharper words, and even began to threaten him more gravely because he was kindling such a charge and so much odium against him. Minucius, who at Syracuse so did business that he remembered his own right and dignity, and who knew that he ought so to increase his estate in the province as not to lose any of his liberty, answered the man what seemed best, and what that moment and case bore: he said he would not defend the case with the council dismissed and sent away. So he departed from the benches, and the same was done by Sopater's other friends and advocates besides the Sicilians.

\ciceroSection{74} Although Verres is of incredible rashness and audacity, yet, suddenly left alone, he was afraid and confused. What to do, where to turn, he did not know. If he had then dismissed the inquiry, he saw that afterwards, with those summoned to his council whom he had sent away, Sopater would be acquitted. But if he had so condemned a wretched and innocent man, when he himself the praetor had been without a council, and the defendant without patron and advocates, and had undone Gaius Sacerdos's judgement, he reckoned he could not bear so great odium. So he was boiling with doubt, turning himself this way and that not only in mind but in body, so that all who were present could understand that fear was fighting with greed in his soul. There was the greatest gathering of men, the highest silence, the highest expectation where his greed would burst forth. Often Timarchides, kindled, leaned down to his ear.

\ciceroSection{75} Then at last Verres said: "Come, speak!" The defendant begs and beseeches that he take cognizance with the council. Then suddenly Verres orders the witnesses to be summoned. One and another speak briefly; nothing is asked; the herald announces that they have spoken. Verres --- as if he feared lest Petilius, that private trial finished or postponed, should return to the council with the others --- so hurrying jumps from his seat, and condemns an innocent man (acquitted by Gaius Sacerdos) without his case being heard, on the opinion of his secretary, his physician, and his haruspex.

\ciceroSection{76} Keep, keep this man in the state, gentlemen, spare and preserve him, that there may be one to judge cases with you, who may give his vote in the senate without any greed about war and peace. Although that is less for us, less for the Roman people to labour over --- of what kind his vote in the senate will be. For what authority will be his? When will he dare or be able to give his opinion? When will a man of so great luxury and idleness, except in the month of February, breathe upon the senate-house? But let him by all means come, decree war on the Cretans, set the Byzantines free, name King Ptolemy, say and feel everything Hortensius wishes; these things less concern us, less the crisis of our life, less the danger of our fortunes.

\ciceroSection{77} This is capital, this is fearful, this is to be feared by every best man: that Verres, if by some force he should snatch himself from this trial, must necessarily be among the jurors, must give his vote on the life of a Roman citizen, must be in the army of the man who wishes to hold the command of the courts as standard-bearer. The Roman people refuses this; cannot bear it. It cries out and gives you leave that, if you delight in such men, if from that kind of man you wish to set up the splendour of your order and the ornament of the senate-house, by all means have him with you as senator, even, if you please, as juror over yourselves; but the men outside that order, on whose lives the splendid Cornelian laws give the power of rejecting no more than three jurors --- they refuse to have this so cruel, so wicked, so unspeakable a man as judge.

\ciceroSection{78} For if it is shameful --- which to me of all things seems most disgraceful and most wicked --- to take money for a judgement, to hold one's faith and conscience put up at a price, how much more shameful, dishonest, unworthy is it to condemn the man from whom you have received money for acquittal --- so that not even a praetor of brigands keeps the customs of faith! It is a crime to receive from a defendant; how much more from a prosecutor; how much more wickedly from both! When you had set out your faith for sale in the province, more weighty with you was he who gave the greater money. I grant it; perhaps someone has at some time done such a thing. But when you have received money and held your faith and conscience now mortgaged to one, you afterwards hand the same to the adversary for greater money: you betray both, and you hand it to whomever you wish, and to him whom you have deceived you do not even return the money?

\ciceroSection{79} What Bulbus, what Staienus do you bring me? What such monster or prodigy have we ever heard or seen --- who transacts with the defendant, then settles with the prosecutor, sends away honourable men who knew the case and dismisses them from his council, and alone condemns the defendant whom he had acquitted, from whom he had received money, and does not return the money? Shall we have this man in the number of jurors? Shall this man hold the second senatorial decury as judge? Shall this man judge concerning a free man's life? Shall the judicial tablet be entrusted to him --- which he will mark not only with wax, but, if he sees fit, with blood?

\ciceroSection{80} For what of these things does he deny he did? That one thing only, what is necessary --- that he received money. Why should he not deny it? But the Roman knight who defended Sopater, who was present at all his counsels and matters, Quintus Minucius, says under oath that money was given; says under oath that Timarchides said greater money was being given by the prosecutors. Many Sicilians will say this; all the Halicyenses will say it; even the praetexta-clad son of Sopater will say --- who was deprived by this most cruel man of a most innocent father and his ancestral money.

\ciceroSection{81} But if I could not by witnesses make plain the matter of money, could you deny --- or will you now deny --- that, with your council dismissed, with the leading men who had been on Gaius Sacerdos's council and used to be on yours removed, you gave judgement on a matter already judged? And that the man whom Gaius Sacerdos with his council, the case being heard, had acquitted, the same man you, your council removed, the case not heard, had condemned? When you confess these things, which were done in the open Forum at Syracuse before the eyes of the province --- deny then by all means, if you wish, that you took money. You will find someone, I imagine, who, when he sees what was done in the open, asks what you did secretly --- or who doubts whether to believe my witnesses or your defenders.

\ciceroSection{82} I have already said that I would not enumerate all his deeds in this kind, but choose those that most stood out. Hear now another famous deed of his, often mentioned in many places, of such a kind that all crimes seem to be in this one. Pay attention diligently; for you will find that crime born from greed, swelled through rape, and finished off and concluded by cruelty.

\ciceroSection{83} Sthenius is here who sits with us, of Thermae --- before, known to many for the highest virtue and nobility, now to all for his own calamity and this man's notable wrong. When Verres had used his hospitality, and had often not only been at his house at Thermae but had even lodged there, he carried off from his house everything which could a little move anyone's heart or eyes. For Sthenius from his youth had with somewhat more zeal acquired these things: a more elegant household furniture in bronze, both Delian and Corinthian; painted panels; even silver of good workmanship --- as the resources of a Thermitan man bore --- enough. Which he had, when a young man in Asia, eagerly (as I said) gathered, not so much for his own pleasure as for the invitation and visits of our men, his friends and guest-friends.

\ciceroSection{84} After Verres had carried all this off --- some by asking, some by demanding, some by taking --- Sthenius bore as he could. He was necessarily anguished in soul that this man had now made his house, adorned and furnished, naked and empty. Yet he shared his grief with no one. He thought a praetor's wrongs should be borne in silence, a guest's calmly.

\ciceroSection{85} Meanwhile that man, with that greed of his known and spread among all, when he had seen at Thermae certain most beautiful and most ancient statues set up in public, fell in love. He began to ask Sthenius to lend his service for taking them away and to help. But Sthenius not only refused, but even showed that it could in no way come about that the most ancient statues, monuments of Publius Africanus, should be taken from the town of the Thermitans, while the city was safe and the empire of the Roman people stood.

\ciceroSection{86} For, that you may at the same time also learn Africanus's humanity and equity --- the Carthaginians had once captured the town of Himera, which had been among the foremost distinguished and adorned cities of Sicily. Scipio --- who thought it worthy of the Roman people, the war finished, that the allies should recover their own through our victory --- with Carthage taken, took care to restore to all the Sicilians what he could. Of those whom the calamity of war had left of the citizens of destroyed Himera, they had settled themselves at Thermae, in the same boundaries of land and not far from the old town; and they reckoned themselves as recovering their fathers' fortunes and dignity when those ornaments of their ancestors were placed in their town.

\ciceroSection{87} There were several bronze statues; among them, of singular beauty, was Himera herself, formed in a woman's figure and dress from the name of the town and the river. There was also a statue of the poet Stesichorus, an old man stooped with a book, finished (as men think) with the highest art --- who was of Himera, but is and was throughout all Greece in highest honour and name for his ability. These Verres had coveted to madness. Even (which I almost passed over) there is a certain little she-goat, marvellously --- as even we who are unskilled in these things can understand --- skilfully and gracefully made. These and other things Scipio had not carelessly thrown away, for an intelligent man like Verres to take, but had restored to the Thermitans --- not because he himself had no garden or suburban estate or any place at all where to put them, but because, if he had taken them home, they would not long have been called Scipio's, but those of whomever they came to by his death. As they are now placed in those places, they always seem to me Scipio's, and so are called.

\ciceroSection{88} When Verres demanded these things, and the matter was being discussed in the senate, Sthenius vehemently resisted, and (since he is among the foremost copious of the Sicilians in speaking) recalled many things: that to leave the city was more honourable for the Thermitans than to suffer to be taken from the city the monuments of their ancestors, the spoils of enemies, the kindnesses of a most distinguished man, the signs of the partnership of the Roman people and friendship. The minds of all were moved. No one was found who did not say it would be better to die. So this town, almost alone in the world, Verres found, from which he could carry off nothing of this kind --- not by force out of the public, not in secret, not by command, not by favour, not by price. But these greeds of his I will set out in another place. Now I shall return to Sthenius.

\ciceroSection{89} Verres, vehemently angry with Sthenius and inflamed, renounces hospitality with him; emigrates from his house; or rather, leaves --- for he had already moved out before. The most hostile enemies of Sthenius at once invite Verres into their own house, that they might inflame his mind against Sthenius by inventing some lie and accusing. These enemies were Agathinus, a noble man, and Dorotheus, who had Callidama, daughter of that very Agathinus, in marriage. Verres had heard about her, and so preferred to move to Agathinus's son-in-law's. One night had passed, when Verres so loved Dorotheus that you would say everything was in common between them; he so respected Agathinus as some kinsman and kinsman by marriage. Even that statue of Himera he now seemed to despise, because the figure and lineaments of his hostess pleased him much more.

\ciceroSection{90} So he began to urge them to create some danger and fabricate some charge against Sthenius. They said they had nothing they could say. Then Verres openly showed and assured them that they would prove against Sthenius whatever they wished, as soon as they had brought it to him. So they did not procrastinate. They at once summon Sthenius, say public records have been corrupted by him. Sthenius demands that, since his own fellow citizens are proceeding against him about corrupted public records, and the action of that matter is by the laws of the Thermitans --- and since the senate and Roman people had given back to the Thermitans (because they had always remained in friendship and faith) their city, fields, and laws, and Publius Rupilius afterwards had given laws to the Sicilians by senatorial decree, by the opinion of the ten legates, that fellow citizens should proceed among themselves by their own laws, and Verres himself had this in his edict --- that on all these grounds he should refer Sthenius to his own laws.

\ciceroSection{91} That man --- of all men the most fair-minded, the most remote from all greed --- assures him he will take cognizance himself, and orders him to be ready at the ninth hour for the case to be heard. It was not obscure what this wicked and unspeakable man was contriving. For he had not himself concealed it well enough, nor had the woman been able to keep silent. It was understood that he was driving at this: that, when he had condemned Sthenius without any proof and without a witness, then this unspeakable man should take from a noble man, of that age, his own guest-friend, the cruelest punishment by the rods. When this was plain, on the advice of his friends and guest-friends Sthenius fled from Thermae to Rome. He preferred to commit himself to winter and the waves rather than not escape that storm and calamity that hung common over all the Sicilians.

\ciceroSection{92} That trustworthy and diligent man is ready and waiting at the ninth hour, orders Sthenius to be summoned. When he sees he is not present, he begins to burn with grief and to rage with anger. He sends Venus's slaves to Sthenius's house; he dispatches horsemen round his fields and farms. So while he waits to hear something certain, he does not leave the Forum before the third hour of night. Next morning he came down. He calls Agathinus to him. He orders him to speak about the public records against Sthenius in his absence. The case was such that, even with no adversary present, he could not before a hostile judge find anything to say.

\ciceroSection{93} So he merely placed it on record in one phrase that, under Sacerdos as praetor, Sthenius had corrupted public records. Hardly had the man said this when our man pronounces: \textsc{Sthenius is held to have corrupted public records}. And to this the man of Venus adds besides --- in a new way, with no precedent --- \textsc{that for that cause he will exact 500{,}000 sesterces from Sthenius's goods for the Erucine Venus}. And he immediately began to sell his goods. And he would have sold them, had there been the slightest delay in the money's being counted out to him.

\ciceroSection{94} After it had been counted out, he was not content with this iniquity. He pronounces openly from his chair and tribunal: \textsc{If anyone wishes to indict the absent Sthenius on a capital charge, he will receive his name}. And at the same time he began to urge Agathinus, his new kinsman by marriage and his guest, to take up the case and lodge the charge. Then Agathinus, in everyone's hearing, said clearly that he would not do it; that he was not so far Sthenius's enemy as to claim that he was answerable for a capital crime. At this point a certain Pacilius suddenly steps forward --- a needy and worthless man --- and says that, if it were lawful, he was willing to lodge the charge against an absent man. Verres says that it is lawful, that it is customary, and that he will receive the name. So the charge is lodged. He at once gives notice that Sthenius is to be present at Syracuse on the Kalends of December.

\ciceroSection{95} Sthenius, having reached Rome, and having sailed (now that the season was against him) reasonably well, and having found everything fairer and gentler than the mind of his praetor and host, laid the matter before his friends. The thing, bitter and unworthy as it was, seemed so to all. So in the senate at once the consuls Gnaeus Lentulus and Lucius Gellius mention it: that it pleased them, if it pleased the conscript fathers, that it be established that absent men should not be made defendants in the provinces. They lay before the senate the whole case of Sthenius and this man's cruelty and iniquity. Verres's father was present in the senate, and weeping, was begging each senator one by one to spare his son. Yet he did not get far with it; for the will of the senate was at the highest. So sentences were being delivered: \textsc{since Sthenius had been made defendant in his absence, no judgement on an absent man should take place; and if any had taken place, that it should not stand}.

\ciceroSection{96} On that day nothing could be transacted, both because the hour was late and because that father of his had found men to consume the time with speaking. Afterwards old Verres met all of Sthenius's defenders and guest-friends; he asks and begs them not to attack his son, not to trouble themselves over Sthenius. He assures them that he will see to it that no harm is done him through his son: that he is sending sure men, for that very purpose, to Sicily by land and sea. And there was a space of nearly thirty days before the Kalends of December, the day on which Verres had given notice that Sthenius should be present at Syracuse.

\ciceroSection{97} Sthenius's friends are moved. They hope that the son will be recalled from his madness by his father's letters and messengers. Afterwards in the senate the case is not pursued. The household messengers come to him; they bring letters from his father before the Kalends of December, when even then to him the whole matter of Sthenius was still open. At the same time several letters on the same matter from many of his friends and connections are brought to him. At this point our man, who out of greed had never reckoned with duty, with danger, with piety, or with humanity --- who thought neither the authority of his father in what he was warned of, nor his father's wish in what he was begged for, ought to come before his lust --- on the morning of the Kalends of December, as he had given notice, orders Sthenius to be summoned.

\ciceroSection{98} Had your father sought this thing of you at the request of some friend, moved by kindness or by the wish to oblige, the will of a father ought still to have weighed most heavily with you. But when he sought it of you for your own life's sake, and had sent sure men from his home, and these had come to you at a time when the whole matter was still open --- not even then could a regard for, if not piety, at least your own safety, recall you to duty and sanity? He summons the defendant; no answer. He summons the prosecutor --- (notice, please, gentlemen, how strongly fortune itself has set herself against this man's madness; and at the same time see what chance came to the aid of Sthenius's case) --- the prosecutor summoned, Marcus Pacilius, by some chance did not answer, did not appear.

\ciceroSection{99} If Sthenius had been made defendant in person, if he had been caught red-handed in his crime, still, the prosecutor not being present, Sthenius ought not to have been condemned. For if a defendant could be condemned in his prosecutor's absence, I should not have come from Vibo to Velia in a tiny boat through the weapons of fugitives and pirates and your own men, at the time when all my haste went with risk to my life --- and for this cause: that you should not be removed from the list of defendants if I did not appear by the day. The thing then which was most desired by you in your own trial --- that I, when summoned, should not be present --- why did you not think it ought to profit Sthenius, when his prosecutor had not appeared? So he saw to it that the end was found most like the beginning: him whom he had made defendant in his absence, that man, in the absence of the prosecutor, he condemns.

\ciceroSection{100} Word was being brought to him in those first days --- as his father too had written him in many words --- that the matter had been agitated in the senate; that even in a public meeting Marcus Palicanus, tribune of the plebs, had complained about Sthenius's case; finally, that I myself, before this college of tribunes of the plebs --- since by their joint edict no man condemned on a capital charge was permitted to be at Rome --- had pleaded Sthenius's case. And that, when I had so set out the matter as I have just now set it out before you, and had shown that this condemnation was not to be admitted, the ten tribunes of the plebs had ruled this, and it had been pronounced from the opinion of all of them: \textsc{Sthenius is not to be held barred by the edict from being lawfully at Rome}.

\ciceroSection{101} When these things were brought to him, he took fright at last and was shaken. He turned his stylus on his own tablets --- by which act he overturned his whole defence. For he left himself nothing that could in any way be defended. For if he had defended himself thus --- "It is lawful to receive the name of an absent man; no law forbids this to be done in a province" --- by an evil and dishonest defence, but at least by some defence, he might have seemed to use one; finally he might have used that most desperate refuge: that he had done it not knowing, that he had thought it was lawful. Although that is the most ruined defence of all, yet something would have seemed to be said. He removes from the tablets what was there, and makes it appear that the case was lodged with him in person.

\ciceroSection{102} Here see in how many snares he has wound himself, from none of which, gentlemen, will he ever extricate himself. First, in Sicily he had himself often and openly said from his bench, and in conversation pointed out to many, that it was lawful to receive the name of an absent man; that he was acting on precedent. That he had repeated this in the previous hearing was said both by Sextus Pompeius Chlorus, of whose virtue I have spoken before, and by Gnaeus Pompeius Theodorus --- a man most approved by the judgement of Gnaeus Pompeius (a most distinguished man) in many and the greatest matters, and most adorned by everyone's esteem --- and by Posides Macro of Solus, a man most adorned in the highest nobility, esteem, and virtue. And in this hearing as many as you wish will say it: men who heard it from this very man, leading men of our order, and others who were present when the name of an absent man was being received. Furthermore at Rome, when this matter was treated in the senate, all this man's friends --- and among them his father --- defended this: that it was lawful for it to be done; that it had often been done; that what this man had done, he had done by the precedent and practice of others.

\ciceroSection{103} Furthermore all Sicily gives testimony, which in the joint petitions of all the cities to the consuls asks and prays the conscript fathers to establish that the names of absent men should not be received. About which matter you have heard Gnaeus Lentulus, the patron of Sicily, that most distinguished young man, say that the Sicilians, when they were instructing him in the case which had to be pleaded for them in the senate, complained of Sthenius's calamity, and on account of this very wrong done to Sthenius had decided that the request which I am speaking of should be lodged.

\ciceroSection{104} When these things stood so, were you possessed of such madness and audacity that, in a matter so plain, so witnessed-to, so spread abroad by yourself, you dared corrupt the public records? But how did you corrupt them? Was it not so that, even were all of us silent, your own tablets could condemn you? Pass the codex to me, please; pass it round; show it. Do you see that this entire entry, where it makes the case lodged in person, is on a place of erasure? What was written there before? What error did that erasure correct? What proofs of this charge, gentlemen, do you await from me? I say nothing. The tablets are in the open, which cry out that they have been corrupted and erased.

\ciceroSection{105} Out of these matters too do you trust that you can escape, when we pursue you not by uncertain conjecture but on your own footprints, which you have left expressed and fresh in the public records? Did this man also pass judgement on Sthenius for corrupting public records, the case unheard --- this man who will not be able to defend himself from having corrupted the public records on Sthenius's very entry?

\ciceroSection{106} See now another madness; see how, while he wishes to extricate himself, he wraps himself in. He registers a representative for Sthenius --- whom? Some kinsman or near relation? No. Some honourable and noble man of Thermae? Not even that. But some Sicilian in whom there was some splendour and dignity? No one. What then? A Roman citizen. To whom can this be made probable? When Sthenius was the most noble of his city, of the widest kinship and the most friendships, and besides could do much through all Sicily by his authority and favour --- could he find no Sicilian who would act as his representative? Will you make this credible? Or did he himself prefer a Roman citizen? Tell me of any Sicilian, when he was made defendant, for whom a Roman citizen was ever made representative. Bring out and unfold the records of all the praetors who were before you. If you find one such, I will grant that the thing was done as you have it written in your tablets.

\ciceroSection{107} But, I suppose, Sthenius thought it splendid for himself to choose, out of the number of Roman citizens, out of the abundance of his friends and guest-friends, the man whom he would give as representative. Whom did he pick? Whose name is written in the tablets? Gaius Claudius, son of Gaius, of the Palatine tribe. I do not ask who this Claudius is; how splendid, how honourable, how fitted, that on account of his authority and dignity Sthenius should depart from the practice of all Sicilians and give a Roman citizen as representative. None of these things I ask; for perhaps Sthenius followed not the man's splendour but their familiarity. What then? If of all mortals no one was more an enemy to Sthenius than this Gaius Claudius, both always and especially in these very matters and at these times --- if he came against him on the charge of corrupted records, if he fought against him in every way --- shall we believe rather that an enemy was made representative on Sthenius's behalf, or that you abused the name of his enemy to bring Sthenius into peril?

\ciceroSection{108} And lest anyone perhaps doubt of what kind this whole business is --- though I trust this man's wickedness has long been clear to all --- yet attend a little further. You see that man with the curly hair, dark, who looks at us with a face such that he seems to himself most sharp-witted, who holds the tablets, who writes, who prompts, who is closest to him. That is Claudius --- the man who in Sicily was reckoned this man's go-between, his interpreter, his finisher of business, almost a colleague of Timarchides. Now he holds such a place that he scarcely seems to yield to that famous Apronius in familiarity --- to him who used to say himself that he was the colleague and partner not of Timarchides but of Verres himself.

\ciceroSection{109} Doubt still, if you can, that this man chose him out of all, in particular, on whom to fasten this dishonest part of false representative --- the man he reckoned both most hostile to Sthenius and most friendly to himself. Will you hesitate, gentlemen, to avenge such audacity, such cruelty, such wrong as this? Will you hesitate to follow the precedent of those judges who, when Gnaeus Dolabella had been condemned, rescinded the condemnation of Philodamus of Opus, not because he had been made defendant in his absence (which is the most iniquitous and bitterest of things), but because, at the time, an embassy to Rome had already been given him by his fellow citizens? What those judges decided in a much lighter case, following equity, will you hesitate to decide in the gravest case --- and one already confirmed by the authority of others?

\ciceroSection{110} But what man, Gaius Verres, have you afflicted with so great and so notable a wrong? What man have you, in his absence, condemned for corrupting the records, the case unheard? Whose name in his absence have you received? What man in his absence have you condemned not only without crime and without witness, but even without prosecutor? What man? Immortal gods! I do not say your friend, which is among men a most distinguished thing; I do not say your guest-friend, which is the most sacred. For there is nothing I less willingly bring up about Sthenius; I find in him no other thing that could be reproached, except that this most frugal and most upright man invited you, a man full of debauchery and disgrace and crime, into his house --- except that, having been and being the guest-friend of Gaius Marius, Gnaeus Pompeius, Gaius Marcellus, Lucius Sisenna (your defender), and the rest of the bravest men, he added your name too to that number of most distinguished men.

\ciceroSection{111} Therefore of the violation of guest-friendship and of that unspeakable crime of yours I make no complaint. I say this not to those who know Sthenius --- that is, to none of those who have been in Sicily, for no one is ignorant in what splendour he stands in his own city, and in what dignity and esteem among all the Sicilians --- but that even those who have not been in that province may understand on what kind of man you have set so great a precedent: a thing which, both for the iniquity of the matter and the dignity of the man, has seemed bitter to all and not to be borne.

\ciceroSection{112} Is not Sthenius the man who, having most easily attained all offices in his own town, discharged them most amply and most magnificently? Who adorned a town not the largest with the largest public spaces and monuments out of his own money? Of whose services to the commonwealth of the Thermitans and to all the Sicilians a bronze tablet was fixed at Thermae in the senate-house, on which his benefits were inscribed and engraved on behalf of the city --- a tablet which was then torn down by your command, but which I have now brought away, that all may know in what honour and amplitude he stood among his own?

\ciceroSection{113} Is not he who, before Gnaeus Pompeius (a most distinguished man), when he was accused (his enemies and prosecutors saying that on account of his familiarity and guest-friendship with Gaius Marius he had taken thought against the commonwealth), and when he was hauled in on a charge more invidious than true, was so acquitted by Gnaeus Pompeius that in that very trial Pompeius judged him most worthy of his own guest-friendship? Praised, moreover, and defended by all the Sicilians, so that the same Pompeius reckoned that he was earning gratitude not only from one man but from a whole province by his acquittal? In short, is not he the man whose mind was such towards the commonwealth, and who could do so much by his authority among his fellow citizens, that he alone in Sicily, with you as praetor, accomplished what no Sicilian --- no, not all Sicily --- could have accomplished: that from the town of Thermae you should touch no statue, no ornament, nothing from sacred things, nothing from public ones --- and this when there were many splendid things and you had coveted them all?

\ciceroSection{114} Then now look at what stands between you --- in whose name festal days are celebrated among the Sicilians and the famous \textit{Verria} are held, to whom gilded statues stand at Rome, given (as we see written on them) by Sicily in common --- look, I say, at what stands between you and this Sicilian, who has been condemned by you, the patron of Sicily. Very many cities in Sicily praise this man publicly by their testimony and by embassies sent for the purpose. You, the patron of all the Sicilians, the one Mamertine city --- ally of your thefts and disgraces --- praises publicly: yet in such a new fashion that the envoys wound while the embassy praises. The other cities all publicly, by letters, embassies, testimonies, accuse, complain, indict. If you should be acquitted, they think themselves utterly overthrown.

\ciceroSection{115} About this man and out of his property, even on Mount Eryx, you set up a monument of your thefts, your disgraces, your cruelty, on which the name of Sthenius of Thermae was inscribed. I have seen the silver Cupid with the lamp. What proof or reason was there in the thing, that on that of all places the prize taken from Sthenius should be set? Did you wish this to be a sign of your greed, or a trophy of intimacy and guest-friendship, or a mark of love? This is what such men do --- those whom in the height of their wickedness not only lust and pleasure delight, but even the very fame of the wickedness itself: that they wish in many places marks and traces of their disgraces to be left.

\ciceroSection{116} He was burning with love for that hostess on whose account he had violated the rights of guest-friendship. This he wished not only to be known then, but always to be remembered. So out of that very business which he had carried through with Agathinus prosecuting, he resolved that the prize should be owed to Venus above all --- the goddess who had set ablaze that whole prosecution and trial. I should have thought you grateful to the gods, had you given this gift to Venus not from Sthenius's goods but from your own, which you ought to have done, especially as in that very year an inheritance had come to you from Chelidon.

\ciceroSection{117} Here, gentlemen, had I not taken up this case at the request of all the Sicilians; had not the whole province demanded this gift of me; had not my mind and love for the commonwealth and the offence given to the standing of our order and of the courts compelled me to do it; and had this been the one cause --- that you had treated my friend and guest-friend Sthenius (whom in my quaestorship I had loved singularly, of whom I had thought best, whom in the province I had known to be most zealous and most desirous of my own standing) so cruelly, so wickedly, so unspeakably --- yet the cause would have seemed worthy enough that I should take up the enmities of the most dishonest of men, to defend the safety and fortunes of a guest-friend.

\ciceroSection{118} Many among our ancestors did this; and lately Gnaeus Domitius, a most distinguished man, did it, who accused Marcus Silanus (a man of consular rank) on account of the wrongs done to his guest-friend Aegritomarus, a Transalpine. I should have thought myself a fit man to follow such an example of humanity and duty, and to set out a hope for my own guest-friends and connections, by which they should reckon themselves to live the safer for my protection. But when in the common wrongs of the whole province Sthenius's case too is contained, and when many of my guest-friends and friends are at one time defended by me both publicly and privately, surely I ought not to fear that anyone will judge what I am doing as undertaken from any motive but compulsion by the highest sense of duty. And, that we may at last leave off speaking of the matters Verres heard and judged, and of the trials he conducted; and since this man's deeds in these kinds are without limit, that we may make some measure and end of our speech and our charges, we shall take a few from other kinds.

\ciceroSection{119} You have heard Quintus Varius say that for the administration of justice his agents gave this man 130,000 sesterces. You remember Quintus Varius's testimony, and that this whole matter is confirmed by the testimony of Gaius Sacerdos, a most distinguished man. You know that Gnaeus Sertius and Marcus Modius, Roman knights, and besides six hundred Roman citizens and many Sicilians, said that they gave this man money for the administration of justice. Why should I dispute about this charge, when the whole matter rests on witnesses? Why should I argue further, when no one can be in doubt about it? Or will any of all of you doubt that the man who at Rome had sold his whole edict and all his decrees had also kept the administration of justice for sale in Sicily? And that he took money from the Sicilians for handing down his decrees, who demanded money from Marcus Octavius Ligus for the administration of justice?

\ciceroSection{120} What kind of money-getting, then, has this man passed by? What that has been passed by by all others has he not devised? Is there any thing sought after among the Sicilian cities --- in which there is any honour, any office, any administration --- that you have not turned to your own gain and the auctioning off of men? Testimonies were given in the previous hearing both privately and publicly. Envoys of Centuripa, Halaesa, Catina, and Panhormus spoke; of many cities besides; and now privately, very many. From whose testimonies you could learn that throughout all Sicily for three years no man in any city was made senator gratis, none --- according to their laws --- by votes, none except by this man's command or letter; and that in the appointment of all these senators not only were no votes taken, but not even were the categories regarded out of which it was lawful to be co-opted into that order: neither census, nor age, nor the rest of the Sicilians' rights had any force.

\ciceroSection{121} Whoever wished to be made senator --- however boyish, however unworthy, from however unlawful a place --- if he won as good enough at this man's price, was at once made one. Not only had the laws of the Sicilians no force in this matter, but not even those given by the senate and Roman people. For the laws which a man holding command from the Roman people, and authority for giving laws from the senate, gives to allies and friends --- these ought to be reckoned the laws of the Roman people and of the senate.

\ciceroSection{122} The Halaesians, in their own right, on account of their many great services and benefits to our commonwealth (their own and their ancestors'), recently --- in the consulship of Lucius Licinius and Quintus Mucius --- when they had disputes among themselves over the co-optation of the senate, sought laws from our senate. The senate by an honourable senatorial decree decreed that Gaius Claudius Pulcher, son of Appius, the praetor, should write up laws for them on the co-optation of the senate. Gaius Claudius, having called in all the Marcelli who were then living, by their advice gave the Halaesians laws, in which he laid down many things about the age of men, that no one under thirty years of age; about livelihood, that no one who had practised one should be chosen; about the census; about the rest. All of which, before this man as praetor, both by the authority of our magistrates and by the strongest will of the Halaesians, held good. Under this man, both the auctioneer (whoever wished) bought that order at a price, and boys of sixteen and seventeen years had the senatorial title sold to them by retail. And what the Halaesians, most ancient and most faithful allies and friends, had obtained at Rome --- that not even by votes should it be lawful to be made senator among them --- this man brought it about that for a price it could be done.

\ciceroSection{123} The Agrigentines have for the co-optation of the senate the ancient laws of Scipio, in which the same things are laid down, and this besides: since there are two stocks of Agrigentines, one of the old, the other of the colonists whom the praetor Titus Manlius led to Agrigentum out of the towns of the Sicilians by senatorial decree, it is provided in Scipio's laws that there should not be more in the senate from the colonists' number than from the old Agrigentines. This man, who had levelled all rights at a price and removed every choice and distinction in things by money, not only mixed up those rules which had to do with age, with rank, with livelihood, but even threw into confusion these two classes of citizens, the new and the old.

\ciceroSection{124} For when a senator from the old number had died, and an equal number remained from each class, by the laws an old one had to be co-opted, that this number should be the larger. Yet, when the matter stood thus, men came nevertheless to buy that senatorial place from him --- not only the old, but the new too. It happens that a new man wins by his price, and brings letters from the praetor to Agrigentum. The Agrigentines send envoys to him to instruct him in the laws and to set out the custom of all the years, that this man might understand he had sold a place to one who ought not even to have been allowed to bid for it. By their speech, since he had now received his price, this man was not moved a fraction.

\ciceroSection{125} The same he did at Heraclea. For there too Publius Rupilius led colonists, and gave like laws on the co-opting of the senate and on the number of old and new. There not only did this man, as among the rest, accept money, but he also threw into confusion the classes of old and new and their numbers. Do not wait for me to go through all the cities in my speech: with this one thing I take in all --- under this man as praetor, no one could be made senator who had not given this man money.

\ciceroSection{126} The same I transfer to magistracies, public charges, and priesthoods. In which matters he repudiated not only the rights of men but all the religious observances of the immortal gods. At Syracuse there is a law on religion which orders the priest of Jupiter to be taken yearly by lot --- which among them is reckoned the most ample priesthood:

\ciceroSection{127} when by votes three men have been chosen out of three classes, the matter is referred to the lot. This man had brought it about by his command that, in lieu of the vote, Theomnastus, his familiar, should be returned in those three. As to the lot, which he had not been able to command, men were waiting to see what he would do. The man --- which was easiest --- first forbids them to draw lots; orders Theomnastus to be returned outside the lot. The Syracusans say that this can in no way be done, by reason of the religious laws of the rites; in short, they say it is not divine right to do so. He orders the law to be read out to him. It is read; in which it was written that, however many had been returned, that many lots should be cast into the urn; whose name had come out, that man should hold the priesthood. This ingenious and most acute man says, "Excellent! It is so written, is it not, \textsc{however many shall be returned}? How many, then," he says, "have been returned?" They answer, "Three." "Need then any but three lots be cast in, and one drawn?" "None." He orders three to be cast in, on all of which was inscribed the name of Theomnastus. There is the loudest outcry, since to all this seemed unworthy and unspeakable. So that most ample priesthood of Jupiter is given to Theomnastus by this device.

\ciceroSection{128} At Cephaloedium there is a fixed month in which the chief priest must be appointed. There was a certain Artemo, surnamed Climachias, eager for that honour, a man certainly well-off and noble at home. But he could in no way be made priest if a certain Herodotus were present: that place and honour was thought so owed to him for that year that not even Climachias spoke against it. The matter is brought to this man and is settled in his usual way: certain well-known and precious embossed silver pieces are carried off. Herodotus was at Rome; he reckoned that he would come in good time to the elections if he came the day before. This man --- so that the elections should not either be held in another month than divine right allowed, or, with Herodotus present, the honour be taken from him (which this man did not care about, but Climachias least of all wanted) --- contrives, as I said long ago, no man is or has been more acute, contrives, I say, by what means the elections might be held in that lawful month and yet Herodotus might not be able to be present.

\ciceroSection{129} It is the custom of the Sicilians and the rest of the Greeks, because they wish their days and months to agree with the reckoning of the sun and moon, sometimes, if anything is out, to take out one day, or at most two, from a month, which they call \textit{exairesimoi} days; in the same way sometimes they make a month a day or two longer. When this man, the new astrologer --- who reckoned not so much the system of the heavens as of embossed silver --- had learned of this, he orders to be taken out, not a day from a month, but from the year one whole-and-a-half month, on this principle: that on the day on which (let us say) the Ides of January ought to have fallen, on that very day he ordered the Kalends of March to be proclaimed. So it is done, with all refusing and weeping. That day was the lawful one for holding the elections. In this way Climachias was returned as priest.

\ciceroSection{130} When Herodotus was returning from Rome --- as he himself reckoned, fifteen days before the elections --- he meets the month that follows the election month, the elections having now been held thirty days back. Then the Cephaloedians made an intercalary forty-five days long, that the rest of the months should return to their proper reckoning. If this could be done at Rome, surely this man would by some means have contrived to remove the forty-five days between the two sets of games --- which were the only days on which the trial could be held.


