% In C. Verrem Actio Secunda V (in-verrem-2-5) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.
%
% Book V of the Actio Secunda --- de suppliciis --- the closing book of
% the prosecution of Verres: the pretended pirate war, the bribed
% release of pirate captains, the executions of the Roman captains of
% the Sicilian fleet at Achradina, the imprisonment and execution of
% Roman citizens in the Latomiae, and the crucifixion of Gavius of
% Consa --- civis Romanus sum.

\ciceroSection{1} I see no man doubts, gentlemen, that Gaius Verres openly despoiled all things sacred and profane, both privately and publicly, in Sicily, and was occupied in every kind of stealing and plundering with no scruple, indeed not even with any pretence of concealment. But a certain magnificent and distinguished defence is being shown me; how I shall withstand it I must consider long beforehand, gentlemen. For the case is so put: that the province of Sicily, by his virtue and singular vigilance in doubtful and fearful times, was kept safe from runaway slaves and from the perils of war.

\ciceroSection{2} What shall I do, gentlemen? Whither shall I bring the reasoning of my prosecution? Whither shall I turn? For against all my onsets the name of a good general is set as a kind of wall. I know the place; I see where Hortensius will toss himself about. He will recall the perils of war, the times of the commonwealth, the want of generals. Then he will plead with you, then he will even contend, as of his right, that you should not allow such a general to be torn from the Roman people by the testimonies of Sicilians; that you should not wish a general's praise to be ground down by charges of greed.

\ciceroSection{3} I cannot pretend, gentlemen, that I do not fear lest, on account of this singular virtue in military affairs, Gaius Verres should have done with impunity all that he has done. For there comes into my mind, in the trial of Manius Aquilius, how much authority, how much weight Marcus Antonius's speech was reckoned to have had: who, as he was in speaking not only wise but also brave, with the case almost concluded himself snatched up Manius Aquilius and set him in the sight of all and tore his tunic from his breast, that the Roman people and the judges might see the scars taken on his front. At the same time about that wound which Aquilius had received in the head from the leader of the enemy he said much, and brought those who were going to judge to fear vehemently lest the man, whom fortune had snatched out of the weapons of enemies (when he had not himself spared himself), should seem kept not for the praise of the Roman people but for the cruelty of the judges.

\ciceroSection{4} The same method and way of defence is now tried by them; the same is sought. Let him be a thief, let him be sacrilegious, let him be the chief of all disgraces and vices: still, he is a good general, still, fortunate and to be reserved for the doubtful times of the commonwealth. I shall not deal with you on the highest right; I shall not say what I should perhaps establish, since the trial is by a fixed law: that not how bravely you have done in military affairs, but in what manner you have kept your hands from others' moneys must be shown by you. I shall not, I say, deal so. But I shall ask, in the manner I see you wish, what your part has been in the war and how great.

\ciceroSection{5} What do you say? That from the war of the runaways Sicily was freed by your virtue? Great praise and an honourable speech; but yet, in what war? For we, after that war which Manius Aquilius brought to an end, so received it: that there was no war of runaways in Sicily. "But there was in Italy." I confess; and a great and vehement one too. Do you then try to seek any part of praise out of that war? Do you reckon that the glory of that victory has been shared with you by Marcus Crassus or Gnaeus Pompeius? I do not think this is wanting yet to your shamelessness, that you should dare to say anything of that kind. You stood in the way, of course, lest the forces of the runaways could cross from Italy into Sicily. Where, when, from what quarter? When they tried to come either by rafts or by ships? For we have never heard of anything; but we have heard this: that by the virtue and counsel of Marcus Crassus, the bravest man, it was brought about that with rafts joined the runaways could not pass through the strait to Messana. From which attempt they need not so greatly have been hindered, if any defences in Sicily had been thought to be set up against their coming.

\ciceroSection{6} But, although there was war in Italy so near to Sicily, yet in Sicily there was none. What wonder? Not even when there was in Sicily, at the same distance, did any part of that war pass over into Italy. For the nearness of places is brought forward in this matter to which point? Whether that the approach was easy to the enemy, or that the contagion of imitating the war was perilous? Every approach to men without any means of ships is not only sundered but even shut, so that for those to whom you say Sicily was near it would have been easier to come to Ocean than to come to Pelorus.

\ciceroSection{7} But that contagion of the slave war --- why is it proclaimed by you rather than by all those who held the rest of the provinces? Or because in Sicily before there had been wars of runaways? But that very thing is the cause why that province is and has been in the least peril. For after Manius Aquilius departed thence, the institutions and edicts of all the praetors were such that no slave should be with a weapon. The thing I shall tell is old, and on account of the strictness of the example perhaps not unheard by any of you: that Lucius Domitius, when he was praetor in Sicily, when a great boar had been brought to him, in wonder asked who had killed it. When he had heard that he was a certain shepherd, he ordered him to be summoned to him. The man hurried eagerly to the praetor as if for praise and reward. Domitius asked who had struck so great a beast. He answered, with a hunting-spear; immediately, by the praetor's order, he was lifted up on the cross. Perhaps this seems harsh; nor do I argue either way. I only see this: that Domitius preferred to seem cruel in animadverting than slack in passing over.

\ciceroSection{8} So with these institutions of the province even when in the war of the allies all Italy was ablaze, a man not most fierce nor most brave, Gaius Norbanus, was in the highest leisure. For Sicily then was easily defending herself, that no war could arise from herself. For when nothing is so joined as our businessmen with the Sicilians by use, by matter, by reason, by concord; and since the Sicilians themselves have their affairs so set up that peace is to their interest, but love the empire of the Roman people so that they would by no means wish it to be lessened or changed; and since these dangers from a slave war have been provided against both by the institutions of praetors and by the discipline of masters --- there is no domestic ill that can arise from the very province.

\ciceroSection{9} What then? Are no movements of slaves under Verres as praetor said to have happened? No conspiracies? Nothing indeed that came to the senate or to the Roman people; nothing which he himself wrote publicly to Rome. And yet I suspect that in some places in Sicily it began to be stirred up. This I learn not so much from the matter as from his own deeds and decrees. And see with what no hostile mind I shall act: I myself shall recall and bring forth those things which he seeks, which up to now you have never heard.

\ciceroSection{10} In the territory of Triocala, a place which the runaways once held before, the household of a certain Sicilian Leonidas was called into suspicion of conspiracy. The matter is reported to him. At once, as was fitting, by his order the men who had been named were arrested and brought to Lilybaeum. Notice was given to the master; the case was pleaded; they were condemned. What then? What do you suppose? Theft perhaps, or some plunder, you await. Do not always look for the same thing. In the fear of war what place can there be for stealing? Even if there was, in this case the chance was passed over. Then he might have taken some money from Leonidas, when he gave him notice to be at hand; there was some bargaining, and not new for him, that they should not plead. There was also another place: that they should be acquitted. But with the slaves once condemned, what reckoning can there be of plunder? They must be brought out for punishment. For there are witnesses who were on the council; the public records are witnesses; the most splendid city of Lilybaeum is witness; the most honourable and largest assize of Roman citizens is witness: nothing can be done; they must be brought out. So they are brought out and bound to the stake.

\ciceroSection{11} Even now, gentlemen, you seem to me to await what was done next, since this man never did anything without some gain and plunder. What in such a matter could be done? Whatever you please, await any deed however dishonest. I will yet outrun the expectation of all. Men condemned of crime and conspiracy, handed over for punishment, bound to the stake, suddenly, with many thousands of men looking on, were loosed and given back to that Triocalan master. What can you say in this place, most senseless of men, save what I do not ask: which finally in a thing so unspeakable, although there can be no doubt, yet, even if there were doubt, ought not to be asked: what or how much or in what manner you took? I leave the whole of this to you and free you from that care. For I do not fear that anyone will be persuaded that you, into a deed to which no one but you could have been led by any sum of money, tried to enter freely. But of that method of stealing and plundering I say nothing: of this generalship's praise of yours I am now disputing.

\ciceroSection{12} What say you, good guardian and defender of the province? Did you dare to snatch from the very midst of death and free those slaves whom you knew had wished to take up arms and make war in Sicily, and whom from the opinion of the council you had judged --- now handed over to punishment by the custom of our ancestors --- that the cross which you had fixed for the condemned slaves you might keep for the unsentenced Roman citizens? Ruined states with all things now despaired of are wont to have these deadly outcomes: that the condemned be restored entire, the bound be loosed, the exiles be brought back, the things judged be rescinded. When these happen, no one fails to understand that the commonwealth is rushing where these things come up; no one thinks that any hope of safety is left.

\ciceroSection{13} And these things, if anywhere they have been done, have been done in such a way that popular men or noblemen were freed from punishment or exile; but not by those very men who had judged, not at once, not those condemned of those deeds which had to do with the life and the fortunes of all. But this is new and of such a kind that it seems credible more on account of the defendant than on account of the matter itself: that he should let go slaves; that he himself, who had judged, should at once let them go from the very midst of punishment; that he should let go slaves condemned of that crime which had to do with the life and blood of all free men.

\ciceroSection{14} O distinguished general, no longer to be compared with Manius Aquilius, that bravest of men, but in truth with the Pauli, the Scipios, the Marii! That he should have foreseen so much in the fear and danger of the province! When he saw the minds of the slaves in Sicily on edge on account of the war of the runaways in Italy, lest any should dare to stir himself, what fear did he throw in! He ordered them to be arrested. Who would not be terrified? The masters to plead the case. What for the slave so dreadful? \textsc{They are seen to have done it} --- so he proclaims. He seems to have put out a flame springing up by the grief and death of a few. What follows next? Floggings and fires and those last things appointed for the punishment of the condemned, the fear of the rest, tortures and the cross. From all these punishments they were freed. Who can doubt that he pressed down the spirits of the slaves by the highest fear, when they saw that praetor of such ease that with him the life of slaves condemned of crime and conspiracy could be ransomed even with the executioner himself as go-between?

\ciceroSection{15} What? In the case of Aristodamus of Apollonia, what? In that of Leon of Imachara, did you not do the same? What? That movement of slaves and that sudden suspicion of war --- did it bring you a diligence in guarding the province, or a new method of most dishonest gain? When the bailiff of Eumenides of Halicyae --- a noble and honourable man, a wealthy man --- had been accused at your prompting, you took 60,000 sesterces from his master, who lately himself sworn taught in what manner the thing was done. From the Roman knight Gaius Matrinius, in his absence, when he was at Rome, because you had said that his bailiffs and shepherds had come into your suspicion, you carried off 600,000 sesterces. This Lucius Flavius said --- the agent of Gaius Matrinius, who counted out that money to you; Matrinius himself said it; the most distinguished man, Gnaeus Lentulus the censor, says it, who, for the sake of Matrinius's honour, in the recent business sent letters to you and saw to their being sent.

\ciceroSection{16} What? About Apollonius, son of Diocles, of Panhormus, whose surname is the Twin --- can it be passed over? Is anything more famous than this in all Sicily, anything more unworthy, anything more manifest can be brought forth? Whom, when he came to Panhormus, he ordered to be summoned to him and called from the tribunal at the great gathering and frequency of the assembly. Men at once began to say: "I wondered that Apollonius, a wealthy man, should have remained for so long untouched by him; he has devised something I know not what, has brought something I know not what; surely a rich man is not without cause suddenly summoned by Verres." There was the highest expectation of all what it was, when, without breath, suddenly Apollonius himself ran up with his young son. For the father, an old man, had long been kept in bed.

\ciceroSection{17} This man names a slave whom he said was the master of the herd; says he conspired and stirred up the households (there was no slave at all by that name in the household). He orders him to be at once produced. Apollonius affirmed that he had no slave at all by that name. This man orders the man to be snatched from the tribunal and thrown into prison. He cried as he was being snatched away that he, wretched, had done nothing, had committed nothing; that his money was in loans, that he did not have it ready in cash. While he was protesting this, in the highest gathering of men, in such a way that anyone could understand that, because he had not given money, he was therefore being afflicted with that bitter wrong --- while, as I say, he was crying out about money, he was thrown into chains.

\ciceroSection{18} See the consistency of the praetor, and a praetor who in these matters is to be defended not as a middling praetor but to be praised as the best general. When a slave war was being feared, with what punishment he was afflicting masters unsentenced, with that he was freeing condemned slaves. Apollonius, a most well-off man, who, if the runaways should make war in Sicily, would lose the most ample fortunes, he threw into chains under the name of the runaways' war, the case unheard. The slaves whom he himself by the opinion of the council had judged to have agreed for the sake of making war, those, without the council's opinion, of his own accord he freed from every punishment.

\ciceroSection{19} What? If something was committed by Apollonius for which he should rightly have been animadverted upon, shall we yet so handle this matter that we should think the defendant ought to be the subject of charge or hatred if he should have judged some man strictly? I shall not deal so bitterly; I shall not use that prosecutor's habit, that, if anything has been done leniently, I should charge it as done loosely; if anything has been punished strictly, I should out of that gather a hatred of cruelty. I shall not deal in that fashion. I shall follow your own judgements; I shall defend your authority, as long as you wish. As soon as you yourself begin to rescind your own judgements, cease to be angry with me; for by my own right I shall contend that he who has been condemned by his own judgement ought to be condemned by the votes of jurors under oath.

\ciceroSection{20} I shall not defend the case of my friend and guest-friend Apollonius, lest I should seem to rescind your judgement. I shall say nothing about the man's thrift, virtue, diligence. I shall pass over even what I said before: that his fortunes were so set up in households, in cattle, in farms, in lent moneys, that nothing was more disadvantageous to anyone than that any tumult or war in Sicily should be stirred up. I shall not even say this: that, although Apollonius were most in fault, yet against the most honourable man of a most honourable city it ought not have been animadverted so heavily, his case unheard.

\ciceroSection{21} I shall stir up no hatred against you even by these things: that, though such a man was in prison, in darkness, in squalor, in filth --- by your tyrannical interdicts the father, an old man, and his young son were never given the power of approaching that wretched man. Even this I shall pass over: how often you came to Panhormus in that year and six months --- for so long Apollonius was in prison --- so often you came as suppliant before the senate of Panhormus, with the magistrates and public priests, asking and beseeching that that wretched and innocent man might at last be freed from that calamity. I leave aside all these things, which, if I should wish to pursue, I should easily show that by your cruelty against others all approaches of pity from the judges have long since been closed off to you.

\ciceroSection{22} All these I shall grant you and remit. For I foresee what Hortensius is about to defend. He will confess that with this man neither the old age of the father nor the youth of the son nor the tears of both prevailed more than the advantage and safety of the province. He will say that the commonwealth cannot be administered without fear and strictness. He will ask why the fasces are carried before praetors, why axes have been given them, why a prison has been built, why so many punishments have been established against the dishonest by the custom of our ancestors. When he has said all these things gravely and strictly, I shall ask why this same Verres ordered this same Apollonius suddenly, with no new thing brought forward, with no defence, without cause, to be let out of prison. I shall affirm that there is so much suspicion in this charge that I shall now allow the very judges to make a guess without my arguing what kind of plundering this is, how dishonest, how unworthy, and to what magnitude of gain it seems boundless and unlimited.

\ciceroSection{23} For what he did in the case of Apollonius --- learn first briefly how many and how great they are; then weigh and assess these in money. You will find that they were established in one wealthy man so as to set the rest fears of like hurts and examples of perils. First a sudden charge of a capital and odious crime --- determine how much you reckon this and how many bought themselves off from it. Next, a charge with no accuser, a sentence with no council, a condemnation with no defence --- assess the prices of all these things and consider that in these injustices Apollonius alone stuck, that the rest, surely many, freed themselves from these hurts by money. Finally darkness, chains, prison, shut-in punishment, and being shut off from the sight of parents and children, finally from free breath and the common light --- these things, indeed, which can rightly even be ransomed with one's life, I cannot assess in money.

\ciceroSection{24} All these things Apollonius too late ransomed, now ruined by sorrow and miseries. But yet he taught the rest to forestall this man's greed and crime. Unless perhaps you think that a most wealthy man was chosen without cause of gain for so incredible a charge, or without the same cause was suddenly let out of prison; or that this kind of plundering was attempted and tried by him in this one alone, and was not through him set out and thrown at all the wealthiest Sicilians.

\ciceroSection{25} I wish, gentlemen, since I am speaking of his military glory, that he should suggest to me, if perhaps I omit anything. For I seem to myself now to have spoken of all his deeds which had to do with the suspicion of a war of runaways. Surely I have not knowingly passed over anything. You have the man's counsels, his diligence, his vigilance, his guardianship and defence of the province. The sum of it points to this: that you should know, since there are several kinds of generals, of which kind he is, lest in such great want of brave men anyone should remain ignorant of such a general. Not to the wisdom of Quintus Maximus, not to the speed in carrying through affairs of that earlier Africanus, not to the singular counsel of him who was afterwards, not to the reckoning and discipline of Paulus, not to the force and virtue of Gaius Marius --- but a different kind of general, surely diligently to be retained and kept, learn, please.

\ciceroSection{26} First the toil of journeyings, which is the greatest in military affairs, gentlemen, and most necessary in Sicily, hear in what manner he made easy and pleasant for himself by reasoning and counsel. First, in winter times against the size of the cold and the force of the storms and the floods, he had prepared this distinguished remedy for himself. He had chosen the city of Syracuse, whose situation and the nature of whose place and sky is said to be such that no day was ever of so great and so troubled a storm but at some moment of that day men have seen the sun. Here this good general so lived in the winter months that no one easily saw him not only out of the house, but not even out of the bed. So the shortness of the day was filled with banquets, the length of the night with debaucheries and disgraces.

\ciceroSection{27} But when spring had begun --- whose beginning he marked not from the west wind nor from any star, but, when he had seen a rose, then he reckoned spring beginning --- he gave himself to labour and to journeys. In which he showed himself so patient and tireless that no one ever saw him sitting on a horse. For, as was the custom of the kings of Bithynia, he was carried in a litter borne by eight, in which there was a most translucent Maltese cushion stuffed with rose. He himself wore one wreath on his head, another on his neck, and held to his nostrils a little net of the finest linen, with small meshes, full of rose. So when, the journey thus made, he had come to some town, in the same litter he was carried even to his bedroom. Thither came the magistrates of the Sicilians, came the Roman knights, which you have heard from many under oath. Disputes were brought to him in secret; a little later openly the decrees were carried off. Then, when he had for a little while in his bedroom apportioned out justice not by equity but by price, he reckoned the rest of the time owed to Venus and Liber.

\ciceroSection{28} Here it seems to me must not be passed over the singular and outstanding diligence of the distinguished general. For know that there was no town in Sicily of those towns in which the praetors are wont to be stationed and to hold the assize, in which town there was not for him a woman chosen for his lust out of some not unrenowned family. So some out of that number were brought openly to his banquets; if any were more chaste, they came at the time, avoided the daylight and the gathering. But the banquets were not of that silence of the praetors and generals of the Roman people, nor of that modesty which is wont to be in the banquets of magistrates, but with the loudest shouting and abuse. Sometimes the matter was even carried to a fight and to fists. For this stern and diligent praetor, who had never obeyed the laws of the Roman people, was diligently obeying those laws which were laid down at the cups. So the outcomes were such that one was carried out of the banquet between hands as out of a battle, another was left as if dead, most lay loose without mind and without any sense --- so that anyone who saw it would think he was looking not at the banquet of a praetor, but at the Battle of Cannae of wickedness.

\ciceroSection{29} But when high summer had begun --- a time which all praetors of Sicily have always been wont to spend in journeys, because then they think that the province must above all be gone over, when the grain is on the threshing-floors, because the households are gathered together and the size of the slave-force is seen and the toil of the work most strikes them; the abundance of grain warns them; the time of the year does not hinder --- then, I say, when the rest of the praetors are running about, this man, by a kind of new fashion, was making for himself standing camp at a most beautiful place at Syracuse.

\ciceroSection{30} For at the very entrance and mouth of the harbour, where first from the deep the bay turns from the shore to the city, he was setting up tents stretched with linen sails. Hither out of that praetorian house which had been King Hiero's he so emigrated that no one through those days could see him outside that place. But to that very place there was approach for no one, except who could be either partner or minister of his lust. Hither all the women came with whom he had been familiar --- of whom an incredible number was at Syracuse. Hither came men worthy of his friendship, worthy of that life and those banquets. Among such men and women, his now grown-up son was at large, that, even if nature drew him from a likeness to his father, custom and his father's discipline might yet compel him to be like.

\ciceroSection{31} Hither that Tertia, brought through guile and snares from the Rhodian flute-player, is said to have made the greatest tumults in his camp; when the wife of Cleomenes the Syracusan, a noble woman, took it ill, and likewise the wife of Aeschrio, born of an honourable family, that the daughter of Isidore the mime had come into their company. But this Hannibal, who reckoned that in his own camp the contest ought to be by virtue not by birth, so loved this Tertia that he carried her with him out of the province. And in those days, when he was wandering in a purple cloak and tunic to the ankles in women's banquets, men were not offended, nor took it heavily that the magistrate was away from the forum, that justice was not given, that trials were not held; that the whole place of the shore resounded with the voices of women and the singing of music, that in the forum there was the highest silence of cases and of right --- men did not bear it heavily; for it did not seem that justice was absent from the forum, nor trials, but force and cruelty and the bitter and unworthy plundering of the goods of good men.

\ciceroSection{32} Will you then defend that this is a general, Hortensius? Will you try to cover his thefts, rapines, greed, cruelty, arrogance, crime, audacity by the size of his deeds and by the praises of generalship? Here surely it is to be feared that, at the end of your defence, that old Antonian method and authority of speaking will be brought forth: that Verres should be roused, should be stripped from his breast, that the Roman people should look at the scars --- of the marks of women's bites, the traces of lust and wickedness. May the gods grant that you should dare to make mention of military service, of war!

\ciceroSection{33} For all those old campaigns of his shall be learned, that you may understand of what kind he was not only in command but also in service. That first service shall be revived, when this man was wont to be carried away from the forum, not (as he himself proclaims) led there. The camps of the Placentine gambler shall be recalled, in which, although he was diligent, he was yet stricken from the rolls. Many losses of his in service shall be brought forth, which were dissolved and made up by him from the fruit of his youth.

\ciceroSection{34} Now indeed, when in such patience of foulness he had grown hardened by another's, not by his own, satiety, what kind of man he was, how many garrisons, how protected of modesty and chastity, he took by force and audacity --- why does it concern me to say or to join with this man's disgrace the dishonour of any other besides? I shall not, gentlemen. I shall pass over all the old. I shall set down only two recent ones with no man's infamy, by which you can make a guess about all: the one which was so notorious and known to all, that no man so country-bred came up to Rome from any town in the consulship of Lucius Lucullus and Marcus Cotta on a bail-bond, but knew that all the rights of the city praetor were ruled by the nod and discretion of the little harlot Chelidon; the other, that, when he had set out in the cloak of war and pronounced his vows for his command and the common commonwealth, he was wont by night, for debauchery's sake, to be carried in a litter into the city to a woman married to one, set out for all --- against divine right, against the auspices, against all divine and human religious obligations!

\ciceroSection{35} O immortal gods! What is the difference between the minds and the thoughts of men! So may my own will and the hope of the rest of my life be approved by your standing and that of the Roman people, as I have so received those magistracies which the Roman people has up to now entrusted to me, that I have reckoned myself bound by the religious obligation of all duties! So I was made quaestor that I thought that honour at that time was given me not only, but committed and entrusted; so I held the quaestorship in the province of Sicily that I thought all eyes were turned on me alone; that I reckoned that I and my quaestorship were occupied as if in some theatre of the world; that I always denied to all those things which seem pleasant, not only to those exceptional desires, but even to nature itself and to necessity.

\ciceroSection{36} Now I am aedile-designate. I have a reckoning of what I have received from the Roman people: that I am to make the most sacred games to Ceres, Liber, and Libera with the highest care and ceremony; that I am to appease Mother Flora to the people and the plebs of Rome by the celebration of games; that I am to make the most ancient games which were first called Roman, with the highest dignity and religious sanctity, to Jupiter, Juno, and Minerva; that to me the procuration of the sacred buildings, that to me the guardianship of the whole city has been committed. For the labour and care of these things have been given those rewards: an earlier place of speaking my opinion in the senate, the bordered toga, the curule chair, the right of an image to be handed down to the memory of posterity.

\ciceroSection{37} Out of all these things, gentlemen --- so may all the gods be propitious to me --- although the honour of the people is most pleasant to me, yet I take by no means as much pleasure as concern and toil, that this very aedileship may seem not given to some candidate because it had to be, but, because so it was right, rightly bestowed and set in its place by the people's judgement.

\ciceroSection{38} You, when you were proclaimed praetor in whatever way --- for I leave aside and pass over what was done then --- but when you were proclaimed, as I said, were you not roused by the very voice of the crier, who so many times in the centuries of the elders and the younger pronounced you affected with that honour, so that you should think this: that some part of the commonwealth was entrusted to you, that for one year you were to be away from a harlot's home? When it had fallen to you by lot to administer justice, did you never consider how much business, how much burden you had? Nor did you take account of this --- if perhaps you could rouse yourself --- that the province, which it was hard to keep safe by singular wisdom and uprightness, had come to the highest stupidity and wickedness? So not only did you not wish to shut Chelidon out of your house in the praetorship, but you brought the whole praetorship into Chelidon's house.

\ciceroSection{39} The province followed; in which it never came into your mind that the fasces and axes and so much force of imperium and so great a dignity of all ornaments had not been given you to break, by the force and authority of those things, all the bars of modesty and duty; that you should hold all men's goods as your plunder; that no man's property could be safe, no man's house closed, no man's life hedged in, no man's chastity guarded against your greed and audacity. In which you so conducted yourself that, when you are caught in all things, you flee to a runaway-slave war; out of which you now understand that no defence has arisen for you, but the greatest force of charges. Unless perhaps you bring forward the remains of the Italic war of runaways and that Tempsan misfortune, to which (when fortune lately had brought you most opportunely) you, if you had had any virtue or industry in you, were found to be the same that you had always been.

\ciceroSection{40} When the Valentini had come to you, and on their behalf an eloquent and noble man, Marcus Marius, was speaking, that you should take up the business; that, since the praetorian command and name was in your power, you should offer yourself as leader and chief for putting down that small band, you not only fled it, but at that very time, when you were on the shore, that Tertia of yours, whom you had brought with you, was in the sight of all. To the Valentini themselves, from so distinguished and noble a town, on so great matters, you gave answer when you were in a dark tunic and cloak. What do you suppose this man as he was setting out, what in the very province he did, who, when he was already departing from the province not to a triumph but to a trial, did not even flee that infamy which without any pleasure he was getting?

\ciceroSection{41} O divine murmuring of the throng of the senate in the temple of Bellona! You hold in memory, gentlemen, when it was growing dark and a little before there had been brought news of this Tempsan hurt, when no one was being found who could be sent to those places with imperium, and someone said that Verres was not far from Tempsa, how strongly all murmured together; how openly the chief men spoke against. And, convicted by these so many charges and testimonies, does he set out for himself any hope in the tablet of those by all of whom openly, his case unheard, he has been condemned by voice?

\ciceroSection{42} Be it so. Out of the war of runaways or the suspicion of war he attained no praise, since neither was such a war nor any peril of war in Sicily; nor was it provided by him that there should be none. But indeed against the war of brigands he had a fleet equipped, and singular diligence in it, and so the province was famously defended by him. Of the war of brigands, of the Sicilian fleet, gentlemen, I shall so speak that I shall affirm this in advance: that in this one kind there are all this man's greatest faults --- of greed, of treason, of madness, of lust, of cruelty. While I briefly set these out, please attend, as you have done up to now, diligently.

\ciceroSection{43} I say first that the naval matter was so administered, not that the province should be defended, but that under the name of fleet money should be sought. When the custom of the previous praetors had been such that ships were demanded from the cities, and a fixed number of sailors and soldiers, on the largest and most well-off Mamertine state you commanded none of these things. For this thing what money the Mamertines secretly gave you, afterwards, if it shall seem good, we shall ask out of their own letters and witnesses.

\ciceroSection{44} But I say that a cargo-ship, the largest, like a trireme, most beautiful and most adorned, was openly built at public expense in your name, publicly, with all Sicily knowing it, through the magistrate and the senate of the Mamertines, and given as a gift to you. This ship, laden with Sicilian plunder (although she herself too was out of the plunder), at the same time when he was leaving, was brought in to Velia with very many things --- and those which this man would not send to Rome with the rest of his thefts, because they were most famous and pleased him most. That ship I myself lately saw at Velia and many others have seen, most beautiful and most adorned, gentlemen: which seemed indeed to all who had seen her now to look out for her master's exile and to scout out his flight.

\ciceroSection{45} What will you answer me in this place? Unless perhaps that which, although it can in no way be proved, yet must be said in a trial for extortion: that the ship was built out of your own money. Dare to say this much at least, which is necessary. Do not fear, Hortensius, lest I ask by what right it was lawful for a senator to build a ship; those laws are old and dead (as you are wont to say) which forbid it. There was once that commonwealth, there was that strictness in trials, that the prosecutor reckoned this thing must be thrown back among the great charges. For what need to you of a ship? Who, if you are setting out anywhere on public business, are furnished vessels for protection and conveyance at public expense; in private, however, you can neither set out anywhere nor send for things from across the sea out of those places in which you may have nothing.

\ciceroSection{46} Then why have you procured anything against the laws? This charge would prevail in that old strictness and dignity of the commonwealth. Now I do not even charge you with this; nor even with that common reproach do I reproach you. Did you yourself never think this would be foul to you, never charged, never odious: that in a most frequented place a cargo-ship should be openly built for you in the province which you were holding with imperium? What did you think those who saw said, what did those who heard reckon? That you would bring that ship empty into Italy? That when you had come to Rome, you would set up as a shipper? Not even this could anyone suspect: that you had a maritime estate in Italy and were procuring a cargo-ship for carrying off harvests. Such was the talk of all about you that you wished to be openly: that you were procuring that ship that should carry plunder out of Sicily, and should make the journey to those thefts which you had left.

\ciceroSection{47} But all these things, if you teach that the ship was built from your own money, I let go and grant. But this, most senseless of men, do you not understand was taken away in the previous hearing by these very Mamertines, your praisers? For Heius, the chief of the city, the chief of the embassy which was sent for your praise, said that the ship was made for you by the public works of the Mamertines, and that a Mamertine senator had been in public charge of its building. The remaining thing is the timber. This, on the Reginians, as they themselves say --- although you cannot deny it --- you publicly commanded, because the Mamertines have no timber. If both that out of which the ship is made, and those who make it, were furnished to you by your command and not at a price, where then is hidden what you say was spent from your own money? But the Mamertines have nothing in their records.

\ciceroSection{48} First, I see it could happen that they should give nothing from their treasury. For even the Capitoline, just as was done by our ancestors, with workmen and labour publicly ordered, could be built up and finished free. Next --- which I see and which I shall show, when I shall produce them, from their own letters --- that very many sums of money were entered as paid out for false and empty contracts of works. Now this is least wonderful: that the Mamertines, from whom they had received the greatest benefit, whom they knew was more friendly to themselves than to the Roman people, had spared his life by their own writings. But if it is a proof that the Mamertines did not give you money because they have not entered it, let it be a proof to you that the ship stood you free, because what you bought or what you let out you cannot show in writing.

\ciceroSection{49} But you did not, you say, command a ship of the Mamertines because they are in alliance. May the gods favour us! We have a man brought up in the hands of the fetial priests, the one above the rest sacred and diligent in the religious bonds of treaties. Let all the praetors who were before you be handed over to the Mamertines, because they had commanded a ship of them against the agreement of the treaty. But yet, you sacred and religious man, why did you command a ship of the Tauromenitans likewise in alliance? Or will you make this credible, that in the equal case of two peoples, without a price, the right was various and the condition unequal?

\ciceroSection{50} What if I show that these two treaties of two peoples, gentlemen, are such that for the Tauromenitans it has been by name provided and excepted in the treaty that they need not give a ship; that for the Mamertines in the very treaty it has been ordained and laid down that it is necessary that they give a ship; that this man, however, against the treaty has both commanded the Tauromenitans and remitted the Mamertines --- could it be in doubt to anyone that under Verres as praetor a cargo-ship aided the Mamertines more than a treaty did the Tauromenitans? Let the treaties be read out. By that benefit of yours, then, as you yourself proclaim it, by a price and wage --- as the matter shows --- you have lessened the majesty of the commonwealth, you have lessened the helps of the Roman people, you have lessened the resources prepared by the virtue and wisdom of our ancestors, you have taken away the right of empire, the condition of the allies, the memory of the treaty. Those who out of the treaty itself ought to send a ship even to Ocean, if we should command it, armed and equipped at their own cost and peril --- these men, that they should not sail in the strait before their own roofs and homes, that they should not defend their own walls and harbours, ransomed from you by a price the right of the treaty and the condition of the imperium.

\ciceroSection{51} What do you suppose the Mamertines wished to spend, in making this treaty, of toil, work, money, that this bireme should not be added in, if they could in any way obtain that from our ancestors? For when this duty so heavy was being laid on the city, there was somehow in that treaty of partnership a kind of mark of slavery. Which thing, then, with their services fresh, the matter still whole, no difficulties of the Roman people, they could not get from our ancestors by treaty: that thing now, with no new service of theirs, so many years afterwards --- a right of our imperium yearly used and always kept --- in the highest difficulty of ships, from Gaius Verres they got by a price. And not only this they got, that they should not give a ship. What sailor, what soldier who should be in the fleet or on guard did the Mamertines under you as praetor for three years give?

\ciceroSection{52} Finally, when by the senate's decree and likewise by the Terentian-Cassian law grain ought to be bought equally from all the cities of Sicily, that duty too, light and common, you remitted the Mamertines. You will say that the Mamertines did not owe grain. How not owe? In the sense that they should not sell? For this kind of grain was not of the kind that was demanded, but of that which was bought. By you then as author and interpreter the Mamertines owed not even market and supply to help the Roman people.

\ciceroSection{53} What city, then, was there which owed it? Those who plough public lands --- it is fixed what they owe by the censorial law. Why did you command of these anything besides out of another kind? What? Do the tithe-payers owe anything beyond the single tithes by the Hieronic law? Why did you also lay down for these how much from this kind of bought grain they should give? What of the immune? These surely owe nothing. But on these you not only commanded; but, that they should give more than they could, those 60,000 modii which you had remitted the Mamertines, you added. Nor do I say this --- that on the rest it was not rightly commanded; but on the Mamertines, who were in the same case, and on whom all the previous praetors had likewise as on the rest commanded and had paid out money by the senate's decree and the law: on these, I say, it was not rightly remitted. And, that he might fix this benefit (as it is said) with a beam-nail, with a council he hears the case of the Mamertines, and from the council's opinion he proclaims that he is not commanding grain of the Mamertines.

\ciceroSection{54} Hear the decree of the hireling praetor from his own commentary, and learn what gravity in writing, what authority in establishing right there is. Read. \textsc{Commentary}. \textsc{From the council's opinion}, he says, \textsc{he gladly does it}, and so he writes. What if you had not used this word "gladly"? We should have thought, of course, that you were making gain unwilling. And "from the council's opinion"! A famous council was read, gentlemen --- did you hear it? Did the council of a praetor seem to you to be read out, when you heard the names, or the partnership and retinue of a most dishonest brigand?

\ciceroSection{55} Lo, the interpreters of treaties, the authors of partnership, of agreement, of religion! Never in Sicily was grain bought publicly without being commanded of the Mamertines in proportion, before this man devised this picked and famous council, that he might receive coins from these and be like himself. So the authority of his decree had as much weight as it ought to have had of that man who, from those from whom he ought to have bought grain, had sold the decree. So at once Lucius Metellus, when he had succeeded him, by the institution and writings of Gaius Sacerdos and Sextus Peducaeus commanded grain of the Mamertines.

\ciceroSection{56} Then those men understood that what they had bought from a bad author they could not longer hold. Come, you, who wished to seem so religious an interpreter of treaties, why did you command grain of the Tauromenitans, why of the Netini? Both of which states are in alliance. And the Netini did not fail themselves; and as soon as you had pronounced that you gladly remitted the Mamertines, they came to you and showed that their cause of treaty was the same. You could not in the same case decree otherwise. You proclaim that the Netini ought not to give grain. And yet you exact it from them. Give me the letters of that same praetor, both of decreed matters and of grain commanded. \textsc{Letters of decreed matters}. What rather, gentlemen, can we suspect in this great and so foul inconsistency than what is necessary: either that the money was demanded by him from the Netini and not given, or that this was being done so that the Mamertines might understand they had bestowed so many prizes and gifts well with him, when others did not get the same right out of the same case?

\ciceroSection{57} And here will he yet dare to make mention of the Mamertine praise! In which how many wounds there are, who of you, gentlemen, does not understand? For first in trials, who cannot give ten praisers, it is more honourable for him to give none than not to fill out that lawful, customary number. So many cities are there in Sicily over which you held command for three years: the rest accuse; a few small ones, by force and fear pressed down, are silent; one praises. What is this save to understand what advantage true praise has, but yet so to have ruled the province that this advantage must necessarily be lacking?

\ciceroSection{58} Next --- which I have said before in another place --- what at last is that praise, of which praise the envoys, the chiefs, said both that a ship had been built for you publicly and that they themselves had been despoiled and plundered by you privately? Finally, what else are these doing, when alone of Sicily they praise you, save that they are witnesses to us that you bestowed all things on them which you took from our commonwealth? What colony in Italy is of such good right, what town so immune, has used in these years so commodious an exemption from all things as the Mamertine state? For three years alone they did not give what they owed by treaty. Alone under you as praetor they were immune from all things; alone in your imperium they lived on this condition: that they should give nothing to the Roman people, that they should refuse nothing to Verres.

\ciceroSection{59} But that I may return to the fleet, whence my speech has gone aside: you have received a ship from the Mamertines against the laws, you have remitted them against the treaties. So in one city you have twice been dishonest, since you have both remitted what you ought not, and received what was not lawful. You ought to have demanded a ship which would sail against brigands, not which sailed with plunder; which would defend the province from being despoiled, not which would carry the spoils of the province. The Mamertines provided you with both a city to which you might bring your thefts from every quarter, and a ship in which you might export them. That town was your receptacle for the plunder; those men the witnesses and watchmen of your thefts; those men prepared for you both a place for thefts and a vehicle of thefts. So not even then, when by your greed and wickedness you lost the fleet, did you dare to command a ship of the Mamertines. At which time, in such great want of ships and in such great calamity of the province, even if they had to be asked as suppliants, yet it would have been obtained from them. For your power of commanding and your attempt of asking was held back by that distinguished bireme not given back to the Roman people, but the cargo-ship given to the praetor. That was the wage of the imperium, of help, of right, of custom, of the treaty.

\ciceroSection{60} You have the firm help of one city lost and sold for a price. Learn now a new method of plundering, devised first by him. The whole expense for the fleet --- in grain, in pay, and in the rest of things --- each city was always wont to give to its own captain. He neither dared to risk being accused by his sailors, and had to render accounts to his fellow citizens, and was occupied in that whole business not only by his own labour but also at his own peril. This was, as I say, always done, and not only in Sicily but in all provinces, and even in the pay and expense of allies and Latins, at the time when we were wont to use their helps. Verres, since the imperium was established, was the first to command that all that money be counted out to him from the cities, that he himself should handle that money whom he himself had set in charge.

\ciceroSection{61} Who can doubt for what reason you both first changed the old custom of all, and neglected so great an advantage of having money handled by others, and undertook so great a difficulty with the charge, so great a trouble with the suspicion? Then other gains are devised: see how many out of the one naval kind! To take money from the cities that they should not give sailors, to discharge the sailors at a fixed price, to take all the pay of the discharged as gain, not to give to the rest what was owed --- learn all these from the testimonies of the cities.

\ciceroSection{62} Read. \textsc{Testimonies of the cities}. This man, this shamelessness, gentlemen, this audacity! To distribute to the cities sums of money in proportion to the number of soldiers, to fix a fixed price --- six hundred sesterces --- of discharging sailors! Those who had given it had taken away leave for the whole summer; he kept as gain whatever he had received in the name of that sailor for pay and grain. So a double gain was being made out of one discharge. And these things this most senseless man, in such great onset of brigands and such great peril of the province, was so doing openly that the brigands themselves knew it and the whole province was witness.

\ciceroSection{63} When on account of this man's greed there was in Sicily a fleet in name --- in fact indeed empty ships, which would bring plunder to the praetor and not fear to the brigands --- yet, when Publius Caesetius and Publius Tadius were sailing with their ten ships half-full, a certain pirate ship laden with plunder they did not capture, but led off, plainly captured and weighed down by its own load. That ship was full of the most beautiful young men, full of silver wrought and stamped, with much bedclothing. This one ship was not captured by our fleet but found by Megaris, which place is not far from Syracuse. When this was reported to him, although he was lying drunk on the shore with the little women, he yet roused himself and at once sent several guards to his quaestor and his legate, that everything should be brought to him whole as soon as possible.

\ciceroSection{64} The ship is brought in to Syracuse. Punishment is awaited by all. He, as if plunder had been brought to him, not pirates captured --- those who were old and ugly, them he reckoned in the number of enemies. Those who had any beauty of age and skill, he led off all. Some he distributed to the clerks, his son, and the staff. Six musicians he sent as a gift to a friend of his at Rome. That whole night was spent in emptying out the ship. The chief pirate himself --- on whom punishment ought to have been taken --- nobody saw. Today all hold this --- you ought by guess to attain how much of it is true --- that this man secretly took money from the pirates for that chief pirate.

\ciceroSection{65} "It is a guess." No one can be a good judge who is not moved by sure suspicion. You know the man; you hold the custom of all --- how willingly he who has captured a leader of brigands or enemies suffers him to be in the open before the eyes of all. I saw at Syracuse, gentlemen, in such a great gathering, no man who said he had seen the chief pirate captured, when all (as is the custom, as is wont to happen) were running together, asking, longing to see. What happened that this man was so greatly hidden that even by chance no one could look on him? Sea-faring men at Syracuse, who had often heard of the name of that leader, often had feared him, when they wished to feed their eyes and to glut their minds with his torture and punishment, were given the power of seeing by no one.

\ciceroSection{66} One man captured more leaders of pirates alive than all before him: Publius Servilius. Was anyone ever therefore deprived of this fruit, that it should not be allowed to see the captured pirate? But on the contrary, wherever he made his journey, he gave this most pleasant spectacle to all of bound and captured enemies. So a gathering was made round him from every quarter, that not only out of those towns through which they were being led, but even out of the neighbouring ones, men came together to look. But the triumph itself --- for what reason was it of all triumphs the most welcome to the Roman people and the most pleasant? Because nothing is sweeter than victory; nothing, however, is a surer witness of victory than to see those whom you have often feared led bound to punishment.

\ciceroSection{67} Why did you not do this? Why was that pirate of yours so hidden as if it were unspeakable for him to be seen? Why did you not exact punishment? For what cause did you keep the man? Do you know any captured chief pirate before in Sicily who was not struck with the axe? Give me one author of your deed; bring forth one example. You were keeping the chief pirate alive: for what? For a triumph, I trust, that you might lead him before your chariot. Nor was anything left save that, with the most beautiful fleet of the Roman people lost and the province torn to pieces, a naval triumph should be decreed for you.

\ciceroSection{68} Come further: it pleased you that the leader of the brigands should be kept under guard by a new fashion rather than struck with the axe by the example of all. What guards are these? Among what men, in what manner was he kept? You have all heard of the Syracusan stone-quarries; most of you know them. The work is huge, magnificent, of kings and tyrants. The whole is from rock sunk to a wonderful depth and cut deeply by the work of many. Nothing so closed for exit, nothing so hedged in on every side, nothing so safe for guarding can either be made or thought of. Into these stone-quarries, if any are to be guarded publicly, men are commanded to be brought even from the rest of the towns of Sicily.

\ciceroSection{69} Because he had thrown into them many captive Roman citizens, because he had ordered the rest of the pirates to be shut up in the same place, he understood that, if he had given this counterfeit chief pirate into the same custody, the true leader would be sought by many in the stone-quarries. So he does not dare commit the man to that best and safest custody; finally he fears all Syracuse. He sends the man off --- whither? To Lilybaeum, perhaps? I see; yet he does not at all fear sea-faring men. By no means, gentlemen. To Panhormus then? I hear; though above all at Syracuse, since he had been captured in Syracusan territory, if not punished, at least he ought to have been kept.

\ciceroSection{70} Nor even at Panhormus. Whither then? Where do you suppose? Among men most far from any fear or suspicion of pirates, most removed from sailing and maritime matters: to the Centuripines, men most inland, the highest farmers, who would never have feared the name of a sea-going brigand --- under you alone as praetor would have shuddered at Apronius, the land-archpirate. And, that any might easily see what was being done by him --- so that this counterfeit might easily and willingly pretend to be the man he was not --- he commands the Centuripines that he should be treated in food and the rest of things most liberally and most commodiously.

\ciceroSection{71} Meanwhile the Syracusans, knowing and humane men, who not only could see those things which were plain but also suspect what was hidden, all daily kept the reckoning of the pirates who were being struck with the axe. How many there ought to be, they were guessing from the very ship that had been captured and from the number of the oars. This man, because he had taken away and led off all who had any skill or beauty, suspected that, if he had bound the rest, as the custom is, all together to the stake, there would be a cry of the people, since so many more had been led off than left. For this reason, when he had set up to bring out some at one time, others at another, yet in such a great gathering there was no one who did not keep the reckoning and the number, and not only miss the rest but even demand and require them.

\ciceroSection{72} When the great number was lacking, then this unspeakable man began to put in their place and to substitute for those whom he had led to his own house out of the pirates Roman citizens, whom he had before thrown into prison; some of whom he was charging had been Sertorian soldiers, and was saying that, fleeing from Spain, they had put in to Sicily; others, who had been captured by brigands when they were doing business or sailing for some cause, he was alleging had been with the pirates of their own will. So some Roman citizens, lest they should be recognized, with their heads veiled were being snatched from the prison to the stake and to death; others, when they were recognized by many Roman citizens and were defended by all, were yet struck with the axe. Of whose bitterest death and most cruel torture I shall speak when I begin to handle that place; and so I shall speak that, if in that complaint I am about to make about this man's cruelty and about the most unworthy death of Roman citizens not only my strength but even my life shall fail me, I shall reckon it distinguished and pleasant for me.

\ciceroSection{73} This therefore is the deed done, this the distinguished victory: a pirate light vessel captured, the leader freed, musicians sent to Rome, beautiful men and youths and craftsmen led off home; in their place and to their number Roman citizens tortured and killed in a hostile manner; all the clothing carried off; all the gold and silver carried off and diverted. But how he himself put himself in it in the previous hearing! Who, when he had been silent through so many days, suddenly at the testimony of M. Annius, a most splendid man --- when he had said that a Roman citizen had been struck with the axe and denied that the chief pirate had been --- leaped up by his consciousness of crime and roused by the frenzy conceived from his evil deeds. He said that, because he knew it would be brought as a charge against him that he had taken money and had not exacted punishment from the true chief pirate, therefore he had not struck him with the axe. He said that he had at his own house two chief pirates.

\ciceroSection{74} O the clemency of the Roman people, or rather their wonderful and singular patience! M. Annius, a Roman knight, says that a Roman citizen was struck with the axe; you are silent: he denies that a chief pirate was; you confess. There is a groan and cry of all, while yet the Roman people held back from your immediate punishment, and reserved the reckoning of its own safety for the strictness of the judges. What? Did you know it would be brought as a charge against you? Why did you know? Why did you even suspect? You had no enemy. Even if you had, yet you had not so lived that you ought to have had the fear of trial set before you. Or did the consciousness of your guilt --- which is wont to happen --- make you fearful and suspicious? You then, when you were in command, even then shuddered at charge and trial: now, when by so many witnesses you are convicted, can you doubt of condemnation?

\ciceroSection{75} But if you feared this charge --- that someone should say a counterfeit had been struck with the axe by you for a chief pirate --- which then did you reckon would be the firmer for your defence: in the trial, by my compulsion and request, to bring forth before strangers, after so long a time, the man whom you should say was the chief pirate; or, the matter fresh, at Syracuse, before those who knew, with almost all Sicily looking on, to strike with the axe? See what difference there is in which had to be done. In the former there would be no reproach; in this there is no defence. So all have always done that; this who has done before you, who beside you, I ask. You held a pirate alive. To what limit? While you were in command. For what cause, by what example, why so long? Why, I say, when the Roman citizens whom the pirates had captured were at once struck with the axe, did you give to the pirates themselves so long-lasting an enjoyment of the light?

\ciceroSection{76} But suppose so. Let all that time when you were in command be free to you. Even as a private man, even as a defendant, even almost condemned, did you keep the leaders of the enemy in a private house? One month, a second, finally almost a year from the time when they were captured, the pirates were at your house, as long as it was allowed me --- that is, as long as it was allowed by Manius Glabrio, who at my request ordered them to be brought forth and shut up in prison. What is the right of this matter, what the custom, what the example? Could the most fierce and most hostile enemy of the Roman people, or rather the common enemy of all races and nations, be kept by any private mortal within the walls of his own house?

\ciceroSection{77} What? If the day before you were forced by me to confess that, with Roman citizens struck with the axe, the leader of the brigands was alive and lived at your house --- if, I say, the day before he had fled from your home, if he had been able to make any band against the commonwealth, what would you say? "He lived at my house, was with me; I kept him alive and unhurt for my own trial, that I might more easily dispel my enemies' charge." Indeed? Will you defend your perils by a common peril? Will you transfer the punishments which are owed to conquered enemies to your own time, not to the commonwealth's? Shall the enemy of the Roman people be kept under the guards of a private man? But even those who triumph, and for that reason longer keep the leaders of the enemy alive, that, when these have been led in triumph, the Roman people may take in the most beautiful spectacle and fruit of victory --- yet, when they begin to turn the chariots from the forum onto the Capitoline, they order them to be led to prison, and the same day makes an end both of imperium for the victors and of life for the conquered.

\ciceroSection{78} And now I trust no one doubts that you would not have committed yourself --- especially when you had decided, as you say, that you would have to plead a case --- to letting the chief pirate be struck with the axe rather than (which was set before your eyes) live at your peril. For if he had been dead, you, who say you feared the charge, I ask, to whom would you prove it? When it was agreed that this man had been seen by no one at Syracuse as the chief pirate, that he was missed by all, when no one doubted that he had been freed by you for money, when commonly men were saying a counterfeit had been put in his place whom you wished to prove instead of him, when you yourself had confessed that you had so long before feared the charge --- if you should say he was dead, who would hear you?

\ciceroSection{79} Now, when you bring forth I know not what living one, you yet see that you are laughed at. What? If he had fled, if he had broken his chains as Nico, that most noble pirate, did, whom Publius Servilius with the same good fortune as he had captured him recovered, what would you say? But this was the truth: if that one true pirate had once been struck with the axe, you would not have that money. If this counterfeit had died or fled, it would not be hard to substitute another in the place of the substitute. I have spoken more than I wished about that chief pirate, and yet I have passed over the surest proofs of this charge. For I wish this whole charge to be kept entire for me; there is a fixed place, a fixed law, a fixed court, for which this is reserved.

\ciceroSection{80} Made richer by this great plunder, enriched by slaves, silver, clothing, he was no more diligent at adorning the fleet, recalling and feeding the soldiers --- although the matter could be not only for the safety of the province but even for the plunder itself. For in the height of summer, at which time the rest of the praetors are wont to make a circuit of the province and to hurry round, or even, in such great fear and peril of pirates, to sail themselves --- at that time he was not content with his own royal house (which was King Hiero's, which the praetors are wont to use) for his luxury and lusts. He ordered tents (as he had been wont in the summer seasons, which I have shown before) stretched with linen sails to be set up on the shore, which is the shore on the Island at Syracuse behind the fountain of Arethusa, near the very entrance and mouth of the harbour, in a pleasant and from witnesses removed place.

\ciceroSection{81} Here through the summer days the praetor of the Roman people, the guardian and defender of the province, so lived that there were daily women's banquets. No man took his place save he himself and the praetexta-clad son --- although I had said rightly without exception "no man," since with these men there, there was none. Sometimes also his freedman Timarchides was brought in. The women, however, were married noblewomen save one, the daughter of the mime Isidorus, whom this man, on account of love, had led off from the Rhodian flute-player. There was a certain Pipa, the wife of Aeschrio of Syracuse, about which woman very many verses, which were made about this man's lust, are spread through all Sicily.

\ciceroSection{82} There was Nice, of singular face (so it is proclaimed), wife of Cleomenes of Syracuse. The husband loved her; but yet against this man's lust he could neither nor dared resist, and at the same time was being bound by him with many gifts and benefits. But at that time this man, although the man's shamelessness is what you know, yet himself, since the husband was at Syracuse, had little chance with a soul free and at ease to keep his wife so many days on the shore with him. So he devises a singular thing. The ships which the legate had commanded he hands over to Cleomenes; he orders Cleomenes the Syracusan to be in command of the fleet of the Roman people. He does this so that the man should not only be away from home while he sailed, but also be willingly absent with great honour and favour, while he himself, with the husband removed and sent away --- not more freely than before (for who has ever stood in the way of his lust?), but with a somewhat freer spirit yet --- might keep her with him, as having removed him not as a husband but as a rival.

\ciceroSection{83} Cleomenes the Syracusan receives the ships of the allies and friends. What shall I first either accuse or complain of? That the power of a legate, of a quaestor, finally of a praetor, the honour, the authority, was given to a Sicilian? If that taking up with banquets and women hindered you, where were the quaestors, where the legates, where the grain assessed at three denarii, where the mules, where the tents, where so many great ornaments granted and given to magistrates and legates by the senate and Roman people; finally, where your prefects, where your tribunes? If no Roman citizen was worthy of that business, what of the cities which had remained perpetually in the friendship and faith of the Roman people? Where was the Segestan state, where the Centuripine? Which both by services, by faith, by age, and even by kinship touch the name of the Roman people.

\ciceroSection{84} O immortal gods! What? If Cleomenes of Syracuse was ordered to command the soldiers, ships, captains of these very cities, was not all honour of dignity, of equity, of duty taken from him by you? Did we ever wage any war in Sicily that we did not have the Centuripines as allies, the Syracusans as enemies? But these things I wish referred to the memory of antiquity, not to the insult of the city. So that most distinguished man and highest general, Marcus Marcellus, by whose virtue Syracuse was taken, by whose pity it was kept safe, did not wish any Syracusan to live in that part of the city which is in the Island; today, I say, no Syracusan is allowed to live in that part. For there is a place which even few could defend. He did not therefore wish to entrust it to least faithful men --- at the same time because from that part of the city the approach to ships from the deep is open. So those who had often shut out our armies, to them he reckoned the keys of the place should not be entrusted.

\ciceroSection{85} See what difference there is between your lust and the authority of our ancestors, between your love and frenzy and their counsel and prudence. They took the approach of the shore from the Syracusans; you have given them maritime command. They did not wish a Syracusan to live in that place from which ships could approach; you wished a Syracusan to be in command of the fleet and ships. Those from whom they took a part of their own city, to those you have given a part of our imperium; and those by whose help as allies the Syracusans obey our word, them you have ordered to obey the word of a Syracusan.

\ciceroSection{86} Cleomenes goes out from the harbour in a Centuripine quadrireme. There follows a Segestan ship, a Tyndaritan, a Herbitensian, a Heraclean, an Apolloniate, a Haluntine --- a fleet famous in show, but lacking and weak on account of the dismissal of the marines and rowers. So long this man, the diligent praetor, saw the fleet in his own command as long as it sailed past his most disgraceful banquet. He himself, who had not been seen for many days, then for a little gave himself into the sight of the sailors. The praetor of the Roman people stood with sandals on, in a purple cloak and tunic to the ankles, leaning on a little woman on the shore. In this dress already this man Sicilians and very many Roman citizens had often seen.

\ciceroSection{87} After the fleet had gone forward a little and on the fifth day at last had put in to Pachynus, the sailors, compelled by hunger, gathered the roots of wild palms (of which there was in those places, as in much of Sicily, a great quantity) and on these the wretched and ruined men were fed. But Cleomenes, who reckoned himself a second Verres in luxury and wickedness as also in command, in like manner spent whole days drinking on the shore in a tent set up. But behold, suddenly, with Cleomenes drunk and the rest hungering, news comes that there are pirate ships in the harbour Odysseia (for so the place is named); but our fleet was at the harbour of Pachynus. But Cleomenes, since there was a land guard not in fact but in name, hoped that with those soldiers whom he had led away from that place he could fill out the number of sailors and rowers. The same reasoning of this most greedy man was found in the garrisons as in the fleet; for very few were left, the rest discharged.

\ciceroSection{88} Cleomenes the chief in the Centuripine quadrireme ordered the mast to be raised, the sails to be made, the anchors to be cut. At the same time he gave the signal that the rest should follow him. This Centuripine ship was of incredible speed under sail; for what each ship could do with oars no one under this man as praetor could know. Although in this quadrireme on account of the honour and favour of Cleomenes very few rowers and soldiers were lacking. The quadrireme had now flown almost out of sight, fleeing, when even then the rest of the ships in one place were struggling.

\ciceroSection{89} There was spirit in the rest. Although they were few, however the matter stood, they cried that they wished to fight, and what hunger had left of life and strength they wished to give back rather by the sword. But if Cleomenes had not fled so much before, there would yet have been some way of resisting. For that ship alone was decked, and so great that it could be a defence to the rest, which, if it had been engaged in a battle of brigands, would seem to have the look of a city among those pirate light vessels. But then the helpless, abandoned by their leader and the prefect of the fleet, of necessity began to hold the same course.

\ciceroSection{90} Toward Helorus, as Cleomenes himself, so the rest sailed; and they were not so much fleeing the assault of the brigands as following the general. Then as each was last in flight, so he was first in danger; for each last ship the pirates attacked first. So the first ship of the Haluntines is taken, of which a Haluntine noble man, the phylarch, was in command, whom afterwards the Locrenses ransomed publicly from those brigands. From which man you have learned, by his oath, the whole matter and case in the previous hearing. Then the Apolloniate ship is taken, and its prefect Anthropinus is killed.

\ciceroSection{91} While these things were going on, meanwhile Cleomenes had now reached the shore of Helorus; he had now thrown himself out of the ship onto land, and had left the quadrireme tossing on the swell. The rest of the prefects of the ships, when their general had gone out onto the land --- since they could neither fight back nor flee by sea in any way --- with their ships put in to Helorus, followed Cleomenes. Then the leader of the brigands, Heracleo, suddenly beyond hope conqueror not by his own virtue but by this man's greed and wickedness, ordered the most beautiful fleet of the Roman people, driven up and cast on the shore, to be set on fire and burned, when first it was growing dark.

\ciceroSection{92} O wretched and bitter time of the province of Sicily! That accident calamitous and deadly to many innocent men! O the singular wickedness and shamefulness of this man! One and the same was the night in which the praetor blazed with the flame of a most foul love, and the fleet of the Roman people with the burning of brigands. The grave news of this ill is brought to Syracuse in the dead of night. Men run to the praetorium, to which the women had brought him back from that famous banquet a little before with song and music. Cleomenes, although it was night, yet does not dare to be in public; he shuts himself up at home; nor was his wife present, who could comfort the man in his ills.

\ciceroSection{93} But the discipline of this distinguished general was so strict at home that in such a great matter, with so grave a message, no one was admitted, no one was there who would dare either to rouse him sleeping or to interrupt him waking. Now, indeed, with the matter known to all, the greatest multitude was running about through the whole city. For (as the custom had always been before) the coming of brigands was not signified by a fire raised on a watchtower or hill, but the flame from the very burning of the ships announced both the calamity received and the peril remaining. When the praetor was being looked for and it was agreed that no one had announced it to him, a running and onset is made to his house with shouting.

\ciceroSection{94} Then this man, roused, hears the whole matter from Timarchides; takes up his cloak --- it was now growing light --- comes forward into the midst, full of wine, sleep, debauchery. He is received by all with such a cry that the likeness of his Lampsacene peril came before his eyes. This even seemed greater, because in like hatred this multitude of men was the largest. Then his deeds were recalled; then those scandalous banquets; then the women were called by the multitude by name; then they asked of him openly through so many continuous days during which he had never been seen, where he had been, what he had done. Then the general appointed by him, Cleomenes, was demanded; nor was anything closer to being done than that the Utican example of Hadrianus should be transferred to Syracuse, that two tombs of two dishonest praetors should be set up in two provinces. But account was taken by the multitude of the time, of the tumult, even of the dignity and common standing, that this assize of Roman citizens at Syracuse is reckoned worthy not only of that province, but even of this commonwealth.

\ciceroSection{95} They confirm themselves, while this man even then half-sleeping was stupid; they take up arms; they fill the whole forum and the Island, which is a great part of the city. The brigands, having tarried that one night alone at Helorus, when they had left even our ships still smoking, begin to approach Syracuse. They, who, of course, had often heard that nothing was more beautiful than the walls and harbours of Syracuse, had decided that, if they had not seen them under Verres as praetor, they would never see them.

\ciceroSection{96} And first they approach those summer haunts of the praetor, that very part of the shore where this man through those days, with tents pitched, had set up the camp of his luxury. After they found this place empty and felt that the praetor had moved his camp from that place, at once without any fear they began to enter into the very harbour. When I say into the harbour, gentlemen --- for it must be more diligently explained for the sake of those who do not know the place --- I say that pirates came into the city and into the inmost part of the city. For that town is not shut in by a harbour; rather the harbour itself is girdled and held in by the city, so that the outer walls are not washed by the sea, but the harbour itself flows into the bosom of the city.

\ciceroSection{97} Here under you as praetor the pirate Heracleo with four small light vessels sailed at his own discretion. By the immortal gods! A pirate light vessel, when the name of the imperium of the Roman people and the fasces were at Syracuse, came as far as the forum of the Syracusans and to all the steps of the city, where neither the most glorious fleets of the Carthaginians, when they were most powerful at sea, in many wars often attempted, were ever able to come; nor was the once unconquered glory of the Roman people at sea, before you were praetor, ever able to penetrate in so many Punic and Sicilian wars. The place is such that the Syracusans saw before within their walls, in their city, in their forum, an armed and victorious enemy than they ever saw a single ship of enemies in the harbour.

\ciceroSection{98} Here, with you as praetor, the little ships of brigands wandered to where the fleet of the Athenians alone, since the memory of men, broke in with three hundred ships by force and multitude, which in that very harbour was conquered and overcome by the nature of the place itself and of the harbour. Here first the resources of that state were broken and pressed down. In this harbour the shipwreck of the nobility, the imperium, the glory of the Athenians is reckoned to have been made. Did a pirate enter where, as soon as he came in, he left a great part of the city not only on his side but also at his back? He sailed past the whole Island, which is a city at Syracuse with its own name and walls, in which place our ancestors, as I said before, forbade a Syracusan to live, because they understood that those who held that part of the city would have the harbour in their power.

\ciceroSection{99} But how he wandered! They were throwing roots of wild palms, which they had found in our ships, that all might know this man's wickedness and the calamity of Sicily. Were the Sicilian soldiers, the children of the farmers, whose fathers ploughed up so much grain by their toil that they could supply the Roman people and all Italy --- they, born on the island of Ceres, where fruits are said to have first been found --- to have used such food as their ancestors removed all others from when fruits had been found! Under you as praetor the Sicilian soldiers were fed on roots of palms, the pirates on Sicilian grain!

\ciceroSection{100} O wretched and bitter spectacle! That the glory of the city, the name of the Roman people, the gathering and multitude of all men should be a mockery to a pirate light vessel! That a pirate should celebrate a triumph in the Syracusan harbour over the fleet of the Roman people, when the oars of the brigands should splash the eyes of a most idle and most worthless praetor! After the pirates had departed from the harbour, not affected by any fear but by satiety, then men began to seek the cause of so great a calamity. All said and openly disputed that it was least wonderful, if the rowers and soldiers were dismissed, the rest ruined by want and hunger, the praetor for so many days carousing with little women, that so great an ignominy and calamity should have been received.

\ciceroSection{101} But this reproach and infamy of his was confirmed by the talk of those who had been put in command of those ships by their own cities. Those of that number who had been left, with the fleet lost, fled back to Syracuse, and were saying how many out of each man's ship he knew had been discharged. The matter was clear; nor only by proofs but even by sure witnesses was this man's audacity caught. The man was made certain that nothing was being done in the forum and in the assize the whole day save this --- that it was being asked of the captains in what manner the fleet had been lost; that they were answering and showing each one --- by the discharge of the rowers, by the hunger of the rest, by the fear and flight of Cleomenes. After this man learned this, he began to take this counsel. He had decided that he must plead a case even before this came to pass --- as you have heard him say in the previous hearing. He saw that, with those captains as witnesses, this charge he could in no way sustain. He takes a counsel at first foolish, but yet mild.

\ciceroSection{102} He orders the captains to be summoned to him; they come. He blames them for having held such talks about him. He asks that each say in his own ship that he had had only as many sailors as he ought, and that no one had been discharged. They show that they will do what he wishes. This man does not put it off; he calls his friends at once. He asks of each in turn how many sailors each had had. Each one in turn answers as he had been instructed. This man enters it on the tablets; he seals them with the seals of his friends, a foreseeing man, that against this charge, if at any time it were needed, he might use this testimony.

\ciceroSection{103} I trust the senseless man was laughed at by his advisers and warned that these tablets would do him no good --- that even more suspicion in that charge would arise from the praetor's excessive diligence. Now this man had been wont in many things to use this stupidity, that he ordered things he wished to be taken out and entered publicly in the writings of the cities. All which now he understands does him no good, after he is convicted by sure writings, witnesses, authorities. When he sees this --- that their confession, his testimony, his tablets will be of no help to him --- he takes a counsel not of a dishonest praetor (for that indeed could be borne), but of a fierce and senseless tyrant. He decides, if he wishes this charge to be lessened (for he did not think it could be wholly taken away), that all the captains, witnesses of his crime, must be deprived of life.

\ciceroSection{104} That reasoning came to him: "What will become of Cleomenes? Shall I be able to animadvert on those whom I ordered to obey his word, and let go him to whom I gave the power and command? Shall I be able to inflict punishment on those who followed Cleomenes, and forgive Cleomenes who ordered them to flee with him and follow him? Shall I be able to be vehement against those who had ships not only empty but even open, slack against him who alone had a decked ship and one less emptied? Let Cleomenes perish too!" Where is the faith, where the oaths, where the right hands and embraces, where that companionship of female military service on that most refined shore? It could in no way be that Cleomenes should not be spared.

\ciceroSection{105} He calls Cleomenes; he tells him that he has decided to animadvert on all the captains; that the reckonings of his own peril so demand. "You alone I shall spare, and shall rather take upon myself the charge of inconsistency in this fault than be cruel to you, or suffer so many such grave witnesses to be alive and unhurt." Cleomenes gives thanks; approves the counsel; says it ought so to be done. He yet warns him of this thing which had escaped him: that against Phalacrus the Centuripine captain it could not be animadverted, because he had been with him in the Centuripine quadrireme. What then? Will a man from such a city, a most noble young man, be left as a witness? "For the present," says Cleomenes, "since it is necessary; but afterwards we shall see something so that he may not be able to stand against us."

\ciceroSection{106} After these things were done and settled, this man comes forth suddenly from the praetorium kindled by crime, frenzy, cruelty. He comes into the forum; orders the captains to be summoned. They, who feared nothing, suspected nothing, run up at once. He orders chains to be cast on the wretched and innocent men. They implore the praetor's faith and ask why he was doing it. Then this man says this for the cause: that they had betrayed the fleet to the brigands. There is a cry and the wonder of the people that there was such great shamelessness and audacity in the man, that he should either ascribe to others the cause of the calamity (which had all happened on account of his own greed); or, when he himself was reckoned the partner of brigands, lay the charge of betrayal against others; then that this charge was born on this fifteenth day after the fleet was lost.

\ciceroSection{107} While these things were going on, men were asking where Cleomenes was --- not because anyone thought him, of whatever kind he be, worthy of punishment for that hurt; for what could Cleomenes do? --- I cannot accuse anyone falsely --- what, I say, could Cleomenes greatly do, with the ships emptied by this man's greed? And they see him sit beside the praetor and whisper familiarly in his ear, as he was wont. Then surely it seemed most unworthy to all that the most honourable men, picked out of their cities, were thrown into iron and chains, while Cleomenes, on account of his partnership in disgraces and shamefulness, was the most familiar with the praetor.

\ciceroSection{108} Yet against them is set as accuser a certain Naevius Turpio, who under Gaius Sacerdos as praetor was condemned for wrongs --- a man well fitted to this man's audacity, whom this man in matters of tithes, in capital matters, in every false charge was wont to have as runner ahead and emissary. The parents and kinsmen of the wretched young men come to Syracuse, moved by this sudden message of their calamity. They see their children bound in chains, when they were sustaining the punishment of his greed on their necks and shoulders. They are present; they defend; they cry out; they implore your faith --- which was nowhere and never had been. The father Dexo of Tyndaris was present, a most noble man, your guest-friend. At whose house you had stayed, whom you had called guest-friend --- when you saw him with such authority lost in misery, could not his tears, not his old age, not the right and name of guest-friendship recall you from crime to some part of humanity?

\ciceroSection{109} But why do I recall the rights of guest-friendship in this monstrous beast? Who entered Sthenius of Thermae, his own guest-friend, whose house he drained and emptied through guest-friendship, in his absence among the defendants, condemned him to capital sentence with the case unheard --- shall we now seek of him the rights and duties of guest-friendships? For do we have to do with a cruel man, or with a wild and monstrous beast? The tears of the father over the peril of an innocent son did not move you. Although you had left a father at home, you had a son with you: did neither your son present, by the dearness of children, nor the absent father, by the indulgence of a parent, warn you?

\ciceroSection{110} Your guest-friend Aristeus, son of Dexo, wore chains. Why so? "He had betrayed the fleet." For what reward? "He had deserted." What of Cleomenes? "He had been a coward." But you had given him a crown for his virtue. "He had discharged the sailors." But you had received from all the wage of discharging. There was on the other side another parent, Eubulides of Herbita, a man at his own home famous and noble; who, because he had hurt Cleomenes in defending his son, was almost left bare. But what was there for anyone to say or to defend? "It is not lawful to name Cleomenes." But the case demands it. "You shall die, if you call him." For he never threatened any man moderately. But there were no rowers. "Will you accuse the praetor? Break his neck." If neither the praetor nor the praetor's rival is allowed to be named, when in these two the whole case rests, what will be?

\ciceroSection{111} Heracleus of Segesta also pleads his case, a man born at his own home in a most noble place. Hear, gentlemen, as your humanity demands; for you shall hear of the great hurts and wrongs of allies. Know that Heracleus was in this case, who, on account of a serious eye disease, had then not sailed, and by the order of him who had power had remained at Syracuse on leave! He surely neither betrayed the fleet, nor, terrified by fear, fled, nor deserted the army. For then this would have to be animadverted upon, when the fleet was setting out from Syracuse. He yet was in the same case as if he had been caught in some manifest crime --- he, on whom the charge could not be laid even falsely.

\ciceroSection{112} There was among those captains a certain Heraclean Furius (for they have not a few names of this Latin sort) --- a man, as long as he lived, famous and noble not only at his own home; after his death, throughout all Sicily. In which man there was so great spirit not only that he should hurt this man freely (for that indeed, since he saw he must die, he understood he was doing without peril), but, with death set before him, when his mother sat by him weeping in the prison nights and days, he wrote a defence of his case. Which now there is no one in Sicily who does not have, who does not read, who is not reminded by that speech of your crime and cruelty. In which he shows how many sailors he had received from his city, how many and at what price he had discharged each one, how many he had had with him; he likewise speaks of the rest of the ships. Which when he was speaking before you, his eyes were beaten with rods. He, with death set before him, easily bore the body's pain. He cried out --- which he left in writing --- that it was an unworthy deed that the tears of a most shameless woman over the safety of Cleomenes should have more weight with you than a mother's over her son's life.

\ciceroSection{113} Then I see this also was said, which (if the Roman people knows you rightly) he proclaimed of you not falsely now in death itself: that Verres could not extinguish testimony by killing witnesses; that he would be a graver witness with wise judges from the dead than if he were brought alive into court; that then he would be a witness of his greed alone, if he lived; now, when he had been so killed, he would be a witness of his crime, audacity, cruelty. Then those distinguished words: that not only crowds of witnesses, when your case was being treated, but from the divine spirits the Avengers of the innocent and the Furies of the wicked would come into your trial. That he therefore reckoned his own lot lighter, because he had already seen the edge of your axes and the face and hand of Sextius, your executioner, when, in the assize of Roman citizens, by your order Roman citizens were being struck with the axe.

\ciceroSection{114} To be brief, gentlemen: of the freedom you have given to allies, this man has used in the most bitter punishment of most wretched slavery. He condemns all from the council's opinion. Yet this man, in such a great matter of so many men, did not call to him Titus Vettius, his quaestor, by whose counsel he should use; nor Publius Cervius, such a man, his legate, who, because he had been legate in Sicily under this man as praetor, was the first man challenged by him as judge: but from the opinion of brigands --- that is, of his own retinue --- he condemns all.

\ciceroSection{115} Here all the Sicilians, the most faithful and most ancient allies, afflicted with very many benefits from our ancestors, are gravely moved and tremble for their own perils and all their fortunes. They take it ill that that clemency and gentleness of our imperium has been turned into such great cruelty and inhumanity --- that so many men are condemned at one time on no charge --- that the dishonest praetor seeks the defence of his own thefts out of the most unworthy death of innocent men. To this dishonesty, frenzy, and cruelty nothing now seems able to be added, gentlemen; and rightly nothing seems so. For if he should contend with the dishonesty of others, he would far and much surpass all.

\ciceroSection{116} But when he himself contends with himself, he does this: that he should always overcome his earlier deed by a new crime. I had said that Phalacrus the Centuripine had been excepted by Cleomenes, because in his quadrireme Cleomenes had sailed. Yet, because the young man had become afraid (since he saw his case the same as theirs who, innocent, were perishing), Timarchides comes to him. He says there is no peril of the axe; warns him to take care lest he be flogged with rods. To be brief, you have heard the young man himself say that, on account of this fear, he counted out money to Timarchides.

\ciceroSection{117} These things are light in this defendant. The captain, a most noble man of a most noble city, ransomed himself from the fear of rods at a price: it is human. Another gave money lest he be condemned: it is customary. The Roman people does not wish Verres to be accused on worn charges; it asks for new ones; it desires unheard ones. It thinks that it is not a trial of a Sicilian praetor, but of an unspeakable tyrant. The condemned are shut up in prison. Punishment is established against them; it is exacted from the wretched parents of the captains. They are forbidden to approach their sons; they are forbidden to bring food and clothing to their own children.

\ciceroSection{118} These fathers whom you see lay on the threshold; the wretched mothers spent the night at the door of the prison, shut out from the last sight of their children: who asked nothing else save that they might be allowed to take in the last breath of their sons in their mouths. There stood the keeper of the prison, the praetor's executioner, the death and terror of allies and Roman citizens, the lictor Sextius, for whom out of every groan and grief a fixed wage was prepared. "That you may approach, you shall give so much; that food may be brought in to you, so much." No one refused. "What? That with one stroke of the axe I should bring death to your son --- what will you give? That he be not long tortured, that he be not stricken oftener, that the breath be not taken away with some sense of pain?" Even on this account money was given to the lictor.

\ciceroSection{119} O great and unbearable grief! O grave and bitter fortune! Not the lives of their children but the swiftness of their death the parents were compelled to ransom at a price. And the young men themselves were speaking with their Sextius about the blow and about that one stroke; and this last thing the children begged of their parents: that, for the sake of lightening their torture, money should be given to the lictor. Many and grave griefs were found for the parents and kinsmen, many. Yet death is the last. It will not be. Is there anything beyond to which cruelty could go? It will be found. For the bodies of those, when they have been struck with the axe and killed, will be cast to the wild beasts. If this is grievous to the parents, let them ransom at a price the power of burying.

\ciceroSection{120} Onasus of Segesta, a noble man, you have heard say that on account of the burial of the captain Heracleus he counted out money to Timarchides; lest you say, "for the fathers come angry over their lost sons" --- a leading man, a most noble man, says, and does not speak about a son. Now who was at Syracuse then who has not heard, who does not know that these agreements of Timarchides about burial were made even with those still alive? Did men not openly speak with Timarchides? Were not all the kinsmen of all called in? Were not the funerals of the living openly being contracted? With all these things done and settled, they are brought out from prison; they are bound.

\ciceroSection{121} Who was so iron at that time, so inhuman save you alone, that he was not moved by their age, nobility, misery? Was there anyone who did not weep, who did not so reckon their calamity that he yet thought their fortune not foreign, the danger common? They are struck with the axe. You rejoice in the groaning of all and triumph; you delight that the witnesses of your greed have been taken away. You erred, Verres, and you erred vehemently, when you reckoned you could wash off the stains of your thefts and disgraces with the blood of innocent allies. You were borne headlong by frenzy, you who reckoned that the wounds of greed could be healed by the remedies of cruelty. For although those men have died, witnesses of your crime, yet their kinsmen are not lacking either to you or to them; yet out of that very number of captains some live and are at hand, whom (as it seems to me) fortune has reserved for the punishment of those innocent men and for this case.

\ciceroSection{122} The phylarch of Haluntium is here, who, because he did not flee with Cleomenes, was overcome by the brigands and captured. To whom calamity was salvation; who, had he not been captured by pirates, would have fallen into this brigand of the allies. He says in evidence about the discharge of the sailors, about the hunger, about the flight of Cleomenes. The Centuripine Phalacrus is here, born in a most ample city in the most ample place. He says the same; in nothing does he disagree.

\ciceroSection{123} By the immortal gods! With what spirit, gentlemen, do you sit, or in what manner do you hear these things? Whether am I out of my senses, and grieve more than enough at so great a calamity and misery of allies; or are you also affected here by this most bitter torture and grief of innocent men with an equal sense of grief? For I, when I say a Herbitan, a Heraclean was struck with the axe, the unworthiness of the calamity turns before my eyes. Were the citizens of those peoples, the nurslings of those fields, out of which the greatest force of grain is yearly sought for the Roman commons by their works and labours --- who from their parents in the hope of our imperium and of our equity were undertaken and brought up --- were they kept for Gaius Verres's unspeakable monstrousness and for his deadly axe?

\ciceroSection{124} When that Tyndaritan comes into my mind, when the Segestan, then I consider together the rights and duties of cities. Those cities which Publius Africanus thought ought to be adorned with the spoils of enemies --- these Gaius Verres deprived not only of those ornaments, but even of most noble men by his unspeakable crime. Lo, what the Tyndaritans willingly proclaim: "We are reckoned among the seventeen peoples of Sicily; we have always in all the Punic and Sicilian wars followed the friendship and faith of the Roman people; from us all things, both helps of war and ornaments of peace, have been ministered to the Roman people." Indeed much did these rights profit them in this man's command and power!

\ciceroSection{125} Your sailors once Scipio led against Carthage; but now an almost empty ship Cleomenes leads against brigands. With you Africanus shared the spoils of the enemy and the rewards of praise; but now, despoiled by Verres, with the ship led off by brigands, we ourselves are reckoned in the place and number of enemies. What further? That kinship of the Segestans, not only handed down in writing or recalled in words, but used and approved by their many services --- what fruits of this connection did it bring in this man's command? Of course this was the right, gentlemen: that out of the bosom of his fatherland a most noble young man was given over to this man's executioner Sextius. To which city our ancestors granted the largest and best fields, which they wished to be free of tribute --- this with you got not even this right by kinship, faithfulness, antiquity, authority: that it should plead for the death and blood of one most honourable and most innocent of its citizens.

\ciceroSection{126} Whither shall the allies flee for refuge? Whom shall they implore? Finally, by what hope, that they should wish to live, will they be held, if you abandon them? Will they come to the senate? What? That it may exact punishment on Verres? That is not customary; not senatorial. Will they flee for refuge to the Roman people? The cause of the people is easy; for the people will say that for the sake of allies it has passed the law and set its guardians and avengers in charge. This place therefore is the only one to which they may flee, this the harbour, this the citadel, this the altar of the allies. To which indeed they now flee not as before they were wont in seeking back their own things. Not silver, not gold, not clothing, not slaves do they ask back; not the ornaments which have been snatched out of cities and shrines. Inexperienced men fear lest now the Roman people grant these things and so wish them to be done. For we suffer many years now and are silent, when we see all moneys of all nations come to a few men. Which we seem the more to bear with calm mind and to grant, because none of those men pretends, no one labours that his own greed should seem hidden.

\ciceroSection{127} In our most beautiful and most adorned city, what statue, what painted panel is there which has not been taken from conquered enemies and brought here? But the villas of these men are adorned and stuffed with very many and most beautiful spoils of the most faithful allies. Where do you suppose are the moneys of the foreign nations, which now all lack, when you see Athens, Pergamum, Cyzicus, Miletus, Chios, Samos, finally all Asia, Achaia, Greece, Sicily shut up in so few villas? But these things, as I say, your allies now leave aside and neglect, gentlemen. Lest they be despoiled publicly by the Roman people, they have provided by their services and faith. To the greed of a few they could yet, when they could not resist, in some way furnish. But now the means not only of resisting but even of furnishing has been taken away. So they leave aside their own affairs; they do not seek back the moneys --- by which name this trial is called; they leave them off. In this dress now they flee for refuge to you.

\ciceroSection{128} Look, gentlemen, look on the squalor and filth of the allies! Sthenius of Thermae here, with this hair and this dress, with his whole house plundered, makes no mention of your thefts; he asks only himself back from you, nothing more. For wholly by your lust and crime you took him from his fatherland, in which by his many virtues and services he was the chief. Dexo here, whom you see, asks not what publicly at Tyndaris, not what privately you snatched from him, but, wretched, his only son, the best and most innocent, you he begs to give back. He wishes to bring home some comfort to the ashes and bones of his son not from the assessed suits of you for money, but from your calamity. This Eubulides, so old, has now in his exhausted age undertaken so much labour and journey, not that he might recover any of his goods, but that, with the same eyes with which he had seen the bloody neck of his son, he might see you condemned.

\ciceroSection{129} If by Lucius Metellus it had been allowed, gentlemen, the mothers of those wretched men and their sisters were coming. One of whom, when I was approaching Heraclea by night, with all the matrons of that state and with many torches met me. And so --- calling me her safety, naming you her executioner, imploring her son's name --- the wretched woman lay at my feet as if I could rouse her son from the dead. Likewise in the rest of the cities the mothers in advanced age and likewise the small children of the wretched did. Of whom each age called for my labour and industry, your faith and pity.

\ciceroSection{130} So Sicily, gentlemen, has brought to me this complaint above the rest. By tears, not by glory, I have come to this work; that false condemnation, that prison, chains, floggings, axes, the tortures of allies, the blood of the innocent, finally even the bloodless bodies of the dead, the grief of parents and kinsmen, may not be a source of gain to our magistrates. If by your faith and truth, gentlemen, I drive away this fear of Sicily by the condemnation of this man, I shall reckon that enough has been done for my duty, enough for the will of those who have asked this of me.

\ciceroSection{131} Wherefore if perhaps you find anyone who tries to defend this naval charge, let him so defend that he leaves aside those common things which have nothing to do with the case --- that I am ascribing fortune to fault, that I am giving calamity for charge, that I am throwing at him the loss of the fleet, when many brave men in the common and uncertain peril of war have often been struck both on land and at sea. I throw no fortune at you. There is nothing for you to bring forth as the less commodiously managed affairs of the rest. There is nothing for you to gather of fortune the shipwrecks of many. I say the ships were empty; the rowers and sailors discharged; the rest had lived on roots of palms; that a Sicilian was in command of the fleet of the Roman people, a Syracusan over allies and friends always. I say that you, at that very time and in all the days before, were carousing on the shore with little women. I produce the authorities and witnesses of all these things.

\ciceroSection{132} Do I seem to insult fortune in calamity, do I seem to cut off Mars's refuges? But if it is their fortunes, do I seem to throw the chances of war at you or to reproach? Although those who have committed themselves to fortune, who have been at large in her perils and her variety, are wont not to wish fortune thrown at them. But of this calamity of yours fortune was not partner; for men are wont to try the fortune of war in battles, not in banquets. But in that calamity we can say there was not Mars in common, but Venus. Fortune ought not to be thrown at you. Why did you not give to the fortune of the innocent any pardon and place?

\ciceroSection{133} You may also cut off this --- that, because you exacted punishment by the custom of our ancestors and struck with the axe, therefore you are called by me into charge and odium. My charge does not turn on the punishment. I do not deny that any man ought to be struck with the axe. I do not say that fear in the military art, that the strictness of imperium, that the punishment of disgrace ought to be taken away. I confess that not only against allies but even against our citizens and soldiers very often it has been animadverted strictly and vehemently. So this also you may pass over. I show that the fault was not in the captains but in you; I prove that you discharged the rowers and soldiers for a price. This the rest of the captains say; this the allied state of the Netini publicly says; this the Amestratini, this the Herbitenses, this the Hennenses, the Agyrines, the Tyndaritans publicly say. Finally your witness, your general, your rival, your guest-friend Cleomenes says this: that he had gone out onto the land that he might gather soldiers from the land guard at Pachynus and place them in the ships. Which surely he would not have done, if the ships had had their proper number; for the manner of equipped and adorned ships is such that not only could not more, but not even single men, come on board.

\ciceroSection{134} I say besides that those very rest of the sailors had been worn out and ruined by hunger and want of all things; I say either that all were free from fault, or, if fault must be ascribed to one, that in him was the most who held the best ship, the most sailors, the highest command; or, if all had been at fault, that Cleomenes ought not to have been set up as spectator of their death and torture. I say even that, in the very punishment, it was unspeakable that there should be set a wage of tears, a wage of the wound and stroke, a wage of the funeral and burial.

\ciceroSection{135} Wherefore if you should wish to answer me, say this: that the fleet was equipped and adorned, that no marine was lacking, that no oar was drawn empty, that the grain administration was supplied; that the captains lie, that so many such grave cities lie, that even all Sicily lies; that you have been betrayed by Cleomenes, who said he had gone out onto the land that he might lead off soldiers from Pachynus; that spirit, not forces, was lacking to them; that Cleomenes, fighting most fiercely, was left and abandoned by them; that no coin was given to anyone for burial. If you say this, you will be caught. But if you say other things, you will not refute what has been said by me.

\ciceroSection{136} Here will you yet dare to say, "There is in the judges that familiar of mine, there is that paternal friend"? Just as each is most one with whom you have anything, are you not in such a charge most ashamed before him? "He is a paternal friend." If your father himself were judging --- by the immortal gods --- what could he do? When he should say to you these things: "You, in a province of the Roman people, as praetor, when you had to administer a maritime war, remitted the Mamertines for three years the ship which by treaty they owed; you had built among them publicly the largest private cargo-ship at public cost; you raked together moneys from the cities under the name of fleet; you discharged the rowers at a price; you, when a ship was captured by the quaestor and the legate from the brigands, removed the chief pirate from the eyes of all; you struck with the axe those who were said to be Roman citizens, who were recognized by many; you led the pirates off to your own house; you dared in court to bring forth the chief pirate from your home;

\ciceroSection{137} you, in a province so splendid, you, among the most faithful allies, the most honourable Roman citizens, in the fear and peril of the province, lay continuously for many days on the shore at banquets; through those days no one could find you at your house, no one could see you in the forum; you brought the matrons of the allies and friends to those banquets; you placed your praetexta-clad son, my grandson, among such women, that the life of the parent might offer examples of wickedness to that age most slippery and uncertain; you, as praetor in the province, were seen in a tunic and a purple cloak; you, on account of your love and lust, took the command of the ships from the Roman people's legate and gave it to a Syracusan; your soldiers in the province of Sicily lacked fruits and grain; by your luxury and greed the fleet of the Roman people was captured and burned by brigands;

\ciceroSection{138} since Syracuse was founded, the harbour into which no enemy had ever entered --- in it under you as praetor for the first time pirates sailed; nor did you wish to cover all these so many great disgraces by your pretence or by men's forgetfulness and silence, but you also dragged the prefects of the ships, without any cause, from the embraces of their parents, your guest-friends, to death and torture. Nor did the recollection of my name in the mourning and tears of the parents soften you. The blood of innocent men was for you not only a pleasure but even a gain!" --- if your father were saying these things to you, could you ask his pardon, could you ask him to forgive you?

\ciceroSection{139} Enough has been done for the Sicilians, enough for duty and connection, enough for our promise and our undertaking. The remaining cause, gentlemen, is that one which is not now received but inborn, not brought to me but deeply fixed and set in my mind and feeling. Which has to do not with the safety of the allies, but with that of Roman citizens --- that is, with the life and blood of each one of us. In which do not, gentlemen, await proofs from me as if anything were doubtful: all I shall say will be so clear that to prove them I could call all Sicily as witness. For a kind of frenzy, the companion of crime and audacity, has so pressed his unbridled mind and his harsh nature with such senselessness that he never hesitated, in an open assembly, to bring forth against Roman citizens those punishments which are established for slaves convicted of crime.

\ciceroSection{140} How many he flogged with rods --- why should I recall? Only this most briefly, gentlemen, I say: there was no distinction at all of citizenship under this praetor in this kind. So now by custom against the bodies of Roman citizens, even without his nod, the very hand of the lictor was carried. Can you deny this, Verres: that in the forum at Lilybaeum in the largest assize Gaius Servilius, a Roman citizen of the Panhormitan assize, an old businessman, before the tribunal at your feet was struck to the ground by rods and floggings? Dare to deny this first, if you can. There was no man at Lilybaeum who did not see; no man in Sicily who did not hear. I say a Roman citizen, worn out by blows from your lictors, fell before your eyes.

\ciceroSection{141} But for what cause, immortal gods! Although I do wrong to the common cause and the right of citizenship; for as if there could be any cause why this should rightly happen to any Roman citizen, so I ask what was the cause in Servilius. Forgive me in this one matter, gentlemen; for in the rest I shall not greatly look for causes. He had spoken somewhat freely about this man's wickedness and dishonesty. As soon as it was reported to him, he ordered the man to give bail to a Venusian slave to come to Lilybaeum. He gives bail; he comes to Lilybaeum. He began to compel him --- when no one was bringing a charge, no one demanding it --- to enter into a sponsion of one thousand sesterces with his lictor: \textit{unless he were not making gain by his thefts}. He said he would give recoverers from his own staff. Servilius both refused and pleaded that, before unjust judges and with no adversary, a capital trial should not be set up against him.

\ciceroSection{142} While he was speaking these very things, six lictors stand round him --- most strong and most practised in beating and flogging men. They cut him most fiercely with rods. Finally the nearest lictor (of whom I have often before spoken), Sextius, with his rod turned around, began most vehemently to thrust at the wretched man's eyes. So he, when blood had filled his face and eyes, fell. They none the less, while he lay, were thrusting at his sides, that at last he might say he gave the bail. So affected, then taken up there as if dead, very shortly afterwards he died. But that man of Venus, overflowing with every charm and grace, out of his goods set up a silver Cupid in the temple of Venus. So even the fortunes of men he was abusing for the night-vows of his lusts.

\ciceroSection{143} For why should I speak of the rest of the punishments of Roman citizens one by one rather than as a class and altogether? That prison which was made by the most cruel tyrant Dionysius at Syracuse, which is called the stone-quarries, in his command was the home of Roman citizens. As each had hurt his mind or eyes, he was at once thrown into the stone-quarries. I see that this seems unworthy to all, gentlemen; and that I knew already in the previous hearing, when the witnesses were saying these things. For you think that the rights of liberty ought to be kept not only here, where there are tribunes of the plebs, where there are the rest of the magistrates, where the forum is full of trials, where the authority of the senate is, where the standing of the Roman people and the throng are; but you decide that wherever in the lands and among the nations the right of Roman citizens is violated, that has to do with the common cause of liberty and dignity.

\ciceroSection{144} Did you dare to shut up so great a number of Roman citizens in the prisons of foreign evildoers and criminals, of brigands and enemies? Never did the trial come into your mind, never the public meeting, never this such great throng which now looks upon you with the most unjust and most hostile spirit? Never did the dignity of the absent Roman people, never the very sight of this multitude move before your eyes and mind? You never thought you would return into the sight of these men, never come into the forum of the Roman people, never fall under the power of the laws and the courts?

\ciceroSection{145} But what was that lust of exercising cruelty, what cause of taking up so many crimes? None, gentlemen, save a new and singular method of plunder. For as those whom we have received from the poets, who are said to have set themselves at certain maritime bays, or to have held certain promontories or jutting rocks, that they might kill those who were brought there by their ships --- so this man, hostile to all the seas, hung over from all parts of Sicily. Whatever ship had come from Asia, what from Syria, what from Tyre, what from Alexandria, was at once held by sure informers and watchmen. All the passengers were thrown into the stone-quarries; their cargoes and goods were carried to the praetor's house. There was at large in Sicily, after a long interval, a second --- not that Dionysius nor Phalaris (for that island once bore many cruel tyrants) --- but a kind of new monster, out of that old monstrousness which is said to have been at large in those same places.

\ciceroSection{146} For I do not think that Charybdis or Scylla was so hostile to sailors as this man was in the same strait. He was even the more hostile, because he had girt himself with many more and more monstrous dogs --- a second Cyclops, much fiercer; for this man besieged the whole island, while the other is said to have held only Aetna and that part of Sicily. But what cause was then offered by him, gentlemen, of this so unspeakable cruelty? The same which now will be brought forward in defence. Whoever had come up to Sicily a little fuller, them he was saying were Sertorian soldiers and were fleeing from Dianium. They, to plead off the danger, brought forth: some Tyrian purple; others incense and perfumes and linen clothing; others gems and pearls; some Greek wines and Asiatic slaves for sale --- so that one could understand from the wares from what places they were sailing. They had not foreseen that those very things were causes of peril for them, by which proofs they reckoned they were using to safety. For this man was saying that they had got these things out of partnership with pirates. He was ordering them to be led to the stone-quarries. He saw to it that their ships and cargoes were diligently kept.

\ciceroSection{147} With these institutions, when the prison was now full of merchants, then those things were happening which you have heard L. Suettius, a Roman knight, a most choice man, say, and which you shall hear from the rest. Necks were being broken in the prison most unworthily of Roman citizens, so that now that voice and that imploring, "I am a Roman citizen" --- which often in many of the most distant lands has brought help and safety among barbarians --- brought to them a more bitter death and a quicker punishment. What is it, Verres? What do you think to answer to these things? That I am lying, that I am feigning anything, that I am swelling the charge? Do you dare to say any of these things to those defenders of yours? Give me, please, from his own bosom the records of the Syracusans, which he himself thinks were drawn up at his own discretion. Give me the prison register, which is most diligently drawn up: on what day each was given into custody, on what he died, on what he was killed. \textsc{Records of the Syracusans}.

\ciceroSection{148} You see Roman citizens thrown by herds into the stone-quarries; you see in a most unworthy place the heaped-up multitude of your citizens. Now look for traces by which their exit from that place might appear. There are none. Did all die? If he could so defend, yet credit would not be given to that defence. But there stands written in those same records what this barbarous and dissolute man could neither attend to nor understand: \textit{edikaiothesan}, he says --- that is, as the Sicilians speak, "they were punished and killed."

\ciceroSection{149} If some king, if any state of the foreign nations, if any nation had done such a thing against Roman citizens, would we not avenge it publicly, would we not pursue it by war? Could we let go this wrong and ignominy of the Roman name unavenged and unpunished? How many wars and how great do you think our ancestors took up, because Roman citizens were said to have been afflicted with wrong, because shippers had been held back, because merchants had been despoiled? But I do not now complain of those held back; despoiled, I think, must be borne. With ships, slaves, goods taken away, I prove that merchants were thrown into chains and that in the chains Roman citizens were killed.

\ciceroSection{150} If I were saying these things among Scythians --- not here in such a great multitude of Roman citizens, not before senators, the most choice of the state, not in the forum of the Roman people about so many and so bitter punishments of Roman citizens --- yet I should move even the spirits of barbarian men. For so great is the amplitude of this empire, so great the dignity of the Roman name among all nations, that this cruelty against our people seems to be granted to no one. Now shall I think you have any safety, any refuge, when I see you entangled in the strictness of the judges, hemmed in by the throng of the Roman people?

\ciceroSection{151} If, by Hercules, what I see cannot happen --- if you should slip out of these snares and disentangle yourself by some way and reasoning --- you must fall into those greater nets, in which you must necessarily be finished and cut to pieces by the same me from a higher place. To whom even if I should wish to grant what he defends, yet that very false defence ought to be no less ruinous to him than my true accusation. For what does he defend? He says that he caught those fleeing from Spain and inflicted punishment on them. Who allowed you this? By what right did you do it? Who has done the same? How was it lawful for you to do it?

\ciceroSection{152} We see the forum and the basilicas full of those men, and we see them with calm mind. For not troublesome is that outcome of civil dissension and (whether of frenzy or of fate or of calamity) in which at least it is allowed to keep the rest of the citizens unhurt. Verres, that old betrayer of his consul, transferrer of his quaestorship, diverter of public money, took on himself such authority in the commonwealth that to those men whom by the senate, by the Roman people, by all the magistrates it was lawful to be at large in the forum, in voting, in this city, in the commonwealth --- to all those he was setting forth a bitter and cruel death, if fortune should bring them to any part of Sicily.

