% In C. Verrem Actio Prima (in-verrem-actio-prima) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus, Clark text).
% Style guide: STYLE.md.

\ciceroSection{1} What was most to be wished for, gentlemen, and what alone bore most to the calming of the resentment against your order and the infamy of the courts, that thing seems --- not by human counsel, but as if divinely --- to have been given and offered to you at the highest moment of the commonwealth. For the opinion has by now grown old, ruinous to the commonwealth and dangerous to you, which has spread by everyone's talk not only among us but among foreign nations: that in these courts as they now are, no man with money, however guilty, can be condemned.

\ciceroSection{2} Now, at the very crisis of your order and your courts, when there are men ready to fan this resentment against the senate by public meetings and by laws, the defendant Gaius Verres has been brought into court --- a man already condemned by his life and deeds in everyone's opinion, but acquitted, by his own hope and boast, by the magnitude of his money. To this case, gentlemen, with the highest goodwill and expectation of the Roman people, I have come forward as prosecutor; not to increase the resentment against the order, but to come to the rescue of our common infamy. For I have brought into court a man through whom you may restore the lost reputation of the courts, return into favour with the Roman people, give satisfaction to foreign nations: a robber of the treasury, a harrier of Asia and Pamphylia, a brigand of city law, the stain and ruin of the province of Sicily.

\ciceroSection{3} If on this man you give a strict and conscientious verdict, the authority which ought to remain in you will hold fast. If, however, his vast wealth shall break through the conscience and truth of the courts, I shall yet attain this much --- that the verdict will appear to have failed the commonwealth, rather than the defendant to have failed the jurors, or the prosecutor the defendant. To confess about myself, gentlemen, although many ambushes were laid for me by Gaius Verres, by land and sea, some of which I avoided by my diligence, others I beat off through the zeal and devotion of my friends, never have I seemed to face so great a peril, never have I been so shaken with fear, as I am now in the trial itself.

\ciceroSection{4} The expectation of my prosecution and the gathering of so great a multitude --- by which I am profoundly disturbed --- do not move me as much as those wicked plots which he is at one and the same time trying to lay against me, against you, against Manius Glabrio, against the Roman people, against the allies, against foreign nations, against our order, against the very name "senator." He keeps saying: "Those should be afraid who have stolen what was enough only for themselves; I have torn off so much that there is enough for many --- nothing is so sacred that money cannot violate it, nothing so well fortified that money cannot storm it."

\ciceroSection{5} If he were as obscure in action as he is bold in attempting, perhaps in some matter he might at some time have deceived us. But it has been very fortunate for us that, with his incredible audacity, a singular stupidity is joined; for as he was open in seizing money, so in the hope of corrupting the court he has made his plans and attempts plain to all. He says he has been afraid in his life only once --- when he was first made a defendant by me; because, while he was fresh from the province, and was burning with resentment and infamy not new but old and lasting, he had hit upon an unfavourable moment for corrupting the court.

\ciceroSection{6} So when I had asked for a most short period for inquiry in Sicily, he found someone to ask for himself, for inquiry in Achaia, a period two days shorter --- not that he might effect by his diligence and industry what I have achieved by my labour and waking nights (indeed his Achaean inquisitor never even reached Brundisium, while I went round all of Sicily in fifty days, in such a way that I learned of all the wrongs done to peoples and to private men) --- but so that it might be plain to anyone that this man was sought out by him not to bring his defendant in, but to consume my time.

\ciceroSection{7} Now this most audacious and senseless man considers this. He understands that I come into court so prepared and equipped that I shall fix his thefts and disgraces not only on your ears but on the eyes of all. He sees that there are many senators witnesses of his audacity, many Roman knights, and besides numerous citizens and allies whom he himself has wronged signally; he sees also that so many heavy embassies have come from cities most friendly to him, with their public authorities.

\ciceroSection{8} Since this is so, his estimate of all good men is so low, his judgement of the senate's courts so wrecked and undone, that he openly declares: "Not without cause have I been greedy of money, since I find that there is so great a defence in money. I have bought" --- the most difficult thing of all --- "the very moment of my trial, that I might more easily buy other things afterwards. Since I could in no way evade the force of the charges, at least I shall escape the storm of the moment."

\ciceroSection{9} If he had placed any hope not only in the case but in any honourable defence, in anyone's eloquence or favour, surely he would not be gathering and grasping after all these things; he would not so despise and contemn the senatorial order that, at his discretion, one should be chosen out of the senate to be made defendant, who, while he was getting ready what was needed, should plead his case before him.

\ciceroSection{10} What he hopes for, what he aims at by these tactics, I easily perceive. But why he should be so confident he can accomplish anything, with this praetor and this council, I cannot understand. This one thing I do understand --- which the Roman people judged in the rejection of jurors --- that he was furnished with this hope: that he had placed all his reckoning of safety in money, and once that defence was torn from him he supposed nothing would be of help to him. For what genius is so great, what capacity for speech or richness of words, that it could in any part defend a life convicted of so many vices and disgraces, already long since condemned by the will and judgement of all?

\ciceroSection{11} To pass over the stains and disgraces of his youth, his quaestorship --- the first step of public office --- what does it have but Gnaeus Carbo despoiled by his own quaestor of public money, a consul stripped and betrayed, an army deserted, a province abandoned, the bond and religion of the lot violated? His legateship was the ruin of all Asia and Pamphylia: in those provinces he plundered many houses, very many cities, every shrine; then with respect to Gnaeus Dolabella he renewed and refreshed his old quaestorian crime, when, the man to whom he had been legate and acting quaestor, he both led into resentment by his own misdeeds and, in the very dangers, not only deserted but assailed and betrayed.

\ciceroSection{12} His city praetorship was the plundering of sacred buildings and public works; at the same time, in administering justice, the assignment and donation of inheritances and possessions against everyone's institutions. And by far the most numerous and greatest monuments and signs of all his vices he has set up in the province of Sicily, which he in three years so harried and ruined that it cannot in any way be restored to its former state, but scarcely seems able through many years and innocent praetors to be in some measure recovered.

\ciceroSection{13} Under this praetor the Sicilians could keep neither their own laws, nor our decrees of the senate, nor the rights common to all. In Sicily each man holds only as much as he either escaped the inadvertence of this most greedy and most lustful of men, or what was left over from his sating. For three years no matter was decided except at his nod; no inheritance was so ancestral and family that it was not taken from a man's possession by his command. Innumerable sums of money were extorted from the goods of the farmers by a new and unspeakable practice; the most faithful allies were reckoned in the number of enemies; Roman citizens were tortured and killed in the manner of slaves; the most guilty men were freed in court for money; the most honourable and upright in their absence were made defendants, condemned without their case being heard, and exiled. The most strongly fortified harbours, the greatest and safest cities, were thrown open to pirates and brigands; the seamen and soldiers of the Sicilians, our allies and friends, were starved to death; the best and most strategic fleets, with great disgrace to the Roman people, were lost and destroyed.

\ciceroSection{14} This same praetor stripped and laid bare the most ancient monuments --- some of them the gifts of the wealthiest kings, who wished them to be ornaments to the cities; others of our own commanders, which the victors either gave or returned to the Sicilian states. Nor did he do this only with statues and public ornaments. He plundered all the shrines consecrated by the most sacred religion: in short, he left the Sicilians not a single god if it seemed somewhat better worked or more antique. As for his rapes and his disgraces, modesty deters me from listing his unspeakable lusts; and at the same time I do not wish, by mention, to add to the calamity of those who were not allowed to keep their children and wives untouched from his shamelessness.

\ciceroSection{15} "But these things were done by him in such a way that they are not all known." There is, I think, no man who has heard his name who could not also recount his unspeakable deeds; so that I should fear rather to leave many charges out than be thought to invent any against him. For I do not think this multitude which has gathered to listen wished to learn the case from me, but rather to recognise with me what they already know. Since this is so, this senseless and ruined man fights with me by another method. He does not match an eloquence of someone against me; he does not lean on anyone's favour, authority, or influence. He pretends to trust in these things; but I see what he is doing, and he does not do it most secretly. He throws empty names of nobility in my way --- that is, names of arrogant men, who do not so much hinder me by being noble as help me by being known. He pretends to trust in their protection while in the meantime he has long been working some other thing.

\ciceroSection{16} What hope he now has in his hands and what he is plotting, I will briefly set out for you, gentlemen. But first learn, please, how the affair was set up by him from the start. As soon as he came back from the province, the buying-off of this trial was made for a great sum of money. The transaction stood firm in agreement and bargain up to the point at which the jurors had been rejected. After the rejection of jurors had taken place --- because both in his lot the fortune of the Roman people had defeated this man's hope, and in the rejection of jurors my diligence had defeated their shamelessness --- the whole bargain was renounced. The matter stood splendidly.

\ciceroSection{17} Lists of your names and of this council were in everyone's hands. No mark, no shading, no smudge could be smeared on these votes. When suddenly, from cheerful and lively, he had become so humble and downcast that he seemed condemned not only to the Roman people but even to himself. But behold, suddenly, in these few days after the consular elections, those same old plans are taken up again with greater money; the same plots against your reputation and the fortunes of all are got ready through the same men. Which matter, gentlemen, was first laid open to us by a very slight argument and indication. After the entrance of suspicion was opened, we came without any error to all of their inmost plans.

\ciceroSection{18} For when Hortensius, consul-elect, was being escorted home from the Campus by a very great crowd, Gaius Curio met that crowd by chance --- a man whom I should rather name out of honour than out of insult; for I shall say what he, if he had not wished it to be mentioned, would not have said in so great a gathering so openly and in public; what nevertheless will be said by me step by step and cautiously, that account may be taken both of our friendship and of his dignity.

\ciceroSection{19} Curio sees Verres in the crowd at the very arch of Fabius. He addresses the man and congratulates him with the loudest voice. To Hortensius himself, who had been made consul, to his close kinsmen and intimates who were present, he says not a word. With this man he stops; this man he embraces; this man he tells to be without anxiety. "I bring you the news," he says, "that you have been acquitted by today's elections." When so many honourable men had heard this, it was at once reported to me; on the contrary, every man who had seen me told it. To some it seemed shameful; to others ridiculous: ridiculous to those who thought his case rested on the credibility of witnesses, on the substance of the charges, on the power of the jurors, not on consular elections; shameful to those who looked deeper and saw that this congratulation looked toward corrupting the trial.

\ciceroSection{20} For these were their reasonings, this is what the most honourable men were saying among themselves and to me: openly and plainly, there are no more courts. The defendant who, the day before, was thinking himself already condemned --- after his defender has been made consul, is acquitted? What of it? That all Sicily is at Rome --- all the Sicilians, all the businessmen, all the public and private letters --- shall this not weigh? Nothing, against the consul-elect's wishes. What? Will the jurors not follow the charges, the witnesses, the estimation of the Roman people? No: everything will turn on the power and discretion of one man. I will speak truly, gentlemen: this matter shook me violently. For each best man was speaking thus: "This man indeed will be snatched from you; but we shall not hold the courts much longer; for who will be able to refuse, after Verres is acquitted, the transfer of the courts?"

\ciceroSection{21} It was painful to all; not so much the sudden joy of that ruined man, as the new congratulation by a most distinguished one, moved them. I wished to dissemble that I bore it ill; I wished to cover the pain of my mind with my face, and to hide it by silence. But behold, on those very same days, when the praetors-elect were drawing lots and it had fallen to Marcus Metellus to preside over the extortion court, it was reported to me that so great a congratulation had been given to Verres that he sent slaves home to bring the news to his wife.

\ciceroSection{22} This too did not at all please me; nor did I altogether understand what was to be feared from that lot. This one thing I learned from sure men, from whom I have learned everything: that several boxes with Sicilian money had been transferred from a certain senator to a Roman knight; that of these about ten boxes had been left with that senator in the name of my elections; that the canvassers of all the tribes had been called to him by night.

\ciceroSection{23} Of these one, who reckoned that he ought to do everything for my sake, came to me on that very night. He showed me the language Verres had used: that he had recalled how generously he had treated them already before, when he himself was canvassing for the praetorship, and at the recent consular and praetorian elections; then that he had immediately promised them as much money as they wished, if they had thrown me out from the aedileship. Some, he said, had refused to dare; others had answered that they did not think it could be carried through. A bold friend, however, was found from the same family --- Quintus Verres of the Romilian tribe, of the best discipline of the canvassers, the pupil and friend of his father --- who, with five hundred thousand sesterces deposited, undertook to do it. There were nevertheless not a few who said they would join in. And so he warned me, with truly friendly feeling, to take great care.

\ciceroSection{24} I was in the greatest matters at the same and a most short time. The elections were pressing; in the elections themselves I was being attacked with great money. The trial was at hand; in that business too the Sicilian boxes were threatening. I was deterred from doing what concerned the trial freely by fear of the elections; I could not, on account of the trial, devote my whole mind to my candidacy. Finally, to threaten the canvassers was no use, because I saw they understood I would be tied down and bound by this trial.

\ciceroSection{25} At this very same moment I hear that the Sicilians had been notified --- first by Hortensius --- to come to him at his house. The Sicilians in this matter were free indeed, who, when they understood for what reason they were summoned, did not come. In the meantime our elections (over which Verres considered himself master, as over the rest this year) began to be held. He goes scampering, this powerful man, with his courteous and obliging son around the tribes; he calls upon and meets all his father's friends --- that is, the canvassers. When this was understood and noticed, the Roman people did most willingly arrange that, by the man whose riches could not have led me from my faith, by the same man's money I should not be cast down from my honour.

\ciceroSection{26} After I was relieved of the great anxiety of the candidacy, with mind much more empty and free I began to do and think nothing else but about the trial. I find, gentlemen, that these plans had been entered into and established by them: that, by whatever method possible, the matter should be drawn out so that the case might be pleaded before Marcus Metellus the praetor. In that there were these advantages: first, Marcus Metellus was very friendly; then there was Hortensius the consul, and not Hortensius alone, but also Quintus Metellus --- consider how friendly he is to Verres. For he gave a foretaste of his goodwill of such a kind that he seems already to have repaid him for past favours.

\ciceroSection{27} Or did you think I would be silent on so great matters, and that I would consult anything in such great peril for the commonwealth and my own reputation rather than my duty and dignity? The other consul-elect summons the Sicilians; some come, since Lucius Metellus is praetor in Sicily. With them he speaks thus: "I am consul; one brother of mine holds the province of Sicily, the other will preside over the extortion court; many provisions have been made that no harm can come to Verres."

\ciceroSection{28} What is corrupting a court, please, Metellus, if this is not: by frightening the witnesses --- especially the Sicilians, fearful and afflicted men --- not only by your authority, but by consular dread and the power of two praetors? What would you do for an innocent man and a kinsman, when for the sake of a most ruined and most alien man you depart from your duty and dignity, and let it appear true to anyone who does not know you, what he keeps saying?

\ciceroSection{29} For they said Verres said this: that you had been made consul not by destiny, like the rest of your family, but by his work. So there will be two consuls and the presiding judge by his will. "We shall not only escape" --- he says --- "a man over-diligent in inquiry, over-attentive to the people's reputation, Manius Glabrio. We shall also have this added to our advantage. Marcus Caesonius is on the bench, the colleague of our prosecutor, a man tested and known in the giving of judgements, whom it least suits us to have on a council we are trying somehow to corrupt --- because, when he was a juror in Junius's council, he not only took ill that disgraceful crime, but brought it into the open. From the first of January we shall not have him as juror. ---

\ciceroSection{30} Quintus Manlius and Quintus Cornificius, two most strict and upright jurors, since they will then be tribunes of the plebs, we shall not have as jurors. Publius Sulpicius, a stern and upright juror, must enter office on the Nones of December. Marcus Crepereius from that fiercest equestrian family and discipline; Lucius Cassius from a family most strict in everything, and especially in giving judgement; Gnaeus Tremellius, a man of the highest religion and diligence --- these three veteran tribunes of the soldiers are designated for office and will not be jurors from the first of January. We shall also draw a substitute for Marcus Metellus's place, since he will be presiding over this very court. So, after the first of January, with the praetor and almost all the council changed, we shall escape the great threats of the prosecutor and the great expectation of the trial, at our own discretion and pleasure."

\ciceroSection{31} Today is the Nones of August. At the eighth hour you have begun to assemble. This day they do not even count. Ten days remain before the votive games which Gnaeus Pompey is going to give; these games will take fifteen days; then immediately the Roman games will follow. So, with about forty days interposed, only then do they think they will have to reply to what we shall have said. Then they think they will easily, by speeches and excuses, draw the matter out to the games of victory; then with these the plebeian games will be joined, after which there will be either no days or very few left for action. So, with the prosecution worn out and grown cold, the matter will come fresh to Marcus Metellus the praetor. Whom, if I had distrusted his good faith, I would not have kept as juror.

\ciceroSection{32} As things stand, however, I am of this mind: that I should rather have him as juror than as praetor; and that I should rather entrust to him under oath my voting tablets than to others not under oath. Now, gentlemen, I consult you about what you think I ought to do. For you will surely give me silently the same counsel which I myself understand I must necessarily take. If I use my own lawful time for speaking, I shall enjoy the fruit of my labour, my industry, and my diligence; and by this prosecution I shall achieve that no one within human memory shall be seen to have come to a trial more prepared, more vigilant, more composed. But in this very praise of my industry the highest danger is that the defendant should slip away. What can be done? It is not, I think, hidden or obscure.

\ciceroSection{33} Let us reserve for other times that fruit of praise which could be reaped from a continuous oration. Now let us prosecute the man with documents, with witnesses, with private and public letters and authorities. The whole matter will be between you and me, Hortensius. I will speak openly. If I thought you were going to fight me with speech and dilution of charges in this case, I too would spend my labour on prosecuting and unfolding the charges. Now, since you have set yourself to fight against me not so much from your own nature as from his moment and case --- and maliciously --- it is necessary to oppose this kind of plan with some counsel of mine.

\ciceroSection{34} Your reasoning is to begin to answer me after the second set of games; mine is to adjourn before the first games. So your reasoning will be reckoned cunning, and this counsel of mine necessary. But what I had begun to say --- that the matter is between you and me --- is of this kind. When I had taken up this case at the request of the Sicilians, and had thought it ample and splendid for me that they wished to test my good faith and diligence who had tested my innocence and abstinence, then, having undertaken the business, I set before myself something greater: in which the Roman people might perceive my goodwill toward the commonwealth.

\ciceroSection{35} For it did not seem to me at all worthy of my industry and effort that this man, already condemned in the judgement of all, should be summoned by me into a trial --- unless that intolerable power of yours, and that greed which through these years you have used in certain trials, should be brought to bear even in the case of this desperate man. Now, indeed, since this whole tyranny and royal rule of the courts pleases you so greatly, and there are men whom their own lust and infamy neither shames nor wearies, who, as if of set purpose, seem to rush against the hatred and offence of the Roman people --- I declare that I have undertaken what is perhaps a great burden, and dangerous to me, but worthy of having all the sinews of my age and industry strained to it.

\ciceroSection{36} Since the whole order is being pressed by the wickedness and audacity of a few, and weighed down by the infamy of the courts, I declare myself a hostile prosecutor of this kind of men --- a hateful, untiring, bitter adversary. This I take upon myself; this I demand for myself --- to act in office, to act from that place where the Roman people, from the first of January, has wished me to act with it on the commonwealth and on dishonest men. This service of my aedileship I promise the Roman people, the most ample and most beautiful. I warn, I forewarn, I declare in advance: those who are accustomed to deposit, to receive, to take charge of, to promise, to be go-betweens or interpreters in corrupting the court; those who have professed for this purpose either their power or their shamelessness --- let them keep their hands and their minds in this trial from this unspeakable crime.

\ciceroSection{37} Hortensius will then be consul with the highest command and power; I shall be aedile --- that is, somewhat more than a private citizen. Yet the matter on which I promise to act is of such a kind, so welcome and pleasing to the Roman people, that the consul himself will, in this case, beside me, seem (if it can be) less even than a private citizen. Everything will not only be mentioned, but, with definite cases set out, will be acted upon: things which in the ten years since the courts were transferred to the senate have been done in giving judgements unspeakably and disgracefully.

\ciceroSection{38} The Roman people will learn from me why, when the equestrian order was sitting in judgement, for nearly fifty continuous years, on no Roman knight as juror was even the slightest suspicion of money taken for a judgement established; but why, after the courts were transferred to the senatorial order and the power of the Roman people over each one of you was abolished, Quintus Calidius, when condemned, said that an ex-praetor could not honourably be condemned for less than three million sesterces; why, when the senator Publius Septimius was condemned under Quintus Hortensius the praetor on extortion, the case was assessed under that very heading --- that he had received money for a judgement.

\ciceroSection{39} What in Gaius Herennius, what in Gaius Popilius, both senators, both condemned for embezzlement; what in Marcus Atilius, who was condemned for treason --- this has been made plain: that they took money for a judgement; that there were senators found who, when Gaius Verres was urban praetor and was drawing lots, came forward to convict an unheard defendant; that a senator was found who, being a juror, in the same trial took money both from the defendant to distribute to the jurors, and from the prosecutor to convict the defendant.

\ciceroSection{40} How shall I lament that ruin, that ignominy, that calamity of the whole order: that this was done in this state, when the senatorial order was sitting in judgement, that the votes of sworn men were marked with different-coloured signs? All these things I promise to do diligently and severely. With what mind, do you suppose, shall I be, if I perceive in this very trial that something has been violated and committed in some similar fashion? --- especially when I can make plain by many witnesses that Gaius Verres in Sicily said many times in the hearing of many that he had a powerful man in whose confidence he was plundering the province; that he was not seeking money for himself alone; that he had so apportioned that triennium of his Sicilian praetorship that he reckoned it would be splendid for him if he turned the gain of one year to his own use; the second to his patrons and defenders; the third, that most lavish and most profitable year, he reserved entirely for the jurors.

\ciceroSection{41} From which it occurs to me to say what, when I had recently mentioned it before Manius Glabrio in rejecting the jurors, I understood the Roman people to be greatly moved by: that I think foreign nations will send embassies to the Roman people for the law on extortion and the trial to be abolished. For if there are no trials, they think, each man will take only as much as he reckons enough for himself and his children. Now, with trials of this kind, each man takes as much as will suffice for him, his patrons, advocates, the praetor, and the jurors --- and this is, indeed, infinite. He says he can satisfy the desire of the most greedy man, but cannot satisfy the victory of the most guilty.

\ciceroSection{42} What memorable courts! What splendid reputation of our order, when the allies of the Roman people are unwilling that trials on extortion --- which were established by our ancestors for the sake of the allies --- should take place at all! Or would Verres ever have had a good hope of himself, unless he had drunk in a bad opinion of you in his mind? For which reason he ought, if it can be, to be even more hated by you than he is by the Roman people, when he supposes that, in greed, in crime, in perjury, you are like himself.

\ciceroSection{43} For this place, by the immortal gods, gentlemen, take care and provide. I warn and proclaim what I understand: that this moment has been given you divinely, that you may free the whole order from hatred, resentment, infamy, and disgrace. There is no longer any thought of strictness in the courts, no scruple, no longer is there even thought to be any "court" at all. So we are despised and looked down on by the Roman people. We are now scorched with serious and lasting infamy.

\ciceroSection{44} For no other cause did the Roman people seek the tribunician power back with such zeal. When it was demanding it, in word it seemed to be demanding the tribunate, but in fact it was demanding the courts. Nor did this escape Quintus Catulus, that wisest and most distinguished man, who, when Gnaeus Pompey, that bravest and most illustrious man, was bringing forward the matter of the tribunician power, and Catulus was asked his opinion --- with the highest authority used this opening: that the senators were maintaining the courts badly and disgracefully; if they had wished to satisfy the estimation of the Roman people in giving judgements, men would not so greatly have desired the tribunician power.

\ciceroSection{45} Finally Gnaeus Pompey himself, when, as consul-elect, he held his first public meeting at Rome, where, what was most awaited, he showed that he would restore the tribunician power, there arose a noise in that meeting and a welcome murmur of approval. The same man at the same meeting, when he had said that the provinces had been plundered and harassed, that the courts were being conducted shamefully and disgracefully, and that he wished to provide for and consult that matter --- then, indeed, not by murmur but by the loudest cry the Roman people signified its will.

\ciceroSection{46} Now men are on the lookout. They are observing how each one of us behaves in retaining religion and preserving the laws. They see that, since the tribunician law, only one senator --- and a most insignificant one --- has been condemned. Which, although they do not censure, yet they do not greatly have anything to praise; for there is no praise in being upright where there is no one who can or tries to corrupt.

\ciceroSection{47} This is the trial in which you will give judgement on the defendant; the Roman people will judge you. In this man it will be settled whether, with senators sitting as jurors, the most guilty and most moneyed man can be condemned. Then he is the kind of defendant in whom there is nothing but the highest crimes and the greatest sum of money: so that, if he is acquitted, no other suspicion can remain than that which is most disgraceful. Not by favour, not by kinship, not by other rightly done deeds, not finally by any moderate vice, will so many and so great vices of his be seen to be lightened.

\ciceroSection{48} Finally, gentlemen, I shall conduct the case in such a way --- I shall produce things so notorious, so witnessed, so great, so manifest --- that no one will try to entreat you out of favour to acquit him. I have besides a definite path and method by which I can investigate and follow up all their attempts. The matter will be conducted by me in such a way that, in all their plans, not only the ears of all but even the eyes of the Roman people shall seem to be present.

\ciceroSection{49} You can wipe out and lift away from this order the disgrace and infamy now contracted for several years. It is agreed among all that, since these courts (with which we now use) were established, no council has been of this splendour and dignity. If anything goes wrong here, all men will think that no longer must others more suitable be sought from the same order (which is impossible) but a different order altogether for giving judgements.

\ciceroSection{50} Wherefore I first beg this from the immortal gods, what I seem to myself to hope for: that in this trial no dishonest man may be found beyond the one who has long since been found. Then, if there shall have been more dishonest men, I assure you, gentlemen, I assure the Roman people, life shall fail me sooner than the force and perseverance of pursuing their dishonesty.

\ciceroSection{51} But what I promise will be strictly pursued by my labour, dangers, and enmities, when the disgrace shall have been committed --- you, Manius Glabrio, can prevent from happening at all by your authority, wisdom, diligence. Take up the cause of the courts. Take up the cause of strictness, integrity, faith, religion. Take up the cause of the senate, that, having been approved by this trial, it may be in praise and in favour with the Roman people. Consider where you stand, what you owe to the Roman people, what you owe to your ancestors. See to it that your father's lex Acilia comes to mind, by which law the Roman people used the best courts and the strictest jurors against extortion.

\ciceroSection{52} The greatest authorities stand around you which will not let you forget your domestic praise; which day and night will remind you that your father was a most brave man, your grandfather a most wise man, your father-in-law a most weighty man. Wherefore, if you take from your father Glabrio his force and keenness in resisting the most audacious men; if from your grandfather Scaevola his prudence in foreseeing the snares that are being prepared against your reputation and that of these gentlemen; if from your father-in-law Scaurus his constancy, that no one can shift you from a true and certain opinion --- the Roman people will understand that, with a most upright and most honourable praetor and a chosen council, the magnitude of money in a guilty defendant has had more weight in suspecting the charge than in calculating his safety.

\ciceroSection{53} I am resolved not to allow that, in this case, our praetor and council shall be changed. I will not suffer the matter to be drawn out to that time when those whom even the slaves of the consuls-elect have not been able to move (when by a new precedent they were summoning them all together) the lictors of the consuls themselves shall summon; that men in distress, before allies and friends of the Roman people, now slaves and suppliants, shall not only by their command lose their right and all their fortunes, but also not have the power of lamenting their right.

\ciceroSection{54} I shall surely not allow that, my speech once finished, with forty days interposed, then at last we should be answered, when our prosecution shall have been brought into the oblivion of long delay. I will not let this matter be judged when this great gathering of all Italy shall have left Rome, which has assembled at one moment from everywhere on account of the elections, the games, and the census. The fruit of praise and the danger of offence in this trial belongs to you, the labour and care to me, the knowledge of what is being done and the memory of what each man has said belongs (I think) to all.

\ciceroSection{55} I shall do this not as something new, but as something done before by those who are now the leading men of our state: I shall use witnesses at once. This you will recognise as new from me, gentlemen: that I shall set up the witnesses in such a way as to unfold the whole charge; when by interrogation, by arguments and by speech, I shall have established it, then I shall fit the witnesses to the charge --- so that there shall be nothing different between the customary prosecution and this new one, except that in the former, when everything has been said, the witnesses are produced; here, on each particular thing they will be produced, so that they too shall have the same opportunity of being interrogated, of arguing, and of speaking. If there is anyone who desires a continuous oration and prosecution, he will hear it in the second hearing. Now what we are doing --- because we are doing it for the reason that we may meet their malice with our counsel --- let him understand is necessary.

\ciceroSection{56} This will be the prosecution in this first hearing. We say that Gaius Verres, when he had done many things wantonly, many cruelly toward Roman citizens and allies, many things impiously toward gods and men, has besides taken from Sicily forty million sesterces against the laws. This by witnesses, by private documents, by public records, by authorities, we will so make plain to you that you may decide that, even if we had had time at our convenience and free days for speaking, yet there had been no need of a long oration. I have spoken.

