% Divinatio in Caecilium (divinatio-in-caecilium) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} If anyone of you, gentlemen of the jury, or of those in attendance, by chance is wondering that I, who for so many years have been engaged in cases and public trials in such a way as to have defended many and harmed no one, am now suddenly with changed purpose descending to prosecute --- if such a man learns the cause and the reasoning of my decision, he will both approve what I am doing and judge that, in this case, no one ought to be set above me as the prosecutor.

\ciceroSection{2} When I had served as quaestor in Sicily, gentlemen, and had taken my leave from the province in such a way as to leave with all Sicilians a pleasant and lasting memory of my quaestorship and my name, it came about that, while they thought their fortunes were chiefly defended by their many old patrons, they thought they had now also some safeguard for them in me. So now, plundered and harried, they have come in numbers in delegations to me, asking me to undertake the defence of all their fortunes. They said I had often promised, often shown, that, if any moment came in which they might require something of me, I would not fail their interests.

\ciceroSection{3} They said the time had come now, not for me to defend their interests, but the life and safety of the entire province; that they no longer had even gods in their cities to whom they might flee, since their most sacred images had been taken by Gaius Verres from their most reverenced shrines. Whatever luxury could effect by way of disgrace, cruelty by way of punishment, greed by way of plunder, arrogance by way of insult, all this they had endured under this one praetor for three years. They asked and begged that I should not despise their suppliant petitions --- they who, while I am still alive, ought to be no man's suppliants.

\ciceroSection{4} I took it heavily and bitterly, gentlemen, that I should be brought to such a position that either men who had sought my help and protection should find their hope deceived, or that I, who had given myself to defending men from the very beginning of my youth, should, by the moment and by duty, be transferred to prosecuting. I said that they had as prosecutor Quintus Caecilius, who had especially been quaestor in the same province after my own quaestorship. The very thing by which I hoped this trouble might be removed from me was for me the chief obstacle. For they would have given way to me much more easily if they had not known him, or if he had not been quaestor among them.

\ciceroSection{5} I was led, gentlemen, by duty, by good faith, by pity, by the example of many honourable men, by the old custom and institution of our ancestors, to think that I ought to take this burden of labour and duty upon myself not at my own moment but at that of my close associates. In this business, however, this consoles me, gentlemen: that what is seen as my prosecution is to be considered not so much a prosecution as a defence. For I am defending many living men, many cities, the whole province of Sicily; and so, since I have only one man to prosecute, I seem to remain almost within my own habit and not to depart altogether from defending and supporting men.

\ciceroSection{6} If I did not have a case so fitting, so distinguished, so weighty --- if either the Sicilians had not asked this of me, or if I had with the Sicilians no bond of so great a connection, and if what I am doing I were merely professing to do for the sake of the commonwealth, that a man endowed with singular cupidity, audacity, and crime --- whose thefts and disgraces we have known to be of the greatest and most shameful kind, not only in Sicily but in Achaia, Asia, Cilicia, Pamphylia, and finally at Rome before the eyes of all --- should at my pleading be summoned into court: who, in the end, would there be who could find fault with my deed or my counsel?

\ciceroSection{7} What is there, by the faith of gods and men, in which I could now do greater service to the commonwealth? What is there which ought to be more pleasing to the Roman people, or more wished for by the allies and foreign nations, or more fitted to the safety and fortunes of all? Plundered, harried, utterly overturned provinces, allies and tributary peoples of the Roman people, beaten down, miserable, are seeking now no longer the hope of their safety but consolation in their ruin.

\ciceroSection{8} Those who wish trials to remain in the senatorial order complain that they do not have suitable prosecutors. Those who can prosecute desire severity in trials. The Roman people meanwhile, although it is afflicted with many inconveniences and difficulties, yet desires nothing in the commonwealth so much as that old force and gravity of the courts. From a desire for proper trials the tribunician power has been demanded back; from the levity of the courts a different order is being asked to do justice; through the fault and disgrace of the jurors even the censorial name --- which used to seem too harsh to the people --- is now demanded; that name has now been made popular and welcome.

\ciceroSection{9} In this licence of the most guilty men, in the daily complaint of the Roman people, in the infamy of the courts, in the offence given to the whole order, since I judged the one remedy for these so many ills to be that suitable and upright men should take up the cause of the commonwealth and the laws, I confess that, for the sake of the safety of all, I have approached that part of the work of relieving the commonwealth which most needs help.

\ciceroSection{10} Now since I have shown by what considerations I came to this case, I must necessarily speak about our contest, that you may have something to follow in choosing the prosecutor. I think the matter is this, gentlemen: when an indictment of any man is laid for extortion, if there is a contest among any parties as to whom delation should chiefly be given, these two things ought to be looked at first --- whom those to whom injuries are alleged to have been done most wish to have as prosecutor, and whom he who is charged with having done those injuries least wishes.

\ciceroSection{11} In this case, gentlemen, although I think both points are clear, I will speak about both, and first about that which ought to count most with you --- the wish of those to whom the injuries have been done, for whose sake the law on extortion was established. It is alleged that Gaius Verres for three years devastated the province of Sicily, sacked the cities of the Sicilians, emptied their houses, plundered their shrines. The Sicilians as a whole are present, and complain. To my good faith, which they have already tested and known, they flee for refuge. They seek through me from you and from the laws of the Roman people the help they need. They have wished me to be the defender of their disasters, the avenger of their injuries, the cognitor of their right, the prosecutor of the whole case.

\ciceroSection{12} Will you, Quintus Caecilius, say either that I am not approaching the case at the request of the Sicilians, or that the wish of the best and most faithful allies ought not to count with these gentlemen? If you dare to say what your enemy --- as you pretend Gaius Verres is --- most wishes to be thought, that the Sicilians have not asked this of me, you will first relieve the case of the man you call your enemy: about whom there is now thought to be not a prejudgement but plainly a judgement, since it is so widely known that all the Sicilians have sought a prosecutor for their case against his injuries.

\ciceroSection{13} If you, his enemy, will deny that this was done, when the man himself, whom this matter most damages, does not dare to deny it --- see that you do not seem to practise enmity too familiarly. Then there are witnesses, the most distinguished men of our state, all of whom I do not need to name. I will appeal to those who are present, whom, if I were lying, I would least wish to be witnesses of my impudence. He who sits on the council, Gaius Marcellus, knows; he whom I see present, Gnaeus Lentulus Marcellinus, knows --- on whose good faith and protection the Sicilians especially lean, since the whole province is attached to the name of the Marcelli.

\ciceroSection{14} These men know that this was not only asked of me, but asked so often and so vehemently that either I had to undertake the case or repudiate the duty of close association. But why do I use these witnesses, as though the matter were doubtful or obscure? There are men present from the whole province, the most distinguished, who beg and beseech you, gentlemen, in choosing the prosecutor of their case, that your judgement should not differ from theirs. Delegations from all the cities of Sicily are present, except for two cities; if those two were present, two great charges would be lessened --- charges which Gaius Verres shared with these two cities.

\ciceroSection{15} "But why," he says, "have they sought this protection chiefly from me?" If it were doubtful whether they had sought it or not, I would say why they sought it. Now, indeed, since it is so evident that you can judge of it with your eyes, I do not see why this should be a disadvantage to me, if it is thrown at me as a reproach that I above others have been chosen.

\ciceroSection{16} But I do not, gentlemen, claim this for myself; I do not put this in my speech, nor do I leave it even in anyone's opinion --- that I have been preferred to all advocates. It is not so. The reasoning has been drawn from each one's circumstances, health, and capacity to act. My wish and view in this matter has always been: that I would prefer anyone of those who are suitable to undertake it rather than myself, but myself rather than no one.

\ciceroSection{17} It remains to ask, since this is established, that the Sicilians have asked me, what weight this matter ought to have with you and with your minds, what authority before you allies of the Roman people, your suppliants, ought to have in seeking back their right. About which, what need is there for me to say more? --- as if it were doubtful that the whole law on extortion was established for the sake of the allies.

\ciceroSection{18} For when money is taken from citizens, it is generally recovered by civil action and private right. This is a law for the allies; this is the right of foreign nations; this is the citadel which they have, somewhat less fortified now than before, but if there is any hope left which can console the souls of the allies, it is wholly placed in this law --- a law for which not only the Roman people but the most distant nations have long been demanding strict guardians.

\ciceroSection{19} Who then can say it is not right to act under the law at the discretion of those for whose sake the law was made? If the whole of Sicily were to speak with one voice, this is what it would say: "Whatever of gold, whatever of silver, whatever of ornaments was in my cities, my dwellings, my shrines; whatever in each thing I held by right by the kindness of the senate and the Roman people, you, Gaius Verres, have torn from me and carried off; under which heading I claim from you a hundred million sesterces under the law." If, as I said, the whole province could speak, it would use these words; since it could not, it has itself chosen as agent of these matters whomever it judged suitable.

\ciceroSection{20} In a matter of this kind will any man be found so impudent as to dare to approach or aspire to another's case against the will of those whose business it is? If, Quintus Caecilius, the Sicilians said this to you: "We do not know you, we do not know who you are, we have never seen you before; let us defend our fortunes through one whose good faith we have known," would they not be saying what they ought to be able to prove to anyone? Now they say this: that they know both of us; that one of us they wish to be the defender of their fortunes, the other they plainly do not.

\ciceroSection{21} They say enough why they do not want him, even if they keep silent; but they do not keep silent. Will you nevertheless thrust yourself upon them most unwilling? Will you nevertheless speak in another's case? Will you nevertheless defend those who would rather be deserted by all than defended by you? Will you nevertheless promise your service to those who think you neither wish to act for them, nor, if you wished, would be able to? Why do you try by force to wrench from them the slender hope of their remaining fortunes, which they have placed in the severity of the law and the trial? Why do you intrude on those most unwilling for whose sake the law most wishes to provide? Why are you now trying to overturn from all their fortunes those of whom in the province you did not deserve very well? Why do you take from them the power not only of pursuing their right, but even of mourning their calamity?

\ciceroSection{22} For with you as prosecutor, which of them do you suppose will be present, when you understand that they are striving not to take vengeance upon another through you, but to take vengeance upon you yourself through someone else? But that is one thing only --- that the Sicilians most wish me to act. The other (I take it) is obscure: by whom Verres least wishes to be prosecuted. Has anyone ever struggled so openly for honour, so vehemently for his life, as he and his friends have struggled to keep this delation from being given me? There are many things, Caecilius, which Verres thinks are in me, which he knows are not in you. What kind of things these are in each of us, I will mention shortly.

\ciceroSection{23} Now I will only say this --- to which you in silence will assent --- that there is nothing in me which he despises, nothing in you which he fears. So that great defender and friend of his supports you, attacks me; openly he asks the jurors that you be set above me, and says he asks this honourably, without odium, without offence. "I do not ask," he says, "what I usually obtain when I have made a vehement effort: I do not ask that the defendant be acquitted, but that he be prosecuted by this man rather than by that. Grant me this; concede what is easy, what is honourable, what is without odium; when you have granted it, without any danger, without any infamy to yourselves, you will have granted that the man for whom I work be acquitted."

\ciceroSection{24} And the same man says --- so that some fear may be added to favour --- that there are certain men on the council to whom he wishes the tablets shown; that this is most easy, for the votes are not cast singly but established collectively; that to each man a wax tablet is given, with lawful wax, not that infamous and unspeakable wax. And he toils not so much on Verres's account as because the whole matter pleases him not at all; for he sees that, if from the noble youths whom he has so far mocked, from the four-time informers whom not without reason he has always despised and reckoned as nothing, the will to prosecute is transferred to brave and tested men, he will no longer be able to hold sway in the courts.

\ciceroSection{25} I now warn this man in advance: if you wish me to plead this case, his whole way of conducting the defence will have to be changed --- and changed so much that, by a better and more honourable course than he himself wishes to take, he should imitate those men whom he himself has seen most distinguished, Lucius Crassus and Marcus Antonius, who thought that nothing should be brought into trials and friends' cases beyond good faith and natural ability. With me as prosecutor he will have nothing to think can be corrupted in court without great danger to many.

\ciceroSection{26} I judge that in this trial the case of the Sicilians has been received by me, the case of the Roman people undertaken --- so that I must crush not one wicked man only (which the Sicilians sought), but altogether all wickedness must be extinguished and destroyed (which the Roman people have long been demanding). What in this I can strive for or accomplish, I prefer to leave to the hopes of others rather than place in my own speech.

\ciceroSection{27} You, however, Caecilius --- what can you do? At what time, in what matter, have you not only given any sample to others, but tested yourself? Has it not occurred to you what business it is to sustain a public case, to unfold the whole life of another, and to expose it not only to the minds of the jurors, but even to the eyes and the gaze of all; to defend the safety of allies, the interests of provinces, the force of laws, the gravity of courts? Learn from me, since this is the first time you have had the chance to learn, how many things must be in him who prosecutes another. If you find any one of these in yourself, I will then, of my own accord, yield to you what you ask. First, integrity and singular innocence; for nothing is less to be tolerated than for one to demand an account of another's life who cannot give an account of his own.

\ciceroSection{28} On this point I will not say more about you. I think everyone notes this one thing: that up to now you have been able to be known by no one but the Sicilians; the Sicilians say that, though they are angry at the same man with whom you say you are at enmity, they will nevertheless not be present at the trial with you as prosecutor. Why they refuse, you will not hear from me. Let these gentlemen suspect what they must. Indeed, as the kind of men they are --- shrewd and over-suspicious --- they think you wish not to bring out from Sicily letters against Verres, but, since both his praetorship and your quaestorship are sealed up by the same letters, they suspect you wish to carry the letters out of Sicily.

\ciceroSection{29} Then, a prosecutor must be firm and truthful. If I supposed you wished to be such, I easily understand you cannot be. Nor do I say what, if I said it, you could not weaken: that before you left Sicily you returned to terms of friendship with Verres; that Potamon, your scribe and intimate, was kept by Verres in the province while you were leaving; that Marcus Caecilius, your brother, a most select and most distinguished young man, is so far from being present and joining with you to pursue your wrongs, that he is with Verres and is living with him on the most familiar and friendly terms. There are these and many other signs in you of a false prosecutor, which I do not now use; I say only this: that, even if you most wished it, you could not be a true prosecutor.

\ciceroSection{30} For I see there are very many charges in which your partnership with Verres is such that you would not dare to touch them in prosecuting. All Sicily complains that Gaius Verres demanded grain from the farmers for his bin, and, when a peck of wheat was costing two sesterces, he exacted twelve sesterces per peck in lieu of the grain. A great charge, vast money, shameless theft, intolerable injury! By this one charge alone I should necessarily condemn him.

\ciceroSection{31} What will you do, Caecilius? Either pass over so great a charge, or bring it forward? If you bring it forward, will you give as a charge against another what you yourself did at the same time in the same province? Will you dare to prosecute another in such a way that you cannot help being condemned yourself? But if you pass it over, what kind of prosecution of yours will it be that, through fear of personal danger, dreads not only making the wager of the gravest and most certain charge, but even the very mention of it?

\ciceroSection{32} Grain was bought from the Sicilians by senatorial decree under the praetor Verres, and the entire purchase money for that grain was not paid. This is a serious charge against Verres --- serious if I prosecute, none if you do. For you were quaestor; you were handling the public money, from which, even if the praetor had wished, the power of preventing any deduction was largely yours. So no mention of this charge will be made with you as prosecutor. The greatest and best-known thefts and injuries of his will be silenced through the whole trial. Believe me, Caecilius: he cannot in prosecuting truthfully defend the allies who is bound to the defendant by partnership in the charges.

\ciceroSection{33} Contractors exacted money from the cities in lieu of the grain. What of it? Was this done only when Verres was praetor? No --- but even when Caecilius was quaestor. What then? Will you give this as a charge, when you both could have prevented it from being done and ought to have done so, or will you leave it untouched altogether? Therefore Verres in his own trial will not even hear about that which, when he was doing it, he could not see how he was going to defend. And these I name which are in plain view; there are other more hidden thefts which he, in order (I take it) to slow down your spirit and impulses, most generously shared with his quaestor.

\ciceroSection{34} You know that I have been informed of these things. If I should wish to bring them out, all would easily understand that there has not only been a common will between you, but that not even the booty has yet been divided. So if you ask that information be given against you on what you and he together did, I grant it, if the law permits. But if we are speaking of prosecution, you must yield to those who are not hindered by any offence of their own from being able to point out the offences of another.

\ciceroSection{35} See how great will be the difference between my prosecution and yours. I will charge against Verres even what you committed without him, on the ground that he did not prevent you, when he himself had the highest power. You, on the other hand, will not bring up even what he did, lest you be found in any way connected with him. What? Caecilius, do those things seem to you to be despicable, without which the case --- especially so great a case --- cannot in any way be sustained? Some capacity for action, some habit of speaking, some method or practice in the Forum, in the courts, in the laws?

\ciceroSection{36} I understand on what rocky and difficult ground I am moving. For all arrogance is hateful, but the arrogance about natural ability and eloquence is by far the most odious. So I say nothing about my own ability; nor is there anything I could say, nor would I say it if there were. For either what is said about me, whatever it is, suffices for me, or, if it is too little, I cannot increase it by mentioning it.

\ciceroSection{37} About you, Caecilius --- I will speak with you familiarly now, by Hercules, beyond this contention and rivalry of ours --- consider again and again how you yourself judge yourself, and gather yourself together, and consider both who you are and what you can do. Do you think you can on the greatest and bitterest matters --- when you have undertaken the case of the allies, and the fortunes of the province, the right of the Roman people, the gravity of the trial and the laws --- sustain so many serious and varied subjects with your voice, your memory, your counsel, your ability?

\ciceroSection{38} Do you think that you can distinguish in charges and in speech, just as they were divided by place and time, what Gaius Verres did wrong as quaestor, what as legate, what as praetor, what at Rome, what in Italy, what in Achaia, in Asia, in Pamphylia? Do you think you can do what is most needful in such a defendant: bring it about that what he did wantonly, what wickedly, what cruelly, shall seem as bitter and unworthy to those who hear of it as it seemed to those who suffered it?

\ciceroSection{39} The things I speak of are great, believe me; do not despise them. Everything must be said, shown, unfolded. Not only must the case be set out, but also gravely and copiously argued. You must achieve, if you mean to do or accomplish anything, that men shall not only hear you but listen gladly and eagerly. Even if nature greatly aided you in this, even if you had pursued and wrought at the best disciplines and arts from boyhood, even if you had learned Greek letters at Athens and not at Lilybaeum, Latin at Rome and not in Sicily --- it would still be a great matter both to compass so great a case, so awaited, by your diligence, to embrace it in your memory, to bring it forth in speech, and to sustain it by voice and force.

\ciceroSection{40} Perhaps you will say: "What then? Are all these things in you?" Would they were! And yet, that they might be, has been my labour with the greatest zeal from boyhood. If, on account of the greatness and difficulty of these things, I have not been able to attain them --- I, who have done nothing else my whole life --- how far do you suppose you stand from these matters, things you not only never thought of before, but cannot suspect even now, when you are entering on them, what they are or how great?

\ciceroSection{41} I myself --- who, as everyone knows, am so engaged in the Forum and in the courts that few or none of my own age have defended more cases, and who spend in these studies and labours all the time given me from my friends' affairs, that I might be more prepared and more ready for forensic use --- so may the gods be propitious to me as, when I think of the moment when, on the day appointed for the case, I must speak, not only am I troubled in mind, but I tremble in my whole body.

\ciceroSection{42} I already foresee in mind and thought what zeal of men, what concourse, will then be there; what expectation the magnitude of the trial will bring; how great a multitude of hearers the infamy of Gaius Verres will rouse; finally, how attentive an audience the wickedness of that man will give to my speech. When I think of this, I am already afraid that I shall not be able to say anything worthy of the offence given to men who are his enemies and bitterly hostile to him, of the expectation of all, and of the magnitude of the matters.

\ciceroSection{43} You fear none of this, think of none of this, labour over none of it. If from some old speech you can learn by heart "I, Jupiter, best and greatest..." or "I should wish, gentlemen, if it could have been..." or something of the sort, you think you will come into court splendidly prepared.

\ciceroSection{44} Even if there were nobody to answer you, still you would not, in my opinion, be able to make plain even the bare facts of the case. As it is, you do not even consider that you will have a contest with a man most eloquent and most prepared for speaking, with whom you must now reason, now fight and contend with every method. His ability I praise in such a way that I am not afraid; I approve it in such a way that I think I can be more easily delighted by him than deceived. He will never overcome me by stratagem, never pervert me by any artifice, never try to weaken and undo me by his ability. I know all his approaches and methods of speaking; we have often been on the same side, often on opposite sides; he will so speak against me, ingenious as he is, that he will think he is undergoing some judgement also of his own ingenuity.

\ciceroSection{45} But you, Caecilius --- how he will play with you, how he will toss you about by every method --- I already seem to see. How often will he give you the option, the choice, of which way you wish to take it: whether the deed was done or not, whether what is alleged is true or false; whichever you say will tell against you. What heat, what bewilderment, what darkness, by the immortal gods, will be in you, the most kindly of men! What when he begins to divide the limbs of your prosecution and lay out the parts of the case on his fingers? What when he settles each point in turn, draws it out, finishes it off? You yourself will surely begin to fear lest you have endangered an innocent man.

\ciceroSection{46} What when he begins to lament, to grieve, to unload some part of the odium from him onto you; to mention the bond established between quaestor and praetor, the custom of our ancestors, the religion of the lot --- will you be able to bear the odium of his speech? Look closely; consider again and again. For it seems to me there will be a danger that he will not only overwhelm you with words, but, by his very gesture and bodily motion, dazzle the edge of your wits and lead you away from your settled purposes and thoughts.

\ciceroSection{47} The judgement on this very point I see is to be made at once. For if you can answer me today on what I am saying, if you depart by even a single word from that book which some schoolmaster put together for you out of others' speeches, I shall judge that you can not fail in that trial either, and can satisfy your case and duty. But if in this prelude with me you have nothing to say, who do we suppose you will be in the actual battle with the keenest adversary? Granted, he himself is nothing, can do nothing; but he comes prepared, with practised and able assistant prosecutors. This is something --- though it is not enough; for in everything he who is the leader in conducting the case ought to be most adorned and most prepared. But I see Lucius Appuleius is the next subscript, a man not in age but in forensic use and practice still a recruit.

\ciceroSection{48} Then, I take it, he has Alienus, but a man from the benches, whose ability in speaking I have never sufficiently noted, but in shouting I see him to be well-vigorous and well-practised. In this man are all your hopes; he, if you are constituted prosecutor, will sustain the whole trial. Even he will not strain in speaking as much as he can, but will consult your praise and reputation, and will from what he himself can do in speaking somewhat hold back, that you may seem to be something. As we see done with Greek actors, often the man who plays the second or third parts, when he could speak somewhat more brilliantly than the lead actor himself, abates much, that the principal may stand out as conspicuously as possible --- so will Alienus do; he will be your servant, your flatterer; he will strain somewhat less than he can.

\ciceroSection{49} Now consider what kind of prosecutors we shall have in so great a trial, when even Alienus from his ability, if he has any, will pull back something, and Caecilius will think he is finally something only if Alienus is less vehement and yields him the leading role in speaking. What fourth man you will have I do not see, unless perhaps from that herd of stallers who demand a subscription for themselves to whomever you might give the delation:

\ciceroSection{50} from these utter strangers you come so prepared that you must take some stranger as guest. To whom I will not pay so much honour as to answer at any specific point or singly to each what they shall have said. So briefly, since I have hit upon mention of them not by design but by chance, I shall, as if in passing, satisfy them all together. Do I seem to you to be so destitute of friends that, instead of being given a subscriptor from those whom I have brought with me, one is to be added from the populace? But to you is there such a dearth of defendants that you should attempt to snatch the case from me rather than find some defendant of your own order at the column of Maenius? "Place me as guard," he says, "over Tullius."

\ciceroSection{51} What? How many guards I shall need, if I once admit you to my boxes! For not only must you be guarded against blabbing, but also against carrying anything off. But about that "guard" of yours I will most briefly answer you all: that men of this kind would not commit, that to a case so great undertaken by me, entrusted to me, anyone should as subscriptor breathe upon it against my will. For my good faith refuses a guard; my diligence shrinks from a spy. But to come back to you, Caecilius: see how many things you lack; how many things are in you which a guilty defendant would wish to be in his prosecutor, you now thoroughly understand.

\ciceroSection{52} What can be said to this? For I do not ask what you will say. I see that not you but this book in your prompter's hand will reply to me. If your prompter wishes to advise you rightly, he will counsel you to depart from here and not answer me a word. For what will you say? Will it be what you are accustomed to say --- that Verres did you an injury? I think so. For it would not be likely that, when he was doing wrong to all the Sicilians, you alone should be the exception for whom he made an effort.

\ciceroSection{53} But the rest of the Sicilians have found an avenger of their wrongs. You, while you try by yourself to pursue your wrongs (which you cannot), are doing this --- making the wrongs of the rest go unpunished and unavenged. And it does not occur to you that not only is it customarily looked at, who ought to take vengeance, but also who is able to. He in whom both qualities are, is to be preferred. Where one of the two is, there it is customary to ask, not what he wishes, but what he can do.

\ciceroSection{54} If, then, you suppose that the power of prosecuting ought to be granted to him to whom Gaius Verres has done the greatest wrong --- which do you think these jurors ought to feel more grievously: that you have been wronged by him, or that the province of Sicily has been harried and ruined? You will, I suppose, concede that this is far more grievous, and that all ought to consider it more grievous. Concede therefore that, in the matter of prosecuting, the province should be set above you. For the province does prosecute, when he conducts the case whom she has adopted as the defender of her right, the avenger of her wrongs, the actor of the whole case.

\ciceroSection{55} "But Verres has done you a wrong of such a kind as could move even the souls of others by another's misfortune." Not at all. For I think this also pertains to the matter --- what kind of wrong is alleged, what cause of enmities is brought forward. Hear it from me, for he himself --- unless he is utterly without sense --- will surely never bring it forward. There is a certain Agonis of Lilybaeum, freedwoman of Erucine Venus; this woman before this quaestor [Caecilius] was plainly well-supplied and rich. From her one of Antonius's prefects was, by violence, leading away musical slaves, whom he said he wished to use in the fleet. Then she --- as is the way in Sicily of all who are slaves of Venus and of those who have been freed by Venus, that they may put a religious scruple before this prefect by Venus's name --- said that both she and her property were Venus's.

\ciceroSection{56} When this was reported to the quaestor Caecilius --- a most excellent man, a most fair person --- he orders Agonis to be summoned to him; he gives a trial at once: "If it appears she said that she and hers were Venus's." The recuperatores judge what was inevitable; for there was no doubt to anyone that she had said it. He then enters into possession of the woman's goods; he adjudges the woman herself into servitude to Venus; then he sells the goods, realizes the money. So while she wished to keep a few slaves under Venus's name and the religious scruple, by Caecilius's wrong she lost all her fortunes and her freedom. Verres comes to Lilybaeum afterwards. He learns of the matter, disapproves of the deed, compels his quaestor to count out and pay back to the woman all the money he had realised from Agonis's goods.

\ciceroSection{57} So far this is, as I see all of you wondering at, not Verres but Quintus Mucius. For what could he have done more elegantly for the regard of men, more equitably for relieving the woman's calamity, more vehemently for restraining the quaestor's lust? All this seems to me supremely to be praised. But suddenly, on the spot, from a man as if by some Circaean cup, he became Verres again; he returned to himself and to his ways. For from that money he turned a great part to himself; to the woman he gave back as much as he saw fit.

\ciceroSection{58} On this point, if you say you have been wounded by Verres, I will allow it and grant it. If you complain that a wrong has been done you, I will defend him and deny it. Finally, on the wrong done to you, none of us ought to be a graver judge than yourself, to whom it is alleged to have been done. If you afterwards came back to friendly terms with him, if you were several times in his house, if he afterwards dined with you --- which would you rather be thought, treacherous or a feigned prosecutor? I see one or the other is necessary; but I will not fight with you about which you wish to choose.

\ciceroSection{59} If even the cause of the wrong, alleged to have been done you by him, does not remain, what have you to say why you should be set above not me only, but anyone? Unless perhaps that thing which I hear you are going to say --- that you were his quaestor. Which would be a serious cause if you were contending with me which of us ought to be more friendly to him. In a contest of undertaking enmities, it is ridiculous to think the cause of bond ought to be reckoned just as a reason for inflicting danger.

\ciceroSection{60} For if you had received the most numerous wrongs from your praetor, you would yet earn more praise by enduring them than by avenging them. But when nothing in his life is more upright than this very thing which you call a wrong --- will these men decide that this case, which they would not approve in another, in you seems just for violating a bond? If you did receive the greatest wrong from him, yet, since you were his quaestor, you cannot accuse him without some censure. But if no wrong has been done you, you cannot accuse him without crime. So when it is uncertain about the wrong, do you think any of these men there is who would not rather you depart without censure than with crime?

\ciceroSection{61} See now what difference there is between my view and yours. You, when you are inferior in everything, think that in this one thing you ought to be set above me --- that you were his quaestor. I, even if you were superior in everything, would think you ought for this very reason alone to be repudiated as prosecutor. For so we have received from our ancestors, that a praetor ought to be in place of a parent to his quaestor. No more just nor weightier cause of close association can be found than the partnership of lot, of province, of duty, of public commission.

\ciceroSection{62} For this reason, if you could lawfully prosecute him, yet, since he had been to you in the place of a parent, you could not piously do so. But when you neither received any wrong and create danger to your praetor, you must necessarily confess that you are trying to wage upon him an unjust and impious war. For your quaestorship is good for this --- that you must labour, in giving an account, why you, who were his quaestor, are prosecuting him; not that for that very reason it should be demanded that the prosecution be given especially to you. And it has scarcely ever come to a contest about prosecuting where one who had been quaestor was not rejected.

\ciceroSection{63} Thus the power of laying an indictment was not granted to Lucius Philo against Gaius Servilius, nor to Marcus Aurelius Scaurus against Lucius Flaccus, nor to Gnaeus Pompey against Titus Albucius. None of them was rejected for any unworthiness, but lest the wish of violating the bond should be approved by the authority of the jurors. And that Gnaeus Pompey contended with Gaius Iulius as you do with me; for he had been Albucius's quaestor, as you Verres's. Iulius brought to the prosecution this authority: that, just as we are now asked by the Sicilians, so he then had been asked by the Sardinians and had taken up the case. Always this cause has had most weight; always this method of prosecuting has been most honourable: to take up enmities for the allies, for the safety of the province, for the interests of foreign nations; to come into danger; to bring effort, zeal, labour to bear.

\ciceroSection{64} For if the case of those is approvable who wish to pursue their own wrongs --- where they serve their own pain, not the interests of the commonwealth --- how much more honourable is that case which not only ought to seem approvable but to be welcome: when, with no wrong privately received, one is moved by the pain and wrongs of allies and friends of the Roman people. Recently, when that bravest and most innocent man Lucius Piso demanded the delation of an indictment against Publius Gabinius, and on the other side Quintus Caecilius was suing for it (saying that he was pursuing old enmities long undertaken), although Piso's authority and dignity counted most, yet that was the most just cause: that the Achaeans had adopted him as their patron.

\ciceroSection{65} For when the law on extortion is itself the patron of the allies and friends of the Roman people, it is unfair not to think him most fitting as the agent of the law and the trial whom the allies have chosen above all to be the agent of their case and the defender of their fortunes. Or what is more honourable to mention, ought it not to be much more equitable to approve? Which is the more splendid, the more illustrious thing to say: "I have prosecuted him whose quaestor I was, with whom the lot and the custom of our ancestors, with whom the judgement of gods and men joined me," or "I have prosecuted at the request of the allies and friends; I have been chosen by the whole province to defend its rights and fortunes"? Can anyone doubt that it is more honourable to act for the sake of those among whom you were quaestor than to prosecute the man whose quaestor you were?

\ciceroSection{66} The most distinguished men of our state, in the best times, considered this their fullest and finest service: to drive away wrongs from their guest-friends and clients, from foreign nations under the protection and dominion of the Roman people, and to defend their fortunes. We hear that wise Marcus Cato, that most distinguished and prudent man, took up serious enmities with many men on account of the wrongs of the Spaniards, among whom he had been consul.

\ciceroSection{67} Recently we know that Gnaeus Domitius cited Marcus Silanus on account of the wrongs of one man, Aegritomarus, his father's friend and guest. For nothing has ever moved the souls of guilty men more than this custom of our ancestors, recovered and renewed after a long interval --- complaints of the allies brought to a man not idle, taken up by one who seemed able by his good faith and diligence to defend their fortunes.

\ciceroSection{68} This is what men fear, this is what they labour over, this is what they take ill at being instituted, or rather at being recalled and renewed. They think that, if this custom should creep on a little and come forth, the laws and trials will be administered through the most honourable men and the bravest, not through unskilled youths or four-time informers of that kind.

\ciceroSection{69} Of which custom and institution our fathers and ancestors did not repent, when Publius Lentulus, who was the chief of the senate, prosecuted Manius Aquillius, with Gaius Rutilius Rufus as subscriptor; or when Publius Africanus --- a man most distinguished in virtue, fortune, glory, achievements --- after he had twice been consul and censor, summoned Lucius Cotta to trial. Then by right the name of the Roman people flourished; by right the authority of this empire and the majesty of the state was reckoned weighty. No one wondered at Africanus then for what now in me --- a man endowed with small means and resources --- they pretend to wonder at, when they take it ill:

\ciceroSection{70} "What does this man want? Does he wish to be thought a prosecutor, who used before to defend, especially now, at this age, when he is seeking the aedileship?" I, however, judge it to belong not only to my age but to a much more advanced one, and to the most ample of honours, both to prosecute the wicked and to defend the wretched and the calamitous. Surely this is the remedy for our sick and almost desperate commonwealth and our courts corrupted and contaminated by the vice and shame of a few: that men as honourable, as upright, as diligent as possible should approach the defence of the laws and the authority of the courts. Or, if not even this can help, surely no medicine for so many troubles will ever be found.

\ciceroSection{71} No greater safety for the commonwealth is there than that those who prosecute another should fear no less for their own praise, honour, and reputation than those prosecuted fear for their lives and fortunes. So they have always prosecuted most diligently and most laboriously who thought their own reputation was at stake. Wherefore you ought to lay this down, gentlemen: that Quintus Caecilius --- of whom there has been no opinion at any time, and of whom there will be no expectation in this very trial; who labours neither to preserve a fame previously gathered, nor to confirm hope for a future time --- will not conduct this case very strictly, not very accurately, not very diligently. For he has nothing to lose by failure. To depart most disgracefully and most flagitiously, he will miss none of his old ornaments.

\ciceroSection{72} The Roman people holds many hostages from us, which, that we may be able to keep, defend, confirm, and recover safe, we shall have to fight in every way. It holds the office I am seeking; it holds the hope which I have set before me; it holds the reputation I have gathered with much sweat, labour, and waking nights. So that, if in this case I have proved my duty and diligence, I may be able to retain through the Roman people what I have said, safe and sound. If even a tiny stumble or hesitation occurs, all those things which were collected one by one and over a long time we may lose at one stroke.

\ciceroSection{73} Wherefore, gentlemen, it is for you to choose whom you judge can most easily sustain the magnitude of the case and the trial by good faith, diligence, counsel, authority. If you set Quintus Caecilius above me, I will not think myself defeated in dignity. Take care that the Roman people do not judge that so honourable, so strict, so diligent a prosecution has not pleased you, and does not please your order.

