% Pro Q. Roscio Comoedo (pro-roscio-comoedo) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.
% The opening of the speech is lost; section 1 begins mid-sentence.

\ciceroSection{1} ...natural malice would have to be assumed. He, naturally, --- excellent man, endowed with singular good faith --- in his own case is trying to use his own account-books as witnesses. Those who have entered a sum charged against an honourable man on their books generally say: "Could I have corrupted such a man, that he should make a false entry in the day-book on my account?" I am waiting for the moment when Chaerea will use this very line: "Could I have driven this hand of mine, full of treachery, and these fingers of mine, to write up a false entry?" If he produces his own books, Roscius will produce his too. The entry will be in his books; in this man's it will not be.

\ciceroSection{2} Why is the one to be believed rather than the other? --- "He would have written it, if he had not entered the sum at this man's bidding?" --- "Would my client not have written what he had given orders to be entered as charged against him? For just as it is shameful to write down what is not owed, so it is dishonest not to enter what you do owe. The book of the man who did not enter what is true is condemned just as much as the book of the man who entered what is false." But, relying on the abundance and capacity of my case, see how far I will go: if Gaius Fannius produces his books of receipts and expenses, written for his own purpose at his own discretion, I will not refuse to have you decide in his favour.

\ciceroSection{3} What brother grants this to brother, what father to son: that whatever he had entered should be held ratified? Roscius will hold it ratified. Produce them. Whatever has persuaded you will persuade him; whatever has been proven to you will be proven to him. A little while ago we were demanding the books of Marcus Perpenna, of Publius Saturius. Now, Gaius Fannius Chaerea, we are pressing for yours alone, and we are not refusing that the case be decided in accordance with them. Why then do you not produce them?

\ciceroSection{4} Does he not keep books? On the contrary, very meticulously. Does he not enter small sums in his day-book? On the contrary, every sum. Is this entry slight and trifling? It is fifty thousand sesterces. How is it that you have so great a sum of money lying out of your account? How is it that fifty thousand sesterces are not in the book of receipts and expenses? By the immortal gods! Is anyone endowed with such audacity that, what he is afraid to enter into his books as a debt, he dares to claim; what he was unwilling to enter into the day-book unsworn, he does not hesitate to swear to in litigation; what he cannot prove to himself, he tries to persuade another of?

\ciceroSection{5} He says I am too quick to be indignant about the books. He admits that this entry is not in the book of receipts and expenses, but contends that it stands in the rough waste-book. Is your love of yourself, your magnificent self-regard, so great that you claim money not from your books but from your waste-book? To recite your own day-book in the place of evidence is arrogance; to produce a waste-book of your own scribbles and erasures --- is that not insanity?

\ciceroSection{6} If the waste-book has the same force, care, and authority as the proper book, what is the point of keeping a day-book, of writing it up, of preserving the order, of handing the antiquity of the records to memory? But if, because we trust the waste-book in nothing, we have therefore established the writing of a day-book --- shall what is even universally regarded as light and weak be considered weighty and inviolate before a judge? What is the reason that we write the waste-book carelessly?

\ciceroSection{7} What is the reason we make up the day-book diligently? For what reason? Because the waste-book is monthly, the books are eternal; the waste-book is rubbed out at once, the books are kept inviolate; the waste-book embraces the memory of a small time, the books embrace the trust and conscience of a perpetual estimation; the waste-book is scattered, the books are made up in order. So no one ever produced his waste-book in court; he produced his day-book; he read out his books. You, Gaius Piso, adorned with such good faith, virtue, gravity, authority, would not dare to claim money out of a waste-book.

\ciceroSection{8} I ought not to dwell longer on what is plain by custom. But what very much concerns the matter, I ask: how long ago, Fannius, did you enter this sum into your waste-book? He blushes; he does not know what to answer; he has no fiction ready on the spur of the moment. "It is two months now," you will say. Even so, it should have been transferred to the book of receipts and expenses. "More than six months." Why has this entry been lying so long in the waste-book? What if it is now more than three years? How is it that, when all who keep books transfer to them their accounts almost monthly, you allow this entry to lie more than three years in the waste-book?

\ciceroSection{9} Have you the rest of your entries digested into a book of receipts and expenses or not? If not, how do you keep books at all? If you have, why, when you were entering all the others in order, did you leave this entry --- which was among the largest --- for more than three years in the waste-book? You did not wish it to be known that Roscius owed you this. Then why were you writing it down at all? You had been asked not to enter it. Then why did you have it written in the waste-book? But although I see that all this is firm enough, still I cannot satisfy myself unless I take from Gaius Fannius's own self the testimony that this money is not owed to him. What I attempt is great; what I promise is hard. Unless Roscius shall have the same man as both adversary and witness, I do not wish him to win.

\ciceroSection{10} A definite sum was owed you, which is now claimed before the iudex (the judge in the case of a definite sum), in which a sponsio (the legal stake) for the lawful share has been entered. If here you have claimed even one sesterce more than was owed you, you have lost your case --- because a iudicium (a trial of definite sum) is one thing, an arbitrium (an arbitration of an indefinite claim) another. The iudicium is for definite money, the arbitrium for indefinite. To the iudicium we come in such a way as to obtain or to lose the entire suit. To the arbitrium we approach in such a frame of mind as to get neither nothing nor as much as we asked for. The very words of the formula are evidence of this.

\ciceroSection{11} What is in the iudicium? It is direct, harsh, simple: "If it appear that one thousand sesterces ought to be paid..." Here, unless he makes it plain that one thousand sesterces are owed him to the last farthing, he loses his case. What is in the arbitrium? Mild, moderate: "However much it is more equitable and better that be paid." There the claimant himself confesses that he is asking for more than is owed; but he says he has more than enough in what is granted to him by the arbitrator. So the one trusts in his case, the other distrusts.

\ciceroSection{12} Since this is so, I ask you why you, of all things, made an arrangement to refer this money --- these very fifty thousand sesterces, the credit of your books --- to a compromise; why you took an arbitrator under the formula "however much it is more equitable and better that be paid," and on the further pledge "if such be found." Who, in this matter, was the arbitrator? Would that he were even at Rome! He is at Rome. Would that he were here in court! He is here. Would that he sat on the council of Gaius Piso! He is Gaius Piso himself. Were you taking the same man as arbitrator and as judge? On the same man you were calling for unlimited indulgence and binding him to the narrowest formula of the sponsio? Who ever obtained from an arbitrator as much as he asked? No one --- for he asked for as much as it was equitable should be given him. The matter on which you went to an arbitrator, you have come on to the iudex! Other people, when they see their case is shaky before the iudex, ---

\ciceroSection{13} take refuge with the arbitrator: this man dared to come from the arbitrator to the iudex! And when he took an arbitrator about the credit of the books for this very money, he had judged that the money was not owed him. Two parts of the case are already wrapped up: he says he did not pay the sum down; he does not say he entered it as a debt --- since he does not read out the books. It remains only that he should say that he had a stipulatio (a formal verbal contract); for besides this I do not see how he can claim a definite sum. You stipulated --- where, on what day, at what time, in whose presence?

\ciceroSection{14} Who says that I made a sponsio? No one. Were I to make an end of speaking here, I should seem to have done enough for my own credit and care, enough for the case and the controversy, enough for the formula and the sponsio, enough also for the iudex --- to show that judgement ought to be given for Roscius. A definite sum was claimed; with the third part the sponsio was made. This money must necessarily either have been given, or entered as a debt, or stipulated. That it was not given, Fannius admits; that it was not entered as a debt, Fannius's books confirm; that it was not stipulated, the silence of the witnesses concedes. What then?

\ciceroSection{15} Because both the defendant is the man for whom money has always been the lightest of things and reputation the most inviolate, and the iudex is one whom we wish to think well of us no less than to give judgement for us, and the body of advocates here is such that, on account of its remarkable distinction, we ought to revere it as a single iudex, we shall speak as if all lawful trials, all honorary arbitrations, all duties of household life were summed up and enclosed in this formula. The earlier speech was necessary; this will be voluntary. The earlier was for the iudex, this is for Gaius Piso; the earlier was for the defendant, this is for Roscius; the earlier was prepared for the sake of victory, this for the sake of his good name.

\ciceroSection{16} You claim money, Fannius, from Roscius. What money? Say it boldly and openly. Is it what is owed you out of the partnership, or what was promised and held out to you out of his liberality? The first is graver and more odious, the latter slighter and easier. What is owed out of the partnership? What say you? This is now no light matter, nor to be carelessly defended. For if there are any private trials of the highest reputation and almost (I would say) of life and death, these are three: of fiducia (trust), of tutela (guardianship), of societas (partnership). For it is equally treacherous and unspeakable to break the faith which sustains life, to defraud a ward who has come into your guardianship, and to deceive a partner who has joined you in business. Since this is so, ---

\ciceroSection{17} let us consider who it is that has cheated and deceived a partner. The life lived already will at this point silently give a firm and weighty witness, in one direction or the other. Quintus Roscius? What say you? Is it not so, just as fire thrown into water is at once put out and cooled, so that a hot false charge dropped into the purest and chastest life is at once extinguished and dies? Roscius cheated a partner! Can this offence cling to this man? --- a man who, by Hercules --- and I say it boldly --- has more good faith than art, more truth than discipline; whom the Roman people consider a better man than an actor; who is so worthy of the stage by his art that he is no less worthy of the senate house by his self-restraint.

\ciceroSection{18} But why am I so foolish as to speak of Roscius before Piso? An unknown man, surely, I am commending in many words! Is there anyone of all men of whom you think more highly? Is there anyone who seems to you purer, more modest, kindlier, more dutiful, more generous? What of you, Saturius, who come against him --- do you think otherwise? As often as you have hit upon his name in the case, have you not as often called him a good man and named him with respect --- which no one is accustomed to do for any but a most honourable or most friendly man?

\ciceroSection{19} In which matter you have struck me as ridiculously inconsistent --- that you wound and praise the same man, and call him both a most excellent gentleman and a most dishonest one. The same man you addressed with respect, called a leading man, and accused of having cheated a partner? But, I take it, you assigned the praise to truth and the charge to favour. About him, you said what you thought; the case you conducted at Chaerea's discretion. Roscius cheated! It is absurd to all ears and minds. What if at last he had encountered some timid, witless, rich, lazy man unable to take legal action, ---

\ciceroSection{20} even so, it would be incredible. But let us look at whom he is supposed to have cheated. Roscius cheated Gaius Fannius Chaerea! I beg and beseech you, you who know them, to compare the lives of the two with each other; you who do not know them, to consider the face of each. Does not his very head, and those eyebrows shaved away to the skin, seem to reek of malice and to cry out cunning? From the tips of his toes to the very crown of his head, if the silent figure of the body offers any inference to men, does he not seem entirely composed of fraud, deceit, and lies? --- a man who has his head and eyebrows always shaved that he may be said to have not a hair of a good man on him; whose character Roscius has finely played on the stage --- and yet receives no equal gratitude in return for the favour. For when he plays that most dishonest and most perjured pander Ballio, he plays Chaerea: that filthy, foul, hateful character is impressed in this man's manners, nature, and life. Why he should think Roscius like himself in fraud and malice I can scarcely see, unless perhaps because he sees how splendidly Roscius plays his part in the role of the pander.

\ciceroSection{21} For this reason, again and again, Gaius Piso, consider who is said to have cheated whom. Roscius Fannius! What is this? An honest man cheating a dishonest one, a modest one an immodest, a chaste one a perjurer, an unskilled one a cunning one, a generous one a greedy one? Incredible. Just as, if Fannius were said to have cheated Roscius, both things would seem likely from each man's character --- both that Fannius did it through malice and that Roscius was deceived through unwariness --- so, when Roscius is charged with having cheated Fannius, both are incredible: both that Roscius sought anything through avarice, and that Fannius lost anything through goodness.

\ciceroSection{22} Such are my opening points; let us look at the rest. Quintus Roscius cheated Fannius of fifty thousand sesterces. Why? Saturius smiles, the old fox, as he supposes; he says, "for the sake of those very fifty thousand." I see; but I ask why he was so violently set on those very fifty thousand. For you, Marcus Perpenna, you, Gaius Piso, surely fifty thousand would not have weighed so much as to make you cheat a partner. Why fifty thousand should have meant so much to Roscius --- I ask the cause. Was he in want? On the contrary, he was wealthy. Was he in debt? On the contrary, he was managing his own money freely. Was he greedy? On the contrary, even before he was wealthy, he was always most generous and most munificent.

\ciceroSection{23} By the faith of gods and men! He who refused to make 250,000 sesterces a year as profit --- for he could and ought to have earned 250,000 sesterces, if Dionysia can earn 200,000 --- he, by the highest fraud and malice and treachery, hunted after fifty thousand? And the one was a vast sum, the other a tiny one; the one honourable, the other sordid; the one pleasant, the other bitter; the one his own, the other staked on a case at law. In these last ten years he could most honourably have earned 6,000,000 sesterces. He refused. He took on the labour of profit; he rejected the profit of his labour. He has not yet stopped serving the Roman people; he stopped serving himself long ago.

\ciceroSection{24} Would you, Fannius, ever have done this? If you could have taken in such profits, would you not at the same moment have been performing both gestures and your last gasp? Now you say that you have been swindled out of fifty thousand sesterces by Roscius --- a man who refused such great and unlimited sums, not from idleness in labour, but from the magnificence of his generosity! Why should I now mention what I am sure occurs to you? Roscius cheated you in a partnership! There are laws, there are formulas established about everything, so that no one may err either in the kind of injury or in the form of the action. Formulas for every man's loss, pain, inconvenience, calamity, injury have been formed by the praetor publicly, to which a private suit is fitted.

\ciceroSection{25} Since this is so, I ask why you did not haul Quintus Roscius before an arbitrator under the formula "for partnership." Did you not know the formula? It was very well known. Did you not wish to test it in a serious trial? Why? On account of old intimacy? Why then are you wounding him? On account of the man's integrity? Why then are you accusing him? On account of the gravity of the offence? Indeed? The man whom you could not catch by arbitrator --- where the trial would have been on the very point --- you will condemn him by iudex, where there is no arbitrium on the point at all? Either bring this charge where you may bring it, or do not throw it about where you ought not. And yet by your own testimony the charge has already been removed; for at the time when you refused to use that formula, you indicated that this man had committed no fraud upon the partnership. For tell me: do you have your books or not? If not, how is there a contract?

\ciceroSection{26} If you have them, why do you not name them? Now say that Roscius asked you to take his close friend as arbitrator! He did not ask. Say that he made a deal that he should be acquitted! He made no deal. Ask why he was acquitted! Because he was a man of the highest innocence and integrity. For what happened? You came of your own accord to Roscius's house, you settled the matter; for what you had rashly entered upon, you asked him to forgive you for serving the citation; you said you would not appear; you cried out that he owed you nothing on account of the partnership. He gave notice to the iudex of this. He was acquitted. And yet you dare to make mention of fraud and theft? He persists in his shamelessness. "He had made a contract with me," he says. To prevent his being condemned, of course. What reason had he to fear condemnation? --- the matter was plain, the theft open.

\ciceroSection{27} Of what theft? He begins, with great expectation, to expound the partnership of the old actor. "Panurgus," he says, "was Fannius's; he becomes shared between him and Roscius." Here Saturius first complained, not lightly, that Panurgus was made joint-property gratis with Roscius, when he had been Fannius's by purchase. Of course Fannius --- generous, dissolute, overflowing with kindness --- gave him as a gift to Roscius. So I take it.

\ciceroSection{28} Since he stopped here a little, I too must dwell here a little. You say, Saturius, that Panurgus was Fannius's own. I, on the contrary, contend that he was wholly Roscius's. For what was Fannius's? The body. What was Roscius's? The training. The face was nothing; the art was costly. As the part that was Fannius's, he was not worth a thousand sesterces; as the part that was Roscius's, he was worth more than 200,000. For nobody looked at him from the trunk of his body; everyone valued him by the comic art --- for those limbs by themselves could not earn more than the prentice's pay; the training, which had been imparted by my client, hired itself out for no less than 200,000.

\ciceroSection{29} What an unfair, unworthy partnership, where one brings a thousand to the partnership and the other 200,000! Unless that is why you take it ill: that you brought a thousand out of the strongbox, and Roscius drew 200,000 from the training and the art. For what hope and expectation, what zeal and what favour Panurgus brought with him onto the stage --- being Roscius's pupil! Those who esteemed him favoured the pupil; those who admired him approved the pupil; those, finally, who had heard his name reckoned the pupil polished and finished. So is the multitude: it judges few things from truth, many from opinion.

\ciceroSection{30} Few took notice of what he knew; everyone asked where he had learned it. They thought that nothing depraved or perverse could be put forth by this man. If Panurgus had come from Statilius, even though he surpassed Roscius in art, no one could have looked at him; for no one would think that, just as a good son cannot be born of a worthless father, a good comic actor can come from a wretched one. Because he came from Roscius, he seemed to know even more than he knew. The same recently happened with Eros the actor: who, after he was hissed off the stage --- not only with whistles, but with abuse --- took refuge as at an altar in this man's house, his discipline, his patronage, his name. So in a very short time, one who had not been even among the latest actors came to the foremost rank of comic players.

\ciceroSection{31} What lifted him up? The single recommendation of this man --- who nevertheless took the famous Panurgus into his home, not just so that he could be called Roscius's pupil, but trained him at the cost of the highest labour, frustration, and suffering. For the cleverer and more talented a man is, the more irritably and laboriously he teaches; for what he himself grasped quickly, when he sees it slowly perceived, he is racked. My speech has been carried a little farther, that you may exactly know the condition of the partnership.

\ciceroSection{32} What followed then? "Panurgus," he says, "this slave of ours in common, was killed by a certain Quintus Flavius of Tarquinii. In this matter," he says, "you appointed me your representative. The suit having been registered, the trial for unlawful damage having been set up, you settled with Flavius without me." For half my share, or for the whole? Let me put it more plainly: for myself, or both for myself and for you? For myself; I could do so by the example of many; it was permitted; many have done so by right; in that I did you no injury. Sue for your own; demand and carry off what is owed; let each man hold and pursue his share of the right. --- "But you have managed your business well." --- Manage your own well too. --- "You settled your half for a great sum." --- Settle your half for a great sum too. --- "You took five hundred thousand sesterces." --- So be it; take you five hundred thousand too.

\ciceroSection{33} But this settlement of Roscius's, you may exalt by speech and by impression; in fact and in truth you will find it ordinary and slight. For he received a piece of land in those times when prices of estates were lying low; an estate which had no farmhouse and was not under cultivation in any part; which is now worth far more than it was then. And no wonder. For at that time, owing to the calamities of the commonwealth, the holdings of all were uncertain; now, by the kindness of the immortal gods, the fortunes of all are settled. Then the land was uncultivated, without a roof; now it is most fully cultivated, with the best of farmhouses.

\ciceroSection{34} But, since you are so malevolent by nature, I will never free you from that worry and care. Roscius transacted his business splendidly, took away a most productive estate; what is that to you? Settle your half-share, however you please. Here he changes his ground and tries to feign what he cannot prove. "You settled the whole matter," he says. So the whole case is reduced to this: whether Roscius made the agreement with Flavius for his own share or for the entire partnership. For I confess that, ---

\ciceroSection{35} if Roscius touched anything in the joint name, he ought to render it to the partnership. He acquired the partnership's suit, not his own private one, when he received the farm from Flavius. Why then did he not give security that no one further should claim a sesterce? He who settles for his own share leaves intact the action of the rest; he who settles on behalf of partners gives security that none of them shall claim afterwards. Why did it not occur to Flavius to take security for himself? He must, of course, not have known that Panurgus was in the partnership. He did know. He did not know that Fannius was Roscius's partner. Excellent --- for Fannius too had a suit registered with him.

\ciceroSection{36} Why then does he settle and not take a counter-stipulation that no one shall claim further? Why does he depart from the farm and is not acquitted in the iudicium? Why does he act so unskilfully as to bind Roscius by no stipulation, nor to be acquitted by Fannius in court?

\ciceroSection{37} This is, first, both from the condition of the law and from the customs of legal precaution, the firmest and weightiest argument, which I would embrace at greater length, were I not to have other surer and clearer testimonies in the case. And, lest you should declare that I have promised this in vain, I will, you yourself, Fannius, raise from your benches as witness against yourself. What is your charge? That Roscius settled with Flavius on behalf of the partnership. When? Fifteen years ago. What is my defence? That Roscius transacted with Flavius for his own share. You give a counter-promise to Roscius three years ago. What? Read out that counter-stipulation more plainly. Pay attention, please, Piso. I am compelling Fannius, unwilling and twisting this way and that, to give witness against himself. For what does the counter-stipulation cry out? "Whatever I shall have got from Flavius, I promise to pay half of it to Roscius." This is your voice, Fannius.

\ciceroSection{38} What can you take from Flavius, if Flavius owes nothing? What is this man now further stipulating about, what he has already long ago himself collected? Why, indeed, will Flavius give you anything, who has paid Roscius all that he owed? Why, in a matter so old, in a business already finished, in a partnership dissolved, is this new counter-stipulation slipped in? Who is the writer of this counter-stipulation, witness, and arbiter? You, Piso. For you asked Quintus Roscius, on account of his service and labour as representative, on account of his attendance at the citations, to give Fannius fifty thousand sesterces, on this condition: that, if he should obtain anything from Flavius, he should pay half of it to Roscius. Does the very counter-stipulation seem to you to say plainly enough that Roscius settled for himself?

\ciceroSection{39} But perhaps it may occur to you that Fannius re-promised Roscius half of what he should obtain from Flavius, but obtained nothing at all. What of it? You ought to look not at the issue of the collection but at the beginning of the re-promise. Even if Fannius judged it not worth pursuing, then on his side he judged that Roscius had won his own private suit, not the partnership's. What if, at long last, I make this plain --- that, after Roscius's old settlement, after this recent counter-promise of Fannius for fifty thousand sesterces, Fannius did obtain that fifty thousand from Flavius in the name of Panurgus? Will he still dare to mock the good name of an excellent man, Quintus Roscius?

\ciceroSection{40} A little while ago I was asking --- and the question was very much to the point --- why Flavius, when he was making a settlement of the whole suit, did not take adequate security from Roscius nor get acquittal in court from Fannius. Now, indeed --- which is wonderful and incredible --- I ask: why, when he had settled the entire matter with Roscius, did he separately pay fifty thousand to Fannius? At this point, Saturius, what you can answer I am eager to know: whether Fannius did not at all obtain fifty thousand from Flavius, or obtained it on some other heading and on some other ground.

\ciceroSection{41} If on some other ground, what business had you with him? None. Was he condemned to you? No. I am wasting my time in vain. "He did not obtain fifty thousand from Flavius at all," he says, "neither in Panurgus's name nor in anyone's." If I make it plain that, after this recent stipulation of Roscius's, you took fifty thousand from Flavius, is there anything for it but that you should depart from court most disgracefully beaten? With what witness, then, shall I make this plain?

\ciceroSection{42} The matter, I take it, had come into court. Certainly. Who was the plaintiff? Fannius. Who was the defendant? Flavius. Who was the iudex? Cluvius. Of these, one I must produce as witness, who shall say that the money was paid. Which is the weightiest of these? Without controversy, the iudex --- a man approved by the verdict of all. Which of these three, then, will you expect from me as witness? The plaintiff? It is Fannius; he will never give evidence against himself. The defendant? It is Flavius. He died long ago; if he were alive, you would hear his words. The iudex? It is Cluvius. What does he say? That Flavius paid fifty thousand sesterces to Fannius in the name of Panurgus. If you regard him by the census, he is a Roman knight; if by his life, a man of the highest distinction; if by good faith, you took him as iudex; if by truth, what he could and ought to know, he has said.

\ciceroSection{43} Deny --- come, deny it now --- that one ought to believe a Roman knight, an honourable man, your own iudex! He looks around, he is in a sweat, he denies that we shall read out Cluvius's testimony. We will read it. You are mistaken; you console yourself with an empty and slender hope. Read the testimony of Titus Manilius and Gaius Luscius Ocrea, two senators, men of the highest distinction, who heard from Cluvius. Testimony of Titus Manilius and Gaius Luscius Ocrea. Are you saying that one ought not to believe Luscius and Manilius, or also that Cluvius is not to be believed? I will say it more plainly and openly. Is it that Luscius and Manilius did not hear anything about the fifty thousand from Cluvius, or that Cluvius told a falsehood to Luscius and Manilius? On this point I am of an untroubled and quiet mind, and I am not greatly concerned how your answer falls; for by the firmest and most inviolate testimonies of the best men, the case for Roscius is fortified.

\ciceroSection{44} If you have already made up your mind whose oath-bound testimony to deprive of credit, answer. Do you say Manilius and Luscius are not to be believed? Say it; dare to. It is the voice of your stubbornness, your arrogance, and your whole life. What are you waiting for --- when am I going to say that Luscius and Manilius are senators by rank, advanced in years, by nature inviolate and devout, well-furnished and rich in family resources? I will not do it; I will not detract from myself anything, when I give back to those of mature age the very strict reward they have earned. My young manhood needs their good opinion more than their most rigorous old age desires my praise.

\ciceroSection{45} You, Piso, must deliberate long and digest carefully which to believe rather: Chaerea, unsworn, in his own suit, or Manilius and Luscius, sworn, in another's case. It remains for him to contend that Cluvius told a falsehood to Luscius and Manilius. If he does that --- with what shamelessness! --- will he reject as witness the man he approved as iudex? Will he say that the man he himself trusted ought not to be believed? Will he weaken before the iudex the trust of the witness on whose good faith and devoutness he chose him as iudex? The man whom, if I tendered him as iudex, he ought not to refuse, will he dare reproach when I produce him as witness?

\ciceroSection{46} "But he speaks unsworn," he says, "to Luscius and Manilius." If he were speaking sworn, would you believe him? But what is the difference between a perjurer and a liar? He who is in the habit of lying is in the habit of perjuring himself. The man whom I can lead to lie, I can easily prevail on to perjure. For he who has once turned aside from truth is accustomed to be brought to perjury with no greater compunction than to lying. For who is moved by a deprecation of the gods, but not by the trust of conscience? Therefore, the punishment which the immortal gods established for the perjurer, the same is established for the liar; for not from the agreement of the words by which the oath is comprehended, but from the treachery and malice by which an ambush is laid for someone --- this is what the immortal gods are accustomed to be angry with men over, and to take ill.

\ciceroSection{47} But I, on the contrary, contend: Cluvius's authority would be slighter if he spoke sworn than it is now, when he speaks unsworn. For then perhaps he might seem to wicked men too eager to be a witness in a matter where he had been iudex; now to all but the unjust he must necessarily seem most chaste and most consistent, who tells what he knows to his close friends.

\ciceroSection{48} Say now, if you can, if the case, if the matter allows, that Cluvius lied! Did Cluvius lie? Truth herself laid her hand on me and forced me to stand and pause a little. Whence comes this whole made-up and confected lie? Roscius, of course, is a clever and crafty man. With this beginning he started to think: "Since Fannius is suing me for fifty thousand sesterces, I will ask Gaius Cluvius, a Roman knight, a most distinguished man, to lie on my account; to say that a settlement was made which was not made; to say that fifty thousand was paid by Flavius to Fannius which was not paid." This is the start of a dishonest mind, of a wretched intellect, of no judgement.

\ciceroSection{49} What then? After he had splendidly fortified himself, he came to Cluvius. What kind of man? A light one? On the contrary, the gravest. A volatile one? On the contrary, the most consistent. An intimate? On the contrary, an utter stranger. After he had greeted him, he began, blandly and elegantly no doubt, to ask: "Lie on my account; in the presence of the best men, your own intimates, say that Flavius settled with Fannius about Panurgus when he made no settlement; say that he paid fifty thousand who paid not a sesterce." What did the other reply? "I will indeed lie eagerly and willingly on your account; and, if you should ever wish me to commit perjury, that you may save yourself a trifle, know that I shall be ready. There was no need to take so much trouble and come to me; you could have transacted by messenger this matter, which was so trifling."

\ciceroSection{50} By the faith of gods and men! Would Roscius ever have asked this from Cluvius, even if it were a hundred million sesterces in court? Would Cluvius have granted it to Roscius's request, even if he were sharing in the whole booty? Hardly, by Hercules, would you, Fannius, dare to ask this of Ballio or someone like him, and obtain it. What is false in fact, is also incredible in reason; for I forget that Roscius and Cluvius are men of the first rank; for the sake of argument I imagine them dishonest.

\ciceroSection{51} Roscius suborned Cluvius as a false witness! Why so late? Why when the second instalment was being paid, not when the first was? For he had already paid fifty thousand before. Then, if Cluvius had already been persuaded to lie, why did he say that fifty thousand --- rather than 300,000 --- had been given by Flavius to Fannius, when by the counter-stipulation half of that would have been Roscius's? You see now, Gaius Piso, that Roscius asked nothing for himself except for himself, nothing for the partnership. Since Saturius perceives this is plain, he does not dare to resist and fight against the truth, but finds another by-way of fraud and ambush in the same track.

\ciceroSection{52} "I confess," he says, "that Roscius asked his own share from Flavius; I grant that he left Fannius's free and intact; but I contend that what he collected for himself became common to the partnership." Nothing more captious or more unworthy can be said. For I ask: could Roscius claim his share out of the partnership or not? If he could not, how did he take it? If he could, how did he not collect it for himself? For what is sued for for one's self, certainly is not collected for another.

\ciceroSection{53} Or is it so: if he had sued for what was the partnership's, all should have shared equally what was then realised; now, since he sued for what was his own share, is what he then took to be regarded as collected for himself alone? What is the difference between the man who litigates on his own behalf and the man who is appointed cognitor (a procurator)? He who registers a suit on his own brings it for himself alone; for another no one can, except one who is made cognitor. Indeed? If he had been your cognitor, what he had won by the trial you would carry off as your own; when he sued in his own name, what he carried off you say he did not collect for himself but for you?

\ciceroSection{54} If anyone can sue for another who has not been made cognitor, I ask: why, when Panurgus was killed and the suit registered with Flavius for unlawful damage, did you become cognitor of Roscius in that suit, especially since on your reasoning whatever you sought for yourself you sought for him, whatever you collected for yourself fell to the partnership? But if nothing of what you obtained from Flavius would have come to Roscius unless he had given you as cognitor in his own suit, nothing of what Roscius collected for his own share ought to come to you, since he has not been made your cognitor.

\ciceroSection{55} For what will you be able to answer, Fannius? When Roscius transacted with Flavius about his own share, did he leave you your action or not? If he did not leave it, how did you afterwards collect fifty thousand from him? If he left it, why do you ask of him what you ought to pursue and demand for yourself? For partnership is most like, and almost the twin of, inheritance. Just as a partner has a share in the partnership, so an heir has a share in the inheritance. As an heir sues for himself alone, not for his coheirs, so a partner sues for himself alone, not for his partners; and just as each sues for his own share, so he discharges for his own share --- the heir from the share by which he came to the inheritance, the partner from the share by which he entered the partnership.

\ciceroSection{56} As Roscius could in his own name forgive his share to Flavius, so that you should not claim it, so when he claimed his own share and left you the action whole, he is not bound to share with you, unless perhaps, by some perverted custom, what is his you cannot extort from another but you can snatch from him. Saturius persists in his opinion, that whatever a partner sues for himself becomes the partnership's. If that is so, by what folly, in heaven's name, did Roscius act, when, on the counsel and authority of jurists, he carefully took back a counter-stipulation from Fannius that he should pay him half of what he should obtain from Flavius --- if indeed, even without security and counter-promise, Fannius nonetheless owed it to the partnership, that is, to Roscius? * * * * * * * * *

% [Lacuna: the rest of the speech is lost.]
