% Pro Sex. Roscio Amerino (pro-roscio-amerino) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} I imagine, gentlemen of the jury, that you wonder what it is that has happened: that with so many of the leading orators and the noblest men in Rome sitting here, I in particular should have stood up --- I, who in years and in talent and in standing am not to be compared with those who sit. Every one of these men whom you see in attendance for the defence holds that an injury contrived by an unprecedented criminality ought to be opposed; but to oppose it themselves they do not dare, given the harshness of the times. So they are present, because duty requires it, and silent, because they want to avoid the danger.

\ciceroSection{2} What then? Am I the boldest of all? Not in the least. Or so much more dutiful than the rest? I am not so eager for that praise either as to want it snatched from others. What is it, then, that has driven me beyond the rest to take up the cause of Sextus Roscius? Because if any of those gentlemen had spoken, men whose authority and standing are very great, and had let fall a single word about the state of the commonwealth --- as in this case must of necessity be done --- he would be thought to have said far more than he had in fact said.

\ciceroSection{3} If, on the other hand, I speak openly everything that needs to be said, my words will not, even so, get out and spread among the public in the same way. Then again, what others say cannot pass unnoticed, given their high birth and standing; nor can it be excused as said rashly, given their age and prudence. As for me, if I have spoken with some freedom, it can either pass unnoticed, since I have not yet entered public life; or it can be pardoned to my youth --- though by now both the principle of pardoning and the practice of taking notice have been wholly abolished from our state.

\ciceroSection{4} There is also this further reason: that perhaps the others were asked to speak in such a way as left them free to think they could either accept or refuse without compromising their duty. I was approached by men whose friendship, whose kindness, whose standing weighs heaviest with me, men whose kindness towards me I could not ignore, whose authority I could not slight, whose request I could not disregard.

\ciceroSection{5} For these reasons I have stood up to speak, not as the man chosen with the most ability, but as the one with the least responsibility left over from other engagements; and not so that Sextus Roscius might be defended by an advocate of standing, but so that he should not be left wholly without one. Perhaps you will ask: what is this case, that scares off so many of the leading men? You will not wonder when you hear what it is. For I think you will judge that no one could possibly have been left more abandoned and helpless than the man for whom I now must speak.

\ciceroSection{6} I will set out the whole charge --- and the manner of the prosecution --- as briefly as I can, gentlemen. Then you will easily understand both why this affair has fallen to me and what the situation is. The father of this Sextus Roscius, a man of Ameria --- by far the first man in his municipality, and not only by birth but by character and means --- and indeed in standing among the leading men of all Italy, given his connections of friendship and hospitality with the most distinguished houses of our city, ---

\ciceroSection{7} this man, with all his rank in the municipality and his reputation in the wider world, with all the praise that high character could earn, was killed; and although his death is, in itself, the matter of accusation, the case of his son is what comes to trial. The son, who in this case is the defendant, is here. He is charged with parricide. What sort of son? You will understand when I have set out the rest in order.

\ciceroSection{8} For the murder, the persons accused are those whom I will name. As for who killed Sextus Roscius the elder --- whether it was those who, as I will show, gained most by his death; or those who feared him most while he was alive and rejoiced most when he was dead; or those who at the time were already implicated in similar atrocities --- this you will judge for yourselves once the full circumstances are laid out.

\ciceroSection{9} Sextus Roscius the father, gentlemen, was a man of the old school, a Roman of rank in his own town, devoted to the leading citizens of Rome, especially to the Metelli, the Servilii, and the Scipios --- houses whose names I mention with respect, since their friendship was both an ornament to him and an honour. While he lived, he visited Rome only as a private citizen who held a place of dignity in their banquets and in their councils. He was a friend, in particular, of the late Caecilia, sister of Nepos, daughter of Balearicus --- a woman whose memory the better sort of citizens are bound to cherish; in her household this Sextus Roscius the younger took refuge when his father had been killed, and there he was sheltered.

\ciceroSection{10} For this woman, by her tenacity in keeping faith, has held to the dictates of an ancient and noble decency in our days, days in which we are scarcely able to recognise such virtue any longer. She has not allowed the son of a friend, the man whom she had received under her protection, to be cast forth into the streets, even when his fortune was at its lowest, when his enemies were at their highest, when by the law of the times nothing was further from prudence than to take in such a man.

\ciceroSection{11} Through her courage, her fidelity, her diligence, what we have on our hands now is a defendant accused while still living, and not a corpse stripped of all rights. For you must know, gentlemen, that those who killed the father of this Sextus Roscius wanted, on top of the killing, to disinherit, despoil, and destroy the son. They are men who can endure no witness to their crimes; they wish them to be shrouded in such a way that not even rumour can reach the public. And so, though they have driven the son from his ancestral lands, they are seeking even now, on top of what they have done, his blood. They could not be content with how they had glutted their cruelty, unless they added this final touch to the rest.

\ciceroSection{12} Bitter is what we suffer, gentlemen; bitter and unworthy: but more endurable, perhaps, if those who do these things were our enemies, or strangers, or barbarians. As it is, the men I will name --- you will judge --- were of our own household: they ate at our table, they shared in our family business, they were treated as kin. From these men this defendant has had to defend his life, his property, and his good name. From these men he must now defend himself before you. Whatever you decide, you must decide with full understanding of what they did, of what they sought, and of what they intend.

\ciceroSection{13} I will go through the matter from the beginning. Sextus Roscius the father, a citizen of Ameria, in standing the foremost man of his town, in property well-furnished, in friendships at Rome very richly endowed --- he was, as I have said, on close terms with the noblest houses of our city. There was, in his own neighbourhood, a man named Titus Roscius Magnus, and another named Titus Roscius Capito, both of his own family, both equally ill-disposed towards him, but unequally distinguished in their lives: Capito the elder thug, Magnus the cleverer one. From these two men, his kinsmen, came the trouble that destroyed him.

\ciceroSection{14} For these men had quarrelled with him over a long-standing grudge, and as their quarrel grew, they began to look about for some way of getting rid of him. Now Sextus Roscius the elder used to come down to Rome from time to time. One night, on his way back from supper, near the Pallacinian baths, he was set upon and killed. The kinsmen, who had long planned for this moment, immediately seized the opportunity that the death gave them.

\ciceroSection{15} For there was at this time a powerful man in Rome named Lucius Cornelius Chrysogonus --- I do not need, I think, to remind you who he was; you know that he was the most influential of all the freedmen of Lucius Sulla, and that for several years he held a very free hand in the public proscriptions. With this Chrysogonus the kinsmen of the elder Roscius opened a private negotiation. Once the murder was done, before the news could reach Ameria by any reliable channel, they had a courier on the road from Rome to the elder Roscius's neighbours --- a courier of theirs, in their pay, with their instructions. The decuriones of Ameria, hearing of the death, sent messengers to Sulla, urging him to set apart the property of the murdered man, since it was illegal to confiscate the goods of one who had not been proscribed. Their messengers reached Sulla's camp at Volaterrae --- but they did not see Sulla.

\ciceroSection{16} For Chrysogonus, in the meantime --- working with the kinsmen --- had pulled it off. Within a few days the property of Sextus Roscius the father, an estate worth some six million sesterces, was put up at auction, sold for two thousand sesterces, and bought in by Chrysogonus himself --- though some of the most valuable farms passed at once to Magnus and Capito as their share. The decuriones' deputation arrived at the camp; they were intercepted by Chrysogonus, given assurances that all would be done as they wished, and sent home. Sulla never even heard of the matter. The corpse was barely cold when the booty was being divided.

\ciceroSection{17} Now of all the murdered man's goods the son got nothing. He was driven from the house, expelled from the lands, denied even the funeral rites of his father in their proper manner. He fled, naked of all support, to Caecilia, of whom I have spoken. Through her, he came at last under the patronage of Marcus Messalla, a young man whose nobility you all see, whose energy and decency are conspicuous. Messalla, finding the son entirely alone, with no one to take his part, with not a pretext of help to be had from anywhere, agreed at last to bring the case before this court.

\ciceroSection{18} While these things were being done in Ameria, Magnus and Capito enjoyed for a time their share of the estates. Then, fearing that someone might at last begin to ask questions, they conceived a fresh plan: to get rid of the son too. Their first attempt was to have him killed on the spot. When that failed, when the son had escaped to Rome, they decided to murder him in Rome. When that too proved impractical, they hit upon a more elegant device --- they prosecuted him for the murder of his own father.

\ciceroSection{19} For if they could secure his condemnation on a charge of parricide, they would not only put a definitive end to the only witness who could testify against them: they would also so cover their own crime under the colour of justice that no one would ever again hear a word said against them. The infamy of the parricide-conviction would silence everyone --- and the goods now in Chrysogonus's hands would be theirs in safety forever. So they have come into court to prosecute, and they have made an arrangement with one Erucius, who carries the prosecution for them, that he is to make speeches and they are to remain in the background while my client is being charged with their crime.

\ciceroSection{20} Now, gentlemen of the jury --- such is the whole matter. The son of a man whom they killed, whom they robbed of his estate, whom they cast from his ancestral lands, this same son they now charge with parricide. Could anything more outrageous be conceived? Could anything more shameless be imagined? In what court of the Roman people, in what age of the Republic, has such a thing been heard? But of all this you will hear more, in detail, as I proceed.

\ciceroSection{21} Let me begin, then, by saying who Sextus Roscius the elder was, what kind of life he led, what kind of household he kept, and how it stood between him and his kinsmen who now charge his son. He was, gentlemen, a man of the old Roman virtues. He had no acquaintance with luxury; he had every acquaintance with industry. He worked his lands; he managed his establishment; he raised a son. He was loyal to the friendships he had inherited from his father; he was generous to the friends he had made for himself. He was very rich, and he carried his riches without ostentation.

\ciceroSection{22} The kind of son he raised, you may judge from the fact that, while his father lived, he never set foot in Rome. He spent his life on the family estates, working with his own hands when work was wanted, supervising others when supervision was wanted, knowing little of city affairs and less of city pleasures. From this you can see what kind of person you are now asked to convict of parricide. A young man, raised in country quietness, ignorant of every luxury, ignorant of every refinement of vice --- and condemned, on no evidence whatever, to a charge that has scarcely been heard at Rome twice in living memory.

\ciceroSection{23} The case for the prosecution, when you strip away its rhetorical flourishes, comes to this: that this defendant murdered his father. The evidence offered? None. The motive supplied? None worth a man's consideration. The means, the place, the moment of the alleged killing? Nothing definite is said. We are simply asked to believe that he did it because Erucius says he did. Erucius! Who, when his fee has been paid, will go home and forget that Sextus Roscius ever existed; whose voice you have only heard so loudly today because he has been hired to make a noise.

\ciceroSection{24} I shall divide my whole defence into three parts. First, I will speak about the charge of parricide, and refute Erucius's argument piece by piece. Second, I will speak about the seizure of the property and the persons involved in that seizure. Third, I will speak about Chrysogonus's part in the matter --- because, although I shall do it at the cost of some of his displeasure, the truth in this case cannot be told without naming him.

\ciceroSection{25} On the first part, then. Erucius made a long speech, but his charge has substance only at one point: he says my client wished to kill his father because of a quarrel that had risen between them. I ask, gentlemen of the jury, whether anyone has ever before been brought into a court on so slender a thread. There was a quarrel between father and son, says Erucius. What sort of quarrel? Was the son disinherited? No: not even when the father was angriest with him was that ever proposed. Was the son banished from the household? No. Was anything done to the son to suggest that the father wished, in any degree, to be rid of him? Nothing: nothing was done.

\ciceroSection{26} They were so afraid, this Sextus Roscius and his neighbours, that they preferred to die rather than have Sulla informed of what was being done. They were men of the old school, who imagined others to be like themselves; and when Chrysogonus assured them that he would have the name of Sextus Roscius struck off the proscription rolls and the estates handed back to the son intact, and when Titus Roscius Capito --- one of the ten delegates --- promised that this would in fact be done, they believed it. Without ever pleading their case, they went home to Ameria. At first these gentlemen merely put off the matter day by day; then they grew lazier still and toyed with their promises; finally --- as was easily perceived --- they began to plot against the life of this Sextus Roscius, judging that they could not long hold the property of another while its owner was still alive.

\ciceroSection{27} As soon as my client realised this, on the advice of his friends and kinsmen he fled to Rome and put himself under the protection of Caecilia, sister of Nepos, daughter of Balearicus --- a name I mention with respect --- with whom his father had had close and constant friendship. In this woman, gentlemen, even now --- as everyone has always thought --- the traces of the ancient sense of duty remain, almost as a specimen of what once was. She received Sextus Roscius into her house, destitute, driven from his home, dispossessed of his property, fleeing the swords and threats of cut-throats; she stood by a guest who had already been crushed and given up for lost by everyone else. By her courage, her loyalty, and her diligence it has come about that this man stands before you alive on a charge, instead of being entered, dead, on the proscription lists.

\ciceroSection{28} For when these gentlemen perceived that Sextus Roscius's life was being guarded with the greatest care, and that no opportunity of murdering him was to be found, they took up a plan full of crime and audacity --- to lay against him a charge of parricide, to procure for the prosecution some old hand who could speak to a matter on which there was in fact no shred of suspicion, and finally, since they could not win on the merits, to win by the timing alone. The talk that was put about ran like this: that since trials had been suspended for so long, the first man to be brought into court was bound to be condemned; that this defendant would have no advocate, owing to Chrysogonus's standing; that no one would breathe a word about the sale of the estates or about the partnership; that the very word "parricide" and the horror of the charge would, by themselves, be enough to ensure that he was disposed of without any difficulty, since he had no one to defend him.

\ciceroSection{29} Driven by this design, by this very madness, those who could not kill him themselves --- though they wanted to --- have handed him over to you for the killing. What shall I lament first, gentlemen? Where shall I begin? Whose help, and against whom, shall I invoke? The faith of the immortal gods? Or of the Roman people? Or of you, who hold the supreme authority of the state at this moment?

\ciceroSection{30} His father has been murdered, foully. His house has been seized by his enemies. His goods have been taken, occupied, plundered. The son's life has been threatened; he has been hunted with steel and ambush. What kind of crime is missing from all these crimes? And yet on top of all this they pile and heap further outrages: they fabricate an incredible charge; they procure witnesses against him; they hire prosecutors with the very money they took from his father. They offer my wretched client this choice: either give his throat up to Titus Roscius's knife, or be sewn into a sack and lose his life in the most disgraceful manner. They thought he would have no advocates: he has none. But, gentlemen, the man who will speak openly, who will defend him with good faith --- which is what this case really requires --- has not, after all, failed him.

\ciceroSection{31} Perhaps in undertaking the case I was driven a little impulsively by the rashness of youth. Yet, having once undertaken it, even if every threat, every terror, every danger were to bear down on me from every quarter, I will see it through; I will face whatever comes. I have made up my mind, and I have settled it: that I will say everything which I judge relevant to the case --- and not just say it, but say it gladly, boldly, freely. Nothing that may arise, gentlemen, will ever weigh upon me with such force that fear can outweigh good faith.

\ciceroSection{32} Truly, who is so spiritless as to see all this and remain silent and indifferent? You murdered my father, though he had not been proscribed. You then put down his name on the list of the proscribed. You drove me from my own house by force. You hold my patrimony. What more do you want? Have you come now into the court itself, with steel and weapons, in order either to murder this defendant or to procure his condemnation?

\ciceroSection{33} We had recently in our city a man as bold as any in living memory, Gaius Fimbria, and --- as everyone agrees, except those who are themselves madmen --- as mad as any. When he had arranged at the funeral of Gaius Marius for Quintus Scaevola to be wounded --- Scaevola, the most upright and most distinguished man of our state, of whose praises this is not the place to say much (and not much could be said, in any case, beyond what the Roman people preserve in memory) --- and when he learned that Scaevola was likely to recover, he laid an indictment against him. When he was asked what he could possibly find to charge against a man whom no one could really praise enough to do him justice, he is said --- being the maniac he was --- to have answered: "for not having taken the whole weapon into his body." The Roman people have seen nothing more shocking than this --- except the death of that same Scaevola, which was so disastrous that, by his murder, all those whom he had wished to save by negotiation perished with him; the men he had hoped to spare by compromise killed him.

\ciceroSection{34} Is the present action so very different from that saying and that deed of Fimbria's? You bring a charge against Sextus Roscius. Why? Because he escaped from your hands. Because he refused to let himself be killed. The other case, since it concerned Scaevola, looks more shocking; but this one, because it is being done by Chrysogonus, is intolerable. By the immortal gods! What in this case requires defence? Where is the talent of an advocate, where the eloquence of an orator, urgently required? Let us simply set the whole matter forth, gentlemen, and consider it laid out before our eyes. So you will most easily see what is at stake in the whole trial, on what points I ought to speak, and in what direction your judgement ought to move.

\ciceroSection{35} There are three things, so far as I can judge, that stand in the way of Sextus Roscius today: the charge brought by his adversaries, their audacity, and their power. The fabrication of the charge has been undertaken by the prosecutor Erucius. The audacity belongs to the Roscii, who have claimed that part for themselves. And Chrysogonus, the man with most influence, fights with his power. About every one of these things I see I must speak.

\ciceroSection{36} But not in the same way about every one. The first --- the charge --- belongs to my office. The other two have been laid upon you by the Roman people. I must dilute the charge to nothing. You must resist their audacity, and you must extinguish and crush, at the very first opportunity, the dangerous and intolerable power of men of this kind.

\ciceroSection{37} Sextus Roscius is accused of having killed his father. By the immortal gods, what a wicked, what an unspeakable deed! What a crime, in which all crimes seem to be wrapped up in one! For if --- as the wise have so finely said --- filial duty can be wounded merely by an expression of the face, what punishment harsh enough can be found for one who has dealt his parent his death? The man for whom, if matters had required it, divine and human law alike would have compelled him to die in his father's place.

\ciceroSection{38} For so great, so atrocious, so singular a crime --- a crime so rare that, if ever it has been heard of, it is reckoned among portents and prodigies --- with what arguments, Gaius Erucius, do you suppose a prosecutor ought to come forward? Ought he not to show the singular audacity of the man brought to trial, his savage character and inhuman nature, a life given up to every vice and every disgrace, and finally everything in the man wrecked and ruined to the point of self-destruction? Of all this, you have brought forward not one shred against Sextus Roscius --- not even by way of slander.

\ciceroSection{39} Sextus Roscius killed his father. Who is the man, then? A young rake, corrupted, led astray by worthless companions? More than forty years old. A practised assassin, no doubt, a daring fellow, often involved in murders? Not a word of any such thing have you heard from the prosecution. So it must have been luxury, or a great burden of debt, or unbridled passions of the heart, that drove him to this crime. Of luxury Erucius cleared him by saying that he was hardly ever even at a dinner party. Debt --- he has never owed anyone a thing. Passions of the heart --- what passions can there be in a man who, as the prosecutor himself has charged, has lived all his life on the land, working a farm? A life as far removed from passion, and as bound up with duty, as any can be.

\ciceroSection{40} What was it, then, that turned this enormous madness against Sextus Roscius? "He did not please his father," he says. He did not please his father? Why? It must have been a serious and obvious cause. For just as this is incredible --- that a son should have brought death upon a father without very many and very weighty reasons --- so this too is unbelievable: that a son should have been hated by a parent without many and weighty and compelling causes.

\ciceroSection{41} Let us return, then, to the same point and ask: what such great failings were there in this only son that he displeased his father? But it is plain there were none. Is the father then to be supposed mad, who hated without cause the son he had begotten? But the father was the most level-headed of men. So this much is now plainly clear: if the father was not mad, and the son was not depraved, then there was neither a cause for hatred on the father's side, nor a cause for crime on the son's.

\ciceroSection{42} "I do not know," he says, "what cause there was for the hatred --- but that there was hatred I gather from this: that the father, while he had two sons, used to keep the one who has since died at his side at all times, and had relegated this one to country estates." What was happening to Erucius in his bad and silly prosecution is happening also to me in my excellent case: he could not find a way to make his fictitious charge stand up; I cannot find a way to refute and dilute objections so flimsy.

\ciceroSection{43} What are you saying, Erucius? So many estates, so beautiful, so productive, Sextus Roscius the elder had handed over to his son to manage and look after as a kind of relegation and punishment? What? Heads of households who have children --- especially men of that rank, from country municipalities --- do they not consider it the most desirable thing for their sons to be most actively engaged in the family business, and to spend all their effort and care on the management of the estates?

\ciceroSection{44} Or had he sent him off in such a way that he was to be on the land merely to be fed at the farmhouse, deprived of every comfort? What if it is established that not only did he have charge of the estates, but that he was accustomed, in his father's lifetime, to enjoy the profits of certain particular farms? Will you still call that life of his in the country a relegation, an exile? You see, Erucius, how far your argument is from the substance and the truth. What fathers do as a matter of custom, you reproach as if it were something new. What is done out of affection, you allege was done out of hatred. What the father, as a mark of honour, granted his son, you say he did as a punishment.

\ciceroSection{45} You are not unaware of all this. But you are so short of substantive accusations that you think you must speak not only against us, but against the nature of things, against the customary practice of mankind, against everyone's settled opinions. "But, with two sons, he kept one always with him, and let the other live in the country." Erucius, please --- I beg you to take this in good part; I will not say it to wound you, but to remind you.

\ciceroSection{46} If fortune did not give you a known father from whom you could learn what a father's feeling for his children is, nature certainly gave you no small share of human feeling; and to that has been added a love of letters, so that you are not a stranger even to literature. Take it from the comic stage. Does the old man in Caecilius's play seem to you to value Eutychus, his country son, less than that other one --- Chaerestratus, I think the name is --- whom he keeps at his side in town as an honour, while he relegates the other to the country as a punishment?

\ciceroSection{47} "Why are you wandering off into these absurdities?" you will ask. As if it were difficult for me to bring forward as many examples as you like by name --- and not to wander too far afield, my own fellow tribesmen and neighbours, who wish to have their children, of whom they think the world, made into hard-working farmers. But it is tiresome to call up real names, partly because it is doubtful whether the people in question would be glad to be named, partly because no one will be more familiar to you than this Eutychus, and partly because in fact it makes no difference whether I name a young man from a comedy or someone from the territory of Veii. Indeed I think these things were composed by the poets so that we might see our own ways set out in others' persons --- a faithful image of daily life.

\ciceroSection{48} Now turn your mind to the truth of the matter. Consider --- not just in Umbria and the country round it, but in those old country municipalities of Italy --- what kinds of pursuits in their sons heads of households especially praise. You will see at once that, for want of charges, you have made the highest praise of Sextus Roscius into a fault and a reproach. And this is not only what their sons do at their fathers' wish: I myself know many --- and unless I am mistaken, every one of you knows men --- who, themselves on fire with enthusiasm for everything connected with the farm, regard this country life --- which you take as something necessarily disgraceful and criminal --- as the most honourable and the most agreeable kind of life there is.

\ciceroSection{49} What do you suppose is the energy and intelligence with which this very Sextus Roscius pursues country matters? From his neighbours --- men of the highest reputation --- whom I have heard speak of him, you, Erucius, are not more skilled in your prosecuting trade than he is in his. But I suppose, as it pleases Chrysogonus, who has not left him a single farm, that he will have to forget his trade and lay aside his pursuits. Wretched and unworthy as that is, he will bear it calmly enough, gentlemen, if through you he can keep his life and his good name. But this is what cannot be borne: that the very excellence and abundance of his estates, and the eagerness with which he cultivated them, should be the chief grounds of his ruin --- as if it were not misery enough that he cultivated them for others, not himself, but he must also be made a defendant for having cultivated them at all.

\ciceroSection{50} You really would have made an absurd prosecutor, Erucius, had you been born in those days when men were called from the plough to be made consul. For if you suppose it a disgrace to be in charge of cultivating a farm, you would surely have judged that Atilius, whom they found scattering seed with his own hand, the most disgraceful and dishonourable of men. But by Hercules our ancestors thought far otherwise of him and of others like him; and so it is that, from the smallest and slenderest of states, they left us the greatest and most flourishing. They cultivated their own fields with care; they did not greedily reach after the fields of other men. By so doing --- in fields and in cities and in nations --- they enlarged the state, and our empire, and the very name of the Roman people.

\ciceroSection{51} I do not bring forward these matters because they are to be set side by side with the case at hand, but to make this point understood: that since among our ancestors the leading and most distinguished men --- men who at every moment ought properly to be at the helm of state --- nevertheless spent some part of their effort and their time on the cultivation of their own land, it is right to forgive a man who admits that he is a countryman, who has lived steadily in the country, especially when there was nothing he could possibly have done that was either more pleasing to his father, or more agreeable to himself, or in real truth more honourable.

\ciceroSection{52} So the bitter hatred of the father against the son is shown, I take it, Erucius, by this single fact: that the father allowed him to live in the country. Is there anything else? "Yes, indeed," he says, "there is. For he had it in mind to disinherit him." I am listening; now you say something to the point. For those other things, I take it, even you grant are flimsy and absurd: "He did not attend dinner parties with his father." Naturally, since he hardly even came to town except very rarely. "Almost no one invited him to his house." No wonder, since he neither lived in the city nor was likely to return the invitation.

\ciceroSection{53} But you yourself understand that all this is trifling. Let us look at what we set out to look at, the surer ground from which an inference of hatred could possibly be drawn. "The father was thinking of disinheriting his son." I will not press to ask why; I ask only how you know. Though it was your duty to say it and to enumerate every reason. It was the office of a serious prosecutor, who was bringing a charge of so monstrous a crime, to set out every fault and every offence of the son, by which the parent could have been so inflamed as to overcome nature itself, to drive that deeply implanted love out of his mind, and finally to forget that he was a father at all. Without grave failings on the son's part, I do not believe such a thing could happen.

\ciceroSection{54} But I will allow you to leave aside what, by your silence, you concede was nothing at all. You must at least make this much plain: that he wished to disinherit. What evidence do you bring forward, that we should suppose this happened? You can produce nothing real; invent something at least, with some skill, so as not to look exactly as you do look --- making sport of this wretched man's misfortune and of the dignity of these distinguished gentlemen. He wished to disinherit his son. For what reason? "I don't know." Did he disinherit him? "No." Who stopped him? "He was thinking about it." Thinking? To whom did he say so? "To no one." What is this but to abuse, for profit and for caprice, the courts and the laws and the dignity of this body --- to bring a charge in this fashion, to assert what you cannot make plain and which you do not even attempt to make plain?

\ciceroSection{55} There is not one of us, Erucius, who does not know that you have no quarrel of your own with Sextus Roscius. Everyone sees what brings you forward as his enemy. Everyone knows that the prospect of his money has drawn you. What of it? You ought to have been moderate enough in your greed at least to think that the opinion of these gentlemen, and the lex Remmia (a law against malicious prosecution) ought to count for something.

\ciceroSection{56} It is useful to have many prosecutors in the state, so that audacity may be held in check by fear. But it is useful in such a way that we are not to be made absolute fools of by prosecutors. A man is innocent --- but, though he is free from guilt, he is nevertheless not free from suspicion: a sad case, but the man who prosecutes him I could in some way pardon, since, having something to which he can give a criminal and suspicious colour, he does not look like one knowingly playing a public game and suing for spite. So we readily put up with as many prosecutors as you like, because the innocent man, if accused, can still be acquitted, while the guilty cannot be condemned unless he is accused. It is more useful that the innocent be acquitted than that the guilty go untried. Geese are publicly fed at the cost of the state, and dogs are kept on the Capitol, that they may give warning if thieves come. They cannot tell thieves apart, but they sound the alarm if anyone has come up onto the Capitol at night; and, since this is suspicious, although they are mere brutes, they err at any rate on the side of caution. But if the dogs barked even by daylight, when worshippers came to greet the gods, I take it their legs would be broken --- because they are sharp at a moment when there is no suspicion at all.

\ciceroSection{57} The case with prosecutors is exactly the same. Some of you are geese, who only make a noise and cannot bite. Others are dogs, who can both bark and bite. We see provisions made for you. But you ought above all to attack those who deserve it. That is what most pleases the people. Then, even when there is some likelihood that someone has done something, you may bark on suspicion --- that too can be allowed. But if you proceed in this fashion, charging a man with the killing of his father, when you can say neither why nor how, and merely bark without any ground for suspicion at all --- nobody, I dare say, will break your legs; but, if I know these gentlemen, they will fix that letter to your forehead --- the letter you so detest that you hate even the Calends --- so firmly that ever afterwards you will be able to prosecute no one but your own ruined fortunes.

\ciceroSection{58} What have you given me to defend, my good prosecutor? And what have you given these gentlemen to suspect? "He was afraid of being disinherited." I am listening; but no one says why he ought to have been afraid. "His father had it in mind." Make it plain. There is nothing. Nothing about with whom he discussed it, whom he informed, by what route this suspicion has come into your minds. When you prosecute in this manner, Erucius, are you not openly saying: "I know what fee I have received; I do not know what to say. The one thing I had in view was what Chrysogonus said --- that this man would have no advocate, that no one in the present time would dare to breathe a word about the sale of the property and the partnership"? This false notion is what pushed you into this dishonesty. By Hercules, you would not have said a word, had you thought anyone would answer you.

\ciceroSection{59} It would have been worth your while, gentlemen, if you had observed it, to consider his negligence in conducting the prosecution. I imagine that, when he saw what kind of men were sitting on these benches, he asked whether this man or that one was likely to undertake the defence; about me he had not even a suspicion, since I had not before pleaded any public case. After he had ascertained that none of those who customarily and capably plead such cases would do so, he began to be so careless that, whenever it occurred to him, he would sit down, then he would walk about, sometimes he would even call a slave-boy --- to give orders for his dinner, I take it --- treating your assembly and this gathering as if it were the merest emptiness. He did at last finish his speech and sit down. Then I rose.

\ciceroSection{60} He seemed to breathe more easily that no one of greater standing was speaking. I began to speak. I noticed, gentlemen, that as long as I had not yet named Chrysogonus, he kept up his joking and his other distractions; but the moment I touched the name, the fellow immediately drew himself up, and seemed astonished. I knew what had stung him. I named Chrysogonus a second time and a third. After that the men did not stop running back and forth, conveying news, I suppose, to Chrysogonus that there was someone in the city bold enough to speak against his wishes, that the case was being conducted otherwise than he had supposed, that the sale of the property was being opened up, that the partnership was being dragged through the mire, that his influence and his power were being slighted, that the jurors were paying close attention, that the Roman people thought the affair scandalous.

\ciceroSection{61} Since these calculations have failed you, Erucius, since you see that everything has been turned upside down, that the case for Sextus Roscius is being argued, if not gracefully, at least freely, that the man you thought delivered up is being defended, that those whom you hoped would betray him you see giving judgement on him --- restore to us, at long last, that old shrewdness and prudence of yours; admit that you came in here in the hope that this would be brigandage, not a trial. The case is one of parricide; no reason has been given by the prosecutor why a son should have killed his father.

\ciceroSection{62} What in the smallest crimes and in those lighter offences which are now more frequent and almost daily is asked first and above all --- "what was the cause of the wrongdoing?" --- Erucius does not think need be asked in a case of parricide. In which crime, gentlemen, even when many causes seem to have come together in one place and to fit in with one another, even then the matter is not lightly believed; nor is the case settled on slight conjecture; nor is an uncertain witness listened to; nor is the matter decided by the prosecutor's cleverness. Many previous offences must be shown, a thoroughly ruined life, a singular audacity --- and not audacity alone, but the highest fury and madness. And though all these be present, still the marks of the crime ought to stand out plainly: where, by what means, by whose hand, at what time the deed was done. Unless these are many and obvious, then a thing so wicked, so atrocious, so unspeakable, simply cannot be believed.

\ciceroSection{63} For the force of human feeling is great. The bond of common blood weighs heavily. Nature herself cries out against suspicions of this sort. It is the surest of portents and prodigies that there should be anyone in human shape and form so far surpassing the beasts in savagery as to deprive most outrageously of the light those through whom he himself looked upon this most pleasant light --- when even wild beasts are bound to one another by birth and rearing and by nature itself.

\ciceroSection{64} Not many years ago, they say, a certain Titus Caelius of Terracina, a man of some standing, having dined and gone to bed in the same room with his two grown sons, was found in the morning with his throat cut. When neither slave nor free man was found on whom suspicion could fall, and when the two sons --- already of an age to be aware of things --- said that, sleeping near him, they had not noticed anything, charges of parricide were laid against the sons. What could be more suspicious? That neither one of them had heard a thing? That anyone should have ventured into that room precisely at the moment when the two grown sons were lying there, who could easily both notice him and defend their father? Furthermore, there was no one against whom the suspicion could fit.

\ciceroSection{65} And yet, when it was made plain to the jurors that they were found asleep with the door open, the young men were acquitted by the verdict and freed of all suspicion. For no one believed that there could be anyone who, having defiled by an unspeakable crime every law human and divine, could immediately afterwards drop into sleep --- because those who have committed so great an offence not only cannot rest without unease, they cannot even draw breath without dread.

\ciceroSection{66} Do you not see how those whom the poets have handed down to us as having taken vengeance on a mother for their father's sake --- though they are said to have done it at the bidding and oracles of the immortal gods themselves --- nevertheless are driven by the Furies and never allowed to rest, because not even the dutiful could perform such a deed without crime? So it stands, gentlemen: a parent's blood, a father's or a mother's, has great force, great urgency, great holiness; and once a stain has been incurred from it, it not only cannot be washed off, but it sinks so deep into the soul that the highest madness and frenzy follow.

\ciceroSection{67} Do not suppose, as you so often see in stories, that those who have done some impious and wicked deed are driven and terrified by the Furies' burning torches. Each man is most tormented by his own deceit and his own dread. Each man is harried and unhinged by his own crime; his own evil thoughts and his own conscience are what frighten him. These are the relentless Furies of the wicked, dwelling in the home, who day and night exact the parents' penalty from sons stained with crime.

\ciceroSection{68} The very magnitude of the offence makes the matter such that, unless the parricide is shown almost manifestly, it is not credible: unless there is a base youth, unless there is a life polluted with every disgrace, unless there are extravagances and shameful and dishonourable expenses, unless there is open audacity, unless such recklessness as is not far from insanity. Add to this hatred of the parent, fear of paternal correction, wicked friends, slaves who are accomplices, a fitting time, a place chosen for the deed. I will almost say: the jurors must see hands actually spattered with a father's blood, if so great a crime, so monstrous, so cruel, is to be believed.

\ciceroSection{69} Wherefore, the less credible the thing is unless it is shown, the more it must be punished if it is convicted. So just as our ancestors are seen to have surpassed other nations not only by force of arms but by judgement and wisdom in many things, so most especially in this: that for impious offenders they devised a singular punishment. Consider in this matter how far they outran in prudence those who are reckoned the wisest among other peoples.

\ciceroSection{70} The most prudent state is said to have been Athens while she held the supremacy; and of that state Solon is said to have been the wisest man --- the same Solon who wrote the laws which they still use today. When he was asked why he had set down no penalty for one who had killed his parent, he replied that he had not supposed anyone would do such a thing. He is said to have done wisely in not having decreed anything against an act not before committed, lest he seem rather to have suggested it than forbidden it. How much more wisely our ancestors did! For when they realised that there is nothing so holy that audacity will not at some time profane it, they devised a singular punishment for parricides, so that those whom nature herself could not hold to their duty might be deterred from the offence by the magnitude of the penalty. They ordered them to be sewn up alive in a sack and so cast into the river.

\ciceroSection{71} What singular wisdom, gentlemen! Do they not seem to have plucked such a man up out of the natural order, suddenly stripping him of sky, sun, water, and earth --- so that he who had killed the man from whom he himself was born should be deprived of all those very things from which all things are said to be born? They were unwilling to throw the body to wild beasts, lest by their having touched so great a crime we should ourselves use beasts grown more savage. They were unwilling simply to fling them naked into the river, lest, when they were carried out into the sea, they should pollute the very element by which all other things that have been violated are thought to be cleansed. There is, in short, nothing so cheap or so common of which they have left them any share whatsoever.

\ciceroSection{72} For what is so common as breath to the living, earth to the dead, the sea to the tossed, the shore to the cast-up? They live, while they can, in such a way that they cannot draw breath from the open sky; they die in such a way that the earth does not touch their bones; they are tossed by the waves in such a way that they are never washed clean; finally they are cast up in such a way that they do not even rest, dead, on the rocks. Of an offence so vast, for which a punishment so distinctive has been instituted, do you suppose, Erucius, that you can persuade men of this sort, when you have not even produced a cause for the offence? If you were prosecuting this man before the very buyers of his property, with Chrysogonus presiding over the court, you would still have come better prepared.

\ciceroSection{73} Do you not see what is in question, or before whom it is being heard? It is a matter of parricide, which cannot be undertaken without many causes; and it is being heard before men of the greatest prudence, who understand that no one commits even the smallest crime without a cause. Very well: you cannot bring forward a cause. Although I ought to have won outright at this point, I will yield some of my right and grant you, in this case, what in any other I would not concede --- relying on my client's innocence. I am not asking you why Sextus Roscius killed his father; I am asking how he killed him. I ask you that, Gaius Erucius, in such a way that I will give you the chance to answer, or to interrupt, or even to put questions, if you will, in your turn.

\ciceroSection{74} How did he kill him? Did he strike him with his own hand, or did he hand him over to others to be killed? If you charge him with the act itself, he was not at Rome. If you say it was done through others, I ask: which others? Slaves or freemen? If freemen, what sort of men? Men from the same place, Ameria? Or these cut-throats from the city? If from Ameria, who are they? Why are they not named? If from Rome, where had Roscius come to know them, who had not been in Rome for many years and was never there for more than three days? Where did he meet them? In what manner did he speak with them? How did he persuade them? "He paid them money." To whom did he pay it? Through whom did he pay it? Where did he get it? How much did he give? Is it not by such tracks that one usually arrives at the source of a crime? And remember at the same time how you yourself painted his life: that this was a wild and rustic man, who had never had a conversation with anyone, who had never set foot in any town.

\ciceroSection{75} I pass over here a point that could have been a very strong argument for his innocence: that in country habits, in a hard fare, in this rough and uncultivated kind of life, crimes of this kind do not generally arise. Just as you cannot find every fruit and every tree in every soil, so not every crime is born from every life. In the city, luxury is bred. From luxury greed must inevitably break out. From greed audacity bursts out, and from there spring all the kinds of crime and offence. But this country life, which you call rustic, is the school of frugality, of diligence, of justice.

\ciceroSection{76} But I let this go. I ask only this: a man who, on your own showing, was never among men, through which men could he have done so great a deed --- so secret, especially from a distance? Many things are false, gentlemen, which can nonetheless be put into a suspicious light. In these matters, if even suspicion is found, I will concede that the guilt is there. Sextus Roscius was killed at Rome while his son was on the estates at Ameria. I suppose he sent letters to some assassin who knew no one in Rome. He summoned someone. Whom or when? He sent a messenger. Whom, or to whom? With money, with favour, with hopes, with promises he induced someone. Of all this, not a thing can even be invented; and yet the case is being argued as parricide.

\ciceroSection{77} It remains for him to have done it through slaves. O immortal gods --- a wretched and ruinous business! What? In so grave a charge, the thing which is usually a help to an innocent man --- to offer up his slaves for examination under torture --- this is something that Sextus Roscius is not allowed to do? You who prosecute him have all his slaves. Of so great a household, not one slave-boy, the manager of his daily provisions, has been left to Sextus Roscius. I appeal to you now, Publius Scipio; to you, Marcus Metellus; with you as advocates, with you as counsel, Sextus Roscius has several times demanded of the opposing party two of his father's slaves for examination. Do you remember Titus Roscius's refusal? What of those slaves? Where are they? They follow Chrysogonus, gentlemen; they enjoy honour and are highly prized by him. Even now I demand that they be examined; my client begs and beseeches it.

\ciceroSection{78} What are you doing? Why do you refuse? Doubt now if you can, gentlemen, by whom Sextus Roscius was killed: by the man who, because of that death, lives in poverty and in danger, who is not even allowed to make inquiry into his father's death --- or by those who flee inquiry, who have his property, who live in murder and from murder. Everything in this case, gentlemen, is wretched and unworthy; but nothing more bitter, nothing more unjust, can be brought forward than this: that a son is not allowed to hold an examination concerning the death of his father by means of the slaves of his father! Will he not even for that long be the master over his own slaves --- long enough to have them questioned about his father's death? I shall come, and not long hence, to this point; for the whole of it concerns the Roscii, of whose audacity I promised to speak when I had refuted the charges of Erucius.

\ciceroSection{79} Now, Erucius, I come to you. We must agree, you and I --- if this offence touches my client at all --- that he either committed it with his own hand (which you deny) or did it through some freemen or slaves. Freemen? You can show neither how he could have met them, nor by what argument he could have induced them, nor where, nor through whom, nor with what hope, nor at what price. I, on the contrary, show that Sextus Roscius not only did none of these things, but could not even have done them, because he was not in Rome for many years, and never went away from the estates without good reason. The only resource left to you appeared to be the title of slaves, into which, like a harbour, you could take refuge from the rest of the suspicions; but here you strike a rock of such a kind that you not only see the charge bouncing off my client, but realise that all the suspicion falls back upon yourselves.

\ciceroSection{80} What is it, then, to which the prosecutor still takes refuge in his lack of arguments? "The times were such," he says, "that men were murdered indiscriminately and with impunity. So you, because of the multitude of cut-throats, could have done this without any trouble." Sometimes you seem to me, Erucius, to want to gain two ends with one fee --- to shower us with the trial, and to accuse the very men from whom you took your fee. What are you saying? That men were murdered indiscriminately? By whom and at whose hands? Do you not realise that you have been brought here by the dealers in confiscated goods? What then? Do we not know that in those times the same men were dealers in throats and in property?

\ciceroSection{81} Will those men, finally, who at that time ran armed about, day and night, who were constantly at Rome, who at every moment were involved in plunder and bloodshed, throw at Sextus Roscius the bitterness and unfairness of those times, and reckon that the multitude of cut-throats, of which they themselves were the leaders and chiefs, will tell against him? --- a man who was not only not in Rome, but knew nothing of what was being done at Rome, since he was steadily, as you yourself admit, in the country.

\ciceroSection{82} I am afraid, gentlemen, of being tedious to you, or of seeming to mistrust your judgement, if I dwell longer on matters so plain. The whole of Erucius's accusation, I take it, has been undone --- unless you are perhaps expecting me to dissolve those things he raised about embezzlement and other made-up accusations, novel and unheard of by us before this moment, which he seemed to me to be reciting from some other speech composed against some other defendant; so little did they relate either to the charge of parricide or to the man whose case is being heard. Since he has charged them only with a word, it is enough to deny them with a word. If there is anything he is keeping in reserve for the witnesses, there too he will find us better prepared, as in the case proper, than he supposed.

\ciceroSection{83} I come now to where it is not eagerness that draws me, but loyalty. For if I wished to prosecute, I would prosecute others, from whom I might rise. But this I am resolved not to do, as long as I have the choice. The man who seems to me most distinguished is the one who reaches a higher place by his own virtue, not the one who climbs by means of another's misfortune and ruin. Let us at last leave off scrutinising what is empty. Let us look for the offence where it is and where it can be found. You will see at once, Erucius, by how many suspicions the certain crime is convicted, even though I shall not say everything and shall touch each thing only lightly. I would not even do this, were it not necessary; and the proof that I am doing it unwillingly will be that I will not pursue the matter further than my client's safety and my own good faith require.

\ciceroSection{84} You found no cause in Sextus Roscius. I find one in Titus Roscius. For the matter is between you and me, Titus Roscius Magnus, since you sit there and openly profess yourself an opponent. Of Capito we shall see later, if --- as I hear --- he comes forward as a witness; then he will become acquainted with other "victories" of his of which he does not even suspect that I have heard. The famous Lucius Cassius --- whom the Roman people thought a most truthful and most wise judge --- used repeatedly to ask in cases: "to whose advantage was it?" For human life is such that no one tries to commit a crime without hope and gain.

\ciceroSection{85} Those who were brought into danger fled and trembled at this judge --- even though he was the friend of truth, his nature seemed not so much inclined to mercy as bent towards severity. I, although the man who presides over this present court is most resolute against audacity and most merciful to innocence, would yet readily plead for Sextus Roscius even if that very harshest of judges were holding the inquiry, or before Cassian jurors, whose name those who must plead before them dread to this day.

\ciceroSection{86} For in this case, when they saw that the others held the very greatest sums of money, while my client was in the depths of want, they would not even ask "to whose advantage was it?" They would attach the charge and the suspicion to the gain rather than to the destitution. What if it is added that you were thin in resources before? What if greedy? What if audacious? What if a particular enemy of the man who was killed? Need we still ask what cause has driven you to so great a crime? Of these things, what can be denied? Your scant means are such that they cannot be hidden; the more you hide them, the more they shine out.

\ciceroSection{87} You display your greed, when you have entered into a partnership with utter strangers over the property of a fellow townsman and a kinsman. How great your audacity is, leaving aside everything else, all could see from this --- that of the whole partnership, that is, of so many cut-throats, you alone were found to sit on the prosecutor's bench, not only showing your face but actually thrusting it forward. That you had quarrels with Sextus Roscius, and grave disputes about family property, you must necessarily concede.

\ciceroSection{88} It remains, gentlemen, only this for us to be in doubt about: which of the two killed Sextus Roscius --- the one to whom riches came through his death, or the one to whom destitution came; the one who was thin in means before, or the one who was reduced afterwards to the most absolute want; the one who, burning with greed, is borne hostile against his own kin, or the one who has always lived in such a way that he knew no kind of gain except the fruit of his own labour; the one who is the boldest of all the partners in confiscations, or the one who, for unfamiliarity with the Forum and the courts, dreads not only the benches but the city itself; finally, gentlemen --- and this is to my mind the chief point of the matter --- whether an enemy or a son.

\ciceroSection{89} If you, Erucius, had hit upon so many and such weighty things in your accused, how long would you not be speaking! How you would strut! By Hercules, time would fail you sooner than your eloquence. For in any one of these matters there is enough material to take up a whole day. Nor am I incapable of doing the same; for I do not so far depreciate myself, even though I make no claim, as to think that you can speak more copiously than I. But I, perhaps, because of the multitude of advocates, may be reckoned simply one of the herd; you the carnage of Cannae has made a sufficiently good prosecutor. We have seen many slain --- not at Lake Trasimene, but at Lake Servilius.

\ciceroSection{90} Who there was not wounded by Phrygian steel? It is unnecessary to call up all the Curtii, Marii, Memmii, whom age was already calling away from battle; finally old Priam himself, Antistius, whom not only age but also the laws forbade to fight. Then those whom no one names because of their obscurity --- there are six hundred of them prosecuting in cases of murder and poisoning. So far as concerns me, I should be glad if every one of them lived. For there is no harm in there being as many dogs as possible where there are very many things to watch over and very much to guard.

\ciceroSection{91} But, as happens, the violence of war and its tumult often work many things by reckless leaders. While the man who held supreme command was occupied with other matters, there were in the meantime those who treated their own wounds; men who, as if a perpetual night had fallen upon the commonwealth, rushed in the dark and threw everything into confusion. From them I am surprised that, just as no trace of the courts was left, even the benches were not burnt; for they did away with prosecutors and jurors alike. There is at least this consolation: they so lived that they could not, if they wished, kill all the witnesses. As long as the human race lasts, there will not be wanting men to bring charges; as long as the state lasts, courts will be held. But, as I began to say, both Erucius, if he had in this case the things I have mentioned, could go on at any length, and so could I, gentlemen; but my plan is, as I said before, to pass lightly over the points and merely to graze each in turn, that everyone may understand that I am prosecuting from no zeal, but defending from duty.

\ciceroSection{92} I see, then, that there are very many causes that might have driven this man on. Let us see now whether he had any opportunity of undertaking the offence. Where was Sextus Roscius killed? --- "At Rome." --- And you, Titus Roscius Magnus: where were you then? --- "At Rome. But what does that prove? Many others, too." --- As if the question now were "which of so great a multitude killed him," and not rather this: of the man who was killed at Rome, is it more probable that he was killed by one who was constantly at Rome at that time, or by one who had not come anywhere near Rome for many years?

\ciceroSection{93} Now consider also the other opportunities. There was at that time, as Erucius mentioned, a multitude of cut-throats, and men were killed with impunity. What sort of multitude? Either, I take it, of those engaged in seizing property, or of those hired by them to kill someone. If you mean those who were after another man's goods, you yourself are in their number, you who are rich at our expense. If you mean those whom under a softer name they call "strikers," ask yourself whose protection and patronage they are under. Believe me, you will find someone from your own partnership; and whatever you say in reply, weigh it against our defence. So you will most easily see Sextus Roscius's case set side by side with yours.

\ciceroSection{94} You will say: "What of it, if I was constantly at Rome?" I will reply: "But I was not there at all." --- "I confess that I am a confiscation-dealer; but so are many others." --- "But I, on your own showing, am a farmer and a countryman." --- "Just because I have entered the gang of assassins, I am not, on that account, an assassin." --- "But I, who do not even know any assassin, am most certainly far from any such charge." There are very many things that could be said to show that you had every facility for undertaking the deed; which I leave aside, partly because I do not gladly accuse you yourself, and even more because, if I wished to relate the killings done in those times in just the manner in which Sextus Roscius was killed, I am afraid my speech would be seen to bear on more men than this one.

\ciceroSection{95} Let us look now briefly, as at the other points, at what was done by you, Titus Roscius Magnus, after the death of Sextus Roscius; things so open and manifest that, by my faith, gentlemen, I say them unwillingly. For I am afraid, whatever sort of man you are, Titus Roscius, that I shall seem to have been so anxious to save my client that I have not spared you at all. Even while I am in this fear and wish to spare you in some respect that good faith permits, again I change my mind --- because your face comes back to me. You, when the rest of your associates were fleeing and hiding themselves, so that this trial should appear to concern not their plunder but my client's offence, demanded, of all people, this part: to be in the trial yourself and to sit with the prosecutor? In which you accomplish nothing else than to make your audacity and your shamelessness known to all mankind.

\ciceroSection{96} When Sextus Roscius was killed, who was the first to bring the news to Ameria? Mallius Glaucia, whom I named before, your client and intimate. What was the point of his bringing the news especially --- since, if you had no plan formed beforehand about the man's death and his property, and had entered into no partnership of crime or of reward with anyone, this least of all concerned you? --- "Mallius is bringing the news of his own accord." What, I ask you, was it to him? Did it happen by chance that, having come to Ameria for some other reason, he should be the first to bring news of what he had heard at Rome? For what reason had he come to Ameria? "I cannot guess," he says. I shall bring matters to such a point that no guessing will be necessary. Why did he bring the news first to Titus Roscius Capito? When at Ameria there was Sextus Roscius's house, his wife, and his children, when there were so many close kin and relatives on the best of terms with him --- by what reasoning did it come about that this client of yours, the messenger of your crime, should bring the news first of all to Titus Roscius Capito?

\ciceroSection{97} He was killed on his way home from supper. It was not yet light when the news was known at Ameria. What does this incredible speed signify, this enormous haste and hurrying? I do not ask who struck the blow. There is no need, Glaucia, for you to be afraid; I will not search you to see whether you happened to have a weapon; I do not strip you. I think it does not concern me; since I find by whose plan he was killed, by whose hand he was struck I do not trouble myself. Just this much I take, which your manifest crime opens to me: where, or from whom, did Glaucia hear it? How did he know so quickly? Granted that he heard immediately --- what was it that compelled him in a single night to cover so great a journey? What necessity pressed him so hard that, if he were going to Ameria of his own accord, he should set off from Rome at that hour and rest no part of the night?

\ciceroSection{98} Even in matters so transparent must we look for argumentation or take up conjecture? Do you not, gentlemen, seem to see with your own eyes the things you have heard? Do you not see that wretched man, ignorant of his own danger, returning from supper? The ambush set, the sudden onslaught? Does not Glaucia hover before your eyes in the murder? Is not this Titus Roscius present? Does he not with his own hands set in his chariot that Automedon --- the messenger of his most cruel crime and his unspeakable victory? Does he not beg him to keep the night through, to take the trouble for his sake, to bring the news to Capito at the earliest possible moment?

\ciceroSection{99} What was it that he wished Capito to know first? I do not know --- except that I see this: that Capito is a partner in this property; that of thirteen farms I see him hold three of the noblest.

\ciceroSection{100} I hear, moreover, that this is not the first time such suspicion has been levelled at Capito. He has many infamous "victories" to his credit; this one, however, is the first that has come to him from Rome with the ribbon of distinction. There is no method of killing a man by which he has not killed several --- many with steel, many with poison. I have besides to say that he threw down off the bridge into the Tiber, against the custom of our ancestors, a man under sixty years of age. If he comes forward --- and he will come forward, for I know he will --- he will hear of these things. Let him only come; let him unfold that volume of his ---

\ciceroSection{101} which volume, I can prove, Erucius drafted; which they say he held over Sextus Roscius, and threatened that he would offer the whole of it as testimony. What an excellent witness, gentlemen! What gravity, equal to the expectation! What an honourable life --- such that you will gladly fit your sworn oath to his testimony! Truly we should not see these men's misdeeds with such clarity, did not greed and avarice and audacity blind them themselves.

\ciceroSection{102} The one sent a swift courier from the very place of the murder to Ameria, to his partner --- or rather to his master --- in such a way that, even if everyone wished to dissemble that they knew to whom the offence belonged, he himself laid his manifest crime out before the eyes of all. The other --- if it pleases the immortal gods --- is even going to give evidence against Sextus Roscius. As though the question now were whether what he says is to be believed, and not what he has done is to be punished. So it has been laid down by ancient custom that, even in the smallest matters, men of the highest rank do not give evidence in their own cause.

\ciceroSection{103} Africanus --- whose very surname declares that he subdued a third part of the world --- nevertheless would not have given evidence, had it been a matter touching himself. For of so great a man I do not dare to say this: that, if he gave evidence, he would not be believed. See now how everything has been turned upside down and changed for the worse. When the matter is one of property and of murder, a man is going to give evidence who is at once a confiscation-dealer and a cut-throat --- that is, who is the buyer and possessor of the very property at issue, and who arranged the killing of the man whose death is being investigated.

\ciceroSection{104} What of you, my excellent fellow? Have you anything to say? Listen to me: see that you do not fail yourself; your own great affair is also at stake. You have done many things wickedly, audaciously, dishonourably, perfidiously --- one thing most foolishly, surely on your own and not on Erucius's advice: there was no need for you to be sitting there. No one uses an accuser dumb, or a witness who rises from the prosecutor's bench. Add to this that your greed of yours might at least have been a little hidden, a little under cover. As it is, what is there that any of you should wish to hear, when what you do is of such a kind that you seem to be doing it on purpose to make my own task against you easy? Now, gentlemen, let us look at what came next.

\ciceroSection{105} At Volaterrae, in the camp of Lucius Sulla, the death of Sextus Roscius was reported to Chrysogonus on the fourth day after he was killed. Even now, must it be asked who sent the courier? Is it not plain that it was the same man who sent the messenger to Ameria? Chrysogonus arranges for the man's property to be sold at once --- a man whom he did not know, and a property he did not know. But how did it occur to him to covet the estates of an unknown man, whom he had certainly never seen? When you hear something of this kind, gentlemen, you usually say at once: "Some fellow townsman or neighbour must have told him; they generally inform; through them most things are betrayed." Here you have nothing to take up by way of conjecture.

\ciceroSection{106} For I shall not argue thus: "It is likely that the Roscii brought the matter to Chrysogonus; for they had a friendship with him beforehand. Although the Roscii had many old patrons and family friends inherited from their ancestors, they ceased to court and respect them all, and entrusted themselves to the protection and clientele of Chrysogonus."

\ciceroSection{107} I could say all these things truthfully; but in this case there is no need of conjecture. I know for certain that they themselves do not deny that Chrysogonus came to this property at their suggestion. If you will see with your own eyes the man who has received a share for the act of informing, can you doubt, gentlemen, who informed? Who, then, are the people in this property to whom Chrysogonus has given a share? Two Roscii. Anyone else besides? No one, gentlemen. Can there then be any doubt that the men who have taken from him a share of the spoils were the men who brought him the spoils?

\ciceroSection{108} Now consider the action of the Roscii in the light of Chrysogonus's own judgement. If the Roscii had done nothing in this affair worth a price, why were they being rewarded by Chrysogonus with such great gifts? If they had done nothing but bring the matter to him, was it not enough to thank them, or, to be very generous, to bestow some honour upon them? Why are three estates, of so much money, given immediately to Capito? Why does this Titus Roscius hold all the rest jointly with Chrysogonus? Is it not plain, gentlemen, that Chrysogonus, with full knowledge of the case, has handed these spoils to the Roscii?

\ciceroSection{109} Capito came among the ten chief delegates as an ambassador to the camp. Learn the man's whole life, character, and habits from this very embassy alone. Unless you understand, gentlemen, that there is no duty, no right so sacred and unblemished that his crime and treachery would not have violated and impaired it, then judge him a most excellent man.

\ciceroSection{110} He stops Sulla from being informed about these matters. He betrays to Chrysogonus the plans and wishes of the rest of the deputation. He warns him to take care that the affair is not openly conducted; he points out that, if the sale of the goods were undone, he would lose a great sum of money, and Capito himself would be in danger of his life. He sharpens Chrysogonus, deceives those who had been sent with him, repeatedly warns the one to be on guard, treacherously holds out a false hope to the others; with the one he forms plans against these others, and reveals their plans to him; with him he bargains for his own share; with the others, relying on some pretext of delay, he constantly blocks every approach to Sulla. Finally, with this man as instigator, author, mediator, the deputation never went to Sulla; they were duped by his trustworthiness --- or rather by his treachery (which you will hear from them, if the prosecutor is willing to call them as witnesses), and so brought home, instead of a settled matter, a false hope.

\ciceroSection{111} In private affairs, if anyone discharged a commission entrusted to him not only with malice for his own gain or convenience, but even just with negligence, our ancestors thought he had committed the most disgraceful kind of disgrace. So a court of mandate was established, no less shameful than that of theft, because, in matters in which we cannot ourselves be present, the trust of friends takes the place of our own action. He who damages that trust attacks the common protection of all, and so far as in him lies, breaks up the partnership of life. We cannot do everything by ourselves; one man is more useful in one matter, another in another. For this reason friendships are entered into, that common advantage may be steered by mutual offices.

\ciceroSection{112} Why do you accept a commission, if you intend either to neglect it or turn it to your own benefit? Why do you offer yourself to me, and, with feigned dutifulness, get in the way of and obstruct my interests? Stand aside; I will deal through someone else. You take up the burden of duty, which you think you can sustain. It seems heaviest to those who are themselves least light-minded. So this fault is shameful for this reason: it violates two of the most sacred things, friendship and good faith. For hardly anyone gives a commission except to a friend, nor entrusts anything except to one whom he thinks faithful. So it is the act of the most utterly ruined man both to dissolve a friendship and to deceive a man who would not have been wronged unless he had trusted.

\ciceroSection{113} Is it not so? In the smallest matters, the man who neglects a commission must necessarily be condemned in the most disgraceful trial. In so great a matter, when the man to whom the reputation of the dead and the fortunes of the living were entrusted has stained the dead with infamy and reduced the living to want, will he be counted among honourable men --- or even among the living? In the smallest, private matters even mere negligence is brought to a charge of mandate and a verdict of infamy, because, if all is done as it ought to be, the man who gave the commission must be the one to neglect it, not he who accepted it. In so great a matter, which was conducted publicly and entrusted publicly --- a man who has not by negligence wronged some private interest, but by treachery has polluted and stained the very ceremony of the embassy itself --- with what penalty in the end will he be visited, in what court will he be condemned?

\ciceroSection{114} If Sextus Roscius had privately given him this charge, to settle and arrange matters with Chrysogonus, and, if necessary, to pledge his good faith to that end, and the man who had undertaken this had then turned even some little of the business to his own profit, would he not, condemned by an arbitrator, both have to restore the property and lose all his good name?

\ciceroSection{115} As it is, Sextus Roscius did not give him this charge. What is much graver: Sextus Roscius himself --- with his reputation, his life, all his goods --- was, by the decuriones, publicly entrusted to Titus Roscius. From this affair Titus Roscius did not turn a little into his own pocket: he overturned the man utterly out of his goods, bargained three estates for himself, and held the will of the decuriones and of all the townsmen as worth precisely as much as he held his own pledged faith.

\ciceroSection{116} See now, gentlemen, the rest, that you may understand that no kind of crime can be invented with which this man has not befouled himself. In smaller matters, to deceive a partner is most shameful, and equally shameful with what I have just spoken of, and not without justice; for he who shares a thing with another reckons that he has joined to himself a help. To whose good faith, then, will he flee for refuge, when he is wounded by the good faith of the man to whom he has entrusted himself? And those wrongs ought above all to be marked which can least be guarded against beforehand. We can be on our guard against strangers; intimates necessarily see many more open things; how can we be on our guard against a partner? Even if we fear him, we wound the very right of duty. Rightly, then, our ancestors thought that he who had cheated a partner ought not to be reckoned in the number of good men.

\ciceroSection{117} But this Titus Roscius cheated not a single partner in a money matter --- a serious matter, but somehow bearable --- but nine men of the highest standing, partners in the same charge, the same embassy, the same duty, the same commission. He led them on, deceived them, abandoned them, betrayed them to the opposite party, cheated them with every sort of fraud and treachery: men who could suspect nothing of his crime, who ought not to have feared a partner in their duty, who did not see his malice, who believed in his empty talk. So now those most honourable men, on account of his treachery, are thought to have been not cautious or foreseeing enough; while he, who at the start was a betrayer, then a deserter, who first betrayed his fellow ambassadors' counsels to the other party and then formed a partnership with that very other party, even now terrifies and threatens us, decked out in three estates --- the rewards of his crime. In a life like this, gentlemen, in so many and such great offences, you will find this offence too --- the one on trial.

\ciceroSection{118} For this is how you must inquire: where you see many things done greedily, audaciously, dishonourably, treacherously, then count that crime too lurks among such offences. Although this very crime is the least concealed; it stands out so plainly that it is not so much understood from the other crimes (which it is agreed are in him) as it is, by itself, used to prove the others, if any of them by chance is in doubt. What say you, gentlemen? Does that gladiator-trainer seem to you to have laid aside the sword altogether yet? Or this pupil to fall short of his master in the art by even a little? Their greed is the same, their dishonesty alike, their shamelessness identical, their audacity twin to each other.

\ciceroSection{119} Since you have learned the master's good faith, learn now the pupil's fairness. I have already said many times that two slaves were demanded by them for examination. You always refused, Titus Roscius. I ask you: were the men demanding it unworthy to obtain it? Or did the man for whom they demanded it not move you? Or did the matter itself seem to you unfair? They who demanded it were the noblest and most upright men of our state, whom I named earlier; men who have so lived, and are so reckoned by the Roman people, that whatever they say no one would not consider fair. They demanded it on behalf of the most wretched and unhappy of men, who himself would gladly hand himself over to torture, if only the matter of his father's death were investigated.

\ciceroSection{120} Furthermore, the request was such that it made no difference whether you refused it or confessed the offence. Since this is so, I ask why you refused. When Sextus Roscius was killed, those slaves were on the spot. Of the slaves themselves I neither bring an accusation nor a defence, so far as concerns me. That I see you fighting against having them given up for examination is suspicious. That they are with you in such great honour, plainly they must know something which it would be ruinous to you if they said. --- "It is unjust to question slaves against their masters." --- But there is no such question; for Sextus Roscius is the defendant, and when he is on trial, the question is not against masters; you say that you are the masters. --- "They are with Chrysogonus." --- I believe it. By their letters and refinement Chrysogonus is so smitten that, among his own boys --- chosen from so many elegant households for every delicacy and every art --- he should wish these to take their place; men barely fit for labour, from the rustic discipline of an Amerian country household.

\ciceroSection{121} It is certainly not so, gentlemen; it is not likely that Chrysogonus has fallen in love with their letters or culture, or has come to know their diligence and faithfulness in domestic affairs. There is something hidden, which the more eagerly it is suppressed and concealed by these men, the more it stands out and shows itself. What then?

\ciceroSection{122} Does Chrysogonus refuse the inquiry into them in order to conceal his own offence? By no means, gentlemen; I do not think everything fits everyone. So far as concerns me, I suspect nothing of that kind in Chrysogonus; nor is this the first time it occurs to me to say so. You remember that I divided the case at the beginning into the charge --- whose whole argument was committed to Erucius --- and the audacity, whose part was put upon the Roscii. Whatever there is of crime, of wickedness, of murder, that must be the property of the Roscii. The excessive influence and power of Chrysogonus we say obstructs us, and cannot in any way be borne, and ought, since the power has been given you, not only to be weakened but to be punished by you.

\ciceroSection{123} I judge thus: he who wishes inquiry to be made of those who, it is agreed, were present when the murder was done, that man wishes the truth to be found; he who refuses, surely, even though he does not dare to say it in words, by the very fact confesses to his own offence. I said at the start, gentlemen, that I was unwilling to say more about their crime than the case demanded and necessity itself compelled. For many things can be brought up, and any one of them can be expounded with many arguments. But what I do unwillingly and from necessity, I cannot do at length nor in detail. Those things which could not be passed over I have lightly touched, gentlemen; those which depend on inferences which, if I begin to speak of them, would have to be set out at length, I leave to your own judgement and conjecture.

\ciceroSection{124} I come now to that golden name of Chrysogonus, under which name the whole partnership has lain hidden; concerning whom, gentlemen, I am unable to find either how I shall speak or how I shall be silent. For if I am silent, I leave out the greatest part. If I speak, I am afraid that not he alone, which does not concern me, but many others may think themselves injured. Although the matter is so situated that, in my view, there is no great need for me to say anything bearing on the common cause of the confiscation-dealers, since this case is plainly new and singular. Chrysogonus is the buyer of Sextus Roscius's property.

\ciceroSection{125} Let us first look at this: in what way was that man's property put up for sale, or in what way could it have been put up? And I shall not ask this in the sense of saying that it is shameful that an innocent man's property was sold --- for if such things may be heard and freely said, Sextus Roscius was not so great a man in the state that we should especially complain of him --- but I ask this: how could it have been, even by that very law on the proscription, whether it is the lex Valeria or the lex Cornelia (for I do not know which, nor do I particularly know), how could the property of Sextus Roscius, even by that very law, have been sold?

\ciceroSection{126} For they say it is written thus: "Either the property of those who have been proscribed shall be sold," in which number Sextus Roscius is not; "or of those who have been killed in the strongholds of the enemy." So long as there were any strongholds, he was in Sulla's. After everyone had laid down arms, while returning from supper at Rome, in a time of complete peace, he was killed. If by law, I confess that the property too was sold by law. But if it is established that he was killed against all laws, not only old but new, by what right or in what way or by what law was the property sold? --- I ask.

\ciceroSection{127} Against whom do I say this, you ask, Erucius? Not against the man you wish and think; for both my speech, from the start, and his own outstanding virtue have always cleared Sulla. I say that all this was done by Chrysogonus: that he lied; that he pretended Sextus Roscius had been a bad citizen; that he said he had been killed in the enemies' lines; and that he did not allow Lucius Sulla to be informed of these matters by the deputation of the Amerians. Finally, I even suspect this: that the property was not sold at all. Which, if you, gentlemen, allow, will be made plain hereafter.

\ciceroSection{128} For I think there is in the law a date by which the proscriptions and sales should take place, undoubtedly the Kalends of June. Several months after that, the man was killed and the property is said to have been sold. Either the property never went into the public records at all (and we are being more cleverly mocked by this rascal than we suppose), or, if it did, the public records have been corrupted somehow; for that the property could not have been sold by law is plain. I am aware, gentlemen, that I am exploring these matters before their time, and almost wandering from my purpose --- since, when I ought to be dressing the wound at the head of Sextus Roscius, I am attending to a hangnail. For he does not labour over money; nor does he take account of any private advantage of his own. He thinks he will easily bear his poverty, if he is freed from this unworthy suspicion and fabricated charge.

\ciceroSection{129} But I beg you, gentlemen, to listen to these few things that remain in such a way that you understand me to be speaking partly for myself, partly for Sextus Roscius. For what seems to me unworthy and intolerable, and what seems to me to concern us all unless we take care, that I declare for myself out of the feeling and the pain of my own mind. What concerns the case and circumstance of his life, and what he wishes to be said in his own behalf, and on what condition he is content --- this you will hear, gentlemen, in the closing part of our speech.

\ciceroSection{130} I, leaving Sextus Roscius aside, ask of Chrysogonus on my own account: first, why was the property of an excellent citizen sold? Then, why was the property of a man who was neither proscribed nor killed in the enemies' lines sold, when the law was written for these alone? Then, why was it sold so long after the day fixed in the law? Then, why was it sold for so paltry a sum? Of all this, if (as worthless and dishonest freedmen often do) he wishes to put the responsibility on his patron, he will achieve nothing; for everyone knows that, owing to the magnitude of events, many men did many things which Lucius Sulla either disapproved of or was unaware of.

\ciceroSection{131} Is it then permissible that something be passed over by oversight in matters of this kind? It is not permissible, gentlemen, but it is necessary. For if Jupiter, the best and greatest, by whose nod and judgement the heavens, the earth, and the seas are ruled, has often by violent winds, by immoderate storms, by excessive heat, by intolerable cold, harmed men, destroyed cities, and ruined crops --- of which we suppose nothing was done for our destruction by divine plan, but by the very force and magnitude of things --- yet by contrast, the advantages we use, the light we enjoy, the breath we draw, all these we see given and shared to us by him: what wonder is it, gentlemen, that Lucius Sulla, while he alone governed the commonwealth and ruled the world, while he was confirming by laws the majesty of empire which he had recovered by arms, could not notice some things? Unless this is wonderful: that human attention should fall short of what divine power cannot reach.

\ciceroSection{132} But, leaving aside what has already been done, can anyone fail to understand from what is being done now, at this very moment, that the whole architect and engineer is one man, Chrysogonus? --- who saw to it that Sextus Roscius's name was laid on the indictment, and for whose sake Erucius said he was bringing the prosecution... [a passage is missing here]... men in the Sallentine country or in Bruttium, where they can scarcely hear news three times in a year, suppose themselves to have a properly arranged household.

\ciceroSection{133} The other comes down to you from the Palatine and his own house. He has, for his pleasure, a country estate near Rome and pleasant; many other estates besides, and not one of them but is splendid and close at hand. His house is stuffed with vessels of Corinth and Delos, among which is that "self-cooking" pot which he recently bought at such a price that, when passers-by heard the auctioneer enumerate the figure, they thought a farm was being sold. What besides of chased silver, of embroidered fabric, of painted panels, of statues, of marble do you suppose he has? As much, of course, as could be heaped together in one house from many splendid families amid the tumult and the plundering. As for his household of slaves, how large it is, and with what varied accomplishments, what shall I say?

\ciceroSection{134} I pass over the common arts: cooks, bakers, litter-bearers. For the soul and the ear he keeps so many men that the whole neighbourhood resounds daily with the music of voices and strings and pipes, and with nightly revels. In this kind of life, gentlemen, what daily expenses, what extravagances do you suppose are going on? What sort of dinners? Honourable, no doubt, in such a house --- if "house" is the word, rather than a workshop of vice and a lodging-place for every disgrace.

\ciceroSection{135} You see, gentlemen, how the man himself flits about everywhere through the Forum, his hair carefully done and shining with oil, with a great train of toga-clad followers; you see how he despises everyone, how he counts no man a man beside himself, how he thinks himself alone happy, alone powerful. If I were to recount what he carries off and what he attempts, I am afraid, gentlemen, that some less informed person might suppose that I wished to wound the cause and the victory of the nobility. Although by my own right I may, if anything in this part of things does not please me, find fault with it; for I do not fear that anyone will think me to have been ill-disposed towards the cause of the nobility.

\ciceroSection{136} Those who know me know that I, in proportion to my slender and weak share, after the thing I most desired could not be done --- a settlement --- chiefly defended this: that those should win who have won. For who was there who did not see that lowness was contending with rank for greatness? In which contest it was the act of a desperate citizen not to join himself to those by whose preservation rank was retained at home and authority abroad. That this has been carried through, that to each man his honour and his place have been restored, I rejoice, gentlemen, and am very glad; and I understand that all this was done by the will of the gods, the zeal of the Roman people, and the planning, command, and good fortune of Lucius Sulla.

\ciceroSection{137} That punishment was visited on those who fought against all reason, I cannot reproach. That brave men, by whose extraordinary work in the conduct of affairs, honour was given, I praise. I judge that the war was waged in order that these things might come about, and I confess that I was on that side in zeal of party. But if this was done, and arms were taken up, in order that the lowest men should be enriched out of others' money and should make assault on each man's fortunes, and if this is something which it is not permitted not only to forbid in fact but even to censure in words --- then truly in this war the Roman people were not regenerated nor restored, but subdued and crushed.

\ciceroSection{138} But it is far otherwise; nothing of all this is so, gentlemen. The cause of the nobility will not be wounded by your resisting these men, but rather adorned. For those who wish to find fault with these things complain that Chrysogonus has so much power. Those who wish to praise them recall that he has not been allowed it. And there is now nothing such that anyone is either so foolish or so wicked as to say: "I should have liked to have leave; I would have said this thing." Say it: you may. "I would have done this." Do it: nobody forbids you. "I would have voted thus." Vote, only correctly: everyone will approve. "I would have judged thus." Everyone will praise you, if you have judged rightly and in order.

\ciceroSection{139} While there was need, and the situation itself compelled it, one man could do all things. Once he created magistrates and established laws, his own administration and authority have been restored to each. If those who have got it back wish to keep it, they will be able to retain it forever. But if they will either commit, or will approve of, these murders and lootings and these so great and so wasted expenditures --- I do not wish to say anything heavier against them, even by way of omen --- I say only this: unless our nobility are watchful and good and brave and merciful, they must necessarily yield those ornaments of theirs to men who have these qualities.

\ciceroSection{140} Wherefore let them at last cease to say that anyone has spoken ill, if anyone has spoken truly and freely; let them cease to make their cause common with Chrysogonus's; let them cease, if he is wounded, to think that anything has been taken from themselves; let them see how shameful and wretched a thing it is that those who could not endure the splendour of the equestrian order should be able to bear the domination of a worthless slave. Which domination, gentlemen, used to operate in other matters before, but now you see what road it is paving, what path it is aiming at: at trust, at the sworn oath, at your courts, at that which alone in our state remains, almost, sound and inviolate.

\ciceroSection{141} Does Chrysogonus think he can do something here too? Does he wish to be powerful here too? O wretched and bitter business! By Hercules, I do not take this with indignation because I fear he can do anything, but because he has dared, because he has hoped that, before such men, he could do anything for the destruction of an innocent man --- this very thing I lament. Was it for this that the awaited nobility, with arms and steel, restored the commonwealth --- that, at their pleasure, the freedmen and slaves of the nobility might harry our property, our fortunes, our very altars?

\ciceroSection{142} If that was done, I confess that I was wrong who preferred this to come about; I confess I was mad who agreed with them. Although unarmed, gentlemen, I did agree. But if the victory of the nobility ought to be an ornament and a benefit to the state and the Roman people, then in truth my speech ought to be most welcome to every best and noblest man. If anyone thinks that he himself and the cause are wounded when Chrysogonus is reproached, he does not understand the cause but knows himself well; for the cause becomes more splendid if every worthless man is resisted. Only that most dishonest supporter of Chrysogonus, who thinks he has a common reckoning with him, is wounded when, from the splendour of the cause, he is set apart.

\ciceroSection{143} But all this speech of mine, as I said before, is mine, which the commonwealth and my own pain and the wrong of these men have compelled me to use. Sextus Roscius thinks none of this unworthy, accuses no one, complains nothing about his patrimony. He thinks --- a man inexperienced in modes of life, a farmer and rustic --- that all those things which you say were done by Sulla were done by custom, by law, by the law of nations. Cleared of guilt and freed from the abominable charge, he wishes to leave you;

\ciceroSection{144} if he is freed from this unworthy suspicion, he says he will bear with an equal mind being deprived of all his goods. He asks and beseeches you, Chrysogonus, if he has converted nothing of his father's most ample fortune to his own use, if in nothing he has defrauded you, if in entire good faith he has handed over, counted, weighed out everything of his to you, if he handed over to you the very clothing he wore and the ring on his finger, if he excepted nothing from anything but himself, naked --- to allow him through your good will to live an innocent life among his friends in poverty.

\ciceroSection{145} You possess my estates, I live by another's pity; I yield, both because my mind is even-tempered and because it is necessary. My house is open to you, closed to me; I bear it. You enjoy my huge household; I have no slave; I endure it and think it bearable. What more do you want? Why do you pursue me, why do you assail me? In what point do you suppose your wishes are wounded by me? Where do I cross your interests? What do I block in your path? If you wish to kill the man for the sake of the spoils, you have spoiled him; what more do you seek? If because of enmity, what enmities do you have with one whose estates you possessed before you knew him? If from fear, do you fear anything from one whom you see unable to ward off so great a wrong from himself? But if it is because the property which was Roscius's has become yours, and on that account you are eager to ruin this son of his --- do you not show that you fear what, beyond all others, you ought not to fear: that some day the patrimony of the proscribed may be restored to their children?

\ciceroSection{146} You do me wrong, Chrysogonus, if you place a greater hope of keeping your purchase in this man's destruction than in the things which Lucius Sulla has done. But if you have no reason why you should wish this wretched man to be afflicted with so great a calamity, if he has handed over to you everything except his life, and has reserved for himself nothing of his father's --- not even by way of memorial --- in the name of the immortal gods, what is this great cruelty, what so wild and savage a nature? What pirate was ever so unspeakable, what corsair so barbarous as, when he could have his prize whole and without bloodshed, to prefer to take the booty bloody?

\ciceroSection{147} You know that this man has nothing, dares nothing, is capable of nothing, has never thought anything against your interest; and yet you assail one whom you cannot fear and ought not to hate, in whom you see nothing left that you can take. Unless you think it unworthy that he sits clothed in court, whom you yourself drove out of his patrimony as out of a shipwreck, naked. As if you did not know that he is fed and clothed by Caecilia, daughter of Balearicus, sister of Nepos, a woman of the greatest distinction, who, though she had a most distinguished father, the most renowned uncles, the most adorned brother, nevertheless, woman though she was, has so wrought by her excellence that, however much honour she received from their dignity, she gave back no smaller ornament to them by her own praise.

\ciceroSection{148} Or do you think that, because he is being defended diligently, that is an unworthy fact? Believe me, if for the sake of the friendship and good will inherited from his father, all his guest-friends were willing to stand by him and to defend him freely, he would be defended copiously enough. But if, in proportion to the magnitude of the wrong, and in proportion to the fact that the very basis of the commonwealth is being tested in his danger, all these men were to claim it as their own, you would not, by Hercules, be permitted to stand in your present place. As it is, he is defended in such a way that the adversaries should not greatly take it amiss, nor consider themselves overborne by power.

\ciceroSection{149} What needs doing at home is done through Caecilia. The business of the Forum and the court Marcus Messalla, as you see, gentlemen, has undertaken; who, if he had now sufficient age and strength, would himself be speaking for Sextus Roscius. Since age stands in the way of his speaking --- and the modesty which adorns his age --- he has handed the case over to me, whom he understood to wish, as I owed it, to take it up for his own sake; and he himself, by his constancy, his counsel, his authority, his diligence, has brought it about that the life of Sextus Roscius, snatched from the hands of the confiscation-dealers, has been committed to the votes of the jurors. For surely, gentlemen, on behalf of this kind of nobility the greater part of the state was in arms; this was what we fought for, that those nobles should be restored to the state who would do what you see Messalla doing: defend the head of the innocent, resist injustice, would rather show what they could do for another's safety than for his ruin. And if all who were born into the same rank did the same, both the commonwealth would suffer less from them, and they less from envy.

\ciceroSection{150} But, gentlemen, if we cannot prevail upon Chrysogonus to be content with our money and not to seek our life --- if he cannot be brought to this, that, having taken from us all that was peculiarly ours, he should not also wish to snatch from us this common light --- if it is not enough for him to glut his greed with money, unless blood too has been offered to his cruelty --- then the one refuge, the one hope left to Sextus Roscius, gentlemen, is the same as that left to the commonwealth: your old goodness and pity. If these remain, we can still even now be saved; but if that cruelty which has been at work in the commonwealth at this time hardens and embitters even your minds --- which surely cannot happen --- it is finished, gentlemen; better to spend one's life among wild beasts than in such great savagery.

\ciceroSection{151} Were you reserved, were you chosen for this purpose --- to condemn those whom the confiscation-dealers and cut-throats could not slaughter? Good generals do this when they engage in battle: in the place where they think the enemy's flight will be, they station soldiers to fall, unforeseen, upon any who flee from the line. Just so, I take it, these confiscation-buyers think that you, men of this calibre, sit here to receive those who have escaped from their hands. May the gods forbid, gentlemen, that what our ancestors wished to be called "the public council" should be reckoned the bodyguard of confiscation-dealers!

\ciceroSection{152} Do you not understand, gentlemen, that nothing else is being done than that the children of the proscribed should be removed by every method, and that the beginning of this is being sought in your sworn oath and in Sextus Roscius's danger? Is it doubtful to whom the offence belongs, when on one side you see a confiscation-dealer, an enemy, an assassin --- and at the same time a prosecutor in this case --- and on the other a destitute son, well-regarded by his own people, in whom not only no fault, but not even any suspicion, has been able to take hold? Do you see anything else here standing against Roscius except that his father's property has been sold?

\ciceroSection{153} If you take this on yourselves, and offer your service to that purpose --- if you sit here so that the children of those whose property was sold shall be brought to you --- by the immortal gods, gentlemen, take care that a new and far more cruel proscription be not seen to be inaugurated through you. That earlier one which was made against those who could take up arms, the Senate nevertheless was unwilling to undertake, lest something sharper than ancestral practice should seem to be done by public counsel. This other one, which extends to their children and to the cradles of their infants --- unless by this present judgement you reject it and turn from it, by the immortal gods, see to what point you suppose the commonwealth must come.

\ciceroSection{154} Wise men, endowed with the authority and power that you have, ought above all to apply remedies to the things from which the commonwealth most suffers. There is no one of you who does not understand that the Roman people, who once were thought most lenient towards their enemies, at this moment suffer from cruelty at home. Take this away from the state, gentlemen; do not allow it to go on any longer in this commonwealth; which has not only this evil in it, that it has carried off so many citizens most atrociously, but even this --- that it has taken away the capacity for pity from the gentlest of men, by the habit of these afflictions. For when at every hour we see or hear of something atrocious being done, even those of us who are by nature most mild lose, by the constancy of these troubles, all sense of human feeling from our minds.
