% Ad Familiares 10.26 (ad-familiares-10-26) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. C. FVRNIO.}

\ciceroSection{1} When I had read your letter, in which you indicated that either the men of Narbonensis must be abandoned, or there must be a fight at hazard, I feared the latter the more --- and I am not sorry to find that it has been avoided. As for what you write about the concord of Plancus and Brutus, in that I place my very greatest hope of victory. About the loyalty of the Gauls we shall, as you write, learn one of these days by whose effort above all it has been stirred up; but already, believe me, we have learned. And so, after your most welcome letter, at the close of it I lost my temper. For you write that, if the elections fall in Sextilis, you will be here quickly; if they are already over, more quickly still --- so as not to remain a fool any longer at the hazard of your life.

\ciceroSection{2} O my Furnius! How little you know your own case, when you can read other men's so easily! Are you now thinking of yourself as a candidate, with this in mind: that you must either run for the elections or, if they are already over, sit at home --- so as not, with the greatest danger (as you write), to be most thoroughly stupid? I do not believe you really think it; for I know all your impulses to credit. But if you really think as you write, it is not so much you I find fault with as my own opinion of you. Shall the headlong haste of getting to a magistracy --- the cheapest and the most common, if you get it as the run of men do --- pull you away from praise so great, with which all men, in justice and in truth, are carrying you to the skies? Is the issue, then, this --- whether you become praetor at this election or the next, not how you might deserve so well of the state as to be judged the worthiest of every honour?

\ciceroSection{3} Do you really not know how high you have climbed, or do you count it as nothing? If you do not know it, I forgive you, we are at fault; but if you do see it, is there any praetorship sweeter to you than duty, which few men follow, or than glory, which all men pursue? Calvisius and I --- and Calvisius is a man of great judgement and most devoted to you --- find fault with you for this every day. The elections, all the same, since you are hanging on them, we are pushing back into the month of January, so far as we can, because for many reasons we think it serves the state. Win, then, and farewell.
