% Ad M. Brutum 1.3 (ad-brutum-01-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} Our situation seemed to be in a better posture: for I know for certain that what has been done has been written to you. The consuls have turned out to be just the kind of men I have so often described to you. As for the boy Caesar, his natural endowment of virtue is extraordinary. I only wish we may guide and hold him as easily when he is in full flower of office and popularity as we have held him so far. That, I grant, is the harder thing; even so, I do not despair. For the young man has been persuaded --- and most of all through me --- that we owe our survival to his efforts. And certainly, if he had not turned Antony away from the city, all would have perished.

\ciceroSection{2} Indeed, three or four days before this most splendid event, the entire citizenry, struck by a kind of panic, was pouring out to you, wives and children and all; the same city, restored on the twelfth day before the kalends of May, preferred your coming here to its going to you. On that day, in fact, I reaped the greatest reward of my many labours and many wakeful nights --- if there is any reward at all in solid and genuine glory. A throng was gathered around me as large as our city can hold; and from it I was escorted up to the Capitol, then placed on the Rostra amid the loudest cheering and applause. There is nothing empty in me, nor should there be; even so, the unanimity of every order, the formal vote of thanks and congratulation, moves me, for this reason --- that to be a man of the people in saving the people is a fine thing indeed.

\ciceroSection{3} But I had rather you have all this from others. As for your own affairs and plans, I should like you to keep me informed with the greatest diligence; and I should like you also to take care that your generosity not look too loose. The Senate's view, and the Roman people's view, is this: that no enemies have ever been worthier of every penalty than those citizens who in this war have taken up arms against their fatherland; and these I, for my part, in every motion I cast, am pursuing in vengeance --- with all loyal men's approval. What you think on this matter is for your own judgement; I think the cause of the three brothers is one and the same.
