% Ad Familiares 10.21 (ad-familiares-10-21) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{PLANCVS CICERONI}

\ciceroSection{1} I would be ashamed of the inconsistency of my letters, if it did not depend on another man's fickleness. I did everything to ensure that, with Lepidus joined to me in the defence of the state, I might withstand desperate men with less anxiety on your part; I accepted everything from him when he asked, and made promises to him of my own accord besides, and wrote to you two days ago that I was confident I should find Lepidus a good man and should conduct the war by a shared strategy. I trusted his autograph letters, and the assurance of Laterensis on the spot, who at the time was with me and begged me to be reconciled to Lepidus and to trust him. I was not allowed to hope well of him any longer; but this at least I have guarded against, and shall guard against: that the supreme interests of the state shall not be betrayed by my credulity.

\ciceroSection{2} When I had thrown a bridge across the river Isara in a single day and led the army over --- applying speed to match the magnitude of the matter, because Lepidus himself had asked by letter that I hasten my coming --- his orderly was at hand with a despatch in which he ordered me \textit{not} to come; he could finish the business by himself; meanwhile I should wait at the Isara. I will lay bare my rash plan to you: I had nonetheless decided to go on, supposing that he was trying to dodge a partner in his glory. I imagined I could both refrain from skimming any of the credit from a man hungry for it, and yet have my forces close at hand, so that if anything went harder than expected I could come up swiftly in support.

\ciceroSection{3} I, no bad man myself, was suspecting this much. But Laterensis, a man of the most upright character, sends me by his own hand a letter despairing only too deeply of himself, of the army, of Lepidus's good faith, complaining that he had been abandoned, in which he openly warns me to look out lest I be deceived; his own pledge was discharged; I was not to fail the state. I have sent a copy of his autograph to Titius; the autographs themselves --- all of them, both those I believed and those I judged should not be trusted --- I shall give to Laevus Cispius to carry through, who was party to the whole affair.

\ciceroSection{4} On top of this it happened that Lepidus's soldiers, while he was haranguing them --- corrupt of themselves and further corrupted by their officers, men of the Canidius and Rufrenus stamp and the rest, whose names you shall know when occasion serves --- shouted with one voice that they, like good men, wanted peace and would not fight against any party, now that two consuls of exceptional quality had been killed, so many citizens lost in the cause of their country, all the enemy declared public enemies at last and their goods confiscated; and Lepidus neither punished this nor cured it.

\ciceroSection{5} To advance into this and to throw a most loyal army, the greatest of auxiliary forces, the chiefs of Gaul, an entire province, into the path of two armies joined together --- I saw this would be the height of folly and recklessness; and that for me, if I were so crushed and betrayed the state along with myself, there would be lacking, in death, not only honour but even pity. So I am about to fall back, and I shall not allow such great prizes to be handed over to desperate men.

\ciceroSection{6} I shall take pains to hold the army in suitable positions, to protect the province even if that other army has defected, and to keep everything intact, until you send up reinforcements and with matching good fortune vindicate the state on your end. No man has been readier to fight to the finish, if occasion offers, to be besieged, if it must be, to die, if it falls so, for your sake. Therefore I urge you, my Cicero, to see to it that that army is sent across as quickly as possible, and to hasten it on, before the enemy gather more strength and our side fall into confusion. If speed is applied to this, the state will remain in possession of victory and the criminals will be destroyed. Take care to keep well, and to hold me in your affection.
