% Ad Familiares 10.20 (ad-familiares-10-20) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO PLANCO}

\ciceroSection{1} Everything that has been coming from your part of the world has been so uncertain that I have not had it in me to know what to write to you. One moment the news about Lepidus is what we should wish, the next the opposite; about you, however, the report is steady --- that you cannot be either deceived or beaten. Of the two, the first is in part a matter of fortune; the second belongs altogether to your own good sense.

\ciceroSection{2} But I have had a letter from your colleague, dated the Ides of May, in which it is said that you had written to him that Antony was not being received by Lepidus. That will be more certain if you write the same thing to us yourself. But perhaps you are less bold to say so on account of the empty rejoicing your earlier letters caused. Yet, however much you could be mistaken, my dear Plancus --- and who could be safe from that? --- who does not see that to be deceived is something you could not be? Now indeed even the occasion for any mistake has been taken away; for that previous slip has been chalked up to the well-known proverb, ``the same man at the same stone twice.'' But if things really stand as you wrote them to your colleague, we are released from all anxiety --- and yet we shall not really be released until you yourself have made us certain that this is so.

\ciceroSection{3} My own opinion, as I have written to you more than once, is this: whoever crushes the remnants of this war will have brought the whole of it to its end. I want this man to be you, and I am confident it will be. As for my zeal for you --- which, certainly, no zeal of mine could ever have exceeded --- that you should find it as pleasing as I supposed it would be does not surprise me in the least, and gladdens me powerfully. Yet, if all goes rightly with you on that side, you will come to know it as something larger and weightier still. The 29th of May.
