% Ad M. Brutum 1.15 (ad-brutum-01-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} You have Messala with you. By what letters, then, however carefully written, can I succeed in setting forth with more precision what is being done and what the state of the commonwealth is, than this man will set it out for you --- a man who knows everything perfectly, and can unfold and convey it all to you with the utmost elegance? Take care not to suppose, Brutus (though there is no need for me to write to you what is already known to you; yet I cannot pass over in silence so consummate an excellence in every kind of distinction) --- take care not to think that anything is comparable to him in integrity, steadiness, devotion, and zeal for the commonwealth, so that his eloquence, in which he is wonderfully preeminent, scarcely seems to find room for praise alongside the rest; and yet in that very faculty his wisdom is more apparent still, so weighty is the judgement and so much the art with which he has trained himself in the truest style of speaking. And so industrious is he, and so wakeful in his application, that the largest part of the gratitude is not, it seems, owed to his natural ability, which in him is supreme.

\ciceroSection{2} But my affection is carrying me away. For the purpose of this letter was not to praise Messala, especially to Brutus, to whom his virtue is no less known than to me, and these very studies which I am praising are still better known. When I sent him from me, I felt it heavily, but I was relieved by this one thing: that in setting out to you, as if to a second self of mine, he was at once performing a duty and pursuing the highest distinction. But enough of this.

\ciceroSection{3} I come now --- after a long interval indeed --- to a certain letter of yours, in which, while granting me much, you reproached me on one count: that I was excessive in voting honours, and as it were prodigal. You say this; another perhaps would say that I was too harsh in censure and punishment --- unless perhaps you say both. If that is the case, I am eager that my judgement on both matters be made fully known to you, †and not merely so that I may borrow Solon's saying† --- Solon, who was the wisest of the Seven, and the only law-giver of the Seven. He said that the commonwealth is held together by two things: reward and punishment. There is, of course, a measure in each, as in all other things, and a certain mean in either kind. But to argue on so great a matter is not my purpose here.

\ciceroSection{4} What, however, I have followed in this war when giving my opinions in the Senate I think not unfitting to disclose. After Caesar's death and your memorable Ides of March, Brutus, you have not forgotten what I said had been left undone by you, and what a storm I said was hanging over the commonwealth. A great pestilence had been driven off by you, a great stain wiped from the Roman people; for you, indeed, a divine glory had been won --- but the instrument of monarchy had been handed over to Lepidus and Antony. Of these the one was more inconstant, the other more foul; both feared peace and were enemies of quiet. With these men burning with the desire to overturn the commonwealth, we had no garrison that could be set against them; for the state had roused itself, united in holding fast to liberty.

\ciceroSection{5} We at that time were perhaps too fierce; you, perhaps more wisely, withdrew from the city which you had liberated, and let Italy go on offering you her own enthusiasms. And so, when I saw that the city was held by parricides, and that neither you nor Cassius could be safe within it, and that it was crushed by Antony's arms, I judged that I too must leave it: for it was a hideous sight --- the state crushed by the impious, the power of bringing aid cut off. But the same spirit that has always been fixed in me, my love of country, could not bear to be parted from her dangers. And so, in mid-course toward Achaia, when in the days of the Etesian winds a south wind had carried me back into Italy, as if to argue me out of my plan, I saw you at Velia and grieved deeply. For you were retreating, Brutus, retreating --- though our Stoics deny that the wise man flees.

\ciceroSection{6} As soon as I reached Rome, I set myself at once against Antony's crime and madness; and when I had drawn his anger down upon me, I began to lay plans, purely in the Brutus manner (for these are the proper habits of your bloodline), for the liberating of the commonwealth. What remains is long, and must be passed over; for it concerns myself. I say only this: that this young Caesar, through whom we are still here, if we are willing to confess the truth, flowed from the wellspring of my counsels.

\ciceroSection{7} The honours I bestowed on him, Brutus, were none but what were owed, none but what were necessary. For when we first began to recall liberty, when not even D. Brutus's divine valour had yet so stirred itself that we could know of it, and our whole defence rested in a boy who had turned Antony away from our throats --- what honour was not to be voted to him? Yet at that time I bestowed on him only verbal praise, and even that in measured terms; I also decreed for him the imperium, which, however honourable it might seem at his age, was nevertheless necessary for one who held an army. For what is an army without imperium? Philippus moved a statue; Servius first moved the lowering of the age for standing for office, and then Servilius an even greater lowering. Nothing at that moment seemed excessive.

\ciceroSection{8} But somehow it is easier to find men generous in fear than grateful in victory. For when, after D. Brutus's liberation, that most joyful day had dawned on the citizenry, and it happened to be Brutus's birthday as well, I decreed that Brutus's name be inscribed in the calendar against that day; and in this I followed the precedent of our ancestors, who paid this honour to the woman Larentia, at whose altar in the Velabrum you pontiffs are accustomed to offer sacrifice. When I gave this to Brutus, I wanted there to be in the calendar an everlasting mark of a most welcome victory. And on that day I learned that there were rather more ill-wishers in the Senate than grateful men. In those very days I poured out honours, if you will have it so, upon the dead --- on Hirtius and Pansa, on Aquila as well. Who shall reproach this, unless one who, having laid down his fear, has forgotten the danger that is past?

\ciceroSection{9} There was added to the welcome remembrance of the benefit that other consideration, which would also be wholesome for posterity. For I wanted there to stand, against the cruellest of enemies, eternal monuments of the public hatred. I suspect that what is less acceptable to you is what was not acceptable to your intimates, excellent men, indeed, but inexperienced in public affairs: that I decreed that Caesar be allowed to enter the city in ovation. As for myself --- but I am perhaps mistaken; and yet I am not the kind of man whom my own counsels especially delight --- I seem to myself in this war to have judged nothing more prudently. Why this is so must not be laid open, lest I seem to have been more provident than grateful. This itself is too much; let us look at the other side. To D. Brutus I voted honours, to L. Plancus I voted honours. Glorious indeed are those natures that are drawn on by glory; but the Senate, too, is wise when it employs, for the helping of the commonwealth, whatever lever it thinks --- provided it is an honourable one --- can move each man. But in the case of Lepidus we are reproached: when we had set up a statue for him on the Rostra, we ourselves overthrew it. We were eager by an honour to recall him from madness. The insanity of a most worthless man overcame our prudence; yet there was not so much harm done in setting up Lepidus's statue as there was good in overthrowing it.

\ciceroSection{10} Enough on the matter of honours; now a few words must be said on punishment. For I have understood from your letters, often enough, that in the case of those whom you have conquered in war you wish your clemency to be praised. For my part, I judge that you do nothing that is not wise; but to pass over the punishment of a crime --- for that is what is called pardoning --- though it may be tolerable in other matters, I think pernicious in this war. For there has been no civil war in our commonwealth, of all those within my memory, in which, whichever side had won, some form of commonwealth was not going to remain. In this war, what commonwealth we who win shall have, I could not easily declare; but for those who lose, there will assuredly never be any. I therefore spoke sternly against Antony, sternly against Lepidus, not so much from a wish to take vengeance, as to deter wicked citizens at the moment by fear from attacking their country, and to set up an example for the future, that no one should wish to imitate such madness.

\ciceroSection{11} And yet this view was no more mine than everyone's. In it, there seems to be this much that is cruel: that the punishment reaches the children, who have deserved nothing. But this is both an ancient practice and one common to all states --- since even the children of Themistocles were in want; and if the same penalty pursues citizens who have been condemned in a court, how could we be milder toward enemies? And what cause can anyone have to complain of me, who, if he had won, must himself confess he would have been harsher toward me? You have the rationale of my opinions, at least on this question of honour and punishment; on other matters, what I felt and what I voted, I think you have heard.

\ciceroSection{12} But these things are not so very pressing; what is altogether pressing, Brutus, is that you come into Italy with your army as soon as possible. The expectation of you is at its height. If you so much as touch Italy, the whole people will come running to you. For either we shall have won --- as indeed we had most splendidly won, had not Lepidus desired to ruin everything and to perish himself together with his followers --- and then we need your authority to set up some kind of constitution for the state; or, if even now the struggle remains, our greatest hope lies both in your authority and in the strength of your army. But hurry, by the gods! You know how much depends on timing, how much on speed.

\ciceroSection{13} How carefully I am looking after your sister's sons I trust you will learn from your mother's and your sister's letters. In this affair I have more regard for your wish, which is most dear to me, than --- as some think --- for my own consistency. But there is nothing in which I would rather be, and seem to be, consistent than in my love for you.
