% Ad Familiares 10.30 (ad-familiares-10-30) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{GALBA CICERONI S.}

\ciceroSection{1} On 14 April, the day Pansa was to be in Hirtius's camp --- I was with him, for I had gone out one hundred miles to meet him, so that he might come up the more quickly --- Antony led out two legions, the Second and the Thirty-Fifth, and two praetorian cohorts, one his own and one Silanus's, and a portion of the re-enrolled veterans. So he came out to meet us, because he supposed we had only four legions of recruits. But by night, so that we could come into camp in greater safety, Hirtius had sent us the Martian Legion (which I used to command) and two praetorian cohorts.

\ciceroSection{2} When Antony's cavalry came in sight, neither the Martian Legion nor the praetorian cohorts could be held back; we were forced to follow them, since we could not call them back. Antony was keeping his forces in hand at Forum Gallorum and did not want it known that he had the legions; he was showing only cavalry and light-armed troops. After he saw that Pansa was letting the legion go forward against his will, he ordered two legions of recruits to follow him. After we had crossed the narrows of marsh and woods, a line of battle was drawn up by us, of twelve cohorts; the two legions had not yet come up:

\ciceroSection{3} suddenly Antony brought his forces out of the village into line and engaged without delay. At first the fighting was such that it could not have been more fierce on either side; though the right wing, on which I was with eight cohorts of the Martian Legion, had at the first onset put to flight Antony's Thirty-Fifth, so that it advanced more than five hundred paces beyond the line on which it had stood. So when the cavalry tried to outflank our wing, I began to fall back and to throw up the light-armed troops against the Moorish cavalry, to keep them from attacking our men in the rear. In the middle of this I see that I am in among Antony's men, and that Antony is some way behind me. Suddenly I put spurs to my horse toward that legion of recruits that was coming from the camp, with my shield slung behind me. Antony's men were chasing me; our men were on the point of throwing their javelins. So by I know not what fate I was saved, in that I was quickly recognized by our own.

\ciceroSection{4} On the Aemilian Way itself, where Caesar's praetorian cohort was, the fighting went on a long while. The left wing, which was weaker --- where the two cohorts of the Martian Legion were, and the praetorian cohort --- began to give ground, because it was being outflanked by the cavalry, the arm in which Antony is by far the strongest. When all our ranks had fallen back, I, last of all, began to fall back to the camp. Antony, as if victorious, thought he could take the camp. When he came up to it, he lost a good many there and accomplished nothing. On hearing the news, Hirtius with twenty veteran cohorts met Antony as he was returning to his own camp, and destroyed all his forces and put them to flight on the very spot where the fighting had been, at Forum Gallorum; Antony with his cavalry got back to his own camp at Mutina at the fourth hour of the night.

\ciceroSection{5} Hirtius returned to the camp from which Pansa had set out, where he had left the two legions that had been attacked by Antony. Thus Antony has lost the greater part of his veteran forces; nor however could this be done without some loss of our own praetorian cohorts and the Martian Legion. Two eagles, sixty standards, of Antony's have been brought in; the business has been well managed. 15 April, from camp.
