% Ad M. Brutum 1.14 (ad-brutum-01-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} A short letter from you --- short, do I say? rather, none at all. Three little lines, in such times as these, from Brutus to me? You would have done better not to write. And you ask after my letters! When did any of your people ever come to you without one of mine? And what dispatch of mine has not had weight? If they have not been delivered to you, I do not suppose your own household letters have been delivered either. You write that you will give my Cicero a longer letter: right enough, but this one too ought to have been fuller. As for me, when you had written to me about Cicero's departure from you, I packed off couriers at once and a letter to Cicero himself, telling him that even if he had reached Italy, he was to return to you; for there is nothing more pleasant to me, nothing more honourable to him. Even though I had several times written to him that the elections of the priestly colleges had, at my own utmost insistence, been put off to the next year --- which I worked for both on Cicero's account, and on that of Domitius, Cato, Lentulus, and the Bibuli; this I had written to you too --- but plainly, when you were sending me that tiny letter of yours, this was not yet known to you.

\ciceroSection{2} For which reason, my Brutus, I beg you with all earnestness not to let my Cicero go, but to take him along with you --- and this very thing, if you have any regard for the commonwealth for whose sake you have been brought up, you must do this very moment. For the war has been reborn, and not slightly, through the crime of Lepidus. And Caesar's army, which was the best, not only is doing no good but is even forcing your army to be clamoured for. If that army shall touch Italy, there will be no citizen --- no one, at least, whom it is lawful to call a citizen --- who does not take himself into your camp. Even though we have Brutus excellently joined with Plancus; but you are not unaware how uncertain men's minds are when tainted by partisanship, and how uncertain are the outcomes of battles. Indeed, even if --- as I hope --- we shall have conquered, still the situation will require great steering by your counsel and authority. Come to our aid, then, by the gods, and that as soon as may be; and persuade yourself that on the Ides of March, when you drove out slavery from your fellow citizens, you did not bring your country more benefit than you will bring by coming in good time. 11 July.
