% Ad Familiares 11.14 (ad-familiares-11-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO D. BRVTO COS. DESIG. S. D.}

\ciceroSection{1} I am beyond measure delighted, my dear Brutus, that my counsels and my proposals --- on the board of ten, on heaping honours on the young man --- have your approval. But what does it matter? Believe me, I am no boaster: I have grown plainly cold by now, Brutus. The Senate was my instrument, and that instrument has been broken up. Your splendid sortie from Mutina, the rout of Antony, the army cut to pieces, had brought such hope of a victory now made sure --- \textit{[the text is corrupt here]} that everyone's spirit has gone slack, and those vehement exertions of mine seem now like so much shadow-boxing \textit{[Greek: skiamachiai]}.

\ciceroSection{2} But to return to the business: as for the Martian legion and the Fourth, those who know them say that they cannot on any terms be brought over to you. As for the money you want, an arrangement can be made, and shall be. On bringing Brutus over and keeping Caesar back to garrison Italy, I am very much of your mind --- but, as you write, you have detractors; I myself bear up under them easily enough, but they do impede the business nonetheless.

\ciceroSection{3} From Africa the legions are awaited. But that this war should have flared up again --- men are astonished. Nothing was ever so far beyond expectation. For on your birthday, with the victory just announced, we were seeing the state set free for ages on ages: these new alarms are unravelling all of it. You wrote to me, in the letter you sent on the Ides of May, that you had only then received from Plancus the news that Antony is not being received by Lepidus. If that is so, everything is easier; if otherwise, the business is large, but I do not fear the outcome of it. The role is yours; I cannot do more than I have done. Still, what I hope is that I may see you ranked greatest and most illustrious of all.
