% Ad Familiares 11.20 (ad-familiares-11-20) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{D. BRVTVS S. D. M. CICERONI}

\ciceroSection{1} What I do not do for myself, my love for you and your kindnesses to me compel me to do for you: to be afraid. For though it had often been said to me, and not despised by me, most recently Labeo Segulius --- a man entirely true to himself --- tells me he was with Caesar and that there was much talk about you. Caesar himself, he says, made no real complaint of you, except that he was reported to have said you had said the young man must be ``praised, decorated, and lifted up'' --- and he, Caesar, would not allow it to come about that he be lifted out of the way. I myself believe Labeo either reported this back to him or invented the saying --- it was not let drop by the young man. Labeo wanted me to believe besides that the veterans talk in the ugliest terms, and that the greatest danger threatens you from them, and that they are most indignant that neither Caesar nor I was placed on the board of ten, and that everything has been entrusted to your discretion.

\ciceroSection{2} When I had heard this, and was already on the road, I did not think I should commit myself to crossing the Alps before I knew what was being done with you there; for as to your danger, believe me, they are hoping that by flinging words about and proclaiming the danger, with you frightened and the young man pushed on, they may secure great rewards for themselves --- the whole song hanging on this, that they may make all the profit they can. Not, for all that, that I would not have you cautious and on your guard against ambushes; for nothing can be sweeter to me than your life, or more precious.

\ciceroSection{3} See, then, that by fearing you are not driven to fear the more; and meet the veterans on the points where they can be met. First, that you do for them what they want about the ten commissioners; then, on the question of rewards, if you think it good, propose that the lands of those soldiers who were veterans with Antonius be assigned to them, with both of us proposing it; concerning the money, that the Senate will decide on the matter slowly and with the state of the treasury taken into account. As for the four legions to which you decreed that lands be given, I see there will be the means in the Sullan lands and the Campanian district; the lands ought, I think, to be assigned to the legions equally or by lot.

\ciceroSection{4} It is not my own prudence that urges me to write you these things, but my love for you and a longing for quiet --- which cannot stand without you. For myself, unless there is the gravest necessity, I shall not leave Italy. I am arming the legions, fitting them out. I hope to have not the worst army for all chances and all the assaults of men. Of the army Pansa had, Caesar will not send back to me the legion. Reply to this letter at once, and send some one of your people, if there is anything more confidential and you think it necessary that I know. The ninth day before the Kalends of June, at Eporedia.
