% Ad Familiares 11.13 (ad-familiares-11-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{D. BRVTVS IMP. COS. DES. S. D. M. CICERONI}

\ciceroSection{1} I no longer thank you in words; for a man to whom I can scarcely repay the thing itself --- to him the matter does not permit satisfaction in words. I want you to attend to what is in hand; for, given your prudence, nothing will escape you if you read my letter carefully. To pursue Antonius at once, Cicero, I could not, for these reasons: I was without cavalry, without pack-animals; I did not know Hirtius had perished, I did not know Aquila had perished; and I would not trust Caesar before I had met and conferred with him. This day passed in this way.

\ciceroSection{2} The next morning I was summoned by Pansa to Bononia. While I was on the road, word was brought to me that he was dead. I hurried back to my little forces --- for so in truth I can call them; they are reduced to almost nothing and very ill served by want of every kind. Antonius got two days ahead of me; he made far longer marches in flight than I made in pursuit, since he moved at random and I in order. Wherever he went, he broke open the slave-barracks and dragged the men off; he made no halt until he reached Vada. I want that place to be familiar to you. It lies between the Apennines and the Alps, and is of all places the most obstructed for travel.

\ciceroSection{3} When I was thirty miles from him, and Ventidius had now joined him, a transcript of his speech to the troops was brought to me. In it he began by asking the soldiers to follow him across the Alps; he had, he said, an understanding with M. Lepidus. He was shouted down repeatedly by Ventidius's soldiers (for of his own he has very few indeed): they cried that they had either to die in Italy or to conquer there, and they began begging him to march on Pollentia. Since he could not hold them, he put off his march to the following day.

\ciceroSection{4} When this had been reported to me, I at once sent five cohorts ahead to Pollentia and turned my own march there. My garrison reached Pollentia one hour before Trebellius with his cavalry. I was overjoyed; for in this, I believe, victory consists.
