% Ad Familiares 12.7 (ad-familiares-12-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CASSIO S.}

\ciceroSection{1} With what zeal I have defended your standing both in the Senate and before the people, I would rather you learn from your own people than from me. My motion would easily have prevailed in the Senate, had not Pansa fiercely stood against it. After that motion was delivered, I was led out to a public meeting by the tribune of the plebs M. Servilius. There I said about you what I could --- with such intensity as the Forum is wide, with such an outburst and unanimous assent on the people's part that I have never seen anything to match it. For this I would have you forgive me, that I did it against the wish of your mother-in-law. The poor lady was timid; she was afraid Pansa's feelings would be offended. Pansa indeed said at the meeting that your mother and your brother too had not wanted that motion of mine to be put. But these considerations did not move me; others weighed more: I was acting for the state, as I always have, and for your own standing and glory.

\ciceroSection{2} As for what I argued at greater length in the Senate, and said in the public meeting, I would have you make my word good. For I promised, and almost guaranteed, that you had not waited for any decree of ours and would not wait, but would yourself, after your own fashion, take up the defence of the state. And although we had as yet heard nothing of where you were or what forces you had, still this was my fixed conviction: that all the resources and forces in those parts were yours, and that through you the province of Asia, I trusted, was already recovered for the state. See to it that, in adding to your glory, you outdo yourself. Farewell.
