% Ad Familiares 12.15 (ad-familiares-12-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{P. LENTVLVS P. F. PROQ. PROPR. S. D. COSS. PR. TR. PL. SENATVI POPVLO PLEBIQVE ROMANAE}

\ciceroSection{1} If you and your children are well, it is well; I and the army are well. When Asia had been crushed by the criminal act of Dolabella, I withdrew into the neighbouring province of Macedonia and to the garrisons of the commonwealth which M. Brutus, that distinguished man, was holding, and set myself to seeing that --- by whatever means it could be done most quickly --- the province of Asia and its revenues should be brought back into your power. Dolabella, in alarm at this, after laying the province waste, after snatching up the revenues, after stripping the Roman citizens above all most cruelly and selling them off, withdrew from Asia faster than a garrison could be brought to it. I did not think it necessary to linger longer or to wait for that garrison, but judged that I should return as soon as possible to my own duty: to collect what revenues remained, to bring in the money I had deposited, to investigate as quickly as I could whatever had been seized of it and by whose fault that had come about, and to make you fully informed on every point.

\ciceroSection{2} Meanwhile, as I was sailing through the islands towards Asia, word came to me that Dolabella's fleet was in Lycia, and that the Rhodians had a good number of ships in the water, rigged and ready. So with the ships I had either brought with me or that Patiscus the proquaestor had procured --- a man bound to me as much by friendship as by shared feeling in matters of state --- I turned aside to Rhodes, trusting in your authority and the decree of the Senate by which you had judged Dolabella an enemy, and trusting too in the treaty which had been renewed with them in the consulship of M. Marcellus and Ser. Sulpicius, by which the Rhodians had sworn to hold the same enemies as the Senate and Roman People. The matter deceived us bitterly. So far were we from strengthening our fleet by support from them, that our soldiers were even shut out by the Rhodians from city, harbour, and the roadstead outside the city, from supplies, finally from water itself; we ourselves were scarcely received, each in a single small boat. This indignity, this diminution of our own standing --- not only as private men but of the very imperium and the Roman People --- we bore for the following reason: from intercepted letters we had learned that Dolabella, if he should despair of Syria and Egypt (as he must), stood ready to embark with all his bandits and all his money and to make for Italy; and that with this in view the cargo-ships --- the smallest among them of two thousand amphorae --- had been gathered in Lycia and were under blockade by his fleet.

\ciceroSection{3} Stricken with fear at this, senators, I chose to endure injuries and to try every means first, even at the cost of insult to myself. So, on their own terms, having been admitted into the city and into their Senate, I pleaded the cause of the commonwealth as carefully as I could, and laid out the whole danger which would press on us if that bandit and all his men should put to sea. But I found the Rhodians sunk in such perversity that they reckoned every strong man worth more than every good one; that they refused to believe this concord and conspiracy of all orders in defence of liberty had been formed in earnest; that they were still confident the patience of the Senate and of every honest man would hold, and that no one could have dared declare Dolabella an enemy; that, in short, they took whatever the wicked were inventing for truer than what had truly been done and was being shown them by us.

\ciceroSection{4} It was in this same temper that, even before our arrival --- after the most monstrous murder of Trebonius and all those other crimes, so many and so unspeakable --- they had twice sent embassies to Dolabella; and this against the precedent of their own laws, with the magistrates then in office trying to prevent it. \dag{} These things, whether from fear, as they keep saying, for the lands they hold on the mainland, or from madness, or from the indulgence of a few --- men who before this had treated the most distinguished of our countrymen with the same insult, and who now, holding the highest magistracies, would not, when they easily could, take steps in the face of any precedent against our peril, the peril of those of us on the spot, and the imminent peril of Italy and our city, should that parricide, driven by his bandits in his ships from Asia and Syria, have made for Italy. \dag{}

\ciceroSection{5} In the case of some of them the magistrates themselves came under suspicion of having detained us and held us up, so that Dolabella's fleet might be informed of our coming. Several subsequent events deepened that suspicion --- above all that, suddenly, from Lycia, Sex. Marius and C. Titius, lieutenants of Dolabella, left the fleet and fled in a warship, abandoning the cargo-vessels, on the gathering of which they had spent no small amount of time and labour. And so when we had come from Rhodes into Lycia with the ships we had, we recovered the cargo-ships and restored them to their owners; and at the same time we ceased to fear what we had most feared --- that Dolabella might reach Italy with his bandits. We pursued the fleeing fleet as far as Sida, which is the furthest district of my province.

\ciceroSection{6} There I learned that part of Dolabella's ships had scattered, while the rest had made for Syria and Cyprus. With them thus dispersed, and knowing that the very large fleet of C. Cassius --- that singular citizen and commander --- would be at hand in Syria, I returned to my own duty; and I will take pains to show you, senators, and the commonwealth, my zeal and my diligence, and to gather all the money I can as quickly as I can and send it to you with full accounting. If, in going through the province, I learn who has kept faith with us and the commonwealth in preserving the money I had deposited, and who by their criminal act, freely turning over the public money, entered into this partnership in crime with Dolabella, I will inform you. If, as they deserve, you decide on these men severely (should that seem good to you) and strengthen our hand by your authority, we shall be the better able to collect the remaining revenues and to keep safe those already collected. In the meantime, that I may guard the revenues the more readily and defend the province from injury, I have raised a force of volunteers as the need required.

\ciceroSection{7} After this letter was written, some thirty soldiers whom Dolabella had levied in Asia came into Pamphylia in flight from Syria. They reported that Dolabella had come to Antiochia in Syria; that he had not been received; that he had several times tried to break in by force; that he had always been driven off with great loss to himself; and so, after losing some six hundred men, leaving the sick behind, he had fled by night from Antiochia in the direction of Laodicea; that on that same night nearly all the Asiatic soldiers had deserted him; that of these some eight hundred had returned to Antiochia and surrendered themselves to those whom Cassius had left in charge of the city, while the rest had come down through Mount Amanus into Cilicia, among whose number they themselves said they belonged; that Cassius, moreover, with all his forces, was reported to be four days' march from Laodicea at the time when Dolabella was making for it. For which reason I am confident that this most criminal bandit will pay his penalty sooner than expectation reckoned. The fourth day before the Nones of June, at Perga.
