% Ad Familiares 11.21 (ad-familiares-11-21) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. D. BRVTO IMP. COS. DESIG.}

\ciceroSection{1} A curse on Segulius from the gods, the worst man of all who are, who have been, or ever will be! What --- do you suppose he said it to you alone, or to Caesar only? He has passed up no one he could get talking with, to whom he has not retailed those very same things. Still, my Brutus, I love you as I should, for wanting me to know that bit of nonsense, whatever it was; you have given a great sign of your affection.

\ciceroSection{2} As for your saying that this same Segulius reports that the veterans are complaining because you and Caesar were not put on the board of ten --- I wish I were not on it myself! What could be more vexing? But all the same, since it had been my view that the motion ought to bring in those who held armies, those same persons who always do raised the cry against it; and so you were even excepted, with me fighting hard against it. For which reason let us pay Segulius no mind --- a man hunting after a fresh fortune, not because he has eaten through his old one (he never had one), but because he has devoured this very recent new one.

\ciceroSection{3} As for what you write --- that, what you do not do for your own sake, you do for mine, in that you are afraid for me on some count --- I free you, my Brutus, best of men and dearest to me, from every fear on my behalf; for in what can be guarded against I shall not be caught off guard, and over what does not admit of precaution I do not greatly trouble myself. I should be shameless indeed if I demanded more than can be granted to a man by the nature of things.

\ciceroSection{4} As for your instructing me to take care that by being afraid I do not force myself to fear the more, you instruct me wisely and like a true friend; but I would have you persuaded that, while it is agreed you excel in this kind of virtue --- that you never take fright, are never thrown into confusion --- I come closest after you in this excellence of yours. So I fear nothing, and I shall take care over everything. But see to it, my Brutus, that the fault will not now be yours, if I am afraid of anything. With your resources and with your consulship, even if we were timid men, we should still cast off all fear, especially when it has been impressed on everyone, and on me above all, that we are loved by you above any other. With your proposals --- about the four legions, and about the lands to be assigned by each of you --- I am in vigorous agreement. And so, when certain of our colleagues were nibbling after the land-assignment commission for themselves, I broke up the affair and kept the whole thing intact for us. If there is anything more confidential and, as you write, ``buried away,'' I shall send one of my own men, so that the letter may be carried to you the more faithfully. The day before the Nones of June.
