% Ad Familiares 12.25 (ad-familiares-12-25) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CORNIFICIO S.}

\ciceroSection{1} On the Liberalia I received your letter, which Cornificius delivered to me, by his own reckoning, on the twenty-second day. That day there was no Senate; nor on the next. On the Quinquatrus, before a crowded Senate, I argued your case, not against Minerva's will --- for that same day the Senate decreed that our Minerva, guardian of the city, whom a whirlwind had toppled, should be restored. Pansa read out your dispatch. Great approval of the Senate followed, with the highest satisfaction and to the discomfiture of the Minotaur --- that is, of Calvisius and Taurus. A complimentary decree was passed about you. It was demanded that those two should also be censured; but Pansa was more lenient.

\ciceroSection{2} My dear Cornificius, the day on which I first set foot upon the hope of liberty --- when the rest hung back and I, on the thirteenth day before the Kalends of January, laid the foundations of the state --- on that very day I took much thought, and made provision for your standing; for the Senate carried my motion on the holding of provinces. Nor have I since left off undermining the man who, with the highest injury to you and contumely to the state, was holding a province in absentia. He did not bear my frequent --- or rather daily --- assaults, but withdrew into the city against his will, and was dislodged not merely from hope but from the certain fact and possession by my just and honourable invective. That you have held your own dignity by your supreme courage, and have been honoured by the province with the highest distinctions, I rejoice exceedingly.

\ciceroSection{3} As for your clearing yourself to me about Sempronius, I accept the excuse: for that was a certain blind season of servitude. I, the prompter of your counsels and the supporter of your standing, in anger at the times was being swept off to Greece in despair of liberty; when the Etesian winds, like good citizens, refused to escort me as I deserted the state, and a south wind contrary to me, blowing hard, carried me back to your fellow-tribesmen at Regium; and from there with winds and oars I hurried home with every haste, and on the next day, amid the utter servitude of the rest, was alone a free man.

\ciceroSection{4} I so attacked Antonius that he could not endure it and poured out all his wine-soaked fury upon me alone, and now sought to draw me out for the sake of a massacre, now attempted me with ambushes; but I cast him, belching and retching, into the nets of Caesar Octavianus. For that excellent boy raised a force first for his own protection and ours, then for the protection of the state at large. If he had not existed, Antonius's return from Brundisium would have been the ruin of our country. What has been done since I take it you know. But let us return to where we turned aside. I accept your excuse about Sempronius: in such confusion you could have nothing settled. Now this day brings a different life, demands different ways, as Terence says. And so, my dear Quintus, embark with us --- and indeed to the stern. There is now one ship for all good men, and we are taking pains to keep it on a straight course; would that the run be prosperous! But whatever the winds prove to be, our craft will certainly not fail us. For what else can virtue answer for? See that you bear yourself with a great and lofty spirit, and reflect that all your dignity must be bound up with the state.
