% Ad Familiares 11.12 (ad-familiares-11-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. D. BRVTO IMP. COS. DES.}

\ciceroSection{1} On a single day I received three letters from you: one short, which you had given to Flaccus Volumnius; two fuller, of which the courier of T. Vibius brought one, and the other Lupus sent on to me. From your letter, and from what Graeceius reports, the war seems not only not put out but actually inflamed afresh. And I do not doubt, in light of your singular prudence, that you see this clearly: if Antonius gets back any solid footing, all those celebrated services of yours to the state will come to nothing. For so it had been reported at Rome, so it had been impressed upon everyone --- that Antonius had fled with a few men, unarmed, panic-stricken, broken in spirit.

\ciceroSection{2} If his condition is in fact such that, as I heard from Graeceius, one cannot engage with him without danger, then he does not seem to me to have fled from Mutina, but to have shifted the seat of his war. And so people have become other men: some indeed are complaining that you did not pursue; they reckon he could have been crushed if speed had been used. It is altogether characteristic of the people --- and of ours above all --- to abuse their liberty most particularly against the very man through whom they have obtained it. Still, one must take care that no just complaint can arise. The matter stands thus: he will have finished the war who crushes Antonius. What force this carries I would rather you weigh than that I write more openly.
