% Ad Familiares 12.29 (ad-familiares-12-29) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CORNIFICIO sal.}

\ciceroSection{1} Not only to you, to whom all my affairs are perfectly familiar, but to no one in the Roman people, I think, is the intimacy between me and L. Lamia unknown. It was witnessed by a great audience when he was banished by the consul A. Gabinius for having defended my safety freely and bravely. Nor was it from that moment that affection between us was born --- it was because the bond was old and great that he hesitated at no danger on my behalf. To these services, or rather to these acts of merit, is added the most delightful intimacy of daily intercourse, so that there is absolutely no man in whom I take more pleasure. I don't suppose you are still waiting to hear in what words I commend him to you: you understand the cause of such great affection, and what words it would call for.

\ciceroSection{2} Take it that I have employed them all. I would only have you reckon this: that if you defend Lamia's affairs, his agents, his freedmen, his household in whatever matters need it, I shall feel more gratitude than if your liberality had been extended to my own private estate. And I do not doubt that, even without my commending him, given your sound judgement of men, you will eagerly do everything for Lamia's own sake. Although it had been reported to me that you supposed Lamia to have been present in the drafting of some senatorial decree that ran against your honour: but he certainly never came to drafting under those consuls; and besides, in those days every kind of falsified decree of the Senate was being put about. Unless perhaps you suppose I too was present at that Sempronian decree --- when I was not even at Rome, as I wrote to you about at the time, while the thing was still fresh.

\ciceroSection{3} But enough of this. My dear Cornificius, I ask you again and again to consider all Lamia's affairs as my own, and to take pains that he should understand that this commendation has been of the greatest service to him. You can do nothing more agreeable to me than this. Take care of your health.
