% Ad Familiares 11.18 (ad-familiares-11-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. D. BRVTO IMP. COS. DESIG.}

\ciceroSection{1} Although from the instructions you gave Galba and Volumnius for the Senate we could guess what it was that you thought there was to fear, still those instructions seemed more timid than was worthy of your victory and the Roman people's. The Senate, my dear Brutus, is firm, and has firm men to lead it. And so it bore it ill that you, whom it judged the bravest man of any who had ever lived, should be judging it timid and lacking in spirit.

\ciceroSection{2} For when, while you were shut in and Antony was at his height, everyone had placed their greatest hope in your courage --- who is there, now that he is shattered and you set free, who should fear anything? Nor in fact did we fear Lepidus. For who could there be who would think him so frenzied that the man who in the greatest of wars had said he wanted peace should, in the most longed-for peace, declare war on the state?

\ciceroSection{3} I do not doubt that you see further than this. But still --- with the thanksgivings so freshly performed at every temple of the gods, in your name and at all of them --- the renewal of fear was bringing us great distress. So my own wish is what I hope for: that Antony lie plainly broken and crushed. But should he by any chance gather some strength again, he will find that the Senate does not lack counsel, nor the Roman people courage, nor the state, while you live, a commander.
\textit{[1 June.]}
