% Ad Familiares 10.8 (ad-familiares-10-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{PLANCVS IMR COS. DESIG. S. D. COS. PR. TR. PL. SENATVI POPVLO PLEBIQVE ROMANAE}

\ciceroSection{1} If to anyone I appear to have held the expectation of mankind and the hope of the state in suspense about my intentions longer than I should, I think I must first excuse myself to him before promising anything to anyone about the duty that is to come; for I do not wish to seem to have redeemed past blame but rather to be giving voice now, at the ripe moment, to thoughts of the soundest mind long since formed.

\ciceroSection{2} It did not escape me, in so great an anxiety of mankind and so disturbed a state of the commonwealth, that to profess good will brings the richest return, and that many had won great honours from such a profession; but since fortune had brought me down to this pass, that I should either, by promising quickly, throw up serious obstacles in the way of my own action, or, if I held back, have larger occasions for being of help, I chose for my path the easier road of the common safety rather than that of my own renown. For who, in the position that is mine, and coming from the kind of life which I think men know in me, and with the prospect that I hold in hand, can either submit to anything dishonourable or desire anything pernicious?

\ciceroSection{3} But a fair stretch of time, and great toils, and many expenses were needed for me to discharge in fact what we kept promising to the state and to all loyal men, and to come to my country's aid not naked save for our good will but provided with the means as well. The army, which had often been tempted with great bribes, had to be steadied so that it should expect from the state a measured return rather than from one man an unlimited one; several communities, which the year before had been bound by largesses and grants of rewards, had to be brought to count those promises empty and to think that the same things ought to be sought from better authorities; the inclinations also of those who had been in charge of the neighbouring provinces and armies had to be drawn out, so that we should form a partnership with a larger number for the defence of liberty, rather than divide with a smaller number a victory disastrous to the whole world.

\ciceroSection{4} We had to fortify ourselves as well, by enlarging the army and multiplying our auxiliary forces, so that, when we openly set forth what we felt, then even against the will of certain parties it should not be a perilous thing for it to be known what we were going to defend. So I shall never deny that, in order to bring these designs to effect, I both unwillingly feigned much and with pain dissembled much, because I had seen, from my colleague's case, how dangerous to a good citizen unprepared is the premature unveiling of his hand.

\ciceroSection{5} For this very reason I gave C. Furnius, my legate --- a man of courage and energy --- more commissions by word of mouth than in writing, so that they should come to you the more secretly and that we should be the safer; and I gave him directions on what arrangements should be used both to secure the common safety and to arm ourselves. From which it can be understood that the care of defending the highest interests of the commonwealth has long been on watch with us.

\ciceroSection{6} Now, by the goodness of the gods, since on every point we are better prepared, we wish men not only to entertain good hope of us but to judge us upon evidence. I have five legions under the standards, bound most closely to the state by their own loyalty and courage, and obedient to us through our generosity; the province most ready, by the consent of all its communities, vying in the fullest exertion of duty; cavalry and auxiliaries in such numbers as these peoples can put together for the defence of their own safety and freedom; and I myself am so prepared in spirit that, whether the call is to hold the province, or to march wherever the state may summon me, or to hand over army, auxiliaries, and province, or to draw down the whole assault of the war upon my own head, I shall not refuse --- if only by my fate I may either secure the safety of the fatherland or buy it time of danger.

\ciceroSection{7} If I am promising these things when all is already in train and the state stands tranquil, then I shall be glad to lose any praise of my own where the gain is the state's; but if I am stepping forward to share in the most untainted and the greatest perils, then I commend my counsels to fair judges, to be defended against the slander of the envious. For my own part, I have the reward of my deserts ready enough in the safety of the commonwealth; but as for those who, following my authority and, far more, your loyalty, could be neither deceived by any hope nor terrified by any fear, I think you ought to be asked to keep them under your protection.
