% Ad M. Brutum 2.2 (ad-brutum-02-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} Plancus's outstanding feeling for the commonwealth, his legions, auxiliaries, and forces, you could see from his letter, of which I expect a copy has been sent to you. As for Lepidus, your kinsman --- who next to his brother counts as his nearest relations the men he most hates --- his lightness and inconstancy, and a mind always inimical to the commonwealth, are by now, I believe, plain to you from your own friends' letters.

\ciceroSection{2} What troubles us is expectation, †which† has by now been brought wholly to its last crisis. For all our hope lies in extricating Brutus, on whose account we were strongly afraid.

\ciceroSection{3} Here I have my hands full with that madman Servilius, whom I endured longer than my dignity allowed --- but I endured him for the commonwealth's sake, so as not to give to the desperate citizens a man, indeed scarcely sane, but for all that of noble blood, to whom they might rally; which they do all the same. But still I did not think him to be alienated from the commonwealth. I have made an end of bearing him. For he had begun to be of such insolence as to count no man free. In the case of Plancus, however, he flared up with extraordinary indignation, contended with me for two days, and was so broken by me that I hope he will be more moderate for ever after. And in this very contest, with the matter at its height, on the fifth day before the Ides of April a letter was delivered to me in the Senate from our friend Lentulus, about Cassius, about the legions, about Syria. As soon as I had read it out, Servilius collapsed, and several others besides --- for there are some men of mark, more than a few, whose convictions are wholly base. But Servilius took it most bitterly that the Senate had voted with me about Plancus. It is a great prodigy in the commonwealth, but in respect of which * * *.
