% Ad Familiares 10.11 (ad-familiares-10-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{PLANCUS CICERONI}

\ciceroSection{1} Undying thanks I give you, and shall give you while I live --- for that I can repay them I cannot affirm. To such great services as yours I do not seem able to make any answering return, unless --- as you yourself, in your most weighty and eloquent way, have written --- you mean to feel it so, and judge that I am repaying the debt when I shall hold the memory of it. Had it been your own son's standing that was at issue, you could surely have done nothing more lovingly. Your first proposals, with their unlimited gifts; the later ones, framed to the moment and to the judgement of my friends; your unfailing and uninterrupted speech on my behalf; your celebrated quarrels with my detractors on my account --- all this is fully known to me. No ordinary care must I take to show myself in public life a citizen worthy of your praises, and in our friendship mindful and grateful. For what remains, look after your own creation; and if you find me in the issue, in deed, the man you wanted me to be, then defend me and stand by me.

\ciceroSection{2} When I had carried the whole force across the Rhone and sent my brother on ahead with three thousand horse, and was myself directing my march toward Mutina, I learned on the road of the battle fought and of Brutus and Mutina freed from siege. I observed that Antony and the remnants who were with him had no other line of retreat than these parts of the country, and that two hopes were held out to him --- one, in Lepidus himself; the other, in his army. Because a certain part of that army is no less frenzied than the men who were with Antony, I have recalled my cavalry, and have myself taken station among the Allobroges, so that I may be prepared accordingly for every contingency as events advise me. If Antony flings himself this way stripped bare, I think I can readily hold him in check by my own resources and administer the state in accordance with your wishes, even though he be received by Lepidus's army; but if he brings any forces of consequence with him, and if the veteran Tenth Legion, which by my exertions was recalled along with the rest, has gone back to the same frenzy, still I shall take all pains that no damage be done, and this I trust I shall make good --- provided your forces are got across from your side and, joined with mine, can the more easily crush the desperate men.

\ciceroSection{3} This I pledge to you, my dear Cicero: that neither spirit nor diligence shall be wanting in me. I could wish, by Hercules, that no anxiety remained at all; but if any remains, I shall yield to no man's spirit, no man's good will, no man's endurance, where you are concerned. I am doing my part, indeed, to draw Lepidus too into the partnership of this enterprise; and I promise him every form of compliance, if only he is willing to look to the state. I am using as my helpers and go-betweens in this my brother and Laterensis and our friend Furnius. Private offences shall not stand in my way: for the safety of the state I will act in concert even with my bitterest enemy. And if I gain nothing by it, I shall still satisfy you, none the less, with the greatest spirit and perhaps with the greater glory to myself. See that you keep well, and that you keep loving me in return.
