% Ad Familiares 10.17 (ad-familiares-10-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{PLANCVS CICERONI}

\ciceroSection{1} Antony reached Forum Iulii with his leading troops on the Ides of May. Ventidius is two days' march behind him. Lepidus has his camp at Forum Voconii, a place twenty-four miles from Forum Iulii, and there he has resolved to wait for me --- so he has written to me himself. If he and fortune together keep everything intact for me, I undertake to you that I shall bring the business to a successful conclusion swiftly.

\ciceroSection{2} I wrote to you earlier that my brother, worn out by unbroken labours and rushings to and fro, had been seriously ill. But for all that, the moment he could begin to walk again, judging that he had recovered no more for his own sake than for the state's, he refused to step back from any first place among the dangers. I myself, however, not only urged but compelled him to set out for Rome, because at that state of health he could exhaust himself more than he could help me in camp, and because, after the terrible loss of the consuls, I judged that the commonwealth, stripped bare, required such a citizen as praetor for civic duties. If any of you do not approve, let them know that the want of foresight in counsel was mine, not the want of loyalty to the fatherland in him.

\ciceroSection{3} Lepidus, however, did what I had been asking: he sent Apella to me, so that I should hold him as a hostage for his good faith and for our partnership in administering the affairs of the state. In this matter L. Gellius gave me proof of his zeal \dag{}from the three brothers Segavianus,\dag{} whom I used most recently as my intermediary with Lepidus. I think I have come to know him as a friend to the commonwealth, and I gladly bear witness to him, and I shall bear witness for all who deserve well of her. Keep well, return my affection, and, if I deserve it, guard my standing --- as up to now you have done with a singular goodwill.
