% Ad M. Brutum 1.18 (ad-brutum-01-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} Although I had often urged you in letters to come to the rescue of the state as soon as you could, and to lead your army into Italy --- and I judged that your own intimates did not doubt it --- I was asked, by a most prudent and most attentive woman, your mother, all of whose cares turn back on you and are spent on you, to come to her on 25 July. This I did, as I was bound to do, without delay. When I had arrived, Casca was there, and Labeo and Scaptius. She brought up the matter and asked what I thought: were we to summon you, and did I judge it the right course for you, or would it be better that you should hold back and stay where you are?

\ciceroSection{2} I gave the answer I felt: that it would be most conducive both to your dignity and to your standing if you brought help at the earliest possible moment to a state that is slipping and almost on the point of toppling. For what evil do you suppose is absent from a war in which the victorious armies refused to pursue the fleeing enemy, and in which a commander who came through it whole --- adorned with the highest honours and greatest fortunes, with a wife, with children, with you for in-laws --- has declared war on the state? What am I to say in the face of so great an accord of Senate and people, when so great an evil sits within the walls?

\ciceroSection{3} The greatest grief I was being afflicted with, as I was writing this, was that, having been accepted by the state as a guarantor for a youth, indeed practically a boy, I seemed scarcely able to make good what I had promised. For an obligation of mind and judgement is heavier and more difficult --- particularly in matters of the highest weight, on another's behalf --- than an obligation of money. The latter can be paid off, and the loss to one's estate is bearable; what one has pledged to the state, how can one discharge it, unless the man on whose behalf one pledged readily allows the payment?

\ciceroSection{4} Yet I shall, I hope, hold on to him too, against the resistance of many. For he seems to have natural ability; but his is a pliant age, and there are many ready to corrupt him --- who are confident that, with the false glitter of a sham honour cast in his way, the edge of a good intellect can be dulled. So this labour has been added to all the rest: to bring every device to bear to keep hold of the young man, that I may not face the reputation of rashness. And yet --- what rashness is there? For I have placed the obligation on him for whom I went surety rather than on myself. Nor can the state regret that I went surety for him, who in carrying things through showed himself, both by his own natural ability and by my pledge, the more consistent man.

\ciceroSection{5} But the greatest knot in the state, unless I am much mistaken, is the want of money. For day by day the good men's ears grow harder to the word \textit{tributum}; and what has been raised at the rate of one per cent on the shameless valuations of the rich is all consumed in bonuses for two legions. Boundless expenses now hang over us, both for the armies by which we are now defended and, with even more force, for yours. Cassius too, our Cassius, seems likely to come well enough equipped. But these matters, and many besides, I long to speak of in person --- and as soon as possible.

\ciceroSection{6} About your sister's sons, Brutus, I have not waited for you to write. The very crisis (for the war will be drawn out) leaves you with the case open and intact. But from the start, since I could not divine how long the war would last, I argued the boys' case in the Senate in the way you can, I imagine, have learned from their mother's letters. And there will never be any matter in which I shall not say and do, even at risk of my life, what I shall judge to be your wish and to bear on your interest. 27 July.
