% Ad M. Brutum 1.4 (ad-brutum-01-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Brutus Ciceroni salutem}

\ciceroSection{1} How great a joy I felt on learning of the doings of our Brutus and of the consuls, it is easier for you to imagine than for me to write. Among the other things I praise and rejoice that they have come to pass, this above all: that Brutus's sortie was not only a saving stroke for himself, but also the greatest help towards the victory.

\ciceroSection{2} When you write that, as far as I am concerned, the three Antonii have one and the same cause, I think the judgement is my own to make. I lay down nothing except this: that the Senate or the Roman people must judge those citizens who have not perished in arms. But this very position --- you will say --- is unjust on my part, who give the name of citizens to men of hostile mind against the state. On the contrary: it is the most just position possible. For what the Senate has not yet decreed, nor the Roman people commanded, I do not arrogantly prejudge nor refer to my own discretion. As for this, I shall not change it: that from the man whom necessity has not compelled me to kill, I have neither stripped anything cruelly, nor remitted anything dissolutely, and I have kept him in my power as long as there was war. Far more honourable, in my view, and easier for the state to allow, is not to harry the fortune of wretched men, than to grant unlimited concessions to the powerful, of the sort that can kindle their greed and arrogance.

\ciceroSection{3} On this point, Cicero --- best and bravest of men, and most dear to me both for your own sake and on behalf of the state --- you seem to me to put too much trust in your own hope, and, the moment anyone has acted rightly in some particular, to grant and entrust him everything, as though a mind once corrupted by lavish gifts could not be drawn over to bad counsels. Such is your humanity, you will allow yourself to be advised with an even temper, especially on the common safety; yet you will do as seems best to you. And I too, when you have set me right * *
