% Ad Familiares 10.10 (ad-familiares-10-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO PLANCO.}

\ciceroSection{1} Although I had learned well enough from our friend Furnius what your will was and what your plan for the state, still on reading your own letter I judged the whole of your mind more clearly. For which reason, although the entire fortune of the state is being decided in a single battle --- and by the time you are reading this, I take it the issue will already have been settled --- still by the mere report which has gone abroad of your loyal intention you have won a great commendation. And so, if we had a consul at Rome, the Senate would have declared, with high honours for you, how welcome your effort and your preparation were. The moment for that has not passed --- nor, in my judgement at any rate, has it even yet arrived. For what I am accustomed to consider true honour is the honour conferred and given to distinguished men not from any hope of future benefit but for the magnitude of services already rendered.

\ciceroSection{2} Therefore --- let there only be some commonwealth left in which honour can shine --- in honours of the very fullest, believe me, you will be steeped in abundance. The kind, however, that can truly be called honour is not an inducement to suit the hour but the prize of unbroken virtue. For which reason, my dear Plancus, lean with the whole of your heart toward glory, come to your country's rescue, stand by your colleague, support the consensus of every nation and the extraordinary spirit that has rallied to it. You shall find me the supporter of your counsels, the favourer of your standing, in every matter the most loving and the most loyal of friends. To the causes by which we are joined --- affection, mutual offices, long acquaintance --- the love of country has now been added; and it is this that has brought me to set your life before my own. The third day before the Kalends of April.
