% Ad Familiares 10.25 (ad-familiares-10-25) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO S. D. FVRNIO.}

\ciceroSection{1} If it matters to the state --- as men suppose it does --- that you continue to throw yourself, as you have begun and as you have been doing, into vigorous service, and that you have a hand in the very great undertakings that bear on extinguishing what is left of the war, then you can do nothing, in my judgement, better, more deserving of praise, or more honourable; and I count that service of yours, that energy of yours, that public spirit of yours, as a thing to be set ahead of haste over the praetorship. For I would not have you ignorant of how great the credit is that you have won: believe me, it is the credit next after Plancus's, and that on Plancus's own testimony --- and besides on the report and the recognition of everyone.

\ciceroSection{2} For which reason, if any work yet remains for you, I judge that it should be carried through with the utmost diligence; for what is more honourable, or what is to be set ahead of the honourable? But if you think enough has been done for the state, then I judge that you should come quickly to the elections, since they are coming early --- only on condition that this ambitious haste of yours not in any way diminish that glory which we have won. Many illustrious men, while they were giving their service to the state, did not stand in the year of their candidacy. And that is the easier in our case, because this year is not the one set down for you, in the sense that, had you been aedile, your year would come two years after. As it is, you will not be seen to have let slip the customary and, as it were, statutory time for standing. And I see that, with Plancus as consul --- even though without him you would have your prospects unencumbered --- your candidacy will be the more brilliant nevertheless, provided only that those affairs of yours have been brought to the conclusion you wish.

\ciceroSection{3} On the whole, with judgement and discernment as great as yours, I did not think it so necessary to write at greater length; but I was unwilling, nonetheless, that my own view should be unknown to you --- and the sum of it is this: I would rather you measured everything by your standing than by ambition, and set the greater profit in lasting credit than in haste over the praetorship. I said these same things at my own house with my brother Quintus called in, and Caecina and Calvisius, who are the most devoted of your friends, while Dardanus your freedman was present. My speech seemed to win the approval of all; but you yourself will judge best.
