% Ad Familiares 10.15 (ad-familiares-10-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{PLANCVS CICERONI}

\ciceroSection{1} After my last letter was written, I thought it concerned the state that you should know what has happened since. My diligence, I hope, has borne fruit for me and for the state both. For by means of continual messengers I urged on Lepidus that, dropping all contention and with our mutual goodwill restored, he should come, in common counsel, to the rescue of the republic; that he should value himself, his children, and the city more highly than one ruined and abandoned brigand; and that, if he did so, he should make full use of my compliance for every purpose. I succeeded.

\ciceroSection{2} And so, through Laterensis as intermediary, he gave me his word that, if he could not keep Antonius out of his province, he would prosecute him by war; and he asked me to come and to join my forces with his, all the more because Antonius was reported to be strong in cavalry and Lepidus had not even a moderate cavalry of his own. Indeed, out of that small force not many days before, ten of the best men had come over to me. When I learned of this, I did not hesitate: while he was on a sound course, I judged Lepidus must be supported.

\ciceroSection{3} I saw what my arrival could achieve: that I could pursue and crush his cavalry with my own, and that, by the presence of my army, I could both correct and curb that part of Lepidus's army which has been corrupted and alienated from the republic. And so, having thrown a bridge in a single day across the Isère, the largest river in the territory of the Allobroges, I led the army across on 12 May. When the news reached me, however, that Lucius Antonius had been sent ahead with cavalry and cohorts and had reached Forum Iulii, I sent my brother on 13 May with four thousand horse to meet him; I myself shall follow by forced marches with four legions in light order and the rest of the cavalry.

\ciceroSection{4} If only moderate fortune assists us in the state's cause, here we shall find the end both of the desperate men's audacity and of our own anxiety. But if the brigand, getting wind of our advance, should begin to fall back into Italy, it will be Brutus's office to meet him --- and Brutus, I know, will lack neither plan nor spirit. Even so, if that happens, I will send my brother with the cavalry to follow him and defend Italy from devastation. Stay well, and love me as I love you.
