% Ad Familiares 10.23 (ad-familiares-10-23) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{PLANCVS CICERONI}

\ciceroSection{1} By Hercules, my Cicero, I shall never repent of having faced the gravest dangers for the sake of my country, provided that, if anything befalls me, I am clear of the reproach of recklessness. I should be admitting that I had slipped through want of foresight, if I had ever trusted Lepidus from the heart; for credulity is an error rather than a fault, and indeed it creeps with the greatest ease into the mind of the best men. But it was not by this failing that I was nearly cheated --- I knew Lepidus through and through. What is the truth, then? Shame, which is the most dangerous thing in a war, forced me into the position I am now in. For if I had stood in one place, I was afraid that I might appear to some of those looking for grounds to carp both excessively obstinate in my offence against Lepidus, and even, by my own passivity, to be feeding the war.

\ciceroSection{2} So I brought my forces almost into sight of Lepidus and Antony, and, leaving a gap of forty miles, took up a position with this design: to be able either to close quickly or to fall back to safety. I added these features in my choice of ground: that I had a river in my front, in which any crossing would mean delay; and that the Vocontii were close at hand, through whose territory the road lay open to me in good faith. Lepidus, his hope of my arrival gone --- and he had been not a little keen for it --- joined himself to Antony on 29 May, and on the same day they moved camp against me. When they were twenty miles off, the news was brought to me.

\ciceroSection{3} By the goodness of the gods I contrived both to fall back quickly and to make this withdrawal in no way resemble a flight: not a single foot-soldier, not a horseman, not a piece of baggage was lost or cut off by those boiling brigands. So on 4 June I led all my forces over the Isara and broke up the bridges I had built, so that my men should have time to pull themselves together and I meanwhile could join up with my colleague --- whom I was expecting in three days as I sent off this letter.

\ciceroSection{4} As for our friend Laterensis, I shall always confess to his exceptional loyalty and patriotism. But it is certain that his excessive indulgence toward Lepidus made him less keen-sighted in seeing these dangers. When he saw himself led into the trap, the hands he should more justly have armed against Lepidus's destruction he turned upon himself; in the act, however, he was interrupted, and he is still alive and is said to be going to live; but on this point I have nothing yet for certain.

\ciceroSection{5} To the great vexation of the parricides I slipped through their fingers; for they were coming against me in the same fury with which they had set themselves against the fatherland, and they carried fresh grudges of their own --- because I had not ceased to lecture Lepidus into putting out the war; because I disapproved of the negotiations going on; because I had forbidden the envoys sent by Lepidus's good faith to come into my sight; because C. Catius Vestinus, a military tribune sent by Antony to him with a letter, I had intercepted and treated as an enemy. In this I take this much pleasure: that the harder they came after me, certainly, the sharper the pain their frustration brought them.

\ciceroSection{6} You, my Cicero, hold steady on what you have done so far --- back us up vigorously and resolutely, who stand in the line. Let Caesar come with the forces he has, which are very strong; or, if something is holding him back personally, let his army be sent. He himself is in great danger as well. Whatever was ever going to be marshalled in the camp of desperate men against the fatherland has now come together; for the actual salvation of the city, then, why should we not use every resource we have? But if you on your side do not fall short, surely, as far as I am concerned, I shall in every respect more than satisfy the state.

\ciceroSection{7} As for you, my Cicero, by Hercules I hold you dearer every day, and your kindnesses sharpen my anxieties more from one day to the next, lest I lose any part of either your affection or your good opinion. My wish is that I may be allowed, by a now actual repayment of devotion in my services, to make your kindnesses still more agreeable to you. 6 June, from Cularo in the territory of the Allobroges.
