% Ad Familiares 12.4 (ad-familiares-12-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CASSIO S.}

\ciceroSection{1} I wish you had invited me to dinner on the Ides of March: there would have been no leftovers. As it is, your leftovers are giving me work --- and giving me more work, by Hercules, than anyone else. Excellent consuls, to be sure, we have; but the consulars are utterly disgraceful. The Senate is steady, but it is precisely the lowest-ranking who are steadiest. As for the people, nothing is steadier, nothing better, and the same goes for all Italy. But nothing is more contemptible than the envoys Philippus and Piso, nothing more scandalous. Sent to deliver specific terms to Antonius on the Senate's authority, when he obeyed none of them they went so far as to bring back to us intolerable demands from him. So people are flocking to me, and in a cause that saves the state I have now turned popular.

\ciceroSection{2} But what you were doing, what you were going to do, where indeed you were --- I had no idea. Rumour reported you were in Syria; no one was vouching for it. About Brutus, the closer he is, the firmer the reports seem. Dolabella was being heavily abused by men of some wit, on the ground that he was so quick to succeed you, when you had scarcely been thirty days in Syria. So it was the settled view that he ought not to be received into Syria. The highest credit goes both to you and to Brutus, that you are reckoned to have raised an army beyond hope. I would write more if I knew the situation and the substance of it; as it is, what I write I write from common opinion and rumour. I wait eagerly for your letter. Farewell.
