% Ad Familiares 12.13 (ad-familiares-12-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{C. CASSIVS Q. S. D. M. CICERONI}

\ciceroSection{1} If you are well, it is well. We rejoice in the Republic's safety, or rather its victory, and at the same time in the renewal of your own glory --- that the greatest consular has surpassed his own greatest consulship --- and we cannot wonder enough at it. Some fated quality, I know not what, has been given to your excellence, as we have already found out often. For your toga has been more fortunate than all men's arms: even now it has snatched the Republic, almost beaten, out of the enemy's hands and given her back to us. So now we shall live free; now we shall have you, greatest of all citizens and dearest of all to me, as the witness of our affection --- as you came to know it in the deepest darkness of the state --- now we shall have you as witness of our affection both toward you yourself and toward the Republic so closely joined to you. And those things which you have often promised --- that you would keep silent so long as we were in slavery, but speak about me at the moment when it would benefit me --- I do not so much desire those things to be said, as to be felt by you yourself. Nor would I rather be recommended to the judgement of all men by you than to have been worthily and as I deserve recommended in your own. So judge that these latest acts of ours have been not sudden, nor matters of convenience, but akin to those purposes of which you yourself are witness; and reckon that I myself am to be brought forward by you no less than the country's best hope.

\ciceroSection{2} You have, M. Tullius, children and kinsmen worthy of you and rightly dearest to you; in the Republic next to them ought to come those who emulate your studies --- of whom I wish you abundance; yet I do not think that I am shut out by so great a crowd that you cannot find time to take me in and bring me forward into every course you wish for and approve. My disposition I have perhaps proved to you; my talent, such as it is, the long servitude has surely allowed to appear less than it was.

\ciceroSection{3} From the coastal strip of the province of Asia and from such of the islands as we could, we drew off ships; the levy of rowers, in the face of great obstinacy from the communities, we still got through fast enough. We followed the fleet of Dolabella, commanded by L. Figulus, who, by repeatedly offering hope of defection and never ceasing to draw back, at last betook himself to Corycus and began to keep himself there with the harbour shut. We left that and made for Cyprus, since we thought it more profitable to get to camp, and besides another fleet was following --- the one which Tillius Cimber had got together in Bithynia the year before, with the quaestor Turullius in command. What I have learned there, I wished to write to you as quickly as I could.

\ciceroSection{4} Dolabella was summoned --- as the people of Tarsus, basest of allies, summoned him, so the people of Laodicea, even more out of their senses, brought him in of their own accord; from these two cities he made up, by the number of Greek troops, the appearance of an army. He has pitched camp before the town of Laodicea, has knocked down a section of the wall, and joined his camp to the town. Our Cassius, with ten legions and twenty cohorts of auxiliaries and a cavalry force of four thousand, has pitched camp at a javelin's cast away \textit{[Greek: paltōi]}, and reckons that he can win without a battle: for already in Dolabella's camp wheat is at three tetrachms a measure. Unless he gets some brought in by Laodicean ships, he must quickly perish of famine; and that he should not be able to get it brought in, both Cassius's fleet --- a good large one, commanded by Sextilius Rufus --- and the three ships we have brought up (I, Turullius, Patiscus) will easily ensure. I want you to be of good hope and to be confident that the Republic, as you on your side have extricated it, can on our side too be extricated quickly. Farewell. Despatched on the Ides of June, from Cyprus at Crommyuacris.
