% Ad Familiares 10.12 (ad-familiares-10-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO PLANCO.}

\ciceroSection{1} Although for the state's sake above all I ought to rejoice that you have brought it so great a protection, so great a relief at what was very nearly the last hour, still my embrace of you, the victor, when the state is recovered will be the warmer because a great share of my joy is owed to your own standing --- a standing that I see is already, and shall be, of the very highest. For do not imagine that any letter was ever read out in the Senate with greater satisfaction than yours; and this came about both through some extraordinary scale in your services to the state and through the weight of your wording and of your sentiments. To me there was nothing in this that was new --- for I knew you, and I remembered the promises in your letters sent to me, and I had your plans deeply known to me through our friend Furnius --- but to the Senate the matter looked greater than what had been expected. Not that the Senate had ever doubted your good will: it simply did not have it sufficiently ascertained how much you were able to do, or how far you meant to go.

\ciceroSection{2} And so when, on the seventh day before the Ides of April, in the morning, M. Varisidius had delivered your letter to me, and I had read it, I was lifted up by an unbelievable joy; and when a great crowd of the best men and citizens was escorting me from my house, I straightway made all of them sharers in my delight. Meanwhile our friend Munatius came to see me, as was his habit. And I to him with your letter --- for he knew nothing yet: it was to me that Varisidius had come first, and this, he said, was what you had charged him to do. A little after, the same Munatius gave me to read the letter you had sent to him, and the one you had sent for public reading.

\ciceroSection{3} We agreed to carry the letter at once to Cornutus, the urban praetor, who, since the consuls were away, was discharging the consul's office in the manner of the ancestors. The Senate was summoned at once, and convened in great numbers, on the strength of the report and the expectation of your letter. When the letter had been read aloud, a religious objection was urged against Cornutus, on the warning of the chicken-keepers, that he had not attended carefully enough to the auspices --- and this was approved by our college. And so the matter was put off to the following day. On that day there was a great struggle for me, on behalf of your standing, with Servilius. He, having by his influence got it arranged that his proposal should be put to the vote first, was abandoned by a packed Senate, which divided into the opposing lobby; and when the Senate was assenting in large numbers to my proposal --- which was put second --- at Servilius's request P. Titius interposed his veto. The matter was put off to the following day.

\ciceroSection{4} Servilius came primed for the fight, with Jupiter himself for an enemy --- in whose temple the business was being conducted. How I broke him, and with what struggle I dislodged the tribune Titius from his veto, I had rather you learned from the letters of others. One thing only from mine: the Senate could not have been graver, steadier, friendlier to your praises than it was that day --- nor, indeed, the Senate friendlier to you than the whole community of the city was; for in a wonderful way the entire Roman people, and the consensus of every class and order, were united in the work of liberating the state.

\ciceroSection{5} Press on, then, as you are pressing, and commend your name to immortality, and despise all these things which have the look of glory got together out of the most empty trappings of show: count them short-lived, painted, fragile. True honour is set in virtue, which is most brilliantly displayed by great services to the state. That capacity you have in the highest degree. And since you have embraced it and hold it, see to it that the state shall owe you no less than you owe the state. You will find me not only the supporter of your standing but the magnifier of it. This I judge that I owe both to the state --- which is dearer to me than my own life --- and to the bond between us. And amid these cares which I have bent to your standing I have taken a great pleasure: that the prudence and loyalty of T. Munatius, already well known to me, I have come to see even more clearly in his extraordinary good will and diligence on your behalf. The third day before the Ides of April.
