% Ad Familiares 12.28 (ad-familiares-12-28) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CORNIFICIO S.}

\ciceroSection{1} I agree with you that those men whom you mention as threatening from Lilybaeum ought to have paid their penalty on the spot; but, as you say, you were afraid you would seem too free-handed in punishing. Afraid, then, of seeming too weighty a citizen, too courageous, too true to yourself.

\ciceroSection{2} That you renew the partnership in preserving the state which you took up with me from your father --- that is welcome; and that partnership, my dear Cornificius, will always stand between us. Welcome, too, that you think no thanks need be returned to me in your name; for indeed we should not do this between ourselves. The Senate would be appealed to more often on behalf of your standing, if it were ever convened while the consuls are absent except for some new matter. And so neither about the two million sesterces nor about the seven hundred thousand can anything now be transacted through the Senate. I rather think you should requisition supplies under the senatorial decree, or borrow on credit. As for what is happening in public affairs, I take it you learn it from the letters of those whose duty it is to write the proceedings out to you in full. I am in good hope, and I do not fail in counsel, in care, or in toil; toward every enemy of the Republic I openly profess myself the bitterest foe. The case, as it now stands, does not seem to me to be in a difficult place, and would have been in the easiest of places, had the fault of certain men been absent.
