% Ad M. Brutum 1.1 (ad-brutum-01-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} L. Clodius, tribune of the plebs designate, is very fond of me --- or, to put it \textit{[Greek: emphatikoteron]} (more emphatically), he loves me dearly. Since I am quite sure of this, I have no doubt (for you know me well) that you judge him to be loved by me too. It seems to me utterly less than human not to answer in kind, in love, those by whom one is challenged to it. He gave me the impression of suspecting --- and with no small pain --- that something had been reported to you by his enemies, or rather through his enemies, with the aim of turning your goodwill away from him. I am not in the habit, my Brutus --- as I think you know --- of making rash claims about another, since it is a hazardous thing on account of the hidden inclinations of men and the many-sidedness of their natures; but Clodius's mind I have inspected, come to know, and judged. I have many proofs of his disposition, but it is not necessary to set them down in writing. For I want this to seem to you a sworn deposition rather than a letter. He has been advanced by a kindness of Antony's. Of that kindness itself, a large share comes from you. And so he would wish him preserved on terms that leave us preserved as well.

\ciceroSection{2} But he understands the matter has been brought to such a point (for, as you know, he is anything but a fool) that the two cannot both come safely through. And so he prefers us; and where you are concerned, his words and his sentiments alike are most friendly. Therefore, if anyone has written to you about him in any other terms, or has spoken so face to face, I beg you again and again to believe me rather than him --- both because I can judge more readily than that nameless someone, and because I love you more. Consider Clodius your warmest friend, and a citizen of the sort that a man of the highest prudence and the most favourable fortune ought to be.
