% Ad Familiares 12.24 (ad-familiares-12-24) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CORNIFICIO S.}

\ciceroSection{1} I let pass no occasion --- as indeed I should not --- not only of praising you but also of advancing your honor; but I would rather have my zeal toward you and my services known to you from your own people's letters than from mine. I urge you all the same to bend every care to the commonwealth. This is the part of your spirit, this of your talent, this of the hope you ought to entertain of enlarging your standing.

\ciceroSection{2} But more of this another time. For as I write, everything stands in suspense; the envoys had not yet returned, whom the Senate sent not to sue for peace but to declare war if their message were not obeyed. I, however, the moment opportunity offered, on my old manner have defended the commonwealth, and have professed myself the leading man for the Senate and the Roman people; nor, since taking up the cause of liberty, have I lost a moment in defending the common safety and freedom. But of this too I prefer that you hear from others.

\ciceroSection{3} T. Pinarius, my very close friend, I commend to you with such zeal that I could not commend him more. With him I am on the most intimate terms, both for all his virtues and for shared studies. He looks after the accounts and affairs of our friend Dionysius, whom you too love much and I more than anyone. These I am under no obligation to commend to you, but commend them I do, nonetheless. You will see to it, then, that from Pinarius's letters --- a most welcome man --- we may make out your zeal both toward him and toward Dionysius.
