% Ad Familiares 10.24 (ad-familiares-10-24) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{PLANCVS IMR COS. DES. S. D. CICERONI}

\ciceroSection{1} I cannot help thanking you individually for each of your services and kindnesses --- but by Hercules I do it with embarrassment. For neither does a tie so great as you have wished there to be between us seem to call for a formal vote of thanks, nor do I willingly discharge what I owe for your greatest benefits with so cheap an offering of speech; I had rather, present in person, by attention, indulgence, and constancy, give you proof that I remember. If life is granted me, in attentiveness, indulgence, and constancy I shall outdo all gracious friendships, and even the dutiful ties of family, in my devotion to you; for whether your affection and good opinion of me is to bring me more standing for ever or more pleasure every day, I could not easily say.

\ciceroSection{2} You took the soldiers' interests to heart; whom I --- not for the sake of my own position (for I am conscious of meaning nothing that is not for the public good) --- wanted to see honoured by the Senate: in the first place because I judged them to deserve it, secondly because I wanted them more closely bound to the state against all contingencies, and lastly so that, turned away from solicitation from any quarter, I might be able to keep them for you the same kind of men they have been so far.

\ciceroSection{3} On our side we have kept everything here so far intact. This strategy of ours --- though I know the appetite of men for a victory of †such a kind is not without its grounds --- I hope wins your approval none the less. For if any setback should happen to these armies, the state has no large reserves ready to hand with which to withstand the sudden attack and brigandage of the parricides. I take it our forces are known to you. In my camp there are three veteran legions, and one of recruits which is, of all of them, perhaps the most magnificent; in Brutus's camp there is one veteran legion, a second of two years' service, and eight of recruits. So the army as a whole is very large in numbers, but slight in reliability; and how much it is safe to entrust to recruits in the line we have learned by experience too often.

\ciceroSection{4} If to this strength of our armies either the African army --- which is a veteran one --- or Caesar's were added, we should be entering on a decisive fight for the supreme stake of the commonwealth with an even mind; and we kept seeing that, as far as Caesar was concerned, he was somewhat the closer of the two. I never ceased to urge him to come by letter, nor did he break off assuring me that he was coming without delay --- yet meanwhile I see that, his mind turned aside from this thought, he has fallen back on other plans. Even so, I have sent our Furnius to him with instructions and a letter, in case anything could possibly be achieved.

\ciceroSection{5} You know, my Cicero, that as far as affection for Caesar goes I have a partnership with you in this, either because, on grounds of my intimacy with the elder Caesar, it was already a duty of mine to look out for him and cherish him while the father was alive, or because the young man himself, so far as I had been able to come to know him, was of a disposition altogether moderate and altogether humane, or because, after a friendship as marked as mine and Caesar's, it does not seem to me dishonourable to hold this one --- substituted, in his judgement and in yours, in the place of a son to that man --- on a corresponding footing.

\ciceroSection{6} But (whatever I am writing to you, by Hercules I do it more in sorrow than in unfriendliness) the fact that Antony is alive today, the fact that Lepidus is at his side, the fact that they have armies not to be despised, the fact that they have hope, the fact that they dare --- they can charge it all to Caesar's account. And I shall not go over earlier matters; but from the moment when he himself professed to me that he was coming, if he had wanted to come, the war would by now either be crushed, or pushed back into the corner of Hispania most hostile to them, at the cost of their heaviest losses. What thinking, or whose counsels, called him aside from glory so great --- and indeed for himself essential and salutary --- and turned him to dreams of a two-month consulship, with the city's people in the utmost terror and his demand for it in tasteless form, I cannot work out.

\ciceroSection{7} In this matter his own intimates seem to me capable of doing much good, for both the state's sake and his own; but most of all, as I imagine, you yourself, to whom he is in such debt for kindnesses as no one but myself. For I shall never forget that I owe you my greatest and most numerous debts. I have given Furnius instructions to press him on these matters. And if I have with him the influence I deserve, I shall have done him himself the greatest service.

\ciceroSection{8} We meanwhile are carrying on the war on a harder footing, because we judge neither that the issue is altogether easy to fight out, nor on the other hand that we shall, by drawing back, allow the state to take a greater loss. But if either Caesar reconsiders, or the African legions come up quickly, we shall make you, on this side, free of anxiety. As for yourself, I ask you to continue as you have begun --- love me, and persuade yourself that you have me for your own. 28 July, from camp.
