% Ad Familiares 11.16 (ad-familiares-11-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO D. BRVTO COS. DES. S. D.}

\ciceroSection{1} It is of the greatest importance at what moment this letter is handed to you --- whether at one when you have some anxiety on your mind or one when you are free from every annoyance. So I have given my orders to the man I am sending you, that he should watch for the right moment to put my letter into your hand. For just as, face to face, people who come up to us at the wrong moment are often a bother, so letters give offence when delivered in the wrong place. But if, as I hope, nothing is upsetting you, nothing standing in your way, and the man to whom I have entrusted this has chosen his moment of approach with sufficient tact and good sense, then I am confident I shall easily get from you what I am after.

\ciceroSection{2} L. Lamia is a candidate for the praetorship. He is the one man of all whom I see the most of: a great length of acquaintance lies between us, a great practice of intimacy, and --- which counts for most --- I have nothing more dear to me than his friendship. Besides this, I am bound to him by a great service and a great deserving of his. For at the time of the Clodian troubles, when he was the leading man of the equestrian order and was contending most fiercely for my preservation, he was banished from the city by the consul Gabinius --- a thing that, until that day, had happened to no Roman citizen at Rome. While the Roman people remembers this, for me myself to forget it would be the most disgraceful thing of all. So persuade yourself of this, my Brutus: I am the one standing for the praetorship.

\ciceroSection{3} For although Lamia is of the highest distinction and the highest standing, with that most magnificent aedile's show to his credit, still --- as if none of this were so --- I have taken on the whole business. As things stand, if you reckon me at the value at which you certainly do reckon me, then, since the centuries of the knights are in your hand --- you are king there --- send word to our friend Lupus to bring those centuries home for us. I shall not detain you with more; I shall set down at the end what I think: there is nothing, Brutus, given that I look to you for everything, that you could do to please me more.
