% Ad M. Brutum 1.9 (ad-brutum-01-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} I would discharge for you the office you discharged for me in my own bereavement, and console you in a letter, if I did not know that, of those remedies by which you eased my grief, you have no need in your own --- and I should wish that you might cure yourself now as easily as I was cured by you then. Yet it is unbecoming to so great a man as you, that he should be unable to do for himself what he has prescribed for another. As for me, both the reasons you had gathered and the weight of your authority deterred me from excessive mourning. For when I seemed to you to bear it more softly than became a man --- and that man one accustomed to consoling others --- you took me to task in a letter, in language sterner than your usual practice allowed.

\ciceroSection{2} And so, prizing your judgement and standing in awe of it, I pulled myself together, and the things I had learned, had read, had heard, I counted weightier with your authority added to them. At that time, Brutus, my duty was to office and to nature only; for you now it is to the people and to the stage, as the saying goes, that you must do service. For since not only the eyes of your own army but those of every citizen and almost of every nation are fixed upon you, it would be least becoming that the very man on whose account the rest of us are stronger should himself be seen weakened in spirit. You have indeed taken on a grief --- for what you have lost had no equal on earth --- and there must be grief in a wound so heavy, lest the very thing, the absence of all sense of grief, should be more wretched than grief itself; but, just as moderate grief is wholesome for the rest of us, so for you it is a necessity.

\ciceroSection{3} I would write more, were these very words not already too many for you. We are awaiting you and your army; without it, even if everything else goes according to our wish, we hardly seem likely to be free enough. About the commonwealth as a whole I shall write more, and perhaps now on better evidence, in the letter I was meaning to entrust to our old friend.
