% Ad M. Brutum 2.4 (ad-brutum-02-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} Having handed Scaptius a letter in the morning of the third day before the Ides of April, I received on the same day your letter sent from Dyrrachium on the kalends of April, in the evening. And so this morning, on the day before the Ides of April, when I had been informed by Scaptius that the bearers to whom I had given the letter the previous day had not set off, but were on the point of starting, I scratched out this little supplement amid the very throng of my morning callers.

\ciceroSection{2} About Cassius I am delighted, and I congratulate the commonwealth --- and myself too, who, with Pansa opposing and angry, delivered the motion that Cassius should pursue Dolabella by war. And indeed I was bold enough to say that, even without a senate-decree of ours, Cassius was waging that war already. About you also I said what I thought ought to be said. The speech itself will be carried to you, since I see that you take pleasure in our Philippics.

\ciceroSection{3} You consult me about Antony: until we have learned how Brutus's case has come out, I think he should be kept in custody. From the letters you sent me, Dolabella seems to be ravaging Asia and behaving most foully in it; but in several letters you have written that Dolabella has been shut out by the Rhodians. If he has approached Rhodes, then he seems to me to have left Asia. If that is so, you should stay where you are; but if †once he has taken it, believe me, the Ides of April will not be†, well, then it is to Asia, in my judgement, that he must be pursued. There is nothing I think you could be doing better at this time.

\ciceroSection{4} You write that you are in need of two essentials, reinforcements and money: this is a hard problem. For I cannot think of resources which I see you can call upon, beyond those which the Senate has decreed --- that you should take loans from the communities. As for reinforcements, I do not see what can be done. So far is Pansa from sparing you anything out of his own army or his levy, that he actually takes it ill that so many volunteers are going to you --- because, as I for my part believe, he judges that, for the operations being decreed in Italy, no forces are too large; though, as many suspect, it is because he does not even want you to be too strong; which I do not suspect.

\ciceroSection{5} You write that you have written to your sister Tertia and to your mother, telling them not to publish what has been done by Cassius until I should have seen fit: I see that you have been afraid --- and rightly so --- that the temper of the Caesarian party (as the party is even now called) would be violently stirred. But before your letter came, the matter had been heard and widely circulated; your own couriers, too, had brought letters to many of your friends. There was therefore no suppressing it, especially since that was impossible; nor, even if it were possible, did we think it ought to be hidden rather than openly spread.

\ciceroSection{6} About my Cicero: if as much is in him as you say, then I rejoice as much, of course, as I am bound to; and if you partly make the more of him because you love him, that very thing --- that he is loved by you --- I rejoice at beyond belief.
