% Ad Familiares 12.14 (ad-familiares-12-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{LENTVLVS CICERONI SVO S. P. D.}

\ciceroSection{1} When I had met up with our Brutus and noticed that he would be coming into Asia rather slowly, I returned to Asia myself, in order to gather up the remaining work I had begun there and to send the money on to Rome as soon as possible. In the meantime I learned that Dolabella's fleet was in Lycia, with more than a hundred transport vessels into which his army could be loaded, and that Dolabella had arranged this with the design that, if his hope of Syria failed him, he could board ship and make for Italy, joining himself with the Antonii and the rest of the brigands. I was in such alarm at this that I dropped everything and tried to go after his fleet with smaller ships and fewer of them.

\ciceroSection{2} If I had not been held up by the Rhodians, the whole business might perhaps have been wiped out; even so, it has been largely broken up, since their fleet was scattered by fear at our approach, the soldiers and commanders escaped, and every one of the transports was captured by us. At any rate, what I most feared I appear to have brought about: Dolabella cannot now get through to Italy, nor put his own confederates on a firmer footing and so make your work harder.

\ciceroSection{3} How thoroughly the Rhodians despaired of us and of the state, you will gather from the official dispatch I have sent. And I have written much more sparingly than I in fact found them to be raving. As for the fact that I have written something about them, do not be surprised: their madness is extraordinary. No private wrongs of my own --- not ever --- but their malice toward our cause, their appetite for the other side, their persistence in contempt for every man of standing --- this I could not bear. I do not, however, think them all corrupt: the same men who once refused entry to my father in his flight, to L. Lentulus, to Pompeius, to the other most illustrious figures, those same men, as if by some destiny, now either hold magistracies or have those who hold them in their power; and so they exercise the same arrogance in their depravity. To curb their wickedness at last, and not allow it to grow by going unpunished, is not only useful for our state but necessary.

\ciceroSection{4} As to my own standing, I should like you to keep it always in mind and, at whatever moment you have the opportunity, to lend your vote to my credit both in the Senate and in everything else. Since Asia has been decreed to the consuls and they have been permitted, until they come in person, to entrust the holding of Asia to someone, I ask you to request of them that they confer this distinction on me above all, and give me the commission to hold Asia until one or the other of them arrives. They have no reason to hasten out in their magistracy, or to send an army. For Dolabella is in Syria, and, as you with your divine insight foresaw and foretold, while they are on their way Cassius will crush him. Driven back from Antioch, and badly mauled in the siege, Dolabella, trusting in no other city, has withdrawn to Laodicea --- the one in Syria --- on the coast. There I expect he will soon pay the penalty: he has nowhere to flee to, and he will not be able to hold out long there against so large an army as Cassius's. Indeed, I expect Dolabella has already been finished off and broken.

\ciceroSection{5} Accordingly, I do not think that Pansa and Hirtius will hurry out of Rome into the provinces during their consulship, but will conduct their consulship at Rome. So if you will request of them that they give us the procuratorship of Asia in the meantime, I expect you can prevail. Besides, Pansa and Hirtius both promised me in person, and wrote to me in my absence, and Pansa assured our Verrius too, that they would see to it that no successor be sent to me during their consulship. As for me, by my faith, I have no wish for my term to be prolonged out of any appetite for the province: this province has been for me one of toil, danger, and loss, and what I am very anxious about is that I shall not have undergone all this in vain, nor be forced to leave before I have brought home the fruits of my labor. For if I had been able to send on the whole sum I had collected, I would myself be asking for a successor; as it is, what I gave to Cassius, what we lost by Trebonius's death, and what I now want to recoup --- through Dolabella's cruelty or the bad faith of those who failed me and the state --- I cannot recover unless I have the time. That I may obtain this through you, please, as is your custom, make it your care.

\ciceroSection{6} For my part, I think I have deserved well enough of the state that I ought not to be looking for the favor of this province alone, but for as much as Cassius and the Bruti --- not only by their share in that famous deed and its danger, but also by the zeal and courage they have shown in this present crisis. For I was the first to break the Antonian laws, the first to bring over Dolabella's cavalry to the state and hand it to Cassius, the first to hold a levy for the safety of all against the most criminal conspiracy, the only one to join Syria and the armies that were there to Cassius and to the state. If I had not given Cassius so much money, such substantial reinforcements, and so quickly, he would not even have dared to enter Syria, and at this moment the state would face from Dolabella no smaller perils than it does from Antonius.

\ciceroSection{7} And all this I did --- I, who was Dolabella's intimate companion and closest friend, bound to the Antonii by blood, and held my province too by their good offices --- but, loving my country more than my own (\textit{[Greek: patrida emēn mallon philōn]}), I was the first to declare war on all that was my own. Though I see that so far this has not borne me any great reward, still I do not despair, nor will I weary of holding fast not only in my zeal for liberty but in toil and dangers too. Yet if, in addition, I am summoned forward by some honor that is just and earned --- by the offices of the Senate and of every man of standing --- I shall stand with greater authority in the eyes of others, and so be able to do that much more good for the state.

\ciceroSection{8} As for your son: when I came to Brutus I was not able to see him, since he had already gone off into winter quarters with the cavalry; but, by my faith, I am glad to hear that he is in such repute, for your sake and his and not least for my own --- for he is to me as a brother, this man who is your son and worthy of you. Farewell. Pergē, 29 May.
