% Ad M. Brutum 1.10 (ad-brutum-01-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero Bruto salutem}

\ciceroSection{1} I have had no letter from you so far, nor even a report to indicate that, on learning the Senate's resolution, you are bringing your army into Italy --- which the commonwealth was urgently asking that you do, and that you do quickly. For the inward malady grows worse by the day, and we are no more harassed by foreign enemies than by domestic ones, who were there from the start of the war but were easier to break. The Senate was firmer in posture, roused not only by my motions but also by my exhortations. In the Senate Pansa was vehement and keen enough, both against the rest of this stamp and most of all against his father-in-law --- a consul to whom from the start spirit was not lacking, nor at the end fidelity.

\ciceroSection{2} The war was being waged at Mutina, in such a way that you could fault nothing in Caesar and something here and there in Hirtius. The fortune of this war was unsteady, as in prosperity it tends to be, but good as in adversity it tends to be. The commonwealth was victorious, Antony's forces cut down, Antony himself driven out. Then so many faults of Brutus's followed that in a way the victory slipped out of our hands. Our generals did not pursue men who were terrified, unarmed, wounded; and time was given to Lepidus in which to make us try out, in graver troubles, the levity of his we had so often seen through. The armies of Brutus and Plancus are sound, but raw; the Gallic auxiliaries are most loyal and very large in number;

\ciceroSection{3} but Caesar, who up to now had been guided by my counsels --- a young man of brilliant native endowment and admirable steadiness --- has been pushed, by certain persons through villainous letters and through deceitful go-betweens and messengers, into a most settled hope of the consulship. The moment I sensed it, I did not cease either to warn him at a distance by letter or to call to account, in person, those of his intimates who were seen to be backing his ambition, nor did I hesitate to lay open in the Senate the springs of the most criminal designs. And on no matter, in fact, do I recall a better Senate or better magistrates; for never, in the matter of an extraordinary honour for a powerful man --- or rather for the most powerful (since power is now lodged in violence and arms) --- has it happened that no tribune of the plebs, no other magistrate, no private citizen took the lead. But amid this firmness and courage the state was nevertheless uneasy. For we are being mocked, Brutus, now by the soldiers' caprices, now by the imperator's effrontery. Each man demands for himself as much power in the commonwealth as he has strength; reason has no force, nor measure, nor law, nor custom, nor duty, nor judgement, nor the esteem of one's fellow citizens, nor reverence for posterity.

\ciceroSection{4} I foresaw all this far in advance, and was running away from Italy at the time when the report of your edicts called me back; you in person spurred me on, Brutus, at Velia. For although I was distressed at going into the city you were fleeing, you who had freed it --- a thing that had once happened to me too, in a like danger, with a sadder outcome --- I nonetheless pressed on and reached Rome and, with no guard at my side, shook Antony, and against his unspeakable arms I made firm the guard that was offered, on Caesar's counsel and authority. If he stands by his word and obeys me, we shall, it seems, have guard enough; but if the counsels of the impious shall prove stronger than mine, or if the weakness of his age cannot bear the weight of these affairs, every hope is in you. Therefore fly here, I beg you, and bring final deliverance, by your exit, to that commonwealth which by your courage and greatness of spirit you have freed more truly than by the outcome of events. To you the whole rush of all men will be. Urge Cassius to the same by letter.

\ciceroSection{5} The hope of liberty has its standing-ground nowhere but on the parade-ground of your camp. We have, in fact, firm generals and firm armies from the West. As for this guard provided by the young man, I am confident, so far, that it is firm; but so many are working to loosen it that I am at times in dread that it may be moved. You have the whole condition of the commonwealth --- as, indeed, it stood at the moment I was sending off this letter. I should wish what comes next to be better. But if it shall turn out otherwise (which heaven avert!), I shall grieve on behalf of the commonwealth, which ought to have been immortal; for myself --- how little is left to me?
