% Ad M. Brutum 2.3 (ad-brutum-02-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Brutus Ciceroni salutem}

\ciceroSection{1} I am keenly awaiting the letter you wrote after the reports of our doings here and of Trebonius's death. For I do not doubt that you will lay out your judgement to me. By an unworthy crime we have lost both an excellent citizen and the hold of our province; but the province is easy to recover, and to have recovered it afterwards will be no less shameful or disgraceful for that.

\ciceroSection{2} Antonius is still with us; but, by my faith, both his entreaties move me and I fear that the fury of certain men may take him out of my hands. I am at a loss, plainly. If only I knew what you thought best, I should be free of anxiety: for I should be persuaded that whatever it was, that was the right course. So tell me, and as soon as you can, what you would have me do.

\ciceroSection{3} Our friend Cassius has Syria and the Syrian legions, summoned --- and indeed of their own motion --- by Murcus, by Marcius, and by the army itself. I have written to my sister Tertia and to my mother not to make public this matter, which Cassius has carried out with the greatest skill and good fortune, until they have heard your judgement and you have decided.

\ciceroSection{4} I have read your two speeches: the one you delivered on the kalends of January, the other about my letters, given by you against Calenus. You are no doubt expecting now that I praise them. I do not know whether the spirit or the talent of yours displayed in those little books is the more praiseworthy; I yield the point now --- let them even be called \textit{Philippics}, as you yourself wrote, jokingly, in a certain letter.

\ciceroSection{5} We are short of two things, Cicero: money and reinforcements. The first you can arrange, by getting some portion of the troops there sent to us --- whether by private understanding against Pansa or by a motion in the Senate; the second, which is the more pressing, and which is no more wanted by my army than by the rest, is what makes me grieve the more that we have lost Asia, which I hear is being so cruelly handled by Dolabella that the murder of Trebonius now does not seem his most savage deed. Veteran Antistius, however, has eased me with money.

\ciceroSection{6} Your son, Cicero, so commends himself to me by industry, endurance, hard work, greatness of spirit, by every dutiful quality in short, that he plainly seems never to let go of the thought whose son he is. Therefore, since I cannot bring you to value more highly one already dearest to you, this much grant to my judgement: persuade yourself that he will not have to lean on your glory to attain his father's honours. Kalends of April, from Dyrrachium.
