% Ad Familiares 10.28 (ad-familiares-10-28) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO TREBONIO S.}

\ciceroSection{1} How I wish you had invited me to that most beautiful banquet on the Ides of March! There would be no leftovers. As it is, we have so much business with what was left over that the divine service you all rendered the state is not without a certain note of complaint. That this plague is still alive --- because he was drawn aside by you, an excellent man, and survives by your kindness --- I am sometimes (though it is scarcely lawful for me) a little angry with you for; for you have left more business on my hands alone than on all other men's put together besides mine. The moment the Senate could meet freely, after Antony's most foul departure, I returned to that old spirit of mine which you, together with that keenest of citizens your father, always had on your lips and in your affection.

\ciceroSection{2} For when the tribunes of the plebs had called the Senate for 20 December and were laying another matter before it, I took up the whole state of the commonwealth and spoke with all force I had, and a Senate already growing slack and weary I called back to its old courage and habit, by powers of spirit rather than of intellect. That day and my exertion and action first brought the Roman people the hope of recovering its liberty; nor for my own part did I afterwards let any time pass without not merely thinking of the commonwealth but acting on it.

\ciceroSection{3} And were I not supposing that the affairs of the city and all that has been done are being reported to you, I should write them out myself, though I have been hindered by the heaviest occupations. But all that you will learn from others; from me only a little, and that in summary. We have a brave Senate; of the consulars some are timid, some ill-disposed; great loss has been suffered in Servius; L. Caesar is of the best disposition, but since he is Antony's uncle he does not deliver his most cutting opinions; the consuls are excellent, D. Brutus splendid, the boy Caesar admirable, and of him I do for my part have hopes for the rest; but hold this for certain --- had he not enlisted the veterans rapidly and had not two legions from Antony's army gone over to his authority, and had not that terror been set against Antony, there is no crime, no cruelty, that Antony would have left undone. This I wanted you to know more clearly, though I supposed you had heard it. I shall write more, if I have more leisure.
