% Ad Atticum 15.19 (ad-atticum-15-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What further is to be done about the Buthrotians? You write that you have laboured in vain. And why on earth does Brutus take it back upon himself? I am sorry, by Hercules, that you are so pulled about, with ten men's business to be reckoned to your account. It is laborious work \textit{[Greek: erg\=odes]}, no doubt, but bearable \textit{[Greek: anekton]}, and most welcome to me. About the arms, I have seen nothing more open. Let us run, then, and, as you say, face to face. What Theophanes wants, I do not know. He had written to me; I have written back as best I could. He writes that he wishes to come to me, both on his own affairs and on certain matters that concern me. I am waiting for a letter from you. See, please, that nothing is done rashly.

\ciceroSection{2} Statius has written to me that Quintus Cicero has spoken with him, asserting most emphatically that he cannot bear these things; that he is determined to go over to Brutus and Cassius. I am very eager just now to learn this --- what it is I cannot interpret. It may be that he is angry with Antony; it may be that he is now seeking new glory; it may be the whole thing is improvised \textit{[Greek: schediasma]}; and so, no doubt, it is. And yet I have my fears, and his father is upset, for he knows what the younger Quintus has said about him, things to me of old beyond words \textit{[Greek: aphata]}. I simply do not know what he wants. From Dolabella I shall have such instructions as seem fit to me --- that is, none. Tell me: did Gaius Antonius wish to be made a member of the board of seven? Certainly he was deserving of it. About Menedemus, things are as you write. You will let me know everything.
