% Ad Atticum 14.9 (ad-atticum-14-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About the state I have learnt much from your letters --- which, indeed, I received in a bundle all at one time from Vestorius's freedman. To the points you ask about I shall reply briefly. First, I am hugely pleased with the Cluvian estates. As for your question why I sent for Chrysippus: two of my shops have collapsed and the rest are gaping with cracks, so that not only the tenants but even the mice have moved out. Other people call this a calamity; I do not even call it an inconvenience. O Socrates, and ye Socratic men! I shall never repay you my debt. Immortal gods, how all that means nothing to me! All the same, a plan of building is being entered into --- with Vestorius as adviser and prime mover --- on terms that should turn this loss to profit.

\ciceroSection{2} Here there is a great crowd, and there will be a greater, I hear. Two of them are practically consuls-designate. O ye good gods! The tyranny lives, the tyrant fallen! We rejoice in the death of the one whose acts we defend! And so how sternly Marcus Curtius takes us to task, that we should be ashamed to be alive --- and not unjustly. For a thousand times better to die than to bear what I see has the look of lasting into our old age.

\ciceroSection{3} Balbus is here too, and much with me. A letter has been sent to him by Vetus, dated the day before the Kalends of January, saying that as Caecilius was being besieged by him and was already in his hands, Pacorus the Parthian came up with very large forces; that Caecilius was thus wrested from him, with many of his own men lost. In this Vetus complains of Volcatius. So that war, it seems to me, is on its way. But let Dolabella and Nicias see to it. The same Balbus has better news from Gaul. On the twenty-first day he had a letter that the Germans and those nations, having heard the news about Caesar, had sent envoys to Aurelius, who has been put in charge by Hirtius, saying that they would do what they were ordered. In short: all full of peace, contrary to what Calvena had told me.
