% Ad Atticum 15.17 (ad-atticum-15-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I received two letters on the day after the Ides --- one sent that same day, one on the Ides itself. Take the earlier one first, then. About Decimus Brutus, when you have news. About the consuls' feigned alarm I had already heard: Sicca, with real warmth of heart \textit{[Greek: philostorg\=os]} but in some agitation, brought that same suspicion to me too. What about you, though? \textit{The things granted ---?} \textit{[Greek: ta men didomena ---?]} For I have had no word from \dag Siregius\dag. I do not approve. About your neighbour Plaeto, I took it very ill that anyone heard of it before I did. About Syrus, prudently done. Lucius Antonius you will most easily put off through his brother Marcus, I think. As for Antronius, I forbade it; but you had not yet received my letter --- not to anyone but the aedile Lucius Fadius. There is no other safe or lawful way. As to what you write about your being short by 100,000 sesterces that have been paid out for young Cicero, please ask Eros where the rent from the apartment blocks has gone. With Arabio I am not at all angry over Sittius. As for the journey, until \dag L \dag\ is sorted out I am thinking nothing of it; and I take it the same view appears to you.

\ciceroSection{2} So much for the earlier letter. Now hear the answer to the other. You do indeed do exactly right in all things in not failing Servilia, that is to say, Brutus. About the queen, I am glad you are not troubling yourself, and that you find the witness convincing too. Eros's accounts I have learned both from Tiro and from the man himself, whom I sent for. Your promise that young Cicero shall lack for nothing is most welcome; about him Messalla, on his way back from Lanuvium, came to me from them and gave a marvellous report, and the boy's own letter, by Hercules, is so warm in feeling \textit{[Greek: philostorg\=os]} and so neatly turned \textit{[Greek: eupin\=os]} that I should dare to read it out at a public recitation. All the more reason, I think, to indulge him. About Bucilianus, I suppose Sestius is not put out. As for me, if Tiro is sent to me, I am thinking of going to the Tusculan villa. You, for your part: anything it is fit for me to know, send it at once.
