% Ad Atticum 14.11 (ad-atticum-14-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The day before yesterday I sent you a longer letter; now to your most recent. I should be glad, by Hercules, if Brutus were at Astura. You write of \textit{their licence} \textit{[Greek: akolasian]}. Did you suppose otherwise? For my part I expect worse still. When I read out a speech of that kind about so great a man, so illustrious a citizen, I cannot bear it. And yet by now such things move me to laughter. But remember: this is how the habit of ruinous public assemblies is fed --- so that our friends, those not heroes but gods-to-be, will indeed have everlasting glory, but not without envy, indeed not without danger. The great consolation for them is the consciousness of an act supreme and most illustrious; for us, what? --- for us, who, the king once killed, are not free. But fortune may look to this, since reason does not govern.

\ciceroSection{2} What you write about Cicero is welcome to me; I hope it goes well. That you take care to supply him liberally for his maintenance and upkeep is most agreeable to me, and I beg you again and again to keep doing so. About the Buthrotians, you for your part are taking the right line, and I do not let go of the concern. I shall take up the whole business in court too --- a business I see grows easier by the day. About the Cluvian estate: since in my own affair you outdo me in diligence, the income is being brought up to a hundred thousand. The collapse has not made it worse, perhaps even more profitable. Here with me are Balbus, Hirtius, Pansa. Octavius has just arrived, into the villa of Philippus right next door, and devoted entirely to me; Lentulus Spinther is with me today; tomorrow morning he leaves.
