% Ad Atticum 14.1 (ad-atticum-14-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I called in on the man we spoke of this morning. Nothing could be more lost; the thing cannot be untangled. For if a mind like his could find no way out, who will find one now? In short, he said everything had gone to ruin --- whether this is so I cannot tell; he said it, at any rate, with pleasure --- and he affirms that within less than twenty days there will be trouble in Gaul. Since the Ides of March, he has spoken with no one but Lepidus. In sum: matters cannot go on like this. What a prudent man Oppius is! He misses Caesar no less, but says nothing that could offend any decent man. But enough of all that.

\ciceroSection{2} You, please, whatever news there is --- and I am waiting for plenty --- do not be slow to write, in particular whether what we hear about Sextus is firm enough, and above all about our Brutus. The man I called on said this about him: Caesar used to say, ``It matters a great deal what this fellow wants, but whatever he wants, he wants intensely''; and that he had noticed this when Brutus spoke at Nicaea on behalf of Deiotarus --- he had struck Caesar as speaking with great force and great freedom. And then this too, for I might as well write whatever comes to mind: recently, when I was at Caesar's at Sestius's request and was sitting waiting to be called in, he said, ``Can I doubt I am thoroughly disliked, when Marcus Cicero sits there and cannot get in to see me at his own convenience? And yet if anyone is approachable, this man is. But I have no doubt he heartily hates me.'' This and much more in the same vein. But to the point. Whatever happens, not only large but small, write to me. I shall not let up for my part.
