% Ad Atticum 15.12 (ad-atticum-15-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Good news, by Hercules, about Buthrotum. As for me, I had sent Tiro to Dolabella with a letter, because you had told me to. What harm in it? About our friends at Antium I thought I had written plainly enough not to leave you in any doubt that they would be sitting on their hands and using Antony's insulting favour. Cassius was spurning the grain-business; Servilia, he said, would have it struck out of the senatorial decree. But ours, very gravely \textit{[Greek: kai mala semnōs]}, was for Asia: after he had agreed with me that he could not be safe at Rome (and he preferred in any case to hold the games in his absence), he said he would set off as soon as he had handed the preparation of the games over to those who would manage them. He was collecting boats; his mind was already in passage. In the meantime they would be staying where they were.

\ciceroSection{2} Brutus said he would be at Astura. As for Lucius Antonius, in a generous letter he bids me have no anxiety. That is one favour I have in hand, and perhaps a second to come if he visits the Tusculan villa. The vexations are unbearable! And yet they are borne. \textit{[Greek: tōnde aitian tōn Broutōn tis echei.] One of the Bruti bears the blame for this.} In Octavian, so far as I could see into him, there is ability enough, spirit enough; and he seemed likely to be as well disposed towards our heroes \textit{[Greek: hērōas]} as we could wish. But what trust to put in his youth, in his name, in his inheritance, in the schooling he is getting \textit{[Greek: katēchēsei]} --- this calls for great deliberation. His stepfather, at any rate, thought nothing of it; I saw him at Astura. Still, he is to be cherished, and at the very least cut off from Antony. Marcellus does admirably, if he is putting our young man through his paces \dag for us\dag; the young man does seem devoted to him. Pansa and Hirtius he was not trusting too far. A good nature, \textit{[Greek: ean diameinēi] if it lasts}.
