% Ad Familiares 11.27 (ad-familiares-11-27) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO MATIO S.}

\ciceroSection{1} I have not yet quite settled whether our friend Trebatius brought me more vexation or more pleasure --- a man as full of dutifulness as he is of affection for us both. For when I had come to my place at Tusculum towards evening, the next day he came to me in the morning, although his health was not yet sufficiently strong. When I scolded him for sparing his health too little, he answered that nothing had felt longer to him than the time before seeing me. ``Is there any news?'' I said. He brought to me your complaint. Before I reply to it, I will set a few things before you.

\ciceroSection{2} So far as I can search the past by memory, no friend of mine is older than you. But long standing has something in common with many; affection does not. I cared for you from the day I came to know you, and I judged that I was cared for by you. Your departure afterwards, and a long one at that, and our public ambitions, and the unlikeness of our manner of life, did not allow our inclinations to be cemented together by daily company; yet I recognised your feeling towards me many years before the civil war, when Caesar was in Gaul. For what you thought to be greatly to my advantage --- and not without advantage to Caesar himself --- you brought about: that he should care for me, court me, count me among his own. I pass over many things which in those days were said, written, shared between us with the utmost intimacy; for graver matters followed.

\ciceroSection{3} At the start of the civil war, when you were on your way to Caesar at Brundisium, you came to me at my villa near Formiae. First of all, that very visit --- how much it was worth, especially in those times! Then, do you suppose that I have forgotten your counsel, your conversation, your humanity? In all of which I remember that Trebatius shared. Nor indeed have I forgotten the letter you sent me when you had gone out to meet Caesar in the territory, I think, of Trebula.

\ciceroSection{4} There followed that time when either my sense of honour, or my duty, or fortune, compelled me to set out to Pompey. What office of yours, what zeal of yours, was wanting either towards me in my absence or towards my people who were present? Whom, moreover, did all who belong to me judge a closer friend to me and to themselves than you? I came to Brundisium. Do you suppose that I have forgotten with what speed --- the moment you heard --- you flew to me from Tarentum? what your sitting beside me was, what your speech was, what the rallying was of my spirit broken by the dread of our common misfortunes?

\ciceroSection{5} At last, eventually, we began to be at Rome. What was lacking to our intimacy? In the greatest questions --- how I should conduct myself with respect to Caesar --- I made use of your counsel; in the rest, of your good office. To whom, except Caesar, besides me, did you grant the favour of calling at his house, of spending many hours there often in the most delightful conversation? It was then too, if you remember, that you urged me to write these works of philosophy. After Caesar's return, what was a greater concern to you than that I should be on the most familiar terms with him? Which you had brought about.

\ciceroSection{6} To what end, then, has this discourse run longer than I had thought? Because I was astonished that you, who ought to know all this, should have believed that anything had been done by me at variance with our friendship. For besides what I have just recounted, which is on record and conspicuous, I have many things more hidden, which I can scarcely follow out in words. Everything about you delights me, but most of all the greatest things: a great loyalty in friendship, judgement, weight, constancy --- and then charm, humanity, learning.

\ciceroSection{7} For which reason --- now I come back to the complaint --- I, in the first place, did not believe that you had cast your vote for that law; and in the second place, had I believed it, I would never have supposed that you had done so without some just cause. Your standing makes it so that whatever you do is noted; while men's ill will makes it so that things are reported as harsher than they really were when done by you. If you do not hear of these things, I do not know what to say. For my part, whenever I hear, I defend you as I know I am accustomed to be defended by you against my enemies. The defence, moreover, is twofold: some things are such as I am accustomed flatly to deny --- as about that very vote --- and others are such that I defend them as having been done by you out of piety and humanity, as about your management of the games.

\ciceroSection{8} But it does not escape you, learned man that you are, that if Caesar was a king --- which to me at least he seems to have been --- one can argue your case on either side: either on the side I am accustomed to take, that your loyalty and humanity are to be praised, since you love a friend even though dead; or on the side some take, that the liberty of one's country is to be set above the life of a friend. Out of such discussions, would that my arguments had been carried back to you! Indeed those two things, which are the greatest of your praises, who recounts either more gladly than I, or more often: that you were the gravest authority both against undertaking the civil war and for moderating the victory? On which point I have found no one who did not agree with me. For which reason I am grateful to our friend Trebatius for giving me cause for this letter; in which, if you do not put trust, you will have judged me devoid of all duty and humanity. Than which nothing can be more grievous to me, or more foreign to you.
