% Ad Familiares 10.3 (ad-familiares-10-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO PLANCO S.}

\ciceroSection{1} I was already most glad to see Furnius for his own sake, and gladder still on this account: that in hearing him I seemed to be hearing you. For he set out for me your manhood in military matters, your justice in the administration of the province, your wisdom in every kind of business; and besides this --- though I knew it already from familiar acquaintance --- your charm; and on top of that, your great generosity toward himself. All this was a pleasure to me, and the last item was a gratitude.

\ciceroSection{2} For my part, Plancus, the bond I struck with your house was established a good while before you were born; my affection for you, from your earliest boyhood; the friendship with you in your now settled years, established both by my own devotion and by your judgment. For these reasons I am wonderfully devoted to your standing, which I hold ought to be a thing held in common between us. Everything of the first rank you have achieved with virtue as your guide and fortune as your companion, and you have won these things as a young man, with many resenting you --- whom by your gifts and your industry you broke. Now, if you listen to me --- who love you more than any man and yield to none in the long-standing claim of intimacy --- you will earn all the standing of the rest of your life from the soundness of the republic's condition.

\ciceroSection{3} You know surely (nothing of this can have escaped you) that there was a certain time when men thought you too compliant with the times; which I should have thought too, had I supposed that you actually approved what you were enduring. But because I understood what your real sentiments were, I judged that you, with prudence, were taking the measure of what was possible. Now the case is different. The judgment in everything is your own, and free. You are consul-designate, in the prime of life, of the highest eloquence, at a moment of the deepest bereavement of such men in the republic. Press on, by the immortal gods! into that care and consideration which will bring you the highest standing and glory; for there is one course --- this above all at the present moment, after so many years of the republic's mauling --- and one only, that leads to glory: governing the republic well.

\ciceroSection{4} These things I thought I should write to you driven more by affection than because I supposed you needed warnings and precepts; for I knew you were drawing them from the same springs from which I had drawn them myself. So I shall hold off. For the present I thought only this should be signalled --- that I should show you my affection rather than display my own prudence. Meanwhile, whatever I judge to bear on your standing, I shall attend to with zeal and care.
