% Ad Familiares 11.1 (ad-familiares-11-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{D. BRVTVS BRVTO SVO ET CASSIO S.}

\ciceroSection{1} Learn what state we are in. Last evening Hirtius was at my house; he laid out what Antony's frame of mind is --- the worst, naturally, and the most faithless. For Antony was saying that he could neither grant the province to me nor judged that any of us was safe in the city, so worked up are the tempers of the soldiers and of the populace. Both of these I take to be false, and I take it to be true what Hirtius was making plain: that Antony is afraid that, if we had any moderate support for our standing, no role at all would be left to his side in the state.

\ciceroSection{2} Caught in these straits, I decided to ask for a \textit{legatio libera} for myself and for the rest of our number, so that some honourable pretext for setting out might be found. He promised that he would secure this; but I do not trust that he will secure it, so great is the insolence of the men and the harrying of us. And if they do grant what we are asking, even so I think that not long afterwards we shall be declared enemies, or interdicted from water and fire.

\ciceroSection{3} ``What then,'' you say, ``is your plan?'' We must give ground to fortune, leave Italy behind, and remove ourselves --- to Rhodes, I think, or somewhere in the world. If the turn of events is better, we shall come back to Rome; if middling, we shall live in exile; if the worst, we shall come down to the last resorts.

\ciceroSection{4} It will occur perhaps to one of you, at this point, why we are waiting for the last extremity rather than setting something in motion now. Because we have no ground to stand on except Sextus Pompeius and Bassus Caecilius, who, I think, will be the steadier once this news about Caesar reaches them. We shall come up to them with time enough to spare, once we know what their strength amounts to. On Cassius's behalf and yours --- if you would have me undertake anything, I shall undertake it; for Hirtius is asking me to do this.

\ciceroSection{5} I beg you to write back to me as soon as possible (for I do not doubt that, on these matters, Hirtius will be informing me before the fourth hour); write back where it is we can meet, and where you would have me come.

\ciceroSection{6} After my last conversation with Hirtius I decided to demand that we be allowed to stay at Rome under a public guard. That, I do not think they will grant us --- we shall be putting them in very great odium. Still, I thought nothing should go unasked that I had judged to be fair.
