% Ad Familiares 12.3 (ad-familiares-12-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CASSIO S.}

\ciceroSection{1} Your friend's frenzy mounts by the day. To begin with, on the statue he set up on the Rostra he inscribed ``TO A PARENT MOST WELL-DESERVING'' --- so that you are now to be judged not merely assassins but actually parricides. Why do I say ``you''? ``We,'' rather; for that maniac says that I was the ringleader of your most splendid deed. Would that I had been! He would not now be a nuisance to us. But that exploit is yours; and since it is past, would that I had something to give you by way of advice! But I cannot even find for myself what is to be done. For what is there that can be done against violence without violence?

\ciceroSection{2} The whole plan of that party is to avenge Caesar's death. And so, on the sixth day before the Nones of October, when he was brought before a public meeting by Cannutius, he came off most shamefully indeed, but for all that he said about the saviors of their country what ought to be said of traitors --- about me, he said without hesitation that you had done all by my counsel and that Cannutius was acting on it now. What the rest of their conduct is like, judge from this: they have stripped your envoy of his travel-money. What do you suppose their construction is, when they do this? That it is being carried to a public enemy, of course. What a wretched business! We could not bear a master; we are slaves to a fellow-slave. And yet, with my favor rather than my hope, hope still rests even now in your courage. But where are the troops? As to the rest, I would rather you talked it over with yourself than learn my views. Farewell.
