% Ad Familiares 9.14 (ad-familiares-09-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO DOLABELLAE CONSVLI SVO S.}

\ciceroSection{1} Although I was content, my Dolabella, with your glory, and from it was taking great joy and pleasure enough, still I cannot help confessing that I am crowned with the greatest delight, because the common opinion of mankind enrolls me as partner in your praises. I have not met a man (and I meet a great many every day --- for there are many of the best sort here for their health, and besides them many of my friends from the country towns) who, after lifting you to the heavens with the highest praises, has not immediately followed up by giving me the warmest thanks; for they say they have no doubt that, obeying my teaching and counsel, you are showing yourself a most outstanding citizen and a consul without peer.

\ciceroSection{2} To these I might answer, with perfect truth, that what you do, you do by your own judgment and of your own accord, and need no one's counsel; yet I neither quite agree --- so as not to diminish your praise, if it should appear to have all proceeded from my counsels --- nor do I emphatically deny; for I am even more greedy of glory than I ought to be. And after all, it is not foreign to your dignity --- since it was an honor to Agamemnon himself, the king of kings --- to have someone of the sort of Nestor among those whose counsel he takes; and to me it is glorious that you, a young man and consul, should flourish with praises as a kind of pupil of my school.

\ciceroSection{3} Indeed Lucius Caesar, when I came to him in his illness at Naples --- although he was overwhelmed by pain in his whole body --- still, before he had even properly greeted me, said, ``My dear Cicero, I congratulate you on having as much influence with Dolabella as, if I had with my sister's son, we could already be safe. As for your Dolabella, I both congratulate him and thank him --- the one man whom, since your consulship, we can truly call a consul.'' Then he said much about your deed and your achievement: that nothing more magnificent, nothing more splendid had ever been done, nothing more salutary to the Republic. And this is everyone's single voice.

\ciceroSection{4} From you, however, I ask that you suffer me to take possession of this seeming windfall, this inheritance of another's glory, and that you let me come, in some part, into partnership in your praises. Although --- my Dolabella (for what I have said so far has been said in jest) --- I would more gladly pour all my praises, if I have any praises of my own, over to you than drain any portion off from yours. For just as I have always loved you as much as you could understand, so by these deeds of yours I am now so kindled that nothing in love has ever been more ardent. For there is nothing, believe me, more beautiful than virtue, nothing fairer, nothing more lovable.

\ciceroSection{5} I have always loved Marcus Brutus, as you know, for his supreme talent, his sweetest character, his singular probity and constancy; yet on the Ides of March so much was added to my love that I marveled there had been any room for increase in something that already seemed to me piled high. Who would have thought that anything could be added to the love I had for you? Yet so much has been added that now at last I seem to myself to love you, and only to have liked you before.

\ciceroSection{6} Why, then, should I urge you to be the servant of your dignity and glory? Should I set before you famous men, as those who urge are accustomed to do? I have no one more famous than yourself; you should imitate yourself, you should compete with yourself; nor is it any longer permitted you, after such great achievements, to be less than your own equal.

\ciceroSection{7} This being so, exhortation is unnecessary; congratulation is more in order. For you have had what has perhaps fallen to no one's lot before: that the highest severity of punishment was not only free of odium but actually popular, and most welcome both to all good men and to the lowest of the people. Had this come to you by some stroke of fortune, I would congratulate your luck; but it has come to you by greatness, of spirit alike, and of talent, and of judgment. For I have read your speech to the assembly; nothing could be wiser; the way you advanced toward the case step by step and by degrees, and then withdrew, so that the affair itself, by everyone's consent, gave you the ripe moment for punishment.

\ciceroSection{8} You have therefore freed the city from danger and the citizen body from fear; nor have you brought the greatest benefit only for the moment, but also as a precedent. By that deed you should understand that the Republic has been placed in your hands, and that you must not only protect, but also exalt, those men through whom liberty's beginning was made. But more of these matters in person, soon, I hope. As for you, since you are saving the Republic and us, see to it, my Dolabella, that you most carefully guard yourself.
