% Ad Atticum 15.26 (ad-atticum-15-26) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On Quintus's business I see you have done everything. He, however, is in distress, hesitating whether to oblige Lepta or to weaken faith with his son. I have heard a whisper that Lucius Piso wants to go out on a legation by means of a forged decree of the Senate \textit{[Greek: pseudengraph\=oi]}. I should like to know what is in it. The courier I had told you I was sending to Brutus came to me at the Anagninum villa on the night just before the Kalends, and brought me a letter --- in which one thing, given his great prudence, was out of keeping, the same point as before: that I should attend his games. I wrote back, of course, that, first, I had already set out, so the question was no longer open; and then, that it was the most absurd thing in the world \textit{[Greek: atop\=otaton]} that I, who since these arms began had not gone near Rome at all (and that not so much because of the danger to myself as because of my dignity), should suddenly turn up for games. For at such a time to put on games is honourable for him, who is bound to do so; for me to watch them is no more honourable than it is necessary. I do most extraordinarily want those games to be well attended and to give as much pleasure as possible, and I am confident it will be so; and I ask you to keep me informed from the very opening, how these games are received, and then in detail the day-by-day course of the rest of them. But enough of the games.

\ciceroSection{2} The remaining part of his letter cuts both ways, but here and there throws off some manly little sparks. So that you might be able to interpret what kind of impression it made on me, I have sent you a copy of the letter --- though our courier had told me he had brought a letter from Brutus to you as well, and that it had been delivered to you from the Tusculan villa.

\ciceroSection{3} I had arranged my stages so as to be at Puteoli on the Nones of July; for I am in great haste, but only so far as that, with all the care a man can manage, I make a cautious crossing.

\ciceroSection{4} You will get Marcus Aelius out of his anxiety. To think that I supposed a few feet --- and those underground --- at the far end of an estate would carry some kind of servitude. That I no longer want this at all, and that the water is not worth so much to me. But, as you used to tell me, do it as gently as possible, so that he is relieved of his worry rather than left to suspect I am at all annoyed. Likewise, about that half-share of the Tullianum, speak frankly with Cascellius. It is a small matter, but you have attended to it well. He was acting too sharply. If he had put one over on me --- as he nearly did, had your shrewdness not stepped in --- I would have taken it ill. So, whatever happens, I would rather the business be blocked. As for the eighth share, \dag of the lights of medium\dag --- remember to whom \dag Caerellia\dag\ --- you may seem to be making the conveyance by mancipation up to that figure that was the highest under the auctioneer's hammer. That, I think, is 380,000.

\ciceroSection{5} If there is anything new, and also if you see anything in prospect that you think will happen, write to me as often as possible. With Varro, as I have instructed you, remember to apologise for the slowness of my letters. What that fellow Mundus has done with Marcus Ennius about the will --- I am curious --- I should like you to let me know. From the Arpinum estate, six days before the Nones.
