% Ad Atticum 14.12 (ad-atticum-14-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} O my Atticus, I am afraid the Ides of March have given us nothing beyond the joy and the satisfaction of our hatred and pain. What is brought to me from your end! What I see here! \textit{Oh!} \textit{[Greek: \=o]}. You know how I love the Sicilians and what an honourable client-relation I judge that to be. Caesar gave them many favours --- not against my wishes, though the grant of Latin status was a step too far. Yet even so ---. But now look: Antony, having taken a large sum of money, has posted up a law carried by the dictator before the assembly, by which the Sicilians become Roman citizens --- a matter of which, in his lifetime, there was no mention. And what of our Deiotarus? Is his case not similar? He deserves any kingdom, certainly --- but not through Fulvia. Six hundred more of the same kind. But I come back to my point. So plain a matter, so well attested, so just, the Buthrotian case --- shall we not carry it at least in part? --- and all the more, given how many things \textit{he} is carrying through? Octavius is here with us, most respectful and most friendly.

\ciceroSection{2} His own men were addressing him as Caesar; Philippus did not, and so neither did we; and I deny that he can be a good citizen. So many men stand around him who threaten our people with death, and they say these things cannot be borne. What do you suppose will happen when the boy comes to Rome, where our liberators cannot be safe? They will always be illustrious, true; in the consciousness of their deed, even blessed. But we, unless I am mistaken, will be flat on our backs. So I long to leave for a place where, as the saying goes, neither the Pelopidae's name . . . I do not much care either for these consuls-designate, who have even forced me to declaim --- so that there is no rest for me even at the baths. But this is my too-easy nature. For once such a thing was a kind of necessity; now, however matters stand, it is not the same.

\ciceroSection{3} How long since I have had anything to write you about! Yet I write, not to give pleasure with this letter but to draw out one of yours. If there is any news about the others --- and about Brutus, anything at all. I have written this on the tenth day before the Kalends, reclining at Vestorius's table, a man remote from dialectic, in arithmetic well practised enough.
