% Ad Atticum 15.2 (ad-atticum-15-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the fifteenth before the Kalends, as I was setting out from the Sinuessan villa, after I had despatched a letter to you and turned aside \dots\ to the Vescine, I received your letter from the courier --- in which there is far too much about Buthrotum. For that business is not, and will not be, more of a care to you than to me. So it ought to be: that you should mind my affairs, and I yours. For that reason this charge has been so taken up by me that I shall hold nothing more in front of it.

\ciceroSection{2} I learn from your letter and from others that L.~Antonius has spoken at a public meeting, and squalidly so; but what sort of thing it was I do not know, for nothing has been written down. About Menedemus, well done. Quintus certainly keeps repeating what you write. That you approve my plan --- that I do not write what you have demanded of me --- I bear easily; and you will approve it much more if you have read the speech I wrote to you about today. What you write about the legions is true. But you do not seem to have convinced yourself of this sufficiently, when you hope our Buthrotian business can be settled through the Senate \dots\ which I think. For this much I see: we do not seem destined to be victors; but, granting that this judgment of mine deceives us, about Buthrotum it will not deceive you.

\ciceroSection{3} About Octavius's public meeting I feel just as you do; and his preparations for the games, and Matius and Postumus as his managers, do not please me; Saserna is a fitting colleague. But all that crowd, as you perceive, fear peace no less than we fear arms. I should be glad if Balbus could be lightened of public hatred through our agency; but even he himself does not trust that it can be done. And so he is meditating other courses.

\ciceroSection{4} That the first book of the \textit{Tusculan Disputations} fortifies you, I am very glad; for indeed there is no refuge either better or more ready to hand. That Flamma speaks well of me, I take in good part. The case of the Tyndaritans, in which he is exerting himself, I do not know what it is. These \textit{five-remaining points} (\textit{[Greek: pentel\=oipon]}) seem to move them in particular --- the outlay of money. About Alexio I grieve; but since he had fallen into so grave an illness, I judge things went well with him. Still, I should like to know who his secondary heirs are and the date of the will.
