% Ad Familiares 11.3 (ad-familiares-11-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{BRVTVS ET CASSIVS PR. S. D. ANTONIO COS.}

\ciceroSection{1} If you are well, it is well. We have read your letter, the very image of your edict --- abusive, threatening, and least of all worthy to be sent from you to us. We, Antony, have provoked you with no injury, and we did not suppose you would think it strange if praetors, and men of our standing, asked something by edict of a consul. But if you take offence that we ventured to do that, allow us to grieve that not even this much is conceded by you to Brutus and Cassius.

\ciceroSection{2} As to the levies you have held and the moneys you have requisitioned, the armies you have tampered with and the messengers you have sent overseas --- since you say you have made no complaint of these, we for our part believe you have acted with the best intention; but we neither acknowledge any of these things, and we wonder, since you held your tongue on these, that you could not contain your anger from throwing Caesar's death in our faces.

\ciceroSection{3} Consider yourself how this is to be borne: that praetors are not to be permitted, by edict, for the sake of concord and liberty, to step back from their own right, without the consul threatening arms. Trusting in those arms, you have no occasion to terrify us; for it is not seemly nor fitting to us to bend our spirit in the face of any danger, and Antony has no business commanding the men by whose work he is a free man. If other considerations were urging us to want to stir up civil war, your letter would accomplish nothing; for the authority of one who threatens has no purchase upon free men. But you understand very well that we cannot be driven anywhere, and perhaps for that reason you act menacingly --- so that our deliberate judgement may look like fear.

\ciceroSection{4} We are of this mind: we would have you great and honoured in a free state; we summon you to no enmities; but nevertheless we value our liberty more highly than your friendship. Consider again and again what you are taking on, what you can bear up under; do not contemplate how long Caesar lived, but how short a time he reigned. We pray the gods that your counsels may be wholesome for the state and for yourself; if they are not, we pray that, with the state safe and in honour, they may do you the least harm possible. The day before the nones of Sextilis.
