% Ad Atticum 15.29 (ad-atticum-15-29) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I have sent on to you a letter of Brutus's. Good gods, what helplessness \textit{[Greek: am\=echania]}! You will see when you read it. About the celebration of Brutus's games I agree with you. To Marcus Aelius, you of course need not go to the house, but if you fall in with him anywhere. About the half-share of the Tullianum you will bring in Marcus Axianus, as you write. That you negotiated with the man from Cosa --- excellent. That you free me not only from my own business but from yours as well --- welcome. That my legation is approved, I am glad. What you promise --- may the gods make it so! For what could be sweeter to me than my own people? But that proviso of yours, I fear. When I have met with Brutus, I will write you everything. About Plancus and Decimus, by all means. As for Sextus throwing his shield away, I would not have wanted that. About Mundus, if you learn anything.

\ciceroSection{2} I have written back to everything in yours; now hear my news. Quintus the son is coming all the way to Puteoli --- a marvellous citizen, as if you were to call \dag Favonius an Asinius\dag --- and that on two counts: \dag both because he wants to make peace with me and to make a treaty\dag\ \textit{[Greek: speisasthai]} with Brutus and Cassius. But what do you say? --- for I know you are close with the Othones. He claims he is making a match with Iulia, since the divorce is now settled. The father asked me what her reputation was. I said I had heard nothing in particular --- I did not know why he was asking --- except concerning her face and her father. ``But to what end?'' I said. He answered that his son was keen. Then I, though I was disgusted \textit{[Greek: ebdelyttom\=en]}, nevertheless denied that the probabilities were true. The aim \textit{[Greek: skopos]} (that is what it is) of our young man is to give nothing to his own people; but she, however, is not on that account \textit{[Greek: ou para touto]}. I suspect, though, that he is rambling, as is his habit. But I should like you to ask --- you can do it easily --- and let me know.

\ciceroSection{3} I beg you, what is this? After the letter was already sealed, the people from Formiae who had been dining with me said that on the day before I was writing this, that is the third day before the Nones, they had seen this Plancus, our friend from Buthrotum, dejected, without his medallions; and that his slaves were saying that he and the squatters had been thrown out by the Buthrotians. Well done! But, I beg you, write me out the whole business in full.
