% Ad Atticum 15.1 (ad-atticum-15-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What a wretched business about Alexio! It is incredible how much grief it has caused me --- and not, by Hercules, chiefly on the ground that most people are with me, namely, ``To what physician, then, will you take yourself?'' What use have I now for a physician? Or, if I do need one, is there such a dearth of them? It is his affection toward me, his cultivated bearing, his sweetness of temper that I miss. There is this too. What is there that is not to be dreaded, when a man so temperate, so consummate a physician, has been overwhelmed by so sudden and so heavy an illness? But against all this there is one consolation: that we are born on these terms, that we ought to refuse nothing which can befall a man.

\ciceroSection{2} About Antony I wrote you before that I had not met with him. For he came to Misenum while I was at the Pompeian villa. From there he set out before I knew he had arrived. But by chance, just as I was reading your letter, Hirtius was with me at the Puteolan villa. I read it to him and pleaded the case. First, as to what concerns \dots\ he conceded me nothing; then, in sum, he was setting me up as arbiter not only of this matter but of his whole consulship. With Antony, however, I shall manage things so that he sees clearly: if he gives me satisfaction in this business, I shall be entirely his. Dolabella, I hope, is at home.

\ciceroSection{3} Let us return to our own men. About them you indicate that you have good hope, on the strength of the humane tone of their edicts. I, however, when on the seventeenth before the Kalends Hirtius set out from me at Puteoli for Naples to meet Pansa, saw straight through to the whole of his sentiment. For I drew him aside and urged him toward peace. He could not, of course, deny that he wanted peace; but he feared our men's arms no less than Antony's, and yet (he said) both sides had reason enough to keep a bodyguard, while he himself dreaded the arms of either. In short: \textit{[Greek: ouden]} (``nothing'').

\ciceroSection{4} About young Quintus I am of your mind. His father at any rate found your charming letter most welcome. As for Caerellia, I satisfied her easily enough; she did not seem to me to be much exercised about it, and even if she were, I certainly should not be. But the other woman whom you write was vexed at hearing of it --- I am altogether amazed that you ever heard such a thing. For where I praised her among friends in the hearing of her three sons and your daughter, \textit{[Greek: to ek toutou]} (``the consequence'') --- ``what is this? and why am I to walk about in a mask? Is the mask of old age itself not foul enough?'' What Brutus asks --- that I come before the Kalends --- he has written to me too, and perhaps I shall do it. But I do not in the least see what he wants. For what counsel can I bring him, when I myself am in want of counsel, and when he has consulted his own immortality better than our peace? About the queen, the rumour is dying down. About Flamma, I implore you, anything you can do.
