% Ad Atticum 15.4 (ad-atticum-15-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the ninth before the Kalends, at about the tenth hour, a courier came from Q.~Fufius. Some little scrap of a note from him, asking me to restore myself to him: silly, as is his way --- unless, perhaps, everything that you do not love seems silly. I wrote back in such terms as I think you will approve. The man delivered to me two letters from you, one of the eleventh, the other of the tenth. To the more recent first --- and the milder. I praise; but if Carfulenus too is involved, \textit{[Greek: an\=o]} (``up'' --- ``up in the air''). Antony's plans, as you tell them, are turbulent. Would that he were acting through the people rather than through the Senate! Which indeed I believe he is. But the whole of his project seems to me to be aimed at war, if in fact the province is being torn from D.~Brutus. However I rate the sinews of that man, it does not seem possible without war. But I do not desire it, since the Buthrotians are being seen to. You laugh? \dots\ I grieve that it has not rather been brought through by my own assiduity, my own diligence, my own influence.

\ciceroSection{2} As to your writing that you do not know what is to be done by our men, that perplexity (\textit{[Greek: aporia]}) has long since been troubling me. And so the consolation of the Ides of March is now a foolish one. For we used manly spirits and, believe me, boyish counsels. The tree was cut down, not torn up by the roots. You see, accordingly, how it sprouts again. Let us return then --- since you so often invoke them --- to the \textit{Tusculan Disputations}. Saufeius let us keep in the dark about you; I shall never give him the hint. As for your writing that Brutus is asking to be informed on what day I shall come to the Tusculan villa --- as I wrote you before, the sixth before the Kalends, and at that I should be glad to see you there as soon as possible. For I think we ought to go to Lanuvium, and indeed not without much conversation. But \textit{[Greek: mel\=esei]} (``it will be a care to me'').

\ciceroSection{3} I come back to the earlier letter. From it I pass over those first things about the Buthrotians; which are shut up in the marrow of my bones, only let there be, as you write, occasion to act. About Brutus's speech, you are plainly straining your point when you press it on me again at such length. Am I to take up the case which he himself wrote? Am I to write unasked by him? No \textit{interference} (\textit{[Greek: parencheir\=esis]}) could be more insulting. But, you say, some \textit{Heraclidean piece} (\textit{[Greek: H\=erakleideion]}). That, I do not refuse; only the argument must be composed, and a riper occasion for writing waited for. For you may think of me as you will (and I should wish, indeed, as well as possible) --- if these things go on streaming out as they seem to (you will bear what I am about to say), the Ides of March do not delight me. For that man would never have returned; fear would not have driven us to confirm his acts; or --- to fall back on the line of Saufeius and let go the \textit{Tusculan Disputations} to which you exhort even Vestorius --- we were so much in his good graces that, since the master being killed we are not free for our years, that master ought not to have been fled from. I blush, believe me; but I had already written it, and I would not delete.

\ciceroSection{4} About Menedemus I could wish it had been true; about the queen I wish it may be true. Other things face to face --- and chiefly what is to be done by our men, what indeed by us, if Antony is going to besiege the Senate with soldiers. If I had given this letter to his courier, I was afraid he would break the seal. So I have sent it by my own. For your own letters needed answering.
