% Ad Familiares 16.21 (ad-familiares-16-21) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO F. TIRONI SVO DVLCISSIMO S.}

\ciceroSection{1} I had been waiting eagerly for the letter-carriers day after day, and at last they came --- on the forty-sixth day after they left you. Their arrival was the most welcome thing in the world; for though I had taken the greatest delight in the letter from my kind, dearest father, your own letter, my sweet Tiro, brought my joy to its full measure. So I no longer regretted the pause in my writing; on the contrary, I was glad of it. For from the silence of my pen I have reaped a great harvest of your kindness. So I am profoundly glad that you accepted my excuse without question.

\ciceroSection{2} The reports about me that reach you are agreeable and welcome to you --- of that I have no doubt, my sweetest Tiro --- and I shall make good on it and strive that this nascent good opinion of me may every day be doubled and doubled again. So when you promise to be the trumpeter of my reputation, you may do so with a firm and steady mind. For the errors of my youth have brought me such pain and torment that not only does my mind shrink from those acts, but my very ears recoil from any mention of them. That you have shared in that distress and that pain I know for certain, and have proof of it; nor is it any wonder, since you wanted everything to turn out well for my sake --- and so also for your own; for I have always wanted you to be partner in my fortunes.

\ciceroSection{3} Since then you suffered through me, I shall now make sure that your joy through me is doubled. Let me tell you: with Cratippus I am bound not as a pupil to his teacher but as a son to his father; for I take pleasure in hearing him lecture, and I am also caught up in the personal charm of the man. I spend whole days with him, and often a good part of the night; I beg him to dine with me as often as he can. With this habit once introduced, he often slips in unannounced while we are at dinner, and, laying aside the gravity of philosophy, he jokes with us in the most genial way. So make every effort to come and see this remarkable man, so delightful and so distinguished, as soon as you possibly can.

\ciceroSection{4} And what shall I say of Bruttius? I never let him leave my side. His way of life is frugal and strict, and at the same time his company is the most agreeable in the world; for in him a sense of fun is not divorced from \textit{[Greek: philologia]} (the love of letters) and our daily \textit{[Greek: suz\=et\=esis]} (joint inquiry). I have rented a place for him in the immediate neighbourhood, and as best I can, out of my own narrow means, I help him keep up his slender resources.

\ciceroSection{5} Besides this, I have started to declaim in Greek with Cassius, and I want to do my Latin exercises with Bruttius. I keep up daily, intimate company with the friends Cratippus brought with him from Mytilene, learned men whose worth he himself thoroughly approves. Epicrates, too, the leading man of Athens, is very often with me, and Leonides, and the rest of that circle. \textit{[Greek: ta men oun kath' h\=emas tade]} (so much, then, for things on our side here).

\ciceroSection{6} As for what you write me about Gorgias: he was indeed useful in daily declamation, but I have put everything else aside in order to obey my father's instructions; for he had written \textit{[Greek: diarrh\=ed\=en]} (in so many words) that I should dismiss the man at once. I did not want to drag my heels, lest excessive eagerness on my part should raise some suspicion in his mind; and besides, it also struck me that it would be a serious thing for me to pass judgement on my father's judgement.

\ciceroSection{7} Still, your concern and your advice are gratifying and welcome to me. The excuse you offer for the shortness of your time I accept; I know how busy you usually are. I am profoundly glad you have bought an estate, and I hope the thing turns out happily for you. (Do not be surprised that I should congratulate you at this point: you wrote and told me of your purchase at much the same point in your letter.) There it is: you must lay aside your city airs; you have become a Roman countryman. How clearly I can see you now before my eyes, the picture of you at your most delightful --- I seem to be watching you buying farm gear, talking with the bailiff, saving fruit-seeds from the second course in a corner of your cloak. But as for the matter that touches the heart of it: I grieve no less than you do that I was unable to help you at the time. But do not doubt, my dear Tiro, that I shall lift you out of it, if only fortune lifts me --- especially since I know that this estate has been bought for the two of us in common.

\ciceroSection{8} I am grateful for the trouble you have taken over my commissions; but I beg you to have a copyist sent to me as quickly as possible, a Greek one above all; for I lose a great deal of labour in transcribing my \textit{[Greek: hupomn\=emata]} (notebooks). Above everything else, please look after your health, so that we may be able to do our \textit{[Greek: sumphilologein]} (joint study) together. I commend Anteros to you.


