% Ad Familiares 12.2 (ad-familiares-12-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CASSIO S.}

\ciceroSection{1} I am intensely glad that my opinion and my speech meet with your approval. If only it were permitted to me to use that liberty more often, there would be no difficulty in recovering our freedom and our state. But this deranged and ruined man, far more worthless than the very man you called the most worthless ever killed, is looking for a pretext to begin a slaughter, and charges me with having been the instigator of Caesar's killing for no other reason than to incite the veterans against me. That danger I do not very greatly fear, so long as he makes the glory of your deed common with the credit of mine. So now there is no safety in coming to the Senate, neither for Piso, who first attacked him with no one rising in support, nor for me, who did the same thirty days later, nor for P. Servilius, who came after me. For the gladiator is looking for blood, and he reckoned to make a start of it with me on 19 September; he had come prepared, after several days of rehearsal at the villa of Metellus. But what rehearsal could there be amid orgies and wine? And so, as I wrote you before, he seemed to everyone, in his usual way, not to be speaking but vomiting.

\ciceroSection{2} As for what you write --- that you trust some progress can be made through our authority and eloquence --- some has indeed been made, considering the magnitude of our troubles. The Roman people understand that there are three men of consular rank who, because they have thought rightly about the state and spoken freely, cannot come to the Senate in safety. Beyond that there is nothing for you to count on. Your kinsman is wrapped up in his new marriage-connection, and so has lost interest in the games; he is fit to burst at the unending applause for your brother. The other in-law, likewise, has been won over by Caesar's new memoranda. But these are tolerable. What is not to be borne is this: that there is a man who thinks his son will be consul in your year, and on that score parades his servitude to this brigand.

\ciceroSection{3} For my dear friend L. Cotta, in a kind of fated despair, as he himself says, scarcely comes into the Senate. L. Caesar, an excellent and most brave citizen, is kept away by ill health. Ser. Sulpicius, a man of the highest authority and the soundest views, is not present. The rest --- the consuls-elect excepted --- forgive me if I do not count among the men of consular rank. You have the counsellors of the public deliberation. Their number would be slender even in good times: what do you think it is in ruined ones? And so all our hope is in you; and if for that reason you are absent, that you may be safe, then it is not even in you. But if you are contemplating something worthy of your glory, I should wish it with ourselves alive; or, if not that, then the state will at any rate, through you and your friends, soon enough recover its rightful station. I am not failing your people and shall not fail them. If they refer any matter to me, my goodwill and my fidelity will be at your disposal. Farewell.
