% Ad Atticum 14.10 (ad-atticum-14-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Is it really so? This is what my Brutus --- and yours --- has brought about: that he should be at Lanuvium; that Trebonius should set off for his province by back-roads; that all of Caesar's acts, writings, sayings, promises, and intentions should carry more weight than if he himself were alive? Do you remember how I cried out on that very first day on the Capitol that the praetors ought to summon the Senate to the Capitoline? Immortal gods, what could have been achieved then, with every honest man rejoicing --- even the merely passably honest --- and the brigands broken! You blame the Liberalia. What could have been done then? We were already long since lost. Do you remember how you cried out that our cause was finished if he was given a public funeral? But in fact he was even burned and eulogized in the Forum, with tears, and slaves and beggars were set loose on our houses with torches. What followed? That they dared to say, ``You against Caesar's wish?'' --- I cannot bear these things and the like. So I am turning over in my mind \textit{one land after another} \textit{[Greek: g\=en pro g\=es]}; though yours is \textit{wind-borne} \textit{[Greek: hyp\=enemios]}.

\ciceroSection{2} Has your nausea now quite passed off? At least from your letter I gathered as much. I come back to the Tebassi, the Scaevae, the Fangones. Do you suppose these men believe they will keep their holdings while we stand? --- men in whom they reckoned more courage than they have found in practice. Lovers of peace, no doubt, these fellows, and not promoters of brigandage. But when I wrote to you about Curtilius and the Sextilian estate, I was also writing about Censorinus, about Messalla, about Plancus, about Postumus, about the whole tribe. It would have been better to have died in his killing --- which would never have happened --- than to see this.

\ciceroSection{3} Octavius arrived at Naples on the fourteenth day before the Kalends. There Balbus saw him early next morning, and on the same day was with me at the Cumean villa; the young man is going to accept the inheritance. But, as you write, there is a great \textit{rixothemin} \textit{[Greek: rhixothemin (text corrupt)]} with Antony. The Buthrotian business is mine, as it must be, and shall have my care. As for your question, whether the Cluvian estate is by now up to a hundred thousand: it seems to be getting there. In the first year, evidently, we have skimmed off eighty thousand.

\ciceroSection{4} Quintus the father writes me bitterly about his son, above all because the boy now indulges his mother --- when he had previously been hostile to her, who had deserved well of him. He has sent me a letter blazing against him. As for what young Quintus is up to: if you know, and you have not yet left Rome, please write me; and by Hercules, anything else as well. I take huge pleasure in your letters.
