% Ad Atticum 15.6 (ad-atticum-15-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} When our Brutus had written to me, and Cassius too, that I should make Hirtius --- who up to now has been honourable (this I knew, and I had no confidence that he would be made any better by my authority; he is perhaps angrier with Antony, but to the cause he is most well-disposed) --- still I wrote to him, and to him I commended the standing of Brutus and Cassius. What he wrote back to me, I wanted you to know, in case you should happen to take the same view as I, that these men even now are afraid that perhaps our own people may have more spirit than they actually have.

\ciceroSection{2} HIRTIUS to his CICERO, greetings.

You ask whether I have gone back into the country by now. Am I, when everyone is on fire with action, to do anything sluggishly? I have actually gone away from the city; I have decided it more advantageous to be absent. I am writing these lines to you as I leave for my Tusculan villa. But do not think me so brisk as to come hurrying back by the Nones. For I see nothing now that requires our concern, since safeguards have been provided for so many years ahead. Brutus and Cassius --- I only wish what they so easily get out of you for me, they might equally well, through you, be talked out of: namely, that they form no hotter design! For you say they wrote that as men giving way --- giving way where, or in what?

\ciceroSection{3} Hold them back, I beg you, Cicero; do not let go to wreck all these things which by my faith are being utterly overthrown by plunderings, burnings, killings. Let them only be on their guard, if they fear anything; let them attempt nothing beyond that. By my faith, with the keenest counsels they will accomplish no more than with the most inactive ones --- so long as those are diligent. For these things that are now in flux are not, of themselves, of long duration; for inflicting harm at the present moment they have force. What you hope concerning them, write to me at my Tusculan villa.

\ciceroSection{4} You have Hirtius's letter. To him I wrote back that they are considering nothing hotter, and gave my word for it. This, such as it is, I wanted you to know. After the letter was sealed, Balbus wrote to me that Servilia has come back; that she gives her word they will not depart. I now await a letter from you.
