% Ad Atticum 14.13 (ad-atticum-14-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the seventh day at last a letter from you was delivered to me, sent on the thirteenth day before the Kalends; in which you ask, and even suppose I myself do not know, whether I take more pleasure in the hills and the view or in the \textit{seaside walk} \textit{[Greek: halitene\^i]}. The truth is, as you say, that the charm of each spot is so great that I am in doubt which should be set first. But \textit{the works of a lovely feast are not our concern; rather an evil all too great, Zeus-nurtured one, we look upon and fear --- in the balance whether to be saved or destroyed} \textit{[Greek: all' ou daitos er\=atou erga mem\=elen, alla li\=en mega p\=ema, diotrephes, eisor\=ontes deidimen; en doi\^ei de sa\=osemen \=e apolesthai]}.

\ciceroSection{2} For granted that you have written things great and gladdening to me about Decimus Brutus's arrival at his legions --- in which I see the greatest hope --- still, if there is going to be a civil war (and there certainly will be if Sextus stays under arms, as I know for certain he will), I do not know what we must do. For it will not now be permitted, as it was in Caesar's war, to go neither this way nor that. For whomsoever this party of the desperate believes to have rejoiced in Caesar's death (and we all of us bore our joy at it most openly), him they will count among their enemies; and that points toward the greatest slaughter. It remains for us to betake ourselves to the camp of Sextus, or perhaps of Brutus. A hateful business, foreign to our age, and the war's outcome uncertain. And somehow I can say to you, and you to me: \textit{my child, the works of war are not given to you; rather, pursue the lovely works of speech} \textit{[Greek: teknon emon, ou toi dedotai pol\=em\=eia erga, alla sy g' himeroenta meterche\^o erga logoio]}.

\ciceroSection{3} But chance may see to all this --- chance, which in such matters can do more than reason. As for us, let us look to what ought to lie within ourselves: that whatever happens, we bear it bravely and wisely, and remember that it has befallen mortals; and may letters, above all, but the Ides of March also, in no small measure console us.

\ciceroSection{4} Now take up my deliberation, which has me uneasy. So many considerations come to mind on either side. I set out, as I had decided, as legate to Greece: I appear to avoid in some degree the danger of impending slaughter, but I shall fall under some censure for having failed the state at so grave a time. If, on the other hand, I stay, I see indeed that I shall be in peril; but I suspect it may happen that I can be of service to the state. There are also private reasons: I am very strongly of opinion that, for confirming Cicero, it is useful for me to go there; nor did I have any other cause for my departure when I conceived the plan of being made legate by Caesar. So consider the whole matter, as you usually do, if you think any of it bears upon me.

\ciceroSection{5} I come back now to your letter. You write that there are rumours that I am going to sell what I have on the lake, and certainly hand over the smaller villa to Quintus, even at an inflated price, so that, as Quintus the son told you, the dowered Aquilia may be brought into it. As for me, I have no thought of selling --- unless I find something that pleases me more. And Quintus has no thought of buying at this time. He is sufficiently tormented by the debt of the dowry, for which he renders extraordinary thanks to Quintus Egnatius; and he shrinks so from taking a wife that he says nothing is more agreeable than a single bed. But enough of this too.

\ciceroSection{6} I come back, then, to the wretched republic --- or rather, to none. Marcus Antonius has written to me about the restoration of Sextus Clodius. How respectfully, so far as concerns me, you will see from his own letter (I have sent you a copy); how dissolutely, how disgracefully, and how perniciously --- so much so that one sometimes finds oneself wishing for Caesar back --- you will judge easily. For things which Caesar would never either have done or have allowed are now being produced out of his forged notebooks. As for me, I made myself the readiest accommodation to Antony. For he, since he has once admitted into his mind that whatever he wishes is permitted him, would have done it no less without my consent. And so I have sent you a copy of my own letter as well.
