% Ad Atticum 15.14 (ad-atticum-15-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the 26th I received a letter from Dolabella, of which I have sent you a copy. In it he said that he had done everything you wished. I wrote back to him at once and gave him thanks at length. But so that he should not wonder why I was doing the same thing twice over, I took this for my pretext: that I had been able to learn nothing of the matter from you yourself in person before. But why prolong it? The letter I sent ran as follows:

\ciceroSection{2} CICERO TO DOLABELLA, CONSUL, HIS FRIEND, GREETING. Earlier, when I had been informed by a letter of our friend Atticus of your supreme generosity and your supreme kindness to him, and when you yourself had also written to me that you had done what we wanted, I gave you thanks by letter in such terms that you should understand nothing could have been more welcome to me than what you had done. But after Atticus himself came to me at Tusculum for this one purpose, that he might thank you in my presence --- having discerned your exceptional and remarkable goodwill in the Buthrotian case and your singular love for him --- I could not contain myself from declaring the same thing to you more openly in this letter. For of all the regard and good offices toward me which are great indeed, know this, my dear Dolabella: the most ample and most welcome to me is that you have brought it about that Atticus should understand how much I love you and how much you love me.

\ciceroSection{3} For what remains: the Buthrotian cause and city, though it has been settled by you (and we are accustomed to protect our own benefactions), I should nevertheless wish that, having been received into your good faith and most earnestly commended again and again by me to you, it should be sheltered by your authority and aid. It will be enough protection for the Buthrotians in perpetuity, and you will have relieved Atticus and me of great care and anxiety, if for my honour's sake you will undertake this --- to wish them always defended by you. That you should do this I beg you very earnestly, again and again.

\ciceroSection{4} The letter written, I have given myself to my \textit{compositions} \textit{[Greek: syntaxeis]} --- which, I fear, your red-lead pencil will have to mark in many places. So adrift \textit{[Greek: meteoros]} am I, and so entangled in great thoughts.
