% Ad Atticum 15.9 (ad-atticum-15-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the evening of the fifth before the Nones, a letter from Balbus reached me: the Senate is to meet on the Nones to arrange that Brutus see to the buying and shipping of grain to the city from Asia, Cassius from Sicily. A wretched business! First, any province at all from that crowd; and then, if any, this \dag legate's-errand\dag\ of a province! And yet perhaps it is better than sitting by the Eurotas. But these things chance will steer. He says, too, that at the same time it will be decreed that the rest of the praetorians, like these, are to have provinces assigned. That is certainly better than the famous Persian Portico; for I would not have Lacedaemon \dag he reckoned as far off as Lanuvium\dag. You will say I am laughing at such matters? What am I to do? I am tired of weeping.

\ciceroSection{2} Immortal gods! What dismay the first page of your letter held me in! What was that business of armed men in your house? Still, I am glad this cloud at any rate passed off quickly. What you have done about your own embassy --- sad in itself, and a hard matter to take counsel on --- I am keenly waiting to hear; for there is no way out of it. We are encircled on every side by their forces. As for me, the letter of Brutus which you say you have read disturbed me so violently that, in want of counsel as I was already, I am slower now from sheer pain of mind. But more, when I have learned how things stand with you. Just now there was nothing to write, and the less so because I was in doubt whether you would even receive this letter; for it was uncertain whether the courier would manage to see you. I am keenly waiting for your letter.
