% Ad Familiares 12.16 (ad-familiares-12-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TREBONIVS CICERONI S.}

\ciceroSection{1} If you are well, it is well. I reached Athens on 22 May, and there saw, as I had most hoped, your son devoted to the best studies and of the highest reputation for self-restraint. What pleasure I have taken in this you can know even without my saying so: for you are not unaware how highly I value you, and how, in keeping with our very oldest and truest affection, I rejoice in all your blessings, even the smallest --- let alone in so great a good as this. Do not imagine, my dear Cicero, that I am saying this to flatter your ear. Of all who are at Athens there is no one more lovable than your young man (and so ours --- for nothing of yours can be set apart from me), no one more devoted to those arts which you yourself love most, that is, to the best. I therefore congratulate you, as I can do in all sincerity, and gladly too; and no less ourselves, that in him whom we must love whatever sort of man he turned out to be we have someone we can love gladly as well.

\ciceroSection{2} When he mentioned to me in conversation that he wished to see Asia, he was not only invited but begged by me to do so above all while I am holding the province. You should not doubt that I will show him your fatherly devotion, in affection and love. I shall also make it my care that Cratippus be there with him, so you need not imagine he will be on holiday in Asia from those studies for which your encouragement spurs him on. For we shall not let off urging him on --- already prepared, as I see, and well into his stride --- so that day by day he may go further in learning and in practice.

\ciceroSection{3} What you are doing in public affairs, when I send this letter, I do not know; I had been hearing of certain disturbances, which of course I hope are false, so that we may at last enjoy a quiet liberty --- which up to now has scarcely fallen to me at all. I have, however, snatched a little leisure in our voyage and put together a small gift for you on my usual plan: I have taken up a saying of yours pronounced to our great honour, rounded it off, and set it down for you below. If in these few verses I shall seem to you in certain expressions a touch \textit{more outspoken} \textit{[Greek: euthyrrh\=emonesteros]}, the disgrace of the person against whom I inveigh somewhat freely will be my defence. You will also forgive me a wrath that is justified against men --- and citizens --- of that stamp. Besides, why should Lucilius have been allowed more licence in such things than I am? --- since, even if his hatred for those he attacked equalled mine, certainly the men he picked were no more deserving objects of so much freedom of speech.

\ciceroSection{4} You, as you promised me, will admit me as soon as possible to your conversations; for I do not doubt that, if you write anything about the death of Caesar, you will not allow me to bear the smallest share of the deed --- nor of your affection. Farewell, and take care of my mother and the rest of mine for me. Sent on 25 May, at Athens.
