% Ad Atticum 15.11 (ad-atticum-15-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I reached Antium on the sixth before the Ides. Brutus was glad of my coming. Then, before a numerous audience --- Servilia, Tertulla, Porcia --- he proceeded to ask what I thought best. Favonius was there too. What I had been turning over on the road was to urge him to take the Asiatic grain commission: nothing was left for us to do but to keep him safe; in that lay the protection of the commonwealth itself. I had got into this speech when Cassius came in. I went over the same ground. At this point Cassius, with eyes flashing (you would have said he was breathing the war-god) declared he would not go to Sicily. Was he to take an insult as a favour? \enquote{Then what are you doing?} I said. He said he would go to Achaea. \enquote{And you, Brutus?} I said. \enquote{To Rome,} he said, \enquote{if you think it right.} I said: not in the least; you will not be safe there. \enquote{What if I could be? Would you approve?} My view was that he should not go to a province at all, either now or after his praetorship; but I would not advise him to entrust himself to the city. I gave the reasons why he would not be safe there --- reasons which are certainly already running through your own mind.

\ciceroSection{2} Then in a long discussion they complained --- and Cassius most fiercely of all --- of opportunities lost, and accused Decimus bitterly. I said one should not go back over what was done; still, I agreed. And when I went on to say what ought to have been done --- not anything new, but what everyone says every day --- and without touching on the head of the matter, namely that others besides should have been struck, but that the Senate should have been summoned, the people, already ablaze with enthusiasm, more violently roused, the whole commonwealth taken in hand --- your dear friend cries out, \enquote{Now that I have never heard anyone say!} I checked myself. But Cassius seemed to me likely to go (for Servilia was promising she would see to it that the grain-commission was struck out of the senatorial decree), and ours was quickly knocked off the empty notion he had \dag said he wanted to embrace\dag. He resolved, then, that the games should go on under his name in his absence. From Antium, however, he seemed to want to set out for Asia.

\ciceroSection{3} In short: nothing in that journey gave me any pleasure beyond the consciousness of my duty. For it would not have been right to let him leave Italy without my having come to him first. With this debt of love and obligation discharged, what was left was for me to say to myself --- \textit{[Greek: hē deur' hodos soi ti dynatai nyn, theoprope?] What does this journey of yours hither avail you now, prophet?} I found the vessel altogether broken up --- shattered, rather. No plan, no method, no order. So even though I had not been in doubt before, all the more now I am bent on flying away from here, and as soon as possible, somewhere where I shall hear neither the deeds nor the name of the Pelopidae.

\ciceroSection{4} And listen --- in case you happen not to know: Dolabella has appointed me his legate on the third before the Nones. I was told of it yesterday evening. The votive embassy did not please even you; and indeed it was absurd that the vows I had made when the commonwealth still stood I should be discharging now that it lies in ruins. And free embassies, I think, have a term fixed by the Lex Iulia (Caesar's law on legates' travel) which cannot easily be extended. What I want is the kind of embassy which lets one come and go as one pleases; and that is just what has now been granted me. And it is a fine licence of right, this five-year span. Though why should I be thinking of five years? The business looks like contracting. But let us put away \textit{[Greek: blasphēma] ill-omened words}.
