% Ad Familiares 11.28 (ad-familiares-11-28) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MATIVS CICERONI S.}

\ciceroSection{1} I took great pleasure from your letter, because I learned that you hold the opinion of me that I had hoped and wished for; and although I had no doubt of it, still, because I valued it most highly, I was anxious that it should remain unspoiled. I was conscious, moreover, that nothing had been done by me that could offend the feeling of any good man. So much the less could I believe that you, adorned with so many and such excellent accomplishments, could have been persuaded of anything rashly --- and least of all about a man toward whom my goodwill had been, and still was, ready and unbroken. Since I know that this is as I wished it, I will answer the charges which you, as was fitting given your singular kindness and our friendship, have so often resisted on my behalf.

\ciceroSection{2} For I know well what they have brought against me since Caesar's death. They count it against me as a fault that I take hard the death of a man who was bound to me, and that I am indignant that a man I loved has perished; for they say that one's country is to be set above friendship --- as though they had already proved that his death was useful to the commonwealth. But I will not argue cunningly: I confess that I have not arrived at that pitch of wisdom. For it was not Caesar I followed in the civil dissension, but a friend; and though I was offended by the cause, still I did not desert him; nor did I ever approve of the civil war, or even of the ground of the dissension, which I strove with all my might to smother while it was still being born. And so, when a man bound to me had won, I was caught by no sweetness of office or of money --- rewards which others, who had less weight with him than I, abused without measure. Indeed my own estate was diminished by a law of Caesar's, by whose benefit very many of those who rejoice at Caesar's death have remained in the state. I labored for sparing the conquered citizens as much as for my own preservation.

\ciceroSection{3} Can I, then, who wished all men safe and sound, fail to be indignant that the man through whom this was obtained has perished? --- especially when the very same men were both his envy and his ruin. ``You will be punished, then,'' they say, ``since you dare to disapprove of our deed.'' What unheard-of arrogance! That some may glory in a crime, while others may not so much as grieve unpunished! Yet these things have always been free even to slaves: that they might fear, rejoice, grieve at their own discretion rather than another's --- the very things which now, as those men keep proclaiming, the ``authors of liberty'' are trying to wring from us by fear.

\ciceroSection{4} But they accomplish nothing. By the terrors of no danger will I ever swerve from duty or from humanity; for I have never thought an honorable death to be shunned, but often even to be sought out. Yet why are they angry with me, if I wish that they may come to repent of their deed? For I desire that Caesar's death be bitter to all. ``But I ought, as a citizen, to wish the commonwealth safe.'' That I do desire this --- unless my past life and my remaining hope prove it while I am silent --- I do not ask to win by speaking.

\ciceroSection{5} Therefore I ask you all the more earnestly to count the fact weightier than the argument, and to believe me, if you feel that to do right is to one's advantage, that there can be no fellowship with the wicked. Or shall I now, with my age running headlong, change and unweave myself --- I, who as a young man held to my course when I might even have erred with excuse? I will not do it, nor will I commit anything that could give offense, except that I grieve at the grievous fate of a man most closely joined to me and of a most distinguished person. But if I were otherwise minded, I would never deny what I was doing, lest I be judged both wicked in my wrongdoing and timid and false in concealing it.

\ciceroSection{6} ``But I superintended the games which the young Caesar held for Caesar's victory.'' Yet that pertains to a private duty, not to the condition of the commonwealth; and it was a service which I owed to the memory and honor of a most beloved man, even now that he is dead, and which I could not refuse to a young man of the highest promise, most worthy of Caesar, when he asked it.

\ciceroSection{7} I have also gone often to the house of the consul Antony to pay my respects; and those who think me too little a lover of my country you will find coming there in crowds, to ask for something or to carry off something. But what arrogance is this? Caesar never interfered to keep me from associating with whom I pleased, even with men he himself did not care for; yet those who have torn my friend from me try, by carping at me, to bring it about that I may not care for whom I please.

\ciceroSection{8} But I have no fear either that the moderation of my life will hereafter prove too weak against false rumors, or even that those who do not love me on account of my constancy toward Caesar will not prefer to have friends like me rather than like themselves. For my part, if what I wish comes to pass, I shall spend what remains of my life in peace at Rhodes; but if some chance interrupts me, I shall be at Rome in such a way that I shall always wish what is rightly done. To our friend Trebatius I give great thanks, for laying open your sincere and friendly feeling toward me, and for making it that I should be the more rightly bound to esteem and respect a man whom I have always gladly loved. Farewell, and hold me dear.
