% Ad Atticum 16.13 (ad-atticum-16-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What a remarkable coincidence! On the fifth before the Ides, having risen before dawn at the Sinuessanum and come at first light to the Tirenus bridge by Minturnae, where the road bends off toward Arpinum, your courier ran into me --- caught me ``pondering a long voyage'' \textit{[Greek: dolichon ploon hormainonta]}. ``Come,'' I said at once, ``hand over anything from Atticus.'' We could not read yet --- we had let the lamps go and it was not light enough. But when it grew light, the earlier of your two letters, written first, was read to me. The most graceful of any --- may I not be well if I write otherwise than I feel: I have read nothing more human. So I shall come where you call --- only with you for an aide. But nothing seemed at first more off-key \textit{[Greek: aprosdionuson]} to me than that to the letter in which I had asked your advice you should write back to me as you did.

\ciceroSection{2} And now look --- the second, in which you urge me ``past windy Mimas'' \textit{[Greek: par' ēnemoenta Mimanta]} --- that is, keeping the Appian Way on my left \textit{[Greek: ep' arister' echonta]}. So I stayed that day at Aquinum --- a longish stretch and a bad road. Setting out next morning I have despatched this letter.

\ciceroSection{3} * * * And indeed it was much against my will that Eros's letter made me let him go. Tiro will tell you the matter. You will see what is to be done. Apart from that, whether I may come closer --- for I would rather be at the Tusculanum or somewhere in the suburbs --- or whether you think I should withdraw even further, please write to me often. There will be someone every day to whom you can hand a letter.

\ciceroSection{4} As for the other matter on which you consult me, what I would judge you should do --- that is hard, being away. Still: if they are evenly matched, sit still; if not, the trouble will spread further --- and to us too; then we act together.
