% Ad Familiares 12.23 (ad-familiares-12-23) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CORNIFICIO S.}

\ciceroSection{1} Tratorius has laid out for me the whole shape of your command and the state of your province. So many intolerable things, in every quarter! But the greater your standing, the less to be borne what has befallen you; for what you tolerate with restraint, in virtue of your greatness of spirit and intellect, must not on that account go unavenged by you, even if it is not to be lamented. But of these matters later. The city's gazette, I am sure, is being sent on to you.

\ciceroSection{2} Were that not so, I would write it all out myself --- and chief of all, the venture of Caesar Octavianus. The multitude regard the charge as concocted by Antonius to give him a pretext for going after the young man's money; the prudent and the good not only believe it was done but approve. In short, there is great hope in him; there is nothing one cannot expect him to do for the sake of praise and glory. Antonius, on the other hand --- our familiar friend --- perceives himself to be so hated that, though he has caught his would-be assassins in his own house, he does not dare bring the matter forward. On the sixth day before the Ides of October he set out for Brundisium to meet the four Macedonian legions, which he was planning to win over with money, lead up to the city, and plant on our necks.

\ciceroSection{3} There you have a sketch of the commonwealth --- if a commonwealth can exist in a camp. For your part in this I often grieve, since for the whole of your generation you have not been able to taste any portion of a sound and undamaged commonwealth. And before now there was at least room to hope; now even that has been torn away. For what hope is there, when Antonius has dared to say in public meeting that Cannutius is looking for a place among men in whose company, while he is alive, no place can exist for him in the state?

\ciceroSection{4} For my own part, I bear these things and everything that can befall a human being so as to feel great gratitude to philosophy, which not only draws me away from anxiety but also arms me against every onset of fortune; and I think you should do the same, and reckon nothing in the catalogue of evils that is free from any fault of one's own. But these things you know better. Tratorius I have always thought highly of; in your affairs above all I have come to know the height of his fidelity, diligence, and prudence. Take care of your health; that you can do nothing for me more welcome than this.
