% Ad Atticum 16.3 (ad-atticum-16-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} You acted wisely --- and now at last I am answering the letter you sent me after your meeting with Antony at Tibur --- wisely, then, in giving him your hand and in volunteering thanks besides. For certainly, as you write, we shall be deserted by our state sooner than by our private estate. As to what you write, that ``\textit{o Tite, si quid}'' --- if anything --- pleases you more and more, you increase my zeal for writing. That you say you are awaiting Eros not without a little present too --- I am glad the matter has not disappointed your expectation. But I have also sent you the same treatise \textit{[Greek: syntagma]} reworked, and indeed the original itself \textit{[Greek: archetypon]} too, with many places overwritten and patched. Read this version, transcribed onto a large sheet, to your dinner-guests privately, but, as you love me, only when they are cheerful and well-fed --- lest they vent their irritation on me when they are really annoyed with you.

\ciceroSection{2} About young Cicero, I hope it is as I hear. About Xeno I shall find out face to face --- though I think nothing of his has been either careless or ungenerous. About Herodes I shall do as you instruct, and what you write of I shall learn from Saufeius and from Xeno.

\ciceroSection{3} About young Quintus, I am glad my letter was delivered to you by my courier before any from him --- though nothing would have escaped you. Still --- well. But I am waiting to hear what he said to you and you to him in return, and I have no doubt each behaved as he is wont. I hope, though, that Curius will deliver his letter to me. Lovable as he is on his own account and dear to me, still your warm recommendation will add a great heap to it.

\ciceroSection{4} Your letter is now answered enough. Now hear what, though I see no need to write it, I write anyway. Many things move me at this departure --- in the first place, by Hercules, that I am being separated from you. The labour of the voyage, too, moves me, foreign to my age and to my standing, and the timing of the departure has something of the absurd. For we are leaving peace to return to war, and the time we might have spent in our little estates, prettily built and amply pleasant, we are spending in voyaging abroad. The consolations are these: either we shall do young Cicero some good, or at least we shall judge how much good can be done; and then, as I hope and you promise, you will join us. Should that come about, everything will be better for me.

\ciceroSection{5} But what most distresses me is the reckoning of my remaining debts. Although they are arranged, still, because Dolabella's name is among them and there are names unknown to me in the assignment, I am thrown into confusion --- nothing distresses me more out of the whole business. Accordingly, I think I made no mistake in writing rather openly to Balbus, asking that, if anything of the kind should happen, that the names not match up, he should come to the rescue; and that I had also charged you that, if anything of that sort happened, you would consult him. This you will do if it seems right to you, and all the more so if you are setting off for Epirus.

\ciceroSection{6} I am writing this as I embark from the Pompeianum on three light boats, each of ten oars. Brutus was still at Nesis, Cassius at Naples. Are you fond of Deiotarus? --- and not fond of Hieras? --- who, after Blesamius came to me, and although he had been instructed to do nothing without our friend Sextius's opinion, referred neither to him nor to any one of us. I long to give our absent Attica a kiss. So sweet did the greeting from her, conveyed through you, seem to me. Give her my warmest greetings in return, and remember me, please, to Pilia.
