% Ad Familiares 11.5 (ad-familiares-11-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. D. BRVTO IMP. COS. DESIG.}

\ciceroSection{1} When our friend Lupus had come from you and was lingering some days at Rome, I was in the places in which I judged I could most safely be. So it came about that Lupus went back to you without a letter from me, although he had taken care that yours should be brought through to me. I came to Rome on the ninth of December, and held nothing before this --- that I should at once meet with Pansa; from whom I learned what about you I most desired to learn. As for that, you of course have no need of urging, if even in the matter which you have carried out --- the greatest within human memory --- you required no urger;

\ciceroSection{2} yet this much, it seems, must briefly be signified: the Roman people expects everything from you, and places in you all its hope of one day recovering liberty. If you remember day and night --- which I am certain that you do --- how great a thing you have done, surely you will not forget how great things you have still now to do. For if that man takes possession of the province --- I, who was always friendly to him before I came to see that he was making war upon the republic, not only openly but indeed gladly --- I see no remaining hope of safety. For which reason I beseech you with the same prayers with which the Senate and the Roman People beseech you, that you set the republic free for perpetuity from a king's domination, so that beginnings and outcomes may answer to each other.

\ciceroSection{3} This is your task, your part; this from you the state --- or rather all the nations --- not merely expects but demands. Although, since you do not need urging, as I wrote above, I will not use more words on it; I will do what is mine to do, which is to pledge you all my offices, my zeal, my care, my thought, in what shall bear on your praise and glory. For which reason, I would have you so persuade yourself: for the sake of the republic, which is dearer to me than my own life, and also because I favour you yourself and would have your standing magnified, I shall fail your most excellent counsels in no point in respect to your dignity and your glory.
