% Ad Atticum 14.19 (ad-atticum-14-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the Nones of May, while I was at the Pompeian villa, I received two letters from you, one on the sixth day, the other on the fourth. To the earlier one first, then. How welcome to me that Barnaeus delivered the letter to you at the right moment! With Cassius, by all means, as in everything else. And how conveniently it fell out that the very thing you urge upon me I had written to him four days earlier --- and I had sent you a copy of my letter! But while I was sunk in deep despair by Dolabella's \textit{†aritia†} (so you had written to me), look here come letters from Brutus too and from yourself! He is meditating exile. As for me, what I had in view was another harbour, one nearer for a man of my age --- though I should have preferred, to be sure, to put in there with Brutus in his prime and with the state set on its feet again. But as things now stand, as you write, it is not a matter of choice. For you agree with me: men of our age have no business with a camp, and least of all a camp of civil war.

\ciceroSection{2} Antony has written back to me on the Clodius business only this much: that my leniency and clemency are pleasing to him and will be a great satisfaction to me. About Pansa, on the other hand, he seems to be in a rage --- both about Clodius and likewise about Deiotarus --- and speaks sternly, if one cares to believe it. But here is something not nicely done, as it seems to me at any rate: that he sharply disapproves of what Dolabella did.

\ciceroSection{3} About the garlanded ones: when your sister's son had been accused by his father, he wrote back that he had worn the garland in honour of Caesar, that he had set it aside in mourning, and lastly that he was happy to face reproach for loving Caesar even dead.

\ciceroSection{4} I have written carefully to Dolabella, in just the way you tell me you would like. I have written to Sicca too; I do not impose this burden on you. I do not want you to have him angry. The line Servius is taking, I understand --- in which I see more of fear than of judgment. But since we are all of us terrified, I am with Servius. Publilius has been playing tricks on you. For Caerellia was sent to me here, an envoy from that lot; whom I had no trouble convincing that what she asked was not so much disagreeable to me as not even permitted me. If I see Antony, I shall press carefully about Buthrotum.

\ciceroSection{5} I come now to the more recent letter --- though about Servius I have already written. That I am making something great of Dolabella's action (an exploit, \textit{[Greek: praxin]}): by Hercules, so it seems to me; nothing greater was possible in such a case and at such a time. But still, whatever I attribute to him I attribute on the strength of your letters. I agree with you, all the same, that his exploit (\textit{[Greek: praxin]}) will be greater if he pays me what he owes me. I should like Brutus to be at Astura.

\ciceroSection{6} As to your praising me for settling nothing about my journey until I see how all this will turn out: I am changing my mind. But nothing before I see you. That my dear Attica is thanking me about her mother delights me; to her I have made over the entire villa and the storeroom, and was thinking of seeing her on the fifth before the Ides. You will give Attica my love. We shall take diligent care of Pilia.
