% Ad Atticum 15.13 (ad-atticum-15-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the 25th I received two letters from you. I shall answer the earlier one first. I agree with you: let us neither lead the column nor bring up the rear --- but let us cheer it on. I have sent you the speech. Whether it should be kept under wraps or put out into the world I leave to your judgement. But when can I imagine the day on which you think it should be published?

\ciceroSection{2} The truce you write of I do not see how it can be made. Better is the \textit{policy of giving no answer} \textit{[Greek: anantiphonesia]}, which I think I shall employ. As for your news that two legions have come to Brundisium, you people have everything before us. So you will write whatever you hear.

\ciceroSection{3} I am waiting for Varro's dialogue. I already approve the \textit{Heraclidean} \textit{[Greek: Herakleideion]} form, particularly since you take such great delight in it; but I should like to know what sort you wish. As to what I wrote to you before --- or rather earlier (since you prefer it so) --- you have made me more eager to write \dag by telling me the truth, \dag for to your judgement, which I already knew, you have added the authority of Peducaeus, weighty indeed with me and singularly grave. I shall make every effort, then, that you should miss in me neither industry nor diligence. Vettienus, as you say, and Faberius I am cultivating. Of Clodius I think nothing malicious --- though --- well, let that be his affair. About preserving freedom, than which there is surely nothing sweeter, I agree with you. So, then, Gallus Caninius? What a worthless fellow! What else can I call him? And Marcellus a cautious man! I in the same way, but still not the most cautious of all.

\ciceroSection{4} I have answered the longer and earlier letter. Now what am I to answer to the shorter and nearer one except that it was most sweet to me? The Spanish business is going very well, if only I may see Balbillus safe and sound, the prop of our old age. The same with Annianus, since Visellia pays me close attention. But these are at any rate human affairs. About Brutus you say you know nothing, but that Servilia says M. Scaptius has arrived, and that he will come to her in secret \dag though not with any ceremony, \dag and that I shall learn everything; which I will pass on at once. Meanwhile she tells the same news --- that a slave of Bassus has come reporting that the Alexandrian legions are under arms, that Bassus is being summoned, that Cassius is being awaited. What more? It looks as though the Republic is going to recover her rights. But not so fast! You know how practised these men are in brigandage, and how insensate.

\ciceroSection{5} Dolabella is the best of fellows --- though just as I was writing, with the second course brought in, I was hearing that he had come to Baiae --- yet from his place at Formiae he wrote to me a letter which I received as I came out of the bath, saying that he had done the utmost about the assignment. He blames Vettienus (the man is dithering like a tradesman over coinage), but says that our friend Sestius has taken on the whole affair, an excellent man indeed and most devoted to me. I ask, though, what on earth Sestius can do in this business that any of us could not do. But if anything beyond hope comes of it, you will let me know; if, as I think, the matter is hopeless, you will write anyway, and the thing will not move me.

\ciceroSection{6} Here we philosophize \textit{[Greek: philosophoumen]} --- what else is there? --- and unfold magnificently the question of \textit{the proper duties} \textit{[Greek: ta peri tou kathekontos]}, and dedicate \textit{[Greek: prosphonoumen]} it to Cicero our son; for on what subject would a father rather address a son? Then other things. What more? The labour of this sojourn will leave something to show for itself. They thought Varro would come today or tomorrow; but I was hurrying on to my place at Pompeii, not that anything here is more beautiful, but that interrupters are less of a nuisance there. But write me out in full, please, what is the story of Myrtilus --- I have heard he has paid the penalty --- and whether it is clear enough by whom he was suborned.

\ciceroSection{7} While I was writing this, I was supposing the speech had just been carried through to you. Oh, how I dread your verdict on it! Though why should I? --- since it will not go out to the world until the Republic is restored. And what I may hope on that score I do not dare to write.
