% Ad Atticum 14.7 (ad-atticum-14-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the day after the Ides I saw Paulus at Caieta. He gave me some thoroughly wretched news about Marius and the state of affairs. From you, of course, nothing --- nor from any of my people. But I hear our Brutus has been seen at Lanuvium. Where in the end is he going to settle? Of this above all, as of the rest, I want full word. I am writing this on leaving the Formian villa on the seventeenth day before the Kalends, on my way to the Puteolan estate by the day after next.

\ciceroSection{2} From my son a letter, thoroughly \textit{[Greek: pepinomenai]}, well-polished, and a good long one. The rest can be feigned, but \textit{[Greek: pinos]}, polish in writing, marks the better-trained mind. Now I urgently ask of you what I spoke of with you lately: see that he wants for nothing. The matter touches our duty and our standing and our dignity --- and I understood that you take the same view. If, as I wish, I am off to Greece in the month of Quintilis, all will be easier; but since the times are such that nothing can be settled --- what is honourable for me, what is open to me, what is to my advantage --- please, see to it that we maintain him as honourably and as amply as we can. On this and the other matters that concern us, as you always do, think things over thoroughly, and write to me --- either on what bears on the business, or, if there is nothing, whatever first comes to mind.
