% Ad Atticum 16.6 (ad-atticum-16-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} So far (I have got as far as Vibo, to Sicca's place) my sailing has been more comfortable than brisk: largely under oar, with no following breezes. This one stroke of luck: there were two gulfs that had to be crossed, the Paestan and the Vibonensian, and we crossed both on an even keel. So I came to Sicca on the eighth day out from the Pompeianum, having stopped for one day at Velia, where indeed I stayed quite gladly at our friend Thalna's, and could not have been received more handsomely --- with Thalna himself away, no less. On the ninth day before the Kalends, then, to Sicca. There, as you may suppose, as though in my own house. And so I have lingered out the day after. But I had been thinking that, once at Regium, we would there work out the long voyage \textit{[Greek: dolichon ploon]} --- to Patrae by merchantman, or to Tarentine Leucopetra by light boat and then on to Corcyra; and, if by freighter, whether to push straight across the strait or by way of Syracuse. I shall write to you about this from Regium.

\ciceroSection{2} By Hercules, my dear Atticus, often I ask myself, oh \textit{[Greek: he]} why am I not with you? why am I not looking at the little eyes of Italy, my own small villas? But it is more than enough that I am not with you --- what am I running from? From danger? Of which now, unless I am mistaken, there is none. To that very thought your own authority calls me back; for you write that my setting out is being praised to the skies --- but only if I return before the Kalends of January; which I shall certainly strive to do. I would rather be at home in fear than at your Athens without it. Still, take note where matters are tending and either write to me or, what I should far prefer, bring word in person. So much for that.

\ciceroSection{3} I should be glad if you would take in good part my pressing on a matter I know is more your worry than mine. My obligations: in the name of the gods, settle them, clear them away. The remainder is a fine thing left to me --- but it needs care that what is owed to the co-heirs for the Cluvian property be paid by the Kalends of August. How to deal with Publilius you will judge. He must not press me, since we are not standing on our legal right; still, I want him too plainly paid off. As for Terentia --- what shall I say? Even before the day, if you can. Indeed, if I get off to Epirus quickly, as I hope, what I owe under bond I ask you to provide for in advance, settle in full, and leave discharged. But enough on this --- I am afraid you will think it too much.

\ciceroSection{4} Now hear of my carelessness. The book \textit{On Glory} I sent to you. But in it the preface is the same as the one in the third book of the \textit{Academica}. This happened because I keep a volume of prefaces. From it I am in the habit of selecting whenever I have set in hand a piece of writing \textit{[Greek: syngramma]}. And so, just now at Tusculum, not remembering that I had already used that preface, I tacked it onto the book I sent you. But when on shipboard I was reading the \textit{Academica}, I recognized my slip. I have at once drafted a new preface and am sending it to you. You will cut out the old one and paste the new in its place. Give my best to Pilia and to Attica, my delight and my love.
