% Ad Atticum 16.14 (ad-atticum-16-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I had nothing at all to write. For while I was at Puteoli, there was every day something new about Octavian, and much that was false about Antony as well. As to what you wrote (I had received three letters from you on the third day before the Ides), I agree with you strongly: if Octavian comes to count for much, the acts of the tyrant will be confirmed far more firmly than they were in the temple of Tellus, and the thing will tell against Brutus. But if he is beaten, you see how intolerable Antony will be --- so that one cannot tell which to wish for. What a worthless fellow Sestius's courier is! The day after he left Puteoli he said he would be at Rome. As for your advising me to go step by step, I agree --- though I was thinking otherwise. Philippus and Marcellus do not sway me. Their situation is different, and even if it is not, it looks so. But in this young man, although there is spirit enough, there is too little authority. Still, see whether, if I can manage to be safe at Tusculum, that would not be the better course. I would be there more gladly: I shall be ignorant of nothing. Or am I to stay here when Antony comes?

\ciceroSection{3} But, to pass from one thing to another: I have no doubt that what the Greeks call \textit{[Greek: kathēkon]} is in our language \textit{officium} --- duty. And why do you doubt that it would apply admirably to public affairs too? Do we not speak of the duty of the consuls, the duty of the Senate, the duty of a commander? It fits admirably; or give me a better word. You give bad news about Nepos's son. I am greatly moved, by Hercules, and bear it with distress. I had not known of the boy's existence at all. I have lost Caninius too --- a man, so far as I was concerned, not without gratitude. There is no need for you to urge Athenodorus on: he has sent me a pretty enough \textit{[Greek: hupomnēma]}. Help my catarrh, I beg you, by every means. The great-grandson of your grandfather writes to the grandson of my father that, from the Nones on which we performed our great deeds, he will open up the temple of Ops, and bring the matter before the people. Look into it, then, and write. I am waiting for Sestius's judgement.
