% Ad Atticum 16.1 (ad-atticum-16-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the Nones of July I arrived at my Puteolan villa. The next day, on my way to Brutus at Nesis, I wrote these lines. But on the day I had come, while I was at dinner, Eros brought me your letter. Really? On the Nones of July? By Hercules, what a crew they are! But one could be irritated all day long. Could anything be more disgraceful than ``July'' for Brutus? So back I come to my old refrain --- ``let it be still'' \textit{[Greek: et' e\=omen]}; I saw nothing.

\ciceroSection{2} But what, pray, is this I hear --- that the land-seekers at Buthrotum have been cut to pieces? And why is Plancus moving so fast --- for so I had it on rumour --- day and night? I am very keen to know what the matter is.

\ciceroSection{3} I am glad my departure is being praised. We must see to it that my staying gets praised too. As for the Dymaeans, driven from their land and making the sea unsafe, no wonder. To be sailing in convoy \textit{[Greek: en homoploia{[i]}]} with Brutus seems to offer some protection --- but I imagine the boats are small. Still, I shall know shortly, and write to you tomorrow.

\ciceroSection{4} On Ventidius, I take it to be a panic \textit{[Greek: panikon]}. On Sextus, the report was held to be sure --- \dag to arms\dag. If that is true, I see we must serve as slaves, but without a civil war. What then? On the Kalends of January, is our hope in Pansa? Much twaddle \textit{[Greek: l\=eros polus]}, in the wine and in the sleep of those fellows.

\ciceroSection{5} On the 210,000, excellent. Let young Cicero's accounts be sorted. For Ovius is just from there. He brings much that I could wish; among it, this is no bad item either --- \dag that on the orders, if he is well-supplied,\dag\ 72,000 sesterces is enough, more than enough, but Xeno is doling it out very meagerly and pinchingly \textit{[Greek: glischr\=os]} --- that is, in driblets. As for your having exchanged more than corresponded to the income of the islands, let that year carry it on which the travelling expenses fell. Henceforth, from the Kalends of April, let it be adjusted to 80,000 sesterces. For at present he has only the islands. We must see what is needed once he is at Rome --- for I think that mother-in-law of his is not to be put up with. I had refused Pindarus about the Cumanum.

\ciceroSection{6} Now, listen to why I have sent off the courier. Young Quintus promises me he will be a Cato. Both father and son begged me to stand surety to you for him --- but only on terms that you should credit it once you had seen for yourself. I shall give him a letter on his own terms. Don't let it move you. I am writing this so that you will not think it has moved me. Heaven grant he does what he promises: the joy would be common to us both. But I --- I will say no more. He is leaving here on the seventh before the Ides. For he says the assignment falls on the Ides, and that he is being pressed hard. From my letter you will judge how to answer him. More when I have seen Brutus and am sending Eros back. I accept dear Attica's excuse and love her dearly for it; give her my greetings, and Pilia's too.
