% Ad Familiares 10.4 (ad-familiares-10-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{PLANCVS CICERONI}

\ciceroSection{1} Your letter was most welcome to me, which I gathered from Furnius's account that you had written. For my part I offer, as my excuse for the past stretch, that I had heard you had set out, and learned that you were back not much before I read it from your letter. For I think I cannot let pass any duty toward you, even the smallest, without the greatest reproach to myself, in keeping up which I have very many grounds --- whether the bond of friendship with my father, or my own attentiveness from boyhood, or our mutual affection.

\ciceroSection{2} So, my Cicero --- as far as your years and mine permit --- be sure of this: that you are the one man in cultivating whom I have set up for myself an obligation as binding as toward my country. All your counsels therefore seem to me no fuller of prudence (and that is the highest) than of fidelity; which I measure by my own awareness. So even if I felt otherwise, your reminder would surely check me; or, if I were in doubt, your urging could carry me forward, to follow what you held best. As things stand --- what is there that could pull me to the other side? Whatever good things are mine, whether granted by the kindness of fortune or earned by my own labor --- though out of affection you have valued them more highly --- are even by the judgment of my bitterest enemy of such a kind that they seem to lack nothing except a good name.

\ciceroSection{3} So be sure of this one thing: that as far as I shall be able by my strength to strain, by counsel to foresee, by authority to admonish, all of this is to belong always to the republic. Your way of thinking is not unknown to me. Even if I had the option (a thing I myself should choose) of your being here in person, I should never depart from your counsels; and now too I shall not allow it that any act of mine should give you just ground for reproach.

\ciceroSection{4} I am on the watch for every development --- what is being done in Cisalpine Gaul, what at Rome in the month of January --- so that I may know. Meanwhile I bear here the very greatest anxiety and care, lest in the midst of others' misdeeds these peoples come to think our troubles are their opportunity. But if it falls out for me as I myself deserve, I shall surely satisfy you --- whom I most desire to satisfy --- and all good men besides. Keep well, and love me in return.
