% Ad Familiares 13.4 (ad-familiares-13-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. Q. VALERIO Q. F. ORCAE LEGATO PROPR.}

\ciceroSection{1} With the townsmen of Volaterrae I am bound by the closest of ties. They received a great service at my hands and have repaid me in heaped measure; never, in my honours or in my hardships, have they failed me. Even if no such ground stood between them and me, still --- because I love you most deeply and feel myself prized by you in equal measure --- I would both urge and exhort you to take thought for their fortunes, the more so because they have what is virtually an unanswerable claim to be left in possession of their right: first, because by the kindness of the immortal gods they escaped the bitterness of the Sullan time, and second, because they were defended by me in my consulship with the keenest zeal of the Roman People.

\ciceroSection{2} When the tribunes of the plebs had published a most iniquitous bill against their lands, I had no difficulty in persuading the Senate and the Roman People to will the preservation of those citizens whom Fortune had spared. C. Caesar in his first consulship ratified this act of mine by his Agrarian Law, and freed the territory and town of Volaterrae from every threat in perpetuity --- so that I do not doubt that he, who is at this moment binding fresh ties to himself, would have his earlier acts of generosity preserved. It is for your good judgement, then, either to follow the lead of that man whose party and command you have followed to your own great honour, or at the very least to reserve the whole question intact for him. But of this you ought to have no doubt: that you would wish to bind to yourself in perpetuity, by an act of the highest generosity, a town so weighty, so steadfast, so honourable as this one.

\ciceroSection{3} But what stands written above is meant to urge and to persuade; what remains belongs to the business of asking. I would have you reckon that I am not only giving you advice in your own interest, but also seeking and requesting from you what I myself need. You will therefore have done me the most welcome thing if you choose that the Volaterrans should be left intact and unharmed in everything. Their dwellings, their homes, their property, their fortunes --- preserved by the immortal gods and by the foremost citizens of our commonwealth with the keenest zeal of the Senate and the Roman People --- I commend to your good faith, your justice, and your goodness.

\ciceroSection{4} If, with the resources I once had, circumstances at this present hour granted me the means to defend the Volaterrans in the way I have been accustomed to defend my own, I would let slip no service, no struggle, in which I could be of use to them. But since I trust that at this hour my weight with you is no less than my weight has always been with all men, I ask of you, by the perfect bond between us and our equal and mutual goodwill, to deserve so well of the Volaterrans that they may conclude it was by a kind of providential design that the man set in charge of this business is the one man with whom we, their lifelong defenders, have most influence.
