% Ad Atticum 13.23 (ad-atticum-13-23) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} To your morning letter I replied at once yesterday; now I answer the evening one. I should have preferred Brutus to send for me. And it would have been more fitting, given that he had a journey before him both sudden and long --- and, by Hercules, now too, when we are in such a frame that we cannot quite live in company (for you surely understand on what point our living-together \textit{[Greek: symbiosis]} chiefly turns), I was glad enough to have us together at Rome rather than at Tusculanum.

\ciceroSection{2} The books for Varro were not being held back; for they are \dag{}finished\dag{}, as you have seen; only the copyists' slips are being weeded out. About these books you know I was in two minds, but let it be as you see fit. Likewise those I am sending to Brutus --- the copyists have them in hand.

\ciceroSection{3} My instructions, you write, are to be unpacked. Yet Trebatius says that those people are all relying on this withholding-clause. What do you make of such men? You know the household. So bring it off smoothly \textit{[Greek: euagogos]}. It is incredible how little I care about all that. I declare it to you with every kind of asseveration --- and I would have you believe me --- my small possessions are more a vexation to me than a pleasure. For I feel more pain at not having a man to leave them to than I take pleasure in having the means to use. And that Trebatius said he had told you; you for your part feared, perhaps, that I should be vexed to hear it. That was a kind thought; but, believe me, those things I no longer care about. So put yourself into talk, cut through and finish, and so speak with Polla as though you thought yourself a Scaevola talking with him --- and do not suppose that men who are accustomed to chase what is not even owed will let go of what is owed. About the date only, look to it, and that too in a decent way.
