% Ad Atticum 13.46 (ad-atticum-13-46) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Pollex, just as he had said, on the Ides of Sextilis --- or rather, the day before the Ides at Lanuvium --- presented himself; but plainly a thumb (\textit{pollex}), not a pointer (\textit{index}).

\ciceroSection{2} So you will learn it from the man himself. I met Balbus --- Lepta, in distress over his management of the games, had taken me to him --- at his Lanuvine place, the one he handed over to Lepidus. There first of all this: a little earlier I had received the letter in which he confirms strenuously that he will be back before the Roman Games. I read the letter through: much in it about my \textit{Cato}, which he says he has been made more copious by reading over and over; that after reading Brutus's \textit{Cato} he had thought himself eloquent.

\ciceroSection{3} Then from him I learned about the entry on Cluvius's inheritance (Vestorius is being negligent!): formal acceptance, with witnesses present, within sixty days. I had been afraid we should have to summon him. As it is, he must be sent so as to enter the inheritance at my instruction. The same task, then, falls to Pollex. About the Cluvian gardens too I dealt with Balbus. Nothing could have been more obliging. He will write to Caesar at once, he says, and that Cluvius leaves a legacy from T. Hordeonius and to Terentia 50,000 sesterces, and for the tomb and many other items, but nothing from us. Please reprove Vestorius a little. What could be less acceptable than that Plotius the perfumer, through his own slaves, should have got everything to Balbus so long in advance, while the man here gets none of it through to me even through my own? About Cossinius I am sorry; I loved the man.

\ciceroSection{4} To Quintus I shall consign anything left over after my own debt and after the purchases for which I think I shall have to take on yet more debt. About the house at Arpinum I know nothing. There is no reason to find fault with Vestorius. For now, when this letter had already been sealed, our courier arrived by night, and brought from him a carefully written letter and a copy of the will.
