% Ad Atticum 13.42 (ad-atticum-13-42) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The man himself came to see me \textit{[and:]} \textit{[Greek: kai]}. And I: \textit{``But you --- why so pensive?''} \textit{[Greek: su de de ti synnous?]} ``Do you ask?'' he says. ``A man with a journey ahead, and a journey to war --- and a war as dangerous as it is shameful!'' ``What's the trouble, then?'' I say. ``Debt,'' he says, ``and not even the travel-money to leave with.'' At this point I borrowed a little something from your style of speaking: I said nothing. But he went on: ``Still, what really tortures me is my uncle.'' ``Why's that?'' I say. ``He's angry with me.'' ``Why do you put up with it?'' I say --- for I prefer to put it that way rather than ``Why bring it on yourself?'' ``I shall not put up with it,'' he says: ``I shall take away the cause.'' And I: ``Very right; but, if it isn't a hardship, I should like to know what the cause is.'' ``Because, while I was hesitating which woman to marry, I did not satisfy my mother --- and so not him either. Now nothing is worth that much to me. I shall do as they want.'' ``Good luck to it,'' I said, ``and I commend you. But when?'' ``As for the timing,'' he says, ``that's nothing to me, since I approve the principle.'' ``But for my part,'' I said, ``I think you should do it before you set out. So you will have humoured your father as well.''

\ciceroSection{2} ``I shall do as you advise,'' he says. So ended the dialogue.

\ciceroSection{3} But, listen --- you know my birthday is the third before the Nones of January; you will be there, of course. I had already written this --- when, look: Lepidus is begging me to come. The augurs want, I gather, to have me there for the ritual marking-out of a temple. So I am to go. \textit{No garlic} \textit{[Greek: me skordou]}. I shall see you, then.
