% Ad Atticum 13.25 (ad-atticum-13-25) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About Andromenes, I had been supposing it just as you write. For you would have known and would have told me. Still, you write to me about Brutus in such a way that you say nothing about yourself. When, then, do you think he will be coming? For my part, I shall be at Rome on the day before the Ides. To Brutus I had meant to write in this sense (but, since you write that you read the letter, I was perhaps a little too obscure \textit{[Greek: asaphesteros]}) --- that I had gathered from your letter that he did not wish me to come to Rome now as if it were to escort him. But, since my coming is already at hand, see to it, I beg you, that the Ides do not in any way prevent him from being at Tusculum at his own convenience. For I was not going to need him for the auction (in such business why are you not enough by yourself?), but for the will I did want him there --- which I now prefer to put off to another day, lest I should seem to have come to Rome on that account. So I have already written to Brutus what I had supposed, that there is no need to make it the Ides. Steer the whole thing, then, in such a way that we obstruct Brutus's convenience in not even the smallest particular.

\ciceroSection{3} But what on earth is it that makes you shudder, because I order, at your own risk, that the books be given to Varro? Even now, if you have doubts, let me know. For nothing is more polished than they are. I want Varro to have them, especially since he himself wants them; but he is, as you know, a formidable man \textit{[Greek: deinos anēr]}: he might bring a charge even against one who has done nothing \textit{[Greek: tacha ken kai anaition aitiōōto]}. So I keep seeing his face before me complaining, perhaps even of this --- that my part is more amply defended in those books than his is, which, by Hercules, you will see is not the case, if ever you come to Epirus. For now we yield place to Alexio's letters. But for all that, I do not despair that the books will meet with Varro's approval; and since we have laid out the expense of large-format paper, I am quite content that this hold. Yet I say it over and over: it shall be at your risk. Therefore if you waver, let us cross over to Brutus; for he too is an Antiochean. O the flighty Academy, true to itself! now here, now there. But, I beg you, did my letter to Varro really please you? Damn me if I ever take such pains over anything again. So I did not even dictate it to Tiro, who is in the habit of running through whole outlines \textit{[Greek: periochas]} at a stretch, but to Spintharus, syllable by syllable.
