% Ad Familiares 5.13 (ad-familiares-05-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. L. LVCCEIO Q. F.}

\ciceroSection{1} Welcome though the consolation of your letter is in itself --- for it makes plain the highest goodwill joined to a like wisdom --- still the fruit I drew from it, the greatest of all, was this: that I came to know you as a man magnificently looking down on human things, and excellently armed and arrayed against fortune. That, I hold, is the supreme praise of wisdom --- not to hang from anything outside, not to keep the reckonings of one's living, well or ill, suspended on external things.

\ciceroSection{2} This thought had not gone clean out of me --- it had sunk deep within --- but it had nonetheless been somewhat shaken loose and torn at by the violence of the storms and the rush of disasters that came together upon me. I see that you are helping me to brace it up again, and I feel that you did so even in your last letter, and to great effect. So this must be said the more often, and not merely indicated to you but openly declared: nothing could have been more welcome to me than your letter.

\ciceroSection{3} As consolation, those things have their force which you gathered with such elegance and abundance; but nothing more than that I saw plainly the firmness and the seriousness of your spirit --- not to imitate which I count the most shameful of things. And so I think myself even braver on this score than you yourself, my teacher in bravery, in that you seem to me to hold some hope that these things will one day be better. The chances of the gladiatorial ring, those comparisons of yours, and the reasonings you gathered up in that disputation, forbade me to despair utterly of the commonwealth. So the one thing is less surprising --- that you, having something to hope for, are braver; the other is the surprise: that you are held by any hope at all. For what is there not so afflicted that you would not confess it has been wiped out and extinguished? Look round on all the limbs of the commonwealth, which are best known to you: you will surely find none that has not been broken or maimed. I would press through them, if either I saw them better than you do, or could call them to mind without pain --- although by your warnings and precepts all pain is to be cast off.

\ciceroSection{4} Therefore I will bear my private losses too, as you bid, and the public ones perhaps even somewhat more bravely than you yourself, who give the precept. For some hope, as you write, consoles you; we shall be brave even in the general despair --- since you, after all, are the man who urges and prescribes it. You give me, that is, pleasant recollections of our shared awareness of what we have done, and of those very deeds in which you above all were the prompter; for we rendered to our country no less, certainly, than we owed, and indeed more than was demanded from any man's spirit or counsel.

\ciceroSection{5} You will forgive my making some claim on my own behalf; the very things by the thought of which you wished to lift us from grief --- by the recalling of those same things we are soothed. So, as you advise, I shall draw myself off, as much as I can, from all vexations and anxieties, and turn my mind to those things by which prosperity is adorned and adversity helped along; and I shall be with you, as much as the age and health of each of us shall allow it --- and if we can be together less than we should wish, still by the joining of our minds and by our shared pursuits we shall enjoy each other so that we never seem not to be together.
