% Ad Familiares 9.8 (ad-familiares-09-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO VARRONI}

\ciceroSection{1} Even the public, no matter how openly someone displays a gift, does not commonly demand it unless worked up; yet I am moved by the anticipation of what you promised, not to demand, but to remind. I have sent you four reminders, however, not overly modest ones --- for surely you know the brashness of that younger Academy. From the very midst of it, then, I have sent these I have stirred up; I am afraid they may actually press you for it, though I have instructed them only to ask. I had been waiting, in any case, for some time now, and was holding myself back from writing you anything myself before I had received something, so that I might repay you with as similar a gift as possible. But since you were taking your time --- that is, in my interpretation, taking more care --- I could not be held back from declaring our partnership in studies and in affection in whatever literary form I could. So I have made a conversation held between us at the Cuman villa, with Pomponius present; I have given you the parts of Antiochus, which I had thought I understood you to approve; I have taken Philo's for myself. I suppose that when you read it, you will be amazed that we spoke between us of things we have never spoken of; but you know the convention of dialogues.

\ciceroSection{2} From now on, however, my Varro, let us have as much as possible of this kind of thing between us, and on ourselves --- though perhaps it is late in the day; but let Fortune answer for the times past on the Republic's account, and we must answer for these present ones ourselves. And how I wish that in quiet times --- and in some settled condition of the state, if not a good one, then at least a sure one --- we could pursue these studies of ours together! Although even then there would be other considerations to give us honorable cares and actions; but now, what reason have we to wish to live, if not for these? With them, life is hardly bearable for me; strip them away, and not even hardly. But all this in person, and more often. I hope the move and the purchase turn out happily for you, and I approve your plan in the matter. Take care of yourself.
