% Ad Atticum 12.40 (ad-atticum-12-40) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What Caesar's attack on Cato is going to look like in answer to my praise of him, I have made out from the book Hirtius has sent me, in which he assembles Cato's faults --- but with the greatest praises of myself. So I have sent the book on to Musca, that he may give it to your copyists; for I want it widely circulated. To make this the easier, you will give the order to your people.

\ciceroSection{2} I often try my hand at a piece of political counsel \textit{[Greek: sumbouleutikon]}. I can find nothing. And indeed I have by me the books of Aristotle \textit{[Greek: Aristotelous]} and of Theopompus \textit{[Greek: Theopompou]} addressed to Alexander \textit{[Greek: pros Alexandron]}. But what kind of parallel is that? Those men wrote things that were honorable to themselves and gratifying to Alexander. Do you find me anything of the sort? To me nothing comes to mind. As for your writing that you fear my influence and standing may be diminished by this mourning of mine --- I do not know what men are blaming or demanding. That I should not grieve? How can I? That I should not lie low? When did anyone do less so? While your house was lifting me up, who was shut out by me? Who came in and met a cold reception? From you I set off for Astura. Those cheerful people who blame me cannot read as many things as I have written. How well, is beside the point; but the kind of writing it was, nobody could have done with a dejected spirit. I was thirty days in the gardens. Did anyone miss either my company or the ease of my conversation? At this very moment I am reading and writing the kind of things that make those who are with me find leisure harder to bear than I find labor.

\ciceroSection{3} If anyone asks why I am not at Rome --- because it is the recess; why I am not at those little estates of mine which are in season --- because I could not easily bear the crowd of visitors. So I am where the man who used to have the finest of Baiae regularly spent this time of year. When I come to Rome, I shall not be faulted in face or in speech. That cheerfulness with which I used to season this bitter time I have lost forever; firmness and steadiness of mind and of speech will not be wanting.

\ciceroSection{4} About the Scapula gardens, this much seems possible --- partly through your influence, partly through mine --- to get them put up by the auctioneer. If that does not happen, we are shut out. But if we come to the auction-table, our eagerness will outbid Otho's resources. As for what you write to me about Lentulus, \dag\ there is nothing in him\dag. Let only the Faberius business be sure, and let you press on, as you are doing, and we shall get what we want.

\ciceroSection{5} You ask how long I shall be here: a few days. But I have not made up my mind. As soon as I have, I shall write; and write to me how long you are going to be at the place outside the city. The same day on which I send you this, about Pilia and Attica I too have the same news as you write, both in letters and by report.
