% Ad Familiares 15.19 (ad-familiares-15-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{C. CASSIVS S. D. M. CICERONI}

\ciceroSection{1} If you are well, it is well. By Hercules, on this travel of mine I do nothing more gladly than write to you; I seem, as I do it, to be talking and joking with you in person. And this does not happen on account of any ``Catian spectres''; for which, in my next letter, I shall pile up on you so many countrified Stoics that you will be saying Catius was born at Athens.

\ciceroSection{2} That our friend Pansa has left the city in his general's cloak with the goodwill of mankind, I am glad on his own account; and, by Hercules, on the account of all our own people too. For I hope men will come to understand how universally hated cruelty is, and how loved integrity and clemency are, and that the very things bad men chiefly seek and covet come round to good men. It is hard, true, to persuade men that the noble \textit{[Greek: to kalon di' hauto haireton]} --- is to be chosen for its own sake; but that pleasure \textit{[Greek: hēdonēn]} and freedom from disturbance \textit{[Greek: ataraxian]} are won by virtue, justice, and the noble \textit{[Greek: tōi kalōi]} is both true and persuasive. For Epicurus himself --- from whom all the Catii and Amafinii, those poor interpreters of words, take their start --- says: \textit{[Greek: ouk estin hēdeōs aneu tou kalōs kai dikaiōs zēn]}, ``there is no living pleasantly without living finely and justly.''

\ciceroSection{3} And so both Pansa, who pursues pleasure \textit{[Greek: hēdonēn]}, holds fast to virtue, and those who are called by you pleasure-lovers \textit{[Greek: philēdonoi]} are in fact lovers of the noble \textit{[Greek: philokaloi]} and lovers of the just \textit{[Greek: philodikaioi]}, and they both cultivate and hold to every virtue. So Sulla --- whose judgement we ought to endorse --- when he saw the philosophers disagreed, did not enquire what was good but bought up all the goods at once. His death, by Hercules, I have borne with a brave heart. Nor will Caesar let us miss him long: he has condemned men whom he can restore to us in his place, nor will he himself want for a buyer at his auctions, once he has had a look at the son.

\ciceroSection{4} Now, to come back to the state: write me what is going on in the Spains. I'm fretting, I will be damned if I am not; and I would rather have my old and clement master than try out a new and cruel one. You know what a fool Cnaeus is; you know how he counts cruelty as virtue; you know how he always thinks we have been laughing at him. I fear he means to give us a country-style sneer back \textit{[Greek: antimuktērisai]} with his sword. What is happening, if you love me, write me back. Oh, how I should like to know whether you are reading this in an anxious mood or a free one! For I shall know in the same moment what I ought to do. Not to make this any longer, farewell; love me, as you do. If Caesar has won, expect me soon.
