% Ad Atticum 13.19 (ad-atticum-13-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Hilarus the copyist had only just left on the fourth before the Kalends, with a letter from me for you, when the courier arrived with yours, dated the day before. What pleased me most in it was that our Attica asks you not to be downcast, and that you write the matter to be without danger \textit{[Greek: akindyna]}.

\ciceroSection{2} The \textit{Pro Ligario}, as I see, your influence has launched splendidly. For Balbus and Oppius have written to me that they wholly approve of it, and that on that account they have sent the little speech to Caesar. So you had already written me the same thing earlier.

\ciceroSection{3} In the case of Varro, what was holding me back was not the worry that I should look fond of fame \textit{[Greek: philendoxos]} --- for I had laid down the rule of including none of the living in my dialogues. But because you write that Varro both wants something from me and rates it highly, I have put the work together and finished it --- with what success I cannot say, but with such care that nothing could go beyond it: the whole Academic inquiry in four books. In those passages where Antiochus had brilliantly assembled arguments against ungraspability \textit{[Greek: akatal\=epsian]}, I have given the part to Varro. To them I myself reply; you are the third in our conversation. If I had made Cotta and Varro debate against each other, as you advise in your most recent letter, mine would have been the mute role \textit{[Greek: k\=ophon pros\=opon]}.

\ciceroSection{4} This works charmingly with figures from the older generation --- as Heraclides has often done it, and as we ourselves did in the six books \textit{On the Republic}. So too in our three books \textit{On the Orator}, of which I greatly approve. In those too the cast is such that I have had to keep silent: Crassus, Antonius, the elder Catulus, C. Julius the brother of Catulus, Cotta, Sulpicius all speak. The conversation is set in my boyhood, so that I could have had no part in it. What I have written in these recent times follows the Aristotelian manner \textit{[Greek: Aristoteleion]}, where the conversation of the others is so brought in that the lead remains with the author himself. On that pattern I have put together five books \textit{On Ends}, so as to give the Epicurean material to L. Torquatus, the Stoic to M. Cato, the Peripatetic to M. Piso. I had thought this would be beyond the reach of jealousy \textit{[Greek: az\=elotyp\=eton]}, since all those men were dead. The present \textit{Academica}, as you know, I had assigned to Catulus, Lucullus, Hortensius. Frankly, the parts did not fit the men: the matter was too technical \textit{[Greek: logik\=otera]} for them to look as if they had ever even dreamed about it. So when I read your letter about Varro, I seized on it as on a godsend \textit{[Greek: hermaion]}. Nothing could have been more apt to that kind of philosophy in which he most appears to take delight --- and so for my part I have not managed to make my own side come out the stronger. For the Antiochean arguments are powerfully persuasive \textit{[Greek: pithana]}; rendered with care by me, they keep Antiochus' sharpness, with our own polish of style, if there is any of that in us. But whether you think these books should after all be given to Varro --- this I should like you to consider again and again. Certain misgivings occur to me; but those when we meet.
