% Ad Atticum 13.22 (ad-atticum-13-22) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About Varro, it is not without reason that I am pressing you so carefully on what you think best. Certain things come to my mind. But of those, face to face. As for you, I wove you in most gladly \textit{[Greek: asmenaitata]}, and I shall do so more frequently. For in your most recent letter I learned for the first time that you do not object to it.

\ciceroSection{2} About Marcellus Cassius had written to me before; Servius has now sent the particulars \textit{[Greek: ta kata meros]}. O bitter business!

\ciceroSection{3} I come back to the first matter. I prefer my writings to be nowhere rather than with you, but only to be put out abroad when both of us think it right. I acquit your copyists of any blame, and I am not accusing you; even so, I had written something else to you --- that Caerellia has certain pieces which she could not have got except from you. To Balbo, certainly, I saw it had to be given; I only did not want either an outdated copy to be given to Brutus or an unfinished one to Balbus. To Varro I shall send it, as soon as I see you, if you think well of it. What it was that made me hesitate, you will know when I see you.

\ciceroSection{4} That you are dunning the assigned debtors, very well done. That you are being put through your paces over Ovia's property I am sorry. Our friend Brutus' business is most tiresome; but life brings such things. The women, however, are scarcely behaving decently, since they take it in a hostile spirit when on both sides he is doing his duty. Tullius my secretary there was no occasion for you to dun; for I should have charged you with it if there had been. Nothing is deposited with him under the head of a vow, but there is a certain sum of mine in his hands. That I have determined to apply to this purpose. So I told you rightly where it was, and he rightly denied it to you. But let us tackle this matter too at once. The grove --- I do not really approve of it for the men's purposes, because the spot is too lonely; but it has its plausibility \textit{[Greek: eulogian]}. Yet this too as you shall judge, since indeed in everything you do. I shall be there, as I have arranged --- and I only wish you too on the same day! But if anything intervenes (many things may), then on the following day at all events. For the co-heirs --- \dag{}from whom to be put down without you is hard service\dag{}. Two letters running now, and nothing to me about Attica. But this I set down to the best hope; what I do complain of is not you but her, that she has not so much as sent a greeting. But you for your part give my warmest to her and to Pilia, and do not let on, however, that I am cross. I have sent Caesar's letter, in case you had not read it.
