% Ad Familiares 6.20 (ad-familiares-06-20) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO TORANIO S.}

\ciceroSection{1} Three days ago I gave a letter for you to Cn. Plancius's slaves; I shall therefore be shorter now, and where before I was offering you consolation, on this occasion I shall give advice. I think nothing more useful for you than to wait it out where you are, until you can know what you ought to do. For apart from the danger you will have avoided of a long voyage, in winter, with few harbours, there is this not inconsiderable advantage too: that the moment you have heard anything definite, you can set out from where you now are. Beyond that, there is no reason to be eager to throw yourself in the way of arrivals. I have many fears, moreover, which I have shared with our friend Cilo.

\ciceroSection{2} Why say more? In these evils you could have been in no more advantageous a place, from which, wherever there is need, you may most easily and most quickly betake yourself. If he comes back to himself in time, you will be on hand; if --- since many things can happen --- some development either hinders or delays him, you will be where you can know everything. This plan suits me thoroughly.

\ciceroSection{3} For the rest, as I have so often urged you by letter, I would have you persuade yourself that you have nothing in this business to fear except the common lot of the state. Heavy though that lot is, still we have lived so long, and are now of such an age, that whatever befalls us not through our own fault we ought to bear bravely. Here all yours are well, and with the deepest devotion they long for you, love you, cherish you. See that you keep well, and that you do not stir from your place without due cause.
