% Ad Familiares 6.19 (ad-familiares-06-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO LEPTAE.}

\ciceroSection{1} I am glad that Macula has done his duty. His place at Falernum has always struck me as a suitable inn --- if only it has roof enough to take in our retinue. Otherwise the spot does not displease me. Not that I shall on that account give up your Petrinum: that villa, and that loveliness of the place, is a property for staying in, not for stopping over.

\ciceroSection{2} About some sort of curatorship of the royal games, I have spoken with Oppius; for Balbus, since you set out, I have not seen. He is so afflicted with pains in the feet that he is not willing to be approached. On the whole matter, frankly, you would in my view do more wisely not to trouble yourself with it. The advantage you mean to win by such pains, you will not win by any means. For so great is the crowd of his intimates that it is more likely one of them will be squeezed out than that a new man will gain admission --- especially one who brings nothing but his services, in which Caesar will think \textit{he} has done \textit{you} the favour (if indeed he learns of you at all), not received one. Still, we shall look for something with show in it; otherwise, in my judgement, the thing is not merely not to be sought after but to be actively avoided. For myself, I think I shall stay at Astura rather longer, until he, whenever it is, comes back. Goodbye.
