% Ad Atticum 13.38 (ad-atticum-13-38) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Before dawn, while I was writing against the Epicureans, I scratched off something or other to you from the same lamp and the same labour, and sent it off before daylight. Then, having gone back to sleep and woken with the sun, a letter is delivered to me from your sister's son --- the very thing I have sent on to you. Its opening is of the most outrageous insolence. But perhaps \textit{he did not stop to consider} \textit{[Greek: ouk epestesen]}. It runs as follows: ``For my part, whatever ill thing can be said against you --- '' He means that much ill could be said against me, but he denies that he himself approves of it. Could anything be filthier than this? You can read the rest for yourself --- I have sent it on --- and judge. Worn down at last by our Brutus's daily and tireless praise of me, of which a great many have reported to me, he has, I think, finally written something to me, and to you as well; you will let me know. What he wrote about me to his father I do not know; about his mother, how lovingly! ``I had wanted,'' he says, ``to be with you as much as possible, and to that end to have a house rented for me. I had written to you about it. You ignored it. So we shall be together less.'' Why? Because I cannot bear the sight of \textit{that} house --- the reason you know.

\ciceroSection{2} His father said the reason was his hatred of his mother. Now help me out, my dear Atticus, with your counsel: \textit{which is it to be} \textit{[Greek: poteron]} --- am I to spurn the man openly and reject him outright, \textit{or with crooked stratagems} \textit{[Greek: e skoliais apatais]}? For as with Pindar, \textit{my mind is divided on speaking the plain truth} \textit{[Greek: dicha moi noos atrekeian eipein]}. The former, certainly, suits my character better; the latter, perhaps, the times. You make up your own mind, and take it as already mine. What I most fear is being caught here at the Tusculan villa. In a crowd these things would be easier. To Astura, then? But what if Caesar drops in unannounced? Help me, please, with your counsel. I will do whatever you decide.
