% Ad Atticum 12.21 (ad-atticum-12-21) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I have read Brutus's letter and send it back to you. What he writes by way of reply to your queries shows little judgment. But he must look to it himself. There is one point, though, on which his ignorance is shameful. He thinks Cato spoke first to the motion on the punishment --- whereas all had spoken before him except Caesar; and since Caesar's own opinion, delivered while he was of praetorian rank, was so severe, Brutus supposes that the consulars, Catulus, Servilius, the Luculli, Curio, Torquatus, Lepidus, Gellius, Volcacius, Figulus, Cotta, Lucius Caesar, Gaius Piso, Manius Glabrio, and even Silanus and Murena, the consuls-elect, spoke more gently. Why then did I put the question on Cato's motion? Because in more luminous words and at greater length he had embraced the same matter. But on this point he praises me for having brought it forward, not for having uncovered it, for having pressed it on, in short for having pronounced judgment myself before I put the question. And because Cato had carried all this to the skies in praise and had moved that it be entered into the record, the division was taken on his motion. The man here even thinks he is paying me a high compliment by writing that I was ``the best of consuls.'' What enemy ever spoke more meagrely? And as to the rest, look how he has answered you! All he asks is that you correct the decree of the senate. He would have done as much had he been prompted by a copyist. But these matters again he must look to himself.

\ciceroSection{2} About the gardens, since you approve, get something done. You know my finances. If indeed something also comes loose from Faberius, the business is nothing. But even without him I think I can manage to stretch to it. The gardens of Drusus are certainly for sale, perhaps also those of Lamia and Cassius. But this when we are face to face.

\ciceroSection{3} About Terentia I cannot write more aptly than you write. Let our obligation be paramount. If anything cheats us, I would rather have it her fault to be sorry for than mine.

\ciceroSection{4} The 100,000 sesterces owed to Ovia, the wife of Gaius Lollius, must be seen to. Eros says it cannot be done without me --- because, I suppose, some valuation has to be received and given. I wish he had told you. For if the matter is, as he wrote me, all arranged, and he is not lying about that as well, it could have been wound up through you. Look into it and wind it up, please.

\ciceroSection{5} As for your summoning me to the Forum, you summon me to the place from which I fled even when my affairs were going well. For what is the Forum to me without the courts, without the senate-house, while there run into my sight men I cannot look at with composure? And as to your writing that men require it of me that I be at Rome and do not allow me to be away, \dag aut quatenus\dag\ that they grant me, know that it has long been so with me that I count you alone more than all of them put together. Nor do I undervalue myself: I would much rather stand by my own judgment than by that of all the rest. Yet I do not go further than the most learned of mankind concede to me; and all their writings, whatever they are upon this matter, I have not only read --- which by itself was the act of a brave invalid, to take his medicine --- but have transferred into my own writings as well, which surely was no act of a stricken and broken mind. From these remedies do not call me back into that throng, or I shall fall back.
