% Ad Familiares 13.7 (ad-familiares-13-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CLVVIO S.}

\ciceroSection{1} When, on your way out to Gaul, you came to my house in keeping with the bond between us and your own great attentiveness towards me, I spoke with you about the tax-paying land which the township of Atella holds in Gaul, and I showed you how strongly I felt that town's case. Since your departure, however, this business --- which is of the highest moment to a most honourable township and one most closely tied to me, and which is at the same time the highest of duties of mine --- has come up for decision; and given your singular goodwill towards me I have thought I ought to write to you about it more carefully. I am not unaware of the demands of the moment, nor of what your authority allows you; and I understand perfectly that what you have from C. Caesar is a commission, not a discretion. I therefore ask of you no more than what I think you are able to do and will gladly do for my sake.

\ciceroSection{2} First, I would have you reckon what is in fact the case: that the whole fortune of the town rests on that revenue, and that in these times the town is weighed down with the heaviest burdens and afflicted with the worst hardships. This may look to be its lot in common with many others, but believe me: peculiar calamities have come upon this town --- which I do not enumerate, for fear that, in lamenting the troubles of those bound to me, I should seem to give offence to men whom I have no wish to offend.

\ciceroSection{3} And so, did I not have great hope of carrying the town's case with C. Caesar, there would be no reason for me to press anything from you at this present hour. But because I have confidence --- I am persuaded of it --- that he will give weight to the town's dignity, to equity, and to its goodwill towards himself, I have not hesitated to press you to keep this case intact for him.

\ciceroSection{4} I would be making this request of you no less even if I had heard of no such thing on your part already; but I have conceived the greater hope of carrying my point since I was told that the men of Rhegium have obtained this same favour from you. They are touched by some bond to you, it is true; but your love for me compels me to hope that what you have granted to those bound to you, you will grant likewise to those bound to me, the more so since I am pressing for this on behalf of these alone, though I have many bound to me labouring under a like case. That I am not doing this without cause, nor pressing my point from any light self-interest, I think you of yourself believe; still, on my own affirmation, take my word that I owe most to this town: never has there been a time, in my honours or in my hardships, when this town's zeal for me has not stood out as singular.

\ciceroSection{5} Therefore again and again, by the perfect bond between us and your unbroken and most generous goodwill towards me, I ask and I press: since you grasp that what is at stake is the fortunes of this town, which is bound to me by the closest of ties, by every service, by every kindness, grant me this much. And it will be of such a kind that, if we obtain from Caesar what we hope, we shall judge it your kindness that we have got it; or if not, even then we shall reckon it as good as obtained, since the effort to obtain it came from you. By this you will not only have done me the most welcome thing, but you will have bound to yourself and to yours, in perpetuity, by the highest of services, men of the finest character and the most honourable standing, men no less grateful, and men worthy in the highest degree of being on your friendship's roll.
