% Ad Familiares 13.8 (ad-familiares-13-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO M. RVTILIO S.}

\ciceroSection{1} Since I was conscious to myself how much I value you, and had had proof of your goodwill towards me, I had no hesitation in asking of you what I had to ask. How much I value P. Sestius I myself know best; how much I owe to value him, you, and indeed everyone, knows. He, having learned from others that you are most warmly devoted to me, asked of me that I should write to you, with all possible care, about the case of the senator C. Albinius. From Albinius's daughter is sprung L. Sestius --- a young man of the finest character, the son of P. Sestius. I have set this out so that you may grasp that not only ought I to labour on P. Sestius's behalf, but Sestius too on Albinius's.

\ciceroSection{2} The matter stands thus. C. Albinius took at valuation, from M. Laberius, certain estates which Laberius had bought from Caesar out of the property of the Plotii. If I were to argue that it is not in the interest of the commonwealth that these be divided up, I should seem to be lecturing you, not asking. Yet since Caesar wishes the Sullan sales and assignations to stand --- so that the firmer his own may be reckoned --- if estates that Caesar himself has sold are now to be divided up, what authority, pray, can there be left in his own sales? But what stands behind this point, you, in your good judgement, will weigh for yourself.

\ciceroSection{3} I ask you plainly --- and with such warmth that I could not ask anything with greater zeal, on juster grounds, or more from the heart --- that you spare Albinius and lay no hand on the Laberian estates. You will have given me great cause, not merely for gladness, but in a way for glory itself, if through me P. Sestius makes good with a man bound to him by the closest of ties, in the way that I, for my own part, owe most to Sestius alone. That you may do this, I ask of you again and again, urgently. No greater service can you do me; and you will see it has been most welcome.
