% Ad Familiares 5.11 (ad-familiares-05-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO VATINIO IMP. S.}

\ciceroSection{1} That my good offices please you does not surprise me; I have come to know that you are the most grateful of all men, and I have never stopped saying so. Nor indeed have you merely felt the obligation to me --- you have repaid it in heaped-up measure. So in all your other affairs you will find me at one with you in the same zeal and the same goodwill.

\ciceroSection{2} As for your commending to me that most distinguished lady, your wife Pompeia: the moment your letter was read I spoke with our friend Sura, that he should tell her, in my name, to send me word whatever she might need, that I would attend to everything she wished with the highest zeal and care. So I shall do, and shall call on her in person if it seems necessary. But I should like you to write to her not to think any matter, great or small, either difficult for me or beneath my dignity. Everything I do in your affairs will seem to me both not laborious and an honour.

\ciceroSection{3} About Dionysius, if you love me, see the thing through. Whatever pledge you have given him I shall make good; but if he turns out to be a scoundrel, as in fact he is, you shall lead him captive in your triumph. May the gods deal ill with the Dalmatians, who are giving you such trouble! But, as you write, they will soon be taken and will set off your exploits in good light --- they have always been counted a warlike people.
