% Ad Atticum 13.49 (ad-atticum-13-49) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} To Attica first my greetings (in the country, I take it; so many greetings), and to Pilia. About Tigellius, if there is anything new. The fellow, so Fadius Gallus writes me, is lodging a most unfair complaint \textit{[Greek: mempsin anapherei]} against me --- that I let Phamea down after I had taken on his case. I had indeed taken it on, against the young Octavii, sons of Gnaeus, and not gladly; but for Phamea's sake I was willing. For he had, if you remember, during my canvass for the consulship, made me a promise through you, of his services should I need them; which I treated as if I had used them. He came to me and said the judge had decided to give him a hearing on the very day on which we were due to retire to render verdict on our friend Sestius under the Lex Pompeia. You will remember that the days of those trials were fixed by the statute. I replied that he was well aware of what I owed Sestius; that if he would take any other day he liked, I would not fail him. With that he went off in anger. I think I told you the story. I gave it no thought, of course, and I did not feel obliged to manage the unfair temper of a man who is nothing to me.

\ciceroSection{2} I told Gallus, the last time I was in Rome, what I had heard, without naming Balbus the younger. Gallus had his own business to handle, as he writes. He says the man is saying that I suspect him out of a guilty conscience over having abandoned Phamea. So this is all I ask of you: about our friend, find out what you can; about me, do not trouble yourself. It is a fine thing to have someone one can hate with pleasure; and just as one is not bound to sleep for everyone, one is not bound to serve everyone. Though by Hercules, as you understand, those very men are more in service to me, if to dance attendance is to serve.
