% Ad Atticum 13.13 (ad-atticum-13-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Stirred by your letter --- by what you had written to me about Varro --- I have taken the whole \textit{Academica} away from those most noble characters, transferred it to our friend, and converted it from two books into four. They are larger, on the whole, than the others were, but a good deal has still been cut out. I should very much like you to write to me how you discovered he wanted it. And I particularly want to know whom you understood him to be jealous of \textit{[Greek: zelotypeisthai]} --- unless perhaps Brutus. That, by Hercules, was all we needed! But still, I should very much like to know. The books, in fact, have come out --- unless our common self-love \textit{[Greek: philautia]} is deceiving me --- so that in this kind of writing not even among the Greeks is there anything like them. You will bear with equanimity that earlier loss --- that the Academica passages you have in copy have been transcribed in vain. These will be much more splendid, more compact, better.

\ciceroSection{2} But now I am at a loss \textit{[Greek: aporo]} which way to turn. I want to do something for Dolabella, who very much wants it; I cannot find what; and at the same time I have a sense of shame before the Trojans \textit{[Greek: aideomai Troas]} --- and if I do produce something, I shall not be able to escape rebuke \textit{[Greek: mempsin]}. Either I must hold off, then, or I must think of something. But why am I troubling myself with these trifles?

\ciceroSection{3} My Attica --- I beg of you --- what is she doing? She torments me. But I keep tasting your letters over again; I rest in them. Still, I am waiting for new ones.

\ciceroSection{4} Brinnius' freedman, our co-heir, has written to me that he and his co-heir Sabinus Albius would like, if it suits me, to come to me. That I plainly do not want. The inheritance is not worth it. And in any case they will be able to make the auction date easily enough (it is the third before the Ides) if they meet me at the Tusculanum on the morning of the day after the Nones. If they wish to put the day off more loosely, they may put it back two or three days or as seems best; it makes no difference. So unless they have already set out, hold them back. About Brutus, if he has done anything; about Caesar, if you know anything; whatever else there may be --- you will write.
