% Ad Atticum 12.41 (ad-atticum-12-41) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} There was nothing for me to write. I wanted, all the same, to know where you were --- whether you are away, or about to be, and when you would be coming back. So you will let me know. And as for what you wanted to know, namely when I am leaving this place: I have decided to stay the day after the Ides at Lanuvium, and from there the day after at the Tusculanum, or at Rome. Which I am going to do, you will know on that very day.

\ciceroSection{2} You know how prone misfortune is to find fault \textit{[Greek: philaition sumphora]} --- least of all toward you, certainly, but all the same I have been seized with greed about the shrine: and unless I see it (I will not say completed, but) on the way to being done --- I shall dare to put it this way, and you will take it as you usually do --- my grief will rush upon you, with no justice in it; but you will bear with this very letter as you bear with all my doings, and as you have always borne. All your consolations I should like you to bring to bear on this one matter. If you ask what I most want: first, Scapula's place; then Clodia's; after that, if Silius will not, and Drusus is going to deal unfairly --- Cusinius' and Trebonius'. I think there is a third owner; I know for certain that Rebilus was one. But if the Tusculanum site suits you, as you hinted in one of your letters, I shall come round to your view. This, in any event, you will bring off, if you wish me lightened --- you who now accuse me even more sharply than your usual gentleness can support, but you do it from the deepest love and, perhaps, overborne by my own fault. Still, if you wish me lightened, this is the supreme relief, or, if you wish to know the truth, the only one.

\ciceroSection{4} If you have read Hirtius' letter --- which seems to me a kind of preliminary sketch \textit{[Greek: proplasma]} of the attack Caesar has written against Cato --- you will let me know what you made of it, if it is convenient. I come back to the shrine. Unless it is brought to completion this summer, which you see still lies open before us, I shall not consider myself freed of the crime.
