% Ad Atticum 13.31 (ad-atticum-13-31) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the fifth before the Kalends, in the morning, I received from Demeas a letter dispatched the day before, from which I ought to be expecting you today or tomorrow. But I, the very man waiting for your arrival, shall, I think, be the one to delay you. For I do not believe the Faberian business will turn out so easily that it does not involve some hold-up, even if it is going to turn out at all.

\ciceroSection{2} When you can, then. Since you are still away, I should very much like you to send me those books of Dicaearchus you mentioned, and add the \textit{Descent} \textit{[Greek: katabaseōs]} as well.

\ciceroSection{3} About the letter to Caesar I have decided \textit{[Greek: kekrika]}. The very thing these people say he is writing --- that he will not set out for the Parthians until affairs are settled --- is what I was urging in that letter. Yes: he might do whatever he wished, on my authority. For this, evidently, is what he is waiting for, and he will not act except on my advice. I beg you, let us throw all that aside and at least be half-free; which we shall achieve by keeping silent and lying low.

\ciceroSection{4} But move on Otho, as you write. Settle that business for me, my dear Atticus. I find nothing else where I can both be out of the Forum and be in your company. As for what it will cost, this comes to mind: Gaius Albanius is next-door neighbour. He bought a thousand iugera from Marcus Pilius for, as I remember, 11,500,000 sesterces. Everything, of course, is cheaper now. But desire is added in; and in that I do not think we shall have any rival but Otho. Yet you yourself will be able to move even him, more easily still if you have Canus on your side. What a tasteless palate! I am ashamed of his father. Write back if you have anything to say.
