% Ad Atticum 13.50 (ad-atticum-13-50) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Prompted by certain letters of yours that I should begin writing fuller letters to Caesar, and after Balbus had told me lately at the Lanuvian villa that he and Oppius had written to Caesar that I had read the books against Cato and had heartily approved them, I have composed a letter to Caesar on those very books, to be delivered to him by Dolabella; but I have sent a copy of it to Oppius and Balbus, and have written to them to give the order for my letter to be delivered to Dolabella only if they themselves approved the copy. They have written back that they had never read anything better, and have given the order for my letter to be handed over to Dolabella.

\ciceroSection{2} Vestorius has written to me to instruct that for my share a slave of his be given by formal conveyance to a certain Hetereius for the Brinnian estate, so that he can himself convey it duly at Puteoli. The slave, if you think it best, will you send to me; for I think Vestorius has written to you as well.

\ciceroSection{3} About Caesar's coming, the same word I had from you I have had from Oppius and Balbus. I am surprised you have nothing yet from Tigellius --- for instance, this point: how much he has received. I am altogether eager to know it, and yet I do not care a straw.

\ciceroSection{4} You ask what I am thinking about going to meet him. What do you suppose, but Alsium? And indeed I had written to Murena about lodgings, but I gather he has set out with Matius. So Sallustius will be pressed into service.

\ciceroSection{5} The previous little line just written, Eros tells me Murena has answered him most handsomely. So let us use his house. For Silius has no cushions. And Dida, I gather, has given over the whole villa to guests.
