% Ad Atticum 12.37 (ad-atticum-12-37) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I received two letters from you yesterday, the one given to Hilarus the day before, the other on the same day to a courier; and I had it on the same day from the freedman Aegypta that Pilia and Attica are in plainly good health. That you sent Brutus' letter --- thank you. He has sent one to me as well; it was delivered to me on the thirteenth day. I have sent on to you that very letter, and with it a copy of my reply.

\ciceroSection{2} As to the shrine, if you find no gardens for me --- and you must find them, if you value me at the rate at which you certainly do --- I thoroughly approve your reasoning about the Tusculanum. Prudent as you are in thinking things over, as you are, still, unless you cared greatly that I should attain what I greatly want, the thing could never have come into your mind so neatly. But somehow I find myself wanting a frequented spot; so you must conclude the gardens deal for me. The greatest throng is at Scapula's, and there is besides the nearness of where you would be --- not the whole day in the villa. So before you leave, I should very much like you to meet Otho, if he is in Rome. If nothing comes of it, although you are used to bearing with my folly, I shall press on so far that you will be exasperated. For Drusus certainly wants to sell. So if nothing else turns up, it will be my fault if I do not buy. Please see to it that I do not slip up here; and the one way to see to it is, if we can do anything about the Scapula heirs. I should like to be informed how long you are to be at the place outside the city.

\ciceroSection{3} With Terentia I have as much need of your influence as of your authority. But do as seems best to you. For I know that, if anything concerns me, it tends to be a greater care to you than to me.
