% Ad Atticum 13.37 (ad-atticum-13-37) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} This is the second letter today. About the debt owed by Xenon and the Epirote affair \dag x x x x \dag --- nothing could be handier or more fitting than what you propose. Balbus the younger had spoken to me to the same effect.

\ciceroSection{2} Nothing really new, except that Hirtius has been quarrelling fiercely with Quintus on my behalf. The man is in a rage wherever he goes, but most of all at dinner parties: a great deal about me, and then back to his father. Nothing, however, is asserted by him more \textit{convincingly} \textit{[Greek: axiopistos]} than that we are utterly estranged from Caesar, that we are not to be trusted, that I in particular am to be watched --- \textit{it would be alarming} \textit{[Greek: phoberon an en]}, if I did not see that the king knows I have no spirit left in me. As for my son Cicero, he is the one being harried; but on that point my nephew can do as he likes.

\ciceroSection{3} I am glad I gave the eulogy of Porcia to Lepta's courier before your letter reached me. So please, if you love me, see to it --- if it is to be sent at all --- that it goes in just the same way to Domitius and to Brutus.

\ciceroSection{4} About the gladiators, and the other matters you describe as \textit{flighty rumours} \textit{[Greek: anemophoreta]}, keep me posted daily. And, if you think it right, please approach Balbus and Offilius. About advertising the auction I have spoken with Balbus myself. He approved (I imagine Offilius has the inventory in full; Balbus has it too) --- but Balbus was for an early date, and at Rome. If Caesar were to delay, the date could be put off. Yet he does seem to be on his way. Think the whole thing through, then; Vestorius is for it as well.
