% Ad Atticum 13.12 (ad-atticum-13-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Your letters about our Attica bit me hard; the same letters, however, also healed me. For the consolation you offered yourself in the very same letter was enough, for me, to soften the pain.

\ciceroSection{2} The \textit{Ligariana} you have sold splendidly. Hereafter whatever I write I shall hand over to you as my crier.

\ciceroSection{3} As to what you write to me about Varro: you know that previously I used to write speeches, or that sort of thing, in such a way that I could nowhere weave Varro in. After I began on these more scholarly things \textit{[Greek: philologotera]}, Varro had already been giving me notice of a very large and weighty address \textit{[Greek: prosphonesin]}. A two-year stretch has gone by during which our friend Callippides \textit{[Greek: Kallippides]}, for all his nonstop running, has not moved one cubit forward; while I, for my part, was getting myself ready for whatever he might send me, to repay him in the same measure \textit{[Greek: auto to metro kai]} --- if at any rate I could. For Hesiod, too, adds this: ``if you can'' \textit{[Greek: ai ke dunai]}. Now that \textit{Treatise on Ends} \textit{[Greek: peri Telon syntaxin]} I am rather pleased with, and we have, at your suggestion, promised to Brutus; and you wrote to me that he was not unwilling. So that \textit{Academic} work \textit{[Greek: Akademiken]} --- in which those characters, men of high station certainly but no philologists, talk too cleverly --- let us transfer to Varro. There are Antiochean doctrines in it which he warmly approves. Catulus and Lucullus we shall set down elsewhere --- provided you approve of this; on which I should like you to write back to me.

\ciceroSection{4} About the Brinnian auction I have had a letter from Vestorius. He says the matter has been referred to me without any dispute. They evidently thought I would be at Rome or at the Tusculanum on the eighth day before the Kalends of Quintilis. You will tell, then, either your friend S.~Vettius, my co-heir, or our Labeo, to put the auction off a little; I shall be at the Tusculanum around the Nones. As for Piso, you have Eros. About the Scapulan gardens, let us think with our whole heart. The day is at hand.
