% Ad Familiares 6.18 (ad-familiares-06-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO LEPTAE.}

\ciceroSection{1} The moment I had the letter from your man Seleucus, I sent to Balbus on the spot, by note, to ask what the law actually said. He wrote back that those who \textit{are} acting as auctioneers are forbidden to sit among the decurions; those who \textit{have} done so are not forbidden. So your friends, and mine, may take heart; for it would have been past bearing that, when men who today practise haruspicy are being enrolled into the Senate at Rome, men who at some point acted as auctioneers should not be allowed to sit on a town council.

\ciceroSection{2} From the Spains there is no news. It is established, however, that Pompeius has a large army; for Caesar himself has sent his people a copy of a letter of Paciaecus, in which it was written that those legions number eleven. Messalla had also written that Q.~Salassus had had his brother, P.~Curtius, put to death on Pompeius's order, in the sight of the army, because he had conspired with certain Spaniards that, if Pompeius came into some town or other on a corn-supply errand, they would seize him and bring him over to Caesar.

\ciceroSection{3} As to your own business --- that you stand surety for Pompeius --- if Galba, your co-surety, comes back (a man not lacking in care over private affairs), I shall not stop conferring with him as to whether anything can be worked out; that, it seemed to me, was something he had confidence in.

\ciceroSection{4} That my \textit{Orator} has your warm approval gives me great delight. For my own part I am convinced that whatever judgement I had about speaking, I put into that book. If it is the kind of thing you write me it seems to you to be, then I too am something; if otherwise, I do not object that as much as is taken from the book's standing should be taken from the reputation of my judgement. I should like our young Lepta to be taking pleasure in such writings already. The ripeness of years is lacking, no doubt; even so, it does no harm to ring his ears with words of this kind.

\ciceroSection{5} What kept me at Rome, more than anything, was my Tullia's confinement. But although she is now, as I hope, reasonably strong, I am still held back here until I get the first instalment out of Dolabella's agents. And, by Hercules, I am not such a traveller now as I used to be. My buildings are my delight --- and my leisure. I have a house that yields to none of my villas; my leisure, to none in the most secluded region. So even my literary work is not interrupted, and I keep at it without any breaking in. So I rather think you will see us here before we see you there. Let our sweetest little Lepta learn his Hesiod by heart and have on his lips \textit{but in front of excellence} \textit{[Greek: t\=es d' aret\=es hidr\=ota]} and the rest.
