% Ad Atticum 13.52 (ad-atticum-13-52) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Oh, what a burdensome guest he was --- and yet I have no regrets \textit{[Greek: ametameleton]}! For it was thoroughly pleasant. When he had arrived at Philippus's on the evening of the second day of the Saturnalia, the villa was so filled with soldiery that there was scarcely a dining-room left vacant for Caesar himself to dine in --- two thousand men, no less. I was, in all candor, much disquieted what was going to happen the next day; but Barba Cassius came to my aid and posted guards. The camp went up in the field; the villa was secured. On the third day of the Saturnalia he stayed at Philippus's until the seventh hour, and admitted no one --- accounts with Balbus, I imagine. Then he took a walk on the shore. After the eighth hour, into the bath. There he heard the news about Mamurra, and did not change his expression. He was anointed; he reclined. He was on the emetic regimen \textit{[Greek: emetiken]}, so he ate and drank without restraint \textit{[Greek: adeos]} and with pleasure, lavishly indeed and elegantly served, and not only that but with good cooking and seasoning --- with good conversation as well, and, if you ask me, in good humor.

\ciceroSection{2} Besides, his retinue \textit{[Greek: hoi peri]} were entertained at three further dining-rooms most generously. To the lower-ranking freedmen and the slaves nothing was lacking; the smarter freedmen I entertained in style. In short, we showed ourselves civilized men. Still, this is not a guest of whom one says, ``Do please drop in again on your way back.'' Once is enough. Nothing serious \textit{[Greek: spoudaion ouden]} came up in the conversation, but plenty of literary talk \textit{[Greek: philologa]}. What more can I tell you? He enjoyed himself and was glad to be there. He said he would spend one day at Puteoli and the next at Baiae. There you have a piece of hospitality --- a billeting \textit{[Greek: epistathmeian]}, rather --- that, as I said, was a burden to me, not a bother. I shall stay here a little while, then on to Tusculum. As he passed Dolabella's villa, the whole armed escort drew up to the right and left of his horse, nowhere else along the way. This from Nicias.
