% Ad Familiares 4.6 (ad-familiares-04-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. SER. SVLPICIO.}

\ciceroSection{1} For my part, Servius, I should have wished, as you write, that you had been with me in this gravest of my misfortunes. How much, by being present, you could have helped me, both by consoling and by grieving with me as if at almost equal share, I see easily enough from this: that on reading your letter I was steadied somewhat. For both you have written things that could lighten a man's mourning, and in consoling me you yourself brought to the work no small grief of your own spirit. Your son Servius, by every office that the moment allowed, has shown me both how much he himself thought of me and how welcome he supposed his bearing toward me would be to you. His services have often, of course, been more pleasant to me; never have they been more welcome. As for me --- it is not your speech alone, and the all-but-fellowship in suffering, but also your authority that is consoling me. For I count it a disgrace not to bear my own case in the way a man of your wisdom thinks it ought to be borne. But sometimes I am overcome, and I can scarcely stand up against the grief, because those consolations fail me which did not fail the other men, whose examples I set before myself, in a like turn of fortune. For both Q. Maximus, who lost a son who had been consul, a distinguished man and great in things accomplished; and L. Paullus, who lost two sons within seven days; and your Galus; and M. Cato, who lost a son of the highest character and the highest virtue --- these men lived in times such that their grief was consoled by the standing which they were attaining from the commonwealth itself.

\ciceroSection{2} But I, with those distinctions you yourself recall taken from me --- which I had won at the greatest pains --- had one consolation left, and that has been wrenched away. My thoughts were no longer hindered by the affairs of my friends, no longer by the administration of the commonwealth; nothing pleased me to do in the Forum; I could not look at the Senate-house; I judged --- and the judgement was true --- that I had lost every fruit both of my industry and of my fortune. But, when I would consider that these losses I shared with you and with certain others, and when I broke myself down and forced myself to bear them with endurance, I had a place to take refuge in, a place to rest, in whose conversation and sweetness I could lay down all my cares and pains. Now, with this so heavy wound on top, even those griefs which had seemed to have healed are breaking open again. For it is not now as it was then, when home received me sad from the commonwealth and lifted me; nor, sorrowing at home, can I now take refuge in the commonwealth, and find rest in its goods. And so I am away both from home and from the Forum: home can no longer console the grief I take from the commonwealth, and the commonwealth cannot console the grief I take from home.

\ciceroSection{3} The more, then, I expect you, the more I long to see you as soon as may be. No reasoning can bring me a greater consolation than the joining-back of our familiar conversation. And yet I was hoping that your arrival --- so I was hearing --- was drawing near. Now, for many reasons I wish to see you at the earliest moment, and above all so that we may take counsel together beforehand on how we are to carry ourselves through this present time, which is to be governed entirely by the will of one man --- a man of prudence, of generous bearing, and (as I think I have made out) not unfriendly to me, and a particularly warm friend to you. This being so, it is nevertheless a matter for grave deliberation what course of conduct we should adopt: not for any action of our own, but for keeping quietly, by his concession and favour. Farewell.
