% Ad Atticum 12.44 (ad-atticum-12-44) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} That Hirtius wrote something to you sympathetically \textit{[Greek: sympath\=os]} about me --- this I take in good part (for he did it humanely); and that you did not send me his letter --- that I take much more in good part (for you did so even more humanely). The book of his on Cato, which he sent to me --- this I want your people to circulate, so that the praise of him may be the greater for those men's reproach.

\ciceroSection{2} As for what you are doing through Mustela: you have a thoroughly suitable man there, sincerely attached to me ever since the time of Pontianus. So accomplish something. And what is there to accomplish, except that there be access for a buyer? That can be effected through any of the heirs. But Mustela, I think, will bring it off if asked. To me you will have given both the spot I wish for, with a view to what I want, and besides a place to grow old in, an \textit{[Greek: eng\=erama]}. For those of Silius and Drusus do not seem to me sufficiently master-of-the-house, \textit{[Greek: oikodespotika]}. What then? sit whole days in a country house? Those, then, I should prefer --- Otho's first, then Clodia's. If nothing comes of it, we must either play Drusus along or make use of the Tusculanum.

\ciceroSection{3} That you shut yourself up at home, you did with reason; but, I beg you, finish, and give yourself back to me a free man. As for me --- from here, as I wrote before, on the day after the Ides to Lanuvium, then the day after that to the Tusculanum. For I have beaten down my spirit, and perhaps conquered it, if only I hold out. So you will know perhaps tomorrow, at latest the day after. But what is this, please? Philotimus reports that Pompey is not being held at Carteia (about which Oppius and Balbus had sent me a copy of the letter sent to Clodius Patavinus, saying they took it for fact), and tells of a sizable war still to come. He is, to be sure, a regular little Fulvinius. Still, if you have any news. I should also like to know what the matter is with Caninius' shipwreck.

\ciceroSection{4} I have completed here two great treatises \textit{[Greek: syntagmata]}; for in no other way can I, so to speak, stray from my misery. As for you --- even if there will be nothing to write, as I see will be the case, still I should like you to write me that very thing: that you had nothing to write of --- provided not in those words.
