% Ad Familiares 5.14 (ad-familiares-05-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{L. LVCCEIVS Q. F. S. D. M. TVLLIO M. F.}

\ciceroSection{1} If you are well, I am glad; I myself am well --- in my usual way, though somewhat worse than usual. I have looked for you more than once, to see you; since you have not been at Rome after my own going away, I have wondered, and now I wonder the same. I have no certain knowledge of what most calls you off from here. If solitude pleases you, in that you may write and undertake some of the things you have always practised, I rejoice and do not find fault with your decision. Nothing could be more delightful than that --- not only in these wretched and grief-laden times, but even in days tranquil and to be wished for --- especially in your case, whether for a spirit worn out, which now seeks rest after great occupations, or for a learned spirit, which always brings something out of itself, things that delight others and set off the man himself with their praise.

\ciceroSection{2} But if, as when you went out from here, you have given yourself over to tears and grief, then I grieve --- because you grieve and are tormented --- and I cannot, if you allow me to speak more freely what I feel, refrain from finding fault with you. For consider: will you alone fail to see what lies in plain view, you who in the very sharpness of your mind see what is most deeply hidden? Will you not understand that you accomplish nothing by daily complaints? Will you not understand that you double the anxieties which your own wisdom requires you to lighten?

\ciceroSection{3} And if I cannot get anywhere by persuasion, I press by the right of friendship, and I ask, if you would do anything at all for our sake: that you loosen yourself from those troubles of yours and come back to our company, and to the way of life that is either our common one, or your own particular and proper one. I want not to grind you down, if I cannot delight you, with my eagerness; I want to deter you from going on with what you have begun. For now these two contrary things vex me, between which I wish, if you can, you may yield to me on the one, or at least not offend on the other. Farewell.
