% Ad Atticum 13.10 (ad-atticum-13-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I am not in the least surprised that you take Marcellus' death hard, and that you fear more than one kind of danger now. For who could have feared a thing of this kind --- one that had never happened before, and seemed by its very nature incapable of happening? So everything is to be feared. But that other thing is against the record \textit{[Greek: para ten historian]} --- coming from you, particularly --- that I am the last consular left. What of it? What is your view of Servius? Though of course this counts to neither side, especially for me, who think that those others have not been so badly served. For what are we, or what can we be? At home or in public? And had it not come into my head to write these things, whatever they are, I should have had nowhere to turn.

\ciceroSection{2} As to Dolabella, your advice is right --- I think it must be done, something rather public, rather political \textit{[Greek: koinotera ... politikotera]}. Something certainly must be done; he is very keen for it.

\ciceroSection{3} If Brutus does anything, you will see to it that I know; and indeed I think he ought to act as soon as possible, especially if he has made up his mind. He will either quench the gossip altogether or quiet it down. For there are people who talk about it even to me. But he himself knows best, especially if he is going to talk it over with you too. I plan to set out on the eleventh before the Kalends. For here I have nothing to do, nor by Hercules anything there either, nor anywhere; but still, something there. Today I am waiting for Spinther --- Brutus has sent him to me. In his letter he clears Caesar of Marcellus' death; on whom not a breath of suspicion would fall even if it had been assassination. As it is, with the case against Magius established, does not his own madness account for the whole affair? I plainly do not understand what it amounts to. You will explain, then. Yet I have no doubt of anything except what cause of derangement Magius himself had --- a man for whom I had even gone surety. And that no doubt was the reason: he was insolvent. I suppose he had asked Marcellus for something and Marcellus, †---†, had answered with more firmness than he liked.
