% Ad Atticum 12.29 (ad-atticum-12-29) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Silius today, as you write. So tomorrow --- or rather, when you can --- write to me, if there is anything once you have seen him. I am not avoiding Brutus, nor on the other hand do I expect any relief from him; but there were reasons why I did not wish to be there at this time. If those reasons hold, I shall have to look for some excuse to Brutus --- and as things stand, they look likely to hold.

\ciceroSection{2} About the gardens, please bring it to a head. The capital point is the one you know. Then comes this: that there must also be something for me myself; for I can neither live in a crowd nor be away from you. For this plan of mine I find nothing better suited than that place, and on this matter --- what is your own counsel? My mind is made up, and the more so because I have understood the same thing seems good to you, that I am very warmly esteemed by Oppius and Balbus. Make clear to them how very much I want the gardens, and on what grounds; but say also that this can only be if that Faberian business is cleared up --- so let them be the prime movers in it. If some sacrifice has to be made in paying ready cash, get them to come up as far as they can; for that whole sum is otherwise written off. In short, you will see whether they show any inclination at all toward helping this plan of mine forward. If they do, it is great help; if not, by whatever way we can, let us push on. You may treat it either as that ``thing to grow old in'' \textit{[Greek: enge\=erama]}, as you wrote, or as a ``thing to be buried in'' \textit{[Greek: entaphion]}. The Ostia place is not to be thought of. If we do not get this --- and I do not think we can get it from Lamia --- we must try Damasippus' offer.
