% Ad Atticum 12.43 (ad-atticum-12-43) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} It had come into my mind to urge you to do the very thing you are doing. For I thought you could carry that same business through more conveniently at home, the interruptions taken out of the way. I have decided, as I wrote to you before, to stop on the day after the Ides at Lanuvium, and from there either to Rome or to the Tusculanum; you will know which beforehand. \dag\ As you shall know, that this affair will rightly be a relief, you do well that\dag\ --- believe me, to a degree you cannot imagine. The thing itself shows how greatly I desire it, since I venture to confess it to you, whom I judge not to approve of it very strongly. But in this you must bear with my error. Bear with? No --- indeed even help.

\ciceroSection{2} Of Otho I have my doubts; perhaps because I want it. But indeed the matter is bigger than our means, especially with an opposing party who is at once eager, wealthy, and an heir. The next best is, I should like Clodia's. But if those cannot be brought off, achieve anything you can. I think myself bound by a more solemn vow than any man ever was by any vow whatever. You will see the Trebonian heirs too, although the owners are away. But, as I wrote to you yesterday, you will also think about the Tusculanum, lest the summer slip away; which must certainly not be allowed to happen.
