% Ad Atticum 12.14 (ad-atticum-12-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About my being excused before Appuleius I sent you a letter yesterday. I take it there is no difficulty. Whomever you approach, none will refuse. But see to Septimius and Laenas and Statilius; three are needed. Laenas, though, had taken the whole business on himself.

\ciceroSection{2} As to your writing that Junius has called upon you --- Cornificius is, in any case, a man of means; even so, I should like to know when I am said to have stood surety, and whether for the father or for the son. None the less, as you say, you will see Cornificius's agents and Appuleius the auctioneer.

\ciceroSection{3} As to your wishing me to be brought back to myself from this sorrow, you do as you do in all things; but that I have not failed myself, you are my witness. Nothing on the subject of lessening sorrow has been written by anyone that I have not read in your house. But grief overpowers every consolation. I have even done what surely no one before me has done --- consoled myself, by writing, in my own person. I shall send the book to you when the copyists have made it out. I assure you, no consolation is of its kind. Whole days I spend writing --- not that I make any progress, but in the meantime I am held back, not enough indeed (the force presses too hard), but I am loosened a little, and I strive in every direction not to repair my mind --- if I can --- so much as my face; and in doing so I sometimes seem to myself to be at fault, sometimes that I would be at fault if I did not. Solitude helps a little, but it would help far more if even so you were here. That is the one reason I have for leaving this place; for as concerns my troubles, it was doing well enough. And yet that itself grieves me. For you will not be able to find in me the same man I was. What you loved is gone.

\ciceroSection{4} About Brutus's letter to me I wrote to you earlier. Shrewdly written, but nothing in it to help me. What he wrote to you, I should wish for: that he were here himself. Surely, since he loves me so much, he would help in some way. If you learn anything, write to me, please --- above all, when Pansa is to set out. About Attica I grieve, but I trust in Craterus. Tell Pilia not to be in anguish. It is enough that I mourn for all.
