% Ad Atticum 12.27 (ad-atticum-12-27) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} As to the Silius business, although the terms are not unknown to me, still I expect to learn everything today from Sicca. The Cotta place, which you say you do not know, lies beyond Silius' villa, which I think you do know --- a sordid little cottage, very cramped, with no land at all, in extent enough for nothing except the one purpose I have in view. What I am after is a frequented spot. But if the Silius gardens come through --- that is, if you bring them through (for the whole thing depends on you) --- there is plainly nothing to think about as regards Cotta's place.

\ciceroSection{2} About my son, I shall do as you say; I shall leave it to him to choose the time. As to how much money will be needed for the transfer, you will see to it. If you have learned anything from Aledius worth writing about, write it. I see from your letters, and you doubtless see from mine, that we have nothing to write about: the same things day after day, things already worn to nothing. Yet I cannot help sending to you every day, so that I may receive your letters in return.

\ciceroSection{3} On Brutus, however --- if you have anything. I think by now it must be known where he is waiting for Pansa. If, as is usual, he is on the nearer side of his province, he looks to arrive about the Kalends. I would rather it were later; for many reasons I am very much avoiding the city. So I am in doubt even on this --- whether to prepare some excuse to him; which I see is easy enough to do. But we have time enough to think about it. My greetings to Pilia and Attica.
