% Ad Atticum 13.17 (ad-atticum-13-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the fourth before the Kalends I was looking for something from Rome --- $\dagger$I had not ordered it. Therefore something from your people.$\dagger$ Now it is still the same things: what Brutus is thinking, or whether he has done anything, and whether anything from Caesar. But why these matters, which I care less about? Our Attica --- how she is doing --- this is what I want to know. Even though your letter (though it is now too old) tells me to hope well, still I am waiting for something fresh.

\ciceroSection{2} You see what proximity carries with it. As for us, let us by all means see the gardens through. When I was at the Tusculanum we seemed to be in conversation, the letters back and forth were so frequent. But that, for now, will be that. In the meantime, on your advice, I have finished books --- decidedly clever little things --- addressed to Varro; but I am still waiting for your answer to what I wrote you. First, on what evidence you have understood that he wants something from me, when he, the most prolific writer of us all \textit{[Greek: polygraph\=otatos]}, has never himself prodded me. Next, whom is he supposed to be jealous of \textit{[Greek: z\=elotypein]}? --- unless perhaps Brutus, and if you are not jealous \textit{[Greek: z\=elotypeis]} of him, then much less of Hortensius, or of those who appear in the political dialogues. Plainly, I should like you to clear this up for me, and above all whether you stand by your view that I should send him what I have written, or whether you now think there is no need. But all this when we meet.
