% Ad Atticum 12.42 (ad-atticum-12-42) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I did not feel the want of any day's letter from you. For I saw what you write, and yet I suspected --- or rather understood --- that there was nothing to write of; and on the sixth day before the Ides I supposed you were away and saw plainly that you had nothing. Still I shall send to you almost every day; for I would rather send in vain than have you with no one to whom you may give a letter, in case there should happen to be anything you think I ought to know. So on the sixth day before the Ides I received your letter --- an empty one. For what did you have to write? Yet even that, such as it was, was not unwelcome to me: if nothing else, to know that nothing was new with you. Still you wrote something or other about Clodia. Where then is she, and when is she coming? The matter pleases me in such a way that next after

\ciceroSection{2} Otho nothing pleases me more. But I do not think this one will sell (she is taken with the place, and well off), and you have not failed to notice how hard the other business is. But, I beg of you, let us strive to think of something that will answer to what I want.

\ciceroSection{3} I think I shall leave here on the day after the Ides --- but either for the Tusculanum or for the house, and from there perhaps to Arpinum. When I know for certain, I shall write to you.
