% Ad Atticum 13.32 (ad-atticum-13-32) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Since I received a second letter from you today, I did not want you to be content with just one from me. By all means proceed, as you write, on the Faberian business. For the whole of what we are planning rests on that; and if that plan had not crossed my mind, believe me, I should not be exerting myself over this place any more than over the rest. So, as you are doing (for nothing can be added to your work as it stands), press, push, finish it.

\ciceroSection{2} Of Dicaearchus's \textit{On the Soul} \textit{[Greek: peri psychēs]}, I should like you to send both volumes, and the \textit{Descent} \textit{[Greek: katabaseōs]} as well. I cannot find the \textit{Tripoliticus} \textit{[Greek: Tripolitikon]}, nor his letter to Aristoxenus. Those three works are what I most need now; they would suit what I have in mind.

\ciceroSection{3} Torquatus is at Rome. I have given orders for the book to be delivered to you. \textit{Catulus} and \textit{Lucullus}, as I think, were sent earlier. To these books new prefaces have been added, in which each of them is praised. I want you to have those passages, and there are certain other things as well. As for what I wrote to you about the ten commissioners, you did not, I believe, fully grasp it, because I had written in shorthand \textit{[Greek: dia sēmeiōn]}. What I was asking about was Gaius Tuditanus, whom I had heard from Hortensius to have been one of the ten. I see in Libo's annal that he was made praetor in the consulship of Publius Popilius and Publius Rupilius. Fourteen years before he was made praetor, could he have served as a commissioner --- unless he was made quaestor remarkably late? Which I do not think. For I see that he took the curule magistracies very easily in the regular years. As for Postumius, whose statue at the Isthmus you say you remember, I did not know he had been there. He is the man who was consul with Lucius Lucullus; whom you have supplied to me, just the right figure for that assembly \textit{[Greek: syllogon]}. So you will see, if you can, about the rest, so that we can stage our procession with the right cast \textit{[Greek: pompeusai kai tois prosōpois]}.
