% Ad Familiares 5.15 (ad-familiares-05-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. L. LVCCEIO Q. F.}

\ciceroSection{1} All your affection shows itself from every quarter in the letter I lately received from you --- not, indeed, an affection unknown to me, but welcome and longed-for nonetheless (I would say ``delightful,'' had I not put that word out of service for every season) --- and not for the one cause which you suspect, and in respect of which, using the most gentle and most loving words, you in fact reproach me sharply, but because, of the remedies which ought to exist for that immense wound, there are none.

\ciceroSection{2} For what then? Am I to take refuge with my friends? How many of them are there? We had them, on the whole, in common; of these some have perished, others have somehow hardened. With you I could indeed live, and would most of all want to; an old tie, affection, habit, the same studies --- what bond, I ask, is missing from our friendship? Can we, then, not be together? I do not, by Hercules, see what stands in the way; but the fact is that until now we have not been --- not when we were neighbours at Tusculum, not at Puteoli; for what shall I say of the city? --- where, the Forum being common ground, neighbourhood is not required.

\ciceroSection{3} But by some mischance our generation has fallen upon such times that, at the very moment when we ought most to be flourishing, we should be ashamed even to be alive. For what refuge could there be left for me, stripped of the adornments and consolations both of home and of the Forum? My letters, I suppose, which I use without ceasing; for what else can I do? But somehow those very studies seem to shut me out from their harbour and refuge, and as it were to reproach me with continuing in a life in which there is nothing but the prolonging of a most wretched time.

\ciceroSection{4} Do you wonder, then, that I am away from the city, in which my house can give me no pleasure, in which I have the utmost loathing for the times, for men, for the Forum, for the Senate-house? Hence I use my books in such a way that I spend all my time on them, not to seek from them an unbroken remedy, but a small forgetfulness of pain.

\ciceroSection{5} If you and I had done what did not even occur to our minds, on account of our daily fears, we should have spent all our time together, and your ill health would not have grieved me nor my sorrow you. What may be done of this, let us make good; for what is more suited to us both? Soon, then, I shall see you.
