% Ad Atticum 13.45 (ad-atticum-13-45) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Lamia was with me after you left, and brought me a letter that had been sent to him by Caesar. Although it had been written earlier than the one to Diochares, it nevertheless made plain that he would be here before the Roman Games. At the end of it was written that Lamia should get everything ready for the games and not let his master hurry in vain. On the face of these lines there seemed to be no doubt that he would arrive before that day, and Lamia said that Balbus too, when he had read the letter, was of the same opinion. I see that holidays have been added to my calendar --- but, if you love me, see to it that I am told just how many. About Baebius you will be able to act, and about the other neighbour Egnatius.

\ciceroSection{2} As for your urging me to spend those days on philosophical exposition --- you are urging on a runner; but you can see that during those days I must live with Dolabella. If it were not Torquatus's case that held me here, there would be enough days for me to make the run down to Puteoli and be back in time.

\ciceroSection{3} Lamia, in any case, had heard from Balbus (so it seemed) that a great deal of money was counted out at the house and ought to be divided as soon as possible, a great weight of silver; that the auction of M. Tullius Cicero's [property], apart from the estates, ought to be held at the earliest opportunity. I should be glad if you would write to me what you think. For my part, if I had to pick from the lot, I should not easily have chosen a more careful or more obliging or, by Hercules, more devoted man than Vestorius, to whom I have written a most considered letter; and the same, I take it, you have done. To me this seems enough. What do you say? One thing only stings me --- that we should not appear too negligent. So I shall be waiting for your letter.
