% Ad Familiares 6.11 (ad-familiares-06-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. TREBIANO.}

\ciceroSection{1} Dolabella up till now I had only held in high regard; I was under no obligation to him (for it had never fallen out that I had occasion to need him, and it was he who owed me, for not having failed him in his dangers); now I am so deeply bound by his kindness --- both before this, in the matter itself, and now, at this moment, in your restoration, on both of which he has done me the fullest possible service --- that I owe more to him than to anyone. On which I congratulate you so warmly that I would rather you congratulated me too than thanked me; thanks I have no use at all for; congratulation you can truly offer.

\ciceroSection{2} For the rest --- since your worth and your standing have opened the way back to your own people --- it lies with your wisdom and your greatness of spirit to forget what you have lost, and to think on what you have recovered. You will live with your own; you will live with us. You have acquired more of standing than you have lost of property; and even that property of yours would be the sweeter to enjoy, if there were any commonwealth left. Vestorius, our friend, has written to me that you give me the warmest thanks. This praise of yours is welcome to me indeed, and I am quite content for you to publish it both before others and, by Hercules, before our friend Siro. For what we do, we want to find approval especially with the wisest men. I am eager to see you, as soon as may be.
