% Ad Familiares 4.5 (ad-familiares-04-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{SERVIVS CICERONI S.}

\ciceroSection{1} After the news was brought to me of the death of your daughter Tullia, I bore it, as indeed I was bound to, with deep grief and pain, and I held the calamity to be one we shared. Had I been there, I should not have been wanting to you; I should have made my own sorrow plain to you face to face. Yet this kind of consolation is itself a miserable and bitter business --- precisely because the people through whom it has to be done, near kinsmen and intimate friends, are themselves struck by the same distress and cannot attempt it without many tears, so that they look more in need of being consoled by others than able to do that office for someone else. Even so, I have resolved to write you briefly the things that for the present have occurred to me --- not because I think they escape you, but because, hindered by grief, you may now see them less clearly.

\ciceroSection{2} What is it that so deeply moves you in this private grief of yours? Consider how fortune has dealt with us up to now: the things have been wrenched from us which ought to be no less dear to a man than his children --- country, honour, public standing, every office. With this one further loss added, what could have been added to your sorrow? Or what spirit, trained in those greater losses, ought not now to be calloused, and to count everything else as smaller? Or is it on \textit{her} behalf, tell me, that you grieve?

\ciceroSection{3} How often the thought must have come to you, as it has often come to us, that in these times it has not been worst with those who were allowed to exchange life for death without pain! What was there at this hour that could greatly invite her to go on living? What good thing, what hope, what comfort to the spirit? That she should keep on a life joined with some young man of the first rank? It was open to you, I imagine, suiting your standing, to choose a son-in-law from the young men of our day, to whose loyalty you would feel you could safely commit your children. Or that she should bear children of her own, whom she could rejoice to see flourishing, who could keep what their parent had handed down by their own efforts, would in due order stand for the offices in the commonwealth, would use their freedom in the affairs of their friends? Of all this, what was there that was not taken away before it was given? --- ``But it is an evil to lose one's children.'' An evil --- unless to undergo and bear out these times be a worse one.

\ciceroSection{4} A thought which has brought me no small consolation I want to recall to you, in case the same thing can lessen your own pain. As I was returning from Asia, sailing from Aegina toward Megara, I began to look at the regions all around. Behind me was Aegina, before me Megara, on my right the Piraeus, on my left Corinth --- cities which at one time were of the highest flourishing, and which now lie thrown down and broken in ruins before the eyes. I began to think with myself thus: ``Ah! we, us little men, take it as outrage if one of us has died or been killed, we whose life ought to be the shorter, when in one place so many corpses of towns lie cast down together. Will you, Servius, contain yourself, and remember that you were born a man?'' Believe me, by that thought I was no little steadied. Do the same, if it seems good to you: put it before your eyes. Just a moment ago, at one stroke, so many men of the highest distinction perished; so great a diminishing of the empire of the Roman people has taken place; all the provinces have been shaken to pieces --- and if some loss has been made in the little life-breath of one little woman, are you so deeply moved? She, if she had not died at this time, would still in a few years have had to die, since she had been born a mortal.

\ciceroSection{5} Even you, then --- call back your spirit and your thought from these things, and rather call to mind what is worthy of the person you are: that she lived as long as it was good for her to live; that she lived together with the commonwealth; that she saw you, her father, praetor, consul, augur; that she was married to young men of the first rank; that she had enjoyed almost every good thing; that she left life when the commonwealth fell. What is it that you, or she, could possibly complain of to fortune on this score? In the end, do not forget that you are Cicero --- the man who has been accustomed to instruct others and to give them counsel; do not imitate the bad doctors, who in the diseases of others profess to hold the science of medicine but cannot themselves cure themselves. Rather, what you are accustomed to prescribe to others, apply that to yourself, set it before your own mind.

\ciceroSection{6} There is no pain which length of time does not lessen and soften. For you to wait for that time, and not to meet the thing in your own wisdom, is a disgrace to you. And if there is any feeling among the dead, the love which she had for you, and the devotion she had to all her own --- certainly she does not want you to be doing what you are doing. Grant this much to her, to that dead woman; grant it to your other friends and intimates, who grieve at your grief; grant it to your country, that, if there should be any need, she may use your effort and counsel. In the end, since we have come into such a state of fortune that we must even be the slaves of this thing, do not give anyone occasion to think that you mourn not so much your daughter as the present times of the commonwealth and the victory of others. I am ashamed to write more to you on this matter, lest I seem to distrust your prudence. So if I set down this one thing, I shall make an end of writing: we have seen you bear good fortune more than once with magnificent grace, and earn great praise from it; now let us at last understand that you can bear bad fortune just as well, and that this burden does not weigh on you heavier than it should --- that out of all the virtues this one alone should not seem to be missing in you. As for myself: when I have heard that your spirit is calmer, I shall let you know how affairs are going here, and how the province is doing. Farewell.
