% Ad Atticum 12.22 (ad-atticum-12-22) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} As to Terentia: in laying the whole burden on me you do not show the indulgence toward me I know. These are the very wounds I cannot handle without the deepest groan. Temper it, then, I beg, so far as you can. I ask of you no more than you can do.

\ciceroSection{2} And you alone can see clearly what the truth of the matter is. About Rutilia, since you seem to be in doubt, write to me as soon as you know --- as soon as possible --- and also whether Clodia outlived her son, the ex-consul Decimus Brutus. The one can be ascertained from Marcellus, or certainly from Postumia; the other from Marcus Cotta, or from Syrus, or from Satyrus.

\ciceroSection{3} About the gardens, I beg you again and again. I must strain after them with all my own resources and those of the men I know will not fail me --- though my own will do. There are even things I could sell easily. But if I do not sell, and pay interest to the seller for no more than a year, I can manage what I want, if you help me. The readiest are Drusus's gardens; he is eager to sell. Lamia's, I think, come next; but he is away. Even so, sniff out what you can. Silius does not even use his, and he will easily be carried on those† interest payments. Treat the business as your own, and weigh not what my finances demand --- of which I take no care --- but what I want and why I want it.
