% Ad Atticum 13.27 (ad-atticum-13-27) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About the letter to Caesar: I have always thought, and quite rightly, that his people there should read it first. Otherwise we would have been remiss toward them and to ourselves, if we were going to give him offence, very nearly reckless. They for their part have spoken candidly; and I am grateful to them for not keeping back what they thought --- and best of all, they want so many things altered that I have no occasion to write the thing afresh. And yet on the Parthian war, what was I bound to look to except what I supposed he wanted? For what else was the burden of our letter but flattery \textit{[Greek: kolakeia]}? Had I wanted to urge what I myself thought best, words would hardly have failed me. So there is no need of the whole letter. Where no great success \textit{[Greek: epiteugma]} can be had, and even a failure \textit{[Greek: apoteugma]}, even one not large, would be troublesome --- what is the use of running the risk \textit{[Greek: parakinduneuein]}? Especially when this occurs to me: that he, since I have written nothing before, will assume I was not going to write anything until the whole war was finished. And I am even afraid he may suppose I meant it as a kind of sop \textit{[Greek: meiligma]} to balance my Cato. What more shall I say? I have come thoroughly to regret the thing, and in this matter nothing more agreeable to me could have happened than that our pains \textit{[Greek: spoud\=e]} were not approved. We would have run foul of those people too, your relative among them.

\ciceroSection{2} But back to the gardens. I really do not want you to go there unless it is much to your convenience; nothing presses. Whatever the outcome, let us put our effort into the Faberius business. About the day of the auction, though, let me know if you hear anything. The man who had come from the Cumanum, since he was reporting that Attica is altogether well and said he had a letter, I sent on to you at once.
