% Ad Atticum 13.20 (ad-atticum-13-20) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I have had a letter from Caesar, a letter of condolence, dated the day before the Kalends of May at Hispalis. What has been promulgated about enlarging the city I did not make out. I should very much like to know. That my good offices are welcome to Torquatus I am quite content to hear, and I shall not leave off adding to them.

\ciceroSection{2} As for the \textit{Pro Ligario}, on Tubero's wife and stepdaughter I can no longer make any addition (the piece is already in wide circulation), nor do I want to give offence to Tubero; he is wonderfully fond of fault-finding \textit{[Greek: philaitios]}. The theatre, certainly, you had a fine one.

\ciceroSection{3} I, although in this place I bear up easily enough, still want to see you. So I shall be there, as I have arranged. My brother has met with you, I take it. I am eager, then, to know what you did.

\ciceroSection{4} About reputation I am not really troubled --- though I had written to you at the time foolishly, ``nothing better''; it is not a thing to care about. And in his whole life one ought not to depart by so much as a nail's breadth from a right conscience. You see how philosophical \textit{[Greek: philosophos]}? Or do you suppose we have these works in hand for nothing? I should not have wanted you to have been so stung \textit{[Greek: dedechthai]} by it, when there was nothing in it. For I come back round to the same point. Do you suppose I care for anything at all but not failing him? That, of course, is what I am at --- the appearance of holding sway over the courts! Far be it from me \textit{[Greek: me gar autois]}. I could wish I bore my own private griefs as easily as I despise these things. Do you imagine I had wanted some piece of work that should not turn out finished? It is not, of course, a man's own opinion that prevails. But still, what was then done I cannot but approve --- and yet I can quite handsomely not care about it, as in fact I do not. But too much about trifles.
