% Ad Atticum 13.26 (ad-atticum-13-26) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About Vergilius's share I thoroughly approve. So that is the way you will proceed. And indeed that one will be the first choice, Clodia's the next. If neither answers, I am afraid I shall make a wreck of things and rush at Drusus. I am beyond restraint in my desire for the thing you know of. So I keep coming back round to the Tusculan villa. Anything rather than have the business not closed this summer.

\ciceroSection{2} As for me, things being as they are with us, I have no place where I can find more ease than at Astura. But since those who are with me, I take it, cannot bear my gloom and are hurrying home, I shall set out from here --- as I wrote to you --- so as not to seem to have been left behind, even though I could have stayed on. Where, though? From Lanuvium I am, for my part, trying to push on to the Tusculan villa. But I shall let you know at once. You will see the letters out. As for me, the amount I am writing is past belief --- even at night, for I get no sleep. Yesterday I even completed the letter to Caesar; that was what you wanted. There was no harm in its being written, if you happened to think it useful; as things stand now, there is really no necessity to send it. But on that point, do as you see fit. I shall send you a copy, from Lanuvium perhaps, unless I do it from Rome. But you shall know tomorrow.
