% Ad Atticum 12.51 (ad-atticum-12-51) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I have Tiro back sooner than I had feared. Nicias has come too, and I was hearing today that Valerius would arrive. Many as they may be, I shall be more alone all the same than I would be if you were the only one with me. But I am waiting for you, at any rate after Peducaeus' --- though you yourself hint at something rather sooner. Well, as you can manage it.

\ciceroSection{2} About Vergilius, as you write. This, though, I should like to know: when the auction. The letter to Caesar I see meets with your approval for sending. What more need I say? It met with my fullest approval too, and all the more because there is nothing in it but what befits the best sort of citizen --- the best, however, that the times allow, which all statesmen \textit{[Greek: politikoi]} enjoin one to obey. But you know our intention was that those men read it first. So you will see to that. But unless you can see plainly that they approve, it is not to be sent. Whether they really feel so, or only pretend, you will perceive. For me, pretence will count as rejection. Probe this point \textit{[Greek: touto de m\=el\=os\=ei]}.

\ciceroSection{3} About Caerellia, Tiro told me what you thought: that to be in debt is not in keeping with my dignity, and that an assignment meets your approval. The one I fear; the other I do not reckon among things to be feared. But of this and much else when we are face to face. The settling of the Caerellian debt, however, should be held off, if you think proper, until we know how things stand with Meto and with Faberius.
