% Ad Atticum 12.23 (ad-atticum-12-23) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I had supposed --- since the opening of your letter was of that kind --- that you were going to write me something new, and that even though I should not care what was happening in Spain you would write it nonetheless; but you have answered my letter, plainly, as if from Forum and Senate-house. Yet the household, as you say, is the Forum. What use to me is the household itself when I am without the Forum? We are finished, finished, Atticus --- finished long since, in fact, but only now confessing it, since we have lost the one thing that held us. So I keep to solitude; and yet, if any business takes me to town, I shall strain, if I can in any way (and I shall be able), that no one but you perceive my grief --- and if it can in any way be done, not even you. There is even this further ground for not coming. You remember what Aledius enquired of you. If they are troublesome to you even now, what do you suppose it will be like if I come?

\ciceroSection{2} About Terentia, see to it as you write, and relieve me of that addition to my greatest sorrows --- not a small addition. And so that you may know I grieve in such a way as not to lie prostrate: in your \textit{Annals} it is written under what consuls Carneades and that embassy came to Rome. What I am asking now is what the cause was (about Oropus, I think, but I am not sure) and, if so, what the points of contention were. Further: who at that time was the eminent Epicurean and had charge of the Gardens at Athens, and who were also the distinguished \textit{statesmen} [Greek: politikoi] at Athens. These things, I think, can be found out from the work of Apollodorus too.

\ciceroSection{3} About Attica: troubling, but since it is light, I am confident she is well. About Gamala: I had no doubt. For from what spring should Ligus, his father, be so favoured? As for me, what am I to say --- when, even if every wish of mine came true, I cannot be relieved? About Drusus's gardens, you write at what price they were on offer; I had heard the same myself, and yesterday, I think, wrote you so; but at whatever price, that is well bought which one must have. For me, however you may judge it (for I know myself what I judge of myself), there is some lightening --- if not of grief, at least of obligation owed. I have written to Sicca, since he is on terms with Lucius Cotta. If nothing comes of the Transtiberine properties, Cotta has a place at Ostia, in a most frequented quarter --- a small place, but for this purpose more than enough. I should be glad if you would think it over. And do not be alarmed by the asking prices of these gardens. I no longer need silver or fine clothes or pleasant retreats of any kind; this is what I need. I see, too, by whom I can be helped. But speak with Silius; nothing would be better. I have given the same charge to Sicca too. He has written back that he has arranged a meeting with him. So he will write to me what he has accomplished, and you will see to it.
