% Ad Atticum 13.40 (ad-atticum-13-40) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Really? Brutus reports that the man is coming over to the loyalists? \textit{Good news indeed} \textit{[Greek: euangelia]}. But where will he find them? --- unless he has hanged himself in the meantime. \dag\ As for this place, it is propped up as it is. \dag\ So where is that artistic conceit of yours \textit{[Greek: philotechnema]} that I saw in the Parthenon --- Ahala and Brutus? But what is he to do? That part is excellent; yet not even the author of every disgrace has done well by us. And I was afraid Brutus too was growing fond of him --- so he had hinted in that letter to me. Well, I wish you had had at least a taste of the talk over dinner. But face to face, as you propose.

\ciceroSection{2} Still --- what is your advice? Am I to fly there, or stay put? For my part, I am stuck in my books and have no wish to receive the man here; to whom, as I hear, the father is going today, to Saxa, \dag\ in bitterness \dag. Astonishing how hostile he was on his way, ready for me to take the boy to task. But I myself \textit{have been duped} \textit{[Greek: kekepphomai]}; well then, no more of that. Do you, though, consider what you make of my coming up to town, and \textit{the whole picture} \textit{[Greek: ta hola]}: tomorrow, if it can be made out, let me know first thing in the morning.
