% Ad Familiares 15.16 (ad-familiares-15-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. C. CASSIO}

\ciceroSection{1} I suspect a little shame is coming over you by now --- this is the third letter of mine to overtake you before you have so much as scratched out a slip or a single letter. But I am not pressing the point; for I shall wait for longer ones from you --- or rather, exact them. As for me, if I always had a courier to give them to, I would hand over three in an hour. For somehow it happens that you seem present, as if face to face, when I write anything to you --- and that not by way of images of mental impressions, as your new friends say, who suppose that even the impressions of thought are excited by Catian spectres. (Lest it slip your notice: Catius the Insubrian, an Epicurean, who died not long ago, calls \textit{spectra} what the man of Gargettus, and Democritus before him, called images \textit{[Greek: eidōla]}; and he calls these things \textit{[Greek: dianoētikas phantasias]} --- impressions of thought --- where his friends speak of mere \textit{[Greek: kat' eidōlōn phantasias]} --- impressions by way of images.)

\ciceroSection{2} Even granting that the eyes could be struck by these spectres, since whatever you wish for runs of itself into them, how the mind could be so struck I do not see. You will have to teach me, when you have come back safely, whether your spectre is in my power --- so that, as soon as I have taken a fancy to think of you, it presents itself --- and whether, not just of you, who cling fast in my marrow, but also if I begin to think of the island of Britain, its image \textit{[Greek: eidōlon]} will come winging into my breast.

\ciceroSection{3} But this for later. I am sounding you out, to see in what spirit you take it. For if you are going to lose your temper and take it amiss, I shall say more, and I shall demand that you be restored to the school \textit{[Greek: hairesei]} from which you have been ejected ``by men under arms.'' To that interdict it is not customary to add ``within the year.'' So if it is now two years or three since you sent virtue packing, seduced by the lures of pleasure, the case will still be open for us. And yet --- to whom am I talking? To a man of the greatest courage, who, from the moment you set foot in the Forum, has done nothing that was not entirely filled with the highest dignity. In that very school \textit{[Greek: hairesei]} of yours I am afraid there is more sinew than I had supposed --- if at any rate you really endorse it. ``How did this come into your head?'' you will say. Because I had nothing else to write about; for of public affairs I can write nothing --- nor does it suit me to write what I think.
