% Ad Atticum 12.19 (ad-atticum-12-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The place here is indeed pleasant, set in the sea itself in such a way that it can be looked across to from Antium and from Circeii; but we must work out some plan by which, in any change of owners --- and these may come without number through endless posterity, if only the present order holds --- that consecrated thing may yet remain. For my part I now have no need of revenues, and can be content with little. I sometimes consider acquiring some gardens across the Tiber, and chiefly for this reason: I see nothing else that can be so much frequented. Which gardens, we will see when we meet --- in such a way, however, that the shrine is finished this summer. You meanwhile, settle with Apella of Chios about the columns.

\ciceroSection{2} As to what you write about Cocceius and Libo, I approve, and most of all your handling of my judgeship. About the surety, if you have made anything out --- and even so, what Cornificius's agents say I should like to know --- arrange it so that you give the matter, busy as you are, not too much labour. About Antonius, Balbus too has written to me jointly with Oppius, and added that this was your wish, that I should not be agitated. I thanked them. But you, as I have already written to you before, I want to know that I was not agitated by that news and shall not now be agitated by any.

\ciceroSection{3} If Pansa has set off today, as you thought, begin from now on to write to me about Brutus's coming --- what you expect, that is to say, by what dates. If you know where he is by now, you will easily come at it by guess.

\ciceroSection{4} As to what you write to Tiro about Terentia, I beg you, my dear Atticus, take the whole business in hand. You see that an obligation of my own is at stake, of which you are conscious, and, as some think, the standing of young Cicero. For me, it is much more the former that moves me, as the more sacred and the older claim, especially since I think the second neither sound nor secure.
