% Ad Atticum 13.1 (ad-atticum-13-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} You have written to Cicero in such terms that nothing could have been written more firmly or more measuredly, nor more nearly as I myself would have most wished; and most prudently of all to the Tullii.

\ciceroSection{2} Either, then, this will have its effect, or we shall try something else. As to the money, I see that every diligence is being applied --- or rather has already been applied --- on your part. If you bring it off, I shall have the gardens from your hand. There is no kind of holding I should prefer, above all of course for that purpose which has been undertaken; you take away from me the haste of it, since you promise --- or rather, undertake --- for the summer. And then too, for the running out of life and for the lessening of grief, no fitter thing can be found for me; the longing for which sometimes drives me to want to urge you on. But I check myself: I do not doubt that, in whatever you suppose I greatly want, you will outrun even me in eagerness. So I count it already as done.

\ciceroSection{3} I am waiting to hear what those people think about the letter to Caesar. Nicias loves you, as he ought, and is much delighted by the memory you keep of him. As for our Peducaeus, I am very fond of him: as much as I valued the father, $\dagger$wholly in this one$\dagger$ I love him too, in himself, as much as I loved the other --- and you most of all, for wishing this to be done by each of us. If you have looked over the gardens and let me know about the letter, you will have given me something to write to you about; if not, I shall write something all the same. There will never be a lack.
