% Ad Familiares 9.6 (ad-familiares-09-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO VARRONI}

\ciceroSection{1} Our friend Caninius, speaking in your name, has reminded me to write to you if there were anything I thought you should know. Well, then: Caesar's arrival is being looked for, as you are no doubt aware. He had written, I believe, that he would come to his place at Alsium; but his people wrote back asking him not to --- many would be a nuisance to him, and he to many; Ostia, they thought, would be a more convenient landing. I could not see what difference it made; but Hirtius told me that he and Balbus and Oppius had written advising it --- men whom I have come to know as great friends of yours.

\ciceroSection{2} I wanted you to know this so that you might know where to prepare your hospitality --- or rather, at both places (since it is uncertain what he will do) --- and at the same time to give you some sign that I am on familiar terms with those men and party to their counsels. Why I should not be, I see no reason: it is not the same thing to bear what must be borne and to approve what is not to be approved. Though, for that matter, what I should now disapprove of I no longer know, beyond the beginnings of the business; for those were a matter of will. I saw with my own eyes, while you were absent, our friends wanting the war, and this man not so much wanting it as not fearing it. Those, then, were matters of choice; the rest was forced. And that one side or the other should win, that was unavoidable.

\ciceroSection{3} I know that you were always with me in our mourning, when we saw both that vast evil --- the destruction of one army or the other and of their commanders --- and beyond that, that the very extremity of all evils was a victory in civil war. And I confess I feared even that victory from the side to which we had come: for the most disengaged of them were making cruel threats, and both your inclination and my talking were hateful to them. As things have turned out, had our side prevailed, they would have been wildly intemperate: they were furious with us, as though we had decided anything for our own safety that we had not equally counselled for theirs, or as though it were more useful to the commonwealth for them to flee even for help to wild beasts than either to die, or to live with some hope ---

\ciceroSection{4} if not the best hope, at any rate some. ``But we live in a disordered commonwealth.'' Who denies it? Let those see to it who have laid up no resources for themselves against every condition of life: I have drifted further than I wished, but the way back to my point is this. As I have always counted you a great man, so now in these tempests you stand almost alone in port, and reap the greatest fruits that learning bears --- that you turn your mind to, and handle in your hands, those things whose use and whose pleasure together are to be preferred to all the doings and all the pleasures of those people. For my part, these Tusculan days of yours seem to me the very pattern of a life; and I would gladly yield up to all men all advantages, if it were granted me, with no force breaking in, to live in that way.

\ciceroSection{5} And we, as far as we can, imitate it; we take our rest most gladly in our studies. For who will not grant us this much: that, since our country either cannot or will not use our service, we should turn back to that life which many learned men, perhaps not rightly, but in any case many, have thought even preferable to the service of the state? Studies, then, which in the verdict of great men carry a kind of furlough from public duty: with the commonwealth's leave, why not draw on them to the full?

\ciceroSection{6} But I am doing more than Caninius charged me with. He had asked, fairly enough, that if I knew anything you did not know, I should tell it to you --- and here I am, narrating to you what you know better than I, the narrator, do. So I will do what I was asked: that you may know nothing of what belongs to the present hour, whatever I shall have learned of it I will report to you.
