% Ad Familiares 6.5 (ad-familiares-06-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. A. CAECINAE.}

\ciceroSection{1} Every time I see your son --- and I see him almost daily --- I promise him my zeal and my effort without any reservation of labour or of engagements or of season, and my good offices and my authority with this one reservation: as far as I have weight and as far as I can. Your book has been read, and continues to be read, by me carefully, and is kept under most careful guard. Your affairs and your fortunes are my deepest concern; and they appear to me, day by day, easier and better; and I see that they are the deep concern of many others; of whose zeal and of whose hopes for you I am certain your son has written you in full.

\ciceroSection{2} In matters which we can only reach by conjecture, I do not presume to see farther myself than I have persuaded myself you see and understand. But because it is possible you may turn these things over with a more disordered spirit, I think it is my part to set out what I think. The nature of things is such, and such is the run of the times, that this fortune of yours, and of the others, cannot be lasting; nor can so bitter an injustice cling to so good a cause and to so good citizens.

\ciceroSection{3} And so, to the hope, beyond the common run, that we hold of you personally --- not only for your worth and your virtue (these are ornaments you share with others), but for special reasons of your own: your outstanding talent and your highest virtue, to which, by Hercules, this man in whose power we stand assigns much weight --- to that hope are added these. And so you would not have spent an instant in that fortune of yours, had he not supposed that he had been injured precisely by the talent of yours by which he is delighted; and even that injury is softened day by day, and word comes to us, from those who live at his side, that this very reputation of yours for talent will count powerfully with him on your behalf.

\ciceroSection{4} Therefore, first, see to it that you are of a great and brave spirit (for you were born so, were so brought up, so taught, and are so known to be, that this must be your bearing); then have, too, the firmest hope, for the reasons I have written. And as for me --- I want you to trust that everything is most fully prepared and ready for you and for your children: this is what the long age of our love, and my way with my own people, and the many services you have rendered me, demand.
