% Ad Familiares 6.14 (ad-familiares-06-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO LIGARIO.}

\ciceroSection{1} Know that I am spending all my labour, all my effort, all my care, all my zeal on your restoration. For I have always loved you most warmly; and now your brothers --- whom I have embraced with the same fullness of goodwill that I bear toward you --- with their singular devotion and brotherly love, leave me no room to let slip any service or any occasion of duty and zeal on your behalf. But what I am doing and have done for you I would rather you learned from their letters than from mine; what I hope, what I trust to, what I hold for certain about your restoration --- that I want to be set out to you in my own words. For if there is any man who is timid in great and dangerous affairs, and always more afraid of the bad outcomes of things than hopeful of the good, I am that man; and if that is a fault, I confess that I am not free of it.

\ciceroSection{2} And yet, this same man, when on the fifth day before the kalends of the first intercalary month I had gone in the early morning, at the entreaty of your brothers, to Caesar, and had endured every indignity and irritation in approaching and obtaining audience with him; and when your brothers and kinsmen lay prostrate at his feet, and I had spoken what your cause, what your present situation required --- not only from Caesar's reply, which was indeed gentle and generous, but from his eyes too, and his expression, and from many other signs besides, which I could more easily make out than write down, I came away with this conviction: that your restoration was not in doubt.

\ciceroSection{3} Therefore see to it that you keep your spirit great and brave, and, if you bore the wildest waters wisely, that you bear the calmer ones with gladness. I, for my part, shall stand by your affairs as if they were the most desperate, and shall on your behalf, as I have done up to now, most gladly entreat not only Caesar but all his friends besides --- whom I have found to be the warmest friends to me. Farewell.
