% Ad Familiares 9.26 (ad-familiares-09-26) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO PAETO S. D.}

\ciceroSection{1} I had been at table since the ninth hour when I scratched out the draft of these lines for you on a notebook. You will say: ``Where?'' At Volumnius Eutrapelus's --- and, mind you, above me Atticus, below me Verrius --- friends of yours both. You wonder that we make our servitude so cheerful? What, then, am I to do? (I consult you, you that go and hear a philosopher.) Should I torment myself, rack myself? What would I get by it? And, in any case, how long should I keep at it? ``Live in books,'' you say. Well, do you suppose I am doing anything else, or that I could go on living, if I were not living in books? But even of books there is, not satiety, but a certain measure; and when I have come away from them --- though my interest in dinner is the least possible (the one \textit{question} \textit{[Greek: z\=et\=ema]} you put to Dion the philosopher) --- still I find nothing better to do, before I betake myself to sleep. Hear the rest: below Eutrapelus reclined Cytheris.

\ciceroSection{2} ``So,'' you say, ``at \emph{that} party was the famous Cicero, at whom they used to stare, on whose face the Greeks turned their faces to gaze?'' By Hercules, I did not suspect she would be there. But then again not even the famous Aristippus, the Socratic, blushed when it was thrown at him that he had Lais. ``I have her,'' he said, ``I am not had by her.'' (It runs better in Greek; translate it yourself, if you like.) None of those things ever moved me even when I was young, much less now that I am old. I take my pleasure in the dinner-party; there I say whatever first falls on the floor, as the saying goes, and I transmute my groaning into the heartiest of laughter.

\ciceroSection{3} Or did you do better, who actually mocked even the philosopher when, on his saying ``if anyone has any question to ask,'' you said that since morning you had been asking for your dinner? That blockhead supposed you were going to ask whether there is one heaven or innumerable. What is that to you? --- and, by Hercules, what is dinner to you either, particularly? So this is how the life runs: every day something is read or written; then, that we may not deny our friends their portion, we feast together --- not only not against the law, if any law there now is, but well within the law, and considerably so. Therefore there is nothing for which to dread my arrival; you will be receiving a guest of little appetite, but plenty of fun.
