% Ad Familiares 15.17 (ad-familiares-15-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO C. CASSIO S.}

\ciceroSection{1} You have couriers running back-to-front --- though for my part they do not offend me. Still, when they leave me, they demand letters; when they come to me, they bring none. And they would do that very thing rather more conveniently if they would give me a little time to write; but they arrive already capped for travel, and say their companions are waiting at the gate. So you will pardon me; this is another short one you will have. But you can expect \textit{[Greek: panta peri pantōn]} --- everything about everything. And yet why am I making excuses to you, when your couriers come empty to me, and return to you with letters?

\ciceroSection{2} We here --- if I am to write you something all the same --- had buried the elder P. Sulla. Some said he was killed by brigands, others that it was indigestion; the people did not care, since it was settled that he was cremated. This, in your wisdom, you will bear with composure --- though we have lost the public face \textit{[Greek: prosōpon poleōs]} of the city. Caesar, they think, will take it amiss, fearing that the auction-spear has cooled off. Mindius the meat-seller and Attius the perfumer were thoroughly delighted to have lost a rival.

\ciceroSection{3} Of Spain nothing new, but expectation runs very high. The rumours are rather grim, but unattributed \textit{[Greek: adespotoi]}. Our friend Pansa set out in his general's cloak three days before the kalends of January, so that anyone could see what you have lately begun to doubt --- that the noble \textit{[Greek: to kalon di' hauto haireton]} is to be chosen for its own sake. For because he has lifted many out of their miseries and shown himself a human being in these evils, a remarkable goodwill of decent men has gone with him.

\ciceroSection{4} As for your staying at Brundisium so long, I thoroughly approve and am glad of it, and by Hercules I think you will be acting wisely if you keep clear of meddling \textit{[Greek: akenospoudos]} --- to us at any rate, who love you, it will be welcome. And please, when you next send anything home in the way of a letter, remember me. I shall never let anyone go to you, when I know of it, without a letter from me. Farewell.
