% Ad Familiares 4.3 (ad-familiares-04-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. SER. SVLPICIO.}

\ciceroSection{1} That you are deeply troubled, and that in our common miseries you are wrung by a peculiar grief of your own, is what many report to me daily. I am hardly surprised at it, and in some way I recognize my own feeling in it; yet I grieve that a man endowed with wisdom almost unique should take less pleasure in his own goods than pain in the misfortunes of others. As for me --- though I yield to no one in the trouble I have drawn from the ruin and the plague of the state --- still many things now console me, and above all the consciousness of my own counsels. For long beforehand, as if from some watchtower, I saw the coming storm; nor did I see it of my own foresight only, but far more on your warning and your prophecy. For though I was absent for the greater part of your consulship, even at a distance I knew what your judgement was as to guarding against this ruinous war and warning of it, and I myself was present in the first months of your consulship, when you most carefully warned the Senate, gathering up all the civil wars, that they should fear what they remembered, and should know that, since earlier men, with no such precedent recorded in the state before them, had nevertheless been so cruel, whoever should later crush the state by arms would be far more intolerable. For what is done by precedent men think also done by right; and they add and contribute, of their own, something more --- indeed much more.

\ciceroSection{2} You must remember, then, that those who did not follow your authority and counsel perished through their own folly, when by your prudence they could have been kept safe. You will say: ``What consolation is that to me, in such darkness, and as it were among the ruined walls of the state?'' The grief, to be sure, scarcely admits consolation: so total is the loss of every good, so total the despair of recovery. Yet even so, Caesar himself judges you as he does, and so do all the citizens: that, with the rest of the lights extinguished, your sanctity, your prudence, your dignity still shine on like some lamp. These ought to weigh greatly with you in easing your distress. And that you are away from your own people is the more easily to be borne for this reason: that at the same time you are away from many and great distresses, all of which I would write out to you in full, but that I fear you would learn, in your absence, things which, just because you do not see them, make me think you in a better case than we who do.

\ciceroSection{3} Thus far I judge my consolation rightly applied: insofar as you could be informed, by a most devoted friend, of those things by which your distress might be eased. The rest lies in yourself, and neither is unknown to me, nor --- as I at any rate feel --- among the lesser comforts, but by far the greatest. I prove them daily in my own experience, and find that they seem to bring me salvation itself. As for you, I recall from boyhood that you were a most ardent student of every branch of learning, and that with the keenest zeal and care you mastered everything which the wisest men had handed down for living well; pursuits which, even in the best of times, could afford both use and pleasure, but in these times we have nothing else in which to find rest. I shall do nothing presumptuous, nor urge a man so endowed with knowledge and with such a nature to return to those arts to which from the earliest period of your life you devoted your study;

\ciceroSection{4} I shall only say --- and this I hope you will approve --- that I myself, once I saw that there was no place left for that art to which I had given my study, neither in the Senate-house nor in the Forum, have turned all my care and labor to philosophy. For your own excellent and singular knowledge little more room is left than for mine. Therefore I do not, indeed, urge it on you; but I have so persuaded myself that you too occupy yourself with the same studies, which, even if they brought less profit, would yet draw the mind away from anxiety. Your son Servius, in all the liberal arts, and most of all in this one in which I have written that I find my rest, so engages himself that he excels in it; by me he is loved in such a way that I would yield place to you, but to no one else; and he repays me with a gratitude in which --- as is easily seen --- he reckons that, by his attention and respect for me, he is doing what is most welcome to you as well.
