% Ad Atticum 12.1 (ad-atticum-12-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the eleventh day after I had left you, I scratched out these few lines, setting out from the villa before daybreak. I was thinking to spend that day at the Anagninum, the next at the Tusculanum, and one day there. So on the fifth before the Kalends I shall be at the appointed meeting. And how I wish I could run straight to my Tullia's embrace, to a kiss from Attica! Write to me, I beg you, about precisely this, so that while I am at the Tusculanum I may know what she is chattering about; or if she is in the country, what she writes to you. Meanwhile, do either send her a greeting from me or give her one, and the same to Pilia. And although we shall be meeting any moment now, write to me anyhow if you have anything.

\ciceroSection{2} As I was folding up this letter, a courier reached me through the night with a letter of yours; on reading it, I was naturally much distressed about Attica's little fever. As for the rest I had been expecting from your letter, I have it all now. But where you write that the morning fire is \textit{[Greek: gerontikon]} --- the mark of an old man --- it is more \textit{[Greek: gerontikon]} still for the little memory to wobble. For I had given a letter to Axius on the fourth before the Kalends, to you on the third, and to Quintus on the day I arrived --- that is, on the fifth before the Kalends. So this is what you will have: nothing new. What need, then, was there for a letter? What need is there, when we are face to face and chatter whatever first comes to mind? Surely there is something in mere conversation \textit{[Greek: lesch\=e]} which has, even when nothing lies beneath it, a sweetness in the very talking.
