% Ad Familiares 7.33 (ad-familiares-07-33) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D VOLVMNIO.}

\ciceroSection{1} That you are missing my declamations costs you nothing; that you would envy Hirtius, if you did not love him, there was no cause for envy --- unless, perhaps, you envy him his eloquence itself rather than the fact that he is being heard by me. For we, my most charming Volumnius, are plainly nothing at all --- or at any rate displease ourselves --- now that we are deprived of those old comrades, the friends to whose applause we used to flourish: so that even when at any time we have launched something worthy of our name, we groan that ``these weapons are being practised on a feathered body, not on an armed one,'' as Philoctetes says in Accius, ``with glory thrown aside.''

\ciceroSection{2} But still, everything will look brighter to me if you come; though you will be arriving, as you know yourself, into what is practically a collision of overwhelming occupations. If we ride those out as we hope to, then truly I shall bid a hearty farewell to the Forum and to the Senate House, and live much with you and with our mutual admirers. For both your Cassius and our Dolabella --- or rather, both of them ours --- are held by the same pursuits, and have in me a most receptive listener. There is a need here of your finely-filed and polished judgement, and of that more recondite learning of yours \textit{[Greek corrupt: lectio incerta]}, by which you often make me more diffident in speaking. For I have made up my mind --- if only Caesar will either permit this or wish it --- to lay aside at last the role in which I have often won approval even from him, and to bury myself wholly in literary studies, and to enjoy in your company and that of others like-minded a leisure of the most honourable kind. I could wish that you had not been afraid that I might read your letter through more than once, if perhaps you sent me longer ones --- as you write you might have done. And I should like you to lay it down henceforth that the longer your letters to me are, the more welcome they will be.
