% Ad Familiares 6.1 (ad-familiares-06-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. A. TORQVATO.}

\ciceroSection{1} The confusion of everything is now such that each man regrets his own fortune above all others, and there is no one who would not rather be anywhere than where he is. And yet I have no doubt that, at this time, for a good man, to be at Rome is the most wretched lot of all. For although wherever a man is, his perception is the same, and the bitterness from the ruin of public and private fortunes the same, still the eyes increase the pain: they force on him the sight of what other men only hear about, and do not let him turn his mind away from the miseries. So although you must be tormented by the longing for many things, still that grief by which I hear you are above all worn down --- that you are not at Rome --- relieve your spirit of it. For though you long for your own and what is yours under great distress, still those things you miss are holding to their condition, and would not hold it any better were you here, and stand under no peril of their own; and you ought not, when you think of your own people, either to demand for them some special fortune or to refuse them the fortune that is common to all.

\ciceroSection{2} As for yourself, Torquatus, the right way to handle your spirit is not to bring despair or fear into the council of your thoughts. For neither has the man who up to now has been more unjust to you than your dignity called for failed to give large signs of a temper softened toward you; and yet that man himself, from whom safety is sought, has not yet a settled or assured plan for his own safety. And since the issues of all wars are uncertain, I see clearly that on the side of the one victory you are in no danger that is at all separate from the destruction of every man, and on the side of the other I know for certain you have never had anything to fear from yourself.

\ciceroSection{3} It remains that the very thing which I put forward as a kind of consolation should torment you most: the common peril of the commonwealth. And for an evil so great, however many things learned men may say, still I fear that no true consolation can be found except that one which is as great as the strength and sinew that is in each man's spirit. For if to think rightly and to act rightly is enough for living well and happily, I fear it is unholy to call wretched the man who can sustain himself by his consciousness of the best counsels. Nor do I think that we, drawn on by the prizes of victory, once left our country and our children and our fortunes; we seemed rather to be following a certain duty, just and pious and owed to the commonwealth and to our own standing; and when we did so, we were not so deranged as to suppose victory was assured to us.

\ciceroSection{4} Therefore, if what has happened is the very thing that, on entering the cause, we set before ourselves as something that could happen, we ought not to fall in spirit as though something had happened we never thought possible. Let us therefore have the disposition that reason and truth prescribe: to think that nothing in life is required of us except to be without fault, and that, being free of that, we should bear all human things with calm and moderation. And this argument tends to this end: that, with everything ruined, virtue itself can still sustain itself. But if there is any hope for the commonwealth, you ought not, whatever the future condition turns out to be, to be without your share in it.

\ciceroSection{5} And as I was writing this, it came into my mind that I am the very man whose despair you used to censure, whom you used to rouse by your authority when I was hesitating and losing faith. At that time it was not our cause that I disapproved of, but our policy. For I saw that we were too late in opposing those arms which had long since been built up by our own hand, and it grieved me that public right was being argued by javelins and swords, not by our policies and authority. Nor when I was saying that what has happened would happen was I divining the future; rather, what I saw could happen --- and what would be ruinous if it did --- that I feared would come about. Especially since, if I were forced to give a sure word on one outcome or the other, I could promise with more confidence that the thing which has happened would happen. For we were superior in those resources which do not come out into the field, but in the use of arms and the strength of soldiers we were inferior. But you, now --- summon up that spirit, I beg, which you then judged I ought to have.

\ciceroSection{6} I have written this much because your man Philargyrus, when I was asking him about everything to do with you, told me --- with what seemed to me the most faithful spirit --- that you are sometimes more deeply troubled than you should be. You should not be so, nor doubt that either, under some form of commonwealth, you will be the man you ought to be, or, with the commonwealth ruined, your case will be no worse than every other man's. And this present time, when we are all without breath and in suspense, you have all the more reason to bear with a steady spirit: because you are in the city in which reason and moderation in living were born and reared, and because you have Servius Sulpicius, whom you have always uniquely loved --- who is, I am sure, consoling you by both his goodwill and his wisdom. Had we followed his authority and his counsel, we should have undergone the power of the man in the toga rather than the victory of the man in arms.

\ciceroSection{7} But these things, perhaps, have run on longer than was necessary. What is greater, I shall set out more briefly. I have no one to whom I owe more than to you; those to whom I owed as much as you understand --- those the chances of this war have torn from me; and what kind of man I am at this moment, I know. But because no one is so beaten down that, if he sets himself to nothing other than what he is doing, he cannot still put a hand to something and accomplish something, I would have you certain that all my counsel, my labour, my zeal is owed to you and to your children.
