% Ad Familiares 13.69 (ad-familiares-13-69) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO P. SERVILIO CONLEGAE S. P.}

\ciceroSection{1} C. Curtius Mithres is, as you know, the freedman of my close friend Postumus; but he attends and respects me as much as he does that very patron of his. At his house in Ephesus I have stayed, every time I have been there, as if it were my own home, and many occasions have arisen in which I have had proof both of his good will toward me and of his fidelity. So if I or any of my people have any need in Asia, it is to him that I have made a habit of writing, and I make use of his services and his fidelity --- and of his house and his property too --- as if they were my own. I have written this to you at greater length so that you might understand that I am not writing of him commonly or perfunctorily, but as for a man who is one of my intimates and bound to me by the closest tie.

\ciceroSection{2} I ask you, then, in the dispute he has with a certain man of Colophon over an estate, and in his other affairs, to oblige him for my sake so far as your good faith will allow and so far as you can do so consistently with your convenience --- though, knowing as I do the moderation of the man, he will be a burden to you in nothing. If, by my recommendation and by his own honesty, he obtains your good opinion of him, he will think he has obtained everything. That you should take him into your trust, then, and count him among your own --- this I most earnestly beg of you, again and again. As for me, whatever I judge to be your wish, and whatever I judge to bear on your interests, I shall attend to with zeal and care in every respect.
