% Ad Familiares 9.17 (ad-familiares-09-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO L. PAPIRIO PAETO.}

\ciceroSection{1} Are you not an absurd fellow, asking me what I think will become of those townships and their land --- you, who have had our friend Balbus staying at your house? As though I knew anything that the man does not know, or, if I ever do know something, did not regularly learn it from him. On the contrary, if you love me, you tell me what is to become of us; for you have had him in your power, in such conditions that you could have got it out of him either sober or, failing that, drunk. As for me, my Paetus, I do not ask these things --- first, because it is now going on four years that we have been living on bonus time, if it is bonus time, or living at all, to outlive the commonwealth; second, because I think I too know what is going to happen. What will happen is whatever those who have the strength want; and the strength will always be with arms. We must therefore be content with whatever is conceded. The man who could not put up with this should have died.

\ciceroSection{2} They are now measuring out the Veientine and Capenate land; this is not far from Tusculum. Yet I am not afraid. I enjoy it while I may; I pray I may always; if that is denied me, then since, brave man and philosopher both as I am, I have counted living the finest of things, I cannot help loving the man through whose kindness I have come by it. As for him --- if he wished the commonwealth to be such as perhaps he himself wishes too, and as we all ought to wish, still he has no way to bring it about: so thoroughly has he bound himself up with so many.

\ciceroSection{3} But I am going on too far; for I am writing to you. Yet know this much: not only do I, who have no part in the deliberations, not know what is to be --- the chief man himself does not know either. We are slaves to him, he to circumstances. So neither can he know what the times will demand, nor we what he is planning. I had not written you back on these matters before --- not that I am usually a slacker, particularly with the pen, but because, having nothing settled, I did not want either to give you anxiety from my own doubt or hope from any assurance. One thing, however, I will add, which is altogether true: I have heard nothing yet, in this present hour, of the danger you speak of. You, for your part, in keeping with your wisdom, should hope for the best, brace for the worst, and bear whatever comes.
