% Ad Atticum 12.7 (ad-atticum-12-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Everything you wanted I have written in a set of notes and given to Eros: briefly, but even more than you asked, on the matter of young Cicero. The first thought of this you yourself put into my head. I have spoken with him in the most generous terms. Find out from him directly, if it suits you, what came of it. But why do I delay? I told him you had brought to me both what he wished and what he was looking for. That he wants Spain; that he is looking for an allowance. On the allowance I named what Publilius gets, what the flamen Lentulus gives his son. On Spain I urged two things --- first, the same I said to you: that I fear the censure. Was it not enough that I laid down arms on that side? Must I lift them on the opposite side? Next, that he will be eaten up at being outdone in intimacy and in every kind of favour by his cousin. I would rather he made use of my generosity than of his own liberty. But for all that I have given way; for I had gathered the plan was not too displeasing to you. I shall keep turning it over, and I ask you to do the same. The matter is great. To remain is straightforward; the other is doubtful. But we shall see.

\ciceroSection{2} About Balbus I had written in the notes, and so I plan: as soon as he comes back. But if he is too slow, I am off for three days at any rate --- and, what I left out, with Dolabella too.
