% Ad Familiares 4.13 (ad-familiares-04-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. P. FIGVLO.}

\ciceroSection{1} I have been asking myself for a long while now what I might best write to you, and not only does no definite matter come to mind, but not even any familiar kind of letter at all. Of the kinds of letter I used to employ in better days, the time has stripped one whole class from me: fortune has so contrived it that I can neither write anything of that sort nor even conceive it. What is left is a kind of letter mournful and miserable and consonant with the present times. That kind too fails me, since in such a letter there should be either the promise of some assistance or the consolation of your grief. The promise was beyond my power; for I, struck down by a like fortune, was myself sustaining my own reverses on the supports of others; and it more often came to my mind to complain that I went on living so, than to rejoice that I went on living at all.

\ciceroSection{2} For although no signal injury has fallen upon me as a private person, and nothing has even come into my mind at such a time to wish for that Caesar has not unprompted bestowed on me, still I am consumed by these cares, so that the very fact that I remain alive I count for a fault in myself. For I am bereft of many of my closest intimates --- some of whom death snatched from me, others flight has put far away --- and of all those friends whose goodwill the state's defence, with you once my partner in it, had won me through my own doing. I move about amid their shipwrecks and the plundering of their goods; and I do not merely hear of it --- which would itself be wretched --- but I see it, than which nothing is bitterer: the fortunes of those men ruined, with whose help we once put out that conflagration; and in that city in which we lately flourished in influence, in standing, in glory, in that city we now lack every one of those things. We have, indeed, Caesar's own utmost humanity towards us, but it cannot prevail against the violence and the upheaval of all things and times.

\ciceroSection{3} And so, stripped of all those things to which nature and inclination and habit had accustomed me, I am out of favour with the rest, as it seems to me at least, and out of favour above all with myself. Born to be always doing something worthy of a man, I now have no way open to me not only for doing but even for thinking; and I, who in earlier days could come to the aid of unknown men or even of guilty ones, can now not even make a kindly promise to P. Nigidius, the most learned and the most upright man living, once of the very highest influence and a most loyal friend to me. So this kind of letter too has been taken from me.

\ciceroSection{4} What is left is for me to console you and bring forward considerations by which I may try to draw you off from your distresses. But that faculty, of consoling yourself or another, is in you supreme, if it ever was in any man. I shall therefore not touch that side of the matter which proceeds from any refined method or learning --- I leave it all to you. What is worthy of a brave and a wise man, what your gravity, the elevation of your spirit, the life you have led, your studies, the arts in which you have flourished since boyhood demand of you --- this you shall see for yourself. What I, being at Rome and looking attentively about me, can understand and feel, this I affirm to you: that in those distresses in which you are at this time, you will not be much longer, but in those in which we too are, perhaps you will always be.

\ciceroSection{5} I think I can see, first, that the disposition of the man himself, who can do most, is well inclined to your restoration. I do not write this without reason; the less close my acquaintance with him is, the more carefully I am at work to discover it. The slower he has so far been to free you from this trouble, the easier it is for him to give a stern answer to those with whom he is more angry. His intimates, and indeed those that are most welcome to him, speak and feel about you wonderfully well. To this is added the goodwill of the people at large, or rather the consensus of everyone. There is the further consideration of the republic --- which, however little it can do at this moment, is bound to be able to do something: whatever it shall have in the way of strength, it shall in a short time, believe me, extort on your behalf from the very men by whom it is now held.

\ciceroSection{6} I come back, then, to this point: that I now promise you something as well, which at first I had passed over. For I shall both embrace those most intimate with him who are deeply attached to me and spend much time with me, and shall force my way into the company of the man himself --- which my own self-respect has hitherto barred to me --- and shall certainly follow up every avenue by which I judge that we can arrive at what we want. In this whole undertaking I shall do more than I dare to write. As to other things, which I know for certain that many men are ready to provide for you, from me they are most ready of all. There is nothing of mine in our household goods which I would not rather were yours than mine. About this, and about this whole subject, I write more sparingly, because I would rather you hoped --- as I myself trust --- that you will have the use of your own.

\ciceroSection{7} The last point is this: that I beg and entreat you to be of the greatest spirit, and to remember not only those lessons which you have received from the other great men, but also those which you have brought forth yourself by your own talent and study. If you gather these together, you will both hope for the best in everything and bear with wisdom whatever shall happen, of whatever sort it may be. But all this you know better than anyone, and best of all men. For my part, in everything I understand to concern you, I will most zealously and most carefully see to all of it, and I shall preserve, in this saddest time of my own, the memory of your services to me.
