% Ad Atticum 12.6 (ad-atticum-12-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On Caelius --- please look that there be no shortfall in the gold. I am not expert in such matters. But surely there is loss enough on the exchange-rate. If the gold too is short --- but why do I go on? You will see to it. There you have the manner of Hegesias, which Varro commends.

\ciceroSection{2} I come to Tyrannio. Really? Is it so? This happened? Without me? And I --- how many times, even at leisure, refused to do it without you? How then will you atone for it? Plainly, with this one thing: send me the book. I ask you again and again to do so. And yet the book itself will give me no greater pleasure than your admiration has given me. For I love any \textit{[Greek: pantha philideimona]} --- a lover of every learning --- and I am glad you so warmly admired that very fine \textit{[Greek: the\=oria]} --- speculative inquiry --- of his. Though indeed all yours is of this sort: you want to know; and on that alone the mind is fed. But, I ask you, what does that acute and weighty matter bring us toward our \textit{[Greek: telos]} --- our end? But it is a long discourse, and you are busy --- in some business of mine, perhaps. And for that bare sunshine you used up in our little meadow, we shall demand back from you a sleek, oiled sun. But I return to first things. Send the book, if you love me. It is yours, surely, since it was sent to you. \textit{Chremes, hast thou so much leisure from thine own affairs that thou readest even the} Orator\textit{?} Good for you! It is welcome to me, and will be more welcome if not only in your own copies but in those of others too, through your copyists, you put Aristophanes back in place of Eupolis.

\ciceroSection{4} Caesar seemed to me to be mocking, I beg you, that remark of yours, which was at once \textit{[Greek: eupines]} --- polished --- and witty. He then bade you be without care, in such terms as to take away all my own doubt. I grieve that Attica has been ill so long; but since she is now past the chill, I hope it is as we wish.
