% Ad Familiares 6.7 (ad-familiares-06-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAECINA CICERONI P. S.}

\ciceroSection{1} That the book has not been put into your hands more quickly --- forgive my fear, and pity my situation. My son, as I hear, was alarmed, and not without reason: that if the book got out (since it matters less in what spirit a thing is written than in what spirit it is received), the very fact of its having appeared might hurt me where I least deserved it --- and that, when I am to this day still paying the penalty of my pen. In this matter my fate is unique. For where a fault in the writing is taken away by the eraser, where folly is fined by reputation, my own error is corrected by exile --- the sum of my crime being that, with arms in my hand, I spoke ill of an adversary.

\ciceroSection{2} There is not one of us, I suppose, who did not vow vows for his side's victory, not one who, even when he was sacrificing for some other reason, did not even at that moment pray that Caesar should be defeated as soon as possible. If this never crosses his mind, he is fortunate in everything; but if he knows it and is persuaded of it, what cause has he to be angry with one man who wrote something against his wishes, when he has pardoned everyone who, in many a prayer, importuned the gods against his safety?

\ciceroSection{3} But, to come back to my point: this was the cause of my fear. What I wrote of you, by my honest faith I wrote sparely and timidly, holding myself back --- almost shying off. And yet who does not know that this kind of writing ought to be not merely unfettered but full-speed and high-flying? To speak ill of another is judged to be free (though one must take care not to slip into mere insolence); to praise oneself is hampered, lest the vice of arrogance follow it; only this is truly free --- to praise another, since whatever you take from that praise must be assigned either to weakness in you or to envy. And I cannot say but that it has fallen out the more kindly and the more conveniently for you: for since I could not do brilliantly, the first thing was not to touch the subject, the second to confer the benefit as sparingly as possible. And as it is, I held myself in; I cut much down, I struck much out, I left a good deal in altogether unwritten. Just so --- if of a flight of stairs you take some steps away, cut others through, and leave some others ill-fastened, you build the danger of collapse, not the means of ascent: in just that way, with the bent for writing bound here and broken there by so many troubles, what worth hearing, what plausible thing, can it bring forward?

\ciceroSection{4} But when I came to Caesar's own name, I trembled in every limb --- not from fear of the punishment, but from fear of his judgement. For I do not know Caesar in his entirety. What sort of mind, do you think, is mine, while I talk it over with myself? --- ``This he will accept; that word is dangerous. What if I change this? But I am afraid it will be worse. Well, suppose I praise someone: do I give no offence? And when, on the other hand, I reproach someone --- what if Caesar would not have me?'' --- He pursues the pen of an armed enemy; what will he do to a defeated man not yet restored? You too increase my fear, you who in your \textit{Orator} take cover behind Brutus and look for a partner in your excuse. When the patron of all men does this, what ought I --- your old client, and now everyone's --- to feel? In this lawsuit of fear, then, and under the rack of blind suspicion, when very many things must be written by guesswork at another man's feeling and not by one's own judgement, how hard it is to come through, I know --- if you have less of the experience yourself, that is because, for everything, your supreme and outstanding talent has armed you. But I had told my son to read the book to you and take it away with him, or else to give it on this condition: that, if you took it, you would correct it --- that is, would make it altogether a different book.

\ciceroSection{5} As for the journey to Asia: though the strictest necessity was pressing, I did as you ordered. What can I urge you, on my own behalf? You see the time has come when a decision must be made about me. There is nothing, my dear Cicero, for which you should wait on my son. He is a young man; he cannot, whether by application or by age or by fear, think everything through. It is on you that the whole business must rest; in you is all my hope. You, in your wisdom, know in what things Caesar takes pleasure, by what he is taken; from you all the moves must come, by you all must be brought to their issue --- you have much weight with the man himself, and the most with all his people.

\ciceroSection{6} If you can persuade yourself of one thing --- that this is not a matter of your doing what you are asked to do (though even that would be great and ample) --- but that the whole load is yours: then you will see it through; unless perhaps in my wretchedness I am laying the load on you too foolishly, or in our friendship too impudently. But the habit of your life is an excuse for either charge. For since you have made it your habit to labour for your friends, your familiars no longer merely hope for it from you, they require it of you. As for the book your son will hand you, I ask of you that it not go out --- or that you correct it so that it does me no harm.
