% Ad Familiares 4.4 (ad-familiares-04-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. SER. SVLPICIO.}

\ciceroSection{1} I accept the excuse you have made for sending me a number of letters in the same wording: but I accept it only in the part where you say that it is owing to the carelessness or the bad faith of those who receive the letters that they do not reach us. The other part of the excuse --- where you write that, from your ``poverty of style,'' as you call it, you have sent the same letters several times in the same words --- I neither recognize nor approve; and I myself, whom you say in jest (so I take it) to have ``riches of style,'' do recognize that I am not so very destitute of words, for there is no need to play the ironist \textit{[Greek: eironeuesthai]}; but I yield, all the same (and not by ironizing \textit{[Greek: eironeuomenos]} either), to the subtlety and elegance of your writing.

\ciceroSection{2} As for the plan by which, as you write, you have not refused this commission in Achaia: I had always approved it, but I approved it far more when I had read your last letter. For all the reasons you mention are most just, and most worthy both of your authority and of your prudence. That you reckon the outcome to have fallen out differently from what you had supposed --- on that I in no way agree with you; but because there is so great a confusion and overthrow of affairs, and everything lies so beaten down and laid low by this most foul war, that each man thinks the place where he himself is the most wretched, and himself the most wretched of all, you are dissatisfied with your own plan, and we who are at home seem to you the fortunate ones; and to us in turn you seem --- not without your share of troubles, certainly, but, as compared with us, fortunate. And in this very point your case is better than ours: that you dare to write what pains you, whereas we cannot even do that with safety; and not by any fault of the victor, who is the soul of moderation, but by the very nature of victory, which in civil wars is always overbearing.

\ciceroSection{3} In one thing we have outdone you: we learned of the safety of Marcellus, your colleague, a little before you did, and, by Hercules, in the way the thing was managed we saw it as well. For this is how you must take it: ever since these miseries began --- that is, ever since the question of public right began to be decided by arms --- nothing has been done with so much dignity. For Caesar himself, after censuring the ``bitterness'' of Marcellus (so he called it) and praising in the most honorific terms both your fairness and your prudence, suddenly and beyond hope said that, when the Senate was making the request on Marcellus's behalf, he would not refuse it, even for the sake of the omen. The Senate had managed it thus: when L. Piso had made mention of M. Marcellus, and C. Marcellus had thrown himself at Caesar's feet, the whole body rose and went up in supplication to Caesar. Do not ask: that day seemed to me so beautiful that I felt I was seeing some image, as it were, of a state coming back to life.

\ciceroSection{4} And so, when all who were called on before me had given thanks to Caesar --- except Volcacius, who said that, had he been in Marcellus's place, he would not have done so --- when I was called on, I changed my plan. For I had resolved, by Hercules, not from sluggishness but from longing for our former standing, to keep silent forever. That resolve was broken in me both by Caesar's greatness of mind and by the Senate's discharge of its duty; and so at some length I delivered my thanks to Caesar, and I am afraid that in other matters as well I have stripped myself of an honorable leisure, which was my one consolation in misery. Still, since I have escaped the displeasure of one who might perhaps suppose that I did not regard this as a commonwealth at all, if I were to keep silent forever, I shall do this in moderation, or even within moderation, so as to serve both his wish and my own pursuits. For though from my earliest youth every art and liberal learning, and most of all philosophy, has been my delight, this study grows weightier in me daily --- both, I think, with the ripening of age into prudence, and with the vices of the times, since nothing else can ease the mind of its troubles.

\ciceroSection{5} I gather from your letter that you are being drawn off from this study by your duties; but at any rate the nights will help you somewhat by now. Your Servius, or rather ours, attends on me with the deepest regard; and I take pleasure in him both for his complete uprightness and highest virtue and for his learning and his studies. He often talks over with me the question of your staying on or coming home. So far I am of this opinion: that we should do nothing except what Caesar will plainly most wish. The situation is such that, if you were at Rome, nothing could give you pleasure except your own family. As for the rest, nothing is better than Caesar himself; everything else is such that, if you must either hear of it or see it, you would rather hear. This counsel of ours is by no means a pleasing one to ourselves, who long to see you; but we are advising in your interest. Farewell.
