% Ad Atticum 12.4 (ad-atticum-12-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} How welcome your letter was to me, how delightful! What more shall I say? A festal day has been restored to me. For I was being squeezed by anxiety because Tiro had told me you had seemed to him \textit{[Greek: enereuthesteron]} (rather flushed). I shall add, then, as you suggest, one more day.

\ciceroSection{2} But on Cato the case is \textit{[Greek: probl\=ema Archim\=edeion]} (an Archimedean problem). I cannot get myself to write what your dinner-companions could read --- I will not say with pleasure, but even with composure. Indeed, even if I were to back away from his opinions delivered, away from every wish and counsel he had on public affairs, and to praise only \textit{[Greek: psil\=os]} (bare-stripped, just so) his gravity and his constancy, this very piece would itself be an \textit{[Greek: akousma]} (a thing to be heard) loathsome to that set. But the man cannot be truly praised unless these things are made part of his glory: that he foresaw what now is and what will yet be, that he fought to keep it from coming about, and that he laid down his life so as not to see it once accomplished. Which of these can we make acceptable to Aledius? But take care of yourself, I beg you, and apply to your recovery, above all things, the same good sense you apply to everything else.
