% Ad Familiares 4.7 (ad-familiares-04-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. M. MARCELLO.}

\ciceroSection{1} Though I see that the course you have followed thus far is one I do not presume to fault --- not that I would not myself dissent from it, but because I judge you to be of such wisdom that I will not set my judgement above your own --- still, the long-standing of our friendship, and that warm goodwill you have shown me from your boyhood onward, have urged me to write to you what I think will conduce to your welfare and what I judge not foreign to your standing.

\ciceroSection{2} I remember vividly that you were the man who foresaw the beginnings of these present miseries long in advance, and who held the consulship with the greatest distinction and the greatest worth. But I saw this much, too: that you did not approve the plan by which the civil war was being conducted, nor the forces of Gnaeus Pompeius, nor the character of his army, and that you mistrusted the whole affair to the utmost. I think you remember that I held that same view. And so you took little part in the campaign, and I made it my business throughout to take no part at all. For we were not fighting with the weapons in which we had the advantage --- counsel, authority, the justice of our cause, in which we were the stronger --- but with sinew and force, in which we were no match. We were beaten, then; or, if standing cannot be beaten, we were broken at any rate, and laid low. In all this no one can fail to praise, in the highest terms, the course you took: that you cast off, together with the hope of victory, even the appetite for further struggle, and showed that the wise and good citizen takes up the opening of a civil war unwillingly, and does not gladly pursue its end.

\ciceroSection{3} Those who did not follow the course you did, I see have been pulled apart into two camps. Either they have tried to revive the war --- these are the men who have taken themselves to Africa --- or, like me, they have entrusted themselves to the victor. Yours has been a middle course, which you may perhaps think the others lowness of spirit on the one hand, and stubbornness on the other. I admit that by most men --- I would even say by all --- your course has been judged wise; by many, indeed, the act of a great and brave spirit. But this reasoning, as it seems to me at least, has a limit; the more so since I believe nothing is wanting to you for the recovery of your whole estate but the will to have it back. For I have understood this clearly: nothing else makes him hesitate --- the man in whose hands the power lies --- except his fear that you would not count it as a benefit at all. What I think on that point there is no need to say, since my own actions make it plain.

\ciceroSection{4} But even so: even had you so settled the matter that you would rather be away forever than see the things you would not wish to see, still you ought to consider this --- that wherever you are, you will be in the power of the very man you flee. If he were going to let you live, deprived of country and fortune, in quiet and freedom, you ought still to weigh whether you would rather live at Rome and in your own home, whatever its state, or at Mytilene or at Rhodes. But since the power of the man we fear extends so widely as to embrace the whole globe, would you not rather be without danger at your own house than with danger at another's? For my part, even if death had to be met, I would meet it at home and in my own country rather than in places foreign and not my own. All who love you feel the same; and the number of those, in proportion to your great and shining virtues, is great.

\ciceroSection{5} We have in mind, too, your household estate, which we do not wish to see scattered. For though it can take no injury that will be permanent --- since neither he who holds the state will allow it, nor will the state itself --- still I do not wish a raid by plunderers to be made upon your fortunes. Who these would be I would venture to write, did I not trust you to understand.

\ciceroSection{6} Here the anxieties of one man, the many and constant tears too of one man, plead for you: of Gaius Marcellus, your splendid brother. In care and in pain we are next to him; in pleading the slower, because we have no right of approach, having needed entreaty on our own behalf; in influence we can do what defeated men can do; but in counsel, in zeal, we do not fail Marcellus. By the rest of your kin we are not called on; for everything we are ready.
