% Ad Familiares 6.13 (ad-familiares-06-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO LIGARIO.}

\ciceroSection{1} Although in this present hour of yours I ought, for the sake of our friendship, to have written you something --- to console you, or to do what I could for you --- yet up to now I have not, because in speech I felt I could neither soften your pain nor lift it. But once I came to hold a strong hope that within a short time we should have you back, safe and whole, I could not but make known to you both my judgement and my will.

\ciceroSection{2} First, then, I shall write what I understand and clearly see: Caesar will not be too hard on you. For events daily, and time, and the talk of men, and, as it seems to me, his own nature, are making him more lenient; and this I feel concerning the rest, and concerning you in particular I hear it too from those closest to him. With these men I have not let off pleading, in company with your brothers, from the time when the first despatch came from Africa; and the singular love of those brothers for you --- their worth and their duty, their constant and unfailing care for your safety --- counts for so much that I believe there is nothing Caesar himself will not grant.

\ciceroSection{3} But if this comes more slowly than we want --- well, the man from whom everything is asked has great occupations, and approach to him has been harder; and at the same time, where the African cause is concerned, in his anger he seems to want to keep men anxious for some time longer, the men by whom he reckons that he has been harassed with the more prolonged annoyances. But on this very point I see that he is daily more relaxed and more appeased. Therefore trust me, and consign it to memory, that I have given you my warranty: you will not be in these troubles long.

\ciceroSection{4} Now that I have set out what I think, what I want for your sake I shall make clear by the deed rather than by speech. If I had as much power as I should have had, given my service to a commonwealth that has so deserved of me (as you yourself reckon), you for your part would not be in this present distress; for the same cause that broke my own resources brought your safety to its crisis. But still, whatever the shadow of my old standing, whatever the remnants of my favour, will avail, my zeal, my counsel, my work, my favour, my good faith will not be lacking anywhere to those excellent brothers of yours.

\ciceroSection{5} See to it that you have the brave spirit you have always had, first for the reasons I have written, and then because in matters of the commonwealth you have always wished and judged in such a way that not only ought you now to hope for prosperity, but even, if everything were against you, you ought, by the consciousness of what you have done and meant, whatever the outcome, to bear it with the bravest and the greatest spirit.
