% Ad Familiares 6.8 (ad-familiares-06-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CAECINAE.}

\ciceroSection{1} When Largus --- a man devoted to your interest --- spoke with me and said the kalends of January had been set as your deadline, since I had perceived in all matters that what Balbus and Oppius transacted in Caesar's absence is usually held by him to be ratified, I urged them strongly to grant me this: that it be open to you to be in Sicily as long as we wished. The men who were used to giving me, with regard to that sort of request, either a gracious promise (if it was something that would not offend their masters' tempers) or a refusal with their reasons for refusing --- to this present request of mine they did not, exceptionally, answer on the spot. Returning to me the same day, however, they granted me this: that you should be in Sicily as long as you wished, and that they would guarantee you should incur no displeasure from it. Now that you know what is permitted you, I think you ought to know what I think best.

\ciceroSection{2} While these matters were going forward, a letter was delivered to me from you, asking my counsel: which would I advise --- that you sit tight in Sicily, or set out to what remains of your business affairs in Asia? This deliberation of yours did not seem to me to agree with what Largus had said. He spoke with me as if you were not at liberty to linger longer in Sicily; but you are deliberating as if it had been granted. Yet, whichever of the two it is, I hold you should linger in Sicily. The nearness of the country helps both with the getting of the favour --- by frequent letters and messengers --- and with the speed of your return, whether by the favour granted (which I hope) or by the business being settled in some other fashion. For which reason I most strongly hold you should linger.

\ciceroSection{3} To Titus Furfanius Postumus, a familiar friend of mine, and to his legates --- likewise my familiar friends --- I shall most carefully commend you, when they have arrived: for they were all at Mutina. They are the best of men, devoted to friends of your sort, and intimates of mine. What occurs to me as bearing on your business, that I shall do unprompted; in matters I am ignorant of, once told of them, I will outdo every other man's zeal. Although when I come face to face with Furfanius I will speak about you in such a way that no letter from me to him need be necessary, still, because your people thought best that you should have a letter of mine to deliver to him, I have humoured them. A copy of that letter is written out below.
