% Ad Familiares 9.2 (ad-familiares-09-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO VARRONI}

\ciceroSection{1} Your Caninius --- and ours --- came to me very late in the evening and said that he was setting out for you early the next morning, so I told him I would give him a letter to take; I asked him to call in the morning. I wrote the letter overnight; he did not come back to me; I supposed he had forgotten. And yet I should have sent that very letter through my own people, had I not heard from the same source that you were setting out from Tusculum early next morning. Then suddenly, a few days later, when I was least expecting it, Caninius came to me in the morning: he said he was leaving for you at once. The letter was by now stale --- \textit{[Greek: heōlos]} --- the more so since such great new events had since been reported; still, I did not want my night's work to go to waste, and I handed that very letter to Caninius. With him, as with a learned man and a great friend of yours, I spoke of the things which I expect he has reported to you.

\ciceroSection{2} For yourself I have the same advice that I give myself: let us avoid men's eyes if we cannot so easily avoid their tongues. For those who pride themselves on the victory look at us as if we were the conquered, and those who take our defeat hard are sorry that we are alive. You will perhaps ask why, with things in this state at the city, I am not absent as you are. You yourself, who outdo me and the rest in foresight, saw, I imagine, all of it; nothing whatever escaped you. But who is so lynx-eyed that in such darkness he does not stumble somewhere, run into something somewhere?

\ciceroSection{3} For my own part the thought has long since occurred to me that it would be a good thing to go off somewhere, so as not to see what was being done here and not to hear what was being said. But I held myself back. I told myself that whoever met me would suppose --- or, suppose or not, would say --- exactly what suited him: ``This man is afraid and is running for that reason,'' or ``He has some plan in hand and has a ship ready.'' In the end, even the man who suspected most lightly --- the one who knew me best, perhaps --- would put it down to my going off because there were certain people my eyes could not bear to see. While I was turning this over, I have stayed on at Rome; and yet, imperceptibly, daily habit has by now grown a callus over my stomach \textit{[Greek: lelēthotōs]}. There you have the reasoning behind my plan.

\ciceroSection{4} For you, then, my advice is this: stay hidden where you are for the time being, while this triumphal congratulation is at its boil, and at the same time while we wait to hear how the business has been settled --- for settled I take it to be. Much will depend on what spirit the victor brings to it, and on how the thing comes out; I have my own guess where my conjecture leads, but I am waiting all the same.

\ciceroSection{5} As for you, I should not have you go to Baiae until the noise has gone hoarse. It will be more becoming for us, even when we have left this place, to be thought to have gone into those regions for weeping rather than for swimming. But you will judge that better. Only let this stand between us: that we live together with our studies, from which we used once to seek only enjoyment, but now even our safety; that we not fail anyone who is willing to take us into the work, not merely as architects but even as common builders for the rebuilding of the commonwealth, but rather run to it gladly; and that, if no one takes us up, we still both write and read of constitutions --- \textit{[Greek: politeias]} --- and, if not in the Senate-house and the Forum, at least in our writing and our books, as the most learned of the ancients did, attend to the commonwealth and inquire into customs and laws. So it seems to me; what you intend to do, and what you yourself approve, I should be very grateful to learn if you will write to me.
