% Ad Familiares 9.3 (ad-familiares-09-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO VARRONI}

\ciceroSection{1} Although I had nothing to write, still I could not let Caninius go off to you with nothing in hand. So what shall I write, most of all? What I take it you want to hear: that I shall be coming to you soon. And yet --- consider, please, whether it is altogether proper that we, in the midst of such a conflagration of the state, should be in those places of yours. We shall give people something to talk about, those who do not know that wherever we are our way of life is the same, our way of eating the same. ``And what does it matter? We shall be talked about anyway.'' Truly, I imagine, we must labour mightily to make sure that, while everyone is wallowing in every kind of crime and outrage, our resting together, or by ourselves, not be censured. For my part, I shall pursue my course and pay no heed to the ignorance of barbarians; for though our condition is wretched --- as wretched as can be --- still our pursuits, somehow, seem now to bear richer fruit than they once bore. Perhaps it is because we find rest in nothing else now; perhaps because the gravity of the disease is what makes us in need of the medicine, and now it shows its power, whose force we did not feel while we were well. But why am I telling you these things now, in whose own house they are home-born --- \textit{[Greek: glauk' eis Athēnas]}? For no reason, surely, except to get a reply from you, to have you waiting for me. Then so you shall do.
