% Ad Familiares 12.18 (ad-familiares-12-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO S. D. CORNIFICIO CONLEGAE.}

\ciceroSection{1} I shall answer first what came at the end of the letter I most recently received from you: I have noticed that this is a trick you great orators use from time to time. You ask for my letters. But whenever I have been notified by your people that someone of theirs was going, I have never failed to send one. As for what I think I gather from your letter --- that you mean to commit yourself to nothing rashly, and to settle nothing for certain until you have learned where this whatever-he-is, Caecilius Bassus, will break out --- I had hoped as much, trusting your prudence; and your most welcome letter has only confirmed my confidence. Do this as often as you possibly can, I earnestly beg, so that I may know both what you are doing and what is being done --- and what too you mean to do. Though I bore your departure from me very ill, still at the time I took this consolation: that, as I thought, you were going off to the deepest quiet, and stepping clear of great affairs that hung over us here.

\ciceroSection{2} Both have turned out the other way. There war has broken out; here peace has followed --- but a peace of the sort in which, were you present, much would not please you, and indeed much that does not please even Caesar himself. For civil wars always end thus: that it is not only the victor's will that is done, but that those by whose support the victory was gained must also be humoured. I myself have grown so callous by now that at our friend Caesar's games I watched T. Plancus, and listened to the verses of Laberius and Publilius, with the most equable mind in the world. Know that I miss nothing so much as someone with whom I can laugh at these things in the right intimate and learned way. You will be that person, if you come as soon as may be; and I think it is in your interest as well as mine that you should.
