% Ad Familiares 13.19 (ad-familiares-13-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO SERVIO S.}

\ciceroSection{1} With Lyso of Patrae I have indeed an old guest-friendship, a bond which I think must be cherished scrupulously; but the same is true for me with a good many others, whereas with no other guest-friend is the closeness so great. It has been so strengthened by his many good services and by daily intimacy besides, that nothing is more tightly knit than the friendship between us. While he was at Rome for nearly a year, living, as it were, under one roof with me, we were in great hopes that you, on the strength of my letter and recommendation, would do most carefully what you have in fact done --- watch over his affairs and his property in his absence. Yet because the whole disposition of things lay in one man's power, and because Lyso had been in our camp and under our protection, there was something we feared every day. As things turned out, by his own distinguished standing and by my zeal, and that of his other guest-friends, we obtained from Caesar everything we wanted; you will see this from the letter Caesar has written to you.

\ciceroSection{2} As for myself, so far from now abating any part of my recommendation, as though we had already gained the whole, I press it on you the more strenuously: I ask you to take Lyso into your good faith and your circle of intimates. While his fortune was uncertain we approached you more tentatively about him, fearing that some such turn of events might come about as not even you could remedy; but now that his safety is assured, I ask everything from you with the highest zeal and care. To avoid enumerating the particulars: I commend to you his whole household, and among them his young son, whom my client C. Maenius Gemellus, when in the calamity of his exile he had been made a citizen of Patrae, adopted under the laws of that city --- that you may guard the very right and the case of the inheritance that this brings him.

\ciceroSection{3} The chief point is this: that Lyso, whom I have come to know as a man of the highest worth and of the warmest gratitude, you receive into your circle of intimates. If you do so, I have no doubt that, in cherishing him --- and in commending others to me hereafter --- you will be of the same mind and the same disposition that I am. The thing I am eager should be done; and yet I have this further fear, that if you should appear to have done anything on his behalf less than fully, he will think I have written without diligence, not that you have been forgetful of me. How much you make of me, he has had means enough to learn, both from my own daily conversation and indeed from your own letters.
