% Ad Atticum 12.3 (ad-atticum-12-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} You are the one man, I think, who flatters less than I do; and if either of us is ever so to someone else, between ourselves we never are. Listen then to what I say without enchantment, \textit{[Greek: ago\=eteut\=os]} (with no spell laid on it). May I not live, my dear Atticus, if my Tusculan villa --- where, in everything else, I am content to be --- if even the \textit{[Greek: makar\=on n\=esoi]} (Isles of the Blessed) are worth so much to me as to be without you for so many days. So I shall harden myself through this three-day stretch, so as to put you, too, into the same \textit{[Greek: pathos]} (state of feeling); and so in fact it is. But I should like to know whether you are coming today, straight off, about the auction, and on what day you come. Meanwhile, I am with my little books; and I am annoyed not to have Vennonius's history. But, not to say nothing about the business: that debt of mine from Caesar has three possible settlements --- either purchase at the auctioneer's spear (I should rather lose it outright, though that, quite apart from its disgrace, is itself what I count as losing it), or assignment from the contractor on a one-year term (and to whom shall I extend credit, or when will that Metonic year of his come round?), or, on Vettienus's terms, the half. \textit{[Greek: skepsai]} (think it over), then. And I am afraid the fellow won't hold any auction at all now, but, the games over, will go running off to the rescue of \textit{[Greek: Atypos]} (the dimwit, our Sir Witless), so that such a man may not be left \textit{[Greek: alog\=eth\=ei]} (without his reckoning). But \textit{[Greek: mel\=esei]} (he'll see to it). You look after Attica, please, and give her my warmest greetings, and Pilia and Tullia's too.
