% Ad Familiares 12.19 (ad-familiares-12-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CORNIFICIO S.}

\ciceroSection{1} I read your letter with the greatest pleasure --- and what pleased me most in it was learning that mine had been delivered to you; for I did not doubt you would read them gladly, but I had my fears as to whether they would be delivered. From your letter I have learned that there is a war in Syria, and that Caesar has assigned the province of Syria to you. I wish the business well and prosperously upon you --- and that it will so turn out I am confident, leaning on your industry and your prudence.

\ciceroSection{2} But what you write of the suspicion of a Parthian war has, I must say, given me real concern. What forces you have I could in part work out for myself by guesswork, and now I have learned from your letter. So I pray that nation may make no move at this present time, while those legions which I hear are being led off are being brought through to you. If you do not have matched forces for a pitched engagement, you will not let slip the strategy of M. Bibulus, who held himself within his most strongly fortified and best-supplied town for as long as the Parthians were in the province. But these things you will settle better as the case and the moment dictate; for my part, I shall be in unbroken anxiety over what you are doing until I know what you have done. I have never had a man to entrust a letter to without entrusting one; I ask you to do the same on your end --- and above all, in writing to your own people to write so that they may know I am yours.
