% Ad Familiares 6.6 (ad-familiares-06-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. A. CAECINAE.}

\ciceroSection{1} I fear you may feel I have failed you in my office of friendship --- an office which, given the bond between us, the many services you have done me, and our shared pursuits, I have no right to leave unperformed; and in particular I fear you may have missed the office of a letter from me. I would have sent you one both long since and often, had I not preferred, day by day expecting better news, to embrace in a letter your congratulation rather than the strengthening of your spirit.

\ciceroSection{2} Now, as I hope, we shall be congratulating before long; so that subject for a letter I am putting off to another time. With this present letter I think your spirit --- which I both hear and trust is anything but weak --- ought nonetheless to be strengthened again and again by the authority of a man who is not the wisest, perhaps, but is the most attached of friends; and that not in such terms as one would use to console a man cast down and stripped of every hope of safety, but in such terms as one would use toward someone whose preservation I doubt no more than I remember your doubting of mine. For when I had been driven from the commonwealth by those who supposed it could not fall while I still stood, I remember hearing, from many a guest who came to me from Asia, where you were, that you firmly predicted my glorious and swift return.

\ciceroSection{3} If that wonderful science of the Etruscan discipline, which you had from your father, a man of the highest nobility and excellence, did not fail you in my case, our own divination will not fail us in yours --- a divination we have arrived at not only through the monuments and precepts of the wisest men, and through that great devotion to learning of which you know, but also through the long practice of managing the commonwealth and through the great variety of our times;

\ciceroSection{4} and in this divination I have the more confidence because in matters as dark and tangled as these it has never deceived us at any point. I would set out what I predicted before the event, did I not fear it should look like I was inventing them after the fact. But there are many witnesses all the same to my warning Pompey first not to join himself with Caesar, and afterwards not to break with him: by their union the strength of the Senate would be broken, by their division civil war would be kindled --- this I saw. And I was on the closest terms with Caesar, and I valued Pompey at the highest; but my counsel was at once loyal to Pompey and salutary to them both.

\ciceroSection{5} What else I foresaw I pass over; I do not want this man, who has deserved the best of me, to think I gave Pompey counsel which, had he obeyed, would have left this one no doubt distinguished in civil life and first man in Rome, but not in possession of the immense power he has now. I held that he ought to go into Spain. Had he done so, there would have been no civil war at all. That his candidacy be heard while he was absent --- I did not so much fight for the legal right of it as for the holding to it, since the people, by act of the consul on the floor, had so ordered. The cause of the war arose. What did I leave unsaid, of warning or of complaint, when I was for setting even the most unjust peace ahead of the most just war?

\ciceroSection{6} My authority was defeated, not so much by Pompey --- for he was moved by it --- as by those who, trusting in Pompey as their leader, supposed the victory in that war would fall conveniently on both their household business and their private appetites. The war was taken up while I held back; it was driven out of Italy while I held my ground, as long as I could; but in the end my shame had more weight with me than my fear. I was ashamed to fail Pompey's safety when he had not, in his day, failed mine. And so, vanquished by duty, or by the reputation of the good men, or by shame, I set out, as Amphiaraus in the plays sets out, ``knowing and aware, to the destruction set before my eyes.'' In which war nothing went badly that I had not predicted.

\ciceroSection{7} Therefore, since (as augurs and astrologers are wont) I too, a public augur, have on the strength of my earlier predictions set up before you the authority of my own augury and divination, our present prediction ought to be believed. I divine for you, then, not from the flight of a bird, nor from the sinister note of the singing-bird (as is our discipline's way), nor from the most full-footed beat or sound of the sacred chickens, but I have other signs to observe; and though they are not more certain than those, they have less of obscurity or of error in them.

\ciceroSection{8} For my divining I note signs along two paths: the one I draw from Caesar himself, the other from the nature and tendency of our civil times. In Caesar these are the signs: a mild and merciful nature, of the kind your famous book of \textit{Querelae} expresses. Add to this that he is wonderfully delighted by men of outstanding talent, of which yours is one. Add too that he yields to the just wishes of many men, born of duty, not to empty or self-seeking ones;

\ciceroSection{9} and in this Etruria, united, will move him strongly. ``Why then have these things availed so little up to now?'' --- Because he does not think he can sustain the cases of many men, if he yields to you, against whom he holds he has the juster claim to be angry. ``Then what hope is there from an angry man?'' you will say. He will come to understand that he is drawing his praises from the same fountain by which he has been lightly splashed. And in any case the man is exceedingly shrewd and has long sight: he understands that you, the man easily the noblest of all in a part of Italy not at all to be slighted, the equal in talent or favour or fame with the Roman people of any of the foremost men of your generation in the commonwealth at large, cannot be kept out of public life any longer. He will not want this hereafter to be a gift of the moment rather than now to be his own.

\ciceroSection{10} So much for Caesar; now I shall speak of the nature of the times and the state of things. There is no man so hostile to the cause which Pompey took up --- with better heart than preparation --- who would dare call us bad citizens or wicked men. On which point I tend to wonder at Caesar's seriousness, justice, and wisdom: he never speaks of Pompey but in the most honourable terms. ``But in the person of Pompey he did many harsher acts.'' Those are the acts of arms and of victory, not of Caesar; and as for us, see how he has embraced us: Cassius he has made his legate, Brutus he has set over Gaul, Sulpicius over Greece, Marcellus --- the man with whom he was angriest --- he has restored with the fullest honour to his standing.

\ciceroSection{11} Where then is all this tending? The nature of things and of the civil times will not endure --- nor will reason allow it, whether things stand or change --- first, that in an equal cause the condition and fortune of all should not be the same; and second, that good men and good citizens, marked with no disgrace, should not return into that city into which so many men condemned of unspeakable crimes have already returned.

\ciceroSection{12} There you have my augury; and if I had any doubt about it, I would have recourse rather to that consolation by which I could easily uphold so brave a man --- namely, that if you had taken up arms for the commonwealth with victory assured (for so you thought at the time), there would not be too much to praise in it; whereas if you had supposed, given the uncertain issues and outcomes of wars, that we might be defeated, then, having been so well prepared for the favourable fortune, you have no business at all of being unable to bear the adverse one. I would argue, too, what great consolation your own conscience of what you did ought to be to you, what great delight your literary work in adversity; and I would call to mind, of the old days and even of these recent ones, the gravest reverses of leaders and of your own companions, and would name many distinguished foreign men besides; for it lightens grief, the recollection of what is, so to speak, the common law and the human condition;

\ciceroSection{13} I would set out, too, in what kind of world we are living here, in what a press and what a confusion of everything; for one must necessarily miss less, with the commonwealth ruined, what one no longer has than if it stood sound. But none of this is needed. Safe and whole I shall see you back, soon, as I hope --- rather, as I clearly see. Meanwhile, to you in your absence, and to this one who is present, the image of your spirit and your body, your son, the most steadfast and excellent young man, I have long since both promised and rendered my zeal, my office, my effort, my labour --- and now this in addition, that Caesar embraces me daily more warmly than the warmest of friends, and his familiars as no one else. With him whatever weight I shall have, by my authority or by his favour, I shall have on your behalf. As for you --- see to it that you sustain yourself by firmness of spirit and by the best of hopes also.
