% Ad Familiares 9.15 (ad-familiares-09-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO PAETO S.}

\ciceroSection{1} I shall answer two letters of yours, the one I had received from Zethus four days ago, the other that Phileros the courier had brought. From your earlier letter I gathered that my concern for your health was very welcome to you, and I am glad you have it in plain view --- though, believe me, you could not see from a letter as it really is. For although I see myself courted and held dear by a fair number of people (I cannot put it otherwise), there is no one of them all whom I find sweeter than you. The fact that you love me, and that you have done so for a long time now and steadily, is itself a great thing --- perhaps the greatest --- but it is a thing you have in common with many; the fact that you yourself are so lovable, so sweet, so delightful in every way --- that is your own.

\ciceroSection{2} Add to that a wit that is not Attic but saltier than the Attics' own --- the old, in-town, Roman wit. As for me (think what you like), I am extraordinarily taken by humour of a homegrown kind, all the more now that I see it has been smeared off, first in Latium when our city was flooded with foreigners, and now also by the trousered, Transalpine tribes, so that no trace of the old charm is left to be seen. And so, when I look at you, I think I am seeing all the Granii, all the Lucilii --- and yes, to tell the truth, the Crassi too, and the Laelii. I shall die if, apart from you, I have anyone left in whom I can recognize the image of the old native gaiety. When such love for me is added to charms like these, do you wonder that I was so heavily knocked out of breath by so serious a turn in your health?

\ciceroSection{3} As for what you say in your second letter, by way of clearing yourself --- that you were not the man who talked me out of the Neapolitan purchase, but the man who urged moderation --- nicely put, and I did not take it otherwise; yet I gathered the same thing then that I gather from this letter, namely that you did not think it open to me, as I had judged it was, to put aside these affairs --- not entirely, but in good part. You bring up Catulus to me, and those times. Where is the likeness? Even I myself, in those days, did not like being too long away from the guard of the commonwealth; for then we sat on the stern and held the helm. Now there is barely room in the bilge.

\ciceroSection{4} Or do you think there will be any fewer decrees of the Senate if I am at Naples? When I am at Rome and dogging the Forum, the senatorial decrees are written up at the house of your admirer, my intimate; and indeed when it crosses his mind, I am put down as one of the drafters, and I hear that a decree has reached Armenia and Syria --- said to have been carried in accordance with my opinion --- before I hear that any mention at all has been made of the matter. And I would not have you suppose I am joking about this; for I would have you know that I have already had letters brought to me from kings on the world's edge, thanking me for having addressed them as kings in my opinion --- kings whom I did not know had been so addressed, indeed whom I did not know had been born at all.

\ciceroSection{5} What of it, then? Still, as long as our prefect of morals is here, I shall bow to your authority; but when he is away, I shall betake myself to your mushrooms. If I have a house of my own, I shall reduce the sumptuary-law allowance to one day in ten; but if I cannot find one that pleases me, I have decided to put up at yours --- for I know I could do you no greater favour. The Sullan house I had been giving up on, as I wrote you most recently, but I have not thrown it over yet. For your part, please, as you write, look it over with builders; for if there is no fault in the walls or the roof, the rest I will be content with.
