% Ad Familiares 6.10 (ad-familiares-06-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO TREBIANO S.}

\ciceroSection{1} How much I think of you, and have always thought of you, and how clearly I have understood you to think of me --- of these things I am my own witness. Your decision, or rather the chance that kept you in arms in the civil war so long, has always been a great grief to me; and this outcome --- that you are recovering your standing and your fortune more slowly than is fair, and than I would have wished --- is no less a care to me than my own reverses have always been to you. And so to Postumulenus and to Sestius, and most often to our Atticus, and most recently to Theudas your freedman, I have opened myself out entirely; and to them, one by one, I have said again and again that I desire to do justice by you and by your children with every resource at my disposal. I wish you would write the same to your people; let them at least be sure of those things which are in my power --- my effort, my counsel, my means, my good faith --- and trust them ready for them in every matter.

\ciceroSection{2} If I had as much authority and favour as I should have, given my service to a commonwealth that has so deserved of me, you too would be the man you were --- not only the worthiest of any high station, but unquestionably the first man of your order. But since, at one and the same time and on one and the same cause, both of us fell, I promise you what I have set down above, which is still mine to give, and along with that whatever I seem to retain in some measure as the remnants of my former standing. For Caesar himself, as I have been able to gather from many indications, is not estranged from us; and almost all his closest familiars, bound to me by my great old services to them when they were in trouble, take careful notice of me and pay me their respect. And so, if ever I have an opening to act on your fortunes --- that is, on the safety in which everything for you rests --- and to this their conversation more and more each day brings me to hope, I shall act for myself, and set the thing in motion.

\ciceroSection{3} To go through the items one by one is not necessary; in sum, I lay my whole zeal and my whole goodwill at your disposal. But it matters greatly to me that all of yours should know this --- which they may come to know through a letter from you --- that all that belongs to Cicero stands open to Trebianus. This is to the end that they may think nothing so difficult that, taken up on your behalf, it will not be a pleasure to me to undertake.

\ciceroSection{4} I would have sent you a letter before now if I could find the kind to write. For at a time like this, the office of friends is either to console or to make promises. Consolation I was not employing, because from many quarters I was hearing how bravely and how wisely you were bearing the injustice of the times, and how powerfully you were being consoled by your own consciousness of what you have done and meant. And if this is what you are doing, you are reaping a great fruit of the best studies --- studies in which I know you have always been engaged. I urge you, again and again, to keep at it.

\ciceroSection{5} At the same time, to you --- a man of the deepest expertise in events and in precedents and in all antiquity --- I, who am not unschooled myself, but who am perhaps less far advanced in study than I would wish and rather farther advanced in affairs and in their practice than I would wish, can give this pledge: that that bitterness and that injustice of yours will not be lasting. For he himself, who has the most power, seems to me daily to drift back toward equity and toward the nature of things; and the cause itself is of such a kind that --- together with the commonwealth, which cannot lie prostrate forever --- it must, of necessity, revive and recover with it; and every day something turns out gentler and more generous than we had feared. And since these matters are set on the tilts of the times, often on small ones, I shall watch every shift, and let pass no opportunity for helping you and for lifting your load.

\ciceroSection{6} And so that other kind of letter I spoke of will, I hope, day by day come more easily to me, so that I may even make promises. Better to fulfil them in the matter than in the words. As for you --- I want you to consider that you have more friends than those who are now, or have been, in your situation, so far as I have been able to make out, and that I yield to none of them. See to it that you keep a brave and a great spirit: this is in you alone. What lies in fortune will be governed by the times, and provided for by our counsels.
