% Ad Familiares 4.8 (ad-familiares-04-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO S. D. M. MARCELLO}

\ciceroSection{1} I do not presume to advise a man of such pre-eminent good sense; nor to give heart to a man of the greatest spirit and unmatched courage; nor, in any way, to console you. For if you bear what has happened as I hear you bear it, I owe your virtue congratulation rather than your sorrow consolation; and if these great public miseries break you, I am not so rich in talent as to console you, when I cannot console myself. What remains, then, is this: that I show myself, and prove myself, your man in every matter, and stand ready for everything those you love would wish, with the conviction that I owe you all things --- not only for your sake but for my own sake too, even what is beyond my power.

\ciceroSection{2} But here is something --- take it as you will: as my advising, or as my judging, or as a thing my goodwill could not keep silent. Bring yourself to the position I have brought myself to. If there is to be any commonwealth, in that commonwealth you must be, by the judgement of men and in fact a leading man, yielding only as necessity yields to the times; and if there is no commonwealth, then this is still the most fitting place even for going into exile. For if it is liberty we seek, what place is empty of this tyranny? And if any place at all, what place is pleasanter than one's own home? Believe me, even the man who holds all things looks kindly upon talent, and nobility and the standing of men he embraces, so far as his own situation and his own cause permit. But I have written more than I meant to. So I come back to that one point: that I am yours; and that I will be with your people, if indeed I shall have any of your people; and if not, that for my part I will satisfy our union and our love in all things, sufficiently and beyond. Farewell.
