% Ad Familiares 12.17 (ad-familiares-12-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO S. D. CORNIFICIO CONLEGAE.}

\ciceroSection{1} The remembrance of me which you signalled in your letter gratifies me deeply; and that you should keep it up I do ask --- not that I have any doubt of your constancy, but because it is the custom to ask in such terms. From Syria some rather disquieting news has been brought to us, which, since it touches you more nearly than it touches us, troubles me more on your account than on my own. At Rome there is the deepest quiet --- of a sort, however, that one would gladly trade for some salutary and honourable business; and that I hope is coming. I see Caesar has it in mind.

\ciceroSection{2} You should know that, while you are away, I have been writing more boldly --- as though I had picked up a kind of opportunity and licence. The rest of what I have written you would perhaps even grant me; but I have recently composed a book \textit{On the Best Kind of Oratory}, in which I have often suspected that you will differ a little from my judgement --- in the way, that is, that a learned man differs from one not unlearned. I should very much like you to vote for this book wholeheartedly, or, failing that, as a favour to me. I will tell your people to copy it out, if they wish, and send it on to you; for I think that, even if you find the substance less than convincing, still in that solitude of yours anything that has come from me will give you pleasure.

\ciceroSection{3} As to your commending your reputation and standing to my care --- in that you do as everyone does; but I should like you to take it this way: that, while I assign the very highest weight to the love I know is mutual between us, still in my judgement of your supreme talent, your excellent studies, and the hope you carry of the most ample distinction, I set no one above you, and few beside you.
