% Ad Familiares 4.9 (ad-familiares-04-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. M. MARCELLO.}

\ciceroSection{1} Though only a very few days ago I gave Quintus Mucius a letter for you, written at greater length, in which I had set out what frame of mind I judged you ought to be in and what I thought you ought to do, still, now that your freedman Theophilus is setting out --- a man whose loyalty and goodwill toward you I have closely observed --- I have not been willing to let him come to you without a letter from me. So I urge you again, and again, in the same terms in which I urged you in my earlier letter: that you make up your mind to be present in this commonwealth, whatever its condition, at the earliest moment. You will see perhaps many things you do not want to see; but not more than you hear of every day. And it is not your way to be moved by the sense of sight alone, and to be less troubled by what you take in by ear, when ear-knowledge generally seems the harder thing.

\ciceroSection{2} But (you will say) I shall have to say something I do not think, or do something I do not approve. To give way to the times --- that is, to obey necessity --- has always been held the wise man's part. Then again, as things stand at present, the fault you fear is not there. To say what you think, perhaps, is not allowed; to be silent is plainly allowed. Everything is referred to one man; and he uses no man's counsel --- not even his own people's --- but his own. This would happen scarcely otherwise had the man whom we followed been holding the state. Are we to imagine that a man who, in the war, when the danger we all shared bound us together, used his own counsel and that of certain men of the least possible prudence --- are we to imagine that he would have been more cooperative in victory than he was in uncertain times? And the man who, when you were consul, did not follow your counsel for all its wisdom, and when your brother held the consulship from the strength of your authority would not be guided by you --- now that he holds all things in his hand, do you suppose he would be looking to our opinions?

\ciceroSection{3} All things in civil wars are miserable --- a thing our ancestors did not feel even once, our own age has by now felt often. But nothing is more miserable than victory itself; for even when it falls to the better side, it makes those very men fiercer and more ungoverned, so that even if they are not such by nature, they are forced to be such by necessity --- since the victor must do many things at the will of the men through whom he has won, even against his own wish. Did you not see at the time, as I did, how cruel that victory would have been? Even then, then, would you have been without your country, so as not to see the things you did not want to see? \enquote{No,} you will say; \enquote{for I myself would have held on to my own resources and my own standing.} But it was the work of your virtue to count your own affairs of slightest moment, and to labour more strenuously for the commonwealth. Then again --- where is the end of this present course of yours? Up to this point your conduct is approved, and even, as one may praise fortune in such a case, your fortune is praised: your conduct, because you followed the opening of the war as you had to, and were unwilling to pursue its end with any gladness; your fortune, because by an honourable retirement you have preserved both the state and the reputation of your own standing. Now, however, no place should be sweeter to you than your country; nor should you love it the less because it is the more disfigured, but should pity it the more --- and not, when it has been bereaved of many illustrious men, deprive it even of the sight of you.

\ciceroSection{4} In short: if it was the part of a great spirit not to be a suppliant to the victor, see whether it is not the part of an arrogant one to spurn that same victor's generosity; and if it is the part of the wise man to go without his country, of the unfeeling one not to long for it; and if you cannot enjoy a public life, that it is stupid to refuse a private one. The central point is this: even if that other way of life seems to you the more agreeable, it must still be considered whether it may not also be the less safe. Great is the licence of swords; and in foreign places there is even less restraint where crime is concerned. As for me, I care for your welfare as much as Marcellus your brother does, or, at any rate, as much as the next man after him; what is yours to do is to look to the times, to your safety, to your life, and to your fortunes.
