% Ad Familiares 7.28 (ad-familiares-07-28) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. CVRIO.}

\ciceroSection{1} I remember the time when you struck me as out of your mind for choosing to live among those people rather than with us. For then the domicile of this City --- when indeed this City was still a city --- was far more suited to your refinement and good cheer than the whole Peloponnese was, let alone Patrae. Now, on the contrary, you seem to me to have seen far ahead, when, almost as our affairs had passed despair, you took yourself off to Greece; and in this present time you seem to me not only wise to be away, but happy. Although --- who, with any sense at all, can now be happy?

\ciceroSection{2} But what you, when you still could, have achieved on foot --- to be ``where neither the Pelopids'' (you know the rest) --- the same I more or less attain in another way. For when I have given the morning to the salutation of my friends --- which now happens, indeed, more thickly than it used to, because they seem to see in a like-minded citizen something of a rare white bird --- I bury myself in the library. So I am turning out work of a quantity you, perhaps, will feel; for I gathered from a remark of yours, when you reproached me at your own house for my dejection and despair, that you were learning from my books to miss the spirit that was in me.

\ciceroSection{3} But, by Hercules, I was then mourning for the commonwealth, which, by reason both of her kindness to me and of mine to her, was dearer to me than my life itself; and now too --- although I am consoled not only by reflection, which ought to count for most, but also by time, which is wont to heal even fools --- still it pains me that the public estate has so collapsed that not even hope is left of better days to come. Nor, indeed, does the fault now lie with the man in whose power everything stands (unless, perhaps, the standing of everything in one man's power was itself what should not have been): part has fallen out by chance, part by our own fault, and in such a way that there is no point lamenting the past. I see no remaining hope. So I return to where I began: wisely you abandoned this scene, if by deliberation; happily, if by chance.
