% Ad Familiares 13.29 (ad-familiares-13-29) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO L. PLANCO S.}

\ciceroSection{1} I do not doubt that you know that of all those friends who were bequeathed to you by your father, I am as closely bound to you as any --- and that not only by the sort of ties which present an outward appearance of close connection, but by those held together by intimacy and habitual exchange, which, as you know, were of the warmest and most agreeable kind between me and your father. From those beginnings my affection for you sprang and grew upon the connection inherited from him, and the more so because, from the first moment you were old enough to form a judgement of what each man was worth to you, I saw that you began, before all others, to mark me out for attention, for cultivation, for affection. There was, besides, no slight bond in our shared interests --- a thing weighty in itself, and weightier still when the interests in question, and the disciplines they involve, are of the kind that of themselves bind those who share the same inclinations also in close personal friendship.

\ciceroSection{2} I imagine you are waiting to see where so long a preamble is tending. Take this as your first point, then: that this rehearsal of mine has not been undertaken without grave and just cause. I am on the most familiar terms with Gaius Ateius Capito. The vicissitudes of my own fortunes are well known to you. Through every sort of office and trial I have undergone, the resolve and the labours and the standing and the goodwill of Gaius Capito --- and his very means as well --- were at hand, and answered to the demands of my circumstances and my fortune.

\ciceroSection{3} A relative of his was Titus Antistius. He, while holding Macedonia by lot as quaestor, and not yet superseded, was overtaken by the arrival of Pompey with his army in the province. Antistius could do nothing; for if he could have done, nothing would have been dearer to him than to return to Capito, whom he loved as a father, especially since he was aware how much Capito valued Caesar, and had always valued him. But, caught in the situation, he took only such part in the business as he could not refuse.

\ciceroSection{4} When silver was being coined at Apollonia, I cannot say that he had charge of the operation, nor can I deny that he was present at it; but only for two or three months. After that he kept away from the camp; he fled every sort of office. I should like you to take this on my word, as a witness; for he saw the depth of my own grief in that war, and shared everything with me. So he buried himself in the most remote part of Macedonia, as far from the camp as he could go, not only that he might have no charge of any business but that he should not even be present at any. After the battle he took himself to a connection of his, Aulus Plautius, in Bithynia. There, when Caesar saw him, he said nothing sharp, nothing bitter, and ordered him to come to Rome. The man fell at once into an illness from which he never recovered; sick, he came as far as Corcyra, and there he died. By the will he had made at Rome in the consulship of Paulus and Marcellus, Capito is heir to a half and a third; in the remaining sixth are those whose share can stand as public property without any complaint from anyone --- and that share amounts to thirty thousand sesterces. But on this point Caesar will see for himself.

\ciceroSection{5} You, my dear Plancus, by the connection inherited from your father, by our mutual affection, by our shared pursuits and the close resemblance of the whole course of our lives, I ask, and ask of you in such terms as I could press no other claim with greater care or greater earnestness, to take this matter up, to count it as my own, to make every exertion, to strain every effort, to bring it about that by my recommendation, by your zeal, and by Caesar's favour, Gaius Capito should obtain his relative's inheritance. Whatever I could have asked of you, with all your standing and influence such as it is, if I had asked it, I shall reckon you have offered me unprompted, if I obtain this one thing.

\ciceroSection{6} One thing, I hope, will be of help to you, and Caesar himself can be its best judge: Capito has always cultivated and esteemed Caesar. But he himself is witness of the fact; I know how good his memory is for a man. So I instruct you in nothing; only assume on Capito's behalf, in Caesar's presence, as much weight as you find that Caesar himself remembers.

\ciceroSection{7} For my own part, I shall lay before you what I have been able to learn by experience, in my own case; how much weight there is in it you will judge. Which side I defended in the republic, and on what men and what orders I rested, and by whom I was supported, you are not unaware. I should like you to take this on my word: that if I did anything in this present war which fell short of Caesar's wishes (and I have understood that Caesar himself knows I did it most unwillingly), I did it on the counsel, the urging, the authority of others; that I was more moderate and more restrained than anyone on that side, was due above all to the authority of Capito --- and if I had had the rest of my connections like him, I should perhaps have done some service to the republic, and certainly the greatest service to myself.

\ciceroSection{8} This thing, my dear Plancus, if you bring it to pass, you will have confirmed the hope I rest in your goodwill toward me, and you will have joined Capito himself --- a man most grateful, most attentive, of the best character --- to your own connection by the greatest of favours.
