% Ad Familiares 15.15 (ad-familiares-15-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. C. CASSIO}

\ciceroSection{1} Although both of us, from hope of peace and from hatred of the spilling of citizen blood, wished to stand aside from the obstinacy that war makes necessary, still --- since I am taken to have been the first author of that plan --- I owe it perhaps to render more to you myself than to expect from you. And yet, as I often have occasion to remind myself, our intimate conversation, mine with you and equally yours with me, brought us both to this same resolution: that we should treat a single battle, if not as deciding the whole cause, then at any rate as settling our personal judgement in the matter. Nor has anyone honestly faulted this view of ours, except those who think it better that the state should be wiped out altogether than survive diminished and crippled. For my part, I held out no hope to myself from its destruction, but a great hope from what would be left.

\ciceroSection{2} What has followed, however, has been such that one wonders less that these things should have happened than that we did not foresee them, or, being mortal men, could not divine. For my own part I confess my reckoning was this: that with that one battle, as if appointed by fate, fought, the victors would mean to consult the common safety, and the defeated their own; and I judged that each of these things turned on the speed of the victor. Had there been such speed, Africa would have known the same clemency that Asia found, and Achaea too --- where you, I think, were yourself the legate and intercessor. As it is, opportunities have been let slip --- and they count for most, particularly in civil wars; a year has come between, and has led some to hope for victory and others to scorn the very fact of being beaten. The whole blame for these evils Fortune carries; for who would have supposed that the Alexandrian war was to be tacked on at such length to this one, or that some Pharnaces or other was going to spread terror through Asia?

\ciceroSection{3} We, then, with the same plan have met different fortunes. You took the part of being in on the counsels and --- the greatest relief to anxiety --- of being able to look ahead in your mind to what was coming; I, who hurried to see Caesar in Italy (for so we used to think), and, on his return after preserving so many men of the highest standing, to ``urge him on toward peace,'' as the saying is, ``while he was already running,'' am as far from him as I could possibly be and have been throughout. As it is, I am taken up with the groaning of Italy and with the most miserable laments of the city; to which, perhaps, I might have brought some help, in my measure, you in yours, every man in his --- had there been any authority on the spot.

\ciceroSection{4} Therefore, in the long-standing goodwill you have shown me, I should be glad if you would write to me what you see, what you feel, what you think we should look forward to and what we should do. Your letters will mean much to me; and would that I had obeyed those earliest ones you sent from Luceria! For then I should have kept my dignity without any vexation at all.
