% Ad Atticum 11.9 (ad-atticum-11-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} It is true: I acted, as you write, both incautiously and more quickly than I should have done, and I have no hope, held as I am by the exceptions made in the edicts. Were these not the work of officious zeal and \dag of goodwill,\dag\ I might have been free to withdraw into some solitude. As it is, not even that is open to me. And what does it profit me that I came in before the entry of the new tribunate, if the very fact of having come avails me nothing? What am I to hope for now from the man who has never been my friend, when I am hemmed in and overborne by his very law as well? Day by day now Balbus's letters to me grow more languid, and many letters from many men reach him, no doubt, against me. I am being ruined by my own fault; chance has brought me no evil, all has been incurred by my own doing. For when I saw the character of the war --- everything on our side unprepared and weak against an enemy in the highest readiness --- I had settled on what I should do, and I had taken a course not so much brave as one that, granted to others, ought also to be granted to me beyond the rest.

\ciceroSection{2} I gave way to my own people --- or rather, I obeyed them. What spirit was in one of those people --- the one you commend to me --- you will gather from his own letters, those he sent to you and to others. I should never have opened them had not events fallen out as they did. A packet was delivered to me; I unsealed it, in case there was anything for me in it. There was nothing: one letter for Vatinius, another for Ligurius. I gave orders for them to be carried to those addressees. They came to me at once, burning with grief, crying out at the man's wickedness; they read me letters full of every kind of reproach against me. At this Ligurius flew into a rage: he knew, he said, that Quintus had been an object of the deepest hatred to Caesar; that even so Caesar had not only favoured him but had given him a great sum of money, and this in consideration of myself. After taking this blow, I wanted to know what he had written to the others, since I thought it would be ruinous for the man himself if so monstrous a piece of villainy on his part should get abroad. I found the rest to be of the same kind. I have sent them on to you. If you think it would be useful for him himself that they should be delivered to him, deliver them. Nothing will hurt me. As for their having been broken open --- Pomponia, I believe, has his seal. By this bitter act, at the very outset of his voyage, he wrought such grief in me that I have been laid low ever since; and now he is said to be exerting himself not so much for his own case as against mine.

\ciceroSection{3} So I am pressed on every side, by burdens which I can scarcely bear, or rather plainly cannot bear at all. Among these miseries, one stands in for all: that I am leaving that poor girl stripped of her patrimony, of all fortune. So I should very much wish, as you promise, to see you in person; for I have no one else to whom I might commend her, since I have come to see that her mother too has prepared for her the same as is prepared for me. But if you do not find me, hold her commended to you nevertheless, and soften her uncle toward her as far as you can. I have written this to you on my birthday. Would that I had never been taken up at birth, or that nothing further had been born to the same mother since! Tears prevent me from writing more.
