% Ad Atticum 11.14 (ad-atticum-11-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The truthfulness of your letter does not offend me --- that you do not even begin, as you used to, to console me, crushed as I am by ills both general and particular, and admit that this can no longer be done. For things are not as they were before, when, if nothing else, I supposed I had companions and partners. For all those in Achaia who were pleading for pardon, and likewise in Asia those who were not pardoned (and even those who were), are said to be on the point of sailing to Africa. So apart from Laelius I have no partner in my fault; and even he is in a better case, in that he has already been received back.

\ciceroSection{2} As for me, I do not doubt he has written to Balbus and Oppius; and if anything more cheerful had come from them, I should have been informed, and they would also have spoken to you. I should like you to confer with them on this very point, and to write back to me what answer they give --- not because any deliverance granted from that quarter will have anything firm about it, but still some plan and provision can be made. Although I shudder at being seen by anyone, especially with this son-in-law of mine, still in such ills I can find nothing else to wish for.

\ciceroSection{3} Quintus is pressing on, as both Pansa and Hirtius have written to me, and he too is said to be making for Africa with the rest. I shall write to Minucius at Tarentum and send him your letter; I shall write to you what I have managed. I should be surprised that thirty thousand sesterces could be raised, were it not for the proceeds of the Fufidian estates. \dag and yet I see... I am waiting\dag; whom I want very much to see, if it is in any way possible (for the situation demands it). Now the final reckoning is closing in; \dag there it is easy to judge, but how grave it is, here is harder to judge.\dag\ Farewell.
