% Ad Atticum 11.25 (ad-atticum-11-25) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I readily assent to your letter, in which you set out at length that no plan stands by which I can be helped by you. There is certainly no consolation that can lighten my grief. Nothing has been brought on by chance --- that would be bearable --- but everything we have done to ourselves, by errors and miseries of mind and body which I only wish those nearest to me had cared to remedy. Therefore, since no hope of counsel from you, or of consolation from anyone, is set before me, I shall not be asking these things of you hereafter; only do not break off --- write to me whatever comes into your mind, when you have someone to whom you can give the letter, and as long as there is someone to whom to send it; and that will not be long.

\ciceroSection{2} The rumour that he has set out from Alexandria first arose, not on firm ground, from a letter of Sulpicius; all the subsequent messengers have confirmed it. Whether it is true or false, since it makes no difference to me, I do not know which I should prefer.

\ciceroSection{3} As to what I wrote you long since about the will, \dag I should wish that in the letters I might be able to face what is against me.\dag\ I am wrecked by this most wretched girl's \dag situation,\dag\ and I am fighting against it. I think nothing was ever like it. If by anything I can take thought for her, I should like to be reminded by you. I see that the difficulty is the same as it was before about giving counsel; yet this disturbs me more than everything else. At the second installment we were blind. I should have wished otherwise; but it is past. I beg you, in our ruin, if anything can be gathered and got into safekeeping --- out of the silver, and a good deal too out of the furniture --- give your effort to it. For now the end seems to be on us, and there will be no terms of peace, and what we have will perish even without an enemy. Speak of these things, if it seems right, with Terentia at a fitting moment. I cannot write everything. Farewell. The third day before the Nones of Quintilis.
