% Ad Atticum 11.24 (ad-atticum-11-24) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What you wrote to me some time ago, and what you also \dag wrote at the same time\dag\ about myself to Tullia, I feel is true. I am the more wretched on that account, though it seemed nothing could be added to my misery, in that, after suffering the gravest injury, I am not allowed even to be angry, indeed not even to grieve, without paying for it. So this we must bear. And when we have borne it, the same things remain to be endured against which you warn us to take care. The fault we have brought on ourselves is such that in every condition and under every regime it appears to be heading toward our ruin.

\ciceroSection{2} But I come now to my own hand, since these things must be transacted more secretly. Look, please, even now to the will --- the one made when she had begun \dag to complain.\dag\ I do not suppose it occurred to him; for he did not even ask --- not even me. But as though that were the case, since you have now broached the subject, you will be able to warn her to commit it to someone whose fortune is out of reach of this war's danger. For my part, I should most prefer you, if she should wish the same. I keep from her, poor thing, that I am afraid of this. As to the other property, I know that nothing can be sold at present, but things can be set aside and concealed so as to lie outside the ruin that is hanging over us.

\ciceroSection{3} As for what you write --- that we shall have our own provided, both yours and Terentia's --- yours, I do not doubt; what of ours can there be? About Terentia, in fact (I pass over the rest, which is past numbering), what can be added to this? You had written that I should have a draft on the bank for twelve thousand sesterces; that was, you said, what remained of the silver-money. She sent me ten thousand, with a note that this was the whole remainder. When she has filched so small a sum out of so small a sum, you see clearly what she has done in the great affair.

\ciceroSection{4} Philotimus has not only failed to arrive, but does not even by letter or by messenger let me know what he has done. People coming from Ephesus report having seen him there going into court over his own disputes --- which (this is the likely thing) are perhaps being put off until Caesar arrives. So either I suppose he has nothing he thinks worth forwarding to me in haste, or my position is so utterly despised that, even if he has something, he does not trouble to send it to me except when all his own business is finished. Of this I take great pain, certainly, but not as much as I seem bound to take. For I think nothing is of less concern to me than what may be coming from there. Why this is so, I am quite sure you understand.

\ciceroSection{5} As to your advice that I adapt my expression and my speech to the times --- though it is hard, still I should make myself do it, if I thought there were anything in it for me. As for your writing in your letter that you think the African business can be brought to an end, I wish you would write why you think so; for nothing comes to my mind on the strength of which I should believe it possible. Still, please, if there will be anything that holds a measure of consolation, write to me; but if, as I see it, there will be nothing, write me that very fact. I, for my part, will write to you if I hear anything more quickly. Farewell. The eighth day before the Ides of Sextilis.
