% Ad Atticum 11.23 (ad-atticum-11-23) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The matter I had written to you about, asking you to share it with Camillus --- Camillus has written to me that you discussed it with him. I was waiting for a letter from you; though indeed, if it is otherwise and must be otherwise, I do not see that this in itself can be altered. But once I had received a letter from him, I missed one from you (even though I supposed you had not been informed) --- if only to know you were well; for you had written that you were being tested by some sort of ill health.

\ciceroSection{2} A man called Agusius had come from Rhodes on the eighth day before the Ides of Quintilis. He reported that my son Quintus had set out for Caesar on the fourth day before the Kalends of June, and that Philotimus had arrived at Rhodes the day before that, and was carrying letters for me. You will hear Agusius himself. But he was travelling rather slowly, so I have arranged for it to be carried by one who is going swiftly. What is in those letters I do not know, but my brother Quintus is offering me very strong congratulations. For my own part, given how grave my error has been, I cannot in even the most distant thought reach anything I could find bearable.

\ciceroSection{3} I beg you, think about this poor girl, and about the matter I most recently wrote to you of --- arranging something to ward off her destitution --- and also about the will itself. I could have wished this too earlier, but we were afraid of everything. Indeed, nothing among our miseries would have been better than a divorce. We could have done something --- by way of a man's role, that is, on grounds of the new account-books, or the nocturnal raids, or Metella, or the whole roster of his crimes; we should not have lost the money, and we should have seemed to have a man's share of indignation. I remember well your letter --- but the times too; though anything would have been better. As it is, the man himself seems to be giving us notice; for we hear about the statue of Clodius. Our son-in-law, of all people --- this, or new account-books? It is my view, then, and yours too, that the bill of divorce be sent. Perhaps he will sue for the third installment of the dowry. Consider, then, whether the moment should be when that comes from him, or sooner. As for me, if I can in any way, even by travelling by night, I shall try to see you. Please write to me on these matters and on anything else it may concern me to know. Farewell.
