% Ad Atticum 11.15 (ad-atticum-11-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Since you bring good reasons why I cannot see you at this present time, I beg you, consider what I ought to do. For he, it seems, has so taken Alexandria in hand that he is even ashamed to write about those matters, while the men from Africa now appear to be at hand; and the Achaian party as well are about to return either to them out of Asia, or to remain in some place that is free. What, then, do you think I ought to do? I see that the decision is hard. For I am alone, or with one other, for whom there is no return to those men either, and from these very ones no hope is held out. Even so, I should like to know what you think; and that, alongside other reasons, was why I wished, if it could be done, to see you.

\ciceroSection{2} I wrote to you before that Minucius had taken care of only twelve thousand. What remains, I should like you to see is taken care of. Quintus has written to me not only with no humble entreaty but in the bitterest terms; and his son with an extraordinary hatred. There is no evil that can be invented with which I am not pressed. Yet all of it is easier to bear than the pain of having done wrong, which is both at its height and unending. If, for that wrong, I were going to have the partners I supposed, even so the consolation would be slight. But every other man's case has an outlet; mine has none. Some, captured, others, cut off, are not called into question over their intention --- still less, of course, once they have got themselves clear and have begun to be together. And even those very men who came of their own will to Fufius can only be judged timid. There are many besides who, by whatever means they wish to take themselves back to those leaders, will be received. So you should be the less surprised that I cannot stand up against so great a pain. For mine alone, of all faults, cannot be repaired, and perhaps Laelius's too. But how does that ease me? For they say even Gaius Cassius has changed his plan of going to Alexandria.

\ciceroSection{3} I write this to you, not \dag to lift some of your worry\dag, but in order to learn whether you have anything to bring to bear on what is undoing me; on top of which there is my son-in-law, and the rest of it, which I am held back from writing of by tears. Even Aesopus's son is racking me. There is positively nothing missing to make me utterly wretched. But I come back to the first question: what do you think I should do --- should I go in some secret way nearer to Rome, or cross the sea? For I cannot stay here any longer.

\ciceroSection{4} On the Fufidian matter, why could nothing be concluded? For the kind of arrangement was of a sort in which there is not usually any dispute, since the share which appears to be the smaller can be filled out by bidding. I do not ask this without reason. I suspect my coheirs think our case is in doubt, and for that reason prefer to leave the matter open. Farewell. The day before the Ides of May.
