% Ad Atticum 11.13 (ad-atticum-11-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} From Murena's freedman I had had no letter as yet. Publius Siser delivered the one I am answering. As for what you write about a letter from old Servius, and what you say about there being people who report that Quintus has come to Syria --- that too is untrue. As for your wanting me to give you a more definite account of the disposition that each of those who came over here had, or has had, towards me: of those who have arrived I have understood none of them to be hostile. But how much that matters to me, I am sure you can judge. Everything is unbearable to my grief, and most of all the fact that I see I have come to such a pass that only those things appear useful to me which I have always refused to want. Publius Lentulus the father is said to be at Rhodes, his son at Alexandria; and that Gaius Cassius set out from Rhodes for Alexandria is well established.

\ciceroSection{2} Quintus by letter makes his apologies to me, in much sharper terms than when he was levelling the gravest accusations. He says that he understands from your letters that you do not approve of his having written rather harshly about me to many people, and accordingly that he is sorry he has offended you; but that he himself has acted within his rights. Then he sets out at length, in the foulest fashion, the reasons for which he did so. But neither at this time nor before would he have laid bare his hatred for me if he had not seen that I was overwhelmed at every turn. Would that even by night-marches, as you had written, I had come closer to you! As things stand I cannot so much as guess where or when I shall see you.

\ciceroSection{3} About the co-heirs in the Fufidian estate there was no need to write to me; their demand is fair, and whatever you have done I should reckon rightly done.

\ciceroSection{4} About buying back the farm at Frusino, you have long understood my wish. Although our affairs then stood in a better place, and there did not seem to be any prospect of such ruin for me, even so I am of the same mind. How it is to be managed you will see. And I should like you, as far as you can, to consider where the necessary expenses can be supplied from. What resources we had, we made over to Pompey at the time when we thought we were acting wisely. Accordingly at that time we drew on your steward and borrowed from elsewhere, even as Quintus complains in his letters that we gave him nothing --- when we were neither asked by him nor laid eyes on the money ourselves. Even so, I should like you to see what can be done, and to give me advice on every part of it; you know the situation.

\ciceroSection{5} Grief keeps me from writing more. If there is anyone you think should be written to in my name, I should like you to do as you usually do; and as often as there is someone to whom you can give a letter to me, I would have you not let the chance pass. Farewell.
