% Ad Familiares 15.21 (ad-familiares-15-21) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. C. TREBONIO}

\ciceroSection{1} I read your letter with pleasure and your book with the greatest pleasure of all; yet in the midst of that delight I felt this pang --- that, after kindling in me a fresh desire to increase our intimacy (for to my love itself nothing could be added), you then leave us, and afflict me with such longing that you leave me only one consolation: that the longing each of us feels for the other in absence will be eased by frequent and long letters. This I can pledge to you not only for my part, but also for yours to me; for you have left in my mind not the slightest doubt of how much you love me.

\ciceroSection{2} Even setting aside those acts to which the whole city was witness --- when you shared my enmities with me, when you defended me in your public speeches, when as quaestor you took up the consuls' side in my cause and in the cause of the state, when as quaestor you refused to obey a tribune of the plebs whom even your own colleague obeyed --- to set aside those recent acts which I shall always remember, even if I forget all else: your anxiety on my behalf when I was in arms, your joy at my return, the care and grief you felt when my own cares and griefs were reported to you, and at the last that you would have come to me at Brundisium, had you not been suddenly dispatched to Spain --- setting aside, then, these things which I am bound to value as I value my life and my preservation, this book that you have sent me, what a declaration of your love it carries! First, that everything I have said seems witty to you, which is perhaps not how it strikes others; next, that those sayings, whether they are witty or merely so-so, become exquisitely charming in your telling --- to the point that nearly all the laughter is used up before one reaches me at all.

\ciceroSection{3} If, in writing these pages, you had spent so much time thinking of me alone with no thought beyond the bare necessity of the work, I should be made of iron not to love you in return. But since you could not have thought through the matter you have followed out in writing without the deepest love, I cannot suppose that any man is loved by himself more than I am loved by you. To that love, would that I could answer in other coin --- but I shall answer it at least in love itself, with which I trust you will be content from me.

\ciceroSection{4} I come now to your letter, to which, copiously and pleasantly written as it is, I have no extensive reply to make. First: I sent that letter to Calvus no more than I sent this one which you are now reading, supposing it would ever go abroad; for we write in one way what we mean only for those to whom we send it, and in another what we suppose many will read. Next: I praised his talent in higher terms than you think the case could truthfully bear --- first, because I judged it so. He had sharpness of mind, he was following a definite school; and though in following it he stumbled in judgement, by the strength he had he still attained what he could put forward as good work. His reading was wide and recondite; what he lacked was force. To that, then, I was urging him on. And in stirring a man up and sharpening him, it counts for the most that you praise the one you are exhorting. There you have my judgement of Calvus, and my policy: the policy, in that I praised him to spur him on; the judgement, in that I really did think very well of his talent.

\ciceroSection{5} It remains for me to attend your departure with my love, await your return with hope, keep your memory in your absence, and ease all my longing by sending and receiving letters. As for you, I should be glad if you would often go over with yourself the kindnesses and services you have rendered me. When you may forget them and I may not, you will conclude not only that I am a good man, but also that I love you with all my heart. Farewell.
