% Ad Atticum 11.12 (ad-atticum-11-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Cephalio delivered a letter from you to me on the eighth day before the Ides of March, in the evening. Earlier the same day, in the morning, I had sent off couriers; with them I had given you a letter. But after reading yours I thought something further had to be written back, especially because you make it plain that you are unsettled in mind as to what reason for my setting out I am going to put before Caesar at the moment of my departure from Italy. I have no need of a new reason. I have written to him often, and given many people the message, that I had been unable, much as I wished, to bear up under what people were saying, and a great deal in that line. There was nothing I less wanted him to think than that in so grave a matter I had failed to act on a judgement of my own. And later, when a letter reached me from the younger Balbus Cornelius saying that Caesar took my brother Quintus to have been the trumpet of my departure (that was his word), I --- who did not yet know what Quintus had been writing about me to many people, though many bitter words and acts had already passed face to face between us --- none the less wrote to Caesar in these words:

\ciceroSection{2} ``About my brother Quintus I am no less concerned than about myself; but to commend him to you at this moment of mine I do not dare. I shall venture only to ask of you what I can: that you suppose nothing to have been done by him to make me less constant in the services I owe you, or less attached to you; and that you take it always to have been he who was the prime mover of our connection and the companion of my journey, not my guide. So in everything else you will allow him as much as your generosity and the friendship between you require. As for me, that he should not stand at any disadvantage with you on my account --- this I entreat of you again and again, most earnestly.''

\ciceroSection{3} Therefore, if any meeting takes place between me and Caesar --- and I do not doubt that he will be lenient toward Quintus, and has already shown himself so --- still I shall be the man I always have been. But, as I see, we have far more to worry about over Africa, which you write of as gathering daily more for the hope of negotiated terms than of victory. Would that it were so! But I understand that it is far otherwise, and I suppose that you yourself think so, and write differently not to deceive me but to keep my courage up, especially since to Africa Spain too is now being added.

\ciceroSection{4} As for your advising me to write to Antony and the others, if you think any of it needful, please do what you have often done. Nothing comes into my mind that I judge worth writing. As for your hearing that my spirits are somewhat raised, what do you think, when you see that on top of my old griefs there have come the splendid performances of my son-in-law? I would have you, even so, not leave off writing to me, as far as you can, even if you have no matter to write about. Your letters always bring me something. I have entered upon Galeo's inheritance --- I take it the formal act of entry must have been the simple one, since none has been sent to me. The eighth day before the Ides of March.
