% Ad Familiares 9.1 (ad-familiares-09-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO M. VARRONI S.}

\ciceroSection{1} From the letter that Atticus read me, the one you sent him, I have learned what you are about and where you are; when, however, we are to see each other, I could gather nothing whatever from the same letter. Still, I let myself hope that your arrival is near; and I pray it may bring me some consolation. Even so, we are pressed by so many and such great troubles that no one but a perfect fool ought to hope for any relief; yet perhaps you can help me, or I you, in something.

\ciceroSection{2} Know, then, that since coming back to the city I have made my peace with my old friends --- that is, with our books. Not that I had cut off their company because I was angry with them, but because I felt some shame in their presence. I thought that, having plunged myself into the most turbulent affairs and the most untrustworthy allies, I had not been obedient enough to their precepts. They forgive me; they call me back into our old intimacy; and they say that you, who held to it, were the wiser man of the two. Since, then, I have them placated, I think I may hope that, if I see you, I shall bear easily both the things that press upon us now and the things that hang over us. So whether you would prefer that I come to you at Tusculum or at Cumae, or --- which I should least like --- at Rome, provided only that we are together, I shall make sure that we settle on whatever is most convenient for us both.
