% Ad Atticum 11.8 (ad-atticum-11-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The cares I am being worn down by you doubtless see for yourself; still you will learn of them from Lepta and Trebatius. I am paying the heaviest penalty for my own rashness --- which you would have me regard as prudence; and I do not deter you from continuing to argue it, and to write to me as often as possible. Your letters relieve me not a little at such a time. There is need that, through those who wish us well and have weight with Caesar, you press with the utmost diligence --- through Balbus and Oppius above all --- that they write on my behalf as diligently as they can. For I am, as I hear, under attack from certain people in his presence and through letters. These attacks must be met with a force answerable to the magnitude of the matter.

\ciceroSection{2} Fufius is there, my bitterest enemy. Quintus has sent his son not only as advocate for himself but as accuser of me. He keeps saying that he is being attacked by me in Caesar's company --- a charge Caesar himself and all his friends refute. Nor in fact does he leave off, wherever he is, heaping every abuse upon me. Nothing more incredible has ever befallen me, nothing in these troubles more bitter. People who had heard it from his own mouth, when at Sicyon he was speaking openly in front of many witnesses certain monstrous things, brought the report to me. You know the type --- you have perhaps tasted it yourself. All of it has now been turned against me. But I am increasing my distress by rehearsing it, and yours too. So I come back to that one point: see that, with this matter expressly in mind, Balbus sends someone. To whoever seems good, please have letters dispatched in my name. Farewell. The thirteenth day before the Kalends of January.
