% Ad Atticum 11.1 (ad-atticum-11-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I received from you the sealed packet that Anteros had brought; from it I was able to learn nothing about my household affairs. About these I am most bitterly tormented, because the man who managed them is neither there with you nor, so far as I know, anywhere on earth. The whole hope I have for my standing and my private affairs rests on the goodwill toward me which I have so thoroughly proved in you. If you sustain that goodwill in these wretched and final times, I shall bear the more bravely these dangers I share with everyone else; and that you should do this I beg and beseech you.

\ciceroSection{2} I have on deposit in Asia, in cistophori, to the amount of twenty-two hundred thousand sesterces. By the transfer of this money you will easily protect my credit; and I should not have left it unsecured --- I trusted the man you know I have long had every reason not to trust --- had I not supposed I was leaving it in good order. I would have stayed a little longer and not left my household affairs in a tangle. And it is for this reason that I write to you so belatedly: I understood too late what there was to fear. Again and again I beg you to take me wholly into your protection, so that, if those with whom I am stand safe, I may stand safe along with them, and may owe my preservation to your goodwill.
