% Ad Atticum 11.2 (ad-atticum-11-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I received your letter on the day before the Nones of February, and on that same day I formally entered upon the inheritance under the terms of the will. Out of my many most wretched anxieties one has been lightened, if, as you write, that inheritance can protect my credit and my good name --- a credit which, I know well enough, you would have defended out of your own resources even without an inheritance.

\ciceroSection{2} As to what you write about the dowry, by all the gods I beseech you, take up the whole business and protect that unhappy girl --- whose ruin is from my own fault and negligence --- out of my resources, if I have any, and out of yours so far as it gives you no trouble. That she lacks for everything, as you write, do not, I implore you, allow it. For into what expenses are the proceeds of the estates running off? As for that sixty thousand sesterces you mention, no one ever told me it had been deducted from the dowry; I would never have allowed it. But this is the smallest of the wrongs I have suffered --- of which I am forbidden by my grief and my tears from writing to you. From the money that was in Asia I have collected nearly half.

\ciceroSection{3} It seemed safer for it to be where it is than with the tax-farmers. As to your urging me to keep my spirits firm, I could wish that you were able to bring forward something to make that possible for me. But if, on top of my other miseries, the report Chrysippus brought me about the house has also come true --- which you gave no hint of --- has any one man yet been more wretched than I? I beg you, I implore you, forgive me. I cannot write more. With what grief I am pressed down, surely you see. If this trouble were shared with the rest who seem to be in the same case, my own fault would look smaller and would for that reason be more bearable. As things stand there is nothing to console me, unless you can bring it about --- if even now it can still be brought about --- that I be not visited with some particular and singular calamity and outrage.

\ciceroSection{4} I have sent your courier back rather slowly because I have had no means of dispatching him. From your people I received seventy thousand sesterces, and the clothing that I needed. Please give letters in my name to whomever you think right. You know my friends. If they look for the seal or the handwriting, you will say that on account of the watchers set on me I avoided these.
