% Ad Atticum 11.7 (ad-atticum-11-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Your letter is welcome to me; you wrote out carefully everything you judged concerned my affairs. So you write that the act itself meets with the approval of that quarter, and that those same people approve of my keeping these lictors, since the same concession was made to Sestius --- though I do not think his were granted to him, but rather assumed by him on his own authority. For I hear that he disapproves of those decrees of the Senate which were passed after the tribunes had withdrawn. So, if he chooses to be consistent, he will be able to approve our lictors as well.

\ciceroSection{2} And yet why do I speak of lictors, when I have all but been ordered to quit Italy? For Antonius sent me a copy of a letter from Caesar to him, in which it was written that he had heard that Cato and Lucius Metellus had come into Italy in order to be openly at Rome; that this did not meet his approval, lest some disturbance arise from it; that everyone was to be barred from Italy except those whose case he himself had examined; and on this point the writing was vehement. Accordingly Antonius asked me by letter to pardon him: he could not but obey those instructions. I then sent Lucius Lamia to him, to point out that Dolabella had told him to write to me to come into Italy as quickly as possible; that I had come on the strength of his letter. Then he issued his edict in terms that made an exception of me, and of Laelius by name --- which I should very much have preferred he had not done; the exception could have been made for the matter itself, without naming anyone.

\ciceroSection{3} O the many heavy mortifications! And you do your best to ease them, and you do not fail of some effect; on the contrary, you lighten my pain by this very fact, that you toil so hard to lighten it; and I should be glad if you would not find it burdensome to do this as often as possible. But you will most fully attain what you wish if you bring me to think that I have not utterly lost the good opinion of honest men. Yet what can you do toward that? Nothing, plainly. Still, if events offer you any opening, that more than anything will be able to console me; though for the moment I see there is none, unless from the way things turn out, as has happened just now. I was being said to have had a duty to set out with Pompey. The outcome of his expedition lessens the reproach of having neglected that duty. But of all charges none is more held against me than that I did not go into Africa. My reasoning in this was that the commonwealth ought not to be defended by the barbarian auxiliaries of a most treacherous nation, especially against an army that has so often been victorious. They do not approve, perhaps; for I hear that many good men have gone into Africa, and I know that many were there before. On this point I am sorely pressed. Here too I need fortune to be with me, that some at least of them --- or, if it can be, all of them --- should set safety above all else. For if they persist and prevail, you see what is to become of me. You will say: what of them, if they are beaten? Their wound is the more honourable. These thoughts rack me.

\ciceroSection{4} As for Sulpicius's policy, you did not write why you should not prefer it to mine. Though it is not as glorious as Cato's, still it is free of danger and free of distress. The worst is the lot of those who are in Achaia; yet even they are better off than I am, in that they are many in one place, and when they come into Italy will at once be at home. You go on, as you are doing, softening matters and winning approval for me from as many as you can.

\ciceroSection{5} When you make your excuses, I really do understand your reasons, but I also think it is in my interest that you should be there, both to deal with those with whom dealings will have to be had on my behalf about what is to be done, and to follow up the dealings you have already had. Above all I should like you to give your attention to this. I believe there are many who have brought reports to Caesar, or will bring them, that I either repent of my course or do not approve of what is being done. Even if both these statements are true, still they are made by men who have set themselves against me, not as people who have actually seen it to be so. But the whole matter rests on this: that Balbus and Oppius hold their ground, and that by their frequent letters Caesar's goodwill toward me is kept firm. You will see to it that this is done in plain fact.

\ciceroSection{6} There is another reason why I do not want you to go away --- you write that Tullia presses you. O wretched business! What am I to write, what am I to want? I shall be brief, for the tears have suddenly burst out. I leave it to you: you decide; only see that no harm can come to her in present circumstances on this account. Forgive me, I beg you. I cannot stay longer on this matter, for weeping and grief. Let me say only this: nothing is dearer to me than that you care for her.

\ciceroSection{7} As for the letters which you think need to be carried, you are taking proper steps to have them sent. I have seen the young Quintus, who had seen his father at Sicyon, after seeing him himself at Samos: their plea for forgiveness is easy. Would that those who saw him before me had been willing to help me with him as much as I should have wished to help them if I had had any power!

\ciceroSection{8} You ask me to take in good part anything in your letters which may sting me: I take it in the very best part, and I ask you, as you do, to write to me openly about everything, and to do so as often as you can. Farewell. The fourteenth day before the Kalends of January.
