% Ad Familiares 9.9 (ad-familiares-09-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{DOLABELLA S. D. CICERONI.}

\ciceroSection{1} If you are well, I am glad; I am well. Our Tullia is well too. Terentia has been less well, but I know for certain that she has now recovered; and otherwise everything on your side of things is going perfectly. Granted, there has never been a time at which you should have suspected me of urging on you what made for our side rather than what made for yours --- of pressing you either to come over to Caesar and to us, or at any rate to retire into quiet --- yet now in particular, with the victory already tipping our way, I cannot fall under any other imputation than this: that I am giving you advice which it would be impious to keep to myself. But you, my dear Cicero, will take what I write in this spirit: that whether you approve it or do not approve it, you will judge that it has been thought and written from a heart that is, at any rate, your most devoted.

\ciceroSection{2} You see for yourself that Cn. Pompeius is safe neither in the renown of his name nor in his record of deeds, nor even in those clientships of kings and nations that he was always so fond of parading; and that this too --- a thing granted to the lowliest man --- cannot be granted to him: to make an honourable escape. He has been driven out of Italy, he has lost the Spains, he has had his veteran army taken from him, and now at last he has been walled in --- a thing which I rather think has never happened to any commander of ours. Consider then, as your good sense will let you, what he can hope for and what you can; for that is the easiest way to take the decision that will serve you best. But this I beg of you: if he has now escaped this danger and shut himself up in his fleet, look to your own affairs, and be at last more of a friend to yourself than to anyone else. You have done enough by now for duty, enough for friendship, enough also for the cause and for that commonwealth which you used to approve.

\ciceroSection{3} What remains is this: that we be where the commonwealth now is, rather than, while we follow the old commonwealth, be in none at all. Therefore, my most delightful Cicero, if it should happen that Pompeius is driven from these regions too and forced again to make for other parts, I would wish you to withdraw either to Athens or to whatever quiet city you please. If you mean to do so, please write to me, so that, if I possibly can, I may fly to your side. Whatever has to be obtained from the commander touching your standing, given Caesar's humanity, will be the easiest thing in the world for you to obtain from him directly; and yet I think my own pleas will carry no small weight with him. It will be a mark of your good faith and your humanity, too, to see that the courier I have sent to you can return to me and bring me a letter back from you.
