% Ad Atticum 11.5 (ad-atticum-11-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The causes that moved me --- how bitter they were, how heavy, how unprecedented, and how they forced me to act by a kind of impulse of the mind rather than by deliberation --- I cannot write to you of without the greatest pain. They were such, in any event, that they produced what you now see. And so I can find nothing to write to you about my affairs, nor anything to ask of you; you see the matter and the whole of the business. From your letters --- both those you wrote together with others and those in your own name --- I understood (and was seeing of my own accord) that you, almost paralysed by the suddenness of the thing, are casting about for new ways of protecting me.

\ciceroSection{2} You write that you think I should come closer, and travel by night through the towns. I do not really see how that can be done. I have no lodgings so suitable that I could pass the whole daylight hours inside them, and, for the purpose you are after, it makes little difference whether people see me in a town or on the road. Still, I will consider this as I do the rest, in the way it seems it can most conveniently be managed.

\ciceroSection{3} Because of incredible distress of both mind and body I could not complete more letters; I have only written back to those from whom I had received letters. I should be grateful if you would write, in my name, to Basilus and to anyone else you think fit --- to Servilius as well --- as seems best to you. That over so long an interval I have written nothing at all to your circle, you will surely understand from this letter is for want of matter to write of, not for want of will.

\ciceroSection{4} As for your asking about Vatinius --- neither his good offices nor anyone else's would be wanting to me, if only they could find some matter in which to help me. Quintus was in a thoroughly hostile frame of mind toward me at Patrae. His son joined him there from Corcyra. From there I suppose they set out, along with the rest.
