% Ad Atticum 8.11 (ad-atticum-08-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} That you think me thrown into a great agitation of mind --- I am, certainly, but not so great as I perhaps seem to you. For all care is eased when either the deliberation has settled or thinking unravels nothing further. One may lament this matter, of course, whole days at a time; but I am afraid that, when I make no progress by it, I shall also bring discredit on my literary studies. So I am spending the whole of my time considering how great is the power of that man whom in my books I have, as it seems to you at any rate, drawn out with sufficient care. Do you hold in mind, then, that ruler of the commonwealth to whom we wished all things to be referred? For so in the fifth book, I think, Scipio speaks: as for the helmsman the favourable course, for the physician health, for the general victory, so for this ruler of the commonwealth the prosperous life of the citizens is the end set before him --- that they should be firm in resources, rich in stores, ample in glory, honourable in virtue. It is of this work, the greatest among men and the best, that I wish him to be the bringer-to-completion.

\ciceroSection{2} Our Cnaeus has never given this a thought before, and least of all in the present cause. Mastery has been the object of both men; the aim has not been that the state should be prosperous and honourable. Nor in fact did he leave the City because he could not defend it, nor Italy because he was being driven from it, but from the first he had this in mind --- to set every land, every sea in motion, to stir up barbarian kings, to bring savage tribes into Italy under arms, to assemble the largest of armies. That kind of Sullan kingship has long been the object of many who are with him, and who want it. Do you suppose that no understanding could be reached between them, no compact made? It can be, today. But for neither of them is the goal \textit{[Greek: skopos]} that we should be prosperous; each of them wants to reign.

\ciceroSection{3} I have set this out briefly because you invited me to. You wanted me to show what I felt about these evils. So I prophesy \textit{[Greek: prothespizō]}, my dear Atticus --- not divining like that woman whom no one believed, but looking ahead by reasoned conjecture: ``and now upon the great sea'' --- by no great margin, I tell you, can I miss in my augury. So great an Iliad of evils \textit{[Greek: Ilias]} hangs over us. And our case, who are at home, is the more burdensome than that of those who crossed over with him by this, that they fear one man, we both.

\ciceroSection{4} ``Then why,'' you say, ``did we stay?'' Either because we obeyed you, or because we did not run out to meet him, or because this was the more right course. You will see, I tell you, wretched Italy trodden underfoot next summer by slave conscripts gathered from every type, and \textit{[the levy]} of which there is so much talk at Luceria is not so much to be feared as the universal ruin. So great I see will be the forces of both men in the colliding. There you have my reckoning. You, perhaps, were expecting some bit of consolation. I find nothing. Nothing more wretched than this can come about, nothing more ruined, nothing fouler.

\ciceroSection{5} As to what you ask --- what Caesar has written to me: the usual thing, that it is most gratifying to him that I have kept quiet, and he begs that I persist in this. The younger Balbus had the same charge. His journey, in fact, was to the consul Lentulus with letters from Caesar and the promise of rewards if he should return to Rome. But when I reckon the days, I think he will cross the sea before he can be met with.

\ciceroSection{6} I wanted you to be aware of the carelessness of Pompey's two letters which he sent to me, and of my own care in the answer. I have sent you copies of them.

\ciceroSection{7} What Caesar's run through Apulia here to Brundisium will produce, I await. Would it were something like the Parthian affair! As soon as I have heard anything, I shall write to you. You, please, write me what the loyal men are saying: their talk is said to be widespread at Rome. I know, indeed, that you do not go out in public; but still you must be hearing many things. I recall that a book by Demetrius of Magnesia was brought to you --- the one I know was sent to you, ``On'' \textit{[Greek: peri]} \textit{[concord]}. I should like you to send it on to me. You see what cause it is I am meditating.
