% Ad Atticum 8.16 (ad-atticum-08-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Everything has been provided for me except a hidden and safe road to the Upper Sea. For this sea here we cannot use at this time of year. To the other --- where my mind looks, and where the matter calls --- by what way shall I come? For one must withdraw quickly, lest by some chance I be hindered and bound fast. And it is not that man who leads me who seems to lead me; I had recognised him long since as the least political of all men \textit{[Greek: apolitikotaton]}, but now also the most lacking in generalship \textit{[Greek: astrategetotaton]}. It is not he, then, who leads me, but the talk of men, which is written to me by Philotimus. He says that I am being torn to pieces by the \textit{optimates}. By what \textit{optimates}, good gods! Who now --- how they hurry to meet him, how they peddle themselves to Caesar! The country towns indeed take him for a god; and they are not pretending, as they were when they were making vows for him in his sickness.

\ciceroSection{2} But plainly, whatever evil this Peisistratus shall not have done will be as welcome as if he had stopped another from doing it. Him they hope to find favourable; the other they think angry. What greetings \textit{[Greek: apanteseis]} do you suppose are being made from the towns, what honours! ``They are afraid,'' you will say. I believe it; but, by Hercules, of the other more. They are delighted by this man's insidious clemency; the other's anger they dread. The jurors out of the three hundred and sixty who took special pleasure in our friend Gnaeus, of whom I see one or another every day, shudder at certain ``Lucerias'' of his. And so I ask --- who are those \textit{optimates} who would drive me out, while they themselves remain at home? But still, whatever they are, I feel shame before them \textit{[Greek: aideomai]}. Although in what hope I should set out, I see; and I am joining myself with a man more prepared to ravage Italy than to win the war, and I am waiting for a master. And indeed, while I was writing these things, on the fourth day before the Nones, I was by now expecting something from Brundisium. But what something? How shamefully the one had fled from there, and by what route the victor here would bring himself back, and where to. When I had heard that, if he was coming by the Appian Way, I was thinking of Arpinum.
