% Ad Familiares 8.16 (ad-familiares-08-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} Stunned by your letter --- in which you showed yourself to be considering nothing but a grim course, and did not write out plainly what it was, and yet did not fail to give away the quality of what you were considering --- I have written this back to you on the spot. By your fortunes, Cicero, by your children I beg and implore you: do not take any graver counsel about your safety and your standing. For I call gods and men and our friendship to witness that I told you ahead of time, and warned you not idly but, after I had met Caesar and learned what his mind would be once victory had been won, made sure you knew. If you suppose Caesar's policy is going to be the same in releasing his adversaries and in offering them terms, you are mistaken: he thinks --- and even speaks --- of nothing but harshness and savagery. He came out of Rome angry at the Senate; by those vetoes he has been plainly inflamed; by Hercules there will be no place for an appeal.

\ciceroSection{2} So, if you matter to yourself, if your only son matters, if your house, if the hopes that are left to you are dear to you, if I count for anything with you, if your son-in-law --- the excellent man --- carries any weight (and you ought not to wish to upend the fortune of these, so as to force us either to hate or to abandon the cause in whose victory our safety lies, or to harbour an impious longing against your own safety) --- and finally consider this: what offence you have already incurred by that hesitation of yours. To act now against a victorious Caesar, whom you would not injure when his cause was in doubt, and to fall in with men in flight, whom you would not follow when they were standing their ground --- that is the height of folly. See to it, while you are ashamed of being too little of an optimate, that you do not choose, too little carefully, what is best.

\ciceroSection{3} But if I cannot persuade you of the whole of this, at least wait until we have learned what we are doing about the Spains --- which I report to you will be ours on Caesar's arrival. What hope those men have once the Spains are lost, I do not know; and what your plan can possibly be --- to fall in with the desperate --- by my faith I cannot make out.

\ciceroSection{4} What you indicated to me by not saying it, Caesar had heard, and as soon as he said ``hail'' to me, he laid out at once what he had heard about you. I said I knew nothing; but I asked him all the same to send you a letter that might above all move you to remain. He is taking me with him to Spain; for had he not, before I came near the city, wherever you were, I would have run to you, and pressed this on you in person, and held you back with all my force.

\ciceroSection{5} Again and again, Cicero, consider --- do not utterly overturn yourself and all of yours, do not knowingly and deliberately send yourself down where you can see there is no way out. If either the cries of the optimates move you, or you cannot bear the insolence and braggadocio of certain men, then choose, I advise, some town free of the war, while these matters are being settled --- and they will soon be settled. If you do that, I shall judge that you have acted wisely, and you will not give Caesar offence.
