% Ad Atticum 10.15 (ad-atticum-10-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} While Servius was with me, Cephalio arrived with your letter on the sixth day before the Ides; it brought us great hope of better things touching the eight cohorts. For even those that are in these parts are said to be wavering. On the same day Funisulanus delivered a letter from you which confirmed the same news more firmly. I gave him a fully favourable answer about his business, with every expression of obligation to you. So far he was not making good; he owes me a great deal of money and is not held to be solvent. Now he says he will pay; but he is delaying the man to whom he had charged the debt; $\dagger$ \textit{tabellarius si apud te esse quas satis fecissest dares} $\dagger$. How much it amounts to, Eros, the agent of Philotimus, will tell you. But let us return to greater matters.

\ciceroSection{2} What you hope for, the affair in Caelius's letter, is ripening. So I am tormented over whether I should wait for a fair wind. A standard is what is needed; they will come flying to it. As to your urging that I go openly, I agree entirely, and so I think I shall set out. Meanwhile, however, I am waiting for your letter. Nothing is settled by Servius's counsel. Every snare in every proposition comes up against him. The one man I have known to be more timid still is Gaius Marcellus, who regrets having been consul. Oh, the depths of ignobility! \textit{[Greek: \=o poll\=es agenneias]} --- the man who is even said to have strengthened Antony in his resolve to obstruct me, all the more honourably for himself, I suppose.

\ciceroSection{3} Antony for his part set out for Capua on the sixth day before the Ides. He sent word to me that he had been deterred by embarrassment from coming to me himself, since he supposed me angry with him. I shall go, then, and indeed just as you advise, unless some prospect of a more weighty role to be taken on is offered me beforehand. But that will hardly come so soon. The praetor Allienus, however, thought that someone --- if not I, then one of his colleagues --- would do. Anyone will serve, so long as it is somebody.

\ciceroSection{4} About your sister, I commend her. About young Quintus, effort is being made; I hope things are looking up. About your brother Quintus, you should know he is in no small trouble over the loan transfer, but so far he has got nothing out of Lucius Egnatius. Axius, about the twelve thousand --- the shamelessness! For he has often written to me to pay Gallius as much as he wished. And if he had not written, could I have done otherwise? I have indeed often promised; only he wanted as much, and quickly. Let them help me, indeed, in these straits! But may the gods take care of such fellows! Another time on this. I am glad you are free of the quartan, and Pilia too. I, while the bread and other provisions are being put aboard ship, am running over to my Pompeian villa. I should be grateful if you would thank Vettienus, who is being attentive; if you can find anyone to carry it, send me a letter before we set sail.
