% Ad Atticum 8.1 (ad-atticum-08-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} After I had sent off my letter to you, a letter was delivered to me from Pompey: as to the rest, on the affairs in Picenum that Vibullius had written to him, on Domitius's levy, things which are known to you and which were yet not so cheerful in his letter as Philotimus had written to me. I should have sent you the letter itself, but my brother's boy was just on the point of setting off. I shall send it, then, tomorrow. But in that letter of Pompey's there was, at the end, in his own hand: ``you, I think, should come to Luceria. Nowhere will you be safer.'' I took this in the sense that he is treating these towns and the seacoast as good as abandoned, and I was not surprised that a man who had given up the head should not spare the remaining limbs.

\ciceroSection{2} I wrote back to him at once, and sent a trusted man from my own retinue --- not failing to enquire where I should be safest; that if he wanted me, for his own sake or for the commonwealth's, to come to Luceria, I would come at once; and I urged him to hold the seacoast, if he wanted his grain supply to be furnished him from the provinces. This I saw I was writing in vain; but as on the holding of the city then, so now on Italy's not being abandoned, I was placing on record my own opinion. For I see the preparations being made are these: that all the forces are to be drawn together at Luceria, and that not even that place will be a stable one, but from there itself, if we are pressed, flight will be prepared.

\ciceroSection{3} The less, then, should you wonder if I am going down unwillingly into a cause in which no calculation of peace or of victory has ever been sought, but always one of disgraceful and ruinous flight: I must go, that I may meet whatever event chance brings me with men who are called loyal, rather than be thought to dissent from the loyalists. And yet I see that very soon the city will be packed with loyalists --- that is, with the well-dressed and the well-to-do --- and, with these country towns abandoned, packed to the rafters. I should be of their number, if I did not have these most vexatious lictors, and if I were not ashamed of having M.~Lepidus, L.~Volcacius, and Ser.~Sulpicius as my companions --- of whom none is either more foolish than L.~Domitius or more inconstant than Ap.~Claudius.

\ciceroSection{4} Pompey alone moves me --- by past favour, not by authority. For what authority can he have in this cause? --- a man who, while all the rest of us were afraid of Caesar, was himself fond of him, and now, the moment he himself has begun to be afraid, thinks everyone else ought to be Caesar's enemy. We shall go to Luceria nonetheless. Nor will my arrival, perhaps, give him any pleasure; for I shall not be able to hide that what has been done up to now displeases me. If I could only get any sleep, I should not be wearying you with such long letters. If your own case is the same, you might very well pay me back in kind.
