% Ad Atticum 9.3 (ad-atticum-09-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Domitius's son passed through Formiae on the eighth day before the Ides, hurrying to his mother at Naples, and ordered word to be brought to me that his father was near the City --- when our slave Dionysius pressed him with questions about him. I, however, had heard that the man had set out either to Pompey or to Spain. I should very much like to know how this stands; for it bears on what I am pondering. If he certainly has gone nowhere, I understand from this that Gnaeus considers our exit from Italy not an easy one --- since all Italy is held by arms and garrisons, and this is winter, on top of everything. For if the season of the year were more accommodating, there would even be the Lower Sea to use; as it is, nothing can be crossed but the Upper, and the road to it is cut off. You will inquire, then, both about Domitius and about Lentulus.

\ciceroSection{2} From Brundisium no rumour had yet come, and today was the seventh day before the Ides, the day on which I suspected Caesar had reached Brundisium, or the day before. For he had stopped at Arpi on the Kalends. But if you cared to listen to Postumus, he was about to follow Gnaeus hard at his heels --- for he thought he had already crossed, judging from the weather and the count of days. I, for my part, did not think he would have the sailors; he was confident, and the more so because the man's generosity had reached the ears of the ship-owners. But on the whole business of Brundisium, how it stands --- I cannot go on much longer in the dark.
