% Ad Atticum 7.14 (ad-atticum-07-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the sixth day before the Kalends of February, setting out from Cales for Capua --- I had a slight inflammation of the eye --- I gave this letter. Lucius Caesar brought Caesar's terms to Pompey on the eighth day before the Kalends, while he was with the consuls at Teanum. The offer was approved, on this condition: that he should withdraw his garrisons from those towns which he has occupied outside his own province. If he does so, the answer was, we shall return to the City and finish the matter through the senate. I hope we can keep the peace for the present; for both that man is somewhat regretting his frenzy and our own is somewhat regretting his forces.

\ciceroSection{2} Pompey wanted me to come to Capua and help with the levy; in which the colonists of Campania are responding with no great readiness. As for Caesar's gladiators at Capua --- about whom I had earlier written to you a false report out of Aulus Torquatus's letter --- Pompey has dispersed them very sensibly, two apiece among heads of households. There were five thousand shields in the school. They were said to have been about to break out. Real provision has been made there for the commonwealth.

\ciceroSection{3} About our womenfolk, your sister among them, I beg you to see whether it is decent enough for us that they should be at Rome, when the others of that rank have left. I have written to them and to you yourself before. I should like you to urge them to leave --- the more so since we have country properties on the seaboard, in the district of which I am in charge, where, as things now are, they can be not uncomfortably. For if we have given any offence in our son-in-law (which indeed I am not bound to be answerable for) --- but it grows the greater because our women, beyond the rest, have stayed at Rome. You yourself, with Sextus, please let me know what you are thinking about leaving, and what you make of the whole business. For my part, I do not cease to urge towards peace; which even unjust is more useful than the most just war with citizens. But these things as chance shall bring them.
