% Ad Atticum 7.22 (ad-atticum-07-22) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I see that there is not a foot of ground in Italy that is not in that man's power. About Pompey I know nothing, and unless he has gotten himself aboard ship I think he will be picked off. What incredible speed! As for our friend --- but I cannot accuse without grief the man over whom I am tormented and racked. As for the bloodshed, you fear it not without reason --- not that anything would be less in Caesar's interest for the durability of his victory and his domination, but I can see at whose direction he is going to act.

\ciceroSection{2} I think the right course is to give way; about whether to leave the towns I need counsel. Whatever seems to you the best thing to do, you will do. Speak with Philotimus; and besides, you will have Terentia with you by the Ides. As for me --- what am I to do? By what road, by what sea am I to follow a man whose whereabouts I do not know? Even by land --- how can I? By sea --- to where? Am I, then, to hand myself over to that one? Granted that it could be done safely (many are urging it), can it also be done honourably? In no way. For my part I shall ask your counsel, as I always do. The matter cannot be untangled; but even so, if anything comes to mind, I should be glad if you would write --- and write what you yourself are going to do.
