% Ad Atticum 9.5 (ad-atticum-09-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On your birthday you wrote me a letter full of counsel and of the highest both good will and prudence. Philotimus delivered it to me the day after he received it from you. The matters you discuss are indeed most difficult: a journey to the Upper Sea, a voyage by the Lower; a withdrawal to Arpinum, lest I should seem to have fled this man; a stay at Formiae, lest we should seem to have offered ourselves to his felicitations --- but nothing more wretched than to see the things that, even so, I tell you, will now have to be seen. Postumus was with me; I have written to you how heavy he was. Quintus Fufius too came to me --- with what a face, with what a spirit! Hurrying to Brundisium, denouncing Pompey's crime, denouncing the levity and stupidity of the Senate. If I cannot bear these things in my own villa, am I to bear Curtius in the Senate-house?

\ciceroSection{2} Come now, suppose me to bear all this with however strong a stomach \textit{[Greek: eustomachos]} --- still, what of the question, ``Tell us, Marcus Tullius: what end will these things have?'' And I leave aside the cause of the commonwealth, which I count as lost both through its own wounds and through the remedies that are being prepared. What am I to do about Pompey? With whom (for why should I deny it?) I have been plainly angry. For the causes of events have always moved me more than the events themselves. Considering, then --- or rather judging --- that these evils (and what can be greater?) have come about through his agency and his fault, I was more hostile to him than to Caesar himself. As our ancestors held the day of the battle of the Allia a more mournful day than the day of the City's capture --- because that disaster proceeded from this one (so that the one day is even now religiously observed and the other is unknown to the common people) --- so I, calling to mind the wrongs of ten years (among which was the year that, with this man not defending us, broke us down --- to put it no more sharply) and recognising the temerity, the cowardice, and the negligence of the present time, was angry with him.

\ciceroSection{3} But these things are now out of my mind; his benefits to me I think on, I think also on his dignity. Later than I could wish, I now understand, because of Balbus's letters and conversations, but I see plainly that nothing else is being aimed at --- nothing has been aimed at from the beginning --- but to do this man to death. Therefore, as that hero in Homer to whom both his mother and a goddess had said, ``At once, then, after Hector your doom is ready'' \textit{[Greek: autika gar toi epeita meth' Hektora potmos hetoimos]}, himself answered his mother, ``Let me die at once, since I was not, after all, to have saved my comrade from being killed'' \textit{[Greek: autika tethnaien, epei ouk ar' emellon hetairoi kteinomenoi epamunai]} --- what if not only ``comrade'' \textit{[Greek: hetairoi]} but ``benefactor'' \textit{[Greek: euergetei]}? add, such a man, conducting such a cause? Such duties, indeed, I count worth purchasing with life itself. As for those ``best men'' of yours, I have no confidence in them; I am not even any longer at their service.

\ciceroSection{4} I see how they give themselves to this one, how they are going to give themselves. Do you suppose those municipal decrees about his health were anything in comparison with these felicitations on his victory? They are afraid, you will say. But they themselves say that they were afraid then too. But let us see what is done at Brundisium. From that, perhaps, new \textit{[Greek: nea]} counsels will be born, and other letters.
