% Ad Atticum 9.6 (ad-atticum-09-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} From Brundisium, so far, we have nothing. From Rome, Balbus has written that he believes Lentulus the consul has now crossed over, and that he was not intercepted by the younger Balbus, because he had already heard this at Canusium; that from there he wrote to him. The six cohorts that had been at Alba have gone over to Curius by the Via Minucia: Caesar wrote this to him, and in a short time he himself will be in the neighborhood of the City. Therefore I shall make use of your advice and not hide myself at Arpinum at this time --- although, since I wanted to give my dear young Cicero the toga of manhood at Arpinum, I had been going to leave precisely this excuse to Caesar. But perhaps on that very ground he will be offended: why not rather at Rome? And yet, if he must be met, here is the best place. Then we shall see to the rest: that is, to where, and by what road, and when.

\ciceroSection{2} Domitius, I hear, is in the territory of Cosa, and indeed, they say, ready to sail. If for Spain, I do not approve; if to join Cnaeus, I praise it. Anywhere, certainly, rather than that he should look upon Curtius --- whom I, his patron, cannot look at. What of the others? But, I think, let us keep quiet, that we may not bring evidence against our own fault, who, while we loved the City, that is to say our country, and while we were thinking the matter would be patched up, have so conducted ourselves as to be plainly cut off and taken captive.

\ciceroSection{3} After the letter was already written, a letter was brought from Capua of the following content: ``Pompey has crossed the sea with all the soldiers he had with him. The number of men is thirty thousand, and the two consuls, and the tribunes of the people, and the senators who were with him, all of them, with their wives and children. He is said to have embarked on the fourth day before the Nones of March. From that day there have been north winds. The ships he did not use they are said either to have cut to pieces or to have burned.'' On this matter a letter has been brought to Capua to Lucius Metellus, tribune of the people, by Clodia his mother-in-law, who herself crossed.

\ciceroSection{4} Before, I was anxious and ground down with care, as the matter itself, of course, compelled, because I could unfold nothing by deliberation; but now, since Pompey and the consuls have gone out of Italy, I am not ground down: I am on fire with grief. ``Nor is my heart steadfast, but I am distraught'' \textit{[Greek: oude moi etor empedon, all' alaluktemai]}. I am not, I tell you, believe me, in possession of my senses; so great a disgrace I seem to myself to have let in. That I should not have gone first with Pompey, whatever his counsel was; and then, with the loyal men, however rashly the cause was begun? Especially when the very persons for whose sake I was committing myself to fortune more timidly --- my wife, my daughter, the young Ciceros --- preferred my following that course, and thought this one shameful and unworthy of me. As for my brother Quintus, he kept saying that whatever pleased me he thought right, and was following it with a mind entirely composed.

\ciceroSection{5} Your letters I now read from the first. They restore me a little. The earliest warn me and beg me not to throw myself away; the most recent show that you rejoice that I have stayed. When I read them, I seem less shameful to myself --- but only so long as I am reading. Then the grief surges up again, and the sense of disgrace \textit{[Greek: aischrou]}. For which reason I beseech you, my Titus, snatch this grief from me, or at least lessen it, by consolation, by counsel, by whatever means you can. But what could you do? Or what could any human being? Scarcely a god, now.

\ciceroSection{6} For my part, I am working at the thing you advise and hope can be done: that Caesar may grant me to be absent when anything against Cnaeus is brought on in the Senate. But I fear I shall not obtain it. Furnius has come from him. And, that you may know what sort of people we are following: he reports that the son of Quintus Titinius is with Caesar --- but that Caesar gives me greater thanks than I should wish. What he asks of me, in his few words but with force \textit{[Greek: en dunamei]} --- learn from his own letter. Wretched me, that you were not well! We would have been together; counsel, at any rate, would not have failed me: ``the two going on together'' \textit{[Greek: sun te du' erchomeno]}\ldots.

\ciceroSection{7} But let us not do over what is done; let us prepare for what is left. These two things have deceived me up to now: first, the hope of a composition, which once made, I wanted to use to live a private life, to free our old age from anxiety; then, I came to understand that the war undertaken by Pompey was cruel and destructive. I thought it the part, by Hercules, of a better citizen and man to be subjected to any punishment than not only to preside over but even to take part in that cruelty. It seems even to have been better to die than to be with these men. To these matters, then, give thought, my Atticus, or rather think them through. Any outcome whatever I shall bear more bravely than this grief.
