% Ad Atticum 8.9 (ad-atticum-08-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} That you write my letter has been put about in public, I do not take ill --- indeed, I myself have given copies to many. For things have already happened, and others are hanging over us, such that I want it on the record where I have stood on the question of peace. And when I was urging peace on him --- a man of his particular sort --- I saw no easier way to move him than by saying that what I was urging would befit his wisdom. If I called that wisdom admirable when I was urging him to save his country, I have no fear of being thought a flatterer toward a man at whose feet, in such a cause, I would gladly have thrown myself. As for the other matter, if you have any time to give it, what I want him to think about is not peace but myself and my own duty. For when I bear witness that I have been no part of the war, even though that has been clear from the facts, I have nevertheless written it so that I should carry more authority in giving advice; and the same is the point of my approving his cause.

\ciceroSection{2} But why all this now? If only something had been achieved by it! I should be willing to have that letter of mine read out at a public assembly --- granted that the man himself, writing back to him, posted up in public the letter in which stands the phrase ``for your most ample achievements in office'' (more ample than his own, than Africanus's? so the moment required it); and granted that you two, gentlemen of your standing, went out to the fifth milestone to a man who is just now taking himself back into the City --- what is he doing, what about to do? --- and not only went out in numbers, but congratulated him with a glad face. Are we, then, doing wrong? You yourselves, by no means; and yet the signs by which true will could be told from pretence are getting muddled. As to the senate's decrees that I see being passed --- but I am opening up more than I had meant to.

\ciceroSection{3} I want to be at Arpinum on the day before the Kalends, and then to wander round our little farms, which I had given up expecting ever to see again. Your noble \textit{[Greek: eugenē]} counsels --- and, given the times, not incautious for all that --- have my warm approval. Lepidus, indeed (for he and I generally pass whole days together \textit{[Greek: syndiēmereuomen]}, which delights him), has never thought leaving Italy a good plan; Tullus much less still --- for his letters keep arriving by way of others. But the opinions of those men moved me less; they had given the republic fewer pledges. Yours, by Hercules, weighs powerfully on me; for it carries with it both a way of recovering the time still left and of safeguarding the present. But, I beg you --- what could be more wretched than this: one man hunting applause in the foulest of causes, the other earning offence in the best? one to be reckoned the preserver of his enemies, the other the deserter of his friends? And, by Hercules, much as we love our Cnaeus --- as we both do, and ought to --- this one thing I cannot praise, that he is not coming to the help of such men. For if he was afraid, what more cowardly? If, as some think, he reckoned his cause would be the better for their slaughter, what more unjust? But let us drop this; we only make the pain worse by going over it.

\ciceroSection{4} On the sixth day before the Kalends, in the evening, the younger Balbus came to me, hurrying by a back road to the consul Lentulus --- sent by Caesar with a letter, with instructions, with the promise of a province, to bring him back to Rome. I do not think he can be persuaded, unless there is a meeting. The same man said that Caesar wants nothing more than to catch up with Pompey --- this I believe --- and be reconciled. That I do not believe; and I am afraid that all this clemency is being heaped up toward the Cinnan kind of cruelty. The elder Balbus, for his part, writes to me that Caesar wants nothing so much as to live without fear under Pompey as first man. You, I take it, believe this. But while I am writing these lines on the fifth day before the Kalends, Pompey might already have reached Brundisium: travelling light, he had gone ahead of his legions on the eleventh day before the Kalends from Luceria. Still, this prodigy \textit{[Greek: teras]} acts with terrifying watchfulness, speed, and energy. What will come of it I plainly do not know.
