% Ad Atticum 10.5 (ad-atticum-10-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About my whole train of thought I wrote to you earlier, carefully enough, as it seemed to me. About the day, nothing certain can possibly be written, save this: not before the new moon.

\ciceroSection{2} Curio's conversation on the next day came to roughly the same total, except that he indicated more openly that he does not see the outcome of these affairs. As for what you charge me with about managing Quintus --- you are asking Arcadia of me \textit{[Greek: Arkadian]}. Still, I shall pass over nothing. Would that you --- but I will not be more troublesome. The letter to Vestorius I delivered at once; he kept asking after it very pressingly.

\ciceroSection{3} Vettienus spoke to you more handsomely than he had written to me. But I cannot wonder enough at the man's negligence. For when Philotimus had told me he could buy that lodging from Canuleius for fifty thousand sesterces, and could buy it even for less if I asked Vettienus, I asked him to take whatever he could off the sum. He promised. Recently he wrote that he had bought it for thirty thousand, and asked to whom I wished the property assigned; the day for payment was to be the Ides of November. I wrote back to him rather testily, but still with the banter of an old friend. Now, since he is dealing generously, I make no complaint against the man, and have written to him that I have been informed by you. As to your own journey --- what you have in mind and when --- please let me know. The fifteenth day before the Kalends of May.
