% Ad Atticum 8.7 (ad-atticum-08-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} One thing still remains to crown our friend's total disgrace: that he not come to Domitius's relief. And yet no one doubts that he will come to the rescue. I do not think so. Will he, then, abandon such a citizen, and the men you know to be with him, when he himself has thirty cohorts at hand? Unless everything deceives me, he will abandon them. He has lost his nerve to an incredible degree; he looks to nothing but flight.

\ciceroSection{2} And it is this man --- for I see what you are thinking --- whom you suppose I ought to accompany. I, for my part, have one to flee from; one to follow I do not have. As for that saying of mine you praise and call memorable --- that I would rather be conquered with Pompey than conquer with those men --- yes, I would prefer it; but with that Pompey who then was, or who seemed to me to be; with this Pompey, who flees before he knows whom he flees or where, who has handed over our position, who has left his country, who is leaving Italy --- if that was my preference, it has come to pass: I have been conquered. As for what remains, I cannot bear the sight of the things I never feared I should see; nor, by Hercules, the sight of this man for whose sake I must do without not only what is mine, but my very self.

\ciceroSection{3} To Philotimus I have written about travel funds --- either from Moneta (for nobody is paying) or from those Oppii who lodge with you. The rest, as the occasions arise, I shall lay upon you.
