% Ad Atticum 9.18 (ad-atticum-09-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Both as you advised. My way of speaking was such that he should think well of me rather than be grateful to me, and I held to the point of not going up to the City. We were mistaken in supposing him easy: I have seen no one less so. He said he was condemned by my judgement, and that the rest would be the slower if I did not come. I said his case was a different one. Then, after much else, ``Come, then,'' he said, ``and act for peace.'' ``On my own judgement?'' said I. ``Or am I,'' he answered, ``to prescribe to you?'' ``This,'' I said, ``is how I shall speak: it is not the Senate's pleasure that there be a crossing into the Spains, nor that armies be transported to Greece; and many things,'' I said, ``I shall lament on Cnaeus's behalf.'' Then he: ``I will not have such things said.'' ``So I supposed,'' I replied; ``and this is precisely why I do not want to be there: either I must speak in this way, and many other things which I could not by any means be silent on if I were present, or I must not come.'' The upshot was that, as if looking for a way out, he asked that I should deliberate. There was no refusing that. So we parted. I take it, then, that this man does not love me. But I have loved myself, which has not happened to me for a long while now.

\ciceroSection{2} The rest --- by the gods! What an entourage, what a \textit{[Greek: n\'ekuia]}, as you are accustomed to call it! And in it the \textit{[Greek: h\=er\=os]} Celer. What a ruined affair! What desperate forces! And then to think that Servius's son, and Titinius's son, were in those camps which had Pompey under siege. Six legions; he watches much, he dares. I see no end to the evil. Now, certainly, you must bring out your counsels. That had been the last thing in reserve.

\ciceroSection{3} That \textit{[Greek: katakleis]} of his, however, which I almost passed over, is hateful --- that if he were not allowed to make use of my counsels he would use the counsels of those whose he could, and would stoop to anything. So you have seen the man, as you had written? You groaned? Surely. The rest --- come, give it. What next? He goes straight on to Norbanus at the Pedian villa, I to Arpinum; from there I await, for my part, that \textit{[Greek: lalageusan]} chatter of a letter of yours. You will say I had rather you didn't go back over a thing already done. Yes --- but even that man we are following has been mistaken in many things.

\ciceroSection{4} But I am waiting for your letter. For now there is no longer any prospect that we, as before, will see where this is going to come out. The end of it was about our meeting; and in that, I have no doubt, I have given offence to the man. All the more reason for acting promptly. I beg you --- a letter, and a \textit{[Greek: politik\=en]} one! I am waiting now keenly for your letter.
