% Ad Atticum 8.3 (ad-atticum-08-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Thrown into confusion by matters of the gravest and most miserable kind, and not having the chance of deliberating with you face to face, I wanted nevertheless to make use of your judgement. The whole deliberation is this: if Pompey leaves Italy, which I suspect he is going to do, what do you think I should do? And so that you can the more easily give me advice, I shall set out briefly what comes into my mind on either side of the question.

\ciceroSection{2} What draws me, on the one hand, is Pompey's signal services on behalf of my own safety, and the close friendship I have with him; and beyond that the cause of the Republic itself; so that it seems to me my course of action must be joined to his course of action, and my fortune to his fortune. There is this further consideration. If I stay, and abandon the company of all those most loyal and most distinguished citizens, I have to fall under the power of one man. He, it is true, gives signs in many ways that he is a friend to me --- and that he should be so I took care for, well in advance, as you well know, on suspicion of the storm now hanging over us; nevertheless two things have to be weighed: how much trust can be reposed in him, and whether, even granting it were perfectly certain that he would be a friend to us, it is the part of a brave man and a good citizen to remain in that city in which he has held the highest offices and commands, has accomplished the greatest things, has been invested with the most distinguished priesthood, but will be none of those things which he has been --- and to take on the risk of some attendant dishonour, should the day ever come when Pompey recovers the Republic.

\ciceroSection{3} On this side of the matter, then, these are the considerations. Look now at what stands on the other side. Nothing has been done by our Pompey wisely, nothing bravely; I add: nothing except against my advice and my authority. I leave aside the old chapters --- that it was he who nursed up Caesar against the Republic, who built him up, who armed him; he the sponsor of laws carried by violence and against the auspices; he the annexer of Further Gaul; he the father-in-law; he the augur at the adoption of P. Clodius; he keener on bringing me back than on never letting me be expelled; he the extender of Caesar's province; he his abettor in everything during his absence; he too, in his own third consulship, when he had at last begun to act the defender of the Republic, who exerted himself to have ten tribunes of the plebs propose that Caesar's candidacy be considered in his absence --- a point he also ratified by a law of his own --- and who, on the Kalends of March, blocked the consul Marcus Marcellus when Marcellus was moving to set a term to Caesar's command of the Gauls. But, leaving those things aside, what could be more shameful, what more disordered, than this departure from the city --- or rather, this most disgraceful flight? What terms would not have been worth accepting, rather than abandoning one's country?

\ciceroSection{4} The terms were bad, I grant; but was anything worse than this? ``But he will recover the Republic.'' When? Or what has been put in readiness toward that hope? Has not the Picene country been lost? Has not the road to the city been thrown open? Has not all the money, both public and private, been handed over to the adversary? In short there is no cause, no force, no place to which any who wish the Republic defended can muster. Apulia has been chosen, the emptiest part of Italy and the one furthest removed from the brunt of this war; it looks like flight, and access to the sea, sought out of despair. I took Capua unwillingly, not that I would shirk that duty, but in a cause in which there was no resentment from any order, no open resentment from private citizens, some from the loyal but blunted, as such resentment commonly is; and, as I myself perceived, the mass of people and every man at the bottom of the heap was leaning to the other side, and many were hungry for an overturn of affairs --- I told the man himself I would take nothing on without a garrison and without money. So I had nothing whatever to do, because I saw from the start that nothing was being looked for but flight. If I follow that flight now, where to? Not with him: when I had set out to join him, I found out that Caesar was in those parts, so that I could not safely reach Luceria. We should have to sail the Lower Sea, on an uncertain course, in the dead of winter. Come now: with my brother, or without him and with my son? But how? In either alternative there will be the greatest difficulty and the greatest pain of mind. And what an onslaught there will be from Caesar against us when we are gone, and against our property! sharper than against the rest, because he will think perhaps that in doing violence to us he has something which will play well with the people. Come now, these chains of mine, these laurelled fasces of mine I mean, how troublesome to carry them out of Italy! And what place will be safe for us, even granting that the waves are calmed for our use, until we have come to Pompey? And by what route, or to where, we have no idea.

\ciceroSection{6} But if I stay behind, and there proves to be a place for me on this side, I shall have done what L. Philippus did in the supremacy of Cinna, what L. Flaccus did, what Q. Mucius did --- however that course of action in the end fell out for the last of them; he used to say even so that he saw that what did happen would happen, but preferred this to coming armed against the walls of his country. Thrasybulus did otherwise, and perhaps better. But there is a settled rationale and view in that course of Mucius's; there is also one in that of Philippus; and one must serve the time, and one must not let the time slip when it is given. But on this very point those same fasces of mine bring their own embarrassment. Suppose he is a friend to me --- which is uncertain, but suppose it: he will offer a triumph. Not to accept may be dangerous; to accept would be invidious in the eyes of the loyalists. ``What a difficult and inextricable matter!'' you say. And yet it must be extricated. For what else can be done? And lest you suppose me more inclined to staying because I have used more words on that side --- it may be, as happens in many investigations, that this side is the more voluble and that the more true. So, as one deliberating with even temper on a matter of the highest moment, please give me your counsel. A ship has been got ready for us both at Caieta and at Brundisium.

\ciceroSection{7} But look --- news, while I am writing this very letter by night here at Cales; look --- a despatch: Caesar at Corfinium, Domitius at Corfinium with a firm army and eager to fight. I do not think our Gnaeus will so far commit himself as to leave Domitius in the lurch --- although he had sent Scipio ahead to Brundisium with two cohorts, and had written to the consuls that the legion which Faustus had levied for him should, by his preference, be led to Sicily by the consul. But to desert Domitius will be shameful when Domitius is calling for his help. I have a hope --- not, indeed, a great one, but in these parts a firm one --- that Afranius has fought Trebonius in the Pyrenees, that Trebonius has been driven off, and that your Fabius too has come over with cohorts; the upshot being that Afranius is approaching with large forces. If that is so, perhaps the war will be carried on in Italy. As for myself, since Caesar's route was uncertain --- he was thought likely to march either to Capua or to Luceria --- I sent Lepta to Pompey with a letter; I myself returned to Formiae so as not to fall into his hands somewhere. I wanted you to know all this, and I have written it with a steadier mind than I wrote with last time, putting in no judgement of my own but asking for yours.
