% Ad Familiares 5.19 (ad-familiares-05-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO RVFO.}

\ciceroSection{1} Though I have never had a doubt that I was very dear to you, I see it more plainly every day, and what you had given me to understand in certain of your letters is now coming out: that you would prove still more devoted in cultivating me than you had been in the province (though, in my judgement, nothing can be added to your service there), since now your judgement could be freer. Your earlier letter, then, gave me great pleasure --- one in which I could see how affectionately you had awaited my arrival, and how greatly, when the affair fell out otherwise than you had thought, you rejoiced at the counsel I had taken; and from this last letter I have taken a great profit, both from your judgement and from your sense of duty: of judgement, in that I see you think --- as all brave and honest men should --- that nothing is useful which is not also right and honourable; of duty, in that you promise to be at my side, whatever counsel I take. Nothing could be more welcome to me, nor, as I judge, more honourable for you.

\ciceroSection{2} My counsel has been taken now for some time; about which I have not written to you earlier --- not because you were to be kept in the dark, but because the sharing of a plan at such a moment looks something like a kind of admonition of duty, or rather a demand that one join in the partnership of the danger or the labour. But since this is your wish, your decency, your goodwill toward me, I gladly embrace such a feeling; only on this condition, though (for I will not lay aside my reserve in asking): if you do what you signify, I shall be greatly grateful; if you do not, I shall forgive, and I shall judge that fear has made you say one thing, and your inability to refuse me the other. For the matter is, in truth, the greatest. What is right is plain; what is expedient is dark --- though even so, in such a way that if we are the men we ought to be, that is, worthy of our studies and our writings, we cannot doubt that those courses are most useful which are most right. So, if you approve the same thing, come to me at once; if you approve the same conclusion and the same destination, but cannot come continuously, I will see to it that you know everything. Whatever you decide, I will judge you my friend; if what I hope for, I will judge you my closest friend.
