% Ad Atticum 9.19 (ad-atticum-09-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Since we are kept from Rome, I gave my Cicero his white toga at Arpinum, preferring it above any other place; and our townsfolk took it kindly. Yet I saw them, and everyone along my route, downcast and stricken. So grim, so savage, is the second look \textit{[Greek: anathe\=or\=esis]} one takes at this enormous evil. Levies are being held, men are being marched into winter quarters. Even when these things are done by honest men, in a just war, with moderation, they are still troublesome in themselves; what bitterness, then, do you suppose they have now, when they are done by ruined men, in a civil and unholy war, with the utmost insolence? And don't suppose, for that matter, that there is any unsavoury man in Italy who is not on this side. I saw them myself at Formiae, in a single mass --- and, by Hercules, I had never thought them men at all. I had known them all, but had never seen them gathered in one place.

\ciceroSection{2} Let us go on, then, where the plan calls, and leave all that is ours behind. Let us set out to one to whom our arrival will be more welcome than if we had been with him from the first. For then we were in the highest hope; now I, at any rate, am in none. No one save myself has quitted Italy, except those who had thought of this man as their enemy. And by Hercules I am not doing this for the commonwealth's sake --- which I judge to be utterly destroyed --- but to keep anyone from thinking me ungrateful to the man who relieved me from those troubles he had himself brought upon me; and at the same time because I cannot bear to see the things that are happening, or that will certainly come. By now, indeed, I think some decrees of the Senate have already been passed --- if only on Volcacius's motion! But what does it matter? For all the votes are one and the same. Servius, though, will be the most pitiless of them all --- he who sent his son to crush Cnaeus Pompeius, or at any rate take him captive, along with Pontius Titinianus. Though this one acted out of fear; but the father? But let us cease grinding our teeth, and at last be sensible that nothing is left to us except --- and this is what I should least have wished --- the breath in our bodies.

\ciceroSection{3} As for us, since the upper sea is blockaded, we shall sail by the lower; and if Puteoli prove difficult, we shall make for Croton, or for Thurii, and have the sea as our hostile country, we good citizens, lovers of our fatherland. I see no other way of carrying on this war. We shall take ourselves off into Egypt. We cannot match them in an army; for peace there is no faith.

\ciceroSection{4} But enough of these lamentations. Please give Cephalio a letter on every matter that has been done, and finally on what people are saying too, unless they have gone wholly silent. I have used your counsels, and chiefly in this --- that I held the gravity at our meeting that I owed myself, and that I held out against going up to the City. As for what remains, write me, I beg, as carefully as you can (we are at the last extremities now) what you think the right course is, what your judgement is; though by now there is no place left for doubt. Still, if anything --- or rather, whatever comes into your mind --- write me, please.
