% Ad Familiares 14.7 (ad-familiares-14-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS TERENTIAE SVAE S. P.}

\ciceroSection{1} All those troubles and anxieties, by which I kept you in the deepest misery --- this being what most distressed me --- and our little Tullia too, who is dearer to us than our own life, I have set down and thrown off. What had caused them I came to understand the day after I parted from you. In the night I brought up pure bile \textit{[Greek: chol\=en akraton]}; at once I felt such relief that it seemed to me some god had done the healing. To that god, then, please give thanks as you are accustomed to do, with piety and with purity --- I mean to Apollo and to Aesculapius.

\ciceroSection{2} The ship I trust we have is a very good one. The moment I came aboard her, I wrote these lines. Next I shall write a great many letters to our friends, in which I shall commend you and our little Tullia to them most earnestly. I would urge you to keep your spirits up, were it not that I know you already braver than any man. And yet I trust that things stand so that I may hope you are where it is most convenient for you to be, and that I shall in time, together with men of our own kind, defend the commonwealth.

\ciceroSection{3} Do you, above all, take care of your health; then, if you think best, use those country houses that will be furthest from the soldiers. The Arpinum estate you can use comfortably with the town household, if grain grows dearer. Our charming little Cicero sends you his warmest greetings. Once again, and again, farewell. Despatched on the seventh day before the Ides of June.
