% Ad Atticum 10.11 (ad-atticum-10-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} My last letter was already sealed when I decided not to give it to the man I had settled on, since he was an outsider; so it was not sent off that day. Meanwhile Philotimus arrived and delivered me a letter from you. What you write about my brother in it --- there is not much firmness in any of it, but there is nothing \textit{[Greek: hypoulon]}, nothing festering underneath, nothing two-faced, nothing that is not pliable toward good, nothing you could not, with a single conversation, bring round to wherever you please. In short: he holds all his own people dear, even those at whom he flares up rather often, and me dearer than he holds himself. As for his writing one thing to you about the boy and another to the mother about her son, I do not fault him for that. What you say about Quintus and about my sister is distressing, the more so because our circumstances are such that I cannot doctor any of it; I most certainly would doctor it; but you see in what evils and in what desperation of affairs we are.

\ciceroSection{2} Those matters about the money accounts are not of such a kind --- for I hear it often from the man himself --- that he does not wish to do well by you, and to take pains in it. But when in this flight of mine Quintus Axius is not repaying me the thirteen thousand sesterces I lent his son in cash, and pleads the times as his excuse, and when Lepta does the same, and the others, I sometimes wonder at hearing him say he is being pressed about some twenty thousand from I do not know whom. For surely you can see what straits we are in. Even so, he gives strict instructions that the business be attended to for you. Or do you suppose him slow or tight-fisted in that sort of thing? No man is less so.

\ciceroSection{3} About my brother, enough. About his son --- his father has always indulged him, certainly; but indulgence does not make a man a liar or a miser, or unloving to his own; it perhaps makes him fierce, arrogant, and aggressive. So the boy has his share of those traits as well that grow from being indulged; but they are tolerable --- what am I to say? --- in this generation of his. As for those other faults of his, which to me, who love him, are more wretched than these very evils we are caught in, they are not from any softness of mine. They have their own roots; which I would tear up, certainly, if I were allowed. But the times are such that I have to endure everything. My own boy I keep easily in hand; nothing is more manageable than he is. It is rather out of pity for him that the plans I have so far adopted are the more languid; and the more the boy wants me to be safe, the more I fear that I shall turn out cruel to him.

\ciceroSection{4} But Antony arrived yesterday evening. He may come to me at any moment now --- or perhaps he will not even do that, since he has written what he wants done. You will know straight away what has been done. We deal in nothing now except covertly. About the boys, what am I to do? Am I to commit them to some small craft? What sort of frame of mind do you suppose I shall be in while sailing? I remember how anxious I was that summer when I sailed with the boy in that Rhodian \textit{[Greek: aphrakt\=o(i)]}, an undecked galley; what sort of state do you think I shall be in, in a small skiff and at this harsh season of the year? What a wretched business every way one turns it! Trebatius was with me, a thoroughly good man and good citizen. What monsters that man reported --- immortal gods! Is even Balbus thinking of coming into the Senate? But I shall give him a letter to you tomorrow. As you say, I take Vettienus to be a friend of mine; with him --- because he had written to me rather \textit{[Greek: apotom\=os]}, peremptorily, about handling the money --- I had joked back rather \textit{[Greek: thumik\=oteron]}, with some heat. You can smooth that down if he has taken it otherwise than he should have. As for my addressing him as ``master of the mint,'' it was because he had addressed me as ``consul.'' But since he is a decent fellow, and fond of me, let him be a friend of mine in turn. Farewell.
