% Ad Familiares 14.18 (ad-familiares-14-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS TERENTIAE SVAE ET PATER SVAVISSIMAE FILIAE, CICERO MATRI ET SORORI S. D. P.}

\ciceroSection{1} My darlings, I think you must consider, again and again and carefully, what you should do --- whether you are to stay at Rome, or be with me, or in some safe place; this is not for me alone to decide, but for you as well. What occurs to me is this: that at Rome you can be safe through Dolabella, and that the fact may be of help to us, if any violence or any plundering should begin. But then again this gives me pause: I see that all loyal men have left Rome, and that they have their womenfolk with them. As for this region in which I am, it is full both of towns of ours and of estates of ours, so that you can spend much of your time with me, and, when you go away, can well stay on one of our own places.

\ciceroSection{2} To me which is better is plainly not yet clear. You for your part should see what the other women of your station are doing, and take care that, when you should want to leave, it not be impossible. The matter I should like you to think over, carefully and again and again, with yourselves and with friends. Tell Philotimus to see that the house has its ramparts and its guard; and please set up reliable letter-carriers, so that I may have some lines from you every day; but above all, take pains to keep well, if you want us to be well. The ninth day before the Kalends, at Formiae.
