% Ad Familiares 4.1 (ad-familiares-04-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. SER. SVLPICIO.}

\ciceroSection{1} G. Trebatius, a close friend of mine, has written to me that you asked him where I was, and that you took it ill that on account of your ill health you had not seen me when I came up to the city, and that at this present moment you wish to confer with me, if I would come a little nearer, about the duty of each of us. If only, Servius, we could have spoken together while things stood safe (so we must put it). Surely we should have brought some aid to the dying state. For I had already learned, even at a distance, that you, foreseeing these evils long beforehand, had been a defender of peace in your consulship and after your consulship; and I, although I approved your counsel and felt the same myself, was making no headway --- I had come late, I was alone, I seemed inexpert in the cause, I had fallen among the madness of men hungry for the fight. Now, since we seem no longer able to help the state, if there is anything by which we can take counsel for ourselves --- not to keep something of our former condition, but to mourn the most honourably we can --- there is no one in the world with whom I would rather take counsel than with you. The examples of those most illustrious men whom we should resemble, and the precepts of the wisest, whom you have always cultivated, are not lost on you. I would myself have written to you earlier that you would come to the Senate, or rather to a meeting of senators, in vain --- had I not feared to give offence to the man who was asking me to imitate your example. To him, when he asked me to attend the Senate, I made it plain that I would say all the same things which you had said about peace and about the Spains.

\ciceroSection{2} You see how matters stand: the whole world is on fire in war over the distribution of commands; the city has been left to plundering and burning, without laws, without courts, without right, without good faith. So nothing can come into my mind --- not only nothing I might hope for, but scarcely yet anything I would dare to wish for. Still, if it seems useful to you, the most prudent of men, that we should confer, although I was indeed thinking of withdrawing further still from the city (the very name of which I now hear against my will), nonetheless I shall come a little closer; and I have charged Trebatius that, if there is anything you would have him bring me, he is not to refuse, and I should like you to do this, or, if you prefer, to send to me some trustworthy man of your own --- so that you need not leave the city or I come up to it. I attribute as much to you as I perhaps arrogate to myself: I take it for certain that whatever we may decide by common judgement, all men will approve. Farewell.
