% Ad Atticum 7.15 (ad-atticum-07-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} From the day I left the city I have not let a single day go by without sending you something by way of a letter --- not that I have much in particular to write, but that I may talk with you in your absence. Since I cannot do it face to face, nothing is more agreeable to me than this.

\ciceroSection{2} When I had arrived at Capua on the sixth day before the Kalends, the day before I gave this letter, I met the consuls and a great many of our order. All were eager that Caesar, with his garrisons withdrawn, should stand by the terms he had himself proposed; Favonius alone was unwilling that the laws be imposed on us by him. But \textdagger{}he was heard in\textdagger{} council. For Cato himself by now prefers to serve rather than to fight; but still he says that he wishes to be present in the Senate when the terms are discussed, if Caesar is brought to withdraw his garrisons. And so he is not eager to do what is most needed, which is to go to Sicily; he wishes to be in the Senate, which I fear will do harm. Postumius, on the other hand, about whom the Senate decreed by name that he should at once go to Sicily and succeed Furfanius, refuses to go without Cato and sets a great value on his own services and weight in the Senate. So the business has come round to Fannius. He is being sent ahead to Sicily with imperium.

\ciceroSection{3} In our discussions the disagreement is total. The majority deny that Caesar will stand by his terms, and say that these demands have been interposed by him in order that we might not get ready what is needed for the war. I, for my part, think he will go through with withdrawing his garrisons. For he will have won, if he is made consul, and will have won by a lesser crime than the one with which he began. But the blow must be taken. For we are shamefully unprepared both in soldiers and in money; and as for that, we have left all of it behind for him --- not only the private money that is in the city, but the public money too, which is in the treasury. Pompey has set out for the Appian legions; he has Labienus with him. I await your opinions on these matters. I had been intending to make for Formiae without delay.
