% Ad Atticum 7.23 (ad-atticum-07-23) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the fifth day before the Ides of February, in the evening, I had a letter from Philotimus: that Domitius has a strong army, that cohorts from Picenum under Lentulus and Thermus have joined Domitius's force, that Caesar can be cut off and is in fear of it, that the spirits of the loyalists have been revived at Rome and the disloyal are as if stunned. These things, I am afraid, may be dreams; even so, Manius Lepidus, Lucius Torquatus, Gaius Cassius, tribune of the plebs --- these are with us, that is, at the Formian villa --- Philotimus's letter has called them back to life. As for me, I am afraid the other things are nearer the truth: that almost all of us are now as good as taken, that Pompey is quitting Italy --- whom (O the bitterness of it!) Caesar is said to be pursuing. Caesar pursuing Pompey? What for? To kill him? O wretched me! And do we not all of us throw our bodies in his way? At which you too groan. But what are we to do? We are beaten, crushed, taken outright.

\ciceroSection{2} Still, on reading Philotimus's letter I have changed my mind about the women. As I had written to you, I was sending them back to Rome; but it came to me that it would be much talked of --- that I had now passed judgement on the public cause, and that, having despaired of it, this in effect was the first step of my own homecoming, that the women had gone back. As for myself, I agree with you: I should not commit myself to an uncertain and dangerous flight when I can do no good to the commonwealth and none to Pompey --- for whom I can die both with piety and with gladness. I shall stay, then; though to go on living ---.

\ciceroSection{3} As to your question of what is happening here: all Capua and every levy here is on the floor; the cause is desperate; everyone is in flight, unless some god help us so that Pompey may join those forces of Domitius's to his own. But we seemed likely to know everything within two or three days. I have sent you a copy of Caesar's letter; you had asked for it. Many have written to me that I am very acceptable to him; which I am content to bear, so long as I do nothing disgraceful, as up to now.
