% Ad Atticum 8.2 (ad-atticum-08-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} To me, indeed, every part of it was welcome --- that you wrote me what you had heard, that you did not believe what was unworthy of the care I take, that you warned me of what you felt. From Capua I sent Caesar one letter, in which I wrote back to him on what he had taken up with me about his gladiators: short, but indicating goodwill, not only without insult to Pompey but with the highest praise of him. For that was what the policy required by which I was urging Caesar toward concord. If he has sent the letter anywhere, I should be glad to have him post it up in public. A second letter I sent on the same day I sent this one to you. I could not but send it, since he himself had written to me, and so had Balbus. I have sent you a copy. There will, I think, be nothing in it for you to criticize. If there is anything, instruct me how to escape \textit{[Greek: mempsis]} --- reproach.

\ciceroSection{2} You will say: write nothing at all. How are you the more to escape the people who want to invent things? Still, I shall do so to the extent that I can. As for your urging me to keep in mind what I have done, said, and even written, you do so as a friend, and most gratefully to me; but you seem to me to be judging some other thing honourable and worthy of me in this cause than I myself judge. For to me it appears that nothing more disgraceful has ever been done by any guide and leader of a commonwealth, among any nation, than has been done by our friend. I grieve at his lot; for he has abandoned the city, that is, the fatherland, for which it was a glorious thing both to die and to die in. You seem to me not to know how great a disaster this is.

\ciceroSection{3} For you are still in your own house, but, against the will of the most utterly lost men, you cannot stay there longer. What can be more wretched, what more disgraceful, than this? We are wandering, with our wives and children and nothing of our own; we have all our hopes placed on the breath of one man dangerously sick every year of his life; we have not been driven out, we have been called out of our fatherland; the fatherland we have left not to be kept for our return, but to be plundered and burned. ``Are so many with us?'' Are they not in their suburban houses? not in their gardens? not in the city itself? And if they are there now, will they not be there still? We meanwhile are not even at Capua, but at Luceria --- and the seacoast itself we shall soon be giving up; we shall be waiting for Afranius and for Petreius. For there is too little dignity in Labienus. Here is where you look for in me the thing you spoke of. About myself I say nothing; let others judge that. But here, in his case --- what is there? You and yours are at home, and you and all the loyalists are going to be at home. Who failed to show himself to me back then? Who is at hand now in this war? --- for that is what it must now be called.

\ciceroSection{4} The achievements of Vibullius have so far been very great. You will learn this from Pompey's letter; in which you must take note of the place where the mark \textit{[Greek: diple]} will stand. You will see what Vibullius himself thinks of our Gnaeus. Where, then, is this discourse of mine pointing? For Pompey I can gladly die; I rate no man on earth higher; but I do not set, as you do, my one hope for the safety of the commonwealth on him alone. For you suggest, rather differently from your wont, that even Italy must be left, if he leaves it. This I think neither useful to the commonwealth nor to my children, and besides, neither right nor honourable. \textit{†sed cur†}. Will you, then, be able to look upon a tyrant? --- as if it made any difference whether I heard him or saw him, or as if I had to find an authority weightier than Socrates, who, when the Thirty Tyrants were in power, did not set foot beyond the gate. Besides this I have a particular reason for staying. Of this I wish I could speak with you some day. I, on the thirteenth day before the Kalends --- when I had written this letter by the same lamp by which I had set yours afire --- setting off from Formiae for Pompey, if it is to be peace; if war, what shall I be?
