% Ad Atticum 10.13 (ad-atticum-10-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Your letter was most welcome to my Tullia and, by Hercules, to me. Your letters always bring some glimmer with them. So write again, and if you can manage anything that points to hope, do not let it slip. Take care not to be frightened by Antony's lions. There is nothing more agreeable than the man. Observe a statesman's policy in action --- \textit{[Greek: praxin politikou]}. He summoned by letter, from the municipalities, the ten leading citizens and the boards of four. They came to his villa in the morning. First he slept until the third hour; then, when it was announced that the Neapolitans and the Cumans had come (since Caesar is angry with them), he ordered them to come back the next day; he wanted to bathe --- and \textit{[Greek: peri]} --- and so on. That was yesterday's accomplishment. Today he has decided to cross over to Aenaria to promise the exiles their return. But enough of this; let us discuss our own affairs.

\ciceroSection{2} I have received a letter from Axius. About Tiro --- thank you. Vettienus I am fond of. To Vestorius I have repaid. Servius is said to have lodged at Minturnae on the day before the Nones of May, and to be lodging today at the Liternum estate with Gaius Marcellus. Tomorrow morning, then, he will see me and will furnish me with the substance of a letter to you; for I am already running out of anything to write to you. What does astonish me is that Antony has not so much as sent me a message, though he had been particularly attentive to me. Plainly the order he has been given about me is rather a brutal one. He does not want to refuse me to my face, when in fact I was neither going to ask him for anything nor, if I had got it, going to believe it. We shall come up with something, even so.

\ciceroSection{3} Please --- anything from Spain. The news must be near now, and everyone is in such a state of expectation that, if it goes the right way, they reckon there will be no further trouble. For my part, I do not think the business finished if our side has held, nor desperate if they have lost. I take it that Silius and Ocella and the rest have been held back. I see that you too are held up by Curio. Though, as I think, you have $\dagger$ \textit{[Greek: EKITAONON]} $\dagger$.
