% Ad Atticum 8.6 (ad-atticum-08-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} When that letter of mine had already been sealed, which I was meaning to send before daybreak --- and did so send (for I had written it the previous evening) --- C. Sosius the praetor came to the Formian villa to M'. Lepidus our neighbour, the man under whom he had served as quaestor. He brought a copy of a letter from Pompey to the consuls:

\ciceroSection{2} ``A letter from L. Domitius was delivered to me on the thirteenth day before the Kalends of March. I have written out its copy below. Now, even if I do not write it, I know that you understand on your own how much it matters to the commonwealth that all the forces converge upon a single place at the earliest possible moment. You, if it seems good to you, will see to coming to us as soon as possible, leaving at Capua whatever garrison you have determined will be sufficient.''

\ciceroSection{3} Then he appended a copy of Domitius's letter, which I had sent to you the day before. Gods immortal, what horror has flooded through me! How anxious I am over what is to come! This much, even so, I hope: that the great name of the commander will count for something, that great will be the terror at his arrival. I hope also that, since up to now no negligence has done us any harm, \textit{$\dagger$nothing will have been changed by negligence in what has been done bravely and carefully --- and indeed, by Hercules$\dagger$.}

\ciceroSection{4} For I have just heard that the quartan fever has left you. I should die, if I were more glad were it to have happened to me. Tell Pilia it is not fair that she keep it any longer, and that it is not in keeping with the concord between you. I hear that our Tiro has been abandoned by the other fever as well. But I see he has borrowed from others for his expenses; whereas I had asked our friend Curius to provide whatever might be needed. I would rather it be Tiro's diffidence that is at fault than Curius's stinginess.
