% Ad Atticum 9.12 (ad-atticum-09-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I had read your letter on the thirteenth day before the Kalends when a dispatch is brought to me from Lepta: that Pompey is shut in by a siege-wall, and that the very exits of the harbour are held by rafts. So help me, for the tears I cannot think out the rest, nor write it. I have sent you a copy. Wretched that we are! Why did we not all run his fate down to the end with him? Then, look, the same news from Matius and Trebatius, whom Caesar's couriers met at Minturnae. Unhappy, I am racked, so that by now I long for that end of Mucius. But how honourable, how clear-cut your counsels are, how thoroughly worked through in your thinking --- the route by land, the route by sea, the meeting and conversation with Caesar! Everything alike honourable and cautious. And the invitation to Epirus --- how sweet, how generous, how brotherly!

\ciceroSection{2} About Dionysius I have been astonished: a man more highly placed in my house than Panaetius was in Scipio's, and our present fortune has been looked down on by him in the foulest way. I hate the man, and I shall go on hating him; would I could take revenge! But his own character will take revenge on him.

\ciceroSection{3} You then, I beg you, now above all consider what I am to do. The army of the Roman people sits down around Cnaeus Pompey, holds him hemmed in by ditch and rampart, cuts off his escape; and we go on living, and that City of ours stands, the praetors give judgement, the aediles get up the games, the honest men enter up their interest payments --- and I myself sit still! Should I try to go there like a madman, and call on the loyalty of the municipal towns? The honest men will not follow; the trivial will laugh; lovers of revolution, the conquering side and armed, will lay hands on me by force.

\ciceroSection{4} What, then, is your opinion? Have you any plan for ending this most wretched life of mine? Now I grieve, now I am racked, when to some I seem either wise, because I did not go with him, or even fortunate. To me, the opposite. I never wished to be a sharer of his victory; I should have preferred to share his calamity. What use is it now for me to call on your letters, on your prudence or your goodwill? It is done with; by nothing now can I be helped, I who no longer have even anything left to wish for --- except to be set free by some mercy of an enemy.
