% Ad Familiares 5.20 (ad-familiares-05-20) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO RVFO.}

\ciceroSection{1} I would have come to meet you somehow or other, if you had been willing to come where you had said you would. Therefore, although you did not want to inconvenience me on your own account, I should like you to take it as certain that, had you sent for me, I would have put your wish before my own convenience. To what you have written, I could in fact reply on each particular more conveniently if my secretary M. Tullius were here. Of him I am sure --- at any rate as far as the rendering of accounts is concerned (in other matters I cannot vouch for him) --- that he has done nothing knowingly contrary either to your interest or to your reputation; and further, that if the old law and custom for the rendering of accounts still held, I would never have rendered them without first having gone over them with you, and finished them with you, as the bond of our intimate connection required.

\ciceroSection{2} What I would have done, then, at Rome had the old practice still stood, that --- since under the Julian law the accounts had to be left in the province and to be entered word for word at the treasury --- I did in the province; and I did it not in such a way as to draw you into my judgement, but giving you exactly that much consideration which I shall never regret having given. The whole of my own secretary, of whom you are now suspicious, I handed over to you. You attached to him M. Mindius your brother. The accounts were finished, with you, in my absence; to them I added nothing beyond a reading. I took the book from my own secretary, on the same footing on which I took it from your brother. If it was a matter of compliment, no greater could I have shown you; if of trust, I trusted you almost more than I trust myself; if of vigilance --- to see that nothing was entered otherwise than was both honourable and useful to you --- there was no one to whom I had rather given the business than the one I did. This much at any rate was done as the law required: that in two states, Laodicea and Apamea, the ones which seemed to us the foremost --- since it had to be so --- we deposited the completed accounts after collation. So to this point I answer first: that although for good reasons I was in haste to render the accounts, I would still have waited for you, had I not regarded accounts left in the province as already rendered. Wherefore ---

\ciceroSection{3} What you write about Volusius is not a matter of accounts. Men expert in the business --- and chief of them, both as the most expert of all and the most attached to me, G. Camillus --- have shown me that the obligation could not be transferred from Valerius to Volusius, and that Valerius's sureties remain bound. Nor was the sum 3{,}000{,}000 sesterces, as you write, but 19{,}000; for the money under the name of Valerius the contractor had been provided to us, and from that I entered in the accounts the balance that remained.

\ciceroSection{4} But by this you cheat me at once of the fruit of my generosity and of my diligence and also --- though about this I least trouble myself --- of any common-sense prudence: of generosity, because you would rather it were by my secretary's favour than by mine that my legate and prefect Q. Lepta has been spared the gravest calamity, especially when they ought not to have been obligated at all; of diligence, because you suppose that, in a service of mine so substantial and at so great a risk, I neither knew nor gave a thought to anything, and that my secretary entered whatever he pleased without even reading it out to me; of prudence, \dag\ when you think the affair was not exactly well thought out on my part. For both the freeing of Volusius was my own counsel, and the contrivance by which so heavy a burden was removed from Valerius's sureties and from T. Marius himself was my own; the device which everyone not only approves but actually praises --- and, if you want the truth, the one person I have noticed it not much pleased was my secretary. But I supposed that it was the part of an honest man, while the public's claim was preserved, to take thought also for the fortunes of so many --- whether friends or fellow citizens.

\ciceroSection{5} As for Lucceius, the arrangement was this: that on the authority of Cn. Pompey the money in question was deposited in a temple. I acknowledged that this was done by my order. The money Pompey used; that which you had deposited, Sestius. But all this, I gather, has nothing to do with you. As for my not having noticed --- I would take it ill if I had not added that you had deposited money in the temple by my order --- if that money were not attested by the gravest and most certain monuments: to whom it was given, by what decree of the Senate, by what letters of yours, by what letters of mine, it was handed to P. Sestius. Seeing it stamped with so many traces that one could not go wrong over them, I did not add what was no concern of yours; still, I would rather have added it, since I see you desire it.

\ciceroSection{6} As you write that you must render this item, I think the same myself, and your accounts will not disagree with mine in any particular: for you will add, on my orders, what I, who did not add it, neither have any reason to deny nor, if there were a reason and you did not wish it, would deny. As to the 900{,}000 sesterces, that has certainly been entered as you or your brother wished. But if, given that the Lucceius item was insufficiently provided for, there is anything I can even now correct in the rendering of the accounts, I must consider --- since I did not act on a senatorial decree --- what the laws permit. Certainly, unless I am mistaken, you ought not to have entered those items as money paid in on the basis of my rendered accounts; for there are others more expert in the matter. But have no doubt that I will do everything which I judge to be in your interest, or even what I judge you wish, if I can do it in any way.

\ciceroSection{7} As to what you write about benefits, please understand that I named as recipients only my own military tribunes, prefects, and members of my staff. In this my reckoning failed me: I supposed time was allowed me to name them, but I was informed afterwards that they had to be named within the thirty days in which the accounts were rendered. I felt it sharply that those benefits had not been kept back for your interests rather than mine, since I have no use for canvassing. As for the centurions, however, and the staff-members of the military tribunes, the matter is open; for that class of benefit had not been defined by law.

\ciceroSection{8} There remains the question of the 100{,}000 sesterces, about which I remember a letter was brought to me from you at Myrina, on a fault which was not mine but yours --- on what seemed an error, if there had been one, of your brother and of Tullius. But since the matter could not be put right --- the accounts having already been deposited and I out of the province --- I believe I wrote back as kindly as I could, in line with my feeling toward you and the prospect of resources we then had. But I do not consider myself bound today by the kindness of that letter, nor do I take your letter today about the 100{,}000 sesterces as those people take letters who in times like these find letters a nuisance.

\ciceroSection{9} You must also consider this: that all the money which had come into my hands consistently with the laws I had deposited at Ephesus with the publicans; that it was 2{,}200{,}000 sesterces; that the whole of this sum Pompey carried off. Whether I bear that equably or not, you ought to bear the matter of the 100{,}000 sesterces equably, and to suppose that by so much less reached you either from your maintenance allowance or from my generosity. If you had charged me those 100{,}000 sesterces, still --- such is your charm and the affection you bear me --- you would not have wished to take payment by valuation from me at such a time; for to pay in cash, even if I wished to, I had no means. But take all this for a joke, just as I take you. Still, when Tullius is back from the country, I will send him to you, if you think it will be to the purpose. As for why I should want this letter torn up, there is no reason at all.
