% Ad Atticum 10.16 (ad-atticum-10-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I had just given you a letter on several matters when, very early in the morning, Dionysius came to me. To him I would have shown myself not just pacified but entirely conciliated, had he come in the frame of mind your letter to me had indicated. For your letter, which I received at Arpinum, said that he would come, and would do what I wished. Now what I wished --- or rather what I longed for --- was that he should be with us. Because, when he came to my Formian villa, he cut that off flatly, I have been used to writing of him to you in harsher terms. But he, after a very few words, gave the substance of his speech in this: that I should forgive him; that, being entangled in his own affairs, he could not go with us. I answered briefly; I felt great pain; I understood that he had looked down on our fortunes. What more can I say? You will be surprised, perhaps: among the greatest griefs of these times, know that this man is one to me. I hope he may be a friend to you. In wishing this for you, I am wishing you happiness; for then it will last just so long.

\ciceroSection{2} I hope our plan will be free of danger. We have dissembled, and, as I think, we shall not be watched too sharply. Only let our voyage be such as I hope; the rest, so far as foresight in planning can provide for it, will be guarded against. While I am here, I should be grateful if you would write me not only what you know or have heard but also what you foresee as coming.

\ciceroSection{3} Cato, who could have held Sicily with no trouble (and, had he held it, all loyal men would have rallied to him), set sail from Syracuse on the eighth day before the Kalends of May, as Curio writes me. Would that Cotta, as they say, may hold Sardinia! For so the rumour goes. Oh, if that should prove true --- the shame of Cato!

\ciceroSection{4} I, to lessen suspicion of my departure or even of the thought of it, set out for my Pompeian villa on the third day before the Ides, to be there while what was needed for the voyage was being made ready. When I had reached the villa, word was brought to me that the centurions of the three cohorts which are at Pompeii wished to meet me the next day --- our friend Ninnius told me this --- and that they wanted to hand over to me both themselves and the town. But I left the villa the next day, before dawn, so that they should not see me at all. For what was in three cohorts? What if there had been more? With what equipment? I thought over those same Caelian considerations I had read in your letter, which I received just as I arrived back at the Cuman villa on the same day, and at the same time it could have been a test laid for me. So I removed all ground for suspicion.

\ciceroSection{5} But while I was returning, Hortensius had come and had stopped off to pay his respects to Terentia. He had spoken about me in honourable terms. I shall see him, as I think, presently; he has sent a slave saying that he is coming to me. This, at any rate, is better than what our colleague Antony does, whose litter, with an actress inside, is carried among his lictors.

\ciceroSection{6} You, since you are free of the quartan, and have got rid of that new ailment too, the head-cold: present yourself to us in Greece full of vigour, and meanwhile something of a letter.
