% Ad Atticum 10.9 (ad-atticum-10-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The arrival of Philotimus --- and what a fellow he is, what a tasteless creature, and what a serial liar on Pompey's behalf! --- left everyone here with me stricken; I myself have gone numb. None of us doubted that Caesar had halted his march --- in fact he is reported to be flying; and that Petreius had joined Afranius --- he brings no such news. What more can I say? It was even believed that Pompey had marched with great forces into Germany by way of Illyricum; for that was being reported \textit{[Greek: authentik\=os]}, on first authority. So let us make for Malta, I think, until something happens in Spain. And in fact I seem to be able to do that almost with Caesar's blessing, judging by his letter, which says that nothing can be more honourable or safer for me than to keep clear of all the strife.

\ciceroSection{2} You will say: where, then, is that spirit of yours from your last letter? It is here, and it is the same; but if only I had only my own head to risk! The tears of my own people sometimes soften me, when they beg me to wait on what comes from Spain. As for the letter from Marcus Caelius --- a piteously written thing, in which he too begged me to wait, not to throw away my fortunes, not to throw away my only son, not to throw away all my people so rashly --- our boys could not read it without bursting into tears. Though mine, indeed, is the stronger of them, and for that very reason moves me the more violently; and he has no concern for anything except my standing. To Malta, then; afterwards, wherever seems best.

\ciceroSection{3} You, even so, send me some scrap of a letter, especially anything from Afranius. If I have a conversation with Antonius, I will write you what was said. I shall, however, be cautious in trusting, as you warn me; for the method of concealment is not only difficult but also dangerous. I am expecting Servius by the Nones; Postumia and young Servius are pressing me. I am glad to hear the quartan fever has eased. I have also sent you a copy of Caelius's letter.
