% Ad Atticum 8.15 (ad-atticum-08-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the fifth day before the Nones of March, Aegypta delivered your letters to me --- one of them an old one, from the fourth day before the Kalends, which you say you gave to Pinarius, whom we have not seen; in it you are waiting to hear what Vibullius, sent on ahead, is doing --- the one who was not seen by Caesar at all (I see from your second letter that you know this is so) --- and how I shall receive Caesar on his return, whom I am thinking of avoiding altogether; and you press the point of your flight to Athens and the change in your way of life, which I think you should do; and you do not know whether Domitius is with the fasces. When you know this, you will see that we know it too. That is the answer to the first letter.

\ciceroSection{2} Two letters followed, both given on the day before the Kalends, which have shaken me from my former position --- already, as I wrote you above, tottering. Nor do I find any weight in your saying that he is unjust even to Jove himself. For the danger lies in the wrath of each of them; the victory, however, is so uncertain that the worse cause seems to me the better-prepared. Nor do the consuls move me; they themselves are moved more easily than a feather or a leaf. The deliberation over duty tortures me, and has been torturing me up to now. Staying is certainly the safer course; crossing is judged the more honourable. I would sometimes rather have many men think I had acted incautiously than a few think I had acted dishonourably. As for what you ask about Lepidus and Tullus --- they have no doubt at all that they will be on hand for Caesar and will come into the Senate. Your most recent letter was given on the Kalends, in which you wish for a meeting and do not despair of peace. But I, while writing these things, was thinking that they would neither meet, nor, if they had met, would Pompey come round to any terms. As for what you seem not to doubt --- that if the consuls cross over, we know what we must do --- certainly they are crossing, or, as things stand now, have crossed. But remember that, with the exception of Appius, there is hardly anyone who has not the right of crossing. For either they are vested with \textit{imperium} --- like Pompey, like Scipio, Sufenas, Fannius, Voconius, Sestius, the consuls themselves, to whom by ancestral custom it is permitted to enter all the provinces --- or they are their legates. But I am deciding nothing; what is pleasing to you, and what is in some manner right, I understand. I should write more, if I could myself. But, as it seems to me, I shall be able to in two days. A copy of the letter of Cornelius Balbus, which I received on the same day as yours, I have sent on to you, that you might grieve on my behalf, when you saw me being made a laughingstock.
