% Ad Atticum 5.1 (ad-atticum-05-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I saw your feelings at our parting, and as for my own I am my own witness. All the more reason for you to see that no new measure is decreed, so that this separation of ours need not be more than a year. About Annius Saturninus you have handled the matter well.

\ciceroSection{2} As for the giving of sureties, please give them yourself as long as you are at Rome. There are several to be given on the transfer of property, for instance on the Mennian and the Atilian estates. The Oppius business has gone as I wished, and especially because you opened the credit of 800{,}000. I want that sum paid in any case, even if it has to be raised by a fresh loan, so that the final calling-in of our debts need not be waited for.

\ciceroSection{3} Now I come to that line drawn across the foot of your last letter in which you remind me about your sister. Here is how matters stand. As soon as I had reached Arpinum and my brother had joined me, our first conversation, and a long one, was about you. From this I came round to what you and I had said to each other about your sister at Tusculum. I never saw anything so mild, anything so placid as my brother was then toward your sister: if there had been any friction over money, it did not show. So passed that day. The next morning we set out from Arpinum. A holiday kept Quintus at Arcanum for the day, but I had to go on to Aquinum; we lunched at Arcanum, though. You know the place. When we arrived, Quintus said in the most courteous way, ``Pomponia, you invite the women; I will gather the men.'' Nothing, at least so it seemed to me, could have been kinder, both in his words and in the spirit and look that went with them. But she, in our hearing, said, ``I am a guest here myself'' --- because, as I suppose, Statius had gone on ahead of us to look after lunch. At that Quintus said to me, ``There --- this is what I put up with every day.''

\ciceroSection{4} You will say, ``What of it, in heaven's name?'' Something. Even I was shaken; the harshness and the absurdity of her words and her face as she gave her answer were that bad. I covered my pain. We all took our places at table except for her; Quintus, however, had food sent to her from the table. She turned it away. To say it briefly: nothing seemed gentler than my brother, nothing harsher than your sister --- and I pass over many things which at the time annoyed me more than they did Quintus himself. I went on from there to Aquinum. Quintus stayed on at Arcanum, and the next morning came to me at Aquinum and told me she had refused to sleep with him, and that when she was on the point of leaving she had been just as I had seen her. In a word, you can even tell her yourself: in my judgement she was lacking in common decency that day. I have written about this to you at more length perhaps than was necessary, so that you might see your share in setting the matter right and giving counsel. What remains is for you to discharge our commissions before you set out, to write to me about everything, to push Pomptinus off, to see that I know when you have started; and to be assured --- by Hercules --- that nothing is dearer or sweeter to me than you. I bade A. Torquatus, an excellent man, a most affectionate farewell at Minturnae; I should like you to drop a hint in conversation that I have written something to you about him.
