% Ad Atticum 10.17 (ad-atticum-10-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the day before the Ides Hortensius came to me, my letter already written. \dag\ I could wish the rest of him matched it\dag\ --- but what extraordinary devotion he shows me, \textit{[Greek: ekten\=eian]}, eagerness; and I do mean to make use of it. Then Serapion came in with your letter. Before I opened it, I told him that you had written to me about him before, as you had. And then the letter \dag did so\dag\ in the fullest possible terms. And by Hercules, I approve of the man; I take him to be both learned and honest. I think, in fact, that I shall make use of his ship, and of him as my fellow-passenger.

\ciceroSection{2} The inflammation of the eyes keeps flaring up afresh --- not unbearably, but enough to hamper my writing. I am glad your health is now firmly restored, both from the old illness and from the new bouts.

\ciceroSection{3} I wish we had Ocella with us; matters here seem likely to be a little easier from now on. As things stand, what holds us up is the equinox, which has been very rough. \dag If it settles tomorrow,\dag\ I hope Hortensius continues as he is; for nothing could be more generous than he has been so far.

\ciceroSection{4} About the warrant, you are astonished, as though I had accused you of some scandal or other. You say you cannot guess what put it into my head. Well: because you had written that you were thinking of setting out (and I had heard that no one was permitted to do so on any other terms), I took it for granted that you had one; and also because you had drawn one for your slaves. There you have the basis of my opinion. Even so, I should like to know what you are thinking, and especially anything new even now. The seventeenth day before the Kalends of June.
