% Ad Atticum 7.18 (ad-atticum-07-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the fourth day before the Nones of February our women came to Formiae and brought me the services you have rendered them, full of your sweet attentiveness. I wished to keep them at the Formian house --- and the Ciceros with them --- until we should know whether we were to make use of a shameful peace or a wretched war. I myself, with my brother, set out for Capua --- to the consuls (for we were ordered to be present on the Nones) --- on the third day before the Nones, the day I gave this letter. Pompey's answers are said to be welcome to the people and approved by the assembly. So I had supposed. If Caesar rejects them, he will be down; if he accepts them ---. ``Then which,'' you will say, ``do you prefer?'' I would answer, if I knew how we were prepared.

\ciceroSection{2} Word had come here that Cassius had been driven out of Ancona and that the city was being held by our side. If there is to be a war, a useful piece of business. As for Caesar, though Lucius Caesar has been sent with instructions about peace, they say even so that he is most fiercely conducting his levy, seizing positions, garrisoning them with troops. The ruined brigand! No quiet, however long, could make good this disgrace to the commonwealth! But let us cease to fume, let us yield to the moment, let us go with Pompey to Spain. This is the best to be made of a bad business, since we did not push back his second consulship from the commonwealth even when the chance was given. But enough of this.

\ciceroSection{3} About Dionysius it slipped my mind to write to you before; but I had settled it as follows --- to wait for Caesar's answers, so that if we were to return to the city he should be awaiting us there, but if that should be a long time happening, then we should send for him. To be sure I know what he ought to have done in our flight, what would have been worthy of a man of learning and of a friend, especially since he had been asked; but I do not exact too much of this from Greeks. You, however, will see to it, if --- which I should not wish --- he is to be sent for, that we are not a nuisance to one who comes against his will.

\ciceroSection{4} My brother Quintus is anxious to settle what he owes you through Egnatius; nor is Egnatius unwilling, nor is he insufficiently solvent --- but the times being what they are (Q. Titinius, who is much with us, says he has no travelling money and has given notice to his debtors that they are to keep the loan at the same rate of interest, and L. Ligus is said to have done the same; and Quintus has at the moment neither cash at home nor any means of getting it in from Egnatius nor of taking up a new loan anywhere) he is surprised you have not taken account of this public difficulty. I, for my part, although I observe that \textit{[Greek: pseudo-Hesiodic]} maxim (for so it is held to be) --- \textit{[Greek: not even a lawsuit]}, especially with you, from whom I have never seen anything done thoughtlessly --- was still moved by his complaint. Such as it is, I wanted you to know.
