% Ad Atticum 10.18 (ad-atticum-10-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} My Tullia was delivered on the fourteenth day before the Kalends of June of a boy, \textit{[Greek: heptam\=eniaios]}, a seven-months child. I am glad the delivery was easy, \textit{[Greek: eutok\=esen]}; what was born, however, is desperately weak. Up to now an extraordinary spell of calm weather has kept me here, and has been more of a hindrance than the watchers who are set on me. As for all that Hortensius business, \dag it has come to nothing,\dag\ \dag and so it will turn out:\dag\ the man is utterly worthless --- corrupted by the freedman Salvius. So from now on I shall not write to you what I am about to do, but what I have done; for every \textit{[Greek: K\=orukaios]}, every eavesdropper of Corycus, seems to be listening in on what I say.

\ciceroSection{2} You, even so, if anything comes through from Spain or anything else, please go on writing; and do not wait for letters from me, except when we have arrived where I wish, or if anything turns up in the course of the journey. But even this I write timidly; so slow and so dense is everything still. As we have made a bad beginning, so the rest follows.

\ciceroSection{3} It is for Formiae now that I am making; perhaps the Furies will follow me there too. From the conversation Balbus had with you, I do not approve about Malta. So can you doubt that he regards me as among his enemies? I have written to Balbus that you have written to me of both his goodwill and his suspicion.

\ciceroSection{4} I gave him my thanks; on the other point, clear me with him. Has there ever been a more unlucky man than I? I shall say no more, to spare you the rack too. I am being torn apart at the realisation that the moment has come when I can no longer do anything either bravely or prudently.
