% Ad Atticum 10.4 (ad-atticum-10-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I received many letters from you on the same day, every one of them carefully written, and one of them in particular --- the one that was the size of a volume --- worth reading over and over, as in fact I do. In writing it you have not undertaken the labour in vain; you have done me a very great kindness. So I ask you most earnestly to do it as often as you possibly can, for as long as it is possible --- that is, for as long as you know where I am. As for lamentation, which is what we do every day, let there now be either an end to it altogether, if that is possible, or at any rate some measure to it, which surely is possible. For I no longer dwell on the standing I have lost, the offices, the position in life; I think instead of what I have achieved, what I have given, by what praise I have lived, and finally --- in these evils --- what difference there is between me and those men on whose account we have lost everything. These are men who thought they could not secure the licence of their appetites unless they drove me from the city. Where the good faith of their compact and their criminal collusion has burst out, you see.

\ciceroSection{2} The one burns with fury and crime, and does not let up at all but grows heavier by the day; just now he drove us from Italy, now from one quarter he tries to pursue us and from another to strip us of our province; nor does he any longer refuse but in a way demands that, being what he is, he should also be called a tyrant.

\ciceroSection{3} The other --- the man who once, when we were prostrate at his feet, did not so much as lift us up; who used to say he could do nothing against this man's will --- having slipped from his father-in-law's hands and his sword, prepares for war by land and sea. It is not an unjust war, granted, and indeed not only dutiful but also necessary; yet for his fellow citizens it will be ruinous if he loses, and disastrous even if he wins.

\ciceroSection{4} I rank the achievements of these supreme commanders not only no higher than my own, but I do not rank even their fortune higher; theirs has been most flourishing, mine, it seems, has been more roughly handled. For who can be happy when his fatherland is either deserted by itself or crushed? And if, as you remind me, we said rightly in those books of mine that nothing is good except what is honourable, nothing evil except what is base, then surely each of these men is most wretched, for to each of them his country's safety and dignity has always come second to his own domination and his own private advantage.

\ciceroSection{5} A splendid self-knowledge, then, sustains me, when I consider that, as regards the commonwealth, I have either served as well as I could, or at any rate have never thought otherwise than dutifully; and that in this very tempest the commonwealth has gone down, the one that I foresaw fourteen years in advance. With this self-knowledge as my companion, then, I shall set out, in very great grief, and that grief not so much for my own sake or my brother's --- our lives are now lived through --- as for the boys, to whom we sometimes think we ought to have presented even a commonwealth to inherit. Of these one tortures me wonderfully precisely because his loyalty is greater; the other (the pity of it! for nothing more bitter has befallen me in all my life), corrupted, no doubt, by our own indulgence, has gone so far that I do not dare to say where. I am waiting for your letter; for you wrote that you would write more when you had seen him in person.

\ciceroSection{6} My indulgence toward him was complete, joined with much severity; and I have suppressed not one offence of his, nor a small one, but many great ones. As for his father's mildness, it ought rather to have been loved by him than to be so cruelly disregarded. The letter he sent to Caesar we took so heavily that we kept it from you, but we seem to have made his own father's life unsweet to him. What sort of thing this journey of his has been, and what sort his pretence of piety, I do not dare to say; this much I know --- after the meeting with Hirtius he was sent for by Caesar, and they conversed about my state of mind, utterly alienated from his calculations, and my plan to leave Italy; and even this timidly. But it is no fault of ours; it is his nature that is to be feared. This is what corrupted Curio, this is what corrupted Hortensius's son, not the fault of their fathers. My brother lies broken in grief and is afraid not so much for his own life as for mine. To this man, to this man bring whatever consolations you can, if you can; most of all I should wish for this one --- that what has been reported to me may turn out either false or less than the truth. But if they are true, what is going to come of it in this life of flight, I do not know. For if we had a commonwealth, I should not be at a loss for counsel either toward severity or toward indulgence. As it is, I have written these things, whether stirred by anger or by grief or by fear, more harshly than either your love for him or mine demanded; if they are true, you will forgive me; if false, you will gladly rescue me from this misapprehension. Whatever way the thing stands, you will lay the blame neither on the uncle nor on the father.

\ciceroSection{7} When I had written this much, word was brought to me from Curio that he was coming to see me. He had reached his place at Cumae the evening before --- that is, on the Ides. So if any matter comes out in his talk that ought to be written to you, I will append it to this letter.

\ciceroSection{8} Curio drove past my villa and sent word that he would come along soon, and he hurried on to Puteoli to address a public meeting there. He held his meeting, came back, was with me for a good long while. A foul business! You know the man; he hid nothing. In the first place, that nothing is more certain than that all those who were condemned under the Pompeian law are to be restored; and so he intended to use their services in Sicily. As for the Spanish provinces, he had no doubt they would be Caesar's. From there, Caesar himself with an army to wherever Pompey was. Pompey's death will be the end of \dag{}the business\dag{}. Nothing came nearer to happening: in a plain burst of anger \dag{}he\dag{} had wanted Metellus the tribune of the plebs killed; and if that had been done, there would have been a great slaughter. There are very many men urging slaughter; he himself, however, is not cruel by inclination or nature, but because he thinks clemency wins popular favour. If he were to lose the people's enthusiasm, he would be cruel. And he was distressed, because he understood that he had given offence to the plebs themselves over the treasury. And so, although it had been wholly settled that he would hold a public meeting before he set out, he had not dared, and had set out in great mental disturbance.

\ciceroSection{9} When I asked him what outcome he foresaw, what kind of commonwealth, he plainly admitted that no hope was left. He feared Pompey's fleet. If it put out to sea, he would quit Sicily. ``What about,'' I said, ``those six fasces of yours? If they are from the Senate, why are they laurelled? If from him, why six?'' ``I wanted them,'' he said, ``filched from a decree of the Senate; for otherwise it could not be done.'' But \emph{Caesar} now hates the Senate more than ever. ``From me,'' he said, ``everything will proceed.'' But why six?

\ciceroSection{10} ``Because I did not want twelve; for I could have had them.'' Then I said, ``How I wish I had asked him for what I hear Philippus has obtained! But I was afraid, because I had been obtaining nothing from him.'' ``He would gladly,'' he said, ``have granted it to you. Indeed, suppose you have obtained it; for I will write to him, just as you yourself wish, that we have spoken together about it. What concern is it of his, anyway, where you are, given that you do not come to the Senate? In fact, at this very moment, you would have done his cause least harm if you had not been in Italy.'' To which I said that I was looking for retirement and solitude, chiefly because I had lictors. He praised the plan. ``What then?'' I said. ``For my route into Greece runs through your province, since there are soldiers on the upper sea.'' ``What,'' he said, ``could be more agreeable to me?'' At this point much was said most generously. So this much, at any rate, has been gained: that we may sail not only safely but openly.

\ciceroSection{11} The rest he deferred to the next day; of which I shall write to you whatever is worth a letter. There are also matters I passed over: whether he is going to wait for an interregnum, or, in the way he himself put it, have the consulship offered to him --- but he does not want it for next year. And there are other matters I shall question him on. To cap it all, he was swearing, a thing he does on no business at all, that Caesar must be most friendly to me. ``Why, when did he ever?'' ``Dolabella wrote to me.'' ``Tell me what.'' He insisted that Dolabella had written --- when I was wishing to come to the City --- that Caesar gave him the greatest thanks and not only approved but rejoiced. What more do you want? I was set at rest. For that suspicion of a domestic evil and of the talk with Hirtius was lightened. How I long for him to be worthy of us, and how unwillingly I bring myself to suspect \dag{}what must be suspected of him\dag{}! But it had to be done --- the meeting with Hirtius. Surely there is something in it, but I should wish it to be as little as possible. And we wonder, too, that he has not yet returned. But we shall see to these things.

\ciceroSection{12} You will assign the Oppii to Terentia. For now the danger to the city is empty. Still, help me with your counsel --- whether on foot to Rhegium, or from here straight aboard ship, and the rest, since I am lingering. I shall have something to write to you at once as soon as I have also seen Curio. About Tiro, please take care, as you do, that I may know how he is faring.
