% Ad Atticum 9.13 (ad-atticum-09-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The tale is not true \textit{[Greek: ouk \'est' \'etumos l\'ogos]}, I think, that one about the rafts. For why would Dolabella write, in the letter he sent from Brundisium on the third day before the Ides of March, of this so-called fine day \textit{[Greek: euh\=emer\'ian]} of Caesar's --- that Pompey was in flight, and would sail with the first wind --- which is utterly at odds with those letters whose copies I sent you earlier? Here, indeed, men talk nothing but villainy; but there could be no more recent authority, nor on this matter a better one, than Dolabella.

\ciceroSection{2} On the eleventh day before the Kalends I received your letter in which you put all decisions off to the time when we shall know what has been done. And surely it is so; in the meanwhile nothing can be settled --- nor even thought through. And yet this letter of Dolabella's bids me return to my old reflections. For the day before the Quinquatria there was a magnificent storm-wind; that, I take it, is what Pompey made use of.

\ciceroSection{3} Your collection \textit{[Greek: sunag\=og\'e]} of counsels was gathered up by me not for complaint but rather for my own consolation. For not so much was I troubled by these present ills as by the suspicion of fault and rashness on my own side. I judge there has been none, since my actions and counsels agree with yours. As for your writing that it has come about more by my own proclamations than by his desert that I seem to owe him so much --- it is true. I have always cried up Caesar's services to me, and the more for this reason: that he might not think me to be remembering the older grievances. Yet even if I were remembering them most sharply, I should still by now have to follow the precedent of that time. He gave me no help when he could; yet afterwards he was a friend, even very much so, and quite why I do not plainly know. Therefore I in turn am his friend. There is even this further pairing between the two of us: that we were drawn in by the same men. But would that I had been able to do as much for him as he was able to do for me! Still, what he did was most welcome to me. And I do not now know in what way I could help him; nor, if I could, when he is preparing so destructive a war, should I think I ought to help him.

\ciceroSection{4} I only do not wish to offend him in spirit by staying here, nor, by Hercules, to see those things which you can already foresee in your mind; nor could I be present at those evils. But I have been the slower in withdrawing for this reason: that it is hard to think of a voluntary departure when there is no hope of return. For I see this man so equipped in infantry, cavalry, fleets, and the auxiliaries of the Gauls --- whom Matius was puffing up \textit{[Greek: el\'apizen]}, I take it, though certainly he was saying so --- to be undertaking on his own resources, for ten years, the upkeep of his infantry and cavalry. But let that be exaggeration \textit{[Greek: l\'apisma]}; he has, for certain, great forces, and will have, not from any revenue of Italy but from citizens' property. Add to this the man's confidence; add the feebleness of the honest men, who, because they think he is rightly enraged at them, hate him --- as you write †I should have liked you to write, who it was you had in mind. But this man too, because† he had displayed more than he carried out, even those who loved him commonly love him no longer; the municipal towns, however, and the country gentry of Rome, fear that one and so far still hold this one dear. And so he is so well prepared that, even if he cannot win, I do not see how he himself can be beaten. As for me, I dread not so much his wizardry \textit{[Greek: go\=ete\'ian]} as his compulsion-through-persuasion \textit{[Greek: peithan\'ank\=en]}. For ``the requests of tyrants,'' says Plato, ``you know are mixed with compulsion'' \textit{[Greek: hai g\`ar t\^on tur\'ann\=on de\'eseis... oisth' h\'oti memigm\'enai an\'ankais]}.

\ciceroSection{5} Those harbourless coasts \textit{[Greek: al\'imena]} I see do not meet with your approval. Nor did they please me; but I had with them concealment, and a trustworthy service-crew \textit{[Greek: hup\=eres\'ian]}. If those were to be had for me at Brundisium, I should prefer it; but there there is no concealment. Still, as you write, when we shall know.

\ciceroSection{6} To the honest men I do not too much excuse myself. The dinners that, as Sextus has written me, they give and attend --- how sumptuous, how seasonable! But let them be as honest as you please, they are no better than ourselves. They would move me if they were braver. About Phamea's place at Lanuvium I was mistaken; I had dreamed of one near Troia. That I did want for Quintus; but the price is higher. That property of yours, all the same, I should be glad to buy, if I saw any hope of enjoying it.

\ciceroSection{7} What monstrosities we are reading every day, you will gather from the little booklet that has been thrown in with the letter. Our Lentulus is at Puteoli, at his wits' end \textit{[Greek: ad\=emon\^on]}, as Caesius reports, what to do. He dreads a repetition \textit{[Greek: diatrop\'en]} of the Corfinium business. By now he thinks Pompey has been done by enough; he is moved by Caesar's kindness; but still he is moved more by looking out for what is to come.

\ciceroSection{8} That you should be able to bear this! Everything wretched, but nothing more wretched than this. Pompey sent N.\ Magius about peace, and even so is being besieged. This I did not believe; but I have a letter from Balbus, a copy of which I have sent you. Read, I beg you, the last paragraph by that excellent Balbus himself --- the man to whom our friend Cnaeus once gave a site to build pleasure-gardens on, whom Cnaeus did not often prefer to any one of us. So now the wretch is in agony. But not to make you read the same thing twice, I refer you to the letter itself. Of peace, however, I have no hope at all. Dolabella, in his letter sent on the third day before the Ides of March, talks unmixed war. Let us stay, then, in that same wretched and despairing mind, since nothing can be more wretched than this.
