% Ad Familiares 16.12 (ad-familiares-16-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS S. D. TIRONI SVO.}

\ciceroSection{1} In what danger our welfare stands --- mine, and that of every loyal man, and of the commonwealth as a whole --- you can tell from this, that we have left our own houses and the fatherland itself to be plundered or burned. The matter has come to that pass, that, unless some god or some accident comes to our rescue, we cannot be saved.

\ciceroSection{2} For my own part, from the moment I came to the city walls, I did not stop perceiving and saying and doing every single thing that bore on concord. But a strange madness had got into not only the unprincipled but even those who pass for loyal men --- they longed to fight, while I was crying out that nothing is more wretched than civil war. And so when Caesar, swept along by a kind of insanity, and oblivious of his name and of the honours done him, had seized Ariminum, Pisaurum, Ancona, and Arretium, we left the city. How wisely or how bravely, there is no purpose now in arguing.

\ciceroSection{3} What straits we are in, you see. Terms, anyhow, are being floated by him: that Pompey go to Spain; that the levies that have been raised, and the troops we have under arms, be disbanded; that he, for his part, will hand over Further Gaul to Domitius and Hither Gaul to Considius Nonianus (for these have drawn them by lot); that he will come to seek the consulship, no longer asking that his candidacy be taken account of in his absence; that he will canvass for the office in person, present for the three market-days. We have accepted the terms, but on this condition: that he withdraw his garrisons from the places he has seized, so that, without fear, on these very terms,

\ciceroSection{4} the senate can be held at Rome. If he does this, there is hope of peace --- not of an honourable peace, for terms are being imposed --- but anything is better than being as we are. If, however, he refuses to stand by his own conditions, then war is prepared --- but a war of such a sort that he cannot sustain it, especially since he himself will have run away from his own terms. It needs only that we cut him off from being able to come at the city, which we were hoping could be done. For we were raising great levies, and we thought he was afraid, in case, if he began to march on the city, he would lose the Gallic provinces, both of which he holds as bitterly hostile (except for the Transpadanes); and he has at his back, from Spain, six legions and great auxiliary forces, under Afranius and Petreius as their commanders. It looks as if, if he goes on in his madness, he can be put down --- so long as the city is safe. The greatest blow he has taken is this: that the man who held the highest authority in his army, T. Labienus, was unwilling to be a partner in his crime. He has left him and is with us; and many are said to be on the point of doing the same.

\ciceroSection{5} For myself, I am at present in charge of the coastline beginning from Formiae. I have wished to take on no larger duty, so that my letters and exhortations toward peace might have the more weight with him. But if there is to be a war, I see that I shall be in command of a camp and of certain legions. I have this further vexation: that our Dolabella is on Caesar's side. I have wanted you to know all this. Take care that it does not upset you and hinder your recovery.

\ciceroSection{6} I have most carefully commended you to A. Varro, whom I have found to be most devoted to me, and a warm friend of yours as well: that he have a care for your health and your passage, and take you wholly into his charge and protection. He will do everything, I am confident: he has promised, and has spoken to me about you in the kindest terms. You, since you could not be with me at the very time when I most missed your hand and your loyalty, take care not to be in a hurry, and not to allow yourself to put to sea either while ill or in winter. I shall never think you arrived too late, if you arrive safe. So far I have seen no one who had seen you later than M. Volusius, from whom I received your letter --- and that did not surprise me; for I do not suppose that my letters get through to you, in such a winter. But take care to get well, and, when you are well and the sailing is good, then sail. My Cicero is at the Formian villa; Terentia and Tullia at Rome. Take care to keep well. The fourth day before the Kalends of February, Capua.
