% Ad Familiares 16.11 (ad-familiares-16-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS ET CICERO, TERENTIA, TVLLIA Q. Q. TIRONI S. P. D.}

\ciceroSection{1} Although I miss the help of your hand at every turn, still it is not so much on my account as on yours that I am grieved at your not being well; but since the force of the illness has turned to a quartan (so Curius writes), I hope that, with care taken, you will already be on the mend. Only do what is your nature, and take no thought for anything else at this time except to recover as comfortably as you can. I am well aware how much you suffer from missing us; but all will be easy if you keep well. I do not want you to hurry, lest you take on the misery of seasickness while still ill and have a dangerous winter passage. I reached the suburbs of Rome on the day before the Nones of January.

\ciceroSection{2} Such was the reception that came out to meet me that nothing more honorific could have been arranged; but I have walked into the very flame of civil discord, or rather of civil war. Although I longed to find a remedy for it, and could have, as I think, the passions of certain men --- for on each side there are those who long to fight --- stood in my way. As it is, Caesar himself, our friend, had sent the senate a letter both menacing and harsh, and was still being shameless in holding on to army and province against the senate's will; and my own Curio was egging him on; while our Antony and Q. Cassius, expelled by no force at all, had set out for Caesar in company with Curio, after the senate had charged the consuls, the praetors, the tribunes of the plebs, and us who hold proconsular rank, to see to it that the commonwealth take no harm.

\ciceroSection{3} Never has the state been in greater danger; never have unprincipled citizens had a leader more prepared. On our side too, in all earnestness, the preparations are being made. This is by the authority and the zeal of our Pompey, who has begun, too late, to fear Caesar. For us, amid these disturbances, the senate, in a full house, has all the same demanded the triumph; but the consul Lentulus, the better to make his own favour the larger, has said that, as soon as he has dispatched what is required for the commonwealth, he will bring the matter on. We for our part press nothing greedily, and for that very reason our authority counts the more. The regions of Italy have been parcelled out, each man's share of the defence. We have taken Capua. I wanted you to know this. You, again and again, see to it that you keep well, and send me letters whenever you have someone to give them to. Once more and again, farewell. Given the day before the Ides of January.
