% Ad Atticum 10.14 (ad-atticum-10-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What a wretched life --- and a greater evil to dread a thing so long than the thing itself when it comes! Servius, as I wrote before, having arrived on the Nones of May, came to me early the next morning. Not to keep you longer: we found no clear way out of any course of action. Never have I seen a man more shaken by fear; and by Hercules, he feared nothing that was not worth fearing --- the one angry with him, the other no friend; the victory of either side a horror, on account of the cruelty of the one, the audacity of the other, and on account of the financial straits of both, which can be relieved nowhere except from the goods of private citizens. And all this he was saying with so many tears that I marvelled they had not dried up after misery so long drawn out. As for me, even this eye-trouble of mine --- on account of which I do not write to you in my own hand --- is without any tears, but it is the more troublesome for the sleepless nights.

\ciceroSection{2} So put together whatever you have by way of consolation, and write me, not the kind drawn from learning and books (that resource is at home; but somehow the medicine proves weaker than the disease), but rather hunt up news of this kind --- about Spain, about Marseilles. What Servius brings on those scores is fair enough; on the two legions, too, he said there were sound authorities. So send me anything of this sort that you have. And in a few days something is bound to be heard.

\ciceroSection{3} But I return to Servius. We put off our talk altogether until tomorrow --- but, slow as he is to set out, he said he would much rather stay in his own little bed, whatever should happen. A tiresome scruple about his son's service at Brundisium. One thing he insisted on most firmly: if the condemned were to be restored, he would go into exile himself. We replied to this both that that very thing was certain to happen, and that what was already happening was no lighter business, and we gathered together many arguments. But these things did not raise his spirit but his fear, so that now it seemed he should rather be kept in the dark about our plan than taken into it. So there is not much in him. As for what you warned me of concerning Caelius, we shall think on it.
