% Ad Atticum 8.12 (ad-atticum-08-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} My eye-inflammation was more troublesome to me even than it had been before. Still, I preferred to dictate this letter rather than give Gallus Fadius --- who is devoted to us both --- nothing in writing to take to you. For only the day before, just as I could, I had myself written that other letter, whose prophecy I want to be false. The reason for this letter, however, is not only that no day may pass without my sending you a letter, but this further and more pressing reason: that I should obtain from you that you take a little time --- the matter is brief --- to lay out your counsel plainly for me, so that I may understand it through and through.

\ciceroSection{2} All my options are still open; nothing has been let slip that does not have a sound excuse, not merely a respectable one. For certainly I did no wrong then, when --- Capua already having been put in my charge --- I declined to accept the charge, escaping not only the imputation of cowardice but also the suspicion of bad faith; nor when, after the terms of peace had been brought through Lucius Caesar and Lucius Fabatus, I took care not to offend the man on whom Pompey was already, with arms in his own hand, conferring the consulship and the triumph of a fellow under arms.

\ciceroSection{3} Nor, indeed, can anyone justly find fault with this latest matter --- that I did not cross the sea. For though it was a matter open to deliberation, still I could not have brought myself to face it. Nor was I bound to suspect what was coming, especially since, from Pompey's own letter, I had no doubt --- and I see you too thought the same --- that he would come to Domitius's aid; and I preferred to spend a little longer thinking what was the right course, and what I had to do.

\ciceroSection{4} First, then, however these matters seem to you, though you have given signs of it, I should like you to write me out in more careful detail; then to look out also a little for what is to come, and to shape the part I should fittingly play, and where you sense I can be of most use to the commonwealth --- whether some role of peacemaker is in demand, or whether everything calls for the man of war.

\ciceroSection{5} And though I measure all things by duty, still I recall your counsels; had I obeyed them, I should not have undergone the bitterness of those times. I remember what you then urged on me through Theophanes and through Culleo, and have often called it to mind with a groan. So now at least let us go back to those calculations which we then threw aside, and use counsels not only honourable but a little more wholesome too. But I prescribe nothing: I should like you to write your view to me carefully in full.

\ciceroSection{6} I want you also to find out, as carefully as you can (and you will have people through whom you can do it), what our Lentulus is doing, what Domitius is doing, what he means to do, how the two of them now bear themselves, whether they accuse anyone, whether they are angry at anyone --- what am I saying, at anyone? at Pompey. Pompey, certainly, puts the whole blame on Domitius, as can be known from his own letter, a copy of which I have sent you. These things, then, you will see to; and, as I wrote to you before, the book of Demetrius of Magnesia on concord which he sent you --- I should like you to send it to me.
