% Ad Familiares 14.14 (ad-familiares-14-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS TERENTIAE ET PATER TVLLIAE, DVABVS ANIMIS SVIS, ET CICERO MATRI OPTIMAE, SVAVISSIMAE SORORI S. R D.}

\ciceroSection{1} If you are well, we are well. It is now for you to decide, no less than for me, what you should do. If that man is going to enter Rome with restraint, you can perfectly well stay in the house for the present; but if a man out of his mind is going to give the city over to plunder, I fear that not even Dolabella himself can help us enough. I am also afraid of this: that by now we may be cut off, so that, when you want to leave, it will not be possible. The remaining question --- which you yourselves will judge best --- is whether women of your standing are still at Rome; for if they are not, you must see to it that you can be in some honourable place. As matters now stand, indeed --- provided we are allowed to hold these regions --- you will be very comfortably placed either with me or on one of our estates. There is this further fear, that in a short time there will be famine in the city.

\ciceroSection{2} On these matters I should like you to consult with Pomponius, with Camillus, with whomever else seems good to you; on the whole, keep up a brave heart. Labienus has bettered the case; Piso too is a help, in that he is leaving the city and condemns his own son-in-law of crime. As for you, my dearest souls, write to me as often as you can, both what you are doing and what is happening there. Quintus, both father and son, and Rufus send their greetings. Farewell. The eighth day before the Kalends of Quintilis, at Minturnae.
