% Ad Atticum 9.14 (ad-atticum-09-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the ninth day before the Kalends I sent you a copy of Balbus's letter to me and of Caesar's to him. Here, on the same day, I received a letter from Capua from Quintus Pedius, with Caesar's letter to him of the day before the Ides of March copied out in these terms: ``Pompey is keeping himself within the town. Our camp is at the gates. We are attempting a great work, the labour of many days, because of the depth of the sea; but still there is nothing better we could be doing. From each horn of the harbour we are throwing out moles, so as either to force him to ferry across what troops he has at Brundisium as soon as possible, or else to shut off his way out.''

\ciceroSection{2} Where is that peace which Balbus wrote he was on the rack for? Could anything be more bitter, anything more cruel? And some fellow has been telling me that Caesar speaks straight out \textit{[Greek: authentik\=os]} of exacting vengeance for Cnaeus Carbo, for Marcus Brutus, and for all those on whom Sulla was cruel with this Pompey as his partner; that nothing he is doing with Curio as his agent has not been done before with Sulla as agent; that $\dagger$by way of canvassing$\dagger$ men for whom exile had not been the penalty under earlier laws have been recalled by him from exile as betrayers of their country; that he complains of Milo, driven out by violence; that he will lay a hand on no one, however, save those who bear arms against him. This is from a certain Baebius, who set out from Curio on the third day before the Ides --- not an inarticulate man, but $\dagger$who would not say to anyone$\dagger$ --- I plainly do not know what to do. From over there I take it Cnaeus has set out by now. Whatever the truth is, in two days we shall know. From you no letter, not even by Anteros; and no wonder. For what is there for us to write about? I, all the same, let no day slip by.

\ciceroSection{3} The letter written, a dispatch was put into my hand before daybreak from Lepta at Capua: that on the Ides of March Pompey had embarked from Brundisium, and that Caesar would be at Capua on the seventh day before the Kalends of April.
