% Ad Atticum 10.7 (ad-atticum-10-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} For my part I approve of your Apulia and Sipontum, and that quiet evasion of yours, and I do not consider your reasoning the same as mine --- not because what is right in public affairs is not one and the same for us both, but because that is not what is at stake. This is a contest about kingship, in which the more modest king has been driven out, and the more upright, and the more sound; the king whom, unless he wins, the very name of the Roman people must be wiped out; whereas if he wins, he will win in Sulla's manner and after Sulla's example. So in this contest you must show no open sympathy for either side, and must give way to the times. My own case is different: I am bound by a kindness and cannot be ungrateful; yet I do not suppose I shall be in the battle-line, but rather on Malta or in some other little town like it. ``You give him no help at all,'' you will say, ``the man you do not wish to be ungrateful to.'' But then perhaps he would have wanted me less. We shall see about that; only let us get away. That we may do so in better weather is the business of Dolabella on the Adriatic side and of Curio on the side of the Strait.

\ciceroSection{2} A certain hope has been cast in my way that Servius Sulpicius wants to confer with me. I have sent Philotimus, my freedman, to him with a letter. If he means to be a man, an admirable companionship \textit{[Greek: synodia]}; but if not ---, we shall be the men we have always been.

\ciceroSection{3} Curio has been with me. He thinks Caesar is on the ground, on account of his unpopularity, and has no confidence in Sicily if Pompey begins to sail. The boy Quintus I received severely. I see there has been greed, and the expectation of a large bounty. This is a great evil, but the crime we had feared has, I hope, been none at all. As for this fault, I take it you judge that it sprang not from any indulgence on our side but from his own nature. Even so we shall manage him by discipline. About the Oppii of Velia, what you think best you will see to with Philotimus. We shall regard our Epirus as ours still; but it had seemed we were going to take other courses.
