% Ad Atticum 9.15 (ad-atticum-09-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} After I had sent off the letter telling you that Caesar would be at Capua on the seventh day before the Kalends, a dispatch was brought to me from Capua $\dagger$and this to me and$\dagger$ that on the fifth day before the Kalends he would be at the Alban estate with Curio. As soon as I have seen him I shall go on to Arpinum. If he grants me the indulgence I am asking, I shall accept his terms; if not, I shall obtain something for myself from myself. He has, as he wrote to me, posted one legion each at Brundisium, Tarentum, and Sipontum. He seems to me to be closing the seaward exits, and yet himself to be looking to Greece rather than Spain.

\ciceroSection{2} But these matters lie further off. For me, what is now pressing is this meeting with him --- and he is at hand --- and I dread his first acts. He will want, I take it, to pass a decree of the Senate; he will want a decree of the augurs (we shall be hauled there, or else, in our absence, harried); or to have the praetor put the question for consular elections, or name a dictator; for neither of these is lawful. And yet --- if Sulla could bring it about, through an interrex, that he was named dictator with a master of the horse, why should this one not be able to? I see no way out, except either, at this man's hands, like Quintus Mucius, or, at that man's, like Lucius Scipio.

\ciceroSection{3} By the time you read this I shall perhaps already have met him. ``Endure, more dog-hearted still'' \textit{[Greek: t\'etlathi k\'unteron]} --- and not even that line is properly ours any more. For there used to be a hope of an early return, there used to be the complaint of the people. Now we are eager to leave, with no hope of return ever entering my head. Not only is there no complaint at all from the country towns and the country folk, but on the contrary they fear him as cruel, as angry. Yet for me nothing is more wretched than to have stayed, nothing more to be wished than to fly out --- not so much for partnership in war as for partnership in flight. $\dagger$But you who$\dagger$ kept putting off all your counsels to the time when we should know what had been done at Brundisium. We know now, surely; we stick fast no less. For I scarcely hope this man will grant me the indulgence I ask, though I bring forward many just grounds for obtaining it. But the whole conversation of his and mine, set down in his words and mine, I shall send to you the moment it has happened.

\ciceroSection{4} You, now, throw the whole force of your love into this: help us with your care and your good sense. Things have rushed on me so suddenly that I cannot even, as I had arranged, see Titus Rebilus; everything must be done by me unprepared. Still, ``some things he himself, some things the divine power will suggest'' \textit{[Greek: \'alla men aut\'os \ldots\ \'alla de kai da\'im\=on hupoth\'esetai]}, as the line has it. Whatever I do, you shall know at once. The mandates of Caesar to the consuls and to Pompey, of which you ask, I have none --- $\dagger$and the copies he brought from along the way$\dagger$ I have sent you already; from those the mandates may be gathered. Philippus is at Naples, Lentulus at Puteoli. About Domitius, please go on as you are doing --- find out where he is, what he means to do.

\ciceroSection{5} As to your writing that I have written more bitterly about Dionysius than my own character would put up with, see what an old-world fellow I am. Heaven help me, I thought you would take this matter more heavily than I do. For beyond the fact that I think you are bound to feel offence at any injury done me by anyone, in this case he in some sort did violence to you yourself when he was so shameless towards me. But at what value you set this is your own judgement; nor in this either am I laying any burden on you. I, for my part, have always thought him not quite in his right mind; now I think him besides foul and vicious, and yet no greater enemy to me than to himself. By Philargyrus you have managed well. You have had a cause indeed both true and good --- that I was left, rather than that I did the leaving.

\ciceroSection{6} When I had sent off my letter on the eighth day before the Kalends, the slaves whom I had sent with Matius and Trebatius brought me a letter in these terms: ``Matius and Trebatius to Cicero \textit{imperator}, greeting. When we had left Capua, on the road we heard that Pompey had set out from Brundisium on the sixteenth day before the Kalends of April with all the troops he had; that Caesar had entered the town the next day, made a speech to his men, and pushed on from there to Rome; that he means to be near the City before the Kalends and to stay there a very few days, then to set out for Spain. It seemed to us not amiss, since we had it for certain about Caesar's arrival, to send back to you your slaves, so that you might know it as soon as possible. Your charges we have at heart, and we shall act on them as the moment calls for. Trebatius is keeping busy to go on ahead. The letter being written, word has been brought us that Caesar will lodge on the eighth day before the Kalends of April at Beneventum, on the seventh at Capua, on the sixth at Sinuessa. This we hold for certain.''
