% Ad Atticum 9.11 (ad-atticum-09-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Do you know that our Lentulus is at Puteoli? I had it from a traveller who said he had recognised him on the Appian Way when Lentulus had opened the curtain of his litter a little. Hardly likely, perhaps; still, I sent boys to Puteoli to track him down, and a letter with them. He was found, scarcely keeping out of sight in his own pleasure grounds, and sent a letter back to me, returning the most extravagant thanks to Caesar; of his own plan, however, he had given his instructions to C.\ Caesius to relay to me. I am expecting him today, that is, on the thirteenth day before the Kalends of April.

\ciceroSection{2} Matius too came to me on the Quinquatria --- a man, by Hercules, as he struck me, level-headed and shrewd; certainly he has always had the reputation of a champion of peace. How plainly he disapproved of all this, it seemed to me, and how he dreaded that \textit{[Greek: n\'ekuia]}, as you call it! In the course of a long conversation I showed him Caesar's letter to me, the one of which I sent you a copy earlier, and asked him to interpret what Caesar meant by writing that he wished to make use of my counsel, my influence, my standing, and my resources of every kind. He answered that he had no doubt Caesar was seeking both my support and my influence for a settlement of peace. May it be granted me, in this calamity of the commonwealth, to accomplish and carry through some statesmanlike work \textit{[Greek: politik\'on]}! Matius for his part was confident that Caesar was of that mind, and undertook to back the proposal himself. The day before, Crassipes had been with me; he said he had set out from Brundisium on the day before the Nones of March and had left Pompey there --- which is what those too who left Brundisium on the eighth day before the Ides reported. But all of them, Crassipes included --- who could, given his shrewdness, take note --- reported menacing talk, hostile to the optimates, hostile to the municipal towns: nothing but proscriptions, nothing but Sullas. Such things as Lucceius was saying, as all Greece was saying --- and Theophanes above all!

\ciceroSection{4} And yet our whole hope of safety lies in those very men, and I lie awake in spirit and take no share of rest, and, to escape these plagues here, I desire to be in the company of men utterly unlike ourselves. For what crime do you think Scipio over there, what Faustus, what Libo, will pass up, men whose creditors are reported to be holding meetings? And what will they do to citizens once they have won? And what \textit{[Greek: mikropsuch\'ian]} of mind on our friend Cnaeus's part! They report that he is thinking of Egypt, of Arabia Felix \textit{[Greek: euda\'imona]}, of Mesopotamia \textit{[Greek: Mesopotam\'ian]}, and that he has now given up Spain. Monstrosities they retail; they may be false, but for certain things here are ruined, and things there are not such as bring safety. By now I am longing for a letter from you. Since our flight there has never yet been so long an interval without one. I have sent you a copy of my letter to Caesar by which I think I may accomplish something.
