% Ad Atticum 10.12 (ad-atticum-10-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What is going to become of me? Or who is more unlucky than I am --- and now even more disgraced? Antony says he has been given specific orders about me, and even so he has not come to see me yet himself; he passed the message through Trebatius. What am I to do now, when nothing comes off, and the things planned most carefully fall out worst of all? I thought I had it all in hand when I got Curio: he had written to Hortensius about me. Reginus was wholly ours. With this fellow we had no suspicion of any trouble at sea. Where am I to turn? I am watched on every side. But enough tears.

\ciceroSection{2} I must steal away, then \textit{[Greek: parakleptéon]}, must creep secretly aboard some merchantman, must not let it appear that I have been stopped by previous arrangement. Sicily is the target; once we make it there, we shall achieve something greater. Only let things go right in Spain! Though as for Sicily itself --- if only the report be true! So far, nothing favourable. There is said to have been a rush of Sicilians to Cato begging him to make a stand, promising everything; that he was moved by it and began holding a levy. I do not believe it --- splendid an authority as he is. That the province could have been held, I do know. But word from Spain will come at any moment now.

\ciceroSection{3} Here with me I have Gaius Marcellus, thinking the same as I do --- truly so, or else giving a good imitation; though I have not seen the man himself, I hear it from a close friend of his. Please, send me anything new, if there is anything; and if we hatch any plan, I shall write you of it at once. I shall handle young Quintus more severely. If only I can make headway with him! Even so, tear up at some point those letters of mine in which I wrote rather sharply about him, in case anything leaks out; I shall do the same with yours.

\ciceroSection{4} I am expecting Servius --- but nothing sound \textit{[Greek: hugiés]} from him. You shall know whatever there is.
