% Ad Atticum 7.16 (ad-atticum-07-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} All your letters, I think, have been delivered to me --- but the first ones out of order, the rest in the sequence in which they were sent by way of Terentia. About Caesar's instructions and the arrival of Labienus and the replies of the consuls and of Pompey I wrote to you in the letter I gave at Capua on the fifth day before the Kalends, and I gathered up a good deal else besides into that same letter.

\ciceroSection{2} For the moment we have these two things to look forward to: first, what Caesar will do when he has received what was given to Lucius Caesar to take back to him; second, what Pompey will do. Pompey, for his part, writes to me that in a few days he will have a sound army, and gives me to hope that, if he himself comes into the Picenum country, we shall be returning to Rome. He has Labienus with him --- a man who has no doubt about the weakness of Caesar's forces; and on his coming our friend Gnaeus has a great deal more spirit. We were ordered by the consuls to come to Capua on the Nones of February. I set out from Capua for Formiae on the third day before the Kalends. That day, when I had received your letter at Cales at about the ninth hour, I dispatched this one at once. As to Terentia and Tullia, I agree with you.

\ciceroSection{3} I had written to them to take you for direction. If they have not yet set out, there is no reason for them to stir until we can see how matters stand.
