% Ad Atticum 7.26 (ad-atticum-07-26) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} It is not now happening to me as you write happens to you, as often as you rise. For I am rising a little now, and for the first time, and most of all thanks to those letters which are coming in from Rome about Domitius, about the cohorts of the Picene country. Everything within the last two days has been made more cheerful. Accordingly the flight that was being prepared has been checked; Caesar's threats --- ``if on a second day's light I find you here'' --- are being rejected; fine news of Domitius, splendid news of Afranius.

\ciceroSection{2} As to your friendly warning, that I keep myself a free hand as long as I can, I am grateful; as to your adding that I should not seem too far inclined toward the dishonourable cause, I can certainly seem so. I have refused to be a leader in civil war for so long as peace was being negotiated --- not that it was not the right course, but because the course that was very much more right was the one that brought me harm. I had simply not wanted to have as an enemy a man on whom our friend would have liked to confer a second consulship and a triumph --- and in what terms! ``for your most ample achievements in office''. I know whom I have to fear, and on what grounds. But if there is to be war, as I see there will be, my part will not be wanting.

\ciceroSection{3} As for the twenty thousand sesterces, Terentia has written you back. As for Dionysius --- while I was supposing we were going to be on the move --- I did not want to be a trouble to him; as for yourself, who write to me again and again about his obligations, I have written nothing back, because I was waiting from day to day, to settle what was to be done. Now, as I see, the boys at least seem likely to winter at the Formian villa. And I? I do not know. For if there is to be a war, I have decided to be with Pompey. What I have settled, I shall make sure you know. I think the war will be the foulest there has been, unless --- as you write --- some Parthian accident should arise.
