% Ad Familiares 2.12 (ad-familiares-02-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. S. D. M CAELIO AEDILI CVRVLI}

\ciceroSection{1} I was indeed anxious about affairs in the city: the public meetings, we were told, were so tumultuous, the Quinquatrus so vexed (for of nearer events we had as yet heard nothing); but nothing was making me more anxious than this, that, in the midst of these vexations, I was not laughing with you at the things that were worth a laugh; for there are many, but I do not dare to write them. This grieves me, that on these matters I have so far had not one letter of yours. So, even though, by the time you read this, I shall already have finished my year's service, I should like your letters to meet me on the way, to school me as to the whole state of the commonwealth, lest I come back a plain stranger. No one can do this better than you.

\ciceroSection{2} Your Diogenes, that modest man, parted from me at Pessinus along with Philo. They were going on to Adiatorix, though they had learned that everything there was neither kindly nor abundant. The City, my dear Rufus, the City: hold to her, and live in her light. All foreign service --- as I have judged it from my earliest manhood --- is obscure and squalid for men whose industry can shine at Rome. Knowing this perfectly well, would that I had stood by my own judgement! With one short walk and one of our conversations, I would not, by Hercules, compare all the rewards of a province.

\ciceroSection{3} I hope I have won the praise of integrity; it was no smaller from a province despised than it is from a province preserved. ``The hope of a triumph?'' you say. I should triumph creditably enough; but at least I should not be so long deprived of what is dearest to me. But, as I hope, before long I shall see you; do you send me letters worthy of you to meet me on the road.
