% Ad Atticum 6.8 (ad-atticum-06-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Just as I had set myself down to write to you and had taken up the pen, Batonius came straight from his ship to me at my house in Ephesus, and delivered your letter, on the day before the Kalends of October. I was glad of the good fortune of your voyage, of Pilia's being to hand, and indeed, by Hercules, glad too of her conversation about my Tullia's marriage.

\ciceroSection{2} But Batonius then brought me sheer terror in the matter of Caesar; he said still more to Lepta. I hope it is all false, but it is certainly horrible: that Caesar will on no account dismiss his army; that he has with him the praetors-designate, the tribune of the plebs Cassius, and the consul Lentulus, all working for him; that Pompey has it in mind to abandon the city.

\ciceroSection{3} But, my dear fellow! Are you not the least bit put out by that man --- who is in the habit of setting himself above the uncle of your sister's son? And by whom he has been beaten! But to the matter.

\ciceroSection{4} The Etesian winds delayed us most violently; the Rhodian open ship cost us a full twenty days. On the Kalends of October, embarking from Ephesus, we put this letter into the hands of Lucius Tarquitius, who was leaving harbour at the same time but sailing in lighter trim. We meant to lie in wait for spells of calm with the Rhodian open ships and the rest of our heavier vessels --- yet hurrying meanwhile as much as ever could be done.

\ciceroSection{5} On the bit of money from Puteoli, thanks. Now I should like you to look about you at the state of things at Rome, and to consider what you think we should make of a triumph --- to which my friends are summoning me. For my part, if it were not for Bibulus --- who, all the while there was a single enemy in Syria, did not set his foot beyond the gate any more than he set it beyond his own front door --- were now exerting himself for a triumph, I should be perfectly calm about it. As things are, though, it would be \textit{[Greek: disgraceful]}. But probe the whole matter, so that on the day we meet we may be able to take counsel. Enough --- I am in a hurry, and I am giving this letter to a man who will arrive either with me or just before me. Cicero sends you his warmest greetings. Greet, in the words of both of us, your own Pilia and our daughter.
