% Ad Familiares 2.11 (ad-familiares-02-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. D. M. CAELIO AEDILI CVRVLI}

\ciceroSection{1} Would you ever have thought it could happen that I should run short of words --- not only of that oratorical kind that is your line, but even of this trivial homespun sort that is mine? They fail me for this reason: that I am marvellously anxious as to what is going to be decreed about the provinces. A strange longing for the city possesses me, an incredible longing for my own people and you above all, and a weariness of the province --- whether because we seem to have attained a reputation in which not so much further increase is to be sought as fortune is to be feared; or because the whole business is not worthy of our strength, who both am able and am wont to sustain greater burdens in the commonwealth; or because the fear of a great war hangs over us, which we look to escape if we depart on the day appointed.

\ciceroSection{2} As for the panthers, the matter is being carefully attended to, on my instructions, by the men who do the hunting; but there is a marvellous scarcity of them, and the ones there are, they say, complain bitterly that no traps are set for anything in my province except themselves. So they are reported to have resolved to leave our province for Caria. But the business is being diligently pursued all the same, and above all by Patiscus. Whatever there is shall be yours; but plainly we know nothing of what it amounts to. Your aedileship, by Hercules, is a great care to me; the very day reminded me of it, for I am writing this on the day of the Megalesia itself. Please write to me as carefully as you can about the whole state of the commonwealth; for I shall reckon those things most certain which I have learned from you.
