% Ad Atticum 7.5 (ad-atticum-07-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I received many letters from you at one time; which, although I was hearing fresher news from those who were coming to me, were welcome all the same, for they declared your zeal and goodwill. Your health worries me, and that Pilia has slipped into the same kind of illness, I sense, adds the greater burden of care to yours.

\ciceroSection{2} Take pains, then, that the two of you get well. About Tiro I see you are concerned. As for him, even though he provides me with marvelous services, when he is in health, in every kind of business or study of mine, still it is for his humanity and modesty that I would rather have him safe than for my own use of him.

\ciceroSection{3} Philogenes has never said a word to me about Luscenius; on the other matters you have Dionysius. That your sister did not come to Arcanum I find strange. About Chrysippus, that my decision meets your approval I am not sorry to hear. I really do not want to be at the Tusculan villa at this season; it lies out of the way for \textit{[Greek: those one meets]} and has other \textit{[Greek: inconveniences]}. From Formiae instead to Tarracina on the day before the Kalends of January; from there to the top of the Pomptine; from there to Pompey's place at Alba. Thus to the City on the third day before the Nones --- my birthday.

\ciceroSection{4} About the commonwealth I grow more afraid by the day. For the loyalists, as people suppose them, are not of one mind. What Roman knights, what senators I have seen, who poured abuse on Pompey for everything else, but above all for this journey of his! Peace is what we need. From victory there will come many evils and surely a tyrant. But of these things shortly, face to face. By now I have nothing left to write to you about; for of the commonwealth there is nothing each of us does not know already, and our domestic affairs are familiar to us both.

\ciceroSection{5} It remains to make jokes, if this man here will allow it. For I am of the school which holds it more useful to concede what he demands than to join battle. We resist too late one whom for ten years we have nourished against ourselves. ``What, then, will be your view?'' you ask. None, of course, but in line with yours, and not even that until I have either finished or laid aside my own business. So take care of your health. \textit{[Greek: Shake off]} at last that quartan ague, by the diligence which in you is supreme.
