% Ad Familiares 8.6 (ad-familiares-08-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} You will certainly have heard by now that Appius has been put on trial by Dolabella --- and to nothing like the storm of ill-will I had expected. Appius has played it shrewdly enough: the moment Dolabella stepped up to the tribunal, he had entered the city and dropped his demand for a triumph. That gesture knocked the talk on the head, and his prosecutor was caught looking less prepared than he had hoped. The man now pins his greatest hope on you. I know that you have no quarrel with him; it is in your power to oblige him as much as you like. If there had been no bad blood between you the whole thing would be simpler. As it is, if you stand on the full strictness of legal right, you will need to take care that you do not look insufficiently frank and open about laying old enmities aside. If you choose to do something gracious in this direction, it is quite safe: no one will say that obligation or friendship deflected you from duty. One thing strikes me: between the demand for an indictment and the actual lodging of the name, his wife has left Dolabella.

\ciceroSection{2} I remember what you charged me with when you left, and I do not think you have forgotten what I wrote you. There is no time to say much more; this much only I can advise you. If the business does not strike you ill, all the same, at this moment do not display any of your real wish in it: wait to see how he comes out of the case. To put it briefly, it will be invidious for you if any hint gets out, and any indication will come out more clearly than is fitting or expedient. He himself will not be able to keep quiet about a thing that has fallen in so handily for his hopes, and that will only become the more conspicuous in the closing of the business --- particularly as the man is one who, even if he knew it was ruinous to speak of the matter, could hardly keep it back.

\ciceroSection{3} Pompey is said to be exerting himself strenuously for Appius --- so much so that people think he will send one or other of his sons out to you. Here all of us are acquitting all comers; in fact, by Hercules, the wickerwork is foul and disgraceful through and through. The consuls are scrupulous to a fault: so far they have managed to pass no senatorial decree except about the Latin Festival. Our Curio's tribunate is frozen solid.

\ciceroSection{4} It is past description how flat everything is. If I were not battling the shopkeepers and the water-board, lethargy would have taken the whole city by now. If the Parthians do not warm you up a bit, we shall freeze to death of cold. All the same, with what he had, and without any Parthians, Bibulus has managed to lose some little band of cohorts in the Amanus. So at any rate the report runs.

\ciceroSection{5} What I wrote above, that Curio is freezing --- well, now he is hot: he is being torn to pieces in the most superheated way. Because he failed of his point about intercalation, in the lightest possible spirit he has gone over to the people and begun to talk for Caesar; and he has flung out a road-building bill not unlike Rullus's agrarian one, and a food-distribution bill that puts the measuring out into the aediles' hands. He had not yet done this when I wrote the first part of this letter. Be a dear: if there is anything you can do for Appius that has to be done, see that I get the credit. As to Dolabella, my advice is, keep your hands free. Both for the thing I mention here, and for your own dignity and your reputation for evenhandedness, that is the line that suits. It will be a disgrace to you if I do not have my Greek panthers.
