% Ad Familiares 2.15 (ad-familiares-02-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. S. D. M. CAELIO AEDILI CVRVLI.}

\ciceroSection{1} Nothing could have been more carefully or more shrewdly handled than what you did with Curio in the matter of the supplication; and, by Hercules, the business turned out to my mind, both for its swiftness and for the fact that the man who was angry --- your competitor, and likewise mine --- gave his assent to the man who decked our successes out with godlike praises. So you may know that I now hope for what is to follow; do you prepare yourself for it.

\ciceroSection{2} That Dolabella is first praised by you, and then even loved, I am glad of; for the things which you hope can be moderated by my Tullia's good sense --- I know what letter of yours they answer. What of it, if you read my own, the one I sent at the time to Appius on the strength of your dispatches? But what is one to do? This is how we live. What is done, may the gods approve. I hope our son-in-law will turn out an agreeable one; in this your humanity will help us much.

\ciceroSection{3} The commonwealth troubles me greatly. I favour Curio; I want Caesar to come off honourably; for Pompey I am willing to die; but for all that nothing is dearer to me than the commonwealth itself --- in which you do not parade yourself overmuch: for you seem to me to be torn this way and that, because you are both a good citizen and a good friend.

\ciceroSection{4} On my departure from the province, I have set the quaestor, Coelius, in charge of it. ``A boy?'' you say. But a quaestor; but a young man of noble family; but as nearly all do. There was no one of higher rank to set over it: Pomptinus had left long before; from my brother Quintus it could not be obtained; and even had I left him, ill-wishers would say that I had not in plain fact stepped down from the province after a year, as the Senate had wished, since I had left another self behind. Perhaps they would add this besides: that the Senate had wished men to be set in charge of provinces who had never been so before, whereas my brother had had charge of Asia for three years. As it is, I am not troubled now; had I left my brother, I should fear everything. Lastly --- not so much of my own motion as on the example of the two most powerful men in the state, who have embraced all the Cassii and Antonii --- I wished, in the case of a young man, not so much to attract him to me as to take care not to alienate him. This decision of mine you must approve, for it cannot now be changed. About Ocella you had written me too little to be clear, and it was not in the dispatches. Your own doings are so well known that even on the far side of the Taurus there has been news of the Matrinius business. As for me, unless the etesian winds delay me at all, I shall, I hope, see you very soon.
