% Ad Familiares 7.32 (ad-familiares-07-32) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. VOLVMNIO.}

\ciceroSection{1} The fact that you sent me a letter without a forename, in the friendly manner you ought, made me first doubt whether it was not from Volumnius the senator, with whom I have a great deal to do; then the urbanity of the writing \textit{[Greek: eutrapelia]} let me see it was yours. In that letter everything was wonderfully pleasing to me except for one thing: that the title to my saltworks is being defended none too diligently by you my agent. For you say that since I left, every man's witticisms --- even Sestius's --- are being fathered on me. What? Do you put up with that? You do not defend me, you do not stand up? I had hoped, for my part, that I had left the genera of my sayings so well marked that they could be recognised of themselves;

\ciceroSection{2} but since there is such dregs in the city that nothing is so witless \textit{[Greek: akythēron]} as not to seem charming to somebody, fight, if you love me, that, unless the ambiguity \textit{[Greek: amphibolia]} is keen, unless the hyperbole \textit{[Greek: hyperbolē]} is elegant, unless the wordplay \textit{[Greek: paragramma]} is neat, unless the surprise turn \textit{[Greek: para prosdokian]} is amusing, unless the other things I went through in book two of the \textit{De Oratore}, through the person of Antonius, about jests, appear artful \textit{[Greek: entechna]} and pointed, you take an oath that they are not mine. As to your complaint about the verdicts, that troubles me much less. Let all the defendants for all I care be dragged off by the heels --- let Selius even be eloquent enough to prove he is a freeman; I am not troubled. The freehold of urbanity --- this, by your leave, let us defend with whatever injunctions there are; in this you are the one man I fear, the rest I despise. You think you are being laughed at: now at last I see that you are wise.

\ciceroSection{3} But, by Hercules, jesting aside: your letter struck me as exceedingly witty and elegant. Those things, however droll they were --- and they were --- did not move me to laugh; for I want our friend, in his tribunate, to have all the gravity he can: not only for his own sake --- I am, as you know, in love with him --- but also, by Hercules, for the sake of the commonwealth, which, ungrateful as she has been towards me, I shall not stop loving. As for you, my Volumnius --- since you have begun, and see that it is welcome to me --- write to me as often as you can about the city, about the commonwealth. Your conversational style on the page is delightful to me. Besides that: Dolabella, whom I look through and judge to be most ardent and most loving towards me, do you exhort and confirm and make wholly mine --- not, by Hercules, because anything is wanting; but because I am keenly bent upon him, I do not feel I take too much trouble.
