% Ad Atticum 6.5 (ad-atticum-06-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} By now you are surely at Rome. If so, I am glad you came safe; for as long as you were away from there, you seemed to me to be more absent than if you had been at home. My own news was less known to me, and so was the city's. So please, although I hope that, by the time you are reading this, I shall have advanced some way on the road, send me letters --- to meet me, frequent, as sharp-witted as may be, on every subject, and above all on the one I wrote to you about before: \textit{[Greek: my wife's freedman, it seemed to me\ldots]} Track this down, as you can, and the more so.

\ciceroSection{2} \textit{[Greek: Coming from the seven-hilled city, he handed over 24 minae and 48 --- a debt owed to Camillus --- and himself owing 24 minae from the Crotonian holdings and 48 from the Chersonesian, and to have inherited 640, 40. Of these not an obol has been paid out, all of it being owed since the first of the second month. His freedman --- the one who shares a name with the father of Conon --- has given the whole matter not a moment's thought. So then, first that everything may be saved, second that you not pass over the interest from the date already set down: how much we have endured at his hands I am very much afraid. For he was here with us, reconnoitring and hoping for something or other; then, despairing for no good reason, he withdrew, throwing out as he went,]} ``\textit{[Greek: I yield --- shame indeed to linger too long ---,]}'' and chided me with the old proverb \textit{[Greek: ``on the one hand\ldots''.]} The rest you can see for yourself, and look into as far as may be done.

\ciceroSection{3} For my part, although I had nearly served out the year (thirty-three days remained), I was being pressed hardest by anxiety for the province. For while Syria was ablaze with war and Bibulus --- in the midst of his great mourning --- was bearing the heaviest part of the burden of that war, and his legates and his quaestor and his friends were writing to me to come to his aid: though my army was weak, with auxiliaries good enough but those Galatian, Pisidian, and Lycian (these being our strength), I judged it my duty to keep the army as close to the enemy as I could for as long as the senate's decree allowed me the command of the province. But what most pleased me was this: Bibulus was no trouble to me; rather he wrote to me on every matter himself. And on me the day of departure was creeping up \textit{[Greek: unobserved]}. When it has come, there is \textit{[Greek: a further problem]}: whom to put in charge --- unless the quaestor Caldus arrives, of whom so far I have nothing certain.

\ciceroSection{4} I wanted, by Hercules, to make this letter longer; but there was nothing to write about, and I could not jest for worry. So fare you well; give my greetings to little Attica and to our Pilia.
