% Ad Familiares 2.18 (ad-familiares-02-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. S. D. Q. THERMO PROPR.}

\ciceroSection{1} That my good offices towards Rhodo, and my other attentions which I have rendered to you and yours, are welcome to you --- a man of the deepest gratitude --- I am exceedingly glad; and you may know that day by day your honour matters more to me. That honour, indeed, you have so enlarged by your own integrity and clemency that there seems nothing left to add.

\ciceroSection{2} But as I think over your situation more and more from day to day, the advice pleases me which I first gave to our Aristo when he came to me: that you would be taking on grave enmities, if a young man, powerful and of noble family, were to be slighted by you. And by Hercules it will, beyond doubt, be a slight: for you have no one of higher rank in office; and the other man, to leave aside his nobility, surpasses your legates --- excellent men, of unblemished character --- on this one ground alone, that he is a quaestor, and your quaestor. That no one in his anger can do you harm I see well enough; but for all that I do not want you to have three brothers angry with you --- men of the highest birth, sharp, not without eloquence, and justly angered too: men whom I see will be tribunes of the plebs in succession over the next three years.

\ciceroSection{3} What sort the times of the commonwealth will be, who knows? To me they look like being turbulent. Why should I want you to walk into the terrors of a tribunician year, especially when you can, without anyone's reproach, prefer a quaestor to legates of quaestorian rank? If he turns out worthy of his ancestors --- as I hope and pray --- some share of the credit will be yours; if he gives offence, the offence will be wholly his, and not in any part yours. What was occurring to me, and what I thought touched your interests, I judged I ought to write you, since I was setting out for Cilicia. As for what you do, may the gods approve it; but, if you will hear me, you will avoid enmities and consult for the peace of your later years.
