% Ad Atticum 7.8 (ad-atticum-07-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What need is there to affirm so very strongly about Dionysius? Would not a nod from you carry conviction with me? But your silence had brought me the greater suspicion, both because you yourself are wont to cement friendships by your own testimonials, and because I was hearing that he had been speaking of me differently with different people. But that the case is just as you write you fully persuade me. And so I am toward him the man you wish me to be.

\ciceroSection{2} I had marked down your fever-day too, from a letter of yours which you had written at the onset of a slight attack, and had noticed that, things being as they are, you could come to me at the Alban villa without inconvenience on the third day before the Nones of January. But, I beg you, do nothing to the inconvenience of your health. What can be so much in a day, or two?

\ciceroSection{3} I see Dolabella, by Livia's will, is among co-heirs in a third with two others, but he is being instructed to change his name. There is a \textit{[Greek: civic problem]} here: whether it is proper for a young man of noble birth to change his name on a woman's will. But that is \textit{[Greek: more philosophical]} a question once we know how much, so to speak, lies in a third of a third.

\ciceroSection{4} What you supposed would happen --- that I would see Pompey before I came up to town --- did happen; for he overtook me at Lavernium on the sixth day before the Kalends. We went together to Formiae and from the eighth hour to evening conferred in private. As to what you ask, whether there is any hope of a peaceful settlement, so far as I could discern from Pompey's long and careful conversation, there is not even the wish for one. He reasons thus: if Caesar, even on dismissing his army, is made consul, there will be \textit{[Greek: a dissolution of the constitution]}; and he thinks moreover that, when Caesar hears that careful preparations are being made against him, he will neglect the consulship this year and will rather hold on to his army and his province. But if Caesar should go mad, Pompey was vehement in his contempt for him, and confident in his own forces and those of the commonwealth. What more would you have? Even though that line \textit{[Greek: Enyalios is impartial]} kept recurring to me, still I was lifted out of my care to hear a brave man, experienced and of supreme authority, discoursing \textit{[Greek: as a statesman]} on the dangers of a sham peace. We had moreover in hand the speech of Antonius delivered before the assembly on the tenth day before the Kalends of January, in which was an indictment of Pompey from the day he took the toga of manhood, a lament for the condemned, the threat of arms. ``With these things,'' Pompey said, ``what do you suppose Caesar himself will do, if he has gained possession of the commonwealth, when his quaestor, weak and resourceless, dares to say such things?'' In short, he seemed not merely not to be seeking that peace, but actually to fear it. He is moved, I think, away from that view of abandoning the City. What weighs most painfully on me is that the money is to be paid back to Caesar, and the apparatus of my triumph must be converted to that end. For it is an \textit{[Greek: ugly thing]} to be debtor to one's \textit{[Greek: political opponent]}. But these things, and many besides, face to face.
