% Ad Familiares 8.13 (ad-familiares-08-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} I congratulate you on your connection by marriage to a man who, on my honour, is of the best --- for so I think of him. The other things, in which he has so far been too little use to himself, both age has now shaken off, and, by your companionship and authority, and by Tullia's modesty, anything that may still remain I am confident will soon be done away with: he is not pugnacious in his vices, nor dull to the perception of what is better. And then --- which is the greatest thing --- I am very fond of him myself.

\ciceroSection{2} You will be glad, Cicero, that our friend Curio has brought to a clean end this business of the veto on the provinces. When the motion was put on the question of obstruction --- the motion was being made by the Senate's own decree --- and the first sentence delivered was that of M. Marcellus, who held that the matter should be taken up with the tribunes of the plebs, a full Senate went against. Pompey the Great is, plainly, so out of sorts and so spiritless now that he can scarcely find what would please himself. They had now come round to the position that the candidacy of a man who would surrender neither army nor provinces should be entertained. How Pompey is going to take this, I shall find out. As to what is to become of the commonwealth, if he does not bestir himself --- well, that will be for you old rich gentlemen to see to. Q. Hortensius, as I write this letter, was at his last gasp.
