% Ad Familiares 2.13 (ad-familiares-02-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. S. D. M. CAELIO AEDILI CVRVLI.}

\ciceroSection{1} Your letters are rare --- perhaps they do not get through --- but how welcome when they come; the most recent above all: how shrewd it was, how full of both kindness and counsel! Even though I had already laid out for myself the very course you were urging on me, all the same our resolutions take firmer shape when we sense that the prudent and the loyal are advising the same thing.

\ciceroSection{2} As to Appius, I am very fond of him, as I have often told you, and I felt that I was being made fond of by him on his side from the moment we set down our quarrel; for as consul he was a man of honour towards me, an agreeable friend, and even encouraging of the very studies that are mine; and that my own good offices did not fail him, you yourself are witness, to whom by now (as I take it) there is added Phania, that \textit{comic-stage witness} \textit{[Greek: k\=omikos martys]}; and by Hercules I have come to value him still more, because I have sensed that you are valued by him. That I am now wholly Pompey's, you know; that Brutus is loved by me, you understand. What reason could there be why I should not have it among my wishes to embrace a man flourishing in years, resources, distinctions, talent, children, kindred, connections, and friends --- my own colleague besides, and, in the very prestige and learning of our college, a partisan of mine? I have written these things at the greater length because your letter gave some hint of a quiet doubt as to my disposition towards him. I take it you have heard something. It is false, believe me, if you have heard anything. The cast of my own arrangements and methods has some unlikeness to his administration of his province; from this some have perhaps suspected that I am at odds with him by clash of feeling, not by difference of view. But I have never done or said anything I would have wish to be against his standing; and since this recent business and the recklessness of our friend Dolabella, I have set myself up as advocate against the danger he is in.

\ciceroSection{3} There was in the same letter the phrase ``the lethargy of the state.'' I was glad indeed, and rejoiced that our friend had frozen over into peace. The bottom of the page pricked me with your own handwriting. What's that you say? Curio defending Caesar now? Who would have thought it, besides me? For --- so may I live --- I did think it. Immortal gods, how I miss our laughter together! I had it in mind, since I had finished the assizes, since I had set the cities on their feet, since I had even kept the arrears of the previous five-year cycle for the tax-farmers without one complaint from our allies, since I had been welcome to private men, both the greatest and the least --- to set out for Cilicia on the Nones of May, and, once I had reached the first summer encampment and placed the soldier in quarters, to depart by decree of the Senate. I am eager to see you aedile, and a marvellous longing for the city possesses me, and for all my own people, and for you above the rest.
