% Ad Familiares 3.7 (ad-familiares-03-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO S. D. APPIO PVLCHRO}

\ciceroSection{1} I shall write to you at greater length when I have come by more leisure; what follows I wrote in haste, when Brutus's slaves had met me at Laodicea and said that they were hurrying to Rome. So I gave them no letters except those to you and to Brutus.

\ciceroSection{2} The ambassadors of the Apameans delivered to me a volume from you full of the most unjust complaint, that I had hindered their building by my letter. In the same letter, moreover, you were asking that I release them as soon as possible, lest they run into winter, for the purpose of building; and at the same time you complained, very sharply, that I had forbidden them to exact the tributes before I had myself, with the matter investigated, given permission --- for this had been, you said, a kind of obstruction, since I could not investigate the matter until I had withdrawn from

\ciceroSection{3} Cilicia toward winter. Take, then, my answer to all this, and recognise the fairness of your expostulation. First: when I had been approached by men who said that intolerable tributes were being exacted of them, what unfairness was there in my writing to your people not to do so, before I had investigated the matter and the case? Not before winter, I take it; for so you write --- as though, indeed, I ought to have gone to them, rather than they to me, for the investigation. ``So far away?'' you say. Well? When you were giving them the letter through which you pressed me not to obstruct them from building before winter, did you not think that they would come to me? Although in that they did make themselves ridiculous: the letter they were carrying, that they might do the work in summer, they delivered to me after the winter solstice. But you should know, both that there are far more men who refuse the tributes than who wish them to be exacted, and that I, all the same, shall do what I take to be your wish. Of the Apameans, enough.

\ciceroSection{4} From Pausanias, the freedman of Lentulus, my orderly, I heard, when he was telling me that you had complained to him that I had not come out to meet you. I despised you, no doubt! No man can be more arrogant than I! When your boy had come to me at about the second watch and had announced that you would be coming to me at Iconium before daylight, and it was uncertain by which road --- there being two --- I sent Varro, your closest friend, by one and Quintus Lepta, my chief of engineers, by the other to meet you. I charged each of them that they should run back to me in time for me to come out and meet you. Lepta came running and brought word that you had already passed beyond the camp. At once I went to Iconium. The rest you know already. Was I to fail to come out and meet you? First, Appius Claudius; next, an imperator; next, in accordance with the custom of our ancestors; next --- the chief point --- a friend! When in matters of that kind I am wont to act with rather more ceremonious eagerness even than my office and my standing require. But enough of this.

\ciceroSection{5} Pausanias also told me this --- that you had said: ``Why not? Appius came out to meet Lentulus; Lentulus came out to meet Ampius; Cicero would not come out to meet Appius?'' Tell me: do even you go in for these trivialities --- a man, in my judgment, of the highest prudence, of much learning besides, of the greatest practical experience, and (I add) of urbanity, which is a virtue, as the Stoics most rightly hold? Do you suppose any ``Appius-ness'' or ``Lentulus-ness'' weighs with me more than the ornaments of virtue? Even when I had not yet attained those things which in men's opinion are the highest, still I never marvelled at those names of yours; I judged the men who had bequeathed them to you to be great. But after I both took up and discharged the greatest commands in such a way that I judged there was nothing further for me to acquire either for honour or for glory, I hoped that I had been made, not your superior --- certainly not --- but your equal. And, by Hercules, I have not seen Gnaeus Pompey, whom I rank above all who have ever been, or Publius Lentulus, whom I rank above myself, judge otherwise. If you judge otherwise, you will not go far wrong if you turn your attention a little more diligently --- so as to understand what \textit{[Greek: eugeneia]} (good birth), what nobility is --- to what Athenodorus, son of Sandon, says on these matters.

\ciceroSection{6} But to return to the point: I should like you to think me not only your friend, but your closest friend. By all my services I shall surely bring it about that you can truly so judge. If, however, what you are aiming at is to seem to owe me less, while I am away, than I have laboured on your behalf, I let you off that concern: ``\textit{[Greek: par' emoige kai alloi hoi ke me timesousi, malista de metieta Zeus]}'' (with me are others too who will honour me, and chief among them Zeus the counsellor). But if by nature you are \textit{[Greek: philaitios]} (fond of fault-finding), this you will not achieve --- you will not make me wish you less well; this much you will compass: that you take it less to heart in whatever way you receive it. I have written to you the more freely, relying on the consciousness of my own good service and goodwill, which, having been taken up by me on settled judgment, I shall keep up so long as you wish.
