% Ad Familiares 3.10 (ad-familiares-03-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO APPIO PVLCHRO S.}

\ciceroSection{1} When the news was brought to us of the recklessness of those who were stirring up trouble for you --- although at the first report I was deeply shaken, in that nothing could have happened more against my expectation --- still, once I had collected myself, the rest seemed to me the easiest thing in the world; for I held the highest hope in yourself and a great hope in your people, and many things came into my mind on account of which I judged that this very labour would even bring you honour. One thing, plainly, I took hard: that I saw the most assured and most rightful triumph wrenched from you by this conspiracy of the envious. But this, if you take it at the value at which I have always judged it must be taken, you will take wisely, and as victor you will conduct, out of the grief of your enemies, the most rightful triumph of all. For I see plainly that, by your sinews, your resources, your wisdom, it will come to pass that your enemies will keenly regret their own want of restraint. Of my own part, calling all the gods to witness, I promise and confirm to you that for your standing (I prefer to put it so, rather than ``for your safety''), in this province over which you presided, I will take upon myself the duties and the role of an intercessor by my pleading, of a kinsman by my labour, of a man dear --- I hope --- by his authority among the communities, of a commander by his weight. I want you to demand and to expect everything from me; I shall outdo your imaginings by my services.

\ciceroSection{2} Q. Servilius delivered me a very short letter from you, which yet seemed to me too long; for I felt I was being wronged in being asked at all. I could wish the occasion had not arisen on which you could see clearly how much I valued you, how much Pompey --- whom alone of all I value, as I ought, most of all --- and how much Brutus (though in the daily intercourse of life you would have seen it, as you will see it); but, since it has arisen --- if anything is omitted by me, I shall confess a crime committed and a disgrace incurred.

\ciceroSection{3} Pomptinus, whom you treated with outstanding and singular loyalty (a kindness of yours of which I am the witness), is paying back the memory and goodwill he owes you. For when, driven by his own most pressing affairs, he had taken his leave of me much against my will, still --- as soon as he saw it concerned you --- with the ship even now being boarded at Ephesus, he turned back to Laodicea. When I see that you will have innumerable acts of zeal like this, plainly I cannot doubt that that high standing of yours will be your due. If indeed you bring it about that censors are elected, and if you so conduct the censorship as you ought and can, I see that not only to yourself but to all your people you will be a sovereign protector for life. Of this one thing, fight and exert yourself: that nothing be added to my term out here, so that, when we have done enough for you in these parts, we may give our goodwill toward you scope to do its work where you are.

\ciceroSection{4} What you write to me of the zeal toward you of men and of all orders is something that surprises me not at all, but pleases me very much; and the same has been written to me in detail by my own intimates. So I take great pleasure in the fact, both that to you --- whose friendship is to me not only ample but a delight --- those tributes are being paid which are owed; and indeed that there should still remain in our state, by what is almost the consensus of all, those forms of zeal toward brave and industrious men which have been the one reward always paid to me too for my own labours and my own vigils.

\ciceroSection{5} But this, indeed, struck me as truly astonishing: that there was so much rashness in that young man, whose own safety I defended in two capital trials with the utmost effort, that, in taking up enmities against you, he should forget almost everything of his own fortunes and reasonings --- especially when you abounded in every ornament and every safeguard, and he himself (to put it most mildly) lacked many things. His foolish and childish talk had been already reported to me in detail by our intimate M. Caelius; and much of the same sort of talk has been written about by you. For my part, I would more readily have severed an old connection with the man who had taken up enmities against you than have brought a new one into being; for you ought not to doubt my zeal toward you, and there is no one in the province, nor was there at Rome, to whom this has been obscure.

\ciceroSection{6} But there is signified, all the same, in your letter a certain suspicion and hesitation of yours, of which it is no time for me to expostulate with you about now, but only necessary for me to clear myself. For where have I ever been an obstacle to any embassy being sent to Rome to your praise? Or in what could I --- if I openly hated you --- have done less of what could harm you, or, if I hated you in secret, have shown my enmity more openly? But if I had been of that perfidy of which those are who report these things against me, still I would surely not have been of that stupidity --- to add to my secret hatred openly declared enmities, or, in a matter in which I could harm you not at all, to display my supreme desire to harm. Certain men did come to me, I recall, from the Epictetus region, saying that excessive expenses were being decreed to the embassies. I did not so much order them, as recommend, that the expenses to the embassies be assessed as nearly as possible to the Cornelian law; and that I did not even hold to that, the accounts of the communities are themselves the witnesses: in which, as much as each community wished, was paid out to your ambassadors --- and is on the books.

\ciceroSection{7} But you --- with what lies have the most worthless of men loaded you down! That not only were the expenses cut off, but that they were even reclaimed and taken back from the agents of those who had already set out, and that this was the reason many at any rate had not gone! I would expostulate with you, and complain --- did I not, as I wrote above, prefer at this moment of yours to clear myself than to accuse you, and judge that the more correct course. So I shall say nothing on your account, of what you have believed; on my own, of why you ought not to have believed it, a few words. For if you hold me established as a good man, as worthy of the studies and the learning to which I have given myself from boyhood, as of mind sufficiently great and of judgment not slight in the greatest matters, you ought to recognise in me not only nothing perfidious, treacherous, or false in friendship, but nothing even mean or paltry.

\ciceroSection{8} But if it pleases you to picture me as cunning and secretive, what is there that could less fall in with a nature of that sort than either to scorn the goodwill of a man in his fullest flourishing, or to attack his standing in the very province whose praise of him one had defended at home, or in a matter in which one could do no harm to show an enemy's mind, or to choose for one's perfidy the very point on which it would be plainest in declaring hatred, slightest in inflicting harm? And what reason was there I should have been so implacable toward you, when I had learned from my brother that not even then, when it was almost a necessity for you to act the part, had you been my enemy? But when, indeed, each of us had sought our reconciliation, what in your consulship did you ask of me in vain that you wished me either to do or to think? What did you commission of me, when I escorted you to Puteoli, in which I did not surpass your expectations by my attentiveness?

\ciceroSection{9} But if this is what is most characteristic of the cunning man --- to refer everything to his own advantage --- what then was more useful to me, what better suited to my own interests, than connection with a man of the highest nobility and the highest honour, whose resources, talent, children, kin, and relations could be to me the greatest ornament or safeguard? All of which, however, I did pursue in seeking out your friendship not by any cunning, but by a certain wisdom. What of those bonds by which I am most gladly bound --- how strong they are! the likeness of our pursuits, the pleasantness of our intercourse, the delight of life and of our way of living, the sharing of conversation, the deeper letters! And these are domestic things. What of those public matters: the illustrious return to goodwill, in which one cannot stray even by inadvertence without suspicion of perfidy; the most ample college of priests, in which not only was the violation of friendship in our ancestors' time held impious, but it was not even permitted to co-opt as a priest anyone who was the enemy of any member of the college?

\ciceroSection{10} But to leave aside those things, many and great as they are --- whom has anyone ever valued, or could or should have valued, so highly as I have valued Gnaeus Pompey, your daughter's father-in-law? For if services count, I think my country, my children, my safety, my standing, my very self restored to me through him; if the pleasantness of intercourse, what consular friendship in our state was ever more closely joined? if the marks of love and obligation, what did he not entrust to me, what did he not share with me, in what matter touching himself --- when he himself was absent from the Senate --- did he prefer business to be conducted through any other than me? in what dignities did he not wish me distinguished to the highest? with what ready good humour, with what humanity, did he bear my struggle on Milo's behalf, opposing at times his own measures? with what zeal did he see to it that none of the odium of that time should touch me, when he sheltered me with his counsel, with his authority, and with his arms at the last? in which times indeed there was this gravity in him, this loftiness of mind: that he would not, not only to some Phrygian or Lycaonian (which is what you did with the legates), but not even to the malicious talk of the highest men against me, give any credit. Since, then, his son is your son-in-law, and since --- besides this connection of \textit{adfinitas} --- I know how dear, how delightful you are to Gnaeus Pompey, with what mind at last ought I to be toward you? --- especially since he has sent me letters of such a kind that, even if I were the enemy of yourself, to whom I am the closest friend, I would still be appeased, and would turn the whole of myself toward the will and the nod of that man who has so deserved of me.

\ciceroSection{11} But enough of this; for these things have been written perhaps in more words than was necessary. Now learn what has been put in motion by me and is being prosecuted; and these things we are pursuing, and shall pursue, with regard rather to your standing than to your danger. For you, as I hope, we shall be hearing called censor any day now --- whose duties of office, requiring as they do the highest spirit and the highest counsel, I think ought to be more carefully and more attentively pondered by you than what we ourselves are pursuing on your behalf.
