% Ad Familiares 16.1 (ad-familiares-16-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS TIRONI SVO S. P. D. ET CICERO MEVS ET FRATER ET FRATRIS F.}

\ciceroSection{1} I thought I could bear missing you a little more easily; but plainly I cannot bear it at all, and although it matters greatly to my standing that I reach the city as soon as possible, I feel all the same that I did wrong to leave you. But because your own wish seemed to be that you would on no account sail unless your strength was assured, I approved your decision, and I do not change it now if you are still of the same mind. If, however, now that you are taking food, you think you can catch up to me, that is for you to judge. I have sent Marion to you for this purpose: either to come back to me with you as soon as possible, or, if you are still delayed, to return to me at once.

\ciceroSection{2} But take it for certain: if it can be done without harm to your health, there is nothing I want more than to have you with me. If, on the other hand, you understand that you need to stay on a little at Patrae for the sake of getting well, there is nothing I want more than that you be well. If you sail at once, you will catch us at Leucas. If you mean to firm up your strength, you will see to it carefully that you have proper companions, the right weather, and a sound ship. One thing, my dear Tiro, see to, if you love me: do not let Marion's arrival or this letter sway you. If you do what is most conducive to your health, you will most fully be obeying my wish.

\ciceroSection{3} Weigh these things with that good sense of yours. We miss you so much because we love you; the love urges us to see you well, the missing to see you as soon as possible --- so the former wins. Take care, then, above all, to be well. Of your countless services to me, this will be the most welcome. The third day before the Nones. The Nones.
