% Ad Familiares 2.17 (ad-familiares-02-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMR S. D. †CANINI SALVSTIO PROQ.}

\ciceroSection{1} Your orderly delivered me a letter from you at Tarsus on the sixteenth day before the Kalends of August. I will answer it in order, as you seem to wish. About a successor for me I have heard nothing, nor do I think there will be one. There is no reason why I should not step down on the day appointed, especially with the Parthian alarm removed. I do not really suppose I shall linger anywhere; I think I shall touch at Rhodes for the sake of the boys, the two Ciceros, but even that is not fixed. I want to come to Rome as soon as possible; but my journey will be governed by the state of the commonwealth and of business in the city. Your successor cannot make such haste that there is any way you could meet me in Asia.

\ciceroSection{2} On the rendering of accounts, it was not inconvenient that you should render none, since (as you write) Bibulus is giving you that liberty; but you scarcely seem to me able to do that under the Julian law, which Bibulus, on some private calculation of his own, does not observe, and which I think you ought most certainly to observe.

\ciceroSection{3} As for what you write, that the garrison at Apamea ought not to have been withdrawn, I saw that others took the same view, and I bore it ill that the talk of malevolent men on that matter ran less in my favour than was convenient. Whether the Parthians have crossed or not, I see no one but you in doubt. Accordingly, all the garrisons --- large and firm --- which I had set up, I dismissed, shaken as I was by the unhesitating talk of men around me.

\ciceroSection{4} The accounts of my quaestor it would not have been right for me to send you, nor were they made up; we were thinking of depositing them at Apamea. Of my plunder no one apart from the city quaestors --- that is, the Roman people --- has touched, nor will touch, so much as a farthing. At Laodicea I expect to take sureties for all the public money, so that both I and the Roman people may be secured against the risks of transport. As to what you write to me about three hundred and thirty-three thousand drachmas, there is nothing in that line that I can do for anyone; for all the money is so handled that the plunder is in the charge of the prefects, while what has been assigned to me is in the charge of the quaestor.

\ciceroSection{5} As to what you ask --- what I think about the legions decreed for Syria --- previously I was in doubt whether they would come; now I have no doubt that, if word goes ahead that quiet has been restored in Syria, they will not come. As for Marius, my successor, I see he will come slowly, for this reason: that the Senate has decreed that he is to march with the legions.

\ciceroSection{6} I have answered one letter; I come to the other. You ask me to commend you as warmly as possible to Bibulus. In that matter goodwill on my part is not wanting, but there seems to be ground for a quiet remonstrance with you. For you alone of all those who are with Bibulus never gave me notice of how fiercely, and without cause, Bibulus's goodwill recoiled from me. Very many men have reported to me that when there was great fear at Antioch and great hope was placed in me and in my army, he was wont to say he would rather endure anything than seem to have stood in need of my help; which behaviour --- since I supposed you were drawn by your duty as quaestor to keep silent about your praetor --- I bore without offence, although I was hearing the way you yourself were being treated. As for him, when he wrote to Thermus about the Parthian war, he never sent a letter to me, to whom he understood the danger of that war belonged; he wrote to me only about the augurate of his son; in which case I, moved by compassion, and because I had always been Bibulus's warmest friend, took pains to write to him as kindly as I could.

\ciceroSection{7} If he is malevolent toward all men --- something I had never supposed --- I take less offence in my own case; but if he is more particularly estranged from me, my letter will be of no use to you; for in the dispatches Bibulus sent to the Senate, what was shared between him and me he assigned to himself alone: he says that he made it his business that the exchange of money should be carried through with a profit to the public; and what was strictly my own --- that I refused to use the auxiliary Transpadane levies --- this too he writes as if he himself had granted to the people; whereas what was his alone he shares with me: ``We made,'' he says, ``with the auxiliary cavalry, larger requisitions of grain.'' But this is the act of a small mind, lean and empty with very malevolence: that Ariobarzanes, since the Senate through me styled him king and commended him to me, he in his dispatches calls not king, but son of King Ariobarzanes. Men of this disposition grow worse when they are asked. But I have indulged you; I have written him a letter, and when you have received it you will do what you wish.
