% Ad Familiares 8.14 (ad-familiares-08-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} It was not worth taking Arsaces and storming Seleucia, only to miss being a spectator at the things that have been done here. Your eyes would never have ached again, if you had seen the look on Domitius's face at his defeat. They were great elections, and quite plainly partisan: feeling broke along the party lines. Very few followed kinship and discharged the obligation. So Domitius is now my bitterest enemy --- he does not hate even a familiar friend of his own as he hates me, all the more because he reckons the augurate, of which I had been the promoter, snatched from him by injury. Now he is in a fury that men so rejoiced in his pain --- and \dots more enthusiastic for Antony. For Cn. Saturninus, a young man, has been put up by Cn. Domitius himself --- a man much exposed by his earlier life --- and the trial of this defendant is now in expectation, and even in good hope, after the acquittal of Sex. Peducaeus.

\ciceroSection{2} On the high politics: I have often written to you that I see no peace for the year, and the closer that struggle approaches --- which it must --- the more clearly that danger shows itself. This is the issue on which the men who hold the controlling hand are going to fight it out: that Cn. Pompey has resolved not to allow C. Caesar to become consul on any other terms than that he hand over his army and his provinces, while Caesar has persuaded himself that he cannot remain safe if he steps down from his army. He proposes, however, this condition: that they both hand over their armies. Thus that famous love-affair and odious conjunction has not subsided into private grumbling, but has burst out into war. As for my own affairs --- I cannot find what counsel to take. This deliberation, I do not doubt, will trouble you as well: I have ties of favour and obligation with these men on the one side, and on the other I hate the cause, not the men.

\ciceroSection{3} It cannot, I think, escape you that in a domestic quarrel men ought, so long as the struggle is conducted by civil means without arms, to follow the more honourable side; once it has come to war and the camp, the stronger --- and to judge the better course the one that is safer. In this discord I see that Cn. Pompey will have the Senate, and those who sit in judgement, on his side; that to Caesar will come over all those who live in fear or in bad hope. Army to army there is no comparison. Altogether, there is room and time enough for examining the forces of each, and choosing one's side.

\ciceroSection{4} I have nearly forgotten what most needed to be written. Did you know that Appius our censor is producing prodigies here --- prosecuting most fiercely on statues and pictures, on assessment of land, on debt? He has it in his head that the censorship is a face-cream or a piece of soap. He is in error, I think: he means to wash off the grime, but he is opening up all his veins and his vitals to view. Run, by gods and men! and come at the earliest to see this laughed at: the Scantinia-law case at Drusus's bench, and Appius prosecuting on pictures and statues. Believe me, it is to be hurried for. Our friend Curio is reckoned to have done a wise thing in his concession on the matter of the pay for Pompey. To sum up: you ask what I think will come of it. If one or the other of these men does not go to the Parthian war, I see great discords hanging over us, which the sword and main force shall decide; either man is ready, both in spirit and in resources. If it could come about without the highest peril, Fortune was preparing for you a great and welcome spectacle.
