% Ad Atticum 6.7 (ad-atticum-06-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Quintus the son --- dutifully enough, and with much urging on my part, certainly, though it was a willing horse I was spurring --- has reconciled his father's heart to your sister. Your letter has stirred him greatly. In short: I trust that matter is as we wish. I have written to you twice before about my private affairs, if the letters have reached you, in Greek and \textit{[Greek: in riddles]}. Plainly, nothing is to be set in motion; but still, by inquiring \textit{[Greek: plainly]} into Milo's debts, and by urging him to make good what he undertook to me, you will do some good.

\ciceroSection{2} I have given orders for the quaestor Mescinius to wait for me at Laodicea, so that I may deposit my accounts, made up under the Julian Law, with two cities. I want to go to Rhodes for the boys' sake, then to Athens as soon as possible --- though the etesian winds blow hard against it. Even so, I positively want to be there under these magistrates, whose goodwill I have proved in the matter of the supplicatio. Still, please send letters to meet me, in case you think we ought to be delayed on any account of public business. Tiro would have given you a letter, had I not had to leave him gravely ill at Issus. They report him better. I am wretched, all the same; for there is no more chaste young man, no more diligent.
