% Ad Atticum 6.6 (ad-atticum-06-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} While in the province I am dressing Appius up in every kind of distinction, I have suddenly been made the father-in-law of his accuser. ``May the gods approve it!'' you say. So I myself wish, and I am sure that is your wish too. But believe me, I had thought nothing less likely --- I, who in the matter of Tiberius Nero (who had been negotiating with me) had sent trusted men to the ladies, men who came to Rome to find the betrothal already made. Still, I hope this match is the better one; for I gather the ladies are quite delighted with the young man's obligingness and charm. The rest --- do not \textit{[Greek: prickle over]}.

\ciceroSection{2} But mind you! \textit{[Greek: A wheat-dole to the people]} at Athens? Does that sit well with you? Though I do confess that my own books at least did not stand in the way --- for that was not a largesse to citizens, but generosity to friends. Even so, you bid me think about the \textit{[Greek: propylon]} of the Academy, when by now Appius is not thinking about his Eleusinion? About Hortensius, I am sure you grieve; I am tormented; for I had decided to live with him in close intimacy.

\ciceroSection{3} I have put Coelius in charge of the province. ``A boy,'' you say, ``and perhaps a fool, and not steady, and not self-controlled!'' I agree --- but it could not have been done otherwise. For those letters of yours which I had received long since, in which you wrote that you yourself were \textit{[Greek: holding back]} over what I ought to do about the leaving, kept pricking me; for I could see what your reasons were for \textit{[Greek: holding back]}, and they were also mine. Hand it to the boy? But to my brother? That would not be in our interest. For apart from my brother there was no one I could rank above the quaestor --- and a quaestor of noble family at that --- without insult. Still, while the Parthians seemed to be hanging over us, I had decided either to leave my brother, or even, for the public good, to stay on myself in contravention of the senate's decree. After they withdrew, with that incredible piece of luck, the doubt was lifted. I could see the talk: ``Oho! he has left his brother! Is that any way of holding the province for no more than a year? And what about the senate's wish that those should preside over the provinces who had not done so before? Yet this man, three years!'' --- so much for the popular case.

\ciceroSection{4} And what about the case in your own ear? I should never be free of care --- if there were some piece of irritability, of insult, of negligence, such things as human life throws up. What if my son the boy, the boy with great confidence in himself, did something? What pain would that be? My brother would not let him go, and he was hurt that you thought he should. As it is, with Coelius --- I don't say what he has done ---, but I worry far less. Add this. Pompey, a man of such strength, with such roots, chose Quintus Cassius without the lot; Caesar, Antonius. Should I take offence with one given me by the lot, so as even to set on foot an inquiry into the one I had left behind? This is better, and there are more precedents for it; for our time of life, certainly, it is more becoming. As for you, with him --- good gods, in what favour I have set you! And I read to him your letter --- not yours, but your secretary's. Letters from my friends call me to a triumph, a thing not to be neglected by me, as I judge, for the sake of this \textit{[Greek: rebirth]} of ours. So you too, my Atticus, begin to want the same, that we may seem the less foolish.
