% Ad Familiares 15.5 (ad-familiares-15-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CATO S. D. M. CICERONI IMP.}

\ciceroSection{1} What both the public interest and our friendship urge upon me, I do gladly: I rejoice that the integrity, the unblemished conduct, the diligence which were known of you in matters of the highest weight at home, in civilian dress, you now administer with equal industry abroad under arms. Accordingly, so far as it lay in my power to act on my judgement, I gave you, both by my vote and by my expressed opinion, the praise that was due --- that by your unblemished conduct and counsel the province was defended, the kingdom of Ariobarzanes preserved together with the king himself, and the goodwill of our allies recalled to a zeal for our rule. That much I have done.

\ciceroSection{2} As for the decreed thanksgiving: if you would rather that we render thanks to the immortal gods for what was secured for the commonwealth by no chance but by your own supreme prudence and self-restraint, than that we credit the matter to your account --- then I am glad. But if you reckon a thanksgiving the foretaste of a triumph, and prefer for that reason that fortune be praised rather than yourself, then I have two things to say: a triumph does not always follow upon a thanksgiving; and it is a much more illustrious thing than a triumph for the Senate to give it as their judgement that a province has been held and kept safe by the mildness and unblemished conduct of its commander, rather than by force of soldiers or by the favour of the gods. That, in my own vote, is what I declared.

\ciceroSection{3} And I have written this to you at greater length than is my custom for this reason: that --- which I most want --- you should reckon I am taking pains to persuade you both that I wished, in the matter of your honour, what I judged to be most ample, and that I rejoice that what you preferred has come to pass. Farewell, and have us in your affection; and on the journey you have set in train, maintain for our allies and for the commonwealth your strictness and your diligence.
