% Ad Familiares 3.9 (ad-familiares-03-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO APPIO PVLCHRO S.}

\ciceroSection{1} At last, with difficulty, I have read a letter worthy of Appius Claudius --- full of humanity, of obligation, of attentiveness. The sight of the city, plainly, has restored you to your old urbanity; for the letters you sent me from the road, before you had left Asia --- the one about ambassadors I had forbidden to set out, the other about the Apameans' building obstructed --- I read with the utmost reluctance, and I wrote back to you in some annoyance, in the consciousness of my unchanging goodwill toward you. But when I had read the letter you gave to my freedman Philotimus, I came to understand --- and to recognise --- that there had been many in the province who would have wished us to be on other terms than those on which we in fact stand; whereas you, when you came to the city, or rather when first you saw your own people, learned from them with what loyalty I had stood by you in your absence, and with what scrupulous and steady regard I had kept up every observance toward you. So at how high a price do you suppose I value what is written in your letter --- that if anything should arise affecting my standing, you would return me a like favour, even though it could hardly be done? You will find it easy: there is nothing zeal and goodwill --- or rather love --- cannot accomplish.

\ciceroSection{2} For my part, although I formed the same opinion myself, and was constantly being kept informed by my people by letter, still I took the greatest pleasure from your letter, in the prospect, by no means in doubt and now plainly settled, of your triumph; and not for this reason --- that I might the more easily attain mine myself (for that is an \textit{[Greek: Epikoureion]} (Epicurean) consideration) --- but, by Hercules, because your standing and eminence are dear to me for their own sake. Wherefore, since you have more people than the rest whom you know to be setting out for this province --- since nearly all of them call on you, in case there is anything you wish --- you will do me the greatest favour if you send me a letter the moment you have attained what you yourself look for with confidence and I desire. The judgment of ``the long bench'' --- as our friend Pompey calls it --- and the delay, if it takes from you a day or two more (for what more can it take?), will leave your standing in its proper place all the same; but if you love me, if you wish to be loved by me, send me a letter, that I may be touched with the joy as soon as may be.

\ciceroSection{3} And I should like you to pay me the rest of what is owed of your promise and gift. While I desire to gain the knowledge itself of augural law, by Hercules I am also incredibly pleased by your kindnesses toward me and your gifts. As for your asking something of the same sort from me --- it really does need thought, in what way most fittingly to pay you back in return. For surely it is not for me --- who, as you are wont to wonder at, put so much labour into writing --- to commit the fault of seeming to have been careless in writing; especially when that would not only be the charge of a careless mind, but even of an ungrateful one.

\ciceroSection{4} But of that we shall see. As for the thing you promise: I should be glad if, by your loyalty and your diligence, and for the sake of our friendship --- not newly set up, but now of long standing --- you would take pains and strive that a supplicatio be decreed me in the most honorific way and at the earliest possible moment. I sent the despatch, on the whole, later than I had wished (in which both the difficulty of sailing was vexing, and I believe my letter fell on the very break-up of the Senate); but I did so prompted by your authority and counsel, and I think I did rightly, in that I sent the despatch not at the moment I was hailed as imperator, but after I had added further actions and finished the summer campaigns. These matters, then, will be your care, in the manner you indicate; and you will hold me, with all that is mine and all my people, in your keeping.
