% Ad Atticum 7.9 (ad-atticum-07-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} ``Are letters,'' you say, ``to be received from you daily?'' If I have someone to send by, then daily. ``But by now you yourself are at hand.'' Then when I have arrived, I shall stop. There is one letter from you, I see, that has not been delivered to me: the one which my close friend Lucius Quinctius was wounded and stripped of, near the tomb of Basilus, while carrying it.

\ciceroSection{2} You will therefore see whether there was anything in it I need to know, and at the same time \textit{[Greek: work out]} for me this \textit{[Greek: thoroughly civic]} \textit{[Greek: problem]}. Given the necessity is that either Caesar's candidacy is to be entertained, while he holds his army, by Senate or by tribunes of the plebs; or Caesar is to be persuaded to hand over his province and army and so be made consul; or, if he cannot be persuaded to this, the elections are to be held without his candidacy, with himself acquiescing and holding on to his province; or, if through tribunes of the plebs he does not allow this, and yet remains quiet, the business is to be brought down to an interregnum; or, if for that cause --- that his candidacy is not entertained --- he brings up his army, war is to be waged with him in arms; further, that he is the one to begin the war either at once, while we are less prepared, or at the moment when his friends, demanding at the elections that under the law his candidacy be entertained, fail to secure it; further, that his going to arms is either on this one ground that his candidacy is not entertained, or on the added ground that a tribune of the plebs, obstructing the Senate or stirring up the People, has been censured or hemmed in by senatorial decree or removed or expelled, or, claiming to have been expelled, has fled to him; further, that, war once undertaken, either the City is to be held, or, abandoning it, he is to be cut off from supplies and the rest of his resources --- of these evils, of which one at least must be undergone, which do you reckon least? You will say, surely, that he is to be persuaded to hand over his army and so be made consul. The matter is altogether such that, if he came down to that, no objection could be made; and I am amazed that, if he cannot obtain that his candidacy be entertained while he keeps his army, he does not do precisely this. To us, however, as some think, nothing is to be more feared than him as consul. ``Yet I prefer it so,'' you will say, ``to him with his army.'' Certainly; but that very ``so'' is what someone, I know, takes to be a great evil, and there is no remedy for it: we must give way, if he so wills it. ``See again as your consul the man you have seen as consul before.'' ``But then,'' he says, ``the weak man counted for more than the whole commonwealth. What do you think now?'' And, with him as consul, Pompey for certain is to be in Spain. O wretched business! --- when the very worst is what cannot be refused, and when, if he were to do this thing, he would at once win supreme favor with all loyal men.

\ciceroSection{4} Let us, then, set aside the option Caesar refuses to be brought to; of the rest, what is the worst? To concede to him what, as he himself says, he most impudently demands. For what is more impudent? ``You have held a province for ten years, granted not by the Senate but by yourself, by force and by faction. The term, not of the law but of your willfulness, has passed --- but let us say, of the law; a successor is decreed; you obstruct, and say, `Entertain my candidacy.''' Entertain ours. Are you to hold an army longer than the People ordered, against the Senate's will? You must fight it out, then, if you do not yield --- with good hope, as he too says, either of winning or of dying in liberty. Now if we must fight, at what time, by what chance, by what plan: that depends on the circumstances. And so I am not putting you to the test on that question; bring to what I have said whatever you have. As for me, I am tortured day and night.
