% Ad Familiares 8.7 (ad-familiares-08-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} How quickly you want to be done with that posting I do not know; for my part, the more successfully you have so far managed things there, the more, while you are still there, I am tortured by the prospect of a Parthian war --- afraid that some fresh alarm should ruin my present good cheer at your expense. I have given these few lines in a rush to a courier of the tax-farmers who was just leaving; one with more in it I sent yesterday by your freedman.

\ciceroSection{2} Nothing in the way of real news has happened, unless you want me to write you what you certainly do want. Young Cornificius has betrothed himself to Orestilla's daughter. Paula Valeria, sister of Triarius, has divorced her husband, for no reason, on the day he was due home from his province; she is to marry D. Brutus. [corrupt: ``you had not yet sent back'' or sim.] Many incredible things of this kind have gone on while you have been away. Servius Ocella could have persuaded no one that he was an adulterer, had he not been caught in the act twice in three days. You will ask where. By Hercules, the very place where I should least have chosen. I leave you to discover the rest from someone else; it actually rather appeals to me that a commander-in-chief should be obliged to ask each man in turn who the lady was with whom so-and-so was caught.
