% Ad Atticum 5.21 (ad-atticum-05-21) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} That you have come safely into Epirus and, as you write, made the crossing to your liking, I rejoice with all my heart; that you are not at Rome at a moment exceedingly necessary to me, I take rather hard. With this one thought, however, I console myself: I hope you are wintering pleasantly where you are and resting yourself with a good will.

\ciceroSection{2} C. Cassius, the brother of your intimate Q. Cassius, had sent more sober letters --- the ones about whose meaning you ask me --- than those he sent afterwards, in which he writes that the Parthian war has been finished off by his own hand. They had indeed withdrawn from Antioch before Bibulus arrived, but with no \textit{[Greek: good day's work]} of ours; today, in fact, they are wintering in Cyrrhestica, and a vast war is hanging over us. For the son of Orodes, king of the Parthians, is in our province, and Deiotarus --- to whose son the daughter of Artavasdes is betrothed, from whom the truth can be known --- does not doubt that he himself will cross the Euphrates with all his forces in early summer. On the very day on which Cassius's victorious dispatches were read out in the Senate, dated the Nones of October, my own were read on the same day, reporting the alarm. Our Axius says that mine were full of authority; the others, he says, were not believed. Bibulus's had not yet arrived; these I know for a fact will be full of fear.

\ciceroSection{3} From all this my fear is that, while Pompey is on no account let go because of the dread of revolution, no honour will be granted to Caesar by the Senate, and until this knot is untied the Senate will think that we ought not to step down before a successor has come, and that in so great an upheaval such great provinces should not be left in the charge of single lieutenants. My horror is that something may be prorogued upon me here, which not even a tribune's veto could ward off; the more so because you are away, who by counsel, influence, and zeal would meet many situations head-on. But you will say I am inventing my own anxieties. I am forced to wish it so; but I fear everything. And though that letter of yours, which you sent off sea-sick from Buthrotum, has the fine \textit{[Greek: closing flourish]} that, as I see and hope, there will be no delay in your stepping down, ``I should rather'' (as I see) was needless; ``I hope'' was enough.

\ciceroSection{4} I had received it fairly promptly at Iconium, by the tax-farmers' couriers, the one sent at the time of Lentulus's triumph. In it you confirm that \textit{[Greek: bittersweet]} thing, that there will be no delay for me; then you add that, if things go otherwise, you will come to me yourself. Your hesitations weigh on me, especially when you see which of your letters I have received. For the ones which you write you sent by the camp-servant of the centurion Hermon I have not received. You had written more than once that you had given letters to Laenius's boys. Those Laenius at last delivered to me at Laodicea, when I had got there, on the third day before the Ides of February --- letters dated the tenth day before the Kalends of October. Your recommendation of Laenius I will at once make good by my words, and by my deeds in the time to come.

\ciceroSection{5} That letter had nothing new in it apart from one thing, about the Cibyran panthers. I love you much for having answered M. Octavius that you did not think so. But from now on, refuse outright as certain anything that is not strictly proper. We, in fact, both of our own accord well steadied, and, by Hercules, fired up by your authority, have outdone everyone --- and this you will find to be so --- in self-restraint as well as in justice, accessibility, and clemency. Don't suppose that anything has ever amazed people more than that, throughout my tenure of the province, not a farthing has been spent --- not on public business, not on any of my staff --- except in the case of L. Tullius the lieutenant. He, otherwise abstinent, does break the Julian law in transit, but only once a day, not, as others were accustomed, in every village (apart from him, only one man took anything, and only once), so that he must be made an exception when I say that not a farthing has been spent. Apart from him no one accepted anything. These soilings I had from our Q. Titinius.

\ciceroSection{6} I, with the summer campaign finished, set my brother Quintus in charge of the winter quarters and of Cilicia. Q. Volusius, the son-in-law of your dear Tiberius, a sound man and astonishingly self-restrained besides, I sent across to Cyprus to be there for just a few days, so that the small number of Roman citizens who do business there could not say that justice had been refused them; for to summon Cyprians out of their

\ciceroSection{7} island is not permitted. I myself set out for Asia from Tarsus on the Nones of January, and, by Hercules, it cannot be told with what admiration of the communities of Cilicia, and above all of the people of Tarsus. After I had crossed the Taurus, there was an extraordinary expectation for me in the dioceses of Asia, which for the six months of my command had had no letter of mine, had never seen me as their guest. Now that season every year before me had been used for this kind of revenue: wealthy communities used to pay large sums of money to keep soldiers from going into winter quarters; the Cyprians paid two hundred Attic talents. From which island (I am not speaking \textit{[Greek: hyperbolically]} but in plainest truth) not a coin will be drawn out while I hold the province. In return for these benefits, at which they stand stupefied, I allow no honours to be decreed me except those of words; I forbid statues, shrines, \textit{[Greek: four-horse chariots]}, and in no other matter am I a burden to the communities --- though perhaps to you, when I keep proclaiming all this about myself. Bear with it, if you love me: you yourself wished me to do these things.

\ciceroSection{8} I made my journey through Asia, then, in such a way that even the famine --- than which nothing is more wretched --- which then weighed on this Asia of mine (for there had been no harvest) became something I had to hope for. Wherever I went, with no force, no judgement, no abuse, by authority and exhortation alone, I brought it about that the Greeks and Roman citizens who had been hoarding grain promised great quantities to the people.

\ciceroSection{9} On the Ides of February, the day on which I sent this letter, I had arranged to hold the assize at Laodicea for the Cibyratic and Apamene dioceses, from the Ides of March in the same place for the Synnadene, the Pamphylian (then I shall test out Phemius's \textit{[Greek: horn]}), the Lycaonian, the Isaurian; from the Ides of May into Cilicia, so that June may be used up there, in peace from the Parthians, I hope. July, if things go as I wish, must be spent on the road, on my way back through the province. For we came into the province at Laodicea in the consulship of Sulpicius and Marcellus on the day before the Kalends of August. From there I must depart on the third day before the Kalends of August. My first effort will be to press my brother Quintus to allow himself to be put in charge, although he himself, like me, will be most unwilling. But honour cannot be served otherwise, since not even now can I retain that excellent man Pomptinus. For Postumius is dragging the man off to Rome --- and perhaps Postumia is too.

\ciceroSection{10} You have my plans; now hear about Brutus. Your friend Brutus has certain intimates who are creditors of the Salaminians of Cyprus, M. Scaptius and P. Matinius; whom he recommended to me in the warmest terms. Matinius I do not know; Scaptius came to me in camp. I promised that, for Brutus's sake, I would see to it that the Salaminians paid him the money. He gave thanks. He asked for a prefecture. I refused, saying I gave one to no man in business (something I had already made plain to you when Cn. Pompeius asked, and which I had proved acceptable to my course of conduct --- what to say of Torquatus on behalf of your M. Laenius, and of many others?); but if he wished to be a prefect for the sake of the bond, I would see to it that he collected. He gave thanks; he withdrew. Our Appius had given this Scaptius some troops of cavalry, by means of which to coerce the Salaminians, and had given him a prefecture too; he was harassing the Salaminians. I ordered the cavalry to withdraw from Cyprus. Scaptius took it hard.

\ciceroSection{11} What more is there to say? So that I might make good my pledge to him, when the Salaminians, Scaptius among them, had come to me at Tarsus, I ordered them to pay the money. Much was said about the bond, much about Scaptius's wrongs. I refused to listen; I urged them, I even begged them, by virtue of my services to their community, to settle the business; I said at last that I would compel them. The men not only did not refuse, but even said this: that they were paying from me. For the sums which they had been accustomed to give to the praetor, since I had not accepted them, they were in some way paying from me, and indeed Scaptius's account was somewhat lower than the praetor's revenue would have been. I commended the men. ``Quite right,'' says Scaptius, ``but let us add up the total.'' Meanwhile, since I had in my standing edict the provision that I would observe twelve per cent simple interest with annual compounding, he was demanding from the bond forty-eight per cent. ``What is this?'' I say. ``Can I go against my own edict?'' But he produces a decree of the Senate, passed in the consulship of Lentulus and Philippus, that whoever held Cilicia should administer justice in accordance with that bond.

\ciceroSection{12} I shuddered at first; for it would have been the ruin of the community. I find two decrees of the Senate by those same consuls about the same bond. The Salaminians, when they had wished to refinance their loan at Rome, could not, because the lex Gabinia (a law on provincial loans) forbade it. Then Brutus's intimates, trusting in Brutus's influence, wished to advance them money at forty-eight per cent, if they could be safeguarded by a decree of the Senate. By Brutus's favour a decree of the Senate is passed that neither the Salaminians nor the men who had given them the money should incur penalty. They paid out the money. But afterwards it came into the moneylenders' minds that that decree did them no good, because the lex Gabinia forbade justice to be administered in accordance with such a bond. Then a decree of the Senate is passed that justice should be administered in accordance with this particular bond --- not so that this bond should be on a different legal footing from any other, but on the same one. When I had set all this out, Scaptius draws me aside; he says he raises no objection to my account, but that they think they owe two hundred talents; that he is willing to take that; and that they owe somewhat less. He asks me to bring them up to two hundred. ``Excellent,'' say I. I summon them to me with Scaptius withdrawn. ``Well, then? How much do you owe?'' I ask. They answer, one hundred and six. I refer the matter back to Scaptius. The man bursts out shouting. ``Well, then,'' say I, ``you must compare your accounts.'' They sit down, they work it out; they agree to the last coin. The Salaminians say they wish to pay; they press him to accept. Scaptius draws me aside again; he asks me to leave the matter where it stands. I yielded to a man asking shamelessly; when the Greeks complained, when they begged to deposit the money in a temple, I refused. All who were present cried out --- some that nothing was more shameless than Scaptius, who was not content with twelve per cent compound; some that nothing was more stupid. To me he seemed more shameless than stupid; for either he was not content with twelve per cent on a good debt, or he was hoping for forty-eight per cent on a bad one. That is my case.

\ciceroSection{13} If Brutus does not approve of it, I see no reason why we should love him. But his uncle will certainly approve --- especially since a decree of the Senate has just been passed (I think, after your departure) in a creditors' case, that twelve per cent should run as the perpetual rate. What this means, if I know your fingers, you have certainly worked out. In this connection --- \textit{[Greek: a thing by the way]} --- L. Lucceius, son of Marcus, complains to me by letter that there is the gravest danger that, through the fault of the Senate in these decrees, the matter may run on to a cancellation of debts; he recalls what mischief C. Iulius once did when he tacked on a little day; never did the commonwealth get worse. But I return to the matter. Rehearse my case against Brutus, if there is a case here against which nothing creditable can be said, especially since I have left the whole matter and the case open.

\ciceroSection{14} The rest is family business. As for the \textit{[Greek: secret-of-the-house matter]}, I am of your mind --- for the son of Postumia, since Pontidia is being trifling. But I wish you were here. Look for nothing from my brother Quintus in these months; for the Taurus, on account of the snows, cannot be crossed before June. Thermus, as you ask, I prop up with the most frequent letters. King Deiotarus says P. Valerius has nothing and that he is being supported by him. When you know whether or not an intercalary month has been added at Rome, I should like you to write me a sure account of the day on which the Mysteries are to be. I look for letters from you a little less than if you were at Rome --- but I look for them still.
