% Ad Familiares 8.11 (ad-familiares-08-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} Your supplications gave us a sharp wringing --- not a long one, but a sharp one. We had run up against an awkward knot. Curio, devoted to you as he is, was being stripped on every count of his comitial days, and would have it that he could on no terms allow supplications to be decreed: he did not want it to look as though, by his own fault, he had thrown away what he had gained from Paulus's fit of temper, and to be reckoned a collusive prosecutor of the public cause. So we came down to a compromise, and the consuls have given their word that this year they will make no use of these supplications. There is plain reason to thank both consuls --- Paulus the more, certainly: for Marcellus's answer to him was that he had no hopes in those supplications anyway, whereas Paulus said outright that he would post no edict at all this year.

\ciceroSection{2} We had had word that Hirrus meant to spin out a long speech. We laid hands on him; he not only stood down, but, when the victims came up for debate and he could have wrecked the business by demanding that they be counted, he held his tongue --- merely siding with Cato, who had spoken honourably of you but voted against decreeing the supplications. Favonius made a third with these. Each man, then, should be thanked according to his nature and his style: these, because they made only a demonstration of goodwill --- their vote went one way, but when they could have obstructed they did not press the fight; Curio, however, because for your sake he deflected the whole course of his proceedings. Furnius and Lentulus, as they ought, made the rounds and put their backs into it as if it were their own affair. I can praise also the diligence and good offices of Cornelius Balbus: he had a fierce word with Curio, told him that if he acted otherwise he would be doing Caesar an injury, and finally cast suspicion on his good faith. The Domitii and Scipios voted for the decree and were unwilling to drop the matter; when these men, by their interrupting, were trying to provoke an interposition, Curio came back at them most charmingly --- that he was all the more glad not to interpose his veto, since he saw that certain of those voting did not wish the thing to go through.

\ciceroSection{3} As to the commonwealth, the whole struggle has come down to a single issue: the provinces. On this front, so far, Pompey and the Senate seem to be bent on having Caesar lay down his command on the Ides of November. Curio is determined to undergo anything sooner than allow that, and has thrown his other proceedings overboard. Our friends, on the other hand --- whom you know well --- did not dare to push the matter to a final showdown. The whole staging is this: Pompey, as though he were not attacking Caesar but only settling what he thinks fair to him, says that Curio is trying to stir up discord. He is, however, strongly opposed to Caesar's becoming consul-designate before he has handed over his army and his province, and frankly afraid of it. Curio is receiving him pretty roughly, and Pompey's whole second consulship is being raked over. Mark this from me: if they press Curio on every point, Caesar will defend his veto-man; if --- as it looks --- they shy off, Caesar will stay on as long as he pleases.

\ciceroSection{4} Who voted which way is in the city journal; from that pick out what is worth your time, and skip much --- in particular the catcalls at the games, the funerals, and the rest of the trivia. It carries plenty of useful matter; on balance I prefer to err on this side --- that you should hear some things you have no need of --- rather than that anything you need should be left out. I am glad you have taken care of the Sittius business; but since you suspect that the men gave you the impression of less than perfect good faith, please act in it as if you were the principal yourself.
