% Ad Atticum 7.6 (ad-atticum-07-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I have absolutely nothing to write to you about; everything is known to you, and I myself have nothing to expect from you. Let us therefore at least keep up that ritual of ours, of letting no one go your way without a letter.

\ciceroSection{2} About the commonwealth I am profoundly afraid, and so far I have scarcely found a man who did not think we should rather concede to Caesar what he demands than fight it out. The demand is indeed an impudent one, stronger than people credit. But why should we now for the first time resist him? \textit{[Greek: For surely this evil that is on us now is no greater]} than when we were prorogating his command for the second five years, or when we were carrying the bill that his candidacy might be entertained in his absence --- unless perhaps we put weapons in his hands then so that we might fight him now when he is well prepared. ``What, then,'' you will say, ``will be your view?'' Not the same as what I shall say. For my view will be that everything must be done to keep it from being decided by arms; what I shall say will be the same as Pompey says, and I shall not say it in a craven spirit. But again, this very thing is a tremendous evil for the commonwealth, and in a certain way it is not right that I, beyond all others, should differ from Pompey in matters of such moment.
