% Ad Atticum 6.3 (ad-atticum-06-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Although I had really nothing new to say of what has happened since I gave a letter for you to your freedman Philogenes, still, as I was sending Philotimus back to Rome, something had to be written to you. And first, what was vexing me most --- not that you could be of any help; how should you be? the matter is in hand, and you are far away in the wide world; \textit{[Greek: and many waves does the south wind roll in the open space of the broad sea]}. The day is creeping up, as you see (for on the third day before the Kalends of Sextilis I must leave the province), and no successor is coming. Whom shall I leave in charge of the province? Reason and the opinion of men alike call for my brother --- first because it seems to be the position of honour, and no one therefore is more eligible; secondly because he is the only man I have of praetorian rank. For Pomptinus, by an agreed bargain (for he came out only on that condition), has now left me; the quaestor no one thinks worthy --- he is a light fellow indeed, loose, and light-fingered.

\ciceroSection{2} As for my brother, the first point is this: I think he cannot be persuaded; for he hates the province, and by Hercules nothing more hateful, nothing more tiresome. Suppose, then, he should be unwilling to refuse me --- what is my duty? Since a great war is reckoned to be in Syria, and that war seems likely to break out upon this province, and here there is nothing in the way of garrison, and an annual stipend has been voted --- does it seem to be the part either of my brotherly affection to leave my brother behind, or of my own diligence to leave any rubbish of a man behind? You see, then, how great is the worry that grips me, how great the lack of counsel. What more would you have? In the whole business we should never have been involved at all. How much better your province is! You will leave it when you wish --- unless indeed you have already left it; whomever you please you will set in charge of Thesprotia and Chaonia. And yet I had not even managed a meeting with Quintus, such that, if that course were chosen, I might know whether it could be got from him; nor, even if it could, did I yet have a clear notion of what I wanted.

\ciceroSection{3} This, then, is the way of it. The rest of the account is still full of praise and good favour, worthy of those books of yours that you so loudly praise: cities preserved, the publicani satisfied to the full, no one offended by an insult, very few by a just and severe sentence, and yet not one of them daring to complain, deeds worthy of a triumph; on which point itself we shall do nothing eagerly, and certainly nothing without your counsel. The closing scene is the difficulty --- the handing over of the province. But some god shall pilot this.

\ciceroSection{4} Of city matters you know more, of course; you hear more often and more reliably; for my part I am sorry I am not kept informed by letters from you. For tiresome news was being brought here of Curio, of Paulus. Not that I see any danger while Pompey stands --- or even sits, provided only he keep his health --- but, by Hercules, I am sorry for the part of Curio and of Paulus, friends of mine. So I should like you to send me, if you are now at Rome or when you are, a sketch of the whole state of the commonwealth, to meet me on the road, that I may shape myself by it and think out beforehand in what spirit to approach the city. It is something, in arriving, not to be a foreigner and a stranger.

\ciceroSection{5} And what I almost passed over: in your Brutus's affair, as I have often written to you, I have done everything. The Cypriots were counting out the money; but Scaptius was not content with interest at twelve percent per annum, renewed yearly. Ariobarzanes is not more forthcoming towards Pompey through his own action than he is towards Brutus through mine. As for Brutus, I could not stand surety for him; for the king was very poor, and I was so far off from him that I could do nothing except by letter; yet by letter I never ceased to fight. The sum of it is this. In proportion to what was owing, Brutus has been more liberally handled than Pompey. To Brutus there has been settled this year about a hundred talents; to Pompey, in six months, two hundred has been promised. As for Appius's affair, how much I have done for Brutus's sake can scarcely be told. What is there, then, for him to be aggrieved at? His friends are pure rubbish, Matinius and Scaptius. The latter, perhaps because he did not get squadrons of cavalry from me with which to harry Cyprus as he had done before my time, may be sulking; or because he was not made a prefect --- an office I have granted to no businessman, not to my own friend C. Vennonius, not to your M. Laenius --- and because I had shown you at Rome that I should keep to this principle, in which I have persevered. But what complaint can a man make who would not take the money when he could? Or as for Scaptius (the one who was in Cappadocia), I think enough has been done. He had accepted a tribunate from me which I had conferred on him at Brutus's request; and afterwards he wrote to me that he did not wish to make use of that tribunate.

\ciceroSection{6} There is a certain Gavius on whom, at Brutus's request, I had bestowed a prefecture: he said and did many things, with a certain insult to me; he is a dog of P. Clodius. He neither escorted me when I was setting out for Apamea, nor, when he had come into camp afterwards and was leaving it again, did he ask whether I wanted anything, and was openly, on some grounds I cannot name, no friend of mine. Had I kept this fellow among my prefects, what sort of man would you take me for? --- I, who, as you know, have never put up with the insolent bearing of the most powerful, was I to put up with this dependent's? And yet this is more than putting up with: it is even to grant him favour and honour. So this Gavius, having lately seen me at Apamea on his way to Rome, addressed me in a way I should hardly have dared to address a Culleolus: ``From where,'' he says, ``do you bid me seek the prefect's rations?'' I answered, more mildly than the bystanders thought I should, that it had not been my practice to give rations to those whose services I had not used. He went off in a rage.

\ciceroSection{7} If Brutus can be moved by the chatter of this scamp, you may love him all by yourself: a rival in me you will not have. But I expect him to be the man he ought to be. Still, I wanted the case to be known to you, and I have written it all out to him most carefully. To say it all (for we are alone): Brutus has never sent me a letter, not even the last one about Appius, in which there has not been something arrogant, something \textit{[Greek: lacking in common feeling]}. You yourself, however, are very fond of having on your tongue: ``but Granius did not think little of himself, and hated proud kings.'' In which, however, the man rather raises a laugh in me than stirs my bile. But he plainly gives too little thought to what he writes, or to whom.

\ciceroSection{8} Quintus Cicero the boy read, as I suppose --- and certainly did read --- a letter addressed to his father. For he is in the habit of opening them, and by my own advice, in case there is anything that needs to be known. In that letter, however, was that same matter about his mother which had come to me. I saw the boy wonderfully upset. In tears he poured out his complaint to me. What more would you have? A wonderful piety, sweetness, and humanity I perceived in him; and so I have all the greater hope that nothing will turn out otherwise than as befits him. This, then, I wanted you to know.

\ciceroSection{9} Nor will I pass over this either. The younger Hortensius was at Laodicea for the gladiatorial games, behaving disgracefully and shamefully. Him, for the father's sake, I invited to dinner the day he came, and for the same father's sake nothing further. He told me he would wait for me at Athens to travel back with me. ``Very well,'' I said; for what was I to say? On the whole I think nothing will come of what he said --- I hope not, indeed, lest I offend the father, whom by Hercules I dearly love. But if he turns out to be in my company, I shall so manage as not to offend the man I least want to offend. So much for that; one thing more. The speech of Q. Celer against M. Servilius please send me. Send a letter as soon as you can; if nothing, then nothing at all by your own courier. Greetings to Pilia and to your daughter. Take care of yourself.
