% Ad Familiares 15.1 (ad-familiares-15-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. TVLLIVS M. F. CICERO PROCOS. S. D. COS. PR., TR. PL., SENATVI.}

\ciceroSection{1} If you are in good health, it is well; I myself am in good health. Though it was being reported to me past doubting that the Parthians had crossed the Euphrates with nearly all their forces, nonetheless, because I judged that fuller information about these matters could be sent to you by M. Bibulus the proconsul, I had determined that there was no necessity for me to write officially about events reported from another man's province. Afterwards, however, when from the surest witnesses --- envoys, messengers, letters --- I had been more fully informed, I thought I ought to write to you what had been brought to me: whether because the matter was so great, or because we had not yet heard that Bibulus had reached Syria, or because the conduct of this war is almost shared between Bibulus and myself.

\ciceroSection{2} The envoys of King Antiochus of Commagene were the first to bring me word that great forces of the Parthians had begun to cross the Euphrates. When this report came in, since there were some who thought that less credit should be given to the king, I determined that I should wait until something more definite was brought. On the 19th of September, when I was leading the army into Cilicia, on the borders of Lycaonia and Cappadocia, a letter was delivered to me from Tarcondimotus, who is reckoned the most loyal of our allies beyond the Taurus and a particular friend of the Roman people, to the effect that Pacorus, son of Orodes the king of the Parthians, with a very great body of Parthian cavalry, and a great force besides drawn from many tribes, had crossed the Euphrates and pitched camp at Tyba, and that a great alarm had been raised in the province of Syria. On the same day a letter was delivered to me on the same matters from Iamblichus, the phylarch of the Arabs, who is held to be well-disposed and a friend to our state.

\ciceroSection{3} When these reports were brought, although I knew that the allies were weakly disposed and held in suspense by the expectation of revolution, I hoped nonetheless that those to whom I had now drawn near, and who had seen plainly our mildness and our integrity, had been made more friendly to the Roman people, and that Cilicia would prove the steadier if it were made a partner in our fairness. Both for that reason, and so that those of the Cilician people who were under arms might be put down, and that the enemy, who was in Syria, might know that the army of the Roman people, far from giving ground at the news, was actually approaching nearer, I gave orders to lead the army to the Taurus.

\ciceroSection{4} But if my authority carries any weight with you --- particularly in matters which you have heard about, but which I almost see in person --- I urge and warn you in the strongest terms that, later than was fitting, but at some point at last, you take thought for these provinces. With what equipment we have been sent, and with what defences fortified, to face the prospect of so great a war, you are not unaware. This task, not from being blinded by stupidity but from being deterred by a sense of shame, I did not refuse; for I have never thought any danger so great that I would rather slip away from it than obey your authority.

\ciceroSection{5} At this point the matter stands as follows: unless you send into these provinces in good time an army as great as you are accustomed to send to the greatest of wars, the gravest danger is that all these provinces, which carry the revenues of the Roman people, must be lost. As for placing any hope in a levy raised here in the province, there is no ground: they are not many, and those there are scatter at the sight of danger; and what kind of soldiers these are has been judged by that most resolute of men, M. Bibulus, in Asia, who, when you had given him the authority, refused to hold a levy. As for the allies' auxiliaries, they are, on account of the harshness and the injustices of our rule, either so feeble that they can be of no great help to us, or so estranged from us that nothing is to be expected of them and nothing is to be entrusted to them.

\ciceroSection{6} The goodwill of King Deiotarus, and his forces, such as they are, I count as ours. Cappadocia is empty; the rest of the kings and princelings are neither in resources sufficiently firm nor in goodwill. In this thinness of soldiery courage, at any rate, will not fail me; counsel, I hope, will not either. What may befall, no one can say. May we be able to look to our preservation! To our standing we shall certainly look.
