% Ad Familiares 13.1 (ad-familiares-13-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. C. MEMMIO}

\ciceroSection{1} Though it had not been quite settled in my own mind whether I should see you at Athens with some distress of feeling or rather with pleasure --- since the wrong done to you would have grieved me, while your wisdom in bearing the wrong would have given me delight --- still I should rather have seen you. For what there is of pain is not really much less for my not seeing you; while what there might have been of pleasure, certainly, had I seen you, would have been more. And so I shall not hesitate to take care that I see you when I can decently do it; in the meantime, what can be transacted by letter --- and, as I think, brought to a finish --- I shall transact.

\ciceroSection{2} Now, in the first place, I shall ask you to do nothing for my sake against your will, but to grant me, only on this condition --- that you see it to matter much to me and not at all to you, on any side --- if you have first persuaded yourself that you would gladly do it. With Patro the Epicurean I have everything in common, except that on points of philosophy I sharply disagree with him. But both at Rome, when he was paying his court to you and to all your circle, he was particularly attentive to me; and lately, when he had obtained what he wanted in the way of his own benefits and rewards, he counted me as nearly the first of his defenders and friends; and indeed by Phaedrus (who, when we were boys, before we had come to know Philo, was greatly approved by us as a philosopher, and was approved later also as a good man, charming and obliging) he was passed on and recommended to me.

\ciceroSection{3} This same Patro, then, having sent letters to me at Rome that I should make his peace with you and ask you to grant him that something-or-other of Epicurus's ruined walls, I wrote nothing to you about it, because I did not want your plan of building to be obstructed by any recommendation of mine; and now, when I came to Athens, the same man having asked that I write to you again to the same effect, he prevailed for this reason: that it was settled among all your friends that you had given up that building project.

\ciceroSection{4} If that is so, and if the matter is now of absolutely no consequence to you, I should like --- in case some little offence has been given by the perverseness of certain men (for I know the tribe) --- that you incline to mildness, whether on account of your own great humanity or even for the sake of my standing. For my part, if you ask what I myself feel, I do not see why he should press so hard, nor why you should resist, except that you can be allowed to give way much less than he can be allowed to make a fuss without cause. Although I am sure that Patro's case and pleading are known to you: he says that he must protect his honour, his duty, the law of testaments, the authority of Epicurus, the entreaty of Phaedrus, the seat, the dwelling, the footprints of the greatest men. We may laugh, if we like, at the whole way of life and the system the man pursues in philosophy, if we want to criticise this insistence of his. But, by Hercules, since we are no great enemies to him and to the others whom such things please, I am not sure whether he ought not to be forgiven if he toils so hard over the matter; for even where he is in the wrong, he is in the wrong by foolishness rather than by depravity.

\ciceroSection{5} But not to say more --- one must come to a stop at some point --- I love Pomponius Atticus as a second brother. Nothing is dearer or more agreeable to me than he is. He (not that he is one of these people --- for he is most polished in every liberal study, though he is very fond of Patro, as he was very fond of Phaedrus), this same Atticus has urged this on me as he has never urged anything --- a man not at all ambitious, not at all importunate in asking --- and he does not doubt that I could obtain it from you by a nod, even if you were planning to build. But now, if he hears that you have abandoned the building and yet that I have not obtained this from you, he will think, not that you are mean-spirited toward me, but that I have been neglectful toward him. I therefore ask you to write to your people that that resolution of the Areopagites --- what they call a memorandum \textit{[Greek: hypomnematismon]} --- may, with your consent, be set aside. But I come back to where I began.

\ciceroSection{6} I would rather you persuaded yourself to do this for my sake gladly than that you simply do it; yet take this for certain --- that if you do what I ask, it will be most welcome to me. Farewell.
