% Ad Atticum 5.14 (ad-atticum-05-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Until I have settled somewhere, you must not expect long letters from me, nor always in my own hand; when there is leisure, I shall manage both. As things stand we were finishing a stage of the road, by a route hot and full of dust. I had posted a letter from Ephesus the day before; this one I post from Tralles. I was reckoning to be in my province on the Kalends of August. From that day, if you love me, set the year-calendar in motion \textit{[Greek: parapegma eniausion]}. Meanwhile, however, the following news, such as I could wish for, was being brought to me: first, that there was peace with Parthia; next, that the contracts of the publicans had been settled; lastly, that a mutiny of the soldiers had been put down by Appius, and their pay made over to them in full down to the Ides of July.

\ciceroSection{2} Asia has received us splendidly. Our arrival has put no one, not even the least, to expense. I hope my whole household is serving my good name. All the same I am in great fear --- but we have good hopes. All my people have now arrived except your Tullius. It was in my mind to set out straight for the army, to give the rest of the summer months to soldiering and the winter months to holding court.

\ciceroSection{3} I should be glad if you, knowing me to be no less curious about public affairs than yourself, would write me everything that is, everything that is to be. There is nothing you can do that would please me more --- unless it be this, that you finish off the business I gave you, and above all that inmost matter \textit{[Greek: endomychon]}, which you know I hold dearer than anything. There you have a letter full of haste and dust; the rest will be more nicely worked.
