% Ad Familiares 15.10 (ad-familiares-15-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. S. D. C. MARCELLO C. R COS.}

\ciceroSection{1} Since it has fallen out as I most wished --- that out of all the Marcelli, and the Marcellini too (for the disposition of your house and name toward me has always been remarkable) --- since, then, it has fallen out so that your consulship of all years can satisfy the zeal of you all, and that into it my campaigns, with their praise and the honour due them, have particularly fallen: I ask of you what is most easily done --- the Senate, as I trust, not refusing --- that you see to it that, when my letter is read out, a decree of the Senate is made in the most honourable terms.

\ciceroSection{2} If I were less close to you than I am to all who belong to you, I would send to you those by whom you understand that I am especially loved. Your father's services to me are very great; for no man can be called a closer friend either to my safety or to my honour. Your brother's regard for me, and the regard he has always shown, there is no one, I think, who does not know. Your whole house, finally, has always attended me with every highest service; nor, in your affection for me, have you yielded to anyone in your family. Wherefore I ask you in the strongest possible terms that you wish me, through your agency, to be as honoured as may be, and that you think my standing, both in the decree of the supplicatio and in other matters, sufficiently commended to your care.
