% Ad Atticum 5.16 (ad-atticum-05-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Although in the very middle of the journey, on the road itself, the publicans' letter-carriers were leaving and we were on the move, I thought I must steal back a little space of time, so as not to seem unmindful of your charge. So I sat down on the very road, while I wrote you out a summary of these matters --- which would require a longer telling.

\ciceroSection{2} I would have you know that, amid the greatest expectation, we have come into a province ruined and quite turned upside down, for good. We arrived on the day before the Kalends of August, stayed three days at Laodicea, three at Apamea, the same at Synnada. We heard nothing else but this: that they cannot pay the poll-taxes \textit{[Greek: epikephalaia]} that have been levied; that the rights of purchase \textit{[Greek: \=onas]} have all been sold off; the groans, the weeping of the cities, a sort of monstrousness not of a man but of some savage beast --- I do not know what. What would you have? They are altogether weary of life.

\ciceroSection{3} The wretched cities are relieved, however, by this --- that no expense is incurred for us, for our legates, for our quaestor, for anyone. I would have you know not only that we do not accept hay or what is wont to be furnished under the Lex Julia, but that we do not even accept firewood; and that, apart from four couches and a roof, no one accepts anything; in many places, not even a roof --- and for the most part we stay in our tents. And so the crowds that flow in --- from the fields, from the villages, from every household --- come in numbers past belief. By Hercules, the people revive at our coming. The justice, the abstinence, the clemency of your Cicero have so risen above the expectations of all.

\ciceroSection{4} Appius, as soon as he heard that we were coming, threw himself off into the furthest corner of the province, all the way to Tarsus. There he is holding court. Of the Parthian there is silence; but still, those who came in were reporting that our cavalry had been cut up by the barbarians. Bibulus was not so much as thinking, even at this point, of coming up into his own province; and they were saying that he was doing this for the reason that he wanted to leave it the later. We were hurrying on to the camp, which was three days' march distant.
