% Ad Familiares 15.4 (ad-familiares-15-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. S. D. M. CATONI}

\ciceroSection{1} Your supreme authority, and the unbroken estimate I have always held of your singular virtue, made me think it of great moment that what I have done should be known to you, and that you should not be in the dark as to the equity and self-restraint with which I have protected our allies and administered the province. With these matters known to you, I judged it would be easier for me to win your approval of what I wish to win it for.

\ciceroSection{2} On entering the province on the last day of July, and seeing that, on account of the season, I must march to the army at once, I spent two days at Laodicea, then four at Apamea, three at Synnada, the same number at Philomelium. In those towns large assizes were held, and I freed many communities from cripplingly bitter taxes, ruinous rates of interest, and debt fraudulently inflated. The army, before my arrival, had been broken up by a kind of mutiny: five cohorts had settled at Philomelium without a legate, without a tribune of the soldiers, indeed without so much as a centurion, while the rest of the army was in Lycaonia. I ordered the legate M. Anneius to lead those five cohorts to the rest of the army, and, having concentrated the whole force in one place, to make camp in Lycaonia near Iconium.

\ciceroSection{3} When that had been carried out by him with care, I myself reached the camp on the 24th of August, having meanwhile, over the previous days, gathered, in accordance with a decree of the Senate, a stout body of recalled veterans, a thoroughly serviceable cavalry, and the volunteer auxiliaries of the free peoples and the allied kings. Then, after the army had been reviewed, on the 30th of August I began my march toward Cilicia. While I was on the road, envoys sent to me from the king of Commagene reported, with great alarm but not without truth, that the Parthians had crossed into Syria.

\ciceroSection{4} On this news I was deeply shaken --- for Syria, for my own province, indeed for the rest of Asia. I therefore judged that I must lead the army through that part of Cappadocia which adjoins Cilicia. For if I had come down into Cilicia proper, Cilicia itself I would easily have held, thanks to the nature of Mount Amanus --- there are two passes from Syria into Cilicia, each of which, on account of its narrowness, can be sealed with small garrisons, and nothing is more defensible against Syria than Cilicia --- but Cappadocia worried me, which lies open from the side of Syria, and has on its borders kings who, though privately friendly to us, do not dare to declare themselves openly hostile to the Parthians. So in the far reach of Cappadocia, not far from the Taurus, I pitched camp by the town of Cybistra, that I might at once protect Cilicia and, by holding Cappadocia, check any new designs of the neighbouring powers.

\ciceroSection{5} Meanwhile, in this great upheaval and the great expectation of a major war, King Deiotarus --- to whom, not without cause, the highest weight has always been allowed, in my own estimate as in yours and the Senate's, a man as exceptional in goodwill and loyalty toward the Roman people as he is preeminent in greatness of spirit and judgement --- sent envoys to me to say that he would come with all his forces into my camp. Moved by his zeal and dutifulness, I sent him thanks in a letter and urged him to hasten.

\ciceroSection{6} Then, having delayed five days at Cybistra on account of the strategy of the campaign, I freed King Ariobarzanes --- whose safety I held under commendation from the Senate at your motion --- from a plot that lay against him, of which he had no inkling; nor was I content with saving his life, but took care that he should reign with authority. Metras, and that Athenaeus whom you had so earnestly recommended to me, who had been punished with exile through the malice of Athenais, I established in the greatest standing and favour with the king; and when a great war was being kindled in Cappadocia, in the event that the priest [the chief priest of Comana] should defend himself by arms --- as he was thought likely to do, being a young man supplied with cavalry, infantry, and money, and supported by all who wanted change --- I so managed it that he withdrew from the kingdom, and the king, without disturbance and without arms, with the whole authority of the court secured, held his throne with dignity.

\ciceroSection{7} Meanwhile I learned, from many letters and messengers, that large forces of Parthians and Arabs had advanced upon the town of Antioch, and that a considerable body of their cavalry, which had crossed into Cilicia, had been cut to pieces by my squadrons of horse and by a praetorian cohort which was on garrison duty at Epiphanea. Therefore, seeing that the Parthian forces, deflected from Cappadocia, were not far from the borders of Cilicia, I marched the army by the longest stages I could to Mount Amanus. On arriving there I found that the enemy had withdrawn from Antioch, and that Bibulus was at Antioch. Deiotarus, who was already on his way to me with a great and reliable force of cavalry and infantry and with all his troops, I informed at once that there seemed no reason for him to be absent from his kingdom, and that I would send him letters and messengers immediately if any new development should arise.

\ciceroSection{8} I had come prepared to relieve either province as the moment might demand; but then, as I had already determined was of great importance to both provinces, I pressed on with the plan of pacifying Amanus and removing from that mountain its perpetual enemy. I made a feint of moving away from the mountain and turning to other parts of Cilicia, and when I had got a day's march from Amanus and had encamped near Epiphanea, on the 12th of October, as evening fell, I marched the army light by night, so that on the 13th of October, at first light, I climbed up onto Amanus. The cohorts and auxiliaries were divided into commands, some under my brother Quintus the legate at my side, others under the legate C. Pomptinus, the rest under the legates M. Anneius and L. Tullius. Most of the enemy we caught unawares; cut off from flight, they were killed or taken.

\ciceroSection{9} Erana, which was not so much a village as a city --- the capital of Amanus --- and likewise Sepyra and Commoris, which Pomptinus, holding that part of Amanus, fought against bitterly and at length, we took, from before dawn until the tenth hour of the day, with a great slaughter of the enemy; and a number of strongholds we took by storm and burned. With this work accomplished we held our camp four days at the foot of Amanus, by the Altars of Alexander, and spent all the remaining time in destroying further parts of Amanus and laying waste the fields --- those portions of the mountain that lie within my province.

\ciceroSection{10} These matters concluded, I marched the army to the town of Pindenissus, of the Free Cilicians. As it stood on a very high and very well-fortified site, and was inhabited by men who had never been subject even to their own kings, who were harbouring runaway slaves and were waiting in eager expectation of the Parthians' coming, I judged it to bear upon the credit of our command that I should crush their boldness, that the spirits of the rest, too, who were estranged from our rule, might more easily be broken. I encircled the place with a rampart and a ditch, hedged it about with six redoubts and a great camp, attacked it with a siege-mound, with mantlets, with towers, and using many siege-engines and many archers, on the fifty-seventh day --- with great toil to myself, and without any trouble or expense to the allies --- I brought the business through: every quarter of the city was thrown down or set on fire, and the inhabitants were driven into my power. Their neighbours, equal in wickedness and audacity, were the Tebarani; from them, after the capture of Pindenissus, I took hostages. I dismissed the army into winter quarters, and put my brother Quintus in charge of the operation, so that the troops should be quartered in the villages, either captured or imperfectly pacified.

\ciceroSection{11} Now I should like you to be persuaded of this: if these matters are referred to the Senate, I shall reckon that the highest possible distinction has been conferred on me if you approve my honour by your vote. Though I know that great men customarily both make and receive requests in such matters, still I think you are rather to be reminded than asked by me. For you are the man who has so often graced me by your speeches in the house --- who by your oratory, your public testimony, your fullest praise, in the Senate, in the public assemblies, has raised me to the skies; and I have always considered the weight of your words so great that, with a single word of yours linked to my praise, I judged I had attained everything. I remember, moreover, when you would not vote a thanksgiving to a certain most distinguished and excellent man, that you said you would vote one if the motion were referred for what he had done as consul within the city: and you yourself voted a thanksgiving to me when I had no armour about me --- not, as to many, for the public business well managed, but, as to none, for the commonwealth saved.

\ciceroSection{12} I pass over the hatred, the dangers, all my storms, which you have not only borne with me, but, had it been in my power, would have been most ready to bear in still ampler measure --- and that you have reckoned my enemy your own enemy, whose ruin, when I could readily see how much you were according to me, you approved in the Senate by defending Milo's cause. From me, on my side, what has gone out is what I would put not in the column of favours, but of true testimony and judgement: not silently to admire your preeminent virtues --- who does not do that? --- but in every speech, in every motion delivered, in every cause pleaded, in all my writings, Greek and Latin, in the whole range of my literary work, to set you ahead not only of those whom we have seen, but of all of whom we have heard.

\ciceroSection{13} You will ask, perhaps, why I make so much of this somewhat-or-other of congratulation and honour from the Senate. I shall now deal with you familiarly, as is fitting between us, with our common studies and our reciprocal services and our high friendship and the bond of family even from our fathers' time. If ever a man has, by his nature, and still more, as I think I am conscious, by reflection and learning, been removed from empty praise and the chatter of the crowd, that man, surely, is I. My consulship is my witness, in which, as in the rest of my life, I confess I have pursued earnestly those things from which true glory could spring --- but glory itself, for its own sake, I have never thought worth the seeking. And so I let pass a rich province and a sure hope of a triumph; and the priesthood, finally, which, as I take it you judge, I could have obtained without great difficulty, I did not seek. The same man, after the injury I sustained --- which you always call a calamity of the commonwealth --- a calamity of my own I do not call it, but rather a glory --- now wanted the verdicts of the Senate and the Roman people upon me to be as splendid as possible. And so I afterwards desired to become augur, which earlier I had let pass; and that honour which the Senate is accustomed to bestow for things done in war, which I once neglected, I now think I should seek.

\ciceroSection{14} I ask you most urgently that you give your support and aid to this wish of mine --- in which there is some force of longing, to heal the wound of that injury --- though a little while ago I had said I would not ask: but ask in such a way that, if this somewhat-or-other I have done shall seem to you not a thin and contemptible business but such and so great that many, on grounds by no means equal, have obtained the highest honours from the Senate, you will give your aid. For my part, I think I have noticed this too --- you know how attentively I am in the habit of listening to you --- that you tend to look not so much at the deeds done as at the character, the principles, and the life of commanders, in granting or withholding such honours. If you will consider these things in my case, you will find that, with a weak army against the threat of a very great war, I have had as my surest defence equity and self-restraint. With these reserves I have achieved what I could not have achieved with any number of legions: that allies most estranged should be turned into most friendly, that allies most faithless should be turned into most steadfast, and that spirits which were trembling in expectation of revolution should be brought back to goodwill toward the old empire.

\ciceroSection{15} But all this about myself is too much, especially to you --- to whom alone, of all men, the complaints of our allies are addressed. You will learn from those who think themselves restored by my institutions; and when nearly all of them, with one voice, declare in your hearing those very things which are most desired by me, then your two greatest clients in particular --- the island of Cyprus and the kingdom of Cappadocia --- will speak to you of me, and I believe King Deiotarus also, who is closer to you than anyone. And if these things are even greater --- and through all ages fewer men have been found who have conquered their own desires than have conquered the enemy's forces --- then surely it is for you, having added to military matters those rarer and more difficult kinds of virtue, to think those same military deeds themselves the more just and the greater.

\ciceroSection{16} One last thing: as though I distrusted my request, I bring philosophy in to plead for me --- than which nothing has ever been dearer to me in life, nor has any greater gift been given by the gods to the race of men. This therefore --- the partnership in our studies and our arts which I share with you, to which from boyhood we have given and bound ourselves, almost alone of men taking that true and ancient philosophy, which to some appears a thing of leisure and idleness, down into the forum and the commonwealth and almost into the line of battle itself --- this partnership pleads with you for my honour, to which I do not think it right that Cato should say no. Wherefore I should like you to be persuaded of this: if the honour shall be assigned me at your motion, on the strength of my letter, I shall reckon that what I most desired has fallen to me both from your authority and from your goodwill toward me.

