% Ad Atticum 5.17 (ad-atticum-05-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I have received from Rome a packet of letters without one from you; in which, if you have only been well and at Rome, I reckon the fault is Philotimus's, not yours. I have dictated this letter sitting in the carriage, on the way to the camp, from which I was two days off. In a few days I should have trustworthy men to whom I could entrust correspondence; and so I have saved myself for that.

\ciceroSection{2} As for us --- though I had rather you heard it from others --- we are conducting ourselves in the province, in the matter of abstinence, in such a way that not a single farthing is spent on anyone. That is being achieved also by the diligence of the legates, the tribunes, the prefects; for all of them are wonderfully sharing my passion for glory \textit{[Greek: symphilodoxousin]}. Our Lepta is a wonder. But for the present I am in a hurry. I shall write you the whole account in a few days.

\ciceroSection{3} Our two Ciceros, the younger Deiotarus has taken with him into his kingdom --- the man who has been styled king by the Senate. While we were in our summer quarters, we judged that for the boys this would be the most charming place possible.

\ciceroSection{4} Sestius wrote to me of the conversation he had had with you about my private and gravest concern, and what was your view of it. Please do, throw yourself into the matter, and write to me what is possible and what you yourself recommend.

\ciceroSection{5} The same Sestius wrote that Hortensius had said something about the prorogation of our province. To me at Cumae he had pledged most punctiliously that he would defend the principle that we should be a one-year man. If you love me at all, fortify this position. It cannot be said how unwillingly I am away from you all; and at the same time I hope that this glory of justice and abstinence will be the more conspicuous if I withdraw soon --- the same thing that fell to Scaevola, who alone presided over Asia for nine months.

\ciceroSection{6} Our Appius, when he saw I was on my way, set off from Laodicea all the way to Tarsus. There he is holding court, while I am in the province. I do not press this injury of his. I have business enough in healing the wounds that have been inflicted on the province; and I am taking pains to do it with as little affront to him as possible. But I should be grateful if you would say this to our Brutus --- that he has not behaved nicely, in that on my arrival he has gone off to the furthest possible distance.
