% Ad Atticum 5.5 (ad-atticum-05-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Plainly there is nothing to write; for I have neither anything to charge you with --- nothing has been left out --- nor anything to tell --- I know nothing new --- nor is this a place for joking, so many cares press on me. Yet know this much: on the Ides of May, as we were setting out from Venusia in the morning, we gave out this letter. On that day, I expect, some business was conducted in the Senate. So let your letters follow after us, so that we may know not only every matter but the rumours too. We shall receive them at Brundisium; for it is the plan to wait there for Pomptinus up to the day you have written.

\ciceroSection{2} The discussions \textit{[Greek: dialogous]} I have at Tarentum with Pompey on the commonwealth I will write out for you in full. Although the very thing I want to know is: up to what point I can rightly write to you --- that is, how long you will be at Rome --- so that I may know either where to give the letters from now on, or not to give them in vain. But before you set out, let that matter of the twenty thousand and the eight hundred thousand sesterces in any case be wound up. I should be glad if you would hold this in the highest and most pressing rank of business: that what, on your urging, I have begun to want, by your help I may attain.
