% Ad Atticum 5.19 (ad-atticum-05-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I had already sealed the letter --- the one, I take it, you have just now read through, written in my own hand, which contains everything --- when suddenly, on the 21st of September, Apella's courier, on the forty-seventh day from Rome (a quick run --- whew, but what a way!) delivered your letter to me. From it I have no doubt that you waited for Pompey until he should come back from Ariminum, and that by now you have set out for Epirus; and I am the more afraid, as you write, that in Epirus you will be no less anxious than we are here. About the Atilian debt I have written to Philotimus not to dun Messalla for it.

\ciceroSection{2} I am glad the news of our march has reached you, and shall be gladder still when you know the rest. I am glad that your little daughter is a delight to you \textit{[at Rome]}, and though I have never seen her, I love her, and I am sure she is lovable. Again and again, farewell.

\ciceroSection{3} About Patro and your fellow-students --- the trouble I took over the ruins at Melita --- I am glad you are pleased. As to your writing that you were happy to see that man defeated at the polls who had contested with the uncle of your sister's son, that is a great sign of love. Indeed, you put me in mind to be glad of it; for it had not entered my head. ``I don't believe you,'' you say. As you please; but I am plainly glad, since \textit{[Greek: there is a difference between rejoicing and envying]}.
