% Ad Familiares 7.2 (ad-familiares-07-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. M. MARIO.}

\ciceroSection{1} Your commission I shall see to diligently. But, sharp man that you are, you have entrusted it to the very person whom it would suit best to have the thing sell for as much as possible. Still, you saw far in setting in advance a ceiling beyond which I am not to buy. If you had left it to my discretion --- such as my love for you is --- I should have settled the matter with the co-heirs; as it is, since I know your price, I shall put up a fictitious bidder rather than let it go for less. But enough of the joke. Your business I shall handle, as I ought, with care.

\ciceroSection{2} That you are glad about Bursa I know for certain; but you congratulate me far too modestly. You suppose, as you write, that on account of the man's squalor I count this no great joy. Believe me, I beg you: I have taken more pleasure in this verdict than I should have in my enemy's death. For in the first place I prefer a judgment to a sword; and next, the glory of a friend to the misfortune of an enemy. Above all, I was delighted that so much zeal for me was shown by good men against the incredible exertions of a most illustrious and most powerful man.

\ciceroSection{3} And finally --- though it may scarcely seem plausible --- I hated this man much worse than Clodius himself. Clodius I had fought; this one I had defended. And Clodius, when the whole commonwealth was about to face its crisis on my head, had some great object in view, and did what he did not of his own initiative but with the help of men who could not stand while I stood; this little ape, for the sake of his own amusement, had picked me out as the man to ride against, and had persuaded certain enviers of mine that he would always be on hand as their thrown weapon to be hurled at me. So I bid you rejoice indeed, and heartily. A great thing has been done. Never were there braver citizens than the men who dared, against the vast resources of one by whom they themselves had been picked as jurors, to condemn him --- which they would never have done, had my pain not been their pain.

\ciceroSection{4} Here we are kept so busy by the crowd and press of trials and by new legislation that we make our vows daily that no intercalary month be inserted, so that we may see you as soon as possible.
