% Ad Atticum 5.10 (ad-atticum-05-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Since I had reached Athens on the sixth day before the Kalends of July, I was now waiting there a fourth day for Pomptinus, and had nothing certain about his coming. I was wholly --- believe me --- with you; and although without those tokens I am with you of myself, all the same, prompted more keenly by your footprints here, I have been thinking about you. What more can I say? By Hercules, no other conversation than about you.

\ciceroSection{2} But perhaps you would rather know something about me myself. Here it is. Up to now no expense, either in my own person or on the public purse or out of pocket, nor on the part of any of my staff. Nothing is being taken under the Julian law, nothing from any host. I have impressed it on all my people that they must serve my reputation. So far it goes well. The thing is being noticed and is talked of widely to the credit of the Greeks. As for what is left, I am putting my back into this, just as I sensed it pleases you. But let us reserve our praises until the speech is concluded.

\ciceroSection{3} The rest is of a kind that I keep blaming my own counsel for not somehow extricating myself from this business. What a thing utterly unsuited to my temper! What truth in that old line, \textit{[Greek: let a man do the work for which he is born]}! You will say --- what so far? You are not yet engaged in the business? I know it well, and I think harder things remain. And yet, even these very things I bear, in face and look, well enough I think; but inwardly I am racked, so many things are said and left unsaid every day, in passion or in arrogance or with foolishness of every variety --- tasteless, insolent --- which I do not write out, not to keep them from you, but because they are \textit{[Greek: hard to unpack]}. And so you will admire my \textit{[Greek: depth of composure]} when we have returned safely; so great is the \textit{[Greek: training]} in this virtue that is being given me.

\ciceroSection{4} So with these things, thus far. And yet I had set myself nothing to write about, since I could not even guess what you were doing or in what part of the world you were. By Hercules, I have never been so long in ignorance of my own affairs, what is being done about the Caesar accounts, what about the Milo accounts. And not just no one from home --- not anyone from Rome --- so that we should know what is going on in the Republic. Therefore if there is anything you know of those matters which you think I should want to know, it will be most welcome to me if you see to its delivery.

\ciceroSection{5} What else is there? Truly nothing except this. Athens has delighted me very much --- in the city itself, that is, and in the city's ornaments, and in the people's affection for you and a certain kindly feeling toward me; \textdagger{}but their philosophy is much up and down\textdagger{} --- if any is in Aristus, with whom I was staying. For your Xeno --- or rather ours --- I had ceded to Quintus; and yet, on account of the closeness, we were together whole days. I should like you to write me your plans as soon as you can, so that I may know what you are doing, where, and at what season --- above all when you will be at Rome.
