% Ad Atticum 5.13 (ad-atticum-05-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} We came to Ephesus on the eleventh day before the Kalends of August, on the five hundred and sixtieth day after the battle of Bovillae. We sailed without fear and without seasickness, but more slowly because of the weakness of the Rhodian undecked vessels. About the gathering of embassies and of private persons, and about the unbelievable throng that was already on hand at Samos but in a wonderful manner at Ephesus --- either I think you have heard of it, or what concern is it of mine? But truly the tithe-farmers behaved as if I had come as their imperator; the Greeks were as forward as if a praetor of Ephesus had come to them. From which I am quite sure you understand that the show-pieces of my many years are now being brought to the test. But, as I hope, we shall use that exercise-ground we learnt from you, and we shall give satisfaction to all --- the more easily because in our own province the bargains have already been struck. But thus far, especially since Cestius has sent me word at dinner that he is setting off in the night.

\ciceroSection{2} Your little errands at Ephesus have been my concern. Thermus, although before my arrival he had promised most generously to all your people, I nonetheless put your Philogenes and Seius in his hands, and I commended to him Xeno of Apollonis. He undertook to do absolutely everything. I also disclosed to Philogenes the account of that exchange of currency I made with you. So this too, thus far.

\ciceroSection{3} I return to city matters. By our fortunes! since you remain at Rome, first prop up and barricade this above all, I beg you --- that ours be a one-year term, and not even an intercalary month added. Next, work off my commissions, and above all whatever can be done about that domestic worry you are not unaware of; then about Caesar --- on whose desire I lay myself down at your prompting, and I do not regret it. And if you understand how much it is mine to know and to take care of what is being done in the Republic --- being done, did I say? on the contrary, what is going to be done --- write me everything, but most carefully, and above all whether the state of the courts, whether of acts done or about to be done, is in any trouble. About the water, if it is on your mind, if Philippus does anything, you will observe it.
