% Ad Familiares 8.5 (ad-familiares-08-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} What anxiety you may be in over the peace of your province and of the neighbouring regions, I do not know; I, for my part, am acutely on edge. For if we could only steer matters in this fashion --- that the scale of the war should rise in step with the strength of your forces, and we should achieve just so much glory and triumph as is necessary, while avoiding that perilous and grave contest --- nothing would be more to be wished for; but as it is, if the Parthian moves at all, I know the struggle will be no light one; and your army can barely cover a single mountain pass. And yet nobody takes this calculation into account: everything is asked of him, as though nothing had been denied him to keep him from being as fully prepared as possible --- this man who has been put in charge of the public business.

\ciceroSection{2} On top of this, I cannot see any prospect of a successor, on account of the dispute over the Gauls. And although on this point I imagine you already have your mind made up on what you would do, still, so that you can make it up more in good time, I thought, foreseeing this outcome, that I should inform you. For you know how these proceedings go, the standard manoeuvres: a resolution will be passed on the Gauls; someone will interpose a veto; then another will rise up who, unless the Senate is given a free hand to decide on all the provinces together, will obstruct the rest. So the game will be played, much and at length, and at such length that the business will be hung up in these snares for more than a full two years.

\ciceroSection{3} If I had anything new in the commonwealth to write to you about, I should be using my usual practice --- to write you out diligently both what had been done and what I hoped would come of it. But everything has now stuck fast, as if in some drainage-ditch. Marcellus is pressing that same point about the provinces, and so far has not managed to muster a full attendance of the Senate. If, with this year passed by, Curio is to be tribune, the same action on the provinces will be set in motion. How easy it is now to obstruct everything --- and how much this \textdagger plays into the hand of Caesar\textdagger, and of those who care nothing for the commonwealth where it stands against their own interest --- does not escape you.
