% Ad Atticum 5.12 (ad-atticum-05-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Sailing is a great business --- and especially in the month of July. We came from Athens to Delos on the sixth day. On the day before the Nones of July from Piraeus to Zoster, with a vexing wind, which held us in the same place on the Nones. Before the eighth day before the Ides, pleasantly to Ceos; thence to Gyaros in a wild but not contrary wind; from there to Syros, then to Delos --- each leg quicker than we should have wished. For you know the Rhodian undecked craft: nothing less able to bear a swell. And so it was in my mind not to hurry at all, and not to stir from Delos unless I had seen all the \textit{[Greek: heights of Gyrae]} clear.

\ciceroSection{2} About Messalla I sent off to you a letter the moment I heard the news at Gyaros, and the same opinion of mine to Hortensius too, for whom I was \textit{[Greek: profoundly anxious]}. But I am waiting for letters from you about the talk surrounding that verdict, and by Hercules about the whole state of the Republic --- letters written, I mean, \textit{[Greek: in a more statesmanlike vein]}, since you are now turning over my books with our Thallumetus --- letters, I say, of the sort by which I may know not what is happening (for that even Helonius, that most weighty person, your client, can supply), but what is about to happen. By the time you read this, we shall have consuls. You will be able to see through everything --- about Caesar, about Pompey, and about the trials themselves.

\ciceroSection{3} Our own affairs, however, since you are lingering at Rome, do please, my dear fellow, unsnarl. The matter that slipped me to write back about, the pile of bricks, I beg you plainly. On the water, if anything can be done, be the man you are always. I prize it very highly, both of my own accord and by your conversations on it. So you will see this through. Besides, if Philippus asks anything --- whatever you would do in your own affair, that I should like you to do. I will write you more when I have set foot somewhere. As I write, I was literally in the middle of the sea.
