% Ad Atticum 5.15 (ad-atticum-05-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I reached Laodicea on the day before the Kalends of August. From this day on you may start clicking off the year. Nothing more longed-for than my arrival, nothing more welcome. But you would not believe how the business sickens me. That well-known galloping pace of mine in mind and effort --- you know it --- has not a field large enough to run in, and so the splendid work goes to waste. Of course --- that I should be giving judgments at Laodicea while A. Plotius gives them at Rome; that, while our friend has so big an army, I should hold the name only of two skeleton legions! And then again --- it is not these things I miss; I miss the daylight, the Forum, the City, my house, you. But I shall bear it as I can, provided only it stay a single year. If it is prorogued, it is finished. Yet it can perfectly well be resisted --- you have only to be at Rome.

\ciceroSection{2} You ask what I am doing here. I'll tell you: I am spending money on a vast scale. I am wonderfully taken with the rule. The abstinence enjoined by your precepts is admirable --- to the point that I fear what I obtained from you on exchange may have to be paid back by a new loan. I do not rub Appius's wounds open, but they are plain to see and cannot be hidden. I was on the march from Laodicea on the third day before the Nones of August, when I was posting this letter, into camp in Lycaonia. From there I was thinking to go to Taurus, so that I might settle, in pitched battle, if I could, with Moeragenes about your slave. \textit{The pack-saddle has been laid on the ox: the load is plainly not mine.} Still, I will bear it, provided that, if you love me, I am a one-year man and you are on the spot when the time comes, to rouse the whole Senate. I am marvellously anxious because for so long now all those affairs of yours have been unknown to me. So, as I have written you before, take care that all the rest, and especially the state of the Republic, is known to me. I would write more, \textit{(of being slowly delivered to you)} but I was posting with a familiar man, of my own household, C. Andronicus of Puteoli. You for your part will often be able to send by the letter-carriers of the publicans, through the masters of pasturage-dues and harbour-dues of our districts.
