% Ad Familiares 8.2 (ad-familiares-08-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} ``Acquitted, I tell you,'' --- and the verdict announced in my presence [corrupt], and indeed by every order, with the votes in each order falling all one way. ``Now laugh,'' you'll say. By Hercules, no. Never has anything happened so wholly against expectation, so plainly intolerable in everyone's eyes. As for me --- though out of friendship I was rooting for him as hard as I could and had already braced myself to grieve --- when the thing was actually done I went numb; I felt I had been struck dumb. What do you suppose the rest did? The jurors, naturally, were set upon with the most enormous shouts and shown beyond doubt that the verdict was not to be borne. So when he comes up under the Lex Licinia, with this still hanging over him, he is reckoned to be in worse danger than before. Add to this that the day after the acquittal Hortensius walked into Curio's theatre --- to make us share his joy, I suppose. Then, my dear fellow: such an uproar, such howling, such thunderclap shouting and rope-end hissing! And it was the more noted because Hortensius had reached old age without once being hissed: but this time so thoroughly that it would have done anyone for a whole lifetime, and he repented of having won.

\ciceroSection{2} On the state of the commonwealth I have nothing to write you. Marcellus's onslaughts have died down --- not from sluggishness, I think, but from policy. About the consular elections, opinion is as unsettled as can be. I have run into a noble competitor, and a noble who is hustling: M. Octavius son of Cnaeus and C. Hirrus are standing against me. I write this for the simple reason that I know you will be hanging on the news of our elections, on account of Hirrus. Anyway, the moment you hear of my being elected designate, do please give some thought to the panthers business. The Sittian bond I commend to you. The first instalment of the city-news bulletin I gave to L. Castrinius Paetus; the second, to the man who is bringing you this letter.
