% Ad Familiares 3.4 (ad-familiares-03-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO S. D. AP. PVLCHRO}

\ciceroSection{1} On the 4th of June, when I was at Brundisium, I received your letter, in which it was written that you had charged L. Clodius with what you wished him to discuss with me. I was indeed expecting him, that I might learn as soon as possible the matters he was bringing from you. My zeal toward you and my dutifulness, though by many things I trust they are by now known to you, I shall make most clearly seen in those by which I shall have most opportunity to indicate that your standing and dignity are most dear to me. Both Q. Fabius Vergilianus and C. Flaccus son of Lucius, and with the greatest particularity M. Octavius son of Cnaeus, have informed me that you make much of me; which I had already judged from many proofs, and most of all from that book on augural law which, written with the warmest affection, you so charmingly sent me. All the highest services of close connection on my side toward you will hold firm.

\ciceroSection{2} For not only have I valued you yourself more highly with every day from the time when you began to be fond of me, but to that have been added the bonds of your relatives --- for of two men I make most account, one of each generation: Cn. Pompeius, your daughter's father-in-law, and M. Brutus, your son-in-law --- and our common membership of the college, the more so as you so honourably gave it your approval, has, in my view at least, brought no slight tie for the binding-together of our affections. But if I meet Clodius, I shall write you more from his account, and shall myself take care that I see you as soon as may be. As for your writing that the reason for your staying was that you might meet me --- that, not to lie, is welcome to me.
