% Ad Atticum 5.18 (ad-atticum-05-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} How I wish you were at Rome, in case you happen not to be! For I had nothing certain except that I had received a letter from you posted on the fourteenth day before the Kalends of August, in which you had written that you would be going to Epirus about the Kalends of August. But whether you are at Rome or in Epirus --- the Parthians have crossed the Euphrates under the lead of Pacorus, son of Orodes, king of the Parthians, with nearly all their forces. Bibulus was not yet heard to be in Syria. Cassius is in the town of Antioch with the whole army; we are in Cappadocia, at the foot of Taurus, with our army, at Cybistra; the enemy is in Cyrrhestice --- which is the part of Syria nearest to my province. I have written to the Senate on these matters; if you are at Rome, you will see this letter and judge whether you think it ought to be delivered. There are many points, indeed all points, of which the main heading \textit{[Greek: kephalaion]} is this: that no further burden of duty or of time be added to me \textit{between the slaying of the victim and the laying-out on the altar} --- as the saying goes. For, given this weakness of an army and want of allies, especially of trustworthy ones, our surest reinforcement is the winter. If that comes on, and they do not cross over before it into my province, the only thing I fear is that the Senate, in its fear for matters in the City, may be unwilling to send Pompey out. If they send some other man by next spring, I am not anxious --- only let no extension of time be granted on us.

\ciceroSection{2} So much, then, in case you are at Rome. But whether you are away, or whether you are present, the affairs here stand as follows. We hold firm in spirit, and, because, as it seems to us, we are taking good counsel, we have hopes also of our right arm. We have settled into a safe position, well supplied with grain, looking across almost to Cilicia, with the ground convenient for moving on, with an army small but, as I hope, of one mind in goodwill towards us. This army we were on the point of doubling by the arrival of Deiotarus with all his own forces. We find our allies far more trustworthy than any man yet has found them; to them our mildness and our abstinence seem incredible. A levy of Roman citizens is being held; grain is being brought in from the fields into safe places. If there is occasion, we shall defend ourselves with our right arm; if not, by our positions.

\ciceroSection{3} So, take heart. For I can see you, and, as if you were present in person, I make out the fellow-feeling \textit{[Greek: sympatheian]} of your love. But I beg you, if it can in any way be managed --- if our case is still standing entire in the Senate up to the Kalends of January --- to be at Rome in the month of January. Surely I shall take no injury, if you are on the spot. We have the consuls as friends, our tribune of the people Furnius. But what is wanted is your constancy, your judgment, your influence. The time is everything. But it is shameful for me to argue at greater length with you.

\ciceroSection{4} Our two Ciceros are with Deiotarus; but, if there is need, they shall be taken away to Rhodes. As for you, if you are at Rome, do as you do --- most punctiliously; if in Epirus, send us, all the same, some letter-carrier of your own, so that both you may know what we are doing, and we may know what you are doing and what you propose to do. As for the business of your Brutus, I am pressing it as he himself would not press his own. But by now I am putting the ward forward on the stand and not defending him; for the proceedings are at once slow and empty. Still, I shall give satisfaction to you, for whom it is harder than for him himself; but I shall give satisfaction, certainly, to both of you.
