% Ad Familiares 2.7 (ad-familiares-02-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. S. D. C. CVRIONI TR. PL.}

\ciceroSection{1} A late congratulation is not usually rebuked, especially when no negligence is to blame for the delay --- for I am far away, and hear things late. But I do congratulate you; and I pray that this tribunate of yours may be a sempiternal praise to you; and I urge you to steer and moderate the whole by your own prudence, lest the counsels of others carry you away. No one can advise you more wisely than you yourself. You will never stumble, if you listen to yourself. I do not write this idly. I see to whom I am writing; I know your spirit, I know your judgement. I do not fear that you will do anything timidly or anything foolishly, if you defend what you yourself shall feel to be right.

\ciceroSection{2} That you have not stumbled into the present hour of the commonwealth, but come into it (for by your own judgement and not by chance you brought your tribunate forward to the very crisis of affairs), you surely see. How great a force in the commonwealth lies in the times, how great a variety in things, how uncertain their issues, how pliable men's wills, what snares, what emptiness in life, I do not doubt you weigh. But, I beg you, take care and think, Curio --- nothing new, but only what I wrote at the start: speak with yourself, take yourself into counsel, listen to yourself, obey yourself. Another who could give you better advice than you yourself can give --- such a one will not easily be found; but better counsel than you yourself certainly no one will give. Immortal gods, why am I not there --- whether as spectator of your praises, or sharer in them, or ally, or minister of your counsels? Though indeed that you least of all lack; still, the magnitude and force of my love would have brought it about that I could help you with counsel.

\ciceroSection{3} I shall write to you more another time; for in a few days I was going to send out my household couriers, so that, since we have managed the affairs of state successfully and according to my own wish, I might write up in a single letter to the Senate all the summer's deeds. As to your priesthood --- how great a care I have taken, and in how difficult an affair and case --- you shall learn from the letter I have given to Thraso, your freedman.

\ciceroSection{4} You, my Curio, by your incredible benevolence toward me, and by my own toward you no less singular, I ask and pray: do not allow any extension of time to be granted me in this provincial vexation. I dealt with you in person, when I did not yet suppose you would be tribune of the plebs that year; and I have made the same request many times by letter, but then as from one who was, so to speak, merely a senator --- though a most noble young man and a most gracious --- and now from a tribune of the plebs, and from Curio the tribune; not that something new be decreed (which is wont to be the harder thing), but that nothing new be decreed, that you defend both the decree of the Senate and the laws, and that the condition under which I set out remain mine. This I urge you again and again, with all the earnestness I have.
