% Ad Familiares 2.10 (ad-familiares-02-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO IMP. S. D. M. CAELIO AEDILI CVRVLI DESIGNATO.}

\ciceroSection{1} See how letters fail to reach me: I cannot be brought to think that you have sent none since you were made aedile, especially as there was such great matter for great congratulation --- about you, because what I had hoped for, about Hillus (for stammerer that I am), because what I had not. And so understand it thus: I have received no letter of yours since those splendid elections, which lifted me out of myself for joy. From which I am afraid the same is befalling my own letters. For my part I have never sent home a single letter without there being a second to you; and nothing is more pleasing or more dear to me than you. But (we are not stammerers) let us return to the matter.

\ciceroSection{2} As you wished, so it has come about. ``You should only wish,'' you said, ``that I have just enough on hand to bring me my little laurel,'' --- you fear the Parthians because you mistrust our forces. So it has fallen out. For when the Parthian war was reported, relying on certain narrow places and the nature of the mountains I led the army to the Amanus, well enough furnished with auxiliaries and with a certain authority among those who did not know me of our name; for it is much in those parts: ``Is this the man who saved the city? whom the Senate?...'' You know the rest. When I came to the Amanus --- which mountain I share with Bibulus, divided by the watershed --- our friend Cassius, much to my pleasure, had thrown the enemy back from Antioch with success, and Bibulus had received his province.

\ciceroSection{3} In the meantime, with all my forces, I harried the Amanienses, our eternal enemies: many were killed, taken captive, the rest scattered; their strongholds, sound as they were, taken and fired by a sudden assault. Thus, after a real victory, I was hailed as imperator at Issus, in that place where, as I have often heard from you, Clitarchus told you Darius was overcome by Alexander, and I led the army away to the most hostile part of Cilicia. There I am now in the twenty-fifth day battering with mounds, mantlets, and towers a most strongly walled town, Pindenissus, with such resources and such effort that nothing is wanting to my crowning glory but the town's name.

\ciceroSection{4} If, as I hope, I take it, then in earnest I shall send the official despatch; this present letter I write you so that you may hope to obtain what you wanted. But to come back to the Parthians: this summer has had a happy enough end; the next is in great fear. For which reason, my Rufus, be watchful: first, that a successor be sent me; if that proves, as you write --- and as I myself judge --- too slow in coming, then for the easier thing, that no prorogation of my time be added on. As for matters of state, from your letters, as I have written you before, I look forward more to what is to come than to what is now present. For which reason I beg you most earnestly to write me out every thing as carefully as may be.
