% Ad Familiares 3.2 (ad-familiares-03-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO PROCOS. S. D. APPIO PVLCHRO IMP.}

\ciceroSection{1} When it had fallen out, against my wishes and beyond my expectation, that I must set out for a province with command, this one consolation kept presenting itself among my many and varied vexations and reflections: that no one could come as your successor more your friend than I am, and that I could receive a province from no one who would prefer to hand it over to me in the most prepared and most fully disengaged condition. And if you too entertain the same hope of my goodwill toward you, that hope assuredly will never fail you. From you, in the name of our great mutual connection and your own singular humanity, I ask, again and again, that by every means in your power (and many will be in your power) you look ahead and consult for my interests.

\ciceroSection{2} You see that, by senatorial decree, the province must be administered. If, so far as you can compass it, you hand it over to me in the freest condition, my passage through my term will be the easier. What in that line you can effect lies in your own judgement; whatever shall occur to you as in my interest, I earnestly ask. I should write to you at greater length, did either your humanity look for a longer speech, or our friendship admit of it, or the matter call for words and not rather speak for itself. Let me put this to you to believe: that if I come to understand my interests have been provided for by you, you will derive from it a great and lasting pleasure. Farewell.
