% Ad Familiares 2.8 (ad-familiares-02-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO PROCOS. S. D. M. CAELIO.}

\ciceroSection{1} What? Do you really suppose I commissioned you for this --- to send me match-ups of gladiators, adjournments of legal sureties, the Chrestus catalogue of pickpocketings, and the kind of things which no one, when I am at Rome, would dare to recount to me? See how much credit I give your judgement (and, by Hercules, not undeservedly: I have known nobody to date more politically perceptive than you, \textit{[Greek: politikoteron]}). I am not asking you to write to me even what is being transacted day by day in the greatest matters of state, unless something is going to bear upon me personally; others will write, many will report, rumour itself will carry much. So from you I look neither for the past nor for the present: I look, as from a man with the longest view of the future, for what is to come, so that, when I have seen the shape of the commonwealth out of your letters, I may know what kind of edifice is going to rise.

\ciceroSection{2} And yet so far I have nothing to charge you with: there has been nothing you could see further on than the rest of us --- myself in particular, since I have been with Pompey for a good many days running, on no other terms of conversation than affairs of state, of which it is not possible to write, and not proper to write. Take only this: Pompey is an outstanding citizen, both in spirit and in judgement prepared for all that must be looked after in the commonwealth. So commit yourself to the man; he will embrace you --- believe me. He now takes the same view, both of good citizens and of bad, that we have always taken.

\ciceroSection{3} I had been at Athens a full ten days, and our friend Caninius Gallus was much with me, when I set out from there on the 6th of July, the day on which I am giving you this letter. I want everything of mine to be very highly commended to you, but nothing more than this: that no extension of time be granted on my province. In that everything for me consists. When, in what way, and through whom this business is to be managed, you will best determine.
