% Ad Familiares 3.6 (ad-familiares-03-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. APPIO PVLCHRO}

\ciceroSection{1} When I compare my own conduct with yours --- though I do not favour myself in the keeping of our friendship more than I favour you --- nonetheless I take much greater satisfaction in what I have done than in what you have done. For at Brundisium I inquired of Phania, whose loyalty toward you I believed I had seen plainly, and whose place with you I knew, into which part of the province he supposed you most wished me to come first in succeeding you. And when he replied that I could do you no greater favour than if I sailed to Side, although that arrival had less of dignity about it and was less convenient for me on several counts, nevertheless I said I would do as he advised.

\ciceroSection{2} When I had likewise met L. Clodius at Corcyra --- a man so closely bound to you that, in talking with him, I seemed to be talking with you --- I told him that I would do this thing, that I would come first to the part Phania had asked for. Then he, after thanking me, urged me strongly to go on directly to Laodicea: that you wished to be in the nearest part of the province, so that you might leave as soon as possible; indeed, that, if I had not been the successor whom you were eager to see, you would already have left before being succeeded. Which agreed well enough with the letters I had received at Rome, from which I had seemed to see plainly how you were hurrying to leave. I answered Clodius that I would do this, and with far greater pleasure than if I had had to do what I had promised Phania. So I both changed my plan and at once sent you a letter written in my own hand; which, as I gathered from yours, was carried to you in time enough.

\ciceroSection{3} With my own conduct I am thoroughly satisfied; nothing could have been done in a friendlier spirit. Now consider in turn your own. Not only were you not in the place where you might have seen me as soon as possible, but you had withdrawn to a point from which I could not even pursue you within the thirty days fixed --- by the Cornelian law, I believe --- as the term for your departure: so that your action, to anyone ignorant of the spirit in which we stand to each other, looks like the action of a stranger --- to put it as gently as may be --- and one avoiding the encounter; mine, like that of the closest and most affectionate of friends.

\ciceroSection{4} And yet, before I came into the province, a letter was delivered to me from you in which, though you showed that you were setting out for Tarsus, you nonetheless held out to me no doubtful hope of a meeting. In the meanwhile, however --- ill-wishers, I do believe, for that vice is widespread and lodged in many men, but who had now hit on a plausible occasion for their talk, ignorant of my own constancy --- were trying to turn my goodwill from you, by saying that you were holding court at Tarsus, settling many things, issuing decrees, sitting in judgement, when you might already suspect that a successor had come, and that not even those who supposed themselves about to be succeeded in a short time were in the habit of doing such things.

\ciceroSection{5} I was not moved by their talk; on the contrary --- believe me, I beg you --- if you were doing anything, I thought myself relieved of trouble, and was glad that my year's province, which had seemed long to me, was now made into a province of nearly eleven months, if a month's labour had been struck off for me in my absence. What does move me, I shall say plainly: that, in so small a force, three cohorts at full strength are absent, and I do not know where they are; and what I bear most painfully is that I do not know where I am to see you. And I have written to you more slowly on that account, because I was expecting you yourself every day, while in the meantime I have received not so much as a letter to tell me what you were doing or where I was to see you. So I have sent to you D. Antonius, a man of courage and one I particularly approve of, prefect of the re-enlisted men, so that, if it seems good to you, you may hand over the cohorts to him, that, while the season of the year is still suitable, I may carry out some piece of business. That I should hope, in this matter, to make use of your counsel, our friendship and your own letters had given me reason to expect, and I do not despair of it even now. But where or when I am to see you I cannot so much as guess, unless you write to me.

\ciceroSection{6} I, for my part, will take care that both fair-minded and unfair-minded men understand that I am most attached to you; about your feeling toward me, you seem to me to have given the unfair-minded some occasion to judge otherwise. If you correct that, I shall be greatly obliged. And, that you may have a way to reckon at what point you can meet me without breach of the Cornelian law: I came into the province on the 31st of July; I am making my journey into Cilicia through Cappadocia; I moved camp from Iconium on the 31st of August. Now you, from the days and the schedule of the route, if you judge that I am worth meeting, will settle yourself at what place this may most conveniently be done, and on what day.
