% Ad Familiares 8.8 (ad-familiares-08-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} I do have things to write you about the commonwealth, but nothing, I think, that will please you more than this: know that Gaius Sempronius Rufus, that honey and darling of yours, has been hit with the calumnia-penalty to a vast round of applause. In what case, you ask? After the Roman Games he charged Marcus Tuccius, his own old prosecutor, under the Lex Plotia on violence; and his calculation was this: that if no extraordinary defendant turned up, he himself would have to stand trial this year. That what was coming for him would otherwise come, he had no doubt; and there was no one he would rather hand the little present to than his prosecutor. So down he went without a co-signer and arraigned Tuccius. The moment I heard, uninvited, I ran to the defence-benches. I rose, and did not say a word about the matter at hand; I worked Sempronius over from end to end, even bringing Vestorius in and telling the famous story --- how he had handed Vestorius over to you as a favour, \textit{[textual crux]} so that what was due for his wrongdoings should be Vestorius's to hold.

\ciceroSection{2} Here is another great contest holding the Forum now. Marcus Servilius, after he had thrown everything into confusion just as he had begun --- left no thing of his unsold for cash --- and had been delivered up into the most savage public hatred, against us: when the praetor Laterensis, on Pausanias's motion, with us as patrons, refused to accept the form ``where this money has come to,'' Q. Pilius, kinsman of our friend Atticus, prosecuted him for extortion. Great fame at once arose, and there began to be hot talk of conviction. On that wind Appius the younger is thrown forward, to argue \textit{[textual crux]} that money from his father's estate had come to Servilius, and to claim 81,000 sesterces had been deposited in collusion. You would marvel at the man's madness --- or rather, you would have, if you had heard his pleading, and the most idiotic confessions about himself, the unspeakable ones about his father.

\ciceroSection{3} Laterensis sends to council the same jurymen who had assessed the damages. When the votes were equal, Laterensis, in ignorance of the laws, announced what each order had judged, and at the end said, as they regularly do, ``I shall not exact.'' After he had withdrawn, and Servilius began to be reckoned as acquitted, and Laterensis read out the hundred-and-first chapter of the law, in which it stood thus: ``Whatever the majority of those jurymen shall have judged --- that shall be the law and stand confirmed,'' he did not enter Servilius onto the tablets as acquitted, but wrote out the verdicts of the orders separately. When Appius again moved, Laterensis came to a settlement with L. Lollius and said that he would refer the matter. So Servilius, now neither acquitted nor condemned, will be handed over to Pilius on the extortion charge, wounded. As for the application to prosecute, Appius, once he had sworn the oath against vexatious accusation, did not dare to contest, and yielded to Pilius; and the man himself has been prosecuted for extortion by the Servilii, and on top of that has been made a defendant for violence by one of his own emissaries, S. Tettius. The pair fit one another, by rights.

\ciceroSection{4} As for matters of state: for many days, in the expectation of the Gauls business, nothing at all was done. At length, however --- the matter having often been deferred, then weightily debated, and Cn. Pompey's mind on it being now plainly seen, namely that it should be his pleasure for Caesar to leave the provinces after the Kalends of March --- the decree of the Senate which I am sending you was passed, and the resolutions reduced to writing.

\ciceroSection{5} \textit{Decree of the Senate; resolutions. On the day before the Kalends of October, in the temple of Apollo. To the drafting were present: L. Domitius Ahenobarbus, son of Cnaeus, of the Fabian tribe; Q. Caecilius Metellus Pius Scipio, son of Quintus, Fabian tribe; L. Villius Annalis, son of Lucius, Pomptine tribe; C. Septimius, son of Titus, Quirine tribe; C. Lucilius Hirrus, son of Caius, Pupinian tribe; C. Scribonius Curio, son of Caius, Poplilian tribe; L. Ateius Capito, son of Lucius, Aniensian tribe; M. Eppius, son of Marcus, Terentine tribe. Whereas M. Marcellus, consul, brought forward the matter concerning the consular provinces, on which thing it was so resolved: that L. Paulus and C. Marcellus, when, as consuls, they shall have entered upon their magistracy, on and after the Kalends of March which shall fall in their magistracy, should refer to the Senate concerning the consular provinces, and should refer no other matter to the Senate from the Kalends of March in priority to this, nor should anything be referred conjointly with this matter \textit{by the advisers}; and that for the sake of this matter they should hold the Senate on assembly-days and should pass decrees of the Senate; and, when the matter should be referred to the Senate, that the advisers --- such of them as should be among the 300 jurymen --- might be brought in \textit{with sureties for appearance}; if anything on this matter required a bill to be put to the people or plebs, that Ser. Sulpicius and M. Marcellus, the consuls, the praetors, tribunes of the plebs, whichever of them should think fit, should put the bill before the people or plebs.}

\ciceroSection{6} \textit{If they did not put it, that whoever should be in office in succession should put it before the people or plebs. On the day before the Kalends of October, in the temple of Apollo. To the drafting were present: L. Domitius Ahenobarbus, son of Cnaeus, Fabian tribe; Q. Caecilius Metellus Pius Scipio, son of Quintus, Fabian tribe; L. Villius Annalis, son of Lucius, Pomptine tribe; C. Septimius, son of Titus, Quirine tribe; C. Lucilius Hirrus, son of Caius, Pupinian tribe; C. Scribonius Curio, son of Caius, Poplilian tribe; L. Ateius Capito, son of Lucius, Aniensian tribe; M. Eppius, son of Marcus, Terentine tribe. Whereas M. Marcellus, consul, brought forward the matter concerning the provinces, on which thing it was so resolved: that the Senate is of the view that none of those who hold the power of intercession or obstruction ought to make delay so that the matter of the commonwealth of the Roman people and the provinces cannot be referred to the Senate and a decree of the Senate be made; and that whoever shall have hindered or prevented it, the Senate judges him to have acted against the commonwealth. If anyone should interpose against this decree of the Senate, it is the pleasure of the Senate that the resolution be reduced to writing, and that on this matter referral be made to the Senate at the earliest opportunity. Against this decree of the Senate there interposed C. Caelius, L. Vinicius, P. Cornelius, C. Vibius Pansa, tribunes of the plebs.}

\ciceroSection{7} \textit{Likewise it is the pleasure of the Senate, concerning the soldiers who are in the army of C. Caesar --- such of them as have their terms of service completed, or have causes for which they ought to be discharged --- that the matter be referred to this body, so that account may be taken of them and their causes inquired into. If anyone should interpose against this decree of the Senate, it is the pleasure of the Senate that the resolution be reduced to writing, and that on this matter referral be made to this body at the earliest opportunity. Against this decree of the Senate there interposed C. Caelius, C. Pansa, tribunes of the plebs.}

\ciceroSection{8} \textit{And likewise it is the pleasure of the Senate, that for the province of Cilicia, and for the eight remaining provinces which men of praetorian rank, as propraetors, hold, those who have been praetors and have not held office with imperium in a province --- such of them as ought, by decree of the Senate, to be sent into the provinces with imperium as propraetors --- be sent into the provinces by lot. If, from that number who ought to go into the provinces by decree of the Senate, there should not be enough to fill the number of those who are to set out into those provinces, then, in the order in which each college of praetors stood, those who had not set out into provinces should accordingly cast lots and proceed; and if those still did not reach the number, then in succession, the praetors of each next-preceding college who had not set out into provinces should be cast into the lot, until that number was made up to which it was required to send out into provinces. If anyone should interpose against this decree of the Senate, the resolution was to be reduced to writing. Against this decree of the Senate there interposed C. Caelius, C. Pansa, tribunes of the plebs.}

\ciceroSection{9} Besides this, the following things from Cn. Pompey were taken notice of, and gave men the greatest reassurance: he said that before the Kalends of March he could not decide about Caesar's provinces without injustice, but after the Kalends of March he would not hesitate. When he was asked what if some should interpose at that point, he said it made no difference whether C. Caesar should refuse to obey the Senate's word, or whether he was now preparing the men who would not allow the Senate to decree. ``What if,'' said another, ``he wants both to be consul and to keep an army?'' To which he answered, ever so mildly: ``What if my son wants to take a stick to me?'' By these words he brought men to think that Pompey has business in hand with Caesar. So now, as I see it, Caesar wishes to come down to one or other of the alternatives: either to stay on and have no account taken of his case this year, or, if he can be elected, to leave.

\ciceroSection{10} Curio is making ready for total war on him. What he can accomplish, I do not know; this much I do see: that a man who has the right view of things, even if he achieves nothing, cannot fall. Curio treats me handsomely, and has by his own gift loaded me with business: for if he had not given me those African beasts which had been brought over for his games, I could have done without. But now, since they have to be put on, I should like you to give the trouble --- which I have always begged of you --- to seeing that we get some beasts from your part of the world. And I commend to you the Sittian bond. I have sent over the freedman Philo, and Diogenes the Greek, with whom I have given a commission and letters for you. I should like you to take both them and the business I have sent them about into your care; for how much it bears upon me, I have written out at length in those letters which they will deliver to you.
