% Ad Familiares 8.10 (ad-familiares-08-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} We have been shaken indeed by the letters of C. Cassius and of Deiotarus: Cassius wrote that the Parthian forces are on this side of the Euphrates; Deiotarus, that they have set out through Commagene into our province. As for me, my chief alarm has been on your account, knowing how poorly furnished you are with an army --- lest this tumult should bring some danger to your standing; for as to your life, had you been better furnished with an army I should have feared more. As it is, the very smallness of your forces gave me reason to expect retreat from you, not a pitched battle. How men will take this, how defensible the necessity will seem --- this I am afraid of even now, and shall not cease to dread until I hear that you have set foot in Italy.

\ciceroSection{2} As for the Parthian crossing, the reports have stirred up varied talk: one man says Pompey should be sent, another that Pompey should not be moved from the city, another Caesar with his own army, another the consuls; no one, however, says any private citizen by decree of the Senate. The consuls, since they fear that the very decree should be made that would have them set out in war-cloak, and that, to their humiliation, the business should be transferred past them to another, are flatly unwilling to have the Senate held at all --- to such a point that they look insufficiently diligent in the commonwealth; but honourably so: whether it is negligence, or inertia, or the very fear I have proposed, lurks beneath this reputation for self-restraint, the wish to have no province. From you no letter has come, and, had Deiotarus's not followed close behind it, Cassius was falling under the suspicion that he had himself plundered what he wished to be seen as ravaged by the enemy, had invented the war and let the Arabs into the province, and then reported to the Senate that these were the Parthians. For which reason I urge you, whatever the state of things there may be, to write me an account that is both careful and cautious, lest you should either be called a sailer-by-favour to anyone, or be charged with having held back something it concerned us to know.

\ciceroSection{3} Now it is the year's end: I am writing this letter on the fourteenth day before the Kalends of December. Plainly I see that nothing can be transacted before the Kalends of January. You know Marcellus, how slow and how little effective; and likewise Servius, what a hesitator. Of what kind do you suppose these men are, or how able to bring to pass what they would not have done, when they do even what they want so coldly that they are reckoned not to want it? With the new magistrates, moreover, if there is a Parthian war, that business will take up the first months. But if either there is, or will be, no war out there, or if it is only such a war that you, or the men succeeding you with small reinforcements, can hold it off --- then I see Curio going to make himself felt in two ways at once: first, to subtract something from Caesar, then, to make a present of something to Pompey --- whatever sort of small bonus, however slight. Paulus too talks about the province in not very humane terms. To this man's grasping our friend Furnius is going to run interference: more than that I cannot guess at.

\ciceroSection{4} This is what I know; the other things that may happen I cannot make out. I know that time brings on much, and changes what has been prepared; but within these limits whatever shall happen will turn. This much I add to the doings of C. Curio: the matter of the Campanian land. About this they say Caesar gives himself no trouble, but Pompey is strongly against it, so that the field may not lie open and empty for Caesar on his coming. As for your departure, this I cannot promise --- that I shall see to it that a successor is sent you; but this much I will undertake: that there shall not be a further prorogation. It is for your own counsel to decide, if the time, if the Senate, presses, if the matter cannot honourably be declined on our side, whether you wish to carry on; mine it is to remember by what adjuration you charged me, on your departure, not to suffer such a thing to happen.
