% Ad Familiares 8.9 (ad-familiares-08-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CAELIVS CICERONI S.}

\ciceroSection{1} ``So that,'' you say, ``is how you handled Hirrus?'' Yes; and if you knew with what ease, and without so much as the smallest contest of any kind, you would be ashamed that he ever dared to come strutting up as if he were your rival. After his defeat, what is more, he is a joke; he plays the good citizen, delivers his opinions against Caesar, snatches at the suspense \textdagger \textit{Curionem prorsus Curionem} \textdagger; chastised not a little even by Curio, he has been thoroughly altered by his repulse. Besides that, he --- who never used to appear in the Forum, who has not been much engaged in the courts --- now pleads liberty-cases, though rarely after the noon hour.

\ciceroSection{2} As for what I had written you about the provinces being brought on on the Ides of August: an interruption from Marcellus, consul-designate, has got in the way. The matter has been put off to the Kalends; they could not even produce a quorum. I am giving you this letter on the 2nd of September, and right up to today nothing has even been advanced. As I see it, the case will be carried over untouched into next year; and, so far as I can divine, you will have to leave behind someone to hold the province; for the succession is not getting through, since the Gauls --- on which there is a veto in standing --- are being brought into the same condition as the other provinces. Of this I have no doubt; which is why I write to you, so that you may prepare yourself for that outcome.

\ciceroSection{3} In almost every letter I have written to you about the panthers. It will be a disgrace to you if Patiscus has sent Curio ten panthers, and you have not sent me many times as many. Curio has made me a present of those very ten, and ten African beasts besides --- lest you suppose he knows only how to make a gift of country estates. If you will simply remember --- and send for those Cibyrans, and likewise dispatch a letter to Pamphylia (for there too, they say, plenty are taken) --- you will succeed in what you intend. I am the more anxious about it now because I imagine I have to make all my preparations on my own account, separately from my colleague. Be a good fellow, give yourself this order. You usually take care of things gladly, just as I, for the most part, take care of nothing. In this business no labour is required of you beyond speaking the word --- which is to say, giving the order and the commission --- for as soon as the beasts have been caught, you have on hand the men I sent on the Sittian bond, who will feed and ship them. I think too, if you give me any encouragement in a letter, I shall send some others over to you.

\ciceroSection{4} M. Feridius, a Roman knight, the son of a friend of mine, an honest and energetic young man who has come out to your part of the world on his own business: I commend him to you, and ask that you count him among your own people. The grounds whose usufruct the cities hold he wants, by your good offices --- which it is both easy and honourable for you to grant --- to be made immune from tax. You will be putting agreeable, honest men under obligation.

\ciceroSection{5} I should not have you suppose that Favonius was passed over by the men at the columns; it was the best people that did not elect him. Your friend Pompey is openly \textdagger \textit{Caesarem et provinciam tenere cum exercitu et consul} \textdagger; his own opinion delivered in the chamber, however, was this --- that at the present time no decree of the Senate should be passed; Scipio's, this: that on the Kalends of March the matter of the Gallic provinces should be put to the house, and nothing else joined with it. This opinion put Balbus Cornelius out of countenance, and I know he registered his complaint with Scipio. Calidius in his own defence was at his most eloquent; in prosecution rather flat.
