% Ad Atticum 5.2 (ad-atticum-05-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the sixth day before the Ides of May, as I was giving out this letter, I was setting out from my Pompeian estate to spend that day at the Trebulan villa with Pontius. From there I was minded, without any delay, to make regular day-stages. While I was at the Cumanum, our Hortensius came to see me, which was a great pleasure --- and when he asked of his own accord for my charges, I committed the rest to him in a general way, but this one specifically: that he should not suffer it, so far as it lay in him, that our provinces be extended. On which point I should be glad if you would brace him up and tell him he has done me a kindness by coming to me and by promising me this and whatever else might be needed. I have braced up for that cause our Furnius as well, whom I saw was going to be tribune of the people for the coming year.

\ciceroSection{2} We had at the Cumanum what was almost a little Rome --- so great was the crowd in those parts. Meanwhile our Rufio, seeing himself watched by Vestorius, struck the man down with a piece of strategy: he simply did not come to me. Really? When Hortensius came to me --- and an unwell Hortensius, and from so far, and Hortensius at that --- with a great crowd besides, he did not come? He did not, I say. ``Then you didn't see the man?'' you will ask. How could I help seeing him, when I was passing through the emporium of Puteoli? I saluted him there, I think, while he was busy at something --- and afterwards even bade him farewell, when he had asked me on coming out from his own villa whether I wanted anything. Could anyone think this man not gracious enough --- or not, on that very ground, deserving of praise, that he gave no trouble to get praised?

\ciceroSection{3} But I return to that other business. Do not suppose that there is any consolation for me in this enormous burden except that I hope it will be no longer than a single year. Many do not believe I really wish it so, judging by the habits of others; you, who do know, will bring all your diligence to bear --- when, that is, the matter must be pressed, after you have come back from Epirus. About the commonwealth I should be glad if you would write to me, if there is anything that \textit{$\dagger$operare$\dagger$}. For there had not yet been brought through to here a sufficient account of how Caesar was taking the decree once written up, and there was a rumour about the Transpadanes that they had been instructed to elect four-men commissioners. If that is so, I fear great upheavals. But I shall learn something from Pompey.
