% Ad Atticum 5.3 (ad-atticum-05-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the sixth day before the Ides of May I came to the Trebulan villa to Pontius's place. There a double letter from you was delivered to me, on the third day from your side. The same day, leaving the Pompeian estate, I had given Philotimus a letter for you; nor in fact is there now anything special to write about. The rumours, please, about the commonwealth --- write to me what they are. In the country towns I see the highest fear; but much of it is empty. What you yourself think about these things, and when, I should like to know.

\ciceroSection{2} To which letter you want me to reply, I do not know. For I had received none up to this point except the double one that was delivered to me at Trebula at one stroke; of these, one contained an edict of P. Licinius (it had been sent on the Nones of May), the other was the reply to my Minturnae letter. I am afraid there may have been something \textit{[Greek: spoudaioteron]} --- more urgent --- in those I did not receive, the ones you want answered. I shall set you in good standing with Lentulus.

\ciceroSection{3} Dionysius is dear to us. Your Nicanor is giving me outstanding service. There is nothing left now to write, and it is getting light. Beneventum is what I have in mind for today. By our own restraint and diligence \textit{$\dagger$esse satis faciemus satis}. From Pontius's at Trebula, on the fifth day before the Ides of May.
