% Ad Atticum 5.8 (ad-atticum-05-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Both an ill turn of health, from which I had now emerged --- since I had laboured under it without fever --- and waiting for Pomptinus, of whom up to this point not even a rumour had come, have held me here at Brundisium for twelve days now; but we were watching for a fair run.

\ciceroSection{2} If you are at Rome --- though I scarcely think so, but if you are --- I should like you to attend to this very keenly. From Rome I had received a letter saying that my Milo was complaining by letter of an injury done him by me, in that Philotimus was a partner in his property. I willed this to be done on C. Duronius's advice, a man I had clearly seen to be most friendly to Milo and had come to know to be of the sort you reckon him. His thinking and mine had been this: first, that the matter should be in our power, so that no malign outsider as buyer might strip him of the very many slaves he has with him; next, that what he had wished secured for Fausta might be made firm. There was this too --- that we ourselves might most easily preserve whatever could be preserved.

\ciceroSection{3} Now I should like you to look the whole matter through; for things are often written to me as larger than they are. If he is complaining, if he is writing to friends, if Fausta wants the same thing, then let Philotimus --- just as I had said face to face to him and he had pledged to me --- not be in the property against Milo's will. Nothing was so much to us. But if these are slighter matters, you shall judge. Speak with Duronius. I have written too to Camillus, to Caelius, to Lamia, and the more so because I had no confidence that you were at Rome. The sum will be this: you shall settle it as it seems to fit with my honour, my reputation, and my interest.
