% Ad Familiares 3.8 (ad-familiares-03-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO S. D. APPIO PVLCHRO}

\ciceroSection{1} Although, so far as I could gather from your letter, I could see that you would be reading this when you were near the city --- by which time the very trivial talk of the provincials would already have cooled --- still, since you had written to me at such length about the talk of ill-disposed men, I thought I ought to send a short answer to your letter.

\ciceroSection{2} But the first two heads of your letter must in a way be passed over by me in silence; for they contain nothing defined or definite, except that, by my expression and my silence, I had given you to understand that I was not your friend, and that this could be seen when something was being transacted on the tribunal, and at certain dinners. This whole thing I can see to be nothing; but, since it is nothing, I cannot even see what is being said. This much I do know: that many talks of mine, illustrious ones, held from a higher place and from level ground alike, with the highest praise of you and with the warmest demonstration of our intimacy, could truly have been reported to you. As for the matter of the embassies, what could have been done by me more decorously or more justly than to reduce the expenditure of the most destitute communities without any diminution of your dignity --- especially since the communities themselves were asking for it? For the whole class of embassies setting out in your name was not known to me. When I was at Apamea, the leading men of many communities reported to me that excessive expenses were being voted to ambassadors at a moment when the communities were not solvent. Here at once I weighed many things together.

\ciceroSection{3} First, I judged that you, a man not only of wisdom but, as we now say, of urbanity, were not pleased with embassies of that sort, and I think I argued at length on the matter from the tribunal at Synnada: first, that Appius Claudius had been commended to the Senate and People of Rome not by the testimony of the men of Midaeum (for the topic had come up in that town), but by his own merit; next, that I had seen this happen in many cases, that embassies came to Rome on people's behalf, but I did not recall any time or place being allowed to those embassies in which they might deliver their praise; that I was pleased with the people's eagerness, in that they were grateful to you for the services you had done; that the whole plan seemed to me by no means necessary; but that if they wished to demonstrate their dutifulness, I would praise the man who performed his duty at his own expense, would permit the one who did it at the legal amount, and would not permit the one who did it at an unlimited rate. What can be censured in this? Unless it is what you add, that to certain people my edict has seemed framed as it were on purpose to obstruct these embassies. By now those who advance these notions do not seem to me to do an injustice so much as those whose ears lie open to such notions.

\ciceroSection{4} I composed the edict at Rome; I added nothing, except that, when the tax-farmers came to me at Samos, they asked me to copy across one head of your edict word for word into my own. The head bearing on the reduction of the expenses of the communities was most carefully written; and in that head are certain new measures, salutary for the communities, with which I am much pleased. But the head out of which the suspicion has arisen that I sought out something in which I might offend you is one of the standard provisions. For I was not so witless as to think that men were going on an embassy for a private matter, when they were being sent --- not to you in private station, and not on a matter that was theirs privately but publicly --- not in a private council, but in the public council of the world, that is, in the Senate, to give thanks; nor, when I issued an edict that no one should set out without my orders, did I shut out those who were unable to follow me into camp and across the Taurus. For that is exactly the laughable thing in your letter. What reason had they to follow me into camp or to cross the Taurus, when from Laodicea as far as Iconium I so made my way that the magistrates and embassies of all those dioceses on this side of the Taurus, and of all their communities, met me?

\ciceroSection{5} Unless perhaps they began to send embassies after I had crossed the Taurus; which is certainly not so. For when I was at Laodicea, at Apamea, at Synnada, at Philomelium, at Iconium --- in all of which towns I stayed --- all embassies of that kind had already been formed. This too I want you to know: that on the question of reducing or remitting the expense of these embassies I decreed nothing except what the leading men of the communities had asked of me, that there should not be added, to the auctioning-off of the tributes and that most bitter exaction of the poll-tax and the door-tax, which is not unknown to you, expenses by no means necessary. For my part, since I had taken up this task moved not only by justice but by pity, that I might lighten the miseries of communities driven to ruin, and driven to ruin chiefly by their own magistrates, I could not afford to be careless about expenditure of that unnecessary kind. You, if reports of this character were brought to you about me, ought not to have believed them; but if you take pleasure in this trick --- of attributing to others what comes into your own head --- you are introducing into friendship a manner of talk by no means liberal. As for me, if I had ever in the province thought of detracting from your reputation, I would not have referred matters to your freedman at Brundisium nor to your chief of engineers at Corcyra, asking to what place you wished me to come. Wherefore you can, on the authority of the most learned writers, whose books on the conduct of friendship are most admirable, do away with this whole manner of speech: ``they were arguing --- I, on the contrary, was contending; they were saying --- I was denying.''

\ciceroSection{6} Or do you think nothing has ever been said to me about you? Not even this --- that, when you had wished me to come to Laodicea, you yourself crossed the Taurus? That on the same days my assize was being held at Apamea, at Synnada, at Philomelium, and yours at Tarsus? I will say no more, lest I should seem to imitate the very thing I am reproaching you with. This I will say, as I feel: if those things which you say others are saying, you yourself feel, the fault is wholly yours; but if others say these things in your presence, even so the fault is not nothing in you, because you hear them. My own course, in the whole of our friendship, will be found steady and serious. As for anyone who pictures me as more cunning --- what could be more cunning than this: that when I was always defending you in your absence, particularly because I did not think it would come to pass that I, too, in my absence should need to be defended by you, I should now commit myself in such a way that you, with the best right, could desert me in my absence?

\ciceroSection{7} There is one class of talk I except, in which something is very often said of a sort that I should suppose you would not wish said --- if anyone of your legates, or your prefects, or your military tribunes is being spoken ill of; but even this, by Hercules, has never yet happened in my hearing, that anyone has spoken more harshly, or against more men, than Clodius spoke to me at Corcyra, in the particular charge that you had been less fortunate, through the dishonesty of others. These talks, since they are many and, as I judge them, do not damage your reputation, I have never set going --- but I have not strenuously repressed them either. If there is anyone who thinks no one can in good faith return to favour, he is not convicting our perfidy, he is betraying his own; and at the same time he is thinking no worse of me than of you. But if there is anyone whom my practices in the province do not please, and who thinks himself injured by some unlikeness of my practices to yours, when each of us has done rightly, but neither has followed the same line: that man I do not care to have as a friend.

\ciceroSection{8} Your liberality, as that of a man of the highest nobility, has shown itself the more amply in the province; if ours is the more narrow (though something has been pared even from your generous and beneficent nature in the later year, on account of a certain harshness of the times), people ought not to be surprised, since by nature I was always more restrained in lavishing what was another's, and since I am moved by those very same times by which others are moved --- ``that I should be bitter to them, that I might be sweet to myself.''

\ciceroSection{9} As for the news of the city, that you have informed me of it --- this in itself was pleasing to me, and the more so because you signified that all my commissions would be a concern to you. Of these I ask especially this one thing: that you take care nothing more of business or burden or time is added to me; and that you ask Hortensius, our colleague and friend, that, if ever he has felt or done anything for my sake, he should now withdraw from that biennial proposal --- than which nothing can be more hostile to me.

\ciceroSection{10} As for our own affairs, since you wish to know: we set out from Tarsus toward Amanus on the Nones of October; I write this on the day after, when I am keeping camp in the territory of Mopsuhestia. If I do anything, I shall write to you; and I shall never send a letter home to my own people without adding one which I want delivered to you. As for what you ask about the Parthians, I take it there have been none; the Arabs who were there, in Parthian guise mingled in, are said to have all turned back; they say there is no enemy in Syria. I should like you to write to me as often as possible, about your own affairs, about mine, and about the whole state of the commonwealth; about which I am the more concerned because I have learned from your letter that our Pompey is going to Spain.
