% Ad Familiares 5.16 (ad-familiares-05-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. TITIO.}

\ciceroSection{1} Of all men I am the least fitted to console you, since I have drawn such grief from your troubles that I have need of consolation myself; and yet, since my own pain stands further from the bitterness of supreme mourning than yours does, I have decided that it belongs to our long-standing bond and to my goodwill toward you not to stay silent so long while you are in so great a sorrow, but to offer some moderate consolation, such as might lighten your pain --- if it could not heal it.

\ciceroSection{2} There is, in the first place, the most commonplace consolation of all --- the one we ought always to have on our lips and in our minds: to remember that we are men, born under this law, that our life is set up as a target for all the weapons of fortune; that we must not refuse to live on the terms on which we were born; and that we should not bear so heavily the blows we cannot escape by any forethought, but, by going over in memory the experiences of others, should consider that nothing new has befallen us.

\ciceroSection{3} Neither these nor the other consolations practised by the wisest men and handed down in the records of literature seem capable of doing as much good as the very state of our commonwealth itself, and this present upheaval of times gone to ruin: in which they are happiest who have not raised children, and those who have lost children in these times are less wretched than they would be if they had lost the same children with the commonwealth still sound, or with any commonwealth left at all.

\ciceroSection{4} If it is your own loss that moves you, or if you grieve from thinking of your own affairs, I do not think this whole pain of yours can easily be drawn off. But if what tortures you is rather the thing that belongs to love --- that you mourn the misery of those who have fallen --- then, not to repeat what I have most often read and heard, that there is no evil in death (from which, if perception remains, what should rather be reckoned is immortality and not death; and if perception has been lost, no wretchedness can seem to exist which is not felt) --- still I can affirm without hesitation that things are being mixed, prepared, hanging over the commonwealth such that whoever has left them behind seems to me in no way to have been cheated. For what room is there now, I do not say for modesty, uprightness, virtue, the right pursuits, the good arts, but for liberty and bare safety at all? By Hercules, I have not heard, in this most calamitous and pestilent year, of any youth or boy dead who did not seem to me to have been snatched away by the immortal gods from these miseries and from this most unjust condition of life.

\ciceroSection{5} Therefore, if this one thing can be subtracted from your view --- that any evil has befallen those whom you have loved --- a great deal will be taken from your sorrow. There will remain only that single concern for your own pain, which will not be shared with them but referred properly to yourself; in which it is not consistent with your gravity and wisdom --- qualities you have shown from boyhood --- to bear too intemperately the blow of your own misfortunes, when that blow is set apart from the wretchedness and harm of those you have loved. You have always conducted yourself in private and in public so that gravity must be guarded, and constancy served. For what mere length of time will bring of itself --- which by its sheer duration removes the greatest griefs --- this we ought to anticipate by deliberation and prudence.

\ciceroSection{6} For if no woman has ever been of so weak a spirit, on the loss of children, that she did not at last set some limit to her mourning, then surely we ought, by counsel, to anticipate what the day will bring, and not wait for the medicine of time --- which we can supply at once, by reason. If by this letter I shall have accomplished anything, I shall consider that I have attained something worth wishing for; if it has not perhaps had the strength to do that, I shall at least have discharged the office of one most goodwilled toward you, and most affectionate as a friend. Such a friend I would have you believe I have always been to you, and feel confident that I shall remain.
