% Ad Familiares 2.1 (ad-familiares-02-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. CVRIONI.}

\ciceroSection{1} Though it pains me to be under suspicion with you on the score of neglect, still I have found it less irksome to be charged by you with shirking my duty than gratifying to be missed --- particularly since on the count of which you accused me I am innocent, while in the matter where you let it be seen that you wanted letters from me, you were displaying to me an affection long well known to me, but for all that sweet and longed-for. For my part, I have not let slip anyone whom I supposed would in fact reach you to whom I did not give a letter; and indeed, who is so untiring in writing as I am? From you, however, I have received at most two or three, and those very short. Therefore, if you are a stern judge against me, I shall condemn you on the same charge; if on the other hand you would rather not have me do so, you will owe it to me to show yourself fair in turn. But enough on letters; for I am not afraid I shall fail to satisfy you in the writing, especially if my zeal in that kind is not despised by you.

\ciceroSection{2} For myself, that you have been so long away from us I have both grieved over --- since I lost the enjoyment of the most delightful companionship --- and rejoiced at, since in your absence you have attained everything with the highest distinction, and since in all your affairs fortune has answered my wishes. There is a brief charge which my incredible love for you compels me to lay upon you. So great is the expectation of your spirit and your talent that I have no hesitation in beseeching and adjuring you: return to us so shaped that you can sustain and uphold the expectation you have stirred up of yourself. And, since no forgetfulness will ever erase from me the memory of your kindnesses to me, I ask you to remember this: whatever accessions of fortune and standing fall to you, you could not have attained them, had you not once as a boy obeyed my counsels --- counsels of the most loyal and most loving kind. For this reason you ought to be disposed toward me thus: that my own years, now growing heavy, may find rest in your love and in your youth.
