% Ad Familiares 2.6 (ad-familiares-02-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. C. CVRIONI.}

\ciceroSection{1} It had not yet been heard that you were on your way to Italy when I sent Sex. Villius, a close friend of my Milo, to you with this letter. But since your arrival was thought to be approaching, and it was now established that you had set out from Asia on the road to Rome, the magnitude of the matter made me unafraid of sending it too soon, when I greatly wished it carried to you as quickly as possible. For my part, Curio, if my services to you were only as great as you yourself are accustomed to proclaim them --- greater than I weigh them at myself --- I should press my case with more reserve, were I asking some large thing of you; for it is a hard thing for a man of any shame to ask anything large from one whom he reckons himself to have served well, lest what he asks should look more like exacting a debt than making a request, and like reckoning the return as wages owed rather than as a kindness given.

\ciceroSection{2} But since your kindnesses to me --- known to everyone, and rendered most signal and most great by the very novelty of my circumstances --- have so stood out, and since it belongs to a generous spirit to wish to owe most to the one to whom one already owes much, I have not hesitated to ask of you by letter the thing that was of all the greatest and most necessary to me. For I have not been afraid that I could not bear up under even countless services from you, especially when I felt sure that there is no favour of yours which my heart could not either contain in the receiving, or in the requital amplify and bring to light.

\ciceroSection{3} I have fixed and lodged every effort of mine, every exertion, care, industry, thought, every faculty of mind in Milo's consular campaign, and I have resolved that in this matter I should seek not the reward of duty alone but the praise of devotion. Nor indeed do I think any man ever cared so much for his own well-being and fortunes as I care for his honour, in which I have decided that everything of mine is placed. To him you alone, if you will, I see can be of such help that nothing further need be sought by us. We have all the rest in hand: the goodwill of the good men, won over from his tribunate on my account, as I hope you understand; the favour of the common run of people, on account of the magnificence of his games and the openhandedness of his nature; the support of the young and of those with weight in the voting, on account of his own outstanding popularity or thoroughness in that domain; my own backing for him --- if not powerful, then at any rate well-regarded, and just, and owed, and perhaps for that very reason carrying some weight.

\ciceroSection{4} What we need is a leader, a sponsor, and what I might call the steersman or pilot of the winds I have set out. If from all the world a single such man were to be chosen, we should have no one to set beside you. Therefore, if you can judge me from this very fact --- that I am toiling so vehemently for Milo --- to be a man who remembers, who is grateful, who is honest, and if you judge me, finally, worthy of your kindnesses, this is what I ask of you: come to the aid of this anxiety of mine, and lend your zeal to this object of my pride or --- to put it more truly --- almost to my safety. About T. Annius himself I promise you only this: that if you choose to embrace the man, you will have no one of greater spirit, gravity, constancy, and goodwill toward you; to me, in turn, you will have added so much grace and so much standing that I shall easily recognise in you the same friend in my fame whom I had in my preservation. Did I not know that you can see in what frame of mind I write this, how heavy the obligation I am bearing, with what effort I must labour in this candidature of Milo's --- not merely contending, but fighting on every front --- I should write more; as it is, I commit and hand over to you the whole affair, the whole cause, and my whole self. One thing alone hold to be so: if I obtain this from you, I shall owe almost more to you than to Milo himself; for my safety, in which I was helped above all by him, was not so dear to me as the devotion of returning thanks for it will be sweet to me --- and that, I am confident, I can attain by your zeal alone.
