% Ad Familiares 2.4 (ad-familiares-02-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO S. D. C. CVRIONI.}

\ciceroSection{1} You know well enough that there are many kinds of letters, but one beyond all doubt, for the sake of which the thing itself was invented: to give information at a distance, if there were something it concerned either us or them to have the other party know. From me you assuredly do not look for a letter of that kind: for your domestic affairs you have your own writers and messengers, and in mine there is really nothing new. There remain two kinds of letter that delight me greatly, the one familiar and playful, the other serious and weighty. Which of the two would less suit me to use, I do not see. Am I to joke with you by letter? By Hercules, I do not think any citizen can laugh in these times. Or am I to write something weightier? What is there that Cicero could write weightily to Curio, except the commonwealth? And yet on that head my own case is this: that I neither dare write what I think, nor wish to write what I do not think.

\ciceroSection{2} Therefore, since I am left no subject to write on, I shall use the closing flourish I am accustomed to, and shall urge you toward the pursuit of the highest praise. For a heavy adversary has been set up against you, and made ready --- a certain extraordinary expectation; which by one thing you will most easily overcome, if you have settled this: that in those arts by which those distinctions are won --- the glory of which you have come to love --- you must put in the labour. On this theme I should be writing more, were I not confident that you are sufficiently spurred on of your own accord. What little I have touched upon, I have done not to inflame you, but to bear witness to my love.
