% Ad Familiares 3.1 (ad-familiares-03-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO APPIO IMP. S. D.}

\ciceroSection{1} If the commonwealth herself could relate to you how she stands, you could not learn it from her more easily than from your freedman Phania; so much is the man not only sensible, but also --- which is welcome --- inquisitive. He, therefore, will set forth everything to you; for this is better suited both to my brevity and to attention to my other affairs. As for my goodwill toward you, although you can learn that too from the same Phania, still some part in the matter seems to belong to me as well. Persuade yourself, then, of this: that you are most dear to me, both on account of the many charms of your character, your good offices, and your humanity, and because I understand from your letters and from the talk of many that everything which has gone out from me toward you has been received by you most gratefully. Since that is so, I shall surely contrive that the long lapse of time during which we have done without our accustomed intercourse I shall make up by the favour, the frequency, and the magnitude of my services; and this, since you so wish, I think I shall do not unwillingly --- nor unwilling, I trust, will Minerva be. Whom indeed, if I happen to take her from your own household, I shall name not only \textit{[Greek: Polias]} but \textit{[Greek: Appias]}.

\ciceroSection{2} Cilix, your freedman, was hitherto less known to me; but when he handed me a letter from you full both of affection and of attentive service, he himself by his own conversation followed up wonderfully the humanity of your letters. His talk was pleasing to me, when he related to me your feeling, and the conversations you have about me every day. What more? In two days he was made familiar to me --- yet so that I shall greatly miss Phania. When you send him back to Rome, which, as we thought, you were going to do shortly, I beg you to give him instructions on every matter concerning which you wish me to act or to take care.

\ciceroSection{3} L. Valerius the jurisconsult I commend warmly to you --- but in such a way that, even if he is not a jurisconsult, I want him better protected than he himself usually protects others. I am very fond of the man; he is one of my own household and intimate friends. He does indeed give you thanks, but he also writes that my letter will carry the greatest weight with you. That this expectation may not deceive him, I ask again and again. Farewell.
