% Ad Familiares 7.11 (ad-familiares-07-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO TREBATIO.}

\ciceroSection{1} If you had not already set out from Rome, you would certainly be leaving it now: who needs a jurist amid so many interregna? My own advice to all defendants would be this --- demand two adjournments from each interrex. Do I seem to have learned my civil law sufficiently from you?

\ciceroSection{2} But come now! How are you doing? Is anything happening? For I see you are already cracking jokes by letter --- these signs are better than the statues in my Tusculan villa. But I want to know what your situation is. You write that Caesar is consulting you; I should have preferred that he were looking after you. If that is happening, or if you think it will, stick out your campaign and stay where you are: I will console myself for the loss of you with hope of your advantage. But if all that is rather hollow, take yourself back to us. For something here will turn up at some point; or, failing that, by Hercules a single conversation between us will be worth more than all the Samarobrivas. In short, if you bring yourself back quickly, there will be no talk; but if you stay away too long, and to no purpose, I am in fear not only of Laberius but of our comrade Valerius too --- for an astonishing character could be put on stage, the British jurisconsult.

\ciceroSection{3} I am not laughing at this, however much you may laugh; I am joking with you, as I always do, about a thoroughly serious matter. Joking aside, I give you this counsel, with the most friendly intention: if out there, on my recommendation, you can hold on to your standing, then bear with our missing you, and increase your honour and your means; but if all that grows cold, take yourself back to us. Still, everything you wish for you will surely attain, both by your own merit and through my own utmost zeal on your behalf.
