% Ad Familiares 7.12 (ad-familiares-07-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO TREBATIO.}

\ciceroSection{1} I had been wondering what it was that had made you break off sending me letters; my friend Pansa has now disclosed it --- you have turned Epicurean. What a splendid camp! What, I ask you, would you have done if I had packed you off to Tarentum, and not to Samarobriva? I had my doubts about you even back then, when you were maintaining the same opinions as my friend Zeius.

\ciceroSection{2} But how, I should like to know, will you uphold the civil law, when you do everything for your own sake, and nothing for the citizens'? Where, again, will that formula of the action on a fiduciary trust come in --- ``as it ought to be done well between honest men''? For who is the honest man, if a man does nothing except for his own sake? What rule will you lay down in the action for partitioning common property, when nothing can be ``common'' among men who measure everything by their own pleasure? And how will it suit you to swear by ``Jupiter of the Stone,'' when you know that Jupiter can be angry with no one? What will become, furthermore, of the people of Ulubrae, if you have decided that one ought not to take part in public life? \textit{[Greek: politeuesthai]} So if you are quite simply defecting from us, I take it ill; but if it is merely convenient to humour Pansa, I forgive you. Only do write to me at some point about what you are up to and what you want done for you, or attended to, at this end.
