% Ad Familiares 2.3 (ad-familiares-02-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. C. CVRIONI.}

\ciceroSection{1} Rupa's zeal in declaring games in your name has not been wanting; but it neither pleased me nor any one of yours that anything should be done in your absence which, when you arrived, would no longer be open for you to decide. My own opinion I shall either write to you afterwards at greater length or, so that you cannot rehearse against it, I shall catch you unprepared and speak my mind face to face against that whole scheme --- in order either to bring you over to my view, or at the least to leave on record with you what I thought, so that, should it ever happen --- as I would not wish --- that your own plan begins to displease you, you can recall mine. For the present, hold this much briefly: your return falls upon such a state of the times that, with the goods nature and study and fortune have given you, you can more easily attain everything that is most distinguished in the commonwealth by other means than by games. No man admires the capacity for these (it is a matter of means, not of merit), nor is anyone who is not by now wearied to the point of disgust with them.

\ciceroSection{2} But I am doing differently from what I had announced, in that I have started to lay out the grounds of my opinion; so the whole of this argument I defer to your coming. Be assured that you stand in the highest expectation, and that what is awaited of you is what is to be awaited of supreme virtue and supreme genius. To which, if you are prepared as you ought to be --- as I am confident you are --- you will gladden us, your friends, and your fellow citizens as a body, and the commonwealth itself with games of the greatest and most numerous kind. This much you will surely come to know: that no one is dearer to me than you, nor anyone whose company is more delightful.
