% Ad Familiares 7.18 (ad-familiares-07-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO TREBATIO S.}

\ciceroSection{1} I received a whole batch of letters from you at one go, though you had sent them off at different times. In them most of what you wrote pleased me --- they showed that you were now bearing that soldiering of yours with a steady mind, that you were a brave and steadfast man. These qualities I had for a little while felt lacking in you, not because I doubted your firmness, but because I supposed you were chafing more out of missing me. So press on as you have begun; endure that camp life with a stout heart. Much you will gain, believe me: I shall renew the recommendation, but at the right moment. Be assured that it is no greater concern to you that this departure of yours from my side prove as profitable as possible than it is to me. And so, since your sort of guarantees are flimsy, I have sent you a little Greek bond in my own hand. I should like you to inform me about the conduct of the Gallic war; for the bigger the coward, the more credit I give him.

\ciceroSection{2} But to come back to your letters: the rest is charming, but this puzzles me --- who is in the habit, when he writes in his own hand, of sending out several copies of the same letter? As for its being on a palimpsest, I commend the thrift; but I wonder what could possibly have been on that scrap of paper that you would rather scrape away than not write this --- unless perhaps it was your own legal forms, since I really do not suppose you are wiping out my letters to put your own in their place. Or do you mean by this that nothing is going on, that you are bored stiff, that you cannot even lay hands on a sheet of paper? Well, that is your own fault, for taking your modesty off with you and not leaving it here behind with us.

\ciceroSection{3} As for myself, when Balbus sets off to join your party I shall commend you to him in proper Roman style. If there is rather a longer gap in my letters than usual, do not be surprised; I was going to be away for the month of April. This letter I am writing in the Pomptine country, having put up at the villa of M. Aemilius Philemo --- from which I had already heard the buzzing of my own constituents, the ones you procured for me: it is well established that at Ulubrae a vast army of little frogs roused itself to do me honour. Look after yourself. The sixth day before the Ides of April, from the Pomptine.

\ciceroSection{4} The letter from you which I received through L. Arruntius I tore up, though it had done no harm: there was nothing in it that could not properly have been read out in a public assembly. But Arruntius said you had given him those instructions, and you yourself had added the same in writing. Very well, let that pass; but I am amazed that you have written nothing more to me since, especially with affairs in such a new state.
