% Ad Familiares 13.11 (ad-familiares-13-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO BRVTO S.}

\ciceroSection{1} Since I have always observed how zealously you take pains that nothing of mine should be unknown to you, I therefore do not doubt that you know not only of what municipality I am, but also how diligently I am accustomed to look after my fellow-townsmen of Arpinum. All their resources, all the means by which they are able to perform the sacred rites and to keep the sacred buildings and the public places in good repair, depend upon those revenues which they hold in the province of Gaul. To inspect these and to collect the moneys which are owed by the tenants, and to look into and administer the whole business, we have sent as legates Roman knights --- Q. Fufidius son of Quintus, M. Faucius son of Marcus, Q. Mamercius son of Quintus.

\ciceroSection{2} I beg of you most particularly, on the strength of our friendship, that you take this matter to heart and see to it that through you the municipality's business is administered as conveniently as possible and concluded as quickly as possible, and that the men whose names I have written you treat as honourably as your own nature dictates and as generously as possible.

\ciceroSection{3} You will have attached good men to your friendship; you will have bound a most grateful municipality to you by your kindness; and you will have done me a yet greater favour, in that --- although I have always made it my custom to look after my fellow-townsmen --- this year in particular touches my care and duty. For with a view to setting our municipality in order, I wanted my son made aedile this year, and my brother's son, and M. Caesius, a man closest to me; for that magistracy --- and no other --- is the one we are accustomed to elect in our municipality. You will be doing honour to them, and above all to me, if the affairs of the municipality have been well administered through your zeal and care. I most earnestly --- again and again --- ask you to do this.
