% Ad Quintum Fratrem 2.11 (ad-quintum-fratrem-02-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} I laughed at your ``black snow,'' and the fact that you are in such good spirits and so ready for a joke pleases me greatly. About Pompey I agree with you --- or rather, you agree with me. For, as you know, I have long sung the praises of your Caesar. Believe me, he is in my heart, and I do not loosen the bond. Now hear the news. It was the Ides of December that was the day appointed for Caelius.

\ciceroSection{2} Domitius did not have the jurors up to the required number. I am afraid that that vile and brutal man, Pola Servius, may come forward as prosecutor; for our Caelius is being attacked vigorously by the Clodian clan. Nothing is certain yet, but we are uneasy. On the same day, then, a crowded Senate gave audience to the Tyrians; against them, in equal numbers, the Syrian tax-farmers. Gabinius was savagely set upon; the tax-farmers, however, were harangued by Domitius for having gone out to escort him in on horseback. Our L. Lamia spoke a little fiercely: when Domitius said, ``This has happened by your fault, knights of Rome --- you are giving slack verdicts,'' he replied, ``\textit{We} give the verdicts; \textit{you} pronounce the praises.'' Nothing was settled that day; night broke it up.

\ciceroSection{3} On the comitial days that follow the Quirinalia, Appius is taking the line that he is not barred by the lex Pupia from holding a meeting of the Senate, and that, since it is sanctioned by the lex Gabinia, he is even compelled, from the Kalends of February right up to the Kalends of March, to grant the Senate daily to foreign embassies. So people think the elections will be pushed off into March. Still, on these very comitial days the tribunes of the plebs say they will move against Gabinius. I am gathering everything I can in order to write you something fresh; but, as you see, the matter itself is running thin on me.

\ciceroSection{4} So I come back to Callisthenes and Philistus, in whom I see you have been rolling about. Callisthenes, of course, is the commonplace and familiar article, in the manner of several Greek writers; that Sicilian of yours, though, is first-rate --- close-packed, sharp, brief, almost a Thucydides in miniature. But which of his books you have had (for they make up two corpora), or whether both, I do not know. The Dionysius pleases me more; for the man is himself a great old hand and an intimate of Philistus. But, since you write of it --- are you actually going at the history? With my encouragement you can; and since you keep me supplied with couriers, you shall have today's doings on the Lupercalia. Enjoy yourself with our dear Cicero as splendidly as possible.
