% Ad Quintum Fratrem 3.1 (ad-quintum-fratrem-03-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} I, after the great heats (we cannot remember greater ones), refreshed myself at Arpinum in the highest pleasantness of the river during the days of the Roman Games, while Philotimus had been commended to my fellow tribesmen. I was at Arcanum on the fourth day before the Ides of September [10 September]. There I saw Mescidius with Philoxenus, and the water which they were leading not far from the villa, flowing very prettily indeed, especially in this great drought; and they said they would gather it considerably more abundant. At the master's everything was in order. At the Manilian estate I came upon Diphilus more dilatory than Diphilus; but still nothing was left to him except the bath-rooms, the walking-gallery, and the aviary. The villa pleased me very much, because the paved colonnade has the highest dignity --- which has now at last revealed itself to me, since both it lies all open and the columns have been polished. The whole point in it, which I shall make my care, is that the stuccowork be neat. The pavements seemed to be properly done; certain vaultings I did not approve and I ordered them changed.

\ciceroSection{2} The place in the colonnade where you write that you want a small atrium made --- it pleased me more as it stands. For there did not seem to be enough room for a small atrium, nor is one usually made except in those buildings in which there is a larger atrium; and it could not have had bedrooms or that sort of room adjoining. Now \dots\ either by the dignity of the vaulting or by being a very fine cool place it will hold up as a summer spot. But still, if you think otherwise, write back as soon as possible. In the bath-rooms I have moved the dry-room to the other corner of the dressing-room, because they had been so set that the steam-flue out of which the fire breaks up was right under the bedrooms. The fairly large bedroom, and the second one, the winter one, I much approved, because they are both ample and set on one side of the walking-gallery, the side nearest the bath-rooms. The columns Diphilus had set neither straight nor in line. He shall, of course, pull them down. He will some day learn how to use plumb-line and ruler. On the whole I hope Diphilus's work will be finished within a few months; for Caesius, who was with me at the time, looks after it most diligently.

\ciceroSection{3} From that place we set off straight by the Vitularian road to the Fufidian property, which we had bought you at the last market days at Arpinum from Fufidius for 101,000 sesterces. I never saw a place more shaded in summer; in very many spots there is water flowing, and in plenty. What more? Caesius judged that you would easily irrigate fifty \textit{iugera} of meadow. As for me, I assert what I better understand: that you will have a wonderfully delightful villa, with a fish-pond and fountains added, and a wrestling-ground and a wood \dots\ As to that Bovillan estate, I hear you wish to keep it. About that, you yourself shall decide what seems best. \textit{[Calibus?]} he was saying that, with the water taken away and the rights of that water established, and a servitude imposed on that estate, we could even so keep up the price if we wished to sell. I had Mescidius with me. He said he had agreed with you at three sesterces a foot, but that he had measured himself the run as 3,000 paces. To me it seemed more; but I will guarantee that the outlay can be put nowhere better. I had sent for Cillo from Venafrum; but on that very day a tunnel had crushed four of his fellow slaves and apprentices at Venafrum.

\ciceroSection{4} On the Ides of September [13 September] I was at Laterium. I inspected the road; which pleased me so much that it seemed to be a public work, except for 150 paces (for I myself measured it) from that little bridge which is at Furina's shrine, towards Satricum. In that place dust has been thrown rather than gravel (and it shall be changed); and that part of the road is much sloped --- but I understood that it could not have been led otherwise, especially when you wanted neither to lead it through Locusta's land nor through Varro's. \textit{Velvinus(?)} had built well in front of his estate; Locusta had not touched it. Him I shall accost at Rome, and, I think, get him moving; and at the same time M. Taurus, who I hear has made you a promise, who was now at Rome, I shall ask about leading the water across his estate.

\ciceroSection{5} Nicephorus, your overseer, I quite approved, and asked of him whether you had given him any commission about that little house at Laterium of which you spoke with me. He answered me that he himself had been the contractor for that work for 16,000 sesterces, but that you had afterwards added many things to the work, and nothing to the price, so that he had given it up. By Hercules, it greatly pleases me that you should add what you had decided on; although the villa as it now stands seems to be a kind of philosopher woman who scolds the madness of the others. But still that addition will be a delight. The garden-keeper I praised; he has so clothed everything with ivy --- the foundation of the villa, the spaces between the columns of the walking-gallery --- that the cloaked Greeks themselves seem to be doing topiary and to be selling ivy. Now \textit{[Greek: apoduterion]} (the dressing-room): nothing cooler, nothing more mossy.

\ciceroSection{6} You have nearly the whole on country matters. The polishing of the town house Philotimus and Cincius press on, but I myself also look in often, which is easy to do. Wherefore I would have you freed from that care.

\ciceroSection{7} About my Cicero (the boy), as you keep asking me --- I forgive you indeed, but I would have you forgive me too. For I do not concede to you that you love him more than I do myself. And how I wish he had been with me these days at Arpinum, as he himself had wanted, and I no less! As to Pomponia, I would have you write, if it seems good to you, that when we go out somewhere she should come with us and bring the boy. I shall make it splendid if I have him with me at leisure; for at Rome there is no place to breathe. You know that I had promised you this before for nothing; what now do you suppose, with such great pay set forth to me from you?

\ciceroSection{8} I now come to your letters, of which I received several while I was at Arpinum; for in one day three were brought to me, and indeed sent, as it seemed, by you at the same time --- one of more words, in which there was first the matter that an earlier date had been set in your letter than in Caesar's. That Oppius does of necessity, sometimes: when he has fixed to send couriers and has received letters from us, if he is hindered by some new matter, he must of necessity send later than he had fixed; and we, the letters being already given, do not trouble to change the date. You write of Caesar's highest love towards us.

\ciceroSection{9} This both you will cherish, and we shall augment by every means we can. About Pompey I am acting diligently, and shall act, as you advise. As to your gratefulness for my permission of your remaining there --- I, with my highest grief and longing, am glad of it in part. About sending for Hippodamus and several others, I do not understand what you have in mind. There is none of those men but expects from you a gift like a suburban estate. As for my Trebatius, that you should mix him in there is unneeded. I sent him to Caesar, who has now satisfied me; if he himself is less satisfied, I owe to perform nothing; and I likewise from him take Trebatius back and free him. As to your writing that you are loved more by Caesar each day, I am everlastingly delighted; and Balbus, who is the assistant in that business as you write, I bear in my eyes. That my Trebonius is loved by you, and you by him, I greatly rejoice.

\ciceroSection{10} About the tribunate, of which you write: I indeed asked by name for Curtius, and Caesar himself wrote back that the post was prepared for Curtius by name, and rebuked my modesty in asking. If I shall ask for any one besides --- which I have also told Oppius to write to him --- I shall easily bear being refused, since those who are troublesome to me will not easily bear to be refused by me. I love Curtius, as I myself told him, not only on account of the request but also on account of your testimony, since from your letter I have plainly perceived his zeal for our preservation. About British matters I have learned from your letters that there is nothing either to fear or to rejoice at. About the public business which you wish Tiro to write to you of, I was earlier writing to you the more carelessly because I knew that everything, smallest and greatest, was being sent to Caesar.

\ciceroSection{11} I have replied to the largest letter; hear now of the little one, in which there is first the matter of Clodius's letter to Caesar. In which I approve Caesar's counsel, that he did not yield to your most loving request to write back any word to that fury. The other matter is about Calventius Marius's speech, which you write of. I am surprised that it should please you that I write back against it, especially when no one will read his speech if I write nothing back, but my speech against him every schoolboy commits to memory as if it were dictation. The books I am working on, which you are awaiting, I have begun, but cannot finish in these days. The orations urgently asked for, on behalf of Scaurus and on behalf of Plancius, I have completed. The poem to Caesar which I had begun I have given up. As for your request, since the very fountains are now thirsting, if I have any spare time, I will write.

\ciceroSection{12} I come to the third letter. That you say Balbus will be coming to Rome before long, well attended, and will be with me constantly until the Ides of May, is to me most welcome and most pleasant. As to your urging me, in the same letter as often before, to canvassing and to labour --- I will indeed do it; but when shall we live?

\ciceroSection{13} The fourth letter was delivered to me on the Ides of September [13 September], which you had given on the fourth day before the Ides of August [10 August] from Britain. In it there was nothing very new, except about Erigona (which, when I receive it from Oppius, I will write to you what I think of it; nor do I doubt that it will please me); and what I almost passed over, about the man you wrote of, who wrote to Caesar about Milo's applause. I indeed easily bear that Caesar should think the applause was as great as possible. And it really was so; and yet that applause which is given to him seems somehow to be given to us.

\ciceroSection{14} A very old letter has also been delivered to me, but late in arriving, in which you remind me about the temple of Tellus and Catulus's portico. Both are being attended to diligently. At the temple of Tellus indeed I have placed your statue too. Likewise as to the gardens, of which you remind me --- I never wanted them very greatly, and now my house supplies the pleasantness of gardens. When I had come to Rome on the thirteenth day before the Kalends of October [19 September], I found the roof finished on your house, the part above the chambers which you had not wished should be of many gables; this now slopes elegantly over the lower portico. Our Cicero (the boy), while I was away, was not idle at the rhetor's. About his learning, that you should worry, there is no need, since you know his talent and I see his industry. The rest of his concerns I take on as I think I ought to perform.

\ciceroSection{15} Gabinius is being prosecuted by three factions: L. Lentulus, the flamen's son, who has already laid the charge of \textit{maiestas}; Ti. Nero with good supporting subscribers; and C. Memmius, tribune of the plebs, with L. Capito. He came up to the city on the twelfth day before the Kalends of October [20 September]. Nothing more shameful, nothing more deserted: but I dare put no confidence in these trials. Because Cato was unwell, he had not yet been prosecuted on extortion. Pompey is pressing me hard to be reconciled with him; but he has so far made no progress, nor, if I shall hold any part of my freedom, will he. I am vehemently awaiting your letter.

\ciceroSection{16} As to your having heard that I took part in a coalition of the consular candidates, that is false. For such bargains were made in that coalition --- which Memmius afterwards laid bare --- that no good man should have taken part in them; and at the same time it was not for me to commit myself to such coalitions in which Messalla would be excluded. To Messalla I am giving every satisfaction, as I think, in all matters; and so to Memmius too. To Domitius himself I have already done many things which he wished and which he has asked of me. By the kindness of my defence I have very greatly bound Scaurus to me. So far it was very uncertain both when the elections would be and who the consuls would be.

\ciceroSection{17} As I was already folding up this letter, the couriers from you came on the eleventh day before the Kalends [21 September], on the twenty-seventh day. How distressed I am! How much I grieved at Caesar's most charming letter! But the more charming it was, the more grief did that misfortune of his bring. But I come to your letter. First, your remaining there I again and again approve, especially as, you write, you have shared this with Caesar. I am surprised that Oppius should have anything to do with Publius; for it had not pleased me.

\ciceroSection{18} As to what you write in the inner letter, that on the Ides of September I shall be made legate to Pompey: that I had not heard, and I wrote to Caesar that neither Vibullius had brought Caesar's instructions about my remaining to Pompey, nor had Oppius. With what counsel? Yet Oppius I held back, because Vibullius's part came first; for with him Caesar had spoken in person, to Oppius he had only written. But for myself I can have no \textit{[Greek: deuteras phrontidas]} (second thoughts) in matters concerning Caesar. He is to me, after you and our children, such that he is almost equal to you. I seem to do this by judgement; for I am now in his debt; but yet I am kindled with love.

\ciceroSection{19} When I had written these last lines in my own hand, your son Cicero came to me to dinner, since Pomponia was dining out. He gave me your letter to read which he had received a little earlier, in the manner of Aristophanes, very pleasing indeed and weighty by Hercules; with which I was extremely delighted. He gave me also the other one, in which you bid him stick to me as to a teacher. How those letters delighted him, how me! Nothing is more charming than that boy, nothing more loving towards us. This I have dictated to Tiro at dinner, that you may not wonder it is in another hand.

\ciceroSection{20} To Annalis your letter was most welcome --- both that you should look after his interests carefully and yet help him with the truest counsel. P. Servilius the elder signifies, from a letter which he said had been sent to him by Caesar, that you have done him a great kindness in speaking, about his goodwill towards Caesar, most courteously and most diligently.

\ciceroSection{21} When I had returned to Rome from Arpinum, it was told me that Hippodamus had set out to you. I cannot write that I was astonished he should have acted so inhumanly as to set out to you without my letter; this I do write, that it has been a trouble to me. For I had now for a long time been thinking, on the basis of what you had written to me, that, if I had anything I wanted to be carried more diligently to you, I would give it to him; for, by Hercules, in these letters which I send to you in the ordinary way I write almost nothing such that, if it should fall into anyone's hands, it would have to be borne ill. I was reserving for Minucius, Salvius, and Labienus. Labeo will either set out late or remain here.

\ciceroSection{22} Hippodamus did not even ask whether I wanted anything. T. Pinarius sends to me about you affectionate letters: that he is most delighted by your letters, conversations, dinners. The man has always been a delight to me, and his brother is much with me. So, as you have begun, embrace the young man.

\ciceroSection{23} As I have had the letter on hand for several days on account of the couriers' delay, on that account many things have been gathered, one at one time, another at another, like this: T. Anicius has often now told me that he would not hesitate to buy you a suburban estate if he found one. In his speech I usually wonder at both: that, when you write to him about buying a suburban property, you not only do not write to me but actually write in a different sense to me about a suburban property; and that, when you write to him, you remember nothing \dots\ those letters which you yourself showed me at his Tusculan villa, nothing of the precepts of Epicharmus: \textit{[Greek: gn\=othi p\=os all\=oi kechr\=etai]} ``know how he uses another'' --- in short, his whole face, his speech, his spirit, as I conjecture, \dots\ But these things you yourself shall see to.

\ciceroSection{24} About the suburban estate, see that I know what you wish, and at the same time see that he does not stir up any trouble. What besides? What? Yes: Gabinius, on the fourth day before the Kalends of October [28 September], entered the city by night; and today, at the eighth hour, when by C. Alfius's edict on \textit{maiestas} he ought to have been present, by a great gathering and by the universal hatred of the people he was almost laid low. Nothing is fouler than he. Yet next is Piso. So I am thinking of putting a marvellous \textit{embolium} in the second book of \textit{My Times}: Apollo speaking in the council of the gods, what kind of homecoming there would be of two imperators --- of whom the one had lost his army, the other had sold his.

\ciceroSection{25} From Britain Caesar gave me a letter on the Kalends of September [1 September], which I received on the fourth day before the Kalends of October [28 September], satisfactory enough about British matters, in which, that I should not be surprised at receiving none from you, he writes that he had been without you when he had advanced to the sea. To this letter of his I wrote nothing back, not even by way of congratulation, on account of his bereavement. I beseech you again and again, my brother, that you keep well.
