% Ad Familiares 13.64 (ad-familiares-13-64) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. P. SILIO PROPR.}

\ciceroSection{1} My dear Nero gave me extraordinary thanks to pass on to you --- quite incredible: he said no honour could possibly have been done him that you had passed over. You will reap great fruit from him, for there is no one more grateful than that young man; but by Hercules you have done me too the greatest favour, since out of the whole nobility I value none higher. So if you do those things which he wished arranged with you through me, you will do me a most welcome favour. First, in the matter of Pausanias of Alabanda, prop up the case until Nero arrives (I perceived that he wished it ardently for the man's sake; for that reason I press this on you most earnestly). Next, the Nysaeans, whom Nero counts among his closest connections and most diligently watches over and defends --- hold them as commended to you, so that that city may understand that it has the highest safeguard in Nero's patronage. Servilius Strabo I have often commended to you; the more earnestly now, because Nero has taken up his case. We ask only this: that you handle the matter so that you do not abandon an innocent man to the profit of someone unlike yourself. This will be welcome to me, and I shall further reckon that you have shown yourself true to your humanity.

\ciceroSection{2} The sum of this letter is this: that you adorn Nero by every means, as you have set out to do and as you have done. That province has a great theatre, not like our own here, for the commendation and glory of a young noble of talent and self-restraint. So if he has had you for his patron --- as he certainly will and has --- he will be able to confirm the most considerable clientships received from his ancestors and to bind them by his services. If in this kind of thing you assist him with the zeal you have shown, you will have placed it most splendidly with him --- and you will have done me too the greatest favour.
