% Ad Familiares 13.16 (ad-familiares-13-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CAESARI S.}

\ciceroSection{1} Of all the nobility, P. Crassus was the young man I loved best; and although I had conceived high hopes of him from his earliest years, I came to think exceedingly well of him once I had learned the outstanding judgements which you had passed on him. His freedman Apollonius --- already, while Crassus still lived --- I both held in high esteem and approved; for he was both devoted to Crassus and strikingly suited to his master's best pursuits, and so was very dear to him.

\ciceroSection{2} After Crassus's death, however, he seemed to me all the more deserving to be received into my trust and friendship, in that he held it his duty to honour and cultivate those whom Crassus had loved and to whom he had been dear. And so he came to me in Cilicia, and was of great use to me in many matters --- both his loyalty and his good sense; and to you, as I believe, in the Alexandrian war, so far as zeal and fidelity could compass it, he did not fail.

\ciceroSection{3} In the hope that you too thought of him in this way, he has set out to you in Spain --- chiefly indeed on his own counsel, but with my encouragement as well. To him I did not promise a commendation: not because I doubted that one of mine would carry weight with you, but because he did not seem to me to need one --- a man who had been with you in the war and who, on account of his memory of Crassus, counted as one of your own; and who, if he wished to use commendations, I saw could obtain that result through others as well. The witness of my own opinion of him --- which he himself prized highly, and which I had had experience of as carrying weight with you --- I gave him gladly.

\ciceroSection{4} He is, then, as I have come to know him, a learned man, and devoted to the best studies, and so from boyhood; for in my house with Diodotus the Stoic --- by my judgement a most erudite man --- he spent much of his boyhood. Now, fired with enthusiasm for your achievements, he was wishing to commit them to writing in Greek. I think he can: he has the talent, he has the practice, he has long worked in that kind of study and writing, and he wonderfully wants to do justice to the immortality of your praises. There you have the witness of my opinion --- but this you will judge much more easily by your own singular discernment. And yet, for all that --- contrary to what I had said I would do --- I commend him to you. Whatever you may oblige him in, that will be in the largest degree welcome to me.
