% Ad Atticum 4.15 (ad-atticum-04-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About Eutychides --- a kindness. With his old praenomen and your new nomen, he will be T. Caecilius, just as out of me and you Dionysius has been compounded into M. Pomponius. By Hercules, I am genuinely grateful that Eutychides has come to know your good will toward me, and that his own fellow-feeling \textit{[Greek: sympatheian]} in my time of grief was neither hidden from me at the time nor ungratefully received afterward.

\ciceroSection{2} Your journey to Asia, I take it, was one you had to undertake; for you would never have wanted, without the most justified cause, to be so far away from so many people and things that are most dear and pleasant to you. But your humanity and your love for your own will be made clear by the speed of your return. Yet I fear that you may be kept longer by the charm of the praetor Clodius, and of Pituanius, a man very learned, as they say, and just now devoted to Greek letters. But if you mean to behave like a human being, take yourself back to us by the time you fixed. With those others you can live at Rome too, once they have come home safe. You write that you are eager to receive some sort of letter from me.

\ciceroSection{3} I have sent letters, and on many subjects, written out day by day \textit{[Greek: hēmerolegdon]} from start to finish; but as I conjecture, since you do not seem to me to have been long in Epirus, I judge they were not delivered to you. The general kind of letter I write to you is such, however, that one does not care to entrust it to just anyone, but only to a man one has tested and knows will deliver it to you.

\ciceroSection{4} Now for the news from Rome. On the fourth day before the Nones of Quintilis, Sufenas and Cato were acquitted, Procilius condemned. From which it has been understood that our thrice-honoured Areopagites \textit{[Greek: trisareiopagitas]} do not care a straw for electoral bribery, the elections, the interregnum, treason --- in short, for the commonwealth in any form; but that we are not, all the same, to allow a head of a household to be murdered in his own home --- and even on that point, not overwhelmingly so; for twenty-two voted to acquit, twenty-eight to condemn. Publius, with a really eloquent peroration accusing me, had stirred the minds of the jurors. Hortalus performed in that case in his usual style. I said not a word; for my little one, who is now in poor health, was afraid I would offend Publius's feelings.

\ciceroSection{5} Once these matters were dispatched, the people of Reate took me off to their own Tempe \textit{[Greek: Tempē]}, so that I might plead their case against the people of Interamna before the consul and the ten commissioners --- the question being that the Lacus Velinus, drained off by M'. Curius through a mountain cut, flows down into the Nar; from which drainage that famous Rosian plain has been dried out, though still moderately moist. I lived with Axius, who also took me to the Seven Waters.

\ciceroSection{6} I came back to Rome on Fonteius's behalf on the seventh day before the Ides of Quintilis. I went into the theatre, first off, to a great and steady round of applause. But never mind that; how silly of me to have written it. Then I attended to Antiphon. He had been manumitted before he was brought on. Not to keep you in suspense any longer, he carried off the prize; but nothing so undersized, nothing so voiceless, nothing so --- well, keep this between us. In the \textit{Andromacha}, however, he was taller than Astyanax; in the rest he had no equal. You ask now about Arbuscula; she gave great satisfaction. The games were magnificent and well received; the wild-beast hunt has been postponed to another time.

\ciceroSection{7} Now follow me into the Campus. Electioneering is on fire; ``and I will tell you the sign'' \textit{[Greek: sēma de toi ereō]}: interest has gone from a third to two-thirds since the Ides of Quintilis. You will say, ``Well, that does not trouble me.'' What a man! What a citizen! Memmius is being shored up by all of Caesar's resources. The consuls have joined Domitius to him --- on what terms I dare not commit to a letter. Pompey roars, complains, throws his support to Scaurus --- but whether on his face or in his mind, no one can tell. There is no preeminence \textit{[Greek: exochē]} in anyone; money is leveling everyone's standing. Messalla is flagging --- not for any want of spirit or of friends, but the consuls' compact and Pompey stand against him. I think these elections are going to be postponed. The tribunician candidates have sworn to canvass under Cato's arbitration. With him each has deposited 500,000 sesterces, on terms that any one of them whom Cato condemns shall forfeit his sum, and it shall be paid over to his competitors.

\ciceroSection{8} I am writing this the day before the elections are thought to be held. But on the fifth day before the Kalends of Sextilis I will write the whole election out for you, if it has come off and the courier has not yet set out. If, as they expect, it goes uncorrupted, Cato single-handed will have done more than all the laws and all the courts put together.

\ciceroSection{9} I was defending Messius, recalled from his commission --- for Appius had attached him to Caesar as legate. Servilius issued an edict that he should appear. He has three tribes: Pomptina, Velina, Maecia. The fight is sharp; the case, all the same, goes well enough. Next I free myself up for Drusus, then for Scaurus. Splendid title-pages are being readied for my speeches. Perhaps the consuls-designate will be added to my list as well. If Scaurus is not among them, he will be in serious trouble at this trial.

\ciceroSection{10} From my brother Quintus's letter I suspect he is now in Britain. With suspense in my heart I wait to hear what he is doing. This much we have at least secured, and can judge by many great signs: that we are exceedingly dear and exceedingly welcome to Caesar. Be so good as to give Dionysius my greeting, and to ask and urge him to come as soon as possible, so that he may go on educating my Cicero --- and me, as well.
