% Ad Familiares 13.56 (ad-familiares-13-56) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO THERMO PROPR. S.}

\ciceroSection{1} Cluvius of Puteoli pays me close attention and is a particular intimate of mine. He has so persuaded himself that, whatever business he has in your province, unless during your governorship he settles it through my recommendations, he will reckon it among his lost and hopeless causes. Now since so heavy a burden is laid on me by a most attentive friend, I in turn shall lay one on you, in proportion to your supreme services to me --- yet only so far as I would not wish to be a nuisance. The Mylasians \textit{[Greek: Mulaseis]} and the Alabandans \textit{[Greek: Alabandeis]} owe Cluvius money. Euthydemus had told me, when I was at Ephesus, that he would see to it that legal representatives \textit{[Greek: ekdikoi]} were sent to Rome from Mylasa. That has not been done. I hear envoys have been sent, but I prefer the legal representatives, so that something may be brought to a conclusion. So I ask of you that you order both them and the Alabandans to send legal representatives to Rome.

\ciceroSection{2} Furthermore, Philocles of Alabanda gave Cluvius mortgages \textit{[Greek: hupothekas]}. They have been forfeited. I should like you to see to it that he either yield up the mortgages and hand them over to Cluvius's agents, or pay the money. Furthermore the Heracleotans and the Bargylietae, who likewise owe, are either to pay the money or satisfy the debt with their produce.

\ciceroSection{3} The Caunians, in addition, owe; but they say they have lodged the money in deposit. I should like you to look into this and, if you find that they have lodged it on deposit neither according to the edict nor according to a decree, to see to it that the interest for Cluvius is preserved according to your established practice. About these matters I am the more concerned, because the affair of Gnaeus Pompey, our intimate friend, is also at stake --- and because he seems to me to take it even more to heart than Cluvius himself, whom I am very anxious that we should satisfy. About these things I beg you again and again most urgently.
