% Ad Familiares 1.9 (ad-familiares-01-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. LENTVLO IMP.}

\ciceroSection{1} Your letter was most pleasing to me, since from it I perceived that you see clearly my devotion toward you --- for why should I speak of goodwill, when that name of devotion itself, the gravest and most sacred of names, seems to me too light a word for what you have deserved of me? That you write you find welcome the zeal I have shown for you --- this you do out of a certain abundance of love, that even those things should be welcome which cannot be omitted without unholy crime. To you, however, my feelings toward you would be far better known and more plainly displayed, if throughout this whole time during which we have been parted we had been both together and at Rome.

\ciceroSection{2} For in that very matter which you say you intend to take up --- a thing for which you have the foremost capacity, and which I most warmly look for from you --- in senatorial debate and in the whole conduct and administration of the commonwealth, we should have flourished together (about which I shall presently show what my own feeling and standing are, and shall write back to you in answer to your questions); but surely both I, with you as my counsellor most friendly and most wise, and you, with me as an adviser perhaps not too inexperienced, and faithful and well-disposed at the least, would have had the use of one another. Although for your own sake, as I should, I rejoice that you are imperator and hold a province with successful affairs and a victorious army; yet surely those fruits which are owed to you from me you could have reaped more richly and more present, had you been present yourself. As for taking vengeance on those whom you understand to be your enemies in part because of your championing of my safety, in part to envy you for the greatness and glory of that action --- I should have shown myself a wonderful comrade to you in it. Although that perennial enemy of his own friends, who, decked out by your greatest kindnesses, brought his broken and crippled force to bear chiefly against you, has avenged our cause for us by his own hand; for he attempted such things that, with them brought to light, he has left himself no share for the future, not of dignity, certainly, but not even of liberty.

\ciceroSection{3} As for you --- though I should have preferred you had made trial of men's loyalty in my affairs rather than in yours as well --- still amid your trouble I rejoice that you have come to know at no very great price what I had come to know at the greatest grief. On this whole matter the time now seems to me given for laying it out, so that I may write back to you in answer to what you ask.

\ciceroSection{4} You write that you have been informed by letter that I am on good terms with Caesar and with Appius, and you add that you do not reprehend it; about Vatinius, however, you show that you wish to know by what considerations I was moved to defend and to praise him. That I may set this out for you more plainly, I must trace the rationale of my counsels a little further back. I, Lentulus, at the outset of your acts and undertakings, considered myself restored not only to my own people but also to the commonwealth; and since I owed to you yourself an incredible love and the sum of singular zeal in every respect, to the commonwealth, which had much aided you in restoring me, I judged that I owed for its own deserving sake at any rate that disposition which before I had merely rendered as the common duty of citizens, not as anything owed for some singular kindness toward me. That this was my mind the Senate heard from me when you were consul, and you yourself saw in our conversations and discussions.

\ciceroSection{5} And yet, even in those first days, my mind was offended by many things, when, while you were acting on behalf of the rest of our dignity, I discerned either the hidden hatreds of certain men or the obscure ill-will toward me of others. For neither concerning my monuments was I aided by those by whom I ought to have been, nor concerning the unholy violence by which I had been driven from my house with my brother, nor, by Hercules, in those very matters which, though they were necessary to me on account of the shipwreck of my estate, were yet held by me of the smallest account --- in the making good of my losses on the authority of the Senate --- did they show that goodwill which I had looked for. Though I saw these things (and they were not obscure), still they did not strike me as bitterly as the things they had done were welcome.

\ciceroSection{6} And so, although I owed a great deal to Pompey --- with you yourself as the publisher and witness of it --- and although I cherished him not only for his kindness but also out of love and a certain settled judgement of my own, still, not reckoning what he might wish, I remained in all my views about the commonwealth as I had been before.

\ciceroSection{7} With Gnaeus Pompey sitting by, when he had come into the city to give testimony in praise of Publius Sestius and Vatinius the witness had said that I had begun to be Caesar's friend because I was moved by Caesar's fortune and happiness, I said that I preferred the fortune of Marcus Bibulus --- which Vatinius thought broken --- to all the triumphs and victories of all men, and I said, with the same man as witness, in another place, that those who had kept Bibulus from leaving his house and those who had compelled me to leave mine were one and the same. Indeed my whole questioning held nothing but the censure of that tribunate of his. In which everything was said with the greatest freedom and spirit, of his violence, of the auspices, of the gift of kingdoms --- and not only in that case, but steadily and often in the Senate.

\ciceroSection{8} What is more, in the consulship of Marcellinus and Philippus, on the Nones of April, the Senate gave its assent to me that the Campanian land should be brought forward for a full Senate on the Ides of May. Could I have invaded more deeply into the citadel of that cause, or have forgotten my own circumstances more, while remembering my acts? When I had spoken this motion, a great stir of feeling was made, both among those whose stir was proper and among those of whom I had never imagined it.

\ciceroSection{9} For this senatorial decree had been passed on my motion when Pompey, having shown me no offence, set out for Sardinia and Africa, and on that journey came to Caesar at Luca. There Caesar complained much of my proposal, since he had even before that seen Crassus at Ravenna and been inflamed by him against me. It was plain that Pompey took it sorely amiss; which when I had heard from others, I learned most clearly from my own brother. For when Pompey had met him in Sardinia, a few days after he had departed from Luca, he said, ``You are the very man I want; nothing more opportune could have happened. Unless you deal carefully with your brother Marcus, you must pay what you have pledged me on his behalf.'' Why say more? He complained gravely; he recalled his own services; he reminded my brother how often he had spoken with him of Caesar's acts and what my brother had undertaken to him about me; and he called my brother himself to witness that what he had done about my safety he had done with Caesar's goodwill. Recommending Caesar's cause and dignity to me, he asked that I should not assault it, if I were unwilling or unable to defend it.

\ciceroSection{10} When my brother had brought these things to me, and when nevertheless Pompey had sent Vibullius to me with charges that I should keep the Campanian question untouched until his return, I collected myself and conferred, as it were, with the commonwealth herself: that she would grant to me, after I had suffered and undergone so much on her behalf, that I should fulfil my duty, and the disposition of my mind mindful of those who had deserved well of me, and the pledge of my brother; and that she would suffer that man, whom she had always held a good citizen, to be also a good man. Now in those acts of mine and in all those motions which seemed to give offence to Pompey, the words of certain men --- whom you ought already to suspect --- were reported to me; men who, though they felt as I did in the commonwealth and had always so felt, nevertheless said they were glad that I was not satisfying Pompey, and that Caesar would prove to me a most bitter enemy. This was to be grieved at; but much more so the fact that they so embraced my enemy --- my enemy, did I say? --- nay rather, the enemy of the laws, of the courts, of public peace, of fatherland, of all good men --- so held him in their hands, so cherished him, so kissed him in my presence; not, indeed, to vex my stomach, which I have lost utterly, but certainly with the intention of thinking they did so. At this point I, as far as human counsel could effect, looking around upon all my affairs and casting up the reckoning, made the sum of all my deliberations; which I shall set out to you, if I can, briefly.

\ciceroSection{11} If I were to see the commonwealth held by base and ruined citizens, as we know happened in the days of Cinna and in not a few other times, I should not, not for rewards (which weigh least with me), nor even for any dangers (by which, however, even the bravest men are moved), join myself to their cause, not even if their highest services to me were beyond question. But since in the commonwealth Gnaeus Pompey was the first man --- a man who had obtained this power and glory by the greatest services to the commonwealth and by the most outstanding achievements, and of whose dignity I had from boyhood been a champion, and in his praetorship and his consulship had even stood as his helper; and since he had aided me with his own authority and vote, and aided me through you with his counsels and zeal, and held my one enemy as his one enemy in the state --- I did not think I had to fear the charge of inconsistency, if in certain of my motions I altered myself a little, and gathered my goodwill onto the side of the dignity of the highest of men, who had deserved most excellently of me.

\ciceroSection{12} In this view Caesar had to be embraced by me, as you see, in joined cause and dignity. Here there weighed much both the old friendship, which you know existed between me and my brother Quintus and Caesar, and his humanity and liberality, in a short space of time perceived and known to us through letters and acts of courtesy. The very state of the commonwealth itself also moved me strongly, which seemed to me unwilling that any contention should arise with those men, especially with the greatest deeds carried through by Caesar; and to refuse vehemently that it should arise. The greatest weight upon my mind in this conviction was the pledge of Pompey, which he had given to Caesar about me, and that of my brother, given to Pompey. Besides, these things had to be considered in the state, things which are written with divine insight in our Plato: that the kind of leading men in a commonwealth tend to determine what kind the rest of the citizens are. I held in memory that, in our own consulship, the foundations of strengthening the Senate had been laid from the very Kalends of January onward --- so that none ought to wonder that on the Nones of December there had been so much spirit and so much authority in that order; and I remembered too that, while we were a private citizen down to the consulship of Caesar and Bibulus, while our motions had great weight in the Senate, the feeling of all good men was almost one.

\ciceroSection{13} Afterward, when you were holding Hither Spain with military command and the commonwealth had no consuls but only traders of provinces and slaves and ministers of sedition, a certain chance flung my head as if for the sake of a contest into the midst of civil strife and dissension. In which crisis, when there had stood out a wonderful resolve of the Senate, an incredible one of all Italy, a singular one of all good men, in my defence --- I shall not say what happened (the fault is many men's, and various). Let me only say briefly: it was not an army that failed me, but generals. In which, granting there is now fault in those who did not defend me, it is no less in those who deserted me; and if those who took fright are to be accused, those who pretended to be afraid deserve still more reproach. Surely that counsel of mine at least deserves to be rightly praised: that I refused to have my citizens, both saved by me and eager to save me, exposed unarmed to slaves stripped of leaders and armed, and chose rather to make clear how great force there could have been in the consensus of good men, had they been permitted to fight for me standing my ground, when they had power to raise me from the dust. Their spirit you not only saw through, when you acted on my behalf, but you confirmed and held it firm.

\ciceroSection{14} In that cause --- and not only will I not deny it but I will always remember it and willingly publish it --- you had the use of certain of the most noble men, braver in restoring me than the same men had been in keeping me. If they had been willing to stand firm in that view, they would have recovered their own authority along with my safety. For when the good men had been revived by your consulship and roused by your most steady and best acts, with Gnaeus Pompey especially joined to the cause --- when even Caesar, decorated by the greatest deeds done with singular honours and new judgements of the Senate, was attaching himself to the authority of that order --- no place could have been left for any base citizen to do violence to the commonwealth.

\ciceroSection{15} But attend, I beg you, to what has followed. First, that fury of feminine rites --- who had set no greater store by the Good Goddess than by his three sisters --- obtained his impunity by the votes of those men who, when the tribune of the plebs was willing to pursue the punishment of a seditious citizen through good men in a regular trial, removed from the commonwealth the most brilliant precedent for vindicating sedition in time to come; and afterwards those same men suffered not my monument (for that was no spoil of mine, but the contracting of the work had been mine) --- the Senate's monument, in truth, branded with a hostile name and with bloody letters, to be allowed to stand. That these men wished my safety is most welcome to me; but I could wish they had been willing to have regard not only to my safety, as physicians do, but also, as trainers do, to my strength and my colour. As Apelles brought to finish the head of his Venus and the upper part of the breast with the most exquisite art but left the rest of the body sketched in and rough, so certain men have laboured upon my head alone, and left the rest of the body unfinished and rude.

\ciceroSection{16} In which I have disappointed the expectation not only of those who envied me but even of my enemies, who, concerning one most fierce and bravest man, in my judgement most outstanding of all in greatness of spirit and steadiness, Quintus Metellus son of Lucius, once took up a mistaken opinion: that he was broken in spirit and downcast after his return (yes, it must be granted: a man who withdrew with the highest goodwill and was absent with outstanding alacrity of spirit and took no real pains to return --- this man was on that very score broken, when in the same matter he had surpassed all men, and in particular Marcus Scaurus that singular man, in steadiness and gravity!). But what they had taken up about him, or even suspected, they were thinking the same of me: that I should be more downcast in spirit, when the commonwealth was giving me even greater spirit than I had ever had, since she had declared that she could not do without me as a single citizen, and since whereas Metellus had been recovered by the proposal of a single tribune of the plebs, I had been recovered by the whole commonwealth, with the Senate as its leader, with Italy as its escort, with all the magistrates promulgating, with you the consul carrying the law, in the centuriate assembly, with all the orders and men leaning on the cause --- with all her own strength, in a word.

\ciceroSection{17} Nor in truth have I taken upon myself anything afterwards, nor do I take upon myself today, that could rightly offend even the most malevolent of men; I only strive that neither to my friends, nor even to those more removed, I may be wanting in service, counsel, and labour. This course of my life offends, perhaps, those who look at the splendour and show of this life but cannot see through to the anxiety and labour of it; but this they openly complain of, that in my motions in which I adorn Caesar I am as if seceding from my old cause. I, however, while I follow what I have proposed a little above, am also moved by this --- not the last among my reasons --- which I had begun to set out. You will not find the same disposition of the good men, Lentulus, that you left: which, confirmed in our consulship, broken at times afterwards, and crushed before your consulship, restored entire by you, has now been deserted by those by whom it ought to have been kept safe. And this they declare not only by brow and countenance, by which dissimulation is most easily upheld, but more often now also by their feeling and their ballots, have shown.

\ciceroSection{18} And so the whole opinion and will of the wise citizens --- such as I both wish to be and to be counted --- ought now to have changed. For that very Plato whom I follow with vehemence as my authority enjoins this: to strive in the commonwealth so far as you can carry your fellow citizens with you; that violence is not rightly brought against parent or fatherland. And this indeed he says was his reason for not approaching the commonwealth: that, when he had found the Athenian people now nearly doting with old age, and saw that it could neither be ruled by persuasion nor by compulsion, when he despaired that it could be persuaded, he judged that it was not lawful that it should be forced. My reasoning was different: for being neither held entangled by a doting people nor by the matter still open for me to deliberate whether I should grasp at the commonwealth, I rejoiced however that it was allowed me in one and the same cause to defend what was useful to me and what was right for any good man. To this was added a certain memorable and god-like liberality of Caesar toward me and toward my brother; who, whatever he were doing, would have to be supported by me, but now, in such felicity and amid such victories, even if he were not toward us what he is, would still seem to deserve honour. For so I would have you think: that, when I have left aside you, the authors of my safety, there is no one to whose services I confess myself, and rejoice to confess myself, so bound. Now that I have set this out to you, what you ask of me about Vatinius and Crassus is easy.

\ciceroSection{19} For as to your writing about Appius, just as about Caesar, that you do not reprehend me, I rejoice that my counsel is approved by you. As for Vatinius: first, a reconciliation had intervened through Pompey, just as soon as he was made praetor --- though, indeed, I had attacked his candidacy with the gravest speeches in the Senate, not so much for the sake of injuring him as for the sake of defending and honouring Cato. Afterwards there followed an extraordinary insistence from Caesar that I should defend him. Why I praised him, I beg you not to require of me either in this defendant or in others, lest I in turn require the same of you when you come back. Although I can do so even with you absent: recall those for whom you have sent testimonials of praise from the ends of the earth. And do not fear this, for those same men are and shall be praised by me as well. But still, of defending Vatinius there was also that goad of which, in the trial when I was defending him, I said that I was doing something which in the \textit{Eunuchus} the parasite advises the soldier:

\begin{quote}
When she shall name Phaedria, do you straightway bring up Pamphila; if ever she shall say, ``Let us bring Phaedria in to revel,'' let us call up Pamphila to sing; if she shall praise the looks of one, do you praise those of the other; in short, give like for like, that it may sting her.
\end{quote}

So I asked of the jurors that, since certain noble men, men also who had deserved most excellently of me, loved my enemy too much, and in my own sight in the Senate sometimes drew him aside severely, sometimes embraced him familiarly and cheerfully --- since they had their own Publius, that they should give to me too another Publius, in whom I might, with mild provocation, prick their feelings a little. And not only have I said this, but I do it often, with gods and men approving.

\ciceroSection{20} You have my account of Vatinius; now hear about Crassus. I, when there was now a great reconciliation between me and him --- because, for the sake of common concord, by a kind of voluntary forgetting, I had rubbed away all the gravest injuries done me by him --- still, when he undertook the sudden defence of Gabinius (whom in the days immediately preceding he had most fiercely attacked), I should yet, if he had taken it up without any insult to me, have borne it. But when he had wounded me as I was debating, not as I was provoking, I blazed out --- not so much, I believe, with present anger (for that would perhaps not have been so vehement) --- but because, that pent-up hatred of his many injuries to me, which I had thought I had wholly poured out, had still been left over, unknown to me, suddenly the whole appeared. At which very time certain men --- the same ones whom I often hint at and do not name --- when they said they had taken the greatest enjoyment of my freedom, and that then at last I seemed to them restored to the commonwealth such as I had been; and when that contention had brought me a great fruit even abroad, they said they rejoiced that I had both that man as an enemy and that those men who were in the same cause with him would never be my friends. Their unjust talk being brought to me by men of the highest standing, and Pompey having pressed --- as never anything more --- that I should be reconciled to Crassus, and Caesar having shown by letter that he was affected with the greatest distress by that quarrel: I had regard not only to my circumstances but also to my own nature; and Crassus, that our reconciliation might be as it were attested before the Roman People, set out for his province almost from my own household gods. For having arranged it beforehand with me, he dined with me in the gardens of my son-in-law Crassipes. For which reason his cause, which you write you have heard of, I undertook, on his strong commendation, and defended in the Senate, as my own good faith required.

\ciceroSection{21} You have received, then, the considerations by which I was moved, and what matter and what cause I have defended, and what my standing is in the commonwealth for the part I undertake. Of which I would have you decide as follows: that I should have felt these same things, had everything been wholly free and intact for me; for I should have judged neither that I must fight against such great resources, nor that the supremacy of the highest citizens must be destroyed, even if that could be done, nor that I must persist in one opinion when circumstances had turned and the wills of good men had altered; rather, that I must yield to the times. For never in eminent men of the governance of the commonwealth has a perpetual standing-fast in one opinion been praised. As in sailing it is the part of skill to give way to the storm, even if you cannot make port --- but when, with sail changed, you can make it, it is folly to keep with danger the course you had begun rather than, by changing it, to come nonetheless to the place you wish: so, when all of us in the administration of the commonwealth ought to have set before us that which has been said by me most often, ``tranquillity with dignity,'' we ought always to look to the same thing, not always to say the same thing. For which reason --- as I laid down a little before --- if all things were freest to me, I should still in the commonwealth be no different from what I now am; and since I am both drawn into this feeling by the kindnesses of men and driven into it by their wrongs, I easily suffer that I should feel and say about the commonwealth those things which I judge to conduce most both to me and also to the reckonings of the commonwealth. And I do this more openly and more often because my brother Quintus is Caesar's legate, and no smallest word of mine, not to mention deed, has been put forward on Caesar's behalf which he has not received with so brilliant a gratitude that I think him bound to me. Accordingly, of all his goodwill, which is the greatest, and of his resources, which you understand to be the greatest, I have the use as of my own; nor do I see how it could have been otherwise possible for me to break the designs of ruined men against me, unless to those defences which I have always had I now joined also the goodwill of the powerful.

\ciceroSection{22} These counsels, if I had had you present, I should, as my judgement bears, have used the same ones; for I know the temperance and moderation of your nature, I know your mind, both most friendly to me and tinged with no ill-will toward others, and on the contrary both great and lofty and also open and simple. I have myself seen in certain men toward you the same dispositions which you might have seen those same men show toward me; the things which moved me would surely have moved you. But, at whatever time the privilege of your presence shall be mine, you shall be the moderator of all my counsels; to you also, who had my safety in care, my dignity shall be in care; me at any rate you shall have as the partner and companion of your acts, your motions, your wishes, in short of all your affairs, nor shall anything in all my life be set before me so much as that you may rejoice every day more deeply that you have deserved most excellently of me.

\ciceroSection{23} As to your asking that I send you my writings which I have composed since your departure --- there are some speeches, which I will give to Menocritus, and not so many that you need be alarmed. I have also written --- for I disengage my mind from speeches for the most part and bring it back to the gentler Muses, which from my earliest youth have been my chief delight --- I have written, then, after the Aristotelian manner (in the way I wished, at any rate), three books in disputation and dialogue \textit{On the Orator}, which I think will not be useless to your young Lentulus; for they depart from the common precepts and embrace the whole rhetorical theory of the ancients, both the Aristotelian and the Isocratean. I have also written in verse three books \textit{On My Times}; which I would have sent to you long ago, if I had thought they were to be published; for they are witnesses, and shall be everlasting witnesses, of your services to me and of my devotion; but, because I was afraid --- not of those who would think themselves wounded (for that I have done sparingly and gently), but of those whom it would be endless to name, all the men who have deserved well of me. These books, however, if I find anyone to whom I may rightly entrust them, I shall see to having brought to you. And that whole portion of my life and habit I make over to you: as much as we shall be able to attain by letters, by studies, by our old delights --- all this most willingly we shall bring under your judgement, who have always loved these things.

\ciceroSection{24} The things you write to me about your domestic affairs, and which you commend to me, are of so great a care to me that I wish not to be reminded, and can hardly be asked without great grief. As to your writing about my brother Quintus's business --- that you could not last summer, because hindered by illness from crossing into Cilicia, complete it, but now will do everything to complete it --- know this to be of such a kind that my brother truly judges that, with that estate added, his patrimony will have been settled through you. I would have you most familiarly and most often inform me of all your affairs, and of your young Lentulus's, and our young Lentulus's, studies and exercises; and reckon that no one has ever been dearer or more pleasant to anyone than you are to me; and that I shall act not only that you may feel it, but that all nations, even that all posterity, may understand it.

\ciceroSection{25} Appius used to say in conversations earlier, and afterwards said openly even in the Senate, that, if he were allowed to carry a curiate law, he would draw lots for the province with his colleague; if there were no curiate law, he would prepare to succeed you in concert with his colleague; and that the carrying of a curiate law was useful for a consul but not necessary; that, since he held a province by senatorial decree, he would have his command under the Cornelian law until he had entered the city. What each of your friends is writing to you, I do not know; I understand opinions are various. Some think you may stay, since you are succeeded without a curiate law; some think also that, if you do depart, someone may be left by you to govern the province. To me there is not so much doubt about this (though even that is not very doubtful) as about this: that it pertains to your highest greatness, dignity, freedom --- in which I know you most willingly take pleasure --- that without any delay you should yield the province to your successor; especially since you cannot refute his ambition without the suspicion of your own. I think both belong to me: both to show what I feel and to defend what you shall have done.

\ciceroSection{26} With my letter already written, I received your letter about the publicans, in which I could not but approve your fairness. I could have wished you had been able by good fortune to attain what I would have wished: that you should not offend the interest or the goodwill of that order which you have always honoured. For my part, I shall not cease to defend your decrees; but you know the habit of men; you know how grievously enemies were even to that famous Quintus Scaevola himself. Yet I am the one who advises you that, if you can by any means, you should either reconcile that order to you or soften it. This, though hard, seems to me to be a matter for your prudence.
