% Ad Familiares 13.55 (ad-familiares-13-55) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO THERMO PROPR. S.}

\ciceroSection{1} Even though I think I made it plain, when I spoke with you at Ephesus about the case of Marcus Anneius, my legate, that you wholeheartedly wish everything done for his own sake, still I value Marcus Anneius so highly that I think I must omit nothing that touches his interests; and I take it that I am so highly valued by you that I do not doubt that to your goodwill my commendation will add a great heap. For though I had long held Marcus Anneius dear, and so esteemed him --- as the fact itself declares --- that I voluntarily offered him a legateship when I had refused it to many petitioners, yet much more, after he had been with me in war and on campaign, I came to know in him such virtue, such prudence, such fidelity, and such goodwill toward me that I rate no man higher. You know that he has a dispute with the people of Sardis. I laid the case before you at Ephesus; in person, however, you will grasp it more readily and better.

\ciceroSection{2} For the rest, by Hercules, I was long in doubt what above all I should write to you. For your way of administering justice --- clear and famously to your credit --- is well known, and in this case I need nothing other than that you administer justice in your established way. And yet, since I am not unaware how great is the authority of a praetor --- particularly one of that integrity, gravity, and clemency which it is universally agreed are yours --- I ask of you, in the name of our most close-knit connection and of the many equal and reciprocal good offices between us, that by your goodwill, your authority, and your zeal you bring it about that Marcus Anneius perceive that you are his friend (which he does not doubt, for he has often said as much to me) and that he has been made far more your friend by this letter of mine. In your whole command and province there is nothing you could do that would be more welcome to me; for as to whether your zeal and service will be well placed with the man himself --- a most grateful man and an excellent one --- I do not suppose you have any doubt.
