% Ad Quintum Fratrem 2.12 (ad-quintum-fratrem-02-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} I have so far received two letters from you: of which one given at the very moment of our parting, the other given at Ariminum; the several others which you write you sent I had not received. I, at my Cuman and my Pompeian estates --- except that I am without you --- was for the rest agreeably amusing myself, and was going to be in the same places until the Kalends of June. I was writing those things which I had spoken of, the \textit{[Greek: politika]} (\textit{Politika}, the Republic) --- a thick and laborious work indeed; but if it succeeds to my satisfaction, the labour will be well spent; if not, into that very sea which we look on as we write we shall throw it, and shall set out on something else, since to keep quiet we cannot.

\ciceroSection{2} I shall pursue your commissions diligently, both in attaching certain men to me, and in not estranging certain others. My greatest care, indeed, will be that I see your son and mine, your Cicero --- daily, of course, but I will look in on what he is learning as often as possible; and, unless he disdains it, I shall even profess myself his teacher; in which I have got some practice in this leisure of these days, by bringing on the smaller Cicero of ours.

\ciceroSection{3} You will indeed do, as you write --- which, even if you did not write, I knew you to be doing most diligently --- that you sort, pursue, complete my commissions. I, when I have come to Rome, will not let pass any courier of Caesar's to whom I do not give a letter for you. In these days (forgive me) there has been no one to whom I might give one before this M. Orfius, a Roman knight, our connection both on his own account and because he is from the township of Atella, which you know is in our trust. I therefore commend him to you in the highest degree, a man splendid at home, popular even outside home; whom see that you bind to yourself by your generosity. He is a military tribune in your army. You will find him a grateful and observant man. That you should love Trebatius greatly, I urgently ask of you.
