% Ad Quintum Fratrem 3.4 (ad-quintum-fratrem-03-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} Gabinius has been acquitted. Nothing on the whole more incoherent than Lentulus as accuser and his supporting counsel, nothing more squalid than that panel. But still, but for an extraordinary effort, the prayers of Pompey, and the rumour of a dictatorship --- a thing full of fear --- not even to Lentulus, such as he was, would he have made any answer; though with that accuser and that panel he was condemned by votes to the number of thirty-two, with seventy on the other side. The trial has on the whole left such a heavy report behind it that he seems sure to perish in the remaining trials, and most of all in the one on extortion. But you see there is no commonwealth, no senate, no courts, no dignity in any of us. What more to say of the jurors? Two ex-praetors sat: Domitius Calvinus (who voted openly to acquit, so that everyone could see), and Cato (who, when the tablets had been sorted, slipped out of the gathering and was first to take the news to Pompey). Some say --- as Sallustius does --- that I ought to have prosecuted.

\ciceroSection{2} Was I to entrust myself to these jurors? What would I have been, if with me conducting the case he had slipped away? But other things moved me. Pompey would have judged that with me his contest was not about Gabinius's safety but about his own dignity; he would have entered the city; the matter would have come to enmity; I should have looked like Pacideianus matched against Aeserninus the Samnite; he might have bitten off an ear, perhaps; with Clodius at any rate he would certainly have come back into favour. As for myself, I strongly approve of my own counsel, especially since you do not disapprove. He had been honoured by me with my singular zeal; and though I owed him nothing and he owed me everything, still in public life he could not bear me dissenting from him (I will say no more), and at that time, when he was less powerful, he showed what he could do against me in my prime. Now, when I do not even take much trouble to be powerful, and the commonwealth certainly can do nothing, and he alone can do everything --- was I to wrestle with him in person? For that is what it would have come to. I do not think that you suppose I ought to have taken that on.

\ciceroSection{3} ``You should have defended one or the other,'' says the same Sallustius, ``and yielded the point to Pompey's pressure; for he begged hard.'' A charming friend Sallustius is, to think that I ought either to have taken on dangerous enmities or undergone everlasting disgrace! For my part I am delighted with this middle course; and I find it agreeable that, when I had given evidence very weightily on my faith and conscience, the defendant said that, if it had been possible for him to remain in the State, he would have given me satisfaction, and asked me no question at all.

\ciceroSection{4} As to the verses you want me to write for you --- I lack, indeed, the leisure, which requires not only time but also a mind empty of every care; but \textit{[Greek: enthousiasmos]} (inspiration) is also absent, for we are by no means free of anxiety about the coming year, even though we are without fear. And at the same time this too (and I speak, by Hercules, without any irony): in writings of that kind I award the first place to you rather than to myself.

\ciceroSection{5} On filling out your Greek library, exchanging books, and procuring the Latin ones, I should very much like those things to be done, especially since they look to my use as well. But I myself have no one through whom I can manage it; for the books worth having are not on sale, and they cannot be procured except through a man both expert and diligent. Still, I will charge Chrysippus and speak with Tyrannio. About what Scipio has done in the matter of the treasury I will inquire; what seems right I will see to. About Ascanio you will do what you like; I do not interpose in any way. That you are not hurrying about the suburban place I praise; that you should have one, I urge.

\ciceroSection{6} I have written this on the ninth day before the Kalends of November [24 October], on which day the games were starting, as I was setting out for Tusculum and taking my Cicero with me --- to a school for learning, not for playing. The reason I am not going further, as I should wish, is that I want to be present for Pomptinus's triumph on the third day before the Nones of November [3 November]. For there will be some little fuss; for Cato and Servilius as praetors threaten to forbid it, and I do not know what they can do (since Pomptinus will have with him the consul Appius and praetors and tribunes of the plebs); but they threaten all the same, and Q. Scaevola first of all, \textit{[Greek: Ar\=e pne\=on]} (breathing Ares). Take care, my sweetest and dearest brother, to keep well.
