% Ad Atticum 4.16 (ad-atticum-04-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Even the fact that this letter is in my secretary's hand will be a sign of how busy I am. I do not reproach you about the frequency of your letters; but most of them only told me where you were, or at most indicated that you were well, by the very fact that they came from you. Of that kind, the two that delighted me most were written from Buthrotum at almost the same time, since I wanted to know that you had had a smooth voyage. So this run of letters delighted me not so much by its fullness as by its frequency. The one substantial letter, full of matter, was the one that your guest-friend M. Paccius delivered to me. To that I shall reply, then; and this first of all. To Paccius, in calculation and in words and in deed, I have made plain what weight your commendation carries. Accordingly he is now among my close intimates, though previously he had not even been an acquaintance.

\ciceroSection{2} Now I will go on to the rest. Varro, about whom you write to me, shall be worked in at some place, if only there is a place for him. But you know the way my dialogues go. Just as in the books on the orator, which you praise to the skies, no one could be mentioned by the speakers except a man known to them or heard by them, so this discussion on the commonwealth which I have set on foot I have assigned to the persons of Africanus and Philus and Laelius and Manilius. I have added some younger men, Q. Tubero, P. Rutilius, the two sons-in-law of Laelius, Scaevola and Fannius. So I have been thinking, since in the separate books I am using proems in the way Aristotle does in the works he calls exoteric \textit{[Greek: ex\=oterikous]}, to contrive some way of bringing him in not without reason --- which I gather is what you want. If only I can carry out what I have undertaken! For, as you are well aware, I have taken on a subject great and weighty and demanding the fullest leisure, and leisure is exactly what I am most in want of.

\ciceroSection{3} As for the fact that in those books you praise you miss the character of Scaevola, I did not move him out of the way casually --- I did what our divine Plato did in his \textit{Republic} \textit{[Greek: Politeia(i)]}. When Socrates had come down to Cephalus, the wealthy and genial old man, in the Piraeus, the old gentleman is present in the conversation only while that first exchange is being held; then, after he too has spoken very gracefully, he says he wants to leave to attend to a sacred rite, and never afterward returns. I suppose Plato thought it would scarcely be in keeping if he had kept a man of that age in so long a conversation any further. Much more, I judged, was the same caution called for in Scaevola, who was of the age and the state of health you remember him to have been, and held offices such that it would hardly have seemed proper for him to be at Crassus's Tusculan villa for many days running. And the conversation of the first book was not foreign to Scaevola's pursuits; the remaining books carry technical exposition \textit{[Greek: technologian]}, as you know. For these I really did not want that jocular old gentleman, as you knew him, to sit in.

\ciceroSection{4} As for the matter of Pilia which you write about, I will see to it. The case is in fact a splendid one, on Aurelianus's evidence, as you write. And in handling it I shall recommend myself to my dear Tullia as well. Vestorius I am not failing. For I see that this is welcome to you, and I take care that he sees it too. But you know how he is? When he has two easy options, nothing is more difficult.

\ciceroSection{5} Now for what you ask about C. Cato. Under the Lex Junia and Licinia (a law on legislative procedure) he has been acquitted, as you know; under the Lex Fufia I forecast to you he will be acquitted, and not with his counsel any more willing than his prosecutors. He has, however, made up with me and with Milo. Drusus has been brought to trial by Lucretius. The day for striking the panel of jurors has been set for the fifth day before the Nones of Quintilis. About Procilius the rumor is bad; but you know what verdicts are like. Hirrus is now in favor with Domitius. The senatorial decree which these consuls have passed about the provinces --- ``whosoever hereafter\,\dots'' --- does not seem to me likely to hold.

\ciceroSection{6} As for what you ask about Messalla --- I do not know what to write. Never have I seen candidates so evenly matched. Messalla's resources you know. Scaurus has been brought to trial by Triarius. If you ask me, no great fellow-feeling \textit{[Greek: sympatheia]} has been stirred up; still, his aedileship is remembered not unfavorably, and his father's memory carries weight with the country voters. The remaining two plebeian candidates are so evenly balanced that Domitius is strong in friends but helped not a little by his thoroughly popular gift to the people, while Memmius is recommended by Caesar's soldiers and relies on Pompey's Gaul. If these supports fail him, they think there will be some man to push the elections off into Caesar's return --- especially with Cato acquitted.

\ciceroSection{7} I have answered the Paccian letter. Now suffer yourself to be scolded, if I am within my rights. For in the letter which C. Decimius brought me, written from Buthrotum, you write that you think you must go on into Asia. By Hercules, it seemed to me that there was nothing in it of such small consequence as whether you handled the business through agents or in person, that you should be away from your own people so often and so far. But I would have preferred to plead this with you while the matter was still open, for then I should certainly have got something done. Now I will hold back the scolding I had taken up. May it prove of use, at least, in hastening your return! I write to you the less often for this reason --- that I do not know for certain where you are or where you will be; this letter, however, I thought I had to entrust to the bearer, whoever he is, since it looked as though he might see you. Since you now think you must go on to Asia, please let me know at what time we may expect you, and what you have done about Eutychides.
