% Ad Quintum Fratrem 2.14 (ad-quintum-fratrem-02-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SAL VTEM.}

\ciceroSection{1} The thing shall be done with a sharpened pen, well-tempered ink, and even ridge-smoothed paper. For you write that you could hardly read my last letter --- in which, my dear brother, there was none of what you suppose. I was not busy, not flustered, not angry with anyone; I simply have this habit, of using whatever pen has come into my hands as if it were a good one.

\ciceroSection{2} But now turn your attention, my best and sweetest brother, while I write back to what you set down in this same brief letter so very \textit{[Greek: pragmatikōs]} --- on matters of business. On the point on which you ask me to write back to you concealing nothing, dissembling nothing, indulging you in nothing, but plainly and in a brother's spirit --- namely, whether you should †\dots† to extricate yourself, if there is reason, stay where you are. If, my Quinte, it were some small matter on which you were asking my mind, even though I would have left you free to do as you chose, I would still have made my own wish plain. But in this matter what you really ask is what sort of year I expect the coming one to be. To my eye plainly tranquil for us, or at any rate most thoroughly fortified --- as the crowds about the house, the forum, the demonstrations at the theatre declare daily; nor do I fear, given my own awareness of our resources, and of the goodwill of Caesar and of Pompey that we hold. These are the things that make me confident. But if some deranged man's frenzy breaks out, all is in readiness to break him.

\ciceroSection{3} This is how I feel, how I judge, how I write to you with my eyes open; I forbid you to doubt --- not as a flatterer, but as a brother. Therefore, for the sake of enjoying our shared sweetness together, I should wish you to come at the time you had named --- but I prefer the other course, what you think more to your purpose. For those things too I value greatly --- that abundance of yours, \textit{[Greek: amphilaphian]}, and the clearing of your debts. Take this for certain: nothing for us, once we are clear, will be more fortunate, if we keep our health. The things that fall short, by our own standards anyway, are small, and most readily settled --- provided we keep our health. Bribery is back, monstrous; there has never been its equal.

\ciceroSection{4} On the Ides of July the interest rate jumped from a third to two-thirds per cent, on the syndicate of Memmius and the consuls with Domitius; Scaurus alone †outbids them†. Messala droops. I do not exaggerate --- they are talking \textit{[Greek: hyperbolas]}, even up to ten million sesterces, to be announced at the praerogativa. The thing is ablaze with public anger. The tribunician candidates have entered into a compact, depositing 500,000 sesterces apiece with Marcus Cato, to canvass on his terms --- so that whoever transgressed should be condemned by him. If indeed these elections turn out free of bribery, as people suppose, Cato alone will have done more than all the laws and all the courts.
