% Ad Quintum Fratrem 3.2 (ad-quintum-fratrem-03-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} On the sixth day before the Ides of October [10 October], Salvius set out from Ostia in the evening by ship with those things which you had wanted sent to you from home. On the same day, before the people, Memmius gave Gabinius a brilliant roasting, such that Calidius was not allowed to say a word for him. On the day after that --- which was then about to come; for I was writing this before light --- there was to be a \textit{divinatio} against Gabinius before Cato's tribunal, between Memmius and Ti. Nero and C. and L. Antonius (sons of Marcus). We thought it would go to Memmius, though Nero was striving extraordinarily. What more? --- the man is being soundly squeezed, unless our Pompey, with gods and men unwilling, upsets the business.

\ciceroSection{2} Hear now of the man's audacity, and take some pleasure in something in this ruined commonwealth. When Gabinius had been saying wherever he came that he was going to demand a triumph, and then suddenly the good imperator stole into the city by night as into a city full of enemies, he would not entrust himself to the Senate. In the meantime on the very tenth day, on which he had to report the number both of the enemy and of the soldiers, he crept in with the lowest attendance. When he wanted to leave, he was held back by the consuls; the publicans were brought in. The man, wounded on every side, when he was being most of all wounded by me, could not bear it, and with a trembling voice called me an exile. Here --- O gods! nothing more honourable has ever happened to us! --- the Senate rose up with a shout to a man, in such a way that it was as if it would go up against his very body; with equal shouting and rush, the publicans. What more? Everyone behaved as if it had been you. Nothing more famous in men's talk out of doors. I, nevertheless, hold myself back from prosecuting --- scarcely, by Hercules, but I do hold myself back, either because I do not wish to fight with Pompey (the Milo trouble pressing on me is enough), or because we have no jurors. \textit{[Greek: apoteugma]} (failure) I dread; I add also men's malevolence, and I fear that with me prosecuting something might happen to him; nor do I despair that the matter can be carried through both without me, and not without me having something to do with it.

\ciceroSection{3} On the charge of canvassing, all who are standing for the consulship have been arraigned: Domitius by Memmius, Memmius by Q. Acutius (a fine and learned young man), Messala by Q. Pompeius, Scaurus by Tnarius. The matter is in great commotion, since either the destruction of the men or the destruction of the laws is set forth. They are working to ensure that the courts do not sit. The matter seems to look to an interregnum. The consuls long to hold the elections; the candidates do not wish it, especially Memmius, who hopes that, with Caesar's coming, he will be consul --- but he stands in a wonderful way prostrate. Domitius with Messala seemed certain; Scaurus had cooled off. Appius, without a curiate law, declares that he will succeed our Lentulus; and Lentulus indeed --- which I almost passed over --- was wonderful that day against Gabinius: he prosecuted him for \textit{maiestas}; the names entered, while Gabinius said not a word. You have the news of the Forum. At home all is well; the house itself the contractors are handling not undiligently.
