% Ad Quintum Fratrem 3.5 (ad-quintum-fratrem-03-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} As to your asking what I have done about those books which I set about writing when I was at the Cumanum --- I have not been idle, nor am I idle, but I have already several times changed the whole design and plan of the writing. For two books had been finished, in which there was a conversation set by me, in the nine-day holidays held under the consulship of Tuditanus and Aquilius [129 BC], between Africanus shortly before his death and Laelius, Philus, Manilius, P. Rutilius, Q. Tubero, and Laelius's sons-in-law Fannius and Scaevola; and the discourse was distributed across nine days and nine books, on the best constitution of the State and the best citizen (the work was being woven up rather brilliantly, indeed, and the standing of the men was bringing no small weight to the speech); but when those books were being read out to me at Tusculum with Sallustius listening, I was warned by him that those subjects could be spoken on with much greater authority if I myself were to speak on the commonwealth, especially since I was not a Heraclides of Pontus but a consular and one who had been involved in the greatest affairs of state; what I was assigning to such ancient men would look fabricated; that in my books on speaking the conversation of orators I had removed gracefully from myself, but had referred it to those whom I myself had seen; and that, finally, what Aristotle writes on the State and the man of excellence, he speaks himself.

\ciceroSection{2} He moved me, and the more so because I could not touch on the greatest movements of our State, since they were later than the lifetimes of those who were speaking. I, however, had been pursuing exactly this in the first place, so as not, by running into our own times, to give offence to anybody. Now I will both avoid that and speak myself with you, and yet I will send to you, if I come to Rome, those books which I had begun. For I think you will conclude that I have left those books behind not without some vexation of my own.

\ciceroSection{3} In the love of Caesar which he has written down to me at length I take a singular delight; in the promises which he holds out I do not greatly hang. I do not thirst for honours nor do I yearn for glory, and I look more for the lastingness of his goodwill than for the issue of his promises; yet I live in that round of canvassing and labour, as if I were expecting what I am not demanding.

\ciceroSection{4} As to your asking me to make verses --- it is incredible, my brother, how I lack the time, and I am scarcely sufficiently stirred in spirit for the matters you want sung. [corrupt] do you really, who in that whole genre of eloquence and expression have surpassed all men, ask me for what I do not attain to even in my thinking? I would do it all the same, if I could; but, as you do not at all forget, for poetry there is need of a certain alacrity of mind, which our times plainly snatch from me. I am withdrawing myself from every care of the State and giving myself to literature; but still, I will reveal to you a thing which, by Hercules, I had wanted above all to keep hidden from you. I am tormented, my sweetest brother, tormented at there being no commonwealth, no courts, and at this period of life of ours which ought to be flourishing in the senatorial authority being either tossed about in the labour of the courts or sustained by literature at home; while that which I had loved from boyhood --- \textit{[Greek: Pollon aristeuein kai hypeirochon emmenai all\=on]} ``always to be best and pre-eminent above the others'' --- has totally gone down; while my enemies have in part not been attacked by me, and in part actually been defended; while not only my spirit, but not even my hatred, is free; and one out of all has been found, Caesar, to love me as much as I should wish --- or, as others think, this is the one man who would wish to. Yet of these things nothing is of such a sort that I cannot daily soothe myself with many a consolation; but that will be the greatest consolation, if we are together. Now to those other things even the longing for you is added, a most weighty consolation in its absence.

\ciceroSection{5} Had I defended Gabinius, as Pansa thought I should have done, I should have been finished. Those who hate him (and that means the whole orders, all of them) would have begun to hate me on account of the man on whose account they hate him. I have held myself in, as I think, splendidly --- doing only that much which everyone could see; and in the whole sum of things, as you urge, I am turning myself strongly towards leisure and peace. About the books, Tyrannio is being dilatory.

\ciceroSection{6} I will instruct Chrysippus; but the business is laborious and requires a most diligent man. I myself perceive it, who, with the highest enthusiasm, get nothing accomplished. As to the Latin books, indeed, I do not know which way to turn; so faulty they are both in the copying and in the selling. Still, whatever can be done, I will not neglect. [corrupt] as I wrote you before, is at Rome, and those who [corrupt] declare [corrupt] that they have a strong account to render to you. From the treasury, I think, it has been settled in my absence.

\ciceroSection{7} When you write that you have finished four tragedies in sixteen days, do you borrow anything from someone else? And do you ask after \textit{[Greek: pathos]} (feeling) when you have written an Electra and a Troilus? Be no idler, and do not think that ``\textit{[Greek: gn\=othi seauton]}'' (know thyself) has been said only towards lessening arrogance, but rather so that we may know our own good things too. But I should like you to send me both those tragedies and the Erigona. You have my reply to your last two letters.
