% Ad Atticum 4.19 (ad-atticum-04-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} How I had longed for your letter! How welcome its arrival! What constancy of promise, what remarkable good faith! What a lovable voyage! --- one which, by Hercules, I had dreaded keenly, remembering the leather screens of your earlier crossing \textit{[Greek: derreis]}. But unless I am mistaken I shall see you sooner than you write. For I believe you supposed your womenfolk to be in Apulia. Since that is otherwise, what is Apulia to detain you? Are days really to be granted to Vestorius, and that Latin Atticism \textit{[Greek: Attikismos]} to be tasted again after so long an interval? Come, fly to us here and look in on the state of that commonwealth of ours --- \textit{$\dagger$Germane putavi de nummis$\dagger$} --- with the cash openly distributed before the elections tribe by tribe in one place, Gabinius thereupon acquitted, the dictatorship --- \textit{$\dagger$fruere iustitio et omnium rerum licentia$\dagger$}.

\ciceroSection{2} Take note of my evenness of mind, and \textit{$\dagger$ludum et contemptionem Seleucianae provinciae$\dagger$}, and, by Hercules, my most agreeable connection with Caesar --- for this is the one plank from the shipwreck that gives me pleasure; he indeed, on my Quintus and yours, good gods! how he handles him --- with honour, with rank, with favour! No differently than if I were the commander myself. The choice of winter quarters for the legion has just been handed to him, as Quintus writes to me. Are you not to love such a man? Whom then, of all those others? But listen --- had I written to you that I have been made a legate to Pompey, and shall in fact be outside the city from the Ides of January? This struck me as fitting to many considerations. But why more? The rest, I think, face to face, so that you have something still to look forward to. Give Dionysius my warmest greeting; for him I have not only kept but actually built up a position. In short: to the great happiness I take from your return, his coming will add a great heap on top. The day you come to me, if you love me, stay with me along with your people.
