% Ad Quintum Fratrem 2.15 (ad-quintum-fratrem-02-15) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} When you have received a letter from me in a secretary's hand, judge that I have not had even a little leisure; when in my own, a little. For know this: I have never been more pulled in pieces by cases and trials, and that at the most oppressive season of the year and in the greatest heats. But since you so prescribe, these things must be borne, and we must not let it appear that I have failed either the hope or the design that the two of you have formed --- especially since, if it proves the more difficult, I shall all the same gather great gratitude and great standing out of this labour. And so, as you would have it, I take pains to offend no one's feeling, and even to be cherished by those very men who grieve that we are so closely joined with Caesar, while by those who are fair --- or even leaning to our side --- to be both vigorously courted and loved.

\ciceroSection{2} When the matter of bribery was being pressed most savagely in the senate for many days --- because the consular candidates had gone so far that it could not be tolerated --- I was not in the senate; I have determined to bring no remedy at all to the commonwealth without strong backing.

\ciceroSection{3} On the day I write this, Drusus was acquitted of collusion by the tribuni aerarii by a margin in all of four votes, although the senators and the knights had condemned him. I am to defend Vatinius the same day in the afternoon. That business is easy. The elections have been put off to the month of September. Scaurus's trial will be set on at once; and at it we shall not be wanting. As for ``The Banqueters of Sophocles'' \textit{[Greek: Sundeipnous Sophokleous]} --- though I can see the little play is by your hand and done in fine style --- I gave it no approval at all.

\ciceroSection{4} I come now to what perhaps ought to have stood first. O how welcome to me was your letter about Britain! I had been fearing the Ocean, fearing the island's shore; the rest I do not at all underrate, but it has more of hope in it than of fear, and I am more anxious in the expectation than in the dread. As for you, I can see you have a magnificent subject \textit{[Greek: hypothesin]} to write of. What landscapes, what natures of things and places, what manners, what peoples, what battles --- and what a commander you have! As you ask, I will gladly help you with whatever materials you wish, and I will send you the verses you ask for --- that is, ``an owl to Athens.''

\ciceroSection{5} But here now! it seems to me you are keeping something from me. Tell me, my brother, what does Caesar say about our verses? For the first book, he wrote to me earlier that he had read it, and the opening parts in such terms that he denies he has read better even in the Greek; the rest, down to a certain place, rather slack \textit{[Greek: rhathumotera]} --- that is the word he uses. Tell me the truth: does either the subject or the style \textit{[Greek: charaktēr]} not please him? You have nothing to fear; for I shall not love myself one hair the less. On this matter, write in love of truth \textit{[Greek: philalēthōs]} --- and, as you usually do, as a brother.
