% Ad Quintum Fratrem 2.10 (ad-quintum-fratrem-02-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} I am glad my letters please you; though even so I should have had no subject for writing now if I had not received yours. For on the day before the Ides, when Appius had convened a thin Senate, the cold was so bitter that, driven by the shouting of the people, he was forced to dismiss us.

\ciceroSection{2} On the Commagenian business --- since I had pulled the whole thing to pieces --- Appius makes wonderful court to me, both in person and through Pomponius; for he sees that if I use this manner of speaking on the rest, February will be barren. Him I rallied playfully enough, and not only wrung from him that little town set at the Zeugma of the Euphrates, but went on to make fun, to the great laughter of those present, of the praetexta he had got in Caesar's consulship.

\ciceroSection{3} ``As to his wanting the same honours renewed,'' said I, ``so that he need not patch up his praetexta every year --- I propose that nothing be decreed; but will you noble gentlemen, who could not endure a Bostrene in the praetexta, endure a Commagenian in it?'' You see the kind of joke and the place for it. I said many things against this obscure little king, by which he was hooted off altogether. Stirred by this style, as I said, Appius is all over me; for nothing is easier than to break up the rest. But I shall not offend him, for fear he invoke the faith of \textit{Jove the Host, and summon all the Greeks} --- through whom he came back into favour with me.

\ciceroSection{4} Theopompus we shall satisfy. About Caesar --- I had forgotten to write to you; for I see the sort of letter you have been waiting for. But he has written to Balbus that the bundle of letters in which mine and Balbus's were enclosed reached him completely sodden, so that he does not even know there was a letter of mine in it. Out of Balbus's letter, though, he had made out a few words, to which he replied in these terms: ``I see you have written something about Cicero which I did not catch; but as far as I could make out by conjecture, it was such that I thought it more to be wished than hoped for.''

\ciceroSection{5} So afterwards I sent Caesar a letter on the same model. Don't take his joke about his own poverty lightly. To him I wrote back that hereafter he had no reason on the strength of our strong-box to throw his accounts into confusion --- and I played upon the theme both intimately and with dignity. His affection for us is reported, by every messenger, as remarkable. The letter on what you are waiting for will follow closely on your return. The day-by-day remainder I shall write you --- provided you supply the couriers. Though such a frost was hanging over us that there was the gravest danger of Appius's having his own house burned down about his ears.
