% Ad Familiares 7.5 (ad-familiares-07-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO CAESARI IMP. S.D.}

\ciceroSection{1} See how I have persuaded myself that you are a second self to me, not only in matters that concern me but also in those that concern my own people. C. Trebatius I had thought of taking with me, wherever I went, that I might bring him home as adorned as possible by all my efforts and kindnesses; but, since both Pompey's stay was longer than I had supposed, and a certain hesitation of mine, not unknown to you, seemed either to hinder my setting out or at least to delay it (see what I have presumed!), I began to wish that Trebatius should expect from you the things he had hoped from me; nor, by Hercules, have I made him any less generous promises about your goodwill than I had been wont to do about my own.

\ciceroSection{2} But indeed a wonderful chance has stepped in, as if a witness of my opinion or a guarantor of your humanity. For while I was speaking with our Balbus at my house about this very Trebatius, more particularly, a letter is delivered to me from you, at the end of which is written: ``M. Itfius, whom you commend to me, I will make even king of Gaul; or, if you wish, hand him over to Lepta. You send me another whom I may dignify.'' We raised our hands, both I and Balbus. So great was the timeliness that the thing seemed I-know-not-what, not chance, but divine. So I send Trebatius to you, and I send him in such a way that I judge him at first to have been sent on my own initiative, and then by your invitation.

\ciceroSection{3} This man, my Caesar, I would have you embrace with all your courtesy, so that all that you can be brought, by my doing, to bestow on my own people, you may bestow on this one. About whom I pledge to you these things, not by that old phrase of mine which, when I had written to you about Milo, you justly made fun of, but in the Roman fashion, in the way men not silly speak: that no one is more upright, no one a better man, no one more modest. To which is added that he leads the front rank in civil law, with a singular memory and the highest knowledge. For him I am asking neither tribunate nor prefecture, nor the certain title of any benefit; I am asking your goodwill and generosity; nor do I block your gracing him, if it so please you, with these little gloriettes too. In short, I hand the whole man over to you, ``from hand to hand,'' as they say, into yours --- a hand outstanding both in victory and in faith. Let us be a little fulsome --- though through you that is scarcely allowed; but, as I see, it will be allowed. See that you keep well; and as you love me, love me.
