% Ad Quintum Fratrem 3.8 (ad-quintum-fratrem-03-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QUINTO FRATRI SALUTEM}

\ciceroSection{1} To your earlier letter there is nothing for me to answer, which was full of vexation and complaints, in which kind you write that you had given another also to Labienus the day before, who has not yet come. The more recent letter has done away with all my distress. So much only do I both admonish and ask of you, that in those vexations and labours and longings you keep recalling the design we had in your setting out. For we were not pursuing certain advantages, small and middling. For what was there which we should have thought worth buying at the price of our parting? The most secure protection we were seeking, out of the goodwill of the best and most powerful man, towards the whole standing of our dignity. More is placed in hope than we are demanding; the rest, let it be reserved as loss to be borne. So, if you frequently bring your mind back to the reckoning both of our old design and of our hope, you will more easily bear those labours of the campaign and the other things that vex you, and you will lay them down all the same when you wish. But the ripeness of that has not yet come, and yet now it is drawing near.

\ciceroSection{2} I also give you this warning: do not entrust to any letter a thing such that, if it were brought out, we should be sorry. There are many things which I would rather not know than be made certain of with some risk. I will write to you more, with a free mind, when, as I hope, my Cicero is going on well. I should like you to see to it that I know to whom we ought to give the letters we shall next send you --- to Caesar's couriers, that he may send them on to you directly, or to Labienus's. For where these Nervii are and how far off they are I do not know.

\ciceroSection{3} On the courage and gravity of Caesar, which he had shown in his deepest grief, from your letter I have received great pleasure. As to your bidding me finish the poem I had begun for him, although I am distracted both by work and, much more, in mind, still, since Caesar has gathered from the letter I had sent you that I had taken something in hand, I will return to my undertaking and finish it in these leisure days of the thanksgivings. I am vehemently delighted that our Messala now and the rest have been relieved of their vexation, and in your numbering him a certain consul along with Domitius you are not at variance from our own opinion. I will deliver Messala over to Caesar. But Memmius's hope is in Caesar's arrival; in which, I think, he is mistaken. He is cold here; Scaurus, on the other hand, Pompey has long since thrown over.

\ciceroSection{4} Things postponed; the elections brought to an interregnum. The rumour of a dictatorship is unpleasant to the good men, and even more so to me are the things they are saying. But the whole matter is both feared and is cooling down. Pompey openly denies that he wants it; before, to me himself, he was not denying it. Hirrus, it seems, will be the proposer --- gods, what an oaf, what a self-lover ``without a rival.'' Crassus Junianus, a man devoted to me, he has put off through me; whether he wants it or not is hard to know; but, with Hirrus as proposer, he will not convince anybody that he does not want it. Of other public business they were saying nothing at that time; nothing in any case was being transacted.

\ciceroSection{5} The funeral of Serranus's domestic son was a most mournful affair on the eighth day before the Kalends of December [24 November]. The father gave the eulogy, on a script written by me.

\ciceroSection{6} Now about Milo: Pompey gives him nothing and gives everything to Cotta, and says that he will bring it about that Caesar comes down on the man. This Milo dreads, not without reason; and, if Pompey shall be made dictator, he almost gives himself up for lost. The vetoer of the dictatorship, if he shall help him by his hand and his bodyguard, he fears Pompey for an enemy; if he shall not help him, he fears that it may be carried by force. He is preparing games of the most magnificent kind, such, I say, that nobody has ever given more sumptuous ones --- a stupid course twice and three times over and not called for: either because he had given a magnificent show, or because the means were not there, or because as administrator (and not as aedile) he could have considered himself such. I have written almost everything. Take care, my dearest brother, to keep well.
