% Ad Familiares 13.73 (ad-familiares-13-73) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO Q. PHILIPPO PROCOS. S.}

\ciceroSection{1} I congratulate you on having brought yourself home safe from your province to your own people, with your reputation and the commonwealth alike intact. Had I been at Rome, I should have seen you in person and thanked you face to face for the care you took of L. Egnatius, my very close friend, in his absence, and of L. Oppius on the spot.

\ciceroSection{2} With Antipater of Derbe there subsists, on my side, not only the tie of guest-friendship but the closest familiarity. I have heard that you were violently angry with him, and I was sorry to hear it. About the matter itself I can pass no judgement, except this much: I am satisfied that you, being the man you are, have done nothing rashly. On the strength, however, of our old connection, I ask you again and again to grant me his sons, who are in your power --- to me before anyone else, unless you think your good name would be touched at all by such a thing. If I thought it would, I should never ask you, and your reputation would weigh with me far more heavily than that old connection does; but my conviction is (and I may be mistaken) that the matter would tell to your credit rather than against you. As to what can be done, and what you can do for my sake --- for that you wish to, I do not doubt --- I should be grateful, if it is not a burden, to be told.
