% Ad Quintum Fratrem 2.2 (ad-quintum-fratrem-02-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} It is not occupation, by which I was indeed quite hindered, but a small inflammation of the eyes that has led me to dictate this letter and not, as I am used to with you, to write it myself. And first I excuse myself to you in the very thing in which I accuse you. For no one yet has asked me whether I want anything sent to Sardinia, while you, I think, often have people asking whether you want anything for Rome. As to what you wrote me about Lentulus and Sestius, I have spoken with Cincius. However the matter stands, it is not the easiest, but Sardinia has surely something fitted to the recollection of past memory. For just as that famous Gracchus the augur, after he came into that province, recalled what had befallen him at the Campus Martius, holding the consular elections against the auspices, so you seem to me at leisure to have re-thought, in Sardinia, of the design from Numisius and of the debts owed to Pomponius. I have so far bought nothing. Culleo's auction has taken place. For the Tusculan villa there was no buyer. If the terms turn out very good, perhaps I will not let it go.

\ciceroSection{2} About your building I do not stop pressing Cyrus. I hope he will do his duty. But everything is the slower because of waiting on this raving aedileship --- for the elections seem to be coming on without delay; they have been proclaimed for the eleventh day before the Kalends of February. Still, I do not want you to be anxious; every kind of precaution will be applied by us.

\ciceroSection{3} About the Alexandrian king a senatus consultum has been passed: that to bring him back with a multitude appears dangerous to the commonwealth. As to the rest, when the contention in the Senate was over whether Lentulus or Pompey should bring him back, Lentulus seemed to be holding the case (in that matter we did wonderful satisfaction to our duty towards Lentulus and clearly enough to Pompey's wishes); but through Lentulus's detractors the matter was drawn out by calumny. Then came the comitial days, on which the Senate cannot be held. What will come of the tribunes' banditry I do not divine; still, I suspect that Caninius will carry his bill through by violence. In this matter what Pompey wishes I do not see clearly; what his intimates desire all see. The king's creditors openly supply money against Lentulus. Without doubt the matter seems to be carried away from Lentulus, with great pain to me; although he has done many things by which, if it were lawful, we could rightly be angry with him.

\ciceroSection{4} You, if it suits, please at the earliest moment, with good and steady weather, embark and come to me. For there are countless matters in which every day I miss you in every kind. Your people and ours are well. The fourteenth day before the Kalends of February.
