% Ad Quintum Fratrem 2.3 (ad-quintum-fratrem-02-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} I wrote to you before about earlier things; now learn what has been done since. From the Kalends of February the embassies were being put off to the Ides of February. On that day the matter was not finished. On the fourth day before the Nones of February Milo appeared. To support him Pompey came; Marcellus spoke at my request; we came off honourably; the day was set forward to the eighth day before the Ides of February. Meanwhile, the embassies put off to the Ides, the question was being moved about the quaestors' provinces and about furnishing the praetors; but with many complaints about the state of the commonwealth being interposed, no business was got through. C. Cato promulgated a law to abrogate Lentulus's command. The young son changed his dress.

\ciceroSection{2} On the eighth day before the Ides of February Milo appeared. Pompey spoke --- or rather wished to. For when he rose, the Clodian gangs raised a shout, and it befell him through his whole speech to be hindered not only by yelling but by railing and abuse. When he had finished --- for in this he was indeed brave; he was not put off; he said everything, and at intervals even drew silence and \dag carried the day with weight\dag --- but when he had finished, Clodius rose. There came on him such a shout from our side (for it had been agreed to repay the favour) that he could keep neither mind nor tongue nor face steady. The thing was carried on, although Pompey had scarcely finished by the sixth hour, all the way to the eighth, with every form of railing, finally lewd verses, said against Clodius and Clodia. He, frantic and bloodless, kept asking his own people in the very midst of the shouting, who it was who was killing the people with hunger. The gangs answered: ``Pompey.'' Who wished to go to Alexandria? They answered: ``Pompey.'' Whom did they want to go? They answered: ``Crassus.'' He was there at the time, with no friendly mind towards Milo. About the ninth hour, as if at a signal given, the Clodians began to spit on our men. Anger flared. They pressed to drive us from our ground; an attack was made by our side; flight of the gangs; Clodius thrown down from the Rostra; and we then fled too, lest there be some mishap in the crowd. The Senate was summoned to the Curia; Pompey home. I, however, did not go to the Senate, lest I either keep silent in such matters or, in defending Pompey (for he was being torn at by Bibulus, Curio, Favonius, the younger Servilius), give offence to the spirits of the good men. The matter was put off to the next day. Clodius set the trial day forward to the Quirinalia.

\ciceroSection{3} On the seventh day before the Ides of February the Senate was at the temple of Apollo, that Pompey might be present. The matter was handled with weight by Pompey. On that day nothing was finished. On the sixth day before the Ides of February, at the temple of Apollo, a senatus consultum was passed: that the things which had been done on the eighth day before the Ides of February had been done against the commonwealth. On that day Cato vehemently inveighed against Pompey, and accused him with a continuous speech as if he were a defendant. About me he said much, against my wish, with the highest praise of me; when he was reproaching that man's treachery against me, he was heard in great silence by the malevolent. Pompey replied vehemently against him, and traced the figure of Crassus, and said openly that he would be more fortified to guard his own life than Africanus had been, whom C. Carbo had put to death.

\ciceroSection{4} And so to me great matters seemed already to be in motion. For Pompey understands these things and shares them with us: that plots are being made against his life; that C. Cato is being kept up by Crassus; that money is being supplied to Clodius; that both are being confirmed both by him and by Curio, Bibulus, and the rest of his detractors; that he must take great precaution lest he be crushed --- with that contio-going people now nearly alienated from him, the nobility hostile, the Senate not even-handed, the youth unprincipled. And so he is making ready, summoning men from the country districts. Clodius, on his side, is strengthening his gangs; a band is being prepared for the Quirinalia. In that we are by far superior to his own forces; but a great band is being expected from Picenum and Gallia, that we may also resist Cato's bills against Milo and Lentulus.

\ciceroSection{5} On the fourth day before the Ides of February Sestius was indicted for canvassing by an informer, Cn. Nerius of the Pupinian tribe, and on the same day by a certain P. Tullius for violence. He was ill. We came at once, as we ought, to him at his house, and gave ourselves over wholly to him; and we did this against the opinion of men who thought we were rightly angry with him --- so that we should appear in the highest degree humane and grateful both to him and to all. So we shall do. But that same Nerius, the informer, gave the names of Cn. Lentulus Vatia and C. Cornelius as accessory. The same day a senatus consultum was passed that the political clubs and the gangs organized by tribes (\textit{decuriati}) should disperse, and that a law should be brought concerning them, that those who did not disperse should be held by the same penalty as for violence.

\ciceroSection{6} On the third day before the Ides of February I spoke for Bestia on a charge of canvassing before the praetor Cn. Domitius in the middle of the Forum, with the largest gathering. And I lit, in the speaking, on the place where Sestius, having taken many wounds in the temple of Castor, had been saved by coming to Bestia's aid. Here I prepared the way for something \textit{[Greek: pro\=oikonom\=esam\=en quiddam eukair\=os]}, opportunely, regarding those charges that were being got up against Sestius, and adorned him with true praises to the great approval of all. The thing was deeply welcome to the man. This I write to you because in your letters you have often warned me to keep Sestius's gratitude.

\ciceroSection{7} The day before the Ides of February I have written this before dawn. On that day I was going to dine at Pomponius's, at his wedding. The rest in our affairs are of this kind --- as you yourself, when I was nearly losing trust, used to predict to me --- full of standing and good will; both of which, indeed, my brother, by your wisdom, patience, virtue, devotion, gentleness too, have been restored both to you and to me. For you the Liciniana house by the Lacus Pisonis has been hired; but in a few months, I hope, after the Kalends of July, you will move into your own. The Lamiae, distinguished tenants, have rented yours in the Carinae. From you, after that letter from Olbia, I have received no letter. I want to know what you are doing and how you are amusing yourself, and most of all to see you yourself as soon as may be. See, my brother, that you keep well, and although it is winter, still bear in mind that this is Sardinia. The fifteenth day before the Kalends of March.
