% Ad Familiares 1.2 (ad-familiares-01-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S.D. P. LENTVLO PROCOS.}

\ciceroSection{1} On the Ides of January nothing was finished in the Senate, because the day was largely consumed by the altercation between Lentulus the consul and Caninius the tribune. On that day I too made many words, and we seemed especially to move the Senate by reminding it of your kind disposition towards that order. So on the next day it was agreed that we should give our opinions briefly --- for the Senate's good will to us seemed restored, as I had perceived both from speaking and from calling on and questioning them one by one. And so, when the first opinion of Bibulus had been put forward, that three legates should bring back the king; the second of Hortensius, that you should bring him back without an army; the third of Volcacius, that Pompey should bring him back --- it was demanded that Bibulus's opinion be divided. So far as it spoke of religion --- a thing which by now could no longer be resisted --- Bibulus was assented to; on the three legates the full house went over to the other side.

\ciceroSection{2} Next was Hortensius's opinion, when Lupus the tribune --- because he himself had moved the question of Pompey --- began to insist that the division ought to be made before him rather than before the consuls. Against his speech there was vehement outcry from all sides, for it was both unfair and unheard of. The consuls neither yielded nor strongly opposed: they wanted the day used up, which is what happened; for they saw clearly that many more would go for Hortensius's opinion, although they were openly assenting to Volcacius. Many were called on, and that very thing was not against the consuls' wishes; for they wanted Bibulus's opinion to prevail.

\ciceroSection{3} The dispute drawn out till nightfall, the Senate was dismissed. I that day chanced to dine at Pompey's, and finding a moment more suitable than ever before --- because after your departure that day had been the most honourable to us in the Senate --- I so spoke with him that I seemed to be drawing the man's mind from every other thought to the protecting of your standing. When I hear him alone, I plainly clear him of all suspicion of grasping; but when I see his intimates of all the orders, I see clearly --- and this is by now open to everyone --- that this whole business has long been corrupted by certain men, the king himself and his counsellors not unwilling.

\ciceroSection{4} This I have written on the sixteenth day before the Kalends of February before dawn. That day there was to be a Senate. We in the Senate, as I hope, shall keep our standing, so far as one can amid such treachery and unfairness of men. As to the popular dimension, this we seem to have brought off: that no business can be done with the people while the auspices stand, while the laws stand, or simply without violence. On these matters, on the day before I wrote this, a most weighty resolution of the Senate was passed; and although Cato and Caninius interposed, it was nevertheless put on record. I think it has been sent to you. About the other things, whatever is done, I will write to you, and I will see to it that everything is done as straightforwardly as possible by my care, exertion, diligence, and influence.
