% Ad Atticum 4.8A (ad-atticum-04-08a) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Apenas had hardly gone, when your letter came. What are you saying? Do you think he will not pass the law? Speak, I beg you, more clearly; for I scarcely seem to have heard. But let me know at once, if only it is convenient for you. Since a day has been added to the games, all the better will we spend that day here with Dionysius.

\ciceroSection{2} About Trebonius I quite agree with you. About Domitius --- ``fig is like fig, by Demeter; no fig has ever been so like a fig'' \textit{[Greek: suk\=oi, ma t\=en D\=em\=etra, sukon oude hen hout\=os homoion gegonen]} as is that situation \textit{[Greek: peristasis]} of his to ours: either because it is from the same men, or because it is beyond expectation, or because the good men are nowhere to be found; one thing only is unlike, that to him it is deserved. As for the misfortune itself, I do not know whether the other reading is the better. For what is more wretched than that the man who for as many years as he has been alive has been a consul-designate cannot become consul --- especially when he canvasses either alone or certainly with no more than one rival? But if it is so (which I do not know whether it is) that he has in his pocket-tablets, in his calendar of consuls-to-be, pages no shorter than of consuls past, what is more wretched than him --- except the commonwealth, in which not even any better is hoped for?

\ciceroSection{3} About Natta I learned first from your letter; I had hated the man. As to the poem you ask about: what if he wants to slip away? What? Would you allow it? About Fabius Luscus, the matter I had begun: he was always a man very fond of us, and never hateful to me; for he is sharp enough, very modest, of good worth. Because I did not see him, I supposed he was out of town. I heard from Gavius the Firmian here that the man is at Rome and has been constantly here. It struck me. You will say, ``so small a cause?'' He had reported to me very many things, on no doubtful authority, about the Firmian brothers. What it is that has made him withdraw from me, if he has withdrawn at all, I do not know.

\ciceroSection{4} As to your warning me to bear myself in the political way \textit{[Greek: politik\=os]} and to keep the inner line \textit{[Greek: t\=en es\=o gramm\=en]}, so I shall do. But greater prudence is needed, which I shall seek from you, as I always do. I should like you, if you have any access, to sniff at Fabius, and to taste that dinner-companion of yours, and to write to me about these things and about everything daily. When there is nothing to write about, write that very thing. Take care of your health.
