% Ad Familiares 1.7 (ad-familiares-01-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S.D. P. LENTVLO PROCOS.}

\ciceroSection{1} I have read your letter, in which you write to me that it is welcome to you to be informed often, through me, of all matters, and that you readily perceive my goodwill toward you. Of these, the one --- that I should love you most --- I must do, if I am to be the man you have wished me to be; the other I do gladly, that, since we are separated by an interval of place and of time, I may converse with you by letter as often as I can. If this is done more rarely than you will look for, the reason will be that my letters are not of the kind I would dare to entrust to chance carriers; whenever I have the chance of trusty men to whom I may rightly give them, I shall not let it pass.

\ciceroSection{2} As to your wishing to know with what good faith and disposition each man stands toward you, it is hard to say of individuals; one thing alone I dare to write, which I have often signalled to you before and now write also having seen and proved the matter: that certain men, and those most who most ought to have helped you and could most have helped you, have envied your distinction; and that the case of your time and of mine, in matters dissimilar, has been the very same: those whom you had wounded for the commonwealth's sake openly attacked you, while those whose authority and standing and wishes you had defended have proved less mindful of your virtue than envious of your praise. At that time, as I have written to you before, I learned that Hortensius was very fond of you, and Lucullus zealous; and among the magistrates Lucius Racilius, in good faith and spirit, stood out alone. For my own championing and defence of your dignity, on account of the magnitude of the kindness you had done me, must seem to most men to carry the weight of duty rather than of judgement.

\ciceroSection{3} Beyond these I can be witness for no consular of zeal toward you, nor of duty, nor of friendly feeling. As for Pompey, who is most in the habit of speaking with me about you (and not only when I have raised the matter, but also of his own accord), you know that in those days he was not often in the Senate; and your last letter, which you sent recently, was --- so much I could easily understand --- very welcome to him. To me your humanity, or rather your supreme wisdom, has appeared not only welcome but admirable: for you have held back, by that letter, an excellent man, bound to you by your outstanding generosity toward him, who was suspecting (because of some men's view of his ambitions) that you had grown estranged from him; he has always seemed to me to favour your praise, even in the most suspicious time of Caninius; but on reading your letter, I saw clearly that he was thinking, with all his soul, about you and about your distinctions and advantages.

\ciceroSection{4} For this reason, regard what is to follow as written from his counsel and authority, the matter often discussed between us: since no senatorial decree exists by which the restoration of the King of Alexandria has been taken from you, and since the resolution that was written about this affair (against which, you know, a veto was interposed, that no one whatever should restore the King) has so much force that it seems rather the partisanship of angry men than the steady policy of the Senate, you can see clearly --- holding Cilicia and Cyprus as you do --- what you can effect and what you can secure: and if the affair seems to admit it, that you may hold Alexandria and Egypt, then it is for your dignity and for that of our empire that, having settled the King at Ptolemais or some place nearby, you should set out for Alexandria with fleet and army, so that, when you have made it firm with peace and garrisons, Ptolemy may return to his kingdom; thus it would come about that he be restored through you, as the Senate at first decreed, and brought back without a multitude, as the religious men have said is pleasing to the Sibyl.

\ciceroSection{5} But this counsel was so approved both by him and by us that we saw men would judge of your policy from the outcome: if it fell out as we wish and pray, all would say you had acted both wisely and bravely; if anything were stumbled at, those same men would say you had acted both greedily and rashly. So what you can attain is not as easy for us to point out as for you, who have Egypt almost in your sight. We for our part feel this: if you have made sure that you can master that kingdom, there must be no delay; if it is in doubt, there must be no attempt. This I affirm to you: if you carry the matter through to your wish, it will come about that, while you are absent, you will be praised by many; when you have returned, by all. The danger of a stumble I see, on account of the religious authority brought into the way. But while I exhort you to certain praise, so I deter you from a hazard, and I come back to what I wrote at the start: that men will pass judgement on your whole act not so much from your counsel as from the outcome.

\ciceroSection{6} If this method of action shall seem to you dangerous, the other was approved: that, if the King has kept faith with your friends who through your province and command have lent him money, you should help him with your auxiliaries and your forces; that the nature and position of your province is such that you may either confirm his restoration by helping or hinder it by neglecting. In this matter, what the case allows, what the cause, what the moment, you yourself will see most easily and best; what we have approved, I thought you should know above all from me.

\ciceroSection{7} As for your congratulating me on my position, on Milo's friendship, on the levity and weakness of Clodius --- we are not at all surprised that you, like an exceptional craftsman, take pleasure in your own brilliant works. Yet the perversity of men is incredible (I do not care to use a heavier word): those whom by favouring us in the common cause they could have kept on their side, they have alienated by envying us; by whose most malevolent disparagements know that we have nearly been driven off our old and long-held position --- not, indeed, that we should forget our standing, but that we should at last take some account of safety. Both could have been done magnificently, if there were good faith, if there were weight, in the consulars; but in most of them so much levity is found that constancy in the commonwealth pleases them less than splendour offends them.

\ciceroSection{8} I write to you the more freely on this account, because not only in those times of mine which I have attained through you, but already long ago, you favoured my praise and dignity even when they were nascent; and at the same time because I see that envy is not, as I had thought before, directed at my newness alone: for in you, the most noble of all men, I have observed the like vices of the envious, who have however readily allowed you to be among the leading men, but have certainly not wished you to fly any higher. I rejoice that your fortune has not been like ours; for there is much difference between praise being diminished and safety being abandoned. That I myself am not too sorry for my own case, your virtue has accomplished: for you have seen to it that more should appear added to the memory of my name than taken from my fortune.

\ciceroSection{9} I exhort you, then, moved both by your kindnesses and by my love, to follow with all care and industry that whole glory for which from boyhood you were inflamed, and never to let the greatness of soul, which I have always admired and always loved, be bent by anyone's wrong. Great is men's opinion of you, great the commendation of your generosity, great the memory of your consulship. You see, surely, how much more vivid these things will be, how much more illustrious, when some praise has been added from the province and the command. And yet I wish you so to handle what must be handled by army and command that you long beforehand meditate these other matters, prepare yourself for them, think of them, train yourself for them; and that you feel --- for since you have always hoped this, I do not doubt, when you have attained it, that you will understand it --- that you can most easily hold the highest and topmost station in the state. And that this exhortation of mine may not seem to you empty or undertaken without cause, this consideration moved me, that I thought you should be reminded by our common experience to consider, in all the rest of your life, whom to trust and whom to beware.

\ciceroSection{10} As to your writing that you wish to know what is the state of the commonwealth: there is great dissension, but the contest is unequal; for those who are stronger in resources, in arms, and in might have, in my view, achieved this --- by the folly and inconstancy of their adversaries --- that they are now stronger even in authority. So, with very few opposing, they have carried through the Senate everything which they had thought they could not have got even through the people without sedition: pay was decreed for Caesar, and ten legates; and that he should not be succeeded under the Sempronian law was easily carried through. I write this to you the more briefly because the present condition of the commonwealth does not delight me; I write it however to admonish you of what I myself, devoted to all literature from boyhood, have nevertheless learned more from experience than from teaching: do you, with your own affairs whole, learn that we must take account neither of safety without dignity, nor of dignity without safety.

\ciceroSection{11} As to your congratulating me on my daughter and on Crassipes, I acknowledge your kindness, and hope and pray that this alliance will be a delight to us. Our young Lentulus, a boy of outstanding promise of the highest virtue, see that you train both in the rest of the arts to which you yourself have always given study, and chiefly in the imitation of yourself --- for no schooling will be more excellent than this. We love him, and hold him dear, in the first rank: both because he is yours, and because he is a son worthy of you, and because he loves us, and has always loved us.
