% Ad Familiares 1.1 (ad-familiares-01-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S. D. P. LENTVLO PROCOS.}

\ciceroSection{1} I, in every duty --- or rather, in my devotion towards you --- give satisfaction to everyone else; to myself I never give satisfaction. For so great is the magnitude of what you have deserved of me that, when you would not rest from my own case until it was carried through, I --- because I do not bring off the same in your case --- think life bitter to me. The state of the case is this. Hammonius, the king's legate, is openly using money against us; the business is being run through the same creditors through whom it was being run when you were here. Of those who wish the king's cause forward (and they are few), all wish the matter referred to Pompey; the Senate confirms the calumny taken from religious scruple, not from religion, but from ill-will and resentment at the king's lavish bribery.

\ciceroSection{2} I do not cease urging Pompey, begging him, by now even accusing and warning him more freely, that he flee the great disgrace; but plainly he leaves no room either for our entreaties or our warnings. For both in daily talk and in the Senate openly he has so pleaded your case that no one could have pleaded with greater eloquence, weight, zeal, or strain, with the most full testimony to the duties he owes you and to his own affection towards you. You know that Marcellinus is angry with you. He shows that, this royal cause excepted, in all other matters he will be the keenest defender of yours. What he gives, we take; from his moving and re-moving the question of religion (and he has already moved it many times) he cannot be turned aside.

\ciceroSection{3} The matter before the Ides was handled thus (for this I wrote on the Ides in the morning): the opinion of Hortensius, of myself, and of Lucullus yields to religious scruple as regards the army --- for the matter cannot otherwise be held; but, according to that decree of the Senate which was passed when you put the question, it decrees to you that you bring back the king, in such manner as you can do consistently with the public interest, so that religious scruple removes the army, but the Senate's authorisation keeps you the agent. Crassus decrees three legates and does not exclude Pompey, for he holds that they may be drawn even from those holding command; Bibulus, three legates from among those who are private citizens. To him the rest of the consulars assent, except Servilius, who says that the king ought not to be brought back at all, and Volcacius, who, when Lupus put the question, decrees Pompey to do it, and Afranius, who assents to Volcacius. This circumstance increases the suspicion of Pompey's wishes, for it was noticed that Pompey's intimate assents to Volcacius. The struggle is hard; the matter is leaning. The not-obscure scurrying-about and contention of Libo and Hypsaeus, and the zeal of all Pompey's intimates, have brought the business to the opinion that Pompey desires it; and those who do not wish him to have it, those same men are not your friends because you advanced him.

\ciceroSection{4} We have less authority in the case, for that reason --- because we are in your debt; the good will we might earn is extinguished by men's suspicion, because they think we are obliging Pompey. Just as in matters that, long before you set out, had been chafed in secret by the king himself and by Pompey's intimates and household, and then openly stirred up by the consulars and brought into the greatest unpopularity, so we are now placed. The loyalty we keep, all will know; the love your friends here bear you in your absence, your own friends will recognize. If there were loyalty in those in whom there ought to have been the greatest, we should not be straining.
