% Ad Familiares 13.6 (ad-familiares-13-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO Q. VALERIO Q. F. ORCAE PROCOS.}

\ciceroSection{1} If you and the army are well, it is well. I trust you remember that I spoke with you in the presence of Publius Cuspius too, when I was seeing you off in your general's cloak, and that I dealt with you afterwards at greater length, asking that whatever connections of his I should commend to you, you would count them among my own. This, in your great kindness towards me and your unbroken regard, you took on most generously and most graciously.

\ciceroSection{2} Cuspius, a man most attentive in services to his own, has a remarkable protective regard for certain men from your province, because he was twice in Africa as the principal officer of the partnership's largest business there. So this attentiveness of his towards them I am in the habit of seconding to the limit of my means and credit. For that reason I thought I should set out the grounds for the commendation of all Cuspius's people in this letter, and that in the rest I shall do only this --- to attach to each that mark of recognition you and I have settled upon, and at the same time to indicate that the man belongs to the number of Cuspius's friends.

\ciceroSection{3} But this commendation, which I have wished to set down in writing, you should know is the most weighty of them all. For Publius Cuspius has pressed me with singular zeal to commend Lucius Iulius to you with the greatest possible care. To his zeal I scarcely seem able to do justice with the words I am accustomed to use when I want to do something most carefully; he calls for new ones, and supposes that I have at my command some special art for the purpose. To him I have promised that I would draw out from the inmost recesses of our craft an extraordinary species of commendation.

\ciceroSection{4} Since I cannot achieve that, I should like you in the matter itself to bring it about that he supposes, from the manner of my letter, that something incredible has been carried through. You will achieve this, if you bring out every kind of generosity which can flow from your courtesy and your power, not only in deed but in words --- yes, in countenance too: how much these things are worth in a province I wish I had tested, but I do suspect it. The man himself, whom I am commending to you, is fully worthy of your friendship, not merely because Cuspius tells me so, though that should be enough; but because I know his judgement in choosing men and friends.

\ciceroSection{5} How great this letter's force has been I shall judge before long, and I shall, as I trust, give you my thanks. As for myself, whatever I shall judge to be your wish and your concern, all of it I shall attend to with zeal and care. Take care of your health.
