% Ad Quintum Fratrem 2.4 (ad-quintum-fratrem-02-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI SALVTEM.}

\ciceroSection{1} Our Sestius was acquitted on the fifth day before the Ides of March; and, since it was a matter of vehement public interest that there should appear to be no division in such a case, he was acquitted by every vote. That thing which I had perceived was often a care to you --- that we should leave no opening to any malevolent man who might say we were ungrateful unless we bore that man's perversity, in certain matters, with the most complete humanity --- be assured that we have so brought it off in that trial that we were judged the most grateful of all. For in defending so morose a man we did him heaped-up satisfaction; and, what he most desired, Vatinius, by whom he was openly attacked, we cut to pieces at our pleasure, while gods and men applauded. Indeed even our Paulus, when he had been brought forward as a witness against Sestius, declared that he would lay an indictment in Vatinius's name if Macer Licinius hesitated; and Macer rose from Sestius's benches and affirmed that he would not fail him. What more would you have? That impudent and reckless man Vatinius went off deeply shaken and weakened.

\ciceroSection{2} Your son Quintus, that excellent boy, is doing his lessons admirably. I notice this the more now because Tyrannio teaches at my house. Both our houses are being built strenuously. To your contractor I have paid over half the money. I hope we shall be in joint quarters before winter. About our Tullia, who, by Hercules, is most affectionate towards you, I hope we have settled the business with Crassipes. Two days remain after the Latin Festival which are religiously kept; \dag the rest had been finished, the Latiar was on \dag.

\ciceroSection{3} As for that ``ample leisure'' of which you used to speak \textit{[Greek: amphilaphian]}, I miss it after a fashion that suits me, just so as to receive it gladly when it comes, but not now while it lurks to summon it forth. I am building in three places, restoring the rest. I live a little more freely than I used to; it was needed. If I had you with me, I should give place to the workmen for a little. But these things too, as I hope, we shall in a short time share between us.

\ciceroSection{4} Roman affairs stand thus: the consul Lentulus is excellent, his colleague not impeding --- so excellent, I say, that I have not seen a better. He has taken away all the comitial days; for even the Latin Festival is being repeated. And yet there have been thanksgivings.

\ciceroSection{5} So the most ruinous laws are being resisted, especially Cato's, on whom, however, our Milo played a fine trick. For that protector of gladiators and beast-fighters had bought beast-fighters from Cosconius and Pomponius, and never went out in public without those armed men. He could not feed them, and so could scarcely keep them. Milo perceived it; gave the commission to a man not of his household to buy the gang from Cato without suspicion. As soon as it was led off, Racilius, who is the one tribune of the plebs at this time, made the matter public, and said that those men had been bought for himself (so it had been agreed), and posted a notice that he would sell ``the Catonian gang.'' Great laughter followed on that notice. This Cato, then, Lentulus has removed from his laws --- and those who promulgated monstrous things about Caesar, who had no one to interpose. For as to what Caninius is moving about Pompey, the matter has gone quite cold; for the thing is not approved of, and our Pompey is being blamed for his friendship with P. Lentulus, and, by Hercules, he is not the same; for among that most ruined and lowest dregs of the people he has half offended on Milo's account, and the good have many things to wish for in him, many to censure. Marcellinus, however, in this one thing alone does not, in my view, satisfy: that he handles him too sharply --- though he does it with the Senate not unwilling, so that I the more gladly withdraw myself from the senate-house and from every part of the commonwealth.

\ciceroSection{6} In trials we are what we were; the house is full as it is at the very fullest. One thing happened, by Milo's lack of forethought, awkwardly --- about Sextus Clodius, whom at this time, and by feeble accusers, I had not wished to be accused. To him three votes were lacking on a most degraded panel. So the people are calling the man back, and he must be brought back; for men do not bear it, and, because when speaking among his own people he was nearly condemned, they see him already condemned. In that very business Pompey's offence stood in our way; for the senators' urn cleared him generously, the knights' was even, the tribuni aerarii condemned. But this discomfort the daily condemnations of our enemies console us for; among which, with my whole-hearted pleasure, Sevius has been dashed to pieces, the rest are being cut down. C. Cato has held a contio: that he will not allow the elections to be held if days for him to deal with the people are taken away. Appius had not yet returned from Caesar.

\ciceroSection{7} I am wonderfully waiting for your letters; and yet I know the sea has been closed up to now. Still, they were saying that some had come from Olbia who praised you uniquely and reported that you stood very high in the province; and the same men were saying that you would cross at the first sailing. This I desire; and although you yourself, of course, most of all --- still, I look forward to your letters even before that. My brother, farewell.
