% Ad Familiares 1.4 (ad-familiares-01-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S.D. P. LENTVLO PROCOS.}

\ciceroSection{1} On the sixteenth day before the Kalends of February, when we were standing very fairly in the Senate --- since we had on the day before broken that opinion of Bibulus about the three legates, and one contest was left, the opinion of Volcacius --- the matter was drawn out by our opponents through various calumnies. For we were holding the cause, in a full Senate, with no great variation, and with great unpopularity falling on those who were transferring the royal cause from you to another. That day we had Curio sour, Bibulus much fairer, even almost a friend; Caninius and Cato declared they would carry no law before the elections. The Senate cannot be held before the Kalends of February under the Lex Pupia, as you know; nor through the whole month of February except after the embassies have been finished or rejected.

\ciceroSection{2} This, however, is the Roman people's opinion: that the name of feigned religious scruple has been brought in by your enviers and detractors, not so much to obstruct you, as to keep anyone from wishing to go to Alexandria out of greed for an army. As to your standing, there is no one who does not judge that the Senate has had regard to it; for there is no one who does not know that the failure to come to a vote was the work of your opponents. If they now, in the people's name --- but in fact through the most criminal banditry of tribunes --- attempt anything, I have provided sufficiently that they should not be able to act with the auspices unbroken, or the laws unbroken, or even without violence.

\ciceroSection{3} I think I need not write about my own zeal nor about the injuries of certain men. For what should I either parade myself --- I who, if I were to pour out my life for your standing, should still seem to have reached no part of what you have deserved --- or complain of others' injuries, which I cannot do without the deepest pain? I cannot guarantee you against violence, especially with magistrates this powerless; violence excepted, I can affirm that, by the highest zeal of the Senate and the Roman people, you will keep your distinction.
