% Ad Familiares 1.5b (ad-familiares-01-05b) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO S.D. P. LENTVLO PROCOS.}

\ciceroSection{1} What is being done here, and what has been done, you have, I think, learned both from many men's letters and from messengers; but what lies in conjecture, and what seems likely to be, this I think it right that you should hear from me. After Pompey, before the people on the eighth day before the Ides of February, when he was speaking for Milo, was thrown out by shouting and railing, and in the Senate was accused harshly and bitterly by Cato in the great silence of his enemies --- he seemed to me deeply shaken. And so the Alexandrian cause, which still stands open to us (for the Senate has stripped nothing from you except what, by the same religious scruple, cannot be given to another), appears plainly to have been laid down by him.

\ciceroSection{2} Now this is what we hope for and are working towards: that the king, when he understands that he cannot achieve what he was scheming --- that he be brought back by Pompey --- and that, unless he is restored through you, he will be deserted and cast off, sets out to come to you. This he will, without any doubt, do, if Pompey only a little shows himself willing; but you know the man's slowness and silence. Nevertheless, we leave nothing pertaining to that business undone. The rest of the injuries put forward by Cato, we shall, I hope, easily resist. Of the consulars, I see you have no friend except Hortensius and Lucullus; the rest are partly more obscurely hostile, partly openly angry. Do you keep your spirit brave and great, and hope that, the rush of this featherweight man broken, you will recover your former standing and glory.
