% Ad Familiares 5.12 (ad-familiares-05-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO L. LVCCEIO Q. F. salutem dicit.}

\ciceroSection{1} When face to face I have often tried to take this matter up with you, a kind of almost rustic shyness has held me back: what I shall now set out the more boldly in absence. For a letter does not blush. I burn with an incredible eagerness, and not, I think, one to be reproached, that my name should be made famous and celebrated in your writings. Although you have often shown me that you would do this, I should like you to forgive this haste of mine. For the kind of writing you do, although it had always been keenly looked for by me, has so surpassed my expectation, has so taken hold of me or set me on fire, that I long for our affairs to be entrusted as quickly as possible to your monuments. For not only does the prospect of remembrance with posterity carry me away to a certain hope of immortality; there is also that other longing --- to enjoy in life either the authority of your testimony, or the proof of your goodwill, or the charm of your art.

\ciceroSection{2} Yet, while writing this, I was not unaware how heavily you are weighed down by the work you have already taken up and put in hand. But because I saw that the history of the Italic and Civil War was already nearly finished by you, and you had told me you were beginning the rest, I did not wish to fail to remind you to consider whether you would prefer to weave our affairs in with the rest, or, as many of the Greeks have done --- Callisthenes the Phocian war, Timaeus the war of Pyrrhus, Polybius the Numantine war, all of whom separated those wars from their continuous histories --- you too in like manner would set apart the civil conspiracy from the foreign and external wars. For my part, I do not see that it makes much difference to my fame; but it does make some difference to my impatience that you should not wait until you come to the place, but seize the whole subject and the whole period at once. And at the same time, if your whole mind is engaged on a single argument and a single person, I see already with the mind's eye how much fuller and more ornate everything is going to be. And yet I am not unaware how shameless I am: first, in laying so great a burden on you (for your occupations may refuse me); next, in even asking you to embellish me. What if you do not think those affairs deserve so much embellishment?

\ciceroSection{3} But still, the man who has once crossed the bounds of modesty had better be shameless boldly and outright. So I beg you, and beg you again, plainly: that you will embellish those affairs even more eagerly than perhaps you feel; that here you will be careless of the laws of history; and as for that bias of which you wrote so charmingly in a certain proem (where you show that you could no more be turned aside from it than the Hercules of Xenophon from Pleasure) --- if it argues for me with you a little vehemently, do not despise it, but lavish on our friendship something even beyond what truth allows. If we draw you to take up this work, it will be, I am persuaded, a subject worthy of your skill and your fullness.

\ciceroSection{4} For from the beginning of the conspiracy to my own return, it seems to me a sort of moderate-sized body could be made; in which you will both be able to use that knowledge of yours of civil revolutions, whether in setting out the causes of upheavals or in the remedies for distresses --- since you will both rebuke those things you think worthy of blame and approve, by setting out the reasons, those that please you; and, if you think you should treat the matter with the freedom you are accustomed to, you will mark out the perfidy, the snares, the betrayal of many men against me. The variety of my own fortunes too will furnish you, in writing, with many things full of a certain pleasure that may strongly hold the minds of men in the reading, you being the writer. For nothing is better fitted to delight the reader than the changes of times and the vicissitudes of fortune. Things which, although in the experiencing they were not desirable for us, in the reading will be enjoyable: for the safe recollection of past pain has its delight.

\ciceroSection{5} For other men, however, who have not been put through any trouble of their own, but look on others' fortunes without any pain, the very pity itself is delightful. Which of us is not delighted, with a certain compassion, by Epaminondas dying at Mantinea? Who only then orders that the spear-point be plucked from his side, when he had been told, on his asking, that the shield was safe --- so that even in the pain of his wound he might die with calm spirit and with praise. Whose attention in reading is not held in suspense by the flight and return of Themistocles? For the order of annals itself holds us only moderately, like a counting-out of fasti; while the doubtful and varied fortunes of often-excellent men have wonder, expectation, joy, distress, hope, fear; and if they are closed by some notable end, the spirit is filled with the pleasantest pleasure of reading.

\ciceroSection{6} I shall therefore have my wish all the more if you will be of this mind: that from your continuous writings, in which you cover the perpetual history of events, you set apart this, as it were a drama, of our affairs and outcomes. For it has its various acts and turnings, both of policy and of times. And I do not fear to seem to be hunting your favour with a small flattery when I declare this --- that I would rather be celebrated and adorned by you than by any other. For you are not the man not to know what you are, nor to think those rather to be envious who do not admire you than those who praise you to be flatterers; nor again am I so out of my mind as to wish to be commended to everlasting glory by one who would not, in the very work of commending me, win his own glory of art.

\ciceroSection{7} For the great Alexander did not wish to be painted especially by Apelles and modelled by Lysippus from any motive of favour, but because he thought their art would be a glory both to themselves and to him. And those artists made the likeness of the body known to the unknowing; which, even if they did not exist, distinguished men would be no more obscure for that. Nor is the Spartan Agesilaus less to be set high --- who would suffer no painted or modelled likeness of himself --- than those who toiled in that craft: for one little book of Xenophon's, in praise of that king, easily surpassed all the images and statues of all men. And this will be more outstanding to my mind, both for joy of heart and for dignity of memory, if I come into your writings rather than into others' --- because not only your art will have been at my service, as Timoleon's was Timaeus's or Themistocles's Herodotus's, but also the authority of a most distinguished and most respected man, tested in the greatest and gravest causes of the commonwealth and approved among the very first --- so that to me will seem given, not just the herald's call which Alexander, when he had come to Sigeum, said had been bestowed by Homer on Achilles, but also the weighty testimony imparted by a famous and great man. For I like that Hector of Naevius, who rejoices not only ``to be praised'' but adds also ``by a praised man.''

\ciceroSection{8} And if I do not get this from you --- that is, if some matter shall have hindered you (for I do not think it lawful that you should not grant me anything I ask) --- I shall perhaps be forced to do what some often reproach: to write about myself, on the example of many distinguished men. But, as does not escape you, here are the faults of this kind of writing: men necessarily write about themselves more bashfully, if there is anything to be praised; and pass over, if there is anything to be reproached. There is added that there is less credit, less authority; and many in the end reproach and say that the heralds of the gymnastic games are more modest, who, when they have placed the crowns on the rest of the victors, and proclaimed their names with a great voice, when they themselves are crowned before the games are dismissed, call in another herald, lest by their own voice they declare themselves to be victors.

\ciceroSection{9} These things we wish to avoid, and, if you take up our case, shall avoid; and we ask that you do so. And lest perhaps you wonder why, since you have so often shown me that you would commit our times' policies and outcomes to writing with the greatest care, we now press you with so much insistence and in so many words --- the spur of that eagerness has fired us of which I wrote at the start, the eagerness of haste: that we are quick of spirit, that others should know us in your books while we are still living, and that we ourselves, while we live, may enjoy our little glory.

\ciceroSection{10} Of these matters, what you intend to do, please, if it is no trouble, write back to me. For if you take up the case, I will draw up notes of all the events; but if you put me off to another time, I will speak with you face to face. You meanwhile will not be idle: you will polish what you have in hand, and love us.
