% Ad Atticum 4.5 (ad-atticum-04-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What are you saying? Do you imagine that I would rather have my writings read and approved by anyone than by you? Why then did I send it to someone else first? I was being pressed by the man to whom I sent it, and I had no copy. And besides --- for I have been gnawing for some time at what I must in the end swallow --- the palinode \textit{[Greek: palinodia]} did seem to me a touch dishonourable. But farewell to upright, true, honourable counsels. It is not credible what perfidy lies in those leading men, as they wish to be and as they would be if they had any good faith in them. I had sensed it, I had known it: led on, abandoned, cast aside by them. Yet I was of this mind, that I would agree with them in the commonwealth. They were what they had been. Scarcely at last, with you to push me, did I come to my senses. You will say that you had warned me and urged me to act in this way, but not also to write it. By Hercules, I wished to lay upon myself the necessity of this new alliance, so that no opening should be left for me to slip back to those who, even when they ought to pity me, do not cease to envy. Yet we were moderate in the case \textit{[Greek: hypothesei]}, as I have written. We shall be more lavish if he receives it gladly --- and those will snarl a little who take it ill that I have the villa which had been Catulus's, and never think that I bought it from Vettius; who say I ought not to have built the house, and say I ought to have sold it. But what does that matter, when in the opinions I delivered I said what they themselves would have approved, and yet they rejoiced that I had spoken against Pompey's wish? Let there be an end of it. Since those who can do nothing do not wish to love me, let us see to it that we are cherished by those who can. You will say, ``I should have wished it long ago.'' I know you wished it, and that I have been a thorough ass. But now it is time that I be loved by myself, since by them I cannot be at all. That you visit my house often is very welcome to me. Crassipes is forestalling the farewell-dinner. You shall come from the road straight to the gardens. It seems more convenient to come to you; the next day, of course; for what difference is it to you? But we shall see. Your men have decorated my library with their construction and the title-labels. I should like you to praise them.
