% Ad Atticum 4.6 (ad-atticum-04-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Of Lentulus, of course, I bear the loss as I ought. We have lost a good man, a great man, in whose surpassing greatness of spirit much humanity was tempered. We comfort ourselves with a poor consolation, but not none: that for his own part we grieve least of all --- not in the manner of Saufeius and your school, but, by Hercules, because he so loved his country that he seems to me, by some kindness of the gods, to have been snatched out of its conflagration. For what is fouler than our life, mine especially? You, indeed, although by nature a public man \textit{[Greek: politikos]}, have no servitude that is properly your own: you enjoy only the common name.

\ciceroSection{2} But I --- who, if I speak about the commonwealth what is right, am thought a madman; if what is needful, a slave; if I am silent, crushed and captive --- with what grief must I be? With as much, of course, as I am --- all the sharper because I cannot even grieve openly without seeming ungrateful. What if I should wish to retire and take refuge in the harbour of leisure? In vain; rather, into war and into the camp. Shall we then be followers \textit{[Greek: opadoi]} who refused to be captains \textit{[Greek: tagoi]}? It must be so; for I see that even you yourself --- whom I wish I had always obeyed --- think this best. What remains now is, ``You have drawn Sparta as your lot: this Sparta adorn'' \textit{[Greek: Spartan elaches, tautan kosmei]}. By Hercules, I cannot. I forgive Philoxenus, who preferred to be sent back to prison. Yet here in this place I am rehearsing with myself, that I may not disapprove of it; and when we are together, you will confirm me. I see that letters are often being written to me by you, but I received them all at once --- which itself increased my grief; for by chance I had previously read three in which it had been written that Lentulus was a little better. And then the lightning of the fourth! But he, as I have written, is not wretched. We, however, are made of iron.

\ciceroSection{3} As for your prompting me to write the piece against Hortensius, I have fallen upon other things, not unmindful of that charge of yours; but, by Hercules, in beginning I shrank back, lest I, who seem to have stupidly not borne that friend's intemperance, should stupidly make his injury famous if I write anything; and at the same time, lest the depth \textit{[Greek: bathytes]} of mine that appeared in the doing should be the more hidden in the writing, and the apology should seem to have something of frivolity in it.

\ciceroSection{4} But we shall see. You meanwhile send me something as often as possible. The letter I have just sent to Lucceius, in which I ask him to write up my affairs, do see that you take it from him (it is a very pretty one); urge him to hasten; and thank him for writing back that he will do as I ask. Visit my house as often as you can. Make some sign to Vestorius: he is very generous to me.
