% Ad Atticum 4.3 (ad-atticum-04-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I know for certain that you are eager both to know what is being done here, and to know my own affairs from me --- not because what is being done before everyone's eyes is more sure if I write it than when others either write it to you or report it, but that you may see clearly from my letters in what spirit I bear what is being done, and what at this time is the feeling of my mind or, in the whole, the state of my life.

\ciceroSection{2} On the third day before the Nones of November the workmen were driven from our building site by armed men. The portico of Catulus, which by the senatus consultum was being rebuilt under the consuls' contract and had nearly reached the roof, was disturbed; the house of my brother Quintus first broken in by stones thrown from our site, then set on fire by Clodius's order with the city looking on, with fires thrown in --- to the great complaint and groan, I shall not say of good men (whether there are any I do not know) but plainly of all men. That madman is rushing on; after this fury he thinks of nothing but the slaughter of his enemies; he canvasses neighbourhood by neighbourhood, openly holds out the hope of liberty to slaves. For before, when he was overturning trials, he had a hard and patent case but yet a case: he could deny, he could shift the matter to others, he could even defend something as lawfully done. After these ruins, fires, plunderings, deserted by his own people he scarcely keeps Decimus the master-of-ceremonies, scarcely Gellius; he uses the counsels of slaves; he sees that, if he openly kills all whom he wishes, his case will be in no way harder than it has been so far. So on the third day before the Ides of November, when I was coming down the Sacra Via, he came after me with his men. A shout, stones, clubs, swords --- all this unforeseen. We withdrew into the vestibule of Tettius Damio. Those who were with me easily kept the gangs from coming in. The man himself could have been killed; but I am beginning to attempt cure by diet, I am tired of surgery. He, when by everyone's words he saw himself being thrust, in person, not into court but to punishment, has since made all the Catilinas into Acidini. For he so attempted to storm and burn Milo's house, the one which is on the Cermalus, on the day before the Ides of November, that openly at the fifth hour he led up men with shields and drawn swords, others with lit torches. He himself had taken the house of P. Sulla as a kind of camp for that assault. Then out of the Annian house of Milo's, Q. Flaccus led out keen men: he killed men most well-known of the whole Clodian banditry, longed for the man himself, but he had withdrawn into the inner chambers of Sulla's house. Then the Senate, on the day after the Ides. Clodius at home. Marcellinus distinguished; all keen. Metellus by his railing dragged out the time of speaking, with Appius helping --- and indeed, by Hercules, your friend, of whose steadiness and courage your own most truthful letter spoke. Sestius furious. The man afterwards threatened to fire the city if his elections were not held. Milo, after Marcellinus's opinion was set out --- which he had read out from a written text in such a way as to comprise our whole case (the site, the fires, my own peril) into a single judgement, and to put all this before the elections --- gave notice that he would watch the heaven on every comitial day.

\ciceroSection{4} The public meetings turbulent under Metellus, rash under Appius, most furious under Publius. Yet this is the upshot: had Milo not announced against on the Field, the elections would have been held. On the twelfth day before the Kalends of December Milo, before midnight, came onto the Field with a great band. Clodius, having forces picked out of fugitives, did not dare to come onto the Field. Milo stayed until midday, with the wonderful joy of men, with the highest glory. The contention of the three brothers shameful; their violence broken, their fury despised. Metellus still demands that, on the next day, the announcement be made to him in the Forum: that there was no need to come onto the Field by night; that he would be in the Comitium at the first hour. So on the eleventh day before the Kalends, Milo came in the night to the Comitium. Metellus, at the first light, ran secretly into the Field by almost off-roads. Milo overtakes the man among the groves, announces against. The man took himself off with great and disgraceful railing from Q. Flaccus. The tenth day before the Kalends, market-days.

\ciceroSection{5} For two days no public meeting. The eighth day before the Kalends I was writing this at the ninth hour of the night. Milo was already holding the Field. Marcellus the candidate was snoring so that I, his neighbour, could hear. Clodius's vestibule was reported to me as quite empty: a few rag-bedecked fellows, a linen lantern. They were complaining that everything was being done by my advice, ignorant of how much spirit was in that hero, and how much judgement too. His courage is wonderful. I leave aside certain new and divine things; but this is the upshot. I do not think the elections will take place; that Publius will be a defendant, unless he is killed first, I think Milo will see to; if he should now offer himself in a crowd to him, I see he will be killed by Milo himself. He does not hesitate to do it; he carries it before him; that fortune of mine he does not fear. For he will never use the counsel of any envious or treacherous man, nor will he trust any inert noble.

\ciceroSection{6} We are in spirit, at least, vigorous --- even more than when we were flourishing; in our family fortune we are broken. Yet, that I might not be entirely emptied out, I, with Quintus my brother refusing, fall back, against my own resources, on the help of friends. What plan to take about my whole standing, in your absence, I do not know. So hurry.
