% Ad Quintum Fratrem 2.1 (ad-quintum-fratrem-02-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI salutem}

\ciceroSection{1} The letter you have read I had given in the morning. But Licinius did me a kindness in coming to me late in the evening, after the Senate had been dismissed, so that, if anything had been done, I might write it out to you, if it should seem fit to me. The Senate was more crowded than we had thought it could be in the month of December, on the eve of the feast-days. We consulars were there --- and the two consuls-elect, P. Servilius, M. Lucullus, Lepidus, Volcacius, Glabrio. Of praetors there was a good number of us; in all about two hundred. Lupus had stirred up expectation; he pleaded the case of the Campanian land with great care. He was heard in great silence. You know the matter of the case. He passed over none of our proceedings. There were certain barbs against Caesar, abuse of Gellius, complaints against Pompey in his absence. The case at last finished, he said he would not call for opinions, lest he lay a burden of quarrel upon us. From the railings of earlier times and from the present silence, he said he understood what the Senate felt. He began to dismiss the house. Then Marcellinus said, ``Do not, Lupus, judge by our silence what we are at this time approving or disapproving. I, for my part --- and I think the rest are the same --- am silent for this reason: because I do not think it fitting that, with Pompey absent, the case of the Campanian land should be moved.'' Then Lupus said he would not detain the Senate.

\ciceroSection{2} Racilius rose, and began to refer about the courts. Marcellinus he asked first. When that man had complained gravely of the Clodian fires, slaughters, stonings, he gave the opinion that he himself should draw the judges by lot through the urban praetor; that, with the lot of the judges drawn, the elections might be held; that whoever should impede the courts would be acting against the commonwealth. The opinion was greatly approved. C. Cato spoke against, and C. Cassius, with the largest applause of the Senate, since they preferred the elections to the courts. Philippus assented to Lentulus. Afterwards Racilius asked me first among the private members.

\ciceroSection{3} I made many words about the whole madness and banditry of P. Clodius; I accused him as a defendant, with many and favourable murmurings of the whole Senate. Old Antistius praised my speech with many words and, by Hercules, not without eloquence; he took up the cause of the courts and said he would treat them as the most ancient. Things were going for that opinion. Then Clodius, when asked, began by speaking to use up the day; he was furious that he had been harassed by Racilius in a stubborn yet polished way. Then his hired ruffians, suddenly, from the Graecostasis and the steps raised a great enough shout --- stirred up, I take it, against Q. Sextilius and Milo's friends. With that fear thrown in, suddenly, with the great complaint of all, we left. There you have the proceedings of one day; the rest, I take it, will be put off into the month of January. Of the tribunes of the plebs, by far the best we have is Racilius. Antistius too looks like he will be our friend; for Plancius is wholly ours. See, if you love me, that you sail with consideration and care in the month of December.
