% Ad Familiares 5.17 (ad-familiares-05-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. CICERO greets P. SITTIVS, son of Publius}

\ciceroSection{1} It was not from any forgetting of our friendship, nor from any breaking of my own custom, that in earlier times I sent you no letter, but because the earlier times lay in ruins of the commonwealth and of mine, and because the later times kept me back from writing by your most unjust and most bitter misfortunes. Since now an interval long enough has passed, and I have recalled to myself with greater care your courage and the greatness of your spirit, I did not think it foreign to my own ways to write you these things.

\ciceroSection{2} I, P. Sittius, both in those first times, in which you, absent, were being called into hatred and into accusation, defended you; and when in the trial and the peril of your closest friend your own charge was being joined to it, I, as best I could, most carefully looked after you and your case. And lately, on my own recent return, when I found the matter set up other than I should have wished had I been there, still in nothing did I fail your safety; and at that time, when the unpopularity over the corn supply, the hostility not only of your enemies but even of your friends', the unfairness of the whole court, and many other faults of the commonwealth, were of more weight than the cause itself and the truth, I failed your son Publius neither in work, nor counsel, nor labour, nor influence, nor testimony.

\ciceroSection{3} For which reason, with all the duties of friendship having been by me diligently and devoutly kept, I did not think this either should be passed over: to urge you and to ask of you that you remember that you are both a man and a citizen --- that is, that you should bear wisely the common and uncertain stroke which none of us can either avoid or guarantee against, and resist your grief bravely and your fortune; and consider that, both in our state and in the other states which have come to power, such a stroke has fallen upon many of the bravest and best men by unjust trials. Would that I were not writing truly, that you are without that commonwealth in which no man of judgement can take any pleasure!

\ciceroSection{4} As for your son: I am afraid that, if I write nothing to you about him, I may seem not to have given the testimony due to his courage; but if I write out all that I feel, I am afraid I shall, by my letter, freshen up your longing and your grief. But still you will most prudently bring it about, if you reckon his devotion, his courage, his industry, wherever you are, to be your own and to be with you. For those things which we encompass in our mind are no less ours than those we look on with our eyes.

\ciceroSection{5} Therefore his outstanding courage and his highest love for you ought to be a great consolation to you, as also we and the rest who esteem you not from fortune but from your own courage, and shall always so esteem you; and most of all the awareness of your own spirit, when you reflect that nothing has fallen upon you by your own deserving, and add this --- that wise men are moved by shame, not by chance; by their own offence, not by another's injury. I, prompted both by the memory of our old friendship and by the courage and devotion of your son, shall fail in no place either to console or to lighten your fortune. If by chance you shall write me anything, I shall bring it about that you do not think you wrote in vain.
