% Ad Familiares 14.1 (ad-familiares-14-01) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS greets his TERENTIA, his TVLLIOLA, his CICERO}

\ciceroSection{1} Both from many people's letters and from everyone's talk, news has come through to me of your incredible courage and strength of spirit, that you are not worn out by labours either of mind or of body. Wretched man that I am! That you, with that courage, that good faith, uprightness, and kindness, should have fallen into such hardships on my account; and that our little Tullia, who used to take such joys from her father, should now take such griefs from him. As for Marcus, what shall I say? --- who, the moment he began to have understanding, took in only the bitterest of pains and miseries. If, as you write, I thought all this had been done by fate, I should bear it a little more easily; but everything has been brought about by my own fault, since I supposed I was loved by men who envied me, and did not follow those who sought me.

\ciceroSection{2} If I had used my own counsel and the talk of foolish friends or wicked ones had not weighed so much with me, we should be living the most blessed of lives. As things are, since my friends bid me hope, I shall take care that my own health does not fail your labour. I see how great the matter is, and how much easier it would have been to remain at home than to come back. But if we have all the tribunes of the plebs, if we have Lentulus as zealous as he seems, and Pompey too, and Caesar --- there is no need to despair.

\ciceroSection{3} About the household slaves we shall do as you write that our friends agreed. As to the place, the pestilence has by now gone away --- though, while it lasted, it did not touch me. Plancius, the most attentive of men, wants to have me with him, and still keeps me. I wanted to be in a more deserted place in Epirus, where neither Piso would come nor the soldiers; but Plancius still holds me back: he hopes that he may leave the province for Italy in my company. If I see the day, and come into your embrace, and recover both you and myself, I shall feel that I have taken a big enough harvest both of your devotion and of mine.

\ciceroSection{4} Piso's kindness, courage, and love towards every one of us is so great that nothing can stand higher. Would that the matter may bring him pleasure! to glory at any rate I see it will. About my brother Quintus I have not been blaming you. But I wished --- you so few as you are --- that you should be the most closely bound up together.

\ciceroSection{5} Those to whom you wanted me to give thanks I have thanked, and have written that you informed me of it. As for what you write, my Terentia --- that you will be selling the village --- I beg you, wretched man that I am --- what will become of us? And if the same fortune presses us, what will become of our poor boy? I cannot write the rest; my tears are too strong; nor will I draw you into the same weeping. I write only this: if our friends shall stand to their duty, money will not be lacking; if they shall not, you will not be able to make it good with your money. For our wretched fortunes' sake --- see that we do not destroy the boy in his ruin. If he has something to keep him from poverty, only middling virtue, only middling fortune, will be needed for him to come into the rest.

\ciceroSection{6} Take care of yourself, and send me letter-carriers, so that I may know what is being done and how you all are. The waiting now is altogether short for me. Greet little Tullia and Cicero. Farewell. Sent the sixth day before the Kalends of December, at Dyrrachium.

\ciceroSection{7} I have come to Dyrrachium, both because it is a free city and is dutiful to me, and the closest to Italy. But if its public crowds disturb me, I shall move elsewhere, and shall write to you.
