% Ad Quintum Fratrem 1.3 (ad-quintum-fratrem-01-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI salutem}

\ciceroSection{1} My brother, my brother, my brother, did you really fear that I, drawn on by some anger, had sent the boys to you without a letter, or even that I had not wanted to see you? I be angry with you? With you, of all men, could I be angry? Of course --- it was you who crushed me; your enemies, your unpopularity, ruined me, and not I you, in my wretchedness. That much-praised consulship of mine has torn from me you, my children, my country, my fortunes; from you let it have torn nothing, I beg, except myself alone. But surely from you all that has come to me has always been honourable and sweet; from me to you, only mourning over my calamity, fear over yours, longing, grief, loneliness. I, not have wanted to see you? Far from it: I did not want to be seen by you. For you would not have seen your brother --- not him whom you had left, not him whom you knew, not him whom you, weeping yourself, sent on his way weeping, as he saw you off on your departure --- not so much as a trace of him, not so much as a likeness, but a kind of breathing effigy of a dead man. And would that you had sooner seen or heard of me dead --- would that I had left you the survivor not only of my life but of my standing.

\ciceroSection{2} But I call all the gods to witness that I have been called back from death by this one saying alone --- that everyone said some part of your life was lodged in mine; in which I sinned, and acted wickedly. For if I had killed myself, death itself would easily have defended my devotion and love towards you. As things stand, I have brought it about that, with me alive, you should be without me; that, with me alive, you should be in need of others; that my voice should fall silent in dangers within my own household, when it had so often been a defence to total strangers. As for the boys having come to you without a letter, since you see now that anger was not the cause, certainly the cause was numbness and a kind of unbounded onrush of tears and grief.

\ciceroSection{3} With what weeping do you suppose I wrote these very lines? With the same, I know full well, with which you are reading them. Can I ever go for any time without thinking of you, or ever think of you without tears? For when I miss you, do I miss only a brother? No --- a brother nearly my own age in his sweetness, a son in his deference, a parent in his counsel. What was ever pleasant to me without you, or to you without me? And what of this --- that at the same time I miss my daughter? Such devotion, such modesty, such wit in her! The image of my face, of my speech, of my mind. And my son, the most charming of children, the dearest in the world to me? Whom I, hardened and made of iron, sent away from my embrace, a wiser little boy than I should have wished --- for the poor child already felt what was being done. And your son, your image, whom my Cicero loved as a brother and already revered as an elder brother? And what of this --- that I would not allow that most wretched woman, my most faithful wife, to follow me as I went, that there might be someone to look after the remains of our common disaster, the children of our two houses?

\ciceroSection{4} But still, in such fashion as I could, I did write, and I gave letters for you to your freedman Philogonus, which I believe were afterwards delivered to you; in which I urge and ask of you the same thing the boys carried in my words --- to push on at once for Rome, and to hurry. For first, I wanted you in a place of safety, in case there were any of our enemies whose cruelty was not yet sated by our calamity. Next, I dreaded the lamentations of our meeting; and the parting I could not have borne; and indeed I feared the very thing you write of --- that you could not have been torn from me. For these reasons this great evil, that I have not seen you --- than which nothing more bitter, nothing more wretched, seems able to have befallen brothers as loving and as close as we are --- has been less bitter, less wretched, than our meeting and above all our parting would have been.

\ciceroSection{5} Now, if you can do what I, who always seemed brave to you, cannot, lift yourself up and steady yourself, if any contest is to be entered. I hope, if my hope has any weight, that your own integrity, and the city's love for you, and some pity even for me will bring you a guard. If you are free of that danger, you will act, of course, if you think anything can be done about my case. Many men write to me of many things, and show themselves hopeful; but I do not see what I should hope for, when our enemies have the most power, friends have partly deserted me, partly even betrayed me --- who perhaps dread, in my return, blame for their own villainy. But what kind of things these are, please look closely and make plain to me. For my part, as long as you have need of me, if you see any danger that must be faced, I shall live. Longer than that, in this life, I cannot be. For there is no prudence and no learning that has strength enough to bear so much grief.

\ciceroSection{6} I know there has been both a more honourable moment for dying and a more useful one; but not that alone --- I have let many other moments slip past, which if I were to begin complaining of as past, I should do nothing but increase your grief, and declare my own folly. But this must not, and cannot, be done: that I should linger in a life so wretched, so disgraceful, longer than your need or firm hope demands. That I, who lately was the most blessed of men in my brother, my children, my wife, my resources, in the very kind of my wealth, in standing, authority, repute, and influence yielding nothing to those who at any time stood at the very top --- I, now in this fortune already afflicted and ruined, can no longer mourn either myself or mine.

\ciceroSection{7} What is the meaning, then, of what you wrote to me about money exchange? --- as if it were not your resources that now keep me up; in which very thing I see, wretchedly, and feel, what offence I have committed, when out of your own flesh and your son's you have to satisfy those to whom you owe, while I shall have flung away to no purpose money drawn from the treasury in your name. Still, M. Antonius and Caepio have each been paid as much as you had written; for myself, for what I am planning, what I have is enough --- whether we are restored, or whether we are given up for lost, no more is needed. If trouble comes upon you, I think you should turn to Crassus and to Calidius. How far Hortensius is to be trusted I do not know.

\ciceroSection{8} Hortensius dealt with me, under the highest pretence of love and with the highest daily assiduity, most wickedly and most treacherously, with Q. Arrius along with him --- by whose counsels, promises, and instructions I was abandoned and fell into this calamity. But you will keep all this hidden, lest it do harm. Beware of this --- and for that reason I think you should warm Hortensius himself by way of Pomponius --- that the verse aimed at you when you were standing for the aedileship, about the lex Aurelia, may not be confirmed by false testimony. For I fear nothing so much as that, when men come to understand how much pity your prayers and your safety will bring to us, they may attack you the more savagely.

\ciceroSection{9} Messalla I take to be well disposed to you; Pompey I think a counterfeit even now. But would that you may not put these things to the test! I should pray the gods so, if they had not ceased to hear my prayers. Yet I do pray them to be content with these unbounded ills of ours --- in which still no infamy of any wrongdoing lies, but the whole of the grief is this: that for my best deeds the greatest punishment has been appointed.

\ciceroSection{10} My daughter, and yours, and our Cicero --- why, my brother, should I commend them to you? Rather I grieve at this, that their orphanhood will bring you no less grief than me. But while you are safe, they will not be orphans. The rest --- so may some safety be granted to me, and the power of dying in my country --- the rest my tears do not let me write. I should wish you also to look after Terentia, and to write back to me about everything. Be brave, so far as the nature of the case allows. On the Ides of June, at Thessalonica.
