% Ad Atticum 3.10 (ad-atticum-03-10) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What had been done up to the eighth day before the Kalends of June I have learned from your letter; the rest I was waiting for, in your judgement, at Thessalonica. With those reports in hand, I shall be better able to settle where to be. For if there is a case, if anything is moving forward, if I see hope, either I shall wait here or I shall come over to you. If, as you write, those rumours have died away, we shall look to something else. So far you all signal me nothing but the discord of those men --- a discord which is between them, however, on every other subject rather than on me. So what good it does me I do not know. But, so long as you all want me to hope, I shall do as you bid.

\ciceroSection{2} As to the fact that you so often, and so vehemently, take me to task and say that I am weak in spirit --- I ask you, what evil is so great that has not fallen in my calamity? Has anyone ever fallen in this way, from so high a station, in so good a cause, with such resources of talent, of judgement, of favour, with such defences from all good men? Can I forget what I was, not feel what I am, what honour I have lost, what glory, what children, what fortunes, what a brother --- whom (mark this new variety of disaster), though I cared for him more than for myself and always had, I avoided seeing, lest I should look upon his mourning and his squalor, or offer my ruined and afflicted self to him who had left me at the height of my flourishing. I leave aside the rest of what is unbearable: weeping stops me. Am I, then, to be blamed for grieving, or for having allowed it to come to this --- that I did not either keep these things, which would have been easy if my own destruction had not been planned within my own walls, or at any rate not lose them while I lived?

\ciceroSection{3} I have written this so that you might raise me up, as you do, rather than judge me deserving of correction or rebuke; and the less I write you the more, both because grief obstructs me and because I have more to expect from your side than to write of myself. If anything is brought, I shall let you know our plan. As you have done up to now, please write me about as many things as possible, that I may be entirely ignorant of nothing. Sent the fourteenth day before the Kalends of July, from Thessalonica.
