% Ad Atticum 3.19 (ad-atticum-03-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} As long as letters of the kind on which I might still expect something from you all were being brought to me, I was held by hope and longing at Thessalonica. After it seemed to me that the action of this whole year was finished, I would not go into Asia, both because crowds are a thing I hate, and because, if anything were done by the new magistrates, I did not want to be far away. So I have decided to come over to you in Epirus --- not that the nature of the place itself can matter to one who avoids the daylight altogether, but because, if life and safety are open, it is from your harbour I shall most gladly set sail for them; and if life and safety are cut off, nowhere shall I more easily either bear out this most wretched life of mine, or --- which is far better --- throw it from me. I shall be with a few; I shall send the rest of the company away.

\ciceroSection{2} Your letters never brought me into so much hope as those of others did; and yet my own hope has always been thinner even than your letters. But still, since the matter has begun to be moved --- in whatever way it has begun, and from whatever cause --- I shall not desert either the wretched and mournful prayers of my best, my only brother, nor the promises of Sestius and the others, nor the hope of that most weighed-down woman Terentia, nor the supplication of that most wretched little Tullia, nor your own faithful letters. Epirus will give me either the road to safety, or what I wrote above.

\ciceroSection{3} I beg you and entreat you, T. Pomponius: if you see me stripped by men's faithlessness of all that was the highest, the dearest, and the sweetest, if you see me betrayed and thrown out by my own counsellors, if you understand that I was forced to destroy both myself and my people --- assist me by your pity, and Quintus my brother (who can be safe) hold up; defend Terentia and my children; wait for me, if you think you will see me there; if not, come and see me, if you can; assign me from your land only as much as my body can take up; and send slaves to me with letters as soon as can be and as often as can be. Sent on the sixteenth day before the Kalends of October.
