% Ad Familiares 14.4 (ad-familiares-14-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS S. D. TERENTIAE ET TVLLIAE ET CICERONI SVIS}

\ciceroSection{1} I write to you less often than I can, because while every hour is wretched to me, in particular when I am either writing to you or reading what you write, I am so undone with weeping that I cannot bear it. Would that we had been less greedy of life! Surely we should have seen no, or no great, evil in life. But if fortune has reserved us for some hope of one day recovering some good thing, my error is the lighter; if these troubles are fixed, then indeed I long to see you, my life, as soon as I can, and to die in your embrace --- since neither the gods, whom you have most chastely worshipped, nor men, whom I have always served, have repaid us.

\ciceroSection{2} We were thirteen days at Brundisium with M. Laenius Flaccus, a most excellent man, who held the danger to his fortunes and to his life of less account than my safety, and was not deterred by the penalty of that most wicked law from doing for me the right and duty of guest-friendship. May we one day be able to render him thanks! at any rate, we shall always feel them.

\ciceroSection{3} We have set out from Brundisium on the second day before the Kalends of May; we were aiming through Macedonia for Cyzicus. Lost me, stricken me! What shall I now ask of you, of you a sick woman, undone in body and mind? Shall I not ask? Am I to be without you, then? I think I shall manage thus: if there is any hope of my return, you confirm it and help the matter on; but if, as I fear, it is over, come to me by whatever means you can. Know this one thing: if I have you, I shall not seem to myself wholly to have perished. But what is to become of my little Tullia? See to that yourselves now; I have no counsel. But certainly, whatever shape the matter takes, the poor little thing's marriage and her good name must be guarded. What of my Cicero? May he, indeed, always be in my bosom and embrace. I cannot write any more now; grief stops me. What you have done I do not know --- whether you keep something, or, as I fear, are wholly stripped. Piso ---

\ciceroSection{4} as you write, I hope will always be ours. About the slaves manumitted, you should not be troubled. First, this was promised to your slaves: that you would do as each had deserved. As yet Orpheus is dutiful, and besides him no one in any great degree; of the rest of the slaves the matter is this --- that, if our property had passed from us, they should be our freedmen if they could prevail; but if it should still be ours, they should remain slaves --- save very few. But these are smaller matters.

\ciceroSection{5} As for your urging me to be of great spirit and to keep hope of recovering my safety, I should wish it might be of such a kind that we could rightly hope. Now in my misery, when shall I receive your letters? Who shall bring them to me? --- letters which I should have waited for at Brundisium, had it been allowed by the sailors, who would not let the favourable weather go by. What remains, hold yourself up, my Terentia, as honourably as you can. We have lived, we have flourished; not our vice but our virtue has cast us down; no fault is ours, unless that we did not lose our life along with our ornaments. But if it was more pleasing to our children that we should live, let us bear the rest, however unbearable. And I, who am putting heart into you, cannot put it into myself.

\ciceroSection{6} Clodius Philhetaerus, because he was hampered by his eyes, a faithful man, I sent back. Sallustius surpasses all others in his service. Pescennius is most well disposed to us; whom I always hope will be observant of you. Sicca had said he would be with me, but he left Brundisium. Take care, as you can, of your health, and reckon thus: that I am moved more vehemently by your misery than by my own. My Terentia, most faithful and best of wives, and my dearest little daughter, and our remaining hope, Cicero --- farewell. The day before the Kalends of May, from Brundisium.
