% Ad Familiares 14.2 (ad-familiares-14-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS greets his TERENTIA, his TVLLIOLA, his CICERO}

\ciceroSection{1} Do not suppose that I write longer letters to anyone, except when someone has written more at length to me, whom I think it right to write back to; for I have nothing to write of, and at this time I do nothing with more difficulty. To you and to our little Tullia I cannot write without floods of tears. For I see you to be the most miserable, whom I always wanted to be the most happy: this I should have made good, and, had we not been so timid, I should have made it good. Our Piso I love most --- as is his deserving.

\ciceroSection{2} Him, as I could, I have urged on by letter, and given him thanks as I should. Among the new tribunes of the plebs, I gather you have hope. That will be firm, if Pompey's will is for it; but I fear Crassus, for all that. From you indeed I see that all things are being done with the highest courage and the highest love --- and I do not wonder. But I grieve that the case is such that my own miseries are lightened by your great miseries. For P. Valerius --- a kind, attentive man --- wrote to me (and I read it with the freest weeping) of how you had been taken from the temple of Vesta to the Valerian Table. Ah --- my light, my heart's longing, the one to whom all used to look for help! That you, my Terentia, should now be so harried, lie now in such tears and in such squalor, and that this should be done by my own fault, who saved everyone else, that we ourselves should perish!

\ciceroSection{3} As to what you write of the house --- I mean the site --- I in truth shall feel myself restored at last only when that has been restored to us. But these things are not in our hand. What I grieve at is this: that out of the cost which has to be borne, you, in your wretchedness and stripped of all you have, must bear the share. If the business is brought through, we shall recover everything; if the same fortune presses us, will you, poor woman, fling away even the remains? I beg you, my life --- as for the cost, let others bear it, those who can if they will; and that weak health of yours, if you love me, do not strain. Day and night you stand before my eyes; I see you taking on every labour; I fear you will not bear it. But I see that everything is in you. So, that we may obtain what you hope and what you work for, look to your health.

\ciceroSection{4} I do not know to whom to write, except to those who write to me, or to those of whom you yourselves write me something. Further off, since it pleases you so, I shall not move; but I should like you to send letters as often as you can, especially if there is anything firmer for us to hope. Farewell, my dear ones, my heart's longings, farewell. Sent the third day before the Nones of October, from Thessalonica.
