% Ad Atticum 3.23 (ad-atticum-03-23) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} On the fifth day before the Kalends of December I received three letters from you. One was given on the eighth day before the Kalends of November, in which you urge me to wait with steady spirit for the month of January, and you write at length about Lentulus's zeal, Metellus's good will, and Pompey's whole reckoning, things you think pertinent to hope. In the second letter, contrary to your custom, you do not put on the date, but you give the time clearly enough: for you write that, on the very day on which the law was promulgated by the eight tribunes of the plebs, you handed in that letter --- that is, on the fourth day before the Kalends of November --- and you write at length on what use you think the promulgation has brought. As to which: if our safety, with this law, is now to be despaired of, I should wish that, by your love for me, you would think this empty diligence of mine pitiable rather than absurd; if there is some hope, give your effort that we may be defended hereafter by our magistrates with greater attention.

\ciceroSection{2} For that bill of the old tribunes had three clauses. The first, on my return, was incautiously drafted: nothing is restored except citizenship and rank --- which for myself, given my case, is enough --- but what should have been provided against, and how, you cannot fail to see. The second clause is the standard one on indemnity, in case anything has been done contrary to other laws by reason of this law. The third clause --- my Pomponius, look with what counsel and by whom it has been pressed in. For you know that Clodius enacted that his own law should hardly, or rather not at all, be open to being weakened either through the Senate or through the people. But you see that the protective clauses of laws to be abrogated are never observed. For if they were, scarcely any could be abrogated --- there is no law which does not hedge itself in by the difficulty of its abrogation. When a law is abrogated, the very rule on how it must be abrogated is abrogated with it.

\ciceroSection{3} Since this is in truth so, and since it has always been so held and so observed, our eight tribunes have set down this clause: ``If anything has been written in this bill which, by the laws or plebiscites'' --- that is, by the Clodian law --- ``it is not permitted, has not been permitted, to promulgate, abrogate, derogate, obrogate without fault to himself, or by reason of which any penalty or fine attaches to him who has promulgated, abrogated, derogated, obrogated --- this is not to redound against him.''

\ciceroSection{4} Now this in those tribunes' own case did them no harm: by the law of their own college they were not bound. So it is the more suspicious that someone meant ill by it, when they wrote into the bill what did not concern them, but was directed against me --- so that the new tribunes, if they should be too timid, would all the more think they had to use that clause. Nor did Clodius let it pass: he said in a contio on the third day before the Nones of November that, by this clause, the tribunes-elect were instructed in what was permitted to them. Yet you cannot fail to see that there is no such clause in any other law: which, if it were needed, all would use in abrogations. Please find out how this escaped Ninnius or the rest, and who brought it forward, and how the eight tribunes did not hesitate to refer the matter to the Senate concerning me --- whether because they thought that clause not worth observing --- and were nevertheless so cautious in abrogating that they feared a thing, free of it as they were themselves, which not even those who are bound by the law need worry about. I should be much against the new tribunes carrying that clause; but let them only carry anything; with one clause, by which I am to be recalled, provided the thing is brought through, I shall be content. I am ashamed by now to write so much; for I am afraid that you will read this when the case is already lost, and what I think is diligence will look pitiable to you and ridiculous to others. But if there is any hope, look at the law that Visellius drafted for T. Fadius. That law I much approve. Sestius's law, which you write satisfies you, does not satisfy me.

\ciceroSection{5} The third letter was given on the day before the Ides of November, in which you set out with judgement and care what it is that seems to be holding back the matter --- on Crassus, on Pompey, on the rest. So I beg you that, if there shall be any hope that the thing can be brought through by the support of good men, by the authority of the few and the multitude of the many gathered together, you will give effort to having it broken through in a single rush, will throw your weight in, and will rouse up the rest. But if --- as I see, both with your reckoning and with my own --- there is no hope, I beg you and entreat you: love my brother Quintus, whom I, wretched, have wretchedly destroyed, and do not let him take any heavier counsel about himself than is good for the son of your sister; protect my Cicero --- the poor little fellow, to whom I leave nothing except the unpopularity and the disgrace of my name --- so far as you can; sustain Terentia, the most weighed-down of all women, by your offices. I shall set out for Epirus when I have heard the news of the first days. I should be glad if in your next letter you will write me how the beginnings have gone. Sent the day before the Kalends of December.
