% Ad Familiares 14.3 (ad-familiares-14-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{TVLLIVS greets his TERENTIA, his TVLLIA, his CICERO}

\ciceroSection{1} I have received three letters from Aristocritus, which I have very nearly washed away with tears. For I am undone by grief, my Terentia; and my own miseries do not torment me more than yours and your children's. And I am the more wretched in this than you, the most wretched of all: that the calamity itself is common to us both, but the fault is mine alone. It was my duty either to avoid the danger by accepting a legateship, or to resist by careful management and by the resources I had, or to fall bravely. Nothing has been more wretched, more disgraceful, more unworthy of us than this.

\ciceroSection{2} So I am undone with grief, but I am undone with shame as well; for I am ashamed not to have shown to my best of wives, and to my sweetest children, the courage and the careful management I owed. Day and night the squalor and the mourning and the weakness of your health stand before my eyes, and the hope of safety is held out only on the thinnest of threads. Enemies are many, those who envy me almost everyone; to throw me out was a great matter, to keep me out is easy. Still, so long as you keep hoping, I shall not give way --- lest everything seem to have fallen by my fault.

\ciceroSection{3} As to your worry that I should be in a place of safety: that for me is now the easiest thing of all, since even my enemies want me to live, in such miseries. Still I shall do what you bid. Those friends to whom you wished I should give thanks I have thanked, and have given the letter to Dexippus, and have written that you informed me of their service. About our Piso: I see, and everyone says, that he is in a remarkable zeal of love and service towards us. May the gods grant that I may, with such a son-in-law present, enjoy his company together with you and our children! All the hope that remains now is in the new tribunes of the plebs, and in the first days at that. For if the matter grows old, it is finished.

\ciceroSection{4} For that reason I have sent Aristocritus straight back, that you may write me at once the beginnings of things and the whole reckoning of the business; though I have given Dexippus the same charge, to come back at once, and have written to my brother to send letter-carriers often. For I am at Dyrrachium for this time on this account: to hear what is being done as quickly as can be, and I am here in safety, since this city has always been defended by me. When my enemies are said to be coming, then I shall go into Epirus.

\ciceroSection{5} As to your writing that you would come to me if I wished --- I, knowing that a great part of that burden is being carried by you, want you to be there. If you bring through what you are working at, I must come to you; if not --- but there is no need to write the rest. From your first letter --- or your second at the latest --- we shall be able to settle what we have to do. I should be glad if you would write me everything in the closest detail; though by now what I should expect is the thing itself rather than letters. Take care of your health, and persuade yourself of this: that nothing is, or ever has been, dearer to me than you. Farewell, my Terentia: I seem to be seeing you, and so I am unmade by tears. Farewell. The day before the Kalends of December.
