% Ad Atticum 3.22 (ad-atticum-03-22) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Although my brother Quintus and Piso had carefully written to me what was being done, I should still have wished that your own engagements had not stood in the way of your writing me, as you used to, about what was going on and what you understood. So far Plancius keeps me here by his generosity, when several times now I have tried to set out for Epirus. The man has had the hope put into him --- the hope, that is, not the same as mine --- that we may leave the province together; which he hopes will be much to his honour. But now, when the soldiers are said to be coming up, I shall have to make ready to leave him. When I do, I shall send to you at once, so that you may know where I am.

\ciceroSection{2} Lentulus, by his courtesies to me --- which he is showing in act and in promise and in letter --- gives me some hope of Pompey's good will: for you have often written to me that he is entirely in Pompey's power. About Metellus, my brother has written to me how far he hopes the matter has been carried by your hand.

\ciceroSection{3} My Pomponius, fight that I may be allowed to live with you and with my own; and write to me everything. I am pressed by grief, by longing both for all things and for my own people, who have always been dearer to me than I am to myself. Take care of your health.

\ciceroSection{4} Because, if I had gone through Thessaly into Epirus, I should have been hearing nothing for a very long time --- and because the men of Dyrrachium are eager for me --- I have made my way to them, when I had written all the foregoing at Thessalonica. From there, when I shall turn back towards you, I shall let you know; and I should like you to write to me everything as closely as you can, of whatever sort the news may be. By now I am waiting for either the thing itself or for not so much as the hope of it. Sent the sixth day before the Kalends of December, at Dyrrachium.
