% Ad Atticum 3.13 (ad-atticum-03-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} What I had written to you, that I would be in Epirus, I changed plan when I saw our hope thinned out and fading away; nor did I stir from Thessalonica, where I had decided to be until you wrote me something on the matter --- as in your last letter you had written that, after the elections, something would be moved about me in the Senate, and that Pompey had said as much to you. Since the elections have been held, and you write me nothing on the question, I shall take it as if you had written that nothing was being done, and shall not bear it heavily that I have been led on by the hope of a not-far-off time. The movement that you had written you saw which seemed likely to be useful for us, men who are coming say will not happen. The remaining hope is in the tribunes designate. If I shall wait on that, there will be no ground for thinking that I have failed my own cause and the wishes of my people.

\ciceroSection{2} As to your accusing me, often, of bearing my own case so heavily, you must forgive me, when you see me afflicted as you have never seen, never heard of, anyone. As for your writing that you hear I am even being shaken in mind by my grief --- no, my mind is sound. I wish only that it had been so sound in the moment of danger! when I dealt with men whom I supposed to hold my safety dearest, and found them my bitterest and cruellest enemies; who, as soon as they saw me bend a little under fear, drove me on so as to use up against my destruction every villainy and treachery they had. Now, since I must go to Cyzicus --- and the rarer therefore my letters will reach you --- I should the more want you to write everything in detail that you may think I have need to know. Make sure you love my brother Quintus; whom if I, wretched as I am, leave unharmed, I shall not think that I have wholly perished. Sent on the Nones of August.
