% Ad Atticum 3.20 (ad-atticum-03-20) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO greets Q. CAECILIVS Q. N. POMPONIANVS ATTICVS}

\ciceroSection{1} As to its being so, and your uncle having performed his duty: that I most strongly approve. I shall say I rejoice over it, if I am still allowed to use the word. Wretched man that I am! How everything would now be as we wished, if I had not been deserted by spirit, by judgement, by the good faith of those I trusted! Which I will not add up, lest I add to my grief; but I know full well there comes into your mind what our life was, what its sweetness, what its dignity. For the recovering of which --- by the gods --- press on, as you do, and see to it that I keep the birthday of my return in your loveliest house, with you and with my own. To this hope and expectation which is set before me I had wished above all to wait at your side in Epirus; but the letters that come to me are such that I think it more convenient for us to remain in these same parts.

\ciceroSection{2} As to the house, and Curio's speech --- it is as you write. In the whole question of safety, if only it shall be restored to me, everything will be contained; among the parts of it I prefer nothing to my house. But I leave you no specific charge: I commit myself wholly to your love and your good faith. That you have got yourself free of every other employment in the midst of so great an inheritance is very welcome to me. As for your promising your resources to my safety, that I may be helped by you in everything before all others --- I see what a defence that is, and I understand both that you take up many parts of my safety yourself, and that you are able to sustain them, and that I need not ask you to do so.

\ciceroSection{3} As to your forbidding me to suppose that anything has come into your mind which would seem from me towards you to have been done or omitted contrary to what was right --- I shall yield to you and free myself of that care, but I shall owe you the more, the more your kindness towards me has been more lofty than mine has been towards you. Please write me what you see, what you understand, what is being done; and urge on all your circle for my safety. The bill of Sestius has neither standing enough nor caution enough: for it ought to be carried with my name in it, and the matter of my goods should be more carefully drafted; and I should be glad if you would attend to that. Sent the fourth day before the Nones of October, from Thessalonica.
