% Ad Atticum 2.13 (ad-atticum-02-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} An unworthy deed! That the letter I wrote back at once \textit{[Greek: auth\=orei]} from Tres Tabernae to your sweetest letters has been delivered to no one! But know that the bundle into which I had thrown it was carried back to my house on the very day on which I had given it, and brought back to me at the Formian villa. So I have ordered your letter to be sent back to you, that you may understand from it how welcome those of yours were to me at the time.

\ciceroSection{2} What you write --- that at Rome it is hushed up --- so I had supposed; but, by Hercules, in the country it is not hushed; and now the very fields cannot bear that kingship of yours. If indeed you come into this Telepylus, this Laestrygonian land \textit{[Greek: T\=elepylon Laistrygoni\=en]} --- I mean Formiae --- what an uproar of men! How angry the spirits! What hatred of our friend Magnus! whose cognomen, together with that of Crassus the Rich, grows old. Believe me, please: I have not yet met any man who would bear these things as patiently as I do. Wherefore, believe me, let us philosophize \textit{[Greek: philosoph\=omen]}. On my oath I can say to you, nothing is worth so much. If you have letters for the Sicyonians, fly to the Formian villa, whence I think to set out the day before the Nones of May.
