% Ad Atticum 2.7 (ad-atticum-02-07) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} About the geography we shall reflect again and again. As to the speeches --- you ask me for two; one of them I had no wish to write [text uncertain], the other lest I praise the man whom I did not love. But this too we shall see. Finally something will appear, lest I quite seem to have been idle.

\ciceroSection{2} What you write to me about Publius is most pleasing to me, and I should be glad if, when you come, you bring with you everything you have hunted out by every track; and meanwhile, if you understand or suspect anything, write it. Especially what he is going to do about the legation. Indeed, before I read your letter, I was eager to ``go for the man'' --- not, by Hercules, that I might put off the case with him (for I am of wonderful eagerness for litigation), but it seemed to me that, if there was any popular favour for him in his having been made a plebeian, he was about to lose it. For what? You went over to the plebs that you might go to greet Tigranes? Tell me, do the kings of Armenia not return greetings to patricians? What more? I had whetted myself for goading this legation of his. If he despises that --- and if, as you write, this annoys both the proposers and the augurs of the curiate law --- it is an outstanding spectacle. And by Hercules, to speak the truth, our Publius has been treated with some contempt: first because, when he had once been the only man at Caesar's house, now he could not even be one of the twenty; next because one legation had been spoken of, another given. That fat one for collecting moneys is being kept, I take it, for Drusus of Pisaurum, or for the priest Vatinius; this thin courier-legation is given to a man whose tribunate is being kept for the season of those men. Light him up, please, what you can. Our one hope of safety is in the dissensions among them; some beginnings of which I have felt in Curio. Now also Arrius rages that the consulship has been snatched from him; Megabocchus and this bloodthirsty youth are most hostile. And then let there be added, let there be added, that quarrel of the augurate. I hope I shall be sending you many fine letters about these matters.

\ciceroSection{4} But I should like to know what that is which you obscurely throw out: that already even some of the five-men are talking. What can it be? For if there is anything, there is more good in it than I had thought. And these things I should like you to consider --- that I am not asking you in the practical way \textit{[Greek: kata to praktikon]}, as if my spirit longed to do something in the commonwealth. I have long been weary of steering, even when it was permitted; now indeed, when I am being forced to leave the ship, with the rudder not laid aside but snatched away, I wish to gaze at their shipwrecks from the land. I wish, as your friend Sophocles says, ``to hear, even under the roof, the heavy raindrop with a sleeping mind'' \textit{[Greek: kan hypo steg\=ei pykn\=es akouein psakados heudous\=ei phreni]}.

\ciceroSection{5} About the wall, see what is needed. The Castricius slip we shall correct, and yet Quintus had written to me of HS 17,000, not HS 30,000 to your sister. Terentia sends greetings. Cicero bids you give the same answer to Aristodemus about himself as you gave about your sister's son, his brother. About the Amalthea, as you remind me, we shall not neglect it. See that you keep well.
