% Ad Atticum 2.8 (ad-atticum-02-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} As I was eagerly waiting for a letter from you toward evening, as is my custom: behold, a messenger announces that boys have come from Rome! I call them, I ask whether they have any letters. They say no. ``What?'' I say. ``Nothing from Pomponius?'' Frightened by my voice and my look, they confessed that they had taken one but had lost it on the road. What more? I bore it most ill; for no letter from you in these days had come empty of some useful and pleasant matter. Now if there was anything in that letter you sent on the seventeenth day before the Kalends of May worth the chronicle, write as soon as possible, that we may not be ignorant; but if nothing but a joke, give that very thing back. And know that the young Curio came to me in greeting. His talk on Publius greatly agreed with your letter; he himself indeed wonderfully hates haughty kings. He told me equally that the youth was inflamed and could not bear these things. We are well off if our hope is in these. I think we should turn to other things. I give myself to history, although you may suppose I am Saufeius. Nothing is lazier than I.

\ciceroSection{2} But learn now my journeys, that you may settle where you will see me. To the Formian villa I wish to come at the Parilia [21 April]; thence, since you think I should pass up that pretty little drinking-cup at this season, on the Kalends of May I shall set out from the Formian villa, so that I may be at Antium on the fifth day before the Nones of May. For there will be games at Antium from the fourth to the day before the Nones of May. These Tullia wishes to watch. Thence I think to the Tusculan villa, then to Arpinum, to Rome on the Kalends of June. See that we may see you either at the Formian villa, or at Antium, or at the Tusculan villa. Restore to us our earlier letter and add something new.
