% Ad Atticum 2.23 (ad-atticum-02-23) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I do not think you have ever before read a letter of mine which was not written in my own hand. From this you can gather with how great an occupation I am pulled. For since I had no leisure, and since for the sake of resting my poor voice it was necessary for me to walk, I dictated this as I walked.

\ciceroSection{2} First, then, I want you to know this: that our friend Sampsiceramus is bitterly sick of his condition, and longs to be restored to that place from which he fell; and that he confides his pain to us and from time to time openly seeks medicine, which I can in no way find. Next, all the authors and allies of that side, with no adversary in the field, are growing old before time; the unanimity of all men, both of will and of speech, has never been greater.

\ciceroSection{3} As for myself --- for I know certainly that you wish to know this --- I take part in no public counsels, and have devoted myself wholly to forensic work and labour. From which, as may easily be understood, I am much engaged in recalling and missing the things I once did. But the kinsman of our Ox-Eyed Hera is hurling threats not slight and announcing his intentions; he denies them to Sampsiceramus, but flaunts and parades them to the rest. Therefore, if you love me as much as you certainly do, if you are asleep, wake; if you are standing, step out; if you are stepping, run; if you are running, fly. It is incredible how much I rest on your counsel and prudence, and --- what is the greatest --- on your love and faith. The greatness of the matter perhaps demands a long speech; the union of our minds is content with brevity. It matters very much to us that, if you cannot be at the elections, at least you be at Rome when the man has been declared. Take care of your health.
