% Ad Atticum 2.5 (ad-atticum-02-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} I do indeed desire, and have long desired, to see Alexandria and the rest of Egypt, and at the same time to leave behind this satiety of men with us, and to come back with some longing felt. But at this time, with these as the senders, ``I am ashamed before the Trojans and the long-robed Trojan women'' \textit{[Greek: aideomai Tr\=oas kai Tr\=oiadas helkesipeplous]} (Iliad 6.442). For what shall our best men, if any are left, say --- that I have been led from my view by some reward? ``Polydamas first will fasten reproach on me'' \textit{[Greek: Pouludamas moi pr\=otos elencheian anath\=esei]} (Iliad 22.100): that Cato of ours, who is alone to me as a hundred thousand. What indeed shall the histories say of us six hundred years hence? --- which I fear far more than the little rumours of the men who live today. But, I take it, let us wait and see. For if it is offered, there will be some power on our side, and we shall then deliberate. By Hercules, there is some glory, also, in not accepting. Wherefore if Theophanes \textit{[Greek: Theophan\=es]} happens to confer with you, do not utterly refuse.

\ciceroSection{2} About these matters I am waiting for your letters: what Arrius reports, in what spirit he bears himself slighted, who are being readied as consuls --- whether, as the people's talk says, Pompey and Crassus, or, as is being written to me, Servius Sulpicius with Gabinius; whether any new laws or anything new at all; and, since Nepos is setting out, on whom the augurate is being conferred (in this one thing alone I can be caught by them). See my levity. But why these things, when I wish to lay them aside, and to philosophize \textit{[Greek: philosophein]} with my whole spirit and all my care? So, I say, it is in my mind. I should have wished it from the beginning; now, since I have learned by trial how empty those things are which I had thought distinguished, I think of taking my reckoning with all the Muses.

\ciceroSection{3} About Curtius, however, write back to me more certainly: who is now being readied in his place; and what is being done about Publius Clodius; and everything, as you promise, write at leisure \textit{[Greek: epi schol\=es]}. And what day you think you are going to leave Rome please write me, that I may inform you in what places I shall be; and send a letter at once on the matters about which I have written to you. For I am eagerly awaiting your letters.
