% Ad Quintum Fratrem 1.2 (ad-quintum-fratrem-01-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{MARCVS QVINTO FRATRI}

\ciceroSection{1} Statius came to me on the eighth day before the Kalends of November. His coming, since you wrote that you would be torn apart by your people while he was away, was troublesome to me; that, however, it took away the expectation of you and the throng which would have gathered if he had come back together with you and not been seen before --- this seemed to me to have happened not inconveniently. For men's talk has been spent and many such voices have been let out: \textit{[Greek: but always some great fellow]} --- which I am glad to have been ended in your absence. As for his being sent to me to clear himself, that was not at all necessary.

\ciceroSection{2} For in the first place, he was never under suspicion with me; nor what I wrote to you about him did I write on my own judgment, but, since the policy and safety of all of us who go into public life depend not on truth alone but on reputation, I always wrote out to you the talk of others, not my own judgments. How frequent and how grave that talk was, Statius himself learnt by his coming. He happened in upon some men's complaints which were being made before me about him personally, and he could perceive that the talk of those who wished us ill burst out particularly upon his name.

\ciceroSection{3} What used to move me most --- when I heard that he had more weight with you than the gravity of your age, your office, your prudence demanded --- how many men, do you think, have approached me to commend themselves to Statius? How often Statius himself \textit{[Greek: in plain language]} laid down in conversation: ``This did not please me; I warned, persuaded, deterred''! In which matters, even if his fidelity is the highest --- which I fully believe, since you so judge --- yet the very appearance of so favoured a freedman or slave can have no dignity. And take this for fact (since I ought neither to speak rashly nor to keep silent in cunning): every material for the talk of those who wished to detract from you was given by Statius. Before, only this much could be perceived: that some were angry at your strictness; once he was manumitted, talk did not fail those who were angry.

\ciceroSection{4} Now I shall reply to those letters which L. Caesius delivered to me, in whom, since I understand it is your wish, I shall fail in nothing. One of these is about Zeuxis of Blaundus, whom you write that I am intimately commending to you, the most certain of matricides. About this matter, and about this whole class of business, lest perhaps you wonder that I have been made too courtier-like to the Greeks, hear a few words. When I felt that the complaints of the Greeks were carrying too much weight on account of men's natural readiness to be deceived, whoever I heard complaining about you, by every reason in my power I appeased him. First, I softened the people of Dionysopolis, who were our bitterest enemies; their leader, Hermippus, I bound to me not by my speech alone but by close friendship. I have embraced with all my civility Hephaestius of Apamea, that most fickle man Megaristus of Antandrus, Nicias of Smyrna, the most trifling characters; even Nymphon of Colophon. All this I did, not because either these men or the whole nation pleased me; I am thoroughly tired of their fickleness, their flattery, their minds serving not their duties but the times.

\ciceroSection{5} But to return to Zeuxis: when, in the conversation about M. Cascellius which you write of, he was uttering those very things, I stopped his speech and received the man into intimacy. What that great desire of yours was I do not know --- since you write that you wished, having sewn up two Mysians at Smyrna in a sack, to set up in the upper part of your province a like example of your strictness, and that for that reason you wished to draw out Zeuxis by every means: the man whom, brought of his own accord into court, you ought perhaps not to have let go, but to have him sought out and drawn by flattery (as you write) to trial was not necessary, especially a man whom I learn daily, both from his fellow-citizens and from many others, to be all but more noble than his city.

\ciceroSection{6} ``But I indulge the Greeks alone.'' What? Have I not appeased Lucius Caecilius by every means? What a man, with what anger, what spirit! What man, in short, except Tuscenius, whose case cannot be cured, have I not softened? Look, here is over our heads a fickle and sordid fellow, but yet of equestrian census, Catienus. He too will be softened. That you were rather sharp with his father I do not blame; for I know certainly that you did it with cause. But what was the use of such a letter as you sent to him --- ``He is setting up his own cross, from which I had previously dragged him down; I shall see to it that he is burnt with smoke amid the applause of the whole province''? What of your letter to one Gaius Fabius (for that letter too Titus Catienus is carrying about): ``I am told that Licinius the kidnapper, with his kite-coloured chick, is gathering the tribute''? You then ask Fabius to burn alive the father and the son if he can; if not, send them to you to be burnt by judgment. These letters, sent by you in jest to Gaius Fabius --- if indeed they are yours --- when they are read have an invidious atrocity of language.

\ciceroSection{7} And if you go back to the precepts of all my letters, you will see that nothing has been blamed by me except sharpness of speech and quickness to anger, and, perhaps occasionally, the carelessness of the letters sent. In which matters, if my authority --- against either your nature, a little keener, or a certain sweetness in anger, or salt and humour in speaking --- had carried more weight with you, there would be nothing for us to be sorry about. And do you think me only moderately pained when I hear in what reputation Gaius Vergilius is, in what reputation your neighbour Gaius Octavius? --- for if you set yourself above your nearer neighbours, the men of Cilicia and Syria, you make much of yourself indeed! And the pain of it is, that, since the men I have named do not surpass you in innocence, they yet surpass you in the art of gathering goodwill --- men who have known neither the Cyrus of Xenophon nor Agesilaus, those kings under whose supreme command no one ever heard a sharper word.

\ciceroSection{8} But, taking precautions on this point with you from the beginning, I am not unaware of how much I have profited. Yet now, leaving the province --- as it seems to me you are doing already --- leave behind, I beg you, a memory of yourself as pleasant as can be. You have a most affable successor; in everything else his arrival will make men miss you greatly. As to letter-sending, as I have often written you, you have shown yourself too open to entreaty. Take away all unfair ones, if you can, take away the unusual, take away those that contradict each other. Statius told me that letters written to you used to be brought to him, read by him, and, if they were unfair, you were informed; but before he himself came to you, no selection of letters had been made; from that came volumes of selected epistles which used to be censured.

\ciceroSection{9} On this kind I do not now warn you (for it is too late), and you can know that I have warned you on much, in different ways, with diligence; but that which I gave Theopompus in charge, when I had been admonished by him --- see to it through your friends, who love you, which is easy, that these classes of letters be removed: first the unfair, then the contradictory, then those absurdly and unusually written, last those insulting toward someone. And I do not think these things are as great as I hear; and if from your occupations they have been less attended to, look into the matter now and clean them up. I read a letter which Sulla your nomenclator was said to have written, not to be approved; I read several angry ones.

\ciceroSection{10} But to come at the right moment to the topic of letters: for as I was holding this page Lucius Flavius the praetor-elect, a man very intimate with me, came to me. He told me that you had sent letters to his procurators, which seemed most unfair to me, that they should diminish nothing of the goods which had been Lucius Octavius Naso's, of whom Lucius Flavius is heir, before they had paid Gaius Fundanius the money; and likewise that you had sent to the people of Apollonis, that they should not allow anything to be diminished of the goods which had been Octavius's, before the debt should be paid to Fundanius. These things did not seem to me likely; for they are far removed from your prudence. Should the heir not diminish? What if he denies the debt? What if he owes nothing at all? What? Is the praetor wont to judge that a debt is owed? What? Do I not love Fundanius, am I not his friend, am I not moved by pity? No one more so; but the path of right is in certain matters such that there is no place for favour. And so Flavius told me it was written in that letter, which he said was yours, that you would either grant favours to your friends, as it were, or bring troubles upon them as enemies.

\ciceroSection{11} Why say more? He took it heavily and made vehement complaint to me, and begged me to write to you as carefully as I could. Which I do, and I beg you very vehemently, again and again: yield to Flavius's procurators on the question of diminishing, and give the people of Apollonis no further direction in writing that would be against Flavius. You will do everything both for Flavius's sake and, of course, for Pompey's. By Hercules, I do not wish to seem generous to him out of an injury done by you. But I beg you also to leave behind some authority and monument of a decree or letter of yours which is suited to Flavius's affair and case; for the man takes it heavily, both as one most observant of me and as a man tenacious of his right and dignity, that he has carried his point with you neither by friendship nor by law. And, as I think, both Pompey and Caesar at one time commended Flavius's case to you, and Flavius himself had written to you, and certainly I had. Therefore, if there is any matter you think should be done at my asking, let it be this. If you love me, take care, work hard, see it through, that Flavius give to you and to me the greatest thanks possible. I beg you so that I could not beg with greater eagerness.

\ciceroSection{12} What you write to me about Hermias was, by Hercules, very troublesome to me. I had written to you a letter scarcely fraternal; that letter, stirred by the speech of Diodotus, the freedman of Lucullus, about the bargain that had just been made --- as I had heard --- I had written rather angrily, and was wishing to call back. To this not-fraternal letter, you ought fraternally to grant pardon.

\ciceroSection{13} About Censorinus, Antonius, Cassius, Scaevola --- that you are deeply loved by them, as you write, I am vehemently glad. The other things in the same letter were graver than I could have wished, \textit{[Greek: keep the ship upright]}, and \textit{[Greek: to die only once]}. Those things will be too great. My censures have been most full of love --- which there are some, but moderate and rather small. I should never have thought you worthy of the slightest reproof in any matter, since you are bearing yourself most scrupulously, had we not many enemies. What I have written to you with any admonition or reproof, I wrote out of the diligence of my caution, in which I both stay and shall stay; and I shall not stop asking you to do likewise.

\ciceroSection{14} Attalus of Hypaepa pleaded with me that you not stand in the way of the disbursement of the sum that had been decreed for the statue of Q. \textit{[Publicenius?]}. About this I both ask you and admonish you, that you not wish the honour of so worthy a man, and one so much our intimate, to be diminished or hindered through you. Besides, Licinus, the slave of Aesopus our tragedian, well known to you, has run away. He was at Athens at the house of Patro the Epicurean, passing as free; thence he came to Asia. Afterwards a certain Plato of Sardis, an Epicurean, who is wont to be much at Athens and who had been at Athens when Licinus arrived there, after he had learned from Aesopus's letters that the man was a runaway, arrested him and handed him over into custody at Ephesus --- but whether public custody or the mill, we could not gather sufficiently from his letter. Whatever it is, since he is at Ephesus, I should like you to investigate the man and bring him with you, with the utmost care. Do not consider how much the man is worth; for one who is so worthless is of small price. But Aesopus is so distressed by the slave's wickedness and audacity that you can do nothing more pleasing to him than to recover him through your help.

\ciceroSection{15} Now learn what you most desire to know. We have utterly lost the commonwealth --- so much so that Cato, a young man of no judgment but yet a Roman citizen, and Cato by name, scarcely escaped alive, in that, when he wished to indict Gabinius for bribery, and the praetors could not be approached for several days nor would they grant him hearing, he mounted the Rostra and called Pompey ``a private dictator.'' He was within an inch of being killed. From this you can see what the condition of the whole commonwealth is.

\ciceroSection{16} Yet to our cause men do not seem likely to be lacking; in a wonderful manner they declare themselves, offer their services, give pledges. For my part I am with the highest hope, and with even greater spirit; with hope, that we shall come out on top; with spirit, that I shall not even fear chance in this commonwealth. Yet the matter stands so: if he names the day to me, all Italy will run together, so that we may come away with our glory multiplied; if, however, he tries to act by force, I hope it will come about by the enthusiasm not only of friends but of strangers besides, that we may resist by force. All men are pledging both themselves and their friends, clients, freedmen, slaves, even their money. Our old band of good men is on fire with zeal and love for me. If any before were rather alien or rather lukewarm, now in hatred of these kings they join themselves to the good men. Pompey promises everything, and Caesar; whom I so trust that I diminish nothing of my own preparation. The tribunes-elect of the plebs are friends to us; the consuls show themselves at their best; we have praetors most friendly and citizens most keen --- Domitius, Nigidius, Memmius, Lentulus; many other singular good men. Therefore make sure you keep up your great spirit and good hope. Yet about each particular thing that is happening from day to day I shall keep informing you frequently.
