% Ad Atticum 2.17 (ad-atticum-02-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} Plainly, as you write, so I judge: Sampsiceramus is in disorder. There is nothing not to be feared. He is fitting up a tyranny by all admission \textit{[Greek: homologoumen\=os tyrannida syskeuazetai]}. For what does that sudden joining of marriage mean, what the Campanian land, what the pouring out of money? If these were the last things, there would still be too much evil; but the nature of the matter is such that these cannot be the last. For what? Can these very things by themselves please them? They would never have come to this point if they were not preparing themselves an entrance to other ruinous things. But, as you write, on the sixth day or so before the Ides of May at the Arpinum estate we shall not lament these things, lest the labour and the oil of our learning go to waste; rather, we shall confer with quiet mind.

\ciceroSection{2} Immortal gods! It is not so much good hope \textit{[Greek: euelpistia]} that consoles me as before, as indifference \textit{[Greek: adiaphoria]} --- of which I make use in nothing so much as in this civil and public affair. Indeed even what little is empty in me, and not unfond of glory \textit{[Greek: ouk aphilodoxon]} (for it is a fine thing to know one's own faults), is affected by a certain pleasure. For it used to sting me that the deeds of Sampsiceramus to his country might in six hundred years' time appear greater than ours. Of this care at any rate I am now free; for he lies cast down so far that even Curio's [text uncertain] looks to be standing up against him. But these things face to face.

\ciceroSection{3} Yet you seem to me to be likely to be at Rome at our coming, which I shall easily bear if it can be at your convenience; but if you come as you write, I should be glad if you would fish out from Theophanes what spirit toward me the Arabarch \textit{[Greek: Arabarch\=es]} has. You will ask, of course, in a tutelary way \textit{[Greek: kata to k\=edemonikon]}, and bring me from him as it were rules \textit{[Greek: hypoth\=ekas]} how to bear myself. From his talk we may suspect something about [text uncertain]\ldots
