% Ad Atticum 2.20 (ad-atticum-02-20) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} I have not failed Anicatus in any place, as I had understood you wished. Numestius, on your zealously-written letters, I have gladly received into friendship. Caecilius, in the things in which I can, I diligently protect. Varro is satisfying me. Pompey loves us and holds us dear. ``You believe him?'' you will say. I do believe him; absolutely he persuades me. But because the practical men, in all their stories and rules and verses too, bid us beware and forbid us to believe, the one I do, that I beware; the other, that I should not believe, I cannot do.

\ciceroSection{2} Clodius is up to now announcing danger to me. Pompey affirms there is no danger; he swears; he even adds that he himself will be killed by him sooner than I be hurt. The matter is being handled. As soon as there is anything certain, I shall write to you. If there is to be a fight, I shall summon you to the partnership of the labour; if rest is given me, I shall not stir you from the Amalthea.

\ciceroSection{3} Of the commonwealth I shall write to you briefly; for I begin to fear that the very paper may betray us. So hereafter, if I have to write you more, I shall darken it with allegories \textit{[Greek: all\=egoriais]}. Now indeed the state is dying of a certain new disease: that, while all disapprove of what has been done --- complain, grieve, and there is no variety of opinion in any matter, and they speak openly and now groan plainly --- yet no remedy is brought. For we judge that without internecine war it cannot be resisted; nor do we see what end of yielding there shall be, save destruction.

\ciceroSection{4} Bibulus, in men's wonder and goodwill, is in the heavens; his edicts and public-meeting speeches are copied out and read. By a new sort of way he has come into the highest glory. Nothing is so popular now as hatred of the populars. Where these things are going to break out I fear; but if I shall begin to discern, I shall write to you more openly. You, if you love me as much as you certainly do, see that you are unencumbered, that, if I shall call out, you may run to me. But I do, and shall, give pains that there be no need. As to what I had written, that I would write to Furius too: there is no need to change your name; I shall make myself Laelius and you Atticus, and I shall use neither my own handwriting nor my seal, if only there will be letters of such a kind that I do not wish them to fall into another's hands.

\ciceroSection{6} Diodotus has died; he has left us perhaps HS 10,000,000(?). Bibulus has put off the elections, by an Archilochean edict, to the fifteenth day before the Kalends of November. From Vibius I have received the books. The poet is silly --- he knows nothing --- but is not useless. I am copying and shall send him back.
