% Ad Atticum 2.4 (ad-atticum-02-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} You have done me a great kindness in sending me Serapion's book; from which (as may be said between us) I scarcely understand a thousandth part. For it I have ordered ready money to be paid you, lest you carry it as expense among your gifts. But since money has been mentioned: please see, in whatever way you can, that you settle with Titinius. If he does not stand by what he had offered, I should much prefer that the things bought ill be returned, if it can be done with Pomponia's goodwill. If even that cannot be, let the money be returned, sooner than that there should be any scruple. I should very much like you to settle this lovingly (as is your habit) and diligently before you set out.

\ciceroSection{2} Clodius then, as you say, is going to Tigranes! I should be glad if it were on those terms (?) [text uncertain]; but I am easily content. For that time is more convenient for our free embassy, when both our Quintus, as we hope, has settled into peace, and we have learned what kind of priest of the Bona Dea he is going to be. Meanwhile we shall delight ourselves with the Muses, with an even mind --- nay rather even with a glad and willing one --- and it will never come into my head either to envy Crassus or to repent that I did not on my own desert myself.

\ciceroSection{3} About the geography I shall give pains that I satisfy you; but I promise nothing certain. It is a great work; but yet, as you bid, I shall see that some product of this travelling of mine remains for you.

\ciceroSection{4} Whatever you have hunted up about public affairs, and especially whom you think will be consuls, see that I know. And yet I am too curious; for I have decided to think about public matters no longer.

\ciceroSection{5} We have inspected Terentia's grove. What more? Apart from the Dodonaean oak, we lack nothing to seem to possess Epirus itself.

\ciceroSection{6} Around the Kalends we shall be either at the Formian or at the Pompeian villa. You, if we are not at the Formian, if you love us, come to the Pompeian. It will both be very pleasant for us and not greatly out of your way.

\ciceroSection{7} About the wall I have given orders to Philotimus that he not hinder what may seem fit to you. Yet I think you should call in Vettius. In times like these, with every best man's life so doubtful, I greatly value one summer's enjoyment of the wrestling-school on the Palatine; but yet so that I should wish nothing less than that Pomponia and the boy should live in fear of its falling.
