% Ad Atticum 2.18 (ad-atticum-02-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} I have received several of your letters, from which I have understood with what an anxious and worried mind you were eager to know what was new. We are held on every side, and we no longer refuse to be slaves; what we fear, as if they were greater things, are death and exile, which are far less. And this present condition is groaned at with one voice by all, and is lifted up by no one's word. The aim, I suspect, of those in power is to leave room for no man's largesse. One alone speaks out and openly opposes them: young Curio. To him are given the loudest applause, the most respectful greeting in the Forum, and many other signs of goodwill from good men. Fufius they pursue with shouting, abuse, and hisses. From these things comes not hope, but greater grief, when you see the will of the citizenry let loose and its courage in chains.

\ciceroSection{2} And so that you may not by chance ask me \textit{[Greek: in detail]} about particular things, the whole matter has been brought to this pass: that there is no longer any hope that not only private men but even magistrates will be free. Yet within this oppression the talk in social circles --- only at dinner-parties --- is freer than it was. Grief is beginning to overcome fear, but in such a way that all is in the deepest despair. The Campanian law even has a curse for the candidates in their public meeting, if they make any mention that the land should be possessed otherwise than by the Julian laws. The rest do not hesitate to swear it; Laterensis is thought to have done handsomely in giving up the candidacy for the tribunate of the plebs to avoid swearing.

\ciceroSection{3} But about the commonwealth I do not care to write more. I am displeased with myself, nor do I write without the deepest pain. I bear myself --- since all are crushed --- not abjectly; given the greatness of what I have done, with too little courage. By Caesar I am very generously invited into that legation: to be his legate; and also a free legation \textit{[for vow's sake]} is offered me. But the latter has no real protection against the shame of little Pulcher, and would put me away at the time of my brother's return; the former is more fortified and does not prevent my being on the spot when I wish. This I am holding on to, but I do not think I shall use it; nor does anyone else know of it. I do not wish to flee; I am eager to fight. Men's enthusiasm for me is great. But I assert nothing; you will keep silent on this.

\ciceroSection{4} About Statius's being manumitted, and certain other matters, I am indeed troubled, but by now I have grown wholly callous. I should wish, indeed I should crave, that you were here; I should lack neither counsel nor consolation. But hold yourself in such readiness that, if I cry out, you fly to me.
