% Ad Atticum 2.9 (ad-atticum-02-09) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} When the quaestor Caecilius had suddenly told me that he was sending a slave to Rome, I have written this in haste, that I might draw out from you those wonderful dialogues with Publius --- both those of which you write, and that one which you hide and say it would be too long to write out the things you replied to him --- and especially that one which has not yet taken place, which the ox-eyed lady \textit{[Greek: bo\=opis]}, when she has come back from Solonium, will report to you. Of this know: that nothing can be more pleasing to me than this. But if the things settled about me are not being kept, I am in heaven; that this our Hierosolymarius transferer-to-the-plebs may know what gratitude he has returned for my most rotten speeches! Of which expect a divine recantation \textit{[Greek: palin\=odian]}. For, as far as we can guess by conjecture, if that scoundrel of his is in favour with these dynasts, he will be able to boast neither of the cynic ex-consul nor of those Tritons of fish-ponds. For we shall not be subject to any envy, stripped of our resources and that senatorial power. But if he disagrees with them, it will be absurd to inveigh against us. Nevertheless let him inveigh. Festively, believe me, and with less noise than I had supposed, this orb of the commonwealth has been turned --- swifter altogether than it could have been [text uncertain]. The fault is Cato's, but again the wickedness of those men, who have neglected the auspices, the Aelian law, the Junian and Licinian, the Caecilian and Didian, who have poured out all the remedies of the commonwealth, who have given kingdoms and lands to tetrarchs, and monstrous sums of money to a few.

\ciceroSection{2} I see now where the envy passes, and where it is going to dwell. You will reckon that I have learned nothing either by experience or from Theophrastus, if you do not see in a brief time a longing for those times of mine. For if there was envy of senatorial power when it was reduced not to the people but to three immoderate men, what do you suppose will be? Wherefore those men may make whom they will consuls, tribunes of the plebs, even dress Vatinius's tumour in a double-dyed \textit{[Greek: dibaph\=oi]} priestly robe; you will see in a brief time how great are not only those who did not waver but even that very man who erred --- Cato.

\ciceroSection{3} For us, if it shall be permitted by your fellow-clansman Publius, we are thinking of playing the sophist \textit{[Greek: sophisteuein]}, and only --- if he thinks of so much --- defending ourselves; and what is proper to this art: ``I declare to ward off any man, when one first does me harm'' \textit{[Greek: epangellomai andr' apamynesthai hote tis proteros chalep\=en\=ei]}. May my country be propitious. She has from us, even if not more than was owed, certainly more than was demanded. I had rather be carried badly with another at the helm than steer well with such ungrateful passengers. But these things face to face will go better.

\ciceroSection{4} Now hear what you ask. I think to take myself from the Formian villa to Antium on the fifth day before the Nones of May; from Antium I wish to set out for the Tusculan villa on the Nones of May. But when I have come back from the Formian villa (where I wish to be up to the day before the Kalends of May), I shall at once let you know. Terentia greets you; ``and little Cicero greets Titus the Athenian.''
