% Ad Atticum 2.19 (ad-atticum-02-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} Many things weigh on me --- both from the great upheaval of the commonwealth and from the dangers held over me, and they are six hundred. But nothing is more troublesome to me than that Statius has been manumitted; that no respect should be paid even to my command --- but command is too much; not even to my displeasure! And I do not know what to do; the matter is not as great as the talk about it. Yet I cannot even be angry with those whom I love so much; I only grieve, and grieve marvellously. \textsection The rest is in great matters. \textsection The threats and contentions of Clodius which are held out to me touch me only moderately; for I think I can either undergo them with the highest dignity, or decline them with no trouble. You will perhaps say: enough of dignity, \textit{[Greek: as of the oak]}; if you love me, look to safety. Wretched me! why are you not here? Nothing surely would escape you. I am perhaps \textit{[Greek: blind]} and stare too much \textit{[Greek: at the seemly]}.

\ciceroSection{2} Be assured: never has anything been so disreputable, so disgraceful, so equally offensive to all kinds, orders, ages as the present condition --- more so, by Hercules, than I should have wished, let alone than I supposed it would be. These ``populares'' have now even taught modest men to hiss. Bibulus is in heaven --- I do not know on what grounds, but he is praised as if ``a single man by his delaying restored our state for us.'' Pompeius, my beloved --- which is the deepest grief to me --- has crushed himself. They hold no man by goodwill; I fear they may have to use fear. I myself neither fight against their cause for our friendship's sake nor approve it, lest I disapprove all that I have done before; I take a middle path. The people's feeling has been seen most plainly at the theatre and the games. For at the gladiators master and supporters alike were torn to pieces with hisses; at the Apollinarian games the tragedian Diphilus inveighed insolently against our Pompeius: ``By our misery you are great'' he was made to say a thousand times; ``A time will come when you will groan heavily for that virtue of yours'' he said amid the shouting of the whole theatre, and the rest besides. For these verses are of such a kind as to seem to have been written for the moment by an enemy of Pompeius: ``If neither laws nor morals can compel,'' and the rest, were said with great roar and clamour. When Caesar entered, his applause was dead; young Curio followed him in. Curio was applauded as Pompeius used to be applauded when the commonwealth was safe. Caesar took it heavily. A letter, it was said, was flying to Pompey at Capua. They were enemies of the knights who had stood up to applaud Curio, hostile to all; they threatened the Roscian law, even the grain law. The whole thing was very disturbed. For my part I should have preferred that what they had taken up should be passed over in silence; but I fear that may not be possible. Men do not endure what nevertheless seems to have to be endured; but already there is a single voice of all, more confirmed by hatred than by force.

\ciceroSection{4} As for our friend Publius --- he threatens me, he is hostile. Trouble hangs over me, to which of course you will fly. I think I have our old consular army of all good men, even of moderately good men, very firm. Pompeius signifies a not slight friendliness towards me; he affirms also that Clodius will not say a word about me. In which it is not he that deceives me; he himself is deceived. On Cosconius's death I was invited to take his place. That was being invited into a dead man's place. Nothing would have been more disgraceful for me before the world, nor in fact more alien to that very \textit{[Greek: security]}. For those men are odious to good citizens; I should have kept my own ill-will among bad men, and taken on another's besides. Caesar wishes me to be his legate.

\ciceroSection{5} This is a more honourable way of declining the danger; yet I do not refuse it. What then? I prefer to fight. Nothing certain, however. Again I say: I wish you were here! But still, if it is necessary, we will summon you. What else? What? This, I think. We are sure that all is lost; for why \textit{[Greek: do we keep up the pretence]} so long? But I have written this in a hurry, and, by Hercules, with timidity. Hereafter I shall write to you either, if I have someone trustworthy enough to whom I may give a letter, plainly all things; or, if I write obscurely, you will yet understand. In those letters I shall make myself Laelius, you Furius; the rest will be \textit{[Greek: in riddles]}. Here we are paying court to Caecilius and observing him diligently. I hear Bibulus's edicts have been sent to you. Over them our Pompeius is burning with grief and anger.
