% Ad Atticum 2.21 (ad-atticum-02-21) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} About the commonwealth, why should I write to you in detail? It has wholly perished, and is the more wretched than when you left it, because then the kind of domination that seemed to oppress the state was such as was pleasing to the multitude, and was so troublesome to the good men that yet it was so without ruin; now suddenly it is in such odium with all that we shudder at where it will burst out. For we have learnt by experience the irascibility and intemperance of those men, who, angered with Cato, ruined everything; but they used such gentle poisons that we seemed able to perish without pain. Now in truth, with the people's hisses, the conversation of the honourable, the murmur of Italy --- I fear they have caught fire.

\ciceroSection{2} For my part I had been hoping --- as I often used to say to you in conversation --- that the wheel of the commonwealth had so turned that we could scarcely hear the sound, scarcely see the imprint of the rut; and so it would have been, if men could have waited for the storm to pass. But, after they had long sighed in secret, after that they began to groan; finally now they have all begun to speak and to shout.

\ciceroSection{3} And so that friend of ours, unaccustomed to ill repute, lifelong frequenter of praise, swimming in glory, now disfigured in body, broken in spirit, knows not whither to take himself; he sees a headlong advance, an inconstant retreat; he has good men as his enemies, the bad men themselves not as his friends. And see the softness of my mind: I could not hold back tears when I saw him on the eighth day before the Kalends of August speaking against Bibulus's edicts. He who before had been wont to bear himself most magnificently in that spot, with the highest love of the people, all favouring him --- how lowered he was then, how dejected, how displeasing even to himself, not to those alone who were present!

\ciceroSection{4} O spectacle pleasing to Crassus alone, but to the rest by no means so! For because he had fallen from the stars, he seemed to have slipped rather than to have advanced; and, as Apelles, if he had seen his Venus, or Protogenes his Ialysus, smeared with mud, would surely have taken great pain, so I, when I saw this man --- whom I had painted and polished with all the colours of my art --- suddenly disfigured, took it not without great pain. Although no one supposed, on account of the Clodian business, that I ought to be friendly to him, yet so great was my love that no injury could exhaust it. And so Bibulus's Archilochian edicts against him are so pleasing to the people that we cannot pass the spot where they are posted up, for the throng of those reading them; to him so bitter that he is wasting away with grief; to me, by Hercules, troublesome too, both because they torture too much the man whom I have always loved, and because I fear, lest a man so vehement and so keen with the sword, and so unaccustomed to insult, may give in with all the impetus of his soul to grief and anger.

\ciceroSection{5} What Bibulus's outcome is going to be I do not know. As things stand now, his glory is wonderful. When he had postponed the elections to October --- a thing that usually offends the people's will --- Caesar thought he could in his speech force the assembly to go to Bibulus; though he spoke in the most seditious vein on many points, he could not extort a sound. Need I say more? They feel that they hold the goodwill of no party. The more we have to fear their violence.

\ciceroSection{6} Clodius is hostile to us. Pompeius assures me he will do nothing against me. For me it is dangerous to believe him; I am preparing to resist. I hope to have the highest enthusiasm of every order. As I miss you, so the moment itself calls you. Counsel, courage, finally protection too I shall gain in the largest measure if I see you in time. Varro satisfies me. Pompeius speaks divinely. I hope we shall come out either with the highest glory or at least without trouble. Take care to let me know what you are doing, how you are amusing yourself, what you have done with the Sicyonians.
