% Ad Atticum 2.14 (ad-atticum-02-14) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} How great an expectation you stir in me about Bibulus's speech, how great about your conversation with the ox-eyed lady \textit{[Greek: bo\=opidos]}, how great about that wanton dinner-party! So come in such a way as to thirsting ears. Although there is now nothing I think to be more feared than that this our Sampsiceramus, when he feels he is being beaten in everyone's talk and when he sees how easy these acts are to overturn \textit{[Greek: euanatreptous]}, may begin to rush. As for me, I am so weakened that, in this leisure in which we are now wasting away, I had rather be tyrannized over \textit{[Greek: entyranneisthai]} than fight even with the best hope.

\ciceroSection{2} As to your urging me, often, to compose --- it cannot be done. I have a basilica, not a country house, with the throng of Formians [text uncertain: ``to which side of the basilica the tribe Aemilia''\ldots]. But I leave aside the multitude. After the fourth hour the rest are not troublesome. Gaius Arrius is the next neighbour; nay, indeed, by now my mess-mate, who even says that he is therefore not going to Rome, that he may philosophize all day here with me. Behold, on the other side, Sebosus, that intimate of Catulus's. Where shall I turn? By Hercules I should at once set out for Arpinum, were I not to see that you are most conveniently to be expected at the Formian villa, at least to the day before the Nones of May. For you see what kind of men my ears have been given over to. O wonderful chance, if anyone, while these men are here at my house, should wish to buy from me the Formian estate! And yet may I bring something to the work --- something great and full of much thinking and much leisure? But yet enough shall be given by us, nor labour spared.
