% Ad Atticum 2.12 (ad-atticum-02-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} Do those men deny that Publius has been made a plebeian? This indeed is kingship, and is in no way to be borne. Let Publius send to me men to seal up the document; I shall swear that our Gnaeus, the colleague of Balbus, told me at Antium that he had been at the auspice. O the sweet letters of yours, two given me at one time, for which I do not know what good news \textit{[Greek: euangelia]} I shall give back: I confess plainly that I owe them. But see what coincidence \textit{[Greek: synkyr\=ema]}.

\ciceroSection{2} I had just come up out of the road from Antium into the Appian Way, at the Tres Tabernae on the very day of the Cerialia, when running into me from Rome came my Curio. There at the same place, instantly, a boy from you with letters. He of me: ``Have you not heard anything new?'' I say no. ``Publius,'' he says, ``is canvassing for the tribunate of the plebs.'' ``What say you?'' ``And as a most bitter enemy of Caesar, and''(he says) ``so as to rescind everything.'' ``What of Caesar?'' I say. ``He says he has carried no measure about that adoption.'' Then he poured out his own hatred --- and that of Memmius and Metellus Nepos. Embracing the young man I sent him off and hurried to my letters. Where are those who say ``the living voice'' \textit{[Greek: z\=os\=es ph\=on\=es]}? How much more I saw, from your letters than from his speech, of what was being done --- of the daily ruminations, of Publius's plans, of the cymbals of the ox-eyed lady \textit{[Greek: bo\=opidos]}, of the standard-bearer Athenio, of the letters sent to Gnaeus, of Theophanes's and Memmius's talk; how great, moreover, the expectation of that wanton dinner-party \textit{[Greek: aselgous]} you have given me! In curiosity I am sharp-hungry \textit{[Greek: oxypeinos]}, but I bear it easily that you do not write me about that drinking-party \textit{[Greek: symposion]}. I prefer to hear in your presence.

\ciceroSection{3} As to your urging me to write something --- the matter, as you say, grows on me; but the whole thing as yet is in flux: ``new wine of the autumn vintage'' \textit{[Greek: kat' op\=or\=en tryx]}. When it has settled, what I write will be more clarified. If you cannot at once carry it from me, you will at least have it first, and for some time alone.

\ciceroSection{4} You rightly love Dicaearchus; he is an excellent man, and not a little better citizen than those rulers-of-injustice \textit{[Greek: adikaiarchoi]} of ours. I wrote at the tenth hour at the Cerialia, as soon as I had read your letter; but I was going to give the letter, as I supposed, the day after to whoever first met me. Terentia is delighted by your letter; she sends you many greetings, ``and little Cicero''.
