% Ad Atticum 2.24 (ad-atticum-02-24) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{CICERO ATTICO salutem}

\ciceroSection{1} The letter which I gave to Numestius summoned you in such a way that nothing could be done sharper or more impetuous. To that swiftness add yet more, if you can. And do not be alarmed --- for I know you, and am not unaware how full of solicitude and anxiety all love is --- but the matter is, as I hope, not so troublesome in its outcome as in its approach.

\ciceroSection{2} That Vettius, that informer of ours, has --- as we see plainly --- promised Caesar that he would see to it that the younger Curio was brought under some suspicion of crime. Accordingly he wormed his way into the young man's intimacy, and after meeting with him often --- as the matter shows --- brought the affair to such a point that he declared he was resolved with his slaves to attack Pompey and kill him. Curio reported this to his father, and the father to Pompey. The matter was brought before the Senate. Vettius, brought in, at first denied that he had ever met with Curio at all; nor did he persist long, for he immediately demanded the public's pledge of safety. There was an outcry. Then he set forth that there had been a band of young men under Curio's leadership, in which Paulus had been at first, and Q. Caepio (this Brutus), and Lentulus, the flamen's son, with his father privy to it; afterwards that C. Septimius the scribe of Bibulus had brought him a dagger from Bibulus. The whole story was laughed at: as if Vettius could have lacked a dagger, had the consul not given him one. The story was the more flatly thrown out because on the third day before the Ides of May Bibulus had warned Pompey to look out for ambushes; for which warning Pompey had thanked him.

\ciceroSection{3} Young Curio, brought in, replied to what Vettius had said. And it was on this point especially that, in the Senate, Vettius was caught out: that he had said the young men's plan was to attack Pompey in the Forum among Gabinius's gladiators, and that the chief in this had been Paulus --- who, it was agreed, had at that time been in Macedonia. A decree of the Senate was passed: that Vettius, since he had confessed himself to have been in possession of a weapon, should be cast into chains; whoever should release him would act against the commonwealth. Opinion held that the design had been that Vettius should be arrested in the Forum with a dagger and his slaves likewise with weapons, and that he should then declare himself an informer. And this would have been done, had not the Curios reported the matter to Pompey beforehand. Then the senatorial decree was read out at a public meeting. But the next day Caesar --- the same who once, when he was praetor, had ordered Q. Catulus to speak from a lower place --- brought Vettius forward onto the Rostra and stood him in that place where Bibulus the consul was not allowed to look. There Vettius said everything he wished about the commonwealth, like a man who had come there made and trained for the purpose. First he removed Caepio from his speech --- the very man whom he had named most sharply in the Senate --- so that it appeared a night and nocturnal entreaty had intervened. Next he named men whom in the Senate he had not touched even with the slightest suspicion: L. Lucullus, by whom (he said) Gaius Fannius used to be sent to him, the same Fannius who had subscribed against P. Clodius; L. Domitius, whose house, he said, had been arranged as the place from which the outbreak was to be made. Me he did not name; but he said that an eloquent consular, the consul's neighbour, had said to him that some Servilius Ahala or Brutus must be found. He added at the end, when --- the assembly already dismissed --- he had been recalled by Vatinius, that he had heard from Curio that my son-in-law Piso and M. Laterensis were privy to these matters.

\ciceroSection{4} Now Vettius was a defendant before Crassus Dives on a charge of violence; and on being condemned was about to demand a hearing as informer. If he had obtained it, it seemed there would have been further trials. We did not despise these things --- for we never despise anything --- but we were not afraid of them either. The highest enthusiasms of men toward us were being signified; but life is utterly tedious to me, so full are all things of every miseries. Lately we had been afraid of slaughter, and a speech of that bravest of old men, Q. Considius, had dispelled it; what we were able to fear daily \textit{[the text is corrupt here]} suddenly came up. Need I say more? Nothing is more unfortunate than I, nothing more fortunate than Catulus, both for the splendour of his life and for this present moment. Yet in these miseries we are of upright mind and minimally disturbed, and we are guarding our dignity most honourably and with great care.

\ciceroSection{5} Pompey bids us be free of care about Clodius, and signifies in every speech the highest goodwill toward me. I want to have you as the author of my counsels, the partner of my anxieties, joined to me in every thought. Therefore, as I instructed Numestius to deal with you, so, and even more sharply if I can, I beg you to fly plainly to us. I shall breathe again if I see you.
