% Ad Atticum 2.2 (ad-atticum-02-02) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} Take care, please, of our little Cicero. To him we seem to be uncles \textit{[Greek: theioi]}.

\ciceroSection{2} I had in my hands the work \textit{On the Pellenians} \textit{[Greek: Pellenai\=on]}, and, by Hercules, I had built up a great heap of Dicaearchus before my feet. O what a great man, and one from whom you may learn far more than from Procilius! The work \textit{On the Corinthians} \textit{[Greek: Korinthi\=on]} and \textit{On the Athenians} \textit{[Greek: Ath\=enai\=on]} I think I have at Rome. Believe me [text corrupt] \ldots he is a wonderful man. Herodes, if he were a man, would read him rather than write a single letter himself --- the man who has assailed me in a letter has, I see, come at you in close combat. I had rather have conspired than have resisted the conspiracy, if I had thought I had to listen to him.

\ciceroSection{3} About the darnel you are not sane; about the wine I praise you. But hold --- do you not see the Kalends coming, Antonius not coming? the judges being summoned? For they send word to me that Nigidius is threatening in a public meeting that he will haul off any judge who has not been present. But please, if there is anything you have heard about Antonius's arrival, write to me; and, since you are not coming here, dine at our house at any rate on the day before the Kalends. Take care not to do otherwise. See that you keep well.
