% Ad Atticum 1.20 (ad-atticum-01-20) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} When I had got back to Rome from the Pompeian villa on the fourth day before the Ides of May, our Cincius delivered me that letter from you which you had given on the Ides of February. To that letter I shall now answer with this. First, I am glad that my judgment of you is clear to you. Next, that you have been most moderate in those things which seemed to me to have been managed somewhat sharply and unpleasantly by us and by our people, I rejoice vehemently, and judge it not the mark of a moderate love and of the highest spirit and wisdom. As to which, since you have written to me so sweetly, diligently, dutifully, humanely, that not only ought I not to exhort you further but could not have expected from you, or from any man, so great a measure of facility and gentleness, I count nothing more convenient than to write nothing more about these matters. When we have met, then, if anything in the matter requires it, we shall confer face to face.

\ciceroSection{2} As to what you write to me about the commonwealth, you argue lovingly and prudently, and your reasoning does not differ from my counsels. For we must not give up the standing of our dignity, nor must we go inside another's defences without our own forces. And he of whom you write has nothing ample, nothing exalted, nothing not low-cast and popular. But yet a calculation has been to me, perhaps, not unprofitable for the tranquillity of my own times --- and, by Hercules, much more useful even to the commonwealth than to me --- that the attacks of the wicked citizens upon me have been beaten back, when I had confirmed the wavering opinion of a man of the highest fortune, authority, favour, and turned him from the hope of the wicked to the praise of my own deeds. And if this had to have been done by me with some lightness, no thing would I have judged so important. But yet all has been so done by me that not I, by agreeing with him, should seem the lighter, but he, by approving me, the weightier.

\ciceroSection{3} The rest is being and shall be so done by me as not to give the appearance that what I did, I did by chance. The good men of mine, those whom you point out, and that Sparta \textit{[Greek: Spartan]} which has fallen, you say, to my lot, I shall not only never desert; even if I were deserted by it, yet I shall remain in my old opinion. But this I should like you to consider: that I hold this aristocratic course, since the death of Catulus, with no defence and with no following. For, as Rhinto says (I think): ``some are nothing, and to others nothing matters'' \textit{[Greek: hoi men par' ouden eisi, tois d' ouden melei]}. As for the way our fish-pond men envy me, I shall either write to you on another occasion or save it for our meeting. From the senate-house, no thing will tear me away --- both because it is so right, and because it is most consonant with my circumstances, and because, of how much the senate values me, I have no regret.

\ciceroSection{4} About the Sicyonians, as I wrote to you before, there is not much hope in the senate; for there is no one who would complain. So if you wait for that, it is a long road. Try another way, if you can. When the matter was acted on, no one paid attention to whom it pertained, and the back-bench senators ran in haste to that opinion. The time for setting aside the senate's decree is not yet ripe, since neither are there men who would complain, nor are many of them, partly out of malice, partly from an opinion of equity, otherwise than pleased.

\ciceroSection{5} Your Metellus is an outstanding consul. One thing I rebuke: that he is not greatly delighted that peace is reported from Gaul. He wishes, I take it, to triumph. I should wish that wish were more moderate; the rest is outstanding. The son of Aulus, on the other hand, conducts himself in such a way that his consulship is not a consulship but the black eye \textit{[Greek: hyp\=opion]} of our Magnus.

\ciceroSection{6} Of my writings, I have sent you my consulship completed in Greek. I have given the book to Lucius Cossinius. I think you take pleasure in my Latin works, but as a Greek you are envious of this Greek one. If others write, I shall send them to you; but believe me, as soon as they read this of ours, somehow they slow down.

\ciceroSection{7} Now, to come back to my own matter: Lucius Papirius Paetus, a good man and a lover of ours, has given me those books which Servius Claudius left behind. Since your friend Cincius said it was permitted to me, by the Cincian law, to take them, I gladly said I would accept them if he brought them. Now, if you love me, if you know yourself loved by me, strive through friends, clients, guests, finally freedmen and slaves of yours, that not a scrap perish. For both those Greek books which I suspect, and the Latin which I know he left, I greatly need. I, indeed, daily the more rest in those studies in proportion as time is given me from the forensic labour. You will do me a great, a very great favour, if you are as diligent in this as you are wont to be in those things which you suppose I greatly wish; and I commend to you the business of Paetus himself, on which he gives you the highest thanks; and that you may now visit us I not only ask but urge.
