% Ad Atticum 1.19 (ad-atticum-01-19) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} Not only if I had so much leisure as you have, but even if I wished to send letters as short as you are wont to, I should easily surpass you and be much more frequent in writing than you. But added to my supreme and incredible occupations is this, that I do not wish a single letter from me to come to you without subject and substance. And first --- as is fair to a citizen who loves his country --- I shall set out for you what is happening in the commonwealth; then, since you are nearest to me in love, I shall write also of myself those things that I take it you would not be unwilling to know.

\ciceroSection{2} In the commonwealth at present the chief thing is the fear of a Gallic war. For the Aedui, our brothers, have lately fought a bad fight, and the Helvetii without a doubt are in arms and are making raids into the province. The senate has decreed that the consuls should draw lots for the two Gauls, that a levy be held, that exemptions should not stand, that legates with full powers should be sent to visit the states of Gaul and to take pains that they not join with the Helvetii. The legates are Quintus Metellus Creticus, Lucius Flaccus, and --- the joke of it --- Lentulus, son of Clodianus.

\ciceroSection{3} And here I cannot pass over this: that, when of the consulars my own lot had come out first, with one voice the full senate decreed I should be kept in the city. The same thing happened to Pompey after me, so that we two seemed to be retained as pledges of the commonwealth. What other men's exclamations \textit{[Greek: epiph\=on\=emata]} should I look for, when these are born at home?

\ciceroSection{4} City matters stand thus. The agrarian law of the tribune Flavius was being vehemently agitated, with Pompey backing it; which had nothing popular but its backer. From this law I, with the assembly's full goodwill, was striking out everything which pertained to the harm of private men: I freed that land which had been public in the consulship of P. Mucius and L. Calpurnius; I confirmed the holdings of the Sullans; I kept the men of Volaterrae and Arretium, whose land Sulla had publicized but had not divided, in their own possession. One reasoning I did not reject: that the land be bought with this fresh money, which over the next five years would come in from the new revenues. To this whole agrarian scheme the senate was opposed, suspecting that some new power was being sought for Pompey. But Pompey had set his heart on the wish to carry the law. I, however, with great gratitude from the agrarian party, was confirming all private men's holdings (for that, as you know, is my own army --- of men of substance), and was satisfying the people, and Pompey too (for I wished that as well), by the purchase scheme; once that was set up, I judged that both the bilge of the city could be carefully drained off, and the wilderness of Italy could be peopled. But this whole matter, interrupted by the war, has cooled. Metellus is a quite good consul, and loves us greatly; the other one is so much nothing that he plainly does not know what he has bought.

\ciceroSection{5} These are the matters in the commonwealth, unless you reckon this also as bearing on it: that a certain Herennius, tribune of the plebs, your fellow tribesman, a wholly worthless and needy fellow, has often now begun to act on transferring Publius Clodius to the plebs. Frequent vetoes are interposed against him. These are, I take it, the public matters.

\ciceroSection{6} As for me: from the moment I attained, on the famous Nones of December, that exceptional and immortal glory joined with envy and the enmity of many, I have not ceased to engage in the commonwealth with the same greatness of mind, and to protect the dignity I had set up and undertaken. But after I first perceived, in Clodius's acquittal, the levity and feebleness of the courts; then saw that our publicans were easily cut off from the senate (although from me at least they were not torn away); and finally saw that the rich men --- those fish-pond owners, I mean, friends of yours --- not obscurely envied us, I judged that I must seek some greater resources and firmer defences.

\ciceroSection{7} So first, I brought Pompey, who had too long kept silent about my affairs, into such a disposition that in the senate not once but often, in many words, he assigned to me the safety of this empire and of the world; which mattered not so much to me (for those things are neither so obscure as to need testimony nor so doubtful as to need an encomium) as to the commonwealth, since there were certain wicked men who supposed that there would be some contention between me and Pompey from the disagreement on those matters. With him I have joined myself in such close friendship that each of us, by this joining, can be more fortified in his own reckoning and steadier in the commonwealth.

\ciceroSection{8} The hatred of that lustful and dainty youth which had been kindled against me has been so softened by a certain courtesy of mine that they all cultivate me alone. Nothing harsh now is done by me against any man, nor anything popular and dissolute either; but the whole reckoning is so tempered that I show steadfastness in the commonwealth, while in my private affairs --- on account of the weakness of the good men, the unfairness of the malevolent, the hatred of the wicked toward me --- I apply a certain caution and diligence; and so, although I am wrapped up in these new friendships, that crafty Sicilian Epicharmus often whispers to me his old refrain: ``Stay sober, and remember to mistrust: these are the joints of the mind'' \textit{[Greek: naphe kai memnas' apistein: arthra tauta tan phren\=on]}. And of our reasoning and our way of life you see, I think, a sort of pattern.

\ciceroSection{9} About your own business, you write to me often. To it I cannot now apply remedy; for that senate's decree was made by the highest goodwill of the back-benchers, by no man's authority of ours. For that I am there to write of it, you can understand from the senate's decree itself: another matter was at that time the topic; this about the free peoples was added without occasion, and so was done by P. Servilius the son, who spoke last in the order. But it cannot at this time be changed. So the assemblies which were at first called have now long ceased to be held. If by your blandishments you have nevertheless squeezed any little money out of the Sicyonians, please let me know.

\ciceroSection{10} I have sent you the memoir of my consulship composed in Greek. In which, if there shall be anything that seems to a man of Athens too little Greek and too little learned, I shall not say what Lucullus, I take it, said to you at Panhormus about his own histories: that, the better to make them pass for the work of a Roman, he had on that account sprinkled in some barbarisms and ``solecisms,'' \textit{[Greek: soloika]}. With me, if there is anything of that kind, it is unwitting and unwilling. The Latin version, when I have finished it, I shall send you. Expect a third poem also, lest any genre of my own praise be passed up by myself. And here take care not to say: ``Who shall praise the father?'' \textit{[Greek: tis patera ain\=esei]}. For if there is anything among men that is more worthy, let it be praised, and let me be reproached for not praising other things instead. But these things which I write are not encomia \textit{[Greek: enk\=omiastika]} but histories \textit{[Greek: historika]}.

\ciceroSection{11} My brother Quintus clears himself with me by letter and affirms that nothing has been said by him to anyone against you. But these things must be handled by us face to face with the highest care and diligence; only come back to us at last. This Cossinius, to whom I have given the letter, seemed to me a thoroughly good man and not a light one, fond of you, and such as your letters had reported him to me. The Ides of March.
