% Ad Atticum 1.18 (ad-atticum-01-18) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} Know that nothing is now so wanting to me as a man with whom I might share, for ever, all the things that affect me with any care --- a man who would love me, who would have wisdom; with whom, when I speak, I would feign nothing, dissemble nothing, hide nothing. For my brother, the most simple and most loving, is away. ``Metellus'' is no man but a shore, and air, and sheer solitude. As for you, who most often have lightened the cares and anguish of my mind by your speech and counsel, who are wont to be both my partner in public affairs and my confidant in all private ones and the sharer of all my conversations and counsels --- where on earth are you? So forsaken am I by all that I have only as much rest as is spent with my wife and little daughter and honey-sweet Cicero. For those ambitious and painted friendships of mine are kept up in some forensic splendour: they have no domestic fruit. So when, in the morning, my house is well filled, when we go down to the Forum hedged about with bands of friends, I cannot find in the great crowd anyone with whom I might either jest freely or sigh familiarly. So we look for you, miss you, even now summon you. For there are many things that vex and trouble me, which, having gained your ear, I think I could draw out in the talk of a single walk together.

\ciceroSection{2} The stings of my domestic anxieties --- their thorns and shards I shall hide; nor shall I commit them to this letter and to an unknown carrier. And these (for I do not want you stirred up) are not very heavy; but they sit and press, and rest neither in the counsel nor in the speech of any loving man. As to the commonwealth, although my mind stands firm, yet \ldots [the cure itself works on it]. For, to put briefly all the things that have been done since your departure, you must by now exclaim that Roman matters can stand no longer. For after your setting out, the first entrance, I take it, was into the case of the Clodian play, in which I, having found, as I supposed, an opportunity for cutting back lust and bridling youth, was vehement, and poured out all the powers of my mind and talent --- not driven by hatred of any man, but by the hope of correcting and healing the state.

\ciceroSection{3} The commonwealth is shattered by the bought and corrupted trial. See what has followed since. There has been imposed on us, as consul, the man whom no one but us philosophers can look upon without a sigh. How great a wound that is! With a senate's decree about bribery, no law has been carried; the senate has been agitated, the Roman equestrians alienated. So that one year overturned the two foundations of the commonwealth which I alone had set up: it threw away the senate's authority, and broke the concord of the orders. Now the present year, that ``outstanding'' year, is upon us. Its beginning was such that the annual rites of Iuventas were not duly performed; for Memmius initiated Marcus Lucullus's wife into rites of his own. Menelaus, taking it ill, divorced her. Although that Idaean shepherd had despised only Menelaus, this our Paris held neither Menelaus nor Agamemnon to be free.

\ciceroSection{4} There is a certain Gaius Herennius, tribune of the plebs, whom you may not even know --- although you can know, for he is your fellow tribesman and his father Sextus used to distribute money to your tribe. He is transferring Publius Clodius to the plebs, and bringing forward that the entire people may vote on Clodius's matter at the Field of Mars. I took him in hand in the senate, as is my custom; but there is nothing slower than that man.

\ciceroSection{5} Metellus is an outstanding consul and loves us; but he diminishes his own authority by having promulgated the same business about Clodius for form's sake. As for the son of Aulus --- O immortal gods! what a coward and spineless soldier! how worthy of, as he in fact does, daily offering his face to be vilified by Palicanus.

\ciceroSection{6} An agrarian law has been promulgated by Flavius, slight enough, almost the same as Plotius's was. But meanwhile, can a public man \textit{[Greek: politikos an\=er]} be found anywhere? He who could --- our friend (and so it is, I want you to know this), Pompey --- guards in silence that little painted toga of his. Crassus says no word against the popular favour. The rest you know now: who are so foolish as to seem to hope that, with the commonwealth lost, their fish-ponds will be safe.

\ciceroSection{7} There is one who cares, by steadiness and integrity rather, as it seems to me, than by counsel or talent: Cato. Who is now harassing the wretched publicans, whom he had as his closest friends, for the third month in a row, and does not allow an answer to be given them by the senate. So we are forced to decree nothing about anything else until the publicans have had their answer. So embassies, I think, will also have to be put off.

\ciceroSection{8} Now you see by what waves we are tossed; and if from what we have written you understand also so much that we have not written, come back to us at last. And though these are things to flee from, to which I summon you, yet make sure that you so value our love as to wish to come, even with these troubles. For that you may not be assessed in your absence, I shall see that an edict is issued and posted in every place; but to be assessed at the very lustration is the proper business of a born tradesman. So see that we may see you as soon as possible. Farewell. Eleven days before the Kalends of February, in the consulship of Quintus Metellus and Lucius Afranius.
