% Ad Atticum 1.12 (ad-atticum-01-12) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} That Teucris affair is quite a slow business, and Cornelius did not come back to Terentia afterwards. I think we shall have to flee to Considius, Axius, Selicius; for from Caecilius our connections cannot raise a sesterce at less than twelve percent. But to come back to that first matter, I have seen nothing more impudent, more cunning, more dilatory than that woman. I send a freedman, I have commissioned Titus. Pretexts \textit{[Greek: sk\=epseis]} and put-offs \textit{[Greek: anabolai]} --- but I do not know whether it is mere chance \textit{[Greek: tautomaton]}. For Pompey's advance-runners are openly announcing that Pompey will move to have Antonius succeeded; and at the same time the praetor will move to the people. The case is of such a kind that I cannot, by either the good men's or the popular party's reputation, honourably defend the man, nor do I wish to --- which is the chief thing. For something has happened of which the whole I commission to you to look into.

\ciceroSection{2} I have a freedman, a sad rascal of a fellow, I mean Hilarus, the accountant and a client of yours. About him Valerius the interpreter informs me, and Thyillus writes that he has heard, this: that the man is with Antonius; that Antonius further keeps saying, in collecting his money, that a share is being sought for me, and that the freedman has been sent by me as guardian of our common gain. I am not slightly disturbed, nor yet have I believed it; but certainly there was some such talk. Investigate the whole, learn, look into it, and that scoundrel, if you can in any way, remove him from those parts. Valerius named Cn. Plancius as the source of this talk. I commission you simply with the whole, that you may see of what kind it is.

\ciceroSection{3} Pompey is, all are agreed, most friendly to us. Mucia's divorce is vehemently approved. As for Publius Clodius son of Appius, I take it you have heard he was caught in women's dress at Gaius Caesar's house when the rite \textit{pro populo} was being held, and was saved and led out through the hands of a maidservant; the matter is of marked notoriety. That you bear this with grief, I am sure.

\ciceroSection{4} I have nothing else to write to you; and indeed, by Hercules, I was in writing the more disturbed. For a delightful boy, our reader Sositheus, had died, and had moved me more than a slave's death seemed to call for. Please write to me often. If you have no news, write whatever first comes to mind. The Kalends of January, in the consulship of Marcus Messalla and Marcus Piso.
