% Ad Atticum 1.13 (ad-atticum-01-13) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} I have now received three letters of yours: one from Marcus Cornelius, which you gave him at Tres Tabernae, I take it; another which your host of Canusium delivered me; the third which, as you write, you handed over with the anchor weighed, from your skiff. All of them, as the rhetoricians' boys say, were both sprinkled with the salt of humanity and marked by the signs of love. By these letters I have indeed been provoked to write back, but I am the slower because I cannot find a faithful letter-carrier. For how few are there who can carry a letter of any weight without lightening it by reading it through? Add to this that nobody is now setting out for Epirus. For I take it that you, when the victims have been killed at your Amalthea, have at once set out for the besieging of Sicyon. But I do not know with any certainty when you set out for Antonius, or how much time you spend in Epirus. So I dare commit a letter of any frankness neither to Achaean nor to Epirote travellers.

\ciceroSection{2} Things have happened, since you went away from me, that deserve a letter; but they are not to be entrusted to such a risk that they could perish, or be opened, or be intercepted. First, then, know this: I have not been the first asked for my opinion in the senate; the pacificator of the Allobroges has been put before me, and that with the senate murmuring and not against my will. For I am free of having to humour a perverse man, and unbound, on the score of my own dignity in the commonwealth, from his goodwill. The second place in speaking holds the authority of nearly the first, and a goodwill not too much chained by a consul's favour. The third is Catulus; the fourth, if you ask, Hortensius. The consul himself is small-spirited and crooked, but a sneerer in that sour way which is laughed at even without wit, ridiculous more by face than by joke, doing nothing for the commonwealth, cut off from the best men --- a man from whom you would hope nothing good for the commonwealth because he does not wish it, nothing bad because he does not dare it. His colleague, however, is most respectful to me, and a zealous defender of the better party.

\ciceroSection{3} The two now slightly differ between themselves. But I fear that this thing left undone may creep further. For I take it you have heard that, at the rite \textit{pro populo} held at Caesar's house, a man came in dressed in women's clothing; and that, when the Vestals had repeated the sacrifice, mention was made in the senate by Quintus Cornificius (he, that you may not chance to think it was one of us, was the first); afterwards the matter, by senate's decree, was referred to the Vestals and the pontiffs, who declared it sacrilege; then, by senate's decree, the consuls promulgated a bill; Caesar has sent his wife notice of divorce. In this case Piso, drawn by his friendship with Publius Clodius, gives his pains that the bill which he himself proposes, and proposes by senate's decree and on a religious matter, should be repudiated. Messalla so far is acting vehemently and severely. The good men are being drawn off from the case by Clodius's prayers; gangs are being got together; we ourselves, who from the start had been Lycurgean, are growing milder by the day; Cato presses on and urges. What more can I say? I fear that this matter, neglected by good men and defended by the wicked, may be the cause of great evils to the commonwealth.

\ciceroSection{4} As for that friend of yours --- you know whom I mean --- of whom you wrote to me that, when he no longer dared to find fault, he had begun to praise: us, as he shows, he is mighty fond of, embraces, loves, openly praises, secretly --- but so that the thing is plain --- envies. Nothing courteous, nothing simple, nothing in public affairs distinguished, \textit{[Greek: en tois politikois]}, nothing honourable, nothing brave, nothing free. But of this I will write to you another time more closely; for I do not yet know it sufficiently, and I dare not commit a letter on such great matters to some son of earth or other.

\ciceroSection{5} The praetors have not yet drawn lots for their provinces. The matter is in the same place where you left it. The topographical sketch \textit{[Greek: topothesian]} of Misenum and Puteoli that you ask for, I shall include in my speech. ``On the third day before the Nones of December'' --- I had noticed that there was a slip there. The things which you praise from the speeches, believe me, were greatly pleasing to me, but I did not dare to say so before; now indeed, since they have been approved by you, they seem to me much more in the Attic style \textit{[Greek: Attik\=otera]}. To that Metelline speech I have added some things. The book will be sent to you, since love of us has made you a lover of rhetoric \textit{[Greek: philorh\=etora]}.

\ciceroSection{6} Of news, what shall I write you? What? this. Messalla the consul has bought Autronius's house for HS 13,400,000 (?). ``What is that to me?'' you will say. So much that, by that purchase, we ourselves have been judged to have bought well, and men have begun to understand that it is permitted, with friends' resources, to attain in a purchase to some position. The Teucris affair is a slow business, but yet it is in hope. Settle that for us. From me expect a freer letter. Six days before the Kalends of February, in the consulship of Marcus Messalla and Marcus Piso.
