% Ad Atticum 1.17 (ad-atticum-01-17) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} A great variety of feeling and an unlikeness of opinion and judgment in my brother Quintus has been shown to me from the letters in which you sent me copies of his letters. By which thing I have been affected both with such pain as my supreme love toward each of you must have brought me, and with wonder at what could have happened to bring upon my brother Quintus either so heavy an offence or so great a change of feeling. And this much I had already long understood --- which I saw you also yourself, when you were leaving us, suspect --- that some uncomfortable opinion lay underneath, and that his mind was wounded, and that some odious suspicions had settled in. To which, when I was eager to apply remedy --- often before, and yet more vehemently after the lots were drawn for the province --- I neither perceived him to be so much offended as your letters declared, nor was getting on as much as I wished.

\ciceroSection{2} But yet I consoled myself with this: that I did not doubt that he would see you either at Dyrrachium or somewhere in those parts. And when that should happen, I trusted, and had persuaded myself, that everything would be made peaceable between you, not only by speech and discussion but by the very sight and meeting of you. For of how great courtesy my brother Quintus is, of how great pleasantness, how mild a spirit both for receiving and laying down offence, it is not for me to write to you, who know it. But by sad mischance you saw him nowhere. For more weight had what had been driven into his head by the artifices of certain men than either duty or connection or that old love between you, which ought to have weighed most.

\ciceroSection{3} And where the fault of this trouble lies I can more easily judge than write; for I fear that, while I defend my own people, I shall not spare yours. For I see this much: that, whether or no any wound had been made by his household, that they certainly could have healed it. But the fault of this whole matter, which extends somewhat further than it seems, I shall set out for you more conveniently when you are present.

\ciceroSection{4} About those letters which he sent you from Thessalonica, and about the conversations which you suppose he had with him, both at Rome among your friends and on the road, what cause there can be I do not know; but all my hope of relieving this trouble is set on your kindness. For if you so resolve --- both that the minds of even the best men are often easily provoked and again easily appeased, and that this nimbleness, so to speak, and softness of nature is most often a sign of goodness, and (the head of the matter) that we, among ourselves, must bear with our discomforts or our faults or our injuries to one another --- these things, as I hope, will easily be softened. And I beg you to do this. For it concerns me, who love you above all, most of all that there be no one of mine who either does not love you or is not loved by you.

\ciceroSection{5} That part of your letter was least necessary in which you set out what occasions of either provincial or urban advantages you have passed up, both at other times and in my own consulship. For your nobility and greatness of mind have been clear to me; nor have I ever reckoned that anything stood between me and you except the choice of the way of life laid out --- since a certain ambition led me to the pursuit of honours, but another reasoning, by no means to be rebuked, led you to honourable leisure. In true praise of probity, of diligence, of scrupulousness, I do not put myself before you, nor anyone; but in love toward me, leaving aside love of brother and of household, I give you the first place.

\ciceroSection{6} For I have seen, and seen deeply, in my various crises, both your anxieties and your joys. To me, often, your congratulation on my praise was pleasing, your consolation on my fear welcome. Indeed now, in your absence, not only does the counsel in which you excel fail me most, but the daily exchange of speech --- which is wont to be the sweetest of all things between us --- is missed in the highest degree. What shall I say? In public matters, in which I cannot be negligent? Or in forensic labour, which I had borne before through ambition and now bear that I may by my favour protect my dignity? Or in those domestic concerns themselves, in which I, both before and especially after my brother's departure, miss you and our daily talk? In short, my labour, my rest, my business, my leisure, the matters of the Forum, those of home, the public, the private --- none of these can do without your sweetest and most loving counsel and conversation any longer.

\ciceroSection{7} And to recall these things has often been hindered by a modesty in each of us; but it has now become necessary on account of that part of your letter in which you wished both yourself and your character to be cleared and approved before me. And in that trouble of his alienated and offended mind there is yet this much of comfort: that, both to me and to your other friends, your wish to give up the province had been known and testified by you some good while in advance, so that what is being done in your not being together may seem to be brought about not by your dissent and breach but by your wish and judgment. So both those bonds which were violated will be expiated, and these bonds of ours, which have been most religiously kept, will hold to their reverence.

\ciceroSection{8} As for us here, we are involved in a weak, miserable, changeable commonwealth. For I take it you have heard that our equestrians are nearly cut off from the senate. Who first bore very gravely that there had been promulgated by senate's decree that an inquiry be made about those who had taken bribes for judging. In the decreeing of which, when I had by chance not been there and had felt that the equestrian order took it ill though did not say so openly, I rebuked the senate, with (as it seemed to me) supreme authority, and was, in a not very becoming case, very weighty and copious.

\ciceroSection{9} Behold another delicacy of the equestrians, scarcely to be borne; which I not only bore but adorned. Those who from the censors took the contract for Asia complained in the senate that they had been carried away by greed and had taken the contract too high; they asked that the contract be set aside. I was leading among their helpers, or rather second; for to make them dare to ask this, Crassus pushed them. An invidious matter, a base demand, a confession of rashness. The supreme danger was that, if they obtained nothing, they would be plainly alienated from the senate. Here too the matter has been most aided by us, and brought it about that they had a most full and most generous senate to deal with; and many things were said by me on the dignity and concord of the orders, on the Kalends of December and the day after. Nor is the matter yet finished, but the senate's wish is clear; for one alone had spoken against, Metellus, the consul-elect; and there was about to speak, but on account of the day's brevity it was not reached, that hero of ours, Cato.

\ciceroSection{10} So, keeping our reasoning and our institution, I protect, as I can, that concord which was glued together by me. But yet, since these things are so weak, a certain safe road, as I hope, is being fortified for us toward keeping our resources --- which I cannot to you in a letter sufficiently set out, but I shall yet show by a small token. I have a most familiar use of Pompey. I see what you would say. I shall guard against what is to be guarded against, and shall write to you on another occasion more about my counsels in handling the commonwealth.

\ciceroSection{11} Know that Lucceius has it in mind to stand for the consulship at once. For only two are said to be standing: Caesar thinks of joining with him through Arrius, and Bibulus thinks he can be joined with this man through Gaius Piso. You laugh? These are not laughable, believe me. What else shall I write to you? What? Many things, but on another occasion. Since you wish, please make me know. Now this much I modestly ask, what most I desire: that you come as soon as possible. The Nones of December.
