% Ad Atticum 1.16 (ad-atticum-01-16) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} You ask of me what happened about the trial that turned out so contrary to all men's expectation, and at the same time you wish to know how I fought less than I am wont. I shall answer you ``later first,'' \textit{[Greek: hysteron proteron]}. As long as the senate's authority had to be defended by me, so keenly and so vehemently did I fight that the shouts and the gathering crowds were made with the highest praise to me. If I have ever seemed to you brave in the commonwealth, certainly in that case you would have wondered at me. For when he had fled to public meetings and was using my name in them to stir up odium --- immortal gods! what battles I gave, what slaughters! what onsets against Piso, against Curio, against that whole gang I made! how I assailed the levity of old men, the lust of the young! Often, so help me the gods, I missed you not only as adviser to my counsels but as spectator of my wonderful battles.

\ciceroSection{2} But after Hortensius hit upon the plan that the law about the sacrilege should be carried by Fufius, tribune of the plebs, in which there was nothing different from the consular bill except in the manner of the judges (and in that, all things were) --- and fought to bring it so about, having persuaded both himself and others that the man could escape no judges --- I drew in my sails, perceiving the poverty of the judges, and said nothing in evidence except what was so well known and so attested that I could not pass it over. So if you ask the cause of the acquittal --- to come back now to ``the earlier'' \textit{[Greek: pros to proteron]} --- it was the destitution of the judges and their disgrace. That this should happen was committed by Hortensius's counsel, who, while he feared that Fufius would interpose his veto against the bill that was being carried by senate's decree, did not see what was the better course: to leave the man in his infamy and disgrace rather than to commit him to a feeble trial. Drawn by hatred, he hurried to bring the matter to trial, saying yet that the man would have his throat cut by a leaden sword.

\ciceroSection{3} But if you ask what kind of trial it was, by an incredible outcome --- so that now, from the event, by others, but by me from the very beginning, Hortensius's counsel is rebuked. For when the rejections were made amid the greatest shouts, the prosecutor like a good censor rejecting the most worthless men, the defendant like a kindly trainer setting aside every most respectable one, as soon as the judges had taken their seats good men began to lose hope. For never was there a fouler bench in a gambling-school: spotted senators, naked equites, tribunes not so much ``of the treasury'' (\textit{aerati}) as, by their proper name, ``of the public account'' (\textit{aerarii}, beggars). A few good men nevertheless were among them whom Clodius had not been able to drive off by rejection, who sat sad among their unlikes and grieving, and were vehemently disturbed by the contagion of the disgrace.

\ciceroSection{4} Here, as each matter was referred to the bench in the first applications, the severity was incredible, with no variety of opinions. The defendant got nothing; more was given to the prosecutor than he asked. Hortensius was triumphing, in short, that he had seen so well; there was no one who did not consider the man a defendant and a thousand times condemned. But when I was produced as witness, you have heard, I take it, from the shouts of Clodius's supporters, what a rising of the judges was made --- how they stood about me, how they openly offered their throats to Publius Clodius for my life. Which thing seemed to me far more honourable than either that, when your fellow-citizens forbade Xenocrates to take an oath when giving evidence, or that, when our judges did not wish to look at the books of Metellus Numidicus when, as is the custom, they were being passed around.

\ciceroSection{5} This our matter was, I say, far the greater. So by the voices of the judges --- when I was so defended by them as the country's safety --- the defendant was broken, and with him all his patrons fell. To me, however, on the next day, the same crowd gathered with which I had been led home in leaving my consulship. The distinguished Areopagites cried out that they would not come if a guard were not appointed. The matter was referred to the bench. Only one vote did not require a guard. The matter was carried to the senate. Most gravely and most splendidly was it decreed; the judges were praised; the business was given to the magistrates. No one supposed that the man would respond. ``Tell me now, Muses, how the fire first fell.'' \textit{[Greek: Espete nyn moi, Mousai, hopp\=os d\=e pr\=oton pyr empese]} (Iliad 16.112). You know that Calvus of the Nanneian estate, that praiser of mine, of whose speech in honour of me I had written to you. In two days, through one slave, and that one out of a gladiator's school, he settled the whole business: he summoned them to his house, he promised, he interposed, he gave. Now indeed --- good gods, what a ruined matter! --- nights with certain women, and introductions of noble youths, were given for the heap of the wage to not a few judges. So with the highest withdrawal of good men, with a Forum full of slaves, twenty-five judges yet were so brave as, at the highest danger laid on them, rather to wish to perish than to ruin everything. There were thirty-one whom hunger more than fame moved. Of whom Catulus, when he had seen one, said: ``What were you asking from us a guard for? Were you afraid that the money would be snatched away from you?''

\ciceroSection{6} You have, as briefly as I could, the kind of trial it was and the cause of the acquittal. You ask next what is now the state of affairs and what is mine. That state of the commonwealth which you, by my counsel --- I, by divine help --- thought confirmed, which seemed fixed and founded by the joining and authority of all good men in my consulship --- unless some god looks back upon us, know that it has slipped from our hands, by this one trial alone, if it is a trial when thirty most worthless and basest men of the Roman people, having received their bits of money, wipe out all law and right, and what all not only men but cattle know to have been done, that Thalna and Plautus and Spongia and the rest of the rubbish of this kind decree never to have been done.

\ciceroSection{7} But yet, that I may console you about the commonwealth: not, as the wicked hoped, when so great a wound had been laid on the commonwealth, does wickedness exult eagerly in its victory. For they thought, plainly, that, when religion had fallen, when chastity had fallen, when the trustworthiness of trials, when the senate's authority, the wickedness and lust openly victorious would seek punishments from every best man for their own grief which the severity of my consulship had branded on every most wicked man.

\ciceroSection{8} I myself, the same man (for I do not seem to myself to be insolently boasting when I speak of myself with you, especially in a letter which I do not wish read by others) --- I, the same, recovered the cast-down spirits of the good, confirming and rousing each one. By insulting and goading the moneyed judges, I have snatched all freedom of speech, \textit{[Greek: parr\=esian]}, from all the studious adherents and partisans of that victory; the consul Piso I have not allowed to take a stand on any matter; Syria, betrothed to him already, I have stripped from him; the senate I have called back to its old severity, and have stirred up from its dejection. Clodius, in his presence, I have broken in the senate --- both by an unbroken speech most full of gravity, and by an altercation of this kind, of which it is permitted you to taste a little: for the rest cannot have the same force or charm with that zeal of contention, which you call the contest \textit{[Greek: ag\=ona]}, removed.

\ciceroSection{9} For when on the Ides of May we had assembled in the senate, I, asked for my opinion, said many things on the supreme matters of the commonwealth, and that passage was brought in by me from divine help: that, when one blow had been received, the senators should not fall, should not give up; that the wound was of such a kind that to me it seemed neither to be dissimulated nor to be too greatly feared, lest by ignoring it we should be judged the most foolish, or by fearing it the most cowardly; twice had Lentulus been acquitted, twice Catiline; this man, the third now, has been let loose by the judges upon the commonwealth. ``You are wrong, Clodius: not for the city did the judges save you, but for the prison; not to keep you in the state did they wish, but to deprive you of exile. Wherefore, members of the senate, raise up your spirits, hold fast to your dignity. That consensus of good men remains in the commonwealth; grief has come to good men, virtue is not diminished; no new harm has been done, but what was already there has been found out. In the trial of one ruined man, more men like him have been discovered.''

\ciceroSection{10} But what am I doing? I have all but locked up a speech in a letter. I come back to the altercation. The pretty little boy stood up and threw at me that I had been at Baiae. False, but in any case --- what of it? ``It is as if you should say,'' I said, ``that I had been at the Mysteries.'' --- ``What has a man of Arpinum,'' he said, ``to do with hot waters?'' --- ``Tell that,'' I said, ``to your patron, who coveted the waters of an Arpine'' (you know I mean Marius's). --- ``How long,'' he said, ``shall we put up with this man as king?'' --- ``You call me king,'' I said, ``when Rex made no mention of you?'' (he had eaten up the inheritance of Rex in hope) --- ``You bought a house,'' he said. ``You would think,'' I said, ``he was saying, `you bought judges'.'' --- ``They did not believe you,'' he said, ``when you swore.'' --- ``Twenty-five judges believed me,'' I said; ``thirty-one, since they took the money first, did not believe you with anything.'' Struck down by great shouts, he fell silent and broke off.

\ciceroSection{11} As to my standing, this is it. With the good, I am the same as when you left me. With the city's dregs and lees, things are now far better than when you left me. For it does not even hurt me that my testimony seems to have had no weight. The blood has been let off the inflamed odium, without pain --- and even more so because all those abettors of the disgrace confess that the manifest matter was bought off from the judges. Add to this that the assembly-leech of the treasury, the wretched and starving plebs, thinks me singularly loved by this Magnus of ours; and, by Hercules, we have been joined in a long and pleasant intimacy, to such a degree that those table-companions in conspiracy of ours, the bearded young men, in their talk call him Gnaeus Cicero. So at the games and at the gladiators we used to carry off wonderful demonstrations of approval \textit{[Greek: epis\=emasias]} without any shepherd's pipe.

\ciceroSection{12} Now there is the expectation of the elections; into which, against everyone's will, our Magnus is shoving the son of Aulus, and in this fights neither by authority nor by favour but by those means by which Philip used to say all fortifications could be stormed which only an ass laden with gold could climb up to. The consul, in fact, like an inferior actor, is said to have undertaken the business and to have agents-of-bribery in his house --- which I do not believe. But two senate's decrees have already been made which are unwelcome, because they are thought to have been made against a consul, on the proposal of Cato and Domitius: one, that it be permitted to investigate at the house of magistrates; the other, that the man at whose house bribery-agents stayed should be reckoned an enemy of the commonwealth.

\ciceroSection{13} Lurco, however, the tribune of the plebs, who entered office --- ``together with another law'' --- was loosed from the Aelian and Fufian law that he might bring forward a law on bribery, which the lame man, with good auspices, has duly promulgated. So the elections have been put off to the sixth day before the Kalends of Sextilis. There is this novelty in the law: that he who has named the price within his tribe, if he should not have given, is unpunished; but if he has given, that for as long as he lives he should owe to each member of the tribe HS 3,000. I said that Publius Clodius had observed this law before, since it was his habit to name the price and not give. But you --- do you see, if this fellow shall be elected, that consulship of ours which Curio used to call ``apotheosis'' \textit{[Greek: apoth\=eosin]} will be a Bean Mime? Wherefore, in my view, one must philosophize \textit{[Greek: philosoph\=eteon]}, as you do, and reckon those consulships not worth a stalk \textit{[Greek: floccou facteon]}.

\ciceroSection{14} As to what you write to me, that you have decided not to go into Asia: I should rather you went, and fear that something less convenient may come of it; but yet I cannot rebuke your counsel, especially since I myself did not go out into a province.

\ciceroSection{15} With the epigrams that you have placed in the Amaltheum we shall be content, especially since both Thyillus has left us, and Archias has written nothing about me. And I fear that, since he has now composed a Greek poem on the Luculli, he will turn next to a Caecilian play.

\ciceroSection{16} I have given Antonius your thanks in your name, and have given that letter to Mallius. To you I have written more rarely before because I had no fit man to whom I might give the letter, nor did I know quite well to where to send it. I have praised you greatly to him.

\ciceroSection{17} If Cincius brings me anything of your business, I shall undertake it; but he is now more occupied in his own, in which I do not fail him. If you are going to be in one place, expect frequent letters from us; but you yourself send even more.

\ciceroSection{18} Please write to me what kind your Amaltheum is, with what adornment, in what landscape \textit{[Greek: topothesi\=ai]}, and send me the poems and stories you have about Amalthea \textit{[Greek: Amalthei\=ai]}. I want to make one in the Arpinum estate. I shall send you something of my writings; I had nothing finished.
