% Ad Familiares 5.5 (ad-familiares-05-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. Cicero to Gaius Antonius, son of Marcus, Imperator, greetings.}

\ciceroSection{1} Although I had decided to send no letters to you except those of recommendation --- not because I supposed they had any great weight with you, but so that I should not show those who asked them of me that anything in our connection had been diminished --- yet, when Titus Pomponius, the man most fully aware of all my zeal and good offices toward you, eager for you, most loving of me, was setting out for you, I thought I should write something, especially since I could not otherwise satisfy Pomponius himself.

\ciceroSection{2} If I am demanding the highest offices from you, that should seem strange to no one. For all things proceeded from me to you that pertained to your interest, your honour, your dignity. That no thanks have been returned to me by you for these things, you yourself are the best witness; that, on the contrary, something even has come from you against me, I have heard from many --- for I do not dare to say that I have ``ascertained'' it, lest I happen to use that very word which men say is wont to be falsely turned against me by you. But the things which have been brought to me, I prefer that you learn of them from Pomponius (to whom they have been no less troublesome) than from any letter of mine. What kind of mind I have had toward you, in singular service, both the senate and the Roman people are witness; how grateful you have been to me, you yourself can judge; how much you owe me, the rest will judge.

\ciceroSection{3} What I did before for your sake, I did being moved by goodwill, and afterwards by my own constancy. But the rest, believe me, demands far greater zeal of me, and greater gravity, and greater labour. If I shall not seem to be pouring all this away and losing it, I shall sustain it with all my strength. But if I begin to feel it is unwelcome, I shall not bring it about that I appear to you yourself to be mad. Of what sort and of what kind these things are, you may learn from Pomponius. And Pomponius himself I commend to you in this way: that, although I trust you will do everything for his own sake, yet I beg this of you --- that, if any love toward me yet remains in you, you display it all in Pomponius's business. Nothing more pleasing can you do for me.
