% Ad Familiares 5.6 (ad-familiares-05-06) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{M. Cicero to Publius Sestius, son of Lucius, Proquaestor, greetings.}

\ciceroSection{1} When Decius the secretary had come to me and dealt with me that I should give my pains that no successor be appointed to you at this time --- although I judged him a worthy man and friendly to you, yet, since I held in memory what kind of letters you had sent to me earlier, I did not quite trust a prudent man that your wish had been so greatly changed. But after both your Cornelia met with Terentia, and I spoke with Quintus Cornelius, I applied diligence: as often as the senate was sitting, I attended; and I had no slight business in compelling Quintus Fufius the tribune of the plebs and the others to whom you had written, to trust me rather than your letter. The whole matter has been thrown back to the month of January, but it was easily held.

\ciceroSection{2} I, moved by your congratulation --- because long ago you had written that you wished it to turn out well that I had bought the house from Crassus --- bought the very house for HS 3,500,000, some little while after your congratulation. So know that now I have so much debt that I should be glad to enter into a conspiracy if anyone would receive me; but, on one side, men induced by hatred shut me out, and openly hate the avenger of conspiracy; on the other, men do not believe me, and fear plots from me, and do not think that money can be lacking to a man who has rescued the financiers from a siege. There is great supply at half-percent. By my deeds, however, I have achieved this: that I am reckoned a good name.

\ciceroSection{3} Your house and the whole building of it I have looked over and have warmly approved. Antonius, although his services to me are missed by all, I yet defended in the senate most gravely and most diligently, and stirred the senate vehemently both by my speech and by my authority. Please send me letters more often.
