% Ad Atticum 1.4 (ad-atticum-01-04) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} You stir up frequent expectations of yourself in us. Lately, when we were already supposing that you were on your way, we were suddenly put off by you to the month of Quintilis. Now in fact I feel: come, so far as you can do so without disrupting your business, at the time you write of. You will be in time for my brother Quintus's election; you will see us after a long interval; you will wind up the Acutilian quarrel. Peducaeus too has reminded me to write this to you, for we think it is time you settled the matter at last.

\ciceroSection{2} My intercession is and has been ready. We here, with an incredible and singular goodwill of the people, have made an end of Gaius Macer. Although we had been fair to him, we yet reaped a far greater fruit from the people's esteem by his condemnation than we should have reaped from his thanks if he had been acquitted.

\ciceroSection{3} As to the Hermathena you write of, I am very grateful. It is an ornament fit for my Academy, since Hermes is common to all and Minerva is the singular emblem of that gymnasium. So I should be glad if, as you write, you adorn that place with as many other things as possible. The statues you sent me earlier I have not yet seen; they are at the Formian villa, where I was just thinking of setting out. All those things I shall transfer to the Tusculan villa. Caieta I shall adorn whenever I begin to have a surplus. Keep your books, and do not despair that I shall be able to make them mine. If I succeed in that, I outdo Crassus in wealth, and I despise the country places and water-meadows of all the others.
