% Ad Atticum 1.3 (ad-atticum-01-03) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} Know that your grandmother has died of longing for you --- and at the same time because she feared the Latin towns would not stand to their duty and would not bring victims up to the Alban Mount. The consolation for this matter, I take it, Lucius Saufeius will be sending to you.

\ciceroSection{2} We here are looking for you in the month of January, on the strength of some rumour or of letters of yours sent to others; for to me you have written nothing of it. The statues which you arranged for us have been unloaded at Caieta. We have not seen them; for I have had no chance of leaving Rome. We have sent a man to pay for the freight. We love you greatly that they have been taken care of by you so diligently and at such small cost.

\ciceroSection{3} As to what you have often written to me about appeasing our friend, I have done it and have tried everything; but in a wonderful way his mind is set against you. From what suspicions, although I think you have heard, you will yet learn from me when you arrive. Sallustius, on the spot, I could not restore to his old standing with him. I write this because he was wont to accuse me on your account. He has himself made trial that the man is less to be moved than he had thought, and that my zeal has not failed either himself or you. Tullia we have betrothed to Gaius Piso son of Lucius, of the Frugi.
