% Ad Atticum 1.11 (ad-atticum-01-11) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} I had been doing this on my own initiative before, and afterwards, by two of your letters written most diligently to the same end, I was strongly stirred. To this was added Sallustius as constant urger that I should treat with Lucceius as carefully as possible about restoring the old friendship between you. But although I have done everything, not only could I not recover the goodwill toward you that he had once had, I could not even draw out the cause of the changed feeling. He boasts of his own decision in the matter and of the things which I had understood, even when you were here, were already offending him; yet he certainly has something deeper-seated in his mind which neither your letters nor my advocacy can so easily wipe away. This you, on the spot, will remove not only by speech but by that familiar look of yours --- if only you reckon it worth the trouble; which, if you listen to me and if you wish to be true to your own kindness, you certainly will. And do not wonder why, when I had earlier signalled by letter that I hoped he would be in our power, I now seem to be losing confidence. It is past belief how more obstinate his feeling toward you appears, how more set in this anger. But this will either be cured when you arrive, or will be vexatious to him, on whichever side the fault lies.

\ciceroSection{2} As to what was written in your letter, that I am already, in your view, declared elected --- know that nothing in Rome is now so harassed as the candidates, with every kind of unfairness, and that it is not even known when the elections will be.

\ciceroSection{3} You will hear of these things from Philadelphus. As to what you have prepared for our Academy, please send it as soon as you can. Wonderful how the use --- and even the very thought --- of that place delights me. But your books, take care that you do not hand them over to anyone; keep them for us, as you write. The greatest passion for them holds me, as does loathing now for everything else; how very much, in how brief a time, those things have grown worse than you left them, you will scarcely believe.
