% Ad Atticum 1.5 (ad-atticum-01-05) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} How great a grief I have taken, and of how great a fruit I have been deprived, both forensic and domestic, by the death of our cousin Lucius --- you, by virtue of our intimacy, can judge as well as anyone. For all that can come to a man as joy from the kindness and character of another came to me from him. So I have no doubt that this too is a sorrow to you, since you both are moved by my grief and have yourself lost a connection and a friend most adorned with every virtue and duty, who loved you both of his own accord and through what I had said.

\ciceroSection{2} As to what you write to me about your sister, she herself will be witness to you of how greatly it has been my care that my brother Quintus's feeling toward her should be what it ought to be. Thinking him still rather offended, I sent him a letter in which I appeased him as a brother, and admonished him as a younger man, and rebuked him as one in error. So, from what he has often written back to me since, I am confident that things are as they ought to be and as we wish.

\ciceroSection{3} About the sending of letters, you accuse me without cause. For I was never told by our Pomponia that there was anyone to whom I might give a letter; nor again has it happened to me to have someone setting out for Epirus, nor were we yet hearing that you were at Athens.

\ciceroSection{4} The Acutilian business which you had commissioned to me I had wound up, as soon as I came back to Rome after parting from you. But it happened that no contention was needed, and that I, judging there to be counsel enough in you, preferred Peducaeus to give you advice by letter rather than I. For when for many days I had given my ears to Acutilius --- whose kind of talk I think is known to you --- I did not think it heavy to write to you of his complaints, having reckoned the listening (which was rather tedious) the lighter task. But from you, who accuse me, know that I have had only one letter delivered, even though you have had both more leisure for writing and a greater chance of sending.

\ciceroSection{5} As for what you write --- that, even if anyone's feeling toward you should be more offended, he ought to be brought back to you by me --- I grasp what you mean and have not neglected it; but the man's mind is in a quite extraordinary state. I, however, did not pass over anything that needed to be said in your defence; how hard I should press the contest, I thought I ought to settle on the basis of your wish. If you write that out to me at length, you will see that I have neither wished to be more diligent than you would be, nor will be more negligent than you wish.

\ciceroSection{6} About the Tadian business, Tadius has spoken with me; you wrote, he said, that there was no need now to be troubled, since the inheritance had been acquired by usucaption. I was surprised that you did not know that, in the case of statutory guardianship --- under which the girl is said to be --- nothing can be acquired by use.

\ciceroSection{7} Of the Epirote purchase I am glad it pleases you. What I have commissioned to you, and what you will yourself see to suit our Tusculan villa, please, as you write, see to --- so far as you can do without trouble. For from all troubles and labours we shall rest in that one place.

\ciceroSection{8} My brother Quintus we are looking for daily. Terentia has great pains in her joints. She is most fond of you and your sister and your mother, and adds her warmest greetings, as does little Tullia, our darling. Take care to be well and to love us and to be sure that you are loved by me as a brother.
