% Ad Atticum 1.8 (ad-atticum-01-08) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroLetterOpener{Cicero to Atticus, greetings}

\ciceroSection{1} At your house things are as we would wish. Your mother and your sister are loved by me and my brother Quintus. With Acutilius I have spoken; he says nothing has been written to him by his agent, and is surprised that this controversy arose --- he, that the agent refused to give security that nothing further would be sought from you. As to your having settled the Tadian business, I gather that this was both welcome and very pleasing to Tadius. That friend of ours --- by Hercules an excellent man and most friendly to me --- is, I must say, very angry with you. If I knew how high you set this, I should know how much I have to put into smoothing it over.

\ciceroSection{2} To Lucius Cincius I have paid out HS 20,400 for the Megarian statues, as you had written to me. Your Pentelic Hermae with the bronze heads, of which you wrote to me, please me very greatly already. So please, send those and the statues and the rest --- as many as you can and as soon as you can --- whatever will seem to you to suit the place and our taste and your own elegance, especially what will seem to you fit for a gymnasium and a colonnade. For in this kind I am so carried away by zeal that I am to be helped on by you, almost rebuked by others. If Lentulus's ship is not available, put them on whatever ship pleases you.

\ciceroSection{3} Little Tullia, our darling, is dunning you for your little present and calls me as guarantor; but for me it is surer to forswear the bond than to pay it.
