% Brutus (brutus) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} When, on my way back from Cilicia, I had come to Rhodes, and word was brought to me there of the death of Q. Hortensius, I took it to heart with a grief greater than anyone supposed I would feel. For with a friend lost I saw myself bereft both of a delightful companionship and of the bond of many shared services; and at the death of so great an augur I grieved that the dignity of our college was diminished. And in this reflection I recalled both that I had been co-opted into the college by him — in which, under oath, he had passed judgment on my worth — and that I had been inaugurated by the same man, on whose account, by the institutions of the augurs, I was bound to honour him in a father's place.

\ciceroSection{2} My distress was deepened, too, because in a great dearth of wise and good citizens this distinguished man, most closely joined to me by partnership in all my counsels, had been struck down at a time most hostile to the commonwealth, and had left us a sorrowful longing for both his authority and his wisdom; and I grieved that I had lost — not, as most supposed, an adversary or a disparager of my own praises — but rather an ally and a partner in glorious labour.

\ciceroSection{3} For if, in the study of lighter arts, it is recorded in memory that noble poets have grieved at the death of poets their equals, then with what feeling, I ask, ought I to have borne the death of a man with whom it was more glorious to contend than to have no rival at all? — especially since not only was his course never hindered by me, nor mine by him, but on the contrary each of us was always helped by the other, by sharing, by counsel, and by encouragement.

\ciceroSection{4} But since, having enjoyed a kind of unbroken good fortune, he departed from life at his own season rather than that of his fellow citizens, and died at a time when, if he had lived, he could more easily have mourned the commonwealth than aided it, and lived as long as it was permitted to live well and happily in the state — let us grieve, if it must be so, at our own loss and harm; but let us attend the timeliness of his death with goodwill rather than with pity, so that, as often as we think of that most illustrious and most fortunate man, we may seem to be cherishing him rather than ourselves.

\ciceroSection{5} For if we grieve at this — that it is no longer permitted us to enjoy him — that is an evil of our own, which we should bear with moderation, lest we seem to refer it not to friendship but to our own private advantage; but if we are tormented as though some bitterness had befallen him himself, then we are not interpreting his supreme good fortune with a sufficiently grateful mind.

\ciceroSection{6} For if Q. Hortensius were alive, he would perhaps feel the want of other things together with the rest of our good and brave citizens; but this one grief he would bear beyond all others, or with only a few, when he saw the Forum of the Roman people — which had been, as it were, the theatre of his genius — stripped and bereft of that cultivated voice worthy of Roman and of Greek ears.

\ciceroSection{7} For my own part, I am tormented in spirit that the commonwealth has need of the arms not of counsel, not of genius, not of authority — arms which I had learned to wield, to which I had accustomed myself, and which belonged both to a man of distinction in public life and to a state of good morals and good order. But if there was ever a time in the commonwealth when the authority and the eloquence of a good citizen could wrest weapons from the hands of angry fellow citizens, it was surely then, when the advocacy of peace was shut out, whether by the error or by the fear of men.

\ciceroSection{8} And so it has befallen us that, although there were far other things more to be mourned, we yet grieved at this: that at the very time when our generation, having discharged the most ample affairs, ought to take refuge as if into a harbour — not of sloth nor of idleness, but of measured and honourable leisure — and when our own eloquence was now turning grey and had reached a certain ripeness and, as it were, an old age, then those weapons were taken up which even the very men who had learned to use them gloriously could not find how to use for the public good.

\ciceroSection{9} And so those men seem to me to have lived fortunately and happily — in other states, but most of all in our own — who were permitted to enjoy not only authority and the glory of their deeds but also the praise of wisdom. The memory and recollection of these men was indeed a pleasure to us amid our greatest and gravest cares, when we lately fell upon it in the course of a certain conversation.

\ciceroSection{10} For while I was walking in the colonnade and was at leisure at home, M. Brutus came to me, as was his custom, together with T. Pomponius — men closely joined to each other and so dear to me and so welcome that at the sight of them all the anxiety which had been gnawing at me over the commonwealth subsided. After I had greeted them: 'What of you, Brutus and Atticus?' I said. 'Is there any news at last?' 'Nothing at all,' said Brutus, 'that you would wish to hear, or that I would dare to report for certain.'

\ciceroSection{11} Then Atticus said: 'We have come to you with this purpose: that there should be silence about the commonwealth, and that we should hear something from you rather than burden you with any trouble.' 'You indeed, Atticus,' I said, 'both relieve me of care while you are present, and have given me great comfort while I was away. For by your letters first of all I was revived and called myself back to my old pursuits.' Then he said: 'I read with great pleasure the letter that Brutus sent you from Asia, in which he seemed to me both to advise you wisely and to console you most affectionately.'

\ciceroSection{12} 'He was rightly so seen,' I said; 'for know that by that letter I was called back, as if to behold the light, out of a long disorder of my whole well-being. And just as, after that disaster at Cannae, the Roman people first lifted itself up by Marcellus' battle at Nola, and afterward many favourable events followed one upon another, so, after the gravest reverses of my own affairs and of the common cause, nothing happened to me before Brutus' letter that I could wish for, or that in any part lightened my anxieties.'

\ciceroSection{13} Then Brutus said: 'I did indeed wish to bring that about, and I reap a great reward, if I have truly achieved in so great a matter what I intended. But I am eager to know what it was in Atticus' letter that delighted you.' 'That letter, Brutus,' I said, 'brought me not only delight but also, as I hope, deliverance.' 'Deliverance?' he said. 'What sort of letter, then, was that, so remarkable?' 'Could any greeting have been more welcome to me,' I said, 'or better suited to this present time, than that book of his by which, addressing me, he roused me as I lay prostrate?'

\ciceroSection{14} Then he said: 'You mean, surely, the one in which he embraced the whole record of history briefly and — as it seemed to me, at least — most carefully?' 'That very book, Brutus,' I said, 'I mean was my deliverance.' Then Atticus said: 'What you say is most welcome to me indeed; but what after all did that book have in it that could be either new to you or of such great use?'

\ciceroSection{15} 'It had, in truth,' I said, 'much that was new to me at least, and the very usefulness I was looking for: that, with the sequences of the ages set out in order, I might see everything in a single view. And when I had begun to handle it eagerly, the very handling of letters proved my salvation, and reminded me, Pomponius, that I should take from you yourself something for my own restoration, and to repay you with a gift that, if not equal, was at least grateful — although that saying of Hesiod is praised by the learned, which bids us return in the same measure as we have received, or even in a fuller one, if we can.

\ciceroSection{16} For my part, I shall surely measure out to you my goodwill in full; but the thing itself I do not yet seem able to give; and that you should pardon this, I ask of you. For I have no fresh fruits, as farmers do, from which to repay you what I received — so checked is all my produce, and the flower of the old abundance has withered, scorched with thirst — nor any from the stores that lie in darkness and to which all approach is now blocked off from us, an approach that lay open to me almost alone. So I shall sow something, as it were, in a soil uncultivated and abandoned; and this I shall tend so carefully that I may even augment, with my own outlay, the bounty of your gift — if only my own mind can do what the field does, which, when it has lain at rest for many years, is wont to yield richer crops.' Then he said:

\ciceroSection{17} 'For my part, I shall both await what you promise, and shall not exact it except at your convenience; and it will be most pleasing to me if you discharge the debt.' 'For me too,' said Brutus, 'what you promise to Atticus is to be awaited — even though perhaps I, his volunteer agent, may demand it of you on his behalf, since the very man to whom you owe it declares he will not exact it to your inconvenience.'

\ciceroSection{18} 'But indeed, Brutus,' I said, 'I shall not discharge the debt to you unless I have first secured from you that no one else shall make any further claim on that account.' 'By Hercules,' he said, 'I would not dare promise you that in return. For this man, who says he will not, I see will be a dun — not a troublesome one to you, I grant, yet a persistent and a keen one.' Then Pomponius said: 'For my part, I think Brutus says nothing false. For I think I shall now dare to address you, since after a long interval I have only just now noticed you a little more cheerful.

\ciceroSection{19} And so, since this man has declared he will exact what was owed to me, I demand of you what you owe to him.' 'And what is that?' I said. 'That you write something,' he said; 'for it is now a long while that your pen has fallen silent. For since you published those books on the commonwealth, we have surely received nothing from you afterward — and it was by them that we ourselves were spurred and kindled to set down a record of our own affairs. But that work, when you can; and that you may be able, I ask;

\ciceroSection{20} now, however,' he said, 'if your mind is free, set out for us what we are asking.' 'And what is that?' I said. 'What you lately began for me at your Tusculan villa, about the orators — when they had begun to be, and who too they were, and of what sort. And when I had reported that conversation to your Brutus, or rather to ours, he said he greatly wished to hear it. And so we have chosen this day, knowing you to be free. Therefore, if it is convenient for you, set forth for Brutus and for me those things you had begun.'

\ciceroSection{21} 'For my part,' I said, 'I shall satisfy you, if I can.' 'You can,' he said; 'only relax your mind a little, or, if you can, set it quite free.' 'The conversation, then, was led from this point, was it not, Pomponius — that mention had been made by me that I had heard the case of Deiotarus, that most faithful and best of kings, defended by Brutus most elegantly and most copiously.' 'I know,' he said, 'that the talk was drawn from that beginning, and that you, grieving Brutus' lot, had as it were lamented the desolation of the courts and of the Forum.' 'I did do that,' I said, 'and I do it often.

\ciceroSection{22} For it often comes into my mind, Brutus, when I look upon you, to fear what course your admirable nature, your exquisite learning, and your singular industry will at length have. For when you had been engaged in the greatest cases, and when our generation was already yielding to you and lowering its fasces, suddenly within the state, as other things fell, so too that very eloquence which we are setting out to discuss fell silent.'

\ciceroSection{23} Then he said: 'For the sake of other things I both grieve at that and think it ought to be grieved at; but as for speaking, it is not so much its fruit and glory that delights me as the pursuit itself and the exercise — which nothing shall snatch from me, you especially being so devoted to me. For no one can speak well unless he understands wisely; so that whoever gives his effort to true eloquence gives it to wisdom, which not even in the greatest wars can anyone do without with an even mind.'

\ciceroSection{24} 'Finely said, Brutus,' I replied; 'and I take all the more delight in that praise of speaking, because, of the other things which were once held in the state as most beautiful, there is no one so lowly that he does not think he can either attain them or has already attained them; but I see that no man has been made eloquent by victory. But that the conversation may unfold more easily, let us conduct it sitting, if it seems good.' When the same had pleased them, we then sat down on a little meadow beside the statue of Plato.

\ciceroSection{25} Here I began: 'To praise eloquence, then, and to set forth how great its power is, and how great a dignity it confers on those who have attained it, is neither my purpose in this place nor necessary. But this I would affirm without any hesitation: that whether eloquence is brought forth by some art, or by a certain practice, or by nature, it is the one most difficult thing of all. For of the five elements out of which it is said to consist, each one is itself a great art in its own right. From this it can be judged how great a power, and how great a difficulty, lies in the convergence of five very great arts.

\ciceroSection{26} Greece is my witness, which, though it has been kindled with zeal for eloquence and has long excelled in it and surpassed all others, yet has all its other arts older — and not only discovered but even perfected long before — than this power and abundance of speaking was wrought out by the Greeks. And when I look upon this, there come to my mind above all, Atticus, and as it were shine forth, your Athens — the city in which the orator first raised himself up, and in which speech first too began to be entrusted to monuments and to letters.'

\ciceroSection{27} And yet before Pericles, of whom some writings are passed down, and before Thucydides — men who belonged to Athens not at her birth but already in her maturity — there is not a single piece of writing that possesses any ornament and seems to be an orator's. Although the opinion holds that Pisistratus, who lived many years before these men, and Solon, somewhat older still, and afterwards Cleisthenes, were each, by the measure of those times, men of great power in speaking.

\ciceroSection{28} Some years after this age, as can be made out from the Attic records, came Themistocles, who, it is agreed, excelled both in practical wisdom and in eloquence; after him Pericles, who, while he flourished in every kind of excellence, was nonetheless most famous for this distinction. Cleon too, in those times, was, it is agreed, a turbulent citizen indeed, but eloquent for all that.

\ciceroSection{29} Close upon this age came Alcibiades, Critias, Theramenes; and what kind of speaking flourished in their day can best be understood from the writings of Thucydides, who himself lived then. They were grand in their words, thick with thoughts, terse in their compression of matter, and for that very reason at times somewhat obscure.

\ciceroSection{30} But once it was understood how great a force a careful and, in a manner, crafted style possessed, then too many teachers of speaking suddenly arose. Then Gorgias of Leontini, Thrasymachus of Chalcedon, Protagoras of Abdera, Prodicus of Ceos, Hippias of Elis stood in great honour; and many others in those same times professed to teach — in arrogant words, to be sure — how the weaker case (for so they put it) could, by speaking, be made the stronger.

\ciceroSection{31} Against these men Socrates set himself, who by a certain subtlety of argument was accustomed to refute their teachings. Out of his most abundant conversations arose the most learned men; and then for the first time philosophy was discovered — not that older philosophy of nature, but this one, in which we debate of good things and bad and of the life and conduct of men. But since this branch is foreign to what we have set before us, let us put off the philosophers to another time, and return to the orators, from whom we have digressed.

\ciceroSection{32} There arose, then, when those men I named a little while ago were already old, Isocrates, whose house lay open to the whole of Greece as a kind of school and workshop of speaking — a great orator and a perfect teacher, though he was without the light of the Forum and reared within his own walls that glory which, in my judgment at least, no one afterwards attained. He both himself wrote much, and splendidly, and taught others; and as he surpassed his predecessors in everything else, so he was the first to understand that even in prose, while you avoid verse, a certain measure and rhythm must nonetheless be kept.

\ciceroSection{33} For before him there was no structure of words, so to speak, and no rounding of the clause to a rhythm, or, if ever there was, it did not appear to have been sought of set purpose — which is perhaps a merit; yet for all that it came about then by nature rather and by chance, never by any method or any conscious attention.

\ciceroSection{34} For nature herself, by a certain framing of the words, grasps and rounds off the thought, and when that thought is bound tight with fitting words, it falls, more often than not, in rhythm too. For the ears themselves judge what is full and what is empty, and by the breath, as by a kind of necessity, the framing of the words is brought to its close; in which to fall short, and even to labour, is shameful.

\ciceroSection{35} Then there was Lysias, a man not himself engaged in cases of the Forum, but an exceptionally subtle and elegant writer, whom you might by now almost venture to call the perfect orator. For as plainly perfect, and lacking nothing whatever, you would readily name Demosthenes. Nothing acute could be found in those cases he wrote, nothing, so to speak, sly, nothing artful, that he did not see; nothing could be said with more subtlety, more compression, more clarity, that anything might be made more finely polished; nothing, on the other side, more grand, more impassioned, more adorned whether by weight of words or of thoughts, that anything could be loftier.

\ciceroSection{36} Next to him came Hyperides and Aeschines and Lycurgus and Dinarchus, and that man of whom no writings survive, Demades, and many others besides. For this age poured forth this abundance; and, as my opinion holds, that sap and uncorrupted blood of the orators lasted right down to this age — an age in which there dwelt a natural, not a painted, brilliance.

\ciceroSection{37} For Demetrius of Phalerum succeeded those old men, a young man and the most learned indeed of all of them, but trained not so much in the camp as in the wrestling-school. And so he delighted the Athenians rather than fired them. For he had advanced into the sun and the dust, not as from a soldier's tent, but as from the shaded walks of Theophrastus, that most learned of men.

\ciceroSection{38} He was the first to bend down his oratory and to make it soft and tender, and he preferred to seem agreeable, as he was, rather than weighty; but with such agreeableness as would steep men's minds, not pierce them through, and only so far as to leave behind the memory of his own neat finish — not, as Eupolis wrote of Pericles, to leave, along with the delight, stings as well in the minds of those who had heard him.

\ciceroSection{39} Do you see, then, even in that very city in which eloquence was both born and reared, how late it came forth into the light? Since indeed, before the age of Solon and Pisistratus, nothing is handed down to memory of anyone as eloquent. And these men, by the reckoning of the Roman people's age, ought to seem old; but by the count of the Athenians' generations, young. For although they flourished in the reign of Servius Tullius, still Athens by then had stood far longer than Rome stands to this present day. Yet I do not doubt that oratory has always had great force.

\ciceroSection{40} For Homer would not have ascribed so much glory in speaking to Ulysses and to Nestor in the times of Troy — wishing the one to have force, the other charm — had not the honour of eloquence already existed even then; nor would the poet himself have been so ornate in speaking, and indeed an orator outright. Though his dates are uncertain, still he lived many years before Romulus; if indeed he was not later than the elder Lycurgus, by whom the discipline of the Lacedaemonians was bound fast with laws.

\ciceroSection{41} But a zeal for that kind of thing, and a greater force of it, is recognized in Pisistratus. Then, in the next generation, Themistocles followed him — by our reckoning very ancient, by the Athenians' not so very old at all. For he lived when Greece was already under settled rule, while our own state had not so long before been freed from the domination of kings. For that gravest of wars against the Volsci, in which the exile Coriolanus took part, fell at nearly the same time as the war against the Persians, and the fortune of the famous men was alike;

\ciceroSection{42} since indeed each, after he had been an outstanding citizen, was driven out by the wrong of an ungrateful people, took himself off to the enemy, and quelled the attempt of his own anger by death. For although it stands otherwise in your account, Atticus, concerning Coriolanus, grant me nonetheless to assent rather to this manner of death. But he, laughing: ``As you please, then,'' he said; ``since indeed it is conceded to the rhetoricians to make false statements in their histories, that they may say something more cleverly. For just as you now invent about Coriolanus, so did Cleitarchus, so did Stratocles invent about Themistocles.

\ciceroSection{43} For the man whom Thucydides — who was both an Athenian and born in the highest station and himself a man of the highest rank, and a little later in age — wrote merely to have died of illness and to have been buried secretly in Attica, adding that there had been a suspicion that he had taken his own life by poison: this man these fellows say, when he had sacrificed a bull, caught the blood in a bowl, and, having drunk it, fell down dead. For this death they could adorn rhetorically and tragically; that ordinary death offered no material for ornament. Therefore, since it suits you so well that everything in Themistocles was matched with Coriolanus, you may take the bowl from me as well; I shall furnish the victim too, so that Coriolanus may be a second Themistocles outright.''

\ciceroSection{44} ``Be it so,'' I said, ``as you like, about that; and I, for my part, shall handle history more cautiously hereafter in your hearing, since I can praise you as the most scrupulous authority on Roman affairs. But it was at about that time that Pericles, the son of Xanthippus, of whom I spoke before, first brought learning to bear; and although there was then no learning of speaking, still, schooled by the natural philosopher Anaxagoras, he had readily transferred the exercise of his mind from recondite and hidden matters to the cases of the Forum and the people. By his agreeableness Athens was most gladdened, his richness and abundance she admired, and the force and terror of his speaking she feared.''

\ciceroSection{45} This age, then, was the first at Athens to bring forth an orator nearly perfect. For neither among men founding a commonwealth, nor among men waging wars, nor among men hampered and bound fast under the domination of kings, does a desire for speaking usually arise. Eloquence is the companion of peace, the partner of leisure, and a kind of nursling, as it were, of a state already well established.

\ciceroSection{46} And so Aristotle says that when the tyrants had been removed in Sicily and, after a long interval, private property was being reclaimed at law, then for the first time — since that people was keen and contentious by nature — the Sicilians Corax and Tisias composed an art and precepts; for before this no one had been accustomed to speak by method or by art, though many spoke carefully nonetheless and in due order. He says too that disputations on celebrated subjects were written out and made ready by Protagoras — what are now called the common topics;

\ciceroSection{47} that Gorgias did the same, having written praises and censures of particular things, since he judged this to be most peculiarly the orator's task: to be able to magnify a thing by praising, and again to dash it down by blaming; that Antiphon of Rhamnus had certain similar things written out, than whom no one ever pleaded any capital case better, when he defended himself — so Thucydides, a weighty authority, wrote, who was himself a hearer of it;

\ciceroSection{48} that Lysias at first was accustomed to profess that speaking was an art; then, because Theodorus was subtler in the art but in his speeches more meagre, Lysias began to write speeches for others and put the art aside; that likewise Isocrates at first had denied that speaking was an art, but was accustomed to write speeches for others to use in the courts; yet, since on this account he was himself often called into court — because he seemed, as it were, to be offending against the law ``whereby anyone should be circumvented in a trial'' — he gave up writing speeches for others and turned himself wholly to the composing of arts.

\ciceroSection{49} You see, then, the birth and the springs of the orators of Greece — old by the reckoning of our annals, but, by their own, recent indeed. For before the Athenian state took its delight in this glory of speaking, it had already accomplished many memorable things both in domestic and in warlike affairs. This pursuit, moreover, was not common to Greece, but peculiar to Athens.

\ciceroSection{50} For who knows of any Argive orator, or Corinthian, or Theban, in those times? — unless one cares to suspect something of Epaminondas, a learned man. Of a Lacedaemonian, indeed, I have heard that down to this time there was none. Menelaus himself, that sweet-spoken man, Homer hands down to us, but as one who said few words. Brevity, however, is at times a merit in some single part of speaking; in eloquence as a whole it earns no praise.

\ciceroSection{51} But outside Greece there were great pursuits of speaking, and the very great honours paid to this glory made the name of the orators illustrious. For once eloquence had sailed out from the Piraeus, it traversed all the islands, and so wandered abroad through the whole of Asia that it smeared itself with foreign manners and lost all that wholesomeness and, so to speak, soundness of Attic diction, and almost unlearned how to speak. From this came the Asiatic orators, not to be despised indeed either for swiftness or for abundance, but too little compressed and too redundant; the Rhodians sounder and more like the Attic men.

\ciceroSection{52} But of the Greeks, enough; for these very matters perhaps were not necessary. Then Brutus: ``As to how necessary they were,'' he said, ``I could not easily say; agreeable they certainly were to me, and not only not long, but even shorter than I should have wished.'' ``Excellent,'' I said; ``but let us come to our own people, about whom it is hard to understand more than what one may surmise from the records.''

\ciceroSection{53} For who would suppose that quickness of mind was lacking to that famous Lucius Brutus, the founder of your nobility? The man who, from the oracle of Apollo, drew so sharp and so witty a conjecture about the kissing of his mother; who veiled the highest prudence under a pretence of stupidity; who drove out the most powerful of kings, the son of a most illustrious king, and bound down the state — freed from perpetual despotism — with annual magistracies, with laws, with courts; who stripped his own colleague of command, that he might remove from the city the very memory of the royal name: which surely could not have been brought about had it not been carried by the persuasion of his speech.

\ciceroSection{54} So too we see that a few years after the kings were expelled, when the plebs had settled near the bank of the Anio at the third milestone and had seized that hill which was called the Sacred Mount, Marcus Valerius the dictator quieted the discord by his speaking, and that for this reason the most ample honours were paid to him, and that he was the first to be called Maximus precisely on this account. Nor do I think that even Lucius Valerius Potitus accomplished nothing by speaking — the man who, after the hatred kindled against the decemvirate, soothed by his laws and his speeches a plebs that had been roused against the senators.

\ciceroSection{55} We may guess that Appius Claudius was a fluent speaker, because he called back a senate already on the very point of inclining toward peace with Pyrrhus; that Gaius Fabricius was, because he was sent as envoy to Pyrrhus to recover the prisoners; Tiberius Coruncanius, because from the records of the pontiffs he appears to have been by far the most powerful in intellect; Manius Curius, because, when he was tribune of the plebs and the interrex Appius Caecus — an eloquent man — was holding the elections against the laws, refusing to accept a consul from the plebs, Curius compelled the senators to give their authorization beforehand: which was a very great thing, the Lex Maenia (a law requiring the senate's prior sanction of elected magistrates) not yet having been passed.

\ciceroSection{56} One may guess something too about the talent of Marcus Popilius, who, when he was consul and at the same time was performing a public sacrifice in the priestly robe — for he was flamen of Carmenta — and the rousing and sedition of the plebs against the senators was announced, came into the assembly just as he was, wrapped in the robe, and quelled the sedition as much by his authority as by his speech. But that these men were held to be orators, or that there was at that time any reward at all for eloquence, I really do not seem to myself to have read anywhere: I am led to my surmise by conjecture alone.

\ciceroSection{57} Gaius Flaminius too is said to have had power over the people by his speaking — the man who, as tribune of the plebs, carried the law for distributing the Gallic and Picene land man by man, and who as consul was killed at Trasimene. Quintus Maximus Verrucosus also was held to be an orator in those times, and Quintus Metellus, the man who was consul with Lucius Veturius Philo in the Second Punic War. But the first of whom there is evidence still surviving, and of whom it has been handed down to memory that he was eloquent and was so reckoned, is Marcus Cornelius Cethegus, for whose eloquence the warrant — and a fit one, in my judgment — is Quintus Ennius, especially since Ennius both heard the man himself and writes of him when dead: from which there is no suspicion that he lied for friendship's sake.

\ciceroSection{58} The passage in him stands, in the ninth book of the \textit{Annals} I think, thus: ``Added there is an orator, Cornelius — sweet of speech / and sweet of mouth — Cethegus Marcus, Tuditanus' colleague, / Marcus' son'' — and he calls him an orator and attributes to him a sweetness of speech, which now indeed is not so common in most men (for by now certain orators bark, they do not speak), but which is surely the greatest praise of eloquence — ``he was called, by those people of long ago, by the men who lived then and passed their span of life, the chosen flower of the people'' — and rightly so;

\ciceroSection{59} for as a man's glory is his talent, so the light of talent itself is eloquence, and the man who excelled in it those men long ago finely called the flower of the people — ``the marrow of Persuasion.'' What the Greeks call \textit{[Greek: Peithō]}, of which the orator is the author, Ennius named Suada [and he means that Cethegus was the marrow of her], so that the goddess whom Eupolis wrote sat upon the lips of Pericles — of her marrow, he said, this orator of ours was made.

\ciceroSection{60} But this Cethegus was consul with Publius Tuditanus in the Second Punic War, and Marcus Cato was quaestor under these consuls, just one hundred and forty years before I was consul; and had this very fact not been known by the single testimony of Ennius, antiquity would have buried this man — as it has perhaps buried many others — under oblivion. What the manner of speaking was in that age can be understood from the writings of Naevius. For under these consuls, as is written in the old records, Naevius died — though our friend Varro, that most diligent investigator of antiquity, thinks an error has been made here, and extends Naevius' life further. For Plautus died under the consuls Publius Claudius and Lucius Porcius, twenty years after those consuls I named before, with Cato as censor.

\ciceroSection{61} This Cethegus, then, was succeeded in time by Cato, who was consul nine years after him. We hold Cato as very ancient — he who died under the consuls Lucius Marcius and Manius Manilius, eighty-six years exactly before I was consul. And in truth I have no one more ancient whose writings, at any rate, I should think worth bringing forward, unless someone happens to be charmed by this very speech of Appius Caecus about Pyrrhus and by some few funeral eulogies of the dead.

\ciceroSection{62} And by Hercules, these last do indeed survive: for the families themselves used to keep them as their ornaments and memorials, both for use, should anyone of the same line have died, and for the memory of their domestic praises and for the illustration of their nobility. Yet by these eulogies the history of our affairs has been made the more faulty. For many things are written in them which were never done: false triumphs, too many consulships, false genealogies too, and crossings-over to the plebs, when men of lower rank were poured into another family of the same name — as if I were to say that I was descended from that Manius Tullius who, a patrician, was consul with Servius Sulpicius in the tenth year after the kings were expelled.

\ciceroSection{63} Cato's speeches, however, are nearly as many as those of Lysias of Attica, whose are, I think, the most numerous of all — for he is an Attic, since he was certainly both born and died at Athens and discharged every duty of a citizen, although Timaeus reclaims him for Syracuse as if by the Lex Licinia and Mucia (a law restoring the wrongly enrolled to their proper community) — and in a certain way there is even some likeness between the two men themselves. They are sharp, elegant, witty, brief; but the Greek is the more fortunate in every kind of praise. For he has his own devotees, who pursue not so much the rich habit of body as the slender forms, men whom — provided only the health be good — the very leanness delights — though in Lysias there are often even muscles, such that nothing could be more vigorous; yet he is certainly, in his whole manner, the leaner — but still he has his admirers, who take great joy in this very subtlety of his.

\ciceroSection{65} Cato, however — which of our orators, of those at least who exist now, reads him? Or who knows him at all? Yet what a man, good gods! I leave aside the citizen, or the senator, or the general; for it is the orator we are seeking in this place: who was weightier than he in praising? more biting in censure? in his judgments more acute? in teaching and in setting forth more subtle? His speeches — more than a hundred and fifty, those at any rate that I have so far found and read — are crowded with brilliant words and brilliant matters. Let them pick out from these the things worthy of remark and praise: all the virtues of the orator will be found in them.

\ciceroSection{66} And then his \textit{Origines} — what flower, what light of eloquence do they not have? Admirers fail him, just as for many ages now they have failed both Philistus of Syracuse and Thucydides himself. For as Theopompus, by the loftiness and elevation of his own style, stands in the way of those men's compressed sentiments — sentiments at times not even clear enough, what with their brevity and their excessive sharpness — just as Demosthenes does the same to Lysias, so this style of the later writers, piled up, as it were, to a greater height, has blocked the light of Cato.

\ciceroSection{67} But there is in our people this ignorance: that these very men who delight in the antiquity of the Greeks and in that subtlety which they call Attic, do not so much as know this same quality in Cato. They wish to be Hyperideses and Lysiases. I commend it.

\ciceroSection{68} But why are they unwilling to be Catos? They say they take joy in the Attic kind of speaking. Wisely, that — and would that they imitated it, not the bones only but the blood as well! Still, what they wish is welcome. — Why, then, is Lysias loved, and Hyperides, while Cato is utterly unknown? His diction is older, and some of his words are rather rough. For that is how they spoke then. Change that — which he, at that time, could not — and add rhythm; and to make the speech more fitting, set the very words in order and, as it were, fit them together, which not even the old Greeks regularly did: then you will set no one before Cato.

\ciceroSection{69} The Greeks think a speech is adorned if they use the changes of words they call \textit{[Greek: tropoi]}, and the figures of thought and of speech they call \textit{[Greek: schēmata]}: it is hardly to be believed how rich and how distinct Cato is in both kinds. Nor indeed am I unaware that this orator is not yet sufficiently polished, and that something more perfect must be looked for — seeing that he is so ancient, by the reckoning of our own times, that no man's writing survives worth reading at all that is older. But in all the arts antiquity is treated with greater honour than in this one art of speaking. For who among those who attend to these lesser things does not understand that the statues of Canachus are too stiff to reproduce the truth; that those of Calamis, hard though they are, are nevertheless softer than Canachus'; that Myron's are not yet brought close enough to the truth, yet are by now such as you would not hesitate to call beautiful; that those of Polyclitus are more beautiful still and by now plainly perfect, as at least they regularly seem to me? The case is similar in painting; in which we praise the shapes and outlines of Zeuxis, Polygnotus, Timanthes, and of those who used no more than four colours; but in Aetion, Nicomachus, Protogenes, Apelles all things are now perfected. And I do not know whether the same thing happens in all other matters too: for nothing is at one and the same time both invented and perfected; nor ought it to be doubted that there were poets before Homer, as can be understood from those very songs which in his pages are sung both at the feasts of the Phaeacians and at those of the suitors. Why, where are our own ancient verses? — ``which once the Fauns and seers used to sing, / when no one had climbed the crags of the Muses, / nor was any man devoted to learned utterance before this one,'' he says of himself — nor does he lie in his boasting: for so the matter stands. For both the Latin \textit{Odyssia} is like some work of Daedalus, and the plays of Livius are not worthy enough to be read a second time.

\ciceroSection{72} And yet this Livius first produced a play in the consulship of Gaius Claudius, son of Caecus, and Marcus Tuditanus, in the very year before Ennius was born, and in the five hundred and fourteenth year after the founding of Rome — as he says, the man whom we follow. For among the writers there is a dispute about the count of years. Accius, however, wrote that Livius was taken prisoner at Tarentum by Quintus Maximus in his fifth consulship, thirty years after that in which both Atticus writes, and we find in the old records, that Livius produced his play;

\ciceroSection{73} and that he produced the play eleven years later, in the consulship of Gaius Cornelius and Quintus Minucius, at the Games of Youth, which Salinator had vowed at the battle of Sena. In which there was so great an error on Accius' part, that under these consuls Ennius would have been forty years old: and if Livius had been his contemporary, then the man who first put on a play would have been somewhat younger than those who had produced many plays before these consuls — both Plautus and Naevius.

\ciceroSection{74} If these things seem less suited to our discussion, Brutus, lay the blame on Atticus, who fired me with a passion for tracing out the ages and the times of illustrious men. But I, said Brutus, both take delight in this charting, so to speak, of the periods, and think that, for the purpose you have set yourself — to distinguish the kinds of orators by their ages — this very precision is well suited.

\ciceroSection{75} You understand rightly, Brutus, I said. And would that those songs still survived which — sung many ages before his own time at banquets, by the several guests in turn, on the praises of famous men — Cato left written in his \textit{Origines}! And yet the Punic War of that poet whom Ennius numbers among the seers and the Fauns delights us like a work of Myron.

\ciceroSection{76} Let Ennius be, as he certainly is, the more perfect; but if he despised that man, as he pretends to, he would not, in tracing out all the wars, have left out that first Punic War, the fiercest of them all. But he himself says why he does it. ``Others,'' he says, ``have written the matter in verse'' — and brilliantly indeed they wrote it, even if less polished than you. Nor in truth ought it to seem otherwise to you, who from Naevius either took much, if you admit it, or, if you deny it, stole it.

\ciceroSection{77} Older in years than this Cato were Gaius Flaminius, Gaius Varro, Quintus Maximus, Quintus Metellus, Publius Lentulus, Publius Crassus, who was consul with the elder Africanus. Scipio himself, we are told, was not without speech. His son, indeed — the one who adopted this younger Scipio from Paullus — would have been held among the foremost as a fluent speaker, had he been sound in body; both his little speeches and a certain history written most charmingly in Greek attest it.

\ciceroSection{78} In the same number was Sextus Aelius, the most expert of all men in the civil law, but also ready for speaking. Of the younger men, there was Gaius Sulpicius Gallus, who of all the nobility was the most devoted to Greek letters; and he both was held in the number of the orators and was, in other respects too, polished and elegant. For by now there was a certain richer and more brilliant habit of speaking. For while this man, as praetor, was holding the games for Apollo, after he had produced the play \textit{Thyestes}, Ennius met his death, in the consulship of Quintus Marcius and Gnaeus Servilius.

\ciceroSection{79} In the same times there was Tiberius Gracchus, son of Publius, who was twice consul and censor, whose Greek speech survives, delivered before the Rhodians; a citizen who, it is agreed, was as weighty as he was eloquent. Publius Scipio Nasica too, who was called Corculum — who likewise was twice consul and censor — they say was held to be eloquent, the son of that Nasica who received the sacred objects; they name also Lucius Lentulus, who was consul with Gaius Figulus; Quintus Nobilior, son of Marcus, who by now, on his father's example, was devoted to the pursuit of letters — the man who even granted citizenship to Quintus Ennius, who had served with his father in Aetolia, when as one of the three commissioners he led out a colony — and Titus Annius Luscus, the colleague of this Quintus Fulvius, they say was not without fluency;

\ciceroSection{80} and even Lucius Paullus, the father of Africanus, easily sustained the part of a leading citizen by his speaking. And in truth, even at that time, while Cato was still alive — who departed life at the age of eighty-five, having indeed in that very year spoken before the people, with the utmost vehemence, against Servius Galba, a speech which moreover he left in writing —

\ciceroSection{81} but in Cato's lifetime many younger orators flourished at one and the same time. For both Aulus Albinus, the man who wrote history in Greek, who was consul with Lucius Lucullus, was a man of letters and fluent; and along with him a certain place was held also by Servius Fulvius, and together with him Servius Fabius Pictor, well versed in law and in letters and in antiquity; and Quintus Fabius Labeo was adorned with much the same praises. For Quintus Metellus, the man whose four sons were of consular rank, was held among the foremost as eloquent — he who spoke for Lucius Cotta when Africanus was the accuser; whose other speeches survive, and whose speech against Tiberius Gracchus is set out in the Annals of Gaius Fannius.

\ciceroSection{82} Then L. Cotta himself was reckoned an old hand; but C. Laelius and P. Africanus were eloquent men of the first rank, and their speeches survive, from which a judgment can be formed about the natural gifts of orators. Yet among these, somewhat older and so a little ahead of them, Ser. Galba beyond dispute stood foremost in eloquence; and he was without doubt the first among the Latins to handle those tasks proper to the orator and, so to speak, sanctioned by his art: to depart from the matter in hand for the sake of ornament, to charm the mind, to stir it, to magnify the case, to make use of appeals to pity and of commonplaces. But somehow the speeches of this man, who is agreed to have excelled in eloquence, are thinner, and smack of antiquity more than do those of Laelius, or of Scipio, or even of Cato himself; and so they have dried up, until now they scarcely come into view at all.

\ciceroSection{83} As for the talent of Laelius and Scipio themselves, although such is their fame that very much is granted to them both, still the praise of speaking is the more brilliant in Laelius. Yet the speech of Laelius on the priestly colleges is no better than any number you please of Scipio's; not that anything is sweeter than that speech of Laelius, or that one could speak of religion with more solemnity, but for all that he is far more old-fashioned and rough than Scipio. And since in speaking men's tastes vary, it seems to me that Laelius takes more pleasure in antiquity, and gladly uses words that are somewhat more archaic.

\ciceroSection{84} But it is the way of men that they are unwilling to have one and the same person excel in several things. For just as no one can aspire to rival Africanus in the glory of war — in that very field, too, we find that Laelius did distinguished service in the war against Viriathus — so, while men assign the first place in talent, in letters, in eloquence, in wisdom, in short, to both alike, they gladly yield the higher place to Laelius. Nor does it seem to me that this was granted to him by the judgment of others only, but by the very concession of the two men between themselves.

\ciceroSection{85} It was altogether the custom then, as in other matters the better, so in this very point the more humane, that men were ready in giving each his due. I remember at Smyrna having heard from P. Rutilius Rufus, who used to say that as a very young man it had been his lot, when by a decree of the Senate P. Scipio and D. Brutus — the consuls, I think — were holding an inquiry into a matter both atrocious and grave. For when a slaughter had been committed in the Silan forest, and well-known men had been killed, and the slave-household was under accusation — and in part, too, the free men of that company which had taken the lease of the pitch-works under the censors P. Cornelius and L. Mummius — the Senate had decreed that the consuls should investigate the affair and pronounce upon it.

\ciceroSection{86} He said that Laelius pleaded the case for the tax-farmers carefully and elegantly, as he was always accustomed to do. When the consuls, the matter heard, had pronounced ADJOURNED on the opinion of their advisory council, and a few days later Laelius had spoken a second time, much more painstakingly and better, and the matter had again in the same way been put off by the consuls — then Laelius, when his clients had escorted him home and given him their thanks and begged him not to wear himself out, spoke as follows: that what he had done he had done for their honour, zealously and with care, but that he judged that case could be defended more weightily and more forcefully by Ser. Galba, because in speaking he was fiercer and keener. And so, on the authority

\ciceroSection{87} of C. Laelius, the tax-farmers carried their case over to Galba; and he, because he had to step into the place of such a man, took it on with modesty and hesitation. There was, as it were, one day in between, the day of adjournment, and the whole of it Galba spent in considering and composing the case; and when the day of the hearing had come and Rutilius himself, at the clients' request, had come early to Galba's house to remind him and bring him round to the time for speaking, he found that, right up until word was brought to him that the consuls had come down, Galba — with everyone shut out — had been at work in a kind of vaulted study with literate slaves, to whom he was accustomed at one and the same moment to dictate, one man one thing, another another. Meanwhile, when word was brought him that it was time, he came out into the house with such a colour and such eyes that you would have thought he had pleaded the case, not rehearsed it.

\ciceroSection{88} He added too — and thought it bore upon the matter — that those writers came out badly battered along with Galba; from which he meant to show that the man had been vehement and fired up, not only in pleading, but even in his preparation. In short: amid great expectation, with very many listening, in the very presence of Laelius, Galba so pleaded that case, with such force and such weight, that hardly any part of the speech was passed over in silence. And so, with many a complaint and much appeal to pity brought to bear, the clients were that day cleared of the charge, with everyone's approval.

\ciceroSection{89} From this account of Rutilius one may surmise that, since there are two supreme merits in an orator — one of arguing subtly so as to instruct, the other of pleading weightily so as to move the minds of the listeners, and since he who inflames the juror accomplishes far more than he who instructs him — the elegance lay in Laelius, the force in Galba. And that very force was then most plainly recognized when, the Lusitanians having been put to death by Ser. Galba as praetor against a pledge given (so it was believed), and L. Libo as tribune of the plebs stirring up the people and bringing forward a bill against Galba much like an act aimed at one man, M. Cato, at the very height of old age, as I said before, urging the law, spoke much against Galba; that speech he set down in his \textit{Origines}, a few days — or was it months? — before he died.

\ciceroSection{90} Then Galba, refusing nothing on his own behalf and imploring the good faith of the Roman people, weeping, commended both his own boys and also the son of C. Gallus, whose orphanhood and tears were wonderfully pitiable on account of the recent memory of his most illustrious father; and he then snatched himself from the flame, the people's compassion having been stirred on the boys' account — just as Cato, the same man, left it written. And I see that Libo himself, too, was no halting speaker, as can be gathered from his speeches.

\ciceroSection{91} When I had said this and paused a little, Brutus said: What then is the reason, if there was so great an excellence in the orator Galba, that none of it shows in his speeches? — though I cannot wonder at this in the case of men who have left nothing whatever in writing. The reason for not writing, I replied, Brutus, is not the same as the reason for not writing so well as one has spoken. For we see that some orators, out of sluggishness, have written nothing, so as not to add domestic labour to the labour of the Forum — for most speeches are written down once already delivered, not in order to be delivered —; that others take no trouble to make themselves better —

\ciceroSection{92} for nothing is so profitable to speaking as writing —; that they do not, however, crave to leave a memorial of their talent for posterity, since they think they have already won glory enough in speaking, and that it will seem even greater if their writings do not come before the judgment of critics; that others, because they think they can speak better than they can write — which generally falls to the lot of men of great native talent but not learned enough, as it did to Galba himself.

\ciceroSection{93} In him, perhaps, it was the force not of talent only but of spirit, and a kind of natural passion, that set him aflame as he spoke and made his speech both impetuous and weighty and vehement; then, when at leisure he had taken up the pen and every motion of the spirit, like a wind, had failed the man, his speech went slack. This does not usually happen to those who pursue a more finished kind of speaking, because good sense never fails the orator, and by using it he can speak and write in one and the same manner; but the ardour of the spirit is not always present, and when it has subsided, all that force and, as it were, flame of the orator is put out.

\ciceroSection{94} For this reason, then, the mind of Laelius seems to breathe even in his writings, while the force of Galba seems to have died. There were also, in the number of middling orators, the brothers L. and Sp. Mummius, the speeches of both of whom survive: Lucius indeed plain and old-fashioned, while Spurius, though no more ornate, was nonetheless more compact, for he was learned in the discipline of the Stoics. There are many speeches of Sp. Albinus. There are also those of L. and C. Aurelius Orestes, whom I see were held in some account among the orators. P. Popilius too

\ciceroSection{95} was at once a distinguished citizen and not without eloquence; his son Gaius, however, was eloquent, and Gaius Tuditanus, refined and polished as he was in his whole way of life and living, had a manner of speech accounted elegant as well. And in the same class was held the man who, when wronged, broke Ti. Gracchus by his patience, a citizen of the utmost steadfastness in the best of causes, M. Octavius. But indeed M. Aemilius Lepidus, the one called Porcina, in much the same period as Galba but a little younger, both was held to be a supreme orator and was, as appears from his speeches, quite a good writer.

\ciceroSection{96} In this man, as a Latin orator, there first appear to me both that smoothness of the Greeks and that rounding-off of words and now, so to speak, an artist's pen. Him two most gifted young men, almost of an age, were accustomed eagerly to hear: C. Carbo and Ti. Gracchus; of whom there will be occasion to speak presently, once I have said a few words about the older men. For Q. Pompeius was no contemptible orator in those days, a man who attained the highest offices on his own recognized merit, without any recommendation of ancestors.

\ciceroSection{97} Then L. Cassius had great power, not by eloquence, but by speaking nonetheless; a man who won popular favour not by generosity, as others did, but by sheer grimness and severity. To his ballot law M. Antius Briso, tribune of the plebs, long held out, with M. Lepidus the consul lending aid; and that affair was held to the discredit of P. Africanus, because Briso was thought to have been brought round from his position by his authority. Then the two Caepiones did much to support their clients by counsel and tongue, but more by authority and influence. There are writings of Sex. Pompeius, not too thin — though like those of the old men — and full of good sense. P. Crassus

\ciceroSection{98} we are told was a much-approved orator in nearly the same period, a man who was strong both in talent and in application, and who had besides certain disciplines within his own household. For he had bound himself by marriage-alliance to that supreme orator Ser. Galba, to whose son Gaius he had given his daughter, and, since he was the son of P. Mucius and had P. Scaevola for a brother, he had learned the civil law at home. In him it is agreed that there was the highest industry and the greatest influence, since he was both consulted very much and pleaded very much.

\ciceroSection{99} Joined to the lifetimes of these men were the two C. Fannii, sons of Gaius and of Marcus; of whom the son of Gaius, who was consul with Domitius, left one speech, on the allies and the Latin name, against C. Gracchus — a thoroughly good and noble one. Then Atticus said: How is that? Is that speech really Fannius's? For when we were boys there was a divided opinion. Some said it was written by C. Persius, a man of letters — the one Lucilius indicates to be very learned —; others said that many nobles had contributed to that speech, each what he could.

\ciceroSection{100} Then I said: I have indeed heard those tales from my elders, but I have never been brought to believe them; and I think that suspicion arose for this reason, that Fannius was held among the middling orators, while the speech was perhaps the very best of all the speeches of that time. But it is not of such a kind that it should seem patched together by several hands — for there is one tone throughout the whole speech and one and the same pen —; nor would Gracchus have kept silent about Persius, when Fannius had reproached him with Menelaus of Marathus and the rest; especially since Fannius was never accounted tongue-tied. For he both defended cases regularly, and his tribunate, conducted by the judgment and authority of P. Africanus, was not obscure. The other, C. Fannius, son of Marcus, son-in-law of C. Laelius, was harsher both in character and in his very manner of speaking.

\ciceroSection{101} He, by the training of his father-in-law — whom, because he had not been co-opted into the college of augurs, he did not much love, especially since Laelius had preferred to him, as son-in-law, the younger Q. Scaevola (though Laelius, excusing himself, said he had given that to the elder daughter, not to the younger son-in-law) — he, by the training of Laelius for all that, had heard Panaetius. His whole faculty in speaking can be discerned from his history, written with no little elegance, which is neither too halting nor perfectly eloquent.

\ciceroSection{102} Mucius the augur, on the other hand, spoke for himself what his own affairs required, as in the case of extortion against T. Albucius. He was not in the number of the orators, but he excelled in understanding of the civil law and in every kind of good sense. L. Caelius Antipater the historian was, as you see, brilliant for those times, very skilled in the law, and the teacher of many, among them L. Crassus.

\ciceroSection{103} Would that in Ti. Gracchus and Gaius Carbo there had been such a mind for governing the commonwealth well as there was talent for speaking well: surely no one would have surpassed these men in glory. But of the two, the one — on account of a most turbulent tribunate, which out of resentment at the treaty of Numantia he had embarked upon, angry at the loyalists — was killed by the very commonwealth; the other — on account of his unbroken fickleness in the popular line — by a voluntary death rescued himself from the severity of his judges. But each was a supreme orator.

\ciceroSection{104} And this we say on the testimony of our fathers' memory; for we have speeches both of Carbo and of Gracchus, not yet splendid enough in their words, but sharp and most full of good sense. Gracchus was, by the diligence of his mother Cornelia, taught from boyhood and schooled in Greek letters. For he always had choice teachers brought from Greece, among them, while still a young man, Diophanes of Mytilene, in those days the most eloquent man of Greece. But he had only a brief span for enlarging and displaying his talent;

\ciceroSection{105} Carbo, whom life held out longer, became known in many trials and cases. The men of good sense who had heard him — among them our friend L. Gellius, who used to tell that he had been his tent-companion during his consulship — said that he was a melodious orator, fluent and keen enough, and at the same time both vehement and very sweet and most witty; he added that he was industrious too, and painstaking, and accustomed to put much labour into exercises and rehearsals.

\ciceroSection{106} He was held the best advocate of those times, and while he commanded the Forum more trials began to take place. For the standing courts were established while he was a young man, courts which before had not existed at all; for L. Piso, tribune of the plebs, was the first to bring forward a law on extortion, in the consulship of Censorinus and Manilius — Piso himself, too, both pleaded cases and was either the proposer or the opposer of many laws, and he left both speeches, which have by now faded away, and annals written quite thinly —; and the people's trials, at which Carbo was present, now stood more in need of an advocate, once the ballot had been introduced; that law L. Cassius brought forward in the consulship of Lepidus and Mancinus.

\ciceroSection{107} Your kinsman too, D. Brutus, son of Marcus — as I used to hear from his intimate, the poet L. Accius — both used to speak with no lack of polish and was learned, for those times, in Latin letters and indeed in Greek as well. The same qualities Accius attributed also to Q. Maximus, grandson of L. Paullus; and indeed, even before that Maximus, of the Scipio under whose leadership, as a private citizen, Ti. Gracchus was slain, he said that he was, as vehement in all things, so also keen in speaking.

\ciceroSection{108} Then too P. Lentulus, that leading man, is said to have had eloquence enough for whatever the commonwealth required, and no more. In the same period L. Furius Philus was reckoned to speak Latin very well, and with more learning than the rest; P. Scaevola with great good sense and acuteness; M'. Manilius a little more fully, and not much less sensibly. The speech of Appius Claudius was fluent, but rather too hot. There were also of some account M. Fulvius Flaccus and C. Cato, the son of Africanus' sister, middling orators — though Flaccus left writings, but as a man devoted to letters does. Flaccus' rival was P. Decius, no tongue-tied man, but, like his life, turbulent also in his speaking.

\ciceroSection{109} M. Drusus, son of Gaius — who in his tribunate broke C. Gracchus, his colleague and tribune for the second time — was a man weighty both in speech and in authority; and most closely joined to him was his brother C. Drusus. Your kinsman too, Brutus, M. Pennus, harried C. Gracchus wittily in his tribunate, being a little older. For in the consulship of M. Lepidus and L. Orestes, Gracchus was quaestor and Pennus tribune — the son of that Marcus who was consul with Q. Aelius; but Pennus, with all the highest hopes before him, died after his aedileship. As for T. Flamininus, whom I myself saw, I have heard nothing except that he spoke Latin with care.

\ciceroSection{110} To these are joined C. Curio, M. Scaurus, P. Rutilius, C. Gracchus. Of Scaurus and Rutilius one may speak briefly: neither of them won the praise of a supreme orator, and both were engaged in many cases. There was, in certain men one praises — even if they were not of the greatest talent — a creditable diligence all the same; though in these two it was not talent altogether, but oratorical talent, that was wanting. For it is no use to see what should be said, unless you can say it fluently and with charm; and not even that is enough, unless what is said is given more savour by voice, by countenance, and by gesture.

\ciceroSection{111} Need I say that learning is required? Without it, even if something is said well with nature's help, still, because it happens by chance, it can never be reliably at hand. In Scaurus' speaking — a man wise and upright — there was the highest gravity and a certain natural authority, so that you would think he was not pleading a case but giving testimony, even when he spoke on a defendant's behalf.

\ciceroSection{112} This manner of speaking seemed only moderately suited to advocacy, but for delivering an opinion in the Senate, where he was the leading man, suited best of all; for it conveyed not good sense alone but — what most held the matter together — credit and good faith. He had this from nature herself, what from teaching one could not easily get; though of this very thing too, as you know, there are rules. There survive both his speeches and three books on the conduct of his own life, written for L. Fufidius — really useful books, which no one reads. Yet they read the life and training of Cyrus, a splendid work, no doubt, but neither so suited to our affairs nor, for all that, to be set above the praises of Scaurus. Fufidius himself, too, was of some account among advocates. Rutilius, on the other hand, worked in a certain grim and severe kind of speaking. Each was by nature vehement and fierce; and so, when they had stood together for the consulship, not only did the one who had suffered defeat charge his successful rival with bribery, but Scaurus, once acquitted, in turn summoned Rutilius into court. Rutilius was a man of much exertion and much industry, the more welcome because he also bore the great service of giving legal responses.

\ciceroSection{114} His speeches are spare; much in them is admirable on points of law. He was a learned man, schooled in Greek letters, a pupil of Panaetius, almost a finished Stoic — and you know that their manner of speaking, keen and full of art though it is, is nonetheless thin and not well fitted to win the assent of the crowd. And so that conviction proper to this school — the philosophers' settled opinion of themselves — was found firm and unshaken in this man.

\ciceroSection{115} When, an utterly blameless man, he had been summoned into court — by which trial, as we know, the commonwealth was shaken to its foundations — although there were at that time men of the greatest eloquence, the consulars L. Crassus and M. Antonius, he was willing to call upon neither of them. He spoke for himself, and C. Cotta said a few words, since he was his sister's son — and Cotta indeed spoke as an orator, though he was still quite young — and Q. Mucius spoke clearly and finely, as was his way, but by no means with that force and abundance which such a trial and the magnitude of the case demanded.

\ciceroSection{116} We have, then, Rutilius among the Stoic orators, Scaurus among those of the old stamp; yet let us praise them both, since through them not even these kinds went without their share of oratorical praise in the state. For I would have it that, as on the stage, so also in the Forum, not only those be praised who use a quick and difficult style, but also those whom they call \textit{stationary} actors — whose mark is that simple, unforced truth in their performance.

\ciceroSection{117} And since mention has been made of the Stoics, there was at that time Q. Aelius Tubero, grandson of L. Paullus; in no count of orators, but austere in life and in keeping with the school he cultivated, indeed a little harsher than that. In his tribunate he ruled — against the testimony of his own uncle P. Africanus — that the augurs had no exemption that would keep them from serving on juries; but, as in life so in speech, he was harsh, uncultivated, rough; and so he could not match the offices his ancestors had held. Yet he was a steadfast and brave citizen, and especially galling to Gracchus, as Gracchus' speech against him shows. There exist also speeches of Tubero against Gracchus; he was middling in speaking, most learned in argument. Then Brutus said:

\ciceroSection{118} How clearly I see that the same thing happens among our own men as among the Greeks — that nearly all the Stoics are most acute in disputation and do it by art, and are well-nigh architects of words, yet these very men, when carried over from disputing to speaking, are found destitute. One I except, Cato, in whom — a most perfect Stoic — I should not miss the highest eloquence, which I see was slight in Fannius, not even great in Rutilius, and in Tubero nonexistent.

\ciceroSection{119} And I said: Not without reason, Brutus, since all the care of those men is spent on dialectic, and that ranging, copious, manifold kind of speaking is not employed. But your uncle, as you know, has from the Stoics what was to be sought from them; yet he learned to speak from teachers of speaking, and trained himself in their manner. For if everything were to be sought from philosophers, speech would be more conveniently shaped by the teachings of the Peripatetics.

\ciceroSection{120} The more do I approve your judgment, Brutus, in having followed the sect of those philosophers — that is, of the Old Academy — in whose teaching and precepts the method of reasoning is joined with charm and abundance of speech; although that very practice of the Peripatetics and Academics, in the matter of speaking, is such that by itself it can neither make the orator complete, nor can the orator be complete without it. For as the speech of the Stoics is more tightly bound and somewhat more contracted than the people's ears require, so theirs is freer and broader than the usage of the courts and the Forum allows. For who is richer in speaking than Plato?

\ciceroSection{121} Jupiter, the philosophers say, if he spoke Greek, would speak as Plato does. Who is more sinewy than Aristotle, who sweeter than Theophrastus? Demosthenes is said to have read Plato closely, and even to have heard him — and this is plain from the kind and the grandeur of his words; he himself says as much about himself in a certain letter — yet his speech, when transferred into philosophy, would seem, so to say, more combative, and theirs, brought into the courts, more peaceable.

\ciceroSection{122} Now let us trace, if it please you, the ages and the gradations of the remaining orators. Indeed we wish it, said Atticus, and earnestly too — that I may answer for Brutus as well. Curio, then, of about the same period, was a notably distinguished orator, whose talent can be judged from his speeches: there are several, and there is the famous speech for Ser. Fulvius, on a charge of incest. When we were boys this was thought the best of all such speeches, which now scarcely shows itself in this throng of new volumes.

\ciceroSection{123} Splendidly, said Brutus, do I know who has produced that throng of volumes. And I said: I understand whom you mean, Brutus; for surely we have brought something good to the young, a manner of speaking grander than there had been, and more ornate — and perhaps we have done harm too, in that the old speeches, after ours, came to be left unread, not by me indeed, for I set them above my own, but by most readers. Count me, he said, among the many; though I see there is much I must now read at your prompting, which before I despised.

\ciceroSection{124} And yet, I said, this praised speech on incest is childish in many passages: on love, on torture, on rumour, the passages are quite empty — yet tolerable still in an age when the ears of our countrymen were not yet worn smooth, nor the state schooled. Curio wrote some other things besides, and said much, and that with distinction, and was of the number of advocates; so that I wonder, since both length of life had served him and standing had not been lacking, that he was never consul.

\ciceroSection{125} But look — here in our hands is a man both of the most outstanding talent and of burning zeal, learned from boyhood: C. Gracchus. Do not suppose, Brutus, that anyone was fuller or richer in speaking. And he said: So I judge entirely, and of the older men I read almost him alone. On the contrary, I said, Brutus, you should read him outright. For in his untimely death the Roman state and Latin letters suffered a loss.

\ciceroSection{126} Would that he had wished to render to his country the devotion he rendered to his brother! How easily, with such talent, had he lived longer, would he have attained the glory of his father, or even of his grandfather! In eloquence, I rather think, he would have had no equal. He is grand in his words, wise in his thoughts, weighty in his whole manner. The final hand was not laid to his works; much is splendidly begun, but not fully finished. This orator is to be read, Brutus, if any is, by the young; for he can not only sharpen but also nourish the mind.

\ciceroSection{127} To this age there succeeded C. Galba, son of that most eloquent man Servius, and son-in-law of P. Crassus, eloquent and skilled in law. Our fathers praised this man, and favoured him too for his father's memory, but he fell in mid-course. For under the Mamilian bill (an inquiry into the Jugurthine conspiracy), amid the odium of that affair, when he had spoken in his own defence, he was crushed. There survives his peroration, which is called the epilogue; it stood in such honour with us as boys that we even learned it by heart. This man, a member of the college of priests, was the first since the founding of Rome to be condemned in a public trial.

\ciceroSection{128} P. Scipio, who died during his consulship, did not speak much, indeed nor often, but in speaking Latin he was a match for anyone, and in wit and humour surpassed all. His colleague L. Bestia, who began his tribunate with good beginnings — for by his own bill he restored P. Popilius, driven out by the violence of C. Gracchus — a man both keen and not without eloquence, had a grim end to his consulship. For under an odious law, the Mamilian inquiry, the Gracchan jurors brought down the priest C. Galba and four consulars: L. Bestia, C. Cato, Sp. Albinus, and that most distinguished citizen L. Opimius, the slayer of Gracchus — acquitted by the people though he had stood against the people's will.

\ciceroSection{129} Unlike him in his tribunate and in all the rest of his life, a wicked citizen, was C. Licinius Nerva — not without eloquence. C. Fimbria, of about the same period, but more advanced in years, was reckoned, so to speak, a muddy advocate, harsh, abusive; in his whole manner a little too hot and excitable, yet by his diligence, his strength of spirit, and his life a good authority in the Senate; likewise a tolerable advocate, and not unschooled in the civil law, and free both in his virtue and in his very style of speaking. His speeches we read as boys, though now we can scarcely find them.

\ciceroSection{130} And of keen talent and elegant conversation, but troubled in health, was C. Sextius Calvinus; who, although when the pains in his feet had let up he did not fail to appear in cases, nonetheless did not do so often. And so men used his counsel when they wished, his advocacy when it was possible. In the same period there was M. Brutus — in whom there was great disgrace to your line, Brutus — who, though he bore so great a name and had a father most excellent and most learned in the law, made a regular trade of prosecution, as Lycurgus did at Athens. He did not seek office, but was a vehement and galling prosecutor, so that you could easily discern that a certain natural good of the stock had degenerated through the vice of a corrupted will.

\ciceroSection{131} And at that same time there was a prosecutor of the plebs, L. Caesulenus, whom I heard when he was already an old man, when he had sued L. Sabellius for a penalty under the Aquilian law (concerning wrongful damage). I would not have made mention of a man almost of the lowest rank, did I not judge that I have heard no one who spoke with more suspicion-stirring or more incriminating force. Learned also in Greek was T. Albucius — or rather plainly a Greek. I speak as I think; but one may judge from his speeches. He had been at Athens as a young man, and had turned out a finished Epicurean — a school least suited to speaking. Now Q. Catulus was schooled not in that old fashion but in our own, unless anything can be made more perfect than this. Wide learning was his, and the highest grace not only of life and nature but of speech as well, a certain unspoiled purity of the Latin tongue; which can be discerned both from his speeches and most easily from that book on his consulship and his own achievements which, composed in the soft, Xenophontic style of discourse, he sent to the poet A. Furius, his intimate friend; a book no better known than those three I spoke of before, the books of Scaurus.

\ciceroSection{133} Then Brutus said: To me, indeed, neither that book is known nor those — but this is my own fault, for they never fell into my hands — yet now I shall both get them from you and seek out such things more carefully hereafter. There was, then, in Catulus a Latin diction; and this praise of speaking, no small one, has been neglected by most orators. As for the sound of his voice and the sweetness with which he pronounced his letters, since you knew his son, do not wait to hear what I might say. And yet the son was not in the number of orators, but in delivering an opinion there was in him, together with good sense, a certain elegant and learned style of speech.

\ciceroSection{134} The father Catulus himself, however, was not held a leading man in the count of advocates, but he was such that, when you heard certain men who were then preeminent, he seemed inferior, but when you heard him by himself, without comparison, you would not only be content, but would seek nothing better.

\ciceroSection{135} Q. Metellus Numidicus and his colleague M. Silanus used to speak on public affairs in a way that was enough for such men and for consular dignity. M. Aurelius Scaurus did not speak often, but spoke with polish; in Latin indeed he spoke with especial elegance. The same praise for good diction belonged to A. Albinus; for Albinus the flamen was even reckoned among the eloquent. Q. Caepio too, a keen and brave man, for whom the fortune of war became a charge, and the people's hatred his ruin.

\ciceroSection{136} In that period too the two Memmii, Gaius and Lucius, were middling orators, but sharp and bitter as prosecutors; and so they hauled many men into capital trials, though they did not often speak for defendants. Spurius Thorius had force enough in the popular style of speaking — the man who relieved the public land of a flawed and useless rent-law. Marcus Marcellus, the father of Aeserninus, was not, indeed, among the leading advocates, but still among those ready and not unpractised in speaking — as was his son

\ciceroSection{137} Publius Lentulus. Lucius Cotta too, of praetorian rank, belonged among the middling orators; in the renown of speaking he had not advanced very far, but deliberately, both in his words and even in his very tone, somewhat rustic as it was, he pursued and imitated antiquity. And in this very Cotta, and in a good many others, I am aware that I have entered into the roll of orators men not so very eloquent, and shall enter more. For my plan is to gather all who have discharged this office in the state, so far as to hold the place of orators; and from what I am about to say one may judge what their ascent was, and how difficult, in everything, is the perfection and completeness of the best.

\ciceroSection{138} For how many orators have now been recalled, and how long we linger in the count of them — and yet, slowly and with difficulty, just as a while ago we came to Demosthenes and Hyperides, so now we have come to Antonius and Crassus! For my judgment is this: that these were the greatest orators, and that in them, for the first time, the abundance of speaking in Latin was made equal to the glory of the Greeks.

\ciceroSection{139} To Antonius everything came to mind; and each thing was set in its own place, where it could do most good and have most force — as the cavalry, the infantry, and the light-armed are stationed by a general, so by him things were posted in the most advantageous parts of a speech. His memory was supreme, and there was no hint of premeditation; he always seemed to come unprepared to speak, yet he was so prepared that the jurors, while he spoke, sometimes seemed to have been not prepared enough to guard themselves.

\ciceroSection{140} His words themselves were not, indeed, of the most elegant diction — and so he lacked the renown of speaking with nicety, though he did not speak at all foully — but rather of that kind which is the proper glory of an orator in his words. For to speak good Latin is itself, as I said a little before, a thing to be set in high renown, yet not so much on its own account as because it is neglected by most: for it is not so splendid to know Latin as it is shameful not to know it, and it seems to me the property not so much of a good orator as of a Roman citizen. But still, Antonius, both in choosing his words — and that not so much for charm as for weight — and in arranging them and binding them tight into a period, directed everything to method and, as it were, to art; and far more so this same care in the ornaments and figures of his thoughts. Because Demosthenes excels all in this kind, he has therefore been judged by the learned the prince of orators. For the figures which the Greeks call \textit{[Greek: schēmata]} are what most adorn an orator, and they carry weight not so much in coloring the words as in lighting up the thoughts. But while these things were great in Antonius, his delivery was singular; and if it is to be divided into gesture and voice, his gesture was one not picturing the words but according with the thoughts: hands, shoulders, sides, the stamp of the foot, his stance, his stride, and his every motion in accord with the words and thoughts; his voice steady, but by nature somewhat hoarse. Yet this single fault he turned to good for himself.

\ciceroSection{142} For it had something plaintive about it, well suited both to winning belief and to stirring pity: so that in him that saying seemed to hold true which they report of Demosthenes, who, when someone asked what was first in speaking, answered "delivery"; what second, the same; what third, the same again. Nothing penetrates more deeply into men's minds, and shapes, forms, and bends them, and makes orators seem such as they themselves wish to seem.

\ciceroSection{143} Some called Lucius Crassus equal to this man, others set him above. One thing at least all judged so: that there was no one who, with either of these two as his advocate, could wish for any other man's talent. For my part, though I grant Antonius as much as I said above, yet I hold that nothing could have been made more perfect than Crassus. His was the highest gravity, and joined to that gravity was a wit and urbanity proper to an orator and not a buffoon's; an elegance accurate and careful, without fuss, in speaking Latin; in argument a marvelous power of unfolding; and when the civil law was debated, or what is fair and good, an abundance of arguments and parallels.

\ciceroSection{144} For just as Antonius had an incredible force in moving conjecture, or lulling suspicion, or rousing it, so in interpreting, in defining, in setting out equity nothing was richer than Crassus; and this was recognized often on other occasions and especially before the centumviri, in the case of Manius Curius.

\ciceroSection{145} For he then said so much, against the letter and for the fair and good, that he overwhelmed Quintus Scaevola — a man most acute, and most thoroughly prepared in the law on which that case turned — with his abundance of arguments and examples; and that case was then so argued by these advocates, equals in age and now of consular rank, each defending the civil law from the opposite side, that Crassus was reckoned the most learned in law of the eloquent, and Scaevola the most eloquent of those learned in law. And Scaevola, though he was most sharp in working out what in law or in equity was true or was not, was yet marvelously fit in words to the matter, with the highest brevity.

\ciceroSection{146} So let him be for us, in this kind of interpreting, explaining, and expounding, a marvelous orator, such that I have seen nothing like him; but in amplifying, in adorning, in refuting, an orator more to be feared as a critic than admired. But let us return to Crassus. Then Brutus said:

\ciceroSection{147} "Although I seemed to myself," he said, "to have come to know Scaevola well enough from the things I often used to hear from Gaius Rutilius — whom I knew through his intimacy with our Scaevola — yet that great renown of his speaking was not known to me; and so I have taken pleasure in learning that a man so polished, and so excellent in talent, lived in our commonwealth."

\ciceroSection{148} At this I said: "Do not suppose, Brutus, that there was anything in our state more outstanding than these two. For as I said a little before, the one was the most eloquent of the lawyers, the other the most learned in law of the eloquent; so in all else they were so unlike each other that you could not decide which of the two you would rather resemble. Crassus was the most sparing of the elegant, Scaevola the most elegant of the sparing; Crassus, in the height of geniality, had also gravity enough, and Scaevola, amid much gravity, did not lack geniality.

\ciceroSection{149} "One might go on so through everything; but I fear these things may seem to be invented, so as to be said by me in a certain balanced way — yet the matter stands so. Since all virtue is, as your old Academy holds, Brutus, a mean, each of these two wished to follow a certain middle course; but it so fell out that each had a share of the other's distinction, while each kept his own entire."

\ciceroSection{150} Then Brutus said: "While from your account I seem to myself to have come to know Crassus and Scaevola well, so too, thinking of you and of Servius Sulpicius, I judge there to be a certain likeness between you and them." "How so?" I said. "Because," he said, "you seem to me to have wished to know just so much of the civil law as was enough for an orator, and Servius to have taken up just so much eloquence that he could easily defend the civil law; and your ages, like theirs, differ in nothing, or in nearly nothing."

\ciceroSection{151} And I said: "About myself there is no need to speak; but of Servius both you speak rightly and I shall say what I think. For I could not easily name anyone who brought more zeal than he both to speaking and to all the disciplines of good learning. For we were together in the same exercises in our early years, and afterward he too set out with me to Rhodes, that he might be the better and the more learned; and from the time he returned, he seems to me to have preferred to be first in the second art rather than second in the first. And I am inclined to think he could have been the equal of the leaders; but perhaps he preferred what he did attain — to be by far, of all, not only of his own age but even of those who had ever been, the prince in the civil law."

\ciceroSection{152} At this Brutus said: "Do you say so? Do you set our Servius even above Quintus Scaevola?" "This is my judgment, Brutus," I said: "that a great practice of the civil law was found both in Scaevola and in many others, but the art of it in this one man alone; which the mere knowledge of the law itself would never have achieved, had he not besides learned that art which teaches one to distribute a whole subject into parts, to unfold the hidden by defining, to make plain the obscure by interpreting, first to see what is ambiguous, then to distinguish it, and last to hold a rule by which the true and the false may be judged, and which things, given which premises, do or do not follow.

\ciceroSection{153} "For he brought this art, greatest of all arts, as a kind of light to bear on the things that by others were either answered or argued in confusion." "You seem to me to mean dialectic," he said. "You understand rightly," I said; "but he joined to it also a knowledge of letters and an elegance of speaking, which can be seen most easily from his writings, to which there are none like."

\ciceroSection{154} And since for the sake of learning he had given his attention to two of the most expert men, Lucius Lucilius Balbus and Gaius Aquilius Gallus, he surpassed by subtlety and care the ready, prepared quickness of Gallus — an acute and practiced man — in pleading and in answering; and he conquered, by the dispatch and finishing of business, the considered slowness of Balbus, a learned and erudite man, in both. So he both has what each of them had, and supplied what each lacked.

\ciceroSection{155} And so, just as Crassus seems to me to have acted more wisely than Scaevola — for the latter eagerly took up cases, in which he was surpassed by Crassus, while the former did not wish to be consulted on law, lest in anything he be inferior to Scaevola — so Servius acted most wisely: since the two civil and forensic arts had the most of both renown and favor, he brought it about that he excelled all in the one, and from the other took up just so much as was enough both for defending the civil law and for securing consular dignity."

\ciceroSection{156} Then Brutus said: "Just so, and I thought as much before — for I heard him recently, with zeal and often, at Samos, when I wished to learn from him our pontifical law, in what part it was joined with the civil law — and now I confirm my judgment far more by your testimony and verdict; and at the same time I rejoice in this, that your equality of age, and your equal grades of office, and the near neighborhood, so to speak, of your arts and pursuits, are so far from the disparagement and envy that usually mangles most men, that they seem not only not to fester your goodwill but even to foster it. For such as I perceive you toward him, such have I come to know him in will and judgment toward you. And so I grieve that the Roman people has been deprived so long both of his counsel and of your voice — which, grievous as it is in itself, is far more so to one who considers into whose hands these things have passed: not handed over, but, I know not by what chance, come down to them." Here Atticus said: "I had said from the beginning that we should be silent about the commonwealth; so let us do so. For if we mean to long for each lost thing in that fashion, we shall find an end not only of complaining but not even of mourning."

\ciceroSection{158} "Let us go on, then," I said, "to what remains, and pursue the order we set ourselves. Crassus, then, came prepared; he was awaited, he was heard; from the very start, since with him everything was always wrought with care, he seemed worthy of the expectation. There was no great tossing of the body, no swaying of the voice, no pacing about, no frequent stamp of the foot; his speaking was forceful and at times angry, and full of just indignation; his wit was abundant and joined with gravity; and — what is hard — he was at once most ornate and most concise; in the give-and-take of debate, moreover, he found no equal.

\ciceroSection{159} He was engaged in nearly every kind of case; early he came into the place of the leading orators. He prosecuted Gaius Carbo, a most eloquent man, while still quite young; and he won not only the highest praise but even admiration for his talent.

\ciceroSection{160} Afterward he defended the Vestal Licinia, when he was twenty-seven years old. In that very case he was most eloquent, and he left behind certain written portions of that speech. As a young man he wished to handle something on the popular side in the colony of Narbo, and himself to lead out that colony, as he did; there survives a speech of his on that law — more mature, so to speak, than that age would bear. Then there were many cases; but his tribunate was so silent that, had he not during that magistracy dined at the house of the auctioneer Granius, and had Lucilius not told us this twice, we should not know he had been a tribune of the people."

\ciceroSection{161} "Just so," said Brutus; "but of Scaevola's tribunate too I seem to have heard nothing, and I believe he was a colleague of Crassus." "In all the other magistracies, indeed," I said, "but he was tribune the year after, and with Scaevola sitting on the Rostra, Crassus spoke in support of the Servilian law; for he held the censorship without Scaevola — since no one of the Scaevolas ever sought that magistracy. But when this speech of Crassus was published — which I am quite sure you have often read — he was then thirty-four years old, and he was older than I by just as many years. For he spoke in support of that law under the very consuls in whose year I was born, he himself having been born under the consuls Quintus Caepio and Gaius Laelius, younger than Antonius by just three years. I have set this down so that one may mark in what age the first ripeness of speaking Latin appeared, and may understand that it had already been carried nearly to the summit, so that scarcely anyone could add anything to it — except one who came better furnished from philosophy, from the civil law, from history."

\ciceroSection{162} "Will there be," said Brutus, "or is there already, that man you await?" "I do not know," I said. "But there is also a speech of Lucius Crassus from his consulship, in defense of Quintus Caepio — joined with a speech, long if taken as a eulogy, but as a speech short; and last his speech as censor, which he made in his forty-eighth year. In all these there is a certain color of truth without any paint; and indeed that compass and circuit of words — if it pleases you so to call it a period, \textit{[Greek: periodon]} — was with him contracted and short, and he more gladly broke up his speech into certain members, which the Greeks call \textit{[Greek: kōla]}."

\ciceroSection{163} At this point Brutus said: "Since you praise those orators so highly, I could wish that Antonius had cared to write something besides that little book on the theory of speaking — a very meager one — and that Crassus had cared to write more: for they would have left to all the memory of themselves, and to us besides the discipline of speaking. As for Scaevola, we have come to know his elegance of speaking well enough from those speeches he did leave."

\ciceroSection{164} And I said: For me, from boyhood, that speech on the Caepian law was a kind of schoolmistress; in it the authority of the senate, on whose behalf those things are spoken, is set off with honour, while ill-feeling is stirred against the faction of the jurors and the prosecutors, against whose power one had then to speak in the popular cause. Many things in that speech were said with weight, many gently, many harshly, many wittily; and more was spoken than was written down, as can be gathered from certain headings that are laid out but not developed. That censorial speech itself, against his colleague Gnaeus Domitius, is not a speech but a kind of heads of the matter, a memorandum of the oration somewhat more fully set out. For never has any altercation been carried on with louder shouting.

\ciceroSection{165} And indeed in this man, too, the popular style of speaking was outstanding; Antonius's manner of speaking was far better suited to the courts than to public meetings. At this point I do not pass over Domitius himself. For though he was not in the number of orators, still I hold that there was enough of eloquence and ability in him to maintain both the dignity of a magistrate and the standing of a man of consular rank. The same I would say of Gaius Caelius: that there was in him the greatest industry and the greatest virtues, and just enough eloquence — enough for his friends in private affairs, enough for his own standing in public life. At the same time Marcus Herennius was counted among the middling orators who spoke pure Latin and with care; yet he defeated Lucius Philippus — a man of the highest nobility, joined to him by kinship, by fellowship and by colleagueship, and possessed besides of the highest eloquence — in the canvass for the consulship. At the same time Gaius Claudius, though he was great by reason of his supreme nobility and his singular power, nonetheless also brought a certain modest measure of eloquence. Of much the same period was the Roman knight

\ciceroSection{167} Gaius Titius, who in my judgment seems to have come as far as a Latin orator could fairly come without Greek letters and without much practice. His speeches have such point, such wealth of examples, such urbanity, that they seem almost to have been written in the Attic manner. These same points he carried over into his tragedies — shrewdly enough, indeed, but with too little of the tragic. Lucius Afranius the poet was eager to imitate him, a most pointed man, and in his plays, as you know, even eloquent.

\ciceroSection{168} There was also Quintus Rubrius Varro, who was judged a public enemy by the senate together with Gaius Marius, a keen and forceful prosecutor, in that line quite acceptable. A kinsman of ours, learned in Greek letters and made for speaking, was Marcus Gratidius, a close intimate of Marcus Antonius — while serving as his prefect in Cilicia he was killed; he was the man who prosecuted Gaius Fimbria, the father of Marcus Marius Gratidianus.

\ciceroSection{169} And among our allies and the Latins these too were held to be orators: Quintus Vettius Vettianus of the Marsi, whom I knew myself, a prudent man and brief in speaking; Quintus and Decimus Valerius of Sora, my neighbours and friends, not so admirable in speaking as learned both in Greek and in Latin letters; Gaius Rusticelius of Bononia, a man both well-practised and by nature fluent; but the most eloquent of all outside this city was Titus Betucius Barrus of Asculum, by whom there are several speeches delivered at Asculum, and one at Rome against Caepio, a famous one indeed — to which speech Aelius replied in Caepio's name, the man who kept writing speeches for many but was never himself an orator.

\ciceroSection{170} And among our ancestors I find that the man held most eloquent out of Latium was Lucius Papirius of Fregellae, of about the age of Tiberius Gracchus the son of Publius; there is also a speech of his, delivered in the senate on behalf of the people of Fregellae and the Latin colonists. Then Brutus said: What, then, do you grant to those foreign quasi-orators? What should I grant, I said, but the very same as to those of the city? — except for one thing, that their speech is not, so to speak, coloured with a certain urbanity. And Brutus:

\ciceroSection{171} What, after all, he said, is that colour of urbanity? I do not know, I said; I only know that there is such a thing. You yourself, Brutus, will understand it once you have come into Gaul; there you will hear certain words not current at Rome — but these can be changed and unlearned. The greater thing is this: that in the voices of our own orators there is a certain ring and resonance that is more of the city. Nor does this appear in orators only, but in everyone else as well.

\ciceroSection{172} I myself remember Titus Tinca of Placentia, a most facetious man, contending in raillery with our friend Quintus Granius the auctioneer. Is that the one, said Brutus, of whom Lucilius says so much? The very man; but though Tinca said no fewer laughable things, Granius would overwhelm him with some homegrown flavour I cannot name; so that I no longer wonder at what is said to have happened to Theophrastus — when he was asking a certain old woman the price of something, and she answered, and added, "Stranger, no less," he took it ill that he could not escape the look of a stranger, when he had spent his life at Athens and spoke the best of all men. So, as I think, there is among our people a certain ring of city-speech, just as there of the Attic. But let us go home — that is, let us return to our own.

\ciceroSection{173} Next, then, after the two supreme men, Crassus and Antonius, came Lucius Philippus — yet next at a long interval. And so, although no one came between to set himself above him, I would call him neither second nor third. For in a chariot-race I would not count second or third the man who has barely come out of the starting-gates when the first has already taken the palm; nor among orators the man who is so far from the first that he scarcely seems to be on the same course. Yet there were qualities in Philippus which one who looked at them without comparing them with those men would call great enough: the greatest freedom in his speech, much wit; quite copious in finding his thoughts, fluent in setting them out; he was also, considering those times, among the foremost trained in Greek learning, and in altercation witty, with a certain sting and a touch of abuse.

\ciceroSection{174} Joined nearly to the age of these men was Lucius Gellius, an orator not so saleable — though you could not tell what was wanting in him; for he was neither unlearned, nor slow to invent, nor forgetful of Roman affairs, and fluent enough in words; but his age had fallen among great orators. Still, he gave much and useful service to his friends, and lived so long that he was bound up with orators of many ages.

\ciceroSection{175} Much engaged in cases too, at about the same period, was Decimus Brutus, the one who was consul with Mamercus, a man learned both in Greek and in Latin letters. Lucius Scipio also spoke, not without skill, and Gnaeus Pompeius the son of Sextus held a certain rank. For his brother Sextus had brought a most outstanding ability to the highest mastery of the civil law and to a perfect knowledge of geometry and of Stoic matters. So too in law, and before these men, Marcus Brutus, and a little after him Gaius Billienus, a man great in his own right, had come out supreme by nearly the same method; he would have been made consul, had he not fallen upon the Marian consulships and upon those narrow straits of canvassing.

\ciceroSection{176} The eloquence of Gnaeus Octavius, which had gone unnoticed before his consulship, was, in his consulship, vehemently approved in many public meetings. But from those who were only in the number of speakers, not of orators, let us now return to the orators. I agree, said Atticus; for you seemed to want to hunt out the eloquent, not the merely diligent.

\ciceroSection{177} In festivity, then, and in wit, I said, Gaius Iulius the son of Lucius surpassed all his elders and all his equals, and was an orator by no means vehement, indeed, but no one was ever more richly seasoned with urbanity, none with charm, none with sweetness. There are several speeches of his, from which, as from his tragedies, his smoothness can be discerned — and not without sinew.

\ciceroSection{178} His equal was Publius Cethegus, who had speech enough for matters of state — for he held the commonwealth so well and had come to know it so thoroughly that in the senate he attained the authority of the consulars; but in public causes he seemed nothing, in private ones a sufficiently practised old hand. In private causes there was Quintus Lucretius Vispillo, both keen and skilled in the law; for Ofella was fitter for public meetings than for the courts. Prudent too was Titus Annius of the Velina tribe, and in causes of that kind a quite tolerable orator. In the same kind of cases there was much in Titus Iuventius — too slow, indeed, in speaking, and almost frigid, but both shrewd and crafty in catching his adversary, and besides neither unlearned and possessed of a great understanding of the civil law.

\ciceroSection{179} His hearer Publius Orbius, about my own equal, was not over-practised in speaking, but in the civil law was no lower than his master. As for Titus Aufidius, who lived to the utmost old age, he wished to be like these men, and was both a good man and blameless, but spoke too little; and his brother Marcus Vergilius did no more — the one who, as tribune of the people, summoned the imperator Lucius Sulla to trial. His colleague Publius Magius was, in speaking, somewhat more copious.

\ciceroSection{180} But of all the orators — or rather wranglers — who were plainly unlearned and unurbane, or even rustic, of those at least whom I have come to know, I judge the most fluent in speaking and the keenest, of our own order, to have been Quintus Sertorius, and of the equestrian, Gaius Gargonius. There was also one easy and ready at speaking, of much splendour of life and a quite acceptable talent, Titus Iunius the son of Lucius, of tribunician rank, on whose accusation Publius Sextius, praetor-designate, was condemned for bribery; he would have advanced further in office, had he not always been of weak and even sickly health.

\ciceroSection{181} And I understand perfectly well that I am dwelling on the recollection of men who were neither held to be orators nor were such, and that I am passing over some of the older men who are worthy of mention or of praise. But this is from ignorance; for what is there of an earlier age that can be written about men of whom no records speak, neither others' nor their own? But of those whom we ourselves have seen, I shall pass over almost no one of those we at some time saw speaking.

\ciceroSection{182} For I wish it to be known that, in so great and so old a commonwealth, with the greatest rewards held out for eloquence, all desired to speak, not very many dared to, and few were able. I shall, however, speak of each one in such a way that it can be understood whom I judge to have been a mere shouter, whom an orator. At about the same period, a little lower in age than Iulius, but very nearly his equals, were Gaius Cotta, Publius Sulpicius, Quintus Varius, Gnaeus Pomponius, Gaius Curio, Lucius Fufius, Marcus Drusus, Publius Antistius; nor in any age was there a richer crop of orators.

\ciceroSection{183} Of these, Cotta and Sulpicius easily carried off first place, both in my judgment and in everyone's. Here Atticus said: How is it that you say that — "both in my judgment and in everyone's"? Does the judgment of the crowd, in approving or disapproving an orator, always agree with the judgment of those who understand? Or are some approved by the multitude, others by those who understand? You ask rightly, Atticus, I said; but you will hear from me, perhaps, something that not everyone will approve.

\ciceroSection{184} Or do you worry about that, he said, if you are only going to win Brutus's approval here? Plainly, Atticus, I said, I would much rather that this discussion of approving or disapproving an orator pleased you and Brutus, but I would wish my own eloquence to be approved by the people. For necessarily the man who speaks so as to be approved by the multitude is the same man approved by the learned. As for what in speaking is right or faulty, I shall judge that — provided I am one who can or knows how to judge it; but what sort of orator a man is can be understood from what he accomplishes by his speaking.

\ciceroSection{185} For there are three things, at least as I feel it, which must be accomplished by speaking: that the one before whom one speaks be instructed, that he be delighted, that he be more vehemently moved. By what virtues of the orator each of these is accomplished, or by what faults the orator either fails to attain them or even, in attempting them, slips and falls — some craftsman will judge that. But whether or not it is accomplished by the orator that those who hear are affected as the orator wishes — that is wont to be judged by the assent of the crowd and the approval of the people. And so there has never been, over a good orator or a bad one, any disagreement between learned men and the people.

\ciceroSection{186} Or do you suppose that, while those flourished whom I named before, the ranking of orators was not the same in the judgment of the crowd as in that of the learned? If you had asked someone of the people thus — "Who in this state is most eloquent?" — about Antonius and Crassus he would either hesitate, or one would name this man, another that. Would no one set Philippus above these — an orator so sweet, so weighty, so witty, whom we ourselves, who wish to weigh these things by some art, have said came next to those men? Surely no one; for this very thing — to seem to the people the supreme orator — is the mark of the supreme orator.

\ciceroSection{187} Therefore, just as the flute-player Antigenidas said to a pupil who was, to be sure, going down coldly with the public, "Play for me and the Muses" — so I would say to Brutus here, when he speaks, as he does, before the multitude, "Play for me and the people, my Brutus," so that the hearers may perceive what is accomplished, while I perceive even why it is accomplished. The man who hears the orator believes what is said, thinks it true, assents, approves; the speech wins his trust: you, the craftsman, what more do you seek?

\ciceroSection{188} The listening multitude is delighted and led on by the speech and steeped, as it were, in a kind of pleasure: what have you to dispute? It rejoices, it grieves, it laughs, it weeps, it favours, it hates, it scorns, it envies, it is led to pity, to shame, to regret; it grows angry, it is softened, it hopes, it fears; and these things befall exactly as the minds of those present are handled by words, by thoughts, and by delivery; what is there to wait for from some learned man's verdict? For what the multitude approves, this same thing must be approved by the learned. In short, this is the test of the people's judgment, in which there has never been disagreement between the people and the learned and discerning.

\ciceroSection{189} When there were many orators of various kinds of speaking, who of them was ever judged by the crowd's judgment to excel, who was not approved by the learned as well? And when would there ever have been any doubt among our fathers, if anyone were given the choice of which advocate should be assigned to him, that he would choose either Antonius or Crassus? Many others were at hand; yet even granting that someone might have hesitated between these two, no one would have hesitated against either. And in our own youth, when there were Cotta and Hortensius, did anyone who had the power of choosing prefer anyone to these?

\ciceroSection{190} Then Brutus said: Why do you ask about others? In your own case, did we not see what litigants wished for, what Hortensius himself judged? — for when he divided cases with you (I was often present), he always left to you the place of the peroration, where speech has its greatest power. He did that, I said, and out of good will, I believe, gave me everything. But what the people's opinion of me may be, I do not know; of the rest I affirm this: that those held by the crowd's opinion to be most eloquent were the very same men most approved in the judgment of the discerning as well.

\ciceroSection{191} For Demosthenes could not have said what they report the famous poet Antimachus said. When he was reading to an assembled audience that great volume of his which you know, and all but Plato had walked out on him as he read, he said: ``I shall read all the same; for to me Plato alone is worth a hundred thousand.'' And rightly so; for a recondite poem owes its claim on the approval of the few, a popular speech on the assent of the crowd. But if Demosthenes had this same Plato as his single listener, and were deserted by all the rest, he could not get out a word.

\ciceroSection{192} What could you do, Brutus, if an assembly were to walk out on you, as one once did on Curio? ``For my part,'' he said, ``to confess it to you, even in those cases where our whole business is with the jurors and not with the people, still, if the ring of bystanders left me, I should not be able to speak.'' That is just how it stands, I said. As a flute-player, if his pipes do not give back the sound when he blows them, thinks he must throw them aside, so to the orator the people's ears are like the pipes: if they will not take the breath he blows into them — or if the listener altogether refuses, like a horse that will not go — there is nothing for it but to leave off driving. There is this difference, though: that the crowd sometimes approves an orator who ought not to be approved, but it approves without comparison; when it is pleased by a middling speaker, or even by a bad one, it is content with that; it does not perceive that there is anything better, and approves what there is, whatever it is. For even a middling orator holds the ears, provided there is something in him; and nothing has more force with the minds of men than the arrangement and adornment of a speech.

\ciceroSection{194} And so, who from among the people, listening to Q. Scaevola speaking for M. Coponius in that case I spoke of before, would either expect anything more polished or more elegant or simply better, or think it could be done?

\ciceroSection{195} When he wished to prove this point — that since M'. Curius had been instituted heir on the terms ``if the ward should die before he had come into his own guardianship,'' there could be no heir, the ward never having been born — what did he not say about the law of wills? about the old formulas? about how it ought to have been drawn up, if a son were to be instituted heir even in the event of his not being born?

\ciceroSection{196} How dangerous a thing it was for the people, that the written word should be set aside, and intentions sought by conjecture, and the writings of plain men perverted by the interpretation of the eloquent?

\ciceroSection{197} How much he said about the authority of his father, who had always maintained that this was the law? How much, in general, about preserving the civil law? And since he said all this with expert knowledge, and likewise with brevity, compression, and enough adornment and great elegance, who was there among the people who would either expect or think it possible that anything better could be done? But then, on the other side, Crassus began with the young dandy who, walking on the shore, had come upon a thole-pin and for that reason conceived a longing to build a ship — and went on that in the same way Scaevola, out of one thole-pin of a quibble, had built up a centumviral suit over an inheritance. With this opening, and with many sallies of the same kind following, he delighted them, and led the minds of all who were present from severity over into mirth — which is one of the three things I said an orator ought to accomplish. Then he argued that the man who had made the will had meant this, had felt this: that in whatever way there should turn out to be no son to come into his own guardianship — whether unborn or dead beforehand — Curius should be the heir; that most men drew up wills so, and that this held good and had always held good. Saying these things and many of the kind, he won belief — which is the second of the orator's three offices.

\ciceroSection{198} Then he defended, on grounds of equity and the good, the meaning and intentions of wills: how great a trap there is in words, in other matters and above all in wills, if intentions are disregarded; how great a power Scaevola was claiming for himself, if no one were to dare make a will thereafter except by his ruling. Setting all this out now with weight, now copiously from examples, now with variety, now even with wit and humour, he stirred such admiration and assent that no one seemed to have spoken against him — and this was, in the division, the orator's third office, in kind the greatest. Here that juror drawn from the people, who had admired each man separately, would, once he had heard the second, despise his own judgment of the first; whereas the discerning and learned hearer, listening to Scaevola, would feel that there is a certain richer and more adorned kind of speaking. But once the case had been pleaded out by both, if the question were asked which was the better orator, surely the wise man's verdict would never diverge from the verdict of the crowd.

\ciceroSection{199} In what, then, does the discerning man surpass the untrained? In a great thing and a hard one; if indeed it is a great thing to know by what means that — whatever it is — which ought to be achieved by speaking is achieved, or that which ought not to be lost is lost. The learned hearer surpasses the unlearned in this too: that often, when two orators or more are approved by the people's judgment, he understands which kind of speaking is best. For what is not approved by the people cannot be approved even by the discerning hearer. As from the sound of the strings on the lyre one can usually tell how skilfully they have been struck, so from the movement of minds it is discerned what the orator accomplishes in handling them.

\ciceroSection{200} And so the discerning judge of speaking, not sitting close and listening intently but at a single glance and in passing, often judges of an orator. He sees the juror yawning, talking with his neighbour, sometimes even gathering a little knot of company, sending to ask the hour, begging the presiding magistrate to dismiss them: he understands that there is no orator present in that case who could bring his speech to bear on the minds of the jurors as a hand is laid upon the strings. The same man, if in passing he has caught sight of the jurors upright and intent, so that they seem either to be learning about the matter and to be approving it with their very faces, or — as a bird is held by some song — to be held, as it were, suspended by the speech, or, what is most needful of all, to be stirred more violently by some emotion of mind, by pity, by hatred: if, in passing, as I said, he has caught sight of this, even if he has heard nothing, he will surely understand that an orator is at work in that trial and that the orator's work is being done, or is already done.

\ciceroSection{201} When I had set all this out, both gave their assent; and I, beginning again as if afresh: Since, then, I said, this whole discourse flowed from Cotta and Sulpicius — for I said that these above all were approved by the judgment of those men and of that age — I shall return to those two themselves; then I shall go through the rest in turn, as I have begun. Since, then, of good orators — for it is these we are looking for — there are two kinds, one of those who speak in a thin and compressed way, the other of those who speak loftily and amply; and although that is the better which is more splendid and more magnificent, still among good men everything that is of the highest is justly praised.

\ciceroSection{202} But the compressed orator must beware of poverty and meagreness, the ample one of an inflated and corrupt kind of speech. Cotta, then, found his matter shrewdly, and spoke purely and fluently; and as, in view of the weakness of his lungs, he had with great judgment given up all straining, so he fitted his kind of speaking to the frailty of his powers. There was nothing in his speech but what was sound, nothing but what was dry and healthy; and this above all, that though by force of speech he could scarcely bend the minds of the jurors, and did not speak at all in that manner, yet by handling them he still drove them to do, moved by him, the same thing they did when roused by Sulpicius.

\ciceroSection{203} For Sulpicius was the most grand and, so to speak, tragic orator of all I have ever heard. His voice was both powerful and sweet and brilliant; his gesture and the movement of his body so graceful that he seemed nonetheless trained for the Forum, not the stage; his speech rapid and flowing, yet neither redundant nor running over. He wished to imitate Crassus; Cotta preferred Antonius; but the one lacked the force of Antonius, the other the charm of Crassus.

\ciceroSection{204} A great art, indeed! said Brutus; if even from those men, supreme orators as they were, the two greatest things were each wanting to the one or the other. And in these orators this is to be noted: that men can be supreme who are unlike one another. For nothing was so unlike as Cotta to Sulpicius, and each far surpassed his own contemporaries. So it is the mark of the discerning teacher to see whither each man's nature carries him, and, using that as his guide, so to train him — as Isocrates is reported to have said that in the case of the keenest talent, Theopompus, and the gentlest, Ephorus, he applied the spur to the one, the rein to the other.

\ciceroSection{205} The speeches of Sulpicius that are in circulation are thought to have been written after his death by P. Cannutius, my contemporary, a man in my judgment the most eloquent outside our own order. Of Sulpicius himself there is no speech, and I often heard from him, when he said that he had neither been in the habit of writing nor was able to. As for Cotta's speech for himself under the Lex Varia, as it is entitled, that L. Aelius wrote at Cotta's request. He was altogether a distinguished man, a Roman knight among the most honourable, and likewise most deeply learned in both Greek and Latin letters, and well versed by his reading in our antiquity, both in its discoveries and in its deeds and in the writers of old. This learning our friend Varro — a man outstanding in talent and in every branch of knowledge — received from him and enlarged by his own efforts, and set forth in more and more illustrious writings.

\ciceroSection{206} But this same Aelius wished to be a Stoic, while as an orator he neither aspired to be one nor was one. He wrote speeches, all the same, for others to deliver: for Q. Metellus, son of †, for Q. Caepio, for Q. Pompeius Rufus — although Pompeius himself also wrote those he used in his own defence, but not without Aelius.

\ciceroSection{207} For I myself was even present at this writing, when as a young man I was with Aelius and used to listen to him most eagerly. As for Cotta, I am surprised that a supreme orator and a man not at all without taste should have wished those slight little speeches of Aelius to be thought his own. To these two of the same age no third was reckoned, but to me Pomponius was most pleasing — or rather, let me say, least displeasing. There was, on the whole, no place in the greatest cases for anyone besides those I have spoken of above, for this reason: that Antonius, who was most sought after, was easy in taking on cases; Crassus more fastidious, but he took them on nonetheless. He who had neither of these would take refuge generally with Philippus or with Caesar; †Cotta† and Sulpicius were in demand. So by these six advocates the famous cases were conducted; nor were as many trials held as in our own age, nor was there this that now happens, that single cases are defended by several advocates — than which nothing is more faulty.

\ciceroSection{208} We answer men we have not heard: in which, first, a thing has often been said one way and reported to us another; and besides, it is of great moment for me to see in person how my adversary asserts each point, and above all how each point is received. But nothing is more faulty than that, when the defence ought to be one body, the case should be born afresh, after it has been pleaded out by another.

\ciceroSection{209} For of all cases there is one natural beginning, one peroration; the remaining parts, like limbs, each set in its own place, keep each its force and its dignity. And since it is hard, in a long speech, not now and then to say something in such a way that one is inconsistent with oneself, how much harder to guard against saying anything that is inconsistent with the speech of the man who spoke before you? But because it is far greater labour to plead the whole case than a part, and because more gratitude is earned if at one time you speak for several, on that account we have gladly taken up this custom.

\ciceroSection{210} There were some, all the same, to whom Curio seemed the third of that age, because he used perhaps more splendid words and because he spoke Latin not at all badly — owing, I believe, to some household usage. For of letters he knew next to nothing; but it is of great moment whom each man hears daily at home, with whom he talks from boyhood, how his father, his tutors, and his mother too speak. We read the letters of Cornelia, the mother of the Gracchi: it is clear that the sons were brought up not so much in their mother's lap as in her speech. I have often heard the speech of Laelia, daughter of Gaius: so I saw her tinged with her father's elegance, and her two daughters, the Muciae, whose speech was known to me, and her granddaughters the Liciniae — both of whom I myself have heard speaking, and this one, the wife of Scipio, you too, Brutus, I believe, heard at some time. ``Indeed I did, and gladly,'' said Brutus; ``and the more gladly because she was the daughter of L. Crassus.''

\ciceroSection{212} What of that Crassus, I said, the son of this Licinia, who was adopted under the will of Crassus — what do you think of him? ``He is said to have been a man of the highest talent,'' he replied; ``and indeed this Scipio, my colleague, seems to me really to speak well, both in conversation and in pleading.'' You judge rightly, Brutus, I said. For that man's stock is engendered from the very root of wisdom itself. For we have already spoken of his two grandfathers, Scipio and Crassus, and of his three great-grandfathers — Q. Metellus, whose were those four sons; P. Scipio, who as a private man rescued the commonwealth into freedom out of the domination of Ti. Gracchus; Q. Scaevola the augur, who was reckoned most expert in the law and likewise most affable. And then of his two great-great-grandfathers, how illustrious is the name —

\ciceroSection{213} P. Scipio, who was twice consul, who was called Corculum, and the other the wisest of all men, C. Laelius. ``O noble stock!'' he said; ``and as on one tree many kinds are grafted, so into that house the wisdom of many is grafted and lit up!'' Similarly, then — to compare small things with great — I suspect that the house of Curio, although he was left a ward, was by his father's training accustomed to pure speech; and I judge this the more because, of those at least who were of any account, I have known no one so unlearned, so raw, in every kind of honourable art.

\ciceroSection{214} He had known no poet, had read no orator, had gathered no record of antiquity; he had learned neither public law nor private and civil law. And yet this was true even of others, and great orators at that, whom we have seen too little furnished with these arts — Sulpicius, for instance, and Antonius. But these, all the same, had that one work of speaking thoroughly worked out; and since it consists of five well-known parts, no one was wholly without anything in any one of them: for in whichever single part he were plainly to fall lame, he could not be an orator; but still, one excelled more in one part, another in another.

\ciceroSection{215} Antonius found out what needed to be said, and how it should be prepared, and in what place set, and grasped this in his memory; but he excelled in delivery. And he had some of these gifts on a par with Crassus, some even superior; yet Crassus's speech shone the more. Nor in truth can we say that of Sulpicius, or of Cotta, or of any good orator, any one thing out of those five parts was plainly and altogether wanting.

\ciceroSection{216} And so in Curio it can most truly be judged that by nothing single is an orator more commended than by the splendour and abundance of words. For he was both slow in invention and disordered in construction. The remaining two are delivery and memory: in both he stirred the guffaws of the mockers. His movement was the kind that C. Iulius branded for good, when, as Curio swayed his whole body from one side to the other, he asked who was speaking out of a skiff; and so did Cn. Sicinius, a foul man but exceedingly funny — and there was nothing else in him resembling an orator.

\ciceroSection{217} When, as tribune of the plebs, he had brought forward the consuls Curio and Octavius, and Curio had said much while his colleague Cn. Octavius sat by — Octavius being bound up in bandages and smeared with many salves for the pain in his joints — Sicinius said: ``You will never repay your colleague's kindness, Octavius: if he had not tossed himself about in his own fashion, today the flies would have eaten you up there.'' His memory, moreover, was so completely nil that several times, when he had set out three points, he would either add a fourth or go searching for the third; and in a private trial, even one of the greatest — when I had pleaded out the case for Titinia, the wife of Cotta, and he was speaking against me for Ser. Naevius — he suddenly forgot the whole case, and said it had been done by the poisons and incantations of Titinia. These are great signs of a forgetful mind; but nothing is more disgraceful than that even in his writings he would forget what he had set down a little before — as in that book where he represents himself coming out of the Senate and conversing both with our friend Pansa and with the younger Curio (after Caesar as consul had held a meeting of the Senate), and that whole conversation is drawn out of the son's inquiry into what had been done in the Senate. There, while Curio inveighed against Caesar at great length and there was a debate between them, as is the custom in dialogues — the conversation having been set after the dismissal of the Senate, the Senate that Caesar as consul had held — he criticizes those measures which that same Caesar carried out a year later, and thereafter in the remaining years, in Gaul.

\ciceroSection{219} Then Brutus, in astonishment: Was the forgetfulness so great, he said, especially in a written work, that not even in reading it did he ever perceive how great a scandal he had committed? And what, I said, Brutus, is more foolish than this: if he wanted to censure the things he did censure, not to set the conversation at the time when the occasions of those matters had not yet passed by? But he goes so wholly astray that in the same conversation he says he does not enter the Senate while Caesar is consul, and says it as he is coming out of the Senate while Caesar himself is consul. Now, when a man was so feeble in this part of the mind — the part that is the guardian of the other parts of the intellect — that not even in a written work did he remember what he had set down a little before, it is hardly any wonder that in speaking off the cuff his wits regularly slipped away from him.

\ciceroSection{220} And so, although he lacked nothing in diligence and burned with a passion for speaking, very few cases were brought to him. As an orator, though, among the living of his own generation he was reckoned next to the best, on account of the soundness of his words, as I said before, and a kind of ready, flowing speed. For that reason I think his speeches still worth a look. They are, to be sure, on the languid side, yet they can enlarge and as it were nourish that good quality which we grant was in him only to a middling degree — a quality of such force that, alone and without the others, it produced in Curio the look of a real orator. But let us return to our plan.

\ciceroSection{221} In the same reckoning, then, and of the same generation, was Gaius Carbo, the son of that most eloquent man. Not a sharp enough orator, but counted an orator all the same. There was weight in his words, he spoke easily, and his speech had a certain natural authority. Sharper at finding his material, and no less ready with his words, was Quintus Varius; a bold pleader, too, and forceful, neither poor nor base in language, and one you would plainly dare to call an orator was Gnaeus Pomponius, fighting from the lungs, driving men's spirits on, keen, biting, full of accusation.

\ciceroSection{222} Far below these stood Lucius Fufius, who nonetheless reaped the fruit of his diligence from the prosecution of Manius Aquilius. As for Marcus Drusus, your great-uncle, a weighty orator — but only when he spoke on public affairs — and Lucius Lucullus, sharp as well, and your own father, Brutus, thoroughly versed besides in both public and private law, and Marcus Lucullus, and Marcus Octavius son of Gnaeus, who had such weight in authority and in speaking that he got the Sempronian grain law repealed by the votes of a thronging people, and Gnaeus Octavius son of Marcus, and Marcus Cato the father, and Quintus Catulus the son too — let us draw these men out of the battle line, that is, from the courts, and station them in the garrisons of the commonwealth, which they could readily serve well enough.

\ciceroSection{223} In the same place I would set Quintus Caepio, had he not, too devoted to the equestrian order, broken with the Senate. Gnaeus Carbo, Marcus Marius, and several of the same stamp, by no means worthy of the ears of a refined gathering, I came to know as best fitted to turbulent public meetings. In which kind, to disturb here the order of the generations, Lucius Quinctius was found not long ago; and fitter still for the ears of the ignorant was Palicanus.

\ciceroSection{224} And since mention has been made of this kind, of all the rabble-rousers after the Gracchi Lucius Appuleius Saturninus seemed the most eloquent — though he caught men more by his look and his movement and his very dress than by any fullness of speech or even modest good sense. By far the most depraved man, however, since men were born, was Gaius Servilius Glaucia, but exceedingly sharp and cunning and among the foremost in raising a laugh. From the lowest dregs of both fortune and life he would have been made consul straight out of his praetorship, had it been ruled that his candidacy could legally be entertained. For he held the common people fast, and had bound the equestrian order to himself by the favour of a law. As praetor he was put to death by public act on the same day as Saturninus the tribune of the people, in the consulship of Marius and Flaccus — a man very like the Athenian Hyperbolus, whose baseness the old comedies of the Attic poets branded.

\ciceroSection{225} These men were followed by Sextus Titius, a man talkative enough and sharp enough, but so loose and soft in his gesturing that a kind of dance arose to which the name Titius was given. So one must take care to do nothing in pleading or speaking whose imitation might be laughed at. But we have been carried back to a slightly earlier generation; now let us return to the one of which we have already spoken a fair amount.

\ciceroSection{226} Joined, then, to the generation of Sulpicius was Publius Antistius, a quite passable hack of the courts, who, after holding his tongue for many years and being not only despised but even laughed at, first won approval in his tribunate, when he pleaded the true case against that extraordinary bid of Gaius Iulius for the consulship; and all the more because, when his colleague — Sulpicius himself, that very man — pleaded the same case, this one said more, and sharper. And so after his tribunate first many cases were brought to him, then all the greatest, whatever they were.

\ciceroSection{227} He saw his matter keenly, composed it carefully, and had a strong memory. He used words that were not, to be sure, ornate, but neither were they base; his speech ran ready and easy, and there was in him a certain manner, not without urbanity. His delivery limped a little, partly from a flaw of voice and partly from clumsiness. He flourished in those days when, between the departure and the return of Lucius Sulla, the commonwealth was without law and without any dignity at all; and he was the more esteemed because there was a kind of solitude of orators in the Forum. Sulpicius had fallen, Cotta and Curio were away, and of the remaining advocates of that age not one was living except Carbo and Pomponius, both of whom he easily surpassed.

\ciceroSection{228} Of the next generation the nearest was Lucius Sisenna, a learned man devoted to the finest studies, speaking good Latin, well acquainted with public affairs, not without wit, but neither a man of much labour nor sufficiently practised in cases; and, set between the two generations of Hortensius and Sulpicius, he could neither come up to the elder nor help yielding to the younger. His whole capacity can be discerned from his history, which, though it easily surpasses all that came before, nonetheless shows how far it falls short of the highest, and how this kind of writing is not yet sufficiently illumined in Latin letters. For the genius of Quintus Hortensius, while still very much a youth, was at once beheld and approved, as a statue of Phidias is. In the consulship of Lucius Crassus and Quintus Scaevola he first spoke in the Forum, and before these very consuls indeed; and he came off approved by the judgment both of those who were present and of the consuls themselves, who outstripped everyone in discernment. He was nineteen years old at that time, and he died in the consulship of Lucius Paullus and Gaius Marcellus: from which we see that he was in the number of the advocates for forty-four years. About this orator we shall say more a little later; here we wished only to fit his lifetime into the differing lifetimes of the orators. Though indeed it must of necessity have befallen all whose lives ran somewhat longer, that they were compared both with men much older than themselves and with men a good deal younger. As Accius says that under the same aediles he and Pacuvius produced a play, when the latter was eighty and he himself thirty years old: so

\ciceroSection{230} Hortensius is joined not only with his own contemporaries but also with my generation, and with yours, Brutus, and with one a good deal earlier — seeing that he used to plead while Crassus was still alive, and that he was already flourishing all the more vigorously when, with Antonius and Philippus now an old man speaking on behalf of the property of Gnaeus Pompeius in that case, he, though still a young man, was foremost; and he had readily come into the number of those I placed in the generation of Sulpicius, and among his own contemporaries — Marcus Piso, Marcus Crassus, Gnaeus Lentulus, Publius Lentulus Sura — he far excelled; and having got hold of me as a young man, eight years younger than he was himself, he kept me at work for many years in the pursuit of the same glory; and together with you, just as I spoke for many, so he spoke for Appius Claudius, a little before his death.

\ciceroSection{231} You see, then, how we have come round to you as an orator, Brutus, with so many orators set between the beginning of my speaking and of yours; and of these, since in this conversation of ours I have resolved to name none who were living — lest you should too inquisitively draw out of me what I judged of each — I shall name those who are now dead. Then Brutus: That is not, he said, the reason you allege, why you are unwilling to say anything of those who live. What, then, said I, is it? I think you fear, he said, that through us this talk of yours may leak out, and that those you have passed over may be angry with you. What? said I; will you not be able to keep silent? We, for our part, he said, most easily; but still I think you would rather keep silent yourself than put our discretion to the test.

\ciceroSection{232} Then I: I shall tell you the truth, Brutus, said I. I did not suppose I would come, in this conversation, all the way down to the present generation; but the sequence of the generations has so drawn my discourse on that I have now reached even the younger men. Bring them in, then, he said, any you think fit; then let us return to you and to Hortensius. No, rather, said I, to Hortensius; others will speak of me, if any wish to. By no means, he said. For although you have easily held me by all your talk, still it seems to me too long, because I am in haste to hear about you — and not so much about your virtues in speaking, which are most fully known both to all and certainly to me, as because I am eager to learn your stages and as it were the steps of your progress in speaking.

\ciceroSection{233} You shall be indulged, said I, since you wish me to be the proclaimer not of my talent but of my labour. But I shall bring in others, as you please, and begin with Marcus Crassus, who was a contemporary of Hortensius. He, then, moderately equipped with learning, and even more cramped by nature, was for several years among the leading advocates by means of toil and industry, and because he brought to the winning of cases both care and personal influence. In his oratory the Latin was sound, the words not base, the matter carefully composed; yet there was no flower, no light at all; a great straining of spirit, a small straining of voice; nearly everything delivered alike and in one manner. For his contemporary and enemy Gaius Fimbria could not show himself off for very long; a man who, saying everything at the top of his voice with a certain headlong rush of words — sound words enough — nevertheless raved so that you might wonder the people were attending to other things, that there should be a madman among eloquent men. But Gnaeus Lentulus created by his delivery a far greater opinion of his speaking than the capacity in him warranted; for though he was neither very sharp — although from his face and his look he seemed so — nor abounding in words, even if in that very point he deceived: yet by his pauses, his exclamations, his sweet and ringing voice, †with a kind of wonder† he kept up his game, and grew warm in pleading, so that what was lacking went unmissed. Thus, as Curio held an orator's place by some fullness of words and no other good thing, so Lentulus hid the mediocrity of his other virtues of speaking by his delivery, in which he was excellent. And not much otherwise did Publius Lentulus, whose slowness both of conceiving and of speaking was screened by the dignity of his appearance, the movement of his body full of both art and grace, and the sweetness and the volume of his voice. So in this man there was nothing but the delivery; the rest was even slighter than in the one before.

\ciceroSection{236} Marcus Piso had whatever he had from training, and was the most learned of all who came before in Greek doctrines. He had from nature a certain kind of acuteness, which he had even polished by art, an acuteness that was deft and skilful in faulting words, but often peevish, sometimes flat, now and then witty too. He did not endure the labour of the Forum — as it were the race-course — for very long, both because he was weak in body and because he would not put up with the follies and stupidities of men, which the rest of us must swallow, but spat them out too irritably, whether out of moroseness, as was thought, or out of a free-born and independent disdain. After he had flourished well enough as a young man, he began afterwards to be reckoned less. Then from the trial of the Vestal Virgins he won great praise, and from that time, recalled as it were into the race, he held his place as long as he could bear the labour; afterwards, as much as he withdrew from the work, so much he lost of his glory.

\ciceroSection{237} Publius Murena, of middling talent, but with a great enthusiasm for ancient times, was a man both fond of letters and not unskilled in them, of much industry and great labour. Gaius Censorinus, learned enough in Greek letters, set out his point readily, was not an ungraceful pleader, but he was inert and no friend to the Forum. Lucius Turius, with small talent but much labour, spoke often in whatever way he could; and so he fell short of the consulship by a few centuries.

\ciceroSection{238} Gaius Macer always lacked authority, but he was about the most diligent of advocates. If his life, his character, his very look, had not overturned every recommendation his talent gave, he would have had a greater name among the advocates. His speech was not abundant, yet not poor; not very polished, not downright rough; his voice, gesture, and whole delivery without charm; but in finding and composing his matter there was a marvellous precision, such that I have not easily known a more careful or a greater man in any one — though you would call it craft of the old hand sooner than oratory. He, though approved even in public cases, nonetheless held a more distinguished place in private ones.

\ciceroSection{239} Next Gaius Piso, a measured orator and full of conversational ease, by no means slow in invention, but who nonetheless seemed, by his look and his pretence, much sharper even than he was. For his contemporary Manius Glabrio, well trained by the diligence of his grandfather Scaevola, had been slowed by his own sluggish and careless nature. Lucius Torquatus too, elegant in speaking, very judicious in his estimates, throughout his whole manner thoroughly urbane. My own contemporary Gnaeus Pompeius, a man born for all the highest things, would have had a greater glory in speaking, had not the craving for a greater glory dragged him off to the praises of war. He was full enough in his oratory, he saw his matter prudently; his delivery, indeed, had both a great splendour in the voice and the highest dignity in the movement.

\ciceroSection{240} Likewise my contemporary Decimus Silanus, your stepfather, had not much application, it is true, but enough of acuteness and of style. Quintus Pompeius son of Aulus, who was called Bithynicus, perhaps two years older than I, was a man of the greatest passion for learning and of much doctrine, of incredible labour and industry — which I am in a position to know, for he was bound to me, and to Marcus Piso, by friendship and also by studies and exercises. His delivery did not commend his oratory sufficiently; for in the latter there was fullness enough, but in the former too little charm.

\ciceroSection{241} His contemporary was Publius Autronius, with a voice exceedingly sharp and loud, but commendable for no other thing; and Lucius Octavius of Reate, who, when he was already pleading many cases, died a young man — though he came to his speaking more boldly than well prepared; and Gaius Staienus, who had adopted himself and made an Aelius out of a Staienus, with a kind of feverish, insolent, and frenzied manner of speaking; and because this was pleasing to many and won approval, he would have climbed to high office, had he not, caught in manifest crime, paid the penalty to the laws and to a court.

\ciceroSection{242} At the same time there were the brothers Gaius and Lucius Caepasius, who by much effort, men unknown and sprung from nowhere, were quickly made quaestors, with a certain provincial and unpolished manner of speaking. Let us add here too, lest we seem to have passed over any voice, Gaius Cosconius Calidianus, who with no acuteness nonetheless served the people that fullness of words he had, if any, with much running about and great shouting. The same thing was done by Quintus Arrius, who was, as it were, Marcus Crassus's second. He ought to be an example to all of how much, in this city, it avails to obey the convenience of many and to be at the service of many men, whether for their honour or in their danger.

\ciceroSection{243} For by these means a man born in the lowest station had attained office, money, and influence, and had even come into some number among the advocates — without learning, without talent. But just as untrained boxers, even if, eager for the Olympic games, they can bear the fists and the blows, still often cannot bear the sun, so that man, when he had now run through every stretch of fortune prosperously and taken up even great labours, did not bear the severity of that judicial year — the sun, as it were.

\ciceroSection{244} Then Atticus: You are drawing now from the dregs, he said, and have been this good while, but I held my tongue; this, though, I did not expect — that you would come all the way down to the Staieni and the Autronii. I do not suppose, said I, that you think I am slipping out of partiality — these men being dead, after all; but following the sequence I run necessarily into a memory familiar and of my own time. And I want this to be seen: that of all who have ever been searched out who dared to speak before a crowd, very few indeed are worthy of memory, and even of those who had any name at all, there were not so very many. But let us return to the conversation we set out upon.

\ciceroSection{245} Titus Torquatus son of Titus, a man both learned in the Rhodian discipline of Molon and by nature fluent and ready enough for speaking — who, had life lasted him, would, with bribery abolished, have been made consul — had more capacity for speaking than inclination. And so he did not satisfy this pursuit; but in duty he was wanting neither in the cases of his own near connections nor in his senatorial vote.

\ciceroSection{246} Marcus Pontidius too, our fellow townsman, pleaded many private cases, rolling out his words quite rapidly and not dull in his cases — or rather I should say more than not dull — but boiling over in his speaking, often with too much heat of bile and anger; so that he would frequently wrangle not only with his adversary but even, what was remarkable, with the judge himself, whose coaxer the orator ought to be. Marcus Messalla, younger than I, by no means poor, but not too ornate in his kind of language; prudent, sharp, by no means an incautious advocate, careful in learning and composing his cases, a man of great labour, of much effort, and of many cases.

\ciceroSection{247} The two Metelli also, Celer and Nepos, somewhat practised in cases, neither without talent nor unlearned, had attained this popular manner of speaking. Gnaeus Lentulus Marcellinus, on the other hand, was never lacking in eloquence, and in his consulship was thought thoroughly eloquent — not slow in his opinions, not poor in words, with a ringing voice, witty enough. Gaius Memmius son of Lucius, accomplished in letters, but Greek ones, quite disdainful of the Latin, a clever orator and sweet in his words, but shrinking from the labour not only of speaking but even of thinking, withdrew as much from his capacity as he subtracted from his industry.

\ciceroSection{248} At this point Brutus: How I wish, he said, you were willing to speak of those orators too who exist today; and, if of others less, yet of two at least whom I know you are accustomed to praise, Caesar and Marcellus, I would hear no less gladly than I have heard about those who are gone. Why, pray? said I; or do you wait to hear what I judge of those men, who are as well known to you as to me? To me, by Hercules, he said, Marcellus is well enough known,

\ciceroSection{249} "But Caesar too little," I said. "For I heard the one often; the other, by the time I could already form some judgment, was away." "What, then, do you judge of the man you have often heard?" I said. "What should I think," he said, "but that you are going to have one like yourself?" "Indeed," I said, "if that is so, I should wish him to please you as much as possible." "And so it is," he said, "and he pleases me intensely; nor without reason. For he both learned, and, setting all other studies aside, pursued this one thing, and trained himself most keenly by daily practice."

\ciceroSection{250} "And so he uses choice words and frequent maxims, and by the splendour of his voice and the dignity of his bearing what he says is made striking and brilliant; and everything is so abundantly at his command that I think no excellence of the orator is wanting in him. And he most deserves praise for this: that at this very time, so far as it is permitted amid this common and, as it were, fated misfortune of ours, he consoles himself both with the consciousness of an excellent mind and with the practice and renewal of his learning. For I saw the man lately at Mytilene — and, as I said, I saw plainly a man. And so, while I had seen him before like you in speaking, now indeed, equipped by Cratippus, that most learned man and, as I understood, your closest friend, with the whole resource of his art, I saw him far more like you."

\ciceroSection{251} Here I said: "Although I gladly hear the praises of a most excellent man and our closest friend, still I run up against the memory of our common miseries — in seeking to forget which I have drawn this very conversation out at greater length. But about Caesar I am eager to hear what, after all, Atticus judges." And he said: "Splendidly do you keep to your principle, in being unwilling to say anything yourself about those now living; and by Hercules, if you were to proceed as you proceeded with those who are now dead — passing over no one — you would run up against many an Autronius and Staienus. So whether you wished to escape that crowd, or feared that someone might complain of being passed over or not praised enough, you could still have spoken about Caesar — especially since both your judgment of his talent is most well known, and his of yours is not obscure. But for all that, Brutus," said Atticus, "as to Caesar, I myself judge thus, and from this most keen appraiser of the kind I hear it most often: that he speaks Latin more elegantly than almost all orators; and this not by household habit alone, as we were lately hearing of the families of the Laelii and the Mucii, but — though I believe that too played its part — still, that this praise of speaking well might be made perfect, he attained it by much literature, and that of a recondite and choice sort, and by the highest zeal and diligence. He even, amid the greatest occupations, wrote to you yourself" — he said, looking toward me — "most carefully on the principle of speaking Latin, and in the first book declared that the choice of words is the origin of eloquence, and granted to this friend of ours, my dear Brutus" — to him who preferred that I, rather than he, should speak about Caesar — "a singular praise; for he wrote in these words, having addressed him by name: 'And if, in order to be able to set forth their thoughts splendidly, some men have laboured with study and practice — and of that resource we must consider you well-nigh the foremost and the inventor, and one who has deserved well of the name and dignity of the Roman people — is the man who has come to know this easy and everyday manner of speech to be reckoned for that reason behind?'"

\ciceroSection{254} Then Brutus said: "Affectionately, by Hercules, and magnificently I think you have been praised — you whom he called not only the foremost and the inventor of resource, which was great praise, but also one who has deserved well of the name and dignity of the Roman people. For in the one thing wherein we were being outdone by a conquered Greece, that has either been wrested from them or at least shared by us with them. And this glory, this testimony of Caesar," he said, "I rank ahead not of your own thanksgiving, indeed, but of the triumphs of many men." "And rightly so, Brutus," I said, "provided this be a testimony of Caesar's judgment, not of his goodwill. For surely whoever he is — if indeed there is anyone — who has not merely illumined but actually given birth to eloquence in this city, has brought more dignity to this people than those who stormed the strongholds of the Ligurians; out of which storming there are, as you know, many triumphs."

\ciceroSection{256} "And if we are willing to hear the truth — setting aside those godlike counsels, by which the wisdom of commanders has often established the safety of the state, whether in war or at home — a great orator far excels paltry commanders. 'But a commander does more good.' Who denies it? Yet for all that — I am not afraid you will shout me down; and here is the place where you may say freely what you feel — I should rather have for myself one speech of L. Crassus on behalf of M'. Curius than two triumphs over strongholds. 'But it mattered more to the commonwealth that a stronghold of the Ligurians be taken than that the case of M'. Curius

\ciceroSection{257} be well defended.' I believe it; but to the Athenians too it mattered more to have weatherproof roofs on their dwellings than that loveliest statue of Minerva in ivory; and yet I should rather be Phidias than even the best carpenter. So it is not how much each man profits us, but how much each man is worth, that must be weighed — especially since few can paint or sculpt to excellence, while labourers and porters can never be lacking.

\ciceroSection{258} "But go on, Pomponius, about Caesar, and render what remains." "You see, indeed," said he, "the ground and, as it were, the foundation of the orator: faultless and Latin speech, in which the praise that has so far rested with men was not a matter of theory or of science, but of good habit, so to speak. I leave aside C. Laelius and P. Scipio: that praise of speaking Latin, like that of innocence, belonged to their age — yet not to all, for we see their contemporaries Caecilius and Pacuvius spoke badly; but nearly all in those days who had neither lived outside this city nor been darkened by some barbarism at home spoke correctly. But this thing antiquity has certainly made worse, both at Rome and in Greece. For many who speak impurely have streamed in, from various places, both into Athens and into this city. The more, therefore, must our speech be purified, and a kind of touchstone of method applied, which cannot be altered, nor must the most depraved rule of habit be used.

\ciceroSection{259} "T. Flamininus, who was consul with Q. Metellus, we saw as boys: he was reckoned to speak good Latin, but he knew nothing of letters. Catulus, indeed, was by no means unlearned, as you said a little while ago; yet still it was the sweetness of his voice and the smooth articulation of his letters that had made his reputation for speaking well. Cotta, who by broadening his vowels strongly had drawn himself away from any resemblance to Greek pronunciation, and sounded the opposite of Catulus — something rather countrified and plainly rustic — had reached the same praise by another route, an uncultivated and woodland one, so to speak. But Sisenna, though he wished to be a kind of reformer of ordinary speech, could not be deterred even by his accuser C. Rusius from using unheard-of words." "What is that, now?" said Brutus; "or who is that

\ciceroSection{260} C. Rusius?" And he replied: "He was an accuser of the old school. When he was prosecuting C. Hirtilius, Sisenna in his defence said that some of the charges were 'spittable.' Then C. Rusius said: 'I am being trapped, jurors, unless you come to my aid. What Sisenna means I do not know; I fear an ambush. Spittable — what is this? What spit is, I know; -table, I do not.' Great laughter; but still, that friend of mine thought that to speak correctly was to speak unusually.

\ciceroSection{261} "Caesar, however, applying method, corrects faulty and corrupt habit by pure and uncorrupted habit. And so, when to this elegance of Latin words — which, even if you are not an orator but are a free-born Roman citizen, is nevertheless necessary — he joins those oratorical ornaments of speaking, then he seems, as it were, to set well-painted panels in a good light. Since he has this praise as his special possession, in the qualities common to all I do not see to whom he ought to yield. He commands a method of speaking that is splendid and not at all that of a hack, and is even, in voice, movement, and presence, magnificent and in a certain way noble."

\ciceroSection{262} Then Brutus said: "His speeches, certainly, win my strong approval. I have read a good many; and also the commentaries which the same man wrote of his own deeds." "Strongly to be approved, indeed," I said; "for they are bare, straight, and graceful, with all the ornament of oratory stripped off like a garment. But while he wished others to have something ready to hand, from which those who wished to write history might draw, he has perhaps done a service to fools who will want to crimp those facts with curling-irons; but sane men he has frightened off from writing — for nothing in history is sweeter than pure and brilliant brevity. But let us return, if you please, to those who have departed from life.

\ciceroSection{263} "C. Sicinius, then — grandson, through a daughter, of that Q. Pompeius who was censor — died of quaestorian rank; a creditable orator, and by now even an approved one, formed by the school of Hermagoras, which is poor in resources for adornment but ready for invention. That school gives sure methods and precepts of speaking, which, if they have less apparatus — for they are meagre — still have order and certain paths that do not allow a man to wander in his speaking. Holding fast to these, and coming prepared to his cases, not at a loss for words, by that very arrangement and discipline of speaking he had already made his way into the ranks of the advocates.

\ciceroSection{264} "There was besides a man learned among the first, C. Visellius Varro, my cousin, who was joined with Sicinius in age. He died while serving, after his curule aedileship, as president of a court; and in his case I confess that the judgment of the crowd differed from my own. For he was not saleable enough to the people: his oratory was headlong, and obscure for the very reason that it was over-sharp, and rapid, and blinded by its own speed; yet I should not readily call another man more apt in words or more crowded with thoughts. Besides, perfect in letters and in the civil law, he held the discipline 265 already handed down by his father Aculeo. The rest who are dead are: L. Torquatus, whom you would not so quickly have called an orator — though oratory was not wanting to him — as, in the Greeks' phrase, a man of public affairs \textit{[Greek: politikon]}. In him there were very many letters, and not common ones, but a kind of deeper and recondite learning, a divine memory, the highest weight and elegance of words; and all this the gravity and integrity of his life adorned. I, for one, took great delight too in the oratory of Triarius, in that age of his so full of lettered old age. What severity in his face! what weight in his words! how nothing came forth from his mouth that was not considered!

\ciceroSection{266} Then Brutus, moved by the mention of Torquatus and Triarius — for he had loved them both dearly — said: "Indeed, to pass over all the rest, which are beyond counting, when I think of those two men I grieve that your unbroken authority on behalf of peace counted for nothing! For the commonwealth would have lost neither those excellent men nor many other outstanding citizens." "Let us be silent, Brutus," I said, "about them, lest we increase our grief. For the recollection of things past is bitter, and the expectation of what is to come is more bitter still. So let us cease to mourn, and only proclaim what each man could do in speaking, since that is what we are inquiring into.

\ciceroSection{267} "There are also, among those who fell in that same war, M. Bibulus, who both wrote carefully — especially since he was no orator — and conducted many affairs with steadiness; Appius Claudius, your father-in-law, my colleague and friend: he was by now both studious enough and very much an orator, both learned and also practised, and well versed both in augural law and in all our public law and antiquity. L. Domitius spoke with no art at all, indeed, but Latin nonetheless, and with much freedom.

\ciceroSection{268} "Two Lentuli besides, of consular rank: of whom that Publius — avenger of my wrongs, author of my restoration — whatever he had, however much it was, all of it he had from training; the equipment of nature was wanting; but so great was the splendour of his spirit and so great his magnanimity that he did not hesitate to claim for himself all that belonged to illustrious men, and held it with full dignity. L. Lentulus, on the other hand, was a stout enough orator, if he was an orator at all, but could not bear the labour of thinking; a tuneful voice, words not at all uncouth, so that his speech was full of spirit and of terror; in the courts you might perhaps look for something better, but in public affairs you would judge what there was to be enough.

\ciceroSection{269} "Nor is T. Postumius to be despised in speaking; and in matters of the commonwealth he was an orator no less vehement than he was a warrior — unbridled and over-fierce, but he had learned the laws and institutions of public law well." At this point Atticus said: "I should think you were courting favour, if, as you said, those whom you have now been gathering for so long were still alive. For you mention everyone who has ever dared to stand up and speak, so that you seem to me, without thinking, to have passed over M. Servilius."

\ciceroSection{270} "I am not unaware of that, Pomponius," I said: "that there were many who never uttered a word in public, though they could speak considerably better than these orators I am gathering; but by recounting these men I attain this further end — that you may understand, first, out of the whole number how few have dared to speak, and then, out of those very men, how few have been worthy of praise.

\ciceroSection{271} "And so I will not pass over even these Roman knights, our friends, who lately died: P. Cominius of Spoletium, on whose prosecution I defended C. Cornelius, in whom there was a style of speaking both composed and keen and ready; T. Accius of Pisaurum, to whose prosecution I replied on behalf of A. Cluentius, who both spoke carefully and copiously enough, and was besides schooled in the precepts of Hermagoras, by which — although ornaments of speaking not rich enough are handed down — still, like javelins fitted with throwing-straps for the light troops, certain arguments apt and ready are supplied for each kind of case.

\ciceroSection{272} "In zeal, moreover, and in industry I have known no one greater — though indeed I should not readily say who surpassed him even in talent — than my son-in-law C. Piso. No time was ever free for him, whether from forensic speaking, or from study at home, or from writing, or from thinking. And so he made such advances that he seemed to fly, not to run; and there was an elegant choice of words and an apt and, as it were, rounded construction; and while many arguments, firm for proving, were thought out by him, there were also neat and acute thoughts; and his gesture was by nature so graceful that even an art, which was not there, and a certain trained movement seemed to be added. I fear I may seem, out of love, to say more than was in him; which is not so — for other and greater things can be said of him. For neither in self-control nor in piety nor in any kind of virtue do I think anyone of the same age fit to be compared with him.

\ciceroSection{273} "Nor, indeed, do I think M. Caelius is to be passed over — whatever, at his end, were his fortune or his mind; who, so long as he obeyed my authority, was such a tribune of the plebs that no one stood more steadfastly against the popular and turbulent madness of ruined citizens, on the side of the Senate and of the good men. His action was old-fashioned; yet his oratory, which was at once brilliant and grand, and likewise witty in the first rank and most urbane, did much to commend it. He delivered several weighty public addresses, and three keen prosecutions, all of them undertaken out of political contention; his defences — though those qualities I have mentioned were better in him — were nonetheless not to be despised, and quite tolerable. When, with the fullest goodwill of the good men, he had been made curule aedile, somehow, at my departure, he departed from himself and fell, after he began to imitate those very men whom he himself had brought down. But let us say something about

\ciceroSection{274} M. Calidius, who was not one orator among many, but rather, among many, well-nigh unique: with such a soft and translucent speech did he clothe recondite and choice thoughts. Nothing so delicate as his joining of words, nothing so supple, nothing more shaped at his own will, so that no orator's words were equally in his power: which, first, were so pure that nothing could be clearer, and flowed so freely that they nowhere stuck fast; you would see no word but one set in its place and, as it were, fitted into mosaic-work, as Lucilius puts it; nor any word that was either harsh, or strange, or low, or drawn out too far; and not the words proper to things, but for the most part transferred ones — yet so transferred that you would say they had not burst into another's place but immigrated into their own; nor was this speech loose or flowing apart, but bound by rhythm, brought to its close not openly nor always in the same way, but variously and with concealment.

\ciceroSection{275} "There were, too, those lights of words and of thoughts which the Greeks call figures \textit{[Greek: schemata]}, by which, as by insignia, his whole oratory was set off in its adornment. And as for that point of 'what the matter at issue is' which is enclosed in so many places in the formulae of the jurisconsults, he saw where it lay.

\ciceroSection{276} "There was added an order of matter full of art, a delivery worthy of a free man, and a wholly calm and sound kind of speaking. And if it is best to speak agreeably, there is nothing you should think it better to seek than this. But since it was said by us a little while ago that there seem to be three things the orator ought to accomplish — to instruct, to delight, to move — two he held in the highest degree: that he both made his matter clear by exposition and bound fast the minds of his hearers with pleasure; but that third praise was wanting to him, by which he should thoroughly stir and rouse their minds — the praise which we have said has the greatest power; nor was there in him any force and intensity: whether by design, because he thought that those whose oratory was loftier and whose action was more ardent were raving and reeling in frenzy; or because by nature he was not so made; or because he was not accustomed to it; or because he could not. This one thing was absent from him, if it had no usefulness; if it was needed, it was lacking.

\ciceroSection{277} "Indeed, I even remember that, when in his prosecution he had charged Q. Gallius with having prepared poison for him, and that this had been detected by him, and said that he was bringing in handwriting, depositions, informations, examinations, the thing manifest — and had argued about that charge carefully and elaborately — I, in replying, after I had argued the case so far as the matter allowed, also put forward this very point as an argument: that he, while saying he had discovered the destruction of his own life, the proofs of his death, manifestly and was holding them in his hand, had pleaded so loosely, so gently, so drowsily. 'Would you, M. Calidius,

\ciceroSection{278} unless you were inventing this, plead it so? — especially since with that eloquence of yours you are wont to defend the perils of other men most keenly, would you have neglected your own? Where is the grief? where the fire of spirit, which is wont to draw out cries and complaints even from the natures of speechless men? No agitation of mind, none of body; the brow not struck, the thigh not smitten; not so much as a stamp of the foot, which is the least of things. And so far was it from your inflaming our minds, that at that point we could scarcely keep ourselves from sleep.' Thus we used the sound health — or the fault — of a supreme orator as an argument toward dissolving the charge."

\ciceroSection{279} Then Brutus said: And do we really doubt whether that was soundness or a fault? For who would not admit that, since of all the orator's merits this by far is the greatest — to set the minds of his hearers ablaze and to bend them whichever way the case demands — a man who has lacked this power has lacked the very thing that was greatest? Let it be so, by all means, I said; but let us return to the one who alone now remains, Hortensius, and then I shall say a few words about myself, since that too, Brutus, is what you ask. And yet I must make mention, as it seems to me at least, of two young men who, had they lived longer, would have won great praise for their eloquence.

\ciceroSection{280} You mean, said Brutus, C. Curio and C. Licinius Calvus, I think. You think rightly, I said. Of these the one rolled out his words so easily and fluently — words at times sharp enough, and at any rate thick-coming in their thoughts — that nothing could be more ornate, nothing more ready to hand. And this man, though too little trained by teachers, had a natural gift for speaking that was admirable. His industry I never tested; his zeal, certainly, was there. Had he been willing to listen to me, as he had begun to, he would have preferred to pursue honours rather than power. What do you mean by that? he said; and how do you draw the distinction? In this way, I said.

\ciceroSection{281} Since an honour is the reward of merit conferred upon someone by the judgment and zeal of his fellow citizens, the man who has gained it by their votes and their ballots seems to me both honourable and honoured. But the man who, by some opportunity and even against the will of his fellow citizens, has got hold of command — as that man desired to do — has, I think, obtained the name of honour, not honour itself. Had he been willing to listen to this, he would have arrived, with the greatest favour and glory, at the highest distinction, climbing by the steps of the magistracies, as his father had done, as the rest of the more illustrious men have done. And I think I often urged this same thing upon P. Crassus, son of Marcus, when in his early years he had attached himself to my friendship — when I exhorted him strongly to count that path to praise the straightest which his ancestors had left worn smooth for him. For he was both excellently trained and also perfectly and plainly learned, and there was in him a wit keen enough and an abundance of language not without elegance; besides this, he seemed grave without arrogance, and modest without sluggishness. But him too a certain tide of glory, uncommon in young men, swept away; because he had served his commander as a soldier with diligence, he wanted at once to be a commander himself — a charge to which the custom of our ancestors has left a fixed age, but an uncertain lot. And so, by his most grievous fall — while he wished to be like Cyrus and Alexander, who had run their course to the end — he was found utterly unlike L. Crassus and the many Crassi.

\ciceroSection{283} But let us return to Calvus — for he was the man we set before ourselves. As an orator he was both more learned in letters than Curio, and he also brought to his speaking a kind that was more careful and more refined; and although he handled it knowingly and with elegance, yet by probing himself too much and watching his own self and fearing to gather in a fault, he kept losing even his true blood. And so his speech, thinned out by an excess of scruple, was brilliant to the learned and the attentive listener, but by the crowd, and by the Forum — for which eloquence was born — it was swallowed up and lost.

\ciceroSection{284} Then Brutus said: Our friend Calvus wished himself to be called an Attic orator: that was the source of that leanness which he pursued of set purpose. He said so, I replied; but he was both mistaken himself and forced others into the same mistake. For if a man thinks that those who speak neither ineptly nor offensively nor with affected niceness speak in the Attic manner, he is right to approve no one who is not an Attic. For he hates the tastelessness and the unrestraint that are a kind of madness in speech, while he approves the soundness and the wholeness that are, as it were, the scruple and modesty of the orator. This ought to be the one judgment of all orators.

\ciceroSection{285} But if a man sets meagreness and dryness and poverty — provided only it be polished, provided it be urbane, provided it be elegant — within the Attic kind, he is right only so far; but because among the Attics some things are better than others, let him take care not to be ignorant of the gradations and the differences and the force and the variety of the Attics. "I wish to imitate the Attics," he says. Which of them? For there is no single kind. What is so unlike as Demosthenes and Lysias? as the same Demosthenes and Hyperides? as Aeschines is unlike all of these? Whom, then, do you imitate? If some one of them, were the rest, then, not speaking in the Attic manner? If all of them, how can you, when they themselves are so utterly unlike one another? On this point I ask this too: whether that famous Demetrius of Phalerum spoke in the Attic manner. To me, at least, his very speeches seem to give off the fragrance of Athens. But he is more florid, so to speak, than Hyperides, than Lysias: there was a certain nature, or a certain inclination, to speak so.

\ciceroSection{286} And indeed there were two men of the same period unlike each other, yet Attics for all that: of whom Charisius, the author of many speeches that he wrote for others, seemed to wish to imitate Lysias; while Demochares, who was the son of Demosthenes' sister, both wrote a number of speeches and set down the history of the things done at Athens in his own lifetime in a manner not so much historical as oratorical. And Hegesias wishes to be like Charisius, and so thinks himself an Attic that, beside himself, he counts those genuine Attics little better than rustics.

\ciceroSection{287} But what is so broken, so chopped fine, so childish in the very neatness which it nonetheless achieves? "We wish to be like the Attics." Excellent; are these men, then, Attic orators? "Who can deny it? These are the men we imitate." How can you, when they are unlike both each other and the rest? "Thucydides," he says, "is whom we imitate." Excellent — if you mean to write history, not if you mean to plead cases. For Thucydides was a chronicler of deeds done, sincere and grand as well; he never handled this forensic, combative, judicial kind. As for the speeches he inserted — and there are many of them — those I am in the habit of praising; imitate them I could not, if I wished, nor would I wish to, perhaps, if I could. It is as if a man should take pleasure in Falernian wine, but neither so new that he would want it born under the latest consuls, nor again so old that he must look for Opimius or Anicius as consul — "and yet those vintages are the very best." I believe it; but excessive age has not the sweetness we seek, nor is it now really tolerable —

\ciceroSection{288} so then, would a man who feels this, if he wished to drink, suppose he must draw for himself straight from the vat? By no means; rather he would seek a certain age. So I should counsel these men too that this new speech of theirs, hot as it were from the must and the vat, must be shunned, and that the splendid speech of Thucydides, too old, like the Anician vintage, must not be pursued. For Thucydides himself, had he come later, would have been much riper and milder. "Let us imitate Demosthenes, then." Good gods — what else, I ask, are we doing, or what else do we desire? But we do not attain it. For these Attics of ours, naturally, do attain what they wish.

\ciceroSection{289} They do not even understand this: that it was not only handed down to memory, but had to be so, that when Demosthenes was going to speak, throngs would gather from all over Greece for the sake of hearing him. But when these Attics speak, they are abandoned not only by the ring of bystanders — which is itself a pitiable thing — but even by the advocates. Therefore, if to speak narrowly and meagrely is the mark of the Attics, let them be Attics, by all means; but let them come into the Comitium, let them speak before a standing judge: the benches call for a grander and fuller voice.

\ciceroSection{290} This is what I wish for the orator: that when it has been heard that he is going to speak, a place on the benches is seized, the tribunal is filled, the clerks are obliging in giving and yielding a place, the ring of bystanders is many-layered, the judge alert; that when the man who is going to speak rises, silence is signalled by the crowd, then come frequent assents, many cries of admiration; laughter when he wishes it, when he wishes it tears — so that a man who watches all this from afar, even if he does not know what is being done, may yet understand that it is going well and that there is a Roscius on the stage. To whomever this falls, know that he speaks in the Attic manner, as we hear of Pericles, of Hyperides, of Aeschines, and of Demosthenes himself most of all.

\ciceroSection{291} But if they approve a kind of speech that is keen and prudent and at the same time sincere and solid and dried out, and do not use that weightier oratorical ornament, and wish this to be the proper mark of the Attics, they praise it rightly. For in an art so great and so varied there is a place even for this fine-spun subtlety. So it will turn out that not all who speak in the Attic manner speak well too, but that all who speak well speak also in the Attic manner. But let us return again to Hortensius.

\ciceroSection{292} By all means, said Brutus; though that digression of yours from the speech you had set out was very pleasant to me. Then Atticus said: Several times I have tried to break in, but I did not wish to interrupt. Now, since your discourse seems already to be looking toward its conclusion, I shall say, I think, what I feel. Do, by all means, Titus, I said. Then he said: I hold that famous irony which they say was in Socrates, which he uses in the books of Plato and Xenophon and Aeschines, to be witty and elegant. For it is the mark of a man by no means inept, and of the same man witty as well, when the discussion is about wisdom, to take this away from himself and to assign it, in mockery, to those who arrogate it to themselves — as in Plato Socrates exalts to the skies with praises Protagoras, Hippias, Prodicus, Gorgias, and the rest, while he feigns himself ignorant of all things and unschooled. This suits him, somehow or other, and I do not agree with Epicurus, who finds fault with it. But in history, which you have employed throughout the whole discourse, when you were setting out what each orator had been like — see, I beg you, he said, that the irony is not as much to be censured here as it would be in giving testimony. To what end is that? I said; for I do not understand.

\ciceroSection{293} Because, first, he said, you praised certain orators in such a way that you could lead the inexperienced into error. For my part, I could scarcely hold back my laughter in some cases, when you were comparing our Cato to the Attic Lysias — a great man, by Hercules, or rather a supreme and singular man; no one will say otherwise — but an orator? but even like Lysias, than whom nothing can be more finished? A pretty irony, if we were joking; but if we are speaking in earnest, see that we must use as much scruple as if we were giving testimony.

\ciceroSection{294} For my part, I approve your Cato as a citizen, as a senator, as a commander, as a man, in short, excelling both in prudence and in diligence and in every virtue; and his speeches, for those times, I praise highly — for they mark out a certain shape of genius, but one quite unpolished and plainly rough. But when you said that the \textit{Origines} were crammed with all the orator's merits, and compared Cato with Philistus and Thucydides, did you suppose you would convince Brutus of that, or me? For men whom not even any of the Greeks can imitate — to these you compare a man of Tusculum who did not yet so much as suspect what it was to speak with abundance and ornament.

\ciceroSection{295} You praise Galba. If as the leading man of that age, I agree — for so we have received it; but if as an orator, produce the speeches, I beg you — for they exist — and tell me that you wish this Brutus, whom you love more than yourself, to speak in that manner. You approve the speeches of Lepidus; here, Paulus, I agree with you, provided you praise them as ancient; the same of Africanus, of Laelius, than whose speech you say nothing can be made sweeter — and you add something or other more majestic still. You take us in by the name of a supreme man and by the truest praises of a most refined life. Remove these: that sweet speech of yours will be so cast down that no one will want to look at it.

\ciceroSection{296} I know that Carbo was held among the foremost orators; but, as in other matters, so in speaking, what is now the best, whatever its quality, is regularly the thing praised. I say the same of the Gracchi, although the things you said of them are ones I agree with. I pass over the rest; I come to those in whom you think eloquence already perfected, men whom I myself heard, and who were beyond dispute great orators, Crassus and Antonius. As to their praises I agree with you entirely, but still not in this fashion: as Lysippus used to say that the Doryphorus of Polyclitus was his master, so you say that the speech in support of the Servilian law (a law on the courts) was your teacher; this is genuine irony. Why I think so I shall not say, lest you suppose I am flattering you.

\ciceroSection{297} I pass over, then, the things you said about these very men, about Cotta, about Sulpicius, about Caelius just now. For these men were certainly orators; how great and of what quality, you must see. For that other matter I care less about — that you have heaped up all the day-labourers; so that it seems to me some of them must have wished to die, in order to be reckoned by you in the number of orators. When he had said this: You have struck up the beginning of a long discourse, I said, Atticus, and stirred up a matter worthy of a fresh debate, which let us put off to another time.

\ciceroSection{298} For the books must be turned over — those of others, and chiefly Cato's. You will understand that nothing was lacking to his outlines except the bloom and colour of those pigments which had not yet been discovered. As for Crassus's speech, my view is this: he himself could perhaps have written better, but no one else, I think. For do not take me to be an ironist in this, that I said that speech was my teacher. For although you seem to judge better of whatever faculty we now possess, still, as young men we had nothing else in Latin that we might rather imitate.

\ciceroSection{299} And that I named more men than I should have — that aimed, as I said a little while ago, at making it understood how few, in that pursuit which all most ardently desire, turn out worthy of the name. Therefore I would not have myself judged an ironist — not even if Africanus was one, as C. Fannius says in his history. As you will, said Atticus. For I did not think a thing alien to you which had been both in Africanus and in Socrates.

\ciceroSection{300} Then Brutus said: Of that later; but you, he said, looking at me, will you unfold for us the old speeches? Yes indeed, Brutus, I said; but at the villa at Cumae or at Tusculum some time, if only it be allowed, since at both places we are neighbours. But now let us return to the point from which we digressed.

\ciceroSection{301} Hortensius, then, having begun to speak in the Forum while quite a young man, was quickly taken on for the greater cases; and although he had fallen into the age of Cotta and Sulpicius, who were ten years older — Crassus and Antonius then standing pre-eminent, then Philippus, afterward Iulius — he was matched in the glory of speaking with these very men. First, a memory so great as I think I have known in no one, so that what he had thought out within himself he would deliver without a script in the very same words in which he had conceived it. This great aid he used in such a way that he remembered both his own thought-out and written matter and, with no one prompting, every word of all his adversaries.

\ciceroSection{302} He burned, moreover, with such eagerness that I never saw in anyone a more blazing zeal. For he would suffer no day to pass without either pleading in the Forum or rehearsing outside it. Very often he did both on the same day. And he had brought in a kind of speaking by no means common; two things, indeed, that no one else had: the divisions of the matters on which he was going to speak, and the recapitulations, in which he kept in mind both what had been said against him and what he himself had said.

\ciceroSection{303} He was elegant in the splendour of his words, apt in his arrangement, copious in his faculty; and these he had attained both by the highest genius and by the greatest exercises. He grasped the matter from memory, divided it keenly, and scarcely passed over anything in the case that served either to confirm or to refute. His voice was tuneful and sweet; his movement and gesture had even more of art than was enough for an orator. As he, then, was coming into flower, Crassus died, Cotta was driven out, the courts were interrupted by the war, and we came into the Forum.

\ciceroSection{304} Hortensius was, in the war, in the first year a soldier, in the second a military tribune; Sulpicius was a legate; M. Antonius too was away. The court under the one Varian law was kept active, the others being interrupted on account of the war; I was a constant attender there, although the orators pleaded for themselves — not, indeed, the foremost orators, L. Memmius and Q. Pompeius, but orators nonetheless, with Philippus at all events as their eloquent witness, whose testimony had both the vehemence of an accuser and his abundance.

\ciceroSection{305} The rest who were then counted among the foremost were in magistracies, and almost daily we heard them in the public assemblies. For the tribune of the plebs at that time was C. Curio — though he, indeed, kept silent, ever since he was once abandoned by the whole assembly; Q. Metellus Celer, not an orator, indeed, but still not tongue-tied; while the eloquent ones were Q. Varius, C. Carbo, Cn. Pomponius — and these men, indeed, lived on the Rostra; C. Iulius too, the curule aedile, held carefully wrought assemblies almost daily. But me, most eager to listen, the first grief struck when Cotta was expelled. Hearing the rest frequently, I was held by the keenest zeal, and daily, both writing and reading and rehearsing, I was not content with oratorical exercises alone. By now, in the following year, Q. Varius, condemned under his own law, had gone into exile;

\ciceroSection{306} As for me, I was giving much effort to the study of the civil law under Quintus Scaevola, son of Quintus, who, though he gave himself to no one as a teacher, nevertheless instructed those eager to learn by answering the men who came to him for counsel. And the year next after this one fell in the consulship of Sulla and Pompeius. Then, while Publius Sulpicius was haranguing the people daily during his tribunate, I came to know his whole manner of speaking thoroughly; and at that same time, when Philo, the head of the Academy, had fled his home with the leading men of the Athenians in the Mithridatic war and come to Rome, I handed myself over to him entirely, fired by a certain marvellous passion for philosophy — in which I lingered the more attentively (even though the variety and the magnitude of the subjects themselves held me with the keenest delight) because it now seemed that the whole system of the courts had been done away with for good.

\ciceroSection{307} Sulpicius had fallen in that year, and in the next the three orators of three generations were most cruelly put to death: Quintus Catulus, Marcus Antonius, Gaius Iulius. In the same year, too, at Rome, I gave my effort to Molon of Rhodes, both a supreme pleader of cases and a teacher. These things, though they seem foreign to the subject I set out on, are nevertheless brought forward by me so that you, Brutus, may be able to trace my course, since you wished it — for to Atticus these things are known — and may see in what manner I followed Quintus Hortensius down the track, in his very footsteps. For some three years the city was without arms; but with the orators either dead, or withdrawn, or in flight — for the young men too were away, Marcus Crassus and the two Lentuli — Hortensius held the first place in the courts, Antistius won more and more approval daily, Piso spoke often, Pomponius less often, Carbo rarely, Philippus once or twice. But I, for all that time, spent my nights and days in the study of every kind of learning.

\ciceroSection{309} I was with the Stoic Diodotus, who, after he had lived under my roof and shared his life with me, died lately in my house. Under him I was exercised, among other things, most diligently in dialectic, which is to be reckoned a kind of eloquence drawn close and bound tight — without which even you, Brutus, have judged that you cannot attain that true eloquence which they hold to be dialectic spread wide. Yet, devoted as I was to this teacher and to his various and many arts, no day was left empty of the exercises of oratory.

\ciceroSection{310} I would prepare and declaim — for so they speak now — often with Marcus Piso and with Quintus Pompeius, or with someone every day; and this I did much in Latin too, but more often in Greek, either because Greek speech, furnishing more ornaments, brought with it the habit of speaking in like manner in Latin, or because by the foremost Greek teachers, unless I spoke in Greek, I could neither be corrected nor taught.

\ciceroSection{311} Meanwhile came the upheaval in the recovery of the commonwealth and the cruel destruction of the three orators Scaevola, Carbo, Antistius; the return of Cotta, Curio, Crassus, the Lentuli, Pompeius; the laws and the courts re-established, the commonwealth recovered; while from the number of the orators Pomponius, Censorinus, Murena were taken off. Then for the first time I began to come forward to cases both private and public — not, as most have done, that I might learn in the Forum, but that I might come into the Forum already trained, so far as I had been able to make myself so.

\ciceroSection{312} At that same time I gave my effort to Molon; for under the dictator Sulla he had come as an envoy to the Senate about the rewards of the Rhodians. And so my first public case, spoken on behalf of Sextus Roscius, had so much to recommend it that there was none which did not seem worthy of my advocacy. Thereafter many followed, which I brought forward carefully wrought and, as it were, worked through by lamplight.

\ciceroSection{313} Now, since you seem to wish to know me whole, not by some single birthmark or rattle but in the whole body, I shall take in some things as well that may perhaps seem less than necessary. There was at that time in me an extreme slenderness and weakness of body, a long and thin neck — a constitution and a frame that are thought to stand not far from danger to life, if hard labour and great straining of the lungs are added. And this troubled the more those to whom I was dear, because I spoke everything without let-up, without variety, with the utmost force of voice and a straining of the whole body.

\ciceroSection{314} And so, when both my friends and my physicians were urging me to give up pleading cases, I thought I must rather face any peril whatever than fall away from the glory of speaking I had set my hopes on. But when I came to judge that, by an easing and moderation of the voice and a changed manner of speaking, I could both escape the danger and speak more temperately, the cause of my setting out for Asia was just this — to change my habit of speaking. And so, when I had been engaged in cases for two years and my name was already much spoken of in the Forum, I set out from Rome.

\ciceroSection{315} When I had come to Athens, I spent six months with Antiochus, the most distinguished and most judicious philosopher of the Old Academy; and the study of philosophy, never broken off, cultivated from earliest youth and always increased, I renewed afresh under this supreme master and teacher. At the same time, however, at Athens, I used to exercise myself diligently with Demetrius the Syrian, an old and not undistinguished teacher of speaking. After that the whole of Asia was traversed by me, in the company indeed of the greatest orators, with whose glad consent I exercised myself; among whom the foremost was Menippus of Stratonicea, in my judgment the most eloquent man in all Asia in those days; and if to have nothing tiresome and nothing in poor taste is the mark of the Attici, this orator can rightly be counted among them.

\ciceroSection{316} But the one most constantly with me was Dionysius of Magnesia; there was also Aeschylus of Cnidos, and Xenocles of Adramyttium. These were then reckoned the chief among the rhetoricians of Asia. Not content with them, I came to Rhodes and attached myself to that same Molon whom I had heard at Rome — a man both an outstanding pleader and writer in real cases, and most judicious in marking and observing faults and in shaping and teaching. He took pains, if only he could achieve it, to check us when we overflowed too far and ran beyond our banks with a certain youthful licence and impunity of speech, and to keep us in, as it were, when we were streaming out past the margins. So I came back two years later not only better trained but well-nigh transformed. For the excessive straining of the voice had settled, the style had, so to speak, boiled itself off, my lungs had gained strength, and my body a moderate condition.

\ciceroSection{317} Two orators then stood out who might rouse me with a desire to imitate them, Cotta and Hortensius — the one easy and gentle, grasping his thought in fitting words, fluently and without strain; the other ornate, keen, and not such as you knew him, Brutus, already in his decline, but more stirring in his kind of language and of delivery. And so I judged that my business lay more with Hortensius, because I was both nearer to him in the ardour of speaking and closer in age. For I had seen, in the very same cases — as for Marcus Canuleius, as for the consular Gnaeus Dolabella — that, though Cotta had been brought in as the leading advocate, Hortensius nevertheless played the chief part. For the throng of men and the din of the Forum call for a keen orator, kindled, active, and ringing of voice.

\ciceroSection{318} For one year, then, after we had come back from Asia, we pleaded notable cases, while I was a candidate for the quaestorship, Cotta for the consulship, Hortensius for the aedileship. Meanwhile the Sicilian year took me up as quaestor; Cotta, his consulship over, set out for Gaul; Hortensius both was, and was held to be, the first. But when, a year later, I had returned from Sicily, that quality in me, whatever it was, now seemed to have reached its full growth and to have a certain ripeness of its own. I seem to be saying too much about myself, especially I myself; but the aim set before this whole discourse is not that you should discern my talent and eloquence — from which I am far removed — but my toil and industry.

\ciceroSection{319} When, therefore, I had been engaged for some five years in very many cases and among the leading advocates, I came, as aedile-elect, into a very great contest, in the Sicilian advocacy, against the consul-elect Hortensius. But since this whole discourse of ours calls not only for an enumeration of the orators but also for certain precepts, I may briefly say what is, as it were, to be marked and observed in Hortensius.

\ciceroSection{320} For after his consulship — because, I suppose, he saw that none of the consulars was to be compared with him, while he disregarded those who had not been consuls — he relaxed that supreme zeal of his with which he had been fired from boyhood, and, in the abundance of all things, chose to live more happily, as he himself thought, and more at ease, certainly. The first year and the second and the third had drawn off from his colour, as from the colour of an old painting, just so much as not any one man of the crowd, but a learned and discerning critic, could perceive. But going on further, sticking fast — as in the other parts of eloquence, so most of all in the speed and the unbroken flow of words — he seemed to be growing more unlike himself with every day.

\ciceroSection{321} But I did not cease, with every kind of exercise and most of all with the pen, to add to that quality of mine, such as it was, whatever I could. And, to pass over much in this stretch and in these years after the aedileship — both praetor at the head of the poll and, by an incredible goodwill of the people, I was made consul. For both by my constancy and industry in cases, and by a more refined and least common kind of speech, I had turned the minds of men towards myself by the novelty of my delivery.

\ciceroSection{322} I shall say nothing about myself: I shall speak of the others, of whom there was not one who seemed to have applied himself to letters — in which the spring of perfect eloquence is contained — with any more refinement than the common run of men; not one who had embraced philosophy, the mother of all things well done and well said; not one who had learned the civil law, a thing most necessary for private cases and for the orator's good judgment; not one who held in his grasp the record of Roman history, out of which, whenever there was need, he might call up from the dead the richest of witnesses; not one who could, by briefly and wittily making a mock of his adversary, loosen the minds of the jurors and lead them for a little while from severity to cheerfulness and laughter; not one who could broaden his theme and carry his speech over from the particular and bounded dispute of a man and a moment to the general question of the whole class; not one who could digress a little from the case for the sake of giving pleasure; not one who could move the juror mightily to anger; not one who could bring him to tears; not one who could drive his mind — the thing that is most peculiarly the orator's own — wherever the matter demanded.

\ciceroSection{323} And so, when Hortensius had now all but faded away and I had been made consul in my own year — the sixth year, however, after he had been consul — he began to call himself back to industry, lest, now that we were equal in office, I should seem superior in any respect. Thus for twelve years after my consulship we were engaged together, in the closest union, in the greatest cases, while I set him before myself and he set me before himself; and my consulship, which at first had slightly grazed him, had at the same time joined us in the praise of my own achievements, which he admired.

\ciceroSection{324} But the training of us both was seen most clearly a little before this study of ours, Brutus, terrified by arms, fell suddenly silent and was struck dumb: when, by the Pompeian law, with three hours each allotted for speaking, we came daily to fresh cases most like to one another, or rather to the same cases. In these cases you too, Brutus, were present, and pleaded several both with us and alone, so that Hortensius, though he did not live long enough, nevertheless completed this course: sixteen years before you were born he began to plead cases; and the same man, in his sixty-fourth year, a very few days before his death, defended together with you your father-in-law Appius. As for what the kind of speaking was in each of us, the speeches of us both will make plain even to those who come after.

\ciceroSection{325} But if we ask why Hortensius flourished more in speaking as a young man than as an older one, we shall find two causes, both perfectly true. The first, that his kind of speech was Asiatic, a kind more allowed to youth than to old age. And there are two kinds of the Asiatic style: one full of maxims and pointed, with maxims not so much weighty and severe as neat and graceful — such as Timaeus in history, and, in speaking, in our boyhood, Hierocles of Alabanda, and still more his brother Menecles, whose speeches both are among the most praiseworthy in the Asiatic kind. The other kind is not so crowded with maxims as winged and impetuous in its words — such as is now all Asia — not only with a flood of speech but also with an ornate and witty kind of diction; in which were Aeschylus of Cnidos and my contemporary, Aeschines of Miletus. In these there was an admirable rush of speech; the ornate, well-turned balance of maxims there was not.

\ciceroSection{326} These kinds of speaking, however, as I have said, are better fitted to young men; in old men they have no weight. And so Hortensius, flourishing in both kinds, drew shouts of applause as a young man. For he had both that Meneclean zeal for frequent and graceful maxims — in which, as in that Greek, so in him there were certain maxims more graceful and sweet than either necessary or, at times, useful — and his speech was at once impetuous and quivering, and also careful and polished. These things did not win approval from the old men — often I saw Philippus now laughing at him, now even angry and out of patience — but the young men admired him, and the crowd was moved. He was outstanding in the judgment of the common people, and as a young man easily held the first place. For even though that kind of speaking had too little authority, it nevertheless seemed suited to his age. And certainly, because both a certain shape of talent shone out and a finished training of words in close-bound compass, he stirred up the highest admiration of men. But when at length the offices and that older man's authority called for something weightier, he remained the same when it no longer became him to be the same; and because he had let go his training and his zeal — the thing in him that had been keenest — that old balance and frequency of maxims remained, but it was no longer adorned with that dress of speech to which it had been accustomed. This, Brutus, perhaps pleased you less than it would have pleased you, if you had been able to hear him blazing with zeal and in the flower of his powers. Then said Brutus: For my part, I both see what kind of thing these things are which you describe, and I always thought Hortensius a great orator, and approved of him most of all when he spoke for Messalla, while you were away. So they say, said I; and the speech itself, written, as they tell us, in just as many words as he spoke, makes it plain. He, then, flourished from the consulship of Crassus and Scaevola down to the consuls Paullus and Marcellus; we held the same course from the dictatorship of Sulla to almost the same consuls. So the voice of Quintus Hortensius was extinguished by his own destiny, mine by the destiny of the state.

\ciceroSection{329} Speak a better omen, I beg you, said Brutus. Let it be just as you wish, said I, and that not so much for my sake as for yours; but fortunate was his ending, who did not see, when they came to pass, the very things he had foreseen would come. For we often wept together over the calamities hanging over us, when we saw the causes of civil war shut up in the appetites of private men and the hope of peace shut out from the counsels of the state. But his own good fortune, which he always enjoyed, seems to have set him free, by his death, from the miseries that followed.

\ciceroSection{330} But as for us, Brutus, since after the death of Hortensius, that most illustrious orator, we have been left as the guardians, so to speak, of an orphaned eloquence, let us hold her at home, fenced within an honourable keeping, and let us reject these unknown and shameless suitors, and protect her like a grown maiden, chastely, and keep her, so far as we can, from the assault of the lovers. And though, for my part, I grieve that, having entered upon life a little late, as upon a road, I have run, before the journey was done, into this night of the commonwealth, yet I am held up by that consolation which you, Brutus, brought me in your most charming letter, in which you reckoned that I ought to be of stout heart, because I had done deeds which would speak of me even though I held my peace, and would live on though I were dead — deeds which, if all went rightly, would bear witness to my counsels for the commonwealth by the safety of the commonwealth, and, if otherwise, by its very ruin.

\ciceroSection{331} But when I look upon you, Brutus, I grieve, in that, as you rode through the midst of praise in your youth, as if in a four-horse chariot, the wretched fortune of the commonwealth came crosswise upon you. This grief touches me, this care troubles me, and troubles this man with me, my partner in the same love and the same judgment. We wish you well; we long that you should enjoy your own worth; we wish for you that commonwealth in which you may be able to renew and to enlarge the memory of two most distinguished houses. For yours was the Forum, yours was that course; you had come there alone — you who had not only sharpened your tongue by the exercise of speaking, but had enriched eloquence itself with the equipment of the weightier arts, and by those same arts had joined every honour of virtue to the highest praise of eloquence.

\ciceroSection{332} From you a twofold anxiety afflicts us: that you yourself are deprived of the commonwealth, and the commonwealth of you. Yet you, Brutus, though this untimely ruin of the state presses upon the course of your talent, hold yourself fast to your unbroken studies, and bring to pass what you had now well-nigh, or rather quite fully, brought to pass — that you should pluck yourself out of that crowd of advocates which I have heaped up into this discourse. For it does not become you, adorned with the richest arts — which, when you could not draw them from home, you sent for out of that city which has always been held the home of learning — to be counted in the common throng of advocates. For to what end did Pammenes, by far the most eloquent man of Greece, train you? to what end that old Academy and its heir, Aristus, my guest and intimate friend, if indeed we are to be like the greater part of the orators?

\ciceroSection{333} Do we not see that, in single generations, scarcely two praiseworthy orators have stood firm? Galba was the one man pre-eminent among so many of his time, to whom, as we have received it, both Cato yielded, the elder man, and those who in those days were younger in age; afterwards Lepidus, then Carbo; for the Gracchi, in their harangues, with a far easier and freer kind of speaking — though even down to their own day the praise of eloquence was not yet perfected; then Antonius, Crassus; after them Cotta, Sulpicius, Hortensius. I say nothing further; only this I say: if it had befallen me to be counted among the many... if the running together of the more well-placed is laborious...
