% Topica ad C. Trebatium --- headnote
% topica, 44 BC, Rhegium / shipboard

\textit{The \textit{Topica} is a handbook of argument written for a
friend and composed, remarkably, from memory at sea. Its addressee is
Gaius Trebatius Testa, the eminent jurist whom Cicero had once
recommended to Caesar and teased in a string of affectionate letters.
Browsing in Cicero's library at Tusculum, Trebatius had come upon
Aristotle's \textit{Topica} and asked what it contained; put off at the
time, and then parted from his books, Cicero made good the promise on his
voyage toward Greece in July of 44 BC, writing the whole treatise out of
his head on shipboard and sending it back from the journey. He presents
it as an exposition of Aristotle, though in truth it owes at least as much
to the later rhetorical and Stoic schools, and bends throughout toward
the law, to suit the man it was written for.}

\textit{A "topic," in this tradition, is the seat of an argument --- a
standing place to which the mind goes to find what will lend conviction
to a doubtful matter. Cicero sorts these places into those that lie in
the subject itself and those drawn from outside it. The first group runs
through definition and partition, etymology and related words, genus and
species, likeness and difference, contraries, adjuncts, antecedents and
consequents and incompatibles, causes and effects, and comparison; the
second turns on testimony and authority. He then passes to the kinds of
question --- the general thesis and the particular case --- to the three
kinds of oratory, and to the doctrine of the "issues," the points on
which a dispute can be made to turn. At every step the illustrations are
Roman and legal, drawn from inheritance, property, water rights, and the
fixed formulae of the courts.}

\textit{For all that it is a small and hurried book, the \textit{Topica}
had an afterlife out of all proportion to its size. Through Boethius, who
wrote a commentary on it, its scheme of the loci passed into the logic
and rhetoric of the Middle Ages and the Renaissance, where the
"commonplaces" became a standard apparatus for invention in argument and
a fixture of education. It is also a precious record of Roman law in its
classical age, preserving the technical vocabulary and the worked
examples of the late-Republican jurists. Behind its dry order one can
still hear the warmth of the occasion: a great advocate, shut out of the
courts and uncertain of the road ahead, repaying a young lawyer's
curiosity with the most useful gift he had.}
