% De Optimo Genere Oratorum --- headnote
% de-optimo-genere-oratorum, 46 BC, Rome

\textit{The \textit{De Optimo Genere Oratorum} (``On the Best Kind
of Orators'') is a short preface --- all that survives of a larger
project. Cicero had translated into Latin the two great paired
speeches of the Crown trial, Aeschines' \textit{Against Ctesiphon}
and Demosthenes' \textit{On the Crown}, the most famous forensic
duel of the Greek world; this essay was to introduce those
translations, which are themselves now lost. Its central argument
is that there are not, properly speaking, many ``kinds'' of orators,
as there are kinds of poets, but one single perfect kind, and that
the rest fall short of it not in kind but in degree. The perfect
orator is the one who, in speaking, does three things at once ---
teaches, delights, and deeply moves: to teach is a debt owed, to
delight an honour conferred, to move a necessity. That single
standard, Cicero insists, is the genuinely Attic one, set by the
orators who flourished at Athens and crowned in Demosthenes; and
he wields it as a weapon against the young Roman ``Atticists,'' who
mistook a bare, thinned-down plainness for the whole of the Attic
ideal.}

\textit{The work is best known for its statement of method. Cicero
declares that he turned the two Greek speeches into Latin ``not as
a translator but as an orator'' --- keeping the same thoughts and
their figures, but in words fitted to Latin usage, and refusing to
render word for word; he did not count the words out to the reader
like coins, he says, but weighed them out. It is one of the
foundational statements of sense-for-sense over word-for-word
translation in the Western tradition. The essay belongs with the
\textit{Brutus} and the \textit{Orator} of the same year, 46 BC,
written in the literary leisure of Caesar's dictatorship: together
they form Cicero's mature defence, against the Atticists, of the
copious and fully armed eloquence he had practised all his life.
The preface breaks off as it turns to the speeches themselves ---
``Let us at last hear Aeschines himself speaking Latin'' --- and the
translations it announces do not survive.}
