% Orator --- headnote
% orator, 46 BC, Rome (dedicated to M. Iunius Brutus)

\textit{The \textit{Orator} is the last and most personal of
Cicero's three great works on the art of speaking, and the
crown of the trilogy that began with the \textit{De Oratore}
and continued in the \textit{Brutus}. Written late in 46 BC,
in the same enforced literary leisure of Caesar's dictatorship
that produced its companions, it answers a question Brutus had
pressed on Cicero again and again: what is the perfect orator
--- not who has ever been one, but what the thing itself would
be, the form than which nothing more excellent can be conceived?
Cicero takes up the challenge knowing it cannot fully be met,
and frames his answer in openly Platonic terms: as Phidias
worked with no living model before his eyes but a vision of
beauty held in the mind, so the ideal orator is an image the
intellect grasps and the hand never quite reproduces. The book
is thus not a history, as the \textit{Brutus} was, but a
portrait of a standard.}

\textit{Its argument turns on the doctrine of the three styles
--- the plain, the middle, and the grand --- and on the
governing claim that the supreme orator is the one who commands
all three and knows, by the unteachable sense of what is
fitting, which to use and when. From this Cicero mounts his
defence against the young men who called themselves
``Attici'': those who prized a bare, thinned-down plainness and
reproached his own copious, fully armed manner as Asiatic
excess. The plain style, he grants, is genuine and difficult;
but to make it the whole of eloquence is to feed on acorns
after the harvest has been found. The true Attic orator is not
the thinnest but the most complete --- and the model, against
the Atticists' own claim, is Demosthenes, who thundered in the
grand style they affect to despise.}

\textit{The longer half of the work is given over to the
subject that was Cicero's most original contribution to
rhetorical theory and his most influential legacy to Latin
prose: the doctrine of prose rhythm (\textit{numerus}).
Oratory, he argues, must be rhythmical without ever becoming
verse; the ear, not the rules of the metricians, is the final
judge, but the ear can be trained to know which feet --- the
cretic, the paean, the dactyl, the spondee --- carry weight,
and where rhythm tells most, which is above all in the close of
the period, the clausula on which a whole sentence is felt to
land. The analysis is technical, sometimes minute, and depends
on quantities of the Latin syllable that no translation can
fully carry across; but its principle --- that the music of a
sentence is built, not stumbled into --- shaped the writing of
Latin prose for centuries. The treatise ends, like the
\textit{Brutus}, in a turn toward Brutus himself: a defence of
the work undertaken, and the confession that, unable to refuse
a friend, Cicero took on the impudence of writing what he knew
could not be perfectly done.}
