% De Oratore --- headnote
% de-oratore, 55 BC, Rome / Tusculum (dedicated to Quintus Cicero)

\textit{De Oratore, Cicero's three-book dialogue on the ideal
orator, written in 55 BC and dedicated to his brother Quintus.
The work is the mature replacement for the schoolboy
\textit{De Inventione} (85 BC), which Cicero by now repudiated;
it is also the first sustained philosophical work of the
corpus, written in the temporary literary leisure between the
exile-year of 58 BC and the consular wars to come.}

\textit{The dialogue is set in 91 BC at the Tusculan villa of
Lucius Licinius Crassus during the Roman games, in the days
just before Crassus's sudden death and the outbreak of the
Social War. The speakers are the two greatest orators of
Cicero's youth: Crassus himself (the Stoic-Academic, master
of cultivated speech) and Marcus Antonius (the practical
forensic master, grandfather of Mark Antony), with Q. Mucius
Scaevola the Augur, the youthful Sulpicius Rufus and Aurelius
Cotta, and on the second day Quintus Catulus and C. Iulius
Caesar Strabo. Each of the three books takes roughly a single
day's conversation: the first on whether the orator must
master a wide knowledge of philosophy, law, and history;
the second on the art of finding, arranging, and remembering
arguments (with the famous digressions on history and humour);
the third on style and delivery.}

\textit{The architectural argument is that the orator
\textit{must} have the encyclopaedic education the philosophers
claim — but he must wear it as a master of practical speech,
not a pedant. The position is Cicero's own programme, set
between the Greek philosophical schools and the workaday Roman
forensic tradition; Crassus's voice carries it, but Antonius's
counterweight is honoured. The dialogue's elegiac frame —
Crassus on the eve of his death, the Republic on the eve of
the Social War — is set against Cicero's own situation in 55
BC, the year of Pompey's stone theatre and the second
Caesar–Pompey–Crassus consensus at Luca: a moment of forced
literary quiet between political storms.}
