% De Fato (de-fato) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} \dots because it bears on character, which they call "character" \textit{[Greek: ēthos]}, we are accustomed to call that part of philosophy "on character," but it is fitting, as one who enlarges the Latin tongue, to name it the moral part. We must also set out the force and the principle of propositions, which the Greeks call \textit{[Greek: axiōmata]}; and as for what force they have when they say something about a future thing and about that which can come to pass or cannot, this is an obscure question, which the philosophers call "on possibles" \textit{[Greek: peri dynatōn]}, and the whole of it is logic \textit{[Greek: logikē]}, which I call the method of reasoning. But what I did in those other books, the ones on the nature of the gods, and likewise in those I published on divination — to set out a continuous discourse on both sides, so that each man might more readily accept what seemed most probable to him — that, in this discussion of fate, a certain accident prevented me from doing.

\ciceroSection{2} For when I was at my place near Puteoli, and our friend Hirtius, the consul-elect, was in the same parts — a man most devoted to me and to those studies in which I have lived since boyhood — we were much together, chiefly indeed inquiring into those counsels that bore on peace and on the concord of the citizens. For since, after the death of Caesar, every cause of fresh disturbances seemed to be in the seeking, and we judged that they had to be met, almost all our talk was spent on these deliberations; and so it was both often on other occasions and, on one day rather freer than usual and more clear of interruptions, when he had come to me, that we began with the matters that were our daily and, as it were, regular business, on peace and on quiet.

\ciceroSection{3} When these had been gone through: "What then?" said he. "Since you have not, I hope, given up your oratorical exercises, but have certainly set philosophy before them — may I hear something?" By all means, said I, either hear or speak; for I have not, as you rightly suppose, deserted those studies of oratory, with which I even kindled you, though I had received you already blazing; nor do the matters I now handle diminish that faculty, but rather enlarge it. For the orator has a great fellowship with this kind of philosophy that I follow; he borrows subtlety from the Academy and gives back to it in turn a richness of speech and the ornaments of expression. And so, said I, since both studies are our possession, today let it be your choice which you would rather enjoy. Then said Hirtius: Most welcome, and like all that is yours; for never has your goodwill refused anything to my eagerness.

\ciceroSection{4} But since your rhetorical works are known to me, and I have both often heard you in them and shall hear you again — and since the \textit{Tusculan Disputations} make plain that you have taken up this Academic practice of arguing against a thesis put forward — I should like, if it is no trouble to you, to lay something down for me to hear argued. Can anything be a trouble to me, said I, that will be welcome to you? But you shall hear me as a Roman, as one stepping timidly into this kind of argument, as one taking up these studies again after a long interval. So I shall hear you arguing, said he, as I read the things you have written. Then begin. Let us sit down here.

\ciceroSection{5} \dots in some of which — as in the poet Antipater, as in those born on a winter's day, as in brothers who fall sick together, as in urine, as in fingernails, as in the rest of that sort — the contagion of nature has its force, which I do not abolish, but there is no force of fate; while in others there may be certain things that come by chance, as in that famous shipwrecked man, as in Icadius, as in Daphitas. Some things, too, Posidonius — let me say it with the master's leave — seems to invent; and they are indeed absurd. For what of it? If it was Daphitas's fate to fall from a horse and so to perish, was it from this horse, which, since it was no horse, bore another's name? Or was Philip warned to beware these little four-horse teams on his sword-hilt? As if, forsooth, he were slain by a hilt. And what is so remarkable about that nameless shipwrecked man having slipped into a stream? — though here at least the writer records it was foretold to him that he would die in water. Nor, by Hercules, do I see any fate at all for the pirate Icadius;

\ciceroSection{6} for he records that nothing was foretold to him. What wonder, then, that a rock from the cave fell upon his legs? For I suppose that, even if Icadius had not then been in the cave, that rock would have fallen all the same. For either nothing whatever is fortuitous, or this very thing could have come about by chance. I ask, therefore — and this point will reach far — if there were no name of fate at all, no nature, no force, and most things or all things came to pass by chance, at random, by accident, would they come to pass otherwise than they now come to pass? What use is it, then, to thrust fate in, when without fate the account of all things may be referred to nature or to fortune?

\ciceroSection{7} But let us dismiss Posidonius, as is fair, with our good will, and return to Chrysippus's snares. To him let us first reply about this contagion of things itself, and pursue the rest afterward. We see how great is the difference among the natures of places: some are healthful, some pestilent; in some men are full of phlegm and, as it were, brimming over, in others dried out and parched; and there are many other things in which place differs greatly from place. At Athens the air is thin, from which the Attic people are reckoned the sharper; at Thebes it is thick, and so the Thebans are heavy and robust. Yet that thin air will not bring it about that a man hears Zeno, or Arcesilaus, or Theophrastus, nor will the thick air bring it about that one seeks victory at Nemea rather than at the Isthmus. Draw the distinction wider.

\ciceroSection{8} For what can the nature of a place contribute to make us walk in Pompey's portico rather than in the open field? With you rather than with another? On the Ides rather than the Kalends? As, then, the nature of a place bears upon certain things somewhat, and upon certain things nothing at all, so let the disposition of the stars have its force, if you like, upon certain things; upon all things it surely will not. But, you say, since there are unlikenesses in the natures of men — so that sweet things delight some, slightly bitter things others, and some are lustful, some prone to anger or cruel or proud, while others shrink from such vices — since, then, says he, one nature stands so far apart from another, what wonder is it that these unlikenesses have been produced by differing causes?

\ciceroSection{9} In arguing thus he does not see what the matter is about, and where the case stands. For if some men are more inclined toward some things on account of natural and antecedent causes, it does not therefore follow that there are natural and antecedent causes also of our acts of will and of our appetites. For nothing would be in our power, were the case so. But as it is, we admit that whether we are sharp or dull, strong or feeble, is not in us. Yet whoever supposes it follows from this that not even our sitting or walking is a matter of will, he does not see what thing follows upon what. For granted that the clever and the slow are born so by antecedent causes, and likewise the strong and the feeble, it still does not follow that their sitting too, and their walking, and their doing of some deed, is fixed and settled by primary causes.

\ciceroSection{10} Stilpo, the Megarian philosopher, we have heard was a sharp man indeed and esteemed in those times. His own intimates write that he was both given to drink and given to women, and they write this not in blame, but rather to his praise; for a vicious nature, they say, had been so tamed and held down in him by learning that no one ever saw him the worse for wine, no one saw in him a trace of lust. What of it? Do we not read how Zopyrus the physiognomist marked Socrates — Zopyrus, who professed to discern men's characters and natures from the body, the eyes, the face, the brow? He said that Socrates was stupid and dull, because he did not have hollows above the collarbone; he said those parts were blocked and stopped up. He added, too, that he was given to women — at which Alcibiades is said to have raised a great laugh.

\ciceroSection{11} But these vices can be born from natural causes; to root them out, however, and to remove them utterly, so that the very man who was inclined toward them is called away from such great vices — that is not lodged in natural causes, but in will, in effort, in discipline. All of which are abolished, if the force and the nature of fate shall be established from the reasoning of divination. For if there is divination, from what perceived principles of the art does it set out? (By "perceived" I mean what are called in Greek "theorems" \textit{[Greek: theōrēmata]}.) For I do not believe that, with nothing perceived, either the rest of the craftsmen are engaged in their proper task, or those who use divination foretell things to come.

\ciceroSection{12} Let, then, the astrologers' perceptions be of this kind: If anyone, say, is born at the rising of the Dog Star, he will not die at sea. Be watchful, Chrysippus, lest you desert your own cause, in which you have a great struggle with Diodorus, a strong dialectician. For if this is true, when so connected — If anyone is born at the rising of the Dog Star, he will not die at sea — then this too is true: If Fabius is born at the rising of the Dog Star, Fabius will not die at sea. These two, then, are at war with each other: that Fabius is born at the rising of the Dog Star, and that Fabius will die at sea; and since it is laid down as certain in Fabius's case that he was born at the rising of the Dog Star, these too are at war: that Fabius exists, and that he will die at sea. Therefore this conjunction too is one of repugnant terms: Both Fabius exists, and Fabius will die at sea — which, as it has been put, cannot even come to pass. Therefore this — Fabius will die at sea — is of that kind which cannot come to pass. Everything, then, that is falsely said of the future cannot come to pass.

\ciceroSection{13} But this, Chrysippus, you are least of all willing to grant, and it is precisely on this very point that your great contest with Diodorus arises. For he says that only that can happen which either is true or will be true, and whatever will be he says must necessarily happen, and whatever will not be he denies can happen. You, on the other hand, say that things which will not be can still happen — that this gem can be shattered, even if that is never going to be the case; and that it was not necessary for Cypselus to reign at Corinth, although this had been declared a thousand years before by an oracle of Apollo. But if you are going to endorse those divine predictions, then you will also have to count things stated falsely about the future among the things that cannot happen — as if it were said that Africanus would take Carthage; and if a thing is said truly about the future, and so is going to come about, you will have to say it is necessary. And that is the whole view of Diodorus, hostile to you.

\ciceroSection{14} For if this connection is true — "If you were born at the rising of the Dog Star, you will not die at sea" — and the first part of the connected statement, "You were born at the rising of the Dog Star," is necessary (for all true statements about the past are necessary, as Chrysippus holds, dissenting from his teacher Cleanthes, because they are unchangeable and the past cannot turn from true to false); if, then, what comes first in the connection is necessary, what follows from it becomes necessary too. Although this does not seem to Chrysippus to hold in all cases; yet still, if there is a natural cause why Fabius should not die at sea, then Fabius cannot die at sea.

\ciceroSection{15} At this point Chrysippus, in a fever, hopes that the Chaldeans and the rest of the diviners are mistaken, and that they will not use connected statements in declaring their findings in this form — "If anyone was born at the rising of the Dog Star, he will not die at sea" — but will rather put it thus: "It is not the case both that someone was born at the rising of the Dog Star and that he will die at sea." What laughable license! To keep from falling into Diodorus's trap himself, he instructs the Chaldeans in the manner they ought to set out their findings. For I ask: if the Chaldeans speak in such a way as to put forward negations of indefinite conjunctions rather than indefinite connected statements, why cannot physicians, why cannot geometers, why cannot the rest do the same? The physician, first of all, will not propose what is clear to him in his art in this form — "If anyone's pulse beats thus, he has a fever" — but rather in this way: "It is not the case both that someone's pulse beats thus and that he does not have a fever." And likewise the geometer will not say, "On a sphere the greatest circles bisect one another," but rather in this way: "It is not the case both that there are greatest circles on a sphere and that they do not bisect one another."

\ciceroSection{16} What is there that cannot be transferred in that fashion from a connected statement to the negation of a conjunction? And indeed we can express the same things in other ways as well. Just now I said, "On a sphere the greatest circles bisect one another"; I can say, "If there are greatest circles on a sphere"; I can say, "Because there are greatest circles on a sphere." There are many forms of statement, and none more contorted than this, in which Chrysippus hopes the Chaldeans will rest content for the Stoics' sake.

\ciceroSection{17} Yet none of them speaks in that way; for it is a greater labor to learn these contortions of speech than the risings and settings of the constellations. But let us return to that thesis of Diodorus, which they call \textit{[Greek: peri dynaton]}, in which the question is examined of what force "the possible" has. Diodorus, then, holds that only that is possible which either is true or will be true. This point touches on the present inquiry — that nothing happens which was not necessary, and that whatever can happen either is already or will be, and that the things which are future can no more be changed from true into false than the things which are past; but that in past things the immutability is plain, while in certain future things, because it is not plain, it does not even seem to be present — as in the case of one gripped by a deadly disease it is true that "He will die of this disease," yet if this same thing be said truly of one in whom no such force of disease appears, it will come about none the less. So it turns out that no change at all from true into false can occur even in the future. For "Scipio will die" has such force that, although it is said about the future, still it cannot be turned into false; for it is said of a man, and a man must necessarily die.

\ciceroSection{18} In the same way, if it were said, "Scipio will die at night in his bedroom, overpowered by violence," it would be said truly; for that would be said to be going to be, which was in fact going to be; and that it was going to be ought to be understood from the fact that it has happened. Nor was "Scipio will die" any more true than "He will die in that manner," nor was it any more necessary for Scipio to die than to die in that manner, nor is "Scipio has been slain" any more unchangeable from true into false than "Scipio will be slain"; nor, since this is so, is there any reason why Epicurus should dread fate and seek protection from the atoms and lead them out of their path and at one stroke take on two insoluble difficulties: one, that something should come about without a cause — from which it follows that something comes from nothing, which neither he nor any natural philosopher allows; the other, that when two indivisible bodies are carried through the void, one should move in a straight line while the other swerves.

\ciceroSection{19} For Epicurus may grant that every proposition is either true or false without fearing that everything must therefore happen by fate; for it is not by eternal causes flowing from nature's necessity that this is true which is stated thus — "Carneades is going down to the Academy" — nor yet is it without causes, but there is a difference between causes that have gone before by chance and causes that contain within themselves a natural efficacy. So it was always true that "Epicurus will die when he has lived seventy-two years, in the archonship of Pytharatus," and yet there were no fated causes why it should so fall out, but, because it so fell out, it was going to fall out, just as it did fall out, by a fixed sequence of causes.

\ciceroSection{20} Nor do those who say that future things are unchangeable, and that a true future cannot be turned into false, thereby confirm the necessity of fate; they are interpreting the force of words. But those who bring in an everlasting sequence of causes — they strip the mind of man of free will and bind it fast with the necessity of fate. But enough of this; let us look at other matters. For Chrysippus argues in this way: "If there is any motion without a cause, then not every proposition — what the logicians call an \textit{[Greek: axioma]} — will be either true or false; for what does not have efficient causes will be neither true nor false; but every proposition is either true or false; therefore there is no motion without a cause. And if this is so, then all things that come about come about by antecedent causes; and if this is so, all things come about by fate. It follows, then, that whatever comes about, comes about by fate."

\ciceroSection{21} Here, in the first place, if I cared to side with Epicurus and deny that every proposition is either true or false, I would rather take that blow than grant that everything comes about by fate; for the former view admits of some debate, but the latter is not to be borne. And so Chrysippus strains every nerve to persuade us that every \textit{[Greek: axioma]} is either true or false. For just as Epicurus is afraid that, if he grants this, he must grant that whatever comes about comes about by fate (for if either of two alternatives is true from eternity, then it is also certain, and if certain, also necessary; thus he thinks both necessity and fate are confirmed) — so Chrysippus is afraid that, if he does not maintain that everything stated is either true or false, he will not hold that everything comes about by fate and from the eternal causes of future things.

\ciceroSection{22} But Epicurus thinks the necessity of fate is escaped by the swerve of the atom. And so a third kind of motion arises, beyond weight and impact, when the atom swerves by the least possible interval — he calls this the \textit{[Greek: elachiston]}. That this swerve happens without a cause, he is forced to admit, if not in words, then in fact. For one atom is not made to swerve by being struck by another. For how can one be struck by another, if individual bodies are carried by their weight straight downward in straight lines, as Epicurus holds? It follows that, if one is never driven aside by another, one does not even touch another. From which it follows that, even if there is an atom and it swerves, it swerves without a cause.

\ciceroSection{23} Epicurus brought in this account for this reason: because he was afraid that, if the atom were always carried by its natural and necessary weight, nothing would be free for us, since the mind would be moved in such a way as to be compelled by the motion of the atoms. Democritus, the author of the atoms, chose rather to accept that all things come about by necessity than to tear from the individual bodies their natural motions. More acutely Carneades, who used to teach that the Epicureans could defend their case without this fictitious swerve. For since they taught that there could be a certain voluntary motion of the mind, it was better to defend that than to bring in the swerve, especially since they could not find its cause; and once that was defended, they could easily withstand Chrysippus. For having granted that there is no motion without a cause, they would not grant that all things which come about come about by antecedent causes; for our will has no external and antecedent causes.

\ciceroSection{24} So we abuse the common usage of speech when we say that anyone wills or refuses something "without a cause"; for we say "without a cause" in such a way as to mean "without an external and antecedent cause," not "without any cause at all" — just as, when we say a vessel is empty, we do not speak as the natural philosophers do, to whom nothing is empty, but in such a way as to say, for example, that a vessel is without water, without wine, without oil; so, when we say the mind is moved without a cause, we mean that it is moved without an antecedent and external cause, not that it is moved with no cause whatever. Of the atom itself it can be said, when it is moved through the void by weight and heft, that it is moved without a cause, because no cause comes to it from outside.

\ciceroSection{25} On the other hand—lest all the natural philosophers laugh us to scorn for saying that anything comes to be without a cause—a distinction must be drawn, and we must put it this way: that it is the very nature of the individual atom to be moved by weight and gravity, and that this itself is the cause of its being carried as it is. In the same way, for the voluntary motions of the mind no external cause need be sought; for voluntary motion contains within itself that nature by which it is in our power and obeys us, and this not without a cause, since the cause of the thing is its own nature.

\ciceroSection{26} Since this is so, what reason is there why every proposition should not be either true or false—unless we have conceded that whatever comes to be, comes to be by fate? Because, he says, things that do not have causes for their being future cannot be true as future things; therefore it is necessary that the things which are true should have causes; and so, when they have come to pass, they will have come to pass by fate. The business is settled, then, if you must concede either that all things come to be by fate, or that something can come to be without a cause.

\ciceroSection{27} Or can this proposition not be true—"Scipio will take Numantia"—unless from all eternity a cause sowing cause from cause is going to bring it about? Or could it have been false, if it had been spoken six hundred ages before? And if at that time this proposition was not true, "Scipio will take Numantia," then neither, once the city has been overthrown, is this proposition true, "Scipio has taken Numantia." Can anything, then, have come to be that it was not true was going to be? For just as we call those things past which were truly present at an earlier time, so we shall call those things future of which it will be true at a later time that they are present.

\ciceroSection{28} Nor, if every statement is either true or false, does it follow at once that there are immutable causes, and those eternal, which forbid anything to fall out otherwise than it is going to fall out. The causes that bring it about that things spoken in this way are spoken truly—"Cato will come into the Senate"—are fortuitous causes, not bound up in the nature of things and the order of the world; and yet that he will come, when it is true, is just as immutable as that he has come, and for that reason fate or necessity need not be dreaded. For surely one must admit this: if the statement "Hortensius will come to his place at Tusculum" is not true, it follows that it is false. These men want neither—which cannot be. Nor will that lazy argument, as it is called, hinder us; for there is a certain argument the philosophers name \textit{[Greek: argos logos]}, the idle argument, which, if we obeyed it, we should do nothing whatever in life.

\ciceroSection{29} For thus they put the question: If it is fated for you to recover from this illness, then whether you call in a doctor or do not call one in, you will recover; likewise, if it is fated for you not to recover from this illness, then whether you call in a doctor or do not call one in, you will not recover; and one or the other is fated; therefore it makes no difference to call in a doctor. Rightly has this kind of questioning been named lazy and inert, since by the same reasoning every action would be stripped out of life. One may even alter it, so as not to attach the word "fate" and yet hold the same opinion, in this way: If from all eternity this was true—"You will recover from that illness"—then whether you call in a doctor or do not call one in, you will recover; and likewise, if from all eternity this was false—"You will recover from that illness"—then whether you call in a doctor or do not call one in, you will not recover; and so on with the rest. This argument is refuted by Chrysippus.

\ciceroSection{30} For some things, he says, among events are simple, some are coupled. Simple is this: "Socrates will die on that day"; for him, whether he has done anything or has done nothing, the day of dying is fixed. But if the fate is going to be this—"Oedipus will be born to Laius"—it will not be possible to say "whether Laius has lain with a woman or has not"; for the matter is coupled, and co-fated—for so he calls it, because it was fated both that Laius would lie with his wife and that he would beget Oedipus from her; just as, if it had been said "Milo will wrestle at Olympia," and someone were to add, "Then whether he has had an opponent or has not had one, he will wrestle," he would be in error; for "he will wrestle" is coupled, since without an opponent there is no wrestling. All catches of that kind, then, are refuted in the same way. "Whether you call in a doctor or do not call one in, you will recover" is a catch; for it is just as much fated to call in a doctor as to recover. These things, as I said, he calls co-fated.

\ciceroSection{31} Carneades did not approve of this whole class of argument, and thought this reasoning was concluded too rashly. And so he pressed his point another way and brought to bear no chicanery; his conclusion was this: If all things come to be by antecedent causes, all things come to be linked and woven together in a natural connection; and if that is so, necessity brings all things about; if that is true, nothing is in our power; but something is in our power; and yet, if all things come to be by fate, all things come to be by antecedent causes; therefore it is not by fate that whatever comes to be, comes to be.

\ciceroSection{32} Reason cannot be drawn tighter than this. For if anyone should wish to make the same point and say it thus—"If everything future is true from all eternity, so that it surely comes about in the very way it is going to be, then it is necessary that all things come to be linked and woven together in a natural connection"—he says nothing. For it makes a great difference whether a natural cause brings it about, from all eternity, that future things are true, or whether even without natural eternity it can be understood that the things which are future are true. And so Carneades used to say that not even Apollo could speak of future events except those whose causes nature contained in such a way that they had to come to be.

\ciceroSection{33} For with what in view would the god himself say that Marcellus, the man who was three times consul, would perish at sea? This indeed was true from all eternity, but it had no causes to bring it about. And so he held that not even past things, of which no signs survived as traces, were known to Apollo; how much less the future! For only when the efficient causes of each thing are known can it at last be known what is going to be. Therefore Apollo could not foretell of Oedipus, either, since no causes had been set down in the nature of things why it should be necessary for his father to be killed by him, nor anything of the kind. And so, if it is consistent for the Stoics, who say that all things come to be by fate, to endorse oracles of this sort and the rest of what is drawn from divination, while the same need not be said by those who say that the things which are future are true from all eternity—consider whether their case may not be the same as the Stoics'; for these men are pressed more narrowly, while the others' reasoning is loose and free.

\ciceroSection{34} But suppose it is granted that nothing can come to pass except by an antecedent cause—what is gained, if that cause is not said to hang upon eternal causes? Now a cause is that which brings about the thing of which it is the cause, as a wound brings death, indigestion an illness, fire heat. And so a cause is not to be understood in such a way that whatever precedes a thing is its cause, but that which efficiently precedes it; nor was my going down into the field the cause of my playing ball, nor was Hecuba the cause of destruction to the Trojans because she bore Alexander, nor Tyndareus the cause to Agamemnon because he begot Clytemnestra. For at this rate the well-dressed traveler too will be called the cause to the highwayman of his being robbed by him.

\ciceroSection{35} Of this kind is that passage of Ennius: "Would that in the grove of Pelion the fir timbers had not fallen to earth, cut down by axes!" He might have gone higher: "Would that no tree had ever been born on Pelion!"—and higher still: "Would that there were no Mount Pelion at all!"—and in the same way, going back to what came before, one may regress without end. "Nor thence had the ship taken the beginning of its launching." To what end these past things? Because there follows this: "For never would my mistress, straying, have carried her foot from home—Medea, sick at heart, wounded by savage love"—though those things did not bring on the cause of her love.

\ciceroSection{36} They say it makes a difference, however, whether a thing is of the kind without which something cannot be brought about, or of the kind with which something must necessarily be brought about. None of these, then, is a cause, since none brings about by its own force the thing of which it is called the cause; nor is that without which something does not come to be a cause, but that which, when it has been added, necessarily brings about the thing of which it is the cause. For while the bite of the serpent had not yet festered upon Philoctetes, what cause was contained in the nature of things that he would be left behind on the island of Lemnos? Afterward, though, the cause was nearer and more closely joined to the outcome.

\ciceroSection{37} It is the outcome, then, that lays bare the cause. But from all eternity this proposition was true: "Philoctetes will be left on the island," and it could not turn from true into false. For of two contrary statements—and by contraries here I mean those of which one affirms something and the other denies it—of these, then, in spite of Epicurus, one must be true and the other false. "Philoctetes will be wounded" was true through all the ages before; "He will not be wounded" was false. Unless, perhaps, we choose to follow the opinion of the Epicureans, who say that such propositions are neither true nor false, or—when they are ashamed of that—say something still more shameless: that disjunctions made of contraries are true, but that of the statements contained in them neither is true.

\ciceroSection{38} What an astonishing license, and what a pitiable ignorance of reasoning! For if anything that is uttered is neither true nor false, then surely it is not true; and what is not true, how can it fail to be false? Or what is not false, how can it fail to be true? So the position Chrysippus defends will hold—that every proposition is either true or false. Reason itself will force the admission both that certain things are true from all eternity, and that these are not bound up with eternal causes, and are free from the necessity of fate.

\ciceroSection{39} To me, at any rate, it seems that, since there had been two opinions among the older philosophers—one held by those who judged that all things come about by fate in such a way that this fate carries the force of necessity, an opinion in which Democritus, Heraclitus, Empedocles, and Aristotle stood; the other held by those to whom the voluntary motions of the mind seemed to be without any fate—Chrysippus, like an honorary arbiter, wished to strike a middle course, but attaches himself rather to those who would have the motions of the mind set free from necessity; yet while he uses his own terms, he slides into such difficulties that, against his will, he ends by confirming the necessity of fate.

\ciceroSection{40} And let us see, if you please, what this comes to in the case of assent, which I treated in my first discourse. For those ancient thinkers, to whom all things seemed to come about by fate, said that assent is brought about by force and by necessity. But those who dissented from them set assent free from fate, and denied that, once fate was applied to assent, necessity could be removed from it; and they argued in this way: If all things come about by fate, all things come about by an antecedent cause; and if appetite does, then the things that follow upon appetite do as well, and therefore so do acts of assent; but if the cause of appetite is not seated in us, then appetite itself is not in our power either; and if this is so, then neither are the things that are brought about by appetite seated in us; therefore neither acts of assent nor actions are in our power. From which it follows that neither praises are just, nor blames, nor honors, nor punishments. Since this is wrong, they think it can be plausibly concluded that not all things come about by fate, whatever comes about.

\ciceroSection{41} Chrysippus, however, since he both rejected necessity and yet would have nothing come about without preceding causes, distinguishes between kinds of causes, so as both to escape necessity and to keep fate. "For of causes," he says, "some are perfect and principal, others are auxiliary and proximate. For which reason, when we say that all things come about by fate through antecedent causes, we do not wish it understood thus: by perfect and principal causes, but by auxiliary and proximate antecedent causes." And so he meets the argument I concluded a little earlier in this way: that, if all things come about by fate, it follows indeed that all things come about by causes set before them, yet not by principal and perfect causes, but by auxiliary and proximate ones. And if these themselves are not in our power, it does not follow that appetite too is not in our power. But this would follow, if we said that all things come about by perfect and principal causes, so that, since those causes were not in our power, appetite would not be in our power either.

\ciceroSection{42} For which reason, against those who so introduce fate that they join necessity to it, that conclusion will hold; but against those who do not call the antecedent causes perfect or principal, it will not hold at all. For as to the saying that acts of assent come about by causes set before them, what this amounts to he thinks is easily explained by him. For although assent cannot come about unless it is stirred by an impression, nevertheless, since that impression has a proximate cause, not a principal one, it has this character (as Chrysippus would have it, as we said a moment ago): not that assent could come about without some force roused from outside—for it is necessary that assent be stirred by an impression—but he comes back to his cylinder and his spinning-top, which cannot begin to move unless they are pushed. But once that has happened, he holds that for the rest the cylinder rolls and the top spins by their own nature.

\ciceroSection{43} "As, therefore," he says, "the man who shoved the cylinder gave it the beginning of its motion, but did not give it its capacity to roll, so the impression presented will indeed press its appearance upon the mind and, as it were, stamp it there, but our assent will be in our power, and—just as was said in the case of the cylinder—once pushed from outside, for the rest it will move by its own force and nature." But if anything were brought about without an antecedent cause, it would be false that all things come about by fate; whereas if it is likely that a cause precedes all things, whatever come about, what can be brought forward to show why it must not be admitted that all things come about by fate? Provided only it is understood what the distinction and difference of causes is.

\ciceroSection{44} Since these things have been so explained by Chrysippus, then, if those who deny that acts of assent come about by fate nevertheless grant that they do not come about without an antecedent impression, the case is different; but if they concede that impressions go before, and yet that acts of assent do not come about by fate—because that proximate and contiguous cause does not move the assent—see whether they are not saying the same thing. For Chrysippus, conceding that the proximate and contiguous cause of assent is lodged in the impression, will not concede that that cause is necessary for assenting, so that, if all things come about by fate, all things come about by antecedent and necessary causes; and likewise those who dissent from him, while admitting that acts of assent do not come about without the prior running-on of impressions, will say that, if all things came about by fate in such a way that nothing came about except by the precedence of a cause, it would have to be admitted that all things come about by fate. From which it is easy to understand that, since both sides, once their opinion has been laid open and explained, arrive at the same outcome, they differ in words, not in substance.

\ciceroSection{45} And altogether, since this is the distinction—that in certain matters it can truly be said that, when these causes have gone before, it is not in our power that those things should not come about whose causes they were, but that in certain matters, though causes have gone before, it is nevertheless in our power that the thing should come about otherwise—this distinction both sides approve; but the one party judges that, in matters where, when the causes have preceded, it is not in our power that those things should come about otherwise, these come about by fate; whereas the things that are in our power, from these fate is absent.

\ciceroSection{46} In this way the question ought to be argued, and not by seeking aid from atoms that wander and swerve from their path. "The atom swerves," he says. First, why? For they had derived a certain other force of motion from Democritus—the force of impulsion, which he calls a blow—and from you, Epicurus, the force of gravity and weight. What new cause is there in nature, then, to make the atom swerve? Or do the atoms cast lots among themselves, which is to swerve and which is not? Or why do they swerve by the least interval and not by a greater? Or why do they swerve by one least interval, and not by two or three?

\ciceroSection{47} This is to wish, not to argue. For you do not say that the atom is moved from its place and swerves by an impulse from outside, nor that in that void through which the atom is carried there was anything to cause it not to be carried straight on, nor that any change was made in the atom itself, on account of which it should not keep the natural motion of its own weight. And so, though he had brought forward no cause to produce that swerve, he yet thinks he is saying something, when he says what the minds of all men spurn and reject.

\ciceroSection{48} Indeed, no one seems to me to confirm more strongly not only fate but even necessity and the compulsion of all things, and to have done away with the voluntary motions of the mind, than this man, who admits he could not have withstood fate in any other way than by taking refuge in these fictitious swerves. For granting that there were atoms—which I can in no way be brought to believe exist—still those swerves would never be explained. For if it is granted to atoms that they are carried by gravity through a necessity of nature, since all weight, with nothing hindering, must of necessity be moved and carried, then this too is of necessity: that they swerve, certain atoms—or, if they like, all of them—by nature . . .

