% De Divinatione (de-divinatione) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroBook{1}

\ciceroSection{1} There is an ancient belief, drawn down all the way from heroic times and confirmed by the agreement both of the Roman people and of all nations, that a certain divination exists among men, which the Greeks call \textit{[Greek: mantikēn]} — that is, the foreknowledge and knowledge of things to come. A magnificent thing it is, and salutary, if only it exists at all, and one by which mortal nature can come nearest to the power of the gods. And so, just as in many other things we have done better than the Greeks, so for this most excellent thing our people drew its name from the gods (\textit{divi}), while the Greeks, as Plato interprets it, drew theirs from frenzy.

\ciceroSection{2} Indeed I see no nation, whether so civilized and learned or so savage and barbarous, that does not hold that things to come are signified and can be understood and foretold by certain men. To begin with the Assyrians — that I may fetch my authority from the most distant peoples — because of the flatness and vastness of the regions they inhabited, since they looked out upon a sky open and exposed on every side, they kept watch over the courses and motions of the stars, and once these were marked, they handed down to memory what was signified to each man. Among that people the Chaldeans — named not from their art but from their nation — are thought to have achieved, by long observation of the constellations, a knowledge whereby it could be foretold what would befall each man and under what fate each was born. The Egyptians, too, are thought to have attained the same art through length of time, over almost numberless ages. The people of the Cilicians and the Pisidians, moreover, and Pamphylia, their neighbor — nations over which I myself held command — believe that things to come are declared by the flights and cries of birds as by the surest of signs.

\ciceroSection{3} And what colony did Greece ever send into Aeolia, Ionia, Asia, Sicily, or Italy without the oracle of the Pythian, or of Dodona, or of Hammon? Or what war was undertaken by her without the counsel of the gods? Nor is it one kind of divination only that has been practiced, both in public and in private. For — to pass over the other peoples — how many kinds has our own embraced! At the very beginning, Romulus, the father of this city, is recorded not only to have founded the city under auspices, but himself to have been a most excellent augur. Then the rest of the kings, too, made use of augurs, and after the kings were driven out nothing was done in public, at home or in the field, without auspices. And since great power seemed to lie, both in seeking and consulting things and in interpreting and averting monstrous signs, in the discipline of the haruspices, they brought all this science from Etruria, that there might be no kind of divination at all which seemed to be neglected by them.

\ciceroSection{4} And since the mind is roused in two ways, without reason and knowledge, by its own free and unfettered motion — the one in frenzy, the other in dreaming — they judged that the divination of frenzy was contained chiefly in the Sibylline verses, and they resolved that ten interpreters of these should be chosen from the state. Of this same kind they often thought that even the frenzied predictions of fortune-tellers and seers should be heeded, as in the Octavian war those of Cornelius Culleolus. Nor indeed were the weightier dreams, if any seemed to bear upon the commonwealth, neglected by the highest council. Why, even within our own memory, Lucius Julius, who was consul with Publius Rutilius, restored the temple of Juno Sospita by decree of the Senate, on the strength of a dream of Caecilia, daughter of Balliaricus.

\ciceroSection{5} And these things, as I judge, the ancients approved, prompted rather by the outcomes of events than taught by reasoning. But certain refined arguments of the philosophers were gathered to show why divination should be true; and of these — to speak of the most ancient — Xenophanes of Colophon alone, while saying that the gods exist, abolished divination root and branch; all the rest, save Epicurus stammering about the nature of the gods, approved divination, though not in one and the same way. For while Socrates and all the Socratics, and Zeno and those who set out from him, abided in the judgment of the ancient philosophers, with the Old Academy and the Peripatetics agreeing — and while Pythagoras had already long before assigned great authority to this thing, he who even wished to be an augur himself, and while Democritus, a weighty authority in very many passages, gave his approval to the foreknowledge of things to come — Dicaearchus the Peripatetic abolished the other kinds of divination, but left those of dreams and of frenzy; and Cratippus, our friend, whom I judge the equal of the foremost Peripatetics, gave credit to these same things, but rejected the remaining kinds of divination.

\ciceroSection{6} But while the Stoics defended almost all of those kinds — because Zeno in his commentaries had scattered, as it were, certain seeds, and Cleanthes had made them somewhat richer — there came a man of the keenest genius, Chrysippus, who unfolded the whole doctrine of divination in two books, and besides one on oracles and one on dreams; and following him, Diogenes the Babylonian, his pupil, brought out one book, Antipater two, and our own Posidonius five. But from the Stoics there fell away one who was almost the chief of that school, Posidonius's teacher and Antipater's pupil — Panaetius; and yet he did not dare to deny that there is a power of divining, but said only that he was in doubt. What he, a Stoic, was allowed to do on one point, much against the Stoics' will — shall we not be granted by the Stoics leave to do the same on the rest? — especially since that of which Panaetius is not certain is seen by the others of the same school more clearly than the light of day.

\ciceroSection{7} But this credit, at any rate, has been confirmed for the Academy by the judgment and testimony of a most excellent philosopher. For when we ourselves were inquiring what should be judged about divination — since Carneades argued many things shrewdly and copiously against the Stoics, and since we were afraid of assenting rashly to a thing either false or not sufficiently known — it seems we ought to compare arguments with arguments carefully, again and again, as we did in those three books which we wrote on the nature of the gods. For while in all things rashness in assenting, and error, is shameful, it is so most of all in that matter where the question to be judged is how much we should grant to auspices, to things divine, and to religion; for there is danger lest, by neglecting them, we be bound by impious wickedness, or, by taking them up, by an old woman's superstition.

\ciceroSection{8} On these matters there was discussion both often at other times and a little more closely of late, when I was with my brother Quintus at the Tusculan villa. For when we had come into the Lyceum for the sake of a walk (for that is the name of the upper gymnasium): "I have read through," said he, "your third book on the nature of the gods a little while ago, in which the disputation of Cotta, though it shook my own opinion, did not nevertheless overthrow it utterly." Excellent, said I; for Cotta himself disputes in such a way as rather to confute the arguments of the Stoics than to destroy men's religion. Then said Quintus: "That is indeed what is said of Cotta — and said too often, I believe, lest he seem to transgress the common laws; but in his zeal for arguing against the Stoics he seems to me to do away with the gods root and branch.

\ciceroSection{9} To his reasoning I do not greatly long to make any reply; for religion has been sufficiently defended in the second book by Lucilius, whose disputation seemed to you yourself, as you write at the end of the third, more inclined toward the truth. But that which was passed over in those books (I suppose because you judged it more convenient that it should be inquired into separately and discussed by itself) — namely, divination, which is the prediction and foreknowledge of those things that are reckoned fortuitous — that, if you please, let us examine, to see what force it has and of what kind it is. For I think thus: if those kinds of divining are true of which we have received report and which we cultivate, then the gods exist; and conversely, if the gods exist, then there are those who divine."

\ciceroSection{10} The very citadel of the Stoics, said I, Quintus, is what you defend, since indeed those things are reciprocal in such a way that, if there is divination, the gods exist, and if the gods exist, there is divination. Yet neither of these is granted so easily as you suppose. For things to come can be signified by nature without a god; and, granting that the gods exist, it may happen that no divination has been bestowed by them on the human race. And he replied: For my part, I have argument enough both that the gods exist and that they take thought for human affairs, in that I judge there to be clear and manifest kinds of divination. About these, if you please, I will set forth what I myself think — yet only if you are free in mind and have nothing that you think should take precedence over this conversation.

\ciceroSection{11} For my part, said I, Quintus, I am always free for philosophy; and at this time, since there is nothing else that I could do with pleasure, I am all the more eager to hear what you think about divination. I have nothing new, indeed, said he, nor anything that I myself think apart from the rest; for I follow the most ancient opinion, and one approved by the agreement of all peoples and nations. There are, then, two kinds of divining: one of which belongs to art, the other to nature.

\ciceroSection{12} For what nation is there, or what state, that is not moved by the prediction of those who interpret entrails, or monstrous signs, or lightning flashes, or of augurs, or of astrologers, or of lots (for these belong, roughly, to art), or of dreams or of prophesyings (for these two are reckoned natural)? And of these matters I think it more fitting that the outcomes be inquired into than the causes. For there is a certain force and nature which, now by significations observed over a long time, now by some divine instinct and inbreathing, foretells things to come. Wherefore let Carneades cease to press the question, which Panaetius too kept asking, whether it was Jupiter who had bidden the crow to caw on the left and the raven on the right. These things have been observed over an immense span of time and, in their signification, noted and marked by their outcomes. There is nothing, however, that length of time, with memory to receive it and records to hand it down, cannot accomplish and attain.

\ciceroSection{13} We may well marvel at the kinds of herbs the physicians have observed, and the kinds of roots good against the bites of beasts, against diseases of the eyes, against wounds — things whose force and nature reason has never explained, yet which by their usefulness have proved both the art itself and its discoverer. Come, let us look at things which, though they belong to another class, are nevertheless more like divination:

And often, too, the swollen sea forewarns of winds to come, \\
when of a sudden, from its very depths, it heaves, \\
and the white rocks, foaming with snowy brine, \\
strive to give back to Neptune their grief-bringing voices, \\
or when, risen from the lofty mountain-peak, \\
a dense shrill whistling swells, flung back and back from the crags.

Of such presentiments of things your \textit{Prognostics} are full. Who, then, can draw out the causes of these presentiments? — although I see that Boethus the Stoic attempted it, and got so far as to explain the principle of those things that come to pass in the sea or the sky.

\ciceroSection{14} But why these others happen, who could say with any probability?

The grey coot likewise, fleeing from the deep of the sea, \\
gives warning, with its cry, that horrible storms are near, \\
pouring no slight songs from its quivering throat. \\
Often, too, from its breast it sings a most mournful strain, \\
and the marsh-bird presses on with its morning calls, \\
presses on with its calls and casts ceaseless complaints from its mouth, \\
when first the dawn sends back the chill dews. \\
And the dusky crow, now and then, coursing along the shores, \\
plunged its head down and took the wave upon its neck.

\ciceroSection{15} We see that these signs almost never lie, yet we do not see why it should so happen.

You, too, behold the signs, you nurslings of the fresh water, \\
when with a clamor you make ready to pour out your empty cries \\
and rouse the springs and standing pools with absurd sound.

Who could suspect that the little frogs see this? Yet there is in the little frogs a force and a certain nature signifying something, plain enough in itself, but rather too obscure for the understanding of men.

And the soft-footed oxen, gazing up at the lights of the sky, \\
drew in the moisture-laden juice from the air through their nostrils.

I do not ask why, since I understand what comes to pass. Now too, ever green and ever weighed down,

The mastic, wont to swell with a triple yield, \\
three times pouring forth its fruit, marks the three seasons of plowing.

Not even this do I ask — why this one tree flowers three times, or why it fits the ripeness for plowing to the sign of its flowering;

\ciceroSection{16} I am content with this, that, even if I am ignorant why each thing comes about, I understand what comes about. For all divination, then, I shall give the same answer that I gave for the things I have just recalled. What the root of scammony can do for purging, what birthwort can do against the bites of serpents — birthwort, which got its name from its discoverer, while the discoverer got the thing itself from a dream — this I see, which is enough; why it can, I do not know. So too the signs of winds and rains that I have spoken of: what account they have I do not adequately perceive; their force and their outcome I recognize, I know, I approve. In the same way, what the cleft in the entrails, what the fiber portends, I take in; what the cause is, I do not know. And of these things life is full, for nearly everyone uses the entrails. Again: can we doubt the force of lightning-bolts? Among many other marvels there is this above all: when Summanus — whose statue, set on the gable of Jupiter Best and Greatest, was at that time of terracotta — had been struck from the sky, and the head of that image could nowhere be found, the haruspices declared it had been hurled down into the Tiber, and it was found in the very spot the haruspices had pointed out.

\ciceroSection{17} But what authority, what witness should I use rather than yourself? — whose verses I have even learned by heart, and gladly so, the ones the Muse Urania pronounces in the second book of your poem on your consulship:

In the beginning Jupiter, kindled with ethereal fire, \\
turns, and with his light surveys the whole world, \\
and with his divine mind reaches toward heaven and earth, \\
which deeply holds within it the senses and the lives of men, \\
hedged about and enclosed in the caverns of the eternal ether. \\
And if you would know the motions of the stars and the courses of the wanderers, \\
which are set in the seats of the constellations \\
— things named, in word and in the false speech of the Greeks, falsely, \\
but in very truth borne on by a fixed gliding and an appointed span — \\
all things you will now discern, marked out by the divine mind.

\ciceroSection{18} For first the swift motions of the stars, in your consulship,

and the heavy gatherings of the stars flashing with fire — \\
you too, when on the Alban mount you purified the snowy hills \\
and with glad milk did sacrifice at the Latin Festival, \\
saw the comets quivering with a bright blaze, \\
and judged that much was being stirred up by nocturnal carnage, \\
since the Latin Festival fell, as it chanced, upon a dread hour, \\
when the moon hid her bright face with congealed light \\
and was of a sudden blotted out in the starry night. \\
And what of the torch of Phoebus, grim herald of war, \\
which flew toward the great summit with a flaming blaze, \\
making for the regions of the headlong sky and its setting? \\
Or when a citizen, struck down by a terrible thunderbolt, \\
gave up his vital light beneath a clearing sky? \\
Or when the earth shook itself with its teeming body? \\
And now, by night, terrible shapes were seen in various forms, \\
warning of war and of upheavals, \\
and across the lands many seers, with frenzied breast, \\
poured forth oracles that threatened grievous disasters,

\ciceroSection{19} and those things which, by the lapse of an age, came at last to pass —

these the very Father of the gods, repeating them \\
with constant and clear signs, sang out from heaven and earth. \\
Now those things which, once, in the consulship of Torquatus and Cotta, \\
the Lydian haruspex of the Tyrrhenian race had declared — \\
all those, gathering them together, your year fixes and determines. \\
For the high-thundering Father, leaning upon starry Olympus, \\
himself once made for his own hills and temples \\
and cast his fires upon the seats of the Capitol. \\
Then the ancient bronze image, revered, of Natta \\
fell, and the laws, by their ancient sanctity, slipped away, \\
and the fire of the thunderbolt destroyed the images of the gods.

\ciceroSection{20} Here was the woodland nurse of the Roman name,

the she-wolf of Mars, who with her teeming udders \\
bedewed with the dew of life the little ones born of the seed of Mavors; \\
she at that time, with the boys, by the flaming stroke of the thunderbolt \\
fell, and left behind the torn-off prints of her feet. \\
Then who, turning over the writings and records of the art, \\
did not draw grief-bringing voices from the Etruscan scrolls? \\
All of them gave warning to beware of a vast calamity and a plague \\
sprung from a noble stock, \\
or with steadfast voice foretold the ruin of the laws, \\
and bade men snatch from the flames the temples and the gods themselves \\
and the city, and dread a horrible slaughter and carnage; \\
and that these things were held fixed by a grievous fate and firmly founded, \\
unless first, fashioned with comeliness, set on a lofty height, \\
the holy image of Jupiter should look toward the bright sunrise. \\
Then it would come to pass that the people and the holy Senate \\
could discern the hidden designs, if, turned thence toward the rising of the sun, \\
the seat of the fathers and of the people should look out.

\ciceroSection{21} This image, long delayed and much postponed,

was at last, in your consulship, set on its lofty seat; \\
and in the very hour of the fixed and appointed time \\
Jupiter on his lofty column made bright his scepter, \\
and the disaster prepared for the fatherland by flame and sword \\
was laid open, by the words of the Allobroges, to the fathers and the people. \\
Rightly, then, did the men of old, whose monuments you keep, \\
who ruled peoples and cities with measure and with valor — \\
rightly too did your own people, whose piety and faith \\
stood foremost, and whose wisdom far surpassed all others — \\
above all, with thriving worship, revere the gods. \\
These things, indeed, they saw deeply, with sagacious care, \\
who, glad, gave their leisure to noble studies,

\ciceroSection{22} and in the shady Academy and the gleaming Lyceum

poured forth the bright arts of a fruitful breast. \\
And you, snatched away from these in the first flower of youth, \\
your fatherland set in the very midst of the mass of virtues. \\
Yet you, relaxing your anxiety-bringing cares with rest, \\
have consecrated to your studies and to me whatever time your fatherland leaves free.

You, then — will you be able to bring your mind to speak against what I argue concerning divination, you who both performed the deeds you performed and wrote out, with the greatest exactness, the very things I have pronounced?

\ciceroSection{23} What? Do you ask, Carneades, why these things happen so, or by what art they can be discerned? I confess I do not know; but that they do happen, I say you yourself see. By chance, you say. Is that so, indeed? Can anything be brought about by chance that has in itself all the marks of truth? Four dice, when thrown, may make the Venus-throw by chance; but do you suppose that, if you threw four hundred dice, even a hundred Venus-throws would come about by chance? Pigments flung at random on a panel can produce the lineaments of a face; but do you suppose that the beauty of the Venus of Cos could be produced by a chance splashing? If a sow should press the letter A in the dirt with its snout, would you on that account be able to suspect that it could write out Ennius's \textit{Andromache}? Carneades used to imagine that, in the quarries of Chios, when a stone was split, the head of a little Pan appeared; I grant some figure not unlike it, but certainly not such a one that you would call it made by Scopas. For surely the matter stands thus, that chance never perfectly imitates the truth.

\ciceroSection{24} But sometimes the things that have been predicted fail to come to pass. And which art, pray, does not have this? I mean those arts which rest on conjecture and are matters of opinion. Or is medicine not to be reckoned an art? Yet how many things deceive it! What? Are pilots not deceived? Or did the army of the Achaeans and the masters of so many ships set sail from Ilium in such a way — joyful at their departure, gazing at the wantonness of the fish, as Pacuvius puts it, that no surfeit of looking could come over them?

Meanwhile, with the sun now nearly setting, the sea grows rough; \\
the darkness is redoubled, and the blackness of night and storm-cloud blinds all.

Did the shipwreck, then, of so many most illustrious leaders and kings abolish the art of steering? Or is the knowledge of generals nothing, because a supreme general lately fled with his army lost? Or is there, on that account, no method and prudence in governing the commonwealth, because many things deceived Gnaeus Pompey, certain things Marcus Cato, and some even yourself? The answer of the haruspices is of like kind, and so is all divination that rests on opinion; for it leans on conjecture, beyond which it cannot advance.

\ciceroSection{25} This perhaps deceives now and then, but for all that it most often points the way to truth; for it has been traced back from all eternity, and since within that span things came to pass innumerable times in the same manner, with the same signs going before, an art was produced by noticing and noting the same things again and again. And how reliable are your auspices! Yet these are now ignored by the Roman augurs (let me say this with your good leave), while they are still kept up by the Cilicians, the Pamphylians, the Pisidians, and the Lycians.

\ciceroSection{26} For why should I bring up our own guest-friend, that most illustrious and excellent man, King Deiotarus? He never does anything except after taking the auspices. Once, when he had set out on a certain journey already planned and fixed upon, he turned back, warned by the flight of an eagle; and the chamber in which he would have lodged, had he gone on, collapsed the very next night.

\ciceroSection{27} And so, as I used to hear from his own lips, he very often turned back from a journey even when he had already advanced many days' travel along the road. The most splendid thing about him is this: that after Caesar had stripped him of his tetrarchy, his kingdom, and his money, he still says he does not regret the auspices that fell out favorably for him as he was setting out to join Pompey; for, he says, the authority of the Senate and the liberty of the Roman people and the dignity of its empire had been defended by his arms, and the birds, on whose prompting he had followed duty and good faith, had counseled him well; for glory had stood higher with him than his own possessions. He, then, seems to me to take the auguries truly. For our own magistrates use auspices that are forced; since a morsel thrown down must necessarily make a scrap fall from the chick's mouth when it feeds;

\ciceroSection{28} but as for what you have set down in writing — that a tripudium occurs if any of it falls to the ground in a solid piece — this too, which I have mentioned, the forced one, you call the most favorable tripudium. And so many auguries, many auspices, as that wise Cato laments, have been quite lost and abandoned through the negligence of the college. Once almost nothing of greater importance was done, even in private life, without taking the auspices; and the wedding auspices still proclaim this even now, though, the practice having lapsed, they keep only the name. For as we now use the entrails for great undertakings (although even this somewhat less than of old), so then great matters were commonly carried through by the birds. And so, while we do not seek out the sinister signs, we run upon dire and faulty ones.

\ciceroSection{29} So Publius Claudius, the son of Appius Caecus, and his colleague Lucius Junius lost the greatest of fleets, having put to sea against the auspices. The same thing befell Agamemnon; for when the Achaeans had begun \\
to mutter among themselves \\
and openly to disparage the art of the readers of entrails, \\
he commands the ships unmoored, \\
with the rumor for him but the bird against. \\
But why these ancient cases? We see what happened to Marcus Crassus, when the announcement of dire omens was disregarded. In this affair Appius, your colleague, a good augur — as I am accustomed to hear from you — did not very knowingly, as censor, mark down Gaius Ateius, a good man and an outstanding citizen, on the charge that he had falsely reported auspices. So be it; this might have been the censor's part, if he judged the man had lied; but it was least of all the augur's part to add as the reason that the Roman people had thereby met with its greatest disaster. For if that was the cause of the disaster, the fault lies not with the man who reported the omen, but with the man who did not obey it. For that the report was true, as that same augur and censor says, the outcome proved; and had it been false, it could have brought no cause of disaster. For dire omens, like the other auspices, like omens, like signs, do not bring about the causes why anything happens, but announce things that will come to pass, unless you take heed.

\ciceroSection{30} The report of Ateius, then, did not fabricate the cause of the disaster, but by setting a sign before him warned Crassus what would come to pass, unless he took care. So either that report had no force, or, if it had force, as Appius judges, this much force it had: that the fault clings not to the man who warned, but to the man who did not heed. And again: that augur's staff of yours, the most illustrious badge of the augurate — from where was it handed down to you? Surely it was with it that Romulus marked off the regions of the sky at the time he founded the city. And that staff of Romulus — that is, a curved rod, slightly bent at the top, which got its name from its likeness to the trumpet by which the signal is sounded — though it had been placed in the chapel of the Salii, which is on the Palatine, and that chapel had burned down, was found intact.

\ciceroSection{31} And again: many years after Romulus, in the reign of Tarquinius Priscus, what writer of the ancients does not tell of the division of the regions made by Attus Navius by means of the augur's staff? When as a boy, on account of his poverty, he was tending swine and lost one of them, he is said to have vowed that, if he recovered it, he would give the god the largest grape there was in the vineyard; and so, when the sow was found, facing south he is said to have taken his stand in the middle of the vineyard, and when he had divided the vineyard into four parts, and the birds had ruled out three parts, in the fourth part, which was left over, distributed into regions, he found, as we see it written, a grape of marvelous size. When this had been noised abroad, since all the neighbors brought their affairs to him, he was held in great name and glory.

\ciceroSection{32} From this it came about that King Priscus summoned him to his court. And when he was testing the man's skill in augury, he said he was thinking of something, and asked whether it could come to pass. Navius, having performed the augury, replied that it could. Then Tarquinius said he had been thinking that a whetstone could be cut through with a razor. Then he bade Attus try. So the whetstone, brought into the Comitium with both the king and the people looking on, was cleft with a razor. From that day forward Tarquinius made use of Attus Navius as his augur, and the people brought their own affairs to him.

\ciceroSection{33} That whetstone and razor, we are told, were buried in the Comitium and a well-curb set over them. Let us deny everything, let us burn the annals, let us call all this invention — anything, in short, rather than confess that the gods care for human affairs. And again: what is written in your own work about Tiberius Gracchus — does it not confirm the discipline of both the augurs and the haruspices? When, unawares, he had pitched his tent under a flaw, because he had crossed the pomerium without taking the auspices, he held the assembly for the election of consuls. The matter is well known and has been committed to the records by you yourself. But Tiberius Gracchus, himself an augur, confirmed the authority of the auspices by his confession of his own error; and great authority accrued to the discipline of the haruspices, who, brought into the Senate just after the assembly, declared that the man who put the election to the vote had not been a lawful one.

\ciceroSection{34} I agree, then, with those who have said that there are two kinds of divination: one that shares in art, the other that lacks art. For there is art in those who pursue new things by conjecture and have learned old things by observation. But those lack art who foreknow things to come not by reasoning or conjecture from signs observed and noted, but by a certain rousing of the mind or a loosed and free motion of it — which often befalls dreamers, and now and then those who prophesy through frenzy, like Bacis the Boeotian, like Epimenides the Cretan, like the Erythraean Sibyl. Of this kind, too, oracles are to be counted — not those that are drawn by lots evenly shuffled, but those that are poured forth by a divine impulse and inspiration; though the lot itself is not to be despised, if it has both the authority of antiquity, as those lots have which we are told came forth from the earth — and yet that, when drawn, they should fall out aptly to the matter, I believe can come about by the gods' working. The interpreters of all these, like the grammarians who interpret the poets, seem to come nearest to the divination of those whom they interpret.

\ciceroSection{35} What sort of cleverness is this, then, that wishes by false accusation to overturn things made robust by their long age? I cannot find the reason. Perhaps it lies hidden, wrapped in the obscurity of nature; for god did not will that I should know these things, but only that I should make use of them. I will make use of them, then, and will not be led to believe either that the whole of Etruria raves over the entrails, or that the same nation goes astray over lightning-flashes, or interprets portents deceitfully — when the rumblings of the earth, often its bellowings, often its tremors, have foretold many grave and true things to our own commonwealth and many to the rest of the states.

\ciceroSection{36} And again: that birth of a mule which is held up to ridicule — was it not, because the offspring appeared where nature was barren, foretold by the haruspices to be an incredible begetting of woes? And again: Tiberius Gracchus, son of Publius, who was twice consul and censor, and likewise a supreme augur and a wise man and an outstanding citizen — did he not, as Gaius Gracchus, his son, has left in writing, when two snakes had been caught in his house, summon the haruspices? When they had answered that if he let the male go, his wife must die within a short time, but if the female, then he himself, he judged it more fitting that he should meet a death already ripe than that the young daughter of Publius Africanus should; he let the female go, and a few days later he himself died. Let us mock the haruspices, let us call them empty and worthless, and let us despise that discipline which both a most wise man and the outcome and the events have confirmed; let us condemn even Babylon and those who, watching the signs of heaven from the Caucasus, track the courses of the stars by numbers and measures; let us condemn, I say, these men of either folly or vanity or impudence, who keep, as they themselves say, four hundred and seventy thousand years recorded in their monuments; and let us judge them liars, with no dread of the judgment of all the ages to come concerning themselves.

\ciceroSection{37} Come now — the barbarians are vain and deceitful; but has the history of the Greeks too lied? What the Pythian Apollo answered to Croesus (to speak of natural divination), what he answered to the Athenians, to the Lacedaemonians, to the Tegeans, to the Argives, to the Corinthians — who does not know it? Chrysippus collected oracles past numbering, and not one without a substantial authority and witness; but since these are known to you, I leave them aside. This one thing I maintain: that famous oracle at Delphi would never have been so frequented and so renowned, nor so heaped with gifts from all peoples and kings, had not every age made trial of the truth of those oracles.

\ciceroSection{38} For a long time now it has done no such thing. As, then, it stands today in lesser glory, because the truth of its oracles is less conspicuous, so back then it would not have stood in such great glory had it not been supreme in truth. And that power of the earth, which used to rouse the mind of the Pythia by a divine exhalation, may have faded away with age, just as we see that certain rivers have faded and dried up, or have been twisted and turned aside into another course. But let it have happened, if you will — for it is a great question; let only this stand fast, which cannot be denied unless we overturn the whole of history: that for many ages that oracle was truthful.

\ciceroSection{39} But let us leave the oracles; let us come to dreams. Discoursing on these, Chrysippus does the very same thing in heaping up many trifling dreams that Antipater does in gathering up those which, explained by Antiphon's interpretation, display indeed the interpreter's acuteness, but it would have been fitting to use grander examples. The mother of Dionysius — the one who was tyrant of the Syracusans — when she was pregnant and carried this very Dionysius in her womb, dreamed that she had given birth to a little satyr; this is recorded in Philistus, a learned and careful man and a contemporary of those times. To her the interpreters of portents, who were then called the Galeotae in Sicily, answered, as Philistus reports, that the one she had borne would be most illustrious in Greece and would enjoy long-lasting good fortune.

\ciceroSection{40} Do I call you back to the fables, whether of our own poets or of the Greeks? For that Vestal in Ennius too relates such things:

When the old woman, summoned, brought a light with trembling limbs, \\
then weeping, terrified out of sleep, she tells this tale: \\
``Daughter of Eurydica, whom our father loved, \\
strength and life now forsake my whole body. \\
For a handsome man seemed to seize me away through pleasant willow-groves \\
and along the banks and into places strange; and so afterward, \\
alone, my own dear sister, I seemed to wander \\
and slowly to track and seek you, and could not \\
lay hold of you in my heart; no path made my foot steady.

\ciceroSection{41} Then my father seems to call me with his voice \\
in these words: ``O daughter, hardships must first \\
be borne by you; afterward your fortune will rise again from a river.'' \\
Having spoken these things, my father, sister, suddenly withdrew \\
and gave not himself to my sight, longed for in my heart, \\
though again and again I stretched my hands, weeping, toward the blue \\
temples of heaven and called him with coaxing voice. \\
Then scarcely did sleep release me from my aching heart.''

\ciceroSection{42} These things, even if they are invented by the poet, are still not far from the way of dreams. Let this too be a fiction, by which Priam was thrown into confusion — that a pregnant mother, Hecuba, seemed in her sleep to bear from herself a blazing torch; at which deed the father, King Priam himself, struck in his mind with fear by the dream, sunk in cares and sighing griefs, made sacrifice with bleating victims. Then, seeking peace, he calls for a conjecture, beseeching Apollo to teach him whither such great fates of his dreams might turn. There, from the oracle, Apollo by a divine voice declared that the boy who should be born to Priam next afterward — let him forbear to rear him; that boy was the ruin of Troy, the plague of Pergamum.

\ciceroSection{43} Let these be, as I said, the dreams of fables; and let there be joined to them the dream of Aeneas too, which in the Greek annals of our Fabius Pictor is of such a kind that all the things done by Aeneas and all that befell him were the very things that had appeared to him in his sleep. But let us look at matters nearer home. Of what kind is the dream of Tarquin the Proud, of which he himself speaks in the \textit{Brutus} of Accius?

\ciceroSection{44} Since I gave my body to rest in the night's onset, \\
soothing my weary limbs with slumber, \\
in my sleep there seemed a shepherd to drive toward me \\
a woolly flock of surpassing beauty; \\
two rams of one blood were chosen from it, \\
and I sacrificed the more splendid of the two; \\
then his brother butted with his horns, \\
charged at me, and by that blow I was thrown down; \\
then, stretched out upon the earth, gravely wounded, \\
lying on my back, I beheld in the sky a great \\
and wondrous deed: the flaming, rayed orb \\
of the sun rolling away to the right on a strange new course.

Let us see, then, what interpretation of this dream was made by the interpreters:

\ciceroSection{45} King, the things that men handle in life, that they think, attend to, see, \\
and the things they do and busy themselves with awake — when these befall a man in sleep, \\
it is less to be wondered at; but the gods do not offer so great a thing rashly and unforeseen. \\
So beware lest the very man whom you reckon to be as dull as a beast \\
should bear in his breast a heart fortified with surpassing wisdom, \\
and drive you from your kingdom; for that which was shown to you concerning the sun \\
portends that a change of affairs for the people will be \\
very near at hand. May these things turn out well for the people. For in that the most mighty sign \\
took its course from the left to the right, it has been most beautifully \\
augured that the Roman commonwealth shall be supreme.

Come now, let us return to foreign matters.

\ciceroSection{46} Heraclides of Pontus, a learned man, a hearer and disciple of Plato, writes that the mother of Phalaris seemed to see in her sleep the images of the gods which she herself had consecrated at home; of these, Mercury seemed to pour blood from the bowl which he held in his right hand, and when it had touched the earth, it seemed to seethe up so that the whole house overflowed with blood. This dream of the mother the monstrous cruelty of the son confirmed. And why should I bring forward, from the \textit{Persica} of Dinon, what the magi interpreted for that first prince Cyrus? For when, as he slept, the sun seemed to be at his feet, three times — he writes — Cyrus tried in vain to grasp it with his hands, while the sun, rolling away, slipped from him and departed; and the magi told him — a class held among the Persians to be wise and learned — that by the threefold attempt at the sun it was portended that Cyrus would reign thirty years. And so it turned out; for he reached his seventieth year, having begun to reign at the age of forty.

\ciceroSection{47} There is surely something even among barbarian peoples that has presentiment and divines, seeing that Callanus the Indian, as he set out toward death, when he was mounting the burning pyre, said: ``O glorious departure from life, when, as it befell Hercules, with the mortal body burned the spirit shall pass forth into the light!'' And when Alexander asked him to say if he wished anything, he said: ``Excellent; before long I shall see you.'' And so it turned out; for at Babylon, a few days later, Alexander died. I depart from dreams for a little, to which I shall soon return. On the same night on which the temple of Diana at Ephesus burned down, it is agreed that Alexander was born of Olympias; and when day had begun to break, the magi cried aloud that a plague and a ruin of Asia had been born the night before. So much concerning the Indians and the magi.

\ciceroSection{48} Let us return to dreams. Coelius writes that when Hannibal wished to carry off a golden column which stood in the shrine of Juno Lacinia, and was in doubt whether it was solid or gilded on the outside, he bored through it, and when he found it solid, resolved to take it away; but in his sleep Juno seemed to warn him not to do it, and to threaten that, if he did, she would see to it that he should lose the one eye too with which he saw well; and this was not disregarded by the keen-witted man; and so out of that gold which had been bored out he had a heifer made, and set it on the top of the column.

\ciceroSection{49} There is, in the same way, a Greek account in Silenus, whom Coelius follows (and he traced the affairs of Hannibal most carefully): that Hannibal, when he had taken Saguntum, seemed in his sleep to be summoned by Jupiter into the council of the gods; and when he had come there, that Jupiter ordered him to carry war into Italy, and gave him one of the council as a guide, whom he took and, using him, began to advance with his army; then that this guide warned him not to look back; but that he could not refrain from doing so for very long, and, carried away by desire, looked back; that then there appeared a vast and monstrous beast, entwined with serpents, overturning every orchard, thicket, and dwelling wherever it moved; and that he, amazed, asked the god what so great a monstrous sign might be; and that the god answered it was the devastation of Italy, and ordered him to press straight on and not trouble himself about what was happening behind him and at his back.

\ciceroSection{50} In Agathocles it is set down in his history that Hamilcar the Carthaginian, while he was besieging Syracuse, seemed to hear a voice telling him that the next day he would dine in Syracuse; but that when that day dawned, a great mutiny arose in his camp between the Carthaginian and the Sicilian soldiers; and that when the Syracusans perceived this, they burst into the camp unexpectedly, and Hamilcar was carried off by them alive. So the event confirmed the dream. History is full of such examples, and so too is common life crowded with them.

\ciceroSection{51} But indeed that famous Publius Decius, son of Quintus, who was the first of the Decii to be consul, when he was a military tribune in the consulship of Marcus Valerius and Aulus Cornelius, and our army was being hard pressed by the Samnites — when he kept entering into the perils of battle too boldly and was warned to be more cautious — said (as stands recorded in the annals) that he had seen in a dream that, when he was in the very midst of the enemy, he died with the greatest glory. And at that time, unharmed, he freed the army from the siege; but three years afterward, when he was consul, he devoted himself and rushed armed into the line of the Latins. By this deed of his the Latins were overcome and destroyed. And his death was so glorious that his son craved the very same.

\ciceroSection{52} But let us come now, if it please you, to the dreams of philosophers. There is in Plato a Socrates who, when he was in the public prison, said to Crito, his close friend, that he must die after the third day; for he had seen in his sleep a woman of surpassing beauty, who, calling him by name, spoke a certain Homeric verse of this sort: \\
On the third day's dawn fair Phthia shall receive you. \\
And, as it was said, so it is written to have come to pass. Xenophon the Socratic — what a man, and how great! — in that campaign which he carried through with the younger Cyrus, writes of his own dreams, whose outcomes proved marvelous.

\ciceroSection{53} Shall we say that Xenophon lies, or that he raves? What of this? Does Aristotle, a man of singular and almost divine genius, himself err, or does he wish others to err, when he writes that Eudemus of Cyprus, his close friend, while making a journey into Macedonia, came to Pherae, which was then a very famous city in Thessaly but was held by the tyrant Alexander under a cruel despotism? In that town, then, Eudemus fell so gravely ill that all the physicians despaired of him; and there seemed to appear to him in his rest a young man of remarkable beauty, who told him that he would very soon recover, that within a few days the tyrant Alexander would perish, and that Eudemus himself would return home five years afterward. And the first of these, Aristotle writes, followed at once: Eudemus recovered, and the tyrant was slain by his wife's brothers; but as the fifth year was drawing to its close, when there was hope, from that dream, that he would return from Sicily to Cyprus, he was killed fighting at Syracuse — from which that dream was so interpreted that, when Eudemus's soul had departed from his body, then he seemed to have returned home.

\ciceroSection{54} Let us add to the philosophers a most learned man, and indeed a divine poet, Sophocles; who, when a heavy golden bowl had been stolen from the temple of Hercules, saw in his sleep the god himself telling him who had done it. This he neglected once and a second time. When the same thing came more often, he went up to the Areopagus and reported the matter; the Areopagites ordered the man whom Sophocles had named to be arrested; he, on being put to the question, confessed and restored the bowl. From this deed that shrine was named the shrine of Hercules the Informer.

\ciceroSection{55} But why do I cite the Greeks? Somehow our own examples please me more. All the historians tell this — the Fabii, the Gellii, but most recently Coelius: When in the Latin War the great votive games were being held for the first time, the state was suddenly roused to arms, and so, the games being interrupted, repeat games were appointed. Before these were held, and when the people had already taken their seats, a slave bearing a yoke was led through the circus while he was beaten with rods. After this, to a certain Roman countryman who was sleeping there seemed to come a figure who said that the lead dancer at the games had not pleased him, and that the man was ordered by him to report this to the Senate; but he did not dare. The same command was given a second time, with a warning not to put the god's power to the test; not even then did he dare. After this his son died, and the same admonition came a third time in his sleep. Then he, having been struck down with paralysis as well, reported the matter to his friends, and on their advice was carried into the Senate-house on a litter; and when he had narrated the dream to the Senate, he returned home sound on his own feet. And so, the dream being confirmed, it has been handed down to memory that the Senate had those games held over again.

\ciceroSection{56} But Gaius Gracchus told many men — as is written in the same Coelius — that, when he was hesitating to seek the quaestorship, his brother Tiberius seemed in his sleep to say to him that, however much he might delay, still he must perish by the same death by which Tiberius himself had perished. Coelius writes both that he himself had heard this before Gaius Gracchus had been made tribune of the plebs, and that Gracchus had told it to many men. What more certain than this dream can be found? And what of those two dreams which are most frequently cited by the Stoics — who, after all, can scorn them? The one is about Simonides: when he had seen the dead body of some unknown man cast out, and had buried it, and had it in mind to embark on a ship, the man whom he had honored with burial seemed to warn him not to do so; that if he sailed, he would perish in shipwreck; and so Simonides turned back, while the rest, who sailed at that time, perished. The other is a very famous dream, handed down thus:

\ciceroSection{57} When two certain Arcadians who were close friends were making a journey together and had come to Megara, the one turned aside to an innkeeper, the other to a guest-friend. After they had dined and gone to rest, in the dead of night the one who was lodging with his guest-friend seemed in his sleep to see the other begging him to come to his aid, because death was being prepared for him by the innkeeper; that he, terrified at first by the dream, started up; then, when he had collected himself and judged that the vision was to be reckoned as nothing, lay down again; that then, as he slept, the same man seemed to him to ask that, since he had not come to his aid while he was alive, he would at least not allow his death to go unavenged: that he had been killed and thrown by the innkeeper into a cart and covered over with dung; he begged him to be at the gate in the morning, before the cart left the town. Stirred indeed by this dream, he was on hand at the gate in the morning before the herdsman, and asked him what was in the cart; the man fled in terror, the dead body was dug out, and the innkeeper, the matter being laid bare, paid the penalty.

\ciceroSection{58} What can be called more divine than this dream? But why do we seek further examples, or ancient ones? I have often told you my own dream, often heard from you yours: that when I was governing Asia as proconsul, I saw in my rest how you, having ridden up on horseback to the bank of a certain great river and being carried forward, suddenly slipped and fell into the river and nowhere appeared, and how I, terrified, trembled with fear; then how you suddenly rose up cheerful and climbed the opposite bank on the same horse, and how we embraced one another. The interpretation of this dream is easy, and it was foretold to me by experts in Asia that there would be those outcomes of affairs which in fact came to pass. I come now to yours.

\ciceroSection{59} I have indeed heard it from you yourself, but our friend Sallustius has told it to me more often: that, while you were staying — in that flight of ours, glorious to us, calamitous to our country — at a certain farmhouse in the plain of Atina, and had kept watch the greater part of the night, at last toward daybreak you began to sleep deeply and heavily; and so, although the journey was pressing, Sallustius nevertheless ordered silence to be kept and did not suffer you to be awakened; but when you had awakened at about the second hour, you told him your dream: that it had seemed to you, while you were wandering sorrowful in lonely places, that Gaius Marius, with his fasces wreathed in laurel, asked you why you were sad, and when you had said that you had been driven from your country by force, that he took you by your right hand, told you to be of good cheer, and handed you over to the nearest lictor, that he might lead you to his own monument, and said that there your safety would be found. Then Sallustius relates both that he himself cried out that a swift and glorious return was being prepared for you, and that you yourself seemed to take delight in the dream. For this much was quickly reported to me: that when you had heard that, in the monument of Marius, that most magnificent decree of the Senate concerning your return had been passed, on the motion of an excellent and most distinguished man, the consul, and that it had been confirmed in a most crowded theater with incredible shouting and applause, you said that nothing could be more divine than that dream of Atina.

\ciceroSection{60} But many dreams are false. No — obscure, perhaps, to us. Yet grant that some are false; what do we say against the true ones? And these indeed would come about far more often, if we went to our rest unimpaired. As it is, weighed down with food and wine, we discern things disturbed and confused. Consider what Socrates says in Plato's \textit{Republic}. For he says: ``When, in those who sleep, that part of the soul which shares in mind and reason lies drowsy and faint, while that part in which there is a certain wildness and a savage brutishness, having been stupefied by immoderate drink and food, leaps up and tosses itself about without measure in sleep — and so all manner of visions are thrown before it, void of mind and reason, so that it seems either to mingle its body with its mother, or with anyone else at all, whether man or god, often a beast, and even to butcher someone and impiously bathe itself in blood, and to do many foul and loathsome things with recklessness and shamelessness.

\ciceroSection{61} But the man who has surrendered himself to rest under a healthful and temperate regimen of life and diet — with that part of the soul which belongs to mind and reflection stirred and roused and filled to the full with the banquet of good thoughts; with that part of the soul which is nourished by pleasure neither wasted by want nor overflowing with surfeit (for either of these is wont to dull the soul's keen edge, whether something be lacking to nature or be too abundant and overflow); and with that third part of the soul, too, in which the heat of anger arises, calmed and quenched — for that man it will come about that, with the two reckless parts of the soul held down, the third part, the part of reason and mind, shines forth and offers itself lively and keen for dreaming, and then the visions that come to him in his rest will be tranquil and truthful. These are Plato's very words, which I have reproduced.

\ciceroSection{62} Shall we, then, listen rather to Epicurus? For Carneades, out of love of dispute, says now this, now that; Epicurus says what he thinks — but he never thinks anything elegant, anything seemly. Will you, then, set this man before Plato and Socrates, who, even if they gave no reasoned account, would still overcome these petty philosophers by their authority alone? Plato, then, bids us go to sleep with our bodies so disposed that there be nothing to bring error and disturbance upon our souls. From this, too, it is thought that the Pythagoreans were forbidden to eat the bean, because that food causes much flatulence, which is contrary to a mind seeking the truth in tranquillity.

\ciceroSection{63} When, therefore, the soul has been called away in sleep from fellowship and contact with the body, then it remembers things past, discerns things present, foresees things to come; for the body of the sleeper lies as if dead, while the soul is vigorous and alive. This it will do far more after death, when it has departed from the body altogether. And so, as death draws near, the soul is far more divine. For those who are afflicted with a grave and deadly sickness see this very thing — that death is upon them; and so for the most part the images of the dead come to meet them, and then above all they are eager for honor, and those who have lived otherwise than they ought then repent most of their sins.

\ciceroSection{64} That the dying possess powers of divination Posidonius confirms by this further example, which he brings forward: that a certain Rhodian, as he was dying, named six men of like age and said which of them would die first, which second, and which thereafter in succession. But he holds that men dream through the impulse of the gods in three ways: first, because the soul of itself foresees, inasmuch as it is held by kinship with the gods; second, because the air is full of immortal souls, in which, as it were, marks of truth stand stamped and apparent; third, because the gods themselves converse with the sleeping. And this, as I said just now, comes about more readily as death draws near, so that souls divine the future.

\ciceroSection{65} From this comes that saying of Callanus, of whom I spoke before, and the saying of Homer's Hector, who in dying foretells to Achilles his death close at hand. For common usage would not have approved that word at random, if the thing were nothing at all: \\ My soul foreboded I went forth in vain, \\ when I went out of doors. For \textit{sagire} means to perceive keenly; from which old women are called \textit{sagae}, because they wish to know much, and keen-scented dogs are called \textit{sagaces}. The man, then, who perceives a thing before it is set before him is said to have a presentiment — that is, to perceive the future beforehand.

\ciceroSection{66} There is, then, in souls a presentiment cast in from without and enclosed within them by divine power. If this has blazed up rather fiercely, it is called frenzy, when the soul, drawn away from the body, is roused by a divine impulse. \\ But why have your eyes of a sudden begun to rave with kindled fire? \\ Where is that wise and maidenly modesty of a moment ago? \\ Mother, of all good women far the best, \\ I am sent forth to utter prophecies of superstition; \\ for Apollo drives me, unwilling, mad with fates that must be told. \\ I am ashamed before the maidens of my own age; I am ashamed of my deed before my father, \\ best of men. Mother, I pity you, I loathe myself. \\ For Priam you bore an excellent offspring, save only me — this is my grief. \\ That I should stand in the way, and they be of service; I a hindrance, they obedient? A tender poem, full of character, and soft! But that is less to the point;

\ciceroSection{67} the thing we want is brought out — that frenzy is wont to prophesy the truth. \\ It is here, it is here, the torch wreathed in blood and fire! \\ For many years it lay hidden; citizens, bring aid and quench it! \\ It is now a god enclosed in a human body that speaks, not Cassandra. \\ And now upon the great sea a swift fleet is being woven; \\ it hurries on a swarm of destruction; \\ there will come a wild host that will fill the shores with sail-flying ships. I seem to be talking tragedies and tales.

\ciceroSection{68} But from you yourself I have heard a thing of the same kind, not made up but true: that Gaius Coponius came to you at Dyrrachium, when he was in command of the Rhodian fleet with praetorian authority — a man pre-eminently prudent and learned — and that he told you how a certain oarsman from a Rhodian quinquereme had prophesied that within less than thirty days Greece would be drenched in blood; that there would be plundering at Dyrrachium and an embarkation onto the ships, and, for those fleeing, a pitiable backward look at the burnings; but that to the Rhodian fleet a swift return and a homeward journey would be granted. And then you yourself were not unmoved, and Marcus Varro and Marcus Cato, learned men who were there at the time, were thoroughly terrified; and very few days later Labienus came in flight from Pharsalus; and when he had brought word of the army's destruction, the rest of the prophecy was soon fulfilled.

\ciceroSection{69} For the grain, plundered and poured out from the storehouses, had spread over all the streets and alleys, and in sudden terror you went on board the ships, and at night, looking back toward the town, you saw the blazing transports which the soldiers had set on fire because they had refused to follow; and at last, deserted by the Rhodian fleet, you perceived that the seer had spoken truly.

\ciceroSection{70} I have set out as briefly as I could the oracles of dreaming and of frenzy, which I had said were without art. For both kinds there is one and the same account, which our friend Cratippus is wont to use: that the souls of men are in some part drawn and drained from without (from which it is understood that there is a divine soul outside, from which the human soul is derived); but that the part of the human soul which has sensation, motion, and appetite is not separated from the action of the body; whereas the part of the soul which shares in reason and intelligence is then most vigorous, when it is furthest removed from the body.

\ciceroSection{71} And so, when examples of true prophecies and dreams have been set out, Cratippus is wont to draw his conclusion in this way: If the function and office of eyes cannot exist without eyes, yet eyes can sometimes fail to perform their office — then whoever has once so used his eyes as to see truly has the sense of eyes that see truly. In the same way, then, if the function and office of divination cannot exist without divination, yet a man who has the power of divination can sometimes err and fail to see truly — it is enough, for the confirmation of divination, that something has once been so divined that nothing can be thought to have fallen out by chance. And of that kind there are countless instances; that divination exists, therefore, must be admitted. But those kinds of divining which are unfolded by conjecture, or have been observed and marked by their outcomes, are called, as I said above, not natural but produced by art;

\ciceroSection{72} and among these are reckoned the haruspices, the augurs, and the interpreters. These are disapproved by the Peripatetics, defended by the Stoics. Some of them rest upon records and a discipline — as the Etruscan books on entrails, on lightning, and on ritual make plain, and your own books of augury as well — while others are unfolded by conjecture suddenly and on the spot, as in Homer Calchas, who from the number of sparrows foretold the years of the Trojan war, and as we see written in the history of Sulla — a thing that happened with you looking on — that, when he was sacrificing in the territory of Nola before his headquarters, a snake suddenly emerged from the bottom of the altar, whereupon the haruspex Gaius Postumius begged him to lead the army out on campaign; and when Sulla had done so, he then captured before the town of Nola the flourishing camp of the Samnites.

\ciceroSection{73} A conjecture was drawn in the case of Dionysius too, a little before he began to reign. As he was making a journey through the Leontine country, he had let his horse down into a river, and the horse, sunk in the whirlpools, did not come back up; though he tried with the greatest effort, he could not drag it out, and so he went off, as Philistus says, taking it hard. But when he had advanced some little way, suddenly he heard a whinny, looked back, and joyfully saw his horse full of spirit, with a swarm of bees settled in its mane. This portent had such force that within a few days afterward Dionysius began to reign.

\ciceroSection{74} What of this? A little before the disaster at Leuctra, what a signification was given to the Spartans, when in the temple of Hercules the arms clashed and the statue of Hercules ran with much sweat! And at the very same time at Thebes, as Callisthenes says, in the temple of Hercules the doors, which had been shut fast with bars, suddenly opened of themselves, and the arms that had been fixed upon the walls were found upon the ground. And at that same time, when the divine rite was being performed at Lebadea for Trophonius, the cocks in that place began to crow so continuously that they never left off; whereupon the Boeotian augurs declared that the victory was the Thebans', for the reason that that bird, when beaten, is wont to be silent, but to crow when it has won.

\ciceroSection{75} And in that same season the disaster of the battle of Leuctra was foretold to the Spartans by many signs. For on the statue of Lysander, who had been the most illustrious of the Spartans, which stood at Delphi, a crown of rough wild herbs suddenly sprang up on the head; and the golden stars that had been set up at Delphi by the Spartans after that naval victory of Lysander's in which the Athenians were brought down — the battle in which Castor and Pollux were said to have been seen alongside the Spartan fleet, and those golden stars, as I have said, were the emblems of those gods — these stars set up at Delphi fell down a little before the battle of Leuctra and were never found again.

\ciceroSection{76} But the greatest portent of all that came to those same Spartiates was this: when they had sought an oracle from Jupiter of Dodona, inquiring about the victory, and the envoys had set in place that vessel in which the lots were, a monkey, which the king of the Molossians kept as a pet, threw into confusion both the lots themselves and the rest of the things that had been made ready for the casting, and scattered one thing here, another there. Then the priestess who was set over the oracle is said to have declared that the Spartans must think about their survival, not about victory.

\ciceroSection{77} What of this? In the Second Punic War, did not Gaius Flaminius, in his second consulship, neglect the signs of things to come, to the great ruin of the republic? When, after the army had been purified, he had moved his camp toward Arretium and was leading his legions against Hannibal, both he himself and his horse fell suddenly and without cause before the statue of Jupiter Stator; and he did not take that as a matter of religious scruple, though it seemed to the skilled that a sign had been set before him not to join battle. The same man, when he was taking the auspices by the sacred feeding, was met by the keeper of the chickens, who kept putting off the day for engaging in battle. Then Flaminius asked him what he advised should be done, even if the chickens still would not feed afterward. When the man answered that he should keep quiet, Flaminius said: Splendid auspices indeed, if business can be done while the chickens are hungry, but nothing can be done when they are full! And so he ordered the standards to be torn up and himself to be followed. At that moment, when the standard-bearer of the first line of spearmen could not move his standard from its place, and nothing was accomplished though several came up to help, Flaminius, when the matter was reported to him, neglected it after his fashion. And so within those three hours the army was cut to pieces and he himself was slain.

\ciceroSection{78} A great thing too is that which Coelius added: that at the very time when this disastrous battle was being fought, such great earthquakes occurred in Liguria, in Gaul, on a number of islands, and throughout all Italy, that many towns collapsed, landslides happened in many places, the ground sank, rivers flowed in contrary directions, and the sea flowed into streams. Sure conjectures of divinations are made by the skilled. To the famous Phrygian Midas, when he was a boy, ants heaped grains of wheat into his mouth as he slept. It was predicted that he would be exceedingly rich; and so it came about. And when bees had settled on the lips of Plato, a little child sleeping in his cradle, the answer was that he would be of singular sweetness in his speech. So his future eloquence was foreseen in the infant.

\ciceroSection{79} What of this? Your love and your delight, Roscius — was either he himself, or all Lanuvium on his behalf, lying about this? When he was in his cradle and being reared at Solonium, which is a plain in the Lanuvine country, his nurse, waking by night with a lamp set near, noticed the sleeping boy wrapped about in the embrace of a serpent. Terrified by that sight, she raised an outcry. The father of Roscius brought the matter to the haruspices, who answered that nothing would be more illustrious than that boy, nothing more renowned. And this scene Pasiteles engraved in silver, and our own Archias set it forth in verse. What, then, are we waiting for? Until the immortal gods are among us in the Forum, in the streets, at home? They do not indeed offer themselves to us in person, but they pour out their power far and wide, now enclosing it in the caverns of the earth, now entwining it with the natures of men. For it was the force of the earth that roused the Pythia at Delphi, the force of nature the Sibyl. For consider: do we not see how varied are the kinds of land? Of which there is even some deadly portion, as at Ampsanctus among the Hirpini and the place of Pluto in Asia, which I have seen; and there are some tracts of country that are pestilent, others healthful, some that breed keen wits, others dull — all of which come about both from the variety of the sky and from the unequal exhalation of the lands.

\ciceroSection{80} It also often happens, by some sight, often by the weight of voices and by songs, that minds are driven more violently, often too by anxiety and by fear, such as that woman who is \\ Distraught in mind, as if frenzied or \\ Stirred by the rites of Bacchus, \\ calling upon her own Teucer among the tombs. And that very excitement, too, declares that there is a divine force in the mind. For Democritus denies that anyone can be a great poet without frenzy, and Plato says the same. Let him, if he likes, call it frenzy — provided only that this frenzy be praised as it was praised in Plato's \textit{Phaedrus}. What of this? Can your own speech in the courts, can the very pleading, be vehement and weighty and full, unless the mind itself is more deeply stirred? Indeed I have often seen, even in you, and — to come to lighter matters — in your friend Aesopus, such a blaze of looks and movements that some force seemed to have carried him away from the awareness of his mind.

\ciceroSection{81} There are also often presented to us shapes that in reality are nothing, yet offer the appearance of something. This is said to have befallen Brennus and his Gallic forces when he had carried wicked war against the temple of Delphic Apollo. For then they relate that the Pythia uttered from the oracle: \\ I shall provide for that matter, and the white maidens. \\ From which it came about that maidens seemed to bear arms against them, and the army of the Gauls was buried under snow. Aristotle, indeed, held that even those who raved through a fault of health and were called melancholic had something in their minds that was prophetic and divine. But I rather doubt whether this should be attributed either to those with heart disease or to the delirious; for divination belongs to a sound mind, not a diseased body.

\ciceroSection{82} That divination really exists is concluded by this reasoning of the Stoics: If there are gods and they do not declare beforehand to men what things are going to be, then either they do not love men, or they are ignorant of what is going to come about, or they think it of no concern to men to know what will be, or they judge it beneath their majesty to signify beforehand to men the things that are to be, or the gods themselves cannot even signify them. But they neither fail to love us (for they are beneficent and friends to the race of men), nor are they ignorant of those things that have been established and appointed by themselves, nor is it of no concern to us to know the things that are going to come about (for we shall be more cautious if we know), nor do they reckon this foreign to their majesty (for there is nothing more excellent than beneficence), nor can they fail to foreknow things to come;

\ciceroSection{83} therefore it is not the case that there are no gods and that they give no sign of things to come; but there are gods; therefore they do give signs; and it is not the case that, if they give signs, they grant us no ways to the knowledge of what is signified (for they would signify in vain); nor, if they grant the ways, is there no divination; therefore there is divination.

\ciceroSection{84} This reasoning is used by Chrysippus and Diogenes and Antipater. What is there, then, why one should doubt that the things I have argued are most true, if reason is on my side, if the outcomes are, if the peoples, if the nations, if the Greeks, if the barbarians, if even our own ancestors, if, in short, it has always been so believed, if the greatest philosophers, if the poets, if the wisest of men, who founded commonwealths, who established cities? Or shall we wait until the beasts speak, not content with the agreeing authority of mankind?

\ciceroSection{85} Nor in truth is any other reason brought forward why the kinds of divination I speak of should be nothing, except that it seems hard to say what the principle of each divination is, what its cause. For what has the haruspex to show why a cut lung, even amid favorable entrails, should break off the appointed time and push the day farther on? What the augur, why a raven on the right and a crow on the left should make a thing ratified? What the astrologer, why the star of Jupiter or of Venus, joined with the moon, should be propitious to the births of children, while that of Saturn or of Mars is the contrary? And why does a god warn us while we sleep, yet neglect us while we wake? What cause is there, next, why Cassandra in her frenzy should look forward into the future, while Priam in his wisdom could not do the same?

\ciceroSection{86} Why each thing happens, you ask. Quite rightly; but that is not what is at issue now: whether it happens or does not happen, that is the question. Just as, if I should say that the magnet is a stone which entices iron toward itself and draws it, and could bring no account why this comes about, you would deny that it comes about at all. The same you do in the matter of divination, which we ourselves both behold and hear and read and have received from our fathers. Nor, before philosophy was opened up — which was only lately discovered — did common life doubt about this matter; and after philosophy went forward, no philosopher in whom there was any authority felt otherwise.

\ciceroSection{87} I have spoken of Pythagoras, of Democritus, of Socrates; among the ancients I excepted no one but Xenophanes; I joined to them the Old Academy, the Peripatetics, the Stoics; Epicurus alone dissents. And what is more shameful than this, that the same man holds no virtue to be unbought? Yet who is there whom antiquity, attested and sealed by the most illustrious records, does not move? Homer writes that Calchas the augur was by far the best, and that he was the guide of the fleets to Ilium — by his knowledge of auspices, I take it, not of places.

\ciceroSection{88} Amphilochus and Mopsus were kings of the Argives, but they were augurs too, and they founded Greek cities on the seacoast of Cilicia; and even before these, Amphiaraus and Tiresias were not lowly and obscure men, nor like those of whom Ennius writes,

Who for their own gain's sake \\
stir up invented pronouncements,

but bright and surpassing men, who, warned by birds and signs, foretold things to come; and of one of these Homer says that even among the dead he alone has wisdom, while the rest flit about after the manner of shadows. As for Amphiaraus, the fame of Greece so honored him that he was held to be a god, and from the very ground in which he lies buried oracles were sought.

\ciceroSection{89} What of this? Did not Priam, king of Asia, have both his son Helenus and his daughter Cassandra as diviners — the one by auguries, the other by the stirring of her mind and a divine derangement? In which kind we find it written that certain Marcius brothers, born of a noble house, lived among our ancestors. What of this? Does not Homer relate that Polyidus the Corinthian foretold death both to many others and to his own son as he set out for Troy? In general, among the ancients those who held power likewise held the auguries; for just as they reckoned it kingly to be wise, so too to divine. Our own state is witness, in which both the kings were augurs, and afterward private men, endowed with that same priesthood, governed the commonwealth by the authority of religion.

\ciceroSection{90} And this account of divinations has not been neglected even among barbarian peoples, since in Gaul there are the Druids, of whom I myself came to know Divitiacus the Aeduan, your guest and admirer — who professed that the principle of nature, which the Greeks call \textit{[Greek: physiologia]}, was known to him, and who foretold what things were to come, partly by auguries, partly by conjecture; and among the Persians the magi practice augury and divination, men who gather in a sanctuary for the sake of meditation and of conversing among themselves — which, indeed, you yourselves were once accustomed to do on the Nones;

\ciceroSection{91} nor can anyone be king of the Persians who has not first taken in the discipline and knowledge of the magi. One may see, moreover, certain families too and whole nations devoted to this knowledge. Telmessus is in Caria, a city in which the discipline of the haruspices is preeminent; and likewise Elis in the Peloponnese has two fixed families, the one of the Iamidae, the other of the Clutidae, surpassing in the renown of their haruspicy. In Syria the Chaldeans excel in the knowledge of the stars and in the shrewdness of their minds.

\ciceroSection{92} Etruria, again, observes most knowingly the things struck from the sky, and likewise interprets what is shown by each of the monstrous signs and portents. Wherefore well did the Senate decree among our ancestors, in the time when the empire was flourishing, that ten of the sons of the leading men should be handed over from each of the peoples of Etruria for instruction, lest so great an art, on account of the poverty of men, be drawn away from the authority of religion to hire and gain. The Phrygians, the Pisidians, the Cilicians, and the Arab nation submit very greatly to the significations of birds — which, we have learned, was likewise practiced in Umbria.

\ciceroSection{93} And to me, indeed, it seems that the opportunities for the several kinds of divination were drawn from the very places that each people inhabited. For the Egyptians and Babylonians, dwelling on the level expanses of open plains, where nothing stood up from the earth that could hinder the contemplation of the sky, set all their care upon the knowledge of the stars; the Etruscans, on the other hand, because, steeped in religion, they sacrificed victims more zealously and more often, gave themselves above all to the knowledge of entrails; and because, on account of the thickness of the air, many things befell among them from the sky, and because for the same reason many unwonted things arose, some from the sky, others from the earth, and certain ones even from the conception and breeding of men or of cattle, they came forth as the most practiced interpreters of portents. The force of which words — as you are wont to say — the very words themselves, wisely fixed by our ancestors, make plain. For because they show forth, portend, point out, and predict, they are called showings, portents, monstrous signs, prodigies.

\ciceroSection{94} The Arabs, the Phrygians, and the Cilicians, because they make the most use of the pasturing of cattle, ranging over the plains and mountains in winter and summer, on that account the more easily marked the songs and the flights of birds; and the same was the cause for Pisidia and for this Umbria of ours. Then all of Caria, and especially the Telmessians whom I mentioned before, because they inhabit the richest and most fertile fields, in which many things, on account of their fruitfulness, can be shaped and brought forth, were diligent in observing portents.

\ciceroSection{95} And who does not see that in every best commonwealth the auspices and the other kinds of divination have had the greatest force? What king was there ever, what people, who did not use divine prediction? And not in peace alone, but in war much more so, the greater the struggle was and the more critical the issue of survival. I pass over our own people, who do nothing in war without the entrails, who hold nothing at home without the auspices; let us look at foreign instances. For both the Athenians always brought certain divine priests, whom they call \textit{[Greek: manteis]}, into all their public counsels, and the Lacedaemonians gave their kings an augur as assessor, and likewise willed that an augur should sit among the elders (for so they name the public council), and these same people, in matters of greater moment, always sought an oracle either at Delphi or from Hammon or from Dodona.

\ciceroSection{96} Lycurgus, indeed, who ordered the commonwealth of the Lacedaemonians, confirmed his laws by the authority of Delphic Apollo; and when Lysander wished to alter them, he was prevented by that same religious scruple. And further, those who governed the Lacedaemonians, not content with their waking cares, kept watch in the sanctuary of Pasiphaë — which is in the country near the city — for the sake of dreaming, because they reckoned the oracles of repose to be true.

\ciceroSection{97} Now I come back to our own affairs. How often has the Senate ordered the decemvirs to consult the books! In how great matters, and how often, has it obeyed the responses of the haruspices! For when two suns were seen, and when three moons, and when torches of fire appeared, and when the sun was seen by night, and when a roaring was heard from the sky, and when the heaven seemed to part asunder and balls of fire were noticed in it; when there was reported to the Senate also the subsidence of the land in the Privernate territory, where the earth had sunk to an unmeasured depth, and Apulia had been shaken by the greatest of earthquakes — by which portents great wars and ruinous seditions were announced to the Roman people; and in all these cases the responses of the haruspices agreed with the verses of the Sibyl — what then?

\ciceroSection{98} When at Cumae Apollo sweated, and Victory at Capua? What then? Was the birth of a hermaphrodite not a kind of monstrous sign of fate? What then? When the river Atratus flowed with blood? What then? When often a rain of stones fell, sometimes of blood, at times of earth, once even of milk? What then? When on the Capitol a Centaur was struck from the sky, and on the Aventine the gates and men were struck, and at Tusculum the temple of Castor and Pollux, and at Rome the temple of Piety — did not the haruspices both give responses that came to pass, and were the same predictions not found in the books of the Sibyl?

\ciceroSection{99} Lately, in the Marsic War, a temple was restored by the Senate to Juno the Savior on account of a dream of Caecilia, daughter of Quintus. This dream Sisenna, when he had argued that it agreed marvelously, word for word, with the event, then argues — presumptuously, led on, I believe, by some Epicurean — that one ought not to believe in dreams. Yet the same man argues nothing against portents, and sets out that at the beginning of the Marsic War the images of the gods sweated, blood flowed, the heaven parted, voices were heard out of the hidden dark announcing the dangers of war, and at Lanuvium the shields — which seemed to the haruspices the most ominous thing of all — were gnawed by mice.

\ciceroSection{100} What of this, that we have it in the annals that in the Veientine War, when the Alban lake had risen beyond measure, a certain Veientine, a man of nobility, deserted to us, and said that — out of the writings of fate which the Veientines held — Veii could not be taken so long as that lake was overflowing; and that if the lake were let out and flowed off in its own gliding course down to the sea, this would be ruinous to the Roman people; but if it were drawn off in such a way that it could not reach the sea, then it would be the salvation of our side? From this came that admirable draining of the Alban water, carried out by our forefathers. And when the Veientines, worn down by the war, had sent envoys to the Senate, then one of them is said to have stated that the deserter had not dared to tell the Senate everything; for in the same writings of fate the Veientines held it written that in a short time Rome would be taken by the Gauls — which indeed we see came to pass six years after the capture of Veii.

\ciceroSection{101} Often, too, both in battles Fauns have been heard, and in troubled times truth-telling voices are said to have been sent out of the hidden dark; of which kind let there be, out of many, two examples, but the greatest. For not long before the city was taken, a voice was heard from the grove of Vesta, which slopes down from the foot of the Palatine to the New Road, that the walls and gates be repaired; that it would come to pass, unless provision were made, that Rome would be taken. This, neglected at the time when it could have been guarded against, was atoned for after that greatest disaster had been suffered; for an altar to Aius the Speaker, which we see fenced in, was consecrated facing that very place. And it has been written by many, too, that when an earthquake had occurred, a voice arose from the temple of Juno on the citadel, that an atonement be made with a pregnant sow; and for that reason that Juno was called Moneta, the Warner. Do we, then, despise these things, both signified by the gods and judged true by our forefathers?

\ciceroSection{102} And it was not only the voices of the gods that the Pythagoreans observed, but those of men too, which they call omens. Because our forefathers reckoned these to have force, for that reason in undertaking all matters they would first pronounce whatever was good, auspicious, lucky, and fortunate, and in the sacred rites that were performed publicly it was commanded that men "favor with their tongues," and in the proclaiming of festivals that they "abstain from suits and quarrels." Likewise, in the lustration of a colony by the man who founded it, and when a commander purified his army or a censor the people, men with good names were chosen to lead the victims. The consuls observe this same rule in the levy, that the first soldier enrolled be of good name.

\ciceroSection{103} And you know that these things were preserved by you, both as consul and as commander, with the utmost religious scruple. Our forefathers also wished the prerogative century to be the omen of a duly conducted election. And let me bring forward famous examples of omens. Lucius Paulus, consul for the second time, when it had fallen to him to wage the war against King Perses, on that very day, returning home toward evening, kissing his little daughter Tertia, who was then quite small, noticed that she was rather sad. "What is it, my Tertia?" he said. "Why are you sad?" "My father," she said, "Persa has died." Then he, embracing the girl more tightly, said, "I accept the omen, my daughter." But it was a little puppy of that name that had died.

\ciceroSection{104} Lucius Flaccus, the priest of Mars, I myself heard, when he related that Caecilia, the wife of Metellus, wishing to settle her sister's daughter in marriage, went into a certain shrine for the sake of taking an omen, as was the custom in the manner of the ancients. When the girl stood and Caecilia sat in a chair, and for a long while no voice at all had arisen, the girl, worn out, asked her aunt to grant that she might rest a little while in her chair. But she said, "Indeed, my girl, I yield my seat to you." This omen the event followed; for she herself shortly died, and the girl married the man to whom Caecilia had been wife. I understand very well that these things can be despised, or even laughed at, but that very laughter is, by despising what the gods signify, to suppose there are no gods.

\ciceroSection{105} Why should I speak of the augurs? That is your part; the patronage of the auspices, I say, ought to be yours. To you, when you were consul, Appius Claudius the augur announced that, since the augury of Safety had been thrown into doubt, there would be a domestic war, grim and turbulent; which, breaking out a few months later, was put down by you in fewer days. With this augur I most vehemently agree; for he alone, within the memory of many years, held to the discipline not of chanting an augury by rote, but of divining. Your colleagues used to mock him, and called him now a Pisidian, now a Soran augur; in whose eyes there seemed to be in the auguries no presentiment or knowledge of future truth; they said, wisely, that the religious observances were invented to suit the opinion of the ignorant. But it is far otherwise; for neither in those shepherds over whom Romulus presided, nor in Romulus himself, could there have been such cunning that they should fashion the semblances of religion to deceive the multitude. No, the difficulty and labor of learning produced an eloquent negligence; for men would rather argue that there is nothing in the auspices than learn thoroughly what it is.

\ciceroSection{106} What is more divine than that augury which is in your own poem, the \textit{Marius}? — to use you, by preference, as my authority:

Here of high-thundering Jupiter the winged squire, \\
Wounded by a serpent's bite from the trunk of a tree, \\
Pierces it through with savage talons, lifting the snake \\
Half-alive and writhing fiercely with mottled neck; \\
And tearing the creature as it twists about, bloodying it \\
With its beak, now sated in spirit, now having avenged its bitter pangs, \\
Casts it down still gasping, and dashes the mangled thing into the wave, \\
And turns from the setting of the sun toward the bright east. \\
When Marius, augur of the divine power, beheld her \\
Gliding in flight on auspicious wings, \\
And marked the favorable signs of his own glory and return, \\
The Father himself thundered on the left quarters of the sky. \\
Thus Jupiter confirmed the eagle's clear omen.

\ciceroSection{107} And that augural office of Romulus was pastoral, not urban, nor invented to suit the opinions of the ignorant, but received from sure sources and handed down to posterity. And so Romulus the augur, as it stands in Ennius, with his brother, also an augur:

Caring with great care, and then desiring \\
The kingdom, they devote their effort together to auspice and augury. \\
On the mountain Remus gives himself to the auspice and alone \\
Watches for a favorable bird. But fair Romulus \\
Searches on the high Aventine; he watches the high-flying breed. \\
They contended whether to call the city Roma or Remora. \\
The care of all the men was which should be ruler. \\
They wait; as, when the consul means to give the signal, \\
All gaze eager toward the barriers' edges,

\ciceroSection{108} how soon he may send forth the chariots from the painted jaws — \\
so the people waited, and feared in their faces \\
to which of the two the victory of great kingship would be given. \\
Meanwhile the white sun withdrew into the depths of night. \\
Then the shining light, struck by the rays, gave itself forth, \\
and at the same time from on high, far off, a most beautiful bird, \\
flying swiftly, soared on the left. At the same time the golden sun rose, \\
and there came down from the sky thrice four sacred bodies \\
of birds, betaking themselves to auspicious and fair places. \\
From this Romulus sees that to him had been given the priority, \\
the throne and seat of the kingdom made firm by augury.

\ciceroSection{109} But to let the discourse return to the same point from which it digressed: if I could argue nothing as to the reason why each thing comes to pass, and only showed that the things I have recounted do come to pass, would I not give answer enough to Epicurus or Carneades? And what if there is also a ready account of foreknowledge by art, while the account of natural divination is a little more obscure? For the things foreknown from entrails, from lightning-flashes, from portents, from the stars — these have been marked by long observation. And antiquity, in all things, by observation over a long stretch of time, brings an incredible store of knowledge; and this can exist even without any motion or impulse of the gods, since it has been discerned by repeated noticing what comes of each thing, and what each thing signifies.

\ciceroSection{110} The other divination is natural, as I said before; and this, by the subtlety of physical reasoning, must be referred to the nature of the gods, from which — as the most learned and wisest men have held — we hold our souls drawn off and, as it were, poured out. And since all things are filled and crammed with the eternal sense and the divine mind, it must be that human souls are stirred by their kinship with the divine souls. But waking souls serve the necessities of life and sever themselves from the divine fellowship, hampered by the chains of the body.

\ciceroSection{111} Rare is a certain kind of men who call themselves away from the body and are swept off, with every care and zeal, to the knowledge of divine things. The auguries of these men come not from a divine impulse but from human reason; for by nature they foreknow things to come — the floodings of waters, and that someday there will be a conflagration of sky and earth. Others, trained in public affairs — as we have learned of Solon the Athenian — foresee the rising tyranny long beforehand; these we may call prudent, that is, foreseeing, but divine we can in no way call them, no more than Thales of Miletus, who, in order to confound those who reproached him and to show that a philosopher too, if he found it convenient, could make money, is said to have bought up the whole olive crop in the territory of Miletus before it had begun to bloom.

\ciceroSection{112} He had perhaps perceived, by some science, that there would be an abundance of olives. And indeed this same man is reported to have been the first to predict the eclipse of the sun that happened in the reign of Astyages. Physicians foreknow much, pilots much, farmers too foreknow much, but I call none of these divination — not even that by which the Spartans were warned by Anaximander the natural philosopher to leave the city and their houses and to keep watch under arms in the countryside, because an earthquake was at hand — at the time when both the whole city collapsed and from Mount Taygetus the end of the mountain was torn away like the stern of a ship. Nor will even Pherecydes, that teacher of Pythagoras, be reckoned divine rather than a natural philosopher, because, when he had seen water drawn from a never-failing well, he said an earthquake was at hand.

\ciceroSection{113} Nor indeed does a man's soul ever divine naturally, except when it is so loosed and emptied that it has plainly nothing to do with the body — which befalls either seers or sleepers. And so those two kinds are approved by Dicaearchus and, as I said, by our Cratippus; if for the reason that they proceed from nature, let them be supreme indeed, only let them not be the sole kinds; but if they think there is nothing in observation, they do away with much by which the conduct of life is held together. Yet since they grant something, and that no small thing — predictions along with dreams — there is no reason why we should contend greatly with these men, especially since there are those who approve no divination whatever.

\ciceroSection{114} Therefore those men too, whose souls, scorning their bodies, fly out and run abroad, kindled and spurred by some burning, surely behold the things they pronounce when they prophesy; and souls of such a kind, which do not cling to bodies, are kindled by many things — as those who are roused by some sound of voices and by Phrygian strains. Many are moved by groves and woods, many by rivers or seas, and their frenzied mind sees long beforehand what is to come. Of this kind are those lines:

Alas, look! \\
Someone will judge a famous judgment among three goddesses, \\
by which judgment a Spartan woman, \\
one of the Furies, will come.

For in the same way many things have often been foretold by those who prophesy, and not only in words but also in verses, such as the Fauns and seers of old sang. In like manner Marcius and Publicius are said to have sung as seers;

\ciceroSection{115} of this kind are the secret sayings of Apollo that have been brought forth. I believe too that there were certain exhalations of the earth, by which minds, breathed full, poured out oracles. And this indeed is the account of seers, and not unlike, surely, that of dreams. For the things that befall seers awake, the same befall us asleep. For the soul thrives in sleep, free from the senses and from every encumbrance of cares, the body lying down and all but dead. And because it has lived from all eternity and has consorted with countless souls, it sees all that is in the nature of things, provided only that it is so disposed — with sparing food and moderate drink — that, the body lulled, it itself keeps watch. This is the divination of one who dreams.

\ciceroSection{116} Here a great thing arises — not natural, but the interpretation of dreams by art, the kind Antiphon practiced — and in the same way there are interpreters of oracles and prophecies too, expounders, as the grammarians are of poets. For just as the divine nature would have brought forth gold and silver, bronze and iron in vain, had it not also taught the way by which one comes to their veins, nor would have given the crops of the earth or the fruits of trees with any benefit to the human race had it not handed down their cultivation and conditions, nor would timber be of any help, had we not the craft of cutting it — so, joined to every benefit the gods have given to men, there is some art by which that benefit can be grasped. Likewise, then, since in dreams, prophecies, and oracles there was much that was obscure, much that was ambiguous, the expositions of interpreters were brought to bear.

\ciceroSection{117} But how either seers or dreamers see the things which as yet exist nowhere at all is a great question. Yet if those points which ought to be inquired into first are settled, the things we are inquiring after will be easier. For this whole question is governed by that account which concerns the nature of the gods, set forth lucidly by you in the second book. If we hold to it, that will stand firm which governs the matter we are treating: that the gods exist, that the world is administered by their providence, and that these same gods take thought for human affairs — not only for all men together, but also for individuals. If we hold to these things, which to me at least do not seem capable of being torn down, then surely it must be that things to come are signified to men by the gods.

\ciceroSection{118} But it seems we must distinguish in what manner. For the Stoics are not pleased that a god should be present at every cleaving of livers or every singing of birds; for it is neither seemly nor worthy of the gods, nor can it happen in any way at all. Rather, the world was so framed from the beginning that certain signs should run ahead of certain events — some in entrails, some in birds, some in lightning-flashes, some in portents, some in stars, some in the visions of dreamers, some in the cries of the frenzied. By those who have well perceived these things, they are not often deceived; what is ill-conjectured and ill-interpreted is false, not through the fault of the things, but through the ignorance of the interpreters. But this once laid down and granted — that there is a certain divine force holding together the life of men — it is not hard to surmise by what reason the things we certainly see happen do happen. For both a certain perceiving force, which is diffused throughout the whole world, can be a guide to the choosing of the victim, and at the very moment when you wish to sacrifice, a change of the entrails can come about, so that something is wanting or something is over and above; for nature, by small movements, often adds or alters or takes away.

\ciceroSection{119} And that we may not be able to doubt this, there is the greatest proof, in what happened a little before the death of Caesar. When he was sacrificing on that day on which he first sat on the golden chair and came forth in the purple robe, in the entrails of the fat ox there was no heart. Do you suppose, then, that any animal which has blood can be without a heart? He was struck by the strangeness of the thing, when Spurinna said that there was cause to fear that both his counsel and his life might fail him; for each of these proceeds from the heart. The next day there was no head on the liver. These things were portended to him by the immortal gods, that he might see his own death, not that he might guard against it. When, therefore, those parts are not found in the entrails without which the victim could not have lived, it must be understood that at the very time of the sacrifice those parts which are absent perished.

\ciceroSection{120} And the divine mind brings it about likewise in birds, that now hither, now thither the birds of omen fly, now in this, now in that quarter they hide themselves, now from the right, now from the left the birds of song sing out. For if every animal uses the motion of its body as it wills — bending forward, sideways, backward — and bends, twists, stretches out, and draws in its limbs wherever it wills, and accomplishes these things almost before it thinks them, how much easier is this for the god, to whose divine will all things are obedient! And this same god sends us signs of that kind, very many of which history has handed down, such as that which we see written:

\ciceroSection{121} that if the moon were eclipsed a little before sunrise in the sign of Leo, it would come to pass that Darius and the Persians would be conquered in battle by Alexander and the Macedonians, and that Darius would die; and that, if a two-headed girl were born, there would be sedition among the people and corruption and adultery in the home; and that, if a woman seemed to have given birth to a lion, the commonwealth in which this had happened would be conquered by foreign nations. Of the same kind, too, is that which Herodotus records: that the son of Croesus spoke while still an infant, and that by this portent his father's kingdom and his house collapsed from the foundations. And what history has not handed down that the head of Servius Tullius caught fire as he slept? Therefore, just as the man who has given himself over to rest with his mind prepared, both by good thoughts and by circumstances suited to tranquillity, perceives in dreams things certain and true, so a chaste and pure mind, in waking, is the more ready for the truth both of the stars and of the birds and of the rest of the signs and of the entrails.

\ciceroSection{122} This, surely, is that of which we have heard concerning Socrates, and which is often said by Socrates himself in the books of the Socratics: that there is something divine, which he calls his \textit{[Greek: daimonion]}, to which he himself always submitted — never urging him on, but often calling him back. And Socrates indeed — and what better authority could we seek? — when Xenophon consulted him whether he should follow Cyrus, after he had set out what seemed best to him, said: But our counsel is human; on matters both obscure and uncertain I think we should refer to Apollo — to whom even the Athenians, on a public footing, always referred their greater concerns.

\ciceroSection{123} It is written, too, that when he had seen the eye of Crito, his intimate friend, bound up, he asked what was the matter; and when Crito answered that, as he was walking in the country, a twig had been pulled back, and on being released had sprung back into his eye, then Socrates said: Yes, for you did not obey me when I called you back, using, as I am wont, my divine presentiment. The same Socrates, too, when there had been ill fighting near Delium under the command of Laches, and he was fleeing along with Laches himself, when they came to a place where three roads met, refused to flee by the same way as the rest. And when they asked why he did not go on by the same road, he said that he was being warned off by the god; and indeed those who had fled by the other road fell in with the enemy's cavalry. Very many instances have been gathered by Antipater of things wonderfully divined by Socrates; which I shall pass over, for they are known to you, and to recount them is not necessary for me.

\ciceroSection{124} Yet this remains a grand and almost divine thing in that philosopher: that, when he had been condemned by impious verdicts, he said he would die with the most equable mind; for neither as he left his house, nor as he mounted that platform on which he pleaded his cause, was any sign given him by the god, as was its custom, of some impending evil. For my part, I hold thus: that even though many things deceive those who seem to divine either by art or by conjecture, there is nonetheless such a thing as divination; but that men, as in the other arts, so in this one, may be deceived. It can happen that some sign given ambiguously is taken for a certain one; it can happen that some sign has escaped notice — either the sign itself, or what was contrary to it. But for me, in proving the point I am arguing, it is enough that there be found things divinely foreknown and foretold, not many merely, but even a few.

\ciceroSection{125} Indeed, I would say this without hesitation: if even one single thing has been so foretold and foreknown that, when it comes about, it falls out exactly as it was foretold, and nothing in it appears to have happened by chance and accident, then there certainly is divination, and this must be acknowledged by all. Wherefore it seems to me that, as Posidonius does, the whole force and principle of divining must be traced first from god, of whom enough has been said, then from fate, then from nature. Reason, then, compels us to confess that all things come to pass by fate. By fate I mean what the Greeks call \textit{[Greek: heimarmenē]}, that is, the order and series of causes, when cause linked to cause begets a thing out of itself. This is an everlasting truth, flowing from all eternity. And since this is so, nothing has come to pass which was not going to be, and in the same way nothing is going to be of which nature does not contain the very causes that bring it about.

\ciceroSection{126} From this it is understood that fate is not what is meant superstitiously, but what is meant in the manner of natural philosophy — the eternal cause of things, why both the things that have passed away came to pass, and the things that press upon us come to pass, and the things that follow are going to be. So it comes about that it can be marked by observation what consequence for the most part follows upon what cause, even if not always (for that, indeed, is hard to affirm); and it is probable that these same causes of things to come are discerned by those who behold them either through frenzy or in rest.

\ciceroSection{127} Moreover, since all things come to pass by fate, as will be shown elsewhere, if there could be some mortal who might perceive in his mind the whole binding-together of all the causes, surely nothing would deceive him. For whoever held the causes of things to come would necessarily hold all the things that are going to be. And since no one but god can do this, it is left to man to foreknow things to come by means of certain signs that declare their consequences. For the things that are going to be do not arise suddenly; rather, the passage of time is like the unwinding of a cable, producing nothing new and unfolding each thing in turn. This both those see to whom natural divination has been given, and those by whom the course of things has been marked by observation. And although these latter do not discern the causes themselves, still they discern the signs and tokens of the causes; and when memory and diligence and the records of earlier men are brought to bear upon them, there is produced that divination which is called divination by art — of entrails, of lightning-flashes, of portents, and of signs in the heavens.

\ciceroSection{128} There is no reason, then, to wonder that those things are foreknown by diviners which are nowhere; for they all exist, but they are absent in time. And as in seeds there is contained the force of those things which are produced from them, so in causes there are stored up things to come, which a mind aroused, or one released by sleep, beholds to be coming, or which reason or conjecture foreknows. And as those who have learned the risings, settings, and motions of the sun and moon and the rest of the stars foretell long beforehand at what time each of these will occur, so those who have marked the course of things and the sequence of events, working it through over a long span, understand what is going to be — either always, or, if that is difficult, for the most part; and if even that is not granted, at least sometimes. These, then, and certain arguments of the same kind, showing why there is divination, are drawn from fate.

\ciceroSection{129} From nature, however, there is another reasoning, which teaches how great is the power of the mind when sundered from the senses of the body — which happens most of all to those asleep or stirred in mind. For just as the minds of the gods perceive among themselves, without eyes, without ears, without tongue, what each of them feels (from which it comes about that men, even when they wish or vow something in silence, do not doubt that the gods hear it), so the minds of men, when, released by sleep, they are free of the body, or, stirred in mind, are moved of themselves, free and roused, behold those things which, when mingled with the body, they cannot see.

\ciceroSection{130} And this reasoning from nature is perhaps difficult to carry over to that kind of divination which we say proceeds from art; yet Posidonius probes even this as far as he can. He holds that there are in nature certain signs of things to come. For we have heard that the people of Ceos are wont to watch carefully every year for the rising of the Dog Star, and to take their conjecture from it, as Heraclides of Pontus writes, whether the year is going to be healthful or pestilent. For if the star comes forth duller and, as it were, dimmed with mist, the air is dense and thickened, so that its breath will be heavy and pestilent; but if the star appears bright and clear through, it is signified that the air is thin and pure, and therefore healthful.

\ciceroSection{131} Democritus, moreover, holds that the ancients established a wise practice in having the entrails of sacrificed victims inspected; from their condition and color signs are perceived now of healthfulness, now of pestilence, and sometimes also of what barrenness or fertility of the fields is to come. And if observation and experience have recognized this as proceeding from nature, time could have brought to light many things which, by being attended to, might be marked — so that that physicist of Pacuvius, brought on in the \textit{Chryses}, would seem to know nothing at all of the nature of things. For: \\
Those men who understand the tongue of birds, \\
who draw more wisdom from another's liver than their own — \\
these I think we should sooner hear than heed. \\
Why so? I ask, when you yourself, a few verses later, say splendidly enough: \\
Whatever this is, it gives all life, shapes, nourishes, increases, creates, \\
buries and takes back into itself all things, and is the father of all, \\
and from the same source the same things rise anew and to the same set. \\
What reason, then, is there — since there is one house of all, and it shared in common, and since the minds of men have always been and always will be — what reason is there that they should not be able to perceive what comes from each thing, and what each thing signifies? This, he said, is what I had to say about divination.

\ciceroSection{132} Now I shall declare this: that I do not recognize fortune-tellers by lots, nor those who prophesy for the sake of gain, nor even the necromancies which your friend Appius was wont to employ; in short, I do not value at a straw the Marsian augur, nor the village haruspices, nor the astrologers of the circus, nor the interpreters of Isis, nor the readers of dreams; for these are diviners neither by knowledge nor by art, but \\
superstitious seers and shameless fortune-tellers, \\
men either idle or insane, or commanded by want, \\
who do not know the path for themselves, yet point out the way to another; \\
who promise riches to those from whom they themselves beg a drachma. \\
Out of these riches, then, let them take their drachma for themselves, and give back the rest. \\
And this, indeed, is Ennius, who a few verses earlier holds that the gods exist, but supposes that they do not care what the human race does. I, however, who hold that they both care and even warn, and foretell many things, approve of divination — once levity, vanity, and malice have been shut out. When Quintus had said this, Admirably, I said, are you prepared

\ciceroBook{2}

\ciceroSection{1} As I asked myself, with much and long reflection, by what means I might be of use to as many people as possible — so that I should never let slip an occasion of serving the commonwealth — none greater offered itself than if I should hand down to my fellow citizens the paths of the noblest arts; and this, I judge, I have already achieved in a good many books. For in that book entitled the \textit{Hortensius} I exhorted men, as forcefully as I could, to the study of philosophy; and in the four books of the \textit{Academica} I set out which manner of philosophizing I judged the least arrogant, and at once the most consistent and the most refined.

\ciceroSection{2} And since the foundation of philosophy is laid in the ends of good and evil, that subject was thoroughly cleared by me in five books, so that what was said by each philosopher, and what was said against each, might be understood. As many books followed, the \textit{Tusculan Disputations}, which laid open the matters most necessary for living happily. For the first is on the contempt of death, the second on the endurance of pain, the third on the assuaging of grief, the fourth on the remaining disturbances of the mind; the fifth embraced that subject which most of all illuminates the whole of philosophy, for it teaches that, for living happily, virtue is sufficient to itself.

\ciceroSection{3} When these had been published, three books were completed \textit{De Natura Deorum}, in which the whole inquiry into that subject is contained. And so that it might be plainly and fully rounded off, I have set about writing these books on divination; and if, as I intend, I add to them a work on fate, ample satisfaction will have been given to this whole inquiry. To these books must be reckoned also the six \textit{De Re Publica}, which I wrote at the time when I held the helm of the commonwealth — a great subject, proper to philosophy, treated most copiously by Plato, Aristotle, Theophrastus, and the whole household of the Peripatetics. For why should I speak of the \textit{Consolation}? — which to me at least brings no small healing, and which I think will profit others greatly besides. Lately there has also been interposed that book which I sent to our friend Atticus, \textit{Cato Maior de Senectute}; and above all, since by philosophy a man is made good and brave, our \textit{Cato} must be placed in the number of these books.

\ciceroSection{4} And since Aristotle, and Theophrastus likewise — men outstanding alike in subtlety and in abundance — joined to philosophy the precepts of speaking as well, my own books on oratory too seem fit to be referred to the same number of books. So there will be three \textit{De Oratore}, a fourth the \textit{Brutus}, a fifth the \textit{Orator}. So far this was the tally; and toward what remained I was pressing with eager spirit, so prepared that, unless some weightier cause had stood in the way, I would have suffered no province of philosophy to lie unillumined by Latin letters and unopened. For what greater or better service can we render the commonwealth than if we teach and instruct the young? — especially in these manners and these times, in which it has so far slipped that it must be reined in and held in check by all our resources.

\ciceroSection{5} Nor indeed do I trust that this can be brought about — what is not even to be demanded — that all the young should turn themselves to these studies. A few, would that there were! Yet their industry might range widely through the commonwealth. For my part I take a harvest of my labor even from those who, already advanced in age, find rest in my books; by their eagerness to read, my own eagerness to write is more keenly spurred day by day; and these, indeed, I have found to be more numerous than I had supposed. This too is a magnificent thing, and a glory to the men of Rome — that they should have no need of Greek letters for philosophy;

\ciceroSection{6} which I shall surely attain, if I complete what I have begun. And to me, in fact, the cause of expounding philosophy was furnished by the grievous calamity of the state, when amid civil arms I could neither guard the commonwealth in my own fashion, nor do nothing, nor find what else, at least worthy of me, I might do. My fellow citizens, then, will grant me their pardon — or rather will be grateful — that, when the commonwealth lay in the power of a single man, I neither hid myself away, nor deserted my post, nor cast myself down, nor bore myself as one angry at a man or at the times; nor, furthermore, did I either flatter or admire another's fortune in such a way that I should repent of my own. For this very thing I had learned from Plato and from philosophy: that there are certain natural revolutions of commonwealths, so that they are held now by leading men, now by the people, sometimes by single rulers.

\ciceroSection{7} And when this had befallen our own commonwealth, then, stripped of my former functions, I began to renew these studies, both that my mind might be lightened of its troubles by this means above all, and that I might be of use to my fellow citizens in whatever way I could. For in my books I delivered my opinion as in the Senate, I addressed the assembly, I supposed philosophy to have been put in place for me in lieu of administering the commonwealth. Now, since I have begun to be consulted on public affairs, my labor must be given to the commonwealth — or rather all my thought and care must be set upon it, and only so much left to this study as shall be free from public office and duty. But of this elsewhere, and at greater length; now let us return to the discussion we set ourselves.

\ciceroSection{8} For when my brother Quintus had discoursed on divination in the terms set down in the preceding book, and we seemed to have walked about enough, we sat down in the library that is in the Lyceum. And I said: You have defended the Stoics' opinion with care, Quintus, and in the Stoic manner; and — what most delights me — you have used very many of our own examples, and those indeed clear and illustrious. I must reply, then, to what you have said, but in such a way that I affirm nothing, that I question everything, doubting for the most part and distrusting even myself. For if I had anything certain to say, I myself would be playing the diviner — I, who deny that divination exists.

\ciceroSection{9} What moves me, in fact, is that question which Carneades above all was wont to ask: of what things divination is concerned — whether of those that are perceived by the senses. But these we do see, hear, taste, smell, touch. Is there, then, anything in these matters that we perceive by foresight or by some stirring of the mind rather than by nature itself? Or could that diviner — whoever he is — if he were robbed of his eyes, as Tiresias was, say which things are white, which black; or, if he were deaf, discern the varieties or the measures of sounds? To none, therefore, of the things that are taken in by sense is divination applied. And yet not even in those matters that are handled by an art is there any need of divination. For we are accustomed to bring to the sick not seers or fortune-tellers but physicians; nor, in truth, do those who wish to use the lyre or the pipe receive instruction in handling them from haruspices, but from musicians.

\ciceroSection{10} The same reasoning holds in letters and in the rest of the subjects that have a discipline. Do you suppose that those said to divine can answer whether the sun is greater than the earth or only as great as it appears; whether the moon uses its own light or the sun's; what motion the sun and the moon have, and what the five stars that are said to wander? Neither do those reckoned diviners profess that they will say these things, nor, of the figures described in geometry, which are true and which false; for these belong to the mathematicians, not to the fortune-tellers. As for those matters that are dealt with in philosophy, is there anything that any of the diviners is wont either to answer or to be consulted on — what is good, what evil, what neither? For these are proper to the philosophers.

\ciceroSection{11} And what of duty? Does anyone consult a haruspex on how he is to live with his parents, with his brothers, with his friends? On how he is to use his money, how his honors, how his command? These matters are usually referred to wise men, not to diviners. And what of the things handled by the dialecticians or by the natural philosophers — can any of them be divined? Whether the world is one or many; what are the first beginnings of things, from which all things are born: that is the wisdom of the natural philosophers. And how you are to undo the Liar — which they call \textit{[Greek: pseudomenon]} — or how you are to withstand the heap (which, if need be, one might call by a Latin word the "piler," but there is no need; for, just as philosophy itself, so many words of the Greeks, and so too "sorites," are sufficiently worn into Latin speech): these too, then, the dialecticians will pronounce upon, not the diviners. And what of when it is asked what the best constitution of the commonwealth is, what laws, what manners are useful or useless — shall haruspices be summoned from Etruria, or shall leading men and chosen experts in civil affairs determine it?

\ciceroSection{12} But if there is no divination of the things that lie under the senses, nor of those contained in the arts, nor of those discussed in philosophy, nor of those that turn upon public affairs, then of what things it may be I do not understand at all; for either it ought to be of all things, or some matter must be assigned to it in which it can be exercised. But there is no divination of all things, as reason has shown, nor is there found any place or matter over which we could set divination in charge. See, then, whether there is no divination at all. There is a certain common Greek verse to this effect:

He who shall conjecture well — that man I shall pronounce the best seer.

Will a seer, then, conjecture better than a helmsman what storm threatens, or attain by conjecture the nature of a disease more keenly than a physician, or the conduct of a war more prudently than a general?

\ciceroSection{13} But I noticed, Quintus, that you carefully drew divination away from those conjectures which involve art and good judgment, and away from those things which are grasped by the senses or by the crafts, and that you defined it thus: divination is the prediction and foreknowledge of those things which are fortuitous. First, you are rolled back to the same point. For the physician and the helmsman and the general all have a foreknowledge of fortuitous things. Could any haruspex or augur or seer or dreamer conjecture better than a physician whether a sick man will recover from his disease, than a helmsman whether a ship will escape from danger, than a general whether an army will escape from an ambush?

\ciceroSection{14} And yet you said that not even this belonged to the diviner — to foresee, by certain signs, impending winds or rains (and on this point some of my own renderings of Aratus were recited by you from memory), even though these very things are fortuitous; for they happen more often than not, but not always. Where, then, is this foreknowledge of fortuitous things which you call divination, and on what does it operate? For the things that can be foreseen by art or by reasoning or by experience or by conjecture, these you hold should be assigned not to diviners but to experts. So it is left that the only fortuitous things that can be divined are those which can be foreseen by no art and no wisdom — as, if anyone had said many years beforehand that the famous Marcus Marcellus, who was three times consul, would perish in a shipwreck, he would surely have divined it; for by no other art and by no wisdom could he have known it. Foreknowledge, then, of such things as are set in the lap of fortune is divination.

\ciceroSection{15} Can there, then, be any foreknowledge of things which have no rational ground at all for their coming to be? For what else are luck, fortune, chance, and outcome, except when something has fallen out and come to pass in such a way that it could also have fallen out and come to pass otherwise? How, then, can a thing that happens at random by blind chance and the rolling of fortune be foreknown and predicted?

\ciceroSection{16} A physician foresees a worsening disease by reasoning, a general an ambush, a helmsman a storm; and yet even these men, who form no opinion without sure reasoning, are often deceived — as the farmer, when he sees the flower of the olive, thinks he will see the fruit as well, and not without reason; yet sometimes even so he is deceived. But if those men are deceived who say nothing without some probable conjecture and reasoning, what is to be thought of the conjecture of those who foresee the future by entrails, or birds, or portents, or oracles, or dreams? I am not yet saying how worthless these signs are — the cleft of the liver, the cawing of the raven, the flight of the eagle, the shooting of a star, the cries of men in frenzy, lots, dreams; of each of these I shall speak in its place; for now I speak of them all together.

\ciceroSection{17} How can anything be foreseen as going to be, when it has neither any cause nor any mark to show that it will be? Eclipses of the sun, and likewise of the moon, are predicted for many years in advance by those who pursue the motions of the heavenly bodies by calculation; for they predict the things which the necessity of nature is going to bring about. They see, from the most regular motion of the moon, when she, set opposite the sun, will run into the shadow of the earth, which is the cone of night, so that she must necessarily be darkened; and when that same moon, placed beneath and over against the sun, will darken its light to our eyes; in what sign each of the wandering stars will be at each time; and what rising of some sign, and what setting, will fall on each day. You see by what reasoning those proceed who say these things beforehand.

\ciceroSection{18} But those who say that a treasure will be found, or an inheritance will come, what do they follow? Or in what nature of things does that future event reside? But if these things, and others of the same kind, have any such necessity, what is there left, then, that we should suppose to happen by chance or by random fortune? For nothing is so contrary to reason and constancy as fortune, so that it does not seem to me to fall even within a god's power to know what is going to happen by chance and at random. For if he knows it, then it will certainly come to pass; but if it will certainly come to pass, there is no such thing as fortune; yet there is such a thing as fortune; therefore there is no foreknowledge of fortuitous things.

\ciceroSection{19} Or, if you deny that there is such a thing as fortune, and say that everything which happens and which will happen has been determined by fate from all eternity, then change your definition of divination, which you said was the foreknowledge of fortuitous things. For if nothing can come to be, nothing befall, nothing come to pass, except what from all eternity has been certain to be at a fixed time, what fortune can there be? And once fortune is taken away, what room is left for divination — which you have called the foreknowledge of fortuitous things? Although you did say that everything which happens or will happen is held fast by fate. The very name of fate is downright old-womanish and full of superstition; yet still, among the Stoics much is said about that fate; of which I shall speak elsewhere; for now, only what is necessary.

\ciceroSection{20} If all things come about by fate, what good does divination do me? For what the diviner predicts is indeed going to be, so that I do not even know what to make of that affair, when an eagle called back our friend Deiotarus from his journey; for if he had not turned back, he would have had to lodge in that chamber which collapsed the very next night, and so he would have been crushed beneath the ruin. But that — if it had been fated — he would not have escaped; and if it had not been fated, he would not have fallen into that mishap. What help, then, is divination? Or what is the point of these warnings to me — whether lots, or entrails, or any prediction? For if it was fated that the fleets of the Roman people in the First Punic War should be destroyed, the one by shipwreck, the other sunk by the Carthaginians, then even if the sacred chickens had given a most favorable feeding-omen to the consuls Lucius Junius and Publius Claudius, the fleets would have been destroyed all the same. But if, had the auspices been obeyed, the fleets were not going to be destroyed, then they were not destroyed by fate; and yet you hold that all things come about by fate;

\ciceroSection{21} therefore there is no such thing as divination. But if it was fated that in the Second Punic War the army of the Roman people should be destroyed at Lake Trasimene, could that have been avoided, had the consul Flaminius obeyed those signs and those auspices by which he was forbidden to fight? Certainly it could. Either, then, the army was not destroyed by fate, or, if it was destroyed by fate (which is certainly what you must say), then even if he had obeyed the auspices, the same thing would have come to pass; for the fates cannot be changed. Where, then, is this divination of the Stoics? For if all things come about by fate, it can give us no warning to be more careful; for in whatever way we conduct ourselves, that will still happen which is going to happen. But if it can be averted, there is no fate; and so there is not even divination, since divination is of things to come. But nothing is going to happen for certain, when by some act of expiation it can be brought about that it does not happen.

\ciceroSection{22} And, for that matter, I do not think that the knowledge of things to come is even useful to us. For what kind of life would Priam have had, if from his youth he had known what outcomes he was going to have in old age? Let us leave the legends and look at things nearer home. In my \textit{Consolation} I gathered together the gravest ends of the most illustrious men of our state. What of it, then? To pass over the earlier ones — do you think it would have been useful to Marcus Crassus, at the time when he flourished with the greatest wealth and fortunes, to know that, after his son Publius was killed and his army destroyed, he was destined to perish beyond the Euphrates in disgrace and dishonor? Or do you suppose that Gnaeus Pompey would have taken joy in his three consulships, his three triumphs, and the glory of his greatest deeds, if he had known that he was going to be butchered in the desolation of Egypt with his army lost, and that after his death those things would follow which we cannot speak of without tears?

\ciceroSection{23} And what of Caesar — what agony of soul, do we think, would he have spent his life in, had he divined that he would lie butchered in that Senate which he himself had for the greater part recruited, in the Hall of Pompey, before the very image of Pompey himself, with so many of his own centurions looking on, slain by the noblest citizens, some of them men he had himself decked out in every distinction, lying there so that not one even of his slaves, let alone of his friends, would come near his body? Surely, then, ignorance of future evils is more useful than knowledge of them.

\ciceroSection{24} For this, at any rate, can in no way be said, especially by the Stoics: Pompey would not have gone to arms, Crassus would not have crossed the Euphrates, Caesar would not have taken up the civil war. Therefore their ends were not fated; yet you hold that all things come about by fate; nothing, then, would it have profited those men to divine; and, what is more, they would have lost all the enjoyment of their earlier life; for what could be a delight to them as they brooded on their own ends? So, whichever way the Stoics turn, all their cleverness must lie prostrate. For if the thing that is going to come to pass can come to pass in either this way or that, fortune has the greatest power; and the things that are fortuitous cannot be certain. But if it is certain what is going to be in each matter at each time, what is there in which the haruspices help me? For when they have said that the most grievous things are portended, they add at the very end that everything will fall out more lightly once the divine rites of expiation have been performed;

\ciceroSection{25} For if nothing happens outside of fate, nothing can be lightened by an act of worship. This is Homer's view, when he brings on Jupiter complaining that he could not snatch his son Sarpedon from death against fate. The same thing is signified by that line of the Greek poet, turned to this sense:

That which is fixed to be \\
overmasters even Jove on high.

The whole notion of fate seems to me to be justly mocked even in a line of Atellan farce; but in matters so grave there is no place for jesting. Let the argument, then, be brought to its conclusion. For if nothing can be foreseen of those things that happen by chance, since they cannot be certain, there is no divination; but if they can be foreseen on this ground, that they are certain and fated, then again there is no divination — for you were saying that divination has to do with fortuitous things.

\ciceroSection{26} But let this have been, as it were, a first sally of my discourse with light-armed troops; now let us engage at close quarters and try, if we can, to dislodge the horns of your argument. For you were saying that there are two kinds of divination, one by art, the other natural; that the kind by art consists partly in conjecture, partly in long-continued observation; that the natural kind is what the mind seizes or receives from without, from the divine power, the source from which we all have our minds drawn off, or received, or poured as a libation. As the kinds of divination by art you laid down roughly these: that of those who read entrails, and of those who foretell from lightning-flashes and portents; then that of the augurs, and of those who use signs or omens; and every conjectural kind you placed roughly within this same class.

\ciceroSection{27} The natural kind, on the other hand, seemed either to be uttered and, as it were, poured out in an excitement of the mind, or to be foreseen by a mind set free, during sleep, from the senses and from cares. Moreover, you derived all divination from three things — from a god, from fate, from nature. Yet even so, since you could explain nothing, you fought with a marvelous abundance of made-up examples. About this I should like to say, first of all, this: I judge it not the part of a philosopher to use witnesses, who may be true by chance or false and counterfeit out of malice; one ought to teach by arguments and reasons why each thing is so, not by outcomes, and least of all by outcomes which I am at liberty not to believe.

\ciceroSection{28} Let me begin from haruspicy, which I think ought to be cultivated for the sake of the commonwealth and of the common religion. But we are alone; it is permitted to search out the truth without giving offense — for me especially, who am in doubt about most of these matters. Let us examine the entrails first, if you please. Can anyone, then, be persuaded that the things said to be signified by entrails have been learned by the haruspices through long-continued observation? How long-continued was that observation? Or over how long a stretch of time could it have been carried on? Or in what way was it agreed among them which part is hostile, which part friendly, which cleft betokens danger, which some advantage? Or did the Etruscan haruspices, the Elians, the Egyptians, the Carthaginians compare these matters among themselves? But that, besides being something that could not have happened, cannot even be imagined; for we see different peoples interpret the entrails in different ways, and that there is no single discipline common to all.

\ciceroSection{29} And surely, if there is in the entrails some force that reveals things to come, that force must necessarily either be bound up with the nature of things or be shaped in some way by the will of the gods and by divine power. With the nature of things — so great, so glorious, diffused through all regions and motions — what can a cock's gall-bladder have in common (for there are those who say that these entrails above all are the most telling), nay, what natural thing has the liver or the heart or the lung of a fat bull, that could reveal what is going to be?

\ciceroSection{30} Democritus, to be sure, trifles not unwittily, in the manner of a natural philosopher — a class than which nothing is more arrogant:

What lies before their feet no man regards; \\
they scan the regions of the sky.

Yet he holds that by the condition and color of the entrails this much, at any rate, is revealed: the kind of fodder, and the abundance or scantiness of the things the earth brings forth; and he thinks that healthfulness too, or pestilence, is signified by the entrails. O blessed mortal! who, I am quite certain, was never at a loss for amusement. That this man should have been delighted by such great trifles, so as not to see that the thing would then be probable only if the entrails of all the beasts changed themselves at the same time into the same condition and color! But if at the same hour one beast's liver is sleek and full, another's rough and shrunken, what is there that could be revealed by the condition and color of the entrails?

\ciceroSection{31} Or is this of the same sort as that saying of Pherecydes which was cited by you? — who, when he had seen water drawn from a well, said an earthquake was going to come. Too little shamelessly, I suppose, do they dare, when an earthquake has occurred, to say what force produced it; do they actually foreknow, even beforehand, that it is going to happen, from the color of the spring-water? Many things of this sort are said in the schools, but take care that it not be unnecessary to believe them all.

\ciceroSection{32} But let those Democritean claims be true, by all means: when do we ever search out such things from the entrails? Or when have we ever heard anything of that kind from a haruspex once the entrails have been inspected? They warn of dangers from water or from fire; now they announce inheritances, now losses; they handle the friendly and vital cleft; they consider the head of the liver from every side most carefully; and if it is in fact not found, they think nothing more grievous could have befallen.

\ciceroSection{33} These things, certainly, could not have been observed, as I taught above. They are therefore inventions of art, not of antiquity — if there is any art at all of things unknown. But what kinship do they have with the nature of things? Even granting that it is joined together and continuous by one accord — which I see has found favor with the natural philosophers, and most of all with those who said that everything which exists is one — what can the universe have bound up with the finding of a treasure? For if by the entrails an increase of money is revealed to me, and that happens by nature, then, first, the entrails are joined to the universe, and second, my gain is contained within the nature of things. Are the natural philosophers not ashamed to say such things? For granted that there is now in the nature of things some contagion — which I concede there is (for the Stoics gather many instances; thus the little livers of mice are said to swell in winter, and dry pennyroyal to flower on the very day of the winter solstice, and inflated bladders to burst, and the seeds of apples enclosed in their middle to turn themselves into opposite directions, and now some strings on a lyre to resound when others are struck, and oysters and all shellfish to wax along with the moon and to wane along with it, and trees in the winter season — when the moon is at the same time waning, because they are then dried out — to be thought timely for cutting.

\ciceroSection{34} Why should I say more about the straits or the tides of the sea? whose ebb and flow are governed by the motion of the moon. Six hundred instances of the same sort might be brought forward, so that the natural kinship of things far apart appears) — let us grant this; for it is no obstacle at all to the present argument. Surely it does not follow that, if there is a cleft of a certain kind in the liver, gain is revealed? By what bond of nature and, as it were, harmony and accord — which the Greeks call \textit{[Greek: sympatheian]} — can either the cleft of a liver agree with my little profit, or my paltry gain with the sky, the earth, and the nature of things? I will concede this very point, if you wish, even though I shall have made a great sacrifice of my case if I concede that there is any agreement at all of nature with the entrails;

\ciceroSection{35} but even granting that, how does it come about that the man who wishes to obtain his request sacrifices a victim suited to his affairs? This was what I did not think could be resolved. But how merrily it is resolved! I am ashamed — not of you, indeed, whose memory I even admire, but of Chrysippus, Antipater, Posidonius, who say this very same thing that has been said by you: that in choosing the victim there is a guide, a certain force, perceptive and divine, which is diffused throughout the whole universe. But this is far better still, what was both employed by you and is said by them: that when someone wishes to sacrifice, then a change of the entrails takes place, so that either something is wanting or something is in excess;

\ciceroSection{36} for all things, they say, obey the divine power of the gods. These things, believe me, not even old women now think true. Or do you suppose that, if one man should choose a given calf he will find its liver without a head, and if another, with a head? Can this loss of the head, or its addition, take place so suddenly that the entrails accommodate themselves to the fortune of the one sacrificing? Do you not see that there is a certain throw of the dice in the choosing of victims, especially when the thing itself instructs us? For when the entrails have been most grievous, without a head — than which nothing seems more dire — often with the very next victim the offering is most beautifully favorable. Where, then, are those threats of the earlier entrails? Or how did so great an appeasement of the gods come about so suddenly? But you bring forward that, in the entrails of a fat bull, when Caesar was sacrificing, there was no heart; and that, because it could not have happened that the victim should have lived without a heart, we must judge that the heart perished at the moment when it was being sacrificed.

\ciceroSection{37} How does it come about that you grasp the one thing — that the ox could not have lived without its heart — but fail to see the other, that the heart could not, all at once, have flown off to who knows where? For my part, I can either not know what power the heart has for living, or suspect that the ox had contracted some disease, and that its heart was thin and shrunken and withered, unlike a heart at all. But you — what reason have you to think that, if a heart was in that fat bull a little while before, it suddenly perished in the very act of sacrifice? Or was it that, having looked upon Caesar, robed in purple and heartless, the heart was itself bereft? Believe me, you are betraying the very city of philosophy while you defend its outposts; for while you would have the haruspex's art be true, you are overturning the whole study of nature. The head is found in the liver, the heart in the entrails; then it will vanish the instant you have sprinkled the meal and the wine; a god will snatch it away, some force will make away with it or eat it up. Nature, then, will not bring about the rise and the passing of all things, and there will be something that either rises out of nothing or sinks suddenly into nothing. What natural philosopher ever said this? The haruspices say it; do you, then, judge that they are to be believed sooner than the natural philosophers?

\ciceroSection{38} And again: when sacrifice is made to several gods, how on earth does it turn out that the offering is favorable to some and not favorable to others? And what inconstancy is this among the gods, that they should threaten by the first set of entrails and promise well by the second? Or what discord so great among them — often even among the nearest kin — that Apollo's entrails should be good and Diana's not good? What is so plain as this: that, since the victims are brought up at random, the entrails are such for each man as the victim that has fallen to him? But — you object — this very thing has something divine in it, namely which victim falls to each man, just as in the casting of lots, which lot is drawn for each. Of lots presently; though you, for your part, do not strengthen the case of the victims by likening it to the lots, but rather weaken the lots by setting them alongside the victims.

\ciceroSection{39} When we have sent a man to the Aequimaelium to fetch a lamb for us to sacrifice, is the lamb that is brought to me really the one that has entrails fitted to my affairs, and is it not by chance but with a god for guide that the slave is led to that lamb? For if you say that in this matter too there is a chance that is, as it were, a kind of lot bound up with the will of the gods, I am sorry that our Stoics have given the Epicureans so great an opening to laugh at them; for you are not unaware how they deride such things.

\ciceroSection{40} And indeed they can do so more easily; for it was by way of a joke that Epicurus brought in his very gods — transparent and ventilated, dwelling between two worlds, as if between two groves, for fear of falling debris, and he holds that they have the same limbs we have, and yet make no use whatever of those limbs. So this man, doing away with the gods by a certain roundabout, has good reason not to hesitate to do away with divination as well; but the Stoics are not, like him, consistent with themselves. For his god, having no business of his own nor of another's, cannot impart divination to men; your god, on the other hand, can withhold it, and none the less govern the world and care for men.

\ciceroSection{41} Why, then, do you entangle yourselves in those snares that you can never unravel? For this is how, when they are in more of a hurry, they are accustomed to draw the conclusion: If there are gods, there is divination; but there are gods; therefore there is divination. Far more probable is this: but there is no divination; therefore there are no gods. See how rashly they commit themselves to the position that, if there is no divination, there are no gods. For divination is plainly done away with, while the existence of the gods must be held fast.

\ciceroSection{42} And once this divination by the inspection of entrails is done away with, the whole of the haruspex's art is done away with. For portents and lightning follow upon it. Now in the case of lightning, long observation prevails; in the case of portents, reasoning and conjecture are for the most part brought to bear. What is it, then, that has been observed in lightning? The Etruscans divided the heaven into sixteen parts. That was easy enough — to double the four that we have, and afterward to do the same again, so as to say from this from which part the bolt had come. First, what difference does that make? And next, what does it signify? Is it not plain that it was out of the first amazement of men — because they had taken fright at thunderclaps and the hurling of lightning — that they believed Jupiter, all-powerful over all things, brings these about? And so in our records we have it written: When Jupiter thunders or makes lightning, it is forbidden to hold an assembly of the people.

\ciceroSection{43} This was perhaps established for the sake of the commonwealth; for they wished there to be grounds for not holding the assemblies. And so lightning is a fault only for the assemblies, though the very same thing we reckon the best of all auspices, if it has come on the left. But of auspices in another place; now of lightning. What, then, ought to be less said by natural philosophers than that anything certain is signified by things uncertain? For I do not suppose you are the kind of man to think the Cyclopes forged Jupiter's thunderbolt on Aetna;

\ciceroSection{44} for it would be a marvel how Jupiter could hurl it so many times, when he had but one; nor in fact did he warn men by lightning what they ought to do or beware. For it is the Stoics' view that those exhalations of the earth which are cold, once they have begun to flow, are winds; but when they have wrapped themselves in a cloud and have begun to part and tear apart its thinnest portions, and to do this more thickly and more violently, then both flashes and thunderclaps arise; and if, by the clashing of the clouds, a heat is forced out and discharges itself, that is the thunderbolt. From a thing, then, which we see brought about by the force of nature, with no constancy, at no fixed time — from this do we look for a signifying of things to follow? Surely, if Jupiter were signifying these things, he would not send out so many bolts in vain! For what does he accomplish when he has hurled a bolt into the middle of the sea?

\ciceroSection{45} What, when into the highest mountains, as for the most part he does? What, when into deserted wastes? What, when onto the shores of those nations among whom these things are not even observed? But — a head was found in the Tiber. As if I denied that these men have some art! It is divination I deny. For the division of the heaven, which I spoke of before, and the marking of certain things, teaches whence the bolt has come, whither it has withdrawn; but what it signifies, no reasoning teaches. Yet you press me with my own verses:

For the high-thundering Father, leaning on starry Olympus, \\
Once aimed his bolts at his own mounds and temples, \\
And cast his fires upon the Capitoline seats. \\
Then Natta's statue, then the images of the gods, \\
And Romulus and Remus with their nursing beast \\
Were struck down by the force of the bolt and fell; \\
And concerning these matters the responses of the haruspices proved most true.

\ciceroSection{46} And there was this marvel besides: that at the very moment when the disclosure of the conspiracy was being made in the Senate, the statue of Jupiter — two years after it had been commissioned — was being set in place on the Capitol. — Will you, then (for this is how you were pressing me), bring yourself to defend that case both against your own deeds and against your own writings? — You are my brother; on that account I hold back. But what, after all, does it harm you here? Is it the matter, which is what it is, or I, who wish the truth to be laid bare? And so I say nothing against you; from you I ask the reasoned account of the whole haruspex's art. But you have flung yourself into a wonderful hiding-place; for, because you understood that you would be hard pressed when I inquired from you the causes of each several kind of divination, you spent many words to the effect that, since you saw the facts, you did not seek the reason and the cause — that what happens, not why it happens, is what bears on the matter. As if I would grant either that it happens, or that it belonged to a philosopher not to ask the cause

\ciceroSection{47} why each thing happens! And at just that point you were quoting my own \textit{Prognostics}, and the kinds of plants, the scammony and the aristolochia-root, whose cause you did not know, though you saw their power and effect. The whole comparison fails. For the causes of weather signs have been pursued, both by Boethus the Stoic, whom you named, and by our own Posidonius too; and even if the causes of those plants are not found, still the facts themselves could be observed and noted. But Natta's statue, or the bronze of the laws struck from the sky — what is there in them that is observed and ancient? The Pinarii Nattae are noble; from the nobility, then, comes danger. So cleverly did Jupiter think it out! Romulus, still at the breast, was struck by the bolt; therefore danger is shown to the city, the very one that he founded. How shrewdly Jupiter informs us by tokens! But at the same time the statue of Jupiter was being set in place at which the conspiracy was being disclosed. And you, of course, would rather hold that this was done by the will of the gods than by chance — and that the contractor who had taken on the making of that column from Cotta and Torquatus was not the slower through idleness or want of means, but was held in reserve by the immortal gods for that very hour.

\ciceroSection{48} I do not, for my part, altogether despair that these things are true; but I do not know, and I wish to learn from you. For when certain things seemed to me to come about by chance in such a way as to have been foretold by the diviners, you said much about chance — that a Venus-throw can be cast by chance with four dice thrown, but that with four hundred throws a hundred Venus-throws cannot come up by chance. First, I do not know why they cannot, but I do not press the point; for you abound in parallels. You have your spattering of paints, and the snout of a pig, and a great many others besides. You say, too, that Carneades invented the story about the head of the little Pan; as if it could not have come about by chance, and as if it were not necessary that in every block of marble there should be heads worthy even of Praxiteles! For those very heads are produced by taking away, and nothing whatever is brought to the marble by Praxiteles; rather, when much has been taken away and the lines of the face have been reached, then you may understand that the thing now polished into shape was within all along.

\ciceroSection{49} Something of that kind, then, could also have come into being of its own accord in the quarries of Chios. But grant that this is a fiction; what of it? Have you never noticed in the clouds the shape of a lion or of a hippocentaur? It is possible, then — the very thing you were just now denying — for chance to imitate the truth. But since we have argued enough about entrails and about lightnings, the portents remain, so that the whole of haruspicy may be handled through. You brought up the case of a mule's giving birth. The thing is marvelous, for the reason that it does not happen often; but if it could not have happened, it would not have happened. And let this hold against all portents: that what could not have come to pass never has come to pass; while if it could, there is nothing to wonder at. For ignorance of causes, in a novel matter, produces wonder; but the same ignorance, when it is about things in common use, does not make us wonder. For the man who wonders that a mule has given birth does not know in what way a mare gives birth, or in general what nature it is that brings about the birth of a living thing. But what he sees often, he does not wonder at, even though he does not know why it happens; what he has not seen before, if it occurs, he reckons to be a portent. Which, then, is the portent — when the mule conceived, or when she gave birth?

\ciceroSection{50} The conception, perhaps, is contrary to nature; but the birth is all but necessary. But why say more? Let us look at the origin of haruspicy; thus we shall most easily judge what authority it has. A certain Tages, it is said, in the territory of Tarquinii, when the earth was being plowed and a furrow had been driven rather deep, suddenly rose up and addressed the man who was plowing. Now this Tages, as it stands in the books of the Etruscans, is said to have appeared with the look of a boy, but to have had the wisdom of an old man. When the plowman, dumbstruck at the sight of him, raised a great cry in his amazement, a crowd gathered, and in a short time the whole of Etruria assembled at that spot; then he spoke at length, with many listening, who took down all his words and committed them to writing; and the whole discourse was that in which the discipline of haruspicy is contained. Afterward this grew, as new matters were learned and referred back to those same first principles. This is what we have received from them; this is what their writings preserve; this is the fountainhead they hold their discipline to have.

\ciceroSection{51} Is there any need, then, of a Carneades to refute these things? any need of an Epicurus? Is there anyone so out of his wits as to believe that this thing was plowed up — shall I call it a god or a man? If a god, why had he hidden himself in the earth against his nature, so that, laid open by the plow, he might look upon the light? Could not the same god, indeed, have handed down the discipline to men from some higher place? But if that Tages was a man, in what way could he live buried under the earth? And from what source, moreover, could he himself have learned the things he was teaching to others? But I am more foolish than the very men who believe these things, in arguing against them at such length. That old saying of Cato's, however, is quite shrewd, who used to say that he wondered a haruspex did not laugh when he caught sight of a haruspex.

\ciceroSection{52} For how few of all the things foretold by these men come to pass? Or, if anything does come to pass, what reason can be brought to show that it did not come about by chance? King Prusias, when Hannibal, in exile at his court, urged that battle be joined, kept saying that he did not dare, because the entrails forbade it. Do you say so? Hannibal replied. You would rather trust a little scrap of veal than an old commander? And what of Caesar himself, when he was warned by the foremost haruspex not to cross over into Africa before the winter solstice — did he not cross over? And had he not done so, all the forces of his adversaries would have gathered into one place. Why should I rehearse the responses of the haruspices — countless ones, indeed, I could — which either had no outcomes at all, or outcomes the very reverse?

\ciceroSection{53} In this civil war, immortal gods, how many tricks they played! What responses of the haruspices were sent to us in Greece from Rome! What things were told to Pompey! For he was very much moved by entrails and by portents. I have no wish to rehearse them, nor indeed is there any need, least of all to you, who were there yourself; yet you see that nearly everything turned out contrary to what had been said. But enough of this; now let us come to the portents.

\ciceroSection{54} You read out many things written by me myself in my consulship; you brought forward many gathered before the Marsic War by Sisenna; you cited many recorded by Callisthenes before the Lacedaemonians' disastrous battle at Leuctra. Of these I shall indeed speak one by one, as far as it seems good; but something must be said about them all together as well. For what is this signifying, as if proceeding from the gods, this giving notice, so to speak, of calamities? And what do the immortal gods intend, in the first place by signifying things which we cannot understand without interpreters, and in the second place by signifying things which we have no power to guard against? Why, not even decent men do this — to foretell to their friends impending calamities which those friends can in no way escape; just as physicians, although they often understand the case, nevertheless never tell the sick that they are going to die of that disease. For every prediction of an evil is approved only when, to the prediction, a means of precaution is attached.

\ciceroSection{55} In what way, then, did either the portents or their interpreters help the Lacedaemonians of old, or our own people lately? And if these are to be reckoned signs of the gods, why were they so obscure? For if it was so that we might understand what was going to happen, the matter ought to have been declared openly; or it ought not to have been declared even covertly, if they did not wish these things to be known. And then again, all conjecture, on which divination rests, is often drawn off by the dispositions of men into many, or differing, or even contrary directions. For as in judicial cases the conjecture of the prosecutor is one thing and that of the defender another, and yet each is plausible, so in all those matters which seem to be tracked down by conjecture, the reasoning is found to face two ways. But where things are brought about now by nature, now by chance, and sometimes resemblance even breeds error, it is great folly to make the gods the authors of those things, and not to seek the causes of things.

\ciceroSection{56} You believe that the Boeotian seers at Lebadia foresaw, from the crowing of the barnyard cocks, that victory would be the Thebans' — because cocks that have been beaten are wont to be silent, while the victors crow. By means of hens, then, was Jupiter giving a sign to so great a state? And do those birds not crow except when they have won? But on that occasion they were crowing, and they had not won. That, you will say, is precisely the portent. A great one, to be sure — as if it had been the fishes that crowed, and not the cocks! But what time is there at which they do not crow, whether of night or of day? And if it is by liveliness, and a kind of gladness, so to speak, that the victors are roused to crow, some other gladness too could have come about, by which they were moved to crowing.

\ciceroSection{57} Democritus, indeed, in the finest terms explains the cause why cocks crow before dawn: it is because, when the food has been driven down from the breast and distributed through the whole body and softened, they utter their song, sated now with rest; cocks which indeed, in the silence of the night, as Ennius says, \\
favor their ruddy throats, \\
and with song and applause press down their wings. \\
Since, then, this creature is so songful of its own accord, what came into Callisthenes's head, to say that the gods had given the cocks a sign for crowing, when either nature or chance could have brought it about?

\ciceroSection{58} It was reported to the Senate that it had rained blood, that the river Atratus too had flowed with blood, that the statues of the gods had sweated. Do you suppose that Thales or Anaxagoras or any natural philosopher would have believed these reports? For neither blood nor sweat comes save from a body. But both a certain discoloration, arising from some contact with the earth, can be very like blood, and moisture gliding on from outside — as we see on plastered walls in a south wind — seems to imitate sweat. And in war these things appear both more numerous and greater to men who are afraid; the same things are not so much remarked upon in peace. There is this besides: that in fear and danger, while such things are the more readily believed, they are also the more freely invented with impunity.

\ciceroSection{59} But are we so frivolous and unconsidering that, if mice have gnawed something — gnawing being their one and only business — we should think it a monstrous sign? Before the Marsic War, because mice had gnawed the shields at Lanuvium, as you have told it, the haruspices said that this was a very great portent — as if, in truth, it made any difference at all whether mice, gnawing something day and night, had gnawed away shields or sieves! For if we are to follow such things, then because mice lately gnawed away my Plato's \textit{Republic}, I ought to have been mightily afraid for the commonwealth; or, if it had been Epicurus's book \textit{On Pleasure} that was gnawed, I should suppose the price of provisions in the market was going to rise.

\ciceroSection{60} Or do those other things really terrify us, whenever some monstrous creatures are said to have been born, either from a beast or from a human being? For all of which — not to be too long-winded — there is one single account. For whatever comes into being, of whatever kind it is, must necessarily have a cause from nature, so that, even if it has come into existence contrary to custom, it nevertheless cannot come into existence contrary to nature. Investigate the cause, therefore, in a novel and wonderful matter, if you can; and if you find none, hold this nevertheless as ascertained, that nothing could have come about without a cause, and drive off that terror which the newness of the thing has brought you by the reasoning of nature. So neither a rumbling of the earth, nor a parting of the sky, nor a rain of stones or of blood, nor the shooting of a star, nor torches seen in the sky shall terrify you.

\ciceroSection{61} If I were to seek from Chrysippus the causes of all these things, that very champion of divination would never call them the work of chance; he would render a natural account of them all. For nothing can come about without a cause; nor does anything happen that cannot happen; nor, if a thing has happened that could happen, ought it to be reckoned a portent. There are, then, no portents at all. For if what rarely happens is to be thought a portent, then a wise man is a portent; for I suppose a mule has given birth more often than a wise man has existed. The argument, therefore, comes to this conclusion: that nothing which could not happen has ever happened, and that nothing which could happen is a portent;

\ciceroSection{62} so that there is no portent whatsoever. This too a certain interpreter and reader of portents is said to have answered, not without wit, to a man who had once brought before him as a kind of showing the fact that a snake at his house had coiled itself around a beam: "It would have been a showing," he said, "if the beam had coiled itself around the snake." By this reply he plainly enough declared that nothing which can happen is to be held a showing. Gaius Gracchus wrote to Marcus Pomponius that, when two snakes had been caught in the house, the haruspices were summoned by his father. Why snakes rather than lizards, rather than mice? Because these last are everyday creatures, and snakes are not — as if it mattered, when a thing can happen, how often it does happen. Yet I do wonder this: if the release of the female snake was bringing death to Tiberius Gracchus, while the release of the male snake was deadly to Cornelia, why he released either of the two; for he writes nothing of what the haruspices answered would come to pass if neither snake were released. But death overtook Gracchus. From some cause, I suppose, of a graver illness, not from the release of a serpent; for the ill luck of the haruspices is not so great that what they have said would come to pass never happens even by chance.

\ciceroSection{63} For I should marvel at that other thing, if I believed it — what you said in Homer, that Calchas augured the years of the Trojan War from the number of the sparrows; of his conjecture Agamemnon speaks thus in Homer, as I in my leisure have rendered it: \\
Endure, men, and bear with steadfast spirit your hard labors, \\
that we may come to know whether the prophecies of our augur Calchas \\
hold true or whether the promptings of his breast were empty. \\
For all of them keep that portent in remembering mind, \\
all who were not robbed of the light by deadly fates. \\
When first Aulis was clothed with the Argive fleets, \\
which were bearing ruin to Priam and destruction to Troy, \\
we, about the cold waters, at the smoking altars, \\
appeasing the gods' will with gold-decked bulls \\
beneath the shady plane, where a spring of water flows forth, \\
saw a dragon of monstrous shape and coil, \\
a terrible thing, dart from the altar at a stroke from Jupiter; \\
which seized the nestlings on the plane-tree's bough, \\
hedged in their covering of leaves; and when it had devoured eight of them, \\
the ninth, the mother, flew above with quivering cry; \\
her too the savage creature mangled, tearing her vitals with monstrous bite.

\ciceroSection{64} Him, when he had slain the birds so tender and their mother, \\
the same Saturnian father who had brought him into the light \\
hid away and shaped into the hard covering of stone. \\
But we, standing in fear, beheld the marvelous monstrous sign \\
turning about in the very midst of the gods' altars. \\
Then Calchas spoke these words with confident voice: \\
"Why are you suddenly stunned and benumbed, Achaeans? \\
The very Creator of the gods has given us these portents, \\
slow and far too late, but with everlasting fame and praise. \\
For as many birds as you see slaughtered by the foul tooth, \\
so many years of war shall we drain to the dregs at Troy; \\
which in the tenth shall fall, and shall sate the Achaeans with vengeance." \\
These things Calchas declared; and these things you now see brought to ripeness. \\
But what sort of augury is that, of years rather than of months or of days, drawn from sparrows?

\ciceroSection{65} And why does he draw his conjecture from the little sparrows, in which there was no monstrous sign, yet stay silent about the dragon, which is said — a thing that could not happen — to have been turned to stone? And lastly, what likeness has a sparrow to years? As for that snake which appeared to Sulla while he was sacrificing, I remember both: that Sulla, when he was about to lead out his army on campaign, sacrificed, and that the snake came forth from the altar, and that on that day the affair was brilliantly conducted — not by the haruspex's counsel, but by the commander's.

\ciceroSection{66} And these kinds of showings have nothing marvelous about them; once they have happened, they are then drawn into a conjecture by some interpretation — so that those grains of wheat heaped into the mouth of the boy Midas, or the bees you said had settled on the lips of the boy Plato, are not so much marvelous as prettily conjectured; and these very things might either have been false themselves, or the things that were predicted might have fallen out by chance. About Roscius himself, that story may indeed be false — that he was coiled about by a snake; but that there was a snake in his cradle is not so much a wonder, especially in the district of Solonium, where snakes are wont to traffic about the hearth. As for the haruspices' answering that nothing would be more illustrious, nothing more noble, than he, I marvel that the immortal gods should have shown a future actor his renown, and shown none to Africanus.

\ciceroSection{67} And then you have gathered up the Flaminian showings too: that he himself and his horse suddenly fell down — hardly a marvel, this; that the standard of the first hastatus could not be pulled up — perhaps the standard-bearer was pulling it up timidly, having planted it confidently. For what wonder did the horse of Dionysius bring, in that it emerged from a river and had bees in its mane? But because in a short time he began to reign, what had happened by chance took on the force of a showing. But at Sparta arms rang out in the temple of Hercules, and of the same god at Thebes the closed doors suddenly opened of themselves, and those shields that had been fixed up high were found upon the ground. Since none of these things could happen without some movement, what reason is there why we should say they were brought about by divine agency rather than by chance?

\ciceroSection{68} But on the head of the statue of Lysander at Delphi there sprang up a crown of rough weeds, and that suddenly. Is that so? Do you suppose the weed's crown sprang up before the seed had been sown? And the rough weed, I believe, came from the droppings of birds, not from human planting; and besides, whatever is on the head can look like a crown. As for what you said — that at the same time the golden stars of Castor and Pollux set up at Delphi fell down, and were nowhere found again — that seems rather the work of thieves than of the gods.

\ciceroSection{69} The wickedness of the ape at Dodona, however, committed to writing in the Greek histories, fills me with wonder. What could be less of a wonder than that that most monstrous beast overturned the urn and scattered the lots? And the historians declare that no showing more grievous than this befell the Spartans! Then there was that prediction of the men of Veii: that if the Alban lake overflowed and flowed down into the sea, Rome would perish; but if it were held back, Veii would — so the Alban water was drawn off to the benefit of the suburban farmland, not to the preservation of citadel and city. But a little later a voice was heard warning men to take care that Rome not be captured by the Gauls; and from this an altar was consecrated on the New Way to Aius the Speaker. What then? This Aius the Speaker, when no one knew him, both spoke and uttered words and from that got his name; but after he found a seat and an altar and a name, he fell dumb? The same may be said of Moneta: from whom, apart from the matter of the pregnant sow, of what have we ever been warned?

\ciceroSection{70} Enough, and more than enough, about showings; there remain the auspices, and those lots which are drawn — not the lots that are poured forth in prophesying, which we more truly call oracles, and about which we shall speak when we come to natural divination. There remains also the matter of the Chaldeans; but first let us look at the auspices. A hard place for an augur, this, to argue against. For a Marsian augur, perhaps; but for a Roman, the easiest. For we are not the kind of augurs who foretell the future by observation of birds and the rest of the signs. And yet I believe that Romulus, who founded the city after taking the auspices, held the opinion that there was a science of augury in foreseeing things — for antiquity went astray in many matters — a science which we now see altered, whether by use, or by learning, or by the lapse of time; but the custom is retained, the religion, the discipline, the augural law, the authority of the college, both for the opinion of the common people and for the great advantages of the state.

\ciceroSection{71} Nor in truth were Publius Claudius and Lucius Junius, the consuls, anything but deserving of every punishment, who sailed against the auspices; for the religion ought to have been obeyed, and the ancestral custom not so defiantly spurned. With justice, then, was the one condemned by the people's verdict, while the other took his own life. Flaminius did not obey the auspices, and so he perished with his army. But a year later Paulus did obey; did he fall any the less, with his army, in the battle of Cannae? For granting that there are auspices, which there are not, at least these that we use — whether by the tripudium or from the sky — are images of auspices, in no way auspices themselves. "Quintus Fabius, I wish you to attend me at the taking of the auspice"; he answers: "I have heard." Among our forefathers, a man skilled in the matter was called in for this; now, anyone you please. But it is necessary that the man understand what silence is; for that is what we call silence in the auspices — what is free of every flaw.

\ciceroSection{72} To understand this belongs to the accomplished augur; but that other man, who is called in for the auspice, when the one taking the auspices has given the command thus — "Declare it, if it shall seem to be silence" — neither looks up nor looks about him; he straightway answers that it seems to be silence. Then the first: "Declare it, if they shall feed." — "They are feeding." — What birds? Or where? "He has brought," he says, "chickens in a cage" — the man who from this very thing is named the chicken-keeper. These, then, are the birds that serve as Jupiter's messengers! Whether they feed or not, what does it matter? Nothing to the auspices; but because, when they feed, something must necessarily fall from the mouth and strike the earth — a "ground-striking" \textit{[Greek: terripavium]} it was called at first, then a "ground-stamping" \textit{[Greek: terripudium]}; and this last is now called a tripudium — when, then, a morsel has fallen from the chicken's mouth, then a tripudium solistimum is announced to the man taking the auspices.

\ciceroSection{73} Can this augury, then, hold anything divine, when it is so forced and squeezed out of the bird? That the most ancient augurs did not use it is shown by this: we have an old decree of the college that any bird whatever can give the tripudium. There would be an augury then, if only the bird were free to give it of itself; then that bird might be seen as the interpreter and attendant of Jupiter. But as it is, when the creature is shut up in a cage and half dead with starvation, if it pounces on a lump of mash and something drops from its mouth — do you take this for an augury, and suppose that Romulus was in the habit of taking the auspices in this fashion?

\ciceroSection{74} Again, do you not think that those who took the auspices used once to watch the sky themselves? Now they give the order to the keeper of the chickens, and he reports back. Lightning on the left we hold to be the best of auguries for all matters except elections — and that exception was established for the sake of the state, so that the leading men of the citizenry should be the interpreters of the auspices, whether in the people's trials, or in the enacting of laws, or in the creation of magistrates. But on the strength of a letter from Tiberius Gracchus, the consuls Scipio and Figulus, when the augurs had judged that they had been elected with a flaw in the proceedings, laid down their office. Who denies that augury is a discipline? It is divination that I deny. But the haruspices are diviners — they whom Tiberius Gracchus, on account of the sudden death of the man who had collapsed without warning while reporting the result of the prerogative century, brought before the Senate, and who declared that the officer who put the question had not been lawful.

\ciceroSection{75} First, take care that they did not pronounce against the man who had been the officer of the century; for it was he who had died, and they could say that without divination, by mere conjecture. Then perhaps it was by chance, which is by no means to be removed from this kind of case. For what could Etruscan haruspices know either about the tent's being correctly taken or about the law of the pomerium? For my part I agree rather with Gaius Marcellus than with Appius Claudius, both of whom were my colleagues, and I judge that the law of the augurs, although it was first established on the belief in divination, was nevertheless afterward preserved and retained for the sake of the state.

\ciceroSection{76} But on this point more elsewhere; for now, so much. Let us look at foreign auguries, which are not so much matters of art as of superstition. They use nearly all birds, we very few; some are unlucky to them, others to us. Deiotarus used to question me about the discipline of our augury, and I him about his. Immortal gods, how great the difference was! So great that some things were even contrary. And he always used those auguries of his, whereas we — except while we hold the auspices taken on behalf of the people — how much do we use ours? Our ancestors would not have war conducted except under the auspices; yet for how many years now have wars been conducted by proconsuls and propraetors

\ciceroSection{77} who hold no auspices at all! And so they neither cross rivers under the auspices nor take auspices by the tripudium. Where, then, is the divination by birds? Since wars are conducted by men who hold no auspices, it appears to have been retained for the affairs of the city and abolished for those of war. As for the auspice of the spear-points — which is wholly a military auspice — the famous Marcus Marcellus, five times consul, gave it up entirely, that same man being both a commander and an augur of the first rank. And indeed he used to say that, whenever he wished to carry through some affair without being hindered by the auspices, he was accustomed to make his journey in a closed litter. Like to this is the rule we augurs lay down, that to keep a yoke-auspice from occurring they should bid the beasts be unyoked.

\ciceroSection{78} What else is it to refuse to be warned by Jupiter than to contrive that either no auspice can come about, or, if it does come about, none can be seen? For that point is quite ridiculous — your saying that Deiotarus does not regret the auspices that were given him when he set out to join Pompey, because, having followed his loyalty, he discharged his duty toward the friendship of the Roman people; and that praise and glory were more precious to him than his kingdom and his possessions. I believe it, indeed; but this has nothing to do with the auspices. For no crow could croak to him that he was acting rightly in preparing to defend the freedom of the Roman people; he felt that himself, as in fact he did feel.

\ciceroSection{79} Birds signify outcomes, whether adverse or favorable. I see that Deiotarus made use of the auspices of virtue, which forbid us to look to fortune so long as loyalty is upheld. But if the birds showed prosperous outcomes, then surely they deceived. He fled from the battle with Pompey — a grievous hour! He parted from him — a sorrowful thing! He saw Caesar at one and the same time enemy and guest — what is more bitter than that? And Caesar, when he had snatched from him the tetrarchy of the Trocmi and given it to some hanger-on of his from Pergamum, and had likewise stripped from him the Armenia granted by the Senate, and when he had been received by him with the most magnificent hospitality, left him plundered, both his host and a king. But I am wandering too far afield; I shall return to my theme. If we are after outcomes, which are what the birds are consulted for, in no way were they prosperous for Deiotarus; but if we are after duties, these were sought from his own virtue, not from the auspices.

\ciceroSection{80} Drop, then, the staff of Romulus, which you say could not be burned in that great fire; despise the whetstone of Attus Navius. There ought to be no place in philosophy for invented little tales. It was rather the part of a philosopher, with respect to the whole of augury, first to see its very nature, then its discovery, then its consistency. What, then, is the nature that brings it about that birds wandering this way and that, here and there, should signify something and now forbid action, now command it, whether by their song or by their flight? And why is it granted to some birds to make a valid auspice from the left, to others from the right? In what manner, again, or when, or by whom shall we say these things were discovered? The Etruscans at least have a boy ploughed up from the earth as the founder of their discipline; whom have we? Attus Navius? But Romulus and Remus were some years older, both of them augurs, as we have received the account. Or shall we say these things were the discoveries of the Pisidians, or the Cilicians, or the Phrygians? Is it your pleasure, then, to have men devoid of humanity as the authorities for divinity?

\ciceroSection{81} But all kings, peoples, and nations use the auspices. As if anything were so very common as to have no wisdom at all, or as if you yourself, in forming a judgment, took your stand on the multitude! How few there are who deny that pleasure is a good — most even call it the highest good! Are the Stoics, then, frightened off from their opinion by the throng of these? Or in most matters does the multitude follow their authority? What wonder is it, then, if in the auspices and in the whole of divination feeble minds take in those superstitions and cannot discern the truth?

\ciceroSection{82} And what coherent, connected consistency is there among the augurs? In accordance with the usage of our augury Ennius said:

Then it thundered well on the left, the weather being clear.

But Homer's Ajax, complaining to Achilles of the ferocity of the Trojans, announces some such thing in this manner:

Jupiter sends forth prosperous things by these flashes on the right.

So to us things on the left appear good, while to Greeks and barbarians things on the right appear the better. And yet I am well aware that the things which are good we call "left-hand," even if they are on the right; but certainly our people named the favorable thing "left" and foreigners named it "right," because for the most part it seemed the better.

\ciceroSection{83} How great a disagreement this is! And again — that they use some birds, others use others, that they observe in different ways, that different birds give different responses — is it not necessary to admit that part of all this was taken up by error, part by superstition, much of it by deception? And to these superstitions you did not hesitate to add omens as well: that Aemilia, by saying Perseus had perished, was an omen for Paulus, an omen her father accepted; that Caecilia, saying she handed over her own seat to her sister's daughter. Then those phrases: "Hold your tongues in good favor," and the prerogative century, the omen of the elections. This is the very thing — to be rich and eloquent against oneself. For when, in observing such things, will you be able to be calm and free in mind, so that in carrying out an affair you have not superstition but reason as your guide? Is it so? If someone says something out of his own affairs and his own talk, and some word of his falls aptly upon what you are doing or thinking, shall that thing bring you either fear or eagerness?

\ciceroSection{84} When Marcus Crassus was loading his army at Brundisium, a man in the harbor selling figs brought from Caunus kept crying out "Cauneans!" Let us say, if you please, that Crassus was warned by him to take care not to go — that he would not have perished, had he heeded the omen. But if we are to take up such things, then a stumble of the foot, the snapping of a shoe-strap, and sneezes will all have to be observed.

\ciceroSection{85} The lots remain, and the Chaldaeans, before we come to seers and to dreams. You think, then, that something must be said about lots? But what is a lot? Much the same thing as casting fingers, as throwing knucklebones, as throwing dice — matters in which recklessness and chance, not reason or deliberation, hold sway. The whole business was contrived by frauds, with an eye to gain, or to superstition, or to error. And just as we did in the case of haruspicy, so let us look at the reputed discovery of the most famous lots. The records of the people of Praeneste declare that one Numerius Suffustius, an honorable man and of good family, was driven by frequent dreams — at the last even threatening ones — being bidden to cut into a flint in a fixed spot, and that, terrified by the visions, while his fellow citizens laughed at him, he set about the task; and so, when the rock was broken through, the lots leapt out, carved in oak, in the characters of an ancient script. That place stands today, fenced about with religious awe, near the statue of the boy Jupiter, who, still at the breast, sits with Juno in the lap of Fortune reaching for her teat, and is worshipped most chastely by mothers.

\ciceroSection{86} At the same time, in that spot where the temple of Fortune now stands, they say that honey flowed from an olive tree, and that the haruspices declared those lots would be of the highest renown, and that at their command a chest was made from that olive, and the lots placed within it — the lots that to this day are drawn out at Fortune's prompting. What, then, can there be of certainty in these, which at Fortune's prompting are shuffled and drawn by a boy's hand? And how, moreover, were those lots laid in that place? Who cut that oak, hewed it, inscribed it? Nothing, they say, that a god cannot bring about. Would that he had brought it about that the Stoics be wise, so that they should not believe everything with superstitious anxiety and misery! But this kind of divination, at least, common life has now hooted off the stage; the beauty and antiquity of the shrine still keeps alive the name of the Praenestine lots — and that only among the crowd.

\ciceroSection{87} For what magistrate, or what man of any distinction, makes use of lots? In other places, indeed, the lots have plainly grown cold. And so Clitomachus writes that Carneades used to say he had nowhere seen Fortune more fortunate than at Praeneste. Let us drop, then, this kind of divination. Let us come to the prodigies of the Chaldaeans; about whom Eudoxus, the pupil of Plato, by the judgment of the most learned men easily first in astrology, holds this opinion — which he left set down in writing — that in the matter of prediction and the marking out of each man's life from his birthday, the Chaldaeans are least of all to be believed.

\ciceroSection{88} Panaetius too, who alone among the Stoics rejected the predictions of the astrologers, names Anchialus and Cassander, the foremost astrologers of the age in which he himself lived, as men who, though they excelled in the other branches of astrology, did not employ this kind of prediction. Scylax of Halicarnassus, an intimate of Panaetius, outstanding in astrology and likewise a leading man in the governance of his own city, repudiated this whole Chaldaean manner of foretelling.

\ciceroSection{89} But to use reasoning and set the witnesses aside, here is how those argue who defend these Chaldaean predictions from the day of birth. They say there is a certain force in the sign-bearing circle, which in Greek is called the \textit{[Greek: zōdiakos]}, of such a kind that each several part of that circle moves and alters the heavens in a different way, according as the stars stand at any given time in each of those parts and the parts adjacent to them; and that this force is variously stirred by those heavenly bodies which are called the wandering ones; and that when they have come into that very part of the circle in which lies the rising of the one being born, or into a part that holds something conjoined or in concord with it, they call those positions "triangles" and "squares." For since, with the approach and recession of the stars, such great turnings and changes of the seasons of the year and of the weather of the sky take place, and since by the force of the sun the things we see are brought about, they hold it not only resembling the truth but even true that, just as the air is tempered, so are children animated and formed as they come into the world, and that from this their talents, their characters, their mind, their body, the conduct of their life, and the chances and outcomes of each are shaped.

\ciceroSection{90} What incredible raving! For not every error is to be called folly. To these men even Diogenes the Stoic concedes something — that they can predict, but only this far, of what nature each man will be and to what pursuit each will be best suited; the rest of what they profess, he denies can in any way be known. For the shapes of twins are alike, while their life and fortune are for the most part unalike. Procles and Eurysthenes, kings of the Lacedaemonians, were twin brothers.

\ciceroSection{91} Yet they did not live the same number of years; for the life of Procles was the shorter, and he far excelled his brother in the glory of his deeds. But that very thing which the excellent Diogenes concedes to the Chaldaeans, by a kind of collusion, I deny can be understood at all. For since, as they themselves say, the rising of those who are born is governed by the moon, and the Chaldaeans observe and mark out as the natal stars whatever stars seem joined to the moon, they judge by the most deceptive sense of the eyes things that they ought to have seen by reason and by the mind. For the reasoning of the mathematicians teaches — and they ought to have known it — at how low a level the moon is borne, almost grazing the earth; how far it is from the nearest star, that of Mercury; how much farther from that of Venus; then by yet another interval how far it stands from the sun, by whose light it is thought to be illumined; while the remaining three intervals are infinite and immense — from the sun to the star of Mars, thence to that of Jupiter, from that to the star of Saturn, and thence to the heaven itself, which is the outermost and uttermost limit of the world.

\ciceroSection{92} What contagion, then, can reach from an almost infinite interval down to the moon, or rather to the earth? Again, when they say — as they must say — that all the births of all men, whoever are begotten in all the inhabited earth, are the same, and that the same things must necessarily befall all who are born under the same disposition of sky and stars, are they not such that it appears these interpreters of the heavens do not know even the nature of the heavens? For since those circles which divide the sky as if down the middle and bound our field of vision — which by the Greeks are called \textit{[Greek: horizontes]}, and may most rightly be called by us the "bounding" circles — have the greatest variety and stand differently in different places, the risings and settings of the heavenly bodies must necessarily not take place at the same time for all.

\ciceroSection{93} But if by their force the sky is tempered now in this way, now in that, how can there be the same force at work upon those being born, when the difference of the sky is so great? In these regions which we inhabit, the Dog Star rises after the solstice, and indeed by some several days; but among the Troglodytes, as is written, before the solstice — so that, even if we now grant that some celestial force does reach to those begotten on earth, it must be confessed to them that those who are born at the same time can fall into unlike natures on account of the unlikeness of the sky; which is by no means to their liking, for they wish all who are born anywhere, having risen at the same time, to be born under the same condition.

\ciceroSection{94} But what madness is so great as to suppose that, amid the greatest motions and changes of the sky, it makes no difference what wind, what rain, what weather there is in any place? — when of these things there are often such great unlikenesses in neighboring places that frequently one storm falls at Tusculum, another at Rome. Those who sail mark this above all, since in rounding the headlands they often feel the greatest changes of the winds. Since, then, there is now this calm, now that disturbance of the sky, is it the part of men in their senses to say that this — which assuredly does not bear upon it — does not bear upon the births of those being born, and yet to say that that something or other, thin and tenuous, which can in no way be felt and can scarcely even be understood — whatever tempering comes from the moon and the rest of the heavenly bodies — does bear upon the births of children? And again: that they do not understand the force of seeds — which counts for most in begetting and procreating — is utterly abolished, is this a middling error? For who does not see that children copy from their parents both their forms and their characters and most of their bearing and movements? — which would not happen if these things were brought about, not by the force and nature of those who beget, but by the tempering of the moon and the governance of the sky.

\ciceroSection{95} And again: that those born at one and the same point of time have unlike natures and lives and fortunes — does this not declare plainly enough that the time of birth has no bearing on the leading of one's life? Unless perhaps we suppose that no one was conceived and born at the very same time as Africanus. Was there ever, then, any man like him?

\ciceroSection{96} And again: is there any doubt that many men, though they were so born as to have certain parts deformed against nature, were restored and corrected by nature, when she had recalled herself, or else by art and medicine? — as those whose tongues so clung that they could not speak were freed by the cutting of the scalpel. Many too have removed a fault of nature by practice and exercise, as the Phalerean writes of Demosthenes, who, when he could not pronounce the letter r, brought it about by exercise that he pronounced it most distinctly. But if these things had been engendered and handed over by a star, no power could change them. And again: does not the unlikeness of places bring with it an unlikeness in the procreation of men? It is easy enough to run over in speech how far Indians and Persians, Ethiopians and Syrians differ in body and in mind, so that the variety and unlikeness is past believing.

\ciceroSection{97} From which it is understood that the situation of lands has more force toward birth than the touch of the moon. For when they say that the Babylonians spent four hundred and seventy thousand years in testing and trying out children, whoever happened to be born, they deceive us; for if it had once been the regular practice, it would not have stopped — yet we have no authority who says it is done or knows it was done. Do you see that I am saying not the things that Carneades said, but the things that Panaetius, the foremost of the Stoics, said? But I also ask this: were all who fell in the battle of Cannae under one star? Their end, at any rate, was one and the same. And what of those singular in talent and spirit — were they too under one star? For what time is there in which countless men are not born? And yet surely no one is like Homer.

\ciceroSection{98} And if it bears on the matter under what disposition of the heaven and arrangement of the stars each living thing comes into being, this must necessarily hold not only among men but among beasts as well — than which what more absurd can be said? Lucius Tarutius of Firmum, our friend, learned above all in the Chaldean calculations, even traced back the birthday of our own city to that festival of the Parilia on which we have received that it was founded by Romulus, and he said that Rome was born when the moon was in the Balance, and did not hesitate to sing her fates.

\ciceroSection{99} O the mighty force of error! Did even a city's birthday have any bearing on the force of the stars and the moon? Grant that in a child it matters from what disposition of the heaven he drew his first breath — could this have any force in the brick or the mortar out of which the city was made? But why say more? Every day they are refuted. How many things spoken by the Chaldeans to Pompey, how many to Crassus, how many to this very Caesar I remember — that not one of them would die except in old age, except at home, except in glory! So that it seems to me most strange that anyone still exists who even now believes those whose predictions he sees refuted every day by the event and the outcome.

\ciceroSection{100} There remain two kinds of divination which we are said to have from nature, not from art: the kind by prophesying and the kind by dreaming; of these, Quintus, let us, if you please, hold our discussion. Indeed I am willing, said he; for I assent wholly to what you have argued thus far, and, to speak truly, although your discourse has confirmed me, still even of my own accord I judged the Stoics' view of divination too superstitious; this reasoning of the Peripatetics moved me more — of the old Dicaearchus and of him who now flourishes, Cratippus — who hold that there is in the minds of men a kind of oracle, from which they have a presentiment of things to come, if the mind is either stirred by divine frenzy or, relaxed in sleep, moved freely and unbound. About these kinds I should very much like to hear what you think and by what reasonings you weaken them.

\ciceroSection{101} When he had said this, I then began again to speak, as if from a fresh starting point. I am not unaware, Quintus, said I, that you have always felt thus — that you doubted the other kinds of divination, but approved those two, of frenzy and of dreaming, which seemed to flow from a free mind. I shall say, then, what I think about those two kinds themselves, when first I have seen what force there is in the Stoics' conclusion of reasoning and in our friend Cratippus. For you said that Chrysippus and Diogenes and Antipater conclude in this manner: If there are gods and they do not declare to men beforehand what things are to come, then either they do not love men, or they are ignorant of what will come to pass, or they think it is no concern of men's to know what will be, or they do not judge it consistent with their majesty to signify beforehand to men what is to come, or even the gods themselves cannot signify those things; but they neither fail to love us (for they are beneficent and friendly to the race of men) nor are ignorant of those things which were established and appointed by themselves;

\ciceroSection{102} nor is it no concern of ours to know the things that are to come (for we shall be more wary if we know); nor do they reckon this foreign to their majesty (for nothing is more excellent than beneficence); nor are they unable to foreknow things to come; the gods, therefore, do not exist, nor do they signify things to come to us; but the gods do exist; therefore they signify; and it is not the case that, if they signify things to come, they give us no ways to the knowledge of their significations (for they would signify in vain); nor, if they give the ways, is there no divination; therefore there is divination.

\ciceroSection{103} O the sharp men! How few the words by which they think the business is finished! They take up, to draw their conclusion, things of which nothing is granted to them. But a conclusion of reasoning is to be approved in which that which is doubted is brought out from things not in doubt. Do you see Epicurus, whom the Stoics are wont to call dull and unschooled — how he concluded that what we say is the whole of nature is infinite? What is finite, he says, has an extremity. Who would not grant this? But what has an extremity is discerned from something else outside it. This too must be conceded. But that which is the whole is not discerned from something else outside it. Not even this can be denied. Since, then, it has no extremity, it must necessarily be infinite.

\ciceroSection{104} Do you see how he arrived at a doubtful matter from things conceded? This you dialecticians do not do; not only do you not take up to draw your conclusion the things that are conceded by all, but you take up things which, even when conceded, do nothing the more to bring about what you wish. For first you take up this: If there are gods, they are beneficent toward men. Who will grant you this? Epicurus? — who denies that the gods care for anything, either another's affairs or their own. Or our own Ennius? — who speaks, with great applause and the people assenting:
I have always said the race of the gods of heaven exists, and shall say so, \\
But I think they do not care what the human race may do.
And indeed he gives the reason why he holds this opinion; but there is no need to say what follows; it is enough to understand only this — that these men take for certain what is doubtful and disputed.

\ciceroSection{105} It follows next that the gods are ignorant of nothing, since all things were established by them. But here how great is the contest of most learned men denying that these things were established by the immortal gods! But, you say, it is our concern to know the things that are going to come to pass. There is a great book of Dicaearchus to the effect that it is better not to know these things than to know them. They deny that this is foreign to the majesty of the gods. As if, forsooth, the gods peered into everyone's hovel to see what would profit each man.

\ciceroSection{106} ``Nor are they unable to foreknow things to come.'' Those deny that they can, to whom it is not acceptable that what is to be is certain. Do you see, then, that the things which are doubtful are taken up as certain and conceded? Then they twist about and conclude thus: Therefore it is not both the case that the gods exist and that they do not signify things to come; for this they now think established. Then they assume: But the gods do exist — which itself is not conceded by all. Therefore they signify. Not even this follows; for they can fail to signify and yet be gods. Nor, if they signify, do they fail to give some ways to the knowledge of the signification. But this too is possible — that they do not give them to man, yet have them themselves; for why should they give them to the Etruscans rather than to the Romans? Nor, if they give the ways, is there no divination. Grant that the gods give them, which is absurd; what does it matter, if we cannot receive them? The last point is: Therefore there is divination. Let it be the last; still it is not brought about; for from false premises, as we learned from them, the true cannot be brought about. The whole conclusion, therefore, lies in ruins.

\ciceroSection{107} Let us now come to that excellent man, our friend Cratippus. If, says he, without eyes the function and office of eyes cannot exist, yet eyes can sometimes fail to perform their office, then whoever has even once so used his eyes that he discerned true things, that man has the sense of eyes that discern truly. Likewise, then, if without divination the function and office of divination cannot exist, yet when someone has divination he can sometimes err and not discern truly, it is enough, to confirm divination, that something has even once been so divined that nothing seems to have fallen out by chance; and of that kind there are countless cases; therefore it must be confessed that divination exists. Wittily and briefly; but since he twice took up what he wished, even if he should find us ready to concede, still that which he assumes can in no way be conceded.

\ciceroSection{108} If, he says, the eyes sometimes err, still, because they have at some time seen rightly, there is in them the power of seeing; likewise, if anyone has once made something out in divination, he, even when he errs, is still to be reckoned to have the power of divining. Consider, I beg you, our Cratippus, how alike these things are; for to me they do not seem alike. For the eyes, in discerning true things, use nature and sense, whereas minds, if ever they have seen true things either by prophesying or by dreaming, have used fortune and chance; unless perhaps you think those will concede to you, who hold dreams to be no more than dreams, that if ever some dream has turned out true, this did not happen by chance. But let us grant you those two premises (the things the dialecticians call lemmata, though we prefer to speak Latin); still the assumption (which the same men call the prosl\=epsis) will not be granted.

\ciceroSection{109} But Cratippus assumes his minor premise in this fashion: There are, however, countless presentiments that are not fortuitous. And I say there are none — see how great the dispute is; and once the minor premise is not granted, there is no conclusion. But we are shameless, you say, in not granting it, when the thing is so plain. What is plain? That many turn out true, he says. What of the fact that far more turn out false? Does not that very inconstancy, which is the property of fortune, teach that fortune, not nature, is the cause? Then, Cratippus, if that conclusion of yours is true — for it is with you that I have to do — do you not see that the same reasoning can be used by the haruspices too, and by the readers of lightning, and the interpreters of portents, and the augurs, and the casters of lots, and the Chaldeans? Of which kinds there is not one out of which something has not turned out just as it was predicted. Therefore either those kinds of divination too exist — the very ones you most rightly reject — or, if they do not exist, I do not see why these two should exist that you leave standing. So, by the very reasoning by which you bring these in, those others can exist that you do away with.

\ciceroSection{110} And what authority does that frenzy carry, which you call divine, that the madman should see what the wise man does not, and that the one who has lost his human senses should attain the senses of the gods? We pay heed to the verses of the Sibyl, which she is said to have poured out in her frenzy. Of these the interpreter was thought, lately, on some false rumor among men, to be going to declare in the Senate that the man whom in plain fact we had as king must also be called king, if we wished to be safe. If this is in the books, to what man does it point, and to what time? For cleverly did the one who composed them bring it about that, whatever should happen, it should seem to have been predicted, since all definition of men and times had been removed.

\ciceroSection{111} He also brought in a screen of obscurity, so that the same verses might seem capable of being fitted now to one matter, now to another. That this song is not the work of one in a frenzy is shown both by the poem itself — for it is more a thing of art and care than of impulse and excitement — and above all by what is called the acrostic \textit{[Greek: akrostichis]}, when in order, out of the first letters of each successive verse, some sense is woven together, as in certain lines of Ennius: \emph{Quintus Ennius made this}. That, surely, is more the mark of an attentive mind than of a frenzied one.

\ciceroSection{112} And in the Sibylline verses the whole sense of each utterance is spelled out from the first letters of the first verse of that utterance. This is the work of a writer, not of a man in a frenzy — of one applying care, not of a madman. For which reason let us keep the Sibyl set apart and shut away, so that — as has been handed down from our ancestors — the books may not even be read without the order of the Senate, and may serve rather for the laying aside of religious scruples than for the taking of them up; and let us deal with the priests in charge, that they bring forth from those books anything whatever rather than a king, whom hereafter neither the gods nor men will suffer to be at Rome. But many, you say, have often prophesied true things, as Cassandra: \\ And now upon the great sea \dots \\ and, the same a little after: \\ Alas, behold \dots \\ Are you then forcing me to believe even the poets?

\ciceroSection{113} Let these things have what charm you will; let them be helped along by words, by sentiments, by measures, by melodies; yet we owe no authority, no credence, to invented matters. And in the same way I judge that we must believe neither some Publicius or other, nor the Marcian seers, nor the secret oracles of Apollo — of which some are openly fabricated, some rashly babbled out, and have never won the approval of anyone even of middling sense, much less of a man of judgment.

\ciceroSection{114} What? you will say — did not that rower from Coponius's fleet foretell the things that came to pass? He did indeed, and the very things that all of us at that time feared might happen. For we heard that camp had been pitched against camp in Thessaly, and it seemed to us that Caesar's army had both more audacity, seeing it had brought war upon its own country, and more strength on account of long service; but as for the issue of the battle, there was not one of us who did not fear it — though, as befitted steady men, not openly. But that Greek — what wonder if, in the greatness of his fear, as commonly happens, he parted from his steadiness, from his mind, from his very self? And in that disturbance of soul he, deranged, declared would come to pass the very things that, when sane, he feared would come to pass. Which, I ask you, by gods and men, is more like the truth: that the crazed rower, or that some one of us who were there at the time — myself, Cato, Varro, Coponius himself — could discern the counsels of the immortal gods?

\ciceroSection{115} But now I come to you: \\ O holy Apollo, you who hold fast \\ the fixed navel of the lands, \\ whence first the savage, superstitious, \\ wild voice came forth. \\ For with your oracles Chrysippus filled a whole volume — partly false, as I think; partly true by chance, as happens very often in every kind of utterance; partly twisting in their speech and obscure, so that the interpreter has need of an interpreter and the lot itself must be referred to lots; partly ambiguous, and such as must be carried before a dialectician. For when that lot was given out to the richest king of Asia: \\ Croesus, crossing the Halys, will overthrow a great might of wealth — \\ he supposed he would overthrow the might of his enemies; but it was his own that he overthrew.

\ciceroSection{116} Yet whichever of the two had happened, the oracle would have been true. But why should I ever believe this was given out to Croesus? Or why should I hold Herodotus more truthful than Ennius? Could the one have invented less about Croesus than Ennius about Pyrrhus? For who is there who would believe that this answer was given to Pyrrhus from the oracle of Apollo: \\ I say that you, descendant of Aeacus, the Romans can conquer? First, Apollo never spoke in Latin; next, that lot is unheard of among the Greeks; besides, by Pyrrhus's time Apollo had already given up making verses; and lastly, although the race of the Aeacidae was ever, as it stands in Ennius, \\ a dull-witted breed, \\ stronger in war than strong in wisdom — \\ still he could have understood this ambiguity of the verse: that "you the Romans can conquer" has no more force toward himself than toward the Romans. For that ambiguity which deceived Croesus could have fooled even Chrysippus; but this one not even Epicurus.

\ciceroSection{117} But — and this is the chief point — why are oracles no longer given out in that fashion at Delphi, not only in our own age but for a long time now, so that nothing can be more contemptible? At this point, when they are hard pressed, they say that the power of that place has vanished with age — the place from which came that breath of the earth, by which the Pythia, her mind stirred up, would give out oracles. You would think they were speaking of wine or pickled fish, which vanish with age. It is the power of a place that is in question, and a power not merely natural but even divine; and how, pray, did it vanish? With age, you will say. What age is there that can wear out a divine power? And what is so divine as a breath out of the earth so moving the mind that it makes it foresee things to come — so that it not only discerns them long before, but even pronounces them in number and verse? But when did that power vanish? Was it after men began to be less credulous?

\ciceroSection{118} Demosthenes indeed, who lived nearly three hundred years ago, even then said that the Pythia "philippized" \textit{[Greek: philippizein]} — that is, as it were, sided with Philip. And by this he meant that she had been corrupted by Philip; from which one may judge that in other Delphic oracles too there was something not sincere. But somehow these superstitious and almost fanatical philosophers seem to prefer anything whatever to admitting they have been fools. You would rather have it that the thing has vanished and been quenched — the thing that, if it had ever existed, would surely be eternal — than not believe what is not to be believed.

\ciceroSection{119} A like error is found in dreams; and how far back is the defense of these fetched! Our souls, they hold, are divine, and are drawn from a source outside, and the world is filled full with a multitude of souls in concord with one another; and so, by this divinity of the mind itself and its conjunction with minds outside, what is to come is discerned. The soul, Zeno thinks, is contracted, and as it were slips and sinks down — and this very thing is to be asleep. And further, Pythagoras and Plato, weightiest of authorities, that we may see surer things in our sleep, bid us go to our rest prepared by a certain regimen of diet — the Pythagoreans, indeed, abstaining from beans at all costs, as if it were the mind, and not the belly, that is bloated by that food. But somehow nothing can be said so absurd that it is not said by some one of the philosophers.

\ciceroSection{120} Do we think, then, that the souls of sleepers are moved of themselves in their dreaming, or, as Democritus thinks, are struck by a vision that comes from without and from elsewhere? For whether it is in this way or in that, very many false things can appear as true to those who dream. For to men sailing, the things that stand still seem to move; and by a certain fixing of the eyes, the one light of a lamp seems two. Why should I speak of the madmen, of the drunk — how many false things appear to them? But if such visions are not to be believed, why dreams should be believed I do not know. For one may dispute about these errors, if you like, just as about dreams: so that, if the things that stand still should seem to move, you might say it signifies an earthquake or some sudden flight; while by the doubled light of the lamp is declared that discord and sedition are afoot. And further, from the visions of the mad or the drunk countless inferences can be drawn by conjecture,

\ciceroSection{121} \dots which seem to be going to come to pass. For who is there who, hurling the javelin all day long, will not sometimes hit the mark? We dream whole nights through, and there is scarcely one on which we do not sleep — and then we marvel when, once in a while, what we dreamed turns out true? What is so uncertain as the throw of the dice? Yet there is no one who, throwing often, does not sometimes throw a Venus, and now and then even twice and three times over. Are we, then, like fools, to prefer to say that this came about by the impulse of Venus rather than by chance? But if at all other times false appearances are not to be believed, I do not see what special property sleep has, in which falsehoods should prevail in place of truths.

\ciceroSection{122} But if nature had been so arranged that sleepers should act out the things they dreamed, then all who went to bed would have to be bound; for as dreamers they would make wilder motions than any madman. And if the appearances of madmen are not to be trusted, because they are false, why one should trust the appearances of dreamers, which are even far more disordered, I do not understand. Or is it because the mad do not report their appearances to an interpreter, whereas those who have dreamed do report them? I ask further: if I should wish to write something, or to read, or to sing with the voice or on the lyre, or to set out something geometrical or physical or dialectical, must I wait for a dream, or must the art be employed — the art without which none of these things can be done or carried through? And further still, not even if I wished to sail would I steer the ship according to what I had dreamed; for the penalty would be immediate.

\ciceroSection{123} How, then, is it fitting that the sick should seek their cure from an interpreter of dreams rather than from a physician? Can Aesculapius, can Serapis prescribe for us a treatment for our health through sleep, while Neptune cannot do so for helmsmen? And if Minerva will give medicine without a physician, will the Muses not give to dreamers the knowledge of writing, of reading, of the other arts? But if a treatment for health were given, these other things I have spoken of would be given too; and since they are not given, medicine is not given either; and once that is taken away, all the authority of dreams is taken away.

\ciceroSection{124} But let these points too be reckoned obvious; now let us look at the deeper matters. For either some divine power, taking thought for us, makes the significations of dreams, or interpreters understand — out of a certain agreement and conjunction of nature, which they call \textit{[Greek: sympatheian]} — what suits each thing from dreams and what follows upon each thing; or neither of these holds, but there is a certain constant and long-continued observation of what tends to come about and what tends to follow, when such-and-such has been seen in sleep. First, then, it must be understood that there is no divine power that produces dreams. And this at least is plain: that no appearances of dreams proceed from the will of the gods; for the gods would do this for our sake, that we might be able to foresee things to come.

\ciceroSection{125} How few there are, then, who obey their dreams, who understand them, who remember them! And how many, on the other hand, who despise them and reckon that superstition the mark of a feeble and old-womanish mind! What reason is there, then, why a god, taking thought for these men, should warn through dreams those who judge such things worthy of no care — nor even of remembrance? For a god cannot be ignorant of what mind each man is of; nor is it worthy of a god to do anything in vain and without cause, which is at odds even with the consistency of a human being. So, if most dreams are either not known or neglected, then either the god does not know this, or he uses the signification of dreams in vain; and neither of these befits a god. We must therefore admit that nothing is signified by a god through dreams.

\ciceroSection{126} This too I ask: why, if a god gives us these appearances for the sake of our foreseeing, he does not rather give them to the waking than to the sleeping. For whether some external and adventitious impulse stirs the minds of sleepers, or the minds move of themselves, or whatever other cause it is that makes us seem in our sleep to see, to hear, to do something — that same cause could have been present in the waking too; and if the gods did this for our sake in our sleep, they would do the same for the waking — especially since Chrysippus, refuting the Academics, says that the things which appear to the waking are by far clearer and more certain than those which appear to dreamers. It was therefore more worthy of divine beneficence, when they were taking thought for us, to give clearer appearances to the waking man than dimmer ones through sleep. And since this is not done, dreams are not to be thought divine.

\ciceroSection{127} And indeed, what need is there of circuit and roundabout, that one should have to use interpreters of dreams, rather than that a god, if he was truly taking thought for us, should say outright, "Do this, do not do this," and give that appearance to the waking rather than to the sleeping? Then again, who would dare to say that all dreams are true?
\begin{quote}
Some dreams are true, \\
but not all of them need be.
\end{quote}
says Ennius. What, pray, is that distinction — what marks the true, what the false? And if the true are sent by a god, where do the false come from? For if these too are divine, what is more inconsistent than a god? And what is more witless than to rouse the minds of mortals with false and lying appearances? But if the true appearances are divine, while the false and empty ones are human, what is that license of assignment, that a god should make this one and nature that one, rather than that a god should make all of them — which you deny — or nature all of them? And since you deny the former, the latter must necessarily be confessed.

\ciceroSection{128} By "nature" I mean that by which the mind can never be empty of agitation and motion, holding still. When, through the languor of the body, it can use neither the limbs nor the senses, it falls upon various and uncertain appearances out of the remnants — as Aristotle says — that cling, of those things which the man did or thought about while awake; and from the disturbance of these there arise at times marvelous shapes of dreams. And if some of these are false, others true, by what mark they are to be told apart I should very much like to know. If there is none, why are we to listen to those interpreters? But if there is some such mark, I am eager to hear what it is — yet they will be stuck.

\ciceroSection{129} For now it comes to a contest: which is the more probable — that the immortal gods, surpassing in the excellence of all things, run about around the couches, and indeed the truckle-beds, of all mortals wherever they are, and, when they have seen someone snoring, cast upon him certain twisted and obscure appearances, which he, terrified out of sleep, carries off in the morning to an interpreter — or that it comes about by nature, that the mind, nimbly agitated, seems in sleep to see what it had seen while awake? Which is more worthy of philosophy — to interpret these things by the superstition of hag-wives, or by the explanation of nature? So that, even if a true conjecture of dreams could now be made, still those who profess it could not make it; for they are drawn from the most worthless and unlettered sort. Yet your Stoics deny that anyone but the wise man can be divine.

\ciceroSection{130} Chrysippus, indeed, defines divination in these words: a power that recognizes and sees and sets forth the signs which are portended to men by the gods; and that its task is to know beforehand of what mind the gods are toward men, and what each thing signifies, and in what manner these things are to be attended to and expiated. And the same man defines the interpretation of dreams in this way: that it is a power discerning and explaining what is signified to men by the gods in their sleep. What then? For these things is there need of moderate good sense, or of both outstanding talent and perfect learning? But of such a sort we have known no one.

\ciceroSection{131} See, then, that even if I should grant you that divination exists — which I shall never do — we still could find no one who is divine. And of what sort is that mind of the gods, if they signify to us in sleep neither the things which we may understand of ourselves, nor the things for which we can have interpreters? For the gods are like men who would cast upon us things of which we have neither the knowledge nor an expounder — as if Carthaginians or Spaniards were to speak in our Senate without an interpreter.

\ciceroSection{132} And indeed, to what end are the obscurities and riddles of dreams? For the gods ought to have wished us to understand the things by which they would warn us for our own sake. What? Is no poet, no natural philosopher obscure?

\ciceroSection{133} Euphorion, to be sure, is excessively obscure; but Homer is not. Which, then, is the better? Heraclitus is very obscure, Democritus least of all. Are they, then, to be compared? You are advising me on my own account — about a thing I do not understand? Why, then, are you advising me at all? It is as if some physician should order a sick man to take "the earth-born, the grass-walker, the house-carrier, the bloodless," rather than say, in the way men speak, "a snail." For when Pacuvius's Amphion had said, more obscurely,

A four-foot beast, slow-paced, of the field, low-set, rough-coated, \\
small of head, with a serpent's neck, fierce of aspect, \\
disemboweled, lifeless, yet with a living sound —

then the Athenians answer: We do not understand, unless you put it plainly. And he, in a single word: A tortoise. Could you not, then, harpist, have said this from the start?

\ciceroSection{134} A certain man reports to an interpreter that he had dreamed an egg was hanging from the strap of his bedroom couch (this dream is in Chrysippus's book); the interpreter answered that a treasure lay buried under the bed. The man digs, finds a quantity of gold, and surrounded by silver; he sent the interpreter as much of the silver as seemed fit. Then the interpreter said: Nothing from the yolk? — for it seemed to him that the gold had been signified by the yolk of the egg, the silver by the rest. Has no one else, then, ever dreamed of an egg? Why, then, did this nobody-or-other alone find a treasure? How many poor men, worthy of the gods' protection, are admonished by no dream to the finding of a treasure! And for what reason was he so obscurely admonished, that a likeness of treasure should be hatched out of an egg, rather than be bidden plainly to seek a treasure, just as Simonides was plainly forbidden to sail?

\ciceroSection{135} Obscure dreams, therefore, are least of all consonant with the majesty of the gods. Let us come to the open and clear ones, such as the story of that man murdered by the innkeeper at Megara, such as that of Simonides, who was forbidden to sail by the man he had buried, such also as that of Alexander, which I am surprised you passed over, Quintus. When Ptolemy, his close friend, had been struck in battle by a poisoned weapon, and was dying of that wound in the greatest pain, Alexander, sitting beside him, was lulled to sleep. Then in his sleep there appeared to him, it is said, that serpent which his mother Olympias kept, carrying a little root in its mouth and at the same time saying in what place it grew (and it was not far from that spot), and that its power was so great that it would easily heal Ptolemy. When Alexander, having awakened, told his friends the dream, men were sent to seek that little root; once found, both Ptolemy is said to have been healed and many soldiers who had been wounded by the same kind of weapon.

\ciceroSection{136} Many dreams, too, have been brought forward by you out of the histories — of the mother of Phalaris, of the elder Cyrus, of the mother of Dionysius, of Hamilcar the Carthaginian, of Hannibal, of Publius Decius; and that famous one about the temple-leader is now widely known, and the dream of Gaius Gracchus too, and the recent one of Caecilia, daughter of Balearicus. But these are foreign and for that reason unknown to us, and some of them perhaps invented. For who is the authority for them? As for our own dreams, what have we to say? You, of me come up from the river and of my horse at the bank; I, of Marius bidding me be conducted into his memorial with the laurel-wreathed rods. There is, Quintus, one and the same account of all dreams; and that account, by the immortal gods! — let us see to it that it is not overborne by our superstition and corruption of judgment.

\ciceroSection{137} For what Marius do you suppose was seen by me? His semblance, I imagine, and his image, as it seems to Democritus. From where did the image set out? For he holds that images flow from solid bodies and from fixed shapes; what, then, was the body of Marius? From that, he says, which had been his body. So this image of Marius was pursuing me into the plain of Atina? — Everything is full of images; for no semblance can be conceived except by the impact of images.

\ciceroSection{138} What of it, then? Are these images so obedient to our command that, the moment we wish it, they come running? And even of those things which are nothing at all? For what shape is so unseen, so non-existent, that the mind cannot fashion it for itself? So that even things we have never seen we yet hold formed within us — the sites of towns, the figures of men.

\ciceroSection{139} Surely, then, when I think either of the walls of Babylon or of Homer's face, no image of them strikes me? Everything, then, that we wish can be known to us; for there is nothing about which we cannot think; no images, therefore, creep into the minds of sleepers from without, nor do any flow at all — and I have never known anyone who, with greater authority, said nothing. Of the soul such is the force and such the nature, that it is vigorous while waking by no incoming impact, but by its own motion of a certain incredible swiftness. When these souls are sustained by the limbs and the body and the senses, they discern, think, and feel all things more surely. But when these supports are withdrawn, and the soul is deserted by the languor of the body, then it is stirred of itself. And so within it both shapes move and actions, and many things seem to be heard, many to be said.

\ciceroSection{140} These things, of course, in a feeble and slackened soul, move about in every way confused and varied, and most of all the remnants are set in motion and agitated within our souls of those matters about which, while waking, we have either thought or busied ourselves — as in me, in those days, Marius dwelt much in my mind, when I recalled with what greatness of soul, with what steadfastness, he had borne that grievous fall of his. This, I believe, was the cause of dreaming about him. And in you, thinking of me with anxiety, I suddenly appeared as though risen from the river. For there were in the souls of both of us the traces of our waking thoughts. But certain things were added — as for me the detail of Marius's memorial, and for you that the horse on which I was riding, having gone down together with me, reappeared.

\ciceroSection{141} Or do you suppose any old woman would be so deranged as to believe in dreams, if these things did not now and then by chance, by accident, at random, fall together? To Alexander a serpent seemed to speak. It may well be that this is false, it may be true; but whichever it is, there is nothing marvelous in it; for he did not hear a serpent speak, but seemed to hear it — and indeed, to make it the greater, it spoke while holding the root in its mouth. But nothing is great for a dreamer. I ask, however, why Alexander had so brilliant a dream, so certain a one, and yet not this same man at other times, nor most men many dreams; I myself, at any rate, have had none worth speaking of besides this one about Marius, so far as I remember. So many nights, then, consumed to no purpose in so long a life!

\ciceroSection{142} Now indeed, on account of my leaving off the work of the Forum, I have cut out my late-night studies and added midday naps, which I was not in the habit of using before; and yet, sleeping so much, I have been admonished by no dream, least of all about such great matters; nor do I ever seem to myself to be dreaming more than when I see the magistrates in the Forum or the Senate in the senate-house. For indeed (this is the second head of the division) what continuity and conjunction of nature is there — which, as I said, they call fellow-feeling \textit{[Greek: sympatheian]} — of such a kind that a treasure ought to be understood out of an egg? For physicians understand the coming on and growth of diseases from certain things, and also certain signs of health, just as that very point — whether we are full or wasted — they say can be understood from a certain kind of dreams. But a treasure, an inheritance, an honor, a victory, and many things of that kind — by what natural kinship are they joined with dreams?

\ciceroSection{143} A certain man is said, when in his sleep he was joined in the embrace of Venus, to have voided gallstones. Here I see fellow-feeling; for such an object was presented to the sleeper that the force of nature, not a mistaken belief, brought about what came to pass. What nature, then, presented to Simonides that semblance by which he was forbidden to sail? Or what bond with nature had that dream of Alcibiades which is recorded? — who, a little before his death, seemed in his sleep to be wrapped in the cloak of his mistress. When he had been cast out unburied and lay deserted by all, his mistress covered his body with her own mantle. Was this, then, present in things to come, and did it have natural causes — or did chance bring it about both that it should be seen and that it should come to pass?

\ciceroSection{144} What of it? Do not the conjectures of the interpreters themselves declare their own ingenuity rather than any force or accord of nature? A runner, intending to set off for the Olympic games, seemed in his sleep to be carried in a four-horse chariot. In the morning he goes to an interpreter. And the man says: You will win; for that is what the swiftness and force of the horses signify. Afterward the same runner goes to Antiphon. But Antiphon says: You must necessarily be beaten; do you not see that four ran ahead of you? Behold another runner (and Chrysippus's book is full of such dreams and the like, and Antipater's full of them) — but I return to the runner: he reported to an interpreter that he had seemed in his sleep to be turned into an eagle. And the man says: You have won; for no bird flies more powerfully than that one. To this same man Antiphon says: Blockhead, do you not see that you have been beaten? For that bird, in pursuing and harrying other birds, is itself always the last.

\ciceroSection{145} A certain married woman, eager to bear a child, was in doubt whether she was pregnant, and seemed in her sleep to have her womb sealed up. She reported it. The interpreter denied that she could have conceived, since it had been sealed. But another said she was pregnant; for, he said, an empty thing is not in the habit of being sealed. What art is this of the interpreter, playing tricks with his own wit? Or do those things I have mentioned, and the countless others the Stoics have gathered together, signify anything beyond the sharpness of men who draw a conjecture from some likeness, now this way, now that? Physicians have certain signs from the veins and the breathing of a sick man, and from many other things they have a presentiment of what is to come; pilots, when they have seen the squids leaping or the dolphins flinging themselves into harbor, judge that a storm is being signaled. These things can readily be explained by reason and traced back to nature; but the things I spoke of a little before, in no way at all.

\ciceroSection{146} But, you say, long-continued observation — for this one part remains — has made an art by the noting of things. Tell me, then: can dreams be observed? In what manner? For their varieties are countless. Nothing can be conceived so preposterous, so disordered, so monstrous, that we cannot dream it; how, then, can we either embrace these endless and ever-new things in memory, or mark them down by observation? The astrologers have noted the motions of the wandering stars; for an order was found in those stars that had not been supposed. Tell me, then, what is the order, or what is the running-together, of dreams; and how can true dreams be distinguished from false? — since the same dreams turn out one way for some, another for others, and not always in the same way even for the same persons. So that it seems strange to me that, when we are not in the habit of believing a liar even when he speaks the truth, these men, if some one dream has come out true, do not rather, out of many, withdraw their trust from the one, than, out of one, confirm a countless multitude.

\ciceroSection{147} If, then, neither is god the maker of dreams, nor is there any fellowship of nature with dreams, nor could a science be found by observation, the result is that nothing whatever is to be attributed to dreams — especially since those very people who see them divine nothing; those who interpret them bring conjecture to bear, not nature; while chance, over almost countless ages, has in all matters produced more marvels than in the visions of dreams; and since nothing is more uncertain than conjecture, which can be drawn into various directions, and sometimes even into contrary ones.

\ciceroSection{148} Let this divination by dreams, then, be hissed off the stage along with the rest. For, to speak truly, superstition, spread abroad among the nations, has crushed the minds of almost all men and seized upon the weakness of mankind. This was said in those books on the nature of the gods, and it is the very thing I have most aimed at in this discussion. For I judged that I should do much good both to myself and to my countrymen, if I tore it up by the roots. Nor indeed — for I wish this to be carefully understood — is religion taken away when superstition is taken away. For it is the part of a wise man to guard the institutions of our ancestors by retaining the rites and ceremonies; and the beauty of the universe and the order of the heavenly things compel us to confess that there is some surpassing and eternal nature, and that the human race must look up to it and wonder at it.

\ciceroSection{149} For which reason, just as religion, which is joined with the knowledge of nature, must be propagated, so all the roots of superstition must be pulled out. For it presses and bears down upon you, and, wherever you turn, it pursues you — whether you have heard a seer or an omen, whether you have sacrificed or caught sight of a bird, whether you have seen a Chaldaean or a haruspex, whether it has lightened, whether it has thundered, whether anything has been struck from the sky, whether anything has been born or done resembling a portent; and some one of these things must for the most part come to pass, so that it is never permitted to rest with a quiet mind.

\ciceroSection{150} Sleep seems to be the refuge from all toils and anxieties. Yet from that very thing the most numerous cares and fears are born; which, indeed, would have less force in themselves and be more despised, had not the philosophers taken up the patronage of dreams — and those not the most contemptible of them, but in the first rank for sharpness, men who see consequence and contradiction, who are reckoned all but finished and perfect. Had Carneades not withstood their license, I am not sure they would not by now be judged the only philosophers. With them is nearly all our debate and contention, not because we hold them in the greatest contempt, but because they seem to defend their views most sharply and most prudently. But since it is proper to the Academy to interpose no judgment of its own, to approve those things that seem most like the truth, to set causes side by side and to bring out what can be said for each view, and, with no authority of its own brought to bear, to leave the judgment of its hearers whole and free — we shall hold to this practice handed down from Socrates, and between us, if it please you, Quintus my brother, we shall use it as often as we can. ``For my part,'' said he, ``nothing could be more welcome to me.'' When this had been said, we rose.
