% Laelius de Amicitia (laelius-de-amicitia) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Quintus Mucius the Augur used to tell many stories about his father-in-law Gaius Laelius, from precise memory and with great charm, and he never hesitated to call him "the Wise" in every conversation. As for me, once I had put on the toga of manhood my father had introduced me to Scaevola in such a way that, as far as I was able and was allowed, I never left the old man's side. And so I committed to memory the many things he argued with such good sense, and the many things, too, that he said briefly and aptly, and I was eager to grow more learned through his wisdom. When he died, I attached myself to Scaevola the Pontiff, the one man of our state whom I dare to call the most distinguished in both ability and justice. But of him another time; for now I return to the Augur.

\ciceroSection{2} Among the many occasions, I remember in particular a day when he was sitting at home in his alcove, as was his habit, and I was there along with a very few close friends, and he fell into that conversation which was then on nearly everyone's lips. For surely you remember, Atticus—all the more because you were so close to Publius Sulpicius—what admiration, or what complaint, ran among people when Sulpicius, as tribune of the plebs, fell into bitter, deadly quarrel with Quintus Pompeius, the consul of that year, with whom he had once lived in the closest and most loving friendship.

\ciceroSection{3} And so on that occasion Scaevola, having stumbled upon this very subject, set out for us the discourse on friendship that Laelius had delivered—to him, and to his other son-in-law, Gaius Fannius the son of Marcus—a few days after the death of Africanus. The substance of that discussion I committed to memory, and have set it out in this book as my own judgment directed; for I have brought the men themselves on, as if speaking, so that the constant "said I" and "said he" need not be inserted, and so that the conversation might seem to be held face to face, by men present before us.

\ciceroSection{4} Since you often pressed me to write something on friendship, the subject seemed to me worthy both of everyone's attention and of our own close bond; and so, at your request, I was glad to be of use to many. But as in the \textit{Cato the Elder}, which I wrote for you on old age, I brought Cato on as an old man holding forth—because no figure seemed more fitting to speak of that time of life than one who had been old for the longest while and had flourished in old age itself beyond all others—so here, since we had received from our fathers that the friendship of Gaius Laelius and Publius Scipio was the most memorable of all, Laelius seemed to me the fitting figure to discourse on that very friendship, the discourse Scaevola remembered him delivering. And this kind of conversation, resting on the authority of men of old and of illustrious men, seems somehow to carry more weight; so much so that, reading my own work, I am sometimes so affected that I imagine it is Cato speaking, not me.

\ciceroSection{5} But as then I wrote, an old man to an old man, about old age, so in this book I have written, a most devoted friend to a friend, about friendship. In that work Cato spoke, than whom scarcely anyone in those days was older, no one wiser; here it is Laelius—a wise man, for so he was reckoned, and outstanding in the glory of his friendship—who will speak about friendship. I would ask you to turn your mind away from me for a little while, and to imagine that Laelius himself is speaking. Gaius Fannius and Quintus Mucius come to their father-in-law after the death of Africanus; from them the conversation arises, Laelius answers, and his whole discourse is about friendship—reading which, you will recognize yourself.

\ciceroSection{6} "That is just so, Laelius," said Fannius. "There was no better man than Africanus, none more illustrious. But you must understand that everyone's eyes are turned upon you alone; it is you they both call and consider wise. Not long ago the same was granted to Marcus Cato; we know that in our fathers' day Lucius Acilius too was called wise—but each in a different way: Acilius because he was thought shrewd in civil law, Cato because he had experience in many matters, and many of his deeds, both in the Senate and in the Forum, were spoken of as wisely foreseen, or steadfastly carried through, or shrewdly answered. And so by old age he already bore 'the Wise' almost as a surname.

\ciceroSection{7} "You, however, are called wise in a somewhat different way: not only by nature and character, but by study and learning as well—and wise not as the crowd uses the word, but as the learned are accustomed to use it, of a kind to be found in no one else in the rest of Greece. (As for the men called the Seven, those who inquire into such things more precisely do not count them among the wise.) At Athens, we are told, there was one—and he, indeed, judged the wisest even by Apollo's oracle. The wisdom they suppose to be in you is this: that you hold all that is yours to rest within yourself, and reckon the accidents of human life inferior to virtue. And so they ask me—and likewise, I imagine, this man Scaevola—how on earth you bear the death of Africanus. They ask all the more because on the last Nones, when we had come, as we usually do, to the gardens of Decimus Brutus the Augur to study together, you were not there—you who had always been most scrupulous to keep that day and that duty."

\ciceroSection{8} "They do indeed ask much, Gaius Laelius," said Scaevola, "as Fannius has said; but I answer what I myself have observed: that you bear with restraint the grief you have received at the death of a man at once supremely great and most dear to you; that you could not but be moved, and that not to be moved would not have been like your humane nature; and as for your absence from our college on the Nones, I answer that the cause was ill health, not sorrow." "Quite right, Scaevola," Laelius replied, "and quite true." For I ought not to have been drawn away by any discomfort of mine from a duty I always performed when I was well, and I do not think it can ever befall a steadfast man, by any accident, that he should let his duty lapse.

\ciceroSection{9} As for you, Fannius—when you say that so much is granted to me as I neither recognize nor claim, you act as a friend; but, as it seems to me, you do not judge rightly about Cato. For either no one was wise, which I am the more inclined to believe, or, if anyone was, it was he. To pass over all else—how he bore the death of his son! I remembered Paulus, I had seen Gallus; but they lost boys, while Cato lost a grown man, tried and proven.

\ciceroSection{10} For this reason, take care not to rank above Cato even that very man whom Apollo, as you say, judged the wisest; for of the one the deeds are praised, of the other only the words. As for myself—to speak now with both of you—hold this for certain: if I were to deny that I am moved by longing for Scipio, the wise may decide how rightly I would do so, but I would surely be lying. For I am moved at being bereft of such a friend as I think there will never be again, and as I can affirm there certainly never was. But I have no need of healing. I console myself, and above all by this comfort: that I am free of the error which torments most people at the passing of their friends. I judge that nothing bad has happened to Scipio; it has happened to me, if anything has. And to be grievously tormented by one's own troubles is the mark not of one who loves his friend, but of one who loves himself.

\ciceroSection{11} As for Scipio—who could deny that all went splendidly with him? Unless he had wished to pray for immortality, which was the last thing in his mind, what did he fail to attain that it is right for a man to pray for? As a mere youth he at once surpassed, by his incredible virtue, the highest hopes his fellow citizens had held of him since boyhood. He never stood for the consulship, yet was made consul twice—the first time before the proper age, the second at a time right for him but almost too late for the Republic. By overthrowing two cities most hostile to this empire, he wiped out not only the present wars but the future ones as well. What shall I say of his most easy-going character, of his devotion to his mother, his generosity to his sisters, his kindness to his own people, his justice toward all? These things are known to you. And how dear he was to the state was shown by the grief at his funeral. What good, then, could the addition of a few more years have done him? For old age, however unburdensome it may be—as I remember Cato arguing, with me and with Scipio, the year before he died—nonetheless strips away that vigor in which Scipio still stood even now.

\ciceroSection{12} His life, then, was such, whether in fortune or in glory, that nothing could be added to it; and the swiftness of his death took away all sense of dying. About the manner of that death it is hard to speak; you see what people suspect. This, however, may truly be said: of the many days Publius Scipio saw in his life, days of the greatest fame and joy, the most brilliant was that day, the day before he departed from life, when, the Senate having been dismissed, he was escorted home toward evening by the senators, by the Roman people, by the allies and the Latins—so that from so lofty a step of honor he seems to have passed up to the gods above rather than down to the dead below.

\ciceroSection{13} For I do not agree with those who have lately begun to argue that souls perish together with bodies, and that all is wiped out in death. With me the authority of the ancients carries more weight—whether of our own ancestors, who assigned to the dead such reverent rites, which they surely would not have done had they thought it of no concern to them; or of those who lived in this land and instructed by their institutions and teachings that Greater Greece which now indeed is destroyed but then was flourishing; or of him who by Apollo's oracle was judged the wisest, who did not say now one thing, now another, as most men do, but always the same: that the souls of men are divine, and that for them, once they have departed from the body, a return to heaven lies open, and most readily of all for the best and most just.

\ciceroSection{14} Scipio saw it the same way. Indeed, as though he had a presentiment, in the very last days before his death—when Philus was there, and Manilius, and a good many others, and you too, Scaevola, had come with me—he discoursed for three days on the commonwealth. The close of that discussion was given over almost entirely to the immortality of the soul, which he said he had heard, in a vision while he slept, from the elder Africanus. If that is so—if it is the soul of every best man that escapes most easily in death, as though from the prison and chains of the body—then whose passage to the gods do we suppose was easier than Scipio's? And so I fear that to grieve at his lot is the mark of envy rather than of friendship. But if, on the other hand, the truer view is that soul and body perish together and no consciousness remains, then just as there is nothing good in death, so surely there is nothing evil. For with consciousness lost, a man becomes exactly as if he had never been born at all—and yet that he was born we rejoice, and this city will go on rejoicing as long as it stands.

\ciceroSection{15} For these reasons, as I said above, all has fallen out best of all for him; less kindly for me, since it would have been fairer, just as I entered life before him, so to leave it before him too. And yet I take such joy in the memory of our friendship that I think I have lived a happy life, because I lived it with Scipio—with whom I shared concern for public affairs and private, with whom I shared a home and military service, and, the thing in which the whole force of friendship lies, the fullest agreement of wishes, of pursuits, of views. And so it is not so much that reputation for wisdom which Fannius mentioned just now that delights me—false, especially, as it is—as the hope that the memory of our friendship will last forever. And this is the dearer to my heart because, out of all the ages, scarcely three or four pairs of friends are named; and it is among such company that I seem to hope the friendship of Scipio and Laelius will be known to posterity.

\ciceroSection{16} "That, Laelius," said Fannius, "must indeed be so. But since you have brought up the subject of friendship, and we are at leisure, you would do me a great kindness—and Scaevola too, I hope—if, in the way you usually treat other questions when they are put to you, you would set out your views on friendship: what you think it is, how you judge it, what rules you would give." "It would be welcome to me as well," said Scaevola, "and just as I was about to take this very thing up with you, Fannius got there first. So you would do both of us a very great kindness."

\ciceroSection{17} "I would not refuse," said Laelius, "if I had confidence in myself; for the subject is a splendid one, and we are, as Fannius said, at leisure." But who am I, and what capacity is in me? That is the practice of the learned—of the Greeks—to have a topic set them to argue on however suddenly. It is a great task, and it calls for no small training. And so, as for what can be argued about friendship, I think you should seek it from those who profess such things; I can only urge you to set friendship above all human concerns. For nothing is so suited to our nature, nothing so fitting whether in fortune's favor or against it.

\ciceroSection{18} But this is the first thing I hold: that friendship cannot exist except among good men. I do not cut this to the quick, as those who argue these matters more subtly do—truly, perhaps, but with too little regard for common usefulness; for they deny that anyone is a good man unless he is wise. So be it, by all means. But the wisdom they expound is one that no mortal has yet attained. We, however, must look to the things that exist in everyday life and practice, not to those that are imagined or wished for. I will never say that Gaius Fabricius, Manius Curius, Tiberius Coruncanius—men whom our ancestors judged wise—were wise by their standard. So let them keep the name of wisdom, both invidious and obscure; let them grant that these men were good men. They will not even do that: they will deny that this can be granted to anyone but the wise man.

\ciceroSection{19} Let us proceed, then, with what they call a thick-witted Minerva. Those who so conduct themselves, so live, that their loyalty, integrity, fairness, and generosity are proven; in whom there is no greed, no lust, no rashness; and who are of great steadfastness, as those were whom I named just now—these men, who have been held to be good men, let us think that they should also be so called, because they follow nature, as far as men can, as the best guide to living well. For this is how I seem to perceive it: that we are so born that a certain bond exists among all, and a stronger one the closer each comes. And so fellow citizens are preferred to foreigners, kinsmen to strangers; for with these nature herself has brought forth a kind of friendship, but one that lacks sufficient firmness. For friendship surpasses kinship in this: that goodwill can be removed from kinship, but not from friendship; for once goodwill is taken away, the name of friendship is gone, while that of kinship remains.

\ciceroSection{20} How great the force of friendship is can be best understood from this: out of the boundless society of the human race, which nature herself has knit together, the matter has been so drawn in and narrowed that all affection is bound up either between two people or among a few. For friendship is nothing other than agreement on all things human and divine, joined with goodwill and affection; and I am inclined to think that, with the exception of wisdom, nothing better has been given to man by the immortal gods. Some put riches first, others good health, others power, others public honors, many even pleasures. This last is the mark of beasts; those earlier things are fleeting and uncertain, set not so much in our own deliberation as in the recklessness of fortune. As for those who place the highest good in virtue—splendid men indeed—it is this very virtue that both begets and sustains friendship, and without virtue friendship cannot exist in any way at all.

\ciceroSection{21} Let us now interpret virtue by the usage of our own life and our own speech, and not measure it, as certain learned men do, by grandeur of words; and let us count among good men those who are held to be such—the Pauluses, the Catos, the Galuses, the Scipios, the Phili. With these our common life is content; let us pass over those who are nowhere to be found at all.

\ciceroSection{22} Friendship, then, among such men holds advantages so great that I can scarcely state them. To begin with, how can life be living—as Ennius puts it—if it does not rest in the mutual goodwill of a friend? What is sweeter than to have someone with whom you may dare to speak of everything as with yourself? What great fruit would there be in good fortune if you had no one to take as much joy in it as you do yourself? And adversity would be hard to bear without someone who bore it even more heavily than you. In short, the other things that are sought after are each suited, more or less, to some single use—riches, to spend; resources, to win regard; honors, to be praised; pleasures, to enjoy; health, to be free of pain and to perform the body's offices. Friendship embraces a great many things: wherever you turn, it is there at hand; it is shut out from no place, never out of season, never a burden. And so we do not, as the saying goes, use water or fire in more places than we use friendship. I am not speaking now of the common or middling sort—though even that both delights and profits—but of the true and perfect kind, such as belonged to those few who are named. For friendship makes good fortune more splendid, and adversity, by dividing and sharing it, lighter.

\ciceroSection{23} And while friendship embraces very many and very great advantages, it surely surpasses them all in this: that it lights the way ahead with good hope, and does not let the spirit be weakened or fall. Indeed, the man who looks upon a friend looks, as it were, upon a kind of likeness of himself. And so the absent are present, the needy are in abundance, the weak are strong, and—what is harder to say—the dead live; such honor, such remembrance, such longing on the part of their friends attends them. From this the death of the one seems blessed, the life of the others praiseworthy. But if you should remove from the nature of things the bond of goodwill, no house, no city could stand; not even the tilling of the fields would endure. If this is grasped too little—how great the force of friendship and concord is—it can be perceived from quarrels and discords. For what house is so stable, what city so firm, that it cannot be utterly overthrown by hatreds and divisions? From this it can be judged how much good there is in friendship.

\ciceroSection{24} They say that a certain learned man of Agrigentum prophesied in Greek verses that whatever in the nature of things and in the whole world stands fixed and whatever moves, friendship draws together and discord scatters. And this, indeed, all mortals both understand and confirm in practice. And so, whenever some service of a friend has shown itself in confronting dangers or sharing them, who is there who does not extol it with the highest praise? What shouts went up lately throughout the whole theater at the new play of my guest-friend and friend Marcus Pacuvius—when, the king not knowing which of the two was Orestes, Pylades declared that he was Orestes, so as to be killed in his place, while Orestes, as he truly was, kept insisting that he was Orestes! The audience stood and applauded at a thing merely staged; what do we suppose they would have done in a real case? Nature easily showed her own force here, in that men judged a thing rightly done in another which they themselves could not do. Thus far, I think, I have been able to say what I feel about friendship; if there is anything more—and I believe there is much—seek it, if you see fit, from those who argue such matters.

\ciceroSection{25} "From you, rather," said Fannius. "Though I have often asked it of those men too, and listened, indeed, not unwillingly—but the thread of your discourse is something different." "You would say so all the more, Fannius," said Scaevola, "if you had been present lately in Scipio's gardens, when there was a debate on the commonwealth. What a champion of justice he was then, against the carefully wrought speech of Philus!" "It was easy enough," said Fannius, "for the most just of men to defend justice." "And friendship?" said Scaevola. "Was that not easy for the man who, by keeping it with the highest fidelity, steadfastness, and justice, won the greatest glory?"

\ciceroSection{26} "This is to use force," said Laelius. "For what does it matter by what means you compel me? Compel me you do." For to resist the eager wishes of one's sons-in-law—especially in a good cause—is not only hard but not even fair. Very often, then, when I reflect on friendship, what seems chiefly to call for consideration is this: whether friendship was longed for out of weakness and want, so that in the giving and receiving of services each might get from another what he was less able to accomplish by himself, and in turn return it; or whether, though this is indeed proper to friendship, there was some other cause, older and more beautiful and sprung more directly from nature herself. For love—from which friendship takes its name—is the chief thing in the joining of goodwill. For advantages, after all, are often gathered even from those who are cultivated and attended out of a pretense of friendship, for the sake of the moment; but in friendship there is nothing feigned, nothing pretended, and whatever there is, it is genuine and freely given.

\ciceroSection{27} For this reason it seems to me that friendship arises from nature rather than from need, from an inclination of the mind joined with a certain feeling of love, rather than from a calculation of how much advantage the thing would bring. What kind of feeling this is can be observed even in certain animals, which up to a certain point love their own offspring and are so loved by them in return that their feeling is plain to see. In a human being this is far more evident: first, in that affection which exists between children and parents, which cannot be severed except by some unspeakable crime; and second, when a like feeling of love arises, if we have come upon someone with whose character and nature we are in harmony, because in him we seem to glimpse a kind of light of integrity and virtue.

\ciceroSection{28} For nothing is more lovable than virtue, nothing that draws us more powerfully to affection, since on account of virtue and integrity we feel, in a way, affection even for those we have never seen. Who does not call up the memory of Gaius Fabricius or Manius Curius with a certain warmth of goodwill, though he never saw them? Who, on the other hand, does not hate Tarquinius Superbus, or Spurius Cassius, or Spurius Maelius? Against two commanders we fought to the death for supremacy in Italy, Pyrrhus and Hannibal: toward the one, on account of his integrity, our feelings are not too unkind; the other, on account of his cruelty, this state will hate forever.

\ciceroSection{29} But if the power of integrity is so great that we feel affection for it even in those we have never seen, and—what is more—even in an enemy, what wonder is it if the minds of men are moved when they seem to glimpse the virtue and goodness of those with whom they can be joined in close dealings? And yet love is strengthened by a kindness received, by zeal observed, and by familiarity added on; and when these are brought to bear on that first stirring of the mind and of love, a remarkable greatness of goodwill bursts into flame. If any suppose that this proceeds from weakness—so that there should be someone through whom each may attain what he lacks—they leave to friendship an origin that is humble indeed and, so to speak, anything but noble, since they would have it born of poverty and need. If that were so, then the less a man judged himself to possess, the better suited he would be for friendship; but the truth is far otherwise.

\ciceroSection{30} For the more a man trusts in himself, and the more he is so fortified by virtue and wisdom that he needs no one and judges everything that is his to rest within himself, the more he excels in seeking out and cultivating friendships. Take Africanus: did he have need of me? By Hercules, not in the least! Nor I, for that matter, of him; but I felt affection for him out of a certain admiration for his virtue, and he in turn out of some opinion, perhaps, that he held of my character; and familiarity increased our goodwill. Yet, though many and great advantages followed, it was not from the hope of these that the reasons for our affection arose.

\ciceroSection{31} For just as we are generous and openhanded not in order to exact gratitude—since we do not lend out a kindness at interest, but are by nature inclined to generosity—so we judge that friendship is to be sought not because we are drawn by the hope of reward, but because all its fruit lies in love itself.

\ciceroSection{32} From these men, who in the manner of cattle refer everything to pleasure, they differ utterly—and no wonder; for they can look up to nothing lofty, nothing grand and divine, who have flung all their thoughts down into a thing so mean and so contemptible. Let us therefore set such men apart from this discussion, and understand for ourselves that the feeling of affection, and the warmth of goodwill, are begotten by nature once a sign of integrity has been given. Those who reach out for it attach themselves and draw nearer, so as to enjoy both the company and the character of the one they have begun to love; and they grow equal and alike in love, more ready to do a good turn than to call one in, and let this be the honorable rivalry between them. In this way both the greatest advantages will be reaped from friendship, and its origin from nature rather than from weakness will be the weightier and the truer. For if advantage were the glue of friendships, the same advantage, once altered, would dissolve them; but because nature cannot be changed, for that very reason true friendships are everlasting. You see, then, the origin of friendship—unless perhaps you have something to add to this.

"Press on, Laelius," said Fannius; "for I answer by my own right, on behalf of this one here, who is the younger."

\ciceroSection{33} "Quite right of you," said Scaevola; "so let us listen." Listen, then, my excellent friends, to the things that Scipio and I most often used to debate between us concerning friendship. And yet he, for his part, used to say that nothing was more difficult than for a friendship to endure all the way to the last day of life: for it often happens either that the same thing is not to one's advantage, or that men do not think alike about public affairs; and he would say that the characters of men too are often altered, sometimes by adversity, sometimes by the weight of advancing age. And he drew an illustration of these matters from the likeness of early life, since the strongest attachments of boyhood are often laid aside along with the boy's toga;

\ciceroSection{34} but even if they had carried them on into young manhood, they were nonetheless sometimes severed by a rivalry—over a marriage match, say, or over some advantage that the two could not both obtain. And if any had advanced further in friendship, still it was often shaken if they fell into a contest for office; for there is no greater bane to friendships than, in most men, the lust for money, and, in the very best, the struggle for office and glory—out of which the bitterest enmities have often arisen between the closest of friends.

\ciceroSection{35} Great rifts too, and for the most part justified ones, arise when something is demanded of friends that is not right—that they serve as agents of lust, or as accomplices in wrongdoing; and those who refuse, however honorably they act, are nonetheless charged with deserting the bond of friendship by the very men whose wishes they would not indulge. Those, on the other hand, who dare to demand anything whatever of a friend, by the very demand profess that they would do anything for their friend's sake. By their complaining, grown chronic, not only are intimacies commonly extinguished, but undying hatreds are begotten as well. So many of these calamities, he used to say, hang like fated dooms over friendships, that to escape them all seemed to him a matter not only of wisdom but of good fortune.

\ciceroSection{36} For this reason let us first consider, if you please, how far love in friendship ought to go. Surely, if Coriolanus had friends, they were not bound to take up arms against their fatherland alongside him? Were his friends bound to aid Vecellinus as he reached for kingship, or Maelius?

\ciceroSection{37} Tiberius Gracchus, indeed, when he was tearing the state apart, we saw abandoned by Quintus Tubero and the friends of his own age. But Gaius Blossius of Cumae, a guest-friend of your family, Scaevola, when he came to me to plead his case—for I was sitting on the council during the consulship of Laenas and Rupilius—offered this as the ground on which I should pardon him: that he had thought so much of Tiberius Gracchus that he believed he ought to do whatever Gracchus wished. "What," I said, "even if he had wished you to carry firebrands to the Capitol?" "He would never have wished that," he answered, "but had he wished it, I would have obeyed." You see what an abominable thing to say. And by Hercules he made good on it, and even went beyond his words; for he did not merely obey the recklessness of Tiberius Gracchus—he led it, offering himself not as the companion of that madness but as its commander. And so, in this derangement, terrified by the new commission of inquiry, he fled into Asia, went over to the enemy, and paid the state a heavy and just penalty. There is, then, no excuse for a wrong if you have done it for a friend's sake; for since it was an opinion of virtue that brought the friendship together, it is hard for the friendship to remain once you have fallen away from virtue.

\ciceroSection{38} But if we should rule it right either to grant friends whatever they wish, or to obtain from them whatever we wish, then—provided we were of perfect wisdom—the matter would carry no flaw. But we are speaking of the friends before our eyes, those we see or of whom we have received the memory, those whom common life has known. From this number must our examples be drawn, and most of all from those who come nearest to wisdom.

\ciceroSection{39} We learn that Aemilius Papus was an intimate of Gaius Luscinus (so we have it from our fathers): twice consuls together, colleagues in the censorship. And it is on record that, both with these men and with one another, Manius Curius and Tiberius Coruncanius were most closely bound. Therefore we cannot so much as suspect that any one of these men ever pressed a friend for anything that ran against good faith, against an oath, against the state. For in the case of such men, what is the point of saying that, had he pressed for it, he would not have prevailed—since they were men of the deepest scruple, and since it is equally an offense both to do such a thing when asked and to ask it? But Tiberius Gracchus, for his part, was followed by Gaius Carbo and Gaius Cato, and least of all at that time by his brother Gaius—who now is the fiercest of them all.

\ciceroSection{40} Let this, then, be enacted as a law in friendship: that we neither ask of a friend what is shameful, nor do it when asked. For the excuse is shameful, and least of all to be accepted—in other wrongdoing certainly, but above all if a man should confess that he had acted against the commonwealth for a friend's sake. We stand in such a position, Fannius and Scaevola, that it is right for us to look far ahead to the dangers coming upon the republic. The discipline of our ancestors has already swerved somewhat from its track and its course.

\ciceroSection{41} Tiberius Gracchus attempted to seize royal power—or rather he did reign, for a few months at least. Had the Roman people ever heard or seen anything like it? The friends and kinsmen who followed his cause even after his death—what they did to Publius Scipio I cannot tell without tears. As for Carbo, we bore with him as best we could, on account of the recent punishment of Tiberius Gracchus. About the tribunate of Gaius Gracchus I have no wish to play the prophet; the thing creeps on, and once it has begun, it slides downhill, headlong toward ruin. You see already what damage was done long ago with the ballot, first by the Gabinian law, and then, two years later, by the Cassian. Already I seem to see the people sundered from the Senate, and the gravest matters decided at the whim of the crowd. For more men will learn how such things are done than how they are to be resisted.

\ciceroSection{42} Why do I say all this? Because no one attempts anything of the kind without allies. The good, then, must be instructed: if by some chance they have stumbled unawares into friendships of this sort, they should not suppose themselves so bound that they cannot break away from friends who are sinning in some great matter against the state. And for the wicked a penalty must be fixed—and no lighter one for those who follow another than for those who have themselves been the leaders in the crime. Who was more illustrious in Greece than Themistocles, who more powerful? Yet when, as commander in the Persian War, he had freed Greece from slavery, and was then driven into exile through envy, he would not bear the wrong done him by an ungrateful country—the wrong he was bound to bear: he did the very thing Coriolanus had done among us twenty years before. For neither of them was a single helper found against his country; and so each took his own life.

\ciceroSection{43} For this reason such a conspiracy of the wicked is not only not to be shielded by the excuse of friendship, but rather to be punished with every penalty, so that no one may think it permitted to follow a friend even when he is making war on his country. And this—the way things have begun to go—may well come to pass one day; for it troubles me no less what the commonwealth will be after my death than what it is today.

\ciceroSection{44} Let this, then, be enacted as the first law of friendship: that we ask of friends only what is honorable, that for a friend's sake we do only what is honorable, that we not even wait to be asked—let the eagerness always be there, hesitation absent; let us dare to give true counsel freely; let the influence of friends who counsel well carry the greatest weight in a friendship, and let it be used to admonish not only openly but even sharply, if the situation demands; and let it be obeyed when so used.

\ciceroSection{45} For certain men—reputed wise in Greece, I am told—held, I believe, some startling opinions (but there is nothing they will not pursue with their subtleties): some that excessive friendships are to be avoided, lest a single man be forced to be anxious on behalf of many; that each has enough and more than enough of his own affairs; that to be too entangled in others' is a burden; that the most convenient course is to hold the reins of friendship as loose as possible, to draw them tight when you wish or to slacken them; for the chief thing in living happily is freedom from care, which the mind cannot enjoy if one man is, as it were, in labor on behalf of many.

\ciceroSection{46} Others, they say, hold a view far more inhuman still—a point I touched on briefly a moment ago—that friendships are to be sought for the sake of protection and support, not of goodwill or affection; and so the less firmness and the less strength a man has, the more he seeks out friendships: from which it comes about that frail women look for the protections of friendship more than men do, and the poor more than the wealthy, and the wretched more than those reckoned fortunate.

\ciceroSection{47} What splendid wisdom! For those who take friendship out of life seem to take the sun out of the world—friendship, than which we have nothing better from the immortal gods, nothing more delightful. For what is this freedom from care of theirs? In appearance, indeed, it is alluring, but in fact, on many counts, to be rejected. For it is not consistent to refuse to undertake any honorable thing or course of action, or having undertaken it to lay it down, merely to spare oneself anxiety. But if we flee care, we must flee virtue, which necessarily, and not without some care, spurns and hates the things contrary to itself—as goodness hates wickedness, temperance lust, courage cowardice. And so you may observe that the just grieve most at unjust deeds, the brave at cowardly ones, the self-controlled at shameful ones. It is therefore the mark of a well-ordered mind both to rejoice at good things and to grieve at their opposites.

\ciceroSection{48} For this reason, if grief of mind can fall upon the wise man—and surely it can, unless we suppose that all human feeling has been rooted out of his heart—what cause is there why we should pull friendship up by the roots out of life, merely to avoid taking on some troubles for its sake? For once feeling is taken away, what difference is there—I will not say between a beast and a man, but between a man and a block of wood, or a stone, or anything of that kind? Nor are those to be heeded who would have virtue be something hard and, as it were, of iron; whereas in fact it is, in many matters and especially in friendship, tender and pliant, so that it expands, as it were, at a friend's good fortune and contracts at his misfortune. For this reason that distress which must often be felt for a friend has not force enough to take friendship out of life—no more than the virtues are to be rejected because they bring with them some cares and troubles. And since, as I said above, friendship draws men together when some sign of virtue shines forth, to which a kindred spirit attaches and binds itself—when that happens, love must of necessity arise.

\ciceroSection{49} For what is so absurd as to take delight in many empty things—in office, in glory, in a fine house, in dress and bodily adornment—and to take no great delight in a living being endowed with virtue, one who can love or, so to speak, love in return? For nothing is more delightful than the repayment of goodwill, nothing than the exchange of attentions and good offices.

\ciceroSection{50} And what if we add this besides, as may rightly be added—that nothing so entices and draws a thing toward itself as likeness draws toward friendship? Surely it will be granted as true that the good love the good and bind them to themselves as though joined by kinship and by nature. For nothing is more eager for what resembles itself, nothing more grasping after it, than nature. So let this much, then, Fannius and Scaevola, stand settled, as I think: that between good men there is, as it were, a necessary goodwill, which is the source of friendship established by nature. But this same goodness extends also to the multitude. For virtue is not inhuman, nor ungiving, nor proud; it is wont even to watch over whole peoples and to take the best counsel for them—which surely it would not do if it shrank from affection for the common throng.

\ciceroSection{51} And to me, indeed, those who invent friendships for the sake of advantage seem to take away the most lovable bond of friendship. For it is not so much the advantage gained through a friend that delights us as a friend's love in itself; and what comes from a friend is then delightful, if it has come from devotion. So far is it from being true that friendships are cultivated out of need, that those who least need another—because they are richest in resources and means, and above all in virtue, in which there lies the greatest protection—are the most generous and the most bountiful. And perhaps it is not even desirable that a friend should never want for anything at all. For where would my devotion have shown itself, if Scipio had never had need of my counsel or my service, at home or in the field? Friendship, then, did not follow upon advantage; advantage followed upon friendship.

\ciceroSection{52} We are not, therefore, to listen to men awash in luxury, if they ever hold forth on friendship—a thing they know neither by practice nor by reflection. For who is there—by the faith of gods and men!—who would wish, on condition of loving no one and being loved by none in return, to swim in every kind of abundance and to live amid a plenty of all things? This, surely, is the life of tyrants—a life in which there can be no loyalty, no affection, no settled trust in any goodwill, where everything is always suspect and full of anxiety, and where there is no place for friendship.

\ciceroSection{53} For who would love a man he fears, or one by whom he believes himself feared? Such men are courted, yes—but only with pretense, and only for a while. And if by chance they fall, as usually happens, then it becomes clear how poor in friends they were. This is what they say Tarquin remarked in exile: that only then did he understand which friends he had held loyal and which disloyal—now that he could no longer repay either kind.

\ciceroSection{54} Though I am surprised that, with such arrogance and such harshness, he could have had any friend at all. And just as this man's character could win him no true friends, so the resources of many men of overwhelming power shut out faithful friendships. For fortune is not only blind herself; she generally makes blind as well those she has embraced. And so they are carried away, as a rule, by disdain and obstinacy—and nothing in the world can be more insufferable than a fool who has been favored by fortune. Indeed you may observe how men once of agreeable character are transformed by command, by power, by prosperity, how they scorn their old friendships and indulge new ones.

\ciceroSection{55} And what could be more foolish than this: that men with the fullest command of means, of resources, of wealth should procure everything else that money procures—horses, servants, splendid clothing, costly vessels—and not procure friends, the finest and most beautiful furnishings of life, if I may put it so? For when they procure those other things, they do not know for whom they procure them, nor for whose sake they labor; each of those things belongs to whoever has prevailed by force. But each man's possession of his friendships remains his own, stable and secure—so that even if those other things endure, those gifts, as it were, of fortune, still a life uncultivated and deserted by friends can hold no joy. But enough of this.

\ciceroSection{56} We must now establish what the limits are in friendship, and the boundaries, so to speak, of our loving. On this I see three opinions advanced, none of which I approve: first, that we should feel toward a friend exactly as we feel toward our own selves; second, that our goodwill toward our friends should answer to their goodwill toward us evenly and in equal measure; third, that a man should be valued by his friends at exactly the value he sets upon himself.

\ciceroSection{57} To none of these three opinions do I assent at all. For that first one is not even true—that a man should be disposed toward his friend just as he is toward himself. Consider how many things we do for our friends' sake that we would never do for our own! To beg from someone unworthy, to plead, then to attack a man more bitterly or to pursue him more fiercely—things that in our own affairs are not quite honorable, but in our friends' affairs are entirely honorable. And there are many matters in which good men sacrifice much of their own advantage, and let it be taken from them, so that their friends rather than themselves may enjoy it.

\ciceroSection{58} The second opinion is the one that defines friendship by equal services and equal goodwill. This is to summon friendship to the accounts too meanly and too meagerly—as if the reckoning of what is received and what is given must balance. True friendship seems to me richer and more abundant than this, not watching narrowly so as never to return more than it received. For there is no need to fear that something will be lost, or spilled onto the ground, or that more than a fair share will be heaped into the friendship.

\ciceroSection{59} But that third limit is the worst of all—that a man should be valued by his friends at exactly the value he sets upon himself. For often in some men the spirit is too dejected, or the hope of bettering their fortune too broken. It is not, then, the part of a friend to be toward such a man as that man is toward himself, but rather to strive and bring it about that he lift up his friend's downcast spirit and lead him into better hope and better thoughts. A different limit, then, must be set for true friendship—once I have first reported what Scipio used most to condemn. He used to say that no saying more hostile to friendship could be found than that of the man who said one ought to love as though one might someday hate. And he could not be brought to believe that this, as people supposed, had been spoken by Bias, who was held to be one of the seven sages: it was the maxim of some unprincipled or self-seeking man, or one who refers everything to his own power. For in what way can anyone be a friend to a man he believes he may become enemy to? Why, he will even be obliged to wish and pray that his friend goes wrong as often as possible, so as to give him more handles, so to speak, for blame—and conversely he will be obliged to feel anguish and grief and envy at his friends' right conduct and good fortune.

\ciceroSection{60} And so this precept, whoever its author is, serves to abolish friendship. What ought rather to have been laid down is this: that we should use such care in forming friendships that we never begin to love a man we might someday come to hate. Indeed, Scipio held that even if we had been less than fortunate in our choosing, the thing was to be borne rather than to keep an eye out for the moment of enmity.

\ciceroSection{61} These, then, are the limits I think we should use: that, the characters of our friends being sound, there should then be among them a sharing of all things, all counsels, all wishes, without any exception—so that even if it should somehow happen that we must support our friends' less than just wishes, in cases where their life or their reputation is at stake, we should turn aside from the straight path, provided no utter disgrace follows. For there is a point up to which indulgence may be granted to friendship. Nor indeed is reputation to be neglected, nor should one reckon the goodwill of one's fellow citizens a trifling weapon for getting things done—though to gather it by flattery and fawning is shameful. Virtue, which affection follows, is by no means to be cast aside.

\ciceroSection{62} But—for I keep returning to Scipio, whose whole talk was of friendship—he used to complain that in every other matter men are more careful. Each can say how many goats or sheep he has, but cannot say how many friends he has; in acquiring the former they take pains, but in choosing friends they are careless and have, as it were, no signs or marks by which to judge who is fit for friendship. We must therefore choose men who are firm and steady and constant—a kind in great scarcity. And to judge is hard indeed without trial, yet the trial must be made within the friendship itself. So friendship outruns the judgment and takes away the chance of trying first.

\ciceroSection{63} It is the part of a prudent man, then, to rein in the rush of his goodwill as he would a chariot's course—so that we may use our friendship as we use horses that have been tested, having put our friends' characters to the proof in some respect. Some men are often seen, over a trifling sum of money, to be how worthless they are; others, whom a small amount could not move, are found out where the stakes are large. But suppose some are discovered who think it base to set money above friendship—where shall we find those who do not set honors, magistracies, military commands, powers, and wealth above friendship, so that, with these proposed on the one side and the claim of friendship on the other, they do not far prefer the former? For nature is too weak to despise power; and even when men have attained it in disregard of friendship, they suppose the act will be obscured, because it was not without grave cause that the friendship was disregarded.

\ciceroSection{64} And so true friendships are found with the greatest difficulty among those who are engaged in office and public affairs. For where will you find the man who sets a friend's advancement above his own? And—to pass over this—how heavy, how hard most men find any partnership in calamity, into which it is not easy to find men who will descend. Though Ennius is right: \textit{a sure friend is discerned in an unsure matter}; still, these two things convict most men of fickleness and weakness: if they look down on a friend in good times, or desert him in bad. Whoever, then, in both situations shows himself grave, constant, and steadfast in friendship—him we ought to judge to be of a kind exceedingly rare, and all but divine.

\ciceroSection{65} The foundation of that stability and constancy which we seek in friendship is good faith; for nothing is stable that is faithless. It is right, besides, to choose a man who is straightforward and companionable and sympathetic—that is, one who is moved by the same things we are; all of which bears on faithfulness. For a nature that is complicated and devious cannot be faithful; nor indeed can a man who is not moved by the same things and not sympathetic by nature be either faithful or stable. We must add this too: that he should take no delight in leveling accusations, nor believe them when they are leveled by others—all of which bears on that constancy I have been treating all along. And so that proves true which I said at the outset: that friendship cannot exist except among good men. For it is the mark of a good man—whom we may equally call a wise one—to hold to these two things in friendship: first, that there be nothing feigned or pretended; for to hate openly is more becoming to a free man than to mask one's feeling behind a smooth face; second, not only to repel accusations brought by another, but not even to be suspicious oneself, forever imagining that one's friend has done some wrong.

\ciceroSection{66} To this should be added a certain charm of conversation and manner, by no means a trivial seasoning of friendship. Gloom and severity in everything do carry a certain weight, it is true, but friendship ought to be more relaxed, freer, sweeter, and more inclined to every kind of courtesy and ease.

\ciceroSection{67} At this point a somewhat difficult question arises: whether new friends, worthy of friendship, should ever be preferred to old ones, as we are accustomed to prefer young horses to aging ones. The doubt is unworthy of a human being. For there should be no surfeit of friendships, as there is of other things; the oldest, like those wines that bear age well, ought to be the sweetest. And that saying is true, that many bushels of salt must be eaten together before the duty of friendship is fulfilled.

\ciceroSection{68} New friendships, though, if they bring promise — so that, as in plants that do not deceive, the fruit shows itself — are not to be rejected; yet the old must be kept in its own place. For the force of age and habit is very great. Even in the horse I mentioned just now, if nothing stands in the way, there is no one who would not more gladly use the one he is used to than one untrained and new. And indeed habit holds sway not only in this, which is a living creature, but in things that are without life as well, since we take delight in the very places — even mountainous and wooded ones — in which we have lingered a long while.

\ciceroSection{69} But the greatest thing in friendship is for the superior to be on equal terms with the inferior. For often there are certain kinds of preeminence, such as Scipio's was in our little flock, if I may call it that. Never did he set himself above Philus, never above Rupilius, never above Mummius, never above his friends of lower rank. As for his brother Quintus Maximus, an outstanding man and yet in no way his equal, since he was the elder in years, Scipio honored him as a superior, and wished all his own people to grow greater through his help.

\ciceroSection{70} This is what everyone should do and imitate: if they have attained any preeminence of virtue, talent, or fortune, let them share it with their own and pass it on to those nearest them. If they were born of humble parents, if they have kinsmen weaker in spirit or in fortune, let them increase their resources and be to them a source of honor and standing. So in the plays, those who for a time were in servitude because their stock and lineage were unknown, when they are recognized and found to be the sons of gods or kings, still keep their affection for the shepherds whom for many years they took to be their fathers. And surely this is to be done all the more in the case of real and certain fathers. For the fruit of talent and virtue and every kind of preeminence is reaped most fully when it is bestowed on those nearest to us.

\ciceroSection{71} As, then, those who are the superiors in a bond of friendship and close connection ought to put themselves on a level with their inferiors, so the inferiors should not be aggrieved at being surpassed by their friends in talent, fortune, or standing. Yet most of them are forever either complaining about something or even reproaching, and all the more so if they think they have something they can claim to have done as a service, in friendship, and with some labor of their own. A thoroughly odious breed of men, this, who throw their services in your face — services that the one who received them ought to remember, and the one who rendered them ought never to mention.

\ciceroSection{72} For this reason, as those who are superiors should lower themselves in friendship, so in a way they should raise up their inferiors. For there are some who make their friendships burdensome by thinking themselves slighted — which scarcely happens except to those who believe they actually deserve to be slighted. Such men must be relieved of this opinion not only by words but by deeds.

\ciceroSection{73} To each, however, only so much should be given: first, as much as you yourself can manage, and then also as much as the one you love and assist can bear. For you cannot, however eminent you are, bring all your friends through to the highest honors — as Scipio could make Publius Rupilius consul, but could not so make his brother Lucius. And even if you could confer anything you pleased on another, you must still consider what that other can sustain.

\ciceroSection{74} On the whole, friendships are to be judged when character and age are already grown firm and settled; nor, if some men in their early years were keen on hunting or ball-playing, should they count as their intimates those whom they loved then for sharing the same pursuit. For by that reasoning our nurses and tutors would, by the right of long acquaintance, demand the largest share of goodwill. They are not, of course, to be neglected, but they are to be valued in some other way; otherwise friendships cannot remain stable. For unlike characters are followed by unlike pursuits, and this unlikeness breaks friendships apart. There is no other reason at all why the good cannot be friends with the wicked, nor the wicked with the good, except that between them lies the greatest possible distance in character and in pursuits.

\ciceroSection{75} A sound rule, too, can be laid down in friendships: that an unmeasured goodwill — which very often happens — should not stand in the way of a friend's great advantage. For, to return to the plays, Neoptolemus could not have taken Troy if he had been willing to listen to Lycomedes, who had brought him up and who tried, with many tears, to block his going. And great occasions often arise that call for parting from friends; whoever wants to obstruct such a parting because he cannot easily bear the loss is weak and soft by nature, and for that very reason too little fair-minded in friendship.

\ciceroSection{76} And in every matter you must consider both what you demand of a friend and what you allow to be obtained from yourself. There is also a certain misfortune in the dissolving of friendships that is sometimes unavoidable — for our discourse now slips down from the close ties of the wise to ordinary friendships. The faults of friends often break out, sometimes against the friends themselves, sometimes against outsiders, while the disgrace nevertheless redounds upon the friends. Such friendships, then, are to be washed away by a slackening of intercourse and, as I have heard Cato say, unstitched rather than torn — unless some quite intolerable wrong has flared up, so that it is neither right nor honorable, nor even possible, not to make the estrangement and rupture at once.

\ciceroSection{77} But if some change of character or of pursuits has come about, as commonly happens, or a disagreement has arisen over political allegiances — for I am speaking now, as I said a little earlier, not of the friendships of the wise but of common ones — then care must be taken that it look not merely as if a friendship has been laid aside, but as if a feud has been taken up. For nothing is more disgraceful than to wage war against a man with whom you have lived on familiar terms. Scipio, as you know, had withdrawn from his friendship with Quintus Pompeius on my account; and on account of the disagreement that there was in public affairs, he became estranged from our colleague Metellus. In both cases he acted with weight and moderation, and with no bitter resentment of spirit.

\ciceroSection{78} For this reason, the first effort must be to see that no ruptures between friends occur; but if anything of the kind does come about, that the friendships seem to have burned out rather than to have been stamped out. Real care must be taken, too, that friendships do not turn into grave feuds, out of which arise quarrels, slanders, and insults. Yet even these, if they are bearable, must be borne, and this honor paid to an old friendship: that the one who commits the wrong, not the one who suffers it, should bear the blame. In short, against all these faults and troubles there is one safeguard, one precaution: not to begin to love too quickly, nor to love those who are unworthy.

\ciceroSection{79} But the ones worthy of friendship are those who carry within themselves the very reason they should be loved. A rare breed! And indeed everything excellent is rare, and nothing is harder than to find what is in every respect perfect of its kind. Yet most people recognize nothing as good in human affairs except what turns a profit, and they love their friends, as they would cattle, chiefly for the return they hope to draw from them.

\ciceroSection{80} And so they go without that most beautiful and most natural friendship, the kind sought for its own sake and on its own account; nor do they have in themselves an example of what this power of friendship is, of its quality and its scope. For each of us loves himself, not to exact from himself some payment for his own affection, but because each is dear to himself in his own right. And unless this same feeling is carried over into friendship, a true friend will never be found; for a true friend is, as it were, a second self.

\ciceroSection{81} If this shows itself in beasts—in the creatures of air, of water, and of land, tame and wild alike—first that they love themselves (for this is born in every living thing equally), and then that they seek out and reach for others of their own kind to attach themselves to, and do so with longing and with something resembling human love—how much more does this happen in man by nature, who both loves himself and looks for another whose mind he may so blend with his own as to make almost one out of two!

\ciceroSection{82} But most people perversely—not to say shamelessly—want to have the kind of friend they cannot themselves be, and they demand from their friends what they do not give them in return. The fair course is, first, to be a good man oneself, and then to look for another like oneself. Among such men the steadiness of friendship I have so long been discussing can be made firm: when men joined by goodwill will first master those desires to which others are slaves; then will delight in fairness and justice; and each will take on every burden for the other; neither will ever ask of the other anything but what is honorable and right; and they will not only cherish and love one another, but also hold one another in reverence. For whoever strips reverence from friendship strips away its greatest ornament.

\ciceroSection{83} And so a ruinous error possesses those who think that in friendship the door stands open to every license of lust and wrongdoing. Nature gave friendship as an ally of the virtues, not a companion of the vices, so that—since virtue alone could not reach the heights—joined and allied with another it might arrive there. Wherever such a partnership exists, or has existed, or will exist among any men, theirs must be reckoned the best and most blessed company on the road to the highest good of nature.

\ciceroSection{84} This, I say, is the partnership in which all things reside that people think worth seeking—honor, glory, peace of mind, and delight—so that, when these are present, life is blessed, and without them it cannot be. Since this is the best and greatest thing, if we want to attain it, we must devote our effort to virtue, without which we can secure neither friendship nor anything else worth desiring. But those who neglect virtue, while supposing themselves to have friends, come to feel their error only at last, when some grievous misfortune forces them to put those friends to the test.

\ciceroSection{85} And so—for it must be said more than once—you should love only after you have judged, not judge after you have loved. But while we pay the penalty for our carelessness in many things, we pay it most of all in the choosing and the cultivating of friends. For we take our counsel back to front, and we do over what is already done, which the old proverb forbids. Bound up with one another, this way and that, by long familiarity or even by mutual services, we suddenly, midcourse, break off our friendships when some offense has arisen.

\ciceroSection{86} All the more, then, is such gross neglect of a thing so necessary to be condemned. For of all things in human life, friendship is the one whose usefulness all agree upon with one voice—though virtue itself is despised by many and called a kind of showmanship and display; many scorn riches, who take pleasure in a meager living and content themselves with little; and as for honors, by whose pursuit some men are set ablaze—how many despise them so completely that they think nothing emptier, nothing more trifling! And likewise with all the other things that some find dazzling: there are very many who count them as nothing. But about friendship all are of one mind, to a man: those who have given themselves to public life, those who delight in knowledge and learning, those who manage their own affairs in leisure, and finally those who have surrendered themselves wholly to pleasure—all hold that without friendship there is no life at all, provided they wish to live in any measure as free men.

\ciceroSection{87} For friendship creeps somehow through all our lives, and suffers no mode of spending one's days to be untouched by it. Indeed, even if a man is so harsh and savage in nature that he flees and hates the company of his fellows—as we are told a certain Timon of Athens once was—even he could not endure to go without someone on whom to spew out the venom of his bitterness. And this would best be judged if something of this kind could happen: that some god should take us out of this throng of men and set us down somewhere in solitude, and there, while furnishing us in abundance and plenty with everything that nature desires, should strip away all power of ever looking on a human being—who would be so made of iron as to bear that life? For whom would solitude not steal away the enjoyment of every pleasure?

\ciceroSection{88} True, then, is that saying which our elders used to recall as the habitual remark, I believe, of Archytas of Tarentum, who had it from elders of his own: that if a man should climb into the heavens and gaze upon the nature of the universe and the beauty of the stars, the wonder of it would hold no sweetness for him—wonder that would have been most delightful had he had someone to tell it to. So nature loves nothing solitary, and always leans toward some support, as it were, which in every truest friend is the sweetest of all. But though nature declares by so many signs what she wishes, what she seeks, what she longs for, still we somehow turn deaf and do not hear the warnings she gives us. For the practice of friendship is varied and many-sided, and many occasions for suspicion and offense arise, which it is the part of a wise man now to avoid, now to make light of, now to bear. There is one offense, though, that must be undergone, so that both the usefulness of friendship and its good faith may be kept: for friends must often be advised and reproved, and these things must be received in a friendly spirit, when they are done out of goodwill.

\ciceroSection{89} But somehow there is truth in what my friend says in the \textit{Andria}: "Compliance breeds friends, candor breeds hatred." Candor is a nuisance, if from it is born the hatred that is the poison of friendship; but compliance is a far worse nuisance, since by indulging a friend's faults it lets him be carried headlong to ruin. The greatest fault, however, lies with the man who spurns the truth and is driven by compliance into self-deception. In this whole matter, then, reason and care must be brought to bear: first, that admonition be free of harshness, and then that reproof be free of insult. As for compliance—since I gladly use Terence's word—let there be courtesy, but let flattery, that abettor of vices, be kept far off, unworthy not only of a friend but even of a free man. For one lives one way with a tyrant, and another way with a friend.

\ciceroSection{90} But the man whose ears are closed to the truth, so that he cannot bear to hear the truth from a friend—his recovery is past hoping for. There is that shrewd saying of Cato's, like so many of his: that some men are better served by bitter enemies than by friends who seem agreeable, since the former often speak the truth, the latter never. And there is this absurdity besides: that those who are admonished do not feel the distress they ought to feel, and feel the distress they ought to be free of. For they are not troubled at having done wrong, but they take it ill to be reproved—when it should have been the other way round: to grieve at the fault and rejoice at its correction.

\ciceroSection{91} As, then, it is the mark of true friendship both to advise and to be advised—and to do the one freely, not harshly, and to receive the other patiently, not with resistance—so it must be held that there is no plague in friendships greater than fawning, cajolery, and flattery: for under however many names this vice is to be branded, it belongs to shallow and deceitful men, who say everything to please and nothing for the truth.

\ciceroSection{92} While pretense in everything is a fault — for it destroys our judgment of the truth and corrupts it — it is most at war with friendship; for it kills the truth, and without truth the very name of friendship can have no force. Since the whole power of friendship lies in this, that out of many a single mind, so to speak, comes into being, how will that be possible if not even in any one man will there be a single mind, and the same mind always, but instead one that is shifting, changeable, manifold?

\ciceroSection{93} For what can be so pliable, so off its course, as the mind of a man who turns himself not only to another's feelings and wishes, but even to his looks and his nods? "Someone says no, I say no; he says yes, I say yes; in the end I have ordered myself to agree with everything" — so says the same Terentius, but in the character of Gnatho. To take on a friend of that kind is the very height of frivolity.

\ciceroSection{94} And there are many men like the Gnathos, who when they are superior in rank, fortune, and reputation make their flattery a heavy thing, when authority is added to their emptiness.

\ciceroSection{95} But the flattering friend can be separated from the true one and told apart, if you bring care to it, just as everything painted over and feigned can be told apart from what is genuine and true. An assembly, made up of the most untrained men, nonetheless is in the habit of judging what difference there is between a demagogue — that is, a flatterer and a worthless citizen — and one who is steady, true, and weighty.

\ciceroSection{96} With what flatteries did Gaius Papirius lately go pouring into the ears of the assembly, when he was proposing his law to make the tribunes of the plebs eligible for re-election! We spoke against it — but I will say nothing of myself; I would rather speak of Scipio. What weight, immortal gods, was in him, what majesty in his speaking! You would easily have called him the leader of the Roman people, not its follower. But you were there, and the speech is in everyone's hands. And so that demagogue's law was thrown out by the votes of the people. And — to come back to myself — you remember that when Quintus Maximus, Scipio's brother, and Lucius Mancinus were consuls, how popular the law on the priesthoods proposed by Gaius Licinius Crassus seemed; for the filling of vacancies in the colleges was being transferred to the people as a favor. (And he was the first to bring in the practice of facing toward the Forum when addressing the people.) Yet, with us defending it, reverence for the immortal gods easily prevailed over his salable oratory. And this was done when I was praetor, five years before I was made consul. So that cause was defended more by its merits than by any high authority.

\ciceroSection{97} But if on the stage — that is, in the assembly, where there is the widest room for things invented and shadowed out — the truth nonetheless prevails, provided only that it is brought into the open and made plain, what ought to happen in friendship, which is weighed wholly by truth? In it, unless you see, as the saying goes, an open heart and show your own, you can hold nothing as trustworthy, nothing as proven — you cannot even love or be loved, since you do not know how truly that is done. And yet that flattery, however ruinous it may be, can harm no one except the man who takes it in and delights in it. So it turns out that the man who opens his ears most to flatterers is the one who flatters himself and delights in himself most of all.

\ciceroSection{98} Virtue, to be sure, loves itself; for it knows itself best of all and understands how lovable it is. But I am not speaking now of virtue, but of the opinion of virtue. For not so many wish to be endowed with virtue itself as wish to seem so. These men flattery delights; when talk shaped to their own wishes is brought to them, they suppose that empty speech to be testimony to their merits. There is, then, no friendship in this, when the one is unwilling to hear the truth and the other is ready to lie. Nor would the flattery of the parasites in our comedies seem witty to us, were there not braggart soldiers. "Does Thais really give me her great thanks?" It would have been enough to answer "Great." "Enormous," he says. The flatterer always magnifies the thing that the man to whose wishes he speaks wants to be great.

\ciceroSection{99} For this reason, although that flattering emptiness has its force among those who themselves lure it on and invite it, even the weightier and steadier men must still be warned to watch that they are not caught by cunning flattery. The man who flatters openly no one fails to see, unless he is utterly out of his wits; but against that cunning, hidden one we must take careful guard, so that he does not worm his way in. For he is not easily recognized, since he often flatters even by opposing, and feigning to quarrel he fawns, and in the end gives in and lets himself be beaten, so that the one who has been fooled may seem to have seen the more. And what is more disgraceful than to be fooled? To keep that from happening one must guard all the harder. "How you have played me today before all the foolish old men of comedy, and fooled me in the grandest style!"

\ciceroSection{100} For even in plays this is the most foolish character, that of the improvident and credulous old men. But somehow or other my discourse has slipped away from the friendships of perfected men — that is, of the wise (I mean here the kind of wisdom that seems able to fall to a human being) — down to the trivial friendships. So let us return to those first matters and at last bring them to a close. Virtue, I say, Gaius Fannius, and you, Quintus Mucius — virtue both wins friendships and keeps them. For in it lies the harmony of things, in it stability, in it constancy; and when it has raised itself up and shown its own light, and has seen and recognized that same light in another, it moves toward it and in turn receives what is in the other, and from this is kindled either love or friendship; for both words come from loving. And to love is nothing other than to cherish the very one you love, with no need felt, no advantage sought. And yet that advantage too flowers out of friendship, even if you have pursued it less.

\ciceroSection{101} With this goodwill we, as young men, loved those old men — Lucius Paulus, Marcus Cato, Gaius Gallus, Publius Nasica, Tiberius Gracchus, the father-in-law of our Scipio. It shines out still more brightly among equals, as between me and Scipio, Lucius Furius, Publius Rupilius, Spurius Mummius; and we in turn, as old men, find rest in the affection of the young, as in yours, as in that of Quintus Tubero; for my part I take delight even in the close friendship of the very young men Publius Rutilius and Aulus Verginius. And since the plan of our life and nature is so arranged that one generation rises after another, surely it is most to be wished that you can reach the finish line, as the saying goes, with the same companions with whom you set out, as if from the starting gate.

\ciceroSection{102} But since human things are fragile and fleeting, we must always be seeking out some whom we may cherish and by whom we may be cherished; for once affection and goodwill are taken away, all delight is taken out of life. For my part, though Scipio was suddenly torn from me, he lives still, and always will live; for it was the virtue of that man that I loved, and that has not been put out. Nor is it set before my eyes alone, who always held it in my hands, but it will be famous and distinguished even for those who come after. No one will ever take up anything greater in mind or hope who does not think he must set before himself the memory and the image of that man.

\ciceroSection{103} For my part, of all the things that either fortune or nature has granted me, I have nothing that I could compare with the friendship of Scipio. In it I found agreement about public affairs, in it counsel in private matters, in that same friendship a rest full of delight. Never did I offend him in even the smallest thing — at least so far as I was aware; never did I myself hear from him anything I would not have wished. We had one home, the same fare and that shared, and not only our military service but our travels abroad and our retreats to the country were shared.

\ciceroSection{104} For why should I speak of our pursuits — always to learn something and to discover it — in which, withdrawn from the eyes of the people, we wore away all our leisure time? If the recollection and memory of those things had fallen with him, in no way could I bear the longing for a man so closely joined to me and so loving. But those things are not put out; rather they are fed and increased by my reflection and my memory, and even if I were utterly bereft of them, my very age brings me great solace; for I can no longer be in this longing for very long, and all brief things ought to be bearable, even if they are great. This is what I had to say about friendship; but I urge you so to place virtue — without which friendship cannot exist — that, virtue excepted, you think nothing more excellent than friendship.
