% De Officiis (de-officiis) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroBook{1}

\ciceroSection{1} Although you, Marcus, my son, have now been studying for a year under Cratippus, and in Athens at that, you must be well supplied by this point with the teachings and principles of philosophy — given both the great authority of your teacher and the great prestige of the city, one of which can enrich you with knowledge, the other with examples. Nevertheless, just as I have always combined Greek learning with Latin for my own benefit, and have done this not only in philosophy but in the practice of oratory as well, I think you should do the same, so that you are equally capable in both languages. And it is we who seem, in this regard, to have done our countrymen great service — so that those who are not merely unacquainted with Greek letters but even those who are learned in them think they have gained something substantial both for speaking and for judging.

\ciceroSection{2} You will learn, then, from the foremost philosopher of this age, and you will go on learning as long as you choose; you ought to choose as long as you feel no dissatisfaction with the progress you are making. But all the same, when you read my writings — which diverge not greatly from those of the Peripatetics, since both they and I claim to follow Socrates and Plato — form your own judgment on the substance (I put no obstacle in your way), but your Latin will certainly be enriched by reading what I have written. I should not wish this to be taken as arrogance. For though I yield to many in knowledge of philosophy, to claim what is the orator's special province — to speak aptly, with precision and with grace — since I have spent a lifetime at that pursuit, I seem to claim it with some right of my own.

\ciceroSection{3} I urge you therefore strongly, my Cicero, to read not only my speeches but these books on philosophy as well, which are now nearly equal to them in number; for though the speeches carry more force, this calm and measured style of philosophical prose is also worth cultivating. And this I see no Greek has yet achieved — that the same man labored in both kinds and pursued both the style of the forum and this quieter mode of disputation — unless perhaps Demetrius of Phalerum may be counted in that number: a subtle debater, not much of an orator in point of force, agreeable nonetheless — enough that you can recognize in him a pupil of Theophrastus. As for how far I have advanced in both, let others decide; I have certainly pursued both.

\ciceroSection{4} For my part I believe that Plato, had he chosen to handle the style of public oratory, could have spoken with the greatest weight and abundance; and I believe Demosthenes, had he retained and been willing to set forth what he had learned from Plato, could have done so with grace and splendor. I judge likewise of Aristotle and Isocrates, each of whom was so taken with his own pursuit that he looked down on the other's. But when I resolved to write something for you at this time — much more to follow — I wished most of all to begin with what is best suited both to your age and to my own standing. For although much that is weighty and profitable has been carefully and abundantly treated in philosophy, the topics that have been handed down and taught concerning duty seem to extend most widely. No part of life — in public affairs, in private, in the forum, in the home, whether you are acting alone or entering into dealings with another — can be free of duty; and in the cultivation of duty lies the whole of life's honor, in the neglect of it, disgrace.

\ciceroSection{5} Now this inquiry is shared by all philosophers — for who would dare call himself a philosopher if he offers no teaching on duty? Yet there are certain schools which, by laying down the limits of good and evil as they do, overturn the whole concept of duty. For a man who defines the highest good in such a way that it has no connection with virtue, and measures everything by his own convenience rather than by what is honorable — if he is consistent with himself and is not sometimes overcome by the goodness of his own nature — he can have no place for friendship, no place for justice, no place for generosity. And a man who counts pain as the highest evil cannot be brave; a man who counts pleasure as the highest good cannot be self-controlled — that at least is impossible.

\ciceroSection{6} Although this is so plain as to need no argument, we have argued it elsewhere. These schools, then, if they wish to remain consistent, can say nothing about duty, nor can any firm, stable, and nature-grounded precepts of duty be transmitted except by those who hold that the honorable is to be sought either for its own sake alone or at the very least above everything else. This teaching accordingly belongs properly to the Stoics, the Academics, and the Peripatetics — since the position of Aristo, Pyrrho, and Erillus was long ago rejected; and yet they would have had a right to discuss duty if they had left some power of discernment between things, so that there might be a path by which duty could be discovered. For the present inquiry, therefore, we shall follow above all the Stoics — not as translators, but drawing as we are accustomed to do, from their springs, by our own judgment and discretion, as much as seems right in each case.

\ciceroSection{7} Since the whole discussion is to concern duty, it is fitting, I think, to define first what duty is — and I am surprised that Panaetius omitted this. For every systematic inquiry undertaken by reason concerning a subject ought to begin from a definition, so that it is understood what the thing under discussion actually is. Every inquiry concerning duty is twofold: one kind pertains to the end of goods, the other is set out in precepts by which conduct may be shaped in every area of life. The following are examples of the first kind: whether all duties are fulfilled duties; whether one duty is greater than another; and others of the same type. The precepts of duty, though they do bear on the end of goods, make that less obvious, because they seem to bear more directly on the shaping of common life; and it is on these that we must set out our account in these books. There is also another division of duty.

\ciceroSection{8} For there is what is called a middle duty and a fulfilled duty. The fulfilled duty I think we may call \textit{right} — since the Greeks call it \textit{[Greek: katorthoma]}, while they call the common duty \textit{[Greek: kathekon]}. And they define these as follows: what is right they define as duty fulfilled; the middle duty, they say, is that for which a plausible reason can be given why it was done.

\ciceroSection{9} The deliberation involved in reaching a decision is therefore threefold, as Panaetius sees it. For what falls under deliberation is a question either of whether the thing is honorable or dishonorable to do — and in weighing this the mind is often pulled in opposite directions; or a question of whether what is being deliberated upon conduces to the convenience and pleasantness of life, to resources and means, to influence and power — things by which a man can help himself and those dear to him; all such deliberation falls under the reasoning of expediency. The third kind of doubt arises when what appears expedient seems to conflict with what is honorable — when expediency seems to pull the mind toward itself while honor calls it back the other way, so that the mind is torn in deliberating and is laden with the anxious burden of indecision.

\ciceroSection{10} In this division, since to pass over anything important is the chief fault in a classification, two cases have been left out. For deliberation is not only about whether something is honorable or dishonorable, but also, when two honorable courses are presented, which is the more honorable; and likewise, when two expedient courses are presented, which is the more expedient. The scheme Panaetius thought was threefold turns out, on examination, to require distribution into five parts. First, then, we must speak about the honorable — but in two aspects; then, on the same plan, about the expedient; then about the comparison of the two.

\ciceroSection{11} To begin with: nature has granted to every living creature the instinct to protect itself, its life and its body; to avoid what seems likely to harm it; and to seek out and obtain everything necessary for life — food, shelter, and things of that sort. Common likewise to all living creatures is the drive toward union for the purpose of reproduction, and a degree of care for the offspring produced. But between man and the beast the chief difference is this: the beast adapts itself, moved only by sensation, only to what is immediately before it, scarcely registering past or future; man, sharing in reason, by which he perceives consequences, sees the causes of things, grasps their antecedents and precursors, compares likenesses, and joins and connects the future to the present — easily sees the whole course of his life and prepares what is needed to live it.

\ciceroSection{12} And the same nature, by the power of reason, draws man to man and into the fellowship of speech and of life; it implants above all else a special tenderness toward those who have been born of us, and moves him to desire that gatherings and assemblies of men should exist and be frequented by him — and for these reasons to strive to provide what serves civilization and sustenance, not for himself alone but for his wife, his children, and all others whom he holds dear and feels bound to protect; this concern stirs the spirit too and makes it the greater for action.

\ciceroSection{13} And it is especially proper to man to search and investigate the truth. And so when we are free from necessary business and cares, we long to see, to hear, to learn something, and we count knowledge of things that are obscure or wonderful as necessary to the happy life. From this we understand that what is true, simple, and sincere is best suited to human nature. To this desire for seeing the truth is joined a kind of longing for leadership — a well-formed mind by nature is unwilling to obey anyone except one who teaches, instructs, or commands with justice and lawfully for the sake of the common good; from this arises greatness of spirit and the contempt of human circumstances.

\ciceroSection{14} Nor is that force of nature and of reason small: that man alone among animals perceives what order is, what fitting conduct is, what measure ought to govern both words and deeds. And so, while no other animal has any sense of beauty, grace, and the harmony of parts in the things perceived by sight, nature and reason transfer this perception from the eyes to the mind and judge that beauty, consistency, and order are to be maintained all the more in plans and actions — taking care to do nothing unseemly or unmanly, and in all opinions and acts to do and think nothing wantonly. From these elements there is fashioned and produced what we are seeking: the honorable — which, even if it should win no recognition, is still honorable; and which, as we rightly say, though praised by no one, is by nature praiseworthy.

\ciceroSection{15} You see now, Marcus, my son, the very form and, as it were, the face of the honorable — which, says Plato, if it could be seen with the eyes, would inspire wondrous love of wisdom \textit{[Greek: sophias]}. But everything that is honorable arises from one of four sources: it belongs either to the perception and discernment of truth; or to the preservation of human fellowship, the rendering of what is due to each person, and fidelity in agreements and dealings; or to the greatness and strength of a great and unconquered spirit; or to order and measure in all that is done and said, in which reside self-control and restraint. Though these four are bound together and interwoven with one another, yet from each separately certain kinds of duty arise — from the first part, which we have described and in which we place wisdom and practical judgment, comes the investigation and discovery of truth; and that is the special function of this virtue.

\ciceroSection{16} For the person who most clearly discerns what in any given matter is most true, and who can most keenly and swiftly both see and set out the reasoned account of it, is rightly considered the most practically wise and sagacious. This virtue has truth, accordingly, as the material it works with and in which it moves.

\ciceroSection{17} The remaining three virtues, however, have practical necessities set before them — to secure and maintain the things by which the conduct of life is sustained: so that the fellowship and bond of human society may be preserved, and that greatness and elevation of spirit may shine forth, both in increasing resources and advantages for oneself and one's own, and still more in looking down on those very things with contempt. Order, consistency, restraint, and whatever is akin to these belong to the category that requires not merely a movement of the mind but a kind of action. For it is by bringing a certain measure and order to the affairs we handle in life that we shall preserve what is honorable and fitting.

\ciceroSection{18} Of the four heads into which we have divided the nature and force of the honorable, the first — which consists in the knowledge of truth — touches human nature most closely. For we are all drawn and led by desire for learning and knowledge, in which to excel we count as fine; while to stumble, to err, to be ignorant, to be deceived we regard as both bad and shameful. In this category — natural and honorable as it is — two faults must be avoided: the first, that we do not take what is unknown for what is known and assent to it rashly; and whoever wishes to escape that fault (as all should wish) will bring both time and care to the examination of matters.

\ciceroSection{19} The second fault is that some men devote too great an enthusiasm and too much effort to subjects that are obscure and difficult, and for that reason unnecessary. When those faults are avoided, the work and care spent on subjects that are honorable and worthy of inquiry will rightly be praised: as we have heard of Gaius Sulpicius in astronomy, as we ourselves have known Sextus Pompeius in geometry, as many men in dialectic, and still more in civil law — all pursuits that move within the investigation of truth. Yet to be drawn away from affairs of life by devotion to such inquiry is against duty. The whole praise of virtue lies in action; and yet action is often interrupted, and ample time is given to return to learning — besides, the activity of the mind, which never rests, can keep us engaged in the pursuits of knowledge even without our deliberate effort. All our thinking and all the movement of the mind will be engaged either in forming plans about honorable matters that bear on living well and happily, or in the pursuit of learning and knowledge. And so much for the first source of duty.

\ciceroSection{20} Of the remaining three, the most wide-ranging is that account by which human society and what we might call the common life of men is held together. Its parts are two: justice, in which virtue's splendor is greatest, and by which good men earn their name; and, joined to it, generosity — which may also be called kindness or liberality. But the first duty of justice is this: that no one should harm another unless provoked by wrong; and then, that the common stock be used for the common good, while what is private is used as one's own.

\ciceroSection{21} Nothing, however, is private by nature; things become private either through ancient occupation — as when men came in earlier times into vacant land — or through conquest in war, or through law, agreement, terms, or lot. Hence it comes about that the land around Arpinum is called the Arpinates', the land around Tusculum the Tusculans', and the division of private holdings follows the same principle. Since what had been common by nature becomes each man's own through the process by which it falls to each, let every man hold what has come to him; and if anyone reaches out to seize what belongs to another, he violates the law of human society.

\ciceroSection{22} But since, as Plato has written with such distinction, we are not born for ourselves alone — our country claims a part of our origin, our friends claim a part — and since, as the Stoics hold, all things that grow upon the earth are created for the use of men, and men are generated for the sake of other men, that they may be able in turn to benefit one another: in this we must follow nature as our guide, to bring common advantages into the common pool, to bind the fellowship of man with man by the exchange of duties — giving and receiving — now through skills, now through labor, now through resources.

\ciceroSection{23} The foundation of justice, however, is good faith — that is, steadiness and truth in what is said and pledged. And so, though this may perhaps strike someone as rather harsh, let us nonetheless dare to imitate the Stoics, who pursue word-derivations with such care, and believe that good faith (\textit{fides}) takes its name from the fact that what is said comes to pass (\textit{fiat quod dictum est}). There are, moreover, two kinds of injustice: the first, of those who commit it; the second, of those who, when they can, fail to drive off injustice from those against whom it is committed. For the man who unjustly attacks another — stirred by anger or some other passion — is as it were laying hands on a fellow member of society; while the man who does not defend or resist injustice when he can is just as much at fault as if he abandoned his parents, his friends, or his country.

\ciceroSection{24} Now those wrongs that are inflicted deliberately for the purpose of harming often spring from fear, when the man who is planning harm to another is afraid that, unless he acts, he will himself suffer some injury. But the greater part of those who go about committing injustice do so in order to get what they have desired — and in this vice, greed spreads most widely of all.

\ciceroSection{25} Wealth is pursued both for the necessities of life and for the enjoyment of its pleasures. In those of larger spirit, however, the desire for money aims at power and the capacity to confer favors — as Marcus Crassus declared not long ago that no sum was large enough for the man who wished to stand first in public life, if he could not from its revenues support an army. Magnificent appointments and a style of life full of elegance and abundance also give their pleasure; and from all this the desire for money becomes boundless. Yet the increase of private fortune — when it harms no one — is not a thing to be condemned; injustice, however, must always be avoided.

\ciceroSection{26} Most men, though, are chiefly led to forget justice when they fall into desire for commands, honors, and glory. For what Ennius has written — \textit{In kingship there is no sacred fellowship, no keeping faith} — applies more widely still. For in whatever matter of this kind it is not possible for several men to excel, the contest that arises is usually so fierce that it becomes extremely difficult to keep the bond of fellowship intact. The recklessness of Gaius Caesar demonstrated this lately: he overturned every law, human and divine, for the sake of that primacy which he had conjured up for himself out of a mistaken notion of greatness. What makes this the more troubling is that it is precisely in the greatest and most brilliant natures that desire for honor, command, power, and glory tends to arise. All the more reason, then, to take care that nothing goes wrong in this regard.

\ciceroSection{27} In all injustice, however, there is the greatest difference between whether the wrong is done under some agitation of the mind — which is generally brief and temporary — or with deliberation and forethought. Those wrongs that arise from some sudden impulse are lighter than those that are planned and prepared in advance. And so much for the doing of injustice.

\ciceroSection{28} For failing to defend others and abandoning one's duty, there are usually many excuses: men are unwilling to take on enmities, or labor, or expense; or they are hindered by negligence, idleness, inertia; or they are so caught up in their own pursuits and occupations that they let those they ought to protect go undefended. One must therefore take care that what Plato says of philosophers — that they are just because they are engaged in investigating truth, and because they despise and count as nothing the things that most men eagerly desire and commonly fight over — may not fall short. For they achieve one kind of justice, in that they commit no injustice themselves, but they fall into the other kind: impeded by the passion for learning, they desert those they ought to protect. And so Plato thinks they will not even enter public life unless compelled. It would be better, though, that they do so willingly — for a just act is truly just only when it is done voluntarily.

\ciceroSection{29} There are also those who claim, either from zeal for managing their own estate or from a kind of dislike of their fellow men, that they attend to their own business and wrong no one, it seems, in any way. They are free from one kind of injustice, but they fall into the other: they desert the fellowship of life, for they contribute nothing to it — no zeal, no labor, no resources. Since, then, we have set out two kinds of injustice and attached the causes belonging to each, and have established beforehand the principles on which justice rests, we shall easily be able — unless we are excessively fond of ourselves — to judge what our duty is in any particular situation;

\ciceroSection{30} for concern for others' affairs is indeed difficult. Yet the Chremes of Terence thinks nothing human is foreign to him; but all the same, because we perceive and feel more keenly what happens to ourselves, whether for good or ill, than what happens to others — which we see as it were across a long interval of distance — we judge those cases differently from our own. Hence good counsel is given by those who forbid us to do anything that we are uncertain is right or wrong. For what is fair shines of itself; hesitation signals the thought of injustice.

\ciceroSection{31} But occasions often arise when things that seem most worthy of a just man — of the man we call good — are reversed and become their opposite: so that to return a deposit, to honor a promise, and all the things that bear on truthfulness and good faith, sometimes become just to depart from and not to observe. For in such cases we must look back to those foundations of justice I laid down at the start: first, that no one be harmed; second, that the common good be served. When circumstances change, duty changes, and it is not always the same.

\ciceroSection{32} For it can happen that some promise or agreement is such that to carry it out would serve neither the person for whom it was made nor the one who made it. For if, as in the old stories, Neptune had not fulfilled what he promised Theseus, Theseus would not have been robbed of his son Hippolytus; for of the three wishes — as the tale runs — this was the third, which he claimed in anger at Hippolytus's death: once it was granted, he fell into the deepest grief. Promises, then, must not be kept when keeping them would harm those to whom the promise was made; and neither is it against duty to weigh a greater interest against a lesser, when keeping your word would harm yourself more than it would benefit the one to whom it was given — as, for instance, if you had arranged to appear in court on someone's behalf and in the meantime your son had fallen seriously ill, it would not be against duty to fail to do what you said; and the man to whom the promise was made would be the one departing from duty, if he complained of being let down. And who cannot see that promises which a man made under duress and fear, or which he was tricked by fraud into making, need not be kept? Most of these are released by the praetor's law, some by statute.

\ciceroSection{33} Injustice arises, too, from a kind of sophistry — an over-clever but malicious interpretation of the law. This is the source of that proverb, now worn smooth by use: the height of right is the height of wrong. In public affairs the same fault is committed in great abundance — as by the general who, when a thirty-day truce had been agreed with the enemy, ravaged the fields by night, on the ground that the truce was for days, not nights. Nor is our own man to be commended, if the story is true — and I have nothing but hearsay — that Quintus Fabius Labeo (or whoever it was) was appointed by the Senate as arbitrator in the boundary dispute between the people of Nola and those of Naples. He went to the site and spoke separately with each party, urging them not to pursue the matter greedily or grasping, but to give ground rather than to press forward. When each side had done so, a stretch of land was left unclaimed in the middle. He fixed their respective boundaries as they had themselves described them; the strip left in between he awarded to the Roman people. That is deception, not arbitration. In every matter, therefore, such clever contrivance must be avoided. There are also duties to be kept toward those by whom you have received an injury. For vengeance and punishment have their proper measure; and I am inclined to think it enough that the man who provoked you should repent his wrong, so that he himself may do nothing of the kind again, and the rest may be slower to inflict injury.

\ciceroSection{34} In public affairs the laws of war must be preserved above all else. For since there are two ways of settling a dispute — by argument and by force — and since argument is the proper method of human beings, while force is the method of beasts, we must have recourse to the latter only when the former is not available to us.

\ciceroSection{35} Wars, then, are to be undertaken for this reason and no other: that we may live in peace without injustice; but once victory is won, those who bore themselves in war without cruelty and without savagery must be spared — as our ancestors received the Tusculans, the Aequi, the Volsci, the Sabines, and the Hernici into the citizen body — while Carthage and Numantia they destroyed utterly. I would have wished Corinth spared; I think, however, they had some purpose in view, above all the strategic position of the place, which they feared might one day of itself invite war. In my own judgment, peace, on terms that contain no treachery, is always to be our aim. Had I been heeded in this, the republic, if not in its finest condition, would at any rate be something — whereas now it is nothing. And as we must have consideration for those whom we have defeated in war, so too must those who lay down their arms and take refuge in the good faith of a commander be received — even when the ram has already breached the wall. In this matter justice was so scrupulously observed by our own people that those who had received cities or nations defeated in war under pledge of good faith became, by ancestral custom, the patrons of those peoples.

\ciceroSection{36} The terms of just war have been set down with the greatest solemnity in the fetial law of the Roman people. From this it may be understood that no war is just unless it is fought either after a formal demand for restitution or with prior announcement and declaration. Popilius was in command of a province, and Cato's son was serving in his army as a recruit. When Popilius decided to discharge one of his legions, he discharged Cato's son as well, since the young man was serving in that same legion. But because the young man remained with the army out of eagerness to fight, his father Cato wrote to Popilius asking that, if he was willing to allow him to remain, he bind him by a fresh military oath — since the first had been dissolved, and without it he had no right in law to fight against the enemy. So strict was the observance in the matter of going to war.

\ciceroSection{37} There is, moreover, a letter of the elder Marcus Cato to his son Marcus, in which he writes that he has heard the young man has been discharged by the consul while serving as a soldier in Macedonia in the war against Perseus. He therefore warns him to take care not to enter battle; for, he says, it is not lawful for a man who is no longer a soldier to fight the enemy. There is something else I note here as well: that the man who would properly be called a public enemy \textit{[Greek: polemios]} was in older usage called \textit{hostis}, the harsher fact softened by the gentler word. For \textit{hostis} in the usage of our ancestors meant what we now call \textit{peregrinus}, a foreigner. The Twelve Tables bear witness: \textit{aut status dies cum hoste} — "or the appointed day with the foreigner" — and again: \textit{adversus hostem aeterna auctoritas} — "against the foreigner, title holds good for ever." What greater clemency could be added than this: to give the man against whom you are fighting so mild a name? Yet time has made the word harsher; it has retreated from the sense of "foreigner" and remained fixed specifically on the man who bears arms against you.

\ciceroSection{38} When the contest is for empire and glory in war is what is sought, the same causes I described a moment ago as making a war just must nevertheless be present. But wars fought for glory and empire are to be prosecuted less harshly. For just as in a quarrel with a fellow citizen we behave differently toward an enemy than toward a rival — with the one it is a contest of honor and standing, with the other of life and reputation — so the war against the Celtiberians and the war against the Cimbrians were fought as against enemies: the question was which people should survive, not which should hold dominion. With the Latins, the Sabines, the Samnites, with the Carthaginians, with Pyrrhus, the struggle was for empire. The Carthaginians were breakers of treaties, and Hannibal was cruel; the rest were men of a more honorable kind. Pyrrhus in particular, on the ransoming of prisoners, spoke those celebrated words:

\begin{quote}
Neither gold I seek, nor price shall you give me; not as traffickers in war but as warriors let us both settle the matter by iron, not gold, which of us Fortune wills to reign and what she brings. And hear this too: those whose valor Fortune has spared in war — their freedom, too, I am resolved to spare. Take them, lead them away, as the great gods will.
\end{quote}

A royal utterance, worthy of the lineage of the Aeacidae.

\ciceroSection{39} Moreover, if individuals have given the enemy their word, under the pressure of the moment, good faith toward it must be kept. So it was in the First Punic War with Regulus, taken prisoner by the Carthaginians and sent to Rome on the question of an exchange of captives, under oath to return. He came, spoke in the Senate against the return of the captives, and then, when his kinsmen and friends tried to detain him, chose to return to his punishment rather than break the good faith he had pledged to an enemy.

\ciceroSection{40} In the Second Punic War, after the battle of Cannae, Hannibal sent to Rome ten men bound by oath to return unless they obtained leave to ransom the prisoners. The censors kept all of those who had broken their oath — every one of them, for as long as each lived — in the roll of the degraded; and no less so the man who had devised a fraudulent way out of his oath. He had left the Carthaginian camp with Hannibal's permission, returned shortly after on the pretext of having forgotten something, and then, having once more left the camp, supposed himself free of the oath on the grounds that he had technically come back. He was right in words; in substance he was not. In keeping faith we must always consider what we meant, not merely what we said. The greatest of all the examples of justice toward an enemy was set by our ancestors when a deserter from Pyrrhus came to the Senate and promised to poison the king and put him to death; the Senate and Gaius Fabricius had the man returned to Pyrrhus. They would not approve even the destruction of a powerful enemy who had brought war upon them, if that destruction came by a crime.

\ciceroSection{41} Enough has been said of the duties that belong to war. Let us remember, however, that justice must be kept even toward the humblest. The humblest condition and fortune is that of slaves; and those who instruct us to treat them as we treat hired men give sound advice — work to be required, what is fair to be rendered. Since injustice is committed in two ways — by force or by fraud — and fraud resembles a fox, force a lion, both are wholly foreign to a human being, but fraud deserves the greater hatred. And of all injustice there is none more ruinous than that of men who, precisely at the moment they are most deeply deceiving, act so as to appear men of integrity. Enough of justice.

\ciceroSection{42} Next, as was proposed, let us speak of generosity and kindness — nothing is more suited to human nature, and yet it is attended by many cautions. First, we must see that our generosity does no harm either to those who seem to benefit from it or to others; second, that it does not exceed our means; and third, that benefits be distributed according to the worth of each recipient — for this is the foundation of justice, to which all of this must be referred. Those who confer a benefit on someone that harms the very person they appear to wish to help are to be judged not generous or kind but destructive flatterers; and those who harm some in order to be liberal toward others are guilty of the same injustice as men who convert another's property to their own use.

\ciceroSection{43} There are many, however — men eager for splendor and glory — who rob some to lavish on others, and who imagine they will pass for generous toward their friends if they enrich them by whatever means they can. This falls so far short of duty that nothing can be more contrary to duty. We must take care, then, to practice a generosity that benefits our friends and harms no one. The transfer of wealth carried out by Lucius Sulla and Gaius Caesar from their rightful owners to others ought not, therefore, to be thought generosity; for nothing is generous that is not at the same time just.

\ciceroSection{44} The second caution was that our generosity should not exceed our means. Those who wish to be more open-handed than their resources allow sin, first, by being unjust toward their own closest kin — for the resources that would more equitably be kept available for them and left to them are instead diverted to strangers. In such generosity there generally lurks also a desire to seize and take by violence, so that there may be enough in hand for largesse. One may observe, too, that most people are driven not so much by a genuinely generous nature as by a certain love of glory, so that they may be seen as beneficent — doing many things that appear to proceed from display rather than from will. Such a pretense has more to do with vanity than with generosity or honor.

\ciceroSection{45} The third point was that there should be discrimination of worth in bestowing benefits. In this the character of the person to be benefited must be weighed, and his feeling toward us, and the community and fellowship of our lives, and the services he has previously rendered toward our interests. It would be ideal if all these requirements converge in one person; where they do not, the greater number of causes, and the weightier, will carry more.

\ciceroSection{46} Since, however, we live not with men who are perfect and fully wise, but with those in whom it is a fine thing if only the semblances of virtue appear, I think we must also understand this: that no one who shows any sign of virtue is to be wholly passed over, but that each person is to be cultivated most whose character is most adorned by these gentler virtues — self-control, restraint, and that very quality of justice of which so much has already been said. For greatness of spirit and courage in a man who is not perfect and not wise tends generally to run too hot; whereas those other virtues are more truly the mark of a good man. So much for character.

\ciceroSection{47} As for goodwill — how much of it a person bears toward us — our first duty is to give most to the one by whom we are most loved; but we should judge goodwill not in the fashion of young men, by some ardent passion, but rather by its steadiness and constancy. If there are services that call not for beginning a kindness but for returning one, still greater care must be taken; for no duty is more binding than the repayment of gratitude.

\ciceroSection{48} If, as Hesiod commands, we ought to return what we have received on loan in even greater measure, when we can, what should we do when we have been moved by a benefit? Should we not imitate the fertile fields, which give back far more than they received? For if we do not hesitate to perform good offices for those whom we hope will be of use to us in the future, what ought we to be toward those who have already been of use? For since there are two kinds of generosity — one that gives a benefit, and one that returns one — whether we give or not lies in our own power; but for a man of integrity, not to return is not permitted, provided he can do so without injustice.

\ciceroSection{49} In the receiving of benefits, however, a selection must be made, and there is no doubt that to whoever has done the most, the most is owed. Yet in this reckoning, what must be weighed above all is the spirit, the eagerness, the goodwill with which each person acted. For many do much out of a kind of recklessness — without judgment, or out of some compulsion toward everyone, or driven by a sudden gust of passion in the mind; such benefits ought not to be counted as great as those that are conferred with deliberation, care, and consistency. But in bestowing a benefit and in returning thanks — if all else is equal — duty demands above all that we assist most urgently whoever is most urgently in need; and yet the majority do exactly the opposite: they pay court above all to the one from whom they have the greatest expectations, even if he has no need of them at all.

\ciceroSection{50} The best way to preserve the fellowship and union of mankind is this: to bestow the greater part of one's kindness on those who stand closest to us. But we must, it seems, go back further and trace the first principles that nature gives us of human community and fellowship. There is, first, that which is seen in the society of the whole human race. Its bond is reason and speech — which, through teaching and learning, through sharing, through debate, through judgment, brings human beings into accord with one another and unites them; and in nothing are we more removed from the nature of wild beasts, in which we often grant the presence of courage — as in horses and lions — but not of justice, equity, or goodness, for they are without reason and without speech.

\ciceroSection{51} The fellowship that reaches widest among human beings — all persons with all persons — is this one. In it, all things that nature has produced for the common use of mankind must be held in common, such that whatever has been distributed by civil law and by statute is held according to what those laws have established, while everything else is observed according to the Greek proverb: among friends, all things are common. And those things which belong to all human beings seem to be of the kind described by Ennius in a single instance that may be extended to a great many: \textit{The man who kindly shows the way to one who wanders, lights, as it were, another's torch from his own; it gives the other light, yet loses none itself.} From this one case he gives sufficient instruction that whatever can be shared without loss ought to be given even to a stranger.

\ciceroSection{52} From this arise those common duties: not to block a flowing stream, to let another take fire from one's fire, to give faithful counsel to one who deliberates — things useful to those who receive them, not burdensome to the one who gives. These, then, we should both practice and always bring something to bear toward the common good. But since the resources of individuals are slender and the multitude of those who need them is without limit, ordinary generosity must be measured by that limit Ennius names: \textit{it loses none itself} — so that we retain the means to be generous to those who belong to us.

\ciceroSection{53} There are, however, several degrees of human fellowship. Leaving aside that one of infinite extent, a narrower bond is that of the same people, nation, and tongue — by which above all else human beings are united. Narrower still is membership in the same city-state: citizens share a great deal — the forum, temples, colonnades, streets, laws, rights, courts, elections, and besides these customs, close associations, and the dealings and transactions of many with many. Closer still is the bond of kinship; from that vast fellowship of the human race, it contracts into something small and confined.

\ciceroSection{54} For since it is nature's gift to all living creatures to possess the drive to reproduce, the first society is in marriage itself; the next, in children; then the single household with all things held in common — and this is the seed-bed of the city and, as it were, the nursery of the commonwealth. Next come the bonds between brothers, then those between cousins and second cousins, who, when they can no longer be contained within one house, go out into other houses as if into colonies. Marriages and ties by alliance follow, from which still more relatives arise; and this branching and propagation is the origin of political communities. Blood-kinship binds human beings and holds them fast through goodwill and affection —

\ciceroSection{55} for it is a great thing to share the same memorials of ancestors, to worship at the same rites, to have burial grounds in common. But of all fellowships, none is more excellent, none more firm, than when men of good character, united in moral likeness, are bound together in intimacy; for that honorable quality which we so often speak of, even when we see it in another, still moves us and makes us friends of the one in whom it appears to dwell.

\ciceroSection{56} And although every virtue draws us toward itself and causes us to love those in whom it appears to reside, yet it is justice and generosity that accomplish this above all. Nothing, moreover, is more lovable or more binding than a likeness of good character; where the same pursuits and the same desires exist, each person takes as much delight in the other as in himself, and what Pythagoras wishes for in friendship is achieved: from many, one is made. Great too is the community of fellowship that is formed by benefits given and received in mutual exchange — which, so long as they remain reciprocal and welcome, binds together in firm fellowship those among whom they pass.

\ciceroSection{57} But when you survey everything by reason and reflection, of all fellowships none is weightier, none more dear, than the one each of us has with the commonwealth. Dear are parents, dear are children, kinsmen, close friends; but all the affections of all are gathered up by the fatherland alone, for which what good man would hesitate to meet death, if by so doing he would be of service to her? How much more hateful, then, is the savagery of those men who have torn the fatherland apart with every kind of wickedness and are, and have been, bent on obliterating her utterly.

\ciceroSection{58} But if we must make a ranking and comparison of those to whom the most duty is owed, first come the fatherland and our parents, to whose vast benefits we are bound; next our children and the whole household, which looks to us alone and can have no other refuge; after them, kinsmen who are well disposed, with whom fortune too is most often shared. The necessities of life, then, are owed above all to those I have named; but the shared life of daily existence, counsel, conversation, encouragement, consolation, and sometimes reproof as well — these thrive most fully in friendships, and that friendship is the most agreeable which a likeness of character has joined together.

\ciceroSection{59} But in discharging all these duties we must look to what each person needs most, and what each person can or cannot achieve without our help. Thus the grades of obligation will not always match the grades of relationship; and there are duties owed more to some than to others — so that you might sooner help a neighbor in gathering his harvest than a brother or a close friend, yet defend a kinsman or a friend rather than a neighbor if the case came to a lawsuit. In all matters of duty, then, circumstances of this kind must be surveyed on every side, and the habit and practice of such reckoning must be acquired, so that we may be good accountants of duty — adding and subtracting to see what remainder is left, and from that understanding how much is owed to each.

\ciceroSection{60} But just as doctors, generals, and orators — however thoroughly they have mastered the rules of their art — can achieve nothing worthy of great praise without practice and experience, so too the precepts for maintaining duty are indeed set out, as I am doing here, but the magnitude of the subject demands practice and exercise as well. I have said, I think, enough about how what is honorable is derived from those matters that fall within the law of human fellowship, and how from it duty in turn is drawn.

\ciceroSection{61} We must understand, however, that when four categories have been set out from which honor and duty flow, the most splendid in appearance is that which is enacted with a great and elevated spirit that looks down on human affairs. And so in reproaches the readiest taunt, when something of this kind can be said, runs: \textit{You young men carry a woman's spirit — that girl, a man's}; and similarly: \textit{Salmacian spoils, won without sweat or blood.} But on the other side, what has been accomplished with greatness of spirit, with courage, with surpassing excellence — these things we praise, I know not how, with something like a fuller voice. Hence the orators' open field of Marathon, Salamis, Plataea, Thermopylae, Leuctra; hence, for us, Cocles, the Decii, Gnaeus and Publius Scipio, Marcus Marcellus, and countless others; and above all the Roman people itself excels in greatness of spirit. The passion for martial glory is evident from the very statues we see, adorned, for the most part, in military dress.

\ciceroSection{62} But that elevation of spirit which shows itself in danger and toil, if it is empty of justice and fights not for the common safety but for its own advantage, is a vice; for it is not merely not virtue — it is rather a brutishness that repels all that is human. The Stoics, therefore, rightly define courage as the virtue that fights in defense of what is equitable. For this reason no one who has won glory for courage through treachery and malice has won genuine praise; nothing can be honorable that is empty of justice.

\ciceroSection{63} Admirable, then, is that saying of Plato: knowledge, he says, when it is cut off from justice, is to be called cunning rather than wisdom; and likewise a spirit ready to face danger, if it is driven by its own desire and not by the common good, ought to bear the name of recklessness rather than courage. And so we require that brave and great-souled men be at the same time good and straightforward — lovers of truth, and least of all deceitful; qualities that flow from the very heart of justice's praise.

\ciceroSection{64} Yet there is this troubling thing: in that elevation and greatness of spirit, obstinacy and an excessive hunger for pre-eminence most easily take root. As Plato says, the entire character of the Spartans was inflamed with a passion for victory; and so the greater a man's greatness of spirit, the more he wishes to be first among all — or rather, to be the only one. It is hard, when you have set your heart on surpassing everyone, to maintain that equity which is most distinctively the property of justice. And so it comes about that such men neither submit to defeat in argument nor endure any established and lawful right; in a city-state they tend to emerge as those who deal out bribes and pursue factions, in order to acquire the greatest possible power and to prevail by force rather than stand level by justice. But the harder the thing, the more glorious; for there is no moment that ought to be empty of justice.

\ciceroSection{65} Those who repel injury, then, are to be counted brave and great of spirit — not those who commit it. True and wise greatness of spirit judges that honorable thing, which nature above all pursues, to reside in actions and not in glory, and prefers to be preeminent rather than to seem so; for anyone who hangs upon the mistaken judgments of an ignorant crowd is not to be reckoned among great men. Yet it is precisely the most exalted souls who are most readily driven to injustice by desire for glory — a slippery ground indeed, since it is hardly possible to find a man who, having undertaken labors and entered into dangers, does not seek glory as something very like the reward of his achievements.

\ciceroSection{66} A spirit that is brave and great is discerned most clearly by two things: of these, the first consists in holding external circumstances in contempt, once one has been persuaded that a man ought to admire, to desire, and to pursue nothing but what is honorable and fitting, and ought to yield to no man, to no disturbance of the mind, to no stroke of fortune. The second is this: that, being so disposed in spirit as I have described, one should accomplish great deeds — great indeed and of the greatest usefulness — but deeds that are strenuous in the highest degree and full of toil and danger, both to one's life and to the many things that belong to life.

\ciceroSection{67} All the brilliance, all the grandeur — and I add the usefulness as well — of these two things resides in the latter; but the cause and the force that produces great men resides in the former: for there lies what makes souls outstanding and contemptuous of human concerns. That very quality is itself discerned in two respects: whether you judge the honorable alone to be good, and whether you are free from every disturbance of the mind. For it takes a brave and great spirit both to reckon as small what most men think exceptional and distinguished, and to hold it in contempt on firm and unwavering grounds; and equally to bear those things that appear bitter — the many blows of various kinds that turn upon men in their lives and fortunes — in such a way that you do not depart at all from nature's own station, nor from the dignity of the wise man. That takes a robust spirit and great steadiness of character.

\ciceroSection{68} But it is hardly consistent that the man who is not broken by fear should be broken by desire, or that the man who has shown himself unconquered by toil should be overcome by pleasure. Therefore both this must be avoided and the desire for money must be shunned; for nothing is so mean of spirit, so petty, as to love wealth — nothing more honorable, more magnificent, than to hold money in contempt: if you lack it, to hold it in contempt; if you have it, to devote it to generosity and kindness. Desire for glory must also be guarded against, as I said above; it robs a man of freedom, and for freedom every man of great spirit ought to be ready to fight with all he has. Nor indeed should command be sought after; rather it should sometimes be declined, sometimes laid aside.

\ciceroSection{69} One must be free from every disturbance of the mind — from desire and from fear alike, and also from grief and from excessive pleasure and from anger — so that tranquility and calm of spirit may be present, bringing with them both steadiness and dignity. Many men there are, and have been, who, in pursuit of the tranquility I speak of, have withdrawn from public affairs and taken refuge in leisure; among these are numbered the most eminent philosophers, who stand far in the lead, and also certain men of strict and serious character who could endure neither the ways of the people nor of those in power — and some lived in the country, delighting in the management of their own estates.

\ciceroSection{70} These men had the same aim as kings: to want for nothing, to obey no one, to enjoy freedom — the essence of which is to live as you choose. And so, though those who desire power share this aim with the men of leisure I have described, the former think they can attain it by possessing great resources, the latter by being content with their own, however modest. In this matter neither view is to be dismissed altogether; but the life of those who keep to leisure is easier and safer, less burdensome and troublesome to others, while the life of those who have dedicated themselves to public affairs and to great undertakings is more productive for humanity and better suited to distinction and renown.

\ciceroSection{71} Perhaps, therefore, pardon should be granted to those who refrain from public life on account of exceptional talent devoted to learning; and equally to those who have withdrawn from public affairs because of bodily weakness or some more serious cause, leaving to others the conduct of government and its attendant credit. But those for whom no such cause exists — if they claim to hold in contempt what most men admire, namely commands and offices — this I consider deserving of blame rather than praise. Their judgment, in that they despise and make nothing of glory, it is hard not to approve; yet they appear to dread the toil and trouble, and then the reproach and shame of rebuffs and defeats, as though these were a kind of disgrace. And indeed there are those who are inconsistent in opposite things: they spurn pleasure with the utmost severity, yet prove softer when it comes to pain; they profess to disregard glory, yet are shattered by disgrace — and in all this they are far from consistent.

\ciceroSection{72} But those who have from nature the endowments required for public life ought, setting aside all hesitation, to seek office and take up the management of the state; for only so can the city be governed, or greatness of spirit be made manifest. And those who enter public life need magnificence and contempt for human concerns — which I speak of often — no less than philosophers do; they need tranquility of mind and calm as well, if they are not to be anxious about the future but to live with gravity and steadiness.

\ciceroSection{73} These things are easier for philosophers, inasmuch as fewer targets are exposed in their lives for fortune to strike, and they need fewer things, and if any adversity comes upon them, they cannot fall so heavily. It is for good reason, therefore, that stronger emotions are stirred and greater eagerness to excel arises in those who manage public affairs than in those who live quietly; and for that very reason such men need all the more both greatness of spirit and freedom from anxiety. Anyone who takes up the management of affairs should take care not only to consider how honorable that particular undertaking may be, but also whether he has the capacity to carry it through — and in judging that, he must take care neither to despair rashly through timidity nor to be reckless with confidence through desire. In all affairs, before you take them on, careful preparation must be applied.

\ciceroSection{74} But since most people think military affairs greater than those of civil life, this opinion must be cut down to size. For many men have often gone looking for wars out of desire for glory — and this is what commonly happens with great souls and great talents, all the more so if they are suited to a military life and eager for campaigns; yet if we wish to judge truly, many achievements in civil life have proven greater and more distinguished than those of war.

\ciceroSection{75} Though Themistocles is rightly praised and his name is more celebrated than Solon's, and Salamis is cited as a witness to a most glorious victory — a victory ranked above Solon's policy in founding the Areopagus — that policy is to be judged no less distinguished than that victory: Salamis served Athens once; the Areopagus will serve it always. Through Solon's counsel the laws of Athens are preserved; through it the institutions of the ancestors are maintained. And as for Themistocles — he could point to no contribution of his own to the Areopagus, whereas the Areopagus could truly say that it had supported Themistocles; for the war was conducted by the counsel of that senate which Solon had established.

\ciceroSection{76} The same may be said of Pausanias and Lysander: though it is supposed that the achievements of these men won empire for Sparta, they are not to be compared even in the smallest measure with the laws and discipline of Lycurgus — and it was precisely because of those laws and that discipline that their armies were more obedient and more courageous. When I was a boy, Marcus Scaurus did not seem to me to yield to Gaius Marius; nor, when I was active in public life, did Quintus Catulus seem to yield to Gnaeus Pompeius; for arms abroad are worth little unless there is counsel at home. Nor did Africanus — a man singular both as a private citizen and as a commander — do more for the state by destroying Numantia than Publius Nasica, then a private citizen, did at the same period by putting Tiberius Gracchus to death; though this last affair was not purely a matter of civil statecraft — it touches on the military as well, since it was settled by force and physical action — yet it was carried out by civil counsel, without an army.

\ciceroSection{77} But there is that line of mine that I hear is often attacked by the wicked and envious: "Let arms yield to the toga; the laurel give way to praise." To pass over all other examples — when I was guiding the ship of state, did not arms yield to the toga? For there was never a graver crisis in the commonwealth, and never a greater peace. Thus, through my counsels and my diligence, the weapons themselves slipped from the hands of those most reckless citizens and fell to the ground of their own accord.

\ciceroSection{78} What achievement in war, then, has ever been so great? What triumph is to be set beside it? For I may boast before you, my son Marcus — to whom this inheritance of glory belongs, and the duty of imitating what was done. Gnaeus Pompeius — a man overflowing with the praise of war — paid me this tribute, in the hearing of many: that his third triumph would have been in vain had it not been for my service to the commonwealth, since it was in the city I saved that he would be celebrating it. Civic courage, then, is not inferior to military courage; and in civic life a man ought to exert even greater effort and devotion than in war.

\ciceroSection{79} That honorable thing which we seek from an exalted and magnificent spirit is the work of the mind, not of the body's strength. The body must nonetheless be exercised and kept in such condition that it can obey counsel and reason in the performance of business and in the endurance of toil. But the honorable thing we are seeking is entirely a matter of the mind's care and thought; and in this those who govern the state in the toga bring no less benefit than those who conduct war. By their counsel wars have often either not been undertaken or been brought to a conclusion — and sometimes, indeed, they have been declared, as when Marcus Cato called for the Third Punic War, whose authority prevailed even after his death.

\ciceroSection{80} Therefore the way of deciding by deliberation is to be preferred to the courage of fighting it out — but we must take care that we do not make this preference out of aversion to fighting rather than out of regard for what is advantageous. War is to be undertaken in such a spirit that nothing appears to be sought except peace. A brave and steadfast spirit is shown in this: not to be thrown into confusion when circumstances are harsh, not to be tumbled from one's footing in a panic — as the saying goes — but to keep one's presence of mind and one's counsel, and not to depart from reason.

\ciceroSection{81} Yet while this belongs to greatness of spirit, the following belongs also to greatness of intellect: to anticipate the future in one's thinking, and well in advance to settle what may happen on either side, and what is to be done when any given event occurs — never to be put in the position of having to say, "I had not expected that." These are the works of a great and lofty spirit, confident in its wisdom and counsel; but to expose oneself recklessly in the field of battle and to close with the enemy hand to hand is something savage and beast-like. Yet when the moment and necessity demand it, one must fight hand to hand, and death is to be preferred to servitude and disgrace.

\ciceroSection{82} On the question of overthrowing and plundering cities, the greatest care must be taken that nothing be done rashly, nothing cruelly. It is the mark of a great man, when affairs are in turmoil, to punish the guilty, to preserve the many, and in all circumstances to hold fast to what is right and honorable. For just as there are those who, as I said above, prefer the affairs of war to those of peace, so you will find many to whom desperate and headlong counsel seems more splendid and more grand than counsel that is calm and deliberate.

\ciceroSection{83} We must never flee danger in such a way as to appear unwarlike and cowardly; yet we must equally avoid thrusting ourselves into dangers without cause, for nothing could be more foolish than that. In confronting dangers, therefore, we should follow the practice of physicians, who treat the mildly ill with mild remedies, but are compelled to apply dangerous and precarious treatments to graver diseases. To pray for a storm in fair weather is the act of a madman; to ride out a storm by whatever means one can is the act of a wise man — and all the more so, if the good gained by seeing the matter through outweighs the harm risked by hesitation. Perilous enterprises fall partly on those who undertake them and partly on the state. Likewise, some men are put at hazard of their lives, others of their glory and the goodwill of their fellow citizens. We ought therefore to be readier to face our own dangers than dangers shared in common, and to fight more willingly for honor and glory than for other advantages.

\ciceroSection{84} Many men have been found who would not only spend their fortunes but give their very lives for their country, yet would not willingly suffer even the slightest loss to their glory — not even when the state demanded it. Such was Callicratidas, who had been the Spartan commander in the Peloponnesian War and had accomplished many fine things, but undid it all in the end by refusing to follow the counsel of those who thought the fleet should be withdrawn from Arginusae and that battle with the Athenians should be avoided; to these men he replied that the Spartans, if that fleet were lost, could build another, but he could not flee without dishonor to himself. And this wound was but moderate for the Spartans; the ruinous blow was when Cleombrotus, through fear of unpopularity, rashly engaged Epaminondas, and the Spartan power collapsed. How much better was Quintus Fabius Maximus! Of him Ennius writes: "One man by his delaying restored our state. / He did not put reports before the safety of Rome. / And so his glory shines ever greater and more bright." This kind of fault must be avoided in civil affairs as well. There are those who, though their opinion may be the best possible, dare not speak it for fear of giving offense.

\ciceroSection{85} In all things, those who are to stand at the head of the state should hold fast to two precepts of Plato: first, to guard the welfare of the citizens so that whatever they do they refer to that, forgetting their own advantage; second, to care for the whole body of the state, not to protect one part of it while abandoning the rest. For a guardianship, and so too the administration of the state, must be conducted for the benefit of those who have been entrusted to it, not of those to whom it has been entrusted. Those who look to one part of the citizens while neglecting another introduce into the state the most destructive of things — faction and strife; and from this it comes about that some appear to favor the popular interest, others to champion the best men, while few care for all.

\ciceroSection{86} From this cause came the great dissensions at Athens, and in our own state not merely factions but ruinous civil wars — things that a serious, courageous citizen worthy of leadership in the state will shun and detest. He will devote himself entirely to the state; he will pursue neither wealth nor power; he will guard the state as a whole, so as to look to the welfare of all. Nor will he drag anyone into hatred or envy by false accusations, and altogether he will cleave so firmly to justice and to what is honorable that, to preserve these things, he will accept the gravest hurt, and will meet death sooner than abandon what I have described.

\ciceroSection{87} The scramble for office and the contest for political power is, taken as a whole, a most wretched thing. On this subject there is a fine passage in the same Plato, where he says that those who contend with each other over which of them shall administer the state act like sailors quarreling over which of them shall take the helm. He also lays down that we should regard as adversaries those who take up arms against us, not those who in their own judgment seek to serve the state — such as the disagreement between Publius Africanus and Quintus Metellus, which was without bitterness.

\ciceroSection{88} Nor again are those to be listened to who think one should nurse a fierce anger against one's enemies, and who will count this the mark of a great and courageous man; for nothing is more praiseworthy, nothing more worthy of a great and distinguished man, than the ability to be appeased and the habit of clemency. In free peoples, moreover, where there is equality before the law, there must also be practiced a readiness of temper and what is called elevation of soul — lest, if we grow angry at those who approach us at the wrong moment or ask something shamelessly, we fall into a peevishness that is both useless and hateful. And yet gentleness and clemency are only to be approved in this way: that strictness be applied when the state requires it, for without strictness a state cannot be governed. All rebuke and all correction, moreover, should be free from contempt, and should be referred not to the advantage of the one who punishes or reproves with words, but to the welfare of the state.

\ciceroSection{89} We must also take care that the punishment not exceed the fault, and that some are not called to account while others who are guilty of the same offenses are not even summoned. Above all, anger must be kept out of the act of punishment; for a man who approaches punishment in anger will never hold that mean between too much and too little which the Peripatetics approve — and rightly approve, were it not that they also praise anger and say it was given by nature to useful purpose. Anger must in truth be rejected in all things, and it is to be desired that those at the head of the state should be like the laws, which are directed to punishment not by anger but by equity.

\ciceroSection{90} And in prosperity too, when things flow as we would wish, we must take great care to flee arrogance, disdain, and conceit. For to bear adversity without self-command and to bear good fortune without self-command are alike the marks of a light character; and a consistency throughout life is an excellent thing, always the same expression, always the same brow — as we are told of Socrates and of Gaius Laelius. Philip of Macedon, I observe, was surpassed in achievement and in glory by his son, yet was his superior in affability and in humanity; and so the one was always great, the other often most disgraceful. There seem to be good grounds for the counsel of those who warn that the higher we stand, the more modestly we should conduct ourselves. Panaetius records that Africanus, his student and intimate friend, used to say that just as horses that have grown unruly through the frequent strain of battle are turned over to trainers so that they may be more easily managed, so men who have been made headstrong by good fortune and who trust too much in themselves should, as it were, be led into the ring of reason and learning, so that they may see clearly the frailty of human affairs and the shifting tides of fortune.

\ciceroSection{91} And in the very height of prosperity it is especially important to take counsel with friends, and to grant them even greater authority than before. In those same times we must also take care not to open our ears to flatterers, and not to suffer ourselves to be cajoled — a thing in which it is easy to be deceived; for we tend to believe ourselves deserving of the praise we receive, and from this springs a multitude of errors, when men, swollen with self-conceit, are most disgracefully made ridiculous and find themselves in the gravest delusions. But enough of this.

\ciceroSection{92} This judgment, however, must stand: that the greatest deeds are done and the greatest spirit shown by those who govern states, since the administration of states reaches the widest extent and concerns the greatest number of people. Yet there have been and are many men of great spirit living a life of retirement, who either investigate or attempt great things and keep themselves within the limits of their own pursuits; or standing midway between the philosophers and those who administer the state, they take their delight in managing their own private affairs — neither building up that estate by every possible means, nor excluding their own people from the use of it, but rather sharing it with friends and with the state whenever occasion arises. Such an estate should first have been well acquired, by no base or hateful profit; then increased by good management, industry, and thrift; and then made as useful as possible to as many as deserve it — answering not to appetite and luxury but to generosity and kindness. The man who keeps these prescriptions may live magnificently, with dignity and spirit, and also with simplicity, with good faith, with love of his fellow men.

\ciceroSection{93} It remains to speak of the one part of the honorable that is still to be discussed — the part in which we find a sense of shame and a certain grace of life, self-control and restraint, the quieting of all disturbances of the mind, and due measure in all things. Here is contained what may be rendered in Latin as \textit{decorum} — for the Greeks call it \textit{[Greek: prepon]}. Its nature is such that it cannot be separated from what is honorable;

\ciceroSection{94} for what is fitting is honorable, and what is honorable is fitting. The difference between the honorable and the fitting can more easily be grasped than explained. Whatever is fitting shows itself when moral rectitude has come first. Thus what is fitting appears not only in this part of the honorable now under discussion, but also in the three that precede it. For to use reason and speech prudently, to act in what you do with deliberation, to discern the true in every matter and to uphold it — all this is fitting; and to be deceived, to err, to slip, to be misled is as unseemly as to rave and to lose one's wits. Everything just is fitting; everything unjust, being at once base, is also unseemly. The same holds for courage: what is done with manliness and greatness of spirit is seen as worthy of a man and fitting; what is done otherwise, being at once base, is also unseemly.

\ciceroSection{95} This fitting quality I describe therefore runs through all that is honorable — it runs through it in such a way that it is seen not by any recondite process of reasoning but plainly before us. For there is a certain element present in every virtue, and recognized in it, that we call fitting; and this can be separated from virtue more easily in thought than in reality. As physical grace and beauty cannot be separated from health, so this fitting quality, of which I speak, is thoroughly blended with virtue as a whole, though it is distinguished from virtue by the mind and in thought.

\ciceroSection{96} Its classification is twofold: for we understand a certain general fitting quality, which is found throughout all that is honorable, and another, subordinate to this, which has to do with the several parts of what is honorable. The general fitting quality is usually defined roughly as follows: that which is consistent with the excellence of a human being, in that in which his nature differs from all other living creatures. The species subordinate to the genus is defined in this way: that what is fitting is that which is so consistent with nature that in it moderation and self-control appear, with a certain liberality of appearance.

\ciceroSection{97} We can understand the nature of this propriety by reference to that which poets observe — a topic on which more is usually said in another context. But what we say when we say that poets preserve the fitting is this: that what is done and said suits each role. If Aeacus or Minos were to say, ``Let them hate, so long as they fear,'' or ``The father himself is the children's tomb,'' it would seem improper — for we understand these men to have been just; but when Atreus says those words, applause breaks out, since the speech is worthy of the role. Poets, then, will judge what is fitting for each person from his role; but on us nature itself has imposed a role with great distinction and superiority over the rest of living things.

\ciceroSection{98} Poets, then, working through the great variety of roles, will see what suits even the vicious and what is fitting for them; but since nature has assigned to us the parts of consistency, self-control, restraint, and reverence, and since that same nature teaches us not to neglect how we conduct ourselves toward other human beings, it follows that the range of propriety — which belongs to all that is honorable — becomes visible in its full breadth, as does the particular form it takes in each kind of virtue. For just as bodily beauty, by the harmonious arrangement of the limbs, draws the eye and gives pleasure by the very fact that all the parts agree with one another in a kind of grace, so this propriety, as it shines out in a life, earns the approval of those among whom one lives, by the order and consistency and restraint of all one's words and deeds.

\ciceroSection{99} We must therefore show a certain deference toward human beings — toward the best of them and toward the rest alike. To be indifferent to what anyone thinks of you is the mark not only of arrogance but of a thoroughly undisciplined character. There is, moreover, a distinction to be drawn in the account we give of human beings between justice and respect. Justice requires that we not do violence to people; respect, that we not give offense — and it is in this latter that the force of propriety is most clearly seen. With these distinctions set out, I take it that we now understand what this thing is that we call the fitting.

\ciceroSection{100} The duty, then, that derives from it has this as its first path: it leads toward agreement with and preservation of nature. If we follow nature as our guide, we shall never go astray — we shall follow what is keen and penetrating in nature, what is fitted to the fellowship of human beings, and what is vigorous and strong. But the greatest force of propriety resides in the part of the subject we are now discussing; for it is not only the bodily movements that conform to nature that win approval, but much more so those of the mind, which likewise accord with nature.

\ciceroSection{101} The force and nature of the mind is twofold: one part lies in desire — in Greek \textit{horm\={e}} — which drives a person this way and that; the other in reason, which teaches and makes plain what should be done and what avoided. The result is that reason commands and desire obeys. All action, moreover, must be free of recklessness and negligence, and nothing should be done for which no satisfactory account can be given; this is very nearly the definition of duty.

\ciceroSection{102} The task, then, is to make the desires obey reason — to have them neither outrun it nor abandon it out of sloth or cowardice — and to keep them tranquil and free of every disturbance of the mind. From this will shine out all consistency, all self-control. When desires wander too far and, exulting as it were in yearning or avoidance, are not restrained sufficiently by reason, they undoubtedly overstep all limit and measure; for they throw off and abandon their obedience and refuse to listen to reason, to which by the law of nature they are subject. By these desires not only are minds thrown into disorder but bodies as well. One has only to look at the faces of people who are angry, or of those disturbed by some passion or some fear, or of those beside themselves with excessive pleasure: their expressions, their voices, their movements and bearing are all changed.

\ciceroSection{103} From which we understand — to return to the form of duty — that all desires must be reined in and quieted, and that attentiveness and care must be roused, so that we do nothing rashly and by chance, nothing without thought and heedlessly. Nature did not make us, after all, to look as though we were born for play and jest; rather for seriousness and for pursuits that are weightier and greater. Play and jest are permitted, of course — but as sleep and other forms of rest are permitted: when we have done justice to serious and demanding work. And the very kind of jesting must not be dissolute and unrestrained, but well-bred and witty. For just as we do not give children unlimited license to play, but only what is not alien from honorable activity, so even in jest some spark of a decent character ought to shine through.

\ciceroSection{104} There are, in sum, two kinds of wit: one coarse, insolent, shameless, obscene; the other refined, urbane, clever, and polished. This latter kind is full not only of our own Plautus and the old comedy of the Athenians, but also of the Socratic philosophers' writings, and many witty sayings of many men — such as the collection made by the elder Cato, which people call \textit{[Greek: apophthegmata]}. The distinction between the wit of a gentleman and that of a boor is easy to draw: the one, if timed properly, is worthy even of the most serious man in a moment of relaxation; the other is not worthy even of a free man, when obscenity of language is brought to bear on squalid subject matter. A limit must also be kept in jesting, so that we do not pour everything out and, carried away by the pleasure, slip into some impropriety. Our own Field of Mars and the pursuits of the hunt provide honorable examples of play.

\ciceroSection{105} But it matters to the whole inquiry into duty always to keep ready at hand how greatly human nature surpasses cattle and the other beasts. They feel nothing but pleasure and rush toward it with every impulse; but the human mind is fed by learning and reflection, always inquiring or doing something, led on by the pleasure of seeing and hearing. Even if someone is rather more given to pleasures — provided he is not of the herd (for there are people who are human in name, not in fact) — even if someone is a little more upright in spirit, however much he is taken by pleasure, he conceals and suppresses his appetite for it out of a sense of shame.

\ciceroSection{106} From this we understand that bodily pleasure is not sufficiently worthy of human excellence, and that it ought to be spurned and rejected. Or if someone does grant some value to pleasure, he must carefully observe the proper measure in enjoying it. And so let the food and care of the body be referred to health and strength, not to pleasure. And further: if we are willing to consider what the excellence and dignity in our nature actually is, we shall understand how base it is to dissolve in luxury and to live a soft and pampered life, and how honorable to live frugally, with self-control, sternly, soberly.

\ciceroSection{107} We must also understand that nature has, so to speak, clothed us in two roles. The first is the role we all share, by virtue of our all being participants in reason and in that superiority whereby we excel the beasts — the role from which all that is honorable and fitting is derived, and from which the method of discovering duty is sought. The second role is the one assigned distinctively to each individual. For just as there are great differences among bodies — we see some men well-endowed for speed in running, others for strength in wrestling, and likewise in appearance some possessing dignity, others grace — so in minds there arise even greater varieties.

\ciceroSection{108} In Lucius Crassus and Lucius Philippus there was abundant wit; even more, and more deliberately deployed, in Gaius Caesar the son of Lucius. In the same era, in Marcus Scaurus and in the young Marcus Drusus there was a remarkable sternness; in Gaius Laelius much cheerfulness; in his intimate friend Scipio, greater ambition and a graver manner of life. Among the Greeks, Socrates we know as charming and polished and entertaining in conversation, a man who in all his talk played the ironist — what the Greeks call the \textit{[Greek: eiron]}. Pythagoras and Pericles, by contrast, attained the highest authority with no lightness at all. Among Carthaginian commanders we know Hannibal for his cunning; among our own, Quintus Maximus — men who could easily conceal, keep silent, dissemble, lay ambushes, forestall the plans of the enemy. In this kind the Greeks rank Themistocles and Jason of Pherae above all others; and especially clever and shrewd was the device of Solon, who, so that his life might be safer and he might serve the state rather better, feigned madness.

\ciceroSection{109} There are others very different from these — open and straightforward, thinking that nothing should be done by stealth or through ambush, devotees of truth, enemies of fraud. There are likewise others who will endure anything and serve anyone, so long as they get what they want — as we saw in Sulla and Marcus Crassus. In this kind, the most cunning and most enduring was the Spartan Lysander, and quite the opposite, Callicratidas, who commanded the fleet directly after Lysander. In conversation, too, we see one man — however powerful he may be — making himself appear as if he were simply one among many: we saw this in Catulus, both father and son, and likewise in Quintus Mucius Mancia. I have heard from older men that this same quality was in Publius Scipio Nasica; and that his father, on the other hand — the man who brought to punishment the ruinous designs of Tiberius Gracchus — had no ease in conversation, nor did Xenocrates, the most austere of philosophers, and he was great and celebrated for that very quality. There are countless other differences of character and temperament, and none of them, by any means, to be censured.

\ciceroSection{110} Each person must, however, hold firmly to what is his own — not what is vicious in him, but still what is distinctively his — so that the propriety we are seeking may more easily be maintained. We must act in this way: that we strive against universal nature in nothing, yet with that preserved, follow our own particular nature; so that even if other pursuits are weightier and better, we still measure our own endeavors by the rule of our own nature. It is no use fighting against nature, or pursuing what you cannot achieve. From this it emerges the more clearly what that propriety is, precisely because nothing is fitting when done, as the saying goes, against Minerva's will — that is, when nature opposes and resists.

\ciceroSection{111} If anything at all is fitting, surely nothing is more so than consistency — consistency in the whole of life and in each individual action — which you cannot maintain if you abandon your own nature in imitating another's. For just as we ought to use the language that is native to us — and not, like certain people who pepper their speech with Greek words, be rightly laughed at — so in action and in the whole of life we ought to introduce no discordance.

\ciceroSection{112} And this difference in natures carries such force that sometimes one man ought to take his own life and another, in the same circumstances, ought not. Was Marcus Cato in any different situation from the others who surrendered to Caesar in Africa? Yet the others would perhaps have been blamed had they killed themselves, because their lives had been gentler and their characters more accommodating; but for Cato, whom nature had endowed with an incredible seriousness and whom he himself had strengthened by lifelong consistency, who had always remained fixed in his resolution and in the course he had undertaken, it was better to die than to look upon the face of a tyrant.

\ciceroSection{113} How much Ulysses endured in that long wandering — submitting to women, if Circe and Calypso are to be called women, and seeking in every conversation to be agreeable and pleasant to all! And at home he even bore the insults of slaves and serving-maids, in order to reach at last what he longed for. But Ajax, with the spirit tradition ascribes to him, would have chosen to die a thousand deaths rather than submit to such things. When we consider all this, it becomes clear that each man must weigh what is his own, regulate it accordingly, and not put to the test how well another's gifts may suit him — for what becomes each man most is precisely what is most fully his own.

\ciceroSection{114} Let each man, then, know his own natural bent, and prove himself a keen judge both of his own virtues and of his own faults, lest actors seem to show more practical wisdom than we do. For they do not choose the best plays but the ones best suited to themselves: those who rely on voice choose \textit{Epigoni} and \textit{Medus}; those who rely on gesture choose \textit{Melanippa} and \textit{Clytemnestra} — Rupilius, as I recall, invariably chose \textit{Antiopa}, Aesopus seldom \textit{Ajax}. Will an actor, then, see this on the stage and a wise man fail to see it in life? We shall therefore apply our greatest effort to those things for which we are best equipped; and if necessity ever thrusts us toward tasks foreign to our natural bent, we must bring to them all possible care, reflection, and diligence, so that we may do them if not with distinction, at least as little undistinguished as may be. Nor is the goal so much to pursue gifts we have not been given as to flee our own faults.

\ciceroSection{115} To the two roles I spoke of above, a third is added — one that some chance or circumstance lays upon us; and a fourth as well, which we take on ourselves by our own deliberate choice. Kingdoms, commands, noble birth, public honors, wealth, resources, and their opposites, all these rest on circumstance and are governed by the times; but the role we wish to play is determined by our own will. And so some men give themselves to philosophy, some to civil law, some to eloquence; and in the virtues themselves, one man prefers to excel in one, another in another.

\ciceroSection{116} Those whose fathers or ancestors have achieved some glory commonly strive to shine in the same field of distinction — as Quintus Mucius, son of Publius, in civil law, or the son of Paulus in the art of war as Africanus. Some add a distinction of their own to the renown they have received from their fathers: this same Africanus, for instance, crowned his military glory with eloquence. Timotheus, son of Conon, did the same: no whit behind his father in military distinction, he added to that renown a glory of learning and intellect. It happens, too, that some set aside the imitation of their ancestors altogether and pursue an enterprise wholly their own — and this is most often the labor of those who set great aims before themselves yet spring from obscure forebears.

\ciceroSection{117} All this, then, we must hold in mind and reflection when we ask what is fitting. Above all we must settle what we wish to be and what manner of men — in what kind of life we mean to live: this is the most difficult deliberation of all. For at the threshold of youth, when judgment is at its weakest, each person fixes on the way of life that has taken his heart most completely; and so he is caught up in some definite course and direction of living before he has been able to judge what is best.

\ciceroSection{118} For the story that Prodicus tells of Hercules — as it appears in Xenophon — that when he first reached manhood (the season nature has appointed for each person to choose which path of life he means to follow), he went apart into solitude and sat there long in deep uncertainty, seeing before him two roads, one of Pleasure and one of Virtue, deliberating which to take: this may perhaps have been possible for Hercules, born of the seed of Jupiter, but it is not so for us. We imitate those who have caught our fancy and are drawn toward their pursuits and ways; most often we are steeped in our parents' teaching and led into their habits and custom; others are carried along by popular opinion and desire most eagerly whatever seems finest to the majority. Yet a few, either by some happy fortune or by native goodness of character, without any discipline from their parents, have followed the right road of life on their own.

\ciceroSection{119} But the rarest kind of all are those who, endowed either with outstanding greatness of intellect or with distinguished culture and learning, or with both together, have also had the leisure to deliberate which course of life they most wished to pursue — and in that deliberation every consideration must be referred back to each person's own nature. For just as in all we do we look for what is fitting on the basis of each man's nature, as was said above, so in the ordering of a whole life far greater care must be given to this, that we may be able to hold consistency with ourselves throughout the course of life and not falter in any duty.

\ciceroSection{120} Now since nature carries the greatest weight in this reckoning and fortune comes next, account must certainly be taken of both in choosing a way of life — but nature must weigh the more, for she is far more firm and constant, so that fortune sometimes seems, as if mortal herself, to be fighting against immortal nature. Whoever has directed his whole plan of living toward a nature that is not corrupt, let him hold to consistency — for that above all is fitting — unless he comes to understand that he made a mistake in his original choice of life. If this happens (and it can happen), a change of habits and pursuits must be made. And if circumstances lend themselves to such a change, we shall make it more easily and comfortably; if not, it must be made gradually and by degrees — just as wise men judge it more fitting to let friendships that have lost their warmth and approval fade out slowly rather than cut them off short.

\ciceroSection{121} But when a way of life has been changed, every effort of reason must be made to show that we acted on sound judgment. And since I said a little earlier that our elders are to be imitated, first let this exception stand: their faults are not to be imitated. Next, if a man's nature does not allow him to imitate certain things — as the elder Africanus's son, who was adopted by the son of Paulus, could not, because of frail health, be as much like his father as that father had been like his — if, then, he cannot plead cases, hold the people by speeches, or conduct wars, he must nonetheless uphold those things that do lie within his power: justice, good faith, generosity, restraint, self-control — so that the deficiency of what he lacks may be the less remarked. The finest inheritance a father can leave his children, surpassing every estate, is the glory of virtue and honorable deeds: to bring disgrace upon this must be judged a wrong and a vice.

\ciceroSection{122} And since the same duties are not assigned to men of different ages — some belonging to the young, others to the old — something must also be said about this distinction. It is the duty of a young man to show deference to his elders and to choose from among them those of the highest reputation and soundest judgment, on whose counsel and authority he may lean; for the inexperience of early life must be formed and guided by the wisdom of age. This time of life above all must be kept from sensual indulgence and exercised in endurance of effort, in body and in mind, so that industry may be vigorous in both military and civic duties. And when young men wish to relax their minds and give themselves over to enjoyment, they should guard against excess and remember modesty — which will be easier if they do not object to having their elders present even on such occasions.

\ciceroSection{123} Old men, on the other hand, should lighten the physical exertions while increasing the exercises of the mind; and they should make it their object to be of as much service as possible to friends, to the young, and above all to the state through counsel and practical wisdom. There is nothing old age should guard against more than yielding to languor and idleness. Self-indulgence is shameful at every age, but at old age it is most degrading of all; and if intemperance in physical desire is added as well, the evil is twofold — for old age itself incurs disgrace, and it makes the intemperance of the young more shameless.

\ciceroSection{124} Nor is it beside the point to speak of the duties of those in office, of private citizens, of the native-born, and of foreigners. It is the peculiar office of a magistrate to understand that he bears the role of the state, and that he must sustain its dignity and honor, uphold the laws, administer justice, and remember that all this has been entrusted to his fidelity. A private citizen, for his part, ought to live on equal and level terms with his fellow citizens — neither submissive and self-abasing, nor puffed up with self-importance — and in public affairs to desire what is peaceable and honorable; such is the man we are accustomed to call, and to regard as, a good citizen.

\ciceroSection{125} The duty of a foreigner or resident alien is to attend to his own affairs, not to inquire into others', and to be as little as possible concerned with matters that belong to another state. In this way duties will generally be discovered when we ask what is fitting and what suits persons, times, and ages. Nothing, however, becomes a man so well as consistency in every action and every decision.

\ciceroSection{126} But since that fitness shows itself in all our actions and words, and in the very movement and bearing of the body, and since it rests on three things — beauty, order, and an adornment suited to action (things difficult to put into words, but sufficient to be understood) — and since under these three heads lies also the care to win approval from those among whom and before whom we live, a few words should be said on these matters too. Nature herself appears to have taken great thought for our body: the form and figure she has given us, whatever in it has a noble appearance, she has set in view; but those parts of the body that have been given to serve the needs of nature, parts whose aspect would be unsightly and unseemly, she has covered and hidden away.

\ciceroSection{127} Human modesty has imitated this careful workmanship of nature. What nature has concealed, all persons of sound mind remove from sight, and they take pains to comply with natural necessity as privately as possible; and the parts of the body whose functions are necessary they do not name, either the parts themselves or their uses, by their proper names. To do what is not in itself shameful, provided it is done privately, is one thing; to speak of it openly is obscene. And so neither the open performance of such acts escapes the charge of wantonness, nor does obscene language.

\ciceroSection{128} Nor are we to listen to the Cynics, or to those among the Stoics who were very nearly Cynics, when they reproach and mock us for thinking it shameful in word what is not shameful in deed, and for naming by their proper names those things that are shameful. Robbery, fraud, adultery — these are shameful in fact, yet are spoken of without obscenity; to beget children is honorable in fact, yet the name is obscene. And much more is argued by these same men against modesty along the same lines. But let us follow nature and flee everything that gives offence to the eyes and ears. In standing, in walking, in sitting, in reclining, in expression, in the eyes, in the movement of the hands, let that fitness hold its ground.

\ciceroSection{129} In these matters two things above all are to be avoided: anything effeminate or soft, and anything harsh or boorish. Nor ought we to concede to actors and orators that such things become them while they are unbecoming for us. The customs of the stage have, under the old discipline, such a degree of modesty that no one comes on without a loincloth; they are afraid that if, by some chance, certain parts of the body should be exposed, they will be seen indecently. Among our own people it is customary that sons who have come of age do not bathe with their fathers, nor sons-in-law with their fathers-in-law. Modesty of this kind, then, is to be preserved — especially with nature herself as teacher and guide.

\ciceroSection{130} Now beauty is of two kinds: in one there is grace, in the other dignity. Grace we should regard as womanly, dignity as manly. Therefore let every adornment unworthy of a man be removed from his person, and let a similar fault be guarded against in gesture and movement. The exercises of the wrestling-ground are often quite repugnant; certain gestures of actors are not free from absurdity; in both arts what is upright and simple is praised. Dignity of appearance is to be maintained by a good complexion, and the complexion by bodily exercise. Beyond this, a certain neatness is called for — not offensive, not too refined, just enough to avoid a rustic and ungentlemanly slovenliness. The same principle holds for dress: here, as in most things, the mean is best.

\ciceroSection{131} We must take care not to walk with steps that are too languid and slow, so that we seem to resemble the bearers in a procession, nor to adopt a pace so hurried that we gasp for breath, our face contorts, our features twist — all of which are strong signals that self-possession is absent. But we must take far greater care that the motions of the mind do not depart from nature. This we shall achieve if we guard against falling into agitations and disturbances, and if we keep our minds attentive to the preservation of what is fitting.

\ciceroSection{132} The motions of the mind are of two kinds: one belongs to thought, the other to desire. Thought is chiefly concerned with seeking out the truth; desire impels us to action. We must therefore take care to employ our thought on the best possible subjects, and to render desire obedient to reason. Now the power of speech is great, and it falls into two kinds: one is disputation, the other conversation. Disputation should be given to the contests of the courts, the assemblies, and the Senate; conversation should move in gatherings, in discussions, in the company of friends — following the dinner-table as well. The rules of disputation belong to the rhetoricians; there are none for conversation, though I am not sure there could not be such rules. But students in the one find teachers; those who seek instruction in the other find none, while everything is crammed with schools of rhetoric — though whatever precepts govern words and thoughts will apply equally to conversation.

\ciceroSection{133} But since we have the voice as the index of the mind, and in the voice two things are to be sought — that it be clear, and that it be pleasing — both must ultimately be sought from nature; yet exercise will improve the one, and imitation of those who speak with precision and ease will improve the other. The Catuli gave no impression — if you were to judge finely — of any particular learning in literature, though they were in fact learned men; but they were thought to use the Latin tongue supremely well. The sound was sweet; letters neither over-pronounced nor swallowed, so that speech was neither obscure nor affected; the voice without strain, neither slack nor singsong. The oratory of Lucius Crassus was richer and no less lively; but the reputation for elegant speech was no less attached to the Catuli. In wit and pleasantry, however, Caesar, the brother of Catulus the elder, surpassed everyone, so that in that very genus of forensic advocacy he overcame the contests of rivals by the quality of his conversation. In all these matters, then, we must take pains, if in everything we are asking what is fitting.

\ciceroSection{134} Let conversation be, then — and it is here that the Socratics most excel — gentle and by no means combative; let it have charm. And let the one who converses not act as though he has come into his own possession and must shut out others, but, as in all other things, so in conversation, let him think a fair give-and-take no bad thing. Let him first notice what subjects he is speaking on: if serious ones, let him bring gravity; if light ones, wit. Above all let him take care that the conversation betrays no fault in his character — which is most likely to happen when people take deliberate pleasure in speaking maliciously of those who are absent, whether in ridicule or in harsh and insulting language, with the aim of blackening their name.

\ciceroSection{135} Conversation is for the most part on domestic affairs, or public business, or the pursuits of learning and scholarship. Care must therefore be taken that, even if it begins to drift elsewhere, it is brought back to one of these; though we must also watch what company is present, since we do not all take equal pleasure in the same subjects, at the same moment, or to the same degree. One must also judge how far conversation holds its delight, and just as there was a way of beginning, so let there be a measure in ending.

\ciceroSection{136} But as the most sound precept in all of life is to flee disturbances — that is, the excessive motions of the mind that do not answer to reason — so conversation ought to be free of such motions, lest anger flare, or some craving, or sluggishness, or faint-heartedness, or anything of that kind appear. Above all we must take care to seem to respect and to esteem those with whom we speak. Rebukes too sometimes become necessary; and in them one must perhaps use a louder and more forceful voice and sharper, weightier words — even seeming to act from anger. But as we come to cauterizing and cutting only rarely and with reluctance, so to this kind of chastisement we shall come seldom and against our will, and never except of necessity, when no other remedy can be found. Yet anger must be far away — for with anger nothing can be done rightly, nothing with deliberation.

\ciceroSection{137} For the most part a mild rebuke is permitted, accompanied by gravity, so that severity is brought to bear while insult is repelled; and even that bitterness which a reproach carries should be made clear to be undertaken for the sake of the very person being reproached. In disputes too with the bitterest of enemies — even when we hear things unworthy of us — it is right to maintain composure and drive out anger. For what is done in a state of agitation cannot be done steadily, and will not win approval from those present. It is also unbecoming to boast about oneself — especially to assert things that are false — and to play the braggart soldier to the laughter of one's audience.

\ciceroSection{138} And since we intend to go through everything — or certainly wish to — something must be said about the kind of house that suits a man of honor and standing. Its purpose is use, and to use the design of building must be adapted; yet a care for comfort and dignity must also be brought to bear. Gnaeus Octavius, who was the first of his family to become consul, is said to have owed his distinction to having built on the Palatine a splendid house, full of dignity: when crowds came to see it, it was thought to have won votes for its owner, a new man, in his campaign for the consulship. It was this house that Scaurus later demolished and added as an extension to his own. So Octavius first brought the consulship into his house; Scaurus, son of a great and celebrated man, brought into his enlarged house not merely defeat but disgrace and ruin as well.

\ciceroSection{139} For dignity is to be adorned by a house, not wholly derived from one; nor should the master be glorified by the house, but the house by the master. And just as in everything else a man must have regard not only for himself but for others, so in the house of a man of note — a house into which many guests must be received and crowds of every kind admitted — care for space is required. Otherwise a large house often brings dishonor on its owner if it stands empty, and especially if it was once accustomed under a different owner to be full. It is unpleasant to hear passersby say: \textit{O ancient house, alas, how unlike a master now controls you} — a line one may apply to many houses in these times.

\ciceroSection{140} Great care must be taken, especially if you are building yourself, not to go beyond measure in expenditure and splendor — and in this there is also much harm in the example set. Most people eagerly imitate the great in this particular direction: who imitates the virtue of Lucius Lucullus, a man of the highest distinction? Yet how many have copied the magnificence of his villas? To those villas a limit must certainly be set and they must be brought back to a mean. And that same principle of moderation must be carried over into every part of the use and furnishing of life. But enough of this.

\ciceroSection{141} In every undertaking three things must be kept in mind: first, that desire obey reason — nothing is better fitted to the preservation of duty; second, that we consider how great is the thing we wish to accomplish, so that neither more nor less care and effort is given to it than the matter demands; third, that we take care to keep whatever pertains to honorable appearance and dignity within due measure. The best measure is to hold to that very fitness of which we spoke earlier, and not to go beyond it. Of these three, however, the most excellent is that desire answer to reason.

\ciceroSection{142} Next, we must speak of the ordering of actions and the opportune moment for each. This is the knowledge contained in what the Greeks call \textit{eutaxia} — not the thing we translate as \textit{modestia}, though the word \textit{modus} is in it, but rather that \textit{eutaxia} in which the preservation of order is understood. Thus, that we may call it \textit{modestia} as well, the Stoics define it as the knowledge of arranging in its proper place each thing that is to be done or said. Order is defined in the same way: the arrangement of things in apt and fitting positions; and the fitting position of an action they say is the opportune moment in time — the opportune moment for action being called in Greek \textit{[Greek: eukairia]}, and in Latin \textit{occasio}. So it comes about that this \textit{modestia}, as we interpret it in the way I have described, is the knowledge of the opportune moments suited to action.

\ciceroSection{143} But the same definition may belong to prudence, of which I spoke at the outset; here, however, we are inquiring into moderation, self-control, and the virtues akin to these. What was proper to prudence has already been said in its place; what belongs to these virtues of which we have long been speaking — the virtues that pertain to modesty and to winning the approval of those with whom we live — must now be set out.

\ciceroSection{144} Such, then, is the order that must be observed in actions: that, just as in a well-knit speech, so in life everything be coherent and consistent with everything else. It is ugly and gravely wrong to introduce into a serious matter any remark fit for the dinner-table, or something self-indulgent. Pericles gave a fine example: when the poet Sophocles was his colleague in the command and they had met on a matter of joint duty, a handsome boy chanced to pass by, and Sophocles said, "What a beautiful boy, Pericles!" — to which Pericles replied: "A general, Sophocles, ought to keep not only his hands but his eyes under control." And yet if Sophocles had said the same thing at a review of athletes, there would have been no just ground for reproach. Such is the force of place and occasion. As when a man who is about to plead a case meditates silently to himself while walking along the road, or thinks carefully about something else, he is not to be blamed; but if he does the same at a dinner party he will seem boorish — ignorant of the demands of the moment.

\ciceroSection{145} Those departures from humanity that are glaring — singing in the Forum, say, or any other gross absurdity — are plain to see and need little by way of counsel or instruction. But the smaller lapses, the ones most people cannot even detect, demand more careful avoidance. Just as with strings or pipes, even a slight dissonance is something a practiced ear catches at once, so we must watch that nothing in our lives slips out of tune — and with all the more vigilance, since the harmony of deeds is greater and better than any harmony of sound.

\ciceroSection{146} Just as the ears of trained musicians catch even the smallest false note, so we, if we wish to be sharp and diligent observers of faults, will often read great things from small signs. From a glance of the eyes, from the ease or tightening of the brow, from sadness, from cheerfulness, from laughter, from speech, from silence, from the raising or lowering of the voice, from all such things, we shall easily judge what in each case is fitting and what falls away from duty and from nature. In this it is no bad thing to judge each quality by what we see in others, so that, if anything unbecoming appears in them, we may avoid it in ourselves. For it happens, I know not how, that we see the faults of others more clearly than our own. And so those pupils whose faults their masters deliberately imitate by way of correction find them easiest to put right.

\ciceroSection{147} Nor is it out of place, when choosing between actions that raise difficulty, to consult men of learning or those experienced by practice, and to enquire what each considers right in every kind of duty. For the larger part of people generally follow wherever nature herself leads. In such consultations we must consider not only what each person says but also what each person thinks, and why he thinks it. Just as painters, sculptors, and poets alike want their own work judged by the public — so that whatever many find fault with may be corrected — and seek out from themselves and from others what is wrong in it, so there are many things that, by the judgment of others, we must do and leave undone, change and correct.

\ciceroSection{148} As for what is to be done according to established custom and civic institutions, no instruction is needed; those customs are themselves the instruction. And no one should be led into this error: that because Socrates or Aristippus did or said something contrary to civic custom and convention, he therefore supposes the same license is open to him. Those men earned that license by great and extraordinary gifts. The Cynic system is to be rejected entirely: it is hostile to that sense of shame without which nothing can be right or honorable.

\ciceroSection{149} Those whose whole lives have been proved in honorable and great pursuits — men who have thought well of the commonwealth and served it well or are serving it still — men invested with some office or command — these we must respect and honor. We must also give much to age, yield to those who hold magistracies, distinguish between citizen and foreigner, and, in the foreigner himself, between one who has come on private and one who has come on public business. To put it briefly, without going through every case: we must cherish, maintain, and preserve the fellowship and community of the whole human race.

\ciceroSection{150} Now, as to trades and livelihoods — which are to be counted respectable, which sordid — this is roughly what we have received by tradition. First, those livelihoods are condemned that earn men's hatred: tax-collecting, for instance, and money-lending. Illiberal and sordid are the occupations of all hired laborers whose work, not whose skill, is purchased — for in their case the wage itself is the contract of servitude. Sordid too must be reckoned those who buy from merchants in order to sell again at once; for they can profit only by lying outright, and there is nothing more base than dishonesty. All craftsmen, moreover, are engaged in a sordid trade, for no workshop can have anything gentlemanly about it. Least of all to be commended are the trades that serve pleasure: fishmongers, butchers, cooks, sausage-makers, fishermen — as Terence says. To these add, if you like, perfumers, dancers, and the whole gaming-table crowd.

\ciceroSection{151} But those pursuits in which either greater wisdom is at work or from which a not inconsiderable benefit is sought — medicine, architecture, the teaching of honorable subjects — these are respectable for those whose station fits them. Commerce, if it is conducted on a small scale, is to be counted sordid; but if it is large and well-stocked, bringing goods in from everywhere and distributing them to many without misrepresentation, it is not altogether to be condemned; and if the merchant, satisfied with — or rather, content with — his profit, sails, so to speak, from open sea into harbor, and then from the harbor into his lands and estates, he seems deserving of the highest praise. Of all the pursuits by which something is gained, however, none is better than farming, none more fruitful, none more pleasant, none more worthy of a free man. Since I have said enough about this in the \textit{Cato Maior}, you will draw from that book what applies here.

\ciceroSection{152} It seems that enough has now been said about how duties are derived from the parts of what is honorable. Among honorable things themselves, however, there often arises a conflict and a question of comparison: when two honorable courses present themselves, which is the more honorable? This is a topic that Panaetius passed over. For since all honor flows from four sources — one being knowledge, another fellowship, a third greatness of spirit, the fourth self-control — it is inevitable that in choosing a duty these will often have to be weighed against one another.

\ciceroSection{153} It is held, then, that duties derived from fellowship are better suited to nature than those derived from knowledge, and this can be confirmed by the following argument: if a wise man's life were arranged so that, with all things flowing to him in abundance, he might at perfect leisure contemplate and investigate by himself everything worth knowing — yet if his solitude were so complete that he could see no other human being, he would quit life entirely. The foremost of all virtues is that wisdom which the Greeks call \textit{sophia} \textgreek{σοφία} — for what we call \textit{prudentia}, which the Greeks name \textit{phronesis} \textgreek{φρόνησις}, we understand as something else: the knowledge of things to be sought and avoided. But that wisdom I called foremost is the knowledge of things divine and human, and it contains within itself the fellowship of gods and men and the community of men with one another. If this is the greatest virtue — as it surely is — it follows necessarily that the duty derived from fellowship is the greatest duty. And indeed, the study and contemplation of nature would be in some way incomplete and unfinished if no action followed upon it. That action shows itself most fully in safeguarding the interests of men; it therefore pertains to the fellowship of the human race. The study of nature must accordingly yield to it.

\ciceroSection{154} This is shown and confirmed by the best of men in their own conduct. What man is so intent on observing and understanding the nature of things that, if while he is examining and contemplating matters supremely worth knowing some sudden danger and crisis faces his country — one he can address and relieve — he will not abandon and cast aside all those studies, even though he supposes himself capable of counting the stars or measuring the magnitude of the universe? And he would do the same for a parent's emergency, or a friend's.

\ciceroSection{155} These considerations make clear that the duties of justice, which bear on men's welfare, must take precedence over the pursuits and duties of knowledge — and that nothing ought to be more prized by a man than that welfare. Even those whose whole study and life has been taken up with the pursuit of knowledge have not withdrawn from adding to men's advantage and benefit; for they educated many, making them better citizens and more useful to their communities — as the Pythagorean Lysis educated Epaminondas of Thebes, as Plato educated Dion of Syracuse, and as many educated many others. And I myself, whatever I contributed to public life — if indeed I contributed anything — came prepared and equipped with the education and teaching I had received from my teachers.

\ciceroSection{156} Nor do learned men instruct and educate their students only while alive and present; they accomplish the very same thing after death through the monuments of their writings. No topic that touches on law, on custom, or on the ordering of the commonwealth has been left untreated by them, so that they seem to have devoted their leisure to our practical concerns. Even those dedicated to the pursuits of learning and wisdom bring their wisdom and intelligence to bear above all for the benefit of men; and for this very reason, to speak with abundance and yet with judgment is better than to think with the greatest acuity but without eloquence — for thought turns in upon itself, while eloquence embraces those with whom we are joined in fellowship.

\ciceroSection{157} And just as swarms of bees do not gather in order to make honeycomb but, being by nature creatures of assembly, make honeycomb as a result, so human beings — and far more so — have by nature gathered together and bring to bear their ingenuity in action and in thought. Thus, if that virtue which consists in protecting men — that is, in the fellowship of the human race — does not touch the study of nature, that study becomes an isolated and barren thing; and in the same way, greatness of spirit, cut off from human community and connection, would be a kind of savagery and brutishness. It follows that the fellowship and community of men overcomes the love of knowledge.

\ciceroSection{158} Nor is it true, as some claim, that community and fellowship with other men were entered into because without others we could not obtain and secure what nature requires for our life. If all that bears on sustenance and comfort were supplied to us by some magic wand, as they say, then every man of fine spirit would lay aside all business and devote himself wholly to knowledge and learning. Not so; for he would flee solitude and seek a partner in study, he would want to teach and to learn, to listen and to speak. Every duty, therefore, that serves the joining of men and the preservation of their community must take precedence over the duty that lies in knowledge and learning.

\ciceroSection{159} It may perhaps be asked whether this fellowship, which is most in accord with nature, must always take precedence even over self-control and modesty. I think not; for there are certain things so foul in part, and in part so flagitious, that a wise man would not do them even for the sake of preserving his country. Posidonius has assembled a great many such things, but some are so repulsive, so obscene, that they seem shameful even to name. These, then, a man will not undertake for his country's sake, nor will his country wish them undertaken on its behalf. But the matter resolves itself more conveniently than one might fear, since there can never arise a time when it is in the interest of the commonwealth for a wise man to do any of those things.

\ciceroSection{160} Let this therefore be settled: in choosing among duties, that class of duty which rests on human fellowship is supreme. And indeed, considered action will follow knowledge and wisdom; so that to act with deliberation is worth more than to think with merely speculative wisdom. So much for that. For the territory has now been opened up enough that it is not difficult to see, in the search for duty, what must take precedence over what. Within the fellowship itself, moreover, there are gradations of duty, from which we can understand what is owed to each: first to the immortal gods, second to one's country, third to one's parents, and then in descending order to the rest.

\ciceroSection{161} From this brief discussion it becomes clear that what men customarily find perplexing is not only whether a thing is honorable or base, but also — when two honorable courses are before them — which of the two is the more honorable. This topic, as I said above, was passed over by Panaetius. But let us now proceed to what remains.

\ciceroBook{2}

\ciceroSection{1} How duty is derived from what is honorable, my son Marcus, and from virtue in all its forms, I believe I have explained sufficiently in the preceding book. What follows is to pursue those kinds of duty that bear on the conduct of life and on the means by which men gain what they use — resources, abundance, wealth. In that inquiry, I said, two questions arise: first, what is expedient and what is not; second, among things expedient, which is more expedient, or most expedient of all. I am going to begin on these matters — but first I must say a few words about my purpose and my approach.

\ciceroSection{2} For although my writings have aroused many readers not only to the desire to read but even to write, still I sometimes fear that certain good men look on the name of philosophy with displeasure and wonder that I devote so much effort and time to it. As long as the state was governed by those to whom it had entrusted itself, I bent all my cares and thoughts upon it; but when everything was held fast by the rule of one man, when there was nowhere any room for counsel or for authority, when I had lost the companions of my work for the state's defense — the best of men — I gave myself neither to the anguish that would have broken me had I not resisted it, nor to pleasures unworthy of an educated man.

\ciceroSection{3} Would that the state had continued in the condition it had reached, and had not fallen into the hands of men who were driven not so much to transform it as to overturn it! In the first place, while the state stood, we should have devoted more effort to action than to writing, as we used to do; and then, what we did commit to writing would have been our own public deeds, not these present works — as we often did before. But when the state, in which all my care and thought and labor used to be invested, was altogether destroyed, those voices fell silent — the voice of the Forum and of the Senate.

\ciceroSection{4} Since my mind could not be idle, and since I had been absorbed in these studies from the start of my life, I judged that the most honorable way to lay down my troubles was to return to philosophy. I had given much of my youth to it for the sake of learning; but once I began to pursue public office and gave myself over entirely to the state, there was only so much room for philosophy as was left over from the time claims of friends and of public service — and all of that was consumed in reading; there was no leisure for writing.

\ciceroSection{5} In the midst of such great evils, then, we seem at least to have gained this good: that we have committed to writing things that were not well enough known to our countrymen and yet were altogether worthy of knowing. What is more to be desired than wisdom, in the name of the gods? What is more excellent? What is better for a human being? What more worthy of a human being? Those who seek wisdom are called philosophers, and philosophy, if you wish to render the word, is nothing other than the love of wisdom. Now wisdom, as defined by the philosophers of old, is the knowledge of all things divine and human, and of the causes by which those things are governed; to censure the study of wisdom is, for my part, something I cannot understand at all — what would such a person think worthy of praise?

\ciceroSection{6} For if it is delight of the mind and relief from care that is sought, what can be compared with the pursuits of those who are always in search of something that bears on living well and happily? Or if it is a principled account of consistency and virtue that is wanted, either this discipline is the one by which we attain them, or there is none at all. To say there is no discipline for the greatest matters, when for the smallest nothing is done without one, is the mark of men who speak without due reflection and go astray in things of the highest importance. If there is any training in virtue, where will it be sought once you have left this kind of study? But these arguments are customarily made more precisely when we are calling men to philosophy — which I have done in another work. For the present I had only to explain why, deprived of public duties, I had turned above all to this pursuit.

\ciceroSection{7} There is, however, an objection raised against us — and raised by learned and cultivated men — who ask whether we appear consistent in our conduct: for although we say that nothing can be grasped with certainty, we still commonly discourse on other subjects and at this very moment pursue the principles of duty. I wish our view were sufficiently understood by them. For we are not those whose mind wanders in confusion and can never find anything to follow. What kind of mind, or rather what kind of life, would that be, if the basis for discussion — and for living — were removed entirely? We, however, as others call some things certain and other things uncertain, so, differing from them, call some things probable and others the contrary.

\ciceroSection{8} What is there, then, to prevent me from following what appears probable to me, rejecting what appears the contrary, and, avoiding the arrogance of assertion, escaping the recklessness that is furthest from wisdom? Our school disputes every proposition precisely because that very probability could not come to light if the case were not argued on both sides. But these matters have been set out carefully enough, as I believe, in my \textit{Academica}. Now as for you, my Cicero — though you are pursuing the oldest and most distinguished school of philosophy under the guidance of Cratippus, who is most like the men who gave birth to those splendid doctrines — still, I did not wish these neighboring investigations of mine to be unknown to you. But let us proceed now to our plan.

\ciceroSection{9} Five principles were set out for pursuing duty: two pertaining to propriety and honor, two to the advantages of life — resources, abundance, wealth — and a fifth to judging between them should they appear to conflict. The part dealing with what is honorable has been completed, and I wish you to know it thoroughly. The matter we are now treating is precisely what is called the expedient. The common habit of thought has here slipped from the path and gradually been led astray to the point of separating the expedient from what is honorable, of maintaining that something may be honorable without being expedient, and expedient without being honorable — a separation that has brought greater ruin to human life than anything else could have.

\ciceroSection{10} Philosophers of the highest authority make the distinction with rigorous integrity, keeping these three concepts — which habit confounds — separate in thought. Whatever is just, they hold, is also expedient, and likewise whatever is honorable is also just; from which it follows that whatever is honorable is also expedient. Those who see this too dimly often admire clever and cunning men and mistake malice for wisdom. Their error must be torn out by the roots, and all their hope turned in this direction: let them understand that they can win what they want by honorable counsels and just actions, not by fraud and malice.

\ciceroSection{11} Now of the things that serve to sustain human life, some are inanimate — gold, silver, the products that come from the earth, and others of that kind — and some are animate, with their own impulses and desires. Of animate things, some lack reason and some possess it. Without reason are horses, cattle, and the other beasts, and bees, whose labor produces something of use and benefit for human life; those who possess reason fall into two classes: gods, and men. The gods are made propitious by piety and devotion; and next after the gods, and second to them, it is human beings who are most able to be of use to human beings.

\ciceroSection{12} The same division applies to things that harm and damage us. But since people do not suppose the gods cause harm, they except the gods and hold that it is human beings who most greatly injure human beings. For most of those things I called inanimate are in fact the product of human labor: we should not have them at all if skilled hands had not been applied, nor could we use them without human management. The care of health, navigation, agriculture, the gathering and storing of crops and all other fruits — none of these could have existed at all without human labor.

\ciceroSection{13} Likewise, the export of what we have in surplus and the import of what we lack — without human beings performing these services there would certainly be none of it. By the same reasoning, the stones the earth yields for our necessary uses would not be quarried, and iron, bronze, gold, and silver — buried deep in the earth — would not be dug out without human toil and hand. And as for houses, which ward off the force of cold and ease the discomfort of heat — where could they originally have been given to the human race, or later supplied when they fell by the violence of a storm, or earthquake, or age, if communal life had not taught men to seek the help of their fellows in such things?

\ciceroSection{14} Add to these the building of aqueducts, the channeling of rivers, the irrigation of fields, the sea-walls raised against the waves, the harbors built by human hand — from where could we have any of these without human labor? From all of this, and much else, it is plain that the benefits and advantages we derive from inanimate things could in no way be reaped by us without human hand and effort. And what yield or profit from animals could be obtained if human beings did not lend their labor? It was surely men who first discovered what use we could make of each kind of creature; and to this day, without human labor, we could neither feed nor tame nor tend them nor take from them the fruits they give in season. Even those creatures that harm us are killed by men, and those that can serve us are caught.

\ciceroSection{15} Why should I enumerate the multitude of crafts without which life could simply not have existed? How would the sick be helped, what pleasure would the healthy have, what food or civilized living — if so many crafts did not serve us? It is through these that human life, so cultivated and refined, stands so far removed from the subsistence and manner of the beasts. Cities, moreover, could neither have been built nor filled with people without human community; it was from this that laws and customs were established, the equitable distribution of rights, and an ordered discipline of living. From these followed the softening of character and a sense of restraint, so that life became more secure, and by giving and receiving, by borrowing means and lending conveniences, we found ourselves in want of nothing.

\ciceroSection{16} I have dwelt longer here than is strictly necessary. For who does not see plainly those truths that Panaetius sets out at such length — that no general in war, no statesman at home, has ever accomplished great and salutary things without the active support of other men? He calls to witness Themistocles, Pericles, Cyrus, Agesilaus, Alexander, none of whom, he says, could have achieved so much without human assistance. He is producing witnesses for a matter that is not in doubt, and so he need not call any. Yet just as we gain great advantages through the agreement and cooperation of men, so there is no plague so destructive as one that springs from man against man. There is a book by Dicaearchus, a Peripatetic of great learning and breadth, \textit{On the Destruction of Mankind}, which gathers the other causes of widespread death — floods, plague, devastation, sudden irruptions of wild beasts, by whose onslaught he shows that whole human populations have been destroyed — and then compares how many more men have been wiped out by the violence of men, that is by wars or civil strife, than by every other disaster combined.

\ciceroSection{17} Since, then, this question admits of no doubt — that men are at once the greatest source of benefit and the greatest source of harm to men — I take it as the particular work of virtue to win men's goodwill and to secure their active support for one's own purposes. Whatever advantages are drawn from inanimate things, and whatever is gained from the handling and use of animals, is credited to the practical arts; but the willing service of men toward the advancement of our affairs is stirred and sustained by the wisdom and virtue of those who stand preeminent.

\ciceroSection{18} Virtue in its entirety, in fact, turns on roughly three things. The first is discerning what in any given matter is true and genuine, what is consistent with each thing, what follows from it, what produces each result, and what the cause of each thing is. The second is restraining the turbulent impulses of the mind — what the Greeks call emotions \textit{[Greek: pathe]} — and the drives they call appetites \textit{[Greek: hormas]}, and making them obedient to reason. The third is dealing with measured judgment and understanding with those in whose company we live, so that through their willing cooperation we may have in abundance what nature requires, and through those same people may ward off whatever inconvenience is brought against us, and may punish those who have attempted to harm us — with such severity as fairness and our common humanity will allow.

\ciceroSection{19} The means by which we may acquire this capacity — to secure men's goodwill and hold it — I shall explain shortly; but a few things must be said first. Everyone knows how great a force fortune exerts in either direction, toward prosperity and toward adversity alike. When we sail before a favorable wind we reach the harbor we aimed for; when it blows against us we are driven under. Now fortune's other strokes are comparatively rare: in the first place those that come from the inanimate world — storms, tempests, shipwrecks, collapses, fires — and then those from wild beasts, their blows, their bites, their attacks; and these, as I have said, are the rarer kind.

\ciceroSection{20} But the destruction of armies — three very recently, many others in times past — the ruin of commanders, such as that most distinguished and singular man whom we lost not long ago, and the popular hatreds that lead to the expulsion, the ruin, the exile of men who have deserved well of the state, and on the other side the prosperous turns — honors, commands, victories — all these, though they depend on fortune, cannot come about in either direction without the resources and the active support of men. With this established, we must say how we can attract and engage men's goodwill to serve our purposes. If the argument runs somewhat long, let it be weighed against the magnitude of what is at stake; it may even come to seem too short.

\ciceroSection{21} Whatever, then, men confer on a man to increase his standing and honor, they do either out of goodwill — because they have some reason to be fond of him — or out of esteem, if they admire his virtue and judge him worthy of the most ample fortune; or they serve a man in whom they have placed their trust and believe he looks well after their affairs; or they fear the power of a man; or they expect something in return from him — as when kings or popular leaders hold out the prospect of some largesse; or finally they are moved by pay and hire, which is the most sordid and corrupt of all motives, both for those who are ruled by it and for those who try to resort to it.

\ciceroSection{22} For it is a poor state of affairs when what ought to be achieved by virtue is attempted with money. Yet since this expedient is sometimes unavoidable, I shall say how it is to be used — after I have first spoken of the things that come closer to virtue. Men also submit to the authority and power of another for various reasons. They are led either by goodwill, or by the greatness of benefits received, or by the distinction of a man's standing, or by the hope that it will serve their own interest, or by fear of being forced to yield to his power, or by the lure of largesse and promises, or finally — as we have so often seen in our own republic — by hired pay.

\ciceroSection{23} Of all things, nothing is more fitted to acquiring and holding power than being loved, nothing more hostile to it than being feared. Ennius says it finely: \textit{Whom they fear, they hate; and whom each man hates, he prays to see destroyed.} That no power can withstand the hatred of many, if it was not known before, has been learned of late. And it is not only the fate of that tyrant — whom the state, crushed beneath his arms, endured, and still obeys though he is newly dead — that shows how great a force men's hatred is for bringing a man to ruin; the similar ends of other tyrants make the same point, for scarcely any one of them escaped such a death. Fear is a poor guardian of anything that must last; goodwill, by contrast, is faithful, and tends toward permanence.

\ciceroSection{24} Those who hold down a subject population by force may have occasion for severity — as masters use it on slaves when they cannot be controlled any other way; but for any man who equips himself to be feared in a free state, nothing could be more insane. However deep the laws may be submerged beneath one man's power, however thoroughly liberty may be intimidated, they rise to the surface again in time — either in the silent verdicts of courts or in the secret ballots of elections. And the bite of a liberty recovered is sharper than that of one never lost. Let us therefore embrace what has the widest reach and serves not only our safety but our power and influence most of all: let us banish fear and hold fast to affection. In this way we shall most easily accomplish what we want both in private life and in the state. For those who wish to be feared must necessarily fear, in turn, those very men by whom they would be feared.

\ciceroSection{25} What agony of terror, do you think, tormented that elder Dionysius — who, fearing the barber's blades, would sear his hair with a glowing coal? And Alexander of Pherae — in what frame of mind did he live? He dearly loved his wife Thebe, as we read; and yet when he came to her bedroom from a dinner he would send ahead a barbarian — branded, as the account says, with Thracian tattoo-marks — with sword drawn, to precede him, and would dispatch members of his bodyguard to search her boxes and rummage through her clothes in case any weapon lay hidden there. A wretched man, who thought a barbarian slave-brand more trustworthy than his wife — and he was not deceived; for it was she herself who killed him, on suspicion of a mistress. No power of command is so great that, pressed down by fear, it can endure.

\ciceroSection{26} Witness Phalaris, whose cruelty is more celebrated than any other's: he did not die from an ambush, as the Alexander I just named, nor at the hands of a few men, as our own — but the entire people of Agrigentum rose against him. Consider: did not the Macedonians abandon Demetrius to a man and go over en masse to Pyrrhus? Did not virtually all the allies of the Spartans desert them the moment they began to govern unjustly, and stand by as idle spectators of the catastrophe at Leuctra? I am more comfortable reaching for foreign examples here than domestic ones. Yet for all that — as long as Rome's dominion was maintained by acts of kindness and not injustice, wars were fought either on behalf of allies or for empire; their endings were either lenient or necessary; the Senate was a harbor and refuge for kings, peoples, and nations.

\ciceroSection{27} And our magistrates and commanders sought to win their greatest distinction from precisely this one thing — defending provinces and allies with fairness and good faith. So that arrangement could more truly be called the patronage of the world than empire. We had already been letting this practice and tradition slip away gradually; but after Sulla's victory we lost them altogether. For it ceased to seem wrong to treat allies unjustly, once so great a cruelty had been shown toward citizens. In that case a just cause was followed by an unjust victory: Sulla had the audacity, when he had set up a spear in the Forum and was selling at auction the goods of men who were upright, wealthy, and at any rate citizens, to call what he was selling his own plunder. What came after him carried an impious cause to a victory still more hideous — confiscating not the property of individual citizens but sweeping whole provinces and regions under a single edict of ruin.

\ciceroSection{28} And so, with the subject peoples devastated and destroyed, we saw Massilia dragged in a triumph as a symbol of an empire already lost, and a triumph celebrated over a city without whose help our commanders never once triumphed in wars beyond the Alps. Many other outrages against our allies I could recount, if the sun had seen anything more shameful than this single act. We are therefore rightly punished. For had we not suffered the unpunished crimes of so many, so much license would never have converged upon one man; from whom the inheritance of his estate passed to a few, and the inheritance of his appetites to the many who have no business with them.

\ciceroSection{29} Nor indeed will the seeds and occasions of civil war ever be wanting, so long as depraved men remember that bloody spear and hope to see it again. It was wielded by Publius Sulla when his kinsman was dictator; and that same man, thirty-six years later, did not flinch from a still more criminal one. Another who had been a secretary under the former dictatorship became city quaestor under the latter. From which it should be understood that, with such prizes on offer, civil wars will never cease to be found. And so only the walls of the city still stand — and those very walls already trembling at crimes of the worst kind — while the republic itself we have lost utterly. Into these disasters we fell — and I must return to my theme — by choosing to be feared rather than loved and cherished. If these things could happen to the Roman people when they ruled unjustly, what should private individuals suppose awaits them? Since it is clear that the power of goodwill is great and that of fear is weak, it follows that we should explain by what means we may most easily acquire that affection — joined to honor and good faith — that we are seeking.

\ciceroSection{30} But we do not all need it in equal measure; for it must be adjusted to the life each person has chosen — whether a man requires the goodwill of many or whether it is enough to be loved by few. Let this much, then, be fixed, and fixed as the first and most indispensable thing: to have close ties of friendship with loyal friends who love us and admire what is ours. For this one thing — unlike most others — is something the greatest and the middling man must equally, and to nearly the same degree, secure for themselves.

\ciceroSection{31} Not everyone stands equally in need of honor and glory and the goodwill of their fellow citizens — yet where these things are available, they assist considerably, both in everything else one undertakes and above all in gathering friends. The subject of friendship has been treated in another book, the one bearing the title \textit{Laelius}; now let us speak of glory — though we have two volumes on that subject as well — but let us touch on it here, since it is of the greatest assistance in managing affairs of consequence. The fullest and most complete glory, then, consists in three things: that the many feel affection for you, that they place their trust in you, and that they consider you worthy of honor with a certain admiration. These things — to speak briefly and plainly — are won from the crowd by much the same means as from individuals. Yet there is also another approach to the multitude, by which we may, as it were, pour ourselves into the hearts of all.

\ciceroSection{32} Let us consider first the principles governing goodwill — the three I named above. Goodwill is won above all by benefactions; in second place it is moved by the will to do good, even when the means to do so happen to be lacking; and the love of the many is stirred very powerfully by reputation and the opinion of generosity, kindness, justice, good faith, and all those virtues that pertain to gentleness and ease of manner. For that very thing which we call what is honorable and fitting — because it pleases us for its own sake, and because it stirs the minds of all men by its nature and its appearance, and shines through above all in those virtues I have mentioned — compels us, by nature itself, to feel affection for those in whom we believe such virtues reside. These are the weightiest grounds for affection; there can be others that are less weighty.

\ciceroSection{33} Trust may be secured by two things: if we are thought to have united wisdom with justice. For we place trust in those whom we judge to understand more than we do, whom we believe to foresee what is coming and, when a matter is under way and the critical moment has arrived, to be able to resolve it and form a plan on the spot — this is what men regard as practical and genuine wisdom. As for those who are just and trustworthy — that is, men of good character — trust is placed in them on the grounds that there can be no suspicion of deceit or wrongdoing in them. So it is to such men that we consider it most right to entrust our safety, our fortunes, and our children.

\ciceroSection{34} Of these two, then, justice has the greater power in producing trust. Coupled with wisdom it needs no assistance, but wisdom without justice counts for nothing in building trust. For the more devious and cunning a man is, the more hated and suspected he becomes, once the belief in his integrity has been stripped away. Therefore, intelligence joined with justice will have as much power as it desires in producing trust; justice without wisdom will have considerable power; wisdom without justice will have none at all.

\ciceroSection{35} But someone may wonder why — when all the philosophers agree, and I myself have often argued, that whoever possesses one virtue possesses them all — I now speak as though it were possible for someone to be just without at the same time being wise. The answer is this: there is one kind of precision when truth itself is being refined in argument, and another when one's whole discourse is accommodated to common opinion. So here I speak as ordinary people speak — calling one man brave, another good, another wise — for when we are speaking of popular opinion, we must use popular, familiar language; and Panaetius did the same. But let us return to what we had proposed.

\ciceroSection{36} The third of those three elements bearing on glory was this: that we be judged by men as worthy of honor, with admiration. People, speaking generally, admire everything they find great and beyond their expectation; and individually, when they discern in some particular person certain unanticipated qualities. And so they look up to and praise to the skies those men in whom they believe they can see exceptional and singular virtues; whereas they look down on and despise those in whom they find no virtue, no spirit, no nerve. The truth is that not everyone they think ill of do they despise. Those they regard as dishonest, slanderous, and fraudulent, equipped to do injury — these they do not despise, but they do think ill of them. It is those, as I said before, who are of use to neither themselves nor anyone else — as the saying goes — in whom there is no effort, no industry, no application: these are the ones who are despised.

\ciceroSection{37} Admiration, on the other hand, is felt for those who are thought to excel others in virtue and to be free of all dishonor — and above all free of those vices to which others cannot easily resist. For pleasures, those most beguiling of mistresses, turn the minds of most men away from virtue; and when the torches of pain are brought close, the majority are terrified beyond measure. Life and death, wealth and poverty: all these move every man most powerfully. When a man of lofty spirit looks down on these things, taking them in either direction with greatness of mind, and when some fine and honorable object presents itself to him and turns him wholly toward itself and carries him away — who would not admire the splendor and beauty of virtue?

\ciceroSection{38} This scorn of the soul, then, produces great admiration; and above all, justice — from which single virtue men of good character take their name — appears to the multitude something almost miraculous, and rightly so. For no one can be just who fears death, pain, exile, or poverty, or who sets anything opposed to these above what is fair. And they admire above all the man who is unmoved by money; one in whom this has been clearly seen they regard as tried by fire. So justice accomplishes all three of those things set out as the elements of glory: goodwill, because it wishes to serve as many as possible; and for the same reason trust; and admiration, because it scorns and disregards the very things toward which most men are driven headlong by greed.

\ciceroSection{39} In my own judgment, every plan and arrangement of life requires the assistance of other men — and above all, one must have people with whom one may share familiar conversation; which is difficult unless one presents the appearance of a good man. Therefore even a man who lives in solitude and on his farm stands in need of the reputation for justice, and all the more so, because without it he will be thought unjust, and, having no protections around him, will be the victim of many wrongs.

\ciceroSection{40} Even those who buy and sell, hire and let, and are entangled in the transactions of business stand in need of justice to conduct their affairs. So powerful is justice that not even those who feed on wrongdoing and crime can live without some portion of it. For if one of those who go briganding together steals or snatches something from another, he loses his place even among the bandits; and the man who is called the chief pirate cannot hold his company unless he divides the plunder fairly, or he will be killed or deserted by his associates. Why, brigands are even said to have laws which they obey and observe. So it was that Bardulis the Illyrian bandit — about whom Theopompus records — amassed great resources through equal division of plunder, and Viriathus the Lusitanian amassed far greater; to him even our armies and their commanders yielded, until Gaius Laelius — he who is called the Wise — as praetor broke and shattered him and suppressed his ferocity so thoroughly that the war he passed on to his successors was already an easy one. Since, then, justice has such power that it even strengthens and enlarges the resources of bandits, how great a force must we suppose it to hold among laws and courts and in a properly constituted commonwealth?

\ciceroSection{41} It seems to me that — not only among the Medes, as Herodotus says, but among our own ancestors as well — kings were established in former times for the sake of enjoying justice. For when the many who were without resources were being crushed by those of greater means, they fled to a single man distinguished in virtue; and he, shielding the weaker from wrong and establishing equality by fixing a rule of fairness, held the highest and the lowest under equal law. And the reason for establishing laws was the same as the reason for establishing kings.

\ciceroSection{42} For the law has always been sought that is impartial — otherwise it would not be law at all. If the people obtained this from a single just and good man, they were content with him; when that was less readily available, laws were devised, which would speak always to all men with one and the same voice. This much, then, is clear: those who were chosen to rule were habitually men of whom the multitude had a high opinion for justice. If in addition they were thought wise as well, there was nothing men believed they could not obtain with such leaders to guide them. Justice must therefore be cultivated and maintained by every means: both for its own sake — since otherwise it would not be justice — and for the enhancement of honor and glory. But just as wealth requires not only a method of acquisition but also of investment — one that provides for continuing expenditure, not merely for necessities but for generosity as well — so glory must be both sought and invested with method.

\ciceroSection{43} Though Socrates made a fine remark: that the nearest road to glory, something of a shortcut, is this — for a man to strive to be, in truth, whatever he wishes to be thought to be. For if anyone supposes that he can win lasting glory by pretense and hollow display and a manner contrived not only in speech but even in expression, he is deeply mistaken. True glory puts down roots and spreads; everything counterfeit falls away quickly, like blossoms; nothing simulated can be long-lasting. There are very many witnesses on both sides, but for the sake of brevity we shall be content with a single family. Tiberius Gracchus, son of Publius, will be praised as long as the memory of Roman affairs endures; but his sons, while alive, were not approved by the good, and in death they hold their place in the reckoning of those justly slain. Whoever wishes to win true glory, then, let him discharge the duties of justice. What those duties are was stated in the preceding book.

\ciceroSection{44} Now, to make it as easy as possible to be seen for what we are — though the greatest force lies in actually being what we wish to be thought — certain rules of guidance ought nonetheless to be given. For if someone from an early age has a basis for celebrity and a name, whether inherited from a father — as I believe has fallen to your lot, my son — or acquired by some chance or fortune, then all eyes are fixed upon him; inquiry is made into what he does and how he lives; and, as though he moved in the brightest light, nothing he says or does can remain in the shadows.

\ciceroSection{45} Those, on the other hand, whose early years pass in obscurity and unrecognized because of lowly and undistinguished origin — once they have begun to be young men, they ought to set their gaze on great things and press toward them with earnest endeavors; and this they will do with the steadier heart, because that time of life is not merely not envied, but is actively encouraged. The first recommendation of a young man to glory, then, is whatever can be won from military service — in which many men distinguished themselves among our ancestors, for there was almost always a war being waged. Your own age fell into a war of which one side had too much of crime and the other too little of fortune. Yet even in that war, when Pompey had placed you in command of a cavalry wing, you were earning great praise — both from a man of the highest eminence and from the army — for your riding, your javelin-work, your endurance of every military hardship. Yet that praise of yours fell along with the commonwealth. This discourse of mine, however, is not undertaken about you in particular, but about the whole class; let us therefore proceed to what remains.

\ciceroSection{46} As in other matters the works of the mind far outweigh those of the body, so the things we pursue through intelligence and reason win more gratitude than those we pursue through strength. The first recommendation, then, comes from a man's modesty joined with piety toward his parents and goodwill toward his own people. But young men are recognized most easily, and in the best light, when they attach themselves to men of distinction and wisdom who give good counsel to the commonwealth. If they are frequently in such company, they create in the public mind the expectation that they will come to resemble those they have chosen to imitate.

\ciceroSection{47} The house of Publius Mucius commended the young Publius Rutilius to a reputation for both integrity and legal knowledge. Lucius Crassus, for his part — though still quite young — did not borrow his greatest distinction from elsewhere: he earned it himself, through that celebrated and glorious prosecution; and at an age when the sort of practice that gave Demosthenes his reputation is usually employed at home, Lucius Crassus showed that he was already doing that very thing in the Forum, and doing it superbly.

\ciceroSection{48} Now since discourse falls into two kinds — one is conversation, the other advocacy — there is no question that advocacy carries greater power for glory (it is what we call eloquence); yet it is remarkable how much affability and ease in conversation win men's goodwill. There survive letters of Philip to Alexander, of Antipater to Cassander, of Antigonus to his son Philip — three of the most prudent men, so we are told — in which they instruct their recipients to attract the goodwill of the multitude through kindly speech and to soften the soldiers with flattering address. But advocacy delivered before a crowd often kindles universal glory; for the admiration felt for one who speaks with fullness and wisdom is great — those who hear him judge that he understands and thinks more deeply than others. If there is in that advocacy a gravity tempered with modesty, nothing more admirable can be imagined; and the more so when these qualities appear in a young man.

\ciceroSection{49} But since there are several kinds of causes that call for eloquence, and many young men in our commonwealth have won distinction by speaking before judges, before the people, and before the Senate, the greatest admiration belongs to the courts. The subject is twofold: there is prosecution and there is defense; and though defense is the more honorable of the two, prosecution too has often won approval. I mentioned Crassus a moment ago; Marcus Antonius did the same when young. The eloquence of Publius Sulpicius was also made famous by a prosecution, when he brought to trial that turbulent and useless citizen, Gaius Norbanus.

\ciceroSection{50} This, however, is not to be done frequently, nor at any time unless it is for the sake of the commonwealth — as those I have named — or for revenge, as the two Luculli, or in fulfillment of an obligation, as I acted for the Sicilians, and Julius for the Sardinians against Albucius. In prosecuting Manius Aquilius the diligence of Lucius Fufius also became known. Once, then, or at any rate not often. But if a man finds he must do it more frequently, let him offer it as a service to the commonwealth, whose enemies it is not blameworthy to pursue more than once; yet even so, moderation must be maintained. For it seems the mark of a hard man — or scarcely a man at all — to bring capital charges against many. It is dangerous to oneself, and dishonorable to one's reputation, to run the risk of being called an accuser; a stigma that befell Marcus Brutus, a man born of the highest lineage and the son of him who was among the foremost authorities on civil law.

\ciceroSection{51} There is also a rule of duty to be kept with care: never arraign an innocent man on a capital charge — that can be done only by a crime. For what is so inhuman as to turn an eloquence given by nature for the preservation and safety of mankind into an instrument of ruin and destruction for the good? Yet just as this must be avoided, so there need be no scruple about defending a guilty man now and then, provided he is not wicked and impious — the public expects it, custom permits it, and human feeling even endorses it. The duty of the judge is always to pursue the truth in a case; the duty of the advocate is sometimes to defend what is plausible, even if it falls short of the truth. I should not venture to write this — especially writing on philosophy as I am — were it not the view of the most weighty of the Stoics, Panaetius himself. The greatest glory and goodwill come above all from defense, and the greater still when one comes to the aid of a man who seems to be hemmed in and crushed by the power of some great personage — as I have often done in other cases, and as I did when young in defending Sextus Roscius of Ameria against the power of Lucius Sulla at the height of his domination; that speech, as you know, is still in circulation.

\ciceroSection{52} Now that I have set out the duties of the young that bear on the winning of glory, I must speak next of beneficence and generosity. Its method is twofold: we can show kindness either through personal service or through money. The latter is the easier kind, particularly for a man of means, but the former is finer, more distinguished, and worthier of a brave and eminent man. For though both spring from a generous wish to confer benefit, the one is drawn from the treasury and the other from character; and the lavish giving that is funded from one's private estate drains the very spring of generosity. Thus generosity is annulled by generosity: the more people you have served with it, the fewer you will be able to serve in the future.

\ciceroSection{53} But those who are beneficent and generous through personal service — that is, through character and effort — will first find that the more people they benefit, the more helpers they will have in doing good; and then, through the habit of beneficence, they will be better prepared and, as it were, more practiced in deserving well of many. Philip upbraids his son Alexander in a famous letter for seeking to buy the goodwill of the Macedonians with gifts: ``What wretched reasoning,'' he writes, ``led you to such a hope as to expect that men whose favor you had purchased with money would be loyal to you? Or is it your aim that the Macedonians should look to you not as their king but as their steward and caterer?'' A fine phrase, ``steward and caterer'' — because such a role is degrading for a king; and even finer was his calling lavish giving a form of corruption. For the man who receives becomes worse thereby, and ever more ready to expect the same again.

\ciceroSection{54} He said this to his son, but let us take it as a lesson for all. It is therefore beyond doubt that the generosity which consists in personal service and effort is both more honorable and more far-reaching and can benefit more people. Yet sometimes money must be given; this kind of generosity is not to be rejected altogether, and we must often contribute from our private means to those in genuine need — but carefully and in measure. Many men have squandered their estates by giving without thought. And what is more foolish than to arrange things so that you can no longer go on doing what you enjoy doing? Furthermore, lavish giving is followed by rapacity: when men begin to run short through giving, they are forced to lay hands on what belongs to others. So it comes about that when they wished to be generous in order to win goodwill, they fall short not so much in the affection of those to whom they gave as in the hatred of those from whom they have taken.

\ciceroSection{55} A man's private resources should therefore not be locked so tight that generosity cannot open them, nor left so open that they lie bare to everyone. A measure must be applied, and that measure must be referred to one's means. We must remember entirely the saying that has become a proverb among our people through constant use: that lavish giving has no bottom. And indeed, what limit can there be, when both those who have grown accustomed to receiving want the same again, and others want the same thing as well? In general there are two kinds of generous men: the prodigal, and the truly liberal. The prodigal are those who pour out money on feasts, distributions of meat, gladiatorial shows, the staging of games and hunts — extravagances from which they will leave behind a memory either brief or none at all.

\ciceroSection{56} The truly liberal, on the other hand, are those who use their resources to ransom captives from bandits, to assume the debts of friends, to help in settling their daughters in marriage, or to give assistance in acquiring or increasing a livelihood. I am therefore puzzled at what moved Theophrastus in that book he wrote on wealth — in which much is excellent — to say something so absurd: he expatiated at great length in praise of the magnificence and apparatus of public entertainments, and treats the means to such expenditure as the fruit of wealth. To my mind that fruit of generosity, of which I have offered a few examples, seems far greater and far more reliable. Aristotle rebukes us far more seriously and truly for not being struck with wonder at those lavish outpourings of money made to win the crowd's favor. For he says that when people besieged by an enemy are forced to buy a pint of water at a mina, this at first seems incredible to us and we all marvel at it — but when we turn our minds to the necessity, we grant it pardon. Yet at these monstrous and limitless expenditures we feel no great surprise — and this though they serve no necessity, they increase no dignity, and even the pleasure the crowd takes in them is of the briefest and most meager duration, dying out even in the most shallow among them together with the satiety itself, leaving no trace of memory behind.

\ciceroSection{57} He also rightly concludes that these things please children and women and slaves and men hardly better than slaves, but can by no means win the approval of a serious man who weighs with sound judgment what is actually happening. Yet I understand that in our own city, even in better times, a tradition has taken hold by which a brilliant aedileship is expected of the best men. And so Publius Crassus — surnamed the Rich, and equal to that name in his resources — discharged the greatest of aedilician shows; then a little later Lucius Crassus, along with the most temperate of all men, Quintus Mucius, gave the most magnificent aedileship on record; then Gaius Claudius, son of Appius, and many after them — the Luculli, Hortensius, Silanus; but Publius Lentulus in my consulship outstripped all his predecessors; Scaurus followed his example; and then came the most magnificent shows of all, those of our Pompey in his second consulship. You see what I think of all of this.

\ciceroSection{58} Nevertheless one must beware of the suspicion of meanness. Mamercus, a man of very great wealth, had the consulship denied him because he had passed over the aedileship. When the people demand it, therefore, it must be done — if not to please even good men, at least with their approval — but in proportion to one's means, as I myself did; and when some more substantial and useful end is to be gained from a public outlay, as recently when Orestes's dinner parties in the streets under the title of the tithe brought him great honor. And Marcus Seius was not blamed for supplying the people with grain at a penny a peck in a time of scarcity; he freed himself at no dishonest price — though not at a very large one either — from a great and long-standing ill will, when he was aedile. But the highest honor recently went to our Milo, who by purchasing gladiators for the defense of the commonwealth — which was bound up with my own safety — suppressed all the plots and outrages of Publius Clodius. The justification for spending, then, lies in necessity or in advantage.

\ciceroSection{59} In all these matters the rule of moderation is the best guide. Lucius Philippus, son of Quintus — a man of great gifts and in the first rank of distinction — used to boast that he had won everything counted most eminent without any public shows at all. Cotta said the same, and Curio. I may take some credit in this regard myself: for, measured against the distinction of the offices I won with the votes of every tribe — in my own year, which fell to none of those I have just named — my expenditure on the aedileship was quite modest.

\ciceroSection{60} And better still are those expenditures made on walls, dockyards, harbors, aqueducts, and all that pertains to the practical needs of the commonwealth. Though what is given directly, as it were, into the hand gives more immediate pleasure, these things prove more welcome in time to come. I hesitate, out of deference to Pompey, to criticize theaters, colonnades, and new temples — but the most learned men have not approved of them: this same Panaetius, whom I have largely followed in these books — though I have not translated him — and Demetrius of Phalerum, who reproaches Pericles, the foremost man of Greece, for pouring such wealth into those splendid Propylaea. The whole subject has been carefully argued in the books I wrote on the commonwealth. All lavish giving of this kind is, taken as a whole, flawed in principle, sometimes made necessary by circumstance — and even then must be fitted to one's means and kept within the rule of moderation.


\ciceroSection{61} In that other kind of giving, the kind that flows from generosity itself, our dispositions ought not to be the same toward every case, since the cases differ. The situation of someone crushed by calamity is one thing; the situation of someone seeking to rise higher while he suffers no particular reverse is quite another.

\ciceroSection{62} Our kindness ought to lean more readily toward the calamity-stricken — unless they happen to deserve their calamity. Yet toward those who want help not to keep from going under but to climb to a higher standing, we must not hold back entirely; but we should bring judgment and care to selecting the right recipients. Ennius put it admirably: \textit{Good deeds badly placed I count as bad deeds.}

\ciceroSection{63} When a benefit is given to a man of integrity and gratitude, the giver gains something both from him and from everyone else. Generosity freed from recklessness is the most welcome thing of all; and most people praise it the more eagerly because the goodness of every man of high standing is a common refuge for all. We must take care, therefore, to confer benefits on as many as possible, in a way whose memory will be handed down to children and posterity, so that ingratitude toward those benefactors becomes impossible. For all men hate one who forgets a benefit, and they regard that ingratitude as an injury done to themselves as well — a deterrent to generosity — and think the man who commits it a common enemy of those who have less. This kind of kindness is also useful to the commonwealth: ransoming captives from slavery, relieving the poor — a practice, as we can see from the published speech of Crassus, that was frequently performed by our own order with great generosity of expression. I therefore set this habitual kindness far above lavish gift-giving; the one belongs to men of weight and greatness, the other is a kind of flattery of the people, tickling the fickleness of the crowd with the pleasure of handouts.

\ciceroSection{64} It will suit a man both to be generous in giving and to be not harsh in collecting, and in every kind of transaction — buying and selling, letting and hiring — in relations with neighbors and boundary-holders, to be fair, accommodating, yielding much of his strict legal right to others, and keeping as far as possible from litigation — further, I am inclined to say, than his strict right requires. For it is not only generous, but sometimes profitable as well, to give ground now and again in one's legal rights. One must also attend to the family estate — to let it be squandered away is disgraceful — but in such a way that no suspicion of meanness or greed attaches; for the greatest gain from wealth is, without doubt, to be able to be generous without stripping oneself of what one has. Theophrastus was right, too, to praise hospitality; it is indeed, as it seems to me, highly fitting that the house of a man of distinction should stand open to distinguished guests, and it is an ornament to the commonwealth as well that foreigners visiting our city should not lack for this form of generosity. For those who wish to have great influence by honorable means, it is also highly advantageous to be well connected through ties of hospitality among foreign peoples. Theophrastus records that Cimon of Athens was hospitable even to his own fellow demesmen, the Laciadae: he arranged it with his farm managers, giving standing orders that any Laciad who came to stay at his estate should be provided with everything he needed.

\ciceroSection{65} Benefits given through effort and service rather than through money are conferred on the commonwealth at large and on individual citizens alike. For giving legal counsel, helping with advice, and offering through this kind of knowledge to as many as possible bears powerfully on the growth of both resources and influence. Our ancestors were distinguished in much, but not least in this: the knowledge and interpretation of civil law — a supremely honored pursuit always — was held throughout that time by men of the first rank in its proper place. Now, as honors and every grade of distinction have been extinguished, so too the luster of this knowledge has been destroyed — and that all the more outrageously, because it came about at the very time when there was a man living who could easily have surpassed all his predecessors, his equals in rank, in learning. These services, then, win the gratitude of many and are well fitted to binding men by obligations.

\ciceroSection{66} Closely related to this pursuit, but weightier, more welcome, and more distinguished, is the gift of eloquence. What is more admirable than eloquence — in the wonder it draws from those who hear it, the hope it gives to those in need, the gratitude of those who have been defended? To eloquence, therefore, our ancestors gave the first place in dignity among the civilian callings. The service and protection of a man who is fluent, who works readily, and who — as the custom of our ancestors held — defends many cases without reluctance and without fee, extend very widely.

\ciceroSection{67} My subject prompted me here to lament the suspension of eloquence — or should I say its destruction — did I not fear I might seem to be complaining about my own situation. And yet we see how few orators are left to carry any hope since those we knew were silenced, how much fewer still have any real capacity, how many have sheer impudence instead. But even if not everyone can be learned in law — and certainly not many — or eloquent, it is still possible to serve many through effort: by seeking favors, by commending clients to judges and magistrates, by keeping watch over another's affairs, by going to those very men who are being consulted or who are conducting the defense and soliciting them in person. Those who do this gain the widest gratitude, and their industry spreads its influence very broadly.

\ciceroSection{68} There is no need to remind anyone — it is perfectly obvious — that in wishing to help others one must take care not to give offense. For people often harm those they ought not to harm, or those it is not to their own advantage to harm; if from inattention, that is negligence; if knowingly, recklessness. One must also offer an apology to those one unavoidably offends — by whatever means one can — explaining why what was done was necessary and could not have been done otherwise; and whatever appearance of wrong remains must be made good through other services and attentions.

\ciceroSection{69} Now, in helping people, it is usual to look at either character or fortune; and easy enough to say — as people commonly do — that in placing benefits we follow a man's character, not his fortune. That is an honorable way to speak; but who, in practice, does not prefer, when it comes to giving effort, the claim of a wealthy and powerful man over that of a virtuous but poor one? For we naturally incline more toward the man from whom repayment seems easier and swifter. But we ought to look more carefully at how things actually stand. That poor man — if he is a man of integrity — even if he cannot repay a favor, can at least feel it. There is a saying, put as neatly as possible, that the man who has money and has not repaid is in debt, while the man who has repaid no longer has it; but gratitude is something a man both possesses when he has shown it and has shown even when he merely possesses it. As for those who think themselves wealthy, honored, and fortunate — they are unwilling to be put under any obligation at all; they even think they have done you a favor when they themselves have received something, however great; and beyond that they suspect that something is being demanded or expected of them, while to have availed themselves of a patron's protection or to be called a client they regard as no better than death.

\ciceroSection{70} The man of modest means, on the other hand — whatever has been done for him, he takes it as a measure of himself, not of his fortune; he tries to appear grateful not only to the man who has deserved it but also to those from whom he hopes for more in the future, for he has need of many; and he does not magnify whatever service he in turn may render but actually understates it. One should keep this in mind as well: if you have defended a wealthy and fortunate man, his gratitude stays with him alone — or at most with his children; but if you defend a man of no means who is nonetheless upright and modest, all men of no means who are not themselves base — and that is a large portion of the people — see in it a shelter prepared for themselves.

\ciceroSection{71} I therefore hold that a benefit is better placed with the good than with the fortunate. We must of course endeavor to give satisfaction to every kind of person; but if a contest comes down to a choice, we should certainly take Themistocles as our guide. When asked whether he would rather give his daughter to a man who was virtuous but poor or to a man who was less reputable but rich, he said: "For my part, I would rather have a man who needs money than money that needs a man." But our standards have been corrupted and perverted by admiration for wealth. What does the size of another man's fortune have to do with any of us? It may benefit the man who holds it — though not even that always. Suppose it does benefit him: does that make him more decent? Grant that he is also a man of character: let his wealth not stand in the way of his being helped — only let it not count in his favor either — and let the whole judgment rest not on how rich a man is but on what kind of man he is. The final rule in giving effort and benefit: never press a claim against equity, never act in support of injustice; for the foundation of lasting reputation and renown is justice, without which nothing can be praiseworthy.

\ciceroSection{72} Since we have discussed the kind of benefits that concern individual persons, we must next consider those that concern people at large and the commonwealth. Of these, some are of a kind that touch the citizenry as a whole, others that reach individual persons — which are the more welcome. We should try, when possible, to attend to both, but no less to take thought for individuals — provided that what is done either benefits the commonwealth or at least does not harm it. The great grain distribution of Gaius Gracchus was emptying the treasury; the modest one of Marcus Octavius was tolerable for the state and necessary for the common people — it was therefore salutary for both citizens and commonwealth alike.

\ciceroSection{73} Above all, the man who administers the commonwealth must take care that each person holds what is his own, and that there is no diminution of private property by public action. Philippus acted ruinously when, as tribune, he proposed his agrarian law — though he let it be rejected without much resistance and showed in that great moderation — but he said many demagogic things in the course of the proceedings, and above all this pernicious thing: that there were not two thousand men of property in the entire state. A fatal speech, tending toward an equalization of wealth; what plague could be greater? It was for this very reason, above all, that states and communities were established — so that each man might hold what is his. For though men were drawn together by nature's guidance, it was in hope of protection for their property that they sought the shelter of cities.

\ciceroSection{74} Care must also be taken that there be no need to exact a property tax — something that used to happen frequently among our ancestors because of the treasury's lean condition and the unbroken succession of wars; and provision for this must be made well in advance. But if some state is overtaken by the necessity of this levy — I would rather speak of some other state than omen it upon our own, though my discussion concerns every commonwealth, not ours alone — every effort must be made so that all understand, if they wish to survive, that they must submit to necessity. And moreover all who govern a commonwealth must take care that there is an adequate supply of those things that are necessary. How such provision is regularly and rightly made there is no need to discuss — it is plain enough; I need only have touched on the subject here.

\ciceroSection{75} The cardinal point, in every conduct of public business and public office, is that even the slightest suspicion of greed be banished. "If only," said Gaius Pontius the Samnite, "fortune had preserved me for those times and I had been born when the Romans began to take bribes — I would not have allowed them to rule much longer." He would have had to wait many centuries yet; this disease has only recently invaded our commonwealth. And so I am quite content that Pontius lived when he did, since he had such strength in him. Not yet a hundred and ten years have passed since the law on extortion was carried by Lucius Piso — there had been none before it. But look at what followed: so many laws, each one harsher than the last, so many defendants, so many convicted, the great Italian War stirred up from fear of the courts, such widespread plundering and pillaging of the allies when the laws and the courts were swept aside — so that it is by the weakness of others, not by our own virtue, that we remain strong.

\ciceroSection{76} Panaetius praises Africanus for his restraint in money matters. Why not? But in that man other qualities were greater; the praise of such restraint belongs not to the man alone but to the age. Paullus, having gained possession of the entire Macedonian treasury — and it was enormous — brought so much money into the public treasury that the spoils of a single commander put an end to the payment of tribute. Yet he brought nothing home to his own house but the eternal memory of his name. Africanus followed his father's example and was no richer for the destruction of Carthage. And what of Lucius Mummius, who was his colleague in the censorship? Did he grow any richer for having razed to the ground the most prosperous city in the world? He preferred to adorn Italy rather than his own house — and yet Italy, so adorned, strikes me as a more splendid house in itself.

\ciceroSection{77} So — to bring the argument back to where it started — there is no vice more hateful than greed, above all in those who lead and govern the commonwealth. To treat the commonwealth as a source of private profit is not merely shameful; it is criminal and impious. And so the oracle that Apollo at Pytho delivered — that Sparta would perish through nothing other than greed — seems to have been a prophecy not for the Spartans alone but for all wealthy peoples. Those who stand at the head of the commonwealth can win the goodwill of the many by no means more readily than by restraint and self-control.

\ciceroSection{78} Those, however, who want to pass for men of the people and for that reason either attack the agrarian laws so as to drive out the current holders, or think that debts owed to creditors should be remitted, undermine the very foundations of the commonwealth — first, harmony, which cannot exist when money is stripped from some and handed to others; and second, equity, which is abolished altogether when each man is not permitted to keep what is his own. For it is the peculiar function of the state and the city, as I said above, that each man's custody of his own property should be free and undisturbed.

\ciceroSection{79} And by bringing this ruin upon the commonwealth they do not even gain the gratitude they expect. For the man whose property has been seized is an enemy; the man to whom it has been given dissembles even his willingness to receive it and — above all in the case of remitted debts — conceals his own satisfaction, lest he be seen to have been insolvent; whereas the man who has suffered the wrong both remembers it and wears his resentment openly. And even if those who have been wrongfully enriched outnumber those who have been unjustly despoiled, the latter are not thereby of less weight: for these things are judged not by numbers but by gravity. And what justice is there in this — that a man who had no land should hold land that has been possessed for many years, even generations, while the man who held it should lose it?

\ciceroSection{80} It was precisely this kind of injustice that led the Spartans to expel the ephor Lysander and to put to death King Agis — something that had never happened among them before — and from that time such great discord followed that both tyrants arose and the best men were driven into exile and the most admirably constituted commonwealth fell apart. Nor did Sparta alone fall: she brought down the rest of Greece as well through the contagion of those evils, which spread outward from Sparta as their source. And what of our own Gracchi — the sons of that great man Tiberius Gracchus, grandsons of Africanus? Were they not destroyed by the agrarian controversies?

\ciceroSection{81} Aratus of Sicyon, on the other hand, is rightly praised. When his city had been held by tyrants for fifty years, he set out from Argos, made his way by stealth into Sicyon, seized the city, overwhelmed the tyrant Nicocles by surprise, restored six hundred exiles — men who had been the wealthiest in that community — and freed the commonwealth by his arrival. But when he observed the great difficulty regarding property and estates — since he judged it most inequitable that those whom he had himself restored should be in want when others were in possession of their goods, while he also thought it not altogether just to disturb possessions of fifty years' standing, because in so long a span much had passed by inheritance, much by purchase, much by dowry, without any wrong being done — he concluded that neither should those current holders be stripped of their property, nor should those to whom it had once belonged be left without satisfaction.

\ciceroSection{82} Having decided, then, that money was needed to settle the matter, he announced his intention of going to Alexandria, ordered that everything remain as it was until his return, and went at once to Ptolemy — a personal guest-friend of his, the second king to reign after the founding of Alexandria. To him he explained that he wished to free his homeland and set out the case; and from so wealthy a king so distinguished a man easily obtained substantial financial assistance. He brought this money to Sicyon, took fifteen leading citizens into his council, examined with them the claims both of those who held what was not theirs and of those who had lost what was theirs, and by valuing the estates effected this: he persuaded some to accept money in exchange for relinquishing possession, and persuaded others that it was more convenient to receive a sum equal to the value of their property than to recover the property itself. So the matter was settled, harmony restored, and all parties went away without complaint.

\ciceroSection{83} What a great man — and how worthy to have been born in our own commonwealth! This is how one should deal with citizens: not, as we have now seen done twice, by planting the spear of confiscation in the Forum and putting citizens' goods up to the auctioneer's cry. But that Greek, with the wisdom and excellence of a truly outstanding man, held that the interests of all must be considered; and this is the supreme method and wisdom of the good citizen — not to tear apart the advantages of his fellow citizens but to hold them all together in the same equity. "Let them live rent-free in another man's house." Why so? So that when I have bought, built, maintained, and spent, you may enjoy what is mine against my will? What else is this but seizing what belongs to some and handing it to others?

\ciceroSection{84} As for debt cancellation — what is its logic, except that you use my money to buy a piece of land, which you then keep while I no longer have my money? One must take care, therefore, that no debt arises to harm the commonwealth, and this can be guarded against by many means; but the remedy, when debt has arisen, is not that the propertied should lose what is theirs while debtors profit from what belongs to others. For nothing holds the commonwealth together more powerfully than good faith, and good faith cannot exist unless the repayment of what has been lent is a binding obligation. During my consulship the agitation to cancel debts was never more violent; men of every class and rank tried to bring it about with arms and encampments; and I resisted with such success that this whole evil was removed from the commonwealth entirely. Never was debt greater, never was it paid off more fully or with greater ease; for once the hope of evading payment was removed, the necessity of paying followed. But the man who is now the victor — who was at that time defeated — carried out what he had planned, once I was no longer of any relevance to him. Such was his passion for wrongdoing that the mere act of doing wrong gave him pleasure, even when he had no purpose to serve.

\ciceroSection{85} Those who protect the commonwealth, then, will keep clear of this kind of largess — taking from some to give to others — and will above all take care that every man keeps what is his by the equity of law and the courts; that the poor are not exploited on account of their low station; that the wealthy are not harmed by prejudice when holding or recovering their property; and that, by whatever means are available in war or in peace, they extend the empire, the territories, and the revenues of the commonwealth. These are the acts of great men; these were habitually performed by our ancestors; and those who pursue these kinds of duty will earn for themselves, alongside the greatest benefit to the commonwealth, great gratitude and glory alike.

\ciceroSection{86} Now, among these precepts on expediency, Antipater of Tyre — a Stoic who died recently at Athens — considers that two things were passed over by Panaetius: the care of health and the management of money. These I believe the supreme philosopher omitted because they were straightforward; they are certainly useful. Health is maintained by knowledge of one's own body and by observing what tends to help it and what tends to harm it; by self-discipline in all aspects of diet and bodily care, through the forgoing of pleasures for the sake of maintaining fitness; and finally by the art of those whose professional knowledge covers these things.

\ciceroSection{87} Private property, moreover, should be acquired by means free of anything shameful, preserved by diligence and thrift, and increased by those same means. These matters Xenophon the Socratic treated most helpfully in the book entitled \textit{Oeconomicus}, which I translated from Greek into Latin when I was roughly the age you are now, my son. But the whole topic of acquiring, investing, and — I could wish — spending money is more conveniently discussed by certain excellent men who sit near the middle arch of the Janus than by any philosopher in any school. These subjects must nonetheless be known; for they bear on expediency, which is the subject of this book.

\ciceroSection{88} Comparison of expedient things — since this was the fourth topic, passed over by Panaetius — is often necessary. The goods of the body are commonly compared with external goods, external goods with bodily goods, and bodily goods with one another, and external goods with one another. Bodily goods are compared with external goods in this way: one would rather be healthy than wealthy; external goods with bodily goods in this way: one would rather be wealthy than possessed of exceptional physical strength. The comparison of bodily goods with one another works like this: good health is ranked above pleasure, strength above speed; and among external goods: glory above riches, and city revenues above those from the country.

\ciceroSection{89} From this kind of comparison comes the famous answer of old Cato. When someone asked him what was most profitable in private estate management, he replied: ``Good pasturage.'' And second? ``Moderately good pasturage.'' Third? ``Poor pasturage.'' Fourth? ``Arable farming.'' And when the questioner asked, ``What of lending at interest?'' Cato replied: ``What of killing a man?'' From this and many other examples it should be understood that comparisons of expedient things are regularly made, and that this fourth class of inquiry into duties has rightly been added. We shall pursue what remains in due order.

\ciceroBook{3}

\ciceroSection{1} Cato — who was roughly Scipio's contemporary — wrote that Publius Scipio, my son, the man who was first called Africanus, used to say that he was never less idle than when at leisure, never less alone than when alone. A magnificent saying, and worthy of a great and wise man; for it tells us that he was accustomed, even in leisure, to think on affairs of state, and even in solitude to converse with himself — so that he was never truly idle, and at times had no need of another's company. The very two things that bring languor to other men sharpened him: leisure and solitude. I wish I could truly say the same of myself; but if I cannot reach so great an eminence of talent by imitation, I am at least near it in intention — for I too pursue leisure, shut out by armed impiety and force from public life and the business of the courts, and for that reason, having left the city to wander through the countryside, I am often alone.

\ciceroSection{2} Yet this leisure cannot be compared with Africanus's leisure, nor this solitude with his. For he would take leisure occasionally as a rest from the most splendid duties of public life, and would withdraw from the crowds and company of men into solitude as into a harbor; whereas my leisure has been imposed by the want of anything to do, not chosen out of desire for rest. When the Senate is extinguished and the courts destroyed, what is there that I, as I should be, can worthily do in the curia or in the Forum?

\ciceroSection{3} And so, having once lived amid the greatest throng and in the eyes of my fellow citizens, I now flee the sight of the criminals who overflow everything, withdraw as much as I may, and am often alone. But since we have learned from men of learning that one ought not merely to choose the least among evils but also to draw out of those evils themselves whatever good may be lodged in them, I therefore make use of my leisure — not, to be sure, the leisure that ought to have been enjoyed by one who had once won leisure for his country — and I do not allow the solitude, which necessity and not choice imposes on me, to sink into languor.

\ciceroSection{4} Though Africanus, in my judgment, achieves the greater glory. For there survive no written records of his genius, no work of his leisure, no gift of his solitude; from which we must understand that he was never truly idle or alone, thanks to the activity of his mind and the investigation of those matters he reached by thought. We, who have not strength enough to be drawn out of solitude by silent reflection, have turned all our zeal and care to this work of writing. And so in a short time of public collapse we have written more than we wrote in many years when the republic stood.

\ciceroSection{5} But though all philosophy, my dear Cicero, is fertile and fruitful and no part of it lies untilled and waste, still no region within it is more productive or more rich than that on duty, from which the precepts for a life of honor and consistency are drawn. Therefore, although I am confident that you hear and receive these things constantly from our friend Cratippus — the foremost philosopher of this age — yet I think it will be advantageous to have your ears resounding with such words on every side, and, if possible, to have them hear nothing else.

\ciceroSection{6} This must be done by all who are minded to enter upon an honorable life, but I am not sure it must be done by anyone more than by you. You bear no small expectation of following in my own industry, a great one of equalling my honors, and some, perhaps, of my name. You have taken on besides the considerable burden of Athens and of Cratippus; since you went out to them as to a mart of the finest accomplishments, it would be most shameful to return empty, bringing discredit on the authority both of the city and of your teacher. Therefore, strain your mind as hard as you can, press as hard in your labor — if studying is a labor and not a pleasure — and see to it that you succeed, and do not allow it to appear that, when everything has been supplied by me, you have been wanting to yourself. But enough of this; I have written to you often by way of encouragement. Let me now return to the remaining part of the division I laid out.

\ciceroSection{7} Panaetius, then — who beyond all question discussed duty with the greatest exactness, and whom I have followed above all others, with some correction — set out three categories in which men ordinarily deliberate and take counsel about duty: first, when they are in doubt whether what they are considering is honorable or base; second, whether it is expedient or inexpedient; and third, when something that bears the appearance of the honorable conflicts with what seems expedient, how those two are to be distinguished. He worked through the first two categories in three books; but of the third he wrote that he would treat it next, and he did not fulfill what he had promised.

\ciceroSection{8} This surprises me the more, because a disciple of his, Posidonius, records that Panaetius lived thirty years after those books were published. I find it remarkable that Posidonius touched on that topic only briefly in certain notes — especially since he writes that there is no topic in all of philosophy so necessary.

\ciceroSection{9} I am by no means persuaded by those who claim that this topic was not passed over by Panaetius but deliberately left aside, and that it should not have been written on at all, because expediency can never conflict with the honorable. Whether it was proper to include this third category, as Panaetius did, or to omit it entirely is a question that admits of debate; but there can be no doubt that Panaetius undertook it and left it unfinished. For a man who has completed two parts of a three-part division must necessarily have a third remaining; and furthermore, at the end of the third book he promises to treat this part next.

\ciceroSection{10} A credible witness is added to this in Posidonius, who also writes in a certain letter that Publius Rutilius Rufus — who had heard Panaetius — used to say that just as no painter had been found to complete that portion of the Coan Venus which Apelles had left unfinished (for the beauty of the face robbed all hope of imitating the rest of the body), so no one had followed up what Panaetius had left aside and unfinished — because of the surpassing excellence of what he had finished.

\ciceroSection{11} On Panaetius's intention, therefore, there can be no doubt. Whether he was right to add this third inquiry into duty, or otherwise, is perhaps open to discussion. For if the honorable alone is good, as the Stoics hold, or if what is honorable is the highest good in such a way — as your Peripatetics think — that everything placed on the other side barely has the weight of the slightest moment, then there can be no doubt that expediency can never contend with the honorable. And so we are told that Socrates used to curse those who first, by mere opinion, had separated things naturally bound together. The Stoics have agreed with him to the extent that they judge whatever is honorable to be expedient, and nothing expedient that is not honorable.

\ciceroSection{12} Now if Panaetius were the kind of man who said virtue is to be cultivated because it produces expediency — as those do who measure desirable objects by pleasure or freedom from pain — it would be open to him to say that expediency sometimes conflicts with the honorable. But since he is one who judges nothing good except what is honorable, and holds that those things which resist it under some appearance of expediency neither make life better by their addition nor worse by their removal, he does not seem to have been right to introduce deliberation of the kind in which what appears expedient is weighed against what is honorable.

\ciceroSection{13} Indeed what the Stoics call the highest good — to live in accordance with nature — contains, as I understand it, this meaning: to be always in harmony with virtue, and to choose the remaining things that accord with nature, so long as they do not conflict with virtue. This being so, some think that this comparison was wrongly introduced and that nothing at all should be prescribed in that category. And indeed that honorable thing properly and truly so called exists in the wise alone and can never be severed from virtue; but among those in whom wisdom is not perfect, perfect honor is in no way possible, though likenesses of the honorable can be.

\ciceroSection{14} For the duties discussed in these books the Stoics call intermediate; they are common and widely shared, and many attain them through natural goodness of character and through progress in learning. But that other duty which they likewise call right is perfect and complete and, as they say, fully rounded, and can fall to none but the wise man.

\ciceroSection{15} But when something is done in which these intermediate duties appear, it tends to look fully perfect — because the common run of men do not generally grasp what is lacking from the perfect, whereas as far as they do grasp it they think nothing has been left out. The same thing happens with poetry, with painting, and many other things: ordinary folk take delight in them and praise what should not be praised, I suppose for this reason — that there is in those works something sound that catches the untrained, who are unable to judge what flaw each work contains; and so, when the learned have instructed them, they readily abandon their earlier view. These duties, then, of which we treat in these books, they call a kind of secondary honorableness — not the property of the wise man alone, but shared with the whole human race.

\ciceroSection{16} And so all in whom there is an aptitude for virtue are moved by them. Yet when the two Decii or the two Scipiones are called brave men, or when Fabricius or Aristides is named as just, it is not from any of these, as from a wise man, that an example of courage or of justice is sought. For none of them is wise in the sense in which we intend the word "wise man," nor were those who were counted and called wise — Marcus Cato and Gaius Laelius — truly wise, nor indeed those seven; but through the constant practice of intermediate duties they carried a certain likeness and appearance of the wise.

\ciceroSection{17} Therefore it is not permissible to weigh what is truly honorable against any competing claim of expediency; nor is that which we call honorable in the common usage — the quality cultivated by those who wish to be regarded as good men — ever to be set against advantage; and that honorable thing which falls within our ordinary understanding is to be kept and preserved by us just as steadfastly as that which is called honorable in the strict and true sense is to be kept by the wise. For no progress toward virtue can be sustained on any other footing. But this concerns those who are judged good men by their observance of duty.

\ciceroSection{18} Those, however, who measure everything by profit and advantage, and refuse to let honor outweigh these, are accustomed in their deliberations to compare what is honorable with what they take to be expedient; good men are not accustomed to do so. For my part, I believe that when Panaetius said men usually hesitate over this comparison, he meant precisely what he said: that they usually do — not that they ought to. In fact, not only to count what seems expedient as worth more than what is honorable, but even to compare the two and to be in any doubt between them, is the most shameful thing imaginable. What is it, then, that occasionally seems to occasion hesitation and to call for consideration? I believe it is when we are in doubt about the actual character of what we are considering.

\ciceroSection{19} For circumstances often bring it about that what is generally held to be disgraceful turns out not to be disgraceful. Take an example that is broad in its application: what greater crime can there be than to kill a man — to kill, moreover, someone close to you? And yet does a man incur the guilt of crime if he kills a tyrant, however intimate his connection with him? The Roman people, at any rate, does not think so — of all splendid deeds they count that the most splendid. Did expediency, then, defeat honor? No: honor followed hard upon expediency. The right rule to follow is therefore this: so that we may judge without error whenever what we call the expedient appears to conflict with what we recognize as honorable, we must establish a kind of formula; and if we hold to that formula in comparing things, we shall never depart from duty.

\ciceroSection{20} That formula will be most fully in accord with the reasoning and discipline of the Stoics. I follow them in these books because, although the older Academics and your Peripatetics — who were once one and the same with the Academics — do place what is honorable above what appears expedient, nevertheless those men argue the point more brilliantly for whom whatever is honorable is also expedient, and nothing expedient that is not honorable, than those for whom something honorable can fail to be expedient and something expedient can fail to be honorable. Our own Academy gives us wide latitude to defend whatever strikes us as most persuasive. But I return to the formula.

\ciceroSection{21} To strip something from another, then, and to enlarge one's own advantage at the cost of another person's loss — this is more contrary to nature than death, than poverty, than pain, than all the other things that can befall the body or external circumstances. For in the first place it destroys human fellowship and community. If we are so disposed that each man pillages or violates another for his own profit, then that fellowship of the human race which is most in accord with nature must inevitably be torn apart.

\ciceroSection{22} Just as, if each limb of the body had this understanding — that it could thrive by drawing the health of the nearest limb into itself — the whole body would necessarily weaken and perish, so, if each of us were to snatch the advantages of others to himself and strip away whatever he could from each for the sake of his own gain, human fellowship and community would inevitably be overturned. For while it is allowed, with nature making no objection, that each person should prefer to acquire for himself rather than for another the things that serve the needs of life, nature does not permit us to enlarge our own resources, wealth, and power by plundering what belongs to others.

\ciceroSection{23} And indeed this is established not only by nature — that is, by the law of nations — but also by the statutes of individual peoples, by which the commonwealth of each city is held together: that it is not lawful to harm another for one's own advantage. For this is the aim of the laws, this their intent: that the bond uniting citizens shall remain unbroken; those who tear it apart they restrain with death, exile, imprisonment, or fine. And nature's own reason — which is the law both divine and human — achieves this far more powerfully. Whoever is willing to obey it (and all who wish to live in accordance with nature will obey it) will never bring himself to covet what belongs to another or to take for himself what he has stripped from someone else.

\ciceroSection{24} For greatness and elevation of soul, and likewise affability, justice, and generosity, are far more in accord with nature than pleasure, than life itself, than riches — things which a great and lofty spirit will hold in contempt and count for nothing when it weighs them against the common advantage. To strip something from another for one's own gain is more contrary to nature than death, than pain, than all else of that kind.

\ciceroSection{25} Likewise it is more in accord with nature to take upon oneself the greatest toils and troubles, if they can be borne, for the sake of all peoples — preserving or aiding them — in imitation of that Hercules whom, in grateful memory of his benefactions, men's tradition placed in the council of the gods above, than to live in solitude not only free from all troubles but surrounded by pleasures in abundance and excelling even in beauty and strength. Therefore every man of the finest and most distinguished character places that life far ahead of this one by a long distance. From this it follows that a person who obeys nature cannot harm another person.

\ciceroSection{26} Then again, whoever wrongs another in order to gain some advantage for himself either thinks he is doing nothing contrary to nature, or considers death, poverty, pain, and even the loss of children, kinsmen, and friends more to be avoided than doing wrong to anyone. If he thinks that nothing contrary to nature is done when men are wronged, how can one argue with someone who strips all humanity out of the human being? But if he does think it is to be avoided, yet judges those other things far worse — death, poverty, pain — he is in error in counting any defect of body or fortune as more serious than defects of character. All people, therefore, should have one and the same goal before them: that the advantage of each and the advantage of all should be identical. If each person snatches that advantage to himself, all human fellowship will be dissolved.

\ciceroSection{27} And further: if nature prescribes that a human being should want what is good for another human being, whoever he may be, simply because he is a human being, then it necessarily follows from that same nature that the advantage of all is held in common. If that is so, we are all bound by one and the same law of nature; and if that in turn is so, we are certainly forbidden by nature's law to harm another. The first premise is true; therefore the conclusion is true as well.

\ciceroSection{28} For this, certainly, is absurd — what some people say: that they would take nothing from a parent or a brother for their own advantage, but that their relationship with the rest of their fellow citizens stands on a different footing. These people settle nothing of right, no fellowship of common interest, between themselves and their fellow citizens — a view that tears apart all the community of the state. Those who say the interests of citizens must be considered but deny the same of foreigners tear apart the common fellowship of the human race; and when that is removed, beneficence, generosity, goodness, and justice are utterly destroyed. Those who destroy these things must also be judged impious toward the immortal gods — for it is a fellowship established among human beings by the gods that they are overturning, and the closest bond of that fellowship is the conviction that it is more contrary to nature to deprive another person of something for one's own advantage than to endure every disadvantage, whether external, bodily, or even of the soul itself — provided those disadvantages are free from injustice; for justice alone is mistress and queen of all the virtues.

\ciceroSection{29} Perhaps someone will say: "Then will the wise man not, if he is himself dying of hunger, take food from another who is utterly useless?" Not at all; my life is not worth more to me than the disposition of soul that keeps me from wronging anyone for my own advantage. "What if a man of good character could, to keep himself from perishing in the cold, strip of his clothing Phalaris — that savage and monstrous tyrant — would he not do it?"

\ciceroSection{30} These cases are the easiest to judge. For if you strip something for your own advantage from a man who is of no use to anyone, you act inhumanly and contrary to nature's law. But if you are the kind of person who, should you remain alive, could bring great advantage to the commonwealth and to human fellowship, then if for that reason you take something from another, it is not to be blamed. In all other cases, each person should bear his own hardship rather than deprive another of his advantage. Disease or poverty or anything of that kind is therefore no more contrary to nature than the act of stripping and seizing what belongs to another; rather it is the abandonment of the common advantage that is contrary to nature — for that is an act of injustice.

\ciceroSection{31} And so nature's law itself, which preserves and upholds the advantage of human beings, will surely determine that the necessities of life be transferred from a man who is idle and useless to one who is wise, good, and brave — a man whose death would subtract greatly from the common advantage — but only on the condition that he who does this does not thereby take high regard for himself and love of self as a justification for wrongdoing. In this way he will always fulfil his duty, serving the advantage of human beings and of that human fellowship which I so often invoke.

\ciceroSection{32} As for the case of Phalaris, the judgment is perfectly straightforward. We have no fellowship with tyrants; on the contrary, there is the sharpest estrangement from them; and it is not contrary to nature to despoil a man whom it is honorable to kill. Indeed the whole of this pestilential and impious breed must be driven out from human community entirely. For just as certain limbs are amputated when they have begun to drain away blood and, as it were, vital spirit from the rest, and when they are doing harm to the other parts of the body, so that ferocity in the form of a man, that savagery of beast, must be cut away from what we might call the common body of humanity. Of this kind are all those questions in which duty must be sought out according to the circumstances of the moment.

\ciceroSection{33} I believe this was the ground Panaetius meant to cover, had not some accident or press of business cut short his design. The earlier books have already supplied rules enough for the kinds of question he addressed, from which it can be seen what is to be avoided on account of its baseness, and what need not be avoided because it is not base at all. But since I am now setting the capstone on a work that has been begun and is almost finished — as geometers are accustomed to demand that certain things be granted them, so as to explain what they want to explain more easily — I ask you, my son, to grant me, if you can, that nothing is to be sought for its own sake except what is honorable. If Cratippus will not permit that, you will at any rate allow me this much: that what is honorable is to be sought above all else for its own sake. Either position satisfies me, and at different times each strikes me as more probable; beyond those two nothing seems probable at all.

\ciceroSection{34} And first Panaetius must be defended on this point: he did not say that the expedient could sometimes stand against what is honorable — that would have been impiety for him — but that what seemed expedient could. He frequently declares that nothing is truly expedient that is not at the same time honorable, and nothing honorable that is not at the same time expedient; and he maintains that no plague has entered human life greater than the opinion of those who have pulled the two apart. What he introduced was not a real conflict but an apparent one — not so that we might sometimes prefer the expedient over the honorable, but so that, when such cases arose, we might judge between them without error. This remaining portion, then, I shall fill in on my own resources, without props — fighting, as they say, under my own standard. For nothing has been worked out on this part of the subject since Panaetius, at least nothing that has come into my hands and won my approval.

\ciceroSection{35} When, then, some semblance of the expedient is set before us, we cannot help being moved. But if, on close attention, you see that baseness is bound up with whatever has offered the appearance of advantage, then it is not the advantage that must be abandoned but the understanding that must be reached: wherever there is baseness, there can be no advantage. And if nothing runs so contrary to nature as baseness — for nature demands what is upright, consistent, and harmonious, and rejects the contrary — and if nothing is so fully in accord with nature as advantage, then surely in one and the same thing advantage and baseness cannot coexist. Moreover, if we are born for what is honorable, and honorable conduct is either the only thing to be sought, as Zeno held, or at any rate to be weighed heavier than everything else, as Aristotle believed, then it follows that what is honorable is either the only good or the supreme good; and what is good is surely advantageous; and so whatever is honorable is advantageous.

\ciceroSection{36} Hence the error of dishonest men: when they seize upon something that has seemed expedient, they at once separate it from what is honorable. From this source spring daggers, poisons, forged wills; from this source thefts, embezzlement, plunder and spoliation of allies and fellow citizens; from this source the greed for excessive wealth, for intolerable power, and finally, even in free states, the desire to reign — desires than which nothing fouler or more repellent can be conceived. For such men see, with their false judgment, the material gains; they do not see the punishment — I mean not that of the laws, which they often break through, but the punishment of baseness itself, which is the most bitter of all.

\ciceroSection{37} That breed of deliberators, therefore, must be driven from our midst entirely — for it is criminal through and through — the kind who deliberate whether to follow what they see to be honorable, or to soil themselves with crime in full knowledge of what they do. In the very act of doubting, the crime is already present, even if they never carry it out. Accordingly, matters should never come to deliberation at all when deliberating itself would be base. And more: from every deliberation the expectation and hope of concealment must be cleared away. For if we have made any progress in philosophy, we ought to be thoroughly persuaded that even if we could hide what we do from all the gods and from all mankind, we must still commit nothing through greed, nothing through injustice, nothing through lust, nothing through want of self-command.

\ciceroSection{38} It is with this in mind that Plato introduces the story of Gyges. When the earth had opened after heavy rains, Gyges descended into the chasm and saw there, as the story goes, a bronze horse, with doors in its flanks; these he opened and found the body of a man of extraordinary size, with a gold ring on his finger. He drew it off and put it on himself — he was a shepherd belonging to the king — and then made his way to the assembly of shepherds. There, whenever he turned the bezel of the ring toward his palm, he became invisible to the rest, while he saw everything; when he turned it back to its place, he was visible again. Using this opportunity the ring afforded, he violated the queen, with her help killed the king his master, did away with those he thought stood in his way, and in none of these crimes could anyone see him. So, suddenly, by the favor of the ring, he was king of Lydia. If the wise man had that very ring, he would think himself no more permitted to do wrong than if he had it not; for what honorable men seek is honorable conduct, not concealment.

\ciceroSection{39} At this point certain philosophers — by no means bad men, but insufficiently sharp — say that the story Plato offers is a fiction and a fabrication. As though Plato were maintaining that the thing happened, or even could have! The force of the ring and of the example is this: if no one were going to know, no one so much as to suspect, when you had done something for the sake of wealth, power, domination, or desire — if it were going to remain forever unknown to gods and men alike — would you do it? They say that is impossible. Quite so; it cannot be done. But I am asking: what would they do if it could be done, given that they deny it can? They press their objection rather crudely; they say it cannot be done and stick to that point, unable to see the force of the word. For when we ask what they would do if they could hide it, we are not asking whether they can hide it, but we apply a kind of pressure, as it were, so that if they answer they would act on what profits them when impunity is on offer, they confess themselves criminals; if they deny it, they concede that everything base is to be avoided for its own sake. But let us return to our subject.

\ciceroSection{40} Many situations arise that trouble minds with the appearance of the expedient — not the question whether honorable conduct is to be abandoned for the greatness of some gain (that would be wicked), but the question whether something that appears expedient can be done without baseness. When Brutus stripped his colleague Collatinus of his command, it might have seemed unjust; for Collatinus had been a partner in Brutus's counsels and a helper in the expulsion of the kings. But when the leading men had resolved that the name of Tarquin, the kinship with Superbus, and all memory of kingship must be extirpated, what was expedient — to consult the good of the fatherland — was at the same time so honorable that it ought to have been welcome to Collatinus himself. And so expediency prevailed on account of its honor; without that honor it could not have been expediency at all. But with the king who founded the city it was otherwise: there the appearance of the expedient misled his mind;

\ciceroSection{41} for when it seemed to him more expedient to reign alone than to reign with another, he killed his brother. He abandoned both fraternal love and common humanity to get what seemed expedient but was not; and yet he offered the wall as his pretext, a show of honor that was neither convincing nor, frankly, adequate. He sinned, therefore — with all respect, be it said, to Quirinus or Romulus.

\ciceroSection{42} And yet we are not to abandon our own interests and hand them over to others when we need them ourselves; each person must attend to his own advantage in whatever can be done without injury to another. Chrysippus put it neatly, as he did so many things: the man who runs in the stadium, he said, must put forth every effort and strain to win, but he has no business tripping up his competitor or pushing him aside with his hand; so in life it is not unjust for each person to seek what falls to his own use, but to wrest it from another is not right.

\ciceroSection{43} It is in friendships, however, that the demands of duty are most disturbed: both to withhold from a friend what you may rightly give, and to give what is not right, are contrary to duty. But on this whole topic the rule is brief and not difficult. What seem to be advantages — offices, riches, pleasures, and the rest of that kind — these are never to be preferred to friendship. On the other hand, a good man will not act against the commonwealth, or against an oath or sworn faith, for a friend's sake — not even if he sits as judge on the friend's very case; for he lays aside the character of friend when he puts on that of judge. He will grant this much to friendship: that he may prefer his friend's cause to be a just one, and arrange the time for pleading the case, so far as the law allows.

\ciceroSection{44} But when he must deliver a verdict on oath, he will remember that he calls a god to witness — that is, in my judgment, his own mind, than which the god himself has given man nothing more divine. And so we have received from our ancestors the fine custom of asking a judge, if he has one in his court, what he can do consistently with his sworn obligation. That question reaches to what I said a moment ago: what a judge may properly grant a friend. If everything a friend wanted had to be done, such relationships ought to be called not friendships but conspiracies.

\ciceroSection{45} I speak, however, of ordinary friendships; among truly wise and perfect men there can be nothing of that kind. They tell of the Pythagoreans Damon and Phintias, who were so disposed toward each other that when the tyrant Dionysius had fixed a day of death for one of them, and that man asked for a few days to provide for his family, the other stood surety for his friend's appearance, on the condition that if the first did not return, the surety must die in his place. When the condemned man came back on the appointed day, the tyrant, in admiration of their loyalty, asked to be enrolled as a third in their friendship.

\ciceroSection{46} When, therefore, what seems expedient in friendship is weighed against what is honorable, let the appearance of advantage fall, let honor prevail. When things that are not honorable are demanded in friendship, let scruple and good faith take precedence over friendship. In this way we shall arrive at the proper judgment of duty that we are seeking. In affairs of state, however, men err by the appearance of expediency all too often — as in the destruction of Corinth by our own countrymen; and still more harshly did the Athenians, who voted that the thumbs of the Aeginetans, strong as they were at sea, should be cut off. This seemed expedient: Aegina, from its nearness, threatened Piraeus too dangerously. But nothing cruel is expedient; cruelty is in the highest degree at war with human nature, which we are bound to follow.

\ciceroSection{47} They also do wrong who bar foreigners from using the cities and expel them — as Pennus did in the time of our fathers, and Papius recently. For it is right not to allow someone who is not a citizen to pass for one; the consuls Crassus and Scaevola, men of the highest wisdom, carried a law to that effect. But to bar foreigners from the use of the city altogether is quite inhuman. There are splendid examples of cases where the appearance of public advantage is despised for the sake of honor. Our commonwealth furnishes many such, none more so than during the Second Punic War; after the disaster at Cannae our ancestors showed more spirit than ever in times of success — not a sign of fear, not a word of peace. So great is the force of what is honorable that it obscures the very appearance of advantage.

\ciceroSection{48} When the Athenians could no way withstand the Persian onslaught and resolved to abandon the city, leaving their wives and children at Troezen, embark their fleet, and defend the freedom of Greece by sea, a certain Cyrsilus who urged them to stay in the city and receive Xerxes was stoned to death. He seemed to be following the expedient; but there was no expediency there, with honor standing against it.

\ciceroSection{49} Themistocles, after the victory in that war against Persia, said in public assembly that he had a plan of service to the state, but that it need not be known; he asked only that the people appoint someone with whom he could share it. Aristides was given. Themistocles told him that the Spartan fleet, which lay beached at Gytheum, could be secretly burned — and that by this act the power of Sparta would necessarily be broken. When Aristides heard this, he came before the assembly to great expectation and said: the plan Themistocles proposed was exceedingly useful, but by no means honorable. The Athenians accordingly judged that what was not honorable was not useful either, and they rejected the whole scheme outright on Aristides' authority — a scheme they had not even heard. Better they than we, who grant pirates immunity and levy tribute on our allies. Let it stand, then: what is base is never useful — not even when you gain by it the very thing you thought useful; for this very thing, thinking it useful when it is base, is ruinous.

\ciceroSection{50} But cases arise, as I said above, in which expediency appears to conflict with what is honorable, so that we must consider whether the conflict is genuine or whether the two can be reconciled. Such questions take this form: suppose, for instance, a good man has shipped a large cargo of grain from Alexandria to Rhodes, where there is famine and scarcity and the price of corn is desperately high; suppose further that this same man knows several other merchants have put out from Alexandria and has seen their ships under sail with full cargoes heading for Rhodes. Will he tell the Rhodians, or hold his tongue and sell his grain as dearly as he can? We are imagining a man both good and wise; the question we are asking concerns the deliberation and judgment of a man who would not conceal the facts from the Rhodians if he held it to be base, but is uncertain whether it is base at all.

\ciceroSection{51} In cases of this kind the view of Diogenes of Babylon, a great and weighty Stoic, regularly differs from that of his pupil Antipater, a man of the keenest mind. Antipater holds that everything must be disclosed, so that the buyer knows nothing the seller knows and has not told him; Diogenes holds that the seller is obliged to declare whatever defects the civil law requires him to declare, and for the rest to proceed without sharp dealing, and since he is selling, to aim at the best price he can get. "I brought the grain, I put it on the market, I am selling my own goods at no higher a price than others — perhaps even lower, given the abundance arriving. Who is wronged?"

\ciceroSection{52} Antipater's line of reasoning rises from the other side: "What are you saying? You who ought to take thought for other men and serve human society, who were born under that law and hold those first principles of nature which you are bound to obey and follow — that your advantage is the common advantage and equally the common advantage is yours — will you conceal from men what abundance and supply are at their door?" Diogenes will perhaps reply as follows: "Concealing is one thing, staying silent another. I am not concealing anything from you right now when I fail to tell you what the nature of the gods is, or what the ultimate good is — knowledge that would benefit you more than information about the price of wheat; but it does not follow that whatever is useful for you to hear is thereby necessary for me to say."

\ciceroSection{53} "On the contrary," Antipater will say, "it is necessary, if you remember that there is a fellowship among men bound together by nature." "I remember it," Diogenes will reply; "but is that fellowship such that nothing belongs to any individual? If it is, then nothing should be sold at all — only given away." You see that throughout this entire debate the point at issue is never stated as: "This, though base, I will do because it is expedient" — but rather, on one side, as: "This is expedient and therefore not base"; and on the other: "This must not be done because it is base." A good man brings these two adversaries before us; let us now resolve the case.

\ciceroSection{54} A good man is selling a house. The house has certain defects which he knows and others do not: it is unhealthy but held to be sound; every bedroom shows snakes; the timber is poor, the structure ruinous — none of which anyone but the owner knows. I ask: if the seller does not disclose any of this to the buyers and sells the house for far more than he expected to get — has he acted unjustly or dishonestly?

\ciceroSection{55} He has, says Antipater. For what else is it than to fail to show a man who has lost his way the road? — an act consecrated at Athens with public curses. If this is not that, then it is something worse: to stand by and let a buyer rush headlong into catastrophe and into the gravest fraud through his own error. More than failing to show the road — it is deliberately leading another astray. Diogenes replies: "Did he compel you to buy? Did he even solicit you? He put on the market what he did not want; you bought what you did want. If those who advertise a villa as fine and well-built are not thought to have cheated, though it is neither fine nor soundly built, how much less has a man cheated who did not praise his house at all? Where the buyer's own judgment operates, what room is there for the seller's fraud? And if not everything one says is warranted, do you think everything one does not say is warranted too? What could be more foolish than for a seller to recite the defects of what he is selling? What more absurd than for an auctioneer, at the owner's instruction, to cry: 'I am selling a house that is unhealthy'?"

\ciceroSection{56} So in certain disputed cases one side defends what is honorable, the other argues expediency in such a way that what appears expedient is shown to be not only compatible with honor to do, but base to leave undone. This is that apparent conflict between the expedient and the honorable which so often seems to arise. These cases must be adjudicated; for we set them out not to leave them open but to resolve them.

\ciceroSection{57} Neither the grain-merchant, then, nor the house-seller ought, it seems to me, to have concealed anything from the buyers. Concealment is not simply keeping silent about something; it is the deliberate choice that those who have an interest in knowing something should remain ignorant of it, for the sake of your own gain. What kind of concealment is that, and what kind of man practices it? Everyone can see plainly enough: not an open man, not a straightforward man, not a candid man, not a just man, not a good man — a schemer rather, a shadow-dweller, a contriver, a deceiver, a cheat, a trickster, a worn old hand, a sly one. Is it not a bad bargain to earn all these names and more besides?

\ciceroSection{58} And if those who stayed silent deserve censure, what are we to think of those who resorted to outright falsehood in their words? Gaius Canius — a Roman knight, a man of some wit and decent learning — had made his way to Syracuse, as he used to say himself, for a holiday, not on business, and was letting it be known that he wanted to buy some small garden property where he could invite friends and amuse himself without being pestered. When this got about, one Pythius, a banker at Syracuse, told him he had no gardens for sale, but that Canius was welcome to use his as his own; and at the same time he invited the man to dinner in the gardens the following day. Canius accepted. Pythius, who as a banker was on good terms with all classes, called his fishermen together and asked them to fish in front of his gardens the next morning; he told them exactly what he wanted them to do. Canius came to dinner on time; Pythius had laid on a lavish spread; a whole flotilla of boats lay before their eyes; each fisherman brought up what he had caught and laid his fish at Pythius's feet.

\ciceroSection{59} Then Canius: "I ask you, Pythius — what is all this? So many fish, so many boats?" And Pythius: "What is surprising? This is where Syracuse keeps everything in the way of fish; this is where their fresh water is; this property is indispensable to those men." Canius was on fire with desire and pressed Pythius to sell. Pythius was reluctant at first. But in short — he got his way. The man, eager and rich, paid whatever Pythius asked, and bought the property furnished as it stood; he had the names entered in the accounts and the transaction completed. The next day Canius invited his friends; he arrived himself early on. He saw not a single thwart-pin. He asked a neighbor whether there was some fishermen's holiday that he could see none of them. "No holiday that I know of," the man said; "but no one fishes here as a rule — so I was puzzled yesterday about what had happened."

\ciceroSection{60} Canius was furious. But what could he do? My colleague and friend Gaius Aquilius had not yet introduced his formulas for fraud — the \textit{dolus malus} actions — though in those very formulas, when he was asked what \textit{dolus malus} was, he used to answer: when one thing is pretended and another done. A definition admirably put, as one would expect from a man skilled in making definitions. So Pythius, and all who do one thing while pretending another, are faithless, dishonest, and malicious. No act of theirs can be useful when it is stained with so many vices.

\ciceroSection{61} And if Aquilius's definition is sound, then all pretense and concealment must be banished from life entire. A good man will not pretend or conceal anything, whether to buy more cheaply or to sell more dearly. Moreover, this \textit{dolus malus} had already been made actionable by law — guardianship under the Twelve Tables, defrauding of young men under the Lex Plaetoria — and without any statutory basis in the courts, where the formula adds the words \textsc{ex fide bona}, "in good faith." Among the other formulas these phrases especially stand out: in the arbitration for recovery of a wife's dowry, \textsc{melius aequius}, "what is better and more equitable"; and in the action on fiduciary deposit, \textsc{ut inter bonos bene agier}, "that dealings be upright between good men." Can there be any trace of fraud in what is "better and more equitable"? Or when dealings are said to be "upright between good men," can anything be done by deceit or malice? Fraud, as Aquilius says, consists in pretense. All lying must therefore be expelled from dealings and contracts. The seller will not plant a shill to bid up the price; the buyer will not plant one to bid against himself. Each, if it comes to the point of a declaration, will make it once and once only.

\ciceroSection{62} Quintus Scaevola, son of Publius, once required that a piece of land of which he was the buyer be named to him at a single price, and when the seller had done so, said he valued it at more, and added a hundred thousand sesterces. There is no one who denies this was the act of a good man; the wise men deny it, on the grounds that he had sold for less than he could have got. This then is that ruinous division: some they count good men, others wise men. Hence Ennius: "Wisdom is wasted on the wise man who cannot profit himself." True enough — if only Ennius and I agreed on what profit means.

\ciceroSection{63} I find that Hecaton of Rhodes, the pupil of Panaetius, says in the books on duty he wrote for Quintus Tubero that the wise man's duty in managing his own estate is to do nothing against custom, law, or established practice. We do not want riches for ourselves alone, after all, but for our children, our relations, our friends, and above all for the state; for the resources and means of individuals are the wealth of the community. Scaevola's act, of which I spoke just now, can in no way satisfy this man; indeed Hecaton flatly says he will do nothing for his own profit that is not lawful. Such a man deserves neither great praise nor gratitude.

\ciceroSection{64} But if both pretense and concealment constitute \textit{dolus malus}, there are very few transactions in which that fraud does not play some part; or if a good man is one who helps whom he can and harms no one, then assuredly such a good man is not easily found. In sum: it is never useful to do wrong, because it is always base; and because it is always honorable to be a good man, it is always useful.

\ciceroSection{65} Now, on the subject of property sales, our civil law has a firm rule: any defects known to the seller must be declared. For under the Twelve Tables it was enough to guarantee what had been formally stated aloud — the seller who denied those statements faced a penalty of double the price — but the jurists went further and established a penalty for silence as well: they ruled that whatever defect existed in a property, if the seller knew of it and did not name it expressly, he was bound to make it good.

\ciceroSection{66} Consider what happened when the augurs were about to conduct an augury on the Citadel and ordered Tiberius Claudius Centumalus, who owned a building on the Caelian Hill, to demolish those parts of it whose height interfered with the taking of auspices. Claudius put the block up for sale and sold it; Publius Calpurnius Lanarius bought it. The augurs gave him the same notice. When Calpurnius had demolished the building and discovered that Claudius had put it up for sale only after being ordered by the augurs to demolish it, he brought him before an arbitrator with the formula: \textsc{whatever he ought by good faith to give or do for him}. The verdict was pronounced by Marcus Cato, father of the Cato who is ours — for as other men are named from their fathers, this one, who was the father of that great light, is rightly named from his son. The judge ruled that since the seller had known of the defect at the time of sale and had not disclosed it, he was bound to make good the buyer's loss.

\ciceroSection{67} He held, then, that it falls under good faith for the buyer to be informed of a defect the seller knows of. If that judgment was right, then the grain merchant I described was not acting rightly, nor was the seller of the plague-ridden house, in keeping silent. But silences of this kind cannot be reached by the civil law; those that can be reached, however, are firmly held. Marcus Marius Gratidianus, a kinsman of mine, had sold to Gaius Sergius Orata a house which he himself had bought from that same man only a few years before. The property was encumbered by a servitude, but Marius had not declared this in the conveyance. The matter was brought to court. Crassus appeared for Orata, Antonius for Gratidianus. Crassus pressed the law: that a seller who knowingly failed to declare a defect was bound to make it good. Antonius argued equity: since the servitude had been no secret to Sergius, who had originally sold the house, there had been no need to declare it, and a man could not be said to have been deceived who held what he had bought with full knowledge of its legal standing.

\ciceroSection{68} What is the point of all this? To show you that our ancestors had no patience for sharp dealing. But legal rules and philosophy do away with sharp dealing by different means: the law, so far as it can lay hold of it with its hand; philosophy, so far as reason and understanding can reach. Reason, then, demands this: no scheming, no dissembling, no deception. Does it count as setting a trap to put out bait — even if you do not intend to flush or drive the quarry? Wild animals themselves, with no one in pursuit, often fall into the net. So when you list a house for sale, put up a sign like a net, and sell a home riddled with defects — if some unwary buyer runs into it, have you set no trap?

\ciceroSection{69} I can see that, given the corruption of our habits, this is held neither shameful by custom nor prohibited by statute or civil law. Yet the law of nature forbids it. For human society — and though I have said this often, I must say it yet again — extends in its widest reach to all men without exception; within that comes the narrower fellowship of those who share a nation, and narrower still of those who share a city. Our ancestors accordingly chose to distinguish the law of nations from the civil law: what is civil law is not necessarily also the law of nations, but what is the law of nations ought also to be civil law. We, however, hold no solid and clearly stamped likeness of true law and genuine justice; we work with shadows and images. Would that we followed even those — for they are drawn from nature's finest and truest patterns.

\ciceroSection{70} Consider what force those words carry: \textsc{that I not be cheated or defrauded through you or your good faith!} And those golden ones: \textsc{that dealings between good men be conducted well and without fraud!} But who the good men are, and what it means to deal well — that is a great question. Quintus Scaevola, the pontifex maximus, used to say that the greatest force lay in all those arbitrations to which the phrase \textit{ex fide bona} — ``in accordance with good faith'' — was appended, and that the name of good faith spread itself across the widest ground: it ran, he said, through guardianships, partnerships, trusts, mandates, purchases, sales, lettings, hirings — all the dealings on which the life of society is built. In these cases, he held, it was the mark of a great judge to determine what each party owed the other, especially since the judgments in most of them could run in either direction.

\ciceroSection{71} Sharp dealing, then, must be rooted out — and that malice which wants to pass itself off as wisdom but stands as far from it as can be. For wisdom is seated in the discernment of good and evil; malice, if everything that is dishonorable is also evil, sets evil before good. And it is not only in the sale of land that the civil law, derived from nature, punishes malice and fraud: in the sale of slaves too all fraud on the seller's part is excluded. The seller who was bound to know about health, liability to flight, and past thefts is answerable under the aediles' edict. The case of heirs is different.

\ciceroSection{72} From this it follows — since nature is the source of law — that it accords with nature for no man to prey on another's ignorance. Nor can any greater ruin of human life be found than in malice that wears the mask of wisdom. It is from this that all those countless cases arise in which the expedient appears to war with what is honorable. How rare is the man who, if he knew he would go unpunished and that no one would ever know, could hold himself back from wrong?

\ciceroSection{73} Let us put it to the test, if you will — and in examples where common people may perhaps see nothing wrong. We need not deal here with assassins, poisoners, forgers, thieves, and embezzlers; they are to be worn down not with the arguments of philosophers but with chains and prison. Let us consider instead what men do who are held to be good. Certain persons brought a forged will from Greece to Rome in the name of Lucius Minucius Basilus, a man of great wealth. To make it easier to establish, they named as co-heirs Marcus Crassus and Quintus Hortensius, the most powerful men of that generation. These men, though they suspected the will was a forgery, were not themselves conscious of any wrongdoing; and so they did not refuse a small gift from another man's crime. Is that enough, then, for them to appear guiltless? In my judgment, no — though the one I loved while he lived, and the other I do not hate now that he is dead.

\ciceroSection{74} But when Basilus had wished his sister's son, Marcus Satrius, to bear his name and had made him his heir — the same Satrius who is patron of the Picene and Sabine territory; what a shameful mark on the age, that that name should survive in such hands! — it was not right that the leading men of the state should get the estate while nothing came to Satrius but the name. For if a man who fails to prevent or repel an injustice when he is able to is himself acting unjustly — as I argued in the first book — what must we think of one who not merely fails to resist an injustice but actually assists it? Even genuine inheritances, to my mind, are not honorable when they are sought through fawning flattery, through the performance of services not in truth but in pretense. And yet in cases of this kind it seems, at times, that expediency and honor pull in opposite directions.

\ciceroSection{75} That is an error. For the rule of expediency and the rule of honor are one and the same. Whoever fails to see this will find no fraud, no crime, beyond his reach. He will think: ``That course is no doubt honorable, but this one pays'' — and with that error he will dare to sever what nature has joined together, which is the very source of all fraud, all wrongdoing, all crime. Suppose a good man had the power that, if he snapped his fingers, his name would slip into the wills of wealthy men — he would not use that power, not even if he were certain no one would ever suspect it. But grant that power to Marcus Crassus, so that by a snap of the fingers he could be written in as heir where he was not truly heir — believe me, he would dance for joy in the Forum. The just man, however, the man we mean when we say a man is good, takes nothing from anyone to transfer to himself. Whoever marvels at this admits he does not know what a good man is.

\ciceroSection{76} But if anyone cares to unroll the conception of a good man that is folded up in his mind, he will come to understand for himself that the good man is one who helps those he can and harms no one except when provoked by wrong. What then? Does the man who contrives — by some poison, as it were — to displace the true heirs and step into their place, not harm anyone? Let him not do, some will say, what pays and profits him? No: let him understand that nothing is profitable, nothing expedient, that is unjust. Whoever has not learned this lesson will never be a good man.

\ciceroSection{77} As a boy I used to hear from Gaius Fimbria, a man of consular rank, that our father had once served as judge in the case of Marcus Lutatius Pinthia, a thoroughly honorable Roman knight, who had entered into a wager on the condition \textsc{that he was a man of good character}. Fimbria declared that he would never pronounce on that matter: otherwise he would either strip a man of established probity of his reputation by ruling against him, or appear to have ruled that someone was a man of good character — when in fact that status rests on an endless accumulation of duties and distinctions. This good man, then — recognized as such by Fimbria, and not only by Socrates — can plainly never suppose that anything is expedient which is not also honorable. And so such a man will dare not only to do nothing but even to think nothing that he would not be willing to proclaim aloud. Is it not shameful that philosophers should hesitate over questions which even plain country people do not hesitate over? From them has come that proverb, worn smooth by age: when they want to praise someone's reliability and goodness they say he is a man worth matching fingers with in the dark. What does that amount to, if not this: nothing is expedient that is not also fitting, even if you could gain it with no one there to refute you?

\ciceroSection{78} Do you see what this proverb implies — that no pardon can be given to that famous Gyges, nor to the man I was imagining a moment ago who could sweep in all the inheritances in the world with a snap of his fingers? For just as what is shameful can never become honorable however deeply concealed, so what is not honorable can never be made expedient when nature herself stands against it and fights back.

\ciceroSection{79} ``But,'' someone will say, ``when the rewards are very great, there is good cause to go wrong.'' Gaius Marius had long been far from any hope of the consulship and had been lying idle for seven years since his praetorship, with no prospect, it seemed, of ever standing for the consulship at all. Yet when Quintus Metellus, that finest of men and citizens, under whom he was serving as legate, sent him to Rome on dispatch, he went before the Roman people and brought false charges against Metellus, his own commander, claiming that Metellus was dragging the war out, and promising that if the people made him consul he would deliver Jugurtha, alive or dead, into the power of Rome within a short time. And so he was indeed made consul — but he departed from good faith and justice, in that he brought an excellent and weighty citizen, his own general, the man who had dispatched him, into popular hatred through a false accusation.

\ciceroSection{80} Nor did our kinsman Gratidianus discharge the duty of a good man at the time when he was praetor and the college of praetors had brought in the tribunes of the plebs to draw up a common policy on the currency — for in those days the coinage had become so unstable that no one could know what he was worth. Together they drafted a joint edict with its penalties and procedures, and agreed that they would all mount the Rostra together after noon to announce it. The others dispersed in various directions; Marius came straight from the benches to the Rostra and announced on his own what had been drawn up jointly. That action, if you want to know, brought him tremendous glory: statues in every neighborhood, incense and candles set before them — in short, no man was ever more beloved by the people.

\ciceroSection{81} These are the cases that sometimes unsettle the mind in deliberation, when the wrong that will be done to justice appears to be no great matter, while the advantage arising from it appears very great — as it seemed to Marius no very disgraceful thing to steal from his colleagues and the tribunes of the plebs a moment of popular favor, but an enormous advantage to be made consul by that means, which was his aim at the time. But there is one rule for all, and I want you to know it thoroughly: either let what seems expedient not be dishonorable, or, if it is dishonorable, let it not seem expedient. Can we then judge either that Marius, or this other man, a good man? Test and examine your own understanding, to see what is the image, the form, the idea of a good man that dwells in it. Does it fall to a good man to lie for his own gain, to slander, to steal a march, to deceive? Surely nothing could be further from it.

\ciceroSection{82} Is there any matter of such weight, or any advantage so much to be desired, that it is worth losing the luster and the name of a good man? What could expediency, so called, possibly bring us that is commensurate with what it takes away, if it has stripped us of the name of a good man and robbed us of good faith and justice? For what difference does it make whether a man transforms himself from a human being into a beast outright, or carries the inward savagery of a beast beneath the outward form of a man? And again: those who trample all that is right and honorable underfoot so long as they may gain power — are they not doing exactly what the man did who chose to have as his father-in-law the one by whose audacity he himself might be powerful? He thought it supremely expedient to have the greatest influence through another man's unpopularity; he did not see how unjust to his country and how dishonorable that was. And the father-in-law himself always had on his lips those Greek verses from the \textit{Phoenissae} — which I shall render as best I can, roughly perhaps, but clearly enough to convey the meaning: \textit{If wrong must be done, let it be done for a kingdom's sake; in all else, keep your piety.} A monstrous utterance — or rather it is Euripides who is monstrous, for making the one exception that is the most criminal thing of all.

\ciceroSection{83} Why then do we go collecting petty cases — inheritances, trading deals, fraudulent sales? Here before you stands a man who lusted to be king of the Roman people and master of all nations, and carried it through! If anyone calls such an ambition honorable, he is out of his mind; for he is endorsing the destruction of law and liberty and taking a hideous, abhorrent tyranny for glory. But if a man admits that it is not honorable to be king in a city that was free and should remain free, yet says that for the man who can manage it, it is expedient — with what rebuke, or with what reproach rather, should I attempt to tear him from so great an error? Can it be expedient, by the immortal gods, for anyone to commit the most abominable and atrocious parricide against his country, even if the man who has bound himself to that crime is hailed as ``father'' by the citizens he has crushed? Expediency, then, must be governed by honor — and so governed that these two are seen to differ in word, but in substance to sound as one.

\ciceroSection{84} In the common opinion I find no advantage that can be greater than being king; and yet when I begin to bring reason back to the truth, I find nothing less expedient for the man who has won it unjustly. For can anguish, anxiety, and fear by day and by night, a life beset by treachery and thick with peril — can any of this be expedient for anyone? ``Many are unjust and faithless to a king; few are well-disposed,'' says Accius. But to what king? The one who held his throne by right, as the heir of Tantalus and Pelops. How much more do you think those words apply to the king who, with the army of the Roman people, had crushed the Roman people themselves, and compelled a city that was not only free but mistress of nations to serve him?

\ciceroSection{85} What stains of conscience do you think he had in his soul, what wounds? Whose life, moreover, can be truly advantageous to him if the terms of that life are such that whoever destroys it will stand in the highest favor and glory? But if things that seem supremely advantageous are not so, because they are full of shame and dishonor, it ought by now to be thoroughly plain that nothing is expedient which is not honorable.

\ciceroSection{86} This principle has indeed been affirmed many times on other occasions, and in particular during the war with Pyrrhus, by the consul Gaius Fabricius in his second consulship and by our Senate. When King Pyrrhus had made war on the Roman people of his own accord, and the contest for supremacy was with a noble and powerful king, a deserter came from his camp to Fabricius and promised that, for a reward, he would return to Pyrrhus's camp as secretly as he had come, and kill him with poison. Fabricius saw to it that the man was taken back to Pyrrhus, and this act of his was commended by the Senate. Yet if it was the appearance of expediency and popular esteem we were after, that one deserter would have removed a great war and a dangerous opponent of Rome's supremacy; but it would have been great dishonor and infamy to have defeated by crime, not by courage, the man with whom we were competing for glory.

\ciceroSection{87} Which was the more expedient, then — for Fabricius, who was in this city what Aristides was in Athens, or for our Senate, which never severed expediency from worth — to contend with the enemy in arms or in poison? If command is to be sought for the sake of glory, let crime be absent, for there can be no glory in crime; if it is power itself that is sought by whatever means, it cannot be expedient with infamy attached to it. That resolution of Lucius Philippus, son of Quintus, was therefore not expedient either: that those states which Lucius Sulla had made free in exchange for money by a decree of the Senate should again become tributaries, and that we should not return the money they had paid for their freedom. The Senate assented. A shameful exercise of power! The good faith of pirates is better than the good faith of the Senate. ``But the revenues were increased — it is therefore expedient.'' How long will people dare to call anything expedient that is not honorable?

\ciceroSection{88} Can hatred and infamy be expedient for any empire, which ought to rest on the glory and goodwill of its allies? I have often disagreed even with my dear Cato: he seemed to me to take too rigid a line in defending the treasury and the revenues, in refusing everything to the tax-collectors, much to the allies — when we ought to have been generous to the latter and to have dealt with the former as we customarily deal with our own tenants. And so too Curio was wrong when he would say that the cause of the Transpadanes was just, but always added: ``Let expediency prevail!'' He should have shown instead that their cause was not just, because it was not in the Republic's interest — not confessed, while saying it was not expedient, that it was just.

\ciceroSection{89} The sixth book of Hecato's \textit{On Duties} is full of questions like these: is it the act of a good man, in a time of severe famine, not to feed his slaves? He argues both sides, but in the end directs duty more by what he takes to be expediency than by humanity. He asks: if a cargo must be thrown overboard at sea, should one sacrifice a valuable horse or a cheap little slave? One consideration pulls toward property, another toward humanity. If a fool has seized the one plank from a shipwreck, will the wise man wrest it from him by force if he can? Hecato says no, because it would be unjust. What, then — will the ship's owner take back what is his? Not at all; no more than he would wish to throw a passenger overboard into the deep because the ship is his. For until the ship has reached its destination, it does not belong to the owner but to the passengers.

\ciceroSection{90} What if there is but one plank, and two shipwrecked men, both of them wise — should each snatch it for himself, or should one yield it to the other? One should yield, but to the man whose life matters more — whether for his own sake or for the sake of the Republic. But if that consideration is equal on both sides? There will be no contest; one will give way to the other, as if by lot or by a throw of the hand. What if a father is plundering temples, driving tunnels to the treasury — should his son report this to the magistrates? That would be an abomination; he should even defend his father if accused. Does country, then, not take precedence over all our duties? Yes, but country itself is served by citizens who are devoted to their parents. But what if the father is attempting to seize a tyranny, or to betray his country — should the son say nothing? On the contrary, he will beg his father not to do it. If that does no good, he will reproach him, he will even threaten him; and in the last resort, if the matter is heading toward the ruin of his country, he will set his country's safety above his father's.

\ciceroSection{91} Hecato also asks: if a wise man has unknowingly accepted counterfeit coins as genuine, and discovers what they are, should he pass them on as genuine in payment of a debt? Diogenes says yes; Antipater says no, and I agree with the latter. The man who is selling wine that he knows is going bad — must he declare it? Diogenes thinks it unnecessary; Antipater judges that a good man must. These are, as it were, the disputed points of Stoic jurisprudence. In selling a slave, must his faults be declared — not those which, if you fail to disclose them, give the buyer legal recourse for recovery, but such things as: the slave is a liar, a gambler, a thief, a drunkard? One side thinks they must be declared, the other does not.

\ciceroSection{92} If a man selling gold thinks he is selling copper, should a good man tell him it is gold, or buy for a denarius what is worth a thousand? By now it is plain enough both what my own view is and what the dispute is between the philosophers I have named. Must agreements and promises always be kept — those, that is, made without force or fraud, as the praetor's formula has it? If a man has given another a certain medicine against dropsy, and stipulated that, once cured, the man shall never use that medicine again; and the man recovers, but some years later falls into the same disease, and the one with whom he made the agreement will not allow him to use it again — what is to be done? The man is inhuman in refusing, and no wrong is done to him by using it; life and health must be consulted.

\ciceroSection{93} What if a wise man has been asked by someone who means to make him his heir — and ten million sesterces will be left to him by the will — to dance openly in the Forum in broad daylight before entering upon the inheritance, and has promised to do so, since the other would not have named him heir otherwise — should he do what he promised, or not? He ought not to have promised; that, I judge, would have been the dignified course. But since he has promised: if he thinks it shameful to dance in the Forum, he will act more honorably by going back on his word and taking nothing from the inheritance than by taking it; unless he were to put that money to some great crisis of the Republic, so that even dancing, in the service of his country, might not be shameful.

\ciceroSection{94} Those promises too must not be kept which are of no advantage to the very persons to whom they were made. To return to myth: the Sun promised his son Phaëthon that he would do whatever he wished; the boy wished to be lifted into his father's chariot. He was lifted into it — and before he could stand firm, he was struck by a thunderbolt and consumed in flames. How much better it would have been in this case if the father's promise had not been kept! And what of Theseus, who exacted his promise from Neptune? When Neptune had granted him three wishes, he prayed for the death of his son Hippolytus, whom he suspected on account of his stepmother; his prayer was granted, and Theseus fell into the deepest grief.

\ciceroSection{95} And Agamemnon, who had vowed to Diana the most beautiful thing born in his kingdom in that year, and sacrificed Iphigenia — than whom nothing more beautiful had been born that year? The promise should have gone unkept rather than so monstrous a crime been committed. Promises, then, are sometimes not to be kept; nor are deposits always to be restored. If a man has deposited a sword with you in his right mind, and demands it back in a fit of madness, to return it would be wrong; it is your duty not to return it. What if the man who deposited money with you is making war on your country — are you to restore the deposit? I think not; you would be acting against the Republic, which ought to be your dearest obligation. Thus many things that seem honorable by nature become, when circumstances change, not honorable: to keep a promise, to stand by an agreement, to restore a deposit — all these, when advantage shifts, cease to be honorable. This, I think, is enough to have said about the apparent advantages that run contrary to justice under the guise of wisdom.

\ciceroSection{96} But since in the first book we derived our duties from the four springs of what is honorable, let us remain within those same springs as we show how hostile to virtue are things that merely seem advantageous but are not. Of wisdom, which wickedness tries to counterfeit, and likewise of justice, which is always expedient, we have already treated. The remaining two parts of what is honorable are these: one is perceived in the greatness and pre-eminence of a high-minded soul; the other in the shaping and governance of self-restraint and temperance.

\ciceroSection{97} It seemed expedient for Ulysses — or so the tragic poets would have us believe; for in Homer, the best of all authorities, there is no such suspicion about him — but the tragedies charge him with wanting to evade military service by feigning madness. Not an honorable plan; yet expedient, someone might perhaps say: to reign and live at ease in Ithaca with his parents, his wife, his son. Do you suppose any distinction earned through daily labors and dangers is to be compared with such tranquility? I for my part hold it entirely contemptible and to be cast aside, since what is not honorable I judge to be not even expedient.

\ciceroSection{98} What do you suppose Ulysses would have heard, had he persisted in that pretense? He who accomplished the greatest deeds in the war is still made to hear this from Ajax: \textit{"He himself was first to take the oath that all of you well know, yet he alone despised the keeping of his pledge; he set about feigning madness, that he might not march to war. And had not Palamedes, with keen discernment, seen through the crafty audacity of this man's scheme, he would have gone on cheating, in perpetuity, the sacred bond of an oath."}

\ciceroSection{99} For him it was surely better to fight not only against the enemy but against the very waves — as he did — than to abandon a Greece united in arms against a barbarian foe. But let us set myths and foreign examples aside and come to what was done, and done by our own people. Marcus Atilius Regulus, serving his second consulship in Africa, was taken prisoner by ambush under the Spartan commander Xanthippus, with Hamilcar — father of Hannibal — as the Carthaginian general in command. He was sent to the Senate under oath: unless certain noble prisoners were returned to the Carthaginians, he was to come back to Carthage himself. When he arrived at Rome, he saw what appeared to be the expedient course — but judged it, as the event makes plain, to be false. That course was this: to remain in his homeland, to stay with his wife and children in his own house, to reckon the disaster he had suffered in war as the common fortune of arms, and to hold the rank of a former consul. Who denies these things are expedient? Who, do you think? Greatness of soul denies it, and courage.

\ciceroSection{100} Do you want weightier authorities? These are the virtues proper to those who fear nothing, look down on all things human, count nothing that can befall a man as intolerable. So what did he do? He came before the Senate, reported his instructions, declined to give his vote — so long as he was bound by the enemy's oath, he said, he was not a senator. And further — "O what a fool!" someone might cry, "fighting against his own interest!" — he declared it was not expedient to return the prisoners: they were young men and good commanders, while he was already spent with age. When his counsel prevailed, the prisoners were kept; he himself returned to Carthage. Neither love of country held him back, nor love of those who were his. Nor was he unaware that he was setting out toward a most savage enemy and toward torture elaborately devised; but he judged that the oath he had sworn must be kept. And so, when he was being killed by sleep-deprivation, his cause was better than it would have been had he remained at home — an old man, a prisoner, a perjurer, a former consul.

\ciceroSection{101} "But it was foolish of him: not only to have declined to vote for returning the prisoners, but even to have spoken against it." Foolish — in what way? Even if it served the commonwealth? Can anything that is of no use to the state be useful to any citizen at all? Men overturn the very foundations of nature when they divorce the expedient from the honorable. For we all seek what is expedient and are swept toward it, and there is no other way for us to be; for who is there who flees what is expedient — or rather, who is there who does not pursue it with the greatest eagerness? But because nowhere can we find the expedient except in what is praiseworthy, seemly, and honorable, we therefore hold those things first and supreme, and regard the name of expediency as less splendid than necessary.

\ciceroSection{102} "What, then," someone may ask, "is the force of an oath? Are we afraid of an angry Jupiter?" Now on this point all philosophers agree — not only those who hold that the god himself has no business of his own and takes no trouble for others, but even those who hold that the god is always active and at work — that the god is never angry and never harms. What more could an angry Jupiter have done to Regulus than Regulus did to himself? The sanction of religion, then, had no power to overturn so great an apparent advantage? "Or was it to avoid acting dishonorably?" — first, we should choose the lesser of evils. Was that disgrace so great an evil as that torture? And then there is that line in Accius: \textit{"Have you broken faith?"} — \textit{"I have given faith to no faithless man, nor do I give it now"} — though it is spoken by an impious king, it is finely said.

\ciceroSection{103} They add this further argument: just as we say that some things appear expedient but are not, so they say that some things appear honorable but are not — as this very thing appears honorable, to have returned to torture for the sake of keeping an oath; but it is not in fact honorable, because what was done under the enemy's compulsion ought never to have been held binding. They add also that whatever is greatly expedient thereby becomes honorable, even if it did not appear so before. These are roughly the arguments against Regulus. But let us consider them one by one.

\ciceroSection{104} Jupiter was not to be feared as one who in his anger would do harm — for it is not his nature to be angry, nor to harm. This argument has no more force against Regulus than against every oath whatsoever. But in an oath what we should understand is not who may be feared, but what the force of the thing is. An oath is a solemn affirmation made with religious sanction; and what you have promised with such affirmation, as it were with the god as witness, must be kept. At stake is no longer the anger of the gods — which does not exist — but justice and good faith. For Ennius says it finely: \textit{"O nurturing Faith, winged and apt, and sworn pledge of Jupiter!"} Whoever violates an oath, then, violates Faith — whom our ancestors wished to dwell, as Cato's speech records, on the Capitol, neighbor to Jupiter Greatest and Best.

\ciceroSection{105} "But even an angry Jupiter could not have harmed Regulus more than Regulus harmed himself." Certainly — if pain is the only evil. But that pain is not the highest evil — not even an evil at all — is affirmed with the greatest authority by philosophers. For that authority do not, I beg you, reproach Regulus — not a slight witness, but perhaps the weightiest of all. What worthier witness shall we seek than the foremost man of the Roman people, who endured torture willingly for the sake of discharging his duty? Now as for their argument — "choose the lesser evil," meaning disgrace rather than calamity — is there any evil greater than disgrace? If disgrace in the body's deformity has some power to offend, how great must the corruption and foulness of a dishonored soul appear!

\ciceroSection{106} And so those who reason more rigorously dare say that the only evil is what is disgraceful; those who reason more loosely do not hesitate to call it the highest evil. As for that line: \textit{"I have given faith to no faithless man, nor do I give it now"} — the poet is right to put it thus, because when Atreus was the character being played, the part had to be served. But if they take from this the principle that no pledge need be kept with a faithless man, let them beware they are not finding a hiding-place for perjury.

\ciceroSection{107} There is, moreover, a law of war, and the keeping of an oath pledged to an enemy is frequently required. For what is sworn with the sincere intention that it ought to be performed must be kept; what is sworn otherwise — if it is not performed, there is no perjury in that. For instance: if you have agreed to pay a price for your life to pirates and have not paid it, there is no fraud, not even if you swore an oath — for a pirate is not a recognized member of the class of public enemies but the common enemy of all; with such a man there ought to be no bond of faith and no shared oath.

\ciceroSection{108} For to swear falsely is not in itself perjury; perjury is to fail to do what you have sworn, as our formula puts it, \textit{from the honest conviction of your mind}. Euripides puts it shrewdly: \textit{"My tongue has sworn; my mind remains unsworn."} \textit{[Greek: hē glōssa omōmoch', hē de phrēn anōmotos.]} Regulus, however, had no business disturbing the terms and agreements of a war by perjury. For the war was being waged against a just and lawful enemy, against whom the entire law of heralds and many shared legal customs apply. Were it otherwise, the Senate would never have surrendered distinguished men in chains to the enemy.

\ciceroSection{109} Consider: Titus Veturius and Spurius Postumius, serving their second consulship, had made peace with the Samnites after our legions had fought badly at the Caudine Forks and been sent under the yoke. They were surrendered to the Samnites — because they had acted without the authorization of the people and the Senate. And at the same time Tiberius Numicius and Quintus Maelius, who were then tribunes of the people and on whose authority the peace had been made, were surrendered so that the peace with the Samnites might be repudiated. Postumius himself, the very man being surrendered, was the advocate and author of his own surrender. Many years later the same course was taken by Gaius Mancinus, who urged passage of the bill that Lucius Furius and Sextus Atilius were carrying by senatorial decree — that he be surrendered to the people of Numantia, with whom he had made a treaty without the Senate's authorization; the bill was passed and he was surrendered to the enemy. More honorable this than the conduct of Quintus Pompeius, who was in the same position, yet pleaded against it and the law was not passed. With Pompeius, what appeared to be the expedient prevailed over what was honorable; with the others, the false appearance of expediency was overcome by the authority of what is honorable.

\ciceroSection{110} "But it ought not to have been binding — it was done under compulsion." As though compulsion can really be applied to a brave man. Why, then, did he go to the Senate at all, especially since he was going to argue against returning the prisoners? The very thing that is greatest in what he did is what you find fault with. He did not stand by his own private judgment: he took up the cause so that the judgment would rest with the Senate; and if he had not been himself the advocate of that course, the prisoners would certainly have been returned to the Carthaginians, and so Regulus would have stayed safe at home in his own country. Because he judged that course not expedient for the state, he believed it honorable both to hold that opinion and to suffer the consequence. And as for their claim that what is greatly expedient thereby becomes honorable — rather, it \textit{is} honorable, not \textit{becomes} so. For nothing is expedient that is not at the same time honorable, and the reason a thing is honorable is not that it is expedient, but the reason it is expedient is that it is honorable. Of all the admirable examples, therefore, it would not be easy to name one more praiseworthy or more distinguished than this.

\ciceroSection{111} Yet in all the glory that surrounds Regulus, there is one thing pre-eminently worthy of admiration: his voting that the prisoners be kept. As for his return — to us now it seems remarkable; but in those times he could not have acted otherwise. That praise accordingly belongs not to the man but to the age. For our ancestors wished there to be no stronger bond for keeping faith than the oath. The laws of the Twelve Tables bear witness to this; the sacred laws bear witness; the treaties bear witness, by which faith is bound even with an enemy; the findings and censures of the censors bear witness — and no subject did they judge with more exacting care than the keeping of an oath.

\ciceroSection{112} When Lucius Manlius, son of Aulus, had been dictator, Marcus Pomponius, a tribune of the people, indicted him on the ground that he had added several days beyond his term to the dictatorship; he also charged him with having banished his son Titus — the one who was later called Torquatus — from human society and ordered him to live in the country. When the young son heard that his father was being harassed, he is said to have rushed to Rome and arrived at Pomponius's house at first light. When it was announced to him, Pomponius supposed the young man was furious and bringing something to lay against his father, and rose from his bed; he had the witnesses removed and bade the young man come in to him alone. But the moment he stepped inside, the son drew his sword and swore on the spot that he would kill him then and there unless Pomponius swore an oath to drop the case against his father. Pomponius, compelled by this terror, swore the oath; he reported the matter to the people, explained why he was obliged to abandon the case, and released Manlius. Such was the force of an oath in those times. And this Titus Manlius is the same man who, at the river Anio, killed the Gaul who had challenged him to single combat, stripped the torque from him, and so found his surname — a very great man indeed; and he who was so indulgent to his father was equally and harshly severe toward his own son.

\ciceroSection{113} But while Regulus deserves praise for keeping faith with his oath, those ten men whom Hannibal, after the battle of Cannae, sent to the Senate under oath to return to the camp in which the Carthaginians held power, should they fail to win the release of the prisoners — those ten, if they did not return, deserve censure. Yet not all of them are judged in the same way. Polybius, a thoroughly reliable authority, records that of the ten noblemen sent at that time, nine returned when the Senate refused their petition; but one of the ten, who had left the camp and then gone back a little while later as though he had forgotten something, remained in Rome — for he supposed that by returning briefly to the camp he had discharged himself from the oath. He supposed wrongly; for deceit tightens the bonds of perjury, it does not loosen them. It was, then, a piece of foolish cunning, a perverse counterfeit of wisdom. And so the Senate decreed that this crafty schemer be put in chains and returned to Hannibal.

\ciceroSection{114} But here is the most telling point of all: Hannibal held eight thousand men — not men taken in the field of battle, nor men who had fled in fear of death, but men left behind in the camp by the consuls Paullus and Varro. The Senate resolved that these should not be ransomed, even though it could have been done at small expense, in order that the lesson be bred into our soldiers: conquer or die. When Hannibal heard this, the same author records that his spirit broke — because even in ruin the Senate and people of Rome had shown such loftiness of soul. So it is that when you set what is honorable in the scales, those things that appear expedient are overcome.

\ciceroSection{115} Gaius Acilius, who wrote a history in Greek, says that there were more who returned to the camp by the same subterfuge, to discharge themselves from the oath — and that the censors marked them with every form of disgrace. Let this be the end of the present topic. For it is plain enough that what is done in a spirit of cowardice, baseness, abjection, and broken will — as Regulus's conduct would have been, had he voted on the prisoners according to his own convenience rather than the public good, or had he chosen to remain at home — is not expedient, precisely because it is shameful, foul, and base.

\ciceroSection{116} There remains the fourth part, which is contained in propriety, moderation, self-restraint, self-command, and temperance. Can anything expedient, then, be opposed to this chorus of such virtues? And yet the Cyrenaics and those named Annicereans, following Aristippus, placed all good in pleasure and held that virtue deserves praise only because it is the producer of pleasure. After these men faded, Epicurus flourished — the champion and co-author, one might say, of much the same position. Against these men we must fight, as the saying goes, with all our horse and foot, if our purpose is to hold and defend what is honorable.

\ciceroSection{117} For if not only expediency but the whole of the happy life rests, as Metrodorus wrote, on a sound constitution of the body and reliable hope of maintaining it, then this expediency — and indeed the highest expediency, as they reckon it — will be at war with what is honorable. Where, first of all, does wisdom find its place? In scouring the world for pleasures? What a wretched servitude that is — virtue waiting upon pleasure. And what is the office of wisdom, then? Selecting pleasures with discernment? Suppose nothing were more agreeable than that; what could be imagined more base? Again: for a man who calls pain the supreme evil, what room is there for courage — which is the contempt of pains and toils? However boldly Epicurus may speak of pain in many passages, as he does, we must look not at what he says but at what is consistent for him to say — a man who has defined good by pleasure and evil by pain. And if I listen to him on self-restraint and temperance, he says much in many places, but, as they put it, the water sticks: for how can a man praise temperance who places the highest good in pleasure? Temperance is the enemy of appetites; appetites, however, are the pursuers of pleasure.

\ciceroSection{118} In these three areas, all the same, they maneuver not without some dexterity, making what shifts they can. They bring in wisdom as the knowledge of how to supply pleasures and drive away pains; they manage courage too after a fashion, by teaching a way of thought that disregards death and endures pain; they even introduce temperance — with no great ease, to be sure, but in whatever way they can — arguing that the magnitude of pleasure is determined by the subtraction of pain. Justice, however, staggers, or rather falls flat entirely, as do all the virtues concerned with community and human fellowship. For goodness, generosity, and courtesy can no more exist than friendship can, if these things are sought not for their own sake but referred to pleasure or advantage. Let us therefore gather our argument into brief compass.

\ciceroSection{119} For just as we have shown that no expediency exists that is contrary to what is honorable, so we say that every pleasure is contrary to what is honorable. The more, then, do I find fault with Calliphon and Dinomachos, who thought they would resolve the controversy by yoking pleasure to what is honorable — as if they were coupling a beast with a human being. What is honorable does not admit that union; it spurns it, it repels it. Nor indeed can the end of good and evil — which must be single and simple — be compounded and blended out of things so utterly unlike each other. But on this question — it is a large one — I shall say more elsewhere; for now, let us return to our subject.

\ciceroSection{120} The question, then, of how to decide when what appears expedient conflicts with what is honorable has been argued sufficiently above. But if pleasure too is to be said to carry the appearance of expediency, there can be no junction between it and what is honorable. For granted that we allow pleasure some standing — perhaps as a kind of seasoning — it will have nothing of expediency, nothing of substance, about it at all.

\ciceroSection{121} You have from your father a gift, my son Marcus — in my own judgment a great one, but its worth will depend on how you receive it. These three books must be welcomed among the lectures of Cratippus as visitors received into a house; but just as, had I come to Athens in person — and I should have done so, had not my country called me back from mid-voyage with a voice too clear to refuse — you would sometimes have heard me as well, so now, since it is my voice that has made its way to you in these volumes, give them as much of your time as you can — and you will be able to give as much as you wish. When I have come to see that this kind of learning gives you joy, I shall speak with you both in person before long, as I hope, and in absence through these pages. And so farewell, my son; be persuaded that you are indeed dearest to me — but will be far dearer still if you take delight in these counsels and instructions.

