% Academica (academica) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroSection{1} Lately at my place near Cumae, while my friend Atticus was with me, word was brought to us by M. Varro that he had arrived from Rome the evening before, and that, had he not been worn out by the journey, he would have come straight on to us. When we heard this, we judged that no delay should be allowed to keep us from seeing a man bound to us both by the same pursuits and by an old friendship; and so we set out at once to go to him. We were only a little way from his villa when we saw the man himself coming toward us; and embracing him, as is the way of friends — for it had been a long enough interval … — we brought him back to his own villa.

\ciceroSection{2} Here a few words at first, and those while we were asking whether there happened to be any news from Rome. Then Atticus said, 'Drop those matters, which we can neither inquire about nor hear of without distress, I beg you, and ask rather whether there is any news from the man himself. The Muses of Varro have been silent longer than they used to be; and yet I judge that he is not idling but concealing what he writes.' 'Not at all,' said he; 'for I count it the mark of an undisciplined man to write what he wishes kept hidden. But I have a great work in hand, one I began long ago; for it is to this very man here' — he meant me — 'that I have addressed certain pieces, which are indeed substantial, and are being polished by me to a finer finish.'

\ciceroSection{3} And I said, 'Those works of yours, Varro, I have long been awaiting, yet I do not dare demand them; for I have heard from our friend Libo, whose enthusiasm you know — nothing of that kind can be hidden from us — that you are not laying them aside but handling them with greater care and never putting them out of your hands. But this other thing it never came into my mind to ask of you before now. Yet now, after I have set about committing to writing the very studies I learned alongside you, and illuminating in Latin letters that ancient philosophy which sprang from Socrates, I ask what the reason is that, while you write much, you pass over this kind of writing — especially since you yourself excel in it, and since this pursuit, this whole subject, far outstrips all other pursuits and arts.'

\ciceroSection{4} Then he said: 'You ask of me a matter I have often deliberated and much turned over. And so I shall answer without hesitation, but shall say what is ready to hand for me, since, as I said, I have thought much and long about this very thing. For when I saw philosophy set out most carefully in Greek letters, I judged that any of our own men who were gripped by zeal for it would, if they were schooled in Greek learning, read the Greek rather than ours, and that if they shrank from the arts and disciplines of the Greeks, they would not care even for these, which cannot be understood without Greek learning. And so I was unwilling to write what neither the unlearned could understand nor the learned care to read.

\ciceroSection{5} You see these things yourself; for you have learned that we cannot be like Amafinius or Rabirius, who, with no method applied, dispute in common speech about matters set before the eyes, define nothing, divide nothing, frame nothing by fitting interrogation, and in the end think there is no art at all, neither of speaking nor of reasoning. We, on the other hand, obeying the precepts of the logicians and of the orators too — since our school holds each faculty to be a virtue — are compelled to use even new words, which the learned, as I said, will prefer to seek from the Greeks, while the unlearned will not accept them even from us, so that all the labour is undertaken in vain.

\ciceroSection{6} As for natural philosophy, then, if I approved of Epicurus, that is, of Democritus, I could write as plainly as Amafinius. For what is the great difficulty, once you have done away with the causes that produce things, in talking about the chance collision of little bodies — for so he calls the atoms? You know our natural philosophy: since it is held together by an active cause and by matter, the matter which that active cause shapes and forms, geometry too must be brought in; and with what words could anyone set this forth, or whom could anyone bring to understand it? These matters of life and conduct, and of things to be sought and shunned, those men treat simply — for they hold the good of a beast and of a man to be the same; but among our own you are not unaware how great and of what kind the subtlety is.

\ciceroSection{7} For if you follow Zeno, it is a great task to bring anyone to understand what that true and unmixed good is which cannot be severed from honour — a good which Epicurus says he cannot so much as conceive of, what its nature might be, apart from the pleasures that stir the senses; but if we are to pursue the Old Academy, which, as you know, I approve, how acutely it will have to be expounded by us, how subtly, how darkly even must we argue against the Stoics. That whole study of philosophy, then, I take upon myself, both for the steadiness of my life so far as I can, and for the delight of my mind; and I judge that no gift greater or better has been given to man by the gods, as it stands in Plato.

\ciceroSection{8} But my friends in whom the zeal is present I send to Greece — that is, I bid them go to the Greeks — so that they may draw from the springs rather than chase after the rivulets. As for the things that no one had yet taught, and from which the eager could find no source of knowledge, these, so far as I could — for I am no great admirer of anything of my own — I have made known to our countrymen; for they could not be sought from the Greeks, and, since the death of our friend L. Aelius, not even from the Latins. And yet in those old works of mine, which, imitating Menippus and not translating him, I sprinkled with a certain gaiety, much is mixed in from the depths of philosophy, much said in the manner of the logicians, which, that the less learned might more easily understand it, I made inviting to read by a kind of pleasantness; and in my eulogies, in these very prefaces to the antiquities, I wished to write after the manner of philosophy — if only I have succeeded.'

\ciceroSection{9} Then I said, 'There is something in that, Varro. For when we were strangers and wanderers in our own city, like guests, your books led us, as it were, home, so that at last we could recognize who and where we were. You it was who laid open the age of our country, the divisions of its eras, the laws of its sacred rites and of its priests, its discipline at home and in war, the seats of its regions and places, the names, kinds, functions, and causes of all things divine and human; you have brought a great light, too, to our poets and to Latin letters and words altogether, and you yourself have made a varied and elegant poem in nearly every measure, and you have begun philosophy in many places — enough to spur, too little to instruct.

\ciceroSection{10} You offer, indeed, a plausible reason: that either the learned will prefer to read the Greek, or that those who do not know the Greek will not even read these works of yours. But you do not really make it good to me; on the contrary, both those who cannot read the Greek will read these, and those who can read the Greek will not despise their own. For what reason is there that men schooled in Greek letters should read the Latin poets but not the philosophers? Is it because Ennius, Pacuvius, Accius, and many others give delight, men who rendered not the words but the force of the Greek poets — how much more will the philosophers delight, if, as those poets imitated Aeschylus, Sophocles, Euripides, so these imitate Plato, Aristotle, Theophrastus? Orators, certainly, I see are praised if any of our own have imitated Hyperides or Demosthenes.

\ciceroSection{11} For my part, Varro — for I shall speak as the matter stands — so long as ambition, so long as honours, so long as cases, so long as not only care for the commonwealth but a certain administration of it, held me entangled and bound fast by many duties, I kept these studies shut up within my mind, and renewed them by reading when I could, that they might not grow stale. But now, struck by the gravest wound of fortune, and freed from the management of the commonwealth, I seek from philosophy a medicine for my grief, and I judge this the most honourable diversion of my leisure. For either this is most fitting to my time of life, or it is above all in keeping with whatever deeds I have done worthy of praise, or there is nothing more useful for the educating of our fellow citizens, or, if these things are not so, I see nothing else that I could do.

\ciceroSection{12} Our friend Brutus, indeed, outstanding in every kind of praise, pursues philosophy in Latin letters so well that you would feel no want of the Greek on the same subjects; and he follows the very view that you do, for he heard Aristus at Athens for some time, whose brother Antiochus you heard. For which reason, give yourself, I beg, to this kind of writing too.'

\ciceroSection{13} Then he said: 'That, indeed, I shall consider — and not, to be sure, without you. But what is this I hear about you yourself?' 'About what?' I said. 'That you have abandoned the Old Academy,' he said, 'and now take up the New.' 'Why, then,' I said, 'should it have been more permissible for our friend Antiochus to move back from the new house into the old, than for us to move into the new from the old? For surely it is always the most recent things that are most corrected and amended. And yet Philo, the teacher of Antiochus — a great man, as you yourself reckon — used to deny in his books, as we also heard from his own lips face to face, that there were two Academies, and refuted the error of those who thought so.' 'It is as you say,' he said; 'but I do not suppose you are unaware of what Antiochus wrote against Philo.'

\ciceroSection{14} 'On the contrary,' I said, 'I should be glad to have both that and the whole Old Academy, from which I have been absent so long, renewed for me by you, if it is no trouble — and at the same time, if you please, let us sit down.' 'By all means as to that,' he said, 'for I am rather frail. But let us see whether Atticus too is content that I should do what I see you wish.' 'For my part, I am,' he said; 'for what could I more desire than to call back to memory the things I heard from Antiochus long ago, and at the same time to see whether they can be put into Latin with sufficient ease?' When this had been said, we all sat down within view of one another.

\ciceroSection{15} Then Varro began thus: 'Socrates, it seems to me — and this is agreed among all — was the first to call philosophy away from things hidden and wrapped up in nature herself, with which all the philosophers before him had been occupied, and to bring it down to common life, so that it should inquire about virtues and about vices and, in a word, about good things and bad; while as for the heavenly bodies, he judged that either they lie beyond our knowledge, or, even if they were known to the fullest, they would still contribute nothing to living well.

\ciceroSection{16} In nearly all the conversations that have been written out, variously and copiously, by those who heard him, he argues in such a way that he affirms nothing himself but refutes others, says that he knows nothing except that one thing itself, and that in this he surpasses the rest: that they suppose they know the things they do not know, while he alone knows this one thing — that he knows nothing; and that for this reason he believes he was pronounced by Apollo the wisest of all men, because this was the one wisdom belonging to a human being, not to suppose one knows what one does not know. When he said these things steadily and remained firm in that conviction, his whole discourse was spent on praising virtue and on exhorting men to the pursuit of virtue — as can be understood from the books of the Socratics, and most of all of Plato.

\ciceroSection{17} But by the authority of Plato, who was varied and many-sided and copious, a single, consistent form of philosophy was established under two names, that of the Academics and that of the Peripatetics, who, agreeing in substance, differed in their names. For when Plato had left Speusippus, his sister's son, as the heir, so to speak, of his philosophy, but also left two men of the most outstanding zeal and learning, Xenocrates of Chalcedon and Aristotle of Stagira — those who were with Aristotle were called Peripatetics, because they used to debate while walking about in the Lyceum, whereas the others, because by Plato's institution they were accustomed to hold their gatherings and conversations in the Academy, which is another gymnasium, took their name from the name of the place. But both groups, filled full with Plato's abundance, framed a certain settled formula of doctrine — and that one full and well-stocked — while they abandoned the old Socratic practice of disputing about all things in doubt and with no affirmation admitted. So there came about — what Socrates least of all approved — a certain art of philosophy, an ordering of subjects, and a systematic plan of doctrine.

\ciceroSection{18} This, as I said, was at first one under two names; for there was no difference between the Peripatetics and that Old Academy. By a certain abundance of genius Aristotle excelled, as it seems to me at least; but the source was the same for both, and the same was their partition of things to be sought and things to be shunned. But what am I doing,' he said, 'or am I in my senses, to be teaching you these things? For though it is not, as they say, the pig teaching Minerva, still whoever teaches Minerva does it ineptly.' Then Atticus said: 'But do go on, Varro; for I dearly love what is ours and our own people, and these things delight me when they are said in Latin, and in that manner.' 'And what of me,' I said, 'who think that I shall now have to display philosophy to our people, since I have already professed it?'

\ciceroSection{19} 'Let us go on, then,' he said, 'since it is agreed. There was, accordingly, a threefold method of philosophizing, already received from Plato: one concerning life and conduct, a second concerning nature and things hidden, a third concerning argument and the judging of what is true, what false, what correct in speech and what faulty, what consistent and what contradictory. And first, that part of living well they drew from nature, and said that nature must be obeyed, and that the highest good, to which all things are referred, must be sought in nothing else than in nature; and they laid down that the limit of things to be desired and the end of goods was to have attained all that comes from nature in mind, in body, and in life. Of the body, they held that some goods lie in the whole, others in the parts: health, strength, and beauty in the whole; and in the parts, unimpaired senses and some excellence of the several parts, as swiftness in the feet, power in the hands, clearness in the voice, and in the tongue too a distinct impression of the words;

\ciceroSection{20} while of the mind, they held that the things suited to grasping virtue lay in our natural gifts, and these they divided into nature and conduct. To nature they assigned quickness in learning and memory, each of which is proper to the mind and to natural endowment; while of conduct they held there were pursuits and, as it were, habit, which they shaped partly by steady exercise, partly by reason — and in these lay philosophy itself. In philosophy, what has been begun but not completed is called a certain progress toward virtue, while what has been completed is virtue itself, the perfection, so to speak, of nature, and the one best thing of all those they place in our minds.

\ciceroSection{21} These, then, belong to the mind; but to life — for that was the third part — they said there were attached those things that bore upon the exercise of virtue. Now virtue is discerned in the goods of the mind and in those of the body, and in certain things attached not so much to nature as to the happy life. For they held that a human being is, as it were, a kind of part of a community and of the whole human race, and that he is bound up with other human beings by a certain human fellowship. And concerning the highest and natural good they argue thus; but the rest they hold to bear upon this — either upon increasing it or upon keeping it — such as riches, resources, glory, and favour. So a threefold account of goods is introduced by them,

\ciceroSection{22} and these are those three kinds which most people suppose the Peripatetics propound. In this they are not mistaken; for this partition is indeed theirs. But this is imprudent: if they suppose that those then called Academics were one set of men and the Peripatetics another. This account was common to both, and to both this seemed the end of goods — to attain the things that are first in nature and that are to be sought for their own sakes, either all of them or the greatest; and the greatest are those that have to do with the mind itself and with virtue itself. And so all that ancient philosophy held that the happy life is placed in virtue alone — yet not the happiest, unless there be joined to it as well the goods of the body and the rest that were mentioned above as suited to the exercise of virtue.

\ciceroSection{23} From this scheme there was also discovered some starting-point for acting in life, and for duty itself, which lay in the preservation of oneself and in the desire for those things that nature prescribes. From this were begotten a flight from sloth and a contempt for pleasures, and from this the taking up of labours and pains, many and great, for the sake of the right and the honourable and of those things that accord with the prescription of nature; from this too there arose friendship and justice and fairness, and these were preferred both to pleasures and to many advantages of life. Such, then, among them was the framing of conduct, and such the form and scheme of that part which I placed first.

\ciceroSection{24} But concerning nature — for that came next — they spoke in such a way as to divide it into two things: one the agent, the other, as it were, that which offers itself to this and out of which something is produced. In that which acts they held that there was force; in that which is acted upon, only a certain matter; yet in each, both — for matter itself could not have held together if it were contained by no force, nor force without some matter, since there is nothing that is not compelled to be somewhere. But that which arises out of both, this they now called body and, as it were, a certain quality \textit{[Greek: poiotes]} — for you will surely grant, in matters out of the ordinary, that we should use words sometimes unheard-of, as the Greeks themselves do, by whom these subjects have long been handled.'

\ciceroSection{25} 'By all means,' said Atticus; 'and indeed you will be free to use Greek whenever you like, if the Latin should happen to fail you.' 'You are kind to say so; but I shall strive to speak Latin, except in words of this kind — so that I may call something "philosophy" or "rhetoric" or "physics" or "dialectic," words which, like many others, custom now employs as Latin. And so I have called "qualities" what the Greeks call \textit{[Greek: poiotetes]}, a term which even among the Greeks belongs not to common speech but to the philosophers — as is the case with many words. The terms of the dialecticians, however, are not public property at all; they use their own. And this is something nearly all the arts have in common: either new names must be made for new things, or names must be transferred from other things. And if the Greeks do this, who have now been engaged in these matters for so many centuries, how much more must it be allowed to us, who are now attempting to handle these things for the first time?'

\ciceroSection{26} 'You will, I think, deserve well of your fellow citizens, Varro,' I said, 'if you enrich them not only with a store of subject matter — as you have done — but with a store of words as well.' 'We shall venture, then, to use new words on your authority,' he said, 'if it should prove necessary. Of these qualities, then, some are primary and some derived from these. The primary ones are of a single kind and simple; those derived from them are various and, as it were, manifold. And so air (this too we use as a Latin word) and fire and water and earth are primary; but derived from these are the forms of living creatures and of those things that are brought forth from the earth. Therefore those primary things are called the beginnings — or, to render it from the Greek, the elements. Of these, air and fire have the power of moving and of producing, while the remaining parts — I mean water and earth — have the power of receiving and, as it were, of undergoing. A fifth kind, out of which came the stars and the minds, Aristotle held to be something distinct and unlike the four I named above.

\ciceroSection{27} But they suppose that underlying all things is a certain matter, without any form of its own and lacking all that quality (let us make this word more familiar and more worn by using it), out of which all things are shaped and produced — a matter able to receive all things entirely, and to be changed in every way and through every part, and thereby even to perish, not into nothing but into its own parts, which can be cut and divided without limit, since there is nothing whatever in the nature of things so small that it cannot be divided. And all the things that are moved, they hold, are moved by intervals, and these intervals likewise can be divided without limit.

\ciceroSection{28} And since that force which we called quality is moved in this way, and since it is turned back and forth in this manner, they suppose that the matter itself is wholly transformed throughout, and that those things are produced which they call qualified bodies; out of which, in a whole nature cohering and continuous with all its parts, a single world has been produced, outside of which there is no portion of matter and no body. And the parts of the world, they say, are all the things contained within it, which are held together by a sentient nature, in which a perfected reason is present — a reason that is everlasting and the same (for there is nothing more powerful by which it could be destroyed);

\ciceroSection{29} and this force, they say, is the soul of the world, and the same is mind and perfected wisdom, which they call god, and a kind of providence, presiding over all the things that are subject to it, caring above all for the things of heaven, and then on earth for those things that concern human beings. This same force they sometimes call necessity, because nothing can come about otherwise than has been ordained by it; sometimes … as it were a fated and unchangeable continuity of an everlasting order; and at times indeed they call it fortune, because it brings about many things unforeseen and unexpected by us, on account of the obscurity and our own ignorance of causes.

\ciceroSection{30} The third part of philosophy, then, which lay in reason and in argument, was handled by both schools as follows. Although knowledge took its rise from the senses, still the judgment of truth did not lie in the senses. They held that the mind is the judge of things; they reckoned it alone fit to be trusted, because it alone discerns that which is always simple and of a single kind and just such as it is (this they called the \textit{[Greek: idea]}, so named already by Plato; we may rightly call it the "form").

\ciceroSection{31} But all the senses they reckoned dull and slow, incapable in any way of perceiving those things that seemed to be subject to the senses, because these were either so small that they could not fall under sense, or so mobile and so driven that nothing was ever single and constant, nor even the same as itself, since all things were continually slipping and flowing away. And so this whole region of things they called the realm of opinion;

\ciceroSection{32} but knowledge, they held, lay nowhere except in the conceptions and reasonings of the mind. For this reason they approved of definitions of things and applied them to everything that came under dispute; the explanation of words too was approved — that is, the account of why each thing was named as it was, which they called \textit{[Greek: etymologia]}. Afterwards they used certain arguments and, as it were, marks of things to guide them in proving and in concluding what they wished to make plain. Under this head was set out the whole discipline of dialectic — that is, of speech bound tight by reasoning; and to it, as if from the other side, was joined the orator's power of speaking, the unfolder of continuous discourse fitted to persuasion.

\ciceroSection{33} This was their first scheme, handed down from Plato; and if you wish, I will set out what dismantlings of it I have learned.' 'We do indeed wish it,' I said, 'and let me answer for Atticus as well.' 'And rightly do you answer,' he said; 'for the authority of the Peripatetics and of the Old Academy is splendidly set forth. Aristotle, then, was the first to shake those forms of which I spoke a little while ago — forms which Plato had embraced so marvellously that he declared there was something divine in them. Theophrastus, moreover, a man both pleasant in his style and so disposed in character that he carries before him a certain uprightness and nobility, broke even more violently, in a manner of speaking, the authority of the old teaching; for he stripped virtue of its proper splendour and made it feeble, when he denied that the happy life rests on virtue alone.

\ciceroSection{34} For Strato, his pupil, though he was a man of sharp intellect, must nonetheless be removed altogether from that school; for though he abandoned the most necessary part of philosophy, which lies in virtue and in conduct, and gave himself wholly to the investigation of nature, even in that very field he diverged in much from his own people. Speusippus, on the other hand, and Xenocrates, who were the first to take up Plato's method and authority, and after them Polemo and Crates, and along with them Crantor, gathered together in the Academy, carefully guarded what they had received from their predecessors. Now Zeno and Arcesilas had been constant hearers of Polemo.

\ciceroSection{35} But Zeno, since he was older than Arcesilas and argued with great subtlety and reasoned with great acuteness, attempted to reform the teaching. This reform too I will set out, if it seems good, just as Antiochus used to do.' 'It does seem good to me,' I said, 'as you see Pomponius indicating likewise.' 'Zeno, then, was by no means a man who, like Theophrastus, cut the sinews of virtue; on the contrary, he placed in virtue alone everything that pertains to the happy life, and counted nothing else among goods, and called that the honourable which was a kind of simple and single and sole good.

\ciceroSection{36} As for the rest, however, though they were neither good things nor bad, he held nonetheless that some were in keeping with nature, others contrary to nature; and between these very ones he reckoned still others, set in between and intermediate. Those that were in keeping with nature, he taught, were to be taken up and held worthy of a certain valuation, and their contraries the reverse; while the intermediates he left in neither class, holding that in them there lay no weight whatever.

\ciceroSection{37} But of the things to be taken up, some were to be valued more highly, others less. Those of higher value he called ``preferred,'' those of lower value ``rejected.'' And just as here he had changed the terms rather than the substance, so between the right act and the wrong, he placed certain things midway between duty and the breach of duty, setting right acts alone among good actions and crooked ones—that is, wrongs—among bad; but duties either kept or neglected he counted, as I said, as intermediate.

\ciceroSection{38} And whereas his predecessors had said that not all virtue lies in reason, but that certain virtues are brought to completion, as it were, by nature or by habit, he placed every virtue in reason. And whereas they thought that the several kinds of virtue I mentioned above could be held apart, he argued that this could in no way be done, and that not the exercise of virtue alone, as the older men held, but the very disposition itself, in and of itself, was a thing of splendour; and yet that virtue could not be present in anyone without his always making use of it. And whereas they did not strip emotional disturbance from man, but said that to feel grief, to feel desire, to dread, and to be carried away by gladness all belong to nature—though they would contract these and draw them down into narrow bounds—he held that the wise man should be wholly free of them, as of so many diseases.

\ciceroSection{39} And whereas the ancients said that these disturbances are natural and devoid of reason, and lodged desire in one part of the soul and reason in another, he did not assent even to this; for he held that the disturbances are voluntary, taken on by a judgment of opinion, and he reckoned the mother of all disturbances to be a certain unbridled lack of restraint. So much, more or less, for ethics. On the subject of nature he held views as follows. First, of those four primary elements of things he did not bring in this fifth nature, out of which his predecessors believed that sensation and mind were produced; for he laid it down that fire is the very nature that begets each thing, mind and the senses included. He differed from them on this point too, that he held it utterly impossible for anything to be produced by that which is without body—of which kind Xenocrates, and his predecessors before him, had declared the soul to be—and that neither what produces something nor what is produced can be other than body.

\ciceroSection{40} But it was in that third part of philosophy that he changed the most. Here, to begin with, he said certain new things about the senses themselves, which he held to be coupled with a kind of impact, so to speak, offered from outside—what he calls a ``presentation'' \textit{[Greek: phantasia]}, and what we may call an ``impression'' (and let us wear this word smooth, for it will have to be used rather often in the rest of the discussion). But to these things that are presented, and as it were received by the senses, he joins an assent of the mind, which he holds to lie within us and to be a matter of the will.

\ciceroSection{41} Not to all impressions did he attach belief, but only to those that possessed a distinctive declaration of their own about the things presented; and that impression, when it was discerned in and of itself, was ``graspable''—will you put up with this? `Indeed we will,' he said; `for in what other way could you render \textit{[Greek: katalēmpton]}?'—but when it had now been received and approved, he called it a ``grasp,'' like things taken hold of by the hand; and from this he had even drawn the word, since before him no one had used that term in such a matter, and he used a great many new words besides (for what he was saying was new). What had been grasped by sensation he called ``sensation'' itself; and if it had been grasped in such a way that it could not be dislodged by reasoning, he named it ``knowledge,'' but otherwise ``ignorance''; out of which there arose also ``opinion,'' which is feeble and shares in what is false and ungrasped.

\ciceroSection{42} But between knowledge and ignorance he set that ``grasp'' I have spoken of, and he counted it neither among right things nor among wrong, but said that it alone is to be trusted. From this he attributed belief to the senses as well, since, as I said above, a grasp made by the senses seemed to him both true and trustworthy—not because it grasped everything that was in the object, but because it left out nothing that could fall within its reach; and because nature had given it as a kind of standard of knowledge and a starting point of itself, from which, afterward, notions of things might be imprinted on the mind; and from these notions are found not only first principles but certain broader roads as well toward the discovery of truth by reason. Error, recklessness, ignorance, opining, suspicion—everything, in a word, that is foreign to firm and steadfast assent—he removed from virtue and wisdom. And it is in these matters, more or less, that the whole of Zeno's alteration and his dissent from his predecessors consists.''

\ciceroSection{43} When he had said this, `It has been set out very briefly indeed, Varro,' I said, `and not in the least obscurely—both the system of the old Academy and that of the Stoics. And I think the truth is, as my friend Antiochus held, that this is to be reckoned a correction of the old Academy rather than any new school.' Then Varro said: `It is now your part—you who break away from the system of the ancients and approve the innovations Arcesilas made—to teach what the rift was, and on what grounds it was made, so that we may see whether that defection of yours is justified enough.'

\ciceroSection{44} Then I said: `It was with Zeno, as we are told, that Arcesilas set up his whole contest—not, as it seems to me at least, out of obstinacy or a passion for winning, but because of the obscurity of those very things that had brought Socrates to a confession of ignorance, and, even before Socrates, Democritus, Anaxagoras, Empedocles, and nearly all the ancients—men who declared that nothing can be known, nothing perceived, nothing understood; that the senses are narrow, the mind feeble, the courses of life short; that truth, as Democritus put it, lies sunk in the depths; that everything is held fast by opinions and conventions, that nothing is left to truth; and, in short, that everything lies wrapped in darkness.

\ciceroSection{45} And so Arcesilas denied that there was anything that could be known—not even that one thing which Socrates had left himself, the knowing that he knew nothing. So thoroughly did he judge that all things lie hidden in concealment, and that there is nothing that can be discerned or understood; and that for these reasons no one ought to profess or affirm anything, or to approve it by assent, but should always rein himself in and hold his recklessness back from every slip—a recklessness that becomes glaring whenever something false or ungrasped is approved, and that there is nothing more shameful than to let assent and approval run ahead of grasping and perception. And he acted consistently with this principle, so that, by arguing against the views of all men, he drew most of them away from their own: so that, when in one and the same question equal weights of reasoning were found on opposing sides, assent might more easily be withheld from either side.

\ciceroSection{46} This is what they call the New Academy, though to me it seems old, if indeed we count Plato as belonging to that old one—Plato, in whose books nothing is affirmed, much is argued on both sides, everything is inquired into, nothing certain is stated. But let it be so: let the one you have set out be called old, and this one new. It was carried down as far as Carneades, who was fourth from Arcesilas, and it abode in that same system of Arcesilas. Now Carneades was ignorant of no part of philosophy, and, as I have learned from those who had heard him and above all from the Epicurean Zeno—who, though he differed from him profoundly, nonetheless admired this one man beyond all others—was a man of a certain incredible power and … And what is Mnesarchus so vexed about, why does Antipater cross swords with Carneades through volume after volume? …
