% De Natura Deorum (de-natura-deorum) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroBook{1}

\ciceroSection{1} There are many questions in philosophy not yet adequately resolved, but one in particular, Brutus — as you are well aware — is exceptionally difficult and exceptionally obscure: the question concerning the nature of the gods, a question both supremely beautiful for the understanding of the mind and necessary for the regulation of religion. On this matter the views of the most learned men are so various and so discordant that the very fact ought to count as strong proof that ignorance is the source and origin of philosophy, and that the Academics have shown prudence in withholding assent from things uncertain. For what is more shameful than rashness, and what so reckless, so unworthy of a wise man's gravity and steadiness, as either to hold a false opinion, or to defend without the slightest hesitation what has not been adequately examined, grasped, and understood?

\ciceroSection{2} Take this very question: most men have said that the gods exist — which is by far the most probable view, the one to which we are all led by nature as our guide; Protagoras declared himself in doubt; Diagoras of Melos and Theodorus of Cyrene held that there are no gods at all. As for those who said the gods exist, they stand in such variety and disagreement that it would be endless to enumerate their opinions. For much is said about the shapes of the gods, about their places and dwellings, and about the manner of their life, and on these points the philosophers contend in the sharpest disagreement. But the one question that above all governs the whole matter is this: whether the gods do nothing, set nothing in motion, and stand free of all care and administration of things; or whether, on the contrary, all things were made and established by them from the beginning, and are governed and moved by them for all time to come. Here, above all, is the great disagreement; and unless it is decided, men must necessarily wander in the gravest error and in ignorance of the greatest matters.

\ciceroSection{3} For there are philosophers — and there have been — who held that the gods exercise no oversight whatever over human affairs. If their view is true, what room is there for piety, what for sanctity, what for religion? For all these are to be rendered, purely and chastely, to the divine power of the gods only on this condition: that the gods take notice of them, and that something has been granted by the immortal gods to the human race. But if the gods can neither help us nor wish to, if they take no care of us at all, take no notice of what we do, and there is nothing that can pass from them into human life — then what reason is there for us to offer the immortal gods any worship, honors, or prayers? Piety, like the rest of the virtues, cannot exist in a mere show of feigned pretense; and with piety, sanctity and religion must necessarily be abolished as well — and once these are abolished, life is thrown into turmoil and great confusion.

\ciceroSection{4} Indeed, I am inclined to think that, with piety toward the gods removed, good faith too is abolished, and the fellowship of the human race, and that one most excellent of the virtues, justice. There are, however, other philosophers — and these great and distinguished men — who hold that the whole world is administered and governed by the mind and reason of the gods; and not this only, but that by those same gods provision and forethought are given to the life of men. For the crops, and all else the earth brings forth, and the seasons, and the changes of weather, and the alterations of the sky, by which all that the earth produces is ripened and brought to its growth — these they hold to be bestowed on the human race by the immortal gods; and they gather many things, to be reported in these books, of such a kind that the immortal gods seem almost to have fashioned them for the use of men. Against these men Carneades argued so much that he roused those of no sluggish mind to a desire for investigating the truth.

\ciceroSection{5} For there is no subject on which there is such disagreement, not only among the unlearned but among the learned as well; and since their opinions are so various and so much at odds with one another, it is surely possible that none of them is true, and certainly not possible that more than one is. In this case we can both placate the well-meaning critics and refute the spiteful detractors, so that the one party may repent of having found fault and the other may rejoice at having learned something; for those who give friendly admonition should be instructed, while those who pursue us with hostility should be repelled.

\ciceroSection{6} Now I observe that much has been said about the books I have published, a good number of them in a short space of time, and that the talk has run various ways: some marveling whence this enthusiasm for philosophy had sprung up in me so suddenly, others eager to know what settled view I held on each particular question. To many, too, I perceived it seemed strange that the philosophy I had above all approved was the very one that snatches away the light and pours, as it were, a kind of darkness over things — and that I had undertaken, against all expectation, to champion a school deserted and long ago abandoned. But I did not begin to philosophize suddenly, nor have I, from the earliest period of my life, spent only middling labor and care on that pursuit; and precisely when I seemed least to be doing so, I was philosophizing most. This is shown by my speeches, crammed with the doctrines of philosophers, and by my close friendships with the most learned men, in which my house has always abounded, and by those leading figures Diodotus, Philo, Antiochus, and Posidonius, under whom I was trained.

\ciceroSection{7} And since all the precepts of philosophy bear upon the conduct of life, I believe that, in public and in private affairs alike, I have performed what reason and learning prescribe. But if anyone asks what cause impelled me to commit these things to writing so late, there is nothing I can explain more easily. For while I was languishing in idleness, and the condition of the commonwealth was such that it had to be governed by the policy and care of one man, I judged that, for the sake of the commonwealth itself, philosophy ought first to be set forth to my countrymen — thinking it a matter of great moment, for the honor and glory of the state, that subjects so weighty and so splendid should be contained in Latin letters too.

\ciceroSection{8} And I regret my undertaking the less because I readily perceive how many men's enthusiasm I have stirred, not only for learning but for writing as well. For a good many, educated in Greek systems of instruction, were unable to share with their fellow citizens what they had learned, because they had no confidence that what they had received from the Greeks could be said in Latin; and in this respect we seem to have made such progress that we are not surpassed by the Greeks even in richness of vocabulary.

\ciceroSection{9} I was urged to apply myself to these studies, besides, by a sickness of spirit stirred by a great and grievous wrong of fortune; and had I been able to find any greater relief from it, I should not have taken refuge in this above all. But there was no way I could better enjoy that very relief than by giving myself not only to the reading of books but to the thorough treatment of philosophy as a whole. And all its parts, all its members, are most easily understood when whole questions are worked out in writing; for there is a certain remarkable continuity and chain in things, such that one is linked to another, and all seem fitted together and bound to one another.

\ciceroSection{10} Those who ask what I myself think on each particular question are more curious than there is any need to be; for in debate it is not so much the weight of authority as the weight of reasoning that should be sought. Indeed, the authority of those who profess to teach is most often an obstacle to those who wish to learn; for they cease to bring their own judgment to bear, and take as settled whatever they see judged by the man they approve. Nor am I in the habit of approving the practice we are told of among the Pythagoreans, who, when they affirmed something in debate and were asked why it was so, are said to have answered, "He himself said it" — and "he himself" was Pythagoras: so great was the force of a preconceived opinion that authority prevailed even without reason.

\ciceroSection{11} As for those who marvel that I have followed this school above all, they will find a sufficient answer, I think, in my four Academic books. Nor have I taken up the championship of a deserted and abandoned cause; for doctrines do not perish along with the death of men, but perhaps await only the light of an exponent. This method in philosophy — of arguing against all things and openly judging none — sprang from Socrates, was taken up again by Arcesilaus, was confirmed by Carneades, and has flourished down to our own age; though now I understand it to be nearly orphaned in Greece itself. This I judge to have come about not through any fault of the Academy but through the slowness of men. For if it is a great thing to master a single school, how much greater to master all — which they must do who have set themselves, for the sake of discovering the truth, to argue both against all the philosophers and on behalf of all.

\ciceroSection{12} I do not profess to have attained the capacity for so great and so difficult a task, but I do profess to have pursued it. And yet it is not possible that those who philosophize by this method should have nothing to follow. The matter has, to be sure, been discussed elsewhere more carefully; but because certain people are excessively slow to learn and dull, they seem to need reminding more than once. For we are not men who think nothing is true, but men who say that to all truths there are attached certain falsehoods so like them that in them there is no sure mark for judging and assenting. From this arises also that further point: that many things are probable which, though they are not grasped with certainty, yet, because they present an appearance distinct and clear, may serve to govern the life of a wise man.

\ciceroSection{13} But now, to free myself entirely from any ill-will, I shall lay out for public view the opinions of the philosophers concerning the nature of the gods. At this point it seems that all should be summoned to judge which of them is the true one; and only then shall I think the Academy presumptuous, if either all shall have agreed, or someone shall have been found who has discovered what is true. And so I am moved to cry out, as in the Synephebi: "By the gods! I call upon, I demand, I beseech, I beg, I plead and implore the protection of all my countrymen, of all the young men" — not over a trifling matter, as that character complains that capital crimes are committed in the state: "a courtesan refuses to take money from a lover who is her friend" — but that they be present, take cognizance, and consider what view is to be held concerning religion, piety, sanctity, ceremonies, faith, the oath sworn; what concerning temples, shrines, and solemn sacrifices; what concerning the very auspices over which we preside (for all these matters are to be referred to this question about the immortal gods).

\ciceroSection{14} Surely the very men who think they hold something certain will be forced into doubt by so great a disagreement among the most learned men on the greatest of subjects.

\ciceroSection{15} This I have remarked often on other occasions, but most of all when there was a discussion at the house of my friend Gaius Cotta, conducted with real care and precision, concerning the immortal gods. For when, during the Latin Festival, I had come to him at his own request and summons, I found him sitting in an alcove and debating with the senator Gaius Velleius, to whom the Epicureans at that time accorded the first place among our countrymen. Quintus Lucilius Balbus was present too, a man who had made such progress among the Stoics that he was ranked with the most distinguished Greeks in that field. Then, as Cotta caught sight of me, he said: "Most opportunely you come; for I have a great dispute arising with Velleius on a great matter, in which it is fitting, given your studies, that you take part."

\ciceroSection{16} "And indeed I too think," said I, "that I have come, as you say, opportunely. For you three leaders of three schools have met together. If Marcus Piso were here, no school of philosophy — none, at least, of those held in honor — would lack its representative." Then Cotta said: "If the book of our friend Antiochus, recently sent by him to this Balbus here, speaks truly, you have no reason to miss your friend Piso; for according to Antiochus the Stoics agree with the Peripatetics in substance and differ only in words. On that book, Balbus, I should like to know what you think." "I, for my part," said he, "wonder that Antiochus, a man of the keenest mind, failed to see that there is the greatest difference between the Stoics, who separate the honorable from the advantageous not in name but in their whole kind, and the Peripatetics, who mingle the honorable with the advantageous, so that these differ from one another in degree and, as it were, by gradations, not in kind. For this is a disagreement not slight, in words, but very great, in things. But of this another time; for now, if you please, let us turn to what we have begun."

\ciceroSection{17} "It does seem so to me," said Cotta. "But so that our latecomer here, looking at me, may not be in the dark as to what we are about: we were discussing the nature of the gods, and since that subject seemed to me quite obscure — as it always does — I was inquiring of Velleius what Epicurus thought of it. And so, Velleius," he said, "if it is no trouble, take up again what you had begun." "I will take it up again," he replied, "though it is not for me but for you that this helper has arrived; for both of you," he said with a smile, "have learned from one and the same Philo to know nothing." Then I said: "What we have learned is Cotta's affair; but as for you, I would not have you suppose I have come as this man's helper, but as a listener — and an impartial one at that, with a free verdict, bound by no necessity of the kind that would compel me, willing or not, to defend some fixed position."

\ciceroSection{18} Then Velleius, with all the confidence those men display — fearing nothing so much as to seem to be in doubt about anything, as though he had just come down from the council of the gods and from the spaces between Epicurus's worlds — said: "Hear me, then. Not idle and fabricated opinions: not the god who is the craftsman and builder of the world in Plato's \textit{Timaeus}; not the Stoics' prophesying old woman, their Providence \textit{[Greek: Pronoia]}, which we may render in Latin as \textit{Providentia}; nor indeed the world itself, endowed with mind and senses, a round, blazing, revolving god — monstrosities and marvels worthy not of philosophers reasoning but of philosophers dreaming.

\ciceroSection{19} For with what eyes of the mind could your Plato look upon that workshop of so vast an undertaking, in which he has the god construct and build the world? What was the contrivance, what the tools, what the levers, what the engines, what the assistants for so great a task? How could air, fire, water, and earth obey and submit to the will of the architect? And from where did those five shapes arise, out of which the rest are formed, fitly disposed to act upon the mind and to give birth to the senses? It would take long to go through it all — matters of such a kind that they seem wished for rather than discovered;

\ciceroSection{20} but the crowning point is this, that a man who has introduced a world not merely born but all but made by hand should declare that it will last forever. Do you suppose that a man who thinks anything that has come to be can be eternal has so much as touched natural philosophy with the tip of his lips, as the saying goes — has so much as tasted the rationale of nature? For what compound is there that cannot be dissolved, or what is there that has a beginning of some kind and has no end? And as for your Providence, Lucilius, if it is the same thing, I demand of it what I demanded a moment ago — the assistants, the engines, the entire plan and apparatus of the whole work; but if it is something else, then why did it make the world mortal, and not everlasting, in the manner of the Platonic god?

\ciceroSection{21} And of both of you alike I inquire: why did these builders of the world spring up so suddenly, after sleeping through countless ages? For it does not follow that, because there was no world, there were no ages. (By 'ages' I do not now mean those that are made up of the number of days and nights in the yearly courses; for I grant that these could not have come about without the revolution of the world; but there was, from infinite time, a certain eternity which no measuring of periods bounded, though it can be understood in respect of its extent, since it does not even fall within thought that there should have been

\ciceroSection{22} some time when there was no time at all.) In so immense a span, then, I ask, Balbus, why did your Providence rest idle. Was it shirking labor? But labor neither touches a god nor was there any, since all the natures — sky, fires, lands, seas — were obeying the divine will. And what was it that the god could desire, to deck out the world with constellations and lights like an aedile? If it was that the god himself might dwell more comfortably, then before this, of course, he had dwelt for infinite time in darkness, as in a hovel. And afterward — are we to suppose he takes delight in the variety with which we see the sky and the lands adorned? What manner of pleasure can this be for a god? And if it were any, he could not have gone without it for so long.

\ciceroSection{23} Or were these things, as you generally say, established by the god for the sake of men? For the sake of the wise? Then so vast an effort of construction was undertaken for the sake of a few. Or for the sake of fools? But, in the first place, there was no reason why he should do a kindness to the worthless; and next, what did he gain by it, since all fools are beyond doubt the most wretched of beings — most of all because they are fools (for what can we name more wretched than folly?), and next because there are so many troubles in life that the wise soften them by setting their advantages against them, while fools can neither avoid the troubles as they come nor bear them once they are present? Now those who said that the world itself is alive and wise saw in no way at all into what figure the nature of an intelligent mind could fall. Of this I shall indeed speak a little later; for now, only so far:

\ciceroSection{24} I shall marvel at the dullness of those who would have it that something living, immortal, and likewise blessed should be round, on the ground that Plato denies there is any shape more beautiful than that — whereas to me a cylinder, or a cube, or a cone, or a pyramid seems more shapely. And what manner of life is granted to this round god? Why, that he should be whirled about with a speed than which none even greater can be conceived; and I do not see how a steady mind and a blessed life could come to rest in such motion. And that which, if it were felt in the smallest part of our own body, would be a distress — why should this same thing not be reckoned a distress in a god? For surely the earth, since it is a part of the world, is also a part of the god; and yet we see that the largest regions of the earth are uninhabitable and untilled, because part of them is scorched by the sun's onset, and part stiffened with snow and frost where the sun has long withdrawn; and these — if the world is a god, since they are parts of the world — must be reckoned the limbs of the god, some of them burning, some of them frozen.

\ciceroSection{25} And these doctrines, Lucilius, are indeed yours; but to show what kind they are, I shall trace them back from the very earliest of the older thinkers. Thales of Miletus, then, who first inquired into such matters, said that water was the first principle of things, and that the god was that mind which fashioned all things out of water — as if gods could exist without sense; and why did he join mind to water, if mind itself can subsist devoid of body? Anaximander's opinion is that there are gods born at long intervals as they rise and set, and that these are the innumerable worlds. But how can we conceive of a god except as everlasting?

\ciceroSection{26} After him, Anaximenes laid it down that air is a god, and that it comes to be and is immeasurable and infinite and ever in motion — as if either air could be a god when it has no shape (especially since it befits a god to be not merely of some appearance but of the most beautiful), or as if mortality did not attend upon everything that has come to be. Next Anaxagoras, who took over his teaching from Anaximenes, was the first who held that the ordering and measure of all things was designed and brought to completion by the force and rationale of an infinite mind. In this he failed to see that there can be no motion joined to sense and continuous within the infinite, and no sense at all whereby nature itself, being unstruck, should feel. Then, if he meant that mind of his to be a kind of living thing, there will be something more inward from which that living thing takes its name; but what is more inward than mind? Let it be girt, then, with an outer body;

\ciceroSection{27} and since this does not satisfy him, a mind open and simple, with nothing joined to it by which it could feel, seems to elude the reach and conception of our understanding. Alcmaeon of Croton, again, who attributed divinity to the sun, the moon, and the rest of the heavenly bodies, and to the soul besides, did not perceive that he was conferring immortality upon mortal things. For Pythagoras, who held that the soul is stretched through and travels across the whole nature of things, and that from it our souls are plucked, did not see that, by the tearing-away of human souls, the god would be rent and mangled, and that, when souls were wretched — as happens to most — then part of the god would be wretched, which cannot be.

\ciceroSection{28} And why should the mind of a man be ignorant of anything, if it were a god? Furthermore, how could that god, if it were nothing but soul, be either fixed in the world or poured through it? Then Xenophanes, who would have it that everything besides — the whole that was infinite — with mind added to it, was god, will be censured concerning mind itself as the others are, and concerning infinity more sharply still, since in it there can be nothing that has sense and nothing that is conjoined. As for Parmenides, he contrives something quite fanciful: he makes a ring resembling a crown (he calls it \textit{[Greek: stephane]}), a continuous circle of glowing light that girds the sky, which he calls god — in which no one can suspect either divine shape or sense. And there are many other monstrosities of his, seeing that he refers war, discord, desire, and the rest of that kind back to a god — things that are abolished by disease, or sleep, or forgetfulness, or age; and the same of the heavenly bodies, which, having been faulted in another, may now be passed over in him.

\ciceroSection{29} Empedocles, again, erring in many other ways, falls most disgracefully in his notion of the gods. For he would have the four natures, out of which he holds all things to be composed, to be divine — though it is plain that they come to be and are extinguished and are wholly without sense. Nor does Protagoras, who declares that he has nothing clear to say about the gods at all — whether they are, or are not, or what they are like — seem to suspect anything about the nature of the gods. And what of Democritus, who at one moment counts the images and their orbits among the number of gods, at another that nature which pours forth and sends out the images, at another our own thinking and intelligence — does he not move in the gravest error? Since the same man, because nothing ever abides in its own state, denies that anything is everlasting, does he not do away with a god altogether, so as to leave no notion of him remaining? And the air, which Diogenes of Apollonia employs as a god — what sense can it have, or what shape of a god?

\ciceroSection{30} As for Plato's inconstancy, it would take long to speak of it: in the \textit{Timaeus} he denies that the father of this world can be named, while in the books of the \textit{Laws} he holds that what god is at all ought not to be inquired into. And as for his wish that god should be without any body (what the Greeks call \textit{[Greek: asomaton]}), it cannot be understood what such a thing could be: for it must of necessity lack sense, must lack practical wisdom, must lack pleasure as well — all of which we comprehend together with the notion of the gods. The same man says, both in the \textit{Timaeus} and in the \textit{Laws}, that the world is a god, and the sky too, and the stars, and the earth, and souls, and the gods we have received by the institutions of our ancestors. These claims are clearly false in themselves and violently at war with one another.

\ciceroSection{31} And Xenophon too, in fewer words, commits nearly the same faults; for in those sayings which he reported as spoken by Socrates, he makes Socrates argue that the shape of god ought not to be inquired into, and the same Socrates call both the sun and the soul a god, and speak now of one god and now of several — errors that are almost the same as those we charge against Plato.

\ciceroSection{32} And Antisthenes too, in the book entitled \textit{The Natural Philosopher}, in saying that there are many gods of the people but one in nature, does away with the power and nature of the gods. Nor much otherwise does Speusippus, following his uncle Plato and speaking of a certain force by which all things are governed, and that force a living one, strive to tear the knowledge of the gods out of our souls.

\ciceroSection{33} Aristotle, too, in the third book of his work \textit{On Philosophy}, throws everything into confusion in his dissent from his master Plato. At one moment he assigns all divinity to mind; at another he says that the world itself is a god; at another he sets some other being over the world and gives it the office of governing and preserving the world's motion by a kind of rolling rotation; then he says that the heat of the sky is a god, failing to grasp that the sky is a part of the world, which he himself in another passage had designated a god. But how can that divine consciousness of the sky be maintained amid such great speed? And then again, where are all those gods of his, if we count the sky too as a god? When, moreover, he holds that this same god is without a body, he strips him of all sensation, and of practical wisdom as well. And how, furthermore, can the world move if it has no body, or how can it be at rest and blessed if it is forever in motion?

\ciceroSection{34} Nor is his fellow pupil Xenocrates any wiser in this field. In his books \textit{On the Nature of the Gods} no divine form is described at all. For he says the gods are eight in number: five are those that are named among the wandering stars; one is to be regarded as a single god made up, as it were, of scattered limbs, out of all the heavenly bodies that are fixed in the sky; the sun he adds as the seventh, and the moon as the eighth — and by what kind of sensation these could be blessed cannot be understood. From the same school of Plato, Heraclides of Pontus crammed his books full of childish fables, and even so he supposes now the world, now the divine mind, to be a god; he assigns divinity to the wandering stars as well, strips the god of sensation, would have his form be subject to change, and in the same book sets earth and sky once more among the gods.

\ciceroSection{35} Nor is the inconsistency of Theophrastus to be borne; for at one moment he assigns the divine primacy to mind, at another to the sky, and then again to the constellations and the heavenly bodies. Nor should we listen to his pupil Strato, the one who is called the natural philosopher, who holds that all divine power resides in nature, which possesses the causes of generation, growth, and diminution, but is devoid of all sensation and all shape.

\ciceroSection{36} Zeno, however — to come now to your school, Balbus — holds that the law of nature is divine, and that it has the force of commanding what is right and forbidding the contrary. How he makes this law a living being we cannot understand; and yet a god, surely, we mean to be a living being. And this same man, in another passage, says that the upper air is a god — if a god that has no sensation can be conceived at all, a god who never meets us in our prayers, our wishes, or our vows. In yet other books he supposes that a certain reason pervading the whole nature of things is endowed with divine power. He assigns this same divinity to the stars, then to the years, the months, and the changing seasons. And when he interprets Hesiod's \textit{Theogony} — that is, the origin of the gods — he abolishes altogether the customary and received conceptions of the gods; for he counts neither Jupiter nor Juno nor Vesta nor anyone called by such a name among the gods, but teaches that these names have been assigned, through a kind of symbolism, to inanimate and voiceless things.

\ciceroSection{37} No less gravely mistaken is the view of his disciple Aristo, who holds that the form of a god cannot be conceived, denies that there is sensation in the gods, and is altogether in doubt whether a god is a living being or not. Cleanthes, again, who heard Zeno together with the man I have just named, at one point says that the world itself is a god, at another assigns this name to the mind and soul of the whole of nature, and at another judges that the surest god is the outermost and highest heat — poured around on every side, encircling and embracing all things at the very edge — which is called the upper air; and the same man, as though raving, in those books he wrote against pleasure now fashions a certain form and appearance of the gods, now assigns all divinity to the stars, now holds that nothing is more divine than reason. The upshot is that this god of his, whom we apprehend with the mind and wish to lodge in the mind's conception as in a footprint, nowhere appears at all. But Persaeus, another pupil of this same Zeno, holds that those have been accounted gods by whom some great benefit toward the conduct of life has been discovered, and that useful and salutary things themselves have been called by the names of gods — so that he does not even say that these were the discoveries of the gods, but that they are themselves divine.

\ciceroSection{38} And what could be more absurd than this — either to invest sordid and unsightly things with the honor of gods, or to set up among the gods men already blotted out by death, the whole worship of whom would consist in mourning?

\ciceroSection{39} And now Chrysippus, who is held to be the shrewdest interpreter of the Stoics' dreams, musters a great throng of unknown gods — and so unknown that we cannot form an idea of them even by conjecture, though our mind seems able to picture anything whatever in thought. For he says that divine power resides in reason and in the soul and mind of universal nature; he says the world itself is a god, and the universal diffusion of its soul; then the ruling part of that very soul, which dwells in mind and reason; then the common nature of all things, universal and embracing everything; then the fated power and necessity of things to come; and besides these, fire, and the upper air I mentioned before; then those things that flow and stream by their own nature, such as water, earth, and air, the sun, the moon, the stars, and the whole sum of things in which all are contained; and even those men who have attained immortality.

\ciceroSection{40} And the same man argues that the upper air is what men call Jupiter, that the air which streams through the seas is Neptune, and that the earth is what is called Ceres; and he pursues the names of the remaining gods by a like method. Likewise he says that Jupiter is the force of the everlasting and eternal law, which is, as it were, the guide of life and the mistress of our duties; and this same thing he calls the fated necessity, the sempiternal truth of things to come — none of which is of such a kind that divine power could be seen to reside in it.

\ciceroSection{41} And this, indeed, is in the first book \textit{On the Nature of the Gods}; in the second, however, he means to accommodate the tales of Orpheus, Musaeus, Hesiod, and Homer to what he himself had said in the first book about the immortal gods, so that even the most ancient poets, who never so much as suspected these things, may be made to look like Stoics. Following him in this, Diogenes of Babylon, in the book entitled \textit{On Minerva}, severs the birth of Jupiter and the rise of the virgin from the realm of fable by transferring them to natural philosophy.

\ciceroSection{42} I have laid out, more or less, not the considered judgments of philosophers but the dreams of madmen. For these are scarcely more absurd than the notions which, poured out in the voices of the poets, have done harm by their very sweetness — poets who brought on gods inflamed with anger and raging with lust, and made us behold their wars, their battles, their combats, their wounds; their hatreds, besides, their estrangements and discords, their births and deaths, their complaints and lamentations, their lusts poured out in every kind of intemperance, their adulteries, their chains, their couplings with the human race and the mortals begotten of immortal stock.

\ciceroSection{43} With the error of the poets one may couple the monstrosities of the magi and the madness of the Egyptians in this same kind, and then the opinions of the common people too, which dwell in the utmost inconstancy through ignorance of the truth. Anyone who considers how thoughtlessly and recklessly these things are said ought to revere Epicurus and to count him among the very beings about whom this inquiry turns. For he alone saw, first, that the gods exist, because nature herself has imprinted a conception of them on the minds of all. For what nation is there, or what kind of human beings, that does not have, without instruction, a certain preconception of the gods — what Epicurus calls \textit{[Greek: prolēpsin]}, that is, a certain outline of a thing conceived in advance in the mind, without which nothing can be understood, or investigated, or argued about? The force and usefulness of this principle we have learned from that celestial volume of Epicurus \textit{On the Standard and the Criterion}.

\ciceroSection{44} You see, then, that the foundation of this inquiry has been splendidly laid. For since the belief has been established not by any institution, custom, or law, and since a firm agreement of all men persists upon this one point, it must be understood that the gods exist, because we have implanted — or rather inborn — knowledge of them; and that on which the nature of all men agrees must of necessity be true; we must therefore confess that the gods exist. And since this is more or less agreed not only among philosophers but also among the unlearned, we acknowledge it to be agreed as well that we possess this — call it preconception, as I said before, or prior notion of the gods (for to new things new names must be assigned, just as Epicurus himself called it \textit{[Greek: prolēpsis]},

\ciceroSection{45} a name by which no one before him had named it) — this, then, we possess: that we hold the gods to be blessed and immortal. For the same nature that gave us an outline of the gods themselves also engraved it on our minds that we should hold them eternal and blessed. And if this is so, then that doctrine of Epicurus was truly set forth — that whatever is blessed and eternal neither has any trouble of its own nor causes any to another, and so is held fast neither by anger nor by favor, since all such things belong to weakness. If we sought nothing further than to worship the gods piously and to be freed from superstition, enough would have been said; for the surpassing nature of the gods would be worshipped by the piety of men, being both eternal and supremely blessed (since whatever excels rightly commands veneration), and every fear arising from the violence and anger of the gods would have been driven off; for it is understood that anger and favor are foreign to a blessed and immortal nature, and that, once these are removed, no fears hang over us from the powers above. But to confirm this conviction, the mind seeks out the form, the life, the action of mind, and the activity of a god.

\ciceroSection{46} As to form, nature in part admonishes us and reason in part instructs us. For from nature we all have, in every nation, no other notion of the gods' appearance than the human one; for what other form ever presents itself to anyone, whether waking or sleeping? But that we may not refer everything to first conceptions, reason itself declares this same thing.

\ciceroSection{47} For since it seems fitting that the most surpassing nature — whether because it is blessed or because it is everlasting — should also be the most beautiful, what composition of limbs, what configuration of features, what figure, what appearance can be more beautiful than the human? You, Lucilius, are accustomed (for my friend Cotta does it now one way, now another), when you represent the craftsmanship and divine workmanship, to describe how all things in the human figure are fitted not only to use but to grace as well;

\ciceroSection{48} and if the human figure surpasses the form of all living creatures, and a god is a living creature, then surely his is the figure that is the most beautiful of all. And since it is agreed that the gods are supremely blessed, and no one can be blessed without virtue, nor can virtue subsist without reason, nor reason reside anywhere except in the human figure, we must confess that the gods have the appearance of a man.

\ciceroSection{49} And yet that appearance is not body but, so to speak, a quasi-body; it has not blood but, so to speak, a quasi-blood. These matters were discovered by Epicurus with greater acuteness, and stated with greater subtlety, than that anyone at all could grasp them; nevertheless, trusting to your intelligence, I treat them more briefly than the subject demands. Epicurus, then — who not only beholds with the mind things hidden and buried deep within, but handles them as though with the hand — teaches that the power and nature of the gods is such that, in the first place, it is discerned not by the senses but by the mind; not by any solidity, nor as a numbered thing, like those bodies which he, on account of their firmness, calls \textit{[Greek: steremnia]}; but apprehended through images grasped by likeness and by passage, since an infinite procession of the most similar images arises from countless atoms and streams toward the gods, and since, with the keenest pleasures, our understanding, its mind fixed and fastened upon those images, takes in what the blessed and eternal nature is.

\ciceroSection{50} The power of infinity, moreover, is supreme, and most worthy of careful and attentive contemplation. In it we must understand that the nature of things is such that all things answer to all in equal balance, like for like; this Epicurus calls \textit{[Greek: isonomia]}, that is, equable distribution. From this it follows that, if the multitude of mortals is so great, the number of immortals is no less; and if the things that destroy are countless, those that preserve must also be infinite. And you, Balbus, are in the habit of asking us what the life of the gods is, and what manner of existence they pass.

\ciceroSection{51} It is, of course, the life than which nothing more blessed, nothing more abounding in every good, can be conceived. For the god does nothing, is entangled in no occupations, sets no labors in motion; he takes joy in his own wisdom and virtue, holding it certain that he will dwell forever amid pleasures both supreme and eternal.

\ciceroSection{52} This god we may rightly call blessed; yours, on the other hand, most toil-burdened. For if the world itself is a god, what can be less at rest than to revolve about the axis of the sky with marvelous speed, not pausing for a single point of time? But nothing that is not at rest is blessed. And if some god dwells within the world itself, to rule it, to steer it, to preserve the courses of the stars, the changes of the seasons, the orderly successions of things, watching over lands and seas to guard the welfare and the lives of men — surely that god is tangled up in irksome and laborious business.

\ciceroSection{53} We, however, place the blessed life in tranquility of mind and in release from all duties. For the same teacher who taught us the rest taught us this: that the world was produced by nature, that there was no need of any workmanship, and that the thing you deny can be accomplished without divine skill is so easy that nature has produced, is producing, and will produce countless worlds. But because you do not see how nature can accomplish this without some mind, you take refuge in a god — like the tragic poets who, unable to work out the resolution of their plot, flee to one.

\ciceroSection{54} Surely you would feel no need of that god's labor if you beheld the measureless and boundless magnitude of regions extending on every side, into which the mind, hurling and straining itself, ranges so far and so wide that it sees nonetheless no farthest shore on which it might come to a halt. In this immensity, then, of breadths and lengths and depths, an infinite host of countless atoms flits about, which, though void lies between them, yet cohere among themselves and, gripping one another, are linked in unbroken chains; and from this are produced those forms and figures of things which you do not think can be made without bellows and anvils. And so you have set over our necks an everlasting master, to fear by day and night. For who would not fear a god who foresees all things, ponders all things, marks all things, supposes all things to concern him — a meddling god, full of business?

\ciceroSection{55} From this there first arose for you that fatal necessity which you call \textit{[Greek: heimarmene]}, so that whatever happens you declare to have flowed from eternal truth and the unbroken chain of causes. But how much is this philosophy to be valued, in which, as to little old women — and unlettered ones at that — all things appear to happen by fate? Then follows your \textit{[Greek: mantike]}, which in Latin is called divination, by which we would be steeped in such superstition, were we willing to listen to you, that we should have to revere soothsayers, augurs, fortune-tellers, seers, and interpreters of omens.

\ciceroSection{56} Freed from these terrors by Epicurus and set at liberty, we neither fear those gods whom we understand to contrive no trouble for themselves nor to seek it for another, and we worship, with piety and reverence, a nature surpassing and preeminent. But carried away by my zeal, I fear I have gone on too long. It was hard, though, to leave so great and so splendid a subject half-finished — although it was not so much my own speaking I should have kept in mind as your listening.''

\ciceroSection{57} Then Cotta, courteously, as was his way, said: ``And yet, Velleius, if you had said nothing at all, you certainly could have heard nothing from me. For it does not usually come into my mind so readily why something is true as why it is false; and this has happened to me often, and not least just now while I was listening to you. Ask me what I say the nature of the gods is: perhaps I shall give no answer. Ask whether I think it is such as you have just set out: I shall say nothing seems to me less so. But before I take up the points you have argued, I will tell you what I think of you yourself.

\ciceroSection{58} For I often think I have heard from that intimate friend of yours, Lucius Crassus, when he placed you without question above all our countrymen and reckoned but few Epicureans from Greece your equal — but, because I understood that he held you in wonderful affection, I supposed he was saying this rather lavishly out of goodwill. As for me, though I am wary of praising a man to his face, I judge nonetheless that on a matter obscure and difficult you have spoken with clarity, and not only abundantly in your thoughts but more elegantly in your words than your school is wont.

\ciceroSection{59} Zeno, whom our friend Philo used to call the chief of the Epicureans, I heard frequently when I was at Athens — and indeed at Philo's own urging, I believe so that I might more easily judge how well those doctrines were refuted, once I had learned from the leading Epicurean how they were stated. So he spoke not as most of them do, but in your manner — distinctly, weightily, with finish. But what often happened to me in his case happened to me just now while I listened to you: I was vexed that so great a talent — you will hear me out, with your good leave — had fallen upon opinions so trifling, not to say so silly.

\ciceroSection{60} Nor do I myself now bring forward anything better. For, as I just said, in nearly all matters, but most of all in natural philosophy, I could sooner say what a thing is not than what it is. Ask me what god is, or of what kind: I shall take Simonides for my authority — for when the tyrant Hiero asked him this same question, he begged one day to deliberate; when Hiero asked him again the next day, he sought two; and when he kept doubling the number of days, and Hiero in astonishment asked why he did so, he said, ``Because the longer I consider it, the more obscure the prospect of an answer seems to me.'' But I think Simonides — who is handed down to us not only as a delightful poet but also as learned and wise in other respects — despaired of the whole truth because, when so many acute and subtle points came into his mind, he was in doubt which of them was truest.

\ciceroSection{61} But your Epicurus — for I would rather argue with him than with you — what does he say that is worthy not merely of philosophy but of ordinary good sense? The first question raised in that inquiry into the nature of the gods is whether the gods exist or do not. ``It is hard to deny it.'' I should think so, if the question were put before a public assembly; but in a discussion of this kind and in a gathering of friends, it is the easiest thing in the world. And so I, a pontiff myself, who hold that the public ceremonies and rites must be guarded with the utmost reverence — I should genuinely wish to be convinced of this, the first point, that the gods exist, not by mere opinion but plainly, in accordance with the truth. For many things present themselves to throw me into confusion, so that at times it seems they do not exist at all.

\ciceroSection{62} But see how generously I deal with you: I shall not touch the matters you hold in common with the rest of the philosophers — such as this very point; for nearly everyone, and I myself above all, agrees that the gods exist. And so I do not contest it; the argument you bring, however, I think not firm enough. You said that, since this is the view of men of all nations and races, it is argument great enough that we should acknowledge the gods exist. But that is not only feeble in itself; it is also false. For, to begin with, how are the opinions of the nations known to you? For my part, I believe there are many peoples so brutalized by savagery that among them there is no inkling of gods at all.

\ciceroSection{63} What of Diagoras, who was called the Atheist, and afterward Theodorus — did they not openly abolish the very nature of the gods? As for Protagoras of Abdera, of whom you made mention just now, the greatest sophist perhaps of those times, when he had set down at the beginning of a book, ``Of the gods I am able to say neither that they exist nor that they do not exist,'' he was by decree of the Athenians banished from their city and territory, and his books were burned in the assembly. From which I judge that many were made the slower to profess this view, since not even doubt could escape the penalty. And what shall we say of the sacrilegious, of the impious and the perjured? Had Lucius Tubulus ever, or Lupus, or Carbo, or the son of Neptune, as Lucilius calls him, believed that the gods exist, would he have been so perjured or so foul?

\ciceroSection{64} That reasoning of yours, then, is not so well established for confirming what you wish as it seems. But since this argument is shared by other philosophers as well, I shall let it go for the present; I prefer to come to what is peculiarly yours.

\ciceroSection{65} I grant that there are gods. Teach me, then, where they come from, where they are, what they are like in body, in mind, in manner of life — for these are the things I long to know. To explain everything, you make free use of the realm and license of the atoms; out of these you mold and fashion whatever, as the saying goes, falls to hand. But in the first place, there are no such things; for there is nothing that is free of body, while every place is occupied by bodies — so that there can be no void, and nothing indivisible.

\ciceroSection{66} These oracles of the natural philosophers I am now pouring forth — whether they are true or false I do not know, yet they are at least more like the truth than yours. For those scandals of Democritus, or even earlier of Leucippus — that there are certain little bodies, some smooth, others rough, some round, others angular and hooked, some curved and as it were bent back, and that out of these the heaven and the earth were brought into being by no compelling natural force but by a kind of chance collision — this is the opinion, Gaius Velleius, that you have carried with you down to this present day; and a man would sooner dislodge you from your whole standing in life than from that authority. For you decided that you ought to be an Epicurean before you came to know these doctrines: so it was necessary either to take these scandals into your mind, or to lose the name of the philosophy you had taken up.

\ciceroSection{67} For what would you take to cease being an Epicurean? "Nothing whatever," you say, "that I should desert the rational scheme of the happy life and the truth." Is that, then, the truth? As for the happy life, I make no objection — though you hold that not even a god possesses it unless he languishes in sheer idleness. But where is the truth? In the countless worlds, I suppose, all coming to birth and all passing away at the smallest points of time, some being born, others falling? Or in those indivisible little bodies that fashion works so splendid with no nature to govern them, no reason to shape them? But I have forgotten the generosity I began to show you a little while ago, and I am taking on too much. I will grant, then, that everything is composed of indivisibles. What has that to do with the matter?

\ciceroSection{68} For the question concerns the nature of the gods. Let them, by all means, be made of atoms; then they are not eternal. For if they are made of atoms, the thing came into being at some point; and if it came into being, there were no gods before they were born; and if the gods had a birth, they must necessarily have a death as well, as you yourself were arguing a little while ago about Plato's world. Where, then, is that "blessed and eternal" of yours — the two words by which you signify a god? When you try to make this good, you scuttle off into the brambles. For you were saying that there is in a god not body but as it were body, not blood but as it were blood.

\ciceroSection{69} This is something you do very often: when you have said something that is not like the truth and you want to escape censure, you bring in something that absolutely cannot even happen — so that it would have been better to concede the very point in dispute than to hold your ground so shamelessly. Just so Epicurus, when he saw that, if the atoms were carried downward by their own weight, nothing would be in our power, since their motion would be fixed and necessary, found a way to escape that necessity — a thing that, plainly, had escaped Democritus: he says that the atom, while it is borne straight downward by weight and gravity, swerves ever so slightly.

\ciceroSection{70} To say this is more disgraceful than to be unable to defend the position he wants to hold. He does the same against the logicians. They have handed down the rule that in every disjunction in which "either yes or no" is set down, one or the other must be true; Epicurus took fright lest, if some proposition of this kind were granted — "Epicurus either will be alive tomorrow or will not be alive" — one or the other of the alternatives should become necessary; and so he denied that the whole "either yes or no" is necessary. What could be said more obtuse than that? Arcesilas pressed Zeno hard, since he himself declared that everything which appears to the senses is false, while Zeno held that some appearances are false but not all; Epicurus took fright lest, if a single appearance were false, none should be true: and so he declared that all the senses are messengers of truth. In none of this was there any shrewdness; for he was taking a heavier blow in order to fend off a lighter one.

\ciceroSection{71} He does the same with the nature of the gods: while he shrinks from any concretion of indivisible bodies, lest dissolution and dispersal should follow, he denies that there is body in the gods but says "as it were body," not blood but "as it were blood." It is held a marvel that a soothsayer does not laugh when he sees another soothsayer; is it not more of a marvel that you can keep from laughing among yourselves? "It is not body but as it were body." I could grasp what this meant if it were modeled in figures of wax or clay; but in a god what "as it were body" might be, or what "as it were blood" might be, I cannot grasp. Nor can you either, Velleius, but you will not admit it.

\ciceroSection{72} For these things are recited by you like dictation, things that Epicurus dreamed up yawning — boasting, as we see in his writings, that he had no teacher. And even if he were not proclaiming it, I would readily believe him, just as I would the owner of a badly built house who boasted he had had no architect; for there is no whiff of the Academy about him, none of the Lyceum, none even of a schoolboy's lessons. He might have heard Xenocrates — what a man, by the immortal gods! — and there are those who think he did hear him; he himself denies it: I believe him, since he would deny it of anyone. He says that he heard a certain Pamphilus, a pupil of Plato, at Samos (for there he lived as a young man with his father and brothers, because his father Neocles had gone there as a settler claiming land, but, since the little farm did not feed him well enough, I suppose, he became a schoolmaster);

\ciceroSection{73} but this Platonist Epicurus holds in remarkable contempt, so afraid is he of ever seeming to have learned anything. On Nausiphanes the Democritean he is caught fast; for though he does not deny having heard him, he yet assails him with every insult. And yet, if he had not heard these Democritean doctrines, what had he heard? What is there in the natural philosophy of Epicurus that is not from Democritus? For although he changed certain things, such as what I said a little while ago about the swerve of the atoms, still he says most of the same things — atoms, the void, the images, the infinity of space and the innumerability of worlds, their risings and perishings, almost everything in which the account of nature is contained. Now this "as it were body" and "as it were blood" — what do you understand by it?

\ciceroSection{74} For I not only acknowledge but readily accept that you know these matters better than I do; and yet, once they have been stated, what is there that Velleius can understand and Cotta cannot? And so I understand what body is and what blood is; what "as it were body" and "as it were blood" are I understand in no way at all. Nor are you hiding it from me, as Pythagoras used to hide things from outsiders, nor are you speaking obscurely on purpose, in the manner of Heraclitus; but — and let this pass between us — you do not understand it yourself.

\ciceroSection{75} What I see is that you are fighting for this: that there should be a certain appearance of the gods, having nothing concrete, nothing solid, nothing distinct, nothing in relief, and that it should be pure, light, transparent. We shall say, then, the same as of the Venus of Cos: that body of hers is not body but the likeness of body; and that flush, suffused and mingled with whiteness, is not blood but a certain semblance of blood; so in the Epicurean god there is not substance but the semblances of substance. Grant that I am persuaded of a thing that cannot even be grasped: give me, then, the outlines and forms of these shadow-sketched gods.

\ciceroSection{76} Here there is no lack of arguments by which you would teach that the gods have human forms: first, that it is so impressed and anticipated in our minds that, when a man thinks of a god, a human form presents itself; next, that since the divine nature surpasses all things, its form too ought to be the most beautiful, and there is no form more beautiful than the human; and you bring forward a third argument, that in no other shape could there be a dwelling-place for mind.

\ciceroSection{77} Consider, then, each of these in turn, what it is worth; for you seem to me to seize, as if by your own right, on a thing in no way to be approved. First of all, who was ever so blind in contemplating the world that he did not see that these human likenesses have been ascribed to the gods either by some deliberate design of wise men, the more easily to turn the minds of the ignorant from depravity of life toward the worship of the gods, or by superstition, so that there should be images which the worshippers believed they were approaching the gods themselves. And these same notions the poets, the painters, and the craftsmen further enlarged; for it was not easy to keep gods that were doing and contriving something in the imitation of other forms. There came in also, perhaps, the opinion that nothing seems more beautiful to a man than a man. But do you not, as a natural philosopher, see this — how coaxing a matchmaker nature is, and as it were a bawd on her own behalf? Or do you think there is any beast, on land or sea, that does not take the greatest delight in a beast of its own kind? If it were not so, why would the bull not be eager at the touch of a mare, or the horse at the cow's? Do you suppose that an eagle, or a lion, or a dolphin prefers any shape to its own? What wonder, then, if in the very same way nature has prescribed to a man that he should think nothing more beautiful than a man? That, I take it, is the reason why we thought the gods like men. What do you suppose — if there were reason in the beasts, would they not each have assigned the most to its own kind?

\ciceroSection{78} But, by Hercules — for I will speak as I feel — much as I love myself, I still do not dare to say that I am more beautiful than that bull was who carried off Europa; for the question here is not about our talents or our discourses, but about appearance and shape. And if we chose to invent and join forms together — such as that sea-creature Triton is painted, riding on swimming beasts joined to a human body — you would not want to be such a thing. I am moving on difficult ground; for so great is the force of nature that no man wants to be like anything but a man — and indeed an ant like an ant.

\ciceroSection{79} But still, like what man? For how few are handsome: when I was at Athens, scarcely one in each band of young men could be found — I see what you smiled at, but the fact is so all the same. Then for us, who, with the indulgence of the ancient philosophers, take delight in the very young, even faults are often pleasing. A mole on a boy's knuckle delighted Alcaeus; yet a mole is a blemish of the body — to him, however, it seemed a point of light. Quintus Catulus, the father of my colleague and friend, loved your fellow townsman Roscius, on whom he even wrote these lines: "I had stopped, by chance saluting the rising Dawn, when suddenly on my left Roscius rises. With your leave, you heavenly powers, may I be permitted to say it: the mortal seemed more beautiful than the god." More beautiful, to him, than a god; and yet he was, as he is to this day, hopelessly cross-eyed. What does it matter, if this very thing seemed witty and charming to Catulus?

\ciceroSection{80} I come back to the gods. Do we suppose there are any of them who, if not quite squint-eyed, are at least a little cast in the eye? Are there any with a mole, any snub-nosed, flap-eared, beetle-browed, or big-headed, as some of us are — or is everything in them flawless? Grant you that: is there, then, also one single face for all of them? For if there are several, one must be more beautiful than another, and so some god is not the most beautiful; but if there is one face for all, then the Academy must flourish in heaven — for if there is no difference between god and god, then among the gods there is no recognition, no perception.

\ciceroSection{81} And what if that claim of yours, Velleius, is altogether false — that when we think of a god no form presents itself to us except a human one: will you even so go on defending so absurd a notion? To us, perhaps, it presents itself as you say; for from childhood we have known Jupiter, Juno, Minerva, Neptune, Vulcan, Apollo, and the rest of the gods with the very faces that painters and sculptors chose to give them — and not their faces only, but their adornment, their age, their dress. But it is not so for the Egyptians, nor the Syrians, nor for nearly the whole barbarian world; for you would find among them firmer convictions about certain beasts than there are among us about the holiest temples and images of the gods.

\ciceroSection{82} For we see many shrines plundered by our own people, and images of the gods carried off from the holiest places; but it has never so much as been heard of, even in report, that a crocodile or an ibis or a cat was ever wronged by an Egyptian. What then do you suppose — that the Egyptians do not regard Apis, that sacred bull of theirs, as a god? Just as surely, by Hercules, as you regard that Savior Juno of yours — whom you never see, even in your dreams, except with her goatskin, her spear, her little shield, and her turned-up little shoes. But the Argive Juno is not like that, nor the Roman. So the Argives have one form of Juno, the people of Lanuvium another. And indeed we have one Capitoline Jupiter, the Africans another in their Jupiter Hammon.

\ciceroSection{83} Is the natural philosopher, then — the observer and tracker of nature — not ashamed to seek the testimony of truth from minds steeped in mere habit? For at that rate one will be free to say that Jupiter is always bearded, Apollo always beardless, that Minerva's eyes are grey and Neptune's blue. And indeed we praise the Vulcan at Athens that Alcamenes made, in whom, standing and clothed, a slight lameness shows itself, not unbecoming: so we shall have a lame god, since this is what we have been told about Vulcan. Come now — do we also make the gods bear the very names by which we call them?

\ciceroSection{84} But, first, there are as many names for the gods as there are languages of men; for it is not as with you, Velleius, who are the same wherever you go — rather, Vulcan is one thing in Italy, another in Africa, another in Spain. Then again, the number of names is not great even in our own pontifical books, while the gods are beyond counting. Or are they nameless? You at least are bound to say so; for what point is there in many names, when there is but a single face? How much finer it would have been, Velleius, to confess that you did not know what you did not know, rather than to make yourself sick blurting out such stuff and to disgust even yourself. Do you really think that a god is like me, or like you? Surely you do not. "What then — am I to call the sun a god, or the moon, or the sky? Then it is also happy: enjoying what pleasures? And wise: how can there be wisdom in a thing of that sort, a mere trunk?"

\ciceroSection{85} These are your own arguments. If, then, a god has neither a human form, which I have refuted, nor some other such form, of which you are so persuaded, why do you hesitate to deny that the gods exist? You do not dare. Wisely so, to be sure — though here it is not the people you fear but the gods themselves. I have known Epicureans who worship every little statuette. And yet I see that some take Epicurus to have left the gods in words while abolishing them in fact, so as not to fall into offense with the Athenians. And so in those select and brief maxims of his, which you call \textit{[Greek: kuriai doxai]} (sovereign opinions), this, I think, is the first: "What is happy and immortal neither has trouble itself nor causes trouble to anyone." On this maxim, as it is laid out, there are those who hold that what he did out of plain clumsiness of expression he did on purpose.

\ciceroSection{86} They think too poorly of a man not the least bit cunning. For it is doubtful whether he is saying that something is happy and immortal, or that, if anything be such, it is of this kind. They do not notice that here he spoke ambiguously, but in many other places both he and Metrodorus spoke as openly as you did a little while ago. Epicurus, however, does believe the gods exist; nor have I ever seen anyone who feared more the very things he denied were to be feared — death, I mean, and the gods. By these the ordinary run of men are not so very greatly moved; yet he proclaims that the minds of all mortals are terrified by them. Thousands turn brigand though death stares them in the face; others ransack every shrine they can: I suppose the one set are driven by the fear of death, the other by the fear of religion. But since you do not dare — for now I shall speak with Epicurus himself — to deny that the gods exist, what is there to keep you from placing either the sun or the universe or some everlasting mind in the number of the gods?

\ciceroSection{87} "I have never seen," he says, "a soul endowed with reason and judgment in any figure but the human." Well, have you ever seen anything resembling the sun, or the moon, or the five wandering stars? The sun, bounding its motion by the two outermost points of a single orbit, completes its yearly course; the moon, kindled by the sun's own rays, fills out this same circuit in the space of a month; and the five planets, holding to that same orbit, some nearer the earth and some farther off, traverse the same distances from the same starting points in unequal stretches of time. Did you, Epicurus, ever see anything of the kind?

\ciceroSection{88} Let there be, then, no sun, no moon, no stars, since nothing can exist except what we have touched or seen. And the god himself — have you seen him? Why, then, do you believe he exists? On this principle let us do away with everything that either history or fresh reasoning brings to us. So it comes about that those who live inland do not believe the sea exists. What narrowness of mind is this — that if you had been born on Seriphos and had never left the island, where you had often seen little hares and little foxes, you would refuse to believe that lions and panthers exist, even when told what they were like, and if anyone spoke to you of an elephant, you would think you were even being laughed at!

\ciceroSection{89} And you indeed, Velleius, drew your conclusion not in your own school's manner but in that of the logicians — a race your school knows nothing of at bottom — by the narrow constraint of an argument. You assumed that the gods are happy: we grant it. But that no one can be happy without virtue: this too we concede, and gladly. That virtue cannot subsist without reason: this also must be agreed. You add that there is no reason except in human form. Who do you think will grant you that? For if it were so, what need was there for you to arrive there step by step? You might have assumed it as your right. But what is this "step by step"? From the happy to virtue, from virtue to reason, I see you came by degrees; but from reason to human form, how do you make the approach? That is to plunge headlong, not to descend.

\ciceroSection{90} Nor indeed do I understand why Epicurus preferred to say that the gods are like men rather than that men are like the gods. You will ask what the difference is; for if the one is like the other, the other is like the first. I see that; but my point is this — that the shape of the figure did not pass from men to the gods. For the gods always existed; they were never born, if indeed they are to be eternal; but men were born. The human form, then, existed before men, and the immortal gods had that form: so it is not their form that should be called human, but ours that should be called divine. Yet on this point let it be as you will; what I ask is what so great a piece of fortune was — for you hold that nothing in the natural world was made by reason —

\ciceroSection{91} but still, what was that mighty chance, whence so happy a concurrence of atoms, that all at once men were born in the form of gods? Are we to suppose that seeds of the gods fell from heaven onto the earth, and that men thus came forth resembling their fathers? I could wish you said so; I would not be unwilling to acknowledge kinship with the gods. But you say nothing of the kind — only that it happened by chance that we came to resemble the gods. And now arguments must be sought by which to refute this. Would that I could find the true as easily as I can convict the false. For you have enumerated, from memory and in full, so that I could not but marvel that such learning should exist in a Roman, the opinions of the philosophers on the nature of the gods, all the way from Thales of Miletus.

\ciceroSection{92} Did they all seem to you to be raving, those who decided that a god can exist without hands and feet? Does not even this give you pause, when you consider what use, what fitness, there is in a man's limbs, so that you might judge the gods to have no need of human limbs? For what need of feet, if there is no walking; what need of hands, if there is nothing to be grasped; what need of the whole other arrangement of all the parts of the body, in which there is nothing idle, nothing without a cause, nothing superfluous — so that no art can imitate the craftsmanship of nature? So a god will have a tongue and not speak; will have teeth, palate, throat to no use; and the parts that nature has attached to the body for the sake of procreation, these a god will have in vain; nor the external parts any more than the inner ones — heart, lungs, liver, and the rest — which, once you strip away their usefulness, what charm have they (since you would have these be in a god for the sake of beauty)?

\ciceroSection{93} Trusting in such dreams, not only did Epicurus and Metrodorus and Hermarchus speak against Pythagoras, Plato, and Empedocles, but even Leontium, a little courtesan, dared to write against Theophrastus — in elegant and Attic style, to be sure, but for all that: such was the license that the garden of Epicurus enjoyed. And you Epicureans are forever complaining; Zeno even went to law. What shall I say of Albucius? As for Phaedrus, nothing was more refined, nothing more humane; yet the old man would grow nettled if I had said anything rather sharp — even though Epicurus assailed Aristotle most insultingly, abused Phaedo the Socratic most shamefully, cut to pieces in whole volumes Timocrates, the brother of his own comrade Metrodorus, because he disagreed on some point or other in philosophy, was ungrateful to Democritus himself, whom he followed, and treated his teacher Nausiphanes, from whom he had learned not a little, very badly indeed. Zeno, for his part, did not merely pierce with abuse those who were then living — Apollodorus, Silus, and the rest — but said that Socrates himself, the very parent of philosophy, had been (to use his Latin word) an Attic buffoon, and never called Chrysippus anything but "Chrysippa."

\ciceroSection{94} You yourself, a little while ago, when you recited what was, as it were, a senate of philosophers, kept saying that the greatest men were dolts, ravers, madmen. But if none of these saw the truth about the nature of the gods, it is to be feared that there is no such nature at all. For those things that you people say are wholly made up, scarcely worth the midnight oil of little old women. For you do not perceive how much you must take on if you should win the point of having us concede that the form of men and of the gods is the same. The whole care and tending of the body that is bestowed on a man will have to be bestowed on a god: walking, running, reclining, leaning, sitting, grasping — and, in the end, conversation and speech as well.

\ciceroSection{95} As for your saying that there are gods both male and female, you see what follows. For my part, I cannot wonder enough whence that founder of yours arrived at such opinions. But you never cease to clamor that this must be held to — that a god is happy and immortal. Yet what stands in the way of his being happy if he is not two-footed? Or this thing — whether it is to be called \textit{beatitas} or \textit{beatitudo} (both quite harsh, but we must soften our words by use) — but whatever this state is, why can it not fall to that sun, or to this universe, or to some eternal mind that is void of figure and bodily limbs?

\ciceroSection{96} You say nothing except "I have never seen a happy sun or universe." But have you ever seen any universe besides this one? You will say no. Why, then, did you make so bold as to say that there are not six hundred thousand worlds but countless ones? "Reason taught me." Then will not reason teach you this too, when the most excellent nature is sought — a nature happy and eternal, which alone is divine — that just as we are surpassed by that nature in immortality, so we are surpassed in excellence of mind, and likewise of body as of mind? Why, then, when we are inferior in all other respects, are we equal in form? For human virtue came nearer to the likeness of the gods than human figure ever did.

\ciceroSection{97} Can anything more childish be said — to press the same point a little longer — than for us to declare that those kinds of beasts which are born in the Red Sea or in India do not exist? Yet not even the most inquisitive of men can, for all their searching, hear of as many things as there are which arise on land, in the sea, in marshes, in rivers. Are we to deny that these exist, because we have never seen them? And as for that very resemblance which delights you most — how utterly beside the point it is! Is not the dog like the wolf? And, as Ennius says, "how like to us is the ape, that most loathsome of beasts"! Yet in both cases the characters are unalike. No beast is shrewder than the elephant;

\ciceroSection{98} but in shape, what is more ungainly? I am speaking of beasts; but among men themselves, do we not find unalike characters in the most alike forms, and unlike shape in those of like character? For once we take up this kind of argument, Velleius, mark where it creeps. You were assuming that reason can dwell nowhere but in the human form; another will assume that it can dwell nowhere but in an earthly being, nowhere but in one that has been born, nowhere but in one that has grown up, nowhere but in one that has learned, nowhere but in one that consists of a mind and of a body that is perishable and frail — in short, nowhere but in a man, and a mortal one. But if you stand firm against all these conditions, why is it that one matter — shape — should throw you into confusion? For with all those conditions I have set out duly attached, you saw that there is reason and mind in man; strip them away, and yet you say you can know a god, provided only the lineaments remain. This is not to reflect on the matter, but to draw lots, so to speak, for what you are to say.

\ciceroSection{99} Unless, perhaps, you do not even attend to this: that, not in a man only but even in a tree, whatever is superfluous or has no use is an encumbrance. How troublesome it would be to have one finger more than the rest! And why so? Because neither in appearance nor in use do the five require another. But your god is redundant not in one finger only, but in head, neck, nape, sides, belly, back, hams, hands, feet, thighs, shins. If it is in order that he be immortal, what have these limbs to do with life — what has the very face? Those inner parts have far more to do with it: the brain, the heart, the lungs, the liver; for these are the dwelling-places of life. The cast of the countenance contributes nothing to the soundness of life.

\ciceroSection{100} And you were finding fault with those who, from works magnificent and splendid — when they had beheld the world itself, and its members the sky, the lands, the seas, and the glories of these, the sun, the moon, and the stars; and when they had come to know the seasons' ripenings and changes and successions — suspected that there is some excellent and surpassing nature which had wrought these things, and moved them, and ruled them, and governed them. Even if such men wander wide of the truth in their conjecture, I still see what they are following after. But you — what great and remarkable work do you have, that might be seen as wrought by a divine mind, from which you can infer that the gods exist? "I had," you say, "a certain conception of a god implanted in my mind." And of a bearded Jupiter, too, of a helmeted Minerva. Do you really suppose, then, that the gods are like that?

\ciceroSection{101} How much better is the conception of the unschooled crowd, who assign to a god not the limbs of a man only, but the use of those limbs as well! For they give him a bow, arrows, a spear, a shield, a trident, a thunderbolt; and if they cannot see what actions the gods perform, still they cannot conceive of a god who does nothing whatever. Even the Egyptians, who are laughed at, consecrated no beast except on account of some benefit they derived from it: the ibises, for instance, dispatch a great host of serpents — being birds tall in stature, with stiff legs and a long, horny beak; they avert a plague from Egypt when they kill and devour the winged snakes carried in by the African wind from the wastes of Libya, and so it comes about that those snakes do no harm by their bite while alive, nor by their stench once dead. I could speak of the usefulness of the ichneumons, of the crocodiles, of the cats, but I do not wish to go on at length. I will conclude, then, in this way: beasts have been consecrated by barbarians on account of the benefit they confer, whereas of your gods not only does no benefit exist, but no deed whatever.

\ciceroSection{102} "He has no business to transact," he says. Surely Epicurus, like a pampered child, thinks nothing better than idleness. Yet even children, when they are idle, take delight in some playful exercise. Are we to wish a god so much on holiday that he grows numb with idleness — so that, if he should but stir himself, we should fear he might be unable to be happy? This way of speaking not only strips the gods of motion and divine action, but even makes men sluggish, if indeed not even a god can be happy while doing something.

\ciceroSection{103} But let it be granted, as you would have it, that a god is the effigy and image of a man: what is his dwelling, what his seat, what his place; what, next, the action of his life; by what means is he happy, as you would have him be? For he who is to be happy must use and enjoy his own goods. And as for place, even those natures which are without souls have each its own proper place: so that the earth holds the lowest position, water floods over it, the air is given the region above, and to the fires of the aether is assigned the highest tract. As for living creatures, some are of the land, some are aquatic, and some, amphibious as it were, live in either element; there are even some that are thought to be born of fire, and that often appear fluttering about in blazing furnaces.

\ciceroSection{104} I ask, therefore, first, where your god dwells; next, what cause moves him from his place — if indeed he ever moves at all; then, since it is the property of living things to seek after something suited to their nature, what your god seeks after; further, toward what end he employs the motion of mind and reason; finally, in what manner he is happy, in what manner eternal. For whichever of these points you touch, it is a running sore: a theory so ill-founded can find no way out.

\ciceroSection{105} For you used to say that the appearance of a god is perceived by thought and not by sense; that it has no solidity, and does not remain the same in number; and that the seeing of him is such that he is discerned by resemblance and by passage, while there never fails a fresh supply of like images coming on from the infinite store of bodies; and that from this it comes about that our mind, intent upon these images, judges that nature to be happy and everlasting. This — by the very gods of whom we are speaking — what in the world is it? For if these images have force only toward thought, and have no solidity, no projection, what difference is there whether we think of a hippocentaur or of a god? Every such conformation of the mind the other philosophers call an empty motion; but you call it the arrival and entry of images into the mind.

\ciceroSection{106} Just as, then, when I seem to see Tiberius Gracchus haranguing the assembly on the Capitol and bringing down the voting-urn against Marcus Octavius, I call that an empty motion of the mind, you on the other hand say that the images of Gracchus and Octavius persist, which, when they have reached the Capitol, are then carried over to my mind — and that this same thing happens in the case of a god, whose frequent appearance strikes our minds, from which the gods are understood to be happy and eternal.

\ciceroSection{107} Grant that there are images by which our minds are struck: still, no more than a certain appearance is presented to us. Is there any reason, too, why that appearance should be happy, why eternal? But these images of yours — what are they, and whence? This license comes entirely from Democritus; but he too has been censured by many, nor do you find any way out, and the whole business wobbles and limps. For what could be less acceptable than that the images of all men should fall upon me — of Homer, Archilochus, Romulus, Numa, Pythagoras, Plato — and not in the form those men actually wore? In what form, then? And whose images are they? Aristotle teaches that the poet Orpheus never existed, and the Pythagoreans report that this Orphic poem was the work of a certain Cercon; yet Orpheus — that is, his image, as you would have it — often rushes into my mind.

\ciceroSection{108} What of the fact that different images of the same man come into my mind and into yours; what of the fact that images come of things that never existed at all and never could have, such as Scylla, such as the Chimaera; what of the fact that images come of men, places, cities which we have never seen; what of the fact that, the very moment it suits me, an image is at hand; what of the fact that they come even to a sleeper, unsummoned? The whole business, Velleius, is a piece of nonsense. Yet you stamp these images not only upon our eyes but upon our minds as well: such is the impunity of your prattling. And with what license!

\ciceroSection{109} "There frequently occurs a passage of streaming visions, so that out of many a single one is seen." I should be ashamed to say that I do not understand, if you who defend these doctrines understood them yourselves. For how do you prove that the images are borne along continuously — or, if continuously, how eternally? "The innumerability of the atoms," he says, "supplies the want." Will that same innumerability, then, also bring it about that all things are everlasting? You take refuge in equilibrium — for so let us render \textit{[Greek: isonomia]}, if you please — and you say that, since there is a mortal nature, there must needs be an immortal one as well. By that reasoning, since men are mortal, let there be some men immortal; and since they are born on land, let them be born in the water too. "And because there are things that destroy, let there be things that preserve." Let there be, by all means; but let them preserve the things that exist. I do not perceive that those gods of yours exist.

\ciceroSection{110} But how does that whole effigy of things arise out of indivisible bodies? Even if such bodies existed — which they do not — they might perhaps be able to push themselves along and be agitated among themselves by collision; but to form, to shape, to color, to animate, they could not. In no way, therefore, do you make your god immortal. Let us now consider his happiness. Without virtue it can in no way be had; but virtue is active; and your god does nothing; he is therefore devoid of virtue;

\ciceroSection{111} and so he is not even happy. What life, then, does he have? "A continual supply," you say, "of good things, with no intrusion of evils." Of what good things, pray? Of pleasures, I suppose — pleasures that pertain to the body, of course; for you have come to know no pleasure of the mind except such as proceeds from the body and returns to the body. I do not think you wish to be like the rest of the Epicureans, who are ashamed of certain sayings of Epicurus, in which he testifies that he cannot even understand any good that is severed from the pleasures of indulgence and lust — pleasures which, without blushing, he pursues every one by name.

\ciceroSection{112} What food, then, or what drinks, or what varieties of sounds or of flowers, or what touches, what scents, will you furnish to the gods, that you may drench them in pleasures? The poets, indeed, prepare nectar, ambrosia, banquets, and either Youth or Ganymede to serve the cup; but you, Epicurus — what will you do? For I do not see whence your god is to have these things, nor how he is to use them. The nature of man, then, is better furnished for living happily than that of the gods, since it enjoys more kinds of pleasures.

\ciceroSection{113} "But you mean those pleasures of a lower order, the ones to which a kind of titillation is applied to the senses" — for the word is Epicurus's own. How long will you go on playing? Why, even our friend Philo could not bear to hear the Epicureans disdain soft and dainty pleasures. For with a prodigious memory he used to recite very many of Epicurus's pronouncements in the very words in which they had been written; and of Metrodorus, who is Epicurus's partner in wisdom, he would quote much that was more shameless still. For Metrodorus accuses his own brother Timocrates of hesitating to measure everything that pertains to the happy life by the belly — and he does not say it once, but again and again. I see you nodding, for these things are familiar to you; I would produce the books if you denied it. And I do not now reproach the doctrine that everything is referred to pleasure — that is another question — but I am demonstrating that your gods are devoid of pleasure, and therefore, by your own judgment, not even happy. "But they are free of pain."

\ciceroSection{114} Is that enough for that life so brimming with goods, that life of supreme happiness? "He is forever thinking," they say, "that he is happy; for he has nothing else to turn over in his mind." Grasp it, then, in your imagination and set before your eyes a god who through all eternity thinks of nothing but "all is well with me" and "I am happy." And yet I do not see how this happy god of yours can fail to fear that he will perish, since without any intermission he is battered and harried by the everlasting bombardment of atoms, and since images are forever streaming off from him. So your god is neither happy nor eternal.

\ciceroSection{115} "But Epicurus actually wrote books on holiness and on piety toward the gods." Yes — and in what manner does he speak in them? In such a way that you would say you were listening to Tiberius Coruncanius or Publius Scaevola, the chief pontiffs, and not to the man who tore up all religion from its very roots, who overturned the temples and altars of the immortal gods — not, like Xerxes, with his hands, but with his arguments. For what reason can there be why you should say the gods ought to be worshipped by men, when the gods not only do not worship men but take no care of anything at all, do nothing at all?

\ciceroSection{116} "But theirs is a certain surpassing and pre-eminent nature, such that of itself, by its own power, it ought to draw the wise man to its worship." Now can there be anything surpassing in a nature that, rejoicing in its own pleasure, neither will ever do anything, nor does anything, nor has ever done anything? Moreover, what piety is owed to one from whom you have received nothing? Or how can anything at all be owed to one who has done you no service? For piety is justice toward the gods; and what claim of right can there be between us and them, when there is no fellowship between man and god? Holiness, again, is the knowledge of how the gods are to be worshipped; but why they should be worshipped I cannot understand, when no good has been received from them or is hoped for.

\ciceroSection{117} And what reason is there why we should revere the gods out of admiration for a nature in which we see nothing remarkable? As for superstition, which you are accustomed to boast of being free from — that is easy to be rid of, once you have done away with all the power of the gods. Unless perhaps you suppose that Diagoras or Theodorus, who flatly denied that the gods exist, could have been superstitious; for my part, I would not say even Protagoras could, who was allowed to affirm neither — neither that the gods exist nor that they do not. For the views of all these men do away not only with superstition, in which there is an empty fear of the gods, but with religion as well, which is contained in the pious worship of the gods.

\ciceroSection{118} What of those who said that the entire belief in the immortal gods was invented by wise men for the sake of the commonwealth, so that religion might lead to their duty those whom reason could not lead — did they not tear up all religion from its very roots? What of Prodicus of Ceos, who said that the things which benefit human life have been reckoned in the number of the gods — what religion, in the end, did he leave standing?

\ciceroSection{119} What of those who hand down that brave or famous or powerful men have, after death, made their way to the gods, and that these very men are the ones we are accustomed to worship and entreat and revere — are they not strangers to all religion? This is the line of reasoning that was most fully developed by Euhemerus, whom our own Ennius interpreted and followed beyond all others; and Euhemerus sets out the deaths and burials of the gods. Does he, then, seem to have confirmed religion, or rather to have utterly abolished the whole of it? I pass over Eleusis, that holy and august place, "where the nations dwelling at the world's farthest shores are initiated"; I leave aside Samothrace and the rites that on Lemnos "are worshipped in secret by night's approach, ringed close with woodland hedges" — rites which, once unfolded and brought back to a rational account, teach us more about the nature of things than about the nature of the gods.

\ciceroSection{120}

\ciceroSection{121} To me, indeed, even Democritus — a great man, and among the very first — from whose springs Epicurus watered his little gardens, seems to waver on the nature of the gods. For at one moment he holds that there exist, within the totality of things, images endowed with divinity; at another he says that the principles of mind which are in that same universe are gods; at another, that they are living images, which are wont either to do us good or to do us harm; and at another, certain enormous images, so vast that they embrace the whole world from without. All of which is more worthy of Democritus's homeland than of Democritus himself; for who can grasp those images in his mind, who admire them, who judge them worthy of worship or of religious devotion? Epicurus, however, tore religion up by the roots from the minds of men when he abolished both aid and favor in the immortal gods. For while he says that the divine nature is the best and most pre-eminent, he denies at the same time that there is any favor in a god — and so he removes the very thing most proper to the best and most pre-eminent nature. For what is better, what more pre-eminent, than goodness and beneficence? Yet when you would have a god lack these, you would have no one — neither god nor man — dear to a god, no one loved by him, no one cherished. And so it comes about that not only are men neglected by the gods, but the gods themselves neglect one another, each indifferent to the rest. How much better the Stoics, whom you reproach! They hold that the wise are friends to the wise even when they are unknown to one another; for nothing is more lovable than virtue, and the man who has attained it will be cherished by us wherever in the world he may be.

\ciceroSection{122} But what harm you do, when you place gracious favor and goodwill in weakness! For — to say nothing of the power and nature of the gods — do you really hold that men too would not have been kind and generous unless they were weak? Is there no natural affection between the good? The very word "dear" belongs to love, from which the name of friendship is derived; and if we shall refer friendship to our own profit, and not to the advantage of the one we love, it will be no friendship but a kind of trading in self-interest. Meadows and fields and herds of cattle are loved in that fashion, because profit is taken from them; the affection and friendship of men is freely given. How much more, then, the affection of the gods, who, in need of nothing, both love one another and look after men. And if this were not so, why do we revere the gods, why pray to them, why do pontiffs preside over the sacred rites and augurs over the auspices, what do we ask of the immortal gods, what do we vow? "But there is even a book of Epicurus on holiness."

\ciceroSection{123} We are toyed with by a man not so much witty as free in the license of his writing. For what holiness can there be if the gods take no care of human affairs? And what is a living nature that cares for nothing? Surely, then, that is nearer the truth which Posidonius, the close friend of us all, set out in the fifth book of his work on the nature of the gods: that in Epicurus's view there are no gods at all, and that what he said about the immortal gods he said to ward off the odium of the thing. For he could not have been so foolish as to fashion a god in the likeness of a paltry little man — endowed only with the barest outlines, not with any solid frame, possessed of all a man's limbs but with not even the slightest use of those limbs, a thin and translucent thing, granting nothing to anyone, doing no one a favor, caring for absolutely nothing, doing nothing at all. Such a nature, in the first place, cannot exist; and seeing this, Epicurus in fact abolishes the gods while in words he keeps them.

\ciceroSection{124} Then again, if a god is in the highest degree such that he is held by no favor and no love of men — farewell to him; for why should I say "may he be gracious"? He can be gracious to no one, since, as you say, all favor and all love reside in weakness.

\ciceroBook{2}

\ciceroSection{1} When Cotta had finished, Velleius said: "What an incautious man I was, to attempt a contest with an Academic who is at the same time a rhetorician. An Academic without eloquence I should not have feared, nor a rhetorician, however eloquent, without that philosophy of his; for I am not thrown into confusion by a flood of empty words, nor by subtlety of thought when the speech is dry. But you, Cotta, prevailed in both respects; only the audience and a panel of judges were wanting to you. Still, of that another time. For now, if it suits him, let us hear Lucilius."

\ciceroSection{2} Then Balbus said: "For my part I should rather hear Cotta again — provided that, with the same eloquence by which he abolished the false gods, he now brings in the true ones. For it befits a philosopher, a pontiff, and a Cotta to hold concerning the immortal gods an opinion not wandering and unsettled, as the Academics do, but firm and certain, as ours is. Against Epicurus enough and more than enough has been said; but I am eager to hear, Cotta, what you yourself think." "Have you forgotten," he said, "what I declared at the outset — that I can more easily say, especially on such matters, what I do not think than what I do think? And even if I had something clear to offer, I should still want to hear you in my turn, now that I myself have said so much."

\ciceroSection{3} Then Balbus said: "I will indulge you, and I will be as brief as I can; for now that the errors of Epicurus stand refuted, much of the length has been taken out of my discourse. Our school divides this whole question of the immortal gods into four parts. First they teach that the gods exist; next, what their nature is; then, that the world is governed by them; and last, that they have a care for human affairs. In this present conversation, however, let us take up the first two; the third and fourth, since they are larger, I think should be put off to another time." "By no means," said Cotta; "for we are at leisure, and we are treating of matters that ought to be set even before our business."

\ciceroSection{4} Then Lucilius said: "The first part scarcely seems to need a speech at all. For what can be so plain and so manifest, when we have looked up at the heavens and contemplated the things of heaven, as that there is some divine power of a most surpassing mind by which these are governed? And if it were not so, how could Ennius, with the assent of all, have said: \textit{Behold this brightness shining on high, whom all men call upon as Jupiter} — yes, and call him both Jupiter and ruler of the universe, governing all things by his motion, and, as that same Ennius has it, \textit{father of gods and of men}, a god present and all-powerful? And the man who doubts this — I really do not see why he could not equally doubt whether the sun exists or does not exist;

\ciceroSection{5} for how is the one more evident than the other? And unless we held this fixed and grasped in our minds, no opinion so stable would persist, nor would it be confirmed by the length of time, nor could it have grown old together with the generations and ages of men. Indeed we see that other opinions, fabricated and empty, have wasted away with the passage of time. For who supposes that a hippocentaur ever existed, or a Chimaera? And what old woman so witless can be found as to dread those monsters of the underworld that once were believed in? Time blots out the fictions of opinion, but confirms the judgments of nature.

\ciceroSection{6} And so, both among our own people and among the rest, the worship of the gods and the sanctities of religion grow greater and better day by day. This happens neither at random nor by chance, but because the gods, often present, declare their power. So at Lake Regillus, in the Latin war, when the dictator Aulus Postumius was joined in battle with Octavius Mamilius of Tusculum, Castor and Pollux were seen fighting on horseback in our ranks; and within more recent memory those same sons of Tyndareus announced the defeat of Perseus. For Publius Vatinius, grandfather of this young man, while coming to Rome by night from the prefecture of Reate, was told by two young men on white horses that King Perseus had been captured that very day; and when he reported this to the Senate, he was at first thrown into prison, as one who had spoken rashly about public affairs — but afterward, when dispatches arrived from Paulus and the same day was found to agree, he was granted both land and exemption from service by the Senate. And again, when the Locrians had overwhelmed the men of Croton in a vast battle at the river Sagra, it is recorded that on that very day the news of the fight was heard at the games of Olympia. Often the voices of the Fauns have been heard, often the shapes of the gods have been seen, compelling anyone not dull-witted or impious to confess that the gods are present.

\ciceroSection{7} As for predictions and forebodings of things to come, what else do they show but that to men are displayed, indicated, portended, and foretold the things that are to be — from which those very things are called displays, indications, portents, and prodigies? But if we believe these to be fictions of unbridled fable — Mopsus, Tiresias, Amphiaraus, Calchas, Helenus (whom, however, not even the fables would have enrolled as augurs, had they utterly repudiated the matter) — shall we, instructed not even by examples from our own home, refuse to acknowledge the divine power of the gods? Will the recklessness of Publius Clodius in the First Punic War move us not at all — who, even mocking the gods in jest, when the chickens released from their coop would not feed, ordered them plunged into the water to drink, since they would not eat? That jest, when the fleet was destroyed, brought many tears upon himself and a great disaster upon the Roman people. And did not his colleague Lucius Junius, in the same war, lose his fleet in a storm when he had not obeyed the auspices? And so Clodius was condemned by the people, and Junius took his own life.

\ciceroSection{8} Coelius writes that Gaius Flaminius, having neglected religion, fell at Lake Trasimene with great injury to the state. From the destruction of such men it may be understood that the state was enlarged under the commands of those who had obeyed religious observance. And if we are willing to compare what is ours with what is foreign, in other respects we shall be found equal or even inferior, but in religion — that is, in the worship of the gods — far superior.

\ciceroSection{9} Or is the famous augural staff of Attus Navius — with which, to track down a sow, he marked off the regions of his vineyard — a thing to be despised? I should believe so, were it not that under his augury King Hostilius waged the greatest wars. But through the negligence of the nobility the discipline of augury has been let go, the truth of the auspices has been scorned, and only the outward show retained; and so the greatest parts of public business, among them the wars on which the safety of the state depends, are conducted with no auspices at all, no observation of the crossing of rivers, no auspices from the points of spears, no men summoned to arms — from which it has come about that wills made under arms are no more; for our generals begin to wage war only when they have laid the auspices aside.

\ciceroSection{10} But among our ancestors the force of religion was so great that certain commanders even devoted themselves, their heads veiled, in set words, to the immortal gods on behalf of the state. I could recount many things from the Sibylline prophecies, many from the responses of the haruspices, by which those matters are confirmed that ought to be in doubt to no one. And indeed the discipline both of our augurs and of the Etruscan haruspices was proved by the event itself, in the consulship of Publius Scipio and Gaius Figulus. When Tiberius Gracchus, consul for the second time, was holding the election of these men, the first returning officer, just as he announced them, dropped dead on the spot. Gracchus nonetheless completed the assembly, but, perceiving that the affair had brought the matter into a religious scruple among the people, referred it to the Senate. The Senate decreed that it should be referred to those to whom it was customary. The haruspices were brought in and replied that the returning officer of the election had not been lawful. Then Gracchus, as I used to hear from my father, blazing with anger:

\ciceroSection{11} "Was I, then, not lawful — I, who both as consul and as augur held the election after taking the auspices? But do you Etruscans, you barbarians, hold the law of the auspices of the Roman people, and can you be the interpreters of our assemblies?" And so on that occasion he ordered them to depart. But afterward he sent a letter from his province to the college, saying that, in reading the books, he had recalled that there had been a flaw in his taking the augural tent — Scipio's gardens — because, when he had later entered the pomerium for the sake of holding a meeting of the Senate, he had forgotten on his return, in crossing that same pomerium, to take the auspices; and that the consuls had therefore been elected with a flaw. The augurs referred the matter to the Senate; the Senate ordered the consuls to abdicate; and they abdicated. What greater examples do we ask? A man of the highest wisdom — and perhaps the most distinguished of all men — chose to confess his own fault, which might have been concealed, rather than that a religious flaw should cling to the state; and the consuls chose to lay down their supreme command at once rather than to hold it for one instant against religion.

\ciceroSection{12} Great is the authority of the augurs; and is not the art of the haruspices divine? Whoever sees these things, and countless others of the same kind, must he not be compelled to confess that the gods exist? For it is surely necessary that those exist whose interpreters certain men are; and the gods are what they interpret; let us confess, then, that the gods exist. But perhaps not everything turns out as it was foretold. Neither, because not all the sick recover, is medicine therefore no art. Signs of things to come are shown by the gods; if any have erred in reading them, it is not the nature of the gods but the conjecture of men that has gone wrong. And so among all peoples there stands one supreme agreement; for it is inborn in all, and as it were engraved upon the mind, that the gods exist. What their nature is, men differ; that they exist, no one denies.

\ciceroSection{13} Our own Cleanthes, indeed, said that conceptions of the gods are formed in the minds of men from four causes. The first he set down as the one I spoke of just now, which arose from the foreboding of things to come; the second, the one we have derived from the magnitude of the benefits we receive through the temperateness of the climate, the fruitfulness of the earth, and the abundance of many other advantages;

\ciceroSection{14} the third, which strikes terror into our minds through thunderbolts, storms, downpours, snows, hailstorms, devastation, pestilence, earthquakes — often through rumblings, and showers of stones, and raindrops as though of blood — then through landslides or sudden chasms in the earth, then through portents of men and beasts contrary to nature, then through torches seen in the sky, then through those stars which the Greeks call \textit{[Greek: kometas]} and our people the long-haired ones, which lately, in the Octavian war, were the heralds of great calamities; then through the doubling of the sun, which, as I heard from my father, had occurred in the consulship of Tuditanus and Aquilius — in that very year in which Publius Africanus, that second sun, was extinguished; and by these things men, terror-struck, came to suspect that there is a certain heavenly and divine power;

\ciceroSection{15} the fourth cause — and that the greatest — is the evenness of the heaven's motion and the unfailing constancy of its revolution, the distinctness, the usefulness, the beauty, the order of the sun, the moon, and all the stars; the very sight of which would be enough to show that these things are not the work of chance. So that, if a man enters some house, or a gymnasium, or a forum, and sees in all things a system, a measure, a discipline, he cannot judge that they came to be without a cause, but understands that there is someone who presides and to whom obedience is given — how much more, amid such great motions and such great alternations, amid the ordered arrangements of so many and so great things, in which the boundless and infinite ages of time have never once played false, must he conclude that motions of nature so vast are governed by some mind.

\ciceroSection{16} Chrysippus, indeed, though he is of the keenest intelligence, yet speaks in such a way that he seems to have learned these things from nature herself rather than to have discovered them himself. "For if," he says, "there is anything in the nature of things that the mind of man, that reason, that human power and capacity cannot bring about, then surely that which brings it about is better than man; but the things of heaven, and all those whose order is everlasting, cannot be wrought by man; therefore that by which they are wrought is better than man. And what would you sooner call this than god? For if the gods do not exist, what can there be in the nature of things better than man? In him alone is reason, than which nothing can be more excellent; but for a man to think that in all the world there is nothing better than himself is the mark of senseless arrogance; therefore there is something better. Therefore god surely exists."

\ciceroSection{17} ``Or consider: if you saw a great and beautiful house, you could not be brought to believe — even if you did not see its master — that it had been built for mice and weasels. So if you should suppose that the great ornament of the world, the great variety and beauty of the heavenly bodies, the great force and vastness of sea and land, were your dwelling and not the dwelling of the immortal gods, would you not plainly seem to be out of your senses? Or do we not even understand this much: that all things above are better, and that the earth is lowest, surrounded by the densest air — so that for this very reason (as we see happen also in certain regions and cities, where men's wits are duller on account of a denser quality of the air) the same thing has befallen the human race, in that it is settled upon the earth, that is, in the densest region of the world.

\ciceroSection{18} And yet from the very ingenuity of men we ought to conclude that there exists some mind, and that a keener and a divine one. For from where did man `lay hold of' this — as Socrates asks in Xenophon? Indeed, if anyone should ask from where we get the moisture and the heat that is diffused throughout the body, and the very earthy solidity of our flesh, and finally that breathing spirit of the soul, it is plain: one part we have taken from earth, another from moisture, another from fire, another from that air which we call breath. But that which surpasses all of these — I mean reason, or, if you prefer more words, mind, deliberation, thought, prudence — where did we find it, from where did we draw it? Shall the world possess all other things and not possess this one thing, which is of the greatest worth? And yet surely there is nothing better than the world among all things, nothing more excellent, nothing more beautiful; and not only is there nothing better, but nothing better can even be conceived. And if there is nothing better than reason and wisdom, these must necessarily be present in that which we concede to be best.

\ciceroSection{19} And what of this great kinship of things — concordant, conspiring, continuous — whom would it not compel to approve what I am saying? Could the earth flower at one season and then in its turn lie barren; could the sun's approaches and departures be known at the solstices and the winter turnings while the things themselves so change; could the ocean tides and the narrows of the straits be set in motion by the rising or setting of the moon; could the unequal courses of the stars be preserved by the single revolution of the whole heaven? These things, with all the parts of the world chiming together, surely could not happen so unless they were held together by a single divine and continuous breath.

\ciceroSection{20} And when these matters are argued more fully and more amply, as I intend to do, they more easily escape the cavils of the Academics; but when they are reasoned out more briefly and more tightly, as Zeno used to do, then they are more exposed to attack. For just as a flowing river is corrupted scarcely or not at all, while enclosed water is corrupted easily, so by a flood of speech the abuse of a critic is washed away, whereas reasoning hemmed into a tight compass does not readily defend itself. These arguments, which I am spreading out, Zeno used to compress in this way:

\ciceroSection{21} ``That which uses reason is better than that which does not use reason; but nothing is better than the world; therefore the world uses reason.'' In like manner it can be made out that the world is wise, in like manner that it is happy, in like manner that it is eternal; for all these are better than the things that lack them, and nothing is better than the world. From which it will follow that the world is god.

\ciceroSection{22} And the same man argued thus: ``No part of anything that lacks sensation can be sentient; but the parts of the world are sentient; therefore the world does not lack sensation.'' He goes on and presses harder, more tightly: ``Nothing,'' he says, ``that is devoid of mind and reason can generate out of itself a thing that is animate and possessed of reason; but the world generates things that are animate and possessed of reason; therefore the world is animate and possessed of reason.'' And the same man, by a comparison, as he often does, brought his reasoning to a conclusion in this way: ``If pipes playing tunefully were born from an olive tree, you would surely not doubt that there was in the olive some knowledge of the flute? And what if plane trees bore little lyres sounding in measure — you would of course judge that music was present in the plane trees too. Why then should the world not be judged animate and wise, when out of itself it brings forth things animate and wise?''

\ciceroSection{23} But since I have begun to proceed otherwise than I had said at the outset — for I had denied that this first part needed argument, since it was clear to all that the gods exist — I nevertheless wish even this very point to be confirmed by physical, that is, natural, reasonings. For the case stands thus: all things that are nourished and that grow contain within themselves a force of heat, without which they could neither be nourished nor grow. For everything that is hot and fiery is roused and driven by its own motion; but that which is nourished and grows employs a certain fixed and steady motion; and as long as this motion remains in us, so long do sensation and life remain, whereas once the heat is cooled and extinguished, we ourselves die and are snuffed out.

\ciceroSection{24} This is what Cleanthes teaches by these further arguments, to show how great a force of heat resides in every body: he denies that there is any food so heavy that it is not digested over a night and a day; and even in its residues there remains heat, in those very parts that nature has rejected. Indeed the veins and arteries do not cease to throb, as if with a certain fiery motion; and it has often been observed that when the heart of some living creature has been torn out, it beats so rapidly that it imitates the swiftness of fire. Everything, therefore, that lives — whether animal or sprung from the earth — lives on account of the heat enclosed within it. From which we ought to understand that this nature of heat possesses within itself a vital force pervading the whole world.

\ciceroSection{25} And we shall discern this more easily when this whole fiery kind, which permeates all things, is explained more closely. All the parts of the world, then (I shall touch on the largest), are upheld and sustained by heat. This can be perceived first in the earthy element. For we see fire struck out by the clashing and rubbing of stones; we see freshly turned earth steam with warmth; and we see warm water drawn from perennial wells, and this happening most of all in the winter season, because a great force of heat is held within the hollows of the earth, and the earth is in winter denser and for that reason holds the heat innate within it more tightly.

\ciceroSection{26} The discourse would be long, and the proofs many, by which it could be shown that all the seeds the earth conceives, and all the things generated of itself which it holds fixed by their roots, both arise and grow by this tempering of heat. And that heat is mingled with water also is declared first by the very fluidity of water and its flowing — which would neither freeze in the cold nor congeal into snow and frost, did not that same water, made liquid and dissolved by mingled heat, spread and pour itself abroad; and so by the north winds and the other added colds the moisture hardens, and the same in turn is softened when warmed and melts away by heat. Indeed the seas too, when stirred by winds, grow so warm that one may easily understand that heat is enclosed within those vast bodies of water; for that warmth must not be reckoned external and brought in from outside, but roused up by agitation from the innermost parts of the sea — which happens to our own bodies as well, when they grow warm again with motion and exercise. And the air itself, which is by nature most cold, is by no means devoid of heat;

\ciceroSection{27} indeed it is mingled with much heat, and that abundantly: for it arises from the exhalation of waters; the air must be regarded as a kind of vapor from them, and this comes into being by the motion of that heat which is contained in the waters — a likeness we can discern in those waters that boil up when fires are set beneath them. And now the remaining fourth part of the world: this too is in its whole nature fiery, and it imparts to all the other natures a wholesome and vital heat.

\ciceroSection{28} From which it is concluded that, since all the parts of the world are sustained by heat, the world itself too is preserved through so long a span by a like and equal nature — and the more so because we ought to understand that the hot and fiery element is so diffused throughout all nature that there resides in it a power of procreation and a cause of begetting, by which both all living creatures and those things whose roots the earth holds must of necessity be born and grow.

\ciceroSection{29} It is nature, therefore, that holds the whole world together and watches over it, and that nature not without sensation and reason. For every nature that is not solitary or simple, but joined and connected with another, must of necessity have within itself some governing principle — as in man the mind, in the beast something resembling mind, from which the appetites for things arise; while in trees and in those things that are brought forth from the earth the governing principle is thought to reside in the roots. By governing principle I mean what the Greeks call \textit{[Greek: hēgemonikon]}, than which there neither can nor ought to be anything more excellent in any kind. Thus it follows of necessity that that in which the governing principle of all nature resides must be the best of all things, and most worthy of authority and dominion over all things.

\ciceroSection{30} Now we see that in the parts of the world (for there is nothing in the whole world that is not a part of the universe) there is sensation and reason. In that part, therefore, in which the governing principle of the world resides, these must necessarily be present, and indeed keener and greater. For which reason the world must necessarily be wise, and the nature that embraces and holds all things together must excel in the perfection of reason; and therefore the world is god, and the whole force of the world is held together by a divine nature. And moreover that fire of the world is far purer, far clearer, far more mobile, and for these reasons far better fitted to stir the senses than this heat of ours, by which the things known to us are sustained and kept alive.

\ciceroSection{31} It is absurd, therefore, to say that, while men and beasts are held together by this heat and on that account move and feel, the world is without sensation, when it is held together by an ardor whole and free and pure, and at the same time keenest and most mobile — especially since that ardor which belongs to the world is moved not by being driven from elsewhere nor by any external blow, but of itself and of its own accord; for what can there be mightier than the world, to drive and move that heat by which it is held together?

\ciceroSection{32} For let us hear Plato, as a kind of god among philosophers; who holds that there are two motions, one its own and the other external, and that what is moved of itself and by its own accord is more divine than what is set going by an alien impulse. This motion, moreover, he places in souls alone, and judges that from them the beginning of motion has been derived. Wherefore, since all motion arises from the ardor of the world, and that ardor is moved not by an alien impulse but of its own accord, it must of necessity be soul; from which it follows that the world is animate. And from this too it will be possible to understand that intelligence resides in it, since the world is certainly better than any other nature. For just as there is no part of our body that is not of less worth than we ourselves are, so the whole universe must necessarily be of more worth than any one part of the universe. And if this is so, the world must necessarily be wise. For if it were not so, man, who is a part of the world, would, since he shares in reason, have to be of more worth than the whole world.

\ciceroSection{33} And further, if we wish to proceed from the first and rudimentary natures to the last and perfected ones, we must necessarily arrive at the nature of the gods. For the first thing we observe is that the things which are born from the earth are sustained by nature, to which nature has granted nothing further than to protect them by feeding and increasing them.

\ciceroSection{34} To the beasts, on the other hand, she gave sensation and motion, and, together with a kind of appetite, an approach toward what is wholesome and a withdrawal from what is harmful. To man she added something more: she gave him reason, by which the appetites of the mind might be governed, now slackened, now held in check. There is, moreover, a fourth and highest grade, belonging to those who are born by nature good and wise, in whom from the beginning is implanted a reason right and constant, which must be reckoned above man and assigned to a god — that is, to the world, in which there must necessarily reside that perfected and consummate reason.

\ciceroSection{35} For it cannot be said that in any established order of things there is nothing supreme and perfected. For just as in a vine, just as in a beast, we see nature — unless some force has stood in her way — make her way by a path of her own to the limit, and just as painting and architecture and the rest of the arts have a certain finished perfection of the completed work, so in all nature, and far more so, something must necessarily be brought to completion and perfected. And indeed, while many external things can stand in the way of other natures and prevent their being perfected, nothing can hinder universal nature, precisely because she herself confines and contains all natures. For this reason there must necessarily be that fourth and highest grade, to which no force can gain access.

\ciceroSection{36} And this is the grade in which the nature of all things is placed; and since that nature is such that it both presides over all things and can be hindered by nothing, the world must necessarily be intelligent, and indeed wise as well. For what is more foolish than to deny that the nature which has embraced all things is the best — or, granting that it is the best, to deny that it is, first, animate, then possessed of reason and counsel, and finally wise? For how can it be best on any other terms? For if it were like the plants or even like the beasts, it would have to be reckoned the worst rather than the best. And again, if it shared in reason and yet were not wise from the beginning, the world's condition would be worse rather than better than man's. For a man can become wise, but the world, if through the whole infinite span of past time it has been without wisdom, will surely never attain wisdom; and so it will be worse than man. Since this is absurd, the world must be held both wise from the beginning and a god.

\ciceroSection{37} For there is nothing whatever besides the world to which nothing is lacking, and which is so fitted together on every side, so perfect and so complete in all its measures and parts. For Chrysippus aptly observes that, just as a sheath exists for the sake of a shield and a scabbard for the sake of a sword, so all other things besides the world have been brought into being for the sake of others: the crops and fruits which the earth produces for the sake of living creatures, living creatures in turn for the sake of men — the horse for carrying, the ox for plowing, the dog for hunting and keeping guard. Man himself, however, came into being to contemplate and imitate the world — in no way perfect, but a certain small portion of the perfect.

\ciceroSection{38} But the world, since it has embraced all things and there is nothing that is not contained within it, is perfect on every side. How then can that which is best be lacking to it? But nothing is better than mind and reason; therefore these cannot be lacking to the world. Rightly, then, does that same Chrysippus, adducing comparisons, teach that all things are better in their perfect and mature state than otherwise — as in a horse than in a colt, in a dog than in a puppy, in a man than in a boy; and likewise that what is best in the whole world must reside in something perfect and consummate.

\ciceroSection{39} But there is nothing more perfect than the world, nothing better than virtue; therefore virtue is proper to the world. Yet man's nature is not perfect, and even so virtue is brought to fulfillment in man; how much more readily, then, in the world. Virtue is therefore in it. The world is therefore wise, and for that reason a god. And once the divinity of the world has been thus discerned, the same divinity must be assigned to the stars, which are produced from the most mobile and purest part of the aether, with no other nature mingled in besides, and are wholly hot and translucent — so that they too may most rightly be said to be living beings, and to have sensation and understanding.

\ciceroSection{40} And that they are wholly fiery is confirmed, in Cleanthes' view, by the testimony of two senses, touch and sight. For the heat and brilliance of the sun is more dazzling than that of any fire, since it shines so far and wide across the immense world; and its touch is such that it not only warms but often even scorches — neither of which it would do were it not fiery. "Therefore," he says, "since the sun is fiery and is nourished by the waters of the Ocean" (because no fire can endure without some sustenance), "it must necessarily be like either the fire we apply to our use and sustenance, or the fire that is contained within the bodies of living creatures.

\ciceroSection{41} Now this fire of ours, which the needs of life require, is the destroyer and consumer of all things, and wherever it has broken in it scatters and disperses everything; whereas that other fire, the bodily, the vital, and the wholesome, preserves all things, feeds them, increases them, sustains them, and endows them with sensation." He says, then, that there is no doubt which of these two fires the sun resembles, since it too brings it about that all things flourish and come to maturity each in its own kind. Therefore, since the fire of the sun resembles those fires that are within the bodies of living creatures, the sun too must be a living being — and likewise the remaining heavenly bodies, which arise in the celestial blaze that is called the aether or the sky.

\ciceroSection{42} Since, then, the origin of some living creatures is on land, of others in water, of others in the air, it seems to Aristotle absurd to suppose that no living thing is born in that region which is most suited to the generation of living things. But the stars occupy the region of the aether; and since this is most rarefied and is forever in agitation and full of vigor, any living thing born in it must necessarily be endowed with the keenest sensation and the swiftest mobility. Therefore, since the heavenly bodies are born in the aether, it is consistent that sensation and understanding reside in them, from which it follows that the heavenly bodies are to be reckoned among the gods. For one may observe that the intellects of those who inhabit lands where the air is pure and rarefied are keener and more apt for understanding than the intellects of those who breathe a thick and dense atmosphere.

\ciceroSection{43} Indeed, they think that the very food one uses makes some difference to the keenness of the mind. It is probable, then, that an excellent intelligence resides in the heavenly bodies, which both inhabit the aetherial part of the world and are nourished by the moistures of sea and land that have been rarefied across a vast interval. But the sensation and understanding of the stars is shown above all by their order and their constancy — for there is nothing capable of being moved by measure and number without counsel — in which there is nothing reckless, nothing variable, nothing fortuitous. And the order of the stars, and their constancy through all eternity, signifies neither nature (for it is full of reason) nor chance, which, being a friend to variety, spurns constancy. It follows, then, that they move of their own accord, by their own sensation and divinity.

\ciceroSection{44} Nor indeed is Aristotle to be denied praise here: he held that all things that are moved are moved either by nature, or by force, or by will; and that the sun and the moon and all the heavenly bodies are moved; and that the things which were moved by nature are borne either downward by weight or upward by lightness — neither of which befalls the stars, because their motion is carried round in a circle. Nor can it be said that it comes about by some greater force that the stars are moved contrary to nature (for what force could be greater?). It remains, therefore, that the motion of the stars is voluntary. And whoever should see this would act not only ignorantly but impiously if he were to deny that the gods exist. Nor does it make much difference whether he denies it or strips them of all governance and activity; for to my mind that which does nothing seems not to exist at all. That the gods exist, then, is so manifest that I should scarcely judge a man to be of sound mind who denied it.

\ciceroSection{45} It remains for us to consider of what kind their nature is; and in this nothing is more difficult than to draw the mind's gaze away from the habit of the eyes. That difficulty has led the common run of the ignorant, and philosophers who are like the ignorant, to be unable to conceive anything about the immortal gods except under figures shaped like men — an opinion whose shallowness, refuted by Cotta, requires no speech of mine. But since by a settled conception of the mind we have a prior sense that a god is of such a kind — first, that he is animate; then, that in all nature there is nothing more excellent than he — to this presentiment and conception of ours I see nothing I might more fittingly accommodate than to judge, first, that this very world, than which nothing more excellent can come to be, is animate and a god.

\ciceroSection{46} Here let Epicurus jest as much as he likes — a man not at all well suited to jesting, and savoring least of all of his native land — and let him say that he cannot understand what a rolling, round god is like; yet by this he will never move me from what even he himself approves. For it is his view that the gods exist, because there must be some excellent nature than which nothing is better. But surely nothing is better than the world; nor is there any doubt that what is animate and possesses sensation and reason and mind is better than what lacks these.

\ciceroSection{47} Thus it follows that the world is animate and is possessed of sensation, mind, and reason; by which reasoning the conclusion is reached that the world is a god. But these matters will be more easily understood a little later from the very things which the world brings about. Meanwhile, Velleius, do not, I beg you, parade the fact that your people are utterly devoid of learning. You say that a cone and a cylinder and a pyramid seem to you more beautiful than a sphere. You have a novel canon of the eyes besides. But granting that these are more beautiful, at least to the sight — which to me they do not even seem to be: for what is more beautiful than that figure which alone, embracing all the others, contains them, and which can have nothing of roughness, nothing to give offense, nothing cut by angles, nothing in folds, nothing jutting out, nothing hollowed in? And since there are two preeminent forms — among solids the globe (for so it is fitting to render \textit{[Greek: sphaira]}), and among plane figures the circle or disc, which in Greek is called \textit{[Greek: kuklos]} — to these two forms alone it belongs that all their parts are most alike to one another, and that the outermost is everywhere equidistant from the center,

\ciceroSection{48} than which nothing more apt can come to be. But if you do not see this, because you have never touched that learned dust, could you not as natural philosophers have understood even this much — that this evenness of motion and constancy of the orbits could not have been preserved in any other figure? And so nothing can be more unlearned than what you are accustomed to affirm. For you do not even say that this very world is certainly round, since it might come to be of some other figure, and there might be countless other worlds of various shapes.

\ciceroSection{49} Had Epicurus once learned how much twice two makes, he certainly would not say this; but while he judges with his palate what is best, he never looked up, as Ennius says, to "the palate of the sky." For since there are two kinds of heavenly bodies — one of which, traveling through unchanging spaces from rising to setting, never bends the track of its course in any direction, while the other completes two continuous revolutions through the same spaces and along the same paths — from both these facts we recognize both the rotation of the world, which could exist only in a spherical form, and the circular orbits of the stars. And first the sun, which holds the chief place among the heavenly bodies, moves in such a way that, when it has filled the lands with abundant light, it darkens those same lands now from these quarters, now from those; for the earth's own shadow, blocking the sun, makes night. The intervals of night, moreover, have the same regularity as those of day. And the moderate approaches and retreats of that same sun temper the measure of cold and of heat. For the circuits of the sun, three hundred and sixty-five orbits with very nearly a fourth part of a day added, complete the annual revolution; and the sun, bending its course now toward the Bear, now toward the south, makes the summers and the winters, and those two seasons of which one is joined to the dying winter, the other to the summer. So from the changes of the four seasons are drawn the beginnings and causes of all things that come to be on land and sea.

\ciceroSection{50} The moon now overtakes the annual courses of the sun in monthly stretches: its faintest light comes when it draws nearest to the sun, its fullest whenever it stands farthest off. And not only its look and shape change, now waxing, now waning back to its starting point, but its region too; and when this is northern or southern, there is in the moon's course as well a certain likeness to the winter and the summer solstice, and many things stream and flow from it by which living creatures are nourished and grow and come to their prime, and by which the things that rise from the earth attain maturity.

\ciceroSection{51} Most wonderful of all, however, are the motions of those five stars falsely called wanderers; for nothing wanders that preserves through all eternity its advances and retreats and its other motions, constant and fixed. And this is the more wonderful in the stars we are speaking of, because now they are hidden, now revealed again; now they approach, now retreat; now they go ahead, now follow behind; now they move more swiftly, now more slowly; now they do not move at all, but for a certain time stand still. From their unequal motions the mathematicians have named the Great Year, which is brought to completion when the sun and the moon and the five wandering stars, all their courses run, have made their revolution back to the same relative position with one another.

\ciceroSection{52} How long this is, is a great question; but that it is fixed and definite is a necessity. For that star which is called Saturn's, and which the Greeks name Phainon \textit{[Greek: Phainon]}, which stands farthest from the earth, completes its course in about thirty years, and in that course, working many wonders — now going ahead, now slowing, now hiding itself in the evening hours, now opening to view again in the morning — it alters nothing through the everlasting ages of the generations, but does the same things at the same times. Below this, nearer to the earth, is borne the star of Jupiter, which is called Phaethon \textit{[Greek: Phaethon]}; and it completes the same circle of the twelve signs in twelve years, and shows the same variations in its course that Saturn's star does.

\ciceroSection{53} Next below this, holding a lower orbit, is Pyroeis \textit{[Greek: Pyroeis]}, the star called Mars's, and it traverses the same circle that the two higher stars do in twenty-four months, less, I think, six days. Below this is the star of Mercury (it is called Stilbon \textit{[Greek: Stilbon]} by the Greeks), which traverses the zodiacal circle in about a turning year and never withdraws from the sun farther than the span of a single sign, now running ahead, now following after. Lowest of the five wanderers and nearest the earth is the star of Venus, which is called Phosphoros \textit{[Greek: Phosphoros]} in Greek and Lucifer in Latin when it goes before the sun, but Hesperos \textit{[Greek: Hesperos]} when it follows; it completes its course in a year, traversing both the breadth of the zodiacal circle and its length — which the higher stars do likewise — and never withdraws from the sun farther than the span of two signs, now going ahead, now following.

\ciceroSection{54} This constancy in the stars, then, this great agreement of seasons through all eternity amid courses so various, I cannot understand apart from mind, reason, and design. And since we see these present in the heavenly bodies, we cannot but place the bodies themselves in the number of the gods. Nor indeed do those stars called fixed fail to signify the same mind and the same intelligence. Their daily revolution is harmonious and constant; and they do not have their courses in the upper air or fixed in the sky, as most people say who are ignorant of natural science. For the nature of the upper air is not such that, embracing the stars by its own force, it whirls them around; for the upper air, thin and translucent and suffused with even warmth, seems hardly fit to hold the stars together.

\ciceroSection{55} The fixed stars, then, have their own sphere, separate and free from any joining with the upper air. And their courses, perennial and perpetual, with their wonderful and incredible constancy, declare that there is in them a divine force and mind, so that whoever does not perceive that these very things possess the force of the gods would seem capable of perceiving nothing at all.

\ciceroSection{56} In the heavens, therefore, there is neither chance nor randomness nor straying nor emptiness, but on the contrary all order, truth, reason, and constancy; and the things that lack these — counterfeit and false and full of error — dwell around the earth, below the moon, which is lowest of all, and upon the earth itself. So the man who thinks that this celestial order, wonderful, and this incredible constancy, from which the preservation and safety of all things wholly arises, is empty of mind, must himself be reckoned destitute of mind.

\ciceroSection{57} I shall not, I think, go wrong, then, if I draw the starting point of this discussion from the leader in the search for truth. Zeno, then, defines nature so as to call it a fire endowed with craft, advancing methodically toward bringing things to birth. For he holds that the most proper work of any craft is to create and to bring forth, and that what the hand accomplishes in the works of our crafts, nature accomplishes far more skillfully — that is, as I said, as a fire endowed with craft, the master of all the other crafts. And by this account every nature is endowed with craft, since it has, as it were, a road and a method that it follows.

\ciceroSection{58} As for the world itself, which encloses and holds together all things in its embrace, the same Zeno calls its nature not merely endowed with craft but plainly a craftsman, taking thought and providing for all things useful and timely. And just as the other natures, each from its own seeds, are brought forth, grow, and are held together, so the nature of the world has all its motions voluntary, and efforts and appetites, which the Greeks call \textit{[Greek: hormas]}, and applies to them the actions that suit them, just as we ourselves do, who are moved by mind and the senses. Since, then, the mind of the world is of this kind, and for that reason can rightly be called either wisdom or providence (for in Greek it is called \textit{[Greek: pronoia]}), it provides above all for these things, and is occupied chiefly with them: first, that the world be as well fitted as possible to endure; next, that it lack nothing; and most of all, that there be in it surpassing beauty and every adornment.

\ciceroSection{59} We have spoken of the world as a whole; we have spoken also of the heavenly bodies, so that now there appears, nearly enough, a multitude of gods who neither rest idle nor carry out their works with toilsome and burdensome labor. For they are not held together by veins and sinews and bones, nor do they feed on these foods or drinks so as to take in humors either too sharp or too thick, nor are they of such bodies as to dread accidents or blows, or fear diseases from the weariness of their limbs — fears that led Epicurus to invent gods of mere outline who do nothing at all.

\ciceroSection{60} But those gods, endowed with the most beautiful form and set in the purest region of the sky, so move and govern their courses that they seem to have agreed together to preserve and protect all things. Many other natures of the gods, moreover, have been established and named — not without cause — for their great benefits, both by the wisest men of Greece and by our own ancestors. For whatever brought great benefit to the human race, they judged was not brought about without divine goodness toward men. And so at one time they called by the name of the god himself the thing that had been born of the god — as when we call the crops Ceres and wine Liber, from which comes that line of Te-

\ciceroSection{61} rence's, "without Ceres and Liber, Venus grows cold" — and at another time a thing in which some greater power resides is so named that the very power is called a god, like Faith, like Mind, which we see dedicated on the Capitol, the most recent dedication by Marcus Aemilius Scaurus, though earlier Faith had been consecrated by Aulus Atilius Calatinus. You see the temple of Virtue, you see the temple of Honor restored by Marcus Marcellus, which many years before had been dedicated in the Ligurian war by Quintus Maximus. What of Plenty, what of Safety, what of Concord, Liberty, Victory? Since in all these things the power was so great that it could not be governed without a god, the thing itself took on the name of a god. From this same class the names of Desire and Pleasure and of Venus Lubentina were consecrated — names of things faulty and not natural, though Velleius judges otherwise; and yet these very vices often strike upon our nature with considerable force.

\ciceroSection{62} By the greatness of their benefits, then, were those gods established who brought forth their several benefits; and by those very names that were mentioned a little earlier, the power that resides in each god is made plain. The life of men and common custom, moreover, took it upon themselves to raise to heaven, by fame and goodwill, the men outstanding in benefits. Hence Hercules, hence Castor and Pollux, hence Aesculapius, hence Liber too (I mean the Liber born of Semele, not the one whom our ancestors solemnly and sacredly consecrated as Liber along with Ceres and Libera — what kind of god that is can be understood from the mysteries; but because we call those born of us our liberi, children, Ceres' offspring were named Liber and Libera, a usage they keep in Libera's case but not likewise in Liber's) — hence Romulus too, whom some hold to be the same as Quirinus. Since their souls survived and enjoyed eternity, they were rightly held to be gods, being both best and eternal.

\ciceroSection{63} From another account too, and indeed a natural one, there flowed a great multitude of gods, who, clothed in human form, furnished the poets with their fables and crammed the life of men with every kind of superstition. And this topic, handled by Zeno, was later set out at greater length by Cleanthes and Chrysippus. For this ancient belief filled Greece: that Caelus was castrated by his son Saturn, and that Saturn himself was bound by his son Jupiter.

\ciceroSection{64} A natural account, and not without elegance, is enclosed within these impious fables. They meant that the celestial nature, highest and of the upper air — that is, fiery — which by itself would bring all things to birth, was free of that part of the body which would need conjunction with another in order to procreate. By Saturn, again, they meant the one who holds within himself the course and revolution of the spaces and seasons. This god has in Greek the very name that fits: he is called Kronos \textit{[Greek: Kronos]}, which is the same as Chronos \textit{[Greek: Chronos]}, that is, a span of time. And he was called Saturn because he was sated with years; for he is imagined as accustomed to devour his own offspring, since time consumes the spans of the seasons and is filled insatiably with the years gone by. He was bound by Jupiter, that he might not run uncontrolled in his courses, and so that Jupiter might fetter him with the bonds of the stars. But Jupiter himself — that is, the helping father, whom by a change of cases we call Jove from "helping" — is named by the poets "father of gods and men," and by our ancestors best and greatest, and indeed best, that is, most beneficent, before greatest, because it is greater, and surely more welcome, to do good to all than to have great wealth.

\ciceroSection{65} ...have. This, then, is the god whom Ennius, as I said above, names when he says, "Behold this radiance gleaming on high, whom all men invoke as Jupiter" --- more plainly than in another passage, where the same poet has, "I will call down upon him whatever lies within my power, this light, whatever it is, that shines." This too is the god our augurs mean when they say "while Jupiter lightens and thunders," for what they mean is "while the sky lightens and thunders." Euripides puts it --- as he puts many things --- splendidly, and here briefly: "You see the boundless ether spread on high, which holds the earth in its soft embrace: hold this to be the highest of the gods, hold this to be Jupiter."

\ciceroSection{66} The air, again, as the Stoics argue, lying between sea and sky, is consecrated under the name of Juno, who is the sister and wife of Jupiter, because the air resembles the ether and is most closely joined to it. They have made it feminine and assigned it to Juno, because nothing is softer than air. But I believe Juno was named from helping --- \textit{a iuvando}. Water and earth remained, that there might be three kingdoms divided up, as the fables have it. To Neptune, then, the brother of Jupiter as we hold, was given the other realm, the entire dominion of the sea; and his name is drawn out, just as Portunus comes from "harbor" --- \textit{a porta} --- so Neptunus comes from "swimming" --- \textit{a nando} --- with a slight change in the first letters. The whole force and nature of the earth was dedicated to Dis Pater --- "the rich one," as among the Greeks he is \textit{[Greek: Plouton]} --- because all things both fall back into the earth and rise out of the earth. To him they assigned Proserpina (which is a Greek name, for she is the one called \textit{[Greek: Persephone]} in Greek) --- whom they take to be the seed of the crops, and whom they imagine to lie hidden and to be sought by her mother.

\ciceroSection{67} The mother is Ceres, as if "Geres," from the bearing of crops --- \textit{a gerendis frugibus} --- the first letter changed by chance, just as it was among the Greeks; for by them too she was named \textit{[Greek: Demeter]}, as if "Earth-mother," \textit{[Greek: ge meter]}. Then there is the one who would overturn great things --- Mavors --- and Minerva, who either diminishes or threatens. And since in all things the first and the last have the greatest force, they wished Janus to be foremost in sacrifice; his name is drawn from "going" --- \textit{ab eundo} --- from which the open passageways are called \textit{iani} and the doorways set in the thresholds of unconsecrated buildings are called \textit{ianuae}. The name Vesta comes from the Greeks (for she is the one whom they call \textit{[Greek: Hestia]}); her power extends to altars and hearths, and so in her, as the keeper of things innermost, every prayer and every sacrifice has its last place.

\ciceroSection{68} Not far removed from this power are the Penates, their name drawn either from \textit{penus} --- for everything men feed on is \textit{penus} --- or from the fact that they dwell deep within, \textit{penitus}; and on that account they are also called \textit{penetrales} by the poets. Then there is the name of Apollo, which is Greek. They take him to be the sun, and they hold Diana and the moon to be one and the same; for the sun is called \textit{sol} either because it alone among all the heavenly bodies is so great, or because, when it has risen, all the rest are blotted out and it alone appears; and the moon is named \textit{luna} from shining, \textit{a lucendo}. She is the same as Lucina; and so, just as among the Greeks they invoke Diana, the Light-bearer, so among us they invoke Juno Lucina in childbirth. This same Diana is called Omnivaga, the All-wandering --- not from hunting, but because she is counted among the seven that are said to wander.

\ciceroSection{69} She is called Diana because by night she makes, as it were, a kind of day. She is brought in to aid at births, because these come to term either after seven, or, as more often, after nine courses of the moon; and because these complete measured spans, they are called \textit{menses}, months. Timaeus, neatly, as he does much else, after recounting in his history that on the very night Alexander was born the temple of Diana at Ephesus burned to the ground, added that this was no wonder at all, since Diana, having wished to be present at the delivery of Olympias, had been away from home. As for the goddess who would come to all things, our people named her Venus --- and \textit{venustas}, charm, comes from her rather than Venus from \textit{venustas}.

\ciceroSection{70} Do you see, then, how from physical things, well and usefully discovered, the account has been carried over to gods invented and feigned? This is what has bred false opinions and turbulent errors and superstitions little better than old wives' tales. For the shapes of the gods are made known to us, and their ages, their dress and adornment, their lineages besides, their marriages and kinships, everything transferred to the likeness of human frailty. They are even brought in with minds disordered: for we have heard of the desires of the gods, their sicknesses, their fits of anger; nor, if the fables are to be believed, have they been free of wars and battles --- not only, as in Homer, where some gods defended one of two opposing armies and some the other, but they have even waged their own wars, as against the Titans, as against the Giants. These things are told and believed in the height of folly, and they are full of triviality and the utmost frivolity.

\ciceroSection{71} Yet even so, with these fables spurned and rejected, a god pervading the nature of each thing --- Ceres throughout the lands, Neptune throughout the seas, others through other domains --- can be understood, both who and of what kind each is, and by what name custom has called him. These are the gods we ought to revere and worship. And the worship of the gods that is best, and likewise purest, holiest, and most full of devotion, is to revere them always with a mind and voice pure, whole, and uncorrupted. For not the philosophers only but our ancestors too separated superstition from religion.

\ciceroSection{72} For those who used to pray and offer sacrifice whole days together, that their own children might survive them --- might be \textit{superstites} --- were called superstitious, a name that afterward spread more widely; while those who carefully went back over everything that pertained to the worship of the gods, and as it were gathered it up again --- \textit{relegerent} --- were called religious, from "gathering again," \textit{relegendo}, just as the elegant are so called from "choosing out," \textit{eligendo}, and the diligent from "valuing," \textit{diligendo}, and the intelligent from "understanding," \textit{intellegendo}; for in all these words there is present the same force of "gathering" that is in "religious." So it came about that of the words superstitious and religious, the one became a name of reproach, the other of praise. And I think I have shown well enough both that the gods exist and what they are.

\ciceroSection{73} What comes next is for me to show that the world is governed by the providence of the gods. This is indeed a great theme, and one your school has harried, Cotta; and surely my whole contest is with you. For to you Epicureans it is less well known in what manner each thing is said: you read only what is your own, you love only what is your own, and you condemn the rest with the case unheard. So, for instance, you yourself said yesterday that a prophesying old woman, Pronoea --- that is, Providence --- is brought in by the Stoics. You said this out of that same error, because you suppose providence to be imagined by them as a kind of single goddess who steers and rules the whole world.

\ciceroSection{74} But the phrase is spoken in shorthand. Just as, if someone were to say that the Athenian republic is ruled by counsel, the words "of the Areopagus" are missing, so when we say the world is governed by providence, you should reckon that "of the gods" is missing, and understand that the full and complete way of saying it is this: the world is governed by the providence of the gods. So do not spend that wit of yours --- the wit your nation lacks --- in mocking us; and, by Hercules, if you will take my advice, do not even attempt it. It does not become you, it has not been granted you, you cannot do it. And this charge does not fall on you alone, polished as you are by the manners of your house and the urbanity of our countrymen, but on the rest of your school, and most of all on the man who fathered these notions --- a man without art, without letters, jeering at everyone, without any keenness, without authority, without grace.

\ciceroSection{75} I say, then, that by the providence of the gods the world and all the parts of the world were both established at the beginning and are governed for all time. This argument our people generally divide into three parts. The first is the one drawn from that reasoning which teaches that the gods exist; and once that is granted, it must be confessed that the world is governed by their counsel. The second is the one which teaches that all things are subject to a sentient nature and are most beautifully ordered by it; and once that is settled, it follows that this nature was generated from living first principles. The third is the line of argument drawn from our wonder at the things of heaven and earth.

\ciceroSection{76} First, then, either it must be denied that the gods exist --- which Democritus, bringing in his "likenesses," and Epicurus, bringing in his "images," in a manner deny --- or those who grant that the gods exist must admit that they do something, and something splendid; but nothing is more splendid than the governance of the world; therefore it is governed by the counsel of the gods. And if it is otherwise, then surely there must necessarily be something better, and endowed with greater force, than a god --- whatever that thing may be, whether an inanimate nature or a necessity driven by great force, producing these most beautiful works that we see;

\ciceroSection{77} the nature of the gods, then, is not all-powerful nor preeminent, if indeed it is subject to that necessity or nature by which sky, seas, and lands are ruled. But nothing is more excellent than a god; by him, then, the world must necessarily be ruled; therefore a god is obedient or subject to no nature; he himself, then, rules all nature. And indeed, if we grant that the gods are intelligent, we grant also that they are provident, and provident over the greatest things at that. Do they then not know which things are the greatest, and in what manner these are to be handled and watched over --- or do they lack the power to sustain and conduct such great things? But ignorance of things is foreign to the nature of the gods, and the difficulty of bearing a charge through weakness in no way befits the majesty of the gods. From which there follows what we maintain: that the world is governed by the providence of the gods.

\ciceroSection{78} And yet it must be that, since the gods exist (if indeed they do, as surely they do), they are living beings --- and not only living beings but also possessed of reason, and joined among themselves as if in a civic fellowship and society, ruling one world as a common commonwealth, a kind of city.

\ciceroSection{79} It follows that the same reason is in them as in the human race, that the same truth is in both, and the same law, which is the prescription of right and the prohibition of wrong. From which it is understood that prudence too, and mind, have come from the gods to men (and it is for this cause that, by the institutions of our ancestors, Mind, Faith, Virtue, and Concord have been consecrated and publicly dedicated; and how can it be consistent to deny that these reside with the gods, when we revere their august and sacred images? For if mind, faith, virtue, and concord are present in the human race, whence could these have flowed down to earth except from the gods above?); and since counsel, reason, and prudence are in us, the gods must necessarily possess these very things in greater measure, and not only possess them but also employ them in the greatest and best of matters.

\ciceroSection{80} But nothing is greater or better than the world; the world, therefore, must necessarily be governed by the counsel and providence of the gods. Finally, when we have shown sufficiently that these are gods whose conspicuous power and resplendent face we behold --- I mean the sun and the moon and the wandering and the fixed stars, and the sky, and the world itself, and the force of those things which are present throughout the whole world with great benefit and advantage to the human race --- it follows that all things are ruled by a divine mind and prudence. And concerning the first part, at least, enough has been said.

\ciceroSection{81} It follows that I should show how all things are subject to nature, and how supremely well she governs them. But first I must explain briefly what nature herself is, so that what I wish to show may be the more readily understood. For some hold that nature is a kind of force without reason, setting in motion the necessary movements in bodies; while others hold that it is a force partaking of reason and order, advancing as if along a road and making plain what it accomplishes as the cause of each thing and what follows from it — a skill which no art, no hand, no craftsman could match by imitation. For so great is the power of a seed that, tiny as it is, yet if it falls upon a nature that conceives and grasps it, and has gained material by which it can be fed and increased, it so shapes and produces each thing in its own kind that some are nourished merely through their own stalks, while others can also move and feel and have appetite and beget from themselves their own likeness.

\ciceroSection{82} There are, however, those who give the name of nature to everything — Epicurus, for instance, who divides the matter thus: that the nature of all things that exist consists of bodies and void and the properties of these. But when we say that the world stands and is administered by nature, we do not mean it in the way one would speak of a clod, or a fragment of stone, or anything of that sort, which has no nature holding it together; we mean it as one speaks of a tree or an animal, in which there is no randomness, but order, and a certain likeness to design.

\ciceroSection{83} Now if the plants that are held in the earth by their roots live and thrive by the art of nature, surely the earth herself is held by that same force, by the art of nature — she who, made pregnant with seeds, brings forth and pours out all things from herself, embraces the roots and feeds and increases them, and is herself fed in turn by the heavenly and outer natures; and by her exhalations the air too is nourished, and the aether, and all the things above. So, if the earth is sustained by nature and thrives, the same reasoning holds for the rest of the world; for the plants cling to the earth, and living creatures are sustained by the breath of air. The air itself sees with us, hears with us, sounds with us — for none of these things can happen without it; indeed, it even moves with us, for wherever we go, wherever we move, it seems to give place, as it were, and yield.

\ciceroSection{84} And those things that are borne toward the central place of the world, which is the lowest, and those that move from the center toward the heights, and those that are carried round about the center in circling revolution — these make up the continuous fabric of the world and a single nature. And since there are four kinds of body, it is by their alternation that the nature of the world is held continuous. For from earth comes water, from water arises air, from air aether; then in reverse turn, from aether comes air, from it water, and from water the lowest earth. Thus, with these natures of which all things are composed passing up and down, back and forth, the conjunction of the world's parts is held together.

\ciceroSection{85} This fabric must either be everlasting in the very order we see, or at any rate of vast duration, enduring for an immense, almost boundless stretch of time. Whichever of the two it is, it follows that the world is administered by nature. For what voyage of fleets, what marshaling of an army — or, to turn again to the works of nature, what generation of vine or tree, what shape of a living creature and conformation of its limbs — declares so great a skill in nature as the world itself? Either, then, there is nothing governed by a nature endowed with sensation, or it must be confessed that the world is so governed.

\ciceroSection{86} For that which contains all other natures and their seeds — how can it not itself be administered by nature? It is as if a man should say that teeth and the growth of manhood come to be by nature, but that the man himself, in whom they come to be, does not consist by nature — failing to understand that the things which bring something forth from themselves have more perfect natures than the things that are brought forth from them. Now of all things that are administered by nature, the world is the sower and planter and, so to speak, the parent and rearer and nourisher; it feeds and contains all things as its own members and parts. And if the parts of the world are administered by nature, the world itself must necessarily be administered by nature. Its administration, indeed, has nothing in it that could be faulted; for out of the natures that there were, what could be made best was made.

\ciceroSection{87} Let someone, then, show that it could have been made better; but no one ever will, and if anyone wishes to correct something he will either make it worse or will be wanting something that could not have been done. But if all the parts of the world are so constituted that they could neither have been better for use nor more beautiful in appearance, let us consider whether these things are the work of chance, or whether they are in a condition in which they could in no way have cohered without a guiding intelligence and divine providence. If, then, the things perfected by nature are better than those perfected by art, and art accomplishes nothing without reason, neither is nature to be held devoid of reason. How then does it make sense that, when you have looked at a statue or a painted panel, you know that art has been applied, and when you see from afar the course of a ship you do not doubt that it is moved by reason and art, and when you contemplate a sundial, whether marked out by the shadow or driven by water, you understand that the hours are declared by art and not by chance — and yet you should suppose the world, which embraces these very arts and their craftsmen and all things besides, to be devoid of design and reason?

\ciceroSection{88} But suppose someone carried into Scythia or Britain that orrery which our friend Posidonius lately made, whose every revolution brings about in the sun and in the moon and in the five wandering stars the same thing that is brought about in the heavens on each day and night — who, in that barbarous land, would doubt that the orrery had been perfected by reason? Yet these men hesitate over the world, from which all things arise and come to be, whether it was itself produced by chance or by some necessity, or rather by reason and a divine mind; and they reckon that Archimedes did more in imitating the revolutions of the sphere than nature did in producing them — although nature accomplished her work far more skillfully, in many respects, than these copies were contrived.

\ciceroSection{89} And like that shepherd in Accius, who had never seen a ship before, when he caught sight from the mountaintop, far off, of the divine and unfamiliar vessel of the Argonauts, and, struck with wonder and terror, spoke in this fashion: "So vast a mass comes gliding, roaring, from the deep, with mighty sound and breath: before itself it rolls the waves, by force whips up the eddies, plunges headlong, sprays the sea and spouts it back; so that now you would believe a storm-cloud broken loose comes rolling, now a crag flung up and snatched aloft by winds or squalls, or globed whirlwinds rising, struck by the dashing waves — unless the sea is stirring up some wreckage of the land, or perhaps Triton, overturning the caverns with his trident beneath their very roots, in the surging strait, heaves a mass of rock from the deep to the sky." At first he doubts what that nature is which he beholds, unknown; and then again, when the young men come into view and the sailors' chant is heard, he says: "Just as the lively, eager dolphins snort through their snouts" — and much else besides — "to my ears they bring a song and a sound much like the strain of Silvanus" —

\ciceroSection{90} so, then, just as this man at first sight thinks he beholds something lifeless and void of sensation, but afterward, from surer signs, begins to surmise what sort of thing it is over which he had been in doubt, in the same way the philosophers, if perhaps the first sight of the world threw them into confusion, ought afterward, when they had seen its motions defined and uniform and all things governed by fixed orders and unchangeable constancy, to have understood that there is present, in this celestial and divine house, not only an inhabitant but a ruler and governor and, as it were, an architect of so great a work and so great a charge. But as it is, they seem to me not even to suspect how great is the marvel of the things of heaven and earth.

\ciceroSection{91} For, to begin with, the earth, set in the middle region of the world, is surrounded on every side by this living, breathable substance whose name is air — a Greek word, to be sure, yet one already taken into use by our own people, for it has become current as though it were Latin. This in turn is embraced by the immense aether, which consists of the highest fires (let us borrow this word too, and let \textit{aether} be said in Latin just as \textit{air} is said, though Pacuvius offers a translation: "this, which I name, our people call the sky, the Greeks call aether" — as if, in fact, the speaker were not himself a Greek. "But he speaks Latin." Yes, if we are not to hear him as one speaking, in effect, Greek; for the same poet shows it elsewhere:

\ciceroSection{92} "Greek-born: of that the speech itself makes plain") — but let us return to greater matters. From the aether, then, arise the countless flames of the stars, of which the chief is the sun, lighting all things with its most brilliant radiance, many times greater and larger than the whole earth; then the rest of the heavenly bodies, of immense magnitudes. And these fires, so great and so many, not only do no harm to the earth and the things of earth, but are of such benefit that, were they moved from their place, the earth would necessarily be set ablaze by such heat, once moderation and tempering were taken away.

\ciceroSection{93} Here, can I keep from wondering that there should be anyone who persuades himself that certain solid and indivisible bodies are borne along by their own force and weight, and that the world, in all its order and beauty, is produced from the chance collision of those bodies? I do not see why a man who thinks this could have happened should not also think that, if countless forms of the one-and-twenty letters, whether of gold or of any material you like, were thrown together somewhere, then from these, shaken out upon the ground, the Annals of Ennius could be produced so as to be read straight through in order — a thing which I doubt chance could so much as accomplish in a single line.

\ciceroSection{94} Yet these men keep insisting that the world was perfected out of little bodies endowed with no color, no quality (what the Greeks call \textit{[Greek: poiotēta]}), no sensation, but rushing together at random and by chance — or rather that countless worlds, in every point of time, are being born and others perishing. But if a concourse of atoms can produce a world, why can it not produce a colonnade, a temple, a house, a city — things less laborious, and indeed many of them far easier? Truly they babble so heedlessly about the world that to me they seem never to have looked up at this marvelous ordering of the heavens (which is the very subject before us).

\ciceroSection{95} Splendidly, then, says Aristotle: "Suppose there were men who had always lived underground, in fine and well-lit dwellings, adorned with statues and paintings and furnished with everything in abundance that those reckoned happy possess, yet who had never come up above the earth, but had learned by report and hearsay that there is a certain divine power and might of the gods; and suppose that at some time the jaws of the earth were opened, and they were able to escape from those hidden seats and come out into these regions which we inhabit: when they had suddenly beheld the earth and the seas and the sky, and had come to know the vastness of the clouds and the force of the winds, and had looked upon the sun and come to know both its magnitude and beauty and also its power — that it makes the day by pouring light over the whole sky — and when, once night had darkened the lands, they then saw the whole sky picked out and adorned with stars, and the changing light of the moon now waxing and now waning, and the risings and settings of all these and their courses fixed and unchangeable through all eternity — when they saw these things, they would surely judge both that the gods exist and that these so great works are

\ciceroSection{96} the works of the gods." So much for him. As for us, let us imagine darkness as great as that which, they say, once obscured the neighboring regions at the eruption of the fires of Aetna, so that for two days no man recognized his fellow, and when on the third day the sun shone out, they then seemed to themselves to have come back to life. But if this same thing should befall us — that out of eternal darkness we should suddenly look upon the light — what an appearance the sky would seem to have! Yet by the daily constancy of it and the habit of our eyes our minds grow used to it, and we neither wonder nor seek out the reasons for the things we always see — just as though it were the novelty, rather than the magnitude, of things that ought to rouse us to inquire into their causes.

\ciceroSection{97} For who would call this creature a man — who, when he has seen motions of the heavens so fixed, orders of the stars so settled, and all things so interconnected and fitted to one another, should deny that any reason is present in them and assert that those things come about by chance, things conducted with a forethought we cannot match by any forethought of our own? When we see something moved by a kind of mechanism — an orrery, say, or a water clock, or a great many other contrivances — we do not doubt that they are the work of reason; and yet, when we see the onrush of the heavens turning and revolving with marvelous speed, bringing about the changes of the seasons year by year with the utmost benefit and preservation of all things, do we doubt that these come about not only by reason, but by a reason surpassing and divine?

\ciceroSection{98} For we may now set aside the subtlety of argument and contemplate, in a manner of speaking, with our very eyes the beauty of those things which we say divine providence has established. And first let the whole earth be brought into view, set in the central seat of the world, solid and globed and gathered into a sphere on every side by its own downward tendencies, clothed in flowers, grasses, trees, and crops — of all of which an incredible multitude is set off by an inexhaustible variety. Add to this the cold perpetual flow of springs, the translucent waters of rivers, the greenest mantling of their banks, the hollow depths of caverns, the roughness of crags, the heights of overhanging mountains, and the immensities of the plains; add too the hidden veins of gold and silver, and the boundless wealth of marble.

\ciceroSection{99} And then the kinds of living creatures — how many and how varied, whether tame or wild — the gliding flight and the song of birds, the grazing of the herds, the life of the woodland creatures. What now shall I say of the race of men, who, appointed as it were to be the tillers of the earth, do not suffer it to run wild with the savagery of beasts or to lie waste under the roughness of brush, and by whose works the fields, the islands, and the shores shine bright, set off with dwellings and cities? If we could see all this with our eyes as we see it with our minds, no one, gazing upon the earth entire, would doubt the divine reason behind it.

\ciceroSection{100} And how great, again, is the beauty of the sea, what the aspect of its whole expanse, what the multitude and variety of its islands, what the loveliness of its coasts and shores, how many and how unlike the kinds of creatures — some submerged, some floating and swimming, some clinging to the rocks in shells of their own growing. The sea itself, reaching out toward the land, so plays along its shores that the two natures seem fused into one.

\ciceroSection{101} Next, bordering on the sea, the air is divided between day and night; now, diffused and rarefied, it is borne aloft, now, condensed, it is packed into clouds and, gathering moisture, swells the earth with rains, and now, flowing this way and that, it makes the winds. The same air produces the yearly alternations of cold and heat; the same both sustains the flights of birds and, drawn in by breath, feeds and supports living things. There remains the last embrace, the highest above our dwellings, which encircles and confines all things — the same that is called the aether, the outermost rim and boundary of the world, in which, with the greatest wonder, the fiery shapes mark out their ordered courses.

\ciceroSection{102} Among these the sun, by whose magnitude the earth is many times surpassed, revolves about that very earth, and rising and setting brings about day and night; and now drawing nearer, now again withdrawing, it makes each year two reversals, opposite to each other, from the farthest points — and in the interval between them it now contracts the earth, as though with a kind of sadness, and now in turn gladdens it, so that it seems to rejoice along with the sky.

\ciceroSection{103} The moon, which is, as the mathematicians demonstrate, larger than half the earth, ranges over the same tracts as the sun, but now meeting the sun, now departing from it, sends down to the lands the light it has received from the sun and itself undergoes various changes of light; and now, lying beneath and set against the sun, it darkens his rays and his light, now, falling itself into the shadow of the earth when it is directly opposite the sun, is suddenly eclipsed by the interposition and intervention of the earth between them. Over the same tracts those stars we call wandering are borne about the earth, and in the same way they rise and set, their motions now quickened, now slowed,

\ciceroSection{104} and often they stand still — than which spectacle nothing can be more wonderful, nothing more beautiful. There follows the vast multitude of the fixed stars, so distinctly grouped that men have found names for them out of their resemblance to familiar figures.'' And here, looking at me, ``I shall make use,'' he said, ``of the Aratean verses, which, since you turned them into Latin in your earliest youth, so delight me precisely because they are Latin that I keep many of them by heart. So then, as we see continually with our eyes: \textit{Without any change or alteration the rest of the heavenly bodies glide on with swift celestial motion, and together with the sky are borne along through the nights and days,}

\ciceroSection{105} \textit{in the contemplation of which the mind of one who longs to see the steadfastness of nature can never have its fill. And at the very farthest point, on the twofold pivot, is what is called the topmost height, the pole.} About this turn the two Bears, that never set. Of these one is called among the Greeks the Dog's Tail, \textit{[Greek: Kynosoura]}, the other is said to be the Twister, \textit{[Greek: Helike]}, whose brightest stars, indeed, we behold all night long — the ones our people are wont to call the Seven Plow-oxen;

\ciceroSection{106} \textit{and with stars set off as evenly and alike, the little Dog's Tail ranges round the same height of the sky. In her the Phoenicians trust as their guide upon the deep at night. But that earlier one shines forth more distinct with stars and is seen broad and foremost the moment night falls. This one, though small, yet to sailors is its use; for it turns in a brief circle on its inner course.} And, that the look of those stars may be the more wonderful, between them, like a river with its rushing flood, \textit{the grim Dragon glides, coiling himself below and above, and shaping the winding folds that issue from his body.}

\ciceroSection{107} \textit{Splendid as is the look of him entire, above all to be marked is the shape of his head and the blaze of his eyes: for him not one star alone shines forth, adorning his head, but his temples are marked with a twofold gleam, and from his fierce eyes two burning fires flame out, and his chin is bright with one radiant star; his head is bent aslant, and turned back upon his rounded neck, you would say he fixes his gaze upon the tail of the Greater Bear.}

\ciceroSection{108} \textit{And the rest of the Dragon's body indeed we behold all night long, but here his head dips and on a sudden hides itself, where rising and setting are blended into one. And, touching that head, there turns — like the image of one weary and grieving — the figure} which the Greeks are wont to call the Kneeler, \textit{[Greek: Engonasin]}, because it is borne along resting upon its knees. \textit{Here, set with surpassing brightness, is placed that Crown. And this one indeed is at its back; but next to its head is the Snake-holder, whom the Greeks proclaim by the famous name of} the Serpent-bearer, \textit{[Greek: Ophiouchos]}.

\ciceroSection{109} \textit{He with the double pressure of his palms holds fast the Serpent, and is himself held bound by its twisted body; for the Serpent girds the man's middle beneath his breast. Yet he, straining, sets his feet down heavily and presses with them the eyes and breast of the Scorpion.} The Seven Plow-oxen, moreover, are followed by the Bear-warder, commonly said to be the Oxherd, \textit{[Greek: Bootes]}, because he drives the Bear before him as though yoked to the pole. Then what follows:

\ciceroSection{110} \textit{For beneath the Oxherd's breast there seems to be fixed a star flashing with rays, Arcturus by its famous name; and set beneath his feet is borne, holding the bright Ear of grain in her shining body, the Maiden.} And so the constellations are laid out by measure, that in arrangements so vast a divine craftsmanship may appear. \textit{And you would have seen the Twin-born below the Bear's head; beneath the middle lies the Crab, and by the feet is held the great Lion, shaking from his body a quivering flame. The Charioteer will be borne along, drawn beneath the left side of the Twins. Opposite, the fierce one of the Twister gazes at his head. But the Goat holds his left shoulder, bright.} Then what follows: \textit{but this is a star endowed with a great and brilliant sign, while over against it the Kids cast their scant fire upon mortals. Beneath whose feet is the horn-bearing Bull, braced upon his sturdy body. His head is sprinkled with thronging stars;}

\ciceroSection{111} these stars the Greeks have been wont to call the Hyades, from raining (\textit{[Greek: hyein]} being to rain), while our people, in their ignorance, call them the Piglets, as though they were named from sows rather than from the showers. \textit{The Lesser Bear, moreover, Cepheus follows close behind with hands outspread; for he himself is turned toward the back of the Dog's-Tail Bear. Before him goes Cassiopeia, dim in the look of her stars. And next to her wheels, with shining body, Andromeda, fleeing the sad gaze of her parent. For her that Horse, shaking his mane with flashing brightness, touches the top of her head with his belly, and one star, joining them, holds the two figures in a shared light, longing to bind an everlasting knot out of the stars. Next clings the Ram with his curled-back horns; near him the Fishes, of which one glides on a little ahead and is touched more by the bristling breezes of the north wind.}

\ciceroSection{112} \textit{At the feet of Andromeda is Perseus traced out, whom the blasts buffet from the topmost region of the north. Near his left knee you will see, with faint light, the Pleiades. Next the Lyre is seen, lightly laid and arched. Next is the winged Bird beneath the broad covering of the sky. And nearest to the Horse's head is the right hand of the Water-pourer, and then, after him, the whole Water-pourer. Then, breathing chill cold from his mighty breast, half-beast in body, Capricorn upon his great orbit; whom, when the Titan has clothed him in unbroken light, wheeling about he drives his chariot at the winter season.}

\ciceroSection{113} \textit{And here is seen, showing itself, the Scorpion rising on high, drawing behind it with the force of its body the curving Bow. Near it the winged Bird wheels in flight, gleaming with its pinions. Hard by, the Eagle carries itself with its burning body. Then the Dolphin. And after it Orion, leaning with his slanting frame.}

\ciceroSection{114} \textit{Following close upon him, that blazing Dog shines out with the light of its stars. After it the Hare comes next, never resting its course on a wearied body. And at the Dog's tail the Argo glides, snaking along. The Ram covers her, and the Fishes with their scaly bodies, while she touches with her gleaming body the banks of the River — which you will see snaking far and streaming on; and you will behold the long Bonds set in the quarter of their tails that hold the Fishes back. Then, beside the Scorpion's flashing sting, you will discern the Altar, which the breath of the south wind caresses with its blast. Near these the Centaur moves, hastening to join the parts of the Horse to the Claws. With his right hand outstretched, where the great four-footed beast is gripped, he reaches and grimly advances toward the bright Altar. Here from the regions below rears up the Hydra, whose body lies poured out far along, and in the middle of its coil the gleaming Bowl shines back. At the end the Crow, straining with its feathered body, strikes with its beak; and here, just beneath the Twins themselves, before the Dog, is that star which goes by the Greek name Procyon.}

\ciceroSection{115} Can this whole ordering of the constellations, this whole great adornment of the heavens, seem to any man of sound mind to have been brought about by bodies coursing this way and that by chance and at random? Or did some other nature, devoid of mind and reason, manage to produce these things, which not only required reason in order to come into being, but cannot even be understood for what they are without the highest reason? Nor is this alone to be marveled at; there is nothing greater than the fact that the world is so stable and coheres so closely that nothing better fitted for its endurance can so much as be conceived. For all its parts, on every side pressing toward the central place, strain equally. And bodies joined to one another remain so most firmly when they are bound together as if by some encompassing chain; and this is the work of that nature which, pervading the whole world, brings all things to completion through mind and reason, drawing them toward the center and turning the outermost parts inward.

\ciceroSection{116} And so, if the world is spherical, and if for that reason all its parts on every side are held together, balanced both in themselves and among themselves, the same must necessarily hold true of the earth: with all its parts inclining toward the center — and the center in a sphere is the lowest point — nothing breaks in to undermine so great a tension of gravity and weight. By the same principle the sea, though it lies above the earth, nevertheless, seeking the central place of the earth, gathers itself into a sphere uniformly on every side, and neither overflows nor pours away.

\ciceroSection{117} The air, which borders upon the sea, is borne aloft by its lightness, yet pours itself out in every direction; and so it is at once continuous and joined with the sea and naturally borne toward the heavens, tempered by whose thinness and warmth it furnishes living things with the breath of life and health. The topmost part of the heavens, which is called the ethereal, embraces the air, and at once retains its own fire — thin and condensed by no admixture — and is joined to the outer limit of the air. In the ether the stars revolve, which both hold themselves together, compacted by their own striving, and sustain their movements by their very form and figure; for they are round, and to forms of this kind, as I think I have said before, no harm can come.

\ciceroSection{118} The stars are by nature fiery; and so they are fed by the vapors of the earth, the sea, and the other waters — vapors raised by the sun from the warmed fields and from the waters. Nourished and renewed by these, the stars and the whole ether pour them out again and draw them back from the same source, so that almost nothing perishes, or only the very smallest amount, consumed by the fire of the stars and the flame of the ether. From this our school holds that what will come about is the very thing they said Panaetius doubted: that in the end the whole world will burst into fire, when, the moisture being consumed, neither can the earth be nourished nor can the air return, since the air's source, once all the water is exhausted, cannot exist; and so nothing will be left but fire, by which in turn, as a living being and a god, the renewal of the world will take place and the same adornment will arise again.

\ciceroSection{119} I do not wish to seem to you to dwell at length on the system of the stars, and above all on those said to wander; their harmony out of the most dissimilar motions is so great that, while the highest, Saturn's, chills, the middle one, Mars's, sets ablaze, and Jupiter's, set between them, gives light and tempers; and below Mars two obey the sun, while the sun itself fills the whole world with its light, and from it the moon, illumined, brings on conceptions and births and the ripening of all that is born. This linking of things, this conjoining of nature consenting, as it were, to the world's preservation — the man whom it does not move, I know for certain has never reflected on any of these matters.

\ciceroSection{120} Come, that we may pass from the things of the heavens to those of the earth — what is there among them in which the reasoning of an intelligent nature does not appear? To begin with, among the things that grow from the earth, the roots both give stability to the things they support and draw from the earth the sap by which the parts held together by the roots are nourished; and the trunks are clothed in bark or rind, so as to be the safer against cold and heat. And then the vines grasp their props with their tendrils, as though with hands, and so raise themselves up like living things; indeed, the cabbages are said to recoil from the stalks of the vine, if they are planted near them, as from things pestilent and harmful, and not to touch them in any part.

\ciceroSection{121} As for living creatures, how great is their variety, how great their capacity for this end — that each should persist within its own kind! Of these, some are covered with hides, some clothed with shaggy hair, some bristling with spines; some we see overlaid with feathers, others with scales, some armed with horns, some furnished with wings for escape. To living creatures, moreover, nature has supplied lavishly and abundantly the food that suited each. I could enumerate, for the seizing and processing of that food, what design there is in the shapes of living things, and how skillful and subtle the disposition of their parts, how admirable the construction of their limbs. For all the parts at least that are enclosed within are so formed and so placed that none of them is superfluous, none not necessary to the preservation of life.

\ciceroSection{122} The same nature has given to the beasts both sensation and appetite, so that by the one they might have the impulse to seize their natural foods, and by the other distinguish the pestilent from the wholesome. Some animals, again, approach their food by walking, others by crawling, others by flying, others by swimming; and they take their food partly with the open mouth and with the teeth themselves, partly seize it with the grip of their claws, partly with the hook of their beaks; some suck, some pluck, some gulp, some chew. Some, too, are of such lowness that they easily reach food on the ground with their mouths, while those that are taller — geese, swans, cranes, camels — are helped by the length of their necks;

\ciceroSection{123} a hand, too, was given to the elephant, because on account of the size of its body it had difficult access to its food. As for the beasts whose food was such that they had to feed on the foods of other kinds, nature gave them either strength or speed. To some, even, a certain contrivance and skill was given — as among the little spiders, some weave a kind of net, so that they may dispatch whatever sticks in it, while others watch unawares and snatch whatever falls in and consume it. The pina, however — for that is its Greek name — opening up two great mussel-shells, enters as it were into a partnership with a little shrimp for the procuring of food; and so, when small fish have swum into its gaping shell, the pina, warned by the shrimp, snaps the shells shut with a bite: thus among the most dissimilar little creatures food is sought in common;

\ciceroSection{124} and here it is to be marveled at whether they were brought together by some encounter with each other, or were associated by nature herself straight from their birth. There is, too, some occasion for wonder in the water-dwelling creatures that are born on land — for instance crocodiles and river tortoises and certain serpents which, born away from the water, as soon as they first can move, make for the water. Indeed, we often place ducks' eggs under hens; and the chicks hatched from them are first nourished by these as by mothers, by whom they were brought forth and warmed, and then they leave them and flee from them as they follow, the moment they have been able to catch sight of water as their natural home: so great a guardianship of self-preservation has nature implanted in living things. I have also read it written that there is a certain bird called the spoonbill; it seeks its food by flying to those birds that plunge themselves into the sea, and when these have emerged and caught a fish, it presses their heads with its bite until they let go their catch, upon which it pounces itself. The same bird is recorded as being in the habit of filling itself with shellfish, and when it has cooked these with the heat of its stomach, of vomiting them up, and so selecting from them the parts that are good to eat.

\ciceroSection{125} Sea-frogs, again, are said to be in the habit of burying themselves in the sand and stirring near the water, and when fish approach them as if to feed, they are dispatched by the frogs and consumed. The kite has a kind of war with the crow, as it were natural to it; and so each, wherever it comes upon them, breaks the other's eggs. But this — observed by Aristotle, by whom most things were observed — who can fail to marvel at? When cranes, seeking warmer regions, cross the seas, they form the shape of a triangle; with its uppermost angle the air opposing them is thrust aside, and then little by little, on either side, as though by oars, the birds' course is lightened by their wings; and the base of the triangle which the cranes make is helped, as though from a stern, by the winds; and they rest their necks and heads on the backs of those flying ahead; and because the leader himself cannot do this, since he has nothing to lean upon, he flies back so that he too may rest, and into his place succeeds one of those that have rested, and this exchange is kept up throughout the whole flight.

\ciceroSection{126} I could bring forward many things of this kind, but you see the type itself. And then there are these still more familiar facts: how greatly the beasts guard themselves, how they look about them while feeding, how they hide away in their lairs. And these are marvelous: that the very remedies discovered only recently — that is, a few centuries ago — by the genius of physicians, the dogs treat themselves with by vomiting, and the Egyptian ibises by purging the bowels. It has been heard that panthers, which in barbarian lands are caught with poisoned meat, have a certain remedy which, when they have used it, they do not die; and that the wild goats in Crete, when they have been pierced with poisoned arrows, seek out a herb called dittany, and that when they have tasted it the arrows are said to fall out of their bodies;

\ciceroSection{127} and that hinds, a little before bearing their young, thoroughly purge themselves with a certain little herb called hartwort. And then we observe how each defends itself with its own weapons against violence and danger: bulls with their horns, boars with their tusks, lions by running, some protect themselves by flight, others by hiding, the cuttlefish by discharging ink, the torpedo-fish by numbing, and many also drive off their pursuers by the intolerable foulness of their stench. And so that the adornment of the world might be perpetual, great care was taken by the providence of the gods that there should always exist both the kinds of beasts and of trees and of all the things held together from the earth by roots — all of which indeed have within themselves such a force of seed that from one many are generated. And this seed is enclosed in the inmost part of those berries which are shed from each plant, and by these same seeds men feed in abundance and the lands are filled by the renewal of plants of the same kind.

\ciceroSection{128} Why should I speak of how great a design appears in the beasts for the perpetual preservation of their kind? For first, some are male and some female, which nature has contrived for the sake of perpetuity; then the parts of the body are most aptly fitted both for begetting and for conceiving, and in male and female alike there are marvelous urges for the coupling of bodies; and when the seed has lodged in its place, it draws to itself nearly all the nourishment and, walled in by it, forms the animal; and when this has slipped from the womb and dropped free, in those living things that are nourished by milk nearly all the mothers' nourishment begins to turn to milk, and the newborn, with no teacher but nature for guide, seek out the teats and are sated by their abundance. And that we may understand that none of these things is by chance and that all of them are the works of a provident and skillful nature, those creatures that bring forth manifold offspring — such as sows and dogs — are given a multitude of teats, the very teats that those beasts have few of which bring forth few.

\ciceroSection{129} What shall I say of how great a love beasts have in rearing and guarding the young they have brought forth, right up to the point at which the young can defend themselves? Fish, it is true, abandon their eggs once they have laid them, since water both buoys those eggs up easily and brings the brood to maturity; turtles and crocodiles, they say, after delivering their young on land, bury the eggs and then depart, so that the young are born and reared entirely by themselves. Hens, on the other hand, and the rest of the birds both seek out a quiet place for laying and build themselves nests and roosts, and line them as softly as they can, so that the eggs may be kept as safely as possible; and once they have hatched the chicks from these, they protect them in such a way as both to cherish them with their feathers, lest they be harmed by cold, and to shield them with their bodies if there is heat from the sun. But when the chicks are able to use their little wings, then the mothers attend their flights, and are released from any further care.

\ciceroSection{130} There is also added, toward the preservation and well-being of certain living creatures and of those things which the earth brings forth, the resourcefulness and diligence of human beings. For there are many kinds of livestock and many plants that cannot survive without human management. Great opportunities, too, for the cultivation and abundance of human life are found in one region or another. The Nile irrigates Egypt, and when it has held the whole country submerged and flooded all summer long, it then withdraws and leaves the fields, softened and silted over, ready for sowing. The Euphrates makes Mesopotamia fertile, into which it each year brings, as it were, fresh fields. And the Indus indeed, which is the greatest of all rivers, not only gladdens and mellows the fields with its water but even sows them; for it is said to carry down with it a great quantity of seeds resembling grain.

\ciceroSection{131} And I could bring forward many other things in other regions worth recording, many fertile fields rich in one crop or another. But how great is that bounty of nature, which brings forth so many things to eat, so various and so delightful, and these not at one season of the year, so that we are always delighted by both their novelty and their plenty. And how seasonable, how salutary the Etesian winds are which she has given, not only to the race of men but also to that of the herds, and indeed to all things that spring from the earth: by their breath excessive heat is tempered, and by them too the swift and steady courses of ships at sea are guided. Many things must be passed over, and yet many are spoken of.

\ciceroSection{132} For the advantages of rivers cannot be enumerated, nor the tides of the sea ebbing and flowing, the mountains clothed and forested, the salt-pans set farthest back from the seacoast, the lands brimming with healing medicines, and finally the countless arts necessary for sustenance and for life. Again, the alternation of day and night preserves living creatures by allotting one time for action and another for rest. Thus on every side, by every line of reasoning, the conclusion is drawn that by a divine mind and design all things in this world are administered, in a manner deserving of wonder, toward the safety and preservation of all.

\ciceroSection{133} But should anyone ask for whose sake so vast a work was contrived—was it for the sake of trees and plants, which, though without sensation, are nevertheless sustained by nature? But that is absurd. Was it for the sake of beasts? It is no more probable that the gods labored so greatly for the sake of creatures dumb and without understanding. For whose sake, then, would one say the world was made? Surely for the sake of those living beings that use reason; these are gods and men, than whom assuredly nothing is better, for reason is what surpasses all things. Thus it becomes credible that the world, and all things that are in that world, were made for the sake of gods and men. And it will be more readily understood that provision has been made by the immortal gods for men, if the whole fabrication of man is examined, and the entire figure and perfection of human nature.

\ciceroSection{134} For since the life of living creatures is sustained by three things—food, drink, and breath—the mouth is best suited for taking in all of these: with the nostrils joined to it, it is augmented by breath, while by the teeth set in array within the mouth the food is chewed and by them broken down and softened. The front teeth, being sharp, divide the food with their bite, while the inner ones, which are called the molars, grind it up; and this grinding seems to be assisted by the tongue as well.

\ciceroSection{135} Next to the tongue, clinging at its roots, the gullet receives what comes; into it first slip down the things taken in by the mouth. Touching the tonsils on either side, it is bounded by the farthest and innermost part of the palate, and when, by the working and movements of the tongue, it has received the food driven and as it were thrust down, it forces it downward. Those parts of it which lie below the thing being swallowed dilate, while those which lie above contract.

\ciceroSection{136} But since the windpipe (for so it is called by the physicians) has its opening joined to the roots of the tongue, a little above the point where the gullet is attached to the tongue, and since it extends all the way to the lungs and receives the air drawn in by breathing, and breathes that same air out again from the lungs and returns it, it is covered by a kind of lid, so to speak, which was given for this purpose: so that, if some bit of food should chance to fall into it, the breath might not be obstructed. But since the belly, set beneath the gullet, is the receptacle of food and drink, while the lungs and the heart draw in breath from without, in the belly many things have been wrought in a manner deserving of wonder, things composed for the most part of sinews; and it is many-folded and winding, and it encloses and holds whatever it has received, whether dry or moist, so that this may be altered and digested, and now it is drawn tight, now relaxed, and it compresses and blends together everything it has received, so that easily—both by the heat, which it has in abundance, and by the grinding of the food, and besides by the breath—all things, cooked and broken down, may be distributed to the rest of the body. In the lungs, on the other hand, there is a certain porousness and a softness resembling that of sponges, most apt for drawing in the breath, which now contract as they exhale, now dilate as the breath flows in, so that the breath of life may be drawn frequently, by which living creatures are above all nourished.

\ciceroSection{137} From the intestines, however, the juice by which we are nourished, separated in the belly from the rest of the food, seeps through to the liver by certain channels drawn and directed from the middle intestine all the way to the gates of the liver (for so they are called), channels which reach to the liver and adhere to it; and from there other passages run on, through which the food, gliding down from the liver, makes its descent. When the bile has been separated from that food, and likewise the humors which are discharged from the kidneys, the rest turns itself into blood and flows together to those same gates of the liver, to which all its channels reach; and through these the food, having glided on, is in this very place poured into the vein which is called the hollow vein, and through it, now broken down and cooked, it slides on to the heart; and from the heart it is distributed into the whole body through veins exceedingly many, reaching into all the parts of the body.

\ciceroSection{138} How the residue of the food is driven out, now by the intestines drawing themselves tight, now by their relaxing, is by no means difficult to tell, yet it must be passed over, lest the discourse contain anything disagreeable. Let this rather be set forth—the incredible workmanship of nature: for the air which is drawn into the lungs by breathing first grows warm from the breath itself, then by contact with the lungs, and from it one part is given back in exhalation, while another part is taken up by a certain part of the heart which they call the heart's ventricle, to which another like it is joined, into which the blood flows from the liver through that hollow vein. And in this way, from these parts, both the blood is diffused through the veins into the whole body and the breath through the arteries; and both kinds, frequent and many, interwoven throughout the whole body, attest to a certain incredible force of skillful and divine workmanship.

\ciceroSection{139} What shall I say of the bones, which, set beneath the body, have wondrous joinings, fitted both for stability and apt for bounding the limbs and for movement and for every action of the body? To this add the sinews, by which the limbs are held together, and their interlacing, which extends through the whole body, and which, like the veins and arteries, are drawn and set out from the heart and led into the whole body.

\ciceroSection{140} To this providence of nature, so diligent and so resourceful, many things can be added, from which it may be understood how great and how surpassing are the gifts bestowed on men by the gods. First, the gods raised men up from the ground and set them tall and erect, so that, gazing at the sky, they might be able to grasp the knowledge of the gods. For men spring from the earth not as its dwellers and inhabitants, but as it were the spectators of things on high and in the heavens, the contemplation of which belongs to no other kind of living creature. The senses, moreover, the interpreters and messengers of things, have been both made and placed in the head, as in a citadel, wonderfully suited to their necessary uses. For the eyes, like watchmen, occupy the highest position, from which, surveying very many things, they may discharge their office;

\ciceroSection{141} and the ears, since they must perceive sound, which by nature is borne upward, have rightly been placed in the high parts of the body; likewise the nostrils—because every odor rises to the upper regions—are rightly set aloft, and because their judgment of food and drink is of great moment, not without reason have they kept close to the neighborhood of the mouth. And taste, which had to perceive the kinds of the things we eat, dwells in that part of the mouth where nature has opened a passage for food and drink. Touch, however, is spread uniformly over the whole body, so that we can feel all blows and all excessive impingements of cold and heat. And just as in buildings architects turn away from the eyes and noses of the masters those outflowing things that would necessarily have something foul about them, so nature has banished such things far from the senses.

\ciceroSection{142} What craftsman, indeed, other than nature—than which nothing can be more cunning—could have pursued such resourcefulness in the senses? First, she clothed and fenced the eyes with the finest of membranes, which she first made transparent so that one might see through them, yet firm so that they might be held in. But she made the eyes slippery and mobile, so that they might both turn aside if something should harm them and easily direct their gaze wherever they wished; and the very point of vision by which we see, which is called the pupil, is so small that it easily avoids the things that might do it harm; and the eyelids, which are the coverings of the eyes, softest to the touch lest they injure the point of vision, have been most aptly made both for closing the pupils, lest anything fall in, and for opening them, and she has provided that this can be done again and again with the utmost speed.

\ciceroSection{143} And the eyelids are fortified, as with a palisade of hairs, by which, when the eyes are open, anything that might fall in is repelled, and, when in sleep they close together—since we did not then need the eyes for seeing—they might rest as though wrapped up. They lie hidden, besides, to good purpose, and are fenced about on every side by lofty parts. For first the upper portions, overhung by the brows, repel the sweat that flows down from the head and forehead; then the cheeks, set beneath on the lower side and slightly projecting, guard them; and the nose is so placed that it seems to stand as a wall set between the eyes.

\ciceroSection{144} The hearing, however, always lies open, for we have need of its sense even while asleep, and when a sound is received by it we are roused even out of sleep. It has a winding path, lest anything be able to enter if it lay open straight and direct; provision has been made, too, that if some tiny creature should try to break in, it would stick fast in the wax of the ears as though in birdlime. And on the outside project what are called the ears, made both for the sake of covering and protecting the sense, and so that incoming sounds might not slip away and stray before the sense had been struck by them. But their passages of entry are hard and, as it were, horny, and beset with many turnings, because by these natural features sound, when carried in, is amplified; this is why a sound resounds in the lyre from its tortoise-shell or its horn, and why from winding and enclosed places sounds are returned more amply.

\ciceroSection{145} The nostrils too, which must stand open for their indispensable uses, have their entrances rather narrowed, so that nothing harmful can force its way into them; and they always hold a moisture not without value for repelling dust and much else besides. The sense of taste is splendidly fenced in, for it is enclosed within the mouth, conveniently placed both for its use and for the safekeeping of our health. And all the senses of human beings far surpass the senses of the beasts. To begin with the eyes: in those arts whose judgment belongs to the eyes — in shapes painted, modeled, and engraved, and in the very motion and gesture of bodies — they discern much with the finer subtlety; the eyes judge too the loveliness and order of colors and figures and, so to speak, their decorum; and they judge things still greater, for they recognize virtues and vices as well — they recognize the angry man and the kindly, the joyful and the grieving, the brave and the cowardly, the bold and the fearful.

\ciceroSection{146} The judgment of the ears, likewise, is a thing wonderful and full of craft, by which is judged, in the singing of the voice and of pipes and strings, the variety of sounds, their intervals, their distinctness, and the very many kinds of voice — ringing and muffled, smooth and rough, low and high, supple and hard — qualities judged by the ears of human beings alone. The judgments of the nostrils too, and of taste, and of touch, are great. To capture and fully enjoy these senses, more arts have been discovered than I could wish; for it is plain enough to what lengths the compounding of perfumes, the seasoning of foods, and the allurements of the body have advanced.

\ciceroSection{147} And now, as for the soul itself and the mind of man, his reason, his judgment, his prudence — whoever does not perceive that these were brought to perfection by divine care seems to me himself to be destitute of those very things. While I argue this point I could wish your eloquence, Cotta, were given to me. For how you would set forth those matters: first, how great an intelligence is in us, and then how, above all, there is a connecting and a comprehending of consequences — from which, plainly, we judge what follows from each set of premises, and conclude it by reason, and define each thing singly and embrace it within precise bounds; from which we understand what power knowledge has and of what kind it is — a thing than which not even in a god is there anything more excellent. And how great are those faculties which you of the Academy weaken and abolish — that by the senses and by the mind we perceive and grasp the things that lie outside us;

\ciceroSection{148} for it is from these, set side by side and compared with one another, that we also fashion the arts, the ones necessary partly for the use of life and partly for its delight. And now, that mistress of all things, as you are accustomed to call it — the power of speech — how splendid it is and how divine! It is this, first, that enables us both to learn what we do not know and to teach others what we do know; it is by this that we exhort, by this that we persuade, by this that we console the afflicted, by this that we lead the terrified back from their fear, by this that we restrain the overexuberant, by this that we quench desires and angers; it is this that has bound us together in the fellowship of law, of statutes, of cities; it is this that has set us apart from a savage and brutish life.

\ciceroSection{149} As for the use of speech, it is incredible — unless you attend to it carefully — how great a work nature has contrived. For first the windpipe runs from the lungs all the way to the inmost mouth, and through it the voice, taking its origin from the mind, is received and poured forth. Then the tongue is set within the mouth, bounded by the teeth; it shapes and gives limit to the voice when it is poured out without measure, and it produces the distinct and well-defined sounds of speech as it strikes both the teeth and the other parts of the mouth. And so our people are accustomed to call the tongue something like a plectrum, the teeth its strings, the nostrils the horns that resonate to the strings in singing.

\ciceroSection{150} And how apt the hands are, and what handmaids of so many arts, that nature gave to man! For the bending of the fingers is easy, and easy their stretching out, on account of their soft joints and articulations, and they labor in no movement. And so, by the application of the fingers, the hand is fitted for painting, for modeling, for carving, for drawing sounds from strings and pipes. And these are the works of delight, those others of necessity — I mean the tilling of fields and the raising of roofs, the coverings of the body whether woven or sewn, and all the fashioning of bronze and iron. From this it is understood that, by applying the workmen's hands to what was discovered by the mind and perceived by the senses, we have obtained everything — so that we could be sheltered, clothed, and safe, and could have cities, walls, dwellings, and shrines.

\ciceroSection{151} And now, by the works of men, that is, by their hands, variety and abundance of food too is procured. For the fields bring forth many things, sought by the hand, to be either consumed at once or stored away and entrusted to keeping; and besides this we feed on beasts of the land, of the water, and of the air, partly by catching them, partly by rearing them. We also contrive, by our taming of them, the carrying-services of four-footed animals, whose swiftness and strength lend strength and swiftness to ourselves. We lay burdens upon certain beasts, we set yokes upon them; we turn to our own advantage the keenest senses of elephants and the sharp scent of dogs; we draw iron out from the caverns of the earth, a thing necessary for the tilling of fields; we find the deeply hidden veins of bronze, silver, and gold, both serviceable for use and lovely for adornment. And by the cutting of trees, and from every material both cultivated and wild, we make use partly — with fire applied — to warm the body and to soften our food, partly for building, so that, fenced in by roofs, we may repel cold and heat.

\ciceroSection{152} And timber brings great uses for the making of ships, by whose voyages all the supplies needed for life are furnished from every quarter; and of those forces which nature has begotten most violent — the sea and the winds — we alone possess the mastery, through our knowledge of seafaring, and we enjoy and make use of very many things from the sea. Likewise the whole dominion over the goods of the land lies in man: we enjoy the plains, we enjoy the mountains; ours are the rivers, ours the lakes; we sow the crops, we plant the trees; we give fertility to the soil by the leading-in of waters; we dam the rivers, we straighten them, we turn them aside; and with our own hands, in short, we strive to bring about within the world of nature a kind of second nature.

\ciceroSection{153} And what of human reason — has it not penetrated even into the heavens? We alone among living creatures have come to know the risings, the settings, and the courses of the stars; by the race of men the day, the month, the year have been bounded; the eclipses of the sun and moon have been recognized and foretold for all time to come — how great they will be, and when. Contemplating these things, the mind arrives at a knowledge of the gods, from which springs piety, to which justice is joined and the remaining virtues; and from these arises a happy life, the equal and the likeness of the gods, yielding to the celestial beings in no other respect than immortality — which has nothing to do with living well. Now that these matters are set out, I think I have shown sufficiently how far the nature of man surpasses all living things. From which it ought to be understood that neither the shape and arrangement of the limbs nor the power of the intellect and mind could have been wrought by chance.

\ciceroSection{154} It remains for me to show, and at last to bring my speech to its close, that everything in this world which men make use of was made and made ready for the sake of men. In the first place, the world itself was made for the sake of gods and men, and whatever is in it was prepared and devised for the enjoyment of men. For the world is, as it were, the common house of gods and men, or the city of both; for they alone, using reason, live by right and by law. As, then, Athens and Sparta must be reckoned to have been founded for the sake of the Athenians and the Spartans, and everything in those cities is rightly said to belong to those peoples, so whatever is in the whole world must be reckoned to belong to the gods and to men.

\ciceroSection{155} And now the circuits of the sun and moon and of the other heavenly bodies, although they belong also to the coherence of the world, nevertheless also afford a spectacle to men; for there is no sight more insatiable, none more beautiful, none more excelling in reason and craftsmanship; for, by measuring out their courses, we have come to know the ripenings of the seasons and their variations and changes. And if these things are known to men alone, it must be judged that they were made for the sake of men.

\ciceroSection{156} And the earth, teeming with crops and with every kind of pulse, which she pours out with the greatest bounty — does she seem to give them birth for the sake of the beasts, or of men? What shall I say of the vines and the olive groves, whose richest and most abundant fruits have nothing whatever to do with the beasts? For the cattle have no knowledge of sowing, or of cultivating, or of reaping and gathering the fruits in due season, or of storing and laying them up; the use and care of all these things belong to men.

\ciceroSection{157} As, then, the lyre and the pipes must be said to have been made for the sake of those who could use them, so the things I have spoken of must be granted to have been prepared for those alone who use them; nor, if certain beasts steal or snatch some of them, shall we say that they were brought into being for the sake of those beasts as well. For men do not store up grain for the sake of mice or ants, but for the sake of their wives and children and households; and so the beasts enjoy it by stealth, as I said, while the masters enjoy it openly and freely.

\ciceroSection{158} It must therefore be admitted that these stores of things were provided for the sake of men. Unless perhaps so great an abundance and variety of fruits, and the pleasantness not only of tasting them but of smelling and seeing them as well, leaves room for any doubt that nature has bestowed them on men alone. And so far is it from the truth that these things were prepared for the sake of the beasts as well, that we see the very beasts themselves were begotten for the sake of men. For what else do sheep provide, except that men may be clothed when their wool has been worked and woven? Indeed they could neither have been nourished nor sustained nor have yielded any produce of themselves without the cultivation and care of men. And as for the dog — its loyalty so faithful in keeping watch, its fawning affection so loving toward its masters, its hatred so great toward strangers, the incredible keenness of its nose for tracking, its great eagerness in the hunt — what does all this signify but that they were begotten for the conveniences of men?

\ciceroSection{159} What shall I say of oxen? Their very backs declare that they were not shaped for receiving a burden, while their necks were born for the yoke, and the strength and breadth of their shoulders for dragging the plow. By these, when the earth was being broken up by the splitting of the clods — in that golden age, as the poets put it — no violence was ever done; but then the iron race sprang up, and was the first that dared to forge the deadly sword and to taste the bullock, joined to it and tamed by its hand: so great was the usefulness reckoned to be reaped from oxen, that it was held a crime to feed upon their flesh. It would be a long task to run through the uses of mules and asses, which were certainly provided for the use of men.

\ciceroSection{160} And the pig — what has it, beyond being food? To it, indeed, Chrysippus says that the very soul was given in place of salt, to keep its flesh from rotting; and of this animal, since it was fit for the eating of men, nature has begotten nothing more prolific. What shall I say of the multitude and the delightfulness of fish, what of birds? From these so great a pleasure is reaped that at times our Stoic Providence seems to have been an Epicurean; and they could not even be caught except by the reason and craftsmanship of men — although certain birds, both the birds of flight and the birds of song, as our augurs call them, we believe were born for the sake of taking auguries.

\ciceroSection{161} Then too we take huge and savage beasts by hunting, both to feed on them and to train ourselves in the chase after the likeness of military discipline, and to put the tamed and well-schooled ones to use, as we do elephants; and we draw from their bodies many remedies for diseases and wounds, just as we do from certain roots and herbs, whose usefulness we have learned by the practice and trial of long ages. You may survey the whole earth and all the seas with your mind as with your eyes: you will see now the fruitful and measureless expanses of the plains and the dense mantling of the mountains, the pastures of the herds, and then the voyages over the sea accomplished with incredible speed.

\ciceroSection{162} And not above the earth only, but even in its innermost darkness there lies hidden the usefulness of very many things, which, having come into being for the use of men, are discovered by men alone. There is a point, however, which each of you will perhaps seize upon to censure — Cotta because Carneades took such pleasure in inveighing against the Stoics, Velleius because Epicurus laughs at nothing so much as the foretelling of things to come — yet this very point seems to me to confirm beyond anything else that human affairs are looked after by the providence of the gods. For there is, surely, such a thing as divination, which shows itself in many places, in many matters, at many times, in private concerns and above all in public ones:

\ciceroSection{163} the haruspices discern much, the augurs foresee much, much is declared by oracles, much by prophecies, much by dreams, much by portents; and through the knowledge of these, many results have often been secured in accordance with men's wishes and to their advantage, and many dangers have been averted as well. This power, then — whether you call it a power or an art or a natural endowment — has surely been granted to man, and to no other creature, by the immortal gods, for the knowledge of things to come. And if these proofs taken singly do not perhaps move you, taken all together, linked and joined to one another, they ought certainly to have moved you.

\ciceroSection{164} And it is not the whole race of men only, but individuals as well, for whom the immortal gods are wont to provide and take thought. For one may draw in the whole community of the human race and reduce it step by step to a smaller number, and at last to single men. For if we judge that the gods take thought for all men, wherever they are, on whatever shore or quarter of the earth, even those separated from the continuous landmass of this earth which we inhabit — and this for the reasons we have given before — then they take thought also for these men who, together with us, dwell in these lands of ours from the rising to the setting of the sun.

\ciceroSection{165} But if they take thought for those who dwell upon what is, as it were, a great island, which we call the circle of the earth, then they also take thought for those who hold the parts of that island, Europe, Asia, and Africa. And so they cherish the parts of these regions as well, such as Rome, Athens, Sparta, and Rhodes; and within those cities they cherish individuals apart from the whole — as, in the war with Pyrrhus, Curius, Fabricius, and Coruncanius; in the first Punic War, Calatinus, Duellius, Metellus, and Lutatius; in the second, Maximus, Marcellus, and Africanus; after these, Paulus, Gracchus, and Cato; and within our fathers' memory, Scipio and Laelius. Many other men of singular worth our own state and Greece besides have produced, and we must believe that none of them was such as he was without the help of a god.

\ciceroSection{166} It was this consideration that drove the poets, and above all Homer, to attach particular gods, as companions in their crises and perils, to the foremost of the heroes — to Ulysses, Diomedes, Agamemnon, and Achilles. Besides this, the visible presence of the gods themselves, of the kind I have mentioned above, makes plain that thought is taken by them both for states and for individual men — a fact understood also through the signs of things to come, which are foreshown sometimes to men asleep, sometimes to men awake. And we are warned, besides, by many prodigies, by many tokens in the entrails of victims, and by many other things which long practice has so marked down that it has produced the art of divination.

\ciceroSection{167} No man, then, was ever great without some breath of the divine. Yet this is not to be refuted in such a way that, if a storm has harmed someone's crops or vineyards, or if chance has stripped away some one of life's comforts, we should judge the man to whom any of these things happened to be either hateful to a god or neglected by a god. The gods attend to great things; the small they overlook. But for great men all things are always prosperous — that is, if enough has been said by our school and by the foremost of philosophers, Socrates, about the abundance and the riches of virtue.

\ciceroSection{168} These, roughly, were the things that came to my mind which I thought ought to be said about the nature of the gods. As for you, Cotta — if you would take my advice, you would plead this same cause, and would consider that you are both a leading citizen and a pontiff, and, since you are free to argue on either side, you would rather take this one, and would bring to it that faculty of discourse which the Academy has enlarged for you out of what you received from your training in rhetoric. For it is a wicked and impious habit to argue against the gods, whether it is done in earnest or only in pretense.''

\ciceroBook{3}

\ciceroSection{1} When Balbus had finished, Cotta said with a smile: "You instruct me too late, Balbus, on what I am to defend. For while you were speaking, I was turning over in my own mind what I might say against you — not so much to refute you as to ask after the things I understood too little. And since each man must use his own judgment, it would be a hard thing to bring me to feel as you would have me feel."

\ciceroSection{2} At this Velleius said: "You have no idea how keenly I look forward to hearing you, Cotta. For our friend Balbus found your discourse against Epicurus delightful; so I in turn will offer myself to you as an attentive listener against the Stoics. For I trust you come, as you usually do, well prepared."

\ciceroSection{3} Then Cotta said: "By Hercules, Velleius, I do — for my reckoning with Lucilius is not the same as it was with you." "How so, then?" said the other. "Because it seems to me that your Epicurus does not put up much of a fight over the immortal gods. He merely does not dare to deny that the gods exist, for fear of incurring some odium or charge; but when he asserts that the gods do nothing and care for nothing, and that they are furnished with human limbs but make no use of those limbs, he seems to be trifling — content to have said that there is a certain blessed and eternal nature.

\ciceroSection{4} From Balbus, on the other hand, you noticed, I am sure, how many things were said, and how — even if less than true — they were nonetheless apt and consistent with one another. And so I intend, as I said, not so much to refute his discourse as to ask after the things I understood too little. So I leave it to you, Balbus: would you rather answer me as I question you on individual points, the things I have grasped too little, or hear my discourse entire?" Then Balbus said: "For my part, if you want anything explained to you, I would rather answer; but if you mean to question me not so much to understand as to refute, I will do whichever you please — either reply at once to each point you raise, or, when you have finished, to all of them together."

\ciceroSection{5} Then Cotta said: "Excellent. So let us proceed as the discourse itself leads us. But before the matter at hand, a few words about myself. I am moved no slightly by your authority, Balbus, and by that part of your discourse which, in your peroration, urged me to remember that I am both a Cotta and a pontiff — which, I believe, came to this: that I should defend the beliefs we have received from our ancestors concerning the immortal gods, the rites, the ceremonies, the observances of religion. These I will indeed defend always, as I have always defended them, and no man's discourse, learned or unlearned, will ever move me from the belief I received from my ancestors concerning the worship of the immortal gods. But when religion is the question, I follow Tiberius Coruncanius, Publius Scipio, Publius Scaevola, men who were chief pontiffs — not Zeno or Cleanthes or Chrysippus; and I have Gaius Laelius, an augur and at the same time a wise man, whom I would sooner hear speaking on religion, in that famous discourse of his, than any leader of the Stoics. And since the whole religion of the Roman people is divided into rites and auspices, with a third thing joined to them — whatever the interpreters of the Sibyl or the soothsayers have advised by way of prediction from portents and prodigies — I have always held that none of these observances is to be despised, and I have convinced myself that Romulus by establishing the auspices, and Numa by establishing the rites, laid the foundations of our state, which surely could never have grown so great without the fullest appeasement of the immortal gods.

\ciceroSection{6} There you have, Balbus, what Cotta thinks, what a pontiff thinks. Now let me understand what you think; for from you, a philosopher, I am bound to receive a reasoned account of religion, whereas in our ancestors I must believe even when no account is rendered." Then Balbus said: "What account, then, do you want from me, Cotta?" And he replied: "Your division was fourfold: first, you proposed to show that the gods exist; next, of what kind they are; then, that the world is governed by them; and last, that they take thought for human affairs. This, if I remember rightly, was the partition." "Quite right," said Balbus; "but I am waiting to hear what you would ask."

\ciceroSection{7} Then Cotta said: "Let us look at each point in turn. And as for the first — if it is first, the thing on which all but the wholly impious agree, and which cannot be burned out of my mind, namely that the gods exist — that very thing, of which I am convinced by the authority of my ancestors, why it should be so, you teach me nothing." "What is this?" said Balbus. "If you are convinced of it, why would you have me teach you?" Then Cotta said: "Because I approach this discussion as though I had never heard anything about the immortal gods, never given them a thought. Take me as a raw and untouched pupil, and teach me the things I ask."

\ciceroSection{8} "Tell me, then," he said, "what you ask." "Why, this first: when you had said that on that side the matter is so plain it does not even need a discourse — that the gods exist being plain and agreed by all — why on that very point you said so much." "Because I have often noticed you too, Cotta," he said, "when you spoke in the Forum, loading the juror with as many arguments as you could, if only the case gave you the means. And philosophers do this same thing, and I have done it as I could. But what you ask is much as if you were to ask me why I look at you with two eyes and do not close the other, when I could attain the same end with one." Then Cotta said: "How apt that comparison is, you may judge for yourself.

\ciceroSection{9} For I am not in the habit, in lawsuits, of arguing for anything that is self-evident and agreed by all — argumentation only weakens what is clear — and even if I did so in the courts, I would not do the same in this finer kind of discourse. As for closing one eye, there would be no cause for that, since the gaze of both is the same, and since nature — which you would have to be wise — willed that we should have two channels bored from the mind to the eyes for light. But because you had no confidence that the thing was as plain as you would have it, for that reason you set out to show by many arguments that the gods exist. For me one thing was enough: that our ancestors had handed it down so. But you despise authorities and fight with reason;

\ciceroSection{10} allow my reason, then, to contend with your reason. You bring forward all these arguments why there should be gods, and by your arguing you make doubtful a matter that in my view is not doubtful in the least; for I have committed to memory not only the number of your arguments but their order as well. The first was that, when we have looked up at the sky, we at once understand that there is some divine power by which all this is governed. From this came also that line: 'Behold this shining vault on high, which all

\ciceroSection{11} invoke as Jupiter' — as if any of us called upon that Jupiter rather than the Jupiter of the Capitol, or as if this were plain and agreed by all, that those are gods which Velleius and many besides will not even grant to be living beings. A weighty argument too you thought it, that the belief in the immortal gods belongs to all men and grows greater every day: so it pleases you, then, that matters so great should be judged by the opinion of fools — and that with you above all, who say those very men are mad? 'But we see the gods present among us, as Postumius did at Lake Regillus, and Vatinius on the Salarian Way' — and there is some tale, too, about the Locrians' battle at the Sagra. Those, then, whom you called the Tyndaridae — that is, men born of a man — and whom Homer, who lived not long after their age, says were buried at Lacedaemon: do you suppose that these came to meet Vatinius on white horses, with no grooms, and announced the victory of the Roman people to Vatinius, a country fellow, rather than to Marcus Cato, who was then the leading man? And do you, in the same way, believe that the mark which appears today in the flint at Regillus is, as it were, the print of a hoof from Castor's horse?

\ciceroSection{12} Would you not rather believe what can be proved — that the souls of illustrious men, such as those Tyndaridae were, are divine and eternal — than that men once cremated could ride horses and fight in the line of battle? Or, if you say this could have happened, you ought to teach how, and not bring out old wives' tales."

\ciceroSection{13} Then Lucilius said: "Do they seem to you mere tales? Do you not see the temple to Castor and Pollux dedicated in the Forum by Aulus Postumius, and the decree of the Senate concerning Vatinius? As for the Sagra, there is even a common proverb among the Greeks, who, when they affirm something, call it 'surer than the things at the Sagra.' Ought you not, then, to be moved by such authorities?" Then Cotta said: "It is with rumors that you fight me, Balbus, while I ask reasons of you. . . . What follows is what is to come;

\ciceroSection{14} for no one can escape what is to be. And often it is not even useful to know what will be; for it is a wretched thing to be tormented to no purpose, and not to have even the last consolation common to all, namely hope — especially since you yourselves say that all things come about by fate, and that what has been true forever from all eternity is fate. What good is it, then, or what does it bring toward taking precaution, to know that something will be, when it certainly will be? And again — where does that divination of yours come from? Who discovered the cleft in a liver, who marked the cry of the crow, who the lots? In these I believe, and I cannot despise the staff of Attus Navius whom you brought up; but how these things came to be understood by the philosophers, that I must learn — especially since those diviners of yours lie about a great many matters.

\ciceroSection{15} 'But physicians too' — for so you were saying — 'are often deceived.' What likeness is there between medicine, whose method I see, and divination, whose source I do not understand? You hold besides that the gods were appeased by the self-devotions of the Decii. What was their injustice so great that they could not be appeased toward the Roman people unless such men had perished? That was a general's stratagem — what the Greeks call \textit{[Greek: stratēgēma]} — but the stratagem of generals who took thought for their country and did not spare their own lives; for they reckoned that the army would follow a general who, spurring his horse, hurled himself against the enemy, and so it fell out. As for the voice of Faunus, I myself have never heard it; if you say that you have heard it, I will believe you, though I have no idea at all what a Faunus is. So far, then, Balbus, at least so far as it rests with you, I do not understand that the gods exist — gods whom I for my part believe to exist, but the Stoics teach me nothing.

\ciceroSection{16} For Cleanthes, as you were saying, holds that the notions of the gods are formed in the minds of men in four ways. The first is the one I have spoken of enough, the one taken from the foreknowledge of things to come; the second, from the disturbances of the weather and the other motions; the third, from the convenience and abundance of the things we enjoy; the fourth, from the order of the stars and the constancy of the heavens. Of foreknowledge we have spoken. Of disturbances in sky, sea, and earth we cannot say, when these occur, that there are not many who fear them and suppose them to be brought about by the immortal gods;

\ciceroSection{17} But that is not the question — whether there are some people who think the gods exist; the question is whether the gods exist or not. For the remaining grounds that Cleanthes brings forward — one of them concerning the abundance of advantages we receive, another concerning the ordering of the seasons and the constancy of the heavens — these I shall take up when we come to dispute about the providence of the gods, on which a great deal was said by you, Balbus;

\ciceroSection{18} and to that same place I shall also defer the point you said Chrysippus made — that since there is something in the natural world which cannot be brought about by man, there exists something better than man — together with the comparisons you drew between the beauty of a fine house and the beauty of the world, and the case you made for the harmony and concord of the whole world. Zeno's brief and rather clever syllogisms, too, I shall defer to that part of the discussion I have just named; and at that same time all those things you stated as a natural philosopher — about the fiery force and about that heat from which you said all things are generated — shall be examined in their proper place; and everything you said the day before yesterday, when you wished to prove the gods exist, as to why the whole world and the sun and the moon and the stars should possess sensation and mind, all of this I shall reserve for the same occasion.

\ciceroSection{19} But of you, again and again, I shall ask this one same thing: by what reasonings do you persuade yourself that the gods exist?'' Then Balbus said: For my part I think I have brought forward my reasonings; but you refute them in such a way that, just when you seem about to question me and I have made myself ready to answer, you suddenly turn your discourse aside and give me no room to reply. And so matters of the greatest weight have passed by in silence — divination, fate — questions on which you indeed speak only in passing, but on which our school is accustomed to say a great deal, though they are kept distinct from the very question now in hand. So, if you please, do not proceed in a confused manner, so that in this disputation we may unravel the thing that is in question. Excellent, said Cotta.

\ciceroSection{20} ``And so, since you divided the whole question into four parts, and we have spoken of the first, let us consider the second; which seemed to me to be of this kind — that, when you wished to show what sort of beings the gods are, you showed there are none at all. You said it was extremely difficult to draw the mind away from the habit of the eyes; yet, since nothing is more excellent than a god, you did not doubt that the world is a god, since there is nothing better than it in the natural world — if only we could conceive of it as a living being, or rather discern this with the mind, as we discern other things with the eyes.

\ciceroSection{21} But when you deny that anything is better than the world, what do you mean by better? If more beautiful, I agree; if better fitted to our advantages, that too I agree; but if you mean that nothing is wiser than the world, I do not in the least agree — not because it is difficult to call the mind away from the eyes, but because the more I call it away, the less I can grasp with the mind what you wish me to. ``Nothing in the natural world is better than the world.'' Not even on earth is anything better than our own city; do you therefore think that on that account reason, thought, and mind reside in the city? Or, since they do not, do you on that account judge that an ant must be set above this most beautiful city, on the ground that in a city there is no sensation, while in an ant there is not only sensation but mind, reason, and memory? You ought to see, Balbus, what is granted to you, and not yourself assume what you would like.

\ciceroSection{22} For that whole position was opened out by that old, brief, and — as it seemed to you — clever syllogism of Zeno's. For Zeno argues thus: ``That which uses reason is better than that which does not use reason; but nothing is better than the world;

\ciceroSection{23} therefore the world uses reason.'' If this is acceptable, you will soon make the world appear to read a book superbly; for following in Zeno's footsteps you will be able to draw your conclusion in this fashion: ``That which is lettered is better than that which is not lettered; but nothing is better than the world; therefore the world is lettered'' — and in that manner the world is also eloquent, and indeed a mathematician, a musician, finally trained in every branch of learning, and at last a philosopher. You said often that nothing comes to be without a god, and that there is no power in nature able to fashion things unlike itself: shall I grant that the world is not only living and wise but also a lyre-player and a trumpeter, since men of those arts too are brought forth from it? Then that father of the Stoics offers nothing to make us think the world uses reason, nor even that it is a living being. The world, therefore, is not a god; and yet there is nothing better than it: for there is nothing more beautiful than it, nothing more wholesome to us, nothing more adorned to look upon or more constant in its motion. But if the world as a whole is not a god, neither are the stars, which you placed, beyond counting, in the number of the gods. Their even and eternal courses delighted you — and by Hercules, not without cause, for they move with an admirable and incredible constancy.

\ciceroSection{24} But not all things, Balbus, that have fixed and constant courses are to be ascribed to a god rather than to nature. What do you suppose can be more constant than the Chalcidian Euripus in its repeated ebb and flow back and forth? Than the Sicilian strait? Than the seething of the Ocean in those regions, ``where the ravening wave divides Europe from Libya''? Are not the sea-tides, whether of Spain or of Britain, with their advances and retreats at fixed times, able to come about without a god? Consider, I beg you: if we call divine all motion and all things that keep their order at fixed times, must we not say that tertian and quartan fevers too are divine — for what can be more constant than their recurrence and movement? But of all such things an account must be rendered;

\ciceroSection{25} and when you cannot do this, you flee to a god as though to an altar. And Chrysippus seemed to you to speak shrewdly — a man without doubt versatile and shrewd (I call those versatile whose mind turns quickly, and those shrewd whose mind has grown hardened by use, as a hand grows hardened by work); so he says: ``If there is something that man cannot bring about, then whatever brings it about is better than man; but man cannot bring about the things that are in the world; therefore whoever was able to do so surpasses man; but who could surpass man except a god? Therefore a god exists.'' All of this turns upon the same error as that argument of Zeno's.

\ciceroSection{26} For what is better, what is more excellent, what is the difference between nature and reason — none of this is distinguished. And the same man, if the gods do not exist, denies that there is anything in all of nature better than man; yet for any human being to think that nothing is better than man he judges to be the height of arrogance. Let it be the mark of an arrogant man to think more of himself than of the world; but to understand that he himself has sensation and reason, while Orion and the Dog-Star do not — that is the mark not of an arrogant man but rather of a prudent one. And ``If a house is beautiful,'' he says, ``let us understand it was built for its masters and not for mice; so, then, we ought to judge the world to be the house of the gods.'' I would judge precisely so, if I thought it built, and not — as I shall show — shaped by nature.

\ciceroSection{27} But, says he, Socrates in Xenophon asks where we should have caught up our soul if there were none in the world. And I ask where we caught up speech, where numbers, where song — unless indeed we suppose the sun talks with the moon when it has drawn nearer, or that the world sings in harmony, as Pythagoras believes. These things, Balbus, belong to nature — nature that walks not artfully, as Zeno says (and we shall presently see just what that claim amounts to), but that sets all things in motion and stirs them by its own movements and changes.

\ciceroSection{28} And so that discourse of yours about the harmony and concord of nature pleased me — nature which you said breathes together, as if held continuous by a kind of kinship; but what I did not approve was your denial that this could come about unless nature were held together by one divine breath. In fact it coheres and endures by the powers of nature, not of the gods, and there is in it that kind of concord which the Greeks call \textit{[Greek: sympatheian]}; but the more it arises of its own accord, the less it is to be thought to come about by divine reason.

\ciceroSection{29} And those arguments which Carneades used to bring forward — how do you dissolve them? If no body is immortal, no body is everlasting; but no body is immortal, not even an indivisible one, nor one that cannot be sundered and torn apart; and since every animate being has a nature capable of being acted upon, there is none of them that escapes the necessity of receiving something from outside — that is, as it were, of bearing and suffering it; and if every animate being is of this kind, none is immortal. So likewise, if every animate being can be cut and divided, none of them is indivisible, none eternal; but every animate being is constituted to receive and bear an external force; therefore every animate being must of necessity be mortal, dissoluble, and divisible.

\ciceroSection{30} For just as, if all wax were changeable, there would be nothing waxen that could not be changed, and likewise nothing silver, nothing bronze, if the nature of silver and bronze were changeable — so, in like manner, if all the things that exist, out of which all things are composed, are changeable, then there can be no body that is not changeable; but those things out of which all things are composed are changeable, as it seems to you; therefore every body is changeable. But if there were any immortal body, not every body would be changeable; and so it follows that every body is mortal. For every body is either water or air or fire or earth, or that which is compounded out of these, or out of some part of them.

\ciceroSection{31} But there is none of these that does not perish; for everything earthen is divided, and moisture is so soft that it can easily be pressed and crushed together; fire indeed and air are most easily driven off by any blow, and are by nature most yielding and most readily dispersed. And besides, all these perish whenever they are turned into another nature, which happens when earth turns itself into water, and when from water air arises, and from air aether, and when these same elements in their turn travel back again. But if it is so, that those things perish out of which every animate being is composed, then no animate being is everlasting.

\ciceroSection{32} And even if we leave all this aside, still no animate being can be found that was never born at any time and will always exist in the future. For every animate being has sensation; it senses, therefore, things hot and cold, sweet and bitter, nor can it through any sense receive what is pleasant and not receive the contrary; if, then, it takes in the sensation of pleasure, it takes in that of pain as well; but whatever receives pain must of necessity also receive destruction; therefore every animate being must be admitted to be mortal. Besides, if there is anything that feels neither pleasure nor pain, that thing cannot be an animate being;

\ciceroSection{33} But if, on the other hand, because a thing is a living creature it must therefore have sensation, then because it has sensation it cannot be eternal; and every living creature has sensation; no living creature, therefore, is eternal. Besides, there can be no living creature in which there is not both a natural appetite and a natural aversion. Now the things that are sought are those that accord with nature, the things that are avoided are their opposites; and every living creature seeks certain things and flees from certain others, and what it flees from is contrary to its nature, and what is contrary to its nature has the power to destroy it. Every living creature, therefore, must of necessity perish.

\ciceroSection{34} The grounds are innumerable from which it can be established and forced upon us that nothing which possesses sensation is exempt from perishing. For those very things that are felt — cold, heat, pleasure, pain, and the rest — destroy when they are intensified; and no living creature is without sensation; no living creature, therefore, is eternal. And indeed, the nature of a living thing is either simple — earthy, say, or fiery, or airy, or watery, what such a thing would be cannot even be conceived — or else compounded of several natures, of which each has its own place toward which it is borne by natural force, one to the lowest, another to the highest, another to the middle. These can cohere for a certain span of time, but in no way can they cohere forever; for each must of necessity be swept by nature into its own place. No living creature, therefore, is everlasting.

\ciceroSection{35} But your school, Balbus, is in the habit of referring everything to the power of fire — following Heraclitus, I suppose, whom not everyone interprets in one and the same way; since what he meant he was unwilling to have understood, let us pass him over. You, however, put it thus: that all force is fire, and accordingly that living things too perish when their heat has failed, and that throughout the whole of nature the thing that lives and the thing that thrives is the thing that is warm. I, for my part, do not understand how bodies perish when their heat is extinguished but do not perish with the loss of moisture or breath — especially since they perish from excessive heat as well.

\ciceroSection{36} On that score, then, the point about the warm is held in common; but let us nonetheless see where it comes out. You hold, I believe, that there is no living thing outside in nature and in the universe except fire: why fire any more than breath, of which the very mind of living things is composed, from which the word "living" is derived? But how do you take it as a thing all but conceded that the mind is nothing but fire? For it seems more probable that the mind is something of this kind: a tempering of fire with breath. "But if fire by itself is a living creature, with no other nature mingled in, then since it, when it is present in our bodies, makes us capable of sensation, it cannot itself be without sensation." Again the same things can be said: for whatever it is that possesses sensation must of necessity feel both pleasure and pain, and the thing to which pain comes also comes to destruction. So it turns out that you cannot make even fire eternal.

\ciceroSection{37} Why, is it not the view of your very own school that all fire needs nourishment and can in no way endure unless it is fed, and that the sun, the moon, and the rest of the stars are fed by waters, some by fresh and some by the sea's? And Cleanthes adduces this as the reason why the sun turns back and goes no farther than the summer circle and likewise the winter one: so that it should not stray too far from its food. What this whole notion amounts to we shall see shortly; for now let this be the conclusion: that what can perish is not eternal by nature; that fire will perish unless it is fed; that fire, therefore, is not by nature everlasting.

\ciceroSection{38} And what sort of god can we conceive who is endowed with no virtue? Why, shall we attribute prudence to a god — prudence, which consists in the knowledge of things good and bad and neither good nor bad? To one who has, and can have, no evil, what need is there of choosing between good and bad, what need of reason, what of intelligence — which we use to attain the obscure by means of the plain? But to a god nothing can be obscure. As for justice, which assigns to each his own, what has it to do with the gods? For it is the fellowship and community of men, as you yourselves say, that begot justice. And temperance consists in foregoing the pleasures of the body: if there is room in heaven for temperance, there is room for pleasures too. For how can a god be conceived as brave — in pain, or in toil, or in danger?

\ciceroSection{39} None of which touches a god. How, then, can we conceive a god who neither employs reason nor is endowed with any virtue? And indeed I cannot despise the ignorance of the common crowd and the unlearned when I consider what is said by the Stoics. For those are the beliefs of the ignorant: the Syrians venerate a fish, the Egyptians have consecrated nearly every kind of beast; and in Greece, again, they have many gods who were once men — Alabandus at Alabanda, Tenes at Tenedos, Leucothea, who was once Ino, and her son Palaemon throughout all Greece — and Hercules, Aesculapius, the sons of Tyndareus, our own Romulus, and a good many others besides, whom they suppose to have been received into heaven as new and enrolled citizens, so to speak.

\ciceroSection{40} So much, then, for the unlearned; what of you philosophers, who do better? I pass over those finer doctrines, for they are splendid: let the world itself, by all means, be a god. This, I take it, is what is meant by "that radiance on high, whom all men invoke as Jove." Why, then, do we add more gods? And how great is their multitude — to me, at any rate, they seem quite many; for you count the single stars as gods and call them either by the names of beasts, like the Goat, the Scorpion, the Bull, the Lion, or by the names of inanimate things, like the Argo, the Altar, the Crown.

\ciceroSection{41} But even granting these, how on earth can the rest not merely be granted but understood at all? When we call grain Ceres and wine Liber, we are using a customary turn of speech; but do you suppose anyone so out of his mind as to believe that the thing he eats is a god? As for those whom you say have passed from being men to being gods, you will render an account of how that could have come about, or why it has ceased to come about, and I shall be glad to learn; for as the case now stands, I do not see how the man for whom "torches were brought to Mount Oeta," as Accius says, passed from that blaze "into the eternal house of his father." Yet Homer has him met among the dead by Ulysses, just like the others who had departed from life.

\ciceroSection{42} Though indeed which Hercules above all we ought to worship I should very much like to know; for those who probe the deeper and more recondite writings hand down to us several. The most ancient was born of Jove — but of Jove the most ancient as well, for in the old writings of the Greeks we find several Joves too: of him, then, and of Lysithoe is that Hercules whom we are told contended with Apollo over the tripod. A second is handed down as Egyptian, born of the Nile, who they say composed the Phrygian letters. A third is from the Idaean Dactyls, to whom they bring funeral offerings. A fourth is the son of Jove and of Asteria, sister of Latona; he is worshipped above all at Tyre, and they hold Carthage to be his daughter. A fifth is in India, who is called Belus. A sixth is this one, born of Alcmena, whom Jupiter begot — but the third Jupiter, since, as I shall presently show, we have received several Joves as well.

\ciceroSection{43} Now that the discourse has brought me to this point, I shall show that I have learned better things about worshipping the immortal gods by pontifical law and the custom of our ancestors — from these little sacrificial bowls that Numa left us, of which Laelius speaks in that golden little oration of his — than from the reasonings of the Stoics. For if I follow you, tell me what I am to answer the man who questions me thus: "If there are gods, are the Nymphs goddesses too? If the Nymphs, then the little Pans and the Satyrs as well; but these are not gods; the Nymphs, then, are not goddesses either. Yet their temples have been publicly vowed and dedicated. Then neither are the rest gods, whose temples have been dedicated. Go further: you count Jupiter and Neptune as gods; therefore Orcus, their brother, is also a god, and those who are said to flow in the underworld — Acheron, Cocytus, Pyriphlegethon — and then Charon and then Cerberus must be reckoned gods.

\ciceroSection{44} But that, surely, must be rejected; then Orcus is no god either; what, then, do you say about his brothers?" This was what Carneades used to say — not in order to do away with the gods, for what could be less becoming to a philosopher, but to convince the Stoics that they explain nothing about the gods. And so he pressed on: "For tell me," he said, "if these brothers are in the number of the gods, surely it cannot be denied of their father Saturn, whom the common people worship above all in the West? And if he is a god, it must be acknowledged that his father Caelus is a god too. And if that is so, then the parents of Caelus must also be held to be gods, Aether and Dies, and their brothers and sisters, who are named thus by the ancient genealogists: Love, Guile, Toil, Envy, Fate, Old Age, Death, Darkness, Misery, Lamentation, Favor, Fraud, Stubbornness, the Fates, the Hesperides, Dreams — all of whom they say were born of Erebus and Night. Either, then, these monstrosities must be approved, or those first gods must be done away with.

\ciceroSection{45} Will you say that Apollo, Vulcan, Mercury, and the rest are gods, yet hesitate about Hercules, Aesculapius, Liber, Castor, Pollux? But these are worshipped just as much as the others — among some peoples, indeed, far more. These, then, must be held to be gods, though born of mortal mothers. What of Aristaeus, who is said to be the discoverer of the olive, son of Apollo; Theseus, son of Neptune; the rest whose fathers are gods — will they not be in the number of the gods? And what of those whose mothers are goddesses? More so, I think; for just as under civil law a man whose mother is free is free, so by the law of nature one who has a goddess for a mother must be a god. And so the islanders of Astypalaea worship Achilles most devoutly; and if he is a god, then Orpheus and Rhesus are gods too, born of a Muse for their mother — unless perhaps marriages of the sea are to be preferred to those of the land. If these are not gods because they are nowhere worshipped, how are the others gods?

\ciceroSection{46} Consider, then, whether these honors are paid to the virtues of men and not to their immortality — which you too, Balbus, seemed to be saying. But how can you, if you think Latona a goddess, refuse to think Hecate one, who is the daughter of Asteria, Latona's sister? Is she too, then, a goddess? For we have seen her altars and shrines in Greece. And if she is a goddess, why not the Eumenides? And if these are goddesses — whose sanctuary there is at Athens, and among us, as I interpret it, the grove of Furina — then the Furies are goddesses, watchers, I suppose, and avengers of crime and wickedness.

\ciceroSection{47} But if the gods are of such a kind that they take part in human affairs, then Natio too must be held a goddess, to whom we are accustomed to perform divine rites when we go round the shrines in the territory of Ardea; she is named Natio because she watches over the childbearing of matrons, from those being born. If she is a goddess, then all those gods you were enumerating are gods — Honor, Faith, Mind, Concord — and so too Hope, Moneta, and everything we can fashion for ourselves in thought. But if that is not plausible, then neither is the source from which these flowed. And what do you say to this: if those whom we worship and have received are gods, why do we not reckon Serapis and Isis in the same class? And if we do that, why should we reject the gods of the barbarians? Oxen, then, and horses, ibises, hawks, asps, crocodiles, fish, dogs, wolves, cats, and many beasts besides we shall put back into the number of the gods. But if we reject these, we shall also reject those from which they sprang.

\ciceroSection{48} What next — shall Ino be reckoned a goddess, and be called Leucothea by the Greeks and Matuta by us, when she is the daughter of Cadmus, while Circe and Pasiphae and Aeetes, born of Perseis the daughter of Ocean and fathered by the Sun, shall not be held in the number of the gods? And yet Circe too is religiously worshipped by our colonists at Circeii. So you reckon her a goddess: what will you answer about Medea, who was born of two gods for grandfathers, the Sun and Ocean, with Aeetes for her father and Idyia for her mother? What about her brother Absyrtus — who in Pacuvius is Aegialeus, but the other name is the more usual in the writings of the ancients? If these are not gods, I fear for the standing of Ino; for all these flowed from the very same spring.

\ciceroSection{49} Will Amphiaraus and Trophonius, then, be gods? Our own tax-farmers, when lands in Boeotia that had been exempted by censorial law as belonging to the immortal gods came into question, denied that any beings were immortal who had once been men. But if these are gods, then surely Erechtheus is one too, whose shrine and whose priest I have myself seen at Athens. And if we make him a god, what doubt can we have about Codrus, or about all the others who fell fighting for their country's freedom? But if that is not credible, then neither are the earlier premises from which these consequences flow to be accepted.

\ciceroSection{50} And in most communities it can be seen that, for the sake of fostering virtue — so that every man of the best sort might face danger more willingly for the commonwealth's sake — the memory of brave men was consecrated with the honor due to the immortal gods. It is for that very reason that Erechtheus at Athens and his daughters are counted among the gods; and likewise there is a shrine at Athens called the Leokorion. The people of Alabanda, indeed, worship Alabandus, the founder of their city, more devoutly than any of the renowned gods; and it was among them that Stratonicus made one of his many witty remarks, when a certain tiresome fellow kept insisting to him that Alabandus was a god and Hercules was not. "Very well, then," he said,

\ciceroSection{51} "let Alabandus be angry with me, and Hercules with you." But as for those gods you derived from the heavens and the stars, Balbus — you do not see how far the thing creeps. The sun is a god, you say, and the moon, the one of which the Greeks take to be Apollo and the other Diana. But if the moon is a goddess, then Lucifer too and the rest of the wandering stars will hold a place in the roll of the gods; and so, then, will the fixed stars. And why should the rainbow not be set down among the gods? For it is beautiful, and on account of that beauty — since its cause is wonderful — it is said to be the daughter of Thaumas. And if the rainbow has a divine nature, what will you do about the clouds? For the rainbow itself is produced, in a manner of speaking, out of clouds that have taken on color; and one cloud, indeed, is even said to have given birth to the Centaurs. And if you reckon clouds among the gods, then storms too will certainly have to be reckoned there, since they have been consecrated in the rites of the Roman people. So rains, downpours, gales, and whirlwinds are to be held gods; at any rate our own commanders, on putting out to sea, made a practice of sacrificing a victim to the waves.

\ciceroSection{52} Again, if Ceres takes her name from "bearing" — for so you were saying — then the earth itself is a goddess (and so it is held to be; for what else is Tellus?); but if the earth, then the sea as well, which you said was Neptune; and so the rivers too, and the springs. And accordingly Massa dedicated a shrine to the Spring from Corsica, and in the augurs' prayer we find Tiberinus, Spino, Anemo, Nodinus, and the names of other neighboring rivers. So this will either creep on without limit, or we shall admit none of it; and that boundless logic of superstition will not be accepted; therefore none of this is to be accepted.

\ciceroSection{53} We must also, then, Balbus, speak against those who say that these gods, translated to heaven from the race of men, exist not in reality but only in opinion — the very gods whom we all venerate with reverence and devotion. To begin with, those who are called theologians count three Jupiters, of whom the first and second were born in Arcadia, the one fathered by Aether (by whom, they say, Proserpina too and Liber were born), the other fathered by Caelus, who is said to have begotten Minerva — said to be the originator and inventor of war — and the third a Cretan, the son of Saturn, whose tomb is shown on that island. The Dioscuri, too, are named in many ways among the Greeks: the first, three in number, who are called the Anactes at Athens, sons of the most ancient King Jupiter and Proserpina — Tritopatreus, Eubuleus, and Dionysus; the second, sons of the third Jupiter and Leda — Castor and Pollux; the third, called by some Alco, Melampus, and Eumolus, the sons of Atreus, who was born of Pelops.

\ciceroSection{54} Then there are the Muses: the first four, born of the second Jupiter — Thelxinoe, Aoede, Arche, and Melete; the second, nine in number, begotten by the third Jupiter and Mnemosyne; the third, born of the third Jupiter and Antiopa, fathered by Pierus, whom the poets are accustomed to call the Pierides and Pieriae, with the same names and the same number as those just above. And whereas you say that the Sun is so called because he is alone, of him too the theologians bring forward how many Suns. One of them is the son of Jupiter and grandson of Aether; a second, son of Hyperion; a third, son of Vulcan the son of the Nile, whose city the Egyptians hold to be the one called Heliopolis; a fourth, the one whom in heroic times Acantho is said to have borne at Rhodes, father of Ialysus, Camirus, and Lindus, from whom the Rhodians are named; a fifth, who is said to have fathered Aeetes and Circe among the Colchians. There are likewise several Vulcans:

\ciceroSection{55} the first born of Caelus, by whom and by Minerva was born that Apollo in whose guardianship the ancient historians wished Athens to stand; the second, born in the Nile, called Opas by the Egyptians, whom they hold to be the guardian of Egypt; the third, of the third Jupiter and Juno, who is said to have presided over the smithy in Lemnos; the fourth, born of Maemalius, who held the islands near Sicily that were called the Vulcanian.

\ciceroSection{56} Mercury is one, born of Caelus his father and Dies his mother, whose nature is said to have been indecently aroused because he was excited at the sight of Proserpina; a second is the son of Valens and Phoronis, the one held to dwell beneath the earth, the same as Trophonius; a third, born of the third Jupiter and Maia, by whom and by Penelope they say Pan was born; a fourth, son of the Nile, whom the Egyptians count it a sin to name; a fifth, whom the people of Pheneus worship, who is said to have killed Argus, and for that reason to have fled to Egypt and to have handed laws and letters to the Egyptians: this one the Egyptians call Theyt, and by the same name the first month of the year is called among them.

\ciceroSection{57} Of the Aesculapii the first is the son of Apollo, worshipped by the Arcadians, who is said to have invented the probe and to have been the first to bind up a wound; the second is the brother of the second Mercury, said to have been struck by lightning and buried at Cynosurae; the third, son of Arsippus and Arsinoe, who first discovered, as they say, the purging of the bowels and the extraction of teeth, whose tomb and grove is shown in Arcadia not far from the river Lusius. Of the Apollos the most ancient is the one I said a little earlier was born of Vulcan, the guardian of Athens; a second is the son of Corybas, born in Crete, who is said to have contended with Jupiter himself for that island; a third, born of the third Jupiter and Latona, who they say came to Delphi from the Hyperboreans; a fourth, in Arcadia, whom the Arcadians call Nomion because they say they received their laws from him.

\ciceroSection{58} There are likewise several Dianas: the first, daughter of Jupiter and Proserpina, said to have given birth to the winged Cupid; a second, better known, whom we have received as born of the third Jupiter and Latona; the father of the third is recorded to be Upis, the mother Glauce — her the Greeks often call Upis by her father's name. We have many Dionysi: the first born of Jupiter and Proserpina; the second of the Nile, said to have killed Nysa; the third fathered by Cabirus, whom they say ruled as king over Asia, in whose honor the Sabazian rites were instituted; the fourth of Jupiter and the Moon, in whose honor the Orphic mysteries are thought to be performed; the fifth born of Nysus and Thyone, by whom the triennial festivals are thought to have been established.

\ciceroSection{59} The first Venus was born of Caelus and Dies, whose shrine I have seen at Elis; a second was begotten from the sea-foam, by whom and by Mercury we have received that the second Cupid was born; a third, born of Jupiter and Dione, who married Vulcan, but is said to have borne Anteros to Mars; a fourth, conceived in Syria and Cyprus, who is called Astarte, who is reported to have married Adonis. The first Minerva is the one we mentioned above as the mother of Apollo; a second, sprung from the Nile, whom the Egyptians of Sais worship; a third, the one we said above was begotten by Jupiter; a fourth, born of Jupiter and Coryphe the daughter of Oceanus, whom the Arcadians call Coria and hold to be the inventor of the four-horse chariot; a fifth, daughter of Pallas, who is said to have killed her father when he tried to violate her virginity, and to whom they fasten winged sandals.

\ciceroSection{60} The first Cupid is said to have been born of Mercury and the first Diana; the second of Mercury and the second Venus; the third, who is the same as Anteros, of Mars and the third Venus. These things, then, and others of the kind, have been gathered from the ancient lore of Greece. You see that they must be resisted, lest religious observances be thrown into confusion; but your school not only does not refute them — it even confirms them by interpreting what each one is supposed to signify. But let us now return to the point from which we digressed to come here. Do you really suppose, then, that any subtler reasoning is needed to refute all this?

\ciceroSection{61} For we see that mind, faith, hope, virtue, honor, victory, safety, concord, and the rest of this kind have the force of things, not of gods. For either they reside within ourselves — as mind, as hope, as faith, as virtue, as concord — or they are things to be wished for by us, as honor, as safety, as victory; and I see the usefulness of these things, I see too their consecrated images; but why there should be in them the power of gods I shall understand only when I have come to know it. In this class above all is Fortune to be counted, whom no one will separate from inconstancy and randomness — qualities surely not worthy of a god.

\ciceroSection{62} But again, what delight do you take in that unraveling of fables and that teasing-out of names? Caelus castrated by his son, Saturn likewise bound by his son — these and other things of the kind you defend in such a way that those who invented them seem to have been not merely sane but actually wise. As for the teasing-out of names, the labor you spend on it is pitiable: "Saturn, because he saturates himself with years; Mavors, because he overturns great things; Minerva, because she diminishes, or because she threatens; Venus, because she comes to all things; Ceres, from bearing." How dangerous a habit! For with many names you will be stuck fast: what will you do with Veiovis, what with Vulcan? And yet, since you think Neptune is named from swimming, there will be no name whose derivation you cannot explain by means of a single letter; and in that, indeed, you seemed to me to be swimming harder than Neptune himself.

\ciceroSection{63} Zeno first took up a great and altogether needless task, and after him Cleanthes, then Chrysippus — that of giving an account of fabricated fables, of explaining the reasons why each thing was called by its particular name. And in doing this you assuredly confess one thing: that the matter stands far otherwise than men's opinion has it; for those who are called gods are the natures of things, not the figures of gods. So great was this error that names of gods were assigned even to pernicious things, and what is more, sacred rites were established for them. For we see a temple of Fever on the Palatine, and one of Orbona at the shrine of the Lares, and an altar consecrated to Mala Fortuna on the Esquiline.

\ciceroSection{64} Let every such error, then, be driven out of philosophy, so that when we discuss the immortal gods we may say what is worthy of immortal gods. On that subject I have my own view, but I do not have grounds for agreeing with you. You say that Neptune is a mind endowed with intelligence, pervading the sea, and you say the same of Ceres about the earth; but that intelligence — whether of the sea or of the earth — I am able not only not to grasp with my mind but not even to touch by conjecture. And so I must seek elsewhere, in order that I may learn both that the gods exist and what sort of beings they are — such as you would have them be * * .

\ciceroSection{65} Let us look at what follows: first, whether the world is governed by the providence of the gods, and second, whether the gods take thought for human affairs. For these are the two points that remain to me out of your division; and on them, if it seems good to you, I think we ought to argue with more care.'' ``It seems very good to me indeed,'' said Velleius; ``for I both look for greater things to come and warmly assent to what has been said.'' Then Balbus said, ``I do not wish to interrupt you, Cotta, but we shall take another occasion; I shall certainly bring you to confess. But '' In the first place, then, it is not probable that the matter of things, from which all things have arisen, was produced by divine providence, but rather that it has, and has always had, a force and a nature of its own. As a builder, then, when he is about to put up some structure, does not himself make his material but uses what is ready to hand, and as the modeler likewise uses his wax, so for that divine providence the material had to be available — not the material it might itself make, but the material it would find prepared. And if matter was not made by a god, then neither were earth nor water nor air nor fire made by a god. ``Men surpass all the beasts'' — by no means will it go that way; there is a great struggle in the matter. For ``that I should so entreat him with such honeyed flattery, were it not to my purpose'': does she not seem to reason quite well, and to be contriving for her own self a monstrous ruin?

\ciceroSection{66} And by how cunning a calculation here: ``Whoever wills what he wills, things so dispose themselves as he will ply his effort'' — this from the man who is the sower of all evils. ``He, with mind turned awry, has handed me today the very bolts by which I shall throw open all my wrath and deal him destruction — to me grief, to him mourning; to me exile, to him annihilation.'' This faculty of reason, which you say has been granted to man alone by the gods' kindness, the beasts plainly do not possess. Do you see, then, with how great a gift of the gods we have been endowed?

\ciceroSection{67} And the same Medea, in flight from her father and her fatherland, when her father drew near and was already preparing all but to seize her, slaughtered her brother in the meantime, hacked his limbs apart joint by joint, and scattered the body abroad through the fields — for this reason: that while her father gathered up the strewn limbs of his child, she herself might escape in the interval, that grief might slow him in pursuit, that she might secure her own safety by a kinsman's murder. To this woman, as wickedness was not wanting, so neither was reason.

\ciceroSection{68} What of the man preparing the deadly banquet for his brother — does he not turn the calculation this way and that in thought? ``A greater pile is mine to raise, a greater evil to be mixed, that I may crush and grind down that man's bitter heart.'' Nor yet is that very man himself to be passed over, who was not content to have lured his brother's wife into adultery — of whom Atreus speaks rightly and most truly: ``What in the highest matter of state I judge the highest peril — that royal mothers be defiled, the line contaminated, the stock confounded.'' But that very thing, with what cunning — the man who would seek a throne by adultery: ``I add to this,'' he says, ``that the father of the gods of heaven sent me a portent and a prodigy, the very prop of my kingdom — a lamb among the flocks, conspicuous with golden fleece, which Thyestes once dared to steal away from the palace'' — and in this affair he took his wife as his accomplice. Does he not seem to have used the utmost depravity, yet not without the utmost reason?

\ciceroSection{69} Nor is the stage alone crammed with these crimes; ordinary life is full of crimes nearly greater. Each man's household feels it, the Forum feels it, the Senate-house, the Campus, our allies, the provinces — so that, just as a thing is rightly done by reason, so it is by reason that men sin; and the one is done both by few and rarely, the other both always and by very many. So it would have been better that no reason at all had been given us by the immortal gods than that it should be given with so great a ruin attending it. As with wine for the sick: because it rarely helps and very often harms, it is better not to administer it at all than to rush into open ruin in the hope of a doubtful recovery — so I rather think it might have been better for the human race that this quick stir of thought, this keenness, this ingenuity, which we call reason, since it is destructive to many and salutary to very few, had not been given at all than that it should be given so munificently and so lavishly.

\ciceroSection{70} For this reason, if the divine mind and will took thought for men precisely because it bestowed reason on them, it took thought only for those whom it endowed with good reason — and these we see, if indeed there are any at all, to be very few. But it cannot be allowed that the immortal gods took thought for only a few; it follows, then, that they took thought for no one. To this point you are accustomed to make this reply: that the gods did not therefore fail to provide for us excellently merely because many would use their gift perversely; that many use their inheritances badly too, and are not on that account held to have had no benefit from their fathers. Who denies it — and what likeness is there in that comparison? For Deianira did not wish to harm Hercules when she gave him the tunic dyed with the Centaur's blood, nor did the man who opened with his sword the abscess of Jason of Pherae — which the physicians had not been able to cure — wish to do him good. For many, even while wishing to do harm, have done good, and while wishing to do good have done harm. And so it does not follow from what is given that the will of the giver is made plain, nor, if the one who received it uses it well, that the one who gave it gave it in friendship. For what lust, what avarice, what crime is either undertaken without a plan having been formed, or carried through without the mind's motion and reflection — that is, without reason?

\ciceroSection{71} For every opinion is reason — good reason if the opinion is true, but bad reason if it is false. Yet from a god we have only reason, if indeed we have it; whether the reason is good or not good comes from ourselves. For reason was not given to man by the kindness of the gods in the way that an inheritance is left to him. For what else could the gods rather have given to men, if they had wished to harm them? And what would be the seeds of injustice, of intemperance, of cowardice, if reason did not underlie these vices? Medea just now and Atreus were called to mind by us — heroic characters who, having reckoned and weighed the account, plotted their unspeakable crimes.

\ciceroSection{72} What of the trivialities of comedy — do they not always turn upon reasoning? Does that fellow in the \textit{Eunuchus} not argue with subtlety enough: ``What, then, shall I do? She shut me out, she calls me back; shall I return? No, not if she begged me.'' And that other in the \textit{Synephebi} does not hesitate, in the manner of the Academics and against the common opinion, to do battle with reason: he says it is ``sweet, in the height of love and the depth of need, to have a father who is grasping, ungracious toward his children, hard — one who neither loves you nor takes pains for you''; and to this incredible thesis he supplies his little arguments:

\ciceroSection{73} ``Either you cheat him of his revenue, or by a letter divert some debt, or through a little slave throw the timid old man into a fright; and in the end, the more sparing the father from whom you filch, the more gladly you squander it.'' And the same fellow argues that an easygoing and generous father is an inconvenience to a son in love: ``one whom I know no way to cheat, nor what to carry off from him, nor what trick or contrivance to set in motion against him — so utterly has my father's obligingness baffled all my tricks, deceits, and sleights of hand.'' Well then — those tricks, those contrivances, those deceits and sleights of hand: could they have existed without reason? O splendid gift of the gods, that Phormio should be able to say: ``Fetch me the old man; all my plans are now drawn up and ready in my heart.'' But let us leave the theater and come into the Forum.

\ciceroSection{74} The praetor goes to take his seat. To try what case? The man who burned down the record office. What crime more hidden? Yet Quintus Sosius, a distinguished Roman knight from the Picene country, confessed that he had done it. The man who falsified the public registers? That too Lucius Alenus did, when he forged the handwriting of the six chief clerks; what was ever more ingenious than this man? Take note of other inquiries: into the gold of Tolosa, into the Jugurthine conspiracy; go back to earlier matters — the case of Tubulus over money taken for rendering a verdict; later matters — the inquiry into incest under the Peducaean bill; then these everyday affairs: daggers, poisonings, embezzlements, and inquiries into wills as well under a new law. From this comes that form of action: ``I declare that a theft was committed by your aid and counsel''; from this come so many trials over bad faith — over guardianship, mandate, partnership, trusteeship — and the rest of the dealings in buying or selling, hiring or letting, that are done against good faith; from this comes the public trial of a private matter under the Laetorian law; from this the dragnet of all knaveries, the action concerning malicious fraud, which our friend Gaius Aquillius brought forward — the very fraud which the same Aquillius holds to be present whenever one thing is feigned and another done.

\ciceroSection{75} Do we then suppose that this vast sowing of evils was done by the immortal gods? For if the gods gave reason to men, they gave malice; for malice is the cunning and deceitful reasoning of doing harm. The same gods gave fraud too, and crime, and the rest, none of which can either be undertaken or accomplished without reason. Would, then — as that old woman wishes that the fir timbers, cut down by axes in the grove of Pelion, had never fallen to earth — would that the gods had not given men this craftiness, which very few use well, and even those few are often overborne by those who use it ill, while countless men use it wickedly: so that this divine gift of reason and counsel seems to have been imparted to men for fraud and not for goodness.

\ciceroSection{76} But you press again and again that this is the fault of men, not of the gods. As if a physician should blame the gravity of the disease, or a pilot the violence of the storm; and even though these are mere mortal creatures, still they are ridiculous: ``for who would have called you in,'' someone might say, ``if these things did not exist?'' Against a god one may argue more freely: ``You say the fault lies in men's vices: you should have given men a reason that would have shut out vice and fault.'' Where, then, was there room for error on the gods' part? For we leave our inheritances in the hope of handing them on well, and in that we can be deceived; but how could a god be deceived? Was it as the Sun was, when he took up his son Phaethon into the chariot, or Neptune, when Theseus destroyed Hippolytus, having been granted by his father Neptune the power of three wishes?

\ciceroSection{77} These are the poets' tales; but we wish to be philosophers, authors of facts, not of fables. And yet even these poets' gods themselves, had they known that those gifts would be ruinous to their sons, would be reckoned to have done wrong in conferring the benefit. And if it is true, as Aristo of Chios used to say, that philosophers do harm to their hearers who put a bad interpretation on things well said — for it is possible for profligates to come out of Aristippus's school, and harsh men out of Zeno's — then surely, if those who had heard were going to depart corrupted because they perversely interpreted the philosophers' discourse, it would be better for the philosophers to keep silent than to harm those who had heard them;

\ciceroSection{78} so, if men turn the reason given them by the immortal gods with good intent into fraud and malice, it would have been better that it had not been given to the human race than that it should be given. As, if a physician should know that the patient ordered to take wine would take it undiluted and perish at once, he would be greatly at fault, so that Providence of yours is to be blamed, which gave reason to those it knew would use it perversely and wickedly. Unless perhaps you say it did not know. Would that it were so; but you will not dare to, for I am well aware how highly you esteem its name.

\ciceroSection{79} But here at any rate this topic can now be brought to a close. For if folly, by the consensus of all philosophers, is a greater evil than all the evils of fortune and of the body weighed in the opposite scale, and if no one attains wisdom, then we are all in the depths of evil — we for whom you say the immortal gods have provided so excellently. For just as it makes no difference whether no one is in good health or no one is able to be in good health, so I do not see what difference it makes whether no one is wise or no one is able to be wise. But we, indeed, have said too much about a thing perfectly plain; Telamo, however, dispatches the whole topic — why the gods neglect men — in a single line: ``For if they cared, it would go well with the good, ill with the wicked; which now is not the case.'' The gods ought indeed to have made all men good, if in truth they were taking thought for the human race;

\ciceroSection{80} or if that less, they ought at least certainly to have taken thought for the good. Why, then, did the Carthaginian crush the two Scipios in Spain, those bravest and best of men? Why did Maximus bury his son, who had held the consulship? Why did Hannibal slay Marcellus? Why did Cannae carry off Paulus? Why was the body of Regulus surrendered to the cruelty of the Carthaginians? Why did his own house's walls not shelter Africanus? But these are old matters, and there are very many others; let us look at things nearer home. Why is my uncle, Publius Rutilius, a man of the purest innocence and at the same time the most learned, in exile? Why was my friend Drusus killed in his own house? Why was Quintus Scaevola, the very pattern of temperance and prudence, the pontifex maximus, butchered before the image of Vesta? Why, still earlier, were so many leading men of the state put to death by Cinna? Why could Gaius Marius, the most treacherous of all men, order Quintus Catulus, a man of the most surpassing dignity, to die?

\ciceroSection{81} The day would fail me if I tried to count those good men to whom evil befell, and no less if I called to mind those scoundrels to whom things fell out best of all. For why did Marius, so happily consul a seventh time, die an old man in his own house? Why did Cinna, cruelest of all men, rule for so long? "But he paid the penalty." It would have been better to stop him and hold him back from killing so many of the greatest men than for him at last to pay for it himself. Quintus Varius, a thoroughly ruthless man, perished under the utmost torture and execution; if it was because he had made away with Drusus by the sword and Metellus by poison, then it would have been better for those two to be preserved than for Varius to pay the penalty of his crime. Dionysius the tyrant ruled a most wealthy and most prosperous city for thirty-eight years;

\ciceroSection{82} and before him, in the very flower of Greece, how many years did Pisistratus rule? "But Phalaris paid the penalty, and Apollodorus too." Yes — after many had first been tortured and put to death. And pirates too often pay the penalty, yet we cannot say that more captives were not cruelly killed than pirates. We are told that Anaxarchus, the follower of Democritus, was butchered by the tyrant of Cyprus, and that Zeno of Elea was killed under torture; and what shall I say of Socrates, over whose death I am accustomed to weep when I read Plato? Do you see, then, that if the gods do regard human affairs, in their judgment the distinction between good men and bad has been abolished?

\ciceroSection{83} Diogenes the Cynic used to say that Harpalus, who in those days was reckoned a fortunate pirate, was a standing witness against the gods, since he lived so long in such prosperity. Dionysius, of whom I spoke before, after he had plundered the shrine of Proserpina at Locri, was sailing to Syracuse; and as he held his course with a most favorable wind, he said with a laugh, "Do you see, friends, what a fine voyage the immortal gods grant to men who rob temples?" And, sharp man that he was, having grasped the matter clearly and fully, he held firmly to the same opinion. When he had brought his fleet to the Peloponnese and had come into the temple of Olympian Jupiter, he stripped from the god a golden cloak of great weight, with which the tyrant Gelo had adorned Jupiter from the spoils of the Carthaginians, and over this he even joked that a golden cloak was heavy in summer and cold in winter, and threw a woolen mantle over the god, saying that it suited every season of the year. The same man ordered the golden beard of Aesculapius at Epidaurus to be taken off, declaring that it was not fitting for the son to be bearded when in all the temples the father stood beardless.

\ciceroSection{84} Then he ordered the silver tables to be carried off from all the sanctuaries, and since, by the custom of old Greece, they bore the inscription "of the good gods," he said he wished to enjoy their goodness. The same man took down without hesitation the little golden Victories and the bowls and crowns held out in the outstretched hands of the images, saying that he was accepting these things, not carrying them off — for it was folly to refuse to take, from gods who held them out and offered them, the very goods we pray to receive from them. They say, too, that the things I have mentioned — taken from the temples — he brought out into the forum and sold by the auctioneer's voice, and that, once the money had been exacted, he proclaimed that whatever anyone held from sacred property he should return, each item to its own temple, before a fixed day: thus to his impiety toward the gods he added injustice toward men. This man, then, neither did Olympian Jupiter strike with the thunderbolt, nor did Aesculapius destroy by some wretched lingering disease; he died in his own bed, was borne to the pyre of his tyranny, and the power that he himself had won by crime he handed on to his son, as though it were a just and lawful inheritance.

\ciceroSection{85} Against my will my argument moves on this ground, for it seems to lend authority to wrongdoing; and it would rightly seem so, were it not that conscience itself carries a heavy weight, in matters both of virtue and of vice, apart from any divine reckoning — and once that is removed, everything collapses. For just as no household and no commonwealth would appear to be ordered by any rational design and discipline if in it there were no rewards set out for right conduct and no punishments for wrongdoing, so there is surely no divine governance of the world over men, if in it there is no distinction between the good and the bad.

\ciceroSection{86} "But the gods neglect small things; they do not attend to the little fields of individual men or their slips of vine, and if blight or hail has harmed someone, that was no business for Jupiter to take notice of; not even kings in their kingdoms see to all the smallest matters" — for so you say. As if I had complained a moment ago about Publius Rutilius's estate at Formiae, and not about the loss of his civic standing! And this, indeed, is how all mortals regard the matter: that external advantages — vineyards, crops, olive groves, abundance of grain and fruit, in short every comfort and prosperity of life — they have from the gods; but no one has ever credited his virtue to a god.

\ciceroSection{87} And rightly, no doubt; for it is on account of virtue that we are justly praised, and in virtue that we rightly glory — which could not be the case if we held it as a gift from a god and not from ourselves. But truly, when we have been raised by honors or in our family fortune, or have gained some other accidental good or fended off some evil, then we give thanks to the gods, and then think nothing has been added to our own credit. Has anyone ever thanked the gods for being a good man? No — but for being rich, for being honored, for being safe; and they call upon Jupiter Best and Greatest for those things, not because he makes us just, temperate, and wise, but because he keeps us safe, sound, wealthy, and well supplied;

\ciceroSection{88} nor did anyone ever vow a tithe to Hercules for having been made wise — though Pythagoras, when he had discovered something new in geometry, is said to have sacrificed an ox to the Muses; but that I do not believe, since he was unwilling even to sacrifice a victim to Apollo at Delos, lest he sprinkle the altar with blood. But to return to the point: this is the judgment of all mortals — that fortune is to be sought from a god, but wisdom to be taken from oneself. Grant that we may dedicate temples to Mind, to Virtue, and to Faith, yet we see that these things reside in ourselves; the capacity for hope, safety, help, and victory must be sought from the gods. The prosperity, then, and the good fortune of the wicked refute, as Diogenes used to say, all the power and might of the gods.

\ciceroSection{89} "But sometimes good men come to a good end." We snatch up such cases, to be sure, and assign them to the immortal gods without any reason. But Diagoras — the one called the Atheist — when he had come to Samothrace, and a certain friend said to him, "You who think the gods neglect human affairs, do you not notice, from all these painted tablets, how many men have escaped the violence of the storm by their vows and reached harbor safely?" — "Just so," he replied, "for those who suffered shipwreck and perished at sea are nowhere painted." And the same man, when he was sailing and his fellow passengers, frightened and panic-stricken by a storm against them, said that this was happening to them not undeservedly, since they had taken him aboard their ship, pointed out to them many other ships laboring on the same course and asked whether they believed that Diagoras was sailing on those ships too. For the truth of the matter is this: that, as regards prosperous or adverse fortune, it makes no difference what sort of man you are or how you have lived.

\ciceroSection{90} "The gods do not attend to everything," he says, "any more than kings do." But where is the likeness? For if kings deliberately overlook things, the fault is great; whereas for a god there is not even the excuse of ignorance. And you defend him splendidly when you say that the power of the gods is such that, even if someone has escaped the penalty of his crime by death, those penalties are exacted from his children, his grandchildren, his descendants. What admirable fairness in the gods! Would any state endure a proposer of a law of this kind — that a son or grandson be condemned if a father or grandfather had transgressed? What end could be set to the slaughter of the house of Tantalus, or what satiety of punishment will ever be granted for the death of Myrtilus to be atoned by penalties? Whether the poets have corrupted the Stoics, or the Stoics have lent authority to the poets, I could not easily say;

\ciceroSection{91} for both alike tell of monstrous and scandalous things. For the man whom an iambic of Hipponax had wounded, or who had been injured by a verse of Archilochus, did not bear a pain sent down by a god but one conceived by himself; and when we observe the lust of Aegisthus or of Paris, we do not look for a god as its cause, since we all but hear the voice of guilt itself; nor do I judge that the recovery of many sick men was given by Aesculapius rather than by Hippocrates; nor shall I ever say that the discipline of the Lacedaemonians was given to Sparta by Apollo rather than by Lycurgus. Critolaus, I say, destroyed Corinth, and Hasdrubal Carthage; these two men gouged out those two eyes of the seacoast — not some angry god, who, you maintain, cannot be angry at all.

\ciceroSection{92} But surely he could have come to their aid and preserved cities so great and so fine; for you yourselves are accustomed to say that there is nothing a god cannot accomplish, and that without any effort at all; for just as the limbs of men are moved by the mind itself and by the will, without any straining, so by the divine power of the gods all things can be fashioned, moved, and changed. And you do not say this in a superstitious, old-wives' way, but on grounds of physics and consistent reasoning: that the matter of things, out of which and in which all things are, is wholly flexible and changeable, so that there is nothing that cannot be fashioned and transformed out of it, however suddenly; and that the fashioner and governess of this whole matter is divine providence; therefore, wherever it moves itself, it can bring about whatever it wills. And so either it does not know what it can do, or it neglects human affairs, or it cannot judge what is best.

\ciceroSection{93} "It does not care for individual men." No wonder: not even for states; not for those: not even for nations and peoples. And if it will despise these too, what wonder is it that the whole human race has been despised by it? But how can you, who say that the gods do not pursue all things, at the same time wish that dreams be distributed and divided among men by the immortal gods (this with you, since the doctrine about the truth of dreams is yours), and likewise say that vows ought to be undertaken? Individuals, surely, make vows; the divine mind, then, hears even individuals; you see, therefore, that it is not so occupied as you supposed. Suppose it is distracted — turning the heavens, watching over the earth, governing the seas: why does it allow so many gods to do nothing and stand idle? Why does it not set over human affairs some of the gods at leisure — those innumerable ones that you, Balbus, have unfolded? This is more or less what I had to say about the nature of the gods, not in order to abolish it, but so that you might understand how obscure it is and what difficult explanations it involves.' When he had said this, Cotta made an end.

\ciceroSection{94} But Lucilius said: "Too vehemently, Cotta, have you inveighed against that doctrine of the Stoics concerning the providence of the gods, which they established most reverently and most wisely. But since evening is coming on, you will grant us some day on which to argue against what you have said. For I have a contest with you on behalf of altars and hearths, on behalf of the temples and shrines of the gods and the walls of the city — those walls which you priests declare to be sacred, and you gird the city more carefully with religion than with its own ramparts; and to abandon them, so long as I can draw breath, I judge to be a sacrilege."

\ciceroSection{95} Then Cotta: 'For my part, I both wish to be refuted, Balbus, and I preferred to set out the things I argued rather than to pass judgment on them, and I know for certain that you can easily defeat me.' "No doubt," said Velleius, "from a man who thinks that even dreams are sent to us by Jupiter — dreams which are nevertheless not so trifling as the Stoics' discourse on the nature of the gods." When these things had been said, we parted in such a way that to Velleius the argument of Cotta seemed the truer, while to me that of Balbus seemed to lean nearer the resemblance of truth.
