% De Re Publica (de-re-publica) --- English translation
% Translated by Claude (Anthropic) from the Latin (Perseus).
% Style guide: STYLE.md.

\ciceroBook{1}

\ciceroSection{1} \ldots had not, by their onset, freed [the state]; nor had Gaius Duilius, Aulus Atilius, Lucius Metellus delivered Rome from the terror of Carthage; nor had the two Scipios put out, with their own blood, the rising flame of the Second Punic War; nor, when it broke out again with greater forces, would either Quintus Maximus have worn it down, or Marcus Marcellus crushed it, or Publius Africanus, having torn it from the very gates of this city, driven it back within the enemy's walls. To Marcus Cato, an obscure new man --- whom all of us who pursue the same things are drawn after, as if to a model, toward industry and virtue --- it was certainly open to enjoy himself in leisure at Tusculum, a healthful spot and close at hand. But the man, mad (as those people think), with no necessity compelling him, preferred to be tossed in these waves and storms into extreme old age rather than to live, in such tranquillity and leisure, most agreeably. I pass over the countless men, each of whom was a saving of this commonwealth: because they are not far off from the memory of this age, I leave their names alone, lest someone complain that he himself, or one of his own, has been overlooked. This one thing I lay down: that so great is the necessity of virtue given by nature to the human race, and so great the love of defending the common safety, that this force has overcome all the enticements of pleasure and ease.

\ciceroSection{2} Nor indeed is it enough to possess virtue as if it were some art, unless you use it; for although an art, even when not in use, can still be held by knowledge alone, virtue lies wholly in its exercise; and the greatest exercise of virtue is the steering of the state and the carrying out --- in fact, not in talk --- of those very things which those people declaim about in corners. For nothing is said by the philosophers, at least when it is rightly and honourably said, that has not been begotten and confirmed by those by whom legal codes were drawn up for cities. Whence comes piety, or from whom religion? Whence comes law --- whether of nations, or this very thing called civil law? Whence justice, faith, equity? Whence shame, restraint, the flight from baseness, the appetite for praise and honour? Whence in toils and dangers comes courage? Surely from those who, having shaped these things by training, have confirmed some by custom and sanctioned others by law.

\ciceroSection{3} Indeed they say of Xenocrates, the foremost philosopher of his rank, that when he was asked what his pupils gained, he answered: ``That they should do of their own will what they would be compelled to do by the laws.'' That citizen, then, who by command and by penalty of the laws compels everyone to do what philosophers can scarcely persuade a few to do by their speech, is to be set above the very teachers who debate the matter. For what speech of theirs is so refined that it is to be preferred to a well-ordered state with its public law and its customs? Just as I think ``great and imperious cities,'' as Ennius calls them, are to be set before villages and forts, so I judge those who preside over these cities by counsel and authority to be set far above, in wisdom itself, those who have no part in any public business. And since we are most of all swept on toward enlarging the resources of the human race, and study to render the life of men safer and richer by our counsels and our labours, and since we are spurred to this very pleasure by the goads of nature herself, let us hold to that course which has always been the course of the best men, and let us not listen to those signals which sound the recall, calling back even those who have already gone forward.

\ciceroSection{4} Against these reasonings, so sure and so brilliant, those who argue on the other side set first the labours which must be borne in defending the commonwealth --- a slight enough hindrance, surely, to the watchful and energetic man, and to be despised not only in such great matters but even in middling pursuits or duties or even in business affairs. Then they add the dangers of life, and they hold up before brave men a base fear of death --- to whom this is wont to seem more wretched, that they should be used up by nature and old age, than that the time should be granted them to give back, on behalf of their country above all, that life which had to be given back to nature in any case. And it is on this ground that they think themselves rich and eloquent, when they gather up the calamities of the most distinguished men and the wrongs done them by their ungrateful fellow citizens.

\ciceroSection{5} For from this come those examples both among the Greeks: that Miltiades, the conqueror and tamer of the Persians, when the wounds were not yet healed which he had taken in his glorious victory by his front turned to the enemy, poured out, in the chains of his own citizens, the life saved from the enemy's weapons; and that Themistocles, driven from the country he had freed and frightened off, took refuge not in the Greek harbours he himself had preserved, but in the gulfs of barbarism, which he had crushed. Nor are examples wanting of Athenian fickleness and cruelty toward their best citizens; and those things, born and rife with them, are said to have spilled over even into our own most weighty state ---

\ciceroSection{6} for they bring up either Camillus's exile, or the affront to Ahala, or the unpopularity of Nasica, or the expulsion of Laenas, or Opimius's condemnation, or Metellus's flight, or the most bitter ruin of Gaius Marius and the slaughter of leading men, or the destruction of those many others which followed not long after. And they no longer keep off my own name; and, I suppose, because they think themselves preserved by my counsel and at my danger to that life and leisure of theirs, they complain of me with even more weight, and more affection. But I cannot easily say why, when these very men cross the seas for the sake of learning and seeing \ldots

\ciceroSection{7} \ldots [had I, swearing] that the commonwealth was safe, when leaving the consulship in the assembly, with the Roman people swearing the same, I should easily compensate the trouble and vexation of all the wrongs done me. Yet our own misfortunes brought more honour than toil, and not so much vexation as glory, and we drew greater joy from the longing of the good than pain from the joy of the wicked. But if it had turned out otherwise, as I have said, how could I have complained? When nothing had befallen me unexpectedly nor more grievously than I had looked for, in return for such great deeds of mine. For I had been such a man that, although it was open to me either to take greater fruits from leisure than others, by reason of the varied sweetness of the studies in which I had lived from boyhood, or, if anything more bitter befell us all, to undergo not a special, but only an equal share with the rest in fortune, I did not hesitate to throw myself headlong against the heaviest storms and almost the very thunderbolts themselves, for the sake of saving my fellow citizens, and by my own personal dangers to procure the common quiet for the rest.

\ciceroSection{8} For our country did not beget us or rear us under such a law that she should expect, as it were, no nourishment from us, and serve only our conveniences, supplying a safe refuge for our leisure and a peaceful place for our quiet --- but rather that she should pledge to herself, for her own use, the most and the largest part of our spirit, our talent, our judgement, and grant back to us, for our private use, only as much as could be left over for herself.

\ciceroSection{9} Then those refuges they take to themselves for an excuse, that they may the more easily enjoy their leisure --- those, too, are by no means to be listened to, when they say that men who come to public affairs are for the most part fellows worthy of nothing good, with whom to be compared is sordid, and to fight, especially when the multitude is roused, is wretched and dangerous. For which reason, they say, it is neither the wise man's part to take up the reins, when he cannot restrain the mad and untamed onsets of the rabble, nor the part of a free man, when he is contending with foul and monstrous opponents, to bear the lashes of insults or to expect injuries which the wise man cannot endure --- as if good men, brave men, men endowed with great spirit had any juster cause for entering on public business than this: not to obey the worthless, nor to suffer the commonwealth to be torn apart by them, when they themselves, even if they wished to bring help, could not.

\ciceroSection{10} And that exception of theirs --- to whom can it ever be approved? --- when they say the wise man will undertake no part in the commonwealth except when time and necessity have compelled him: as if a greater necessity could befall any man than has befallen me. In that necessity, what could I have done if I had not been then consul? But how could I have been consul, if I had not held that course of life from boyhood by which, born of equestrian rank, I might come to the most distinguished honour? There is no power, then, to bring help to the commonwealth on the spur of the moment, or when you wish, however hard pressed it is by dangers, unless you are in the place where you may do so.

\ciceroSection{11} And what most often seems strange to me in the speech of these learned men is this: that those who deny they can steer in a calm sea --- because they have neither learned the art nor ever cared to know it --- the very same men profess that they will set hand to the helm when the greatest waves have been raised. For these people are wont to say openly, and even to glory in the saying, that they have neither ever learned nor teach anything about the principles of either founding or preserving states; and they think the knowledge of these things is to be conceded not to learned and wise men, but to those skilled in that other line. How then is it fitting to promise their service to the commonwealth only when they are forced by necessity --- since, what is much easier, when no necessity presses, they cannot govern the state? For my part, even if it were true that the wise man does not commonly come down of his own will to public business, but if compelled by the times then does not refuse the duty, I should still judge that this knowledge of public affairs is by no means to be neglected by the wise man, since he ought to have everything ready for use, when he might not know whether at some time he must use it.

\ciceroSection{12} I have spoken of these matters at greater length for this reason: that in these books a discussion has been undertaken and taken on by me concerning the commonwealth; and lest it be conducted in vain, I had first of all to remove this hesitation about entering on public business. Yet if there are any who are moved by the authority of philosophers, let them give heed for a little while and listen to those whose authority and glory is highest among the most learned men --- whom I judge, even if some of them did not themselves manage public business, still, since they have inquired and written much about the commonwealth, to have discharged some duty to the commonwealth. As to those seven whom the Greeks named the Wise Men, I see that almost all of them moved in the very midst of public business. For there is no thing in which human virtue draws nearer to the divine power of the gods than in either founding new cities or preserving those already founded.

\ciceroSection{13} On these matters, since it has been our lot both to have achieved something worthy of remembrance in the conduct of the commonwealth and to be authorities, by some skill in the unfolding of the principles of public affairs, not only by use but also by the study of learning and teaching --- whereas earlier men were either polished in their disputations, but no deeds of theirs are to be found, or excellent in action but unpracticed in argument --- still no new doctrine of our own, invented by us, is to be set up here: rather, we must call back to memory the discussion of the most distinguished and wisest men of our state, of one age, which was reported to me, and to you when you were a young man, by Publius Rutilius Rufus when we were together at Smyrna for several days; in which scarcely anything was passed over which had any great bearing on the principles of the whole subject.

\ciceroSection{14} For when this Publius Africanus, son of Paulus, in the consulship of Tuditanus and Aquilius, had decided to be in his gardens for the Latin holidays, and his closest friends had said they would visit him often during those days, on the morning of the holidays itself the first to come to him was the son of his sister, Quintus Tubero. When Scipio had hailed him kindly and was glad to see him, he said: ``Why so early, Tubero? These holidays were giving you a fine opportunity for unfolding your books.'' Then Tubero: ``For me indeed every time is free for my books; for they are never busy. But to find you at leisure is no small thing, especially with this present commotion of the commonwealth.'' Then Scipio: ``You have caught me, but, by Hercules, more at leisure in business than in mind.'' And Tubero: ``Yet you ought to relax your mind too. We are many, as we determined, ready, if it can be done at your convenience, to enjoy this leisure with you.'' ``Gladly will I --- so that we may at last be put in mind of something out of the studies of learning.''

\ciceroSection{15} Then he said: ``Do you wish, then, since you in some way invite me and give me hope of yourself, that we look first into this, Africanus, before the others come --- what is the meaning of that other sun reported in the senate? For neither few nor light are those who say they have seen two suns, so that the matter is less one of crediting than of looking for the cause.'' Here Scipio: ``How I should wish we had our Panaetius with us! He, with the rest, is wont to investigate these celestial things most eagerly. But for my part, Tubero --- to speak openly to you what I think --- I do not greatly agree, in this whole subject, with that intimate of ours, who, when we can scarcely guess by conjecture what these things are, so confidently affirms them as if he saw them with his eyes or fully handled them with his hand. For this reason I am wont to judge Socrates the wiser, who set aside all care of this kind, and said that the things asked about nature were either greater than the human mind could attain, or had no bearing at all on the life of men.''

\ciceroSection{16} Then Tubero: ``I do not know, Africanus, why it has been so handed down to memory --- that Socrates rejected all that kind of inquiry and was wont to ask only about life and morals. For what authority on him fuller than Plato can we cite? In whose books, in many places, Socrates so speaks that even when he is debating about morals, about the virtues, in fine about the commonwealth, he yet busies himself in joining numbers and geometry and harmony, in the manner of Pythagoras.'' Then Scipio: ``It is as you say. But I believe you have heard, Tubero, that Plato, after Socrates's death, went first to Egypt for the sake of learning, then made for Italy and Sicily, that he might thoroughly learn the inventions of Pythagoras; that he was much with Archytas of Tarentum and with Timaeus of Locris, and got hold of the commentaries of Philolaus; and since at that time the name of Pythagoras flourished in those parts, he gave himself to the Pythagoreans and to those studies. So, since he loved Socrates uniquely and wished to assign everything to him, he wove together the Socratic charm and subtlety of speech with the obscurity of Pythagoras and with that gravity of so many arts.''

\ciceroSection{17} When Scipio had said these things, he caught sight of Lucius Furius coming up suddenly; and when he had greeted him, he took him by the hand most warmly and placed him on his own couch. And when Publius Rutilius came in at the same time --- he who is our authority for this discussion --- he greeted him too, and made him sit down beside Tubero. Then Furius: ``What are you doing? Has our coming in cut short any conversation of yours?'' ``Not at all,'' said Africanus, ``for you are wont to investigate these matters with care, of the kind on which Tubero had begun to inquire a little while ago; and our Rutilius indeed used at times, even under the very walls of Numantia, to look into something of this sort with me.'' ``What was it that had come up?'' said Philus. Then he: ``About those two suns, on which I am eager, Philus, to hear from you what you think.''

\ciceroSection{18} He had said this, when a slave-boy announced that Laelius was on his way to him, and had already left his house. Then Scipio, having put on his shoes and clothes, went out from the bedchamber, and after he had walked a little while in the colonnade, greeted Laelius as he came up, together with those who had come with him --- Spurius Mummius, whom he loved most of all, and Gaius Fannius and Quintus Scaevola, sons-in-law of Laelius, learned young men, already at the age for the quaestorship. When he had greeted them all, he turned in the colonnade and put Laelius in the middle; for there was, in their friendship, a certain rule between them: that on campaign, on account of his outstanding glory in war, Laelius should worship Africanus as a god, while at home in turn Scipio should regard Laelius, because he was the elder, in the place of a parent. Then, after a few words exchanged in one or two turns up and down, and the coming of these had been most pleasant and most welcome to Scipio, it was agreed that they should sit down in the most sun-warmed spot of the meadow --- since it was winter time. As they were about to do so, in came Manius Manilius, a man of judgement, and to all of them welcome and dear; and after most friendly greetings from Scipio and the rest, he sat down next to Laelius.

\ciceroSection{19} Then Philus: ``It does not seem to me,'' he said, ``because these men have come, that we ought to look for some other subject of conversation, but rather that we should go on more carefully and say something worthy of their ears.'' Here Laelius: ``What were you doing, then? On what conversation have we broken in?'' ``Scipio had asked me what I thought of this --- that two suns were agreed to have been seen.'' ``Do you mean it, Philus? Have we already explored the things that pertain to our households and to the commonwealth, that we should be asking what is going on in the heavens?'' And Philus: ``Do you not think it pertains to our households to know what is being done and what is happening at home? --- which is not just what our walls enclose, but this whole world, which the gods have given us as a common dwelling and a common country with themselves; especially since, if we are ignorant of these things, many great matters must be unknown to us. As for me indeed --- and you too, Laelius, by Hercules, and all who are eager for wisdom --- the very knowledge and contemplation of things delights me.''

\ciceroSection{20} Then Laelius: ``I do not stand in your way, especially since we are on holiday. But can we hear something, or have we come too late?'' ``Nothing has been argued out yet, and since the matter is open, I shall gladly give way to you, Laelius, that you may speak on it.'' ``Nay rather, let us hear you --- unless perchance Manilius thinks some interdict is to be drawn up between the two suns, that they might possess the heavens just as each had possessed it.'' Then Manilius: ``Do you go on, Laelius, to make sport of that art in which you yourself excel above all, and without which no one can know what is his own and what another's? But that for later. Now let us hear Philus, whom I see consulted on greater matters than I am, or than Publius Mucius.''

\ciceroSection{21} Then Philus: ``I shall bring you nothing new, nor anything thought up or discovered by me; for I remember that Gaius Sulpicius Gallus --- a most learned man, as you know --- when this same thing was said to have been seen, and he happened to be at the house of Marcus Marcellus, who had been consul with him, ordered to be brought out the celestial globe which the grandfather of that Marcellus, when Syracuse was taken, had carried off from that wealthiest and most adorned city; though out of so great a spoil he had brought home nothing else. Of which globe, although for Archimedes's glory I had often heard the name, I did not so greatly admire its appearance; for the other one was more elegant and more famous in popular fame, which the same Marcellus had set up in the temple of Virtue, made by the same Archimedes.

\ciceroSection{22} But after Gallus began most knowingly to lay out the design of this work, I judged there had been more genius in that Sicilian than human nature could be thought to bear. For Gallus said that the invention of that other solid and full globe was an old one, and that it had first been turned by Thales of Miletus, then later by Eudoxus of Cnidus, the disciple, as he reported, of Plato; that it had been mapped out with the stars and constellations which are fixed in the sky; and that all its ornament and design, taken from Eudoxus, had many years afterwards been put into verse by Aratus --- not by his knowledge of astronomy, but by a certain poetic gift. But this kind of globe, in which were the motions of the sun and moon and of those five stars which are called wandering and, as it were, vagrant, could not be marked out on that solid globe; and in this was the marvellous invention of Archimedes, that he had thought out how, in motions so unlike, one revolution should preserve unequal and varied courses. When Gallus moved this globe, it would happen that the moon followed the sun by as many revolutions in that bronze as it does in days in the heavens itself; from which both that same eclipse of the sun would occur in the artificial heaven and the moon would fall upon that mark which is the shadow of the earth, when the sun, on the other side \ldots

\ciceroSection{23} \ldots [I was glad,] both because I myself loved the man, and because I knew that he had been highly approved and dear to my father Paulus. I remember when I was quite a young man, when my father was consul in Macedonia and we were in the camp, that our army was thrown into religious panic and fear, because on a clear night, suddenly, the bright and full moon was eclipsed. Then he, when he had been our legate about a year before he was declared consul, did not hesitate the next day, openly in the camp, to teach that this was no portent, and that it had then so happened, and would always happen at fixed times, when the sun should so be placed that it could not reach the moon with its light. ``Do you really mean it?'' said Tubero. ``Could that man teach this to almost rustic men, and dared he say these things in the presence of the unlearned?'' ``Indeed he did, and with great \ldots''

\ciceroSection{24} \ldots[was] no insolent display, nor a speech ill-suited to the character of a most weighty man; for he had achieved a great thing, in that he had cast off vain religious fear from troubled men.

\ciceroSection{25} And something of this sort, even in that greatest of wars which the Athenians and the Lacedaemonians waged against each other in the highest contention, the famous Pericles --- the leading man of his own state in authority and eloquence and judgement --- when on the sun being darkened sudden shadows had fallen, and the highest fear had seized the minds of the Athenians, is said to have taught his fellow citizens what he himself had taken from Anaxagoras, whose pupil he had been: that this happens at a fixed and necessary time, when the whole moon has put itself under the sun's orb; and that, although it does not happen in every interlunar period, still it cannot happen except at a certain interlunar time. When he had taught this by argument and reasoning, he freed the people from fear; for at that time this was a new and unknown explanation, that the sun is wont to fail by the moon's interposition --- which they say Thales of Miletus first saw. Nor afterwards did this escape even our Ennius; who, as he writes, in about the three hundred and fiftieth year after the founding of Rome, ``On the Nones of June the moon stood in the sun's path, and night.'' And in this matter so great is the calculation and skill that, from this day, which we see marked down in Ennius and in the greatest annals, prior eclipses of the sun have been reckoned back even to that one which fell on the Nones of Quintilis when Romulus was reigning --- in which darkness, even though nature had snatched Romulus to a human end, virtue is said to have raised him to heaven.

\ciceroSection{26} Then Tubero: ``Do you see, Africanus, that what a little while ago seemed otherwise to you \ldots'' --- skilled in those things which others see. What further could anyone count splendid in human affairs, who has looked into these realms of the gods? Or lasting, who has known what is eternal? Or glorious, who has seen how small the earth is --- first the whole, then that part of it which men inhabit; and how, fastened to a tiny portion of it, unknown to most peoples, we yet hope our name will fly and roam abroad most widely?

\ciceroSection{27} As for fields and buildings and cattle and an immense weight of silver and gold --- the man who is wont neither to count nor to call them goods, because their fruit seems to him slight, their use small, their lordship uncertain, often the boundless property of the worst of men --- how fortunate he is to be reckoned! He alone may truly claim all things for his own --- not by Quiritary right, but by the right of the wise --- not by civil bond, but by the common law of nature, which forbids anything to belong to anyone except him who knows how to handle and use it. He thinks our military commands and consulships are to be undertaken among the necessary, not the desirable, things, for the sake of discharging a duty, not to be sought for their rewards or for glory. He, finally --- as Cato writes my grandfather Africanus was wont to say --- can declare the same of himself: that he is never busier than when he has nothing to do, and never less alone than when he is alone.

\ciceroSection{28} For who can truly think that Dionysius did more, when by all his contriving he tore liberty from his fellow citizens, than that citizen of his Archimedes, who, when he seemed to be doing nothing, fashioned that very globe of which we were just speaking? Or who that the men more alone are not those who, in the forum and the crowd, have no one with whom they may care to converse, than those who, with no witness, either talk with themselves, or, as it were, are present in the council of the most learned, when they delight themselves with their inventions and writings? Who, indeed, can think anyone richer than him to whom nothing is lacking that nature desires; or more powerful than him who attains all that he seeks; or more blessed than him who is freed from every disturbance of mind; or of firmer fortune than him who possesses what he can carry off with himself, as they say, even from a shipwreck? What command, what magistracy, what kingdom can be set above one who --- looking down on all human things and counting them lower than wisdom --- turns nothing in his mind ever, except what is everlasting and divine? --- one to whom it is settled that the rest are men only in name, and that those alone are men who have been polished with the proper arts of humanity ---

\ciceroSection{29} so that to me that saying of Plato's, or whoever else said it, seems most elegant: when a storm had carried him to unknown lands and a deserted shore from the deep, while the rest were afraid because of their ignorance of the place, they say he noticed certain geometrical figures drawn in the sand. When he had seen them, he cried out that they should be of good cheer; for he saw the marks of men --- which he indeed read not from any cultivation of the field, which he could see, but from the signs of learning. For which reason, Tubero, learning and learned men and those studies of yours have always pleased me.

\ciceroSection{30} Then Laelius: ``I dare not indeed, Scipio, speak in opposition; nor do I either you or Philus or Manilius \ldots'' [\ldots] in his very father's house was that friend of ours, worthy to be imitated by the man here, ``the eminently sensible man, the shrewd Aelius Sextus,'' who was both ``eminently sensible'' and ``shrewd,'' as he was called by Ennius, not because he sought after things he would never find, but because he answered things by which those who had asked were freed from their care and trouble; and to whom, when he debated against Gallus's pursuits, that line was always on his lips, of Achilles in the \textit{Iphigenia}: ``What watching is in heaven among the astrologers' signs, when the goat or the scorpion or some name of beasts rises! What is before their feet, no man looks at; the regions of heaven they search.'' And he himself --- for I heard him much, and gladly --- used to say that the Zethus of Pacuvius was too hostile an enemy of learning. He was more delighted with the Neoptolemus of Ennius, who says he ``wishes to philosophize, but in a few things; for it pleases him not at all in full.'' And if the studies of the Greeks delight you so greatly, there are others freer and spread more widely, which we can apply either to the use of life or even to the commonwealth itself. Those arts of yours, if they have any value, have only this value: that they sharpen and, as it were, prick the talents of boys, that they may more easily learn greater things.

\ciceroSection{31} Then Tubero: ``I do not disagree with you, Laelius, but I ask, what do you understand to be the greater things?'' ``I will tell you, by Hercules --- and perhaps I shall be despised by you, when you put those celestial questions to Scipio, while I think that these things which are seen before our eyes are rather to be inquired about. For why does the grandson of Lucius Paulus, with this man for an uncle, born in the noblest of families and in this so brilliant commonwealth, ask how two suns were seen, and not ask why in one commonwealth there are two senates and almost now two peoples? For, as you see, the death of Tiberius Gracchus, and even before that the whole strategy of his tribunate, divided one people into two parts. The detractors and enviers of Scipio, with the start being made by Publius Crassus and Appius Claudius, none the less, even after their death, hold the other part of the senate, dissenting from you under the lead of Metellus and Publius Mucius; nor do they suffer this man --- who alone can do it --- to bring help in matters so dangerous: with the allies and the Latin name stirred up, the treaties violated, the most seditious triumvirs daily contriving something new, the rich and good men thrown into confusion.

\ciceroSection{32} For which reason, if you will hear me, young men, do not fear the second sun. Either it cannot exist, or, if it does --- as it was seen --- so be it, only let it not be troublesome; we either know nothing of these matters, or, even if we know to the highest degree, we cannot be the better or the more blessed for that knowledge. But that we should have one senate and one people --- both can come to pass, and it is most troublesome if it does not, and we know and see that it is otherwise; and if that should be brought about, we shall live both better and more blessedly.''

\ciceroSection{33} Then Mucius: ``What then, Laelius, do you think we ought to learn, in order that we may bring to pass that very thing which you call for?'' ``Those arts which bring it about that we are of use to the state. For I think this is the most distinguished office of wisdom and the greatest either testimony or duty of virtue. For which reason, that these holidays of ours may be given to the most useful conversations on public matters, let us ask Scipio to set out what he judges to be the best constitution of a state. After that we shall ask other things; and from the knowledge of these I hope we shall arrive at the very matters that now press, and explain the reasoning of them.''

\ciceroSection{34} When Philus and Manilius and Mummius had heartily approved \ldots ``There is no example to which we should rather liken the commonwealth.'' \ldots ``For which reason, if you please, bring down your speech from heaven to these nearer things.'' \ldots I wished this to be done, not only for that reason --- because it was fitting that on the commonwealth, above all, a leading man of the commonwealth should speak --- but also because I remembered that you used very often to discuss with Panaetius, in the presence of Polybius, two Greeks most experienced in civil affairs, and to gather and to teach many things: that by far the best constitution of a state is that which our ancestors have left us. Since you are the better prepared in this argument, you will oblige all of us if, by setting out what you think about the commonwealth, you let me speak also on behalf of these others.''

\ciceroSection{35} Then he said: ``I cannot indeed say that I am wont to busy myself in any thought more keenly or more carefully than in this very one which is set before me, Laelius, by you. For when I see that every craftsman in his own art --- the one who excels --- has nothing else to think of, study, attend to, except how he may be better in his line, then how should I, when this one work has been left to me by my parents and ancestors --- the care and conduct of the commonwealth --- not confess myself more idle than any workman, if I had spent less labour in the greatest of arts than they in the smallest?

\ciceroSection{36} But I am not content with what the highest and wisest men out of Greece have left us in writing on this subject; nor do I dare to put my own views before theirs. For which reason I ask you to hear me as one neither wholly without acquaintance with Greek matters nor putting them, in this kind of subject especially, before our own; but as one of the togate, brought up not illiberally by his father's care, and from boyhood fired with the love of learning, yet trained much more by use and by the teachings of home than by literature.''

\ciceroSection{37} Here Philus: ``By Hercules, Scipio, I have no doubt that no one has gone before you in talent, and that in actual experience of the greatest matters in the commonwealth you have easily surpassed all; but the studies that have always been yours, we know. For which reason, if, as you say, you have applied your mind also to this method, this art, I am most grateful to Laelius. For I hope that what is going to be said by you will be much richer than all that has been written by the Greeks.'' Then he: ``You lay a great expectation --- the heaviest of all burdens upon a man who is to speak on great matters --- upon my speech.'' And Philus: ``However great it is, you will yet exceed it, as you are wont to do; for there is no danger that speech should fail you when discoursing on the commonwealth.''

\ciceroSection{38} Here Scipio: ``I shall do as you wish, as I am able, and shall enter on the discussion under that rule which I think must be used in arguing every kind of subject if you wish to remove error: that the name of the thing about which the inquiry is to be, if it is agreed, should be unfolded as to what is signified by that name. If this is agreed, then at length it will be fitting to enter into discourse; for what kind of thing is in question can never be understood unless what it is has first been understood. Therefore, since we are inquiring about the commonwealth, let us first see what that very thing is which we are seeking.'' When Laelius had approved, ``Yet I shall not so discuss,'' said Africanus, ``a matter so brilliant and so well known that I roll back to those elementary points which learned men are wont to use in these matters --- to begin from the first meeting of male and female, then from offspring and kinship, and to define more often, in words, what each thing is and how many ways it may be said. For when I am speaking among prudent men, in the greatest of commonwealths, men who have moved with the highest glory in war and at home, I will not commit the fault of having my speech less brilliant than the very thing under debate. Nor have I taken this on, that I should treat all things like a schoolmaster, nor do I promise to bring it about that no particular shall be passed over in this discussion.'' Then Laelius: ``That very kind of speech which you promise is what I look for.''

\ciceroSection{39} ``The commonwealth, then,'' said Africanus, ``is the affair of the people; but the people is not every assembly of men gathered together in any way whatever, but an assembly of a multitude associated by an agreement of right and a sharing of common interest. The first cause of its coming together is not so much weakness as a certain natural --- as it were --- gathering of men. For this kind is not solitary nor wandering alone, but is so brought into being that, even amid an abundance of all things \ldots

\ciceroSection{40fr} \textit{[Augustine quotes Cicero:]} In a brief time, the dispersed and wandering multitude had become, by concord, a single state.

\ciceroSection{41} \ldots certain seeds, as it were, are to be found, nor any institution of the other virtues nor of the commonwealth itself. These assemblies, then, having been formed for the cause I have set out, established their seat first in a fixed place, for the sake of dwellings; and when they had hedged it about with sites and with the work of their hands, they called such a connection of roofs a \textit{town} or a \textit{city}, marked off with sanctuaries and common spaces. Every people, therefore, which is such an assembly of a multitude as I have described --- every state, which is the constitution of a people --- every commonwealth, which, as I said, is the affair of a people --- must be ruled by some counsel, that it may last. And this counsel must always first of all be referred to that cause which begot the state.

\ciceroSection{42} Then it must be entrusted either to one, or to certain chosen men, or it must be taken on by the multitude and by all. When the supreme power over all things is in the hands of one, we call him a king and the constitution of that commonwealth a kingdom. When it is in the hands of chosen men, then that state is said to be ruled by the will of the best men. But that state is the people's (so they call it) in which all things are in the people. Of these three kinds, any one --- if it should hold that bond by which men were first bound together in the partnership of a commonwealth --- is not, indeed, perfect, nor in my judgement the best, but is yet endurable, in such a way that one may be more excellent than another. For either a just and wise king, or chosen and leading citizens, or even the people itself --- though this is least to be commended --- can yet, when no inequalities or appetites have come between, seem to be in some not uncertain state.

\ciceroSection{43} But both in monarchies the rest are too far excluded from the common law and counsel; and in the rule of the best men the multitude can hardly share in liberty, when it is cut off from all common counsel and power; and when all things are done by the people, however just and moderate, that very evenness is unfair, when it has no degrees of standing. So, if the famous Cyrus the Persian was the most just and wise of kings, still it does not seem to me that an affair of the people (for so it is, as I said before, the public affair) was greatly to be sought, since it was being ruled by the nod and at the will of one man. If the Massiliotes, our clients, are ruled with the highest justice by chosen and leading citizens, there is yet in that condition of the people a kind of likeness of slavery. If at certain times the Athenians, the Areopagus having been removed, did nothing except by decrees and resolutions of the people --- since they had no marked degrees of standing --- the state did not keep its proper ornament.''

\ciceroSection{44} ``And I am speaking now of these three kinds of commonwealth, not as disordered and mixed up, but as keeping each its own constitution. Now these kinds, first of all, have in themselves the faults I just spoke of, and afterwards have other ruinous faults; for there is no kind of commonwealth which has not, set toward some neighbouring evil, a road both headlong and slippery. For under that king --- to name him by preference, the endurable or, if you will, even the lovable Cyrus --- there lies, ready for the moment when his mind turns wanton, the most cruel Phalaris, into whose likeness the rule of one man slips by an easy and downhill course. To that Massiliote administration of the state by a few leading men there is neighbour the conspiracy and faction of the Thirty which, at one time, the Athenians had. As for the Athenian people's power over all things --- without seeking other examples --- itself, when turned to the frenzy and licence of the multitude, [becomes] a plague \ldots

\ciceroSection{45} \ldots the worst, and out of this either of optimates or of factions a tyranny --- whether kingly or even very often popular --- and likewise out of that some kind of one of those I named before is wont to flower forth; and wonderful are the cycles and, as it were, the circuits of changes and vicissitudes in commonwealths. To know these is the part of a wise man; but to look ahead to those impending, in steering the commonwealth, moderating its course and keeping it in his own power, is the part of a great citizen, and almost of a divine man. Therefore I think a fourth kind of commonwealth is to be most approved, which is moderate and mixed out of the three I named first.

\ciceroSection{46} Here Laelius: ``I know it is so pleasing to you, Africanus; for I have often heard it from you. But still, if it is no trouble, I should like to know which of the three kinds of commonwealth you judge the best. For it will be of some help toward understanding \ldots''

\ciceroSection{47} \ldots and every commonwealth is such as the nature or will of him who rules it. Therefore in no other state does liberty have any home except in that in which the people's power is supreme; than which, indeed, there can certainly be nothing sweeter, and which, if it is not equal, is not even liberty. But how can it be equal --- I forbear to say in a kingdom, where servitude is not even hidden or doubtful, but in those states in which all are free in word? They cast their votes, they entrust commands and magistracies, they are canvassed, asked: but they give what, even if they did not wish to, must be given, and what they themselves do not have, others ask of them. For they are without a share in command, in public counsel, in the judgement of chosen judges, weighed as those things are by the antiquity of families or by money. But in a free people, as at Rhodes, as at Athens, there is no citizen who \ldots

\ciceroSection{48} \ldots [when] in a people one or more had stood out as richer and more powerful, then out of their disdain and arrogance, they record, that situation arose --- with the cowardly and weak yielding, and giving way to the arrogance of the rich. But if a people hold their own right, they say nothing is more excellent, freer, more blessed --- since they are masters of laws, of judgements, of war, peace, treaties, of every man's life and money. This alone, they think, is rightly called a commonwealth, that is, the affair of a people. So both from the dominion of kings and from that of the senate the affair of the people is wont to be claimed back into liberty: but kings are not asked for from free peoples, nor the power and resources of optimates.

\ciceroSection{49} And indeed they say that the whole kind of free people ought not to be cast off because of the fault of an unbroken people: where the people is concordant and refers everything to its own safety and liberty, nothing is more unchanging, nothing more firm. But concord is most easily found in that commonwealth in which the same thing is to the advantage of all; from varieties of interest, when one thing is helpful to some and another to others, discords are born. So when the senators got hold of affairs, the constitution of the state was never fixed; and much less so in monarchies, of which Ennius says: ``no holy partnership of a kingdom is there, nor any faith.'' Wherefore, since law is the bond of civil society, and the right of law is equal, by what right can a society of citizens be held together when the condition of the citizens is not equal? For if it is not pleasing that monies be made equal, if the talents of all cannot be equal, the rights, certainly, of those who are citizens in the same commonwealth ought to be equal among themselves. For what is a state, if not a partnership of right?

\ciceroSection{50} The other commonwealths, indeed, they think are not even to be called by those names with which they wish themselves to be called. For why should I call by the name of Jupiter Optimus a man eager for domination, or for one-man rule, lording it over an oppressed people --- and not rather a tyrant? For a tyrant can be as merciful as a king burdensome; and the only difference for peoples is whether they serve a kindly master or a harsh one --- but that they should serve, that cannot fail to happen. And how was that famous Lacedaemon able, at the time when its constitution was held to be excellent, to make use of good and just kings, when whoever was born of royal stock had to be held to be king? For the optimates --- who would put up with men who, not by the people's grant but by their own elections, have arrogated this name to themselves? For how is that man judged the best? By learning, arts, pursuits \ldots

\ciceroSection{51} \ldots if it does this by chance, it will be overturned as quickly as a ship that has come to the helm with a man drawn by lot from among the passengers. But if a free people choose to whom to entrust themselves --- and they will choose, if only they wish to be safe, the best men --- then certainly the safety of states lies in the counsels of the best men, especially since nature has so brought it about that not only do those highest in virtue and spirit preside over the weaker, but the weaker also wish to obey the highest. But this best constitution, they say, has been overthrown by the perverse opinions of men, who, through ignorance of virtue --- which is found among few, and is judged and discerned by few --- think that wealthy and well-stocked men, or those born of noble family, are the best. By this error of the crowd, when the wealth of a few, not their virtues, has begun to hold the commonwealth, those leading men still hold tooth and nail to the name of optimates --- but in fact they are without that name. For riches, name, and resources, devoid of counsel and of any decent way of living and of ruling others, are full of disgrace and of insolent pride; and there is no uglier face of a state than that in which the wealthiest are accounted the best.

\ciceroSection{52} But when virtue rules the commonwealth, what can be more brilliant? When he who commands others is himself slave to no desire; when he himself has embraced all those things to which he calls and trains the citizens; when he imposes no laws on the people which he himself does not obey, but sets up his own life as a law before his fellow citizens. If one such man could attain everything sufficiently, there would be no need of more; if all could see what was best and could agree on it, no one would seek out chosen leaders. The difficulty of taking counsel transferred the matter from the king to the many; the error and rashness of peoples have transferred it from the multitude to the few. So between the weakness of one and the rashness of many, the optimates have held the middle place, than which nothing can be more measured. With them as guardians of the commonwealth, peoples must necessarily be most blessed, free of all care and thought, when their leisure has been entrusted to others by whom it must be guarded; nor must it ever be allowed that the people should think their own concerns neglected by the leading men.

\ciceroSection{53} For that evenness of right which free peoples embrace cannot even be preserved (for the very peoples themselves, however loose and unrestrained, give much in the matter of high office to many men, and in them there is a great selecting of men and of standings); and what is called \textit{evenness} is most uneven. For when equal honour is held for the highest and the lowest --- and these necessarily exist in every people --- that very fairness is most unfair. This cannot happen in those states which are ruled by the best men. These things, more or less, Laelius, and certain things of the same kind, are wont to be argued by those who most praise that form of commonwealth.''

\ciceroSection{54} Then Laelius: ``What of you, Scipio? Of those three, which do you most approve?'' ``You ask rightly which of the three I most approve, since I do not approve any of them by itself, separately; and I prefer to each one alone that which has been blended out of them all. But if one and simple [must be approved], I should approve the kingly \ldots [textual gap] \ldots in this place is named, the name of the king as if of a father comes up first --- as one taking thought, like a parent, for citizens born from himself, and preserving them \ldots [\ldots] sustained by the diligence of one best and supreme man.

\ciceroSection{55} ``Up come the optimates, who profess that they will do this same thing better, and say that there will be more counsel in many than in one, and yet the same fairness and trust. And lo, the people calls out with the loudest voice that it does not wish to obey one man, nor a few; that nothing is sweeter than liberty, even to wild beasts; and that of this all are deprived, whether they serve king or optimates. Thus kings take us by affection, optimates by counsel, peoples by liberty, so that in comparing it is hard to choose what one would most desire.'' ``I believe you,'' he said, ``but it will scarcely be possible to set out the rest, if you leave this matter half-finished.''

\ciceroSection{56} ``Let us imitate then Aratus, who, beginning to speak of great matters, thinks one must begin from Jupiter.'' ``What Jupiter? Or what does this discourse of ours have like that poem?'' ``This much,'' said he, ``that we may rightly take from him the beginning of speech --- whom alone, of all gods and men, the learned and the unlearned alike consent to call king. For why?'' said Laelius. And he: ``What do you suppose, except what is before our eyes? Whether these things have been ordained for the use of life by the leaders of commonwealths, that there should be thought to be one king in heaven, who, as Homer says, with a nod turns the whole of Olympus, and is held to be both king and father of all --- great is this authority, many are these witnesses, if all those many can be called witnesses; so all peoples have agreed, plainly by the decrees of leading men, that nothing is better than a king, since they think that all the gods are ruled by the will of one. Or, if these are matters of the ignorant in error and we have learned them as like fables, let us hear, as it were, the common teachers of learned men, who have seen with their eyes, as it were, what we scarcely know by hearing.'' ``Who, pray,'' said Laelius, ``are these?'' And he: ``Those who, in tracking out the nature of all things, have understood that this whole world by mind \ldots

\ciceroSection{58} But, if you wish, Laelius, I will give you witnesses neither too ancient nor in any way barbarous.'' ``Those,'' he said, ``I want.'' ``Do you see, then, that this city has been less than four hundred years without kings?'' ``Indeed less.'' ``Well then, is this age of four hundred years for a city or a state very long?'' ``That, indeed,'' he said, ``has scarcely come of age.'' ``Therefore four hundred years ago there was a king at Rome?'' ``And a proud one at that.'' ``And before him?'' ``A most just king, and so back in succession to Romulus, who was king six hundred years from this time.'' ``Then not even he is very ancient?'' ``Hardly, and almost when Greece was already growing old.'' ``Tell me, was Romulus, Scipio, a king of barbarians?'' ``If, as the Greeks say, all are either Greeks or barbarians, I fear he was a king of barbarians; but if that name is to be given by morals, not by tongues, I think the Greeks no less barbarians than the Romans.'' And Scipio: ``But for the matter at hand we are not asking about race; we are asking about minds. For if prudent men, and not ancient ones, wished to have kings, I am using witnesses neither very ancient nor uncivilized and savage.''

\ciceroSection{59} Then Laelius: ``I see you, Scipio, well furnished with witnesses; but with me, as with a good judge, arguments are worth more than witnesses.'' Then Scipio: ``Use, then, an argument, Laelius --- your own feeling.'' ``What feeling?'' said he. ``If ever, if perchance, you have seemed to yourself to be angry at someone.'' ``More often, indeed, than I should wish.'' ``What then? When you are angry, do you allow that anger lordship over your mind?'' ``By Hercules, no,'' he said; ``but I imitate that famous Archytas of Tarentum, who, when he had come to his country house and had found everything otherwise than he had ordered, said to his bailiff: ``Wretch, I should have killed you with stripes, if I were not angry.'''' ``Excellent,'' said Scipio.

\ciceroSection{60} ``Therefore Archytas, plainly, considered anger to be at odds with reason, a kind of sedition of the mind, and wished it to be quieted by counsel. Add greed; add the desire of empire, the desire of glory; add the lusts. And you see this: that if there is to be a kingly command in the minds of men, it will be the lordship of one --- of counsel, plainly (for that is the best part of the mind) --- and that, with counsel ruling, no place will be left to the lusts, none to anger, none to rashness.'' ``It is so,'' he said. ``Do you approve, then, the mind so disposed?'' ``Indeed, nothing more.'' ``Then you would not approve, if, with counsel driven out, the lusts (which are countless) or the angers held all things?'' ``I should think nothing more wretched than that mind, no man more so than one so disposed.'' ``Are you content, then, that all parts of the mind be under a kingdom, and ruled by counsel?'' ``So I am content.'' ``Why, then, do you hesitate as to your judgement on the commonwealth? In which, if the matter has been transferred to many, it can already be understood that there will be no command to preside over them; for command, unless it be one, can be no command.''

\ciceroSection{61} Then Laelius: ``What is the difference, pray, between one and many, if there is justice in the many?'' And Scipio: ``Since I have understood, Laelius, that you are not greatly moved by my witnesses, I shall not stop using you yourself as a witness, to prove what I am saying.'' ``Me,'' said he; ``in what way?'' ``Because I noticed lately, when we were at Formiae, that you very strictly forbade your household to obey anyone but one.'' ``The bailiff, of course.'' ``What? At home, are there several over your business?'' ``Indeed, no, but one.'' ``What? Does anyone else besides you rule the whole household?'' ``Not at all.'' ``Why, then, do you not concede that the same in a commonwealth --- the rule of single men --- is best, if only they are just?'' ``I am led,'' he said, ``almost to assent.''

\ciceroSection{62} And Scipio: ``Then you will assent the more, Laelius, if --- to leave aside likenesses --- it is more proper to entrust the ship to one helmsman, the sick man to one doctor (provided they are worthy of those arts), than to many; and I shall come to greater things.'' ``What things are these?'' ``Don't you see that, by the unbearableness and pride of one Tarquin, the name of king came to be hated by this people?'' ``I do see it.'' ``Then you also see this --- on which I think I shall say more as the discussion goes on --- that, with Tarquin driven out, the people exulted in a wonderful insolence of liberty: then innocent men were driven into exile, then the goods of many were plundered, then consuls were yearly, then the rods were lowered to the people, then there were appeals on every matter, then the secessions of the plebs --- then in short most things were done in such a way that all things were in the people.'' ``It is,'' he said, ``as you say.''

\ciceroSection{63} ``So it is,'' said Scipio, ``in peace and leisure: for there is room for being playful, while you fear nothing --- as on a ship, and often even in a slight illness. But just as the man who is sailing, when the sea has begun to grow rough, and the sick man, when the disease grows worse, calls upon the help of one --- so our own people, in peace and at home, commands and threatens the magistrates themselves, refuses, appeals, exercises provocation; but in war it obeys as if a king. For safety counts more than caprice. In heavier wars our forefathers wished even all command, without a colleague, to lie in the hands of single men: whose very name shows the force of their power. For \textit{dictator} is so called from \textit{dico} (`I declare'), but in our books, Laelius, you see him called the master of the people.'' ``I do see it.'' And Scipio: ``Wisely then, those men of old \ldots

\ciceroSection{64} \ldots when a people is bereaved of a just king, ``their hearts long while are held by yearning,'' as Ennius says, ``after the death of the best king''; and they say among themselves: ``O Romulus, divine Romulus, what guardian of his country the gods made you! O father, O begetter, O blood sprung from the gods!'' They did not call those whom they justly obeyed masters, nor lords, nor even kings, but guardians of their country, but fathers, but gods; and not without cause. For what do they add? ``You led us forth into the regions of light.'' Life, honour, dignity --- they thought were given them by the king's justice. The same goodwill would have remained in their descendants, had the likeness of the kings remained; but you see, by the injustice of one, that whole kind of commonwealth fell.'' ``I see it indeed,'' he said, ``and I am eager to learn these courses of changes, no more in our own commonwealth than in every commonwealth.''

\ciceroSection{65} And Scipio: ``After I have said what I think on that kind of commonwealth which I most approve, I shall need to speak more carefully on the changes in commonwealths, although I think they will be by no means easy to come about in this commonwealth. But the first and surest change of this kingly form is this: when the king has begun to be unjust, that kind perishes on the spot, and there is the very same man as a tyrant --- the worst kind, and neighbour to the best. If the optimates have crushed him --- as commonly happens --- the commonwealth holds the second of the three constitutions; for it is, as it were, kingly, that is, the fatherly counsel of leading men taking thought for the people's good. But if the people of itself has slain or cast out a tyrant, it is moderate, so long as it feels and is wise; it rejoices in what it has done and wishes to preserve the commonwealth set up by itself. Whenever the people has either done violence to a just king, or stripped him of his kingdom, or even (which more often happens) tasted the blood of optimates and laid the whole commonwealth under its own caprice (and beware lest you think any sea or any flame so great that it cannot be calmed more easily than a multitude unrestrained by insolence), then comes about that which is brilliantly said in Plato --- if I can reproduce it in Latin; it is hard to do, but I shall try.

\ciceroSection{66} ``For,'' he says, ``when the unfillable throats of the people have parched with the thirst of liberty, and, using ill servants, the people, in its thirst, has drained --- not moderately tempered, but too unmixed --- this liberty, then it pursues the magistrates and leading men, unless they are very mild and lax and minister liberty largely to it; arraigns them, accuses them, and calls them domineering, kings, tyrants.'' I think these are known to you.'' ``Indeed,'' he said, ``most known.''

\ciceroSection{67} ``Then this follows: ``those who obey the leading men are harassed by that people and called voluntary slaves; but those who, in office, wish to be like private men, and those private men who bring it about that nothing should differ between private and magistrate --- they bear them with praises and load them with honours; so that of necessity in such a commonwealth, full of liberty, every private home is empty of any rule, and this evil reaches even to the beasts; finally, the father fears the son, the son disregards the father, all shame is gone, that they may plainly be free; nothing differs whether one be a citizen or a foreigner, the master fears his pupils and flatters them, and the pupils despise their masters; the youths take on the weight of the old, the old come down to the play of the youths so as not to be hateful and heavy to them. From which it happens that even slaves carry themselves more freely; the wives are by the same right as their husbands; and amid such liberty even the dogs and the horses, and finally the asses, run about so freely that one must give way to them in the street. Therefore from this boundless,'' he says, ``licence, this sum is gathered: that the minds of citizens grow so squeamish and soft that, if the slightest force of authority be applied, they grow angry and cannot bear it. From which they begin even to neglect the laws, that they may plainly be without any master.'''

\ciceroSection{68} Then Laelius: ``You have rendered exactly,'' he said, ``what was said by him.'' ``And, to come back at last to the author of my own discourse, from this excessive licence, which they alone think liberty, he says that, as if from a stock, the tyrant springs up and is, as it were, born. For, just as from the excessive power of leading men comes the destruction of leading men, so this people, too free, is afflicted with slavery by liberty itself. Thus all things in excess --- when whether in weather or in fields or in bodies they have been too prosperous --- mostly turn into their opposites; and this happens above all in commonwealths. And that excessive liberty, both for peoples and for private men, falls into excessive servitude. Therefore from this greatest liberty is begotten the tyrant and that most unjust and harshest servitude. For from this unbroken, or rather wild people, some leader is generally chosen, against those leading men already cast down and driven from their place: bold, foul, savagely persecuting men often well-deserving of the commonwealth, gratifying the people with both his own and others' goods. Because, as a private man, fears stand in his way, commands are given him, and these are continued; he is even hedged about with bodyguards (as Pisistratus at Athens), and at last from those by whom he was brought forward arise the tyrants of those very men. If good men have crushed them (as often happens), the state recovers; but if bold ones, then arises that faction, another kind of tyrants. And the same thing arises often even from that brilliant condition of the optimates, when some perversity has turned the leading men themselves from the way. So, like a ball, tyrants snatch the constitution of the commonwealth from kings, leaders or peoples from them, and from these in turn factions or tyrants; and never is the same form of commonwealth held for long.

\ciceroSection{69} ``Such being the case, of the three primary kinds the kingly is, in my judgement, by far the best; but to the kingly itself will be preferred that which is evened out and tempered out of the three best forms of commonwealths. For it is fitting that there should be in the commonwealth something outstanding and kingly; that there should be another part shared with and granted to the authority of leading men; and that certain things should be reserved to the judgement and will of the multitude. This constitution has, first, a certain great evenness, without which free men can scarcely live for long; secondly, firmness; for those primary forms easily turn into their opposite vices, so that out of the king there comes a master, out of the optimates a faction, out of the people a mob and confusion; and the kinds themselves are often changed for new kinds. But in this combined and moderately mixed conformation of the commonwealth this generally does not happen without great vices in the leading men. For there is no cause of overthrow when each is firmly placed in his own grade, and there is nothing into which he might fall headlong.

\ciceroSection{70} ``But I fear, Laelius, and you most friendly and most prudent men, that if I dwell longer on this kind, my speech will seem to be that of one prescribing or teaching, and not considering with you. Therefore I shall enter on those things which are known to all, and inquired after by us long since. For so I judge, so I think, so I affirm: no commonwealth, in either constitution or arrangement or discipline, is to be compared with that which our fathers have left us, having received it from our forefathers. And this --- if you please, since you wished to hear from me also that which you yourselves already grasped --- I shall show, both what it is and that it is the best; and, with our own commonwealth set out as an example, I shall fit to it, if I can, that whole speech which I have to give on the best constitution of a state. If I am able to hold and follow this through, I shall, in my opinion, have abundantly discharged this duty which Laelius laid upon me.''

\ciceroSection{71} Then Laelius: ``Yours, indeed, Scipio,'' he said, ``and yours alone. For who better than you should speak either of the institutions of our forefathers, since you yourself are sprung from the most distinguished forefathers? Or of the best constitution of a state? --- which, if we have it (though hardly even now), then who could be more flourishing than you? Or of counsels for the future, since you, having driven away two terrors of this city, have looked forward to all time?''

\bigskip
\noindent\textit{[Book 1 fragments preserved by other authors]}

\ciceroSection{1fr} \textit{[Pliny, Natural History, preface 7:]} Nor [is this] for the most learned. ``I do not wish a Manius Persius to read these things; I want a Junius Congus.''

\ciceroSection{2fr} \textit{[Nonius p. 426M:]} So, since one's country contains more benefits and is an older parent than he who begot us, surely greater gratitude is owed to it than to a parent.

\ciceroSection{3fr} \textit{[Nonius p. 526M:]} Nor would Carthage have had so great resources for almost six hundred years without counsels and discipline.

\ciceroSection{4fr} \textit{[Nonius p. 276M:]} ``By Hercules,'' he said, ``[I want] to come to know that habit and pursuit of speech.''

\ciceroSection{5fr} \textit{[Lactantius, Divine Institutes 3.16.5:]} Surely the whole disputation of those men, although it contains the richest sources of virtue and knowledge, when set beside their actions and finished deeds, brings to mankind, I fear, less of business than of delight.

\ciceroSection{6fr} \textit{[Arusianus Messius:]} ``From which those people were calling [him] away.''

\ciceroBook{2}

\ciceroSection{1} When all were burning with eagerness to hear, Scipio began to speak as follows: ``This is the saying of old Cato, whom, as you know, I loved uniquely and most of all admired, and to whom, by the judgement of both my fathers and by my own zeal, I gave myself wholly from boyhood; whose talk could never tire me; so great was the man's experience of public business --- which he had handled both at home and on campaign, both excellently and very long --- and so was the moderation in his speaking, the charm mingled with weight, the highest passion either for learning or for teaching, and a life corresponding fully with his speech.

\ciceroSection{2} ``He used to say that the constitution of our state stood out before others for this reason: that in those others there had usually been single men who had each set up his own state by his own laws and ordinances --- as Minos for the Cretans, Lycurgus for the Lacedaemonians, and at Athens (which had been so often changed) now Theseus, now Draco, now Solon, now Cleisthenes, then many others, finally the learned Demetrius of Phalerum, who had propped up that state already pale and lying down. But our commonwealth had been set up by the genius not of one but of many, nor in the lifetime of one man, but over several ages and generations. For he said that no genius had ever existed so great that nothing escaped him; nor could all the geniuses gathered into one foresee enough at one time to embrace all things, without the use and antiquity of affairs.

\ciceroSection{3} ``For which reason, as he was wont to do, so now my speech will go back to the origin of the people; for I gladly use Cato's word too. I shall the more easily attain what I propose, if I show you our commonwealth being born and growing and grown adult and now firm and strong, than if I should myself fashion some commonwealth, as in Plato Socrates does.''

\ciceroSection{4} When all had approved this, ``Have we,'' he said, ``a beginning of a state's establishment so famous and so well known to all as the founding of this city, set forth from Romulus? Born of his father Mars (for let us grant the report of men, especially when it is not only ancient but also wisely handed down by our forefathers, that those well-deserving of the common cause should be thought, even by descent, divine in their nature, not only in their genius) --- he, then, as soon as he was born, is said, with his brother Remus, to have been ordered exposed beside the Tiber by Amulius, king of Alba, on account of his fear of the kingdom's being undermined. In which spot, when he had been suckled by the dugs of a wild beast of the woods, and shepherds had taken him up and reared him in rustic life and toil, it is reported that, when he had grown up, he so excelled the others both in the strength of his body and the fierceness of his spirit that all who at that time inhabited those fields, where this city now stands, gladly and willingly obeyed him. With whose forces, when he had given himself out as their leader (to come now from fables to facts), he is said to have crushed Long Alba, a strong and powerful city in those times, and to have killed King Amulius.

\ciceroSection{5} ``With this glory won, he is said to have first considered, with proper auspices, founding and confirming the city as his commonwealth. The site of the city --- which must be most carefully provided for by him who tries to plant a long-lasting commonwealth --- he chose with incredible advantage. He did not bring it down to the sea --- which would have been very easy for him with that band and those troops, that he should advance into the field of the Rutulians and the Aborigines, or himself found a city at the mouth of the Tiber, where many years later King Ancus led a colony --- but this man, with outstanding foresight, perceived and saw that maritime sites are not the most advantageous for those cities which are founded in the hope of long endurance and of empire: first, because maritime cities are not only exposed to many dangers, but also to hidden ones.''

\ciceroSection{6} ``For dry land announces beforehand the coming of enemies --- not only those expected, but even sudden ones --- by many signs and as it were by some crash and noise itself; nor indeed can any enemy come flying upon us by land but that we can know not only that he is, but also who and from where he is. But that maritime and naval enemy can be present before anyone could even suspect he is going to come; nor, when he has come, does he display himself, who he is or whence he comes or even what he wants; finally, no mark whatever can distinguish or judge whether he is at peace or at war.

\ciceroSection{7} ``There is, moreover, in maritime cities also a certain corruption and changing of morals; for new tongues and new disciplines are mixed in, and not only foreign goods are brought in but also foreign morals, so that nothing in the country's institutions can remain whole. Those who inhabit such cities do not stick fast in their seats, but are always swept away by winged hope and thought from home toward the further off; and even when they remain in body, in mind they are exiled and wander. Nor indeed did anything more upset Carthage and Corinth, long shaken, and at last overthrew them, than this error and dispersal of citizens: that, in the desire of trading and seafaring, they had abandoned the cultivation of fields and arms.

\ciceroSection{8} ``Many enticements to luxury, ruinous to states, are also supplied by the sea, whether captured or imported; and the very pleasantness has many lures, costly or slothful, to desire. And what I said of Corinth, I do not know whether it may be said most truly of the whole of Greece. For the Peloponnese itself is almost entirely on the sea, and except for the Phliasians none have fields that do not touch the sea; and outside the Peloponnese only the Aenianes, the Dorians, and the Dolopes are away from the sea. What shall I say of the islands of Greece? Which, girt round by waves, almost float --- and the institutions and morals of the cities themselves with them.

\ciceroSection{9} ``And these things, as I said above, are of old Greece. As for the colonies sent by the Greeks to Asia, Thrace, Italy, Sicily, Africa --- which is there, except Magnesia alone, that is not bathed by the wave? So the rim of Greece seems, as it were, to be sewn onto the lands of barbarians; for among the barbarians themselves none were before this maritime, except the Etruscans and Carthaginians: the one for the sake of trade, the other of piracy. The cause of the misfortunes and changes of Greece is plain through these vices of maritime cities, which I touched on briefly a little before. Yet there is in these vices that great convenience, that whatever is born anywhere can sail to the city you inhabit, and again that what your fields produce, you can carry and send to whatever lands you wish.

\ciceroSection{10} ``How then could Romulus more divinely have embraced the advantages of the sea, and avoided its vices, than by placing the city on the bank of an unfailing and even-flowing river broadly emptying into the sea? --- a river by which the city could both receive from the sea what it lacked, and send back what overflowed; and by the same river not only \ldots from the sea, but also receive things imported from the land, the things most necessary for living and culture; so that he seems to me, even then, to have divined that this city would one day be the seat and home of supreme empire. For hardly could a city set in any other part of Italy have held so great a power.

\ciceroSection{11} ``As for the city's own natural defences, who is so careless as not to hold them noted in his mind and plainly known? --- whose track and circuit of wall, marked out by the wisdom of Romulus and of the rest of the kings, is on every side hemmed by steep and sheer mountains, so that the one approach which lay between the Esquiline and the Quirinal hills was protected by a most vast embankment thrown up against it and a most extensive ditch surrounding it. The citadel, so fortified by the surrounding cliff, and as it were by sheer rock cut on every side, gleams forth so safely that, even in that horrible time of the Gallic coming, it remained whole and untouched. He chose, too, a place rich in springs, healthful in a pestilent region; for there are hills which, being themselves blown through by winds, also bring shade upon the valleys.

\ciceroSection{12} ``And these things he did very swiftly. For he both founded the city, which he ordered to be named Rome from his own name, and, to strengthen the new state, followed a counsel new and somewhat rustic, but for fortifying the resources of his kingdom and his people the counsel of a great man, and even then far-foreseeing: when the Sabine virgins, born of an honest place, had come to Rome for the games which he had then first established as yearly in the circus, he ordered them to be carried off at the Consualia, and he placed them in the marriages of the most distinguished families.

\ciceroSection{13} ``For which cause, when the Sabines had brought war upon the Romans, and the contest of battle had been varying and doubtful, he struck a treaty with Titus Tatius, king of the Sabines, with the very matrons who had been carried off pleading for it. By this treaty he both took the Sabines into the citizenship, with the rites shared, and joined his kingdom with their king's.

\ciceroSection{14} ``After Tatius's death, when all rule had fallen back to him, although he had with Tatius chosen leading men into the kingly council (who were called \textit{patres} on account of the affection borne them) and had marked off the people, in his own and Tatius's name and in the name of Lucumo, who as Romulus's ally had fallen in the Sabine battle, into three tribes and thirty curiae (which curiae he named from the names of those who had been ravished from the Sabines and afterwards were entreaters of peace and of the treaty) --- but although these things had been so arranged in Tatius's lifetime, yet, with him slain, Romulus reigned much more by the authority and counsel of the senators.

\ciceroSection{15} ``By which deed he first saw and judged the same thing which Lycurgus had seen at Sparta a little before: that with one command and royal power states are then better steered and ruled, when the authority of every best man is added to that force of dominion. Therefore, supported and protected by this counsel, this senate as it were, he both waged many wars most fortunately with neighbours, and, although he himself carried back nothing of the spoil to his own house, did not cease to enrich his fellow citizens.

\ciceroSection{16} ``Then --- a thing we hold today with great safety to the commonwealth --- Romulus most of all paid heed to auspices. For he himself --- and this was the beginning of the commonwealth --- founded the city after taking the auspices; in instituting all public matters, he co-opted, as augurs to be in his auspices, single men out of the single tribes; and he held the plebs marked off into the clienteles of the leading men (of which I shall see later how great was the use); and he restrained men by the imposition of fines in sheep and oxen (for at that time wealth was in herds and possession of land --- whence men of money and substance were called \textit{pecuniosi} and \textit{locupletes}), not by force or punishments.

\ciceroSection{17} ``And Romulus, when he had reigned thirty-seven years and had brought forth those two excellent supports of the commonwealth, the auspices and the senate, attained so much that, when, the sun being suddenly darkened, he had not appeared, he was thought to have been placed in the number of the gods --- which opinion no mortal could ever attain without an extraordinary glory of virtue.

\ciceroSection{18} ``And this is the more to be wondered at in Romulus, because the others who are said to have been made gods out of men were of less learned ages of mankind, so that the impulse of fashioning was easy, since the unlearned were readily driven to belief; but we see that Romulus's age was less than six hundred years ago, when letters and learning were already long-established, and all that error of the old uncultivated life of men had been removed. For if --- as is investigated in the annals of the Greeks --- Rome was founded in the second year of the seventh Olympiad, into that age fell the time of Romulus, when Greece was already full of poets and musicians and less faith was placed in fables, except about old matters. For one hundred and eight years after Lycurgus had begun to write his laws, the first Olympiad was held --- which some, by an error of the name, think was set up by the same Lycurgus. As for Homer, those who place him latest set him about thirty years before the age of Lycurgus.

\ciceroSection{19} ``From which it can be understood that Homer was many years before Romulus, so that already, when men were learned and the times themselves educated, there was scarcely any place left for fashioning. For antiquity received fictitious fables, sometimes even \textit{[as Augustine quotes:]} crudely; but this present age, already cultivated, especially when it makes sport of all that cannot happen, rejects them.

\ciceroSection{20} \ldots his grandson, as some have said, by his daughter. And in the same year as that man died Simonides was born, in the fifty-sixth Olympiad --- so that it can the more easily be understood that the immortality of Romulus was believed in then, when the life of men had grown old and was handled and known. But surely such was in him the force of genius and of virtue, that what was believed of Romulus was believed by Proculus Julius, a country man, which men in many earlier ages had believed of no other mortal: who at the urging of the senators (that they might drive away from themselves the suspicion of having killed Romulus) is reported to have said in the assembly that Romulus had been seen by him on that hill which is now called the Quirinal; that he had charged him to ask the people that a shrine be made him on that hill; that he was a god and was called Quirinus.

\ciceroSection{21} ``Do you see, then, that by the counsel of one man not only was a new people brought to birth, nor left wailing in its cradle, but already grown adult and almost come of age?'' Then Laelius: ``We see indeed; and that you have entered upon a new method of arguing, which is nowhere in the books of the Greeks. For that prince of writers, than whom none was more excellent in writing, took for himself an empty plot in which to build up a state by his own choice --- splendid no doubt, but at odds with the life and morals of men.

\ciceroSection{22} The rest discoursed without any fixed example or pattern of a commonwealth on the kinds and the methods of states; you seem to me about to do both. For you have so begun that what you yourself have found you would rather assign to others than (as Socrates does in Plato) yourself fashion, and recall to system those things about the city's site which were done by Romulus by chance or necessity; and you debate not with a wandering speech, but with one fixed in one commonwealth. Therefore go on as you have begun. For I seem to look ahead, while you trace through the rest of the kings, to a commonwealth as it were complete.''

\ciceroSection{23} ``Therefore,'' said Scipio, ``when that senate of Romulus, made up of optimates --- to whom the king himself had granted so much that he wished them to be called \textit{patres} and their children patricians --- tried, after Romulus's departure, themselves to rule the commonwealth without a king, the people would not bear it, and out of yearning for Romulus did not afterward cease to demand a king. Then those leading men prudently devised a new way of taking up the interregnum, unheard of among other peoples, that until a certain king should be declared, the state should neither be without a king nor be one for long under a single king, nor should it be allowed that anyone, by long-rooted power, should grow either too slow to lay down the imperium or too well-founded to keep it.

\ciceroSection{24} ``At which time that new people none the less saw what escaped Lacedaemonian Lycurgus, who held that the king was not to be chosen --- if this could be in Lycurgus's power --- but was to be had, of whatever kind he should be, provided only he was begotten of Hercules's stock. Our own people --- even then rustic --- saw that virtue and royal wisdom, not lineage, were what must be sought.

\ciceroSection{25} ``Since rumour reported that Numa Pompilius excelled in these, the people themselves, passing over their own citizens, took to themselves on the senators' urging a foreign-born king, and called this Sabine man from Cures to reign at Rome. Who, when he had come hither --- although the people had ordered him by curiate comitia to be king --- yet himself proposed a curiate law about his own command; and seeing the Roman men kindled by Romulus's establishment in their warlike pursuits, judged that they should be called back somewhat from that habit.

\ciceroSection{26} ``And first he divided the fields which Romulus had taken in war, man by man, among the citizens, and showed that without ravaging and plunder they could, by tilling the fields, abound in all conveniences, and instilled in them a love of leisure and peace, by which justice and faith most easily strengthen, and by whose protection the cultivation of fields and the gathering of crops are most defended. The same Pompilius added two augurs to the original number, with greater auspices having been found, and put five priests, drawn from the leading men, in charge of the rites; and the spirits of men, hot with the habit and desire of war, he softened with the ceremonies of the religions, by laws set forth such as we have in the records; and he added, besides, the flamines, the Salii, the Vestal virgins, and set up every part of religion most solemnly.

\ciceroSection{27} ``He wished the diligence of the rites themselves to be hard, the apparatus very easy. For he laid down many things to be thoroughly learned and observed, but those things without expense. So in cultivating the religions he added work, removed cost; and he also invented markets, games, and all the causes of meeting and the festal days. By the institution of which he called back to humanity and gentleness the spirits of men by then become wild and savage in the pursuits of war. Thus, when he had reigned thirty-nine years in the highest peace and concord (for let us follow above all our Polybius, than whom no one was more careful in working out times), he passed from life --- having confirmed two most splendid things for the long endurance of the commonwealth: religion and clemency.''

\ciceroSection{28} When Scipio had said this, ``Is this truly handed down to memory, Africanus,'' said Manilius, ``that this King Numa was a disciple of Pythagoras himself, or at least a Pythagorean? For we have often heard this from our elders, and so understand it commonly thought; nor do we see it sufficiently shown by the authority of the public annals.'' Then Scipio: ``It is false, Manilius,'' he said, ``the whole of it, and not only made up, but ignorantly and absurdly made up. Things in falsehood are not to be borne when we see that they are not only fictitious but cannot even have happened. For Pythagoras is found to have come to Sybaris and Croton and into those parts of Italy in the fourth year of Lucius Tarquinius Superbus's reign; for the same sixty-second Olympiad marks both the beginning of Superbus's reign and Pythagoras's coming.

\ciceroSection{29} ``From which it can be understood --- when the years of the kings have been counted up --- that about a hundred and forty years after Numa's death Pythagoras first set foot in Italy; nor has this been doubted among those who have most carefully traced the annals of times.'' ``Immortal gods,'' said Manilius, ``how great and how old is this error of men! And yet I take it gladly that we have been educated not by overseas or imported arts, but by native and homely virtues.''

\ciceroSection{30} ``But you will know this much more easily,'' said Africanus, ``if you see the commonwealth advancing, and coming by some natural road and course into the best constitution; nay, you will set down this very thing as praiseworthy in the wisdom of our forefathers --- that you will understand that many things, even taken from elsewhere, became much better with us than they had been there whence they were taken hither, and where they had first arisen; and you will understand that the Roman people was made firm not by chance, but by counsel and discipline --- nor yet with fortune turning against it.

\ciceroSection{31} ``With King Pompilius dead, the people, on the interrex's proposal, by the curiate comitia made Tullus Hostilius king; and he, after the example of Pompilius, consulted the people in the curies on his command. His glory in military matters was outstanding, and great were his deeds in war. He both made and hedged about, from his spoils, the Comitium and the Curia. He set up the law by which wars should be declared --- which, most justly invented by him, he sanctioned with the religion of the fetiales, that every war which had not been announced and declared should be judged unjust and impious. And, that you may turn your minds to how wisely our kings already saw that something must be granted to the people (for many things on that subject must be said by us), Tullus did not even dare use the kingly insignia except by command of the people. For that he might have twelve lictors with the rods marching before him \ldots

\ciceroSection{33} \ldots for the commonwealth does not creep but flies into the best state, your discourse so set up. After him Ancus Marcius, grandson of Numa Pompilius by his daughter, was set up by the people as king, and likewise carried a curiate law about his command. He, having conquered the Latins in war, took them into the citizenship; he added the Aventine and the Caelian hill to the city; he divided the lands he had captured, made public all the maritime forests he had taken, and founded a city at the mouth of the Tiber, strengthening it with colonists. And so, after he had reigned twenty-three years, he died.'' Then Laelius: ``This king too is to be praised. But Roman history is obscure, since we have his mother but do not know his father.'' ``So it is,'' he said. ``But of those times only the names of the kings are illuminated.''

\ciceroSection{34} ``At this point first the state seems to have been made more learned by a certain grafted-on discipline. For there flowed into this city no thin streamlet from Greece, but a most abundant river of those disciplines and arts. They say there was a certain Demaratus of Corinth, easily by honour, authority, and fortune the leading man of his own state, who, when he could not bear Cypselus, the Corinthian tyrant, fled, it is said, with much money, and betook himself to Tarquinii, the most flourishing city of Etruria. When he heard that Cypselus's domination was being confirmed, this free and brave man left his country and was received as a citizen by the men of Tarquinii, and there set down a home and a seat. There, when he had begotten two sons of a Tarquinian matron, he educated them in all arts after the discipline of the Greeks \ldots

\ciceroSection{35} \ldots [his son] was easily received into the citizenship; on account of his cultivation and learning, he became intimate with King Ancus, even to such a degree that he was thought to share in all his counsels and to be almost a partner in the kingdom. There was in him besides the highest courtesy, the greatest generosity in giving help, aid, defence, even bounty to all citizens. So when Marcius died, by all the people's votes Lucius Tarquinius was made king. For so he had bent his name from a Greek name, that in every kind of thing he might seem to have imitated the custom of this people. And, when he had carried a law on his command, he first doubled that original number of the senators, and called the old senators those of the major clans (whom he asked first for their opinion), those taken in by himself those of the minor.

\ciceroSection{36} ``Then he set up the cavalry on the form which is still kept to this day; he could not change the names of the Titienses, the Ramnenses, and the Luceres --- though he wished to --- because the augur Attus Navius, of supreme glory, was not for it. I see also that the Corinthians were once careful in assigning public horses and feeding them out of taxes on orphans and widows. But still, with later units of horsemen added to the earlier, he made eighteen hundred horsemen, and doubled the number. Afterwards in war he subdued the Aequians, a great and fierce people pressing upon the affairs of the Roman people; and likewise he routed the Sabines, when he had repulsed them from the walls of the city, with the cavalry, and conquered them in war. We have also received that he was the first to hold the great games which were called the Roman; and that in the Sabine war, in the very battle, he vowed to make a temple on the Capitoline to Jupiter Optimus Maximus; and that he died, after reigning thirty-eight years.''

\ciceroSection{37} Then Laelius: ``Now Cato's saying becomes more sure --- that the constitution of the commonwealth is the work neither of one time nor of one man; for it is plain how great an addition of good and useful things comes to be made in each single king. But there follows that king who, of all those in our commonwealth, seems to me to have seen the most.'' ``So it is,'' said Scipio. ``For after him Servius Tullius is reported to have been the first to reign without the people's command. They say he was born of a Tarquinian slave-woman, having been conceived from one of the king's clients. Brought up among the slaves, he attended at the king's banquets; the spark of genius did not lie hidden, which even then shone out in the boy: such was his cleverness in every duty and speech. So Tarquinius, who at that time had very small children, so loved Servius that he was commonly held to be his son; and with the highest zeal he educated him in all those arts which he himself had learned, after the most cultivated custom of the Greeks.

\ciceroSection{38} ``But when Tarquinius had perished by the plot of Ancus's sons, and Servius, as I said before, had begun to reign, not by command but by the will and consent of the citizens (because, when Tarquinius had been falsely said to be sick from his wound and yet alive, Servius had pronounced judgement in royal robes, freed debtors with his own money, and, by the use of the greatest courtesy, had made it credible that he gave judgement at Tarquinius's command), he did not commit himself to the senators; but, with Tarquinius buried, he himself consulted the people about himself, and being ordered to reign carried a curiate law about his command. And first he avenged in war the wrongs of the Etruscans; out of which when \ldots

\ciceroSection{39} \ldots eighteen of the highest assessment. Then, with a great number of cavalrymen separated out from the whole sum of the people, he distributed the rest of the people into five classes, and divided the elders from the younger; and so disposed them that the votes were not in the power of the multitude but of the wealthy. He took care --- which must always be held to in the commonwealth --- that the most should not have the most weight. Were this distribution unknown to you, it would be set out by me; now, you see the principle is such that the centuries of horsemen, with their six votes and the first class with the added century which is given to the most useful service of the city, the carpenters --- has eighty-nine centuries; to which, if from the hundred and four centuries (for so many remain) only eight are added, the strength of the whole people is made up; and the much greater rest of the multitude --- ninety-six centuries --- would neither be excluded from the votes (lest it be insolent), nor have weight too great (lest it be dangerous).

\ciceroSection{40} ``In which matter he was diligent even in words and the very names. For when he had called the wealthy \textit{assidui} from \textit{ass-giving}, he named those who had brought either no more than fifteen hundred asses, or nothing at all into their assessment except a head, \textit{proletarii} --- so that from them, as it were, offspring (\textit{proles}), that is, as it were, a progeny of the state, would seem to be expected. Of those ninety-six centuries, in one century at that time more were rated than in almost the whole first class. So no one was kept from the right of voting; and he had most weight in voting whose interest it was most that the state be in the best estate. Indeed, the \textit{accensi velati}, the trumpeters, the horn-players, the proletarii \ldots

\ciceroSection{41} \textit{[Nonius p. 342M:]} The commonwealth is set up in the best estate when, mixed in measure out of those three kinds --- kingly, optimate, popular --- it does not, by punishing, stir up the savage and wild spirit.

\ciceroSection{42} ``[Rome was] sixty-five years more ancient [than the Lacedaemonians' constitution], for it had been founded thirty-nine years before the first Olympiad. And that most ancient Lycurgus saw almost the same things. So this evenness, and this triple kind of commonwealths, seems to me to have been common to us with those peoples. But that which is proper to our commonwealth, than which nothing can be more excellent, this I shall trace out, if I can, with more refinement; which will be of such a kind that nothing of the sort is to be found in any other commonwealth. For these things, which I have so far set out, were so mixed in this state and among the Lacedaemonians and the Carthaginians, that they were in no way tempered.

\ciceroSection{43} ``For in a commonwealth in which there is some one man with perpetual power, especially royal --- though there be also a senate, as at Rome when there were kings, or as at Sparta under Lycurgus's laws, and even though there be some right of the people, as there was under our kings --- the kingly name still stands out: nor can a commonwealth of that kind not be a kingdom and be called one. But that form of state is most changeable, for this reason: that, dashed by the fault of one, it most easily falls into the most ruinous part. The kingly kind of state in itself, however, is not only not to be condemned, but, I do not know whether, far to be set above the other simple forms --- if I were to approve any simple kind of commonwealth: but only so long as it keeps its own constitution. That constitution, however, is this: that by the perpetual power and justice and total wisdom of one man, the safety, evenness, and quiet of the citizens are governed. There are altogether wanting to that people which is under a king many things, and first of all liberty --- which lies not in this, that we have a just master, but in this, that we have none.

\ciceroSection{44} \ldots they bore. For some time too success in business attended even that unjust and harsh master. For he both conquered all Latium in war and took Suessa Pometia, a wealthy and well-stocked city; and, enriched by a great booty of gold and silver, he discharged his father's vow by building the Capitol; he led colonies; and after the institutions of those from whom he was sprung, he sent splendid gifts, like libations of the spoils, to Apollo at Delphi.

\ciceroSection{45} ``Here that orbit will now be turned, whose natural motion and round you must learn to recognize from its beginning. For this is the head of civic prudence, in which all this our discourse is busy: to see the courses and bendings of commonwealths, that, when you know whither each thing inclines, you may either keep them back or run to meet them in advance. For that king of whom I speak, first stained by the slaying of the best of kings, was not whole in mind; and, while he himself feared the highest punishment of his own crime, he wished himself to be feared. Then, propped up by victories and riches, he exulted in insolence, and could neither rule his own morals nor the lusts of his own.

\ciceroSection{46} ``Therefore, when his elder son had done violence to Lucretia, daughter of Tricipitinus, wife of Collatinus, and the modest and noble woman had punished herself by death for that wrong, then a man outstanding in genius and virtue, Lucius Brutus, drove from his fellow citizens the unjust yoke of harsh servitude. Though a private man, he upheld the whole commonwealth, and was the first to teach in this state that, in preserving the citizens' liberty, no one is a private man. With him as authority and leader, the state was stirred up; and by this fresh complaint of Lucretia's father and kinsmen, and by recollection of the pride of Tarquin and the many wrongs done both by him and his sons, it ordered that the king himself, his children, and the whole Tarquin clan be in exile.

\ciceroSection{47} ``Do you see, then, how out of a king there came a master, and by the fault of one the kind of commonwealth was changed from the best to the worst? For this is the master of the people, whom the Greeks call \textit{tyrannus}; for they wish that man to be \textit{rex} who, like a parent, takes thought for the people and preserves those over whom he is set in the best possible condition of living --- a kind of commonwealth indeed good, as I said, but yet inclined and as it were leaning toward the most ruinous state.

\ciceroSection{48} ``For as soon as this king has bent himself toward an unjust dominion, he becomes at once a tyrant: than which no animal more foul, more loathsome, more hateful both to gods and men can be thought of. He, although in figure a man, yet by the savagery of his morals surpasses the most monstrous beasts. For who could rightly call \textit{man} him who would have no sharing of right with his fellow citizens, no fellowship of humanity with the whole race of men? But there will be another fitter place for us to speak of this kind, when the matter itself shall remind us, that we may speak of those who, even when the state was already freed, sought a tyranny.

\ciceroSection{49} ``So you have the first rise of the tyrant; for the Greeks wished this name to belong to an unjust king. Our own people called all those kings who alone had perpetual power over peoples. Hence both Spurius Cassius and Marcus Manlius and Spurius Maelius are said to have wished to seize a kingdom, and lately Tiberius Gracchus \ldots

\ciceroSection{50} ``Lycurgus called those at Lacedaemon \textit{gerontes} --- too few indeed, twenty-eight --- in whose hands he wished the chief counsel to be, while the king held the chief command. From whom our forefathers, following and interpreting the same thing, named those whom he called the \textit{senes} the senate, as we said Romulus had already done with the chosen senators (\textit{patres}). Yet there the kingly force, power, and name stand out and are eminent. You allot some power to the people, as both Lycurgus and Romulus did; you do not satisfy them with liberty, but inflame them with the desire of liberty, when you have only given them the power of tasting it; and there will always hang over you the fear that the king (which often happens) may turn out to be unjust. Therefore that fortune of a people is fragile which is set in the will or morals of one man, as I said before.

\ciceroSection{51} ``Wherefore let this be the first form, and image, and origin of a tyrant, found by us in that commonwealth which Romulus founded with auspices, not in that one which (as Plato has set out) Socrates himself depicted in that tripartite discourse: how, like Tarquinius, having gained no new power but using unjustly that which he had, he overturned this whole kind of regal state. Let there be set against him another, good and wise and skilled in the advantage and dignity of the citizens, as it were the guardian and steward of the commonwealth: for so let him be called, whoever shall be ruler and steersman of the state. Make sure you recognize that man; for he is the one who can guard the state by counsel and effort. Since this name is up to now less worn in our speech, and the kind of this man will more often have to be handled in the rest of our discourse \ldots

\ciceroSection{52} \ldots [Plato sought] foundations, and arrived at a state to be hoped for rather than wished for, as small as he could --- not what could exist, but in which the system of civic affairs could be looked into. But I, if I can in any way attain it, by the same reasonings which he saw, will strive --- not in the shadow and image of a state, but in the most extensive commonwealth --- to seem to touch each cause of public good and ill as if with a wand. For when those two hundred and forty-odd kingly years had passed, with the interregnums included, and Tarquin had been driven out, the Roman people held so great a hatred of the kingly name as it had held longing for it after the death --- or rather the departure --- of Romulus. Therefore, as it had then been unable to do without a king, so, with Tarquin driven out, it could not bear to hear the name of king. Here, when \ldots

\ciceroSection{53fr} \textit{[Nonius p. 526M:]} ``Therefore that splendid constitution of Romulus, when it had remained firm for almost two hundred and twenty years \ldots''

\ciceroSection{53} \ldots that whole law was annulled. With this in mind our forefathers then both drove out the innocent Collatinus on the suspicion of his kinship, and the rest of the Tarquinii on the offence of the name; and with the same mind Publius Valerius first ordered the rods to be lowered, when he had begun to speak in the assembly, and brought down his house under the Velia, after he saw the people's suspicion stirred because he had begun to build on a higher place of the Velia, in that very spot where King Tullus had lived. The same man --- in which Publicola most especially showed his name --- carried the law to the people, the first carried in the centuriate comitia: that no magistrate should put a Roman citizen to death or scourge him against his appeal.

\ciceroSection{54} ``That appeal lay even from kings is shown by the books of the pontifices, and ours of the augurs make the same plain; and that one might appeal from every judgement and penalty the Twelve Tables show in many laws. And what has been handed down to memory --- that the decemvirs who wrote the laws were created without right of appeal --- shows clearly enough that other magistrates were not without appeal; and the consular law of Lucius Valerius Potitus and Marcus Horatius Barbatus, men wisely friendly to the people for the sake of concord, sanctioned that no magistrate should be created without right of appeal. Nor did the Porcian laws --- which are three, of three Porcii, as you know --- bring anything new beyond a sanction.

\ciceroSection{55} ``So Publicola, after that law on appeal had been carried, immediately ordered the axes to be taken from the rods; and on the next day he had Spurius Lucretius substituted to himself as colleague, and ordered his own lictors to pass over to him, because he was older; and he was the first to lay it down that, for each consul in turn, the lictors should march before him in alternate months, that there should not be more insignia of command in a free people than there had been under a kingdom. This man, as I at least understand, was no middling man, who, by giving the people moderate liberty, more easily kept the authority of the leading men. Nor do I now without cause sing to you these so old and so worn things, but in distinguished persons and times I lay down examples of men and matters, to which the rest of my speech may be directed.

\ciceroSection{56} ``In this state, then, the senate held the commonwealth in those times, that in a free people few things were done by the people, most by the authority and ordinance and custom of the senate; and that the consuls held a power that was indeed yearly in time, but kingly in kind itself and in right. And what was perhaps the chief means of keeping power among the nobility was strenuously kept up: that the people's comitia were not valid unless the authority of the senators had approved them. And in these very times, almost ten years after the first consuls, a dictator was instituted --- Titus Larcius --- and that new kind of command seemed near to the likeness of a kingdom. But still all things were held with the highest authority by the leading men, while the people gave way; and great matters in those times were carried on in war by the bravest men, equipped with supreme command, dictators and consuls.

\ciceroSection{57} ``But what the very nature of things compelled --- that the people, freed from kings, should arrogate to itself a little more right --- it attained, no long interval after, in about the sixteenth year, in the consulship of Postumius Cominius and Spurius Cassius. In which there was wanting, perhaps, system; but the very nature of commonwealths often overcomes system. For hold to this --- which I said at the beginning --- that, unless there be in the state an even balance of right and duty and office, so that there is enough power in the magistrates, enough authority in the council of leading men, and enough liberty in the people, this unchanging condition of the commonwealth cannot be preserved.

\ciceroSection{58} ``For when the state was disturbed by debt, the plebs took possession first of the Sacred Mount, then of the Aventine. Not even the discipline of Lycurgus held those reins among the Greeks; for at Sparta also, in the reign of Theopompus, there are likewise five whom they call ephors; in Crete ten, who are called \textit{cosmoe} --- as the tribunes of the plebs were set against the consular command, so they were set against the kingly force.

\ciceroSection{59} ``There had been perhaps some way for our forefathers of remedying that debt, which had escaped neither the Athenian Solon not long before, nor afterwards by no long time our own senate; when, on account of one man's lust, all bonds of citizens were freed, and afterwards the binding ceased; and always for this kind, when the plebs, weakened by the public disaster of expenditures, was failing, some relief and remedy has been sought for the safety of all. With which counsel then passed by, the cause arose for the people, with two tribunes of the plebs created in the sedition, that the power and authority of the senate should be lessened. Yet that authority remained heavy and great, because the wisest and bravest men, guarding the state with arms and counsel, flourished above all in authority --- because, while they far surpassed the rest in office, they were inferior in pleasures and not generally superior in money; and so the virtue of each was the more welcome in public matters, because, in private matters, they most diligently watched over each citizen by service, counsel, and substance.

\ciceroSection{60} ``In which condition of the commonwealth, when Spurius Cassius, flourishing in the highest favour with the people, was working to seize a kingdom, the quaestor accused him; and as you have heard, when his father said he had found him to be in that fault, with the people giving way, he had him put to death. And in about the fifty-fourth year after the first consuls, the consuls Spurius Tarpeius and Aulus Aternius carried that grateful law on a fine and bond in the centuriate comitia. Twenty years later, after the censors Lucius Papirius and Publius Pinarius, by demanding many fines, had turned the substance of cattle from private men to the public, a light valuation of cattle in the law on fines was established by the consuls Gaius Iulius and Publius Papirius.

\ciceroSection{61} ``But some years before, when the highest authority lay in the senate and the people were patient and obedient, a method was devised: that both the consuls and the tribunes of the plebs should resign their magistracy, and that ten men should be created with supreme power and without right of appeal, who should both hold the chief command and write the laws. When they had written ten tables, with the highest fairness and prudence in the laws, they substituted, for the next year, other decemvirs, whose faith and justice were not similarly praised. Yet from that college that singular praise belongs to Gaius Iulius, who, when he himself held supreme power because he was a decemvir without right of appeal, yet demanded sureties of a noble man, Lucius Sestius, in whose bedroom he said he had seen a dead body dug up, in his own presence: because he denied he would neglect that famous law which forbade the trial of a Roman citizen on a capital charge except by the centuriate comitia.

\ciceroSection{62} ``A third year of the decemvirs followed, when the same men were [in office] and would not have others substituted. In this condition of the commonwealth --- which I have already often said could not last long, because it was not equal in all orders of the state --- the whole commonwealth was in the hands of the leading men, with the most noble decemvirs presiding, no tribunes of the plebs set against them, no other magistrates joined to them, no right of appeal to the people left against death and stripes.

\ciceroSection{63} ``Therefore, out of the injustice of these men, suddenly arose the greatest disturbance and a total change of the commonwealth. They had added two tables of unjust laws, in which they sanctioned with most inhuman law that even those marriages which are wont to be granted to peoples disjoined from us should not exist between plebeians and patricians (this was afterwards annulled by Canuleius's plebiscite); and they presided over the people in every command lustfully, harshly, and greedily. Famous, of course, is that affair, celebrated in many records of letters: when one Decimus Verginius slew his maiden daughter with his own hand in the Forum on account of the lust of one of those decemvirs, and, mourning, fled to the army (which was then on Algidus); and how the soldiers left the war they had in hand, and first occupied, in arms, the Sacred Mount (as had been done before in like cause), and afterwards the Aventine \ldots

\ciceroSection{63fr} \textit{[Philargyrius, Servius on Georgics 3.125:]} \ldots Lucius Quinctius being named dictator. ``I judge that our forefathers both most approved [this institution] and most wisely retained it.''

\ciceroSection{64} When Scipio had said this, and the rest of his discourse was awaited in the silence of all, then Tubero: ``Since these elders of ours ask nothing of you, Africanus, you shall hear from me what I find wanting in your speech.'' ``Indeed,'' said Scipio, ``and gladly.'' Then he: ``You seem to me to have praised our commonwealth, when Laelius had asked you not about ours, but about every commonwealth. Yet I have not learned from your speech by what discipline, what morals, what laws we may set up or preserve that very commonwealth which you praise.''

\ciceroSection{65} Here Africanus: ``I think there will soon be a fitter place for us, Tubero, to discuss the founding and preserving of states; on the best constitution, however, I thought I had answered enough to what Laelius had asked. For first I had distinguished by number three approved kinds of states, and three ruinous opposed to those three; that none of these by itself was the best, but that there stood out above the single ones that which was moderately tempered out of the first three.

\ciceroSection{66} ``As for my having used the example of our own state, this had its force not toward defining the best constitution (for that could be done without an example), but that, in our greatest of states, the matter described by reasoning and speech might be discerned in fact. But if you ask, apart from the example of any people, the very kind of the best constitution, we must use the image of nature, since you \ldots [will not be content with] this image of city and people \ldots

\ciceroSection{67} \ldots whom I have long sought and to whom I am eager to come. Are you perhaps asking for the prudent man?'' Then he: ``That very one.'' ``You have, of those very men present, a fine supply --- or to begin from yourself.'' Then Scipio: ``And would that the supply were proportionate from the whole senate! But still, that man is prudent who, as we have often seen in Africa, sitting on the savage and vast beast, restrains and rules the beast wherever he wishes, and with a slight prompting, not blow, bends that wild creature. I have known him, and when I was your legate I often saw him. So that Indian or Carthaginian restrains one beast --- and that one teachable and accustomed to human ways. But that thing which lies hidden in the minds of men, that part of the mind which is called \textit{mens}, must rein in and tame not one nor easy creature for subduing, when ever it can do this --- which it very rarely can. For also that fierce thing must be held in \ldots

\ciceroSection{68} \ldots \textit{[Nonius:]} which is fed with blood, which exults in every cruelty so much that it is scarcely glutted with the bitter funerals of men. \textit{[Nonius:]} \ldots and to the desirous, lustful, dissolute man wallowing in pleasures. \textit{[Nonius:]} and the fourth, anxiety, prone to mourning and grieving, and always troubling itself. \textit{[Nonius:]} that they are afflicted with anguish, with misery, or cast down by timidity and cowardice. \textit{[Nonius:]} as an unskilled charioteer is dragged from the chariot, trodden on, torn, dashed in pieces.

\ciceroSection{69} \ldots could be said. Then Laelius: ``I see now the man whom I was awaiting --- whom you would set up for office and duty.'' ``To this one, plainly,'' said Africanus, ``almost alone (for in this one are nearly all the rest): that he should never depart from looking at and contemplating his own self, that he should call others to imitation of himself, that by the splendour of his mind and life he should offer himself as a mirror to the citizens. For just as in lyres or pipes, and in the very song and voices, a certain harmony must be kept out of distinct sounds, which when changed or discordant the cultivated ear cannot bear; and that harmony, by the modulation of the most unlike voices, is yet made concordant and agreeing --- so out of the highest and the lowest and the middle orders, set between like sounds, the state, by the moderate ratio, sounds together by the consent \textit{[as Augustine quotes:]} of the most unlike. And what is called by musicians \textit{harmony} in song, is in the state \textit{concord} --- the closest and best bond of safety in every commonwealth; and this can in no way exist without justice.''

\ciceroSection{70} \ldots [the commonwealth must be] full of justice. Then Scipio: ``I assent indeed, and I declare to you that we should suppose nothing said as yet about the commonwealth, nor anywhere we can go further, unless it is established not only that this is false, that without injustice the state cannot be carried on, but also that this is most true, that without the highest justice the state can in no way be carried on. But, if you please, this far for today; let us put off the rest --- enough remains --- till tomorrow.'' This being agreed, an end of debate was made for that day.

\ciceroBook{3}

\ciceroSection{1} \textit{[Lacuna at the head of Book 3.]}

\ciceroSection{3} \ldots and [a remedy] for the slowness of carriages; and the same [reason], when it had received men sounding from disordered voices something inchoate and confused, cut these voices and divided them into parts, and as it were stamped certain signs --- words --- upon things, and bound up by the most pleasant bond of speech men who before had been disjoined. By a like genius of voice, the sounds, which seemed boundless, were all marked and expressed by the discovery of a few letters, by which conversations with the absent, the signs of wills, and the records of past matters might be kept. Number was added thereto, a thing both necessary to life and alone unchanging and eternal; which first impelled us also to look up at the heaven and not in vain to behold the motions of the stars, and by the numberings of nights and of days \ldots

\ciceroSection{4} \ldots whose minds raised themselves higher and could either bring to pass or think out something worthy of the gift, as I said before, of the gods. Therefore let those who discuss the system of living be for us great men, as they are; let them be learned, let them be teachers of truth and virtue, only provided that there be this matter, whether discovered by men engaged in the variety of public affairs, or even handled in their leisure and in letters, since it is the case --- by no means to be despised --- the civic system and the discipline of peoples, which brings forth in good geniuses (which thing it has very often brought forth) that there should arise a certain incredible and divine virtue.

\ciceroSection{5} ``But if anyone has thought that to those instruments of the mind which one has from nature and from civil institutions, doctrine and a richer knowledge of things must be added by him --- as those very men who appear in the discussion of these books --- there is no one who would not be bound to set them above all others. For what can be more brilliant than when the handling and use of great matters is joined with the studies and knowledge of those arts? Or what more perfect can be thought of than Publius Scipio, than Gaius Laelius, than Lucius Philus? --- who, that they might pass over nothing pertaining to the highest praise of distinguished men, added to the homely custom of our forefathers this also, the doctrine come from Socrates.''

\ciceroSection{6} ``Therefore the man who has both wished and been able to do both --- that is, to equip himself with the institutions of the forefathers and with doctrine --- I think has attained everything that belongs to praise. But if one of two ways of prudence is to be chosen, then, even if to some that quiet mode of life among the best studies and arts will seem the more blessed, this civic mode is certainly the more praiseworthy and brilliant; out of which life such great men are adorned, as Manius Curius, ``whom no man could surpass with iron or with gold,'' or \ldots

\ciceroSection{7} \ldots [there will have] been wisdom; yet this difference there was between the system of either kind: that those nourished by words and arts the beginnings of nature, these by institutions and laws. But more such men this one state has brought forth --- if not wise men (since they hold that name so strictly), at any rate men worthy of the highest praise, since they cultivated the precepts and discoveries of the wise. And likewise: how many states are or were praiseworthy --- since this is by far the greatest counsel in the nature of things, to set up that commonwealth which can be lasting --- if we count single men in single states, what a multitude of excellent men is then to be found! But if either Latium of Italy, or Sabine or Volscian people of the same, if Samnium, if Etruria, if that great Greece we should wish to range over with our minds; if then the Assyrians, if the Persians, if the Carthaginians, if these \ldots

\ciceroSection{8} \ldots [Laelius and the others to Philus, asking him to take up the case for injustice]. And Philus: ``A splendid case indeed you bring to me, when you wish me to take up the patronage of wickedness.'' ``And yet,'' said Laelius, ``you must indeed be wary of this --- if you say what is wont to be said against justice --- that you should not seem also to think it; for you yourself are, as it were, the unique example of ancient probity and faith, and your habit of arguing for opposite sides is not unknown, because you think that thus the truth is most easily found.'' And Philus: ``Well then, I shall yield to you, and shall knowingly soil myself; for, since those who seek for gold do not think they should refuse this, we, when we seek justice --- a thing far dearer than all gold --- ought certainly to refuse no trouble. And would that, just as I am about to use the speech of another, so I might be allowed to use another's mouth! Now those things must be said by Lucius Furius Philus which Carneades, a Greek man and accustomed to whatever is convenient, [said] in words \ldots

\ciceroSection{9} \textit{[Nonius:]} \ldots that you may answer Carneades, who is wont often to make sport of the best causes by the sophistry of his wit.

\ciceroSection{11} \textit{[Nonius:]} Justice looks outward, and is wholly thrust out and stands forth. \textit{[Nonius:]} A virtue which beyond the rest stretches and unfolds itself entirely toward the advantages of others.

\ciceroSection{12} \ldots [the one] would both find and protect; the other filled four sizable books on justice itself. From Chrysippus I expected nothing great or magnificent; he speaks in his own way, so as to weigh all things by the weight of words, not by the weight of things. It was for those heroes to summon up that virtue --- which is the one most bountiful and liberal, if it exists at all, the one that loves all rather than itself, born for others rather than for itself --- and to set it, as it lay, on that divine seat not far from wisdom.

\ciceroSection{13} ``Nor was either the will or the genius wanting in those men --- for what other cause of writing or counsel could be theirs? --- but their cause overcame even their will and their abundance. For the right concerning which we are inquiring is something civil, none natural; for if there were [a natural one], then, as hot and cold things, as bitter and sweet things, so just and unjust would be the same for all.

\ciceroSection{14} ``But now, if anyone --- like that man in Pacuvius ``borne in a chariot of winged serpents'' --- could look down upon and survey with his eyes many and varied peoples and cities, he would see, first, in that most uncorrupted people of the Egyptians, who hold the memory of so many ages and events in their writings, that a certain ox is held a god (which the Egyptians call Apis), and many other portents are with them, and beasts of every kind consecrated to the number of the gods. Then in Greece (just as with us) splendid shrines consecrated with human likenesses --- which the Persians thought wicked. And for that one cause Xerxes is said to have ordered the temples of the Athenians to be set on fire, because he thought it impious that the gods, whose home was this whole world, should be confined and shut up by walls.

\ciceroSection{15} ``Afterwards, against the Persians both Philip, who planned, and Alexander, who waged, brought this cause of war: that he wished to punish the temples of Greece. These the Greeks did not even think to be repaired, that there might be before the eyes of posterity an everlasting witness of the Persians' crime. How many --- as the Tauri at the Black Sea, as Busiris king of Egypt, as the Gauls, as the Carthaginians --- have thought human sacrifice both pious and most pleasing to the immortal gods! The institutions of life are so different that the Cretans and the Aetolians think brigandage honourable; the Lacedaemonians used to declare all those fields theirs which they could touch with a spear. The Athenians used even publicly to swear that all the land was theirs which bore olive or grain. The Gauls think it disgraceful to seek grain by hand, and so go armed and reap others' fields.

\ciceroSection{16} ``We Romans, of course --- the most just of men --- do not allow the peoples beyond the Alps to plant olive or vine, that our olive groves and vineyards may be worth more. When we do this we are said to act prudently, but not to act justly --- so that you may understand that wisdom differs from equity. As for Lycurgus, that founder of the best laws and most equable right, he gave the lands of the wealthy to the plebs to be tilled like slaves.

\ciceroSection{17} ``If I should choose to set out the kinds of right, of institutions, of morals and customs, I should show that they are not only varied among so many peoples, but in one city --- in this very city of ours --- have been changed a thousand times: as our interpreter of right Manilius here pronounces certain rules now on women's legacies and inheritances, while as a young man before the Voconian law was carried he was wont to pronounce others. Which very law, asked for the sake of men's interest, is full of injustice toward women. For why should a woman not have money? Why may a Vestal virgin be heir, and not her own mother? Why, indeed, if a limit of money had to be set for women, could the daughter of Publius Crassus, if she were her father's only child, lawfully have a hundred million asses, while my daughter could not have three million?

\ciceroSection{18} ``[If nature, as I were saying,] had laid down our laws, then all would use the same and the same men would not use one set at one time and another at another. But I ask: if it is the part of a just man and of a good man to obey the laws --- which? Whatever they shall be? But virtue does not admit inconstancy, nor nature suffer variety; and laws are kept by penalty, not by our own justice. Therefore right has nothing natural; from which it follows that there are not even just men by nature. Or do they say that there is variety in laws, while by nature good men follow that justice which truly is, not that which is thought to be? --- for that they say is the part of a good and just man, to give to each what is worthy of him.

\ciceroSection{19} ``What, then, shall we first allot to dumb beasts? For not common men but the greatest and most learned, Pythagoras and Empedocles, declare that there is one condition of right for all living things, and they shout that unappeasable penalties hang over those by whom an animal has been violated. It is, then, a crime to harm a beast, and the man who would so do \ldots

\ciceroSection{23} ``For all those who hold the power of life and death over a people are tyrants; but they prefer to be called kings, by the name of Jupiter Optimus. When certain men, on account of riches or birth or some resources, hold the commonwealth, that is a faction; but they are called optimates. If, indeed, the people has the most power and all things are ruled by its judgement, that is called liberty, but it is licence. But when one fears another --- man fears man, order fears order --- then, because no one trusts himself, a kind of pact is made between the people and the powerful; out of which arises that mixed form of state which Scipio was praising. So the mother of justice is not nature nor will, but weakness. For when of three one must be chosen --- either to do wrong and not suffer it, or both to do and to suffer it, or neither --- the best is to do it, if you can do it without punishment; the second is neither to do nor to suffer; most wretched of all is to be ever fighting it out, now in doing wrongs and now in suffering them. Therefore he who first attains \ldots

\ciceroSection{24fr} \textit{[Nonius:]} ``For when it was asked of him by what crime he was driven to make the sea unsafe with one little galley, he said: ``By the same crime by which you make the world unsafe.'''' \textit{[The pirate's reply to Alexander.]}

\ciceroSection{24} ``\ldots [says wisdom]. \textit{Wisdom} bids us increase resources, multiply riches, push out the boundaries (for whence would there have been that praise carved on the monuments of the supreme commanders, ``He extended the bounds of empire,'' if something had not been added from another's territory?), to rule over as many as possible, to enjoy pleasures, to be powerful, to reign, to be lord. \textit{Justice}, however, instructs us to spare all, to take thought for the human race, to give each his own, to touch nothing of the sacred, the public, the foreign. What, then, comes about if you obey wisdom? Riches, powers, resources, honours, commands, kingdoms --- whether for private men or for peoples. But since we are speaking of the commonwealth, those things are more brilliant which happen publicly; and since the same system of right is in either, I think we must speak of the wisdom of a people. Now to leave aside others --- this very people of ours, whom Africanus traced from its stock in yesterday's discourse, by whose command now the round earth is held: was it by justice or by wisdom that, from the smallest of all, it was made the greatest?

\ciceroSection{25} ``\ldots except the Arcadians and the Athenians, who, fearing, I suppose, this prohibition of justice should ever come about, made up the story that they had sprung from the earth like field-mice from a furrow.

\ciceroSection{26} ``Against these things this is wont to be said first by those who are by no means evil in argument --- and who in this cause have the more authority, because, when a good man is asked about, whom we wish to be open and simple, they are not crafty in arguing, not old foxes, not malicious. They deny that the wise man is therefore good, because goodness and justice delight him by themselves and on their own; but because the life of good men is empty of fear, care, anxiety, danger, while in the wicked there is always some pebble sticking in the soul: always before their eyes are judgements and punishments. There is, however, no profit, no prize won by injustice so great that you should always fear, always think some penalty present, always think some hanging over \ldots

\ciceroSection{27fr} \textit{[Lactantius:]} ``I ask --- if there be two men, one of whom is the best, the most fair, of the highest justice, of singular faith; the other of marked crime and audacity --- and if a state be in such error that it thinks the good man criminal, foul, abominable, and on the contrary holds him who is most wicked to be of the highest probity and faith, and on the strength of this opinion of all citizens that good man be tormented, dragged off, hands at last laid upon him, his eyes gouged out, condemned, bound, burned, exiled, made needy \ldots''

\ciceroSection{27} ``\ldots [until at last] he should also be thought, by the very justest right, of all men the most wretched; and on the contrary that wicked man should be praised, cultivated, loved by all, and all honours, all commands, all resources, all wealth from every side be heaped upon him --- finally that he be judged the best of men in all men's estimation and worthy of every best fortune --- who at last will be so mad as to doubt which he would rather be?

\ciceroSection{28} ``What [is true] in single men, the same is in peoples: there is no state so foolish as not to prefer to rule unjustly than to serve justly. Nor shall I go far afield. As consul I asked --- when you were sitting with me in council --- about the Numantine treaty. Who did not know that Quintus Pompeius had made a treaty, that Mancinus stood in the same case? The one, the best of men, even urged the bill that I, on the senate's resolve, was carrying; the other defended himself most fiercely. If shame is sought, if probity, if faith, Mancinus brought these; if reason, counsel, prudence, Pompeius excels. Which \ldots

\ciceroSection{32} \textit{[Aulus Gellius 1.22.8:]} ``I should not refuse, Laelius, were I not to think that they too wished it, and were I not myself eager that you also should touch some part of this our discussion, especially since yesterday you yourself said that you would have something left for us.'' ``But indeed that cannot be done; that you should not be wanting, we all ask you.'' \textit{[Nonius:]} \ldots ``but he is by no means to be listened to by our youth; for if he meant what he said, he is a foul man; but if otherwise (which I prefer), still his speech was monstrous.''

\ciceroSection{33} \textit{[Lactantius, Divine Institutes 6.8.6:]} ``True law is right reason, agreeing with nature, diffused in all, constant, everlasting, which calls to duty by commanding and deters from fraud by forbidding; which yet bids the upright neither in vain nor forbids them, nor moves the wicked by commanding or forbidding. Against this law it is sin to legislate; nor is it lawful to derogate anything from it; nor can it be wholly abolished; nor can we be released from this law either by the senate or by the people; nor must any other expounder or interpreter of it be sought; nor will there be one law at Rome, another at Athens; one law now, another hereafter; but one and everlasting and unchanging law will hold all nations and at all time, and there will be one common, as it were, master and commander of all --- God: He, the founder of this law, its arbiter, its proposer; whoever shall not obey it will flee himself, and despising the nature of man will by this very fact suffer the greatest penalties, even if he should escape the other punishments which are thought to be.''

\ciceroSection{34fr} \textit{[Augustine, City of God 22.6:]} ``No war is to be undertaken by the best state except either for faith or for safety.''

\ciceroSection{34} ``But from those penalties which even the most foolish feel --- want, exile, chains, scourges --- private men often slip away by the swiftness of death set before them. For states, however, death itself is the penalty, which seems to free single men from a penalty. For the state ought so to be set up that it be eternal. So there is no destruction of a commonwealth that is natural, as of a man, in whom death is not only necessary but very often even to be desired. But when a state is taken away, destroyed, blotted out, it is somehow alike --- to compare small things with great --- as if all this world should perish and fall.

\ciceroSection{35afr} \textit{[Isidore, Origins 18.1.2sq.:]} ``Those wars are unjust which are undertaken without cause; for outside the cause of avenging or driving back enemies no war can be justly waged.'' \textit{[Isidore:]} ``No war is held just unless it be announced, declared, with restoration of property demanded.''

\ciceroSection{35bfr} \textit{[Nonius:]} ``Our people, however, by defending its allies, has now mastered all the lands.''

\ciceroSection{36} \textit{[Augustine, City of God 19.21:]} ``Why, then, does God rule man, the mind the body, reason lust and the other faulty parts of the mind?''

\ciceroSection{37} \textit{[Augustine, Against Julian 4.12.61:]} ``Do we not see that lordship has been given by nature itself to every best man, with the highest advantage of the weak? Why, then, does God rule man, the mind the body, reason lust and anger and the other faulty parts of the same mind?'' \textit{[Augustine:]} ``But the dissimilitudes both of commanding and of serving are to be known. For as the mind is said to command the body, so it is said also to command lust; but the body as a king his citizens or a parent his children, but lust as a master slaves --- because it represses and breaks it. Just so the commands of kings, of generals, of magistrates, of fathers, of peoples preside over citizens and allies as the mind over the body; whereas masters wear out their slaves as the best part of the mind, that is, wisdom, [wears out] the faulty and weak parts of the same mind, the lusts and angers and other passions.'' \textit{[Augustine, City of God 14.23:]} ``so that the children's command [is exercised] over the body's members on account of their ease of obeying; the faulty parts of the mind, however, [are restrained] like slaves with a harsher rule.'' \textit{[Nonius:]} ``For there is a kind of unjust slavery, when those are another's who could be their own; but when they serve [a master, that is just]\ldots''

\ciceroSection{39} \textit{[Priscian:]} ``\ldots in which I assent that anxious and dangerous justice is not the wise man's [concern].''

\ciceroSection{40} \textit{[Lactantius:]} ``Virtue plainly wishes for honour, and there is no other reward of virtue. Yet she takes it easily and demands it not bitterly. What riches will you set before this man? What commands? What kingdoms? --- one who counts these things human, and judges his own goods divine. But if either ungrateful peoples, or many enviers, or powerful enemies, strip virtue of its rewards, then virtue solaces itself with many comforts and most of all sustains itself by its own dignity.'' \textit{[Augustine, City of God 22.4:]} ``Their bodies are not raised to heaven; for nature would not allow that what was of the earth should remain anywhere except on earth.'' \textit{[Nonius:]} ``Never have the bravest men's fortitude, energy, patience [\ldots]'' \textit{[Nonius:]} ``Pyrrhus's bounty, indeed, was not wanting to Fabricius, nor the Samnites' wealth to Curius.'' \textit{[Nonius:]} ``Whose hearth our Cato, when he had gone to him in the Sabine country, used to visit (as we used to hear from himself), at whose seat that man, sitting, had once refused the gifts of the Samnites, then his enemies, now his clients.''

\ciceroSection{41} \ldots [in] Asia Tiberius Gracchus persisted [in his ways] toward our citizens; he despised the rights of the allies and the Latin name and the treaties. If this habit and licence shall begin to spread further and shall translate our command from right to force, so that those who up to now obey us by their will may be held by terror --- although for those of us who are of this age, the watch is well-nigh over, yet I am anxious for our descendants and for that immortality of the commonwealth, which could have been perpetual, were the country to live by its ancestral institutions and morals.''

\ciceroSection{42} When Laelius had said this, although all who were present made known that they were greatly delighted by him, yet beyond the rest Scipio, raised up as if by some joy: ``You have, Laelius, defended many causes so often, that I should not compare with you not only Servius Galba, our colleague, whom you, while he lived, set above all others, but not even any of the Attic orators in his \ldots [textual gap] \textit{[Nonius:]} ``Two things, he said, were lacking in him to keep him from speaking among the people and in the forum: confidence, and a voice.'' \textit{[Scholiast on Juvenal:]} ``With the groan of the men shut up inside, the bull lowed.''

\ciceroSection{43} ``\ldots [Phalaris's bull] to bring back. Therefore who would call that a thing of the people --- that is, a commonwealth --- when, by the cruelty of one, all are crushed, and there is neither one bond of right nor any agreement and association of an assembly, which is what a people is? And the same at Syracuse: that splendid city, which Timaeus calls the greatest of Greek cities, and the most beautiful of all, with a citadel to be seen, harbours flowing into the very bays of the town and to the city's quays, broad streets, colonnades, temples, walls --- not at all the more, while Dionysius held it, was it a commonwealth: for nothing belonged to the people, and the people itself belonged to one. Therefore where there is a tyrant, there must be said --- not (as I said yesterday) a corrupt commonwealth, but, as reason now compels, plainly no commonwealth at all.

\ciceroSection{44} ``You speak excellently,'' said Laelius; ``for I see now whither your speech is going.'' ``Do you see, then, that not even that one which is wholly in the power of a faction can truly be called a commonwealth?'' ``So plainly I judge.'' ``And most rightly you judge. For what was that thing of the Athenians, when, after that great Peloponnesian war, those thirty men presided most unjustly over the city? Did either the city's old glory, or the splendid look of the town, or the theatre, the gymnasia, the colonnades, or the noble propylaea, or the citadel, or the wonderful works of Phidias, or that magnificent Piraeus make it a commonwealth?'' ``Not at all,'' said Laelius, ``since indeed it was no thing of the people.'' ``What? When the decemvirs at Rome were without right of appeal in that third year, when liberty itself had lost the right of claiming security?'' ``It was no thing of the people --- nay, the people itself acted to recover its own thing.''

\ciceroSection{45} ``I come now to that third kind, in which there will perhaps appear straits. When all things are said to be done by the people and to be in the power of the people, when the multitude takes punishment of whomever it wishes, when they drive on, snatch up, hold, dissipate the things they wish --- can you then deny, Laelius, that it is a commonwealth? When all is the people's, since indeed we wish a commonwealth to be the people's affair?'' Then Laelius: ``And nothing more swiftly than that would I deny to be a commonwealth, which is wholly \ldots'' [textual gap] \ldots ``But \ldots [we have agreed] that there was no commonwealth at Syracuse, nor at Agrigentum, nor at Athens, when there were tyrants; nor here, when there were the decemvirs. Nor do I see how the name of commonwealth can the more appear under the lordship of a multitude. For first, a people, as you have most excellently defined, Scipio, is no people unless it is held together by the agreement of right; but that mob is as much a tyrant as if it were one --- the more foul because no beast is more savage than that thing which imitates the appearance and the name of a people. Nor is it fitting that, when by the laws the goods of madmen are in the power of their kinsmen [\ldots]\ldots

\ciceroSection{46} \ldots [the same things] could be said: why is that a commonwealth and an affair of the people, which has been said about a kingdom?'' ``And much more so,'' said Mummius; ``for the likeness of a master falls more on a king, since he is one. But where many good men shall hold the matters of the commonwealth, nothing can be more blessed than it. Yet I would rather have a kingdom than a free people; that third kind of most corrupted commonwealth remains for you.''

\ciceroSection{47} Here Scipio: ``I recognize,'' he said, ``Spurius, that custom of yours, averse from the system of the people; and although it can be borne more mildly than you are wont to bear it, still I assent that there is none of these three kinds which is less to be approved. But on this I do not assent to you, that the optimates are to be preferred to a king; for if it is wisdom that steers the commonwealth, what does it matter whether this be in one or in many? But we are deceived by some error in arguing thus; for when men are called optimates, nothing can seem more excellent --- for what better thing can be thought than the best? But when mention is made of a king, an unjust king too comes to mind. We, however, are saying nothing now of an unjust king, since we are inquiring about the kingly commonwealth itself. Therefore think of Romulus or Pompilius or Tullus as king, and perhaps you will be less displeased with that commonwealth.''

\ciceroSection{48} ``What praise, then, do you leave for a popular commonwealth?'' Then he: ``What? Does the commonwealth of the Rhodians, Spurius, with whom we lately stayed together, seem to you no commonwealth at all?'' ``It seems to me a commonwealth indeed, and by no means to be reproached.'' ``You speak rightly. But, if you remember, all were the same men, sometimes from the plebs, sometimes senators; and they had alternations, in which months they performed the popular service, in which the senatorial. In either they received their pay; both in the theatre and in the curia they judged capital and all other matters; the senate was as powerful, and worth as much as, the multitude \ldots




































