% Laelius de Amicitia --- headnote
% laelius-de-amicitia, 44 BC, Rome

\textit{\textit{Laelius de Amicitia} --- Laelius, On Friendship --- is
the companion piece to \textit{On Old Age}, written in the same year,
44 BC, and dedicated to the same friend, Titus Pomponius Atticus. As he
had put the discourse on old age into the mouth of Cato the Censor,
Cicero now gives the discourse on friendship to Gaius Laelius, surnamed
"the Wise," whose friendship with the younger Scipio Africanus had
become for the Romans the very pattern of the thing. The dramatic date
is 129 BC, a few days after Scipio's sudden and suspicious death; the
setting is Laelius's home, where his two sons-in-law, Gaius Fannius and
Quintus Mucius Scaevola the Augur, come to him and, marveling at the
calm with which he bears the loss, draw him into speaking about
friendship itself. Cicero frames the whole as something he had heard,
as a young man, from Scaevola, who had heard it from Laelius --- a chain
of memory that lets him say, in the preface, that the reader should
forget the author and imagine Laelius himself is speaking.}

\textit{The discourse moves from definition to practice. Friendship,
Laelius holds, is "agreement on all things human and divine, joined
with goodwill and affection," and it can exist only among good men, for
it is born not of weakness or calculated advantage but of nature and of
love --- the \textit{amor} from which \textit{amicitia} takes its name.
Advantage follows friendship but is never its cause; the proof is that
friendship, unlike a bargain, is sought for its own sake and lasts
because nature does not change. From this root Laelius works outward to
the hard questions: the first law, that one must never ask a friend to
do wrong nor do wrong for a friend's sake, not even for those one loves
most, since loyalty to a friend can never excuse harm to the
commonwealth; how friends are to be chosen and tested; how far
indulgence should go; how a friendship that has soured should be
unstitched rather than torn; and, at the close, the menace of flattery,
against which the true friend's candor is the surest guard. He ends as
he began, with Scipio --- commending virtue and the memory of that
friendship to the two young men as the best of all human goods.}

\textit{Like its companion, the dialogue has been read and loved far
beyond the world that made it. Its blend of Stoic principle and Roman
warmth, carried by the voice of a man speaking of a friend he has just
lost, gave the Latin West its classic statement on the subject;
medieval and Renaissance writers on friendship --- Aelred of Rievaulx,
Montaigne in his essay on La Boétie --- wrote in its shadow, and its
formulas passed into common speech. Beneath the doctrine lies the
private occasion of all these last works: Cicero, robbed of the public
life that had been his element and soon to lose his own life as well,
setting down through the mouth of an older and serener Roman his
conviction that, after virtue, nothing the gods have given is better
than a friend.}
