% De Fato --- headnote
% de-fato, 44 BC, Puteoli

\textit{\textit{De Fato} --- On Fate --- belongs to the same astonishing
run of philosophical writing that Cicero produced in 45 and 44 BC, and
forms the third panel of a connected inquiry into religion and the gods,
after \textit{De Natura Deorum} and \textit{De Divinatione}. He composed
it in the spring of 44, in the unsettled months after Caesar's
assassination, and set it as a conversation held at his villa near
Puteoli with Aulus Hirtius, the consul-designate and an old friend, with
whom he says he was spending long days in search of some path back to
peace and concord in the state. The dialogue frame is light: Hirtius
asks to hear something, and Cicero, taking up a single proposition for
examination in the manner of the Academy, holds forth.}

\textit{The question is whether everything that happens is fixed by fate,
and if it is, whether anything is left in our power. Cicero works through
the Stoic case for an unbroken chain of causes --- above all the
arguments of Chrysippus --- and the famous puzzles that cluster around
it: the "lazy argument" \textit{[Greek: argos logos]}, which would have a
man do nothing on the ground that the outcome is settled either way; the
dilemma of Diodorus over what is possible; and the Stoic claim that every
proposition about the future is already true or false. Against the Stoics
he sets the Epicurean answer, the unprovoked swerve of the atom, by which
Epicurus hoped to rescue the freedom of the will --- and finds it no
better, a remedy worse than the disease. Cicero's own conclusion, so far
as the broken text lets us follow it, is that the will is moved by causes
but not compelled by them: some things are in our power, and a man's
assent is his own.}

\textit{The work survives only as a torso. Its opening is lost, so that
the treatise begins in mid-argument, and the end breaks off in
mid-sentence; there are gaps within as well, and a handful of later
quotations preserve scraps of what is missing. Even so, it is the fullest
surviving discussion from antiquity of determinism and responsibility,
and a chief channel through which Stoic and Epicurean thinking on the
subject reached later philosophy. The technical vocabulary Cicero forged
here to carry Greek logic into Latin --- and his refusal to purchase
freedom either at the price of the Stoic chain or by the arbitrary lurch
of an atom --- gave the problem of fate and the will much of the shape it
would keep for centuries.}
