% De Divinatione --- headnote
% de-divinatione, 44 BC, Rome

\textit{\textit{De Divinatione} --- On Divination --- is the central panel
of the connected inquiry into religion that Cicero composed at the end of
his life, standing between \textit{De Natura Deorum}, which asks whether
the gods exist and govern the world, and \textit{De Fato}, which asks
whether all things are bound by an unbroken chain of cause. Its question
follows directly from the first: granted that there are gods who care for
mankind, do they disclose the future, and can men read it? Cicero casts
the work as a conversation held at his Tusculan villa with his brother
Quintus, and divides it cleanly between them --- a debate of two books,
one for and one against, in the manner of the Academy he professed, which
sets the strongest case on each side and leaves the verdict to the
reader.}

\textit{In the first book Quintus puts the case for divination, and puts
it well. He distinguishes its two kinds --- the divination of art, which
reasons from observed signs (the entrails of victims, the flight and cry
of birds, lightning, portents, lots, the stars), and the divination of
nature, which comes unbidden through frenzy and through dreams --- and
heaps up a great store of attested instances, Roman and foreign, from the
oracles of Delphi to the dreams of famous men, before grounding the whole
in the Stoic doctrine of a world bound together by a divine spirit, in
which the soul, kindred to the gods, can foreknow what is to come. In the
second book Cicero himself replies, and takes the structure apart piece by
piece. He asks what subject-matter such an art could possibly have; he
shows that a slaughtered beast's organs cannot be made to carry the fate
of states, that the augur's birds and the astrologer's stars predict
nothing, that the oracles have fallen silent, that no rule can be drawn
from the boundless chaos of dreams; and again and again he sets chance and
conjecture, not the gods, as the true authors of every hit the diviners
boast of.}

\textit{The work was written across the convulsion of the age: begun
under Caesar's dictatorship and finished after the Ides of March, it
carries a sharper political edge than its companions, for Cicero, himself
an augur, knew exactly how the auspices were used and abused in the
service of power. Yet his attack is not on religion. The famous close
draws the distinction on which the whole rests: superstition, which
hounds and enslaves the mind, must be torn out by the roots, while true
religion --- bound up with the knowledge of nature and the rites of the
ancestors --- is to be cherished and enlarged. It is the fullest
treatment of the subject to survive from antiquity, and the most
unsparing: a skeptic's anatomy of belief, written by a Roman priest who
would not give up the gods even as he dismantled the proofs that men
claimed to read their will.}
