% Tusculanae Disputationes --- headnote
% tusculanae-disputationes, 45 BC, Tusculum (dedicated to M. Junius Brutus)

\textit{\textit{Tusculanae Disputationes} --- the Tusculan Disputations
--- belongs to the great philosophical year of 45 BC, written at
Cicero's villa near Tusculum in the months after the death of his
daughter Tullia, when, shut out of the Forum under Caesar's
dictatorship, he turned the whole force of his mind to giving Rome a
philosophy in her own tongue. Dedicated to Marcus Brutus and conceived
as a companion to the \textit{De Finibus} just finished, the work casts
five afternoons of argument as a dialogue of a peculiar kind: a friend
proposes a thesis --- that death is an evil, that pain cannot be borne
--- and Cicero, in the manner of the schools, takes up the contrary and
argues it down. The interlocutors go unnamed, marked in the Latin only
by the turn of the speech; the setting is the colonnade and the walks of
the villa, the hour the slack of the afternoon when the morning's
declaiming is done.}

\textit{The five books pose five questions, each the matter of a single
day. The first asks whether death is an evil, and answers that it is
not, ranging over the nature and the immortality of the soul to free us
of the fear of it. The second asks whether pain is the greatest of
evils, and teaches the endurance that strips it of its power. The third
turns inward, to distress --- \textit{aegritudo}, the grief that gnaws
--- and undertakes its cure; the fourth completes the account with the
rest of the passions, setting out the Stoic anatomy of the soul's
diseases, the genera and species of fear and desire and gladness and
grief, and the discipline by which they are torn out at the root. The
fifth, the book Brutus loved best, defends the hardest and grandest of
the claims of philosophy: that virtue is sufficient of itself for the
happy life, that the good man is happy on the rack and the tyrant
wretched on his throne.}

\textit{Throughout, philosophy is conceived as the medicine of the soul
--- \textit{animi medicina} --- and the work is its dispensary: a course
of consolation, written by a man in fresh mourning, made general and
raised into doctrine. Cicero argues in the manner of the New Academy,
taking up whatever thesis is offered and refuting it without binding
himself to a creed, yet in ethics he leans always toward the Stoic
dignity of the \textit{honestum}, the conviction that the honorable is
the one thing needful. The books are studded with the poetry he loved
--- Ennius and Accius, Pacuvius and the Greek tragedians, often turned
into Latin verse of his own --- and with set pieces that have never lost
their hold: the great apostrophe to philosophy with which the fifth book
rises, \textit{O vitae philosophia dux}; the sword of Damocles hung by a
single hair above the tyrant's feast; and the discovery of the lost and
overgrown tomb of Archimedes, which Cicero, as quaestor in Sicily, found
by its carved sphere and cylinder when the Syracusans themselves had
forgotten where it lay. It is at once the most personal and the most
consoling of his philosophical works --- the record of a mind teaching
itself, in grief, that the happy life lies within its own power.}
