% De Natura Deorum --- headnote
% de-natura-deorum, 45 BC, Rome

\textit{\textit{De Natura Deorum} --- On the Nature of the Gods --- belongs
to the great burst of philosophical writing of 45 BC, the year in which
Cicero, shut out of public life under Caesar's dictatorship and grieving
the death of his daughter Tullia, set himself to give Rome in her own
tongue the whole of Greek thought. Dedicated, like the \textit{De
Finibus}, to Marcus Brutus, it takes up the question on which all the
schools divided and which Cicero calls at once the most beautiful for the
understanding and the most necessary for the ordering of religion: whether
the gods exist, what their nature is, and whether they govern and care for
the world. As befits a man of the New Academy, Cicero does not pronounce
but stages a contest, setting an able advocate of each of the three living
positions to argue his case and be put to the question, and reserving for
himself the part of a near-silent listener.}

\textit{The three books fall to three speakers, met at the house of Gaius
Cotta. In the first, Velleius the Epicurean expounds with great confidence
the theology of his school --- gods of human form, built of the finest
atoms, blessed and deathless, dwelling at ease in the spaces between the
worlds and troubling themselves with nothing --- and then Cotta the
Academic takes that theology apart, exposing the idleness of such gods and
the incoherence of a piety that asks nothing of them. In the second book,
the longest and the richest, Quintus Lucilius Balbus sets out the Stoic
case in four movements: that the gods exist, proved from the heavens and
the consent of all peoples; what their nature is; that the world itself is
a rational and divine being governed by providence; and that this
providence extends its care to humankind. Here belong the famous arguments
from design --- the ordered procession of the stars, the long passage of
the \textit{Aratea} on the constellations, the intricate fitness of plants
and animals and the human frame --- and the rational reverence that Balbus
opposes to mere superstition.}

\textit{In the third book Cotta replies again, now against the Stoics. He
speaks as a Roman pontiff who holds the existence of the gods firmly on the
authority of his ancestors and the rites of the state, yet denies that the
philosophers' arguments establish what tradition already gives; and he
turns the Stoic proofs against themselves, presses the tangled genealogies
of the traditional gods, and raises with real force the problem that has
never lost its sting --- that reason, the supposed gift of a provident
deity, is the very instrument of human crime, and that in this world the
wicked so often flourish while the good are crushed. The book is damaged,
its long attack on providence surviving only in part, and it breaks off
into the famous closing frame: the company parts, and to Cicero himself the
case of Balbus seemed to lean nearer the resemblance of truth. With its
companions \textit{De Divinatione} and \textit{De Fato}, the work is
Cicero's fullest reckoning with religion --- skeptical of the proofs,
unwilling to abandon the gods, and content, as the Academy must be, to set
the strongest cases side by side and leave the reader to weigh where the
truth most nearly lies.}
