% Paradoxa Stoicorum --- headnote
% paradoxa-stoicorum, 46 BC, Rome

\textit{Six of the Stoa's most deliberately counter-intuitive
theses, taken up and made to speak Latin in the open air. The
school had called them \textit{[Greek: paradoxa]} --- marvels
contrary to common opinion --- and had been content to establish
them by the dry, pin-prick dialectic Cicero describes in the
proem: that the morally good is the only good; that virtue is
sufficient of itself for happiness; that all offences and all
right actions are equal; that every fool is mad; that only the
wise man is free and every fool a slave; that only the wise man
is rich. Cicero's experiment, dedicated to Brutus and dashed off
(he says) in the short nights of the year, is to drag these
arguments out of the gymnasium and into the Forum --- to test
whether what the Stoics can barely defend at leisure can be made
plausible to a Roman audience when dressed in the colour of
oratory. He grants at the outset that this is the reverse of
Cato's practice: where Cato spoke Stoically of great themes with
the ornaments of the orator added, Cicero takes the school's
least popular theses and gives them the same treatment as a
game.}

\textit{The result is the lightest piece in the great burst of
46 BC --- the year of the \textit{Brutus} and, soon after, the
\textit{Orator} --- and the most openly rhetorical. The
reflective calm of the proem does not last: as each paradox is
argued, the second person sharpens, and the later essays turn
into something close to invective, with the demagogue of the
fourth paradox (the wrecker of courts and Senate, the captain of
brigandage in the Forum) and the grasping politician of the
sixth standing in plainly for Cicero's enemies of the previous
decade. The text carries the marks of its transmission: a short
lacuna opens the fourth paradox (\textsection 27), and a few
phrases survive only as editorial cruxes; these are left as gaps
rather than filled. What holds the collection together is the
single claim of the proem --- that nothing is so incredible that
speaking cannot make it plausible, nothing so rough that oratory
cannot polish it --- offered here as a demonstration rather than
a doctrine.}
