% De Re Publica --- headnote
% de-re-publica, composed 54-51 BC, published 51 BC, set dramatic date 129 BC

\textit{De Re Publica is Cicero's first major philosophical
treatise, composed in his Tusculan and Cuman villas between
54 and 51 BC and published just before he left for the
Cilician proconsulship. Six books in dialogue form, set on
the three days of the Latin holidays in 129 BC, three years
before the death of the dramatic host Publius Cornelius
Scipio Aemilianus Africanus the Younger --- the conqueror of
Carthage and Numantia. The interlocutors are Scipio's circle:
Laelius, Philus, Manilius, Mummius, Tubero, Furius, Rutilius
Rufus (whom Cicero had heard report the conversation in his
own youth at Smyrna). The work is Cicero's response to
Plato's \textit{Republic}, with two large differences. First,
the best constitution is not constructed in speech --- it
is exhibited in fact, in the historical Roman state. Second,
the closing dream-vision is not the myth of Er but the
Somnium Scipionis: the elder Africanus appears to the younger
in a dream from the Milky Way and shows him the smallness of
the earth, the harmony of the spheres, and the immortal
reward awaiting those who have served the commonwealth.}

\textit{The transmission is broken. Books 1 and 2 survive
substantially intact through the Vatican palimpsest discovered
by Cardinal Mai in 1819, with parts of book 3 and scattered
sections of books 4--6 also recovered. The argumentative
heart of book 3 --- Carneades's case against natural justice,
Laelius's reply, the famous formulation of \textit{vera lex}
as right reason agreeing with nature, the analysis of just
war --- comes mainly through Lactantius and Augustine. Books
4 and 5 are almost entirely lost; what remains comes from
late grammarians (Nonius), the Christian Fathers (Augustine,
Lactantius), and the scholiasts. The Somnium Scipionis (6.9--29)
survives in full only because Macrobius transmitted it
separately, with his own neo-Platonic commentary; through
that channel it became, with Boethius, one of the central
philosophical texts of the medieval Latin tradition.}

\textit{Book 1 is the dialogue's setting and the formal
argument: Cicero's defence in his own voice of the active
political life against philosophical \textit{otium}; Scipio's
classification of the three simple kinds of state (kingly,
optimate, popular), the three corruptions (tyrant, faction,
mob), and the cycles between them; and the case for the
mixed constitution as steadier than any pure form. Book 2
is the case study, treating the Roman state's gradual
construction by the seven kings and the early Republic ---
Cicero's most concentrated reading of his own constitutional
history, with the deep Catonian thesis that ``the
commonwealth is the work neither of one time nor of one
man.''}

\textit{Book 3 is the philosophical centre. On the second
day, at Tubero's prompting, Philus takes the part of
Carneades, who at Rome in 155 BC argued first for and then
against justice: empires grow by injustice, the just man
allows himself to be cheated, the prudent state prefers to
rule unjustly than serve justly. Laelius answers --- in the
fragments we have, with the famous formulation that ``true
law is right reason agreeing with nature, diffused in all,
constant, everlasting.'' Books 4--5 take up the formation
of citizens through education, custom, the shame-discipline,
and the figure of the \textit{rector rei publicae} --- the
statesman as guardian and steerer of the state. Book 6,
through the Dream of Scipio, frames the whole: the rewards
of the public man are not earthly fame (the earth itself,
seen from the Milky Way, is a point) but the eternal seat
in heaven reserved for those who have preserved their
commonwealths.}

\textit{The text in this volume is rendered from the
Perseus edition (PHI 0474.043), which preserves the
palimpsest text of books 1--3 followed by the fragments of
books 4--6 in their standard Powell / Ziegler numbering.
Where the Latin breaks mid-thought (frequent in books 1, 2,
and 3), the English follows the same breaks, marked
\textit{[\ldots]}. Fragmentary passages preserved by other
authors are cited at the head of each section in italics.}
