% Ad Familiares 10.27 --- headnote
% ad-familiares-10-27, 20 March 43 BC (evening), Rome

\textit{Cicero to M. Aemilius Lepidus, from Rome on the
evening of 20 March 43 BC --- Perseus dateline
\textit{Scr. Romae xiii K. Apr. vesperi a. 711 (43)}.
Lepidus is governor of Hispania Citerior and
Narbonensis and has, in February, written to the
Senate urging a negotiated peace with Antony. The
Senate has just adorned him with the honours that
section 1 alludes to (a gilded equestrian statue on
the Rostra, voted on 30 December 44 and again
celebrated in early 43), and the present letter is
Cicero's reply to a man who has accepted the honours
without thanking the body that bestowed them.}

\textit{The letter is two short paragraphs of cold,
exact diplomacy. The opening compliment is genuine ---
Cicero has supported Lepidus's standing on the floor
and in the Philippics --- but every sentence after the
first is a warning. The fork in section 1 is the moral
hinge of the letter: peace separated from servitude
serves the state; peace that restores Antony to
``utterly intolerable despotism'' restores the worst
master Rome has ever known. The closing line ``you, in
your prudence, will see what is best to be done''
($\textit{tu pro tua prudentia quid optimum factu sit
videbis}$) is the polite formula stripped of its
warmth: a public man writing to another public man on
the record, putting him on notice that the Senate, the
people, and ``every honest man'' will read what he
does next. Months later Lepidus would defect to
Antony; the wariness in this letter is already
visible.}
