% Ad Familiares 10.20 --- headnote
% ad-familiares-10-20, 29 May 43 BC, Rome

\textit{Cicero to L. Munatius Plancus, written at Rome on 29
May 43 BC --- Perseus dateline \textit{Scr. Romae iv K. Iun.
a. 711 (43)}. The dateline ``iiii K. Iun.'' is even repeated
at the foot of the letter itself, in Plancus's own hand or
Cicero's secretary's, a rare in-text dating note. Cicero
writes against a steady drumbeat of contradictory reports
from Gaul: news of Lepidus is wavering --- one day he is
holding firm against Antony, the next he is letting Antony's
men through.}

\textit{The hinge of the letter is the rumour, picked up by
D. Brutus and forwarded by him to Cicero, that ``Antony is
not being received by Lepidus.'' Cicero will not credit it
until Plancus himself confirms it; Plancus, it seems, has
been burned once already by a previous false report of
Lepidus's loyalty (\textit{inanis laetitia litterarum
superiorum}, ``the empty rejoicing your earlier letters
caused''), and is now reluctant to put anything sanguine in
writing. The proverbial tag in the second section, ``the
same man at the same stone twice'' (\textit{his ad eundem},
short for \textit{bis ad eundem lapidem offendere} ---
``to stub your toe twice on the same stone''), is one of
the few openly proverbial flashes in this correspondence.
The third section repeats Cicero's standing claim, here in
its sharpest form: whoever crushes the remnants of the war
will be the one who has finished it --- and he wants this
man to be Plancus.}

\textit{Behind the polite uncertainty about Lepidus, the
shape of what is actually coming is already visible. Three
days earlier (29 May, the date of this letter) was in fact
the day Lepidus's army went over to Antony at the Argenteus
river. The dispatch crossed Plancus's news in transit;
within the week Cicero will know that everything the rumour
mill had been hinting at was true, and Plancus is back
across the Is\`ere with a force that is now, by itself, all
that stands between Antony and Italy.}
