% Ad Familiares 11.21 --- headnote
% ad-familiares-11-21, 4 June 43 BC, Rome

\textit{Cicero to D. Junius Brutus Albinus, imperator and
consul-elect, from Rome on 4 June 43 BC --- Perseus dateline
\textit{Scr. Romae prid. Non. Iun. a. 711 (43)}. The
political weather has just turned again. On 30 April the
senate received the news of Antony's defeat at Mutina; on
30 May, on Cicero's motion, Lepidus was declared a public
enemy after his junction with Antony in Gaul. The
\textit{supplicationes} for Mutina had been voted, but the
honours and the practical follow-on --- the planned board
of ten to settle veterans' lands, the four legions Brutus
proposed to keep up, the relations between Brutus's army
and Octavian's --- were all live business in early June.
Brutus is still in the field, advancing into Cisalpine Gaul
to take up the pursuit of Antony; Cicero is in Rome,
managing the senate.}

\textit{The letter is a reply on four fronts. First, a piece
of poisonous gossip retailed by one Segulius Labeo --- a
man Cicero had encountered before --- to the effect that
the veterans were murmuring against Brutus and Octavian
because they had not been placed on the agrarian
commission of ten. Cicero, who \textit{was} on the
commission, brushes the gossip aside: it was on his own
motion that commanders in the field were exempted from the
board, against the opposition of the usual obstructors.
Second, Brutus has expressed personal anxiety for Cicero's
safety; Cicero releases him from it with a Stoic dignity
he has earned the right to claim. Third, Brutus has urged
him to keep his nerve --- ``do not let fear of fear push
you into fear'' --- and Cicero accepts the counsel while
gently turning it round: with Brutus's resources and his
prospective consulship, fear should have no foothold.
Fourth, the concrete agenda --- the four legions to be
kept on the books, the lands to be assigned by Brutus and
Octavian jointly --- has Cicero's energetic backing; he
has personally beaten off a play by other senators to take
the land commission for themselves, and offers to send
sensitive matter back by a private courier. The closing
date, \textit{Pr. Non. Iun.}, is 4 June.}
