% Ad M. Brutum 2.2 --- headnote
% ad-brutum-02-02, 11 April 43 BC, Rome

\textit{Cicero to M. Junius Brutus, written from Rome ---
Perseus dateline \textit{Scr. Romae iii Id. Apr. a. 711
(43)}, i.e.\ 11 April 43 BC. The date is internally
corroborated by the reference in section 3 to a Senate
session held \textit{a.\ d.\ v Idus Aprilis} (9 April),
which Cicero now relates as a recent event. The dateline
matches the meta date. The letter is days from the
battle of Forum Gallorum (14 April) and Mutina has not
yet been relieved: the besieged Decimus Brutus is the
``Brutus'' on whose extrication ``all our hope lies''
(section 2). Section 1 reports the contents of a letter
from L. Munatius Plancus, governor of Transalpine Gaul,
who has at last declared firmly for the Senate; against
him is set Lepidus, governor of Nearer Spain and
Narbonensis, ``your kinsman'' (M. Brutus's wife Porcia
was a half-sister of Lepidus's wife) --- whose brother
L. Aemilius Paullus had been on the proscription list
prepared for Caesar's funeral, the family fracture lying
just under Cicero's elegant antithesis about ``the men
he most hates as his nearest relations.''}

\textit{Section 3 is the heart of the letter and a
revealing piece of senatorial reportage. P.\ Servilius
Isauricus --- a consular, a Caesarian by descent, and
the same Servilius whose motion of 27 April will appear
in 1.5 --- has fought Cicero over the senatorial honours
for Plancus, has been beaten in the contest, and at the
critical moment a dispatch from Cn.\ Cornelius Lentulus
Spinther in Asia arrived, announcing Cassius's seizure
of the legions in Syria; Cicero read it out in session
to crushing effect (\textit{cecidit Servilius, complures
praeterea}). The half-finished sentence at the end ---
``It is a great prodigy in the commonwealth, but in
respect of which * * *'' --- breaks off in the manuscript
with a clear lacuna. The phrase \textit{magnum monstrum
in re publica} refers to Servilius, whose rage Cicero
reads as politically diagnostic; what completed the
sentence is lost. Throughout, the voice is the most
intimate available: a senior statesman, the day after
a Senate floor-fight, writing privately to a younger
ally about who has just been broken on the floor and
how.}
