% Ad Familiares 10.8 --- headnote
% ad-familiares-10-08, shortly after mid-March 43 BC, Transalpine Gaul

\textit{L. Munatius Plancus, imperator and consul-designate,
to the consuls, praetors, tribunes of the plebs, the Senate,
the Roman people and plebs, written from Transalpine Gaul
shortly after the middle of March 43 BC --- Perseus dateline
\textit{Scr. in Gallia Transalpina paulo post med. Mart.
a. 711 (43)}. This is the formal public dispatch which the
brief covering note Fam. 10.7 was sent on top of. It is by
far the longest letter in the Plancus correspondence, the
most carefully composed, and one of the principal public
documents of the spring of 43.}

\textit{Plancus's situation is delicate. He is the
designated consul for 42 and the holder of Transalpine
Gaul, with five legions of his own and a province whose
communities had been pre-bribed under Antony's Lex
Cornelia of November 44; he had been a Caesarian for ten
years; his brother Plotius is in Rome; his colleague-elect
Decimus Brutus is under siege at Mutina. He has been slow
to declare himself, and the slowness has drawn comment in
the Senate. The letter's first task is to defend the
slowness as deliberate strategy --- ``thoughts of the
soundest mind long since formed'' --- not as the
opportunism of a man waiting to see which way the wind
blows. The argument runs: an early profession would have
made the army harder to steady, the bribed Gallic
communities harder to pull back, the neighbouring
governors (Lepidus above all) harder to draw into a
partnership; \textit{premature denuntiatio} would have
been ruinous, as it had proved for his colleague ---
Decimus Brutus, now besieged. So he ``feigned much
unwillingly and dissembled much with pain'' to keep his
options open while building up his force.}

\textit{The structure of the letter is rhetorically careful:
the apologia of sections 1--2; the catalogue of preparation
in section 3 (the army, the provincial communities, the
neighbouring commanders); the candid admission of
dissimulation in section 4; the reference to the secret
oral commissions carried by C. Furnius in section 5; and
the public declaration in section 6 --- five legions, a
province in consensus, cavalry and auxiliaries, and a
governor prepared for anything from defending his
province to dying in its cause. The closing section
recommends those who have followed his lead in Gaul to
the Senate's protection, and frames the whole as an
intervention to be judged \textit{aequis iudicibus},
``by fair judges,'' against the slander of envy ---
already, before the campaign opens, a glance at the
faction-fighting that will swallow him later in the year
when he goes over to Antony at Lepidus's camp.}
