% Ad Familiares 10.10 --- headnote
% ad-familiares-10-10, 30 March 43 BC, Rome

\textit{Cicero to L. Munatius Plancus, governor of Transalpine
Gaul, written from Rome on 30 March 43 BC --- Perseus dateline
\textit{Scr. Romae iii K. Apr. a. 711 (43)}. The letter is short
and answers one of Plancus's reports; the battle at Mutina, on
which ``the entire fortune of the state is being decided,'' has
not yet been fought, and Cicero writes in the certainty that by
the time the letter is read the issue will have been settled
one way or the other.}

\textit{The substance is twofold. Plancus's commitment to the
republican cause --- already known to Cicero from his
conversations with Furnius and now confirmed in writing --- has
won him \textit{magnam laudem} ``great commendation''; the
Senate, were the consuls in Rome and not in the field, would
already have voted him honours. But the moment is not yet ripe,
and Cicero wants Plancus to understand his theory of public
recognition: real honour is not a bait dangled before the
service, it is the prize of unbroken virtue (\textit{perpetuae
virtutis praemium}) awarded after it. The closing imperatives
--- \textit{incumbe toto pectore ad laudem, subveni patriae,
opitulare conlegae} --- press Plancus to lean his whole weight
into the work, with Lepidus (the colleague of the title) named
as the man he must stand by. The final pledge --- that to the
old causes of their friendship, public spirit has now been added,
and Cicero now sets Plancus's life above his own --- is a degree
of warmth Cicero only reaches when the political stakes are
unbearable.}
