% Ad Familiares 12.14 --- headnote
% ad-familiares-12-14, 29 May 43 BC, Perga

\textit{P. Cornelius Lentulus Spinther (the younger) to Cicero,
from Perga in Pamphylia on 29 May 43 BC --- Perseus dateline
\textit{Scr. Pergae iiii K. Iun. a. 711 (43)}.
Lentulus is proquaestor of Asia, left in charge of the province
after the murder of Trebonius at Smyrna by Dolabella in January,
and writing as one of the new men of the cause: son of the consul of
57 BC who had championed Cicero's recall, intimate of Dolabella in
peacetime, kin to the Antonii by marriage, holding his commission
through their favor --- and now, by his own telling, the first man
to break with all of them on the state's behalf. The dispatch is a
long act of self-presentation cast as a field report: the chase
after Dolabella's hundred-ship transport fleet in Lycia (broken up
when the Rhodians delayed him); the Rhodians themselves, who shut
their gates against his father in the civil war and now shut them
again; a request that Cicero secure him a continuation of his
command in Asia from Pansa and Hirtius; an accounting of what he
has done for Cassius in Syria; and the central claim, sealed with
a fragment of Homer, that he loved his country more than his own.}

\textit{The political moment is that brief window in the spring of
43 when the Liberators' eastern coalition still looked like a state
in waiting. Cassius has Dolabella penned in at Laodicea on the
Syrian coast; M.~Brutus is consolidating Macedonia and Greece;
Lentulus, holding Asia in the absence of the consuls, has redirected
the province's revenues and cavalry to Cassius. His confidence here
that ``Dolabella has already been finished off and broken'' is
about to be confirmed: Dolabella will take his own life at Laodicea
within weeks. The confidence about Pansa and Hirtius is already, at
the moment of writing, retrospective and tragic --- both consuls
were dead at Mutina before this letter left Perga, though the news
had not yet reached Asia. The note about Cicero's young son with
M.~Brutus, in winter quarters with the cavalry, is the personal
seal at the end of an otherwise official dispatch: Lentulus is
asking the father of one of the cause's officers to look after the
career of another. The conventional Perseus heading dates the letter
1 June; the dateline itself reads \textit{iiii K. Iun.} = 29 May,
which is followed here.}
