% Ad Familiares 10.21 --- headnote
% ad-familiares-10-21, around early June 43 BC, camp in Allobrogan country

\textit{L. Munatius Plancus to Cicero, written from camp
shortly before the Ides of May 43 BC according to the
Perseus dateline \textit{Scr. in castris pnd. Id. Mai.
a. 711 (43)}; but the events the letter narrates ---
the Isara bridge and the warning from Laterensis ---
belong to the days immediately preceding Lepidus's
defection on 29 May, so a dating in late May or the very
beginning of June is what the contents require.}

\textit{The letter is the apologia for an alliance that
fell apart in Plancus's hands. He had written
confidently two days earlier that Lepidus was a good man
and the war would be conducted by joint counsel; this
letter unwrites that. The Isara has been bridged in a
single day on Lepidus's urging; an orderly then arrives
with a fresh dispatch telling Plancus \textit{not} to
come --- Lepidus, Plancus suspects, wants to keep the
glory to himself. Plancus's tone is candidly
self-incriminating about this --- ``I will lay bare my
rash plan to you'' --- and the candour is part of the
self-defence. Then comes the letter of Laterensis, the
upright Senate-appointee inside Lepidus's camp, warning
that Lepidus's good faith is gone and that he, Laterensis,
considers his own pledge to vouch for it discharged. The
mutinous shout from Lepidus's troops --- ``we want peace,
we will not fight'' --- which Lepidus neither punishes
nor cures, is the proof. Plancus pulls back, keeps the
army intact, and asks Cicero to hurry an army up from
Italy. The closing imperatives --- send the army quickly,
before the enemy gather and our side falter --- are the
operative request the letter is built to deliver.}
