% Ad M. Brutum 1.18 --- headnote
% ad-brutum-01-18, 27 July 43 BC, Rome

\textit{Cicero to M. Brutus, written from Rome on
27 July 43 BC --- Perseus dateline \textit{Scr. Romae vi
K. Sex., a. 711 (43)}, confirmed by the closing subscript
``\textit{vi Kal. Sextilis}.'' The meta entry's
year-precision placeholder of 7 November is wrong by
more than three months and is corrected here. This is one
of the very latest of Cicero's surviving letters: he has
under four months to live.}

\textit{The setting is a private meeting at the house of
Servilia, Brutus's mother. Two days before this letter
(25 July), Servilia summoned Cicero with Casca, Labeo,
and Scaptius --- the inner circle of Brutus's friends in
Rome --- to consult on whether Brutus should be called
back from the East with his army. Cicero gave the answer
the meeting wanted: yes, at once, before the cause
collapses. The grim picture of section 2 --- the
victorious consular armies that refused to pursue
Antony after Mutina, and Lepidus declaring war on the
state from a position of safety, wife and children
intact --- is the strategic background; the answer is
Brutus's eleven legions in Macedonia. The famous central
section is Cicero's confession of his anxiety about
Octavian: he has gone surety to the state for a young man
of pliant age surrounded by many ready to corrupt him,
and he is no longer sure he can make good on the pledge.
This is the same anxiety Brutus had diagnosed in letter
1.17 some weeks earlier, now seen from the other side.}

\textit{The closing material is practical and personal.
Section 5 names the structural problem of the war:
\textit{tributum}, the property-tax levied on Roman
citizens that was supposed to fund the legions, has been
shamelessly under-declared by the rich and is being
swallowed up in legionary bonuses; the cost of supplying
Brutus's army and the others is now beyond what taxation
will reach. Cassius's army, by contrast, will arrive
adequately funded. Section 6 returns to the case of
Brutus's nephews (the sons of his sister Junia by
Lepidus, recently declared a public enemy): Cicero has
already, before being asked, argued for the boys in the
Senate, and tells Brutus so. Within days of this letter
Octavian's army would begin its march on Rome to extort
the consulship; within five months Cicero would be dead
in the proscriptions; within nineteen months Brutus would
be dead at Philippi. The letter is among the last where
the strategic horizon of the senatorial party is still
intact on paper.}
