% Ad Familiares 10.23 --- headnote
% ad-familiares-10-23, 6 June 43 BC, Cularo in the territory of the Allobroges

\textit{L. Munatius Plancus to Cicero, written from
Cularo (modern Grenoble) in the territory of the
Allobroges on 6 June 43 BC --- Perseus dateline
\textit{Scr. Cularone in finibus Allobrogum viii Id. Iun.
a. 711 (43)} and closing subscript in the letter itself.
This is the dispatch in which Plancus reports Lepidus's
defection to Antony, the most consequential single piece
of news to reach Rome between Mutina in April and the
formation of the triumvirate in November.}

\textit{Lepidus had joined Antony on 29 May. The
nominally Senate-loyal commander of Narbonensis and
Nearer Spain, with seven legions, had crossed over into
open coalition with the man the Senate had declared a
public enemy. Plancus, advancing from Transalpina to
support Lepidus, learnt of the defection when the joined
armies were already on the move against him; his account
of the withdrawal --- bridges broken on the Isara, not a
soldier lost, baggage train intact, the bitterness of
the ``parricides'' that he had slipped through their
fingers --- is the work of a commander who knows he
will be judged on whether he can extract himself in
order. The first section is an explicit defence against
the charge of \textit{temeritas}: he had never, he says,
really trusted Lepidus; what drove him to advance into
contact was \textit{pudor} --- the fear of being charged
with feeding the war by his own inaction. The
self-portrait is part of the letter's political work.}

\textit{The middle of the letter does three further
pieces of business. It absolves M. Iuventius Laterensis,
the Senate's man inside Lepidus's camp, from blame and
reports his attempted suicide --- a small but
politically careful kindness toward the family. It
itemises the grievances Lepidus's troops carried against
Plancus (refusing to receive Antony's envoys; intercepting
the tribune Catius Vestinus with Antony's letter; refusing
to negotiate) so that the Senate will know on what side
of the war Plancus has been all along. And it calls
urgently for reinforcement: Octavian, with his five
legions and growing ambition, must be sent up the
peninsula, or his army at least dispatched, or the
republican cause in Gaul will not hold. The closing
personal coda to Cicero --- ``I hold you dearer every
day'' --- is the conventional intimacy that frames the
substance: \textit{I have been loyal; I have been
prudent; I need you to send the legions north now}.}
