% Ad M. Brutum 1.14 --- headnote
% ad-brutum-01-14, 11 July 43 BC, Rome

\textit{Cicero to M. Brutus, from Rome, 11 July 43 BC
--- Perseus dateline \textit{Scr. Romae v Id. Quint., a.
711 (43)}, the date confirmed by the closing subscript
``\textit{v Idus Quintiles}.'' Day-precision in
\texttt{meta/works.yaml}. The opening is one of the
sharpest reproaches in the correspondence: ``A short
letter from you --- short, do I say? rather, none at all.
Three little lines, in such times as these, from Brutus
to me? You would have done better not to write.''
Brutus's previous note had been a thin scrap; it had
also reported that young Marcus Cicero, who has been
studying with Brutus in Greece, was about to leave him
for Italy. Cicero has packed off couriers to head his
son back. He has been working hard to postpone the
priestly elections to the next year for his son's sake
(and for Domitius, Cato --- Brutus's brother-in-law ---
Lentulus, and ``the Bibuli'').}

\textit{Section 2 is the political plea, sharpened to
something close to an order. The war has been reborn
through Lepidus's crime; Octavian's army, ``which was
the best,'' is no longer a help and is now actively
forcing the demand for Brutus's army to be heard;
``if that army shall touch Italy, there will be no
citizen --- no one, at least, whom it is lawful to call
a citizen --- who does not take himself into your
camp.'' The closing sentence is the famous one ---
the appeal that turns the meaning of the Ides on its
head: ``persuade yourself that on the Ides of March,
when you drove out slavery from your fellow citizens,
you did not bring your country more benefit than you
will bring by coming in good time.'' Brutus will not
come. Eight days after this letter, on 19 August,
Octavian will march on Rome and seize the consulship.
The next surviving letter to Brutus, 1.15, is later
and longer; 1.14 stands as the last unstrained appeal.}
