% Ad M. Brutum 1.12 --- headnote
% ad-brutum-01-12, mid-July 43 BC, Rome

\textit{Cicero to M. Brutus, from Rome, mid-July 43 BC
--- Perseus dateline \textit{Scr. Romae in. m. Quint.
a. 711 (43)}. The meta entry rounds the month-precision
date to 15 July; the dateline's \textit{in.~m.~Quint.}
(``at the beginning of July'') would actually place it
a few days earlier than 1.10 (\textit{med.~m.~Quint.}),
but the two letters are clearly of a piece in tone and
political moment. The bearer is the \textit{vetus
noster} --- the ``old friend'' --- whom Cicero
mentioned at the close of 1.9 as the future courier of
a fuller dispatch; in this letter he is praised
explicitly. Messala Corvinus is also about to carry a
letter east.}

\textit{Two things happen here. The first is the
mother-and-sister scene of section 1. Servilia and
Junia have come to Cicero to plead for Lepidus's
children, who stand to be ruined if Lepidus is declared
a public enemy (as he will be, on 30 June, when the
Senate hostis-declares him). Brutus's mother and
half-sister are also Lepidus's wife's relations ---
this is a tightly intermarried world --- and Cicero is
distressed that he could not yield to their pleas. His
reasoning is rigorous: Lepidus's case cannot be
distinguished from Antony's; the laws have wisely
arranged that love of children should make a parent a
better citizen, not the reverse; ``it is Lepidus who is
cruel to his children, not the man who pronounces
Lepidus an enemy.'' The second thing is the renewed
imperative of section 2: come into Italy, with your
forces and with your counsel, as soon as may be. The
closing paragraph passes on news of young Marcus
Cicero, studying with Brutus in Greece: Cicero hopes
to see his son ``before long,'' with Brutus bringing
him in person. That hope, like the larger hope, will
not be redeemed.}
