% Ad M. Brutum 1.6 --- headnote
% ad-brutum-01-06, 19 May 43 BC, from camp at the foot of Candavia

\textit{M. Junius Brutus to Cicero, written on 19 May 43
BC. The Perseus header gives the place as Rome (\textit{Scr.\
Romae xiv K.\ Iun.\ a.\ 711 (43)}), but this is a Perseus
error: the letter's own closing subscript reads
\textit{x iiii K.\ Iunias ex castris ad imam Candaviam}
--- ``the 14th day before the kalends of June, from camp
at the foot of Candavia'' --- which fixes both the date
(19 May) and the place (the mountain pass on the Via
Egnatia between Dyrrachium and Macedonia). The launch
metadata in the meta entry gives Dyrrachium, plausibly
the staging city from which Brutus had set out; the
closing subscript is authoritative and is used in the
parallel sidecar. Brutus is on the march from the
Adriatic into Macedonia.}

\textit{The letter has four practical items, each handled
with Brutus's characteristic plainness. Section 1 waves
off thanks --- the relationship is past such formalities
--- and reports that young Marcus, currently leading the
cavalry from Ambracia through Thessaly, will meet Brutus
at Heraclea; what to do about his return to Rome for
office (whether to stand in person or to be set up in
absence) will be decided jointly when they meet. Section
2 is a personal recommendation of Glyco, Pansa's
physician, accused of poisoning his patient: Brutus
defends him on motive (Glyco had everything to lose by
Pansa's death) and on character, and asks bluntly for
his release. Section 3, the most historically loaded of
the four, reports the news --- still arriving piecemeal
--- that Dolabella has been cut down and put to flight,
sent on by Satrius, Trebonius's former lieutenant, via
Tillius and Deiotarus; the actual death of Dolabella at
Laodicea, by his own hand under siege by Cassius, will
be confirmed in fuller detail later. Section 4 closes
with a private matter: a contested inheritance at
Dyrrachium in which Flavius (a friend of Brutus's) is
heir and the city denies the debt as forgiven by
Caesar; Brutus asks Cicero, as agreed judge, to settle
it. The Greek letter of Cicereius mentioned at the end
of section 3, sent to Satrius and enclosed by Brutus,
is not preserved.}
