% Ad Familiares 10.24 --- headnote
% ad-familiares-10-24, late July 43 BC, camp in Gaul

\textit{L. Munatius Plancus to Cicero, written from camp
on 28 July 43 BC --- Perseus dateline \textit{Scr. in
castris v K. Sext. a. 711 (43)}, the date confirmed by
the closing subscript ``\textit{v.~K.~Sext. ex castris}''.
This is the last surviving letter of Plancus's
correspondence with Cicero, and one of the most clear-
sighted political documents of the late summer of 43. By
the time of writing, Plancus has joined his army to
that of Decimus Brutus, ten legions are camped together
in Gaul against Lepidus and Antony, and the situation
is paradoxically the worst it has been: Octavian, on
whose march north all the republican hopes turn, has
stopped marching.}

\textit{The arithmetic of the war is the heart of the
letter. Plancus has three veteran legions in his own
camp and one fine legion of recruits; Decimus Brutus
has one veteran, one of two years' service, and eight
of recruits --- a total ``very large in numbers, but
slight in reliability,'' which is a candid admission
for a commander to commit to paper. The Africanus
exercitus, the four veteran legions from Africa under
Q. Cornificius, is one of the two forces that could
tip the balance; the other is Octavian, the closer
and more obvious. Plancus has written, Furnius has
been sent in person with letters and instructions,
Octavian has assured him he is coming without delay
--- and then has turned aside to ``other plans,''
which Plancus identifies in section 6 with painful
precision: the demand for a two-month consulship
extorted from the city ``with the utmost terror and
in tasteless form.'' That demand will be granted on
19 August. The triumvirate is two months in the
future as Plancus writes; the bitter sentence
``Antony is alive today, Lepidus is at his side,
they have armies, they have hope, they dare ---
they can charge it all to Caesar's account'' is the
clearest contemporary diagnosis of what is happening.}

\textit{The personal frame is unsparingly careful.
Section 1 is a courtly thank-you for senatorial
attentions, qualified by the diffidence of a man
who knows that the formal vote is the cheap currency
he is forced to use; section 5 is the disclaimer that
he, like Cicero, has been a friend to the young
Caesar in good faith, on grounds Plancus is at pains
to list (his intimacy with the elder Caesar; the
young man's moderate disposition; the propriety of
treating an adopted son with the affection due his
father's friend). The complaint of section 6 is thus
not the complaint of an enemy. It is the complaint of
an ally writing for the record --- ``more in sorrow
than in unfriendliness'' --- as the alliance between
the Senate and its young Caesarian commander begins
visibly to dissolve. Within a month of this letter
Octavian would march on Rome; within four months
Plancus, like Lepidus before him, would be in the
triumviral camp; within five, Cicero would be dead.
The Perseus text in section 3 carries an unresolved
crux at ``\textit{talis victoriae},'' here preserved
as ``a victory of such a kind.''}
