% Ad Familiares 10.33 --- headnote
% ad-familiares-10-33, late May or early June 43 BC, Corduba

\textit{C. Asinius Pollio to Cicero, written from Corduba
in the last days of May or the first days of June 43 BC
--- Perseus dateline \textit{Scr. Cordubae vel ex. m.
Maio vel in. Iun. a. 711 (43)}, a range deliberately left
open because the letter itself reports the news of the
Mutina battles as received only after a forty-day delay,
and the chronology of Pollio's dispatches has to be
worked back from the date Lepidus detained his couriers.
Pollio, governor of Hispania Ulterior with three veteran
legions, is the southernmost commander on the
republican side. He has just learned of the deaths of
both consuls and of the destruction of the Martian
legion; he has also been told --- erroneously --- that
Octavian has fallen. The letter is a longer, more
reflective companion to Fam.~10.32: where 10.32 was the
sketch of Balbus's outrages, this one is Pollio
attempting, with elaborate care, to lay out his
political position.}

\textit{The position is the awkward one of a Caesarian
who has been left, by accidents of distance and timing,
on the senatorial side. The letter opens with the
forty-day lag in news --- Lepidus held Pollio's
couriers for nine days, which Pollio reports as a fact
and explains by Lepidus's wish to keep him in the dark
--- and circles back to it at the close. In between
come the two main pleas. First: had the Senate's
recall, which brought Plancus and Lepidus into Italy,
extended to him as well, the catastrophe at Mutina
might have been avoided. Second: he could not move
toward Lepidus without flattering him, given the
letters Lepidus has been writing and the speeches he is
said to have made at Narbo, because his supplies on the
march would otherwise have been cut off; and he could
not move freely on his own initiative without giving
his detractors at Rome --- who know his earlier
friendship with Antony --- the chance to read the move
the wrong way. The recital of the Mutina news in
section 4 is the longest connected piece of military
reporting in the Pollio correspondence: it carries the
news of Pansa and Hirtius dead, the Fourth legion cut
to pieces after taking Antony's camp, Ventidius
joining Antony with three more legions, Parma
sacked, L. Antonius across the Alpine passes. The
report that Octavian has been killed is the one
substantial inaccuracy; Pollio's grief at it (``which,
if it be true, and may the gods forbid, grieves me
beyond measure'') is one of the more striking
expressions of personal loyalty to the young Caesar
that survive from the side of Cicero's allies.}

\textit{The closing sentence --- ``I wish neither to fail
the commonwealth nor to outlive it'' --- is the
formulation by which Pollio has been remembered. Within
six months he would join Antony, then enter Octavian's
camp; he would survive both consulships and the
proscriptions, and would be writing the history of the
civil wars decades later. The Perseus dateline's range
(\textit{vel ex.\ m.\ Maio vel in.\ Iun.}) sits awkwardly
against the day-precision \mbox{-0043-06-13} carried for
this letter in the project manifest; the manifest entry
likely wants relaxing to month precision or to a range
spanning the last week of May into the first week of
June. The salutation as transmitted, \textit{POLLIO
CICERONI S.~N}, is preserved verbatim from the Latin
file --- the standard expansion is \textit{salutem nuntiat},
``Pollio sends greeting to Cicero,'' though the
reconstruction is not certain. The closing prayer
\textit{di prohibeant} (``may the gods forbid'') is here
the wish that the report of Octavian's death prove
false.}
