% Ad M. Brutum 1.4 --- headnote
% ad-brutum-01-04, mid-May 43 BC, Dyrrachium

\textit{M. Brutus to Cicero, written from Dyrrachium in
mid-May 43 BC. The Perseus dateline as printed reads
\textit{Scr. Dyrrachi iii antprid. id. Mai. a. 711 (43)};
the printed string ``iii antprid.'' is a corruption of
``\textit{iii ante prid. id. Mai.}'' --- ``the day before
the day before the Ides of May,'' i.e. 13 May 43.
The meta entry's placeholder of 21 September is far
too late and is corrected here. Dyrrachium (modern
Durrës) is Brutus's base in Macedonia while he
consolidates his eastern command.}

\textit{The letter is Brutus's first congratulation on
the victory at Forum Gallorum and Mutina (14 and 21 April
43): his nephew Decimus Brutus's break-out from Mutina,
he says, was both the saviour of Decimus himself and the
decisive contribution to the success. The body of the
letter is the famous statement of Brutus's juridical
principle. Cicero had written that ``the three Antonii''
--- M. Antonius, his brother L. Antonius the tribune, and
his other brother C. Antonius, whom Brutus held captive
at Apollonia --- are now to be treated as a single case.
Brutus refuses. He has not killed C. Antonius, though he
could have, because the Senate has not condemned him; he
has not stripped him of his property cruelly, nor remitted
his confinement dissolutely; and he insists that the
question of citizens taken in arms is for the Senate and
people to decide, not for a commander in the field. The
closing section, broken off in the manuscripts after
``when you have set me right * *,'' is the first surviving
note of the disagreement that will widen through the
summer: Brutus's accusation that Cicero is too quick to
trust ``a mind once corrupted by lavish gifts'' --- meaning
Octavian --- to ``bad counsels'' will become the central
theme of letter 1.17 to Atticus six weeks later.}
