% Ad Familiares 11.10 --- headnote
% ad-familiares-11-10, 5 May 43 BC, camp at Dertona

\textit{Decimus Brutus to Cicero, written from camp at Dertona on 5 May 43 BC --- Perseus dateline \textit{Scr. in castris Dertonae iii Non. Mai. a. 711 (43)}. A week after the dispatch from Regium (11.9), D. Brutus has pushed his pursuit westward across the Apennines into Liguria. Antony, who had come out of the defeat at Mutina with a small unarmed band of foot, has gathered a force by emptying slave-pens and sweeping up irregulars; Ventidius, having made a punishing crossing of the Apennines, has joined him at Vada with a body of veterans and armed men. The two armies that ought to have caught Antony between them --- D. Brutus from the east, Octavian from the south --- have not closed, and Brutus's frustration with the young Caesar (``Caesar cannot be commanded, nor Caesar his own army --- both of which are very bad indeed'') is the political tell at the centre of the letter.}

\textit{The dispatch is in two registers. The opening paragraphs are personal and political --- thanks to Cicero, scorn for the senators who are already withholding the honours Brutus has earned, a tight summary of the danger left by the consuls' deaths and the vacancy. Sections 3 and 4 are the strategic appreciation: Antony's three possible courses (to Lepidus, to the mountains, back into an undefended Etruria), the failure of co-operation with Octavian, the fear that even what Cicero may untangle will be tangled again. The closing section is the supply crisis stated baldly: seven legions to feed, his personal fortune already spent, friends bound in debt --- a senator with an army in the field who cannot keep it in the field. He had four hundred thousand sesterces of his own when he began the war; he tells Cicero that not Varro's treasuries would now meet the bill. The figure in the manuscript transmission is uncertain (most modern editors emend to a much larger sum, consistent with what a man of his standing would have had), but the rhetorical point is the same: the senatorial commander in the field is bankrupt.}
