% Ad Familiares 12.13 --- headnote
% ad-familiares-12-13, 13 June 43 BC, Cyprus (Crommyuacris)

\textit{C. Cassius, proquaestor, to Cicero, from Cyprus at Crommyuacris on
13 June 43 BC --- Perseus dateline \textit{Scr. in Cypro Crommyuacride Id. Iun.
a. 711 (43)}. (The sender is C. Cassius Parmensis, the tyrannicide and
proquaestor in the eastern campaign --- not the more famous C. Cassius Longinus,
``noster Cassius'' of \S 4. The two are commonly confused in manuscripts.)
Cassius writes from the Cypriot promontory of Crommyuacris, where his squadron
has put in after pulling levies of ships and rowers out of the Asian coast and
the islands. The first two sections are tribute, almost a prose hymn: the
\emph{toga} of Cicero has been ``more fortunate than all men's arms,'' has
snatched a beaten Republic back from the enemy, and the writer asks only to be
recommended to Cicero's own judgement and brought forward as part of the
country's best hope. The factual report follows. The conspirators' main fleet
under L. Figulus had drawn off into the closed harbour at Corycus; Cassius
Parmensis broke off pursuit to join the main force on Cyprus and link up with
Tillius Cimber's Bithynian squadron under the quaestor Turullius.}

\textit{The decisive intelligence is in \S 4. Dolabella is now besieging
Laodicea, walled in by the disloyal Tarsenes and Laodicenes; ``our Cassius'' ---
C. Cassius Longinus --- is encamped at javelin-cast distance (the Greek dative
\textgreek{πάλτῳ} ``with a javelin's cast'' is what the manuscripts give,
though some editors have suspected a place-name) with ten legions, twenty
auxiliary cohorts, and four thousand horse, and counts on starving Dolabella
out: grain inside the besieged camp is already at three tetrachms a measure. The
allied fleets (Sextilius Rufus's, and the three ships of the writer, Turullius,
and Patiscus) will keep Laodicean supply ships off. The closing line is a
deliberate matching of fronts: as the Republic has been extricated at Rome, so
on this side it will be extricated quickly --- a confidence the autumn was to
betray.}
