% Ad Familiares 10.32 --- headnote
% ad-familiares-10-32, 8 June 43 BC, Corduba

\textit{C. Asinius Pollio to Cicero, written from Corduba
on 8 June 43 BC --- Perseus dateline \textit{Scr. Cordubae
vi Id. Iun. a. 711 (43)}, the date confirmed by the
closing subscript ``\textit{vi Idus Iun. Corduba}''. Pollio,
governor of Hispania Ulterior, has just watched his own
quaestor, L. Cornelius Balbus the younger --- nephew of
the Augustan diplomat of the same name, and the first
non-Italian ever to hold a Roman triumph --- empty the
provincial treasury, refuse to pay the legions out of it,
and decamp on the Kalends of June into the kingdom of
Bogudes of Mauretania. The bulk of the letter is the
catalogue of outrages Balbus had committed in the months
before he fled: a single coiled indictment, mounting
clause by clause from peculation to mock-magistracy to
the burning alive of a Roman citizen in the arena, and
delivered in some of the most caustic prose Pollio left
us. Of all the Pollio dispatches preserved in Book 10
this is by some margin the longest and the most
unsparing.}

\textit{The set-piece is the gladiatorial show at Gades.
Fadius, a former Pompeian soldier pressed into Balbus's
gladiatorial school, had been made to fight twice
without payment; when he refused to take the
\textit{auctoramentum} and threw himself on the
people, Balbus first sent in Gallic cavalry to disperse
the crowd that had begun to throw stones, then had the
man dragged off, buried up to the neck in the sand, and
burnt alive --- pacing about the meanwhile, after a good
lunch, barefoot, tunic loose, hands behind his back, and
answering Fadius's repeated cry ``\textit{civis Romanus
natus sum}'' with the smiling reply, ``Off you go, then,
go appeal to the people's good faith.'' The other
exhibits are scarcely less choice. Balbus had extended
his own quattuorvirate, held the elections of two years
in a single forty-eight-hour stretch (declaring elected
whomever he liked), recalled the exiles of Sextus Varus
the proconsul under Caesar's enemies, and put on as a
festival play a \textit{praetexta} drawn from his own
journey to suborn the proconsul L. Lentulus --- at the
performance of which, Pollio adds with magnificent dryness,
he wept, ``stirred by the memory of his exploits.'' At
the games at Gades he had also presented the actor
Herennius Gallus with a gold ring and seated him in the
equestrian rows, after carving out fourteen such rows for
the purpose. ``With a monster of this kind I have had my
dealings.''}

\textit{Section 4 turns abruptly from the indictment to
Pollio's own position. He has three legions; Antony has
been trying to buy off the Twenty-eighth at five hundred
denarii a head ``on the day it enters camp''; Lepidus has
been pressing him by letter to send up the Thirtieth.
Pollio has held them all. He insists on the strict
construction of his duty --- not a foot outside his
province, not a soldier dispatched, deserters caught and
put to death --- and closes by reminding Cicero, with the
slightly exasperated note of a man who feels the Senate
has misread him, that ``the commonwealth, if it had known
me well enough, would have got a larger return from me.''
The letter to Balbus that Pollio encloses for Cicero's
inspection does not survive; the \textit{praetexta} he
suggests Cicero borrow from Cornelius Gallus has not
survived either. The reference is the first datable
mention in the correspondence of the elegist Gallus, then
a young officer in Pollio's circle.}
