% Ad M. Brutum 2.1 --- headnote
% ad-brutum-02-01, mid-April 43 BC, Rome

\textit{Cicero to M. Junius Brutus, written from Rome ---
Perseus dateline \textit{Scr. Romae ex.\ m.\ Mart.\ ant in.\
Apr.\ a.\ 711 (43)}, ``end of March or beginning of April
43 BC.'' This is broader than the meta date of 13 April,
which has been imposed by editorial convention rather than
by the dateline itself; the letter is in any case
contemporaneous with the most anxious phase of the war
at Mutina, before the battle of Forum Gallorum (14 April).
The opening words \textit{cum haec scribebam, res
existimabatur in extremum adducta discrimen} --- ``while
I was writing these lines, the situation was judged to
have come to its last crisis'' --- and the ``grim letters
and reports about our friend Brutus'' (i.e.\ Decimus
Brutus, besieged at Mutina) place the letter at the
moment when no news of any relief has yet arrived. The
Perseus dateline is broader than the meta date; this is
worth flagging.}

\textit{The letter is short, dignified, and tightly
controlled. Section 1 is Cicero's analysis of the
situation: the armies and commanders are sufficient, but
the consuls' delays --- not their loyalty --- have been
the problem, and the most general truth in his political
toolkit (\textit{quanta momenta sint in re publica
temporum}, ``how much in public affairs hinges on
timing'') is invoked to explain why measures resolved
too slowly amount to no measures at all. Section 2 is
Cicero's account of himself, framed as a defence even
before the charge has been brought: he has rendered to
the commonwealth not only the qualities that anyone may
be asked for --- \textit{fides}, \textit{vigilantia},
\textit{patriae caritas} --- but the \textit{prudentia}
appropriate to one who has taken the helm; if his
counsels prove unprofitable, he claims to deserve as
much censure as if they had been disloyal. The
self-portrait is a man who places himself, with
deliberation, at the centre of the political action and
asks to be judged by that placement. Section 3 turns
to Brutus and Cassius, whose distant armies are the
republican reserve: ``my mind is in the line of battle,''
but \textit{the majority's} eyes look East. The closing
sentence is a kind of commission: either Brutus will
have a better commonwealth to remake, or he will have
to recover the one that is lost --- a sentence the
younger Brutus will still be reading when Cicero is dead
and Mark Antony triumvir.}
