% Ad M. Brutum 1.2 --- headnote
% ad-brutum-01-02, mid-May 43 BC, Rome

\textit{Cicero to M. Brutus, from Rome, late May 43 BC ---
Perseus dateline \textit{Scr. Romae ex. m. Mai. a. 711 (43)}.
The meta entry assigns this to mid-May (15 May) at month
precision; the dateline's \textit{ex.~m.~Mai.} (``at the end
of May'') would place it a week or two later, but within
the same month. The letter is a hurried postscript: Cicero
has already written and sealed a previous dispatch when
Brutus's fresh news arrives. Antony has been defeated at
Mutina (April), Decimus Brutus has burst out from siege,
and the immediate eastern question is what Dolabella ---
who fled the consulship into Asia after murdering Trebonius
in Smyrna --- will do next. The astonishing item: Dolabella
has sent five cohorts across into the Thracian Chersonese,
into the teeth of M. Brutus's five legions, cavalry, and
auxiliaries. Cicero's tone is incredulous, almost amused
(``the bandit has been so far out of his senses'').}

\textit{Section 2 endorses Brutus's caution in not leaving
Apollonia and Dyrrhachium before the news from Italy made
the strategic picture clear, and his subsequent resolve to
move east into the Chersonese. Section 3 is the one that
makes the letter sting: Brutus has reported a mutiny in
the Fourteenth Legion engineered by C. Antonius (Mark
Antony's brother, captured at Apollonia), and Cicero ---
in the polite preface ``you will take this in good part''
--- tells Brutus bluntly that he approves the soldiers'
severity more than Brutus's own. The letter then breaks
off; the Perseus text marks the lacuna ``** *''. The
indulgent handling of C. Antonius is the recurring
political quarrel that will shape Cicero's last letters
to Brutus over the coming weeks.}
