% Ad M. Brutum 1.17 --- headnote
% ad-brutum-01-17, May or June 43 BC, in the East

\textit{M. Brutus to T. Pomponius Atticus --- not to
Cicero --- written from somewhere in Brutus's eastern
command, in May or June 43 BC. The Perseus dateline reads
\textit{Scr. m. Maio, ut videtur, a. 711 (43)}, ``written
in the month of May, so it appears,'' 43. The meta entry's
year-precision placeholder of 6 October must be corrected
to mid-43; the letter has nothing to do with the autumn
proscriptions and predates them by months. It belongs in
\textit{Ad M. Brutum} only because, alone among Brutus's
letters to Atticus, it survived in the same archive as
the Brutus--Cicero correspondence and was inserted into
the book on textual rather than chronological grounds.}

\textit{This is the most politically charged document in
the surviving corpus of Brutus. The occasion is Atticus's
report that Cicero is hurt at Brutus's silence on his
political conduct; Brutus, under Atticus's pressure,
writes out at last what he really thinks. The criticism
is comprehensive. (1) Cicero has irritated the young
Caesar's appetite for power rather than checked it.
(2) Cicero's strategy of cultivating Octavian as the
counterweight to Antony has been not merely a mistake but
a folly: he has stood forth as the avenger of one evil
only to become the founder of another whose roots will go
deeper. (3) The triumph voted to Octavian, the soldiers'
pay he has authorised, the decrees encouraging him not to
``be ashamed of coveting the fortune of the man whose name
he has taken up'' --- this is not the conduct of a
consular, of \textit{Cicero}. (4) Cicero, like all old men
who have property, children, and honours, is afraid of the
three evils Brutus lists with a Stoic philosopher's
unflinchingness: death, exile, poverty. Out of that fear
he has come to think it the worst of all evils to lose
what he has, and so will tolerate \textit{servitus}, if
only it is honourable. (5) The famous taunt of the central
section: ``We do not boast of the Ides of March at every
hour, as Cicero has the Nones of December always on his
tongue,'' contrasts Brutus's tyrannicide and Cicero's
suppression of the Catilinarian conspiracy on equal
historical footing --- and Brutus on the better side of
the comparison.}

\textit{The letter is dated by its political content to
the weeks when news from Italy made the trajectory clear:
Octavian had been refused the consulship in the spring,
the Senate had begun showering him with extraordinary
honours rather than concede the office, and Brutus had
begun to see what that combination of refusal and
flattery would yield. The closing section, with its
solicitous question about Atticus's daughter Attica's
betrothal and its concern for Brutus's wife Porcia, gives
the personal frame: Atticus and Brutus are the closest of
friends, and the letter is an attempt to enlist Atticus
in the cause of restraining Cicero. The attempt failed.
Within a year, Brutus's prediction --- that Cicero would
fall under the proscriptions of the master he had helped
to make --- came true. The text in section 6 carries an
unresolved crux at ``\textit{Antonius},'' here preserved.}
