% Ad Familiares 10.35 --- headnote
% ad-familiares-10-35, 30 May 43 BC, camp at Pons Argenteus

\textit{M. Aemilius Lepidus to the praetors, tribunes of
the plebs, the Senate, and the People and Plebs of Rome,
written from Pons Argenteus on 30 May 43 BC --- Perseus
dateline \textit{Scr. in Ponte Argenteo iii K. Iun. a.
711 (43)}, the date confirmed by the subscript ``D.~iii
Kal.~Jun.~a Ponte Argenteo.'' This is the dispatch by
which Lepidus, governor of Gallia Narbonensis and
Hispania Citerior, formally announced to the Senate
that he had merged his army with Mark Antony's. He
writes as \textit{imperator iterum} and \textit{pontifex
maximus}, and addresses the full constitutional ladder
of Rome --- the magistrates, the Senate, and the
populus and plebs collectively. The letter is two
sections long. By the end of June he had been declared
a public enemy.}

\textit{The argument, stripped of its official Latin,
is that Lepidus did not defect: his soldiers did.
``My whole army, an uprising having broken out, kept
up its accustomed practice in preserving fellow-
citizens and the common peace, and compelled me to
take up the cause of the safety and survival of so
great a multitude of Roman citizens.'' This is the
classic exculpatory formula of a Roman commander who
has lost control of his men --- \textit{seditione
facta} (``an uprising having broken out''),
\textit{coegit} (``compelled me''). The fact that
Lepidus's army had been fraternizing across the lines
with Antony's veterans for weeks, that Lepidus had
done nothing to stop it, that twelve days earlier he
had written to Cicero in tones of bland competence
(see 10.34) reporting Antony's collapsing position,
and that the dispatch was sent only after his junction
with Antony was a fait accompli --- all of this the
Senate read between the lines on the day the letter
arrived. The meaning of the document, in the way of
such documents, sits in what cannot be said.}

\textit{The historical weight of the letter is hard
to overstate. With Lepidus's seven legions and
Antony's surviving forces combined, the strategic
picture of the war was inverted in a stroke: the
twelve legions that Plancus and Decimus Brutus could
field in Gaul were now matched, and Octavian --- whose
march north Plancus had still been hoping for in late
July (see 10.24) --- had no reason left to march. On
30 June the Senate, on Cicero's motion, declared
Lepidus a \textit{hostis publicus}. In November
Lepidus, Antony, and Octavian met on the island near
Bononia and constituted themselves as the
triumvirate \textit{rei publicae constituendae}. The
proscription lists were drawn up. Cicero's name was
on them within weeks; on 7 December he was killed at
Formiae. The phrase \textit{nostram humanitatem in
civili dissensione sceleris loco ponatis} (``[do not]
count our pity in a civil disagreement as a crime'')
--- the central plea of section 2 --- is the
self-justification of a man who knows he has, in
fact, committed a crime, and whose only defence is to
ask the Senate to redescribe it.}
