% Ad Familiares 10.12 --- headnote
% ad-familiares-10-12, 11 April 43 BC, Rome

\textit{Cicero to L. Munatius Plancus, written from Rome on 11
April 43 BC --- Perseus dateline \textit{Scr. Romae iii Id. Apr.
a. 711 (43)}. This is one of the great letters of the spring of
43 BC and one of Cicero's happiest. Plancus's despatch
declaring his loyalty to the republican cause has reached Rome,
been read aloud in a packed Senate, and --- after a two-day
struggle in which the urban praetor Cornutus had to defer
business because the auspices had been fumbled, and the
tribune P. Titius interposed his veto on behalf of Servilius
Isauricus --- has carried a vote of public honours moved by
Cicero. The letter is the account of that triumph, sent back
to its author in Gaul.}

\textit{The narrative is dated to the hour: Varisidius delivers
Plancus's letter on the morning of 7 April; a crowd of
\textit{optimorum virorum et civium} escorts Cicero from his
house and is made to share the joy; Munatius arrives and is
shown both the private letter and the public dispatch; the
party goes to Cornutus, then in charge of consular business
\textit{more maiorum} (the consuls Hirtius and Pansa being in
the field at Mutina); the Senate is summoned, then deferred on
auspicial grounds, with the College of Augurs --- Cicero's own
college --- confirming the religious objection. On the second
day Cicero wrestles Servilius's prior motion off the floor;
on the third he breaks both Servilius and the tribunician
veto. Plancus is to read the details in other correspondents'
letters; from Cicero he gets the one line that matters: the
Senate could not have been ``graver, steadier, friendlier to
your praises,'' and ``the whole community of the city'' was
more loyal still.}

\textit{The closing section is the moral peroration. Plancus is
to press on, to despise the \textit{inanissimis splendoris
insignibus} ``the most empty trappings of show'' --- those
short-lived, painted, fragile things that look like glory ---
and to fix on \textit{verum decus}, true honour, which is set in
virtue and is most brilliantly displayed by great services to
the state. Cicero will be not merely the supporter of Plancus's
standing but the magnifier of it. The credit Cicero gives to T.
Munatius (Plancus's brother, sometimes called Bursa, here ranged
firmly on the right side) is a courtesy and a piece of
political housekeeping: the letter that makes Plancus's career
also names the friend who has been working the chambers for him
in Rome.}
