% Ad Familiares 10.26 --- headnote
% ad-familiares-10-26, late June 43 BC, Rome

\textit{Cicero to C. Furnius, from Rome between 24 and 29
June 43 BC --- Perseus dateline \textit{Scr. Romae inter
viii et iii K. Quint. a. 711 (43)}. This is the second
of the two surviving letters from Cicero to Furnius
(10.25 preceded it by a month), and the second-last
letter of the Plancus dossier. By the time of writing,
Plancus has been celebrated by the Senate, his
$\textit{supplicationes}$ are voted, his concord with
Decimus Brutus is established, and the army of
Narbonensis has been preserved from a costly engagement
--- the news that opens the letter.}

\textit{The body of the letter, however, is a sharp
rebuke. Furnius has written again about the praetor's
elections (the same subject as 10.25): if they fall in
Sextilis he will come down quickly to stand; if they
are already over, more quickly still ``so as not to
remain a fool any longer at the hazard of his life.''
Cicero reads the line for what it is --- a man at his
post pretending to himself that it is the post that
makes him foolish --- and answers with a structured
indignation that runs the length of section 2:
exclamation, rhetorical questions in tricolon, the
contemptuous flattening of the praetorship to ``the
cheapest and the most common'' of magistracies if won
at the run of common men. The closing offer in section
3 is Cicero at his most practical: ``the elections,
since you are hanging on them, we are pushing back into
the month of January, so far as we can'' --- standing
and circumstance both being served at once. ``Win,
then, and farewell'' (\textit{vince igitur et vale})
collapses the two valedictions, the political and the
epistolary, into a single line.}
