% Ad Familiares 10.30 --- headnote
% ad-familiares-10-30, 15 April 43 BC, camp at Mutina

\textit{Servius Sulpicius Galba to Cicero, written from camp
at Mutina on 15 April 43 BC --- Perseus dateline \textit{Scr.
in castris ad Mutinam a. d. xvii K. Mai. a. 711 (43)}. Galba
is one of the conspirators of the Ides (the great-grandfather
of the future emperor) and at this date a senior officer with
the consul Pansa's army. He is writing the morning after the
battle of Forum Gallorum (14 April 43), the first of the two
engagements --- with the larger battle of Mutina a week later
on 21 April --- that ended Antony's siege of Decimus Brutus
and broke his army. It is one of the very few first-person
eyewitness accounts of a Roman battle anywhere in the
surviving Latin corpus, and the only one written for
publication purposes none, by a senior officer to a
political principal at Rome who needed an exact picture of
what had happened.}

\textit{The register is military and reportorial, not
Ciceronian. The sentences are short, the chronology is
strict, the technical detail is unflinching --- legions and
cohorts named by number (the Second, the Thirty-Fifth, the
Martian), the praetorian cohorts of Antony, of Silanus, and
of Caesar (Octavian) tracked separately, the topography of
marsh, woods, and the Aemilian Way given as a soldier would
need them. Galba commands the eight cohorts of his old
Martian Legion on the right wing of Pansa's force; he routs
the Thirty-Fifth at first onset, finds himself overshot
beyond his own line, has to wheel back to deal with
flanking Moorish cavalry, and in section 3 finds himself
suddenly behind enemy lines --- the celebrated moment in
which he charges back toward his own raw recruits with his
shield flung behind him, and is saved ``by I know not what
fate'' because his own men recognize him before they throw
their javelins. The understatement is the point: the senior
commander writing in soldier's prose about almost being
killed by his own legion.}

\textit{Section 4 records what would prove decisive ---
Hirtius's arrival with twenty veteran cohorts after Antony,
thinking himself victorious, had pursued Pansa's broken left
wing back to the camp. Hirtius cut Antony's veterans to
pieces at the very ground of the morning's battle, and Antony
got back to Mutina with only the cavalry, at the fourth hour
of the night. The losses on the senatorial side are owned in
the same plain register: praetorian cohorts gone, the
Martian Legion bled, two eagles and sixty standards taken.
The closing \textit{res bene gesta est} --- ``the business
has been well managed'' --- is the official communiqué's
formula, the senatorial reply to which would be the granting
of a public thanksgiving. Pansa was carried from the field
seriously wounded and died of those wounds about a week
later; Hirtius himself fell at Mutina on 21 April; both
consuls dead in a week, the army intact in the field but
ownerless, was the structural opening that Octavian would
exploit through the summer.}
