% Ad Familiares 10.31 --- headnote
% ad-familiares-10-31, 16 March 43 BC, Corduba

\textit{C. Asinius Pollio to Cicero, written from Corduba on
16 March 43 BC --- Perseus dateline \textit{Scr. Cordubae
xvii K. Apr. a. 711 (43)}. This is Pollio's first surviving
letter to Cicero. He is governor of Hispania Ulterior, with
one veteran legion (the Thirtieth) and an army that includes
recent recruits; between him and Italy stand the Alps, the
army and province of M. Aemilius Lepidus in Narbonese Gaul,
and Lepidus's open declaration that he is in agreement with
Antony. The letter is written the day after the first
anniversary of Caesar's assassination and addresses a man
who has known the writer only by reputation; Pollio is
positioning himself, carefully and on the record, in the
moment before he has any real news of Mutina.}

\textit{The prose is Pollio's own --- the prose of the man
whose ``Pollionian'' style ancient critics set against
Cicero's middle style as a denser, more clipped, more
self-consciously austere idiom. The sentences are short
where Cicero's would build; the political ironies are tight
(``the man whom, though no one wants to look at him, men
nonetheless do not hate as he deserves'' for Antony in
section 2); the self-justifications are placed exactly. He
is at pains to construct a record: that he was forced into
the Caesarian camp by having enemies on both sides; that he
did under orders only what he could not avoid, and visibly
against his will; that he declared in his provincial
assembly at Corduba that he would deliver the province to
no one but a senatorial commissioner; that he resisted the
demand to surrender the Thirtieth Legion, with reasons; that
his nature and his studies (\textit{natura autem mea et
studia} --- the literary man's defence) draw him to peace
and to liberty.}

\textit{The crucial admission is section 4: the consuls had
sent him no instructions, only Pansa's single private letter
after the Ides urging him to declare for the Senate. With
Lepidus haranguing in public that he was with Antony, the
roads searched and the couriers detained, Pollio could not
march into Italy through Lepidus's province even if he wished
to. The closing decision in section 6 is therefore taken on
his own initiative and in the abstract: that, since legions
are now wanted more than provinces, he has resolved to set
out with the army. Within months Pollio would in fact, like
Lepidus and Plancus, throw in with Antony rather than with
the Senate; the letter is the most careful contemporary
record of how a Caesarian provincial governor of literary
ambitions positioned himself in the spring of 43, before
events forced his hand.}
