% Ad Familiares 9.14 --- headnote
% ad-familiares-09-14, 3 May 44 BC, Pompeii

\textit{Cicero to P. Cornelius Dolabella, then suffect consul,
written from the villa at Pompeii on 3 May 44 BC (Perseus:
\textit{v Non. Mai.}). The salutation gives the recipient explicitly:
\textit{Cicero Dolabellae consuli suo}. The letter celebrates
Dolabella's violent suppression of the spontaneous cult of Caesar
that had grown up at the so-called pseudo-Marius's altar in the
Forum in the weeks after the assassination, an action carried out
in late April. Dolabella had been Cicero's son-in-law (the marriage
to Tullia had ended before her death in early 45) and remained a
warmly addressed correspondent; here the warmth is the warmer for
the political opening Cicero sees in the consul's stance.}

\textit{The letter weaves together private affection and public
exhortation. Cicero quotes the dying Lucius Caesar's praise of
Dolabella as ``the one man whom, since your consulship, we can
truly call a consul'' --- the older Caesarian moderate transferring
to Caesar's son-in-law the title he denies his own nephew Antony.
The Homeric figure of Agamemnon and Nestor is offered as cover for
the open vanity of the wish that Dolabella should pass as Cicero's
pupil. The political turn at the end --- ``the Republic has been
placed in your hands,'' and the urging to honor ``those men through
whom liberty's beginning was made'' (the assassins) --- is among
the most committed Republican statements of these volatile weeks
before Cicero finally despaired of Dolabella. The final image,
\textit{custodias} (``guard yourself''), strikes the note that
will preoccupy the correspondence of the coming months.}
