% Ad Atticum 14.17 --- headnote
% ad-atticum-14-17, 4 May 44 BC, in Pompeiano

\textit{Cicero to Atticus, written at the Pompeian villa on 4
May 44 BC --- Perseus dateline \textit{Scr. in Pompeiano iv Non.
Mai. a. 710 (44)}. The day after sending 14.16 from Puteoli,
Cicero has crossed to Pompeii (having stopped at the Cumean to
install Pilia the previous afternoon). Atticus's most recent
letter, by the freedman Demetrius, has caught up with him at
dinner: many wise things, but resting in the end on \textit{omne
consilium in fortuna positum} --- ``every plan depends on
fortune.''}

\textit{Three pieces of news dominate. First, Buthrotum: Cicero
still wants a meeting with Antony, but Antony is reported to be
sticking at Capua, raising veterans; the dying Lucius Caesar
(Antony's uncle, seen the day before at Naples) takes the same
view, that this stay at Capua bodes ill. Second, a domestic
crisis in the Quintus family: young Quintus has written his
father a savage letter announcing he will not tolerate his
stepmother Aquilia and --- worse --- that he had had everything
from Caesar, nothing from his father, and now hopes for the rest
from Antony. Cicero's verdict is single-word: \textit{o perditum
hominem!} The Greek tag \textit{mel\=esei} (``it will be seen
to'') shrugs the matter forward. Third, in a striking aside,
Cicero says that to speak against ``that wicked party'' was
safer with the tyrant alive than now that he is dead --- because
every move is now ``called back to the acts, even the thoughts,
of Caesar.'' The unfinished \textit{anekdoton} (his unpublished
book, probably the \textit{De Gloria} or an earlier draft of
material to be reworked) is also mentioned.}
