% Ad Atticum 14.7 --- headnote
% ad-atticum-14-07, 15 April 44 BC, in Formiano

\textit{Cicero to Atticus, written at the Formian villa on 15
April 44 BC --- Perseus dateline \textit{Scr. in Formiano xvii
K. Mai. a. 710 (44)}, the body confirming \textit{e Formiano
exiens xvii Kal.} as he prepares to leave for Puteoli. A month
after the Ides, Cicero is still on the road south. He has just
seen L. Aemilius Paulus at Caieta and picked up further bad news
about ``Marius'' --- the impostor Herophilus who had been
trading on a claimed descent from the great Marius; Antony,
having earlier protected him, has just had him killed without
trial (cf. \textit{Phil.} 1.5). Brutus, banished in effect from
the city, is seen at Lanuvium, and no one yet knows where he
will settle.}

\textit{The second section turns to the domestic anxiety that
threads through these weeks: the welfare of young Marcus,
studying in Athens, whose latest letter pleases Cicero by its
\textit{pinos} --- the patina of stylistic finish. Three Greek
phrases cluster on the same metaphor (\textit{pepinomenai},
\textit{pinos}), then the closing instruction to Atticus uses a
classic Latin idiom for casual writing --- \textit{quod in
buccam venerit}, ``whatever first comes to mind'' (the image of
words rising onto the tongue is lost in English but the function
is preserved, per STYLE.md). The letter closes with the public
plan that floats through this whole April: a quiet departure
\textit{mense Quintili in Graeciam}, in July, by sea ---
contingent on a political settlement that does not come.}
