% Ad Atticum 14.3 --- headnote
% ad-atticum-14-03, 9 April 44 BC, Tusculum

\textit{Cicero to Atticus, written on 9 April 44 BC at the
Tusculan villa --- Perseus dateline \textit{Scr. in Tusculano v
Id. Apr. a. 710 (44)}. Atticus's last letter has been
``tranquil''; Cicero half-prays it stays so, but Matius's verdict
from two days earlier is still in his ear. Then the day's
incident: his estate's builders, sent into the city for grain,
have come back empty-handed with a rumour that the whole supply
is being carried off to Antony's house. Cicero diagnoses the
rumour as a piece of panic --- a Pan-fright (\textit{panikon}) ---
and brushes it aside; if it were real, Atticus would have
written. The vignette catches the texture of these days: country
seclusion, urban hearsay, rapid sceptical adjudication.}

\textit{The second section asks Atticus to take Antony's measure
--- \textit{[Greek: diathesin]}, his disposition, his frame of
mind --- and offers Cicero's own first sketch of him: more
absorbed in his dinner parties than in serious mischief, an
estimate which the summer will keep revising. The closing is the
running joke of the post-Ides correspondence: substantive news
if Atticus has any (\textit{pragmatikon}), and if not, the
\textit{[Greek: episēmasian]} of the crowd and the jokes of the
mimes --- the temper of the theatre, in lieu of the temper of
the senate. Greek phrases here are deployed for that
half-affectionate, half-conspiratorial register the two men
reach for when politics is exactly the topic.}
