% Ad Atticum 15.17 --- headnote
% ad-atticum-15-17, 14 June 44 BC, in Antiati

\textit{Cicero to Atticus, written at Antium on 14 June 44 BC ---
Perseus dateline \textit{Scr. in Antiati postr. Id. Iun. a. 710
(44)}, i.e. the day after the Ides of June. Cicero is replying
to two letters from Atticus that arrived together, one written
on the Ides and one the day before, and he works through them
in turn. The first section gathers up business: Decimus Brutus,
one of the assassins, is now in his Gallic command and Cicero
wants whatever news Atticus can supply; the consuls Antony and
Dolabella have been putting about an alarmist story which Sicca
has already passed on; a Greek-tagged half-quotation
(\textit{ta men didomena}, ``the things being granted ---'')
hangs unfinished, and the sequel \dag Siregius\dag\ is a
manuscript ruin. The practical questions concern who is to be
trusted with what intelligence: Lucius Antonius (Mark Antony's
brother) is to be diverted through Marcus rather than approached
directly, Antronius is not to be used at all, and only the
aedile Lucius Fadius is to be informed of the underlying
arrangement.}

\textit{The second section turns to family and household. Cicero
has not failed Servilia --- mother of Marcus Brutus --- which
he warmly approves; he is no longer fretting over ``the queen''
(Cleopatra, then still in Rome, or just leaving it); the
financial accounts kept by Eros, his land-agent, are being
straightened out; and his son Cicero in Athens has sent a
letter so affectionate \textit{[philostorg\=os]} and so neatly
finished \textit{[eupin\=os]} that the father says he would
read it out in public. The closing question of a possible move
to the Tusculan villa is acted on the next day: Letter 15.18 is
written from the road.}
