% Ad Atticum 15.1 --- headnote
% ad-atticum-15-01, 17 May 44 BC, in Puteolano

\textit{Cicero to Atticus, written at the Puteolan villa on 17
May 44 BC --- Perseus dateline \textit{Scr. in Puteolano xvi K.
Iun. a. 710 (44)}. The letter opens with a cry of grief over
the sudden death of Alexio, Cicero's physician and friend ---
``what a wretched business about Alexio!'' --- carrying the
familiar Stoic consolation that we are born on terms which
permit us to refuse nothing that can befall a man. The rest is
political reconnaissance from Campania on the eve of his
departure inland. Antony has slipped past Misenum unseen; but
Hirtius, who happened to be at Puteoli reading Atticus's
letter with him, has been pumped both on Antony's intentions
and on the Caesarian camp's mood.}

\textit{The closing section is a knot of domestic and political
threads characteristic of these letters: young Quintus,
Caerellia, an unnamed lady offended at a kindness; the
Sophoclean line \textit{quid est autem cur ego personatus
ambulem?} (``why am I to walk about in a mask?'') glancing at
the indignity of dissembling in old age; Brutus's invitation to
Rome before the Kalends, which Cicero may accept but finds
opaque (``he has consulted his own immortality better than our
peace''); the queen --- Cleopatra, still in or recently
departed from Rome --- whose rumour is dying down; and Flamma,
on whose behalf Atticus is being pressed. Greek surfaces three
times --- \textit{[Greek: ouden]} (``nothing,'' Hirtius's
verdict on peace prospects), \textit{[Greek: to ek toutou]}
(``the consequence''), and a half-line of dramatic
verse --- each a marker of intimacy in the running shorthand
of these two old friends.}
