% Ad Atticum 14.21 --- headnote
% ad-atticum-14-21, 11 May 44 BC, in Puteolano

\textit{Cicero to Atticus, written at the Puteolan villa later
on the same day as 14.20 --- Perseus dateline \textit{Scr. in
Puteolano v Id. Mai. a. 710 (44)}. The manuscripts preserve
both letters of 11 May; Cicero's own opening sentence
(``when I had just a little while ago handed a letter to
Cassius's courier'') anchors this one as the second. His own
slave-courier has arrived without a letter from Atticus
(``like a portent''), but bearing a reply from Dolabella ---
not yet on the dowry, but answering an earlier letter handsomely
enough.}

\textit{Then Balbus comes to the door, and Cicero's account
of him sharpens into one of the bitter set pieces of the
post-Ides correspondence. Balbus is too guarded to be read,
but a man who fears peace is unmistakable; his whole speech
is a covert apology for Antony. Cicero diagnoses the
underlying error in plain Latin and Greek both: the
Liberators acted with the spirit of men (\textit{animo
virili}) and the planning of children (\textit{consilio
puerili}) --- they left the heir of the kingship alive (Antony
himself, here glanced at), feared the wrong man, and let the
right one breathe. The Greek phrase \textit{[Greek:
huposolokia]} (``quite a bit barbarous'') marks the
political solecisms multiplying around him --- including the
fact that the house of Pontius at Naples is now occupied by
the mother of Caesar's assassin. Cicero recommends himself a
re-reading of his own \textit{Cato Maior}, since old age is
making him sour, and signs off with \textit{[Greek:
bebi\=otai]} --- ``my life has been lived; let the young see
to it.'' The closing sketch --- the letter dictated over
dessert at Vestorius's, dinner tomorrow at Hirtius's, a
five-course meal (\textit{[Greek: pentelopion]}), and the
joke about ``looking for our travelling shoes'' (\textit{
talaria}, Mercury's winged sandals) rather than enlisting in
any camp --- has the speed of the bay-of-Naples letters at
their best.}
