% Ad Atticum 15.4 --- headnote
% ad-atticum-15-04, 24 May 44 BC, in Arpinati

\textit{Cicero to Atticus, written at Arpinum on 24 May 44 BC
--- Perseus dateline \textit{Scr. in Arpinati ix K. Iun. a. 710
(44)}. (The launch prompt's ``Astura'' for this letter is
superseded by the Perseus dateline; Cicero is still inland at
Arpinum, planning to reach Tusculum on the sixth before the
Kalends, 27 May.) The letter opens at the tenth hour of the
day with a courier from Q.~Fufius bearing ``some little scrap
of a note'' that Cicero finds silly --- ``unless, perhaps,
everything that you do not love seems silly'' --- and two of
Atticus's letters, the more recent of which Cicero answers
first. Antony's plans look turbulent, and his stripping of
Cisalpine Gaul from Decimus Brutus seems to Cicero to aim at
war; he would not desire it if the Buthrotian business were
not already being seen to, and even so he grudges that it is
not being settled through his own offices.}

\textit{Sections 2 and 3 deliver the most explicit retraction
of the Ides of March in the whole correspondence. ``We used
manly spirits and, believe me, boyish counsels. The tree was
cut down, not torn up by the roots'' --- the conspirators
killed the man without removing the system, and now ``you see
how it sprouts again.'' Atticus has pressed Cicero to write a
companion piece to Brutus's recently published Capitoline
speech; Cicero refuses for the third time and at length, calls
the request a \textit{[Greek: parencheir\=esis]} (``interference,''
literally ``putting in one's hand among another's work''), but
will consider an independent \textit{[Greek: H\=erakleideion]}
(a dialogue in the manner of Heraclides Ponticus) when the
argument and the occasion are riper. The famous admission
follows: ``the Ides of March do not delight me'' --- for had
Caesar lived, fear would not have driven the Senate to confirm
his acts, and Cicero, who stood high in his favour, would not
now be unfree under the heirs. ``I blush, believe me; but I had
already written it, and I would not delete.'' Other Greek
markers: \textit{[Greek: an\=o]} (``up in the air,'' of news
involving Carfulenus), \textit{[Greek: aporia]} (the perplexity
about what the Liberators should do next), and \textit{[Greek:
mel\=esei]} (``it will be my care''). The closing precaution ---
that he sent the letter by his own courier rather than Fufius's
agent, lest the latter break the seal --- registers the level
of suspicion that May 44 has now reached.}
