% Ad Familiares 16.21 --- headnote
% ad-familiares-16-21, c. 15 July 44 BC, Athens

\textit{Marcus Cicero the younger from Athens to Tiro,
mid-summer 44 BC, per the Perseus dateline \textit{inter ex. m.
Quint. et ex. Oct. a.~710 (44)} --- between late Quintilis and
late October, here taken at the early end of that window. The
sender is the orator's son, then in his early twenties and
studying philosophy at Athens; the salutation \textit{CICERO F.
TIRONI SVO DVLCISSIMO S.} marks him as \textit{filius} (the son),
not the great Cicero. The letter is by some distance the longest
extant piece in young Marcus's hand, and gives the fullest
surviving picture of his Athens year.}

\textit{The letter reports on the son's studies in eight short
sections: he is intimate with the philosopher Cratippus, lives
next door to the rhetor Bruttius (whose support he is helping out
of his own narrow means), declaims daily in Greek with Cassius
and in Latin with Bruttius, and keeps company with the circle
Cratippus brought from Mytilene and with Epicrates and Leonides
at Athens. He has dismissed his old teacher Gorgias on his
father's explicit instructions. He also responds to Tiro's news
of a country purchase, and asks for a Greek copyist to relieve
him of transcribing his philosophical notebooks.}

\textit{The voice is studious-anxious and a little performatively
grown-up --- the young Cicero writing to his father's freedman
with a slightly theatrical seriousness about his progress, his
``errors of youth'' now firmly behind him, and his desire that
Tiro be ``trumpeter of his reputation.'' The letter is studded
with Greek phrases (\textit{philologia}, \textit{suz\=et\=esis},
\textit{diarrh\=ed\=en}, \textit{hupomn\=emata},
\textit{sumphilologein}) in the manner of a student steeped in
Greek lectures. The closing commendation of Anteros (presumably a
slave or freedman) is the warm Tironian sign-off he has learned
from his father.}
