% Ad Atticum 15.20 --- headnote
% ad-atticum-15-20, 21 June 44 BC, in Tusculano

\textit{Cicero to Atticus, written at the Tusculan villa about
21 June 44 BC --- Perseus dateline \textit{Scr. in Tusculano
inter xv et xi K. Quint. a. 710 (44)}, the same window as 15.19,
and probably the same day or the day after. This is one of the
hardest letters of the year. Cicero has decided to leave Italy
under cover of a legation, and the decision has crystallised
into bleakness. He has been losing faith in Rome since the day
he heard ``that tyrant'' --- Caesar --- publicly called ``the
most illustrious of men''; and when he saw, with Atticus at
Lanuvium, that the assassins held only as much hope of
remaining alive as Antony chose to grant them, he despaired.
Departure now is, he confesses bluntly, an act of despair, not
of embassy. He is leaving the lobster-pot not in flight but in
the hope of a better death than the one Antony seems to
intend. The whole of the blame, he says, is Marcus Brutus's
--- for sparing Antony at the Ides, and for handing the state
to him in the days that followed.}

\textit{The letter then moves to logistics. Pompeius the
younger has been received at Carteia in Spain: there is again
an army against the Caesarian regime. Which camp, then ---
that one, or Brutus's and Cassius's? Antony has crushed any
middle position. Cicero asks Atticus's advice on the practical
question (Brundisium or Puteoli for embarkation) and turns to
money: a loan of 200,000 sesterces will have to be raised for
five months to cover the interval until Quintus's debts fall
in. The final section he writes out in his own hand --- a
small mark of secrecy and weight. The closing instruction is
to write back today.}
