% Ad Atticum 16.5 --- headnote
% ad-atticum-16-05, 9 July 44 BC, Puteolanum

\textit{Cicero to Atticus, written from the Puteolan villa on
9 July 44 BC --- Perseus dateline \textit{Scr. in Puteolano
vii Id. Quint. a. 710 (44)}. The letter is from the day before
16.4; the manuscript ordering does not match the chronology.
Brutus is staging the Apollinarian games \textit{in absentia}
(he holds the urban praetorship but cannot return to Rome),
and Cicero is south on the bay of Naples, waiting on news and
on a convoy.}

\textit{The heart of the letter is the central section on
young Quintus, Cicero's nephew. After years of unsteadiness
and disloyalty --- he had taken Caesar's part against his
elders --- the young man has spent several days with Cicero
and emerged, in Cicero's account, wholly changed: changed by
``certain writings of mine I had to hand'' (almost certainly
the recently-finished \textit{De Officiis} drafts), by daily
conversation, and by precept. Cicero has stood up for him to
Brutus, who took the boy on faith, refused a guarantor,
embraced him, and dismissed him with a kiss. Cicero now asks
Atticus to credit this reformation and to give it the weight
of his authority. The transformation will not, in the event,
hold: by autumn the boy will be back to his old courses.}

\textit{Around this core are the lighter pieces: Accius's
\textit{Tereus} at the games (Brutus had supposed it was a
play by his ancestor, the Brutus who expelled the kings ---
the joke turns on the name); the question of a joint voyage
with Brutus (the Greek \textit{homoploiai}, ``sailing in
company''), which Brutus is in no hurry to commit to; the
asides of plotted alternative ports --- Venusia, Hydruntum,
back here again --- ``I will die if anyone but you holds me
here'' --- and the closing literary banter with Atticus about
the collection of Cicero's letters, of which Tiro already has
about seventy and the rest are to be retrieved from Atticus
and corrected by Cicero himself before publication.}
