% Ad Atticum 15.21 --- headnote
% ad-atticum-15-21, 22 June 44 BC, in Tusculano

\textit{Cicero to Atticus, written at the Tusculan villa on
22 June 44 BC --- Perseus dateline \textit{Scr. in Tusculano x
K. Quint. a. 710 (44)}. The opening is acid family comedy.
Quintus the elder is exulting over a letter from his son ---
who has resurfaced after the recent rupture --- in which the
young man claims he wanted to flee to Brutus precisely because
he had refused Antony's commission to help make him dictator
and seize a garrison; that he refused only to spare his
father's feelings; that this is why Antony has turned hostile;
that he ``pulled himself together'' and patched things up
again. Statius, the elder Quintus's freedman, adds that young
Quintus now wants to live with his father. Cicero's
exclamation --- ``did you ever know a more thorough fraud than
that fellow?'' --- speaks for itself. The two Quinti have been
running through a long series of these performances all year.}

\textit{The second section turns to routine matters. Atticus
has rightly suspended decision on a piece of business
involving one Canus; Cicero approves the holdup. He had not
yet been alerted to a problem about some accounts and had
assumed everything restored without compromise. The courier
schedule is in Atticus's hands; he is busy. Xeno's affair is
in good order. Tiro reports that Atticus now disfavours
Brundisium as an embarkation point and is raising some
concern about the soldiers there; Cicero had already settled
on Hydruntum (Otranto), put off by Atticus's earlier remark
about the five-hour crossing from Brundisium --- though, he
notes, the alternative route is itself a serious voyage. He
is pressing on, so as to leave before Sextus Pompeius
arrives.}
