% Ad Atticum 15.10 --- headnote
% ad-atticum-15-10, 5 June 44 BC, in Tusculano

\textit{Cicero to Atticus, written at the Tusculan villa on 5
June 44 BC --- Perseus dateline \textit{Scr. in Tusculano Non.
Iun. aut postridie a. 710 (44)}. The dateline allows the Nones
themselves or the day after; we follow Perseus's first option.
Atticus has now forwarded an affectionate letter from Marcus
Brutus, the chief Liberator, and Cicero answers in a single
breathless section that turns each of Brutus's options over and
finds every one unsatisfactory. To accept the grain commission
from Antony and his circle is shameful. To attempt something
fresh --- another stroke against the Caesarians --- is
impossible: the moment has passed, the men neither dare nor can
act. To stay quiet on the conspirators' own advice gives no
guarantee of safety, especially if news arrives that Decimus
Brutus, then in Cisalpine Gaul, has come to grief.}

\textit{Particularly galling to Cicero is the choice between
two indignities: to miss the praetorian games Brutus is bound
by office to put on, or to take the grain errand --- which he
compares to the legate's job that Pompey gave the Academic
philosopher Dio of Alexandria, a charge proverbially humiliating
for a man of dignity. Cicero is reluctant to give written
counsel in any case, since giving counsel in such a matter is
itself unsafe; reluctant, too, because Brutus is taking his
direction from his mother Servilia, whose entreaties are not
counsel that another voice can easily interpose itself against.
Yet silence Cicero cannot manage. He will think over how to
write, and send a courier at once to Brutus, who is at Antium
or perhaps already at Circeii.}
