% Ad Atticum 15.3 --- headnote
% ad-atticum-15-03, 22 May 44 BC, in Arpinati

\textit{Cicero to Atticus, written at Arpinum on 22 May 44 BC ---
Perseus dateline \textit{Scr. in Arpinati xt K. Iun. a. 710
(44)}, where the corrupt ``xt'' is conventionally read as
\textit{xi K. Iun.}, the eleventh before the Kalends of June.
(The \texttt{works.yaml} entry places the letter on 1 June and
the launch prompt at Astura; both are superseded by the
Perseus dateline as restored.) On reaching Arpinum Cicero finds
two of Atticus's letters waiting, and answers the earlier of
them first. Atticus, planning to be at Tusculum when Cicero
arrives there on the sixth before the Kalends, has urged him
that ``one must obey the victors''; Cicero refuses the
parallel his friend has drawn with the temple-of-Apollo
negotiations under the consulship of Lentulus and Marcellus,
since the situation is not comparable --- especially now that
Marcellus and others are quitting Italy.}

\textit{The second section turns to the later letter: Alexio's
posthumous arrangements, Hirtius's loyalty, a barbed wish for
the worse against Antony (``since he is what he is, I want it
the worse for him''), young Quintus, and --- at length ---
Brutus's recently published speech on the Capitoline. Atticus
has asked Cicero to write a version of it as if delivered by
Brutus, and Cicero declines with mounting impatience: it
cannot be reconstructed in retrospect from outside, and he is
not yet ready to write ``of a tyrant most justly slain.'' That
will come, ``in another manner and at another time.'' The
letter ends in the rapid telegraphic register of these
exchanges --- the tribunes' suppression of the gilded chair
voted to Caesar, the fourteen rows of seats reserved at the
games --- and with the affectionate hope that Brutus stayed at
Tusculum willingly and long enough.}
