% Ad Atticum 16.7 --- headnote
% ad-atticum-16-07, 19 August 44 BC, navigans ad Pompeianum

\textit{Cicero to Atticus, written at sea on the run up the
Italian coast toward the Pompeianum on 19 August 44 BC ---
Perseus dateline \textit{Scr. navigans ad Pompeianum xiv K.
Sept. a. 710 (44)}. This is the turnabout letter. Having
sailed south for Greece in late July (the journey he was
embarking on at the end of 16.3), Cicero had got as far as
Leucopetra, the southern tip of Bruttium, and on 6 August had
put out across the strait toward Patrae; some three hundred
stadia out, a violent south wind drove him back to the same
Leucopetra. The villa was Valerius's, and while he waited for
his wind, fresh visitors arrived from Rome carrying the
edict of Brutus and Cassius, the news of a full senate
summoned for the Kalends, and the report that Cicero himself
was missed and half-blamed for his absence. The plan to leave
Italy, in which he says he had taken no pleasure even before,
was thrown aside on the spot.}

\textit{The bulk of the letter is a long, surprisingly raw
reply to a letter from Atticus that had reached him only
after he turned back --- a letter in which Atticus, having
earlier urged the departure, now rebuked it in Greek tags
(\textit{euthanasian}, \textit{scholion}, the famously
wounding \textit{Brutus noster silet}, ``our Brutus is
silent''). Cicero half-defends the original decision, half
welcomes its reversal, and in the central paragraph
recounts the meeting that ratified everything: at Velia, on
17 August, Brutus heard he was there and came on foot from
his ships at the river Heles three miles off, ``and poured
out all that he had kept silent before.'' Brutus had been
chiefly grieved that Cicero was not in the Senate on the
Kalends; he praised Piso (who alone had spoken against
Antony at that meeting); and above all he rejoiced that
Cicero had escaped the reproach of being thought to be
travelling to Olympia. The south wind, Cicero says, turned
him back from a dishonour past all defence.}

\textit{The closing sections turn brief and practical: the
edict-war between Antony and the liberators is opening but
its aim is not yet plain; Cicero does not yet come to take
the state in hand (``what can be done? did anyone vote with
Piso?''); credit is hard to get because everyone fears arms.
The very last paragraph is the human postscript --- a
disturbed inquiry about Pilia's health, which Brutus has
mentioned in a phrase Cicero now turns over (\textit{Piliam
peirazesthai paralysei}, that she was being assailed by a
stroke). Within weeks of this letter, on 2 September, Cicero
will deliver the First Philippic. The turnabout at Leucopetra
is, in a real sense, where the Philippics begin.}
