% Ad Atticum 16.4 --- headnote
% ad-atticum-16-04, 10 July 44 BC, Puteolanum

\textit{Cicero to Atticus, written from the Puteolan villa on
10 July 44 BC --- Perseus dateline \textit{Scr. in Puteolano
vi Id. Quint. a. 710 (44)}. Cicero is still on the bay of
Naples, working his way south by sea toward Brundisium and
embarkation for Greece. The letter is a day's news from
Nesis: he has been to see Brutus, who is wounded to the quick
by the reception of his Apollinarian games on the Nones of
July --- so much so that he is arranging to have the
beast-hunt re-advertised under a different date. Libo arrives
with a letter from Sextus Pompey, addressed only to the
consuls; Cicero and the others insist on the address being
broadened to praetors, tribunes, and Senate so that the
consuls cannot suppress it.}

\textit{News from Spain follows: Sextus has taken Baria with
a single legion and a swell of popular feeling, but has
fallen back on the six legions he kept in reserve in Further
Spain, demanding as the price of any settlement that all
armies everywhere be disbanded. The Buthrotian land
question --- Atticus's perennial worry --- yields nothing
definite from inquiry: one report says the settlers have been
massacred, another that Plancus has absconded with the money.
The letter closes with travel arithmetic: the land road to
Brundisium is closed by troop movements, so Cicero will sail
in convoy with Brutus, whose squadron of two-banked vessels
(his own and Domitius's, plus the lighter craft of Sestius
and Bucilianus) is in better trim than report had suggested.
Cassius's flotilla is already past the strait. The one
remaining trouble is that Brutus is in no hurry --- but a
slow voyage, Cicero decides, beats no voyage at all.}
