% Ad Atticum 16.8 --- headnote
% ad-atticum-16-08, 2 November 44 BC, in Puteolano

\textit{Cicero to Atticus, written at the Puteolan villa on
2 November 44 BC --- Perseus dateline \textit{Scr. in
Puteolano iv Non. Nov. a. 710 (44)}. The months between this
letter and 16.7 have collapsed in the surviving
correspondence: Cicero has been back, has delivered the
First Philippic on 2 September, has weathered Antony's
furious reply of 19 September, and has retreated to the Bay
of Naples to write the Second Philippic (still unpublished
at this date) and the second book of the \textit{De
Officiis}. The news this letter carries is that Octavian,
nineteen years old, has begun to recruit Caesar's veterans
for himself --- five hundred denarii apiece to the men of
Casilinum and Calatia, with the rest of the Campanian
colonies to follow --- and to assemble a private army with
which to wage war on Antony.}

\textit{Octavian has asked for a secret meeting with Cicero
at or near Capua (``a childish thing, if he thinks it can be
done in secret''); he has sent the Volaterran Caecina with
the news that Antony is marching on Rome with the Alauda
legion, demanding moneys of the towns, and that the three
Macedonian legions on the upper sea have refused Antony's
donative and walked out on him. The famous cry of the letter
is \textit{quem autem sequamur? vide nomen, vide aetatem} ---
``but whom are we to follow? Look at the name, look at the
age.'' The name is Caesar; the age is nineteen. The closing
question --- Rome, Puteoli, or Arpinum? --- frames the
political vertigo: ``never have I been in greater
\textit{aporia}.''}
