% Ad Atticum 15.9 --- headnote
% ad-atticum-15-09, 2 June 44 BC, in Tusculano

\textit{Cicero to Atticus, written at the Tusculan villa on 2
June 44 BC --- Perseus dateline \textit{Scr. in Tusculano iv
Non. Iun. a. 710 (44)}. Late on the previous evening a letter
from Balbus has arrived with the latest political news: on the
Nones (the 5th) the Senate will meet to assign Marcus Brutus a
grain commission in Asia and Cassius another in Sicily. Both
men are praetors and ought to be administering Italy; both are
also the chief assassins of Caesar, and the proposal is plainly
a manoeuvre by Antony to push them out of Italy under a face-
saving title. Cicero is appalled --- a grain-buying errand is
no fit charge for a praetor, and certainly not for the man who
killed the tyrant. Yet, he admits, anything is better than
sitting at the Eurotas (the river of Sparta, then a stock
refuge for Romans contemplating exile or a self-imposed retreat
abroad). The daggered cruxes around the Persian Portico --- a
famous building at Sparta --- and the comparison to Lanuvium
mark a textually wrecked passage where Cicero seems to be
mocking the idea of treating Lacedaemon as a near and pleasant
resort.}

\textit{The second section turns to Atticus's own anxieties.
The previous letter from Atticus had begun with some alarming
report of armed men at his house, which Cicero is relieved to
hear has blown over. Atticus too is angling for an embassy of
his own, and Cicero awaits news of how that has gone. Cicero
adds that a letter from Brutus, which Atticus had read and
forwarded, has so unsettled him that, already in want of
counsel, he is now slower for sheer grief. The letter ends
abruptly: there was little to write, and even now uncertainty
whether the courier will reach Atticus at all.}
