% Ad Atticum 16.6 --- headnote
% ad-atticum-16-06, 25 July 44 BC, Vibo Valentia

\textit{Cicero to Atticus, written from Vibo Valentia on the
toe of Italy on 25 July 44 BC --- Perseus dateline
\textit{Scr. Vibone viii K. Sext. a. 710 (44)}. He is en
route by sea toward Brundisium and the embarkation for
Greece. Eight days out from his Pompeian villa, with one
day's stop at Velia (where Thalna, in absentia, kept him
handsomely), he has put in at the house of Sicca at Vibo,
stayed an extra day, and is now planning the next, longer
leg --- the \textit{long voyage} from Regium: by merchantman
to Patrae, or by light boat to Tarentine Leucopetra and on to
Corcyra; and if by freighter, whether to push across the
strait or go round by Syracuse.}

\textit{The middle of the letter is one of the most candid
moments in the whole sequence of the abortive Greek voyage:
``By Hercules, my dear Atticus, often I ask myself, oh, why
am I not with you? why am I not looking at the little eyes of
Italy, my own small villas?'' The answer he supplies himself
--- there is no danger now to run from --- and the question
why he is sailing at all is given an Attican answer: because
Atticus has written that his setting out is being praised to
the skies, but only on condition of return before the Kalends
of January. The familiar litany of debts follows: the
co-heirs for the Cluvian property, Publilius (his second wife
Publilia's brother, to whom he owes the return of her dowry),
and Terentia, his first wife --- Atticus is to settle
everything before the journey to Epirus.}

\textit{The letter closes with a confession of authorial
carelessness: the recently-sent \textit{De Gloria} has, by
oversight, the same preface as the third book of the
\textit{Academica}. Cicero kept a volume of prefaces and drew
from it whenever he started a new work; at Tusculum he had
forgotten that this one was already used. Now, having
recognized the slip while reading the \textit{Academica} on
shipboard, he has drafted a replacement preface and encloses
it: Atticus is to cut out the old and paste in the new.}
