% Ad Atticum 14.20 --- headnote
% ad-atticum-14-20, 11 May 44 BC, in Puteolano

\textit{Cicero to Atticus, written at the Puteolan villa on 11
May 44 BC --- Perseus dateline \textit{Scr. in Puteolano v Id.
Mai. a. 710 (44)}. Cicero has crossed the bay from Pompeii by
boat, stopped a night at Lucullus's, and arrived at his own
Puteolan house, where two further letters from Atticus reach
him on top of the one handed to him as he stepped ashore.
The letter works through them in series. Antony has slipped
through Misenum into Samnium before Cicero could intercept
him, so the Buthrotum business --- Atticus's persistent
grievance about the colony being planted on his estates ---
must wait for Rome. Lucius Antonius has been haranguing
horribly; Dolabella, by contrast, ``brilliantly'' (the
clearing of the Forum column is meant). On the dowry: he can
keep the money, Cicero says, so long as he pays the agreed
instalment on the Ides.}

\textit{The centre of the letter, section 3, is a small ars
poetica. Asked to write up a speech for Brutus, Cicero begs
off with a \textit{[Greek: katholikon theorema]} ``general
theorem'' about authors: no poet or orator has ever thought
any other man better than himself --- even bad ones think so;
so must Brutus, with his gift and learning. Cicero cites
their recent disagreement over the funeral edict and over
\textit{De Optimo Genere Dicendi}, and ends with a snatch of
Atilius (``to each his own betrothed, to me my own''), which
he calls inelegant in the same breath that he quotes it. He
then turns to political weather: the edict of Brutus and
Cassius approved, Hirtius coachable but lodging with Balbus,
Dolabella's two recent strokes a real advance. Section 5
catches Atticus in a half-blasphemy against Epicurus
(\textit{[Greek: m\=e]}) and teases him that Brutus's slight
scowl ought to deter him. Cicero is dashing off the letter
because Cassius's courier is leaving; he is about to call on
Pilia, then sail across to Vestorius's for dinner.}
