% Ad Atticum 15.8 --- headnote
% ad-atticum-15-08, 31 May 44 BC, in Tusculano

\textit{Cicero to Atticus, written at the Tusculan villa on 31
May 44 BC --- Perseus dateline \textit{Scr. in Tusculano prid.
K. Iun. a. 710 (44)}. Atticus has now left Cicero; since the
departure two letters have arrived from Balbus (nothing new in
them) and a matching pair from Hirtius, reporting that he is
violently at odds with the veterans. Cicero is preoccupied with
how to behave on the Kalends of June, the next sitting of the
Senate; he has dispatched Tiro with several other couriers, so
that Atticus may send a letter back by each as each piece of
news comes in. He has also written to Antony about the embassy
he is seeking --- not just to Dolabella, lest the
\textit{iracundus homo} take offence --- and has arranged to
have his letter handed to Antony through Eutrapelus, since
direct access is difficult. The \textit{legatio votiva}, an
embassy nominally to discharge a vow, is the more honourable
form, but either type will serve.}

\textit{Section 2 closes with anxiety about Atticus's own
safety and a piece of alarming intelligence: Graeceius has
reported that Gaius Cassius has reported in turn that armed
men are being mustered to be sent into the Tusculan country
where Cicero is. The Latin around the assessment of this
report and the precaution Cicero proposes is in textual ruin
--- two daggered cruxes in the manuscripts --- and the sense
must be left provisional. The letter ends in mid-thought, with
the hope that the next day will bring something further to
weigh.}
