% Ad Atticum 14.8 --- headnote
% ad-atticum-14-08, 15 April 44 BC, in Sinuessano

\textit{Cicero to Atticus, written on 15 April 44 BC at a wayside
inn at Sinuessa --- Perseus dateline \textit{Scr. in Sinuessano
xvii K. Mai. a. 710 (44)}, the body confirming the day and the
\textit{deversoriolum Sinuessanum}. The letter sits on the same
day as 14.7 (written that morning at Formiae as Cicero left for
Puteoli) and is the second instalment of the day's
correspondence, written on the road. The news is mixed. The
pseudo-Marius Herophilus has been put down (\textit{de Mario
probe}) --- ``well done,'' though Cicero spares a line of regret
for L. Licinius Crassus's grandson; Brutus is now finding even
Antony acceptable; Cleopatra, the \textit{regina} who has been
in Rome since 46, has fled, and Cicero is unmoved
(\textit{reginae fuga mihi non molesta est}, his famous tart
half-sentence on the queen).}

\textit{The register is fragmentary and rapid --- a wayside
dispatch built from short clauses and queries, the kind of
letter Atticus is meant to skim and answer item by item. A
sentence is left hanging at the end: \textit{verum tamen---},
``and yet---,'' a deliberate aposiopesis on Brutus's safety
that the Latin marks with a long dash and that the English
preserves. The whole letter sits in the post-Ides ambivalence
that runs through the 14th book: things are calming, the
Caesarians are not yet united, and the Liberators may yet be
able to move in Rome --- but Cicero, on the way to Cumae, will
not say so without the qualifier.}
