% Ad Atticum 14.13 --- headnote
% ad-atticum-14-13, 26 April 44 BC, in Cumano

\textit{Cicero to Atticus, written at the Cumean villa on 26
April 44 BC --- Perseus dateline \textit{Scr. in Cumano vi K.
Mai. a. 710 (44)}. The longest of this Cumean cluster, and the
most heavily Homeric. Two passages of the \textit{Iliad} carry
the argument: the council-of-the-Achaeans line on which a
greater evil presses than feasting (\textit{Il.}~9.228--230,
adapted), and Aphrodite's instruction to Aeneas to leave war
to others and pursue the works of speech (\textit{Il.}~5.428--429,
re-pointed as the consolation Cicero and Atticus might offer
each other in the looming civil war). The seaside-walk word
\textit{halitene\^i} (\textit{Greek: ``stretched along the
sea''}) names a specific delight of the Bay of Naples.}

\textit{The middle of the letter contains the clearest
statement so far of the choice Cicero will worry at all
spring: stay in Italy and risk slaughter, or accept the
embassy to Greece and face the charge of having abandoned the
state. The last section turns to a piece of business that
crystallises the new regime's character --- Antony's
restoration of Sextus Clodius, justified out of Caesar's
\textit{commentarii}, which Cicero believes to be forged.
Cicero has, against his judgement, agreed: ``\textit{he, since
he has once admitted into his mind that whatever he wishes is
permitted him, would have done it no less without my
consent.}'' Copies of both letters --- Antony's and his own
reply --- are enclosed.}
