% Ad Atticum 15.11 --- headnote
% ad-atticum-15-11, 8 June 44 BC, in Antiati

\textit{Cicero to Atticus, written at Antium on 8 June 44 BC ---
Perseus dateline \textit{Scr. in Antiati a. d. vi Id. Iun. a. 710
(44)}. Cicero has gone to Antium to attend the council Brutus
and Cassius have convened on the question of what to do next.
Two months after the Ides of March the assassins are stuck:
Antony has the consulship and the veterans, the conspirators
have praetorships and a country in which they cannot safely
appear, and the proposal to send Brutus and Cassius abroad on a
grain commission --- ostensibly an honour, in fact an exile ---
is on the table. The council is held in front of the Liberators'
formidable women: Servilia (Brutus's mother and Caesar's old
mistress), Tertulla (Cassius's wife, Brutus's half-sister), and
Porcia (Brutus's wife and Cato's daughter); the Stoic Favonius
is also present. This letter is Cicero's eye-witness report to
Atticus, and one of the most vivid pieces of political reportage
in the corpus.}

\textit{Cicero comes with advice prepared: accept the grain
commission, take Asia, stay alive --- in Brutus's survival lies
the survival of the commonwealth. Cassius bursts in with the
glance of Mars and refuses Sicily; he will go to Achaea instead.
Brutus asks whether he might come to Rome; Cicero says no. The
recriminations turn on Decimus Brutus and on opportunities
missed; when Cicero says what ought to have been done --- not
only Caesar struck, but the Senate summoned, the people roused,
the whole commonwealth taken in hand --- Servilia (the
\textit{tua familiaris}, your dear friend, addressed to Atticus
the family intimate) cuts him off: \textit{nobody has ever said
that to me}. The Greek line in section 3, from Euripides'
\textit{Iphigeneia in Aulis}, is what Cicero says to himself on
the journey home: \textit{what does this journey of yours hither
avail you now, prophet?} The vessel of state is shattered; he
will get away. The Pelopidae --- the cursed house of Pelops ---
are Caesar's heirs and the conspirators alike. Section 4 brings
the news Cicero has been angling for: Dolabella has named him
his \textit{legatus}, an embassy of the most useful kind ---
five years' worth of freedom to come and go --- though, as Cicero
breaks off with a Greek word for ill-omened speech, who knows
how much of the five he will have.}
