% Ad Familiares 11.28 --- headnote
% ad-familiares-11-28, September 44 BC, Rome

\textit{C. Matius to Cicero, from Rome, a little after
the preceding letter (11.27) --- the Perseus dateline
places it \textit{Scr. Romae paulo post ep. 27}, so that
the meta year-placeholder is here corrected to 44 BC. This
is Matius's reply to Cicero's defense of their friendship:
the complaint that Trebatius had carried to Cicero, and
that Cicero had answered, now draws from Matius himself a
full and deliberate response. Matius was an old equestrian
friend of both men and a devoted personal intimate of
Julius Caesar; after the Ides he had taken on the
management of the \textit{ludi Victoriae Caesaris} for
Octavian and was widely reported to be the leading
Caesarian voice in private conversation at Rome.}

\textit{The letter is Matius's celebrated apologia. Point
by point he takes up the charges laid against him --- that
he grieves openly for a murdered man, that he superintended
Octavian's games, that he frequents Antony's house --- and
he answers each not by political argument but by an appeal
to the constancy of private friendship: it was a friend he
followed in the civil dissension, not Caesar's cause, of
which he says he disapproved and which he strove to smother
at its birth. He claims for himself the right, granted even
to slaves, to grieve at his own discretion, and turns the
language of the ``authors of liberty'' against them. The
prose is built with great care --- balanced, proud,
wounded but unfailingly courteous --- and the result is one
of the most admired letters in the whole correspondence:
the finest surviving statement from antiquity of personal
loyalty held steady above political faction, and the
dignified counterweight to Cicero's own casuistry in
11.27.}
