% Ad Familiares 10.5 --- headnote
% ad-familiares-10-05, mid-December 44 BC, Rome

\textit{Cicero to L. Munatius Plancus, written from Rome in
the middle of December 44 BC --- Perseus dateline
\textit{Scr. Romae med. m. Dec. a. 710 (44)}. This is the
first letter of the surviving correspondence with Plancus,
governor of Transalpine Gaul and consul-designate for 42,
that will run through the spring and summer of 43 and form
one of the central documents of Cicero's last political
campaign. Plancus had written two identical letters
(``the same content'' --- Cicero's gloss is that this was
itself a sign of his anxiety to get the message through)
declaring his good faith toward the republic and recalling
his father's friendship with Cicero. Cicero's reply is a
brisk piece of management: he warmly accepts the personal
note, redirects to the political one, and then closes with
a sharp piece of moral counsel.}

\textit{The argument of section three is the one Cicero
will press on Plancus all spring: that what he has so far
attained is owed in part to ``fortune and the times,'' and
that the present crisis is his chance to make his own
\textit{dignitas} on his own merit, by coming down on the
right side against Antony. The \textit{latrones}, the
``brigands,'' are Antony's circle --- the standing
\textit{Philippics} term that runs through these letters.
The hand-on-hilt mood --- ``in the gods' name, do not let
the moment slip'' --- is the same urgency Cicero will turn
on Decimus Brutus and on the Senate itself over the next
three months.}
