% Ad Atticum 14.6 --- headnote
% ad-atticum-14-06, 12 April 44 BC, Fundis

\textit{Cicero to Atticus, written at Fundi on the evening of 12
April 44 BC --- Perseus dateline \textit{Scr. Fundis prid. Id.
Apr. a. 710 (44)}, with the body confirming \textit{pridie
Idus}. The letter carries an earlier date than 14.5 (13 April,
from Astura): Perseus's letter-ordering follows the manuscript
sequence, not strict chronology. The post-Ides week is not yet
a month old. Antony has called a meeting with the Liberators
(\textit{Antonio conloquium cum heroibus nostris}, ``our heroes''
with the affectionate sting Cicero gives the word at this date);
the calendar of magistracies Caesar laid down --- consuls and
tribunes set two years ahead --- still stands; the country towns
are exultant at Caesar's death, but in Rome no decree of the
Senate has issued.}

\textit{The famous formulation of the post-Ides paradox falls in
the second section: ``the tyrannicides in heaven, the tyrant's
acts defended''. The Greek is denser here than in 14.5
(\textit{politeuesthai}, \textit{soloikon},
\textit{pepoliteumetha}, \textit{politikotera}) and clusters
around a single root --- the verb of civic action --- as Cicero
turns it over and finds no way to use it: he cannot play the
citizen if the citizen's role is to ratify the murdered
tyrant's decrees. The dispatch is hurried, written at table
with the second course in front of him (\textit{apposita secunda
mensa}), in the brisk, anxious register of the moving carriage:
notes, queries, and one sharp line of political diagnosis ---
\textit{ut victos metueremus}, ``so that we go in fear of the
beaten.''}
