% Ad Atticum 14.5 --- headnote
% ad-atticum-14-05, 13 April 44 BC, Asturae

\textit{Cicero to Atticus, sent from Astura on 13 April 44 BC ---
Perseus dateline \textit{Scr. Asturae it Id. Apr. a. 710 (44)}
(the body adds \textit{Astura iii Idus}, the day of writing as
he sets out). Less than a month after the Ides of March, Cicero
is moving south through his estates, and the news from Rome is
all bad. The Liberators are still effectively penned up: Brutus
and Cassius have left the Capitol but cannot move freely; the
legions Caesar had collected for the Parthian campaign are
returning, standards and all, from Gaul, and may yet be turned
to dynastic ends; the freedmen of Caesar's household have begun
to plot on their own account, ``which could easily be put down,
if Antony had any sense.'' The young Octavius has just landed in
Italy, and Cicero already wants to know whether anyone is
rallying to him.}

\textit{The letter is hurried --- written on the road, sprinkled
with Greek (\textit{esitesas} for his friend's fast; the
self-correction \textit{mnemonikon hamartema} when he muddles
Calvena with Asinius; \textit{phurmos polus}, the bath-keeper's
``great muddle''; \textit{euripista} for the easily-shifted props
of the new regime; \textit{neoterismou} for a possible
disturbance) --- in the rapid, ironic register of the daily
correspondence under threat. The third paragraph's tricolon
\textit{vides magistratus, vides tyranni satellites, vides
exercitus} is the architecture of the piece: the apparatus of
state still standing, but staffed by Caesar's men, and the
Liberators reduced to ``only praised and only loved''
(\textit{tantum modo laudari atque amari}) --- a bitter
diminutive in a single line.}
