% Ad Familiares 10.4 --- headnote
% ad-familiares-10-04, end of December 44 BC, Transalpine Gaul

\textit{L. Munatius Plancus to Cicero, written from
Transalpine Gaul at the end of December 44 BC --- Perseus
dateline \textit{Scr. in Gallia Transalpina ex. m. Dec.
a. 710 (44)}. This is the first of Plancus's own letters
to survive in the correspondence, and it answers
\textit{Fam.}\ 10.3 (Cicero's substantive piece of
political advice, sent earlier in December). The opening
explains the delay: Plancus had heard Cicero was on his
way to Greece and only learned from the letter itself
that he was back in Italy.}

\textit{The body is, by the conventions of late-republican
correspondence, an unusually direct profession of loyalty:
Plancus binds himself to Cicero with the formula
``cultivating you I have set up for myself an obligation
as binding as toward my country,'' and promises that
``whatever good things are mine \dots\ seem to lack
nothing except a good name'' --- meaning that his
\textit{fama}, his standing among good men, is what he
now intends to make. He undertakes that his strength,
counsel, and authority will all be at the republic's
service, and reports that he is watching closely what is
happening in Cisalpine Gaul --- Antony will shortly cross
into the province against D. Brutus --- and at Rome in
the month of January, when the new tribunes and consuls
take office. The local anxiety is that the Gallic peoples,
seeing the disorder among the Romans, will think it their
chance. The letter closes with the standard intimate
sign-off: \textit{fac valeas meque mutuo diligas} ---
``keep well, and love me in return.''}
