% Ad Familiares 10.3 --- headnote
% ad-familiares-10-03, shortly after 10 December 44 BC, Rome

\textit{Cicero to L. Munatius Plancus, written from Rome
shortly after 10 December 44 BC --- Perseus dateline
\textit{Scr. Romae paulo post iv Id. Dec. a. 710 (44)}.
Furnius, the courier of the Plancus correspondence, has
returned from Gaul with a report on Plancus's
administration. The letter opens with the report
itself --- ``manhood in military matters, justice in the
administration of the province, wisdom in every kind of
business'' --- and slides from compliment into the longest
and most directive piece of political advice in the
sequence so far.}

\textit{The frank centre of the letter is its third
section. Cicero acknowledges, in print and to Plancus's
face, what everyone had thought during the Caesarian
years: ``there was a certain time when men thought you too
compliant with the times.'' The exoneration is on the
ground that Plancus had not approved what he endured but
had been gauging what was possible. Now, with Plancus
consul-designate and the senatorial bench thinned of men
of his calibre, the calculation is different. ``Press on,
by the immortal gods!\ into that care and consideration
which will bring you the highest standing and glory; for
there is one course \dots\ and one only, that leads to
glory: governing the republic well.'' Cicero closes by
disclaiming any need to instruct Plancus --- they have
drunk from the same springs --- and promising continued
attention to his interests at Rome.}
