% Ad Atticum 14.4 --- headnote
% ad-atticum-14-04, 10 April 44 BC, Lanuvium

\textit{Cicero to Atticus, written on 10 April 44 BC at Lanuvium
--- Perseus dateline \textit{Scr. Lanuvi iv Id. Apr. a. 710 (44)}.
The fourth letter in four days. The itinerary set in 14.2 has
been kept: from Matius's suburban villa to Tusculum, now to
Lanuvium, with Astura still ahead. Cicero is out of Rome and
hungry for news; he assumes Atticus is hearing something every
day. Then the central sentence of the early post-Ides
correspondence: \textit{equidem doleo, quod numquam in ulla
civitate accidit non una cum libertate rem publicam reciperatam}
--- ``I grieve, what has never happened in any state, that the
commonwealth has not been recovered along with liberty.''
Liberty without a state to house it is the political diagnosis
the rest of the year will not escape.}

\textit{The second movement of the letter is the famous balance
the Liberators set against that grief. \textit{Idus Martiae
consolantur}: the Ides of March are a consolation. \textit{Nostri
\textit{[Greek: hērōes]}, quod per ipsos confici potuit,
gloriosissime et magnificentissime confecerunt; reliquae res
opes et copias desiderant, quas nullas habemus} --- our heroes
did what they could do by themselves, with the greatest glory
and the greatest magnificence; the rest demands resources and
forces which we do not have. The Greek tag (\textit{hērōes}) is
already, ten days after the act, doing the work Cicero will keep
asking it to do: dignifying the deed at the moment of its
political insufficiency. The register is what these letters
have begun to settle into --- short, paratactic, anxious,
intimate, with the standing closing pledge that the daily notes
between the two men will not lapse.}
