% Ad Atticum 14.9 --- headnote
% ad-atticum-14-09, 17 April 44 BC, in Cumano

\textit{Cicero to Atticus, written at the Cumean villa on 17
April 44 BC --- Perseus dateline \textit{Scr. in Cumano xv K.
Mai. a. 710 (44)}. Cicero has finally reached the bay of Naples,
and a bundle of Atticus's letters has caught up with him.
The Cluvian inheritance is settling well; some Puteolan
properties --- the shops Cicero rents out --- have collapsed,
\textit{tenants and mice} (\textit{inquilini sed mures}) alike
having fled, and Cicero turns the loss into a Stoic flourish
(\textit{o Socrate et Socratici viri!}, with a passing
ventriloquism of the Phaedo's address) before noting drily that
Vestorius has a scheme to make the rebuilding profitable.}

\textit{The political middle section carries the famous line of
the post-Ides spring: \textit{vivit tyrannis, tyrannus
occidit!} --- ``the tyranny lives, the tyrant fallen!''. The
formula is built on a tight chiasmus and a pair of cognates, and
the English keeps the bare structure (subject-verb,
verb-subject). Marcus Curtius has been shaming the Liberators
and their friends for staying alive; Cicero accepts the charge.
The final section turns to news from the East and the West:
Vetus's setback in Syria against Pacorus's Parthians (the
trouble Cassius will inherit), and Balbus's report from Gaul
that the German tribes have submitted to Aurelius --- ``all full
of peace,'' which is, as Cicero notes with his last word, the
opposite of what Calvena had told him.}
