% Ad Atticum 15.14 --- headnote
% ad-atticum-15-14, 27 June 44 BC, in Tusculano

\textit{Cicero to Atticus, written at the Tusculan villa on 27
June 44 BC --- Perseus dateline \textit{Scr. in Tusculano v K.
Quint. a. 710 (44)}. The subject is the Buthrotian affair: the
free Greek city of Buthrotum on the Epirote coast had been
threatened with a punitive land-confiscation, and Atticus, who
owned estates there, had spent months trying to head it off.
Caesar before his death had granted Atticus's request; after
the Ides, the matter fell to the consul Dolabella, and a
favourable judgement had now been secured. Atticus had come to
Tusculum in person to thank Cicero for his part in the
intercession; Cicero, having already written once to Dolabella,
now writes again so that the consul should not think his
earlier thanks perfunctory.}

\textit{The letter encloses a full copy of the second letter to
Dolabella --- a study in elaborate Ciceronian gratitude, with
the political subtext that Dolabella's continuing protection of
Buthrotum is what Cicero is really after. The closing
section returns to Cicero's own preoccupation of the summer: he
is at work on his \textit{compositions} --- the philosophical
writing of 44, the De Finibus already in circulation and the
Tusculan Disputations and De Natura Deorum in progress --- and
warns Atticus, his usual editor, to keep his red-lead pencil
ready. The Greek term \textit{meteoros}, ``suspended in
mid-air,'' is Cicero's wry self-description of a mind too taken
up with great thoughts to attend properly to small ones.}
