% Ad Atticum 15.29 --- headnote
% ad-atticum-15-29, 6 July 44 BC, in Formiano

\textit{Cicero to Atticus from the Formian villa, 6 July 44 BC
--- Perseus dateline \textit{Scr. in Formiano prid. Non. Quint. a.
710 (44)}. (The ``prict'' in the fetched dateline is an OCR bleed
for ``prid.''; the date is fixed internally by section 3, where
Cicero says his Formian dinner-guests reported seeing Plancus
``III Non.'' = 5 July, ``the day before I was writing this.'') The
letter is in three sections and was written on the journey south:
Cicero is at his Formian estate, between the Arpinum letters
(15.26--15.28) and the planned arrival at Puteoli on the Nones.}

\textit{Section 1 is a string of replies to Atticus's previous
letter and a forwarding of Brutus's --- ``good gods, what
helplessness'' \textit{[Greek: am\=echania]}: the perplexity of
the assassin party, who can neither stay nor leave. Section 2
brings Quintus the younger into the picture: the wretched nephew
is rushing all the way to Puteoli to make a show of patriotism, on
two pretexts, both suspect --- a reconciliation with his uncle and
an alliance with Brutus and Cassius. The interview with Quintus's
father over a possible marriage to a Iulia of the Othones is
sketched with sardonic precision; Cicero suspects the boy is
fantasising. Section 3 is a postscript opened after the letter was
sealed: Formian dinner-guests have reported that Plancus, the
agent of the Buthrotian colonists whose case has bedevilled the
year, has been thrown out of Buthrotum stripped of his
\textit{phalerae} (the medallions of a soldier's harness).
Cicero's ``Well done!'' \textit{(macte!)} is unadulterated.}
