% Ad Atticum 15.19 --- headnote
% ad-atticum-15-19, 21 June 44 BC, in Tusculano

\textit{Cicero to Atticus, written at the Tusculan villa about
21 June 44 BC --- Perseus dateline \textit{Scr. in Tusculano
inter a. d. xv et xi K. Quint. a. 710 (44)}, between 17 and 21
June. The first section gathers up two recurring threads.
Buthrotum --- the small Epirote town whose interests Atticus
has championed against a planned Caesarian settlement of
veterans --- has come to nothing in spite of his efforts;
Cicero asks why Brutus, as urban praetor, is now reopening it.
Brutus's affairs are altogether overburdening Atticus, the work
of ten men loaded onto one --- laborious
\textit{[erg\=odes]} but bearable
\textit{[anekton]}, in the Greek terms Cicero reaches for.
The reference to ``the arms'' is glancing and conspiratorial:
Cicero will not commit anything more explicit to paper, and
proposes that he and Atticus run together and confer face to
face.}

\textit{The second section reports an extraordinary piece of
intelligence. Statius, Quintus Cicero's freedman, has written
to say that the younger Quintus --- Cicero's nephew, an
intemperate and unreliable young man --- has announced his
intention of going over to Brutus and Cassius. The three
possible motives Cicero offers (anger with Antony, fresh
ambition, sheer improvisation
\textit{[schediasma]}) shade quickly into the fourth, that the
last is no doubt the truth. The boy's father, Quintus the
elder, is upset, for he remembers things his son has said about
him --- unspeakable things \textit{[aphata]}. The note closes
with two quick questions: did Gaius Antonius (Mark Antony's
younger brother) want a place on the board of seven for land
distribution, and what is happening with Menedemus.}
