% Ad Atticum 15.12 --- headnote
% ad-atticum-15-12, 10 June 44 BC, in Antiati

\textit{Cicero to Atticus, written at Antium on 10 June 44 BC
--- Perseus dateline \textit{Scr. in Antiati v aut iv Id. Iun.
a. 710 (44)}, the 9th or 10th. Two days after the Antium
council, Cicero is still on the coast and tying up the threads
of that meeting for Atticus. Buthrotum --- the old Epirote town
where Atticus held land and was trying to head off a colonial
settlement of Caesarian veterans --- has been spared, at least
for the moment, which is the day's one piece of unambiguously
good news. Tiro has been sent to Dolabella with the letter
Atticus asked for. The verdict on the conspirators is already
written out elsewhere, but Cicero restates it for clarity:
Cassius, having spurned Sicily, expects Servilia to get the
grain-commission revoked; Brutus, very gravely (Cicero reaches
for the Greek \textit{kai mala semnōs}) is bent on Asia and has
begun collecting ships. The games will go on at Rome under
Brutus's name in his absence.}

\textit{The second section steps from immediate politics to the
question of who will be left to do anything. Lucius Antonius
(the consul's brother and tribune) has written generously and
tells Cicero to be at ease; one favour from a Caesarian, perhaps
a second on the way. Then the Greek tag from tragedy, \textit{One
of the Bruti bears the blame for this} --- Decimus Brutus, who
let Antony live --- and the first considered sketch of Octavian
in the Ciceronian corpus. Cicero has now seen the young
nineteen-year-old at Astura: ability, spirit, sympathy for the
\textit{hērōes} of the Ides. What weight to give his youth, his
adoptive name, his Caesarian inheritance, and the people putting
ideas into his head \textit{(katēchēsei)} --- this needs careful
thought. The stepfather (L. Marcius Philippus) thinks nothing of
him. Still, the young man should be cultivated --- if only to
prise him away from Antony. C. Claudius Marcellus is doing well
by him, if (the text is uncertain here, two daggered words)
\textit{he is putting our young man through his paces for us}.
He distrusts Pansa and Hirtius. A good nature, Cicero closes in
Greek --- \textit{if it lasts}.}
