% Ad Atticum 16.15 --- headnote
% ad-atticum-16-15, c. 9 December 44 BC, Arpinum

\textit{Cicero to Atticus, written from Arpinum on or shortly
before 9 December 44 BC --- Perseus dateline
\textit{Scr. in Arpinati ante v Id. Dec. a. 710 (44)}. This is
the last surviving letter to Atticus that the manuscript
tradition preserves; nothing further runs forward into 43 BC,
and the archive that gave us sixteen books of intimate
correspondence breaks off here, on the eve of the Mutina
campaign. The letter is also one of the longest of the year,
covering legal business, the constitutional crisis, and an
ugly personal money problem in a single sweep.}

\textit{The opening is one of Cicero's familiar pieces of
self-deprecating play: he is not writing in his own hand, and
yes, it really is from laziness. Then to business. The first
section is on his quarrel with Dolabella, his former
son-in-law, who had ``begun to defend the republic with my
encouragement, and then, bought off with money, not only
deserted her but, so far as it lay in him, actually overthrew
her.'' Cicero will now press the full rigour of the law, but
asks Atticus to consider whether bringing in his agent rather
than calling the sureties would be the more dignified course.
The legal vocabulary --- \textit{sponsores}, \textit{litem
contestabuntur}, \textit{satis dato} --- is technical, and the
text is in places visibly corrupt (daggers preserved).}

\textit{The middle section turns to the political situation
and produces the letter's most famous moment: a speech of
Octavian's has reached Cicero in which the young heir, with
hand stretched out toward the statue of his adoptive father,
swears ``as I hope to attain the honours of my father.''
Cicero's response is in Greek, the disgust unbroken by the
register-shift: \textit{[Greek: m\=ede s\=othei\=en hupo ge
toioutou]} --- ``may I never be saved by the likes of him.''
This is December 44, and the calculation is the one that will
prove fatal: that Octavian can be used against Antony and then
discarded. Cicero has not yet committed to that gamble but is
moving toward it; he will look to Casca's tribunate (Casca,
one of the assassins) as the index of which way the young man
will go. The cover-letter from Lepta which he is sending on
suggests Antony --- ``that \textit{[Greek: Stratullax]},''
that strutting comic-stage soldier --- has been knocked off
his pedestal.}

\textit{The closing two sections drop the public matter
abruptly --- ``in cases beyond hope even Hippocrates forbids
the application of medicine'' --- and turn to a humiliating
private one: a debt of 25{,}000 sesterces to Montanus, which
Cicero had promised to pay on his son's honour, has gone
unpaid, Aurelius has been forced to borrow at ruinous
interest, and the money Atticus had thought would come from
Dolabella is not coming. ``It is more shameful to fall
privately than publicly.'' The letter ends with Cicero's
decision to come down from Arpinum: \textit{adsum igitur} ---
``So --- I am coming.'' It is the last line we have from him
to Atticus.}
