% Ad Familiares 6.11 --- headnote
% ad-familiares-06-11, early June 45 BC, Tusculum

\textit{Cicero to Trebianus, written from the Tusculan villa
in early June 45 BC (works.yaml \texttt{-0045-06-15} at
month-precision; the Perseus dateline reads \textit{in Tusculano
in.~m.~Iun.~a.~709 (45)}, ``at the beginning of the month of June''
--- a tightening to \textit{circa} ${-}0045$-$06$-$01$ is plausible,
but month-precision covers it). This is the companion piece to
\textit{Fam.}~6.10 of nearly a year before, the case now closed:
Trebianus has been restored, the intervention of Dolabella has
carried the day, and Cicero writes from Tusculum a short letter
of congratulation and Stoic stocktaking.}

\textit{No Greek. The register lightens after the long working
note of 6.10: Dolabella has done what Cicero himself could not
quite do, and Cicero is unembarrassed to mark the debt. The
philosophical pivot in section~2 is the standard
\textit{consolatio}-flipped-into-thanksgiving turn of Book~6 ---
the recovered \textit{dignitas} now matters more than the lost
\textit{res familiaris}, and even property would only be the
sweeter to enjoy ``if there were any commonwealth left''. The
parenthetical greeting to Siro the Epicurean (in whose Naples-bay
circle Virgil and Varius were just then reading philosophy) is
the unobtrusive marker of how thin the line is, in this cluster,
between consolatory Stoicism and an Epicurean refuge of letters
and friends.}

\textit{One crux: in section~1 the Perseus text carries a dagger
at \textit{nec enim acciderat mihi $\dagger$opus esse}; I have
rendered the sense as ``for it had never fallen out that I had
occasion to need him,'' following the standard supplement
(\textit{mihi <eo> opus esse} or similar). The line of thought is
secure even if the precise words are not.}
