% Ad Atticum 13.44 --- headnote
% ad-atticum-13-44, 21 July 45 BC, Tusculum

\textit{Cicero to Atticus, written at the Tusculan villa on 21
July 45 BC --- Perseus dateline \textit{Scr. in Tusculano xiii aut
xii K. Sext. a. 709 (45)}, which fixes the day as 20 or 21 July.
Three short sections, packed with the daily threads of the
summer's correspondence. The opening flash is a glance at the
\textit{ludi Victoriae Caesaris}: Caesar's statue had been carried
in procession among the gods, and the people had refused to
applaud Victory because they had been forced to applaud the
``bad neighbour'' (Caesar) alongside her. ``Splendid of them''
(\textit{populum praeclarum}) is as openly political as Cicero
ever gets in these notes. The procession is also what makes him
flinch from finishing the letter to Caesar that Brutus had come
out to Tusculum to press on him --- the much-discussed
``letter to Caesar'' of advice, never sent.}

\textit{The middle section is the now-running joke about Atticus
having handed the \textit{Academica} to Varro --- ``yet you
ventured!'' (\textit{ausus es}) --- and a passing approval of
something pious that Attica has done. The third drops into the
purest house-keeping mode: Cicero corrects a factual slip in the
already-circulating \textit{Pro Ligario} (Lucius Corfidius was
dead before the speech's date), and asks Atticus to set the
copyists Pharnaces, Antaeus, and Salvius to scrubbing the name
out of every copy. The Greek tag for the mistake is
\textit{mnemonikon hamartema}, a slip of memory --- a small
candid window into how Cicero's published speeches circulated
and got patched after release.}
