% Ad Atticum 13.42 --- headnote
% ad-atticum-13-42, late December 45 BC, Tusculum

\textit{Cicero to Atticus, written at the Tusculan villa in late
December 45 BC --- Perseus dateline \textit{Scr. in Tusculano
ex. m. Dec. a. 709 (45)}. Three sections, the first of which is
one of the small masterpieces of the correspondence: a comic
dialogue, set out in direct speech, between Cicero and his
nephew. Four months on from the bitter August exchange, the
young man arrives at the Tusculan villa preparing to march off to
the war in Spain --- the long aftermath of Munda --- in debt and
without even travel-money. He pivots, in the middle of the
encounter, to the real wound: ``What really tortures me is my
uncle.'' Cicero, with a phrase he attributes to Atticus's
manner, ``borrows a little from your style of speaking: I said
nothing.'' The boy promises to ``take away the cause'' by
marrying as his mother wishes; Cicero, dryly, presses for a
date.}

\textit{The opening Greek tag \textit{kai} is left dangling --- a
manuscript scar where some Greek pleasantry has dropped out ---
followed by Cicero's own Greek opening to the boy:
\textit{su de de ti synnous?} ``But you --- why so pensive?'' (a
phrase with a faintly tragic register). The closing tag
\textit{me skordou} (``no garlic'') is one of Cicero's running
in-jokes with Atticus; the augurs' ritual of \textit{templum
effandum} --- marking out a sacred precinct --- is what Lepidus
is calling him to. The register is comic, controlled, almost
miniature-dramatic: dialogue, frame, and outward business in
three crisp sections.}
