% Ad Atticum 13.33 --- headnote
% ad-atticum-13-33, 3 June 45 BC, Tusculum

\textit{Cicero to Atticus, written at the Tusculan villa on 3 June 45
BC --- Perseus dateline \textit{Scr. in Tusculano iii Non. Iun. a.
709 (45)}. Three short sections of pure business chatter from the
Tusculan daily-letter run: an outburst of irritation at Balbus
and Faberius for having had to be told repeatedly that the
property declaration was filed; an inventory of pending
transactions --- a Vergilius purchase, a debt with Cispius and
Plancus, the haggling with Otho over the Scapulan gardens, a
valuation of an estate; and, hanging on to the end, a request
for Atticus to look up the consulship of Cnaeus Cornelius and
Lucius Mummius (146 BC) in the senatorial decrees, with a
prosopographical question about whether Hortensius's Tuditanus
was a quaestor, a military tribune, or simply on the staff in
the Achaean war.}

\textit{Three Greek phrases punctuate the letter, all colloquial.
\textit{dys\=opia} --- ``embarrassment'' or ``awkwardness'' --- captures
Cicero's relief at having no scruple about buying a property out
from under Vergilius, since he owes the man nothing. \textit{katabase\=os}
--- ``of the descent'' --- is the title of one of Dicaearchus's works
(the \textit{Descent into Trophonius}'s cave), here as a book Cicero
is still waiting to receive. \textit{eulogon}, ``plausible,'' is
the everyday Stoic-philosophical adverb that Cicero reaches for
when adjudicating a historical guess. The Latin carries several
cruxes preserved as daggers --- \textit{H in Capitolio}, the
half-recoverable note on Balbus's business; \textit{exspecto},
where the verb governing \textit{katabase\=os} is uncertain; and
two near the close of section 3 where the text breaks down on the
question of Antiochus's position in the army at Corinth.}
