% Ad Atticum 13.45 --- headnote
% ad-atticum-13-45, 11 August 45 BC, Tusculum

\textit{Cicero to Atticus, written at the Tusculan villa on 11
August 45 BC --- Perseus dateline \textit{Scr. in Tusculana iii
Id. Sext. a. 709 (45)}. Three sections, each on a different
piece of the August calendar. First, the news that Caesar is on
the move: Lamia has been by with a letter the dictator sent him,
making it plain he means to be in Rome before the
\textit{Ludi Romani}; the public holidays attached to the games
have been quietly extended again, and Cicero wants to know
exactly by how many. Second, Atticus has urged him to use the
extra days for ``unfolding philosophy'' --- and gets the dry
Greek-flavoured retort that he is ``urging on a runner''
(\textit{currentem hortari}), the standard tag for telling
someone to do what he is already doing. The catch is Dolabella,
his son-in-law of the moment, who must be entertained, and the
court case of Torquatus, which keeps him from running down to
Puteoli and back.}

\textit{The third section is the one that ties this letter to
several others around it: the affairs of the deceased Cluvius,
whose estate Cicero has inherited a share in, with Vestorius
acting as agent in Puteoli. The careful tricolon ``a more
careful or more obliging or, by Hercules, more devoted man''
(\textit{nec diligentiorem nec officiosiorem nec me hercule
nostri studiosiorem}) is the kind of small architectural
flourish Cicero produces in passing even in a money-and-auction
note. The letter closes with the worry that animates so much of
the financial correspondence of these months ---
\textit{ne neglegentiores esse videamur}, ``that we should not
appear too negligent'' --- and the open invitation for Atticus
to write back at once.}
