% Ad Atticum 13.12 --- headnote
% ad-atticum-13-12, 24 or 25 June 45 BC, Arpinum

\textit{Cicero to Atticus, written from the Arpinate
estate on or about the eighth or seventh day before
the Kalends of Quintilis 709 AUC --- 24 or 25 June 45
BC. The Perseus dateline is OCR-garbled
(\textit{Scr.\ in Arpinati vh i K.\ Quint.\ a.\ 709
(45)}); the surrounding sequence is fixed --- 23 June
(13.11), 26 June (13.13) --- and the corruption is
best read as \textit{viii K.\ Quint.} (24 June) or
\textit{vii K.\ Quint.} (25 June). Section 4 itself
fixes the auction date as \textit{a.\ d.\ viii Kal.\
Quint.} (24 June), and Cicero says ``they thought I'd
be at Rome or the Tusculanum then,'' which reads more
naturally if the letter is written a day later than
the auction notice. (The works.yaml entry currently
carries the invalid date \texttt{-0045-06-31} for this
letter; flagged for PM correction to 24 or 25 June.)}

\textit{The letter's interest is concentrated in
section 3, the celebrated dedication-shuffle of the
\textit{Academica}. Cicero had drafted the dialogue
with Catulus and Lucullus as interlocutors ---
``high-born men, but no \textit{[Greek: philologoi]}''
--- and is now, at Atticus' prompting, transferring
the dedication to Varro, whose Antiochean
sympathies suit the dialogue's doctrine. The
two-year wait on Varro's own promised
\textit{[Greek: prosphonesin]} is glossed by the
Callippides image (the Aristophanic figure who runs
ceaselessly without progressing a cubit) and a tag of
Hesiod, \textit{Works and Days} 350: \textit{[Greek:
auto to metro kai ai ke dunai]} --- ``in the same
measure, if you can.'' The \textit{De Finibus}
\textit{[Greek: peri Telon syntaxin]} has meanwhile
been promised to Brutus. Section 1 hides a private
grief --- Attica's illness --- behind the formula of
the consolation that consoles its consoler. Section 2
acknowledges Atticus' role as Cicero's publisher: the
\textit{Pro Ligario}, finished a few weeks earlier,
has sold well, and Cicero promises to deliver his
future writing to him as \textit{praeconium}, his
auctioneer's crier. Section 4 turns to the Brinnius
inheritance and the Scapulan gardens, the
property-search that has been running since
13.07--08.}
