% Ad Atticum 13.52 --- headnote
% ad-atticum-13-52, 19 December 45 BC, Puteoli

\textit{Cicero to Atticus, written at the Puteolan villa on the
fourteenth day before the Kalends of January 45 BC --- Perseus
dateline \textit{Scr. in Puteolano xiv K. Ian. a. 709 (45)} ---
the morning after one of the best-known social encounters in
Roman literature: Julius Caesar's dinner with Cicero on 19
December 45 BC, the third day of the Saturnalia. Caesar, on his
way through Campania four months before the Ides of March, has
quartered himself with two thousand troops at Philippus's villa
next door and has come over to Cicero for the evening meal. The
letter, written immediately afterward, is Cicero's account to
Atticus of how it went. The famous opening note ---
\textit{o hospitem mihi tam gravem [Greek: ametameleton]} ---
sets the register: a guest who weighed on the host, but a visit
he is glad to have got through. The visit went off pleasantly,
the food was elegant, the conversation was literary; Caesar even
heard the news of his henchman Mamurra (almost certainly bad
news) and did not change his expression. He ate and drank
freely on the emetic regimen
(\textit{[Greek: emetiken... adeos]}) that allowed Roman
dignitaries to manage a banquet.}

\textit{This is the most political of the letters of this run not
because Cicero says anything political --- almost the opposite:
\textit{[Greek: spoudaion ouden] in sermone,
[Greek: philologa] multa}, nothing serious in the talk, plenty
of literary chat --- but because of what the silence cost.
Caesar comes to dinner under the protection of two thousand
men, and the armed escort draws up on either side of his horse
as he passes Dolabella's villa next door, ``nowhere else along
the way.'' The closing line --- ``a billeting
(\textit{[Greek: epistathmeian]}), as I said, was a burden, not
a bother'' --- gives a name in Greek to the political
condition: not Roman \textit{hospitium} between equals, but
something nearer the imposed quartering of a military master on
his subject. Cicero saves the political point for the Greek
word and keeps the Latin host's pose intact.}
