% Ad Familiares 6.19 --- headnote
% ad-familiares-06-19, late August 45 BC, Astura

\textit{Cicero to Quintus Lepta, written at Astura at the end of
August 45 BC --- Perseus dateline \textit{Scr. Asturae ex. m.
Sext. a. 709 (45)} (i.e. \textit{exeunte mense Sextili}, the
closing days of the month, not the middle as the works.yaml
metadata had it). Two short sections of unflashy, practical
counsel from one well-placed friend to another. Lepta is in
search of some footing in Caesar's circle, and Cicero --- writing
from his settled retreat on the seacoast --- thinks the project
not worth pursuing.}

\textit{The first section is travel logistics: Macula, a mutual
acquaintance, has done his duty as host, his Falernum estate
might serve as a roadside inn for the party so long as the roof
covers everyone, and even so Cicero will not give up Lepta's
Petrinum --- ``that villa, and that loveliness of the place, is a
property for staying in, not for stopping over.'' The second
turns to the substantive request: a \textit{curatio aliqua munerum
regiorum}, ``some sort of curatorship of the royal games'' (the
spectacles Caesar was preparing on his return from Spain). Cicero
has spoken to Oppius about it; Balbus, gouty and unwilling to be
approached, he has not seen. The advice is unsparing: Caesar's
inner court is full, anyone Lepta might displace will resent it,
and Caesar himself --- supposing he even registers Lepta's
existence --- will count the appointment as a favour conferred,
not received. ``We shall look for something with show in it,''
Cicero concludes, ``otherwise, in my judgement, the thing is not
merely not to be sought after but to be actively avoided.'' He
will wait at Astura until the dictator returns.}
