% Ad Atticum 13.35 --- headnote
% ad-atticum-13-35, 13 July 45 BC, Tusculum

\textit{Cicero to Atticus, written at the Tusculan villa on 13 July
45 BC --- Perseus dateline \textit{Scr. in Tusculano iii Id.
Quint. a. 709 (45)}. (Modern editions usually print 13.35 and
13.36 as a single composite letter; the Perseus segmentation
keeps them apart, and this translation follows that
segmentation.) Two short sections. The first opens with a flare
of indignation: ``What a disgrace! Your kinsman is enlarging
the city which he first laid eyes on only two years ago, and
which seemed to him too small to contain him in person.'' The
target is one of Atticus's Pomponian relatives, evidently
attached to Caesar's circle and now busy with the dictator's
plans for Rome --- a glancing reminder of how much of the daily
chatter between these two slides past the politics it skirts.}

\textit{The second section turns to two of the running threads of
the summer correspondence: the still-undecided dedication of the
\textit{Academica} (now given to Varro, ``at risk to yourself,''
the playful charge Cicero keeps levelling at Atticus); and the
deepening friendship between Atticus and Marcus Brutus, whose
walks together at the Tusculan villas have become a recurring
topic. The closing tricolon --- the more often I hear it, the
more gladly; the keener pleasure because you take pleasure in
it; the more certain because you say it --- is the kind of
small architectural flourish that the daily letters carry off
in passing.}
