% Ad Atticum 12.25 --- headnote
% ad-atticum-12-25, 22 March 45 BC, Astura

\textit{Cicero to Atticus, written from Astura on the
eleventh day before the Kalends of April 709 AUC ---
22 March 45 BC (the manuscript dateline: \textit{Scr.\
Asturae xi K. Apr.\ a.\ 709 (45)}). A short letter
focused on the gardens negotiation, two days after
12.24. Sicca has written to Cicero about Silius's
property and reports that the matter has been referred
to Atticus; Cicero is pleased with the property and
the terms, but wants the price in cash rather than in
kind, since he can barely live off the income-bearing
properties he already has. He works through the
financing aloud --- six hundred thousand sesterces to
be pressed out of Hermogenes, six hundred thousand he
sees lying at home, the remainder to be carried on
interest until Faberius or one of Faberius's debtors
can settle.}

\textit{The second section is briefer and more
revealing. Drusus's gardens, Cicero says, he prefers
much more than Silius's --- the two are not even
comparable. One thing only moves him in the matter,
and in it he knows he is \textit{infatuated}
(\textit{tetyphōsthai}, a Stoic-Cynic term for being
puffed up by vanity); but he asks Atticus to indulge
this delusion of his. As for what Atticus has
written about \textit{something to grow old with}
(\textit{engērama}) --- a quiet philosophical
counsel, that he should choose a property for the
long quiet years ahead --- Cicero answers flatly:
that matter is over (his daughter being dead, those
long years no longer interest him), and he is seeking
other things. The two Greek words carry the whole
weight: Atticus tactfully naming what the search is
properly for; Cicero answering that he knows his
motive is irrational, and asking that it be indulged
anyway.}
