% Ad Familiares 7.25 --- headnote
% ad-familiares-07-25, c. 24 August 45 BC, Tusculum

\textit{Cicero to M. Fadius Gallus, written from the Tusculan villa
around 24 August 45 BC, a few days after 7.24 and continuing the
same conversation. The torn-up letter Gallus had been distressed
about is safe at Cicero's house in town and can be retrieved at
will. The rest of the section turns on the season's open
secret --- the literary contest of \textit{Catones}, in which
Cicero himself, Brutus, Fadius Gallus, and others had each produced
encomia of Cato Uticensis --- and on the danger that hung over those
encomia: Caesar, returning from Spain, was nearer than they had
supposed. The proverbial Greek tag is rendered in transliteration:
``a Sardonic laugh,'' the bitter laugh of those whose joke goes
wrong. \textit{Manum de tabula} --- ``hand off the slate'' ---
is the schoolmaster's shout to the boy still scribbling; the
schoolmaster here, of course, is Caesar, and Cicero pictures the
``Cato-men'' \textit{[Catoniani]} being hoisted up for a flogging
\textit{[in catomum]}.}

\textit{The second section pivots from worry to delight in
Gallus's prose itself. The phrase \textit{cetera labuntur} ---
``and as for the rest, it falls away'' --- is something Cicero
wants kept utterly quiet, not even Gallus's freedman Apella to be
told: only the two of them write in that vein, and whatever the
strain, ``it is ours.'' Cicero closes with the proverbial
exhortation a finger's-breadth (\textit{transversum unguem}) never
from the pen --- \textit{the pen is the workman of speech} --- and
the matter-of-fact note that he is by now claiming back some hours
of the night for it himself. The Perseus dateline is \textit{circa
ix K. Sept.}, i.e.\ about 24 August 45 BC.}
