% Ad Atticum 12.44 --- headnote
% ad-atticum-12-44, 13 May 45 BC, Astura

\textit{Cicero to Atticus, written from Astura on the
third day before the Ides of May 709 AUC --- 13 May
45 BC (the manuscript dateline: \textit{Scr.\ Asturae
iii Id.\ Mai.\ a.\ 709 (45)}). One of the more
densely textured letters of the Astura sequence,
moving from literary politics to estate-buying to
the daily war news to the philosophical writing that
is the actual work of these months. Cicero received
Hirtius' anti-Cato pamphlet --- a piece of Caesarian
propaganda answering his own \textit{Cato} of the
previous year --- and has decided to have it
circulated through Atticus' copying-shop: the worse
Hirtius' attack, the louder the praise of Cato will
sound. The opening play on \textit{humane} (kindly,
in the human manner) --- Hirtius writes sympathetically
to Atticus, Atticus does not forward the letter to
Cicero, and each is praised for the deeper humanity
of his act --- is one of the small moments of
warmth that survive in the grief sequence.}

\textit{The second section is the estate search, now
running through the freedman Mustela: the access
needed is simply legal access for a buyer, which any
heir can provide. The Greek vocabulary turns
explicitly to what the estate is for ---
\textit{[Greek: eng\=erama]}, a place to grow old in;
\textit{[Greek: oikodespotika]}, befitting a
householder. Otho first, then Clodia; failing that,
Drusus must be played along (\textit{Druso ludus est
suggerendus}) or the Tusculanum used. The third
section turns to the war: Philotimus brings news
that Pompey the Younger has not been pinned at
Carteia and that a substantial war remains, but
Cicero is sceptical --- the informant is a regular
little Fulvinius, the proverbial false-news pedlar.
The final section is the one that gives the sequence
its second arc beneath the grief: Cicero has
finished two great treatises \textit{[Greek:
syntagmata]} at Astura, ``for in no other way can
I, so to speak, stray from my misery.'' This is
the \textit{Consolatio} alongside another work,
probably the lost \textit{Hortensius}, taking shape
as the medicine and the distraction at once.}
