% Ad Familiares 5.9 --- headnote
% ad-familiares-05-09, 11 July 45 BC, Narona (camp)

\textit{Publius Vatinius to Cicero, written from his camp at
Narona on the Dalmatian coast on 11 July 45 BC (Perseus:
\textit{in castris Naronae a.~d.~v Id.~Quint.~a.~709 (45)}). The
letter is incoming, not outgoing: Vatinius, hailed
\textit{imperator} by his troops after operations against the
Illyrian tribes, writes to his old patron in Rome. He had been
Caesar's man in 59 BC, the tribune through whom Caesar's
extraordinary commands were carried, and the target of Cicero's
fiercest invective in \textit{In Vatinium}; yet when Vatinius
came to trial in 54 BC Cicero --- by then forced into the
triumviral orbit --- defended him, and the two men have ever
since been on the warmest terms. \S 1 punningly registers the
inversion: ``Publius Vatinius, your client, has come,'' the
former defendant turned suppliant to the orator who once
prosecuted him.}

\textit{The letter opens with the soldier's epistolary formula
\textit{S.~V.~B.~E.~E.~V.}\ (``if you are well, I am glad; I
am well''), and runs on in Vatinius's brisk Caesarian voice:
ingratiating, slightly arch, but plain about what it wants.
What it wants is two things. First, that Cicero in Rome should
defend his standing against the malicious tongues at home while
he is on campaign --- the same office Cicero once performed at
his trial. The dispatch to the Senate on his \textit{res gestae},
he says, he has copied below; the Senate vote on a
\textit{supplicatio} will turn on how it is received. Second,
in the short closing section, he reports on a small private
favour: Cicero's runaway reader, last seen among the Vardaeans
of the Illyrian coast, will be hunted down by Vatinius's
soldiers and returned, even if the chase has to be carried into
Dalmatia. The two halves of the letter --- patronage at Rome and
manhunt in the field --- sit side by side without embarrassment,
which is itself an index of how completely Vatinius now belongs
to Cicero's circle.}
