% Ad Familiares 4.12 --- headnote
% ad-familiares-04-12, 31 May 45 BC, Athens

\textit{Servius Sulpicius Rufus to Cicero, written from
Athens on the day before the Kalends of June 45 BC ---
Perseus dateline \textit{Scr. Athenis pr. K. Iunias a. 709
(45)}, the closing line \textit{D. pr. K. Iun. Athenis}
confirming both date and place. The salutation
\textit{SERVIVS CICERONI sal. PLVR.}\ flips the usual
direction of the corpus: this is one of the few letters in
the \textit{Familiares} written \textit{to} Cicero rather
than \textit{by} him, and is preserved alongside Cicero's
own grief-stricken correspondence about the murder. Servius
Sulpicius Rufus is the great jurist and Caesarian
proconsul of Achaia in 46--45; M. Claudius Marcellus, the
Pompeian consul of 51 whose pardon by Caesar in September
46 occasioned Cicero's \textit{Pro Marcello}, was on his
way back to Italy after years of exile. He met his end at
the Piraeus, struck down at dinner by his own intimate
P. Magius Cilo --- a death so theatrical and unexpected
that Cicero will return to it again and again over the
coming months.}

\textit{The letter is one of the masterpieces of the
corpus. Sulpicius writes as a jurist and as a witness:
the chronology is tight (``the tenth day before the
Kalends of June \ldots the next day \ldots the third day
after that \ldots about the tenth hour of the night''),
the narrative reported in the indirect speech of the man
who brought the news (\textit{nuntiavit \ldots se ipsum
interfecisse \ldots se a Marcello ad me missum esse}), the
two wounds catalogued with anatomical exactness
(\textit{unum in stomacho, alterum in capite secundum
aurem}). The third section turns from chronicle to
epitaph: the antithesis \textit{vir clarissimus ab homine
deterrimo} (``a most illustrious man by a most worthless
one'') is the sentence that has fixed the murder in
literary memory, and the elaborate burial in the
Academy --- ``the most famous gymnasium in the world''
--- gives the philosopher-statesman a Platonic resting
place. This letter, together with Cicero's reply (Fam.\
4.6, already written) and Sulpicius's later
\textit{consolatio} (Fam.\ 4.5), constitutes a triptych on
the death of the Republic.}
