% Ad Atticum 12.21 --- headnote
% ad-atticum-12-21, 17 March 45 BC, Astura

\textit{Cicero to Atticus, written from Astura on
the sixteenth day before the Kalends of April 709
AUC --- 17 March 45 BC (the manuscript dateline:
\textit{Scr.\ Asturae xvi K.\ Apr.\ a.\ 709 (45)}).
Two days after 12.20, and the longest letter of the
cluster. The first section is an indignant
roll-call: Brutus, in a returned letter, has
suggested that Cato was the first speaker on
the punishment of the Catilinarian conspirators, and
that the consulars before him had been more lenient
than the praetorian-rank Caesar. Cicero recites
the consulars one by one --- Catulus, Servilius,
the Luculli, Curio, Torquatus, Lepidus, Gellius,
Volcacius, Figulus, Cotta, Lucius Caesar, Gaius
Piso, Manius Glabrio, Silanus and Murena --- to
correct the record. He put the question on Cato's
motion because Cato had embraced the same matter in
more luminous words and at greater length; the
division was taken on Cato's motion because Cato had
moved that the proceedings be entered into the
public record. Brutus's compliment that he was
``the best of consuls'' Cicero cuts to the bone:
``what enemy ever spoke more meagrely?''}

\textit{The remainder is the now-familiar cluster of
practical and personal business. The Roman gardens
for the \textit{fanum} are to be pursued (Drusus,
Lamia, Cassius are possible vendors). Terentia is
handled in a single hard sentence: ``If anything
cheats us, I would rather have it her fault to be
sorry for than mine.'' Ovia's 100,000-sesterce
debt is to be wound up through Eros. The letter
closes on the deepest note of the cluster: Atticus
has summoned Cicero to the Forum, but Cicero refuses
on principle. ``What is the Forum to me without the
courts, without the senate-house, while there run
into my sight men I cannot look at with composure?''
The therapy is his philosophical reading and writing
--- ``the act of a brave invalid, to take his
medicine'' --- and he begs Atticus not to call him
back from those remedies, lest he relapse. The
daggered crux in \S5 (\textit{aut quatenus}) is
preserved at the obelus; the construction of the
clause remains uncertain, but the broad sense
(``and do not allow me to be away, or with what
licence they grant'') is clear enough that the
argument carries.}
