% Ad Atticum 12.41 --- headnote
% ad-atticum-12-41, 11 May 45 BC, Astura

\textit{Cicero to Atticus, written from Astura on the
fifth day before the Ides of May 709 AUC --- 11 May
45 BC (the manuscript dateline: \textit{Scr.\ Asturae
v Id.\ Mai.\ a.\ 709 (45)}). Section 1 is logistical:
Cicero has fixed his itinerary, day after the Ides at
Lanuvium, then on to the Tusculanum or to Rome. The
Latin transmits sections 1, 2 and 4 only --- there is
no section 3 in the Perseus text --- and the
numbering is preserved as the editorial tradition
gives it.}

\textit{Section 2 is one of the most exposed moments in
the cluster. Cicero opens with a Greek tag,
\textit{[Greek: philaition sumphora]} ---
``misfortune is a great fault-finder'' --- to forgive
in advance the unfairness of the pressure he is about
to apply: unless he sees the \textit{fanum} actually
being built (not even completed), his
\textit{dolor} will rush upon Atticus, unjustly. The
ranked list of sites that follows --- first
Scapula's, then Clodia's, then Cusinius' and
Trebonius' if Silius refuses and Drusus deals
unfairly --- is the most explicit prioritisation in
the whole estate-search. The closing self-correction
is harder still: ``you, who now accuse me even more
sharply than your usual gentleness can support, but
you do it from the deepest love and, perhaps,
overborne by my own fault.'' The shrine is now
described as \textit{summa levatio vel ... una}: the
supreme relief, or in truth the only one. The
cluster's grief register (\textit{dolor},
\textit{levari}) is now bound directly to the shrine
project --- the building has become the only viable
form of consolation.}

\textit{Section 4 returns to Hirtius' draft of the
anti-Cato. Cicero now names it: a \textit{[Greek:
proplasma]}, a preliminary plaster sketch of the
attack Caesar is preparing. The letter closes with a
striking ethical conditional: unless the shrine is
finished this summer, ``I shall not consider myself
freed of the crime'' --- the \textit{scelus} being,
presumably, a debt of piety to Tullia still unpaid.}
