% Ad Familiares 9.11 --- headnote
% ad-familiares-09-11, ~20 April 45 BC, Atticus's Ficuleanum estate

\textit{Cicero to P. Cornelius Dolabella, written at
Atticus's Ficuleanum estate a little after the twelfth
day before the kalends of May --- Perseus: \textit{in
Attici Ficuleano paulo post xii K. Mai. a.~709 (45)}, that
is, shortly after 20 April 45 BC. Tullia had died in
mid-February. Dolabella, though he and Tullia had
divorced, had written Cicero a letter of condolence; this
short note is Cicero's reply. Dolabella was abroad on
campaign with Caesar; Cicero was passing his bereavement
between his own and Atticus's country houses, working on
the \textit{Consolatio} for himself and beginning the
\textit{Academica}.}

\textit{The letter is two short paragraphs, both bent
around the same thought: gratitude to Dolabella for
loving him visibly, and a frank statement that Cicero is
not himself. The opening sentence is a load-bearing
hyperbole --- \textit{vel meo ipsius interitu mallem
litteras meas desiderares}, ``I should rather you had
missed letters from me by reason of my own death than by
this calamity'' --- and the disavowal in the middle of
section~1 is the philosophical pivot: he has not forgotten
\textit{hominem me esse}, that he is a man, and he does
not think he must \textit{succumb to fortune}; but the
old cheerfulness is gone. \textit{Proelia te mea causa
sustinere} at the head of section~2 evidently refers to
disputes in which Dolabella has defended Cicero's name in
Cicero's absence --- Cicero asks not to be vindicated so
much as to be loved. The brevity is signalled at the
close: he is not yet \textit{satis confirmatus ad
scribendum}, not yet steady enough for writing.}
