% Ad Familiares 9.22 --- headnote
% ad-familiares-09-22, June or July 45 BC, Tusculum (probably)

\textit{Cicero to L. Papirius Paetus, written probably at the
Tusculan villa in June or July of 45 BC --- Perseus dateline
\textit{Scr. fort. in Tusculano m. Iun. aut Quint. a. 709 (45)},
month-precision uncertain between the two. Provoked, evidently,
by a freely worded letter of Paetus that used a bawdy term
without apology, Cicero responds with a mock-Stoic disquisition,
in five sections, on whether obscenity inheres in the word or in
the thing it names. The structure is a small Academic exercise:
expound Zeno's thesis, marshal counter-examples from comedy and
tragedy, work the conclusion through pairs of words for the same
object, then end with a Stoic tag and a comic concession. The
closing date-pun --- \textit{honorem igitur K. Martiis} ---
treats the Kalends of March as something one must be polite to,
just as one is polite to the obscene words.}

\textit{The voice is the late Paetus voice: high learning poured
into low diction, mock-formal philosophising about latrine
matters, the sage and the bath both invoked in a single
paragraph. One Greek phrase closes the argument: \textit{ho
sophos euthyrr\=emon\=esei}, ``the wise man will speak
straight-forwardly,'' a Stoic technical formula on free speech
(\textit{parrh\=esia}). Cruxes: \textit{libertatem loquendi} I
take as ``freedom of speech,'' the technical Stoic
\textit{parrh\=esia} that Paetus has just exhibited. The
fragmentary stage quotations --- one from Naevius's
\textit{Demiurgus}, one a song associated with Roscius, several
unattributed --- I render closely enough for the punch-lines to
land without footnotes; the point each time is that the matter
is more brazen than the wording. The famous catalogue of false
or shifting obscenities (\textit{penis}, \textit{anus},
\textit{bini}, \textit{mentula}, \textit{divisio},
\textit{intercapedo}, \textit{batuit}, \textit{depsit},
\textit{testes}, \textit{colei}) I have left the Latin tags in
place where the joke turns on the Latin word, with the gloss
inline, since a fully Englished version would lose the linguistic
point of the letter. \textit{Suppedit} I have construed in the
sense ``pisses under (himself)'' that the bath-contrast requires;
others (Shackleton Bailey) take it of farting and emend, but the
manuscript reading works once the comic shape is admitted.}
