% Ad Atticum 13.19 --- headnote
% ad-atticum-13-19, 30 June 45 BC, Arpinum

\textit{Cicero to Atticus, written from Arpinum on
\textit{prid.\ K.\ Quint.} --- 30 June 709 AUC, 45
BC, the last day before the Kalends of Quintilis and
the close of the four-letter Arpinum sequence (13.14,
13.16, 13.17, 13.19). The framing items are
domestic: Attica is doing well enough that she has
asked Atticus not to be gloomy, and the situation
generally is \textit{akindyna}, ``without danger.''
Balbus and Oppius approve of the just-published
\textit{Pro Ligario} and have forwarded it to Caesar
--- Atticus' part in that placement is now
explicit.}

\textit{The heart of the letter, in section 3 and the long
section 4, is one of the most lucid statements
Cicero ever made about his own practice as a writer
of dialogue. He explains the rule that had kept him
from putting Varro in (\textit{philendoxos}, no
flattery of the living); the casting of the
\textit{Academica} now redone with Varro as
Antiochean speaker against Cicero himself (with
Atticus the third interlocutor); the contrasting
Heraclidean manner of the \textit{De Re Publica} and
\textit{De Oratore}, where the speakers are old
masters and the author keeps silent
(\textit{k\=ophon pros\=opon}, the ``mute mask''); and
finally the Aristotelian manner of the
\textit{De Finibus}, where the author leads. The
final clause confesses the awkwardness Cicero has
just created for himself: Antiochus' arguments are
\textit{pithana} (``persuasive''), and rendered with
care by Cicero they do not let his own Academic
position come out as the stronger one. Greek terms
run thick: \textit{akindyna}, \textit{philendoxos},
\textit{akatal\=epsian}, \textit{k\=ophon pros\=opon},
\textit{Aristoteleion}, \textit{az\=elotyp\=eton},
\textit{logik\=otera}, \textit{hermaion},
\textit{pithana}; together with the Greek title
\textit{peri tel\=on} (\textit{De Finibus}, ``On
Ends'') and the adjective \textit{Peripat\=etika}
attached to Piso's part. One textual crux:
\textit{$\dagger$easque$\dagger$}, daggered by
editors, between the two sentences on the
Antiochean parts; I render the surrounding sense
without trying to repair it.}
