% Ad Atticum 12.32 --- headnote
% ad-atticum-12-32, 28 March 45 BC, Astura

\textit{Cicero to Atticus, written from Astura on the
fifth day before the Kalends of April 709 AUC ---
28 March 45 BC (the manuscript dateline:
\textit{Scr.\ Asturae v K.\ Apr.\ a.\ 709 (45)}). The
opening paragraph is the most personal in the
cluster. After a few lines on the Silius
negotiation --- Egnatius is now the preferred
channel, and the Silius deal looks dead --- Cicero
shifts (``this part to you in my own hand'') to the
Publilia question. His young second wife has
written to say that her mother and Publilius (her
brother) propose to come to him at Astura, with
Publilia in the party, if he will allow it. Cicero
suspects the letter was not really hers; he has
written back firmly that the visit is impossible.
The one way to prevent their coming, he tells
Atticus, is to leave Astura himself: ``I would
rather not, but it is necessary.'' He asks Atticus
to find out how long he can safely stay. The
text carries a long-standing crux at \dag cum
Publilio loqueretur \dag, preserved here.}

\textit{Section 2 turns to the practicalities of
young Marcus' Athens studies. The expenses are to
be capped at the rents from the Argiletum and
Aventine properties --- enough, Cicero promises, to
keep his son at the level of Bibulus, Acidinus, or
Messalla, the other young aristocrats then at
Athens. Atticus is asked to vet the tenants, secure
prompt payers, and budget travel-money and outfit.
A characteristic Ciceronian detail closes the
letter: no horse will be needed at Athens, and as
for the road, the household has more mounts than it
ever required.}
