% Ad Familiares 5.15 --- headnote
% ad-familiares-05-15, December 45 BC, Astura

\textit{Cicero to L. Lucceius son of Quintus, written from
Astura a few days after the previous letter to him (Fam.\
5.14), in December 45 BC --- Perseus dateline
\textit{Scr. Asturae paucis diebus post ep. xiv a. 709 (45)}.
Lucceius is the historian (and addressee of the famous
\textit{Fam.}\ 5.12, in which Cicero had asked him to write
up his consulship in monograph form) and a friend of long
standing, now likewise in retirement and in poor health. He
had evidently written to scold Cicero, in ``the most gentle
and most loving words,'' for his continued absence from
Rome --- the ``one cause which you suspect'' --- when in
fact the deeper grief was the death of Tullia in February
of this year, the wound for which Cicero in the opening
line confesses there is no remedy.}

\textit{The letter is a small, finely shaped piece of
self-portraiture. It moves through four refuges in turn
and lets each fail: friends (\S 2), then the
companionship of Lucceius himself (the rest of \S 2),
then literary work (\S 3 --- ``those very studies seem to
shut me out from their harbour''), and finally retreat
from Rome itself (\S 4). The architecture is the central
philosophical move of these months in Astura, the same
move that powers the \textit{Tusculans}: take stock of the
remedies, find them insufficient, and continue. The
parenthesis ``(I would say `delightful,' had I not put
that word out of service for every season)'' is one of
the most quietly devastating lines in the
correspondence --- the lexicon itself has narrowed under
grief. The closing promise \textit{propediem te igitur
videbo} (``soon, then, I shall see you'') stands at the
end of a letter whose whole argument is that being
together has not in fact happened, even when it was
geographically simple.}
