% Ad Familiares 5.14 --- headnote
% ad-familiares-05-14, 9 May 45 BC, Rome

\textit{Lucius Lucceius to Cicero, written at Rome on 9 May
45 BC (Perseus: \textit{Romae vii.~Id.~Mai.~a.~709 (45)}). The
letter is incoming, one of the surviving consolations sent to
Cicero in the weeks after Tullia's death at Tusculum in
mid-February. Lucceius, historian and senator and a friend of
many years (he is the same man Cicero in 56 BC had begged in
\textit{Fam.}~5.12 to write up his consulship as a monograph),
addresses his friend with the same patrician formality the
heading registers --- \textit{L. Lucceius Q.~f.}\ to \textit{M.
Tullius M.~f.}\ --- but with a brisk and impatient affection,
the affection of a man who has been waiting for his friend to
come back to him and to himself.}

\textit{The argument of the letter runs in two halves and an
appeal. \S 1 takes the more generous reading: if Cicero has
absented himself from Rome to write and to take up again the
old learned occupations, Lucceius rejoices and does not find
fault. But \S 2 turns to the other possibility, the one
Lucceius plainly suspects --- that Cicero has surrendered
himself to tears and is doubling the anxieties his own wisdom
should lighten --- and the three anaphoric questions
(\textit{tu solus aperta non videbis ... tu non intelleges ...
non intelleges ...}) press the philosopher's reproach in the
voice of a man who knows how to lean on his friend. \S 3 drops
the argument and pleads: come back to our company, to the way
of life that is either our common one or your own proper one
(\textit{vel nostram communem vel tuam solius ac propriam}).
Cicero's reply, the brief and grateful \textit{Fam.}~5.15, was
sent within days; he came back to Rome but not, as his other
letters of this period show, to himself.}
