% Ad Atticum 13.9 --- headnote
% ad-atticum-13-09, mid-to-late June 45 BC (Perseus dateline garbled), Tusculanum

\textit{Cicero to Atticus, written from the Tusculanum
in the second half of June 709 AUC --- mid-to-late
June 45 BC. The Perseus dateline is garbled
(\textit{Scr.\ in Tusculano xttv K.\ Quint.\ a.\ 709
(45)}); the editorial tradition has variously
restored it as \textit{xiv K.\ Quint.} (18 June) or
read the day-element as ``Kalends of Quintilis'' (1
July). Either reading places the letter in the
Tusculanum sequence that runs from Atticus' visit
through Cicero's planned removal to Arpinum.
Dolabella has come up from Rome and stayed for a
long morning's talk; Trebatius is now an
houseguest; and Torquatus has arrived in time to
hear Dolabella report the warmth with which Cicero
had pleaded his cause.}

\textit{The texture is unusually thick with Greek ---
\textit{[Greek: ektenesteron]} ``more attentive,''
\textit{[Greek: philostorgoteron]} ``more
affectionate,'' \textit{[Greek: aphata]}, \textit{[Greek:
adiegeta]} ``unspeakable, not to be recounted,''
\textit{[Greek: eukairos]} ``at just the right
moment'' --- the standard private-language of the
Atticus correspondence at moments of feeling and of
discretion. The reference to Quintus is one of the
half-confidences Cicero will not commit to writing
even to Tiro's dictation; the brother quarrel is at
this point the running grief beneath the philosophical
calm. Section 2 turns to the Brutus--Claudia divorce
(Nicias' news, the question whether it will be ratified)
and to Cicero's plan to get to Arpinum before
Caesar's return puts travel out of reach.}
