% Ad Atticum 13.31 --- headnote
% ad-atticum-13-31, 28 May 45 BC, Tusculanum

\textit{Cicero to Atticus, written at the Tusculan villa on 28 May
45 BC --- Perseus dateline \textit{Scr. in Tusculano v K. Iun. a.
709 (45)}. The second of the day's letters: Cicero opens with the
dating, ``On the fifth before the Kalends, in the morning,'' and
moves through four short sections covering the gardens
financing, a request for Dicaearchus's books, a sudden settled
decision about the letter to Caesar, and the rival bidder Otho.}

\textit{Three Greek phrases punctuate the page. \textit{katabaseōs},
``of the \textit{Descent},'' is a title --- Dicaearchus's lost
treatise (perhaps \textit{The Descent into the Cave of Trophonius})
which Cicero wants alongside the political dialogues for his
Varro-revision work. \textit{kekrika}, ``I have decided,'' is the
clipped Greek perfect that Cicero uses to mark a settled
intellectual judgement: he has decided not to send the proposed
letter of advice to Caesar after all, because the only counsel he
could honestly give --- wait to settle affairs at home before the
Parthian campaign --- is what Caesar is reportedly saying anyway.
The line that follows is bitter: ``Let us throw all that aside
and at least be half-free; which we shall achieve by keeping
silent and lying low.'' The closing paragraph returns to the
Scapulan gardens, with a recalled price-comparison (Albanius paid
11,500,000 HS for a thousand iugera next door), and a jab at the
rival Otho's father --- \textit{o gulam insulsam! pudet me
patris}: ``What a tasteless palate! I am ashamed of his father.''}
