% Ad Atticum 13.21 --- headnote
% ad-atticum-13-21, 29 July 45 BC, Astura

\textit{Cicero to Atticus, written from the seaside villa at Astura
on 29 July 45 BC --- Perseus dateline \textit{Scr. Asturae iv K.
Sext. a. 709 (45)}. The letter falls outside the dense Tusculan
daily-letter run of mid-May to early June and belongs to the later
Astura stretch of the same summer: Cicero has retreated to the coast,
is awaiting news from Dolabella and from Quintus, and is meanwhile
chasing political gossip --- Torquatus, Pansa, Critonius, Metellus
and Balbinus --- as the messages from town come in.}

\textit{The heart of the letter, however, is a piece of pure
scholarly second-guessing: in the \textit{Academica} Cicero had
adopted Atticus's suggested rendering of the Greek
\textit{epoch\=e} (the Sceptic ``suspension of judgement'') by the
Latin verb \textit{inhibere}, on the assumption that it described
holding the oars steady. Watching a ship come in at the villa
yesterday, he has discovered that the term in fact denotes a
reverse-stroke that moves the ship --- the opposite of suspension.
He wants the original wording restored in the book, and Atticus
to warn Varro in case he has already passed on the change. Three
Greek phrases sit at the joints: \textit{epoch\=ei} twice, for
the Sceptic technical term itself, and \textit{probol\=en}, the
boxer's defensive guard, in Carneades's analogy. The closing
section preserves a daggered crux in the Perseus text
(\textit{si quid \dag{}esset certe ne\dag{}}) which I have left
daggered.}
