% Ad Familiares 13.66 --- headnote
% ad-familiares-13-66, late January or early February 45 BC, Rome

\textit{Cicero at Rome to P. Servilius Isauricus, proconsul of Asia,
written at the end of January or the beginning of February 45 BC
(the manuscript dateline: \textit{Scr. Romae vel ex. m. Ian. vel in.
Febr. a. 709}). Servilius had been Caesar's consular colleague in
48 BC; the formula \textit{conlegae}, which Cicero uses elsewhere
in this cluster (13.68, 13.72), is the Roman address-title that
marks ex-consular peerage, since Cicero had himself held the
consulship in 63 BC. Here the salutation is the simpler
\textit{M. Cicero P. Servilio s.} The letter is a recommendation
on behalf of A. Caecina, son of Cicero's old friend and client
of the Caecinae --- an Etruscan family long attached to the
Servilii. The younger Caecina had been a partisan of Pompey and
the author of a pamphlet against Caesar, for which he had been
condemned and exiled; after surrendering himself to Servilius's
\textit{iustitia} he had crossed to the safer ground of the
province of Asia, and Cicero asks Servilius both to help him
gather in what is left of his old business interests and to
protect him in everything else, against the hope that ``the
clemency of your colleague'' --- that is, of Caesar --- may yet
restore him outright.}

\textit{The date sits painfully near the death of Cicero's daughter
Tullia, which fell about the middle of February 45 BC. If, as the
day-precision range suggests, the letter belongs to early February,
it was written within days --- on either side --- of that loss. The
letter itself betrays no grief; it is a workmanlike act of
patronage, undertaken for a family Cicero owed, in a register that
elsewhere in the same months he could not always sustain. The
Perseus dateline is range-precision; \texttt{meta/works.yaml} may
carry a year-precision placeholder of \texttt{-0045-01-01} that
should be tightened toward \texttt{-0045-02-01} (or similar) with
the file prefix kept at \texttt{045bc-}.}
