% Ad Atticum 12.37 --- headnote
% ad-atticum-12-37, 4 May 45 BC, Astura

\textit{Cicero to Atticus, written from Astura on the
fourth day before the Nones of May 709 AUC --- 4 May
45 BC (the manuscript dateline: \textit{Scr.\ Asturae
iv Non.\ Mai.\ a.\ 709 (45)}). The opening section is
domestic relief: two letters from Atticus arrived
together, Pilia and Attica are in good health, and
Brutus' long-delayed letter (thirteen days in
transit) has arrived. Cicero forwards it with a copy
of his own reply --- one of the few moments in the
sequence where the wider Roman correspondence breaks
through the grief.}

\textit{The second section is the conceptual hinge of
the estate search. Cicero now admits that he wants a
\textit{celebritas} --- a frequented spot --- not the
\textit{solitudo} Astura had given him. The shrine
needs witnesses: a \textit{fanum} surrounded by
strangers will outlast a \textit{fanum} in a private
garden, and the Scapula property offers the greatest
throng, with the further advantage of being near
where Atticus would be staying. Otho is to be
approached if he is in Rome; Drusus is selling and
will not be passed over if nothing better turns up;
the Scapula heirs are the way to keep all options
live. The closing section turns to the unresolved business
with Terentia --- the post-divorce financial settlement
that has dragged through this whole period --- where
Cicero needs Atticus' \textit{gratia} (influence) as
much as his \textit{auctoritas} (authority). The
ex-wife's affairs, like the shrine and the second
wife Publilia, are being negotiated through the same
patient intermediary.}
