% Ad Atticum 13.30 --- headnote
% ad-atticum-13-30, 28 May 45 BC, Tusculanum

\textit{Cicero to Atticus, written at the Tusculan villa on 28 May
45 BC --- Perseus dateline \textit{Scr. in Tusculano post op xxxi v
K. Iun. a. 709 (45)}. The fifth Kalends of June. The day's note is
short and two-thirds practical: the Faberian debt is still not
discharged; the auction (for the Scapulan gardens) is two days
off; Atticus must keep Faberius courted, however much such
courtship grates on Cicero, who calls it a near-crime. He expects
Atticus himself the day after tomorrow.}

\textit{The second paragraph is a research request that lights up
the literary project running underneath the daily letters: the
Academica being re-cast for Varro. Cicero is staging a dialogue in
the manner of Dicaearchus --- a \textit{politikon syllogon},
``political assembly,'' a set-piece gathering of named historical
figures at Olympia or some similar venue --- and he needs to know
who the ten Roman commissioners sent to Mummius Achaicus in 146 BC
actually were. Polybius does not list them. Hortensius gave him
Tuditanus, but Libo's chronology makes that impossible (Tuditanus
was praetor only fourteen years after Mummius's consulship, hence
too young in 146). Two Greek phrases: \textit{kolakeiai},
``flatteries,'' the deprecating word for the social courtship
Atticus must keep paying Faberius; and \textit{politikon syllogon},
``political assembly,'' the technical name for Dicaearchus's
literary form. The Mummius-commission research will continue in
13.32 (where Cicero proposes Postumius instead).}
