% Ad Familiares 9.7 --- headnote
% ad-familiares-09-07, end of May 46 BC, Rome

\textit{Cicero to Varro, written at Rome at the end of May 46 BC
(Perseus: \textit{Romae ex.~m.~Maio a.~708 (46)}). The canonical
numbering of book 9 places this letter after \textit{Fam.}~9.6, but
the Perseus dateline puts it earlier in the calendar than either
\textit{Fam.}~9.5 (early June) or \textit{Fam.}~9.6 (last week of
June); chronologically it falls before both. The substance bears
this out: Caesar's return from Africa is still being mapped out
("when, by what route, where, we know nothing as yet"), whereas
in 9.6 his arrival is being looked for any day. The letter is the
first of the Varro sequence in which the war is openly behind them
and the question is how the regime will treat its returned
opponents: Cicero has heard about the killing of the younger
Lucius Caesar in Africa, says to himself ``What will this fellow
do to me, given what he did to his own father?'' --- and goes on
dining, all the same, at the houses of ``these men who are now
masters.'' \textit{tempori serviendum est}: one must serve the
time.}

\textit{The letter is the densest cluster of Greek and verse
tag-ends in the surviving Varro correspondence. The half-line from
\textit{Iliad} 10.224 (Diomedes to Nestor, ``when two go
together'') stands in for Cicero's wish to be at Varro's side for
whatever was coming; the Ennian hexameter on the bristling land of
Africa marks the recent campaign; the Stoic technical
\textit{apoproēgmena} (``dispreferred,'' the negative half of the
indifferents) names the whole present situation in one word; and
the proverbial \textit{polloi mathētai kreissones didaskalōn},
``many pupils outdo their teachers,'' is turned, with light irony,
on Dolabella as the latest source of regime news. The register is
that of two of the most learned Romans alive trading shorthand
across a political vacuum.}
