% Ad Familiares 4.9 --- headnote
% ad-familiares-04-09, 13 August 46 BC, Rome

\textit{Cicero to M. Marcellus, written at Rome --- Perseus:
\textit{Romae post Id.~Sext.~a.~708 (46)}. The Perseus dateline
is identical to that of 4.7, and the conventional dating places
the two letters on the same day: 13 August 46 BC. The opening
sentence states the relation between them. A few days earlier
Cicero had given a letter for Marcellus to Quintus Mucius
Scaevola; now, with the freedman Theophilus about to set out for
Mytilene, he writes a second on the same theme. The two letters
should be read as a pair.}

\textit{This second letter pushes harder than 4.7. Cicero
anticipates Marcellus's principled objection --- that to return
is to commit oneself to saying things one does not believe and
doing things one does not approve --- and meets it with the
oldest Stoic-Roman commonplace: \textit{tempori cedere}, to
yield to the times, is the wise man's part, and silence is still
permitted even where speech is not. The long middle paragraph
turns the argument toward Caesar's victory. A counterfactual ---
what kind of victor would Pompeius have been? --- gives way to
the unsparing claim that all victory in civil war is miserable,
even when it goes to the better side, because the victor must
do many things at the will of those who won for him. The closing
section, with its near-Stoic argument from contraries (\textit{si
magni animi non esse supplicem \dots vide ne superbi sit
aspernari}), prefigures the rhetorical shape of the speech
\textit{Pro Marcello} that Cicero would deliver only weeks
later before Caesar in the Senate.}
