% Ad Familiares 9.6 --- headnote
% ad-familiares-09-06, between 20 and 25 June 46 BC, Rome

\textit{Cicero to Varro, written at Rome between the twelfth and
the seventh day before the Kalends of Quintilis (= July) 46 BC
(Perseus: \textit{Romae inter xii et vii K.~Quint.~a.~708 (46)}),
that is, in the last week of June, with Caesar's return from the
African campaign now expected at any moment. The letter opens as
a piece of practical intelligence --- where to prepare hospitality
for the dictator: he had written that he would come to his villa
at Alsium, but Hirtius, Balbus, and Oppius have advised him to
land at Ostia instead, and Cicero passes this on so that Varro can
have a guest-room ready in either place. The naming of the three
Caesarian intimates is itself the point of the passage. Cicero is
careful to record both that he is on familiar terms with them and
that they have shown themselves friends of Varro; he is mapping, in
the simplest possible terms, the channel through which two
ex-Pompeians of consular and senatorial rank now communicate with
the regime.}

\textit{The middle of the letter is the most candid statement
Cicero ever made about his own retrospective judgment of the war.
``It is not the same thing,'' he says, ``to bear what must be borne
and to approve what is not to be approved''; and then, with a
remarkable concession: he no longer even knows what he would now
disapprove of, beyond the war's beginnings, ``for those were a
matter of will.'' Our friends, he says, wanted the war; Caesar did
not so much want it as not fear it. He was always with Varro, he
adds, in dreading that the extreme of all evils was a victory in a
civil war, whichever side won --- and he confesses he had feared
even the victory of the side to which they had come, since the most
idle of them were making cruel threats. The end of the letter is
the manifesto of \textit{Fam.}~9.2 brought back in different
language: in tempests like these, Varro stands almost alone in
port; his Tusculan days are ``the very pattern of a life''; and
since studies, in the verdict of great men, carry a kind of
furlough from public duty, with the commonwealth's leave they will
draw on them to the full.}
