% Ad Familiares 9.13 --- headnote
% ad-familiares-09-13, late 46 or early 45 BC, Rome

\textit{Cicero to P. Cornelius Dolabella, written at Rome at the
turn of the year --- Perseus: \textit{Romae ex.~a.~708 (46) aut
in.~a.~709 (45)}, that is, late December 46 BC or the opening
weeks of 45. The letter is a recommendation on behalf of two men
of Cales, C. Subernius and the heir of M. Planius, both close
connections of Cicero's intimate Lepta. The two had gone out to
Spain with Varro before the civil war to escape it, on the
calculation --- universal at the time --- that after Afranius's
defeat no more fighting could break out there; instead they were
caught up in the rising of the Pompeian remnant under T. Quinctius
Scapula, which Cn. Pompeius the younger then took over and
hardened into the war that would end at Munda in March 45.
Caesar's lieutenant Dolabella was in a position to procure their
return.}

\textit{The argument is built in three steps: the misfortune was
fortune's not the men's, Caesar has already granted them their
lives so there is no political stake left in keeping them out, and
all that remains in the way of their coming home to die among
their own is the length of the journey. The letter is fluent and
warm but not relaxed; the run of triple intensifiers ---
\textit{maiore cura, studio, sollicitudine animi} --- and the
closing \textit{vehementer te etiam atque etiam rogo} mark how
hard Cicero is leaning on the relationship. \textbf{Metadata
note: the \texttt{meta/works.yaml} entry carries the date
\texttt{-0046-02-18}, which conflicts with the Perseus dateline
of late 46 or early 45 BC; the entry should be corrected.}}
