% Ad Familiares 9.15 --- headnote
% ad-familiares-09-15, late 46 BC, Rome (intercalary month)

\textit{Cicero to L. Papirius Paetus, written at Rome at the beginning
of the earlier intercalary month inserted by Caesar at the end of 46
BC --- Perseus: \textit{Romae in.~interc.~priore a.~708 (46)}.
Caesar's calendar reform of 46 carried two intercalary months between
the old November and December to bring the civil year back into step
with the solar year, so the dateline locates the letter very near the
true end of 46 BC, after the more substantial Paetus letters of the
preceding summer (9.16, 9.17) and not long before the New Year letter
9.10 to Dolabella. \textbf{Metadata note: the \texttt{meta/works.yaml}
entry carries the year-precision placeholder \texttt{-0046-04-20};
the Perseus dateline locates the letter more precisely at the
opening of the earlier intercalary month and so at the very close
of 46 BC. The entry should be revised when the metadata is
consolidated.}}

\textit{Five short sections in the established Paetus register, all
appetite, wit, and political ruefulness. Cicero acknowledges two
recent letters from Paetus, takes the occasion of a worry about
Paetus's health to praise him as the last living specimen of the old
in-town Roman gaiety --- a \textit{salt} fresher and sharper than the
Attic kind, and now smeared off Latium by an influx first of foreigners
and then of \textit{bracatae et Transalpinae nationes}, the trousered
Transalpine tribes. He runs through the canonical witty Romans
(Granius, Lucilius, the Crassi, the Laelii) and lands the joke as
flattery. The second half turns rueful: Paetus had urged him not to
retreat to a country house at Naples, citing Catulus as precedent;
Cicero replies that the comparison fails, because in Catulus's day
the steersmen still sat on the stern, whereas now there is barely
room left in the bilge. The Senate's decrees, he adds, are now
drafted in the house of one of Caesar's intimates and reach the kings
of Armenia and Syria over his name before he has heard the matter was
even raised --- and he has had thank-you letters from kings on the
world's edge whom he did not know had been so addressed, indeed whom
he did not know had been born. So: while Caesar (\textit{noster
praefectus moribus}, our prefect of morals) remains in town, Cicero
will behave; the moment Caesar leaves, he is for Paetus's mushrooms.
The closing lines turn practical --- the Sullan house, the builders,
the sumptuary-law allowance --- and shade back into banter.}
