% Ad Familiares 6.12 --- headnote
% ad-familiares-06-12, late November 46 BC, Rome

\textit{Cicero to Ampius Balbus, written at Rome at the end of
November 46 BC (works.yaml \texttt{-0046-11-15} at month-precision;
the Perseus dateline reads \textit{Romae ex.~m.~Nov.~a.~708 (46)},
``at the end of the month of November'' --- a tightening to
\textit{circa} ${-}0046$-$11$-$28$ is plausible, but
month-precision covers it). Titus Ampius Balbus, an old Pompeian
who had been tribune in 63 with Cicero and was consequently in
exile after Pharsalus, has now had his recall set in motion;
Cicero, with the lobbying of Pansa, Hirtius, Balbus, Oppius,
Matius, Postumius, and Tillius Cimber, has secured the pardon ---
the formal \textit{diploma} (the imperial warrant of safe-conduct
and recall) is being held back for tactical reasons, but the
substance is done. The letter is a Trebianus-cluster sibling: the
news is good, but Balbus is reading it in exile with his wife
Eppuleia and his daughter Ampia away from him, and Cicero writes
with one eye on the husband and father whose nerves had begun to
go before the despatches caught up with him.}

\textit{No Greek. The roll of Caesar's intimates in section~2 is
a small Who's Who of the regime at the close of 46: Pansa,
Hirtius, the other Balbus (L.~Cornelius), Oppius, Matius,
Postumius --- with Tillius Cimber, three years later one of the
Liberators, here briefly the man who got Balbus his pardon. The
nickname \textit{tuba belli civilis}, ``the trumpet of the civil
war,'' is the live coinage of the moment, and Cicero quotes it
exactly. The doctrinal pivot is the Trebianus pivot in another
key: the brave-and-wise man's consciousness of his own conduct
(\textit{conscientia factorum et consiliorum}, recalling
\textit{Fam.}~6.10.4) is now the consolation, and ``learning and
letters'' (\textit{doctrina ac litterae}) is the single refuge ---
formerly a pleasure of prosperity, now a means of survival. The
closing ring-form (\textit{ut ad initium revertar}) is Cicero
deliberately marking the rhetorical shape: open with the
ratified pardon, close with it.}
