% Ad Familiares 13.37 --- headnote
% ad-familiares-13-37, 46 BC, Rome

\textit{Cicero to Manius Acilius Glabrio, proconsul of
Sicily, written from Rome in 46 BC (the manuscript
dateline: \textit{Scr. Romae, ut videtur, a. 708 (46)}).
The sixth surviving letter of the Acilius cluster (Fam.\
13.30--39). The beneficiary is Hippias, son of
Philoxenus, of Calacte --- a small Greek town on the
north coast of Sicily --- with whom Cicero stands in the
double relation of \textit{hospes} and \textit{necessarius}.
The reported grievance is a confiscation: Hippias's
property, Cicero understands, has been taken into the
town's possession under another man's name, in
contravention of the local statutes of Calacte. He asks
the proconsul to set the matter right.}

\textit{Distinct in the cluster for being a substantive
appeal rather than a routine commendation. Cicero
structures the asking on two registers: the equity of
the case (if the facts are as reported, the proconsul's
\textit{aequitas} should suffice without intervention)
and, redundantly, the writer's standing (which is
invoked anyway, against whatever facts may emerge).
The qualifier \textit{quoquo modo autem se res habet} is
characteristic --- Cicero hedges over the facts, which he
knows only at one remove, while not hedging at all on
the request. The closing formula echoes the cluster's
common cadence (\textit{quantum tua fides dignitasque
patietur}, with the deference to the governor's
scruple), and the asseverative \textit{vehementer
gratum} carries the strong-thanks register.}
