% Ad Familiares 6.6 --- headnote
% ad-familiares-06-06, 1 October 46 BC, Rome

\textit{Cicero to Aulus Caecina, written at Rome around the
kalends of October 46 BC (Perseus: \textit{Romae circ.~K.~Oct.
a.~708 (46)}). Caecina, son of the Caecina whom Cicero had
defended in 69 BC (the \textit{Pro Caecina}), was a man of the
Etruscan nobility whose family had long preserved the
\textit{disciplina Etrusca} of haruspicy; he was himself a
learned antiquarian. In the late fifties and early forties he
had circulated an anti-Caesarian pamphlet --- the
\textit{Querelae} (``Complaints'') --- and after Pharsalus he
remained in exile, more dangerous to Caesar than the other
Pompeians for that pamphlet's sake. He was now in Sicily,
waiting on Cicero's mediation. The letter is the substantive
piece of the small Caecina cluster (\textit{Fam.}~6.5--8), longer
and politically more candid than its companions, and the closest
register-cousin in this group to \textit{Fam.}~6.14 to Ligarius
of late November.}

\textit{The body is built as a public augury. Cicero, the
official augur, takes up Caecina's own Etruscan \textit{disciplina}
and answers it with his own: since Caecina's haruspical reading
once divined Cicero's restoration from exile correctly, Cicero
will now divine Caecina's. The augury proper occupies \S\S~7--11.
It runs along two paths, in the regular augural manner: one drawn
from Caesar's character (mild, clement, drawn to talent, swayed
by united constituencies --- and Etruria will move united for
Caecina), the other from the nature of the times (Caesar has
restored Cassius, Brutus, Sulpicius, and even Marcellus with whom
he had been angriest; the logic of the present settlement will
not permit men of clean record to remain shut out while men
condemned of nameless crimes have already returned). The praise
of Caesar in \S\S~9--10 is the most candid piece of public-private
overlap on Caesar's likely course that Cicero allows himself in
the autumn of 46 --- not flattery, but the policy reading on
which the petition for Caecina rests.}

\textit{Two passages bear special weight. In \S~6 Cicero
quotes the tragic Amphiaraus on himself --- \textit{prudens et
sciens ad pestem ante oculos positam} --- to describe his own
going out to the Pompeian war: knowing the outcome, foreseeing
the ruin, and going anyway, defeated not by hope but by shame at
the prospect of failing Pompey. In \S~8 the praise of Caecina's
\textit{Querelae} as expressing Caesar's own \textit{mitis
clemensque natura} is a delicate piece of misdirection: the
pamphlet had attacked Caesar, but Cicero recasts it as a portrait
of the very clemency Caesar should now show its author. The
closing promise (\S~13) that Cicero's standing with Caesar grows
daily, and that whatever weight he has with him will be spent on
Caecina, is the substantive offer behind the augury.}
