% Ad Familiares 5.21 --- headnote
% ad-familiares-05-21, early or mid-April 46 BC, Rome

\textit{Cicero to L. Mescinius Rufus, written from Rome
early or mid-April 46 BC --- the Perseus dateline reads
\textit{Scr. Romae in. aut med. m. April. a. 708 (46)},
\textit{in.\ aut med.\ m.\ April.}\ marking it as the
opening or middle of April. Mescinius had served as Cicero's
quaestor in Cilicia in 51--50 and had attached himself,
with some friction at the time, to Cicero's circle; he
had taken the Pompeian side in the civil war and was now
living in retirement on his estates, in failing health, and
oppressed by the wrong of his political reverses. The
``African news'' in section 3 is the campaign that would
end at Thapsus on 6 April; the letter must have been written
just before, or in the first days after, that news reached
Rome, and Cicero's tone --- the war is for him a contest
between two victories from which good men have nothing to
hope --- gives no sign that the outcome was yet known.}

\textit{The letter is one of the small group from this
spring in which Cicero, under Caesar's clemency but
without political function, writes consolation to other
Pompeians and in the same breath consoles himself. The
movement of the letter is from intimacy (``one day with you
more gladly than all this present time with most of those
among whom I have to live'') to political stocktaking
(``preferred to accept peace on any tolerable terms rather
than fight against a stronger force'') to the bleak central
sentence of section 4 --- that death, which even in
happiness ought to have been despised as without sensation,
in present circumstances ought to be wished for ---
followed by a closing turn back to the practical care of
Mescinius's broken health. ``He whom you never loved (since
you loved me)'' at the head of section 2 is the standard
elliptical reference, between two Pompeians, to Caesar.}
