% Ad Familiares 13.27 --- headnote
% ad-familiares-13-27, 46 BC, Rome

\textit{Cicero to Servius Sulpicius Rufus, proconsul of
Achaia, written from Rome in 46 BC (the manuscript
dateline: \textit{Scr. Romae, ut videtur, a. 708 (46)}).
A thank-you letter and partial follow-on to Fam. 13.21:
Servius has, in the meantime, treated Hammonius and the
estate of his patron M. Aemilius Avianius generously,
Hammonius has written a profuse letter of thanks back
to Rome, and Cicero is now relaying that thanks to
Servius. The letter doubles as a renewed light
recommendation that the same beneficiaries' remaining
Achaian business be wrapped up before Servius's
proconsulship ends. The close adds a small grace note:
Cicero is enjoying the company of Servius's son
(the younger Servius Sulpicius), who is in Rome and
evidently spending time at Cicero's house.}

\textit{The opening is the most striking piece in the
letter --- a small piece of \textit{commendaticia} wit
addressed jurist-to-jurist. Servius was the most
eminent jurist of the age, and Cicero invokes the
jurists' technical practice of the
\textit{formula}\,---\,the set legal pattern under which
a case was framed --- as a metaphor for his own
recurrent letters of thanks: ``as you jurists do with
the formulae of legal pleading, I shall put it
\textit{de eadem re alio modo}'' (``on the same matter,
in another mode''). The clause is broken in the
manuscripts (Mendelssohn marks a lacuna at this point in
the Teubner text), and the precise verb completing the
metaphor is lost, but the joke is clear. Cicero is
performing exactly the variation he describes: same
business, fresh formula. The piece is the warmest of
the cluster, with the personal disclosure about
Servius's son operating as its quietly affectionate
seal.}
