% Ad Familiares 13.19 --- headnote
% ad-familiares-13-19, 46 BC, Rome

\textit{Cicero to Servius Sulpicius Rufus, proconsul of
Achaia, written from Rome around the beginning of December
46 BC (the manuscript dateline: \textit{Scr. Romae, ut
videtur, a. 708 (46)}). Part of the Servius-recommendation
sheaf (Fam.\ 13.17--28), and the longest letter of the
cluster. The beneficiary is Lyso of Patrae, Cicero's
guest-friend of long standing and (as the letter
emphasises) more than that --- a man who had spent nearly
a year living almost in Cicero's own household at Rome.
Lyso had been on the Pompeian side in the civil war and
was therefore politically exposed under Caesar's
dispensation; the letter records, in passing, that the
intercession on his behalf has succeeded and that Caesar
has sent Servius a covering letter restoring Lyso's
position. A separate sub-recommendation is interleaved
for Lyso's son, whom Cicero's own client C. Maenius
Gemellus had adopted at Patrae and whose inheritance
right Cicero asks Servius to safeguard.}

\textit{The letter sits at the top of the
\textit{commendaticia} register. Cicero's intimacy with
Lyso is the most explicit and personal of any in the
Servius sheaf --- \textit{nihil sit familiaritate nostra
coniunctius}, ``nothing is more tightly knit than the
friendship between us'' --- and the closing turn is
unusually candid about the patronage stakes: if Servius
fails to deliver, Cicero's standing with Lyso will be
diminished, not Servius's standing with Cicero. That
inversion of the usual flow of obligation, in which the
patron protects the writer's authority over his client,
is the precise calculus of a recommendation between equals.}
