% Ad Familiares 13.52 --- headnote
% ad-familiares-13-52, 46 BC, Rome

\textit{Cicero to one Rex --- the salutation in the
manuscripts is the bare \textit{Cicero Regi s.}, with no
office or further title --- written from Rome about the
middle of September 46 BC (the manuscript dateline:
\textit{Scr. Romae circ. med. m. Sept. a. 708 (46)}). The
recipient's identity is uncertain: he is normally taken to
be a Marcius Rex, perhaps the propraetor Q. Marcius Rex,
governing a province in the East --- Asia or Cilicia ---
where the addressee can plausibly affect the fortunes of
a Maltese in transit; but the editors hedge, and Cicero
himself, with the casual \textit{mi Rex} at the close,
gives no formal cue. The brevity, the bare salutation,
and the easy intimacy of the close all point to an
addressee Cicero knows well enough not to need
ceremony with.}

\textit{The beneficiary is A.\ Licinius Aristoteles of
Malta, named in the opening line as Cicero's
\textit{antiquissimus hospes} --- ``my oldest guest-friend''
--- a tier-marker Cicero reserves for the most established
of his hospitium relationships. Aristoteles had been a
Pompeian and had stayed in the cause even longer than
Cicero, who got his pardon from Caesar. The recommendation
is therefore not so much an introduction as a request
that an existing favour be visibly delivered: the
addressee already knows the man's standing, and the
letter exists to make sure Aristoteles can carry it away
as a tangible benefit. The flattery in the middle
sentence --- \textit{ex multis cognosco meam
commendationem plurimum apud te valere}, ``my
recommendation carries the greatest weight with you'' ---
is the standard pivot of such letters between equals.}
