% Ad Familiares 4.4 --- headnote
% ad-familiares-04-04, late September or early October 46 BC, Rome

\textit{Cicero to Servius Sulpicius Rufus, in Achaia, written at
Rome at the very end of September or the beginning of October 46
BC --- Perseus dateline: \textit{Romae ex. m. Sept. aut in. Oct.
a. 708 (46)}. This is the second of Cicero's surviving letters
to Sulpicius from the autumn of 46 (after 4.3 in August) and
together with 4.7 it sits at the centre of the Marcellus
business. M. Marcellus, Sulpicius's consular colleague of 51 BC,
had been in self-imposed exile at Mytilene since Pharsalus; in a
Senate session of September, on a motion that began with L.
Piso's reference to Marcellus and culminated in C. Marcellus
prostrating himself at Caesar's feet, Caesar unexpectedly
restored him. Cicero's response, delivered when his turn came to
speak, was the speech we now read as \textit{Pro Marcello} ---
his first public oratory after a near two-year silence.}

\textit{The letter opens with a piece of literary banter ---
Sulpicius has excused his repetitive letters as ``poverty of
style,'' \textit{paupertate orationis}, and joked that Cicero
has ``riches of style''; Cicero will not let either compliment
stand, and the exchange runs into two Greek tags playing on
\textit{eir\=oneia}. The political body of the letter is the
account of Marcellus's restoration: ``that day seemed to me so
beautiful that I felt I was seeing some image, as it were, of a
state coming back to life'' --- \textit{speciem aliquam viderer
videre quasi reviviscentis rei publicae} --- the famous line
that records the moment, and the moment that gave Cicero the
warrant to break his silence. The closing sections balance two
ironies. Cicero has parted with his ``honorable leisure'' and
fears he may now be drawn into speaking on other matters too;
and his advice to Sulpicius, who is asking whether to stay in
Achaia or come home, is that he should stay --- not because Rome
is intolerable, but because at Rome only one's own family can
give pleasure, and of the public business one would rather hear
than see.}
