% Ad Familiares 6.14 --- headnote
% ad-familiares-06-14, late November 46 BC, Rome

\textit{Cicero to Q. Ligarius, written at Rome on the fifth day
before the kalends of the first intercalary month of 46 BC ---
Perseus: \textit{Romae a.~d.~v K. intercal.~priores a.~708 (46)}
--- one of the special intercalary months Caesar inserted that
year before the Julian calendar reform took effect. Ligarius, a
former legate in Africa, had stayed with the Pompeian army at
Thapsus, and a private prosecutor, Q. Tubero, had brought against
him before Caesar a charge that would amount to treason. The
trial was approaching, and Cicero was preparing the defence ---
the speech that survives as the \textit{Pro Ligario}. This is the
letter he sent in the meantime to Ligarius himself.}

\textit{The piece is brief, and it is among the gravest of the
letters of 46. Cicero begins by confessing the temperament he
always confesses --- a man more afraid of bad outcomes than
hopeful of good ones --- and then sets that temperament against
what he himself has seen and heard. Going to Caesar at dawn, with
the brothers and kinsmen of Ligarius prostrate at the dictator's
feet, he read in Caesar's expression and bearing more than in his
words that the matter was settled. The closing tricolon ---
\textit{si turbidissima sapienter ferebas, tranquilliora laete
feras} --- is the heart of the letter: bear the calmer waters
gladly, having borne the wildest waters wisely. The promise to
keep entreating not only Caesar but all Caesar's friends is what
Cicero would in fact do on the day of the trial, with the speech
that turned the case.}
