% Ad Familiares 9.17 --- headnote
% ad-familiares-09-17, after the Ides of August 46 BC, Rome

\textit{Cicero to L. Papirius Paetus, written at Rome after the Ides
of August (13 August) 46 BC --- Perseus: \textit{Romae post Id.~Sext.~a.~708 (46)}. Paetus has written to Cicero with a worry that
Caesar's land settlements for his veterans may sweep up his own
holdings (or those of friends and townships near him); the surveyors
have indeed come close, into the Veientine and Capenate fields just
north of Tusculum. \textbf{Metadata note: the \texttt{meta/works.yaml}
entry gives day-precision \texttt{-0046-08-13}, the Ides of Sextilis
itself; the Perseus dateline says \textit{post Id.~Sext.}, so the
true date is between 14 and 31 August. The entry should be revised to
year-or-month precision when the metadata is consolidated.}}

\textit{Three short sections, drier than the great gastronomic
letters of mid-46 but in the same Paetus voice. The opening jab is
perfect: are you not an absurd fellow, asking me what is to become of
the townships and the land, when you have had Balbus staying with
you --- the very man from whom I learn everything I know, and whom you
could have got it out of, sober or drunk? Cicero then turns serious:
he himself does not ask these things, because for nearly four years
now they have been living \textit{de lucro}, on bonus time, if it can
be called bonus time to outlive the commonwealth; and because he
already knows the answer anyway --- what will happen is what those
who have the strength want, and the strength is in arms. The
surveyors at the Veientine and Capenate land have come close, but
\textit{nihil tamen timeo}: he is enjoying things while he can,
prepared, as \textit{vir fortis idemque philosophus} --- brave man
and philosopher together --- to love the man through whose kindness
he is still alive. The closing flourish is the great Stoic-tinged
triad: \textit{optare optima, cogitare difficillima, ferre quaecumque
erunt} --- hope for the best, brace for the worst, bear what comes;
even Caesar, he adds, does not know what is to be, since
\textit{nos enim illi servimus, ipse temporibus} --- we are slaves to
him, he to circumstances.}
