% Ad Familiares 9.18 --- headnote
% ad-familiares-09-18, around 20 July 46 BC, Tusculanum

\textit{Cicero to L. Papirius Paetus, written at the Tusculan
villa around 20 July 46 BC --- Perseus: \textit{in Tusculano
circ.~xiii K.~Sext.~a.~708 (46)}, that is, the thirteenth day
before the Kalends of Sextilis, counted inclusively. With the
courts shut down under Caesar's dictatorship and his
\textit{forense regnum} --- his kingdom of the Forum --- gone,
Cicero has begun keeping a kind of declamation school for the
new ruling generation: Hirtius, Pansa, Dolabella and others come
to his villa to declaim with him in Greek and Latin, and he
eats with them in return. The conceit of the letter is the
parallel with the tyrant Dionysius of Syracuse, who, when
driven into exile, is said to have opened a school at Corinth ---
a stock anecdote for fallen greatness reduced to schoolmastering.}

\textit{Four short sections, all of them in Paetus mode. The
opening is the Dionysius simile and a list of consolations for
his new life: he fortifies himself against the times, his health
returns with the exercise, and his oratorical \textit{facultas},
such as it ever was, comes back too. The set-piece is the death
catalogue --- \textit{in lectulo, fateor, sed non accidit; in
acie non fui; ceteri quidem, Pompeius, Lentulus tuus, Scipio,
Afranius foede perierunt. at Cato praeclare} --- four
Republicans dead foully, one gloriously, and the option of
joining them ``when we wish'' kept dryly in reserve. Then the
gastronomic boast in chiastic balance: \textit{pluris iam pavones
confeci quam tu pullos columbinos. tu istic te Hateriano iure
delectas, ego me hic Hirtiano} --- more peacocks downed than you
have wood-pigeon squabs, your Haterian gravy against my
Hirtian. The Greek tag \textit{prolegomenas} (rhetorical
preludes) is the school joke, capped by the proverbial
\textit{sus Minervam} --- a pig teaching Minerva. The close
turns to Paetus's finances: if you cannot sell off your
\textit{aestimationes} (Caesar's debt-relief settlement
valuations) or fill your pot of denarii, come back to Rome on
the mule you have left after eating the gelding, and take the
junior master's bench beside mine --- with a cushion thrown in.}
