% Ad Familiares 13.79 --- headnote
% ad-familiares-13-79, 46 BC, Rome

\textit{Cicero to Decimus Allienus, proconsul-designate (or
incoming proconsul) of Sicily, written from Rome around
the beginning of December 46 BC (the manuscript dateline:
\textit{Scr. Romae in. a. 708 (46)}, which Perseus prints
as ``the beginning of 708'' but which on the substantive
reasoning belongs to the threshold between the outgoing
and incoming governors). Allienus succeeded Manius
Acilius Glabrio in Sicily at the close of 46, and this
letter belongs to the same handoff moment as Fam.\ 13.78:
the recommendation patterns that Cicero had been using
with Acilius across the cluster of Fam.\ 13.30--39 are now
being redirected to his successor. The opening salutation
\textit{D. Allieno procos.} reflects that turn, and the
two surviving Allienus letters (Fam.\ 13.78--79) read
naturally as the start of a new cycle on Cicero's part
for the same province.}

\textit{The beneficiaries are the two sons of C.\ Avianius
Flaccus, one of Cicero's older intimates and a
beneficiary of Allienus's earlier good will. The elder
son, also C.\ Avianius, is in Sicily already; the
younger, Marcus, is at Rome with Cicero. The letter is
unusually short and concentrated: a single paragraph of
testimonial, request, and double affirmation of the
request --- \textit{te rogo} \dots \textit{te vehementer
etiam atque etiam rogo} --- a closing reduplication that
in this cluster marks the warmer end of the genre. The
old debt Allienus owed Flaccus, registered in the second
sentence, supplies the ground for the recommendation: the
sons are inheriting the favour, and Cicero is the
mediating voice that turns the inheritance into a
present obligation.}
