% Ad Familiares 6.10 --- headnote
% ad-familiares-06-10, mid-August 46 BC, Rome

\textit{Cicero to Trebianus, written at Rome about August 46 BC
(works.yaml \texttt{-0046-08-15} at month-precision; the Perseus
dateline reads \textit{Romae circ.~m.~Sext.~a.~708 (46)} ---
``around the month of Sextilis''). Trebianus, a Roman of some
standing whose name surfaces only in this short cluster of letters
(\textit{Fam.}~6.10--11), had stayed in arms with the Pompeians
longer than most --- the line ``your decision, or rather the chance
that kept you in arms in the civil war so long'' implies he had not
laid them down at any of the conventional moments --- and was now
in exile, with his fortune and standing under cloud. The case is
not desperate; Caesar has not refused him, and intermediaries are
in play. This is the first of the two surviving letters, the
second (\textit{Fam.}~6.11) celebrating his recall a little less
than a year later.}

\textit{The register is the Trebianus-counterpart of the Caecina
note of December (\textit{Fam.}~6.5): a short, working letter that
catalogues the legwork Cicero is doing on the case (Postumulenus,
Sestius, Atticus, the freedman Theudas) and lifts Trebianus's
spirits with a Stoic-consolatio note that the philosophical
resources of his studies are now repaying him. No Greek. The
doctrinal pivot is the standard one of the cluster --- the
\textit{conscientia factorum et consiliorum} (the consciousness of
having acted and counselled rightly) carries the philosophical
freight, with the assurance that ``the nature of things'' and ``the
tilt of the times'' will not let so bitter an injustice cling to
so good a cause for long. The closing tricolon ---
great-spirited, fortune-governed-by-the-times, our-counsels-will-
provide --- is the standing reassurance of Book~6.}
