% Ad Familiares 13.68 --- headnote
% ad-familiares-13-68, mid-early September 46 BC, Rome

\textit{Cicero at Rome to P. Servilius Isauricus, proconsul of
Asia, written about the middle of early September 46 BC (the
manuscript dateline: \textit{Scr. Romae circ. med. in. Sept. a.
708}). Servilius had been consul with Caesar in 48 BC --- the
``colleague'' (\textit{conlega}) of the salutation --- and had now
gone out to govern the senatorial province of Asia for the
Caesarian regime. The first paragraph thanks him for a letter
reporting the legs of his voyage out, and asks for fuller news of
provincial conditions and policy; the second touches, with a
careful glance at the danger of frank letters being intercepted,
on Cicero's faint hope that ``our colleague'' Caesar will leave
some sort of commonwealth at all, and accepts that Servilius's
post is the right one for him; the third closes with a promise
to attend to Servilius's interests at Rome, and in particular to
his father --- the elder Servilius, princeps senatus and one of
the most senior Optimate survivors.}

\textit{This is the opening letter in the small Servilius cluster
preserved at the end of \textit{Ad Familiares} 13: a personal
exchange between two ex-consuls (Cicero himself had been consul
in 63, Servilius in 48, and the formula \textit{conlegae} hails
that shared dignity), followed by a sequence of recommendations
(13.69--72) sent to Servilius in his Asian capacity. The piece is
plain and short, but the brittleness behind it is audible: ``what
I think on the largest questions of the commonwealth, on account
of the danger of letters of that sort, I shall not often write,''
and the half-hope --- ``Caesar, our colleague, will see to it that
we have some sort of commonwealth at all'' --- is the most one
could safely write down in the autumn of 46. The Perseus dateline
sets the letter to mid-early September 46 BC; \texttt{meta/works.yaml}
carries a year-precision placeholder of \texttt{-0046-01-25} that
should be corrected to \texttt{-0046-09-05} (or similar) and the
file prefix kept at \texttt{046bc-}.}
