% Ad Familiares 13.24 --- headnote
% ad-familiares-13-24, 46 BC, Rome

\textit{Cicero to Servius Sulpicius Rufus, proconsul of
Achaia, written from Rome in 46 BC (the manuscript
dateline: \textit{Scr. Romae, ut videtur, a. 708 (46)}).
The fifth letter in the Servius-recommendation cluster
of Book 13, and the first of a second sub-sequence
(Fam. 13.24--28) which continues the same patronage
business begun in 13.20--23. The beneficiary here is
Lyso of Patrae, a Greek \textit{hospes} (formal guest-
friend) of Cicero, already commended to Servius in an
earlier letter no longer extant. Someone at Achaia had
evidently carried a story to the governor that Lyso
spoke disparagingly of him at Rome; Lyso had written
back to Cicero to report that the first commendatory
letter had been enough to dislodge the suspicion.}

\textit{The letter is essentially a thank-you note
folded inside a renewed recommendation. The structure
is careful: a movement of gratitude in section 1 to
section 2, a categorical denial of the slander on Lyso's
behalf and on everyone's behalf ("there is no one alive
who has ever made mention of you except with the highest
praise" --- a calibrated hyperbole that nevertheless
performs the loyalty-formula correctly), and then a
return to the original commendation, asking that Servius
extend his good offices further. The piece is a useful
specimen of how a Roman of standing managed a small
provincial crisis between absent friends: not by
demanding anything, but by repairing the social fabric
in the standard idiom of the \textit{commendaticia} and
trusting his correspondent to read the gesture
correctly.}
