% Ad Atticum 11.9 --- headnote
% ad-atticum-11-09, 3 January 47 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium on
the third day before the Nones of January 707 AUC ---
3 January 47 BC (the manuscript dateline: \textit{Scr.\
Brundisi iii Non.\ Ian.\ a.\ 707 (47)}). The letter is
written on Cicero's birthday, as he tells us in the
last section, and the day weighs on every line of it.
Two weeks have passed since 11.7--11.8; the new
tribunate has now entered office, and Cicero realises
he is held in Italy by the exceptions of Caesar's
edicts and by some clause of the recent legislation
(``cum iam lege etiam sim confectus et oppressus'').
Balbus's letters, his lifeline to headquarters, are
``daily more languid''; many other men's letters are
reaching Caesar against him. The accusation against
himself he answers in the most absolute terms he ever
puts to Atticus: ``I am being ruined by my own
fault; chance has brought me no evil, all has been
incurred by my own doing.'' He had read the war
correctly --- weak forces against a most prepared
enemy --- and had taken the course that ought
``granted to others, also to be granted'' to him; but
he gave way to his own people, ``or rather obeyed
them,'' and now sees what that obedience has cost.}

\textit{The middle section narrates a small,
devastating discovery. A bundle of letters was
delivered to Cicero; he opened it on the off-chance
some piece was for him. There was none --- only one
letter to Vatinius, another to Ligurius, both written
by his brother Quintus. He had them carried on; the
two addressees came back to him at once, ``burning
with grief,'' read him the letters, full of reproaches
against him. Ligurius flew into a rage: he knew that
Caesar had once hated Quintus, and had even so given
him a great sum of money for Cicero's sake. Cicero,
struck, asked to see what Quintus had written to
others; he found more of the same. He encloses
copies for Atticus, leaves it to him whether to
return them to Quintus, and notes drily that the
seals were broken --- Pomponia, Quintus's wife and
Atticus's sister, has his seal-ring. The letter
closes with Tullia. Among all his miseries the one
that stands in for all is that he is leaving her
``stripped of her patrimony, of all fortune''; her
mother Terentia, he now perceives, has prepared for
her the same fate that is prepared for him. He begs
Atticus to come, or failing that to hold Tullia
commended to him and to soften Quintus, her uncle,
toward her. ``I have written this to you on my
birthday. Would that I had never been taken up at
birth, or that nothing further had been born to the
same mother since! Tears prevent me from writing
more.'' The faint daggered crux in \S1 (``benivolentie
va\dag'') is preserved at the obelus; the sense
followed is ``goodwill,'' the most natural reading of
the corruption.}
