% Ad Atticum 11.21 --- headnote
% ad-atticum-11-21, 27 or 28 August 47 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium on
the sixth or fifth day before the Kalends of September
47 BC --- 27 or 28 August (the manuscript dateline:
\textit{Scr.\ Brundisii vi aut v K.\ Sept.\ a.\ 707
(47)}). Atticus's letter of the twelfth day before the
Kalends has arrived, and with it --- forwarded ---
a letter from Cicero's brother Quintus that has
returned the grief Cicero had finally laid down. He
will not blame Atticus for sending it, but says he
wishes it had not been sent. Quintus's letter, which
the next book of the correspondence will keep
revisiting, evidently included an attack on Cicero
delivered into Caesar's camp through young Quintus.
The remaining business of \S1 is brisk: Atticus is to
see to the will, on terms to be settled; Terentia has
written about money in the same sense Cicero had
already reported, and Cicero will draw on the source
Atticus has named if need arises.}

\textit{Section 2 reports the latest news of Caesar.
He will not be at Athens by the Kalends of September;
many things hold him in Asia, especially Pharnaces;
the twelfth legion, the first that Sulla approached,
is said to have driven him off with stones, and the
view in camp is that not a legion will be moved.
Previously he had been thought to be heading from
Patrae straight to Sicily; if so, he must touch at
Brundisium, and Cicero would have preferred the
\dag other course\dag (here a small crux), since some
opportunity of escape would have come with it. As it
stands, he must wait, and the foulness of the place
must be borne on top of the rest. Section 3 takes up
Atticus's tactical urging --- accommodate what you do
to the moment --- and refuses it on principle: amid
his own great faults and the great wrongs done him by
his own household, there is nothing he can either do
or pretend to do worthy of himself. Atticus's
parallel with Sulla's day Cicero rejects: those
proscriptions were of their kind exceedingly
distinguished, only somewhat less tempered in
moderation --- a sentence that has the dryness of the
late letters, and that the apparatus has long puzzled
over. The closing lines are characteristic: write
often; everyone else is silent; Quintus's son got
everything from Caesar on the spot, and Cicero was
not so much as mentioned.}
