% Ad Familiares 15.15 --- headnote
% ad-familiares-15-15, mid-August 47 BC, Brundisium

\textit{Cicero to C. Cassius Longinus, written from Brundisium
around the middle of Sextilis --- Perseus dateline
\textit{Scr. Brundisii circ. med. m. Sext. a. 707 (47)}, that
is, mid-August 47 BC. Cicero has been stuck at Brundisium for
the better part of a year, ever since he came back from
Pompey's camp after Pharsalus (August 48). Caesar, after
Pharsalus, has spent the intervening months tangled in the
Alexandrian war, then turning to deal with Pharnaces of
Pontus at Zela --- whence the famous \textit{veni, vidi, vici}
in early Sextilis --- and is at last expected back in Italy.
The African war, with Cato, Scipio Nasica, and Labienus
holding out at Utica, is openly impending. Cassius, future
tyrannicide and at this date already detached from the lost
Pompeian cause, has been writing to Cicero from Italy and ---
as a former legate of Pompey who had reconciled with Caesar
at the Hellespont --- speaks with some authority as a fellow
survivor.}

\textit{The letter is the political confession of a man who
has staked everything on one battle and lost. The first two
sections defend the original logic of going over to Pompey
(``a single battle, if not deciding the whole cause, then at
any rate settling our personal judgement'') and trace what
went wrong --- not the engagement itself, but Caesar's
inability to follow up Pharsalus quickly, the lost year of
Alexandria and Pontus, and the fact that some on the losing
side now ``scorn the very fact of being beaten'' and are
regrouping in Africa. The third section is a piece of bitter
self-mockery: Cicero rushed back to Italy to ``urge Caesar on
toward peace while he was already running,'' a proverbial
phrase --- and then Caesar was nowhere to be seen for a year.
The closing line --- ``would that I had obeyed those earliest
letters you sent from Luceria!'' --- refers to Cassius's
advice, in the days just after Caesar crossed the Rubicon, not
to take the Pompeian side at all; the dignity that has been
worn out in the long wait at Brundisium would, on that
counsel, have been preserved without any of this misery.}
