% Ad Atticum 11.10 --- headnote
% ad-atticum-11-10, 21 January 47 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium on the
twelfth day before the Kalends of February 47 BC --- 21
January (the manuscript dateline: \textit{Scr.\ Brundisi
xii K.\ Febr.\ a.\ 707 (47)}). Cicero has now been
marooned at Brundisium for the better part of a year,
having recrossed from the Pompeian camp after Pharsalus
and lacking either Caesar's clear pardon or the nerve to
press for it. The new wound this letter bears is family:
Publius Terentius, an associate who has been deputy
collector of the Asian port-duties and pasture-tax, saw
young Quintus at Ephesus on 8 December and reports that
the boy spoke of himself as Cicero's bitterest enemy and
showed Terentius a written speech he meant to deliver
against his uncle before Caesar. The father, Cicero's
brother Quintus, has spoken in the same vein at Patrae;
Cicero has already forwarded Atticus the letters in
which Quintus's \textit{furor} is on display. He knows
this grieves Atticus too. The thought that he will not
even have room to lodge a complaint with the two of them
is what twists the knife.}

\textit{The second paragraph turns to the political
weather, and it is uniformly bad. Reports from Africa
contradict the more hopeful note Atticus had sent: the
Pompeian remnant there is said to be in formidable
order. Add Spain, an alienated Italy, legions
whose strength and loyalty are no longer what they
were, and the affairs of the city in ruin: Cicero finds
his only rest in Atticus's letters and begs him not to
let them lapse, and to denounce his enemies even if he
cannot bring himself to hate them --- not to accomplish
anything, but so that they feel that Cicero is dear to
him. The letter ends on the bare date with no other
business; this is the tone of all of Book 11.}
