% Ad Atticum 11.22 --- headnote
% ad-atticum-11-22, c. 1 September 47 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium around
the Kalends of September 47 BC (the manuscript dateline:
\textit{Scr.\ Brundisi circ.\ K.\ Sept., ut videtur, a.\
707 (47)}). A short, querulous note in two sections.
A packet has come through from Balbus's courier, bringing
letters from Quintus to Caesar attacking Cicero ---
``that whole style of letter,'' as Cicero calls it,
which by now is so notorious that it would have told
no one anything new. Atticus has been worried that
Quintus's offensive may rebound on Quintus himself,
and has asked Cicero to intercede on his brother's
behalf; Cicero replies that Caesar would not even
\textit{allow} a petition about the man, and the
brutal addendum is that he himself counts for so little
that his interventions, when granted at all, ``carry
no weight.''}

\textit{The second paragraph is Brundisium logistics
at the end of a hot summer. Sulla and Messalla, the
two senatorial intermediaries on whom the camp's pay
strike depends, are expected tomorrow on their way
to Caesar --- the troops have driven them out and
will not move without their stipendium. Caesar
therefore will come this way after all, though slowly,
``spending many days in one \dag town\dag,'' and
Pharnaces' war in Pontus will impose further delay.
Cicero's body is failing under the climate; he asks
Atticus, for once, to break a long silence and actually
\textit{advise} him --- should he send his excuses
by Sulla's party and move closer himself? The plea
``to see you'' closes the letter, with a brief
instruction about the will. Two short cruxes
(\dag oppidum\dag, and the Greek-like Pharnaces clause)
are preserved as \dag\ markers; the sense given is
the most natural reading.}
