% Ad Atticum 11.20 --- headnote
% ad-atticum-11-20, 17 August 47 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium on
the sixteenth day before the Kalends of September 47 BC
--- 17 August (the manuscript dateline: \textit{Scr.\
Brundisi xvi K.\ Sept.\ a.\ 707 (47)}). The letter
opens on news that has finally crossed the Adriatic.
On the day before this one, the seventeenth day before
the Kalends, a freedman of C.\ Trebonius arrived,
twenty-eight days out of Seleucia in Pieria, having
seen Cicero's nephew Quintus with Hirtius in Caesar's
company at Antioch: the two have obtained for Quintus
the elder everything they wanted, and got it without
difficulty.}

\textit{The relief is cut short the moment Cicero
takes it in. The pardon would do more to ground a
considered hope if there were not other Quintuses to
fear --- and if what this one Caesar grants did not
remain in his power, like a master's, to take back.
Even Sallustius has been pardoned. Caesar is said to
refuse no one absolutely, which is itself suspect:
recognition of him (the formal notitio, the audience
in which the new dispensation is acknowledged and the
suppliant's standing fixed) is put off. M.\ Gallius
has restored the slaves to Sallustius; he is in
Brundisium to ferry legions across to Sicily. Caesar
himself will go straight on from Sicily to Patrae ---
if so, Cicero will shift somewhere closer, which is
what he had wanted all along. He is waiting urgently
for Atticus's reply to his last request for advice.
The signed dateline is preserved.}
