% Ad Familiares 9.1 --- headnote
% ad-familiares-09-01, late 47 or early 46 BC, Rome

\textit{Cicero to M. Terentius Varro, written at Rome late in 47
or at the start of 46 BC (Perseus: \textit{Romae ex.~a.~707 (47)
aut in.~a.~708 (46)}). This is the opening letter of the great
Varro sequence of \textit{Ad Familiares} book 9 --- four or five
letters in which Cicero, lately back in Italy after the disaster
at Pharsalus and the awkward year of waiting at Brundisium, is
beginning to feel his way back into a civic and literary life
under Caesar's sole power. The two men are at the same point:
both senior consulars and former Pompeians, both now pardoned
and at Rome, both deciding what use to make of an unwanted
leisure. Varro is the older, the polymath, the man whose
learning Cicero has all his life treated with a half-rivalrous
deference; the tone of these letters keeps just that note.}

\textit{The pivot of the letter is one of Cicero's most candid
sentences about the years of the civil war: ``since coming back
to the city I have made my peace with my old friends --- that
is, with our books.'' The conceit is allegorical and unforced.
He had not stopped reading because he was angry with the
philosophers; he had stopped reading because he was ashamed to
face them. He had ignored their precepts, joined turbulent
affairs and untrustworthy allies, and now comes back, contrite,
and is forgiven. Varro, who had stayed with the books all along,
is acknowledged as the wiser man --- a graceful piece of staged
deference that also frames the whole project of the literary
years to come. The close is practical: Tusculum, Cumae, or even
Rome, anywhere they can be together.}
