% Ad Atticum 11.25 --- headnote
% ad-atticum-11-25, 5 July 47 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium on
the third day before the Nones of Quintilis 47 BC ---
5 July (the manuscript dateline: \textit{Scr.\ Brundisi
iii Non.\ Quint.\ a.\ 707 (47)}). A short, very dark
letter in three sections that opens with Atticus
having finally confessed, at length, that there is no
plan by which he can help. Cicero accepts the verdict
without protest. The bleakness of \S1 is the bleakest
in all of Book 11: nothing has been brought on him by
chance (which, he says, would be bearable); everything
has been done to himself, by errors and miseries of
mind and body which he wishes those nearest him had
cared to remedy. He asks Atticus to stop offering
counsel or consolation, and merely to keep writing
whatever comes to mind --- ``as long as there is
someone to whom to send it; and that will not be
long.''}

\textit{\S2 is a single Brundisium-room sentence: a
rumour, started by a letter of Servius Sulpicius and
confirmed by every subsequent messenger, that Caesar
has left Alexandria. ``Whether it is true or false,
since it makes no difference to me, I do not know
which I should prefer.'' \S3 is the longest and the
worst, returning to Tullia and the divorce. Cicero
is wrecked by the girl's situation; at the second
installment of the dowry, he says, they were blind.
He begs Atticus to gather what can be saved from
silver and furniture into safekeeping --- ``for now
the end seems to be on us, and there will be no
terms of peace, and what we have will perish even
without an enemy'' --- and to find a moment to speak
to Terentia about it. The letter closes with the
bare admission that he cannot write everything.
Two short cruxes in \S3 are preserved as \dag\
markers; the sense given is the most natural reading.}
