% Ad Familiares 16.18 --- headnote
% ad-familiares-16-18, c. May 47 BC, Rome

\textit{Cicero to his freedman Tiro, written from Rome. The
Perseus dateline is the very wide bracket \textit{Scr. Romae
inter med. m. Oct. a. 707 (47) et Id. Mai. a. 710 (44)} ---
somewhere between mid-October 47 and mid-May 44. The standing
editorial guess, followed in \textit{meta/works.yaml} and
adopted here, lodges the letter in the spring of 47, while
Tiro is at the Tusculan villa convalescing and Cicero is in
the city. The opening sentence is a fragment of a private
joke about the form of the address line --- whether the
salutation should read ``Tullius to his Tiro'' with or without
``his own'' (\textit{suo}) tacked on --- and the rest is the
domestic chatter of a man who knows his correspondent's body
and his correspondent's reading habits equally well.}

\textit{Almost every line carries a Greek word, since Greek
is the technical language of medicine and Tiro's prescription
is being checked off in it: a sweating-cure
\textit{[Greek: diaphoresin]} that has already done some good,
and then the regimen to be kept up --- digestion
\textit{[Greek: pepsin]}, the loosening-rub
\textit{[Greek: akopian]}, a measured walk
\textit{[Greek: peripaton symmetron]}, a rubdown
\textit{[Greek: tripsin]}, and an easy bowel
\textit{[Greek: eulysian koilias]}. The middle business about
the gardener Parhedrus, Helico's old lease, and the
Aqua Crabra (a small stream that supplied the Tusculan
estates) is estate management talk passed between owner and
freedman as between equals. The closing tease ---
``no little books with you? or are you composing something
Sophoclean?'' --- is Cicero teasing Tiro about his own
literary ambitions, of which Cicero in fact thoroughly
approved.}
