% Ad Atticum 11.11 --- headnote
% ad-atticum-11-11, 8 March 47 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium. The
manuscript dateline as transmitted is corrupt
(\textit{Scr.\ Brundisi vir i Id.\ Mart.\ a.\ 707 (47)}),
but the postscript at the end of \S2 gives the date
plainly: \textit{viii Idus Mart.}, the eighth day before
the Ides of March --- 8 March 47 BC. This is the same
day as 11.12, and indeed the two letters were probably
dispatched together (cf.\ 11.12.1, where Cicero notes
that he had sent off the morning's letters before
Cephalio brought Atticus's of the previous evening).
The works.yaml entry carries a day-precision date of
15 March, evidently a guess at the corrupt
\textit{vir i Id.\ Mart.}; this should be corrected to
8 March in the consolidation pass.}

\textit{The letter is brief and almost without business.
Cicero is so worn down by his ``greatest of pains'' that
he can no longer easily compose answers, and is no
longer even waiting for Atticus's letters (though they
always bring something he wants). The single piece of
operational content concerns money: thirty thousand
sesterces received from Gnaeus Sallustius are to be paid
out to Publius Sallustius, and Atticus is to see, with
Terentia, that there is some further sum left for
Cicero's own use; he has not dared draw on anyone at
Brundisium until he knows funds are in hand at Rome.
The closing sentences turn to brother Quintus
(``the other man, in Achaea''), who continues to run
Cicero down behind his back: Atticus's letters of
remonstrance, evidently, did no good. Cicero's verdict
on himself --- ``the grief is the heavier in proportion
as the fault is the greater'' --- is the Book 11
keynote.}
