% Ad Familiares 8.17 --- headnote
% ad-familiares-08-17, late January or February 48 BC, written on the Caesarian side

\textit{M. Caelius Rufus to Cicero, the last letter of the
correspondence to survive (Perseus: \textit{Scr. Romae vel
ex. m. Ian. vel m. Febr. a. 706 (48)}; the Perseus dateline
records the Roman tradition, but the contents make plain
that Caelius is writing not from the city but from the
Caesarian camp, probably soon after the close of the Ilerda
campaign and the surrender of Pompey's Spanish armies). The
civil war that the previous letters in book 8 had watched
gathering has now broken open. Cicero, after months of
agonised hesitation, finally crossed to Pompey in Greece in
the summer of 49; Caelius, the protégé of his early career
and his political ally through the long Cilician
correspondence, took the other road and went to Caesar.}

\textit{The note is brittle, accusatory, and unmistakably
Caelian: a wishing-aloud rather than a considered argument,
all the rapid-fire flexes of the gossip-letters now turned
to recrimination. He half-regrets having been at Formiae,
not Spain, when Cicero set out for Pompey --- as if
distance alone would have spared him the choice he made;
he blames Curio, whose friendship pulled him by degrees
into ``this ruined business'' (\textit{hanc perditam causam}
--- the very word Cicero used in private of the Caesarian
side); he charges Cicero himself with having neglected the
duty of a friend on that night ride to Ariminum, when
Caelius came to him on his way to Caesar's camp and Cicero
was full of peace-errands and grand citizenly poses, but
gave the younger man no counsel that would have held him
back. The second paragraph turns to bitter analysis: he
does not lack faith in the Caesarian cause, he says, but
he loathes the men around it; the towns are Pompeian
through and through except for a handful of money-lenders;
he has been working to make the people Caesar's, and means
to make Caesar's side win in spite of itself. The piece is
heavily corrupted in the manuscripts, and three short
phrases are bracketed [corrupt] in the translation; the
sense, even so, is the unmistakable sound of a young
politician who has chosen wrong, knows it, and is reckoning
publicly with the friend whose advice he failed to ask.
Within months Caelius would quarrel with Caesar over his
debt-relief politics, raise a revolt with Milo in Italy,
and die there.}
