% Ad Atticum 11.8 --- headnote
% ad-atticum-11-08, 19 December 48 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium. The
Perseus dateline reads \textit{Scr.\ Brundisi x iiii K.\
Ian.\ a.\ 706 (48)} (a.d.\ xiv K.\ Ian.\ = 19 December
48 BC), which matches the body's farewell ``xiii Kal.\
Ian.'' --- 20 December --- to within a day; the
manuscript subscription on the heading thus dates the
letter to the same brief Brundisian window as 11.7.
(The entry in \texttt{meta/works.yaml} currently carries
the placeholder date \texttt{-0048-01-01}, which is
plainly wrong and should be corrected to
\texttt{-0048-12-19}, in line with the dateline.)
The letter is short, two sections, and operational.
Lepta and Trebatius, both with the writer at
Brundisium, will tell Atticus the rest in person;
what Cicero himself writes is a renewed appeal that
Atticus push Balbus and Oppius to keep writing to
Caesar on his behalf, against the attacks --- present
and epistolary --- now being mounted at headquarters.}

\textit{The second section names them. Fufius Calenus
(Quintus Fufius Calenus, the consul-designate Caesarian)
is in Caesar's circle and is Cicero's bitterest enemy
there. Worse, Cicero's own brother Quintus has sent
his son not merely to plead his own case before Caesar
but to accuse Cicero, and Quintus himself, wherever
he is, will not let off heaping abuse: people who heard
him speak ``certain monstrous things'' at Sicyon in
front of many witnesses have carried the report back
to Brundisium. Cicero rehearses the bitterness only
briefly, knowing that rehearsal deepens both his own
distress and Atticus's, and returns to the practical
request: Balbus should be moved to send someone with
this specific matter in mind, and Atticus should
arrange letters in Cicero's name to whoever he judges
useful. The letter closes with the farewell ``xiii
Kal.\ Ian.,'' the thirteenth day before the Kalends
of January.}
