% Ad Atticum 11.7 --- headnote
% ad-atticum-11-07, 19 December 48 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium on the
fourteenth day before the Kalends of January 706 AUC ---
19 December 48 BC (the manuscript dateline: \textit{Scr.\
Brundisi xiv Kal.\ Ian.\ a.\ 706 (48)}). Pharsalus is
five months past. Pompey has been murdered in Egypt;
Cato and the surviving optimates have regrouped in
Africa; Caesar is still in Alexandria. Cicero, having
declined to follow the war into Africa, has crossed
back to Italy on Dolabella's advice and is held at
Brundisium under what amounts to internal exile, his
lictors still in attendance --- an ambiguous honour
he half wants to keep and half wants gone. The letter
is long, eight sections, and ranges through the whole
of his post-Pharsalus predicament: the question of
whether he was entitled to retain his lictors (Sestius's
precedent cuts both ways); the bare fact, just notified
him by Antonius, that Caesar has ordered Cato and L.
Metellus barred from Italy and all returning Pompeians
held off pending personal review, with Cicero and
Laelius named by name as exceptions in Antonius's edict
--- a singling-out he would have preferred to do without.
Then the longer self-justification of \S3: he is being
reproached for not having sailed to Africa with the
republican remnant; his answer is that the commonwealth
ought not to be defended by the barbarian auxiliaries
of a treacherous nation (Juba's Numidians) against a
veteran Caesarian army. He sees the trap clearly: if
the Africans win, he is finished; if they lose, their
wound is at least the honourable one. ``Haec me
excruciant.''}

\textit{The second half turns to Atticus's role as
intermediary. Sulpicius's quietism, though less
glorious than Cato's stand, is free of both danger
and distress; the Achaian remnant is better off than
Cicero only because there are many of them in one
place and a return to Italy means a return home.
Atticus is to keep mitigating, keep winning approval
where he can, and above all keep Balbus and Oppius
writing to Caesar on Cicero's behalf, against the
unnamed enemies (his brother Quintus, his nephew, and
others) who are reporting to Caesar that Cicero
repents his course or disapproves of what is being
done. Section 6 breaks the letter open: Atticus has
written that Tullia is demanding he come, and Cicero,
who has not seen his daughter since before Pharsalus
and knows she is in a wretched marriage and worse
finances, cannot keep writing for tears. ``\textit{Lacrimae
enim se subito profuderunt.}'' He leaves the
decision to Atticus, asks his pardon, and signs off
with the closing of an exile who has been brought
to depend on a friend for everything: write openly,
write often. The dateline ``x iiii K.\ Ian.'' makes
this and 11.8 effectively a same-day pair from
Brundisium.}
