% Ad Atticum 7.12 --- headnote
% ad-atticum-07-12, 23 January 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian
villa on the tenth day before the Kalends of February,
23 January 49 BC (the manuscript dateline:
\textit{Scr. in Formiano x K. Febr. a. 705 (49)}).
The third letter of the Rubicon-week sequence: Cicero
has come down off the road into the villa at Formiae
and is taking stock --- waiting for Atticus's letters,
trying to read Pompey's mind, sounding the country-town
gossip, and turning the question on himself.}

\textit{Section 1 registers the post: one letter has
arrived, dated 21 January, which mentions an earlier
one that has not. The request is the recurring one of
this week --- write often, even on suspicion, and above
all on what to do. Section 2 is the political readout.
Cicero has seen the consul Lentulus and Libo at Formiae;
the cohorts are at Luceria and Teanum and the rest in
Apulia; whether Pompey will stand at Larinum or cross
the sea, nobody knows. Of Caesar's \emph{Phalarismos}
(the cruelty of Phalaris, the proverbial Sicilian tyrant
who roasted his victims in a bronze bull) Cicero
expects the worst.}

\textit{Section 3 turns the question on himself in the
sharpest form of the week: \emph{demittamne me penitus
in causam?} --- shall I throw myself wholly into the
cause? The Homeric tag \emph{aideomai Trōas} (``I shrink
from the Trojans'') is Hector's line at \emph{Iliad}
6.442; Cicero uses it, with characteristic compression,
for the social and political pull of public reputation.
Sections 4--6 close on what he needs from Atticus:
Manius Lepidus and Lucius Torquatus both set
Pompey-leaves-Italy as the limit of their endurance;
Cicero wants Atticus's own \emph{aporia}, and a
description of the City as it now looks --- whether
Pompey is missed, whether Caesar is hated --- and
counsel on whether Terentia and Tullia should remain
in Rome, come to him, or go somewhere safe.}
