% Ad Atticum 8.9 --- headnote
% ad-atticum-08-09, 25 February 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian
villa on the fifth day before the Kalends of March 49
BC (the manuscript dateline: \textit{Scr.\ in Formiano
v K.\ Mart.\ a.\ 705 (49)}). Cicero's much-discussed
letter urging Caesar toward peace --- written some
weeks earlier --- is now circulating in copies, and
Caesar has answered it in a piece of correspondence
that he in turn has posted publicly. The first half of
this letter is Cicero's defence of what he had written
and an attempt to put the political tactics into
focus; the second half closes on the news of Caesar's
envoy slipping through Formiae by night, and the
appalling speed with which Caesar is advancing on
Brundisium.}

\textit{Sections 1--2 work through the embarrassment
of public correspondence in a civil war. Cicero is not
sorry his letter has been read --- on the contrary, he
wants it on the record that he stood for peace --- but
the published reply has carried in it Pompey's formula
\textit{pro tuis rebus gestis amplissimis}, the very
words in which Pompey had once meant to move Cicero's
own second consulship and triumph; and Atticus and his
friends have gone out cheerfully to greet Caesar at
the fifth milestone. The signs by which loyalty might
be told from pretence, Cicero writes, are getting
muddled. Section 3 reports the daily company of
Lepidus and the steady flow of letters from Tullus,
weighs Atticus's advice against theirs, and reaches
the despairing antithesis at the heart of the letter:
one man (Caesar) winning applause in the foulest
cause, the other (Pompey) earning offence in the best.
Section 4 is the night dispatch: the younger Balbus
hurrying through Formiae to suborn the consul Lentulus
with the promise of a province, and Pompey, travelling
light from Luceria, perhaps already at Brundisium.
Caesar, Cicero ends, is a \textit{teras} of vigilance,
speed, and energy.}
