% Ad Atticum 7.15 --- headnote
% ad-atticum-07-15, 28 January 49 BC, Capua

\textit{Cicero to Atticus, written from Capua on the fifth
day before the Kalends of February in 49 BC (the manuscript
dateline: \textit{Scr. Capuae v K. Febr. a. 705 (49)}).
Cicero, in flight from Rome since Caesar crossed the
Rubicon, has now reached the Pompeian headquarters at
Capua. He has met the consuls and a great many of the
senatorial order, and reports back what the conference
looks like from inside.}

\textit{The picture is grim. The consensus is to accept
Caesar's terms --- only Favonius wants to refuse them ---
and Cato, of all people, ``prefers to serve rather than to
fight'' (\emph{servire quam pugnare mavult}) and is
unwilling to leave the Senate even to take up the urgent
governorship of Sicily. Postumius will not go without him;
the assignment falls instead on Fannius. In section 3
Cicero gives Atticus his own reading: most think Caesar
is bluffing to keep the Pompeian side from arming, but
Cicero believes Caesar will actually withdraw his
garrisons, since once he is consul ``he will have won''
(\emph{vicerit enim}), and won by a lesser crime than the
one he began with. The cold sentence at the close is the
strategic confession: ``we are shamefully unprepared'' in
soldiers and in money, and have left both private and
public funds behind in the city for Caesar to take. Pompey
himself has gone off to the two old Caesarian legions
(the Appianae) with Labienus at his side; Cicero plans to
slip back to his Formian villa at once.}
