% Ad Atticum 10.16 --- headnote
% ad-atticum-10-16, 14 May 49 BC, Cumae

\textit{Cicero to Atticus, written from the Cuman villa
on the day before the Ides of May 49 BC --- 14 May
(the manuscript dateline: \textit{Scr.\ in Cumano prid.\
Id.\ Mai.\ a.\ 705 (49)}). The letter opens on a fresh
private wound. Dionysius --- the freedman tutor of the
boys --- has just come early in the morning. Atticus's
earlier letter, received at Arpinum, had given Cicero
to expect that Dionysius would arrive ready to do what
Cicero wished, namely to go with him; instead, after a
very few words, Dionysius asked to be excused, pleading
his own entangled affairs. Cicero's reaction is sharp
and unguarded: he understood that Dionysius had looked
down on his fortunes; among the greatest griefs of these
days, he reckons this one. The grace note is for
Atticus's sake --- ``I hope he may be a friend to you.
In wishing this for you, I am wishing you happiness;
for then it will last just so long.''}

\textit{The rest of the letter is operational. Cicero
hopes his plan to slip away to Pompey will pass without
danger; he has dissembled and trusts he is not being
watched too sharply. Cato has sailed from Syracuse on
24 April, having given up a Sicily he could have held;
Cotta is rumoured to hold Sardinia (``oh, the shame of
Cato'' if so). On 13 May Cicero went to his Pompeian
villa to lessen suspicion. There the centurions of the
three cohorts at Pompeii sent Ninnius to say they wished
to hand themselves and the town over to him; he left
the villa before dawn so as not to be seen. Three
cohorts were too few, the apparatus too small, and the
overture (as he reflected on a Caelian letter from
Atticus that had just reached him at Cumae) might well
have been a trap. Hortensius has called on Terentia in
his absence and spoken honourably of him --- a contrast
sharpened in the closing thrust at Antony, whose litter,
with an actress inside, is carried in state among his
lictors. The letter ends as the previous one did, with a
nudge to Atticus, now clear of the quartan and a
head-cold besides, to show himself vigorous in Greece
and to keep the letters coming meanwhile.}
