% Ad Atticum 8.12 --- headnote
% ad-atticum-08-12, 28 February 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian
villa on the day before the Kalends of March 49 BC (the
manuscript dateline: \textit{Scr.\ in Formiano prid.\ K.\
Mart.\ a.\ 705 (49)}). Dictated, because the eye-
inflammation (\textit{lippitudo}) that has been bothering
Cicero for days is worse; carried by Gallus Fadius, a
mutual friend; written the day after the long meditation
on the \textit{moderator rei publicae}, and answering it
from a different angle. There the question was the
political diagnosis; here it is Cicero's own conduct
and what role he should now play.}

\textit{Section 1 sets the dictation in its physical
constraint and the request in its urgency: a little
time, the matter is brief, lay out the counsel plainly.
Sections 2--3 are Cicero's own brief in his defence.
\textit{Omnia integra}: nothing has been let slip that
does not have a sound excuse, not merely a respectable
one --- not the refusal of Capua, not the care taken
not to give Caesar cause for offence while Pompey was
already shaping a second consulship and a triumph for
him, not even the latest decision not to follow the
army across the Adriatic. Section 4 is the question
itself: shape me, look out for what is to come, tell me
where I can be of most use --- as peacemaker, or as
soldier. Section 5 returns to an earlier moment in the
friendship: Atticus's advice once relayed through
Theophanes and Culleo, which Cicero now wishes he had
taken. Section 6 closes on the daily business: find
out, by whatever channels, what Lentulus and Domitius
are doing and saying --- whether they are angry at
anyone --- ``what am I saying, at anyone? at Pompey.''
The book of Demetrius of Magnesia ``On Concord,'' asked
for the day before, is asked for again.}
