% Ad Familiares 8.15 --- headnote
% ad-familiares-08-15, around 9 March 49 BC, location uncertain

\textit{M. Caelius Rufus to Cicero, written around the
seventh day before the Ides of March (about 9 March) 49
BC (manuscript dateline \textit{Scr. circ. vii Id. Mart.
a. 705 (49)}). The shortest of Caelius's surviving
newsletters of this year and the first sent after he had
gone over to Caesar. The place of writing is not given
in the dateline --- the \textit{circ.} is a hedge on the
date (``written around''), not a location. The works
register has carried the abbreviation into the
location-written slot in error; the letter was very
likely composed somewhere on the march in north Italy as
Caesar's column moved toward Brundisium and the Spanish
campaign.}

\textit{The voice is unmistakably Caelius's: gossipy,
quick, half-mocking, half-anxious. Section 1 opens with
the rhetorical contrast of the two principals --- Pompey
ineffectual, Caesar keen and temperate --- and the
gibe that Caesar's troops, who walked a war shut by mere
ambulation through the worst of winter, were not exactly
feeding on apples. Then the mood drops: ``if you knew
how anxious I am, you would laugh at this glory of mine,
which has nothing to do with me.'' Section 2 turns to
the immediate trouble --- a local affair at Intimilium
on the Ligurian coast, where a slave-officer named
Bellienus took money to murder a guest-friend of
Caesar's named Domitius, and the town has gone to arms
in protest. Caelius is being sent over the Alpine snows
to put it down. He closes with a Roman witticism on the
two Domitii: the one ``sprung from Venus'' is Caesar
(Iulus, Aeneas, Venus); ``your Domitius'' is L. Domitius
Ahenobarbus, whom Caesar had pardoned at Corfinium; the
one born of Psecas (a slave-woman's son) is Bellienus,
who at least had the spirit to dispatch his Domitius.
The wit covers a complaint: Caesar's clemency at
Corfinium has already begun to look, to Caelius, like
political weakness.}
