% Ad Atticum 7.23 --- headnote
% ad-atticum-07-23, 9 February 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian
villa on the evening of the ninth or the morning of
the tenth of February 49 BC (the manuscript dateline:
\textit{Scr. in Formiano v Id. Febr. vesp. aut iv Id.
a. 705 (49)}). A day after \emph{Ad Atticum} 7.22 a
letter has reached Cicero from his freedman Philotimus
at Rome with the kind of news that bends the room
back upright: Domitius has a strong army, cohorts
from Picenum have joined him, Caesar can perhaps be
cut off. Manius Lepidus, Lucius Torquatus and the
tribune Cassius, all of whom have washed up at the
Formian villa, are revived by it.}

\textit{Cicero is not. He records the bulletin, says
he is afraid it is a dream, and then writes the
sentence the dream is hiding: \emph{Caesarem
Pompeium persequi}. ``Caesar pursuing Pompey? What
for? To kill him? O wretched me!'' The triple
\emph{victi, oppressi, capti plane sumus} --- ``we
are beaten, crushed, taken outright'' --- is the
honest reading. Section 2 is the housekeeping that
the bulletin alters: he had been about to send the
women back to Rome but reverses the order, because
sending them would be read in town as his having
already given up the public cause, the first step of
his own homecoming. He will stay; for Pompey he can
die ``both with piety and with gladness'' --- and the
unfinished half-line, \emph{etsi vivere---}, ``though
to go on living ---,'' is one of the most desolate
sentences he ever wrote. Section 3 closes with the
levy at Capua flat on the floor, a copy of Caesar's
letter enclosed because Atticus had asked, and the
private settlement: many tell me I am acceptable to
him, which I am content to bear, \emph{dum ut adhuc
nihil faciam turpiter}.}
