% Ad Atticum 8.14 --- headnote
% ad-atticum-08-14, 2 March 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian
villa on the sixth day before the Nones of March 49 BC
(the manuscript dateline: \textit{Scr.\ in Formiano vi
Non.\ Mart.\ a.\ 705 (49)}). The daily letters from
Formiae continue. Cicero apologises for them in the
opening sentence and then justifies them: with a
household courier going in any case, he cannot bring
himself to send the man empty-handed, and the writing
itself is a small rest from the misery. Caesar set out
from Corfinium on the Feralia, the same day Pompey
left Canusium; the whole contest of this first phase
is turning at Brundisium, and Cicero is afraid Caesar
will arrive there sooner than he should.}

\textit{Section 2 is the central confession: \textit{labare
meum consilium illud quod satis iam fixum videbatur}
--- the plan, which had seemed by now fairly fixed, is
tottering. The man-by-man case for crossing the sea
with the senatorial leadership does not hold; the
\textit{auctores} Atticus has been citing are not heroes
of the state, and Cicero cannot pretend they are. What
holds him is one man --- Pompey --- whose companion in
flight he seems bound to be. Section 3 returns to
ordinary intelligence-gathering: Domitius's
whereabouts (still unknown; even Lepidus does not
know), Domitius's son (also unknown), and a
disagreeable money matter --- a quite large sum
Domitius had at Corfinium has not been returned. The
manuscript carries a couple of obeli in these
sentences; the sense is recoverable, the wording is
not.}
