% Ad Atticum 9.9 --- headnote
% ad-atticum-09-09, 17 March 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian villa
on the seventeenth day before the Kalends of April 49 BC
(the manuscript dateline: \textit{Scr.\ in Formiano xvi
K.\ Apr.\ a.\ 705 (49)}); the date the letter is sent off
is given in the body as the Liberalia (17 March), which
agrees. Three of Atticus's letters had reached Formiae the
day after the Ides, and Cicero answers them one by one,
oldest first. This is also the day on which, unbeknownst
to Cicero at the moment of writing, Pompey was preparing
to sail from Brundisium: by the next morning the war would
have crossed into a new phase.}

\textit{Section 1 takes up the eldest of Atticus's three
letters --- Formiae as the best place to remain, the Upper
Sea route, and the question whether Cicero can refrain
from touching public affairs with Caesar's goodwill. The
philosophical asides (\textit{sophisteu\=o}, \textit{theseis})
catch his routine of turning over set propositions on the
road from town to country. Section 2 moves to the second
letter: Clodia's troop-numbers were double the truth, the
ships were not destroyed, Lentulus the consul was right in
spirit but wrong in judgment to scatter the senatorial
party, and a famine-war is now certain --- a catalogue of
the eastern fleet (Alexandria, Tyre, Cyprus, Rhodes, and
the rest) being assembled to cut off Italy's grain. The
section ends in a fastidious resolution about how he
should bear himself when meeting Caesar: with weight, not
cringing. Section 3 is the third letter: a long, sharp
reading of Caesar's request for Cicero's \textit{consilium,
gratia, dignitas, ope omnium rerum} --- with the legal
point, drawn from the \textit{De Republica}, that consuls
cannot lawfully be elected by a praetor. Caesar will press
for a senatorial decree to that effect; Cicero will not be
one of those who give it. Section 4 closes on the
miscellany of the Lanuvine villa once owned by Phamea, the
hoped-for visit of Trebatius, and the gathering tempest.
The closing tag, ``Given on the Liberalia'' (\textit{D.
Liberalibus}), fixes the dispatch date.}
