% Ad Atticum 9.11 --- headnote
% ad-atticum-09-11, 20 March 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian villa
on the thirteenth day before the Kalends of April 49 BC
(the manuscript dateline: \textit{Scr.\ in Formiano xiii
K.\ Apr.\ a.\ 705 (49)}). Three days have passed since
Pompey sailed from Brundisium; the news is now beginning
to reach Formiae through travellers, intermediaries, and
Caesar's go-betweens. The letter opens with the
sighting of P.\ Cornelius Lentulus Spinther --- the
consul of 57, Cicero's recall champion --- hiding at
Puteoli; he has slipped away from Pompey's camp and is
sending fulsome thanks to Caesar. The middle of the
letter records a visit from Matius, Caesar's
level-headed friend, on the Quinquatria of Minerva
(19 March); Matius reads Caesar's letter to Cicero as
an overture for peace-broking, and Cicero half-allows
himself the hope of \textit{[Greek: politik\'on opus]}
--- a piece of statesmanship to set against the wreck.
Section numbering jumps from 2 to 4 in the Perseus
text; the standard scholarly numeration is preserved.}

\textit{The letter is dense with Greek. Atticus's tag
\textit{[Greek: n\'ekuia]} --- the underworld
visitation of Odyssey 11 --- has become the shorthand
between them for the company Pompey now keeps;
Cicero echoes it back. Crassipes, Cicero's
former son-in-law, has come from Brundisium with
reports of the Pompeian camp's tone: pure
proscription-talk, pure Sulla, hatred of the optimates
and the municipalities, with Theophanes of Mytilene
loudest of all. The closing paragraph is the bleakest:
the men in whom safety must lie --- Scipio Nasica,
Faustus Sulla, Scribonius Libo --- are reported to
have their creditors meeting in town. Pompey is now
dreaming of Egypt, of Arabia Felix \textit{[Greek:
euda\'imona]}, of Mesopotamia, and has written off
Spain. Cicero closes by sending Atticus a copy of his
own letter to Caesar.}
