% Ad Atticum 5.1 --- headnote
% ad-atticum-05-01, early May 51 BC, Minturnae

\textit{Cicero to T. Pomponius Atticus, written from Minturnae
on 5 or 6 May 51 BC --- Perseus dateline \textit{Scr. Menturnis
iii aut prid. Non. Mai. a. 703 (49)}, where the parenthetical
year is a misprint: AUC 703 = 51 BC, and the contents place the
letter firmly in the first stage of Cicero's reluctant journey
to his governorship of Cilicia. The first two sections wrap up
Roman business at Atticus's door --- the desire that no
extension of the proconsular term be decreed, the giving of
sureties on land transfers, a credit Atticus has just opened
with Oppius for 800{,}000 sesterces.}

\textit{The body of the letter --- and the reason it is so often
quoted --- is the third and fourth sections: a small, exact
account of a domestic blow-up between Cicero's brother Quintus
and his wife Pomponia, who is Atticus's sister. The party stops
overnight at the family estate at Arcanum on the road to
Aquinum; Quintus, in front of Cicero, asks Pomponia to invite
the women in to lunch while he gathers the men; Pomponia says
in everyone's hearing that she is ``a guest here myself,''
declines food sent from the table, locks Quintus out of her
bedroom. Cicero is mortified, takes his brother's side, and
writes to Atticus partly so Atticus will speak to her ---
``you, too, have a part in setting this right.'' Statius is
Quintus's freedman steward, whose having gone on ahead to see to
the lunch is what Pomponia is reacting to. The closing requests
are practical: Atticus is to finish a list of errands before
sailing for Epirus, dislodge the dilatory legate Pomptinus, and
report his own movements.}
