% Ad Atticum 10.15 --- headnote
% ad-atticum-10-15, 12 May 49 BC, Cumae

\textit{Cicero to Atticus, written from the Cuman villa
on the fourth day before the Ides of May 49 BC --- 12 May
(the manuscript dateline: \textit{Scr.\ in Cumano iv
Id.\ Mai.\ a.\ 705 (49)}). Two of Atticus's letters
arrived on the same day, one with Cephalio and one with
Funisulanus, both bringing better news about the eight
cohorts and about the wavering of the local Caesarian
forces. Cicero gives a fully favourable answer to
Funisulanus's business --- the man owes him a large
debt, is not held to be solvent, and now claims he will
pay, though he is delaying the third party to whom he
had charged the obligation; the text here is corrupt
(daggers preserved). Eros, the agent of Philotimus, will
tell Atticus the sum. The closing line of the section
is the pivot: \textit{sed ad maiora redeamus} --- ``let
us return to greater matters.''}

\textit{The greater matter is the secret plan in
Caelius's letter (10.9.2), now ripening. Cicero is
tormented whether to wait for a fair wind; what is
needed is a standard, and men will come flying to it.
He agrees with Atticus that he should go openly, and
intends to set out. Servius's counsel (10.14) has
yielded nothing: every proposition meets some snare; the
only man Cicero has known more timid is Gaius Marcellus,
who, in a striking turn, ``regrets having been consul''
--- prompting an exasperated Greek aside, \textit{[Greek:
\=o poll\=es agenneias]} (``oh, the depths of
ignobility!''). Antony has set out for Capua but sent
word that embarrassment kept him from calling in person.
Sections 4 and 5 turn to domestic and financial chatter:
news of Atticus's sister and of young Quintus; Quintus
the elder's troubles over a loan transfer with Lucius
Egnatius; the brazen Axius and the twelve-thousand
business with Gallius; relief that Atticus and Pilia
are both free of the quartan. Cicero is dashing over
to his Pompeian villa while the ship is being
provisioned; he asks Atticus to thank Vettienus and to
send one more letter, if any courier can be found,
before he sails.}
