% Ad Atticum 7.18 --- headnote
% ad-atticum-07-18, 3 February 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian villa
on the third day before the Nones of February in 49 BC
(the manuscript dateline as Perseus prints it reads
\textit{Scr. imi Formiano iii Non. Febr. a. 705 (49)},
where \emph{imi} is an OCR slip for \emph{in}; the place
is Formiae). Written the same morning as Att.~7.19 ---
the women have just arrived from Rome and Cicero is on
the point of leaving them at the villa to head back to
Capua for the Nones.}

\textit{Section 1 names the dilemma exactly: whether
``a shameful peace or a wretched war'' --- \emph{turpi
pace nobis an misero bello}. Pompey's reply is reported
welcome at the contio, and if Caesar refuses it he is
finished; but the unfinished sentence (``if he accepts
them ---'') concedes that he probably will. To Atticus's
imagined question --- which outcome would you prefer? ---
Cicero answers that he would say, if only he knew what
state the Pompeian preparation was in.}

\textit{Section 2 is the day's military news: Cassius
driven from Ancona, the town now ours. But Caesar, even
while peace envoys are in flight, is reported to be
conducting his levy \emph{acerrime}, seizing positions
and garrisoning them. The outburst is the most acid of
the whole sequence --- \emph{o perditum latronem!} (``the
ruined brigand!'') --- before Cicero collects himself:
\emph{stomachari desinamus}, ``let us cease to fume,
let us yield to the moment, let us go with Pompey to
Spain.'' Section 3 picks up an old loose end (the
philosopher Dionysius, the boys' tutor, who had failed
to follow Cicero out of Rome) with the dry coda that he
does not exact too much of this kind from Greeks.
Section 4 closes with a small but exact picture of the
credit market collapsing under the crisis: even
solvent debtors cannot pay because no one has cash on
hand and no one will renew a loan. The \emph{pseudo-
Hesiodic} maxim Cicero quotes (\emph{mēde dikēn}, ``not
even a lawsuit'') warns against pressing a friend ---
yet his brother's complaint, he confesses, has moved
him.}
