% Ad Atticum 10.18 --- headnote
% ad-atticum-10-18, 19 or 20 May 49 BC, Cumae

\textit{Cicero to Atticus, written from the Cuman villa
on the fourteenth or thirteenth day before the Kalends
of June 49 BC --- 19 or 20 May (the manuscript dateline:
\textit{Scr.\ in Cumano xiv aut xiii K.\ Iun.\ a.\ 705
(49)}). The letter opens with the household's one piece
of private news: Tullia, on the fourteenth day before
the Kalends (19 May), has given birth to a seven-months
boy. The technical Greek does the social work --- the
delivery itself was easy (\textit{[Greek: eutok\=esen]}),
which Cicero is glad of, but the child born is
\textit{perimbecillum}, ``desperately weak,'' and is
not expected to live. Around this domestic fact the
political weather hardens. A spell of extraordinary
calm at sea has held Cicero in the bay --- more
effectively, he notes drily, than the watchers Caesar
has set on him. The Hortensius hopes of the previous
letter have collapsed: the young proconsul has been
corrupted by his freedman Salvius and is now worthless.
Cicero announces a change of practice that the rest
of Book 10 will keep: he will report only what he has
done, never what he is about to do, because every
\textit{[Greek: K\=orukaios]} --- every eavesdropper
of Corycus, in the proverb --- seems to be listening
in on what he says.}

\textit{The rest is bleak operational housekeeping.
Cicero still wants any news from Spain or anywhere
else, and tells Atticus to write without waiting on
letters back; he himself will write only when he has
arrived where he means to go, or if something turns up
en route --- and even this he sets down timidly, so
slow and so thick is everything. ``As we have made a
bad beginning, so the rest follows.'' He is now leaving
for Formiae (perhaps, he says, the Furies will follow
him there too). From Balbus's conversation with
Atticus he gathers that the Malta plan is disapproved
in Caesar's camp: clearly Caesar reckons him among the
enemy. He has written to Balbus relaying what Atticus
reported, both the goodwill and the suspicion;
Atticus is to thank Balbus on his behalf and clear
him on the second point. The letter ends in the
darkest register of the whole book --- ``Has there ever
been a more unlucky man than I?'' --- with Cicero
saying he will spare Atticus the rest, and confessing
that the moment has come when he can no longer act
either bravely or prudently. Two short cruxes in \S1
are preserved as \dag\ markers; the sense given is the
most natural reading of the corrupted text.}
