% Ad Atticum 9.15 --- headnote
% ad-atticum-09-15, 25 March 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian villa
on the eighth day before the Kalends of April 49 BC
(the manuscript dateline: \textit{Scr.\ in Formiano viii
K.\ Apr.\ a.\ 705 (49)}), and continued through the day:
a follow-on to letter 9.14 written within hours. Pompey has
sailed from Brundisium on the Ides of March; Caesar is
moving up Italy; the dreaded meeting is at hand. A fresh
dispatch from Capua corrects yesterday's schedule:
Caesar will now be at the Alban villa with Curio on the
fifth day before the Kalends --- that is, the
twenty-eighth, three days off. Cicero has decided that
after the interview he will go on to Arpinum; if Caesar
grants him the indulgence he is asking (the leave not to
sit in a Senate summoned by Caesar's hand) he will take
the terms, and if not he will fall back on himself. He
notes the legions posted at Brundisium, Tarentum, and
Sipontum, and the inference that Caesar is closing the
seaward exits and aiming for Greece rather than Spain.}

\textit{The second section is the constitutional dread.
Caesar will want a senatus consultum, a decree of the
augurs, a praetor to put the consular question, perhaps
a dictator named: ``neither of these is lawful,'' but if
Sulla could have it done through an interrex, why not
this one? The grim parallel between Sulla and Caesar runs
under the whole letter. Cicero foresees only one of two
fates --- to be killed by Caesar like Quintus Mucius
Scaevola, or by Pompey like Lucius Cornelius Scipio.
Section 3 takes its motto from Odysseus's address to his
own heart: \textit{[Greek: t\'etlathi k\'unteron]},
``endure, more dog-hearted still'' --- and the bitterness
is that even that line, once a private possession, no
longer fits his case, because there is no hope of return
to wait through. Section 4 closes with another Homeric
tag, Telemachus's resignation that ``some things he
himself, some things the divine power will suggest.''
Section 5 is a domestic eruption: Atticus has hinted that
Cicero was too sharp about Dionysius (the freedman tutor
who had abandoned the boys at a bad moment); Cicero
declines to soften. The letter is closed in section 6 by
the dispatch of Matius and Trebatius, who set down
Caesar's itinerary day by day --- Beneventum, Capua,
Sinuessa --- and confirm that Pompey is gone. Several
daggered cruxes in the transmitted text have been kept
as obeli.}
