% Ad Atticum 9.17 --- headnote
% ad-atticum-09-17, 27 March 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian
villa on the sixth day before the Kalends of April 49 BC
(the manuscript dateline: \textit{Scr.\ in Formiano vi
K.\ Apr.\ a.\ 705 (49)}). A very short letter, scribbled
on the eve of the long-foretold interview: Trebatius is
expected today, and from his report --- together with
Matius's letter --- Cicero will rehearse how he must
speak with Caesar himself when the moment comes. Caesar
has had it proclaimed even at Formiae that he wants the
Senate to attend in numbers on the Kalends of April;
Cicero foresees that the central pressure of the
interview will be the demand that he come up to the
City. The line \textit{ergo ei negandum est? sed quid
praeripio?} --- ``am I, then, to refuse him? but why do
I run ahead of myself?'' --- catches the whole posture
of the letter: the question to be put has been
formulated, but the answer is held off until the man
himself is in the room.}

\textit{Two small matters fill the closer of the letter.
Cicero has decided to give his son the \textit{toga
pura} (the plain white toga of manhood) at Arpinum
rather than at Rome, and is turning over where exactly
to go after the interview is past. And he is worried
about Tiro, his freedman secretary, whose own letter
makes him uneasy and whose state those coming up from
the south describe as \textit{[Greek: kindyn\=od\=e]},
``perilous.'' In the present fortune, Cicero writes,
both Tiro's service and his fidelity would be most
useful --- a small private anxiety threaded through the
larger political dread. The letter is the last word
from Formiae before the meeting; what follows in 9.18
is the post-mortem.}
