% Ad Familiares 5.20 --- headnote
% ad-familiares-05-20, early January 49 BC, ad urbem (outside Rome)

\textit{Cicero to M. Mescinius Rufus, written from the
neighbourhood of Rome a few days after the Nones of January 49
BC --- Perseus dateline \textit{Scr. ad urbem paulo post prid.
Non. Ian. a. 705 (49)}. Cicero, returned from his Cilician
governorship in late 50, has not yet entered the city: a
triumph is still notionally in prospect and he is holding
\textit{imperium} outside the pomerium. The political weather
has already broken --- by 7 January the Senate will pass the
\textit{senatus consultum ultimum} against Caesar --- but the
business of this letter is the one Cicero left unfinished
behind him: the rendering of his provincial accounts under the
Lex Iulia of Caesar's consulship, which required them to be
deposited at two cities in the province and then transcribed,
word for word, at the treasury.}

\textit{The recipient is Cicero's quaestor of 51--50, M.
Mescinius Rufus, who is unhappy about how those accounts were
finished. The argument runs through every standard category of
the dispute: the role of Cicero's secretary M. Tullius and of
Mescinius's brother M. Mindius in producing the books; an
arrangement that freed one Volusius from a heavy liability and
shifted it back onto the original contractor Valerius's
sureties (a piece of legal contrivance Cicero is proud of and
credits to his friend the jurist G. Camillus); a deposit, on
Pompey's authority, of money belonging to one Lucceius in a
temple, used by Pompey himself; the missed deadline within
which a returning governor had to name his beneficiaries for
their share of the official allowance; and a final 100{,}000
sesterces over which Mescinius is still aggrieved. Cicero's
closing point is sharp: of the 2{,}200{,}000 sesterces he had
deposited with the publicans at Ephesus, Pompey has carried
the lot off (for the war), so the 100{,}000 in dispute should
be borne in proportion. The tone moves between formal
self-defence and the intimacy of an old patronage relationship
--- ``take all this for a joke, just as I take you'' --- and
the last line refuses Mescinius's apparent suggestion that the
letter be destroyed.}
