% Ad Atticum 7.11 --- headnote
% ad-atticum-07-11, 21 January 49 BC, Campania

\textit{Cicero to Atticus, written somewhere in
Campania around 21 January 49 BC (the manuscript
dateline: \textit{Scr. in Campania inter xiv et ix
K. Febr. a. 705 (49)} --- between 19 and 24 January;
Shackleton Bailey settles on the midpoint). Cicero is
on the roads south, in the days after the
\emph{subito-consilium} letter, with the news shifting
under him as he goes.}

\textit{Section 1 is a single sustained outburst against
Caesar, triggered by the bulletins coming in
(``Cingulum we hold; Ancona we have lost; Labienus has
left Caesar'') and pivoting on the question of whether
the man on the march is a commander of the Roman people
or a Hannibal. Cicero charges him with claiming
\emph{dignitas} where there is no \emph{honestas}, with
seizing cities of fellow-citizens to ease his road home,
with \emph{chreōn} (cancellations of debt) and
\emph{phugadōn kathodous} (recalls of exiles) and the
final, openly tragic Greek phrase that names tyranny as
the greatest of the gods. The personal aside is the
heart of it: he would rather take an hour's sun with
Atticus in his \emph{lucrativum solem} (the ``windfall''
sun-trap of the Roman house) than all the kingships in
the world. Sections 2--3 turn the same alarm on
Pompey: was abandoning the City the right counsel?
Themistocles abandoned Athens, yes, but only because
one city could not bear the wave of a whole barbarian
world; Pericles fifty years later did not, and the
Romans of old, when the City fell, held the Capitol.}

\textit{Sections 4--5 are the steadier civic register.
From the talk of the country-towns Cicero records that
the case is shifting --- the complaint that the City
sits without magistrates and senate is moving people in
Pompey's favour, and now they think nothing is to be
conceded to Caesar. The letter closes with the position
Pompey has assigned Cicero (\emph{episkopos}, overseer,
of Campania and the seaboard, with the levy and the
chief business referred to him) and the request for
news of the \emph{hormē}, Caesar's onset, as often as
Atticus can manage --- since the situation is changeable
and the writing and the reading are themselves what give
him rest.}
