% Ad Atticum 7.14 --- headnote
% ad-atticum-07-14, 27 January 49 BC, Cales

\textit{Cicero to Atticus, written at Cales on the sixth
day before the Kalends of February, 27 January 49 BC
(the manuscript dateline: \textit{Scr. Calibus a. d. vi
K. Febr. a. 705 (49)}), as Cicero sets out from Cales
for Capua. The last letter of the Rubicon-week sequence:
the southward retreat from the City has become the
inland march to join Pompey, and a tentative peace
proposal has reached the camp.}

\textit{Section 1 records the breaking news. Lucius
Caesar (Caesar's distant cousin, then acting as his
envoy) has brought terms to Pompey at Teanum: Caesar to
withdraw his garrisons from towns outside his province,
and in that case the senate at Rome to be allowed to
settle the matter. Cicero hopes the peace will hold for
the moment --- the symmetrical clause is the sharp one:
\emph{illum furoris et hunc nostrum copiarum suppaenitet},
``that man is somewhat regretting his frenzy and our own
is somewhat regretting his forces.'' Section 2 is the
levy at Capua, where the Campanian colonists are slow,
and a small bright spot: Caesar's gladiators in the
training school at Capua have been quietly distributed
two apiece among householders --- five thousand shields
were in the school, and a breakout was rumoured.}

\textit{Section 3 returns to the women. Cicero wants
Terentia and Tullia (and Pomponia, Atticus's sister)
out of Rome --- it does not look decent for them to
stay when their peers have gone; the seaboard estates
in his own district can hold them. The closing line is
the one to underline: \emph{ad pacem hortari non desino;
quae vel iniusta utilior est quam iustissimum bellum cum
civibus} --- I do not cease to urge peace; which even
unjust is more useful than the most just war with citizens.
The Att 7.10--7.14 sequence ends here, on that sentence.}
