% Ad Familiares 16.11 --- headnote
% ad-familiares-16-11, 12 January 49 BC, ad urbem (the suburbs of Rome)

\textit{Cicero to his freedman Tiro, written from the
suburbs of Rome on the day before the Ides of January
49 BC (the manuscript dateline, garbled in Perseus as
\textit{pdd. Id. Ian.}, is \textit{prid. Id. Ian.}).
Tiro is still at Patrae, recovering from the illness
that kept him behind when Cicero sailed home from
Cilicia; this is the first letter back to him from
Italy since Cicero reached the city walls. He is at
the suburbs and not within them because, holding the
\textit{imperium} of a returning proconsul, he is
still keeping his claim to a triumph alive.}

\textit{Section 1 is the affectionate medical opening
the Tiro letters have settled into --- the quartan
fever, the order not to risk a winter crossing, the
worry that goes through every line. Sections 2 and 3
are the news. The senate's emergency decree
(\textit{ut curaremus ne quid res p. detrimenti
caperet}, the \textit{senatus consultum ultimum}) has
just passed; Antony and Cassius have left the city to
join Caesar; Curio has been egging Caesar on; the
regions of Italy have been distributed among the
commanders, and Cicero has been given Capua. This is
the news he is sending in parallel to Terentia and
Tullia at home --- Tiro abroad is being kept on the
same footing as the household. The closing
\textit{etiam atque etiam vale} is the standard sign-off
of these letters; it does double duty here, written as
the world tips toward war.}
