% Ad Atticum 9.18 --- headnote
% ad-atticum-09-18, 28 March 49 BC, Arpinum

\textit{Cicero to Atticus, written from Arpinum on the
fifth day before the Kalends of April 49 BC (the
manuscript dateline: \textit{Scr.\ Arpini v K.\ Apr.\ a.\ 705
(49)}). This is one of the most consequential letters in
the whole correspondence: Cicero's eyewitness report,
written the same day, of his face-to-face meeting with
Caesar at the Formian villa on the eve before. Caesar
on his march up from Brundisium had stopped at Formiae;
Cicero, who had spent weeks rehearsing the encounter in
letter after letter, finally had it; the interview
followed two lines of advice Atticus had given --- speak
so that he should think well of you rather than be
grateful, and hold the line against going up to the
City --- and on both Cicero says he succeeded. He had
expected Caesar to be \textit{facilem}, easy; he found
him the opposite (\textit{nihil vidi minus}, ``I have
seen no one less so'').}

\textit{The heart of the letter is the reported exchange
in section 1, in which the demand at the centre of the
whole correspondence finally crystallises. Caesar wants
Cicero to come to Rome and speak in the Senate on the
Kalends; Cicero answers that if he comes, he will say
things Caesar will not want said --- that the Senate
does not consent to a crossing into Spain, does not
consent to armies being transported into Greece, that he
will lament much on Pompey's behalf. Caesar's reply ---
\textit{ego vero ista dici nolo}, ``I will not have such
things said'' --- is the substance of the breach,
delivered without ornament. Caesar leaves the room
asking that Cicero deliberate; Cicero understands that
asking-to-deliberate as a way of seeking an exit. The
two part. The letter's most-quoted line follows: ``I
take it this man does not love me. But I have loved
myself --- which has not happened to me for a long
while now.'' The closing \textit{[Greek: katakleis]}
that Cicero says he almost passed over is Caesar's
parting threat: if my counsels are unavailable he will
use whose he can, and stoop to anything.}

\textit{Section 2 turns to the entourage --- the
\textit{[Greek: n\'ekuia]}, the ``calling-up of the
dead,'' a tag from Homer that Atticus had used of
Caesar's circle, with the ``\textit{[Greek: h\=er\=os]}
Celer'' singled out for sarcasm --- and to the bitter
recognition that the sons of Servius Sulpicius Rufus
and of Titinius were in the camps that had Pompey under
siege at Brundisium. Section 3 records the day's
itinerary: Caesar going on to Norbanus at the Pedian
villa, Cicero turning back inland to Arpinum, where he
will give his son the white toga. Section 4 closes on
the demand the letter itself is sending up the chain:
\textit{epistulam et \textit{[Greek: politik\=en]}!},
``a letter --- and a political one!'' The unbroken
Formiae diary of weeks ends here; the next letter is
from Arpinum.}
