% Ad Atticum 7.17 --- headnote
% ad-atticum-07-17, 2 February 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian villa
on the fourth day before the Nones of February in 49 BC
(the manuscript dateline: \textit{Scr. in Formiano iv
Non. Febr. a. 705 (49)}). Cicero is back at Formiae,
having returned from Capua a day or two earlier and
waiting for Terentia and Tullia to join him; the consuls
have summoned him back to Capua again for the Nones,
three days off.}

\textit{The letter is the longest of this five-day cluster
and the most analytic. Section 1 opens with a dry joke
about city property values (no one, Cicero says, has
knocked so much off them as Pompey has): Atticus need not
flee Rome with him. Section 2 is the political reading.
Lucius Caesar, sent as intermediary, is shuttling
Pompey's reply back to Caesar; the reply is conciliatory
enough to be posted in public, and Cicero is furious that
Pompey --- himself an able writer --- has let it be drafted
by Sestius, whose stilted style yields the coinage
\emph{Sēstiōdesteron} (``more Sestius-like'' than
anything ever read). The substance is worse than the
style: ``nothing is being refused to Caesar.'' Caesar
will likely take the deal --- yet Cicero already fears he
will not be content with terms this favourable, since he
is reported to be \emph{acerrimus}, at his most ferocious,
even as the answers come back.}

\textit{Sections 3--4 turn personal. Trebatius writes that
Caesar, on hearing of Cicero's withdrawal from Rome,
began to fret at the senatorial absences and asked
Trebatius to recall him to the city; Cicero is struck
that Caesar approached him via the obscure jurist
Trebatius rather than directly or through Dolabella or
Caelius. He writes back that he will stay on his own
estates so long as there is hope of peace --- but if war
comes, he will not fail his duty or his dignity, with the
boys (his son and nephew) \emph{hypektithemenos},
``smuggled out for safekeeping,'' to Greece. Section 5
records the day's domestic detail: the women are on the
road, Rome is reported more frightened than before, and
he must turn round again for Capua at the Nones.}
