% Ad Atticum 8.2 --- headnote
% ad-atticum-08-02, 17 February 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian
villa on the thirteenth day before the Kalends of
March 49 BC (the manuscript dateline: \textit{Scr. in
Formiano xiii K. Mart. a. 705 (49)}). One day after
\textit{Att.}~8.1, with Atticus's reply now in hand
and burned at the lamp by which Cicero is writing
this one back. The two letters together are the
clearest record we have of Cicero arguing with himself
about whether to follow Pompey out of Italy.}

\textit{Section 1 takes up the practical matters of
the day: a short, conciliatory letter sent to Caesar
from Capua, on the matter of his gladiators, which
Cicero would be content to have posted up in public;
a second letter sent the same day, prompted by
notes from Caesar and from Balbus, copy enclosed
for Atticus's inspection. Section 2 turns serious.
Atticus has counselled him to remember what he has
done, said, and written; Cicero will not be lectured
on his own honour, because what Pompey has done in
abandoning the city is, to Cicero's mind, the most
disgraceful act any leader of any nation has ever
committed: ``the fatherland, for which it was a
glorious thing both to die and to die in.''}

\textit{Section 3 is the bitter catalogue of the
flight: the wandering with wives and children, the
hopes placed on the breath of one ailing man, the
loyalists who will in fact stay home. Section 4
brings the famous comparison with Socrates, who under
the Thirty Tyrants did not set foot beyond the gate
of Athens, and ends with the lamp-conceit and the
unanswered question that hangs over book 8 ---
\textit{si de pace agetur, profecturus, si de bello,
quid ero?} ``setting off for Pompey, if it is to be
peace; if war, what shall I be?'' The manuscript has
a crux at \textit{†sed cur†}, preserved here with the
daggers; the Greek tags --- \textit{mempsis}
``reproach'' in section 1, \textit{diple} for the
editorial mark to look out for in section 4 --- are
of a piece with the conspiratorial register of these
days.}
