% Ad Atticum 8.16 --- headnote
% ad-atticum-08-16, 4 March 49 BC, Formian villa

\textit{Cicero to Atticus, written from the Formian
villa on the fourth day before the Nones of March 49
BC (the manuscript dateline: \textit{Scr.\ in Formiano
iv Non.\ Mart.\ a.\ 705 (49)}). This is the closing
letter of \textit{Ad Atticum} book 8 --- the same
end-of-book dramatic function that 7.26 served the
month before, and a much darker note to close on.
Everything has been provided, Cicero begins, except a
hidden and safe road to the Adriatic; the Tyrrhenian
sea is unusable in early March. Section 1 hardens the
verdict on Pompey: not the man who seems to lead me,
\textit{apolitikotaton} (least political) of all men,
recognised so long since, but now also
\textit{astrategetotaton} (most lacking in
generalship). What pulls him on is \textit{sermo
hominum} --- talk of men --- carried in reports from
Philotimus that the \textit{optimates} are tearing him
to pieces for staying. The reply to that is the
withering parenthesis on those same \textit{optimates},
who are hurrying out to meet Caesar and selling
themselves to him by the town.}

\textit{Section 2 calls Caesar \textit{hic Pisistratus}
--- the tyrant of Athens analogy, the textbook example
of a \textit{tyrannos} who came in clemently and stayed
on as ruler --- and observes that whatever evil he does
not do will be received as a favour. The 360 jurors who
once acquitted Milo and stood with Pompey now shudder
at \textit{nescio quas eius Lucerias} --- ``certain
Lucerias of his,'' the rumoured proscription list
Pompey was said to be drawing up at his Apulian camp.
The Homeric \textit{aidéomai} (``I feel shame before
them'') concedes the pull of the senatorial cause even
as the sentence that follows guts it: he is joining
himself with a man more prepared to ravage Italy than
to win the war, and waiting for a master. If Caesar
should come up the Appian Way, Cicero is thinking of
Arpinum --- his own hill town, far from any road
between Brundisium and Rome. Book 8 ends on that
contingency: the master's route is the geometry of
Cicero's next move.}
