% Ad Familiares 9.25 --- headnote
% ad-familiares-09-25, 13 February 50 BC, Laodicea

\textit{Cicero to L. Papirius Paetus, written at Laodicea in
Cilicia about the thirteenth of February 50 BC (Perseus:
\textit{Scr.~Laudiceae post iii Id.~Febr.~a.~704}). Though the
letter stands in Book 9 --- which is otherwise the cluster of
genial notes to Paetus and Varro from 46 BC --- it belongs four
years earlier, to Cicero's year as proconsul of Cilicia, when he
held the title of imperator after his operations in the Amanus.
The salutation accordingly reads \textsc{Cicero Imp.~Paeto}, not
the bare \textsc{Cicero Paeto} of the later correspondence; the
editors have grouped it with the rest of the Paetus letters by
addressee rather than by date.}

\textit{The two movements are sharply distinct. First the joke:
Paetus, the cultivated stay-at-home Epicurean, has evidently sent
his soldier-friend some military advice, and Cicero plays it back
with mock gravity --- the letter has made a ``supreme commander''
of him, he has been studying the treatises of Pyrrhus and of
Cineas, he will keep light craft along the coast against the
Parthian horse. ``But why are we playing? You do not know what
kind of general you are dealing with'' --- and the punch line is
bookish, not martial: the general's true campaign manual is
Xenophon's Cyropaedia, the \textit{Paideia Kyrou}, which he has
read to tatters and now ``unfolds entire'' in his province. Then,
with \textit{sed iocabimur alias}, the tone turns and the real
errand emerges: Cicero asks Paetus to intervene for his friend
M.~Fadius, whose brother Q.~Fadius has put up for sale a jointly
held estate at Herculaneum. Cicero wants Paetus to use his
authority, counsel, and influence to keep the brothers out of
disgraceful litigation against the enemies Mato and Pollio. The
letter thus moves from the lightest banter to a pointed request
for a favor, the characteristic register of the Paetus
correspondence carried back into the Cilician year.}
