% Ad Familiares 3.13 --- headnote
% ad-familiares-03-13, 10 August 50 BC, Rhodes

\textit{Cicero to Appius Claudius Pulcher, written from
Rhodes around the 10th of August 50 BC (Perseus
dateline: \textit{Scr. Rhodi circ. iv Id. Sext. a. 704
(50)}). It is a short follow-up only six days after
\textit{Ad Familiares} 3.12, written from the next
stop on Cicero's return voyage. The letter answers
fresh news that has reached Cicero of Appius's
service to him during the prosecution: not only had
Appius spoken and voted in his favour, but he had
gone to Cicero's house, met with Cicero's people, and
left no piece of friendship's office for anyone else
to perform. It is the moment in the correspondence at
which Cicero, having now folded the Dolabella
engagement into his calculations, openly says that the
weight has been returned to him in fuller measure than
he sent it out.}

\textit{The body of the letter is a small set-piece on
\textit{amicitia} as its own reward: ``the outward
marks of virtue many have attained even without
virtue; the zeal of such men, virtue alone can
attain.'' The fruit of friendship, Cicero says, is
the friendship itself. Beneath this lies the practical
note that the new tie of \textit{adfinitas} through
Tullia's marriage to Dolabella has not subtracted from
his goodwill toward Appius, but added to it. The
letter is shorter and (Cicero says) the more modest
in form because he is writing to a man he hopes is
now \textit{censor} --- master of public morals --- a
status Appius would in fact assume not long after.}
