% Ad Atticum 6.2 --- headnote
% ad-atticum-06-02, May 50 BC, Laodicea

\textit{Cicero to Atticus, written at Laodicea in the
month of May 50 BC (the manuscript dateline:
\textit{Scr. Laodiceae in m. Mai. a. 704 (50)}). The
sequel to \textit{Ad Atticum} 6.1: the great assize
circuit of the province is finished, the proconsular
year is running out, and the Salaminian loan affair is
still on the table. Sent off with the same Philogenes,
Atticus's freedman, who is returning by sea to Italy.
The letter is laid out backwards from Atticus's
preceding bulletin: Cicero first answers the last
page of it (the gossip about Statius, his brother
Quintus's freedman, and the rumour that some plan
against the family is approved by Cicero himself),
then comes round to the first (a learned point about
the Peloponnese, drawn from Dicaearchus). Only then
does he turn to the substance.}

\textit{Sections 4--5 are the great central panel: the
first connected account of Cicero's own provincial
administration. The cities of the assize district have
been freed or relieved of their debts by two means ---
not a farthing of expense charged to them, and the
return of plundered public funds by their own Greek
magistrates. The publicani are paid in full for the
current lustrum and even the arrears of the previous
one. The portrait is grave and pleased; the technical
vocabulary (\textit{continentia}, \textit{integritas},
\textit{αὐτονομία}) sits inside a confident
moral self-portrait, not a defensive one. Sections
7--9 then come back to the scandal at the heart of
\textit{Ad Atticum} 6.1: Brutus's bond on the
Salaminians at \textit{quaternae centesimae}
(forty-eight percent per annum), in defiance of the
Lex Gabinia, enforced by the agent Scaptius with a
cavalry squadron lent by Appius Claudius. Cicero has
forced the Salaminians to pay at the lawful twelve
percent renewed yearly; Scaptius has refused; and
Cicero has refused, in turn, both the squadron and the
prefecture for Scaptius. The letter closes (\S 10)
with the usual spray of news --- Caelius the
incoming quaestor, Pompey's good offices for Appius,
Sempronius Rufus, the Pammenian business --- and the
hurried farewell as the crowd presses and Philogenes
sets sail.}
