% Ad Familiares 3.10 --- headnote
% ad-familiares-03-10, April 50 BC, Laodicea

\textit{Cicero to Appius Claudius Pulcher, written from
Laodicea in April 50 BC (Perseus dateline:
\textit{Scr. Laudiceae m. April a. 704 (50)}). This is
the longest of the Appius letters, written while the
\textit{maiestas} prosecution was still in prospect ---
some weeks before the acquittal celebrated in
\textit{Ad Familiares} 3.11. Cicero has had two
communications from Appius about the gathering case:
the short note delivered by Q. Servilius (``too long
for me,'' Cicero says, ``because I felt I was being
wronged in being asked at all''), and a further
letter listing the marks of zeal being shown for
Appius by the orders at Rome. The present letter is
Cicero's full reply: a pledge of total support, a
defence of his own conduct in the province against
the suspicions Appius's letter had let fall, and a
long demonstration --- almost a forensic brief --- of
why it would be psychologically absurd to suppose
Cicero his enemy.}

\textit{Section 1 sets out the case in high register
and gives the letter its memorable formula. The most
rightful triumph, Cicero says, has indeed been taken
from Appius by the conspiracy of the envious; but if
he bears the prosecution as he should, ``as victor
you will conduct, out of the grief of your enemies,
the most rightful triumph of all'' (\textit{triumphum
iustissimum ex inimicorum dolore}). Cicero promises to
play four roles for him at once: \textit{deprecator}
(intercessor), \textit{propinquus} (kinsman by zeal),
\textit{auctoritas cari hominis} (a dear man's
authority among the communities), and the
\textit{gravitas} of a sitting commander. Sections 3
through 5 fill out the immediate context. Pomptinus,
Cicero's lieutenant, had been on the point of sailing
from Ephesus when news of Appius's trouble reached
him, and turned back to Laodicea on Appius's account.
The ``young man'' of section 5 is the prosecutor
Dolabella, whom Cicero himself had twice defended in
capital trials and whose folly in taking on Appius's
enmities Cicero finds astounding.}

\textit{Sections 6 through 9 are the defensive heart
of the letter. Appius's note had carried a hint of
suspicion --- that Cicero, by certain provisions of
his proconsular edict, had been less than friendly to
the embassies being sent on Appius's behalf. Cicero
demolishes the suggestion in a piece of close
reasoning: the only intervention he had made on the
embassies was to suggest, not order, that their
expenses be reckoned by the Cornelian law; the accounts
of the communities are the witnesses. To paint him as
cunning is incoherent --- a cunning man would not
choose for his perfidy the very point that would
declare hatred plainest and inflict damage least.
Section 10 closes the demonstration with the great
peroration on Pompey: Cicero values Pompey alone, as
he ought, most of all --- ``\textit{quem unum ex
omnibus facio, ut debeo, plurimi}'' --- and since
Pompey's son is now Appius's son-in-law (the marriage
of the elder Pompey's son to Appius's daughter
Claudia is meant), the \textit{adfinitas} alone makes
any other posture toward Appius impossible. The
letter closes in section 11 with the hope that Appius
will soon hold the censorship, the duties of which
Cicero pointedly says will require more careful
pondering on Appius's part than the labour Cicero is
putting in for him at Rome.}

\textit{Note: \textit{works.yaml} carried this letter
at year-precision November 50 BC, which is wrong.
Perseus's dateline puts it in April, and the internal
references (the trial still in prospect, Pomptinus
turning back from Ephesus, the censorship still
``hoped for'') confirm April. The corpus date should
be \textit{-0050-04}, place Laodicea, ordered before
\textit{Ad Familiares} 3.11.}
