% Ad Atticum 6.9 --- headnote
% ad-atticum-06-09, 15 October 50 BC, Athens

\textit{Cicero to Atticus, written from Athens on the
Ides of October 50 BC (the manuscript dateline:
\textit{Scr. Athenis Id. Oct. a. 704 (50)}). Cicero
landed at the Piraeus on the 14th, picked up Atticus's
letter from his slave Acastus at once, and on the
following day wrote this reply: the last letter of
\textit{Ad Atticum} book 6, sent off just before he
sails for Italy. By the time it lands in Rome Caesar
has crossed the Rubicon four months later; the world the
letter belongs to will have ended. He does not yet know
that, but the alarm is in every paragraph.}

\textit{Section 1 opens with what he can read in the
physical packet itself: the seal short, the hand
disordered (\textit{sunchusin tōn grammatōn},
``confusion of the little letters'') --- and at the
close the explanation, that Atticus had reached Rome
with a fever. The diagnostic is loving and exact and
ends in the trust that good sense and self-discipline
will have the upper hand. Section 2 turns to a piece
of domestic management --- a legacy a certain
manipulative party is not to be allowed to touch ---
wrapped in dense Greek (\emph{paraphylaxon},
\emph{philotimian}, \emph{autotata}, \emph{kenon},
\emph{atyphon}) of the kind Cicero uses with Atticus
only. Section 3 is the running argument about the
nephew Quintus: Atticus had urged that the boy not be
left behind in the province, and had registered his
broader judgement about the handover to brother Quintus
under the verb \emph{epechein}, ``to suspend judgement.''
Cicero, having now done both, reads the suspension as
in fact a rejection (\emph{athetēsis}) rather than a
withholding (\emph{epochē}) and is amused, and grateful,
that they reached the same conclusion as though they
had talked it out. Section 4 closes with the homecoming
checklist: a letter from Atticus to follow him on Tullia's
marriage to Dolabella, on the censors, on the question
of the statues and pictures and whether the matter is to
be raised --- and the cold news he is recording as he
writes, \textit{quo die, ut scribis, Caesar Placentiam
legiones IIII}, ``on which day, as you write, Caesar
reaches Placentia with four legions.'' The next clause
is the one he says aloud: \textit{quaeso, quid nobis
futurum est?} ``I ask you, what is to become of us?''
The very last sentence is a half-jest, half-prayer ---
his station now, he says, is the citadel at Athens.}
