% Ad Familiares 3.7 --- headnote
% ad-familiares-03-07, mid-February 50 BC, Laodicea

\textit{Cicero to Appius Claudius Pulcher, his
predecessor as governor of Cilicia, written hastily
from Laodicea around the Ides of February 50 BC
(Perseus dateline: \textit{Scr. Laudiceae circiter Id.
Febr. a. 704 (50)}). The Apamean ambassadors had
brought Cicero a long letter of complaint from Appius:
that Cicero had stopped the Apameans' building work
with one of his letters, that he had forbidden them to
collect their tributes before he could investigate, and
--- in a private aside relayed by Lentulus's freedman
Pausanias --- that Cicero had snubbed his predecessor
by failing to come out to meet him on the road into
Iconium.}

\textit{Cicero replies in three movements. He defends
the order about the tributes as fair (and as exactly
what the Apameans themselves were asking for); he gives
the practical narrative of the missed meeting at
Iconium, where he had sent both Varro and Lepta out to
intercept Appius on the two roads, only for Appius to
slip past; and he refuses, in the most studied way,
Appius's recital of how every Claudius and Lentulus
came out to meet every other Claudius and Lentulus.
``Do you suppose any \textit{Appietas} or \textit{Lentulitas}
weighs with me more than the ornaments of virtue?'' is
the centre of the letter --- a Stoic point made
politely. Cicero closes with a Greek tag from the
\textit{Iliad} (Achilles to the embassy), and signs off
with the assurance that the goodwill he has shown is
``taken up on settled judgment'' and will be maintained
as long as Appius wishes. The dossier of patching-up
in the Cilicia handover.}
