% Ad Familiares 8.13 --- headnote
% ad-familiares-08-13, late May or early June 50 BC, Rome

\textit{M. Caelius Rufus to Cicero in Cilicia, written from
Rome late in May or early in June 50 BC (manuscript dateline
\textit{Scr. Romae ex. m. Mai. aut in. Iun. a. 704 (50)}).
A short two-section bulletin. The first sentence congratulates
Cicero on the new marriage of his daughter Tullia: in his
absence and without his foreknowledge, Tullia and Terentia
had betrothed her to P. Cornelius Dolabella --- a young
patrician of brilliant connections but spoiled habits, by
his second wife already. Caelius, who knew Dolabella from
the same political and convivial circle, recommends the man
on a calculation: he will be steadied by Cicero's authority
and Tullia's modesty; he is not pugnacious in his vices, and
he understands what would be better. The same matrimonial
news Cicero is meanwhile reporting in his own correspondence
with stoic resignation.}

\textit{The political paragraph reports the second great
crisis of the year: the proposed recall of one of Caesar's
legions and of one of Pompey's, ostensibly for the Parthian
war but in fact a manoeuvre to disarm Caesar at a stroke.
Curio has interposed his veto on the senatorial action
against Caesar's tenure, and the Senate has gone "in alia
omnia" --- voted the other way --- on the motion that would
have constrained him. Caelius's verdict on Pompey is
withering and exact: he is "languenti, ut vix id quod sibi
placeat reperiat" --- so spiritless that he can hardly even
find what he himself wants. The closing tag, "you old rich
gentlemen will see to it," is a young politician's swipe
at the senior generation who control the Senate but cannot
or will not lead it. The notice that Hortensius is dying as
the letter is being written closes an era: with him goes the
last of Cicero's great rivals at the bar.}
