% Ad Familiares 8.6 --- headnote
% ad-familiares-08-06, late February 50 BC, Rome

\textit{M. Caelius Rufus to Cicero, written from Rome before
the end of February 50 BC (Perseus: \textit{Scr. Romae ante
ex. m. Febr. a. 704 (50)}). Cicero is in the last weeks of his
year as proconsul of Cilicia; Appius Claudius Pulcher, his
predecessor in the province and the man whose accounts he
spent his governorship trying to disentangle, has just been
indicted in Rome by P. Cornelius Dolabella --- soon to be
Cicero's son-in-law --- on charges of maiestas and ambitus.
Caelius opens with the politics of the case: Appius has
handled himself adroitly, scotching the worst of the
ill-will by quietly dropping his pending triumph and walking
into the city as a private citizen. He now looks to Cicero
for support, and Caelius takes care to spell out the line
Cicero must walk. He has no personal feud with Appius and
can oblige him as he likes, but if he stands on the strict
law he will look churlish about laying past enmity aside;
the safe ground is to be generous, since no one will accuse
him of having sacrificed duty to friendship. A pointed
delicate note follows: between the demand for indictment
and the formal naming, Dolabella's wife (Fabia) has left
him --- the move that will shortly clear the way for his
marriage to Tullia.}

\textit{The remainder is a survey of Roman political weather.
Pompey is working hard for Appius and may send one of his
sons out to plead with Cicero in person. The juries are
acquitting indiscriminately --- ``the wickerwork is foul
through and through.'' The consuls of 50, L.~Aemilius
Paullus and C. Claudius Marcellus, are so cautious that the
only senatorial decree they have managed to pass concerns
the Latin Festival. Curio's tribunate, on which the
optimates had pinned their hopes against Caesar, is
described first as ``frozen solid'' --- and then, in a
postscript, as suddenly red hot: angered by being denied his
intercalary month, Curio has gone over to Caesar and is
flinging out a road bill (compared, dismissively, with the
notorious agrarian proposal of Rullus that Cicero crushed in
63) and a corn bill that bestows the dole-measuring on the
aediles. The closing line returns to the running gag of the
correspondence: Cicero, hunting Cilician beasts for Caelius's
upcoming aedilician games, must not let him down on the
panthers.}
