% Ad Atticum 6.3 --- headnote
% ad-atticum-06-03, before 27 June 50 BC, Cilicia

\textit{Cicero to Atticus, written from somewhere in
Cilicia before the fifth day before the Kalends of
Quintilis (before 27 June) 50 BC (the manuscript
dateline: \textit{Scr. in Cilicia ... ante v K. Quint.
a. 704 (50)}). The next letter after \textit{Ad
Atticum} 6.2: sent off with the freedman Philotimus,
who is travelling back to Rome. The proconsular year
is now visibly running out --- Cicero must leave the
province on the third day before the Kalends of
Sextilis (30 July) --- and there is still no
successor, and no obvious candidate to leave in
charge. The whole letter is held in the gravitational
field of that single structural problem.}

\textit{Three threads weave through. First, the
succession (\S\S 1--3): the only acceptable candidate
of praetorian rank is Cicero's brother Quintus, who
hates the province and cannot well be asked; Pomptinus
has already gone home by prior agreement; the quaestor
is \textit{levis, libidinosus, tagax} --- and a great
Parthian war is presumed to be coming. Second, the
political weather at Rome (\S 4): tiresome news of
Curio and Paulus; Cicero, anxious about the
Caesar/Pompey trajectory, asks Atticus to send him a
sketch of the state of the commonwealth so that he can
think his way back into a city he has begun to feel
foreign to. Third, the Brutus-Scaptius-Salaminian
affair, which is now Cicero's fixed thorn (\S\S 5--7):
the Cypriots were counting out the money, but Scaptius
would not have it at twelve percent renewed yearly;
the prefecture has been refused again, both for
Scaptius and for the Clodian dependant Gavius, who
parted from Cicero at Apamea with open insolence; and
Brutus, in every letter, has had ``something
arrogant, something \textit{ἀκοινονόητον}.'' The
closing paragraphs (\S\S 8--9) drop into the family
news: Quintus the younger, having read the troubling
letter to his father about his mother, has wept --- a
moment of unexpected sweetness in a boy Cicero has
otherwise begun to worry about --- and the younger
Hortensius is making a disgraceful spectacle of himself
at the Laodicea gladiatorial games.}
