% Ad Familiares 2.18 --- headnote
% ad-familiares-02-18, early May 50 BC, Laodicea

\textit{Cicero to Q. Minucius Thermus, propraetor of Asia, written
from Laodicea at the beginning of May 50 BC (the manuscript dateline:
\textit{Scr. Laudiceae in. m. Maio a. 704 (50)}). One of the cluster
of letters Cicero addressed to Thermus across the months of his own
Cilician proconsulship; this one carries advice on a piece of
provincial succession.}

\textit{Thermus is approaching the end of his year, and the question
is whom to leave in charge of the province in the interval between
his own departure and his successor's arrival --- his own quaestor
(young, well-born, the proper man by rank) or one of his three
veteran legates. The legates are Cicero's friends and excellent men;
the quaestor is unnamed in the letter, but a youth from a powerful
house with three brothers in line for the tribunate of the plebs over
the next three years. Cicero's counsel is unambiguous and personal:
do not slight the young man and his family. The argument is delivered
in the voice of practical wisdom rather than principle --- the quaestor
outranks legates of quaestorian standing, so preferring him incurs no
reproach; if he turns out worthy, Thermus shares the credit; if he
gives offence, the offence is wholly his own. The closing sentence is
a small Ciceronian formula: \textit{tu quod egeris, id velim di
adprobent} --- whatever you do, may the gods approve it --- which
acknowledges the decision is Thermus's, even as the recommendation
has been pressed home.}

\textit{The thinking parallels the choice Cicero would make for his
own province four months later, when he set his quaestor Caelius
Caldus over Cilicia on departure (\textit{Ad Familiares} 2.15.4) ---
on closely related reasoning about not alienating a young man of
standing.}
