% Ad Atticum 7.8 --- headnote
% ad-atticum-07-08, 27/28 December 50 BC, Formiae

\textit{Cicero to Atticus, written from his Formian villa on
the sixth or fifth day before the Kalends of January, i.e.
27 or 28 December 50 BC (Perseus dateline:
\textit{Scr. in Formiano vi aut v K. Ian. a. 704 (50)}). The
report letter after a face-to-face meeting with Pompey: the
political content of the cluster suddenly thickens, because
for the first time in months Cicero has spoken with the man
he is bound to follow.}

\textit{The first three sections discharge domestic items.
He drops the Dionysius quarrel on Atticus's say-so; he is
glad Atticus can come to him at the Alban villa, but begs
him not to make the journey if the quartan is on him; and he
turns over, as a \emph{politikon skemma} ``civic problem,''
whether Dolabella, just installed as Tullia's husband, ought
to take a new name under Livia's will. The lawyer's note is
characteristic: settle the principle once we know the actual
sum involved (``how much, so to speak, lies in a third of a
third'').}

\textit{Section 4 is the heart. Pompey caught up with him at
Lavernium on the 26th and the two rode on to Formiae, where
from the eighth hour to evening (about 2 p.m. until dark)
they conferred in private. The headline finding: Pompey does
not wish for a peaceful settlement. Either Caesar, even on
dismissing his army, would be a consul who would produce
\emph{sunchusin tēs politeias}, ``a dissolution of the
constitution''; or, hearing of the preparations against him,
he would simply hold on to province and army. Pompey is
confident in his own forces and in the commonwealth's, and
contemptuous of Caesar should he ``go mad.'' Cicero records
his own private flicker --- the Homeric \emph{xunos
Enualios}, ``Enyalios is impartial'' (\emph{Iliad} 18.309),
his caveat against confidence in war --- but says he was
lifted, all the same, by hearing a brave man speak
\emph{politikōs} of the dangers of a sham peace. They read
together the speech that Antony, as Caesar's quaestor and
tribune, had delivered before the assembly on the 23rd of
December, an indictment of Pompey from boyhood, with menace
of arms. The letter ends in self-disgust: Cicero owes
Caesar money, and the dowry-apparatus he had earmarked for
his triumph must now go to pay the debt. ``It is an ugly
thing to be debtor to one's political opponent.''}
