% Ad Atticum 5.21 --- headnote
% ad-atticum-05-21, 13 February 50 BC, Laodicea

\textit{Cicero to Atticus, written from Laodicea on the
Ides of February (13 February) 50 BC (Perseus dateline:
\textit{Scr. Laodiceae Id. Febr. a. 704 (50)}). It is
the second winter of the Cilician proconsulship.
Atticus has finished his crossing into Epirus and is
wintering there; Cicero has come down out of the Taurus
to the assize town of Laodicea to begin the spring
circuit of his courts. The letter is a long, settled,
deliberate one --- the kind he writes when he has the
table to himself and his secretary at his elbow --- and
covers in turn the Parthian situation, the worry that
his command will be prorogued, the moral economy of his
administration, the year's judicial itinerary, the
great affair of Brutus and the Salaminian creditors,
and a closing flurry of household business. It is one
of the load-bearing letters of the proconsular
correspondence: the moneylending case in particular
will hover over the Brutus correspondence to come.}

\textit{The first half of the letter is governed by an
old Ciceronian theme dressed in new clothes. He has
governed his province at three-and-a-quarter percent
where his predecessors took whatever the market would
bear; he has refused to billet troops in winter quarters
on communities that would pay to be spared them; he has
forbidden statues, shrines, four-horse chariots and the
rest of the honorific apparatus by which Roman governors
were customarily flattered into avarice. He says this to
Atticus because Atticus is the one man in Rome with
whom he can say it without performance: ``you yourself
wished me to do these things.'' And he says it because
the famine that has fallen on Asia made the test sharp:
he extracted grain not by force or judgement but by
authority and exhortation. The detail about L. Tullius
--- a lieutenant who broke the Julian law in transit
``only once a day,'' not, as others did, in every
village --- is meant as comedy, not as confession.}

\textit{Then comes the great business with which the
letter is most identified. Brutus, through his agents
M. Scaptius and P. Matinius, had advanced money to the
community of Salamis in Cyprus at forty-eight percent
compound interest, four percent per month. This was
flagrantly above the twelve percent ceiling of the
\textit{lex Gabinia} on provincial loans; the loan had
been protected only by two senatorial decrees of 56 BC,
which Brutus's connections had procured and which
exempted this one bond from the cap. Appius Claudius,
Cicero's predecessor, had given Scaptius a prefecture
and a troop of cavalry with which to bully the
Salaminians into paying. Cicero finds, on arrival, that
his own standing edict caps interest at twelve percent
with annual compounding; that the Salaminians can in
fact pay; that they offer to deposit the money in a
temple if Scaptius will not take it at the legal rate;
and that Scaptius is fighting on, demanding the higher
rate by force of the senatorial decree. Cicero withdraws
the cavalry, refuses Scaptius a new prefecture (``I give
one to no man in business''), and freezes the case at
his edict's twelve percent. The episode is the great
hinge of his self-image as governor: Brutus's
friendship was at stake, and Cicero would not bend the
rule for him. The reader of the Brutus correspondence
later --- and of the letters that record Atticus's own
embarrassment at having to mediate --- should keep this
scene in view.}

\textit{The closing flicker is the Tullia question. The
\textit{[Greek: secret-of-the-house matter]} is the
choice of a third husband for his daughter; Cicero is
canvassing Atticus's preference between the son of
Postumia (Servius Sulpicius's wife) and an arrangement
proposed through Pontidia. (The choice will in the end
fall on neither --- Tullia and her mother will settle
on Dolabella in his absence, to his complicated
later feelings.) Then a note about the Eleusinian
Mysteries, and the calendar question --- whether Rome
has inserted an intercalary month --- on which the date
of the Mysteries, and therefore Atticus's movements,
will turn. The letter trails off, as the long
deliberate letters often do, into the small fabric of
practical concerns through which the great matters had
to be carried.}
