% Ad Familiares 8.14 --- headnote
% ad-familiares-08-14, around 24 September 50 BC, Rome

\textit{M. Caelius Rufus to Cicero in Cilicia, written from
Rome around the eighth day before the Kalends of October
(roughly 24 September) 50 BC (manuscript dateline
\textit{Scr. Romae circ. viii K. Oct. a. 704 (50)}).
The longest of the four newsletters Caelius sends in 50 BC,
and the most consequential. The opening paragraph reports
the augural election: the seat had been the prize, M. Antony
won it backed by Curio, L. Domitius Ahenobarbus was crushingly
defeated, and the city took partisan pleasure in his pain.
Caelius, who had backed Antony, is now Domitius's bitterest
enemy --- the same Domitius who in section 1 of \textit{Fam.}
8.12 was already plotting with Appius against Caelius's
aedileship. The political fault lines for the coming war are
now drawn through the smaller magistracies and priesthoods.}

\textit{The middle of the letter is the famous analysis: he
sees no peace for the year, Pompey and Caesar are no longer
in private rivalry but on the edge of open war over the
single question of who lays down his army first, and Caelius
cannot find what counsel to take for his own conduct. Section
3 contains the political-realism formula that the letter is
remembered for: in a domestic quarrel, while the conflict is
civil and unarmed, one should follow the more honourable
side; once it has come to war and the camp, the stronger ---
and judge the better course the safer one. He sees the
Senate and the courts staying with Pompey, and all the
discontented and the desperate going over to Caesar; if the
forces are measured army to army, there is no comparison.
The closing paragraph turns to Appius's reckless censorship
("the censorship is a face-cream or a piece of soap" --- he
will scrub away his own veins) and ends, half in jest, with
the wish that Cicero come at speed to be present for the
spectacle. The "great and welcome spectacle" Fortune is
preparing is the civil war.}
