% Ad Atticum 7.2 --- headnote
% ad-atticum-07-02, 27 November 50 BC, Brundisium

\textit{Cicero to Atticus, written from Brundisium on the fifth
day before the Kalends of December 50 BC (the manuscript
dateline: \textit{Scr. Brundisi v K. Dec. a. 704 (50)}). This
is the first letter from Italian soil after the homeward voyage
from Cilicia: the crossing from Epirus was easy, the proconsul
is back, and Atticus has been ill at Rome. The book-opening
joke is technical --- a soft south-easterly wind from Onchesmos
whose name happens to fall into a spondaic hexameter, which
Cicero offers Atticus to pass off as his own.}

\textit{The body of the letter braids together everything
Cicero has been carrying since he landed. Atticus's quartan
fevers: Terentia met him at the forum the moment he came
through the gate with a report from Trebula that the second
fever has broken. The bundle of letters waiting at Brundisium,
some in Atticus's hand and some in the hand of Alexis (who
imitates his patron so well Cicero loves the script for the
resemblance). Tiro, left sick at Patrae --- the boy Cicero
calls \emph{probus}, of the soundest character, and to whom
he is about to write the letter that follows this one in the
collection. Manius Curius's will sealed with the signets of
the three Ciceros and the praetorian cohort, with Atticus
named heir of an as and Cicero of a quarter-as. The young
Marcus's irrepressible desire to see the river Thyamis. The
report Atticus had sent of Pompey's conversation at Naples
about the triumph --- the news that warmed Cicero most.}

\textit{The political temper rises sharply in the second half.
The triumph itself: Cicero had not coveted one until Bibulus's
preposterous dispatches were rewarded with a supplicatio of the
most lavish kind, and the inequity is now intolerable --- the
disgrace, he tells Atticus, is \emph{nostrum}, ours, binding
the two of them together. Then a flash of resentment against
Cato, who had refused him the supplicatio while voting Bibulus
twenty days, and on whose snub Caesar (who is writing him
flattering letters and promising everything) is openly gloating.
And finally the runaway slave Chrysippus, the one he had
liked for the boy's sake --- the offence is the flight, not
the petty thefts. The letter closes with the one news that
keeps the panic at bay: the Parthians have suddenly left
Bibulus half-alive, and the danger that mattered last summer
is, for the moment, off the board.}
