% Ad Familiares 2.17 --- headnote
% ad-familiares-02-17, 17 July 50 BC, Tarsus

\textit{Cicero to his correspondent of the moment ---
the salutation in the manuscript is corrupt
(\textit{†CANINI SALVSTIO PROQ.}), and editors have
generally identified the addressee as C. Caelius
Caldus, then serving as proquaestor with Bibulus in
Syria. The Perseus dateline gives Tarsus,
\textit{a.\ d.\ xvi K. Sext. a. 704 (50)}, that is, the
sixteenth day before the Kalends of August --- 17 July
50 BC --- ``or shortly after,'' written from Tarsus as
Cicero is preparing to depart the province at the end
of his year. The letter is the considered reply to two
of Caelius's: technical, deliberate, methodically
numbered through six topics, with one personal
remonstrance reserved for the close.}

\textit{The first half of the letter is administrative
in temper. Cicero has heard nothing of a successor and
plans to step down on the day he arrived: the Parthian
alarm has lifted, the legions decreed for Syria will
not now march, Marius (his successor designate, if any)
is being held to come up only with the legions. He will
not linger; he will perhaps stop at Rhodes for the sake
of the two boys (his son Marcus and his nephew
Quintus); he wants Rome as soon as possible. Then in
turn: the rendering of accounts under the Julian law
(Cicero advises Caelius to render them, whatever
Bibulus does); the controversy of the Apamea garrison,
which was withdrawn when the Parthian fear was lifting
and which Caelius thinks should have been kept; the
quaestor's accounts and Cicero's careful handling of
public money; the question of three hundred and
thirty-three thousand drachmas, on which Cicero firmly
declines to oblige Caelius; the legions; and Marius. It
is the cool prose of a man closing his books.}

\textit{The closing section --- the second of Caelius's
two letters answered --- shifts the temperature. Caelius
has asked Cicero to commend him as warmly as possible
to Bibulus, his praetor. Cicero will do so, but he uses
the occasion for a measured complaint: Caelius alone,
of all those around Bibulus, has never warned Cicero of
how venomously Bibulus speaks of him. Cicero rehearses
Bibulus's pettiness: refusing to write to Cicero about
the Parthian war though the danger touched him; writing
only about his son's augurate; in his dispatches to the
Senate, claiming Cicero's measures as his own and
shrugging off shared ones as Cicero's; insulting
Ariobarzanes (whom the Senate, through Cicero, had
named king) by addressing him in the despatches only as
``son of King Ariobarzanes.'' The letter to Bibulus
will be written; Caelius can do with it what he likes.
The acidity of these lines toward Bibulus --- earlier a
political rival, now a fellow proconsul whose
incompetence has shadowed the eastern command --- is
the kind of private accounting Cicero generally
reserves for Atticus; here he lets a younger man see
it, in proportion to a younger man's request.}
