% Ad Familiares 15.5 --- headnote
% ad-familiares-15-05, late April or early May 50 BC, Rome

\textit{Marcus Porcius Cato to Cicero at Tarsus, written
from Rome in the last days of April or the first days of
May 50 BC (the manuscript dateline: \textit{Scr. Romae
vel ex. m. Apr. vel in. Mai. a. 704 (50)}). This is
Cato's answer to Cicero's request, sent from Cilicia in
Fam. 15.4, that Cato support a public thanksgiving for
the Pindenissus campaign. Cicero had asked for two
things at once: the \textit{supplicatio} itself, the
formal religious vote which (he hoped) would be the
foretaste of a triumph, and a personal moral
endorsement of the way he had governed his province.
Cato gives him the second, in extravagant terms, and
withholds the first.}

\textit{The architecture is the meaning. Section 1
declares Cato's own vote --- a citation for integrity,
defended province, preserved client-king, recovered
allied goodwill --- in language pitched as high as a
Roman public testimonial can be pitched. Section 2 turns
on a pair of conditional clauses, doubled in perfect
balance: \emph{if} you prefer that thanks be rendered to
the gods rather than credited to your account, I am
glad; \emph{but if} you take the thanksgiving as the
foretaste of a triumph and prefer that fortune be
praised rather than yourself, then --- and Cato then
springs the trap --- a thanksgiving does not always
issue in a triumph, and what the Senate has already
given you (a moral judgement that a province was held
by mildness rather than by force) is a more illustrious
thing than a triumph could be. The implication, courteous
on its surface and pitiless underneath, is that to
prefer the supplicatio to Cato's testimonium is to
prefer the lesser honour. Section 3 closes with the
admission, made as if in apology for unusual length,
that Cato has written this carefully precisely so that
Cicero will reckon him to have wished what he himself
judged most ample --- and to rejoice, all the same, in
what Cicero preferred. The Senate would in fact vote
the supplicatio; Cato's letter, with the elegance of
its refusal, is the one Cicero kept.}
