% Ad Familiares 3.9 --- headnote
% ad-familiares-03-09, c. 20 February 50 BC, Laodicea

\textit{Cicero to Appius Claudius Pulcher, written from
Laodicea a little after the tenth day before the
Kalends of March 50 BC (Perseus dateline: \textit{Scr.
Laudiceae paulo post x K. Mart. a. 704 (50)}). Appius
is now back at Rome and under prosecution by P.
Cornelius Dolabella. He has just sent Cicero, by way of
the freedman Philotimus, a long and warmly written
letter; in it he promises Cicero his support, hopes for
the supplicatio that will lead to Cicero's triumph, and
asks Cicero in turn to back him in the prosecution.
The Dolabella-Tullia match has only just been made
(news of the betrothal reaches Cicero from Rome at
about this date), and the political tangle of son-in-law
prosecuting Appius hangs unspoken over the page.}

\textit{Cicero's reply opens with the courtly admission
that the earlier letters from the road (about the
embassies of thanks and the Apamean building works)
had been ill-tempered enough to draw a sharp answer ---
his \textit{Fam.} 3.7 of the week before --- but that
this present letter from Rome is finally ``worthy of
Appius Claudius.'' The body of the letter is the
counter-offer: Cicero will not be drawn on the
Dolabella prosecution by name, but declares his alliance
warmly and at length. He celebrates the now-secure
prospect of Appius's own triumph; presses Appius, in
turn, to take in hand the decree of the supplicatio
for Cicero's victory on Mount Amanus (sent late, with
explanation, after the Pomptinus letters had cleared
the season); promises to repay in kind the gift of
augural learning Appius has begun to send him; and ends
with the formula of total commitment --- ``you will
hold me, with all that is mine and all my people, in
your keeping.'' The Appius file has turned: from the
expostulation of \textit{Fam.} 3.8, through the
patching-up of \textit{Fam.} 3.7, to this declared
alliance.}
