% Ad Atticum 6.6 --- headnote
% ad-atticum-06-06, about the fourth day before the Ides of Sextilis (~10 August) 50 BC, Rhodes

\textit{Cicero to Atticus, written at Rhodes about the
fourth day before the Ides of Sextilis (10 August) 50
BC --- the manuscript dateline: \textit{Scr. Rhodi
circ. iv Id. Sext. a. 704 (50)}. The province is
behind him; he is on the long way home, the boys with
him, the etesian winds making the next leg slow. The
piece of news that opens the letter is that he has, by
arrangement made at Rome in his absence, become
father-in-law to Cornelius Dolabella --- the man who
not long ago prosecuted his predecessor Appius
Claudius, whom Cicero is busy ``dressing up in every
kind of distinction'' from the province. Cicero had
been negotiating about Tiberius Nero; the women at
home settled it the other way before he could
intervene. He hopes the marriage will be the better
one.}

\textit{The rest of the letter moves through the
business he has not been able to manage from the
field. Section 2: the Athenian grain-dole and his
unease about Atticus's part in it; the commission for
a propylon at the Academy, set against Appius's
abandoned Eleusinion; the loss of Hortensius (whose
death he has only just heard of and feels as a
personal injury, since he had meant at last to live
with him in close intimacy). Section 3 is the
defence of the succession he has had to improvise:
not Quintus filius (a boy), not Quintus frater
(the cost too high, the popular case too damaging),
but Coelius --- because the choice as in fact
constrained, and because Pompey leaving Quintus
Cassius and Caesar leaving Antonius without lot are
the precedents he can stand behind. Section 4 is
the private case for the same choice: the worry of
ever leaving family there, the worry above all of
Quintus filius, and at the close the triumph he has
begun to want --- the \textit{palingenesia} of a
reputation made new in Cilicia, which he tells Atticus
to want with him.}
