% Ad Atticum 7.4 --- headnote
% ad-atticum-07-04, 10 or 11 December 50 BC, Pompeian villa

\textit{Cicero to Atticus, written from the Pompeian villa on
the fourth or third day before the Ides of December 50 BC
(the manuscript dateline: \textit{Scr. in Pompeiano iv aut iii
Id. Dec. a. 704 (50)}). One or two days after the long Trebula
letter, with Cicero now further up the coast at his own villa
near Pompeii. The opening discharges a small piece of
household business: he is sending Dionysius back to Atticus
(``burning with longing for you'') with a careful character
reference that ends in the half-deprecating, half-emphatic
\emph{plane virum bonum} --- plainly a good man, even though
he is a freedman.}

\textit{The substance is the report of an interview with
Pompey himself: two hours together on the fourth day before
the Ides of December. On the triumph, Pompey was warm and
helpful, with practical advice not to enter the Senate before
the matter was settled. On the commonwealth, he was already
talking as though the war were not in doubt and concord
beyond reach --- the proof he offered was a small but
diagnostic incident: Hirtius, Caesar's intimate, had come
into Rome on the evening of the eighth, an arrangement had
been made for him to meet Scipio before dawn on the seventh,
and he had instead slipped away in the dead of night to
Caesar. Pompey called this \emph{tekmeriōdes}, evidentiary,
of the breach. The letter closes with the same forced
consolation Cicero will repeat for weeks --- that the man on
whom even his enemies have bestowed a second consulship will
surely not be \emph{tam amens}, so deranged, as to push it
to the extremity --- and the bald aside that he fears many
things he does not dare put in writing. His own plan: to be
at the city by the third day before the Nones of January.}
