% Ad Familiares 16.9 --- headnote
% ad-familiares-16-09, 28 November 50 BC, Brundisium

\textit{Cicero (with his son Marcus, jointly named in the
salutation) to Tiro, written from Brundisium on the fourth day
before the Kalends of December 50 BC (the manuscript dateline:
\textit{Scr. Brundisi iv K. Dec. a. 704 (50)}). The day after
the Atticus letter from the same harbour, with Tiro still left
behind sick at Patrae. The opening is a day-by-day log of the
crossing --- Patrae to Leucas, Leucas to Actium, the storm,
Corcyra, Cassiope, the run across to Hydrus on a gentle south
wind, the fourth-hour arrival at Brundisium and Terentia
walking into the town beside them.}

\textit{The second half is what the log is in service of. A
letter from Tiro dated the Ides of November has at last
reached him by Plancius's slave, with Asclapo the doctor's
plain assurance of recovery. The fatherly anxiety modulates
into instruction: don't push it; you were too modest at
Lyson's concert and risked a fourth-week relapse; Curius has
orders to see the doctor paid; a horse and mule are waiting
for you at Brundisium when you come. And the closing
imperative is the one that recurs in every letter of this
series --- do not sail rashly. Travel with Mescinius if you
can, with someone of standing if you cannot, and present
yourself to us safe and sound. The intimacy of address
(\emph{mi Tiro}), the worry, and the practical disposition
all hold steady from the Cilician correspondence into the
new chapter that is about to open.}
