% Ad Familiares 7.32 --- headnote
% ad-familiares-07-32, after 11 February 50 BC, Laodicea

\textit{Cicero to P. Volumnius Eutrapelus, written from Laodicea
after the third day before the Ides of February (after 11 February)
50 BC, deep into the Cilician proconsulship (the manuscript dateline:
\textit{Scr. Laudiceae post iii Id. Febr. a. 704 (50)}). Volumnius
carries the cognomen Eutrapelus --- ``the witty,'' from the Greek
\textit{eutrapelia} --- and was the city's connoisseur of urbane
talk; Cicero in this letter is asked to defend his own copyright over
that domain.}

\textit{The opening conceit is a small game. Volumnius has sent his
letter without a forename, in friendly informality; Cicero affects
to wonder whether it might not be from the senator Volumnius until
the wit of the prose reveals the writer's hand. From there the joke
turns substantive: while Cicero has been out of Rome, every clever
saying in the city --- even Sestius's --- has been getting fathered
on him. He casts Volumnius as his ``agent'' in possession of the
``saltworks'' (\textit{possessio salinarum}), and asks him to put up
a real defence: unless an ambiguity is keen, a hyperbole elegant, a
wordplay neat, a turn against expectation amusing --- in short,
unless the techniques he had set out through Antonius in the second
book of the \textit{De Oratore} are visible --- let Volumnius swear
under oath that the joke is not his. The Greek terminology is
deliberate: the technical vocabulary of comic theory recurs at every
beat (\textit{akythēron}, \textit{amphibolia}, \textit{hyperbolē},
\textit{paragramma}, \textit{para prosdokian}, \textit{entechna}),
the language of the schools deployed in mock-legal idiom.}

\textit{The note about ``our friend'' in his tribunate is for Curio,
whose year Cicero is following with hope; the closing request for
Volumnius to encourage Dolabella is the early note of Cicero's new
son-in-law's place in the household. The whole short letter is a
specimen of the Ciceronian register the addressee was reputed for ---
playful, technical, and unmistakably his own.}
