% Ad Atticum 7.9 --- headnote
% ad-atticum-07-09, 28/29 December 50 BC, Formiae

\textit{Cicero to Atticus, written from his Formian villa on
the fifth or fourth day before the Kalends of January, i.e.
28 or 29 December 50 BC (Perseus dateline:
\textit{Scr. in Formiano v aut iv K. Ian. a. 704 (50)}). The
day or two after the Pompey conference of 7.8, and the most
analytical of the December letters: an attempt to lay out
exhaustively the branching possibilities of the next several
weeks, and to ask which of these evils is the least.}

\textit{Section 1 opens on the bantering reproach: a daily
letter? if I have a courier, daily; and besides, in a few
days you will be standing here in person. There is also a
chilling small datum --- a letter from Atticus has gone
missing, because the courier Lucius Quinctius was attacked
and stripped near the tomb of Basilus on the Appian Way. The
road from Formiae to Rome is no longer safe.}

\textit{Section 2 is the \emph{politikon problema}: a
catalogue, in Cicero's most formal logical manner, of the
constitutional options as of the last days of December.
Either Caesar's candidacy in absence is entertained by
Senate or tribunes while he holds his army; or he is brought
to surrender province and army and stand for consul properly;
or he refuses and the elections proceed without him; or his
tribunes block them and matters fall to an interregnum; or
he marches on Rome --- and within that last option, all the
further branches: when he begins, on what pretext, whether
the City is to be held or abandoned. The ``best'' option ---
that Caesar agree to lay down his army and stand for consul
--- Pompey (per 7.8) has just told Cicero is not in fact in
prospect. So the analysis falls to: which is the least
unbearable of the remaining? Section 4 closes on the harshest
acknowledgement: even conceding everything would mean
``See again as your consul the man you have seen as consul
before,'' i.e. a Caesarian second consulship with Pompey
exiled to Spain. The famous final line: \textit{equidem
dies noctesque torqueor}, ``As for me, I am tortured day
and night.''}
