% Ad Familiares 8.7 --- headnote
% ad-familiares-08-07, late April or early May 50 BC, Rome

\textit{M. Caelius Rufus to Cicero, written from Rome at the
end of April or beginning of May 50 BC (Perseus: \textit{Scr.
Romae ex. m. Apr. aut in. Mai. a. 704 (50)}). Cicero's year
in Cilicia is almost over; he is openly counting the days
until he can leave the province, and Caelius is counting them
along with him. The opening confesses an anxiety as
characteristic of Caelius as the gossip that follows it ---
``the more successfully you have so far managed things
there, the more I am tortured by the prospect of a Parthian
war'' --- afraid that some fresh alarm in the east will pin
Cicero down beyond the year and ruin the campaign he is
already counting upon. The letter is short and scribbled in
haste, slipped to a courier of the publicani who was leaving
that minute; Caelius had sent the fuller weekly dispatch via
Cicero's own freedman the day before.}

\textit{All Caelius has time for, then, is the salacious
gleanings of the city. Young Cornificius has betrothed
himself to Orestilla's daughter; Paula Valeria, sister of
Triarius, has divorced her husband for no reason on the very
day he was due home from his province, in order to marry D.
Junius Brutus (the future tyrannicide). Servius Ocella has
been caught \textit{in flagrante} twice in three days --- with,
Caelius hints, a woman whose identity is too good for the
page, and which an enquiring imperator may, if he likes,
extract from each Roman in turn on his return. The letter is
the purest vein of Caelian gossip-letter-writing in the
collection: short, friend-to-friend, with a sting of mock-
prurience and not a word of public business.}
