% Ad Familiares 8.11 --- headnote
% ad-familiares-08-11, end of April / beginning of May 50 BC, Rome

\textit{M. Caelius Rufus to Cicero in Cilicia, written from
Rome at the end of April or the beginning of May 50 BC
(manuscript dateline \textit{Scr. Romae ex. m. Apr. aut in.
Mai. a. 704 (50)}). The letter is given over almost entirely
to the senatorial fight over the public supplications voted
for Cicero's Amanus campaign. The new tribune C. Scribonius
Curio, hitherto an ally of the optimates, has crossed the
floor to Caesar's interest, and his veto-power over
supplicationes had become the bargaining chip in a wider
struggle over Caesar's provincial command. Caelius, brokering
between Curio and the consuls, secures Cicero his decree
only by getting the consuls to undertake not to act on it ---
neither posting the edict nor wearing the laurelled robe ---
this year. Even so the inventory of who voted how, who
spoke honourably, and who obstructed is precisely the kind
of city report Cicero, far away in Tarsus, has been asking
him for.}

\textit{The political news in the second half of the letter
is the bombshell of 50 BC: Curio has switched sides, the
quarrel between Pompey and Caesar has now narrowed to the
single question of when Caesar lays down his army and his
province, and the optimate party --- "our friends, whom you
know well" --- is failing to push the matter to a head.
Caelius's voice is younger and terser than Cicero's, less
oratorical, more straightforwardly political: he reports
without sentiment, hands out his thanks "according to each
man's nature and style," and closes with the verdict that
Caesar will keep his command "as long as he pleases" unless
the Senate finds the nerve to act. The closing paragraph
attends to Cicero's running private business with Sittius
and refers him to the \textit{acta diurna} for the rest.}
