% Ad Atticum 7.7 --- headnote
% ad-atticum-07-07, c. 21 December 50 BC, Formiae

\textit{Cicero to Atticus, written from his Formian villa
about the 21st of December 50 BC. The Perseus dateline as
transmitted reads
\textit{Scr. in Formiano inter xiiii et x K. Iun. a. 704
(50)} --- ``between the 14th and 10th days before the Kalends
of June'' --- which is impossible: in May 50 BC Cicero was in
Cilicia, not at Formiae. The accepted reading, following
Shackleton Bailey, is that \textit{Iun.} here is a copyist's
or OCR error for \textit{Ian.}; the letter belongs to the
December cluster, slotting between 7.6 (19 December) and 7.8
(27/28 December). The text itself fixes Dionysius's arrival
to the 15th day before the Kalends of January, i.e. 18
December, and points forward to the journey toward Rome
already laid out in 7.5. We date the sidecar to 21 December,
a mid-range placeholder.}

\textit{The letter survives with gaps: sections 2 and 5 are
absent from the transmitted text, and a noun in section 1 is
locally corrupt (\textit{†putato†}). The argument runs all
the same. Section 1 picks up the long-running thread of
Atticus's freedman Dionysius, who has now reached Rome and
delivered an Atticus letter; Cicero notes wryly that his
earlier warm testimony forbids him a \emph{palinōidia}, a
recantation. Section 3 dovetails into travel logistics: he
will not arrive at Pompey's Alban villa on the
\textit{Compitalia} (so as not to disrupt the household
festival) and asks Atticus not to exert himself on account
of the quartan fever. Section 4 opens the political core:
rumor has it that Pompey's council mean to send Cicero into
Sicily on the strength of the still-active Cilician imperium
he carries; Cicero retorts that this is \emph{Abdēritikon},
``Abderite logic'' (the Abderites were a Thracian byword for
stupidity), since neither Senate nor people have decreed any
such thing. He then turns to a withering anatomy of who
exactly the ``loyalists'' (\emph{boni}) are supposed to be:
the senators who failed to break Curio? the publicans now
on Caesar's payroll? the financiers? the farmers, who care
only for quiet?}

\textit{Sections 6 and 7 are the famous resolution. Resisting
Caesar now is resisting a man we ourselves armed, by every
prior concession; the moment to have stopped him was when
he was weak, and that moment is past. Behind him stand
eleven legions, the Transpadanes, the urban plebs, ``so
many tribunes of the plebs, a youth so utterly corrupted, a
commander of such authority, of such audacity.'' Either we
fight or we let his candidacy stand under the law; ``fight,
then, rather than be a slave,'' Atticus is imagined as
saying --- but if you lose you are proscribed and if you win
you are still a slave. So Cicero will do as cattle do: he
will follow the herd of his kind. The play of registers ends
on the Senate-floor formula, \emph{dic, M. Tulli}, ``Say your
piece, Marcus Tullius'' --- and the punch-line: ``I assent to
Gnaeus Pompeius, that is, to Titus Pomponius.'' He owes the
loyalist line not to its merits but to his oldest friend.
And then, mid-thought, the lamp fails (\emph{me lucerna
deserit}) and the letter ends with greetings to Atticus's
slave-boy Alexis.}
