% Ad Familiares 13.57 --- headnote
% ad-familiares-13-57, early April 50 BC, Laodicea

\textit{Cicero to Q. Minucius Thermus, propraetor of Asia, written
from Laodicea at the beginning of April 50 BC (the manuscript dateline:
\textit{Scr. Laudiceae in. m. Apr. a. 704 (50)}). The setting is the
late phase of Cicero's proconsular year, with reports of a Parthian
war breaking over Syria and Cicero preparing to take the field along
the Taurus.}

\textit{The letter has two requests, woven together. The first is
military: Cicero needs M. Anneius, his \textit{legatus}, back at
once. Anneius had been on detached leave with Thermus in Asia, settling
a private case at Sardis; Cicero, who is to march for Cilicia about
the Kalends of May, asks for him to be returned by then because, of
all his staff, it is on Anneius's service, judgement, and military
experience that he and the commonwealth chiefly depend. The second is
forensic: that Thermus see Anneius's business with the Sardians through
to a settlement worthy of the truth of his case and his own dignity.
Cicero recalls a conversation with Thermus at Ephesus on this point,
and renews the request twice over (\textit{etiam atque etiam rogo})
in the formula of a recommendation that has been pressed before.}

\textit{The letter is a working document of the Cilician proconsulship
at its most strained: the war news from across the Taurus, the
schedule for striking camp, the recall of the man Cicero most relies
on. Thermus is the one provincial neighbour whom Cicero treats
throughout as collegial; the tone is brisk and uncomplicated.}
