% Ad Familiares 8.12 --- headnote
% ad-familiares-08-12, around 20 September 50 BC, Rome

\textit{M. Caelius Rufus to Cicero in Cilicia, written from
Rome around the twelfth day before the Kalends of October
(roughly 20 September) 50 BC (manuscript dateline
\textit{Scr. Romae circ. xii K. Oct. a. 704 (50)}). Caelius
is now curule aedile; the games of section 3 ("at my games
at the Circus") are the Ludi Romani of mid-September, which
he is presiding over for the first time. The letter is given
over entirely to a single grievance: the campaign of his old
patron, the censor Appius Claudius Pulcher, against him.
Appius --- whom Cicero had succeeded in Cilicia, and whose
daughter Caelius had once been about to marry --- has refused
to pay the kindnesses he owes to Caelius, has tampered with
the augural college, and has put up a stooge prosecutor to
indict him under the Lex Scantinia, a sumptuary statute
covering acts contra naturam. Caelius's reply has been to
serve Appius, in his capacity as censor, under the same law.}

\textit{The voice throughout is Caelius's at his most direct:
short clauses, colloquial scorn ("that ape"), the relish of
the political man who has just landed a public counter-blow
("nothing could have fallen out better"). The closing
paragraph turns from grievance to the slow communications
that have held this letter back forty days, to the looming
consular elections (Domitius, who has thrown in with Appius
and against him, is dreading them), and to a personal
appeal: Caelius asks Cicero to grieve for his injuries as
Caelius has been used to grieve for Cicero's, and to avenge
them. The wider crisis --- Caesar, Pompey, the provinces ---
is not in this letter; Caelius will return to it four days
later in \textit{Ad Familiares} 8.14.}
