% Ad Atticum 6.5 --- headnote
% ad-atticum-06-05, 27 June 50 BC, in camp (Cilicia)

\textit{Cicero to Atticus, written in camp on the
fifth day before the Kalends of Quintilis (27 June) 50
BC --- the manuscript dateline: \textit{Scr. in castris
v K. Quint. a. 704 (50)}. Thirty-three days remain of
the proconsular year. Syria is in flames; Bibulus, in
the personal grief of his sons' murder, is shouldering
the heaviest part of the Parthian war; Cicero, with a
weak army of legionaries and good Galatian, Pisidian,
and Lycian auxiliaries, is keeping his forces as close
to the enemy as the law of the decree allows. The
question that will not let him sleep is who takes
Cilicia after him. The senate's failure to nominate a
successor --- the leitmotif of late \textit{Ad Atticum}
6 --- raises the spectre of a second year in the
province, the thing he has feared from the moment he
came out.}

\textit{The letter is short and uncharacteristically
constrained: he says at the end he wanted to write
more, but had nothing else to say and was too anxious
to jest. Section 1 picks up the encrypted business of
6.4 --- Terentia's freedman and what Cicero suspects of
him --- and asks Atticus to dig further. Section 2
runs that cipher openly: a long Greek-script passage,
half ledger half proverb, on what looks to be the
Scaptius / Brutus money in Salaminian Cyprus (the
debt-figures in minae, the second-month interest, the
recriminations of an agent who has come up to camp,
played the suppliant, and gone away muttering quotations
from Homer and proverb when he failed to get what he
wanted). Section 3 is the procuratorial heart: the
day of decessio creeping up \textit{le-le-thotos}, the
\textit{allo problema} of whom to put in charge if
Caldus does not arrive in time. Section 4 closes with
the apology --- no jesting today --- and the family
greetings.}
