% Ad Atticum 5.6 --- headnote
% ad-atticum-05-06, around 19 May 51 BC, near Tarentum

\textit{Cicero to Atticus, written from Tarentum on the
eighteenth of May 51 BC (the manuscript dateline:
\textit{Scr. Tarenti xiv K. Iun. a. 703}). Cicero is on
his outward journey to the province of Cilicia, of
which he is now proconsul --- a posting he had not
sought, foisted on him by the lottery procedures that
followed the Lex Pompeia de provinciis of 52 BC. He has
reached the heel of Italy and there encountered Pompey,
who happens to be in residence at his estate near
Tarentum; the encounter, which Cicero presents as
incidental, is in fact a small political event ---
Pompey is the senior figure in the state and Cicero is
his man on the eastern road --- and he uses the few
days of waiting for his legate Pomptinus to take soundings
about how the Senate sees his commission.}

\textit{The letter is short, distracted, and businesslike.
Cicero is uncertain whether Atticus is still in Rome or
has already set out for Epirus; he writes anyway, out of
the prudential rule of the late letters that any open
channel must be used. The one piece of substance ---
``leave the business of Caesar's debt settled'' --- is
the loan that Cicero made (or that he allowed his
brother Quintus to take on his behalf) from Caesar
before departure: an obligation that will hang over the
correspondence of the next eighteen months and which
the political earthquake of 49 BC will turn into a
liability of a different order. Here it is still simply
a piece of housekeeping to be cleared before Atticus
sails.}
