% Ad Familiares 15.3 --- headnote
% ad-familiares-15-03, 30 August 51 BC, camp near Iconium

\textit{Cicero to M. Porcius Cato, written from camp
near Iconium on the 30th of August 51 BC --- one day
before he broke camp and pushed on through Cappadocia,
the same day he was writing to Appius (3.6). The
manuscript dateline (\textit{Scr. in castris ad
Iconium iii K. Sept. aut paulo post a. 703}) gives the
date as the 30th of August or shortly after. This is
an early personal dispatch from the Cilician command,
written to a senator whom Cicero wants in his corner
when the time comes for honours to be voted on his
governorship --- Cato's incorruptible reputation makes
his support uniquely valuable. The letter is the first
of several to Cato in this period and the only one of
the surviving group sent in mid-campaign, before the
operations and the legal--administrative reforms that
will fill the autumn.}

\textit{The news is the same news that will be sent
officially to the Senate three weeks later (15.1):
Pacorus has crossed the Euphrates with a great Parthian
host; the Armenian king is said to be moving against
Cappadocia. Cicero explains why he is not writing
\textit{publice} --- Antiochus has already sent the
Senate his own dispatch, and Bibulus, proconsul of
Syria, will have reached his province by now and will
report from there. (Bibulus had not in fact yet
arrived, which is why the official dispatch of 15.1
becomes necessary.) The closing sentences carry the
two notes that will be the leitmotifs of Cicero's
Cilician self-presentation: in so weak a force, the
strategy is to hold the East by Roman mildness and
\textit{continentia} --- decent conduct toward the
provincials --- and by the loyalty those win, more
than by arms; and a personal request to Cato to think
well of the absent governor and to defend him in his
absence.}
