% Ad Familiares 13.1 --- headnote
% ad-familiares-13-01, around late June / early July 51 BC, Athens

\textit{Cicero to C. Memmius, written from Athens in late
June or early July 51 BC (the manuscript dateline:
\textit{Scr. Athenis inter vii K. et prid. Non. Quint. a.
703}). On his outward passage to Cilicia, Cicero broke
his journey at Athens for a few days, lodged with his
friend Aristus, and walked the philosophical sites of
the city. The matter of this letter is one such site:
the ruins of Epicurus's old house and garden in the
Melite quarter, of which the title was held by Memmius
--- the dedicatee of Lucretius's \textit{De Rerum
Natura}, now exiled to Athens after his conviction for
electoral corruption in 52 BC --- and which Memmius
appeared to have meant to demolish and rebuild on. Patro,
the current head of the Epicurean school at Athens, had
appealed first by letter to Cicero in Rome and now in
person at Athens, asking him to dissuade Memmius from
clearing the site; the Areopagus had registered a
formal memorandum (\textit{hypomnematismon} in the
Greek) recognising the school's claim. The letter is
Cicero's gracefully managed intervention.}

\textit{The interest of the piece is the calibration of
its philosophical politeness. Cicero is an Academic
sceptic; he is not an Epicurean and is not pretending
to be; he says so plainly --- ``with Patro the Epicurean
I have everything in common, except that on points of
philosophy I sharply disagree with him.'' But he treats
the Epicurean school as one of the houses of philosophy
to which the proprieties are owed, and Patro's
loyalty-formula (``honour, duty, the law of testaments,
the authority of Epicurus, the entreaty of Phaedrus,
the seat, the dwelling, the footprints of the greatest
men'') is reported with affectionate accuracy, not with
the mockery the disagreement would have allowed. Behind
the letter is the friendship with Atticus: it is
Atticus who, though not an Epicurean himself, cares about
Patro and Phaedrus and has asked Cicero to press this,
and Cicero is doing it as much for Atticus as for the
school. The request is, in the event, the smallest of
favours: Memmius has reportedly already abandoned the
building project, so all Cicero is asking for is the
formal letter to Athens releasing the site. The letter
survives among the books of testimonials and
recommendations gathered as \textit{Ad Familiares} 13.}
