% Ad Familiares 8.2 --- headnote
% ad-familiares-08-02, June 51 BC, Rome

\textit{M. Caelius Rufus to Cicero, written from Rome in June
51 BC (Perseus: \textit{Scr. Romae m. Iun. a. 703 (51)}). The
sensational headline is the acquittal of M. Valerius Messalla
Rufus, consul of 53, on a charge of electoral bribery
(\textit{ambitus}) --- a verdict so plainly bought, and so
plainly out of step with public feeling, that Caelius is left
literally speechless when it is announced. The jurors are
shouted down on the spot; Hortensius, who had led the
defence, walks into Curio's wooden games-theatre the very
next day and is hissed off the benches in a din the like of
which his unmarred old age had never seen. Messalla, Caelius
notes, still has a second indictment hanging over him under
the Lex Licinia de sodaliciis (a Pompeian law against
electioneering combines), and is now reckoned more likely to
fall on the second prosecution than he was on the first.}

\textit{The political shorthand of the second paragraph
gathers up everything Cicero needs to track from his
provincial post in Cilicia. The consul Marcellus has
quieted down the offensive against Caesar's Gallic command;
the consular elections are entirely up in the air; Caelius
himself is now standing for an aedileship (he uses
\textit{designatum} of his expected return) against M.
Octavius and the obnoxious C. Hirrus, the cousin of Pompey
whose name Cicero loathed. ``On account of Hirrus,'' as
Caelius puts it dryly, Cicero will be especially attentive
to the result. The closing requests show the friendship at
work: the panthers Cicero is to capture and ship for
Caelius's aedilician games at Rome, and a bond on the
freedman Sittius's account, which Cicero is to honour. The
\textit{commentarius rerum urbanarum} --- the rolling Roman
newsletter that Caelius engaged a professional to compile,
described in \textit{Fam.} 8.1 --- is now in regular instalments,
the first sent by the courier Castrinius Paetus, the second
by this letter's bearer.}
