% Ad Familiares 8.4 --- headnote
% ad-familiares-08-04, 1 August 51 BC, Rome

\textit{M. Caelius Rufus to Cicero, written from Rome on
the Kalends of August 51 BC (the manuscript dateline:
\textit{Scr. Romae K. Sext. a. 703 (51)}). The third
Caelius newsletter, and the first that really earns the
description Cicero set it: a future-leaning political
bulletin from the sharpest eye in town. The opening
catalogue is a Roman summer in five clauses --- Messala
acquitted (of ambitus, then condemned by a second
court on a parallel charge), the conservative C.
Claudius Marcellus elected consul for 50, M. Calidius
indicted, P. Dolabella made a quindecimvir
\textit{sacris faciundis}; and as the centrepiece,
the loss of Lentulus Crus's bid for the same priestly
college, which Caelius savours with undisguised
schadenfreude. The third paragraph reaches the news
this volume was made for: the entry of the younger C.
Scribonius Curio --- son of the elder Curio whose death
was the heart of \textit{Fam} 8.2 --- as a candidate
for the vacant tribunate, and the early signs that he
is leaning towards the Senate rather than Caesar.
Caelius's diagnosis (Caesar has miscalculated by
treating Curio with contempt; the elegance of the
unintended manoeuvre amuses everyone) will be embarrassed
by the events of the next year, when Curio's tribunate
becomes one of Caesar's principal instruments.}

\textit{The political heart of the letter is the fourth
section: the senatorial debate of 22 July on Pompey's
pay, which forced Pompey to declare publicly that he
would withdraw from Gaul the legion he had lent to
Caesar, and brought the question of the succession to
Caesar's provinces openly into the chamber for the
first time. The Senate has tabled the substantive vote
until Pompey's return, and Caelius is bracing for the
session of the Ides of August. The Pompey \textit{vox}
he records --- ``all must obey the Senate's word'' ---
will read very differently nine months on. The closing
paragraph drops back into the domestic register: the
Sittian bond (a debt Caelius is trying to recover from
the African operator P. Sittius, on which he wants
Cicero's pressure as governor); the famous request for
Cibyran panthers for his aedilician games; and a query
about the political consequences of the death of
Ptolemy XII Auletes of Egypt, news of which is just
reaching Rome.}
