% Ad Familiares 8.10 --- headnote
% ad-familiares-08-10, 18 November 51 BC, Rome

\textit{M. Caelius Rufus to Cicero, written from Rome
on the fourteenth day before the Kalends of December
(18 November) 51 BC (the manuscript dateline:
\textit{Scr. Romae x iiii K. Dec. a. 703 (51)}; section 3
gives the same date in the body). The first paragraph
records the impact at Rome of the dispatches from
C. Cassius and from king Deiotarus that the Parthians
have crossed the Euphrates and are pouring into the
Roman East --- the very news whose despatch from
Cicero's own camp at Cybistra in late September is
preserved in \textit{Ad Atticum} 5.18. Caelius's
private alarm is not for Cicero's life --- the army
is too small to fight a real battle, so retreat is
expected --- but for his reputation, since a Roman
proconsul who retreats has to defend the retreat in
the city.}

\textit{The middle paragraphs report the political
indecision at Rome that has greeted the news. Names
are being floated for an extraordinary command ---
Pompey, Caesar, the consuls --- but no private citizen
will be sent by decree. The consuls themselves do not
wish to be sent and so refuse to hold the Senate at all,
hiding their reluctance under a reputation for civic
self-restraint. The standing private suspicion appears:
without confirmation from Cicero, the man-on-the-spot,
the Roman political class is half-disposed to believe
that Cassius has manufactured the whole crisis. Caelius
urges Cicero to write to the Senate with careful and
cautious accuracy. The closing paragraph turns again
to Cicero's own succession --- Caelius cannot promise a
successor by the Kalends, but does undertake to prevent
a further prorogation, the one commission he had been
given on Cicero's departure (\textit{Fam} 2.8) and
which he has restated in nearly every letter since.
The Curio of section 4 is now plainly looked to as the
hinge of the new year's politics.}
