% Ad Familiares 15.4 --- headnote
% ad-familiares-15-04, around 21 May 51 BC, Tarsus / Laodicea

\textit{Cicero to M. Porcius Cato, written from his
proconsular province late in 51 BC (the manuscript
dateline: \textit{Scr. Tarsi vel ex. a. 703 (51) vel in.
704 (50)}). The letter is the formal proconsular
dispatch in which Cicero reports the military campaign
of his single year of office in Cilicia: the season's
movements, the false alarm of a Parthian invasion of
Syria, the relief of Ariobarzanes of Cappadocia, the
operation on Mount Amanus, the fifty-seven-day siege of
the hill-town of Pindenissus, and the eventual quartering
of the army for the winter. It is, in its first half,
the most extended piece of military narrative in the
correspondence: a Roman public-record register, with
dates by the Roman calendar, named legates, and the
running list of pacified places. Cicero is not
campaigning in any first-rate sense --- Pindenissus is a
hill-fort, not Pharsalus --- but he is on a real
military footing, the Parthian danger is real, and the
operations earned his men's salutation of him as
\textit{imperator}.}

\textit{The whole apparatus of narration, however, is
preliminary to the second half of the letter, which is
the political point: Cicero is asking Cato for his vote
in the forthcoming senatorial motions on a public
thanksgiving (\textit{supplicatio}) and the implicit
later motion for a triumph. The request is made with
elaborate, slightly anxious care --- Cato is famously
parsimonious with such honours, has refused them to
others Cicero does not name, and is the one senator
whose endorsement would settle the matter. Cicero
therefore lays out three arguments in turn: that he has
governed with equity and self-restraint (Cato's own
test), that the military achievement, modest in absolute
terms, is comparable to operations for which others have
received the highest honours, and --- the third
argument, made with characteristic candour --- that he
who once let a triumph pass after his consulship now
wants one, to heal the wound of his exile in 58 BC and
the public recognition it took from him.}

\textit{The closing move, philosophy summoned in to plead
for him, has often been read as ornament; it is not. It
is Cicero's most precise statement of where he and Cato
stood together in the politics of the late 50s: as the
two senators who tried to make the inherited Greek
philosophical tradition do real work in the running of
the commonwealth. ``Almost alone of men, we took that
true and ancient philosophy, which to some appears a
thing of leisure and idleness, down into the forum and
the commonwealth and almost into the line of battle
itself.'' Cato did, in the event, vote against the
supplicatio; the historical reading is that he did so
on principle, not from any cooling of the friendship,
and Cicero's surviving reply (\textit{Fam.} 15.6) takes
exactly that view. The triumph itself never came: the
Civil War overtook the question.}
