% Ad Familiares 8.3 --- headnote
% ad-familiares-08-03, around 10 June 51 BC, Rome

\textit{M. Caelius Rufus to Cicero, written from Rome
about the 10th of June 51 BC (the manuscript dateline:
\textit{Scr. Romae circ. iv Id. Iun. a. 703 (51)}). The
second of the Caelius newsletters: a fortnight or so
after \textit{Fam} 8.1, with Cicero now well on the road
east, Caelius keeps the promise of regular dispatch and
turns from the formal cover-letter idiom of the first
into something looser and more characteristic --- a
short, gossipy run of paragraphs in his own voice, with
the hired bulletin-compiler's roll presumably travelling
separately. The chief topic is the contest for the
augurate, vacant since the death of Hortensius the
previous year. Caelius is standing for it himself
against Sextus Hirrus (C. Lucilius Hirrus, a Pompeian
loyalist of cordially disliked personal qualities);
M. Octavius is the other declared candidate. Cicero ---
absent on his way to Cilicia --- is being held in
reserve as the wittiest possible patron of Caelius's
campaign; his correspondent's pleasure at the prospect
of beating Hirrus, who had also been Cicero's
unsuccessful rival for the augurate, is undisguised.}

\textit{The remaining business is editorial. The middle
paragraph is the briefest possible reassurance that
Cicero's freedman Philotimus, charged with managing the
property of the exiled Milo, is performing well and
not damaging Cicero's good name --- a touchy file, since
Philotimus's competence was already in dispute. The
closing paragraph is the most attractive moment in the
letter: Caelius asks Cicero, in the otium of his
provincial command, to write him some small treatise
that will preserve their friendship in writing alongside
\textit{De Re Publica} and the rest of the monuments.
The request is sincere and unfulfilled --- no surviving
work answers to it --- but it confirms that Caelius,
even at his most flippant, thinks of his old patron
first as a producer of books for posterity.}
