% Ad Familiares 15.1 --- headnote
% ad-familiares-15-01, 19 September 51 BC, on the Lycaonia-Cappadocia border

\textit{Cicero, proconsul of Cilicia, to the consuls,
praetors, tribunes of the plebs, and senate of Rome.
The dispatch was written in the field on or about the
19th of September 51 BC, on the borders of Lycaonia and
Cappadocia as Cicero's army marched toward the Taurus
(the manuscript dateline: \textit{Scr. in Cilicia xiii
K. Oct. a. 703}). It is one of the very few surviving
specimens of Cicero's official prose as a Roman
magistrate addressing the Senate by letter --- the
\textit{litterae publice missae} of a serving
proconsul. The register is distinct from everything
else in the correspondence: the formulaic
health-greeting \textit{S. v. v. h. e. e. q. v.}
(``if you are in good health it is well; I myself am
in good health'') opens the letter; the periods are
longer and more architected; the personal voice is
disciplined into the impersonal posture of the
magistrate reporting.}

\textit{The news is grave. Pacorus, son of Orodes the
Parthian king, has crossed the Euphrates with a great
Parthian cavalry force and a large mixed levy and has
pitched camp at Tyba; Syria is in alarm; the Parthian
campaign that the consular M. Crassus had opened with
his disaster at Carrhae two years earlier is now coming
out the other way, against a Roman East that has not
been re-equipped. Cicero has the news from three
independent sources --- King Antiochus of Commagene's
envoys, Tarcondimotus the trans-Tauran prince, and
Iamblichus the Arab phylarch --- and reports each in
turn. He has marched his army toward the Taurus rather
than withdrawing, both to put down the Cilician tribes
still in arms and to show the enemy in Syria that the
Roman force is advancing. The body of the dispatch
(sections 4--5) is a structured appeal: the provinces
must be reinforced, the local levy is useless, the
allies are either too feeble or too alienated by the
abuses of Roman rule (\textit{propter acerbitatem
atque iniurias imperi nostri}) to be relied on; M.
Bibulus, the proconsul of Syria with shared
responsibility for the war, refused a levy in Asia
when he had the authority for one, which Cicero offers
as a judgement on the quality of provincial recruits.
Section 6 closes on the kings of the East --- Deiotarus
loyal, Cappadocia empty, the rest neither in resources
nor in will sufficient --- and on a closing line that
will become famous in Roman Stoic memory:
\textit{utinam saluti nostrae consulere possimus!
dignitati certe consulemus}. The plain shape of the
Latin is preserved: he hopes for survival, he is
certain of honour. Bibulus, in the event, would not
reach Syria for another month.}
