% Ad Atticum 5.18 --- headnote
% ad-atticum-05-18, 21 September 51 BC, written in camp at Cybistra in Cappadocia

\textit{Cicero to Atticus, written from the army camp at Cybistra
in Cappadocia, beneath Mount Taurus, on the eleventh day before
the Kalends of October (21 September) 51 BC. Of all the letters
of Cicero's governorship this is the one in which the orator
who had hoped to leave Cilicia in nine months as ``a second
Scaevola'' becomes, for the first time, a frontline general.
The Parthian invasion which the earlier letters had reported as
rumour is now fact: Pacorus, son of King Orodes, has crossed
the Euphrates with nearly all his forces; Cassius, the
quaestor-in-charge after Crassus's death, is shut up in Antioch
with the army of Syria; Bibulus, who was meant to be the new
governor of Syria, has not yet so much as reached his province;
and the enemy is now in Cyrrhestice, the strip of Syria
immediately adjacent to Cicero's own province. He has written
to the Senate on the crisis and asks Atticus, if he is at
Rome, to read the despatch and judge whether to have it
delivered.}

\textit{The substance of the letter is double. To the Senate
(through Atticus) the appeal is for relief: not, above all,
that any extension of his term be added to him \textit{``between
the slaying of the victim and the laying-out on the altar''} ---
the proverb taken from a Roman sacrifice and applied with grim
exactness to a year that was meant to end in January. He wants
Pompey sent out, or some other man by the spring, but no
prorogation. His clearest hope is the winter, which would close
the passes before the Parthians could cross over. To Atticus
himself the report is steadier than one would expect: the army
is small but united, the position safe and well-supplied with
grain, the allies extraordinarily loyal in response to his
mildness and abstinence, a Roman levy is under way, the grain
is being moved into strongholds, and Deiotarus is on the way
with reinforcements that will nearly double the column. ``We
hold firm in spirit, and, because, as it seems to us, we are
taking good counsel, we have hopes also of our right arm.'' The
closing paragraphs --- on the safety of the two boys with
Deiotarus, on the postal arrangements through Epirus, on the
wardship of young Brutus, which Cicero is now reduced to
exhibiting rather than defending --- show how the private fabric
of his life is being held together in the same hand that is
preparing a battle.}
