% Ad Familiares 7.2 --- headnote
% ad-familiares-07-02, ca. 15 February 51 BC, Rome

\textit{Cicero to M. Marius, written at Rome around the Ides of
February 51 BC (Perseus: \textit{Romae circ. id. m. Febr. 703}).
Marius --- the cultivated invalid friend of Cicero's
Campanian neighbourhood, addressee of the great theatrical-
review letter \textit{Fam.} 7.1 of the year before --- has sent
two requests across the bay: a commission to handle a piece of
inherited property at Rome, and a half-apologetic note of
congratulation on the recent conviction of T. Munatius Plancus
Bursa. The two halves of the letter are pitched in different
registers but in the same intimate voice. The opening is light,
mock-mercantile teasing: Marius has, as it happens, picked for
the job the one man with an interest in the property going for
as much as possible (Cicero is himself a co-heir), but his
ceiling-price will protect him, and Cicero, foiled in his
mock-conspiracy, will plant a fictitious bidder rather than let
the thing go for less.}

\textit{The second movement is the real news. Bursa --- one of
the tribunes of 52 who had inflamed the mob against Cicero's
client Milo after the killing of Clodius --- had just been
convicted under the lex Pompeia on public violence, with Cicero
prosecuting and Pompey, the very man whose law it was, exerting
himself for the defence. The verdict was a personal vindication
on a scale rare in Cicero's later career, and his pleasure in
it is open. He distinguishes it from any merely vengeful
satisfaction: he prefers a judgment to a sword, a friend's
glory to an enemy's ruin, and he was struck above all by the
ranks of good men who stood with him against the incredible
exertion of Pompey. And then, with one of his hard private
confessions, he says he hated Bursa worse than Clodius himself
--- Clodius at least had a public motive and acted with men who
needed Cicero down; this one was a \textit{simiolus}, a ``little
ape,'' attacking him for his own amusement at the prompting of
Cicero's lesser enviers. The close is the quiet preoccupation
that will shape all his correspondence of these months: he is
about to leave for Cilicia under the \textit{lex Pompeia} of 52,
and prays daily that no intercalary month be inserted, so that
he may see Marius again before the long absence begins.}
