% Ad Familiares 2.10 --- headnote
% ad-familiares-02-10, 15 November 51 BC, camp at Pindenissus

\textit{Cicero to M. Caelius Rufus, written from the
army camp at Pindenissus in the Cilician highlands on
the seventeenth day before the Kalends of December
(15 November) 51 BC (the manuscript dateline:
\textit{Scr. in castris ad Pindenissum a. d. xvii K.
Decembr. a. 703 (51)}). Cicero is now in the twenty-
fifth day of the siege of the mountain town from which
\textit{Fam} 2.9 was already written; the news of
Caelius's election as aedile, which had reached him
just before the previous letter, has been followed by
silence from Rome, and he opens this one complaining
that no letter of Caelius has come through since the
elections themselves --- and worrying that perhaps his
own letters are similarly failing to arrive. The dry
self-mockery of \textit{balbus enim sum} (``stammerer
that I am'') is a sideways gesture at the
Hirrus-pastiche of \textit{Fam} 2.9; the joke is
allowed one more pass and then dropped: ``we are not
stammerers --- let us return to the matter.''}

\textit{The body of the letter is Cicero's first
narrative summary of the Cilician campaign to a
correspondent who is neither Atticus nor a military
colleague. It compresses the work of three months
into one paragraph: the rumour of the Parthian war,
the position taken on the Amanus, Cassius's
successful repulse of the Parthians from Antioch
(which has spared Cicero's province the brunt of the
invasion), the punitive operation against the
Amanienses, the salutation as imperator at Issus on
the field of Alexander's victory, and the
twenty-five-day investment of Pindenissus that is
still in progress as he writes. The closing turn is
the standing private commission: be watchful, my
Rufus --- send me a successor, or at least no
prorogation. Caelius had written that the matter would
not move; Cicero accepts the assessment but still
asks. The letter is dated four days before Caelius's
\textit{Fam} 8.10, and the two will have crossed at
sea.}
