% Ad Familiares 8.5 --- headnote
% ad-familiares-08-05, before 13 August 51 BC, Rome

\textit{M. Caelius Rufus to Cicero, written from Rome
shortly before the Ides of August 51 BC (the manuscript
dateline: \textit{Scr. Romae ante Id. Sext. a. 703
(51)}, usually placed about the 13th of August). A
short, sober letter from a writer who is usually neither:
Caelius has been reading the dispatches from the east
and has decided that Cicero's safety in Cilicia is in
real question. The first section sets out the
dilemma --- a tidy little Parthian campaign of just the
size to earn Cicero a triumph would be ideal, but the
army he has been given would scarcely hold a mountain
pass, and Rome refuses to face the disproportion
between what is asked of the proconsul and what has
been put at his disposal.}

\textit{The remainder of the letter is the gloomy
political diagnosis behind it. With the consul C.
Claudius Marcellus pressing the question of the
succession to Caesar's Gallic provinces, and with the
familiar machinery of tribunician veto on its way to
stop him, Caelius foresees that the dispute will jam in
the chamber for ``more than two full years'' --- which
is to say, no successor for Cicero is in sight, and
none will be sent. The structural diagnosis is acute:
the obstruction is easy because every interested party
can find someone willing to interpose; and if Curio
secures his tribunate of 50 BC, the same action will
simply restart there. Within the year Curio will indeed
restart it --- not, however, in the direction Caelius
in \textit{Fam} 8.4 had hoped. The fatalism of the
closing sentence is the letter's real news: Cicero
should not expect to be replaced.}
