% Ad Familiares 15.14 --- headnote
% ad-familiares-15-14, ca. mid-October 51 BC, from camp at the Altars of Alexander

\textit{Cicero to C. Cassius Longinus, written from the field
near the Altars of Alexander in the Amanus range, around the
middle of October 51 BC (Perseus: \textit{in castris ad Aras
Alexandri \dots\ med. ut Oct. a. 703}). The opening
salutation --- \textit{M. Cicero imp. s. d. C. Cassio proq.} ---
fixes the political shape of the moment. Cicero is now imperator,
acclaimed by his troops after the campaign he is about to bring
to its head at Pindenissus; Cassius, who as Crassus's quaestor
in 53 had held the broken Roman position in Syria after the
disaster at Carrhae, is still proquaestor and has just left
his province in good odour, with the dispersal of the Parthian
incursion to his credit. The letter accordingly knits two
strands together: routine business of recommendation, and the
machinery of two men in command across the East congratulating
each other and exchanging political requests.}

\textit{The structure is laid out openly --- four items, taken
in order. \S 1--2 dispatch the recommendation of M. Fadius, with
the courteous form-of-art Cicero brings to such letters: he
makes no profit from being asked, since Fadius is already in his
debt-book, but the recommendation has earned its keep by
revealing Fadius's regard for Cassius. \S 3 is the
congratulation that could not be paid in person. \S 4 is the
political advice: Cassius should hurry back to Rome, if his
own household burdens allow; the arrival of a victorious
proquaestor at Rome will be brilliantly received. \S 5 is the
substantive request --- the standard refrain of Cicero's
Cilician correspondence of these months: that nothing be done
to extend the term of his command beyond the year the Senate and
people set for it. Paullus, Curio, and Furnius are named as
allies on the ground at Rome. \S 6 is the closing courtesy: the
friendship is old, was reinforced in Cicero's worst days, and
has lately extended to Cassius's brother Brutus; it asks only
for letters, continuously now and frequently when Cassius is
back in the city.}
