% Ad Familiares 3.6 --- headnote
% ad-familiares-03-06, 31 August 51 BC, in camp (moving from Iconium)

\textit{Cicero to Appius Claudius Pulcher, written from
camp on the last day of August 51 BC (the manuscript
dateline: \textit{Scr. in castris prid. K. Sept. a.
703}). Cicero has by now crossed into his province,
held court at Laodicea, marched east to join the army,
and broken camp at Iconium to push on through
Cappadocia. Appius, the outgoing proconsul, was
supposed to wait for him so that the handover might be
done in person and within the thirty days the Lex
Cornelia (a law fixing the term within which a
proconsul must leave his province after his successor's
arrival) allowed. He has not waited; he has gone to
Tarsus, at the far end of the province, and gone on
holding court, issuing decrees, sitting in judgement
--- the kind of business a man who knew his successor
was almost on him would normally lay aside.}

\textit{The letter is the longest and most direct of
the series. Cicero is now openly remonstrating with
Appius, while preserving the form of the friendship.
The structure is a piece of forensic accountancy ---
\textit{cum meum factum cum tuo comparo}, ``when I
compare my own conduct with yours.'' Sections 1--2 set
out what Cicero himself did: he asked Appius's freedman
Phania at Brundisium where Appius wanted him to come
first, was told Side; he met L. Clodius at Corcyra,
who redirected him to Laodicea; he changed his plan
accordingly. Section 3 turns the page: \textit{considera
nunc vicissim tuum}, ``now consider in turn your own.''
Appius has gone in the wrong direction; ignorant
observers would read his conduct as that of a man
avoiding the meeting. Section 4 raises the rumour that
Appius is still actively governing --- holding court,
deciding cases --- after his successor is in the
country, which is the substantive complaint underneath
the personal one. Section 5 names the operational
grievance: three cohorts at full strength are detached
from Cicero's already-thin army and Cicero does not
know where they are; he has sent D. Antonius to recover
them. Section 6 closes by giving Appius, with pointed
precision, every datum he needs to set up a meeting
within the legal term --- the date of Cicero's entry
into the province, the route through Cappadocia, the
date of breaking camp at Iconium. The courtesy is
intact; the patience is not. The reference to
``ill-wishers'' (\textit{malevoli homines}) in section
4 is the polite shape into which Cicero packs the
suspicion that the rumours about Appius's conduct are
in fact true.}
