% Ad Atticum 5.13 --- headnote
% ad-atticum-05-13, 26 July 51 BC, written at Ephesus

\textit{Cicero to Atticus, written at Ephesus on 26 July 51 BC
(the sixth day before the Kalends of August), four days
after his landing in the province. The opening sentence
gives the journey-letters' precise milestone: arrival on
22 July, ``the five hundred and sixtieth day after the
battle of Bovillae'' --- the clash on the Appian Way in
January 52 in which Milo's slaves killed Clodius, and from
which Cicero dates his exile from Roman politics. The
voyage from Delos has been without fear or seasickness but
slow, because the Rhodian flotilla is what it is. At Samos
and then in greater force at Ephesus the proconsul has been
met by embassies, private petitioners, and an
``unbelievable throng''; the tax-farming \textit{publicani}
already treat him as their imperator, the Greek provincials
as a returning praetor of Ephesus. After many years of
ornamental rhetoric on integrity in provincial government,
Cicero is now on trial against his own \textit{ostentationes},
his showpieces.}

\textit{The second section is one of Cicero's running services
to Atticus, this time in Atticus's business interests in the
province of Asia (governed by Q. Minucius Thermus): Cicero
has commended Atticus's freedman Philogenes and another
agent named Seius to Thermus, along with Xeno of Apollonis,
and has settled an exchange-of-currency account. The third
section turns home, and back to the journey-letters' refrain:
\textit{ut simus annui}, that the Cilician year be only a
year, and that no intercalary month be allowed to extend it.
Then the cluster of commissions: the ``domestic worry'' is
generally read as Tullia's marriage prospects, on which
Cicero and Atticus have been corresponding; the ``Caesar''
business is Cicero's pending loan from him --- begun at
Atticus's prompting and, the proconsul hastily adds, not
yet regretted; the trials, the water-rights at Tusculum,
Philippus the family connection: the same anxious agenda
that will recur through the year.}
