% Ad Atticum 5.8 --- headnote
% ad-atticum-05-08, ca. 2 June 51 BC, written at Brundisium

\textit{Cicero to Atticus, written at Brundisium on the second or
third of June 51 BC. Cicero has been stuck on the heel of Italy
for twelve days, waiting on his legate C. Pomptinus and shaking
off a minor illness, before he can sail for his unwanted
Cilician command. The opening section is the briefest
travel-bulletin, but the news that has just reached him from
Atticus --- the death of Atticus's mother, alluded to by
Shackleton Bailey but in this surviving text passing without
the explicit notice it must have had in the lost portion or
companion letter --- and the press of his own affairs leave
little room for ceremony. In the manuscript order of Atticus
book 5 the letter falls late; in chronology it is the first of
the Cilician series, written before the sailing.}

\textit{The body of the letter is Cicero's anxious request that
Atticus, if still at Rome, intervene in the financial tangle
surrounding the property of T. Annius Milo. Milo, exiled to
Massilia after the trial for the killing of Clodius, had had
his estate sold up; Cicero had agreed --- on C. Duronius's
advice --- to have his own freedman-agent Philotimus take a
share in the purchase, with the aim of keeping the property
within friendly hands, protecting Milo's slaves from being
broken up by a hostile buyer, and securing the settlement Milo
wished for his wife Fausta. Word has come back to Brundisium
that Milo is complaining of the arrangement, and Cicero
backpedals at once: if Milo and Fausta are unhappy, Philotimus
is to be withdrawn. The closing formula --- ``you shall settle
it as it seems to fit with my honour, my reputation, and my
interest'' --- is the standard Ciceronian delegation, but
sharpened here by distance and haste.}
