% Ad Atticum 5.4 --- headnote
% ad-atticum-05-04, 12 May 51 BC, written at Beneventum

\textit{Cicero to Atticus, written at Beneventum on 12 May 51 BC,
the day after he had reached the town from the Trebulan villa
of Pontius (Ad Atticum 5.3 of 11 May). The Perseus dateline has
a year-number typo --- ``701'' for 703 --- but the day-of-month
is secure: iv Id. Mai., 12 May. The letter answers a double
delivery from Atticus that came in at Beneventum within hours of
each other, one carried by Funisulanus before dawn, the other
by the secretary Tullius. The first and greatest business is
the search for a husband for Tullia: Atticus has been working
on it, but his own approaching departure for Epirus, and the
fact that the most plausible candidate is one whom Atticus calls
\textit{dysdiagnoston} --- hard to read --- leaves Cicero with
little leverage from a thousand miles away. A wistful aside
imagines what Servilia might have done with Servius Sulpicius
Rufus if either Atticus or Cicero had been at Rome to push the
match.}

\textit{The rest of the letter is the running mass of small
business that Cicero's setting-out has thrown up at once.
Section 2 disposes of a string of items in a single page: M.
Marcellus's affair, the credit for which is to be entered to
Cicero and Bibulus jointly; Torquatus; the absent Maso and
Ligus; the petitioner Chaerippus, on whom Atticus has now ``lifted
the \textit{prosneusin}'' --- removed the favourable inclination
--- and at whom Cicero throws an exasperated \textit{o
provincia!}; the speed at which his legate Pomptinus is now
expected at Brundisium; the prefects Pompey will appoint. The
practical heart is in \S 3 with the eight hundred thousand
sesterces owed Oppius, which Cicero wants cleared up before
Atticus leaves Rome. \S 4 closes with an embarrassed joke
about Atticus's complaint that he is short of paper: Cicero
will send two hundred sheets, though the cramped little page of
this very letter is a comment on his own thrift.}
