% Ad Familiares 2.8 --- headnote
% ad-familiares-02-08, 6 July 51 BC, Athens

\textit{Cicero to M. Caelius Rufus, written from Athens
on the 6th of July 51 BC (the manuscript dateline:
\textit{Scr. Athenis prid. Non. Quint. a. 703 (51)}),
on the day of his departure for the province of Cilicia.
This is, in effect, the charter for the run of
Roman-news letters from Caelius which will run from
\textit{Fam} 8.1 through the rest of Cicero's eastern
year. Cicero has now received Caelius's first dispatch
--- the cover-letter with the hired bulletin-compiler's
roll attached --- and is writing back to redirect the
service. He does not want the gladiator-matches and
court-day-postponements that a professional Roman
news-sheet supplied; he wants the political weather, and
in particular the future, read by the sharpest political
intelligence in his circle. The line about Caelius being
the most \textit{[Greek: politikoteron]} man Cicero
knows is half-flattery, half-genuine, and entirely
characteristic of the relationship.}

\textit{The middle paragraph reports the substance of
Cicero's long stop with Pompey at Tarentum on the way
east: many days of conversations on the state, which
cannot be written down, and a verdict that Pompey is
sound. Caelius is told to attach himself to him --- a
strategic prescription that the events of the next
eighteen months will sorely test. The final paragraph
gives the practical commission that frames the whole
correspondence: Cicero is leaving Athens for his
province on this very day; the one thing he wants
Caelius to manage in his absence is that the year of
the Cilician command not be extended. The means and
moment of doing so are left to Caelius's discretion ---
a vote of confidence, and a piece of business that will
recur in nearly every letter Cicero writes back from
Cilicia.}
