% Ad Atticum 5.10 --- headnote
% ad-atticum-05-10, late June / 1 July 51 BC, written at Athens

\textit{Cicero to Atticus, written at Athens on the eve of his
crossing into Asia --- the Perseus dateline gives ``the day
before the Kalends or the Kalends of July, 51 BC,'' that
is 30 June or 1 July. Cicero has now been four days at
Athens, still waiting for his legate Pomptinus, and is
filling the time by walking among the haunts that bear
Atticus's footprints from the years his friend lived there.
The first section is pure feeling: every conversation
turns to Atticus. The second is the centrepiece of the
journey-letters' running self-portrait --- a proconsul who
has impressed on his whole staff that they must serve his
\textit{fama}, who is taking nothing for himself either
under the lex Julia de repetundis (Caesar's extortion law
of 59) or from his hosts; and a man who knows enough not
to take a victory lap until the year is over.}

\textit{The third section is the inward counterweight, full of
the Greek tags Cicero scatters more freely in letters to
Atticus than to anyone else. He quotes the Homeric tag
\textit{[Greek: erdoi tis hēn hekastos eideiē technēn]} ---
``let each man practise the craft he was born for,'' from
the lost \textit{Margites} (a line later proverbial) --- to
sigh that the proconsulship is a craft he was not born
for. He bears the daily affronts of his entourage with a
straight face, but the strain on his \textit{[Greek:
bathytēs]}, his ``depth of composure,'' is real, and the
year's hard \textit{[Greek: meletē]} (training, discipline)
will be its own philosophical exercise. The closing
section deflates philosophy back to gossip: Athens itself is
charming, its present-day Academic chair Aristus less so
(the corrupt \textit{sursum deorsum} ``up and down'' shows
the manuscript already troubled by Cicero's joke), and the
gentler Xeno of the Epicurean garden, ceded to Quintus, is
his pleasanter daily company. He ends with the same anxious
question of every letter on this road: when will Atticus
next be at Rome to attend to his affairs?}
