% Ad Familiares 2.9 --- headnote
% ad-familiares-02-09, late October or early November 51 BC, camp at Pindenissus

\textit{Cicero to M. Caelius Rufus, written from the
camp at Pindenissus in the highlands of Cilicia in late
October or early November 51 BC (the manuscript dateline:
\textit{Scr. in castris ad Pindenissum vel ex m. Oct.
vel in. Nov. a. 703 (51)}). It is the prompt reply
to the news --- belatedly arrived at the besieging
army --- that Caelius has been elected curule aedile.
The greater part of the elation is at Caelius's other,
unhoped-for victory: that of his old rival Hirrus,
defeated by Caelius in the augural election earlier in
the year. Cicero is sending congratulations from a
camp under a Cilician hill-fort, with brigand-country
between him and Rome and his news running months
late.}

\textit{The conceit on which the letter turns is that
of an actor working a part. Cicero, hearing of Hirrus's
defeat, claims to have become Hirrus himself --- the
pompous orator from Lucilian satire, with his
\textit{balbus} (stammering) speech and his roster of
showy young proteges --- and to have played out his
mannerisms scene by scene. The three short scraps of
verse in section 2 are quotations or near-quotations of
old Latin comedy or tragedy (\textit{ne edepol quantam
rem egeris...}, \textit{incredibile hoc factu obicitur},
\textit{ego voluptatem animi nimiam}), each set in
Hirrus's mouth and broken off at a comic instant.
Cicero closes by collecting himself in his proper voice:
he loves Caelius as the man fortune has given him as
the magnifier of his own standing and the avenger of
his enemies --- a glance back at the political shaming
of Hirrus, of which the augural defeat had been only
the beginning.}
