% Ad Familiares 3.8 --- headnote
% ad-familiares-03-08, 8 October 51 BC, camp in the territory of Mopsuhestia

\textit{Cicero to Appius Claudius Pulcher, his
predecessor as governor of Cilicia, written from camp
in the territory of Mopsuhestia on the 8th of October
51 BC (Perseus dateline: \textit{Scr. in castris in agro
Mopsuhestiae a. d. viii Id. Oct. a. 703 (51)}). It is
the long and sharply controlled letter in which Cicero
finally expostulates with Appius for the whole sequence
of failed meetings, broken promises of contact, and now
the imputation --- circulated by ``ill-disposed men''
and politely conveyed in Appius's last letter --- that
Cicero has shown by expression and silence that he is
no friend of his predecessor. It is the second great
Appius confrontation in the correspondence, the first
being the run of letters in which Cicero traced the
successive places at which Appius had failed to meet him
on the road. Here the patience has thinned.}

\textit{The substantive complaint Cicero answers is
Appius's: that Cicero's edict on the conduct of the
province has been read as a deliberate frustration of
the embassies of thanks being sent to Rome on Appius's
behalf, that those embassies have been deprived by
Cicero of their proper expenses, and that Cicero's
public manner toward Appius has been less than friendly.
Cicero answers in three movements. First (sections
2--5), the procedural defence: he composed the edict at
Rome, copied a clause from Appius's own edict at the
tax-farmers' request, and limited the expenses of
embassies only at the express request of the
communities themselves --- communities driven to ruin
by exactions their own magistrates had piled up. Second
(section 6), the courteously violent reminder: ``do you
think nothing has ever been said to me about you?''
Cicero crossed into the province expecting a meeting at
Iconium; Appius was at Tarsus. While Cicero held his
assize at Apamea, Synnada, and Philomelium, Appius was
holding his at Tarsus on the same days. The reproach
is not made for its own sake, Cicero says, but to make
the point that it is the listener who hears the
detraction, as much as the speaker, who is at fault.
Third (sections 7--8), the principled distinction:
Cicero will not be drawn into the small currency of
counter-detraction. His liberality, he concedes, has
been narrower than Appius's --- but in the lighter
times of Appius's first year. In the hardened later
year, Cicero ends with the quotation that he must be
bitter to the provincials in order to be sweet to
himself: ``\textit{med esse acerbum sibi, uti sim
dulcis mihi}.'' The letter closes with practical
business: a renewed request that Hortensius withdraw an
unwelcome biennial proposal, the report of Cicero's own
march toward Mount Amanus, and the assurance that the
Parthian alarm appears for now to be over. The Appius
correspondence will continue, but the bottom of the
relationship has been reached here.}
