% Ad Familiares 15.10 --- headnote
% ad-familiares-15-10, late November 51 BC, from camp / Tarsus

\textit{Cicero to M. Claudius Marcellus, consul of 51
BC, written from Cilicia late in the year (Perseus
dateline: \textit{Scr. Tarsi vel ex. a. 703 (51) vel
in. 704 (50)}, i.e. either the close of 51 or the
opening of 50). The letter is a short request:
Marcellus, in his last weeks of office as consul,
should see that when Cicero's dispatch of victory is
read in the Senate, the corresponding decree of a
public thanksgiving --- \textit{supplicatio} --- is
voted in the most honourable terms. It is the
companion piece to the long letter to Cato (\textit{Fam.}
15.4) and part of the larger lobbying campaign by which
Cicero, after the Pindenissus operation, sought
senatorial recognition for his year of command.}

\textit{The letter's currency is the unbroken history
of services between the Marcelli and Cicero. The
opening sentence, with its careful cataloguing of
``the Marcelli, and the Marcellini too,'' acknowledges
that the entire family --- this consul, his father, his
brother, his cousins --- has stood by Cicero in matters
of his standing and safety. The implication is the
quietly insistent one of the genre: I have never asked
you for anything that I have not earned the right to
ask. The hard practical move is the request that the
consul take charge of the procedural moment when the
dispatch is read out, and shepherd the decree through
in good form: Marcellus held the fasces in November and
December and could control the order of business. The
\textit{supplicatio} did in fact pass, after some
contest, but only after Cato voted against it on
principle; for that further chapter see \textit{Fam.}
15.4 and 15.6.}
