% Ad Familiares 8.9 --- headnote
% ad-familiares-08-09, 2 September 51 BC, Rome

\textit{M. Caelius Rufus to Cicero, written from Rome
on the 2nd of September 51 BC (the manuscript dateline:
\textit{Scr. Romae iv Non. Sept. a. 703 (51)}). The
fullest Caelius newsletter so far, and the one in which
the writer is plainly enjoying himself: he opens with
the news he had hinted at in \textit{Fam} 8.4 ---
Hirrus, beaten in the augural election, has been
publicly and totally demolished --- and gives himself
the indulgence of a brief vignette of the loser's
afterlife (delivering anti-Caesar speeches in the
chamber, taking up minor liberty-suits in the Forum
``rarely after noon''). The conventional politics of
Roman shaming are very visible: the defeated rival is
no longer a man to take seriously.}

\textit{The middle section returns to the public
business of the August letters and reports a fresh
deadlock: the consul-designate Marcellus has interrupted
the Ides-of-August session on the provinces; the matter
has been pushed to the Kalends of September, and even
those have come and gone without progress, as a quorum
could not be raised. Caelius now states, plainly and as
his own settled view, that no successor will be sent
to Cicero in his proconsular year, and that Cicero will
have to leave behind a deputy --- the prediction from
\textit{Fam} 8.5 hardening into a fact. The rest of the
letter is the usual mixture of practical business and
political colour: the panthers for the aedilician
games, which Curio has now upstaged with a gift of ten
plus ten African beasts; a commendation of one M.
Feridius, a young Roman knight on business in Cilicia
who wants the city-leased lands he holds exempted from
tribute; the failure of M. Favonius, the noisy Catonian,
to win the praetorship; and, most ominously, Pompey's
\textit{vox} that no decree should be passed at present,
counterposed to Scipio's motion that the Gallic question
be brought up on the Kalends of March 50, by itself,
disjoined from all other provinces. That formula will
shape the next year of senatorial business.}
