% Ad Familiares 7.27 --- headnote
% ad-familiares-07-27, around March 52 BC, Rome

\textit{Cicero to T. Fadius Gallus, written from Rome
around March 52 BC (the manuscripts: \textit{Scr. Romae,
ut videtur, post m. Mart. a. 702}). The salutation in the
manuscripts reads only \textit{Gallo}; some early editors,
misled by the cognomen, supposed the addressee to be the
Epicurean Volumnius Eutrapelus, but the content fixes the
recipient as Cicero's old quaestor T. Fadius Gallus,
tribune of the plebs in 58 BC --- one of the tribunes who
had stood with Cicero against Clodius in the run-up to
the exile. Fadius had since been convicted (we do not
know on what charge) and was angling for restoration; he
had evidently asked Cicero to use his influence, been
refused, and now written an aggrieved letter accusing
Cicero of cowardice and ingratitude. This is the
reply.}

\textit{The register is unique in the surviving book 7
--- the playful, affectionate correspondence with
Trebatius and the wits has nothing in it like this. The
letter is short, sharp, and deliberately cutting. The
opening lines reduce Fadius's complaints to single
points and dispatch each: he ``attended dutifully''
during the consulship (a small service over against
Cicero's claim that he, Cicero, is the reason the rest
are \textit{free}); Caesar will restore him (``no one
believes you''); he sought the tribunate for Cicero's
sake (``I wish you were tribune still --- you would not
be looking about for a veto,'' i.e.\ you would be the
veto). The closing turn is the cruelest: if the reports
you say you brought me about Clodius's faction were
false, I owe you nothing; if they were true, then you
yourself are the best witness to what the Roman People
owes me. The whole letter is a study in Cicero putting
an ingrate in his place with no waste of words.}
