% Ad Familiares 5.18 --- headnote
% ad-familiares-05-18, around 20 March 52 BC, Rome

\textit{Cicero to T. Fadius, written from Rome around
20 March 52 BC (the manuscripts: \textit{Scr. Romae post
med. m. Mart. a. 702}). The recipient is T. Fadius,
Cicero's quaestor in 63 BC during his consulship and
afterward tribune of the plebs in 57 BC, when he worked
energetically for Cicero's recall from exile. He has now
himself been convicted in a court --- on what charge the
letter does not specify, though the language of
``laws, courts, and times'' and the single doubtful
vote points to the wave of prosecutions that followed
Pompey's sole consulship in the wake of Clodius'
killing, when new laws on \textit{vis} and
\textit{ambitus} were rapidly producing condemnations.
The verdict has cost Fadius his place in the
commonwealth; his property and his children remain.}

\textit{The letter is short, and the consolation is
political rather than philosophical. Cicero does not
reach for the schools, as he had a few weeks earlier in
\textit{Fam.} 5.16 to Titius; here the argument runs
the other way --- the times are so ruinous that the man
who has been removed from public life at the lightest
penalty has been best served by his fate. The line
\textit{plus tibi virtus tua dedit quam fortuna
abstulit} (``your virtue has given you more than
fortune has taken away'') turns on the contrast
between \textit{novus homo} and \textit{nobilissimus}:
Fadius the new man has gained office --- something most
new men never reach --- and has lost only what the
greatest of the nobility have also lost. The closing
pledge of unchanged goodwill toward Fadius and his
children is the form a consolation took when there was
nothing else to be said.}
