% Ad Familiares 13.75 --- headnote
% ad-familiares-13-75, late 52 BC or early 51 BC, Rome

\textit{Cicero to T. Titius the legate, written from
Rome at the end of 52 BC or the beginning of 51 BC (the
manuscripts: \textit{Scr. Romae vel ex. a. 702 vel in.
a. 703}). T. Titius is serving as a \textit{legatus} on
Pompey's grain commission --- the \textit{cura
annonae} which Pompey had been voted in 57 BC and was
still discharging through subordinates --- with
responsibility for places of delivery and schedules of
shipment. The subject is the same C. Avianius Flaccus
whom Cicero had commended to Pompey himself for the
three preceding years (\textit{triennium, dum Pompeius
isti negotio praefuit}): a grain-shipper or
\textit{publicanus}-class merchant whose business
turned on which port he was directed to and on when
his cargoes had to be in.}

\textit{The letter is a renewal, not a new request,
and Cicero says so frankly. He has already spoken with
Titius in person --- and been answered \textit{humanissime}
--- and has already written at length; the present
letter exists because Avianius wants the recommendation
repeated, and Cicero apologises in advance for what
might look like distrust of Titius' \textit{constantia}
(his reliability in keeping a promise). The closing
formula --- \textit{ut Avianius, quoniam se a me amari
putat, me a te amari sciat} (``so that, since
Avianius thinks himself loved by me, he may know
himself loved by you'') --- is one of the most polished
turns in Book 13: it makes Titius' favour to Avianius
double as a favour to Cicero, and lets Cicero close on
a note of triangulated affection rather than on a
specific request.}
