% Ad Familiares 5.16 --- headnote
% ad-familiares-05-16, around March 52 BC, Rome

\textit{Cicero to Titius, written from Rome around March
52 BC (the manuscripts: \textit{Scr. Romae, ut videtur,
circ. m. Mart. a. 702}). The recipient --- a Titius
otherwise unknown, perhaps the senator of that name
attested in the early 40s --- has lost children, more
than one, in what Cicero calls ``this most calamitous
and pestilent year'' (\textit{hoc gravissimo et
pestilentissimo anno}): the early months of 52 BC, in
which Clodius had just been killed on the Appian Way,
the Senate house had been burnt over his pyre, and Rome
was without proper magistrates, sliding toward the sole
consulship of Pompey. The plague Cicero glances at is
real but secondary; what makes this year ``pestilent''
in the writing is the political ruin behind the plague.
The letter is a letter of consolation in the strict
ancient sense --- the \textit{consolatio} as literary
form, with its inherited topoi --- and it is unusually
substantive among Cicero's surviving examples of the
genre, set late in his fifty-fourth year, well before
his own great bereavement (the death of Tullia in 45
BC, which would produce the lost \textit{Consolatio} and
shape book 1 of the \textit{Tusculans}).}

\textit{The argument moves through the standard
philosophical commonplaces, knowingly. The opening
topos --- we are men, born under this law, our life set
up as a target for all the weapons of fortune --- is
the Stoic-Academic commonplace Cicero will rework into
the Latin theoretical literature of his last years (cf.
\textit{Tusc.} 3, the lost \textit{Consolatio},
\textit{De Senectute}). The classical disjunction on
death --- if perception remains, what is in question is
immortality, not death; if it does not, no wretchedness
can be felt --- is the schoolroom argument Cicero will
attribute in different forms to Socrates and to
Epicurus and rework in his own voice in
\textit{Tusc.} 1. What is distinctive here is the third
move, the political one: the standard commonplaces no
longer carry their old force, because the times do not
let them. The state itself has become a worse evil than
death; the children Titius has lost have been spared
\textit{by the immortal gods} from a condition of life
worse than the loss. That argument --- not in the
philosophers --- is Cicero's contribution, and it tells
us where he stood in March 52 BC: still able to
mobilise the schools, no longer able to believe what
their consolation was originally for. The letter closes
on the topos of \textit{anticipating} the medicine of
time --- doing now, by reason, what duration will do
of itself --- a figure that points forward to the
explicit programme of the \textit{Tusculans}.}
