% Ad Familiares 2.1 --- headnote
% ad-familiares-02-01, 53 BC, Rome

\textit{Cicero to C. Scribonius Curio, written from Rome on 6
October 53 BC (the manuscripts give only the year, \textit{Scr.
Romae a. 701}; the conventional date follows Shackleton
Bailey's chronology). Curio is by now finishing his quaestor-
service in Asia and on the slow road home --- the same return
journey that Fam 2.6 looks forward to in the spring. The
ostensible occasion is a complaint Curio had evidently raised
that Cicero was not writing often enough. Cicero's reply is a
small set-piece of friendly counter-litigation: the charge is
welcome because it shows the affection, but on the merits it is
false; he has given a letter to every plausible courier, and it
is Curio who has sent only two or three short ones. He
threatens, in mock-forensic form, to bring the same charge
back the other way.}

\textit{The second section turns to the standard \textit{
clausula} that recurs through the book 2 letters: a careful
mixture of compliment and counsel. Curio's long absence
abroad has been a loss to Cicero personally and a gain to
Curio's career; fortune, Cicero says, has answered all his
wishes for his prot\'eg\'e. The note of warning is barely
veiled. The \textit{expectatio} that Curio has built up around
himself --- the same heavy adversary set up against him in
Fam 2.4 --- will have to be sustained when he comes home; he
should return \textit{conformatus}, shaped for the part. The
closing turn is the most disclosing line in the small Curio
correspondence: whatever Curio has won, Cicero says, he could
not have won had he not as a boy obeyed Cicero's counsels, and
so Cicero's now-heavy years deserve to find their rest in
Curio's love and youth. The patronage is being invoiced. Within
three years, the prot\'eg\'e who is here being reminded of his
debts would be Caesar's man at the tribune's bench, and the
debt would not be paid.}
