% Ad Familiares 2.2 --- headnote
% ad-familiares-02-02, 53 BC, Rome

\textit{Cicero to C. Scribonius Curio, written from Rome on 7
November 53 BC (the manuscripts give only the year, \textit{Scr.
Romae a. 701}; the conventional date follows the standard
chronology). The occasion is the death of the elder Curio, C.
Scribonius Curio (consul 76 BC), the young man's father, who
had died at Rome while Curio was still abroad on his return
journey from Asia. The note is a single short paragraph and
performs the smallest possible Roman \textit{consolatio}: an
elegant compliment to the dead (he would have outdone all the
world in fortune had he only lived to see his son once more),
a brief pious wish that the gods bless the patrimony Curio now
inherits, and a quiet relocation of patronage in Cicero's own
person --- the father had been a great witness to Cicero's love
for the son, but the friendship needs no witness, and Curio
will find Cicero himself as devoted to him as the father was.}

\textit{The political subtext is heavy. The elder Curio had
been one of Cicero's most prominent senatorial defenders during
the exile of 58--57; his death now removes from the young
Curio's life the conservative father who had largely held him
to the \textit{boni}, and Cicero is the obvious replacement
\textit{auctor} on offer. The letter is at once a courtesy and
a quiet bid, written in the same November that follows the
spring's larger appeal of Fam 2.6 (Milo's consulship) and the
autumn's gentler counsel of Fam 2.1 (the reminder of debts).
Within months Curio would be back in Rome, the inheritance in
hand, and on the path that would take him through Clodius'
murder, the Milo trial, and his sale to Caesar.}
