% Ad Familiares 2.6 --- headnote
% ad-familiares-02-06, 53 BC, Rome

\textit{Cicero to C. Scribonius Curio, written from Rome in
March 53 BC (the manuscripts give the year only: \textit{Scr.
Romae a. 701}; the conventional date 11 March follows
Shackleton Bailey's reconstruction of the run-up to Curio's
return from quaestor-service in Asia). This is the substantive
letter of the small Curio sub-correspondence preserved in book
2 of the \textit{Familiares}. The frame is a campaign favour:
T. Annius Milo, Cicero's saviour-tribune of 57 and patron-in-
return, is running for the consulship of 52, and Cicero is
calling in every chip he holds with the young Curio. The
\textit{exordium} (section 1) is a textbook Ciceronian
captatio: he is embarrassed to ask anything large of one to
whom he is already so much in debt, because to ask would look
like dunning a creditor; but Curio's kindnesses to him --- those
performed during his exile and recall, the \textit{novitas
meorum temporum} --- have been so signal and so much spoken of
that to owe still more would only be generous. The whole
paragraph is an ornament on a single rhetorical move: I will
ask, and the asking will be a compliment.}

\textit{The body (sections 3--4) lays out the case for Milo
plainly. Cicero has staked everything on the consular campaign;
he has the \textit{boni}, won over by Milo's tribunate; the
\textit{plebs}, won over by Milo's games and openhandedness;
the young and the well-connected on the voting rolls. What he
lacks is a single great patron to be \textit{dux et auctor}, a
steersman for the winds already blowing. Curio, just returning
from Asia, is being asked to be that man. The closing turn is
striking: Cicero will owe more to Curio for backing Milo than
he owes to Milo himself, because Milo only saved his life,
whereas to back the saviour is the sweeter pious return. The
reader who knows what is coming --- Clodius dead on the Appian
Way in January 52, Milo prosecuted and exiled, Curio bought by
Caesar for the tribunate of 50 --- reads this as a piece of
careful, doomed political husbandry. The flattery is
interested, and the patient courtship of a vain young patrician
that runs through book 2 is here at its most concentrated.}
