% Ad Familiares 2.5 --- headnote
% ad-familiares-02-05, 53 BC, Rome

\textit{Cicero to C. Scribonius Curio, written from Rome in 53
BC (the manuscripts give only the year: \textit{Scr. Romae
a. 701}). The second of the small sub-correspondence with
Curio in book 2 of the \textit{Familiares}, and a notch
darker in tone than 2.4. Where the prior letter circled around
the impossibility of writing seriously about the commonwealth,
this one says outright that affairs at Rome cannot be set down
even in a letter --- and then, with the same elaborate
courtesy, draws Curio's distance into a kind of consolation:
you are away, so you do not see what we see, and meanwhile your
reputation rises in the unanimous voice of allies and citizens
alike. The compliment is real and the political flattery is
plain.}

\textit{The second section turns the screw. The
\textit{exspectatio reditus} --- the extraordinary expectation
of Curio's return --- is double-edged: Cicero fears not that
Curio's virtue will fall short of it, but that by the time
he arrives there will be nothing left to act on. The
state is so weakened, so nearly extinguished
(\textit{ita sunt omnia debilitata et iam prope exstincta})
that even saying so feels unsafe for a letter. The close
returns to Curio's own \textit{clausula} from 2.4: prepare,
rehearse, consider the qualities of the citizen and man who
will vindicate the commonwealth, struck down and oppressed,
back into its old dignity and freedom. The vocabulary
(\textit{adflictam et oppressam}, \textit{veterem dignitatem
et libertatem}) is the formal language Cicero uses for the
republic he believes is dying around him.}
