% Ad Familiares 2.3 --- headnote
% ad-familiares-02-03, 8 December 53 BC, Rome

\textit{Cicero to C. Scribonius Curio, written from Rome
late in 53 BC (the manuscripts give only the year:
\textit{Scr. Romae a. 701}; the conventional date is 8
December). Curio --- the future tribune of 50 BC, at this
date still a brilliant young man abroad in the East --- is
about to come home into a quaestorship that will turn into
something larger. His agent Rupa had set in motion the
declaration of public games \textit{in his name}, a
familiar Roman expedient by which a young man rising
toward the magistracies advertised himself to the people
with a costly spectacle. Cicero has stopped Rupa, and is
writing to explain why before the matter is fixed.}

\textit{The shape of the letter is the keynote of the
whole sub-correspondence with Curio (\textit{Fam.}
2.1--7): patient, flattering management of a brilliant
patrician's vanity, executed with such formal courtesy
that the management almost disappears. The argument
itself is sober: games are a matter of means, not of
merit (\textit{copiarum, non virtutis}); no one admires
the capacity to put them on; the audience is by now
wearied to disgust. Curio possesses what games cannot
give him --- nature, study, fortune --- and the
commonwealth needs him in the qualities those gifts
have built. The closing flourish is the same
\textit{clausula} Cicero will use again in 2.4: the
exhortation to the highest praise, the assurance that no
one is dearer. The careful management is the message;
the letter exists to forestall a young man's mistake
without letting him feel managed.}
