% Ad Familiares 7.18 --- headnote
% ad-familiares-07-18, 8 April 53 BC, Pomptine marshes (villa of M. Aemilius Philemo)

\textit{Cicero to C. Trebatius Testa, written on 8 April 53 BC
from a friend's villa in the Pomptine marshes south of Rome,
where Cicero had broken his journey. Trebatius, the young jurist
whom Cicero had pressed upon Caesar, was still on the Gallic
staff. The letter opens with brisk encouragement --- Trebatius is
at last bearing camp life like a man --- and a teasing promise to
``renew the recommendation, but at the right moment.'' Because
the legal guarantees Trebatius can give from Gaul are worthless
at this distance, Cicero says he is enclosing a little Greek IOU
in his own hand: a joke on the fashion among Roman lawyers for
\textit{cautiones} dressed up in Greek formulas. He then asks
for real news of the Gallic war, professing his usual rule that
the bigger the coward, the more reliable his battle-reports.}

\textit{The middle section turns to Trebatius's recent letters
themselves: why send several copies of the same letter in your
own hand, and why scrape down a palimpsest to do it? Cicero
spins a small comic essay out of the question --- either Trebatius
is wiping out his own legal forms, or he is so unoccupied that
he cannot even get hold of paper, in which case it is his own
fault for leaving his Roman modesty behind. The closing
paragraph descends from joke to particulars: Cicero will commend
Trebatius to Balbus when Balbus next goes north; he is on the
road for April, so the post will be irregular; and at Ulubrae a
multitude of frogs has croaked in honour of their new patron ---
a passing dig at the desolate little Pomptine town and its
electorate. A postscript notes that he has torn up an earlier
letter from Trebatius, harmless in itself, on the writer's own
instructions.}
