% Ad Familiares 15.13 --- headnote
% ad-familiares-15-13, January 50 BC, Tarsus

\textit{Cicero from Tarsus, on or just after 1 January 50 BC, to L.
Aemilius Paullus, now actual consul (the salutation has changed since
\textit{Fam.} 15.12 from \textit{cos. desig.} to \textit{cos.}).
Perseus groups the letter with \textit{Fam.} 15.4 and 15.10, of late
51, but the address makes the new year's opening days the obvious
moment of composition. The proximate aim is sharp and specific:
Cicero has won a modest victory on Mount Amanus, has been hailed
\textit{imperator} by his troops, and is now lobbying the consuls and
senate for a \textit{supplicatio} --- a formal public thanksgiving ---
to be voted on his behalf, with all the political weight that such a
vote will carry when he comes home to face the deteriorating
republic.}

\textit{The letter is a small masterclass in how to ask without
appearing to ask. Cicero builds an architecture of mutual indebtedness
--- Paullus lent lustre to his consulship, lent lustre to his return
from exile, and now finds Cicero's third great public moment falling
in the year of Paullus's own consulship --- and then declines, with
elaborate civility, to press his request with the long speech the
case would warrant. He merely begs Paullus to procure the decree
\textit{quam honorificentissime} and \textit{quam celerrime}, to
manage his reputation in absentia, and (the constant refrain of these
months) to see that his year-long commission not be extended. The
official dispatches mentioned at the close are \textit{Fam.} 15.1
(to the senate) and 15.2 (to the consuls and praetors).}
