% Ad Quintum Fratrem 2.13 --- headnote
% ad-quintum-fratrem-02-13, beginning of June 54 BC, Rome

\textit{Cicero to his brother Quintus, written at Rome on 29 May
54 BC --- ``the day I came to Rome'' --- after he had been at
Cumae for the spring. Quintus is by now in Caesar's camp in Gaul,
on the eve of the second British expedition. Two letters have just
reached Cicero, the second carrying a personal letter from Caesar
himself: ``full of every kindness and care and charm.'' The
letter's first section is a small declaration of Cicero's settled
heart on the matter --- the honours Caesar promises him are
great, but what he prizes most he has already: Quintus's devotion
to their shared standing, and Caesar's affection.}

\textit{Section 2 is the famous Cicero pledge to write a poem on
Caesar's British expedition. ``You are urging on a runner''
--- Quintus's exhortation to bring all of Cicero's zeal to bear
on Caesar is unnecessary; he is already running. The metaphor
unfolds: a slow start can be made up by a rush; he will sprint
the cultivation he has too long slept on; with horses, indeed,
and (since Caesar approves his poem) with the four-horse chariot
of poetry. ``Only give me Britain, that I may paint it with your
colours and my own brush.'' Quintus is to send the eyewitness
report; Cicero, the verses. He breaks off self-mocking on time:
where, with Caesar wanting him to stay at Rome, will he find
the hours?}

\textit{Section 3 turns to two business items. Caesar has thanked
Cicero (``wittily and graciously'') for sending him Trebatius:
in such a crowd, he writes, no one knew how to draw up a
\textit{vadimonium}, a forensic recognizance, until Trebatius
came. The tribunate-question for M. Curtius is more delicate.
Domitius, the consul of 54, was famously close-fisted with such
favours --- he liked to say he was not even making anyone a
tribune of the soldiers, and joked publicly that his colleague
Appius had gone to Caesar to fetch some tribunate or other.
Cicero has therefore asked Domitius to grant Curtius a tribunate
for the year after, on Curtius's own suggestion.}

\textit{Section 4 is the political summary. Cicero promises
Quintus that he is, and will be, ``softer than the lowest
ear-tip'' both in public affairs and in private quarrels. The
state of the city is the year's familiar fog: hopes of elections,
suspicions of a dictatorship, leisure in the forum but the
leisure of a state aging rather than resting; the senate's
proceedings reach a position where others assent to Cicero rather
than Cicero to himself. The closing Greek tag --- \textit{such is
the work of wretched war} --- is from Aristophanes (or a tragic
echo); the war Cicero has in mind is the larger civic war just
out of view.}
