% Ad Atticum 4.15 --- headnote
% ad-atticum-04-15, 27 July 54 BC, Rome

\textit{Cicero to Atticus, written from Rome on the sixth day before
the Kalends of Sextilis --- 27 July 54 BC. Atticus is in the East
on the journey to Asia that 4.14 had assumed; Cicero, back in town
after a circuit out to Reate and the Velinus drainage country, is
writing the long Roman bulletin Atticus had asked for. The first
three sections handle private business --- the manumission of
Eutychides, paralleled to Atticus's earlier freeing of Dionysius;
a tease about Atticus lingering in Greek company; the working
explanation for why earlier letters never arrived. Then, at
section 4, the dispatch from the Forum begins.}

\textit{The middle and back half of the letter is one of the
sharpest contemporary pictures we have of the 54 BC consular
campaign collapsing into open corruption: Sufenas and Cato
acquitted, Procilius condemned, and a jury of ``thrice-honoured
Areopagites'' who care, in Cicero's accounting, about nothing
except domestic murder. Memmius has Caesar's money behind him;
the sitting consuls have secretly cut Domitius into the same
syndicate (Cicero will not put the terms in writing); Pompey
backs Scaurus but no one knows whether sincerely; Messalla is
sinking under the consuls' combine. The Homeric tag \textit{sēma
de toi ere\=o} (``and I will tell you the sign,'' from Iliad
23.326) introduces, with mock grandeur, the index of corruption
that interest rates have gone from 4 to 8 percent in a fortnight
as cash is hoovered up for bribes. The closing sections itemize
Cicero's own defense work --- Messius now, Drusus and Scaurus next
--- and end with the standing anxiety: Quintus is now somewhere in
Britain with Caesar's army, and the household waits for word.}
