% Ad Familiares 1.10 --- headnote
% ad-familiares-01-10, end of 54 BC, Rome

\textit{Cicero to L. Valerius the jurist, written from Rome
late in 54 BC (Perseus dateline ``ex a. 700 (54)'' = end of
year). The letter is one section: a teasing, very short
recall-summons to a friend who has been spending too long in
the provinces. The Lentulus referenced is P. Lentulus Spinther
proconsul of Cilicia (Cicero's correspondent of Fam. 1.1--1.7),
to whom Cicero has, on Valerius's behalf, written thanks.}

\textit{The body is the comic-rhetorical mode of the letters
to friends not currently embroiled in politics: the playful
``audacity instead of wisdom'' framing of the opening ---
i.e., I will speak boldly because the times demand it; the
double charge from the home-comers, that Valerius is either
haughty when he says nothing, or insulting when he answers
badly; the wish to ``joke face to face,'' the close on the
Ulysses-recognition gag --- ``if you come there, you, like
Ulysses, will recognize none of your own.'' The line is a
glance at the Odyssean homecoming: Odysseus, returning to
Ithaca after twenty years, was not recognized by his own
people. Apulia, in Valerius's case, has perhaps been the
posting; his Roman friends will go on without him if he
delays.}

\textit{The recipient is identified in the heading as
\textit{iurisconsultus} (jurist) --- one of the practical
legal experts, in the line of Q. Mucius Scaevola the augur and
Trebatius. The closing is the small Ciceronian intimacy
addressed to a man of his own profession.}
