% Ad Atticum 4.17 --- headnote
% ad-atticum-04-17, 1 October 54 BC, Rome

\textit{Cicero to Atticus, written at Rome on the Kalends of October
54 BC --- ``it is already growing light,'' he says in section 4, as he
finishes the letter at first light on the day the Senate is to
meet. Atticus is still away in the east, his itinerary unknown
even to Cicero: Epirus, Athens, Asia are all on the table, and
Cicero will not send the letter except by a trusted bearer going
to him in person, because the contents are too compromising to
risk. The opening is a frank statement of the new conditions of
the correspondence: his letters now carry such confidences that
not even his own copyists are quite safe with them.}

\textit{And the confidences are spectacular. The bulk of the letter
is the great electoral-bribery scandal of 54: the consular
candidate Memmius, having broken with his coalition partner
Calvinus and seen his prospects collapse, has read out in the
Senate --- on Pompey's urging --- the actual bargain by which he and
Domitius (the other Domitius, not Cicero's friend Ahenobarbus) had
agreed to pay the sitting consuls 400,000 sesterces if elected, to
be forfeit unless the consuls could produce false witnesses for a
\textit{lex curiata} that had never been passed and for a senatorial
decree drafted at a meeting that had never occurred. The detail
that the compact was inscribed not in vague words but in named
sureties and in many men's account-books is what gives the affair
its weight. Cicero tracks the wreckage in real time: the panic
among the candidates when the Senate orders a silent verdict before
the vote; the appeal to the tribunes; the tribunician veto by
Terentius; the consuls' feeble handling; his own outburst on the
Senate floor (``Abdera!''); the Gabinius-like court manoeuvres
around the younger Scaurus, freshly acquitted thanks to Cicero's
defence of the elder; the prospect of three more candidates being
indicted; and beneath all of it the awareness that Caesar in
Britain is the political weather. The letter closes on the great
building programme Cicero is putting up on Caesar's behalf with
Oppius --- Paulus's basilica, the marble \textit{saepta} in the
Campus, the expansion of the Forum to the Atrium of Liberty, all
for sixty million sesterces --- and on a glance at the trials and at
the \textit{lustrum} that everyone has already given up on. The
manuscripts are corrupt in three places (the end of section 1, the
end of section 7, the name of the law at the very end); these are
marked in the translation.}
