% Ad Familiares 7.5 --- headnote
% ad-familiares-07-05, April 54 BC, Rome

\textit{Cicero to C. Iulius Caesar, imperator, written from
Rome in April 54 BC. The recommendation of C. Trebatius
Testa, the young jurist, to Caesar in Gaul --- the founding
letter of what becomes the Fam.\ 7 sequence on Trebatius's
years in the Gallic command. The letter is short and the
register the formal recommendation that, between intimates,
turns immediately into the affectionate fulsome.}

\textit{The body has the famous coincidence of \textsection 2:
while Cicero was discussing Trebatius with Balbus at his
house, a letter from Caesar arrived containing what amounted
to a request --- ``send me another whom I may dignify.'' Both
men raised their hands at the timeliness, ``not chance, but
divine.'' Trebatius was therefore sent both on Cicero's
initiative and by Caesar's own invitation.}

\textit{The closing pledge --- ``no one is more upright, no
one a better man, no one more modest'' (in plain Roman fashion,
not the old high-rhetorical phrase Caesar had teased him about
in the Milo case) --- and the technical compliment that
Trebatius ``leads the front rank in civil law with a singular
memory and the highest knowledge'' --- mark the recommendation
as a serious one. The Roman handshake idiom, \textit{de manu in
manum}, ``from hand to hand,'' completes it. The closing
\textit{cura ut valeas et me ut amas ama} is the soft closure
of intimacy.}

\textit{The whole sequence --- this letter, then \textit{Fam.}
7.6, 7.7, 7.8, 7.9, 7.10, 7.13, 7.14, 7.15, 7.16, 7.17, 7.18,
running from April 54 BC into 53 BC --- forms one of the
sustained light-comedy strands of the correspondence: the
urbane Roman jurist transposed into the camp on the edge of
the world, the joke of the man-of-Roman-law hunted by
British essedarii. Trebatius, against the joke, did
distinguish himself; he survived to be one of the great
jurists of the Augustan age, the dedicatee of the
\textit{Topica}.}
