% Ad Quintum Fratrem 3.4 --- headnote
% ad-quintum-fratrem-03-04, 24 October 54 BC, Rome

\textit{Cicero to Quintus, written from Rome on the ninth day
before the Kalends of November (24 October 54 BC), on the day
Cicero set out for Tusculum with his son for the start of the
ludi. The letter's first three sections are the running
report on the \textit{maiestas} trial of Gabinius, which had
just ended in acquittal; \textsection\textsection 4--5 are
the literary and library postscript; \textsection 6 the
closing news about Pomptinus's contested triumph.}

\textit{Gabinius's acquittal on the \textit{maiestas} charge
was the bitter conclusion of a half-year campaign that the
last several Q. fr. letters have tracked --- the rumour of
his approach, his night-entry to Rome ``as into a city full
of enemies,'' the universal hatred at his appearance before
the people, the divinatio that named L. Lentulus the elder
as accuser. Cicero gave the principal witness-testimony.
The acquittal followed: thirty-two votes to convict against
seventy to acquit, despite the incoherence of Lentulus and
his subscribers and the squalor of the jury. The decisive
weight, Cicero says, was three: the intensity of effort by
the defence; the prayers (\textit{preces}) of Pompey, whose
political stake in Gabinius's safety was direct; and the
rumour of an impending dictatorship, which the Senate
``feared.'' Even so the verdict carries a heavy report behind
it, and Gabinius's coming trial for extortion (where Cicero
himself would in fact be conscripted to defend him in 54/53,
\textit{Pro Rabirio Postumo} being the surviving relict of
that affair) seems certain to convict.}

\textit{The candid moment at \textsection 2 is Cicero's
private accounting of why he had not prosecuted Gabinius
himself: the contest would have been one with Pompey, not
with the defendant, and Cicero's standing then was such that
he could not undertake to be matched (the gladiatorial
metaphor) against Pacideianus's most famous opponent for
Pompey's sake. Sallustius --- the historian's namesake,
probably C. Sallustius Crispus the future historian as a
young man in Cicero's circle --- had been pressing him to
prosecute one of the two recent triumvirate-clients (Gabinius
or Vatinius) and yield the other to Pompey; the rebuke at
\textsection 3 (``a charming friend Sallustius is, to think
that I ought either to have taken on dangerous enmities or
undergone everlasting disgrace'') is the only known stick
Cicero took to him. The defendant's parting word --- that, if
he had remained in the State, he would have given Cicero
satisfaction --- is what Cicero recovers from the ordeal.}

\textit{The closing sections are domestic and literary. The
\textit{[Greek: enthousiasmos]} (\textsection 4) for the
poetry Quintus has been asking after is absent: the year is
anxious, even if not fearful. The library notes
(\textsection 5) introduce Chrysippus and Tyrannio --- the
freedmen and scholars who handled Cicero's book-procurement
--- as the channels through which Quintus's Greek collection
in Gaul will be brought up to strength. \textsection 6 is
the report on Pomptinus's triumph: C. Pomptinus, praetor of
63 BC who had defeated the Allobroges in the year of Cicero's
consulship, had been hovering outside the city for years
seeking the triumph, and would finally hold it on 3 November
54, against the threatened veto of the praetors Cato and
Servilius (and Q. Scaevola the tribune of the plebs, ``breathing
Ares,'' \textit{[Greek: Ar\=e pne\=on]}, in the Homeric
phrase). The first private notice of P. Crassus's death in
Parthia would arrive in Cicero's circle in the days following
this letter; here it is not yet known, the rumour of a
dictatorship is still the principal alarm.}
