% Ad Familiares 5.8 --- headnote
% ad-familiares-05-08, January 54 BC, Rome

\textit{Cicero to M. Licinius Crassus, son of Publius, written
from Rome in January 54 BC. The recipient is the triumvir
himself, second consul of 70 and 55 BC, by this date already
en route to his Syrian command for the Parthian campaign that
would end fifteen months later in his death at Carrhae (May
53 BC). The letter is not the routine recommendation of
\textit{Fam.} 5.5 (62 BC), but the formal repair of a
political relationship: the public profession of an alliance
that has, until now, been at best uneven.}

\textit{The chain of recent events is there in the
\textsection 1 phrase ``perpetual championship of all your
honours.'' Crassus's consulship of 55 BC --- the second
consulship of the Luca compact, which Cicero had returned
from his Antium retreat to support after Luca --- had been
followed by the senatorial debates over the Syrian command
and its prerogatives, in which Cicero (against the consuls of
54, Ap. Claudius and L. Domitius Ahenobarbus, the latter
Crassus's bitter enemy) defended Crassus's settlement. The
letter records, after the fact, the change of side that the
post-Luca compact had required. The candour of the
\textsection 2 admission ---``certain plagues of men, those
who grieve at another's praise, sometimes turned you against
me and sometimes changed me towards you'' --- is the only
direct acknowledgement Cicero ever wrote to Crassus of the
years of mutual coolness.}

\textit{The structural turn at \textsection 4--5 is to the
two Crassi --- M. Crassus the elder son (the future consul of
30 BC) and the famously brilliant younger son P. Crassus, who
had served in Caesar's Gallic command and would die in the
Parthian campaign with his father. Cicero's particular bond
is with Publius (\textsection 4: ``observes and loves me as a
second parent'' --- the line on which much of Cicero's
literary judgement of the boy as the heir-apparent of
republican \textit{ingenium} has rested). Tertulla,
Crassus's wife, and the boys are the connecting persons; the
political alliance is mediated through them.}

\textit{The closing note --- ``this letter has the force of a
treaty, not of a letter'' --- is the formal commitment, by
Cicero, to act as Crassus's standing agent in absentia, on
all matters public, private, forensic, domestic, of friends,
guests, and clients. The letter would in fact prove the
sincere foundation of Cicero's late friendship with the
younger Crassus and, after Carrhae, of his careful working
relations with the family. With Crassus dead at Carrhae and
the triumvirate broken, the alliance Cicero pledged here would
turn within five years into the political balance that he
would in turn lose: when Caesar crossed the Rubicon (January
49 BC), the senate had no third triumvir to lean on.}
