% Ad Quintum Fratrem 2.10 --- headnote
% ad-quintum-fratrem-02-10, 13 February 54 BC, written at Rome

\textit{Marcus to Quintus, written at Rome on the Ides of February
54 BC. (Older editions number the letter 2.9; the Teubner
numbering is 2.10, sometimes 2.11.) Quintus has by this point left
Italy: he will spend the spring and summer of the year as a
legate on Caesar's staff in Gaul, the engagement that will keep
the brothers writing to each other across the campaigning season
and around which much of Marcus's correspondence in this year
turns. The Perseus dateline carries 13 February --- two days
after the bribery decree reported in 2.7 --- and the substantive
news here is the next sitting: a thin Senate convened by the
consul Appius Claudius Pulcher, broken up by a roaring crowd in
freezing weather before any decree could be passed.}

\textit{The body of the letter is the Commagene business and the
Caesar business. Antiochus I of Commagene had asked the Senate
to confirm the honours --- including the right to wear the
\textit{toga praetexta} --- granted him in Caesar's consulship of
59 along with title to the small fortress of Zeugma on the
Euphrates. Cicero, irritated and amused, broke up the proposal
piece by piece in open Senate, getting laughs at the Bostrene
analogy and the praetexta-patching joke and reducing the king's
case to ridicule; Appius, the consul whose business it was to
shepherd the decree through, is now suing Cicero with flattery
to drop the obstruction. The verse Cicero quotes ---
\textit{Iovis Hospitalis fidem, Graios omnis convocet}, a
fragment of Roman tragedy --- is half-mock-threat, half-allusion
to the reconciliation Pompey had brokered between him and
Appius. The closing movement is Caesar in Gaul: a previous packet
of letters had arrived from the front so water-soaked that
Caesar could not even tell Cicero had written; the joke about
his own poverty and Cicero's pretended caution about the family
strong-box is the warm political theatre of the moment when the
two men were on cordial terms. The whole letter is Cicero in
high holiday: hurried, gossipy, pleased with his own performance
in a freezing Curia, certain his brother on Caesar's staff will
relish the news.}
