% Ad Atticum 4.18 --- headnote
% ad-atticum-04-18, ca. 15 October 54 BC, Rome

\textit{Cicero to Atticus, written from Rome in October 54 BC ---
the Perseus dateline gives ``end of October 700 (54).'' Atticus is
still in the East and has been there longer than Cicero wished; the
letter opens, after a lacuna in the manuscripts, in the middle of
Cicero's account of his own conduct in court. Two trials hang over
the letter. The first --- almost certainly that of Drusus, or possibly
Vatinius --- has just ended in a near-split verdict (thirty-two for
conviction, thirty-eight for acquittal) brought in by jurors Cicero
calls squalid, against prosecution he calls incompetent, and only
after extraordinary pressure from Pompey. The second is the trial of
Gabinius, the ex-consul of 58 BC and one of the engineers of
Cicero's exile, whom Cicero is now reluctantly defending at Pompey's
insistence: Gabinius has just been acquitted on a charge of treason,
even as the courts run hot on the new bribery laws and a freedman of
his is condemned within the hour under the Lex Papia. The Greek
\textit{[transliterated as porpapymna]} in section 1 is a textual
crux; the manuscripts have lost the noun.}

\textit{Section 2 is the heart of the letter: a long, balanced piece
of self-examination addressed to Atticus by his old family name
``my dear Pomponius.'' The state has lost not only sap and blood but
colour and outward form; the Republic is unrecognizable; the man who
saved her once is now ornamental. Cicero's response is a deliberate
retreat into private life --- letters, oratory, the house, the
country places, the company of brother and friend --- and an unbroken
recollection of the height from which he fell answered, in
characteristic Ciceronian antithesis, by the height to which he
rose. The Greek verb \textit{[Greek: emphilosophēsai]} ``to
philosophize with [you]'' marks the move: with Quintus back and
Atticus returned, the new posture is philosophic withdrawal. The
political bulletin that fills sections 3 and 4 (electoral standstill,
the whiff of a dictatorship, Pomptinus's contested triumph, Appius
planning to leave for Cilicia without a law) is precisely the kind
of currency the philosophic withdrawal claims to have stopped
caring about, and goes on noting move by move all the same. Section
5 closes with seventeen letters arriving at once from Quintus and
Caesar off the coast of Britain --- the second invasion finished,
hostages but no plunder --- and the standing demand for Atticus to
come home.}
