% Ad Quintum Fratrem 3.5 --- headnote
% ad-quintum-fratrem-03-05, end October or early November 54 BC, Tusculum

\textit{Cicero to Quintus, written from Tusculum at the close
of October or the beginning of November 54 BC. The letter
that the older numbering presents as Q. fr. 3.5 is, in modern
editions (Shackleton Bailey and the Teubner), the combined
letter Q. fr. 3.5+6 or 3.5b; the seven sections preserved
here are the surviving body of what was, in the Perseus
text, two letters run together. The piece is the most
important single document for the composition of
\textit{De Republica}, and one of the most candid private
moments of Cicero's literary career: the brother-correspondence
hinge on which a draft was scrapped and the work as we know
it was begun.}

\textit{\textsection\textsection 1--2 are the famous narrative.
At Cumae in the summer Cicero had begun a dialogue on the
constitution of the State, structured on the nine-day
holidays (\textit{novendiales feriae}) of the year of
Tuditanus and Aquilius (129 BC), with Scipio Africanus the
Younger speaking shortly before his death to Laelius, Philus,
Manilius, P. Rutilius Rufus, Q. Tubero, and Fannius and Q.
Mucius Scaevola (the latter two Laelius's sons-in-law); the
work was to be distributed over nine days and nine books, on
the best constitution and the best citizen. Two books had
been finished. But when those books were being read aloud at
Tusculum with Sallustius listening, Sallustius advised that
the matter could be spoken on with much greater authority if
Cicero spoke in his own person, especially since he was not
``a Heraclides of Pontus but a consular and a man who had
been involved in the greatest affairs of state'' --- a
distinct compliment from a sharp listener, and the moment at
which the conversational dialogue of the Cumaean draft began
to be reshaped into the work that came down to us in
fragments. The reservation Cicero attaches at \textsection 2
is the one objection he could not get around in the new
plan: he could not touch ``the greatest movements of our
State'' (the consulate of 63, the exile, the return) since
they were later than the lifetimes of his speakers; and he
had been pursuing precisely that effacement, ``so as not, by
running into our own times, to give offence to anybody.''
The compromise --- speak in propria persona to Quintus
directly, by way of a dedication --- was the final shape of
\textit{De Republica}'s six books as published in 51.}

\textit{The other movements of the letter are characteristic
of the autumn 54 register. \textsection 3 is the Caesar note:
the Gallic dispatches arriving in October contained
remarkable declarations of love (\textit{Caesaris amor}); the
promises annexed to them Cicero takes more cautiously
(\textit{non valde pendeo}), ``I do not thirst for honours nor
do I yearn for glory.'' \textsection 4, the densest paragraph
of the letter, is the broken cry under the polished surface
of the year: ``I am tormented, my sweetest brother, tormented
at there being no commonwealth, no courts'' --- and the
beautiful Homeric self-quotation (Iliad 6.208), the line
Glaucus's father had given him at his arming, has gone down
in him: ``\textit{[Greek: Pollon aristeuein kai hypeirochon
emmenai all\=on]} --- always to be best and pre-eminent above
the others.'' The final clause is the most affectionate
single sentence in the brother-correspondence: only Caesar,
out of all men, has been found to love him as much as he
could wish --- or, as others say, who would want to.}

\textit{\textsection 5 records the political moment of the
year: had Cicero defended Gabinius on Pansa's urging, he would
have been ``finished'' (\textit{concidissem}), because all the
orders that hated Gabinius would have transferred their
hatred to his defender. The fact that Cicero in fact escaped
the request and only gave testimony is what the previous
two letters (Q. fr. 3.2, 3.4) have already established;
here, with Sallustius's reproof of \textit{Q. fr.} 3.4
already received, Cicero adds the obverse note --- defending
would have been worse. The library note follows on Tyrannio
(``dilatory'') and Chrysippus; the closing \textsection 7 is
the affectionate flourish on Quintus's four tragedies dashed
off in sixteen days (Electra, Troilus, and Erigone, with a
fourth not named): ``do you really ask after \textit{[Greek:
pathos]} when you have written an Electra and a Troilus?''
The closing \textit{[Greek: gn\=othi seauton]} (``know
thyself'') is twisted on the way out: it is not only about
diminishing arrogance, Cicero says, but about knowing our own
good things, too. The line is the period's most concentrated
expression of the two brothers' working relation.}
