% Ad Quintum Fratrem 2.14 --- headnote
% ad-quintum-fratrem-02-14, late July 54 BC, written at Rome

\textit{Marcus to Quintus, written at Rome around 27 July 54 BC.
Quintus is in Gaul on Caesar's staff --- the proconsul's
legate, posted there since the spring --- and is weighing
whether to come home at the date he had named or stay on
with the army to keep clearing his debts. The letter opens
with a self-deprecating joke about Marcus's notoriously
illegible last note: nothing was wrong, he says, he simply
uses whatever pen comes to hand. The brother had taken the
scrawl as a sign of distraction or anger and asked. The
warmth of the deflection is the warmth of the relationship.}

\textit{The substance is a political reading and a piece of
financial intelligence. On the political side, Marcus
forecasts the year ahead as calm, or at worst easily
defended: the daily crowds about his house, the noise in the
forum, the demonstrations at the theatre all favour him,
and his standing with both Caesar and Pompey covers him
against any sudden \textit{furor amentis hominis} ---
unnamed but transparently Clodius. He therefore advises
Quintus to do what is best for his estate rather than rush
back for fraternity's sake; the
\textit{\textit{[Greek: amphilaphian]}} --- the abundance
of money that staff service under Caesar was now generating
--- is worth the wait. The closing paragraph is the news
from the consular and tribunician elections: a corrupt
syndicate of Memmius, the sitting consuls, and Domitius
has driven the interest rate from a third to two-thirds
per cent on the Ides of July, with bids of ten million
sesterces being floated to be announced at the
\textit{praerogativa}; Scaurus alone is said to outbid
them, Messala has wilted. The single bright spot is the
extraordinary pact among the tribunician candidates: each
has deposited 500,000 sesterces with Cato as surety to
canvass under his terms, on penalty of his judgement. The
remark that, if it holds, Cato will have done more than
all the laws and all the courts is the heart of the
letter.}
