% Ad Familiares 1.9 --- headnote
% ad-familiares-01-09, mid-December 54 BC, written at Rome

\textit{Cicero to P. Cornelius Lentulus Spinther, by now
\textit{imperator} in Cilicia, written at Rome in mid-December
54 BC. The longest and most rhetorically laboured of the
Lentulus correspondence and one of the central political
documents of the post-exile years: a full-dress apologia,
delivered in the form of a private letter, for Cicero's
recent and visible alignment with Caesar and Pompey. Lentulus
had been the consul of 57 who carried Cicero's recall; the
Egyptian commission of 56 had been removed from him by the
optimate faction; by 54 the Luca conference had reset Roman
politics and Lentulus's own provincial command was about to be
broken short by Appius Claudius. The letter responds to a
Lentulus letter (now lost) that had asked, gently but
pointedly, why Cicero was now on good terms with Caesar, with
Appius, and even with Vatinius --- why he had defended
Vatinius, why he had praised Caesar in \textit{De Provinciis
Consularibus}, why he had defended Cornelius Balbus on the
strength of Caesar's and Pompey's commendations.}

\textit{The architecture of the answer is the architecture of
a forensic narrative. Sections 1--3 are exordium: the bond
between writer and recipient, the regret that they have not
been together to face these things side by side. Sections
4--10 are the \textit{narratio} proper: how Cicero had returned
from exile resolved to keep his old course; how the
optimates' coolness over his monument, the failed restitution
of his estate, and the violence he had suffered showed him
that his protectors were not what they had been; how
nevertheless he held to his old motions, attacked Vatinius
under questioning, censured the gift of kingdoms, brought the
Campanian land back into question on the Nones of April ---
and how that motion drove Pompey to Caesar at Luca, drew the
celebrated rebuke through Quintus in Sardinia (``\textit{te
ipsum cupio}''), and forced the reckoning. Sections 11--18
are the apology proper: Pompey was \textit{princeps}; Cicero's
debt to him was overwhelming; the old friendship with Caesar
was real; the Platonic precept --- the kind of leading men a
commonwealth has determines the kind of citizens --- counselled
not opposition but consent; and the good men themselves had
deserted the position. The Apelles simile in \S 15 is the
diagnostic image of the whole letter: my saviours finished my
head and left the body rough. Sections 19--20 dispatch the
specific cases of Vatinius and Crassus, with the famous
Terentian quotation from the \textit{Eunuchus} as the
intellectual frame for the Vatinius defence. The peroration
in \S 21 produces the navigator-simile (``\textit{tempestati
obsequi}'') and the great political formula \textit{cum
dignitate otium} --- tranquillity with dignity --- as the
constant aim that justifies a changed course.}

\textit{Beneath the rhetoric is a precise debt of loyalty.
Lentulus is the one optimate consul of 57 to whom Cicero
remained bound by an unrepayable kindness; he is also the
optimate now most damaged by the new disposition Cicero is
defending. The whole letter is an attempt to keep the man who
restored him as a friend while telling him that the cause they
shared has been deserted by its own. The closing four
sections (23--26) drop register: a list of the works Cicero
has written since Lentulus left (\textit{De Oratore} in three
books, three books of \textit{De Temporibus Suis} which he is
afraid to publish), a note on Quintus's estate-business,
sober and rather grim advice on the Appius Claudius succession
question (better to yield without delay than to defend
oneself against the suspicion of \textit{cupiditas}), and a
postscript about the \textit{publicani}. Shackleton Bailey
fixes the letter to mid-December 54 BC; the Perseus dateline
gives ``\textit{mense Dec. a. 700}.'' The cura annonae has
made Pompey indispensable, Crassus is east toward Parthia, and
Cicero is writing, in the voice of the man who had to bend,
to the man whose consulship had set him on his feet.}
