% Ad Quintum Fratrem 2.15 --- headnote
% ad-quintum-fratrem-02-15, late August 54 BC, written at Rome

\textit{Marcus to Quintus, written at Rome at the end of Sextilis
(late August) 54 BC, in the worst of the summer heats. Quintus is
on Caesar's staff in Gaul, on the eve of (or just back from) the
second crossing to Britain. The opening is Marcus's wry running
joke about secretarial labour: if the hand is the scribe's he has
had no leisure at all, if his own he has had a little. He has never
been more harried by cases and trials; but since Quintus and Caesar
have between them charted the line, he will follow it --- and even
seek out the goodwill of the very men who resent his new closeness
to Caesar. The political note is bleak: the consular candidates
for 53 are so brazen in bribery that the senate's debate has become
a spectacle Marcus will not lend himself to without backing.}

\textit{The middle of the letter is courtroom news and a brotherly
literary tease. Drusus has been acquitted of \textit{praevaricatio}
(collusive prosecution) by the tribuni aerarii by four votes, with
the senators and knights for conviction; Marcus is defending
Vatinius the same afternoon; Scaurus's trial is up next. Then the
turn to Quintus's Britannic letter: Marcus had been afraid of the
Ocean and the alien shore, and is anxious in the expectation of
news rather than in the dread of it; Quintus has --- as the Greek
puts it --- a magnificent \textit{hypothesis} for writing, with
the commander himself for a subject, and Marcus will send the verses
asked for, ``an owl to Athens.'' The close is at once the most
private and the most exposing thing in the letter: Caesar has
written that the first book of Marcus's poem is finer than any
Greek, but that the rest, down to a certain point, is
\textit{rhathumotera}, ``rather slack.'' Marcus asks Quintus to
report frankly what Caesar really thinks of subject and style ---
he will not, he promises, love himself a hair the less for it.}
