% Ad Atticum 4.19 --- headnote
% ad-atticum-04-19, late November 54 BC, Rome

\textit{Cicero to Atticus, written from Rome at the end of November
54 BC, in answer to a letter just arrived announcing Atticus's safe
crossing back from Epirus. The opening is pure performance: four
exclamations in a single rush, capped with relief that the voyage
went better than the previous one (the corrupt Greek word
\textit{derreis} preserves Cicero's nautical worry, though what
exactly he remembered fearing is unrecoverable). The pun on Atticus's
``Latin Atticism'' is the same affectionate teasing that runs
through every letter of this year. Behind the high spirits, the
political news is grim: open bribery, tribe by tribe, in the
run-up to the elections; Gabinius acquitted on his return from
Syria; talk of a dictatorship. Cicero registers it all in a single
sentence that the manuscript has badly mangled --- two obeli mark
passages no editor has restored with confidence.}

\textit{Section 2 turns to the consolation. Cicero's ``evenness of
mind'' in the face of all this is real, and rests on one plank:
his connection with Caesar, who is treating Quintus magnificently
in Gaul. The metaphor --- \textit{una ex hoc naufragio tabula} ---
is the letter's emotional centre; the shipwreck is the republic,
the plank is Caesar's friendship, and Cicero is candid enough to
say so to Atticus. The other news, dropped almost in passing, is
that he himself has been made a legate to Pompey and will be
formally outside the city from the Ides of January --- a useful
fiction. Closing on Dionysius the tutor, and the urgent
request that Atticus and his household stay with him the day he
arrives.}
