% Ad Quintum Fratrem 3.1 --- headnote
% ad-quintum-fratrem-03-01, September 54 BC, Arpinum and Rome

\textit{Cicero to Quintus, written largely from Arpinum and
Rome over the course of mid-September 54 BC. The letter is
substantial --- twenty-five sections --- and was put together
in stages over several days, with the couriers held up at
Rome (\textsection 23: ``as I have had the letter on hand for
several days on account of the couriers' delay, on that
account many things have been gathered, one at one time,
another at another''). Quintus is in Caesar's Gallic command
as a legate, and through the summer had been with Caesar in
Britain (the second British expedition); he is now back on
the Continent.}

\textit{The letter falls into two parts. The first six
sections are a tour of Quintus's country properties: Arcanum,
the Manilian villa with Diphilus the dilatory architect, the
new Fufidian estate (101,000 sesterces) with its summer shade
and water in plenty, the Lateran property with its public-work
road, the Bovillan estate, and Nicephorus the overseer's
account of the philosopher-woman villa that ``scolds the
madness of the others.'' \textsection 5's topiary --- ``the
cloaked Greeks themselves seem to be doing topiary and to be
selling ivy'' --- and the cool mossy dressing-room close the
estate-tour. The mode is the late-Republican landed gentleman's
correspondence on works in progress, intimate and competent.}

\textit{From \textsection 7 the letter turns to Roman public
business, organised as point-by-point replies to four letters
of Quintus's that arrived at staggered times. The chief
material: the absent boy Cicero (whose education Cicero is
overseeing, with the rhetor Paeonius?); Caesar's growing
``love'' for Quintus, the running theme of the year; the
tribunate sought for Curtius (granted by name, with Caesar's
gentle reproof of Cicero's modesty in asking for so little);
the Pro Scauro and Pro Plancio finished, the poem to Caesar
abandoned for now; the resigned ``but when shall we live?''
(\textsection 12) under Quintus's repeated urgings to
canvassing. The Erigona of \textsection 13 is Quintus's verse
tragedy, sent for criticism. Two political reports are the
heaviest material: Gabinius's homecoming after the Syrian
governorship, with three factions prosecuting him on
\textit{maiestas} and the universal hatred of the people
nearly laying him low at the praetor's tribunal
(\textsection\textsection 15, 24); and the consular elections
of 53 BC, with the bribery-coalition Cicero refused to join
(\textsection 16). Pompey's pressure for a reconciliation
between Cicero and Clodius is rebuffed: ``so long as I shall
hold any part of my freedom'' (\textsection 15).}

\textit{The most poignant thing in the letter is the news of
Caesar's grief. Julia, Caesar's daughter and Pompey's wife,
had died in late August or early September of complications
in childbirth (the infant son also died within days). The
news reaches Cicero through Caesar's own letter on 21
September (the moment is preserved at \textsection 17: ``how
distressed I am! How much I grieved at Caesar's most charming
letter; but the more charming it was, the more grief did that
misfortune of his bring''). Cicero does not write back even by
way of congratulation on the British campaign, because of
Caesar's bereavement. The political consequences would be
enormous: the family bond between Caesar and Pompey, the only
private hinge of the triumvirate after Crassus's departure for
Syria, was broken. Carrhae would follow within seven months.}

\textit{The final flourish, at \textsection 24, is the planned
\textit{embolium} (mythological digression) in the second book
of \textit{De Temporibus Suis}, Cicero's now-lost autobiographical
poem in three books: Apollo in the council of the gods will
speak of the homecoming of two imperators, ``of whom the one
had lost his army, the other had sold his'' --- Gabinius (Syria,
who had let Crassus's expedition begin in disrepair) and Piso
(Macedonia, the war-imperium converted to plunder). The poem
itself does not survive; this passage is the principal
testimony for what its second book contained.}
