% Ad Quintum Fratrem 3.7 --- headnote
% ad-quintum-fratrem-03-07, late October or early November 54 BC, Tusculum

\textit{Cicero to Quintus, written from Tusculum at the close
of October or beginning of November 54 BC. A very short note,
two sections only --- the briefest letter of the year --- and
written before dawn by the light of a little wooden lampstand
that Quintus had had made for his brother on Samos.}

\textit{The body of \textsection 1 is the flood. The Tiber
broke in great violence; Crassipes's promenade (Crassipes was
Cicero's son-in-law by the marriage of Tullia of 56 BC) was
carried off, with the gardens and many shops, the water
running all the way up to the Piscina Publica near the Porta
Capena. The text at this point is corrupt --- the manuscripts'
\textit{et Appia ad Martis mira luvies} is the standard
restoration: ``and on the Appian Way, by the temple of Mars
[just outside the Porta Capena], a wonderful flood.'' Cicero
reaches for two Homeric tags to make the moral. The first
(\textit{Iliad} 16.385--386) is a famous simile of Zeus
loosing his autumn rains on men who have angered him; the
second (16.387--388) is the gloss on the moral cause, the
ones who in the assembly give crooked judgments and drive
out right. The fit, Cicero says drily, is the acquittal of
Gabinius, which had just been delivered (compare Q. fr. 3.4)
under Pompey's pressure: heaven and the city have rendered
their respective verdicts on the same day.}

\textit{The closing sentences are the small domestic
counter-rhythm of the political letters of 54: the resolution
to ``not care about these things,'' the plan to write fuller
news about the dictatorship from Rome, the letters promised to
Labienus and Ligurius (both with Quintus in Gaul), and the
detail of the lampstand --- a personal gift sent from Samos
which the brother had had made for him during the eastern
travels of an earlier year. The closing salutation, ``my
sweetest and best brother,'' is the period's stock formula
between the two men.}
