% Ad Quintum Fratrem 2.7 --- headnote
% ad-quintum-fratrem-02-07, mid-February 55 BC, written at Rome

\textit{Marcus to Quintus, written at Rome shortly after 11
February 55 BC. Quintus has just left Rome --- probably for
the country --- and Marcus reports the political week
overnight. The frame is the brother's compliment to a
\textit{liber} of Marcus's that has just appeared, almost
certainly the third book of \textit{De Temporibus Suis} --- the
poem on his exile and recall whose interlocutor is the muse
\textit{Vrania} and which closed with a speech of Jupiter on
the orator's vindication. The remark \textit{illa omnia mihi
magis scripsi quam ceteris} --- ``I wrote all that more for
myself than for the rest'' --- is one of the most candid
glosses Cicero ever gave on his own self-glorifying verses.}

\textit{The substantive news is the working of the new
consulships. Pompey and Crassus, consuls for the second time,
have taken Rome in hand; the senatorial fight Cicero had
waged in 56 is over. The little vignette of the night call
on Pompey, then the daytime escort of Crassus from the Senate
\textit{(consulem ex senatu domum reduxi)}, then the offer
to broker --- if Cicero will not stand in the way of some
profitable Clodian legation --- is the theatre of the new
politics. The young Publius Crassus, the host's son and
Cicero's admirer, is the silent witness; the Brogitaran
embassy and the ``free'' legation to Byzantium are the kinds
of money-rich provincial errands a tribune of 58 BC like
Clodius could expect to convert to coin.}

\textit{The closing paragraph is the second movement: the
Senate's bribery decree of 11 February in the formula
proposed by Afranius, with the consuls (Pompey and Crassus)
refusing to add the praetor-elect provision that would have
required sixty private days before assuming office --- the
provision Cato had pressed and they rejected outright. The
verdict at the end is sharp: \textit{tenent omnia idque ita
omnis intellegere volunt} --- ``they hold everything, and
they want everyone to know it.''}
