% Ad Familiares 7.1 --- headnote
% ad-familiares-07-01, October 55 BC, Rome

\textit{Cicero to M. Marius, written at Rome in October 55 BC,
just after the lavish dedication-games of Pompey's stone
theatre — the first permanent theatre at Rome, on the Campus
Martius. The literary friend at Pompeii, ill of body but happy
in his bedroom-window-cut-to-Stabiae, has stayed home from the
games. Cicero writes the letter that has stayed with this
friendship in literary history: a wry account of the games for
the man who was right not to come.}

\textit{The first half is the games themselves: Sp. Maecius
Tarpa's stage-plays (a forgivable sin against Marius's taste,
since the elderly Aesopus's voice failed him in the oath
\textit{si sciens fallo}), the six hundred mules in Accius's
\textit{Clytaemnestra}, the three thousand mixing-bowls in the
revival of the \textit{Equus Troianus}, the gladiatorial shows
on which (Pompey himself confessed) ``both his pains and his
oil were lost,'' and the dual-five-day beast-hunts. Cicero's
most famous comment closes section 3: on the last day's
elephants, ``a kind of pity followed, and the impression that
there is some kind of fellowship between that beast and the
human race'' --- the locus classicus of Roman compassion for
the elephants of the games (cf. Pliny \textit{NH} 8.21).}

\textit{The second half is the moral. Cicero almost burst
himself defending Caninius Gallus, a maiestas case; the law's
yoke, and the inability to say no when good friends ask, are
the present grievances. He looks forward to true leisure --- to
visit Marius's villa with him on the litter, and to learn from
the man who has long studied nothing else what it is to live as
a human being. The closing pun is Cicero at his lightest:
Marius half-invited him to write so as to not regret missing
the games, and now Marius will have to come to the games next
time --- because his absence cannot expect another such letter.}
