% Ad Atticum 4.11 --- headnote
% ad-atticum-04-11, late April or 1 May 55 BC, written at Cumae

\textit{Cicero to Atticus, written at Cumae at the end of
April --- the manifest assigns 1 May (the Kalends), the
Perseus dateline ``a. d. v. K. Mai.'' (27 April). Cicero is
still in the bay, reading literature with the philosopher
Dionysius and waiting for political news from Rome. Pompey
has reported that Crassus is expected at the Alban villa on
28 April, that the two consuls will then come together to
Rome to clear the publican accounts; and the question Cicero
puts is whether all this will happen in time for ``the
gladiators'' --- almost certainly the games for the dedication
of Pompey's theatre and the temple of Venus Victrix later
that year.}

\textit{The second paragraph is the Atticus joke at the heart
of the letter: of news, the Greek tag \textit{[Greek:
ouden]} --- nothing --- so write to me as to a curious man
what the first day, what the second, what the censors are
doing, what Appius (Appius Claudius Pulcher, censor of 50
BC), what \textit{your Appulean lady the people} --- a
mock-formal title, the Pulcra plebs of the Appian gens. The
Demetrius of Magnesia at the close is the standard handy
literate slave whom Cicero is sending Atticus's way precisely
so that there will always be a courier on hand to bring back
a return letter. Lucceius, the historian, is to receive the
copy of Cicero's book that Atticus has been holding (cf.
\textit{Att.} 4.6, the long letter on what Cicero wants
Lucceius to write of his consulship).}
