% Ad Familiares 1.1 --- headnote
% ad-familiares-01-01, 13 January 56 BC, written at Rome

\textit{Cicero to P. Cornelius Lentulus Spinther, proconsul of
Cilicia, written from Rome on the Ides of January (13 January)
56 BC. The first surviving letter to Lentulus and the opening
of the long Egyptian-question correspondence of Fam. book 1.
Lentulus had been the consul of 57 BC who carried Cicero's
recall from exile; in return Cicero was bound, as he says here,
in a debt that nothing he could do would discharge. The
question on the table is the restoration of Ptolemy XII Auletes
to the Egyptian throne. Lentulus, now in his Cilician province,
held a Senate decree (passed late 57 BC on his own motion as
consul) commissioning him to lead the king back. But by January
56 the optimate front against the assignment had hardened.
The tribune C. Cato had read out a Sibylline oracle ---
``if the king of Egypt come asking for help, do not refuse him
friendship, but do not aid him with a multitude'' --- which gave
the Senate a religious pretext to strip Lentulus of his army.
Hammonius, the Egyptian envoy, was buying senators openly;
Pompey's intimates Libo and Hypsaeus were running about lobbying
that the commission go to Pompey. Cicero's plain reading: those
who wish the king restored want Pompey to do it, the Senate
hides behind religion, Pompey's public speech for Lentulus is
contradicted by the actions of his entourage. The opinion he
proposes (with Hortensius and Lucullus) is the compromise that
saves Lentulus's standing while yielding the army: Lentulus
remains the named agent under the original decree; the king
is brought back without troops; the Sibyl's letter is
satisfied. The letter is the keynote of the whole 56 BC
sequence: the Egyptian question is the opening through which
the optimate-Pompeian alliance of the previous year breaks.}
