% Ad Familiares 5.12 --- headnote
% ad-familiares-05-12, June 56 BC, Antium (?)

\textit{To L. Lucceius, written probably from Antium in
June 56 BC. Lucceius is the optimate-leaning historian
who had run with Caesar for the consulship of 59 BC and
been beaten down by the optimate cash-bid for Bibulus;
his lost \textit{History of the Italic and Civil War}
(the Social War of 91--88 BC and the wars of Marius and
Sulla) was nearly finished, and he had told Cicero he was
beginning ``the rest'' --- the period after Sulla. The
letter is the famous request to write a separate
monograph on the Catilinarian conspiracy and Cicero's
exile and recovery: ``a kind of moderate-sized body''
running from late 64 BC to September 57.}

\textit{The letter is the most self-conscious surviving
specimen of Cicero's interest in his own historical
reputation. The opening apology for shamelessness ---
``a letter does not blush'' (\textit{epistula non
erubescit}, \textsection 1) --- pre-empts the principal
embarrassment, which is the request that Lucceius set
aside the laws of history (\textsection 3:
``\textit{leges historiae neglegas}'') and even go beyond
what truth permits in praising him. The two arguments are
that the matter is rich in dramatic shape (the
conspiracy and the recovery as a self-contained tragedy
of \textit{varios actus mutationesque}, \textsection 6),
and that the author of his praise must be one whose own
glory the praise will increase. The catalogue of
exempla in \textsection 7 is unusually concentrated:
Alexander chose Apelles and Lysippus for his own glory's
sake, Agesilaus refused all images but is known by
Xenophon's little book, Timoleon by Timaeus, Themistocles
by Herodotus, Achilles by Homer (the herald's call at
Sigeum), and Naevius's Hector who is glad to be praised,
``but by a praised man.''}

\textit{The closing fallback (\textsection 8) is that, if
Lucceius declines, Cicero will be driven to do what some
men reproach --- write the history himself; the figure of
the gymnastic-games herald who must call in another to
proclaim his own victory is the wittiest moment of the
letter. The Greek monograph (\textit{De Consiliis Suis}
or, in fragments, the lost Greek \textit{commentarius}
written ``in the manner of Theopompus'' that Cicero would
in fact send to Posidonius) is the survival of this
fallback. Lucceius did not, so far as we know, write the
monograph Cicero asked for; the labour of the request
itself is the document.}

\textit{The letter is the centerpiece of the early Lucceius
correspondence (\textit{Fam.} 5.12--5.15) and is the
classic statement of the ancient distinction between
\textit{historia} as truth and \textit{laudatio} as
embellishment --- in this case explicitly inviting the
historian to fold one into the other. The acid test of
the letter's seriousness is the closing line of
\textsection 10: if Lucceius takes the case, Cicero will
\textit{conficiam commentarios rerum omnium} ---- ``I will
draw up notes of all the events'' --- the explicit
\textit{commentarius} hand-off whose surviving Latin
shadow is presumably the \textit{De Consulatu Suo}.}
