% Ad Familiares 1.7 --- headnote
% ad-familiares-01-07, late July 56 BC, written at Rome

\textit{Cicero to Publius Cornelius Lentulus Spinther,
proconsul in Cilicia and Cyprus, written at Rome late in
July 56 BC. The Lentulus correspondence had opened in
January with the daily reports on the Senate's struggle over
the Egyptian commission (\textit{Fam.} 1.1--1.5b); it now
returns to that question, but in a transformed political
landscape. The Luca conference is past, the triumvirate has
re-asserted itself, and Cicero is writing as a man whose own
position has been forced to shift.}

\textit{The first three sections concern Cicero's role as
informant. Letters of this kind he dares not entrust to chance
carriers; among the consulars only Hortensius and Lucullus are
reliably friendly to Lentulus's cause; among the magistrates
of the year, only L. Racilius. Pompey, on the other hand, is
warmer than Lentulus may suspect; the recent letter Lentulus
sent him has been very welcome. The compliment to Lentulus's
\textit{humanitas vel summa potius sapientia} in writing it is
the gentlest rebuke for an earlier coolness.}

\textit{Sections 4--6 are the substantive political advice on
Egypt --- and they are, in effect, Pompey's counsel relayed
through Cicero. The legal position: no \textit{senatus
consultum} has stripped Lentulus of the commission to restore
Ptolemy XII Auletes; the obstructive resolution forbidding
restoration ``with a multitude'' was vetoed and represents
\textit{magis iratorum hominum studium quam constantis senatus
consilium}. Lentulus, holding Cilicia and Cyprus, can take
the king to Ptolemais and march on Alexandria with fleet and
army; once it is held with garrisons, the king can be
restored, and the Sibylline oracle is technically respected.
The whole calculation is governed by outcome (\S 5): success
will be praised, a stumble condemned. The fallback (\S 6) is
to give military aid only if the king has kept faith with
those of Lentulus's friends who have lent him money on the
provincial credit.}

\textit{Sections 7--10 are the post-Luca political dispatch.
The optimates are perverse: by envying Cicero rather than
favouring him in the common cause, they have nearly driven
him off the position he had so long held. Lentulus, the
most noble of men, has had the same vices of the envious
turned against him --- the optimates allowed him among the
\textit{principes}, but \textit{evolare altius certe
noluerunt}. The famous summary of the year's politics is in
\S 10: pay decreed for Caesar, ten legates, and the
suspension of the Sempronian succession --- everything the
Senate of the previous year had refused, the Senate of this
year passes ``with very few opposing.'' The lesson Cicero
draws from his own experience for Lentulus to learn from
his own affairs whole: \textit{neque salutis nostrae
rationem habendam nobis esse sine dignitate neque
dignitatis sine salute}.}

\textit{The closing section congratulates Lentulus on
Tullia's marriage to Crassipes (announced in \textit{Q. fr.}
2.4) and commends Lentulus's son, the young Lentulus, then
in his teens, with the hope that he will be educated chiefly
by the imitation of his father.}
