% Ad Atticum 4.6 --- headnote
% ad-atticum-04-06, April or May 56 BC, written at Antium

\textit{Cicero to Atticus, written at Antium in April or May
56 BC. The letter is built on two losses --- the death of
P. Cornelius Lentulus Niger, the flamen Martialis, and what
Cicero is now starting to admit is the loss of his political
freedom. The two are stitched together in the great consoling
formula of \textsection 1: Lentulus is fortunate, because he
loved his country and the gods snatched him out of the
\textit{incendium} before he had to watch any more of it.}

\textit{The lament of \textsection 2 is one of the most cited
passages in the correspondence: ``if I speak about the
commonwealth what is right, I am thought a madman; if what is
needful, a slave; if I am silent, crushed and captive.'' The
trichotomy is the political life of these months in three
clauses. The Greek tag from Euripides's \textit{Telephus} ---
\textit{[Greek: Spartan elaches, tautan kosmei]}, ``the Sparta
you have drawn, adorn it'' --- is the resignation that hardens
into program: take the lot you have been given and make
something of it. The exemplum of Philoxenus, who would rather
go back to the quarries than flatter Dionysius once more, is
the wish that runs against the resignation. The whole letter
is the conversation between the two.}

\textit{\textsection 3 turns to Atticus's standing request for
a literary attack on Hortensius (who had recently given Cicero
some grievance now lost to us). Cicero declines, with a
beautifully tortuous reason: to write the thing would only
make a small injury famous, and would invite the suspicion
that what looked like depth (\textit{[Greek: bathytes]}) in his
forbearance was only frivolity (\textit{levitas}) underneath.
\textsection 4 closes with the famous mention of the letter
to Lucceius (\textit{Fam.} 5.12), in which Cicero is asking
the historian to set down the consulship.}
