% Ad Quintum Fratrem 2.5 --- headnote
% ad-quintum-fratrem-02-05, 8 April 56 BC, en route to Anagnia

\textit{Cicero to Quintus, written before daylight on 8 April
56 BC at the start of the journey out from Rome that would
take him through T. Titius's house in the Anagnine country
to Laterium, then to Arpinum, the Pompeian and the Cuman
estates, returning to Rome the day before the Nones of May
for Milo's trial. Section 2 is missing in the manuscript
tradition; the letter survives in three sections, the loss
falling between the building-site visit (\textsection 1) and
the building-site account (\textsection 3).}

\textit{The news is the betrothal of Tullia to Furius Crassipes
(her second marriage --- her first husband, C. Calpurnius
Piso Frugi, had died in 57 BC during Cicero's exile). The
betrothal-feast was on 6 April; the engagement formal date
was 4 April. Three pieces of public news fill out the body:
the senatorial corn-decree of 5 April giving Pompey the
\textit{cura annonae} 40 million sesterces; the same-day
contio-like dispute over the Campanian land (the second great
political fault-line of 56 BC, which would force the
post-Luca about-face); and the casting-out of M. Furius
Flaccus from the college of the Capitolini and Mercuriales,
with the man lying at each member's feet in vain. The visit
to Pompey in his gardens, after dinner at Crassipes's, has
the practical errand --- get Quintus released from Sardinia
quickly --- and the practical reply: Pompey was to embark for
Sardinia on 11 April from Labro or Pisa.}

\textit{The chronology of the letter is one of its great
documentary uses: it is the latest fixed-date letter before
the conference at Luca (15--16 April), at which Caesar,
Pompey, and Crassus would reset the political terms. Cicero
in this letter knows nothing of any such meeting; Pompey is
simply ``going to Sardinia.'' The casual mention of the corn
authority and the route via Pisa places Pompey within fifty
miles of Luca; the Senate's recriminations on the Campanian
land are the very stalking-horse Caesar would use at Luca to
compel the post-Luca compact.}

\textit{The closing note --- ``come as soon as possible''
(\textit{quam primum venias}) --- belongs to the running theme
of Q. fr. book 2: Cicero needing his brother home, the
political climate having turned, and the family business
(Tullia's marriage, the boy's health, Pomponia's complaints,
two house-building sites) all wanting Quintus's hand on the
ground. He would return at midsummer, in time for the
post-Luca debate.}
