% Ad Quintum Fratrem 2.4 --- headnote
% ad-quintum-fratrem-02-04, mid-March 56 BC, written at Rome

\textit{Cicero to his brother Quintus, written from Rome in
March 56 BC, shortly after the acquittal of Sestius on the
fifth day before the Ides of March (11 March). The Pro
Sestio --- the speech that survives, with its great political
digression on the \textit{boni} and the \textit{tota Italia}
--- was given just before this letter. Vatinius, prosecution
witness, was cross-examined to pieces by Cicero in what
became the surviving \textit{In Vatinium}, and the gods and
men applauded. \textsection 2: domestic notes --- Tyrannio
teaches the young Quintus at Cicero's house; both brothers'
houses are being rebuilt; the engagement of Tullia (Cicero's
daughter) to Furius Crassipes appears to be settled. The
Greek tag \textit{amphilaphian} (``ample leisure / abundance'')
in \textsection 3 is a term Quintus used to use to Cicero in
better days. \textsection 4--5: the political news is the
new consul Lentulus Marcellinus, who has been blocking the
``most ruinous'' laws --- Cato's bills against Lentulus
Spinther and Milo, and the bills against Caesar --- and the
beautiful trick by which Milo bought up Cato's beast-fighter
gang via a third party (Racilius), then posted a notice that
he would sell ``the Catonian gang'' to general laughter.
\textsection 5--6: the contention between Pompey and the
Senate continues; Marcellinus handles Pompey ``too sharply,''
the Senate happily looks on; the trial of Sextus Clodius
went the wrong way by three votes (the senators acquitted,
the knights split, the \textit{tribuni aerarii} condemned).
\textsection 7: still no letter from Quintus since the one
from Olbia, only word from incoming travellers that he
stands very high in Sardinia.}
