% Ad Familiares 7.3 --- headnote
% ad-familiares-07-03, summer 46 BC, written at Rome (placeholder 56 BC)

\textit{Cicero to Marcus Marius, written at Rome before the
Kalends of September. The Perseus dateline records ``a. 708,''
which converts to 46 BC, not the manifest's 56 BC placeholder;
the content (Pompey defeated, the camp lost, the survivors
forced to Juba in Africa or to Caesar's hands) places the
letter unmistakably in the summer of 46 BC, after Pharsalus
and before the Thapsus aftermath was fully known. The
chronology will be corrected in the next manifest pass.}

\textit{The occasion is a retrospect of the war from the man
who saw it from the inside. The remembered evening is the day
before Cicero left his Pompeian villa to follow Pompey east
(10 May 49, in the consulship of Lentulus and Marcellus); the
addressee Marius is the same friend whose anxious mind that
evening Cicero now thanks. The frame is an apologia: he chose
\textit{honour and reputation} over safety, and once at the
camp found nothing good but the cause --- forces unfit, leaders
rapacious in the war and cruel in talk afterwards, all of them
swimming in debt. He urged peace, then a long war; Pompey
shifted, and after one engagement gained false confidence, and
\textit{from that moment that great man was no general at all},
\textit{vir ille summus nullus imperator fuit}.}

\textit{The closing pages are the famous Ciceronian post-war
self-justification: he wished no one man to overshadow the
state; when one did, he wished for peace; with the army lost
and the one hope dead, he wished to end the war for himself.
The classical line is the saying he calls old: \textit{ubi non
sis qui fueris, non esse cur velis vivere} --- ``when you are
not the man you were, there is no reason to want to live.''
The two consolations he keeps are knowledge of the noblest
pursuits and the renown of the greatest deeds, the second of
which not even death can take. The end is testy: a wordier
letter than Marius might have liked, but Marius is to send a
longer one back, and meanwhile Cicero has work to finish in
Rome before he can come south.}
