% Ad Quintum Fratrem 2.2 --- headnote
% ad-quintum-fratrem-02-02, 19 January 56 BC, written at Rome

\textit{Cicero to his brother Quintus, dictated --- not written
in his own hand, on account of an eye inflammation --- from
Rome on the fourteenth day before the Kalends of February
(19 January) 56 BC. Quintus is now in Sardinia, where Pompey
sent him as legate after the second consulship of 55 BC was
already in view. The letter braids the brothers' running
domestic correspondence (Numisius's villa design, the debts
owed to Atticus's brother Pomponius, the Tusculan property at
Culleo's auction, the building work at Rome and pressing the
contractor Cyrus) with the political dispatch. The political
half of the letter (\textsection 3) is the Quintus-version of
the Egyptian-question report Cicero is also sending to Lentulus
that same week (Fam. 1.1--1.4): the Senate decree that the king
must not be brought back ``with a multitude'' (the Sibyl-induced
no-army condition), the calumnies that have stalled the
Lentulus-vs-Pompey contention, the comitial days that close
the Senate, and the suspicion that the tribune Caninius will
carry the bill (transferring the commission to Pompey) by
violence. Cicero's gauge of Pompey is the same as in Fam. 1.2
--- the man himself opaque, his entourage transparent. The
\textit{xiv K. Feb.} dating both opens and closes the letter.
The augury anecdote of \textsection 1 is the elder Gracchus
in 162 BC, who while governing Sardinia recalled having held
his own consular elections against the auspices and reported
the irregularity to the Senate.}
