% Ad Atticum 4.8A --- headnote
% ad-atticum-04-08a, autumn 56 BC, written at Rome

\textit{Cicero to Atticus, written at Rome in the autumn of
56 BC, after the Luca conference and as the consular
elections for 55 are taking shape. Four short, hurried
sections of close-mouthed political gossip, the kind of
letter that cannot be entrusted to a doubtful carrier (compare
the cautious opening of \textit{Fam.} 1.7).}

\textit{Section 1 turns on a piece of news Atticus has just
sent: someone --- almost certainly Pompey or one of the
triumviral candidates --- is uncertain whether to bring forward
his law. Cicero asks for clarification. The added day to the
games will be spent at home with Dionysius (the freedman tutor).
Section 2 is the most quoted: the comparison of L. Domitius
Ahenobarbus to a fig --- ``no fig was ever so like a fig'' as
his case is to Cicero's own, in respect of: those bringing it,
the surprise, the absence of \textit{viri boni}; only one thing
unlike, that Domitius deserves it. Then the famous epigram on
his blocked candidacy: he has been consul-designate as many
years as he is old, has more pages of future consuls in his
calendar-tablets than of past ones --- and what is more wretched
than him, except the commonwealth, in which not even any better
is hoped?}

\textit{Section 3 chases the small mystery of Fabius Luscus,
who has been at Rome and has not visited Cicero --- a small
withdrawal, but it strikes (\textit{percussit animum}) because
Fabius had been a useful informant on the Firmian brothers.
Section 4 closes on Atticus's standing political advice:
\textit{[Greek: politik\=os me geram et t\=en es\=o
gramm\=en teneam]} --- bear yourself with political prudence
and keep the inner line. The chariot-racing metaphor of
hugging the post is the perfect image for the Cicero of the
post-Luca months: stay clear of the danger zone. Send word
about everything every day; and when there is nothing to send
word of, send that very thing.}
