% Ad Quintum Fratrem 2.3 --- headnote
% ad-quintum-fratrem-02-03, 12--15 February 56 BC, written at Rome

\textit{Cicero to his brother Quintus, written from Rome
between the day before the Ides of February (12 February) ---
the principal dispatch of \textsection 1--6 was finished
before dawn that day --- and the fifteenth day before the
Kalends of March (15 February), when the closing line was
added. The letter is the great Q.fr. companion-piece to
Fam. 1.5b: the same week of public riot and senatorial
scrambling, but in much fuller daily detail. \textsection 2 is
the famous report of the Milo-Pompey contio of 8 February,
the same scene Fam. 1.5b summarizes in two lines: Pompey
shouted down for two solid hours by Clodius's claque, then
Clodius rising to be roared down in turn by Cicero's side
(with lewd verses against Clodius and his sister Clodia, the
same Clodia who would be Caelius's prosecutor two months
later), and the famous chant in the gangs --- ``who is killing
the people with hunger? Pompey. Who wishes to go to Alexandria?
Pompey. Whom do you want to go? Crassus.'' Crassus stood by
``with no friendly mind towards Milo.'' \textsection 3 carries
the Senate of the next two days at the temple of Apollo: the
SC declaring the 8 February proceedings \textit{contra rem
publicam}, Cato's ``continuous speech'' indicting Pompey to
the great silence of his enemies, and Pompey's reply ---
naming Crassus, swearing that he would be better guarded
against Carbo than Africanus had been. \textsection 4 is the
single clearest narrative of how the senators-vs-Pompey
front of the autumn has, in three weeks, become an
optimate-Crassan partnership against Pompey: Crassus is
sustaining C. Cato; money is going to Clodius; Curio,
Bibulus, Favonius, young Servilius are tearing at Pompey
in the chamber. \textsection 5--6: the indictment of P.
Sestius for canvassing and for violence, the political
clubs (\textit{sodalitates}, \textit{decuriati}) ordered to
disperse, and Cicero's anticipatory advocacy for Sestius
worked into a Bestia speech (the Greek phrase
\textit{pro\=oikonom\=esam\=en} marks the rhetorical move).
The closing date, with the brother's wedding-dinner that
evening at Pomponius's, the Lentulan house secured by the
Lacus Pisonis, the Lamiae as Quintus's distinguished
tenants in the Carinae, and the gentle joke ``although it
is winter, still bear in mind that this is Sardinia.''}
