% Ad Atticum 4.5 --- headnote
% ad-atticum-04-05, April or May 56 BC, written at Antium

\textit{Cicero to Atticus, written at Antium in April or May
56 BC, the famous letter on the palinode. Atticus has
reproached him --- on two grounds, by the look of it: that he
sent the offending speech (almost certainly the
\textit{De Provinciis Consularibus}, or perhaps an earlier draft
of one of the post-Luca compositions) to a third party before
showing it to him; and behind that, that he wrote it at all.
The defence is awkward and not pretended to be otherwise:
\textit{subturpicula mihi videbatur esse palinodia} ---
``the palinode did seem to me a touch dishonourable.''}

\textit{What follows is the most candid account in the
correspondence of Cicero's break with the optimate front of
the year before. The men he had defended in the Sestius
trial, the men whose ideal he had set out in the
\textit{cum dignitate otium} digression, have abandoned him
once and for all: \textit{senseram noram inductus, relictus,
proiectus ab iis} --- ``I had sensed it, I had known it: led
on, abandoned, cast aside by them.'' He has therefore done
the prudent thing and committed himself in writing to the
new coniunctio, so that he cannot now slip back. The
optimates' true grievance is laid bare in one bitter
sentence: even when he votes their way they delight in
seeing him cross Pompey. Resolution: \textit{tempus est me
ipsum a me amari} --- it is time I started loving myself.}

\textit{The letter then turns abruptly to domestic news.
Crassipes (Tullia's new husband, the engagement reported in
\textit{Q. fr.} 2.4) is hosting the farewell dinner; Atticus
will come straight from the road to his gardens; Cicero will
visit him there the day after. And the closing image, which
links straight back to \textit{Att.} 4.4A: the slaves Atticus
sent have decorated the Antium library with their construction
and the title-labels (\textit{constructione et sillybis}); they
deserve to be praised.}
