% Ad Atticum 4.3 --- headnote
% ad-atticum-04-03, 26 November 57 BC, written at Rome

\textit{Cicero to Atticus, written from Rome on the sixth day
before the Kalends of December (26 November) 57 BC. The
single most violent of the surviving exile-aftermath letters,
a daily report on the Clodian assault. \textsection 2: on
3 November Clodius's gangs drove the workmen off the Palatine
site, broke up Catulus's portico (then nearly at roof level),
and set fire to Quintus's house. On 11 November Clodius
ambushed Cicero in person on the Sacra Via; on 12 November
his men assaulted Milo's house with shields and drawn swords
and torches, with P. Sulla's house as their forward base. The
Senate met the next day; Marcellinus delivered the opinion
that bound up the whole case --- the site, the fires, the
peril of Cicero --- into a single judgement to be settled
ahead of the consular elections.}

\textit{\textsection 4--5 are the dispatch on Milo's
counter-strategy: announcing against the heavens on every
comitial day to block the elections, holding the Campus
Martius from before midnight, overtaking Metellus among the
groves. The famous note in \textsection 5: ``Marcellus the
candidate was snoring so that I, his neighbour, could hear.
Clodius's vestibule was reported to me as quite empty: a few
rag-bedecked fellows, a linen lantern.'' Cicero's bet that
Milo will kill Clodius before any trial settles the matter
will hold for another five years and come true on the Appian
Way at Bovillae in January 52. The closing line is the
practical confession: family fortune broken, drawing on
friends' help against Quintus's protest, no plan possible
without Atticus.}
