% Ad Atticum 4.2 --- headnote
% ad-atticum-04-02, mid-October 57 BC, written at Rome

\textit{Cicero to Atticus, written from Rome in mid-October
57 BC --- the report on the success of \textit{De Domo Sua}.
\textsection 2--5 is the only ancient narrative we have of
the case's working through the Senate after the speech.
The pontiffs delivered, on the day after Cicero spoke, the
narrow ruling that he had asked for: that the dedication
could be set aside without religious offence, since the
dedicator had not been put in charge of the matter by name
through people or plebs. Clodius then went to the Forum and
declared the opposite at a public meeting Appius had given
him. The Senate the next day --- with all the pontiffs who
were senators present, Marcellinus speaking first, M. Lucullus
on behalf of the college (the pontiffs the judges of religion,
the Senate of law) --- voted in favour of restoration. Serranus
vetoed; Cornicinus threw off his toga and threw himself at
his son-in-law's feet (a play borrowed from the January
proceedings); a night was granted, and on the next day the
senatus consultum passed.}

\textit{The valuations granted to Cicero by the consuls'
council are visibly grudging --- two million sesterces for
the Palatine house, five hundred thousand for the Tusculan
villa, two hundred and fifty thousand for the Formian.
Cicero's reading: ``the same men who had clipped my wings
are unwilling that the same wings should grow back''
--- the optimate consulars led by Hortensius and the rest
who had let the Clodian bill stand. \textsection 6 reports
on the new shape of his year: he has accepted a Pompeian
legateship that lets him stand for office or leave Rome at
the start of summer, but for the moment will keep himself
in the people's sight. \textsection 7 is the domestic
strain --- the Palatine house being rebuilt, the Formian
being repaired, the Tusculan up for sale, friends' kindness
exhausted, and ``other things \textit{[Greek: more for
mysteries]}'' that he will not entrust to a letter.}
