% Ad Atticum 4.1 --- headnote
% ad-atticum-04-01, mid-September 57 BC, written at Rome

\textit{Cicero to Atticus, written from Rome in mid-September
57 BC --- the first letter of the post-exile correspondence,
opening Atticus book 4. The narrative \textsection 4--5 is
the journey home: setting out from Dyrrachium on the day
the centuriate assembly carried the recall (4 August),
arriving at Brundisium on the Nones (5 August), where Tullia
was waiting on her birthday --- the same day as the
foundation-day of the colony of Brundisium and of the
neighbouring temple of Salus. The reception at Rome on
4 September is the famous procession to the Capitol, every
order represented except the few enemies whom the law would
not let dissemble.}

\textit{\textsection 6 carries the news that has consumed the
following ten days. A grain shortage has been blamed on
Cicero by Clodius's crowd; Cicero has answered by moving in
the Senate that Pompey be given charge of the corn supply for
five years over the whole world. The consular drafting
(\textit{lex Cornelia Caecilia} of the consuls Cn. Cornelius
Lentulus Spinther and Q. Caecilius Metellus Nepos, on Cicero's
opinion) is modest; the more sweeping rival, the Messian
bill, would have given Pompey command of all moneys and a
fleet and army and greater \textit{imperium} in the provinces
than their existing holders --- the prefiguring of what
Pompey will get in 52 BC. The consulars under Favonius's lead
are murmuring against both. The other suspended question is
the rebuilding of the Palatine house: Clodius had consecrated
the site as a Temple of Liberty, and the pontiffs have not
yet ruled. \textsection 8 closes with the line that defines
the post-exile years: ``We are beginning a kind of second
life.''}
