% Ad Atticum 3.23 --- headnote
% ad-atticum-03-23, 29 November 58 BC, written at Dyrrachium

\textit{Cicero to Atticus, written at Dyrrachium on the day
before the Kalends of December (29 November) 58 BC --- a
detailed legal letter on the bill that the eight outgoing
tribunes have promulgated to recall him. Three letters from
Atticus have arrived together; Cicero answers each. Only the
first carries good news --- to wait with steady spirit for the
month of January, when Lentulus Spinther will enter on his
consulship and the new tribunes on their year. The second and
third turn out to be the technical reckoning on the bill
itself.}

\textit{The law as drafted by the eight tribunes
\textsection 2--4 has three clauses, and it is the third that
Cicero now thinks may sink it: a clause that says that what is
done in carrying the bill, contrary to the Clodian law against
him, shall not redound against the bill's promulgators. Cicero
sees the clause as a snare. The outgoing tribunes are
themselves protected by their college's collective immunity;
the clause was therefore unnecessary for them, but written in.
The new tribunes, more timid, will read it as a binding
instruction not to do anything that contravenes the Clodian
law --- which is to say, not to do the very thing the bill is
meant to do. Cicero asks Atticus to find out who introduced
the clause, and recommends instead the alternative draft that
Visellius wrote for T. Fadius. \textsection 5 is the alternative
ending: if the bill cannot be broken through in one rush, then
Atticus is to look after Quintus, Marcus, and Terentia. The
move from Dyrrachium to Epirus will follow once the first
news of the new year arrives.}
