% Ad Atticum 3.1 --- headnote
% ad-atticum-03-01, March or April 58 BC, written setting out into exile

\textit{Cicero to Atticus, written as he was setting out into
exile, March or April 58 BC. The shortest of all the
Atticus letters, and the first of the exile correspondence.
The political picture has by this point collapsed: P.
Clodius is tribune (entered office 10 December 59 BC), has
carried at the start of his year the lex Clodia de capite
civis Romani --- a bill threatening exile for any magistrate
who had put a Roman citizen to death without trial --- and
Cicero, the consul of 63 who had executed the Catilinarian
conspirators on the senatorial decree of 5 December, is
that bill's target.}

\textit{The ``bill'' Cicero refers to is the second Clodian
law, naming Cicero by name, banning him from fire and water
within four hundred miles of Italy. ``The law on the province
of Macedonia having been carried'' refers to Clodius's
gift-bill assigning Macedonia to L. Calpurnius Piso (the
consul of 58 BC, Caesar's father-in-law) --- which had the
practical effect of removing him from Rome and freeing
Atticus, owner of estates in Epirus, to come and meet Cicero
on the way. The route Cicero is contemplating is the same
overland route through Epirus that he in fact took: across
to Brundisium, across the strait, north along the coast to
Epirus or Macedonia, eventually to Thessalonica, where he
would spend most of his exile. The bill of 3.1 cleared the
last political obstacle to that route.}

\textit{The next surviving letters --- \textit{Att.} 3.2--3.27,
\textit{Q. fr.} 1.3--1.4, \textit{Fam.} 14.1--14.4 --- are the
exile correspondence of 58 BC, written from the journey,
from Thessalonica, and from Dyrrachium, and are the darkest
stretch of Cicero's surviving prose.}
