% Ad Atticum 2.19 --- headnote
% ad-atticum-02-19, mid-Quintilis (mid-July) 59 BC, written at Rome

\textit{Cicero to Atticus, mid-July 59 BC, from Rome. The
letter is the great document of the moment when the
triumvirate's standing collapsed in the public mood. \textsection
1 returns again to Statius (the freedman of Quintus, lately
manumitted, which Cicero cannot stop grieving over) and to
Clodius's threats, which Cicero still claims he can either
endure or escape. The Greek of \textsection 1 is Cicero's
self-judgment: ``perhaps I am blind and stare too much at the
seemly.''}

\textit{\textsection 2 is the celebrated theatre passage. At
the Apollinarian games of early July the tragedian Diphilus
seized the lines of his role to taunt Pompey to his face:
``by our misery you are great,'' ``a time will come when you
will groan heavily for that virtue of yours,'' the whole
theatre roaring along. Caesar entered to dead silence; young
Curio entered after him to ovations of the kind that used to
greet Pompey when the commonwealth was safe. Pompey writes
furious letters from Capua. The hostility extends to the
\textit{equites} who applauded Curio, and the triumvirs are
threatening to repeal the Roscian law (reserved seats for
knights at the games, 67 BC) and even the grain law. The
sense of \textsection 2 is that the Roman political world is
held now ``more by hatred than by force,'' and that ``the
populares have taught even modest men to hiss.''}

\textit{\textsection 4 returns to private peril. Clodius
threatens; Cicero counts his \textit{exercitus consularis}, the
army of every good man who saved the state in 63, as still
firm; Pompey assures him that not a word will be said against
him --- which Cicero takes as Pompey deceiving himself, not
deceiving Cicero. On Cosconius's death Cicero has been
invited to fill his vacancy in the agrarian commission of
twenty; he refuses, since to step into a dead triumvir's
place would brand him in front of the world. Caesar's offer
of a Gallic legateship reappears as the more honourable way
out. \textsection 5 closes with the cipher arrangement: in
future delicate letters Cicero will sign as Laelius, Atticus
as Furius, the rest \textit{in riddles}. Bibulus's edicts ---
the consul of 59 issuing his denunciations against the
triumvirs from his own house, the year-long retort that gave
the year its nickname ``the consulship of Julius and Caesar''
--- are burning Pompey's heart out. Note: section 3, present
in some editions, is here absent (as in Perseus); the
numbering jumps from 2 to 4.}
