% Ad Atticum 2.21 --- headnote
% ad-atticum-02-21, after 25 July 59 BC, written at Rome

\textit{Cicero to Atticus, written at Rome shortly after 25
July 59 BC. The letter dates itself: \textsection 3 records
the contio of \textit{a. d. viii Kal. Sextilis} (25 July) at
which Pompey, then at the height of public unpopularity, had
to defend himself against Bibulus's edicts. The piece is
Cicero's most extended portrait of Pompey's collapse from
glory: ``unaccustomed to ill repute, lifelong frequenter of
praise, swimming in glory, now disfigured in body, broken in
spirit, knows not whither to take himself.''}

\textit{\textsection 1--2 set the political picture: the
triumvirate's domination has shifted from a thing pleasing to
the multitude into a thing universally hated. Cicero's hope
had been that the wheel had turned silently and the storm
would pass without men acting; instead, after long secret
sighing, the city has begun to groan, then to speak, then to
shout. \textsection 3--4 give the famous portrait of Pompey
disfigured. The simile is the painter's: as Apelles would
grieve to see his Venus smeared with mud, or Protogenes his
Ialysus, so Cicero grieves to see Pompey, whom (as the
consul who had carried the Manilian law) Cicero had painted
and polished with all the colours of his art, suddenly
disfigured. Bibulus's edicts are described as ``Archilochian'' ---
that is, of the savagery of Archilochus's iambics --- so
pleasing to the public that the crowd in front of the posting
boards prevents passers-by from getting through. \textsection 5
records Caesar's failed contio against Bibulus (he could not
extort a sound from the crowd) and the elections postponed
into October. \textsection 6 closes on the personal threat:
Clodius is hostile, Pompey gives reassurances Cicero dares
not believe, Cicero is preparing to resist; the resources
he counts on are Varro and Pompey, and he ends with the now
constant refrain --- come, Atticus, come.}
