% Ad Atticum 2.22 --- headnote
% ad-atticum-02-22, Sextilis (August) 59 BC, written at Rome

\textit{Cicero to Atticus, written at Rome probably in
August 59 BC. The letter is the first in the surviving
correspondence to take Clodius's coming tribunate (he is to
enter office on 10 December) as the imminent threat that
will dictate Cicero's political conduct. \textsection 1 is the
character sketch: ``He flits about, he rages; he has nothing
certain, he denounces many; what chance shall offer, that he
seems about to do.'' Clodius oscillates between attacking the
triumvirs (whose unpopularity he sees) and attacking Cicero
(when he recalls the triumvirs' resources). \textsection 2
records Pompey's intervention: Pompey has reproached Clodius
to his face, telling him he himself --- the man who had armed
Clodius by letting him pass to the plebs --- would be in the
highest infamy of perfidy if Clodius were now to attack
Cicero. Both Clodius and Appius (Clodius's elder brother) have
given Pompey their word; Clodius gave in at the end of the
interview. But Clodius has not stopped speaking bitterly
about Cicero in public.}

\textit{\textsection 3--4 turn from policy to person. Cicero's
reputation is rising daily: the house is thronged in the
morning salutations, men salute him in the streets, the
\textit{memoria consulatus} is renewed --- and the prospect of
the contest with Clodius now seems, ``sometimes,'' a thing
worth meeting. He needs Atticus: at his side, by Varro, by
Publius himself, by reading the truth in Pompey through
\textit{[Greek: Ox-Eyed Hera]} --- the in-joke for Pompey's
wife Mucia, or in some readings for Clodia, with the
Homeric epithet for Hera carrying the implication of marital
intrigue. \textsection 5--6 close the political account:
Pompey is sick of his own creation, the city's hatred is
universal, no foreseeable outcome but a bursting out somewhere.
\textsection 7 is the letter's coda: returned books, a new
friend.}
