% Ad Atticum 2.24 --- headnote
% ad-atticum-02-24, before 18 October 59 BC, written at Rome

\textit{Cicero to Atticus, written at Rome shortly before 18
October 59 BC. The letter is the earliest surviving narrative
of the \textbf{Vettius affair} --- the murky episode in which
the informer L. Vettius (Cicero's old witness against Catiline
in 62) was either set up or set himself up to denounce a
plot, allegedly by young aristocratic ``optimates'' under
Curio the younger, against Pompey's life. The affair fell in
August 59 BC; this is Cicero's report to Atticus from the
weeks afterwards.}

\textit{The narrative \textsection 2--3 is one of the
clearest political set pieces in Cicero's correspondence. The
sequence: (1) Vettius worms his way into the young Curio's
intimacy, brings him to the point of saying he intended to
attack Pompey with his slaves. (2) Curio reports to his
father, the father to Pompey. (3) The Senate hears Vettius;
he denies, then demands the public's pledge of safety; gets
shouted down; then accuses, naming Paulus, the young Brutus
(Q. Servilius Caepio, who within a year would change his name
to Brutus), Lentulus, with the consul Bibulus alleged to have
supplied the dagger through his scribe Septimius. The Bibulus
detail collapses: Bibulus had on 13 May warned Pompey to
beware ambushes, and Pompey had thanked him. (4) Curio's
counter-evidence reveals that Paulus, named as the chief, was
in Macedonia at the relevant time. The Senate orders Vettius
into chains. (5) The next day Caesar takes Vettius onto the
Rostra in person and has him repeat the story with the names
revised: Caepio dropped (a night's entreaty has intervened),
Lucullus, Domitius, Cicero himself implied (``an eloquent
consular, the consul's neighbour, said to me Ahala or Brutus
must be found''), Piso (Cicero's son-in-law) and Laterensis
brought in at the end at Vatinius's prompt. The whole thing
turns out to be Caesar's: an attempt to discredit the
opposition by manufacturing a Pompey-assassination plot to be
laid at the optimates' door. Vettius was found dead in
prison shortly after, ending the matter in the way the
Catilinarian conspirators ended.}

\textit{\textsection 4 closes the political account: Vettius
on trial before Crassus Dives for violence; Cicero's life
\textit{taedet} (tedious to him); the contrast Catulus
\textit{fortunatus} (the great consular dead two years before,
spared this) against Cicero \textit{infortunatior}. \textsection
5 is the constant refrain: come, Atticus.}
