% Ad Atticum 2.25 --- headnote
% ad-atticum-02-25, October 59 BC, written at Rome

\textit{Cicero to Atticus, written at Rome around October 59
BC. The letter is the closing piece of \textit{Ad Atticum} 2,
and Cicero's last surviving letter before the Clodian crisis
breaks open. Two characters dominate \textsection 1: Varro,
whom Cicero is courting (see the contemporary \textit{Att.}
2.21--2.24, where Varro recurs as one of the bulwarks Cicero
is counting on for the contest with Clodius), and Hortensius
(``Hortalus''), Cicero's ancient rival at the bar and now
co-defender, with the report that Hortensius in the
\textit{Pro Flacco} (mid-59 BC, see the preceding entry in
this volume) had lifted Cicero's praises to the stars when he
spoke of Flaccus's praetorship and ``that time of the
Allobroges'' --- Hortensius covering the whole 5 December
narrative as a way of defending the consular against the
implicit charge of the trial. Two Greek tags: \textit{[Greek:
helikta kai oudhen hugies]}, ``tortuous and not at all sound,''
of Varro; and \textit{[Greek: tas t\=on krat\=ount\=on
\=eth\=e ph\=erein]}, ``to bear with the manners of the
powerful,'' Cicero's working maxim of the year.}

\textit{\textsection 2 closes the book with the standard
appeal: Atticus is on his way; come and free me of trouble or
share it. Of the commonwealth: nothing more desperate; of
those who hold it: nothing in greater hatred. The next
surviving letters --- if the editorial dating is right ---
are the exile letters of 58 BC, opening with \textit{Att.}
3.1 from Cicero in flight or already in flight from Rome.}
