% Ad Atticum 1.19 --- headnote
% ad-atticum-01-19, Ides (15) March 60 BC, written at Rome

\textit{Cicero to Atticus, written at Rome on the Ides of
March 60 BC. The longest political bulletin yet sent to
Atticus and the most candid of Cicero's accounts of the
post-consular settlement. Three pieces:}

\textit{(\textsection 2--5) the Gallic war scare. The Aedui
have been beaten, the Helvetii are in arms, the senate has
voted that the consuls (Metellus Celer and Afranius) draw
lots for the two Gauls; embassies are being sent under
Q. Metellus Creticus, L. Flaccus, and ``Lentulus, son of
Clodianus'' (the joke of the third name lay in his lack of
gravity). Cicero notes that, when the consulars drew lots,
his own came out first and the senate by acclamation kept
him in the city; the same happened to Pompey next, ``so
that we two seemed to be retained as pledges of the
commonwealth'' --- the closest the corpus comes to Cicero
naming his own dignity in print without irony. Flavius's
agrarian bill (Pompey's: to settle Pompey's veterans) is
under negotiation; Cicero has worked it from the inside,
extracting Sullan and Volaterran and Arretine private
holdings, and reshaping the funding so the land would be
bought with future revenues.}

\textit{(\textsection 6--8) the new strategy. After the
Bona Dea trial, the alienation of the equestrians, and the
malice of the fish-pond men, Cicero ``judged that I must
seek some greater resources and firmer defences,'' and has
brought Pompey to praise him publicly in the senate, ``the
safety of this empire and of the world.'' The closing tag
of the paragraph is the famous Epicharmus quotation in
Greek that runs through the corpus from this point on:
\textit{naphe kai memnas' apistein --- arthra tauta tan
phren\=on}, ``stay sober, and remember to mistrust: these
are the joints of the mind.'' Cicero's two poles in the
1860s --- the new alliance with Pompey, the inward
philosophical caution against it --- meet in the same
sentence.}

\textit{The closing paragraphs return to private business:
the Sicyonian creditors (Atticus's chronic concern), the
Greek and Latin versions of the consular memoir going off,
the third poem (the lost \textit{Consulatus Suus}, three
books in Latin verse on Cicero's own consulship; only the
fragment ``O fortunatam natam me consule Romam'' has come
down to us), and the Quintus--Pomponia quarrel which
Cicero pushes once again to a face-to-face meeting.}
