% Ad Atticum 1.13 --- headnote
% ad-atticum-01-13, 27 January 61 BC, written at Rome

\textit{Cicero to Atticus, written at Rome on 27 January 61
BC --- six days before the Kalends of February, in the
consulship of Marcus Messalla and Marcus Piso, as the
closing line dates the letter exactly. The first long
``newsletter'' of the corpus and one of the central
political documents in Cicero's hand. Atticus is now in
Epirus, on his way to settle business with Cicero's old
colleague Antonius in Macedonia; Cicero answers three
letters at once but is wary of who carries replies (``how
few who can carry a letter of any weight without lightening
it by reading it through''), and the substantive paragraphs
are written under the constant awareness that they could
be intercepted.}

\textit{The body of the letter is the new political map.
\textsection 2 is Cicero's new place: he is no longer the
first asked for his opinion (the new consul Pupius Piso has
shifted the order of consulars), and he claims to be relieved.
The four leaders of senatorial opinion now run: the
``pacificator of the Allobroges'' (Piso himself), Cicero,
Catulus, Hortensius. \textsection 3 is the Bona Dea affair
in compressed narrative: P. Clodius Pulcher in early December
62 entered Caesar's house in women's dress during the
all-female rite \textit{pro populo}, was caught, the Vestals
repeated the sacrifice, the senate referred the matter, the
pontiffs and Vestals declared it sacrilege, the consuls
promulgated a special bill, Caesar divorced his wife
Pompeia. The political battle is now whether the bill
passes, and Piso the consul (Clodius's friend) is sabotaging
his own bill from the inside. Cicero closes the paragraph
with the foreboding sentence that runs through every later
crisis of his life: ``I fear that this matter, neglected by
good men and defended by the wicked, may be the cause of
great evils to the commonwealth.''}

\textit{\textsection 4 is the famous portrait of Pompey,
just back from the East, addressed obliquely (``you know
whom I mean''): ``nothing courteous, nothing simple, nothing
in public affairs distinguished, nothing honourable, nothing
brave, nothing free.'' The closing paragraph mixes literary
shop --- Atticus has asked for a topographical sketch of
Misenum and Puteoli for the speech Cicero is writing, has
praised the speeches Cicero had sent (``now they seem much
more Attic to me'') --- with Roman house-price gossip:
Messalla the consul has bought Autronius's old house for
some thirteen million sesterces, the second great recent
purchase that vindicates Cicero's own (Pal.) HS 3,500,000
the year before. ``Expect a freer letter from me'' is the
sign-off: this one is already too dangerous.}
