% Ad Atticum 1.17 --- headnote
% ad-atticum-01-17, 5 December 61 BC, written at Rome

\textit{Cicero to Atticus, written at Rome on the Nones (5)
of December 61 BC. The first long letter of the Pomponia
quarrel: Atticus, on his way back from Epirus, has sent
Cicero copies of letters from his brother Quintus, the
proconsul of Asia, complaining about Pomponia (Atticus's
sister) --- letters that show ``a great variety of feeling
and an unlikeness of opinion and judgment.'' Quintus and
Pomponia's marriage, brokered by Cicero seven years before,
is fraying for the first time on the record. Cicero answers
in the central register of the corpus's domestic dimension
(\textsection 1--7): he has long known something was wrong;
he had hoped Atticus would meet Quintus on the road and
mend it in person but the meeting did not happen; he asks
Atticus to be the soft party and to receive Quintus's
``mildness of nature ready to take and to lay down offence''
in good faith.}

\textit{The second half of the letter (\textsection 8--11)
is the political news. The senate has just promulgated, by
its own decree, an inquiry against equestrian judges who
took bribes in the Clodius trial; the equestrian order is
incensed, threatening to break with the senate. Cicero has
opposed the inquiry from the floor and saved the day. Then
the publicans who had taken the Asian tax-farm at too high
a price (Crassus pushing) ask the senate to release them
from the contract; Cicero, swallowing his judgment of the
demand (``an invidious matter, a base demand, a confession
of rashness''), throws his weight behind them too, lest the
equestrian order be lost to the senate altogether. The
matter awaits a final speech from Cato, ``that hero of
ours,'' who is about to make the speech that, in fact, will
sink the publicans' suit and break Cicero's policy. The
closing sentence about Pompey is the new strategic axis:
``I have a most familiar use of Pompey. I see what you
would say. I shall guard against what is to be guarded
against, and shall write to you on another occasion more
about my counsels.'' The First Triumvirate is forming
underground; Cicero's ``concord of the orders'' is its
political opponent.}
