% Ad Atticum 1.14 --- headnote
% ad-atticum-01-14, Ides (13) February 61 BC, written at Rome

\textit{Cicero to Atticus, written at Rome on the Ides of
February (13 February) 61 BC. The second long Bona Dea
letter and the most vivid surviving narrative of a senate
day in Cicero's hand. The contest is over the special
bill against Clodius's sacrilege, which Piso the consul
(its sponsor by senatorial decree) is now sabotaging. The
letter walks through the day in order: Pompey's first
public meeting, said to have fallen flat (``not pleasing
to the wretched, empty to the wicked, not welcome to the
rich, not weighty to the good''); the contio in the Circus
Flaminius where Piso pushed Pompey into approving the
trial procedure under the praetor; the senate session
where Pompey replied vaguely about the senate's authority,
and Crassus then rose and made a great speech of praise
for Cicero's consulship --- ``that he was a senator, a
citizen, free, alive, he counted up to me; as often as he
saw his wife, his house, his country, so often did he see
my service.'' Cicero, sitting next to Pompey, registers
that Pompey was moved (whether by Crassus's gallantry or
by the obvious popularity of Cicero's consulship in the
house). Cicero himself then took the floor and ``showed
off'' before his new auditor with a performance of his
favourite consular set-piece --- ``the gravity of the
order, the concord of the equestrians, the consensus of
Italy, the dying remnants of the conspiracy, cheap grain,
peace.''}

\textit{The fifth paragraph is the famous account of the
contio at which the bill was withdrawn: Clodius's gangs
holding the bridges, the voting tablets distributed
without ``As you propose,'' Cato vaulting onto the Rostra
to upbraid Piso (``a voice full of gravity, full of
authority, full at last of safety''), Hortensius and
Favonius coming up beside him. The senate was then
summoned and voted four hundred to fifteen against the
bill in its sabotaged form. Clodius retreated to public
meetings of insult, where he attacked Lucullus, Hortensius,
Gaius Piso, the consul Messalla; Cicero only ``with having
ascertained everything,'' the running joke of the
\textit{comperi} verb that haunts the corpus through these
years. Greek tags, mostly drawn from rhetorical theory,
sprinkle the body --- Cicero's stylish bilingualism with
the friend who had just lived a decade in Athens; the
self-mocking \textit{eneperpereusam\=en} (``how I showed
off'') is the letter's affectionate signature.}
